### 2023-04-27 chapter1.tex
Base project setup

\documentclass[../main.tex]{subfiles}
\graphicspath{{\subfix{../images/}}}

\begin{document}
\title{The model for observing infections}
\author{}
\date{}

\begin{document}
\maketitle

\section{Introduction}\label{introduction}

This paper discusses a model for observing duration of PCR positivity in
the Coronavirus (COVID-19) Infection Survey (CIS). The CIS continually
enrols individuals and tests them on an ongoing basis, following a
specified testing protocol. The protocol specifies that the initial five
tests are spaced every 7 days, then the spacing is every 28 days.
However, real-world considerations often lead to variations in test
times and missed tests. Our analysis focuses solely on the binary result
(positive or negative) of these tests, assuming no misclassification
bias

Two major challenges arise when dealing with this data: double interval
censoring and left truncation. Double interval censoring arises because
we can only bound the start and end of an infection by observing an
individual's test results transitioning from negative to positive and
vice versa. Left truncation occurs because an individual is infected at
time \(t\) is only observed if their infection duration is at least
\(t^N - t\) days long, where \(t^N\) is the time of their next test.

To address these challenges, we employ two distinct approaches: the
``individual'' model and the ``total'' model.

The individual model considers only individuals with observed (i.e., not
truncated) episodes. This approach is similar to methods used in
previous studies, such as \protect\hyperlink{ref-heiseyModelling}{Heisey
and Nordheim} (\protect\hyperlink{ref-heiseyModelling}{1995}),
\protect\hyperlink{ref-dempsterMaximum}{Dempster, Laird, and Rubin}
(\protect\hyperlink{ref-dempsterMaximum}{1977}), and
\protect\hyperlink{ref-turnbullEmpirical}{Turnbull}
(\protect\hyperlink{ref-turnbullEmpirical}{1976}). The conceptual
approach here is to form a cohort of individuals for which we observe at
least one positive test, and then consider a study which had only
enrolled these individuals. The modelling approach then corrects for
this selection bias.

In contrast, the total model also takes into account data from
individuals without any observed episodes. These individuals may have
truncated episodes or may not have been infected at all. This approach
utilises their test times to estimate the probability of truncated
infections, ultimately providing an interpretable estimate of the number
of truncated episodes in the cohort.

Unlike most Bayesian methods (eg:
\protect\hyperlink{ref-heBayesiana}{He}
(\protect\hyperlink{ref-heBayesiana}{2003}),
\protect\hyperlink{ref-heBayesian}{He and Sun}
(\protect\hyperlink{ref-heBayesian}{2005}), and
\protect\hyperlink{ref-caoModeling}{Cao et al.}
(\protect\hyperlink{ref-caoModeling}{2009})) we avoid augmenting the
data with unobserved times of infection or any information on the
truncated infections; this is due to the large number of infections
(\textgreater4000) making this approach computationally expensive.

\hypertarget{data}{%
\section{Data}\label{data}}

Episodes are taken from the data extract of the 3rd Jan 2023, as
computed by SARAH using her updated episode definition. Only tests taken
as part of CIS are included (ie: no pillar 2 testing). The following
criteria are applied.

\begin{itemize}
\tightlist
\item
  The first positive in the episode occurred between 16th Oct 2020 and
  5th Dec 2020 inclusive.
\item
  The individual had recorded a negative test prior to their episode
  beginning (this could have happened at any time, including prior to
  16th Oct).
\item
  A negative test following the final positive in the episode has been
  recorded.
\end{itemize}

\hypertarget{notation-and-assumptions}{%
\section{Notation and assumptions}\label{notation-and-assumptions}}

\begin{itemize}
\tightlist
\item
  We initially assume perfect testing (no misclassification), although
  later relax this assumption.
\item
  Time is discrete: \(1, \dots, T\).
\item
  We observe \(n_a\) episodes of infection, indexed by \(i\).
\item
  The \(i\) in which episode \(i\) occurs is tested at times
  \(t_i = \{ t_{i,1}, \dots, t_{i,m_i} \}\).
\item
  Each observed episode has an unknown start of infection time \(b_i\)
  and end of infection time \(e_i\) (the first and last day that the
  individual would test positive due to this episode respectively).
  These are realisations of the random variables \(B_i\) and \(E_i\)
  respectively.
\item
  The duration of the infection is the number of days for which an
  individual tests positive, \(D_i = E_i - B_i + 1\).
\item
  We assume that, for all episodes \(i\), \(D_i\) is iid and independent
  of the time of the infection. Define the survival function
  \(Pr(D_i \geq t \mid B_i = b, \theta) = Pr(D_i \geq t \mid \theta) = S_\theta(t)\),
  where \(\theta\) are the parameters controlling the survival
  distribution (the discussion in this document is valid regardless of
  the model specified for \(S_\theta\) and hence we consider \(\theta\)
  as an arbitrary vector of parameters).
\item
  The beginning of episode \(i\) is known to occur in the interval
  \([l_i^{(b)}, r_i^{(b)}]\), and similarly for the end of the infection
  in \([l_i^{(e)}, r_i^{(e)}]\).
\item
  \(y_i(t)\) for \(t \in t_i\) is a binary indicator giving the test
  result for individual \(i\) at time \(t\). Under the perfect testing
  assumption, \(y_i(t) = 1\) if and only if \(b_i \leq t \leq e_i\).
\item
  Throughout, the convention that lower-case letters are realisations of
  upper-case random variables is used.
\end{itemize}

\hypertarget{individual-model}{%
\section{Individual model}\label{individual-model}}

\hypertarget{likelihood}{%
\subsection{Likelihood}\label{likelihood}}

Episode \(i\) can be fully characterised by the pair \((b_i, e_i)\) of
start and end dates of the episode, which belong to the state space
\(T \times T\). If the individual in which the episode occurs is tested
between \(b_i\) and \(e_i\) (inclusive), then we observe the episode,
otherwise it is truncated. For the \(i\)th episode that we observed, we
consider there is an unknown number, \(n_{it}\), where \((b, e)\) are
drawn from the same distribution as those that lead to episode \(i\),
and occur in an (imagined) individual identical to \(i\), but were
truncated. Episodes are also truncated if there is no negative test
prior to their start, that is, the individual the episode occurred in
had no tests before \(b_i\). The truncated episodes are often referred
to as the ``ghosts'' of \(i\).

Formally (and following \protect\hyperlink{ref-heiseyModelling}{Heisey
and Nordheim} (\protect\hyperlink{ref-heiseyModelling}{1995})), we
define three sets which partition the space of possible infection and
recovery (ie: are strict subsets of \(T \times T\)) given the
observations associated with episode \(i\) (graphically shown in Figure
\ref{fig:partitionSpace}). Each episode, whether observed or not, must
fall into one of these three classes with a probability that is a
function of \(\theta\) (see section \ref{sec:deriv-probs}).

\begin{itemize}
\tightlist
\item
  Admissible episodes, \(\alpha_i\), which have an infection and
  recovery time within their respective intervals observed.
  \(n_{ia} =1\) of these occurred, each with probability
  \(p_{ia} = Pr((b, e) \in \alpha_i \mid \theta)\).
\item
  Truncated episodes, \(\Omega_i^C\), which have an infection and
  recovery time such that they would not have tested positive. An
  unknown number, \(n_{it}\) of these observed, each with probability
  \(p_{it} = Pr((b, e) \in \Omega^C_i \mid \theta)\).
\item
  Inadmissible episodes, \(\beta_i\), which do not have an infection and
  recovery time within their respective intervals but would have been
  observed (not truncated). This is all remaining episodes not in the
  previous sets. \(n_{iu} = 0\) of these occurred, where they would have
  occurred with probability
  \(p_{iu} = Pr((b, e) \in \beta_i \mid \theta)\).
\end{itemize}

The untruncated region (where we could have observed an episode) is
\(\Omega_i = \alpha_i \cup \beta_i\) (the notation here is used for
consistency with \protect\hyperlink{ref-heiseyModelling}{Heisey and
Nordheim} (\protect\hyperlink{ref-heiseyModelling}{1995})), and is the
complement of \(\Omega^C_i\).

\begin{figure}
\centering
\includegraphics{20230216_123954.jpg}
\caption{Admissible, \(\alpha_i\), inadmissible, \(\beta_i\) and
untruncated, \(\Omega_i^C\) regions for an episode \(i\) in an
individual whose first test was at time 0 and negative, with subsequent
negative tests at times 7, 14, and 56, and positive tests at times 21
and 28. Purple region is \(\alpha_i\), orange region (which includes all
negative infection times) is \(\Omega_i^C\) and unshaded region is
\(\beta_i\).\label{fig:partitionSpace}}
\end{figure}

These three classes span all possible episodes and are mutually
exclusive, and hence the total number of episodes,
\(n_i = n_{ia} + n_{iu} + n_{it}\), and
\(p_{ia} + p_{iu} + p_{it} = 1\).

Episode \(i\) and its ghosts independently occur belong to one of these
classes. Therefore, conditional on \(n_i\), the number of episodes in
each class (that is, the counts \(n_{ia}, n_{iu}, n_{it}\)) are
distributed multinomially. Hence: \[
\begin{aligned}
&p(n_{ia} = 1, n_{iu} = 0, n_{it} = n_i - 1 \mid n_i, \theta) \\
&= \frac{n_i!}{n_{ia}! n_{iu}! (n_i- n_{ia} - n_{it})!} p_{ia}^{n_{ia}} p_{ua}^{n_{ia}} p_{it}^{n_i- n_{ia} - n_{it}} \\
&= \frac{n_i!}{(n_i-1)!} p_{ia} p_{it}^{n_i- 1} &\text{as $n_{ia} = 1$ and $n_{iu} = 0$}\\
&= n_i p_{ia} p_{it}^{n_i- 1}
\end{aligned}
\]

The posterior we are interested in is
\(p(\theta \mid n_{ia} = 1, n_{iu} = 0, \dots, n_{n_a,a} = 1, n_{n_a,u} = 0)\),
where \(\theta\) are the parameters governing the survival distribution
and hence allow derivation of the \(p\)s. \[
\begin{aligned}
&p(\theta \mid n_{ia} = 1, n_{iu} = 0, \dots, n_{n_a,a} = 1, n_{n_a,u} = 0) \\
&\propto p(\theta) \prod_{i=1}^{n_a} p(n_{ia} = 1, n_{iu} = 0 \mid \theta) \\
&= p(\theta) \prod_{i=1}^{n_a} \sum_{n_i=1}^\infty p(n_{ia} = 1, n_{iu} = 0, n_{it} = n_i - 1 \mid \theta, n_i) p(n_i \mid \theta) \\
&= p(\theta) \prod_{i=1}^{n_a} \sum_{n_i=1}^\infty n_i p_{ia} p_{it}^{n_i- 1} p(n_i) \\
&= p(\theta) \prod_{i=1}^{n_a} p_{ia} \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i)
\end{aligned}
\] where the last line follows by assuming prior independence.

In the following sections, we derive the expressions for \(p_{ia}\) and
\(p_{it}\) in terms of the data, and analytical solutions for the sum
under different priors.

\hypertarget{solution-for-the-sum}{%
\subsubsection{Solution for the sum}\label{solution-for-the-sum}}

This section is concerned with finding analytical solutions to
\(\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i)\) under two different
assumptions for \(p(n_i)\). First, we consider the improper prior
\(p(n_i) \propto 1/n_i\) which has several convenient properties. Then,
we consider a negative binomial distribution (including the geometric
and Poisson distributions as a special and limiting case respectively).

The prior \(p(n_i) \propto 1/n_i\) means that the model coincides with
the frequentist likelihood-based approach, up to the priors on other
parameters (\protect\hyperlink{ref-dempsterMaximum}{Dempster, Laird, and
Rubin} (\protect\hyperlink{ref-dempsterMaximum}{1977}) section 4.2;
\protect\hyperlink{ref-heiseyModelling}{Heisey and Nordheim}
(\protect\hyperlink{ref-heiseyModelling}{1995});
\protect\hyperlink{ref-gelmanBayesian}{Gelman et al.}
(\protect\hyperlink{ref-gelmanBayesian}{2013}) section 8.7.5), and is
the reference prior for this quantity
(\protect\hyperlink{ref-heBayesiana}{He}
(\protect\hyperlink{ref-heBayesiana}{2003})). Under this prior,
\(\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) \propto \sum_{n_i=1}^\infty p_{it}^{n_i-1} = 1/(1-p_{it})\).

Now consider \(N_i \sim \text{NegBin}(\mu, r)\) using the
mean/dispersion parameterisation of the negative binomial. Equivalently:
\[
\begin{aligned}
N_i \mid \lambda &\sim \text{Poisson}(\lambda) \\
\lambda &\sim \text{Gamma}(a, b)
\end{aligned}
\] where \(b = r / \mu\) and \(a = r\). Hence: \[
\begin{aligned}
\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) 
&= \int \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i \mid \lambda) p(\lambda) d\lambda \\
&= \int \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} \frac{\lambda^{n_i} e^{-\lambda}}{n_i!} p(\lambda) d\lambda \\
&= \int e^{-\lambda} \sum_{n_i=1}^\infty p_{it}^{n_i- 1} \frac{\lambda^{n_i-1}\lambda }{(n_i-1)!} p(\lambda) d\lambda \\
&= \int e^{-\lambda} \lambda p(\lambda) \sum_{n_t=0}^\infty p_{it}^{n_t} \frac{\lambda^{n_t} }{n_t!} d\lambda &n_t = n_i - 1 \\
&= \int e^{-\lambda} \lambda p(\lambda) e^{p_{it}\lambda} d\lambda &\text{Poisson pmf} \\
&= \int e^{\lambda (p_{it} - 1)} \lambda p(\lambda) d\lambda \\
&= \int e^{\lambda (p_{it} - 1)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda \\
&= \frac{\Gamma(a+1)b^a}{\Gamma(a) (b+1-p_{it})^{a+1}} \int \frac{(b+1-p_{it})^{a+1}}{\Gamma(a+1)} \lambda^{(a+1)-1} e^{-(b+1-p_{it})\lambda} d\lambda \\
&= \frac{\Gamma(a+1)b^a}{\Gamma(a) (b+1-p_{it})^{a+1}} &\text{as gamma pdf} \\
&\propto (b+1-p_{it})^{-(a+1)} \\
&= (r/\mu+1-p_{it})^{-(r+1)} \\
&\propto (r+\mu (1-p_{it}))^{-(r+1)}
\end{aligned}
\] Note that \(r=0\) recovers the previous derivation (up to
proportionality).

The above allows us to sample from the posterior without needing to
sample the \(n_i\)s. This is convenient because there are as many
\(n_i\)s as episodes observed and hence is computationally costly to
sample. Furthermore, each \(n_i\) is a discrete parameter, and hence
cannot be sampled within the current Stan implementation.

Having estimated the posterior of the other parameters, we might want to
consider the posterior of \(n_i\) (normally for diagnostic or model
debugging purposes), which we can sample from the full conditional.
Specifically, for each posterior sample of \(\theta\), we sample one
draw from \(n_i \mid \theta, y\) where \(y\) represents all the data.

Due to the conditional independence of the \(n_i\)s, we find that: \[
\begin{aligned}
p(n_i \mid \theta, n_{ia} = 1, n_{iu} = 0)
&\propto p(n_{ia} = 1, n_{iu} = 0, n_{it} = n_i - 1 \mid n_i, p_{iu}, p_{ia}, p_{it}) p(n_i) \\
&\propto n_i p_{it}^{n_i- 1} p(n_i)
\end{aligned}
\] With the prior, \(p(n_i) \propto 1/n_i\) then this is a geometric
distribution with parameter \(1 - p_{it}\). With a negative binomial
prior, we have: \[
\begin{aligned}
n_i p_{it}^{n_i- 1} p(n_i)
&\propto n_i p_{it}^{n_i- 1} \frac{\Gamma(r + n_i)}{n_i!} \left( \frac{\mu}{r+\mu} \right)^{n_i} \\
&\propto \frac{\Gamma((r + 1) + (n_i - 1))}{(n_i-1)!} \left( \frac{\mu p_{it}}{r+\mu} \right)^{n_i-1}
\end{aligned}
\] which is a negative binomial pmf with size parameter \(r+1\) and
probability parameter \(\frac{r + \mu (1 - p_{it})}{r+\mu}\). The mean
of this is \[
\frac{(r+1)\mu p_{it}}{r+\mu(1-p_{it})}
\]

\hypertarget{derivations-of-probabilities}{%
\subsubsection{\texorpdfstring{Derivations of probabilities
\label{sec:deriv-probs}}{Derivations of probabilities }}\label{derivations-of-probabilities}}

By definition, we have \(p_{ia} = Pr((b, e) \in \alpha_i)\) and
\(\alpha_i = \{ (b, e) : l_i^{(b)} \leq b \leq r_i^{(b)} \wedge l_i^{(e)} \leq e \leq r_i^{(e)}\}\).
Hence: \[
\begin{aligned}
p_{ia}
&= Pr \left( l_i^{(b)} \leq B_i \leq r_i^{(b)}, l_i^{(e)} \leq E_i \leq r_i^{(e)} \right) \\
&= Pr \left( l_i^{(e)} \leq E_i \leq r_i^{(e)} \mid l_i^{(b)} \leq B_i \leq r_i^{(b)} \right) Pr \left( l_i^{(b)} \leq B_i \leq r_i^{(b)} \right) \\
&=\sum_{b = l_i^{(l)}}^{r_i^{(b)}} Pr \left( l_i^{(e)} \leq E_i \leq r_i^{(e)} \mid B_i = b \right) Pr \left(B_i = b \right) \\
&=\sum_{b = l_i^{(b)}}^{r_i^{(b)}} Pr \left( l_i^{(e)} - b + 1 \leq D_i \leq r_i^{(e)} - b + 1 \right) Pr \left(B_i = b \right) \\
&=\sum_{b = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b + 1) - S_\theta(r_i^{(e)} - b + 2) \right) Pr \left(B_i = b \right) \\
&\propto \sum_{b = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b + 1) - S_\theta(r_i^{(e)} - b + 2) \right)
\end{aligned}
\] under the assumption of uniform probability of infection time.

Next, we derive \(p_{it} = Pr((b, e) \in \Omega^C_i)\), the probability
that episode \(i\) was truncated. A truncated episode means that no test
was performed during the episode or there was no negative test prior to
the episode. Denote by \(t_i\) the set of testing times for the
individual in which episode \(i\) was observed, and define the time
until the next test on the individual after time \(t'\) as
\(t^N_{it} = \min \{ t' \in t_i : t' \geq t \} - t\). Then
\(\Omega^C_i = \{ (b, e) : \nexists t \in t_i. b \leq t \leq e \vee b \leq \min(t_i) \} = \{ (b, e) : e - b < t^N_{ib} \vee b \leq \min(t_i) \}\).
Hence: \[
\begin{aligned}
1 - p_{it}
&= 1 - Pr(E_i - B_i < t_{iB_i}^N \vee B_i \leq \min(t_i)) \\
&= 1 - Pr(E_i - B_i < t_{iB_i}^N \wedge B_i > \min(t_i)) - Pr(B_i \leq \min(t_i)) \\
&= 1 - \sum_{b=\min(t_i) + 1}^T Pr(E_i - b + 1 < t_{ib}^N \mid B_i = b) Pr(B_i = b) - \sum_{t=1}^{\min(t_i)} Pr(B_i = b)\\
&= 1 - \frac{1}{T} \sum_{b=\min(t_i)+1}^T (1 - S_\theta(t_{ib}^N + 1)) - \frac{\min(t_i)}{T} \\
&= 1 - \frac{T-\min(t_i)}{T} + \frac{1}{T} \sum_{b=\min(t_i)+1}^T S_\theta(t_{ib}^N + 1)) - \frac{\min(t_i)}{T} \\
&= \frac{1}{T} \sum_{b=\min(t_i)+1}^T S_\theta(t_{ib}^N + 1)
\end{aligned}
\]

\hypertarget{full-posterior}{%
\subsection{Full posterior}\label{full-posterior}}

Under the prior \(p(n_i) \propto 1/n_i\): \[
\begin{aligned}
&p(\theta \mid n_{ia} = 1, n_{iu} = 0, \dots, n_{n_a,a} = 1, n_{n_a,u} = 0) \\
&\propto p(\theta) \prod_{i=1}^{n_a} p_{ia} \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) \\
&= p(\theta) \prod_{i=1}^{n_a} \frac{p_{ia}}{1-p_{it}} \\
&\propto p(\theta) \prod_{i=1}^{n_a} \frac{\sum_{b = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(r_i^{(e)} - b - 1) - S_\theta(l_i^{(e)} - b - 1) \right)}{\sum_{b=\min(t_i)}^T S_\theta(t_{ib}^N + 1)} \\
\end{aligned}
\]

Under the prior \(N_i \sim \text{NegBinom}(\mu, r)\): \[
\begin{aligned}
&p(\theta \mid n_{ia} = 1, n_{iu} = 0, \dots, n_{n_a,a} = 1, n_{n_a,u} = 0) \\
&\propto p(\theta) \prod_{i=1}^{n_a} p_{ia} \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) \\
&= p(\theta) \prod_{i=1}^{n_a} \frac{p_{ia}}{(r+\mu (1-p_{it}))^{(r+1)}} \\
&= p(\theta) \prod_{i=1}^{n_a} \frac{\sum_{b=l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(r_i^{(e)} - b - 1) - S_\theta(l_i^{(e)} - b - 1) \right)}{\left( r+\mu/T \left( \sum_{b=\min(t_i)}^T S_\theta(t_{ib}^N + 1) \right) \right)^{(r+1)}} \\
\end{aligned}
\]

\hypertarget{total-model}{%
\section{Total model}\label{total-model}}

We now consider all \(N\) individuals in the CIS cohort who have at
least one test during the period of interest, this includes those
without an observed infection, and assume that each is equally likely to
be infected. As before, \(i = 1, \dots, n_a\) indexes episodes but now
\(j = 1, \dots, N\) indexes individuals. All episodes have an associated
individual, each individual may have any number (including 0) episodes.
Admissible regions are only defined for episodes, and have not changed
from the previous section. The truncated region is defined for each
individual, in addition we consider the combined truncated region which
is episodes that would be truncated without conditioning on which
individual the episode occurs in:
\(\Omega^C = \{ (b, e, j) : (b, e) \in \Omega_j^C \}\).

Denote by \(n_\text{tot}\) the total number of episodes across all \(N\)
individuals, regardless of whether they are admissible or truncated (we
know there are no inadmissible episodes by definition).

For any infection, the probability of it being admissible for episode
\(i\) is \(\frac{1}{N} p_{ia}\). That is, the probability that the
infection occurs in the individual corresponding to the individual in
which episode \(i\) (\(1/N\) by the assumption that episodes are equally
likely to occur in any of the \(N\) individuals) multiplied by the
probability that the episode is admissible for episode \(i\) conditional
on it occurring in the relevant individual (\(p_{ia}\)).

The total number of truncated infections that occur is
\(n_t = n_\text{tot} - n_a\). Conditional on an infection occurring in
individual \(j\), its probability of being truncated is \(p_{jt}\) (as
defined previously). Therefore, the overall probability of an episode
being truncated is \(p_t = \frac{1}{N} \sum_{j=1}^N p_{jt}\).

Consider our data as there being one admissible episode for each
observed episode, and all other episodes being truncated. As in the
previous section, conditional on \(n_\text{tot}\), we have a multinomial
likelihood. \[
\begin{aligned}
p(\text{data} \mid \theta)
&= \sum_{n_\text{tot}=n_a}^\infty p(\text{data} \mid n_\text{tot}, \theta) p(n_\text{tot}) \\
&= \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} \left( \prod_{i=1}^{n_a} \frac{1}{N} p_{ia} \right) p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \\
&= \left( \prod_{i=1}^{n_a} \frac{1}{N} p_{ia} \right) \left( \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \right) \\
\end{aligned}
\]

In this case, it would be computationally feasible to augment the data
with \(n_\text{tot}\) directly (as it is only a single parameter).
However, implementation is easier if we can use a closed-form for the
sum (as in the previous section) as it allows the use of standard
implementations of NUTS such as Stan, which cannot handle discrete
parameters.

Assuming \(N_\text{tot} \sim \text{NegBin}(\mu, r)\):

\[
\begin{aligned}
\sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot})
&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot} \mid \lambda) p(\lambda) d\lambda \\
&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} \frac{\lambda^n_\text{tot} e^{-\lambda}}{n_\text{tot}!} p(\lambda) d\lambda \\
&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{1}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} \lambda^{n_\text{tot}-n_a} \lambda^{n_a} e^{-\lambda} p(\lambda) d\lambda \\
&= \int \lambda^{n_a} e^{-\lambda} p(\lambda) \sum_{n_t=0}^\infty \frac{1}{n_t!} p_t^{n_t} \lambda^{n_t} d\lambda &n_t = n-n_a\\
&= \int \lambda^{n_a} e^{-\lambda} p(\lambda) e^{\lambda p_t} d\lambda \\
&= \int \lambda^{n_a} e^{-\lambda(1 - p_t)} p(\lambda) d\lambda \\
&= \int \lambda^{n_a} e^{-\lambda(1 - p_t)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda \\
&= \int \frac{b^a}{\Gamma(a)} \lambda^{a+n_a-1} e^{-(b+1-p_t)\lambda} \lambda d\lambda \\
&= \frac{b^a}{\Gamma(a)} \frac{\Gamma(a+n_a)}{(b+1-p_t)^{a+n_a}} \\
&\propto (b+1-p_t)^{-(a+n_a)} \\
&= (r/\mu + 1 - p_t)^{-(r+n_a)} \\
&\propto(r + \mu (1- p_t))^{-(r+n_a)}
\end{aligned}
\]

Which gives the full posterior density as: \[
\begin{aligned}
p(\theta \mid \text{data})
&\propto \frac{p(\theta) \prod_{i=1}^{n_a} p_{ia}}{(r + \mu(1 - p_t))^{r+n_a}}
\end{aligned}
\]

As before, being able to reconstruct the posterior of \(n_\text{tot}\)
using its full conditional is useful. \[
\begin{aligned}
p(n_\text{tot} \mid \text{data}, \theta)
&\propto p(\text{data} \mid \theta, n_\text{tot}) p(n_\text{tot}) \\
&= \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} \left( \prod_{i=1}^{n_a} \frac{1}{N} p_{ia} \right) p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \\
&\propto \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}} \frac{\Gamma(r + n_\text{tot})}{n_\text{tot}!} \left( \frac{\mu}{r + \mu} \right)^{n_\text{tot}}  \\
&\propto \frac{\Gamma(r + n_\text{tot})}{(n_\text{tot}-n_a)!} \left( \frac{\mu p_t}{r + \mu} \right)^{n_\text{tot}}  \\
&\propto \frac{\Gamma((r + n_a) + (n_\text{tot}- n_a))}{(n_\text{tot}-n_a)!} \left( \frac{\mu p_t}{r + \mu} \right)^{n_\text{tot}-n_a}  \\
\end{aligned}
\] which is a negative binomial pmf with size parameter \(r + n_a\) and
probability parameter \(\frac{r+\mu(1-p_t)}{r+\mu}\).

\hypertarget{false-negatives}{%
\section{False negatives}\label{false-negatives}}

In this section, we relax the assumption of perfect testing to allow
some false negatives. We use a fairly simplified model to ensure the
likelihood is tractable. This requires modifying both \(p_{ia}\) to
allow for the episode possibly being longer than observed, and
\(p_{it}\) to allow for additional episodes being missed.

Specifically, we introduce a constant test sensitivity \(p_\text{sens}\)
such that: \[
Pr(y_i(t) = 1) = \begin{cases} p_\text{sens} &b_i \leq t \leq e_i \\ 0 &\text{otherwise.}\end{cases}
\]

In practice, it is likely that the test sensitivity is a function of the
viral load of an individual (the concentration of viral material in the
part of the body being sampled). For this initial attempt at modelling
the sensitivity, we ignore this complication.

\hypertarget{modifying-p_ia}{%
\subsection{\texorpdfstring{Modifying
\(p_{ia}\)}{Modifying p\_\{ia\}}}\label{modifying-p_ia}}

The idea when modifying \(p_{ia}\) is to allow for the negative test
following the last positive to be either a true or false negative. If it
is a true negative, then we have a bound on the length of the episode.
If it is a false negative, then we consider the episode's length
right-censored. The likelihood contribution for an episode is then a
mixture of these two scenarios, with the mixture probability determined
by the test sensitivity.

For tractability, we assume that the tests bounding the start of the
infection (\(l_i^{(b)}\) and \(r_i^{(b)}\)) are true results and hence
\(b_i \in [l_i^{(b)}, r_i^{(b)}]\). Furthermore, we consider the tests
only between \(r_i^{(b)}\) and \(l_i^{(e)}\) inclusive (the positive
tests providing a lower bound on the length of the episode). Denote
these tests by
\(t'_i = \{ t \in t_i : r_i^{(b)} \leq t \leq l_i^{(e)} \}\) and their
results by \(y_i'\). We know that the test results at times in \(t_i'\)
are either true positives or false negatives. By definition, the test at
\(r_i^{(e)}\) is a false negative if and only if \(e_i > r_i^{(e)}\). We
proceed by considering the two cases.

First, if \(e_i \leq r_i^{(e)}\). In this case, the test at
\(r_i^{(e)}\) is a true negative, as are all other tests not in
\(t_i'\), and these occur with probability 1. \[
\begin{aligned}
&p(y_i', e_i \leq r_i^{(e)} | b_i, p_\text{sens}, \theta) \\
&= p(y_i', l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) &\text{as no false pos} \\
&= p(y_i' \mid l_i^{(e)} \leq e_i \leq r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
&= \left( \prod_{t \in t_i'} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) \right)
\end{aligned}
\]

Second, if \(e_i > r_i^{(e)}\). In this case, the test test at
\(r_i^{(e)}\) is a false negative, occurring with probability
\((1 - p_\text{sens})\) and we have no upper bound on the end of the
infection. \[
\begin{aligned}
&p(y_i', e_i > r_i^{(e)} | b_i, p_\text{sens}, \theta) \\
&= p(y_i \mid e_i > r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(e_i > r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
&= \left( \prod_{t \in t_i'} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) (1 - p_\text{sens}) S_\theta(r_i^{(e)} - b_i - 1)
\end{aligned}
\]

Combining the above, the replacement for \(p_{ia}\) is : \[
\begin{aligned}
p_{ia}'
&= p(y_i' \mid p_\text{sens}, \theta) \\
&= \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( p(y_i', b_i, e_i \leq r_i^{(e)} \mid p_\text{sens}, \theta) p(y_i', b_i e_i > r_i^{(e)} \mid p_\text{sens}, \theta) \right) p(b_i \mid p_\text{sens}, \theta) \\
&= \left( \prod_{t \in t_i'} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \\ & \ \times \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) + (1 - p_\text{sens}) S_\theta(r_i^{(e)} - b_i - 1) \right) p(b_i \mid p_\text{sens}, \theta) \\
&= \left( \prod_{t \in t_i'} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right)\sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - p_\text{sens} S_\theta(r_i^{(e)} - b_i - 1) \right) p(b_i \mid p_\text{sens}, \theta).
\end{aligned}
\] Note that if \(p_\text{sens} = 1\) then \(p_{ia}' = p_{ia}\).

\hypertarget{modifying-p_it}{%
\subsection{\texorpdfstring{Modifying
\(p_{it}\)}{Modifying p\_\{it\}}}\label{modifying-p_it}}

With false negatives, an episode with can be truncated in the following
ways:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The episode begins before \(\min(t_i)\) (assuming there is a
  negligible probability of a false negative test at this point).
\item
  The duration of the episode is less than \(t_{ib_i}^N\). These first
  two mechanisms are how truncation previously occurred, with
  probability \(p_{it}\).
\item
  The duration of the episode is at least as long as \(t_{it}^N\) but
  less than the time to the second test after \(t\), \(t_{it}^{2N}\) ,
  and a false negative episode occurred at the first test after \(t\).
  This occurs with probably
  \((1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(t_i)}^T \left( S_\theta(t_{ib}^N + 1) - S_\theta(t_{ib}^{2N} + 1)\right)\).
\item
  The duration of the episode is at least as long as \(t_{it}^{2N}\) ,
  and all the tests within the episode (at least two) are false
  negatives. We assume this occurs with negligible probability.
\end{enumerate}

Hence the replacement for \(1 - p_{it}\) is: \[
\begin{aligned}
1 - p_{it}'
&= 1 - p_{it} - (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(t_i)}^T \left( S_\theta(t_{ib}^N + 1) - S_\theta(t_{ib}^{2N} + 1)\right) \\
&= \frac{1}{T} \sum_{b=\min(t_i)}^T \left( p_\text{sens} S_\theta(t_{ib}^N + 1) + (1 - p_\text{sens}) S_\theta(t_{ib}^{2N} + 1)\right) \\
\end{aligned}
\]

\end{document}

### 2023-04-27 main.tex
Base project setup

\documentclass{article}
\usepackage{graphicx}
\graphicspath{{images/}}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{textcomp} % provide euro and other symbols

\def\reals{\mathbb{R}}
\def\nats{\mathbb{N}}
\def\sampSpace{\mathcal{X}}
\def\dist{\sim}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\V}{\mathbb{V}}
\DeclareMathOperator{\prob}{\mathbb{P}}
\DeclareMathOperator{\var}{\mathbb{V}}
\DeclareMathOperator{\indicator}{\mathbb{I}}
\DeclareMathOperator{\cov}{Cov}
\DeclareMathOperator{\Ber}{Bernoulli}
\DeclareMathOperator{\Bin}{Binomial}
\DeclareMathOperator{\Poi}{Poisson}
\def\matr{\bm}
\def\set{\mathcal}
\renewcommand{\vec}{\bm}

\usepackage{subfiles} % Best loaded last in the preamble

\title{Estimating the incidence of SARS-CoV-2 from prevalence data}
\author{Joshua Blake}
\date{2023}

\begin{document}

\maketitle
abc
\section{Chapter 1}

\subfile{chapter1/chapter1}

\begin{align}
    \cov
\end{align}

\end{document}

### 2023-05-05 cis-perfect-testing.tex
Format TeX and working pipeline

\documentclass[main.tex]{subfiles}

\title{The model for observing infections}
\author{Joshua Blake}
\date{\today}

\begin{document}
\maketitle

\section{Introduction}\label{introduction}

This paper discusses a model for observing duration of PCR positivity in
the Coronavirus (COVID-19) Infection Survey (CIS). The CIS continually
enrols individuals and tests them on an ongoing basis, following a
specified testing protocol. The protocol specifies that the initial five
tests are spaced every 7 days, then the spacing is every 28 days.
However, real-world considerations often lead to variations in test
times and missed tests. Our analysis focuses solely on the binary result
(positive or negative) of these tests, assuming no misclassification
bias

Two major challenges arise when dealing with this data: double interval
censoring and left truncation. Double interval censoring arises because
we can only bound the start and end of an infection by observing an
individual's test results transitioning from negative to positive and
vice versa. Left truncation occurs because an individual is infected at
time $t$ is only observed if their infection duration is at least
$t^N - t$ days long, where $t^N$ is the time of their next test.

To address these challenges, we employ two distinct approaches: the
``individual'' model and the ``total'' model.

The individual model considers only individuals with observed (i.e., not
truncated) episodes. This approach is similar to methods used in
previous studies, such as \textcite{heiseyModelling},
\textcite{dempsterMaximum}, and
\textcite{turnbullEmpirical}. The conceptual
approach here is to form a cohort of individuals for which we observe at
least one positive test, and then consider a study which had only
enrolled these individuals. The modelling approach then corrects for
this selection bias.

In contrast, the total model also takes into account data from
individuals without any observed episodes. These individuals may have
truncated episodes or may not have been infected at all. This approach
utilises their test times to estimate the probability of truncated
infections, ultimately providing an interpretable estimate of the number
of truncated episodes in the cohort.

Unlike most Bayesian methods (eg: \textcite{heBayesiana}, \textcite{heBayesian}, and \textcite{caoModeling}) we avoid augmenting the
data with unobserved times of infection or any information on the
truncated infections; this is due to the large number of infections
(>4000) making this approach computationally expensive.

\section{Data}\label{data}

Episodes are taken from the data extract of the 3rd Jan 2023, as
computed by SARAH using her updated episode definition. Only tests taken
as part of CIS are included (ie: no pillar 2 testing). The following
criteria are applied.

\begin{itemize}
\item
  The first positive in the episode occurred between 16th Oct 2020 and
  5th Dec 2020 inclusive.
\item
  The individual had recorded a negative test prior to their episode
  beginning (this could have happened at any time, including prior to
  16th Oct).
\item
  A negative test following the final positive in the episode has been
  recorded.
\end{itemize}

\section{Notation and assumptions}\label{notation-and-assumptions}

\begin{itemize}
\item
  We initially assume perfect testing (no misclassification), although
  later relax this assumption.
\item
  Time is discrete: $1, \dots, T$.
\item
  We observe $n_a$ episodes of infection, indexed by $i$.
\item
  The $i$ in which episode $i$ occurs is tested at times
  $t_i = \{ t_{i,1}, \dots, t_{i,m_i} \}$.
\item
  Each observed episode has an unknown start of infection time $b_i$
  and end of infection time $e_i$ (the first and last day that the
  individual would test positive due to this episode respectively).
  These are realisations of the random variables $B_i$ and $E_i$
  respectively.
\item
  The duration of the infection is the number of days for which an
  individual tests positive, $D_i = E_i - B_i + 1$.
\item
  We assume that, for all episodes $i$, $D_i$ is iid and independent
  of the time of the infection. Define the survival function
  $\prob(D_i \geq t \mid B_i = b, \theta) = \prob(D_i \geq t \mid \theta) = S_\theta(t)$,
  where $\theta$ are the parameters controlling the survival
  distribution (the discussion in this document is valid regardless of
  the model specified for $S_\theta$ and hence we consider $\theta$
  as an arbitrary vector of parameters).
\item
  The beginning of episode $i$ is known to occur in the interval
  $[l_i^{(b)}, r_i^{(b)}]$, and similarly for the end of the infection
  in $[l_i^{(e)}, r_i^{(e)}]$.
\item
  $y_i(t)$ for $t \in t_i$ is a binary indicator giving the test
  result for individual $i$ at time $t$. Under the perfect testing
  assumption, $y_i(t) = 1$ if and only if $b_i \leq t \leq e_i$.
\item
  Throughout, the convention that lower-case letters are realisations of
  upper-case random variables is used.
\end{itemize}

\section{Individual model}\label{individual-model}

\subsection{Likelihood}\label{likelihood}

Episode $i$ can be fully characterised by the pair $(b_i, e_i)$ of
start and end dates of the episode, which belong to the state space
$T \times T$. If the individual in which the episode occurs is tested
between $b_i$ and $e_i$ (inclusive), then we observe the episode,
otherwise it is truncated. For the $i$th episode that we observed, we
consider there is an unknown number, $n_{it}$, where $(b, e)$ are
drawn from the same distribution as those that lead to episode $i$,
and occur in an (imagined) individual identical to $i$, but were
truncated. Episodes are also truncated if there is no negative test
prior to their start, that is, the individual the episode occurred in
had no tests before $b_i$. The truncated episodes are often referred
to as the ``ghosts'' of $i$.

Formally (and following \textcite{heiseyModelling}), we
define three sets which partition the space of possible infection and
recovery (ie: are strict subsets of $T \times T$) given the
observations associated with episode $i$ (graphically shown in Figure
\ref{fig:partitionSpace}). Each episode, whether observed or not, must
fall into one of these three classes with a probability that is a
function of $\theta$ (see section \ref{sec:deriv-probs}).

\begin{itemize}
\item
  Admissible episodes, $\alpha_i$, which have an infection and
  recovery time within their respective intervals observed.
  $n_{ia} =1$ of these occurred, each with probability
  $p_{ia} = \prob((b, e) \in \alpha_i \mid \theta)$.
\item
  Truncated episodes, $\Omega_i^C$, which have an infection and
  recovery time such that they would not have tested positive. An
  unknown number, $n_{it}$ of these observed, each with probability
  $p_{it} = \prob((b, e) \in \Omega^C_i \mid \theta)$.
\item
  Inadmissible episodes, $\beta_i$, which do not have an infection and
  recovery time within their respective intervals but would have been
  observed (not truncated). This is all remaining episodes not in the
  previous sets. $n_{iu} = 0$ of these occurred, where they would have
  occurred with probability
  $p_{iu} = \prob((b, e) \in \beta_i \mid \theta)$.
\end{itemize}

The untruncated region (where we could have observed an episode) is
$\Omega_i = \alpha_i \cup \beta_i$ (the notation here is used for
consistency with \textcite{heiseyModelling}), and is the
complement of $\Omega^C_i$.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{cis-perfect-testing/images/likelihood_areas.jpg}
\caption{Admissible, $\alpha_i$, inadmissible, $\beta_i$ and
untruncated, $\Omega_i^C$ regions for an episode $i$ in an
individual whose first test was at time 0 and negative, with subsequent
negative tests at times 7, 14, and 56, and positive tests at times 21
and 28. Purple region is $\alpha_i$, orange region (which includes all
negative infection times) is $\Omega_i^C$ and unshaded region is
$\beta_i$.\label{fig:partitionSpace}}
\end{figure}

These three classes span all possible episodes and are mutually
exclusive, and hence the total number of episodes,
$n_i = n_{ia} + n_{iu} + n_{it}$, and
$p_{ia} + p_{iu} + p_{it} = 1$.

Episode $i$ and its ghosts independently occur belong to one of these
classes. Therefore, conditional on $n_i$, the number of episodes in
each class (that is, the counts $n_{ia}, n_{iu}, n_{it}$) are
distributed multinomially. Hence:
\begin{align}
&p(n_{ia} = 1, n_{iu} = 0, n_{it} = n_i - 1 \mid n_i, \theta) \\
&= \frac{n_i!}{n_{ia}! n_{iu}! (n_i- n_{ia} - n_{it})!} p_{ia}^{n_{ia}} p_{ua}^{n_{ia}} p_{it}^{n_i- n_{ia} - n_{it}} \\
&= \frac{n_i!}{(n_i-1)!} p_{ia} p_{it}^{n_i- 1} &\text{as $n_{ia} = 1$ and $n_{iu} = 0$}\\
&= n_i p_{ia} p_{it}^{n_i- 1}
\end{align}

The posterior we are interested in is
$p(\theta \mid n_{ia} = 1, n_{iu} = 0, \dots, n_{n_a,a} = 1, n_{n_a,u} = 0)$,
where $\theta$ are the parameters governing the survival distribution
and hence allow derivation of the $p$s.
\begin{align}
&p(\theta \mid n_{ia} = 1, n_{iu} = 0, \dots, n_{n_a,a} = 1, n_{n_a,u} = 0) \\
&\propto p(\theta) \prod_{i=1}^{n_a} p(n_{ia} = 1, n_{iu} = 0 \mid \theta) \\
&= p(\theta) \prod_{i=1}^{n_a} \sum_{n_i=1}^\infty p(n_{ia} = 1, n_{iu} = 0, n_{it} = n_i - 1 \mid \theta, n_i) p(n_i \mid \theta) \\
&= p(\theta) \prod_{i=1}^{n_a} \sum_{n_i=1}^\infty n_i p_{ia} p_{it}^{n_i- 1} p(n_i) \\
&= p(\theta) \prod_{i=1}^{n_a} p_{ia} \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i)
\end{align}
where the last line follows by assuming prior independence.

In the following sections, we derive the expressions for $p_{ia}$ and
$p_{it}$ in terms of the data, and analytical solutions for the sum
under different priors.

\subsubsection{Solution for the sum}\label{solution-for-the-sum}

This section is concerned with finding analytical solutions to
$\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i)$ under two different
assumptions for $p(n_i)$. First, we consider the improper prior
$p(n_i) \propto 1/n_i$ which has several convenient properties. Then,
we consider a negative binomial distribution (including the geometric
and Poisson distributions as a special and limiting case respectively).

The prior $p(n_i) \propto 1/n_i$ means that the model coincides with
the frequentist likelihood-based approach, up to the priors on other
parameters \cites[section 4.2]{dempsterMaximum}{heiseyModelling}[section 8.7.5]{gelmanBayesian}, and is
the reference prior for this quantity\autocite{heBayesiana}. Under this prior,
$\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) \propto \sum_{n_i=1}^\infty p_{it}^{n_i-1} = 1/(1-p_{it})$.
% (\protect\hyperlink{ref-dempsterMaximum}{Dempster, Laird, and
% Rubin} (\protect\hyperlink{ref-dempsterMaximum}{1977}) section 4.2;
% \protect\hyperlink{ref-heiseyModelling}{Heisey and Nordheim}
% (\protect\hyperlink{ref-heiseyModelling}{1995});
% \protect\hyperlink{ref-gelmanBayesian}{Gelman et al.}
% (\protect\hyperlink{ref-gelmanBayesian}{2013}) section 8.7.5)

Now consider $N_i \sim \text{NegBin}(\mu, r)$ using the
mean/dispersion parameterisation of the negative binomial. Equivalently:
\begin{align}
N_i \mid \lambda &\sim \text{Poisson}(\lambda) \\
\lambda &\sim \text{Gamma}(a, b)
\end{align}
where $b = r / \mu$ and $a = r$.
Hence:
\begin{align}
\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) 
&= \int \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i \mid \lambda) p(\lambda) d\lambda \\
&= \int \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} \frac{\lambda^{n_i} e^{-\lambda}}{n_i!} p(\lambda) d\lambda \\
&= \int e^{-\lambda} \sum_{n_i=1}^\infty p_{it}^{n_i- 1} \frac{\lambda^{n_i-1}\lambda }{(n_i-1)!} p(\lambda) d\lambda \\
&= \int e^{-\lambda} \lambda p(\lambda) \sum_{n_t=0}^\infty p_{it}^{n_t} \frac{\lambda^{n_t} }{n_t!} d\lambda &n_t = n_i - 1 \\
&= \int e^{-\lambda} \lambda p(\lambda) e^{p_{it}\lambda} d\lambda &\text{Poisson pmf} \\
&= \int e^{\lambda (p_{it} - 1)} \lambda p(\lambda) d\lambda \\
&= \int e^{\lambda (p_{it} - 1)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda \\
&= \frac{\Gamma(a+1)b^a}{\Gamma(a) (b+1-p_{it})^{a+1}} \\ 
  &\; \times \int \frac{(b+1-p_{it})^{a+1}}{\Gamma(a+1)} \\
  &\; \lambda^{(a+1)-1} e^{-(b+1-p_{it})\lambda} d\lambda \\
&= \frac{\Gamma(a+1)b^a}{\Gamma(a) (b+1-p_{it})^{a+1}} &\text{as gamma pdf} \\
&\propto (b+1-p_{it})^{-(a+1)} \\
&= (r/\mu+1-p_{it})^{-(r+1)} \\
&\propto (r+\mu (1-p_{it}))^{-(r+1)}
\end{align}
Note that $r=0$ recovers the previous derivation (up to
proportionality).

The above allows us to sample from the posterior without needing to
sample the $n_i$s. This is convenient because there are as many
$n_i$s as episodes observed and hence is computationally costly to
sample. Furthermore, each $n_i$ is a discrete parameter, and hence
cannot be sampled within the current Stan implementation.

Having estimated the posterior of the other parameters, we might want to
consider the posterior of $n_i$ (normally for diagnostic or model
debugging purposes), which we can sample from the full conditional.
Specifically, for each posterior sample of $\theta$, we sample one
draw from $n_i \mid \theta, y$ where $y$ represents all the data.

Due to the conditional independence of the $n_i$s, we find that:
\begin{align}
&p(n_i \mid \theta, n_{ia} = 1, n_{iu} = 0) \\
&\propto p(n_{ia} = 1, n_{iu} = 0, n_{it} = n_i - 1 \mid n_i, p_{iu}, p_{ia}, p_{it}) p(n_i) \\
&\propto n_i p_{it}^{n_i- 1} p(n_i)
\end{align}
With the prior, $p(n_i) \propto 1/n_i$ then this is a geometric
distribution with parameter $1 - p_{it}$. With a negative binomial
prior, we have:
\begin{align}
n_i p_{it}^{n_i- 1} p(n_i)
&\propto n_i p_{it}^{n_i- 1} \frac{\Gamma(r + n_i)}{n_i!} \left( \frac{\mu}{r+\mu} \right)^{n_i} \\
&\propto \frac{\Gamma((r + 1) + (n_i - 1))}{(n_i-1)!} \left( \frac{\mu p_{it}}{r+\mu} \right)^{n_i-1}
\end{align}
which is a negative binomial pmf with size parameter $r+1$ and
probability parameter $\frac{r + \mu (1 - p_{it})}{r+\mu}$. The mean
of this is
\begin{align}
\frac{(r+1)\mu p_{it}}{r+\mu(1-p_{it})}
\end{align}

\subsubsection{Derivations of probabilities} \label{sec:deriv-probs}

By definition, we have
\begin{align}
p_{ia} &= \prob((b, e) \in \alpha_i) \\
\alpha_i &= \{ (b, e) : l_i^{(b)} \leq b \leq r_i^{(b)} \wedge l_i^{(e)} \leq e \leq r_i^{(e)}\}.
\intertext{Hence:}
p_{ia}
&= \prob \left( l_i^{(b)} \leq B_i \leq r_i^{(b)}, l_i^{(e)} \leq E_i \leq r_i^{(e)} \right) \\
&= \prob \left( l_i^{(e)} \leq E_i \leq r_i^{(e)} \mid l_i^{(b)} \leq B_i \leq r_i^{(b)} \right) \prob \left( l_i^{(b)} \leq B_i \leq r_i^{(b)} \right) \\
&=\sum_{b = l_i^{(l)}}^{r_i^{(b)}} \prob \left( l_i^{(e)} \leq E_i \leq r_i^{(e)} \mid B_i = b \right) \prob \left(B_i = b \right) \\
&=\sum_{b = l_i^{(b)}}^{r_i^{(b)}} \prob \left( l_i^{(e)} - b + 1 \leq D_i \leq r_i^{(e)} - b + 1 \right) \prob \left(B_i = b \right) \\
&=\sum_{b = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b + 1) - S_\theta(r_i^{(e)} - b + 2) \right) \prob \left(B_i = b \right) \\
&\propto \sum_{b = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b + 1) - S_\theta(r_i^{(e)} - b + 2) \right)
\end{align}
under the assumption of uniform probability of infection time.

Next, we derive $p_{it} = \prob((b, e) \in \Omega^C_i)$, the probability
that episode $i$ was truncated. A truncated episode means that no test
was performed during the episode or there was no negative test prior to
the episode. Denote by $t_i$ the set of testing times for the
individual in which episode $i$ was observed, and define the time
until the next test on the individual after time $t'$ as:
\begin{align}
t^N_{it} &= \min \{ t' \in t_i : t' \geq t \} - t.
\intertext{Then:}
\Omega^C_i
&= \{ (b, e) : \nexists t \in t_i. b \leq t \leq e \vee b \leq \min(t_i) \} \\
&= \{ (b, e) : e - b < t^N_{ib} \vee b \leq \min(t_i) \}.
\intertext{Hence:}
1 - p_{it}
&= 1 - \prob(E_i - B_i < t_{iB_i}^N \vee B_i \leq \min(t_i)) \\
&= 1 - \prob(E_i - B_i < t_{iB_i}^N \wedge B_i > \min(t_i)) - \prob(B_i \leq \min(t_i)) \\
&= 1 - \sum_{b=\min(t_i) + 1}^T \prob(E_i - b + 1 < t_{ib}^N \mid B_i = b) \prob(B_i = b) \\
  &\; - \sum_{t=1}^{\min(t_i)} \prob(B_i = b)\\
&= 1 - \frac{1}{T} \sum_{b=\min(t_i)+1}^T (1 - S_\theta(t_{ib}^N + 1)) - \frac{\min(t_i)}{T} \\
&= 1 - \frac{T-\min(t_i)}{T} + \frac{1}{T} \sum_{b=\min(t_i)+1}^T S_\theta(t_{ib}^N + 1)) - \frac{\min(t_i)}{T} \\
&= \frac{1}{T} \sum_{b=\min(t_i)+1}^T S_\theta(t_{ib}^N + 1)
\end{align}

\subsection{Full posterior}\label{full-posterior}

Under the prior $p(n_i) \propto 1/n_i$:
\begin{align}
&p(\theta \mid n_{ia} = 1, n_{iu} = 0, \dots, n_{n_a,a} = 1, n_{n_a,u} = 0) \\
&\propto p(\theta) \prod_{i=1}^{n_a} p_{ia} \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) \\
&= p(\theta) \prod_{i=1}^{n_a} \frac{p_{ia}}{1-p_{it}} \\
&\propto p(\theta) \prod_{i=1}^{n_a} \frac{\sum_{b = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(r_i^{(e)} - b - 1) - S_\theta(l_i^{(e)} - b - 1) \right)}{\sum_{b=\min(t_i)}^T S_\theta(t_{ib}^N + 1)} \\
\end{align}

Under the prior $N_i \sim \text{NegBinom}(\mu, r)$:
\begin{align}
&p(\theta \mid n_{ia} = 1, n_{iu} = 0, \dots, n_{n_a,a} = 1, n_{n_a,u} = 0) \\
&\propto p(\theta) \prod_{i=1}^{n_a} p_{ia} \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) \\
&= p(\theta) \prod_{i=1}^{n_a} \frac{p_{ia}}{(r+\mu (1-p_{it}))^{(r+1)}} \\
&= p(\theta) \prod_{i=1}^{n_a} \frac{\sum_{b=l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(r_i^{(e)} - b - 1) - S_\theta(l_i^{(e)} - b - 1) \right)}{\left( r+\mu/T \left( \sum_{b=\min(t_i)}^T S_\theta(t_{ib}^N + 1) \right) \right)^{(r+1)}} \\
\end{align}

\section{Total model}\label{total-model}

We now consider all $N$ individuals in the CIS cohort who have at
least one test during the period of interest, this includes those
without an observed infection, and assume that each is equally likely to
be infected. As before, $i = 1, \dots, n_a$ indexes episodes but now
$j = 1, \dots, N$ indexes individuals. All episodes have an associated
individual, each individual may have any number (including 0) episodes.
Admissible regions are only defined for episodes, and have not changed
from the previous section. The truncated region is defined for each
individual, in addition we consider the combined truncated region which
is episodes that would be truncated without conditioning on which
individual the episode occurs in:
$\Omega^C = \{ (b, e, j) : (b, e) \in \Omega_j^C \}$.

Denote by $n_\text{tot}$ the total number of episodes across all $N$
individuals, regardless of whether they are admissible or truncated (we
know there are no inadmissible episodes by definition).

For any infection, the probability of it being admissible for episode
$i$ is $\frac{1}{N} p_{ia}$. That is, the probability that the
infection occurs in the individual corresponding to the individual in
which episode $i$ ($1/N$ by the assumption that episodes are equally
likely to occur in any of the $N$ individuals) multiplied by the
probability that the episode is admissible for episode $i$ conditional
on it occurring in the relevant individual ($p_{ia}$).

The total number of truncated infections that occur is
$n_t = n_\text{tot} - n_a$. Conditional on an infection occurring in
individual $j$, its probability of being truncated is $p_{jt}$ (as
defined previously). Therefore, the overall probability of an episode
being truncated is $p_t = \frac{1}{N} \sum_{j=1}^N p_{jt}$.

Consider our data as there being one admissible episode for each
observed episode, and all other episodes being truncated. As in the
previous section, conditional on $n_\text{tot}$, we have a multinomial
likelihood.
\begin{align}
p(\text{data} \mid \theta)
&= \sum_{n_\text{tot}=n_a}^\infty p(\text{data} \mid n_\text{tot}, \theta) p(n_\text{tot}) \\
&= \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} \left( \prod_{i=1}^{n_a} \frac{1}{N} p_{ia} \right) p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \\
&= \left( \prod_{i=1}^{n_a} \frac{1}{N} p_{ia} \right) \left( \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \right) \\
\end{align}

In this case, it would be computationally feasible to augment the data
with $n_\text{tot}$ directly (as it is only a single parameter).
However, implementation is easier if we can use a closed-form for the
sum (as in the previous section) as it allows the use of standard
implementations of NUTS such as Stan, which cannot handle discrete
parameters.

Assuming $N_\text{tot} \sim \text{NegBin}(\mu, r)$:
\begin{align}
&\sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \\
&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot} \mid \lambda) p(\lambda) d\lambda \\
&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} \frac{\lambda^n_\text{tot} e^{-\lambda}}{n_\text{tot}!} p(\lambda) d\lambda \\
&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{1}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} \lambda^{n_\text{tot}-n_a} \lambda^{n_a} e^{-\lambda} p(\lambda) d\lambda \\
&= \int \lambda^{n_a} e^{-\lambda} p(\lambda) \sum_{n_t=0}^\infty \frac{1}{n_t!} p_t^{n_t} \lambda^{n_t} d\lambda &n_t = n-n_a\\
&= \int \lambda^{n_a} e^{-\lambda} p(\lambda) e^{\lambda p_t} d\lambda \\
&= \int \lambda^{n_a} e^{-\lambda(1 - p_t)} p(\lambda) d\lambda \\
&= \int \lambda^{n_a} e^{-\lambda(1 - p_t)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda \\
&= \int \frac{b^a}{\Gamma(a)} \lambda^{a+n_a-1} e^{-(b+1-p_t)\lambda} \lambda d\lambda \\
&= \frac{b^a}{\Gamma(a)} \frac{\Gamma(a+n_a)}{(b+1-p_t)^{a+n_a}} \\
&\propto (b+1-p_t)^{-(a+n_a)} \\
&= (r/\mu + 1 - p_t)^{-(r+n_a)} \\
&\propto(r + \mu (1- p_t))^{-(r+n_a)}
\end{align}

Which gives the full posterior density as:
\begin{align}
p(\theta \mid \text{data})
&\propto \frac{p(\theta) \prod_{i=1}^{n_a} p_{ia}}{(r + \mu(1 - p_t))^{r+n_a}}
\end{align}

As before, being able to reconstruct the posterior of $n_\text{tot}$
using its full conditional is useful.
\begin{align}
p(n_\text{tot} \mid \text{data}, \theta)
&\propto p(\text{data} \mid \theta, n_\text{tot}) p(n_\text{tot}) \\
&= \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} \left( \prod_{i=1}^{n_a} \frac{1}{N} p_{ia} \right) p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \\
&\propto \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}} \frac{\Gamma(r + n_\text{tot})}{n_\text{tot}!} \left( \frac{\mu}{r + \mu} \right)^{n_\text{tot}}  \\
&\propto \frac{\Gamma(r + n_\text{tot})}{(n_\text{tot}-n_a)!} \left( \frac{\mu p_t}{r + \mu} \right)^{n_\text{tot}}  \\
&\propto \frac{\Gamma((r + n_a) + (n_\text{tot}- n_a))}{(n_\text{tot}-n_a)!} \left( \frac{\mu p_t}{r + \mu} \right)^{n_\text{tot}-n_a}  \\
\end{align}
which is a negative binomial pmf with size parameter $r + n_a$ and
probability parameter $\frac{r+\mu(1-p_t)}{r+\mu}$.

\section{False negatives}\label{false-negatives}

In this section, we relax the assumption of perfect testing to allow
some false negatives. We use a fairly simplified model to ensure the
likelihood is tractable. This requires modifying both $p_{ia}$ to
allow for the episode possibly being longer than observed, and
$p_{it}$ to allow for additional episodes being missed.

Specifically, we introduce a constant test sensitivity $p_\text{sens}$
such that:
\begin{align}
\prob(y_i(t) = 1) = \begin{cases} p_\text{sens} &b_i \leq t \leq e_i \\ 0 &\text{otherwise.}\end{cases}
\end{align}

In practice, it is likely that the test sensitivity is a function of the
viral load of an individual (the concentration of viral material in the
part of the body being sampled). For this initial attempt at modelling
the sensitivity, we ignore this complication.

\subsection{Modifying $p_{ia}$} \label{modifying-p_ia}

The idea when modifying $p_{ia}$ is to allow for the negative test
following the last positive to be either a true or false negative. If it
is a true negative, then we have a bound on the length of the episode.
If it is a false negative, then we consider the episode's length
right-censored. The likelihood contribution for an episode is then a
mixture of these two scenarios, with the mixture probability determined
by the test sensitivity.

For tractability, we assume that the tests bounding the start of the
infection ($l_i^{(b)}$ and $r_i^{(b)}$) are true results and hence
$b_i \in [l_i^{(b)}, r_i^{(b)}]$. Furthermore, we consider the tests
only between $r_i^{(b)}$ and $l_i^{(e)}$ inclusive (the positive
tests providing a lower bound on the length of the episode). Denote
these tests by
$t'_i = \{ t \in t_i : r_i^{(b)} \leq t \leq l_i^{(e)} \}$ and their
results by $y_i'$. We know that the test results at times in $t_i'$
are either true positives or false negatives. By definition, the test at
$r_i^{(e)}$ is a false negative if and only if $e_i > r_i^{(e)}$. We
proceed by considering the two cases.

First, if $e_i \leq r_i^{(e)}$. In this case, the test at
$r_i^{(e)}$ is a true negative, as are all other tests not in
$t_i'$, and these occur with probability 1.
\begin{align}
&p(y_i', e_i \leq r_i^{(e)} | b_i, p_\text{sens}, \theta) \\
&= p(y_i', l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\ % &\text{as no false positives}
&= p(y_i' \mid l_i^{(e)} \leq e_i \leq r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
&= \left( \prod_{t \in t_i'} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) \right)
\end{align}

Second, if $e_i > r_i^{(e)}$. In this case, the test at
$r_i^{(e)}$ is a false negative, occurring with probability
$(1 - p_\text{sens})$, and we have no upper bound on the end of the
infection.
\begin{align}
&p(y_i', e_i > r_i^{(e)} | b_i, p_\text{sens}, \theta) \\
&= p(y_i \mid e_i > r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(e_i > r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
&= \left( \prod_{t \in t_i'} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) (1 - p_\text{sens}) S_\theta(r_i^{(e)} - b_i - 1)
\end{align}

Combining the above, the replacement for $p_{ia}$ is:
\begin{align}
p_{ia}'
&= p(y_i' \mid p_\text{sens}, \theta) \\
&= \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( p(y_i', b_i, e_i \leq r_i^{(e)} \mid p_\text{sens}, \theta) p(y_i', b_i e_i > r_i^{(e)} \mid p_\text{sens}, \theta) \right) p(b_i \mid p_\text{sens}, \theta) \\
&= \left( \prod_{t \in t_i'} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \\ & \ \times \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) + (1 - p_\text{sens}) S_\theta(r_i^{(e)} - b_i - 1) \right) p(b_i \mid p_\text{sens}, \theta) \\
&= \left( \prod_{t \in t_i'} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right)\sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - p_\text{sens} S_\theta(r_i^{(e)} - b_i - 1) \right) p(b_i \mid p_\text{sens}, \theta).
\end{align}
Note that if $p_\text{sens} = 1$ then $p_{ia}' = p_{ia}$.

\subsection{Modifying $p_{it}$} \label{modifying-p_it}

With false negatives, an episode with can be truncated in the following
ways:

\begin{enumerate}
\item
  The episode begins before $\min(t_i)$ (assuming there is a
  negligible probability of a false negative test at this point).
\item
  The duration of the episode is less than $t_{ib_i}^N$. These first
  two mechanisms are how truncation previously occurred, with
  probability $p_{it}$.
\item
  The duration of the episode is at least as long as $t_{it}^N$ but
  less than the time to the second test after $t$, $t_{it}^{2N}$,
  and a false negative episode occurred at the first test after $t$.
  This occurs with probability:
  \begin{math}
    (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(t_i)}^T \left( S_\theta(t_{ib}^N + 1) - S_\theta(t_{ib}^{2N} + 1)\right).
  \end{math}
\item
  The duration of the episode is at least as long as $t_{it}^{2N}$,
  and all the tests within the episode (at least two) are false
  negatives. We assume this occurs with negligible probability.
\end{enumerate}

Hence, the replacement for $1 - p_{it}$ is:
\begin{align}
1 - p_{it}'
&= 1 - p_{it} - (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(t_i)}^T \left( S_\theta(t_{ib}^N + 1) - S_\theta(t_{ib}^{2N} + 1)\right) \\
&= \frac{1}{T} \sum_{b=\min(t_i)}^T \left( p_\text{sens} S_\theta(t_{ib}^N + 1) + (1 - p_\text{sens}) S_\theta(t_{ib}^{2N} + 1)\right) \\
\end{align}

\end{document}

### 2023-05-05 cis-imperfect-testing.tex
Sort chapter structure

--- 2023-05-05 cis-perfect-testing.tex
@@
 
-\title{The model for observing infections}
+\title{Estimating duration in the presence of misclassification}
 \author{Joshua Blake}
@@
 
-\section{Introduction}\label{introduction}
-
-This paper discusses a model for observing duration of PCR positivity in
-the Coronavirus (COVID-19) Infection Survey (CIS). The CIS continually
-enrols individuals and tests them on an ongoing basis, following a
-specified testing protocol. The protocol specifies that the initial five
-tests are spaced every 7 days, then the spacing is every 28 days.
-However, real-world considerations often lead to variations in test
-times and missed tests. Our analysis focuses solely on the binary result
-(positive or negative) of these tests, assuming no misclassification
-bias
-
-Two major challenges arise when dealing with this data: double interval
-censoring and left truncation. Double interval censoring arises because
-we can only bound the start and end of an infection by observing an
-individual's test results transitioning from negative to positive and
-vice versa. Left truncation occurs because an individual is infected at
-time $t$ is only observed if their infection duration is at least
-$t^N - t$ days long, where $t^N$ is the time of their next test.
-
-To address these challenges, we employ two distinct approaches: the
-``individual'' model and the ``total'' model.
-
-The individual model considers only individuals with observed (i.e., not
-truncated) episodes. This approach is similar to methods used in
-previous studies, such as \textcite{heiseyModelling},
-\textcite{dempsterMaximum}, and
-\textcite{turnbullEmpirical}. The conceptual
-approach here is to form a cohort of individuals for which we observe at
-least one positive test, and then consider a study which had only
-enrolled these individuals. The modelling approach then corrects for
-this selection bias.
-
-In contrast, the total model also takes into account data from
-individuals without any observed episodes. These individuals may have
-truncated episodes or may not have been infected at all. This approach
-utilises their test times to estimate the probability of truncated
-infections, ultimately providing an interpretable estimate of the number
-of truncated episodes in the cohort.
-
-Unlike most Bayesian methods (eg: \textcite{heBayesiana}, \textcite{heBayesian}, and \textcite{caoModeling}) we avoid augmenting the
-data with unobserved times of infection or any information on the
-truncated infections; this is due to the large number of infections
-(>4000) making this approach computationally expensive.
-
-\section{Data}\label{data}
-
-Episodes are taken from the data extract of the 3rd Jan 2023, as
-computed by SARAH using her updated episode definition. Only tests taken
-as part of CIS are included (ie: no pillar 2 testing). The following
-criteria are applied.
-
-\begin{itemize}
-\item
-  The first positive in the episode occurred between 16th Oct 2020 and
-  5th Dec 2020 inclusive.
-\item
-  The individual had recorded a negative test prior to their episode
-  beginning (this could have happened at any time, including prior to
-  16th Oct).
-\item
-  A negative test following the final positive in the episode has been
-  recorded.
-\end{itemize}
-
-\section{Notation and assumptions}\label{notation-and-assumptions}
-
-\begin{itemize}
-\item
-  We initially assume perfect testing (no misclassification), although
-  later relax this assumption.
-\item
-  Time is discrete: $1, \dots, T$.
-\item
-  We observe $n_a$ episodes of infection, indexed by $i$.
-\item
-  The $i$ in which episode $i$ occurs is tested at times
-  $t_i = \{ t_{i,1}, \dots, t_{i,m_i} \}$.
-\item
-  Each observed episode has an unknown start of infection time $b_i$
-  and end of infection time $e_i$ (the first and last day that the
-  individual would test positive due to this episode respectively).
-  These are realisations of the random variables $B_i$ and $E_i$
-  respectively.
-\item
-  The duration of the infection is the number of days for which an
-  individual tests positive, $D_i = E_i - B_i + 1$.
-\item
-  We assume that, for all episodes $i$, $D_i$ is iid and independent
-  of the time of the infection. Define the survival function
-  $\prob(D_i \geq t \mid B_i = b, \theta) = \prob(D_i \geq t \mid \theta) = S_\theta(t)$,
-  where $\theta$ are the parameters controlling the survival
-  distribution (the discussion in this document is valid regardless of
-  the model specified for $S_\theta$ and hence we consider $\theta$
-  as an arbitrary vector of parameters).
-\item
-  The beginning of episode $i$ is known to occur in the interval
-  $[l_i^{(b)}, r_i^{(b)}]$, and similarly for the end of the infection
-  in $[l_i^{(e)}, r_i^{(e)}]$.
-\item
-  $y_i(t)$ for $t \in t_i$ is a binary indicator giving the test
-  result for individual $i$ at time $t$. Under the perfect testing
-  assumption, $y_i(t) = 1$ if and only if $b_i \leq t \leq e_i$.
-\item
-  Throughout, the convention that lower-case letters are realisations of
-  upper-case random variables is used.
-\end{itemize}
-
-\section{Individual model}\label{individual-model}
-
-\subsection{Likelihood}\label{likelihood}
-
-Episode $i$ can be fully characterised by the pair $(b_i, e_i)$ of
-start and end dates of the episode, which belong to the state space
-$T \times T$. If the individual in which the episode occurs is tested
-between $b_i$ and $e_i$ (inclusive), then we observe the episode,
-otherwise it is truncated. For the $i$th episode that we observed, we
-consider there is an unknown number, $n_{it}$, where $(b, e)$ are
-drawn from the same distribution as those that lead to episode $i$,
-and occur in an (imagined) individual identical to $i$, but were
-truncated. Episodes are also truncated if there is no negative test
-prior to their start, that is, the individual the episode occurred in
-had no tests before $b_i$. The truncated episodes are often referred
-to as the ``ghosts'' of $i$.
-
-Formally (and following \textcite{heiseyModelling}), we
-define three sets which partition the space of possible infection and
-recovery (ie: are strict subsets of $T \times T$) given the
-observations associated with episode $i$ (graphically shown in Figure
-\ref{fig:partitionSpace}). Each episode, whether observed or not, must
-fall into one of these three classes with a probability that is a
-function of $\theta$ (see section \ref{sec:deriv-probs}).
-
-\begin{itemize}
-\item
-  Admissible episodes, $\alpha_i$, which have an infection and
-  recovery time within their respective intervals observed.
-  $n_{ia} =1$ of these occurred, each with probability
-  $p_{ia} = \prob((b, e) \in \alpha_i \mid \theta)$.
-\item
-  Truncated episodes, $\Omega_i^C$, which have an infection and
-  recovery time such that they would not have tested positive. An
-  unknown number, $n_{it}$ of these observed, each with probability
-  $p_{it} = \prob((b, e) \in \Omega^C_i \mid \theta)$.
-\item
-  Inadmissible episodes, $\beta_i$, which do not have an infection and
-  recovery time within their respective intervals but would have been
-  observed (not truncated). This is all remaining episodes not in the
-  previous sets. $n_{iu} = 0$ of these occurred, where they would have
-  occurred with probability
-  $p_{iu} = \prob((b, e) \in \beta_i \mid \theta)$.
-\end{itemize}
-
-The untruncated region (where we could have observed an episode) is
-$\Omega_i = \alpha_i \cup \beta_i$ (the notation here is used for
-consistency with \textcite{heiseyModelling}), and is the
-complement of $\Omega^C_i$.
-
-\begin{figure}
-\centering
-\includegraphics[width=\textwidth]{cis-perfect-testing/images/likelihood_areas.jpg}
-\caption{Admissible, $\alpha_i$, inadmissible, $\beta_i$ and
-untruncated, $\Omega_i^C$ regions for an episode $i$ in an
-individual whose first test was at time 0 and negative, with subsequent
-negative tests at times 7, 14, and 56, and positive tests at times 21
-and 28. Purple region is $\alpha_i$, orange region (which includes all
-negative infection times) is $\Omega_i^C$ and unshaded region is
-$\beta_i$.\label{fig:partitionSpace}}
-\end{figure}
-
-These three classes span all possible episodes and are mutually
-exclusive, and hence the total number of episodes,
-$n_i = n_{ia} + n_{iu} + n_{it}$, and
-$p_{ia} + p_{iu} + p_{it} = 1$.
-
-Episode $i$ and its ghosts independently occur belong to one of these
-classes. Therefore, conditional on $n_i$, the number of episodes in
-each class (that is, the counts $n_{ia}, n_{iu}, n_{it}$) are
-distributed multinomially. Hence:
-\begin{align}
-&p(n_{ia} = 1, n_{iu} = 0, n_{it} = n_i - 1 \mid n_i, \theta) \\
-&= \frac{n_i!}{n_{ia}! n_{iu}! (n_i- n_{ia} - n_{it})!} p_{ia}^{n_{ia}} p_{ua}^{n_{ia}} p_{it}^{n_i- n_{ia} - n_{it}} \\
-&= \frac{n_i!}{(n_i-1)!} p_{ia} p_{it}^{n_i- 1} &\text{as $n_{ia} = 1$ and $n_{iu} = 0$}\\
-&= n_i p_{ia} p_{it}^{n_i- 1}
-\end{align}
-
-The posterior we are interested in is
-$p(\theta \mid n_{ia} = 1, n_{iu} = 0, \dots, n_{n_a,a} = 1, n_{n_a,u} = 0)$,
-where $\theta$ are the parameters governing the survival distribution
-and hence allow derivation of the $p$s.
-\begin{align}
-&p(\theta \mid n_{ia} = 1, n_{iu} = 0, \dots, n_{n_a,a} = 1, n_{n_a,u} = 0) \\
-&\propto p(\theta) \prod_{i=1}^{n_a} p(n_{ia} = 1, n_{iu} = 0 \mid \theta) \\
-&= p(\theta) \prod_{i=1}^{n_a} \sum_{n_i=1}^\infty p(n_{ia} = 1, n_{iu} = 0, n_{it} = n_i - 1 \mid \theta, n_i) p(n_i \mid \theta) \\
-&= p(\theta) \prod_{i=1}^{n_a} \sum_{n_i=1}^\infty n_i p_{ia} p_{it}^{n_i- 1} p(n_i) \\
-&= p(\theta) \prod_{i=1}^{n_a} p_{ia} \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i)
-\end{align}
-where the last line follows by assuming prior independence.
-
-In the following sections, we derive the expressions for $p_{ia}$ and
-$p_{it}$ in terms of the data, and analytical solutions for the sum
-under different priors.
-
-\subsubsection{Solution for the sum}\label{solution-for-the-sum}
-
-This section is concerned with finding analytical solutions to
-$\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i)$ under two different
-assumptions for $p(n_i)$. First, we consider the improper prior
-$p(n_i) \propto 1/n_i$ which has several convenient properties. Then,
-we consider a negative binomial distribution (including the geometric
-and Poisson distributions as a special and limiting case respectively).
-
-The prior $p(n_i) \propto 1/n_i$ means that the model coincides with
-the frequentist likelihood-based approach, up to the priors on other
-parameters \cites[section 4.2]{dempsterMaximum}{heiseyModelling}[section 8.7.5]{gelmanBayesian}, and is
-the reference prior for this quantity\autocite{heBayesiana}. Under this prior,
-$\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) \propto \sum_{n_i=1}^\infty p_{it}^{n_i-1} = 1/(1-p_{it})$.
-% (\protect\hyperlink{ref-dempsterMaximum}{Dempster, Laird, and
-% Rubin} (\protect\hyperlink{ref-dempsterMaximum}{1977}) section 4.2;
-% \protect\hyperlink{ref-heiseyModelling}{Heisey and Nordheim}
-% (\protect\hyperlink{ref-heiseyModelling}{1995});
-% \protect\hyperlink{ref-gelmanBayesian}{Gelman et al.}
-% (\protect\hyperlink{ref-gelmanBayesian}{2013}) section 8.7.5)
-
-Now consider $N_i \sim \text{NegBin}(\mu, r)$ using the
-mean/dispersion parameterisation of the negative binomial. Equivalently:
-\begin{align}
-N_i \mid \lambda &\sim \text{Poisson}(\lambda) \\
-\lambda &\sim \text{Gamma}(a, b)
-\end{align}
-where $b = r / \mu$ and $a = r$.
-Hence:
-\begin{align}
-\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) 
-&= \int \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i \mid \lambda) p(\lambda) d\lambda \\
-&= \int \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} \frac{\lambda^{n_i} e^{-\lambda}}{n_i!} p(\lambda) d\lambda \\
-&= \int e^{-\lambda} \sum_{n_i=1}^\infty p_{it}^{n_i- 1} \frac{\lambda^{n_i-1}\lambda }{(n_i-1)!} p(\lambda) d\lambda \\
-&= \int e^{-\lambda} \lambda p(\lambda) \sum_{n_t=0}^\infty p_{it}^{n_t} \frac{\lambda^{n_t} }{n_t!} d\lambda &n_t = n_i - 1 \\
-&= \int e^{-\lambda} \lambda p(\lambda) e^{p_{it}\lambda} d\lambda &\text{Poisson pmf} \\
-&= \int e^{\lambda (p_{it} - 1)} \lambda p(\lambda) d\lambda \\
-&= \int e^{\lambda (p_{it} - 1)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda \\
-&= \frac{\Gamma(a+1)b^a}{\Gamma(a) (b+1-p_{it})^{a+1}} \\ 
-  &\; \times \int \frac{(b+1-p_{it})^{a+1}}{\Gamma(a+1)} \\
-  &\; \lambda^{(a+1)-1} e^{-(b+1-p_{it})\lambda} d\lambda \\
-&= \frac{\Gamma(a+1)b^a}{\Gamma(a) (b+1-p_{it})^{a+1}} &\text{as gamma pdf} \\
-&\propto (b+1-p_{it})^{-(a+1)} \\
-&= (r/\mu+1-p_{it})^{-(r+1)} \\
-&\propto (r+\mu (1-p_{it}))^{-(r+1)}
-\end{align}
-Note that $r=0$ recovers the previous derivation (up to
-proportionality).
-
-The above allows us to sample from the posterior without needing to
-sample the $n_i$s. This is convenient because there are as many
-$n_i$s as episodes observed and hence is computationally costly to
-sample. Furthermore, each $n_i$ is a discrete parameter, and hence
-cannot be sampled within the current Stan implementation.
-
-Having estimated the posterior of the other parameters, we might want to
-consider the posterior of $n_i$ (normally for diagnostic or model
-debugging purposes), which we can sample from the full conditional.
-Specifically, for each posterior sample of $\theta$, we sample one
-draw from $n_i \mid \theta, y$ where $y$ represents all the data.
-
-Due to the conditional independence of the $n_i$s, we find that:
-\begin{align}
-&p(n_i \mid \theta, n_{ia} = 1, n_{iu} = 0) \\
-&\propto p(n_{ia} = 1, n_{iu} = 0, n_{it} = n_i - 1 \mid n_i, p_{iu}, p_{ia}, p_{it}) p(n_i) \\
-&\propto n_i p_{it}^{n_i- 1} p(n_i)
-\end{align}
-With the prior, $p(n_i) \propto 1/n_i$ then this is a geometric
-distribution with parameter $1 - p_{it}$. With a negative binomial
-prior, we have:
-\begin{align}
-n_i p_{it}^{n_i- 1} p(n_i)
-&\propto n_i p_{it}^{n_i- 1} \frac{\Gamma(r + n_i)}{n_i!} \left( \frac{\mu}{r+\mu} \right)^{n_i} \\
-&\propto \frac{\Gamma((r + 1) + (n_i - 1))}{(n_i-1)!} \left( \frac{\mu p_{it}}{r+\mu} \right)^{n_i-1}
-\end{align}
-which is a negative binomial pmf with size parameter $r+1$ and
-probability parameter $\frac{r + \mu (1 - p_{it})}{r+\mu}$. The mean
-of this is
-\begin{align}
-\frac{(r+1)\mu p_{it}}{r+\mu(1-p_{it})}
-\end{align}
-
-\subsubsection{Derivations of probabilities} \label{sec:deriv-probs}
-
-By definition, we have
-\begin{align}
-p_{ia} &= \prob((b, e) \in \alpha_i) \\
-\alpha_i &= \{ (b, e) : l_i^{(b)} \leq b \leq r_i^{(b)} \wedge l_i^{(e)} \leq e \leq r_i^{(e)}\}.
-\intertext{Hence:}
-p_{ia}
-&= \prob \left( l_i^{(b)} \leq B_i \leq r_i^{(b)}, l_i^{(e)} \leq E_i \leq r_i^{(e)} \right) \\
-&= \prob \left( l_i^{(e)} \leq E_i \leq r_i^{(e)} \mid l_i^{(b)} \leq B_i \leq r_i^{(b)} \right) \prob \left( l_i^{(b)} \leq B_i \leq r_i^{(b)} \right) \\
-&=\sum_{b = l_i^{(l)}}^{r_i^{(b)}} \prob \left( l_i^{(e)} \leq E_i \leq r_i^{(e)} \mid B_i = b \right) \prob \left(B_i = b \right) \\
-&=\sum_{b = l_i^{(b)}}^{r_i^{(b)}} \prob \left( l_i^{(e)} - b + 1 \leq D_i \leq r_i^{(e)} - b + 1 \right) \prob \left(B_i = b \right) \\
-&=\sum_{b = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b + 1) - S_\theta(r_i^{(e)} - b + 2) \right) \prob \left(B_i = b \right) \\
-&\propto \sum_{b = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b + 1) - S_\theta(r_i^{(e)} - b + 2) \right)
-\end{align}
-under the assumption of uniform probability of infection time.
-
-Next, we derive $p_{it} = \prob((b, e) \in \Omega^C_i)$, the probability
-that episode $i$ was truncated. A truncated episode means that no test
-was performed during the episode or there was no negative test prior to
-the episode. Denote by $t_i$ the set of testing times for the
-individual in which episode $i$ was observed, and define the time
-until the next test on the individual after time $t'$ as:
-\begin{align}
-t^N_{it} &= \min \{ t' \in t_i : t' \geq t \} - t.
-\intertext{Then:}
-\Omega^C_i
-&= \{ (b, e) : \nexists t \in t_i. b \leq t \leq e \vee b \leq \min(t_i) \} \\
-&= \{ (b, e) : e - b < t^N_{ib} \vee b \leq \min(t_i) \}.
-\intertext{Hence:}
-1 - p_{it}
-&= 1 - \prob(E_i - B_i < t_{iB_i}^N \vee B_i \leq \min(t_i)) \\
-&= 1 - \prob(E_i - B_i < t_{iB_i}^N \wedge B_i > \min(t_i)) - \prob(B_i \leq \min(t_i)) \\
-&= 1 - \sum_{b=\min(t_i) + 1}^T \prob(E_i - b + 1 < t_{ib}^N \mid B_i = b) \prob(B_i = b) \\
-  &\; - \sum_{t=1}^{\min(t_i)} \prob(B_i = b)\\
-&= 1 - \frac{1}{T} \sum_{b=\min(t_i)+1}^T (1 - S_\theta(t_{ib}^N + 1)) - \frac{\min(t_i)}{T} \\
-&= 1 - \frac{T-\min(t_i)}{T} + \frac{1}{T} \sum_{b=\min(t_i)+1}^T S_\theta(t_{ib}^N + 1)) - \frac{\min(t_i)}{T} \\
-&= \frac{1}{T} \sum_{b=\min(t_i)+1}^T S_\theta(t_{ib}^N + 1)
-\end{align}
-
-\subsection{Full posterior}\label{full-posterior}
-
-Under the prior $p(n_i) \propto 1/n_i$:
-\begin{align}
-&p(\theta \mid n_{ia} = 1, n_{iu} = 0, \dots, n_{n_a,a} = 1, n_{n_a,u} = 0) \\
-&\propto p(\theta) \prod_{i=1}^{n_a} p_{ia} \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) \\
-&= p(\theta) \prod_{i=1}^{n_a} \frac{p_{ia}}{1-p_{it}} \\
-&\propto p(\theta) \prod_{i=1}^{n_a} \frac{\sum_{b = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(r_i^{(e)} - b - 1) - S_\theta(l_i^{(e)} - b - 1) \right)}{\sum_{b=\min(t_i)}^T S_\theta(t_{ib}^N + 1)} \\
-\end{align}
-
-Under the prior $N_i \sim \text{NegBinom}(\mu, r)$:
-\begin{align}
-&p(\theta \mid n_{ia} = 1, n_{iu} = 0, \dots, n_{n_a,a} = 1, n_{n_a,u} = 0) \\
-&\propto p(\theta) \prod_{i=1}^{n_a} p_{ia} \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) \\
-&= p(\theta) \prod_{i=1}^{n_a} \frac{p_{ia}}{(r+\mu (1-p_{it}))^{(r+1)}} \\
-&= p(\theta) \prod_{i=1}^{n_a} \frac{\sum_{b=l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(r_i^{(e)} - b - 1) - S_\theta(l_i^{(e)} - b - 1) \right)}{\left( r+\mu/T \left( \sum_{b=\min(t_i)}^T S_\theta(t_{ib}^N + 1) \right) \right)^{(r+1)}} \\
-\end{align}
-
-\section{Total model}\label{total-model}
-
-We now consider all $N$ individuals in the CIS cohort who have at
-least one test during the period of interest, this includes those
-without an observed infection, and assume that each is equally likely to
-be infected. As before, $i = 1, \dots, n_a$ indexes episodes but now
-$j = 1, \dots, N$ indexes individuals. All episodes have an associated
-individual, each individual may have any number (including 0) episodes.
-Admissible regions are only defined for episodes, and have not changed
-from the previous section. The truncated region is defined for each
-individual, in addition we consider the combined truncated region which
-is episodes that would be truncated without conditioning on which
-individual the episode occurs in:
-$\Omega^C = \{ (b, e, j) : (b, e) \in \Omega_j^C \}$.
-
-Denote by $n_\text{tot}$ the total number of episodes across all $N$
-individuals, regardless of whether they are admissible or truncated (we
-know there are no inadmissible episodes by definition).
-
-For any infection, the probability of it being admissible for episode
-$i$ is $\frac{1}{N} p_{ia}$. That is, the probability that the
-infection occurs in the individual corresponding to the individual in
-which episode $i$ ($1/N$ by the assumption that episodes are equally
-likely to occur in any of the $N$ individuals) multiplied by the
-probability that the episode is admissible for episode $i$ conditional
-on it occurring in the relevant individual ($p_{ia}$).
-
-The total number of truncated infections that occur is
-$n_t = n_\text{tot} - n_a$. Conditional on an infection occurring in
-individual $j$, its probability of being truncated is $p_{jt}$ (as
-defined previously). Therefore, the overall probability of an episode
-being truncated is $p_t = \frac{1}{N} \sum_{j=1}^N p_{jt}$.
-
-Consider our data as there being one admissible episode for each
-observed episode, and all other episodes being truncated. As in the
-previous section, conditional on $n_\text{tot}$, we have a multinomial
-likelihood.
-\begin{align}
-p(\text{data} \mid \theta)
-&= \sum_{n_\text{tot}=n_a}^\infty p(\text{data} \mid n_\text{tot}, \theta) p(n_\text{tot}) \\
-&= \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} \left( \prod_{i=1}^{n_a} \frac{1}{N} p_{ia} \right) p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \\
-&= \left( \prod_{i=1}^{n_a} \frac{1}{N} p_{ia} \right) \left( \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \right) \\
-\end{align}
-
-In this case, it would be computationally feasible to augment the data
-with $n_\text{tot}$ directly (as it is only a single parameter).
-However, implementation is easier if we can use a closed-form for the
-sum (as in the previous section) as it allows the use of standard
-implementations of NUTS such as Stan, which cannot handle discrete
-parameters.
-
-Assuming $N_\text{tot} \sim \text{NegBin}(\mu, r)$:
-\begin{align}
-&\sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \\
-&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot} \mid \lambda) p(\lambda) d\lambda \\
-&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} \frac{\lambda^n_\text{tot} e^{-\lambda}}{n_\text{tot}!} p(\lambda) d\lambda \\
-&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{1}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} \lambda^{n_\text{tot}-n_a} \lambda^{n_a} e^{-\lambda} p(\lambda) d\lambda \\
-&= \int \lambda^{n_a} e^{-\lambda} p(\lambda) \sum_{n_t=0}^\infty \frac{1}{n_t!} p_t^{n_t} \lambda^{n_t} d\lambda &n_t = n-n_a\\
-&= \int \lambda^{n_a} e^{-\lambda} p(\lambda) e^{\lambda p_t} d\lambda \\
-&= \int \lambda^{n_a} e^{-\lambda(1 - p_t)} p(\lambda) d\lambda \\
-&= \int \lambda^{n_a} e^{-\lambda(1 - p_t)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda \\
-&= \int \frac{b^a}{\Gamma(a)} \lambda^{a+n_a-1} e^{-(b+1-p_t)\lambda} \lambda d\lambda \\
-&= \frac{b^a}{\Gamma(a)} \frac{\Gamma(a+n_a)}{(b+1-p_t)^{a+n_a}} \\
-&\propto (b+1-p_t)^{-(a+n_a)} \\
-&= (r/\mu + 1 - p_t)^{-(r+n_a)} \\
-&\propto(r + \mu (1- p_t))^{-(r+n_a)}
-\end{align}
-
-Which gives the full posterior density as:
-\begin{align}
-p(\theta \mid \text{data})
-&\propto \frac{p(\theta) \prod_{i=1}^{n_a} p_{ia}}{(r + \mu(1 - p_t))^{r+n_a}}
-\end{align}
-
-As before, being able to reconstruct the posterior of $n_\text{tot}$
-using its full conditional is useful.
-\begin{align}
-p(n_\text{tot} \mid \text{data}, \theta)
-&\propto p(\text{data} \mid \theta, n_\text{tot}) p(n_\text{tot}) \\
-&= \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} \left( \prod_{i=1}^{n_a} \frac{1}{N} p_{ia} \right) p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \\
-&\propto \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}} \frac{\Gamma(r + n_\text{tot})}{n_\text{tot}!} \left( \frac{\mu}{r + \mu} \right)^{n_\text{tot}}  \\
-&\propto \frac{\Gamma(r + n_\text{tot})}{(n_\text{tot}-n_a)!} \left( \frac{\mu p_t}{r + \mu} \right)^{n_\text{tot}}  \\
-&\propto \frac{\Gamma((r + n_a) + (n_\text{tot}- n_a))}{(n_\text{tot}-n_a)!} \left( \frac{\mu p_t}{r + \mu} \right)^{n_\text{tot}-n_a}  \\
-\end{align}
-which is a negative binomial pmf with size parameter $r + n_a$ and
-probability parameter $\frac{r+\mu(1-p_t)}{r+\mu}$.
-
-\section{False negatives}\label{false-negatives}
-
-In this section, we relax the assumption of perfect testing to allow
-some false negatives. We use a fairly simplified model to ensure the
-likelihood is tractable. This requires modifying both $p_{ia}$ to
-allow for the episode possibly being longer than observed, and
-$p_{it}$ to allow for additional episodes being missed.
-
-Specifically, we introduce a constant test sensitivity $p_\text{sens}$
-such that:
-\begin{align}
-\prob(y_i(t) = 1) = \begin{cases} p_\text{sens} &b_i \leq t \leq e_i \\ 0 &\text{otherwise.}\end{cases}
-\end{align}
-
-In practice, it is likely that the test sensitivity is a function of the
-viral load of an individual (the concentration of viral material in the
-part of the body being sampled). For this initial attempt at modelling
-the sensitivity, we ignore this complication.
-
-\subsection{Modifying $p_{ia}$} \label{modifying-p_ia}
-
-The idea when modifying $p_{ia}$ is to allow for the negative test
-following the last positive to be either a true or false negative. If it
-is a true negative, then we have a bound on the length of the episode.
-If it is a false negative, then we consider the episode's length
-right-censored. The likelihood contribution for an episode is then a
-mixture of these two scenarios, with the mixture probability determined
-by the test sensitivity.
-
-For tractability, we assume that the tests bounding the start of the
-infection ($l_i^{(b)}$ and $r_i^{(b)}$) are true results and hence
-$b_i \in [l_i^{(b)}, r_i^{(b)}]$. Furthermore, we consider the tests
-only between $r_i^{(b)}$ and $l_i^{(e)}$ inclusive (the positive
-tests providing a lower bound on the length of the episode). Denote
-these tests by
-$t'_i = \{ t \in t_i : r_i^{(b)} \leq t \leq l_i^{(e)} \}$ and their
-results by $y_i'$. We know that the test results at times in $t_i'$
-are either true positives or false negatives. By definition, the test at
-$r_i^{(e)}$ is a false negative if and only if $e_i > r_i^{(e)}$. We
-proceed by considering the two cases.
-
-First, if $e_i \leq r_i^{(e)}$. In this case, the test at
-$r_i^{(e)}$ is a true negative, as are all other tests not in
-$t_i'$, and these occur with probability 1.
-\begin{align}
-&p(y_i', e_i \leq r_i^{(e)} | b_i, p_\text{sens}, \theta) \\
-&= p(y_i', l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\ % &\text{as no false positives}
-&= p(y_i' \mid l_i^{(e)} \leq e_i \leq r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
-&= \left( \prod_{t \in t_i'} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) \right)
-\end{align}
-
-Second, if $e_i > r_i^{(e)}$. In this case, the test at
-$r_i^{(e)}$ is a false negative, occurring with probability
-$(1 - p_\text{sens})$, and we have no upper bound on the end of the
-infection.
-\begin{align}
-&p(y_i', e_i > r_i^{(e)} | b_i, p_\text{sens}, \theta) \\
-&= p(y_i \mid e_i > r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(e_i > r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
-&= \left( \prod_{t \in t_i'} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) (1 - p_\text{sens}) S_\theta(r_i^{(e)} - b_i - 1)
-\end{align}
-
-Combining the above, the replacement for $p_{ia}$ is:
-\begin{align}
-p_{ia}'
-&= p(y_i' \mid p_\text{sens}, \theta) \\
-&= \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( p(y_i', b_i, e_i \leq r_i^{(e)} \mid p_\text{sens}, \theta) p(y_i', b_i e_i > r_i^{(e)} \mid p_\text{sens}, \theta) \right) p(b_i \mid p_\text{sens}, \theta) \\
-&= \left( \prod_{t \in t_i'} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \\ & \ \times \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) + (1 - p_\text{sens}) S_\theta(r_i^{(e)} - b_i - 1) \right) p(b_i \mid p_\text{sens}, \theta) \\
-&= \left( \prod_{t \in t_i'} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right)\sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - p_\text{sens} S_\theta(r_i^{(e)} - b_i - 1) \right) p(b_i \mid p_\text{sens}, \theta).
-\end{align}
-Note that if $p_\text{sens} = 1$ then $p_{ia}' = p_{ia}$.
-
-\subsection{Modifying $p_{it}$} \label{modifying-p_it}
-
-With false negatives, an episode with can be truncated in the following
-ways:
-
-\begin{enumerate}
-\item
-  The episode begins before $\min(t_i)$ (assuming there is a
-  negligible probability of a false negative test at this point).
-\item
-  The duration of the episode is less than $t_{ib_i}^N$. These first
-  two mechanisms are how truncation previously occurred, with
-  probability $p_{it}$.
-\item
-  The duration of the episode is at least as long as $t_{it}^N$ but
-  less than the time to the second test after $t$, $t_{it}^{2N}$,
-  and a false negative episode occurred at the first test after $t$.
-  This occurs with probability:
-  \begin{math}
-    (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(t_i)}^T \left( S_\theta(t_{ib}^N + 1) - S_\theta(t_{ib}^{2N} + 1)\right).
-  \end{math}
-\item
-  The duration of the episode is at least as long as $t_{it}^{2N}$,
-  and all the tests within the episode (at least two) are false
-  negatives. We assume this occurs with negligible probability.
-\end{enumerate}
-
-Hence, the replacement for $1 - p_{it}$ is:
-\begin{align}
-1 - p_{it}'
-&= 1 - p_{it} - (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(t_i)}^T \left( S_\theta(t_{ib}^N + 1) - S_\theta(t_{ib}^{2N} + 1)\right) \\
-&= \frac{1}{T} \sum_{b=\min(t_i)}^T \left( p_\text{sens} S_\theta(t_{ib}^N + 1) + (1 - p_\text{sens}) S_\theta(t_{ib}^{2N} + 1)\right) \\
-\end{align}
-
 \end{document}

### 2023-05-05 cis-perfect-testing.tex
Sort chapter structure

@@
 
-\title{The model for observing infections}
+\title{Estimating duration using survival analysis}
 \author{Joshua Blake}
@@
 \maketitle
+
+This chapter aims to estimate the duration of SARS-CoV-2 PCR positivity from CIS data if perfect testing can be assumed (ie: no misclassification bias).
+I take a survival analysis approach, considering that neither the start nor end time of positivity are known exactly (double interval censoring) and that infections may be missed (truncation); these issues are explained fully below.
+Through a simulation study, I show the applicability of these methods.
+However, application in reality must account for misclassification bias, the subject of the following chapter.
+
+This chapter proceeds as follows.
+\begin{itemize}
+\item  In section~\ref{cpt:sec:model}, I develop a model accounting for double interval censoring and truncation.
+This builds on previous work in both biomedical and bird nesting studies.
+The arbitrary nature of truncation in CIS, where each individual has a unique truncation pattern which is neither simple left nor right truncation, has not previously been explored in the literature.
+Furthermore, I explore a Bayesian framework where the amount of explicit data augmentation is limited allowing the method to scale to the large number of episodes present in CIS without requiring vast computing resources (such as a high-performance cluster) which are unavailable within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
+\item  In section~\ref{cpt:sec:parameters-priors} I discuss parameterisation of this model and the challenges of specifying a vague prior.
+Whether a prior is vague or not depends on the parameterisation chosen, and for independent priors on the parameters, it is not possible to be vague on all sensible parameterisations.
+I also discuss the appeal of informative priors, and how the estimates in the previous chapter could be incorporated.
+\item  Section~\ref{cpt:sec:simulation-study} is a simulation study showing the applicability of this method to the CIS.
+I show that the model developed previously is able to estimate the duration distribution in this setting under vague or informative priors.
+However, I also show that, within the simulation considered, only the survival, and not the hazard, function are identifiable.
+\item  Section~\ref{cpt:sec:conclusion} concludes the chapter, discussing the limitations of this model. The following chapter will motivate and consider relaxing the perfect testing assumption.
+\end{itemize}
 

### 2023-05-15 cis-imperfect-testing.tex
Start drawing together writings into appropriate sections

@@
 
+In this section, we relax the assumption of perfect testing to allow
+some false negatives. We use a fairly simplified model to ensure the
+likelihood is tractable. This requires modifying both $p_{ia}$ to
+allow for the episode possibly being longer than observed, and
+$p_{it}$ to allow for additional episodes being missed.
+
+Specifically, we introduce a constant test sensitivity $p_\text{sens}$
+such that:
+\begin{align}
+\prob(y_i(t) = 1) = \begin{cases} p_\text{sens} &b_i \leq t \leq e_i \\ 0 &\text{otherwise.}\end{cases}
+\end{align}
+
+In practice, it is likely that the test sensitivity is a function of the
+viral load of an individual (the concentration of viral material in the
+part of the body being sampled). For this initial attempt at modelling
+the sensitivity, we ignore this complication.
+
+\section{Modifying $p_{ia}$} \label{modifying-p_ia}
+
+The idea when modifying $p_{ia}$ is to allow for the negative test
+following the last positive to be either a true or false negative. If it
+is a true negative, then we have a bound on the length of the episode.
+If it is a false negative, then we consider the episode's length
+right-censored. The likelihood contribution for an episode is then a
+mixture of these two scenarios, with the mixture probability determined
+by the test sensitivity.
+
+For tractability, we assume that the tests bounding the start of the
+infection ($l_i^{(b)}$ and $r_i^{(b)}$) are true results and hence
+$b_i \in [l_i^{(b)}, r_i^{(b)}]$. Furthermore, we consider the tests
+only between $r_i^{(b)}$ and $l_i^{(e)}$ inclusive (the positive
+tests providing a lower bound on the length of the episode). Denote
+these tests by
+$t'_i = \{ t \in t_i : r_i^{(b)} \leq t \leq l_i^{(e)} \}$ and their
+results by $y_i'$. We know that the test results at times in $t_i'$
+are either true positives or false negatives. By definition, the test at
+$r_i^{(e)}$ is a false negative if and only if $e_i > r_i^{(e)}$. We
+proceed by considering the two cases.
+
+First, if $e_i \leq r_i^{(e)}$. In this case, the test at
+$r_i^{(e)}$ is a true negative, as are all other tests not in
+$t_i'$, and these occur with probability 1.
+\begin{align}
+&p(y_i', e_i \leq r_i^{(e)} | b_i, p_\text{sens}, \theta) \\
+&= p(y_i', l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\ % &\text{as no false positives}
+&= p(y_i' \mid l_i^{(e)} \leq e_i \leq r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
+&= \left( \prod_{t \in t_i'} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) \right)
+\end{align}
+
+Second, if $e_i > r_i^{(e)}$. In this case, the test at
+$r_i^{(e)}$ is a false negative, occurring with probability
+$(1 - p_\text{sens})$, and we have no upper bound on the end of the
+infection.
+\begin{align}
+&p(y_i', e_i > r_i^{(e)} | b_i, p_\text{sens}, \theta) \\
+&= p(y_i \mid e_i > r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(e_i > r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
+&= \left( \prod_{t \in t_i'} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) (1 - p_\text{sens}) S_\theta(r_i^{(e)} - b_i - 1)
+\end{align}
+
+Combining the above, the replacement for $p_{ia}$ is:
+\begin{align}
+p_{ia}'
+&= p(y_i' \mid p_\text{sens}, \theta) \\
+&= \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( p(y_i', b_i, e_i \leq r_i^{(e)} \mid p_\text{sens}, \theta) p(y_i', b_i e_i > r_i^{(e)} \mid p_\text{sens}, \theta) \right) p(b_i \mid p_\text{sens}, \theta) \\
+&= \left( \prod_{t \in t_i'} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \\ & \ \times \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) + (1 - p_\text{sens}) S_\theta(r_i^{(e)} - b_i - 1) \right) p(b_i \mid p_\text{sens}, \theta) \\
+&= \left( \prod_{t \in t_i'} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right)\sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - p_\text{sens} S_\theta(r_i^{(e)} - b_i - 1) \right) p(b_i \mid p_\text{sens}, \theta).
+\end{align}
+Note that if $p_\text{sens} = 1$ then $p_{ia}' = p_{ia}$.
+
+\section{Modifying $p_{it}$} \label{modifying-p_it}
+
+With false negatives, an episode with can be truncated in the following
+ways:
+
+\begin{enumerate}
+\item
+  The episode begins before $\min(t_i)$ (assuming there is a
+  negligible probability of a false negative test at this point).
+\item
+  The duration of the episode is less than $t_{ib_i}^N$. These first
+  two mechanisms are how truncation previously occurred, with
+  probability $p_{it}$.
+\item
+  The duration of the episode is at least as long as $t_{it}^N$ but
+  less than the time to the second test after $t$, $t_{it}^{2N}$,
+  and a false negative episode occurred at the first test after $t$.
+  This occurs with probability:
+  \begin{math}
+    (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(t_i)}^T \left( S_\theta(t_{ib}^N + 1) - S_\theta(t_{ib}^{2N} + 1)\right).
+  \end{math}
+\item
+  The duration of the episode is at least as long as $t_{it}^{2N}$,
+  and all the tests within the episode (at least two) are false
+  negatives. We assume this occurs with negligible probability.
+\end{enumerate}
+
+Hence, the replacement for $1 - p_{it}$ is:
+\begin{align}
+1 - p_{it}'
+&= 1 - p_{it} - (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(t_i)}^T \left( S_\theta(t_{ib}^N + 1) - S_\theta(t_{ib}^{2N} + 1)\right) \\
+&= \frac{1}{T} \sum_{b=\min(t_i)}^T \left( p_\text{sens} S_\theta(t_{ib}^N + 1) + (1 - p_\text{sens}) S_\theta(t_{ib}^{2N} + 1)\right) \\
+\end{align}
+
 \end{document}

### 2023-05-15 cis-perfect-testing.tex
Start drawing together writings into appropriate sections

@@
 This chapter aims to estimate the duration of SARS-CoV-2 PCR positivity from CIS data if perfect testing can be assumed (ie: no misclassification bias).
-I take a survival analysis approach, considering that neither the start nor end time of positivity are known exactly (double interval censoring) and that infections may be missed (truncation); these issues are explained fully below.
+I take a survival analysis approach, considering that neither the start nor end time of positivity are known exactly (double interval censoring) and that infections may be missed (truncation), explained in the following paragraphs.
 Through a simulation study, I show the applicability of these methods.
 However, application in reality must account for misclassification bias, the subject of the following chapter.
-
-This chapter proceeds as follows.
-\begin{itemize}
-\item  In section~\ref{cpt:sec:model}, I develop a model accounting for double interval censoring and truncation.
-This builds on previous work in both biomedical and bird nesting studies.
-The arbitrary nature of truncation in CIS, where each individual has a unique truncation pattern which is neither simple left nor right truncation, has not previously been explored in the literature.
-Furthermore, I explore a Bayesian framework where the amount of explicit data augmentation is limited allowing the method to scale to the large number of episodes present in CIS without requiring vast computing resources (such as a high-performance cluster) which are unavailable within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
-\item  In section~\ref{cpt:sec:parameters-priors} I discuss parameterisation of this model and the challenges of specifying a vague prior.
-Whether a prior is vague or not depends on the parameterisation chosen, and for independent priors on the parameters, it is not possible to be vague on all sensible parameterisations.
-I also discuss the appeal of informative priors, and how the estimates in the previous chapter could be incorporated.
-\item  Section~\ref{cpt:sec:simulation-study} is a simulation study showing the applicability of this method to the CIS.
-I show that the model developed previously is able to estimate the duration distribution in this setting under vague or informative priors.
-However, I also show that, within the simulation considered, only the survival, and not the hazard, function are identifiable.
-\item  Section~\ref{cpt:sec:conclusion} concludes the chapter, discussing the limitations of this model. The following chapter will motivate and consider relaxing the perfect testing assumption.
-\end{itemize}
-
-\section{Introduction}\label{introduction}
-
-This paper discusses a model for observing duration of PCR positivity in
-the Coronavirus (COVID-19) Infection Survey (CIS). The CIS continually
-enrols individuals and tests them on an ongoing basis, following a
-specified testing protocol. The protocol specifies that the initial five
-tests are spaced every 7 days, then the spacing is every 28 days.
-However, real-world considerations often lead to variations in test
-times and missed tests. Our analysis focuses solely on the binary result
-(positive or negative) of these tests, assuming no misclassification
-bias
 
@@
 
-\section{Data}\label{data}
+This chapter proceeds as follows.
+\begin{itemize}
+\item  I introduce the motivating dataset in section~\ref{cpt:sec:data}. This is the CIS dataset previously encountered, processed to form "episodes" of infection.
+\item  In section~\ref{cpt:sec:model}, I develop a model accounting for double interval censoring and truncation.
+This builds on previous work in both biomedical and bird nesting studies.
+The arbitrary nature of truncation in CIS, where each individual has a unique truncation pattern which is neither simple left nor right truncation, has not previously been explored in the literature.
+Furthermore, I explore a Bayesian framework where the amount of explicit data augmentation is limited allowing the method to scale to the large number of episodes present in CIS without requiring vast computing resources (such as a high-performance cluster) which are unavailable within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
+\item  In section~\ref{cpt:sec:parameters-priors} I discuss parameterisation of the survival function and the challenges of specifying a vague prior.
+Whether a prior is vague or not depends on the parameterisation chosen, and for independent priors on the parameters, it is not possible to be vague on all sensible parameterisations.
+I also discuss the appeal of informative priors, and how the estimates in the previous chapter could be incorporated.
+\item  Section~\ref{cpt:sec:simulation-study} is a simulation study showing the applicability of this method to the CIS.
+I show that the model developed previously is able to estimate the duration distribution in this setting under vague or informative priors.
+However, I also show that, within the simulation considered, only the survival, and not the hazard, function are identifiable.
+\item  Section~\ref{cpt:sec:conclusion} concludes the chapter, discussing the limitations of this model. The following chapter will motivate and consider relaxing the perfect testing assumption.
+\end{itemize}
+
+\section{Data}\label{cpt:sec:data}
+
+The CIS continually enrols individuals and tests them on an ongoing basis, following a
+specified testing protocol. The protocol specifies that the initial five
+tests are spaced every 7 days, then the spacing is every 28 days.
+However, real-world considerations often lead to variations in test
+times and missed tests. Our analysis focuses solely on the binary result
+(positive or negative) of these tests, assuming no misclassification
+bias
 
@@
 
-\section{Notation and assumptions}\label{notation-and-assumptions}
+\section{Modelling the duration}\label{cpt:sec:model}
 
@@
 
-\section{Individual model}\label{individual-model}
+\subsection{Individual model}\label{cpt:sec:individual-model}
 
-\subsection{Likelihood}\label{likelihood}
+\subsubsection{Likelihood}\label{cpt:sec:likelihood}
 
@@
 observations associated with episode $i$ (graphically shown in Figure
-\ref{fig:partitionSpace}). Each episode, whether observed or not, must
+\ref{cpt:fig:partitionSpace}). Each episode, whether observed or not, must
 fall into one of these three classes with a probability that is a
-function of $\theta$ (see section \ref{sec:deriv-probs}).
+function of $\theta$ (see section \ref{cpt:sec:deriv-probs}).
 
@@
 negative infection times) is $\Omega_i^C$ and unshaded region is
-$\beta_i$.\label{fig:partitionSpace}}
+$\beta_i$.\label{cpt:fig:partitionSpace}}
 \end{figure}
@@
 
-In the following sections, we derive the expressions for $p_{ia}$ and
+In the remainder of the section, we derive the expressions for $p_{ia}$ and
 $p_{it}$ in terms of the data, and analytical solutions for the sum
@@
 
-\subsubsection{Solution for the sum}\label{solution-for-the-sum}
-
-This section is concerned with finding analytical solutions to
-$\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i)$ under two different
-assumptions for $p(n_i)$. First, we consider the improper prior
-$p(n_i) \propto 1/n_i$ which has several convenient properties. Then,
-we consider a negative binomial distribution (including the geometric
-and Poisson distributions as a special and limiting case respectively).
+First, I derive an analytical solution to $\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i)$ under two different assumptions for $p(n_i)$.
+First, we consider the improper prior $p(n_i) \propto 1/n_i$, this prior has several attractive properties.
+Then, we consider a negative binomial distribution (which includes the geometric and Poisson distributions as a special and limiting case respectively).
 
@@
 
-\subsubsection{Derivations of probabilities} \label{sec:deriv-probs}
+Next, I derive $p_{ia}$ and $p_{it}$.
 
@@
 
-\subsection{Full posterior}\label{full-posterior}
+\subsubsection{Full posterior}\label{cpt:sec:full-posterior}
 
@@
 
-\section{Total model}\label{total-model}
+\subsection{Total model}\label{cpt:sec:total-model}
 
@@
 
-\section{False negatives}\label{false-negatives}
+\section{Parameterisation and priors for the survival function} \label{cpt:sec:parameters-priors}
 
-In this section, we relax the assumption of perfect testing to allow
-some false negatives. We use a fairly simplified model to ensure the
-likelihood is tractable. This requires modifying both $p_{ia}$ to
-allow for the episode possibly being longer than observed, and
-$p_{it}$ to allow for additional episodes being missed.
+\section{Simulation study} \label{cpt:sec:simulation-study}
 
-Specifically, we introduce a constant test sensitivity $p_\text{sens}$
-such that:
-\begin{align}
-\prob(y_i(t) = 1) = \begin{cases} p_\text{sens} &b_i \leq t \leq e_i \\ 0 &\text{otherwise.}\end{cases}
-\end{align}
-
-In practice, it is likely that the test sensitivity is a function of the
-viral load of an individual (the concentration of viral material in the
-part of the body being sampled). For this initial attempt at modelling
-the sensitivity, we ignore this complication.
-
-\subsection{Modifying $p_{ia}$} \label{modifying-p_ia}
-
-The idea when modifying $p_{ia}$ is to allow for the negative test
-following the last positive to be either a true or false negative. If it
-is a true negative, then we have a bound on the length of the episode.
-If it is a false negative, then we consider the episode's length
-right-censored. The likelihood contribution for an episode is then a
-mixture of these two scenarios, with the mixture probability determined
-by the test sensitivity.
-
-For tractability, we assume that the tests bounding the start of the
-infection ($l_i^{(b)}$ and $r_i^{(b)}$) are true results and hence
-$b_i \in [l_i^{(b)}, r_i^{(b)}]$. Furthermore, we consider the tests
-only between $r_i^{(b)}$ and $l_i^{(e)}$ inclusive (the positive
-tests providing a lower bound on the length of the episode). Denote
-these tests by
-$t'_i = \{ t \in t_i : r_i^{(b)} \leq t \leq l_i^{(e)} \}$ and their
-results by $y_i'$. We know that the test results at times in $t_i'$
-are either true positives or false negatives. By definition, the test at
-$r_i^{(e)}$ is a false negative if and only if $e_i > r_i^{(e)}$. We
-proceed by considering the two cases.
-
-First, if $e_i \leq r_i^{(e)}$. In this case, the test at
-$r_i^{(e)}$ is a true negative, as are all other tests not in
-$t_i'$, and these occur with probability 1.
-\begin{align}
-&p(y_i', e_i \leq r_i^{(e)} | b_i, p_\text{sens}, \theta) \\
-&= p(y_i', l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\ % &\text{as no false positives}
-&= p(y_i' \mid l_i^{(e)} \leq e_i \leq r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
-&= \left( \prod_{t \in t_i'} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) \right)
-\end{align}
-
-Second, if $e_i > r_i^{(e)}$. In this case, the test at
-$r_i^{(e)}$ is a false negative, occurring with probability
-$(1 - p_\text{sens})$, and we have no upper bound on the end of the
-infection.
-\begin{align}
-&p(y_i', e_i > r_i^{(e)} | b_i, p_\text{sens}, \theta) \\
-&= p(y_i \mid e_i > r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(e_i > r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
-&= \left( \prod_{t \in t_i'} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) (1 - p_\text{sens}) S_\theta(r_i^{(e)} - b_i - 1)
-\end{align}
-
-Combining the above, the replacement for $p_{ia}$ is:
-\begin{align}
-p_{ia}'
-&= p(y_i' \mid p_\text{sens}, \theta) \\
-&= \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( p(y_i', b_i, e_i \leq r_i^{(e)} \mid p_\text{sens}, \theta) p(y_i', b_i e_i > r_i^{(e)} \mid p_\text{sens}, \theta) \right) p(b_i \mid p_\text{sens}, \theta) \\
-&= \left( \prod_{t \in t_i'} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \\ & \ \times \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) + (1 - p_\text{sens}) S_\theta(r_i^{(e)} - b_i - 1) \right) p(b_i \mid p_\text{sens}, \theta) \\
-&= \left( \prod_{t \in t_i'} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right)\sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - p_\text{sens} S_\theta(r_i^{(e)} - b_i - 1) \right) p(b_i \mid p_\text{sens}, \theta).
-\end{align}
-Note that if $p_\text{sens} = 1$ then $p_{ia}' = p_{ia}$.
-
-\subsection{Modifying $p_{it}$} \label{modifying-p_it}
-
-With false negatives, an episode with can be truncated in the following
-ways:
-
-\begin{enumerate}
-\item
-  The episode begins before $\min(t_i)$ (assuming there is a
-  negligible probability of a false negative test at this point).
-\item
-  The duration of the episode is less than $t_{ib_i}^N$. These first
-  two mechanisms are how truncation previously occurred, with
-  probability $p_{it}$.
-\item
-  The duration of the episode is at least as long as $t_{it}^N$ but
-  less than the time to the second test after $t$, $t_{it}^{2N}$,
-  and a false negative episode occurred at the first test after $t$.
-  This occurs with probability:
-  \begin{math}
-    (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(t_i)}^T \left( S_\theta(t_{ib}^N + 1) - S_\theta(t_{ib}^{2N} + 1)\right).
-  \end{math}
-\item
-  The duration of the episode is at least as long as $t_{it}^{2N}$,
-  and all the tests within the episode (at least two) are false
-  negatives. We assume this occurs with negligible probability.
-\end{enumerate}
-
-Hence, the replacement for $1 - p_{it}$ is:
-\begin{align}
-1 - p_{it}'
-&= 1 - p_{it} - (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(t_i)}^T \left( S_\theta(t_{ib}^N + 1) - S_\theta(t_{ib}^{2N} + 1)\right) \\
-&= \frac{1}{T} \sum_{b=\min(t_i)}^T \left( p_\text{sens} S_\theta(t_{ib}^N + 1) + (1 - p_\text{sens}) S_\theta(t_{ib}^{2N} + 1)\right) \\
-\end{align}
+\section{Conclusion} \label{cpt:sec:conclusion}
 

### 2023-05-25 cis-perfect-testing.tex
Move text from sim study doc across

@@
 \maketitle
+
+\listoftodos
+
+\pagebreak
 
@@
 $t^N - t$ days long, where $t^N$ is the time of their next test.
+\todo[inline]{Add figures demonstrating the censoring and truncation here}
 
@@
 ``individual'' model and the ``total'' model.
+\todo[inline]{\textbf{Should I cut one of the models?}}
 
-The individual model considers only individuals with observed (i.e., not
-truncated) episodes. This approach is similar to methods used in
-previous studies, such as \textcite{heiseyModelling},
-\textcite{dempsterMaximum}, and
-\textcite{turnbullEmpirical}. The conceptual
-approach here is to form a cohort of individuals for which we observe at
-least one positive test, and then consider a study which had only
-enrolled these individuals. The modelling approach then corrects for
-this selection bias.
-
-In contrast, the total model also takes into account data from
-individuals without any observed episodes. These individuals may have
-truncated episodes or may not have been infected at all. This approach
-utilises their test times to estimate the probability of truncated
-infections, ultimately providing an interpretable estimate of the number
-of truncated episodes in the cohort.
+\todo[inline]{Need to think about what of the below belongs here in the introduction, and turn these parts into a proper literature review}
+\todo[inline]{\textbf{Is it better to have a large literature review here, or to discuss relevant literature in each section (the papers on survival analysis in the model section, the papers on priors in that section, COVID estimates where I use them, etc.)?}}
 
@@
 \begin{itemize}
-\item  I introduce the motivating dataset in section~\ref{cpt:sec:data}. This is the CIS dataset previously encountered, processed to form "episodes" of infection.
-\item  In section~\ref{cpt:sec:model}, I develop a model accounting for double interval censoring and truncation.
+\item  I introduce the motivating dataset in section~\ref{perf-test:sec:data}. This is the CIS dataset previously encountered, processed to form "episodes" of infection.
+\item  In section~\ref{perf-test:sec:model}, I develop a model accounting for double interval censoring and truncation.
 This builds on previous work in both biomedical and bird nesting studies.
@@
 Furthermore, I explore a Bayesian framework where the amount of explicit data augmentation is limited allowing the method to scale to the large number of episodes present in CIS without requiring vast computing resources (such as a high-performance cluster) which are unavailable within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
-\item  In section~\ref{cpt:sec:parameters-priors} I discuss parameterisation of the survival function and the challenges of specifying a vague prior.
+\item  In section~\ref{perf-test:sec:parameters-priors} I discuss parameterisation of the survival function and the challenges of specifying a vague prior.
 Whether a prior is vague or not depends on the parameterisation chosen, and for independent priors on the parameters, it is not possible to be vague on all sensible parameterisations.
 I also discuss the appeal of informative priors, and how the estimates in the previous chapter could be incorporated.
-\item  Section~\ref{cpt:sec:simulation-study} is a simulation study showing the applicability of this method to the CIS.
+\item  Section~\ref{perf-test:sec:simulation-study} is a simulation study showing the applicability of this method to the CIS.
 I show that the model developed previously is able to estimate the duration distribution in this setting under vague or informative priors.
 However, I also show that, within the simulation considered, only the survival, and not the hazard, function are identifiable.
-\item  Section~\ref{cpt:sec:conclusion} concludes the chapter, discussing the limitations of this model. The following chapter will motivate and consider relaxing the perfect testing assumption.
+\item  Section~\ref{perf-test:sec:conclusion} concludes the chapter, discussing the limitations of this model. The following chapter will motivate and consider relaxing the perfect testing assumption.
 \end{itemize}
 
-\section{Data}\label{cpt:sec:data}
+\section{Problem description}\label{perf-test:sec:data}
 
-The CIS continually enrols individuals and tests them on an ongoing basis, following a
-specified testing protocol. The protocol specifies that the initial five
-tests are spaced every 7 days, then the spacing is every 28 days.
-However, real-world considerations often lead to variations in test
-times and missed tests. Our analysis focuses solely on the binary result
-(positive or negative) of these tests, assuming no misclassification
-bias
+Consider a series of tests on the same individuals (longitudinal data), where some individuals have a series of positive tests.
+We wish to estimate the distribution of time for which an individual is positive for.
+An individual could have multiple infections, in which case we refer to "infection episodes", where each episode corresponds to a single infection event.
 
-Episodes are taken from the data extract of the 3rd Jan 2023, as
-computed by SARAH using her updated episode definition. Only tests taken
-as part of CIS are included (ie: no pillar 2 testing). The following
-criteria are applied.
+I define an individual as \emph{detectable} if they would test positive if tested, in the absence of misclassification error (which is dealt with in \autoref{imperf-test}).
+In this chapter I will assume that the time between infection and becoming detectable is negligible.
+The episode's \emph{duration of positivity} is the number of days that they are detectable for; I use time discretised to days and hence this is at least 1 day.
+Note that the relationship between positivity and infectiousness is complex\todo{ref for PCR pos vs infectiousness} and beyond the scope of this thesis.
+An episode is \emph{detected} if they test positive at least once; episode $i$ is detected at time $t$ if and only if: $i$ is detectable at $t$, $i$ is tested at $t$, and the test returns positive (a false negative does not occur).
 
+I take a survival analysis approach to analysing this data.
+Survival analysis is the area of statistics which estimate the distribution of times between events.
+It is broadly applicable to many domains, such as the time-to-failure for a mechanical system or the effect of a treatment on time spent in hospital.
+In this context, the distribution to estimate is that of duration of positivity.
+
+The reason for this approach, over modelling biomarkers (as in chapter\todo{ref chapter on ATACCC}) is...
+
+The methods in this chapter deal with modelling two issues that arise from the data. False negatives are not dealt with and left to chapter X.
+
+The first issue is double interval censoring.
+Explain issue, brief overview of literature.
+
+The second issue is truncation.
+
+These combination of issues are rarely encountered in biomedical data (ref).
+
+I use the CIS data described in section\todo{ref description of CIS data}, with positive tests heuristically divided into episodes.
+The episode definition was developed by Sarah Walker and is as follows~\citePersonalComms:
+\todo[inline]{Insert episode definition}
+
+I filter the episodes as follows.
 \begin{itemize}
@@
 \item
-  A negative test following the final positive in the episode has been
-  recorded.
+  A negative test to end the episode (ie: following the final positive in the episode) has been recorded.
 \end{itemize}
 
-\section{Modelling the duration}\label{cpt:sec:model}
+\todo[inline]{Descriptive analyses of the data here}
 
+\section{Modelling the duration}\label{perf-test:sec:model}
+
+The individual model considers only individuals with observed (i.e., not truncated) episodes.
+This approach is similar to methods used in previous studies, such as \textcite{heiseyModelling}, \textcite{dempsterMaximum}, and \textcite{turnbullEmpirical}.
+The conceptual approach here is to form a cohort of individuals for which we observe at least one positive test, and then consider a study which had only enrolled these individuals.
+The modelling approach then corrects for this selection bias.
+
+In contrast, the total model also takes into account data from individuals without any observed episodes.
+These individuals may have truncated episodes or may not have been infected at all.
+This approach utilises their test times to estimate the probability of truncated infections, ultimately providing an interpretable estimate of the number of truncated episodes in the cohort.
+
+\subsection{Notation and assumptions}
+\todo[inline]{Some of this will possibly be defined elsewhere, such as in the introduction where CIS is introduced}
 \begin{itemize}
-\item
-  We initially assume perfect testing (no misclassification), although
-  later relax this assumption.
 \item
@@
 
-\subsection{Individual model}\label{cpt:sec:individual-model}
+\subsection{Individual model}\label{perf-test:sec:individual-model}
 
-\subsubsection{Likelihood}\label{cpt:sec:likelihood}
+\subsubsection{Likelihood}\label{perf-test:sec:likelihood}
 
@@
 observations associated with episode $i$ (graphically shown in Figure
-\ref{cpt:fig:partitionSpace}). Each episode, whether observed or not, must
+\ref{perf-test:fig:partitionSpace}). Each episode, whether observed or not, must
 fall into one of these three classes with a probability that is a
-function of $\theta$ (see section \ref{cpt:sec:deriv-probs}).
+function of $\theta$ (see section \ref{perf-test:sec:likelihood}).
 
@@
 negative infection times) is $\Omega_i^C$ and unshaded region is
-$\beta_i$.\label{cpt:fig:partitionSpace}}
+$\beta_i$.\label{perf-test:fig:partitionSpace} \todo[inline]{Digitalise figure of regions.}}
 \end{figure}
@@
 
-\subsubsection{Full posterior}\label{cpt:sec:full-posterior}
+\subsubsection{Full posterior}\label{perf-test:sec:full-posterior}
 
@@
 
-\subsection{Total model}\label{cpt:sec:total-model}
+\subsection{Total model}\label{perf-test:sec:total-model}
 
@@
 
-\section{Parameterisation and priors for the survival function} \label{cpt:sec:parameters-priors}
+\section{Parameterisation and priors for the survival function} \label{perf-test:sec:parameters-priors}
 
-\section{Simulation study} \label{cpt:sec:simulation-study}
+\todo[inline]{Section is currently notes: expand into prose which can largely come from previous documents}
 
-\section{Conclusion} \label{cpt:sec:conclusion}
+Define hazard and survival (maybe this should be done previously?).
+
+The hazard is convenient to use because it is unconstrained except in the interval [0 ,1]; include REFs.
+Survival needs to be monotonic.
+
+Discuss literature on Beta Process priors (maybe this should be in a broader literature review part?).
+
+\subsection{Independent priors}
+
+Traditional uninformative priors, such as Beta(1, 1) or Beta(0.5, 0.5), are unsuitable.
+
+Instead propose Beta(0.1, 1.9), linking to previous literature for a point estimate and showing vague figures.
+
+\subsection{Smoothing priors}
+
+\subsection{Informative priors}
+
+ATACCC can be used to provide prior information for the CIS analysis.
+
+Intuition: ATACCC informs early part, CIS informs late part.
+
+Problem: do not believe late part of ATACCC analysis.
+
+First, transform ATACCC analsysis into a posterior over survival time.
+
+Second, approximate ATACCC analysis with multivariate normal distribution.
+Show that this is a good approximation.
+
+Finally, use a Beta Process to "discount" some of the information in ATACCC towards a vague prior (as in the previous section).
+
+\section{Simulation study} \label{perf-test:sec:simulation-study}
+
+\subsection{Setup}
+
+\subsubsection{Testing and infection times}
+
+\subsubsection{Duration of positivity}
+
+We require a ground truth for the distribution giving duration of positivity.
+The majority of this distribution is based on \autoref{ATACCC}'s estimates from the ATACCC data.
+However, the tail is modified based on Sarah Walker's (unpublished) work based on CIS data.
+ATACCC is thought to estimate accurately the majority of the distribution but not the tail due to a small sample size and model uncertainty.
+These curves and the combined curve are compared in figure \@ref(fig:duration-curves).
+Individual $i$'s duration of positivity, $d_i$, is then an independent draw from this distribution.
+
+Sarah uses survival analysis to estimate the duration from CIS data.
+The initiating event is assumed known as the time the episode was detected.
+The final event is assumed interval-censored between the time of the final positive test and the subsequent negative test, or right-censored if a negative test has not yet been observed.
+A flexible, spline-based form is assumed for the baseline survival function (@roystonSTPM; @roystonFlexible) with covariates introduced via proportional odds.
+
+We know of two biases introduced from this analysis: “There is a bias in estimating the clearance distribution because the analysis used to estimate how long a person stays positive only starts from their first positive test.
+Since (most) people will have become positive on an earlier day, this will bias the clearance curves downwards (making the estimates too short).
+However, there is another bias due to the survey missing positive episodes entirely if they are short.
+This means that our dataset has fewer short positive episodes than in the population as a whole, and that the sample used to run the survival analysis is biased towards people with longer positive episodes.
+This will bias the clearance curves upwards (making the estimates too long).”~\autocite{cisMethodsONS}.
+
+To form the duration distribution used in the simulation, we combine these two estimates.
+The first 30 days of the distribution is proportional to the ATACCC estimate, with the rest proportional to this CIS-based estimates.
+The CIS-based estimates are shifted 3 days in order to make this smooth; this counteracts the above bias of missing the start of the infection (for times above 30 days, the other bias, missing short infections, is negligible).
+Denote by $f_A(t)$ the distribution function estimated from ATACCC and $f_C(t)$ that from these CIS-based estimates.
+Define:
+$$
+f_S'(t) = \begin{cases}
+	f_A(t) &t \leq 30 \\
+	f_C(t-3) &t > 30
+\end{cases}
+$$
+Then the distribution used in the simulation is the normalised version of this: $f_S(t) = f'_S(t)/\sum_i f_S'(i)$.
+
+\todo[inline]{Figure with duration distributions}
+
+\subsubsection{Test results and episode summaries}
+\todo[inline]{Make notation here consistent with elsewhere}
+
+The first day individual $i$ can be positive is their day of infection, $t_i^\text{onset}$, the last day is $t_i^\text{final} = t_i^\text{onset} + d_i - 1$ where $d_i$ is the duration of positivity for individual $i$.
+Outside of this period, all tests are negative (ie: no false positives occur). 
+Any test between $t_i^\text{onset}$ and $t_i^\text{final}$ (inclusive) is positive with probability $p_{sens}$ (that is, each test is an independent Bernoulli trial).
+
+Each individual is then summarised into an "episode" (when working with CIS data, the same individual can have multiple episodes due to reinfections, these are split heuristically but a negligible number in the pre-Omicron period and hence ignored within the simulation).
+An episode is characterised by:
+
+- The total number of positive tests observed per individual, $n_i^\text{pos}$.
+- The time of an individual's first positive test $V'_i$.
+- The time of an individual's last negative test prior to $V_i'$, $E'_i$.
+- The time of an individual's last positive test, $L_i'$.
+- The time of an individual's first negative test after $L_i'$, $R_i'$.
+
+\subsubsection{Restrictions and sample size}
+
+When using the real testing schedules, we then filter the episodes to replicate the process used when working with the CIS data.
+First, we discard any episodes where $n_i^\text{pos}=0$ as they are never detected.
+We then apply the same criteria as that applied to the modelling dataset (see Section \@ref(design)).
+Finally, of those remaining episodes, we select a random subset to match the number of episodes in the simulation to the number of episodes in the modelling cohort.
+The last step is required because we infect everyone in the simulation which does not happen in reality.
+
+In the fixed testing schedules, there is no automatic choice of sample size.
+We simulate a large number of episodes, and then choose to keep 4800 episodes\todo{Check how many this is} where at least one positive test was returned.
+
+\subsection{Descriptive analyses of simulated data}
+
+\subsection{Results}
+
+\todo[inline]{Need to decide what exactly to include... There's lots of possible results but need to form some coherent story here!}
+
+\section{Conclusion} \label{perf-test:sec:conclusion}
 

### 2023-05-25 main.tex
Format TeX and working pipeline
Sort chapter structure
Start drawing together writings into appropriate sections
Move text from sim study doc across

@@
-\documentclass{article}
+\documentclass[a4paper]{report}
 \usepackage{graphicx}
-\graphicspath{{images/}}
 
@@
 \usepackage[T1]{fontenc}
+\usepackage{hyperref}
 \usepackage[utf8]{inputenc}
-\usepackage{textcomp} % provide euro and other symbols
+\usepackage{textcomp} % provide symbols
+\usepackage{todonotes}
+% \usepackage{microtype}
 
@@
 
-\usepackage{subfiles} % Best loaded last in the preamble
+\newcommand{\citePersonalComms}{(personal communication)}
 
@@
 \author{Joshua Blake}
-\date{2023}
+\date{\today}
+
+\usepackage[backend=biber,style=authoryear,uniquename=false]{biblatex}
+
+\usepackage{subfiles} % Best loaded last in the preamble
+\addbibresource{references.bib}
 
@@
 \maketitle
-abc
-\section{Chapter 1}
 
-\subfile{chapter1/chapter1}
+\chapter{Text to use}
 
-\begin{align}
-    \cov
-\end{align}
+\section{Data}
+The CIS continually enrols individuals and tests them on an ongoing basis, following a
+specified testing protocol. The protocol specifies that the initial five
+tests are spaced every 7 days, then the spacing is every 28 days.
+However, real-world considerations often lead to variations in test
+times and missed tests. Our analysis focuses solely on the binary result
+(positive or negative) of these tests, assuming no misclassification
+bias.
+
+The CIS (Coronavirus Infection Survey) is a longitudinal study on a representative sample of households.
+A full description of the study can be found in the supplementary materials of [Pouwels et al.
+2020](https://www.thelancet.com/journals/lanpub/article/PIIS2468-2667(20)30282-6/fulltext#supplementaryMaterial) or [the study protocol](https://www.ndm.ox.ac.uk/covid-19/covid-19-infection-survey/protocol-and-information-sheets).
+In brief, households are invited to the study from databases held by the ONS.
+The number invited was to satisfy various target number of individuals swabbing per fortnight; relevant to the discussion in this document is that the target increased from 2,500 by the end of May 2020 to 150,000 in October 2020.
+Within a household, all indvidiuals aged 2 and over were invited to participate.
+Once invited, an enrolement swab would be taken followed by 4 further weekly swabs (giving a total of 5 swabs on days 0, 7, 14, 21, 28 relative to enrolement) after which monthly swabs are taken.
+
+During the period we consider, the total cohort size expands due to the continuous recruitment into the study.
+Following this, the number then decreases and stabilises as those recruited to meet the October target transition from weekly onto monthly testing.
+See figure \@ref(fig:test-schedules) for further detail.
+
+\todo[inline]{Add some descriptive figures eg those in ~/COVID/ons-incidence/duration_estimation/reproducible-sims/reports/simulation-studies.pdf}
+
+\section{Producing duration from ATACCC}
+
+The ATACCC analysis produces a posterior distribution for some population-level parameters, $\phi$.
+Taking a posterior sample for this, $\phi^{(i)}$, we can then draw the parameters which determine the viral
+load trajectory for a random individual $j$, $\theta_j$, from which that individual's duration $d_j^{(i)}$ can be calculated ([see this Overleaf paper for further details](https://www.overleaf.com/9549594748zbvpgjgvmgsd)).
+By drawing $\theta_1, \theta_2, \dots, \theta_n$ from the distribution $p(\theta_j \mid\phi^{(i)})$, the density of the duration distribution, $f_A(t)$, can be
+calculated for the posterior sample $i$ as:
+$$
+f^{(i)}_A(t) = \frac{1}{n}\sum_{j=1}^n \mathbb{I} \left( t-0.5 < d_j^{(i)} < t+0.5 \right)
+$$
+Given $f_A(t)$, the cumulative density and hazard functions can be calculated.
+Repeating this for $N$ posterior samples of $\phi$ gives the posterior distribution.
+I have used $N = 1000$ and $n = 100,000$, 
+Since implementing this solution, I have found that Monte Carlo integration to calculate durations given $\theta_i$ is more accurate by avoiding discretisation errors.
+However, the results here do not use this for consistency.
+
+\chapter{Survival analysis with perfect testing} \label{perf-test}
+
+\subfile{cis-perfect-testing}
+
+\chapter{Survival analysis with misclassification} \label{imperf-test}
+
+\subfile{cis-imperfect-testing}
 

### 2023-05-26 cis-perfect-testing.tex
Keep evolving text

@@
 
-This chapter aims to estimate the duration of SARS-CoV-2 PCR positivity from CIS data if perfect testing can be assumed (ie: no misclassification bias).
+This chapter aims to estimate the duration of SARS-CoV-2 PCR positivity from CIS data if perfect testing can be assumed (i.e.: no misclassification bias).
 I take a survival analysis approach, considering that neither the start nor end time of positivity are known exactly (double interval censoring) and that infections may be missed (truncation), explained in the following paragraphs.
@@
 However, application in reality must account for misclassification bias, the subject of the following chapter.
-
-Two major challenges arise when dealing with this data: double interval
-censoring and left truncation. Double interval censoring arises because
-we can only bound the start and end of an infection by observing an
-individual's test results transitioning from negative to positive and
-vice versa. Left truncation occurs because an individual is infected at
-time $t$ is only observed if their infection duration is at least
-$t^N - t$ days long, where $t^N$ is the time of their next test.
-\todo[inline]{Add figures demonstrating the censoring and truncation here}
-
-To address these challenges, we employ two distinct approaches: the
-``individual'' model and the ``total'' model.
-\todo[inline]{\textbf{Should I cut one of the models?}}
-
-\todo[inline]{Need to think about what of the below belongs here in the introduction, and turn these parts into a proper literature review}
-\todo[inline]{\textbf{Is it better to have a large literature review here, or to discuss relevant literature in each section (the papers on survival analysis in the model section, the papers on priors in that section, COVID estimates where I use them, etc.)?}}
-
-Unlike most Bayesian methods (eg: \textcite{heBayesiana}, \textcite{heBayesian}, and \textcite{caoModeling}) we avoid augmenting the
-data with unobserved times of infection or any information on the
-truncated infections; this is due to the large number of infections
-(>4000) making this approach computationally expensive.
 
@@
 I define an individual as \emph{detectable} if they would test positive if tested, in the absence of misclassification error (which is dealt with in \autoref{imperf-test}).
-In this chapter I will assume that the time between infection and becoming detectable is negligible.
 The episode's \emph{duration of positivity} is the number of days that they are detectable for; I use time discretised to days and hence this is at least 1 day.
 Note that the relationship between positivity and infectiousness is complex\todo{ref for PCR pos vs infectiousness} and beyond the scope of this thesis.
-An episode is \emph{detected} if they test positive at least once; episode $i$ is detected at time $t$ if and only if: $i$ is detectable at $t$, $i$ is tested at $t$, and the test returns positive (a false negative does not occur).
+An episode is \emph{detected} when they test positive at least once, if this occurs; episode $i$ is detected at time $t$ if and only if: $i$ is detectable at $t$, $i$ is tested at $t$, and the test returns positive (a false negative does not occur).
 
@@
 In this context, the distribution to estimate is that of duration of positivity.
+The \emph{initiating event} is when the duration starts, here becoming detectable.
+The \emph{terminating event} is when the duration starts, here becoming no longer detectable.
 
@@
 
-The methods in this chapter deal with modelling two issues that arise from the data. False negatives are not dealt with and left to chapter X.
+The methods in this chapter deal with modelling two issues that arise from the data.
+False negatives are not dealt with and left to chapter X.
 
 The first issue is double interval censoring.
-Explain issue, brief overview of literature.
+An event (either the initiating or terminating) is \emph{interval censored} if its value is only to lie within a range.
+\emph{Double interval censoring} means both the initiating and the terminating events are interval censored.
+The initiating event is censored because we only observe the change from being not detectable (i.e.: negative) at one test, to detectable (i.e.: positive) at the following test; at the terminating event we see the opposite occurring (see\todo{interval censoring figure}).
+
+Interval censoring is a well-studied area in the biostatistics and broader statistical literature\todo{interval censoring refs}.
+Double interval censoring is less frequent, although has been considered previously\todo{double interval censoring refs}.
 
 The second issue is truncation.
+Truncation occurs because we do not observe some episodes: an episode can occur between tests and hence the episode is never detected (see figure\todo{truncation figure}).
+We do not know how
 
-These combination of issues are rarely encountered in biomedical data (ref).
+This combination of issues are rarely encountered in biomedical data\todo{refs to combining double interval censoring with truncation}.
 
@@
 \section{Modelling the duration}\label{perf-test:sec:model}
+Now I develop a model accounting for double interval censoring and truncation, building on prior work in both biomedical and bird nesting studies.
+The arbitrary nature of truncation in CIS, where each individual has a unique truncation pattern and is neither simple left nor right truncation, has not previously been explored in the literature.
+Furthermore, I explore a Bayesian framework where the amount of explicit data augmentation is limited allowing the method to scale to the large number of episodes present in CIS without requiring vast computing resources (such as a high-performance cluster).
+Only limited resources are unavailable within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
+Therefore, unlike most Bayesian methods (eg: \textcite{heBayesiana}, \textcite{heBayesian}, and \textcite{caoModeling}) we avoid augmenting the data with unobserved times of infection or any information on the truncated infections; this is due to the large number of infections (>4000 in the time period considered) making this approach computationally expensive.
 
@@
 \begin{figure}
-\centering
 \includegraphics[width=\textwidth]{cis-perfect-testing/images/likelihood_areas.jpg}
@@
 the frequentist likelihood-based approach, up to the priors on other
-parameters \cites[section 4.2]{dempsterMaximum}{heiseyModelling}[section 8.7.5]{gelmanBayesian}, and is
+parameters~\cites[section 4.2]{dempsterMaximum}{heiseyModelling}[section 8.7.5]{gelmanBayesian}, and is
 the reference prior for this quantity\autocite{heBayesiana}. Under this prior,
@@
 
+In this section, I consider the various parameterisation of the survival function that are available.
+This has implications for the choice of prior  it is not possible to be vague on all sensible parameterisations.
+Informative priors might also be attractive, and allow the incorporation of the estimates from \autoref{ATACCC}.
+
 Define hazard and survival (maybe this should be done previously?).
 
-The hazard is convenient to use because it is unconstrained except in the interval [0 ,1]; include REFs.
+The hazard is convenient to use because it is unconstrained except in the interval [0, 1]\todo{ref where been used previously}.
 Survival needs to be monotonic.
 
-Discuss literature on Beta Process priors (maybe this should be in a broader literature review part?).
+The main building basis of the priors I propose is the Beta Process\todo{Beta Process refs}.
+The Beta Process is...
 
@@
 
-Instead propose Beta(0.1, 1.9), linking to previous literature for a point estimate and showing vague figures.
+Instead, I propose Beta(0.1, 1.9), linking to previous literature for a point estimate and showing vague figures.
 
@@
 
-First, transform ATACCC analsysis into a posterior over survival time.
+First, transform ATACCC analysis into a posterior over survival time.
 
@@
 
-Finally, use a Beta Process to "discount" some of the information in ATACCC towards a vague prior (as in the previous section).
+Finally, use a Beta Process to discount some information in ATACCC towards a vague prior (as in the previous section).
 
@@
 ATACCC is thought to estimate accurately the majority of the distribution but not the tail due to a small sample size and model uncertainty.
-These curves and the combined curve are compared in figure \@ref(fig:duration-curves).
+These curves and the combined curve are compared in \autoref{perf-test:fig:duration-dist}.
 Individual $i$'s duration of positivity, $d_i$, is then an independent draw from this distribution.
+
+\begin{figure}
+  \todo[inline]{Find figure comparing different duration distributions}
+  \caption{Comparison of the duration distributions, see main text for details of how each is derived. \label{perf-test:fig:duration-dist}}
+\end{figure}
 
@@
 The final event is assumed interval-censored between the time of the final positive test and the subsequent negative test, or right-censored if a negative test has not yet been observed.
-A flexible, spline-based form is assumed for the baseline survival function (@roystonSTPM; @roystonFlexible) with covariates introduced via proportional odds.
+A flexible, spline-based form is assumed for the baseline survival function~\autocite{roystonSTPM,roystonFlexible} with covariates introduced via proportional odds.
 
-We know of two biases introduced from this analysis: “There is a bias in estimating the clearance distribution because the analysis used to estimate how long a person stays positive only starts from their first positive test.
+We know of two biases introduced from this analysis.
+\enquote{There is a bias in estimating the clearance distribution because the analysis used to estimate how long a person stays positive only starts from their first positive test.
 Since (most) people will have become positive on an earlier day, this will bias the clearance curves downwards (making the estimates too short).
@@
 This means that our dataset has fewer short positive episodes than in the population as a whole, and that the sample used to run the survival analysis is biased towards people with longer positive episodes.
-This will bias the clearance curves upwards (making the estimates too long).”~\autocite{cisMethodsONS}.
+This will bias the clearance curves upwards (making the estimates too long).}~\autocite{cisMethodsONS}.
 
@@
 
-\todo[inline]{Figure with duration distributions}
-
 \subsubsection{Test results and episode summaries}
@@
 The first day individual $i$ can be positive is their day of infection, $t_i^\text{onset}$, the last day is $t_i^\text{final} = t_i^\text{onset} + d_i - 1$ where $d_i$ is the duration of positivity for individual $i$.
-Outside of this period, all tests are negative (ie: no false positives occur). 
+Outside this period, all tests are negative (ie: no false positives occur). 
 Any test between $t_i^\text{onset}$ and $t_i^\text{final}$ (inclusive) is positive with probability $p_{sens}$ (that is, each test is an independent Bernoulli trial).
@@
 
-In the fixed testing schedules, there is no automatic choice of sample size.
-We simulate a large number of episodes, and then choose to keep 4800 episodes\todo{Check how many this is} where at least one positive test was returned.
-
 \subsection{Descriptive analyses of simulated data}

### 2023-05-26 main.tex
Keep evolving text

@@
 \usepackage{graphicx}
+\usepackage[utf8]{inputenc}
 
@@
 \usepackage{bm}
+\usepackage[style=british]{csquotes}
 \usepackage[T1]{fontenc}
 \usepackage{hyperref}
-\usepackage[utf8]{inputenc}
 \usepackage{textcomp} % provide symbols
@@
 
-\chapter{Text to use}
-
-\section{Data}
-The CIS continually enrols individuals and tests them on an ongoing basis, following a
-specified testing protocol. The protocol specifies that the initial five
-tests are spaced every 7 days, then the spacing is every 28 days.
-However, real-world considerations often lead to variations in test
-times and missed tests. Our analysis focuses solely on the binary result
-(positive or negative) of these tests, assuming no misclassification
-bias.
-
-The CIS (Coronavirus Infection Survey) is a longitudinal study on a representative sample of households.
-A full description of the study can be found in the supplementary materials of [Pouwels et al.
-2020](https://www.thelancet.com/journals/lanpub/article/PIIS2468-2667(20)30282-6/fulltext#supplementaryMaterial) or [the study protocol](https://www.ndm.ox.ac.uk/covid-19/covid-19-infection-survey/protocol-and-information-sheets).
-In brief, households are invited to the study from databases held by the ONS.
-The number invited was to satisfy various target number of individuals swabbing per fortnight; relevant to the discussion in this document is that the target increased from 2,500 by the end of May 2020 to 150,000 in October 2020.
-Within a household, all indvidiuals aged 2 and over were invited to participate.
-Once invited, an enrolement swab would be taken followed by 4 further weekly swabs (giving a total of 5 swabs on days 0, 7, 14, 21, 28 relative to enrolement) after which monthly swabs are taken.
-
-During the period we consider, the total cohort size expands due to the continuous recruitment into the study.
-Following this, the number then decreases and stabilises as those recruited to meet the October target transition from weekly onto monthly testing.
-See figure \@ref(fig:test-schedules) for further detail.
-
-\todo[inline]{Add some descriptive figures eg those in ~/COVID/ons-incidence/duration_estimation/reproducible-sims/reports/simulation-studies.pdf}
+\chapter{Duration from biomarkers} \label{ATACCC}
 
@@
 
+\appendix
+
+\chapter{Text to use}
+
+Random bits of text that I've written that don't currently fit anywhere but will likely be useful in future
+
+\section{CIS study description}
+The CIS continually enrols individuals and tests them on an ongoing basis, following a
+specified testing protocol. The protocol specifies that the initial five
+tests are spaced every 7 days, then the spacing is every 28 days.
+However, real-world considerations often lead to variations in test
+times and missed tests. Our analysis focuses solely on the binary result
+(positive or negative) of these tests, assuming no misclassification
+bias.
+
+The CIS (Coronavirus Infection Survey) is a longitudinal study on a representative sample of households.
+A full description of the study can be found in the supplementary materials of \textcite{pouwelsCommunity} or the study protocol~\autoref{cisProtocol}.
+In brief, households are invited to the study from databases held by the ONS.
+The number invited was to satisfy various target number of individuals swabbing per fortnight; relevant to the discussion in this document is that the target increased from 2,500 by the end of May 2020 to 150,000 in October 2020.
+Within a household, all individuals aged 2 and over were invited to participate.
+Once invited, an enrolment swab would be taken followed by 4 further weekly swabs (giving a total of 5 swabs on days 0, 7, 14, 21, 28 relative to enrolment) after which monthly swabs are taken.
+
+During the period we consider, the total cohort size expands due to the continuous recruitment into the study.
+Following this, the number then decreases and stabilises as those recruited to meet the October target transition from weekly onto monthly testing.
+See figure \@ref(fig:test-schedules) for further detail.
+
+\todo[inline]{Add some descriptive figures e.g. those in ~/COVID/ons-incidence/duration_estimation/reproducible-sims/reports/simulation-studies.pdf}
+
 \end{document}

### 2023-05-26 cis-perfect-testing.tex
Rearrange chapter

@@
 
-\title{Estimating duration using survival analysis}
-\author{Joshua Blake}
-\date{\today}
-
 \begin{document}
-\maketitle
-
-\listoftodos
-
-\pagebreak
+\chapter{Survival analysis with perfect testing} \label{perf-test}
 
@@
 However, application in reality must account for misclassification bias, the subject of the following chapter.
-
-This chapter proceeds as follows.
-\begin{itemize}
-\item  I introduce the motivating dataset in section~\ref{perf-test:sec:data}. This is the CIS dataset previously encountered, processed to form "episodes" of infection.
-\item  In section~\ref{perf-test:sec:model}, I develop a model accounting for double interval censoring and truncation.
-This builds on previous work in both biomedical and bird nesting studies.
-The arbitrary nature of truncation in CIS, where each individual has a unique truncation pattern which is neither simple left nor right truncation, has not previously been explored in the literature.
-Furthermore, I explore a Bayesian framework where the amount of explicit data augmentation is limited allowing the method to scale to the large number of episodes present in CIS without requiring vast computing resources (such as a high-performance cluster) which are unavailable within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
-\item  In section~\ref{perf-test:sec:parameters-priors} I discuss parameterisation of the survival function and the challenges of specifying a vague prior.
-Whether a prior is vague or not depends on the parameterisation chosen, and for independent priors on the parameters, it is not possible to be vague on all sensible parameterisations.
-I also discuss the appeal of informative priors, and how the estimates in the previous chapter could be incorporated.
-\item  Section~\ref{perf-test:sec:simulation-study} is a simulation study showing the applicability of this method to the CIS.
-I show that the model developed previously is able to estimate the duration distribution in this setting under vague or informative priors.
-However, I also show that, within the simulation considered, only the survival, and not the hazard, function are identifiable.
-\item  Section~\ref{perf-test:sec:conclusion} concludes the chapter, discussing the limitations of this model. The following chapter will motivate and consider relaxing the perfect testing assumption.
-\end{itemize}
 
@@
 \item
-  A negative test to end the episode (ie: following the final positive in the episode) has been recorded.
+  A negative test to end the episode (i.e.: following the final positive in the episode) has been recorded.
 \end{itemize}
@@
 Only limited resources are unavailable within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
-Therefore, unlike most Bayesian methods (eg: \textcite{heBayesiana}, \textcite{heBayesian}, and \textcite{caoModeling}) we avoid augmenting the data with unobserved times of infection or any information on the truncated infections; this is due to the large number of infections (>4000 in the time period considered) making this approach computationally expensive.
+Therefore, unlike most Bayesian methods (e.g.: \textcite{heBayesiana}, \textcite{heBayesian}, and \textcite{caoModeling}) we avoid augmenting the data with unobserved times of infection or any information on the truncated infections; this is due to the large number of infections (>4000 in the time period considered) making this approach computationally expensive.
 
@@
 define three sets which partition the space of possible infection and
-recovery (ie: are strict subsets of $T \times T$) given the
+recovery (i.e.: are strict subsets of $T \times T$) given the
 observations associated with episode $i$ (graphically shown in Figure
@@
 \includegraphics[width=\textwidth]{cis-perfect-testing/images/likelihood_areas.jpg}
+\todo[inline]{Digitalise figure of regions.}
 \caption{Admissible, $\alpha_i$, inadmissible, $\beta_i$ and
@@
 negative infection times) is $\Omega_i^C$ and unshaded region is
-$\beta_i$.\label{perf-test:fig:partitionSpace} \todo[inline]{Digitalise figure of regions.}}
+$\beta_i$.\label{perf-test:fig:partitionSpace}}
 \end{figure}
@@
 
-The prior $p(n_i) \propto 1/n_i$ means that the model coincides with
-the frequentist likelihood-based approach, up to the priors on other
-parameters~\cites[section 4.2]{dempsterMaximum}{heiseyModelling}[section 8.7.5]{gelmanBayesian}, and is
-the reference prior for this quantity\autocite{heBayesiana}. Under this prior,
+The prior $p(n_i) \propto 1/n_i$ means that the model coincides with the frequentist conditional likelihood approach, up to the priors on other parameters~\cites[section 4.2]{dempsterMaximum}{heiseyModelling}[section 8.7.5]{gelmanBayesian}.
+Furthermore, this is the reference prior for this quantity\autocite{heBayesiana}.
+Under this prior,
 $\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) \propto \sum_{n_i=1}^\infty p_{it}^{n_i-1} = 1/(1-p_{it})$.
@@
 The first day individual $i$ can be positive is their day of infection, $t_i^\text{onset}$, the last day is $t_i^\text{final} = t_i^\text{onset} + d_i - 1$ where $d_i$ is the duration of positivity for individual $i$.
-Outside this period, all tests are negative (ie: no false positives occur). 
+Outside this period, all tests are negative (i.e.: no false positives occur). 
 Any test between $t_i^\text{onset}$ and $t_i^\text{final}$ (inclusive) is positive with probability $p_{sens}$ (that is, each test is an independent Bernoulli trial).
@@
 
+\listoftodos
+
 \end{document}

### 2023-05-26 main.tex
Follow UoC formatting rules

@@
-\documentclass[a4paper]{report}
+\documentclass[a4paper,12pt,custombib,times]{PhDThesisPSnPDF} % add print option for printing
 \usepackage{graphicx}
@@
 \author{Joshua Blake}
+\supervisor{Prof.~D.~De Angelis \newline Dr.~P.~Birrell}
 \date{\today}
+\dept{MRC Biostatistics Unit}
+\university{University of Cambridge}
+\crest{\includegraphics[width=0.45\textwidth]{University_Crest_Long}}
+\college{Jesus College}
+\collegeshield{\includegraphics[width=0.2\textwidth]{jesus}}
+\degreetitle{Doctor of Philosophy}
 
@@
 
+\frontmatter
+
 \maketitle
+
+\begin{declaration}
+    This thesis is my own work and contains nothing which is the outcome of work done in collaboration with others except as specified in the text and Acknowledgements.
+    It is not substantially the same as any work that has already been submitted before for any degree or other qualification except as declared in the preface and specified in the text.
+    The thesis contains no more than 60,000 words (excluding figures, photographs, tables, appendices, and bibliography).
+\end{declaration}
+
+\begin{acknowledgements}      
+    \todo[inline]{Acknowledgements here}    
+\end{acknowledgements}
+
+\tableofcontents
+
+\listoffigures
+
+\mainmatter
 
@@
 However, the results here do not use this for consistency.
-
-\chapter{Survival analysis with perfect testing} \label{perf-test}
 
@@
 
-\todo[inline]{Add some descriptive figures e.g. those in ~/COVID/ons-incidence/duration_estimation/reproducible-sims/reports/simulation-studies.pdf}
+\todo[inline]{Add some descriptive figures e.g. those in \url{~/COVID/ons-incidence/duration_estimation/reproducible-sims/reports/simulation-studies.pdf}}
 

### 2023-05-30 cis-perfect-testing.tex
various editing and figure creation

@@
 We wish to estimate the distribution of time for which an individual is positive for.
-An individual could have multiple infections, in which case we refer to "infection episodes", where each episode corresponds to a single infection event.
+An individual could have multiple infections, in which case define \emph{infection episodes}, corresponding to a single infection event.
 
@@
 An episode is \emph{detected} when they test positive at least once, if this occurs; episode $i$ is detected at time $t$ if and only if: $i$ is detectable at $t$, $i$ is tested at $t$, and the test returns positive (a false negative does not occur).
+
+Series of positive tests within CIS were heuristically categorised into episodes by Sarah Walker (see \autoref{episode-def}).
+These tests are assumed to have been due to a single infection event.
 
@@
 The reason for this approach, over modelling biomarkers (as in chapter\todo{ref chapter on ATACCC}) is...
+\todo[inline]{Justify approach}
 
 The methods in this chapter deal with modelling two issues that arise from the data.
-False negatives are not dealt with and left to chapter X.
+False negatives are not dealt with and left to \autoref{imperf-test}.
 
 The first issue is double interval censoring.
-An event (either the initiating or terminating) is \emph{interval censored} if its value is only to lie within a range.
-\emph{Double interval censoring} means both the initiating and the terminating events are interval censored.
-The initiating event is censored because we only observe the change from being not detectable (i.e.: negative) at one test, to detectable (i.e.: positive) at the following test; at the terminating event we see the opposite occurring (see\todo{interval censoring figure}).
+An event is \emph{censored} if the time of the event can only be bounded; it is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
+Censoring is a very widely studied problem, and a central consideration in any survival analysis.
+Interval censoring arises in CIS because we only observe the change from being not detectable (i.e.: negative) at one test, to detectable (i.e.: positive) at the following test and vice versa; \textcite{bogaertsSurvival} review many statistical methods applicable to interval censoring (see \autoref{perf-test:fig:double-interval-censor}).
+\emph{Double censoring} means that both the originating and terminating event are censored, and is also a well-studied problem, reviewed by \textcite{sunStatisticala}.
+In CIS, we have \emph{double interval censoring} meaning that both the initiating and the terminating events are interval censored.
+\begin{figure}
+  \todo[inline]{Copy over figure demonstrating double interval censoring}
+  \caption{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval. \label{perf-test:fig:double-interval-censor}}
+\end{figure}
 
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{cis-perfect-testing/images/likelihood_areas.jpg}
-\todo[inline]{Digitalise figure of regions.}
+\includegraphics[width=\textwidth]{cis-perfect-testing/regions_diag}
 \caption{Admissible, $\alpha_i$, inadmissible, $\beta_i$ and

### 2023-05-30 cis-perfect-testing/regions_diag.R

@@ -1,0 +1,86 @@
+library(ggplot2)
+
+neg_test_dates = c(0, 7, 14, 56)
+pos_test_dates = c(28, 21)
+all_test_dates = sort(c(neg_test_dates, pos_test_dates))
+
+test_lines = function(values, result) {
+    if (result) {
+        label = "Positive"
+        colour = "Red"
+    } else {
+        label = "Negative"
+        colour = "Green"
+    }
+    list(
+        geom_vline(aes(colour = label, xintercept = !!values), linetype = "dotted"),
+        geom_hline(aes(colour = label, yintercept = !!values), linetype = "dotted")
+    )
+}
+
+shade_alpha = 0.3
+ggplot() +
+    test_lines(pos_test_dates, TRUE) +
+    test_lines(neg_test_dates, FALSE) +
+    theme_minimal() +
+    scale_x_continuous(breaks = all_test_dates, minor_breaks = -100:100, limits = c(-10, 60)) +
+    scale_y_continuous(breaks = all_test_dates, minor_breaks = -100:100, limits = c(0, 60)) +
+    labs(
+        x = "Infection time",
+        y = "Recovery time",
+        colour = "Test results",
+        fill = "Regions"
+    ) +
+    scale_colour_manual(
+        values = c(
+            "Positive" = "red",
+            "Negative" = "seagreen"
+        )
+    ) +
+    geom_rect(
+        aes(fill = "Truncated"),
+        xmin = -Inf,
+        xmax = 0,
+        ymin = -Inf,
+        ymax = Inf,
+        alpha = shade_alpha
+    ) +
+    purrr::map2(
+        all_test_dates + 1, dplyr::lead(all_test_dates) - 1,
+        ~geom_polygon(
+            aes(x = c(.x, .x, .y), y = c(.x, .y, .y), fill = "Truncated"),
+            alpha = shade_alpha,
+        )
+    ) +
+    geom_polygon(
+        aes(
+            fill = "Impossible",
+            x = c(0, 60, Inf),
+            y = c(0, 60, 0)
+        ),
+        alpha = shade_alpha
+    ) +
+    geom_rect(
+        aes(fill = "Impossible"),
+        xmin = 0,
+        xmax = Inf,
+        ymin = -Inf,
+        ymax = 0,
+        alpha = shade_alpha
+    ) +
+    geom_rect(
+        aes(fill = "Admissible"),
+        xmin = 15,
+        xmax = 21,
+        ymin = 28,
[…]

### 2023-05-30 cis-perfect-testing.tex
Make i.e. prettier

@@
 
-This chapter aims to estimate the duration of SARS-CoV-2 PCR positivity from CIS data if perfect testing can be assumed (i.e.: no misclassification bias).
+This chapter aims to estimate the duration of SARS-CoV-2 PCR positivity from CIS data if perfect testing can be assumed (\ie: no misclassification bias).
 I take a survival analysis approach, considering that neither the start nor end time of positivity are known exactly (double interval censoring) and that infections may be missed (truncation), explained in the following paragraphs.
@@
 I define an individual as \emph{detectable} if they would test positive if tested, in the absence of misclassification error (which is dealt with in \autoref{imperf-test}).
-The episode's \emph{duration of positivity} is the number of days that they are detectable for; I use time discretised to days and hence this is at least 1 day.
-Note that the relationship between positivity and infectiousness is complex\todo{ref for PCR pos vs infectiousness} and beyond the scope of this thesis.
+The episode's \emph{duration of positivity} is the number of days that they are detectable for; I use time discretised to days, therefore, the duration is at least 1 day.
+Note that the relationship between positivity and infectiousness is complex, \todo{ref for PCR pos vs infectiousness} and beyond the scope of this thesis.
+
 An episode is \emph{detected} when they test positive at least once, if this occurs; episode $i$ is detected at time $t$ if and only if: $i$ is detectable at $t$, $i$ is tested at $t$, and the test returns positive (a false negative does not occur).
 
-Series of positive tests within CIS were heuristically categorised into episodes by Sarah Walker (see \autoref{episode-def}).
-These tests are assumed to have been due to a single infection event.
-
-I take a survival analysis approach to analysing this data.
-Survival analysis is the area of statistics which estimate the distribution of times between events.
+I utilise survival analysis for analysing this data.
+Survival analysis is the area of statistics concerned with estimating the distribution of the times between events.
 It is broadly applicable to many domains, such as the time-to-failure for a mechanical system or the effect of a treatment on time spent in hospital.
-In this context, the distribution to estimate is that of duration of positivity.
-The \emph{initiating event} is when the duration starts, here becoming detectable.
-The \emph{terminating event} is when the duration starts, here becoming no longer detectable.
+In the context of this thesis, the distribution to estimate is that of duration of positivity.
+The \emph{initiating event} is when the duration starts, \ie becoming detectable.
+The \emph{terminating event} is when the duration starts, \ie becoming no longer detectable.
 
@@
 Censoring is a very widely studied problem, and a central consideration in any survival analysis.
-Interval censoring arises in CIS because we only observe the change from being not detectable (i.e.: negative) at one test, to detectable (i.e.: positive) at the following test and vice versa; \textcite{bogaertsSurvival} review many statistical methods applicable to interval censoring (see \autoref{perf-test:fig:double-interval-censor}).
+Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa; \textcite{bogaertsSurvival} review many statistical methods applicable to interval censoring (see \autoref{perf-test:fig:double-interval-censor}).
 \emph{Double censoring} means that both the originating and terminating event are censored, and is also a well-studied problem, reviewed by \textcite{sunStatisticala}.
@@
 I use the CIS data described in section\todo{ref description of CIS data}, with positive tests heuristically divided into episodes.
-The episode definition was developed by Sarah Walker and is as follows~\citePersonalComms:
-\todo[inline]{Insert episode definition}
+Prior work categorised a series of positive tests within the same individual CIS were heuristically categorised into episodes (see \autoref{episode-def}).
 
@@
 \item
-  A negative test to end the episode (i.e.: following the final positive in the episode) has been recorded.
+  A negative test to end the episode (\ie: following the final positive in the episode) has been recorded.
 \end{itemize}
@@
 
-The individual model considers only individuals with observed (i.e., not truncated) episodes.
+The individual model considers only individuals with observed (\ie, not truncated) episodes.
 This approach is similar to methods used in previous studies, such as \textcite{heiseyModelling}, \textcite{dempsterMaximum}, and \textcite{turnbullEmpirical}.
@@
 define three sets which partition the space of possible infection and
-recovery (i.e.: are strict subsets of $T \times T$) given the
+recovery (\ie: are strict subsets of $T \times T$) given the
 observations associated with episode $i$ (graphically shown in Figure
@@
 The first day individual $i$ can be positive is their day of infection, $t_i^\text{onset}$, the last day is $t_i^\text{final} = t_i^\text{onset} + d_i - 1$ where $d_i$ is the duration of positivity for individual $i$.
-Outside this period, all tests are negative (i.e.: no false positives occur). 
+Outside this period, all tests are negative (\ie: no false positives occur). 
 Any test between $t_i^\text{onset}$ and $t_i^\text{final}$ (inclusive) is positive with probability $p_{sens}$ (that is, each test is an independent Bernoulli trial).

### 2023-05-31 cis-perfect-testing.tex
Problem description

@@
 
-\section{Problem description}\label{perf-test:sec:data}
+\section{Problem description and approach}\label{perf-test:sec:data}
 
@@
 The episode's \emph{duration of positivity} is the number of days that they are detectable for; I use time discretised to days, therefore, the duration is at least 1 day.
-Note that the relationship between positivity and infectiousness is complex, \todo{ref for PCR pos vs infectiousness} and beyond the scope of this thesis.
+Note that the relationship between positivity (\ie the result of a PCR test) and infectiousness (\ie an individual's ability to pass on the virus) is complex, \todo{ref for PCR pos vs infectiousness} and beyond the scope of this thesis.
 
@@
 
-The reason for this approach, over modelling biomarkers (as in chapter\todo{ref chapter on ATACCC}) is...
-\todo[inline]{Justify approach}
+The reason for this approach, over modelling biomarkers (as in \autoref{ATACCC}) is due to two factors.
+First, the large sample size in CIS means that approaches where the number of parameters scales with the number of observed infections (as it does with random effect based approaches) would be computationally challenging; this challenge is compounded by the data being stored in a trusted research environment limiting the computational power and tools available.
+Second, the data is coarse with most episodes having only a single positive test; therefore, the trajectory of the viral load would not be well-identified in an individual.
+Combined, these factors outweighed the downside of survival analysis only considering the binary outcome (\ie positive or negative) or the test rather than the additional information on the viral load.
 
-The methods in this chapter deal with modelling two issues that arise from the data.
+The methods in this chapter deal with modelling two elements of the data-generating process, double interval censoring and truncation.
 False negatives are not dealt with and left to \autoref{imperf-test}.
 
-The first issue is double interval censoring.
+Double interval censoring is a special case of censoring, a common and well-studied challenge encountered when dealing with survival data.
 An event is \emph{censored} if the time of the event can only be bounded; it is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
-Censoring is a very widely studied problem, and a central consideration in any survival analysis.
 Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa; \textcite{bogaertsSurvival} review many statistical methods applicable to interval censoring (see \autoref{perf-test:fig:double-interval-censor}).
@@
 \begin{figure}
-  \todo[inline]{Copy over figure demonstrating double interval censoring}
-  \caption{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval. \label{perf-test:fig:double-interval-censor}}
+  \centering \includegraphics{cis-perfect-testing/double-interval-censor}
+  \caption{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval.
+Demonstrated here by a participant who transitions negative at time 7 to positive at time 14, bounding the start of the episode; similarly, the transition positive at time 28 to negative at time 56 bounds the end of the episode.
+\label{perf-test:fig:double-interval-censor}}
 \end{figure}
@@
 
-The second issue is truncation.
-Truncation occurs because we do not observe some episodes: an episode can occur between tests and hence the episode is never detected (see figure\todo{truncation figure}).
-We do not know how
+The second element is truncation.
+Truncation occurs because we do not observe some episodes: an episode can occur between tests and hence the episode is never detected (see \autoref{perf-test:fig:truncation}).
+We do not know how many infections we do not detect.
+Failure to model truncation will lead to estimating too long durations because the detected episodes are a biased sample with a longer duration.
+To see why they are shorter, consider the undetected infection in the bottom of \autoref{perf-test:fig:truncation}.
+This individual was infected at time 33 for 10 days.
+If instead there infection was 24 days (or longer), the individual would still have been detectable at their next test, and therefore would have been detected.
 
-This combination of issues are rarely encountered in biomedical data\todo{refs to combining double interval censoring with truncation}.
+Truncation is a large problem when individuals move to the infrequent testing schedule within CIS.
+Consider that, based on the estimates in \autoref{ATACCC}, the median length of an episode is 14 days \todo{check this number} we would not expect many episodes to last the 28 days that can occur between tests.
+\begin{figure}
+  \centering \includegraphics{cis-perfect-testing/truncation}
+  \caption{Episodes data in CIS is truncated, meaning that an unknown number of infections are never detected.
+Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was.
+See the main text for why this is an issue.
+\label{perf-test:fig:truncation}}
+\end{figure}
 
-I use the CIS data described in section\todo{ref description of CIS data}, with positive tests heuristically divided into episodes.
+This combination of censoring and truncation are rarely encountered in biomedical data\todo{refs to combining double interval censoring with truncation}.
+
+I use the CIS data described in \autoref{intro:sec:cis}, with positive tests heuristically divided into episodes.
 Prior work categorised a series of positive tests within the same individual CIS were heuristically categorised into episodes (see \autoref{episode-def}).
@@
 
+\section{Discussion}
+
+\todo[inline]{Add discussion after deciding on results}
+
 \section{Conclusion} \label{perf-test:sec:conclusion}

### 2023-05-31 cis-perfect-testing/dgp-challenges.R

@@ -1,0 +1,90 @@
+library(dplyr)
+library(ggplot2)
+library(tibble)
+library(tidyr)
+
+plot_testing_schedule = function(x, break_modifier = 0) {
+  scale_colours = c(
+      test = "black",
+      positive = "red",
+      negative = "seagreen",
+      infected = "blue",
+      recovered = "blue"
+  )
+  scale_colours = scale_colours[unique(x$type)]
+  scale_shapes = c(
+    test = 4,
+    positive = 4,
+    negative = 4,
+    infected = 20,
+    recovered = 20
+  )
+  scale_shapes = scale_shapes[unique(x$type)]
+  
+  inf_boxes = x %>% 
+    filter(type %in% c("infected", "recovered")) %>% 
+    pivot_wider(individual, names_from = type, values_from = time)
+  
+  x_breaks = c((0:3) * 7, (1:5) * 28) - break_modifier
+  
+  p = x %>% 
+    ggplot(aes(time, individual)) +
+    geom_point(aes(shape = type, colour = type), size = 3, stroke = 2) +
+    geom_hline(aes(yintercept = individual), colour = "grey") +
+    scale_y_discrete() +
+    scale_x_continuous(breaks = x_breaks, minor_breaks = NULL) +
+    scale_shape_manual(values = scale_shapes) +
+    scale_colour_manual(values = scale_colours) +
+    xlab("Time") +
+    theme_minimal() +
+    theme(
+      axis.title.y = element_blank(),
+      axis.text.y = element_blank(),
+      axis.ticks.y = element_blank(),
+      legend.title = element_blank(),
+      legend.position = "bottom",
+    #   axis.text = element_text(size = 18),
+    #   axis.title = element_text(size = 18),
+      panel.grid.major = element_line(colour = "black"),
+    #   legend.text = element_text(size = 18)
+    )
+  
+  if (nrow(inf_boxes) > 0) {
+    p = p +
+      geom_rect(aes(xmin = infected, xmax = recovered,
+                  ymin = individual - 0.5, ymax = individual + 0.5,
+                  x = NULL),
+              data = inf_boxes, fill = "red", alpha = 0.1)
+  }
+  
+  return(p)
+}
+
+p_censor = tibble(
+  time = c((0:3) * 7, (1:3) * 28),
+) %>% 
+  mutate(
+    type = if_else(time >= 10 & time <= 30,
[…]

### 2023-05-31 main.tex
Push chapter name into subfile
various editing and figure creation
Make i.e. prettier
Swap to new thesis template based on cam-thesis
Provides better formatting, particularly page numbering.
Has been kept up-to-date better with changed (relaxed) requirements.

Correct decleration

Finalise and credit new thesis template

--- 2023-05-26 main.tex
@@
-\documentclass[a4paper]{report}
+\documentclass{cam-thesis}
+\usepackage{setspace}
+\onehalfspacing
 \usepackage{graphicx}
@@
 % \usepackage{microtype}
+
+% Macros for common abbreviations to get the spacing right
+% See: https://stackoverflow.com/questions/3282319/correct-way-to-define-macros-etc-ie-in-latex
+\usepackage{xspace}
+\makeatletter
+\DeclareRobustCommand\onedot{\futurelet\@let@token\@onedot}
+\def\@onedot{\ifx\@let@token.\else.\null\fi\xspace}
+
+\def\eg{\emph{e.g}\onedot} \def\Eg{\emph{E.g}\onedot}
+\def\ie{\emph{i.e}\onedot} \def\Ie{\emph{I.e}\onedot}
+\def\cf{\emph{c.f}\onedot} \def\Cf{\emph{C.f}\onedot}
+\def\etc{\emph{etc}\onedot} \def\vs{\emph{vs}\onedot}
+\def\wrt{w.r.t\onedot} \def\dof{d.o.f\onedot}
+\def\etal{\emph{et al}\onedot}
+\makeatother
 
@@
 \date{\today}
+\subjectline{MRC Biostatistics Unit}
+\college{Jesus College}
+\collegeshield{CollegeShields/jesus}
+% \keywords{one two three}
+
+%% Submission date [optional]:
+% \submissiondate{November, 2042}
+
+%% You can redefine the submission notice [optional]:
+% \submissionnotice{A badass thesis submitted on time for the Degree of PhD}
 
@@
 
+\abstract{%
+    TODO
+}
+
+\acknowledgements{%
+  TODO
+}
+
 \begin{document}
 
-\maketitle
+\frontmatter{}
+
+\listoffigures
 
@@
 
-\chapter{Survival analysis with perfect testing} \label{perf-test}
-
 \subfile{cis-perfect-testing}
-
-\chapter{Survival analysis with misclassification} \label{imperf-test}
 
@@
 
-\todo[inline]{Add some descriptive figures e.g. those in ~/COVID/ons-incidence/duration_estimation/reproducible-sims/reports/simulation-studies.pdf}
+\todo[inline]{Add some descriptive figures e.g. those in \url{~/COVID/ons-incidence/duration_estimation/reproducible-sims/reports/simulation-studies.pdf}}
+
+\chapter{CIS episode definitions} \label{episode-def}
+
+\todo[inline]{Get this from SARAH}
 

### 2023-06-01 cis-imperfect-testing.tex
Follow UoC formatting rules
Push chapter name into subfile
Rename main file to thesis
Misc edits / bugfixes

@@
-\documentclass[main.tex]{subfiles}
+\documentclass[thesis.tex]{subfiles}
 
@@
 \begin{document}
-\maketitle
+
+\chapter{Survival analysis with misclassification} \label{imperf-test}
 
@@
 $p_{it}$ to allow for additional episodes being missed.
+
+\section{Data}
+
+I use the CIS data described in \cref{intro:sec:cis}, with positive tests heuristically divided into episodes.
+Prior work categorised a series of positive tests within the same individual CIS were heuristically categorised into episodes (see \cref{episode-def}).
+
+I filter the episodes as follows.
+\begin{itemize}
+\item
+  The first positive in the episode occurred between 16th Oct 2020 and
+  5th Dec 2020 inclusive.
+\item
+  The individual had recorded a negative test prior to their episode
+  beginning (this could have happened at any time, including prior to
+  16th Oct).
+\item
+  A negative test to end the episode (\ie: following the final positive in the episode) has been recorded.
+\end{itemize}
+
+\todo[inline]{Descriptive analyses of the data here}
+
+\section{Modelling}
 
@@
 
-\section{Modifying $p_{ia}$} \label{modifying-p_ia}
+\subsection{Modifying $p_{ia}$} \label{modifying-p_ia}
 
@@
 
-\section{Modifying $p_{it}$} \label{modifying-p_it}
+\subsection{Modifying $p_{it}$} \label{modifying-p_it}
 

### 2023-06-01 cis-perfect-testing.tex
Rename main file to thesis
Misc edits / bugfixes

@@
 
-This chapter aims to estimate the duration of SARS-CoV-2 PCR positivity from CIS data if perfect testing can be assumed (\ie: no misclassification bias).
-I take a survival analysis approach, considering that neither the start nor end time of positivity are known exactly (double interval censoring) and that infections may be missed (truncation), explained in the following paragraphs.
-Through a simulation study, I show the applicability of these methods.
-However, application in reality must account for misclassification bias, the subject of the following chapter.
+The estimation of incidence from prevalence is sensitive to the tail of the duration distribution.
+In \cref{ATACCC} I estimated the duration distribution from the ATACCC data, however, this study suffered from a small sample size and only 20 days of follow-up.
+Therefore, the estimation of the tail of the duration distribution is driven by model assumptions.
+This chapter develops methods to estimate the distribution using CIS data, which has less frequent sampling but a large sample size and long follow-up.
 
-\section{Problem description and approach}\label{perf-test:sec:data}
+I take a survival analysis approach to model the coarse sampling of the CIS (see \cref{perf-test:sec:problem}).
+The problem of priors can be important to the results here, due to the lack of information in the data; in addition, the prior can be used to combine the information in the CIS with the results of \cref{ATACCC} (see \cref{perf-test:sec:parameters-priors}).
+Through a simulation study (see \cref{perf-test:sec:simulation-study}), I show that these methods perform well, and in this context are not sensitive to model or prior specification.
+However, application in reality must account for misclassification bias, the subject of \cref{imperf-testing}.
 
-Consider a series of tests on the same individuals (longitudinal data), where some individuals have a series of positive tests.
+\section{Problem description and approach} \label{perf-test:sec:problem}
+
+\todo[inline]{Some of these definitions likely need to be moved elsewhere as they will be needed in other chapters (e.g. ATACCC analysis).}
+Consider the CIS data.
+It consists of a series of tests on the same individuals (longitudinal data), where some individuals have a series of positive tests.
 We wish to estimate the distribution of time for which an individual is positive for.
-An individual could have multiple infections, in which case define \emph{infection episodes}, corresponding to a single infection event.
+An individual could have multiple infections, and hence I refer to \emph{infection episodes}.
 
-I define an individual as \emph{detectable} if they would test positive if tested, in the absence of misclassification error (which is dealt with in \autoref{imperf-test}).
-The episode's \emph{duration of positivity} is the number of days that they are detectable for; I use time discretised to days, therefore, the duration is at least 1 day.
+I define an individual as \emph{detectable} if they would test positive if tested, in the absence of misclassification error (which is dealt with in \cref{imperf-test}).
+Episode $i$'s \emph{duration of positivity}, $D_i$ is the number of days that they are detectable for; I use time discretised to days, therefore, the duration is at least 1 day.
 Note that the relationship between positivity (\ie the result of a PCR test) and infectiousness (\ie an individual's ability to pass on the virus) is complex, \todo{ref for PCR pos vs infectiousness} and beyond the scope of this thesis.
+I define an episode as starting at time $B_i$ and ends at $E_i$, hence $D_i = E_i - B_i + 1$.
 
 An episode is \emph{detected} when they test positive at least once, if this occurs; episode $i$ is detected at time $t$ if and only if: $i$ is detectable at $t$, $i$ is tested at $t$, and the test returns positive (a false negative does not occur).
+
+Formally, the survival time function, parameterised by $\theta$, is the probability that an episode lasts at least $t$ days: $S_\theat(t) = \prob(D_i \geq t \mid B_i = b, \theta)$ where $D_i$ and $B_i$ are the duration and starting time of episode $i$ respectively.
+We make the standard assumption that $D_i$ and $B_i$ are independent and hence: $S_\theta(t) = \prob(D_i \geq t \mid \theta)$.
 
@@
 
-The reason for this approach, over modelling biomarkers (as in \autoref{ATACCC}) is due to two factors.
+The reason for this approach, over modelling biomarkers (as in \cref{ATACCC}) is due to two factors.
 First, the large sample size in CIS means that approaches where the number of parameters scales with the number of observed infections (as it does with random effect based approaches) would be computationally challenging; this challenge is compounded by the data being stored in a trusted research environment limiting the computational power and tools available.
@@
 The methods in this chapter deal with modelling two elements of the data-generating process, double interval censoring and truncation.
-False negatives are not dealt with and left to \autoref{imperf-test}.
+False negatives are not dealt with and left to \cref{imperf-test}.
 
@@
 An event is \emph{censored} if the time of the event can only be bounded; it is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
-Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa; \textcite{bogaertsSurvival} review many statistical methods applicable to interval censoring (see \autoref{perf-test:fig:double-interval-censor}).
+Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa; \textcite{bogaertsSurvival} review many statistical methods applicable to interval censoring (see \cref{perf-test:fig:double-interval-censor}).
 \emph{Double censoring} means that both the originating and terminating event are censored, and is also a well-studied problem, reviewed by \textcite{sunStatisticala}.
@@
 The second element is truncation.
-Truncation occurs because we do not observe some episodes: an episode can occur between tests and hence the episode is never detected (see \autoref{perf-test:fig:truncation}).
+Truncation occurs because we do not observe some episodes: an episode can occur between tests and hence the episode is never detected (see \cref{perf-test:fig:truncation}).
 We do not know how many infections we do not detect.
 Failure to model truncation will lead to estimating too long durations because the detected episodes are a biased sample with a longer duration.
-To see why they are shorter, consider the undetected infection in the bottom of \autoref{perf-test:fig:truncation}.
+To see why they are shorter, consider the undetected infection in the bottom of \cref{perf-test:fig:truncation}.
 This individual was infected at time 33 for 10 days.
@@
 Truncation is a large problem when individuals move to the infrequent testing schedule within CIS.
-Consider that, based on the estimates in \autoref{ATACCC}, the median length of an episode is 14 days \todo{check this number} we would not expect many episodes to last the 28 days that can occur between tests.
+Consider that, based on the estimates in \cref{ATACCC}, the median length of an episode is 14 days \todo{check this number} we would not expect many episodes to last the 28 days that can occur between tests.
 \begin{figure}
@@
 
-I use the CIS data described in \autoref{intro:sec:cis}, with positive tests heuristically divided into episodes.
-Prior work categorised a series of positive tests within the same individual CIS were heuristically categorised into episodes (see \autoref{episode-def}).
-
-I filter the episodes as follows.
-\begin{itemize}
-\item
-  The first positive in the episode occurred between 16th Oct 2020 and
-  5th Dec 2020 inclusive.
-\item
-  The individual had recorded a negative test prior to their episode
-  beginning (this could have happened at any time, including prior to
-  16th Oct).
-\item
-  A negative test to end the episode (\ie: following the final positive in the episode) has been recorded.
-\end{itemize}
-
-\todo[inline]{Descriptive analyses of the data here}
+While this chapter sets up the framework for analysing the CIS data, applying the method to CIS data requires modelling false negatives.
+That is the subject of \cref{imperf-test} and hence a full description of the data and processing of it is deferred to that chapter.
 
@@
 
-The individual model considers only individuals with observed (\ie, not truncated) episodes.
+The individual model considers only individuals with detected episodes.
 This approach is similar to methods used in previous studies, such as \textcite{heiseyModelling}, \textcite{dempsterMaximum}, and \textcite{turnbullEmpirical}.
@@
 negative tests at times 7, 14, and 56, and positive tests at times 21
-and 28. Purple region is $\alpha_i$, orange region (which includes all
-negative infection times) is $\Omega_i^C$ and unshaded region is
-$\beta_i$.\label{perf-test:fig:partitionSpace}}
+and 28. \label{perf-test:fig:partitionSpace}}
 \end{figure}
@@
 
-First, I derive an analytical solution to $\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i)$ under two different assumptions for $p(n_i)$.
+I derive an analytical solution to $\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i)$ under two different assumptions for $p(n_i)$.
 First, we consider the improper prior $p(n_i) \propto 1/n_i$, this prior has several attractive properties.
@@
 
+The form of $S_\theta(t)$ is yet to be specified.
+As frequently observed in the literature\todo{refs for parameterisation}, it is more convenient to operate with the hazard than the survival or probability mass function directly.
+This is because the mass function must sum to one and the survival must be monotonically decreasing, while there are no constraints across multiple hazards.
+The only constraint on the hazard is to be in the interval $[0, 1]$, since they are probabilities.
+
+Flat or uninformative priors on the hazard lead to very informative priors on the survival, favouring short survival times.
+This is because...
+Our prior belief is not well-represented by this prior, and in fact a priori viewed as extremely unlikely (see \cref{perf-test:fig:flat-priors}),
+Therefore, I favour a prior which is weakly informative on the survival, or an informative prior used to represent our belief from the \cref{ATACCC} analysis.
+
 In this section, I consider the various parameterisation of the survival function that are available.
 This has implications for the choice of prior it is not possible to be vague on all sensible parameterisations.
-Informative priors might also be attractive, and allow the incorporation of the estimates from \autoref{ATACCC}.
-
-Define hazard and survival (maybe this should be done previously?).
+Informative priors might also be attractive, and allow the incorporation of the estimates from \cref{ATACCC}.
 
@@
 We require a ground truth for the distribution giving duration of positivity.
-The majority of this distribution is based on \autoref{ATACCC}'s estimates from the ATACCC data.
+The majority of this distribution is based on \cref{ATACCC}'s estimates from the ATACCC data.
 However, the tail is modified based on Sarah Walker's (unpublished) work based on CIS data.
 ATACCC is thought to estimate accurately the majority of the distribution but not the tail due to a small sample size and model uncertainty.
-These curves and the combined curve are compared in \autoref{perf-test:fig:duration-dist}.
+These curves and the combined curve are compared in \cref{perf-test:fig:duration-dist}.
 Individual $i$'s duration of positivity, $d_i$, is then an independent draw from this distribution.

### 2023-06-02 cis-imperfect-testing.tex
Work on priors

@@
 
+\section{The issues of false negatives}
+
+Demonstrate the likely issue of false negatives (Armitage talk).
+
+Discuss a bit why this is a problem (single positives and bias).
+
+Discuss very low sensitivity needed and changing sensitivity over time (ATACCC and CIS plots).
+
 In this section, we relax the assumption of perfect testing to allow
@@
 $p_{it}$ to allow for additional episodes being missed.
-
-\section{Data}
-
-I use the CIS data described in \cref{intro:sec:cis}, with positive tests heuristically divided into episodes.
-Prior work categorised a series of positive tests within the same individual CIS were heuristically categorised into episodes (see \cref{episode-def}).
-
-I filter the episodes as follows.
-\begin{itemize}
-\item
-  The first positive in the episode occurred between 16th Oct 2020 and
-  5th Dec 2020 inclusive.
-\item
-  The individual had recorded a negative test prior to their episode
-  beginning (this could have happened at any time, including prior to
-  16th Oct).
-\item
-  A negative test to end the episode (\ie: following the final positive in the episode) has been recorded.
-\end{itemize}
-
-\todo[inline]{Descriptive analyses of the data here}
 
@@
 
+\section{Simulation study}
+
+\subsection{Setup}
+
+Each test that was previously positive, is now positive with probability $p_\text{sens}$.
+
+\section{Application to CIS data}
+
+\subsection{Data}
+
+\section{Data}
+
+I use the CIS data described in \cref{intro:sec:cis}, with positive tests heuristically divided into episodes.
+Prior work categorised a series of positive tests within the same individual CIS were heuristically categorised into episodes (see \cref{episode-def}).
+
+I filter the episodes as follows.
+\begin{itemize}
+\item
+  The first positive in the episode occurred between 16th Oct 2020 and
+  5th Dec 2020 inclusive.
+\item
+  The individual had recorded a negative test prior to their episode
+  beginning (this could have happened at any time, including prior to
+  16th Oct).
+\item
+  A negative test to end the episode (\ie: following the final positive in the episode) has been recorded.
+\end{itemize}
+
+\todo[inline]{Descriptive analyses of the data here}
+
 \end{document}

### 2023-06-02 cis-perfect-testing.tex
Work on priors

@@
 However, application in reality must account for misclassification bias, the subject of \cref{imperf-testing}.
+
+\todo[inline]{%
+  Somehwere, I need to fit in what exactly is novel here, and issues with the pre-existing work.
+What to cover is the application to both double interval censored and arbitrarily truncated data (which existed in some of the theoretical frameworks but not used); the issues with the conditional likelihood method of \textcite{bacchettiNonparametric}; and the lack of much work this century.
+}
 
@@
 
-Formally, the survival time function, parameterised by $\theta$, is the probability that an episode lasts at least $t$ days: $S_\theat(t) = \prob(D_i \geq t \mid B_i = b, \theta)$ where $D_i$ and $B_i$ are the duration and starting time of episode $i$ respectively.
+Formally, the survival time function, parameterised by $\theta$, is the probability that an episode lasts at least $t$ days: $S_\theta(t) = \prob(D_i \geq t \mid B_i = b, \theta)$ where $D_i$ and $B_i$ are the duration and starting time of episode $i$ respectively.
 We make the standard assumption that $D_i$ and $B_i$ are independent and hence: $S_\theta(t) = \prob(D_i \geq t \mid \theta)$.
@@
 Furthermore, I explore a Bayesian framework where the amount of explicit data augmentation is limited allowing the method to scale to the large number of episodes present in CIS without requiring vast computing resources (such as a high-performance cluster).
-Only limited resources are unavailable within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
+Only limited resources are available within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
 Therefore, unlike most Bayesian methods (e.g.: \textcite{heBayesiana}, \textcite{heBayesian}, and \textcite{caoModeling}) we avoid augmenting the data with unobserved times of infection or any information on the truncated infections; this is due to the large number of infections (>4000 in the time period considered) making this approach computationally expensive.
@@
 The individual model considers only individuals with detected episodes.
-This approach is similar to methods used in previous studies, such as \textcite{heiseyModelling}, \textcite{dempsterMaximum}, and \textcite{turnbullEmpirical}.
+This approach is based on long-standing methodology~\autocites{heiseyModelling}{dempsterMaximum}{turnbullEmpirical}.
 The conceptual approach here is to form a cohort of individuals for which we observe at least one positive test, and then consider a study which had only enrolled these individuals.
-The modelling approach then corrects for this selection bias.
+The modelling approach then corrects for this selection bias, by considering \emph{ghost} infections that occur in identical individuals but were truncated.
 
@@
 prior to their start, that is, the individual the episode occurred in
-had no tests before $b_i$. The truncated episodes are often referred
-to as the ``ghosts'' of $i$.
+had no tests before $b_i$. The truncated episodes are the ghosts of $i$.
 
-Formally (and following \textcite{heiseyModelling}), we
-define three sets which partition the space of possible infection and
+\Textcite{heiseyModelling} provide a framework for formalising this.
+Define three sets which partition the space of possible infection and
 recovery (\ie: are strict subsets of $T \times T$) given the
@@
 
-Now consider $N_i \sim \text{NegBin}(\mu, r)$ using the
+Now consider $N_i \dist \NegBin(\mu, r)$ using the
 mean/dispersion parameterisation of the negative binomial. Equivalently:
 \begin{align}
-N_i \mid \lambda &\sim \text{Poisson}(\lambda) \\
-\lambda &\sim \text{Gamma}(a, b)
+N_i \mid \lambda &\dist \Poi(\lambda) \\
+\lambda &\dist \GamDist(a, b)
 \end{align}
@@
 
-Under the prior $N_i \sim \text{NegBinom}(\mu, r)$:
+Under the prior $N_i \dist \text{NegBinom}(\mu, r)$:
 \begin{align}
@@
 
-Assuming $N_\text{tot} \sim \text{NegBin}(\mu, r)$:
+Assuming $N_\text{tot} \dist \text{NegBin}(\mu, r)$:
 \begin{align}
@@
 
-\todo[inline]{Section is currently notes: expand into prose which can largely come from previous documents}
-
 The form of $S_\theta(t)$ is yet to be specified.
@@
 The only constraint on the hazard is to be in the interval $[0, 1]$, since they are probabilities.
-
-Flat or uninformative priors on the hazard lead to very informative priors on the survival, favouring short survival times.
-This is because...
-Our prior belief is not well-represented by this prior, and in fact a priori viewed as extremely unlikely (see \cref{perf-test:fig:flat-priors}),
-Therefore, I favour a prior which is weakly informative on the survival, or an informative prior used to represent our belief from the \cref{ATACCC} analysis.
 
@@
 
-The main building basis of the priors I propose is the Beta Process\todo{Beta Process refs}.
-The Beta Process is...
-
 \subsection{Independent priors}
 
-Traditional uninformative priors, such as Beta(1, 1) or Beta(0.5, 0.5), are unsuitable.
+A common assumption throughout Bayesian analyses is independence of the parameters in the prior.
+This is convenient because we can then use univariate distributions.
+\todo{This feels a bit clumsy/patronising to the reader. Maybe change the phrasing?}
+A Beta distribution for each hazard parameter, $\lambda_t$, is a natural choice due to it being a probability.
 
-Instead, I propose Beta(0.1, 1.9), linking to previous literature for a point estimate and showing vague figures.
+An uninformative prior would be $\lambda_t \dist \Beta(\alpha, \alpha)$, with common choices of $\alpha$ being $0.5$, Jeffreys' Prior, or $1$, a flat prior.
+Flat or uninformative priors on the hazard lead to very informative priors on the survival, favouring short survival times.
+The sharp falloff can be seen by noting that these priors have expectation 0.5, hence $\E(S(t)) = \E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right) = 0.5^{t-1}$ which decreases very quickly.
+Our prior belief is not well-represented by this prior, and in fact a priori viewed as extremely unlikely (see \cref{perf-test:fig:flat-priors}),
+Therefore, I favour a prior which is weakly informative on the survival, or an informative prior used to represent our belief from the \cref{ATACCC} analysis.
+\begin{figure}
+  \centering \includegraphics{cis-perfect-testing/flat-prior}
+  \todo[inline]{Create flat priors figure}
+  \caption{The prior predictive survival time (shown) is very short when using either a Beta(1, 1) flat prior or Beta(0.5, 0.5) uninformative prior on the hazard at each time.
+\label{perf-test:fig:flat-prior}}
+\end{figure}
+
+Instead, I propose a weakly informative prior Beta(0.1, 1.9), linking to previous literature for a point estimate and showing vague figures.
+
+A Beta(0.1, 1.9) prior has mean 0.05 while not having much information (the informativeness of a Beta distribution is related to $\alpha + \beta$ which here equals 2, the same as the flat prior case); and the central 95\% mass is 0.00-0.47.
+The central estimate reflects previous estimates that the median duration is in the range 15-20 days~\autocite{cevikShedding}, which would translate to hazard of around 0.05.
+The prior predictive distribution on the survival time implied by this prior is very vague (see \cref{perf-test:fig:vague-prior}).
+\begin{figure}
+  \centering %\includegraphics{cis-perfect-testing/vague-priors}
+  \todo[inline]{Create vague prior figure}
+  \caption{The prior predictive survival time (shown) is very vague using a Beta(0.1, 1.9) prior. \label{perf-test:fig:vague-prior}}
+\end{figure}
+
+\subsection{Informative priors}
+
+The analysis in \cref{ATACCC} can be used to provide prior information for the CIS analysis.
+In particular, the ATACCC study frequently samples individuals early in the infection.
+Therefore, the survival distribution from that analysis should be reliable through this period.
+CIS can then inform the remainder of the distribution where the previous analysis is based on extrapolation.
+
+The model structure assumed in \cref{ATACCC} leads to positive correlation in the posterior estimates of the hazard.
+It is important to propagate this correlation as part of the prior uncertainty into the present analysis.
+We do this in two parts: first, approximate the previous posterior estimate of the logit-hazard as a multivariate normal (which can be viewed as an approximation of Markov melding\cref{cite Markov melding}).
+Then, use a Beta process to add additional uncertainty to this.
+
+There are two reasons to increase the uncertainty relative to the \cref{ATACCC} posterior.
+First, we do not have full belief in these estimates due to the extrapolation problem previously discussed.
+Second, the results may not fully generalise between the two studies.
+The samples are tested using different laboratories, with different PCR machines, and hence may not have the same characteristics~\autocite{punchooLaboratory}.
+The study participants, while both targeting a general population cohort, may also differ because ATACCC recruited through NHS Test and Trace while CIS recruited by sampling from a database of addresses (see \cref{intro:sec:data}).
+
+The first step, approximating the \cref{ATACCC} posterior as a multivariate normal, is done through the method of moments.
+The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC}).
+
+The discrete Beta process prior generalises the form of prior used in the previous section, allowing the central estimate of the hazard to vary over time~\autocites{ibrahimBayesian}{sunStatisticala}.
+\begin{align}
+  \lambda_t &\sim \text{Beta}(\alpha_t, \beta_t) &t = 1, 2, \dots \\
+  \alpha_t &= c_t h_t + \alpha_0 \\
+  \beta_t &= c_t (1 - h_t) + \beta_0
+\end{align}
+where $k_t$, $\alpha_0$, and $\beta_0$ are hyperparameters; and $h_t$ is a point estimate of the hazard at time $t$ from ATACCC.
+An intuition for what this distribution represents can be gained by considering the conjugate model for $\lambda_t$ with a beta prior and a binomial likelihood.
+If $\lambda_t$ is given the prior distribution $\text{Beta}(\alpha_0, \beta_0)$, and we then have $k_t$ observations with $k_t h_t$ successes, then the posterior distribution for $\lambda_t$ is $\text{Beta}(\alpha_t, \beta_t)$ (as defined above).
 
@@
 
-\subsection{Informative priors}
+A final possible prior are ones to favour smoothness in the hazard.
+This is the Bayesian equivalent of a penalised likelihood, used frequently in the non-parametric frequentist setting.
+Biological and epidemiological considerations mean that it would be surprising if the hazard varied significantly day-to-day.
+A smoothing prior encodes this information within the analysis.
 
-ATACCC can be used to provide prior information for the CIS analysis.
+I consider two forms of smoothing priors: first-order and second-order random walks on the logit-hazard, $\logit(\lambda_t) = \log(\lambda_t) - \log(1 - \lambda_t)$.
+The logit transformation maps from the $[0, 1]$ interval to the full real line.
+\todo{The logit intro will be in the previous subsection}
+Random walks are a very common form of smoothing prior for discrete time.
 
-Intuition: ATACCC informs early part, CIS informs late part.
+A first-order random walk prior encodes that the hazard should be constant with some random changes at each time step.
+\begin{align}
+  \logit\lambda_{t+1} &= \logit\lambda_t + \sigma \epsilon_t &\text{for $t \geq 1$} \\
+  \epsilon_t &\dist N(0, 1) \\
+  \logit \lambda_1 &\dist N(\mu_1, \sigma_1^2)
+\end{align}
+The hyperparameter $\sigma$ controls the standard deviation of the random walk, and the prior on $\lambda_1$ gives the starting point of the random walk.
+Smaller values favour a smoothing process.
 
-Problem: do not believe late part of ATACCC analysis.
-
-First, transform ATACCC analysis into a posterior over survival time.
-
-Second, approximate ATACCC analysis with multivariate normal distribution.
-Show that this is a good approximation.
-
-Finally, use a Beta Process to discount some information in ATACCC towards a vague prior (as in the previous section).
+A second-order random walk prior encodes that the hazard should be linearly changing with some random changes at each time step.
+\begin{align}
+  \logit\lambda_{t+1}
+  &= \logit\lambda_t + (\logit\lambda_t - \logit\lambda_{t-1}) + \sigma \epsilon_t &\text{for $t \geq 2$} \\
+  &= 2\logit\lambda_t - \logit\lambda_{t-1} + \sigma \epsilon_t \\
+  \epsilon_t &\dist N(0, 1) &\text{for $t \geq 2$}  \\
+  \logit\lambda_2 &= \logit\lambda_1 + \epsilon_1 \\\
+  \epsilon_1 &\dist N(\mu_2, \sigma_2^2) \\
+  \logit \lambda_1 &\dist N(\mu_1, \sigma_1^2)
+\end{align}
+The hyperparameter $\sigma$ controls the standard deviation of the random walk, and there is an additional prior on $\epsilon_1$ for the initial gradient of the change in the hazard.
 
@@
 
-\subsubsection{Testing and infection times}
+I simulate each individual enrolled in CIS who was tested at least once between ?? and ??.
+All their tests between ?? and ?? are considered, and each individual is assigned an infection time $B_i$ uniformly at random from ??.
+\todo{check dates and periods}
 
-\subsubsection{Duration of positivity}
-
-We require a ground truth for the distribution giving duration of positivity.
-The majority of this distribution is based on \cref{ATACCC}'s estimates from the ATACCC data.
-However, the tail is modified based on Sarah Walker's (unpublished) work based on CIS data.
-ATACCC is thought to estimate accurately the majority of the distribution but not the tail due to a small sample size and model uncertainty.
+I take as the ground truth distribution for duration of positivity a combination of the estimates from \cref{ATACCC} with an inflated tail to represent what is seen within CIS.
+The tail is modified based on Sarah Walker's (unpublished) work based on CIS data.
 These curves and the combined curve are compared in \cref{perf-test:fig:duration-dist}.
-Individual $i$'s duration of positivity, $d_i$, is then an independent draw from this distribution.
+Individual $i$'s duration of positivity, $D_i$, is then an independent draw from this distribution.
 
@@
 A flexible, spline-based form is assumed for the baseline survival function~\autocite{roystonSTPM,roystonFlexible} with covariates introduced via proportional odds.
+By not accounting for either the truncation or the interval censoring of the initiating event, this analysis has competing biases which makes them hard to interpret~\autocite{cisMethodsONS}.
 
-We know of two biases introduced from this analysis.
-\enquote{There is a bias in estimating the clearance distribution because the analysis used to estimate how long a person stays positive only starts from their first positive test.
-Since (most) people will have become positive on an earlier day, this will bias the clearance curves downwards (making the estimates too short).
-However, there is another bias due to the survey missing positive episodes entirely if they are short.
-This means that our dataset has fewer short positive episodes than in the population as a whole, and that the sample used to run the survival analysis is biased towards people with longer positive episodes.
-This will bias the clearance curves upwards (making the estimates too long).}~\autocite{cisMethodsONS}.
+% \todo[inline]{The following seems important to justify why we can't just use these estimates but is also basically an aside}
+% We know of two biases introduced from this analysis.
+% \enquote{There is a bias in estimating the clearance distribution because the analysis used to estimate how long a person stays positive only starts from their first positive test.
+% Since (most) people will have become positive on an earlier day, this will bias the clearance curves downwards (making the estimates too short).
+% However, there is another bias due to the survey missing positive episodes entirely if they are short.
+% This means that our dataset has fewer short positive episodes than in the population as a whole, and that the sample used to run the survival analysis is biased towards people with longer positive episodes.
+% This will bias the clearance curves upwards (making the estimates too long).}~\autocite{cisMethodsONS}.
 
@@
 The first 30 days of the distribution is proportional to the ATACCC estimate, with the rest proportional to this CIS-based estimates.
-The CIS-based estimates are shifted 3 days in order to make this smooth; this counteracts the above bias of missing the start of the infection (for times above 30 days, the other bias, missing short infections, is negligible).
+% The CIS-based estimates are shifted 3 days in order to make this smooth; this counteracts the above bias of missing the start of the infection (for times above 30 days, the other bias, missing short infections, is negligible).
+ERR...
+I'M NOT DOING THIS ANYMORE OH NO!!
 Denote by $f_A(t)$ the distribution function estimated from ATACCC and $f_C(t)$ that from these CIS-based estimates.
@@
 	f_A(t) &t \leq 30 \\
-	f_C(t-3) &t > 30
+	f_C(t) &t > 30
 \end{cases}
@@
 
-\subsubsection{Test results and episode summaries}
-\todo[inline]{Make notation here consistent with elsewhere}
+The first day individual $i$ can be positive is their day of infection, $B_i$, the last day is $E_i = B_i + D_i - 1$ where $D_i$ is the duration of positivity for individual $i$, drawn from $f_S$.
+Any test between $B_i$ and $E_i$ (inclusive) is positive; outside this period, all tests are negative.
 
-The first day individual $i$ can be positive is their day of infection, $t_i^\text{onset}$, the last day is $t_i^\text{final} = t_i^\text{onset} + d_i - 1$ where $d_i$ is the duration of positivity for individual $i$.
-Outside this period, all tests are negative (\ie: no false positives occur). 
-Any test between $t_i^\text{onset}$ and $t_i^\text{final}$ (inclusive) is positive with probability $p_{sens}$ (that is, each test is an independent Bernoulli trial).
+I then filter the episodes to replicate the process which will be used when working with the CIS data.
+First, discard any episodes where there are no positive tests as they are never detected.
+Next, discard any episodes where there are no negative tests prior to the first positive; since there is no lower bound on their infection time, they contain little information, this selection effect was included when constructing $p_{it}$.
+Finally, of those remaining episodes, we select a random subset to match the number of episodes in the simulation to the number of episodes in the modelling cohort.
+The last step is required because everyone is infected in the simulation which does not happen in reality.
 
-Each individual is then summarised into an "episode" (when working with CIS data, the same individual can have multiple episodes due to reinfections, these are split heuristically but a negligible number in the pre-Omicron period and hence ignored within the simulation).
-An episode is characterised by:
-
-- The total number of positive tests observed per individual, $n_i^\text{pos}$.
-- The time of an individual's first positive test $V'_i$.
-- The time of an individual's last negative test prior to $V_i'$, $E'_i$.
-- The time of an individual's last positive test, $L_i'$.
-- The time of an individual's first negative test after $L_i'$, $R_i'$.
-
-\subsubsection{Restrictions and sample size}
-
-When using the real testing schedules, we then filter the episodes to replicate the process used when working with the CIS data.
-First, we discard any episodes where $n_i^\text{pos}=0$ as they are never detected.
-We then apply the same criteria as that applied to the modelling dataset (see Section \@ref(design)).
-Finally, of those remaining episodes, we select a random subset to match the number of episodes in the simulation to the number of episodes in the modelling cohort.
-The last step is required because we infect everyone in the simulation which does not happen in reality.
+For each preserved infection episode, calculate $(l_i^{(b)}, r_i^{b}, l_i^{(e)}, r_i^{(e)})$ by taking the day after the last negative prior to any positives, the first positive, the last positive, and the day before the negative following the last positive respectively.
 
 \subsection{Descriptive analyses of simulated data}
+
+\subsection{Inference and hyperparameters}
 

### 2023-06-02 cis-perfect-testing/priors.R

@@ -1,0 +1,92 @@
+library(dplyr)
+library(ggplot2)
+library(patchwork)
+library(purrr)
+library(tidybayes)
+library(tidyr)
+source(here::here("utils.R"))
+
+flat_params = c(0.5, 1)
+dist_levels = c(
+  map_chr(flat_params, ~glue::glue("Beta({.x}, {.x})")),
+  "ATACCC"
+)
+
+sample_priors = function(alpha, beta, n_samples = 1000) {
+  stopifnot(length(alpha) == length(beta))
+  tibble(
+    time = 1:length(alpha),
+    alpha = alpha,
+    beta = beta
+  ) %>%
+    mutate(
+      lambda = map2(alpha, beta, ~rbeta(n_samples, .x, .y)),
+      .draw = list(1:n_samples),
+    ) %>%
+    unnest(c(.draw, lambda)) |>
+    group_by(.draw) |>
+    reframe(
+      time = c(time, max(time) + 1),
+      S = c(1, cumprod(1 - lambda)),
+      lambda = c(lambda, 0)
+    )
+}
+
+sample_constant_prior = function(alpha, beta, prior_len, n_samples = 1000) {
+  stopifnot(length(alpha) == 1)
+  stopifnot(length(beta) == 1)
+  sample_priors(rep(alpha, prior_len), rep(beta, prior_len), n_samples)
+}
+
+beta_plot_dens = function(alpha, beta) {
+  reduce2(
+    alpha,
+    beta,
+    function(p, a, b) p + stat_function(aes(colour = glue::glue("Beta({a}, {b})")),
+                                        fun = dbeta, args = list(a, b)),
+    .init = ggplot(NULL) + labs(colour = "Dist", x = "Value", y = "Density")
+  ) +
+  standard_plot_theming()
+}
+
+beta_plot_surv = function(alpha, beta, prior_len = 30) {
+  purrr::map2(
+    alpha, beta,
+    ~sample_constant_prior(.x, .y, prior_len) |>
+      mutate(dist = glue::glue("Beta({.x}, {.y})"))
+  ) |>
+  bind_rows(
+    ataccc_posterior |>
+      mutate(dist = "ATACCC")
+  ) |>
+  group_by(dist, time) |>
+  mean_qi(S, .simple.names = FALSE) |>
+  mutate(dist = factor(dist, levels = dist_levels)) |>
[…]

### 2023-06-02 utils.R

@@ -1,0 +1,5 @@
+standard_plot_theming = function() {
+    rlang::list2(
+        theme_minimal()
+    )
+}

### 2023-06-02 cis-perfect-testing.tex
Move total model to appendix

@@
 In \cref{ATACCC} I estimated the duration distribution from the ATACCC data, however, this study suffered from a small sample size and only 20 days of follow-up.
+\todo{Ref ATACCC chapter and in following paragraph}
 Therefore, the estimation of the tail of the duration distribution is driven by model assumptions.
@@
 Through a simulation study (see \cref{perf-test:sec:simulation-study}), I show that these methods perform well, and in this context are not sensitive to model or prior specification.
-However, application in reality must account for misclassification bias, the subject of \cref{imperf-testing}.
+However, application in reality must account for misclassification bias, the subject of \cref{imperf-test}.
+\todo{Ref chapter incl misclassification}
 
@@
 I define an individual as \emph{detectable} if they would test positive if tested, in the absence of misclassification error (which is dealt with in \cref{imperf-test}).
+\todo{Ref chapter incl misclassification}
 Episode $i$'s \emph{duration of positivity}, $D_i$ is the number of days that they are detectable for; I use time discretised to days, therefore, the duration is at least 1 day.
@@
 The reason for this approach, over modelling biomarkers (as in \cref{ATACCC}) is due to two factors.
+\todo{Ref ATACCC chapter}
 First, the large sample size in CIS means that approaches where the number of parameters scales with the number of observed infections (as it does with random effect based approaches) would be computationally challenging; this challenge is compounded by the data being stored in a trusted research environment limiting the computational power and tools available.
@@
 False negatives are not dealt with and left to \cref{imperf-test}.
+\todo{Ref chapter incl misclassification}
 
@@
 Consider that, based on the estimates in \cref{ATACCC}, the median length of an episode is 14 days \todo{check this number} we would not expect many episodes to last the 28 days that can occur between tests.
+\todo{Ref ATACCC chapter}
 \begin{figure}
@@
 That is the subject of \cref{imperf-test} and hence a full description of the data and processing of it is deferred to that chapter.
+\todo{Ref chapter incl misclassification}
 
@@
 
-The individual model considers only individuals with detected episodes.
+I model only individuals with detected episodes.
 This approach is based on long-standing methodology~\autocites{heiseyModelling}{dempsterMaximum}{turnbullEmpirical}.
@@
 
-In contrast, the total model also takes into account data from individuals without any observed episodes.
+I have also considered a model including individuals without any observed episodes (see \cref{appendix:total-model} for details).
 These individuals may have truncated episodes or may not have been infected at all.
-This approach utilises their test times to estimate the probability of truncated infections, ultimately providing an interpretable estimate of the number of truncated episodes in the cohort.
+However, I found that this model has little difference in results while increasing computational complexity and requiring assumptions on reinfection (discussed in \cref{perf-test:sec:discussion}).
 
@@
 
-\subsection{Individual model}\label{perf-test:sec:individual-model}
-
-\subsubsection{Likelihood}\label{perf-test:sec:likelihood}
+\subsection{Likelihood}\label{perf-test:sec:likelihood}
 
@@
 
-\subsubsection{Full posterior}\label{perf-test:sec:full-posterior}
+\subsection{Full posterior}\label{perf-test:sec:full-posterior}
 
@@
 
-\subsection{Total model}\label{perf-test:sec:total-model}
-
-We now consider all $N$ individuals in the CIS cohort who have at
-least one test during the period of interest, this includes those
-without an observed infection, and assume that each is equally likely to
-be infected. As before, $i = 1, \dots, n_a$ indexes episodes but now
-$j = 1, \dots, N$ indexes individuals. All episodes have an associated
-individual, each individual may have any number (including 0) episodes.
-Admissible regions are only defined for episodes, and have not changed
-from the previous section. The truncated region is defined for each
-individual, in addition we consider the combined truncated region which
-is episodes that would be truncated without conditioning on which
-individual the episode occurs in:
-$\Omega^C = \{ (b, e, j) : (b, e) \in \Omega_j^C \}$.
-
-Denote by $n_\text{tot}$ the total number of episodes across all $N$
-individuals, regardless of whether they are admissible or truncated (we
-know there are no inadmissible episodes by definition).
-
-For any infection, the probability of it being admissible for episode
-$i$ is $\frac{1}{N} p_{ia}$. That is, the probability that the
-infection occurs in the individual corresponding to the individual in
-which episode $i$ ($1/N$ by the assumption that episodes are equally
-likely to occur in any of the $N$ individuals) multiplied by the
-probability that the episode is admissible for episode $i$ conditional
-on it occurring in the relevant individual ($p_{ia}$).
-
-The total number of truncated infections that occur is
-$n_t = n_\text{tot} - n_a$. Conditional on an infection occurring in
-individual $j$, its probability of being truncated is $p_{jt}$ (as
-defined previously). Therefore, the overall probability of an episode
-being truncated is $p_t = \frac{1}{N} \sum_{j=1}^N p_{jt}$.
-
-Consider our data as there being one admissible episode for each
-observed episode, and all other episodes being truncated. As in the
-previous section, conditional on $n_\text{tot}$, we have a multinomial
-likelihood.
-\begin{align}
-p(\text{data} \mid \theta)
-&= \sum_{n_\text{tot}=n_a}^\infty p(\text{data} \mid n_\text{tot}, \theta) p(n_\text{tot}) \\
-&= \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} \left( \prod_{i=1}^{n_a} \frac{1}{N} p_{ia} \right) p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \\
-&= \left( \prod_{i=1}^{n_a} \frac{1}{N} p_{ia} \right) \left( \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \right) \\
-\end{align}
-
-In this case, it would be computationally feasible to augment the data
-with $n_\text{tot}$ directly (as it is only a single parameter).
-However, implementation is easier if we can use a closed-form for the
-sum (as in the previous section) as it allows the use of standard
-implementations of NUTS such as Stan, which cannot handle discrete
-parameters.
-
-Assuming $N_\text{tot} \dist \text{NegBin}(\mu, r)$:
-\begin{align}
-&\sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \\
-&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot} \mid \lambda) p(\lambda) d\lambda \\
-&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} \frac{\lambda^n_\text{tot} e^{-\lambda}}{n_\text{tot}!} p(\lambda) d\lambda \\
-&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{1}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} \lambda^{n_\text{tot}-n_a} \lambda^{n_a} e^{-\lambda} p(\lambda) d\lambda \\
-&= \int \lambda^{n_a} e^{-\lambda} p(\lambda) \sum_{n_t=0}^\infty \frac{1}{n_t!} p_t^{n_t} \lambda^{n_t} d\lambda &n_t = n-n_a\\
-&= \int \lambda^{n_a} e^{-\lambda} p(\lambda) e^{\lambda p_t} d\lambda \\
-&= \int \lambda^{n_a} e^{-\lambda(1 - p_t)} p(\lambda) d\lambda \\
-&= \int \lambda^{n_a} e^{-\lambda(1 - p_t)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda \\
-&= \int \frac{b^a}{\Gamma(a)} \lambda^{a+n_a-1} e^{-(b+1-p_t)\lambda} \lambda d\lambda \\
-&= \frac{b^a}{\Gamma(a)} \frac{\Gamma(a+n_a)}{(b+1-p_t)^{a+n_a}} \\
-&\propto (b+1-p_t)^{-(a+n_a)} \\
-&= (r/\mu + 1 - p_t)^{-(r+n_a)} \\
-&\propto(r + \mu (1- p_t))^{-(r+n_a)}
-\end{align}
-
-Which gives the full posterior density as:
-\begin{align}
-p(\theta \mid \text{data})
-&\propto \frac{p(\theta) \prod_{i=1}^{n_a} p_{ia}}{(r + \mu(1 - p_t))^{r+n_a}}
-\end{align}
-
-As before, being able to reconstruct the posterior of $n_\text{tot}$
-using its full conditional is useful.
-\begin{align}
-p(n_\text{tot} \mid \text{data}, \theta)
-&\propto p(\text{data} \mid \theta, n_\text{tot}) p(n_\text{tot}) \\
-&= \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} \left( \prod_{i=1}^{n_a} \frac{1}{N} p_{ia} \right) p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \\
-&\propto \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}} \frac{\Gamma(r + n_\text{tot})}{n_\text{tot}!} \left( \frac{\mu}{r + \mu} \right)^{n_\text{tot}}  \\
-&\propto \frac{\Gamma(r + n_\text{tot})}{(n_\text{tot}-n_a)!} \left( \frac{\mu p_t}{r + \mu} \right)^{n_\text{tot}}  \\
-&\propto \frac{\Gamma((r + n_a) + (n_\text{tot}- n_a))}{(n_\text{tot}-n_a)!} \left( \frac{\mu p_t}{r + \mu} \right)^{n_\text{tot}-n_a}  \\
-\end{align}
-which is a negative binomial pmf with size parameter $r + n_a$ and
-probability parameter $\frac{r+\mu(1-p_t)}{r+\mu}$.
-
 \section{Parameterisation and priors for the survival function} \label{perf-test:sec:parameters-priors}
@@
 Informative priors might also be attractive, and allow the incorporation of the estimates from \cref{ATACCC}.
+\todo{Ref ATACCC chapter}
 
@@
 Therefore, I favour a prior which is weakly informative on the survival, or an informative prior used to represent our belief from the \cref{ATACCC} analysis.
+\todo{Ref ATACCC chapter}
 \begin{figure}
@@
 The analysis in \cref{ATACCC} can be used to provide prior information for the CIS analysis.
+\todo{Ref ATACCC chapter here and next few}
 In particular, the ATACCC study frequently samples individuals early in the infection.
@@
 The model structure assumed in \cref{ATACCC} leads to positive correlation in the posterior estimates of the hazard.
+\todo{Ref ATACCC chapter}
 It is important to propagate this correlation as part of the prior uncertainty into the present analysis.
-We do this in two parts: first, approximate the previous posterior estimate of the logit-hazard as a multivariate normal (which can be viewed as an approximation of Markov melding\cref{cite Markov melding}).
+We do this in two parts: first, approximate the previous posterior estimate of the logit-hazard as a multivariate normal (which can be viewed as an approximation of Markov melding~\autoref{goudieJoining}).
 Then, use a Beta process to add additional uncertainty to this.
@@
 I take as the ground truth distribution for duration of positivity a combination of the estimates from \cref{ATACCC} with an inflated tail to represent what is seen within CIS.
+\todo{Ref ATACCC chapter}
 The tail is modified based on Sarah Walker's (unpublished) work based on CIS data.
@@
 
-\section{Discussion}
+\section{Discussion} \label{perf-test:sec:discussion}
 
@@
 
+\appendix
+
+\chapter{Including all CIS individuals in survival analysis}\label{perf-test:sec:total-model} \label{appendix:total-model}
+
+We now consider all $N$ individuals in the CIS cohort who have at
+least one test during the period of interest, this includes those
+without an observed infection, and assume that each is equally likely to
+be infected. As before, $i = 1, \dots, n_a$ indexes episodes but now
+$j = 1, \dots, N$ indexes individuals. All episodes have an associated
+individual, each individual may have any number (including 0) episodes.
+Admissible regions are only defined for episodes, and have not changed
+from the previous section. The truncated region is defined for each
+individual, in addition we consider the combined truncated region which
+is episodes that would be truncated without conditioning on which
+individual the episode occurs in:
+$\Omega^C = \{ (b, e, j) : (b, e) \in \Omega_j^C \}$.
+
+Denote by $n_\text{tot}$ the total number of episodes across all $N$
+individuals, regardless of whether they are admissible or truncated (we
+know there are no inadmissible episodes by definition).
+
+For any infection, the probability of it being admissible for episode
+$i$ is $\frac{1}{N} p_{ia}$. That is, the probability that the
+infection occurs in the individual corresponding to the individual in
+which episode $i$ ($1/N$ by the assumption that episodes are equally
+likely to occur in any of the $N$ individuals) multiplied by the
+probability that the episode is admissible for episode $i$ conditional
+on it occurring in the relevant individual ($p_{ia}$).
+
+The total number of truncated infections that occur is
+$n_t = n_\text{tot} - n_a$. Conditional on an infection occurring in
+individual $j$, its probability of being truncated is $p_{jt}$ (as
+defined previously). Therefore, the overall probability of an episode
+being truncated is $p_t = \frac{1}{N} \sum_{j=1}^N p_{jt}$.
+
+Consider our data as there being one admissible episode for each
+observed episode, and all other episodes being truncated. As in the
+previous section, conditional on $n_\text{tot}$, we have a multinomial
+likelihood.
+\begin{align}
+p(\text{data} \mid \theta)
+&= \sum_{n_\text{tot}=n_a}^\infty p(\text{data} \mid n_\text{tot}, \theta) p(n_\text{tot}) \\
+&= \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} \left( \prod_{i=1}^{n_a} \frac{1}{N} p_{ia} \right) p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \\
+&= \left( \prod_{i=1}^{n_a} \frac{1}{N} p_{ia} \right) \left( \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \right) \\
+\end{align}
+
+In this case, it would be computationally feasible to augment the data
+with $n_\text{tot}$ directly (as it is only a single parameter).
+However, implementation is easier if we can use a closed-form for the
+sum (as in the previous section) as it allows the use of standard
+implementations of NUTS such as Stan, which cannot handle discrete
+parameters.
+
+Assuming $N_\text{tot} \dist \NegBin(\mu, r)$:
+\begin{align}
+&\sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \\
+&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot} \mid \lambda) p(\lambda) d\lambda \\
+&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} \frac{\lambda^n_\text{tot} e^{-\lambda}}{n_\text{tot}!} p(\lambda) d\lambda \\
+&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{1}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} \lambda^{n_\text{tot}-n_a} \lambda^{n_a} e^{-\lambda} p(\lambda) d\lambda \\
+&= \int \lambda^{n_a} e^{-\lambda} p(\lambda) \sum_{n_t=0}^\infty \frac{1}{n_t!} p_t^{n_t} \lambda^{n_t} d\lambda &n_t = n-n_a\\
+&= \int \lambda^{n_a} e^{-\lambda} p(\lambda) e^{\lambda p_t} d\lambda \\
+&= \int \lambda^{n_a} e^{-\lambda(1 - p_t)} p(\lambda) d\lambda \\
+&= \int \lambda^{n_a} e^{-\lambda(1 - p_t)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda \\
+&= \int \frac{b^a}{\Gamma(a)} \lambda^{a+n_a-1} e^{-(b+1-p_t)\lambda} \lambda d\lambda \\
+&= \frac{b^a}{\Gamma(a)} \frac{\Gamma(a+n_a)}{(b+1-p_t)^{a+n_a}} \\
+&\propto (b+1-p_t)^{-(a+n_a)} \\
+&= (r/\mu + 1 - p_t)^{-(r+n_a)} \\
+&\propto(r + \mu (1- p_t))^{-(r+n_a)}
+\end{align}
+
+Which gives the full posterior density as:
+\begin{align}
+p(\theta \mid \text{data})
+&\propto \frac{p(\theta) \prod_{i=1}^{n_a} p_{ia}}{(r + \mu(1 - p_t))^{r+n_a}}
+\end{align}
+
+As before, being able to reconstruct the posterior of $n_\text{tot}$
+using its full conditional is useful.
+\begin{align}
+p(n_\text{tot} \mid \text{data}, \theta)
+&\propto p(\text{data} \mid \theta, n_\text{tot}) p(n_\text{tot}) \\
+&= \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} \left( \prod_{i=1}^{n_a} \frac{1}{N} p_{ia} \right) p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \\
+&\propto \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}} \frac{\Gamma(r + n_\text{tot})}{n_\text{tot}!} \left( \frac{\mu}{r + \mu} \right)^{n_\text{tot}}  \\
+&\propto \frac{\Gamma(r + n_\text{tot})}{(n_\text{tot}-n_a)!} \left( \frac{\mu p_t}{r + \mu} \right)^{n_\text{tot}}  \\
+&\propto \frac{\Gamma((r + n_a) + (n_\text{tot}- n_a))}{(n_\text{tot}-n_a)!} \left( \frac{\mu p_t}{r + \mu} \right)^{n_\text{tot}-n_a}  \\
+\end{align}
+which is a negative binomial pmf with size parameter $r + n_a$ and
+probability parameter $\frac{r+\mu(1-p_t)}{r+\mu}$.
+
 \end{document}

### 2023-06-03 cis-perfect-testing.tex
Sketch results

@@
 }
+
+\todo[inline]{Need to discuss interpretation of the ghosts in the individual model, why this is hard, and the implications for setting a prior here. Not even clear in simulation what it should be!}
 
@@
 The sharp falloff can be seen by noting that these priors have expectation 0.5, hence $\E(S(t)) = \E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right) = 0.5^{t-1}$ which decreases very quickly.
-Our prior belief is not well-represented by this prior, and in fact a priori viewed as extremely unlikely (see \cref{perf-test:fig:flat-priors}),
+Our prior belief is not well-represented by this prior, and in fact a priori viewed as extremely unlikely (see \cref{perf-test:fig:flat-prior}),
 Therefore, I favour a prior which is weakly informative on the survival, or an informative prior used to represent our belief from the \cref{ATACCC} analysis.
@@
   \centering \includegraphics{cis-perfect-testing/flat-prior}
-  \todo[inline]{Create flat priors figure}
-  \caption{The prior predictive survival time (shown) is very short when using either a Beta(1, 1) flat prior or Beta(0.5, 0.5) uninformative prior on the hazard at each time.
+  \caption{The density of Beta(1, 1) flat prior and Beta(0.5, 0.5) uninformative priors (left) with the prior predictive survival time (right).
+The implied prior on the survival time is very short when using either prior on the hazard at each time.
 \label{perf-test:fig:flat-prior}}
@@
 
-A Beta(0.1, 1.9) prior has mean 0.05 while not having much information (the informativeness of a Beta distribution is related to $\alpha + \beta$ which here equals 2, the same as the flat prior case); and the central 95\% mass is 0.00-0.47.
-The central estimate reflects previous estimates that the median duration is in the range 15-20 days~\autocite{cevikShedding}, which would translate to hazard of around 0.05.
+A Beta(0.1, 1.9) prior has mean 0.05 while not having much information (the informativeness of a Beta distribution is related to $\alpha + \beta$ which here equals 2, the same as the flat prior case); and the central 95\% mass is 0.00--0.47.
+The central estimate reflects previous estimates that the median duration is in the range 15--20 days~\autocite{cevikShedding}, which would translate to hazard of around 0.05.
 The prior predictive distribution on the survival time implied by this prior is very vague (see \cref{perf-test:fig:vague-prior}).
 \begin{figure}
-  \centering %\includegraphics{cis-perfect-testing/vague-priors}
-  \todo[inline]{Create vague prior figure}
-  \caption{The prior predictive survival time (shown) is very vague using a Beta(0.1, 1.9) prior. \label{perf-test:fig:vague-prior}}
+  \centering \includegraphics{cis-perfect-testing/vague-prior}
+  \caption{The density of Beta(0.1, 1.9) vague prior (left) and the prior predictive survival time (right).
+The prior predictive survival time is very vague using a Beta(0.1, 1.9) prior.
+Note the different x-axis to \cref{perf-test:fig:flat-prior}.
+\label{perf-test:fig:vague-prior}}
 \end{figure}
@@
 The model structure assumed in \cref{ATACCC} leads to positive correlation in the posterior estimates of the hazard.
-\todo{Ref ATACCC chapter}
 It is important to propagate this correlation as part of the prior uncertainty into the present analysis.
-We do this in two parts: first, approximate the previous posterior estimate of the logit-hazard as a multivariate normal (which can be viewed as an approximation of Markov melding~\autoref{goudieJoining}).
+We do this in two parts: first, approximate the previous posterior estimate of the logit-hazard as a multivariate normal (which can be viewed as an approximation of Markov melding~\autocite{goudieJoining}).
 Then, use a Beta process to add additional uncertainty to this.
@@
 
+Stan including settings
+
+Priors on missed infections.
+
+Survival prior hyperparameters (maybe this should just go in the previous section though?)
+
+Criteria for assessing convergence (Rhat, ESS, divergent transitions)
+
 \subsection{Results}
 
-\todo[inline]{Need to decide what exactly to include... There's lots of possible results but need to form some coherent story here!}
+Points to make:
+
+\begin{enumerate}
+  \item Convergence assessment
+  \item Repeated simulation showed very little variation (large sample size)
+  \item Recovers true value of survival well with little difference between priors, \cref{perf-test:fig:survival-results}
+  \item RW2 performs poorly considering hazard and when little information in prior very uncertain, \cref{perf-test:fig:hazard-results}; not clear why this is (\cref{perf-test:fig:hazard-pairs-results})
+  \item Do I need some quantitive performance measures or is just showing plots enough?
+\end{enumerate}
+
+\begin{figure}
+  \centering \includegraphics{cis-perfect-testing/survival-results}
+  \caption{Comparison of the posterior estimate of the survival distribution (pointwise median and 95\% CrI) when using different priors on the hazard. \label{perf-test:fig:survival-results}}
+\end{figure}
+
+\begin{figure}
+  \centering \includegraphics{cis-perfect-testing/hazard-results}
+  \caption{Comparison of the posterior estimate of the hazard (pointwise median and 95\% CrI) when using different priors on the hazard. \label{perf-test:fig:hazard-results}}
+\end{figure}
+
+\begin{figure}
+  \centering \includegraphics{cis-perfect-testing/hazard-pairs-results}
+  \todo[inline]{Probably scrap this (correlation) plot}
+  \caption{Correlation between posterior estimates of the hazard when using the vague prior. \label{perf-test:fig:hazard-pairs-results}}
+\end{figure}
 
@@
 
-\todo[inline]{Add discussion after deciding on results}
+Discussion points:
+\begin{enumerate}
+  \item Other priors (GPs?)
+  \item Inclusion of covariates
+  \item Individual vs total model
+  \item Consideration of smaller sample size
+\end{enumerate}
+
+Limitations to include:
+\begin{enumerate}
+  \item Assumption of constant incidence
+\end{enumerate}
 
 \section{Conclusion} \label{perf-test:sec:conclusion}
-
-\listoftodos
 
@@
 \chapter{Including all CIS individuals in survival analysis}\label{perf-test:sec:total-model} \label{appendix:total-model}
+
+\todo[inline]{Add an intro to this and maybe a plot of results in simulation}
 

### 2023-06-03 cis-perfect-testing/priors.R
Sketch results

@@ -9,6 +9,7 @@ source(here::here("utils.R"))
 flat_params = c(0.5, 1)
 dist_levels = c(
   map_chr(flat_params, ~glue::glue("Beta({.x}, {.x})")),
+  "Beta(0.1, 1.9)",
   "ATACCC"
 )
 
@@ -49,30 +50,35 @@ beta_plot_dens = function(alpha, beta) {
   standard_plot_theming()
 }
 
-beta_plot_surv = function(alpha, beta, prior_len = 30) {
-  purrr::map2(
+beta_plot_surv = function(alpha, beta, prior_len = 30, include_ATACCC = FALSE) {
+  priors = purrr::map2(
     alpha, beta,
     ~sample_constant_prior(.x, .y, prior_len) |>
       mutate(dist = glue::glue("Beta({.x}, {.y})"))
   ) |>
-  bind_rows(
-    ataccc_posterior |>
-      mutate(dist = "ATACCC")
-  ) |>
-  group_by(dist, time) |>
-  mean_qi(S, .simple.names = FALSE) |>
-  mutate(dist = factor(dist, levels = dist_levels)) |>
-  ggplot(aes(time, S, ymin = S.lower, ymax = S.upper, group = dist)) +
-  geom_line(aes(colour = dist)) +
-  geom_ribbon(aes(fill = dist), alpha = 0.5) +
-  labs(
-    x = "Day",
-    y = "Survival",
-    colour = "Distribution",
-    fill = "Distribution"
-  ) +
-  coord_cartesian(c(0, prior_len)) +
-  standard_plot_theming()
+  bind_rows()
+  if (include_ATACCC) {
+    priors = bind_rows(
+      priors,
+      ataccc_posterior |>
+        mutate(dist = "ATACCC")
+    )
+  }
+  priors |>
+    group_by(dist, time) |>
+    mean_qi(S, .simple.names = FALSE) |>
+    mutate(dist = factor(dist, levels = dist_levels)) |>
+    ggplot(aes(time, S, ymin = S.lower, ymax = S.upper, group = dist)) +
+    geom_line(aes(colour = dist)) +
+    geom_ribbon(aes(fill = dist), alpha = 0.5) +
+    labs(
+      x = "Day",
+      y = "Survival",
+      colour = "Distribution",
+      fill = "Distribution"
+    ) +
+    coord_cartesian(c(0, prior_len)) +
+    standard_plot_theming()
 }
 
[…]

### 2023-06-06 cis-imperfect-testing.tex
Intro and structure for imperfect testing

@@
 
-\chapter{Survival analysis with misclassification} \label{imperf-test}
+\chapter{Survival analysis with false negatives} \label{imperf-test}
 
-\section{The issues of false negatives}
+In \cref{perf-test}, a framework for analysing the CIS to produce estimates of the duration of positivity was developed.
+However, initial application of this framework produced implausibly short estimates; I identified the issue as being due to the presence of false negatives, a form of misclassification bias, within the CIS (\cref{imperf-test:sec:issue}).
+There exists little prior work on the incorporation of misclassification bias within similar survival analysis frameworks.
+\todo[inline]{Check exactly what has been done and verify exactly what I can claim here}
+In this chapter, I modify the previous model to include false negatives (\cref{imperf-test:sec:modelling}).
+I show in simulation that this model will recover the duration distribution (\cref{imperf-test:sec:sim-study}) and apply it to the CIS data (\cref{imperf-test:sec:application}).
+Plausible estimates of the duration distribution are produced, however, they require assuming that the test sensitivity (the probability of testing positive given being detectable) is known; I discuss what would be required to relax this assumption in \cref{imperf-test:sec:discussion}.
+
+\section{The issues of false negatives} \label{imperf-test:sec:issue}
 
@@
 
-\section{Modelling}
+\section{Modelling} \label{imperf-test:sec:modelling}
 
@@
 Combining the above, the replacement for $p_{ia}$ is:
+\todo[inline]{Fix the length of the lines of these equations}
 \begin{align}
@@
 
-\section{Simulation study}
+\section{Simulation study} \label{imperf-test:sec:sim-study}
 
@@
 
-\section{Application to CIS data}
+\section{Application to CIS data} \label{imperf-test:sec:application}
 
 \subsection{Data}
-
-\section{Data}
 
@@
 
+\subsection{Methods}
+
+Think about what I need to include here.
+Might include hyperparameters etc or assumptions made?
+
+\subsection{Results}
+
+\section{Discussion} \label{imperf-test:sec:discussion}
+
+\section{Conclusion} \label{imperf-test:sec:conclusion}
+
 \end{document}

### 2023-06-07 cis-imperfect-testing.tex
Move test to right section
Outline chapter

@@
 In \cref{perf-test}, a framework for analysing the CIS to produce estimates of the duration of positivity was developed.
-However, initial application of this framework produced implausibly short estimates; I identified the issue as being due to the presence of false negatives, a form of misclassification bias, within the CIS (\cref{imperf-test:sec:issue}).
+However, initial application of this framework produced implausibly short estimates; I identified the issue as being due to the presence of false negatives, a form of misclassification bias, within the CIS (\cref{imperf-test:sec:problem}).
 There exists little prior work on the incorporation of misclassification bias within similar survival analysis frameworks.
@@
 
-\section{The issues of false negatives} \label{imperf-test:sec:issue}
+\section{The problem} \label{imperf-test:sec:problem}
 
-Demonstrate the likely issue of false negatives (Armitage talk).
+Applying the method of \cref{perf-test} to CIS data gave implausible estimates of the duration distribution (see \cref{imperf-test:fig:problem-cis-estimates}).
+In particular, I estimate a very high probability of the episode lasting less than five days, in conflict with the results of \cref{ATACCC} and other prior work (\eg: \autocite{cevikShedding}).
 
-Discuss a bit why this is a problem (single positives and bias).
+\begin{figure}
+  \centering %\includegraphic{}
+  \todo[inline]{Figure showing estimates without false negatives}
+  \caption{\label{imperf-test:fig:problem-cis-estimates}}
+\end{figure}
 
-Discuss very low sensitivity needed and changing sensitivity over time (ATACCC and CIS plots).
+Comparing the simulated and CIS data suggested that this is due to the high frequency of single positive episodes in the CIS data.
+A single positive episode is one in which the episode contains exactly one positive test, intuitively these cause short estimates because there is no lower bound on the length of time the episode lasts.
+Furthermore, due to the truncation, there is little or no ability to upper bound the number of very short episodes because these would almost all be missed.
+\Textcite{shenNonparametrica} (building on \textcite{panNote}) showed that in a related setting that the maximum likelihood (frequentist) estimator is inconsistent and can lead to severe underestimation (bias) of the survival function.
+While the result is not directly applicable, and can not simply be extended, to the more complex scenario within CIS, it does still build an intuition that these single positive episodes could be causing this issue.
+Furthermore, simulations building in even a small number of excess single positive episodes led to problematic estimation even within the CIS setting.
+\todo[inline]{My previous application and discussion of \textcite{shenNonparametrica} in this setting would need quite significant reworking to include; is this needed or is this slightly hand-wavy discussion enough?}
+
+Discussion and interrogation of the data led me to believe the most likely reason for an unexpectedly high number of single positive episodes was the presence of false negatives, meaning a false negative as a negative test result returned by a detectable individual.
+This can occur for a variety of reasons, most commonly that an individual swabs themselves poorly and hence the concentration of virus on the swab is too low to be detected by the PCR process.
+\todo{Maybe I need a better description of the swabbing process?}
+Plausibly, a variety of other reasons could also lead to a false negative such as logistical issues or mislabelling of samples.
+Consideration of these reasons are beyond the scope of this thesis, and I group all of these possible reasons together.
+
+There are two strands of evidence that support this hypothesis.
+First, it is well known that PCR testing can return false negatives \todo{refs for sensitivity} and I needed to explicitly include these in the viral load model of \cref{ATACCC}; they can be directly observed within CIS sa intermittent negatives (see below).
+Second, adjusting the simulation to include false negatives can reproduce both the CIS data more realisticly and the same issues when estimating the duration distribution (see \cref{imperf-test:sec:simulate})
+
+\emph{Intermittent negatives} are the clearest example of a false negative.
+An intermittent negative is when an individual tests negative but tested positive previously and subsequently.
+I explicitly included these in the viral load model of \cref{ATACCC} but ignored in \cref{perf-test}.
+Intermittent negatives stripped out when creating the dataset used for all duration analyses in this chapter (as described in \cref{episode-def}), but demonstrate that false negatives do occur.
+
+Whether a future positive is part of the same episode of a reinfection is not always trivial; I rely on a process developed previously by Sarah Walker, see \cref{episode-def}.
+However, up until the emergence of the Omicron variant, reinfections are rare, especially in a short time frame, and hence the tricky cases are rare until this time (late 2021).
+
+\section{Simulating false negatives} \label{imperf-test:sec:simulate}
+
+The simplest model for false negatives is by introducing a constant \emph{test sensitivity}, a probability of returning a negative test even when detectable.
+This can be simulated by...
+
+However, an implausibly low test sensitivity is required to reproduce the rate of single positive episodes seen in CIS.
+\todo[inline]{Figure demonstrating this and references to what reasonable test sensitivities are, including the estimate from the ATACCC chapter}
+More plausibly, the rate of false negatives varies over the course of an episode.
+False negatives are more likely to occur when the viral load of an individual is low, because there is less virus in their body to be sampled.
+Following the short, initial period of growth, viral load largely declines monotonically with time (see \cref{ATACCC}).
+\todo[inline]{Reference to ATCCCC chapter, although maybe the specific section with this result would be better}
+This suggests that a model with a declining test sensitivity as a function of time since infection might be more suitable.
+
+Examination of the CIS data also provides evidence for a declining test sensitivity.
+We can bound the test sensitivity by...
+\todo[inline]{Include the analysis of CIS data with bounding test sensitivity}
+
+A simple declining test sensitivty might be:
+\todo[inline]{Equation for linearly decreasing test sensitivity}
+Simulated data using this test sensitivity is much more similar ot the data, while being more plausible than very low probabilities of false negatives.
+
+\todo[inline]{Add the evidence that false negatives do create the ssame issue in inference in real and simulated data}
+
+\section{Modelling} \label{imperf-test:sec:modelling}
 
@@
 $p_{it}$ to allow for additional episodes being missed.
-
-\section{Modelling} \label{imperf-test:sec:modelling}
 
@@
 
-\subsection{Modifying $p_{ia}$} \label{modifying-p_ia}
+\subsection{Modifying \texorpdfstring{$p_{ia}$}{pia}} \label{modifying-p_ia}
 
@@
 
-\subsection{Modifying $p_{it}$} \label{modifying-p_it}
+\subsection{Modifying \texorpdfstring{$p_{it}$}{pit}} \label{modifying-p_it}
 
@@
 
-\section{Simulation study} \label{imperf-test:sec:sim-study}
-
-\subsection{Setup}
+\section{Simulation study results} \label{imperf-test:sec:sim-study-results}
 

### 2023-06-07 cis-imperfect-testing.tex
Write first half of chapter

@@
 \date{\today}
+
+\newcommand{\psens}{p_\text{sens}}
 
@@
 In this chapter, I modify the previous model to include false negatives (\cref{imperf-test:sec:modelling}).
-I show in simulation that this model will recover the duration distribution (\cref{imperf-test:sec:sim-study}) and apply it to the CIS data (\cref{imperf-test:sec:application}).
+I show in simulation that this model will recover the duration distribution (\cref{imperf-test:sec:sim-study-results}) and apply it to the CIS data (\cref{imperf-test:sec:application}).
 Plausible estimates of the duration distribution are produced, however, they require assuming that the test sensitivity (the probability of testing positive given being detectable) is known; I discuss what would be required to relax this assumption in \cref{imperf-test:sec:discussion}.
@@
 While the result is not directly applicable, and can not simply be extended, to the more complex scenario within CIS, it does still build an intuition that these single positive episodes could be causing this issue.
-Furthermore, simulations building in even a small number of excess single positive episodes led to problematic estimation even within the CIS setting.
+Furthermore, simulations building in even a few excess single positive episodes led to problematic estimation even within the CIS setting.
 \todo[inline]{My previous application and discussion of \textcite{shenNonparametrica} in this setting would need quite significant reworking to include; is this needed or is this slightly hand-wavy discussion enough?}
@@
 There are two strands of evidence that support this hypothesis.
-First, it is well known that PCR testing can return false negatives \todo{refs for sensitivity} and I needed to explicitly include these in the viral load model of \cref{ATACCC}; they can be directly observed within CIS sa intermittent negatives (see below).
-Second, adjusting the simulation to include false negatives can reproduce both the CIS data more realisticly and the same issues when estimating the duration distribution (see \cref{imperf-test:sec:simulate})
+First, it is well known that PCR testing can return false negatives \todo{refs for sensitivity} and I needed to explicitly include these in the viral load model of \cref{ATACCC}; they can be directly observed within CIS as intermittent negatives (see below).
+Second, adjusting the simulation to include false negatives can reproduce both the CIS data more faithfully and the same issues when estimating the duration distribution (see \cref{imperf-test:sec:simulate})
 
@@
 The simplest model for false negatives is by introducing a constant \emph{test sensitivity}, a probability of returning a negative test even when detectable.
-This can be simulated by...
+Denoting the test sensitivity by $\psens$, we replace the deterministic $y_i(t)$ with a random binary variable such that:
+\begin{equation}
+  \prob(y_i(t) = 1) \begin{cases}
+      \psens &b_i \leq t \leq e_i \\
+      0 &\text{otherwise}
+  \end{cases} 
+\end{equation}
+with $y_i(t)$ still only defined for $t \in t_i$.
 
-However, an implausibly low test sensitivity is required to reproduce the rate of single positive episodes seen in CIS.
+However, an implausibly low value for $\psens$ is required to reproduce the rate of single positive episodes seen in CIS.
 \todo[inline]{Figure demonstrating this and references to what reasonable test sensitivities are, including the estimate from the ATACCC chapter}
@@
 Following the short, initial period of growth, viral load largely declines monotonically with time (see \cref{ATACCC}).
-\todo[inline]{Reference to ATCCCC chapter, although maybe the specific section with this result would be better}
+\todo{Reference to ATCCCC chapter, although maybe the specific section with this result would be better}
 This suggests that a model with a declining test sensitivity as a function of time since infection might be more suitable.
@@
 Examination of the CIS data also provides evidence for a declining test sensitivity.
-We can bound the test sensitivity by...
+We can approximately bound the test sensitivity by assuming that all intermittent negatives are false negatives, that the negative following the last positive in an episode could be either a true or false negative, and all other negatives are true negatives.
+The point estimate for the test sensitivity is the true positives / (true positives + false negatives).
+In \cref{imperf-test:fig:bounding-cis-sensitivity} we consider the values produced by assuming the negative following the last positive in an episode could be either a true or false negative as a function of time since the individual was detected (\ie: the first positive test in the episode).
+While the bounds are broad, they clearly suggest a declining test sensitivity over the course of an episode.
+\begin{figure}
   \todo[inline]{Include the analysis of CIS data with bounding test sensitivity}
+  \caption{Bounding the test sensitivity using CIS data as a function of time since the infection was detected (see main text for explanation of how the bounds are formed).\label{imperf-test:fig:bounding-cis-sensitivity}}
+\end{figure}
 
-A simple declining test sensitivty might be:
-\todo[inline]{Equation for linearly decreasing test sensitivity}
-Simulated data using this test sensitivity is much more similar ot the data, while being more plausible than very low probabilities of false negatives.
+I propose the following model for test sensitivity compatible with \cref{imperf-test:fig:bounding-cis-sensitivity}, consisting of a linearly declining period then a constant period:
+\begin{equation}
+  p_\text{sens}(t) = \begin{cases}
+    0.9 - \frac{0.9-0.5}{50}t &t \leq 50 \\
+    0.5 &t > 50
+  \end{cases}
+\end{equation}
+where $t$ is the time in days since infection
+Simulated data using this test sensitivity is much more similar to the data, while being more plausible than very low probabilities of false negatives.
+\todo{Figure comparing the variable test sensitivity to fixed and CIS data}
 
@@
 
-Specifically, we introduce a constant test sensitivity $p_\text{sens}$
-such that:
-\begin{align}
-\prob(y_i(t) = 1) = \begin{cases} p_\text{sens} &b_i \leq t \leq e_i \\ 0 &\text{otherwise.}\end{cases}
-\end{align}
-
-In practice, it is likely that the test sensitivity is a function of the
-viral load of an individual (the concentration of viral material in the
-part of the body being sampled). For this initial attempt at modelling
-the sensitivity, we ignore this complication.
+The model is only extended to model a constant test sensitivity.
+Further work is required to introduce the more realistic varying test sensitivity (discussed in \cref{imperf-test:sec:discussion}).
 

### 2023-06-09 cis-perfect-testing.tex
Notes on changes needed
Edit perf chapter

@@
 I take a survival analysis approach to model the coarse sampling of the CIS (see \cref{perf-test:sec:problem}).
-The problem of priors can be important to the results here, due to the lack of information in the data; in addition, the prior can be used to combine the information in the CIS with the results of \cref{ATACCC} (see \cref{perf-test:sec:parameters-priors}).
+Flat priors can be unintentionally very informative, and hence weakly informative priors should be used; in addition, the prior can be used to combine the information in the CIS with the results of \cref{ATACCC} (see \cref{perf-test:sec:parameters-priors}).
 Through a simulation study (see \cref{perf-test:sec:simulation-study}), I show that these methods perform well, and in this context are not sensitive to model or prior specification.
@@
 
-Formally, the survival time function, parameterised by $\theta$, is the probability that an episode lasts at least $t$ days: $S_\theta(t) = \prob(D_i \geq t \mid B_i = b, \theta)$ where $D_i$ and $B_i$ are the duration and starting time of episode $i$ respectively.
-We make the standard assumption that $D_i$ and $B_i$ are independent and hence: $S_\theta(t) = \prob(D_i \geq t \mid \theta)$.
+The survival time function, parameterised by $\theta$, is the probability that an episode lasts at least $t$ days: $S_\theta(t) = \prob(D_i \geq t \mid B_i = b, \theta)$, where $D_i$ and $B_i$ are the duration and starting time of episode $i$ respectively.
+I make the standard assumption that $D_i$ and $B_i$ are independent and hence: $S_\theta(t) = \prob(D_i \geq t \mid \theta)$.
 
@@
 The \emph{initiating event} is when the duration starts, \ie becoming detectable.
-The \emph{terminating event} is when the duration starts, \ie becoming no longer detectable.
+The \emph{terminating event} is when the duration ends, \ie becoming no longer detectable.
 
@@
 An event is \emph{censored} if the time of the event can only be bounded; it is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
-Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa; \textcite{bogaertsSurvival} review many statistical methods applicable to interval censoring (see \cref{perf-test:fig:double-interval-censor}).
+Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa (see \cref{perf-test:fig:double-interval-censor}); \textcite{bogaertsSurvival} review many statistical methods applicable to interval censoring.
 \emph{Double censoring} means that both the originating and terminating event are censored, and is also a well-studied problem, reviewed by \textcite{sunStatisticala}.
 In CIS, we have \emph{double interval censoring} meaning that both the initiating and the terminating events are interval censored.
+\todo{double interval censoring refs}.
 \begin{figure}
@@
   \caption{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval.
-Demonstrated here by a participant who transitions negative at time 7 to positive at time 14, bounding the start of the episode; similarly, the transition positive at time 28 to negative at time 56 bounds the end of the episode.
+Demonstrated here by a participant who is recorded negative at time 7 to positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode.
 \label{perf-test:fig:double-interval-censor}}
 \end{figure}
-
-Interval censoring is a well-studied area in the biostatistics and broader statistical literature\todo{interval censoring refs}.
-Double interval censoring is less frequent, although has been considered previously\todo{double interval censoring refs}.
 
@@
 
-Truncation is a large problem when individuals move to the infrequent testing schedule within CIS.
+Truncation is a large problem within CIS, especially when individuals move to the more infrequent testing schedule (described in \cref{intro:sec:cis}).
+\todo{ref to description of CIS}
 Consider that, based on the estimates in \cref{ATACCC}, the median length of an episode is 14 days \todo{check this number} we would not expect many episodes to last the 28 days that can occur between tests.
@@
 This combination of censoring and truncation are rarely encountered in biomedical data\todo{refs to combining double interval censoring with truncation}.
+A more familiar setting for this work
 
@@
 \section{Modelling the duration}\label{perf-test:sec:model}
-Now I develop a model accounting for double interval censoring and truncation, building on prior work in both biomedical and bird nesting studies.
+In this section, I develop a model accounting for double interval censoring and truncation, building on prior work in both biomedical and bird nesting studies.
 The arbitrary nature of truncation in CIS, where each individual has a unique truncation pattern and is neither simple left nor right truncation, has not previously been explored in the literature.
-Furthermore, I explore a Bayesian framework where the amount of explicit data augmentation is limited allowing the method to scale to the large number of episodes present in CIS without requiring vast computing resources (such as a high-performance cluster).
-Only limited resources are available within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
-Therefore, unlike most Bayesian methods (e.g.: \textcite{heBayesiana}, \textcite{heBayesian}, and \textcite{caoModeling}) we avoid augmenting the data with unobserved times of infection or any information on the truncated infections; this is due to the large number of infections (>4000 in the time period considered) making this approach computationally expensive.
+Furthermore, I explore a Bayesian framework where the amount of explicit data augmentation is limited allowing the method to scale to the large number of episodes present in CIS without requiring vast computing resources (such as a high-performance cluster), as only limited resources are available within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
+Therefore, unlike most Bayesian methods (e.g.: \textcite{heBayesiana}, \textcite{heBayesian}, and \textcite{caoModeling}) we avoid augmenting the data with unobserved times of infection or any information on the truncated infections; this would be computationally expensive with the large number of episodes in CIS (>4000 in the time period considered in the main analysis of \cref{imperf-test}).
 
@@
 \includegraphics[width=\textwidth]{cis-perfect-testing/regions_diag}
+\todo[inline]{Fix the impossible region to fill out the RHS; label the inadmissible regoin.}
 \caption{Admissible, $\alpha_i$, inadmissible, $\beta_i$ and
@@
 \subsection{Independent priors}
-
-A common assumption throughout Bayesian analyses is independence of the parameters in the prior.
-This is convenient because we can then use univariate distributions.
-\todo{This feels a bit clumsy/patronising to the reader. Maybe change the phrasing?}
-A Beta distribution for each hazard parameter, $\lambda_t$, is a natural choice due to it being a probability.
+Take the standard assumption of independent priors on each parameter.
+A Beta distribution is then a natural choice for each hazard parameter, $\lambda_t$, as they are probabilities.
 
 An uninformative prior would be $\lambda_t \dist \Beta(\alpha, \alpha)$, with common choices of $\alpha$ being $0.5$, Jeffreys' Prior, or $1$, a flat prior.
-Flat or uninformative priors on the hazard lead to very informative priors on the survival, favouring short survival times.
+Flat or uninformative priors on the hazard lead to very informative priors on the survival, favouring short survival times (see \cref{perf-test:fig:flat-prior}).
 The sharp falloff can be seen by noting that these priors have expectation 0.5, hence $\E(S(t)) = \E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right) = 0.5^{t-1}$ which decreases very quickly.
-Our prior belief is not well-represented by this prior, and in fact a priori viewed as extremely unlikely (see \cref{perf-test:fig:flat-prior}),
+This prior expresses extreme scepticism in estimates from previous studies, such as \cref{ATACCC} and the meta-analysis of \textcite{cevikShedding}.
 Therefore, I favour a prior which is weakly informative on the survival, or an informative prior used to represent our belief from the \cref{ATACCC} analysis.
@@
 
-The first step, approximating the \cref{ATACCC} posterior as a multivariate normal, is done through the method of moments.
+The first step, approximating the \cref{ATACCC} posterior hazard as a multivariate normal on the logit scale, is done through the method of moments.
 The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC}).
@@
 \subsection{Smoothing priors}
-
-A final possible prior are ones to favour smoothness in the hazard.
-This is the Bayesian equivalent of a penalised likelihood, used frequently in the non-parametric frequentist setting.
+The final class of priors I consider are ones that aim to favour smoothness in the hazard.
+This is the Bayesian equivalent of a penalised likelihood, used frequently in the non-parametric frequentist setting (\eg: \autocite{bacchettiNonparametric}).
 Biological and epidemiological considerations mean that it would be surprising if the hazard varied significantly day-to-day.
@@
 Denote by $f_A(t)$ the distribution function estimated from ATACCC and $f_C(t)$ that from these CIS-based estimates.
-Define:
+Then define:
 $$
@@
   \item Other priors (GPs?)
-  \item Inclusion of covariates
+  \item Inclusion of covariates: age, previous exposure status, variant.
   \item Individual vs total model

### 2023-06-09 cis-imperfect-testing.tex
Fix cross-file referencing

@@
 
-\newcommand{\psens}{p_\text{sens}}
-
 \begin{document}
@@
 
-In \cref{perf-test}, a framework for analysing the CIS to produce estimates of the duration of positivity was developed.
+In \cref{E-perf-test}, a framework for analysing the CIS to produce estimates of the duration of positivity was developed.
 However, initial application of this framework produced implausibly short estimates; I identified the issue as being due to the presence of false negatives, a form of misclassification bias, within the CIS (\cref{imperf-test:sec:problem}).
@@
 Applying the method of \cref{perf-test} to CIS data gave implausible estimates of the duration distribution (see \cref{imperf-test:fig:problem-cis-estimates}).
-In particular, I estimate a very high probability of the episode lasting less than five days, in conflict with the results of \cref{ATACCC} and other prior work (\eg: \autocite{cevikShedding}).
+In particular, I estimate a very high probability of the episode lasting less than five days, in conflict with the results of \cref{E-ATACCC} and other prior work (\eg: \autocite{cevikShedding}).
 
@@
 Discussion and interrogation of the data led me to believe the most likely reason for an unexpectedly high number of single positive episodes was the presence of false negatives, meaning a false negative as a negative test result returned by a detectable individual.
-This can occur for a variety of reasons, most commonly that an individual swabs themselves poorly and hence the concentration of virus on the swab is too low to be detected by the PCR process.
-\todo{Maybe I need a better description of the swabbing process?}
+This can occur for a variety of reasons, most commonly that an individual swabs themselves poorly and hence the concentration of virus on the swab is too low to be detected by the PCR process (see \cref{E-intro:sec:PCR}).
+\todo{Ref to section in intro explaining PCR testing}
 Plausibly, a variety of other reasons could also lead to a false negative such as logistical issues or mislabelling of samples.
@@
 There are two strands of evidence that support this hypothesis.
-First, it is well known that PCR testing can return false negatives \todo{refs for sensitivity} and I needed to explicitly include these in the viral load model of \cref{ATACCC}; they can be directly observed within CIS as intermittent negatives (see below).
-Second, adjusting the simulation to include false negatives can reproduce both the CIS data more faithfully and the same issues when estimating the duration distribution (see \cref{imperf-test:sec:simulate})
+First, it is well known that PCR testing can return false negatives \todo{refs for sensitivity} and I needed to explicitly include these in the viral load model of \cref{E-ATACCC}; they can be directly observed within CIS as intermittent negatives (see below).
+Second, adjusting the simulation to include false negatives can reproduce both the CIS data more faithfully and exhibits the same issues when estimating the duration distribution (see \cref{imperf-test:sec:simulate})
 
@@
 An intermittent negative is when an individual tests negative but tested positive previously and subsequently.
-I explicitly included these in the viral load model of \cref{ATACCC} but ignored in \cref{perf-test}.
-Intermittent negatives stripped out when creating the dataset used for all duration analyses in this chapter (as described in \cref{episode-def}), but demonstrate that false negatives do occur.
+I explicitly included these in the viral load model of \cref{E-ATACCC} but ignored in \cref{E-perf-test}.
+Intermittent negatives are stripped out when creating the dataset used for all duration analyses in this chapter (as described in \cref{E-episode-def}), but demonstrate that false negatives do occur.
 
-Whether a future positive is part of the same episode of a reinfection is not always trivial; I rely on a process developed previously by Sarah Walker, see \cref{episode-def}.
+Whether a future positive is part of the same episode of a reinfection is not always trivial; I rely on a process developed previously by Sarah Walker, see \cref{E-episode-def}.
 However, up until the emergence of the Omicron variant, reinfections are rare, especially in a short time frame, and hence the tricky cases are rare until this time (late 2021).
@@
 False negatives are more likely to occur when the viral load of an individual is low, because there is less virus in their body to be sampled.
-Following the short, initial period of growth, viral load largely declines monotonically with time (see \cref{ATACCC}).
+Following the short, initial period of growth, viral load largely declines monotonically with time (see \cref{E-ATACCC}).
 \todo{Reference to ATCCCC chapter, although maybe the specific section with this result would be better}
@@
 
-I use the CIS data described in \cref{intro:sec:cis}, with positive tests heuristically divided into episodes.
-Prior work categorised a series of positive tests within the same individual CIS were heuristically categorised into episodes (see \cref{episode-def}).
+I use the CIS data described in \cref{E-intro:sec:cis}, with positive tests heuristically divided into episodes.
+Prior work categorised a series of positive tests within the same individual CIS were heuristically categorised into episodes (see \cref{E-episode-def}).
 

### 2023-06-09 thesis.tex
Misc edits / bugfixes
Work on priors
Fix cross-file referencing

--- 2023-05-31 main.tex
@@
 \usepackage{bm}
+\usepackage[nameinlink]{cleveref}
 \usepackage[style=british]{csquotes}
@@
 \usepackage{todonotes}
+\usepackage{xr}
 % \usepackage{microtype}
@@
 \def\@onedot{\ifx\@let@token.\else.\null\fi\xspace}
-
-\def\eg{\emph{e.g}\onedot} \def\Eg{\emph{E.g}\onedot}
-\def\ie{\emph{i.e}\onedot} \def\Ie{\emph{I.e}\onedot}
-\def\cf{\emph{c.f}\onedot} \def\Cf{\emph{C.f}\onedot}
-\def\etc{\emph{etc}\onedot} \def\vs{\emph{vs}\onedot}
+\def\eg{e.g\onedot} \def\Eg{\emph{E.g}\onedot}
+\def\ie{i.e\onedot} \def\Ie{\emph{I.e}\onedot}
+\def\cf{c.f\onedot} \def\Cf{\emph{C.f}\onedot}
+\def\etc{etc\onedot} \def\vs{\emph{vs}\onedot}
 \def\wrt{w.r.t\onedot} \def\dof{d.o.f\onedot}
-\def\etal{\emph{et al}\onedot}
+\def\etal{et al\onedot}
 \makeatother
 
+% Generic maths commands
 \def\reals{\mathbb{R}}
@@
 \DeclareMathOperator{\cov}{Cov}
+\DeclareMathOperator{\logit}{logit}
 \DeclareMathOperator{\Ber}{Bernoulli}
@@
 \DeclareMathOperator{\Poi}{Poisson}
+\DeclareMathOperator{\Beta}{Beta}
+\DeclareMathOperator{\NegBin}{NegBin}
+\DeclareMathOperator{\GamDist}{Gamma}
 \def\matr{\bm}
@@
 \renewcommand{\vec}{\bm}
+
+% Thesis-specific maths commands
+\newcommand{\psens}{p_\text{sens}}
 
@@
 \addbibresource{references.bib}
+\externaldocument[E-]{\subfix{thesis}}
 
@@
 
+\usepackage{geometry}
+\setlength{\marginparwidth}{3cm}
+\geometry{margin=3.8cm}
+
 \begin{document}
@@
 
-\section{Producing duration from ATACCC}
+\section{Producing duration from parameter estimates}
 
@@
 Taking a posterior sample for this, $\phi^{(i)}$, we can then draw the parameters which determine the viral
-load trajectory for a random individual $j$, $\theta_j$, from which that individual's duration $d_j^{(i)}$ can be calculated ([see this Overleaf paper for further details](https://www.overleaf.com/9549594748zbvpgjgvmgsd)).
+load trajectory for a random individual $j$, $\theta_j$, from which that individual's duration $d_j^{(i)}$ can be calculated.
 By drawing $\theta_1, \theta_2, \dots, \theta_n$ from the distribution $p(\theta_j \mid\phi^{(i)})$, the density of the duration distribution, $f_A(t)$, can be
@@
 
+\subfile{appendix-cis-perfect-testing}
+
+\chapter{CIS episode definitions} \label{episode-def}
+
+\todo[inline]{Get this from SARAH}
+
 \chapter{Text to use}
@@
 The CIS (Coronavirus Infection Survey) is a longitudinal study on a representative sample of households.
-A full description of the study can be found in the supplementary materials of \textcite{pouwelsCommunity} or the study protocol~\autoref{cisProtocol}.
+A full description of the study can be found in the supplementary materials of \textcite{pouwelsCommunity} or the study protocol~\cref{cisProtocol}.
 In brief, households are invited to the study from databases held by the ONS.
@@
 
-\chapter{CIS episode definitions} \label{episode-def}
-
-\todo[inline]{Get this from SARAH}
-
 \end{document}

### 2023-06-09 appendix-cis-perfect-testing.tex
Fix cross-file referencing

--- 2023-06-02 cis-perfect-testing.tex
@@
 \begin{document}
-\chapter{Survival analysis with perfect testing} \label{perf-test}
 
-The estimation of incidence from prevalence is sensitive to the tail of the duration distribution.
-In \cref{ATACCC} I estimated the duration distribution from the ATACCC data, however, this study suffered from a small sample size and only 20 days of follow-up.
-\todo{Ref ATACCC chapter and in following paragraph}
-Therefore, the estimation of the tail of the duration distribution is driven by model assumptions.
-This chapter develops methods to estimate the distribution using CIS data, which has less frequent sampling but a large sample size and long follow-up.
+\chapter{Including all CIS individuals in survival analysis} \label{appendix:total-model}
 
-I take a survival analysis approach to model the coarse sampling of the CIS (see \cref{perf-test:sec:problem}).
-The problem of priors can be important to the results here, due to the lack of information in the data; in addition, the prior can be used to combine the information in the CIS with the results of \cref{ATACCC} (see \cref{perf-test:sec:parameters-priors}).
-Through a simulation study (see \cref{perf-test:sec:simulation-study}), I show that these methods perform well, and in this context are not sensitive to model or prior specification.
-However, application in reality must account for misclassification bias, the subject of \cref{imperf-test}.
-\todo{Ref chapter incl misclassification}
-
-\todo[inline]{%
-  Somehwere, I need to fit in what exactly is novel here, and issues with the pre-existing work.
-What to cover is the application to both double interval censored and arbitrarily truncated data (which existed in some of the theoretical frameworks but not used); the issues with the conditional likelihood method of \textcite{bacchettiNonparametric}; and the lack of much work this century.
-}
-
-\section{Problem description and approach} \label{perf-test:sec:problem}
-
-\todo[inline]{Some of these definitions likely need to be moved elsewhere as they will be needed in other chapters (e.g. ATACCC analysis).}
-Consider the CIS data.
-It consists of a series of tests on the same individuals (longitudinal data), where some individuals have a series of positive tests.
-We wish to estimate the distribution of time for which an individual is positive for.
-An individual could have multiple infections, and hence I refer to \emph{infection episodes}.
-
-I define an individual as \emph{detectable} if they would test positive if tested, in the absence of misclassification error (which is dealt with in \cref{imperf-test}).
-\todo{Ref chapter incl misclassification}
-Episode $i$'s \emph{duration of positivity}, $D_i$ is the number of days that they are detectable for; I use time discretised to days, therefore, the duration is at least 1 day.
-Note that the relationship between positivity (\ie the result of a PCR test) and infectiousness (\ie an individual's ability to pass on the virus) is complex, \todo{ref for PCR pos vs infectiousness} and beyond the scope of this thesis.
-I define an episode as starting at time $B_i$ and ends at $E_i$, hence $D_i = E_i - B_i + 1$.
-
-An episode is \emph{detected} when they test positive at least once, if this occurs; episode $i$ is detected at time $t$ if and only if: $i$ is detectable at $t$, $i$ is tested at $t$, and the test returns positive (a false negative does not occur).
-
-Formally, the survival time function, parameterised by $\theta$, is the probability that an episode lasts at least $t$ days: $S_\theta(t) = \prob(D_i \geq t \mid B_i = b, \theta)$ where $D_i$ and $B_i$ are the duration and starting time of episode $i$ respectively.
-We make the standard assumption that $D_i$ and $B_i$ are independent and hence: $S_\theta(t) = \prob(D_i \geq t \mid \theta)$.
-
-I utilise survival analysis for analysing this data.
-Survival analysis is the area of statistics concerned with estimating the distribution of the times between events.
-It is broadly applicable to many domains, such as the time-to-failure for a mechanical system or the effect of a treatment on time spent in hospital.
-In the context of this thesis, the distribution to estimate is that of duration of positivity.
-The \emph{initiating event} is when the duration starts, \ie becoming detectable.
-The \emph{terminating event} is when the duration starts, \ie becoming no longer detectable.
-
-The reason for this approach, over modelling biomarkers (as in \cref{ATACCC}) is due to two factors.
-\todo{Ref ATACCC chapter}
-First, the large sample size in CIS means that approaches where the number of parameters scales with the number of observed infections (as it does with random effect based approaches) would be computationally challenging; this challenge is compounded by the data being stored in a trusted research environment limiting the computational power and tools available.
-Second, the data is coarse with most episodes having only a single positive test; therefore, the trajectory of the viral load would not be well-identified in an individual.
-Combined, these factors outweighed the downside of survival analysis only considering the binary outcome (\ie positive or negative) or the test rather than the additional information on the viral load.
-
-The methods in this chapter deal with modelling two elements of the data-generating process, double interval censoring and truncation.
-False negatives are not dealt with and left to \cref{imperf-test}.
-\todo{Ref chapter incl misclassification}
-
-Double interval censoring is a special case of censoring, a common and well-studied challenge encountered when dealing with survival data.
-An event is \emph{censored} if the time of the event can only be bounded; it is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
-Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa; \textcite{bogaertsSurvival} review many statistical methods applicable to interval censoring (see \cref{perf-test:fig:double-interval-censor}).
-\emph{Double censoring} means that both the originating and terminating event are censored, and is also a well-studied problem, reviewed by \textcite{sunStatisticala}.
-In CIS, we have \emph{double interval censoring} meaning that both the initiating and the terminating events are interval censored.
-\begin{figure}
-  \centering \includegraphics{cis-perfect-testing/double-interval-censor}
-  \caption{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval.
-Demonstrated here by a participant who transitions negative at time 7 to positive at time 14, bounding the start of the episode; similarly, the transition positive at time 28 to negative at time 56 bounds the end of the episode.
-\label{perf-test:fig:double-interval-censor}}
-\end{figure}
-
-Interval censoring is a well-studied area in the biostatistics and broader statistical literature\todo{interval censoring refs}.
-Double interval censoring is less frequent, although has been considered previously\todo{double interval censoring refs}.
-
-The second element is truncation.
-Truncation occurs because we do not observe some episodes: an episode can occur between tests and hence the episode is never detected (see \cref{perf-test:fig:truncation}).
-We do not know how many infections we do not detect.
-Failure to model truncation will lead to estimating too long durations because the detected episodes are a biased sample with a longer duration.
-To see why they are shorter, consider the undetected infection in the bottom of \cref{perf-test:fig:truncation}.
-This individual was infected at time 33 for 10 days.
-If instead there infection was 24 days (or longer), the individual would still have been detectable at their next test, and therefore would have been detected.
-
-Truncation is a large problem when individuals move to the infrequent testing schedule within CIS.
-Consider that, based on the estimates in \cref{ATACCC}, the median length of an episode is 14 days \todo{check this number} we would not expect many episodes to last the 28 days that can occur between tests.
-\todo{Ref ATACCC chapter}
-\begin{figure}
-  \centering \includegraphics{cis-perfect-testing/truncation}
-  \caption{Episodes data in CIS is truncated, meaning that an unknown number of infections are never detected.
-Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was.
-See the main text for why this is an issue.
-\label{perf-test:fig:truncation}}
-\end{figure}
-
-This combination of censoring and truncation are rarely encountered in biomedical data\todo{refs to combining double interval censoring with truncation}.
-
-While this chapter sets up the framework for analysing the CIS data, applying the method to CIS data requires modelling false negatives.
-That is the subject of \cref{imperf-test} and hence a full description of the data and processing of it is deferred to that chapter.
-\todo{Ref chapter incl misclassification}
-
-\section{Modelling the duration}\label{perf-test:sec:model}
-Now I develop a model accounting for double interval censoring and truncation, building on prior work in both biomedical and bird nesting studies.
-The arbitrary nature of truncation in CIS, where each individual has a unique truncation pattern and is neither simple left nor right truncation, has not previously been explored in the literature.
-Furthermore, I explore a Bayesian framework where the amount of explicit data augmentation is limited allowing the method to scale to the large number of episodes present in CIS without requiring vast computing resources (such as a high-performance cluster).
-Only limited resources are available within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
-Therefore, unlike most Bayesian methods (e.g.: \textcite{heBayesiana}, \textcite{heBayesian}, and \textcite{caoModeling}) we avoid augmenting the data with unobserved times of infection or any information on the truncated infections; this is due to the large number of infections (>4000 in the time period considered) making this approach computationally expensive.
-
-I model only individuals with detected episodes.
-This approach is based on long-standing methodology~\autocites{heiseyModelling}{dempsterMaximum}{turnbullEmpirical}.
-The conceptual approach here is to form a cohort of individuals for which we observe at least one positive test, and then consider a study which had only enrolled these individuals.
-The modelling approach then corrects for this selection bias, by considering \emph{ghost} infections that occur in identical individuals but were truncated.
-
-I have also considered a model including individuals without any observed episodes (see \cref{appendix:total-model} for details).
-These individuals may have truncated episodes or may not have been infected at all.
-However, I found that this model has little difference in results while increasing computational complexity and requiring assumptions on reinfection (discussed in \cref{perf-test:sec:discussion}).
-
-\subsection{Notation and assumptions}
-\todo[inline]{Some of this will possibly be defined elsewhere, such as in the introduction where CIS is introduced}
-\begin{itemize}
-\item
-  Time is discrete: $1, \dots, T$.
-\item
-  We observe $n_a$ episodes of infection, indexed by $i$.
-\item
-  The $i$ in which episode $i$ occurs is tested at times
-  $t_i = \{ t_{i,1}, \dots, t_{i,m_i} \}$.
-\item
-  Each observed episode has an unknown start of infection time $b_i$
-  and end of infection time $e_i$ (the first and last day that the
-  individual would test positive due to this episode respectively).
-  These are realisations of the random variables $B_i$ and $E_i$
-  respectively.
-\item
-  The duration of the infection is the number of days for which an
-  individual tests positive, $D_i = E_i - B_i + 1$.
-\item
-  We assume that, for all episodes $i$, $D_i$ is iid and independent
-  of the time of the infection. Define the survival function
-  $\prob(D_i \geq t \mid B_i = b, \theta) = \prob(D_i \geq t \mid \theta) = S_\theta(t)$,
-  where $\theta$ are the parameters controlling the survival
-  distribution (the discussion in this document is valid regardless of
-  the model specified for $S_\theta$ and hence we consider $\theta$
-  as an arbitrary vector of parameters).
-\item
-  The beginning of episode $i$ is known to occur in the interval
-  $[l_i^{(b)}, r_i^{(b)}]$, and similarly for the end of the infection
-  in $[l_i^{(e)}, r_i^{(e)}]$.
-\item
-  $y_i(t)$ for $t \in t_i$ is a binary indicator giving the test
-  result for individual $i$ at time $t$. Under the perfect testing
-  assumption, $y_i(t) = 1$ if and only if $b_i \leq t \leq e_i$.
-\item
-  Throughout, the convention that lower-case letters are realisations of
-  upper-case random variables is used.
-\end{itemize}
-
-\subsection{Likelihood}\label{perf-test:sec:likelihood}
-
-Episode $i$ can be fully characterised by the pair $(b_i, e_i)$ of
-start and end dates of the episode, which belong to the state space
-$T \times T$. If the individual in which the episode occurs is tested
-between $b_i$ and $e_i$ (inclusive), then we observe the episode,
-otherwise it is truncated. For the $i$th episode that we observed, we
-consider there is an unknown number, $n_{it}$, where $(b, e)$ are
-drawn from the same distribution as those that lead to episode $i$,
-and occur in an (imagined) individual identical to $i$, but were
-truncated. Episodes are also truncated if there is no negative test
-prior to their start, that is, the individual the episode occurred in
-had no tests before $b_i$. The truncated episodes are the ghosts of $i$.
-
-\Textcite{heiseyModelling} provide a framework for formalising this.
-Define three sets which partition the space of possible infection and
-recovery (\ie: are strict subsets of $T \times T$) given the
-observations associated with episode $i$ (graphically shown in Figure
-\ref{perf-test:fig:partitionSpace}). Each episode, whether observed or not, must
-fall into one of these three classes with a probability that is a
-function of $\theta$ (see section \ref{perf-test:sec:likelihood}).
-
-\begin{itemize}
-\item
-  Admissible episodes, $\alpha_i$, which have an infection and
-  recovery time within their respective intervals observed.
-  $n_{ia} =1$ of these occurred, each with probability
-  $p_{ia} = \prob((b, e) \in \alpha_i \mid \theta)$.
-\item
-  Truncated episodes, $\Omega_i^C$, which have an infection and
-  recovery time such that they would not have tested positive. An
-  unknown number, $n_{it}$ of these observed, each with probability
-  $p_{it} = \prob((b, e) \in \Omega^C_i \mid \theta)$.
-\item
-  Inadmissible episodes, $\beta_i$, which do not have an infection and
-  recovery time within their respective intervals but would have been
-  observed (not truncated). This is all remaining episodes not in the
-  previous sets. $n_{iu} = 0$ of these occurred, where they would have
-  occurred with probability
-  $p_{iu} = \prob((b, e) \in \beta_i \mid \theta)$.
-\end{itemize}
-
-The untruncated region (where we could have observed an episode) is
-$\Omega_i = \alpha_i \cup \beta_i$ (the notation here is used for
-consistency with \textcite{heiseyModelling}), and is the
-complement of $\Omega^C_i$.
-
-\begin{figure}
-\includegraphics[width=\textwidth]{cis-perfect-testing/regions_diag}
-\caption{Admissible, $\alpha_i$, inadmissible, $\beta_i$ and
-untruncated, $\Omega_i^C$ regions for an episode $i$ in an
-individual whose first test was at time 0 and negative, with subsequent
-negative tests at times 7, 14, and 56, and positive tests at times 21
-and 28. \label{perf-test:fig:partitionSpace}}
-\end{figure}
-
-These three classes span all possible episodes and are mutually
-exclusive, and hence the total number of episodes,
-$n_i = n_{ia} + n_{iu} + n_{it}$, and
-$p_{ia} + p_{iu} + p_{it} = 1$.
-
-Episode $i$ and its ghosts independently occur belong to one of these
-classes. Therefore, conditional on $n_i$, the number of episodes in
-each class (that is, the counts $n_{ia}, n_{iu}, n_{it}$) are
-distributed multinomially. Hence:
-\begin{align}
-&p(n_{ia} = 1, n_{iu} = 0, n_{it} = n_i - 1 \mid n_i, \theta) \\
-&= \frac{n_i!}{n_{ia}! n_{iu}! (n_i- n_{ia} - n_{it})!} p_{ia}^{n_{ia}} p_{ua}^{n_{ia}} p_{it}^{n_i- n_{ia} - n_{it}} \\
-&= \frac{n_i!}{(n_i-1)!} p_{ia} p_{it}^{n_i- 1} &\text{as $n_{ia} = 1$ and $n_{iu} = 0$}\\
-&= n_i p_{ia} p_{it}^{n_i- 1}
-\end{align}
-
-The posterior we are interested in is
-$p(\theta \mid n_{ia} = 1, n_{iu} = 0, \dots, n_{n_a,a} = 1, n_{n_a,u} = 0)$,
-where $\theta$ are the parameters governing the survival distribution
-and hence allow derivation of the $p$s.
-\begin{align}
-&p(\theta \mid n_{ia} = 1, n_{iu} = 0, \dots, n_{n_a,a} = 1, n_{n_a,u} = 0) \\
-&\propto p(\theta) \prod_{i=1}^{n_a} p(n_{ia} = 1, n_{iu} = 0 \mid \theta) \\
-&= p(\theta) \prod_{i=1}^{n_a} \sum_{n_i=1}^\infty p(n_{ia} = 1, n_{iu} = 0, n_{it} = n_i - 1 \mid \theta, n_i) p(n_i \mid \theta) \\
-&= p(\theta) \prod_{i=1}^{n_a} \sum_{n_i=1}^\infty n_i p_{ia} p_{it}^{n_i- 1} p(n_i) \\
-&= p(\theta) \prod_{i=1}^{n_a} p_{ia} \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i)
-\end{align}
-where the last line follows by assuming prior independence.
-
-In the remainder of the section, we derive the expressions for $p_{ia}$ and
-$p_{it}$ in terms of the data, and analytical solutions for the sum
-under different priors.
-
-I derive an analytical solution to $\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i)$ under two different assumptions for $p(n_i)$.
-First, we consider the improper prior $p(n_i) \propto 1/n_i$, this prior has several attractive properties.
-Then, we consider a negative binomial distribution (which includes the geometric and Poisson distributions as a special and limiting case respectively).
-
-The prior $p(n_i) \propto 1/n_i$ means that the model coincides with the frequentist conditional likelihood approach, up to the priors on other parameters~\cites[section 4.2]{dempsterMaximum}{heiseyModelling}[section 8.7.5]{gelmanBayesian}.
-Furthermore, this is the reference prior for this quantity\autocite{heBayesiana}.
-Under this prior,
-$\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) \propto \sum_{n_i=1}^\infty p_{it}^{n_i-1} = 1/(1-p_{it})$.
-% (\protect\hyperlink{ref-dempsterMaximum}{Dempster, Laird, and
-% Rubin} (\protect\hyperlink{ref-dempsterMaximum}{1977}) section 4.2;
-% \protect\hyperlink{ref-heiseyModelling}{Heisey and Nordheim}
-% (\protect\hyperlink{ref-heiseyModelling}{1995});
-% \protect\hyperlink{ref-gelmanBayesian}{Gelman et al.}
-% (\protect\hyperlink{ref-gelmanBayesian}{2013}) section 8.7.5)
-
-Now consider $N_i \dist \NegBin(\mu, r)$ using the
-mean/dispersion parameterisation of the negative binomial. Equivalently:
-\begin{align}
-N_i \mid \lambda &\dist \Poi(\lambda) \\
-\lambda &\dist \GamDist(a, b)
-\end{align}
-where $b = r / \mu$ and $a = r$.
-Hence:
-\begin{align}
-\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) 
-&= \int \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i \mid \lambda) p(\lambda) d\lambda \\
-&= \int \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} \frac{\lambda^{n_i} e^{-\lambda}}{n_i!} p(\lambda) d\lambda \\
-&= \int e^{-\lambda} \sum_{n_i=1}^\infty p_{it}^{n_i- 1} \frac{\lambda^{n_i-1}\lambda }{(n_i-1)!} p(\lambda) d\lambda \\
-&= \int e^{-\lambda} \lambda p(\lambda) \sum_{n_t=0}^\infty p_{it}^{n_t} \frac{\lambda^{n_t} }{n_t!} d\lambda &n_t = n_i - 1 \\
-&= \int e^{-\lambda} \lambda p(\lambda) e^{p_{it}\lambda} d\lambda &\text{Poisson pmf} \\
-&= \int e^{\lambda (p_{it} - 1)} \lambda p(\lambda) d\lambda \\
-&= \int e^{\lambda (p_{it} - 1)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda \\
-&= \frac{\Gamma(a+1)b^a}{\Gamma(a) (b+1-p_{it})^{a+1}} \\ 
-  &\; \times \int \frac{(b+1-p_{it})^{a+1}}{\Gamma(a+1)} \\
-  &\; \lambda^{(a+1)-1} e^{-(b+1-p_{it})\lambda} d\lambda \\
-&= \frac{\Gamma(a+1)b^a}{\Gamma(a) (b+1-p_{it})^{a+1}} &\text{as gamma pdf} \\
-&\propto (b+1-p_{it})^{-(a+1)} \\
-&= (r/\mu+1-p_{it})^{-(r+1)} \\
-&\propto (r+\mu (1-p_{it}))^{-(r+1)}
-\end{align}
-Note that $r=0$ recovers the previous derivation (up to
-proportionality).
-
-The above allows us to sample from the posterior without needing to
-sample the $n_i$s. This is convenient because there are as many
-$n_i$s as episodes observed and hence is computationally costly to
-sample. Furthermore, each $n_i$ is a discrete parameter, and hence
-cannot be sampled within the current Stan implementation.
-
-Having estimated the posterior of the other parameters, we might want to
-consider the posterior of $n_i$ (normally for diagnostic or model
-debugging purposes), which we can sample from the full conditional.
-Specifically, for each posterior sample of $\theta$, we sample one
-draw from $n_i \mid \theta, y$ where $y$ represents all the data.
-
-Due to the conditional independence of the $n_i$s, we find that:
-\begin{align}
-&p(n_i \mid \theta, n_{ia} = 1, n_{iu} = 0) \\
-&\propto p(n_{ia} = 1, n_{iu} = 0, n_{it} = n_i - 1 \mid n_i, p_{iu}, p_{ia}, p_{it}) p(n_i) \\
-&\propto n_i p_{it}^{n_i- 1} p(n_i)
-\end{align}
-With the prior, $p(n_i) \propto 1/n_i$ then this is a geometric
-distribution with parameter $1 - p_{it}$. With a negative binomial
-prior, we have:
-\begin{align}
-n_i p_{it}^{n_i- 1} p(n_i)
-&\propto n_i p_{it}^{n_i- 1} \frac{\Gamma(r + n_i)}{n_i!} \left( \frac{\mu}{r+\mu} \right)^{n_i} \\
-&\propto \frac{\Gamma((r + 1) + (n_i - 1))}{(n_i-1)!} \left( \frac{\mu p_{it}}{r+\mu} \right)^{n_i-1}
-\end{align}
-which is a negative binomial pmf with size parameter $r+1$ and
-probability parameter $\frac{r + \mu (1 - p_{it})}{r+\mu}$. The mean
-of this is
-\begin{align}
-\frac{(r+1)\mu p_{it}}{r+\mu(1-p_{it})}
-\end{align}
-
-Next, I derive $p_{ia}$ and $p_{it}$.
-
-By definition, we have
-\begin{align}
-p_{ia} &= \prob((b, e) \in \alpha_i) \\
-\alpha_i &= \{ (b, e) : l_i^{(b)} \leq b \leq r_i^{(b)} \wedge l_i^{(e)} \leq e \leq r_i^{(e)}\}.
-\intertext{Hence:}
-p_{ia}
-&= \prob \left( l_i^{(b)} \leq B_i \leq r_i^{(b)}, l_i^{(e)} \leq E_i \leq r_i^{(e)} \right) \\
-&= \prob \left( l_i^{(e)} \leq E_i \leq r_i^{(e)} \mid l_i^{(b)} \leq B_i \leq r_i^{(b)} \right) \prob \left( l_i^{(b)} \leq B_i \leq r_i^{(b)} \right) \\
-&=\sum_{b = l_i^{(l)}}^{r_i^{(b)}} \prob \left( l_i^{(e)} \leq E_i \leq r_i^{(e)} \mid B_i = b \right) \prob \left(B_i = b \right) \\
-&=\sum_{b = l_i^{(b)}}^{r_i^{(b)}} \prob \left( l_i^{(e)} - b + 1 \leq D_i \leq r_i^{(e)} - b + 1 \right) \prob \left(B_i = b \right) \\
-&=\sum_{b = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b + 1) - S_\theta(r_i^{(e)} - b + 2) \right) \prob \left(B_i = b \right) \\
-&\propto \sum_{b = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b + 1) - S_\theta(r_i^{(e)} - b + 2) \right)
-\end{align}
-under the assumption of uniform probability of infection time.
-
-Next, we derive $p_{it} = \prob((b, e) \in \Omega^C_i)$, the probability
-that episode $i$ was truncated. A truncated episode means that no test
-was performed during the episode or there was no negative test prior to
-the episode. Denote by $t_i$ the set of testing times for the
-individual in which episode $i$ was observed, and define the time
-until the next test on the individual after time $t'$ as:
-\begin{align}
-t^N_{it} &= \min \{ t' \in t_i : t' \geq t \} - t.
-\intertext{Then:}
-\Omega^C_i
-&= \{ (b, e) : \nexists t \in t_i. b \leq t \leq e \vee b \leq \min(t_i) \} \\
-&= \{ (b, e) : e - b < t^N_{ib} \vee b \leq \min(t_i) \}.
-\intertext{Hence:}
-1 - p_{it}
-&= 1 - \prob(E_i - B_i < t_{iB_i}^N \vee B_i \leq \min(t_i)) \\
-&= 1 - \prob(E_i - B_i < t_{iB_i}^N \wedge B_i > \min(t_i)) - \prob(B_i \leq \min(t_i)) \\
-&= 1 - \sum_{b=\min(t_i) + 1}^T \prob(E_i - b + 1 < t_{ib}^N \mid B_i = b) \prob(B_i = b) \\
-  &\; - \sum_{t=1}^{\min(t_i)} \prob(B_i = b)\\
-&= 1 - \frac{1}{T} \sum_{b=\min(t_i)+1}^T (1 - S_\theta(t_{ib}^N + 1)) - \frac{\min(t_i)}{T} \\
-&= 1 - \frac{T-\min(t_i)}{T} + \frac{1}{T} \sum_{b=\min(t_i)+1}^T S_\theta(t_{ib}^N + 1)) - \frac{\min(t_i)}{T} \\
-&= \frac{1}{T} \sum_{b=\min(t_i)+1}^T S_\theta(t_{ib}^N + 1)
-\end{align}
-
-\subsection{Full posterior}\label{perf-test:sec:full-posterior}
-
-Under the prior $p(n_i) \propto 1/n_i$:
-\begin{align}
-&p(\theta \mid n_{ia} = 1, n_{iu} = 0, \dots, n_{n_a,a} = 1, n_{n_a,u} = 0) \\
-&\propto p(\theta) \prod_{i=1}^{n_a} p_{ia} \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) \\
-&= p(\theta) \prod_{i=1}^{n_a} \frac{p_{ia}}{1-p_{it}} \\
-&\propto p(\theta) \prod_{i=1}^{n_a} \frac{\sum_{b = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(r_i^{(e)} - b - 1) - S_\theta(l_i^{(e)} - b - 1) \right)}{\sum_{b=\min(t_i)}^T S_\theta(t_{ib}^N + 1)} \\
-\end{align}
-
-Under the prior $N_i \dist \text{NegBinom}(\mu, r)$:
-\begin{align}
-&p(\theta \mid n_{ia} = 1, n_{iu} = 0, \dots, n_{n_a,a} = 1, n_{n_a,u} = 0) \\
-&\propto p(\theta) \prod_{i=1}^{n_a} p_{ia} \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) \\
-&= p(\theta) \prod_{i=1}^{n_a} \frac{p_{ia}}{(r+\mu (1-p_{it}))^{(r+1)}} \\
-&= p(\theta) \prod_{i=1}^{n_a} \frac{\sum_{b=l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(r_i^{(e)} - b - 1) - S_\theta(l_i^{(e)} - b - 1) \right)}{\left( r+\mu/T \left( \sum_{b=\min(t_i)}^T S_\theta(t_{ib}^N + 1) \right) \right)^{(r+1)}} \\
-\end{align}
-
-\section{Parameterisation and priors for the survival function} \label{perf-test:sec:parameters-priors}
-
-The form of $S_\theta(t)$ is yet to be specified.
-As frequently observed in the literature\todo{refs for parameterisation}, it is more convenient to operate with the hazard than the survival or probability mass function directly.
-This is because the mass function must sum to one and the survival must be monotonically decreasing, while there are no constraints across multiple hazards.
-The only constraint on the hazard is to be in the interval $[0, 1]$, since they are probabilities.
-
-In this section, I consider the various parameterisation of the survival function that are available.
-This has implications for the choice of prior it is not possible to be vague on all sensible parameterisations.
-Informative priors might also be attractive, and allow the incorporation of the estimates from \cref{ATACCC}.
-\todo{Ref ATACCC chapter}
-
-The hazard is convenient to use because it is unconstrained except in the interval [0, 1]\todo{ref where been used previously}.
-Survival needs to be monotonic.
-
-\subsection{Independent priors}
-
-A common assumption throughout Bayesian analyses is independence of the parameters in the prior.
-This is convenient because we can then use univariate distributions.
-\todo{This feels a bit clumsy/patronising to the reader. Maybe change the phrasing?}
-A Beta distribution for each hazard parameter, $\lambda_t$, is a natural choice due to it being a probability.
-
-An uninformative prior would be $\lambda_t \dist \Beta(\alpha, \alpha)$, with common choices of $\alpha$ being $0.5$, Jeffreys' Prior, or $1$, a flat prior.
-Flat or uninformative priors on the hazard lead to very informative priors on the survival, favouring short survival times.
-The sharp falloff can be seen by noting that these priors have expectation 0.5, hence $\E(S(t)) = \E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right) = 0.5^{t-1}$ which decreases very quickly.
-Our prior belief is not well-represented by this prior, and in fact a priori viewed as extremely unlikely (see \cref{perf-test:fig:flat-priors}),
-Therefore, I favour a prior which is weakly informative on the survival, or an informative prior used to represent our belief from the \cref{ATACCC} analysis.
-\todo{Ref ATACCC chapter}
-\begin{figure}
-  \centering \includegraphics{cis-perfect-testing/flat-prior}
-  \todo[inline]{Create flat priors figure}
-  \caption{The prior predictive survival time (shown) is very short when using either a Beta(1, 1) flat prior or Beta(0.5, 0.5) uninformative prior on the hazard at each time.
-\label{perf-test:fig:flat-prior}}
-\end{figure}
-
-Instead, I propose a weakly informative prior Beta(0.1, 1.9), linking to previous literature for a point estimate and showing vague figures.
-
-A Beta(0.1, 1.9) prior has mean 0.05 while not having much information (the informativeness of a Beta distribution is related to $\alpha + \beta$ which here equals 2, the same as the flat prior case); and the central 95\% mass is 0.00-0.47.
-The central estimate reflects previous estimates that the median duration is in the range 15-20 days~\autocite{cevikShedding}, which would translate to hazard of around 0.05.
-The prior predictive distribution on the survival time implied by this prior is very vague (see \cref{perf-test:fig:vague-prior}).
-\begin{figure}
-  \centering %\includegraphics{cis-perfect-testing/vague-priors}
-  \todo[inline]{Create vague prior figure}
-  \caption{The prior predictive survival time (shown) is very vague using a Beta(0.1, 1.9) prior. \label{perf-test:fig:vague-prior}}
-\end{figure}
-
-\subsection{Informative priors}
-
-The analysis in \cref{ATACCC} can be used to provide prior information for the CIS analysis.
-\todo{Ref ATACCC chapter here and next few}
-In particular, the ATACCC study frequently samples individuals early in the infection.
-Therefore, the survival distribution from that analysis should be reliable through this period.
-CIS can then inform the remainder of the distribution where the previous analysis is based on extrapolation.
-
-The model structure assumed in \cref{ATACCC} leads to positive correlation in the posterior estimates of the hazard.
-\todo{Ref ATACCC chapter}
-It is important to propagate this correlation as part of the prior uncertainty into the present analysis.
-We do this in two parts: first, approximate the previous posterior estimate of the logit-hazard as a multivariate normal (which can be viewed as an approximation of Markov melding~\autoref{goudieJoining}).
-Then, use a Beta process to add additional uncertainty to this.
-
-There are two reasons to increase the uncertainty relative to the \cref{ATACCC} posterior.
-First, we do not have full belief in these estimates due to the extrapolation problem previously discussed.
-Second, the results may not fully generalise between the two studies.
-The samples are tested using different laboratories, with different PCR machines, and hence may not have the same characteristics~\autocite{punchooLaboratory}.
-The study participants, while both targeting a general population cohort, may also differ because ATACCC recruited through NHS Test and Trace while CIS recruited by sampling from a database of addresses (see \cref{intro:sec:data}).
-
-The first step, approximating the \cref{ATACCC} posterior as a multivariate normal, is done through the method of moments.
-The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC}).
-
-The discrete Beta process prior generalises the form of prior used in the previous section, allowing the central estimate of the hazard to vary over time~\autocites{ibrahimBayesian}{sunStatisticala}.
-\begin{align}
-  \lambda_t &\sim \text{Beta}(\alpha_t, \beta_t) &t = 1, 2, \dots \\
-  \alpha_t &= c_t h_t + \alpha_0 \\
-  \beta_t &= c_t (1 - h_t) + \beta_0
-\end{align}
-where $k_t$, $\alpha_0$, and $\beta_0$ are hyperparameters; and $h_t$ is a point estimate of the hazard at time $t$ from ATACCC.
-An intuition for what this distribution represents can be gained by considering the conjugate model for $\lambda_t$ with a beta prior and a binomial likelihood.
-If $\lambda_t$ is given the prior distribution $\text{Beta}(\alpha_0, \beta_0)$, and we then have $k_t$ observations with $k_t h_t$ successes, then the posterior distribution for $\lambda_t$ is $\text{Beta}(\alpha_t, \beta_t)$ (as defined above).
-
-\subsection{Smoothing priors}
-
-A final possible prior are ones to favour smoothness in the hazard.
-This is the Bayesian equivalent of a penalised likelihood, used frequently in the non-parametric frequentist setting.
-Biological and epidemiological considerations mean that it would be surprising if the hazard varied significantly day-to-day.
-A smoothing prior encodes this information within the analysis.
-
-I consider two forms of smoothing priors: first-order and second-order random walks on the logit-hazard, $\logit(\lambda_t) = \log(\lambda_t) - \log(1 - \lambda_t)$.
-The logit transformation maps from the $[0, 1]$ interval to the full real line.
-\todo{The logit intro will be in the previous subsection}
-Random walks are a very common form of smoothing prior for discrete time.
-
-A first-order random walk prior encodes that the hazard should be constant with some random changes at each time step.
-\begin{align}
-  \logit\lambda_{t+1} &= \logit\lambda_t + \sigma \epsilon_t &\text{for $t \geq 1$} \\
-  \epsilon_t &\dist N(0, 1) \\
-  \logit \lambda_1 &\dist N(\mu_1, \sigma_1^2)
-\end{align}
-The hyperparameter $\sigma$ controls the standard deviation of the random walk, and the prior on $\lambda_1$ gives the starting point of the random walk.
-Smaller values favour a smoothing process.
-
-A second-order random walk prior encodes that the hazard should be linearly changing with some random changes at each time step.
-\begin{align}
-  \logit\lambda_{t+1}
-  &= \logit\lambda_t + (\logit\lambda_t - \logit\lambda_{t-1}) + \sigma \epsilon_t &\text{for $t \geq 2$} \\
-  &= 2\logit\lambda_t - \logit\lambda_{t-1} + \sigma \epsilon_t \\
-  \epsilon_t &\dist N(0, 1) &\text{for $t \geq 2$}  \\
-  \logit\lambda_2 &= \logit\lambda_1 + \epsilon_1 \\\
-  \epsilon_1 &\dist N(\mu_2, \sigma_2^2) \\
-  \logit \lambda_1 &\dist N(\mu_1, \sigma_1^2)
-\end{align}
-The hyperparameter $\sigma$ controls the standard deviation of the random walk, and there is an additional prior on $\epsilon_1$ for the initial gradient of the change in the hazard.
-
-\section{Simulation study} \label{perf-test:sec:simulation-study}
-
-\subsection{Setup}
-
-I simulate each individual enrolled in CIS who was tested at least once between ?? and ??.
-All their tests between ?? and ?? are considered, and each individual is assigned an infection time $B_i$ uniformly at random from ??.
-\todo{check dates and periods}
-
-I take as the ground truth distribution for duration of positivity a combination of the estimates from \cref{ATACCC} with an inflated tail to represent what is seen within CIS.
-\todo{Ref ATACCC chapter}
-The tail is modified based on Sarah Walker's (unpublished) work based on CIS data.
-These curves and the combined curve are compared in \cref{perf-test:fig:duration-dist}.
-Individual $i$'s duration of positivity, $D_i$, is then an independent draw from this distribution.
-
-\begin{figure}
-  \todo[inline]{Find figure comparing different duration distributions}
-  \caption{Comparison of the duration distributions, see main text for details of how each is derived. \label{perf-test:fig:duration-dist}}
-\end{figure}
-
-Sarah uses survival analysis to estimate the duration from CIS data.
-The initiating event is assumed known as the time the episode was detected.
-The final event is assumed interval-censored between the time of the final positive test and the subsequent negative test, or right-censored if a negative test has not yet been observed.
-A flexible, spline-based form is assumed for the baseline survival function~\autocite{roystonSTPM,roystonFlexible} with covariates introduced via proportional odds.
-By not accounting for either the truncation or the interval censoring of the initiating event, this analysis has competing biases which makes them hard to interpret~\autocite{cisMethodsONS}.
-
-% \todo[inline]{The following seems important to justify why we can't just use these estimates but is also basically an aside}
-% We know of two biases introduced from this analysis.
-% \enquote{There is a bias in estimating the clearance distribution because the analysis used to estimate how long a person stays positive only starts from their first positive test.
-% Since (most) people will have become positive on an earlier day, this will bias the clearance curves downwards (making the estimates too short).
-% However, there is another bias due to the survey missing positive episodes entirely if they are short.
-% This means that our dataset has fewer short positive episodes than in the population as a whole, and that the sample used to run the survival analysis is biased towards people with longer positive episodes.
-% This will bias the clearance curves upwards (making the estimates too long).}~\autocite{cisMethodsONS}.
-
-To form the duration distribution used in the simulation, we combine these two estimates.
-The first 30 days of the distribution is proportional to the ATACCC estimate, with the rest proportional to this CIS-based estimates.
-% The CIS-based estimates are shifted 3 days in order to make this smooth; this counteracts the above bias of missing the start of the infection (for times above 30 days, the other bias, missing short infections, is negligible).
-ERR...
-I'M NOT DOING THIS ANYMORE OH NO!!
-Denote by $f_A(t)$ the distribution function estimated from ATACCC and $f_C(t)$ that from these CIS-based estimates.
-Define:
-$$
-f_S'(t) = \begin{cases}
-	f_A(t) &t \leq 30 \\
-	f_C(t) &t > 30
-\end{cases}
-$$
-Then the distribution used in the simulation is the normalised version of this: $f_S(t) = f'_S(t)/\sum_i f_S'(i)$.
-
-The first day individual $i$ can be positive is their day of infection, $B_i$, the last day is $E_i = B_i + D_i - 1$ where $D_i$ is the duration of positivity for individual $i$, drawn from $f_S$.
-Any test between $B_i$ and $E_i$ (inclusive) is positive; outside this period, all tests are negative.
-
-I then filter the episodes to replicate the process which will be used when working with the CIS data.
-First, discard any episodes where there are no positive tests as they are never detected.
-Next, discard any episodes where there are no negative tests prior to the first positive; since there is no lower bound on their infection time, they contain little information, this selection effect was included when constructing $p_{it}$.
-Finally, of those remaining episodes, we select a random subset to match the number of episodes in the simulation to the number of episodes in the modelling cohort.
-The last step is required because everyone is infected in the simulation which does not happen in reality.
-
-For each preserved infection episode, calculate $(l_i^{(b)}, r_i^{b}, l_i^{(e)}, r_i^{(e)})$ by taking the day after the last negative prior to any positives, the first positive, the last positive, and the day before the negative following the last positive respectively.
-
-\subsection{Descriptive analyses of simulated data}
-
-\subsection{Inference and hyperparameters}
-
-\subsection{Results}
-
-\todo[inline]{Need to decide what exactly to include... There's lots of possible results but need to form some coherent story here!}
-
-\section{Discussion} \label{perf-test:sec:discussion}
-
-\todo[inline]{Add discussion after deciding on results}
-
-\section{Conclusion} \label{perf-test:sec:conclusion}
-
-\listoftodos
-
-\appendix
-
-\chapter{Including all CIS individuals in survival analysis}\label{perf-test:sec:total-model} \label{appendix:total-model}
+\todo[inline]{Add an intro to this and maybe a plot of results in simulation}
 

### 2023-06-09 cis-perfect-testing.tex
Fix cross-file referencing

@@
 The estimation of incidence from prevalence is sensitive to the tail of the duration distribution.
-In \cref{ATACCC} I estimated the duration distribution from the ATACCC data, however, this study suffered from a small sample size and only 20 days of follow-up.
-\todo{Ref ATACCC chapter and in following paragraph}
+In \cref{E-ATACCC} I estimated the duration distribution from the ATACCC data, however, this study suffered from a small sample size and only 20 days of follow-up.
 Therefore, the estimation of the tail of the duration distribution is driven by model assumptions.
@@
 I take a survival analysis approach to model the coarse sampling of the CIS (see \cref{perf-test:sec:problem}).
-Flat priors can be unintentionally very informative, and hence weakly informative priors should be used; in addition, the prior can be used to combine the information in the CIS with the results of \cref{ATACCC} (see \cref{perf-test:sec:parameters-priors}).
+Flat priors can be unintentionally very informative, and hence weakly informative priors should be used; in addition, the prior can be used to combine the information in the CIS with the results of \cref{E-ATACCC} (see \cref{perf-test:sec:parameters-priors}).
 Through a simulation study (see \cref{perf-test:sec:simulation-study}), I show that these methods perform well, and in this context are not sensitive to model or prior specification.
-However, application in reality must account for misclassification bias, the subject of \cref{imperf-test}.
-\todo{Ref chapter incl misclassification}
+However, application in reality must account for misclassification bias, the subject of \cref{E-imperf-test}.
 
@@
 
-I define an individual as \emph{detectable} if they would test positive if tested, in the absence of misclassification error (which is dealt with in \cref{imperf-test}).
-\todo{Ref chapter incl misclassification}
+I define an individual as \emph{detectable} if they would test positive if tested, in the absence of misclassification error (which is dealt with in \cref{E-imperf-test}).
 Episode $i$'s \emph{duration of positivity}, $D_i$ is the number of days that they are detectable for; I use time discretised to days, therefore, the duration is at least 1 day.
@@
 
-The reason for this approach, over modelling biomarkers (as in \cref{ATACCC}) is due to two factors.
-\todo{Ref ATACCC chapter}
+The reason for this approach, over modelling biomarkers (as in \cref{E-ATACCC}) is due to two factors.
 First, the large sample size in CIS means that approaches where the number of parameters scales with the number of observed infections (as it does with random effect based approaches) would be computationally challenging; this challenge is compounded by the data being stored in a trusted research environment limiting the computational power and tools available.
@@
 The methods in this chapter deal with modelling two elements of the data-generating process, double interval censoring and truncation.
-False negatives are not dealt with and left to \cref{imperf-test}.
-\todo{Ref chapter incl misclassification}
+False negatives are not dealt with and left to \cref{E-imperf-test}.
 
@@
 
-Truncation is a large problem within CIS, especially when individuals move to the more infrequent testing schedule (described in \cref{intro:sec:cis}).
-\todo{ref to description of CIS}
-Consider that, based on the estimates in \cref{ATACCC}, the median length of an episode is 14 days \todo{check this number} we would not expect many episodes to last the 28 days that can occur between tests.
-\todo{Ref ATACCC chapter}
+Truncation is a large problem within CIS, especially when individuals move to the more infrequent testing schedule (described in \cref{E-intro:sec:cis}).
+Consider that, based on the estimates in \cref{E-ATACCC}, the median length of an episode is 14 days \todo{check this number} we would not expect many episodes to last the 28 days that can occur between tests.
 \begin{figure}
@@
 While this chapter sets up the framework for analysing the CIS data, applying the method to CIS data requires modelling false negatives.
-That is the subject of \cref{imperf-test} and hence a full description of the data and processing of it is deferred to that chapter.
-\todo{Ref chapter incl misclassification}
+That is the subject of \cref{E-imperf-test} and hence a full description of the data and processing of it is deferred to that chapter.
 
@@
 Furthermore, I explore a Bayesian framework where the amount of explicit data augmentation is limited allowing the method to scale to the large number of episodes present in CIS without requiring vast computing resources (such as a high-performance cluster), as only limited resources are available within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
-Therefore, unlike most Bayesian methods (e.g.: \textcite{heBayesiana}, \textcite{heBayesian}, and \textcite{caoModeling}) we avoid augmenting the data with unobserved times of infection or any information on the truncated infections; this would be computationally expensive with the large number of episodes in CIS (>4000 in the time period considered in the main analysis of \cref{imperf-test}).
+Therefore, unlike most Bayesian methods (e.g.: \textcite{heBayesiana}, \textcite{heBayesian}, and \textcite{caoModeling}) we avoid augmenting the data with unobserved times of infection or any information on the truncated infections; this would be computationally expensive with the large number of episodes in CIS (>4000 in the time period considered in the main analysis of \cref{E-imperf-test}).
 
@@
 
-I have also considered a model including individuals without any observed episodes (see \cref{appendix:total-model} for details).
+I have also considered a model including individuals without any observed episodes (see \cref{E-appendix:total-model} for details).
 These individuals may have truncated episodes or may not have been infected at all.
@@
 This has implications for the choice of prior it is not possible to be vague on all sensible parameterisations.
-Informative priors might also be attractive, and allow the incorporation of the estimates from \cref{ATACCC}.
-\todo{Ref ATACCC chapter}
+Informative priors might also be attractive, and allow the incorporation of the estimates from \cref{E-ATACCC}.
 
@@
 The sharp falloff can be seen by noting that these priors have expectation 0.5, hence $\E(S(t)) = \E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right) = 0.5^{t-1}$ which decreases very quickly.
-This prior expresses extreme scepticism in estimates from previous studies, such as \cref{ATACCC} and the meta-analysis of \textcite{cevikShedding}.
-Therefore, I favour a prior which is weakly informative on the survival, or an informative prior used to represent our belief from the \cref{ATACCC} analysis.
-\todo{Ref ATACCC chapter}
+This prior expresses extreme scepticism in estimates from previous studies, such as \cref{E-ATACCC} and the meta-analysis of \textcite{cevikShedding}.
+Therefore, I favour a prior which is weakly informative on the survival, or an informative prior used to represent our belief from the \cref{E-ATACCC} analysis.
 \begin{figure}
@@
 
-The analysis in \cref{ATACCC} can be used to provide prior information for the CIS analysis.
-\todo{Ref ATACCC chapter here and next few}
+The analysis in \cref{E-ATACCC} can be used to provide prior information for the CIS analysis.
 In particular, the ATACCC study frequently samples individuals early in the infection.
@@
 
-The model structure assumed in \cref{ATACCC} leads to positive correlation in the posterior estimates of the hazard.
+The model structure assumed in \cref{E-ATACCC} leads to positive correlation in the posterior estimates of the hazard.
 It is important to propagate this correlation as part of the prior uncertainty into the present analysis.
@@
 
-There are two reasons to increase the uncertainty relative to the \cref{ATACCC} posterior.
+There are two reasons to increase the uncertainty relative to the \cref{E-ATACCC} posterior.
 First, we do not have full belief in these estimates due to the extrapolation problem previously discussed.
@@
 The samples are tested using different laboratories, with different PCR machines, and hence may not have the same characteristics~\autocite{punchooLaboratory}.
-The study participants, while both targeting a general population cohort, may also differ because ATACCC recruited through NHS Test and Trace while CIS recruited by sampling from a database of addresses (see \cref{intro:sec:data}).
+The study participants, while both targeting a general population cohort, may also differ because ATACCC recruited through NHS Test and Trace while CIS recruited by sampling from a database of addresses (see \cref{E-intro:sec:data}).
 
-The first step, approximating the \cref{ATACCC} posterior hazard as a multivariate normal on the logit scale, is done through the method of moments.
+The first step, approximating the \cref{E-ATACCC} posterior hazard as a multivariate normal on the logit scale, is done through the method of moments.
 The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC}).
@@
 
-I take as the ground truth distribution for duration of positivity a combination of the estimates from \cref{ATACCC} with an inflated tail to represent what is seen within CIS.
-\todo{Ref ATACCC chapter}
+I take as the ground truth distribution for duration of positivity a combination of the estimates from \cref{E-ATACCC} with an inflated tail to represent what is seen within CIS.
 The tail is modified based on Sarah Walker's (unpublished) work based on CIS data.
@@
 
+\ifSubfilesClassLoaded{
   \appendix
-
-\chapter{Including all CIS individuals in survival analysis}\label{perf-test:sec:total-model} \label{appendix:total-model}
-
-\todo[inline]{Add an intro to this and maybe a plot of results in simulation}
-
-We now consider all $N$ individuals in the CIS cohort who have at
-least one test during the period of interest, this includes those
-without an observed infection, and assume that each is equally likely to
-be infected. As before, $i = 1, \dots, n_a$ indexes episodes but now
-$j = 1, \dots, N$ indexes individuals. All episodes have an associated
-individual, each individual may have any number (including 0) episodes.
-Admissible regions are only defined for episodes, and have not changed
-from the previous section. The truncated region is defined for each
-individual, in addition we consider the combined truncated region which
-is episodes that would be truncated without conditioning on which
-individual the episode occurs in:
-$\Omega^C = \{ (b, e, j) : (b, e) \in \Omega_j^C \}$.
-
-Denote by $n_\text{tot}$ the total number of episodes across all $N$
-individuals, regardless of whether they are admissible or truncated (we
-know there are no inadmissible episodes by definition).
-
-For any infection, the probability of it being admissible for episode
-$i$ is $\frac{1}{N} p_{ia}$. That is, the probability that the
-infection occurs in the individual corresponding to the individual in
-which episode $i$ ($1/N$ by the assumption that episodes are equally
-likely to occur in any of the $N$ individuals) multiplied by the
-probability that the episode is admissible for episode $i$ conditional
-on it occurring in the relevant individual ($p_{ia}$).
-
-The total number of truncated infections that occur is
-$n_t = n_\text{tot} - n_a$. Conditional on an infection occurring in
-individual $j$, its probability of being truncated is $p_{jt}$ (as
-defined previously). Therefore, the overall probability of an episode
-being truncated is $p_t = \frac{1}{N} \sum_{j=1}^N p_{jt}$.
-
-Consider our data as there being one admissible episode for each
-observed episode, and all other episodes being truncated. As in the
-previous section, conditional on $n_\text{tot}$, we have a multinomial
-likelihood.
-\begin{align}
-p(\text{data} \mid \theta)
-&= \sum_{n_\text{tot}=n_a}^\infty p(\text{data} \mid n_\text{tot}, \theta) p(n_\text{tot}) \\
-&= \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} \left( \prod_{i=1}^{n_a} \frac{1}{N} p_{ia} \right) p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \\
-&= \left( \prod_{i=1}^{n_a} \frac{1}{N} p_{ia} \right) \left( \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \right) \\
-\end{align}
-
-In this case, it would be computationally feasible to augment the data
-with $n_\text{tot}$ directly (as it is only a single parameter).
-However, implementation is easier if we can use a closed-form for the
-sum (as in the previous section) as it allows the use of standard
-implementations of NUTS such as Stan, which cannot handle discrete
-parameters.
-
-Assuming $N_\text{tot} \dist \NegBin(\mu, r)$:
-\begin{align}
-&\sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \\
-&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot} \mid \lambda) p(\lambda) d\lambda \\
-&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} \frac{\lambda^n_\text{tot} e^{-\lambda}}{n_\text{tot}!} p(\lambda) d\lambda \\
-&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{1}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} \lambda^{n_\text{tot}-n_a} \lambda^{n_a} e^{-\lambda} p(\lambda) d\lambda \\
-&= \int \lambda^{n_a} e^{-\lambda} p(\lambda) \sum_{n_t=0}^\infty \frac{1}{n_t!} p_t^{n_t} \lambda^{n_t} d\lambda &n_t = n-n_a\\
-&= \int \lambda^{n_a} e^{-\lambda} p(\lambda) e^{\lambda p_t} d\lambda \\
-&= \int \lambda^{n_a} e^{-\lambda(1 - p_t)} p(\lambda) d\lambda \\
-&= \int \lambda^{n_a} e^{-\lambda(1 - p_t)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda \\
-&= \int \frac{b^a}{\Gamma(a)} \lambda^{a+n_a-1} e^{-(b+1-p_t)\lambda} \lambda d\lambda \\
-&= \frac{b^a}{\Gamma(a)} \frac{\Gamma(a+n_a)}{(b+1-p_t)^{a+n_a}} \\
-&\propto (b+1-p_t)^{-(a+n_a)} \\
-&= (r/\mu + 1 - p_t)^{-(r+n_a)} \\
-&\propto(r + \mu (1- p_t))^{-(r+n_a)}
-\end{align}
-
-Which gives the full posterior density as:
-\begin{align}
-p(\theta \mid \text{data})
-&\propto \frac{p(\theta) \prod_{i=1}^{n_a} p_{ia}}{(r + \mu(1 - p_t))^{r+n_a}}
-\end{align}
-
-As before, being able to reconstruct the posterior of $n_\text{tot}$
-using its full conditional is useful.
-\begin{align}
-p(n_\text{tot} \mid \text{data}, \theta)
-&\propto p(\text{data} \mid \theta, n_\text{tot}) p(n_\text{tot}) \\
-&= \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} \left( \prod_{i=1}^{n_a} \frac{1}{N} p_{ia} \right) p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \\
-&\propto \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}} \frac{\Gamma(r + n_\text{tot})}{n_\text{tot}!} \left( \frac{\mu}{r + \mu} \right)^{n_\text{tot}}  \\
-&\propto \frac{\Gamma(r + n_\text{tot})}{(n_\text{tot}-n_a)!} \left( \frac{\mu p_t}{r + \mu} \right)^{n_\text{tot}}  \\
-&\propto \frac{\Gamma((r + n_a) + (n_\text{tot}- n_a))}{(n_\text{tot}-n_a)!} \left( \frac{\mu p_t}{r + \mu} \right)^{n_\text{tot}-n_a}  \\
-\end{align}
-which is a negative binomial pmf with size parameter $r + n_a$ and
-probability parameter $\frac{r+\mu(1-p_t)}{r+\mu}$.
-
+  \subfile{appendix-cis-perfect-testing}
   \listoftodos
-
+}{}
 \end{document}

### 2023-06-09 cis-perfect-testing.tex
Bit of editing and organising

@@
   \centering \includegraphics{cis-perfect-testing/double-interval-censor}
-  \caption{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval.
+  \caption[Double-interval censoring in CIS data]{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval.
 Demonstrated here by a participant who is recorded negative at time 7 to positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode.
@@
   \centering \includegraphics{cis-perfect-testing/truncation}
-  \caption{Episodes data in CIS is truncated, meaning that an unknown number of infections are never detected.
+  \caption[Truncation in CIS data]{Episodes data in CIS is truncated, meaning that an unknown number of infections are never detected.
 Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was.
@@
 \todo[inline]{Fix the impossible region to fill out the RHS; label the inadmissible regoin.}
-\caption{Admissible, $\alpha_i$, inadmissible, $\beta_i$ and
+\caption[Admissible, inadmissible, and untruncated infections]{Admissible, $\alpha_i$, inadmissible, $\beta_i$ and
 untruncated, $\Omega_i^C$ regions for an episode $i$ in an
@@
   \centering \includegraphics{cis-perfect-testing/flat-prior}
-  \caption{The density of Beta(1, 1) flat prior and Beta(0.5, 0.5) uninformative priors (left) with the prior predictive survival time (right).
+  \caption[Uninformative priors for the hazard]{The density of Beta(1, 1) flat prior and Beta(0.5, 0.5) uninformative priors (left) with the prior predictive survival time (right).
 The implied prior on the survival time is very short when using either prior on the hazard at each time.
@@
   \centering \includegraphics{cis-perfect-testing/vague-prior}
-  \caption{The density of Beta(0.1, 1.9) vague prior (left) and the prior predictive survival time (right).
+  \caption[Weakly informative priors for the hazard]{The density of Beta(0.1, 1.9) vague prior (left) and the prior predictive survival time (right).
 The prior predictive survival time is very vague using a Beta(0.1, 1.9) prior.
@@
 The samples are tested using different laboratories, with different PCR machines, and hence may not have the same characteristics~\autocite{punchooLaboratory}.
-The study participants, while both targeting a general population cohort, may also differ because ATACCC recruited through NHS Test and Trace while CIS recruited by sampling from a database of addresses (see \cref{E-intro:sec:data}).
+The study participants, while both targeting a general population cohort, may also differ because ATACCC recruited through NHS Test and Trace while CIS recruited by sampling from a database of addresses (see \cref{E-intro:sec:studies}).
 
@@
   \todo[inline]{Find figure comparing different duration distributions}
-  \caption{Comparison of the duration distributions, see main text for details of how each is derived. \label{perf-test:fig:duration-dist}}
+  \caption[Comparison of duration distributions]{Comparison of the duration distributions, see main text for details of how each is derived. \label{perf-test:fig:duration-dist}}
 \end{figure}
@@
   \centering \includegraphics{cis-perfect-testing/survival-results}
-  \caption{Comparison of the posterior estimate of the survival distribution (pointwise median and 95\% CrI) when using different priors on the hazard. \label{perf-test:fig:survival-results}}
+  \caption[Comparison of survival function estimates under different priors]{Comparison of the posterior estimate of the survival distribution (pointwise median and 95\% CrI) when using different priors on the hazard.
+\label{perf-test:fig:survival-results}}
 \end{figure}
@@
   \centering \includegraphics{cis-perfect-testing/hazard-results}
-  \caption{Comparison of the posterior estimate of the hazard (pointwise median and 95\% CrI) when using different priors on the hazard. \label{perf-test:fig:hazard-results}}
+  \caption[Comparison of hazard estimates under different priors]{Comparison of the posterior estimate of the hazard (pointwise median and 95\% CrI) when using different priors on the hazard.
+\label{perf-test:fig:hazard-results}}
 \end{figure}
@@
   \todo[inline]{Probably scrap this (correlation) plot}
-  \caption{Correlation between posterior estimates of the hazard when using the vague prior. \label{perf-test:fig:hazard-pairs-results}}
+  \caption[Correlation between hazard estimates]{Correlation between posterior estimates of the hazard when using the vague prior. \label{perf-test:fig:hazard-pairs-results}}
 \end{figure}

### 2023-06-09 cis-imperfect-testing.tex
Bit of editing and organising

@@
   \todo[inline]{Figure showing estimates without false negatives}
-  \caption{\label{imperf-test:fig:problem-cis-estimates}}
+  \caption[Estimating survival using CIS data assuming perfect testing]{Estimating the survival function using CIS data and the model of \cref{perf-test} gives implausibly short estimates.
+\label{imperf-test:fig:problem-cis-estimates}}
 \end{figure}
@@
   \todo[inline]{Include the analysis of CIS data with bounding test sensitivity}
-  \caption{Bounding the test sensitivity using CIS data as a function of time since the infection was detected (see main text for explanation of how the bounds are formed).\label{imperf-test:fig:bounding-cis-sensitivity}}
+  \caption[Bounding test sensitivity using CIS data]{Bounding the test sensitivity using CIS data as a function of time since the infection was detected (see main text for explanation of how the bounds are formed).\label{imperf-test:fig:bounding-cis-sensitivity}}
 \end{figure}

### 2023-06-09 thesis.tex
Bit of editing and organising

@@
 
+\chapter{Introduction} \label{intro}
+
+\section{The COVID-19 pandemic and ongoing burden}
+
+\section{SARS-CoV-2 and COVID-19}
+
+\section{PCR testing} \label{intro:sec:PCR}
+
+\section{Incidence, prevalence, and duration}
+
+\section{Studies used in this thesis} \label{intro:sec:studies}
+
+\subsection{Assessment of Transmission and Contagiousness of COVID-19 in Contacts}
+
+\subsection{Coronavirus (COVID-19) Infection Survey} \label{intro:sec:cis}
+
+The CIS (Coronavirus (COVID-19) Infection Survey) continually enrols individuals and tests them on an ongoing basis, following a
+specified testing protocol. The protocol specifies that the initial five
+tests are spaced every 7 days, then the spacing is every 28 days.
+However, real-world considerations often lead to variations in test
+times and missed tests. Our analysis focuses solely on the binary result
+(positive or negative) of these tests, assuming no misclassification
+bias.
+
+The CIS (Coronavirus Infection Survey) is a longitudinal study on a representative sample of households.
+A full description of the study can be found in the supplementary materials of \textcite{pouwelsCommunity} or the study protocol~\cref{cisProtocol}.
+In brief, households are invited to the study from databases held by the ONS.
+The number invited was to satisfy various target number of individuals swabbing per fortnight; relevant to the discussion in this document is that the target increased from 2,500 by the end of May 2020 to 150,000 in October 2020.
+Within a household, all individuals aged 2 and over were invited to participate.
+Once invited, an enrolment swab would be taken followed by 4 further weekly swabs (giving a total of 5 swabs on days 0, 7, 14, 21, 28 relative to enrolment) after which monthly swabs are taken.
+
+During the period we consider, the total cohort size expands due to the continuous recruitment into the study.
+Following this, the number then decreases and stabilises as those recruited to meet the October target transition from weekly onto monthly testing.
+See figure \@ref(fig:test-schedules) for further detail.
+
+\todo[inline]{Add some descriptive figures e.g. those in \url{~/COVID/ons-incidence/duration_estimation/reproducible-sims/reports/simulation-studies.pdf}}
+
+\section{Thesis outline}
+
 \chapter{Duration from biomarkers} \label{ATACCC}
+
+\section{Background}
+
+\section{Original analysis}
+
+\section{Comparison to \texorpdfstring{\textcite{hakkiOnset}}{Hakki \etal (2022)}}
+
+\section{Final analysis}
 
@@
 
+\chapter{Statistical backcalculation}
+
+\chapter{Mechanistic modelling}
+
+\chapter{Conclusion}
+
 \appendix
@@
 
-\chapter{Text to use}
-
-Random bits of text that I've written that don't currently fit anywhere but will likely be useful in future
-
-\section{CIS study description}
-The CIS continually enrols individuals and tests them on an ongoing basis, following a
-specified testing protocol. The protocol specifies that the initial five
-tests are spaced every 7 days, then the spacing is every 28 days.
-However, real-world considerations often lead to variations in test
-times and missed tests. Our analysis focuses solely on the binary result
-(positive or negative) of these tests, assuming no misclassification
-bias.
-
-The CIS (Coronavirus Infection Survey) is a longitudinal study on a representative sample of households.
-A full description of the study can be found in the supplementary materials of \textcite{pouwelsCommunity} or the study protocol~\cref{cisProtocol}.
-In brief, households are invited to the study from databases held by the ONS.
-The number invited was to satisfy various target number of individuals swabbing per fortnight; relevant to the discussion in this document is that the target increased from 2,500 by the end of May 2020 to 150,000 in October 2020.
-Within a household, all individuals aged 2 and over were invited to participate.
-Once invited, an enrolment swab would be taken followed by 4 further weekly swabs (giving a total of 5 swabs on days 0, 7, 14, 21, 28 relative to enrolment) after which monthly swabs are taken.
-
-During the period we consider, the total cohort size expands due to the continuous recruitment into the study.
-Following this, the number then decreases and stabilises as those recruited to meet the October target transition from weekly onto monthly testing.
-See figure \@ref(fig:test-schedules) for further detail.
-
-\todo[inline]{Add some descriptive figures e.g. those in \url{~/COVID/ons-incidence/duration_estimation/reproducible-sims/reports/simulation-studies.pdf}}
-
 \end{document}

### 2023-06-10 cis-imperfect-testing.tex
Add literature

@@
 Furthermore, due to the truncation, there is little or no ability to upper bound the number of very short episodes because these would almost all be missed.
-\Textcite{shenNonparametrica} (building on \textcite{panNote}) showed that in a related setting that the maximum likelihood (frequentist) estimator is inconsistent and can lead to severe underestimation (bias) of the survival function.
+\Textcite{shenNonparametrica} (building on \textcite{panNote}) showed, in a related setting, that the maximum likelihood (frequentist) estimator is inconsistent and can lead to severe underestimation (bias) of the survival function.
 While the result is not directly applicable, and can not simply be extended, to the more complex scenario within CIS, it does still build an intuition that these single positive episodes could be causing this issue.
@@
 This can occur for a variety of reasons, most commonly that an individual swabs themselves poorly and hence the concentration of virus on the swab is too low to be detected by the PCR process (see \cref{E-intro:sec:PCR}).
-\todo{Ref to section in intro explaining PCR testing}
 Plausibly, a variety of other reasons could also lead to a false negative such as logistical issues or mislabelling of samples.
@@
 First, it is well known that PCR testing can return false negatives \todo{refs for sensitivity} and I needed to explicitly include these in the viral load model of \cref{E-ATACCC}; they can be directly observed within CIS as intermittent negatives (see below).
-Second, adjusting the simulation to include false negatives can reproduce both the CIS data more faithfully and exhibits the same issues when estimating the duration distribution (see \cref{imperf-test:sec:simulate})
+Second, adjusting the simulation to include false negatives can reproduce both the CIS data more faithfully and exhibits the same issues when estimating the duration distribution (see \cref{imperf-test:sec:simulate}).
+
+\todo[inline]{These next two paragraphs are kinda disjointed but not sure how to bring them in}
 
@@
 An intermittent negative is when an individual tests negative but tested positive previously and subsequently.
-I explicitly included these in the viral load model of \cref{E-ATACCC} but ignored in \cref{E-perf-test}.
 Intermittent negatives are stripped out when creating the dataset used for all duration analyses in this chapter (as described in \cref{E-episode-def}), but demonstrate that false negatives do occur.
@@
 
-Each test that was previously positive, is now positive with probability $p_\text{sens}$.
-
 \section{Application to CIS data} \label{imperf-test:sec:application}
@@
 
+\ifSubfilesClassLoaded{
+  \listoftodos
+}{}
+
 \end{document}

### 2023-06-10 cis-perfect-testing.tex
Add literature

@@
 An event is \emph{censored} if the time of the event can only be bounded; it is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
-Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa (see \cref{perf-test:fig:double-interval-censor}); \textcite{bogaertsSurvival} review many statistical methods applicable to interval censoring.
-\emph{Double censoring} means that both the originating and terminating event are censored, and is also a well-studied problem, reviewed by \textcite{sunStatisticala}.
-In CIS, we have \emph{double interval censoring} meaning that both the initiating and the terminating events are interval censored.
+Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa (see \cref{perf-test:fig:double-interval-censor}).
+Left and right censoring can be viewed as special cases of interval censoring where the lower bound is negative infinity or the upper bound is positive infinity respectively.
+\emph{Double censoring} means that both the originating and terminating event are censored, and is also a well-studied problem.
+Doubly censored data is more challenging to handle because the initiating event time and duration must be jointly modelled~\autocite[and references therein]{liSemiparametric}.
+In CIS, we have \emph{double interval censoring} meaning that both the initiating and the terminating events are interval censored; approaches are reviewd by \textcite{sunAnalysis,bogaertsSurvival}.
 \todo{double interval censoring refs}.
@@
 
-This combination of censoring and truncation are rarely encountered in biomedical data\todo{refs to combining double interval censoring with truncation}.
-A more familiar setting for this work
+Few studies consider the combination of both double censoring (of any type) and truncation, especially within the human biostatistical literature, although the general frameworks developed in the 1970s do include this case \autocites{turnbullEmpirical}{dempsterMaximum}.
+Perhaps the main use of human studies involving double interval censoring and truncation are those estimating the incubation period of HIV/AIDS (the incubation period in that context is the time from being infected with HIV to developing AIDS symptoms).
+\Textcite{sunEmpirical,bacchettiNonparametric} discuss this, however, they use a conditional likelihood approach which conditions on both the truncation (a selection effect) and the time of the infection.
+The reason for the latter is unclear, perhaps though the infection time is considered to provide little information on the incubation period~\citePersonalComms{Nick Jewell}; a similar approach is taken by \textcite{shenNonparametric}.
+Within CIS, detected infections are more likely to occur in the period where there is frequent testing, and hence conditioning on the infection time ovestates the probability of the infection being detected.
+Furthermore, the interpretation and justification of a conditional likelihood approach is unclear in a Bayesian framework.
+
+A more common setting for this work is estimating the survival time of bird nests~\autocite{heiseyABCs}.
+In this setting, \textcite{heiseyModelling} develop a general framework allowing for arbitrary censoring patterns and double interval censoring; this is the approach I adopt in \cref{perf-test:sec:model}.
+In bird nesting studies, the visit times are normally global (all nests could be detected on the same days), while in CIS each individual has their own visit schedule; with global visit times the probability of being detected is the same for all participants in the study, simplifying the likelihood.
+
+I use a Bayesian framework for this analysis for two reasons.
+First, asymptotic results for the variance of frequentist estimators do not apply.
+\todo{Check why asymptotic results don't work}
+This complicates the construction of confidence intervals, which may require bootstrapping or other computationally intensive methods.
+Second, a prior naturally allows the inclusion information from previous studies, such as that from \cref{E-ATACCC}, and/or a belief that the hazard changes smoothly (see \cref{perf-test:sec:parameters-priors}).
+These assumptions may be needed to combat the lack of information in the data due to the trunction~\autocite{caoBias}.
+
+Bayesian methods have previously been applied to nest survival~\autocites{heBayesiana}{heBayesian}{caoModeling} but these augment the data with unobserved times of infection; this would be computationally expensive with the large number of episodes in CIS (>4000 in the time period considered in the main analysis of \cref{E-imperf-test}).
+I develop a Bayesian framework where there is no explicit data augmentation, allowing the method to scale to the large number of episodes present in CIS without requiring vast computing resources (such as a high-performance cluster), as only limited resources are available within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
 
@@
 \section{Modelling the duration}\label{perf-test:sec:model}
-In this section, I develop a model accounting for double interval censoring and truncation, building on prior work in both biomedical and bird nesting studies.
-The arbitrary nature of truncation in CIS, where each individual has a unique truncation pattern and is neither simple left nor right truncation, has not previously been explored in the literature.
-Furthermore, I explore a Bayesian framework where the amount of explicit data augmentation is limited allowing the method to scale to the large number of episodes present in CIS without requiring vast computing resources (such as a high-performance cluster), as only limited resources are available within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
-Therefore, unlike most Bayesian methods (e.g.: \textcite{heBayesiana}, \textcite{heBayesian}, and \textcite{caoModeling}) we avoid augmenting the data with unobserved times of infection or any information on the truncated infections; this would be computationally expensive with the large number of episodes in CIS (>4000 in the time period considered in the main analysis of \cref{E-imperf-test}).
-
+In this section, I develop a model accounting for double interval censoring and truncation.
 I model only individuals with detected episodes.

### 2023-06-10 thesis.tex
Add literature

@@
 
-\newcommand{\citePersonalComms}{(personal communication)}
+\newcommand\citePersonalComms[1]{(#1, personal communication)}
 
@@
 The CIS (Coronavirus Infection Survey) is a longitudinal study on a representative sample of households.
-A full description of the study can be found in the supplementary materials of \textcite{pouwelsCommunity} or the study protocol~\cref{cisProtocol}.
+A full description of the study can be found in the supplementary materials of \textcite{pouwelsCommunity} or the study protocol~\autocite{cisProtocol}.
 In brief, households are invited to the study from databases held by the ONS.
@@
 
+\listoftodos
+
+\printbibliography
+
 \appendix

### 2023-06-11 cis-perfect-testing.tex
Remove completed to do

@@
 However, application in reality must account for misclassification bias, the subject of \cref{E-imperf-test}.
-
-\todo[inline]{%
-  Somehwere, I need to fit in what exactly is novel here, and issues with the pre-existing work.
-What to cover is the application to both double interval censored and arbitrarily truncated data (which existed in some of the theoretical frameworks but not used); the issues with the conditional likelihood method of \textcite{bacchettiNonparametric}; and the lack of much work this century.
-}
 

### 2023-06-11 cis-perfect-testing.tex
Refs for infectiousness vs PCR

@@
 Episode $i$'s \emph{duration of positivity}, $D_i$ is the number of days that they are detectable for; I use time discretised to days, therefore, the duration is at least 1 day.
-Note that the relationship between positivity (\ie the result of a PCR test) and infectiousness (\ie an individual's ability to pass on the virus) is complex, \todo{ref for PCR pos vs infectiousness} and beyond the scope of this thesis.
+Note that the relationship between positivity (\ie the result of a PCR test) and infectiousness (\ie an individual's ability to pass on the virus) is complex~\autocites{lascolaViral}{singanayagamDuration}, and beyond the scope of this thesis.
 I define an episode as starting at time $B_i$ and ends at $E_i$, hence $D_i = E_i - B_i + 1$.
@@
 \includegraphics[width=\textwidth]{cis-perfect-testing/regions_diag}
-\todo[inline]{Fix the impossible region to fill out the RHS; label the inadmissible regoin.}
-\caption[Admissible, inadmissible, and untruncated infections]{Admissible, $\alpha_i$, inadmissible, $\beta_i$ and
-untruncated, $\Omega_i^C$ regions for an episode $i$ in an
+\caption[Admissible, inadmissible, and untruncated infections]{The regions (as defined in the main text) for an episode $i$ in an
 individual whose first test was at time 0 and negative, with subsequent
 negative tests at times 7, 14, and 56, and positive tests at times 21
-and 28. \label{perf-test:fig:partitionSpace}}
+and 28. Inadmissible region is unshaded. The black (impossible) region is because it does not make sense for recovery to occur prior to infection. \label{perf-test:fig:partitionSpace}}
 \end{figure}

### 2023-06-11 cis-perfect-testing.tex
Clearer discussion of ghost epsisodes

@@
 However, application in reality must account for misclassification bias, the subject of \cref{E-imperf-test}.
-
-\todo[inline]{Need to discuss interpretation of the ghosts in the individual model, why this is hard, and the implications for setting a prior here. Not even clear in simulation what it should be!}
 
@@
 In CIS, we have \emph{double interval censoring} meaning that both the initiating and the terminating events are interval censored; approaches are reviewd by \textcite{sunAnalysis,bogaertsSurvival}.
-\todo{double interval censoring refs}.
 \begin{figure}
@@
 Few studies consider the combination of both double censoring (of any type) and truncation, especially within the human biostatistical literature, although the general frameworks developed in the 1970s do include this case \autocites{turnbullEmpirical}{dempsterMaximum}.
-Perhaps the main use of human studies involving double interval censoring and truncation are those estimating the incubation period of HIV/AIDS (the incubation period in that context is the time from being infected with HIV to developing AIDS symptoms).
+Perhaps the main use of human studies involving both double censoring and truncation are those estimating the incubation period of HIV/AIDS (the incubation period in that context is the time from being infected with HIV to developing AIDS symptoms).
 \Textcite{sunEmpirical,bacchettiNonparametric} discuss this, however, they use a conditional likelihood approach which conditions on both the truncation (a selection effect) and the time of the infection.
@@
 
-\section{Modelling the duration}\label{perf-test:sec:model}
-In this section, I develop a model accounting for double interval censoring and truncation.
-I model only individuals with detected episodes.
-This approach is based on long-standing methodology~\autocites{heiseyModelling}{dempsterMaximum}{turnbullEmpirical}.
-The conceptual approach here is to form a cohort of individuals for which we observe at least one positive test, and then consider a study which had only enrolled these individuals.
-The modelling approach then corrects for this selection bias, by considering \emph{ghost} infections that occur in identical individuals but were truncated.
-
-I have also considered a model including individuals without any observed episodes (see \cref{E-appendix:total-model} for details).
-These individuals may have truncated episodes or may not have been infected at all.
-However, I found that this model has little difference in results while increasing computational complexity and requiring assumptions on reinfection (discussed in \cref{perf-test:sec:discussion}).
-
-\subsection{Notation and assumptions}
-\todo[inline]{Some of this will possibly be defined elsewhere, such as in the introduction where CIS is introduced}
+\section{Notation and assumptions}
+\todo[inline]{To be rewritten with some or all moved to the thesis introduction. Currently just being left here as a reference.}
 \begin{itemize}
@@
 
+\section{Modelling the duration}\label{perf-test:sec:model}
+
 \subsection{Likelihood}\label{perf-test:sec:likelihood}
 
-Episode $i$ can be fully characterised by the pair $(b_i, e_i)$ of
-start and end dates of the episode, which belong to the state space
-$T \times T$. If the individual in which the episode occurs is tested
-between $b_i$ and $e_i$ (inclusive), then we observe the episode,
-otherwise it is truncated. For the $i$th episode that we observed, we
-consider there is an unknown number, $n_{it}$, where $(b, e)$ are
-drawn from the same distribution as those that lead to episode $i$,
-and occur in an (imagined) individual identical to $i$, but were
-truncated. Episodes are also truncated if there is no negative test
-prior to their start, that is, the individual the episode occurred in
-had no tests before $b_i$. The truncated episodes are the ghosts of $i$.
+In this section, I develop a model accounting for double interval censoring and truncation.
+I model only individuals with detected episodes.
+This approach is based on long-standing methodology~\autocites{heiseyModelling}{dempsterMaximum}{turnbullEmpirical}.
+The conceptual approach here is to form a cohort of individuals for which we observe at least one positive test, and then consider a study which had only enrolled these individuals.
+The modelling approach then corrects for this selection bias, by considering \emph{ghost} infections that occur in identical individuals but were truncated.
+
+I have also considered a model including individuals without any observed episodes (see \cref{E-appendix:total-model} for details).
+These individuals may have truncated episodes or may not have been infected at all.
+However, I found that this model has little difference in results while increasing computational complexity and requiring assumptions on reinfection (discussed in \cref{perf-test:sec:discussion}).
+
+Ghost infections are a mathematical convenience to correct for truncated episodes.
+Consider that we draw many episodes independently, from a distribution of infection times and durations with all other characteristics identical (most notably the testing schedule and, if the model were to be extended, other covariates of the individual in which the episode occurs).
+This probability space is equivalent to any two random variables which can be over any two of infection time, recovery time, and duration.
+Conceptually, we will condition on the total number of episodes drawn (which is unknown) corresponding to the detected episode $i$, $n_i$, then consider the probability of observing exactly one episode and this episode occurring in the interval censored ranges observed (being \emph{admissible}), and then finally marginalise over $n_i$ to form the likelihood contribution for this episode.
+
+Unfortunately, there is no clean interpretation of the number of ghost episodes in terms of real-world quantities; in this sense, it is a pure model parameter.
+In reality, there is only one individual who is identical to the individual in which the episode occurs in all relevant ways.
+Even within the simulation study (see \cref{perf-test:sec:simulation-study}) these are not infections that are simulated, and understanding the ground truth number of ghost infections is not straightforward.
+
+To formalise this, note that all relevant information about episode $i$ can be fully characterised by the (unobserved) pair $(b_i, e_i)$ of start and end dates of the episode, which belong to the state space $T \times T$.
+If the individual in which the episode occurs is tested between $b_i$ and $e_i$ (inclusive), then the episode is detected, otherwise it is a ghost episode.
+For the $i$th detected episode, assume that there are $n_{it}$ ghost episodes.
+Further, we discard episodes if there is no negative test prior to their start, and hence also view any such episodes as ghost episodes.
 

### 2023-06-11 cis-perfect-testing.tex
Minor text editing

@@
 
-To formalise this, note that all relevant information about episode $i$ can be fully characterised by the (unobserved) pair $(b_i, e_i)$ of start and end dates of the episode, which belong to the state space $T \times T$.
+To formalise this, within the framework of \textcite{heiseyModelling}, note that all relevant information about episode $i$ can be fully characterised by the (unobserved) pair $(b_i, e_i)$ of start and end dates of the episode, which belong to the state space $T \times T$.
 If the individual in which the episode occurs is tested between $b_i$ and $e_i$ (inclusive), then the episode is detected, otherwise it is a ghost episode.
@@
 
-\Textcite{heiseyModelling} provide a framework for formalising this.
-Define three sets which partition the space of possible infection and
-recovery (\ie: are strict subsets of $T \times T$) given the
-observations associated with episode $i$ (graphically shown in Figure
-\ref{perf-test:fig:partitionSpace}). Each episode, whether observed or not, must
-fall into one of these three classes with a probability that is a
-function of $\theta$ (see section \ref{perf-test:sec:likelihood}).
+To relate this to the data, define three sets which partition the space of possible infection and recovery times (\ie: are strict subsets of $T \times T$), based on the observations associated with episode $i$ (graphically shown in Figure \ref{perf-test:fig:partitionSpace}).
+Each episode, whether observed or not, must fall into one of these three classes with a probability that is a function of $\theta$.
 

### 2023-07-10 cis-imperfect-testing.tex
Fix cross references

@@
 
-Applying the method of \cref{perf-test} to CIS data gave implausible estimates of the duration distribution (see \cref{imperf-test:fig:problem-cis-estimates}).
-In particular, I estimate a very high probability of the episode lasting less than five days, in conflict with the results of \cref{E-ATACCC} and other prior work (\eg: \autocite{cevikShedding}).
+Applying the method of \cref{E-perf-test} to CIS data gave implausible estimates of the duration distribution (see \cref{imperf-test:fig:problem-cis-estimates}).
+In particular, I estimate a very high probability of the episode lasting less than five days, in conflict with the results of \cref{E-ATACCC} and other prior work~\autocite[e.g.:][]{cevikShedding}.
 
@@
   \todo[inline]{Figure showing estimates without false negatives}
-  \caption[Estimating survival using CIS data assuming perfect testing]{Estimating the survival function using CIS data and the model of \cref{perf-test} gives implausibly short estimates.
+  \caption[Estimating survival using CIS data assuming perfect testing]{Estimating the survival function using CIS data and the model of \cref{E-perf-test} gives implausibly short estimates.
 \label{imperf-test:fig:problem-cis-estimates}}
@@
 with $y_i(t)$ still only defined for $t \in t_i$.
-
 However, an implausibly low value for $\psens$ is required to reproduce the rate of single positive episodes seen in CIS.
 \todo[inline]{Figure demonstrating this and references to what reasonable test sensitivities are, including the estimate from the ATACCC chapter}
+
 More plausibly, the rate of false negatives varies over the course of an episode.
@@
 Following the short, initial period of growth, viral load largely declines monotonically with time (see \cref{E-ATACCC}).
-\todo{Reference to ATCCCC chapter, although maybe the specific section with this result would be better}
 This suggests that a model with a declining test sensitivity as a function of time since infection might be more suitable.
 
+\todo[inline]{Next paragraph needs editing for clarity}
 Examination of the CIS data also provides evidence for a declining test sensitivity.
@@
 \end{equation}
-where $t$ is the time in days since infection
+where $t$ is the time in days since infection.
 Simulated data using this test sensitivity is much more similar to the data, while being more plausible than very low probabilities of false negatives.
@@
 
-In this section, we relax the assumption of perfect testing to allow
-some false negatives. We use a fairly simplified model to ensure the
-likelihood is tractable. This requires modifying both $p_{ia}$ to
-allow for the episode possibly being longer than observed, and
-$p_{it}$ to allow for additional episodes being missed.
+In this section, I introduce false negatives into the model of \cref{E-perf-test:sec:model}.
+We use a fairly simplified model to ensure the likelihood is tractable.
+This requires modifying both $p_{ia}$ to allow for the episode possibly being longer than observed, and $p_{it}$ to allow for additional episodes being missed.
 
@@
 
-The idea when modifying $p_{ia}$ is to allow for the negative test
-following the last positive to be either a true or false negative. If it
-is a true negative, then we have a bound on the length of the episode.
-If it is a false negative, then we consider the episode's length
-right-censored. The likelihood contribution for an episode is then a
-mixture of these two scenarios, with the mixture probability determined
-by the test sensitivity.
+The idea when modifying $p_{ia}$ is to allow for the negative test following the last positive to be either a true or false negative.
+If it is a true negative, then we have a bound on the length of the episode.
+If it is a false negative, then we consider the episode's length right-censored.
+The likelihood contribution for an episode is then a mixture of these two scenarios, with the mixture probability determined by the test sensitivity.
 
-For tractability, we assume that the tests bounding the start of the
-infection ($l_i^{(b)}$ and $r_i^{(b)}$) are true results and hence
-$b_i \in [l_i^{(b)}, r_i^{(b)}]$. Furthermore, we consider the tests
-only between $r_i^{(b)}$ and $l_i^{(e)}$ inclusive (the positive
-tests providing a lower bound on the length of the episode). Denote
-these tests by
-$t'_i = \{ t \in t_i : r_i^{(b)} \leq t \leq l_i^{(e)} \}$ and their
-results by $y_i'$. We know that the test results at times in $t_i'$
-are either true positives or false negatives. By definition, the test at
-$r_i^{(e)}$ is a false negative if and only if $e_i > r_i^{(e)}$. We
-proceed by considering the two cases.
+For tractability, we assume that the tests bounding the start of the infection ($l_i^{(b)}$ and $r_i^{(b)}$) are true results and hence $b_i \in [l_i^{(b)}, r_i^{(b)}]$.
+Furthermore, we consider the tests only between $r_i^{(b)}$ and $l_i^{(e)}$ inclusive (the positive tests providing a lower bound on the length of the episode).
+Denote these tests by $t'_i = \{ t \in t_i : r_i^{(b)} \leq t \leq l_i^{(e)} \}$ and their results by $y_i'$.
+We know that the test results at times in $t_i'$ are either true positives or false negatives. 
+By definition, the test at $r_i^{(e)}$ is a false negative if and only if $e_i > r_i^{(e)}$.
+We proceed by considering the two cases.
 
-First, if $e_i \leq r_i^{(e)}$. In this case, the test at
-$r_i^{(e)}$ is a true negative, as are all other tests not in
-$t_i'$, and these occur with probability 1.
+First, if $e_i \leq r_i^{(e)}$.
+In this case, the test at $r_i^{(e)}$ is a true negative, as are all other tests not in $t_i'$, and these occur with probability 1.
 \begin{align}
@@
 
-Second, if $e_i > r_i^{(e)}$. In this case, the test at
-$r_i^{(e)}$ is a false negative, occurring with probability
-$(1 - p_\text{sens})$, and we have no upper bound on the end of the
-infection.
+Second, if $e_i > r_i^{(e)}$.
+In this case, the test at $r_i^{(e)}$ is a false negative, occurring with probability $(1 - p_\text{sens})$, and we have no upper bound on the end of the infection.
 \begin{align}

### 2023-07-11 cis-perfect-testing.tex
Heavily edit the section on priors

@@
 
-\subsection{Independent priors}
-Take the standard assumption of independent priors on each parameter.
-A Beta distribution is then a natural choice for each hazard parameter, $\lambda_t$, as they are probabilities.
+\subsection{Independent priors} \label{perf-test:sec:independent-priors}
+Start with the standard assumption of independent priors on each parameter.
+Each hazard, $\lambda_t$, is a (conditional) probability, and hence a Beta distribution is a natural choice.
 
-An uninformative prior would be $\lambda_t \dist \Beta(\alpha, \alpha)$, with common choices of $\alpha$ being $0.5$, Jeffreys' Prior, or $1$, a flat prior.
-Flat or uninformative priors on the hazard lead to very informative priors on the survival, favouring short survival times (see \cref{perf-test:fig:flat-prior}).
-The sharp falloff can be seen by noting that these priors have expectation 0.5, hence $\E(S(t)) = \E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right) = 0.5^{t-1}$ which decreases very quickly.
+Uninformative Beta priors, with equal parameters, on the hazard are problematic in this context.
+An uninformative prior would be of the form $\lambda_t \dist \Beta(\alpha, \alpha)$.
+Commonly, $\alpha$ is chosen to be $0.5$ (Jeffreys' Prior) or $1$ (a uniform prior).
+Even though these priors are uninformative or flat on the hazard, they become highly informative on the survival. 
+Specifically, they tend to favour shorter survival times (see \cref{perf-test:fig:flat-prior}).
+An intuitive explanation of this can be drawn from the fact that these types of priors have an expected value of 0.5.
+The expected survival time, $\E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right)$, is therefore equal to $0.5^{t-1}$, a quantity that declines rapidly.
 This prior expresses extreme scepticism in estimates from previous studies, such as \cref{E-ATACCC} and the meta-analysis of \textcite{cevikShedding}.
-Therefore, I favour a prior which is weakly informative on the survival, or an informative prior used to represent our belief from the \cref{E-ATACCC} analysis.
 \begin{figure}
@@
 
-Instead, I propose a weakly informative prior Beta(0.1, 1.9), linking to previous literature for a point estimate and showing vague figures.
-
-A Beta(0.1, 1.9) prior has mean 0.05 while not having much information (the informativeness of a Beta distribution is related to $\alpha + \beta$ which here equals 2, the same as the flat prior case); and the central 95\% mass is 0.00--0.47.
-The central estimate reflects previous estimates that the median duration is in the range 15--20 days~\autocite{cevikShedding}, which would translate to hazard of around 0.05.
-The prior predictive distribution on the survival time implied by this prior is very vague (see \cref{perf-test:fig:vague-prior}).
+Instead, I propose a weakly informative prior Beta(0.1, 1.9), which has mean 0.05 and minimal information.
+The amount of information in a Beta distribution is related to the sum of its parameters.
+Here, the sum equals 2, the same as the flat prior case.
+The central 95\% probability mass of Beta(0.1, 1.9) is 0.00--0.47.
+The central estimate, of 0.05, is in line with previous estimates that the median duration is in the range 15--20 days~\autocite{cevikShedding}.
+This prior gives a very vague prior predictive distribution on the survival time (see \cref{perf-test:fig:vague-prior}).
 \begin{figure}
@@
 
-\subsection{Informative priors}
+\subsection{Smoothing priors}
 
-The analysis in \cref{E-ATACCC} can be used to provide prior information for the CIS analysis.
-In particular, the ATACCC study frequently samples individuals early in the infection.
-Therefore, the survival distribution from that analysis should be reliable through this period.
-CIS can then inform the remainder of the distribution where the previous analysis is based on extrapolation.
-
-The model structure assumed in \cref{E-ATACCC} leads to positive correlation in the posterior estimates of the hazard.
-It is important to propagate this correlation as part of the prior uncertainty into the present analysis.
-We do this in two parts: first, approximate the previous posterior estimate of the logit-hazard as a multivariate normal (which can be viewed as an approximation of Markov melding~\autocite{goudieJoining}).
-Then, use a Beta process to add additional uncertainty to this.
-
-There are two reasons to increase the uncertainty relative to the \cref{E-ATACCC} posterior.
-First, we do not have full belief in these estimates due to the extrapolation problem previously discussed.
-Second, the results may not fully generalise between the two studies.
-The samples are tested using different laboratories, with different PCR machines, and hence may not have the same characteristics~\autocite{punchooLaboratory}.
-The study participants, while both targeting a general population cohort, may also differ because ATACCC recruited through NHS Test and Trace while CIS recruited by sampling from a database of addresses (see \cref{E-intro:sec:studies}).
-
-The first step, approximating the \cref{E-ATACCC} posterior hazard as a multivariate normal on the logit scale, is done through the method of moments.
-The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC}).
-
-The discrete Beta process prior generalises the form of prior used in the previous section, allowing the central estimate of the hazard to vary over time~\autocites{ibrahimBayesian}{sunStatisticala}.
-\begin{align}
-  \lambda_t &\sim \text{Beta}(\alpha_t, \beta_t) &t = 1, 2, \dots \\
-  \alpha_t &= c_t h_t + \alpha_0 \\
-  \beta_t &= c_t (1 - h_t) + \beta_0
-\end{align}
-where $k_t$, $\alpha_0$, and $\beta_0$ are hyperparameters; and $h_t$ is a point estimate of the hazard at time $t$ from ATACCC.
-An intuition for what this distribution represents can be gained by considering the conjugate model for $\lambda_t$ with a beta prior and a binomial likelihood.
-If $\lambda_t$ is given the prior distribution $\text{Beta}(\alpha_0, \beta_0)$, and we then have $k_t$ observations with $k_t h_t$ successes, then the posterior distribution for $\lambda_t$ is $\text{Beta}(\alpha_t, \beta_t)$ (as defined above).
-
-\subsection{Smoothing priors}
-The final class of priors I consider are ones that aim to favour smoothness in the hazard.
-This is the Bayesian equivalent of a penalised likelihood, used frequently in the non-parametric frequentist setting (\eg: \autocite{bacchettiNonparametric}).
-Biological and epidemiological considerations mean that it would be surprising if the hazard varied significantly day-to-day.
-A smoothing prior encodes this information within the analysis.
-
-I consider two forms of smoothing priors: first-order and second-order random walks on the logit-hazard, $\logit(\lambda_t) = \log(\lambda_t) - \log(1 - \lambda_t)$.
-The logit transformation maps from the $[0, 1]$ interval to the full real line.
-\todo{The logit intro will be in the previous subsection}
-Random walks are a very common form of smoothing prior for discrete time.
-
-A first-order random walk prior encodes that the hazard should be constant with some random changes at each time step.
-\begin{align}
-  \logit\lambda_{t+1} &= \logit\lambda_t + \sigma \epsilon_t &\text{for $t \geq 1$} \\
-  \epsilon_t &\dist N(0, 1) \\
-  \logit \lambda_1 &\dist N(\mu_1, \sigma_1^2)
-\end{align}
-The hyperparameter $\sigma$ controls the standard deviation of the random walk, and the prior on $\lambda_1$ gives the starting point of the random walk.
-Smaller values favour a smoothing process.
+Smoothing priors encode the biological consideration that the hazard should not vary significantly from day-to-day.
+It is the Bayesian equivalent of a penalised likelihood, used frequently in the non-parametric frequentist setting (\eg: \autocite{bacchettiNonparametric}).
+Many possible forms of smoothing priors exist, including splines, Gaussian Processes, and random walks.
+I use a second-order random walk because it is simple and can produce sensible prior predictive distributions.
 
@@
 The hyperparameter $\sigma$ controls the standard deviation of the random walk, and there is an additional prior on $\epsilon_1$ for the initial gradient of the change in the hazard.
+
+\subsection{Model combination priors}
+
+I propose combining the information in CIS with \cref{E-ATACCC}'s analysis by using \cref{E-ATACCC}'s analysis as a prior for this analysis.
+This is desirable because the ATACCC study frequently samples individuals early in the infection.
+Therefore, that analysis's estimate of the survival distribution in the early infection period should be reliable.
+In the later infection period, the ATACCC study has fewer observations and hence the estimates are less reliable; CIS can then inform the posterior in this region.
+
+In the combination methodology, two aspects need to be considered.
+Firstly, the model structure from \cref{E-ATACCC} leads to positive correlation in the posterior estimates of the hazard, which should be propagated into this analysis.
+That is, the prior used in this analysis cannot be independent across the hazards.
+Secondly, the uncertainty in the estimates from \cref{E-ATACCC} should be increased for this analysis.
+The additional uncertainty is because the \cref{E-ATACCC} analysis extrapolates beyond the data using strong model assumptions (see \todo{insert location}).
+Furthermore, there are differences in the study design and laboratories used between the two studies (see \cref{E-intro:sec:studies}) which may mean that results do not generalise between the studies.
+I form the prior for the combination in two steps.
+
+I first approximate the \cref{E-ATACCC} posterior estimate of the hazard as a multivariate normal on the logit scale.
+This can be viewed as an approximation of Markov melding~\autocite{goudieJoining}.
+Using a multivariate normal, as opposed to multiple univariate distribution, ensures that the correlation between the hazards is preserved.
+The logit transformation maps from the $[0, 1]$ interval to the full real line.
+To estimate the parameters of the multivariate normal, I use the method of moments.
+The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC}).
+
+Having approximated the estimate as a multivariate normal, I add additional uncertainty using a discrete Beta process.
+The discrete Beta process prior generalises the form of prior used in \cref{perf-test:sec:independent-priors} by allowing the central estimate of the hazard to vary over time~\autocites{ibrahimBayesian}{sunStatisticala}.
+It is:
+\begin{align}
+  \lambda_t &\sim \text{Beta}(\alpha_t, \beta_t) &t = 1, 2, \dots \\
+  \alpha_t &= c_t h_t + \alpha_0 \\
+  \beta_t &= c_t (1 - h_t) + \beta_0
+\end{align}
+where $k_t$, $\alpha_0$, and $\beta_0$ are hyperparameters; and $h_t$ is a point estimate of the hazard at time $t$ from ATACCC.
+An intuition for what this distribution represents can be gained by considering the conjugate model for $\lambda_t$ with a beta prior and a binomial likelihood.
+If $\lambda_t$ is given the prior distribution $\text{Beta}(\alpha_0, \beta_0)$, and we then have $k_t$ observations with $k_t h_t$ successes, then the posterior distribution for $\lambda_t$ is $\text{Beta}(\alpha_t, \beta_t)$ (as defined above).
 

### 2023-07-12 cis-perfect-testing.tex
Edit priors section

@@
 
-I propose combining the information in CIS with \cref{E-ATACCC}'s analysis by using \cref{E-ATACCC}'s analysis as a prior for this analysis.
+I propose combining the information in CIS with \cref{E-ATACCC}'s analysis by using \cref{E-ATACCC}'s analysis as a prior for the CIS analysis.
 This is desirable because the ATACCC study frequently samples individuals early in the infection.
@@
 I simulate each individual enrolled in CIS who was tested at least once between ?? and ??.
-All their tests between ?? and ?? are considered, and each individual is assigned an infection time $B_i$ uniformly at random from ??.
-\todo{check dates and periods}
+The individuals' testing schedules are exactly as observed in CIS.
+The infection times and durations of positivity are simulated.
+From this, a set of test results can be generated.
+\todo{check dates}
 
-I take as the ground truth distribution for duration of positivity a combination of the estimates from \cref{E-ATACCC} with an inflated tail to represent what is seen within CIS.
-The tail is modified based on Sarah Walker's (unpublished) work based on CIS data.
-These curves and the combined curve are compared in \cref{perf-test:fig:duration-dist}.
-Individual $i$'s duration of positivity, $D_i$, is then an independent draw from this distribution.
+To generate the durations, an assumption is required for the distribution of the duration of positivity.
+I base this assumption on the estimate from \cref{E-ATACCC} with an inflated tail to represent what is seen within CIS.
+The tail is modified based on an unpublished analysis of CIS data by Sarah Walker (described below).
+Sarah uses survival analysis to estimate the duration from CIS data.
+The initiating event is assumed known as the time the episode was detected.
+The final event is assumed interval-censored between the time of the final positive test and the subsequent negative test, or right-censored if a negative test has not yet been observed.
+A flexible, spline-based form is assumed for the baseline survival function~\autocite{roystonSTPM,roystonFlexible} with covariates introduced via proportional odds.
+By not accounting for either the truncation or the interval censoring of the initiating event, this analysis has competing biases which makes them hard to interpret~\autocite{cisMethodsONS}.
 
@@
 \end{figure}
-
-Sarah uses survival analysis to estimate the duration from CIS data.
-The initiating event is assumed known as the time the episode was detected.
-The final event is assumed interval-censored between the time of the final positive test and the subsequent negative test, or right-censored if a negative test has not yet been observed.
-A flexible, spline-based form is assumed for the baseline survival function~\autocite{roystonSTPM,roystonFlexible} with covariates introduced via proportional odds.
-By not accounting for either the truncation or the interval censoring of the initiating event, this analysis has competing biases which makes them hard to interpret~\autocite{cisMethodsONS}.
 
@@
 
-To form the duration distribution used in the simulation, we combine these two estimates.
+To form the duration distribution used in the simulation, we combine the two estimates.
 The first 30 days of the distribution is proportional to the ATACCC estimate, with the rest proportional to this CIS-based estimates.
@@
 Then the distribution used in the simulation is the normalised version of this: $f_S(t) = f'_S(t)/\sum_i f_S'(i)$.
+These curves and the combined curve are compared in \cref{perf-test:fig:duration-dist}.
+Individual $i$'s duration of positivity, $D_i$, is then an independent draw from this distribution.
 
-The first day individual $i$ can be positive is their day of infection, $B_i$, the last day is $E_i = B_i + D_i - 1$ where $D_i$ is the duration of positivity for individual $i$, drawn from $f_S$.
-Any test between $B_i$ and $E_i$ (inclusive) is positive; outside this period, all tests are negative.
+Individual $i$ is first detectable on their day of infection, $B_i$.
+$B_i$ is drawn uniformly between 100 days prior to the first day of testing included and the final day of testing included (inclusive).
+$i$ is detectable up to and including the last day of infection, $E_i = B_i + D_i - 1$, where $D_i$ is the duration of positivity for individual $i$, drawn from $f_S$.
+Therefore, tests between $B_i$ and $E_i$ (inclusive) are positive, and all other tests are negative.
 
-I then filter the episodes to replicate the process which will be used when working with the CIS data.
+I then replicate the process used when working with the CIS data by filtering the episodes.
 First, discard any episodes where there are no positive tests as they are never detected.
 Next, discard any episodes where there are no negative tests prior to the first positive; since there is no lower bound on their infection time, they contain little information, this selection effect was included when constructing $p_{it}$.
-Finally, of those remaining episodes, we select a random subset to match the number of episodes in the simulation to the number of episodes in the modelling cohort.
+Finally, of those remaining episodes, select a random subset to match the number of episodes in the simulation to the number of episodes in the modelling cohort.
 The last step is required because everyone is infected in the simulation which does not happen in reality.

### 2023-07-12 cis-perfect-testing.tex
First pass editing problem section
Done:
- Restructuring section so that it makes sense
- Majority of edits to section (see below)

Still outstanding:
- Edit pargraph introducting survival analysis
- Edit from "few studies" onwards
- Reread whole section for flow and clarity

@@
 \begin{document}
-\chapter{Survival analysis with perfect testing} \label{perf-test}
+\chapter{Estimating duration with infrequent testing} \label{perf-test}
 
@@
 
-\todo[inline]{Some of these definitions likely need to be moved elsewhere as they will be needed in other chapters (e.g. ATACCC analysis).}
-Consider the CIS data.
-It consists of a series of tests on the same individuals (longitudinal data), where some individuals have a series of positive tests.
-We wish to estimate the distribution of time for which an individual is positive for.
-An individual could have multiple infections, and hence I refer to \emph{infection episodes}.
-
-I define an individual as \emph{detectable} if they would test positive if tested, in the absence of misclassification error (which is dealt with in \cref{E-imperf-test}).
-Episode $i$'s \emph{duration of positivity}, $D_i$ is the number of days that they are detectable for; I use time discretised to days, therefore, the duration is at least 1 day.
-Note that the relationship between positivity (\ie the result of a PCR test) and infectiousness (\ie an individual's ability to pass on the virus) is complex~\autocites{lascolaViral}{singanayagamDuration}, and beyond the scope of this thesis.
-I define an episode as starting at time $B_i$ and ends at $E_i$, hence $D_i = E_i - B_i + 1$.
-
-An episode is \emph{detected} when they test positive at least once, if this occurs; episode $i$ is detected at time $t$ if and only if: $i$ is detectable at $t$, $i$ is tested at $t$, and the test returns positive (a false negative does not occur).
-
-The survival time function, parameterised by $\theta$, is the probability that an episode lasts at least $t$ days: $S_\theta(t) = \prob(D_i \geq t \mid B_i = b, \theta)$, where $D_i$ and $B_i$ are the duration and starting time of episode $i$ respectively.
-I make the standard assumption that $D_i$ and $B_i$ are independent and hence: $S_\theta(t) = \prob(D_i \geq t \mid \theta)$.
+The CIS infrequently tests each individual, with tests occurring either weekly or monthly (see \todo{refer to section with CIS intro}).
+The infrequent testing complicates analysis of the data because the infection episode endpoints (start and end times) are known imprecisely, furthermore, infection episodes can be undetected.
+This chapter develops a method to account for the infrequent testing; this section describes my approach and its background within the relevant statistical literature.
 
@@
 
-The reason for this approach, over modelling biomarkers (as in \cref{E-ATACCC}) is due to two factors.
-First, the large sample size in CIS means that approaches where the number of parameters scales with the number of observed infections (as it does with random effect based approaches) would be computationally challenging; this challenge is compounded by the data being stored in a trusted research environment limiting the computational power and tools available.
-Second, the data is coarse with most episodes having only a single positive test; therefore, the trajectory of the viral load would not be well-identified in an individual.
-Combined, these factors outweighed the downside of survival analysis only considering the binary outcome (\ie positive or negative) or the test rather than the additional information on the viral load.
+This chapter uses survival analysis rather than the biomarker approach in \autoref{E-ATACCC} for computational efficiency and data limitations.
+Computational efficiency is an issue because the number of random effects in the biomarker approach scales with the number of detected infection episodes.
+Therefore, the biomarker approach would be computationally challenging with the large sample size in CIS.
+The computational challenge is compounded by the data being stored in a trusted research environment limiting the computational power and tools available.
+CIS data is limited by the infrequent sampling, meaning that most episodes have only a single positive test.
+With only a single positive test, there is no information on the viral load trajectory, and hence the random effects are unlikely to be well-identified.
+The biomarker approach does use more information per sample, taking into account the viral load as well as the binary test result (\ie whether the result is positive or negative), however, this is unlikely to be sufficient to overcome these problems.
 
-The methods in this chapter deal with modelling two elements of the data-generating process, double interval censoring and truncation.
-False negatives are not dealt with and left to \cref{E-imperf-test}.
-
-Double interval censoring is a special case of censoring, a common and well-studied challenge encountered when dealing with survival data.
-An event is \emph{censored} if the time of the event can only be bounded; it is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
-Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa (see \cref{perf-test:fig:double-interval-censor}).
-Left and right censoring can be viewed as special cases of interval censoring where the lower bound is negative infinity or the upper bound is positive infinity respectively.
-\emph{Double censoring} means that both the originating and terminating event are censored, and is also a well-studied problem.
-Doubly censored data is more challenging to handle because the initiating event time and duration must be jointly modelled~\autocite[and references therein]{liSemiparametric}.
-In CIS, we have \emph{double interval censoring} meaning that both the initiating and the terminating events are interval censored; approaches are reviewd by \textcite{sunAnalysis,bogaertsSurvival}.
-\begin{figure}
-  \centering \includegraphics{cis-perfect-testing/double-interval-censor}
-  \caption[Double-interval censoring in CIS data]{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval.
-Demonstrated here by a participant who is recorded negative at time 7 to positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode.
-\label{perf-test:fig:double-interval-censor}}
-\end{figure}
-
-The second element is truncation.
-Truncation occurs because we do not observe some episodes: an episode can occur between tests and hence the episode is never detected (see \cref{perf-test:fig:truncation}).
-We do not know how many infections we do not detect.
-Failure to model truncation will lead to estimating too long durations because the detected episodes are a biased sample with a longer duration.
-To see why they are shorter, consider the undetected infection in the bottom of \cref{perf-test:fig:truncation}.
-This individual was infected at time 33 for 10 days.
-If instead there infection was 24 days (or longer), the individual would still have been detectable at their next test, and therefore would have been detected.
-
-Truncation is a large problem within CIS, especially when individuals move to the more infrequent testing schedule (described in \cref{E-intro:sec:cis}).
-Consider that, based on the estimates in \cref{E-ATACCC}, the median length of an episode is 14 days \todo{check this number} we would not expect many episodes to last the 28 days that can occur between tests.
+The first complication is that infections episodes can be undetected, meaning no there are no positive tests associated with them.
+This happens when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
+For these undetected episodes, there is no information on them meaning that we do not even know how many of them occurred.
+Undetected episodes are important to account for because they are systematically shorter than detected episodes.
+To see why this is the case, consider the undetected infection in the bottom of \cref{perf-test:fig:truncation}.
+The infection episode depicted begins at time 33 and has a 10-day duration.
+If the duration was 24 days (or longer), the individual would still have been detectable at their next test, and therefore would have been detected.
+Undetected infections are a large problem in CIS because individuals are tested rarely compared to the duration of an infection episode, particularly when tested only every 28 days.
+However, we do still know the testing schedule of the individual as they are in the CIS cohort.
+Undetected events where we have some data on the events in which these events occur is an unusual situtation.
+Far more common is that there is no data on these events, a situation known as \emph{truncation}.
+A situation that is particularly relevant is \emph{left truncation}, when only events lasting a minimum amount of time are observed.
 \begin{figure}
@@
 \label{perf-test:fig:truncation}}
+\end{figure}
+
+The second complication is that the endpoints of detected infection episodes are known imprecisely.
+An event, such as an endpoint, is known \emph{censored} if the time of the event can only be bounded
+Censored data is present in almost all survival analyses.
+An event is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
+%Left and right censoring can be viewed as special cases of interval censoring where the lower bound is negative infinity or the upper bound is positive infinity respectively.
+Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa (see \cref{perf-test:fig:double-interval-censor}).
+\emph{Double censoring} means that both the originating and terminating event are censored, and is also a well-studied problem.
+Doubly censored data is more challenging to handle because the initiating event time and duration must be jointly modelled~\autocite[and references therein]{liSemiparametric}.
+CIS data is \emph{double interval censoring}, meaning that both the initiating and the terminating events are interval censored; approaches for double interval censoring are reviewed by \textcite{sunAnalysis,bogaertsSurvival}.
+\begin{figure}
+  \centering \includegraphics{cis-perfect-testing/double-interval-censor}
+  \caption[Double-interval censoring in CIS data]{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval.
+Demonstrated here by a participant who is recorded negative at time 7 to positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode.
+\label{perf-test:fig:double-interval-censor}}
 \end{figure}
@@
 
+I perform inference within the Bayesian paradigm.
 I use a Bayesian framework for this analysis for two reasons.
@@
 I develop a Bayesian framework where there is no explicit data augmentation, allowing the method to scale to the large number of episodes present in CIS without requiring vast computing resources (such as a high-performance cluster), as only limited resources are available within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
-
-While this chapter sets up the framework for analysing the CIS data, applying the method to CIS data requires modelling false negatives.
-That is the subject of \cref{E-imperf-test} and hence a full description of the data and processing of it is deferred to that chapter.
 

### 2023-07-12 thesis.tex
First pass editing problem section
Done:
- Restructuring section so that it makes sense
- Majority of edits to section (see below)

Still outstanding:
- Edit pargraph introducting survival analysis
- Edit from "few studies" onwards
- Reread whole section for flow and clarity

@@
 
+\section{Thesis outline}
+
+\chapter{Incidence, prevalence, and duration}
+We wish to estimate the distribution of time for which an individual is positive for.
+An individual could have multiple infections, and hence I refer to \emph{infection episodes}.
+
+I define an individual as \emph{detectable} if they would test positive if tested, in the absence of misclassification error (which is dealt with in \cref{E-imperf-test}).
+Episode $i$'s \emph{duration of positivity}, $D_i$ is the number of days that they are detectable for; I use time discretised to days, therefore, the duration is at least 1 day.
+Note that the relationship between positivity (\ie the result of a PCR test) and infectiousness (\ie an individual's ability to pass on the virus) is complex~\autocites{lascolaViral}{singanayagamDuration}, and beyond the scope of this thesis.
+I define an episode as starting at time $B_i$ and ends at $E_i$, hence $D_i = E_i - B_i + 1$.
+
+An episode is \emph{detected} when they test positive at least once, if this occurs; episode $i$ is detected at time $t$ if and only if: $i$ is detectable at $t$, $i$ is tested at $t$, and the test returns positive (a false negative does not occur).
+
+The survival time function, parameterised by $\theta$, is the probability that an episode lasts at least $t$ days: $S_\theta(t) = \prob(D_i \geq t \mid B_i = b, \theta)$, where $D_i$ and $B_i$ are the duration and starting time of episode $i$ respectively.
+I make the standard assumption that $D_i$ and $B_i$ are independent and hence: $S_\theta(t) = \prob(D_i \geq t \mid \theta)$.I utilise survival analysis for analysing this data.
+Survival analysis is the area of statistics concerned with estimating the distribution of the times between events.
+It is broadly applicable to many domains, such as the time-to-failure for a mechanical system or the effect of a treatment on time spent in hospital.
+In the context of this thesis, the distribution to estimate is that of duration of positivity.
+The \emph{initiating event} is when the duration starts, \ie becoming detectable.
+The \emph{terminating event} is when the duration ends, \ie becoming no longer detectable.
+
+\chapter{Studies used in this thesis} \label{intro:sec:studies}
+
 \section{PCR testing} \label{intro:sec:PCR}
 
-\section{Incidence, prevalence, and duration}
+\section{Assessment of Transmission and Contagiousness of COVID-19 in Contacts}
 
-\section{Studies used in this thesis} \label{intro:sec:studies}
-
-\subsection{Assessment of Transmission and Contagiousness of COVID-19 in Contacts}
-
-\subsection{Coronavirus (COVID-19) Infection Survey} \label{intro:sec:cis}
+\section{Coronavirus (COVID-19) Infection Survey} \label{intro:sec:cis}
 
@@
 
-\section{Thesis outline}
+Consider the CIS data.
+It consists of a series of tests on the same individuals (longitudinal data), where some individuals have a series of positive tests.
 

### 2023-07-12 cis-perfect-testing.tex
Finishing editing of problem intro section

@@
 \begin{document}
+\section{Definitions}
+\todo[inline]{These will be defined elsewhere in the thesis, but are here so that this draft is understandable}
+\begin{itemize}
+\item
+  Time is discrete: $1, \dots, T$.
+\item
+  We observe $n_a$ episodes of infection, indexed by $i$.
+\item
+  The $i$ in which episode $i$ occurs is tested at times
+  $t_i = \{ t_{i,1}, \dots, t_{i,m_i} \}$.
+\item
+  Each observed episode has an unknown start of infection time $b_i$
+  and end of infection time $e_i$ (the first and last day that the
+  individual would test positive due to this episode respectively).
+  These are realisations of the random variables $B_i$ and $E_i$
+  respectively.
+\item
+  The duration of the infection is the number of days for which an
+  individual tests positive, $D_i = E_i - B_i + 1$.
+\item
+  We assume that, for all episodes $i$, $D_i$ is iid and independent
+  of the time of the infection. Define the survival function
+  $\prob(D_i \geq t \mid B_i = b, \theta) = \prob(D_i \geq t \mid \theta) = S_\theta(t)$,
+  where $\theta$ are the parameters controlling the survival
+  distribution (the discussion in this document is valid regardless of
+  the model specified for $S_\theta$ and hence we consider $\theta$
+  as an arbitrary vector of parameters).
+\item
+  The beginning of episode $i$ is known to occur in the interval
+  $[l_i^{(b)}, r_i^{(b)}]$, and similarly for the end of the infection
+  in $[l_i^{(e)}, r_i^{(e)}]$.
+\item
+  $y_i(t)$ for $t \in t_i$ is a binary indicator giving the test
+  result for individual $i$ at time $t$. Under the perfect testing
+  assumption, $y_i(t) = 1$ if and only if $b_i \leq t \leq e_i$.
+\item
+  Throughout, the convention that lower-case letters are realisations of
+  upper-case random variables is used.
+\end{itemize}
+
 \chapter{Estimating duration with infrequent testing} \label{perf-test}
@@
 
-The CIS infrequently tests each individual, with tests occurring either weekly or monthly (see \todo{refer to section with CIS intro}).
+The CIS infrequently tests each individual, with tests occurring either weekly or monthly (see\todo{refer to section with CIS intro}).
 The infrequent testing complicates analysis of the data because the infection episode endpoints (start and end times) are known imprecisely, furthermore, infection episodes can be undetected.
@@
 
-I utilise survival analysis for analysing this data.
+I use survival analysis for analysing this data.
 Survival analysis is the area of statistics concerned with estimating the distribution of the times between events.
@@
 
-Few studies consider the combination of both double censoring (of any type) and truncation, especially within the human biostatistical literature, although the general frameworks developed in the 1970s do include this case \autocites{turnbullEmpirical}{dempsterMaximum}.
-Perhaps the main use of human studies involving both double censoring and truncation are those estimating the incubation period of HIV/AIDS (the incubation period in that context is the time from being infected with HIV to developing AIDS symptoms).
-\Textcite{sunEmpirical,bacchettiNonparametric} discuss this, however, they use a conditional likelihood approach which conditions on both the truncation (a selection effect) and the time of the infection.
-The reason for the latter is unclear, perhaps though the infection time is considered to provide little information on the incubation period~\citePersonalComms{Nick Jewell}; a similar approach is taken by \textcite{shenNonparametric}.
-Within CIS, detected infections are more likely to occur in the period where there is frequent testing, and hence conditioning on the infection time ovestates the probability of the infection being detected.
-Furthermore, the interpretation and justification of a conditional likelihood approach is unclear in a Bayesian framework.
+Few studies consider the combination of both double censoring (of any type) and undetected or truncated events
+This is especially true within the human biostatistical literature.
+This literature though did generate general theoretical frameworks in the 1970s which include double censored and truncated case~\autocites{turnbullEmpirical}{dempsterMaximum}.
+Double censored and truncated data do occur in studies estimating the time from being infected with HIV to being diagnosed with AIDS (the disease that HIV causes) such as \textcite{sunEmpirical,bacchettiNonparametric}.
+These studies (and some others~\autocite{shenNonparametric}) take a conditional likelihood approach, conditioning on both the truncation (a selection effect) and the period where the infection is censored to\todo{clumsy sentence}.
+They condition on the time of infection is because this information was considered to provide little information on the incubation period~\citePersonalComms{Nick Jewell}.
+The period in which the infection is detected in CIS is likely to be in the period where there is frequent testing.
+Therefore, conditioning on this period overstates the probability of the infection being detected.
 
-A more common setting for this work is estimating the survival time of bird nests~\autocite{heiseyABCs}.
-In this setting, \textcite{heiseyModelling} develop a general framework allowing for arbitrary censoring patterns and double interval censoring; this is the approach I adopt in \cref{perf-test:sec:model}.
+Double interval censored and truncated data frequently occur when estimating the survival time of bird nests~\autocite{heiseyABCs}.
+In this context, \textcite{heiseyModelling} develop a general framework allowing for arbitrary censoring patterns and double interval censoring.
 In bird nesting studies, the visit times are normally global (all nests could be detected on the same days), while in CIS each individual has their own visit schedule; with global visit times the probability of being detected is the same for all participants in the study, simplifying the likelihood.
+I use this framework as the basis of the model I derive in \cref{perf-test:sec:model}.
 
-I perform inference within the Bayesian paradigm.
-I use a Bayesian framework for this analysis for two reasons.
+I perform inference within the Bayesian paradigm for two reasons.
 First, asymptotic results for the variance of frequentist estimators do not apply.
@@
 Second, a prior naturally allows the inclusion information from previous studies, such as that from \cref{E-ATACCC}, and/or a belief that the hazard changes smoothly (see \cref{perf-test:sec:parameters-priors}).
-These assumptions may be needed to combat the lack of information in the data due to the trunction~\autocite{caoBias}.
+These assumptions may be needed to combat the lack of information in the data due to the undetected episodes~\autocite{caoBias}.
+They can also be viewed as combining the information from the two analyses (see \cref{perf-test:sec:informative-priors}).
 
-Bayesian methods have previously been applied to nest survival~\autocites{heBayesiana}{heBayesian}{caoModeling} but these augment the data with unobserved times of infection; this would be computationally expensive with the large number of episodes in CIS (>4000 in the time period considered in the main analysis of \cref{E-imperf-test}).
-I develop a Bayesian framework where there is no explicit data augmentation, allowing the method to scale to the large number of episodes present in CIS without requiring vast computing resources (such as a high-performance cluster), as only limited resources are available within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
-
-\section{Notation and assumptions}
-\todo[inline]{To be rewritten with some or all moved to the thesis introduction. Currently just being left here as a reference.}
-\begin{itemize}
-\item
-  Time is discrete: $1, \dots, T$.
-\item
-  We observe $n_a$ episodes of infection, indexed by $i$.
-\item
-  The $i$ in which episode $i$ occurs is tested at times
-  $t_i = \{ t_{i,1}, \dots, t_{i,m_i} \}$.
-\item
-  Each observed episode has an unknown start of infection time $b_i$
-  and end of infection time $e_i$ (the first and last day that the
-  individual would test positive due to this episode respectively).
-  These are realisations of the random variables $B_i$ and $E_i$
-  respectively.
-\item
-  The duration of the infection is the number of days for which an
-  individual tests positive, $D_i = E_i - B_i + 1$.
-\item
-  We assume that, for all episodes $i$, $D_i$ is iid and independent
-  of the time of the infection. Define the survival function
-  $\prob(D_i \geq t \mid B_i = b, \theta) = \prob(D_i \geq t \mid \theta) = S_\theta(t)$,
-  where $\theta$ are the parameters controlling the survival
-  distribution (the discussion in this document is valid regardless of
-  the model specified for $S_\theta$ and hence we consider $\theta$
-  as an arbitrary vector of parameters).
-\item
-  The beginning of episode $i$ is known to occur in the interval
-  $[l_i^{(b)}, r_i^{(b)}]$, and similarly for the end of the infection
-  in $[l_i^{(e)}, r_i^{(e)}]$.
-\item
-  $y_i(t)$ for $t \in t_i$ is a binary indicator giving the test
-  result for individual $i$ at time $t$. Under the perfect testing
-  assumption, $y_i(t) = 1$ if and only if $b_i \leq t \leq e_i$.
-\item
-  Throughout, the convention that lower-case letters are realisations of
-  upper-case random variables is used.
-\end{itemize}
+Bayesian methods have previously been applied to nest survival~\autocites{heBayesiana}{heBayesian}{caoModeling}.
+These methods augment the data with the unobserved times of infection, creating a parameter per detected infection.
+This would be many parameters for CIS (which has over 4000 in the time period considered in the main analysis of \cref{E-imperf-test}) and therefore computationally expensive.
+I develop a Bayesian framework where there is no explicit data augmentation, allowing the method to scale to the large number of episodes present in CIS.
 
@@
 
-\subsection{Model combination priors}
+\subsection{Model combination priors} \label{perf-test:sec:informative-priors}
 

### 2023-07-13 cis-imperfect-testing.tex
Some edits on imperfect testing

@@
 
-The simplest model for false negatives is by introducing a constant \emph{test sensitivity}, a probability of returning a negative test even when detectable.
-Denoting the test sensitivity by $\psens$, we replace the deterministic $y_i(t)$ with a random binary variable such that:
+The simplest model for false negatives is a constant probability of returning a negative test even when detectable.
+The probability of testing positive given being detectable is known as the test sensitivity.
+The impact of a test sensitivity, $\psens$, less than perfect (\ie less than 100\%) is formalised by replacing the deterministic $y_i(t)$ with a random binary variable such that:
 \begin{equation}
-  \prob(y_i(t) = 1) \begin{cases}
+  \prob(Y_i(t) = 1) \begin{cases}
       \psens &b_i \leq t \leq e_i \\
@@
 \end{equation}
-with $y_i(t)$ still only defined for $t \in t_i$.
+with $Y_i(t)$ still only defined for $t \in t_i$.
 However, an implausibly low value for $\psens$ is required to reproduce the rate of single positive episodes seen in CIS.
@@
 
-In this section, I introduce false negatives into the model of \cref{E-perf-test:sec:model}.
-We use a fairly simplified model to ensure the likelihood is tractable.
+In this section, I introduce a simple model of false negatives into the model of \cref{E-perf-test:sec:model}.
+The simple model ensures the likelihood remains tractable.
 This requires modifying both $p_{ia}$ to allow for the episode possibly being longer than observed, and $p_{it}$ to allow for additional episodes being missed.
 
-The model is only extended to model a constant test sensitivity.
+A constant test sensitivity is assumed throughout this section.
 Further work is required to introduce the more realistic varying test sensitivity (discussed in \cref{imperf-test:sec:discussion}).
@@
 
-The idea when modifying $p_{ia}$ is to allow for the negative test following the last positive to be either a true or false negative.
-If it is a true negative, then we have a bound on the length of the episode.
+I modify $p_{ia}$ to allow the negative test following the last positive to be a false negative.
 If it is a false negative, then we consider the episode's length right-censored.
-The likelihood contribution for an episode is then a mixture of these two scenarios, with the mixture probability determined by the test sensitivity.
+However, if it is in fact a true negative, then we are in the same case as before with a bound on the length of the episode.
+A mixture of these scenarios then forms the episode's likelihood contribution, with the mixture probability determined by the test sensitivity.
 
-For tractability, we assume that the tests bounding the start of the infection ($l_i^{(b)}$ and $r_i^{(b)}$) are true results and hence $b_i \in [l_i^{(b)}, r_i^{(b)}]$.
-Furthermore, we consider the tests only between $r_i^{(b)}$ and $l_i^{(e)}$ inclusive (the positive tests providing a lower bound on the length of the episode).
+For tractability, assume that the negative test bounding the start of the infection, at $l_i^{(b)}$, is a true negative.
+This assumption is reasonable because the test sensitivity is high early in an infection, therefore $l_i^{(b)}$ is unlikely to be a false negative.
+\todo{Are these bounds correct?}
+We will consider only the tests between $r_i^{(b)}$ and $l_i^{(e)}$ inclusive (the positive tests providing a lower bound on the length of the episode).
 Denote these tests by $t'_i = \{ t \in t_i : r_i^{(b)} \leq t \leq l_i^{(e)} \}$ and their results by $y_i'$.
@@
 
-With false negatives, an episode with can be truncated in the following
+With false negatives, an episode with can be undetected in the following
 ways:

### 2023-07-14 cis-perfect-testing.tex
Edits for CIS perfect testing chapter

@@
 
-I take a survival analysis approach to model the coarse sampling of the CIS (see \cref{perf-test:sec:problem}).
+I take a Bayesian survival analysis approach to model the coarse sampling of the CIS (see \cref{perf-test:sec:problem}).
 Flat priors can be unintentionally very informative, and hence weakly informative priors should be used; in addition, the prior can be used to combine the information in the CIS with the results of \cref{E-ATACCC} (see \cref{perf-test:sec:parameters-priors}).
-Through a simulation study (see \cref{perf-test:sec:simulation-study}), I show that these methods perform well, and in this context are not sensitive to model or prior specification.
-However, application in reality must account for misclassification bias, the subject of \cref{E-imperf-test}.
+Through a simulation study, I show that these methods perform well, and in this context are not sensitive to model or prior specification (see \cref{perf-test:sec:simulation-study}).
+However, application in reality must account for false negatives, the subject of \cref{E-imperf-test}.
 
@@
 The CIS infrequently tests each individual, with tests occurring either weekly or monthly (see\todo{refer to section with CIS intro}).
-The infrequent testing complicates analysis of the data because the infection episode endpoints (start and end times) are known imprecisely, furthermore, infection episodes can be undetected.
-This chapter develops a method to account for the infrequent testing; this section describes my approach and its background within the relevant statistical literature.
+The infrequent testing complicates analysis of the data in two ways: infection episodes can be undetected and the infection episode endpoints (start and end times) are known imprecisely.
+This chapter develops a method to account for the infrequent testing.
+This section describes my approach and its background within the relevant statistical literature.
 
-I use survival analysis for analysing this data.
-Survival analysis is the area of statistics concerned with estimating the distribution of the times between events.
+I analyse the data using survival analysis.
+Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
+The event at the start of the time of interest is the \emph{initiating event}, and the event at the end of the time of interest is the \emph{terminating event}.
 It is broadly applicable to many domains, such as the time-to-failure for a mechanical system or the effect of a treatment on time spent in hospital.
-In the context of this thesis, the distribution to estimate is that of duration of positivity.
-The \emph{initiating event} is when the duration starts, \ie becoming detectable.
-The \emph{terminating event} is when the duration ends, \ie becoming no longer detectable.
+In the context of this thesis, the distribution to estimate is the duration of positivity.
+Therefore, the initiating event is when the duration starts, \ie becoming detectable; and the terminating event is when the duration ends, \ie becoming no longer detectable.
 
-This chapter uses survival analysis rather than the biomarker approach in \autoref{E-ATACCC} for computational efficiency and data limitations.
-Computational efficiency is an issue because the number of random effects in the biomarker approach scales with the number of detected infection episodes.
+\todo[inline]{I think the following paragraph is needed somewhere, to answer why I'm not doing the same thing as in ATACCC, but it feels like an aside here since the previous paragraph and the one after this one are more linked.}
+Here, I use survival analysis rather than the biomarker approach taken in \autoref{E-ATACCC} for computational efficiency and data limitations.
+In the biomarker approach, computational efficiency is an issue because the number of random effects scales with the number of detected infection episodes.
+There are a large number of detected episodes in CIS 
 Therefore, the biomarker approach would be computationally challenging with the large sample size in CIS.
-The computational challenge is compounded by the data being stored in a trusted research environment limiting the computational power and tools available.
-CIS data is limited by the infrequent sampling, meaning that most episodes have only a single positive test.
+The computational challenge is compounded by the data being stored in a trusted research environment which limits the computational power and tools available.
+The data limitation is due to the infrequent sampling in the CIS study design, meaning that most episodes have only a single positive test.
 With only a single positive test, there is no information on the viral load trajectory, and hence the random effects are unlikely to be well-identified.
-The biomarker approach does use more information per sample, taking into account the viral load as well as the binary test result (\ie whether the result is positive or negative), however, this is unlikely to be sufficient to overcome these problems.
+While the biomarker approach does use more information per sample by taking into account the viral load as well as the binary test result (\ie whether the result is positive or negative), however, this is unlikely to be sufficient to overcome these problems.
 
-The first complication is that infections episodes can be undetected, meaning no there are no positive tests associated with them.
+The first complication in the analysis is that infections episodes can be undetected, meaning no there are no positive tests associated with them.
 This happens when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
@@
 The second complication is that the endpoints of detected infection episodes are known imprecisely.
-An event, such as an endpoint, is known \emph{censored} if the time of the event can only be bounded
+An event, such as an endpoint, is known as \emph{censored} if the time of the event can only be bounded.
 Censored data is present in almost all survival analyses.
@@
 Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa (see \cref{perf-test:fig:double-interval-censor}).
-\emph{Double censoring} means that both the originating and terminating event are censored, and is also a well-studied problem.
+If both the initiaing and terminating events are censored, then the event is \emph{doubly censored}.
 Doubly censored data is more challenging to handle because the initiating event time and duration must be jointly modelled~\autocite[and references therein]{liSemiparametric}.
-CIS data is \emph{double interval censoring}, meaning that both the initiating and the terminating events are interval censored; approaches for double interval censoring are reviewed by \textcite{sunAnalysis,bogaertsSurvival}.
+In CIS data both the initiating and the terminating events are interval censored.
+Therefore, the CIS data is \emph{double interval censored}; approaches for double interval censoring are reviewed by \textcite{sunAnalysis,bogaertsSurvival}.
 \begin{figure}
@@
 
-Few studies consider the combination of both double censoring (of any type) and undetected or truncated events
+Few studies consider the combination of both double interval censoring and undetected or truncated events
 This is especially true within the human biostatistical literature.
-This literature though did generate general theoretical frameworks in the 1970s which include double censored and truncated case~\autocites{turnbullEmpirical}{dempsterMaximum}.
-Double censored and truncated data do occur in studies estimating the time from being infected with HIV to being diagnosed with AIDS (the disease that HIV causes) such as \textcite{sunEmpirical,bacchettiNonparametric}.
-These studies (and some others~\autocite{shenNonparametric}) take a conditional likelihood approach, conditioning on both the truncation (a selection effect) and the period where the infection is censored to\todo{clumsy sentence}.
-They condition on the time of infection is because this information was considered to provide little information on the incubation period~\citePersonalComms{Nick Jewell}.
+While this literature does contain theoretical frameworks which include double censored and truncated case~\autocites{turnbullEmpirical}{dempsterMaximum}, they were normally applied when the terminating event was either uncensored or right censored~\autocite{sunEmpirical,bacchettiNonparametric}.
+The approach of these studies (and elsewhere~\autocite{shenNonparametric}) was to take a conditional likelihood approach, conditioning on both the truncation (a selection effect) and the period in which the initiating event is known to occur.
+They justify conditioning on the time of the initiating event by arguing that this information provides little information on the distribution being estimated~\citePersonalComms{Nick Jewell}.
 The period in which the infection is detected in CIS is likely to be in the period where there is frequent testing.
@@
 
-The hazard is convenient to use because it is unconstrained except in the interval [0, 1]\todo{ref where been used previously}.
-Survival needs to be monotonic.
+The hazard is convenient to use because it is unconstrained except in the interval [0, 1].
+The [0, 1] interval is easily be mapped to the unconstrained reals using a logit transformation (where $\logit(x) = \log(x/(1-x))$).
+Conversely, the survival function, $S_\theta(t)$, needs to be monotonically decreasing.
+Therefore, there is a constraint on $S_\theta(t)$ which depends on $S_\theta(t-1)$.
+This complicates the inference process and cannot be removed through a simple transformation.
+
+The rest of this section considers appropriate priors for the hazard in this context.
 
 \subsection{Independent priors} \label{perf-test:sec:independent-priors}
-Start with the standard assumption of independent priors on each parameter.
+Take the standard assumption of independent priors on each parameter.
 Each hazard, $\lambda_t$, is a (conditional) probability, and hence a Beta distribution is a natural choice.
 
-Uninformative Beta priors, with equal parameters, on the hazard are problematic in this context.
+Uninformative Beta priors on the hazard are problematic in this context.
 An uninformative prior would be of the form $\lambda_t \dist \Beta(\alpha, \alpha)$.
 Commonly, $\alpha$ is chosen to be $0.5$ (Jeffreys' Prior) or $1$ (a uniform prior).
-Even though these priors are uninformative or flat on the hazard, they become highly informative on the survival. 
+Even though these priors are uninformative on the hazard, they become highly informative on the survival. 
 Specifically, they tend to favour shorter survival times (see \cref{perf-test:fig:flat-prior}).
@@
 A second-order random walk prior encodes that the hazard should be linearly changing with some random changes at each time step.
+Specifically, the difference between $\lambda_t$ to $\lambda_{t+1}$ is the difference between $\lambda_{t-1}$ and $\lambda_t$ plus some Gaussian-distributed noise.
+Formally:
 \begin{align}
@@
   \logit\lambda_2 &= \logit\lambda_1 + \epsilon_1 \\\
-  \epsilon_1 &\dist N(\mu_2, \sigma_2^2) \\
-  \logit \lambda_1 &\dist N(\mu_1, \sigma_1^2)
+  \epsilon_1 &\dist N(\mu_{\epsilon_1}, \sigma_{\epsilon_1}^2) \\
+  \logit \lambda_1 &\dist N(\mu_{\lambda_1}, \sigma_{\lambda_1}^2) \\
+  \sigma &\dist \Exponential(1/\mu_\sigma).
 \end{align}
-The hyperparameter $\sigma$ controls the standard deviation of the random walk, and there is an additional prior on $\epsilon_1$ for the initial gradient of the change in the hazard.
+The hyperparmeters $\mu_{\lambda_1}$ and $\sigma_{\lambda_1}$ specify the prior on the hazard at time 1; $\mu_{\epsilon_1}$ and $\sigma_{\epsilon_1}$ the prior on the initial gradient; and $\mu_\sigma$ the smoothness of the random walk.
+I use $\mu_{\lambda_1} = -17.5$, $\sigma_{\lambda_1} = 6$, $\mu_\epsilon = 1.09$, $\sigma_\epsilon = 0.03$, and $\mu_\epsilon = 0.1$ based on the analysis in \cref{E-ATACCC}.
 
@@
 
-I propose combining the information in CIS with \cref{E-ATACCC}'s analysis by using \cref{E-ATACCC}'s analysis as a prior for the CIS analysis.
+I propose combining the information from the CIS data with \cref{E-ATACCC}'s analysis by using \cref{E-ATACCC}'s analysis as a prior for the CIS analysis.
 This is desirable because the ATACCC study frequently samples individuals early in the infection.
@@
 Secondly, the uncertainty in the estimates from \cref{E-ATACCC} should be increased for this analysis.
-The additional uncertainty is because the \cref{E-ATACCC} analysis extrapolates beyond the data using strong model assumptions (see \todo{insert location}).
+The additional uncertainty is because the \cref{E-ATACCC} analysis extrapolates beyond the data using strong model assumptions (see \todo{reference discussion on extrapolation in ATACCC chapter}).
 Furthermore, there are differences in the study design and laboratories used between the two studies (see \cref{E-intro:sec:studies}) which may mean that results do not generalise between the studies.
@@
 The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC}).
+\todo[inline]{Add figure showing the approximation is good}
 
@@
 
-\subsection{Descriptive analyses of simulated data}
+\subsection{Implementation}
 
-\subsection{Inference and hyperparameters}
+Simulation was performed in R~4.2.0~\autocite{R4-2-0} using tidyverse~2.0.0~\autocite{tidyverse} and targets~1.1.3~\autocite{targetsPackage}.
 
-Stan including settings
+Inference was implemented in Stan via RStan~2.21.8~\autocite{rstan2-21-8} using default settings.
+Convergence was assessed using Rhat and ESS as described in \cref{E-ATACCC:sec:convergence}, and all runs checked for divergent transitions.
 
-Priors on missed infections.
+The prior on the total number of infections was TBD\todo{fill in once decided on total vs individual model}.
 
-Survival prior hyperparameters (maybe this should just go in the previous section though?)
-
-Criteria for assessing convergence (Rhat, ESS, divergent transitions)
+A framework for performing CIS-like simulations is available at \url{https://github.com/joshuablake/cisSimulation}.
+A framework for fitting the class of models described here and in \cref{E-imperf-test} is available at \url{https://github.com/joshuablake/cisDurationModel}.
+Code utilising these frameworks to perform the analysis described in this chapter is available at \url{https://github.com/joshuablake/cisRuns}.
 
@@
 
-Points to make:
+All analyses converged satisfactorily\todo{put numbers once confirmed exact analyses to use}.
+Repeating simulations showed very little variation in the results, due to the large sample size in CIS; therefore, only results from a single simulation are shown.
 
-\begin{enumerate}
-  \item Convergence assessment
-  \item Repeated simulation showed very little variation (large sample size)
-  \item Recovers true value of survival well with little difference between priors, \cref{perf-test:fig:survival-results}
-  \item RW2 performs poorly considering hazard and when little information in prior very uncertain, \cref{perf-test:fig:hazard-results}; not clear why this is (\cref{perf-test:fig:hazard-pairs-results})
-  \item Do I need some quantitive performance measures or is just showing plots enough?
-\end{enumerate}
+The true value of the survival function is very well recovered with little difference between priors (see \cref{perf-test:fig:survival-results}).
+
+The independent prior has the fewest assumptions and least prior information.
+Therefore, it leads to posterior estimates that are most uncertain.
+
+The second-order random walk is the only prior that smooths the hazard throughout the entire infection.
+This leads to a near-constant hazard, and hence exponential-like survival, in areas where there is little information (long infections).
 
@@
   \caption[Comparison of survival function estimates under different priors]{Comparison of the posterior estimate of the survival distribution (pointwise median and 95\% CrI) when using different priors on the hazard.
+Results using each of the three different priors discussed in \cref{perf-test:sec:parameters-priors} are shown.
 \label{perf-test:fig:survival-results}}
@@
   \caption[Comparison of hazard estimates under different priors]{Comparison of the posterior estimate of the hazard (pointwise median and 95\% CrI) when using different priors on the hazard.
+Results using each of the three different priors discussed in \cref{perf-test:sec:parameters-priors} are shown.
 \label{perf-test:fig:hazard-results}}
-\end{figure}
-
-\begin{figure}
-  \centering \includegraphics{cis-perfect-testing/hazard-pairs-results}
-  \todo[inline]{Probably scrap this (correlation) plot}
-  \caption[Correlation between hazard estimates]{Correlation between posterior estimates of the hazard when using the vague prior. \label{perf-test:fig:hazard-pairs-results}}
 \end{figure}
@@
 
-Discussion points:
-\begin{enumerate}
-  \item Other priors (GPs?)
-  \item Inclusion of covariates: age, previous exposure status, variant.
-  \item Individual vs total model
-  \item Consideration of smaller sample size
-\end{enumerate}
+This chapter shows that the model I propose can recover the true survival function well, is not sensitive to the prior used for the survival distribution, nor requires information on the number of infections that were undetected.
+However, the problem becomes substantially harder if false negatives are present, as we shall see in the next chapter.
+Unfortunately, false negatives do occur in the CIS data.
 
-Limitations to include:
-\begin{enumerate}
-  \item Assumption of constant incidence
-\end{enumerate}
+Various extensions could be worth exploring.
+Most importantly, covariates (\eg age, immunity from vaccination and/or previous infection, or variant) could be included.
+Immunity in particular affects the duration substantially, although age and variant may also make a contribution~\autocites{hakkiOnset}{russellWithinhost}.
+Covariates can be included with a variety of model, the simplest being a proportional hazards model.
+
+The form of priors used for the survival function could also be extended.
+In particular, the form of smoothing chosen here is very simple and could be replaced with a more flexible model, such as a Gaussian Process for additional flexibility and smoothing and different time-scales~\autocite{saulGaussian}.
+
+The prior would likely be more important if the number of detected infections was smaller.
+Therefore, exploring its impact would be important for designing future surveys.
+A particular application would be estimation in the early stages of an epidemic of a novel pathogen.
+Here, little would be known about the pathogen, but gathering information with robust estimates of uncertainty is important to inform the response.
+
+I have assumed constant and independent incidence (a priori).
+In this case, a sensitivity analysis (not shown) showed minimal impact on survival estimates.
+Furthermore, I chose a period where the incidence was approximately constant.
+However, this assumption could be important in other contexts, especially if the incidence is changing rapidly.
+The assumption of constant incidence can lead to biased estimates\todo{cite}.
 
 \section{Conclusion} \label{perf-test:sec:conclusion}
+This chapter has shown the feasibility of estimating the survival function from a study with a CIS-like testing schedule.
+While the infrequent testing schedule is not ideal, appropriate modelling can recover the true survival function well.
+Next, we will see that this model is inadequate when false negatives are present, such as they are within the CIS data.
 

### 2023-07-14 cis-perfect-testing/results.R
Edits for CIS perfect testing chapter

@@ -6,26 +6,48 @@ library(tidybayes)
 library(tidyr)
 source(here::here("utils.R"))
 
-tbl_posteriors = readRDS(here::here("cisRuns-output/perfect_posteriors.rds"))
+tbl_posteriors = readRDS(here::here("cisRuns-output/perfect_posteriors.rds")) |>
+    filter(
+        .width == 0.95, # Credible intervals are narrow anyway so only show 95%
+        survival_prior %in% c(
+            "ATACCC-new",
+            "RW2-infer-sigma",
+            "vague"
+        ),
+        missed_prior == "vague", # Use the reference prior on the number of missed infections as this doesn't matter
+    ) |>
+    mutate(
+        survival_prior = case_match(
+            survival_prior,
+            "ATACCC-new" ~ "Combination",
+            "RW2-infer-sigma" ~ "Smoothing (RW2)",
+            "vague" ~ "Independent (vague)"
+        ),
+    )
 truth = readRDS(here::here("cisRuns-output/input_curves.rds")) |>
   filter(source == "Combined")
 
 theme_survival_time_series = function() {
     rlang::list2(
         standard_plot_theming(),
-        scale_x_continuous(breaks = 0:100*14, minor_breaks = 0:100*2, limits = c(0, 50))
+        scale_x_continuous(breaks = 0:100*14, minor_breaks = 0:100*2, limits = c(0, 100)),
+        labs(
+            x = "t",
+            fill = "Hazard prior",
+            colour = "Hazard prior"
+        )
     )
 }
 
 surv_plot = tbl_posteriors |>
-    filter(.width == 0.95, survival_prior != "ATACCC-old", missed_prior == "vague") |>
       ggplot() +
       ggdist::geom_lineribbon(
         aes(time, S, ymin = S.lower, ymax = S.upper, fill = survival_prior, colour = survival_prior),
         linewidth = 1, alpha = 0.3
     ) +
     geom_line(aes(time, S), data = truth) +
-    theme_survival_time_series()
+    theme_survival_time_series() +
+    ylab("S(t)")
 ggsave(
     filename = here::here("cis-perfect-testing/survival-results.png"),
     plot = surv_plot,
@@ -36,14 +58,14 @@ ggsave(
 )
 
 hazard_plot = tbl_posteriors |>
-    filter(.width == 0.95, survival_prior != "ATACCC-old", missed_prior == "vague") |>
       ggplot() +
       ggdist::geom_lineribbon(
[…]

### 2023-07-14 cis-perfect-testing.tex
Figure for ATACCC approximation

@@
 To estimate the parameters of the multivariate normal, I use the method of moments.
-The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC}).
-\todo[inline]{Add figure showing the approximation is good}
+The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC-hazard} and \cref{perf-test:fig:approximate-ATACCC-survival}).
+\begin{figure}
+  \thisfloatpagestyle{empty}
+  \vspace{-2.5cm}
+  \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{cis-perfect-testing/ataccc-approximation-hazard}}
+  \caption[Approximating the ATACCC posterior hazard as a logit-normal]{Comparison of the posterior estimate of the hazard from \cref{E-ATACCC} and its approximation with a logit-normal distribution (kernal density estimate smoothed) for the first 15 hazards.
+\label{perf-test:fig:approximate-ATACCC-hazard}}
+\end{figure}
+\begin{figure}
+  \centering \includegraphics{cis-perfect-testing/ataccc-approximation-survival}
+  \caption[Approximating the ATACCC posterior survival as a logit-normal]{Comparison of the posterior estimate of the survival from \cref{E-ATACCC} and its approximation with a logit-normal distribution (median and central 95\% interval).
+The survival is a function of all the hazards up until that point, therefore the correlation between the hazards must be well-approximated to have these agree.
+\label{perf-test:fig:approximate-ATACCC-survival}}
+\end{figure}
 
@@
 While the infrequent testing schedule is not ideal, appropriate modelling can recover the true survival function well.
-Next, we will see that this model is inadequate when false negatives are present, such as they are within the CIS data.
 

### 2023-07-14 cis-perfect-testing/ataccc-approximation.R

@@ -5,7 +5,7 @@ library(tidybayes)
 library(tidyr)
 source(here::here("utils.R"))
 
-n_lambdas_to_plot = 30
+n_lambdas_to_plot = 41
 
 output_file = function(..., read_from_RDS = TRUE) {
     filename = here::here("ATACCC-distributions", ...)
@@ -34,7 +34,7 @@ mvnorm_to_dist = function(draw_matrix) {
 logit_normal_mean = output_file("logit_hazard_mean.rds")
 logit_normal_cov = output_file("logit_hazard_cov.rds")
 tbl_logit_normal = mvtnorm::rmvnorm(
-    1e3,
+    20e3,
     mean = logit_normal_mean[1:n_lambdas_to_plot],
     sigma = logit_normal_cov[1:n_lambdas_to_plot,1:n_lambdas_to_plot]
 ) |>
@@ -47,10 +47,48 @@ tbl_priors = bind_rows(
         mutate(version = "Normal approx"),
 )
 
-tbl_priors |>
+p_surv = tbl_priors |>
+    filter(time <= n_lambdas_to_plot - 1) |>
     group_by(time, version) |>
     median_qi(S) |>
     ggplot(aes(time, S, ymin = .lower, ymax = .upper, colour = version,
                 fill = version)) +
     geom_lineribbon(alpha = 0.4) +
-    facet_wrap(~version)
+    standard_plot_theming()
+ggsave(
+    filename = here::here(
+        "cis-perfect-testing/ataccc-approximation-survival.png"
+    ),
+    plot = p_surv,
+    width = 15,
+    height = 9,
+    units = "cm",
+    dpi = 300
+)
+
+p_hazard = tbl_priors |>
+    filter(time <= 11) |>
+    ggplot(aes(lambda, colour = version)) +
+    geom_density() +
+    facet_wrap(~time, scales = "free", ncol = 4) +
+    standard_plot_theming() +
+    labs(
+        x = expression(lambda),
+        y = "Density",
+        colour = ""
+    ) +
+    theme(
+        legend.position = "bottom",
+        axis.text.x = element_text(angle = 45)
+    )
+
+ggsave(
+    filename = here::here(
+        "cis-perfect-testing/ataccc-approximation-hazard.png"
+    ),
+    plot = p_hazard,
+    width = 15,
+    height = 20,
+    units = "cm",
+    dpi = 300
+)

### 2023-07-14 cis-perfect-testing.tex
Add figure with simulation durations

@@
 
-\begin{figure}
-  \todo[inline]{Find figure comparing different duration distributions}
-  \caption[Comparison of duration distributions]{Comparison of the duration distributions, see main text for details of how each is derived. \label{perf-test:fig:duration-dist}}
-\end{figure}
-
 % \todo[inline]{The following seems important to justify why we can't just use these estimates but is also basically an aside}
@@
 Individual $i$'s duration of positivity, $D_i$, is then an independent draw from this distribution.
+\begin{figure}
+  \centering \includegraphics{cis-perfect-testing/input-duration-dists}
+  \caption[Comparison of duration distributions]{Comparison of the duration distribution used for the simulation.
+ATACCC is the posterior mean from \cref{E-ATACCC}'s analysis, CIS is Sarah Walker's analysis, and Combined is the combination of these used for the simulation.
+See main text for details of each analysis.
+\label{perf-test:fig:duration-dist}}
+\end{figure}
 

### 2023-07-14 cis-perfect-testing/input-duration-dists.R
Add figure with simulation durations

@@ -1,0 +1,34 @@
+library(dplyr)
+library(ggplot2)
+library(patchwork)
+library(purrr)
+library(tidybayes)
+library(tidyr)
+source(here::here("utils.R"))
+
+ataccc_posterior = readRDS(here::here("cisRuns-output/input_curves.rds"))
+
+plot = ataccc_posterior |>
+  filter(between(time, 0, 40)) |>
+  mutate(source = case_match(
+    source,
+    "SARAH (CIS)" ~ "CIS",
+    .default = source
+  )) |>
+  ggplot(aes(time, S, colour = source)) +
+  geom_line() +
+  standard_plot_theming() +
+  scale_x_continuous(breaks = 0:100*14, minor_breaks = 0:100*2) +
+  labs(
+    x = "t",
+    colour = ""
+  )
+
+ggsave(
+  filename = "cis-perfect-testing/input-duration-dists.png",
+  plot = plot,
+  width = 15,
+  height = 9,
+  units = "cm",
+  dpi = 300
+)

### 2023-07-14 cis-perfect-testing.tex
Final edits for first draft of chapter

@@
 I perform inference within the Bayesian paradigm for two reasons.
-First, asymptotic results for the variance of frequentist estimators do not apply.
-\todo{Check why asymptotic results don't work}
-This complicates the construction of confidence intervals, which may require bootstrapping or other computationally intensive methods.
+First, the standard methods for estimating the parameter's variance and covariance are poor.
+Commonly used methods are based on the Fisher information matrix or resampling, but neither have theoretical justification~\autocite[230]{sunAnalysis}.
 Second, a prior naturally allows the inclusion information from previous studies, such as that from \cref{E-ATACCC}, and/or a belief that the hazard changes smoothly (see \cref{perf-test:sec:parameters-priors}).
@@
 The form of $S_\theta(t)$ is yet to be specified.
-As frequently observed in the literature\todo{refs for parameterisation}, it is more convenient to operate with the hazard than the survival or probability mass function directly.
+As frequently observed in the literature~\autoref[e.g.:][]{heBayesian}, it is more convenient to operate with the hazard than the survival or probability mass function directly.
 This is because the mass function must sum to one and the survival must be monotonically decreasing, while there are no constraints across multiple hazards.
@@
 
-I simulate each individual enrolled in CIS who was tested at least once between ?? and ??.
+I simulate each individual enrolled in CIS who was tested at least once between 10th Oct 2020 and 6th Dec 2020 (inclusive, the same period used for the later analysis using real CIS data).
 The individuals' testing schedules are exactly as observed in CIS.
@@
 From this, a set of test results can be generated.
-\todo{check dates}
 
@@
 However, this assumption could be important in other contexts, especially if the incidence is changing rapidly.
-The assumption of constant incidence can lead to biased estimates\todo{cite}.
+The assumption of constant incidence can lead to biased estimates~\autocite{degruttolaAnalysis}.
 

### 2023-07-18 incidence-prevalence.tex
Create incidence and prevalence chapter

--- 2023-06-09 appendix-cis-perfect-testing.tex
@@
 \begin{document}
-
-\chapter{Including all CIS individuals in survival analysis} \label{appendix:total-model}
-
-\todo[inline]{Add an intro to this and maybe a plot of results in simulation}
-
-We now consider all $N$ individuals in the CIS cohort who have at
-least one test during the period of interest, this includes those
-without an observed infection, and assume that each is equally likely to
-be infected. As before, $i = 1, \dots, n_a$ indexes episodes but now
-$j = 1, \dots, N$ indexes individuals. All episodes have an associated
-individual, each individual may have any number (including 0) episodes.
-Admissible regions are only defined for episodes, and have not changed
-from the previous section. The truncated region is defined for each
-individual, in addition we consider the combined truncated region which
-is episodes that would be truncated without conditioning on which
-individual the episode occurs in:
-$\Omega^C = \{ (b, e, j) : (b, e) \in \Omega_j^C \}$.
-
-Denote by $n_\text{tot}$ the total number of episodes across all $N$
-individuals, regardless of whether they are admissible or truncated (we
-know there are no inadmissible episodes by definition).
-
-For any infection, the probability of it being admissible for episode
-$i$ is $\frac{1}{N} p_{ia}$. That is, the probability that the
-infection occurs in the individual corresponding to the individual in
-which episode $i$ ($1/N$ by the assumption that episodes are equally
-likely to occur in any of the $N$ individuals) multiplied by the
-probability that the episode is admissible for episode $i$ conditional
-on it occurring in the relevant individual ($p_{ia}$).
-
-The total number of truncated infections that occur is
-$n_t = n_\text{tot} - n_a$. Conditional on an infection occurring in
-individual $j$, its probability of being truncated is $p_{jt}$ (as
-defined previously). Therefore, the overall probability of an episode
-being truncated is $p_t = \frac{1}{N} \sum_{j=1}^N p_{jt}$.
-
-Consider our data as there being one admissible episode for each
-observed episode, and all other episodes being truncated. As in the
-previous section, conditional on $n_\text{tot}$, we have a multinomial
-likelihood.
-\begin{align}
-p(\text{data} \mid \theta)
-&= \sum_{n_\text{tot}=n_a}^\infty p(\text{data} \mid n_\text{tot}, \theta) p(n_\text{tot}) \\
-&= \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} \left( \prod_{i=1}^{n_a} \frac{1}{N} p_{ia} \right) p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \\
-&= \left( \prod_{i=1}^{n_a} \frac{1}{N} p_{ia} \right) \left( \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \right) \\
-\end{align}
-
-In this case, it would be computationally feasible to augment the data
-with $n_\text{tot}$ directly (as it is only a single parameter).
-However, implementation is easier if we can use a closed-form for the
-sum (as in the previous section) as it allows the use of standard
-implementations of NUTS such as Stan, which cannot handle discrete
-parameters.
-
-Assuming $N_\text{tot} \dist \NegBin(\mu, r)$:
-\begin{align}
-&\sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \\
-&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot} \mid \lambda) p(\lambda) d\lambda \\
-&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} \frac{\lambda^n_\text{tot} e^{-\lambda}}{n_\text{tot}!} p(\lambda) d\lambda \\
-&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{1}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} \lambda^{n_\text{tot}-n_a} \lambda^{n_a} e^{-\lambda} p(\lambda) d\lambda \\
-&= \int \lambda^{n_a} e^{-\lambda} p(\lambda) \sum_{n_t=0}^\infty \frac{1}{n_t!} p_t^{n_t} \lambda^{n_t} d\lambda &n_t = n-n_a\\
-&= \int \lambda^{n_a} e^{-\lambda} p(\lambda) e^{\lambda p_t} d\lambda \\
-&= \int \lambda^{n_a} e^{-\lambda(1 - p_t)} p(\lambda) d\lambda \\
-&= \int \lambda^{n_a} e^{-\lambda(1 - p_t)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda \\
-&= \int \frac{b^a}{\Gamma(a)} \lambda^{a+n_a-1} e^{-(b+1-p_t)\lambda} \lambda d\lambda \\
-&= \frac{b^a}{\Gamma(a)} \frac{\Gamma(a+n_a)}{(b+1-p_t)^{a+n_a}} \\
-&\propto (b+1-p_t)^{-(a+n_a)} \\
-&= (r/\mu + 1 - p_t)^{-(r+n_a)} \\
-&\propto(r + \mu (1- p_t))^{-(r+n_a)}
-\end{align}
-
-Which gives the full posterior density as:
-\begin{align}
-p(\theta \mid \text{data})
-&\propto \frac{p(\theta) \prod_{i=1}^{n_a} p_{ia}}{(r + \mu(1 - p_t))^{r+n_a}}
-\end{align}
-
-As before, being able to reconstruct the posterior of $n_\text{tot}$
-using its full conditional is useful.
-\begin{align}
-p(n_\text{tot} \mid \text{data}, \theta)
-&\propto p(\text{data} \mid \theta, n_\text{tot}) p(n_\text{tot}) \\
-&= \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} \left( \prod_{i=1}^{n_a} \frac{1}{N} p_{ia} \right) p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \\
-&\propto \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}} \frac{\Gamma(r + n_\text{tot})}{n_\text{tot}!} \left( \frac{\mu}{r + \mu} \right)^{n_\text{tot}}  \\
-&\propto \frac{\Gamma(r + n_\text{tot})}{(n_\text{tot}-n_a)!} \left( \frac{\mu p_t}{r + \mu} \right)^{n_\text{tot}}  \\
-&\propto \frac{\Gamma((r + n_a) + (n_\text{tot}- n_a))}{(n_\text{tot}-n_a)!} \left( \frac{\mu p_t}{r + \mu} \right)^{n_\text{tot}-n_a}  \\
-\end{align}
-which is a negative binomial pmf with size parameter $r + n_a$ and
-probability parameter $\frac{r+\mu(1-p_t)}{r+\mu}$.
+\chapter{Incidence and prevalence} \label{incidence-prevalence}
 

### 2023-07-19 incidence-prevalence.tex
Start outlining chapter

@@
 
+This chapter provides the required understanding of incidence and prevalence, and the mathematical relationship between these quantities.
+These quantities have long been of interest, and the duration of infection is the quantity that relates them.
+The relationship under constant incidence was reviewed as early as 1980~\autocite{freemanPrevalence}.
+
+Want to be clear between: incidence number and incidence rate.
+
+Incidence is
+
+Prevalence is
+
+Throughout this thesis, time will almost always be considered discrete.
+This is formed by grouping any data by the day of the test, and assuming both incidence and prevalence are constant over that day.
+Most epidemiological data is only available at daily granularity or, when more granular data exists, is often unreliable.
+Furthermore, the variation in transmission that occurs due to daily patterns (\eg: sleep) is not of interest but complicates any analysis.
+
+Define the first day of an infection episode in individual $i$ as $B_i$, and assume that the probability of $i$ being infected multiple times within the period of interest is negligible.
+The time between being infected and first being detectable is short (see ), and therefore I often assume that $B_i$ is the same as the time of infection.
+I keep this assumption for the remainder of this chapter.
+
+The probability of being positive can be modelled in different ways depending on what assumptions are most reasonable.
+The most important assumption from a statistical perspective is the whether the probability of testing positive on a given day is independent conditional on the infection time.
+
+\todo[inline]{Is the discussion of this model just a distraction?}
+Assuming that the probability of testing positive depends only on the infection time implies no individual variation.
+This model is formalised as $Pr(Z_i(t) = 1 \mid B_i, Z_i(1), Z_i(2) \dots) = Pr(Z_i(t) \mid B_i) = p_{t-B_i}$, that is the probability of individual $i$ testing positive $t$ days after infection depends only on the infection start time and no other quantities.
+A slight relaxation of this model is introducing covariates, allowing some variation based on an individual's characteristics.
+
+The model with the most dependence between times, while remaining realistic, is a multi-state model.
+A multi-state model assumes that an individual always tests positive between the start of their infection, $B_i$, and the end of their infection, $E_i$.
+Therefore, if an individual has tested negative before the present time $t$ but after $B_i$ then they will test negative at time $t$.
+The complexity of the transitions between positive and negative can be very complex.
+I consider a slight extension to allow for the possibility of false negatives, meaning a negative test between $B_i$ and $E_i$, and that this probability may change.
+This leads to the following model.
+\begin{align}
+  \prob(Z_i(t) = 1 \mid B_i, \leq E_i) &= \begin{cases}
+    \psens(t - B_i) &B_i \leq t \leq E_i\\
+    0 &\text{otherwise}
+  \end{cases}\\
+  \prob(E_i = t \mid B_i) = f_i(B_i)
+\end{align}
+Conditional on both $B_i$ and $E_i$, then $Z_i(t)$ and $Z_i(t')$ are independent for $t \neq t'$.
+However, unlike the previous model, conditioning only on $B_i$ is not sufficient for $Z_i(t)$ and $Z_i(t')$ to be independent because each $Z_i(t)$ provides information on $E_i(t)$ (in particular, $Z_i(t) = 1$ implies $E_i > t$).
+For SARS-CoV-2, the latter model is more appropriate, although, as we shall see, the assumption that $\psens$ is independent of $t$ is violated.
+
+Backcalculation makes use of this relationship to estimate incidence from prevalence, assuming that the duration is known.
+
+Deterministic backcalculation assumes that the variance in the population is negligible, and therefore that the prevalence is equal to the mean prevalence.
+This is justified because the variance due to sampling is much larger than the variance due to the true prevalence.
+
+Robust, unbiased prevalence estimates already exist (see ), and incidence estimates are the most useful (see ).
+Therefore, a large portion of this thesis concerns itself with estimating the duration of PCR positivity.
+
+Two approaches are commonly used to estimate durations\todo{cite Sweeney}: either modelling an underlying biomarker or modelling the duration directly.
+For SARS-CoV-2, both approaches would be based on the results of PCR testing.
+Modelling a biomarker would mean modelling the viral load, measured as a Ct value.
+Modelling the duration would consider only the binary result at each test (positive or negative); in this thesis a survival analysis framework is adopted but other related approaches, such as a multi-state model, could also be considered.
+Therefore, the biomarker approach considers more information per observation, however, a model for the biomarker is required which generally requires stronger assumptions.
+Modelling a biomarker uses more information per observation, but requires stronger assumptions.
+
 \end{document}

### 2023-07-20 incidence-prevalence.tex
Add (messy) text
Currently pretty confused how to do this chapter,
just playing round with things hoping it sticks...

@@
 
-Prevalence is
+Prevalence, $P_t$, is the number of people in a given population with an infection at time $t$.
+For SARS-CoV-2, it is hard to define what it means to be infected.
+An infected individual could mean: one that is infectious; one that is symptomatic; or one that, if they were tested, would return a positive PCR test (see \cref{E-SARS-CoV-2} for what each of these mean and their relation).
+In this thesis, prevalence is defined as the number of people in the population who would return a positive PCR test at time $t$.
+
+A \emph{prevalence survey} is a survey that tests a random sample of the population for the presence of an infection.
+Prevalence surveys are the form of measuring prevalence that this thesis is concerned with.
+This thesis aims to infer the incidence rate from prevalence surveys, and therefore the likelihood of the results of the survey given the incidence is the most important quantity.
+At a time $t$ we sample $n_t$ individuals from the population of interest returning $x_t$ positive tests.
+Therefore, our likelihood contribution for each sample is $\p(x_t \mid n_t, \vec{I})$.
+We assume that the samples are independent of each other and $\vec{I}$, conditional on the true population prevalence at that time.
+Therefore, our overall likelihood is $\p(\vec{x} \mid \vec{n}, \vec{I}) = \int \p(\vec{P} \mid \vec{I}) \prod_t \p(x_t \mid n_t, \vec{P}_t) d\vec{P}$.
+
+Typically, the uncertainty in $\vec{P} \mid \vec{I}$ is negligible.
+This can be seen by considering the variance in $x_t$.
+$x_t$ is binomially distributed with parameters $n_t$ and $P_t / N$, where $N$ is the size of the population.
+Therefore:
+\begin{align}
+  \var\left( x_t \mid \vec{I} \right)
+    &= \var\left[\E\left( x_t \mid \vec{I}, \vec{P} \right) \mid \vec{I} \right] + \E\left[\var\left( x_t \mid \vec{I}, \vec{P} \right) \mid \vec{I} \right] \\
+    &= \var\left( n_t \frac{P_t}{N} \mid \vec{I} \right) + \E\left( n_t \frac{P_t}{N} \left(1 - \frac{P_t}{N} \right) \mid \vec{I} \right) \\
+    &= \left( \frac{n}{N} \right)^2 \var\left(P_t \mid \vec{I} \right) + \frac{n_t}{N} \E\left(P_t \mid \vec{I} \right)  - \frac{n_t}{N^2} \E\left(P_t^2 \mid \vec{I} \right)
+\end{align}
+since, in almost all settings, $n_t << N$ and $\E(P_t) \approx \var(P_t)$, the first term is negligible.
+Hence, assuming that $\var(\vec{P}) = \matr{0}$ is a reasonable approximation.
+This is equivalent to assuming that $\vec{P}$ is a deterministic function of $\vec{I}$, generally $\vec{P} = \E( \vec{P} \mid \vec{I})$.
+
+In \emph{deterministic backcalculation}, the observed prevalence, $x_t/n_t$, is assumed to be equal to $P_t/N$.
+However, this is often a poor approximation.
+\emph{Statistical backcalculation} retains the sampling distribution of $x_t$ (or an approximation of it, such as a Poisson).
 
@@
 Define the first day of an infection episode in individual $i$ as $B_i$, and assume that the probability of $i$ being infected multiple times within the period of interest is negligible.
-The time between being infected and first being detectable is short (see ), and therefore I often assume that $B_i$ is the same as the time of infection.
+The time between being infected and first being detectable is short (see\todo{ref relevant part}), and therefore I often assume that $B_i$ is the same as the time of infection.
 I keep this assumption for the remainder of this chapter.
@@
 
-Robust, unbiased prevalence estimates already exist (see ), and incidence estimates are the most useful (see ).
+Robust, unbiased prevalence estimates already exist (see\todo{ref prevalence estimates}), and incidence estimates are the most useful (see\todo{ref intro with importance of incidence}).
 Therefore, a large portion of this thesis concerns itself with estimating the duration of PCR positivity.

### 2023-08-10 cis-perfect-testing.tex
Updates from Overleaf

@@
 The estimation of incidence from prevalence is sensitive to the tail of the duration distribution.
-In \cref{E-ATACCC} I estimated the duration distribution from the ATACCC data, however, this study suffered from a small sample size and only 20 days of follow-up.
+In \cref{E-ATACCC} I estimated the duration distribution from the ATACCC dataset, however, this study suffered from a small sample size and only 20 days of follow-up.
 Therefore, the estimation of the tail of the duration distribution is driven by model assumptions.
@@
 I take a Bayesian survival analysis approach to model the coarse sampling of the CIS (see \cref{perf-test:sec:problem}).
-Flat priors can be unintentionally very informative, and hence weakly informative priors should be used; in addition, the prior can be used to combine the information in the CIS with the results of \cref{E-ATACCC} (see \cref{perf-test:sec:parameters-priors}).
-Through a simulation study, I show that these methods perform well, and in this context are not sensitive to model or prior specification (see \cref{perf-test:sec:simulation-study}).
+Flat priors can be unintentionally very informative, and hence (weakly) informative priors should be used.
+Of particular interest is a strongly informative prior to combine the information in the CIS with the results of \cref{E-ATACCC} (see \cref{perf-test:sec:parameters-priors}).
+Through a simulation study, I show that these methods perform well, and in this context are not sensitive to model or whether the prior is weakly or strongly informative (see \cref{perf-test:sec:simulation-study}).
 However, application in reality must account for false negatives, the subject of \cref{E-imperf-test}.
@@
 The CIS infrequently tests each individual, with tests occurring either weekly or monthly (see\todo{refer to section with CIS intro}).
-The infrequent testing complicates analysis of the data in two ways: infection episodes can be undetected and the infection episode endpoints (start and end times) are known imprecisely.
-This chapter develops a method to account for the infrequent testing.
-This section describes my approach and its background within the relevant statistical literature.
+The infrequent testing complicates analysis of the data in two ways: infection episodes can go undetected and the infection episode endpoints (start and end times) are known imprecisely.
+This chapter develops a method to account for the infrequent testing pattern.
 
-I analyse the data using survival analysis.
-Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
+I approach the problem through survival analysis, the area of statistics concerned with estimating the distribution of the times between two events.
 The event at the start of the time of interest is the \emph{initiating event}, and the event at the end of the time of interest is the \emph{terminating event}.
@@
 In the context of this thesis, the distribution to estimate is the duration of positivity.
-Therefore, the initiating event is when the duration starts, \ie becoming detectable; and the terminating event is when the duration ends, \ie becoming no longer detectable.
+Therefore, the initiating event is when the duration starts; and the terminating event is when the duration ends.
 
@@
 In the biomarker approach, computational efficiency is an issue because the number of random effects scales with the number of detected infection episodes.
-There are a large number of detected episodes in CIS 
+There are a large number of detected episodes in CIS; in the period I will consider, around 50 times more than in ATACCC.
 Therefore, the biomarker approach would be computationally challenging with the large sample size in CIS.
 The computational challenge is compounded by the data being stored in a trusted research environment which limits the computational power and tools available.
-The data limitation is due to the infrequent sampling in the CIS study design, meaning that most episodes have only a single positive test.
-With only a single positive test, there is no information on the viral load trajectory, and hence the random effects are unlikely to be well-identified.
-While the biomarker approach does use more information per sample by taking into account the viral load as well as the binary test result (\ie whether the result is positive or negative), however, this is unlikely to be sufficient to overcome these problems.
+Due to the infrequent sampling in the CIS study design, most detected episodes have only a single positive test, and consequently,
+there is no information on the viral load trajectory, and hence the random effects are unlikely to be well-identified.
+While the biomarker approach does use more information per sample by taking into account the viral load as well as the binary test result (\ie whether the result is positive or negative), however, the available information is unlikely to be sufficient to overcome these problems.
 
-The first complication in the analysis is that infections episodes can be undetected, meaning no there are no positive tests associated with them.
+The first complication in the analysis is that infection episodes can be undetected, meaning no there are no positive tests associated with them.
 This happens when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
 For these undetected episodes, there is no information on them meaning that we do not even know how many of them occurred.
-Undetected episodes are important to account for because they are systematically shorter than detected episodes.
+It is important to account for undetected episodes because they are systematically of shorter duration than detected episodes.
 To see why this is the case, consider the undetected infection in the bottom of \cref{perf-test:fig:truncation}.
@@
 If the duration was 24 days (or longer), the individual would still have been detectable at their next test, and therefore would have been detected.
-Undetected infections are a large problem in CIS because individuals are tested rarely compared to the duration of an infection episode, particularly when tested only every 28 days.
+Undetected infections are a large problem in CIS because the interval between tests is often 28 days (see\todo{ref section on the survey design}), large relative to the duration of an infection episode.
 However, we do still know the testing schedule of the individual as they are in the CIS cohort.
-Undetected events where we have some data on the events in which these events occur is an unusual situtation.
+Undetected events where we have information on missed events is unusual within the survival analysis literature.
 Far more common is that there is no data on these events, a situation known as \emph{truncation}.
@@
 An event, such as an endpoint, is known as \emph{censored} if the time of the event can only be bounded.
-Censored data is present in almost all survival analyses.
+Censored data are present in almost all survival analyses.
 An event is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
@@
 Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa (see \cref{perf-test:fig:double-interval-censor}).
-If both the initiaing and terminating events are censored, then the event is \emph{doubly censored}.
+If both the initiating and terminating events are censored, then the event is \emph{doubly censored}.
 Doubly censored data is more challenging to handle because the initiating event time and duration must be jointly modelled~\autocite[and references therein]{liSemiparametric}.
@@
 This is especially true within the human biostatistical literature.
-While this literature does contain theoretical frameworks which include double censored and truncated case~\autocites{turnbullEmpirical}{dempsterMaximum}, they were normally applied when the terminating event was either uncensored or right censored~\autocite{sunEmpirical,bacchettiNonparametric}.
+While this literature does contain theoretical frameworks which include the double censored and truncated case~\autocites{turnbullEmpirical}{dempsterMaximum}, they have been applied when the terminating event was either uncensored or right censored~\autocite{sunEmpirical,bacchettiNonparametric}.
 The approach of these studies (and elsewhere~\autocite{shenNonparametric}) was to take a conditional likelihood approach, conditioning on both the truncation (a selection effect) and the period in which the initiating event is known to occur.
 They justify conditioning on the time of the initiating event by arguing that this information provides little information on the distribution being estimated~\citePersonalComms{Nick Jewell}.
-The period in which the infection is detected in CIS is likely to be in the period where there is frequent testing.
-Therefore, conditioning on this period overstates the probability of the infection being detected.
+However, this is not the case here.
+In CIS, infection episodes are more likely to be detected when an individual is in the weekly testing phase.
+This phase is not representative of the overall study: infections are more likely to be detected during this phase.
+Therefore, estimating the probability of being detected conditional on being infected in this period (which would be implictly done in this previous method)  will overestimate the probability of being detected.
 
-Double interval censored and truncated data frequently occur when estimating the survival time of bird nests~\autocite{heiseyABCs}.
+Censored and truncated data frequently occur when estimating the survival time of bird nests, with the censoring taking a variety of forms~\autocite{heiseyABCs}.
 In this context, \textcite{heiseyModelling} develop a general framework allowing for arbitrary censoring patterns and double interval censoring.
-In bird nesting studies, the visit times are normally global (all nests could be detected on the same days), while in CIS each individual has their own visit schedule; with global visit times the probability of being detected is the same for all participants in the study, simplifying the likelihood.
+In their application, the visit times are global (all nests could be detected on the same days), while in CIS each individual has their own visit schedule; with global visit times the probability of being detected is the same for all participants in the study, simplifying the likelihood.
 I use this framework as the basis of the model I derive in \cref{perf-test:sec:model}.
 
-I perform inference within the Bayesian paradigm for two reasons.
+Inference will be performed within the Bayesian paradigm for two reasons.
 First, the standard methods for estimating the parameter's variance and covariance are poor.
-Commonly used methods are based on the Fisher information matrix or resampling, but neither have theoretical justification~\autocite[230]{sunAnalysis}.
+Commonly used methods are based on the Fisher information matrix or resampling, but neither have theoretical justification~\autocite[230]{sunAnalysis}\todo{Try and understand why this is the case}.
 Second, a prior naturally allows the inclusion information from previous studies, such as that from \cref{E-ATACCC}, and/or a belief that the hazard changes smoothly (see \cref{perf-test:sec:parameters-priors}).
-These assumptions may be needed to combat the lack of information in the data due to the undetected episodes~\autocite{caoBias}.
+These prior structures may be needed to combat the lack of information in the data due to the undetected episodes~\autocite{caoBias}.
 They can also be viewed as combining the information from the two analyses (see \cref{perf-test:sec:informative-priors}).
 
-Bayesian methods have previously been applied to nest survival~\autocites{heBayesiana}{heBayesian}{caoModeling}.
-These methods augment the data with the unobserved times of infection, creating a parameter per detected infection.
-This would be many parameters for CIS (which has over 4000 in the time period considered in the main analysis of \cref{E-imperf-test}) and therefore computationally expensive.
+Bayesian methods have previously been applied to nest survival settings~\autocites{heBayesiana}{heBayesian}{caoModeling}, augmenting the data with the unobserved times of infection.
+The augmentation creates a parameter per detected infection.
+Therefore, data augmentation applied to the 4000 detected episodes in the main analysis of \cref{E-imperf-test} greatly increases the dimensionality of the problem, making it computationally infeasible.
 I develop a Bayesian framework where there is no explicit data augmentation, allowing the method to scale to the large number of episodes present in CIS.
@@
 
-In this section, I develop a model accounting for double interval censoring and truncation.
-I model only individuals with detected episodes.
-This approach is based on long-standing methodology~\autocites{heiseyModelling}{dempsterMaximum}{turnbullEmpirical}.
+In this section, I develop a model accounting for double interval censoring and truncation, modelling only individuals with detected episodes.
+This approach is based on long-standing methodology~\autocites{heiseyModelling}{dempsterMaximum}{turnbullEmpirical}; in particular, I adopt the framework of \textcite{heiseyModelling}.
 The conceptual approach here is to form a cohort of individuals for which we observe at least one positive test, and then consider a study which had only enrolled these individuals.
-The modelling approach then corrects for this selection bias, by considering \emph{ghost} infections that occur in identical individuals but were truncated.
+Previous studies consider the case where nothing was observed regarding individuals without a positive test (left truncation).
+I do not make use of this additional information here because it makes little difference but increases the computational requirements and strength of assumptions regarding reinfections/immunity required (see \cref{E-appendix:total-model} for details).
+The modelling approach then corrects for this selection bias, by considering \emph{ghost} infections that occur in identical individuals but were undetected.
 
-I have also considered a model including individuals without any observed episodes (see \cref{E-appendix:total-model} for details).
-These individuals may have truncated episodes or may not have been infected at all.
-However, I found that this model has little difference in results while increasing computational complexity and requiring assumptions on reinfection (discussed in \cref{perf-test:sec:discussion}).
+Ghost infections are a mathematical convenience to correct for the undetected episodes which arise through the following thought experiment.
+For each detected episode $i$, let $n_i-1$ be the number of undetected episodes, such that $n_i$ is the total number of episodes.
+Further, assume that each of these $n_i$ infections episodes were drawn independently from $i$'s distribution of episode beginning and end times and all other characteristics of these infection episodes are identical; most notably, the testing schedule and, if the model were to be extended, other covariates of the individual in which the episode occurs.
+If we had detected all $n_i$ episodes, then our sample of detected episodes would be unbiased; however, we did not, and the detected episode is systematically longer than the other $n_i-1$ episodes (as explained in \cref{perf-test:sec:problem}).
 
-Ghost infections are a mathematical convenience to correct for truncated episodes.
-Consider that we draw many episodes independently, from a distribution of infection times and durations with all other characteristics identical (most notably the testing schedule and, if the model were to be extended, other covariates of the individual in which the episode occurs).
-This probability space is equivalent to any two random variables which can be over any two of infection time, recovery time, and duration.
-Conceptually, we will condition on the total number of episodes drawn (which is unknown) corresponding to the detected episode $i$, $n_i$, then consider the probability of observing exactly one episode and this episode occurring in the interval censored ranges observed (being \emph{admissible}), and then finally marginalise over $n_i$ to form the likelihood contribution for this episode.
+In the following method, I consider $n_i$ and the duration of the missing episodes as missing data.
+With the full dataset of $n_i$ infection episodes, the problem reduces to double interval censored survival analysis, which has a known likelihood.
+Therefore, I will consider the posterior augmented with all the $n_i$s and then integrate over the $n_i$s to form the desired posterior.
 
-Unfortunately, there is no clean interpretation of the number of ghost episodes in terms of real-world quantities; in this sense, it is a pure model parameter.
-In reality, there is only one individual who is identical to the individual in which the episode occurs in all relevant ways.
-Even within the simulation study (see \cref{perf-test:sec:simulation-study}) these are not infections that are simulated, and understanding the ground truth number of ghost infections is not straightforward.
+\todo[inline]{Maybe move the following paragraph elsewhere}
+It is important to note that there is not a correspondence between the number of ghost episodes and the number of infection episodes that were missed within the true cohort.
+Within the true cohort, the episodes in which the undetected episodes occurred are different to those in which detected episodes occurred.
+In particular, the individuals in which undetected episodes occurred would (on average) have longer spacing between their tests.
 
-To formalise this, within the framework of \textcite{heiseyModelling}, note that all relevant information about episode $i$ can be fully characterised by the (unobserved) pair $(b_i, e_i)$ of start and end dates of the episode, which belong to the state space $T \times T$.
+The framework of \textcite{heiseyModelling} formalises this thought experiment.
+Start by noting that all relevant information about episode $i$ can be fully characterised by the (unobserved) pair $(b_i, e_i)$ of start and end dates of the episode, which belong to the state space $T \times T$.
 If the individual in which the episode occurs is tested between $b_i$ and $e_i$ (inclusive), then the episode is detected, otherwise it is a ghost episode.
@@
 To relate this to the data, define three sets which partition the space of possible infection and recovery times (\ie: are strict subsets of $T \times T$), based on the observations associated with episode $i$ (graphically shown in Figure \ref{perf-test:fig:partitionSpace}).
-Each episode, whether observed or not, must fall into one of these three classes with a probability that is a function of $\theta$.
+Each episode, whether detected or not, must fall into one of these three classes with a probability that is a function of $\theta$.
 
@@
   Admissible episodes, $\alpha_i$, which have an infection and
-  recovery time within their respective intervals observed.
+  recovery time consistent with the data observed. That is, $b_i \in [l_i^{(b)}, r_i^{(b)}]$, and $e_i \in [l_i^{(e)}, r_i^{(e)}]$.
   $n_{ia} =1$ of these occurred, each with probability
@@
 \item
-  Inadmissible episodes, $\beta_i$, which do not have an infection and
-  recovery time within their respective intervals but would have been
+  Inadmissible episodes, $\beta_i$, which are not consistent with the data observed but would have been
   observed (not truncated). This is all remaining episodes not in the
@@
 
-Episode $i$ and its ghosts independently occur belong to one of these
+Episode $i$ and its ghosts independently belong to one of these
 classes. Therefore, conditional on $n_i$, the number of episodes in
@@
 The posterior we are interested in is
-$p(\theta \mid n_{ia} = 1, n_{iu} = 0, \dots, n_{n_a,a} = 1, n_{n_a,u} = 0)$,
+$p(\theta \mid n_{1a} = 1, n_{1u} = 0, \dots, n_{n_a,a} = 1, n_{n_a,u} = 0)$,
 where $\theta$ are the parameters governing the survival distribution
@@
 &= p(\theta) \prod_{i=1}^{n_a} \sum_{n_i=1}^\infty n_i p_{ia} p_{it}^{n_i- 1} p(n_i) \\
-&= p(\theta) \prod_{i=1}^{n_a} p_{ia} \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i)
+&= p(\theta) \prod_{i=1}^{n_a} p_{ia} \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i).\label{eqn:sum}
 \end{align}
-where the last line follows by assuming prior independence.
 
-In the remainder of the section, we derive the expressions for $p_{ia}$ and
-$p_{it}$ in terms of the data, and analytical solutions for the sum
+In the remainder of the section, expressions for $p_{ia}$,
+$p_{it}$, and analytical solutions for the sum in \cref{eqn:sum} are derived in terms of the observed data, 
 under different priors.
@@
 The prior $p(n_i) \propto 1/n_i$ means that the model coincides with the frequentist conditional likelihood approach, up to the priors on other parameters~\cites[section 4.2]{dempsterMaximum}{heiseyModelling}[section 8.7.5]{gelmanBayesian}.
-Furthermore, this is the reference prior for this quantity\autocite{heBayesiana}.
+Furthermore, this is the reference prior for this quantity \autocite{heBayesiana}.
 Under this prior,
@@
 sample. Furthermore, each $n_i$ is a discrete parameter, and hence
-cannot be sampled within the current Stan implementation.
+cannot be sampled with widely-used inference algorithms such as Stan's Hamiltonian Monte Carlo (HMC)\todo{ref}.
 
@@
 \end{align}
-which is a negative binomial pmf with size parameter $r+1$ and
+Therefore, by comparing this expression to the pmf of a negative binomial, we find that $n_i - 1$ is distributed negative binomial with size parameter $r+1$ and
 probability parameter $\frac{r + \mu (1 - p_{it})}{r+\mu}$. The mean
-of this is
+of this distribution is
 \begin{align}

### 2023-08-10 cis-perfect-testing.tex
Finish responding to Paul's comments

@@
 The prior $p(n_i) \propto 1/n_i$ means that the model coincides with the frequentist conditional likelihood approach, up to the priors on other parameters~\cites[section 4.2]{dempsterMaximum}{heiseyModelling}[section 8.7.5]{gelmanBayesian}.
-Furthermore, this is the reference prior for this quantity \autocite{heBayesiana}.
+Furthermore, this is the reference prior for this quantity~\autocite{heBayesiana}.
 Under this prior,
@@
 \begin{align}
-N_i \mid \lambda &\dist \Poi(\lambda) \\
-\lambda &\dist \GamDist(a, b)
+N_i \mid \lambda_i &\dist \Poi(\lambda_i) \\
+\lambda_i &\dist \GamDist(a, b)
 \end{align}
@@
 \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) 
-&= \int \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i \mid \lambda) p(\lambda) d\lambda \\
-&= \int \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} \frac{\lambda^{n_i} e^{-\lambda}}{n_i!} p(\lambda) d\lambda \\
-&= \int e^{-\lambda} \sum_{n_i=1}^\infty p_{it}^{n_i- 1} \frac{\lambda^{n_i-1}\lambda }{(n_i-1)!} p(\lambda) d\lambda \\
-&= \int e^{-\lambda} \lambda p(\lambda) \sum_{n_t=0}^\infty p_{it}^{n_t} \frac{\lambda^{n_t} }{n_t!} d\lambda &n_t = n_i - 1 \\
-&= \int e^{-\lambda} \lambda p(\lambda) e^{p_{it}\lambda} d\lambda &\text{Poisson pmf} \\
-&= \int e^{\lambda (p_{it} - 1)} \lambda p(\lambda) d\lambda \\
-&= \int e^{\lambda (p_{it} - 1)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda \\
+&= \int \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i \mid \lambda_i) p(\lambda_i) d\lambda_i \\
+&= \int \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} \frac{\lambda_i^{n_i} e^{-\lambda_i}}{n_i!} p(\lambda_i) d\lambda_i \\
+&= \int e^{-\lambda_i} \sum_{n_i=1}^\infty p_{it}^{n_i- 1} \frac{\lambda_i^{n_i-1}\lambda_i }{(n_i-1)!} p(\lambda_i) d\lambda_i \\
+&= \int e^{-\lambda_i} \lambda_i p(\lambda_i) \sum_{n_t=0}^\infty p_{it}^{n_t} \frac{\lambda_i^{n_t} }{n_t!} d\lambda_i &n_t = n_i - 1 \\
+&= \int e^{-\lambda_i} \lambda_i p(\lambda_i) e^{p_{it}\lambda_i} d\lambda_i &\text{Poisson pmf} \\
+&= \int e^{\lambda_i (p_{it} - 1)} \lambda_i p(\lambda_i) d\lambda_i \\
+&= \int e^{\lambda_i (p_{it} - 1)} \frac{b^a}{\Gamma(a)} \lambda_i^{a-1} e^{-b\lambda_i} \lambda_i d\lambda_i \\
 &= \frac{\Gamma(a+1)b^a}{\Gamma(a) (b+1-p_{it})^{a+1}} \\ 
   &\; \times \int \frac{(b+1-p_{it})^{a+1}}{\Gamma(a+1)} \\
-  &\; \lambda^{(a+1)-1} e^{-(b+1-p_{it})\lambda} d\lambda \\
+  &\; \lambda_i^{(a+1)-1} e^{-(b+1-p_{it})\lambda_i} d\lambda_i \\
 &= \frac{\Gamma(a+1)b^a}{\Gamma(a) (b+1-p_{it})^{a+1}} &\text{as gamma pdf} \\
@@
 The above allows us to sample from the posterior without needing to
-sample the $n_i$s. This is convenient because there are as many
-$n_i$s as episodes observed and hence is computationally costly to
-sample. Furthermore, each $n_i$ is a discrete parameter, and hence
+sample the $n_i$s, which is much more efficient due to the previously-discussed dimensionality issue.
+Furthermore, each $n_i$ is a discrete parameter, and hence
 cannot be sampled with widely-used inference algorithms such as Stan's Hamiltonian Monte Carlo (HMC)\todo{ref}.
 
-Having estimated the posterior of the other parameters, we might want to
-consider the posterior of $n_i$ (normally for diagnostic or model
-debugging purposes), which we can sample from the full conditional.
+The posterior of $n_i$, while inconvenient to sample from and not of epidemiological interest directly, can be useful for diagnostic or model debugging purposes (for example, see\todo{ref where I use this in the next chapter}).
+This posterior can be found by sampling from the full conditional.
 Specifically, for each posterior sample of $\theta$, we sample one

### 2023-08-11 cis-perfect-testing.tex
Restructure first section of perfect testing chapter

@@
 
-\section{Problem description and approach} \label{perf-test:sec:problem}
+\section{Problem description} \label{perf-test:sec:problem}
 
@@
 The infrequent testing complicates analysis of the data in two ways: infection episodes can go undetected and the infection episode endpoints (start and end times) are known imprecisely.
-This chapter develops a method to account for the infrequent testing pattern.
-
-I approach the problem through survival analysis, the area of statistics concerned with estimating the distribution of the times between two events.
-The event at the start of the time of interest is the \emph{initiating event}, and the event at the end of the time of interest is the \emph{terminating event}.
-It is broadly applicable to many domains, such as the time-to-failure for a mechanical system or the effect of a treatment on time spent in hospital.
-In the context of this thesis, the distribution to estimate is the duration of positivity.
-Therefore, the initiating event is when the duration starts; and the terminating event is when the duration ends.
-
-\todo[inline]{I think the following paragraph is needed somewhere, to answer why I'm not doing the same thing as in ATACCC, but it feels like an aside here since the previous paragraph and the one after this one are more linked.}
-Here, I use survival analysis rather than the biomarker approach taken in \autoref{E-ATACCC} for computational efficiency and data limitations.
-In the biomarker approach, computational efficiency is an issue because the number of random effects scales with the number of detected infection episodes.
-There are a large number of detected episodes in CIS; in the period I will consider, around 50 times more than in ATACCC.
-Therefore, the biomarker approach would be computationally challenging with the large sample size in CIS.
-The computational challenge is compounded by the data being stored in a trusted research environment which limits the computational power and tools available.
-Due to the infrequent sampling in the CIS study design, most detected episodes have only a single positive test, and consequently,
-there is no information on the viral load trajectory, and hence the random effects are unlikely to be well-identified.
-While the biomarker approach does use more information per sample by taking into account the viral load as well as the binary test result (\ie whether the result is positive or negative), however, the available information is unlikely to be sufficient to overcome these problems.
+This section explains these issues fully, and the remainder of the chapter develops a method to model them.
 
@@
 This happens when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
-For these undetected episodes, there is no information on them meaning that we do not even know how many of them occurred.
+For these undetected episodes, there is no information on them episode, meaning that we do not even know how many of them occurred.
 It is important to account for undetected episodes because they are systematically of shorter duration than detected episodes.
@@
 Undetected infections are a large problem in CIS because the interval between tests is often 28 days (see\todo{ref section on the survey design}), large relative to the duration of an infection episode.
-However, we do still know the testing schedule of the individual as they are in the CIS cohort.
-Undetected events where we have information on missed events is unusual within the survival analysis literature.
-Far more common is that there is no data on these events, a situation known as \emph{truncation}.
-A situation that is particularly relevant is \emph{left truncation}, when only events lasting a minimum amount of time are observed.
 \begin{figure}
@@
 
-The second complication is that the endpoints of detected infection episodes are known imprecisely.
-An event, such as an endpoint, is known as \emph{censored} if the time of the event can only be bounded.
-Censored data are present in almost all survival analyses.
+The second complication is that the beginning and end times of detected infection episodes are known imprecisely.
+The imprecision in the beginning of the infection episode arises because the only observation that an episode has started is that an individual was negative at one test, on day $l_i^{(b)}-1$ (the minus one is because the first day that the infection could have begun is the day after the negative test), and positive at the next test, on day $r_i^{(b)}$.
+Similarly, the end of the infection episode is bounded by the final positive test on day $l_i^{(e)}$ and the following negative test on day $r_i^{(e)}+1$ (the plus one is because the last day that the infection could have ended is the day before the negative test).
+\Cref{perf-test:fig:double-interval-censor} shows this issue graphically.
+\begin{figure}
+  \centering \includegraphics{cis-perfect-testing/double-interval-censor}
+  \caption[Double-interval censoring in CIS data]{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval.
+Demonstrated here by a participant who is recorded negative at time 7 to positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode.
+\label{perf-test:fig:double-interval-censor}}
+\end{figure}
+
+I will base my analysis on the 4,800\todo{check number} of episodes that were first detected between 10th Oct 2020 and 6th Dec 2020 inclusive and have negatives bounding the start and end time of the episode.
+That is, the individual in which the infection episode occurred tested negative at least once in the survey prior to the start of their infection episode, and they tested negative at least once after their infection episode ended.
+Without a negative before the episode there is little information on the episode's length because it's start time could have occurred arbitrarily long ago.
+The only episodes without a negative following the end of the episode are those in individuals lost to follow-up before their episode ended, which is extremely rare.
+This filtering could bias the results of the study but will be considered in the same way as undetected episodes to avoid this.
+
+In the remainder of this and the following chapter, I use survival analysis rather than the biomarker approach taken in \cref{E-ATACCC} for computational efficiency and data limitations arising from the study design.
+In the biomarker approach, computational efficiency is an issue because the number of random effects scales with the number of detected infection episodes.
+There are a large number of detected episodes in CIS; in the period I will consider, around 50 times more than in ATACCC.
+Therefore, the biomarker approach would be computationally challenging with the large sample size in CIS.
+The computational challenge is compounded by the data being stored in a TRE with the associated limited access to computational power and software.
+The study design of CIS, with the infrequent sampling in the CIS study design, means that most detected episodes have only a single positive test.
+Hence, there is no information on the viral load trajectory required to identify the random effects in the model used in \cref{E-ATACCC} or similar approaches.
+While the biomarker approach does use more information per sample by taking into account the viral load as well as the binary test result (\ie whether the result is positive or negative), however, the available information is unlikely to be sufficient to overcome these problems.
+
+\section{Survival analysis}
+
+In survival analysis, the event at the start of the time of interest is the \emph{initiating event}, and the event at the end of the time of interest is the \emph{terminating event}.
+It is broadly applicable to many domains, such as the time-to-failure for a mechanical system or the effect of a treatment on time spent in hospital.
+In the context of this thesis, the distribution to estimate is the duration of positivity.
+Therefore, the initiating event is when the duration starts; and the terminating event is when the duration ends.
+
+Undetected events where we have information on missed events is unusual within the survival analysis literature.
+Far more common is that there is no data on these events, a situation known as \emph{truncation}.
+A situation that is particularly relevant is \emph{left truncation}, when only events lasting a minimum amount of time are observed.
+A subtlety that makes the CIS different from standard left truncation is that information on individuals in which undetected infections occurred is available because they were enrolled in the study.
+This information could be incorporated into the analysis, although doing so increases the computational burden and requires strong assumptions regarding reinfections without noticeably affecting the results (see \cref{appendix:total-model}).
+In standard left truncation, nothing is known about the missed events.
+As the information is not being incorporated here, standard left truncation methods remain applicable, and they can be modified to incorporate the additional information if desired.
+
+When the time of the initiating and/or terminating events are not known exactly, the survival analysis literature refers to the event(s) as \emph{censored}.
+Censored events are present in almost all survival analyses.
 An event is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
@@
 Therefore, the CIS data is \emph{double interval censored}; approaches for double interval censoring are reviewed by \textcite{sunAnalysis,bogaertsSurvival}.
-\begin{figure}
-  \centering \includegraphics{cis-perfect-testing/double-interval-censor}
-  \caption[Double-interval censoring in CIS data]{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval.
-Demonstrated here by a participant who is recorded negative at time 7 to positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode.
-\label{perf-test:fig:double-interval-censor}}
-\end{figure}
 
@@
 While this literature does contain theoretical frameworks which include the double censored and truncated case~\autocites{turnbullEmpirical}{dempsterMaximum}, they have been applied when the terminating event was either uncensored or right censored~\autocite{sunEmpirical,bacchettiNonparametric}.
-The approach of these studies (and elsewhere~\autocite{shenNonparametric}) was to take a conditional likelihood approach, conditioning on both the truncation (a selection effect) and the period in which the initiating event is known to occur.
+The approach of these studies \autocite[and elsewhere, e.g.:][]{shenNonparametric} was to take a conditional likelihood approach, conditioning on both the truncation (a selection effect) and the period in which the initiating event is known to occur.
 They justify conditioning on the time of the initiating event by arguing that this information provides little information on the distribution being estimated~\citePersonalComms{Nick Jewell}.
@@
 Inference will be performed within the Bayesian paradigm for two reasons.
-First, the standard methods for estimating the parameter's variance and covariance are poor.
+First, the standard frequentist methods for estimating the parameter's variance and covariance are poorly developed in this area.
 Commonly used methods are based on the Fisher information matrix or resampling, but neither have theoretical justification~\autocite[230]{sunAnalysis}\todo{Try and understand why this is the case}.
-Second, a prior naturally allows the inclusion information from previous studies, such as that from \cref{E-ATACCC}, and/or a belief that the hazard changes smoothly (see \cref{perf-test:sec:parameters-priors}).
+Second, a prior naturally allows the inclusion information from previous studies, such as that from \cref{E-ATACCC}, and/or a belief that the hazard changes smoothly in a way that can be encoded in a prior (these options are explored in \cref{perf-test:sec:parameters-priors}).
 These prior structures may be needed to combat the lack of information in the data due to the undetected episodes~\autocite{caoBias}.
@@
 For the $i$th detected episode, assume that there are $n_{it}$ ghost episodes.
-Further, we discard episodes if there is no negative test prior to their start, and hence also view any such episodes as ghost episodes.
 
@@
 
-The prior $p(n_i) \propto 1/n_i$ means that the model coincides with the frequentist conditional likelihood approach, up to the priors on other parameters~\cites[section 4.2]{dempsterMaximum}{heiseyModelling}[section 8.7.5]{gelmanBayesian}.
-Furthermore, this is the reference prior for this quantity~\autocite{heBayesiana}.
+The reference prior for $n_i$ is $p(n_i) \propto 1/n_i$~\autocite{heBayesiana}.
 Under this prior,
 $\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) \propto \sum_{n_i=1}^\infty p_{it}^{n_i-1} = 1/(1-p_{it})$.
+This likelihood is equal to the conditional likelihood used in the frequentist approaches previously mentioned.
+It has been used elsewhere for Bayesian inference~\autocite{zhouUnified}, but the justification within a Bayesian paradigm has not been made explicitly.
+%The prior $p(n_i) \propto 1/n_i$ means that the model coincides with the frequentist conditional likelihood approach, up to the priors on other parameters~\cites[section 4.2]{dempsterMaximum}{heiseyModelling}[section 8.7.5]{gelmanBayesian}.
 % (\protect\hyperlink{ref-dempsterMaximum}{Dempster, Laird, and
@@
 The form of $S_\theta(t)$ is yet to be specified.
-As frequently observed in the literature~\autoref[e.g.:][]{heBayesian}, it is more convenient to operate with the hazard than the survival or probability mass function directly.
+As frequently observed in the literature~\autocite[e.g.:][]{heBayesian}, it is more convenient to operate with the hazard than the survival or probability mass function directly.
 This is because the mass function must sum to one and the survival must be monotonically decreasing, while there are no constraints across multiple hazards.
@@
 
-I simulate each individual enrolled in CIS who was tested at least once between 10th Oct 2020 and 6th Dec 2020 (inclusive, the same period used for the later analysis using real CIS data).
-The individuals' testing schedules are exactly as observed in CIS.
-The infection times and durations of positivity are simulated.
-From this, a set of test results can be generated.
+I simulate a dataset of detected episodes that has the same characteristics as that in the CIS by the following procedure.
+\begin{enumerate}
+    \item Extract the test schedules for each individual who had at least one test during the period of interest.
+    \item Draw an episode start time, $b_i$ for each individual uniformly at random between 2nd July 2020 (100 days before the period where a detected episode would be included) and 6th December 2020 (the end of this period).
+    \item Draw a duration of episode for each individual, $d_i$, based on the distribution described below. Then calculate the end of their infection episode, $e_i = b_i + d_i - 1$.
+    \item Simulate the test results based on the test schedule, $b_i$, and $e_i$. Tests between $b_i$ and $e_i$ (inclusive) are positive, all other tests are negative.
+    \item Discard episodes where there are no positive tests (\ie undetected episodes) and then apply the same inclusions/exclusions as in \cref{perf-test:sec:problem}.
+    \item Of these remaining episodes, sample 4,800 to match the sample size of the true dataset.
+This is needed because in step 2 the entire cohort was infected, while in the real study only a (unknown) portion is infected.
+    \item For this final set of episodes, calculate $(l_i^{(b)}, r_i^{b}, l_i^{(e)}, r_i^{(e)})$ by taking the day after the last negative prior to any positives, the first positive, the last positive, and the day before the negative following the last positive respectively.
+\end{enumerate}
 
@@
 \end{figure}
-
-Individual $i$ is first detectable on their day of infection, $B_i$.
-$B_i$ is drawn uniformly between 100 days prior to the first day of testing included and the final day of testing included (inclusive).
-$i$ is detectable up to and including the last day of infection, $E_i = B_i + D_i - 1$, where $D_i$ is the duration of positivity for individual $i$, drawn from $f_S$.
-Therefore, tests between $B_i$ and $E_i$ (inclusive) are positive, and all other tests are negative.
-
-I then replicate the process used when working with the CIS data by filtering the episodes.
-First, discard any episodes where there are no positive tests as they are never detected.
-Next, discard any episodes where there are no negative tests prior to the first positive; since there is no lower bound on their infection time, they contain little information, this selection effect was included when constructing $p_{it}$.
-Finally, of those remaining episodes, select a random subset to match the number of episodes in the simulation to the number of episodes in the modelling cohort.
-The last step is required because everyone is infected in the simulation which does not happen in reality.
-
-For each preserved infection episode, calculate $(l_i^{(b)}, r_i^{b}, l_i^{(e)}, r_i^{(e)})$ by taking the day after the last negative prior to any positives, the first positive, the last positive, and the day before the negative following the last positive respectively.
 

### 2023-08-11 cis-perfect-testing.tex
Use (un)detected rather than (un)truncated

@@
   \centering \includegraphics{cis-perfect-testing/truncation}
-  \caption[Truncation in CIS data]{Episodes data in CIS is truncated, meaning that an unknown number of infections are never detected.
+  \caption[Undetected episodes in CIS data]{An unknown number of infection episodes in CIS are undetected.
 Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was.
@@
 
-In this section, I develop a model accounting for double interval censoring and truncation, modelling only individuals with detected episodes.
+In this section, I develop a model accounting for double interval censoring and undetected episodes, modelling only individuals with detected episodes.
 This approach is based on long-standing methodology~\autocites{heiseyModelling}{dempsterMaximum}{turnbullEmpirical}; in particular, I adopt the framework of \textcite{heiseyModelling}.
@@
 \item
-  Truncated episodes, $\Omega_i^C$, which have an infection and
+  Undetected (or, for consistency with the literature, truncated) episodes, $\Omega_i^C$, which have an infection and
   recovery time such that they would not have tested positive. An
@@
   Inadmissible episodes, $\beta_i$, which are not consistent with the data observed but would have been
-  observed (not truncated). This is all remaining episodes not in the
+  detected. This is all remaining episodes not in the
   previous sets. $n_{iu} = 0$ of these occurred, where they would have
@@
 
-The untruncated region (where we could have observed an episode) is
+The detected region (where we could have observed an episode) is
 $\Omega_i = \alpha_i \cup \beta_i$ (the notation here is used for
@@
 \includegraphics[width=\textwidth]{cis-perfect-testing/regions_diag}
-\caption[Admissible, inadmissible, and untruncated infections]{The regions (as defined in the main text) for an episode $i$ in an
+\caption[Admissible, inadmissible, and undetected infections]{The regions (as defined in the main text) for an episode $i$ in an
 individual whose first test was at time 0 and negative, with subsequent
@@
 Next, we derive $p_{it} = \prob((b, e) \in \Omega^C_i)$, the probability
-that episode $i$ was truncated. A truncated episode means that no test
+that episode $i$ was detected. A detected episode means that no test
 was performed during the episode or there was no negative test prior to
@@
 A flexible, spline-based form is assumed for the baseline survival function~\autocite{roystonSTPM,roystonFlexible} with covariates introduced via proportional odds.
-By not accounting for either the truncation or the interval censoring of the initiating event, this analysis has competing biases which makes them hard to interpret~\autocite{cisMethodsONS}.
+By not accounting for either the undetected infections or the interval censoring of the initiating event, this analysis has competing biases which makes them hard to interpret~\autocite{cisMethodsONS}.
 

### 2023-08-18 incidence-prevalence.tex
Start writing incidence/prevalence chapter

--- 2023-07-18 incidence-prevalence.tex
@@
 \begin{document}
-\chapter{Incidence and prevalence} \label{incidence-prevalence}
+\chapter{Statistical background} \label{incidence-prevalence}
+%\todo[inline]{This chapter is pretty short. Not sure if this matters or it needs to merged with another chapter? For example, one background chapter split into epi/biology and stats.}
+
+This chapter explains incidence and prevalence, and the relationship between these quantities.
+These quantities have been of interest as long as disease has been studied, and estimating the prevalence from incidence is the fundamental motivation for this thesis.
+Their exact definition can vary throughout the literature, here I follow the definitions of the seminal paper \textcite{freemanPrevalence}.
+As I will explain, the duration of infection is the quantity that relates them.
+
+Timescale underlies all these quantities.
+Throughout this thesis, time will generally\todo{revisit phrasing once finished - do I actually use continuous time anywhere (probably in SEIR model)} be considered discrete in units of days.
+Most epidemiological data is only available at daily granularity or, when more granular data exists, is often unreliable.
+Furthermore, within-day patterns produce variations in transmission (\eg due to sleep) and data (\eg due to logistical considerations around collection); these variations are not of interest in most contexts.
+
+\section{Incidence, prevalence, and duration}
+
+\emph{Incidence} at time $t$, $Z_t$, is the number of people in a given population who become infected at time $t$ (I use $Z_t$ rather than $I_t$ to avoid confusion with the convention that $i$ is an indexing variable).
+Incidence expressed as a proportion of the population, $Z_t/N$ where $N$ is the number of people in the population, is the \emph{incidence rate}.
+As discussed in \cref{E-intro}\todo{might want to point to a specific section}, infection events are not directly observable and must be inferred from other data.
+Despite this issue, it remains the single most important quantity for informing the response to an epidemic or pandemic.
+
+\emph{Prevalence} of an infectious disease at time $t$, $P_t$, is the number of people in a given population currently infected at time $t$, assumed constant over the short period $t$.
+Each of these individuals is known as a \emph{prevalent} individual.
+The \emph{prevalence proportion} is the same quantity expressed as a proportion of the population.
+For SARS-CoV-2, there are several definitions of infected.
+An infected individual could mean: one that is infectious; one that is symptomatic; or one that, if they were tested, would return a positive PCR test if tested (discussed in\todo{ref appropriate section}).
+In this thesis, prevalence is defined as the number of people in the population who would return a positive PCR test (if everyone was tested) at time $t$.
+
+The relationship between incidence and prevalence can be derived by considering the probability that an individual is prevalent at any time following an infection.
+The period the individual is prevalent for is the \emph{infection episode}, which I will index with $i$, associated with that infection event.
+Denote the time that the infection episode begins as $B_i$ and the time it ends as $E_i \geq B_i$.
+The individual is prevalent from $B_i$ to $E_i$ inclusive.
+The duration of the infection episode, the number of days for which they are prevalent, is $D_i = E_i - B_i + 1$.
+In this thesis, it is assumed that the $D_i$ for each infection episode is independent and $\prob(D_i = d) = f_d$ for some pmf $f$ over $1, 2, \dots, d_\text{max}$.
+
+From this, we can derive the relationship between incidence and prevalence.
+Denote by $R_{t,t'}$ be the number of infection episodes that begin at time $t$ and end at time $t'$.
+For each infection beginning at $t$, the probability that it ends at $t'$ is $f_d(t' - t + 1)$.
+Since these durations are assumed independent, the number of episodes of each duration is multinomially distributed.
+Therefore:
+\begin{align}
+\begin{bmatrix}
+  R_{t,t} \\ R_{t,t+1} \\ \vdots \\ R_{t,t+d_\text{max}-1}
+\end{bmatrix} \mid Z_t
+\sim \MN \left(
+  Z_t, 
+  \begin{bmatrix}
+    f_d(1) \\ f_d(2) \\ \vdots \\ f_d(d_\text{max})
+  \end{bmatrix}
+\right).
+\end{align}
+Now, denote by $P_{t,t'}$ the number of episodes beginning at time $t$ that are prevalent at time $t'$ (where $t' \geq t$).
+This is $Z_t$ minus those that have recovered:
+\begin{align}
+    P_{t,t'} = \begin{cases}
+      0 &t' < t\\
+      Z_t - \sum_{i=0}^{t'-t-1} R_{t,t+i} &t \leq t' < t + d_\text{max}\\
+      0 &t' > t + d_\text{max}.
+  \end{cases}
+\end{align}
+Finally, the prevalence at $t$ can be expressed as $P_t = \sum_{t'=-\infty}^t P_{t',t}$.
+
+In general, all these random variables will be dependent for an infectious disease.
+This is because the probability of being infected at time $t$ depends on the number of infectious individuals at time $t$, which itself depends on the number of individuals infected before time $t$.
+However, if we condition on the vector $\vec{Z} = (Z_1, Z_2, \dots)^T$, then we will induce some independence.
+In particular, because the duration of individuals is independent, $R_{t_1,t_2}$ and $R_{t_1',t_2'}$ will be independent if $t_1 \neq t_1'$; importantly, this holds even if $t_2 = t_2'$.
+
+A quantity that will be important is $\E(P_t \mid \vec{Z})$, which can be derived from its constituent parts.
+First, the properties of a multinomial distribution give the marginal distribution:
+\begin{align}
+  \sum_{i=t}^{t'-1} R_{t,i} \mid \vec{Z} &\sim \text{Binomial}(Z_t, \sum_{i=1}^{t-t'} f_d(i)). \\
+\intertext{Therefore:}
+\E(P_{t-t'} \mid \vec{Z})
+  &= \E(Z_t - \sum_{i=t}^{t'-1} R_{t,i} \mid \vec{Z}) \\
+  &= Z_t (1 - \sum_{i=1}^{t-t'} f_d(i)).
+\intertext{Finally:}
+\E(P_t \mid \vec{Z})
+&= \E(\sum_{t'=-\infty}^t P_{t-t'} \mid \vec{Z}) \\
+&= \sum_{t'=-\infty}^t Z_t (1 - \sum_{i=1}^{t-t'} f_d(i)).
+\end{align}
+The quantity $1 - \sum_{i=1}^{t-t'} f_d(i)$ is the probability that an infection episode's duration is at least $t - t' + 1$, which is the survival function of $D$, denoted $S(t - t' + 1)$.
+This equation, normally expressed in terms of the survival function, forms the basis of discrete-time \emph{backcalculation}.
+
+% The probability of being positive can be modelled in different ways depending on what assumptions are most reasonable.
+% The most important assumption from a statistical perspective is the whether the probability of testing positive on a given day is independent conditional on the infection time.
+
+% \todo[inline]{Is the discussion of this model just a distraction?}
+% Assuming that the probability of testing positive depends only on the infection time implies no individual variation.
+% This model is formalised as $Pr(Z_i(t) = 1 \mid B_i, Z_i(1), Z_i(2) \dots) = Pr(Z_i(t) \mid B_i) = p_{t-B_i}$, that is the probability of individual $i$ testing positive $t$ days after infection depends only on the infection start time and no other quantities.
+% A slight relaxation of this model is introducing covariates, allowing some variation based on an individual's characteristics.
+
+% The model with the most dependence between times, while remaining realistic, is a multi-state model.
+% Therefore, if an individual has tested negative before the present time $t$ but after $B_i$ then they will test negative at time $t$.
+% The complexity of the transitions between positive and negative can be very complex.
+% I consider a slight extension to allow for the possibility of false negatives, meaning a negative test between $B_i$ and $E_i$, and that this probability may change.
+% This leads to the following model.
+% \begin{align}
+%   \prob(Z_i(t) = 1 \mid B_i, \leq E_i) &= \begin{cases}
+%     \psens(t - B_i) &B_i \leq t \leq E_i\\
+%     0 &\text{otherwise}
+%   \end{cases}\\
+%   \prob(E_i = t \mid B_i) = f_i(B_i)
+% \end{align}
+% Conditional on both $B_i$ and $E_i$, then $Z_i(t)$ and $Z_i(t')$ are independent for $t \neq t'$.
+% However, unlike the previous model, conditioning only on $B_i$ is not sufficient for $Z_i(t)$ and $Z_i(t')$ to be independent because each $Z_i(t)$ provides information on $E_i(t)$ (in particular, $Z_i(t) = 1$ implies $E_i > t$).
+% For SARS-CoV-2, the latter model is more appropriate, although, as we shall see, the assumption that $\psens$ is independent of $t$ is violated.
+
+\section{Prevalence data}
+
+A \emph{prevalence survey} is a survey that tests a sample of the population for the presence of an infection.
+Series of prevalence surveys are the form of measuring prevalence that this thesis is concerned with.
+This thesis aims to infer the incidence rate from prevalence surveys.
+I use Bayesian inference, therefore the data contributes through the likelihood.
+
+The likelihood contribution from the survey at a time $t$ is based on a sample of $n_t$ individuals from the population of interest returning $x_t$ positive tests.
+For now, I assume that the sample is chosen uniformly at random from the population of interest.
+Assume that the surveys are independent of each other and $\vec{Z}$, conditional on the population prevalence at that time.
+The population prevalence is a latent quantity.
+The likelihood is then $\p(\vec{x} \mid \vec{n}, \vec{Z}) = \int \p(\vec{P} \mid \vec{Z}) \prod_t \p(x_t \mid n_t, \vec{P}_t) d\vec{P}$, where $\vec{x} = (x_1, x_2, \dots)^T$ and similarly for $\vec{n}$.
+
+In settings I consider, the uncertainty arising from the distribution $\vec{P} \mid \vec{Z}$ is negligible.
+This can be seen by considering the variance in the likelihood.
+$x_t$ is binomially distributed with parameters $n_t$ and $P_t / N$, where $N$ is the size of the population.
+Therefore, by the law of total variance:
+\begin{align}
+  \var\left( x_t \mid \vec{n}, \vec{Z} \right)
+    &= \var\left[\E\left( x_t \mid \vec{n}, \vec{Z}, \vec{P} \right) \mid \vec{n}, \vec{Z} \right] + \E\left[\var\left( x_t \mid \vec{n}, \vec{I}, \vec{P} \right) \mid \vec{n}, \vec{Z} \right] \\
+    &= \var\left( n_t \frac{P_t}{N} \mid \vec{n}, \vec{Z} \right) + \E\left( n_t \frac{P_t}{N} \left(1 - \frac{P_t}{N} \right) \mid \vec{n}, \vec{Z} \right) \\
+    &= \left( \frac{n}{N} \right)^2 \var\left(P_t \mid \vec{n}, \vec{Z} \right) + \frac{n_t}{N} \E\left(P_t \mid \vec{n}, \vec{Z} \right)  - \frac{n_t}{N^2} \E\left(P_t^2 \mid \vec{n}, \vec{Z} \right).
+\end{align}
+For large populations and reasonable sample sizes, $n_t << N$ and $\E(P_t) \approx \var(P_t)$, the first term is negligible.
+Hence, assuming that $\var(\vec{P}) = \matr{0}$ is a reasonable approximation.
+This is equivalent to assuming that $\vec{P}$ is a deterministic function of $\vec{Z}$, generally $\vec{P} = \E( \vec{P} \mid \vec{Z})$.
+
+\section{Backcalculation}
+
+\emph{Backcalculation} is a procedure that combines the elements of the previous two sections to estimate incidence based on a measurement of prevalence
+
+In \emph{deterministic backcalculation}, the observed prevalence, $x_t/n_t$, is assumed to be equal to $P_t/N$.
+However, this is often a poor approximation.
+\emph{Statistical backcalculation} retains the sampling distribution of $x_t$ (or an approximation of it, such as a Poisson).
+
+Define the first day of an infection episode in individual $i$ as $B_i$, and assume that the probability of $i$ being infected multiple times within the period of interest is negligible.
+The time between being infected and first being detectable is short (see\todo{ref relevant part}), and therefore I often assume that $B_i$ is the same as the time of infection.
+I keep this assumption for the remainder of this chapter.
+
+Backcalculation makes use of this relationship to estimate incidence from prevalence, assuming that the duration is known.
+
+Deterministic backcalculation assumes that the variance in the population is negligible, and therefore that the prevalence is equal to the mean prevalence.
+This is justified because the variance due to sampling is much larger than the variance due to the true prevalence.
+
+\section{Estimating the duration}
+\todo[inline]{Could be conclusion this section?}
+
+Robust, unbiased prevalence estimates already exist (see\todo{ref prevalence estimates}), and incidence estimates are the most useful (see\todo{ref intro with importance of incidence}).
+Therefore, a large portion of this thesis concerns itself with estimating the duration of PCR positivity.
+
+The distribution of durations can be expressed in several equivalent ways.
+Let $D_i$ be the random variable for the duration of episode $i$, then the distribution of interest is that of $D_i$.
+I will assume that each $D_i$ is an independent and identically distributed realisation of a random variable $D$; extending the methodology to account for covariates is left to further work.
+For an infection in discrete time, the duration is a non-negative integer, \ie $D \in \{1, 2, \dots\}$.
+The first way of expressing the distribution is as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=0}^d f_{D}(i)$.
+The \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$.
+The \emph{hazard} at time $t$ is the probability of recovering at time $t$ conditional on having survived to time $t$: $\lambda(t) = \prob(D = t \mid D \geq t) = f_{D}(d) / S(d)$.
+Throughout this thesis, the survival function will be the primary quantity of interest, and it will normally be parameterised in terms of the hazard; these two quantities are related by $S(t) = \prod_{i=1}^{t-1} (1 - \lambda(i))$.
+
+Two approaches are commonly used to estimate durations\todo{cite Sweeney}: either modelling an underlying biomarker or modelling the duration directly.
+For SARS-CoV-2, both approaches would be based on the results of PCR testing.
+Modelling a biomarker would mean modelling the viral load, measured as a Ct value.
+Modelling the duration would consider only the binary result at each test (positive or negative); in this thesis a \emph{survival analysis} framework is adopted.
+Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
+Other related approaches, such as a multi-state model, could also be considered (multi-state models can be viewed as a generalisation of survival analysis~\autocite{jacksonMSM}).
+Therefore, the biomarker approach considers more information per observation, however, a model for the biomarker is required which generally requires stronger assumptions.
+Modelling a biomarker uses more information per observation, but requires stronger assumptions.
 

### 2023-09-12 cis-imperfect-testing.tex
Notes on what should go in imperfect testing chapter

@@
 
+\begin{itemize}
+  \item When sensitivity is too bad (0.6), cannot recover the true distribution. The assumptions in the model, that the first test is a true positive and no missed infections prior, cannot hold.
+  \item Assuming too low a test sensitivity (bottom left triangle) means that number of missed infections will be overestimated and the duration distribution will be overestimated (\ie episodes are too long on average).
+\end{itemize}
+
 \section{Application to CIS data} \label{imperf-test:sec:application}
@@
 
+\begin{itemize}
+  \item Survival prior matters for quantitative details but not qualitative shape. 
+  \item Number of missed infections prior is more important.
+  \item Unclear why this is the case, probably model misspecification meaning that fewer missed infections than expected.
+  \item Two possible misspecifications: test sensitivity (not constant) and constant incidence.
+\end{itemize}
+
 \section{Conclusion} \label{imperf-test:sec:conclusion}

### 2023-09-12 incidence-prevalence.tex
Try and make this chapter reasonable...

@@
 \end{align}
-Now, denote by $P_{t,t'}$ the number of episodes beginning at time $t$ that are prevalent at time $t'$ (where $t' \geq t$).
+Now, denote by $P_{t_z,t_p}$ the number of episodes beginning at time $t_z$ that are prevalent at time $t_p$ (where $t' \geq t$).
 This is $Z_t$ minus those that have recovered:
 \begin{align}
-    P_{t,t'} = \begin{cases}
-      0 &t' < t\\
-      Z_t - \sum_{i=0}^{t'-t-1} R_{t,t+i} &t \leq t' < t + d_\text{max}\\
-      0 &t' > t + d_\text{max}.
-  \end{cases}
+    P_{t_z,t_p} = \begin{cases}
+      0 &t_p < t_z\\
+      Z_t - \sum_{i=0}^{t_p-t_z-1} R_{t_z,t_z+i} &t_z \leq t_p < t_z + d_\text{max}\\
+      0 &t_p \geq t_z + d_\text{max}.
+  \end{cases} \label{incprev:eq:Ptt-to-R}
 \end{align}
-Finally, the prevalence at $t$ can be expressed as $P_t = \sum_{t'=-\infty}^t P_{t',t}$.
+Finally, the prevalence at $t$ can be expressed as the total number of people prevalent at time $t$, regardeless of when they were infected:
+\begin{align}
+  P_t
+  &= \sum_{t_z=\infty}^\infty P_{t_z,t} \\
+  &= \sum_{i=0}^{d_\text{max}-1} P_{t-i,t} \label{incprev:eq:Pt-to-Ptt}.
+\end{align}
 
@@
 \begin{align}
-  \sum_{i=t}^{t'-1} R_{t,i} \mid \vec{Z} &\sim \text{Binomial}(Z_t, \sum_{i=1}^{t-t'} f_d(i)). \\
-\intertext{Therefore:}
-\E(P_{t-t'} \mid \vec{Z})
-  &= \E(Z_t - \sum_{i=t}^{t'-1} R_{t,i} \mid \vec{Z}) \\
-  &= Z_t (1 - \sum_{i=1}^{t-t'} f_d(i)).
-\intertext{Finally:}
+  \sum_{i=0}^{t'} R_{t,t+i} \mid \vec{Z} &\sim \text{Binomial}(Z_t, F_d(t'+1)).
+\end{align}
+where $F_d(t) = \sum_{i=1}^t f_d(t) = \prob(D \leq t)$ is the CDF of $D$.
+Therefore:
+\begin{align}
+  \E(\sum_{i=0}^{t'} R_{t,t+i} \mid \vec{Z}) = Z_t F_d(t'+1) \label{incprev:eq:EsumRt}
+\end{align}
+
+Now consider $P_t$:
+\begin{align}
 \E(P_t \mid \vec{Z})
-&= \E(\sum_{t'=-\infty}^t P_{t-t'} \mid \vec{Z}) \\
-&= \sum_{t'=-\infty}^t Z_t (1 - \sum_{i=1}^{t-t'} f_d(i)).
+  &= \E\left(\sum_{i=1}^{d_\text{max}-1} P_{t-i,t} \mid \vec{Z} \right) &\text{by \cref{incprev:eq:Pt-to-Ptt}} \\
+  &= \E\left(\sum_{i=1}^{d_\text{max}-1} \left( Z_{t-i} - \sum_{j=0}^{i-1} R_{t-i,t-i+j} \right) \mid \vec{Z} \right) &\text{by \cref{incprev:eq:Ptt-to-R}}\\
+  &= \sum_{i=1}^{d_\text{max}-1} \left( Z_{t-i} - \sum_{j=0}^{i-1} \E( R_{t-i,t-i+j} \mid \vec{Z}) \right) \\
+  &= \sum_{i=1}^{d_\text{max}-1} \left( Z_{t-i} - Z_{t-i} F(i) \right) &\text{by \cref{incprev:eq:EsumRt}}\\
+  &= \sum_{i=1}^{d_\text{max}-1} Z_{t-i} (1 - F(i)) \\
+  &= \sum_{i=1}^{d_\text{max}-1} Z_{t-i} S(i+1)
 \end{align}
-The quantity $1 - \sum_{i=1}^{t-t'} f_d(i)$ is the probability that an infection episode's duration is at least $t - t' + 1$, which is the survival function of $D$, denoted $S(t - t' + 1)$.
-This equation, normally expressed in terms of the survival function, forms the basis of discrete-time \emph{backcalculation}.
+where $S(i+1) = 1-F(i) = \prob(D \geq i+1)$ is the \emph{survival function} of $D$ (see \cref{inprev:sec:duration}).
 
@@
 
+\todo[inline]{Need some conclusion text for this section}
+
 \section{Prevalence data}
 
-A \emph{prevalence survey} is a survey that tests a sample of the population for the presence of an infection.
-Series of prevalence surveys are the form of measuring prevalence that this thesis is concerned with.
-This thesis aims to infer the incidence rate from prevalence surveys.
-I use Bayesian inference, therefore the data contributes through the likelihood.
+% A \emph{prevalence survey} is a survey that tests a sample of the population for the presence of an infection.
+% Series of prevalence surveys are the form of measuring prevalence that this thesis is concerned with.
+% This thesis aims to infer the incidence rate from prevalence surveys.
+% I use Bayesian inference, therefore the data contributes through the likelihood.
 
-The likelihood contribution from the survey at a time $t$ is based on a sample of $n_t$ individuals from the population of interest returning $x_t$ positive tests.
-For now, I assume that the sample is chosen uniformly at random from the population of interest.
-Assume that the surveys are independent of each other and $\vec{Z}$, conditional on the population prevalence at that time.
-The population prevalence is a latent quantity.
-The likelihood is then $\p(\vec{x} \mid \vec{n}, \vec{Z}) = \int \p(\vec{P} \mid \vec{Z}) \prod_t \p(x_t \mid n_t, \vec{P}_t) d\vec{P}$, where $\vec{x} = (x_1, x_2, \dots)^T$ and similarly for $\vec{n}$.
+Prevalence surveys, as described in\todo{insert ref}, sample individuals at random from the population of interest and test if they are prevalence.
+Therefore, we view the number of positive tests on day $t$ as a random variable $X_t$ based on a sample of $n_t$ individuals.
+The $X_t$s are conditionally independent, given the population prevalences at each time, $\vec{P}_t$. 
+%For now, I assume that the sample is chosen uniformly at random from the population of interest.
+%Assume that the surveys are independent of each other and $\vec{Z}$, conditional on the population prevalence at that time.
+%The population prevalence is a latent quantity.
+From these assumptions, $X_t \mid \vec{P} \sim \Bin(n_t, P_t / N)$.
+I base inference on the likelihood, $p(\vec{x} \mid \vec{Z}) = \int p(\vec{x} \mid \vec{P}) p(\vec{P} \mid \vec{Z}) d\vec{P} = \int \prod_t p(x_t \mid P_t) p(\vec{P} \mid \vec{Z}) d\vec{P}$, where $\vec{x} = (x_1, x_2, \dots)^T$.
 
-In settings I consider, the uncertainty arising from the distribution $\vec{P} \mid \vec{Z}$ is negligible.
+For most prevalence surveys, the uncertainty arising from the distribution $\vec{P} \mid \vec{Z}$ is negligible.
 This can be seen by considering the variance in the likelihood.
-$x_t$ is binomially distributed with parameters $n_t$ and $P_t / N$, where $N$ is the size of the population.
+Assume $x_t \sim \Bin(n_t, P_t / N)$, where $N$ is the size of the population.
 Therefore, by the law of total variance:
@@
 Hence, assuming that $\var(\vec{P}) = \matr{0}$ is a reasonable approximation.
-This is equivalent to assuming that $\vec{P}$ is a deterministic function of $\vec{Z}$, generally $\vec{P} = \E( \vec{P} \mid \vec{Z})$.
+This is equivalent to assuming that $\vec{P}$ is a deterministic function of $\vec{Z}$, $\vec{P} = \E( \vec{P} \mid \vec{Z})$.
 
@@
 
-\section{Estimating the duration}
-\todo[inline]{Could be conclusion this section?}
+\todo[inline]{Explain backcalculation in HIV/AIDS context, and how it relates to what is above.}
+
+\section{Estimating the duration} \label{inprev:sec:duration}
 
@@
 
+\todo[inline]{Comment on previous estimates of duration for SARS-CoV-2, but not sure where exactly this goes...}
+
 \end{document}

### 2023-09-19 cis-perfect-testing.tex
Lots of text on new model

@@
 In the biomarker approach, computational efficiency is an issue because the number of random effects scales with the number of detected infection episodes.
-There are a large number of detected episodes in CIS; in the period I will consider, around 50 times more than in ATACCC.
+There are many detected episodes in CIS; in the period I will consider, around 50 times more than in ATACCC.
 Therefore, the biomarker approach would be computationally challenging with the large sample size in CIS.
@@
 This phase is not representative of the overall study: infections are more likely to be detected during this phase.
-Therefore, estimating the probability of being detected conditional on being infected in this period (which would be implictly done in this previous method)  will overestimate the probability of being detected.
+Therefore, estimating the probability of being detected conditional on being infected in this period (which would be implicitly done in this previous method)  will overestimate the probability of being detected.
 
@@
 
-In this section, I develop a model accounting for double interval censoring and undetected episodes, modelling only individuals with detected episodes.
-This approach is based on long-standing methodology~\autocites{heiseyModelling}{dempsterMaximum}{turnbullEmpirical}; in particular, I adopt the framework of \textcite{heiseyModelling}.
+In this section, I develop a model accounting for double interval censoring and undetected episodes.
+This approach extends the framework of \textcite{heiseyModelling} to include individuals without detected episodes.
+This framework is itself based long-standing methodology~\autocites{dempsterMaximum}{turnbullEmpirical}.
 The conceptual approach here is to form a cohort of individuals for which we observe at least one positive test, and then consider a study which had only enrolled these individuals.
-Previous studies consider the case where nothing was observed regarding individuals without a positive test (left truncation).
-I do not make use of this additional information here because it makes little difference but increases the computational requirements and strength of assumptions regarding reinfections/immunity required (see \cref{E-appendix:total-model} for details).
-The modelling approach then corrects for this selection bias, by considering \emph{ghost} infections that occur in identical individuals but were undetected.
 
-Ghost infections are a mathematical convenience to correct for the undetected episodes which arise through the following thought experiment.
-For each detected episode $i$, let $n_i-1$ be the number of undetected episodes, such that $n_i$ is the total number of episodes.
-Further, assume that each of these $n_i$ infections episodes were drawn independently from $i$'s distribution of episode beginning and end times and all other characteristics of these infection episodes are identical; most notably, the testing schedule and, if the model were to be extended, other covariates of the individual in which the episode occurs.
-If we had detected all $n_i$ episodes, then our sample of detected episodes would be unbiased; however, we did not, and the detected episode is systematically longer than the other $n_i-1$ episodes (as explained in \cref{perf-test:sec:problem}).
+\todo[inline]{Add an overview to the approach taken}
 
-In the following method, I consider $n_i$ and the duration of the missing episodes as missing data.
-With the full dataset of $n_i$ infection episodes, the problem reduces to double interval censored survival analysis, which has a known likelihood.
-Therefore, I will consider the posterior augmented with all the $n_i$s and then integrate over the $n_i$s to form the desired posterior.
+For our purposes, an infection episode $i$ is fully characterised by the day it begins, $b_i$, the day it ends, $e_i$, and the testing schedule of the individual in which it occurs, $\sched_i$.
+Denote by $\tilde{y}_i$ the triplet $(b_i, e_i, \sched_i)$.
+Even when an episode is detected, $b_i$ and $e_i$ are unobserved due to the interval censoring.
 
-\todo[inline]{Maybe move the following paragraph elsewhere}
-It is important to note that there is not a correspondence between the number of ghost episodes and the number of infection episodes that were missed within the true cohort.
-Within the true cohort, the episodes in which the undetected episodes occurred are different to those in which detected episodes occurred.
-In particular, the individuals in which undetected episodes occurred would (on average) have longer spacing between their tests.
+The observations of detected episode $i$'s are fully characterised by the time of the negative prior to the episode, $t_{i,1}$, the time of the negative following the episode, $t_{i,2}$, and the test schedule of the individual in which it occurs, $\sched_i$.
+$t_{i,1}$ and $t_{i,2}$ are deterministic functions of $\tilde{y}_i$.
+Specifically, $t_{i,1} = \tbefore_{\sched_i}(b_i)$ and $t_{i,2} = \tbefore_{\sched_i}(e_i)$.
+Let $y_i = (t_{i,1}, t_{i,2}, \sched_i)$.
+The possible values of $y_i$ require that $t_{i,1}$ and $t_{i,2}$ both appear in the test schedule $\sched_i$, and that there is a positive test between them.
+Formally, this set of possible values is $\mathbb{D} = \{ (t_{i,1}, t_{i,2}, \sched_i) : t_{i,1} \in \sched_i \wedge t_{i,2} \in \sched_i \wedge \tafter_{\sched_i}(t_{i,1}) < t_{i,2} \}$.
 
-The framework of \textcite{heiseyModelling} formalises this thought experiment.
-Start by noting that all relevant information about episode $i$ can be fully characterised by the (unobserved) pair $(b_i, e_i)$ of start and end dates of the episode, which belong to the state space $T \times T$.
-If the individual in which the episode occurs is tested between $b_i$ and $e_i$ (inclusive), then the episode is detected, otherwise it is a ghost episode.
-For the $i$th detected episode, assume that there are $n_{it}$ ghost episodes.
+For any $\vec{d} \in \mathbb{D}$, define $\mathbb{A}_{\vec{d}}$ as the set of triplets $(b, e, \sched_i)$ such that, if an infection episode began at $b$, and ended at $e$, in an individual with test schedule $\sched$, then we would observe $\vec{d}$.
+Denote by $n_{\vec{d}}$ the number of detected episodes with observations $\vec{d}$.
 
-To relate this to the data, define three sets which partition the space of possible infection and recovery times (\ie: are strict subsets of $T \times T$), based on the observations associated with episode $i$ (graphically shown in Figure \ref{perf-test:fig:partitionSpace}).
-Each episode, whether detected or not, must fall into one of these three classes with a probability that is a function of $\theta$.
+Undetected (or excluded) episodes have a start time before the individual's first test, an end time after their last test, or began and ended between tests.
+Formally, define undetected episodes as $\mathbb{U} = \{ (b_i, e_i, \sched_i) : b_i \leq \min(\sched_i) \vee e_i \geq \max(\sched_i) \vee e_i < \tafter_{\sched_i}(b_i) \}$.
+Denote by $n_u$ the unknown number of undetected episodes which I will treat as missing data.
 
-\begin{itemize}
-\item
-  Admissible episodes, $\alpha_i$, which have an infection and
-  recovery time consistent with the data observed. That is, $b_i \in [l_i^{(b)}, r_i^{(b)}]$, and $e_i \in [l_i^{(e)}, r_i^{(e)}]$.
-  $n_{ia} =1$ of these occurred, each with probability
-  $p_{ia} = \prob((b, e) \in \alpha_i \mid \theta)$.
-\item
-  Undetected (or, for consistency with the literature, truncated) episodes, $\Omega_i^C$, which have an infection and
-  recovery time such that they would not have tested positive. An
-  unknown number, $n_{it}$ of these observed, each with probability
-  $p_{it} = \prob((b, e) \in \Omega^C_i \mid \theta)$.
-\item
-  Inadmissible episodes, $\beta_i$, which are not consistent with the data observed but would have been
-  detected. This is all remaining episodes not in the
-  previous sets. $n_{iu} = 0$ of these occurred, where they would have
-  occurred with probability
-  $p_{iu} = \prob((b, e) \in \beta_i \mid \theta)$.
-\end{itemize}
+The set of all possible infection episodes is $\Omega = \mathbb{D} \cup \mathbb{U}$.
+For any possible infection episode, $\tilde{y}_i \in \Omega$:
+\begin{align}
+  \prob(\tilde{y}_i = (b, e, \sched))
+  &= \prob(E_i = e \mid B_i = b) \prob(B_i = b) \prob(\sched_i = \sched) \\
+  &= \prob(D_i = e - b + 1) \prob(B_i = b) \prob(\sched_i = \sched) \label{perf-test:eq:tildeyi-prob}
+\end{align}
+since the testing schedule is determined by the study design.
+The distribution of $D_i$ is the quantity of interest and parameterised by $\theta$ (see \cref{perf-test:sec:parameters-priors}); the distribution of $B_i$ is uniform over the possible infection times; and I assume the probability of any testing schedule is proportional to the number of people in the cohort with that schedule.
+Furthermore, I assume that this probability is independent for each infection episode.
+Therefore, the counts of each possible infection episode triplet conditional on the total number of infection episodes is multinomially distributed.
 
-The detected region (where we could have observed an episode) is
-$\Omega_i = \alpha_i \cup \beta_i$ (the notation here is used for
-consistency with \textcite{heiseyModelling}), and is the
-complement of $\Omega^C_i$.
+Due to the censoring and missed infections, the observations of episodes are incomplete.
+The observations are equivalent to counts of the number of episodes in each set $\mathbb{A}_{\vec{d}}$, $n_{\vec{d}}$.
+Additionally, I will impute the total number of episodes (including undetected episodes) $n_u$.
+Let $n = n_u + \sum_{\vec{d} \in \mathbb{D}} n_{\vec{d}}$.
+For any infection episode, consider the probability of it being in each set $\mathbb{A}_{\vec{d}}$ or $\mathbb{U}$.
+\begin{align}
+  p_{\vec{d}} 
+  &= \prob(\tilde{y}_i \in \mathbb{A}_{\vec{d}}) \\
+  &= \sum_{(b, e, \sched) \in \mathbb{A}_{\vec{d}}} \prob(\tilde{y}_i = (b, e, \sched)) \\
+  p_U
+  &= \prob(\tilde{y}_i \in \mathbb{U}) \\
+  &= \sum_{(b, e, \sched) \in \mathbb{U}} \prob(\tilde{y}_i = (b, e, \sched))
+\end{align}
+
+Each infection episode is independent of the others, condtional on the model parameters.
+Which set each episode is in is a categorical random variable.
+Therefore, the vector of all counts, $\vec{n}$ (including the unknown $n_u$), is distributed multinomially, conditional on its total, $n$.
+Furthermore, $n_d = n - \sum_{\vec{d} \in \mathbb{D}} n_{\vec{d}}$.
+
+Denote by $\mathbb{A}_{t_1,t_2,\sched}$ the set of all pairs $(b, e)$ such that, if an infection episode began at $b$, and ended at $e$, in an individual with test schedule $\sched$, then the negative immediately before the start of the episode occurs at time $t_1$ and the negative immediately following the episode occurs at time $t_2$ (see \cref{perf-test:fig:partitionSpace}).
+Define the function $\tbefore_\sched(t)$ and $\tafter_\sched(t)$ as the time of the next test before and after time $t$ (excluding $t$) for an individual with test schedule $\sched$ respectively.
+Since $b \leq e$, for any biologically-possible infection episode, $t_1 \leq t_2$.
+If $\tafter_\sched(t_1) = t_2$ then the infection would be undetected because there are no positive tests.
+Otherwise,$\tafter_\sched(t_1) < t_2$ and, for any $(b, e) \in \mathbb{A}_{t_1,t_2,\sched}$, our observations bounding the beginning of the episode are the negative at $t_1$ and the positive at $\tafter_\sched(t_1)$.
+Similarly, our observations bounding the end of the episode are the positive at $t_2$ and the negative at $\tafter_\sched(t_2)$.
+Therefore, we know from our observations $t_1 < b \leq \tafter_\sched(t_1)$ and $\tbefore_\sched(t_2) \leq e < t_2$.
+
+I assume that infection episode $i$ occurs, independent of all other infection episodes, in an individual with test schedule $\sched_i$ with probability proportional to the number of such individuals in the cohort, $n_{\sched_i}$.
+This assumption is equivalent to assuming that all individuals are equally likely to be infected and there is no immunity.
+The latter assumption is unrealistic, however, since the cohort is much larger than the number of infection episodes, reinfections occur with negligible probability, and this assumption is a good approximation.
+Further, assume that $b_i$ is drawn iid from a uniform distribution over possible infection times.
+The time of $e_i$, conditional on $b_i$ is determined by the distribution function.
+% Should already have this bit written out somewhere.
+
+From this, the probability that $\tilde{y}_i$ is in the set $\mathbb{A}_{t_1,t_2,\sched}$ is determined only by $\theta$.
+Conditional on the total number of infection episodes, $\ntot$, the count of episodes in each set is multinomial.
+Define by $y$ as the vector of counts for each observed set, and by $\tilde{y}$ the count of undetected episodes, \ie the sum of the counts of sets with $t_1 = t_2$.
+Then, $\tilde{y} = \ntot - \sum_j \vec{y}_j$.
+
+For any individual $i$, there is a probability space $B_i \times E_i$ of possible infection episode start and end times.
+\todo{Check this notation}
+A subset of this space is detectable $\mathbb{D}_i \subset B_i \times E_i$.
+The space $\mathbb{D}_i$ can be partitioned by the first and last positive that would be observed for that episode.
+Label this partition $\mathbb{A}_{i,1}, \dots \mathbb{A}_{i,A_i}$.
+For each partition, there are $n_{i,a}$ episodes in it, and each episode has a probability $p_{i,a}$ of being in that partition.
+In addition, there are $y_{i,0}$ episodes that are undetected, and each has a probability $p_{i,0}$ of being undetected.
+$y_{i,0}$ is unknown and considered as missing data, with the remaining $y_{i,1}, \dots, y_{i,A_i}$ being observed.
+Let $\vec{y_i} = (y_{i,1}, \dots, y_{i,A_i})$ be the vector of observed counts and $\vec{y} = (\vec{n_1}, \dots, \vec{y_{n}})$ be the vector of all observed counts, which is equivalent to our data.
+Similarly for $\vec{p_i}$ and $\vec{p}$.
+The counts can be augmented by the unobserved $y_{i,0}$ to give $\yaug_i = (y_{i,0}, \dots, y_{i,A_i})$.
+I model $\vec{y_i}$ conditional on $n_i = \sum_j y_{i,j}$ and $\vec{p}$, then integrate over $\vec{p}$ and express $\vec{p}$ as a deterministic function of the model parameters $\vec{\theta}$ to obtain the posterior.
+
+The augmented counts $\yaug_i$ are, conditional on their sum, distributed multinomially.
+We must have that $y_{i,0} = n_i - \sum_{j=1}^{A_i} y_{i,j}$, to fulfil the support of the multinomial.
+Since there is at most one infection in each area $\mathbb{A}_{i,j}$ for $j \neq 0$, we have:
+\begin{align}
+p(\vec{y_i} \mid n_i, \vec{p_i})
+  &= p(y_{i,0}, \vec{y_i} \mid n_i, \vec{p_i}) \\
+  &= \frac{n_i!}{(n_i - \sum_{j=1}^{A_i} y_{i,j})!} \prod_j p_{i,j}^{y_{i,j}}.
+\end{align}
+
+Now assume that each individual in the cohort is equally likely to be infection.
+Define $n = \sum_i n_i$.
+I model the partition that each infection occurs in as $\yaug \mid n \dist \MN(n, \vec{p})$.
+Since we do not observe any of the $y_{i,0}$, their counts can be combined.
+Let $n_o = \sum_{i,j\geq1} y_{i,j}$ be the total number of observed infection and $n_u = \sum_{i} y_{i,0}$ be the total number of undetected infections.
+The probability of an infection being undetected is $p_u = \frac{1}{n} \sum_i p_{i,0}$.
+Hence, $p(\vec{y} \mid n, \vec{p}) = \frac{n!}{n_u!} p_u^{n_u} \prod_{i} \prod_{j=1}^{A_i} p_{i,j}^{y_{i,j}}$.
+
+The posterior of interest is $p(\theta \mid \vec{y})$.
+Recalling that $\vec{p}$ is a deterministic function of $\theta$ and integrating over $n$, we have:
+\begin{align}
+  p(\theta \mid \vec{y})
+    &\propto p(\theta) p(\vec{y} \mid \theta) \\
+    &= p(\theta) \sum_{n=n_o}^\infty p(\vec{y} \mid n, \theta) p(n) \\
+    &= p(\theta) \sum_{n=n_o}^\infty \frac{n!}{(n - n_o)!} p_u^{n_u} \prod_{i} \prod_{j=1}^{A_i} p_{i,j}^{y_{i,j}} p(n) \\
+    &= p(\theta) \prod_{i} \prod_{j=1}^{A_i} p_{i,j}^{y_{i,j}} \sum_{n=n_o}^\infty \frac{n!}{(n - n_o)!} p_u^{n_u} p(n).
+\end{align}
+Next, I will show that this sum has a closed form under a negative binomial prior for $p(n)$, and derive expressions for $p_{i,j}$ in terms of the survival function parameterised by $\theta$.
 
@@
   \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{cis-perfect-testing/ataccc-approximation-hazard}}
-  \caption[Approximating the ATACCC posterior hazard as a logit-normal]{Comparison of the posterior estimate of the hazard from \cref{E-ATACCC} and its approximation with a logit-normal distribution (kernal density estimate smoothed) for the first 15 hazards.
+  \caption[Approximating the ATACCC posterior hazard as a logit-normal]{Comparison of the posterior estimate of the hazard from \cref{E-ATACCC} and its approximation with a logit-normal distribution (kernel density estimate smoothed) for the first 15 hazards.
 \label{perf-test:fig:approximate-ATACCC-hazard}}

### 2023-09-20 cis-perfect-testing.tex
Minor editing attempt

@@
 
-The observations of detected episode $i$'s are fully characterised by the time of the negative prior to the episode, $t_{i,1}$, the time of the negative following the episode, $t_{i,2}$, and the test schedule of the individual in which it occurs, $\sched_i$.
+The observations of detected episode $i$'s are fully characterised by the time of the negative prior to the episode, $t_{i,1}$, the time of the final positive in the episode, $t_{i,2}$, and the test schedule of the individual in which it occurs, $\sched_i$.
 $t_{i,1}$ and $t_{i,2}$ are deterministic functions of $\tilde{y}_i$.
-Specifically, $t_{i,1} = \tbefore_{\sched_i}(b_i)$ and $t_{i,2} = \tbefore_{\sched_i}(e_i)$.
+Specifically, $t_{i,1} = \tbefore_{\sched_i}(b_i)$ and $t_{i,2} = \tbefore_{\sched_i}(e_i+1)$ where $\tbefore_{\sched}(t) = \max \{ t' \in \sched \ssep t' < t \}$ is the test immediately before time $t$.
 Let $y_i = (t_{i,1}, t_{i,2}, \sched_i)$.
 The possible values of $y_i$ require that $t_{i,1}$ and $t_{i,2}$ both appear in the test schedule $\sched_i$, and that there is a positive test between them.
-Formally, this set of possible values is $\mathbb{D} = \{ (t_{i,1}, t_{i,2}, \sched_i) : t_{i,1} \in \sched_i \wedge t_{i,2} \in \sched_i \wedge \tafter_{\sched_i}(t_{i,1}) < t_{i,2} \}$.
+Formally, this set of possible values is $\mathbb{D} = \{ (t_{i,1}, t_{i,2}, \sched_i) : t_{i,1} \in \sched_i \wedge t_{i,2} \in \sched_i \wedge \tafter_{\sched_i}(t_{i,1}) < t_{i,2} \}$, where $\tafter_{\sched}(t) = \min\{ t' \in \sched \ssep t' > t \}$ is the next test at or after time $t$.
 
@@
 Undetected (or excluded) episodes have a start time before the individual's first test, an end time after their last test, or began and ended between tests.
-Formally, define undetected episodes as $\mathbb{U} = \{ (b_i, e_i, \sched_i) : b_i \leq \min(\sched_i) \vee e_i \geq \max(\sched_i) \vee e_i < \tafter_{\sched_i}(b_i) \}$.
+Formally, define undetected episodes as $\mathbb{U} = \{ (b_i, e_i, \sched_i) : b_i \leq e_i \left( b_i \leq \min(\sched_i) \vee e_i \geq \max(\sched_i) \vee e_i < \tafter_{\sched_i}(b_i) \right) \}$.
 Denote by $n_u$ the unknown number of undetected episodes which I will treat as missing data.
@@
   p_{\vec{d}} 
-  &= \prob(\tilde{y}_i \in \mathbb{A}_{\vec{d}}) \\
-  &= \sum_{(b, e, \sched) \in \mathbb{A}_{\vec{d}}} \prob(\tilde{y}_i = (b, e, \sched)) \\
+  &= \prob(\tilde{y}_i \in \mathbb{A}_{\vec{d}} \mid \theta) \\
+  &= \sum_{(b, e, \sched) \in \mathbb{A}_{\vec{d}}} \prob(\tilde{y}_i = (b, e, \sched) \mid \theta) \\
   p_U
-  &= \prob(\tilde{y}_i \in \mathbb{U}) \\
-  &= \sum_{(b, e, \sched) \in \mathbb{U}} \prob(\tilde{y}_i = (b, e, \sched))
+  &= \prob(\tilde{y}_i \in \mathbb{U} \mid \theta) \\
+  &= \sum_{(b, e, \sched) \in \mathbb{U}} \prob(\tilde{y}_i = (b, e, \sched) \mid \theta)
 \end{align}

### 2023-09-25 cis-perfect-testing.tex
Rewrite likelihood explanation

--- 2023-08-11 cis-perfect-testing.tex
@@
 
-When the time of the initiating and/or terminating events are not known exactly, the survival analysis literature refers to the event(s) as \emph{censored}.
+The time of the initiating and/or terminating events are not known exactly.
+This is known as \emph{censord} in the survival analysis literature.
 Censored events are present in almost all survival analyses.
@@
 
-In this section, I develop a model accounting for double interval censoring and undetected episodes, modelling only individuals with detected episodes.
-This approach is based on long-standing methodology~\autocites{heiseyModelling}{dempsterMaximum}{turnbullEmpirical}; in particular, I adopt the framework of \textcite{heiseyModelling}.
-The conceptual approach here is to form a cohort of individuals for which we observe at least one positive test, and then consider a study which had only enrolled these individuals.
-Previous studies consider the case where nothing was observed regarding individuals without a positive test (left truncation).
-I do not make use of this additional information here because it makes little difference but increases the computational requirements and strength of assumptions regarding reinfections/immunity required (see \cref{E-appendix:total-model} for details).
-The modelling approach then corrects for this selection bias, by considering \emph{ghost} infections that occur in identical individuals but were undetected.
+In this section, I develop a model accounting for double interval censoring and undetected episodes.
+This approach extends the framework of \textcite{heiseyModelling} to include individuals without detected episodes.
+This framework is itself based long-standing methodology~\autocites{dempsterMaximum}{turnbullEmpirical}.
+I will augment the data set with the full number of episodes, including those that are undetected.
+The augmented dataset is standard double interval censored data, with a well-known likelihood.
+I show that integrating over the number of missed epsidoes is analytically tractable.
+Therefore, inference is tractable and implementable within modern, high-performance Bayesian software such as Stan.
 
-Ghost infections are a mathematical convenience to correct for the undetected episodes which arise through the following thought experiment.
-For each detected episode $i$, let $n_i-1$ be the number of undetected episodes, such that $n_i$ is the total number of episodes.
-Further, assume that each of these $n_i$ infections episodes were drawn independently from $i$'s distribution of episode beginning and end times and all other characteristics of these infection episodes are identical; most notably, the testing schedule and, if the model were to be extended, other covariates of the individual in which the episode occurs.
-If we had detected all $n_i$ episodes, then our sample of detected episodes would be unbiased; however, we did not, and the detected episode is systematically longer than the other $n_i-1$ episodes (as explained in \cref{perf-test:sec:problem}).
+Let the total number of infection episodes in the cohort be $\ntot = \ndet + \nnodet$ where $\ndet$ is the number of detected infections and $\nnodet$ is the number of undetected infections.
+There are $\Ncis$ individuals in the cohort.
+Over the short period of time considered, no individual has more than one detected episode. 
+Without loss of generality, let the individuals with a detected episode be indexed by $i = 1, \dots, \ndet$ and the individuals without a detected episode be indexed by $i = \ndet + 1, \dots, \Ncis$.
+Let $j = 1, \dots \ntot$ index infection episodes, ordered such that for $j \leq \ndet$ the $j$th infection episode occurs in the $i$th individual.
 
-In the following method, I consider $n_i$ and the duration of the missing episodes as missing data.
-With the full dataset of $n_i$ infection episodes, the problem reduces to double interval censored survival analysis, which has a known likelihood.
-Therefore, I will consider the posterior augmented with all the $n_i$s and then integrate over the $n_i$s to form the desired posterior.
+All the relevant information about episode $j$ can be fully characterised by the (unobserved) triplet $(b_j, e_j, \sched_j)$ where $b_j$ and $e_j$ are the times the episode being and end respectively and $\sched_j$ is the schedule of tests for the individual in which the episode occurs.
+If $\sched_j$ contains at least one test between $b_i$ and $e_i$ (inclusive), then the episode is detected, otherwise it is undetected.
+The triplet belongs to a state space where $b_j$ and $e_j$ span possible infection and recovery times, and $\sched_j$ is one of the observed test schedules (assuming other test schedules occur with probability zero).
 
-\todo[inline]{Maybe move the following paragraph elsewhere}
-It is important to note that there is not a correspondence between the number of ghost episodes and the number of infection episodes that were missed within the true cohort.
-Within the true cohort, the episodes in which the undetected episodes occurred are different to those in which detected episodes occurred.
-In particular, the individuals in which undetected episodes occurred would (on average) have longer spacing between their tests.
-
-The framework of \textcite{heiseyModelling} formalises this thought experiment.
-Start by noting that all relevant information about episode $i$ can be fully characterised by the (unobserved) pair $(b_i, e_i)$ of start and end dates of the episode, which belong to the state space $T \times T$.
-If the individual in which the episode occurs is tested between $b_i$ and $e_i$ (inclusive), then the episode is detected, otherwise it is a ghost episode.
-For the $i$th detected episode, assume that there are $n_{it}$ ghost episodes.
-
-To relate this to the data, define three sets which partition the space of possible infection and recovery times (\ie: are strict subsets of $T \times T$), based on the observations associated with episode $i$ (graphically shown in Figure \ref{perf-test:fig:partitionSpace}).
+To relate this to the data, define three sets which partition the space of possible infection and recovery times (\ie: are strict subsets of $T \times T$), for individual $i$ (graphically shown in Figure \ref{perf-test:fig:partitionSpace}).
 Each episode, whether detected or not, must fall into one of these three classes with a probability that is a function of $\theta$.
@@
 \item
-  Admissible episodes, $\alpha_i$, which have an infection and
-  recovery time consistent with the data observed. That is, $b_i \in [l_i^{(b)}, r_i^{(b)}]$, and $e_i \in [l_i^{(e)}, r_i^{(e)}]$.
-  $n_{ia} =1$ of these occurred, each with probability
-  $p_{ia} = \prob((b, e) \in \alpha_i \mid \theta)$.
+  Admissible episodes, $\alpha_i$, which have an infection and recovery time consistent with the data observed.
+  This space is empty if individual $i$ had no detected episode (\ie: $i > \ndet$).
+  Otherwise, $b_i \in [l_i^{(b)}, r_i^{(b)}]$, and $e_i \in [l_i^{(e)}, r_i^{(e)}]$.
+  If $i \leq \ndet$ then $n_{ia} =1$ of these occurred in individual $i$, otherwise $n_{ia} = 0$.
+  Conditional on the episode occurring in individual $i$, it is admissible with probability $p_{ia} = \prob((b, e) \in \alpha_i \mid \theta)$.
 \item
-  Undetected (or, for consistency with the literature, truncated) episodes, $\Omega_i^C$, which have an infection and
-  recovery time such that they would not have tested positive. An
-  unknown number, $n_{it}$ of these observed, each with probability
-  $p_{it} = \prob((b, e) \in \Omega^C_i \mid \theta)$.
+  Undetected episodes, $\Omega_i^C$, which have an infection and recovery time such that they would not have tested positive.
+  An unknown number, $n_{iu}$ of these observed in individual $i$.
+  Conditional on the episode occurring in individual $i$, it is undetected with probability $p_{iu} = \prob((b, e) \in \Omega^C_i \mid \theta)$.
 \item
   Inadmissible episodes, $\beta_i$, which are not consistent with the data observed but would have been
-  detected. This is all remaining episodes not in the
-  previous sets. $n_{iu} = 0$ of these occurred, where they would have
-  occurred with probability
-  $p_{iu} = \prob((b, e) \in \beta_i \mid \theta)$.
+  detected.
+  This is all remaining episodes not in the previous sets.
+  For an individual with no detected episodes, $\beta_i$ is the area of possible infections that would have been detected.
+  $n_{i\bar{a}} = 0$ of these occurred in any individual $i$.
+  Conditional on the episode occurring in individual $i$, it is inadmissible with probability $p_{i\bar{a}} = \prob((b, e) \in \beta_i \mid \theta)$.
 \end{itemize}
 
-The detected region (where we could have observed an episode) is
-$\Omega_i = \alpha_i \cup \beta_i$ (the notation here is used for
-consistency with \textcite{heiseyModelling}), and is the
-complement of $\Omega^C_i$.
+The detected region, $\Omega_i = \alpha_i \cup \beta_i$, for individual $i$ is the region where we could have observed an episode. 
+The notation is for consistency with \textcite{heiseyModelling}.
+$\Omega_i$ is the complement of $\Omega^C_i$.
 
@@
 
-These three classes span all possible episodes and are mutually
-exclusive, and hence the total number of episodes,
-$n_i = n_{ia} + n_{iu} + n_{it}$, and
-$p_{ia} + p_{iu} + p_{it} = 1$.
+These three classes span all possible episodes and are mutually exclusive.
+Hence, $\ntot = \ndet + \nnodet = \sum_{i=1}^{\ndet} n_{ia} + \sum_{i=1}^{\Ncis} n_{iu}$.
+For all $i$, $p_{ia} + p_{iu} + p_{i\bar{a}} = 1$.
 
-Episode $i$ and its ghosts independently belong to one of these
-classes. Therefore, conditional on $n_i$, the number of episodes in
-each class (that is, the counts $n_{ia}, n_{iu}, n_{it}$) are
-distributed multinomially. Hence:
+Any episode $j$ independently belongs to one of these classes.
+I assume the individual in which each episode occurs is iid uniform across the cohort.
+Therefore, conditional on $\ntot$, the number of episodes in each class (that is, the counts $n_{ia}, n_{iu}, n_{i\bar{a}}$) are distributed multinomially.
+The conditional probabilities given above are multiplied by $1/\Ncis$ to give the unconditional probability that an episode belongs to each class in a specific individual.
+However, the undetected classes cannot be distinguished by any observable data.
+Therefore, only their total is relevant to the data's likelihood.
+The overall probability of an episode being unobserved is $\pnodet = \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} p_{iu}$.
+Furthermore, the inadmissible classes are guaranteed to have zero episodes and hence do not contribute to the multinomial probability.
+Therefore, I combine the undetected classes into a single class and ignore the inadmissible classes.
+Let $\na = (n_{1a}, \dots, n_{\Ncis a})^T$ be the vector of counts of episodes in the admissible classes.
+The counts $\na$ and $\nnodet$ are distributed multinomially with probabilities $\pcis = (p_{1a}, \dots, p_{\Ncis a}, \pnodet)^T$.
+This is equivalent to the full likelihood with all the counts and marginalising over the different undetected classes.
+The support of the distribution requires $\nnodet = \ntot - \sum_{i=1}^{\Ncis} n_{ia} = \ntot - \ndet$ and $\ntot \geq \ndet$.
+Hence:
 \begin{align}
-&p(n_{ia} = 1, n_{iu} = 0, n_{it} = n_i - 1 \mid n_i, \theta) \\
-&= \frac{n_i!}{n_{ia}! n_{iu}! (n_i- n_{ia} - n_{it})!} p_{ia}^{n_{ia}} p_{ua}^{n_{ia}} p_{it}^{n_i- n_{ia} - n_{it}} \\
-&= \frac{n_i!}{(n_i-1)!} p_{ia} p_{it}^{n_i- 1} &\text{as $n_{ia} = 1$ and $n_{iu} = 0$}\\
-&= n_i p_{ia} p_{it}^{n_i- 1}
+&p(\na \mid \ntot, \theta) \\
+&= p(\na, \nnodet = \ntot - \na \mid \ntot, \theta) \\
+&= \frac{\ntot!}{\prod_{i=1}^{\Ncis} n_{ia}! \nnodet!}  \left(\frac{1}{\Ncis}\pnodet\right)^{\nnodet}\prod_{i=1}^{\Ncis} \left(\frac{1}{\Ncis}p_{ia}\right)^{n_{ia}} \\
+&\propto \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i=1}^{\ndet} p_{ia} \label{perf-test:eq:augmented-likelihood}.
 \end{align}
 
-The posterior we are interested in is
-$p(\theta \mid n_{1a} = 1, n_{1u} = 0, \dots, n_{n_a,a} = 1, n_{n_a,u} = 0)$,
-where $\theta$ are the parameters governing the survival distribution
-and hence allow derivation of the $p$s.
+In forming the probability $\p_{ia}$, the data is used.
+This is not quite correct\todo{better word} because forming a likelihood should not depend on the data.
+However, an alternative where each combination of infection and recovery times is considered its own class is possible.
+Then, the counts for all the classes that previously contained undetected episodes would be zero.
+The resulting likelihood would be the same.
+Therefore, I present this version as it is clearer.
+
+The posterior we are interested in is $p(\theta \mid \na)$ where $\theta$ are the parameters governing the survival distribution.
+The probabilities $\pcis$ are functions of the parameters $\theta$.
 \begin{align}
-&p(\theta \mid n_{ia} = 1, n_{iu} = 0, \dots, n_{n_a,a} = 1, n_{n_a,u} = 0) \\
-&\propto p(\theta) \prod_{i=1}^{n_a} p(n_{ia} = 1, n_{iu} = 0 \mid \theta) \\
-&= p(\theta) \prod_{i=1}^{n_a} \sum_{n_i=1}^\infty p(n_{ia} = 1, n_{iu} = 0, n_{it} = n_i - 1 \mid \theta, n_i) p(n_i \mid \theta) \\
-&= p(\theta) \prod_{i=1}^{n_a} \sum_{n_i=1}^\infty n_i p_{ia} p_{it}^{n_i- 1} p(n_i) \\
-&= p(\theta) \prod_{i=1}^{n_a} p_{ia} \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i).\label{eqn:sum}
+&p(\theta \mid \na) \\
+&\propto p(\theta) p(\na \mid \theta) \\
+&= p(\theta) \sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) p(\na \mid \theta, \ntot) \\
+&\propto p(\theta) \sum_{\ntot=\ndet}^\infty \left( p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i=\ndet}^{\ndet} p_{ia} \right) \\
+&= p(\theta) \left( \prod_{i=\ndet}^{\ndet} p_{ia} \right) \left( p(\ntot \mid \theta) \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \right) \label{perf-test:eq:full-posterior}.
 \end{align}
 
-In the remainder of the section, expressions for $p_{ia}$,
-$p_{it}$, and analytical solutions for the sum in \cref{eqn:sum} are derived in terms of the observed data, 
-under different priors.
+In the remainder of the section, expressions for each $p_{ia}$, $p_{it}$, and analytical solutions for the sum in \cref{perf-test:eq:full-posterior} are derived in terms of the observed data, under different priors.
 
-I derive an analytical solution to $\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i)$ under two different assumptions for $p(n_i)$.
-First, we consider the improper prior $p(n_i) \propto 1/n_i$, this prior has several attractive properties.
-Then, we consider a negative binomial distribution (which includes the geometric and Poisson distributions as a special and limiting case respectively).
-
-The reference prior for $n_i$ is $p(n_i) \propto 1/n_i$~\autocite{heBayesiana}.
-Under this prior,
-$\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) \propto \sum_{n_i=1}^\infty p_{it}^{n_i-1} = 1/(1-p_{it})$.
-This likelihood is equal to the conditional likelihood used in the frequentist approaches previously mentioned.
-It has been used elsewhere for Bayesian inference~\autocite{zhouUnified}, but the justification within a Bayesian paradigm has not been made explicitly.
-%The prior $p(n_i) \propto 1/n_i$ means that the model coincides with the frequentist conditional likelihood approach, up to the priors on other parameters~\cites[section 4.2]{dempsterMaximum}{heiseyModelling}[section 8.7.5]{gelmanBayesian}.
-% (\protect\hyperlink{ref-dempsterMaximum}{Dempster, Laird, and
-% Rubin} (\protect\hyperlink{ref-dempsterMaximum}{1977}) section 4.2;
-% \protect\hyperlink{ref-heiseyModelling}{Heisey and Nordheim}
-% (\protect\hyperlink{ref-heiseyModelling}{1995});
-% \protect\hyperlink{ref-gelmanBayesian}{Gelman et al.}
-% (\protect\hyperlink{ref-gelmanBayesian}{2013}) section 8.7.5)
-
-Now consider $N_i \dist \NegBin(\mu, r)$ using the
-mean/dispersion parameterisation of the negative binomial. Equivalently:
+I derive an analytical solution to $p(\ntot \mid \theta) \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under a negative binomial prior for $\ntot$.
+This prior is $\NegBin(\mu, r)$ using the mean/dispersion parameterisation of the negative binomial.
+Equivalently:
 \begin{align}
-N_i \mid \lambda_i &\dist \Poi(\lambda_i) \\
-\lambda_i &\dist \GamDist(a, b)
+\ntot \mid \lambda &\dist \Poi(\lambda) \\
+\lambda &\dist \GamDist(a, b)
 \end{align}
@@
 \begin{align}
-\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) 
-&= \int \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i \mid \lambda_i) p(\lambda_i) d\lambda_i \\
-&= \int \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} \frac{\lambda_i^{n_i} e^{-\lambda_i}}{n_i!} p(\lambda_i) d\lambda_i \\
-&= \int e^{-\lambda_i} \sum_{n_i=1}^\infty p_{it}^{n_i- 1} \frac{\lambda_i^{n_i-1}\lambda_i }{(n_i-1)!} p(\lambda_i) d\lambda_i \\
-&= \int e^{-\lambda_i} \lambda_i p(\lambda_i) \sum_{n_t=0}^\infty p_{it}^{n_t} \frac{\lambda_i^{n_t} }{n_t!} d\lambda_i &n_t = n_i - 1 \\
-&= \int e^{-\lambda_i} \lambda_i p(\lambda_i) e^{p_{it}\lambda_i} d\lambda_i &\text{Poisson pmf} \\
-&= \int e^{\lambda_i (p_{it} - 1)} \lambda_i p(\lambda_i) d\lambda_i \\
-&= \int e^{\lambda_i (p_{it} - 1)} \frac{b^a}{\Gamma(a)} \lambda_i^{a-1} e^{-b\lambda_i} \lambda_i d\lambda_i \\
-&= \frac{\Gamma(a+1)b^a}{\Gamma(a) (b+1-p_{it})^{a+1}} \\ 
-  &\; \times \int \frac{(b+1-p_{it})^{a+1}}{\Gamma(a+1)} \\
-  &\; \lambda_i^{(a+1)-1} e^{-(b+1-p_{it})\lambda_i} d\lambda_i \\
-&= \frac{\Gamma(a+1)b^a}{\Gamma(a) (b+1-p_{it})^{a+1}} &\text{as gamma pdf} \\
-&\propto (b+1-p_{it})^{-(a+1)} \\
-&= (r/\mu+1-p_{it})^{-(r+1)} \\
-&\propto (r+\mu (1-p_{it}))^{-(r+1)}
-\end{align}
-Note that $r=0$ recovers the previous derivation (up to
-proportionality).
-
-The above allows us to sample from the posterior without needing to
-sample the $n_i$s, which is much more efficient due to the previously-discussed dimensionality issue.
-Furthermore, each $n_i$ is a discrete parameter, and hence
-cannot be sampled with widely-used inference algorithms such as Stan's Hamiltonian Monte Carlo (HMC)\todo{ref}.
-
-The posterior of $n_i$, while inconvenient to sample from and not of epidemiological interest directly, can be useful for diagnostic or model debugging purposes (for example, see\todo{ref where I use this in the next chapter}).
-This posterior can be found by sampling from the full conditional.
-Specifically, for each posterior sample of $\theta$, we sample one
-draw from $n_i \mid \theta, y$ where $y$ represents all the data.
-
-Due to the conditional independence of the $n_i$s, we find that:
-\begin{align}
-&p(n_i \mid \theta, n_{ia} = 1, n_{iu} = 0) \\
-&\propto p(n_{ia} = 1, n_{iu} = 0, n_{it} = n_i - 1 \mid n_i, p_{iu}, p_{ia}, p_{it}) p(n_i) \\
-&\propto n_i p_{it}^{n_i- 1} p(n_i)
-\end{align}
-With the prior, $p(n_i) \propto 1/n_i$ then this is a geometric
-distribution with parameter $1 - p_{it}$. With a negative binomial
-prior, we have:
-\begin{align}
-n_i p_{it}^{n_i- 1} p(n_i)
-&\propto n_i p_{it}^{n_i- 1} \frac{\Gamma(r + n_i)}{n_i!} \left( \frac{\mu}{r+\mu} \right)^{n_i} \\
-&\propto \frac{\Gamma((r + 1) + (n_i - 1))}{(n_i-1)!} \left( \frac{\mu p_{it}}{r+\mu} \right)^{n_i-1}
-\end{align}
-Therefore, by comparing this expression to the pmf of a negative binomial, we find that $n_i - 1$ is distributed negative binomial with size parameter $r+1$ and
-probability parameter $\frac{r + \mu (1 - p_{it})}{r+\mu}$. The mean
-of this distribution is
-\begin{align}
-\frac{(r+1)\mu p_{it}}{r+\mu(1-p_{it})}
+&\sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} p(\ntot) \\
+&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} p(\ntot \mid \lambda) p(\lambda) d\lambda &\text{$\lambda$ explicit}\\
+&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \frac{\lambda^{\ntot} e^{-\lambda}}{\ntot!} p(\lambda) d\lambda &\text{Poisson pmf}\\
+%&= \int \sum_{\ntot=\ndet}^\infty \frac{1}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \lambda^{\ntot-\ndet} \lambda^{\ndet} e^{-\lambda} p(\lambda) d\lambda \\
+&= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) \sum_{\nnodet=0}^\infty \frac{(\pnodet \lambda)^{\nnodet}}{\nnodet!} d\lambda &\nnodet = \ntot-\ndet\\
+&= \int \lambda^{\ndet} e^{-\lambda} e^{\lambda \pnodet} p(\lambda) d\lambda \\
+&= \int \lambda^{\ndet} e^{-\lambda(1 - \pnodet)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda &\text{Gamma pdf}\\
+&= \int \frac{b^a}{\Gamma(a)} \lambda^{a+\ndet-1} e^{-(b+1-\pnodet)\lambda} \lambda d\lambda \\
+&= \frac{b^a}{\Gamma(a)} \frac{\Gamma(a+\ndet)}{(b+1-\pnodet)^{a+\ndet}} &\text{Gamma pdf}\\
+&\propto (b+1-\pnodet)^{-(a+\ndet)} \\
+&= (r/\mu + 1 - \pnodet)^{-(r+\ndet)} &\text{sub in $\mu$ and $r$}\\
+&\propto(r + \mu (1- \pnodet))^{-(r+\ndet)}.
 \end{align}
 
-Next, I derive $p_{ia}$ and $p_{it}$.
+Substituting this into \cref{perf-test:eq:full-posterior} gives the following expression for the posterior.
+\begin{align}
+&p(\theta \mid \na) \\
+&\propto p(\theta) \left( \prod_{i=\ndet}^{\ndet} p_{ia} \right) (r + \mu (1- \pnodet))^{-(r+\ndet)} \label{perf-test:eq:full-posterior-simplified}.
+\end{align}
+
+$\ntot$ is generally a nuisance parameter, however, its posterior distribution can be useful for diagnostic purposes (\eg: see X\todo{ref where I use this in the next chapter}).
+Given a sample from the posterior of $\theta$, it can be reconstructed by using its full conditional.
+One sample from the full conditional is drawn for each sample from the posterior of $\theta$.
+This gives the joint posterior of $\ntot$ and $\theta$, from which the marginal distribution can be obtained.
+The full conditional is given by:
+\begin{align}
+p(n_\text{tot} \mid \ncis, \theta)
+&\propto p(\ntot) p(\ncis \mid \theta, \ntot) \\
+&\propto p(\ntot) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i=1}^{\ndet} p_{ia} &\text{by \cref{perf-test:eq:augmented-likelihood}} \\
+&\propto \frac{\Gamma(r + \ntot)}{\ntot!} \left( \frac{\mu}{r + \mu} \right)^{\ntot} \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot} \\
+&= \frac{\Gamma(r + \ntot)}{(\ntot-\ndet)!} \left( \frac{\mu \pnodet}{r + \mu} \right)^{\ntot}  \\
+&= \frac{\Gamma((r + \ndet) + (\ntot- \ndet))}{(\ntot-\ndet)!} \left( \frac{\mu \pnodet}{r + \mu} \right)^{\ntot-\ndet}.
+\end{align}
+Comparing this expression to the pmf of a negative binomial, we find that $\ntot - \ndet$ is distributed negative binomial with size parameter $r+\ndet$ and probability parameter $(r + \mu(1 - \pnodet))/(r + \mu)$.
+The mean of this distribution is $(r+\ndet)\mu \pnodet/r+\mu(1-\pnodet)$.
+
+Next, I derive $p_{ia}$ and $p_{iu}$.
 
@@
 &= \frac{1}{T} \sum_{b=\min(t_i)+1}^T S_\theta(t_{ib}^N + 1)
-\end{align}
-
-\subsection{Full posterior}\label{perf-test:sec:full-posterior}
-
-Under the prior $p(n_i) \propto 1/n_i$:
-\begin{align}
-&p(\theta \mid n_{ia} = 1, n_{iu} = 0, \dots, n_{n_a,a} = 1, n_{n_a,u} = 0) \\
-&\propto p(\theta) \prod_{i=1}^{n_a} p_{ia} \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) \\
-&= p(\theta) \prod_{i=1}^{n_a} \frac{p_{ia}}{1-p_{it}} \\
-&\propto p(\theta) \prod_{i=1}^{n_a} \frac{\sum_{b = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(r_i^{(e)} - b - 1) - S_\theta(l_i^{(e)} - b - 1) \right)}{\sum_{b=\min(t_i)}^T S_\theta(t_{ib}^N + 1)} \\
-\end{align}
-
-Under the prior $N_i \dist \text{NegBinom}(\mu, r)$:
-\begin{align}
-&p(\theta \mid n_{ia} = 1, n_{iu} = 0, \dots, n_{n_a,a} = 1, n_{n_a,u} = 0) \\
-&\propto p(\theta) \prod_{i=1}^{n_a} p_{ia} \sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) \\
-&= p(\theta) \prod_{i=1}^{n_a} \frac{p_{ia}}{(r+\mu (1-p_{it}))^{(r+1)}} \\
-&= p(\theta) \prod_{i=1}^{n_a} \frac{\sum_{b=l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(r_i^{(e)} - b - 1) - S_\theta(l_i^{(e)} - b - 1) \right)}{\left( r+\mu/T \left( \sum_{b=\min(t_i)}^T S_\theta(t_{ib}^N + 1) \right) \right)^{(r+1)}} \\
 \end{align}

### 2023-09-26 cis-perfect-testing.tex
Edit chapter, removing appendix

@@
 This framework is itself based long-standing methodology~\autocites{dempsterMaximum}{turnbullEmpirical}.
-I will augment the data set with the full number of episodes, including those that are undetected.
+I augment the data set with the full number of episodes, including those that are undetected.
 The augmented dataset is standard double interval censored data, with a well-known likelihood.
-I show that integrating over the number of missed epsidoes is analytically tractable.
-Therefore, inference is tractable and implementable within modern, high-performance Bayesian software such as Stan.
 
+Integrating over the number of missed episodes is analytically tractable making inference tractable within modern, high-performance Bayesian software such as Stan.
 Let the total number of infection episodes in the cohort be $\ntot = \ndet + \nnodet$ where $\ndet$ is the number of detected infections and $\nnodet$ is the number of undetected infections.
 There are $\Ncis$ individuals in the cohort.
+Without loss of generality, let the individuals with a detected episode be indexed by $i = 1, \dots, \ndet$ and the individuals without a detected episode be indexed by $i = \ndet + 1, \dots, \Ncis$.
 Over the short period of time considered, no individual has more than one detected episode. 
-Without loss of generality, let the individuals with a detected episode be indexed by $i = 1, \dots, \ndet$ and the individuals without a detected episode be indexed by $i = \ndet + 1, \dots, \Ncis$.
 Let $j = 1, \dots \ntot$ index infection episodes, ordered such that for $j \leq \ndet$ the $j$th infection episode occurs in the $i$th individual.
@@
 To relate this to the data, define three sets which partition the space of possible infection and recovery times (\ie: are strict subsets of $T \times T$), for individual $i$ (graphically shown in Figure \ref{perf-test:fig:partitionSpace}).
-Each episode, whether detected or not, must fall into one of these three classes with a probability that is a function of $\theta$.
+Each episode, whether detected or not, must fall into one of these three classes with a probability that is a function of the model parameters, $\theta$.
 
@@
   Conditional on the episode occurring in individual $i$, it is admissible with probability $p_{ia} = \prob((b, e) \in \alpha_i \mid \theta)$.
+  \label{perf-test:def:admissible}
 \item
@@
 These three classes span all possible episodes and are mutually exclusive.
-Hence, $\ntot = \ndet + \nnodet = \sum_{i=1}^{\ndet} n_{ia} + \sum_{i=1}^{\Ncis} n_{iu}$.
-For all $i$, $p_{ia} + p_{iu} + p_{i\bar{a}} = 1$.
+Therefore, these definitions are consistent with the previous definition of $\ntot$.
+That is, $\ntot = \ndet + \nnodet + n_{i\bar{a}} = \sum_{i=1}^{\ndet} n_{ia} + \sum_{i=1}^{\Ncis} n_{iu} + 0$.
+Furthermore, for all $i$, $p_{ia} + p_{iu} + p_{i\bar{a}} = 1$.
 
@@
 Therefore, conditional on $\ntot$, the number of episodes in each class (that is, the counts $n_{ia}, n_{iu}, n_{i\bar{a}}$) are distributed multinomially.
-The conditional probabilities given above are multiplied by $1/\Ncis$ to give the unconditional probability that an episode belongs to each class in a specific individual.
-However, the undetected classes cannot be distinguished by any observable data.
-Therefore, only their total is relevant to the data's likelihood.
+The above probabilities ($p_{iu}$, $p_{ia}$, and $p_{i\bar{a}}$) are conditional on the episode occurring in individual $i$.
+These probabilities are multiplied by $1/\Ncis$ to give the probability that an episode belongs to each class, without conditioning on a specific individual.
+
+The undetected classes for each individual cannot be distinguished by any observable data.
+Each class's count needs to be marginalised over to calculate the likelihood.
+Because the counts have a joint multinomial distribution, this is equivalent to only considering the sum of the counts.
 The overall probability of an episode being unobserved is $\pnodet = \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} p_{iu}$.
-Furthermore, the inadmissible classes are guaranteed to have zero episodes and hence do not contribute to the multinomial probability.
+Furthermore, the inadmissible classes are guaranteed to have zero episodes.
+Counts of zero do not contribute to a multinomial pmf (they multiply by 1).
 Therefore, I combine the undetected classes into a single class and ignore the inadmissible classes.
+
 Let $\na = (n_{1a}, \dots, n_{\Ncis a})^T$ be the vector of counts of episodes in the admissible classes.
-The counts $\na$ and $\nnodet$ are distributed multinomially with probabilities $\pcis = (p_{1a}, \dots, p_{\Ncis a}, \pnodet)^T$.
-This is equivalent to the full likelihood with all the counts and marginalising over the different undetected classes.
+The counts $\na$ and $\nnodet$ are jointly distributed multinomially.
 The support of the distribution requires $\nnodet = \ntot - \sum_{i=1}^{\Ncis} n_{ia} = \ntot - \ndet$ and $\ntot \geq \ndet$.
@@
 &p(\na \mid \ntot, \theta) \\
-&= p(\na, \nnodet = \ntot - \na \mid \ntot, \theta) \\
-&= \frac{\ntot!}{\prod_{i=1}^{\Ncis} n_{ia}! \nnodet!}  \left(\frac{1}{\Ncis}\pnodet\right)^{\nnodet}\prod_{i=1}^{\Ncis} \left(\frac{1}{\Ncis}p_{ia}\right)^{n_{ia}} \\
-&\propto \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i=1}^{\ndet} p_{ia} \label{perf-test:eq:augmented-likelihood}.
+&= p(\nnodet, \na\mid \ntot, \theta) \\
+&= \frac{\ntot!}{\nnodet! \prod_{i=1}^{\Ncis} n_{ia}!}  \left(\frac{1}{\Ncis}\pnodet\right)^{\nnodet}\prod_{i=1}^{\Ncis} \left(\frac{1}{\Ncis}p_{ia}\right)^{n_{ia}} \\
+&\propto \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i=1}^{\ndet} p_{ia} \label{perf-test:eq:augmented-likelihood}
 \end{align}
+the final line follows because $\nnodet = \ntot - \ndet$, $n_{ia} = 1$ for $i \leq \ndet$, and $n_{ia} = 0$ for $i > \ndet$.
 
-In forming the probability $\p_{ia}$, the data is used.
-This is not quite correct\todo{better word} because forming a likelihood should not depend on the data.
+\todo{better phrasing needed for this paragraph, mauybe try email to Shaun?}
+In forming the probability $p_{ia}$, the data is used.
+This is not quite correct because forming a likelihood should not depend on the data.
 However, an alternative where each combination of infection and recovery times is considered its own class is possible.
@@
 
-In the remainder of the section, expressions for each $p_{ia}$, $p_{it}$, and analytical solutions for the sum in \cref{perf-test:eq:full-posterior} are derived in terms of the observed data, under different priors.
+In the remainder of the section, expressions for each $p_{ia}$, $p_{iu}$, and analytical solutions for the sum in \cref{perf-test:eq:full-posterior} are derived in terms of the observed data, under different priors.
 
-I derive an analytical solution to $p(\ntot \mid \theta) \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under a negative binomial prior for $\ntot$.
+I derive an analytical solution to $p(\ntot \mid \theta) = \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under a negative binomial prior for $\ntot$.
 This prior is $\NegBin(\mu, r)$ using the mean/dispersion parameterisation of the negative binomial.
@@
 &= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} p(\ntot \mid \lambda) p(\lambda) d\lambda &\text{$\lambda$ explicit}\\
-&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \frac{\lambda^{\ntot} e^{-\lambda}}{\ntot!} p(\lambda) d\lambda &\text{Poisson pmf}\\
+&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \frac{\lambda^{\ntot} e^{-\lambda}}{\ntot!} p(\lambda) d\lambda &\ntot \dist \Poi\\
 %&= \int \sum_{\ntot=\ndet}^\infty \frac{1}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \lambda^{\ntot-\ndet} \lambda^{\ndet} e^{-\lambda} p(\lambda) d\lambda \\
 &= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) \sum_{\nnodet=0}^\infty \frac{(\pnodet \lambda)^{\nnodet}}{\nnodet!} d\lambda &\nnodet = \ntot-\ndet\\
-&= \int \lambda^{\ndet} e^{-\lambda} e^{\lambda \pnodet} p(\lambda) d\lambda \\
+&= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) e^{\lambda \pnodet} d\lambda &\text{Maclaurin series of $e$} \\
 &= \int \lambda^{\ndet} e^{-\lambda(1 - \pnodet)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda &\text{Gamma pdf}\\
@@
 \begin{align}
-&p(\theta \mid \na) \\
+p(\theta \mid \na)
 &\propto p(\theta) \left( \prod_{i=\ndet}^{\ndet} p_{ia} \right) (r + \mu (1- \pnodet))^{-(r+\ndet)} \label{perf-test:eq:full-posterior-simplified}.
@@
 $\ntot$ is generally a nuisance parameter, however, its posterior distribution can be useful for diagnostic purposes (\eg: see X\todo{ref where I use this in the next chapter}).
-Given a sample from the posterior of $\theta$, it can be reconstructed by using its full conditional.
-One sample from the full conditional is drawn for each sample from the posterior of $\theta$.
-This gives the joint posterior of $\ntot$ and $\theta$, from which the marginal distribution can be obtained.
-The full conditional is given by:
+I reconstruct the posterior using a posterior sample of $\theta$.
+For each posterior sample of $\theta$, I sample from the full conditional of $\ntot$.
+This gives the joint posterior of $\ntot$ and $\theta$.
+The full conditional of $\ntot$ is given by:
 \begin{align}
 p(n_\text{tot} \mid \ncis, \theta)
-&\propto p(\ntot) p(\ncis \mid \theta, \ntot) \\
-&\propto p(\ntot) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i=1}^{\ndet} p_{ia} &\text{by \cref{perf-test:eq:augmented-likelihood}} \\
+&\propto p(\ntot \mid \theta) p(\ncis \mid \theta, \ntot) \\
+&\propto p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} &\text{by \cref{perf-test:eq:augmented-likelihood}} \\
 &\propto \frac{\Gamma(r + \ntot)}{\ntot!} \left( \frac{\mu}{r + \mu} \right)^{\ntot} \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot} \\
@@
 
-By definition, we have
+By definition (see~\cpageref{perf-test:def:admissible}), we have
 \begin{align}
@@
 &=\sum_{b = l_i^{(l)}}^{r_i^{(b)}} \prob \left( l_i^{(e)} \leq E_i \leq r_i^{(e)} \mid B_i = b \right) \prob \left(B_i = b \right) \\
-&=\sum_{b = l_i^{(b)}}^{r_i^{(b)}} \prob \left( l_i^{(e)} - b + 1 \leq D_i \leq r_i^{(e)} - b + 1 \right) \prob \left(B_i = b \right) \\
-&=\sum_{b = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b + 1) - S_\theta(r_i^{(e)} - b + 2) \right) \prob \left(B_i = b \right) \\
-&\propto \sum_{b = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b + 1) - S_\theta(r_i^{(e)} - b + 2) \right)
+&=\sum_{b = l_i^{(b)}}^{r_i^{(b)}} \prob \left( l_i^{(e)} - b + 1 \leq D_i \leq r_i^{(e)} - b + 1 \right) \prob \left(B_i = b \right) &\text{by def of $D_i$} \\
+&=\sum_{b = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b + 1) - S_\theta(r_i^{(e)} - b + 1) \right) \prob \left(B_i = b \right) &\text{by def of $S_\theta$} \\
+&\propto \sum_{b = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b + 1) - S_\theta(r_i^{(e)} - b + 1) \right)
 \end{align}
+\todo[inline]{Fill in references}
 under the assumption of uniform probability of infection time.
 
-Next, we derive $p_{it} = \prob((b, e) \in \Omega^C_i)$, the probability
-that episode $i$ was detected. A detected episode means that no test
-was performed during the episode or there was no negative test prior to
-the episode. Denote by $t_i$ the set of testing times for the
-individual in which episode $i$ was observed, and define the time
-until the next test on the individual after time $t'$ as:
+Next, I derive $p_{iu} = \prob((B_i, E_i) \in \Omega^C_i)$, the probability that episode $i$ was undetected.
+An episode is undetected if and only if no tests are performed during the episode or if there was no negative test prior to the episode.
+Equivalently, that the first test at or after $b$ is after $e$, or that there is no negative test prior to $b$.
+To formalise this, first define $\tau_{\sched_i}(t)$ as the time until the next test at or after time $t$ in the schedule $\sched_i$:
 \begin{align}
-t^N_{it} &= \min \{ t' \in t_i : t' \geq t \} - t.
-\intertext{Then:}
-\Omega^C_i
-&= \{ (b, e) : \nexists t \in t_i. b \leq t \leq e \vee b \leq \min(t_i) \} \\
-&= \{ (b, e) : e - b < t^N_{ib} \vee b \leq \min(t_i) \}.
-\intertext{Hence:}
-1 - p_{it}
-&= 1 - \prob(E_i - B_i < t_{iB_i}^N \vee B_i \leq \min(t_i)) \\
-&= 1 - \prob(E_i - B_i < t_{iB_i}^N \wedge B_i > \min(t_i)) - \prob(B_i \leq \min(t_i)) \\
-&= 1 - \sum_{b=\min(t_i) + 1}^T \prob(E_i - b + 1 < t_{ib}^N \mid B_i = b) \prob(B_i = b) \\
-  &\; - \sum_{t=1}^{\min(t_i)} \prob(B_i = b)\\
-&= 1 - \frac{1}{T} \sum_{b=\min(t_i)+1}^T (1 - S_\theta(t_{ib}^N + 1)) - \frac{\min(t_i)}{T} \\
-&= 1 - \frac{T-\min(t_i)}{T} + \frac{1}{T} \sum_{b=\min(t_i)+1}^T S_\theta(t_{ib}^N + 1)) - \frac{\min(t_i)}{T} \\
-&= \frac{1}{T} \sum_{b=\min(t_i)+1}^T S_\theta(t_{ib}^N + 1)
+\tau_{\sched_i}(t) &= \min \{ t' \in \sched_i : t' \geq t \} - t.
+\end{align}
+Then $\Omega^C_i$ can be written as:
+\begin{align}
+\Omega_i^C = \{ (b, e) \ssep \tau_{\sched_i}(b) + b > e \vee b \leq \min(\sched_i) \}.
+\end{align}
+Only $1 - p_u = 1 - \frac{1}{N} \sum_i^{\Ncis} p_{iu} = \frac{1}{N} \sum_i^{\Ncis} (1 - p_{iu})$ is required (see \cref{perft-test:eq:full-posterior-simplified}).
+Therefore, I derive $1 - p_{iu}$:
+\begin{align}
+1 - p_{iu}
+&= 1 - \prob((B_i, E_i) \in \Omega^C_i) \\
+&= 1 - \prob( \tau_{\sched_i}(B_i) + B_i > E_i \vee B_i \leq \min(\sched_i) ) \\
+&= 1 - \prob( \tau_{\sched_i}(B_i) > D_i , B_i > \min(\sched_i) )  - \prob( B_i \leq \min(\sched_i) ) \\
+&= 1 - \sum_{b={\min\sched_i} + 1}^T \prob( \tau_{\sched_i}(b) > D_i) \prob(B_i = b) - \sum_{t=1}^{\min\sched_i} \prob(B_i = b)\\
+&= 1 - \frac{1}{T} \sum_{b=\min(t_i)+1}^T (1 - S_\theta(\tau_{\sched_i}(b + 1))) - \frac{1}{T} \min\sched_i \\
+&= \frac{1}{T} \left(T - \sum_{b=\min\sched_i+1}^T (1 - S_\theta(\tau_{\sched_i}(b + 1))) - \min\sched_i \right) \\
+&= \frac{1}{T} \left(T - (T - \min\sched_i) + \sum_{b=\min\sched_i+1}^T S_\theta(\tau_{\sched_i}(b + 1)) - \min\sched_i \right) \\
+&= \frac{1}{T} \sum_{b=\min(t_i)+1}^T S_\theta(\tau_{\sched_i}(b + 1)).
 \end{align}
@@
 \ifSubfilesClassLoaded{
-  \appendix
-  \subfile{appendix-cis-perfect-testing}
   \listoftodos

### 2023-09-27 incidence-prevalence.tex
Rearrange chapter

@@
 
-\section{Incidence, prevalence, and duration}
+\section{Definitions}
 
@@
 In this thesis, prevalence is defined as the number of people in the population who would return a positive PCR test (if everyone was tested) at time $t$.
+
+\section{Backcalculation}
+
+\emph{Backcalculation} is the procedure for estimating incidence based on a measurement of prevalence.
+It was first developed in the context of HIV (human immunodeficiency virus) and AIDS (acquired immune deficiency syndrome)~\autocite{brookmeyerMethod}.
+HIV is a virus transmitted sexually or through blood, and causes the disease AIDS.
+AIDS has killed around 40 million people; at its peak, around 2 million people died from AIDS each year\todo{cite WHO}.
+In this context, backcalculation estimates the incidence of HIV infection from data on the number of AIDS diagnoses and the distribution of the time between infection and AIDS diagnosis~\autocite{brookmeyerBackcalculation}{brookmeyerMeasuring}.
+\todo{explain how AIDS diagnosis is "prevalence" data (is it?)}
+
+Two processes are involved in backcalculation: the \emph{prevalence process} and the \emph{observation process}.
+The prevalence process describes the relationship between incidence and prevalence.
+The observation process describes the relationship between prevalence and the observed data.
+
+In \emph{deterministic backcalculation}, the observed prevalence, $x_t/n_t$, is assumed to be equal to $P_t/N$.
+However, this is often a poor approximation.
+\emph{Statistical backcalculation} retains the sampling distribution of $x_t$ (or an approximation of it, such as a Poisson).
+
+Define the first day of an infection episode in individual $i$ as $B_i$, and assume that the probability of $i$ being infected multiple times within the period of interest is negligible.
+The time between being infected and first being detectable is short (see\todo{ref relevant part}), and therefore I often assume that $B_i$ is the same as the time of infection.
+I keep this assumption for the remainder of this chapter.
+
+Backcalculation makes use of this relationship to estimate incidence from prevalence, assuming that the duration is known.
+
+Deterministic backcalculation assumes that the variance in the population is negligible, and therefore that the prevalence is equal to the mean prevalence.
+This is justified because the variance due to sampling is much larger than the variance due to the true prevalence.
+
+\todo[inline]{Explain backcalculation in HIV/AIDS context, and how it relates to what is above.}
+
+\section{Prevalence process}
 
@@
 
-\section{Prevalence data}
+\section{Observation process}
 
@@
 
-\section{Backcalculation}
-
-\emph{Backcalculation} is a procedure that combines the elements of the previous two sections to estimate incidence based on a measurement of prevalence
-
-In \emph{deterministic backcalculation}, the observed prevalence, $x_t/n_t$, is assumed to be equal to $P_t/N$.
-However, this is often a poor approximation.
-\emph{Statistical backcalculation} retains the sampling distribution of $x_t$ (or an approximation of it, such as a Poisson).
-
-Define the first day of an infection episode in individual $i$ as $B_i$, and assume that the probability of $i$ being infected multiple times within the period of interest is negligible.
-The time between being infected and first being detectable is short (see\todo{ref relevant part}), and therefore I often assume that $B_i$ is the same as the time of infection.
-I keep this assumption for the remainder of this chapter.
-
-Backcalculation makes use of this relationship to estimate incidence from prevalence, assuming that the duration is known.
-
-Deterministic backcalculation assumes that the variance in the population is negligible, and therefore that the prevalence is equal to the mean prevalence.
-This is justified because the variance due to sampling is much larger than the variance due to the true prevalence.
-
-\todo[inline]{Explain backcalculation in HIV/AIDS context, and how it relates to what is above.}
-
 \section{Estimating the duration} \label{inprev:sec:duration}

### 2023-09-28 cis-perfect-testing.tex
Convergence diagnostics
finish chapter rewrite

@@
 I take a Bayesian survival analysis approach to model the coarse sampling of the CIS (see \cref{perf-test:sec:problem}).
-Flat priors can be unintentionally very informative, and hence (weakly) informative priors should be used.
+Flat priors can be unintentionally very informative, and hence uninformative priors should be used.
 Of particular interest is a strongly informative prior to combine the information in the CIS with the results of \cref{E-ATACCC} (see \cref{perf-test:sec:parameters-priors}).
-Through a simulation study, I show that these methods perform well, and in this context are not sensitive to model or whether the prior is weakly or strongly informative (see \cref{perf-test:sec:simulation-study}).
+Through a simulation study, I show that these methods perform well, and in this context are not sensitive to model or whether the prior is vague or strongly informative (see \cref{perf-test:sec:simulation-study}).
 However, application in reality must account for false negatives, the subject of \cref{E-imperf-test}.
@@
 \begin{figure}
-  \centering \includegraphics{cis-perfect-testing/truncation}
+  \makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{cis-perfect-testing/truncation}}
   \caption[Undetected episodes in CIS data]{An unknown number of infection episodes in CIS are undetected.
@@
 A subtlety that makes the CIS different from standard left truncation is that information on individuals in which undetected infections occurred is available because they were enrolled in the study.
-This information could be incorporated into the analysis, although doing so increases the computational burden and requires strong assumptions regarding reinfections without noticeably affecting the results (see \cref{appendix:total-model}).
 In standard left truncation, nothing is known about the missed events.
@@
 \end{align}
-Only $1 - p_u = 1 - \frac{1}{N} \sum_i^{\Ncis} p_{iu} = \frac{1}{N} \sum_i^{\Ncis} (1 - p_{iu})$ is required (see \cref{perft-test:eq:full-posterior-simplified}).
+Only $1 - p_u = 1 - \frac{1}{N} \sum_i^{\Ncis} p_{iu} = \frac{1}{N} \sum_i^{\Ncis} (1 - p_{iu})$ is required (see \cref{perf-test:eq:full-posterior-simplified}).
 Therefore, I derive $1 - p_{iu}$:
@@
 
-Instead, I propose a weakly informative prior Beta(0.1, 1.9), which has mean 0.05 and minimal information.
+Instead, I propose a weakly informative (or vague) prior Beta(0.1, 1.9), which has mean 0.05 and minimal information.
 The amount of information in a Beta distribution is related to the sum of its parameters.
@@
   \centering \includegraphics{cis-perfect-testing/vague-prior}
-  \caption[Weakly informative priors for the hazard]{The density of Beta(0.1, 1.9) vague prior (left) and the prior predictive survival time (right).
+  \caption[Vague prior for the hazard]{The density of Beta(0.1, 1.9) vague prior (left) and the prior predictive survival time (right).
 The prior predictive survival time is very vague using a Beta(0.1, 1.9) prior.
@@
   \lambda_t &\sim \text{Beta}(\alpha_t, \beta_t) &t = 1, 2, \dots \\
-  \alpha_t &= c_t h_t + \alpha_0 \\
-  \beta_t &= c_t (1 - h_t) + \beta_0
+  \alpha_t &= k_t h_t + \alpha_0 \\
+  \beta_t &= k_t (1 - h_t) + \beta_0
 \end{align}
@@
 If $\lambda_t$ is given the prior distribution $\text{Beta}(\alpha_0, \beta_0)$, and we then have $k_t$ observations with $k_t h_t$ successes, then the posterior distribution for $\lambda_t$ is $\text{Beta}(\alpha_t, \beta_t)$ (as defined above).
+
+\todo{Add plot of $k_t$}
 
@@
 
-All analyses converged satisfactorily\todo{put numbers once confirmed exact analyses to use}.
-Repeating simulations showed very little variation in the results, due to the large sample size in CIS; therefore, only results from a single simulation are shown.
+Repeating the simulation showed negligible variation in the results (not shown).
+This is caused by the large sample size (4800 detected infections).
+Therefore, only results from a single simulation are shown and dicussed.
 
-The true value of the survival function is very well recovered with little difference between priors (see \cref{perf-test:fig:survival-results}).
+All analyses converged.
+Rhat were below 1.01, all parameters had an effective sample size of 1000, and there were no divergent transitions.
 
-The independent prior has the fewest assumptions and least prior information.
-Therefore, it leads to posterior estimates that are most uncertain.
+The true value of the survival function was generally well recovered (see \cref{perf-test:fig:survival-results}).
+This was especially true for the first 30 days of the distribution.
 
-The second-order random walk is the only prior that smooths the hazard throughout the entire infection.
-This leads to a near-constant hazard, and hence exponential-like survival, in areas where there is little information (long infections).
+The second-order random walk prior performed worst.
+It showed oversmoothing, especially in the later parts of the distribution.
+It smooths towards a linearly changing hazard; here, linearly decreasing (see \cref{perf-test:fig:hazard-results}).
+This means that the survival function decreases slower than an exponential distribution in the tail.
+
+The informative and vague independent priors performed similarly.
+The informative prior had tighter credible intervals, as expected.
+This prior was centered on the correct value for the initial part of the distribution.
+Therefore, the narrow credible intervals are unsurprising.
+For hazards after day X\todo{check and insert}, the informative and vague priors are the same.
+Therefore, the posterior distributions are almost identical.
 
@@
 Results using each of the three different priors discussed in \cref{perf-test:sec:parameters-priors} are shown.
+y-axis has either natural (left) or log (right) scale.
+Only showing first 100 days of the distribution.
 \label{perf-test:fig:survival-results}}
@@
 
-This chapter shows that the model I propose can recover the true survival function well, is not sensitive to the prior used for the survival distribution, nor requires information on the number of infections that were undetected.
+This chapter shows that the model I propose can recover the true survival function well.
 However, the problem becomes substantially harder if false negatives are present, as we shall see in the next chapter.
 Unfortunately, false negatives do occur in the CIS data.
+
+Reasonable choices of the prior for the hazard are vague (although not flat) or informative.
+A smoothing prior performed poorly, although that may be influenced by the choice of hyperparameters.
+Reducing the amount of smoothing or type of smoothing may help.
+More flexible smoothers, such as a Gaussian Process for additional flexibility and smoothing and different time-scales~\autocite{saulGaussian}.
 
@@
 Immunity in particular affects the duration substantially, although age and variant may also make a contribution~\autocites{hakkiOnset}{russellWithinhost}.
-Covariates can be included with a variety of model, the simplest being a proportional hazards model.
-
-The form of priors used for the survival function could also be extended.
-In particular, the form of smoothing chosen here is very simple and could be replaced with a more flexible model, such as a Gaussian Process for additional flexibility and smoothing and different time-scales~\autocite{saulGaussian}.
+Covariates can be included with a variety of models, the simplest being a proportional hazards model.
+A major challenge here is that the calculation of $p_t$ is greatly complicated, because it is no longer linear in the survival function.
 
 The prior would likely be more important if the number of detected infections was smaller.
-Therefore, exploring its impact would be important for designing future surveys.
+Therefore, exploring its impact could be important for future surveys.
 A particular application would be estimation in the early stages of an epidemic of a novel pathogen.
@@
 
+I have developed a flexible simulation and inference framework for this type of analysis.
+This framework could be used to simulate alternate study designs.
+The design of CIS was created on a very short timescale in March 2020, in response to the pandemic's rapid spread.
+Therefore, it is likely that there are more efficient designs.
+These could be more cost-effective.
+Improved cost-effectiveness could allow more rapid response to potential pandemics because the threshold for policy-makers to approve the study would be lower.
+A long-term, preparatory effort to develop a more efficient design in preparation of a future pandemic would be worthwhile.
+
+Other useful preparatory work would be developing reasonable priors.
+These could be based on seasonal viruses that are of the same family as those likely to cause future pandemics.
+For example, seasonal influenzas and coronaviruses.
+These would allow analyses such as those performed in these chapters to be performed more rapidly in a future pandemic scenario.
+
 \section{Conclusion} \label{perf-test:sec:conclusion}

### 2023-09-28 cis-perfect-testing/results.R

@@ -1,23 +1,11 @@
-library(dplyr)
+suppressMessages(library(dplyr))
 library(ggplot2)
 library(patchwork)
 library(purrr)
 library(tidybayes)
 library(tidyr)
 source(here::here("utils.R"))
-readRDS(here::here("cisRuns-output/perfect_posteriors.rds")) |>
-    filter(
-        .width == 0.95, # Credible intervals are narrow anyway so only show 95%
-        survival_prior %in% c(
-            "ATACCC-new",
-            "RW2-infer-sigma",
-            "vague"
-        ),
-        # Use the reference prior on the number of missed infections as this doesn't matter
-        missed_prior == "vague", 
-        missing_model == "total"
-    ) |>
-    distinct(survival_prior)
+
 tbl_posteriors = readRDS(here::here("cisRuns-output/perfect_posteriors.rds")) |>
     filter(
         .width == 0.95, # Credible intervals are narrow anyway so only show 95%
@@ -55,7 +43,7 @@ theme_survival_time_series = function() {
     )
 }
 
-surv_plot = tbl_posteriors |>
+surv_nat_plot = tbl_posteriors |>
       ggplot() +
       ggdist::geom_lineribbon(
         aes(time, S, ymin = S.lower, ymax = S.upper, fill = survival_prior, colour = survival_prior),
@@ -64,6 +52,20 @@ surv_plot = tbl_posteriors |>
     geom_line(aes(time, S), data = truth) +
     theme_survival_time_series() +
     ylab("S(t)")
+
+surv_log_plot = surv_nat_plot +
+    scale_y_log10(
+        minor_breaks = map(
+            -1:-3,
+            ~seq(from = 10^.x, to = 10^(.x+1), by = 10^.x)
+        ) |>
+        unlist()
+    ) +
+    coord_cartesian(c(0, 100), c(0.001, 1))
+
+surv_plot = surv_nat_plot + surv_log_plot + plot_layout(guides = "collect") &
+    theme(legend.position = "bottom")
+
 ggsave(
     filename = here::here("cis-perfect-testing/survival-results.png"),
     plot = surv_plot,

### 2023-09-28 incidence-prevalence.tex
AIDS intro and refs

@@
 HIV is a virus transmitted sexually or through blood, and causes the disease AIDS.
-AIDS has killed around 40 million people; at its peak, around 2 million people died from AIDS each year\todo{cite WHO}.
-In this context, backcalculation estimates the incidence of HIV infection from data on the number of AIDS diagnoses and the distribution of the time between infection and AIDS diagnosis~\autocite{brookmeyerBackcalculation}{brookmeyerMeasuring}.
+AIDS has killed around 40 million people; at its peak, around 2 million people died from AIDS each year~\autocite{unaids2023}.
+In this context, backcalculation estimates the incidence of HIV infection from data on the number of AIDS diagnoses and the distribution of the time between infection and AIDS diagnosis~\autocites{brookmeyerBackcalculation}{brookmeyerMeasuring}.
 \todo{explain how AIDS diagnosis is "prevalence" data (is it?)}
 
-Two processes are involved in backcalculation: the \emph{prevalence process} and the \emph{observation process}.
+Three processes are involved in backcalculation: the \emph{infection process}, the \emph{prevalence process} and the \emph{observation process}.
+The infection process describes how individuals become infected.
 The prevalence process describes the relationship between incidence and prevalence.

### 2023-09-28 incidence-prevalence.tex
tweaking

@@
 It was first developed in the context of HIV (human immunodeficiency virus) and AIDS (acquired immune deficiency syndrome)~\autocite{brookmeyerMethod}.
-HIV is a virus transmitted sexually or through blood, and causes the disease AIDS.
-AIDS has killed around 40 million people; at its peak, around 2 million people died from AIDS each year~\autocite{unaids2023}.
-In this context, backcalculation estimates the incidence of HIV infection from data on the number of AIDS diagnoses and the distribution of the time between infection and AIDS diagnosis~\autocites{brookmeyerBackcalculation}{brookmeyerMeasuring}.
+HIV is transmitted sexually or through blood.
+It causes the disease AIDS.
+AIDS has killed around 40 million people, peaking at 2 million annual deaths~\autocite{unaids2023}.
+In this context, backcalculation estimates the incidence of HIV infection from: data on the number of AIDS diagnoses, and the distribution of the time between infection and AIDS diagnosis~\autocites{brookmeyerBackcalculation}{brookmeyerMeasuring}.
 \todo{explain how AIDS diagnosis is "prevalence" data (is it?)}
@@
 Three processes are involved in backcalculation: the \emph{infection process}, the \emph{prevalence process} and the \emph{observation process}.
-The infection process describes how individuals become infected.
+The infection process determines incidence.
 The prevalence process describes the relationship between incidence and prevalence.
 The observation process describes the relationship between prevalence and the observed data.
+
+The infection process is a point process, where each infection is an event.
+The simplest plausible model is a time-inhomogenous Poisson process.
+That is, a Poisson process where the intensity varies over time.
+The target of inference can be either the process's intensity or the counts of the specific realisation that occurred.
+The realised counts are normally of greater interest because we care about the realised epidemic, not what may happen if there were future realisations of the same epidemic.
+
+Two features of a Poisson process can be violated for epidemics.
+Epidemics commonly exhibit overdispersion and have feedback loops.
+Overdispersion means that the variance in the number of infections in a time interval is higher than its mean.
+In a Poisson process, counts in a time interval have a Poisson distribution.
+Therefore, the mean and variance are equal.
+Feedback loops mean that the stochastically low or high numbers of infections, possibly due to overdispersion, affect the future number of infections.
+In a Poisson process, the intensity (mean number of events per unit time) is not influenced by the realisation of the process.
+Both of these features violate the assumptions of a Poisson process.
+
+A common cause of overdispersion is super-spreading events.
+\todo{ref}
+A super-spreading event is an event where many more people than expected are infected.
+Coronaviruses, such as SARS-CoV-2, have particularly high overdispersion.
+\todo{insert a SARS-CoV-2 superspreading example}
+Overdispersion could be incorporated by modelling the counts in a given time interval as a negative binomial, rather than Poisson, distribution.
+The mean of these distributions would be equal, but the negative binomial introduces an additional overdispersion parameter, commonly notated $k$.
+These distributions are related in that the negative binomial tends to Poisson as $k\to\infty$.
+Generally, $k<1$ indicates significant levels of overdispersion.
+Estimates for SARS-CoV-2 indicate $k \approx 0.1$, and can be increased by interventions such as lockdowns.
+However, the negative binomial distribution loses much of the mathematical convenience of a Poisson distribution.
+For a negative binomial process, the count's standard deviation scales on the order of the mean.
+Conversely, for a Poisson process, the standard deviation is on the order of the square root of the mean.
+A particular implication of this is that the coefficient of variation (the standard deviation divideed by the mean) tends to zero for a Poisson process, but tends to a constant for a negative binomial process.
+
+The independence assumption is violated because if the number of infections is stochastically high, this feeds back to the process intensity at future times.
+This can be seen simply in the event of a super-spreading event.
+If a super-spreading event occurs, then many more individuals than suggested by the process intensity will be infected.
+These additional infections will propagate the epidemic further, increasing the intensity at future times.
 

### 2023-09-28 SEIR.tex
Add files etc for SEIR

--- 2023-09-27 incidence-prevalence.tex
@@
 \begin{document}
-\chapter{Statistical background} \label{incidence-prevalence}
-%\todo[inline]{This chapter is pretty short. Not sure if this matters or it needs to merged with another chapter? For example, one background chapter split into epi/biology and stats.}
-
-This chapter explains incidence and prevalence, and the relationship between these quantities.
-These quantities have been of interest as long as disease has been studied, and estimating the prevalence from incidence is the fundamental motivation for this thesis.
-Their exact definition can vary throughout the literature, here I follow the definitions of the seminal paper \textcite{freemanPrevalence}.
-As I will explain, the duration of infection is the quantity that relates them.
-
-Timescale underlies all these quantities.
-Throughout this thesis, time will generally\todo{revisit phrasing once finished - do I actually use continuous time anywhere (probably in SEIR model)} be considered discrete in units of days.
-Most epidemiological data is only available at daily granularity or, when more granular data exists, is often unreliable.
-Furthermore, within-day patterns produce variations in transmission (\eg due to sleep) and data (\eg due to logistical considerations around collection); these variations are not of interest in most contexts.
-
-\section{Definitions}
-
-\emph{Incidence} at time $t$, $Z_t$, is the number of people in a given population who become infected at time $t$ (I use $Z_t$ rather than $I_t$ to avoid confusion with the convention that $i$ is an indexing variable).
-Incidence expressed as a proportion of the population, $Z_t/N$ where $N$ is the number of people in the population, is the \emph{incidence rate}.
-As discussed in \cref{E-intro}\todo{might want to point to a specific section}, infection events are not directly observable and must be inferred from other data.
-Despite this issue, it remains the single most important quantity for informing the response to an epidemic or pandemic.
-
-\emph{Prevalence} of an infectious disease at time $t$, $P_t$, is the number of people in a given population currently infected at time $t$, assumed constant over the short period $t$.
-Each of these individuals is known as a \emph{prevalent} individual.
-The \emph{prevalence proportion} is the same quantity expressed as a proportion of the population.
-For SARS-CoV-2, there are several definitions of infected.
-An infected individual could mean: one that is infectious; one that is symptomatic; or one that, if they were tested, would return a positive PCR test if tested (discussed in\todo{ref appropriate section}).
-In this thesis, prevalence is defined as the number of people in the population who would return a positive PCR test (if everyone was tested) at time $t$.
-
-\section{Backcalculation}
-
-\emph{Backcalculation} is the procedure for estimating incidence based on a measurement of prevalence.
-It was first developed in the context of HIV (human immunodeficiency virus) and AIDS (acquired immune deficiency syndrome)~\autocite{brookmeyerMethod}.
-HIV is a virus transmitted sexually or through blood, and causes the disease AIDS.
-AIDS has killed around 40 million people; at its peak, around 2 million people died from AIDS each year\todo{cite WHO}.
-In this context, backcalculation estimates the incidence of HIV infection from data on the number of AIDS diagnoses and the distribution of the time between infection and AIDS diagnosis~\autocite{brookmeyerBackcalculation}{brookmeyerMeasuring}.
-\todo{explain how AIDS diagnosis is "prevalence" data (is it?)}
-
-Two processes are involved in backcalculation: the \emph{prevalence process} and the \emph{observation process}.
-The prevalence process describes the relationship between incidence and prevalence.
-The observation process describes the relationship between prevalence and the observed data.
-
-In \emph{deterministic backcalculation}, the observed prevalence, $x_t/n_t$, is assumed to be equal to $P_t/N$.
-However, this is often a poor approximation.
-\emph{Statistical backcalculation} retains the sampling distribution of $x_t$ (or an approximation of it, such as a Poisson).
-
-Define the first day of an infection episode in individual $i$ as $B_i$, and assume that the probability of $i$ being infected multiple times within the period of interest is negligible.
-The time between being infected and first being detectable is short (see\todo{ref relevant part}), and therefore I often assume that $B_i$ is the same as the time of infection.
-I keep this assumption for the remainder of this chapter.
-
-Backcalculation makes use of this relationship to estimate incidence from prevalence, assuming that the duration is known.
-
-Deterministic backcalculation assumes that the variance in the population is negligible, and therefore that the prevalence is equal to the mean prevalence.
-This is justified because the variance due to sampling is much larger than the variance due to the true prevalence.
-
-\todo[inline]{Explain backcalculation in HIV/AIDS context, and how it relates to what is above.}
-
-\section{Prevalence process}
-
-The relationship between incidence and prevalence can be derived by considering the probability that an individual is prevalent at any time following an infection.
-The period the individual is prevalent for is the \emph{infection episode}, which I will index with $i$, associated with that infection event.
-Denote the time that the infection episode begins as $B_i$ and the time it ends as $E_i \geq B_i$.
-The individual is prevalent from $B_i$ to $E_i$ inclusive.
-The duration of the infection episode, the number of days for which they are prevalent, is $D_i = E_i - B_i + 1$.
-In this thesis, it is assumed that the $D_i$ for each infection episode is independent and $\prob(D_i = d) = f_d$ for some pmf $f$ over $1, 2, \dots, d_\text{max}$.
-
-From this, we can derive the relationship between incidence and prevalence.
-Denote by $R_{t,t'}$ be the number of infection episodes that begin at time $t$ and end at time $t'$.
-For each infection beginning at $t$, the probability that it ends at $t'$ is $f_d(t' - t + 1)$.
-Since these durations are assumed independent, the number of episodes of each duration is multinomially distributed.
-Therefore:
-\begin{align}
-\begin{bmatrix}
-  R_{t,t} \\ R_{t,t+1} \\ \vdots \\ R_{t,t+d_\text{max}-1}
-\end{bmatrix} \mid Z_t
-\sim \MN \left(
-  Z_t, 
-  \begin{bmatrix}
-    f_d(1) \\ f_d(2) \\ \vdots \\ f_d(d_\text{max})
-  \end{bmatrix}
-\right).
-\end{align}
-Now, denote by $P_{t_z,t_p}$ the number of episodes beginning at time $t_z$ that are prevalent at time $t_p$ (where $t' \geq t$).
-This is $Z_t$ minus those that have recovered:
-\begin{align}
-    P_{t_z,t_p} = \begin{cases}
-      0 &t_p < t_z\\
-      Z_t - \sum_{i=0}^{t_p-t_z-1} R_{t_z,t_z+i} &t_z \leq t_p < t_z + d_\text{max}\\
-      0 &t_p \geq t_z + d_\text{max}.
-  \end{cases} \label{incprev:eq:Ptt-to-R}
-\end{align}
-Finally, the prevalence at $t$ can be expressed as the total number of people prevalent at time $t$, regardeless of when they were infected:
-\begin{align}
-  P_t
-  &= \sum_{t_z=\infty}^\infty P_{t_z,t} \\
-  &= \sum_{i=0}^{d_\text{max}-1} P_{t-i,t} \label{incprev:eq:Pt-to-Ptt}.
-\end{align}
-
-In general, all these random variables will be dependent for an infectious disease.
-This is because the probability of being infected at time $t$ depends on the number of infectious individuals at time $t$, which itself depends on the number of individuals infected before time $t$.
-However, if we condition on the vector $\vec{Z} = (Z_1, Z_2, \dots)^T$, then we will induce some independence.
-In particular, because the duration of individuals is independent, $R_{t_1,t_2}$ and $R_{t_1',t_2'}$ will be independent if $t_1 \neq t_1'$; importantly, this holds even if $t_2 = t_2'$.
-
-A quantity that will be important is $\E(P_t \mid \vec{Z})$, which can be derived from its constituent parts.
-First, the properties of a multinomial distribution give the marginal distribution:
-\begin{align}
-  \sum_{i=0}^{t'} R_{t,t+i} \mid \vec{Z} &\sim \text{Binomial}(Z_t, F_d(t'+1)).
-\end{align}
-where $F_d(t) = \sum_{i=1}^t f_d(t) = \prob(D \leq t)$ is the CDF of $D$.
-Therefore:
-\begin{align}
-  \E(\sum_{i=0}^{t'} R_{t,t+i} \mid \vec{Z}) = Z_t F_d(t'+1) \label{incprev:eq:EsumRt}
-\end{align}
-
-Now consider $P_t$:
-\begin{align}
-\E(P_t \mid \vec{Z})
-  &= \E\left(\sum_{i=1}^{d_\text{max}-1} P_{t-i,t} \mid \vec{Z} \right) &\text{by \cref{incprev:eq:Pt-to-Ptt}} \\
-  &= \E\left(\sum_{i=1}^{d_\text{max}-1} \left( Z_{t-i} - \sum_{j=0}^{i-1} R_{t-i,t-i+j} \right) \mid \vec{Z} \right) &\text{by \cref{incprev:eq:Ptt-to-R}}\\
-  &= \sum_{i=1}^{d_\text{max}-1} \left( Z_{t-i} - \sum_{j=0}^{i-1} \E( R_{t-i,t-i+j} \mid \vec{Z}) \right) \\
-  &= \sum_{i=1}^{d_\text{max}-1} \left( Z_{t-i} - Z_{t-i} F(i) \right) &\text{by \cref{incprev:eq:EsumRt}}\\
-  &= \sum_{i=1}^{d_\text{max}-1} Z_{t-i} (1 - F(i)) \\
-  &= \sum_{i=1}^{d_\text{max}-1} Z_{t-i} S(i+1)
-\end{align}
-where $S(i+1) = 1-F(i) = \prob(D \geq i+1)$ is the \emph{survival function} of $D$ (see \cref{inprev:sec:duration}).
-
-% The probability of being positive can be modelled in different ways depending on what assumptions are most reasonable.
-% The most important assumption from a statistical perspective is the whether the probability of testing positive on a given day is independent conditional on the infection time.
-
-% \todo[inline]{Is the discussion of this model just a distraction?}
-% Assuming that the probability of testing positive depends only on the infection time implies no individual variation.
-% This model is formalised as $Pr(Z_i(t) = 1 \mid B_i, Z_i(1), Z_i(2) \dots) = Pr(Z_i(t) \mid B_i) = p_{t-B_i}$, that is the probability of individual $i$ testing positive $t$ days after infection depends only on the infection start time and no other quantities.
-% A slight relaxation of this model is introducing covariates, allowing some variation based on an individual's characteristics.
-
-% The model with the most dependence between times, while remaining realistic, is a multi-state model.
-% Therefore, if an individual has tested negative before the present time $t$ but after $B_i$ then they will test negative at time $t$.
-% The complexity of the transitions between positive and negative can be very complex.
-% I consider a slight extension to allow for the possibility of false negatives, meaning a negative test between $B_i$ and $E_i$, and that this probability may change.
-% This leads to the following model.
-% \begin{align}
-%   \prob(Z_i(t) = 1 \mid B_i, \leq E_i) &= \begin{cases}
-%     \psens(t - B_i) &B_i \leq t \leq E_i\\
-%     0 &\text{otherwise}
-%   \end{cases}\\
-%   \prob(E_i = t \mid B_i) = f_i(B_i)
-% \end{align}
-% Conditional on both $B_i$ and $E_i$, then $Z_i(t)$ and $Z_i(t')$ are independent for $t \neq t'$.
-% However, unlike the previous model, conditioning only on $B_i$ is not sufficient for $Z_i(t)$ and $Z_i(t')$ to be independent because each $Z_i(t)$ provides information on $E_i(t)$ (in particular, $Z_i(t) = 1$ implies $E_i > t$).
-% For SARS-CoV-2, the latter model is more appropriate, although, as we shall see, the assumption that $\psens$ is independent of $t$ is violated.
-
-\todo[inline]{Need some conclusion text for this section}
-
-\section{Observation process}
-
-% A \emph{prevalence survey} is a survey that tests a sample of the population for the presence of an infection.
-% Series of prevalence surveys are the form of measuring prevalence that this thesis is concerned with.
-% This thesis aims to infer the incidence rate from prevalence surveys.
-% I use Bayesian inference, therefore the data contributes through the likelihood.
-
-Prevalence surveys, as described in\todo{insert ref}, sample individuals at random from the population of interest and test if they are prevalence.
-Therefore, we view the number of positive tests on day $t$ as a random variable $X_t$ based on a sample of $n_t$ individuals.
-The $X_t$s are conditionally independent, given the population prevalences at each time, $\vec{P}_t$. 
-%For now, I assume that the sample is chosen uniformly at random from the population of interest.
-%Assume that the surveys are independent of each other and $\vec{Z}$, conditional on the population prevalence at that time.
-%The population prevalence is a latent quantity.
-From these assumptions, $X_t \mid \vec{P} \sim \Bin(n_t, P_t / N)$.
-I base inference on the likelihood, $p(\vec{x} \mid \vec{Z}) = \int p(\vec{x} \mid \vec{P}) p(\vec{P} \mid \vec{Z}) d\vec{P} = \int \prod_t p(x_t \mid P_t) p(\vec{P} \mid \vec{Z}) d\vec{P}$, where $\vec{x} = (x_1, x_2, \dots)^T$.
-
-For most prevalence surveys, the uncertainty arising from the distribution $\vec{P} \mid \vec{Z}$ is negligible.
-This can be seen by considering the variance in the likelihood.
-Assume $x_t \sim \Bin(n_t, P_t / N)$, where $N$ is the size of the population.
-Therefore, by the law of total variance:
-\begin{align}
-  \var\left( x_t \mid \vec{n}, \vec{Z} \right)
-    &= \var\left[\E\left( x_t \mid \vec{n}, \vec{Z}, \vec{P} \right) \mid \vec{n}, \vec{Z} \right] + \E\left[\var\left( x_t \mid \vec{n}, \vec{I}, \vec{P} \right) \mid \vec{n}, \vec{Z} \right] \\
-    &= \var\left( n_t \frac{P_t}{N} \mid \vec{n}, \vec{Z} \right) + \E\left( n_t \frac{P_t}{N} \left(1 - \frac{P_t}{N} \right) \mid \vec{n}, \vec{Z} \right) \\
-    &= \left( \frac{n}{N} \right)^2 \var\left(P_t \mid \vec{n}, \vec{Z} \right) + \frac{n_t}{N} \E\left(P_t \mid \vec{n}, \vec{Z} \right)  - \frac{n_t}{N^2} \E\left(P_t^2 \mid \vec{n}, \vec{Z} \right).
-\end{align}
-For large populations and reasonable sample sizes, $n_t << N$ and $\E(P_t) \approx \var(P_t)$, the first term is negligible.
-Hence, assuming that $\var(\vec{P}) = \matr{0}$ is a reasonable approximation.
-This is equivalent to assuming that $\vec{P}$ is a deterministic function of $\vec{Z}$, $\vec{P} = \E( \vec{P} \mid \vec{Z})$.
-
-\section{Estimating the duration} \label{inprev:sec:duration}
-
-Robust, unbiased prevalence estimates already exist (see\todo{ref prevalence estimates}), and incidence estimates are the most useful (see\todo{ref intro with importance of incidence}).
-Therefore, a large portion of this thesis concerns itself with estimating the duration of PCR positivity.
-
-The distribution of durations can be expressed in several equivalent ways.
-Let $D_i$ be the random variable for the duration of episode $i$, then the distribution of interest is that of $D_i$.
-I will assume that each $D_i$ is an independent and identically distributed realisation of a random variable $D$; extending the methodology to account for covariates is left to further work.
-For an infection in discrete time, the duration is a non-negative integer, \ie $D \in \{1, 2, \dots\}$.
-The first way of expressing the distribution is as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=0}^d f_{D}(i)$.
-The \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$.
-The \emph{hazard} at time $t$ is the probability of recovering at time $t$ conditional on having survived to time $t$: $\lambda(t) = \prob(D = t \mid D \geq t) = f_{D}(d) / S(d)$.
-Throughout this thesis, the survival function will be the primary quantity of interest, and it will normally be parameterised in terms of the hazard; these two quantities are related by $S(t) = \prod_{i=1}^{t-1} (1 - \lambda(i))$.
-
-Two approaches are commonly used to estimate durations\todo{cite Sweeney}: either modelling an underlying biomarker or modelling the duration directly.
-For SARS-CoV-2, both approaches would be based on the results of PCR testing.
-Modelling a biomarker would mean modelling the viral load, measured as a Ct value.
-Modelling the duration would consider only the binary result at each test (positive or negative); in this thesis a \emph{survival analysis} framework is adopted.
-Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
-Other related approaches, such as a multi-state model, could also be considered (multi-state models can be viewed as a generalisation of survival analysis~\autocite{jacksonMSM}).
-Therefore, the biomarker approach considers more information per observation, however, a model for the biomarker is required which generally requires stronger assumptions.
-Modelling a biomarker uses more information per observation, but requires stronger assumptions.
-
-\todo[inline]{Comment on previous estimates of duration for SARS-CoV-2, but not sure where exactly this goes...}
-
+\chapter{Mechanistic modelling of incidence and transmission}
 \end{document}

### 2023-10-02 cis-perfect-testing.tex
Finish chapter rewrite/edit

@@
 The estimation of incidence from prevalence is sensitive to the tail of the duration distribution.
-In \cref{E-ATACCC} I estimated the duration distribution from the ATACCC dataset, however, this study suffered from a small sample size and only 20 days of follow-up.
-Therefore, the estimation of the tail of the duration distribution is driven by model assumptions.
-This chapter develops methods to estimate the distribution using CIS data, which has less frequent sampling but a large sample size and long follow-up.
+In \cref{E-ATACCC} I estimated the duration distribution from the ATACCC dataset.
+However, this study suffered from a small sample size and only 20 days of follow-up.
+Therefore, model assumptions determined the size of the tail estimated.
+In this chapter, I develop a method to estimate the distribution using CIS data. 
+CIS sampling is less frequent, but with a large sample size and long follow-up.
 
 I take a Bayesian survival analysis approach to model the coarse sampling of the CIS (see \cref{perf-test:sec:problem}).
-Flat priors can be unintentionally very informative, and hence uninformative priors should be used.
+Flat priors are unintentionally very informative.
+Hence, at least weakly informative priors should be used.
 Of particular interest is a strongly informative prior to combine the information in the CIS with the results of \cref{E-ATACCC} (see \cref{perf-test:sec:parameters-priors}).
-Through a simulation study, I show that these methods perform well, and in this context are not sensitive to model or whether the prior is vague or strongly informative (see \cref{perf-test:sec:simulation-study}).
-However, application in reality must account for false negatives, the subject of \cref{E-imperf-test}.
+I conduct a simulation study to demonstrate this method's performance.
+It succesfully recovers the truth when using reasonable priors (see \cref{perf-test:sec:simulation-study}).
+In reality, the method must account for false negatives, discussed in \cref{E-imperf-test}.
 
@@
 
-The CIS infrequently tests each individual, with tests occurring either weekly or monthly (see\todo{refer to section with CIS intro}).
-The infrequent testing complicates analysis of the data in two ways: infection episodes can go undetected and the infection episode endpoints (start and end times) are known imprecisely.
-This section explains these issues fully, and the remainder of the chapter develops a method to model them.
+The CIS tests each individual infrequently, either weekly or monthly (see\todo{refer to section with CIS intro}).
+The infrequent testing complicates analysis of the data in two ways.
+First, infection episodes can go undetected.
+Second, the infection episode start and end times are known imprecisely.
+This section explains these issues.
+The rest of the chapter develops a method to model them.
 
-The first complication in the analysis is that infection episodes can be undetected, meaning no there are no positive tests associated with them.
-This happens when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
-For these undetected episodes, there is no information on them episode, meaning that we do not even know how many of them occurred.
-It is important to account for undetected episodes because they are systematically of shorter duration than detected episodes.
+The first complication in the analysis is that infection episodes can be undetected.
+This means there are no positive tests associated with them.
+Undetected episodes occur when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
+For undetected episodes, there is no information on them episode; even the number that occurred is unknown.
+Undetected episodes are important because they are shorter than detected episodes.
 To see why this is the case, consider the undetected infection in the bottom of \cref{perf-test:fig:truncation}.
 The infection episode depicted begins at time 33 and has a 10-day duration.
-If the duration was 24 days (or longer), the individual would still have been detectable at their next test, and therefore would have been detected.
-Undetected infections are a large problem in CIS because the interval between tests is often 28 days (see\todo{ref section on the survey design}), large relative to the duration of an infection episode.
+If the duration was 24 days or more, the individual would still be detectable at their next test and detected then.
+CIS has many undetected infections. This is because frequency of testing (up to 28 days, see\todo{ref section on the survey design}) is longer than most infections (median around 15 days).
 \begin{figure}
@@
 
-The second complication is that the beginning and end times of detected infection episodes are known imprecisely.
+The second complication is that the exact beginning and end times of detected infection episodes are unknown.
 The imprecision in the beginning of the infection episode arises because the only observation that an episode has started is that an individual was negative at one test, on day $l_i^{(b)}-1$ (the minus one is because the first day that the infection could have begun is the day after the negative test), and positive at the next test, on day $r_i^{(b)}$.
@@
 
-I will base my analysis on the 4,800\todo{check number} of episodes that were first detected between 10th Oct 2020 and 6th Dec 2020 inclusive and have negatives bounding the start and end time of the episode.
-That is, the individual in which the infection episode occurred tested negative at least once in the survey prior to the start of their infection episode, and they tested negative at least once after their infection episode ended.
-Without a negative before the episode there is little information on the episode's length because it's start time could have occurred arbitrarily long ago.
-The only episodes without a negative following the end of the episode are those in individuals lost to follow-up before their episode ended, which is extremely rare.
-This filtering could bias the results of the study but will be considered in the same way as undetected episodes to avoid this.
+I base my analysis on the 4,800\todo{check number} episodes that were first detected between 10th Oct 2020 and 6th Dec 2020 inclusive with negatives bounding the start and end time of the episode.
+That is, the individual tested negative before and after the episode.
+Without a negative before the episode there is no lower bound on the episode's start time.
+It could be arbitrarily long and there is little information on its length.
+A negative test follows each episode unless the individual was lost to follow-up before their episode ended.
+This is rare.
+Episodes removed in this step are incorporated in the adjustment for undetected episodes.
 
-In the remainder of this and the following chapter, I use survival analysis rather than the biomarker approach taken in \cref{E-ATACCC} for computational efficiency and data limitations arising from the study design.
-In the biomarker approach, computational efficiency is an issue because the number of random effects scales with the number of detected infection episodes.
-There are a large number of detected episodes in CIS; in the period I will consider, around 50 times more than in ATACCC.
-Therefore, the biomarker approach would be computationally challenging with the large sample size in CIS.
-The computational challenge is compounded by the data being stored in a TRE with the associated limited access to computational power and software.
-The study design of CIS, with the infrequent sampling in the CIS study design, means that most detected episodes have only a single positive test.
-Hence, there is no information on the viral load trajectory required to identify the random effects in the model used in \cref{E-ATACCC} or similar approaches.
-While the biomarker approach does use more information per sample by taking into account the viral load as well as the binary test result (\ie whether the result is positive or negative), however, the available information is unlikely to be sufficient to overcome these problems.
+In the rest of this and the following chapter, I use survival analysis rather than the biomarker approach taken in \cref{E-ATACCC}.
+This choice is motivated by computational efficiency and limitations of the CIS's study design.
+The biomarker approach is computational challenging because the number of random effects scales with the number of detected infection episodes.
+There are many detected episodes in CIS; in the period I will consider, around 50 times more than in ATACCC.
+Therefore, the biomarker approach would be computationally challenging.
+The computational challenge is compounded by the data being stored in a TRE, which has limited computational power and software.
+The issue with random effects is made worse by the CIS's study design.
+Infrequent testing means that most detected episodes have only a single positive test.
+Hence, there is no information on the viral load trajectory and the random effects are unidentifiable.
 
@@
 In survival analysis, the event at the start of the time of interest is the \emph{initiating event}, and the event at the end of the time of interest is the \emph{terminating event}.
-It is broadly applicable to many domains, such as the time-to-failure for a mechanical system or the effect of a treatment on time spent in hospital.
-In the context of this thesis, the distribution to estimate is the duration of positivity.
-Therefore, the initiating event is when the duration starts; and the terminating event is when the duration ends.
+It is used in many areas, including how long a machine lasts before breaking down or how long a patient spends in hospital after a treatment.
+Here, I am estimating the distribution of the duration of positivity.
+The initiating event is the episode's start; the terminating event is the episode's end.
 
-Undetected events where we have information on missed events is unusual within the survival analysis literature.
-Far more common is that there is no data on these events, a situation known as \emph{truncation}.
+Undetected events where there is no information on these events is known as \emph{truncation}.
 A situation that is particularly relevant is \emph{left truncation}, when only events lasting a minimum amount of time are observed.
-A subtlety that makes the CIS different from standard left truncation is that information on individuals in which undetected infections occurred is available because they were enrolled in the study.
+There is a subtlety that makes the CIS different from standard left truncation.
+In CIS, individuals in which undetected infections occurred are enrolled in the study.
+Therefore, some data has been collected about them.
 In standard left truncation, nothing is known about the missed events.
-As the information is not being incorporated here, standard left truncation methods remain applicable, and they can be modified to incorporate the additional information if desired.
+I incorporate the additional information into the analysis.
+I refer to these events as ``undetected'' to emphasise this difference.
 
 The time of the initiating and/or terminating events are not known exactly.
-This is known as \emph{censord} in the survival analysis literature.
+This is known as \emph{censored} in the survival analysis literature.
 Censored events are present in almost all survival analyses.
-An event is \emph{left censored} if it is known to occur before a certain time, \emph{right censored} if it is known to occur after a certain time, and \emph{interval censored} if it is known to occur within an interval.
+An event is \emph{left censored} if it is known to occur before a certain time.
+An event is \emph{right censored} if it is known to occur after a certain time.
+An event is \emph{interval censored} if it is known to occur within an interval.
 %Left and right censoring can be viewed as special cases of interval censoring where the lower bound is negative infinity or the upper bound is positive infinity respectively.
-Interval censoring arises in CIS because we only observe the change from being not detectable (\ie: negative) at one test, to detectable (\ie: positive) at the following test and vice versa (see \cref{perf-test:fig:double-interval-censor}).
-If both the initiating and terminating events are censored, then the event is \emph{doubly censored}.
-Doubly censored data is more challenging to handle because the initiating event time and duration must be jointly modelled~\autocite[and references therein]{liSemiparametric}.
+As explained in \cref{perf-test:sec:problem}, the initiating and terminating events are only bounded, and hence interval censored.
+The data is \emph{doubly censored} because both the initiating and terminating events are censored.
+To handle doubly censored data, the times need to be jointly modelled~\autocite[and references therein]{liSemiparametric}.
 In CIS data both the initiating and the terminating events are interval censored.
-Therefore, the CIS data is \emph{double interval censored}; approaches for double interval censoring are reviewed by \textcite{sunAnalysis,bogaertsSurvival}.
+Therefore, the CIS data is \emph{double interval censored}.
+\Textcite{sunAnalysis,bogaertsSurvival} review methods for double interval censored data.
 
-Few studies consider the combination of both double interval censoring and undetected or truncated events
+Few studies look at the combination of both double interval censoring and undetected (or truncated) events
 This is especially true within the human biostatistical literature.
-While this literature does contain theoretical frameworks which include the double censored and truncated case~\autocites{turnbullEmpirical}{dempsterMaximum}, they have been applied when the terminating event was either uncensored or right censored~\autocite{sunEmpirical,bacchettiNonparametric}.
-The approach of these studies \autocite[and elsewhere, e.g.:][]{shenNonparametric} was to take a conditional likelihood approach, conditioning on both the truncation (a selection effect) and the period in which the initiating event is known to occur.
-They justify conditioning on the time of the initiating event by arguing that this information provides little information on the distribution being estimated~\citePersonalComms{Nick Jewell}.
-However, this is not the case here.
+The literature does include theoretical frameworks which include the double censored and truncated case~\autocites{turnbullEmpirical}{dempsterMaximum}.
+The frameworks have only been applied when the terminating event was either uncensored or right censored~\autocite{sunEmpirical,bacchettiNonparametric}.
+These studies \autocite[and elsewhere, e.g.:][]{shenNonparametric} take a conditional likelihood approach.
+They condition on the truncation (a selection effect) and the period where the initiating event occurs.
+They condition on the initiating event time because it provides negligible information on the distribution of interest~\citePersonalComms{Nick Jewell}.
+This is not the case for the CIS.
 In CIS, infection episodes are more likely to be detected when an individual is in the weekly testing phase.
-This phase is not representative of the overall study: infections are more likely to be detected during this phase.
-Therefore, estimating the probability of being detected conditional on being infected in this period (which would be implictly done in this previous method)  will overestimate the probability of being detected.
+CIS is more likely to detect infection episodes during the weekly testing phase.
+Therefore, the probability of an infection episode being detected conditional on being infected is higher than the overall probability of being detected.\todo{this is a bit unclear, try and rephrase?}
 
-Censored and truncated data frequently occur when estimating the survival time of bird nests, with the censoring taking a variety of forms~\autocite{heiseyABCs}.
-In this context, \textcite{heiseyModelling} develop a general framework allowing for arbitrary censoring patterns and double interval censoring.
-In their application, the visit times are global (all nests could be detected on the same days), while in CIS each individual has their own visit schedule; with global visit times the probability of being detected is the same for all participants in the study, simplifying the likelihood.
-I use this framework as the basis of the model I derive in \cref{perf-test:sec:model}.
+Censored and truncated data are normal for estimation of the survival time of bird and insect nests~\autocite{heiseyABCs}.
+\textcite{heiseyModelling} developed a framework for handling this situation.
+The framework is general, allowing for arbitrary censoring patterns and double interval censoring.
+They apply it to a situation the visit times are global (all nests could be detected on the same days), simplifying the likelihood.
+I adapt this framework for CIS, when each individual has their own visit schedule (see section \cref{perf-test:sec:model}).
 
-Inference will be performed within the Bayesian paradigm for two reasons.
-First, the standard frequentist methods for estimating the parameter's variance and covariance are poorly developed in this area.
-Commonly used methods are based on the Fisher information matrix or resampling, but neither have theoretical justification~\autocite[230]{sunAnalysis}\todo{Try and understand why this is the case}.
-Second, a prior naturally allows the inclusion information from previous studies, such as that from \cref{E-ATACCC}, and/or a belief that the hazard changes smoothly in a way that can be encoded in a prior (these options are explored in \cref{perf-test:sec:parameters-priors}).
-These prior structures may be needed to combat the lack of information in the data due to the undetected episodes~\autocite{caoBias}.
-They can also be viewed as combining the information from the two analyses (see \cref{perf-test:sec:informative-priors}).
+I use Bayesian inference for two reasons.
+First, frequentist methods for estimating the parameter's variance and covariance are poorly developed.
+Commonly used methods are based on the Fisher information matrix or resampling.
+Neither have theoretical justification~\autocite[230]{sunAnalysis}\todo{Try and understand why this is the case}.
+Second, a prior can include information from previous studies, such as from \cref{E-ATACCC}.
+Alternatively, the prior can favour a smoothly changing hazard.
+I explore priors in section 1.4.
+These prior structures overcome issues caused by the undetected episodes ~\autocite{caoBias}.
 
-Bayesian methods have previously been applied to nest survival settings~\autocites{heBayesiana}{heBayesian}{caoModeling}, augmenting the data with the unobserved times of infection.
+Bayesian methods been used for nest survival before~\autocites{heBayesiana}{heBayesian}{caoModeling}, augmenting the data with the unobserved times of infection.
 The augmentation creates a parameter per detected infection.
-Therefore, data augmentation applied to the 4000 detected episodes in the main analysis of \cref{E-imperf-test} greatly increases the dimensionality of the problem, making it computationally infeasible.
-I develop a Bayesian framework where there is no explicit data augmentation, allowing the method to scale to the large number of episodes present in CIS.
+This increases the dimensionality of the problem greatly, making it computationally infeasible here.
+I develop a novel Bayesian framework without explicit data augmentation.
+This method scales to the large number of episodes present in CIS.
 
 \section{Modelling the duration}\label{perf-test:sec:model}
-
-\subsection{Likelihood}\label{perf-test:sec:likelihood}
 
@@
 
-Integrating over the number of missed episodes is analytically tractable making inference tractable within modern, high-performance Bayesian software such as Stan.
+Integrating over the number of missed episodes is analytically tractable.
+This makes inference tractable within modern, high-performance Bayesian software such as Stan.
 Let the total number of infection episodes in the cohort be $\ntot = \ndet + \nnodet$ where $\ndet$ is the number of detected infections and $\nnodet$ is the number of undetected infections.
-There are $\Ncis$ individuals in the cohort.
+Let $\Ncis$ be the number of individuals in the cohort.
 Without loss of generality, let the individuals with a detected episode be indexed by $i = 1, \dots, \ndet$ and the individuals without a detected episode be indexed by $i = \ndet + 1, \dots, \Ncis$.
@@
 
-All the relevant information about episode $j$ can be fully characterised by the (unobserved) triplet $(b_j, e_j, \sched_j)$ where $b_j$ and $e_j$ are the times the episode being and end respectively and $\sched_j$ is the schedule of tests for the individual in which the episode occurs.
-If $\sched_j$ contains at least one test between $b_i$ and $e_i$ (inclusive), then the episode is detected, otherwise it is undetected.
-The triplet belongs to a state space where $b_j$ and $e_j$ span possible infection and recovery times, and $\sched_j$ is one of the observed test schedules (assuming other test schedules occur with probability zero).
+All the relevant information about episode $j$ can be fully characterised by the (unobserved) triplet $(b_j, e_j, \sched_j)$ where $b_j$ and $e_j$ are the times the episode being and end respectively and $\sched_j$ is the set of test times for the individual in which the episode occurs.
+If $\sched_j$ contains at least one time between $b_i$ and $e_i$ (inclusive), one time prior to $b_i$, and one time after $e_i$, then the episode is detected, otherwise it is undetected.
+The triplet belongs to a state space.
+$b_j$ and $e_j$ span possible infection and recovery times.
+$\sched_j$ is one of the observed test schedules, with any other test schedule assumed to occur with probability zero.
 
-To relate this to the data, define three sets which partition the space of possible infection and recovery times (\ie: are strict subsets of $T \times T$), for individual $i$ (graphically shown in Figure \ref{perf-test:fig:partitionSpace}).
+To relate the state space to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie: are strict subsets of $T \times T$), for any individual based on their test times.
+Figure \ref{perf-test:fig:partitionSpace} shows this graphically.
 Each episode, whether detected or not, must fall into one of these three classes with a probability that is a function of the model parameters, $\theta$.
@@
 I assume the individual in which each episode occurs is iid uniform across the cohort.
-Therefore, conditional on $\ntot$, the number of episodes in each class (that is, the counts $n_{ia}, n_{iu}, n_{i\bar{a}}$) are distributed multinomially.
+Therefore, conditional on $\ntot$, the number of episodes in each class, $(n_{ia}, n_{iu}, n_{i\bar{a}})^T$, are distributed multinomially.
 The above probabilities ($p_{iu}$, $p_{ia}$, and $p_{i\bar{a}}$) are conditional on the episode occurring in individual $i$.
@@
 
-\todo{better phrasing needed for this paragraph, mauybe try email to Shaun?}
+\todo{Not sure if the followin paragraph is understandable}
 In forming the probability $p_{ia}$, the data is used.
@@
 
-The posterior we are interested in is $p(\theta \mid \na)$ where $\theta$ are the parameters governing the survival distribution.
-The probabilities $\pcis$ are functions of the parameters $\theta$.
+The posterior we are interested in is $p(\theta \mid \na)$.
+The posterior density can be written as follows.
 \begin{align}
@@
 \end{align}
+The rest of this section derives expressions for each $p_{ia}$, $p_{u}$ and the sum.
 
-In the remainder of the section, expressions for each $p_{ia}$, $p_{iu}$, and analytical solutions for the sum in \cref{perf-test:eq:full-posterior} are derived in terms of the observed data, under different priors.
-
-I derive an analytical solution to $p(\ntot \mid \theta) = \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under a negative binomial prior for $\ntot$.
-This prior is $\NegBin(\mu, r)$ using the mean/dispersion parameterisation of the negative binomial.
+Now I derive an analytical solution to $p(\ntot \mid \theta) = \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under the prior $\ntot \dist \NegBin(\mu, r)$ (using the mean/dispersion parameterisation of the negative binomial).
 Equivalently:
@@
 
-Substituting this into \cref{perf-test:eq:full-posterior} gives the following expression for the posterior.
+Substituting this into \cref{perf-test:eq:full-posterior} gives:
 \begin{align}
@@
 
-$\ntot$ is generally a nuisance parameter, however, its posterior distribution can be useful for diagnostic purposes (\eg: see X\todo{ref where I use this in the next chapter}).
-I reconstruct the posterior using a posterior sample of $\theta$.
-For each posterior sample of $\theta$, I sample from the full conditional of $\ntot$.
-This gives the joint posterior of $\ntot$ and $\theta$.
+$\ntot$ is generally a nuisance parameter, however, its posterior distribution can be useful for diagnostic purposes (as in X\todo{ref where I use this in the next chapter}).
+Its posterior can be reconstructed using a posterior sample of $\theta$.
+For each posterior sample of $\theta$, sample from the full conditional of $\ntot$ to give the joint posterior of $\ntot$ and $\theta$.
 The full conditional of $\ntot$ is given by:
@@
 \end{align}
-Comparing this expression to the pmf of a negative binomial, we find that $\ntot - \ndet$ is distributed negative binomial with size parameter $r+\ndet$ and probability parameter $(r + \mu(1 - \pnodet))/(r + \mu)$.
-The mean of this distribution is $(r+\ndet)\mu \pnodet/r+\mu(1-\pnodet)$.
+Comparing this expression to the pmf of a negative binomial, we find that $\nnodet = \ntot - \ndet$ is distributed negative binomial with size parameter $r+\ndet$ and probability parameter $(r + \mu(1 - \pnodet))/(r + \mu)$.
+%The mean of this distribution is $(r+\ndet)\mu \pnodet/(r+\mu(1-\pnodet))$.
 
-Next, I derive $p_{ia}$ and $p_{iu}$.
-
-By definition (see~\cpageref{perf-test:def:admissible}), we have
+Next, I derive $p_{ia}$.
+By definition (see~\cpageref{perf-test:def:admissible}), we have:
 \begin{align}
-p_{ia} &= \prob((b, e) \in \alpha_i) \\
+p_{ia} &= \prob((b, e) \in \alpha_i \mid \theta) \\
 \alpha_i &= \{ (b, e) : l_i^{(b)} \leq b \leq r_i^{(b)} \wedge l_i^{(e)} \leq e \leq r_i^{(e)}\}.
-\intertext{Hence:}
+\intertext{Suppressing the conditioning on $\theta$, this gives:}
 p_{ia}
@@
 \end{align}
-\todo[inline]{Fill in references}
 under the assumption of uniform probability of infection time.
 
-Next, I derive $p_{iu} = \prob((B_i, E_i) \in \Omega^C_i)$, the probability that episode $i$ was undetected.
+Next, I derive $p_{iu} = \prob((B_i, E_i) \in \Omega^C_i \mid \theta)$, the probability that episode $i$ was undetected.
 An episode is undetected if and only if no tests are performed during the episode or if there was no negative test prior to the episode.
@@
 Only $1 - p_u = 1 - \frac{1}{N} \sum_i^{\Ncis} p_{iu} = \frac{1}{N} \sum_i^{\Ncis} (1 - p_{iu})$ is required (see \cref{perf-test:eq:full-posterior-simplified}).
-Therefore, I derive $1 - p_{iu}$:
+Therefore, I derive $1 - p_{iu}$, again suppressing the conditioning on $\theta$:
 \begin{align}
@@
 As frequently observed in the literature~\autocite[e.g.:][]{heBayesian}, it is more convenient to operate with the hazard than the survival or probability mass function directly.
-This is because the mass function must sum to one and the survival must be monotonically decreasing, while there are no constraints across multiple hazards.
+The mass function must sum to one.
+The survival must be monotonically decreasing
+But, there are no constraints across multiple hazards.
+This would complicate the inference process in a way that cannot be removed through a simple transformation.
 The only constraint on the hazard is to be in the interval $[0, 1]$, since they are probabilities.
+The [0, 1] interval is easily be mapped to the unconstrained reals using a logit transformation (where $\logit(x) = \log(x/(1-x))$).
 
-In this section, I consider the various parameterisation of the survival function that are available.
-This has implications for the choice of prior it is not possible to be vague on all sensible parameterisations.
-Informative priors might also be attractive, and allow the incorporation of the estimates from \cref{E-ATACCC}.
-
-The hazard is convenient to use because it is unconstrained except in the interval [0, 1].
-The [0, 1] interval is easily be mapped to the unconstrained reals using a logit transformation (where $\logit(x) = \log(x/(1-x))$).
-Conversely, the survival function, $S_\theta(t)$, needs to be monotonically decreasing.
-Therefore, there is a constraint on $S_\theta(t)$ which depends on $S_\theta(t-1)$.
-This complicates the inference process and cannot be removed through a simple transformation.
-
-The rest of this section considers appropriate priors for the hazard in this context.
+Priors cannot be vague on both the hazard and the survival, the implications of which I consider in what follows.
+Informative priors are also attractive to allow the incorporation of the estimates from \cref{E-ATACCC}.
 
@@
 Smoothing priors encode the biological consideration that the hazard should not vary significantly from day-to-day.
-It is the Bayesian equivalent of a penalised likelihood, used frequently in the non-parametric frequentist setting (\eg: \autocite{bacchettiNonparametric}).
+It is the Bayesian equivalent of a penalised likelihood, used frequently in the non-parametric frequentist setting~\autocite[e.g.:][]{bacchettiNonparametric}.
 Many possible forms of smoothing priors exist, including splines, Gaussian Processes, and random walks.
-I use a second-order random walk because it is simple and can produce sensible prior predictive distributions.
+I use a second-order random walk because it is simple and produces plausible prior predictive distributions.
 
 A second-order random walk prior encodes that the hazard should be linearly changing with some random changes at each time step.
-Specifically, the difference between $\lambda_t$ to $\lambda_{t+1}$ is the difference between $\lambda_{t-1}$ and $\lambda_t$ plus some Gaussian-distributed noise.
+Specifically, the difference on the logit scale between $\lambda_t$ to $\lambda_{t+1}$ is the difference between $\lambda_{t-1}$ and $\lambda_t$ plus some Gaussian-distributed noise.
 Formally:
@@
 I use $\mu_{\lambda_1} = -17.5$, $\sigma_{\lambda_1} = 6$, $\mu_\epsilon = 1.09$, $\sigma_\epsilon = 0.03$, and $\mu_\epsilon = 0.1$ based on the analysis in \cref{E-ATACCC}.
+\todo[inline]{Add RW2 prior predictive}
 
@@
 
-I propose combining the information from the CIS data with \cref{E-ATACCC}'s analysis by using \cref{E-ATACCC}'s analysis as a prior for the CIS analysis.
-This is desirable because the ATACCC study frequently samples individuals early in the infection.
-Therefore, that analysis's estimate of the survival distribution in the early infection period should be reliable.
-In the later infection period, the ATACCC study has fewer observations and hence the estimates are less reliable; CIS can then inform the posterior in this region.
+The information from \cref{E-ATACCC} can be incorporated through the prior.
+This is desirable because the ATACCC study frequently sampled individuals early in the infection.
+Therefore, \cref{E-ATACCC}'s estimate of the survival distribution in the early infection period should be reliable.
+In the later infection period, the ATACCC study has fewer observations and hence the estimates are less reliable.
+CIS can inform the posterior in this region.
 
-In the combination methodology, two aspects need to be considered.
+When constructing the prior, two aspects need consideration.
 Firstly, the model structure from \cref{E-ATACCC} leads to positive correlation in the posterior estimates of the hazard, which should be propagated into this analysis.
@@
 The additional uncertainty is because the \cref{E-ATACCC} analysis extrapolates beyond the data using strong model assumptions (see \todo{reference discussion on extrapolation in ATACCC chapter}).
-Furthermore, there are differences in the study design and laboratories used between the two studies (see \cref{E-intro:sec:studies}) which may mean that results do not generalise between the studies.
+%Furthermore, there are differences in the study design and laboratories used between the two studies (see \cref{E-intro:sec:studies}) which may mean that results do not generalise between the studies.
 I form the prior for the combination in two steps.
@@
 where $k_t$, $\alpha_0$, and $\beta_0$ are hyperparameters; and $h_t$ is a point estimate of the hazard at time $t$ from ATACCC.
-An intuition for what this distribution represents can be gained by considering the conjugate model for $\lambda_t$ with a beta prior and a binomial likelihood.
+An intuition for what this distribution represents can be gained from a conjugate model for $\lambda_t$ with a beta prior and a binomial likelihood.
 If $\lambda_t$ is given the prior distribution $\text{Beta}(\alpha_0, \beta_0)$, and we then have $k_t$ observations with $k_t h_t$ successes, then the posterior distribution for $\lambda_t$ is $\text{Beta}(\alpha_t, \beta_t)$ (as defined above).
@@
 I base this assumption on the estimate from \cref{E-ATACCC} with an inflated tail to represent what is seen within CIS.
-The tail is modified based on an unpublished analysis of CIS data by Sarah Walker (described below).
+The tail is modified based on an unpublished analysis of CIS data by Sarah Walker.
 Sarah uses survival analysis to estimate the duration from CIS data.
@@
 Inference was implemented in Stan via RStan~2.21.8~\autocite{rstan2-21-8} using default settings.
-Convergence was assessed using Rhat and ESS as described in \cref{E-ATACCC:sec:convergence}, and all runs checked for divergent transitions.
+Convergence was assessed using Rhat and ESS as described in \cref{E-ATACCC:sec:convergence}\todo{add ref to place explaining diagnostics}, and all runs checked for divergent transitions.
 
-The prior on the total number of infections was TBD\todo{fill in once decided on total vs individual model}.
+The prior on the total number of infections was centered on the true value with a dispersion parameter of 1.
+This meant that the prior was very diffuse.
 
@@
 I have assumed constant and independent incidence (a priori).
-In this case, a sensitivity analysis (not shown) showed minimal impact on survival estimates.
-Furthermore, I chose a period where the incidence was approximately constant.
-However, this assumption could be important in other contexts, especially if the incidence is changing rapidly.
-The assumption of constant incidence can lead to biased estimates~\autocite{degruttolaAnalysis}.
+Prevalence in the CIS data was approximately constant over the period of interest (see \cref{E-intro:fig:prevalence}).
+A sensitivity analysis, where an epidemic in exponential growth or decline was simulated (not shown), showed minimal impact on survival estimates.
+An incorrect assumption of constant incidence can lead to biased estimates~\autocite{degruttolaAnalysis}.
+Therefore, the assumption could be important in other contexts, especially if the incidence is changing rapidly.
 

### 2023-10-03 incidence-prevalence.tex
Form full chapter

@@
 \emph{Backcalculation} is the procedure for estimating incidence based on a measurement of prevalence.
-It was first developed in the context of HIV (human immunodeficiency virus) and AIDS (acquired immune deficiency syndrome)~\autocite{brookmeyerMethod}.
-HIV is transmitted sexually or through blood.
-It causes the disease AIDS.
-AIDS has killed around 40 million people, peaking at 2 million annual deaths~\autocite{unaids2023}.
+\Textcite{brookmeyerMethod} developed it in the context of HIV (human immunodeficiency virus) and AIDS (acquired immune deficiency syndrome).
+HIV is transmitted sexually or through blood and causes the disease AIDS.
+It has killed 40 million people, peaking at 2 million annual deaths~\autocite{unaids2023}.
 In this context, backcalculation estimates the incidence of HIV infection from: data on the number of AIDS diagnoses, and the distribution of the time between infection and AIDS diagnosis~\autocites{brookmeyerBackcalculation}{brookmeyerMeasuring}.
-\todo{explain how AIDS diagnosis is "prevalence" data (is it?)}
+AIDS is incurable, therefore, AIDS prevalence is equivalent to cumulative incidence.
 
@@
 
+The original formulations of backcalculation neglected the observation process.
+AIDS is almost always diagnosed and recorded in developed countries.
+That is, the observations of AIDS diagnoses are the same as the true incidence of AIDS.
+An exception is that reporting delay may be important~\autocite{paganoHIV}.
+Reporting delay is not relevant to the studies in this thesis, and I ignore it in this and further discussion.
+Therefore, two processes need to be considered: infection and prevalence.
+
+For SARS-CoV-2, measurements of prevalence are noisy because they are based on population sampling.
+The sampling is described as the observation process.
+The observation process noise dominates the noise of the prevalence process (see \cref{incidence-prevalence:sec:observation-process}).
+Therefore, the latter is negligible and again two processes need to be considered: infection and observation.
+
+\section{Infection process}
+
 The infection process is a point process, where each infection is an event.
-The simplest plausible model is a time-inhomogenous Poisson process.
+A simple and commonly-used model is the time-inhomogenous Poisson process~\autocites{brookmeyerMethod}{paganoHIV}{rosenbergBackcalculation}{brookmeyerBackcalculation}.
 That is, a Poisson process where the intensity varies over time.
@@
 Both of these features violate the assumptions of a Poisson process.
+Despite these issues, satisfactory inference on the incidence can still be produced, although the uncertainty in the estimates may be understated~\autocite{beckerDependent}.
 
-A common cause of overdispersion is super-spreading events.
-\todo{ref}
+A common cause of overdispersion is super-spreading events~\autocite{lloyd-smithSuperspreading}.
 A super-spreading event is an event where many more people than expected are infected.
-Coronaviruses, such as SARS-CoV-2, have particularly high overdispersion.
-\todo{insert a SARS-CoV-2 superspreading example}
+Coronaviruses, such as SARS-CoV-2, have particularly high overdispersion~\autocites{endoEstimating}{adamClustering}{mccloskeySARS}.
 Overdispersion could be incorporated by modelling the counts in a given time interval as a negative binomial, rather than Poisson, distribution.
@@
 Generally, $k<1$ indicates significant levels of overdispersion.
-Estimates for SARS-CoV-2 indicate $k \approx 0.1$, and can be increased by interventions such as lockdowns.
+Estimates for SARS-CoV-2 indicate $k \approx 0.1$~\autocite{endoEstimating}, and can be increased by interventions such as lockdowns~\autocites{quiltyReconstructing}{quiltyUnderstanding}.
 However, the negative binomial distribution loses much of the mathematical convenience of a Poisson distribution.
@@
 
-In \emph{deterministic backcalculation}, the observed prevalence, $x_t/n_t$, is assumed to be equal to $P_t/N$.
-However, this is often a poor approximation.
-\emph{Statistical backcalculation} retains the sampling distribution of $x_t$ (or an approximation of it, such as a Poisson).
+An alternative to a Poisson process is \emph{mechanistic models}.
+Mechanistic models explicitly describe the population and transmission between individuals.
+They vary greatly in their details and realism~\autocite{murilloMultiscale}.
+Mechanistic models are discussed further in \cref{E-SEIR}.
 
-Define the first day of an infection episode in individual $i$ as $B_i$, and assume that the probability of $i$ being infected multiple times within the period of interest is negligible.
-The time between being infected and first being detectable is short (see\todo{ref relevant part}), and therefore I often assume that $B_i$ is the same as the time of infection.
-I keep this assumption for the remainder of this chapter.
+% In \emph{deterministic backcalculation}, the observed prevalence, $x_t/n_t$, is assumed to be equal to $P_t/N$.
+% However, this is often a poor approximation.
+% \emph{Statistical backcalculation} retains the sampling distribution of $x_t$ (or an approximation of it, such as a Poisson).
 
-Backcalculation makes use of this relationship to estimate incidence from prevalence, assuming that the duration is known.
+% Define the first day of an infection episode in individual $i$ as $B_i$, and assume that the probability of $i$ being infected multiple times within the period of interest is negligible.
+% The time between being infected and first being detectable is short (see\todo{ref relevant part}), and therefore I often assume that $B_i$ is the same as the time of infection.
+% I keep this assumption for the remainder of this chapter.
 
-Deterministic backcalculation assumes that the variance in the population is negligible, and therefore that the prevalence is equal to the mean prevalence.
-This is justified because the variance due to sampling is much larger than the variance due to the true prevalence.
+% Backcalculation makes use of this relationship to estimate incidence from prevalence, assuming that the duration is known.
 
-\todo[inline]{Explain backcalculation in HIV/AIDS context, and how it relates to what is above.}
+% Deterministic backcalculation assumes that the variance in the population is negligible, and therefore that the prevalence is equal to the mean prevalence.
+% This is justified because the variance due to sampling is much larger than the variance due to the true prevalence.
 
@@
 For each infection beginning at $t$, the probability that it ends at $t'$ is $f_d(t' - t + 1)$.
-Since these durations are assumed independent, the number of episodes of each duration is multinomially distributed.
+Since these durations are assumed independent, the number of episodes of each duration is multinomially distributed~\autocite{paganoHIV}.
 Therefore:
@@
 
-A quantity that will be important is $\E(P_t \mid \vec{Z})$, which can be derived from its constituent parts.
-First, the properties of a multinomial distribution give the marginal distribution:
+Only $\E(P_t \mid \vec{Z})$ will be relevant (see \cref{incidence-prevalence:sec:observation-process}), which can be derived from its constituent parts.
+Start by considering the recoveries.
+The sum of recoveries of infections that occurred on the same day $t$ is the sum of multinomial cell probabilities, and hence binomially distributed~\autocite{alamAnalysis}.
+Specifically:
 \begin{align}
@@
 \end{align}
-
 Now consider $P_t$:
@@
 \end{align}
-where $S(i+1) = 1-F(i) = \prob(D \geq i+1)$ is the \emph{survival function} of $D$ (see \cref{inprev:sec:duration}).
+where $S(i+1) = 1-F(i) = \prob(D \geq i+1)$ is the \emph{survival function} of $D$ (see \cref{incidence-prevalence:sec:duration}).
 
@@
 
-\todo[inline]{Need some conclusion text for this section}
+The equation derived here is a discrete convolution equation, specifically a Volterra equation of the first kind~\autocite{brookmeyerBackcalculation}.
+Clearly, knowledge of the survival function, $S$, will be important; I discuss this further in \cref{incidence-prevalence:sec:duration}.
 
-\section{Observation process}
+\section{Observation process} \label{incidence-prevalence:sec:observation-process}
 
@@
 Prevalence surveys, as described in\todo{insert ref}, sample individuals at random from the population of interest and test if they are prevalence.
-Therefore, we view the number of positive tests on day $t$ as a random variable $X_t$ based on a sample of $n_t$ individuals.
+Therefore, I model the number of positive tests on day $t$ as a random variable $X_t$ based on a sample of $n_t$ individuals.
 The $X_t$s are conditionally independent, given the population prevalences at each time, $\vec{P}_t$. 
@@
 
-\section{Estimating the duration} \label{inprev:sec:duration}
+\section{Estimating duration} \label{incidence-prevalence:sec:duration}
+
+\todo[inline]{This section fills a bit out of place, could maybe be a more general conclusion? Although these ideas and notation should probably stay...}
 
@@
 The \emph{hazard} at time $t$ is the probability of recovering at time $t$ conditional on having survived to time $t$: $\lambda(t) = \prob(D = t \mid D \geq t) = f_{D}(d) / S(d)$.
-Throughout this thesis, the survival function will be the primary quantity of interest, and it will normally be parameterised in terms of the hazard; these two quantities are related by $S(t) = \prod_{i=1}^{t-1} (1 - \lambda(i))$.
+Throughout this thesis, the survival function will bne the primary quantity of interest, and it will normally be parameterised in terms of the hazard; these two quantities are related by $S(t) = \prod_{i=1}^{t-1} (1 - \lambda(i))$.
 
-Two approaches are commonly used to estimate durations\todo{cite Sweeney}: either modelling an underlying biomarker or modelling the duration directly.
+Two approaches are commonly used to estimate durations: either modelling an underlying biomarker or modelling the duration directly~\autocite{sweetingEstimating}.
 For SARS-CoV-2, both approaches would be based on the results of PCR testing.
-Modelling a biomarker would mean modelling the viral load, measured as a Ct value.
-Modelling the duration would consider only the binary result at each test (positive or negative); in this thesis a \emph{survival analysis} framework is adopted.
-Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
-Other related approaches, such as a multi-state model, could also be considered (multi-state models can be viewed as a generalisation of survival analysis~\autocite{jacksonMSM}).
+Modelling a biomarker would mean modelling the viral load, measured as a Ct value (see \cref{E-ATACCC}).
+Modelling the duration would consider only the binary result at each test (positive or negative); in this thesis a \emph{survival analysis} framework is adopted (see \cref{E-perf-test,E-imperf-test}).
+% Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
+Other related approaches, such as a multi-state model~\autocite{jacksonMSM}, could also be considered (multi-state models can be viewed as a generalisation of survival analysis).
 Therefore, the biomarker approach considers more information per observation, however, a model for the biomarker is required which generally requires stronger assumptions.
@@
 
-\todo[inline]{Comment on previous estimates of duration for SARS-CoV-2, but not sure where exactly this goes...}
-
 \end{document}

### 2023-10-04 cis-perfect-testing.tex
todo

@@
 
-I base my analysis on the 4,800\todo{check number} episodes that were first detected between 10th Oct 2020 and 6th Dec 2020 inclusive with negatives bounding the start and end time of the episode.
+I base my analysis on the 4,800 episodes that were first detected between 10th Oct 2020 and 6th Dec 2020 inclusive with negatives bounding the start and end time of the episode.
 That is, the individual tested negative before and after the episode.

### 2023-10-04 cis-perfect-testing.tex
add RW2 prior predictive

@@
 Many possible forms of smoothing priors exist, including splines, Gaussian Processes, and random walks.
-I use a second-order random walk because it is simple and produces plausible prior predictive distributions.
+I use a second-order random walk (RW2) because it is simple and produces plausible prior predictive distributions.
 
-A second-order random walk prior encodes that the hazard should be linearly changing with some random changes at each time step.
+RW2 prior encodes that the hazard should be linearly changing with some random changes at each time step.
 Specifically, the difference on the logit scale between $\lambda_t$ to $\lambda_{t+1}$ is the difference between $\lambda_{t-1}$ and $\lambda_t$ plus some Gaussian-distributed noise.
@@
 The hyperparmeters $\mu_{\lambda_1}$ and $\sigma_{\lambda_1}$ specify the prior on the hazard at time 1; $\mu_{\epsilon_1}$ and $\sigma_{\epsilon_1}$ the prior on the initial gradient; and $\mu_\sigma$ the smoothness of the random walk.
-I use $\mu_{\lambda_1} = -17.5$, $\sigma_{\lambda_1} = 6$, $\mu_\epsilon = 1.09$, $\sigma_\epsilon = 0.03$, and $\mu_\epsilon = 0.1$ based on the analysis in \cref{E-ATACCC}.
-\todo[inline]{Add RW2 prior predictive}
+I use $\mu_{\lambda_1} = -17.5$, $\sigma_{\lambda_1} = 6$, $\mu_\epsilon = 1.09$, $\sigma_\epsilon = 0.03$, and $\mu_\epsilon = 0.1$ based on the analysis in \cref{E-ATACCC} (see \cref{perf-test:fig:rw2-prior}).
+\begin{figure}
+  \centering \includegraphics{cis-perfect-testing/rw2-prior}
+  \caption[RW2 prior for the hazard]{The of prior predictive distribution (mean and 95\% central interval) for the RW2 prior specified in the main text.
+The posterior mean estimate using ATACCC data in \cref{E-ATACCC} are shown for comparison.
+\label{perf-test:fig:rw2-prior}}
+\end{figure}
 

### 2023-10-04 cis-perfect-testing/priors.R
add RW2 prior predictive

@@ -10,7 +10,8 @@ flat_params = c(0.5, 1)
 dist_levels = c(
   map_chr(flat_params, ~glue::glue("Beta({.x}, {.x})")),
   "Beta(0.1, 1.9)",
-  "ATACCC"
+  "ATACCC",
+  "RW2"
 )
 
 sample_priors = function(alpha, beta, n_samples = 1000) {
@@ -77,6 +78,7 @@ beta_plot_surv = function(alpha, beta, prior_len = 30, include_ATACCC = FALSE) {
       colour = "Distribution",
       fill = "Distribution"
     ) +
+    scale_x_continuous(breaks = seq(0, 100) * 14, minor_breaks = seq(0, 100) * 2) +
     coord_cartesian(c(0, prior_len)) +
     standard_plot_theming()
 }
@@ -105,6 +107,109 @@ p_vague = p_vague_dens + p_vague_surv
 ggsave(
   filename = "cis-perfect-testing/vague-prior.png",
   plot = p_vague,
+  width = 15,
+  height = 9,
+  units = "cm",
+  dpi = 300
+)
+
+logits_to_dist = function(x) {
+    x |>
+        mutate(lambda = expit(value)) |>
+        group_by(.draw) |>
+        arrange(time, .by_group = TRUE) |>
+        mutate(
+            S = c(1, cumprod(1 - lambda)[-n()]),
+            F = c(lead(1 - S)[-(n())], 1),
+            f = diff(c(0, F))
+        ) |>
+        ungroup()
+}
+
+rw1_from_known_start = function(start, sd, n_steps) {
+    n_samples = length(start)
+    # Generate random steps matrix
+    random_steps = matrix(
+        rnorm(n_samples * (n_steps - 1), mean = 0, sd = sd),
+        nrow = n_samples
+    )
+
+    # Compute cumulative sum along rows
+    cumulative_random_steps = t(apply(random_steps, 1, cumsum))
+
+    # Add starting points to the random walks
+    random_walks = sweep(cumulative_random_steps, 1, start, FUN = "+")
+
+    # Add starting points as the first time step
+    return(cbind(start, random_walks))
+}
+
+sim_rw2 = function(seed = 123, n_samples = 1000, n_steps = 35, start_mean = -17.5,
+                    start_sd = 6, gradient_mean = 1.09, gradient_sd = 0.03,
+                    walk_sd_rate = 10) {
+
[…]

### 2023-10-04 cis-perfect-testing.tex
Finish perfect testing chapter to dos

@@
 
-\todo{Add plot of $k_t$}
+$k_t$ reflects the subjective belief that the estimates from \cref{E-ATACCC} are very reliable in the early infection period but unreliable by day 30.
+The equation for $k_t$ is below and shown in \cref{perf-test:fig:kt}.
+\begin{align}
+k_t = \begin{cases}
+  \expit(-0.4 * (t - 20)) &\text{for $t \leq 39$} \\
+  0 &t > 39
+\end{cases}
+\end{align}
+where $\expit(x) = 1 / (1 + \exp(-x))$.
+
+\begin{figure}
+  \centering \includegraphics{cis-perfect-testing/kt-prior}
+  \caption{Function used for $k_t$. \label{perf-test:fig:kt}}
+\end{figure}
 
@@
 Therefore, the narrow credible intervals are unsurprising.
-For hazards after day X\todo{check and insert}, the informative and vague priors are the same.
+For hazards after day 40, the informative and vague priors are the same.
 Therefore, the posterior distributions are almost identical.
+There remains a difference because the hazard estimates are partially dependent on earlier estimates.
 

### 2023-10-04 cis-perfect-testing/priors.R

@@ -215,3 +215,23 @@ ggsave(
   units = "cm",
   dpi = 300
 )
+
+expit = function(x) 1 / (1 + exp(-x))
+k_t = 1000 * expit(-0.4 * ((1:40) - 20))
+p_kt = ggplot() +
+  geom_line(aes(x = 0:39, y = k_t)) +
+  scale_x_continuous(breaks = seq(0, 100) * 14, minor_breaks = seq(0, 100) * 2) +
+  standard_plot_theming() +
+  labs(
+    x = "t",
+    y = expression(k[t])
+  )
+ggsave(
+  filename = "cis-perfect-testing/kt-prior.png",
+  plot = p_kt,
+  width = 15,
+  height = 9,
+  units = "cm",
+  dpi = 300
+)
+

### 2023-10-04 cis-perfect-testing.tex
Updates from Overleaf

--- 2023-10-04 cis-perfect-testing.tex
@@
 However, this study suffered from a small sample size and only 20 days of follow-up.
-Therefore, model assumptions determined the size of the tail estimated.
-In this chapter, I develop a method to estimate the distribution using CIS data. 
-CIS sampling is less frequent, but with a large sample size and long follow-up.
+Therefore, model assumptions determined the size of the estimated tail estimated.
+In this chapter, I develop a method to estimate the duration distribution using CIS data. CIS sampling is less frequent, but with a large sample size and long follow-up.
 
@@
 
-The CIS tests each individual infrequently, either weekly or monthly (see\todo{refer to section with CIS intro}).
+In the CIS each individualis tested  infrequently, either weekly or monthly (see\todo{refer to section with CIS intro}).
 The infrequent testing complicates analysis of the data in two ways.
@@
 Second, the infection episode start and end times are known imprecisely.
-This section explains these issues.
-The rest of the chapter develops a method to model them.
+This section explains these issues. The rest of the chapter develops a method to tackle  them.
 
@@
 The infection episode depicted begins at time 33 and has a 10-day duration.
-If the duration was 24 days or more, the individual would still be detectable at their next test and detected then.
+If the duration had been 24 days or more, the individual would still be detectable at their next test and detected then.
 CIS has many undetected infections. This is because frequency of testing (up to 28 days, see\todo{ref section on the survey design}) is longer than most infections (median around 15 days).
@@
 
-I base my analysis on the 4,800 episodes that were first detected between 10th Oct 2020 and 6th Dec 2020 inclusive with negatives bounding the start and end time of the episode.
+I base my analysis on the 4,800\todo{check number} episodes that were first detected between 10th Oct 2020 and 6th Dec 2020 inclusive with negatives bounding the start and end time of the episode.
 That is, the individual tested negative before and after the episode.

### 2023-10-06 cis-perfect-testing.tex
Correct maths errors (from Shaun)

--- 2023-10-04 cis-perfect-testing.tex
@@
 &\propto p(\theta) \sum_{\ntot=\ndet}^\infty \left( p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i=\ndet}^{\ndet} p_{ia} \right) \\
-&= p(\theta) \left( \prod_{i=\ndet}^{\ndet} p_{ia} \right) \left( p(\ntot \mid \theta) \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \right) \label{perf-test:eq:full-posterior}.
+&= p(\theta) \left( \prod_{i=\ndet}^{\ndet} p_{ia} \right) \left( \sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \right) \label{perf-test:eq:full-posterior}.
 \end{align}
@@
 
-Now I derive an analytical solution to $p(\ntot \mid \theta) = \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under the prior $\ntot \dist \NegBin(\mu, r)$ (using the mean/dispersion parameterisation of the negative binomial).
-Equivalently:
+Now I derive an analytical solution to $\sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under the prior $\ntot \dist \NegBin(\mu, r)$ (using the mean/dispersion parameterisation of the negative binomial).
+I assume that $\theta$, which are only the parameters of the survival distribution, is a priori independent of $\ntot$, the number of infections that occurred in the cohort.
+Therefore, $p(\ntot \mid \theta) = p(\ntot)$.
+Putting a negative binomial prior on $\ntot$ is equivalent to the following gamma-Poisson composite.
+Using the composite simplifies the derivation.
 \begin{align}

### 2023-10-06 incidence-prevalence.tex
Finish initial chapter draft

@@
 \begin{document}
-\chapter{Statistical background} \label{incidence-prevalence}
+\chapter{Statistical background} \label{inc-prev}
 %\todo[inline]{This chapter is pretty short. Not sure if this matters or it needs to merged with another chapter? For example, one background chapter split into epi/biology and stats.}
@@
 This chapter explains incidence and prevalence, and the relationship between these quantities.
-These quantities have been of interest as long as disease has been studied, and estimating the prevalence from incidence is the fundamental motivation for this thesis.
 Their exact definition can vary throughout the literature, here I follow the definitions of the seminal paper \textcite{freemanPrevalence}.
@@
 
-\emph{Incidence} at time $t$, $Z_t$, is the number of people in a given population who become infected at time $t$ (I use $Z_t$ rather than $I_t$ to avoid confusion with the convention that $i$ is an indexing variable).
-Incidence expressed as a proportion of the population, $Z_t/N$ where $N$ is the number of people in the population, is the \emph{incidence rate}.
-As discussed in \cref{E-intro}\todo{might want to point to a specific section}, infection events are not directly observable and must be inferred from other data.
+\emph{Incidence} at time $t$, $Z_t$, is the number of people in a given population who become infected at time $t$.
+Incidence expressed as a proportion of the population, $Z_t/\Npop$ where $\Npop$ is the number of people in the population, is the \emph{incidence rate}.
+As discussed in \cref{E-intro}\todo{might want to point to a specific section}, the timing of infection events is not directly observable and must be inferred from other data.
 Despite this issue, it remains the single most important quantity for informing the response to an epidemic or pandemic.
 
-\emph{Prevalence} of an infectious disease at time $t$, $P_t$, is the number of people in a given population currently infected at time $t$, assumed constant over the short period $t$.
+\emph{Prevalence} of an infectious disease at time $t$, $P_t$, is the number of people in a given population currently infected at time $t$.
 Each of these individuals is known as a \emph{prevalent} individual.
+As I use discrete time, I assume that an individual's prevalent status cannot change over the period represented by $t$, normally one day.
+Therefore, prevalence is constant throughout $t$.
 The \emph{prevalence proportion} is the same quantity expressed as a proportion of the population.
-For SARS-CoV-2, there are several definitions of infected.
-An infected individual could mean: one that is infectious; one that is symptomatic; or one that, if they were tested, would return a positive PCR test if tested (discussed in\todo{ref appropriate section}).
-In this thesis, prevalence is defined as the number of people in the population who would return a positive PCR test (if everyone was tested) at time $t$.
+In this thesis, a prevalent individual is one that would return a positive PCR test if tested at time $t$, not other possible definitions of ``currently infected'' discussed in\todo{ref section}.
 
@@
 It has killed 40 million people, peaking at 2 million annual deaths~\autocite{unaids2023}.
-In this context, backcalculation estimates the incidence of HIV infection from: data on the number of AIDS diagnoses, and the distribution of the time between infection and AIDS diagnosis~\autocites{brookmeyerBackcalculation}{brookmeyerMeasuring}.
+In this context, backcalculation estimates the incidence of HIV infection from: data on the number of AIDS diagnoses, and the incubation distribution (the time between infection and AIDS diagnosis)~\autocites{brookmeyerBackcalculation}{brookmeyerMeasuring}.
+\todo{maybe move some of this to intro}
 AIDS is incurable, therefore, AIDS prevalence is equivalent to cumulative incidence.
@@
 That is, the observations of AIDS diagnoses are the same as the true incidence of AIDS.
-An exception is that reporting delay may be important~\autocite{paganoHIV}.
+An exception is that the reporting delay may be important~\autocite{paganoHIV}.
 Reporting delay is not relevant to the studies in this thesis, and I ignore it in this and further discussion.
-Therefore, two processes need to be considered: infection and prevalence.
+Therefore, two processes needed to be considered in the context of HIV/AIDS: infection and prevalence.
 
@@
 The sampling is described as the observation process.
-The observation process noise dominates the noise of the prevalence process (see \cref{incidence-prevalence:sec:observation-process}).
+The observation process noise is much larger than the noise of the prevalence process (see \cref{inc-prev:sec:observation-process}).
 Therefore, the latter is negligible and again two processes need to be considered: infection and observation.
@@
 A simple and commonly-used model is the time-inhomogenous Poisson process~\autocites{brookmeyerMethod}{paganoHIV}{rosenbergBackcalculation}{brookmeyerBackcalculation}.
-That is, a Poisson process where the intensity varies over time.
+That is, a Poisson process where the intensity varies over time, and possibly with other covariates.
 The target of inference can be either the process's intensity or the counts of the specific realisation that occurred.
@@
 A common cause of overdispersion is super-spreading events~\autocite{lloyd-smithSuperspreading}.
-A super-spreading event is an event where many more people than expected are infected.
+A super-spreading event is an event where many more people than average are infected.
 Coronaviruses, such as SARS-CoV-2, have particularly high overdispersion~\autocites{endoEstimating}{adamClustering}{mccloskeySARS}.
+An infamous example of a SARS-CoV-2 super-spreading event was the Skagit County choir practice in March 2020~\autocite{hamnerHigh}.
+One infected individual is thought to have infected at least 32 further individuals during a single choir practice.
+For comparison, the average individual infected around three others in the early stages of the pandemic~\autocite{pellisChallenges}. 
 Overdispersion could be incorporated by modelling the counts in a given time interval as a negative binomial, rather than Poisson, distribution.
-The mean of these distributions would be equal, but the negative binomial introduces an additional overdispersion parameter, commonly notated $k$.
-These distributions are related in that the negative binomial tends to Poisson as $k\to\infty$.
-Generally, $k<1$ indicates significant levels of overdispersion.
-Estimates for SARS-CoV-2 indicate $k \approx 0.1$~\autocite{endoEstimating}, and can be increased by interventions such as lockdowns~\autocites{quiltyReconstructing}{quiltyUnderstanding}.
-However, the negative binomial distribution loses much of the mathematical convenience of a Poisson distribution.
+The mean of these distributions is equal, but the negative binomial introduces an additional overdispersion parameter, commonly notated $k$.
+The negative binomial tends to Poisson as $k\to\infty$.
+Generally, $k<1$ is considered as significant levels of overdispersion.
+Estimates for SARS-CoV-2 indicate $k \approx 0.1$~\autocite{endoEstimating}, increased by some interventions including lockdowns~\autocites{quiltyReconstructing}{quiltyUnderstanding}.
 For a negative binomial process, the count's standard deviation scales on the order of the mean.
@@
 A particular implication of this is that the coefficient of variation (the standard deviation divideed by the mean) tends to zero for a Poisson process, but tends to a constant for a negative binomial process.
+The disadvantage of the negative binomial distribution is that it loses much of the mathematical convenience of a Poisson distribution.
 
-The independence assumption is violated because if the number of infections is stochastically high, this feeds back to the process intensity at future times.
+The independence assumption of a Poisson or negative binomial process is violated because if the number of infections is stochastically high (or low), this feeds back to the process intensity at future times.
 This can be seen simply in the event of a super-spreading event.
@@
 
-% In \emph{deterministic backcalculation}, the observed prevalence, $x_t/n_t$, is assumed to be equal to $P_t/N$.
+% In \emph{deterministic backcalculation}, the observed prevalence, $x_t/n_t$, is assumed to be equal to $P_t/\Npop$.
 % However, this is often a poor approximation.
@@
 The relationship between incidence and prevalence can be derived by considering the probability that an individual is prevalent at any time following an infection.
-The period the individual is prevalent for is the \emph{infection episode}, which I will index with $i$, associated with that infection event.
+The period the individual is prevalent for is the \emph{infection episode}.
+I index infection episodes with $i$.
 Denote the time that the infection episode begins as $B_i$ and the time it ends as $E_i \geq B_i$.
@@
 The duration of the infection episode, the number of days for which they are prevalent, is $D_i = E_i - B_i + 1$.
-In this thesis, it is assumed that the $D_i$ for each infection episode is independent and $\prob(D_i = d) = f_d$ for some pmf $f$ over $1, 2, \dots, d_\text{max}$.
+In this thesis, it is assumed that the $D_i$ for each infection episode is an independent draw from a discrete distribution such that $\prob(D_i = d) = f_d(d)$ for $d = 1, \dots, d_\text{max}$ and $\prob(D_i = d) = 0$ otherwise.
+Denote by $F_D$ the cdf of $D_i$, that is $F_D(d) = \prob(D_i \leq d)$.
 
@@
 Denote by $R_{t,t'}$ be the number of infection episodes that begin at time $t$ and end at time $t'$.
-For each infection beginning at $t$, the probability that it ends at $t'$ is $f_d(t' - t + 1)$.
+For each infection beginning at $t$, the probability that it ends at $t'$ is $\prob(E_i = t' \mid B_i = t) = \prob(D_i = t' - t + 1 \mid B_i = t) = f_d(t' - t + 1)$.
 Since these durations are assumed independent, the number of episodes of each duration is multinomially distributed~\autocite{paganoHIV}.
@@
 \end{align}
-Now, denote by $P_{t_z,t_p}$ the number of episodes beginning at time $t_z$ that are prevalent at time $t_p$ (where $t' \geq t$).
-This is $Z_t$ minus those that have recovered:
+Now, denote by $P_{t_z,t_p}$ the number of episodes beginning at time $t_z$ that are prevalent at time $t_p$.
+This is $Z_t$ minus those that recover before $t_p$ (\ie at or up to $t_p - 1$).
+Hence:
 \begin{align}
@@
       0 &t_p < t_z\\
-      Z_t - \sum_{i=0}^{t_p-t_z-1} R_{t_z,t_z+i} &t_z \leq t_p < t_z + d_\text{max}\\
+      Z_{t_z} - \sum_{i=0}^{t_p-t_z-1} R_{t_z,t_z+i} &t_z \leq t_p < t_z + d_\text{max}\\
       0 &t_p \geq t_z + d_\text{max}.
-  \end{cases} \label{incprev:eq:Ptt-to-R}
+  \end{cases} \label{inc-prev:eq:Ptt-to-R}
 \end{align}
-Finally, the prevalence at $t$ can be expressed as the total number of people prevalent at time $t$, regardeless of when they were infected:
+Finally, the prevalence at $t$ can be expressed as the total number of people prevalent at time $t$, regardless of when they were infected.
+The limits of the sum are limited by the zeroes in \cref{inc-prev:eq:Ptt-to-R}.
 \begin{align}
   P_t
-  &= \sum_{t_z=\infty}^\infty P_{t_z,t} \\
-  &= \sum_{i=0}^{d_\text{max}-1} P_{t-i,t} \label{incprev:eq:Pt-to-Ptt}.
+  &= \sum_{t_z=-\infty}^\infty P_{t_z,t} \\
+  &= \sum_{i=0}^{\dmax-1} P_{t-i,t} \label{inc-prev:eq:Pt-to-Ptt} \\
+  &= \sum_{i=0}^{\dmax-1} \left(Z_{t_z} - \sum_{j=0}^{i-1} R_{t-i,t-i+j} \right) &\text{by \cref{inc-prev:eq:Ptt-to-R}}\label{inc-prev:eq:Pt-to-Rtt}.
 \end{align}
 
-In general, all these random variables will be dependent for an infectious disease.
-This is because the probability of being infected at time $t$ depends on the number of infectious individuals at time $t$, which itself depends on the number of individuals infected before time $t$.
-However, if we condition on the vector $\vec{Z} = (Z_1, Z_2, \dots)^T$, then we will induce some independence.
-In particular, because the duration of individuals is independent, $R_{t_1,t_2}$ and $R_{t_1',t_2'}$ will be independent if $t_1 \neq t_1'$; importantly, this holds even if $t_2 = t_2'$.
+The infection process means that these random variables are dependent.
+However, if we condition on the vector of incidence, $\vec{Z} = (Z_1, Z_2, \dots)^T$, then we will induce some independence.
+This is essentially isolating only the prevalence process.
+In particular, because the duration of individuals is independent, $R_{t_z,t_p}$ and $R_{t_z',t_p'}$ will be independent if $t_z \neq t_z'$; importantly, this holds even if $t_p = t_p'$.
+Therefore, $P_{t_z,t_p}$ and $P_{t_z',t_p'}$ will be independent if $t_z \neq t_z'$.
 
-Only $\E(P_t \mid \vec{Z})$ will be relevant (see \cref{incidence-prevalence:sec:observation-process}), which can be derived from its constituent parts.
+Only $\E(P_t \mid \vec{Z})$ and bounds on $\V(P_t \mid \vec{Z})$ will be relevant (see \cref{inc-prev:sec:observation-process}).
+These can be derived from their constituent parts.
+
 Start by considering the recoveries.
@@
 \begin{align}
-  \sum_{i=0}^{t'} R_{t,t+i} \mid \vec{Z} &\sim \text{Binomial}(Z_t, F_d(t'+1)).
+  \sum_{i=0}^{t'} R_{t,t+i} \mid \vec{Z} &\sim \text{Binomial}(Z_t, F_d(t'+1)). \label{inc-prev:eq:binomialRt}
 \end{align}
@@
 \begin{align}
-  \E(\sum_{i=0}^{t'} R_{t,t+i} \mid \vec{Z}) = Z_t F_d(t'+1) \label{incprev:eq:EsumRt}
+  \E \left( \sum_{i=0}^{t'} R_{t,t+i} \mid \vec{Z} \right) &= Z_t F_d(t'+1) \label{inc-prev:eq:EsumRt} \\
+  \V \left( \sum_{i=0}^{t'} R_{t,t+i} \mid \vec{Z} \right) &= Z_t F_d(t'+1) (1 - F_d(t'+1)) \label{inc-prev:eq:VsumRt}
 \end{align}
-Now consider $P_t$:
+
+The first moment follows.
 \begin{align}
 \E(P_t \mid \vec{Z})
-  &= \E\left(\sum_{i=1}^{d_\text{max}-1} P_{t-i,t} \mid \vec{Z} \right) &\text{by \cref{incprev:eq:Pt-to-Ptt}} \\
-  &= \E\left(\sum_{i=1}^{d_\text{max}-1} \left( Z_{t-i} - \sum_{j=0}^{i-1} R_{t-i,t-i+j} \right) \mid \vec{Z} \right) &\text{by \cref{incprev:eq:Ptt-to-R}}\\
-  &= \sum_{i=1}^{d_\text{max}-1} \left( Z_{t-i} - \sum_{j=0}^{i-1} \E( R_{t-i,t-i+j} \mid \vec{Z}) \right) \\
-  &= \sum_{i=1}^{d_\text{max}-1} \left( Z_{t-i} - Z_{t-i} F(i) \right) &\text{by \cref{incprev:eq:EsumRt}}\\
-  &= \sum_{i=1}^{d_\text{max}-1} Z_{t-i} (1 - F(i)) \\
-  &= \sum_{i=1}^{d_\text{max}-1} Z_{t-i} S(i+1)
+  &= \E\left(\sum_{i=0}^{d_\text{max}-1} \left( Z_{t-i} - \sum_{j=0}^{i-1} R_{t-i,t-i+j} \right) \mid \vec{Z} \right) &\text{by \cref{inc-prev:eq:Pt-to-Rtt}}\\
+  &= \sum_{i=0}^{d_\text{max}-1} \left( Z_{t-i} - \sum_{j=0}^{i-1} \E( R_{t-i,t-i+j} \mid \vec{Z}) \right) \\
+  &= \sum_{i=0}^{d_\text{max}-1} \left( Z_{t-i} - Z_{t-i} F_d(i) \right) &\text{by \cref{inc-prev:eq:EsumRt}}\\
+  &= \sum_{i=0}^{d_\text{max}-1} Z_{t-i} (1 - F_d(i)) \\
+  &= \sum_{i=0}^{d_\text{max}-1} Z_{t-i} S(i+1)
 \end{align}
-where $S(i+1) = 1-F(i) = \prob(D \geq i+1)$ is the \emph{survival function} of $D$ (see \cref{incidence-prevalence:sec:duration}).
+where $S(i+1) = 1-F_d(i) = \prob(D \geq i+1)$ is the \emph{survival function} of $D$ (see \cref{inc-prev:sec:duration}).
+
+The second moment is simplified by the independence.
+\begin{align}
+\V(P_t \mid \vec{Z})
+  &= \V\left(\sum_{i=1}^{d_\text{max}-1} P_{t-i,t} \mid \vec{Z} \right) &\text{by \cref{inc-prev:eq:Pt-to-Ptt}} \\
+  &= \sum_{i=1}^{d_\text{max}-1} \V\left(P_{t-i,t} \mid \vec{Z} \right) &\text{by the conditionl independence} \\
+  &= \sum_{i=1}^{d_\text{max}-1} \V\left(\sum_{j=0}^{i-1} R_{t-i,t-i+j} \mid \vec{Z} \right) &\text{by~\cref{inc-prev:eq:Ptt-to-R}}\\
+  &= \sum_{i=1}^{d_\text{max}-1} Z_{t-i} F_d(i) (1 - F_d(i)) &\text{by~\cref{inc-prev:eq:VsumRt}} \\
+  &\leq \sum_{i=1}^{d_\text{max}-1} Z_{t-i} (1 - F_d(i)) &\text{as $F_d \leq 1$}\\
+  &= \E(P_t \mid \vec{Z}). \label{inc-prev:eq:boundVPt}
+\end{align}
 
@@
 
-The equation derived here is a discrete convolution equation, specifically a Volterra equation of the first kind~\autocite{brookmeyerBackcalculation}.
-Clearly, knowledge of the survival function, $S$, will be important; I discuss this further in \cref{incidence-prevalence:sec:duration}.
+The expectation derived here is a discrete convolution equation, specifically a discrete version of the Volterra equation of the first kind~\autocite{brookmeyerBackcalculation}.
+The variance is upper bounded by this expectation, a fact that will prove useful in \cref{inc-prev:sec:observation-process}.
+Clearly, knowledge of the survival function, $S$, will be important; I discuss this further in \cref{inc-prev:sec:duration}.
 
-\section{Observation process} \label{incidence-prevalence:sec:observation-process}
+\section{Observation process} \label{inc-prev:sec:observation-process}
 
@@
 Therefore, I model the number of positive tests on day $t$ as a random variable $X_t$ based on a sample of $n_t$ individuals.
+That is, $X_t \mid P_t, n_t, \Npop \dist \Bin(n_t, P_t/\Npop)$.
+The conditioning on $n_t$ and $\Npop$ is implicit in what follows.
+
 The $X_t$s are conditionally independent, given the population prevalences at each time, $\vec{P}_t$. 
@@
 %The population prevalence is a latent quantity.
-From these assumptions, $X_t \mid \vec{P} \sim \Bin(n_t, P_t / N)$.
-I base inference on the likelihood, $p(\vec{x} \mid \vec{Z}) = \int p(\vec{x} \mid \vec{P}) p(\vec{P} \mid \vec{Z}) d\vec{P} = \int \prod_t p(x_t \mid P_t) p(\vec{P} \mid \vec{Z}) d\vec{P}$, where $\vec{x} = (x_1, x_2, \dots)^T$.
+I base inference on the likelihood, $p(\vec{x} \mid \vec{Z})$.% = \int p(\vec{x} \mid \vec{P}) p(\vec{P} \mid \vec{Z}) d\vec{P} = \int \prod_t p(x_t \mid P_t) p(\vec{P} \mid \vec{Z}) d\vec{P}$, where $\vec{x} = (x_1, x_2, \dots)^T$.
 
-For most prevalence surveys, the uncertainty arising from the distribution $\vec{P} \mid \vec{Z}$ is negligible.
-This can be seen by considering the variance in the likelihood.
-Assume $x_t \sim \Bin(n_t, P_t / N)$, where $N$ is the size of the population.
-Therefore, by the law of total variance:
+This section argues that $p(\vec{x} \mid \vec{Z}) \approx \prob(\vec{X} = \vec{x} \mid \vec{P} = \E(\vec{P} \mid \vec{Z}))$.
+As the $x_t$s are independent conditional on $\vec{P}$, it follows that $p(\vec{x} \mid \vec{Z}) \approx \prod_t \prob(X_t = x_t \mid P_t = \E(P_t \mid \vec{Z}))$.
+This relationship will make inference for $\vec{Z}$ much simpler because the latent vector $\vec{P}$ and the dependence of the $x_t$s can be ignored.
+
+The basis of the argument is that the first two moments of these distributions have negligible difference.
+The first moment follows directly from the tower property of expectations.
+The majority of the argument is to show that the variance-covariance matrix $\V(\vec{X} \mid \vec{Z}) \approx \V(\vec{X} \mid \vec{P} = \E(P \mid Z))$.
+There are two steps to this argument.
+I start with the diagonal elements, the variances.
+I show that $\V(X_t \mid \vec{Z}) \approx \V(X_t \mid \vec{P} = \E(\vec{P} = \vec{Z}))$.
+The conditional independence of the $x_t$s implies that the conditional covariances are 0.
+Therefore, I show that the correlation between $X_t$ and $X_{t'}$ conditional only on $\vec{Z}$ is negligible.
+
+Starting with the variances, by the law of total variance:
 \begin{align}
-  \var\left( x_t \mid \vec{n}, \vec{Z} \right)
-    &= \var\left[\E\left( x_t \mid \vec{n}, \vec{Z}, \vec{P} \right) \mid \vec{n}, \vec{Z} \right] + \E\left[\var\left( x_t \mid \vec{n}, \vec{I}, \vec{P} \right) \mid \vec{n}, \vec{Z} \right] \\
-    &= \var\left( n_t \frac{P_t}{N} \mid \vec{n}, \vec{Z} \right) + \E\left( n_t \frac{P_t}{N} \left(1 - \frac{P_t}{N} \right) \mid \vec{n}, \vec{Z} \right) \\
-    &= \left( \frac{n}{N} \right)^2 \var\left(P_t \mid \vec{n}, \vec{Z} \right) + \frac{n_t}{N} \E\left(P_t \mid \vec{n}, \vec{Z} \right)  - \frac{n_t}{N^2} \E\left(P_t^2 \mid \vec{n}, \vec{Z} \right).
+  &\var\left( X_t \mid \vec{Z} \right) \\
+    &= \var\left[\E\left( X_t \mid \vec{Z}, \vec{P} \right) \mid \vec{Z} \right] + \E\left[\var\left( X_t \mid \vec{Z}, \vec{P} \right) \mid \vec{Z} \right] \\
+    &= \var\left( n_t \frac{P_t}{\Npop} \mid \vec{Z} \right) + \E\left( n_t \frac{P_t}{\Npop} \left(1 - \frac{P_t}{\Npop} \right) \mid \vec{Z} \right) \\
+    &= \left( \frac{n_t}{\Npop} \right)^2 \var\left(P_t \mid \vec{Z} \right) + \frac{n_t}{\Npop} \E\left(P_t \mid \vec{Z} \right)  - \frac{n_t}{\Npop^2} \E\left(P_t^2 \mid \vec{Z} \right) \\
+    &= \left( \frac{n_t}{\Npop} \right)^2 \var\left(P_t \mid \vec{Z} \right) + \frac{n_t}{\Npop} \E\left(P_t \mid \vec{Z} \right)  - \frac{n_t}{\Npop^2} \left(\E\left(P_t \mid \vec{Z} \right) ^ 2 + \V(P_t \mid \vec{X})\right) \\
+    &= \frac{n_t(n_t - 1)}{\Npop^2} \var\left(P_t \mid \vec{Z} \right) + n_t \frac{\E\left(P_t \mid \vec{Z} \right)}{\Npop}\left(1 - \frac{\E\left(P_t \mid \vec{Z} \right)}{\Npop} \right) \\
+    &= \frac{n_t(n_t - 1)}{\Npop^2} \var\left(P_t \mid \vec{Z} \right) + \V(X_t \mid P_t = \E(P_t \mid \vec{Z})).
 \end{align}
-For large populations and reasonable sample sizes, $n_t << N$ and $\E(P_t) \approx \var(P_t)$, the first term is negligible.
-Hence, assuming that $\var(\vec{P}) = \matr{0}$ is a reasonable approximation.
-This is equivalent to assuming that $\vec{P}$ is a deterministic function of $\vec{Z}$, $\vec{P} = \E( \vec{P} \mid \vec{Z})$.
+The first term of this expression is negligible in the contexts I consider.
+These contexts have $n_t << \Npop$.
+$\Npop$ is the population of England (around 56 million) while $n_t$ is the daily sample size (10s of thousands).
+Further $\V(P_t \mid \vec{Z}) \leq \E(P_t \mid \vec{Z})$ (\cref{inc-prev:eq:boundVPt}).
+Therefore, $\var\left( X_t \mid \vec{Z} \right) \approx \V(X_t \mid P_t = \E(P_t \mid \vec{Z}))$.
 
-\section{Estimating duration} \label{incidence-prevalence:sec:duration}
+A similar argument will show that the correlation between $x_t$ and $x_{t'}$ ($t \neq t'$) is negligible.
+The basic structure of this argument is as the previous one.
+The result follows because the observation noise dominates, and this is uncorrelated.
+In the following, all are implicitly conditioned on $\vec{Z}$.
+However, for clarity this is omitted from the notation.
+\begin{align}
+  &\lvert \cor(X_t, X_{t'}) \rvert \\
+  &= \frac{\lvert\cov(X_t, X_{t'})\rvert}{\sqrt{\var(X_t) \var(X_{t'})}} &\text{by definition}\\
+  &= \frac{\lvert \E(\cov(X_t, X_{t'} \mid \vec{P})) + \cov(\E(X_t \mid \vec{P}), \E(X_{t'} \mid \vec{P})) \rvert}{\sqrt{\var(X_t) \var(X_{t'})}} &\text{by law of total covariance} \\
+  &= \frac{\lvert 0 + \cov(n_t P_t / \Npop, n_{t'} P_{t'} / \Npop) \rvert}{\sqrt{\var(X_t) \var(X_{t'})}} \\
+  &= \frac{n_t n_{t'} \lvert \cov(P_t, P_{t'}) \rvert}{\Npop^2 \sqrt{\var(X_t) \var(X_{t'})}}  \\
+  &<< \frac{n_t n_{t'} \lvert \cov(P_t, P_{t'}) \rvert}{\Npop^2 \sqrt{\var(P_t) \var(P_{t'})}} &\text{as $\V(X_t) >> \V(P_t)$} \\
+  &= \frac{n_t n_{t'}}{\Npop^2} \lvert \cor(P_t, P_{t'}) \rvert \\
+  &\leq \frac{n_t n_{t'}}{\Npop^2}.
+\end{align}
+$n_t n_{t'} / \Npop^2$ is very small, for the same reason that $n_t / \Npop$ is small.
+The argument above shows that $\lvert \cor(X_t, X_{t'}) \rvert$ is much smaller than this.
+Therefore, neglecting the correlation between $X_t$ and $X_{t'}$ is a good approximation.
 
-\todo[inline]{This section fills a bit out of place, could maybe be a more general conclusion? Although these ideas and notation should probably stay...}
+This section showed that, when using a prevalence survey such as CIS, the likelihood $p(\vec{x} \mid \vec{Z})$ is very well approximated by $\prob(\vec{X} = \vec{x} \mid \vec{P} = \E(\vec{P} \mid \vec{Z}))$.
 
-Robust, unbiased prevalence estimates already exist (see\todo{ref prevalence estimates}), and incidence estimates are the most useful (see\todo{ref intro with importance of incidence}).
+\section{Estimating duration} \label{inc-prev:sec:duration}
+
+Robust, unbiased prevalence estimates already exist (see\todo{ref prevalence estimates}), and incidence estimates are the objective of this thesis.
+The quantity that relates these is the distribution of the duration of PCR positivity.
+However, no estimates of this quantity in the general population exist.
 Therefore, a large portion of this thesis concerns itself with estimating the duration of PCR positivity.
@@
 The distribution of durations can be expressed in several equivalent ways.
-Let $D_i$ be the random variable for the duration of episode $i$, then the distribution of interest is that of $D_i$.
-I will assume that each $D_i$ is an independent and identically distributed realisation of a random variable $D$; extending the methodology to account for covariates is left to further work.
-For an infection in discrete time, the duration is a non-negative integer, \ie $D \in \{1, 2, \dots\}$.
+As before, $D_i$ is the duration of episode $i$.
+I model each $D_i$ as a discrete independent and identically distributed random variable; extending the methodology to account for covariates is left to further work.
+The duration is a positive integer, \ie $D \in \{1, 2, \dots\}$.
 The first way of expressing the distribution is as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=0}^d f_{D}(i)$.
@@
 
+\todo[inline]{Maybe add a conclusion?}
+
 \end{document}

### 2023-10-10 SEIR.tex
Outline SEIR chapter
Start of SEIR chapter

@@
 \begin{document}
-\chapter{Mechanistic modelling of incidence and transmission}
+\chapter{Mechanistic modelling of incidence and transmission} \label{SEIR}
+
+Mechanistic models are...
+
+This chapter focuses on mechanistic models based on ordinary differential equations (ODEs).
+They are useful because...
+
+Describing other mechanistic models is beyond the scope of this thesis.
+Models vary greatly in complexity and form, depending on the disease and the question of interest.
+I recommend \Textcite{kretzschmarMathematical} for an introduction to the breadth of the field or \Textcite{keelingModeling} for a book-length treatment.
+
+Notation in this chapter differs from the rest of the thesis.
+I adopt the conventions and notation standard in the infectious disease modelling field.
+In particular, upper-case letters denote populations, and lower-case letters denote proportions.
+Whether a variable is random or not is explictly stated if not clear from the context.
+
+Sections X\todo{which?} give the background on ODE-based mechanistic models.
+Section X\todo{which?} describes the model that I use for the CIS data, and section X the inference procedure.
+I verify the model and inference procedure in a simulation study in section X.
+I apply the model to the autumn 2020 CIS data in section X.
+Finally, section X discusses the results, and possible extensions, including those needed to extend to the 2021 period which includes vaccination.
+
+\section{ODE-based compartmental transmission models} \label{SEIR:sec:transmission}
+
+This section first introduces a simple ODE-based model.
+ODE-based models are \emph{deterministic}.
+Deterministic models have the property that once the parameters are set, there is no randomness in the transmission model.
+The transmission model corresponds to the infection process (see \cref{E-inc-prev:sec:infection-process}).
+There remains randomness in the observation model (described in \cref{SEIR:sec:observation}).
+
+ODE models are mathematically described by a system of ODEs.
+The derivative is taken with respect to time, $t$.
+Therefore, in an ODE model, time is necessarily continuous.
+However, time is discretised when solving (see \cref{SEIR:sec:inference-implementation})
+
+The number of individuals is also continuous.
+This is a reasonable approximation when the population is large.
+
+The other type of model is \emph{stochastic}.
+Stochastic models have randomness in the transmission model, and often use discrete time.
+For example, the number of individuals infected on day $t+1$, conditional on the model state on day $t$ might be a Poisson distribution.
+When the number of infected individuals is large, then deterministic models are a good approximation to stochastic models.
+Deterministic models are also known as \emph{mean-field} models because they can be derived from stochastic models by taking the mean of the distribution of the number of infected individuals.
+Further discussion of stochastic models is beyond the scope of this thesis.
+
+The class of models considered here is \emph{compartmental} models.
+A compartmental model divides all members of the population into compartments, each representing a disease state.
+Compartmental models are the most common type of mechanistic infectious disease model.
+Compartmental models can be deterministic or stochastic.
+
+\subsection{The SIR model}
+
+The simplest compartmental model is the \emph{susceptible-infected-recovered} (SIR) model.
+The name refers to the three compartments in the model.
+Individuals in the susceptible compartment can be infected.
+Individuals in the infected compartment can infect others, and eventually recover.
+Individuals in the recovered compartment are immune to the disease, and cannot be infected again.
+
+The SIR model is described by the following system of ODEs:
+\begin{align}
+\frac{ds}{dt} &= -\beta si \\
+\frac{di}{dt} &= \beta si - \gamma i \\
+\frac{dr}{dt} &= \gamma i
+\end{align}
+$s$, $i$ and $r$ are the proportions of the population in the susceptible, infected and recovered compartments respectively.
+These are the state of the system.
+The state is a function of time, $t$, although this is not shown explicitly.
+Two parameters are present in this model: $\beta$ and $\gamma$.
+$\beta$ is the transmission rate, or the number of individuals an infected individual infects per day.
+$\gamma$ is the recovery rate, or the proportion of infected individuals that recover per day; hence, $1/\gamma$ is the mean infectious period.
+These parameters are assumed to be constant over time (this assumption is relaxed in \cref{SEIR:sec:time-varying-foi}).
+
+The term $\beta si$ is the instantaneous number of individuals infected.
+That is, the number of individuals moving from the susceptible state to the infected state.
+This term can be arrived at by considering a single infected individual.
+Assume this individual makes contact with other individuals in the population at a constant rate.
+Further, assume that the population is \emph{well-mixed}, that is that the individual is equally likely to come into contact with any other individual.
+Therefore, the probability that the individual comes each of these contacts is a susceptible individual is $s$.
+Finally, assume that there is a constant probability of transmission upon contact between an infected and susceptible individual.
+The two constant terms here (number of contacts and probability of transmission) are absorbed into the transmission rate, $\beta$.
+Therefore, the instantaneous number of infections per infected individual is $\beta s$.
+The overall proportion of the population instantaneously infected is $\beta si$.
+
+The term $\gamma i$, the instantaneous number of individuals recovering, follows directly from the definitions of $\gamma$ and $i$.
+
+A few properties of the SIR model are worth noting.
+First, the population is closed.
+That is, the total population size is constant, and no infections are imported from outside the population.
+Births and deaths of individuals are also excluded.
+Second, there is no analytical solution to the system of ODEs.
+Numerical methods are required to solve the system.
+Third, in addition to the parameters, the state of the system at some time (normally the initial state) is required to solve the system.
+
+\subsection{Non-exponential waiting times}
+
+The basic SIR model implies that the distribution of the time spent in the infected compartment is exponential.
+This is an unrealistic assumption and leads to underestimating the reproduction number~\autocites{lloydRealistic}{wearingAppropriate}.
+A standard extension to the SIR model is to allow the infectious period to follow a gamma distribution.
+This is done by adding multiple infected compartments.
+The system can now be written as follows
+\begin{align}
+\frac{ds}{dt} &= -\beta si \\
+\frac{di_1}{dt} &= \beta si - n\gamma i_1 \\
+\frac{di_2}{dt} &= n\gamma i_1 - n \gamma i_2 \\
+&\vdots \nonumber \\
+\frac{di_n}{dt} &= n\gamma i_{n-1} - n \gamma i_n \\
+\frac{dr}{dt} &= n\gamma i_n
+\end{align}
+where $n$ is the number of infected compartments.
+The infectious period is now distributed $\GamDist(n, n\gamma)$.
+A gamma distribution with the first paramter equal to 1 is an exponential distribution, connecting this model to the basic SIR model.
+$1/\gamma$ remains the mean infectious period because $n\gamma$ is used as the rate of transition between the compartment.
+
+Using a two parameter distribution allows the first two moments of the infectious period to be specified.
+This is normally sufficient to give realistic dynamics within a mean-field model.
+
+\subsection{The SEIR model}
+
+The SIR model assumes that individuals are immediately infectious upon infection.
+This is unrealistic, as there is a period between infection and infectiousness, known as the \emph{latent period}.
+Ignoring the latent period will again overestimate the reproduction number~\autocite{wearingAppropriate}.
+
+The SIR model can be extended to include a latent period by adding a latent compartment.
+This is also known as an \emph{exposed} compartment.
+Therefore, we now form the susceptible-exposed-infected-recovered (SEIR) model.
+
+Similar to the exponential infectious period, the latent period can be modelled as a gamma distribution by adding multiple latent compartments.
+\begin{align}
+\frac{ds}{dt} &= -\beta si \\
+\frac{de_1}{dt} &= \beta si - m\sigma e_1 \\
+\frac{de_2}{dt} &= m\sigma e_1 - m \sigma e_2 \\
+&\vdots \nonumber \\
+\frac{de_m}{dt} &= m\sigma e_{m-1} - m \sigma e_m \\
+\frac{di_1}{dt} &= m\sigma e_m - n\gamma i_1 \\
+\frac{di_2}{dt} &= n\gamma i_1 - n \gamma i_2 \\
+&\vdots \nonumber \\
+\frac{di_n}{dt} &= n\gamma i_{n-1} - n \gamma i_n \\
+\frac{dr}{dt} &= n\gamma i_n
+\end{align}
+where $m$ is the number of latent compartments and $1/\sigma$ is the mean latent period.
+The latent period is now distributed $\GamDist(m, m\sigma)$.
+The infectious period is still distributed $\GamDist(n, n\gamma)$.
+
+\subsection{Structured populations}
+
+\todo[inline]{Introduction about the importance of age and mixing patterns, maybe show a contact matrix}
+
+To allow for age, the population is divided into age groups.
+The state of each comparment is now a vector.
+The proportion of the population that is susceptible is $\vec{s}$ where $s_i$ is the proportion of the $i$th age group that is susceptible.
+Similarly for the other compartments.
+
+The modelling approach here can be used to stratify the population by any charcteristic.
+For example, using deprivation or ethnicity.
+In this case, $i$ simply represents the $i$th strata.
+
+The assumption of a well-mixed population is relaxed by using a contact matrix.
+Now, only conditional on age are all individuals identical.
+
+To incorporate the strata into the model, first I introduce some notation.
+denote the set of strata, normally age groups, as $\set{A}$. $N_i$ notates the population in stratum $i$ and hence $s_i N_i$ is the number of susceptible people in stratum $i$.
+The other compartments are notated similarly.
+
+We want to derive a matrix $\matr{B}$ which we can use to calculate the instantaneous number of new infections.
+We do this by considering a susceptible individual $a$ in stratum $i$.
+I use the notation $\rate(X)$ to mean the rate at which the event $X$ occur.
+\begin{align}
+    &\rate(\text{Individual $a$ infected}) \\
+    &\approx \sum_{j \in \set{A}} \rate(\text{Individual $a$ is infected by an individual in stratum $j$}) \\
+    &\approx \sum_{j \in \set{A}} \rate(\text{Contacts $a$ has with individuals in stratum $j$}) \\ &\, \times \prob(\text{$a$ infected} \mid \text{contact between $a$ and individual from stratum $j$})\\
+    &= \sum_{j \in \set{A}} M_{ij} \beta_{ij} i_j
+\end{align}
+where $M_{ij}$ is the instantaneous number of contacts an individual in strata $i$ has with individuals in stratum $j$ (assumed homogeneous mixing within strata); and $\beta_{ij}$ is the probability of transmission from an infectious individual in stratum $j$ to a susceptible individual in stratum $i$ given a contact between these individuals.
+Note that, unlike previously, here we have explicitly separated out the number of contacts $M_{ij}$ from the probability of transmission $\beta_{ij}$.
+The first two approximations are from mass-action dynamics which are valid as long as the transmission probability is small, meaning that we can ignore that the events are not mutually exclusive.
+Therefore, the expected rate of new infections in strata $i$ is $s_i N_i \sum_{j \in \set{A}} M_{ij} \beta_{ij} i_j$, which is equivalent to:
+\begin{align}
+    \frac{d\vec{s}}{dt} = -\vec{s} \circ ((\matr{\beta} \circ \matr{M}) \vec{i})
+    \label{eq:matrix:theory}
+\end{align}
+where $\circ$ is the element-wise product (\ie: $(\matr{A} \circ \matr{B})_{ij} = A_{ij} B_{ij}$).
+
+Clearly $M_{ij}$ and $\beta_{ij}$ are not individually identifiable.
+Normally, $M_{ij}$ is assumed to be known.
+$\beta_{ij}$ is then estimated, although often with some simplifying assumptions.
+These could be that the probability of transmission is independent of age, and therefore all $\beta_{ij}$ are equal.
+Alternatively, that each stratum has a relative probability of being infected independent of who is infecting them, and therefore $\beta_{ij} = \beta \beta_j$.
+
+Now, we assume that $\beta_{ij} = \beta(t) \beta_{ij}$ where $\beta(t)$ is a non-strata-specific, time-varying component representing precautions taken by the population, for example masking; and $\beta_{ij}$ represents a constant probability of transmission from an infectious individual in strata $j$ to a susceptible individual in strata $i$ given contact, which could, for instance, be affected by differential susceptibility to infection for different age-groups.
+We arbitrarily choose one $\beta_{ij}$ to be equal to 1 for identifiability and hence the other $\beta_{ij}$ values are relative to this reference class.
+
+I assume that all individuals have the same \emph{infectiousness} but that there is a difference in \emph{susceptibility}.
+Infectiousness is how likely an individual is to infect others.
+Susceptibility is how likely an individual is to be infected.
+Therefore, $\beta_{ij} = \beta \beta_i$ where $\beta_i$ is the relative susceptibility of stratum $i$.
+I assume that the only difference is that children and adults have differing susceptibility.
+For identifiability, I set $\beta_i$ to 1 for adults and $\beta_i = \beta_c$ for children.
+
+\subsection{Geographic structure}
+
+The well-mixed population assumption is more plausible for smaller scales.
+Additionally, due to non-response bias, CIS is not geographically representative.
+Therefore, I stratify the population by region.
+
+I assume that each region is a well-mixed population and closed.
+A more complex model would allow for interaction between regions.
+However, for a large epidemic, non-interacting populations is a good approximation and can even provide better fits~\autocite{birrellRealtimea}.
+
+This is implemented by running the model separately and independently for each region.
+England is split into nine regions (previously known as Government Offices for the Regions): South East, London, North West, East of England, West Midlands, South West, Yorkshire and The Humber, East Midlands, and North East.
+
+Each element of the vector $\vec{s}$ refers to the proportion of that region's population of the relevant age group.
+The other vectors are interpreted similarly.
+
+\subsection{Time-varying behaviour} \label{SEIR:sec:time-varying-foi}
+
+During the pandemic, behaviour clearly changed over time.
+This was in response to government policy, also individual's perceived risk, and a variety of other factors.
+I introduce these changes by allowing both the contact matrix $\matr{M}$ and $\beta_{ij}$ to depend on time.
+
+How between-age-group contacts occur have previously been inferred from a combination of contact surveys, time use data, and mobility data~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}.
+I use the outputs from this analysis as $\matr{M}$.
+To summarise, baseline rates and contexts of contacts between age groups are taken from the contact survey POLYMOD~\autocite{mossongSocial}.
+How these contacts relate to time use is taken from the UK Time Use Survey.
+How time use changed over the pandemic is taken from the Google mobility data, which measured the locations of Android users over the period of interest.
+
+Some behavioural changes (\eg mask wearing) are not captured by the contact survey.
+To allow for these changes, $\beta$ is allowed to vary over time.
+In particular, $\beta$ follows a random walk in log space on the weekly timescale:
+\begin{align}
+    \beta(t) &= \beta_{w(t)} \\
+    \beta_1 &= 1 \\
+    \log\beta_{w+1} &= \log\beta_w + \sigma_\epsilon \epsilon_w \\
+    \epsilon_w &\sim N(0, 1)
+\end{align}
+where $w(t)$ is the week number of time $t$, with week 1 being the first week modelled.
+
+\section{Observation model} \label{SEIR:sec:observation}
+
+A variety of different approaches can be taken to link a transmission model to observations.
+Here, I extend the system of ODEs to include comparments representing PCR positive individuals.
+Whenever an individual is infected, a copy of them is placed in the pre-positive compartment, $p_0$.
+These individuals then transition into a set of compartment with their transition rates chosen to match the duration of positivity estimates in \cref{E-imperf-test}, $p_1, \dots, p_l$.
+The modelled proportion of individuals that are PCR positive in the can then be linked to the data using a Beta-Binomial likelihood.
+
+I use the method of \textcite{osogamiClosed} to determine $l$ and the transition rates between the compartments.
+The class of distributions which can be represented as a series of compartments are known as \emph{phase-type} distributions.
+
+\Textcite{osogamiClosed} define a subset of phase-type distributions which they refer to as Erlang-Coxian (EC) distributions (see following e).
+A $l$-phase EC distribution is a convoluation of a $(l-2)$-phase gamma distribution and two-phase acyclic, arbitrary phase-type distribution (with some additional constraints).
+The arrival times at the absorbing state (after the final state) is the duration.
+Therefore, the absorbing state is the PCR negative state, which does not need to be tracked.
+
+\Textcite{osogamiClosed} prove that EC distributions have the following properties:
+* Any distribution that can be well-matched (defined as having the same first three moments) by an acyclic
+phase-type distribution can be well-matched for some $P$
+which is a EC distribution.
+* The number of phases used by $P$ is at most one more than the optimal number.
+* The parameters of $P$ are available in closed form, and the expression for these is derived.
+
+A total of five parameters are required to specify a $l$-phase EC distribution.
+\begin{itemize}
+    \item $l$, the number of phases.
+    \item $\kappa_Y$, the rate parameter of the Erlang distribution.
+    \item $\kappa_{X1}$, the rate parameter of the first state after the Erlang distribution.
+    \item $\kappa_{X2}$, the rate parameter of the second state after the Erlang distribution.
+    \item $p_x$, the probability of moving from the first state after the Erlang distribution to the second (as opposed to directly to the absorbing state).
+\end{itemize}
+\begin{figure}
+\begin{tikzpicture}[
+    node distance = 2.5cm,
+    on grid,
+    auto,
+    ->,>=stealth',
+    every state/.style={draw,rectangle,very thick},
+    ]
+
+    \node[state] (1) {1};
+    \node[state, right of=1] (2) {2};
+    \node[state, right of=2, draw=none] (dots) {$\dots$};
+    \node[state, right of=dots] (lm2) {$l-2$};
+    \node[state, right of=lm2] (lm1) {$l-1$};
+    \node[state, right of=lm1] (l) {$l$};
+    \node[state, right of=l, align=center] (absorbing) {absorbing\\state};
+
+    \path (1) edge node {$\kappa_Y$} (2)
+          (2) edge node {$\kappa_Y$} (dots)
+          (dots) edge node {$\kappa_Y$} (lm2)
+          (lm2) edge node {$\kappa_Y$} (lm1)
+          (lm1) edge node {$p_x \kappa_{X1}$} (l)
+                edge [out=-45,in=-135] node[below] {$(1 - p_x) \kappa_{X1}$} (absorbing)
+          (l) edge node {$\kappa_{X2}$} (absorbing);
+\end{tikzpicture}
+\caption{A $l$ phase EC distribution.}
+\end{figure}
+
+\section{Full model} \label{SEIR:sec:full-model}
+
+\begin{figure}
+\begin{tikzpicture}[
+    node distance = 2cm
+    on grid,
+    auto,
+    ->,>=stealth',
+    every state/.style={draw,rectangle,very thick},
+    ]
+
+    \node[state] (S) {$s$};
+    \node[state, right of=S] (E1) {$e_1$};
+    \node[state, right of=E1] (E2) {$e_2$};
+    \node[state, right of=E2] (I1) {$i_1$};
+    \node[state, right of=I1] (I2) {$i_2$};
+    \node[state, right of=I2] (R) {$r$};
+
+    \path (S) edge node {$\beta$} (E1)
+          (E1) edge node {$\sigma$} (E2)
+          (E2) edge node {$\sigma$} (I1)
+          (I1) edge node {$\gamma$} (I2)
+          (I2) edge node {$\gamma$} (R);
+
+    \node[state, below=3cm of E1] (P0) {P0};
+    \node[state, right of=P0] (P1) {P1};
+    \node[state, right of=P1] (P2) {P2};
+    \node[state, right of=P2] (P3) {P3};
+    \node[state, right of=P3, draw=none] (Pdots) {...};
+
+    \path (P0) edge (P1)
+          (P1) edge (P2)
+          (P2) edge (P3);
+
+    \node[draw, thick, inner sep=0.3cm, fit=(P1) (P3), fill=blue!10,opacity=0.2] {};
+\end{tikzpicture}
+  \caption{SEIR model}
+\end{figure}
+
+The model I use for the CIS data is shown in figure X.
+
+The transmission model is a SEIR moel with two latent and two infectious states.
+It is based on \textcite{birrellRealtime}.
+
+The observation model contains one infected pre-positive state, and X infected positive state.
+
+I link the positive states to the data with a Beta-Binomial likelihood.
+
+\section{Inference and implementation} \label{SEIR:sec:inference-implementation}
+
+\subsection{MCMC}
+
+I use an adaptive random-walk Metropolis-Hastings algorithm.
+The algorithm is based on \textcite[algorithm 4]{andrieuTutorial}.
+Its implementation is adapted from \textcite{ghoshApproximate}.
+
+\subsection{Parameterisation}
+
+Following \textcite{birrellBayesian}, I parameterise the model in terms of the initial growth rate.
+
+\subsection{Priors}
+
+\subsection{Approximation of duration distribution}
+
+\section{Simulation study} \label{SEIR:sec:sim-study}
+
+\section{Application to CIS data} \label{SEIR:sec:application}
+
+\section{Discussion} \label{SEIR:sec:discussion}
+
+\section{Conclusion}
+
 \end{document}

### 2023-10-10 thesis.tex
Edits for CIS perfect testing chapter
Figure for ATACCC approximation
Add (messy) text
Currently pretty confused how to do this chapter,
just playing round with things hoping it sticks...
Start writing incidence/prevalence chapter
Lots of text on new model
Rewrite likelihood explanation
Edit chapter, removing appendix
Add files etc for SEIR
Fix title
Finish perfect testing chapter to dos
Finish initial chapter draft
Start of SEIR chapter

--- 2023-06-10 thesis.tex
@@
 \usepackage{amssymb}
+\usepackage{floatpag}
 \usepackage{bm}
@@
 % \usepackage{microtype}
+
+% Diagrams
+\usepackage{tikz}
+\usetikzlibrary{arrows,positioning,automata,fit}
 
@@
 \DeclareMathOperator{\prob}{\mathbb{P}}
+\DeclareMathOperator{\p}{\pi}
 \DeclareMathOperator{\var}{\mathbb{V}}
@@
 \DeclareMathOperator{\cov}{Cov}
+\DeclareMathOperator{\cor}{Cor}
 \DeclareMathOperator{\logit}{logit}
@@
 \DeclareMathOperator{\Beta}{Beta}
+\DeclareMathOperator{\Exponential}{Exponential}
 \DeclareMathOperator{\NegBin}{NegBin}
 \DeclareMathOperator{\GamDist}{Gamma}
-\def\matr{\bm}
-\def\set{\mathcal}
+\DeclareMathOperator{\MN}{Multinomial}
+\DeclareMathOperator{\expit}{expit}
+\newcommand\matr{\bm}
+\newcommand\set{\mathcal}
 \renewcommand{\vec}{\bm}
+\newcommand{\ssep}{:}
 
 % Thesis-specific maths commands
+\newcommand{\dmax}{d_\text{max}}
 \newcommand{\psens}{p_\text{sens}}
+\newcommand{\ntot}{n_\text{tot}}
+\newcommand{\ndet}{n_\text{d}}
+\newcommand{\nnodet}{n_\text{u}}
+\newcommand{\pnodet}{p_\text{u}}
+\newcommand{\Npop}{N_\text{pop}}
+\newcommand{\Ncis}{N_\text{CIS}}
+\newcommand{\ncis}{\vec{n_\text{CIS}}}
+\newcommand{\na}{\vec{n_\text{a}}}
+\newcommand{\pcis}{\vec{p_\text{CIS}}}
+\newcommand{\sched}{\mathbb{T}}
+\newcommand{\rate}{\mathfrak{R}}
 
@@
 
-\title{Estimating the incidence of SARS-CoV-2 from prevalence data}
+\title{Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys}
 \author{Joshua Blake}
@@
 
+% Only number labelled equations
+% \forlistloop workaround due to incompatability with biblatex
+% See: https://tex.stackexchange.com/questions/220268/biblatex-and-autonum-dont-work-together
+\usepackage{etoolbox} % get the good meaning of \forlistloop
+\let\etoolboxforlistloop\forlistloop % save the good meaning of \forlistloop
+\usepackage{autonum}
+\let\forlistloop\etoolboxforlistloop % restore the good meaning of \forlistloop
+\makeatletter
+\let\blx@noerroretextools\empty
+\makeatother
 \usepackage[backend=biber,style=authoryear,uniquename=false]{biblatex}
+\addbibresource{references.bib}
 
@@
 
+\section{Thesis outline}
+
+\subfile{incidence-prevalence}
+
+% We wish to estimate the distribution of time for which an individual is positive for.
+% An individual could have multiple infections, and hence I refer to \emph{infection episodes}.
+
+% I define an individual as \emph{detectable} if they would test positive if tested, in the absence of misclassification error (which is dealt with in \cref{E-imperf-test}).
+% Episode $i$'s \emph{duration of positivity}, $D_i$ is the number of days that they are detectable for; I use time discretised to days, therefore, the duration is at least 1 day.
+% Note that the relationship between positivity (\ie the result of a PCR test) and infectiousness (\ie an individual's ability to pass on the virus) is complex~\autocites{lascolaViral}{singanayagamDuration}, and beyond the scope of this thesis.
+% I define an episode as starting at time $B_i$ and ends at $E_i$, hence $D_i = E_i - B_i + 1$.
+
+% An episode is \emph{detected} when they test positive at least once, if this occurs; episode $i$ is detected at time $t$ if and only if: $i$ is detectable at $t$, $i$ is tested at $t$, and the test returns positive (a false negative does not occur).
+
+% The survival time function, parameterised by $\theta$, is the probability that an episode lasts at least $t$ days: $S_\theta(t) = \prob(D_i \geq t \mid B_i = b, \theta)$, where $D_i$ and $B_i$ are the duration and starting time of episode $i$ respectively.
+% I make the standard assumption that $D_i$ and $B_i$ are independent and hence: $S_\theta(t) = \prob(D_i \geq t \mid \theta)$.I utilise survival analysis for analysing this data.
+% Survival analysis is the area of statistics concerned with estimating the distribution of the times between events.
+% It is broadly applicable to many domains, such as the time-to-failure for a mechanical system or the effect of a treatment on time spent in hospital.
+% In the context of this thesis, the distribution to estimate is that of duration of positivity.
+% The \emph{initiating event} is when the duration starts, \ie becoming detectable.
+% The \emph{terminating event} is when the duration ends, \ie becoming no longer detectable.
+
+\chapter{Studies used in this thesis} \label{intro:sec:studies}
+
 \section{PCR testing} \label{intro:sec:PCR}
 
-\section{Incidence, prevalence, and duration}
+\section{Assessment of Transmission and Contagiousness of COVID-19 in Contacts}
 
-\section{Studies used in this thesis} \label{intro:sec:studies}
-
-\subsection{Assessment of Transmission and Contagiousness of COVID-19 in Contacts}
-
-\subsection{Coronavirus (COVID-19) Infection Survey} \label{intro:sec:cis}
+\section{Coronavirus (COVID-19) Infection Survey} \label{intro:sec:cis}
 
@@
 
-\chapter{Mechanistic modelling}
+\subfile{SEIR}
 
@@
 
-\subfile{appendix-cis-perfect-testing}
-
 \chapter{CIS episode definitions} \label{episode-def}

### 2023-10-11 cis-perfect-testing.tex
table of distributions with consistent notation

@@
 
-Now I derive an analytical solution to $\sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under the prior $\ntot \dist \NegBin(\mu, r)$ (using the mean/dispersion parameterisation of the negative binomial).
+Now I derive an analytical solution to $\sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under the prior $\ntot \dist \NB(\mu, r)$ (using the mean/dispersion parameterisation of the negative binomial).
 I assume that $\theta$, which are only the parameters of the survival distribution, is a priori independent of $\ntot$, the number of infections that occurred in the cohort.
@@
   &= 2\logit\lambda_t - \logit\lambda_{t-1} + \sigma \epsilon_t \\
-  \epsilon_t &\dist N(0, 1) &\text{for $t \geq 2$}  \\
+  \epsilon_t &\dist \N(0, 1) &\text{for $t \geq 2$}  \\
   \logit\lambda_2 &= \logit\lambda_1 + \epsilon_1 \\\
-  \epsilon_1 &\dist N(\mu_{\epsilon_1}, \sigma_{\epsilon_1}^2) \\
-  \logit \lambda_1 &\dist N(\mu_{\lambda_1}, \sigma_{\lambda_1}^2) \\
+  \epsilon_1 &\dist \N(\mu_{\epsilon_1}, \sigma_{\epsilon_1}^2) \\
+  \logit \lambda_1 &\dist \N(\mu_{\lambda_1}, \sigma_{\lambda_1}^2) \\
   \sigma &\dist \Exponential(1/\mu_\sigma).

### 2023-10-11 distributions.tex
table of distributions with consistent notation

\documentclass[thesis.tex]{subfiles}

\begin{document}
\begin{landscape}
\chapter{Tables of distributions}
\section{Continuous distributions}
\begin{tabular}{llp{3.5cm}lll}
Name & Notation & Parameters & Mean & Variance & pdf\\
Normal & $\N(\mu, \sigma^2)$ & $\mu$ (mean)\newline $\sigma^2$ (variance) & $\mu$ & $\sigma^2$ & $\frac{1}{\sqrt{2\pi\sigma^2}}e^{-\frac{(x-\mu)^2}{2\sigma^2}}$ \\
Lognormal & $\text{LN}(\mu, \sigma^2)$ & $\mu$ (mean of the logarithm)\newline $\sigma^2$ (variance of the logarithm) & $e^{\mu + \frac{\sigma^2}{2}}$ & $(e^{\sigma^2} - 1)e^{2\mu + \sigma^2}$ & $\frac{1}{x\sigma\sqrt{2\pi}}e^{-\frac{(\ln(x)-\mu)^2}{2\sigma^2}}$ \\
Gamma & $\GamDist(k, \theta)$ & $k$ (shape)\newline $\theta$ (scale) & $k\theta$ & $k\theta^2$ & $\frac{1}{\Gamma(k)\theta^k}x^{k-1}e^{-\frac{x}{\theta}}$ \\
Beta & $\Beta(\alpha, \beta)$ & $\alpha$ (shape)\newline $\beta$ (shape) & $\frac{\alpha}{\alpha+\beta}$ & $\frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}$ & $\frac{1}{B(\alpha,\beta)}x^{\alpha-1}(1-x)^{\beta-1}$ \\
Exponential & $\Exponential(\lambda)$ & $\lambda$ (rate) & $\frac{1}{\lambda}$ & $\frac{1}{\lambda^2}$ & $\lambda e^{-\lambda x}$ \\
\end{tabular}

\section{Discrete distributions}
\begin{tabular}{llp{3.5cm}lll}
Name & Notation & Parameters & Mean & Variance & pmf \\
Binomial & $\Bin(n, p)$ & $n$ (number of trials)\newline $p$ (probability of success) & $np$ & $np(1-p)$ & ${n \choose x}p^x(1-p)^{n-x}$ \\
Poisson & $\Poi(\lambda)$ & $\lambda$ (average rate) & $\lambda$ & $\lambda$ & $\frac{e^{-\lambda}\lambda^x}{x!}$ \\
Negative Binomial & $\NB(r, p)$ & $r$ (number of failures until stopping)\newline $p$ (probability of success) & $\frac{pr}{1-p}$ & $\frac{pr}{(1-p)^2}$ & ${x + r - 1 \choose x} p^x (1-p)^r$ \\
Beta-Binomial (centered) & $\BB(n, p, \rho)$ & $n$ (number of trials)\newline $p$ (success probability)\newline $\rho$ (overdispersion) & $np$ & $np(1-p)[1+(n-1)\rho]$ & ${n \choose x} \frac{B(x+\alpha, n-x+\beta)}{}$ \\
Multinomial & $\MN(n, \vec{p})$ & $n$ (number of trials)\newline $\vec{p}$ (probabilities) & $n \vec{p}$ & See below & $\frac{n!}{x_1!x_2!\ldots x_k!}p_1^{x_1}p_2^{x_2}\ldots p_k^{x_k}$ \\
\end{tabular}
\end{landscape}
\end{document}

### 2023-10-11 distributions.tex
consistent notation for distributions

@@
 \begin{document}
+\chapter{Tables of distributions}
 \begin{landscape}
-\chapter{Tables of distributions}
 \section{Continuous distributions}
@@
 Name & Notation & Parameters & Mean & Variance & pdf\\
-Normal & $\N(\mu, \sigma^2)$ & $\mu$ (mean)\newline $\sigma^2$ (variance) & $\mu$ & $\sigma^2$ & $\frac{1}{\sqrt{2\pi\sigma^2}}e^{-\frac{(x-\mu)^2}{2\sigma^2}}$ \\
-Lognormal & $\text{LN}(\mu, \sigma^2)$ & $\mu$ (mean of the logarithm)\newline $\sigma^2$ (variance of the logarithm) & $e^{\mu + \frac{\sigma^2}{2}}$ & $(e^{\sigma^2} - 1)e^{2\mu + \sigma^2}$ & $\frac{1}{x\sigma\sqrt{2\pi}}e^{-\frac{(\ln(x)-\mu)^2}{2\sigma^2}}$ \\
-Gamma & $\GamDist(k, \theta)$ & $k$ (shape)\newline $\theta$ (scale) & $k\theta$ & $k\theta^2$ & $\frac{1}{\Gamma(k)\theta^k}x^{k-1}e^{-\frac{x}{\theta}}$ \\
-Beta & $\Beta(\alpha, \beta)$ & $\alpha$ (shape)\newline $\beta$ (shape) & $\frac{\alpha}{\alpha+\beta}$ & $\frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}$ & $\frac{1}{B(\alpha,\beta)}x^{\alpha-1}(1-x)^{\beta-1}$ \\
+Normal & $\N(\mu, \sigma^2)$ & $\mu$ (mean)\newline $\sigma^2 > 0$ (variance) & $\mu$ & $\sigma^2$ & $\frac{1}{\sqrt{2\pi\sigma^2}}e^{-\frac{(x-\mu)^2}{2\sigma^2}}$ \\
+Lognormal & $\text{LN}(\mu, \sigma^2)$ & $\mu$ (mean of the logarithm)\newline $\sigma^2 > 0$ (variance of the logarithm) & $e^{\mu + \frac{\sigma^2}{2}}$ & $(e^{\sigma^2} - 1)e^{2\mu + \sigma^2}$ & $\frac{1}{x\sigma\sqrt{2\pi}}e^{-\frac{(\ln(x)-\mu)^2}{2\sigma^2}}$ \\
+Gamma & $\GamDist(\alpha, \beta)$ & $\alpha > 0$ (shape) \newline $\beta > 0$ (rate) & $\frac{\alpha}{\beta}$ & $\frac{\alpha}{\beta^2}$ & $\frac{\beta^\alpha}{\Gamma(\alpha)\theta}x^{\alpha-1}e^{-\beta x}$ \\
+Beta & $\BetaDist(\alpha, \beta)$ & $\alpha > 0$ (shape)\newline $\beta > 0$ (shape) & $\frac{\alpha}{\alpha+\beta}$ & $\frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}$ & $\frac{1}{B(\alpha,\beta)}x^{\alpha-1}(1-x)^{\beta-1}$ \\
 Exponential & $\Exponential(\lambda)$ & $\lambda$ (rate) & $\frac{1}{\lambda}$ & $\frac{1}{\lambda^2}$ & $\lambda e^{-\lambda x}$ \\
@@
 Name & Notation & Parameters & Mean & Variance & pmf \\
-Binomial & $\Bin(n, p)$ & $n$ (number of trials)\newline $p$ (probability of success) & $np$ & $np(1-p)$ & ${n \choose x}p^x(1-p)^{n-x}$ \\
-Poisson & $\Poi(\lambda)$ & $\lambda$ (average rate) & $\lambda$ & $\lambda$ & $\frac{e^{-\lambda}\lambda^x}{x!}$ \\
+Binomial & $\Bin(n, p)$ & $n$ (number of trials, integer)\newline $0 \leq p \leq 1$ (probability of success) & $np$ & $np(1-p)$ & ${n \choose x}p^x(1-p)^{n-x}$ \\
+Poisson & $\Poi(\lambda)$ & $\lambda > 0$ (average rate) & $\lambda$ & $\lambda$ & $\frac{e^{-\lambda}\lambda^x}{x!}$ \\
 Negative Binomial & $\NB(r, p)$ & $r$ (number of failures until stopping)\newline $p$ (probability of success) & $\frac{pr}{1-p}$ & $\frac{pr}{(1-p)^2}$ & ${x + r - 1 \choose x} p^x (1-p)^r$ \\
-Beta-Binomial (centered) & $\BB(n, p, \rho)$ & $n$ (number of trials)\newline $p$ (success probability)\newline $\rho$ (overdispersion) & $np$ & $np(1-p)[1+(n-1)\rho]$ & ${n \choose x} \frac{B(x+\alpha, n-x+\beta)}{}$ \\
-Multinomial & $\MN(n, \vec{p})$ & $n$ (number of trials)\newline $\vec{p}$ (probabilities) & $n \vec{p}$ & See below & $\frac{n!}{x_1!x_2!\ldots x_k!}p_1^{x_1}p_2^{x_2}\ldots p_k^{x_k}$ \\
+Beta-Binomial (centered) & $\BB(n, p, \rho)$ & $n$ (number of trials, integer)\newline $0 \leq p \leq 1$ (success probability)\newline $\rho > 0$ (overdispersion) & $np$ & $np(1-p)(1+n\rho)(1+\rho)^{-1}$ & ${n \choose x} \frac{B(x+p/\rho, n-x+(1-p)/\rho)}{B(p/\rho, (1-p)/\rho)}$ \\
+Multinomial & $\MN(n, \vec{p})$ & $n$ (number of trials, integer)\newline $0 < \vec{p}$ (probabilities, $\sum_i p_i = 1$) & $n \vec{p}$ & See below & $\frac{n!}{x_1!x_2!\ldots x_k!}p_1^{x_1}p_2^{x_2}\ldots p_k^{x_k}$ \\
 \end{tabular}
+$B(x,y)$ is the beta function, defined in terms of the Gamma function $\Gamma(x)$ as $B(x,y) = \frac{\Gamma(x)\Gamma(y)}{\Gamma(x+y)}$.
+For $X \dist \MN(n, \vec{p})$, $\V(X_i) = np_i(1-p_i)$ and $\cov(X_i, X_j) = -np_ip_j$ for $i \neq j$.
 \end{landscape}

### 2023-10-11 SEIR.tex
EC fitting details

@@
     \log\beta_{w+1} &= \log\beta_w + \sigma_\epsilon \epsilon_w \\
-    \epsilon_w &\sim N(0, 1)
+    \epsilon_w &\sim \N(0, 1)
 \end{align}
@@
 \Textcite{osogamiClosed} prove that EC distributions have the following properties:
-* Any distribution that can be well-matched (defined as having the same first three moments) by an acyclic
-phase-type distribution can be well-matched for some $P$
-which is a EC distribution.
-* The number of phases used by $P$ is at most one more than the optimal number.
-* The parameters of $P$ are available in closed form, and the expression for these is derived.
+\begin{itemize}
+    \item Any distribution that can be well-matched (defined as having the same first three moments) by an acyclic phase-type distribution can be well-matched for some $P$ which is an EC distribution.
+    \item The number of phases used by $P$ is at most one more than the optimal number.
+    \item The parameters of $P$ are available in closed form, and the expression for these is derived.
+\end{itemize}
 
@@
 
+I use the mapfit package~\autocite{mapfit} to compute $P$ for the posterior mean of the distribution estimated in \cref{E-imperf-test}.
+The parameter estimates are in \cref{SEIR:table:ec-params}.
+\begin{table}
+    \centering
+    \begin{tabular}{c c c c c}
+        $l$ & $\kappa_Y$ & $\kappa_{X1}$ & $\kappa_{X2}$ & $p_x$ \\
+        3 & 0.0820 & 0.126 & 0.0223 & 0.00973  \\
+    \end{tabular}
+    \caption{Parameter estimates for $P$, the EC distribution fit to the posterior mean of the duration distribution estimated in \cref{E-imperf-test}.}
+    \label{SEIR:table:ec-params}
+\end{table}
+% end table of ec params
+
+To link the model with prevalence measurements, I use a beta-binomial likelihood.
+If on each day $t$, each strata $i$ is observed to have $y_{it}$ positives out of $n_{it}$ tests on days $t$, then the likelihood is:
+\begin{align}
+    \prod_{i,t} \BB (y_{it} \mid n_{it}, \sum_{k=1}^l p_{ki}(t), \rho)
+\end{align}
+using the mean-dispersion parametrisation of the beta-binomial distribution.
+$\rho$ is a nuisance parameter controlling the dispersion of the beta-binomial distribution.
+
 \section{Full model} \label{SEIR:sec:full-model}
@@
 \begin{tikzpicture}[
-    node distance = 2cm
+    node distance = 2.5cm,
     on grid,
@@
     \node[state] (S) {$s$};
-    \node[state, right of=S] (E1) {$e_1$};
-    \node[state, right of=E1] (E2) {$e_2$};
-    \node[state, right of=E2] (I1) {$i_1$};
-    \node[state, right of=I1] (I2) {$i_2$};
-    \node[state, right of=I2] (R) {$r$};
+    \node[state, right=of S] (E1) {$e_1$};
+    \node[state, right=of E1] (E2) {$e_2$};
+    \node[state, right=of E2] (I1) {$i_1$};
+    \node[state, right=of I1] (I2) {$i_2$};
+    \node[state, right=of I2] (R) {$r$};
 
-    \path (S) edge node {$\beta$} (E1)
+    \path (S) edge node {$\sum_j \beta_{ij} i_j$} (E1)
           (E1) edge node {$\sigma$} (E2)
@@
 
-    \node[state, below=3cm of E1] (P0) {P0};
-    \node[state, right of=P0] (P1) {P1};
-    \node[state, right of=P1] (P2) {P2};
-    \node[state, right of=P2] (P3) {P3};
-    \node[state, right of=P3, draw=none] (Pdots) {...};
+    \node[state, below=3cm of E1] (P0) {$p_0$};
+    \node[state, right=of P0] (P1) {$p_1$};
+    \node[state, right=of P1] (P2) {$p_2$};
+    \node[state, right=of P2] (P3) {$p_3$};
+    \node[state, right=of P3, draw=none] (P4) {};
 
-    \path (P0) edge (P1)
-          (P1) edge (P2)
-          (P2) edge (P3);
+    \path  (S) edge node {$\sum_j \beta_{ij} i_j$} (P0)
+          (P0) edge node {$\kappa_0$} (P1)
+          (P1) edge node {$\kappa_Y$} (P2)
+          (P2) edge node {$p_x \kappa_{X1}$} (P3)
+               edge [out=-45,in=-135] node[below] {$(1 - p_x) \kappa_{X1}$} (P4)
+          (P3) edge node {$\kappa_{X2}$} (P4);
 
@@
 \end{tikzpicture}
-  \caption{SEIR model}
+  \caption[SEIR model]{SEIR model used for the analysis in this chapter. The shaded box represents PCR-positive individuals. Shown for a single strata $i$.}
 \end{figure}

### 2023-10-13 SEIR.tex
More SEIR chapter

@@
 
-Describing other mechanistic models is beyond the scope of this thesis.
-Models vary greatly in complexity and form, depending on the disease and the question of interest.
-I recommend \Textcite{kretzschmarMathematical} for an introduction to the breadth of the field or \Textcite{keelingModeling} for a book-length treatment.
-
 Notation in this chapter differs from the rest of the thesis.
@@
 
-Sections X\todo{which?} give the background on ODE-based mechanistic models.
-Section X\todo{which?} describes the model that I use for the CIS data, and section X the inference procedure.
-I verify the model and inference procedure in a simulation study in section X.
-I apply the model to the autumn 2020 CIS data in section X.
-Finally, section X discusses the results, and possible extensions, including those needed to extend to the 2021 period which includes vaccination.
+\Cref{SEIR:sec:transmission} gives the background on ODE-based mechanistic models.
+The literature on mechanistic models is vast, going back to at least \textcite{kermackContribution}.
+Here, I explain only the background necessary to understand the model used in this chapter.
+For further details I recommend the tutorial paper \textcite{kretzschmarMathematical} or the textbook \textcite{keelingModeling}.
+\Cref{SEIR:sec:observation} describes the novel observation model that I use for the CIS data.
+The model's implementation and fitting is explained in \cref{SEIR:sec:inference-implementation}.
+I verify the inference procedure in a simulation study in \cref{SEIR:sec:sim-study} before applying it to the autumn 2020 CIS data in \cref{SEIR:sec:application}.
+Finally, \cref{SEIR:sec:discussion} discusses the results, and possible extensions, including those needed to extend to the period including vaccination.
 
@@
 
-\subsection{Non-exponential waiting times}
+\subsection{Non-exponential waiting times} \label{SEIR:sec:non-exponential}
 
@@
 
-\subsection{Structured populations}
+\subsection{Structured populations} \label{SEIR:sec:structured-populations}
 
@@
 
-I assume that all individuals have the same \emph{infectiousness} but that there is a difference in \emph{susceptibility}.
+Based on prior work\todo{cite}, I assume that all individuals have the same \emph{infectiousness} but that there is a difference in \emph{susceptibility}.
 Infectiousness is how likely an individual is to infect others.
 Susceptibility is how likely an individual is to be infected.
-Therefore, $\beta_{ij} = \beta \beta_i$ where $\beta_i$ is the relative susceptibility of stratum $i$.
+Therefore, $\beta_{ij} = \beta_i$ where $\beta_i$ is the relative susceptibility of stratum $i$.
 I assume that the only difference is that children and adults have differing susceptibility.
@@
 
-\subsection{Geographic structure}
+\subsection{Geographic structure} \label{SEIR:sec:geography}
 
@@
 This was in response to government policy, also individual's perceived risk, and a variety of other factors.
-I introduce these changes by allowing both the contact matrix $\matr{M}$ and $\beta_{ij}$ to depend on time.
+I follow \textcite{birrellRealtime}, using a combination of Google moblity data and a time-varying transmission intensity to include these changes.
 
@@
     \beta(t) &= \beta_{w(t)} \\
-    \beta_1 &= 1 \\
     \log\beta_{w+1} &= \log\beta_w + \sigma_\epsilon \epsilon_w \\
@@
 where $w(t)$ is the week number of time $t$, with week 1 being the first week modelled.
+Two parameters are unspecified here: the initial random of the random walk, $\beta_1$, and the standard deviation of the random walk, $\sigma_\epsilon$.
+See \cref{perf-test:sec:parameters-priors} for the implementation and parameterisation of these.
 
@@
 \end{align}
-using the mean-dispersion parametrisation of the beta-binomial distribution.
-$\rho$ is a nuisance parameter controlling the dispersion of the beta-binomial distribution.
+using the mean-dispersion parametrisation of the beta-binomial distribution (see \cref{E-distributions}).
+$\rho$ is a nuisance parameter controlling the overdispersion of the beta-binomial distribution.
+At $\rho=0$, the beta-binomial coincides with the binomial distribution.
 
@@
   \caption[SEIR model]{SEIR model used for the analysis in this chapter. The shaded box represents PCR-positive individuals. Shown for a single strata $i$.}
+  \label{SEIR:fig:full-model}
 \end{figure}
 
-The model I use for the CIS data is shown in figure X.
+The full model combines each of the elements discussed in the previous sections.
+The transmission model follows \textcite{birrellRealtime}.
+It is a SEIR model with two latent and two infectious states (see \cref{SEIR:sec:non-exponential}), producing a gamma distribution.
+The population is stratified by age, using contact matrices to  (see \cref{SEIR:sec:structured-populations}).
+The force-of-infection and contact matrices changes over time as described in \cref{SEIR:sec:time-varying-foi}.
+The population in the model is that of one region of England.
+To form whole of England estimates, each region is modelled independently with the results then combined (see \cref{SEIR:sec:geography}).
 
-The transmission model is a SEIR moel with two latent and two infectious states.
-It is based on \textcite{birrellRealtime}.
+The observation model is novel, to relate the transmission to CIS data.
+It uses a phase-type model to represent the duration distribution of PCR positivity as explained in \cref{SEIR:sec:observation}.
 
-The observation model contains one infected pre-positive state, and X infected positive state.
-
-I link the positive states to the data with a Beta-Binomial likelihood.
+A diagram of the model is shown in figure \cref{SEIR:fig:full-model}.
+The system of differential equations is below.
+\begin{align}
+    \label{SEIR:eq:fullODEs}
+    \frac{d\vec{s}}{dt} &= -\vec{\Delta} \\
+    \frac{d\vec{e_1}}{dt} &= \vec{\Delta} - 2\sigma \vec{e_1} \\
+    \frac{d\vec{e_2}}{dt} &= 2\sigma \vec{e_1} - 2\sigma \vec{e_2} \\
+    \frac{d\vec{i_1}}{dt} &= 2\sigma \vec{e_2} - 2\gamma \vec{i_1} \\
+    \frac{d\vec{i_2}}{dt} &= 2\gamma \vec{i_1} - 2\gamma \vec{i_2} \\
+    \frac{d\vec{r}}{dt} &= 2\gamma \vec{i_2} \\
+    \frac{d\vec{p_0}}{dt} &= \vec{\Delta} - \kappa_0 \vec{p_0} \\
+    \frac{d\vec{p_1}}{dt} &= \kappa_0 \vec{p_0} - \kappa_Y \vec{p_1} \\
+    \frac{d\vec{p_2}}{dt} &= \kappa_Y \vec{p_1} - \kappa_{X1} \vec{p_2} \\
+    \frac{d\vec{p_3}}{dt} &= p_x \kappa_{X1} \vec{p_2} - \kappa_{X2} \vec{p_3}
+\end{align}
+where $\vec{\Delta} = \vec{s} \circ (\matr{\beta} \circ \matr{M}) (\vec{i_1} + \vec{i_2})$ is the vector of instantaneous incidence.
 
@@
 I use an adaptive random-walk Metropolis-Hastings algorithm.
-The algorithm is based on \textcite[algorithm 4]{andrieuTutorial}.
-Its implementation is adapted from \textcite{ghoshApproximate}.
+The proposal distribution is a multivariate normal that adapts to the covariance of the posterior.
+The adaption algorithm is based on \textcite[algorithm 4]{andrieuTutorial}, as implemented in \textcite{ghoshApproximate}.
+Proposals are accepted or rejected using the standard Metropolis-Hastings (MH) acceptance probability.
+See \cref{SEIR:MCMC-algorithm} for full details.
+\begin{algorithm}
+ set $\vec{X_0}$ to an initial value of the parameter vector \;
+ $\vec{\mu_0} = \vec{X_0}$ \;
+ $\matr{\Sigma_0} = \text{diag}(\vec{\mu_0})$ \;
+ $\lambda_0 = 1$ \;
+ \For{$i = 1, \dots, M$}{
+  sample $\vec{Y_}i \sim \N(\vec{\mu_{i-1}}, \lambda_{i-1}\matr{\Sigma_{i-1}})$\;
+  set $\vec{X_i}$ to $\vec{Y_i}$ or $\vec{X_{i-1}}$ using a MH acceptance step\;
+  update the proportion of proposals accepted so far, $\alpha$ \;
+  \eIf(\tcc{No adaptation for 200 iterations}){$i \leq 200$}{
+    $\lambda_i = \lambda_{i-1}$ \;
+    $\vec{\mu_i} = \vec{\mu_{i-1}}$ \;
+    $\matr{\Sigma_i} = \matr{\Sigma_{i-1}}$ \;
+   }{
+    $\gamma_i = (i - 200)^{-0.6}$ \;
+    $\log \lambda_i = \log \lambda_{i-1} + \gamma_i(\alpha - 0.234)$ \;
+    $\vec{\mu_i} = (1 - \gamma_i) \vec{\mu_{i-1}} + \gamma_i \vec{X_i}$ \;
+    $\matr{\Sigma_i} = (1 - \gamma_i) \matr{\Sigma_{i-1}} + \gamma_i (\vec{X_i} - \vec{\mu_i})(\vec{X_i} - \vec{\mu_i})^T$ \;
+  }
+ }
+ \caption{Algorithm for adaptive random-walk Metropolis-Hastings.
+$\vec{\mu_i}$ and $\matr{\Sigma_i}$ are the mean and covariance of the proposal distribution at iteration $i$.
+They adapt to the shape and location of the posterior.
+$\text{diag}(\vec{\mu_0})$ is the diagonal matrix with diagonal entries equal to $\vec{\mu_0}$.
+$\lambda_i$ is the scale parameter of the proposal distribution at iteration $i$.
+$\gamma_i$ is the learning rate, which determines how much adaptation occurs.
+$\gamma_i \to 0$ as $i \to \infty$ so the rate of adaptation is \emph{vanishing}.
+Vanishing adaptation is required for the algorithm to converge to the target distribution~\autocite[section 3]{andrieuTutorial}.}
+ \label{SEIR:MCMC-algorithm}
+\end{algorithm}
 
@@
 
-Following \textcite{birrellBayesian}, I parameterise the model in terms of the initial growth rate.
+Following \textcite{birrellBayesian}, I reparameterise the model to improve identifiability.
+Rather than directly inferring $\beta_1$, I infer the initial growth rate of the epidemic.
+I use a parsimonious parameterisation of the initial conditions using the initial proportion of the population that are infected and the initial proportion that are susceptible only.
+Both of these rely on an assumption that the epidemic is in the steady-state exponential growth phase at time 0.
+
+For an age-structured model, as the one here, the initial phase of the epidemic, when most individuals are susceptible, is approximately exponential.
+In fact, all the $\vec{e}$, $\vec{i}$ and $\vec{p}$ compartments grow at the same initial exponential growth rate, $\psi$.
+The basic reproduction number, $\R$ of the epidemic is related to this growth rate as follows~\autocites{birrellRealtimea}{wearingAppropriate}:
+\begin{align}
+    \R = \frac{\psi \left( \frac{\psi}{2\sigma} + 1 \right)^2}{\gamma \left(1 - \frac{1}{\left(\frac{\psi}{2 \gamma} + 1 \right)^2} \right)} \label{SEIR:eq:rtoR}.
+\end{align}
+
+$\R$ can be linked to $\beta_1$ because $\R$ is also equal to the dominant eigenvalue of the next-generation matrix, $\matr{G} = \matr\beta \circ \matr{M}$~\autocites{diekmannDefinition}{birrellRealtimea}.
+Therefore, the following procedure finds $\beta_1$ such that the growth rate $\psi$ is achieved.
+First calculate $\R$ from the model parameters using \cref{SEIR:eq:rtoR}.
+Denote this $\hat{R}$.
+Then calculate $R^* = D(\matr{G})$ when $\beta_1=1$, where $D(\cdot)$ gives the dominant eigenvalue.
+Note that multiplying a matrix by a scalar multiplies its eigenvalues by that same scalar.
+Finally, set $\beta_1 = \hat{R} / R^*$.
+
+To form a parsimonious parameterisation of the initial state, first focus on the initial number of infected individuals.
+In steady-state exponential growth, the relative proportion of each strata that is infected is proportional to the dominant eigenvector of $\matr{G}$, $\vec{d}$.
+I assume this is the case at time 0.
+Let the proportion of the population in the $i$ states at time 0 in the strata with the highest proportion infected, $i^+$, be a model parameter.
+Then, $\vec{i_0} = \vec{i_1} + \vec{i_2} = \frac{i^+}{\max \vec{d}} \vec{d}$.
+I introduce an additional vector of parameters, $\vec{\pi}$, where $\pi_i$ is the proportion of strata $i$ without immunity at time 0.
+That is, at time 0, $\vec{r} = 1 - \vec{\pi}$.
+
+The system of ODEs and the result that all the comparments are growing at the same rate give the following for the initial state of the compartments:
+\begin{align}
+    \vec{i_1} &= \vec{i_0} \left(1 + \frac{2\gamma}{\psi + 2\gamma} \right)^{-1} \\
+    \vec{i_2} &= \vec{i_0} - \vec{i_1} \\
+    \vec{e_2} &= \vec{i_1} \frac{\psi + 2\gamma}{2\sigma} \\
+    \vec{e_1} &= \vec{e_2} \frac{\psi + 2\sigma}{2\sigma} \\
+    \vec{s} &= \vec{\pi} - \vec{i_0} - \vec{e_1} - \vec{e_2} \\
+    \vec{p_0} &= \vec{e_1} \frac{\psi + 2\sigma}{\psi + \kappa_0} \\
+    \vec{p_1} &= \vec{p_0} \frac{\kappa_0}{\psi + \kappa_Y} \\
+    \vec{p_2} &= \vec{p_1} \frac{\kappa_Y}{\psi + \kappa_{X1}} \\
+    \vec{p_3} &= \vec{p_2} \frac{p_x \kappa_{X1}}{\psi + \kappa_{X2}}.
+\end{align}
 
 \subsection{Priors}
+The priors for most parameters are shown in \cref{SEIR:table:priors}.
+The parameters governing the duration of positivity ($\kappa$s) were explained previous (see \cref{SEIR:table:ec-params}).
 
-\subsection{Approximation of duration distribution}
+\begin{landscape}
+\begin{table}
+\begin{tabular}{l c l l}
+    Parameter description & Symbol & Distribution & Comment \\
+    \hline \\
+    Mean latent period & $1/\sigma$ & 3.5 days (fixed) & From a paper \\
+    Mean infectious period & $1/\gamma$ & 3.5 days (fixed) & From a paper \\
+    Initial proportion without immunity & $\pi$ & See \cref{SEIR:table:immunity} & \\
+    Initial proportion infected & $i^+$ & $\BetaDist(0.5, 1000)$ & Weakly informative \\
+    Relative susceptibility of children & $\beta_c$ & $\LN(-4.325, 0.1174)$ &  \\
+    Initial growth rate & $\psi$ & $\N(0.048, 0.0035)$ & Weakly informative, centered on growth rate of deaths \\
+    Standard deviation of random walk & $\sigma$ & $\Exponential(80)$ & Weakly informative, NCP \\
+    Overdispersion of observations & $\rho$ & $\Exponential(2 \times 10^5)$ & Weakly informative
+
+\end{tabular}
+\todo[inline]{Find citation for children, check growth rate, check what NCP is}
+\caption[SEIR model priors]{Priors for each parameter. For details of the distributions and their parameterisations see \cref{E-distributions}.}
+\label{SEIR:table:priors}
+\end{table}
+\end{landscape}
+
+\subsection{Solving the systems of ODEs}
 

### 2023-10-13 distributions.tex
More SEIR chapter

@@
 \begin{document}
-\chapter{Tables of distributions}
+\chapter{Tables of distributions} \label{E-distributions}
 \begin{landscape}
@@
 Normal & $\N(\mu, \sigma^2)$ & $\mu$ (mean)\newline $\sigma^2 > 0$ (variance) & $\mu$ & $\sigma^2$ & $\frac{1}{\sqrt{2\pi\sigma^2}}e^{-\frac{(x-\mu)^2}{2\sigma^2}}$ \\
-Lognormal & $\text{LN}(\mu, \sigma^2)$ & $\mu$ (mean of the logarithm)\newline $\sigma^2 > 0$ (variance of the logarithm) & $e^{\mu + \frac{\sigma^2}{2}}$ & $(e^{\sigma^2} - 1)e^{2\mu + \sigma^2}$ & $\frac{1}{x\sigma\sqrt{2\pi}}e^{-\frac{(\ln(x)-\mu)^2}{2\sigma^2}}$ \\
+Lognormal & $\LN(\mu, \sigma^2)$ & $\mu$ (mean of the logarithm)\newline $\sigma^2 > 0$ (variance of the logarithm) & $e^{\mu + \frac{\sigma^2}{2}}$ & $(e^{\sigma^2} - 1)e^{2\mu + \sigma^2}$ & $\frac{1}{x\sigma\sqrt{2\pi}}e^{-\frac{(\ln(x)-\mu)^2}{2\sigma^2}}$ \\
 Gamma & $\GamDist(\alpha, \beta)$ & $\alpha > 0$ (shape) \newline $\beta > 0$ (rate) & $\frac{\alpha}{\beta}$ & $\frac{\alpha}{\beta^2}$ & $\frac{\beta^\alpha}{\Gamma(\alpha)\theta}x^{\alpha-1}e^{-\beta x}$ \\

### 2023-10-16 SEIR.tex
Finish chapter draft 1

@@
 
-Mechanistic models are...
-
-This chapter focuses on mechanistic models based on ordinary differential equations (ODEs).
-They are useful because...
+Mechanistic models explicitly represent the process of how an infectious disease spreads.
+Infectious diseases have inherent feedback loops: current infectious individuals generate future infectious individuals.
+Explicitly representing this system can generate insights not possible with the statistical methods used in the previous chapter.
 
@@
 I adopt the conventions and notation standard in the infectious disease modelling field.
-In particular, upper-case letters denote populations, and lower-case letters denote proportions.
-Whether a variable is random or not is explictly stated if not clear from the context.
+In particular, upper-case letters denote numbers of people, and lower-case letters denote proportions.
+Whether a variable is random or not should be clear from the context, or will be explicitly stated if needed.
 
-\Cref{SEIR:sec:transmission} gives the background on ODE-based mechanistic models.
+The first sections of this chapter give the background and methodology of the model I use.
 The literature on mechanistic models is vast, going back to at least \textcite{kermackContribution}.
@@
 For further details I recommend the tutorial paper \textcite{kretzschmarMathematical} or the textbook \textcite{keelingModeling}.
-\Cref{SEIR:sec:observation} describes the novel observation model that I use for the CIS data.
-The model's implementation and fitting is explained in \cref{SEIR:sec:inference-implementation}.
-I verify the inference procedure in a simulation study in \cref{SEIR:sec:sim-study} before applying it to the autumn 2020 CIS data in \cref{SEIR:sec:application}.
-Finally, \cref{SEIR:sec:discussion} discusses the results, and possible extensions, including those needed to extend to the period including vaccination.
 
@@
 
-This section first introduces a simple ODE-based model.
-ODE-based models are \emph{deterministic}.
-Deterministic models have the property that once the parameters are set, there is no randomness in the transmission model.
+This section gradually builds complexity into a mechanistic model
+The class of models considered here is \emph{compartmental} models.
+A compartmental model divides all members of the population into compartments, each representing a disease state.
+Compartmental models are the most common type of mechanistic infectious disease model.
+They can be \emph{deterministic} or \emph{stochastic}.
+
+Deterministic models have the property that once the parameters are set, there is no stochasticity in the transmission model.
+All other model variables are deterministic functions of the parameters.
 The transmission model corresponds to the infection process (see \cref{E-inc-prev:sec:infection-process}).
@@
 
+The other type of model is \emph{stochastic}.
+Stochastic models have randomness in the transmission model, and often use discrete time.
+For example, the number of individuals infected on day $t+1$, conditional on the model state on day $t$, might be a Poisson distribution.
+The mean of the distribution is a function of the model state on day $t$.
+
+I will use a deterministic model, specifically an ODE model
 ODE models are mathematically described by a system of ODEs.
 The derivative is taken with respect to time, $t$.
-Therefore, in an ODE model, time is necessarily continuous.
-However, time is discretised when solving (see \cref{SEIR:sec:inference-implementation})
+Therefore, in an ODE model, time is necessarily continuous although discretised when solving (see \cref{SEIR:sec:inference-implementation})
+The number or proportion of individuals in each compartment is also continuous, which is a reasonable approximation when the population is large.
+For determinstic models, equivalent fomulations exist for the proportion or number of individuals.
 
-The number of individuals is also continuous.
-This is a reasonable approximation when the population is large.
-
-The other type of model is \emph{stochastic}.
-Stochastic models have randomness in the transmission model, and often use discrete time.
-For example, the number of individuals infected on day $t+1$, conditional on the model state on day $t$ might be a Poisson distribution.
 When the number of infected individuals is large, then deterministic models are a good approximation to stochastic models.
@@
 Further discussion of stochastic models is beyond the scope of this thesis.
-
-The class of models considered here is \emph{compartmental} models.
-A compartmental model divides all members of the population into compartments, each representing a disease state.
-Compartmental models are the most common type of mechanistic infectious disease model.
-Compartmental models can be deterministic or stochastic.
+\todo{maybe add a reference to a review or textbook on stochastic models}
 
 \subsection{The SIR model}
+\begin{figure}[h]
+\begin{tikzpicture}[
+    node distance = 2.5cm,
+    on grid,
+    auto,
+    ->,>=stealth',
+    every state/.style={draw,rectangle,very thick},
+    ]
+
+    \node[state] (S) {$s$};
+    \node[state, right=of S] (I) {$i$};
+    \node[state, right=of I] (R) {$r$};
+
+    \path (S) edge node {$\beta i$} (I)
+          (I) edge node {$\gamma$} (R);
+\end{tikzpicture}
+  \caption[The SIR model]{Schematic of the basic SIR model.}
+  \label{SEIR:fig:SIR}
+\end{figure}
 
 The simplest compartmental model is the \emph{susceptible-infected-recovered} (SIR) model.
-The name refers to the three compartments in the model.
+The name refers to the three compartments in the model, displayed in \cref{SEIR:fig:SIR}.
 Individuals in the susceptible compartment can be infected.
@@
 
-The term $\beta si$ is the instantaneous number of individuals infected.
-That is, the number of individuals moving from the susceptible state to the infected state.
+The term $\beta si$ is the instantaneous proportion of the population becoming infected.
+That is, the proportion of the population moving from the susceptible state to the infected state.
 This term can be arrived at by considering a single infected individual.
-Assume this individual makes contact with other individuals in the population at a constant rate.
-Further, assume that the population is \emph{well-mixed}, that is that the individual is equally likely to come into contact with any other individual.
-Therefore, the probability that the individual comes each of these contacts is a susceptible individual is $s$.
-Finally, assume that there is a constant probability of transmission upon contact between an infected and susceptible individual.
-The two constant terms here (number of contacts and probability of transmission) are absorbed into the transmission rate, $\beta$.
+Assume this individual makes contact with other individuals in the population at a constant rate, $c$, in units of contacts per day.
+Further, assume the population is \emph{well-mixed}, meaning the individual is equally likely to come into contact with any other individual.
+For a randomly selected individual, the probability that they are susceptible is $s$.
+Each of the infected individual's contacts is therefore with a susceptible individual with probability $s$, by the well-mixed assumption.
+Therefore, in expectation, the infected individual makes contact with susceptible individuals at a rate of $cs$.
+Assume that there is a constant probability of transmission upon contact between an infected and susceptible individual, $p_{si}$.
+The expected rate of infections caused by this infected individual is therefore $cs p_{si}$.
+The two constant terms here (number of contacts and probability of transmission) are absorbed into the transmission rate, $\beta = c p_{si}$.
 Therefore, the instantaneous number of infections per infected individual is $\beta s$.
@@
 
-The term $\gamma i$, the instantaneous number of individuals recovering, follows directly from the definitions of $\gamma$ and $i$.
+The term $\gamma i$, the instantaneous proportion of the population recovering, follows directly from the definitions of $\gamma$ and $i$.
 
@@
 \subsection{Non-exponential waiting times} \label{SEIR:sec:non-exponential}
+\begin{figure}[h]
+\begin{tikzpicture}[
+    node distance = 2.5cm,
+    on grid,
+    auto,
+    ->,>=stealth',
+    every state/.style={draw,rectangle,very thick},
+    ]
+
+    \node[state] (S) {$s$};
+    \node[state, right=of S] (I1) {$i_1$};
+    \node[state, right=of I1] (I2) {$i_2$};
+    \node[state, right=of I2, draw=none] (I3) {$\cdots$};
+    \node[state, right=of I3] (I4) {$i_n$};
+    \node[state, right=of I4] (R) {$r$};
+
+    \path (S) edge node {$\beta i$} (I1)
+          (I1) edge node {$n\gamma$} (I2)
+          (I2) edge node {$n\gamma$} (I3)
+          (I3) edge node {$n\gamma$} (I4)
+          (I4) edge node {$n\gamma$} (R);
+\end{tikzpicture}
+  \caption[The SIR model with non-exponential infectious period]{Schematic of the SIR model with non-exponential infectious period. The infectious period is modified by having multiple I states.}
+  \label{SEIR:fig:SIR-gamma}
+\end{figure}
 
@@
 A standard extension to the SIR model is to allow the infectious period to follow a gamma distribution.
-This is done by adding multiple infected compartments.
-The system can now be written as follows
+This is done by adding multiple infected compartments (see \cref{SEIR:fig:SIR-gamma}).
+The system can now be written as follows:
 \begin{align}
@@
 \subsection{The SEIR model}
+\begin{figure}[h]
+\begin{tikzpicture}[
+    node distance = 2cm,
+    on grid,
+    auto,
+    ->,>=stealth',
+    every state/.style={draw,rectangle,very thick},
+    remember picture, overlay
+    ]
+
+    \node[state, xshift=-2cm] (S) {$s$};
+    \node[state, right=of S] (E1) {$e_1$};
+    \node[state, right=of E1] (E2) {$e_2$};
+    \node[state, right=of E2, draw=none] (E3) {$\cdots$};
+    \node[state, right=of E3] (E4) {$e_m$};
+    \node[state, right=of E4] (I1) {$i_1$};
+    \node[state, right=of I1] (I2) {$i_2$};
+    \node[state, right=of I2, draw=none] (I3) {$\cdots$};
+    \node[state, right=of I3] (I4) {$i_n$};
+    \node[state, right=of I4] (R) {$r$};
+
+    \path (S) edge node {$\beta i$} (E1)
+          (E1) edge node {$m\sigma$} (E2)
+          (E2) edge node {$m\sigma$} (E3)
+          (E3) edge node {$m\sigma$} (E4)
+          (E4) edge node {$m\sigma$} (I1)
+          (I1) edge node {$n\gamma$} (I2)
+          (I2) edge node {$n\gamma$} (I3)
+          (I3) edge node {$n\gamma$} (I4)
+          (I4) edge node {$n\gamma$} (R);
+\end{tikzpicture}
+  \caption[The SEIR model]{Schematic of the SEIR model.}
+  \label{SEIR:fig:SEIR}
+\end{figure}
 
@@
 This is also known as an \emph{exposed} compartment.
-Therefore, we now form the susceptible-exposed-infected-recovered (SEIR) model.
+Therefore, we now form the susceptible-exposed-infected-recovered (SEIR) model, shown in \cref{SEIR:fig:SEIR}.
 
@@
 
-\todo[inline]{Introduction about the importance of age and mixing patterns, maybe show a contact matrix}
+\todo[inline]{Possibly add a paragraph about the importance of age and mixing patterns, maybe show a contact matrix?}
 
@@
 To incorporate the strata into the model, first I introduce some notation.
-denote the set of strata, normally age groups, as $\set{A}$. $N_i$ notates the population in stratum $i$ and hence $s_i N_i$ is the number of susceptible people in stratum $i$.
-The other compartments are notated similarly.
+Denote the set of strata, normally age groups, as $\set{A}$. $N_i$ notates the total population of stratum $i$ and hence $s_i N_i$ is the number of susceptible people in stratum $i$.
 
-We want to derive a matrix $\matr{B}$ which we can use to calculate the instantaneous number of new infections.
-We do this by considering a susceptible individual $a$ in stratum $i$.
-I use the notation $\rate(X)$ to mean the rate at which the event $X$ occur.
-\begin{align}
-    &\rate(\text{Individual $a$ infected}) \\
-    &\approx \sum_{j \in \set{A}} \rate(\text{Individual $a$ is infected by an individual in stratum $j$}) \\
-    &\approx \sum_{j \in \set{A}} \rate(\text{Contacts $a$ has with individuals in stratum $j$}) \\ &\, \times \prob(\text{$a$ infected} \mid \text{contact between $a$ and individual from stratum $j$})\\
-    &= \sum_{j \in \set{A}} M_{ij} \beta_{ij} i_j
-\end{align}
-where $M_{ij}$ is the instantaneous number of contacts an individual in strata $i$ has with individuals in stratum $j$ (assumed homogeneous mixing within strata); and $\beta_{ij}$ is the probability of transmission from an infectious individual in stratum $j$ to a susceptible individual in stratum $i$ given a contact between these individuals.
-Note that, unlike previously, here we have explicitly separated out the number of contacts $M_{ij}$ from the probability of transmission $\beta_{ij}$.
-The first two approximations are from mass-action dynamics which are valid as long as the transmission probability is small, meaning that we can ignore that the events are not mutually exclusive.
-Therefore, the expected rate of new infections in strata $i$ is $s_i N_i \sum_{j \in \set{A}} M_{ij} \beta_{ij} i_j$, which is equivalent to:
+I will derive a matrix which can be used to replace the scalar $\beta$ in the single stratum case.
+The derivation is similar to the single stratum case, except that the contact rates and probabilities of transmission will be allowed to vary for each pair of strata.
+Consider an infected individual in stratum $j$.
+Assume they make contacts with individuals in stratum $i$ at a rate of $M_{ij}$ per day.
+Therefore, they make contact with susceptible individuals in stratum $i$ at a rate of $s_i M{ij}$ per day.
+Assume that the probability of transmission from an infected individual in stratum $j$ to a susceptible individual in stratum $i$ upon contact is $\beta_{ij}$.
+Therefore, $M_{ij} \beta_{ij} s_i$ is the instantaneous proportion of individuals in stratum $i$ infected by an individual in stratum $j$.
+Considering infected individuals across all strata, $\sum_{j \in \set{A}} M_{ij} \beta_{ij} s_i$ is the instantaneous proportion of individuals in stratum $i$ infected.
+
 \begin{align}
@@
 Clearly $M_{ij}$ and $\beta_{ij}$ are not individually identifiable.
-Normally, $M_{ij}$ is assumed to be known.
-$\beta_{ij}$ is then estimated, although often with some simplifying assumptions.
-These could be that the probability of transmission is independent of age, and therefore all $\beta_{ij}$ are equal.
-Alternatively, that each stratum has a relative probability of being infected independent of who is infecting them, and therefore $\beta_{ij} = \beta \beta_j$.
+Normally, $M_{ij}$ is assumed to be known, for example from a previous contact survey~\autocite[such as][]{mossongSocial}.
+$\beta_{ij}$ is then estimated, although often with a parsimonious model.
+A common assumption is that $\beta_{ij}$ can be decomposed into two independent components such that $\beta_{ij} = \beta \beta^\text{susc}_{i} \beta^\text{inf}_j$.
+For any stratum $i$, $\beta^\text{susc}_i$ is known as the \emph{relative susceptibility} of stratum $i$.
+Informally, it is how vulnerable to infection individuals in stratum $i$ are.
+Formally, it is the probability that a susceptible individual in stratum $i$ will be infected upon contact with an infectious individual, relative to a individual in the reference stratum.
+Similarly, $\beta^\text{inf}_i$ is known as the \emph{relative infectiousness} of stratum $i$.
+Informally, it is how likely individuals in stratum $i$ are to pass on the infection.
+Formally, it is the probability that an infectious individual in stratum $i$ will infect a susceptible individual upon contact, relative to an individual in the reference stratum.
+These measures could be biological, \eg due to differences in immune systems, or behavioural, \eg if infants are less hygenic than adults.
 
-Now, we assume that $\beta_{ij} = \beta(t) \beta_{ij}$ where $\beta(t)$ is a non-strata-specific, time-varying component representing precautions taken by the population, for example masking; and $\beta_{ij}$ represents a constant probability of transmission from an infectious individual in strata $j$ to a susceptible individual in strata $i$ given contact, which could, for instance, be affected by differential susceptibility to infection for different age-groups.
-We arbitrarily choose one $\beta_{ij}$ to be equal to 1 for identifiability and hence the other $\beta_{ij}$ values are relative to this reference class.
+%Now, we assume that $\beta_{ij} = \beta(t) \beta_{ij}$ where $\beta(t)$ is a non-strata-specific, time-varying component representing precautions taken by the population, for example masking; and $\beta_{ij}$ represents a constant probability of transmission from an infectious individual in strata $j$ to a susceptible individual in strata $i$ given contact, which could, for instance, be affected by differential susceptibility to infection for different age-groups.
+%We arbitrarily choose one $\beta_{ij}$ to be equal to 1 for identifiability and hence the other $\beta_{ij}$ values are relative to this reference class.
 
 Based on prior work\todo{cite}, I assume that all individuals have the same \emph{infectiousness} but that there is a difference in \emph{susceptibility}.
-Infectiousness is how likely an individual is to infect others.
-Susceptibility is how likely an individual is to be infected.
-Therefore, $\beta_{ij} = \beta_i$ where $\beta_i$ is the relative susceptibility of stratum $i$.
-I assume that the only difference is that children and adults have differing susceptibility.
-For identifiability, I set $\beta_i$ to 1 for adults and $\beta_i = \beta_c$ for children.
+Therefore, $\beta^\text{inf}_{j} = 1$ for all $j$.
+The prior literature suggests that children are less susceptible than adults, but otherwise there is little difference in susceptibility.
+Therefore, I set $\beta^\text{susc}_i$ to 1 for adult-aged strata and $\beta^\text{susc}_i = \beta_c$ for child-aged strata.
 
@@
 
-The well-mixed population assumption is more plausible for smaller scales.
+The well-mixed population assumption is implausible for geographically dispersed populations.
 Additionally, due to non-response bias, CIS is not geographically representative.
@@
 A more complex model would allow for interaction between regions.
-However, for a large epidemic, non-interacting populations is a good approximation and can even provide better fits~\autocite{birrellRealtimea}.
+However, for a large epidemic, non-interacting populations is a good approximation and can even fit the data better~\autocite{birrellRealtimea}.
 
-This is implemented by running the model separately and independently for each region.
+I implement this by running the model separately and independently for each region.
 England is split into nine regions (previously known as Government Offices for the Regions): South East, London, North West, East of England, West Midlands, South West, Yorkshire and The Humber, East Midlands, and North East.
@@
 The other vectors are interpreted similarly.
+For simplicity, I do not introduce additional notation for the regions as it is clear from context which region is being referred to.
+Vectors referring to different regions will never be referred to in the same expression because the regions are modelled independently without any interaction.
 
@@
 To summarise, baseline rates and contexts of contacts between age groups are taken from the contact survey POLYMOD~\autocite{mossongSocial}.
-How these contacts relate to time use is taken from the UK Time Use Survey.
-How time use changed over the pandemic is taken from the Google mobility data, which measured the locations of Android users over the period of interest.
+How these contacts relate to time use is taken from the UK Time Use Survey\todo{cite time use survey}.
+How the population spent their time each week during the pandemic is taken from the Google mobility data\todo{cite mobility data}, which measured the locations of Android users over the period of interest.
+\todo[inline]{is the above paragraph too hand wavy?}
 
@@
 These individuals then transition into a set of compartment with their transition rates chosen to match the duration of positivity estimates in \cref{E-imperf-test}, $p_1, \dots, p_l$.
-The modelled proportion of individuals that are PCR positive in the can then be linked to the data using a Beta-Binomial likelihood.
+These compartments do not have an interpretation.
+Their only purpose is to give the time spent in the PCR positive compartments the desired distribution.
 
@@
 
-\Textcite{osogamiClosed} define a subset of phase-type distributions which they refer to as Erlang-Coxian (EC) distributions (see following e).
-A $l$-phase EC distribution is a convoluation of a $(l-2)$-phase gamma distribution and two-phase acyclic, arbitrary phase-type distribution (with some additional constraints).
-The arrival times at the absorbing state (after the final state) is the duration.
-Therefore, the absorbing state is the PCR negative state, which does not need to be tracked.
+\Textcite{osogamiClosed} define a subset of phase-type distributions which they refer to as Erlang-Coxian (EC) distributions.
+EC distributions are convenient because they are flexible enough to approximate any distribution, and have a closed form expression for the parameters.
 
-\Textcite{osogamiClosed} prove that EC distributions have the following properties:
-\begin{itemize}
-    \item Any distribution that can be well-matched (defined as having the same first three moments) by an acyclic phase-type distribution can be well-matched for some $P$ which is an EC distribution.
-    \item The number of phases used by $P$ is at most one more than the optimal number.
-    \item The parameters of $P$ are available in closed form, and the expression for these is derived.
-\end{itemize}
+A $l$-phase EC distribution is a convoluation of a $(l-2)$-phase gamma distribution and a two-phase acyclic, arbitrary phase-type distribution.
+The time between entering the 1st state and the $l+1$th state, the absorbing state, is the duration.
+A schematic of an EC distribution and its parameters is shown in \cref{SEIR:fig:EC}
 
-A total of five parameters are required to specify a $l$-phase EC distribution.
-\begin{itemize}
-    \item $l$, the number of phases.
-    \item $\kappa_Y$, the rate parameter of the Erlang distribution.
-    \item $\kappa_{X1}$, the rate parameter of the first state after the Erlang distribution.
-    \item $\kappa_{X2}$, the rate parameter of the second state after the Erlang distribution.
-    \item $p_x$, the probability of moving from the first state after the Erlang distribution to the second (as opposed to directly to the absorbing state).
-\end{itemize}
 \begin{figure}
@@
 \caption{A $l$ phase EC distribution.}
+\label{SEIR:fig:EC}
 \end{figure}
+
+\Textcite{osogamiClosed} prove that EC distributions have the following properties:
+\begin{itemize}
+    \item Any distribution that can be well-matched (defined as having the same first three moments) by an acyclic phase-type distribution can be well-matched for some $P$ which is an EC distribution.
+    \item The number of phases used by $P$ is at most one more than the optimal number.
+    \item The parameters of $P$ are available in closed form, and the expression for these is derived.
+\end{itemize}
+
+A total of five parameters are required to specify a $l$-phase EC distribution.
+\begin{itemize}
+    \item $l$, the number of phases.
+    \item $\kappa_Y$, the rate parameter of the Erlang distribution.
+    \item $\kappa_{X1}$, the rate parameter of the first state after the Erlang distribution.
+    \item $\kappa_{X2}$, the rate parameter of the second state after the Erlang distribution.
+    \item $p_x$, the probability of moving from the first state after the Erlang distribution to the second (as opposed to directly to the absorbing state).
+\end{itemize}
 
@@
     \end{tabular}
-    \caption{Parameter estimates for $P$, the EC distribution fit to the posterior mean of the duration distribution estimated in \cref{E-imperf-test}.}
+    \caption{Parameter estimates for the EC distribution approximating the posterior mean of the duration distribution estimated in \cref{E-imperf-test}.}
     \label{SEIR:table:ec-params}
@@
 
-To link the model with prevalence measurements, I use a beta-binomial likelihood.
-If on each day $t$, each strata $i$ is observed to have $y_{it}$ positives out of $n_{it}$ tests on days $t$, then the likelihood is:
+The modelled proportion of individuals that are PCR positive in the can then be linked to the data using a Beta-Binomial likelihood (see\todo{ref section on clustering} for the issues with using a binomial likelihood).
+If on each day $t$, each strata $i$ is observed to have $y_{it}$ positives out of $n_{it}$ tests, then the likelihood is:
 \begin{align}
@@
 
-    \path (S) edge node {$\sum_j \beta_{ij} i_j$} (E1)
+    \path (S) edge node {$\Delta_i$} (E1)
           (E1) edge node {$\sigma$} (E2)
@@
 
-    \path  (S) edge node {$\sum_j \beta_{ij} i_j$} (P0)
+    \path  (S) edge node {$\Delta_i$} (P0)
           (P0) edge node {$\kappa_0$} (P1)
@@
 
-A diagram of the model is shown in figure \cref{SEIR:fig:full-model}.
-The system of differential equations is below.
+A diagram of the model for a single stratum is shown in figure \cref{SEIR:fig:full-model}.
+The system of differential equations for each region is below.
 \begin{align}
@@
 where $\vec{\Delta} = \vec{s} \circ (\matr{\beta} \circ \matr{M}) (\vec{i_1} + \vec{i_2})$ is the vector of instantaneous incidence.
+Recall that each region is run independenlty (see \cref{SEIR:sec:geography}), $\matr{beta}_{ij} = \beta(t) \beta^\text{susc}_i$ with $\beta^\text{susc}_i = \beta_c$ for child-aged strata and $\beta^\text{susc}_i = 1$ for adult-aged strata (see \cref{SEIR:sec:structured-populations}), and that $\beta(t)$ and $\matr{M}$ are time-varying (see \cref{SEIR:sec:time-varying-foi}).
 
@@
 Rather than directly inferring $\beta_1$, I infer the initial growth rate of the epidemic.
-I use a parsimonious parameterisation of the initial conditions using the initial proportion of the population that are infected and the initial proportion that are susceptible only.
+In addition, I use a parsimonious parameterisation of the initial conditions using the initial proportion of the population that are infected and the initial proportion that are susceptible only.
 Both of these rely on an assumption that the epidemic is in the steady-state exponential growth phase at time 0.
 
-For an age-structured model, as the one here, the initial phase of the epidemic, when most individuals are susceptible, is approximately exponential.
+For any short period of time when $\vec{s}$ is approximately constant, the system of ODEs can be solved analytically to produce exponential growth.
 In fact, all the $\vec{e}$, $\vec{i}$ and $\vec{p}$ compartments grow at the same initial exponential growth rate, $\psi$.
-The basic reproduction number, $\R$ of the epidemic is related to this growth rate as follows~\autocites{birrellRealtimea}{wearingAppropriate}:
+If $\vec{s} = 1$ then the basic reproduction number can be linked to the growth rate of this epidemic as follows~\autocites{birrellRealtimea}{wearingAppropriate}:
 \begin{align}
@@
 Then calculate $R^* = D(\matr{G})$ when $\beta_1=1$, where $D(\cdot)$ gives the dominant eigenvalue.
-Note that multiplying a matrix by a scalar multiplies its eigenvalues by that same scalar.
-Finally, set $\beta_1 = \hat{R} / R^*$.
+Multiplying a matrix by a scalar multiplies its dominant eigenvalue by that same scalar, therefore, setting $\beta_1 = \hat{R} / R^*$ gives the desired $\R$.
 
-To form a parsimonious parameterisation of the initial state, first focus on the initial number of infected individuals.
-In steady-state exponential growth, the relative proportion of each strata that is infected is proportional to the dominant eigenvector of $\matr{G}$, $\vec{d}$.
-I assume this is the case at time 0.
+To form a parsimonious parameterisation of the initial state, first focus on the initial proportion of the population that is in infected compartments.
+In steady-state exponential growth, the relative proportion of each strata that is infected is proportional to the dominant eigenvector of $\matr{G}$, which I denote $\vec{d}$.
 Let the proportion of the population in the $i$ states at time 0 in the strata with the highest proportion infected, $i^+$, be a model parameter.
@@
     Mean latent period & $1/\sigma$ & 3.5 days (fixed) & From a paper \\
-    Mean infectious period & $1/\gamma$ & 3.5 days (fixed) & From a paper \\
-    Initial proportion without immunity & $\pi$ & See \cref{SEIR:table:immunity} & \\
+    Mean infectious period & $1/\gamma$ & 4 days (fixed) & From a paper \\
+    Initial proportion without immunity & $\vec\pi$ & See \cref{SEIR:table:immunity} & \\
     Initial proportion infected & $i^+$ & $\BetaDist(0.5, 1000)$ & Weakly informative \\
-    Relative susceptibility of children & $\beta_c$ & $\LN(-4.325, 0.1174)$ &  \\
-    Initial growth rate & $\psi$ & $\N(0.048, 0.0035)$ & Weakly informative, centered on growth rate of deaths \\
-    Standard deviation of random walk & $\sigma$ & $\Exponential(80)$ & Weakly informative, NCP \\
+    Relative susceptibility of children & $\beta_c$ & $\LN(-0.4325, 0.1174)$ &  \\
+    Initial growth rate & $\psi$ & $\N(0.06, 0.04)$ & Weakly informative \\
+    Standard deviation of random walk & $\sigma$ & $\Exponential(80)$ & Weakly informative penalised model complexity prior~\autocite{simpsonPenalising} \\
     Overdispersion of observations & $\rho$ & $\Exponential(2 \times 10^5)$ & Weakly informative
-
 \end{tabular}
-\todo[inline]{Find citation for children, check growth rate, check what NCP is}
+\todo[inline]{Find citation for children, add priors for $\pi$}
 \caption[SEIR model priors]{Priors for each parameter. For details of the distributions and their parameterisations see \cref{E-distributions}.}
@@
 \subsection{Solving the systems of ODEs}
+I solve the system of ODEs using Euler's method, discretised at half day timesteps.
+That is to find the state of the system at time $t+1/2$ given the state at time $t$, I add half the derivative to the state at time $t$.
+
+The prevalence on day $t$, used for the likelihood, is the mean of the solution on the two corresponding days.
 
 \section{Simulation study} \label{SEIR:sec:sim-study}
+I ran a simulation study of the East of England region.
+This checked that the inference procedure (described in \cref{SEIR:sec:inference-implementation}) was working correctly and that the model was identifiable.
+I ran 100 simulations, with parameters drawn from the priors (see \cref{SEIR:table:priors}) except that the prior on $\psi$ was $\LN(0.048, 0.0035)$.
+This is because the prior used for the main analysis was too weak on this prior to create realistic simulations.
+
+Simulating from the prior then attempting to recover the parameters is the basis of simulation-based calibration~\autocite{taltsValidating}.
+The coverage of the posterior credible intervals should be their nominal value.
+Furthermore, the combination of the posteriors across all simulations should recover the prior.
+
+To estimate the posterior, I ran a single MCMC chain for 500,000 iterations, discarding the first half as burn-in.
+On some occasions, the proposal distribution's covariance matrix became singular (or close to singular).
+This means it is not possible to propose a new value.
+In these cases, I restarted the chain with a new seed.
 

### 2023-10-20 cis-perfect-testing.tex
consistent notation for distributions
Incorporate Dani's comments

@@
 
-The estimation of incidence from prevalence is sensitive to the tail of the duration distribution.
+When estimating incidence from prevalence, the mean duration, not its median, is the most important feature of the duration distribution~\autocite{freemanPrevalence}.
+Duration distributions tend to be right-skewed.
+Therefore, estiamtes of the mean are sensitive to the tail of the duration distribution.
 In \cref{E-ATACCC} I estimated the duration distribution from the ATACCC dataset.
 However, this study suffered from a small sample size and only 20 days of follow-up.
-Therefore, model assumptions determined the size of the estimated tail estimated.
-In this chapter, I develop a method to estimate the duration distribution using CIS data. CIS sampling is less frequent, but with a large sample size and long follow-up.
+Therefore, model assumptions drive the size of the estimated tails.
+In this chapter, I develop a method to estimate the duration distribution using CIS data.
+The CIS sampling is less frequent, but with a large sample size and long follow-up.
 
 I take a Bayesian survival analysis approach to model the coarse sampling of the CIS (see \cref{perf-test:sec:problem}).
-Flat priors are unintentionally very informative.
-Hence, at least weakly informative priors should be used.
-Of particular interest is a strongly informative prior to combine the information in the CIS with the results of \cref{E-ATACCC} (see \cref{perf-test:sec:parameters-priors}).
-I conduct a simulation study to demonstrate this method's performance.
-It succesfully recovers the truth when using reasonable priors (see \cref{perf-test:sec:simulation-study}).
-In reality, the method must account for false negatives, discussed in \cref{E-imperf-test}.
+I use a simulation study (see \cref{perf-test:sec:simulation-study}) to evaluate a range of priors for the survival distribution (described in \cref{perf-test:sec:parameters-priors}).
+These include a prior to combine the information in the CIS with the results of \cref{E-ATACCC}.
+Reasonable priors successfully recover the true distribution.
+However, the method must be able to account for false negatives, discussed in \cref{E-imperf-test}.
 
@@
 
-In the CIS each individualis tested  infrequently, either weekly or monthly (see\todo{refer to section with CIS intro}).
+The CIS tests each individual infrequently, either weekly or monthly (see\todo{refer to section with CIS intro}).
 The infrequent testing complicates analysis of the data in two ways.
@@
 
+\subsection{Undetected infections}
+
 The first complication in the analysis is that infection episodes can be undetected.
-This means there are no positive tests associated with them.
-Undetected episodes occur when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
-For undetected episodes, there is no information on them episode; even the number that occurred is unknown.
+Undetected infection episodes are episodes with no associated positive tests.
+Therefore, there is no information directly observed about the episode; even the number that occur is unobserved.
+They occur when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
 Undetected episodes are important because they are shorter than detected episodes.
@@
 If the duration had been 24 days or more, the individual would still be detectable at their next test and detected then.
-CIS has many undetected infections. This is because frequency of testing (up to 28 days, see\todo{ref section on the survey design}) is longer than most infections (median around 15 days).
+CIS has many undetected infections.
+This is because interval between tests in the same individual (up to 28 days, see\todo{ref section on the survey design}) is longer than most infections (median around 15 days, see\todo{ref ATACCC chapter}).
 \begin{figure}
@@
 \end{figure}
+
+\subsection{Double interval censoring}
 
@@
 \end{figure}
+
+\subsection{Approach}
 
@@
 Uninformative Beta priors on the hazard are problematic in this context.
-An uninformative prior would be of the form $\lambda_t \dist \Beta(\alpha, \alpha)$.
+An uninformative prior would be of the form $\lambda_t \dist \BetaDist(\alpha, \alpha)$.
 Commonly, $\alpha$ is chosen to be $0.5$ (Jeffreys' Prior) or $1$ (a uniform prior).

### 2023-10-20 cis-perfect-testing.tex
Spellcheck

@@
 Duration distributions tend to be right-skewed.
-Therefore, estiamtes of the mean are sensitive to the tail of the duration distribution.
+Therefore, estimates of the mean are sensitive to the tail of the duration distribution.
 In \cref{E-ATACCC} I estimated the duration distribution from the ATACCC dataset.
@@
 CIS has many undetected infections.
-This is because interval between tests in the same individual (up to 28 days, see\todo{ref section on the survey design}) is longer than most infections (median around 15 days, see\todo{ref ATACCC chapter}).
+This is because the interval between tests in the same individual (up to 28 days, see\todo{ref section on the survey design}) is longer than most infections (median around 15 days, see\todo{ref ATACCC chapter}).
 \begin{figure}
@@
 \end{align}
-The hyperparmeters $\mu_{\lambda_1}$ and $\sigma_{\lambda_1}$ specify the prior on the hazard at time 1; $\mu_{\epsilon_1}$ and $\sigma_{\epsilon_1}$ the prior on the initial gradient; and $\mu_\sigma$ the smoothness of the random walk.
+The hyperparameters $\mu_{\lambda_1}$ and $\sigma_{\lambda_1}$ specify the prior on the hazard at time 1; $\mu_{\epsilon_1}$ and $\sigma_{\epsilon_1}$ the prior on the initial gradient; and $\mu_\sigma$ the smoothness of the random walk.
 I use $\mu_{\lambda_1} = -17.5$, $\sigma_{\lambda_1} = 6$, $\mu_\epsilon = 1.09$, $\sigma_\epsilon = 0.03$, and $\mu_\epsilon = 0.1$ based on the analysis in \cref{E-ATACCC} (see \cref{perf-test:fig:rw2-prior}).
@@
 Secondly, the uncertainty in the estimates from \cref{E-ATACCC} should be increased for this analysis.
-The additional uncertainty is because the \cref{E-ATACCC} analysis extrapolates beyond the data using strong model assumptions (see \todo{reference discussion on extrapolation in ATACCC chapter}).
+The additional uncertainty is because the \cref{E-ATACCC} analysis extrapolates beyond the data using strong model assumptions (see\todo{reference discussion on extrapolation in ATACCC chapter}).
 %Furthermore, there are differences in the study design and laboratories used between the two studies (see \cref{E-intro:sec:studies}) which may mean that results do not generalise between the studies.
@@
 
-The prior on the total number of infections was centered on the true value with a dispersion parameter of 1.
+The prior on the total number of infections was centred on the true value with a dispersion parameter of 1.
 This meant that the prior was very diffuse.
@@
 This is caused by the large sample size (4800 detected infections).
-Therefore, only results from a single simulation are shown and dicussed.
+Therefore, only results from a single simulation are shown and discussed.
 
@@
 The informative prior had tighter credible intervals, as expected.
-This prior was centered on the correct value for the initial part of the distribution.
+This prior was centred on the correct value for the initial part of the distribution.
 Therefore, the narrow credible intervals are unsurprising.

### 2023-10-23 cis-perfect-testing.tex
finish redraft

@@
 \section{Definitions}
-\todo[inline]{These will be defined elsewhere in the thesis, but are here so that this draft is understandable}
+\todo[inline]{These will be defined in the introductory chapters, but are here so that this draft is understandable}
 \begin{itemize}
-\item
-  Time is discrete: $1, \dots, T$.
-\item
-  We observe $n_a$ episodes of infection, indexed by $i$.
-\item
-  The $i$ in which episode $i$ occurs is tested at times
-  $t_i = \{ t_{i,1}, \dots, t_{i,m_i} \}$.
-\item
-  Each observed episode has an unknown start of infection time $b_i$
-  and end of infection time $e_i$ (the first and last day that the
-  individual would test positive due to this episode respectively).
-  These are realisations of the random variables $B_i$ and $E_i$
-  respectively.
-\item
-  The duration of the infection is the number of days for which an
-  individual tests positive, $D_i = E_i - B_i + 1$.
 \item
@@
   in $[l_i^{(e)}, r_i^{(e)}]$.
-\item
-  $y_i(t)$ for $t \in t_i$ is a binary indicator giving the test
-  result for individual $i$ at time $t$. Under the perfect testing
-  assumption, $y_i(t) = 1$ if and only if $b_i \leq t \leq e_i$.
-\item
-  Throughout, the convention that lower-case letters are realisations of
-  upper-case random variables is used.
 \end{itemize}
@@
 In CIS, infection episodes are more likely to be detected when an individual is in the weekly testing phase.
-CIS is more likely to detect infection episodes during the weekly testing phase.
 Therefore, the probability of an infection episode being detected conditional on being infected is higher than the overall probability of being detected.\todo{this is a bit unclear, try and rephrase?}
@@
 I augment the data set with the full number of episodes, including those that are undetected.
-The augmented dataset is standard double interval censored data, with a well-known likelihood.
 
@@
 Let the total number of infection episodes in the cohort be $\ntot = \ndet + \nnodet$ where $\ndet$ is the number of detected infections and $\nnodet$ is the number of undetected infections.
-Let $\Ncis$ be the number of individuals in the cohort.
-Without loss of generality, let the individuals with a detected episode be indexed by $i = 1, \dots, \ndet$ and the individuals without a detected episode be indexed by $i = \ndet + 1, \dots, \Ncis$.
-Over the short period of time considered, no individual has more than one detected episode. 
-Let $j = 1, \dots \ntot$ index infection episodes, ordered such that for $j \leq \ndet$ the $j$th infection episode occurs in the $i$th individual.
+Let $\Ncis$ be the number of individuals in the cohort and $n_\sched$ be the number of test schedules.
+%Without loss of generality, let the individuals with a detected episode be indexed by $i = 1, \dots, \ndet$ and the individuals without a detected episode be indexed by $i = \ndet + 1, \dots, \Ncis$.
+Let $i = 1, \dots \ntot$ index infection episodes.
 
-All the relevant information about episode $j$ can be fully characterised by the (unobserved) triplet $(b_j, e_j, \sched_j)$ where $b_j$ and $e_j$ are the times the episode being and end respectively and $\sched_j$ is the set of test times for the individual in which the episode occurs.
-If $\sched_j$ contains at least one time between $b_i$ and $e_i$ (inclusive), one time prior to $b_i$, and one time after $e_i$, then the episode is detected, otherwise it is undetected.
-The triplet belongs to a state space.
-$b_j$ and $e_j$ span possible infection and recovery times.
-$\sched_j$ is one of the observed test schedules, with any other test schedule assumed to occur with probability zero.
+All the relevant information about episode $i$ can be fully characterised by the (unobserved) triplet $(b_i, e_i, \sched_i)$ where $b_i$ and $e_i$ are the times the episode being and end respectively and $\sched_i$ is the set of test times for the individual in which the episode occurs.
+The episode's duration, the number of days the individual in which the episode occurred, was positive for is $d_i = e_i - b_i + 1$.
+This triplet is assumed iid.
+If $\sched_i$ contains at least one time between $b_i$ and $e_i$ (inclusive), one time prior to $b_i$, and one time after $e_i$, then the episode is detected, otherwise it is undetected.
+$b_i$ and $e_i$ span possible infection and recovery times.
+$\sched_i$ is one of the observed test schedules, with any other test schedule assumed to occur with probability zero.
+Over the short period of time considered, each unique test schedule has zero or one detected episodes.
+Let $\set{D}$ be the set of test schedules with a detected infection in the cohort.
 
-To relate the state space to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie: are strict subsets of $T \times T$), for any individual based on their test times.
+To relate the state space to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie: are strict subsets of $T \times T$), for each test schedule.
 Figure \ref{perf-test:fig:partitionSpace} shows this graphically.
@@
   Admissible episodes, $\alpha_i$, which have an infection and recovery time consistent with the data observed.
-  This space is empty if individual $i$ had no detected episode (\ie: $i > \ndet$).
-  Otherwise, $b_i \in [l_i^{(b)}, r_i^{(b)}]$, and $e_i \in [l_i^{(e)}, r_i^{(e)}]$.
-  If $i \leq \ndet$ then $n_{ia} =1$ of these occurred in individual $i$, otherwise $n_{ia} = 0$.
-  Conditional on the episode occurring in individual $i$, it is admissible with probability $p_{ia} = \prob((b, e) \in \alpha_i \mid \theta)$.
+  This space is empty for $i \notin \set{D}$.
+  Otherwise, define $l_i^{(b)}$ and $r_i^{(b)}$ as the first and last days $b_i$ could have occurred.
+  $l_i^{(b)}$ is the day after the last negative test prior to the first positive test in the episode.
+  $r_i^{(b)}$ is the day of the first positive test.
+  Similarly, define $l_i^{(e)}$ and $r_i^{(e)}$ as the first and last days $e_i$ could have occurred.
+  If $i \in \set{D}$ then $n_{ia} =1$ admissible episodes occurred in individuals with test schedule $\sched{i}$, otherwise $n_{ia} = 0$.
+  Conditional on the episode occurring in individuals with test schedule $\sched_i$, it is admissible with probability $p_{ia} = \prob((b, e) \in \alpha_i \mid \theta)$.
   \label{perf-test:def:admissible}
@@
   Undetected episodes, $\Omega_i^C$, which have an infection and recovery time such that they would not have tested positive.
-  An unknown number, $n_{iu}$ of these observed in individual $i$.
-  Conditional on the episode occurring in individual $i$, it is undetected with probability $p_{iu} = \prob((b, e) \in \Omega^C_i \mid \theta)$.
+  An unknown number, $n_{iu}$ of these observed in individuals with test schedule $\sched_i$.
+  Conditional on the episode occurring in individuals with test schedule $\sched_i$, it is undetected with probability $p_{iu} = \prob((b, e) \in \Omega^C_i \mid \theta)$.
 \item
@@
   This is all remaining episodes not in the previous sets.
-  For an individual with no detected episodes, $\beta_i$ is the area of possible infections that would have been detected.
-  $n_{i\bar{a}} = 0$ of these occurred in any individual $i$.
-  Conditional on the episode occurring in individual $i$, it is inadmissible with probability $p_{i\bar{a}} = \prob((b, e) \in \beta_i \mid \theta)$.
+  $n_{i\bar{a}} = 0$ of these occurred in any individuals with test schedule $\sched_i$.
+  Conditional on the episode occurring in these individuals, it is inadmissible with probability $p_{i\bar{a}} = \prob((b, e) \in \beta_i \mid \theta)$.
 \end{itemize}
 
-The detected region, $\Omega_i = \alpha_i \cup \beta_i$, for individual $i$ is the region where we could have observed an episode. 
+The detected region, $\Omega_i = \alpha_i \cup \beta_i$, for test schedule $\sched_i$ is the region where we could have observed episodes. 
 The notation is for consistency with \textcite{heiseyModelling}.
@@
 \includegraphics[width=\textwidth]{cis-perfect-testing/regions_diag}
-\caption[Admissible, inadmissible, and undetected infections]{The regions (as defined in the main text) for an episode $i$ in an
-individual whose first test was at time 0 and negative, with subsequent
+\caption[Admissible, inadmissible, and undetected infections]{The regions (as defined in the main text) for an individual with test schedule $\sched_i$ whose first test was at time 0 and negative, with subsequent
 negative tests at times 7, 14, and 56, and positive tests at times 21
@@
 Therefore, these definitions are consistent with the previous definition of $\ntot$.
-That is, $\ntot = \ndet + \nnodet + n_{i\bar{a}} = \sum_{i=1}^{\ndet} n_{ia} + \sum_{i=1}^{\Ncis} n_{iu} + 0$.
+That is, $\ntot = \ndet + \nnodet + n_{i\bar{a}} = \sum_{i=1}^{\ndet} n_{ia} + \sum_{i=1}^{\ntot} n_{iu} + 0$.
 Furthermore, for all $i$, $p_{ia} + p_{iu} + p_{i\bar{a}} = 1$.
 
-Any episode $j$ independently belongs to one of these classes.
+Any episode $i$ independently belongs to one of these classes.
 I assume the individual in which each episode occurs is iid uniform across the cohort.
 Therefore, conditional on $\ntot$, the number of episodes in each class, $(n_{ia}, n_{iu}, n_{i\bar{a}})^T$, are distributed multinomially.
-The above probabilities ($p_{iu}$, $p_{ia}$, and $p_{i\bar{a}}$) are conditional on the episode occurring in individual $i$.
-These probabilities are multiplied by $1/\Ncis$ to give the probability that an episode belongs to each class, without conditioning on a specific individual.
+The above probabilities ($p_{iu}$, $p_{ia}$, and $p_{i\bar{a}}$) are conditional on the episode occurring with test schedule $\sched_i$.
+These probabilities are multiplied by $\eta_i/\Ncis$, where $\eta_i$ is the number of individuals with test schedule $\sched_i$, to give the probability that an episode belongs to each class, without conditioning on a specific individual.
 
@@
 Because the counts have a joint multinomial distribution, this is equivalent to only considering the sum of the counts.
-The overall probability of an episode being unobserved is $\pnodet = \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} p_{iu}$.
+The overall probability of an episode being unobserved is $\pnodet = \frac{1}{\Ncis} \sum_{i=1}^{n_\sched} \eta_i p_{iu}$.
 Furthermore, the inadmissible classes are guaranteed to have zero episodes.
@@
 
-Let $\na = (n_{1a}, \dots, n_{\Ncis a})^T$ be the vector of counts of episodes in the admissible classes.
+Let $\na = (n_{1a}, \dots, n_{n_\sched a})^T$ be the vector of counts of episodes in the admissible classes.
 The counts $\na$ and $\nnodet$ are jointly distributed multinomially.
-The support of the distribution requires $\nnodet = \ntot - \sum_{i=1}^{\Ncis} n_{ia} = \ntot - \ndet$ and $\ntot \geq \ndet$.
+The support of the distribution requires $\nnodet = \ntot - \sum_{i=1}^{n_\sched} n_{ia} = \ntot - \ndet$ and $\ntot \geq \ndet$.
 Hence:
@@
 &= p(\nnodet, \na\mid \ntot, \theta) \\
-&= \frac{\ntot!}{\nnodet! \prod_{i=1}^{\Ncis} n_{ia}!}  \left(\frac{1}{\Ncis}\pnodet\right)^{\nnodet}\prod_{i=1}^{\Ncis} \left(\frac{1}{\Ncis}p_{ia}\right)^{n_{ia}} \\
-&\propto \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i=1}^{\ndet} p_{ia} \label{perf-test:eq:augmented-likelihood}
+&= \frac{\ntot!}{\nnodet! \prod_{i=1}^{n_\sched} n_{ia}!}  \pnodet^{\nnodet}\prod_{i=1}^{n_\sched} \left(\frac{1}{n_\sched}p_{ia}\right)^{n_{ia}} \\
+&\propto \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i \in \set{D}} p_{ia} \label{perf-test:eq:augmented-likelihood}
 \end{align}
-the final line follows because $\nnodet = \ntot - \ndet$, $n_{ia} = 1$ for $i \leq \ndet$, and $n_{ia} = 0$ for $i > \ndet$.
+the final line follows because $\nnodet = \ntot - \ndet$, $n_{ia} = 1$ for $i \in \set{D}$, and $n_{ia} = 0$ otherwise.
 
@@
 
-Now I derive an analytical solution to $\sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under the prior $\ntot \dist \NB(\mu, r)$ (using the mean/dispersion parameterisation of the negative binomial).
+Now I derive an analytical solution to $\sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under the prior $\ntot \dist \NB(\mu, r)$ (using the mean/dispersion parameterisation of the negative binomial, see \cref{E-distributions}).
 I assume that $\theta$, which are only the parameters of the survival distribution, is a priori independent of $\ntot$, the number of infections that occurred in the cohort.

### 2023-10-23 distributions.tex
Add centred negbin and fix notation

@@
 Name & Notation & Parameters & Mean & Variance & pmf \\
-Binomial & $\Bin(n, p)$ & $n$ (number of trials, integer)\newline $0 \leq p \leq 1$ (probability of success) & $np$ & $np(1-p)$ & ${n \choose x}p^x(1-p)^{n-x}$ \\
+Binomial & $\Bin(n, p)$ & $n \in \nats$ (number of trials, integer)\newline $0 \leq p \leq 1$ (probability of success) & $np$ & $np(1-p)$ & ${n \choose x}p^x(1-p)^{n-x}$ \\
 Poisson & $\Poi(\lambda)$ & $\lambda > 0$ (average rate) & $\lambda$ & $\lambda$ & $\frac{e^{-\lambda}\lambda^x}{x!}$ \\
-Negative Binomial & $\NB(r, p)$ & $r$ (number of failures until stopping)\newline $p$ (probability of success) & $\frac{pr}{1-p}$ & $\frac{pr}{(1-p)^2}$ & ${x + r - 1 \choose x} p^x (1-p)^r$ \\
-Beta-Binomial (centered) & $\BB(n, p, \rho)$ & $n$ (number of trials, integer)\newline $0 \leq p \leq 1$ (success probability)\newline $\rho > 0$ (overdispersion) & $np$ & $np(1-p)(1+n\rho)(1+\rho)^{-1}$ & ${n \choose x} \frac{B(x+p/\rho, n-x+(1-p)/\rho)}{B(p/\rho, (1-p)/\rho)}$ \\
-Multinomial & $\MN(n, \vec{p})$ & $n$ (number of trials, integer)\newline $0 < \vec{p}$ (probabilities, $\sum_i p_i = 1$) & $n \vec{p}$ & See below & $\frac{n!}{x_1!x_2!\ldots x_k!}p_1^{x_1}p_2^{x_2}\ldots p_k^{x_k}$ \\
+Negative Binomial (standard) & $\NB(r, p)$ & $r > 0$ (number of failures until stopping)\newline $0 < p < 1$ (probability of success) & $\frac{pr}{1-p}$ & $\frac{pr}{(1-p)^2}$ & $\frac{\Gamma(x+r)}{x!
+\Gamma(r)} p^x (1-p)^r$ \\
+Negative Binomial (centred) & $\NB(\mu, r)$ & $\mu > 0$ (mean) \newline r (dispersion) & $\frac{pr}{1-p}$ & $\frac{pr}{(1-p)^2}$ & $\frac{\Gamma(x+r)}{x!
+\Gamma(r)} \left( \frac{r}{r+\mu} \right)^\mu \left( \frac{\mu}{r+\mu} \right)^k$ \\
+Beta-Binomial (centred) & $\BB(n, p, \rho)$ & $n \in \nats$ (number of trials, integer)\newline $0 \leq p \leq 1$ (success probability)\newline $\rho > 0$ (overdispersion) & $np$ & $np(1-p)(1+n\rho)(1+\rho)^{-1}$ & ${n \choose x} \frac{B(x+p/\rho, n-x+(1-p)/\rho)}{B(p/\rho, (1-p)/\rho)}$ \\
+Multinomial & $\MN(n, \vec{p})$ & $n \in \nats$ (number of trials, integer)\newline $\vec{p} > 0$ (vector of $k$ probabilities, $\sum_{i=1}^k p_i = 1$) & $n \vec{p}$ & See below & $\frac{n!}{x_1!x_2!\ldots x_k!}p_1^{x_1}p_2^{x_2}\ldots p_k^{x_k}$ \\
 \end{tabular}

### 2023-10-26 ATACCC.tex
Copy/paste old text for ATACCC

\documentclass[thesis.tex]{subfiles}

\begin{document}
\ifSubfilesClassLoaded{
  \setcounter{chapter}{4}
}

\chapter{Estimating duration using biomarkers} \label{ATACCC}

\section{Background}

Two major approaches have been taken to model SARS-CoV-2 viral load trajectories.
There are either using ordinary differential equations (ODEs) with a biological interpretation or using mathematical functions that are flexible enough to capture the dynamics.

ODE-based approaches have parameters represent biological properties of the virus and host, such as the immune response (eg:~\autocites{ejimaEstimation,keViro,kimQuantitative,goncalvesTiming,perelsonMechanistic}).
Such models are rarely analytically-tractable and the system's evolution is computed using numeric methods.
The second approach is to approximate the dynamics using simple functions.
Most common is a piecewise model, where the logarithm of viral load increases linearly, peaks instantaneously, and then decreases linearly~\autocites{clearyUsing,kisslerSARSCoV2,larremoreTest}; although more flexible functions have also been suggested~\autocites{quiltyQuarantine}.
In these models, parameters are clearly related to the data, such as the time from first detection to peak viral load, and the viral load at the peak.
In both approaches, random effects are used to represent variation between hosts within the population, capturing the non-independence of longitudinal data from the same host.
Here, we focus on the latter class of models, adopting a piecewise-linear approach.

Piecewise linear, random effect models have a long history of use for longitudinal data in biostatistics~\autocites{slateStatistical}.
Often, as here, they are used to model the temporal evolution of biomarkers, including:
CD4 counts following HIV infection~\autocites{langeHierarchical,lynchPredicting}, parasite density in
blood samples following malaria treatment~\autocites{fogartyBayesian}, and measurements
of cognitive decline due to Alzheimer's disease~\autocites{bealisleBayesian}.
Various extensions to the model exist including the incorporation of multiple biomarkers~\autocites{inoueModeling,giardinaGetting}, time-to-event data [REF], and relaxing the piecewise linear assumption to a semi-parameteric form~\autocites{zegerSemiparametric}.

Viral loads are infrequently measured exactly.
The most common form of data obtained is cycle threshold (Ct) values from real-time quantitative  reverse-transcriptase polymerase chain reaction (PCR) tests.
The Ct value is proportional to the negative log of the viral load which is convenient when modelling log viral load as a piecewise linear function, since Ct values themselves are piecewise linear under this assumption.
When a host's viral load is too low (below the limit of detection of the PCR test), no Ct value is available and hence viral load can be considered censored at the limit of detection.
The relationship between Ct and viral load is noisy and can vary based on many factors including quality of swab obtained, primers used (even between batches), and the PCR system set-up~\autocites{dahdouhCt,hanRTPCR}.

\section{Original analysis}

\subsection{Model of viral load}

For an individual $i$, their trajectory is determined by two points and one slope parameter.
The points are the time at which the individual has a 50\% chance of testing positive, $t_{i,0}$, and the time at which their viral load
peaks, $t_{i,\text{peak}}$ after $t_{i,0}$; the slope parameter, $\beta_i$ determines the rate their Ct increases (viral load decreases) following the peak.
The Ct value for an individual at time $t_{i,0}$ is the known limit of detection, $C_\text{lod}$,
Their Ct value at the peak is $C_{i,\text{peak}}$ and inferred.
The slope parameter, $\beta_i$, determines how quickly an individual's Ct value increases following the peak viral load.
We denote the vector of the parameters defined here as $\vec\theta_i = \begin{bmatrix} t_{i,0} & t_{i,\text{peak}} & C_{i,\text{peak}} &  \beta_i \end{bmatrix}^T$.

This parameterisation leads to the following Ct value trajectory for individual $i$:
$$
\hat{C}_i(t, \vec\theta_i) = \begin{cases}
  C_{\text{lod}} + (C_{i,\text{peak}} - C_{\text{lod}}) \frac{t - t_{i,0}}{t_{i,\text{peak}}}
    &t \leq t_{i,\text{peak}} + t_{i,0} \\
  C_{i,\text{peak}} + \beta_i (t - t_{i,\text{peak}} - t_{i,0})
    &t > t_{i,\text{peak}} + t_{i,0}.
\end{cases}
$$

Day 0 for an individual is the day they were enrolled in the study and first swabbed.

\subsection{Observation model and likelihood}

Observed Ct values are assumed to be drawn from a Normal distribution with mean $\hat{C}_i(t)$ and standard deviation $\sigma_\text{obs}$.
censored at the limit of detection.
Additionally, we introduce a probability of a false negative $\eta$, to account for negative test results seen near the peak viral load.
Denoting an observed Ct value for individual $i$ at time $t$ as $y_{i,t}$ with
$y_{i,t} = \infty$ for a negative test, this gives the following likelihood.
$$
\pi(y_{i,t} \mid \vec\theta_i, \sigma_\text{obs}, \eta) = \begin{cases}
  (1 - \eta) f_N(y_{i,t} \mid \hat{C}_i(t, \vec\theta_i), \sigma_\text{obs}^2) &y_{i,t} < \infty \\
  \eta + (1 - \eta) (1 - F_N(C_\text{lod} \mid \hat{C}_i(t, \vec\theta_i), \sigma_\text{obs}^2)) & y_{i,t} = \infty
\end{cases}
$$
where $\sigma_\text{obs}$ is a model parameter determining the variability in
test results, and $f_N(y \mid \mu, \sigma^2)$ and $F_N(y \mid \mu, \sigma^2)$ are the pdf
and cdf (respectively) of a Normal distribution with mean $\mu$ and variance $\sigma^2$.

We assume that all observations (both of the same and different individuals) are independent conditional on the model parameters.
The parameter $\vec\theta_i$ introduces dependence between observations of the same individuals.
\autoref{ATACCC:sec:hierarchy} explains the hierarchical structure which borrows information between individuals.

\subsection{Hierarchical structure}\label{ATACCC:sec:hierarchy}

\begin{figure}
\begin{tikzpicture}[
    node distance = 3cm and 3cm,
    on grid,
    auto,
    ->,>=stealth',
    every state/.style={draw,circle,minimum size=1.5cm},
    ]

% Define the nodes
\node[state] (mu-tpeak) {$\mu_{\log t_\text{peak}}$};
\node[state, right=of mu-tpeak] (sigma-logtpeak) {$\sigma_{\log t_\text{peak}}$};
\node[state, below right=of mu-tpeak] (ti-peak) {$t_{i,\text{peak}}$};
\node[state, right=of sigma-logtpeak] (mu-Cpeak) {$\mu_{C_\text{peak}}$};
\node[state, right=of mu-Cpeak] (sigma-Cpeak) {$\sigma_{C_\text{peak}}$};
\node[state, below right=of mu-Cpeak] (Ci-peak) {$C_{i,\text{peak}}$};
\node[state, right=of sigma-Cpeak] (mu-log-Bi) {$\mu_{\log \beta_i}$};
\node[state, right=of mu-log-Bi] (sigma-log-Bi) {$\sigma_{\log \beta_i}$};
\node[state, below right=of mu-log-Bi] (Bi) {$\beta_i$};
\node[state, left=of ti-peak] (ti0) {$t_{i,0}$};
\node[draw, inner sep=0.3cm, fit=(ti0) (Bi), fill=blue!10,opacity=0.2] {$\theta_i$};
\node[state, below left=of Ci-peak] (Ci-t) {$\hat{C}_i(t)$};
\node[state, right=of Ci-t] (sigma-obs) {$\sigma_\text{obs}$};
\node[state, right=of sigma-obs] (eta-i) {$\eta_i$};
\node[rectangle, draw, left=of Ci-t,minimum size=1.5cm] (Clod) {$C_\text{lod}$};
\node[state, double, below=of sigma-obs] (Yi-t) {$y_{i,t}$};

% Draw the edges
\draw[->] (mu-tpeak) -- (ti-peak);
\draw[->] (sigma-logtpeak) -- (ti-peak);
\draw[->] (mu-Cpeak) -- (Ci-peak);
\draw[->] (sigma-Cpeak) -- (Ci-peak);
\draw[->] (mu-log-Bi) -- (Bi);
\draw[->] (sigma-log-Bi) -- (Bi);
\draw[->,dotted] (ti0) -- (Ci-t);
\draw[->,dotted] (ti-peak) -- (Ci-t);
\draw[->,dotted] (Ci-peak) -- (Ci-t);
\draw[->,dotted] (Bi) -- (Ci-t);
%\draw[->] (theta-i) -- (Ci-t);
\draw[->] (Ci-t) -- (Yi-t);
\draw[->] (eta-i) -- (Yi-t);
\draw[->,dotted] (Clod) -- (Ci-t);
\draw[->] (Clod) -- (Yi-t);
\draw[->] (sigma-obs) -- (Yi-t);

\end{tikzpicture}
\caption{Hierarchical structure}
\end{figure}

\subsection{Inference}

Inference was performed by Hamiltonian Monte Carlo (HMC) implemented in Stan 2.26.1~[REF].
To aid with mixing, the hierarchical parameters were implemented using the non-centred parameterisation~[REF].
In addition, $t_{i,0}$ was reparameterised in terms of the absolute time of peak.
Convergence was assessed by checking R-hat < 1.1, and visual inspection of traceplots, and effective sample size.

\subsection{Priors}

Weakly informative priors were used on all parameters except $\mu_{\log t_\text{peak}}$ and $\sigma_{\log t_\text{peak}}$.
The latter were chosen to reflect the prior belief that viral load peaks shortly before symptom onset, which occurs a median of 5.1 days after infection~\autocite{mcaloonIncubation}.

\begin{table}[ht]
\centering
\begin{tabular}{llrrr}
  \hline
    Parameter & Distribution & 0.05 & 0.5 & 0.95 \\ 
  \hline
    $\eta$ & $\text{Beta}(0.95, 9)$ & $0$ & $0.07$ & $0.28$ \\ 
    $\sigma_\text{obs}$ & $\text{Half-Cauchy}(0, 0.67)$ & $0.05$ & $0.67$ & $8.51$ \\ 
    $t_{0,i}$ & $N(0, 5^2)$ & $-8.22$ & $0$ & $8.22$ \\ 
    $\mu_{\log t_\text{peak}}$ & $N(\log(3), \log(2.5)^2)$ & $\log (0.66)$ & $\log (3)$ & $\log (13.54)$ \\ 
    $\sigma_{\log t_\text{peak}}$ & $\text{Half-Cauchy}(0, 0.67)$ & $0.05$ & $0.67$ & $8.51$ \\ 
    $\mu_{C_\text{peak}}$ & $N(17, 5^2)$ & $8.78$ & $17$ & $25.22$ \\ 
    $\sigma_{C_\text{peak}}$ & $\text{Half-Cauchy}(0, 4)$ & $0.31$ & $4$ & $50.82$ \\ 
    $\mu_{\log\beta}$ & $N(\log(3.5), \log(1.4)^2)$ & $\log (2.01)$ & $\log (3.5)$ & $\log (6.09)$ \\ 
    $\sigma_{\log\beta}$ & $\text{Half-Cauchy}(0, 0.67)$ & $0.05$ & $0.67$ & $8.51$ \\
   \hline
\end{tabular}
\caption{Priors used, columns 3 to 5 gives quantiles of the distributions specified. See main text for description of the parameters.}
\label{tab:paper:priors}
\end{table}

\section{Comparison to \texorpdfstring{\textcite{hakkiOnset}}{Hakki \etal (2022)}}

\section{Final analysis}

\section{Producing duration from parameter estimates}

The ATACCC analysis produces a posterior distribution for some population-level parameters, $\phi$.
Taking a posterior sample for this, $\phi^{(i)}$, we can then draw the parameters which determine the viral
load trajectory for a random individual $j$, $\theta_j$, from which that individual's duration $d_j^{(i)}$ can be calculated.
By drawing $\theta_1, \theta_2, \dots, \theta_n$ from the distribution $p(\theta_j \mid\phi^{(i)})$, the density of the duration distribution, $f_A(t)$, can be
calculated for the posterior sample $i$ as:
$$
f^{(i)}_A(t) = \frac{1}{n}\sum_{j=1}^n \mathbb{I} \left( t-0.5 < d_j^{(i)} < t+0.5 \right)
$$
Given $f_A(t)$, the cumulative density and hazard functions can be calculated.
Repeating this for $N$ posterior samples of $\phi$ gives the posterior distribution.
I have used $N = 1000$ and $n = 100,000$, 
Since implementing this solution, I have found that Monte Carlo integration to calculate durations given $\theta_i$ is more accurate by avoiding discretisation errors.
However, the results here do not use this for consistency.

\end{document}

### 2023-10-26 ATACCC.tex
Fix references

@@
 
-ODE-based approaches have parameters represent biological properties of the virus and host, such as the immune response (eg:~\autocites{ejimaEstimation,keViro,kimQuantitative,goncalvesTiming,perelsonMechanistic}).
+ODE-based approaches have parameters represent biological properties of the virus and host, such as the immune response ~\autocites[e.g.:][]{ejimaEstimation,keVivo,kimQuantitative,goncalvesTiming,perelsonMechanistic}.
 Such models are rarely analytically-tractable and the system's evolution is computed using numeric methods.
-The second approach is to approximate the dynamics using simple functions.
-Most common is a piecewise model, where the logarithm of viral load increases linearly, peaks instantaneously, and then decreases linearly~\autocites{clearyUsing,kisslerSARSCoV2,larremoreTest}; although more flexible functions have also been suggested~\autocites{quiltyQuarantine}.
+
+The second approach is to approximate the dynamics using general mathematical functions.
+Most common is a piecewise model, where the logarithm of viral load increases linearly, peaks instantaneously, and then decreases linearly~\autocites{clearyUsing,kisslerViral,larremoreTest}; although more flexible functions have also been suggested~\autocites{quiltyQuarantine}.
 In these models, parameters are clearly related to the data, such as the time from first detection to peak viral load, and the viral load at the peak.
 In both approaches, random effects are used to represent variation between hosts within the population, capturing the non-independence of longitudinal data from the same host.
-Here, we focus on the latter class of models, adopting a piecewise-linear approach.
+This chapter uses the latter class of models, adopting a piecewise-linear approach.
 
@@
 of cognitive decline due to Alzheimer's disease~\autocites{bealisleBayesian}.
-Various extensions to the model exist including the incorporation of multiple biomarkers~\autocites{inoueModeling,giardinaGetting}, time-to-event data [REF], and relaxing the piecewise linear assumption to a semi-parameteric form~\autocites{zegerSemiparametric}.
-
-Viral loads are infrequently measured exactly.
-The most common form of data obtained is cycle threshold (Ct) values from real-time quantitative  reverse-transcriptase polymerase chain reaction (PCR) tests.
-The Ct value is proportional to the negative log of the viral load which is convenient when modelling log viral load as a piecewise linear function, since Ct values themselves are piecewise linear under this assumption.
-When a host's viral load is too low (below the limit of detection of the PCR test), no Ct value is available and hence viral load can be considered censored at the limit of detection.
-The relationship between Ct and viral load is noisy and can vary based on many factors including quality of swab obtained, primers used (even between batches), and the PCR system set-up~\autocites{dahdouhCt,hanRTPCR}.
+Various extensions to the model exist including the incorporation of multiple biomarkers~\autocites{inoueModeling,giardinaGetting} and relaxing the piecewise linear assumption to a semi-parameteric form~\autocites{zegerSemiparametric}.
 

### 2023-10-26 thesis.tex
table of distributions with consistent notation
consistent notation for distributions
More SEIR chapter
Fix references

@@
 \usepackage{bm}
+\usepackage{algorithm2e}
 \usepackage[nameinlink]{cleveref}
@@
 \usepackage{xr}
+\usepackage{pdflscape}
+\usepackage{afterpage}
 % \usepackage{microtype}
@@
 \DeclareMathOperator{\Poi}{Poisson}
-\DeclareMathOperator{\Beta}{Beta}
+\DeclareMathOperator{\BetaDist}{Beta}
 \DeclareMathOperator{\Exponential}{Exponential}
-\DeclareMathOperator{\NegBin}{NegBin}
+\DeclareMathOperator{\NB}{NegBin}
+\DeclareMathOperator{\BB}{BetaBin}
 \DeclareMathOperator{\GamDist}{Gamma}
 \DeclareMathOperator{\MN}{Multinomial}
+\DeclareMathOperator{\N}{N}
+\DeclareMathOperator{\LN}{LN}
 \DeclareMathOperator{\expit}{expit}
@@
 \newcommand{\rate}{\mathfrak{R}}
+\newcommand{\R}{\mathcal{R}_{0}}
 
@@
 \makeatother
-\usepackage[backend=biber,style=authoryear,uniquename=false]{biblatex}
+\usepackage[backend=biber,style=authoryear,uniquename=false,sortcites=true]{biblatex}
 \addbibresource{references.bib}
@@
 \section{SARS-CoV-2 and COVID-19}
+
+Viral loads are infrequently measured exactly.
+The most common form of data obtained is cycle threshold (Ct) values from real-time quantitative  reverse-transcriptase polymerase chain reaction (PCR) tests.
+The Ct value is proportional to the negative log of the viral load which is convenient when modelling log viral load as a piecewise linear function, since Ct values themselves are piecewise linear under this assumption.
+When a host's viral load is too low (below the limit of detection of the PCR test), no Ct value is available and hence viral load can be considered censored at the limit of detection.
+The relationship between Ct and viral load is noisy and can vary based on many factors including quality of swab obtained, primers used (even between batches), and the PCR system set-up~\autocites{dahdouhCt,hanRTPCR}.
 
@@
 
-\chapter{Duration from biomarkers} \label{ATACCC}
-
-\section{Background}
-
-\section{Original analysis}
-
-\section{Comparison to \texorpdfstring{\textcite{hakkiOnset}}{Hakki \etal (2022)}}
-
-\section{Final analysis}
-
-\section{Producing duration from parameter estimates}
-
-The ATACCC analysis produces a posterior distribution for some population-level parameters, $\phi$.
-Taking a posterior sample for this, $\phi^{(i)}$, we can then draw the parameters which determine the viral
-load trajectory for a random individual $j$, $\theta_j$, from which that individual's duration $d_j^{(i)}$ can be calculated.
-By drawing $\theta_1, \theta_2, \dots, \theta_n$ from the distribution $p(\theta_j \mid\phi^{(i)})$, the density of the duration distribution, $f_A(t)$, can be
-calculated for the posterior sample $i$ as:
-$$
-f^{(i)}_A(t) = \frac{1}{n}\sum_{j=1}^n \mathbb{I} \left( t-0.5 < d_j^{(i)} < t+0.5 \right)
-$$
-Given $f_A(t)$, the cumulative density and hazard functions can be calculated.
-Repeating this for $N$ posterior samples of $\phi$ gives the posterior distribution.
-I have used $N = 1000$ and $n = 100,000$, 
-Since implementing this solution, I have found that Monte Carlo integration to calculate durations given $\theta_i$ is more accurate by avoiding discretisation errors.
-However, the results here do not use this for consistency.
+\subfile{ATACCC}
 
@@
 
+\subfile{distributions}
+
 \chapter{CIS episode definitions} \label{episode-def}

### 2023-10-27 ATACCC.tex
Edit notation

@@
 \chapter{Estimating duration using biomarkers} \label{ATACCC}
+
+Overall summary of chapter.
+Need to include: how the different sections and analyses interact.
+
+\Cref{ATACCC:sec:background} explains how biomarkers are typically modelled, both for SARS-CoV-2 and historically in the biostatistical literature.
+I explain an analysis, \textcite{hakkiOnset}, applying this to the ATACCC data in \cref{ATACCC:sec:hakki}.
+However, this analysis did not converge fully.
+I modify their observation model in section \cref{ATACCC:sec:observation-modification} to improve the convergence.
+This modification is based on an earlier analysis I did with a slightly different model and a preliminary version of the dataset, which is described in \cref{ATACCC:sec:original-analysis}.
+The model and parameter estimates of the biomarker need to be converted to a duration of PCR positivity to be used in backcalculation, which I do in \cref{ATACCC:sec:duration}.
 
@@
 
+This section introduces the piecewise linear model I initially used to model viral loads.
+It consists of two steps:...
+
 \subsection{Model of viral load}
@@
 For an individual $i$, their trajectory is determined by two points and one slope parameter.
-The points are the time at which the individual has a 50\% chance of testing positive, $t_{i,0}$, and the time at which their viral load
-peaks, $t_{i,\text{peak}}$ after $t_{i,0}$; the slope parameter, $\beta_i$ determines the rate their Ct increases (viral load decreases) following the peak.
-The Ct value for an individual at time $t_{i,0}$ is the known limit of detection, $C_\text{lod}$,
-Their Ct value at the peak is $C_{i,\text{peak}}$ and inferred.
-The slope parameter, $\beta_i$, determines how quickly an individual's Ct value increases following the peak viral load.
-We denote the vector of the parameters defined here as $\vec\theta_i = \begin{bmatrix} t_{i,0} & t_{i,\text{peak}} & C_{i,\text{peak}} &  \beta_i \end{bmatrix}^T$.
+The points are the time at which the individual has a 50\% chance of testing positive, $t_{i,0}$, and the time at which their viral load peaks relative to their first swab, $t_{i,\text{peak}}$ after $t_{i,0}$; the slope parameter, $\beta_i$ determines the rate their viral load decreases following the peak.
+The viral load value for an individual at time $t_{i,0}$ is the known limit of detection, $v_\text{lod}$,
+Their viral load value value at the peak is $v_{i,\text{peak}}$ and inferred.
+The slope parameter, $\beta_i$, determines how quickly an individual's viral load decreases following the peak viral load.
+We denote the vector of the parameters defined here as $\vec\theta_i = \begin{bmatrix} t_{i,0} & t_{i,\text{peak}} & v_{i,\text{peak}} &  \beta_i \end{bmatrix}^T$.
 
-This parameterisation leads to the following Ct value trajectory for individual $i$:
+This parameterisation leads to the following value for individual $i$'s log viral load:
 $$
-\hat{C}_i(t, \vec\theta_i) = \begin{cases}
-  C_{\text{lod}} + (C_{i,\text{peak}} - C_{\text{lod}}) \frac{t - t_{i,0}}{t_{i,\text{peak}}}
+\hat{v}_i(t, \vec\theta_i) = \begin{cases}
+  v_{\text{lod}} + (v_{i,\text{peak}} - v_{\text{lod}}) \frac{t - t_{i,0}}{t_{i,\text{peak}}}
     &t \leq t_{i,\text{peak}} + t_{i,0} \\
-  C_{i,\text{peak}} + \beta_i (t - t_{i,\text{peak}} - t_{i,0})
+  v_{i,\text{peak}} - \beta_i (t - t_{i,\text{peak}} - t_{i,0})
     &t > t_{i,\text{peak}} + t_{i,0}.
@@
 
-Observed Ct values are assumed to be drawn from a Normal distribution with mean $\hat{C}_i(t)$ and standard deviation $\sigma_\text{obs}$.
-censored at the limit of detection.
+Observed log viral load values are assumed to be drawn from a Normal distribution with mean $\hat{v}_i(t)$ and standard deviation $\sigma_\text{obs}$, censored at the limit of detection.
 Additionally, we introduce a probability of a false negative $\eta$, to account for negative test results seen near the peak viral load.
-Denoting an observed Ct value for individual $i$ at time $t$ as $y_{i,t}$ with
+Denoting an observed log viral load value for individual $i$ at time $t$ as $y_{i,t}$ with
 $y_{i,t} = \infty$ for a negative test, this gives the following likelihood.
@@
 \pi(y_{i,t} \mid \vec\theta_i, \sigma_\text{obs}, \eta) = \begin{cases}
-  (1 - \eta) f_N(y_{i,t} \mid \hat{C}_i(t, \vec\theta_i), \sigma_\text{obs}^2) &y_{i,t} < \infty \\
-  \eta + (1 - \eta) (1 - F_N(C_\text{lod} \mid \hat{C}_i(t, \vec\theta_i), \sigma_\text{obs}^2)) & y_{i,t} = \infty
+  (1 - \eta) f_N(y_{i,t} \mid \hat{v}_i(t, \vec\theta_i), \sigma_\text{obs}^2) &y_{i,t} < \infty \\
+  \eta + (1 - \eta) F_N(v_\text{lod} \mid \hat{v}_i(t, \vec\theta_i), \sigma_\text{obs}^2) & y_{i,t} = \infty
 \end{cases}
@@
 \node[state, below right=of mu-tpeak] (ti-peak) {$t_{i,\text{peak}}$};
-\node[state, right=of sigma-logtpeak] (mu-Cpeak) {$\mu_{C_\text{peak}}$};
-\node[state, right=of mu-Cpeak] (sigma-Cpeak) {$\sigma_{C_\text{peak}}$};
-\node[state, below right=of mu-Cpeak] (Ci-peak) {$C_{i,\text{peak}}$};
-\node[state, right=of sigma-Cpeak] (mu-log-Bi) {$\mu_{\log \beta_i}$};
+\node[state, right=of sigma-logtpeak] (mu-vpeak) {$\mu_{v_\text{peak}}$};
+\node[state, right=of mu-vpeak] (sigma-vpeak) {$\sigma_{v_\text{peak}}$};
+\node[state, below right=of mu-vpeak] (vi-peak) {$v_{i,\text{peak}}$};
+\node[state, right=of sigma-vpeak] (mu-log-Bi) {$\mu_{\log \beta_i}$};
 \node[state, right=of mu-log-Bi] (sigma-log-Bi) {$\sigma_{\log \beta_i}$};
@@
 \node[draw, inner sep=0.3cm, fit=(ti0) (Bi), fill=blue!10,opacity=0.2] {$\theta_i$};
-\node[state, below left=of Ci-peak] (Ci-t) {$\hat{C}_i(t)$};
-\node[state, right=of Ci-t] (sigma-obs) {$\sigma_\text{obs}$};
+\node[state, below left=of vi-peak] (vi-t) {$\hat{v}_i(t)$};
+\node[state, right=of vi-t] (sigma-obs) {$\sigma_\text{obs}$};
 \node[state, right=of sigma-obs] (eta-i) {$\eta_i$};
-\node[rectangle, draw, left=of Ci-t,minimum size=1.5cm] (Clod) {$C_\text{lod}$};
+\node[rectangle, draw, left=of vi-t,minimum size=1.5cm] (vlod) {$v_\text{lod}$};
 \node[state, double, below=of sigma-obs] (Yi-t) {$y_{i,t}$};
@@
 \draw[->] (sigma-logtpeak) -- (ti-peak);
-\draw[->] (mu-Cpeak) -- (Ci-peak);
-\draw[->] (sigma-Cpeak) -- (Ci-peak);
+\draw[->] (mu-vpeak) -- (vi-peak);
+\draw[->] (sigma-vpeak) -- (vi-peak);
 \draw[->] (mu-log-Bi) -- (Bi);
 \draw[->] (sigma-log-Bi) -- (Bi);
-\draw[->,dotted] (ti0) -- (Ci-t);
-\draw[->,dotted] (ti-peak) -- (Ci-t);
-\draw[->,dotted] (Ci-peak) -- (Ci-t);
-\draw[->,dotted] (Bi) -- (Ci-t);
-%\draw[->] (theta-i) -- (Ci-t);
-\draw[->] (Ci-t) -- (Yi-t);
+\draw[->,dotted] (ti0) -- (vi-t);
+\draw[->,dotted] (ti-peak) -- (vi-t);
+\draw[->,dotted] (vi-peak) -- (vi-t);
+\draw[->,dotted] (Bi) -- (vi-t);
+%\draw[->] (theta-i) -- (vi-t);
+\draw[->] (vi-t) -- (Yi-t);
 \draw[->] (eta-i) -- (Yi-t);
-\draw[->,dotted] (Clod) -- (Ci-t);
-\draw[->] (Clod) -- (Yi-t);
+\draw[->,dotted] (vlod) -- (vi-t);
+\draw[->] (vlod) -- (Yi-t);
 \draw[->] (sigma-obs) -- (Yi-t);
@@
 \end{tikzpicture}
-\caption{Hierarchical structure}
+\caption[Hierarchical structure]{Hierarchical structure}
 \end{figure}
@@
     $\sigma_{\log t_\text{peak}}$ & $\text{Half-Cauchy}(0, 0.67)$ & $0.05$ & $0.67$ & $8.51$ \\ 
-    $\mu_{C_\text{peak}}$ & $N(17, 5^2)$ & $8.78$ & $17$ & $25.22$ \\ 
-    $\sigma_{C_\text{peak}}$ & $\text{Half-Cauchy}(0, 4)$ & $0.31$ & $4$ & $50.82$ \\ 
+    $\mu_{v_\text{peak}}$ & $N(17, 5^2)$ & $8.78$ & $17$ & $25.22$ \\ 
+    $\sigma_{v_\text{peak}}$ & $\text{Half-Cauchy}(0, 4)$ & $0.31$ & $4$ & $50.82$ \\ 
     $\mu_{\log\beta}$ & $N(\log(3.5), \log(1.4)^2)$ & $\log (2.01)$ & $\log (3.5)$ & $\log (6.09)$ \\ 
@@
 \section{Comparison to \texorpdfstring{\textcite{hakkiOnset}}{Hakki \etal (2022)}}
+
+\subsection{Viral load model}
+
+\subsection{Observation model}
+
+\subsection{Within-individual correlation}
 

### 2023-10-31 ATACCC-appendix-original-analysis.tex
Bulk of ATACCC chapter (no results yet)

--- 2023-10-26 ATACCC.tex
@@
 \begin{document}
-\ifSubfilesClassLoaded{
-  \setcounter{chapter}{4}
-}
 
-\chapter{Estimating duration using biomarkers} \label{ATACCC}
-
-\section{Background}
-
-Two major approaches have been taken to model SARS-CoV-2 viral load trajectories.
-There are either using ordinary differential equations (ODEs) with a biological interpretation or using mathematical functions that are flexible enough to capture the dynamics.
-
-ODE-based approaches have parameters represent biological properties of the virus and host, such as the immune response ~\autocites[e.g.:][]{ejimaEstimation,keVivo,kimQuantitative,goncalvesTiming,perelsonMechanistic}.
-Such models are rarely analytically-tractable and the system's evolution is computed using numeric methods.
-
-The second approach is to approximate the dynamics using general mathematical functions.
-Most common is a piecewise model, where the logarithm of viral load increases linearly, peaks instantaneously, and then decreases linearly~\autocites{clearyUsing,kisslerViral,larremoreTest}; although more flexible functions have also been suggested~\autocites{quiltyQuarantine}.
-In these models, parameters are clearly related to the data, such as the time from first detection to peak viral load, and the viral load at the peak.
-In both approaches, random effects are used to represent variation between hosts within the population, capturing the non-independence of longitudinal data from the same host.
-This chapter uses the latter class of models, adopting a piecewise-linear approach.
-
-Piecewise linear, random effect models have a long history of use for longitudinal data in biostatistics~\autocites{slateStatistical}.
-Often, as here, they are used to model the temporal evolution of biomarkers, including:
-CD4 counts following HIV infection~\autocites{langeHierarchical,lynchPredicting}, parasite density in
-blood samples following malaria treatment~\autocites{fogartyBayesian}, and measurements
-of cognitive decline due to Alzheimer's disease~\autocites{bealisleBayesian}.
-Various extensions to the model exist including the incorporation of multiple biomarkers~\autocites{inoueModeling,giardinaGetting} and relaxing the piecewise linear assumption to a semi-parameteric form~\autocites{zegerSemiparametric}.
-
-\section{Original analysis}
-
-\subsection{Model of viral load}
-
-For an individual $i$, their trajectory is determined by two points and one slope parameter.
-The points are the time at which the individual has a 50\% chance of testing positive, $t_{i,0}$, and the time at which their viral load
-peaks, $t_{i,\text{peak}}$ after $t_{i,0}$; the slope parameter, $\beta_i$ determines the rate their Ct increases (viral load decreases) following the peak.
-The Ct value for an individual at time $t_{i,0}$ is the known limit of detection, $C_\text{lod}$,
-Their Ct value at the peak is $C_{i,\text{peak}}$ and inferred.
-The slope parameter, $\beta_i$, determines how quickly an individual's Ct value increases following the peak viral load.
-We denote the vector of the parameters defined here as $\vec\theta_i = \begin{bmatrix} t_{i,0} & t_{i,\text{peak}} & C_{i,\text{peak}} &  \beta_i \end{bmatrix}^T$.
-
-This parameterisation leads to the following Ct value trajectory for individual $i$:
-$$
-\hat{C}_i(t, \vec\theta_i) = \begin{cases}
-  C_{\text{lod}} + (C_{i,\text{peak}} - C_{\text{lod}}) \frac{t - t_{i,0}}{t_{i,\text{peak}}}
-    &t \leq t_{i,\text{peak}} + t_{i,0} \\
-  C_{i,\text{peak}} + \beta_i (t - t_{i,\text{peak}} - t_{i,0})
-    &t > t_{i,\text{peak}} + t_{i,0}.
-\end{cases}
-$$
-
-Day 0 for an individual is the day they were enrolled in the study and first swabbed.
-
-\subsection{Observation model and likelihood}
-
-Observed Ct values are assumed to be drawn from a Normal distribution with mean $\hat{C}_i(t)$ and standard deviation $\sigma_\text{obs}$.
-censored at the limit of detection.
-Additionally, we introduce a probability of a false negative $\eta$, to account for negative test results seen near the peak viral load.
-Denoting an observed Ct value for individual $i$ at time $t$ as $y_{i,t}$ with
-$y_{i,t} = \infty$ for a negative test, this gives the following likelihood.
-$$
-\pi(y_{i,t} \mid \vec\theta_i, \sigma_\text{obs}, \eta) = \begin{cases}
-  (1 - \eta) f_N(y_{i,t} \mid \hat{C}_i(t, \vec\theta_i), \sigma_\text{obs}^2) &y_{i,t} < \infty \\
-  \eta + (1 - \eta) (1 - F_N(C_\text{lod} \mid \hat{C}_i(t, \vec\theta_i), \sigma_\text{obs}^2)) & y_{i,t} = \infty
-\end{cases}
-$$
-where $\sigma_\text{obs}$ is a model parameter determining the variability in
-test results, and $f_N(y \mid \mu, \sigma^2)$ and $F_N(y \mid \mu, \sigma^2)$ are the pdf
-and cdf (respectively) of a Normal distribution with mean $\mu$ and variance $\sigma^2$.
-
-We assume that all observations (both of the same and different individuals) are independent conditional on the model parameters.
-The parameter $\vec\theta_i$ introduces dependence between observations of the same individuals.
-\autoref{ATACCC:sec:hierarchy} explains the hierarchical structure which borrows information between individuals.
-
-\subsection{Hierarchical structure}\label{ATACCC:sec:hierarchy}
-
-\begin{figure}
-\begin{tikzpicture}[
-    node distance = 3cm and 3cm,
-    on grid,
-    auto,
-    ->,>=stealth',
-    every state/.style={draw,circle,minimum size=1.5cm},
-    ]
-
-% Define the nodes
-\node[state] (mu-tpeak) {$\mu_{\log t_\text{peak}}$};
-\node[state, right=of mu-tpeak] (sigma-logtpeak) {$\sigma_{\log t_\text{peak}}$};
-\node[state, below right=of mu-tpeak] (ti-peak) {$t_{i,\text{peak}}$};
-\node[state, right=of sigma-logtpeak] (mu-Cpeak) {$\mu_{C_\text{peak}}$};
-\node[state, right=of mu-Cpeak] (sigma-Cpeak) {$\sigma_{C_\text{peak}}$};
-\node[state, below right=of mu-Cpeak] (Ci-peak) {$C_{i,\text{peak}}$};
-\node[state, right=of sigma-Cpeak] (mu-log-Bi) {$\mu_{\log \beta_i}$};
-\node[state, right=of mu-log-Bi] (sigma-log-Bi) {$\sigma_{\log \beta_i}$};
-\node[state, below right=of mu-log-Bi] (Bi) {$\beta_i$};
-\node[state, left=of ti-peak] (ti0) {$t_{i,0}$};
-\node[draw, inner sep=0.3cm, fit=(ti0) (Bi), fill=blue!10,opacity=0.2] {$\theta_i$};
-\node[state, below left=of Ci-peak] (Ci-t) {$\hat{C}_i(t)$};
-\node[state, right=of Ci-t] (sigma-obs) {$\sigma_\text{obs}$};
-\node[state, right=of sigma-obs] (eta-i) {$\eta_i$};
-\node[rectangle, draw, left=of Ci-t,minimum size=1.5cm] (Clod) {$C_\text{lod}$};
-\node[state, double, below=of sigma-obs] (Yi-t) {$y_{i,t}$};
-
-% Draw the edges
-\draw[->] (mu-tpeak) -- (ti-peak);
-\draw[->] (sigma-logtpeak) -- (ti-peak);
-\draw[->] (mu-Cpeak) -- (Ci-peak);
-\draw[->] (sigma-Cpeak) -- (Ci-peak);
-\draw[->] (mu-log-Bi) -- (Bi);
-\draw[->] (sigma-log-Bi) -- (Bi);
-\draw[->,dotted] (ti0) -- (Ci-t);
-\draw[->,dotted] (ti-peak) -- (Ci-t);
-\draw[->,dotted] (Ci-peak) -- (Ci-t);
-\draw[->,dotted] (Bi) -- (Ci-t);
-%\draw[->] (theta-i) -- (Ci-t);
-\draw[->] (Ci-t) -- (Yi-t);
-\draw[->] (eta-i) -- (Yi-t);
-\draw[->,dotted] (Clod) -- (Ci-t);
-\draw[->] (Clod) -- (Yi-t);
-\draw[->] (sigma-obs) -- (Yi-t);
-
-\end{tikzpicture}
-\caption{Hierarchical structure}
-\end{figure}
-
-\subsection{Inference}
-
-Inference was performed by Hamiltonian Monte Carlo (HMC) implemented in Stan 2.26.1~[REF].
-To aid with mixing, the hierarchical parameters were implemented using the non-centred parameterisation~[REF].
-In addition, $t_{i,0}$ was reparameterised in terms of the absolute time of peak.
-Convergence was assessed by checking R-hat < 1.1, and visual inspection of traceplots, and effective sample size.
-
-\subsection{Priors}
-
-Weakly informative priors were used on all parameters except $\mu_{\log t_\text{peak}}$ and $\sigma_{\log t_\text{peak}}$.
-The latter were chosen to reflect the prior belief that viral load peaks shortly before symptom onset, which occurs a median of 5.1 days after infection~\autocite{mcaloonIncubation}.
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{llrrr}
-  \hline
-    Parameter & Distribution & 0.05 & 0.5 & 0.95 \\ 
-  \hline
-    $\eta$ & $\text{Beta}(0.95, 9)$ & $0$ & $0.07$ & $0.28$ \\ 
-    $\sigma_\text{obs}$ & $\text{Half-Cauchy}(0, 0.67)$ & $0.05$ & $0.67$ & $8.51$ \\ 
-    $t_{0,i}$ & $N(0, 5^2)$ & $-8.22$ & $0$ & $8.22$ \\ 
-    $\mu_{\log t_\text{peak}}$ & $N(\log(3), \log(2.5)^2)$ & $\log (0.66)$ & $\log (3)$ & $\log (13.54)$ \\ 
-    $\sigma_{\log t_\text{peak}}$ & $\text{Half-Cauchy}(0, 0.67)$ & $0.05$ & $0.67$ & $8.51$ \\ 
-    $\mu_{C_\text{peak}}$ & $N(17, 5^2)$ & $8.78$ & $17$ & $25.22$ \\ 
-    $\sigma_{C_\text{peak}}$ & $\text{Half-Cauchy}(0, 4)$ & $0.31$ & $4$ & $50.82$ \\ 
-    $\mu_{\log\beta}$ & $N(\log(3.5), \log(1.4)^2)$ & $\log (2.01)$ & $\log (3.5)$ & $\log (6.09)$ \\ 
-    $\sigma_{\log\beta}$ & $\text{Half-Cauchy}(0, 0.67)$ & $0.05$ & $0.67$ & $8.51$ \\
-   \hline
-\end{tabular}
-\caption{Priors used, columns 3 to 5 gives quantiles of the distributions specified. See main text for description of the parameters.}
-\label{tab:paper:priors}
-\end{table}
-
-\section{Comparison to \texorpdfstring{\textcite{hakkiOnset}}{Hakki \etal (2022)}}
-
-\section{Final analysis}
-
-\section{Producing duration from parameter estimates}
-
-The ATACCC analysis produces a posterior distribution for some population-level parameters, $\phi$.
-Taking a posterior sample for this, $\phi^{(i)}$, we can then draw the parameters which determine the viral
-load trajectory for a random individual $j$, $\theta_j$, from which that individual's duration $d_j^{(i)}$ can be calculated.
-By drawing $\theta_1, \theta_2, \dots, \theta_n$ from the distribution $p(\theta_j \mid\phi^{(i)})$, the density of the duration distribution, $f_A(t)$, can be
-calculated for the posterior sample $i$ as:
-$$
-f^{(i)}_A(t) = \frac{1}{n}\sum_{j=1}^n \mathbb{I} \left( t-0.5 < d_j^{(i)} < t+0.5 \right)
-$$
-Given $f_A(t)$, the cumulative density and hazard functions can be calculated.
-Repeating this for $N$ posterior samples of $\phi$ gives the posterior distribution.
-I have used $N = 1000$ and $n = 100,000$, 
-Since implementing this solution, I have found that Monte Carlo integration to calculate durations given $\theta_i$ is more accurate by avoiding discretisation errors.
-However, the results here do not use this for consistency.
+\chapter{Original ATACCC analysis} \label{ATACCC:sec:original-analysis}
 

### 2023-10-31 ATACCC.tex
Bulk of ATACCC chapter (no results yet)

--- 2023-10-26 ATACCC.tex
@@
 
-\section{Background}
+\todo[inline]{Needs intro paragraph, remember to mention using ATACCC data and ref section}
+
+\Cref{ATACCC:sec:background} explains how biomarkers are typically modelled, both for SARS-CoV-2 and historically in the biostatistical literature.
+I explain \textcite{hakkiOnset}'s application of these methods to the ATACCC data in \cref{ATACCC:sec:hakki}.
+However, this analysis did not converge fully.
+I modify their observation model in section \cref{ATACCC:sec:observation-modification} to improve the convergence.
+This modification is based on an earlier analysis I did with a slightly different model and a preliminary version of the dataset, described in \cref{ATACCC:sec:original-analysis}.
+I convert the model and parameter estimates of the biomarker to a duration of PCR positivity in \cref{ATACCC:sec:duration}.
+This can then be used for backcalculation and as a prior for the CIS analysis.
+
+\section{Background} \label{ATACCC:sec:background}
 
@@
 There are either using ordinary differential equations (ODEs) with a biological interpretation or using mathematical functions that are flexible enough to capture the dynamics.
+Both of these approaches have been used with other viruses and biomarkers.
 
@@
 Most common is a piecewise model, where the logarithm of viral load increases linearly, peaks instantaneously, and then decreases linearly~\autocites{clearyUsing,kisslerViral,larremoreTest}; although more flexible functions have also been suggested~\autocites{quiltyQuarantine}.
-In these models, parameters are clearly related to the data, such as the time from first detection to peak viral load, and the viral load at the peak.
+In these models, parameters are clearly related to features of the biomarker's trajectory, such as the time from first detection to peak viral load, and the viral load at the peak.
 In both approaches, random effects are used to represent variation between hosts within the population, capturing the non-independence of longitudinal data from the same host.
-This chapter uses the latter class of models, adopting a piecewise-linear approach.
+This chapter uses the model of \textcite{hakkiOnset}.
+This model is approximately piecewise log-linear, however, a slight adjustment is made which means that the model is continuos (see \cref{ATACCC:sec:hakki} for details).
+
+Random effects are used to capture the variation between individuals and the correlation between measurements within the same individual.
+In this context, a random effect means that parameter values are specific to each individual, rather than being a population average.
+The parameter's distribution across the population is assumed to be some known distribution, normally a normal or log-normal distribution~\autocite[chapter 24]{lashModern}.
+Specifically, if $\phi_j$ is the value of some parameter for individual $j$, $\mu_\phi$ is the population mean of $\phi$, and $\sigma_\phi$ is the population standard deviation, then we model $\phi_j$ as: $\phi_j \sim \N(\mu_\phi, \sigma_\phi^2)$.
+$\mu_\phi$ and $\sigma_\phi$ are parameters; in a Bayesian setting, a prior is placed on them in the normal way.
+This generalises to a vector of parameters per individual, $\vec{\phi}_j$, in the obvious way.
+In this chapter, $\mu_\phi$ and $\sigma_\phi$ are the parameters of interest, because we want to generalise our model across the population.
+The $\phi_j$s, which tells us about the parameter only for individuals in our sample, are considered nuisance parameters.
+
+Following much of the literature referenced in this section, I will take a Bayesian approach in this chapter.
+In the Bayesian context, random effect models are often referred to as hierarchical or multilevel models because they can be interpreted as placing a prior at two levels of a parameter hierarchy~\autocite[chapter 5]{gelmanBDA}.
+$\phi_j$ would be referred to as the bottom level of the hierarchy, with $\mu_j$ and $\sigma_j$ being the top level.
+More complex models could have more levels, for instance if the individuals are grouped into subpopulations.
 
 Piecewise linear, random effect models have a long history of use for longitudinal data in biostatistics~\autocites{slateStatistical}.
-Often, as here, they are used to model the temporal evolution of biomarkers, including:
-CD4 counts following HIV infection~\autocites{langeHierarchical,lynchPredicting}, parasite density in
-blood samples following malaria treatment~\autocites{fogartyBayesian}, and measurements
-of cognitive decline due to Alzheimer's disease~\autocites{bealisleBayesian}.
-Various extensions to the model exist including the incorporation of multiple biomarkers~\autocites{inoueModeling,giardinaGetting} and relaxing the piecewise linear assumption to a semi-parameteric form~\autocites{zegerSemiparametric}.
+Often, as here, they are used to model the temporal evolution of biomarkers.
+A few example applications are: CD4 counts following HIV infection~\autocites{langeHierarchical,lynchPredicting}, parasite density in blood samples following malaria treatment~\autocites{fogartyBayesian}, and measurements of cognitive decline due to Alzheimer's disease~\autocites{bealisleBayesian}.
 
-\section{Original analysis}
+\section{\texorpdfstring{\textcite{hakkiOnset}}{Hakki \etal (2022)}} \label{ATACCC:sec:hakki}
 
-\subsection{Model of viral load}
+The model of \textcite{hakkiOnset} has two components: a viral load model and an observation model.
+The viral load model produces a determinstic trajectory for the viral load of an individual, based on the model parameters.
+The observation model describes how the PCR test noisly observes this viral load.
+It includes the possibility of false positive and negatives, and the limit of detection of the PCR test.
+\todo{Make sure limit of detection is introduced previously when I first talk about PCR}
+% The limit of detection means that viral loads below a certain threshold are not detected by the PCR test.
+% They can be represented as a censoring of the viral load at the limit of detection.
 
-For an individual $i$, their trajectory is determined by two points and one slope parameter.
-The points are the time at which the individual has a 50\% chance of testing positive, $t_{i,0}$, and the time at which their viral load
-peaks, $t_{i,\text{peak}}$ after $t_{i,0}$; the slope parameter, $\beta_i$ determines the rate their Ct increases (viral load decreases) following the peak.
-The Ct value for an individual at time $t_{i,0}$ is the known limit of detection, $C_\text{lod}$,
-Their Ct value at the peak is $C_{i,\text{peak}}$ and inferred.
-The slope parameter, $\beta_i$, determines how quickly an individual's Ct value increases following the peak viral load.
-We denote the vector of the parameters defined here as $\vec\theta_i = \begin{bmatrix} t_{i,0} & t_{i,\text{peak}} & C_{i,\text{peak}} &  \beta_i \end{bmatrix}^T$.
-
-This parameterisation leads to the following Ct value trajectory for individual $i$:
-$$
-\hat{C}_i(t, \vec\theta_i) = \begin{cases}
-  C_{\text{lod}} + (C_{i,\text{peak}} - C_{\text{lod}}) \frac{t - t_{i,0}}{t_{i,\text{peak}}}
-    &t \leq t_{i,\text{peak}} + t_{i,0} \\
-  C_{i,\text{peak}} + \beta_i (t - t_{i,\text{peak}} - t_{i,0})
-    &t > t_{i,\text{peak}} + t_{i,0}.
-\end{cases}
-$$
-
-Day 0 for an individual is the day they were enrolled in the study and first swabbed.
-
-\subsection{Observation model and likelihood}
-
-Observed Ct values are assumed to be drawn from a Normal distribution with mean $\hat{C}_i(t)$ and standard deviation $\sigma_\text{obs}$.
-censored at the limit of detection.
-Additionally, we introduce a probability of a false negative $\eta$, to account for negative test results seen near the peak viral load.
-Denoting an observed Ct value for individual $i$ at time $t$ as $y_{i,t}$ with
-$y_{i,t} = \infty$ for a negative test, this gives the following likelihood.
-$$
-\pi(y_{i,t} \mid \vec\theta_i, \sigma_\text{obs}, \eta) = \begin{cases}
-  (1 - \eta) f_N(y_{i,t} \mid \hat{C}_i(t, \vec\theta_i), \sigma_\text{obs}^2) &y_{i,t} < \infty \\
-  \eta + (1 - \eta) (1 - F_N(C_\text{lod} \mid \hat{C}_i(t, \vec\theta_i), \sigma_\text{obs}^2)) & y_{i,t} = \infty
-\end{cases}
-$$
-where $\sigma_\text{obs}$ is a model parameter determining the variability in
-test results, and $f_N(y \mid \mu, \sigma^2)$ and $F_N(y \mid \mu, \sigma^2)$ are the pdf
-and cdf (respectively) of a Normal distribution with mean $\mu$ and variance $\sigma^2$.
-
-We assume that all observations (both of the same and different individuals) are independent conditional on the model parameters.
-The parameter $\vec\theta_i$ introduces dependence between observations of the same individuals.
-\autoref{ATACCC:sec:hierarchy} explains the hierarchical structure which borrows information between individuals.
-
-\subsection{Hierarchical structure}\label{ATACCC:sec:hierarchy}
-
+Viral loads exhibit exponential growth followed by exponential decline.
+\Textcite{hakkiOnset} use a model which is approximately piecewise linear but continuous.
+A continuous viral load trajectory should give better performance when using gradient-based inference methods such as Hamiltonian Monte Carlo (HMC)\todo{cite paper or ref section}.
+The model for the viral load at time $t$, $\hat{v}_i(t)$, in individual $i$ is:
+\begin{align}
+\hat{v}_i(\tau) = v_{\max,i} \frac{a_i+b_i}{b_ie^{-a_i(\tau-\tau_{\max,i})} + a_ie^{b_i(\tau-\tau_{max,i})}} \label{ATACCC:eq:viral-load}
+\end{align}
+Time is continuous and relative to the time that individual $i$ was first tested.
+$v_{\max,i}$ is the maximum viral load, which occurs at time $\tau_{\max,i}$.
+As time moves away from the peak (\ie becomes large in magnitude), the viral load trajectory tends towards exponential growth (before the peak) or decline (after the peak) in viral load at rates $a_i$ and $b_i$ respectively.
+Before the peak, this is because $b(\tau-\tau_{\max})$ is small and hence $a\exp(b(\tau-\tau_{\max}))$ is negligible.
+After the peak, this is because $b\exp(-a(\tau-\tau_{\max}))$ is negligible.
+This happens quickly for realistic values of the parameters.
+Therefore, the log viral load is approximately piecewise linear (see \cref{ATACCC:fig:viral-load-model} for a representative viral load trajectory).
+$\vec{\phi}_i = (v_{\max,i}, a_i, b_i)^T$ are the random effects for individual $i$.
+$\tau_i$ is inferred independently for each individual because the time from first test to peak is not a biological concept shared between individuals.
 \begin{figure}
-\begin{tikzpicture}[
-    node distance = 3cm and 3cm,
-    on grid,
-    auto,
-    ->,>=stealth',
-    every state/.style={draw,circle,minimum size=1.5cm},
-    ]
-
-% Define the nodes
-\node[state] (mu-tpeak) {$\mu_{\log t_\text{peak}}$};
-\node[state, right=of mu-tpeak] (sigma-logtpeak) {$\sigma_{\log t_\text{peak}}$};
-\node[state, below right=of mu-tpeak] (ti-peak) {$t_{i,\text{peak}}$};
-\node[state, right=of sigma-logtpeak] (mu-Cpeak) {$\mu_{C_\text{peak}}$};
-\node[state, right=of mu-Cpeak] (sigma-Cpeak) {$\sigma_{C_\text{peak}}$};
-\node[state, below right=of mu-Cpeak] (Ci-peak) {$C_{i,\text{peak}}$};
-\node[state, right=of sigma-Cpeak] (mu-log-Bi) {$\mu_{\log \beta_i}$};
-\node[state, right=of mu-log-Bi] (sigma-log-Bi) {$\sigma_{\log \beta_i}$};
-\node[state, below right=of mu-log-Bi] (Bi) {$\beta_i$};
-\node[state, left=of ti-peak] (ti0) {$t_{i,0}$};
-\node[draw, inner sep=0.3cm, fit=(ti0) (Bi), fill=blue!10,opacity=0.2] {$\theta_i$};
-\node[state, below left=of Ci-peak] (Ci-t) {$\hat{C}_i(t)$};
-\node[state, right=of Ci-t] (sigma-obs) {$\sigma_\text{obs}$};
-\node[state, right=of sigma-obs] (eta-i) {$\eta_i$};
-\node[rectangle, draw, left=of Ci-t,minimum size=1.5cm] (Clod) {$C_\text{lod}$};
-\node[state, double, below=of sigma-obs] (Yi-t) {$y_{i,t}$};
-
-% Draw the edges
-\draw[->] (mu-tpeak) -- (ti-peak);
-\draw[->] (sigma-logtpeak) -- (ti-peak);
-\draw[->] (mu-Cpeak) -- (Ci-peak);
-\draw[->] (sigma-Cpeak) -- (Ci-peak);
-\draw[->] (mu-log-Bi) -- (Bi);
-\draw[->] (sigma-log-Bi) -- (Bi);
-\draw[->,dotted] (ti0) -- (Ci-t);
-\draw[->,dotted] (ti-peak) -- (Ci-t);
-\draw[->,dotted] (Ci-peak) -- (Ci-t);
-\draw[->,dotted] (Bi) -- (Ci-t);
-%\draw[->] (theta-i) -- (Ci-t);
-\draw[->] (Ci-t) -- (Yi-t);
-\draw[->] (eta-i) -- (Yi-t);
-\draw[->,dotted] (Clod) -- (Ci-t);
-\draw[->] (Clod) -- (Yi-t);
-\draw[->] (sigma-obs) -- (Yi-t);
-
-\end{tikzpicture}
-\caption{Hierarchical structure}
+  \centering \includegraphics{ATACCC/typical_trajectory}
+  \caption[Typical viral load model trajectory.]{Typical viral load trajectory produced by the \textcite{hakkiOnset} viral load model.
+The majority of the trajectory is indistinguishable from linear but there is smoothing at the transition between growth and decline in viral load around time 0.
+\label{ATACCC:fig:viral-load-model}}
 \end{figure}
 
-\subsection{Inference}
+PCR tests are noisy observations of the viral load.
+A positive PCR test has a Ct value.
+Ct values are linearly related to viral load (see\todo{ref intro section}).
+A negative PCR test signals either the viral load is too low to be detected, \ie below the limit of detection, or the test is a false negative.
+Positive PCR tests, especially those with low viral loads, could be false positives, for example caused by contamination.
 
-Inference was performed by Hamiltonian Monte Carlo (HMC) implemented in Stan 2.26.1~[REF].
-To aid with mixing, the hierarchical parameters were implemented using the non-centred parameterisation~[REF].
-In addition, $t_{i,0}$ was reparameterised in terms of the absolute time of peak.
-Convergence was assessed by checking R-hat < 1.1, and visual inspection of traceplots, and effective sample size.
+The noisy observations are modelled by assuming each measurement of log viral load has iid Normally-distributed error with mean 0 and standard deviation $\sigma_v$.
+The measurements are censored at the limit of detection, $\vlod$, meaning that any observation which would give a log viral load of less than $\vlod$ is instead observed as negative.
+False positives and negatives are modelled by allowing a test failure with probability $\rho$.
+Test failures are modelled as producing a log viral load reading drawn from $\N(x_0, \sigma_0^2)$ where $x_0$ and $\sigma_0^2$ are parameters to be estimated.
+These readings are censored is the same way as true readings.
+A censored test failure represents a false negative, otherwise the test failure represents a false positive.
+\textcite{hakkiOnset} uses a prior for $x_0$ centred on $\vlod$, which encodes a prior belief that false negatives and false positives occur with equal probability.
 
-\subsection{Priors}
+Let the observed log viral load for individual $i$ at time $t$ be $y_{i,t}$, with $y_{i,t}=0$ if the test is negative.
+Then, its likelihood is:
+\begin{align}
+p(y_{i,t} \mid \cdot) &= \begin{cases}
+    \rho f_N(y_{i,t} \mid x_0, \sigma_0^2) + (1 - \rho) f_N(y_{i,t} \mid \log \hat{v}_i(t), \sigma_v^2) & \text{if $y_{i,t} > 0$}\\
+    \rho F_N(v_\text{log} \mid x_0, \sigma_0^2) + (1 - \rho) F_N(\vlod \mid \log \hat{v}_i(t), \sigma_v^2) & \text{if $y_{i,t} = 0$} \label{ATACCC:eq:hakki-likelihood}
+\end{cases}
+\end{align}
+where $f_N(y \mid \mu, \sigma^2)$ and $F_N(y \mid \mu, \sigma^2)$ are the pdf and cdf at $x$ of a Normal distribution with mean $\mu$ and variance $\sigma^2$ respectively.
+The first term on each line is the likelihood contribution from test failures, either a false positive (top line) of false negative (bottom line).
+The second term on the first line is the likelihood of the observed log viral load from non-failed test results.
+The second term on the bottom line is the likelihood of the negative (censored) tests from non-failed test results.
 
-Weakly informative priors were used on all parameters except $\mu_{\log t_\text{peak}}$ and $\sigma_{\log t_\text{peak}}$.
-The latter were chosen to reflect the prior belief that viral load peaks shortly before symptom onset, which occurs a median of 5.1 days after infection~\autocite{mcaloonIncubation}.
+Conditional on all $\hat{v}_i$s, the observations are assumed independent.
+Because $\hat{v}_i$ is a deterministic transform of $\vec{\phi}_i$ and $\tau_i$, the observations are independent conditional on the model parameters, and the full likelihood is simply the product of the likelihoods of each observation.
 
-\begin{table}[ht]
-\centering
-\begin{tabular}{llrrr}
-  \hline
-    Parameter & Distribution & 0.05 & 0.5 & 0.95 \\ 
-  \hline
-    $\eta$ & $\text{Beta}(0.95, 9)$ & $0$ & $0.07$ & $0.28$ \\ 
-    $\sigma_\text{obs}$ & $\text{Half-Cauchy}(0, 0.67)$ & $0.05$ & $0.67$ & $8.51$ \\ 
-    $t_{0,i}$ & $N(0, 5^2)$ & $-8.22$ & $0$ & $8.22$ \\ 
-    $\mu_{\log t_\text{peak}}$ & $N(\log(3), \log(2.5)^2)$ & $\log (0.66)$ & $\log (3)$ & $\log (13.54)$ \\ 
-    $\sigma_{\log t_\text{peak}}$ & $\text{Half-Cauchy}(0, 0.67)$ & $0.05$ & $0.67$ & $8.51$ \\ 
-    $\mu_{C_\text{peak}}$ & $N(17, 5^2)$ & $8.78$ & $17$ & $25.22$ \\ 
-    $\sigma_{C_\text{peak}}$ & $\text{Half-Cauchy}(0, 4)$ & $0.31$ & $4$ & $50.82$ \\ 
-    $\mu_{\log\beta}$ & $N(\log(3.5), \log(1.4)^2)$ & $\log (2.01)$ & $\log (3.5)$ & $\log (6.09)$ \\ 
-    $\sigma_{\log\beta}$ & $\text{Half-Cauchy}(0, 0.67)$ & $0.05$ & $0.67$ & $8.51$ \\
-   \hline
+Individuals are split into groups based on their vaccination status.
+The first group is unvaccinated and infected with a variety of Pre-Alpha, Alpha, and Delta variants.
+The second group is vaccinated and all were infected by the Delta variant.
+My focus is on the first group since they best correspond to the infections in the time period used throughout this thesis (Autumn 2020).
+
+The parameters $\vec{\phi}_i$ are constrained to the positive reals.
+They are modelled as log-normal random effects with the hierarchical prior $\phi_i \dist \LN(\vec{\mu}_{\phi,g(i)}, \matr{\Sigma}_\phi)$.
+$g(i)$ is the group of individual $i$.
+$\vec{\mu}_{\phi,g} = (\mu_{v_{\max},g}, \mu_{a,g}, \mu_{b,g})^T$ is a vector of the population means of the parameters.
+The covariance matrix $\matr{\Sigma}_\phi$ controls the variability of the population and allows for the parameters to be correlated.
+For example, an individual with a high peak viral load may also clear the virus slowly because both have an underlying cause of a weak immune system.
+To improve sampling efficiency, the multivariate version of the non-centred parameterisation\todo{REF NCP} is used.
+This can be expressed as follows.
+\begin{align}
+  \vec{\eta}_i &\dist \N(0, \matr{C}_\phi) \\
+  \log \vec{\phi}_i &= \vec{\mu}_\phi + \vec{\sigma}_\phi \circ \vec{\eta}_i
+\end{align}
+where $\vec{\sigma}_\phi$ is a vector of the standard deviations of the parameters, $\matr{C}_\phi$ is a correlation matrix, and $\circ$ is element-wise multiplication.
+Combining $\matr{C}$ and $\vec{\sigma}_\phi$ allows recovering $\matr\Sigma_\phi$ because $(\matr{\Sigma}_\phi)_{ij} = (\vec{\sigma}_\phi)_i (\vec{\sigma}_\phi)_j (\matr{C}_\phi)_{ij}$.
+Correlation matrices have diagonal elements all 1, therefore this reparameterisation has the same number of parameters as the original model.
+
+\textcite{hakkiOnset} uses weakly informative priors, similar to those previously used~\autocite{singanayagamCommunity}.
+These are given in \cref{ATACCC:table:hakki-priors}.
+I reran \textcite{hakkiOnset}'s analysis using R~4.3.1~\autocite{R-4-3-1}.
+The inference was performed using RStan~2.32.3~\autocite{RStan-2-32-3} and Stan~2.32.2~\autocite{Stan-2-32-2}.
+Inference settings were the same as \textcite{hakkiOnset}.
+\begin{table}
+\begin{tabular}{l c l l}
+    Parameter description & Symbol & Distribution \\
+    \hline \\
+    Mean max viral load & $\mu_{v_\text{max}, g}$ & $\N(15, 15^2)$ \\
+    Standard deviation of max log viral load & $\sigma_{v_\text{max}}$ & $\N_+(0, 10^2)$ \\
+    Mean rate of viral load increase & $\mu_{a, g}$ & $\N(1.25, 0.75^2)$ \\
+    Standard deviation of rate of viral load increase & $\sigma_{a}$ & $\N_+(0, 1)$ \\
+    Mean rate of viral load decrease & $\mu_{b, g}$ & $\N(0.5, 1.4^2)$ \\
+    Standard deviation of rate of viral load decrease & $\sigma_{b}$ & $\N_+(0, 1)$ \\
+    Correlation of viral load parameters & $\matr{C}_\phi$ & $\LKJ(1)$ \\
+    Negative log of rate of test failure & $-\log \rho$ & $\N(3, 0.5)$ \\
+    Mean log viral load for test failures & $x_0$ & $\N(0, 3^2)$ \\
+    Standard deviation of log viral load for test failures & $\sigma_0$ & $\N_+(3, 3^2)$ \\
+    Standard deviation of log viral load observations & $\sigma_v$ & $1 + \N_+(2, 3^2)$ \\
+    Time from first test to peak & $\tau_{\max,i}$ & $\N(5,5)$
 \end{tabular}
-\caption{Priors used, columns 3 to 5 gives quantiles of the distributions specified. See main text for description of the parameters.}
-\label{tab:paper:priors}
+\caption[SEIR model priors]{Priors for each parameter.
+For details of the distributions and their parameterisations see \cref{E-distributions}.
+The elements of $\vec{\mu}_{\phi,g}$ and $\vec{\sigma}_\phi$ have independent priors placed on them, as specified in the first lines of the table.
+Priors are the same for all $g$ and $i$ where not specified.
+$\N_+$ indicates a normal distribution truncated to the non-negative reals.
+\label{ATACCC:table:hakki-priors}}
 \end{table}
 
-\section{Comparison to \texorpdfstring{\textcite{hakkiOnset}}{Hakki \etal (2022)}}
+\section{Modifying the observation model} \label{ATACCC:sec:observation-modification}
 
-\section{Final analysis}
+The model of \textcite{hakkiOnset} did not converge fully (see \cref{ATACCC:sec:convergence}).
+In particular, the Rhat values, traceplots, and effective sample size for the parameters of the observation model were poor.
 
-\section{Producing duration from parameter estimates}
+Based on my previous analysis (see \cref{ATACCC:sec:original-analysis}) and the fit to the data, I hypothesised that the model for test failures was overcomplicated.
+False positives do not appear to be present in the dataset.
+Therefore, the model for test failures can be simplified to simply have a constant probability of a false negative.
+% These false negatives could represent a variety of scenarios, such as poor swabbing technique.
+Formally, the model can be expressed by replacing \cref{ATACCC:eq:hakki-likelihood} with the following.
+\begin{align}
+p(y_{i,t} \mid \cdot) &= \begin{cases}
+    (1 - \rho) f_N(y_{i,t} \mid \log \hat{v}_i(t), \sigma_v^2) & \text{if $y_{i,t} > 0$}\\
+    \rho + (1 - \rho) F_N(\vlod \mid \log \hat{v}_i(t), \sigma_v^2) & \text{if $y_{i,t} = 0$}
+\end{cases}
+\end{align}
 
-The ATACCC analysis produces a posterior distribution for some population-level parameters, $\phi$.
-Taking a posterior sample for this, $\phi^{(i)}$, we can then draw the parameters which determine the viral
-load trajectory for a random individual $j$, $\theta_j$, from which that individual's duration $d_j^{(i)}$ can be calculated.
-By drawing $\theta_1, \theta_2, \dots, \theta_n$ from the distribution $p(\theta_j \mid\phi^{(i)})$, the density of the duration distribution, $f_A(t)$, can be
-calculated for the posterior sample $i$ as:
+I kept all priors the same, including the prior on $\rho$, and performed inference with the same settings.
+The parameters $x_0$ and $\sigma_0$ are no longer used in the model.
+
+\section{Results and discussion} \label{ATACCC:sec:results-discussion}
+
+\subsection{Convergence} \label{ATACCC:sec:convergence}
+
+The original \textcite{hakkiOnset} model has convergence issues (Rhat > 1.07 and ESS < 100) for the parameters of the test failure distribution.
+No such issues occur in the modified model introduced in \cref{ATACCC:sec:observation-modification}.
+Both model have convergence issues (Rhat >1.4 and ESS<20) for some individual-level parameters, although with the simplified model this is only individual 55.
+
+Individual 55 has an extremely noisy set of observations.
+This means that two viral load trajectories fit the data well (shown in \cref{ATACCC:fig:individual-55}).
+This first, found by three chains, increases very quickly, peaks 4.0 days after the first test, then declines slowly
+The second, found by the other five chains, is more symmetrical, peaking 7.2 days after the first test.
+Because each individual's trajectory is not of interest for this analysis, and the population-level parameters agree between the two sets of chains, I consider this lack of convergence acceptable.
+\begin{figure}
+  \todo[inline]{Create figure with individual 55's trajectories}
+  \caption{\label{ATACCC:fig:individual-55}}
+\end{figure}
+
+\subsection{Parameter estimates} \label{ATACCC:sec:estimates}
+
+\subsection{Duration} \label{ATACCC:sec:duration}
+
+The ultimate aim of this chapter is to produce an estimate of the distribution for the duration of PCR positivity.
+So far, only the viral load trajectory has been considered.
+The time that individual $i$ first and last positive are the two values of $\tau$ such that $\log \hat{v}_i(\tau) = 0$.
+Denote these times as $\tau_{0,i}$ and $\tau_{1,i}$.
+Hence, their duration of PCR positivity, in continuous time, is $d_i = \tau_{1,i} - \tau_{0,i}$.
+
+At $\tau_{0,i}$ and $\tau_{1,i}$, the approximation as exponential growth (described in \cref{ATACCC:sec:hakki}) is good.
+Therefore, set $\log \hat{v}(\tau)$ and one of the exponential terms to 0 in \cref{ATACCC:eq:viral-load}.
+This gives the following expressions:
 $$
-f^{(i)}_A(t) = \frac{1}{n}\sum_{j=1}^n \mathbb{I} \left( t-0.5 < d_j^{(i)} < t+0.5 \right)
+\begin{aligned}
+\tau_{0,i}
+&= \tau_{\max,i} - \frac{1}{a_i} \left(\log v_{\max,i} + \log(a_i + b_i) - \log b_i \right) \\
+\tau_{1,i}
+&= \tau_{\max,i} + \frac{1}{b_i} \left(\log v_{\max,i} + \log(a_i + b_i) - \log a_i \right) \\
+d_i
+&= \tau_{1,i} - \tau_{0,i}  \\
+&= \frac{1}{a_i b_i} \left( (a_i + b_i) (\log v_{\max,i} + \log(a_i + b_i)) - a_i \log a_i - b_i \log b_i \right).
+\end{aligned}
 $$
-Given $f_A(t)$, the cumulative density and hazard functions can be calculated.
-Repeating this for $N$ posterior samples of $\phi$ gives the posterior distribution.
-I have used $N = 1000$ and $n = 100,000$, 
-Since implementing this solution, I have found that Monte Carlo integration to calculate durations given $\theta_i$ is more accurate by avoiding discretisation errors.
-However, the results here do not use this for consistency.
+Note that $d_i$ does not depend on $\tau_{\max,i}$, only the random effects.
+
+For later use, a discretised version is preferred (as described in \cref{E-inc-prev}).
+To calculate the discrete version of this distribution, $F_D$, a Monte Carlo integration is used with the following relation.
+\begin{align}
+  F_D(d)
+  &= \prob(d - 0.5 \leq D_i \leq d + 0.5) \\
+  &= \E_{D_i}(\I(d - 0.5 \leq D_i \leq d + 0.5)) 
+\end{align}
+where $\I$ is the indicator function.
+Here, $D_i$ is drawn conditional on the population-level parameters $\vec{\phi} = (\vec{\mu}_\phi, \matr{\Sigma}_\phi)^T$.
+A large number of Monte Carlo draws are needed to ensure stability of the estimates of extreme values of $F_D$.
+I found 100,000 draws produced stable estimates.
+
+\section{Conclusion}
+
+\ifSubfilesClassLoaded{
+  \appendix
+  \subfile{distributions}
+  \subfile{ATACCC-appendix-original-analysis}
+  \listoftodos
+}{}
 

### 2023-10-31 ATACCC/posteriors.R

@@ -1,0 +1,127 @@
+library(dplyr)
+library(ggplot2)
+library(purrr)
+library(rstan)
+library(tidybayes)
+library(tidyverse)
+source("utils.R")
+
+hakki = readRDS("ATACCC/fit.rds")
+modified = readRDS("ATACCC/fit2.rds")
+
+parameter_names = c("vmax", "a", "b")
+
+fit_to_cov = function(fit) {
+    corr_chol = rstan::extract(fit, pars = "Lc")[["Lc"]]
+    corr = apply(corr_chol, 1, function(x) x %*% t(x), simplify = FALSE)
+    sds = spread_draws(
+        hakki,
+        v_sd[i]
+    ) |>
+        ungroup()
+        
+    full_join(
+        sds |> rename(sd_i = v_sd),
+        sds |>
+            rename(j = i, sd_j = v_sd),
+        by = c(".chain", ".iteration", ".draw"),
+        relationship = "many-to-many"
+    ) |>
+        mutate(
+            cor = pmap_dbl(
+                list(.draw, i, j),
+                function(d, i, j) corr[[d]][i, j]
+            ),
+            name_i = parameter_names[i],
+            name_j = parameter_names[j],
+            cov = cor * sd_i * sd_j,
+        )
+}
+
+p_cov = bind_rows(
+    fit_to_cov(hakki) |>
+        mutate(
+            model = "Hakki"
+        ),
+    fit_to_cov(modified) |>
+        mutate(
+            model = "Modified"
+        )
+) |>
+    ggplot() +
+    geom_density(aes(cov, colour = model)) +
+    facet_grid(name_i ~ name_j)
+
+fit_to_mean = function(fit) {
+    fit |>
+        spread_draws(vgrow[i, g]) |>
+        filter(g == 1) |>
+        mutate(name = parameter_names[i])
+}
+
+bind_rows(
+    fit_to_mean(hakki) |>
+        mutate(
+            model = "Hakki"
+        ),
+    fit_to_mean(modified) |>
+        mutate(
+            model = "Modified"
+        )
+) |>
+    ggplot() +
+    geom_density(aes(vgrow, colour = model)) +
+    facet_grid(. ~ name, scales = "free_x")
+
+tibble_diagnostics = function(sims) {
+  tibble(
+    `Bulk-ESS` = ess_bulk(sims),
[…]

### 2023-11-02 ATACCC-appendix-original-analysis.tex
First full draft of ATACCC chapter

@@
 
+This chapter describes an early analysis I did on an initial version of the ATACCC dataset.
+This dataset is not publicly available.
+It contains 79 individuals, all those in the study at that time with at least one positive test.
+The duration distribution estimated is very similar to that in \cref{E-ATACCC} but more certain because it assumes that the random effects are drawn independently.
+
+\section{Methods}
+
+\subsection{Model of viral load}
+
+For an individual $i$, their trajectory is determined by two points and one slope parameter.
+The points are the time at which the individual has a 50\% chance of testing positive, $t_{i,0}$, and the time at which their viral load
+peaks, $t_{i,\text{peak}}$ after $t_{i,0}$; the slope parameter, $\beta_i$ determines the rate their Ct increases (viral load decreases) following the peak.
+The Ct value for an individual at time $t_{i,0}$ is the known limit of detection, $C_\text{lod}$,
+Their Ct value at the peak is $C_{i,\text{peak}}$ and inferred.
+The slope parameter, $\beta_i$, determines how quickly an individual's Ct value increases following the peak viral load.
+We denote the vector of the parameters defined here as $\vec\theta_i = \begin{bmatrix} t_{i,0} & t_{i,\text{peak}} & C_{i,\text{peak}} &  \beta_i \end{bmatrix}^T$.
+
+This parameterisation leads to the following Ct value trajectory for individual $i$:
+$$
+\hat{C}_i(t, \vec\theta_i) = \begin{cases}
+  C_{\text{lod}} + (C_{i,\text{peak}} - C_{\text{lod}}) \frac{t - t_{i,0}}{t_{i,\text{peak}}}
+    &t \leq t_{i,\text{peak}} + t_{i,0} \\
+  C_{i,\text{peak}} + \beta_i (t - t_{i,\text{peak}} - t_{i,0})
+    &t > t_{i,\text{peak}} + t_{i,0}.
+\end{cases}
+$$
+
+Day 0 for an individual is the day they were enrolled in the study and first swabbed.
+
+\subsection{Observation model and likelihood}
+
+Observed Ct values are assumed to be drawn from a Normal distribution with mean $\hat{C}_i(t)$ and standard deviation $\sigma_\text{obs}$.
+censored at the limit of detection.
+Additionally, we introduce a probability of a false negative $\eta$, to account for negative test results seen near the peak viral load.
+Denoting an observed Ct value for individual $i$ at time $t$ as $y_{i,t}$ with
+$y_{i,t} = \infty$ for a negative test, this gives the following likelihood.
+$$
+\pi(y_{i,t} \mid \vec\theta_i, \sigma_\text{obs}, \eta) = \begin{cases}
+  (1 - \eta) f_N(y_{i,t} \mid \hat{C}_i(t, \vec\theta_i), \sigma_\text{obs}^2) &y_{i,t} < \infty \\
+  \eta + (1 - \eta) (1 - F_N(C_\text{lod} \mid \hat{C}_i(t, \vec\theta_i), \sigma_\text{obs}^2)) & y_{i,t} = \infty
+\end{cases}
+$$
+where $\sigma_\text{obs}$ is a model parameter determining the variability in
+test results, and $f_N(y \mid \mu, \sigma^2)$ and $F_N(y \mid \mu, \sigma^2)$ are the pdf
+and cdf (respectively) of a Normal distribution with mean $\mu$ and variance $\sigma^2$.
+
+We assume that all observations (both of the same and different individuals) are independent conditional on the model parameters.
+The parameter $\vec\theta_i$ introduces dependence between observations of the same individuals.
+\autoref{sec:paper:hierarchy} explains the hierarchical structure which borrows information between individuals.
+
+\subsection{Hierarchical structure}\label{sec:paper:hierarchy}
+
+We assume that $\log t_{i,\text{peak}}$, $C_{i,\text{peak}}$, and $\log \beta_i$ are drawn independently from univariate normal distributions (details below).
+Other parameters are not hierarchical and have prior distributions placed on them in the normal way; \autoref{fig:paper:model_DAG} shows these relationships.
+As in \textcite{fogartyBayesian}, a log transform is applied to the parameters which must be positive for biological plausibility.
+As well as restricting these parameters to be positive, this enables the skewed distribution observed in SARS-CoV-2 with some individuals testing positive for much longer than the median.
+
+The formal specification of this, including the required population-level parameters is:
+\begin{align}
+    \log t_{i,\text{peak}} &\dist N(\mu_{\log t_\text{peak}}, \sigma_{\log t_\text{peak}}^2) 
+    \label{eq:paper:tpeak}\\
+    C_{i,\text{peak}} &\dist N(\mu_{C_\text{peak}}, \sigma_{C_\text{peak}}^2) 
+    \label{eq:paper:Cpeak}\\
+    \log \beta_i &\dist N(\mu_{\log \beta}, \sigma_{\log \beta}^2).
+    \label{eq:paper:beta}
+\end{align}
+
+\begin{figure}
+    \centering
+    \makebox[\textwidth][c]{\includegraphics[scale=0.8]{ATACCC/appendix/Bayesian_piecewise_DAG}}
+    \caption{Directed acyclic graph showing the relationship between parameters in the model.
+Dotted lines represent deterministic (functional) relationships and solid lines stochastic (distributional) relationships.
+The box represents a fixed (known) quantity and double circle an observed quantity (the data).
+Index $i$ ranges over individuals, and $t$ over time.
+Parameters indexed by $i$ exist once for each individual.
+See main text for equations governing the relationships and the distributions used.}
+    \label{fig:paper:model_DAG}
+\end{figure}
+
+\subsection{Inference}
+
+Inference was performed by Hamiltonian Monte Carlo (HMC) implemented in Stan 2.26.1~[REF].
+To aid with mixing, the hierarchical parameters were implemented using the non-centred parameterisation~[REF].
+In addition, $t_{i,0}$ was reparameterised in terms of the absolute time of peak.
+Convergence was assessed by checking R-hat < 1.1, and visual inspection of traceplots, and effective sample size.
+
+\subsection{Priors}
+
+Weakly informative priors were used on all parameters except $\mu_{\log t_\text{peak}}$ and $\sigma_{\log t_\text{peak}}$.
+The latter were chosen to reflect the prior belief that viral load peaks shortly before symptom onset, which occurs a median of 5.1 days after infection~\autocite{mcaloonIncubation}.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{llrrr}
+  \hline
+    Parameter & Distribution & 0.05 & 0.5 & 0.95 \\ 
+  \hline
+    $\eta$ & $\text{Beta}(0.95, 9)$ & $0$ & $0.07$ & $0.28$ \\ 
+    $\sigma_\text{obs}$ & $\text{Half-Cauchy}(0, 0.67)$ & $0.05$ & $0.67$ & $8.51$ \\ 
+    $t_{0,i}$ & $N(0, 5^2)$ & $-8.22$ & $0$ & $8.22$ \\ 
+    $\mu_{\log t_\text{peak}}$ & $N(\log(3), \log(2.5)^2)$ & $\log (0.66)$ & $\log (3)$ & $\log (13.54)$ \\ 
+    $\sigma_{\log t_\text{peak}}$ & $\text{Half-Cauchy}(0, 0.67)$ & $0.05$ & $0.67$ & $8.51$ \\ 
+    $\mu_{C_\text{peak}}$ & $N(17, 5^2)$ & $8.78$ & $17$ & $25.22$ \\ 
+    $\sigma_{C_\text{peak}}$ & $\text{Half-Cauchy}(0, 4)$ & $0.31$ & $4$ & $50.82$ \\ 
+    $\mu_{\log\beta}$ & $N(\log(3.5), \log(1.4)^2)$ & $\log (2.01)$ & $\log (3.5)$ & $\log (6.09)$ \\ 
+    $\sigma_{\log\beta}$ & $\text{Half-Cauchy}(0, 0.67)$ & $0.05$ & $0.67$ & $8.51$ \\
+   \hline
+\end{tabular}
+\caption{Priors used, columns 3 to 5 gives quantiles of the distributions specified. See main text for description of the parameters.}
+\label{tab:paper:priors}
+\end{table}
+
+\section{Results}
+
+\subsection{Goodness of fit}
+
+The model generally shows a good fit to the observed data (see \autoref{fig:paper:goodness_of_fit}).
+For individuals with few or no observations before their peak viral load, information is borrowed from other individuals to infer when this is likely to have occurred.
+Most observations, excluding those that look to be false negatives (negatives surrounded by two high viral load positives) fall within the 95\% predictive intervals.
+
+\begin{figure}
+    \centering
+    \vspace*{-3cm}
+    \makebox[\textwidth][c]{\includegraphics[scale=0.9]{ATACCC/appendix/goodness_of_fit}}
+    \caption{Predictive median (solid line), 50\% and 95\% predictive intervals (light and dark shading respectively) for test results (ignoring false negatives). Points show actual observations.}
+    \label{fig:paper:goodness_of_fit}
+\end{figure}
+
+\subsection{Parameter estimates}
+
+We estimate that the time from the time from when an individual has a 50\% chance of detection to peak viral load (equivalently: when $\hat{C} = C_\text{lod}$) is 2.4 (posterior median; with 95\% central CrI 1.5--3.3) days with a population standard deviation on the log scale of 0.69 (0.42--1.06).
+The mean peak viral corresponds to 17.8 (17.1--18.6) Ct units, with a standard deviation of 2.2 (1.5--3.2) Ct units, after which the decrease in viral corresponds to Ct values increasing by 1.6 (1.5--1.8) Ct / day, with a standard deviation on the log scale of 0.50 (0.40--0.61).
+Predictive distributions for these parameters for a new individual are shown in \autoref{fig:paper:indiv_predict}.
+We estimate a 6.8 (5.0--8.8) \% probability of a false negative independent of the viral load of an individual.
+
+\begin{figure}
+    \centering
+    \makebox[\textwidth][c]{\includegraphics{ATACCC/appendix/indiv_predictive}}
+    \caption{Posterior predictions of model parameters for a new, previously unobserved individual.}
+    \label{fig:paper:indiv_predict}
+\end{figure}
+
+The time between the two points when $\hat{C} = C_\text{lod}$ for an individual can be considered a duration of infection.
+This gives a mean duration estimated at 18.4 (16.5--20.9) days but the distribution is skewed, as seen by the median of 16.6 (14.8--18.5) days.
+The full distribution is in \autoref{fig:paper:duration}.
+\begin{figure}
+    \centering
+    \makebox[\textwidth][c]{\includegraphics{ATACCC/appendix/duration}}
+    \caption{Proportion of people with $\hat{C} > C_\text{lod}$ a given number of days after this first occurred. Shading indicates central 95\% credible interval.}
+    \label{fig:paper:duration}
+\end{figure}
+
+\section{Discussion}
+
+This study presents estimates of viral load dynamics based on a general population sample with daily swabbing.
+Our results show a peak Ct value around 5 Ct lower than \textcite{kisslerViral}, that is a higher viral load, although occurring at a similar time (see \autoref{tab:paper:compare}) and a longer time from first detection to peak viral load than \textcite{jonesEstimating}.
+This could be because of the different populations in each study.
+\Textcite{kisslerViral}'s data is collected from longitudinal swabbing of individuals professionally involved in the National Basketball Association (NBA).
+Of the individuals who were included in their analysis, around half were in professional athletes, likely in peak physical condition, and 90\% were male.
+Our data is collected from the general population who were reported as contacts of individuals detected through surveillance open to all members of the public.
+Therefore, our sample is likely to be, on average, in worse physical condition which may lead to a higher viral load (lower Ct).
+Another possible difference could be the lab the tests were performed in.
+Ct values are not fully comparable between different laboratories due to different testing methodologies~\autocites{dahdouhCt,hanRTPCR}.
+
+Our estimate for the duration of detectable infection, 18.4 (16.5--20.9) days, is similar to a recent meta-analysis~\autocite{cevikShedding}.
+However, since individuals were tested for a maximum of 21 days in this study, a large part of the distribution is extrapolated.
+
+\begin{table}[]
+    \centering
+    \makebox[\textwidth][c]{\begin{tabular}{p{7cm}ccc}
+    Quantity & This study & \Textcite{kisslerViral} \\ 
+    \hline
+    Mean time from first detectable to peak viral load / days & 2.4 (1.5--3.3) & 3.2 (2.4--4.2) \\
+    Mean Ct at peak viral load / Ct & 17.8 (17.1--18.6) & 22.4 (20.7--24.0) \\
+    Mean increase in Ct following peak / Ct / day & 1.6 (1.5--1.8) & 2.1 (1.7--2.6)
+    \end{tabular}}
+    \caption{Summary of results in comparison to previous studies. \Textcite{jonesEstimating} do not give values in terms of Ct.}
+    \label{tab:paper:compare}
+\end{table}
+
 \end{document}

### 2023-11-02 ATACCC.tex
First full draft of ATACCC chapter

@@
 
-\todo[inline]{Needs intro paragraph, remember to mention using ATACCC data and ref section}
+\Cref{E-inc-prev:sec:duration} showed that the duration of PCR positivity is a key parameter for estimating the incidence of SARS-CoV-2 infections.
+This chapter estimates the duration of PCR positivity using the viral load data from the ATACCC study (described in\todo{ref section describing ATACCC}).
+I model a biomarker, the log viral load, over the course of an infection.
+From this model, I derive the distribution of the duration of PCR positivity.
 
 \Cref{ATACCC:sec:background} explains how biomarkers are typically modelled, both for SARS-CoV-2 and historically in the biostatistical literature.
-I explain \textcite{hakkiOnset}'s application of these methods to the ATACCC data in \cref{ATACCC:sec:hakki}.
+Next, \cref{ATACCC:sec:hakki} explains \textcite{hakkiOnset}'s application of these methods to the ATACCC data.
 However, this analysis did not converge fully.
@@
   \centering \includegraphics{ATACCC/typical_trajectory}
-  \caption[Typical viral load model trajectory.]{Typical viral load trajectory produced by the \textcite{hakkiOnset} viral load model.
-The majority of the trajectory is indistinguishable from linear but there is smoothing at the transition between growth and decline in viral load around time 0.
+  \caption[Example viral load model trajectory.]{Example viral load trajectory produced by \textcite{hakkiOnset}'s viral load model, using realistic parameter values.
+The increase in log viral load is approximately linear with a gradient of $a_i$, here 5.
+The decrease is also linear, with gradient $-b_i$, here -1.4.
+The peak is at time $\tau_{\max,i}$, here 0, and log viral load $\log v_{\max,i}$, here = 15.
+The smoothing (lack of linearity) near the peak is visible.
 \label{ATACCC:fig:viral-load-model}}
@@
 $g(i)$ is the group of individual $i$.
-$\vec{\mu}_{\phi,g} = (\mu_{v_{\max},g}, \mu_{a,g}, \mu_{b,g})^T$ is a vector of the population means of the parameters.
+$\vec{\mu}_{\phi,g} = (\mu_{v_{\max},g}, \mu_{a,g}, \mu_{b,g})^T$ is a vector of the population means of the log-parameters.
+The random effects have a log-normal, rather than normal, distribution.
+Therefore, $\exp(\vec{\mu}_{\phi,g})$ is the population median of the parameter value but not the mean.
 The covariance matrix $\matr{\Sigma}_\phi$ controls the variability of the population and allows for the parameters to be correlated.
 For example, an individual with a high peak viral load may also clear the virus slowly because both have an underlying cause of a weak immune system.
-To improve sampling efficiency, the multivariate version of the non-centred parameterisation\todo{REF NCP} is used.
+To improve sampling efficiency, the multivariate version of the non-centred parameterisation~\autocites{papaspiliopoulosGeneral,stanReparameterization} is used.
 This can be expressed as follows.
@@
 \end{tabular}
-\caption[SEIR model priors]{Priors for each parameter.
-For details of the distributions and their parameterisations see \cref{E-distributions}.
+\caption[Viral load model priors]{Priors for each parameter, taken from \textcite{hakkiOnset}.
+$\N_+$ indicates a normal distribution truncated to the non-negative reals.
+Details of the distributions, including their parameterisations, are in \cref{E-distributions}.
 The elements of $\vec{\mu}_{\phi,g}$ and $\vec{\sigma}_\phi$ have independent priors placed on them, as specified in the first lines of the table.
-Priors are the same for all $g$ and $i$ where not specified.
-$\N_+$ indicates a normal distribution truncated to the non-negative reals.
+Priors are the same for all $g$ and $i$.
 \label{ATACCC:table:hakki-priors}}
@@
 
-The model of \textcite{hakkiOnset} did not converge fully (see \cref{ATACCC:sec:convergence}).
+The model of \textcite{hakkiOnset} did not converge fully (see \cref{ATACCC:sec:results}).
 In particular, the Rhat values, traceplots, and effective sample size for the parameters of the observation model were poor.
@@
 
-I kept all priors the same, including the prior on $\rho$, and performed inference with the same settings.
 The parameters $x_0$ and $\sigma_0$ are no longer used in the model.
+For all other parameters, I kept the priors the same as \textcite{hakkiOnset}, including the prior on $\rho$, and performed inference with the same settings.
 
-\section{Results and discussion} \label{ATACCC:sec:results-discussion}
-
-\subsection{Convergence} \label{ATACCC:sec:convergence}
-
-The original \textcite{hakkiOnset} model has convergence issues (Rhat > 1.07 and ESS < 100) for the parameters of the test failure distribution.
-No such issues occur in the modified model introduced in \cref{ATACCC:sec:observation-modification}.
-Both model have convergence issues (Rhat >1.4 and ESS<20) for some individual-level parameters, although with the simplified model this is only individual 55.
-
-Individual 55 has an extremely noisy set of observations.
-This means that two viral load trajectories fit the data well (shown in \cref{ATACCC:fig:individual-55}).
-This first, found by three chains, increases very quickly, peaks 4.0 days after the first test, then declines slowly
-The second, found by the other five chains, is more symmetrical, peaking 7.2 days after the first test.
-Because each individual's trajectory is not of interest for this analysis, and the population-level parameters agree between the two sets of chains, I consider this lack of convergence acceptable.
-\begin{figure}
-  \todo[inline]{Create figure with individual 55's trajectories}
-  \caption{\label{ATACCC:fig:individual-55}}
-\end{figure}
-
-\subsection{Parameter estimates} \label{ATACCC:sec:estimates}
-
-\subsection{Duration} \label{ATACCC:sec:duration}
-
+\section{Deriving duration from the parameters} \label{ATACCC:sec:duration}
 The ultimate aim of this chapter is to produce an estimate of the distribution for the duration of PCR positivity.
@@
 To calculate the discrete version of this distribution, $F_D$, a Monte Carlo integration is used with the following relation.
+\section{Results} \label{ATACCC:sec:results-discussion}
+
+The original \textcite{hakkiOnset} model has convergence issues (Rhat > 1.07 and ESS < 100) for the parameters of the test failure distribution.
+No such issues occur in the modified model introduced in \cref{ATACCC:sec:observation-modification}.
+Both model have convergence issues (Rhat > 1.4 and ESS < 20) for some individual-level parameters, although with the simplified model this is only individual 55.
+
+Individual 55 has an extremely noisy set of observations.
+This means that two viral load trajectories fit the data well (shown in \cref{ATACCC:fig:individual-55}).
+This first, found by two chains, increases very quickly, peaks 4.0 days after the first test, then declines slowly
+The second, found by the other six chains, is more symmetrical, peaking 7.2 days after the first test.
+Because each individual's trajectory is not of interest for this analysis, and the population-level parameters agree between the two sets of chains, I consider this lack of convergence acceptable.
+\begin{figure}
+  \centering \includegraphics{ATACCC/fit_individual_55}
+  \caption[Goodness-of-fit for individual 55]{Fit to individual 55's data coloured by MCMC chain.
+Two chains favour the quickly increasing viral load trajectory, peaking at day 4.0, the other six favour the slower trajectory, peaking at day 7.2.
+\label{ATACCC:fig:individual-55}}
+\end{figure}
+
+The modified model produced good fits to the data (see \cref{ATACCC:fig:goodness-of-fits}).
+Almost all test results are captured in the 95\% credible intervals.
+Individual 55 has some fitting issues, as explored above.
+Individual 57 appears to have far noisier measurements than the other individuals, suggesting that the observation noise might vary between individuals.
+Including a random effect on $\sigma_v$ could capture this affect.
+However, it would complicate the model, be unlikely to change the estimates of the parameters of interest, and may not be identifiable.
+
+\begin{figure}
+  \thisfloatpagestyle{empty}
+  \vspace{-3cm}
+  \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{ATACCC/fits}}
+  \captionsetup{width=0.8\paperwidth}
+  \captionof{figure}[Posterior predictive viral load fits]{Posterior predictive observed viral loads (median and 95\% CrI, lines and ribbon) for each individual, overlaid with their observed test results (dots).
+Posterior estimates are conditional on the result not being a false negative.
+Negative PCR tests are plotted at the limit of detection, \ie a log viral load of 0.
+Each plot is labelled arbitrarily with the individual's position in the dataset.
+\label{ATACCC:fig:goodness-of-fits}}
+\end{figure}
+
+The two models give very similar estimates for the population-level parameters (see \cref{ATACCC:fig:compare-hakki-modified}).
+These are the main parameters of interest (as explained in \cref{ATACCC:fig:duration}), suggesting that the results are robust to the choice of model.
+As the estimates are similar and the modified model converges better, I use the modified model for the remainder of this chapter.
+\begin{figure}
+  \centering \includegraphics{ATACCC/compare_hakki_modified}
+  \caption[Comparison of population-level parameters between models.]{Kernel-density smoothed posterior densities for the population mean (top), $\vec{\mu}_{\phi,1}$ and covariance matrix (bottom), $\matr{\Sigma}_\phi$, for both the \textcite{hakkiOnset} model (described in \cref{ATACCC:sec:hakki}) and the model with a modified observation model to improve convergence (described in \cref{ATACCC:sec:observation-modification}).
+Mean only shown for the unvaccinated group, the primary group of interest, the covariance matrix is shared.
+These parameters, which are the primary ones of interest (see \cref{ATACCC:sec:duration}) have very similar estimates between the models.
+\label{ATACCC:fig:compare-hakki-modified}}
+\end{figure}
+
+Clear differences emerge between the vaccinated and unvaccinated groups (see \cref{ATACCC:fig:median-trajectories} and \cref{ATACCC:table:population-estimates}).
+The rates of viral load increase at similar rates in the two groups, the vaccinated group have higher peak viral loads and clear the virus faster.
+This leads to a shorter duration of PCR positivity in the vaccinated group.
+\begin{figure}
+  \centering \includegraphics{ATACCC/mean_trajectories}
+  \caption[Mean viral load trajectories]{Trajectory (posterior median and 95\% CrI) using the median parameter values, $\exp \vec{\mu}_{\phi,g}$, excluding the within-population variation.
+These assume $\tau_{\max,i}=0$, however modifying this simply translates the trajectories left or right.
+\label{ATACCC:fig:mean-trajectories}}
+\end{figure}
+\begin{table}
+  \makebox[\linewidth]{
+  \begin{tabular}{lcc}
+    Parameter & Unvaccinated population & Vaccinated population \\
+    Max log viral load, $\mu_{v_\text{max},g}$ & 14.9 (14.2--15.7) & 16.0 (15.2--16.8) \\
+    Rate of log viral load increase, $\exp(\mu_{a,g})$ & 5.07 (3.70--7.06) & 1.25 (1.18--1.32) \\
+    Rate of log viral load decrease, $\exp(\mu_{b,g})$ & 1.33 (1.09--1.63) & 1.89 (1.50--2.36) \\
+  \end{tabular}
+  }
+  \caption{Population-level estimates of the viral load parameters (posterior medians and 95\% CrI).
+The vaccinated population have higher peak viral loads but clear the virus faster.
+\label{ATACCC:table:population-estimates}}
+\end{table}
+
 \begin{align}
   F_D(d)
-  &= \prob(d - 0.5 \leq D_i \leq d + 0.5) \\
-  &= \E_{D_i}(\I(d - 0.5 \leq D_i \leq d + 0.5)) 
+  &= \prob(D_i \leq d + 0.5) \\
+  &= \E_{D_i}(\I(D_i \leq d + 0.5)) 
 \end{align}
 where $\I$ is the indicator function.
-Here, $D_i$ is drawn conditional on the population-level parameters $\vec{\phi} = (\vec{\mu}_\phi, \matr{\Sigma}_\phi)^T$.
-A large number of Monte Carlo draws are needed to ensure stability of the estimates of extreme values of $F_D$.
+Here, $D_i$ is drawn conditional on the population-level parameters $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$.
+Many Monte Carlo draws are needed to ensure stability of the estimates of extreme values of $F_D$.
 I found 100,000 draws produced stable estimates.
 
-\section{Conclusion}
+The $F_D$ estimates for the two groups are in \cref{ATACCC:fig:duration}.
+In \cref{ATACCC:sec:estimates}, the two models gave very similar results.
+Therefore, I show results only from the modified model (which had better convergence).
+\begin{figure}
+  \centering \includegraphics{ATACCC/duration}
+  \caption[Duration of PCR positivity.]{Estimated distribution (survival function, defined in \cref{E-inc-prev:sec:duration}) of duration of PCR positivity for the two populations.
+The vaccinated population have shorter durations of PCR positivity.
+\label{ATACCC:fig:duration}}
+\end{figure}
+
+The vaccinated population have shorter durations of PCR positivity than the unvaccinated population (see \cref{ATACCC:fig:duration}).
+The median duration of PCR positivity for the vaccinated population is 13 days (95\% CrI: 11--16 days), but it is 16 days (95\% CrI: 14--19 days) for the unvaccinated population.
+
+\section{Discussion}
+
+This chapter has estimated the distribution of PCR positive duration based on the viral load data from the ATACCC study.
+
+Different estimates were produced for vaccinated and unvaccinated individuals in the study.
+However, there are many differences between individuals in these two groups.
+For example, the vaccinated individuals are likely to be older and/or have co-morbidities because the roll-out of the vaccine in the UK prioritised these groups~\autocite{naoCovidVaccination}.
+They were also infected by different variants because the vaccinated individuals were infected later.
+All these factors could change the viral dynamics~\autocite{russellWithinhost} and hence the duration.
+Therefore, the relationships shown here cannot be considered causal, or even generalisable associations to contexts where these other confounders are not present.
+
+A strength of the ATACCC data is that the sample is drawn from the general population.
+This is unlike most estimates (see\todo{ref review of lit}) which are from hospitalised patients or healthcare workers.
+This means that, notwithstanding the issues highlighted in the previous paragraph, the estimates are likely to be generalisable to the general population.
+
+In the following chapters, I will use the estimates from the unvaccinated individuals.
+The CIS data I use is from autumn 2020 and early 2021 when few individuals were vaccinated and the majority of infections were by pre-Alpha or Alpha variants.
+Therefore, this group of people should be fairly representative of those within CIS.
+
+A major issue with the estimates in this chapter is that they are extrapolated.
+In ATACCC, individuals were tested for at most 20 days.
+Yet, the estimates here sill have 33\% (95\% CrI: 21--46\%) of individuals positive at 20 days following infection and less than 1\% positive at 50 days.
+The remainder of the estimate for $f_D$ is driven by the assumed form of the viral load trajectory.
+However, the CIS has individuals testing positive for very long periods of time.
+In periods of low incidence following a large peak, the number of individuals who are currently positive is dominated by those who have been positive for a long time.
+Therefore, understanding if these probabilities are correct is important for understanding transmission in the period after a peak.
+These effects are even more prominent when incidence falls suddenly to low levels due to an intervention such as a lockdown.
+This is exactly the situation in the UK in early 2021.
+The next few chapters will incorporate CIS data to improve the estimates of these tail probabilities.
 

### 2023-11-02 distributions.tex
Bulk of ATACCC chapter (no results yet)
First full draft of ATACCC chapter

@@
 \begin{document}
-\chapter{Tables of distributions} \label{E-distributions}
+\chapter{Tables of distributions} \label{distributions}
+
+This appendix contains details of the distributions used in this thesis.
+
 \begin{landscape}
@@
 Name & Notation & Parameters & Mean & Variance & pdf\\
+\hline \\
 Normal & $\N(\mu, \sigma^2)$ & $\mu$ (mean)\newline $\sigma^2 > 0$ (variance) & $\mu$ & $\sigma^2$ & $\frac{1}{\sqrt{2\pi\sigma^2}}e^{-\frac{(x-\mu)^2}{2\sigma^2}}$ \\
@@
 Exponential & $\Exponential(\lambda)$ & $\lambda$ (rate) & $\frac{1}{\lambda}$ & $\frac{1}{\lambda^2}$ & $\lambda e^{-\lambda x}$ \\
+LKJ* & $\LKJ(\eta)$ & $\eta > 0$ (shape) & $\matr{I}$ (identity matrix) & N/A & $\det(\matr{x})^{\eta-1}$ \\
 \end{tabular}
+
+* The LKJ distribution has support over positive definite symmetric matrices $\matr{x}$.
+These matrices are exactly the matrices that are valid correlation matrices.
+The normalising constant, $C$, has a complex form and is rarely needed.
 
@@
 Name & Notation & Parameters & Mean & Variance & pmf \\
+\hline \\
 Binomial & $\Bin(n, p)$ & $n \in \nats$ (number of trials, integer)\newline $0 \leq p \leq 1$ (probability of success) & $np$ & $np(1-p)$ & ${n \choose x}p^x(1-p)^{n-x}$ \\
@@
 \end{tabular}
+
 $B(x,y)$ is the beta function, defined in terms of the Gamma function $\Gamma(x)$ as $B(x,y) = \frac{\Gamma(x)\Gamma(y)}{\Gamma(x+y)}$.

### 2023-11-02 ATACCC/duration.R

@@ -1,0 +1,45 @@
+library(dplyr)
+library(ggplot2)
+library(patchwork)
+library(purrr)
+library(tidybayes)
+library(tidyr)
+source(here::here("utils.R"))
+
+posterior_draws = readRDS(here::here("ATACCC-distributions/posterior_samples.rds"))
+posterior_draws2 = readRDS(here::here("ATACCC-distributions/posterior_samples2.rds"))
+
+# posterior_draws |>
+#     group_by(.draw) |>
+#     arrange(time, .by_group = TRUE) |>
+#     summarise(median = time[min(which(F >= 0.5))]) |>
+#     median_qi()
+
+# posterior_draws2 |>
+#     group_by(.draw) |>
+#     arrange(time, .by_group = TRUE) |>
+#     summarise(median = time[min(which(F >= 0.5))]) |>
+#     median_qi()
+
+p_durations = bind_rows(
+    posterior_draws |>
+        mutate(Group = "Unvaccinated"),
+    posterior_draws2 |>
+        mutate(Group = "Vaccinated")
+) |>
+    ggplot() +
+    stat_lineribbon(
+        aes(time, S, fill = Group, colour = Group),
+        alpha = 0.3,
+        .width = 0.95
+    ) +
+    theme_survival_time_series () +
+    coord_cartesian(xlim = c(0, 40))
+ggsave(
+    filename = here::here("ATACCC/duration.png"),
+    plot = p_durations,
+    width = 15,
+    height = 9,
+    units = "cm",
+    dpi = 300
+)

### 2023-11-02 ATACCC/goodness_fit.R

@@ -1,0 +1,104 @@
+library(dplyr)
+library(ggplot2)
+library(purrr)
+library(rstan)
+library(patchwork)
+library(tidybayes)
+library(tidyverse)
+source(here::here("ATACCC/trajectory_functions.R"))
+source(here::here("utils.R"))
+
+fit = here::here("ATACCC/fit.rds") |>
+    readRDS()
+
+tbl_data = here::here("ATACCC/data.rds") |>
+    readRDS() |>
+    mutate(group = case_when(
+        (vaccinated == FALSE) & (WGS == "Pre-Alpha") ~ 1,
+        (vaccinated == FALSE) & (WGS == "Alpha") ~ 1,
+        (vaccinated == FALSE) & (WGS == "Delta") ~ 1,
+        (vaccinated == TRUE) & (WGS == "Delta") ~ 2
+    )) |>
+    filter(!is.na(group), !is.na(copy)) |> 
+    mutate(
+        i = as.integer(factor(id_sub, levels=unique(id_sub))),
+        y = if_else(
+            copy == 1,
+            0,
+            log(copy) - 3.43
+        ),
+        result = if_else(
+            y > 0,
+            "Positive",
+            "Negative"
+        ),
+    )
+
+theme_trajectory = function() {
+  list(
+    standard_plot_theming(),
+    coord_cartesian(c(-1, 20), c(0, 20)),
+    scale_y_continuous(breaks = c(0, 10, 20), minor_breaks = seq(0, 20, 5)),
+    theme(
+      legend.position = "bottom"
+    ),
+    labs(
+      x = "Time since first test (days)",
+      y = "Log viral load",
+      colour = ""
+    ) 
+  )
+}
+
+posterior_predictive = spread_draws(fit, pred_vl[m], v_s) |>
+  ungroup() |>
+  mutate(
+    i = (m - 1) %/% 42 + 1,
+    day = (m - 1) %% 42 - 14,
+    yrep = rnorm(n(), pred_vl, v_s),
+  )
+
+p_55 = posterior_predictive |>
+  filter(i == 55) |>
+  group_by(.chain, day) |>
+  median_qi(yrep) |>
+  mutate(
+    .chain = factor(.chain),
+  ) |>
+  ggplot() +
+  geom_point(aes(day, y, colour = result), data = filter(tbl_data, i == 55)) +
+  geom_lineribbon(
[…]

### 2023-11-02 ATACCC/posteriors.R

@@ -2,12 +2,14 @@ library(dplyr)
 library(ggplot2)
 library(purrr)
 library(rstan)
+library(patchwork)
 library(tidybayes)
 library(tidyverse)
-source("utils.R")
+source(here::here("ATACCC/trajectory_functions.R"))
+source(here::here("utils.R"))
 
-hakki = readRDS("ATACCC/fit.rds")
-modified = readRDS("ATACCC/fit2.rds")
+hakki = readRDS(here::here("ATACCC/fit.rds"))
+modified = readRDS(here::here("ATACCC/fit2.rds"))
 
 parameter_names = c("vmax", "a", "b")
 
@@ -38,6 +40,13 @@ fit_to_cov = function(fit) {
         )
 }
 
+fit_to_mean = function(fit, g_to_show = 1) {
+    fit |>
+        spread_draws(vgrow[i, g]) |>
+        filter(g == g_to_show) |>
+        mutate(name = parameter_names[i])
+}
+
 p_cov = bind_rows(
     fit_to_cov(hakki) |>
         mutate(
@@ -50,16 +59,18 @@ p_cov = bind_rows(
 ) |>
     ggplot() +
     geom_density(aes(cov, colour = model)) +
-    facet_grid(name_i ~ name_j)
+    facet_grid(name_i ~ name_j) +
+    standard_plot_theming() +
+    labs(
+        x = "(Co)variance",
+        y = "Posterior density",
+        colour = "Model"
+    ) +
+    geom_vline(xintercept = 0) +
+    coord_cartesian(xlim = c(NA, 5))
 
-fit_to_mean = function(fit) {
-    fit |>
-        spread_draws(vgrow[i, g]) |>
-        filter(g == 1) |>
-        mutate(name = parameter_names[i])
-}
 
-bind_rows(
+p_mean = bind_rows(
     fit_to_mean(hakki) |>
         mutate(
             model = "Hakki"
@@ -71,57 +82,107 @@ bind_rows(
 ) |>
     ggplot() +
     geom_density(aes(vgrow, colour = model)) +
-    facet_grid(. ~ name, scales = "free_x")
+    facet_grid(. ~ name, scales = "free_x") +
+    standard_plot_theming() +
+    labs(
+        x = "Parameter value",
+        y = "Posterior density",
+        colour = "Model"
+    )
+p_combined = p_mean / p_cov + plot_layout(
+    guides = "collect",
[…]

### 2023-11-02 ATACCC/trajectories.R

@@ -1,18 +1,9 @@
 library(ggplot2)
-source("utils.R")
-
-logVL <- function(t, vp) {
-  vl <- log(vp[1]) + log(vp[2] + vp[3]) - log_sum_exp(log(vp[3]) - vp[2] * t, log(vp[2]) + vp[3] * t)
-  return(vl)
-}
-
-log_sum_exp <- function(x, y) {
-  max_value <- pmax(x, y)
-  max_value + log1p(exp(pmin(x, y) - max_value))
-}
+source(here::here("ATACCC/trajectory_functions.R"))
+source(here::here("utils.R"))
 
 p_typical_trajectory = ggplot(NULL) +
-  stat_function(fun = logVL, args = list(exp(c(15, 1.6, 0.3))), xlim = c(-5, 15)) +
+  stat_function(fun = logVL, args = list(c(exp(15), 5, 1.4)), xlim = c(-5, 15), n = 1000) +
   coord_cartesian(ylim = c(0, NA)) +
   labs(
     x = "Time from peak viral load",
@@ -26,8 +17,3 @@ ggsave(
   units = "cm",
   dpi = 300
 )
-
-params = exp(c(15, 1.6, 0.3))
-logVL(c(-3.34, 0, 12), params)
--(log(params[1])+log(params[2]+params[3])-log(params[3]))/params[2]
-(log(params[3])-log(params[2]+params[3])-log(params[1]))/params[2]

### 2023-11-02 ATACCC/trajectory_functions.R

@@ -1,0 +1,9 @@
+log_sum_exp <- function(x, y) {
+  max_value <- pmax(x, y)
+  max_value + log1p(exp(pmin(x, y) - max_value))
+}
+
+logVL <- function(t, vp) {
+  vl <- log(vp[1]) + log(vp[2] + vp[3]) - log_sum_exp(log(vp[3]) - vp[2] * t, log(vp[2]) + vp[3] * t)
+  return(vl)
+}

### 2023-11-02 cis-perfect-testing/results.R

@@ -29,20 +29,6 @@ tbl_posteriors = readRDS(here::here("cisRuns-output/perfect_posteriors.rds")) |>
 truth = readRDS(here::here("cisRuns-output/input_curves.rds")) |>
   filter(source == "Combined")
 
-theme_survival_time_series = function() {
-    rlang::list2(
-        standard_plot_theming(),
-        scale_x_continuous(breaks = 0:100*14, minor_breaks = 0:100*2),
-        labs(
-            x = "t",
-            fill = "Hazard prior",
-            colour = "Hazard prior"
-        ),
-        theme(legend.position = "bottom"),
-        coord_cartesian(xlim = c(0, 100))
-    )
-}
-
 surv_nat_plot = tbl_posteriors |>
       ggplot() +
       ggdist::geom_lineribbon(

### 2023-11-02 SEIR.tex
Fix chapter numbering
First full draft of ATACCC chapter
Updates from Overleaf

@@
 
-Mechanistic models explicitly represent the process of how an infectious disease spreads.
+Mechanistic models explicitly represent the underlying processes driving the spread of infectious disease.
 Infectious diseases have inherent feedback loops: current infectious individuals generate future infectious individuals.
@@
 The first sections of this chapter give the background and methodology of the model I use.
-The literature on mechanistic models is vast, going back to at least \textcite{kermackContribution}.
+The literature on mechanistic models is vast, going back to \textcite{kermackContribution}.
 Here, I explain only the background necessary to understand the model used in this chapter.
@@
 
-This section gradually builds complexity into a mechanistic model
-The class of models considered here is \emph{compartmental} models.
-A compartmental model divides all members of the population into compartments, each representing a disease state.
+This section gradually builds complexity into a mechanistic model.
+The mechanistic models considered here are \emph{compartmental} models.
+A compartmental model divides all members of a population into compartments, each representing a disease state.
 Compartmental models are the most common type of mechanistic infectious disease model.
@@
 
-I will use a deterministic model, specifically an ODE model
-ODE models are mathematically described by a system of ODEs.
+I will use a deterministic model mathematically described by a system of Ordinary Differential Equations (ODE).
 The derivative is taken with respect to time, $t$.
-Therefore, in an ODE model, time is necessarily continuous although discretised when solving (see \cref{SEIR:sec:inference-implementation})
+Therefore, in an ODE model, time is necessarily continuous although discretised when solving (see \cref{SEIR:sec:inference-implementation}).
 The number or proportion of individuals in each compartment is also continuous, which is a reasonable approximation when the population is large.
-For determinstic models, equivalent fomulations exist for the proportion or number of individuals.
+Equivalent formulations exist for the proportion or number of individuals.
 
@@
 Further discussion of stochastic models is beyond the scope of this thesis.
-\todo{maybe add a reference to a review or textbook on stochastic models}
+\todo{maybe add a reference to a review or textbook on stochastic models} YES!
 
@@
 \begin{figure}[h]
+\centering
 \begin{tikzpicture}[
@@
 
-The simplest compartmental model is the \emph{susceptible-infected-recovered} (SIR) model.
+The simplest compartmental model is the \emph{susceptible-infected-recovered} (SIR) model (REF?).
 The name refers to the three compartments in the model, displayed in \cref{SEIR:fig:SIR}.
@@
 \begin{align}
-\frac{ds}{dt} &= -\beta si \\
+\frac{ds(t)}{dt} &= -\beta s(t)i \\
 \frac{di}{dt} &= \beta si - \gamma i \\
@@
 Two parameters are present in this model: $\beta$ and $\gamma$.
-$\beta$ is the transmission rate, or the number of individuals an infected individual infects per day.
+$\beta$ is the effective? transmission rate, or the number of individuals an infected individual infects per day.
 $\gamma$ is the recovery rate, or the proportion of infected individuals that recover per day; hence, $1/\gamma$ is the mean infectious period.
@@
 This term can be arrived at by considering a single infected individual.
-Assume this individual makes contact with other individuals in the population at a constant rate, $c$, in units of contacts per day.
+Assume this individual makes contact with any other individual in the population at a constant rate, $c$, in units of contacts per day.
 Further, assume the population is \emph{well-mixed}, meaning the individual is equally likely to come into contact with any other individual.
@@
 The basic SIR model implies that the distribution of the time spent in the infected compartment is exponential.
-This is an unrealistic assumption and leads to underestimating the reproduction number~\autocites{lloydRealistic}{wearingAppropriate}.
+This is an unrealistic assumption and leads to underestimation of the reproduction number~\autocites{lloydRealistic}{wearingAppropriate}.
 A standard extension to the SIR model is to allow the infectious period to follow a gamma distribution.
@@
 The infectious period is now distributed $\GamDist(n, n\gamma)$.
-A gamma distribution with the first paramter equal to 1 is an exponential distribution, connecting this model to the basic SIR model.
-$1/\gamma$ remains the mean infectious period because $n\gamma$ is used as the rate of transition between the compartment.
+A gamma distribution with the first parameter equal to 1 is an exponential distribution, connecting this model to the basic SIR model.
+$1/\gamma$ remains the mean infectious period by construction.
 
@@
 To allow for age, the population is divided into age groups.
-The state of each comparment is now a vector.
+The state of each compartment is now a vector.
 The proportion of the population that is susceptible is $\vec{s}$ where $s_i$ is the proportion of the $i$th age group that is susceptible.
@@
 
-The modelling approach here can be used to stratify the population by any charcteristic.
+The modelling approach here can be used to stratify the population by any characteristic.
 For example, using deprivation or ethnicity.
@@
 
-To incorporate the strata into the model, first I introduce some notation.
-Denote the set of strata, normally age groups, as $\set{A}$. $N_i$ notates the total population of stratum $i$ and hence $s_i N_i$ is the number of susceptible people in stratum $i$.
+To incorporate the stratification into the model, first I introduce some additional notation.
+Denote the set of strata, for example age groups, as $\set{A}$. $N_i$ notates the total population of stratum $a$ and hence $N_i s_i$ is the number of susceptible people in stratum $i$.
 
@@
 Informally, it is how likely individuals in stratum $i$ are to pass on the infection.
-Formally, it is the probability that an infectious individual in stratum $i$ will infect a susceptible individual upon contact, relative to an individual in the reference stratum.
+Formally, it is the relative risk of an infectious individual in stratum $i$ infecting a susceptible individual upon contact, relative to an individual in the reference stratum.
 These measures could be biological, \eg due to differences in immune systems, or behavioural, \eg if infants are less hygenic than adults.
@@
 
-I assume that each region is a well-mixed population and closed.
+I assume that in each region the population is well-mixed and closed.
 A more complex model would allow for interaction between regions.
-However, for a large epidemic, non-interacting populations is a good approximation and can even fit the data better~\autocite{birrellRealtimea}.
+However, for a large epidemic, non-interacting populations is a good approximation ~\autocite{birrellRealtimea}.
 
@@
 
-Each element of the vector $\vec{s}$ refers to the proportion of that region's population of the relevant age group.
+Each element of the vector $\vec{s}$ refers to the proportion of that region's population of the relevant age group that are susceptible.
+For example, $\vec{s}_1$ is...
 The other vectors are interpreted similarly.
@@
 During the pandemic, behaviour clearly changed over time.
-This was in response to government policy, also individual's perceived risk, and a variety of other factors.
-I follow \textcite{birrellRealtime}, using a combination of Google moblity data and a time-varying transmission intensity to include these changes.
+This was in response to government policy, an individual's perceived risk, and a variety of other factors.
+I follow \textcite{birrellRealtime}, using a combination of Google mobility data and a time-varying transmission intensity to include these changes.
 
-How between-age-group contacts occur have previously been inferred from a combination of contact surveys, time use data, and mobility data~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}.
+Between-age-group contacts rates have previously been inferred from a combination of contact surveys, time use data, and mobility data~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}.
 I use the outputs from this analysis as $\matr{M}$.
@@
 
-Some behavioural changes (\eg mask wearing) are not captured by the contact survey.
+Some behavioural changes (\eg mask wearing) and their consequent impacts on transmission are not captured by the contact survey.
 To allow for these changes, $\beta$ is allowed to vary over time.
-In particular, $\beta$ follows a random walk in log space on the weekly timescale:
+In particular, $\beta$ could? follows a random walk in log space on the weekly timescale:
 \begin{align}
@@
 A variety of different approaches can be taken to link a transmission model to observations.
-Here, I extend the system of ODEs to include comparments representing PCR positive individuals.
+Here, I extend the system of ODEs to include compartments representing PCR positive individuals.
 Whenever an individual is infected, a copy of them is placed in the pre-positive compartment, $p_0$.
@@
 
-A $l$-phase EC distribution is a convoluation of a $(l-2)$-phase gamma distribution and a two-phase acyclic, arbitrary phase-type distribution.
+A $l$-phase EC distribution is a convolution of a $(l-2)$-phase gamma distribution and a two-phase acyclic, arbitrary phase-type distribution.
 The time between entering the 1st state and the $l+1$th state, the absorbing state, is the duration.
-A schematic of an EC distribution and its parameters is shown in \cref{SEIR:fig:EC}
+A schematic of an EC distribution and its parameters is shown in \cref{SEIR:fig:EC}.
 

### 2023-11-06 incidence-prevalence.tex
Start of SEIR chapter
Fix chapter numbering
Edit backcalc chapter

@@
 \begin{document}
-\chapter{Statistical background} \label{inc-prev}
+\ifSubfilesClassLoaded{
+  \setcounter{chapter}{2}
+}
+
+\chapter{Statistical and epidemiological background} \label{inc-prev}
+
 %\todo[inline]{This chapter is pretty short. Not sure if this matters or it needs to merged with another chapter? For example, one background chapter split into epi/biology and stats.}
 
-This chapter explains incidence and prevalence, and the relationship between these quantities.
-Their exact definition can vary throughout the literature, here I follow the definitions of the seminal paper \textcite{freemanPrevalence}.
-As I will explain, the duration of infection is the quantity that relates them.
+This chapter defines incidence and prevalence, and the relationship between these quantities.
+In the course of explaining this relationship, it will become clear that the duration of PCR positivity, and its distribution in the population, is the vital parameter which will allow estimation of incidence from prevalence, a major aim of this thesis.
+I start by defining these terms in \cref{inc-prev:sec:definitions}.
+I then move on to explaining the three processes that are involved: the \emph{infection process}, the \emph{prevalence process} and the \emph{observation process}.
+The infection process (\cref{inc-prev:sec:infection-process}) models how infections arise.
+The concepts here are well-established within infectious disease epidemiology (see\todo{ref intro chapter}).
+The prevalence process describes the relationship between incidence and prevalence.
+Here, I adapt previous work to the SARS-CoV-2 context.
+In particular, the recoveries need explicit modelling which was not required for HIV/AIDS.
+The observation process describes the relationship between prevalence and the observed data.
+Unlike in HIV/AIDS backcalculation, the CIS has a sampling mechanism that needs to be modelled.
+This model is guided by its specific study design of the CIS.
+Therefore, the model is specific to this context, although much would be applicable to other, similarly designed prevalence surveys.
+Finally, \cref{inc-prev:sec:MCMC} will introduce Markov chain Monte Carlo (MCMC), an inference technique I will make extensive use of in this thesis, before \cref{inc-prev:sec:conclusion} concludes the chapter.
 
-Timescale underlies all these quantities.
-Throughout this thesis, time will generally\todo{revisit phrasing once finished - do I actually use continuous time anywhere (probably in SEIR model)} be considered discrete in units of days.
+\section{Definitions} \label{inc-prev:sec:definitions}
+
+Time will be considered discrete in this chapter.
+I use days, however, the results generalise in the obvious way to other time units.
 Most epidemiological data is only available at daily granularity or, when more granular data exists, is often unreliable.
@@
 
-\section{Definitions}
-
-\emph{Incidence} at time $t$, $Z_t$, is the number of people in a given population who become infected at time $t$.
-Incidence expressed as a proportion of the population, $Z_t/\Npop$ where $\Npop$ is the number of people in the population, is the \emph{incidence rate}.
-As discussed in \cref{E-intro}\todo{might want to point to a specific section}, the timing of infection events is not directly observable and must be inferred from other data.
+\emph{Incidence} on day $t$, $Z_t$, is the number of people in a given population who become infected on day $t$; when expressed as a proportion of the population it is the \emph{incidence proportion}~\autocite[89]{lashModern}.
+The incidence proportion is $Z_t/\Npop$, where $\Npop$ is the number of people in the population.
+As discussed in\todo{ref intro}, the timing of infection events is not directly observable and must be inferred from other data.
 Despite this issue, it remains the single most important quantity for informing the response to an epidemic or pandemic.
 
-\emph{Prevalence} of an infectious disease at time $t$, $P_t$, is the number of people in a given population currently infected at time $t$.
-Each of these individuals is known as a \emph{prevalent} individual.
-As I use discrete time, I assume that an individual's prevalent status cannot change over the period represented by $t$, normally one day.
-Therefore, prevalence is constant throughout $t$.
-The \emph{prevalence proportion} is the same quantity expressed as a proportion of the population.
-In this thesis, a prevalent individual is one that would return a positive PCR test if tested at time $t$, not other possible definitions of ``currently infected'' discussed in\todo{ref section}.
+\emph{Prevalence} of a disease on day $t$, is the number of people with the disease on that day~\autocite[90]{lashModern}.
+What it means to have a disease can vary by context (see\todo{ref section discussing what it means to have covid}).
+I define an individual to have SARS-CoV-2 on day $t$ if, had they been swabbed on day $t$, that swab would have returned a positive PCR test.
+I refer to these individuals as \emph{prevalent} individuals.
+I assume that an individual's prevalent status cannot change over the course of a day.
+Therefore, prevalence is constant over any single day $t$.
+I denote the number of prevalent individuals on day $t$ as $P_t$.
+Therefore, the prevalence is $P_t/\Npop$.
 
-\section{Backcalculation}
+Duration distributions can be expressed in several equivalent ways.
+The duration distribution I will be concerned with is the distribution of the number of days for which an individual is prevalent.
+Consider an individual $i$ infected on day $B_i$ that is prevalent during the interval $[B_i, E_i]$.
+I define their \emph{infection episode} as beginning on day $B_i$ and ending on day $E_i$.
+$D_i = E_i - B_i + 1$ is their \emph{duration of positivity}, the number of days for which they are prevalent.
+I model each $D_i$ as a discrete independent and identically distributed random variable; extending the methodology to account for covariates is left to further work.
+The duration is a positive integer, \ie $D \in \{1, 2, \dots\}$.
+The first way of expressing the distribution of $D_i$ is as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=0}^d f_{D}(i)$.
+The \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$~\autocite{yanDistribution}.
+The \emph{hazard} at time $t$ is the probability of recovering at time $t$ conditional on having survived to time $t$: $\lambda(t) = \prob(D = t \mid D \geq t) = f_{D}(d) / S(d)$.
+For discrete variables, the convention that the survival function is $\prob(D >d)$, rather than $\prob(D \geq d)$, is also commonly used, I adapt that latter convention.
+Throughout this thesis, the survival function will be the primary quantity of interest, and it will normally be parameterised in terms of the hazard; these two quantities are related by $S(t) = \prod_{i=1}^{t-1} (1 - \lambda(i))$.
 
-\emph{Backcalculation} is the procedure for estimating incidence based on a measurement of prevalence.
-\Textcite{brookmeyerMethod} developed it in the context of HIV (human immunodeficiency virus) and AIDS (acquired immune deficiency syndrome).
-HIV is transmitted sexually or through blood and causes the disease AIDS.
-It has killed 40 million people, peaking at 2 million annual deaths~\autocite{unaids2023}.
-In this context, backcalculation estimates the incidence of HIV infection from: data on the number of AIDS diagnoses, and the incubation distribution (the time between infection and AIDS diagnosis)~\autocites{brookmeyerBackcalculation}{brookmeyerMeasuring}.
-\todo{maybe move some of this to intro}
-AIDS is incurable, therefore, AIDS prevalence is equivalent to cumulative incidence.
+\section{Infection process} \label{inc-prev:sec:infection-process}
+The infection process is a \emph{point} or \emph{counting} process, where each infection is an event.
 
-Three processes are involved in backcalculation: the \emph{infection process}, the \emph{prevalence process} and the \emph{observation process}.
-The infection process determines incidence.
-The prevalence process describes the relationship between incidence and prevalence.
-The observation process describes the relationship between prevalence and the observed data.
+A point process is a random function of continuous time $K(\tau)$.
+$K(t)$ must take non-negative integer values and be non-decreasing in $\tau$.
+The interpretation of $K(t)$ is the number of events, here infections, that have occurred in the interval $(-\infty, t]$~\autocites[244]{yanDistribution}.
+In the context of a disease, the point process is discretised into incidence for integer $t$ as $Z_t = K(t) - K(t-1)$.
 
-The original formulations of backcalculation neglected the observation process.
-AIDS is almost always diagnosed and recorded in developed countries.
-That is, the observations of AIDS diagnoses are the same as the true incidence of AIDS.
-An exception is that the reporting delay may be important~\autocite{paganoHIV}.
-Reporting delay is not relevant to the studies in this thesis, and I ignore it in this and further discussion.
-Therefore, two processes needed to be considered in the context of HIV/AIDS: infection and prevalence.
+A counting process has \emph{independent increments} if the number of events in disjoint intervals are independent.
+That is, for any $t_1 < t_2 < t_3 < t_4$, $K(t_2) - K(t_1)$ and $K(t_4) - K(t_3)$ are independent.
 
-For SARS-CoV-2, measurements of prevalence are noisy because they are based on population sampling.
-The sampling is described as the observation process.
-The observation process noise is much larger than the noise of the prevalence process (see \cref{inc-prev:sec:observation-process}).
-Therefore, the latter is negligible and again two processes need to be considered: infection and observation.
+A commonly-used point process for infectious disease epidemiology is the time-inhomogenous Poisson process~\autocites{brookmeyerMethod}{paganoHIV}{rosenbergBackcalculation}{brookmeyerBackcalculation}.
+A Poisson process is a point process with independent increments, and the property that $K(s) - K(t) \dist \Poi\left(\int_s^t \lambda(r) dr \right)$ where $\lambda(r) \geq 0$ is the \emph{intensity} of the epidemic at time $r$~\autocites[244]{yanDistribution}.
+That is, a Poisson process where the intensity varies over time, and possibly with other covariates.
+Extensions to this model can incorporate covariates, such as space or age~\autocite[e.g.][]{diggleModeling}.
 
-\section{Infection process}
-
-The infection process is a point process, where each infection is an event.
-A simple and commonly-used model is the time-inhomogenous Poisson process~\autocites{brookmeyerMethod}{paganoHIV}{rosenbergBackcalculation}{brookmeyerBackcalculation}.
-That is, a Poisson process where the intensity varies over time, and possibly with other covariates.
 The target of inference can be either the process's intensity or the counts of the specific realisation that occurred.
 The realised counts are normally of greater interest because we care about the realised epidemic, not what may happen if there were future realisations of the same epidemic.
+Furthermore, there is only one realisation of the epidemic, and hence
 
@@
 Overdispersion means that the variance in the number of infections in a time interval is higher than its mean.
-In a Poisson process, counts in a time interval have a Poisson distribution.
-Therefore, the mean and variance are equal.
+In a Poisson process these counts have a Poisson distribution, hence, their mean and variance are equal.
 Feedback loops mean that the stochastically low or high numbers of infections, possibly due to overdispersion, affect the future number of infections.
-In a Poisson process, the intensity (mean number of events per unit time) is not influenced by the realisation of the process.
-Both of these features violate the assumptions of a Poisson process.
-Despite these issues, satisfactory inference on the incidence can still be produced, although the uncertainty in the estimates may be understated~\autocite{beckerDependent}.
+This violates the assumption of independent increments.
+Both of these features violate the properties of a Poisson process.
+Despite these issues, consistent inference on the incidence can still be produced using the Poisson process model, although the uncertainty in the estimates may be understated~\autocite{beckerDependent}.
 
@@
 A super-spreading event is an event where many more people than average are infected.
+That is, $Z_t >> \E(Z_t)$ for some $t$ during which a super-spreading event occurs.
 Coronaviruses, such as SARS-CoV-2, have particularly high overdispersion~\autocites{endoEstimating}{adamClustering}{mccloskeySARS}.
@@
 Conversely, for a Poisson process, the standard deviation is on the order of the square root of the mean.
-A particular implication of this is that the coefficient of variation (the standard deviation divideed by the mean) tends to zero for a Poisson process, but tends to a constant for a negative binomial process.
+A particular implication of this is that the coefficient of variation (the standard deviation divided by the mean) tends to zero for a Poisson process, but tends to a constant for a negative binomial process.
 The disadvantage of the negative binomial distribution is that it loses much of the mathematical convenience of a Poisson distribution.
@@
 These additional infections will propagate the epidemic further, increasing the intensity at future times.
+Point processes can be extended to \emph{branching processes} to account for this feedback~\autocite[246]{yanDistribution}, although these are beyond the scope of this thesis.
 
@@
 The relationship between incidence and prevalence can be derived by considering the probability that an individual is prevalent at any time following an infection.
-The period the individual is prevalent for is the \emph{infection episode}.
+I define the period the individual is prevalent for as their \emph{infection episode}.
 I index infection episodes with $i$.
 Denote the time that the infection episode begins as $B_i$ and the time it ends as $E_i \geq B_i$.
-The individual is prevalent from $B_i$ to $E_i$ inclusive.
+The individual is prevalent during the interval $[B_i, E_i]$ (\ie including $B_i$ and $E_i$).
 The duration of the infection episode, the number of days for which they are prevalent, is $D_i = E_i - B_i + 1$.
-In this thesis, it is assumed that the $D_i$ for each infection episode is an independent draw from a discrete distribution such that $\prob(D_i = d) = f_d(d)$ for $d = 1, \dots, d_\text{max}$ and $\prob(D_i = d) = 0$ otherwise.
+I assume that the $D_i$s are iid, with a discrete distribution defined by $\prob(D_i = d) = f_D(d)$ for $d = 1, \dots, d_\text{max}$, for some maximum episode length $d_\text{max}$, and $\prob(D_i = d) = 0$ otherwise.
 Denote by $F_D$ the cdf of $D_i$, that is $F_D(d) = \prob(D_i \leq d)$.
@@
 Denote by $R_{t,t'}$ be the number of infection episodes that begin at time $t$ and end at time $t'$.
-For each infection beginning at $t$, the probability that it ends at $t'$ is $\prob(E_i = t' \mid B_i = t) = \prob(D_i = t' - t + 1 \mid B_i = t) = f_d(t' - t + 1)$.
+For each infection beginning at $t$, the probability that it ends at $t'$ is $\prob(E_i = t' \mid B_i = t) = \prob(D_i = t' - t + 1 \mid B_i = t) = f_D(t' - t + 1)$.
 Since these durations are assumed independent, the number of episodes of each duration is multinomially distributed~\autocite{paganoHIV}.
@@
   \begin{bmatrix}
-    f_d(1) \\ f_d(2) \\ \vdots \\ f_d(d_\text{max})
+    f_D(1) \\ f_D(2) \\ \vdots \\ f_D(d_\text{max})
   \end{bmatrix}
@@
 \end{align}
-Now, denote by $P_{t_z,t_p}$ the number of episodes beginning at time $t_z$ that are prevalent at time $t_p$.
-This is $Z_t$ minus those that recover before $t_p$ (\ie at or up to $t_p - 1$).
+Now, denote by $P_{t_z,t_p}$ the number of infection episodes with $B_i = t_z$ and $E_i \geq t_p$.
+Therefore, the individuals in which these infection episodes occur are prevalent at time $t_p$.
+$P_{t_z,t_p}$ is $Z_{t_z}$ minus those that recover before $t_p$ (\ie at or up to $t_p - 1$).
 Hence:
@@
 \end{align}
-Finally, the prevalence at $t$ can be expressed as the total number of people prevalent at time $t$, regardless of when they were infected.
+Finally, the $P_t$ can be expressed as the total number of people prevalent at time $t$, regardless of when they were infected.
 The limits of the sum are limited by the zeroes in \cref{inc-prev:eq:Ptt-to-R}.
@@
 
-The infection process means that these random variables are dependent.
+In general, all the random variables in this section are dependent.
+This is because they are linked by the $Z_t$s, which are themselves dependent (as explained in \cref{inc-prev:sec:infection-process}).
 However, if we condition on the vector of incidence, $\vec{Z} = (Z_1, Z_2, \dots)^T$, then we will induce some independence.
-This is essentially isolating only the prevalence process.
+This reduces the stochasticity to only that attributable to the prevalence process.
 In particular, because the duration of individuals is independent, $R_{t_z,t_p}$ and $R_{t_z',t_p'}$ will be independent if $t_z \neq t_z'$; importantly, this holds even if $t_p = t_p'$.
@@
 
-Only $\E(P_t \mid \vec{Z})$ and bounds on $\V(P_t \mid \vec{Z})$ will be relevant (see \cref{inc-prev:sec:observation-process}).
+In \cref{inc-prev:sec:observation-process} I will show that only $\E(P_t \mid \vec{Z})$ and bounds on $\V(P_t \mid \vec{Z})$ will be relevant.
 These can be derived from their constituent parts.
@@
 \begin{align}
-  \sum_{i=0}^{t'} R_{t,t+i} \mid \vec{Z} &\sim \text{Binomial}(Z_t, F_d(t'+1)). \label{inc-prev:eq:binomialRt}
+  \sum_{i=0}^{t'} R_{t,t+i} \mid \vec{Z} &\sim \text{Binomial}(Z_t, F_D(t'+1)). \label{inc-prev:eq:binomialRt}
 \end{align}
-where $F_d(t) = \sum_{i=1}^t f_d(t) = \prob(D \leq t)$ is the CDF of $D$.
+where $F_D(t) = \sum_{i=1}^t f_D(t) = \prob(D \leq t)$ is the CDF of $D$.
 Therefore:
 \begin{align}
-  \E \left( \sum_{i=0}^{t'} R_{t,t+i} \mid \vec{Z} \right) &= Z_t F_d(t'+1) \label{inc-prev:eq:EsumRt} \\
-  \V \left( \sum_{i=0}^{t'} R_{t,t+i} \mid \vec{Z} \right) &= Z_t F_d(t'+1) (1 - F_d(t'+1)) \label{inc-prev:eq:VsumRt}
+  \E \left( \sum_{i=0}^{t'} R_{t,t+i} \mid \vec{Z} \right) &= Z_t F_D(t'+1) \label{inc-prev:eq:EsumRt} \\
+  \V \left( \sum_{i=0}^{t'} R_{t,t+i} \mid \vec{Z} \right) &= Z_t F_D(t'+1) (1 - F_D(t'+1)) \label{inc-prev:eq:VsumRt}
 \end{align}
@@
   &= \sum_{i=0}^{d_\text{max}-1} \left( Z_{t-i} - \sum_{j=0}^{i-1} \E( R_{t-i,t-i+j} \mid \vec{Z}) \right) \\
-  &= \sum_{i=0}^{d_\text{max}-1} \left( Z_{t-i} - Z_{t-i} F_d(i) \right) &\text{by \cref{inc-prev:eq:EsumRt}}\\
-  &= \sum_{i=0}^{d_\text{max}-1} Z_{t-i} (1 - F_d(i)) \\
-  &= \sum_{i=0}^{d_\text{max}-1} Z_{t-i} S(i+1)
+  &= \sum_{i=0}^{d_\text{max}-1} \left( Z_{t-i} - Z_{t-i} F_D(i) \right) &\text{by \cref{inc-prev:eq:EsumRt}}\\
+  &= \sum_{i=0}^{d_\text{max}-1} Z_{t-i} (1 - F_D(i)) \\
+  &= \sum_{i=0}^{d_\text{max}-1} Z_{t-i} S(i+1) \label{inc-prev:eq:EPt}
 \end{align}
-where $S(i+1) = 1-F_d(i) = \prob(D \geq i+1)$ is the \emph{survival function} of $D$ (see \cref{inc-prev:sec:duration}).
+where $S(i+1) = 1-F_D(i) = \prob(D \geq i+1)$ is the \emph{survival function} of $D$.
+The survival function will be further explored in \cref{inc-prev:sec:duration}.
 
@@
   &= \sum_{i=1}^{d_\text{max}-1} \V\left(\sum_{j=0}^{i-1} R_{t-i,t-i+j} \mid \vec{Z} \right) &\text{by~\cref{inc-prev:eq:Ptt-to-R}}\\
-  &= \sum_{i=1}^{d_\text{max}-1} Z_{t-i} F_d(i) (1 - F_d(i)) &\text{by~\cref{inc-prev:eq:VsumRt}} \\
-  &\leq \sum_{i=1}^{d_\text{max}-1} Z_{t-i} (1 - F_d(i)) &\text{as $F_d \leq 1$}\\
+  &= \sum_{i=1}^{d_\text{max}-1} Z_{t-i} F_D(i) (1 - F_D(i)) &\text{by~\cref{inc-prev:eq:VsumRt}} \\
+  &\leq \sum_{i=1}^{d_\text{max}-1} Z_{t-i} (1 - F_D(i)) &\text{as $F_D \leq 1$}\\
   &= \E(P_t \mid \vec{Z}). \label{inc-prev:eq:boundVPt}
@@
 
-The expectation derived here is a discrete convolution equation, specifically a discrete version of the Volterra equation of the first kind~\autocite{brookmeyerBackcalculation}.
+The expectation in \cref{inc-prev:eq:EPt} is a discrete convolution equation, specifically a discrete version of the Volterra equation of the first kind~\autocite{brookmeyerBackcalculation}.
+\todo{Go back to this reference and clarify why this matters}
 The variance is upper bounded by this expectation, a fact that will prove useful in \cref{inc-prev:sec:observation-process}.
-Clearly, knowledge of the survival function, $S$, will be important; I discuss this further in \cref{inc-prev:sec:duration}.
+\Cref{inc-prev:eq:EPt} shows that knowledge of the survival function, $S$, is crucial to the prevalence process; I will discuss $S$ further in \cref{inc-prev:sec:duration}.
 
@@
 % I use Bayesian inference, therefore the data contributes through the likelihood.
+The previous sections have explained the processes leading to the population prevalence.
+However, we do not directly observe the population prevalence but a sample of it.
+This section describes the observation process, that is, the process by which the prevalence surveys are conducted.
+The survey design considered here reflects the CIS, although is applicable to many randomly sampled prevalence survey with samples for all $t$.
 
-Prevalence surveys, as described in\todo{insert ref}, sample individuals at random from the population of interest and test if they are prevalence.
-Therefore, I model the number of positive tests on day $t$ as a random variable $X_t$ based on a sample of $n_t$ individuals.
-That is, $X_t \mid P_t, n_t, \Npop \dist \Bin(n_t, P_t/\Npop)$.
+On day $t$ (for $t = 1, \dots, T$) a prevalence survey such as the CIS samples $n_t$ individuals at random from the population of interest and tests if they are prevalent.
+Therefore, I model the number of positive tests, $X_t$, as $X_t \mid P_t, n_t, \Npop \dist \Bin(n_t, P_t/\Npop)$.
 The conditioning on $n_t$ and $\Npop$ is implicit in what follows.
 
-The $X_t$s are conditionally independent, given the population prevalences at each time, $\vec{P}_t$. 
+The $X_t$s are conditionally independent, given the population prevalences at each time, $\vec{P} = (P_1, \dots, P_t)^T$. 
 %For now, I assume that the sample is chosen uniformly at random from the population of interest.
@@
 %The population prevalence is a latent quantity.
-I base inference on the likelihood, $p(\vec{x} \mid \vec{Z})$.% = \int p(\vec{x} \mid \vec{P}) p(\vec{P} \mid \vec{Z}) d\vec{P} = \int \prod_t p(x_t \mid P_t) p(\vec{P} \mid \vec{Z}) d\vec{P}$, where $\vec{x} = (x_1, x_2, \dots)^T$.
+Prevalence surveys are used in this thesis to infer incidence.
+I base this inference on the likelihood, $p(\vec{x} \mid \vec{Z})$.% = \int p(\vec{x} \mid \vec{P}) p(\vec{P} \mid \vec{Z}) d\vec{P} = \int \prod_t p(x_t \mid P_t) p(\vec{P} \mid \vec{Z}) d\vec{P}$, where $\vec{x} = (x_1, x_2, \dots)^T$.
+Here $\vec{x} = (x_1, \dots, x_T)$ is the vector of observed number of positive tests in the prevalence survey on each day and $\vec{Z} = (Z_{-\dmax}, Z_{-\dmax+1} \dots, Z_T)$ is the vector of incidence on each day.
+The incidence back to day $-\dmax$ will affect $P_1$ and therefore need to be included in $\vec{Z}$.
 
@@
 
-The basis of the argument is that the first two moments of these distributions have negligible difference.
+The basis of the argument is that the first two moments of $\vec{x} \mid \vec{Z}$ and $\vec{X} \mid \vec{P} = \E(\vec{P} \mid \vec{Z})$ have negligible difference.
 The first moment follows directly from the tower property of expectations.
@@
 
-\section{Estimating duration} \label{inc-prev:sec:duration}
+\section{Markov chain Monte Carlo} \label{inc-prev:sec:MCMC}
+
+\section{Conclusion} \label{inc-prev:sec:conclusion}
 
 Robust, unbiased prevalence estimates already exist (see\todo{ref prevalence estimates}), and incidence estimates are the objective of this thesis.
-The quantity that relates these is the distribution of the duration of PCR positivity.
+This chapter has shown that the key quantity relating these quantities is the distribution of the duration of PCR positivity.
 However, no estimates of this quantity in the general population exist.
-Therefore, a large portion of this thesis concerns itself with estimating the duration of PCR positivity.
+Therefore, I now turn to estimating the duration of PCR positivity.
 
-The distribution of durations can be expressed in several equivalent ways.
-As before, $D_i$ is the duration of episode $i$.
-I model each $D_i$ as a discrete independent and identically distributed random variable; extending the methodology to account for covariates is left to further work.
-The duration is a positive integer, \ie $D \in \{1, 2, \dots\}$.
-The first way of expressing the distribution is as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=0}^d f_{D}(i)$.
-The \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$.
-The \emph{hazard} at time $t$ is the probability of recovering at time $t$ conditional on having survived to time $t$: $\lambda(t) = \prob(D = t \mid D \geq t) = f_{D}(d) / S(d)$.
-Throughout this thesis, the survival function will bne the primary quantity of interest, and it will normally be parameterised in terms of the hazard; these two quantities are related by $S(t) = \prod_{i=1}^{t-1} (1 - \lambda(i))$.
-
-Two approaches are commonly used to estimate durations: either modelling an underlying biomarker or modelling the duration directly~\autocite{sweetingEstimating}.
-For SARS-CoV-2, both approaches would be based on the results of PCR testing.
-Modelling a biomarker would mean modelling the viral load, measured as a Ct value (see \cref{E-ATACCC}).
-Modelling the duration would consider only the binary result at each test (positive or negative); in this thesis a \emph{survival analysis} framework is adopted (see \cref{E-perf-test,E-imperf-test}).
-% Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
-Other related approaches, such as a multi-state model~\autocite{jacksonMSM}, could also be considered (multi-state models can be viewed as a generalisation of survival analysis).
-Therefore, the biomarker approach considers more information per observation, however, a model for the biomarker is required which generally requires stronger assumptions.
-Modelling a biomarker uses more information per observation, but requires stronger assumptions.
-
-\todo[inline]{Maybe add a conclusion?}
+\ifSubfilesClassLoaded{
+  \listoftodos
+}{}
 

### 2023-11-06 introduction.tex
Edit backcalc chapter

\section{Backcalculation}

\emph{Backcalculation} is the procedure for estimating incidence based on a measurement of prevalence.
\Textcite{brookmeyerMethod} developed it in the context of HIV (human immunodeficiency virus) and AIDS (acquired immune deficiency syndrome).
HIV is transmitted sexually or through blood and causes the disease AIDS.
It has killed 40 million people, peaking at 2 million annual deaths~\autocite{unaids2023}.
In this context, backcalculation estimates the incidence of HIV infection from: data on the number of AIDS diagnoses, and the incubation distribution (the time between infection and AIDS diagnosis)~\autocites{brookmeyerBackcalculation}{brookmeyerMeasuring}.
\todo{maybe move some of this to intro}
AIDS is incurable, therefore, AIDS prevalence is equivalent to cumulative incidence.

The original formulations of backcalculation neglected the observation process.
AIDS is almost always diagnosed and recorded in developed countries.
That is, the observations of AIDS diagnoses are the same as the true incidence of AIDS.
An exception is that the reporting delay may be important~\autocite{paganoHIV}.
Reporting delay is not relevant to the studies in this thesis, and I ignore it in this and further discussion.
Therefore, two processes needed to be considered in the context of HIV/AIDS: infection and prevalence.

For SARS-CoV-2, measurements of prevalence are noisy because they are based on population sampling.
The sampling is described as the observation process.
The observation process noise is much larger than the noise of the prevalence process (see \cref{inc-prev:sec:observation-process}).
Therefore, the latter is negligible and again two processes need to be considered: infection and observation.

\section{Durations} \label{inc-prev:sec:duration}

Two approaches are commonly used to estimate durations: either modelling an underlying biomarker or modelling the duration directly~\autocite{sweetingEstimating}.
For SARS-CoV-2, both approaches would be based on the results of PCR testing.
Modelling a biomarker would mean modelling the viral load, measured as a Ct value (see \cref{E-ATACCC}).
Modelling the duration would consider only the binary result at each test (positive or negative); in this thesis a \emph{survival analysis} framework is adopted (see \cref{E-perf-test,E-imperf-test}).
% Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
Other related approaches, such as a multi-state model~\autocite{jacksonMSM}, could also be considered (multi-state models can be viewed as a generalisation of survival analysis).
Therefore, the biomarker approach considers more information per observation, however, a model for the biomarker is required which generally requires stronger assumptions.
Modelling a biomarker uses more information per observation, but requires stronger assumptions.

### 2023-11-06 ATACCC.tex
Fix chapter numbering
Edit ATACCC chapter

@@
 \ifSubfilesClassLoaded{
-  \setcounter{chapter}{4}
+  \setcounter{chapter}{3}
 }
@@
 
-\Cref{E-inc-prev:sec:duration} showed that the duration of PCR positivity is a key parameter for estimating the incidence of SARS-CoV-2 infections.
+\Cref{E-inc-prev} showed that the duration of PCR positivity is a key parameter for estimating the incidence of SARS-CoV-2 infections.
 This chapter estimates the duration of PCR positivity using the viral load data from the ATACCC study (described in\todo{ref section describing ATACCC}).
@@
 I convert the model and parameter estimates of the biomarker to a duration of PCR positivity in \cref{ATACCC:sec:duration}.
-This can then be used for backcalculation and as a prior for the CIS analysis.
+I will refine this estimate in future chapters before its use in estimating transmission.
 
@@
 This chapter uses the model of \textcite{hakkiOnset}.
-This model is approximately piecewise log-linear, however, a slight adjustment is made which means that the model is continuos (see \cref{ATACCC:sec:hakki} for details).
+This model is approximately piecewise log-linear, however, a slight adjustment is made which means that the model is continuous (see \cref{ATACCC:sec:hakki} for details).
 
@@
 The parameter's distribution across the population is assumed to be some known distribution, normally a normal or log-normal distribution~\autocite[chapter 24]{lashModern}.
-Specifically, if $\phi_j$ is the value of some parameter for individual $j$, $\mu_\phi$ is the population mean of $\phi$, and $\sigma_\phi$ is the population standard deviation, then we model $\phi_j$ as: $\phi_j \sim \N(\mu_\phi, \sigma_\phi^2)$.
+Specifically, if $\phi_j$ is the value of some parameter for individual $j$, $\mu_\phi$ is the population mean of $\phi$, and $\sigma_\phi$ is the population standard deviation, then we model $\phi_j$ as: $\phi_j \sim \N(\mu_\phi, \sigma_\phi^2)$ or $\phi_j \sim \LN(\mu_\phi, \sigma_\phi^2)$.
 $\mu_\phi$ and $\sigma_\phi$ are parameters; in a Bayesian setting, a prior is placed on them in the normal way.
@@
 A continuous viral load trajectory should give better performance when using gradient-based inference methods such as Hamiltonian Monte Carlo (HMC)\todo{cite paper or ref section}.
-The model for the viral load at time $t$, $\hat{v}_i(t)$, in individual $i$ is:
+The model for the viral load at time $t$, $\hat{v}_i(t)$, in individual $i$ is shown in \cref{ATACCC:fig:viral-load-model} and given by:
 \begin{align}
-\hat{v}_i(\tau) = v_{\max,i} \frac{a_i+b_i}{b_ie^{-a_i(\tau-\tau_{\max,i})} + a_ie^{b_i(\tau-\tau_{max,i})}} \label{ATACCC:eq:viral-load}
+\hat{v}_i(\tau) = v_{\max,i} \frac{a_i+b_i}{b_ie^{-a_i(\tau-\tau_{\max,i})} + a_ie^{b_i(\tau-\tau_{max,i})}}. \label{ATACCC:eq:viral-load}
 \end{align}
@@
 As time moves away from the peak (\ie becomes large in magnitude), the viral load trajectory tends towards exponential growth (before the peak) or decline (after the peak) in viral load at rates $a_i$ and $b_i$ respectively.
-Before the peak, this is because $b(\tau-\tau_{\max})$ is small and hence $a\exp(b(\tau-\tau_{\max}))$ is negligible.
-After the peak, this is because $b\exp(-a(\tau-\tau_{\max}))$ is negligible.
+Before the peak, this is because $b(\tau-\tau_{\max,i})$ is small and hence $a\exp(b(\tau-\tau_{\max,i}))$ is negligible.
+After the peak, this is because $b\exp(-a(\tau-\tau_{\max,i}))$ is negligible.
 This happens quickly for realistic values of the parameters.
@@
 $\vec{\phi}_i = (v_{\max,i}, a_i, b_i)^T$ are the random effects for individual $i$.
-$\tau_i$ is inferred independently for each individual because the time from first test to peak is not a biological concept shared between individuals.
+$\tau_{\max,i}$ is inferred independently for each individual because the time from first test to peak is not a biological concept shared between individuals.
 \begin{figure}
@@
 \end{align}
-where $f_N(y \mid \mu, \sigma^2)$ and $F_N(y \mid \mu, \sigma^2)$ are the pdf and cdf at $x$ of a Normal distribution with mean $\mu$ and variance $\sigma^2$ respectively.
+where $f_N(y \mid \mu, \sigma^2)$ and $F_N(y \mid \mu, \sigma^2)$ are the pdf and cdf at $y$ of a Normal distribution with mean $\mu$ and variance $\sigma^2$ respectively.
 The first term on each line is the likelihood contribution from test failures, either a false positive (top line) of false negative (bottom line).
@@
 Individuals are split into groups based on their vaccination status.
-The first group is unvaccinated and infected with a variety of Pre-Alpha, Alpha, and Delta variants.
+The first group is unvaccinated and infected with a variety of pre-Alpha, Alpha, and Delta variants.
 The second group is vaccinated and all were infected by the Delta variant.
-My focus is on the first group since they best correspond to the infections in the time period used throughout this thesis (Autumn 2020).
+My focus is on the first group since they best correspond to the infections in the time period used throughout this thesis, Autumn 2020.
 
-The parameters $\vec{\phi}_i$ are constrained to the positive reals.
-They are modelled as log-normal random effects with the hierarchical prior $\phi_i \dist \LN(\vec{\mu}_{\phi,g(i)}, \matr{\Sigma}_\phi)$.
-$g(i)$ is the group of individual $i$.
+Biological plausibility requires constraining $\vec{\phi}_i$ to the positive reals.
+Each $\phi_i$ is modelled as a log-normal random effects by using the hierarchical prior $\phi_i \dist \LN(\vec{\mu}_{\phi,g(i)}, \matr{\Sigma}_\phi)$ where $g(i)$ is the group of individual $i$.
 $\vec{\mu}_{\phi,g} = (\mu_{v_{\max},g}, \mu_{a,g}, \mu_{b,g})^T$ is a vector of the population means of the log-parameters.
-The random effects have a log-normal, rather than normal, distribution.
-Therefore, $\exp(\vec{\mu}_{\phi,g})$ is the population median of the parameter value but not the mean.
+As the random effects have a log-normal, rather than normal, distribution $\exp(\vec{\mu}_{\phi,g})$ is the population median of the parameter value but not the mean.
 The covariance matrix $\matr{\Sigma}_\phi$ controls the variability of the population and allows for the parameters to be correlated.
@@
 
-Based on my previous analysis (see \cref{ATACCC:sec:original-analysis}) and the fit to the data, I hypothesised that the model for test failures was overcomplicated.
-False positives do not appear to be present in the dataset.
+Based on my previous analysis (see \cref{ATACCC:sec:original-analysis}) and the fit to the data, I hypothesised that the cause of this was overcomplicating the model for test failures.
+Specifically, false positives do not appear to be present in the dataset.
 Therefore, the model for test failures can be simplified to simply have a constant probability of a false negative.
@@
     (1 - \rho) f_N(y_{i,t} \mid \log \hat{v}_i(t), \sigma_v^2) & \text{if $y_{i,t} > 0$}\\
-    \rho + (1 - \rho) F_N(\vlod \mid \log \hat{v}_i(t), \sigma_v^2) & \text{if $y_{i,t} = 0$}
+    \rho + (1 - \rho) F_N(\vlod \mid \log \hat{v}_i(t), \sigma_v^2) & \text{if $y_{i,t} = 0$.}
 \end{cases}
@@
 The ultimate aim of this chapter is to produce an estimate of the distribution for the duration of PCR positivity.
-So far, only the viral load trajectory has been considered.
-The time that individual $i$ first and last positive are the two values of $\tau$ such that $\log \hat{v}_i(\tau) = 0$.
+So far, only the viral load trajectories have been considered.
+The time that individual $i$ are first and last positive are the two values of $\tau$ such that $\log \hat{v}_i(\tau) = 0$.
 Denote these times as $\tau_{0,i}$ and $\tau_{1,i}$.
@@
 To calculate the discrete version of this distribution, $F_D$, a Monte Carlo integration is used with the following relation.
+\begin{align}
+  F_D(d)
+  &= \prob(D_i \leq d + 0.5) \\
+  &= \E_{D_i}(\I(D_i \leq d + 0.5)) 
+\end{align}
+where $\I$ is the indicator function.
+Here, $D_i$ is drawn conditional on the population-level parameters $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$.
+Equivalent forms of $F_D$, such as the survival and hazard functions, can be calculated using the relations in\todo{ref section defining survival functions}.
+I found 100,000 draws produced stable estimates of the tail of $F_D$.
+
 \section{Results} \label{ATACCC:sec:results-discussion}
@@
 The original \textcite{hakkiOnset} model has convergence issues (Rhat > 1.07 and ESS < 100) for the parameters of the test failure distribution.
-No such issues occur in the modified model introduced in \cref{ATACCC:sec:observation-modification}.
-Both model have convergence issues (Rhat > 1.4 and ESS < 20) for some individual-level parameters, although with the simplified model this is only individual 55.
+These issues do not occur in the modified model introduced in \cref{ATACCC:sec:observation-modification}, although both model have convergence issues (Rhat > 1.4 and ESS < 20) for some individual-level parameters; with the simplified model this is only individual 55.
 
@@
 This means that two viral load trajectories fit the data well (shown in \cref{ATACCC:fig:individual-55}).
-This first, found by two chains, increases very quickly, peaks 4.0 days after the first test, then declines slowly
+The first, found by two chains, increases very quickly, peaks 4.0 days after the first test, then declines slowly.
 The second, found by the other six chains, is more symmetrical, peaking 7.2 days after the first test.
-Because each individual's trajectory is not of interest for this analysis, and the population-level parameters agree between the two sets of chains, I consider this lack of convergence acceptable.
+Because the individual trajectories is not of interest for this analysis, and the population-level parameters agree between the two sets of chains, I consider this lack of convergence acceptable.
 \begin{figure}
@@
 Negative PCR tests are plotted at the limit of detection, \ie a log viral load of 0.
-Each plot is labelled arbitrarily with the individual's position in the dataset.
+Each plot is labelled with the individual's position in the dataset, an arbitrary value.
 \label{ATACCC:fig:goodness-of-fits}}
@@
 The two models give very similar estimates for the population-level parameters (see \cref{ATACCC:fig:compare-hakki-modified}).
-These are the main parameters of interest (as explained in \cref{ATACCC:fig:duration}), suggesting that the results are robust to the choice of model.
+These are the main parameters of interest (as explained in \cref{ATACCC:sec:duration}), suggesting that the results are robust to the choice of model.
 As the estimates are similar and the modified model converges better, I use the modified model for the remainder of this chapter.
@@
   \centering \includegraphics{ATACCC/compare_hakki_modified}
-  \caption[Comparison of population-level parameters between models.]{Kernel-density smoothed posterior densities for the population mean (top), $\vec{\mu}_{\phi,1}$ and covariance matrix (bottom), $\matr{\Sigma}_\phi$, for both the \textcite{hakkiOnset} model (described in \cref{ATACCC:sec:hakki}) and the model with a modified observation model to improve convergence (described in \cref{ATACCC:sec:observation-modification}).
-Mean only shown for the unvaccinated group, the primary group of interest, the covariance matrix is shared.
-These parameters, which are the primary ones of interest (see \cref{ATACCC:sec:duration}) have very similar estimates between the models.
+  \caption[Comparison of population-level parameters between models.]{Kernel-density smoothed posterior densities for the population mean (top), $\vec{\mu}_{\phi,1}$, and covariance matrix (bottom), $\matr{\Sigma}_\phi$, for both the \textcite{hakkiOnset} model (described in \cref{ATACCC:sec:hakki}) and the model with a modified observation model to improve convergence (described in \cref{ATACCC:sec:observation-modification}).
+Mean only shown for the unvaccinated group, the primary group of interest.
+The groups share the covariance matrix.
+These parameters, which are the primary ones of interest (see \cref{ATACCC:sec:duration}), have very similar estimates between the models.
 \label{ATACCC:fig:compare-hakki-modified}}
@@
   \centering \includegraphics{ATACCC/mean_trajectories}
-  \caption[Mean viral load trajectories]{Trajectory (posterior median and 95\% CrI) using the median parameter values, $\exp \vec{\mu}_{\phi,g}$, excluding the within-population variation.
+  \caption[Median viral load trajectories]{Trajectory (posterior median and 95\% CrI) using the median parameter values, $\exp \vec{\mu}_{\phi,g}$, excluding the within-population variation.
 These assume $\tau_{\max,i}=0$, however modifying this simply translates the trajectories left or right.
-\label{ATACCC:fig:mean-trajectories}}
+\label{ATACCC:fig:median-trajectories}}
 \end{figure}
@@
 
-\begin{align}
-  F_D(d)
-  &= \prob(D_i \leq d + 0.5) \\
-  &= \E_{D_i}(\I(D_i \leq d + 0.5)) 
-\end{align}
-where $\I$ is the indicator function.
-Here, $D_i$ is drawn conditional on the population-level parameters $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$.
-Many Monte Carlo draws are needed to ensure stability of the estimates of extreme values of $F_D$.
-I found 100,000 draws produced stable estimates.
-
 The $F_D$ estimates for the two groups are in \cref{ATACCC:fig:duration}.
-In \cref{ATACCC:sec:estimates}, the two models gave very similar results.
+In \cref{ATACCC:sec:results-discussion}, the two models gave very similar results.
 Therefore, I show results only from the modified model (which had better convergence).
@@
   \centering \includegraphics{ATACCC/duration}
-  \caption[Duration of PCR positivity.]{Estimated distribution (survival function, defined in \cref{E-inc-prev:sec:duration}) of duration of PCR positivity for the two populations.
+  \caption[Duration of PCR positivity.]{Estimated distribution (survival function) of duration of PCR positivity for the two populations.
 The vaccinated population have shorter durations of PCR positivity.
@@
 All these factors could change the viral dynamics~\autocite{russellWithinhost} and hence the duration.
-Therefore, the relationships shown here cannot be considered causal, or even generalisable associations to contexts where these other confounders are not present.
+Therefore, the relationships shown here cannot be considered causal, or even generalisable to contexts where these other confounders are not present.
 
@@
 This is exactly the situation in the UK in early 2021.
-The next few chapters will incorporate CIS data to improve the estimates of these tail probabilities.
+The next two chapters will incorporate CIS data to improve the estimates of these tail probabilities.
 

### 2023-11-07 incidence-prevalence.tex
Editing

@@
 In the course of explaining this relationship, it will become clear that the duration of PCR positivity, and its distribution in the population, is the vital parameter which will allow estimation of incidence from prevalence, a major aim of this thesis.
+
 I start by defining these terms in \cref{inc-prev:sec:definitions}.
@@
 The concepts here are well-established within infectious disease epidemiology (see\todo{ref intro chapter}).
-The prevalence process describes the relationship between incidence and prevalence.
+The prevalence process (\cref{inc-prev:sec:prevalence-process}) describes the relationship between incidence and prevalence.
 Here, I adapt previous work to the SARS-CoV-2 context.
@@
 Unlike in HIV/AIDS backcalculation, the CIS has a sampling mechanism that needs to be modelled.
-This model is guided by its specific study design of the CIS.
+This model is guided by the specific study design of the CIS.
 Therefore, the model is specific to this context, although much would be applicable to other, similarly designed prevalence surveys.
@@
 I use days, however, the results generalise in the obvious way to other time units.
+I make this choice for pragmatic reasons.
 Most epidemiological data is only available at daily granularity or, when more granular data exists, is often unreliable.
@@
 
-\emph{Incidence} on day $t$, $Z_t$, is the number of people in a given population who become infected on day $t$; when expressed as a proportion of the population it is the \emph{incidence proportion}~\autocite[89]{lashModern}.
-The incidence proportion is $Z_t/\Npop$, where $\Npop$ is the number of people in the population.
-As discussed in\todo{ref intro}, the timing of infection events is not directly observable and must be inferred from other data.
+\emph{Incidence} on day $t$, $Z_t$, is the number of people in a given population who become infected on day $t$; when expressed as a proportion of the population it is the \emph{incidence proportion}~\autocite[89]{lashModern}, $Z_t/\Npop$, where $\Npop$ is the number of people in the population.
+As discussed in\todo{ref intro}, the timing of infection events is very rarely directly observable and must be inferred from other data.
 Despite this issue, it remains the single most important quantity for informing the response to an epidemic or pandemic.
@@
 
-Duration distributions can be expressed in several equivalent ways.
+Duration, or time-to-event, distributions can be expressed in several equivalent ways.
 The duration distribution I will be concerned with is the distribution of the number of days for which an individual is prevalent.
 Consider an individual $i$ infected on day $B_i$ that is prevalent during the interval $[B_i, E_i]$.
-I define their \emph{infection episode} as beginning on day $B_i$ and ending on day $E_i$.
+I define this period as an \emph{infection episode}, beginning on day $B_i$ and ending on day $E_i$.
 $D_i = E_i - B_i + 1$ is their \emph{duration of positivity}, the number of days for which they are prevalent.
-I model each $D_i$ as a discrete independent and identically distributed random variable; extending the methodology to account for covariates is left to further work.
-The duration is a positive integer, \ie $D \in \{1, 2, \dots\}$.
+I model each $D_i$ as a discrete independent and identically distributed (iid) random variable; extending the methodology to account for covariates is left to further work.
+The duration is a positive integer, \ie $D_i \in \{1, 2, \dots\}$.
 The first way of expressing the distribution of $D_i$ is as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=0}^d f_{D}(i)$.
@@
 
-A point process is a random function of continuous time $K(\tau)$.
-$K(t)$ must take non-negative integer values and be non-decreasing in $\tau$.
+A point process is a random function of continuous time $K(t)$.
+$K(t)$ must take non-negative integer values and be non-decreasing in $t$.
 The interpretation of $K(t)$ is the number of events, here infections, that have occurred in the interval $(-\infty, t]$~\autocites[244]{yanDistribution}.
@@
 
-A counting process has \emph{independent increments} if the number of events in disjoint intervals are independent.
+A counting process has \emph{independent increments} if the number of events in disjoint intervals are independent~\autocite[244]{yanDistribution}.
 That is, for any $t_1 < t_2 < t_3 < t_4$, $K(t_2) - K(t_1)$ and $K(t_4) - K(t_3)$ are independent.
@@
 A commonly-used point process for infectious disease epidemiology is the time-inhomogenous Poisson process~\autocites{brookmeyerMethod}{paganoHIV}{rosenbergBackcalculation}{brookmeyerBackcalculation}.
-A Poisson process is a point process with independent increments, and the property that $K(s) - K(t) \dist \Poi\left(\int_s^t \lambda(r) dr \right)$ where $\lambda(r) \geq 0$ is the \emph{intensity} of the epidemic at time $r$~\autocites[244]{yanDistribution}.
-That is, a Poisson process where the intensity varies over time, and possibly with other covariates.
-Extensions to this model can incorporate covariates, such as space or age~\autocite[e.g.][]{diggleModeling}.
+A Poisson process is a point process with independent increments, and the property that $K(s) - K(t) \dist \Poi\left(\int^s_t \lambda(r) dr \right)$ where $\lambda(r) \geq 0$ is the \emph{intensity} of the epidemic at time $r$~\autocites[244]{yanDistribution}.
+A time-homogenous Poisson process is the special case where $\lambda(r)$ is constant.
+The independent increment property says that, conditional on $\lambda(r)$, the number of infections in disjoint intervals are independent.
+This implies $Z_t$ and $Z_{t'}$ are independent for $t \neq t'$.
+Extensions to this model can incorporate covariates, such as geography or age~\autocite[e.g.][]{diggleModeling}.
 
 The target of inference can be either the process's intensity or the counts of the specific realisation that occurred.
-The realised counts are normally of greater interest because we care about the realised epidemic, not what may happen if there were future realisations of the same epidemic.
-Furthermore, there is only one realisation of the epidemic, and hence
+The realised counts are normally of greater interest because we care about the realised epidemic, not what may happen if there were future realisations of the same epidemic~\autocites{beckerDependent}{brookmeyerMethod}.
 
-Two features of a Poisson process can be violated for epidemics.
-Epidemics commonly exhibit overdispersion and have feedback loops.
-Overdispersion means that the variance in the number of infections in a time interval is higher than its mean.
-In a Poisson process these counts have a Poisson distribution, hence, their mean and variance are equal.
+Two features of a Poisson process are violated for epidemics.
+Epidemics commonly exhibit overdispersion and have feedback loops~\autocite{beckerDependent}.
+Overdispersion means that $\V(Z_t) > \E(Z_t)$ for any day $t$.
+In a Poisson process these counts have a Poisson distribution, hence, $\V(Z_t) = \E(Z_t)$.
 Feedback loops mean that the stochastically low or high numbers of infections, possibly due to overdispersion, affect the future number of infections.
 This violates the assumption of independent increments.
+In particular, $\lambda(t+\delta t)$ is dependent on $K(t)$ for $\delta t > 0$.
 Both of these features violate the properties of a Poisson process.
@@
 For comparison, the average individual infected around three others in the early stages of the pandemic~\autocite{pellisChallenges}. 
-Overdispersion could be incorporated by modelling the counts in a given time interval as a negative binomial, rather than Poisson, distribution.
-The mean of these distributions is equal, but the negative binomial introduces an additional overdispersion parameter, commonly notated $k$.
-The negative binomial tends to Poisson as $k\to\infty$.
+Overdispersion can be incorporated by replacing the assumption that $K(s) - K(t)$ is Poisson distributed with assuming it is negative binomially distributed.
+This would not affect $\E(K(s) - K(t))$, but the negative binomial introduces an additional overdispersion parameter, commonly notated $k$.
+%The negative binomial tends to Poisson as $k\to\infty$.
 Generally, $k<1$ is considered as significant levels of overdispersion.
-Estimates for SARS-CoV-2 indicate $k \approx 0.1$~\autocite{endoEstimating}, increased by some interventions including lockdowns~\autocites{quiltyReconstructing}{quiltyUnderstanding}.
-For a negative binomial process, the count's standard deviation scales on the order of the mean.
-Conversely, for a Poisson process, the standard deviation is on the order of the square root of the mean.
+Estimates for SARS-CoV-2 indicate $k \approx 0.1$~\autocite{endoEstimating}, and may be decreased by some interventions including lockdowns~\autocites{quiltyReconstructing}{quiltyUnderstanding}.
+If $K(s) - K(t)$ has a negative binomial distribution, then its standard deviation scales on the order of the mean.
+If it is Poisson distributed, then the standard deviation is on the order of the square root of the mean.
 A particular implication of this is that the coefficient of variation (the standard deviation divided by the mean) tends to zero for a Poisson process, but tends to a constant for a negative binomial process.
@@
 
-An alternative to a Poisson process is \emph{mechanistic models}.
+An alternative to explicit use of counting processes is using \emph{mechanistic models}.
 Mechanistic models explicitly describe the population and transmission between individuals.
@@
 
-\section{Prevalence process}
+\section{Prevalence process} \label{inc-prev:sec:prevalence-process}
 
 The relationship between incidence and prevalence can be derived by considering the probability that an individual is prevalent at any time following an infection.
-I define the period the individual is prevalent for as their \emph{infection episode}.
-I index infection episodes with $i$.
-Denote the time that the infection episode begins as $B_i$ and the time it ends as $E_i \geq B_i$.
-The individual is prevalent during the interval $[B_i, E_i]$ (\ie including $B_i$ and $E_i$).
-The duration of the infection episode, the number of days for which they are prevalent, is $D_i = E_i - B_i + 1$.
-I assume that the $D_i$s are iid, with a discrete distribution defined by $\prob(D_i = d) = f_D(d)$ for $d = 1, \dots, d_\text{max}$, for some maximum episode length $d_\text{max}$, and $\prob(D_i = d) = 0$ otherwise.
-Denote by $F_D$ the cdf of $D_i$, that is $F_D(d) = \prob(D_i \leq d)$.
+%I define the period the individual is prevalent for as their \emph{infection episode}.
+%I index infection episodes with $i$.
+%Denote the time that the infection episode begins as $B_i$ and the time it ends as $E_i \geq B_i$.
+%The individual is prevalent during the interval $[B_i, E_i]$ (\ie including $B_i$ and $E_i$).
+%The duration of the infection episode, the number of days for which they are prevalent, is $D_i = E_i - B_i + 1$.
+%I assume that the $D_i$s are iid, with a discrete distribution defined by $\prob(D_i = d) = f_D(d)$ for $d = 1, \dots, d_\text{max}$, for some maximum episode length $d_\text{max}$, and $\prob(D_i = d) = 0$ otherwise.
+%Denote by $F_D$ the cdf of $D_i$, that is $F_D(d) = \prob(D_i \leq d)$.
+From this, we can derive the relationship between incidence and prevalence.
 
-From this, we can derive the relationship between incidence and prevalence.
-Denote by $R_{t,t'}$ be the number of infection episodes that begin at time $t$ and end at time $t'$.
+Denote by $R_{t,t'}$ be the number of infection episodes with $B_i=t$ and $E_i=t'$.
 For each infection beginning at $t$, the probability that it ends at $t'$ is $\prob(E_i = t' \mid B_i = t) = \prob(D_i = t' - t + 1 \mid B_i = t) = f_D(t' - t + 1)$.
@@
 \end{align}
+by the properties of the binomial distribution (see \cref{E-distributions}).
 
@@
 where $S(i+1) = 1-F_D(i) = \prob(D \geq i+1)$ is the \emph{survival function} of $D$.
-The survival function will be further explored in \cref{inc-prev:sec:duration}.
 
@@
 
-The expectation in \cref{inc-prev:eq:EPt} is a discrete convolution equation, specifically a discrete version of the Volterra equation of the first kind~\autocite{brookmeyerBackcalculation}.
-\todo{Go back to this reference and clarify why this matters}
-The variance is upper bounded by this expectation, a fact that will prove useful in \cref{inc-prev:sec:observation-process}.
-\Cref{inc-prev:eq:EPt} shows that knowledge of the survival function, $S$, is crucial to the prevalence process; I will discuss $S$ further in \cref{inc-prev:sec:duration}.
+The expectation in \cref{inc-prev:eq:EPt} is a discrete convolution equation.
+The equation is of the same as the equation used for HIV/AIDS backcalculation, which itself is often found in a variety of applications~\autocite[and references therein]{brookmeyerBackcalculation}.
+The system can be written in the form $\E(\vec{P} \mid \vec{Z}) = \matr{L} \vec{Z}$ where $\matr{L}$ is a lower-triangular matrix.
+Numerical methods for solving systems of this form, that is solving for $\vec{Z}$ given $\vec{P}$, are well-established~\autocite[e.g.:][section 8.2]{highamAccuracy}.
+However, these algorithms can be unstable, meaning that numerical errors are large.
+This is a particular issue if $S(1)$ is small.
+Since, by definition, $S(1)=1$ here, this is not an issue in this context.
+%However, if instead of estimating the incidence of PCR positives, the aim was to estimate the incidence of infections, then $S(1)$ would be small and this would be an issue.
+%While this quantity is more epidemiologically relevant, the time from infection to PCR positivity is short and, for now, I will neglect the difference.
+
+The variance in \cref{inc-prev:eq:boundVPt} is upper bounded by \cref{inc-prev:eq:EPt}, a fact that will prove useful in \cref{inc-prev:sec:observation-process}.
+\Cref{inc-prev:eq:EPt} shows that knowledge of the survival function, $S$, is crucial to the prevalence process.
 
@@
 However, we do not directly observe the population prevalence but a sample of it.
-This section describes the observation process, that is, the process by which the prevalence surveys are conducted.
+The sampling is described mathematically by the observation process.
 The survey design considered here reflects the CIS, although is applicable to many randomly sampled prevalence survey with samples for all $t$.
@@
 The conditioning on $n_t$ and $\Npop$ is implicit in what follows.
+These quantities are assumed to be fixed by the study design and population choice, and independent of $P_t$ and $Z_t$.
 
@@
 
+MCMC is a class of algorithms for sampling from a distribution where the normalising constant for the distribution is unknown.
+I will use it to sample a set of parameters, say $\psi$ (which may be a scalar or vector), from a posterior distribution $p(\psi \mid y)$ where $y$ is the data.
+MCMC algorithms produce a sample $\psi^1, \psi^2, \dots, \psi^M$ from $p(\psi \mid \vec{y})$ by constructing a Markov chain with $p(\psi \mid \vec{y})$ as its stationary distribution~\autocite[275]{gelmanBDA}.
+It can be proved that as $M\to\infty$, the sample tends towards a sample from the stationary distribution, here the target posterior.
+This section outlines some basic principles of MCMC and considerations when using it in practice.
+Any reader interested in more technical details or proofs of the results here should consult the references within this section.
+
+The Metropolis-Hastings algorithm~\autocite{hastingsMCMC} describes a general method for constructing such a Markov chain.
+The algorithm requires a proposal distribution $q(\psi' \mid \psi)$, which proposes moving to a new state $\psi'$ from the current state $\psi$, and a calculation of $p(\psi \mid \vec{y})$ up to some constant, $\pi(\psi)$.
+The proposal distribution is most commonly a Normal distribution centred on $\psi$, this is used in \cref{E-SEIR}.
+The proposal is then either accepted or rejected based on the \emph{acceptance ratio}, which is computed using $\pi$.
+The full algorithm is given in \cref{inc-prev:algorithm:MH}, with further details in available in a wide-range of sources~\autocites[e.g.][]{brooksMCMCNotes}[chapter 11]{gelmanBDA}.
+\begin{algorithm}
+ set $\psi^0$ to some initial value\;
+ \For{$i = 1, \dots, M$}{
+  sample a proposal $\psi' \sim q(\psi' \mid psi)$ \;
+  calculate the acceptance probability $\alpha(\psi, \psi') = \min \left( 1, \frac{\pi(\psi') q(\psi \mid \psi')}{\pi(\psi) q(\psi' \mid \psi)} \right)$ \;
+  with probability $\alpha(\psi, \psi')$ set $\psi^i$ to $\psi'$, otherwise set $\psi^i$ to $\psi^{i-1}$ \;
+ }
+ \caption{The Metropolis-Hastings algorithm.}
+ \label{inc-prev:algorithm:MH}
+\end{algorithm}
+
+\todo[inline]{Talk about HMC and NUTS}
+
+Two practical issues arise when applying MCMC: \emph{convergence} and \emph{stability}\todo{cite BUGS course}.
+
+Convergence is how quickly the Markov chain approaches the stationary distribution.
+Early draws from the Markov chain reflect the choice of the starting point, $\psi^0$, rather than the stationary distribution that the chain will eventually converge to~\autocite[282]{gelmanBDA}.
+The simplest technique for assessing convergence is graphically, using traceplots~\autocite[81]{brooksMCMCNotes}.
+Traceplots show the value of $\psi^i$ against $i$.
+If the chain has converged, then the traceplot will show a random walk around the stationary distribution.
+This can be improved by using multiple chains with different starting points~\autocite[283]{gelmanBDA}.
+If both chains have converged, then they will be in the same location.
+
+A more complex measure of convergence when running multiple MCMC chains is comparing the within-chain variance to the between-chains variance.
+The principle here is that if all chains have converged then they are sampling from the same distribution.
+This means that the within-chain and between-chain variances are the same~\autocite[82]{brooksMCMCNotes}.
+In particular, I will make use of the improved Rhat statistic~\autocite{vehtariRhat}.
+%The Rhat statistic is defined for a scalar parameter $\psi$.
+%Let $\psi^{(ij)}$ denote the $i$th (of $M$) sample from the $j$th chain (of $N$).
+%Let $\bar{\psi}^j$ denote the mean value of the samples of $\psi$ from the $j$th chain, $\bar{\psi}^j = \frac{1}{M} \sum_{i=1}^M \psi^{(ij)}$.
+%Let $\bar{\psi}$ denote the mean value of the samples of $\psi$ from all chains, $\bar{\psi} = \frac{1}{N} \sum_{j=1}^N \bar{\psi}^j$.
+%Then the between-chain variance is $B = \frac{M}{N-1} \sum_{j=1}^N (\bar{\psi}^j - \bar{\psi})^2$.
+%The within-chain variance is $W = \frac{1}{N} \sum_{j=1}^N \frac{1}{M-1} \sum_{i=1}^M (\psi^{(ij)} - \bar{\psi}^j)^2$.
+A Rhat of 1 means that the within-chain and between-chain variance are equal, informally that convergence is perfect, as far as can be measured by this technique.
+Various rules-of-thumb exist for what should be considered an acceptable Rhat.
+Historically, 1.1 has commonly been recommended, although more recent work suggests that 1.01 is more appropriate~\autocite{vehtariRhat}.
+
+The stability of estimates considers the error arising because only a finite sample of the distribution of interest is available.
+Furthermore, the sample will tend to be autocorrelated meaning that estimates using the example have higher variance than if it were an iid sample from the posterior~\autocite[286]{gelmanBDA}.
+The \emph{effective sample size} corrects for the autocorrelation.
+The effective sample size can be interpreted as the number of iid samples that would have the same variance as the MCMC sample~\autocite[286]{gelmanBDA}{vehtariRhat}.
+Various techniques have been proposed for computing the effective sample size, I will follow the approach of \textcite{vehtariRhat}.
+
 \section{Conclusion} \label{inc-prev:sec:conclusion}

### 2023-11-07 SEIR.tex
Fix chapter numbering
Start addresssing comments

@@
 \begin{document}
+\ifSubfilesClassLoaded{
+    \setcounter{chapter}{7}
+}
 \chapter{Mechanistic modelling of incidence and transmission} \label{SEIR}
+
+\todo[inline]{Add an intro with an overview of what this chapter contains. Include why I discuss mechanistic models, how I will use them, and what they add to the previous analyses.}
 
@@
 
-Notation in this chapter differs from the rest of the thesis.
-I adopt the conventions and notation standard in the infectious disease modelling field.
-In particular, upper-case letters denote numbers of people, and lower-case letters denote proportions.
-Whether a variable is random or not should be clear from the context, or will be explicitly stated if needed.
-
 The first sections of this chapter give the background and methodology of the model I use.
@@
 For further details I recommend the tutorial paper \textcite{kretzschmarMathematical} or the textbook \textcite{keelingModeling}.
+
+Notation in this chapter differs from the rest of the thesis.
+I adopt conventions and notation from the infectious disease modelling field, although this is not consistent in the literature.
+In particular, upper-case letters denote numbers of people, and lower-case letters denote proportions.
+Whether a variable is random or not should be clear from the context, or will be explicitly stated if needed.
 
@@
 
-Deterministic models have the property that once the parameters are set, there is no stochasticity in the transmission model.
-All other model variables are deterministic functions of the parameters.
+A compartmental model consists of a state vector $\vec{x}(t)$ and a set of parameters $\vec{\theta}$.
+The state vector is the proportion of the population (or number of individuals) in each compartment at time $t$.
+
+Deterministic models have the property that, for any $t$, the state vector $\vec{x}(t)$ is a deterministic function of the parameters $\vec{\theta}$ and the initial state $\vec{x}(0)$.
 The transmission model corresponds to the infection process (see \cref{E-inc-prev:sec:infection-process}).
@@
 The other type of model is \emph{stochastic}.
-Stochastic models have randomness in the transmission model, and often use discrete time.
+Stochastic models specify a probability distribution $p(\vec{x} \mid \vec{\theta})$ for the state vector $\vec{x}(t)$ at time $t$.
+Unlike the models previously encountered in this thesis, there will be dependence between $\vec{x}(t)$ and $\vec{x}(t')$.
 For example, the number of individuals infected on day $t+1$, conditional on the model state on day $t$, might be a Poisson distribution.
@@
 Therefore, in an ODE model, time is necessarily continuous although discretised when solving (see \cref{SEIR:sec:inference-implementation}).
+For concreteness, I use units of days throughout, although the model can use any unit of time.
 The number or proportion of individuals in each compartment is also continuous, which is a reasonable approximation when the population is large.
@@
 Deterministic models are also known as \emph{mean-field} models because they can be derived from stochastic models by taking the mean of the distribution of the number of infected individuals.
-Further discussion of stochastic models is beyond the scope of this thesis.
-\todo{maybe add a reference to a review or textbook on stochastic models} YES!
+Further discussion of stochastic models is beyond the scope of this thesis; \textcite[chapter 6]{keelingModeling} provides an introduction to these models.
 
-\subsection{The SIR model}
+\subsection{The SIR model} \label{SEIR:sec:SIR}
 \begin{figure}[h]
@@
 
-The SIR model is described by the following system of ODEs:
+The SIR model is described by the following system of ODEs~\autocite[19]{keelingModeling}:
 \begin{align}
 \frac{ds(t)}{dt} &= -\beta s(t)i \\
-\frac{di}{dt} &= \beta si - \gamma i \\
-\frac{dr}{dt} &= \gamma i
+\frac{di(t)}{dt} &= \beta s(t)i(t) - \gamma i(t) \\
+\frac{dr(t)}{dt} &= \gamma i(t)
 \end{align}
-$s$, $i$ and $r$ are the proportions of the population in the susceptible, infected and recovered compartments respectively.
+$s(t)$, $i(t)$ and $r(t)$ are the proportions of the population in the susceptible, infected and recovered compartments respectively.
 These are the state of the system.
@@
 
-The term $\beta si$ is the instantaneous proportion of the population becoming infected.
-That is, the proportion of the population moving from the susceptible state to the infected state.
+The term $\beta s(t)i(t)$ is the rate at which susceptible individuals (as a proportion of the total population size) are infected~\autocite[18]{keelingModeling}.
+Susceptible individuals move from the $s(t)$ to $i(t)$ when they are infected.
 This term can be arrived at by considering a single infected individual.
-Assume this individual makes contact with any other individual in the population at a constant rate, $c$, in units of contacts per day.
+Assume this individual makes contact with any other individual in the population at a constant rate, $\kappa$, in units of contacts per day.
 Further, assume the population is \emph{well-mixed}, meaning the individual is equally likely to come into contact with any other individual.
-For a randomly selected individual, the probability that they are susceptible is $s$.
-Each of the infected individual's contacts is therefore with a susceptible individual with probability $s$, by the well-mixed assumption.
-Therefore, in expectation, the infected individual makes contact with susceptible individuals at a rate of $cs$.
+For a randomly selected individual, the probability that they are susceptible is $s(t)$.
+Each of the infected individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
+Therefore, in expectation, the infected individual makes contact with susceptible individuals at a rate of $\kappa s(t)$.
 Assume that there is a constant probability of transmission upon contact between an infected and susceptible individual, $p_{si}$.
-The expected rate of infections caused by this infected individual is therefore $cs p_{si}$.
-The two constant terms here (number of contacts and probability of transmission) are absorbed into the transmission rate, $\beta = c p_{si}$.
-Therefore, the instantaneous number of infections per infected individual is $\beta s$.
-The overall proportion of the population instantaneously infected is $\beta si$.
+The expected rate of infections caused by this infected individual is therefore $\kappa s p_{si}$.
+The two constant terms here (number of contacts and probability of transmission) are absorbed into the transmission rate, $\beta = \kappa p_{si}$.
+Therefore, the instantaneous number of infections per infected individual is $\beta s(t)$.
+The overall proportion of the population instantaneously infected is $\beta s(t)i(t)$.
 
-The term $\gamma i$, the instantaneous proportion of the population recovering, follows directly from the definitions of $\gamma$ and $i$.
+The term $\gamma i$, the rate at which infected individuals (as a proportion of the total population size) recover, follows directly from the definitions of $\gamma$ and $i$.
+
+Other quantities of this model are of interest.
+The \emph{force of infection}, $\lambda(t) = \beta i(t)$, is the instantaneous risk of infection for a susceptible individual~\autocite[17]{keelingModeling}.
+The basic reproduction number (discussed previously in\todo{ref intro}) for this model is $R_0 = \beta / \gamma$~\autocite[20]{keelingModeling}.
+Intuitively, this can be thought of as the average number of infections caused by an infectious individual per day, $\beta$, multiplied by the average number of days they are infectious for, $1/\gamma$.
 
@@
 That is, the total population size is constant, and no infections are imported from outside the population.
-Births and deaths of individuals are also excluded.
-Second, there is no analytical solution to the system of ODEs.
-Numerical methods are required to solve the system.
-Third, in addition to the parameters, the state of the system at some time (normally the initial state) is required to solve the system.
+Births and deaths of individuals are also excluded, although they can easily be incorporated in situations where they are relevant~\autocites[26]{keelingModeling}[214]{kretzschmarMathematical}.
+Second, there is no analytical solution to the system of ODEs~\autocite[25]{keelingModeling}: numerical methods are required to solve the system.
+Third, in addition to the parameters, the state of the system at some time (normally the initial state, the state at time $t=0$) is required to solve the system.
 
@@
 
-The basic SIR model implies that the distribution of the time spent in the infected compartment is exponential.
+The basic SIR model implies that the distribution of the time spent in the infected compartment is exponential~\autocite[96]{keelingModeling}.
 This is an unrealistic assumption and leads to underestimation of the reproduction number~\autocites{lloydRealistic}{wearingAppropriate}.
@@
 This is done by adding multiple infected compartments (see \cref{SEIR:fig:SIR-gamma}).
-The system can now be written as follows:
+The system can now be written as follows~\autocite[94]{keelingModeling}:
 \begin{align}
@@
 The infectious period is now distributed $\GamDist(n, n\gamma)$.
-A gamma distribution with the first parameter equal to 1 is an exponential distribution, connecting this model to the basic SIR model.
-$1/\gamma$ remains the mean infectious period by construction.
+A gamma distribution with the first parameter equal to 1 is an exponential distribution, connecting this model to the basic SIR model in \cref{SEIR:sec:SIR}.
+The distribution for the time spent in the infected compartments tends to a constant (\ie the same for all individuals) as $n \to \infty$~\autocite{lloydRealistic}.
+The form of these equations is chosen so that $1/\gamma$ remains the mean infectious period.
 
@@
 The state of each compartment is now a vector.
-The proportion of the population that is susceptible is $\vec{s}$ where $s_i$ is the proportion of the $i$th age group that is susceptible.
-Similarly for the other compartments.
+The proportion of the population that is susceptible is $\vec{s} = (s_1, \dots, s_A)$ where $s_a$ is the proportion of the $a$th age group that is susceptible.
+Similarly, for the other compartments.
 
 The modelling approach here can be used to stratify the population by any characteristic.
-For example, using deprivation or ethnicity.
+For example, HIV/AIDS models incorporated subgroups with different patterns of sexual activity~\autocite{kretzschmarMathematical}.
+In other contexts, transmission is strongly influenced by socio-economic status, and the model can be stratified by income levels~\autocite{menkirIncorporating}.
 In this case, $i$ simply represents the $i$th strata.
@@
 Consider an infected individual in stratum $j$.
-Assume they make contacts with individuals in stratum $i$ at a rate of $M_{ij}$ per day.
-Therefore, they make contact with susceptible individuals in stratum $i$ at a rate of $s_i M{ij}$ per day.
+Assume they make contacts with individuals in stratum $i$ at a rate of $K_{ij}$ per day.
+Therefore, they make contact with susceptible individuals in stratum $i$ at a rate of $s_i K{ij}$ per day.
 Assume that the probability of transmission from an infected individual in stratum $j$ to a susceptible individual in stratum $i$ upon contact is $\beta_{ij}$.
-Therefore, $M_{ij} \beta_{ij} s_i$ is the instantaneous proportion of individuals in stratum $i$ infected by an individual in stratum $j$.
-Considering infected individuals across all strata, $\sum_{j \in \set{A}} M_{ij} \beta_{ij} s_i$ is the instantaneous proportion of individuals in stratum $i$ infected.
+Therefore, $K_{ij} \beta_{ij} s_i$ is the instantaneous proportion of individuals in stratum $i$ infected by an individual in stratum $j$.
+Considering infected individuals across all strata, $\sum_{j \in \set{A}} K_{ij} \beta_{ij} s_i$ is the instantaneous proportion of individuals in stratum $i$ infected.
 
 \begin{align}
-    \frac{d\vec{s}}{dt} = -\vec{s} \circ ((\matr{\beta} \circ \matr{M}) \vec{i})
+    \frac{d\vec{s}}{dt} = -\vec{s} \circ ((\matr{\beta} \circ \matr{K}) \vec{i})
     \label{eq:matrix:theory}
@@
 
-Clearly $M_{ij}$ and $\beta_{ij}$ are not individually identifiable.
-Normally, $M_{ij}$ is assumed to be known, for example from a previous contact survey~\autocite[such as][]{mossongSocial}.
+$K_{ij}$ and $\beta_{ij}$ appear only as a product and therefore are not individually identifiable.
+Normally, $K_{ij}$ is assumed to be known, for example from a previous contact survey~\autocite[such as][]{mossongSocial}.
 $\beta_{ij}$ is then estimated, although often with a parsimonious model.
@@
 The parameters governing the duration of positivity ($\kappa$s) were explained previous (see \cref{SEIR:table:ec-params}).
+\todo[inline]{Explain other priors (identifiability or weakly informative)}
 

### 2023-11-10 cis-perfect-testing.tex
Fix chapter numbering
Bulk of ATACCC chapter (no results yet)
Various edits

@@
 \begin{document}
+\ifSubfilesClassLoaded{
+
 \section{Definitions}
@@
 \end{itemize}
+
+\setcounter{chapter}{4}
+}
 
@@
 
-Now I derive an analytical solution to $\sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under the prior $\ntot \dist \NB(\mu, r)$ (using the mean/dispersion parameterisation of the negative binomial, see \cref{E-distributions}).
+Now I derive an analytical solution to $\sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under the prior $\ntot \dist \NB_c(\mu, r)$.
 I assume that $\theta$, which are only the parameters of the survival distribution, is a priori independent of $\ntot$, the number of infections that occurred in the cohort.
 Therefore, $p(\ntot \mid \theta) = p(\ntot)$.
-Putting a negative binomial prior on $\ntot$ is equivalent to the following gamma-Poisson composite.
-Using the composite simplifies the derivation.
+Putting a negative binomial prior on $\ntot$ is equivalent to the following gamma-Poisson composite; its use simplifies the derivation.
 \begin{align}
@@
 RW2 prior encodes that the hazard should be linearly changing with some random changes at each time step.
-Specifically, the difference on the logit scale between $\lambda_t$ to $\lambda_{t+1}$ is the difference between $\lambda_{t-1}$ and $\lambda_t$ plus some Gaussian-distributed noise.
+Specifically, the difference on the logit scale between $\lambda_t$ to $\lambda_{t+1}$ is the difference between $\lambda_{t-1}$ and $\lambda_t$ plus some Normally-distributed noise.
 Formally:
@@
 
-Simulation was performed in R~4.2.0~\autocite{R4-2-0} using tidyverse~2.0.0~\autocite{tidyverse} and targets~1.1.3~\autocite{targetsPackage}.
+Simulation was performed in R~4.2.0~\autocite{R-4-2-0} using tidyverse~2.0.0~\autocite{tidyverse} and targets~1.1.3~\autocite{targetsPackage}.
 

### 2023-11-10 distributions.tex
Various edits

@@
 Poisson & $\Poi(\lambda)$ & $\lambda > 0$ (average rate) & $\lambda$ & $\lambda$ & $\frac{e^{-\lambda}\lambda^x}{x!}$ \\
-Negative Binomial (standard) & $\NB(r, p)$ & $r > 0$ (number of failures until stopping)\newline $0 < p < 1$ (probability of success) & $\frac{pr}{1-p}$ & $\frac{pr}{(1-p)^2}$ & $\frac{\Gamma(x+r)}{x!
+Negative Binomial (standard) & $\NBs(r, p)$ & $r > 0$ (number of failures until stopping)\newline $0 < p < 1$ (probability of success) & $\frac{pr}{1-p}$ & $\frac{pr}{(1-p)^2}$ & $\frac{\Gamma(x+r)}{x!
 \Gamma(r)} p^x (1-p)^r$ \\
-Negative Binomial (centred) & $\NB(\mu, r)$ & $\mu > 0$ (mean) \newline r (dispersion) & $\frac{pr}{1-p}$ & $\frac{pr}{(1-p)^2}$ & $\frac{\Gamma(x+r)}{x!
+Negative Binomial (centred) & $\NBc(\mu, r)$ & $\mu > 0$ (mean) \newline r (dispersion) & $\frac{pr}{1-p}$ & $\frac{pr}{(1-p)^2}$ & $\frac{\Gamma(x+r)}{x!
 \Gamma(r)} \left( \frac{r}{r+\mu} \right)^\mu \left( \frac{\mu}{r+\mu} \right)^k$ \\

### 2023-11-10 incidence-prevalence.tex
Various edits

@@
 Here, I adapt previous work to the SARS-CoV-2 context.
-In particular, the recoveries need explicit modelling which was not required for HIV/AIDS.
+In particular, recoveries need considering, which was not the case for HIV/AIDS.
 The observation process describes the relationship between prevalence and the observed data.
@@
 Therefore, the model is specific to this context, although much would be applicable to other, similarly designed prevalence surveys.
-Finally, \cref{inc-prev:sec:MCMC} will introduce Markov chain Monte Carlo (MCMC), an inference technique I will make extensive use of in this thesis, before \cref{inc-prev:sec:conclusion} concludes the chapter.
+Finally, \cref{inc-prev:sec:MCMC} will introduce Markov chain Monte Carlo (MCMC), an inference technique I will make extensive use of in this thesis.
+\Cref{inc-prev:sec:conclusion} concludes the chapter.
 
@@
 I use days, however, the results generalise in the obvious way to other time units.
-I make this choice for pragmatic reasons.
+I choose discrete time for pragmatic reasons.
 Most epidemiological data is only available at daily granularity or, when more granular data exists, is often unreliable.
@@
 What it means to have a disease can vary by context (see\todo{ref section discussing what it means to have covid}).
-I define an individual to have SARS-CoV-2 on day $t$ if, had they been swabbed on day $t$, that swab would have returned a positive PCR test.
+Here, an individual is defined to have SARS-CoV-2 on day $t$ if, had they been swabbed on day $t$, that swab would return a positive PCR test.
 I refer to these individuals as \emph{prevalent} individuals.
@@
 I model each $D_i$ as a discrete independent and identically distributed (iid) random variable; extending the methodology to account for covariates is left to further work.
-The duration is a positive integer, \ie $D_i \in \{1, 2, \dots\}$.
-The first way of expressing the distribution of $D_i$ is as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=0}^d f_{D}(i)$.
-The \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$~\autocite{yanDistribution}.
-The \emph{hazard} at time $t$ is the probability of recovering at time $t$ conditional on having survived to time $t$: $\lambda(t) = \prob(D = t \mid D \geq t) = f_{D}(d) / S(d)$.
-For discrete variables, the convention that the survival function is $\prob(D >d)$, rather than $\prob(D \geq d)$, is also commonly used, I adapt that latter convention.
-Throughout this thesis, the survival function will be the primary quantity of interest, and it will normally be parameterised in terms of the hazard; these two quantities are related by $S(t) = \prod_{i=1}^{t-1} (1 - \lambda(i))$.
+The duration is a positive integer, \ie $D_i \in \{1, 2, \dots \dmax \}$, where $\dmax$ is the maximum duration.
+The first way of expressing the distribution of $D_i$ is as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
+The \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$ (in some of the literature on survival with discrete variables, the convention that the survival function is $\prob(D >d)$, rather than $\prob(D \geq d)$, is used).
+The \emph{hazard} on day $d$ is the probability of recovering at time $d$ conditional on having survived to time $d$: $\lambda(d) = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
+Throughout this thesis, the survival function will be the primary quantity of interest, and it will normally be parameterised in terms of the hazard; these two quantities are related by $S(d) = \prod_{i=1}^{d-1} (1 - \lambda(i))$.
+
+Since $D_i$ is a discrete random variable, $f_D(d) \geq 0$ for all $d$ and $\sum_{i=1}^{\dmax} f_D(i) = 1$.
+This implies that $F_D(d)$ is monotonically increasing.
+Therefore, $S(d)$ is monotonically decreasing.
+Both $F_D$ are $S$ are in the interval $[0, 1]$.
+$\lambda(d)$ is in the interval $[0, 1]$, and can otherwise vary arbitrarily (\ie there is no monotonic or sum constraint).
 
@@
 A point process is a random function of continuous time $K(t)$.
-$K(t)$ must take non-negative integer values and be non-decreasing in $t$.
+$K(t)$ must take non-negative integer values and be monotonically increasing in $t$.
 The interpretation of $K(t)$ is the number of events, here infections, that have occurred in the interval $(-\infty, t]$~\autocites[244]{yanDistribution}.
-In the context of a disease, the point process is discretised into incidence for integer $t$ as $Z_t = K(t) - K(t-1)$.
+In the context of a disease, the point process can be discretised into the incidence on day $t$ as $Z_t = K(t) - K(t-1)$, where $t$ is an integer.
 
-A counting process has \emph{independent increments} if the number of events in disjoint intervals are independent~\autocite[244]{yanDistribution}.
+A counting process has \emph{independent increments} if the number of events in disjoint intervals are independent conditional on the intensity~\autocite[244]{yanDistribution}.
 That is, for any $t_1 < t_2 < t_3 < t_4$, $K(t_2) - K(t_1)$ and $K(t_4) - K(t_3)$ are independent.
@@
 A commonly-used point process for infectious disease epidemiology is the time-inhomogenous Poisson process~\autocites{brookmeyerMethod}{paganoHIV}{rosenbergBackcalculation}{brookmeyerBackcalculation}.
-A Poisson process is a point process with independent increments, and the property that $K(s) - K(t) \dist \Poi\left(\int^s_t \lambda(r) dr \right)$ where $\lambda(r) \geq 0$ is the \emph{intensity} of the epidemic at time $r$~\autocites[244]{yanDistribution}.
-A time-homogenous Poisson process is the special case where $\lambda(r)$ is constant.
-The independent increment property says that, conditional on $\lambda(r)$, the number of infections in disjoint intervals are independent.
+A Poisson process is a point process with independent increments, and the property that $K(s) - K(t) \dist \Poi\left(\int^s_t \nu(r) dr \right)$ where $\nu(r) \geq 0$ is the \emph{intensity} of the epidemic at time $r$~\autocites[244]{yanDistribution}.
+A time-homogenous Poisson process is the special case where $\nu(r)$ is constant.
+The independent increment property says that, conditional on $\nu(r)$, the number of infections in disjoint intervals are independent.
 This implies $Z_t$ and $Z_{t'}$ are independent for $t \neq t'$.
@@
 Two features of a Poisson process are violated for epidemics.
-Epidemics commonly exhibit overdispersion and have feedback loops~\autocite{beckerDependent}.
-Overdispersion means that $\V(Z_t) > \E(Z_t)$ for any day $t$.
-In a Poisson process these counts have a Poisson distribution, hence, $\V(Z_t) = \E(Z_t)$.
-Feedback loops mean that the stochastically low or high numbers of infections, possibly due to overdispersion, affect the future number of infections.
-This violates the assumption of independent increments.
-In particular, $\lambda(t+\delta t)$ is dependent on $K(t)$ for $\delta t > 0$.
-Both of these features violate the properties of a Poisson process.
-Despite these issues, consistent inference on the incidence can still be produced using the Poisson process model, although the uncertainty in the estimates may be understated~\autocite{beckerDependent}.
+Epidemics commonly exhibit \emph{overdispersion} and have feedback loops~\autocite{beckerDependent}.
+
+Overdispersion means that $\V(Z_t) > \E(Z_t)$.
+In a Poisson process $Z_t$ has a Poisson distribution, hence, $\V(Z_t) = \E(Z_t)$.
+Therefore, overdispersed epidemics cannot be modelled by a Poisson process.
 
@@
 For comparison, the average individual infected around three others in the early stages of the pandemic~\autocite{pellisChallenges}. 
-Overdispersion can be incorporated by replacing the assumption that $K(s) - K(t)$ is Poisson distributed with assuming it is negative binomially distributed.
-This would not affect $\E(K(s) - K(t))$, but the negative binomial introduces an additional overdispersion parameter, commonly notated $k$.
+Overdispersion can be incorporated by replacing the assumption that $K(s) - K(t) \dist \Poi(\int_{t}^{s} \nu(r) dr)$ with $K(s) - K(t) \dist \NBc(\int_{t}^{s} \nu(r) dr, k)$ where $k$ is the overdispersion parameter, normally assumed to be constant absent interventions.
+Note that the expected number of events has not changed, but the variance has increased (see \cref{E-distributions} for details of parameterisations and moments).
 %The negative binomial tends to Poisson as $k\to\infty$.
@@
 
-The independence assumption of a Poisson or negative binomial process is violated because if the number of infections is stochastically high (or low), this feeds back to the process intensity at future times.
-This can be seen simply in the event of a super-spreading event.
+Feedback loops mean that stochastically low or high numbers of infections affect the future number of infections.
+That is, $\nu(t+\Delta t)$ is dependent on $K(t)$, even after conditioning on $\nu(t)$, for $\Delta t > 0$.
+This violates the assumption of independent increments.
+Both of these features violate the properties of a Poisson process.
+Despite these issues, consistent inference on the incidence can still be produced using the Poisson process model, although the uncertainty in the estimates may be understated~\autocite{beckerDependent}.
+
+The independence assumption of a Poisson or negative binomial process is violated by an infectious disease because stochastically high (or low) or low incidence, feeds back to the process intensity at future times.
+This can be seen most obviously in the event of a super-spreading event.
 If a super-spreading event occurs, then many more individuals than suggested by the process intensity will be infected.
@@
 
-Denote by $R_{t,t'}$ be the number of infection episodes with $B_i=t$ and $E_i=t'$.
+Denote by $R_{t,t'}$ the number of infection episodes with $B_i=t$ and $E_i=t'$.
 For each infection beginning at $t$, the probability that it ends at $t'$ is $\prob(E_i = t' \mid B_i = t) = \prob(D_i = t' - t + 1 \mid B_i = t) = f_D(t' - t + 1)$.
@@
 \end{align}
-Now, denote by $P_{t_z,t_p}$ the number of infection episodes with $B_i = t_z$ and $E_i \geq t_p$.
-Therefore, the individuals in which these infection episodes occur are prevalent at time $t_p$.
+Now, denote by $P_{t_z,t_p}$ the number of individuals infected at $t_z$ and prevalent at $t_p$.
+Formally, $P_{t_z,t_p}$ is the infection episodes with $B_i = t_z$ and $E_i \geq t_p$.
 $P_{t_z,t_p}$ is $Z_{t_z}$ minus those that recover before $t_p$ (\ie at or up to $t_p - 1$).
@@
 The algorithm requires a proposal distribution $q(\psi' \mid \psi)$, which proposes moving to a new state $\psi'$ from the current state $\psi$, and a calculation of $p(\psi \mid \vec{y})$ up to some constant, $\pi(\psi)$.
-The proposal distribution is most commonly a Normal distribution centred on $\psi$, this is used in \cref{E-SEIR}.
+The proposal distribution is most commonly a (possibly multivariate) Normal distribution centred on $\psi$, this is used in \cref{E-SEIR}.
 The proposal is then either accepted or rejected based on the \emph{acceptance ratio}, which is computed using $\pi$.
-The full algorithm is given in \cref{inc-prev:algorithm:MH}, with further details in available in a wide-range of sources~\autocites[e.g.][]{brooksMCMCNotes}[chapter 11]{gelmanBDA}.
+Tuning the proposal distribution to propose ``good'' moves is a key part of MCMC.
+The ideal proposal distribution is the posterior of interest, however, this is impossible because that is what MCMC is trying to estimate~\autocite[296]{gelmanBDA}.
+One alternative is \emph{adaptive} algorithms which tune the proposal distribution based on previous draws in an attempt to approximate the posterior; I use such a scheme in \cref{E-SEIR:sec:MCMC}.
+The full Metropolis-Hastings algorithm is given in \cref{inc-prev:algorithm:MH}, with further details in available in a wide-range of sources~\autocites[e.g.][]{brooksMCMCNotes}[chapter 11]{gelmanBDA}.
 \begin{algorithm}
@@
 
-\todo[inline]{Talk about HMC and NUTS}
+MCMC using a Normal proposal distribution can be inefficient.
+In particular, it often fails to propose moves into the tails of the posterior distribution\todo{any suggested refs for this (Paul?) - we've discussed this lots but I've never read it anywhere}.
+\emph{Hamiltonian Monte Carlo} (HMC)~\autocites{nealMCMC}{nealImproved}{duaneHybrid} improves the proposals by using information from the gradient of the posterior density (with respect to the parameters).
+HMC simulates a physical system travelling through the posterior space with momentum.
+The momentum increases in the direction the posterior density increases, using the gradient.
+The \emph{No U-Turn Sampler} (NUTS)~\autocite{hoffmanNUTS} is a variant of HMC that automatically tunes the number of steps to take in the Hamiltonian dynamics.
+NUTS has been implemented in a variety of software packages making it easy to apply to arbitrary posterior distributions.
+I will make use of the implementation in the package \emph{Sampling through adaptive neighborhoods} (Stan)~\autocite{Stan-2-32-2}.
+Stan implements \emph{automatic differentiation} to calculate the gradient of the posterior density without any user input.
+Automatic differentiation uses the chain rule to calculate the gradient of a function from its code by breaking it down into a series of elementary operations, for which the gradient is known.
+For a more detailed introduction to HMC, NUTS, and Stan see \textcite[chapter 12]{gelmanBDA}.
 
@@
 This can be improved by using multiple chains with different starting points~\autocite[283]{gelmanBDA}.
-If both chains have converged, then they will be in the same location.
+If all chains have converged, then they will be in the same location.
 
@@
 The \emph{effective sample size} corrects for the autocorrelation.
-The effective sample size can be interpreted as the number of iid samples that would have the same variance as the MCMC sample~\autocite[286]{gelmanBDA}{vehtariRhat}.
-Various techniques have been proposed for computing the effective sample size, I will follow the approach of \textcite{vehtariRhat}.
+The effective sample size can be interpreted as the number of iid samples that would have the same variance as the MCMC sample~\autocites[286]{gelmanBDA}{vehtariRhat}.
+Various techniques have been proposed for computing the effective sample size, I will follow the approach of \textcite{vehtariRhat}, implemented in RStan~\autocite{RStan-2-32-3}.
 

### 2023-11-10 cis-perfect-testing.tex
Fix typo

@@
 
-Now I derive an analytical solution to $\sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under the prior $\ntot \dist \NB_c(\mu, r)$.
+Now I derive an analytical solution to $\sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under the prior $\ntot \dist \NBc(\mu, r)$.
 I assume that $\theta$, which are only the parameters of the survival distribution, is a priori independent of $\ntot$, the number of infections that occurred in the cohort.

### 2023-11-13 SEIR.tex
Various edits
Fix formatting

@@
 }
+
 \chapter{Mechanistic modelling of incidence and transmission} \label{SEIR}
@@
 The infectious period is now distributed $\GamDist(n, n\gamma)$.
+Gamma distributions are a realistic model for a variety of diseases~\autocite{wearingAppropriate}.
 A gamma distribution with the first parameter equal to 1 is an exponential distribution, connecting this model to the basic SIR model in \cref{SEIR:sec:SIR}.
@@
 The form of these equations is chosen so that $1/\gamma$ remains the mean infectious period.
-
-Using a two parameter distribution allows the first two moments of the infectious period to be specified.
-This is normally sufficient to give realistic dynamics within a mean-field model.
 
@@
 \todo[inline]{is the above paragraph too hand wavy?}
+
+I use previously developed estimates of how contact matrices change over time.
+These matrices were produced for \textcite{birrellRealtime}.
+These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
+These use data from the POLYMOD study~\autocite{mossongSocial} and UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
+POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on age and location (home, work, school, leisure, transport, or other) of these contacts.
+UKTUS asked a representative sample of the UK population about how much time they spent doing various activities each day.
+\Textcite{vanleeuwenAugmenting} combined these two datasets to estimate the rate of contacts between age groups during each activity.
+\Textcite{birrellRealtime} then modify these contact rates based on measured changes in time use during the pandemic.
+These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
+Google mobility data measures the location of Android users over time, relative to a pre-pandemic baseline.
+These data sources are then used to down or upweight the number of contacts in each setting, as estimated by \textcite{vanleeuwenAugmenting}.
+For mathematical details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.
 
@@
 
-\subsection{MCMC}
+\subsection{MCMC} \label{SEIR:sec:MCMC}
 

### 2023-11-13 incidence-prevalence.tex
Further edits

@@
 In particular, recoveries need considering, which was not the case for HIV/AIDS.
-The observation process describes the relationship between prevalence and the observed data.
+The observation process (\cref{inc-prev:sec:observation-process}) describes the relationship between prevalence and the observed data, ultimately allowing inference of incidence based on this data.
 Unlike in HIV/AIDS backcalculation, the CIS has a sampling mechanism that needs to be modelled.
@@
 I model each $D_i$ as a discrete independent and identically distributed (iid) random variable; extending the methodology to account for covariates is left to further work.
-The duration is a positive integer, \ie $D_i \in \{1, 2, \dots \dmax \}$, where $\dmax$ is the maximum duration.
+The duration is a positive integer, \ie $D_i \in \{1, 2, \dots, \dmax \}$, where $\dmax$ is the maximum duration.
 The first way of expressing the distribution of $D_i$ is as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
-The \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$ (in some of the literature on survival with discrete variables, the convention that the survival function is $\prob(D >d)$, rather than $\prob(D \geq d)$, is used).
+The \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$ (a common alternative convention for survival analysis with discrete variables is to define the survival function as $\prob(D >d)$).
 The \emph{hazard} on day $d$ is the probability of recovering at time $d$ conditional on having survived to time $d$: $\lambda(d) = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
@@
 Therefore, $S(d)$ is monotonically decreasing.
-Both $F_D$ are $S$ are in the interval $[0, 1]$.
+Both $F_D$ and $S$ are in the interval $[0, 1]$.
 $\lambda(d)$ is in the interval $[0, 1]$, and can otherwise vary arbitrarily (\ie there is no monotonic or sum constraint).
@@
 \section{Infection process} \label{inc-prev:sec:infection-process}
-The infection process is a \emph{point} or \emph{counting} process, where each infection is an event.
+The infection process is the mathematical description of incidence.
+It is a \emph{point} or \emph{counting} process, where each infection is an event.
 
@@
 In a Poisson process $Z_t$ has a Poisson distribution, hence, $\V(Z_t) = \E(Z_t)$.
-Therefore, overdispersed epidemics cannot be modelled by a Poisson process.
+Therefore, an overdispersed epidemic modelled as a Poisson process will be less variable than the realised epidemic.
 
@@
 This violates the assumption of independent increments.
-Both of these features violate the properties of a Poisson process.
-Despite these issues, consistent inference on the incidence can still be produced using the Poisson process model, although the uncertainty in the estimates may be understated~\autocite{beckerDependent}.
-
-The independence assumption of a Poisson or negative binomial process is violated by an infectious disease because stochastically high (or low) or low incidence, feeds back to the process intensity at future times.
+The reasons infectious diseases generate feedback loops is because stochastically high or low incidence feeds back to the process intensity at future times.
 This can be seen most obviously in the event of a super-spreading event.
@@
 Point processes can be extended to \emph{branching processes} to account for this feedback~\autocite[246]{yanDistribution}, although these are beyond the scope of this thesis.
+
+Both overdispersion and feedback loops violate the properties of a Poisson process.
+Despite these issues, consistent inference on the incidence can still be produced using the Poisson process model, although the uncertainty in the estimates may be understated~\autocite{beckerDependent}.
 
@@
 
-In \cref{inc-prev:sec:observation-process} I will show that only $\E(P_t \mid \vec{Z})$ and bounds on $\V(P_t \mid \vec{Z})$ will be relevant.
+In \cref{inc-prev:sec:observation-process} I will show that only $\E(P_t \mid \vec{Z})$ and bounds on $\V(P_t \mid \vec{Z})$ will be relevant for inferring incidence.
 These can be derived from their constituent parts.
@@
   &= \V\left(\sum_{i=1}^{d_\text{max}-1} P_{t-i,t} \mid \vec{Z} \right) &\text{by \cref{inc-prev:eq:Pt-to-Ptt}} \\
-  &= \sum_{i=1}^{d_\text{max}-1} \V\left(P_{t-i,t} \mid \vec{Z} \right) &\text{by the conditionl independence} \\
+  &= \sum_{i=1}^{d_\text{max}-1} \V\left(P_{t-i,t} \mid \vec{Z} \right) &\text{by the conditional independence} \\
   &= \sum_{i=1}^{d_\text{max}-1} \V\left(\sum_{j=0}^{i-1} R_{t-i,t-i+j} \mid \vec{Z} \right) &\text{by~\cref{inc-prev:eq:Ptt-to-R}}\\
@@
 The system can be written in the form $\E(\vec{P} \mid \vec{Z}) = \matr{L} \vec{Z}$ where $\matr{L}$ is a lower-triangular matrix.
-Numerical methods for solving systems of this form, that is solving for $\vec{Z}$ given $\vec{P}$, are well-established~\autocite[e.g.:][section 8.2]{highamAccuracy}.
+Numerical methods for solving systems of this form, that is solving for $\vec{Z}$ (non-random in this expression) for a known $\E (\vec{P} \mid \vec{Z} )$, are well-established~\autocite[e.g.:][section 8.2]{highamAccuracy}.
 However, these algorithms can be unstable, meaning that numerical errors are large.
@@
 The variance in \cref{inc-prev:eq:boundVPt} is upper bounded by \cref{inc-prev:eq:EPt}, a fact that will prove useful in \cref{inc-prev:sec:observation-process}.
-\Cref{inc-prev:eq:EPt} shows that knowledge of the survival function, $S$, is crucial to the prevalence process.
+\Cref{inc-prev:eq:EPt} shows that knowledge of the survival function, $S$, is crucial to modelling the prevalence process.
 
@@
 
-On day $t$ (for $t = 1, \dots, T$) a prevalence survey such as the CIS samples $n_t$ individuals at random from the population of interest and tests if they are prevalent.
+On day $t$ (for $t = 1, \dots, T$), a prevalence survey such as the CIS samples $n_t$ individuals at random from the population of interest and tests if they are prevalent.
 Therefore, I model the number of positive tests, $X_t$, as $X_t \mid P_t, n_t, \Npop \dist \Bin(n_t, P_t/\Npop)$.
@@
 MCMC is a class of algorithms for sampling from a distribution where the normalising constant for the distribution is unknown.
-I will use it to sample a set of parameters, say $\psi$ (which may be a scalar or vector), from a posterior distribution $p(\psi \mid y)$ where $y$ is the data.
-MCMC algorithms produce a sample $\psi^1, \psi^2, \dots, \psi^M$ from $p(\psi \mid \vec{y})$ by constructing a Markov chain with $p(\psi \mid \vec{y})$ as its stationary distribution~\autocite[275]{gelmanBDA}.
-It can be proved that as $M\to\infty$, the sample tends towards a sample from the stationary distribution, here the target posterior.
+I will use it to sample a set of parameters, say $\psi$ (which may be a scalar or vector), from a posterior distribution $p(\psi \mid y)$ where $y$ is the data (again, scalar or vector).
+MCMC algorithms produce a sample $\psi^1, \psi^2, \dots, \psi^M$ from $p(\psi \mid y)$ by constructing a Markov chain with $p(\psi \mid y)$ as its stationary distribution~\autocite[275]{gelmanBDA}.
+It can be proved that as $M\to\infty$, the sample tends towards a sample from the stationary distribution, \ie the target posterior.
 This section outlines some basic principles of MCMC and considerations when using it in practice.
@@
 
-The Metropolis-Hastings algorithm~\autocite{hastingsMCMC} describes a general method for constructing such a Markov chain.
-The algorithm requires a proposal distribution $q(\psi' \mid \psi)$, which proposes moving to a new state $\psi'$ from the current state $\psi$, and a calculation of $p(\psi \mid \vec{y})$ up to some constant, $\pi(\psi)$.
-The proposal distribution is most commonly a (possibly multivariate) Normal distribution centred on $\psi$, this is used in \cref{E-SEIR}.
+A common way of constructing a Markov chain with $p(\psi \mid y)$ as its stationary distribution is using the Metropolis-Hastings algorithm~\autocite{hastingsMCMC}.
+The algorithm requires a proposal distribution $q(\psi' \mid \psi)$, which proposes that the chain moves to a new state $\psi'$ from its current state $\psi$, and a calculation of $p(\psi \mid y)$ up to some constant, $\pi(\psi)$.
 The proposal is then either accepted or rejected based on the \emph{acceptance ratio}, which is computed using $\pi$.
-Tuning the proposal distribution to propose ``good'' moves is a key part of MCMC.
-The ideal proposal distribution is the posterior of interest, however, this is impossible because that is what MCMC is trying to estimate~\autocite[296]{gelmanBDA}.
-One alternative is \emph{adaptive} algorithms which tune the proposal distribution based on previous draws in an attempt to approximate the posterior; I use such a scheme in \cref{E-SEIR:sec:MCMC}.
-The full Metropolis-Hastings algorithm is given in \cref{inc-prev:algorithm:MH}, with further details in available in a wide-range of sources~\autocites[e.g.][]{brooksMCMCNotes}[chapter 11]{gelmanBDA}.
+The full algorithm is given in \cref{inc-prev:algorithm:MH}, with further details in available in a wide-range of sources~\autocites[e.g.][]{brooksMCMCNotes}[chapter 11]{gelmanBDA}.
 \begin{algorithm}
@@
  \For{$i = 1, \dots, M$}{
-  sample a proposal $\psi' \sim q(\psi' \mid psi)$ \;
+  sample a proposal $\psi' \sim q(\psi' \mid \psi)$ \;
   calculate the acceptance probability $\alpha(\psi, \psi') = \min \left( 1, \frac{\pi(\psi') q(\psi \mid \psi')}{\pi(\psi) q(\psi' \mid \psi)} \right)$ \;
@@
 
+The proposal distribution $q$ is most commonly a (possibly multivariate) Normal distribution centred on $\psi$; this scheme is used in \cref{E-SEIR}.
+Tuning the proposal distribution to propose ``good'' moves is a key part of MCMC.
+The ideal proposal distribution is the posterior of interest, however, this is impossible because that is what MCMC is trying to estimate~\autocite[296]{gelmanBDA}.
+One alternative is \emph{adaptive} algorithms which tune the proposal distribution based on previous draws in an attempt to approximate the posterior; I use such a scheme in \cref{E-SEIR:sec:MCMC}.
+
 MCMC using a Normal proposal distribution can be inefficient.
 In particular, it often fails to propose moves into the tails of the posterior distribution\todo{any suggested refs for this (Paul?) - we've discussed this lots but I've never read it anywhere}.
-\emph{Hamiltonian Monte Carlo} (HMC)~\autocites{nealMCMC}{nealImproved}{duaneHybrid} improves the proposals by using information from the gradient of the posterior density (with respect to the parameters).
+\emph{Hamiltonian Monte Carlo} (HMC)~\autocites{nealMCMC}{nealImproved}{duaneHybrid} improves the proposals by using information from the gradient of the posterior density (taken with respect to the parameters).
 HMC simulates a physical system travelling through the posterior space with momentum.
 The momentum increases in the direction the posterior density increases, using the gradient.
-The \emph{No U-Turn Sampler} (NUTS)~\autocite{hoffmanNUTS} is a variant of HMC that automatically tunes the number of steps to take in the Hamiltonian dynamics.
+\todo{add something about leapfrog steps}
+This means that the sampler tends to propose moves with higher posterior density.
+The Metropolis-Hastings algorithm is then used to accept or reject the proposed move.
+
+The \emph{No U-Turn Sampler} (NUTS)~\autocite{hoffmanNUTS} is a variant of HMC that automatically tunes the number of leapfrog steps to take in the Hamiltonian dynamics.
 NUTS has been implemented in a variety of software packages making it easy to apply to arbitrary posterior distributions.
 I will make use of the implementation in the package \emph{Sampling through adaptive neighborhoods} (Stan)~\autocite{Stan-2-32-2}.
-Stan implements \emph{automatic differentiation} to calculate the gradient of the posterior density without any user input.
+Stan implements \emph{automatic differentiation}\todo{ref automatic diff} to calculate the gradient of the posterior density without any user input.
 Automatic differentiation uses the chain rule to calculate the gradient of a function from its code by breaking it down into a series of elementary operations, for which the gradient is known.
@@
 Various rules-of-thumb exist for what should be considered an acceptable Rhat.
-Historically, 1.1 has commonly been recommended, although more recent work suggests that 1.01 is more appropriate~\autocite{vehtariRhat}.
+Historically, 1.1 has commonly been recommended, although more recent work suggests that 1.01 or 1.05 is more appropriate~\autocite{vehtariRhat}.
 

### 2023-11-13 SEIR.tex
Updates from Overleaf

--- 2023-11-07 SEIR.tex
@@
 
-\todo[inline]{Add an intro with an overview of what this chapter contains. Include why I discuss mechanistic models, how I will use them, and what they add to the previous analyses.}
+\todo[inline]{Add an intro with an overview of what this chapter contains.
+Include why I discuss mechanistic models, how I will use them, and what they add to the previous analyses.
+Differentiate the transmission model from the observation model.}
 
-Mechanistic models explicitly represent the underlying processes driving the spread of infectious disease.
+Mechanistic models explicitly represent the infection process driving the spread of infectious disease.
 Infectious diseases have inherent feedback loops: current infectious individuals generate future infectious individuals.
@@
 The first sections of this chapter give the background and methodology of the model I use.
-The literature on mechanistic models is vast, going back to \textcite{kermackContribution}.
+The literature on mechanistic models is vast, a major early contribution being \textcite{kermackContribution}.
 Here, I explain only the background necessary to understand the model used in this chapter.
@@
 This section gradually builds complexity into a mechanistic model.
-The mechanistic models considered here are \emph{compartmental} models.
-A compartmental model divides all members of a population into compartments, each representing a disease state.
-Compartmental models are the most common type of mechanistic infectious disease model.
+The class of mechanistic models considered here is \emph{compartmental} models, the most common type of mechanistic infectious disease model.
+A compartmental model divides all members of the population under study into compartments, each representing a disease state.
 They can be \emph{deterministic} or \emph{stochastic}.
@@
 The state vector is the proportion of the population (or number of individuals) in each compartment at time $t$.
-
 Deterministic models have the property that, for any $t$, the state vector $\vec{x}(t)$ is a deterministic function of the parameters $\vec{\theta}$ and the initial state $\vec{x}(0)$.
-The transmission model corresponds to the infection process (see \cref{E-inc-prev:sec:infection-process}).
-There remains randomness in the observation model (described in \cref{SEIR:sec:observation}).
-
+\todo{Maybe say something about still having a stochastic observation model, if not clear enough from intro}
+%The transmission model corresponds to the infection process (see \cref{E-inc-prev:sec:infection-process}).
+%There remains randomness in the observation model (described in \cref{SEIR:sec:observation}).
 The other type of model is \emph{stochastic}.
 Stochastic models specify a probability distribution $p(\vec{x} \mid \vec{\theta})$ for the state vector $\vec{x}(t)$ at time $t$.
-Unlike the models previously encountered in this thesis, there will be dependence between $\vec{x}(t)$ and $\vec{x}(t')$.
-For example, the number of individuals infected on day $t+1$, conditional on the model state on day $t$, might be a Poisson distribution.
-The mean of the distribution is a function of the model state on day $t$.
+Unlike a Poisson process (see\todo{ref intro section}), there is a dependence between $\vec{x}(t)$ and $\vec{x}(t')$.
+Further discussion of stochastic models is beyond the scope of this thesis; see \textcite[chapter 6]{keelingModeling} for an introduction.
 
-I will use a deterministic model mathematically described by a system of Ordinary Differential Equations (ODE).
-The derivative is taken with respect to time, $t$.
-Therefore, in an ODE model, time is necessarily continuous although discretised when solving (see \cref{SEIR:sec:inference-implementation}).
+This chapter focuses on deterministic models mathematically described by a system of ODEs, specifically $d\vec{x}/dt$.
+Therefore, in an ODE model, time is necessarily continuous, although discretised when solving (see \cref{SEIR:sec:inference-implementation}).
 For concreteness, I use units of days throughout, although the model can use any unit of time.
 The number or proportion of individuals in each compartment is also continuous, which is a reasonable approximation when the population is large.
-Equivalent formulations exist for the proportion or number of individuals.
 
@@
 Deterministic models are also known as \emph{mean-field} models because they can be derived from stochastic models by taking the mean of the distribution of the number of infected individuals.
-Further discussion of stochastic models is beyond the scope of this thesis; \textcite[chapter 6]{keelingModeling} provides an introduction to these models.
 
 \subsection{The SIR model} \label{SEIR:sec:SIR}
+
 \begin{figure}[h]
@@
 
-The simplest compartmental model is the \emph{susceptible-infected-recovered} (SIR) model (REF?).
+The simplest compartmental model is the \emph{susceptible-infected-recovered} (SIR) model\todo{ref for SIR being simplest}.
 The name refers to the three compartments in the model, displayed in \cref{SEIR:fig:SIR}.
@@
 Two parameters are present in this model: $\beta$ and $\gamma$.
-$\beta$ is the effective? transmission rate, or the number of individuals an infected individual infects per day.
+$\beta$ is the transmission rate\todo{check what beta is called: effective transmission rate or transmission rate or something else?}, or the number of individuals an infected individual infects per day.
 $\gamma$ is the recovery rate, or the proportion of infected individuals that recover per day; hence, $1/\gamma$ is the mean infectious period.
@@
 This term can be arrived at by considering a single infected individual.
-Assume this individual makes contact with any other individual in the population at a constant rate, $\kappa$, in units of contacts per day.
-Further, assume the population is \emph{well-mixed}, meaning the individual is equally likely to come into contact with any other individual.
-For a randomly selected individual, the probability that they are susceptible is $s(t)$.
+Assume this individual makes $\kappa$ contacts with individuals in the population per day.
+Further, assume the population is \emph{well-mixed}, meaning the each contact is with an individual chosen uniformly at random from the population.
+For such a randomly selected individual, the probability that they are susceptible is $s(t)$.
 Each of the infected individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
@@
 
-\todo[inline]{Possibly add a paragraph about the importance of age and mixing patterns, maybe show a contact matrix?}
+\todo[inline]{Possibly add a paragraph about the importance of age and mixing patterns and show a contact matrix}
 
@@
 To incorporate the stratification into the model, first I introduce some additional notation.
-Denote the set of strata, for example age groups, as $\set{A}$. $N_i$ notates the total population of stratum $a$ and hence $N_i s_i$ is the number of susceptible people in stratum $i$.
+Denote the set of strata, for example age groups, as $\set{A}$. $N_i$ notates the total population of stratum $a$ and hence $N_a s_a$ is the number of susceptible people in stratum $a$.
 
@@
 Considering infected individuals across all strata, $\sum_{j \in \set{A}} K_{ij} \beta_{ij} s_i$ is the instantaneous proportion of individuals in stratum $i$ infected.
-
+Therefore, modify\todo{refer to right equations and add missing bit} with the following:
 \begin{align}
     \frac{d\vec{s}}{dt} = -\vec{s} \circ ((\matr{\beta} \circ \matr{K}) \vec{i})
+    \frac{d\vec{i}_1}{dt} = \vec{s} \circ ((\matr{\beta} \circ \matr{K}) \vec{i}) -
     \label{eq:matrix:theory}
 \end{align}
-where $\circ$ is the element-wise product (\ie: $(\matr{A} \circ \matr{B})_{ij} = A_{ij} B_{ij}$).
+where $\circ$ is the element-wise product (\ie: $(\matr{A} \circ \matr{B})_{ij} = A_{ij} B_{ij}$) and $\vec{i} = \sum_{k=1}^n \vec{i}_k$.
 
@@
 A more complex model would allow for interaction between regions.
-However, for a large epidemic, non-interacting populations is a good approximation ~\autocite{birrellRealtimea}.
+However, for a large epidemic, non-interacting populations is a good approximation~\autocite{birrellRealtimea}.
 
@@
 How the population spent their time each week during the pandemic is taken from the Google mobility data\todo{cite mobility data}, which measured the locations of Android users over the period of interest.
-\todo[inline]{is the above paragraph too hand wavy?}
+\todo[inline]{add a bit more detail to the above}
 

### 2023-11-13 SEIR.tex

@@
 
+I use previously developed estimates of how contact matrices change over time.
+These matrices were produced for \textcite{birrellRealtime}.
+These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
+These use data from the POLYMOD study~\autocite{mossongSocial} and UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
+POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on age and location (home, work, school, leisure, transport, or other) of these contacts.
+UKTUS asked a representative sample of the UK population about how much time they spent doing various activities each day.
+\Textcite{vanleeuwenAugmenting} combined these two datasets to estimate the rate of contacts between age groups during each activity.
+\Textcite{birrellRealtime} then modify these contact rates based on measured changes in time use during the pandemic.
+These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
+Google mobility data measures the location of Android users over time, relative to a pre-pandemic baseline.
+These data sources are then used to down or upweight the number of contacts in each setting, as estimated by \textcite{vanleeuwenAugmenting}.
+For mathematical details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.
+
 Some behavioural changes (\eg mask wearing) and their consequent impacts on transmission are not captured by the contact survey.

### 2023-11-14 SEIR.tex
add in comments

--- 2023-11-13 SEIR.tex
@@
 
+\todo[inline]{double check correct terms for the below, especially rates vs instantaneous proportion and per unit time vs per day}
 The term $\beta s(t)i(t)$ is the rate at which susceptible individuals (as a proportion of the total population size) are infected~\autocite[18]{keelingModeling}.
@@
 
+To relax the assumption of a well-mixed population, the population can be stratified by some characteristic.
+
+\todo[inline]{start with a generic intro on structured populations and use of matrices, then give age as a motivating example, finally specifics go elsehwere}
+
 \todo[inline]{Possibly add a paragraph about the importance of age and mixing patterns, maybe show a contact matrix?}
@@
 Normally, $K_{ij}$ is assumed to be known, for example from a previous contact survey~\autocite[such as][]{mossongSocial}.
-$\beta_{ij}$ is then estimated, although often with a parsimonious model.
+Ideally, all the $\beta_{ij}$ terms would be estimated, but there is normally inadequate data to do so, requiring assumptions to be made to reduce the number of parameters~\autocite[65]{keelingModeling}.
+% $\beta_{ij}$ is then estimated, although often with a parsimonious model\todo{ref / give examples / delete on how age-specfic susceptibility / transmissibility happens}.
 A common assumption is that $\beta_{ij}$ can be decomposed into two independent components such that $\beta_{ij} = \beta \beta^\text{susc}_{i} \beta^\text{inf}_j$.
@@
 Informally, it is how vulnerable to infection individuals in stratum $i$ are.
-Formally, it is the probability that a susceptible individual in stratum $i$ will be infected upon contact with an infectious individual, relative to a individual in the reference stratum.
+Formally, define an $a_0$ and $a'_0$ and enforce $\beta^\text{susc}_{a_0}=1$ for identifiability.
+Then $\beta^\text{susc}_i$ is the relative risk of an individual in stratum $i$ being infected, relative to an individual in stratum $a_0$.
+\todo{clarify notation and explaantion here of reference class}
+it is the probability that a susceptible individual in stratum $i$ will be infected upon contact with an infectious individual, relative to a individual in the reference stratum.
 Similarly, $\beta^\text{inf}_i$ is known as the \emph{relative infectiousness} of stratum $i$.
@@
 
+\todo[inline]{move the below into a context-specific section}
 Based on prior work\todo{cite}, I assume that all individuals have the same \emph{infectiousness} but that there is a difference in \emph{susceptibility}.
@@
 \subsection{Geographic structure} \label{SEIR:sec:geography}
+\todo[inline]{remove this section and just explain what I'm doing in this case in an implementation section}
+
+To allow for regional hetereogeneity....
+This assumes non-interaction between regions...
 
@@
 Between-age-group contacts rates have previously been inferred from a combination of contact surveys, time use data, and mobility data~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}.
-I use the outputs from this analysis as $\matr{M}$.
+I use the outputs from this analysis as $\matr{K}$ in equation\todo{ref where this contact matrix fits in}.
 To summarise, baseline rates and contexts of contacts between age groups are taken from the contact survey POLYMOD~\autocite{mossongSocial}.
@@
 \section{Observation model} \label{SEIR:sec:observation}
+\todo[inline]{Be clear this section is to fit to CIS data}
 
@@
 
+\todo[inline]{need more explanation here: what is a phase-type distribution, be less terse in the following text}
 A $l$-phase EC distribution is a convolution of a $(l-2)$-phase gamma distribution and a two-phase acyclic, arbitrary phase-type distribution.
@@
 In fact, all the $\vec{e}$, $\vec{i}$ and $\vec{p}$ compartments grow at the same initial exponential growth rate, $\psi$.
+\todo{move R definition for each model into where they are introduced}
 If $\vec{s} = 1$ then the basic reproduction number can be linked to the growth rate of this epidemic as follows~\autocites{birrellRealtimea}{wearingAppropriate}:
@@
     Mean infectious period & $1/\gamma$ & 4 days (fixed) & From a paper \\
-    Initial proportion without immunity & $\vec\pi$ & See \cref{SEIR:table:immunity} & \\
+    Initial proportion without immunity & $\vec\pi$ & See \cref{SEIR:table:immunity-prior} & \\
     Initial proportion infected & $i^+$ & $\BetaDist(0.5, 1000)$ & Weakly informative \\
@@
 \label{SEIR:table:priors}
+\end{table}
+
+\begin{table}[]
+\centering
+\begin{tabular}{l|lllll}
+         & $[0,16)$ & $[16,25)$ & $[25,50)$ & $[50,70)$ & $[70,\infty)$ \\
+        \hline
+        East Midlands & $\BetaDist(8.5, 301)$ & $\BetaDist(18, 256)$ & $\BetaDist(19, 475)$ & $\BetaDist(20, 493)$ & $\BetaDist(5, 337)$ \\
+        East of England & $\BetaDist(7.9, 236)$ & $\BetaDist(21, 248)$ & $\BetaDist(35, 677)$ & $\BetaDist(26, 613)$ & $\BetaDist(5.6, 332)$ \\
+        London & $\BetaDist(9.7, 131)$ & $\BetaDist(50, 328)$ & $\BetaDist(154, 1331)$ & $\BetaDist(106, 990)$ & $\BetaDist(7.6, 204)$ \\
+        North East & $\BetaDist(6.1, 180)$ & $\BetaDist(12, 145)$ & $\BetaDist(14, 274)$ & $\BetaDist(14, 295)$ & $\BetaDist(4.2, 240)$ \\
+        North West & $\BetaDist(12, 252)$ & $\BetaDist(38, 353)$ & $\BetaDist(72, 885)$ & $\BetaDist(44, 695)$ & $\BetaDist(6.3, 264)$ \\
+        South East & $\BetaDist(8.9, 404)$ & $\BetaDist(17, 319)$ & $\BetaDist(20, 577)$ & $\BetaDist(19, 586)$ & $\BetaDist(4.4, 356)$ \\
+        South West & $\BetaDist(6.9, 415)$ & $\BetaDist(11, 277)$ & $\BetaDist(12, 556)$ & $\BetaDist(11, 559)$ & $\BetaDist(4, 492)$ \\
+        West Midlands & $\BetaDist(8, 167)$ & $\BetaDist(19, 151)$ & $\BetaDist(32, 475)$ & $\BetaDist(33, 504)$ & $\BetaDist(5.8, 241)$ \\
+        Yorkshire and The Humber & $\BetaDist(9.2, 314)$ & $\BetaDist(19, 242)$ & $\BetaDist(27, 589)$ & $\BetaDist(18, 474)$ & $\BetaDist(4.9, 316)$ \\
+    \end{tabular}
+\caption{Priors for $\vec{\pi}$, the proportion of the population without prior immunity, in each age group and region combination.
+Based on seroprevalence in blood donors~\autocite{amirthalingamSeroprevalence} and a representative sample of children~\autocite{ratcliffeCommunity} in August 2020.}
+\label{SEIR:table:immunity-prior}
 \end{table}

### 2023-11-14 SEIR.tex

@@
 
-\todo[inline]{Add an intro with an overview of what this chapter contains. Include why I discuss mechanistic models, how I will use them, and what they add to the previous analyses.}
+\todo[inline]{Add an intro with an overview of what this chapter contains.
+Include why I discuss mechanistic models, how I will use them, and what they add to the previous analyses.
+Differentiate the transmission model from the observation model.}
 
-Mechanistic models explicitly represent the underlying processes driving the spread of infectious disease.
+Mechanistic models explicitly represent the infection process driving the spread of infectious disease.
 Infectious diseases have inherent feedback loops: current infectious individuals generate future infectious individuals.
-Explicitly representing this system can generate insights not possible with the statistical methods used in the previous chapter.
+Explicitly representing this system can generate insights not possible with the statistical methods used in the previous chapter\todo{give explicit examples of advantages}.
 
 The first sections of this chapter give the background and methodology of the model I use.
-The literature on mechanistic models is vast, going back to \textcite{kermackContribution}.
+The literature on mechanistic models is vast, a major early contribution being \textcite{kermackContribution}.
 Here, I explain only the background necessary to understand the model used in this chapter.
@@
 
+\todo[inline]{integrate this paragraph into the intro part}
 This section gradually builds complexity into a mechanistic model.
-The mechanistic models considered here are \emph{compartmental} models.
-A compartmental model divides all members of a population into compartments, each representing a disease state.
-Compartmental models are the most common type of mechanistic infectious disease model.
+The class of mechanistic models considered here is \emph{compartmental} models, the most common type of mechanistic infectious disease model.
+A compartmental model divides all members of the population under study into compartments, each representing a disease state.
 They can be \emph{deterministic} or \emph{stochastic}.
@@
 
-Some behavioural changes (\eg mask wearing) and their consequent impacts on transmission are not captured by the contact survey.
+Some behavioural changes (\eg mask wearing) impact transmission.
+However, these are not captured by the contact survey.
 To allow for these changes, $\beta$ is allowed to vary over time.
-In particular, $\beta$ could? follows a random walk in log space on the weekly timescale:
+In particular, $\beta$ follows a random walk in log space on the weekly timescale:
 \begin{align}

### 2023-11-14 SEIR.tex
address comments

--- 2023-11-13 SEIR.tex
@@
 Infectious diseases have inherent feedback loops: current infectious individuals generate future infectious individuals.
-Explicitly representing this system can generate insights not possible with the statistical methods used in the previous chapter.
+Explicitly representing this system can generate insights not possible with the statistical methods used in the previous chapter\todo{give explicit examples of advantages}.
 
@@
 
+\todo[inline]{integrate this paragraph into the intro part}
 This section gradually builds complexity into a mechanistic model.

### 2023-11-15 SEIR.tex
restructure SEIR chapter

--- 2023-11-14 SEIR.tex
@@
 
-\chapter{Mechanistic modelling of incidence and transmission} \label{SEIR}
+\chapter{Estimating transmission using mechanistic modelling} \label{SEIR}
 
-\todo[inline]{Add an intro with an overview of what this chapter contains.
-Include why I discuss mechanistic models, how I will use them, and what they add to the previous analyses.
-Differentiate the transmission model from the observation model.}
+This chapter uses mechanistic models for estimating the transmission of SARS-CoV-2.
+This is an alternative approach to the statistical methods used in \cref{E-backcalc}.
+Mechanistic models explicitly represent the infection process driving the spread of infectious disease.
+The explicit representation means that a mechanistic model's state and parameters have biological and/or epidemiological interpretations; these interpretations provide scientific understanding of the disease.
+Of particular interest is that the basic reproduction number, $R_0$, and the time-varying reproduction number, $R_t$, can both be calculated as simple functions of the model parameters; in \cref{E-backcalc} additional modelling is required to calculate these numbers.
+Furthermore, mechanistic models can be used for scenario-based modelling, whereby the effect of different interventions can be simulated.
+All of these properties make mechanistic models a useful tool for understanding and controlling infectious diseases.
+However, they make stronger assumptions about the disease than statistical models, and often perform worse in short-term forecasting\todo{add a citation}.
 
-Mechanistic models explicitly represent the infection process driving the spread of infectious disease.
-Infectious diseases have inherent feedback loops: current infectious individuals generate future infectious individuals.
-Explicitly representing this system can generate insights not possible with the statistical methods used in the previous chapter\todo{give explicit examples of advantages}.
+\Cref{SEIR:sec:transmission-generic} introduces a generic framework for mechanistic modelling of infectious diseases.
+% This framework is then applied to SARS-CoV-2 in \cref{SEIR:sec:transmission-application}.
+It describes the \emph{transmission model}, corresponding to the infection process (see \cref{E-inc-prev:sec:infection-process}).
+The transmission model needs to be linked to observations; in this chapter, that will be the CIS prevalence data.
+The \emph{observation model}, introduced in \cref{SEIR:sec:observation}, is this link; it incorporates both the prevalence process (see \cref{E-inc-prev:sec:prevalence-process}) and the observation process (see \cref{E-inc-prev:sec:observation-process}).
+\Cref{SEIR:sec:inference-implementation} describes the inference process, parameterisation, and priors for this model.
+I ensure that the model can recover the parameters successfully in a simulation study in \cref{SEIR:sec:sim-study} before applying it to the CIS data in \cref{SEIR:sec:application}, which includes comparisons to the results in \cref{E-backcalc}.
+Finally, \cref{SEIR:sec:discussion} discusses the results, limitations, and extensions of this work.
 
-The first sections of this chapter give the background and methodology of the model I use.
-The literature on mechanistic models is vast, a major early contribution being \textcite{kermackContribution}.
-Here, I explain only the background necessary to understand the model used in this chapter.
+The literature on mechanistic models is vast, having been developed for over a century; the model in \cref{SEIR:sec:SIR} was first formulated in the 1920s~\autocite{kermackContribution}.
+This chapter explains only the background I will use in the application to SARS-CoV-2.
 For further details I recommend the tutorial paper \textcite{kretzschmarMathematical} or the textbook \textcite{keelingModeling}.
 
-Notation in this chapter differs from the rest of the thesis.
+Notation in this chapter differs from the rest of this thesis.
 I adopt conventions and notation from the infectious disease modelling field, although this is not consistent in the literature.
@@
 
-\section{ODE-based compartmental transmission models} \label{SEIR:sec:transmission}
+\section{Transmission models} \label{SEIR:sec:transmission-generic}
 
-\todo[inline]{integrate this paragraph into the intro part}
-This section gradually builds complexity into a mechanistic model.
-The class of mechanistic models considered here is \emph{compartmental} models, the most common type of mechanistic infectious disease model.
-A compartmental model divides all members of the population under study into compartments, each representing a disease state.
-They can be \emph{deterministic} or \emph{stochastic}.
+This section introduces the class of mechanistic transmission models known as \emph{compartmental} models.
+Compartmental models are the most widely-used class of mechanistic infectious disease model.
+A compartmental model is named because they place every member of the population under study into a compartment; each compartment represents a disease state (\eg susceptible).
 
 A compartmental model consists of a state vector $\vec{x}(t)$ and a set of parameters $\vec{\theta}$.
-The state vector is the proportion of the population (or number of individuals) in each compartment at time $t$.
-Deterministic models have the property that, for any $t$, the state vector $\vec{x}(t)$ is a deterministic function of the parameters $\vec{\theta}$ and the initial state $\vec{x}(0)$.
-\todo{Maybe say something about still having a stochastic observation model, if not clear enough from intro}
+The elements of $\vec{x}(t)$ are the proportion of the population (or number of individuals) in each compartment at time $t$.
+The parameters $\vec{\theta}$ are constants that determine, along with the model structure, how the state vector changes over time.
+Compartmental models can be \emph{deterministic} or \emph{stochastic}.
+Deterministic models are defined by the state vector $\vec{x}(t)$ being a deterministic, although rarely analytically-tractable, function of the parameters $\vec{\theta}$ and the initial state $\vec{x}(0)$.
 %The transmission model corresponds to the infection process (see \cref{E-inc-prev:sec:infection-process}).
 %There remains randomness in the observation model (described in \cref{SEIR:sec:observation}).
-The other type of model is \emph{stochastic}.
 Stochastic models specify a probability distribution $p(\vec{x} \mid \vec{\theta})$ for the state vector $\vec{x}(t)$ at time $t$.
-Unlike a Poisson process (see\todo{ref intro section}), there is a dependence between $\vec{x}(t)$ and $\vec{x}(t')$.
-Further discussion of stochastic models is beyond the scope of this thesis; see \textcite[chapter 6]{keelingModeling} for an introduction.
+In general, this means, there is a dependence between $\vec{x}(t)$ and $\vec{x}(t')$.
+This relaxes the assumption of independent increments in the counting process formulation of the infection process (see \cref{E-inc-prev:sec:infection-process}).
 
-This chapter focuses on deterministic models mathematically described by a system of ODEs, specifically $d\vec{x}/dt$.
+Deterministic models are also known as \emph{mean-field} models because they can are the behaviour corresponding to a stochastic model's mean.
+When the number of infected individuals is large, then this mean-field behaviour is a good approximation of the fully stochastic model, due to the law of large numbers\todo{ref mean-field and approximation error}.
+An alternative justification, when making noisy observations of the system, is that the additional noise from the stochastic model is negligible compared to this observation noise.
+This argument is similar to the justification for neglecting the stochasticity of the prevalence process in \cref{E-inc-prev:sec:observation-process}.
+The remainder of this chapter uses only deterministic models, because throughout the time period this thesis considers there were enough infectious individuals.
+For an introduction to stochastic models, see \textcite[chapter 6]{keelingModeling}.
+% \todo{Maybe say something about still having a stochastic observation model, if not clear enough from intro}
+
+In particular, this chapter discusses deterministic models whose evolution is specified by a system of ODEs, $d\vec{x}/dt$.
 Therefore, in an ODE model, time is necessarily continuous, although discretised when solving (see \cref{SEIR:sec:inference-implementation}).
@@
 
-When the number of infected individuals is large, then deterministic models are a good approximation to stochastic models.
-Deterministic models are also known as \emph{mean-field} models because they can be derived from stochastic models by taking the mean of the distribution of the number of infected individuals.
+\Cref{SEIR:sec:SIR} begins by introducing the \emph{susceptible-infected-recovered} (SIR) model, a simple compartmental model for a disease which confers immunity.
+The SIR model is named because each individual in the population is divided into one of three states: susceptible, infectious, or recovered.
+This model has several assumptions which make it inappropriate to model SARS-CoV-2; therefore, I introduce extensions to relax these assumptions.
+First, \cref{SEIR:sec:non-exponential} relaxes the assumption of an exponentially-distributed infectious period to a gamma distribution.
+Then, \cref{SEIR:sec:SEIR} relaxes the assumption that individuals are infectious immediately after they are infected; it does this by extending the SIR model to the SEIR model, including an \emph{exposed} compartment.
+Next, \cref{SEIR:sec:structured-populations} relaxes the homogeneous population assumption, introducing stratification by important characteristic(s).
+Finally, \cref{SEIR:sec:time-varying-foi} relaxes the assumption of a constant transmission rate to allow for time-varying behavioural changes.
 
@@
 
-The simplest compartmental model is the \emph{susceptible-infected-recovered} (SIR) model\todo{ref for SIR being simplest}.
+The simplest compartmental model for a disease which confers immunity is the (SIR) model\todo{ref for SIR being simplest}.
 The name refers to the three compartments in the model, displayed in \cref{SEIR:fig:SIR}.
-Individuals in the susceptible compartment can be infected.
-Individuals in the infected compartment can infect others, and eventually recover.
-Individuals in the recovered compartment are immune to the disease, and cannot be infected again.
+Individuals in the susceptible compartment can be infected; $s(t)$ is the proportion of the population that is susceptible at time $t$.
+Individuals in the infectious compartment can infect others, and eventually recover; $i(t)$ is the proportion of the population that is infectious at time $t$.
+Individuals in the recovered compartment are immune to the disease, and cannot be infected again; $t(t)$ is the proportion of the population that have recovered by time $t$.
+The model's state is $\vec{x}(t) = (s(t), i(t), r(t))^T$.
 
@@
 \begin{align}
-\frac{ds(t)}{dt} &= -\beta s(t)i \\
-\frac{di(t)}{dt} &= \beta s(t)i(t) - \gamma i(t) \\
+\frac{ds(t)}{dt} &= -\lambda(t) s(t) \\
+\frac{di(t)}{dt} &= \lambda(t) s(t) - \gamma i(t) \\
 \frac{dr(t)}{dt} &= \gamma i(t)
 \end{align}
-$s(t)$, $i(t)$ and $r(t)$ are the proportions of the population in the susceptible, infected and recovered compartments respectively.
-These are the state of the system.
-The state is a function of time, $t$, although this is not shown explicitly.
-Two parameters are present in this model: $\beta$ and $\gamma$.
-$\beta$ is the transmission rate\todo{check what beta is called: effective transmission rate or transmission rate or something else?}, or the number of individuals an infected individual infects per day.
-$\gamma$ is the recovery rate, or the proportion of infected individuals that recover per day; hence, $1/\gamma$ is the mean infectious period.
-These parameters are assumed to be constant over time (this assumption is relaxed in \cref{SEIR:sec:time-varying-foi}).
+where $\lambda(t) = \beta i(t)$, is the \emph{force of infection}: the risk of infection for a susceptible individual at time $t$~\autocite[17]{keelingModeling}.
+Two parameters are present in this model: $\beta$ and $\gamma$, \ie $\vec{\theta} = (\beta, \gamma)^T$.
+$\beta$ is the transmission rate, absorbing several terms as described below.
+%\todo{check what beta is called: effective transmission rate or transmission rate or something else?}, or the number of individuals an infected individual infects per day.
+$\gamma$ is the recovery rate, or the proportion of infected individuals that recover per day; hence, $1/\gamma$ is the mean infectious period~\autocite[367]{keelingModeling}.
+These parameters are assumed to be constant (an assumption relaxed in \cref{SEIR:sec:time-varying-foi}).
 
-\todo[inline]{double check correct terms for the below, especially rates vs instantaneous proportion and per unit time vs per day}
 The term $\beta s(t)i(t)$ is the rate at which susceptible individuals (as a proportion of the total population size) are infected~\autocite[18]{keelingModeling}.
 Susceptible individuals move from the $s(t)$ to $i(t)$ when they are infected.
-This term can be arrived at by considering a single infected individual.
-Assume this individual makes $\kappa$ contacts with individuals in the population per day.
-Further, assume the population is \emph{well-mixed}, meaning the each contact is with an individual chosen uniformly at random from the population.
+This term can be arrived at by considering a single infected individual~\autocite[214]{kretzschmarMathematical}.
+Assume this individual comes into contact with other individuals in the population at a rate of $\kappa$ per day.
+Further, assume the population is \emph{well-mixed}, meaning that each contact is with an individual chosen uniformly at random from the population.
 For such a randomly selected individual, the probability that they are susceptible is $s(t)$.
@@
 Therefore, in expectation, the infected individual makes contact with susceptible individuals at a rate of $\kappa s(t)$.
-Assume that there is a constant probability of transmission upon contact between an infected and susceptible individual, $p_{si}$.
-The expected rate of infections caused by this infected individual is therefore $\kappa s p_{si}$.
-The two constant terms here (number of contacts and probability of transmission) are absorbed into the transmission rate, $\beta = \kappa p_{si}$.
-Therefore, the instantaneous number of infections per infected individual is $\beta s(t)$.
-The overall proportion of the population instantaneously infected is $\beta s(t)i(t)$.
+Assume that there is a constant probability of transmission upon contact between an infected and susceptible individual, $q$.
+Combined with the above, the expected rate of infections caused by this infected individual is $\kappa s q$.
+The two constant terms here (number of contacts and probability of transmission) are absorbed into the transmission rate, $\beta = \kappa q$.
+Therefore, the rate of new infections generated by a single infectious individual is $\beta s(t)$.
+Multiplying by the number of infected individuals and dividing by the population size gives the rate at which susceptible individuals are infected as a proportion of the population, $\beta s(t) i(t)$.
 
-The term $\gamma i$, the rate at which infected individuals (as a proportion of the total population size) recover, follows directly from the definitions of $\gamma$ and $i$.
+The term $\gamma i(t)$, the rate at which infected individuals (as a proportion of the total population size) recover, follows directly from the definitions of $\gamma$ and $i(t)$.
 
-Other quantities of this model are of interest.
-The \emph{force of infection}, $\lambda(t) = \beta i(t)$, is the instantaneous risk of infection for a susceptible individual~\autocite[17]{keelingModeling}.
-The basic reproduction number (discussed previously in\todo{ref intro}) for this model is $R_0 = \beta / \gamma$~\autocite[20]{keelingModeling}.
+A few properties and assumptions of the SIR model are worth noting.
+\begin{itemize}
+    \item The basic reproduction number (discussed previously in\todo{ref intro}) for this model is $R_0 = \beta / \gamma$~\autocite[20]{keelingModeling}.
     Intuitively, this can be thought of as the average number of infections caused by an infectious individual per day, $\beta$, multiplied by the average number of days they are infectious for, $1/\gamma$.
-
-A few properties of the SIR model are worth noting.
-First, the population is closed.
+    \item The population is closed.
     That is, the total population size is constant, and no infections are imported from outside the population.
     Births and deaths of individuals are also excluded, although they can easily be incorporated in situations where they are relevant~\autocites[26]{keelingModeling}[214]{kretzschmarMathematical}.
-Second, there is no analytical solution to the system of ODEs~\autocite[25]{keelingModeling}: numerical methods are required to solve the system.
-Third, in addition to the parameters, the state of the system at some time (normally the initial state, the state at time $t=0$) is required to solve the system.
+    \item There is no analytical solution to the system of ODEs~\autocite[25]{keelingModeling}: numerical methods are required to solve the system.
+    \item In addition to the parameters, the state of the system at some time (normally the initial state, the state at time $t=0$) is required to solve the system.
+    \item Immunity is assumed to last until the end of the modelled period.
+    For many diseases, including SARS-CoV-2, this is a reasonable assumption as long as the modelled period is short, as it will be in this chapter\todo{ref that SARS-CoV-2 immunity is long enough to not be an issue}.
+    For modelling longer time periods, waning immunity can be incorporated\todo{ref waning immunity models}.
+\end{itemize}
 
@@
 The basic SIR model implies that the distribution of the time spent in the infected compartment is exponential~\autocite[96]{keelingModeling}.
-This is an unrealistic assumption and leads to underestimation of the reproduction number~\autocites{lloydRealistic}{wearingAppropriate}.
-A standard extension to the SIR model is to allow the infectious period to follow a gamma distribution.
+This is an unrealistic assumption for most diseases and leads to an underestimation of the reproduction number if the model is fit to data~\autocites{lloydRealistic}{wearingAppropriate}.
+A standard extension to the SIR model is to allow the infectious period to follow a gamma distribution~\autocite[94]{keelingModeling}.
 This is done by adding multiple infected compartments (see \cref{SEIR:fig:SIR-gamma}).
-The system can now be written as follows~\autocite[94]{keelingModeling}:
+The system can now be written as follows:
 \begin{align}
-\frac{ds}{dt} &= -\beta si \\
-\frac{di_1}{dt} &= \beta si - n\gamma i_1 \\
-\frac{di_2}{dt} &= n\gamma i_1 - n \gamma i_2 \\
+\frac{ds(t)}{dt} &= -\lambda(t) s(t)\\
+\frac{di_1(t)}{dt} &= \lambda(t) s(t) - n\gamma i_1(t) \\
+\frac{di_2(t)}{dt} &= n\gamma i_1(t) - n \gamma i_2(t) \\
 &\vdots \nonumber \\
-\frac{di_n}{dt} &= n\gamma i_{n-1} - n \gamma i_n \\
-\frac{dr}{dt} &= n\gamma i_n
+\frac{di_n(t)}{dt} &= n\gamma i_{n-1}(t) - n \gamma i_n(t) \\
+\frac{dr(t)}{dt} &= n\gamma i_n(t)
 \end{align}
-where $n$ is the number of infected compartments.
-The infectious period is now distributed $\GamDist(n, n\gamma)$.
-Gamma distributions are a realistic model for a variety of diseases~\autocite{wearingAppropriate}.
+where $n$ is the number of infected compartments, and redefining $\lambda(t) = \beta \sum_{j=1}^n i_j(t)$ 
+In this model the infectious period is distributed $\GamDist(n, n\gamma)$, a realistic model for a variety of diseases~\autocite{wearingAppropriate}.
+This parameterisation is chosen to give $1/\gamma$ as the mean infectious period (the gamma distribution and its expectation is defined in \cref{E-distributions}).
+% (it is slightly unclear what it means for a variable to have a distribution in a deterministic model, the clearest explanation is that this is the distribution which the ).
 A gamma distribution with the first parameter equal to 1 is an exponential distribution, connecting this model to the basic SIR model in \cref{SEIR:sec:SIR}.
 The distribution for the time spent in the infected compartments tends to a constant (\ie the same for all individuals) as $n \to \infty$~\autocite{lloydRealistic}.
-The form of these equations is chosen so that $1/\gamma$ remains the mean infectious period.
+The reproduction number in terms of the parameters is unchanged by this extension, \ie $R_0 = \beta / \gamma$.
 
-\subsection{The SEIR model}
+\subsection{The SEIR model} \label{SEIR:sec:SEIR}
+
 \begin{figure}[h]
@@
 The SIR model assumes that individuals are immediately infectious upon infection.
-This is unrealistic, as there is a period between infection and infectiousness, known as the \emph{latent period}.
+This is unrealistic for most diseases, as there is a period between an individual being infected and them being infectious, known as the \emph{latent period}.
+\todo{perhaps link with where define phases of infection in background chapters}
 Ignoring the latent period will again overestimate the reproduction number~\autocite{wearingAppropriate}.
 
-The SIR model can be extended to include a latent period by adding a latent compartment.
+The SIR model can be extended to include a latent period by adding a latent compartment~\autocite[41]{keelingModeling}.
 This is also known as an \emph{exposed} compartment.
@@
 where $m$ is the number of latent compartments and $1/\sigma$ is the mean latent period.
-The latent period is now distributed $\GamDist(m, m\sigma)$.
+The latent period is distributed $\GamDist(m, m\sigma)$.
 The infectious period is still distributed $\GamDist(n, n\gamma)$.
+
+The reproduction number in terms of the parameters is unchanged by this extension, \ie $R_0 = \beta / \gamma$, however, the growth rate in the exponential phase will be slower than before~\autocite[41]{keelingModeling}.
 
@@
 
-To relax the assumption of a well-mixed population, the population can be stratified by some characteristic.
+To relax the assumption of a homogeneous population, the population can be stratified.
+If the strata have differing transmission patterns, this can change the dynamics of the epidemic.
+The different patterns could be due to behavioural factors (\eg high-numbers of sexual partners in a sexually-transmitted disease~\autocite[69]{keelingModeling} or younger age groups making many more contacts for respiratory diseases~\autocite[176]{andersonInfectious}) or biological factors (\eg genetics reducing the susceptibility of some individuals~\autocite[208]{andersonInfectious}).
+I will assume that the disease progression (\eg length of the exposed and infectious periods) is the same for all strata, although relaxing this assumption is easy.
+Stratification requires two major changes model changes: extension of the state vector and describing the between-strata interactions.
+
+To incorporate the stratification into the model, first I introduce some additional notation.
+Denote the set of strata of size $A$, as $\set{A}$.
+A stratum will be denoted by $a \in \set{A}$ whose total population is $N_a$.
+
+First, the state vector is now a vector of vectors, one for each of the compartments.
+For example, in the SEIR model of \cref{SEIR:sec:SEIR}, $\vec{x} = (\vec{s}, \vec{e_1}, \dots, \vec{e_m}, \vec{i_1}, \dots, \vec{i_n}, \vec{r})^T$.
+The $a$th element of each vector is the proportion of strata $a$ in that compartment.
+For example, $\vec{s} = (s_1, \dots s_A)^T$ where $s_a$ is the proportion of the $a$th strata that are susceptible.
+Hence, $N_a s_a$ is the number of susceptible people in stratum $a$.
+These vectors simply replace the scalars in the previous equations.
+Various other definitions are possible and equivalent, for example defining $s_a$ as the proportion of the total population that are both in strata $a$ and susceptible or parameterised using numbers of individuals\todo{cite Keeling}.
+
+Second, the interactions between strata must be described.
+This is done by replacing the scalar $\beta$ with a \emph{Who Acquires Infection From Whom} (WAIFW) matrix, $\matr{\beta}$.
+The force of infection is now a vector, $\vec{\lambda} = \matr{\beta} \sum_{j=1}^n \vec{i_j}$.
+The matrix element $\beta_{aa'}$ is the rate at which individuals in stratum $a$ infect individuals in stratum $a'$.
+In practical situations, estimating all of these parameters is infeasible, so some assumptions must be made to reduce the number of parameters.
+Common assumption include assuming symmetry, \ie $\beta_{aa'} = \beta{a'a}$ or a decomposable independence, \ie $\beta_{aa'} = \beta \beta^\text{inf}_a \beta^\text{susc}_{a'}$~\todo{ref A+M on these parameterisations}.
+Alternatively, as in the scalar case, $\beta_{aa'}$ can be decomposed into the contact rate between strata $a$ and $a'$, $\kappa_{aa'}$ and the probability of transmission upon contact, $q_{aa'}$, and then assumptions made about the components.
 
@@
 
-To incorporate the stratification into the model, first I introduce some additional notation.
-Denote the set of strata, for example age groups, as $\set{A}$. $N_i$ notates the total population of stratum $a$ and hence $N_a s_a$ is the number of susceptible people in stratum $a$.
-
-I will derive a matrix which can be used to replace the scalar $\beta$ in the single stratum case.
 The derivation is similar to the single stratum case, except that the contact rates and probabilities of transmission will be allowed to vary for each pair of strata.
@@
 
+\todo[inline]{this paragraph probably just goes and its obvious from context}
 Each element of the vector $\vec{s}$ refers to the proportion of that region's population of the relevant age group that are susceptible.
@@
 
-\subsection{Time-varying behaviour} \label{SEIR:sec:time-varying-foi}
+% \section{A transmission model for SARS-Cov-2} \label{SEIR:sec:transmission-specific}
 
-During the pandemic, behaviour clearly changed over time.
-This was in response to government policy, an individual's perceived risk, and a variety of other factors.
-I follow \textcite{birrellRealtime}, using a combination of Google mobility data and a time-varying transmission intensity to include these changes.
-
-Between-age-group contacts rates have previously been inferred from a combination of contact surveys, time use data, and mobility data~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}.
-I use the outputs from this analysis as $\matr{K}$ in equation\todo{ref where this contact matrix fits in}.
-To summarise, baseline rates and contexts of contacts between age groups are taken from the contact survey POLYMOD~\autocite{mossongSocial}.
-How these contacts relate to time use is taken from the UK Time Use Survey\todo{cite time use survey}.
-How the population spent their time each week during the pandemic is taken from the Google mobility data\todo{cite mobility data}, which measured the locations of Android users over the period of interest.
-\todo[inline]{add a bit more detail to the above}
-
-I use previously developed estimates of how contact matrices change over time.
-These matrices were produced for \textcite{birrellRealtime}.
-These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
-These use data from the POLYMOD study~\autocite{mossongSocial} and UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
-POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on age and location (home, work, school, leisure, transport, or other) of these contacts.
-UKTUS asked a representative sample of the UK population about how much time they spent doing various activities each day.
-\Textcite{vanleeuwenAugmenting} combined these two datasets to estimate the rate of contacts between age groups during each activity.
-\Textcite{birrellRealtime} then modify these contact rates based on measured changes in time use during the pandemic.
-These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
-Google mobility data measures the location of Android users over time, relative to a pre-pandemic baseline.
-These data sources are then used to down or upweight the number of contacts in each setting, as estimated by \textcite{vanleeuwenAugmenting}.
-For mathematical details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.
-
-Some behavioural changes (\eg mask wearing) impact transmission.
-However, these are not captured by the contact survey.
-To allow for these changes, $\beta$ is allowed to vary over time.
-In particular, $\beta$ follows a random walk in log space on the weekly timescale:
-\begin{align}
-    \beta(t) &= \beta_{w(t)} \\
-    \log\beta_{w+1} &= \log\beta_w + \sigma_\epsilon \epsilon_w \\
-    \epsilon_w &\sim \N(0, 1)
-\end{align}
-where $w(t)$ is the week number of time $t$, with week 1 being the first week modelled.
-Two parameters are unspecified here: the initial random of the random walk, $\beta_1$, and the standard deviation of the random walk, $\sigma_\epsilon$.
-See \cref{perf-test:sec:parameters-priors} for the implementation and parameterisation of these.
+% I follow \textcite{birrellRealtimea} in parameterising $\matr{\beta}$, which includes introducing a time-dependency.
 
@@
 
-\section{Full model} \label{SEIR:sec:full-model}
+\section{Inference and implementation} \label{SEIR:sec:inference-implementation}
+
+\todo[inline]{mention that the model is run independently for each region, as in the previous chapter}
+\todo[inline]{maybe put something about the likelihood here}
+
+\subsection{MCMC} \label{SEIR:sec:MCMC}
+
+I use an adaptive random-walk Metropolis-Hastings algorithm (as introduced in \cref{E-inc-prev:sec:MCMC}).
+The proposal distribution is a multivariate normal that adapts to the covariance of the posterior.
+The adaption algorithm is based on \textcite[algorithm 4]{andrieuTutorial}, as implemented in \textcite{ghoshApproximate}.
+Proposals are accepted or rejected using the standard Metropolis-Hastings (MH) acceptance probability.
+See \cref{SEIR:MCMC-algorithm} for full details.
+\begin{algorithm}
+ set $\vec{X_0}$ to an initial value of the parameter vector \;
+ $\vec{\mu_0} = \vec{X_0}$ \;
+ $\matr{\Sigma_0} = \text{diag}(\vec{\mu_0})$ \;
+ $\lambda_0 = 1$ \;
+ \For{$i = 1, \dots, M$}{
+  sample $\vec{Y_}i \sim \N(\vec{\mu_{i-1}}, \lambda_{i-1}\matr{\Sigma_{i-1}})$\;
+  set $\vec{X_i}$ to $\vec{Y_i}$ or $\vec{X_{i-1}}$ using a MH acceptance step\;
+  update the proportion of proposals accepted so far, $\alpha$ \;
+  \eIf(\tcc{No adaptation for 200 iterations}){$i \leq 200$}{
+    $\lambda_i = \lambda_{i-1}$ \;
+    $\vec{\mu_i} = \vec{\mu_{i-1}}$ \;
+    $\matr{\Sigma_i} = \matr{\Sigma_{i-1}}$ \;
+   }{
+    $\gamma_i = (i - 200)^{-0.6}$ \;
+    $\log \lambda_i = \log \lambda_{i-1} + \gamma_i(\alpha - 0.234)$ \;
+    $\vec{\mu_i} = (1 - \gamma_i) \vec{\mu_{i-1}} + \gamma_i \vec{X_i}$ \;
+    $\matr{\Sigma_i} = (1 - \gamma_i) \matr{\Sigma_{i-1}} + \gamma_i (\vec{X_i} - \vec{\mu_i})(\vec{X_i} - \vec{\mu_i})^T$ \;
+  }
+ }
+ \caption{Algorithm for adaptive random-walk Metropolis-Hastings.
+$\vec{\mu_i}$ and $\matr{\Sigma_i}$ are an estimate of the mean and covariance of the posterior distribution using information up to iteration $i$.
+$\text{diag}(\vec{\mu_0})$ is the diagonal matrix with diagonal entries equal to $\vec{\mu_0}$.
+$\lambda_i$ is the scale parameter of the proposal distribution at iteration $i$, tuned to try and ensure an optimal proportion of proposals are accepted (23.4\%).
+$\gamma_i$ is the learning rate, which determines how much adaptation occurs.
+$\gamma_i \to 0$ as $i \to \infty$ so the rate of adaptation is \emph{vanishing}.
+Vanishing adaptation is required for the algorithm to converge to the target distribution~\autocite[section 3]{andrieuTutorial}.}
+ \label{SEIR:MCMC-algorithm}
+\end{algorithm}
+
+\subsection{Parameterisation}
+
+\subsubsection{WAIFW matrix}
+
+\todo[inline]{define the strata in this section, and edit the below text to make sense in this new context}
+
+During the pandemic, behaviour clearly changed over time.
+This was in response to government policy, an individual's perceived risk, and a variety of other factors.
+I follow \textcite{birrellRealtime}, using a combination of Google mobility data and a time-varying transmission intensity to include these changes.
+I follow \textcite{birrellRealtimea}'s parameterisation of $\matr{\beta}$, which includes introducing a time-dependency.
+
+Between-age-group contacts rates have previously been inferred from a combination of contact surveys, time use data, and mobility data~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}.
+I use the outputs from this analysis as $\matr{K}$ in equation\todo{ref where this contact matrix fits in}.
+To summarise, baseline rates and contexts of contacts between age groups are taken from the contact survey POLYMOD~\autocite{mossongSocial}.
+How these contacts relate to time use is taken from the UK Time Use Survey\todo{cite time use survey}.
+How the population spent their time each week during the pandemic is taken from the Google mobility data\todo{cite mobility data}, which measured the locations of Android users over the period of interest.
+\todo[inline]{add a bit more detail to the above}
+
+I use previously developed estimates of how contact matrices change over time.
+These matrices were produced for \textcite{birrellRealtime}.
+These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
+These use data from the POLYMOD study~\autocite{mossongSocial} and UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
+POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on age and location (home, work, school, leisure, transport, or other) of these contacts.
+UKTUS asked a representative sample of the UK population about how much time they spent doing various activities each day.
+\Textcite{vanleeuwenAugmenting} combined these two datasets to estimate the rate of contacts between age groups during each activity.
+\Textcite{birrellRealtime} then modify these contact rates based on measured changes in time use during the pandemic.
+These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
+Google mobility data measures the location of Android users over time, relative to a pre-pandemic baseline.
+These data sources are then used to down or upweight the number of contacts in each setting, as estimated by \textcite{vanleeuwenAugmenting}.
+For mathematical details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.
+
+Some behavioural changes (\eg mask wearing) impact transmission.
+However, these are not captured by the contact survey.
+To allow for these changes, $\beta$ is allowed to vary over time.
+In particular, $\beta$ follows a random walk in log space on the weekly timescale:
+\begin{align}
+    \beta(t) &= \beta_{w(t)} \\
+    \log\beta_{w+1} &= \log\beta_w + \sigma_\epsilon \epsilon_w \\
+    \epsilon_w &\sim \N(0, 1)
+\end{align}
+where $w(t)$ is the week number of time $t$, with week 1 being the first week modelled.
+Two parameters are unspecified here: the initial random of the random walk, $\beta_1$, and the standard deviation of the random walk, $\sigma_\epsilon$.
+See \cref{perf-test:sec:parameters-priors} for the implementation and parameterisation of these.
+
+\subsubsection{Identifiability}
+
+Following \textcite{birrellBayesian}, I reparameterise the model to improve identifiability.
+Rather than directly inferring $\beta_1$, I infer the initial growth rate of the epidemic.
+In addition, I use a parsimonious parameterisation of the initial conditions using the initial proportion of the population that are infected and the initial proportion that are susceptible only.
+Both of these rely on an assumption that the epidemic is in the steady-state exponential growth phase at time 0.
+
+For any short period of time when $\vec{s}$ is approximately constant, the system of ODEs can be solved analytically to produce exponential growth.
+In fact, all the $\vec{e}$, $\vec{i}$ and $\vec{p}$ compartments grow at the same initial exponential growth rate, $\psi$.
+\todo{move R definition for each model into where they are introduced}
+If $\vec{s} = 1$ then the basic reproduction number can be linked to the growth rate of this epidemic as follows~\autocites{birrellRealtimea}{wearingAppropriate}:
+\begin{align}
+    \R = \frac{\psi \left( \frac{\psi}{2\sigma} + 1 \right)^2}{\gamma \left(1 - \frac{1}{\left(\frac{\psi}{2 \gamma} + 1 \right)^2} \right)} \label{SEIR:eq:rtoR}.
+\end{align}
+
+$\R$ can be linked to $\beta_1$ because $\R$ is also equal to the dominant eigenvalue of the next-generation matrix, $\matr{G} = \matr\beta \circ \matr{M}$~\autocites{diekmannDefinition}{birrellRealtimea}.
+Therefore, the following procedure finds $\beta_1$ such that the growth rate $\psi$ is achieved.
+First calculate $\R$ from the model parameters using \cref{SEIR:eq:rtoR}.
+Denote this $\hat{R}$.
+Then calculate $R^* = D(\matr{G})$ when $\beta_1=1$, where $D(\cdot)$ gives the dominant eigenvalue.
+Multiplying a matrix by a scalar multiplies its dominant eigenvalue by that same scalar, therefore, setting $\beta_1 = \hat{R} / R^*$ gives the desired $\R$.
+
+To form a parsimonious parameterisation of the initial state, first focus on the initial proportion of the population that is in infected compartments.
+In steady-state exponential growth, the relative proportion of each strata that is infected is proportional to the dominant eigenvector of $\matr{G}$, which I denote $\vec{d}$.
+Let the proportion of the population in the $i$ states at time 0 in the strata with the highest proportion infected, $i^+$, be a model parameter.
+Then, $\vec{i_0} = \vec{i_1} + \vec{i_2} = \frac{i^+}{\max \vec{d}} \vec{d}$.
+I introduce an additional vector of parameters, $\vec{\pi}$, where $\pi_i$ is the proportion of strata $i$ without immunity at time 0.
+That is, at time 0, $\vec{r} = 1 - \vec{\pi}$.
+
+The system of ODEs and the result that all the comparments are growing at the same rate give the following for the initial state of the compartments:
+\begin{align}
+    \vec{i_1} &= \vec{i_0} \left(1 + \frac{2\gamma}{\psi + 2\gamma} \right)^{-1} \\
+    \vec{i_2} &= \vec{i_0} - \vec{i_1} \\
+    \vec{e_2} &= \vec{i_1} \frac{\psi + 2\gamma}{2\sigma} \\
+    \vec{e_1} &= \vec{e_2} \frac{\psi + 2\sigma}{2\sigma} \\
+    \vec{s} &= \vec{\pi} - \vec{i_0} - \vec{e_1} - \vec{e_2} \\
+    \vec{p_0} &= \vec{e_1} \frac{\psi + 2\sigma}{\psi + \kappa_0} \\
+    \vec{p_1} &= \vec{p_0} \frac{\kappa_0}{\psi + \kappa_Y} \\
+    \vec{p_2} &= \vec{p_1} \frac{\kappa_Y}{\psi + \kappa_{X1}} \\
+    \vec{p_3} &= \vec{p_2} \frac{p_x \kappa_{X1}}{\psi + \kappa_{X2}}.
+\end{align}
+
+\subsection{Full model} \label{SEIR:sec:full-model}
 
@@
 
-The full model combines each of the elements discussed in the previous sections.
+The full model combines the elements discussed in the previous sections.
 The transmission model follows \textcite{birrellRealtime}.
@@
 Recall that each region is run independenlty (see \cref{SEIR:sec:geography}), $\matr{beta}_{ij} = \beta(t) \beta^\text{susc}_i$ with $\beta^\text{susc}_i = \beta_c$ for child-aged strata and $\beta^\text{susc}_i = 1$ for adult-aged strata (see \cref{SEIR:sec:structured-populations}), and that $\beta(t)$ and $\matr{M}$ are time-varying (see \cref{SEIR:sec:time-varying-foi}).
-
-\section{Inference and implementation} \label{SEIR:sec:inference-implementation}
-
-\subsection{MCMC} \label{SEIR:sec:MCMC}
-
-I use an adaptive random-walk Metropolis-Hastings algorithm.
-The proposal distribution is a multivariate normal that adapts to the covariance of the posterior.
-The adaption algorithm is based on \textcite[algorithm 4]{andrieuTutorial}, as implemented in \textcite{ghoshApproximate}.
-Proposals are accepted or rejected using the standard Metropolis-Hastings (MH) acceptance probability.
-See \cref{SEIR:MCMC-algorithm} for full details.
-\begin{algorithm}
- set $\vec{X_0}$ to an initial value of the parameter vector \;
- $\vec{\mu_0} = \vec{X_0}$ \;
- $\matr{\Sigma_0} = \text{diag}(\vec{\mu_0})$ \;
- $\lambda_0 = 1$ \;
- \For{$i = 1, \dots, M$}{
-  sample $\vec{Y_}i \sim \N(\vec{\mu_{i-1}}, \lambda_{i-1}\matr{\Sigma_{i-1}})$\;
-  set $\vec{X_i}$ to $\vec{Y_i}$ or $\vec{X_{i-1}}$ using a MH acceptance step\;
-  update the proportion of proposals accepted so far, $\alpha$ \;
-  \eIf(\tcc{No adaptation for 200 iterations}){$i \leq 200$}{
-    $\lambda_i = \lambda_{i-1}$ \;
-    $\vec{\mu_i} = \vec{\mu_{i-1}}$ \;
-    $\matr{\Sigma_i} = \matr{\Sigma_{i-1}}$ \;
-   }{
-    $\gamma_i = (i - 200)^{-0.6}$ \;
-    $\log \lambda_i = \log \lambda_{i-1} + \gamma_i(\alpha - 0.234)$ \;
-    $\vec{\mu_i} = (1 - \gamma_i) \vec{\mu_{i-1}} + \gamma_i \vec{X_i}$ \;
-    $\matr{\Sigma_i} = (1 - \gamma_i) \matr{\Sigma_{i-1}} + \gamma_i (\vec{X_i} - \vec{\mu_i})(\vec{X_i} - \vec{\mu_i})^T$ \;
-  }
- }
- \caption{Algorithm for adaptive random-walk Metropolis-Hastings.
-$\vec{\mu_i}$ and $\matr{\Sigma_i}$ are the mean and covariance of the proposal distribution at iteration $i$.
-They adapt to the shape and location of the posterior.
-$\text{diag}(\vec{\mu_0})$ is the diagonal matrix with diagonal entries equal to $\vec{\mu_0}$.
-$\lambda_i$ is the scale parameter of the proposal distribution at iteration $i$.
-$\gamma_i$ is the learning rate, which determines how much adaptation occurs.
-$\gamma_i \to 0$ as $i \to \infty$ so the rate of adaptation is \emph{vanishing}.
-Vanishing adaptation is required for the algorithm to converge to the target distribution~\autocite[section 3]{andrieuTutorial}.}
- \label{SEIR:MCMC-algorithm}
-\end{algorithm}
-
-\subsection{Parameterisation}
-
-Following \textcite{birrellBayesian}, I reparameterise the model to improve identifiability.
-Rather than directly inferring $\beta_1$, I infer the initial growth rate of the epidemic.
-In addition, I use a parsimonious parameterisation of the initial conditions using the initial proportion of the population that are infected and the initial proportion that are susceptible only.
-Both of these rely on an assumption that the epidemic is in the steady-state exponential growth phase at time 0.
-
-For any short period of time when $\vec{s}$ is approximately constant, the system of ODEs can be solved analytically to produce exponential growth.
-In fact, all the $\vec{e}$, $\vec{i}$ and $\vec{p}$ compartments grow at the same initial exponential growth rate, $\psi$.
-\todo{move R definition for each model into where they are introduced}
-If $\vec{s} = 1$ then the basic reproduction number can be linked to the growth rate of this epidemic as follows~\autocites{birrellRealtimea}{wearingAppropriate}:
-\begin{align}
-    \R = \frac{\psi \left( \frac{\psi}{2\sigma} + 1 \right)^2}{\gamma \left(1 - \frac{1}{\left(\frac{\psi}{2 \gamma} + 1 \right)^2} \right)} \label{SEIR:eq:rtoR}.
-\end{align}
-
-$\R$ can be linked to $\beta_1$ because $\R$ is also equal to the dominant eigenvalue of the next-generation matrix, $\matr{G} = \matr\beta \circ \matr{M}$~\autocites{diekmannDefinition}{birrellRealtimea}.
-Therefore, the following procedure finds $\beta_1$ such that the growth rate $\psi$ is achieved.
-First calculate $\R$ from the model parameters using \cref{SEIR:eq:rtoR}.
-Denote this $\hat{R}$.
-Then calculate $R^* = D(\matr{G})$ when $\beta_1=1$, where $D(\cdot)$ gives the dominant eigenvalue.
-Multiplying a matrix by a scalar multiplies its dominant eigenvalue by that same scalar, therefore, setting $\beta_1 = \hat{R} / R^*$ gives the desired $\R$.
-
-To form a parsimonious parameterisation of the initial state, first focus on the initial proportion of the population that is in infected compartments.
-In steady-state exponential growth, the relative proportion of each strata that is infected is proportional to the dominant eigenvector of $\matr{G}$, which I denote $\vec{d}$.
-Let the proportion of the population in the $i$ states at time 0 in the strata with the highest proportion infected, $i^+$, be a model parameter.
-Then, $\vec{i_0} = \vec{i_1} + \vec{i_2} = \frac{i^+}{\max \vec{d}} \vec{d}$.
-I introduce an additional vector of parameters, $\vec{\pi}$, where $\pi_i$ is the proportion of strata $i$ without immunity at time 0.
-That is, at time 0, $\vec{r} = 1 - \vec{\pi}$.
-
-The system of ODEs and the result that all the comparments are growing at the same rate give the following for the initial state of the compartments:
-\begin{align}
-    \vec{i_1} &= \vec{i_0} \left(1 + \frac{2\gamma}{\psi + 2\gamma} \right)^{-1} \\
-    \vec{i_2} &= \vec{i_0} - \vec{i_1} \\
-    \vec{e_2} &= \vec{i_1} \frac{\psi + 2\gamma}{2\sigma} \\
-    \vec{e_1} &= \vec{e_2} \frac{\psi + 2\sigma}{2\sigma} \\
-    \vec{s} &= \vec{\pi} - \vec{i_0} - \vec{e_1} - \vec{e_2} \\
-    \vec{p_0} &= \vec{e_1} \frac{\psi + 2\sigma}{\psi + \kappa_0} \\
-    \vec{p_1} &= \vec{p_0} \frac{\kappa_0}{\psi + \kappa_Y} \\
-    \vec{p_2} &= \vec{p_1} \frac{\kappa_Y}{\psi + \kappa_{X1}} \\
-    \vec{p_3} &= \vec{p_2} \frac{p_x \kappa_{X1}}{\psi + \kappa_{X2}}.
-\end{align}
 

### 2023-11-17 SEIR.tex
Edit figures, talk about structured populations

@@
 The explicit representation means that a mechanistic model's state and parameters have biological and/or epidemiological interpretations; these interpretations provide scientific understanding of the disease.
-Of particular interest is that the basic reproduction number, $R_0$, and the time-varying reproduction number, $R_t$, can both be calculated as simple functions of the model parameters; in \cref{E-backcalc} additional modelling is required to calculate these numbers.
+Of particular interest is that the basic reproduction number, $R_0$, and the effective reproduction number, $R_e(t)$, can both be calculated as simple functions of the model parameters; using \cref{E-backcalc} additional modelling would be required to estimate these numbers.
 Furthermore, mechanistic models can be used for scenario-based modelling, whereby the effect of different interventions can be simulated.
@@
 I adopt conventions and notation from the infectious disease modelling field, although this is not consistent in the literature.
-In particular, upper-case letters denote numbers of people, and lower-case letters denote proportions.
-Whether a variable is random or not should be clear from the context, or will be explicitly stated if needed.
+In particular, case no longer signifies whether a variable is random or not, but will be clear from the context.
 
@@
 
+\todo[inline]{ref intro to compartmental model setup}
 This section introduces the class of mechanistic transmission models known as \emph{compartmental} models.
@@
 %There remains randomness in the observation model (described in \cref{SEIR:sec:observation}).
-Stochastic models specify a probability distribution $p(\vec{x} \mid \vec{\theta})$ for the state vector $\vec{x}(t)$ at time $t$.
-In general, this means, there is a dependence between $\vec{x}(t)$ and $\vec{x}(t')$.
+Stochastic models are most commonly specified in discrete time, using a Markov assumption.
+That is, they specify a probability distribution $p(\vec{x}(t) \mid \vec{x}(t-1), \vec{\theta})$.
 This relaxes the assumption of independent increments in the counting process formulation of the infection process (see \cref{E-inc-prev:sec:infection-process}).
@@
 Deterministic models are also known as \emph{mean-field} models because they can are the behaviour corresponding to a stochastic model's mean.
-When the number of infected individuals is large, then this mean-field behaviour is a good approximation of the fully stochastic model, due to the law of large numbers\todo{ref mean-field and approximation error}.
+When the number of infected individuals is large, the law of large numbers means that the mean-field behaviour is a good approximation of the fully stochastic model~\autocite[20]{diekmannMathematical}.
 An alternative justification, when making noisy observations of the system, is that the additional noise from the stochastic model is negligible compared to this observation noise.
-This argument is similar to the justification for neglecting the stochasticity of the prevalence process in \cref{E-inc-prev:sec:observation-process}.
+This second justification is of the same form as the one made for neglecting the stochasticity of the prevalence process in \cref{E-inc-prev:sec:observation-process}.
 The remainder of this chapter uses only deterministic models, because throughout the time period this thesis considers there were enough infectious individuals.
@@
 
-\Cref{SEIR:sec:SIR} begins by introducing the \emph{susceptible-infected-recovered} (SIR) model, a simple compartmental model for a disease which confers immunity.
+\Cref{SEIR:sec:SIR} introduces the \emph{susceptible-infected-recovered} (SIR) model, a simple compartmental model for a disease which confers immunity.
 The SIR model is named because each individual in the population is divided into one of three states: susceptible, infectious, or recovered.
-This model has several assumptions which make it inappropriate to model SARS-CoV-2; therefore, I introduce extensions to relax these assumptions.
+This model has several assumptions which make it inappropriate to model SARS-CoV-2; therefore, the remainder of this sections relaxes these assumptions.
 First, \cref{SEIR:sec:non-exponential} relaxes the assumption of an exponentially-distributed infectious period to a gamma distribution.
 Then, \cref{SEIR:sec:SEIR} relaxes the assumption that individuals are infectious immediately after they are infected; it does this by extending the SIR model to the SEIR model, including an \emph{exposed} compartment.
-Next, \cref{SEIR:sec:structured-populations} relaxes the homogeneous population assumption, introducing stratification by important characteristic(s).
-Finally, \cref{SEIR:sec:time-varying-foi} relaxes the assumption of a constant transmission rate to allow for time-varying behavioural changes.
+Finally, \cref{SEIR:sec:structured-populations} allows heterogeneity in the population, introducing stratification by important characteristic(s).
+% Finally, \cref{SEIR:sec:time-varying-foi} relaxes the assumption of a constant transmission rate to allow for time-varying behavioural changes.
 
@@
 \begin{figure}[h]
-\centering
+\makebox[\textwidth][c]{
 \begin{tikzpicture}[
@@
     ->,>=stealth',
-    every state/.style={draw,rectangle,very thick},
+    every state/.style={draw,rectangle},
     ]
@@
 
-    \path (S) edge node {$\beta i$} (I)
+    \path (S) edge node {$\lambda(t)$} (I)
           (I) edge node {$\gamma$} (R);
 \end{tikzpicture}
-  \caption[The SIR model]{Schematic of the basic SIR model.}
+}
+  \caption[The SIR model]{Schematic of the basic SIR model. Arrows are labelled with the rates at which individuals move between compartments.}
   \label{SEIR:fig:SIR}
@@
 The name refers to the three compartments in the model, displayed in \cref{SEIR:fig:SIR}.
-Individuals in the susceptible compartment can be infected; $s(t)$ is the proportion of the population that is susceptible at time $t$.
-Individuals in the infectious compartment can infect others, and eventually recover; $i(t)$ is the proportion of the population that is infectious at time $t$.
-Individuals in the recovered compartment are immune to the disease, and cannot be infected again; $t(t)$ is the proportion of the population that have recovered by time $t$.
+Individuals in the susceptible compartment are those that could be infected (they have no immunity).
+The proportion of the population that is susceptible at time $t$ is $s(t)$.
+Individuals in the infectious compartment can infect others, and eventually recover.
+The proportion of the population that is infectious at time $t$ is $i(t)$.
+Individuals in the recovered compartment are immune to the disease, and cannot be infected again.
+The proportion of the population that have recovered by time $t$ is $r(t)$.
 The model's state is $\vec{x}(t) = (s(t), i(t), r(t))^T$.
+In the SIR model, individuals are assumed to be homogeneous, \ie they are all identical.
 
@@
 $\gamma$ is the recovery rate, or the proportion of infected individuals that recover per day; hence, $1/\gamma$ is the mean infectious period~\autocite[367]{keelingModeling}.
-These parameters are assumed to be constant (an assumption relaxed in \cref{SEIR:sec:time-varying-foi}).
+These parameters are assumed to be constant.
 
-The term $\beta s(t)i(t)$ is the rate at which susceptible individuals (as a proportion of the total population size) are infected~\autocite[18]{keelingModeling}.
-Susceptible individuals move from the $s(t)$ to $i(t)$ when they are infected.
-This term can be arrived at by considering a single infected individual~\autocite[214]{kretzschmarMathematical}.
-Assume this individual comes into contact with other individuals in the population at a rate of $\kappa$ per day.
-Further, assume the population is \emph{well-mixed}, meaning that each contact is with an individual chosen uniformly at random from the population.
+% The term $\beta s(t)i(t)$ is the rate at which susceptible individuals (as a proportion of the total population size) are infected~\autocite[18]{keelingModeling}.
+% Susceptible individuals move from the $s(t)$ to $i(t)$ when they are infected.
+The expression for the force of infection, $\lambda(t)$ can be arrived at by considering a single infected individual~\autocite[214]{kretzschmarMathematical}.
+Assume this individual, on average, comes into contact with other individuals in the population at a rate of $\kappa$ per day.
+Further, assume the population is \emph{well-mixed}, meaning that each contact is with an individual chosen uniformly at random from the population; this is implied by assuming a homogeneous population.
 For such a randomly selected individual, the probability that they are susceptible is $s(t)$.
-Each of the infected individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
+% Each of the infected individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
 Therefore, in expectation, the infected individual makes contact with susceptible individuals at a rate of $\kappa s(t)$.
 Assume that there is a constant probability of transmission upon contact between an infected and susceptible individual, $q$.
-Combined with the above, the expected rate of infections caused by this infected individual is $\kappa s q$.
+Combined with the above, the expected rate of infections caused by this infected individual is $\kappa s(t) q$.
 The two constant terms here (number of contacts and probability of transmission) are absorbed into the transmission rate, $\beta = \kappa q$.
 Therefore, the rate of new infections generated by a single infectious individual is $\beta s(t)$.
-Multiplying by the number of infected individuals and dividing by the population size gives the rate at which susceptible individuals are infected as a proportion of the population, $\beta s(t) i(t)$.
+Multiplying by the number of infected individuals and dividing by the population size gives the rate at which susceptible individuals are infected as a proportion of the population, $\beta s(t) i(t) = \lambda(t) s(t)$.
 
-The term $\gamma i(t)$, the rate at which infected individuals (as a proportion of the total population size) recover, follows directly from the definitions of $\gamma$ and $i(t)$.
+% The term $\gamma i(t)$, the rate at which infected individuals (as a proportion of the total population size) recover, follows directly from the definitions of $\gamma$ and $i(t)$.
 
@@
 \begin{itemize}
-    \item The basic reproduction number (discussed previously in\todo{ref intro}) for this model is $R_0 = \beta / \gamma$~\autocite[20]{keelingModeling}.
+    \item The basic reproduction number for this model is $R_0 = \beta / \gamma$~\autocite[20]{keelingModeling}.
     Intuitively, this can be thought of as the average number of infections caused by an infectious individual per day, $\beta$, multiplied by the average number of days they are infectious for, $1/\gamma$.
+    \item The effectie reproduction number for this model is $R_e(t) = \beta / \gamma s(t) = R_0 s(t)$~\autocite{pellisEstimation}.
     \item The population is closed.
@@
     Births and deaths of individuals are also excluded, although they can easily be incorporated in situations where they are relevant~\autocites[26]{keelingModeling}[214]{kretzschmarMathematical}.
-    \item There is no analytical solution to the system of ODEs~\autocite[25]{keelingModeling}: numerical methods are required to solve the system.
+    \item There is no analytical solution to the system of ODEs: numerical methods are required to solve the system~\autocite[25]{keelingModeling}.
     \item In addition to the parameters, the state of the system at some time (normally the initial state, the state at time $t=0$) is required to solve the system.
-    \item Immunity is assumed to last until the end of the modelled period.
-    For many diseases, including SARS-CoV-2, this is a reasonable assumption as long as the modelled period is short, as it will be in this chapter\todo{ref that SARS-CoV-2 immunity is long enough to not be an issue}.
-    For modelling longer time periods, waning immunity can be incorporated\todo{ref waning immunity models}.
+    \item Immunity is assumed to last until the end of the modelled period~\autocite[61]{andersonInfectious}.
+    In many contexts, including here, this assumption is justified because the modelled period is shorter than the length of immunity conferred~\autocite{milneImmunity}.
+    For modelling longer time periods, waning immunity can be incorporated by allowing individuals to move from the recovered compartment back to the susceptible compartment~\autocite[40]{keelingModeling}.
 \end{itemize}
@@
 \begin{figure}[h]
+\makebox[\textwidth][c]{
 \begin{tikzpicture}[
@@
 \end{tikzpicture}
+}
   \caption[The SIR model with non-exponential infectious period]{Schematic of the SIR model with non-exponential infectious period. The infectious period is modified by having multiple I states.}
@@
 \end{align}
-where $n$ is the number of infected compartments, and redefining $\lambda(t) = \beta \sum_{j=1}^n i_j(t)$ 
+where $n$ is the number of infected compartments, and redefining $\lambda(t) = \beta \sum_{j=1}^n i_j(t)$.
 In this model the infectious period is distributed $\GamDist(n, n\gamma)$, a realistic model for a variety of diseases~\autocite{wearingAppropriate}.
-This parameterisation is chosen to give $1/\gamma$ as the mean infectious period (the gamma distribution and its expectation is defined in \cref{E-distributions}).
+This parameterisation is chosen to give $1/\gamma$ as the mean infectious period (the gamma distribution's definition and its expectation are in \cref{E-distributions}).
 % (it is slightly unclear what it means for a variable to have a distribution in a deterministic model, the clearest explanation is that this is the distribution which the ).
-A gamma distribution with the first parameter equal to 1 is an exponential distribution, connecting this model to the basic SIR model in \cref{SEIR:sec:SIR}.
+A gamma distribution with the first parameter equal to 1 is an exponential distribution, in which case this model collapses to the basic SIR model in \cref{SEIR:sec:SIR}.
 The distribution for the time spent in the infected compartments tends to a constant (\ie the same for all individuals) as $n \to \infty$~\autocite{lloydRealistic}.
@@
     ->,>=stealth',
-    every state/.style={draw,rectangle,very thick},
-    remember picture, overlay
+    every state/.style={draw,rectangle},
+    % remember picture, overlay
     ]
@@
     \node[state, right=of E3] (E4) {$e_m$};
-    \node[state, right=of E4] (I1) {$i_1$};
+    % \node[state, right=of E4] (I1) {$i_1$};
+    \node[state, below=of E1] (I1) {$i_1$};
     \node[state, right=of I1] (I2) {$i_2$};
@@
           (E3) edge node {$m\sigma$} (E4)
-          (E4) edge node {$m\sigma$} (I1)
           (I1) edge node {$n\gamma$} (I2)
@@
           (I4) edge node {$n\gamma$} (R);
+    \draw[->] (E4.east) -- ++(1,0) |- ([yshift=-1cm]E4.east) node [near start, right] {$m\sigma$} -|  (I1.north);
 \end{tikzpicture}
+}
   \caption[The SEIR model]{Schematic of the SEIR model.}
@@
 This is unrealistic for most diseases, as there is a period between an individual being infected and them being infectious, known as the \emph{latent period}.
-\todo{perhaps link with where define phases of infection in background chapters}
+\todo{I will probably end up defining the latent period in the intro, so should just ref that}
 Ignoring the latent period will again overestimate the reproduction number~\autocite{wearingAppropriate}.
@@
 If the strata have differing transmission patterns, this can change the dynamics of the epidemic.
-The different patterns could be due to behavioural factors (\eg high-numbers of sexual partners in a sexually-transmitted disease~\autocite[69]{keelingModeling} or younger age groups making many more contacts for respiratory diseases~\autocite[176]{andersonInfectious}) or biological factors (\eg genetics reducing the susceptibility of some individuals~\autocite[208]{andersonInfectious}).
-I will assume that the disease progression (\eg length of the exposed and infectious periods) is the same for all strata, although relaxing this assumption is easy.
+The different patterns could be due to behavioural factors (\eg high-numbers of sexual partners in a sexually-transmitted disease~\autocite[69]{keelingModeling}, younger age groups making many more contacts for respiratory diseases~\autocite[176]{andersonInfectious}), or biological factors (\eg genetics reducing the susceptibility of some individuals~\autocite[208]{andersonInfectious}).
+I will assume that the disease progression (\eg length of the exposed and infectious periods) is the same for all strata, although this assumption can be relaxed.
 Stratification requires two major changes model changes: extension of the state vector and describing the between-strata interactions.
 
-To incorporate the stratification into the model, first I introduce some additional notation.
-Denote the set of strata of size $A$, as $\set{A}$.
-A stratum will be denoted by $a \in \set{A}$ whose total population is $N_a$.
+The state vector is now a vector of vectors, one for each of the compartments.
+I denote the set of strata, as $\set{A}$, with the set's size as $A$.
+For example, in the basic SIR model, $\vec{x}(t) = (\vec{s}(t), \vec{i}(t), \vec{r}(t))^T$.
+The $a$th element of each vector is the proportion of strata $a$ in that compartment.
+For example, $\vec{s}(t) = (s_1(t), \dots, s_A(t))^T$ where $s_a(t)$ is the proportion of the $a$th strata that are susceptible at time $t$.
+Hence, $N_a s_a(t)$ is the number of susceptible people in stratum $a$.
+These vectors simply replace the scalars in the previous equations.
+Various other definitions are possible and equivalent, for example defining $s_a$ as the proportion of the total population that are both in strata $a$ and susceptible or parameterised using numbers of individuals~\autocite[57]{keelingModeling}.
 
-First, the state vector is now a vector of vectors, one for each of the compartments.
-For example, in the SEIR model of \cref{SEIR:sec:SEIR}, $\vec{x} = (\vec{s}, \vec{e_1}, \dots, \vec{e_m}, \vec{i_1}, \dots, \vec{i_n}, \vec{r})^T$.
-The $a$th element of each vector is the proportion of strata $a$ in that compartment.
-For example, $\vec{s} = (s_1, \dots s_A)^T$ where $s_a$ is the proportion of the $a$th strata that are susceptible.
-Hence, $N_a s_a$ is the number of susceptible people in stratum $a$.
-These vectors simply replace the scalars in the previous equations.
-Various other definitions are possible and equivalent, for example defining $s_a$ as the proportion of the total population that are both in strata $a$ and susceptible or parameterised using numbers of individuals\todo{cite Keeling}.
+The interactions between strata also need describing.
+This is done by replacing the scalar $\beta$ with a \emph{Who Acquires Infection From Whom} (WAIFW) matrix, $\matr{\beta}$~\autocite[58]{keelingModeling}; its interpretation is described below.
+The force of infection is now a vector, $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i_j}(t)$.
+Its expression can be arrived at by considering a single infected individual in strata $a'$.
+Assume any such individual comes into contact with individuals in strata $a$ at a rate of $\kappa_{aa'}$ in units of contacts per day.
+% Further, assume that each contact is with an individual chosen uniformly at random from within the strata.
+% For such a randomly selected individual, the probability that they are susceptible is $s_a(t)$.
+% Each of the infected individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
+Following the same argument as in \cref{SEIR:sec:SIR} finds that, in expectation, the infected individual makes contact with susceptible individuals in stratum $a$ at a rate of $\kappa_{aa'} s_a(t)$.
+Assume that there is a constant probability of transmission upon contact between an infected in strata $a'$ and susceptible in strata $a$ individual, $q_{aa'}$.
+Therefore, the expected rate of infections in strata $a$ caused by this infected individual is $\kappa_{aa'} s_a(t) q_{aa'}$.
+Multiplying by the number of infected individuals in strata $a'$, $N_{a'} i_{a'}(t)$, gives $\kappa_{aa'} q_{aa'} s_a(t) N_{a'} i_{a'}(t)$.
+The constant terms here are absorbed into the WAIFW matrix, giving $\beta_{aa'} = \kappa_{aa'} q_{aa'} N_{a'}$.
+The total force of infection on strata $a$ from all strata is, therefore, $\lambda_a(t) = \sum_{a' \in \set{A}} \beta_{aa'} i_{a'}(t)$.
+Equivalently, $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i_j}(t)$.
+If all $\beta_{aa'}$ are equal, then the model is equivalent to the homogeneous model (\ie without any structure).
+The rate that individuals move between the susceptible and exposed compartments is now $\vec{s}(t) \circ \vec\lambda(t)$, where $\circ$ denotes element-wise multiplication.
 
-Second, the interactions between strata must be described.
-This is done by replacing the scalar $\beta$ with a \emph{Who Acquires Infection From Whom} (WAIFW) matrix, $\matr{\beta}$.
-The force of infection is now a vector, $\vec{\lambda} = \matr{\beta} \sum_{j=1}^n \vec{i_j}$.
-The matrix element $\beta_{aa'}$ is the rate at which individuals in stratum $a$ infect individuals in stratum $a'$.
-In practical situations, estimating all of these parameters is infeasible, so some assumptions must be made to reduce the number of parameters.
-Common assumption include assuming symmetry, \ie $\beta_{aa'} = \beta{a'a}$ or a decomposable independence, \ie $\beta_{aa'} = \beta \beta^\text{inf}_a \beta^\text{susc}_{a'}$~\todo{ref A+M on these parameterisations}.
+In practical situations, estimating all of the $\beta_{aa'}$ parameters is infeasible, so some assumptions must be made to reduce the number of parameters~\autocite[176]{andersonInfectious}.
+The infeasibility arises because the progression of the epidemic depends on $\vec{\lambda}(t)$, which has dimension $A$, while $\matr{\beta}$ has dimension $A^2$.
+Common assumption include assuming symmetry, \ie $\beta_{aa'} = \beta_{a'a}$, or that the direction of transmission is irrelevant, \ie $\beta_{aa'} = \beta_{a} \beta_{a'}$.
 Alternatively, as in the scalar case, $\beta_{aa'}$ can be decomposed into the contact rate between strata $a$ and $a'$, $\kappa_{aa'}$ and the probability of transmission upon contact, $q_{aa'}$, and then assumptions made about the components.
+The choice of assumptions depends on the context, including the disease being studied and the data available.
 
-\todo[inline]{start with a generic intro on structured populations and use of matrices, then give age as a motivating example, finally specifics go elsehwere}
+Reproduction numbers in models with structured populations need to be careful to consider the distribution of infections across the strata.
+The WAIFM allows calculation of the number of secondary infections generated by a single infection in the $a$th strata.
+This follows directly from its definition as $\frac{1}{\gamma} \sum_{a'} \beta_{aa'}$. 
+However, how to average these contributions to form the reproduction is not obvious.
+The standard solution to this was proposed by \textcite{diekmannDefinition} (although \textcite[chapter 7]{diekmannMathematical} provides a more gentle, but still rigorous, explanation).
+First, they show that the distribution of infections across the strata during the initial period of exponential growth (when $\vec{s} \approx \vec{0}$) relies only on $\matr{\beta}$.
+The distribution is proportional to the \emph{dominant eigenvector} (the eigenvector corresponding to the largest eigenvalue) of $\matr{\beta}$.
+Denote by $\rho(\matr{\beta})$ the \emph{dominant eigenvalue} (the largest eigenvalue).
+The basic reproduction number is then defined as $R_0 = \rho(\matr{\beta}) / \gamma$.
+This $R_0$ keeps the same interpretation as in the homogeneous case, \ie the average number of secondary infections caused by a single infection in a fully susceptible population, where the average is taken over the distribution implied by the dominant eigenvector.
+The important threshold that exponential growth occurs (in expectation) iff $R_0 > 1$ remains.
 
-\todo[inline]{Possibly add a paragraph about the importance of age and mixing patterns, maybe show a contact matrix?}
-
-To allow for age, the population is divided into age groups.
-The state of each compartment is now a vector.
-The proportion of the population that is susceptible is $\vec{s} = (s_1, \dots, s_A)$ where $s_a$ is the proportion of the $a$th age group that is susceptible.
-Similarly, for the other compartments.
-
-The modelling approach here can be used to stratify the population by any characteristic.
-For example, HIV/AIDS models incorporated subgroups with different patterns of sexual activity~\autocite{kretzschmarMathematical}.
-In other contexts, transmission is strongly influenced by socio-economic status, and the model can be stratified by income levels~\autocite{menkirIncorporating}.
-In this case, $i$ simply represents the $i$th strata.
-
-The assumption of a well-mixed population is relaxed by using a contact matrix.
-Now, only conditional on age are all individuals identical.
-
-The derivation is similar to the single stratum case, except that the contact rates and probabilities of transmission will be allowed to vary for each pair of strata.
-Consider an infected individual in stratum $j$.
-Assume they make contacts with individuals in stratum $i$ at a rate of $K_{ij}$ per day.
-Therefore, they make contact with susceptible individuals in stratum $i$ at a rate of $s_i K{ij}$ per day.
-Assume that the probability of transmission from an infected individual in stratum $j$ to a susceptible individual in stratum $i$ upon contact is $\beta_{ij}$.
-Therefore, $K_{ij} \beta_{ij} s_i$ is the instantaneous proportion of individuals in stratum $i$ infected by an individual in stratum $j$.
-Considering infected individuals across all strata, $\sum_{j \in \set{A}} K_{ij} \beta_{ij} s_i$ is the instantaneous proportion of individuals in stratum $i$ infected.
-Therefore, modify\todo{refer to right equations and add missing bit} with the following:
-\begin{align}
-    \frac{d\vec{s}}{dt} = -\vec{s} \circ ((\matr{\beta} \circ \matr{K}) \vec{i})
-    \frac{d\vec{i}_1}{dt} = \vec{s} \circ ((\matr{\beta} \circ \matr{K}) \vec{i}) -
-    \label{eq:matrix:theory}
-\end{align}
-where $\circ$ is the element-wise product (\ie: $(\matr{A} \circ \matr{B})_{ij} = A_{ij} B_{ij}$) and $\vec{i} = \sum_{k=1}^n \vec{i}_k$.
-
-$K_{ij}$ and $\beta_{ij}$ appear only as a product and therefore are not individually identifiable.
-Normally, $K_{ij}$ is assumed to be known, for example from a previous contact survey~\autocite[such as][]{mossongSocial}.
-Ideally, all the $\beta_{ij}$ terms would be estimated, but there is normally inadequate data to do so, requiring assumptions to be made to reduce the number of parameters~\autocite[65]{keelingModeling}.
-% $\beta_{ij}$ is then estimated, although often with a parsimonious model\todo{ref / give examples / delete on how age-specfic susceptibility / transmissibility happens}.
-A common assumption is that $\beta_{ij}$ can be decomposed into two independent components such that $\beta_{ij} = \beta \beta^\text{susc}_{i} \beta^\text{inf}_j$.
-For any stratum $i$, $\beta^\text{susc}_i$ is known as the \emph{relative susceptibility} of stratum $i$.
-Informally, it is how vulnerable to infection individuals in stratum $i$ are.
-Formally, define an $a_0$ and $a'_0$ and enforce $\beta^\text{susc}_{a_0}=1$ for identifiability.
-Then $\beta^\text{susc}_i$ is the relative risk of an individual in stratum $i$ being infected, relative to an individual in stratum $a_0$.
-\todo{clarify notation and explaantion here of reference class}
-it is the probability that a susceptible individual in stratum $i$ will be infected upon contact with an infectious individual, relative to a individual in the reference stratum.
-Similarly, $\beta^\text{inf}_i$ is known as the \emph{relative infectiousness} of stratum $i$.
-Informally, it is how likely individuals in stratum $i$ are to pass on the infection.
-Formally, it is the relative risk of an infectious individual in stratum $i$ infecting a susceptible individual upon contact, relative to an individual in the reference stratum.
-These measures could be biological, \eg due to differences in immune systems, or behavioural, \eg if infants are less hygenic than adults.
-
-%Now, we assume that $\beta_{ij} = \beta(t) \beta_{ij}$ where $\beta(t)$ is a non-strata-specific, time-varying component representing precautions taken by the population, for example masking; and $\beta_{ij}$ represents a constant probability of transmission from an infectious individual in strata $j$ to a susceptible individual in strata $i$ given contact, which could, for instance, be affected by differential susceptibility to infection for different age-groups.
-%We arbitrarily choose one $\beta_{ij}$ to be equal to 1 for identifiability and hence the other $\beta_{ij}$ values are relative to this reference class.
-
-\todo[inline]{move the below into a context-specific section}
-Based on prior work\todo{cite}, I assume that all individuals have the same \emph{infectiousness} but that there is a difference in \emph{susceptibility}.
-Therefore, $\beta^\text{inf}_{j} = 1$ for all $j$.
-The prior literature suggests that children are less susceptible than adults, but otherwise there is little difference in susceptibility.
-Therefore, I set $\beta^\text{susc}_i$ to 1 for adult-aged strata and $\beta^\text{susc}_i = \beta_c$ for child-aged strata.
-
-\subsection{Geographic structure} \label{SEIR:sec:geography}
-\todo[inline]{remove this section and just explain what I'm doing in this case in an implementation section}
-
-To allow for regional hetereogeneity....
-This assumes non-interaction between regions...
-
-The well-mixed population assumption is implausible for geographically dispersed populations.
-Additionally, due to non-response bias, CIS is not geographically representative.
-Therefore, I stratify the population by region.
-
-I assume that in each region the population is well-mixed and closed.
-A more complex model would allow for interaction between regions.
-However, for a large epidemic, non-interacting populations is a good approximation~\autocite{birrellRealtimea}.
-
-I implement this by running the model separately and independently for each region.
-England is split into nine regions (previously known as Government Offices for the Regions): South East, London, North West, East of England, West Midlands, South West, Yorkshire and The Humber, East Midlands, and North East.
-
-\todo[inline]{this paragraph probably just goes and its obvious from context}
-Each element of the vector $\vec{s}$ refers to the proportion of that region's population of the relevant age group that are susceptible.
-For example, $\vec{s}_1$ is...
-The other vectors are interpreted similarly.
-For simplicity, I do not introduce additional notation for the regions as it is clear from context which region is being referred to.
-Vectors referring to different regions will never be referred to in the same expression because the regions are modelled independently without any interaction.
+Based primarily on the study of influenza, age-based contact patterns have been found to be important for respiratory diseases~\autocites{birrellRealtimea}{jacksonEffects}.
+\todo{maybe better ref for influence of age contacts on respiratory diseases}
+Respiratory diseases can be spread by conversational contacts, which show very clear age patterns (see \cref{SEIR:fig:age-contacts}).
+In particular, the contacts are \emph{associative} (contacts are more likely between individuals of similar ages) and younger age groups have more contacts; this will tend to concentrate the epidemic in these younger age groups~\autocite[67]{keelingModeling}.
+\begin{figure}
+  \centering \includegraphics[width=\textwidth]{SEIR/contact_matrices}
+  \caption[Age-based contact matrices]{Number of contacts reported between age groups in two contact surveys.
+Left shows the number of contacts reported in the POLYMOD survey~\autocite{mossongSocial}.
+Right shows the number of contacts reported in the BBC Pandemic survey~\autocite{klepacContacts}, which only collected data on those aged over 12.
+A clear diagonal, showing associative mixing is visible, and children have more contacts.
+Parallel diagonals off the main diagonal show mixing between parents and their children.
+Figure adapted from \textcite{klepacContacts}.}
+  \label{SEIR:fig:age-contacts}
+\end{figure}
 
@@
 A variety of different approaches can be taken to link a transmission model to observations.
-Here, I extend the system of ODEs to include compartments representing PCR positive individuals.
-Whenever an individual is infected, a copy of them is placed in the pre-positive compartment, $p_0$.
+% Here, I extend the system of ODEs to include compartments representing PCR positive individuals.
+% The proportion of PCR positive individuals in each strata is then linked to the CIS using a Beta-Binomial likelihood, with the modelled proportion that are PCR positive as the mean of the Beta distribution.
+One approach would be to convolve the incidence based on the transmission model with the PCR duration distribution.
+\todo{tidy up this notation}
+Essentially, using the notation of \cref{E-inc-prev}, calculate the expected proportion of positive tests on day $t$ as $P_t = \sum_{j=1}^{\dmax} Z_{t-j} S(j)$.
+$Z_{t-j}$ can be related to the ODEs because $Z_t / \Npop = s(t) - s(t-1)$, as $s(t)$.
+However, this approach is computationally expensive.
+
+A much more efficient option is to approximate this convolution by adding compartments to the ODEs.
+This approach also allows the infected, pre-positive compartment to be represented.
+I will denote this new compartment as $p_0$.
+$p_0$ increases at the same rate as individuals flow into $e_1$.
 These individuals then transition into a set of compartment with their transition rates chosen to match the duration of positivity estimates in \cref{E-imperf-test}, $p_1, \dots, p_l$.
-These compartments do not have an interpretation.
-Their only purpose is to give the time spent in the PCR positive compartments the desired distribution.
+These compartments do not have a biological interpretation, unlike the other compartments in the system, but are purely a mathematical tool so that the time spent in these compartments matches the duration distribution.
+This is the same idea behind having multiple exposed or latent compartments (introduced in \cref{SEIR:sec:non-exponential}).
 
-I use the method of \textcite{osogamiClosed} to determine $l$ and the transition rates between the compartments.
 The class of distributions which can be represented as a series of compartments are known as \emph{phase-type} distributions.
+Discussions of general phase-type distributions are beyond the scope of this thesis; a review in a biological context can be found in \textcite{hobolthPhasetype}.
+I will restrict myself to the subset of phase-type distributions defined by \textcite{osogamiClosed} as Erlang-Coxian (EC) distributions.
+EC distributions are convenient because they are flexible enough to approximate almost any distribution, and the parameters of the approximation are given by closed-form functions of the distribution's moments.
 
-\Textcite{osogamiClosed} define a subset of phase-type distributions which they refer to as Erlang-Coxian (EC) distributions.
-EC distributions are convenient because they are flexible enough to approximate any distribution, and have a closed form expression for the parameters.
+% Formally, a phase-type distribution as the distribution of \emph{absorption time} in a \emph{continuous-time Markov process} with a single absorbing state~\autocite{hobolthPhasetype}.
+% Given a probability distribution over starting states, a rate at which 
+% The process can be defined by its size, $l$, a $l \times l$ \emph{transition matrix}, $\matr{M} \in \reals^{l \times l}$, an \emph{exit rate} vector, $\vec{\alpha}$ 
+% The process's state at time $t$, $W_t$, is a scalar $W_t \in \{ 1, \dots, l+1 \}$.
+% Its transition matrix is defined, for $j \neq k$, as $\prob(W_{t+\delta t} = k \mid ) = M_{jk} \delta t$ for small, positive $\delta t$.
+% The diagonal is defined as $M_{jj} = - \sum_{k \neq j} M_{jk}$ so that the matrix $\matr{M}$ has row sums 0.
+% The state $l+1$ is an absorbing state, that is $\prob(W_{t+t'} = l+1 \mid W_t = l_1) = 1$ for any non-negative $t'$ (once the process enters it, it cannot leave).
+% The exit rate gives the transition rates into the absorbing state: $\prob(W_{t+\delta t} = l + 1 \mid W_t = j) = \alpha_j \delta t$ for small $\delta t$.
+% The process's absorption time is when it first arrives at the absorbing state, $\tau = \inf \{ t \ssep W_t = l+1 \}$.
+% That is, $P$ is a phase-type distribution if it can be represented as a process 
+% They are additions of arbitrary exponential distributions, and can (if there is no limit on the number of compartments used) represent any distribution.
+% \todo{check/cite these claims}
+% I use the method of \textcite{osogamiClosed} to determine $l$ and the transition rates between the compartments.
 
-\todo[inline]{need more explanation here: what is a phase-type distribution, be less terse in the following text}
-A $l$-phase EC distribution is a convolution of a $(l-2)$-phase gamma distribution and a two-phase acyclic, arbitrary phase-type distribution.
-The time between entering the 1st state and the $l+1$th state, the absorbing state, is the duration.
-A schematic of an EC distribution and its parameters is shown in \cref{SEIR:fig:EC}.
+% A $l$-phase EC distribution is the sum of a $(l-2)$-phase gamma distribution and a two-phase acyclic, arbitrary phase-type distribution.
+An EC distribution $P$ can be defined as the sum of three random variables $K$, $X_1$, and $X_2$.
+It is shown schematically in \cref{SEIR:fig:EC}.
+Here, $K \dist \GamDist(l-2, \kappa_Y)$, $X_1 \dist \Exponential(\kappa_{X1})$ and $X_2$ is a mixture distribution taking the value 0 with probability $1-p_x$ or otherwise (\ie with probability $p_x$) distributed $\Exponential(\kappa_{X2})$.
+% The time between entering the 1st state and the $l+1$th state, the absorbing state, is the duration.
+Gamma distributions with integer first parameters (as $K$ has) are also known as Erlang distributions.
+The distribution $X_1 + X_2$ is a special case of the Coxian distribution.
+Hence, the overall distribution is known as a Erlang-Coxian distribution.
 
 \begin{figure}
+\makebox[\textwidth][c]{
 \begin{tikzpicture}[
@@
 \end{tikzpicture}
-\caption{A $l$ phase EC distribution.}
+}
+\caption[A $l$ phase EC distribution.]{Representation of an EC distribution as a series of compartments.
+The distribution is the time from entering state 1 and entering the absorbing state.
+The time between entering state 1 and state $l-2$ is distributed $K \dist \GamDist(l-2, \kappa_Y)$, also known as an Erlang distribution.
+The time between entering state $l-2$ and the absorbing state has a Coxian distribution (see the main text for details).}
 \label{SEIR:fig:EC}
@@
 
-\Textcite{osogamiClosed} prove that EC distributions have the following properties:
+\Textcite{osogamiClosed} define $P$ as well-matching an arbitrary distribution $G$ if $P$ and $G$ have the same first three moments.
+They then prove that EC distributions have the following properties:
 \begin{itemize}
-    \item Any distribution that can be well-matched (defined as having the same first three moments) by an acyclic phase-type distribution can be well-matched for some $P$ which is an EC distribution.
-    \item The number of phases used by $P$ is at most one more than the optimal number.
+    \item If there exists a phase-type distribution $P'$ that well-matches $G$ then there exists an EC distribution $P$ that well-matches $G$.
     \item The parameters of $P$ are available in closed form, and the expression for these is derived.
+    \item The number of phases used by the $P$ calculated by these expressions is at most one more than the minimum for any such $P'$.
 \end{itemize}
@@
     \item $\kappa_{X2}$, the rate parameter of the second state after the Erlang distribution.
-    \item $p_x$, the probability of moving from the first state after the Erlang distribution to the second (as opposed to directly to the absorbing state).
+    \item $p_x$, the probability of moving from the end state of the Erlang distribution to the second (as opposed to directly to the absorbing state).
 \end{itemize}
@@
 I use the mapfit package~\autocite{mapfit} to compute $P$ for the posterior mean of the distribution estimated in \cref{E-imperf-test}.
+This is the same duration distribution used in \cref{E-backcalc}.
 The parameter estimates are in \cref{SEIR:table:ec-params}.
@@
 
-\todo[inline]{mention that the model is run independently for each region, as in the previous chapter}
-\todo[inline]{maybe put something about the likelihood here}
+\todo[inline]{maybe put something about the likelihood somewhere in this section}
 
@@
 
+$K_{ij}$ and $\beta_{ij}$ appear only as a product and therefore are not individually identifiable.
+Normally, $K_{ij}$ is assumed to be known, for example from a previous contact survey~\autocite[such as][]{mossongSocial}.
+Ideally, all the $\beta_{ij}$ terms would be estimated, but there is normally inadequate data to do so, requiring assumptions to be made to reduce the number of parameters~\autocite[65]{keelingModeling}.
+% $\beta_{ij}$ is then estimated, although often with a parsimonious model\todo{ref / give examples / delete on how age-specfic susceptibility / transmissibility happens}.
+A common assumption is that $\beta_{ij}$ can be decomposed into two independent components such that $\beta_{ij} = \beta \beta^\text{susc}_{i} \beta^\text{inf}_j$.
+For any stratum $i$, $\beta^\text{susc}_i$ is known as the \emph{relative susceptibility} of stratum $i$.
+Informally, it is how vulnerable to infection individuals in stratum $i$ are.
+Formally, define an $a_0$ and $a'_0$ and enforce $\beta^\text{susc}_{a_0}=1$ for identifiability.
+Then $\beta^\text{susc}_i$ is the relative risk of an individual in stratum $i$ being infected, relative to an individual in stratum $a_0$.
+\todo{clarify notation and explaantion here of reference class}
+it is the probability that a susceptible individual in stratum $i$ will be infected upon contact with an infectious individual, relative to a individual in the reference stratum.
+Similarly, $\beta^\text{inf}_i$ is known as the \emph{relative infectiousness} of stratum $i$.
+Informally, it is how likely individuals in stratum $i$ are to pass on the infection.
+Formally, it is the relative risk of an infectious individual in stratum $i$ infecting a susceptible individual upon contact, relative to an individual in the reference stratum.
+These measures could be biological, \eg due to differences in immune systems, or behavioural, \eg if infants are less hygenic than adults.
+
+%Now, we assume that $\beta_{ij} = \beta(t) \beta_{ij}$ where $\beta(t)$ is a non-strata-specific, time-varying component representing precautions taken by the population, for example masking; and $\beta_{ij}$ represents a constant probability of transmission from an infectious individual in strata $j$ to a susceptible individual in strata $i$ given contact, which could, for instance, be affected by differential susceptibility to infection for different age-groups.
+%We arbitrarily choose one $\beta_{ij}$ to be equal to 1 for identifiability and hence the other $\beta_{ij}$ values are relative to this reference class.
+
+\todo[inline]{move the below into a context-specific section}
+Based on prior work\todo{cite}, I assume that all individuals have the same \emph{infectiousness} but that there is a difference in \emph{susceptibility}.
+Therefore, $\beta^\text{inf}_{j} = 1$ for all $j$.
+The prior literature suggests that children are less susceptible than adults, but otherwise there is little difference in susceptibility.
+Therefore, I set $\beta^\text{susc}_i$ to 1 for adult-aged strata and $\beta^\text{susc}_i = \beta_c$ for child-aged strata.
 \subsubsection{Identifiability}
@@
 \begin{figure}
+\makebox[\textwidth][c]{
 \begin{tikzpicture}[
@@
 
-    \path (S) edge node {$\Delta_i$} (E1)
+    \path (S) edge node {$\vec{\lambda} \circ \vec{s}$} (E1)
           (E1) edge node {$\sigma$} (E2)
@@
 
-    \path  (S) edge node {$\Delta_i$} (P0)
+    \path  (S) edge node {$\vec{\lambda} \circ \vec{s}$} (P0)
           (P0) edge node {$\kappa_0$} (P1)
@@
 \end{tikzpicture}
+}
   \caption[SEIR model]{SEIR model used for the analysis in this chapter. The shaded box represents PCR-positive individuals. Shown for a single strata $i$.}
@@
     \label{SEIR:eq:fullODEs}
-    \frac{d\vec{s}}{dt} &= -\vec{\Delta} \\
-    \frac{d\vec{e_1}}{dt} &= \vec{\Delta} - 2\sigma \vec{e_1} \\
-    \frac{d\vec{e_2}}{dt} &= 2\sigma \vec{e_1} - 2\sigma \vec{e_2} \\
-    \frac{d\vec{i_1}}{dt} &= 2\sigma \vec{e_2} - 2\gamma \vec{i_1} \\
-    \frac{d\vec{i_2}}{dt} &= 2\gamma \vec{i_1} - 2\gamma \vec{i_2} \\
-    \frac{d\vec{r}}{dt} &= 2\gamma \vec{i_2} \\
-    \frac{d\vec{p_0}}{dt} &= \vec{\Delta} - \kappa_0 \vec{p_0} \\
-    \frac{d\vec{p_1}}{dt} &= \kappa_0 \vec{p_0} - \kappa_Y \vec{p_1} \\
-    \frac{d\vec{p_2}}{dt} &= \kappa_Y \vec{p_1} - \kappa_{X1} \vec{p_2} \\
-    \frac{d\vec{p_3}}{dt} &= p_x \kappa_{X1} \vec{p_2} - \kappa_{X2} \vec{p_3}
+    \frac{d\vec{s}(t)}{dt} &= -\vec{\lambda}(t) \circ \vec{s}(t) \\\\
+    \frac{d\vec{e_1}(t)}{dt} &= \vec{\lambda}(t) \circ \vec{s}(t) - 2\sigma \vec{e_1}(t) \\
+    \frac{d\vec{e_2}(t)}{dt} &= 2\sigma \vec{e_1}(t) - 2\sigma \vec{e_2}(t) \\
+    \frac{d\vec{i_1}(t)}{dt} &= 2\sigma \vec{e_2}(t) - 2\gamma \vec{i_1}(t) \\
+    \frac{d\vec{i_2}(t)}{dt} &= 2\gamma \vec{i_1}(t) - 2\gamma \vec{i_2}(t) \\
+    \frac{d\vec{r}(t)}{dt} &= 2\gamma \vec{i_2}(t) \\
+    \frac{d\vec{p_0}(t)}{dt} &= \vec{\lambda} \circ \vec{s}(t) - \kappa_0 \vec{p_0}(t) \\
+    \frac{d\vec{p_1}(t)}{dt} &= \kappa_0 \vec{p_0}(t) - \kappa_Y \vec{p_1}(t) \\
+    \frac{d\vec{p_2}(t)}{dt} &= \kappa_Y \vec{p_1}(t) - \kappa_{X1} \vec{p_2}(t) \\
+    \frac{d\vec{p_3}(t)}{dt} &= p_x \kappa_{X1} \vec{p_2}(t) - \kappa_{X2} \vec{p_3}(t) \\
+    \vec{\lambda}(t) &= (\matr{\beta}(t) \circ \matr{\kappa}(t)) (\vec{i_1}(t) + \vec{i_2}(t))
 \end{align}
-where $\vec{\Delta} = \vec{s} \circ (\matr{\beta} \circ \matr{M}) (\vec{i_1} + \vec{i_2})$ is the vector of instantaneous incidence.
-Recall that each region is run independenlty (see \cref{SEIR:sec:geography}), $\matr{beta}_{ij} = \beta(t) \beta^\text{susc}_i$ with $\beta^\text{susc}_i = \beta_c$ for child-aged strata and $\beta^\text{susc}_i = 1$ for adult-aged strata (see \cref{SEIR:sec:structured-populations}), and that $\beta(t)$ and $\matr{M}$ are time-varying (see \cref{SEIR:sec:time-varying-foi}).
+Recall that each region is run independently (see \cref{SEIR:sec:geography}), $\matr{beta}(t)_{ij} = \beta(t) \beta^\text{susc}_i$ with $\beta^\text{susc}_i = \beta_c$ for child-aged strata and $\beta^\text{susc}_i = 1$ for adult-aged strata.
+
+\subsection{Solving the systems of ODEs}
+I solve the system of ODEs using Euler's method, discretised at timesteps of $\delta t = 0.5$days.
+That is to find the state of the system at time $t+1/2$ given the state at time $t$, I add half the derivative to the state at time $t$.
+\todo[inline]{add these equations}
+
+The prevalence on day $t$, used for the likelihood, is the mean of the solution on the two corresponding days.
 
 \subsection{Priors}
-The priors for most parameters are shown in \cref{SEIR:table:priors}.
-The parameters governing the duration of positivity ($\kappa$s) were explained previous (see \cref{SEIR:table:ec-params}).
-\todo[inline]{Explain other priors (identifiability or weakly informative)}
+Priors for the parameters are given in \cref{SEIR:table:priors}.
+The parameters governing the duration of positivity ($\kappa$s) were explained previously (see \cref{SEIR:table:ec-params}).
+The initial conditions and transition rate parameters are unindentifiable in a SEIR model~\autocite{dankwaStructural}, therefore I used assumed values or strongly informative priors on them.
+All other parameters had weakly informative priors.
 
@@
     \hline \\
-    Mean latent period & $1/\sigma$ & 3.5 days (fixed) & From a paper \\
-    Mean infectious period & $1/\gamma$ & 4 days (fixed) & From a paper \\
+    Mean latent period & $1/\sigma$ & 3.5 days (fixed) & \Textcite{zhaoEstimating} \\
+    Mean infectious period & $1/\gamma$ & 4 days (fixed) & \Textcite{zhaoEstimating} \\
     Initial proportion without immunity & $\vec\pi$ & See \cref{SEIR:table:immunity-prior} & \\
@@
     Initial growth rate & $\psi$ & $\N(0.06, 0.04)$ & Weakly informative \\
-    Standard deviation of random walk & $\sigma$ & $\Exponential(80)$ & Weakly informative penalised model complexity prior~\autocite{simpsonPenalising} \\
+    Standard deviation of random walk & $\sigma$ & $\Exponential(80)$ & Weakly informative \\ %penalised model complexity prior~\autocite{simpsonPenalising} \\
     Overdispersion of observations & $\rho$ & $\Exponential(2 \times 10^5)$ & Weakly informative
 \end{tabular}
-\todo[inline]{Find citation for children, add priors for $\pi$}
+\todo[inline]{Find citation for children}
 \caption[SEIR model priors]{Priors for each parameter. For details of the distributions and their parameterisations see \cref{E-distributions}.}
@@
 
-\begin{table}[]
+\begin{table}
 \centering
@@
 
-\subsection{Solving the systems of ODEs}
-I solve the system of ODEs using Euler's method, discretised at half day timesteps.
-That is to find the state of the system at time $t+1/2$ given the state at time $t$, I add half the derivative to the state at time $t$.
+\subsection{Data}
 
-The prevalence on day $t$, used for the likelihood, is the mean of the solution on the two corresponding days.
+As in \cref{E-backcalc}, I use CIS data to estimate the model for each region independently.
+As well as the motivation given previously, this allows heterogeneity in the parameters between regions.
+This heterogeneity could arise from a range of socioeconomic factors.
+Furthermore, the assumption of homogeneous mixing within age groups is more plausible for regions than the whole of England.
+
+By having indpendent models for each region, I assume that the epidemics in each region do not interact.
+Clearly, this is not true as travel between regions of England is common and unrestricted.
+However, for a large epidemic, the non-interaction is a good approximation~\autocite{birrellRealtimea}.
+
+The test results are aggregated to age group and region on each day.
 

### 2023-11-18 SEIR.tex
first go at CIS SEIR figs

@@
 
+For each region, I ran 4 chains for 1,000,000 iterations.
+The first 500,000 were discarded as burn-in, with the remaining chains thinned to every 100th iteration.
+This gives 20,000 posterior samples per region.
+This was further thinned to 2,000 samples when calculating posterior predictive quantities (incidence and prevalence) due to their computational cost.
+
 \subsection{Parameterisation}
@@
     Initial growth rate & $\psi$ & $\N(0.06, 0.04)$ & Weakly informative \\
-    Standard deviation of random walk & $\sigma$ & $\Exponential(80)$ & Weakly informative \\ %penalised model complexity prior~\autocite{simpsonPenalising} \\
+    Standard deviation of random walk steps on log scale & $\sigma_\epsilon$ & $\Exponential(80)$ & Weakly informative \\ %penalised model complexity prior~\autocite{simpsonPenalising} \\
     Overdispersion of observations & $\rho$ & $\Exponential(2 \times 10^5)$ & Weakly informative
 \end{tabular}
-\todo[inline]{Find citation for children}
+\todo[inline]{Dani/Paul: I think I took the prior for $\beta_c$ from RTM, but can't find the reference now - any ideas?}
 \caption[SEIR model priors]{Priors for each parameter. For details of the distributions and their parameterisations see \cref{E-distributions}.}
@@
 This is because the prior used for the main analysis was too weak on this prior to create realistic simulations.
+Priors and population sizes were matched to the East of England region, over the same time period as the main analysis.
 
-Simulating from the prior then attempting to recover the parameters is the basis of simulation-based calibration~\autocite{taltsValidating}.
-The coverage of the posterior credible intervals should be their nominal value.
+Simulating from the prior then attempting to recover the parameters is the basis of simulation-based calibration~\autocites{cookValidation}{taltsValidating}.
+The coverage of the posterior credible intervals should be their nominal value (\eg 95\% of the 95\% credible intervals should contain the true paramter value).
 Furthermore, the combination of the posteriors across all simulations should recover the prior.
 
-To estimate the posterior, I ran a single MCMC chain for 500,000 iterations, discarding the first half as burn-in.
+To estimate the posterior, I ran a single MCMC chain per simulated data set for 500,000 iterations, discarding the first 300,000 iterations as burn-in, and thinning every 100 iterations.
+Therefore, posterior calculations are based on 2,000 samples per data set.
 On some occasions, the proposal distribution's covariance matrix became singular (or close to singular).
 This means it is not possible to propose a new value.
-In these cases, I restarted the chain with a new seed.
+In these cases, I restarted the chain with a new seed, discarding all information from the original chain; this was required at most once per chain.
 
 \section{Application to CIS data} \label{SEIR:sec:application}
+
+All regions' MCMC chains converged satisfactorily.
+Two chains had to be restarted with new seeds to achieve this as they converged to locations which were clearly incorrect (not shown).
+Following this, the Rhat for all parameters except $\sigma_\epsilon$ had Rhat < 1.01.
+All $\sigma_\epsilon$ parameters had Rhat < 1.05.
+All parameters had ESS > 500, except $\sigma_\epsilon$ in London which had ESS of 182.
+Since $\sigma_\epsilon$ is not a parameter of interest, and all other parameters had sufficient ESS and Rhat, I considered the chains to have converged.
+
+\begin{figure}
+    \thisfloatpagestyle{empty}
+    \vspace{-3cm}
+    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_young}}
+    \captionsetup{width=0.8\paperwidth}
+    \captionof{figure}[Posterior predictive prevalence (younger ages)]{%
+        Posterior predictive prevalence by region and age for the younger three age groups (see \cref{SEIR:fig:prev-old} for the others).
+        The central ribbon and line shows the posterior median and 95\% CrI estimate for the proportion of the relevant strata PCR positive on each day.
+        The error bar is the 95\% CrI including the noise due to the sampling (\ie observation process), aggregated by week.
+        The dots show the observed prevalence in CIS, aggregated by week.
+    }
+    \label{SEIR:fig:prev-young}
+\end{figure}
+\begin{figure}
+    \thisfloatpagestyle{empty}
+    \vspace{-3cm}
+    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_old}}
+    \captionsetup{width=0.8\paperwidth}
+    \captionof{figure}[Posterior predictive prevalence (older ages)]{%
+        Posterior predictive prevalence for the older three age groups (see \cref{SEIR:fig:prev-young} for the others and details of the plot).
+    }
+    \label{SEIR:fig:prev-old}
+\end{figure}
 

### 2023-11-18 SEIR/CIS/posterior_predictive.R

@@ -1,0 +1,133 @@
+library(dplyr)
+library(ggplot2)
+library(patchwork)
+library(purrr)
+library(stringr)
+library(tidybayes)
+library(tidyr)
+source(here::here("utils.R"))
+
+predictions = readr::read_csv(here::here("SEIR/CIS/predictive.csv")) |>
+    mutate(
+        .chain = factor(chain + 1),
+        .iteration = iteration + 1,
+        .draw = max(.iteration) * (as.integer(.chain) - 1) + .iteration,
+        age = str_replace(age, fixed("+"), "Inf")
+    ) |>
+    select(!c(chain, iteration))
+
+data = readr::read_csv(here::here("SEIR/CIS/data.csv")) |>
+    mutate(
+        obs_prevalence = obs_positives / num_tests,
+        age = str_replace(age, fixed("("), "[") |>
+            str_replace(fixed("]"), ")")
+    ) |>
+    rename(day = date) |>
+    filter(
+        day >= min(predictions$day),
+        day <= max(predictions$day),
+    )
+
+prediction_intervals = predictions |>
+    group_by(region, day, age) |>
+    median_qi()
+
+# get theta parameter values
+thetas = readr::read_csv(here::here("SEIR/CIS/params.csv")) |>
+    filter(parameter == "theta") |>
+    transmute(
+        region,
+        .chain = factor(chain + 1),
+        .iteration = iteration + 1,
+        theta = exp(value)
+    )
+
+noisy_predict = predictions |>
+    left_join(data, by = join_by(region, day, age)) |>
+    left_join(thetas, by = join_by(region, .chain, .iteration)) |>
+    mutate(
+        predict_out = VGAM::rbetabinom.ab(
+            n = n(),
+            size = num_tests,
+            shape1 = prevalence / theta,
+            shape2 = (1 - prevalence) / theta,
+        )
+    )
+
+weekly_predictions = noisy_predict |>
+    group_by(
+        region, .draw, age,
+        week = lubridate::floor_date(day, unit = "week", week_start = "Monday") + 3,
+    ) |>
+    summarise(
+        n = sum(num_tests),
+        y = sum(obs_positives),
+        p = y / n,
+        pred_y = sum(predict_out),
+        pred_p = pred_y / n,
+    ) |>
+    group_by(region, age, week) |>
+    median_qi()
+
+create_prev_plot = function(age_groups, label) {
+    plot = weekly_predictions |>
+        filter(age %in% age_groups) |>
+        ggplot(aes(week)) +
[…]

### 2023-11-18 SEIR/CIS_params.R

@@ -1,29 +1,0 @@
-library(dplyr)
-library(ggplot2)
-library(patchwork)
-library(purrr)
-library(tidybayes)
-library(tidyr)
-source(here::here("utils.R"))
-
-
-posterior = readr::read_csv(here::here("SEIR/CIS_params.csv")) |>
-    extract(
-        parameter,
-        into = c("parameter", "i"), 
-        regex = "^([^\\[]+)(\\[[0-9]+\\])?$"
-    ) |>
-    transmute(
-        region,
-        .chain = factor(chain + 1),
-        .iteration = iteration + 1,
-        .draw = max(.iteration) * (as.integer(.chain) - 1) + .iteration,
-        parameter,
-        i = if_else(
-            i == "",
-            NA_integer_,
-            # Remove first and last character from i
-            as.integer(substr(i, 2, nchar(i) - 1))
-        ),
-        value
-    )

### 2023-11-18 SEIR/CIS_predictive.R

@@ -1,29 +1,0 @@
-library(dplyr)
-library(ggplot2)
-library(patchwork)
-library(purrr)
-library(tidybayes)
-library(tidyr)
-source(here::here("utils.R"))
-
-
-posterior = readr::read_csv(here::here("SEIR/CIS_predictive.csv")) |>
-    extract(
-        parameter,
-        into = c("parameter", "i"), 
-        regex = "^([^\\[]+)(\\[[0-9]+\\])?$"
-    ) |>
-    transmute(
-        region,
-        .chain = factor(chain + 1),
-        .iteration = iteration + 1,
-        .draw = max(.iteration) * (as.integer(.chain) - 1) + .iteration,
-        parameter,
-        i = if_else(
-            i == "",
-            NA_integer_,
-            # Remove first and last character from i
-            as.integer(substr(i, 2, nchar(i) - 1))
-        ),
-        value
-    )

### 2023-11-20 SEIR.tex
editing SEIR and adding results

@@
 
+Overall the model fits the data well (see \cref{SEIR:fig:prev-young,SEIR:fig:prev-old}).
+There are a few features of the data which are not captured by the model though.
+In particular, prevalence during October is higher in the 16--25s than the model predicts and lower in the 11--16s.
 \begin{figure}
@@
         The central ribbon and line shows the posterior median and 95\% CrI estimate for the proportion of the relevant strata PCR positive on each day.
-        The error bar is the 95\% CrI including the noise due to the sampling (\ie observation process), aggregated by week.
+        The lighter outer ribbon is the 95\% CrI including the noise due to the sampling (\ie observation process), aggregated by week.
         The dots show the observed prevalence in CIS, aggregated by week.
+        For a well-calibrated model, 95\% of the dots would be within the outer ribbon.
     }
@@
     \captionof{figure}[Posterior predictive prevalence (older ages)]{%
-        Posterior predictive prevalence for the older three age groups (see \cref{SEIR:fig:prev-young} for the others and details of the plot).
+        Posterior predictive prevalence for the older three age groups (see \cref{SEIR:fig:prev-young} for the other age groups and details).
     }
@@
 
+The 11--16 age group (roughly corresponding to compulsory secondary school education) is estimated to have the highest incidence (see \cref{SEIR:fig:incidence}).
+Incidence is smooth within each week but shows discontinuities between weeks.
+This is because the parameters within $\vec{\lambda}(t)$ are constant within each week.
+A constant $\vec{\lambda}(t)$ leads to exponential growth over short time periods, once the initial dynamics have died out.
+When the peak in incidence is estimated to occur varied by region, but due to this reason, is almost always on a Friday (see \cref{SEIR:fig:peak-incidence}).
+
+\begin{figure}
+    \includegraphics{SEIR/CIS/incidence}
+    \caption[Posterior estimates of incidence]{%
+        Posterior estimates (95\% CrI) of incidence by region and age.
+        See main text for discussion.
+    }
+    \label{SEIR:fig:incidence}
+\end{figure}
+
+The parameters $\beta_t$ absorb all the time-varying changes in transmission not captured by Google mobility and school attendance data.
+They trend downwards up until early December when they start increasing again.
+Each region is estimated independently, however correlation is clear across the regions.
+\begin{figure}
+    \includegraphics{SEIR/CIS/beta_walk}
+    \caption[Posterior estimates of $\beta_t$]{%
+        Posterior estimates $\beta_t$ for each week.
+        Shown are "violins" where the width is proportional to the kernel-density estimated posterior density at that point.
+        Values trend downwards until early December when an increase starts.
+    }
+    \label{SEIR:fig:beta-walk}
+\end{figure}
+
 \section{Discussion} \label{SEIR:sec:discussion}
+
+Parameterisation of this model is limited by the inference techniques.
+The algorithm used here tends to struggle with a large number of parameters.
+Alternative proposal schemes, such as blocking\todo{cite what RTM does?}, could help, or a more advanced algorithm such as NUTS (used elsewhere in this thesis).
+
+The current model is not very flexible with regards to the age distribution of infections.
+The lack of flexibility is visible in \cref{SEIR:fig:prev-young} where how well the model reproduces the data varies by age group.
+In some regions, there appears to be a trade-off between the 11--16s and 16--25s, with the model misfitting in opposite directions.
+The age distribution of infections determined by $\matr{\beta}$, and to a lesser extent $\vec{\pi}$, the former of which only contains one inferred parameter, $\beta_c$.
+More parameters would help to capture these dynamics more accurately.
+Increasing the number of parameters inferred in the contact matrices would allow for more flexibility in the age distribution of infections, either by introducing additional susceptibility or infectiousness parameters, or by allowing these parameters to vary over time.
+For example, the introduction and relaxation of control measures within schools over time is a plausible mechanism changing the age distribution of infections that cannot be captured in the current model because it only uses school attendance data with no allowance for the changing risk of within-school contacts.
+
+The weekly patterns in incidence are another limitation of the current parameterisation.
+The random walk and contact matrices both vary only on the weekly timescale.
+A daily random walk could provide more flexibility here.
+However, this would greatly increase the number of parameters.
+An alternative would be a smoothly-varying function or an approximation of a random walk such as that proposed by \textcite{ghoshApproximate}.
+
+The parameters $\beta_t$ absorb all the time-varying factors not captured by Google mobility and school attendance data.
+These factors are likely to be correlated across the country, giving the similar patterns for different regions (see \cref{SEIR:fig:beta-walk}).
+Exploiting these correlations by linking their values across regions could improve the precision of the estimates.
+This would require a more complex model without the independence of parameters between regions.
+
+Various factors could explain the pattern in the $\beta_t$ values, which have a minima near the start of December.
+Some hypothesis follow, although this analysis is unable to distinguish causality.
+Increased caution as cases and deaths increased could explain the decrease (\eg masking or stricter adherence to social distancing).
+The increase coincides with the rise of the Alpha variant, known to be more transmissible.
+This is a likely cause of the increase.
+A multiple variant model might be able to capture this more accurately.
+This would stratify positive tests in CIS by pre-Alpha or Alpha variant.
+Then, each variant would be explicitly included in the model to fit to this data.
+
+Extending the estimates here to the rest of the pandemic would clearly be of interest.
+However, this requires a more complex model structure.
+I chose the end of the modelling period to coincide with the vaccination rollout.
+Vaccination dramatically affects the transmission dynamics by introducing some additional immunity into the population.
+Mechanistic models require explicit representation of this immunity\todo{cite something about vaccination}.
+
+\todo[inline]{move the following discussion to conclusion chapter as relevant to whole thesis}
+Household structure is not included in this model.
+The household-based design of the survey is likely to induce clustering in results, increasing the variance of the measured prevalence.
+The beta-binomial likelihood structure should capture some of this variance within each strata\todo{ref beta-binomial as a model for clustering}, but not between strata.
 

### 2023-11-20 SEIR/CIS/posterior_predictive.R

@@ -1,4 +1,5 @@
 library(dplyr)
+library(lubridate)
 library(ggplot2)
 library(patchwork)
 library(purrr)
@@ -57,7 +58,7 @@ noisy_predict = predictions |>
 weekly_predictions = noisy_predict |>
     group_by(
         region, .draw, age,
-        week = lubridate::floor_date(day, unit = "week", week_start = "Monday") + 3,
+        week = floor_date(day, unit = "week", week_start = "Monday") + 3,
     ) |>
     summarise(
         n = sum(num_tests),
@@ -74,7 +75,7 @@ create_prev_plot = function(age_groups, label) {
         filter(age %in% age_groups) |>
         ggplot(aes(week)) +
         geom_point(aes(y = p)) +
-        geom_errorbar(aes(ymin = pred_p.lower, ymax = pred_p.upper)) +
+        geom_ribbon(aes(ymin = pred_p.lower, ymax = pred_p.upper), alpha = 0.1) +
         geom_lineribbon(
             aes(day, prevalence, ymin = prevalence.lower, ymax = prevalence.upper),
             data = filter(prediction_intervals, age %in% age_groups),
@@ -105,29 +106,57 @@ old_ages = unique(weekly_predictions$age)[4:6]
 create_prev_plot(young_ages, "young")
 create_prev_plot(old_ages, "old")
 
-create_incidence_plot = function(age_groups, label) {
-    plot = prediction_intervals |>
-        filter(age %in% age_groups) |>
-        ggplot() +
-        geom_lineribbon(
-            aes(day, incidence),
-            data = prediction_intervals,
-            alpha = 0.4
-        ) +
-        labs(
-            x = "Date",
-            y = "Incidence"
-        ) +
-        standard_plot_theming() +
-        theme(legend.position = "none")
-    ggsave(
-        filename = here::here(
-            glue::glue("SEIR/CIS/incidence_{label}.png")
+p_incidence = prediction_intervals |>
+    ggplot() +
+    geom_ribbon(
+        aes(
+            day,
+            ymin = incidence.lower,
+            ymax = incidence.upper,
+            fill = age
         ),
-        plot = plot,
-        width = 15,
-        height = 20,
-        units = "cm",
-        dpi = 300
-    )
-}
+        alpha = 0.4
+    ) +
+    facet_wrap(~region, ncol = 2) +
+    labs(
+        x = "Date",
[…]

### 2023-11-20 SEIR/CIS/random_walk.R

@@ -1,0 +1,89 @@
+library(dplyr)
+library(ggplot2)
+library(patchwork)
+library(purrr)
+library(lubridate)
+library(tidybayes)
+library(tidyr)
+source(here::here("utils.R"))
+
+start_date = ymd("2020-08-31")
+
+posterior = readr::read_csv(here::here("SEIR/CIS/params.csv")) |>
+    filter(stringr::str_starts(parameter, "beta")) |>
+    extract(
+        parameter,
+        into = c("parameter", "i"), 
+        regex = "^([^\\[]+)(\\[[0-9]+\\])?$"
+    ) |>
+    transmute(
+        region,
+        .chain = factor(chain + 1),
+        .iteration = iteration + 1,
+        .draw = max(.iteration) * (as.integer(.chain) - 1) + .iteration,
+        i = if_else(
+            i == "",
+            NA_integer_,
+            # Remove first and last character from i
+            as.integer(substr(i, 2, nchar(i) - 1))
+        ),
+        value = if_else(
+            i == 0,
+            0,
+            value
+        )
+    )
+
+walk = posterior |>
+    group_by(region, .draw) |>
+    arrange(i, .by_group = TRUE) |>
+    mutate(
+        beta = value |>
+            cumsum() |>
+            exp()
+    ) |>
+    ungroup()
+
+walk_summary = walk |>
+    group_by(region, i) |>
+    median_qi()
+
+walk_summary |>
+    mutate(date = start_date + i * 7) |>
+    ggplot(aes(date, beta, color = region)) +
+    geom_errorbar(aes(ymin = beta.lower, ymax = beta.upper), alpha = 0.5) +
+    facet_wrap(~region) +
+    standard_plot_theming() +
+    scale_y_log10(
+        breaks = c(0.01, 0.1, 1, 10, 100)
+    )
+
+
+# A violin plot of the posterior distribution of beta for each region and week.
+# The x-axis is "continuous" so that ggplot2 does something sensible with the
+# tick labels but "discrete" for the violins so that there is a separate violin
+# per week.
+p_walk = walk |>
+    mutate(date = start_date + i * 7) |>
+    ggplot(aes(factor(date), beta)) +
+    # geom_violin(aes(factor(date)))
+    # stat_eye(point_size = 0.1) +
+    stat_slab(side = "both") +
+    facet_wrap(~region, ncol = 2) +
+    standard_plot_theming() +
[…]

### 2023-11-20 SEIR.tex

@@
 
-This chapter uses mechanistic models for estimating the transmission of SARS-CoV-2.
-This is an alternative approach to the statistical methods used in \cref{E-backcalc}.
-Mechanistic models explicitly represent the infection process driving the spread of infectious disease.
-The explicit representation means that a mechanistic model's state and parameters have biological and/or epidemiological interpretations; these interpretations provide scientific understanding of the disease.
-Of particular interest is that the basic reproduction number, $R_0$, and the effective reproduction number, $R_e(t)$, can both be calculated as simple functions of the model parameters; using \cref{E-backcalc} additional modelling would be required to estimate these numbers.
-Furthermore, mechanistic models can be used for scenario-based modelling, whereby the effect of different interventions can be simulated.
+This chapter uses mechanistic models for estimating the transmission of SARS-CoV-2, an alternative to the statistical approach in \cref{E-backcalc}.
+Mechanistic models explicitly represent the infection process driving the spread of infectious disease, giving model's state and parameters have a biological and/or epidemiological interpretation.
+Of particular interest is that the basic reproduction number, $R_0$, and the effective reproduction number, $R_e(t)$, which can both be calculated as functions of the model parameters; using \cref{E-backcalc}'s approach, additional modelling would be required to estimate these numbers.
+Furthermore, mechanistic models can be used for scenario-based modelling, which simulates the effect of interventions.
 All of these properties make mechanistic models a useful tool for understanding and controlling infectious diseases.
@@
 % This framework is then applied to SARS-CoV-2 in \cref{SEIR:sec:transmission-application}.
-It describes the \emph{transmission model}, corresponding to the infection process (see \cref{E-inc-prev:sec:infection-process}).
+It describes the \emph{transmission model}, corresponding to the infection process (from \cref{E-inc-prev:sec:infection-process}).
 The transmission model needs to be linked to observations; in this chapter, that will be the CIS prevalence data.
-The \emph{observation model}, introduced in \cref{SEIR:sec:observation}, is this link; it incorporates both the prevalence process (see \cref{E-inc-prev:sec:prevalence-process}) and the observation process (see \cref{E-inc-prev:sec:observation-process}).
-\Cref{SEIR:sec:inference-implementation} describes the inference process, parameterisation, and priors for this model.
-I ensure that the model can recover the parameters successfully in a simulation study in \cref{SEIR:sec:sim-study} before applying it to the CIS data in \cref{SEIR:sec:application}, which includes comparisons to the results in \cref{E-backcalc}.
+The \emph{observation model}, introduced in \cref{SEIR:sec:observation}, is this link; it incorporates both the prevalence process (from \cref{E-inc-prev:sec:prevalence-process}) and the observation process (from \cref{E-inc-prev:sec:observation-process}).
+\Cref{SEIR:sec:inference-implementation} describes how I will apply the model to the CIS prevalence data, including the inference process, parameterisation, and priors for this model.
+\Cref{SEIR:sec:sim-study} checks this recovers the parameters successfully using a simulation study.
+\Cref{SEIR:sec:application} then describes the results with the CIS data, including making comparisons to the results in \cref{E-backcalc}.
 Finally, \cref{SEIR:sec:discussion} discusses the results, limitations, and extensions of this work.
@@
 The literature on mechanistic models is vast, having been developed for over a century; the model in \cref{SEIR:sec:SIR} was first formulated in the 1920s~\autocite{kermackContribution}.
-This chapter explains only the background I will use in the application to SARS-CoV-2.
+\Cref{SEIR:sec:transmission-generic} explains only the parts relevant to the remainder of the chapter.
 For further details I recommend the tutorial paper \textcite{kretzschmarMathematical} or the textbook \textcite{keelingModeling}.
@@
 Notation in this chapter differs from the rest of this thesis.
-I adopt conventions and notation from the infectious disease modelling field, although this is not consistent in the literature.
-In particular, case no longer signifies whether a variable is random or not, but will be clear from the context.
+I adopt conventions and notation familiar to the infectious disease modelling field, although the field's notation is not consistent between sources.
+In particular, upper or lower case no longer signifies whether a variable is random or not, but this is clear from the context.
 
@@
 
-\todo[inline]{ref intro to compartmental model setup}
+\todo[inline]{I think this intro needs references but I'm struggling to find suitable ones}
 This section introduces the class of mechanistic transmission models known as \emph{compartmental} models.
 Compartmental models are the most widely-used class of mechanistic infectious disease model.
-A compartmental model is named because they place every member of the population under study into a compartment; each compartment represents a disease state (\eg susceptible).
+A compartmental model is named because they place every member of the population under study into a compartment; each compartment representing a disease state (\eg susceptible).
 
-A compartmental model consists of a state vector $\vec{x}(t)$ and a set of parameters $\vec{\theta}$.
+A compartmental model consists of a state vector $\vec{x}(t)$ and a set of parameters $\vec{\theta}$~\autocite{birrellEvidence}.
 The elements of $\vec{x}(t)$ are the proportion of the population (or number of individuals) in each compartment at time $t$.
-The parameters $\vec{\theta}$ are constants that determine, along with the model structure, how the state vector changes over time.
+The parameters $\vec{\theta}$ determine, along with the model structure, how the state vector changes over time.
+Inference about $\vec{x}(t)$ and $\vec{\theta}$ is through an observation model, $p(\vec{y}(t) \mid \vec{x}(t), \vec{\theta})$ which links $\vec{x}(t)$ to some observation $\vec{y}(t)$~\autocite{birrellEvidence}; I discuss observation models further in \cref{SEIR:sec:observation}.
+
 Compartmental models can be \emph{deterministic} or \emph{stochastic}.
-Deterministic models are defined by the state vector $\vec{x}(t)$ being a deterministic, although rarely analytically-tractable, function of the parameters $\vec{\theta}$ and the initial state $\vec{x}(0)$.
+Deterministic models are defined by the state vector $\vec{x}(t)$ being a deterministic function of the parameters $\vec{\theta}$ and the initial state $\vec{x}(0)$; equivalently, $\var(\vec{x}(t) \mid \theta) = 0$~\autocite{birrellEvidence}.
+In most situations, the function is not analytically-tractable and requires numeric methods to solve.
 %The transmission model corresponds to the infection process (see \cref{E-inc-prev:sec:infection-process}).
@@
 Stochastic models are most commonly specified in discrete time, using a Markov assumption.
-That is, they specify a probability distribution $p(\vec{x}(t) \mid \vec{x}(t-1), \vec{\theta})$.
+That is, they specify a probability distribution $p(\vec{x}(t) \mid \vec{x}(t-1), \vec{\theta})$ for integer $t$ and assume that, having conditioned on $\vec{x}(t-1)$, $\vec{x}(t)$ is independent of $\vec{x}(t')$ for all $t' < t - 1$~\autocite{birrellEvidence}.
 This relaxes the assumption of independent increments in the counting process formulation of the infection process (see \cref{E-inc-prev:sec:infection-process}).
@@
 The remainder of this chapter uses only deterministic models, because throughout the time period this thesis considers there were enough infectious individuals.
-For an introduction to stochastic models, see \textcite[chapter 6]{keelingModeling}.
+For an introduction to stochastic models, see \textcite[chapter 6]{keelingModeling} or \textcite{birrellEvidence}.
 % \todo{Maybe say something about still having a stochastic observation model, if not clear enough from intro}

### 2023-11-21 ATACCC.tex
editing

@@
 \Cref{ATACCC:sec:background} explains how biomarkers are typically modelled, both for SARS-CoV-2 and historically in the biostatistical literature.
-Next, \cref{ATACCC:sec:hakki} explains \textcite{hakkiOnset}'s application of these methods to the ATACCC data.
-However, this analysis did not converge fully.
-I modify their observation model in section \cref{ATACCC:sec:observation-modification} to improve the convergence.
+Next, \cref{ATACCC:sec:hakki} explains \textcite{hakkiOnset}'s application of these methods to the ATACCC data and in Section \cref{ATACCC:sec:observation-modification} the observation model of Hakki is modified  to improve convergence.
 This modification is based on an earlier analysis I did with a slightly different model and a preliminary version of the dataset, described in \cref{ATACCC:sec:original-analysis}.
@@
 
-The second approach is to approximate the dynamics using general mathematical functions.
+The second approach is to approximate the trajectories from these models using general mathematical functions.
+In these models, the function's parameters are clearly related to features of the biomarker's trajectory, such as the time from first detection to peak viral load, and the viral load at the peak.
 Most common is a piecewise model, where the logarithm of viral load increases linearly, peaks instantaneously, and then decreases linearly~\autocites{clearyUsing,kisslerViral,larremoreTest}; although more flexible functions have also been suggested~\autocites{quiltyQuarantine}.
-In these models, parameters are clearly related to features of the biomarker's trajectory, such as the time from first detection to peak viral load, and the viral load at the peak.
+
 In both approaches, random effects are used to represent variation between hosts within the population, capturing the non-independence of longitudinal data from the same host.
@@
 In the Bayesian context, random effect models are often referred to as hierarchical or multilevel models because they can be interpreted as placing a prior at two levels of a parameter hierarchy~\autocite[chapter 5]{gelmanBDA}.
-$\phi_j$ would be referred to as the bottom level of the hierarchy, with $\mu_j$ and $\sigma_j$ being the top level.
+$\phi_j$ would be referred to as the bottom level of the hierarchy, with $\mu$ and $\sigma$ being the top level.
 More complex models could have more levels, for instance if the individuals are grouped into subpopulations.
@@
 
-The model of \textcite{hakkiOnset} has two components: a viral load model and an observation model.
-The viral load model produces a determinstic trajectory for the viral load of an individual, based on the model parameters.
-The observation model describes how the PCR test noisly observes this viral load.
-It includes the possibility of false positive and negatives, and the limit of detection of the PCR test.
+The model of \textcite{hakkiOnset} has two components: a latent viral load model and an observation model.
+The viral load model produces a deterministic trajectory for the viral load of an individual, based on the model parameters.
+The observation model then describes how the PCR testing process produces noisy observations of this viral load, including the possibility of false positive and negatives, and the limit of detection of the PCR test.
 \todo{Make sure limit of detection is introduced previously when I first talk about PCR}
@@
 
-Viral loads exhibit exponential growth followed by exponential decline.
+Viral load exhibits exponential growth followed by exponential decline.
 \Textcite{hakkiOnset} use a model which is approximately piecewise linear but continuous.
@@
 \begin{align}
-\hat{v}_i(\tau) = v_{\max,i} \frac{a_i+b_i}{b_ie^{-a_i(\tau-\tau_{\max,i})} + a_ie^{b_i(\tau-\tau_{max,i})}}. \label{ATACCC:eq:viral-load}
+v_i(\tau) = v_{\max,i} \frac{a_i+b_i}{b_ie^{-a_i(\tau-\tau_{\max,i})} + a_ie^{b_i(\tau-\tau_{max,i})}}. \label{ATACCC:eq:viral-load}
 \end{align}
@@
 $v_{\max,i}$ is the maximum viral load, which occurs at time $\tau_{\max,i}$.
-As time moves away from the peak (\ie becomes large in magnitude), the viral load trajectory tends towards exponential growth (before the peak) or decline (after the peak) in viral load at rates $a_i$ and $b_i$ respectively.
+As $|\tau - tau_\max|$ becomes large, the viral load trajectory tends towards exponential growth (before the peak) or decline (after the peak) in viral load at rates $a_i$ and $b_i$ respectively.
 Before the peak, this is because $b(\tau-\tau_{\max,i})$ is small and hence $a\exp(b(\tau-\tau_{\max,i}))$ is negligible.
@@
 
-The noisy observations are modelled by assuming each measurement of log viral load has iid Normally-distributed error with mean 0 and standard deviation $\sigma_v$.
-The measurements are censored at the limit of detection, $\vlod$, meaning that any observation which would give a log viral load of less than $\vlod$ is instead observed as negative.
+The noisy observations are modelled by assuming each measurement of log viral load is a realisation of a iid Normally-distributed random variable error with mean 0 and standard deviation $\sigma_v$.
+The measurements are left-censored at the limit of detection, $\vlod$, meaning that any observation which would give a log viral load of less than $\vlod$ is instead observed as negative.
 False positives and negatives are modelled by allowing a test failure with probability $\rho$.
@@
     \rho f_N(y_{i,t} \mid x_0, \sigma_0^2) + (1 - \rho) f_N(y_{i,t} \mid \log \hat{v}_i(t), \sigma_v^2) & \text{if $y_{i,t} > 0$}\\
-    \rho F_N(v_\text{log} \mid x_0, \sigma_0^2) + (1 - \rho) F_N(\vlod \mid \log \hat{v}_i(t), \sigma_v^2) & \text{if $y_{i,t} = 0$} \label{ATACCC:eq:hakki-likelihood}
+    \rho F_N(v_\text{lod} \mid x_0, \sigma_0^2) + (1 - \rho) F_N(\vlod \mid \log \hat{v}_i(t), \sigma_v^2) & \text{if $y_{i,t} = 0$} \label{ATACCC:eq:hakki-likelihood}
 \end{cases}
@@
 
-Biological plausibility requires constraining $\vec{\phi}_i$ to the positive reals.
+Biological plausibility requires constraining $\vec{\phi}_i$ to positive real values.
 Each $\phi_i$ is modelled as a log-normal random effects by using the hierarchical prior $\phi_i \dist \LN(\vec{\mu}_{\phi,g(i)}, \matr{\Sigma}_\phi)$ where $g(i)$ is the group of individual $i$.
@@
 
-\section{Modifying the observation model} \label{ATACCC:sec:observation-modification}
+\section{Modifying the Hakki observation model} \label{ATACCC:sec:observation-modification}
 
@@
 The second, found by the other six chains, is more symmetrical, peaking 7.2 days after the first test.
-Because the individual trajectories is not of interest for this analysis, and the population-level parameters agree between the two sets of chains, I consider this lack of convergence acceptable.
+Because the individual trajectories are not of interest for this analysis, and the population-level parameters agree between the two sets of chains, I consider this lack of convergence acceptable.
 \begin{figure}

### 2023-11-21 SEIR.tex
editing

@@
 All of these properties make mechanistic models a useful tool for understanding and controlling infectious diseases.
-However, they make stronger assumptions about the disease than statistical models, and often perform worse in short-term forecasting\todo{add a citation}.
+However, they make stronger assumptions about the disease and population behaviour than statistical models.
 
@@
 
-In particular, this chapter discusses deterministic models whose evolution is specified by a system of ODEs, $d\vec{x}/dt$.
+This chapter discusses deterministic models whose evolution is specified by a system of ODEs, $d\vec{x}/dt$.
 Therefore, in an ODE model, time is necessarily continuous, although discretised when solving (see \cref{SEIR:sec:inference-implementation}).
@@
 
-\Cref{SEIR:sec:SIR} introduces the \emph{susceptible-infected-recovered} (SIR) model, a simple compartmental model for a disease which confers immunity.
+\Cref{SEIR:sec:SIR} introduces the \emph{susceptible-infected-recovered} (SIR) model, a simple compartmental model for a disease where individuals are immune to reinfection following their recovery.
 The SIR model is named because each individual in the population is divided into one of three states: susceptible, infectious, or recovered.
 This model has several assumptions which make it inappropriate to model SARS-CoV-2; therefore, the remainder of this sections relaxes these assumptions.
-First, \cref{SEIR:sec:non-exponential} relaxes the assumption of an exponentially-distributed infectious period to a gamma distribution.
-Then, \cref{SEIR:sec:SEIR} relaxes the assumption that individuals are infectious immediately after they are infected; it does this by extending the SIR model to the SEIR model, including an \emph{exposed} compartment.
+First, \cref{SEIR:sec:non-exponential} relaxes the assumption that the infectious period is exponentially-distributed to allow gamma distributions.
+Then, \cref{SEIR:sec:SEIR} relaxes the assumption that individuals are infectious immediately after they are infected; it does this by extending the SIR model to the SEIR model with the additional E signifying an \emph{exposed} compartment.
 Finally, \cref{SEIR:sec:structured-populations} allows heterogeneity in the population, introducing stratification by important characteristic(s).
@@
 where $\lambda(t) = \beta i(t)$, is the \emph{force of infection}: the risk of infection for a susceptible individual at time $t$~\autocite[17]{keelingModeling}.
-Two parameters are present in this model: $\beta$ and $\gamma$, \ie $\vec{\theta} = (\beta, \gamma)^T$.
+Two parameters are present in this model, $\vec{\theta} = (\beta, \gamma)^T$.
 $\beta$ is the transmission rate, absorbing several terms as described below.
@@
 \begin{itemize}
+    \item For any short period of time when $s(t)$ is approximately constant, the system of ODEs can be solved analytically to produce exponential growth\todo{cite exponential growth}.
+    A notable example is the early part of an epidemic when $s(t) \approx 1$.
+    This is the phase of initial exponential growth of an epidemic, which produces the characteristic early epidemic curves.
+    It is in this phase that the basic reproduction number provides a good approximation of the epidemic dynamics.
     \item The basic reproduction number for this model is $R_0 = \beta / \gamma$~\autocite[20]{keelingModeling}.
     Intuitively, this can be thought of as the average number of infections caused by an infectious individual per day, $\beta$, multiplied by the average number of days they are infectious for, $1/\gamma$.
-    \item The effectie reproduction number for this model is $R_e(t) = \beta / \gamma s(t) = R_0 s(t)$~\autocite{pellisEstimation}.
+    \item The effective reproduction number for this model is $R_e(t) = \beta / \gamma s(t) = R_0 s(t)$~\autocite{pellisEstimation}.
+    This is the basic reproduction number reduced because some infections are avoided due to individuals being immune.
     \item The population is closed.
@@
 A standard extension to the SIR model is to allow the infectious period to follow a gamma distribution~\autocite[94]{keelingModeling}.
-This is done by adding multiple infected compartments (see \cref{SEIR:fig:SIR-gamma}).
+This is done by adding multiple infected compartments, shown in \cref{SEIR:fig:SIR-gamma}.
 The system can now be written as follows:
@@
 This parameterisation is chosen to give $1/\gamma$ as the mean infectious period (the gamma distribution's definition and its expectation are in \cref{E-distributions}).
+A gamma distribution with the first parameter equal to 1 is an exponential distribution, in which case this model collapses to the basic SIR model in \cref{SEIR:sec:SIR}.
+As $n$ increases, the between-individual variability in infectious period length reduces, tending to a constant as $n \to \infty$~\autocite{lloydRealistic}.
+
+The new compartments are considered biologically identical here.
+They should be considered as a mathematical trick so that the ODE system produces a gamma distribution rather than an interpretable quantity.
+
 % (it is slightly unclear what it means for a variable to have a distribution in a deterministic model, the clearest explanation is that this is the distribution which the ).
-A gamma distribution with the first parameter equal to 1 is an exponential distribution, in which case this model collapses to the basic SIR model in \cref{SEIR:sec:SIR}.
-The distribution for the time spent in the infected compartments tends to a constant (\ie the same for all individuals) as $n \to \infty$~\autocite{lloydRealistic}.
+
 The reproduction number in terms of the parameters is unchanged by this extension, \ie $R_0 = \beta / \gamma$.
@@
 
-The SIR model can be extended to include a latent period by adding a latent compartment~\autocite[41]{keelingModeling}.
-This is also known as an \emph{exposed} compartment.
-Therefore, we now form the susceptible-exposed-infected-recovered (SEIR) model, shown in \cref{SEIR:fig:SEIR}.
+The SIR model can be extended to include a latent period by adding a latent compartment~\autocite[41]{keelingModeling}, also known as an \emph{exposed} compartment.
+This forms the susceptible-exposed-infected-recovered (SEIR) model, shown in \cref{SEIR:fig:SEIR}.
 
-Similar to the exponential infectious period, the latent period can be modelled as a gamma distribution by adding multiple latent compartments.
+Similar to the infectious period, the latent period can be modelled as a gamma distribution by adding multiple latent compartments.
 \begin{align}
@@
 \end{align}
-where $m$ is the number of latent compartments and $1/\sigma$ is the mean latent period.
-The latent period is distributed $\GamDist(m, m\sigma)$.
-The infectious period is still distributed $\GamDist(n, n\gamma)$.
+where $m$ is the number of latent compartments and $1/\sigma$ is the mean duration of the latent period.
+The latent period is distributed $\GamDist(m, m\sigma)$ and the infectious period is still distributed $\GamDist(n, n\gamma)$.
 
@@
 
+\todo[inline]{add structured pop refs}
 To relax the assumption of a homogeneous population, the population can be stratified.
 If the strata have differing transmission patterns, this can change the dynamics of the epidemic.
-The different patterns could be due to behavioural factors (\eg high-numbers of sexual partners in a sexually-transmitted disease~\autocite[69]{keelingModeling}, younger age groups making many more contacts for respiratory diseases~\autocite[176]{andersonInfectious}), or biological factors (\eg genetics reducing the susceptibility of some individuals~\autocite[208]{andersonInfectious}).
-I will assume that the disease progression (\eg length of the exposed and infectious periods) is the same for all strata, although this assumption can be relaxed.
+The different patterns could be due to behavioural factors (\eg high-numbers of sexual partners in a sexually-transmitted disease~\autocite[69]{keelingModeling} or younger age groups making many more contacts for respiratory diseases~\autocite[176]{andersonInfectious}), or biological factors (\eg genetics reducing the susceptibility of some individuals~\autocite[208]{andersonInfectious}).
+I will assume that the disease progression (\ie length of the exposed and infectious periods) is the same for all strata, although this assumption can be relaxed.
 Stratification requires two major changes model changes: extension of the state vector and describing the between-strata interactions.
@@
 For example, $\vec{s}(t) = (s_1(t), \dots, s_A(t))^T$ where $s_a(t)$ is the proportion of the $a$th strata that are susceptible at time $t$.
-Hence, $N_a s_a(t)$ is the number of susceptible people in stratum $a$.
-These vectors simply replace the scalars in the previous equations.
+Hence, $N_a s_a(t)$ is the number of susceptible people in stratum $a$, where $N_a$ is the number of individuals in stratum $a$.
+These vectors replace the scalars in the previous equations.
 Various other definitions are possible and equivalent, for example defining $s_a$ as the proportion of the total population that are both in strata $a$ and susceptible or parameterised using numbers of individuals~\autocite[57]{keelingModeling}.
@@
 The force of infection is now a vector, $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i_j}(t)$.
-Its expression can be arrived at by considering a single infected individual in strata $a'$.
+Its interpretation can be arrived at by considering a single infected individual in strata $a'$, analogous to the previous explanation in the single stratum case.
 Assume any such individual comes into contact with individuals in strata $a$ at a rate of $\kappa_{aa'}$ in units of contacts per day.
@@
 In practical situations, estimating all of the $\beta_{aa'}$ parameters is infeasible, so some assumptions must be made to reduce the number of parameters~\autocite[176]{andersonInfectious}.
-The infeasibility arises because the progression of the epidemic depends on $\vec{\lambda}(t)$, which has dimension $A$, while $\matr{\beta}$ has dimension $A^2$.
-Common assumption include assuming symmetry, \ie $\beta_{aa'} = \beta_{a'a}$, or that the direction of transmission is irrelevant, \ie $\beta_{aa'} = \beta_{a} \beta_{a'}$.
+It is infeasible because the progression of the epidemic depends on $\vec{\lambda}(t)$, which has dimension $A$, while $\matr{\beta}$ has dimension $A^2$.
+Common assumptions include assuming symmetry, \ie $\beta_{aa'} = \beta_{a'a}$, or that the direction of transmission is irrelevant, \ie $\beta_{aa'} = \beta_{a} \beta_{a'}$.
 Alternatively, as in the scalar case, $\beta_{aa'}$ can be decomposed into the contact rate between strata $a$ and $a'$, $\kappa_{aa'}$ and the probability of transmission upon contact, $q_{aa'}$, and then assumptions made about the components.
@@
 Reproduction numbers in models with structured populations need to be careful to consider the distribution of infections across the strata.
-The WAIFM allows calculation of the number of secondary infections generated by a single infection in the $a$th strata.
-This follows directly from its definition as $\frac{1}{\gamma} \sum_{a'} \beta_{aa'}$. 
-However, how to average these contributions to form the reproduction is not obvious.
-The standard solution to this was proposed by \textcite{diekmannDefinition} (although \textcite[chapter 7]{diekmannMathematical} provides a more gentle, but still rigorous, explanation).
+The WAIFM allows calculation of the number of secondary infections generated by a single infection in the $a$th strata; this follows directly from its definition as $\frac{1}{\gamma} \sum_{a' \in \set{A}} \beta_{aa'}$.
+However, how to average these contributions to calculate the basic reproduction number is not obvious.
+The standard solution to this was proposed by \textcite{diekmannDefinition} (\textcite[chapter 7]{diekmannMathematical} provides a more gentle, but still rigorous, explanation).
 First, they show that the distribution of infections across the strata during the initial period of exponential growth (when $\vec{s} \approx \vec{0}$) relies only on $\matr{\beta}$.
-The distribution is proportional to the \emph{dominant eigenvector} (the eigenvector corresponding to the largest eigenvalue) of $\matr{\beta}$.
+This distribution is proportional to the \emph{dominant eigenvector} (the eigenvector corresponding to the largest eigenvalue) of $\matr{\beta}$.
 Denote by $\rho(\matr{\beta})$ the \emph{dominant eigenvalue} (the largest eigenvalue).
@@
 
-Based primarily on the study of influenza, age-based contact patterns have been found to be important for respiratory diseases~\autocites{birrellRealtimea}{jacksonEffects}.
-\todo{maybe better ref for influence of age contacts on respiratory diseases}
+Age-based contact patterns are important for the transmission dynamics of respiratory diseases~\autocites{birrellBayesian}{jacksonEffects}.
 Respiratory diseases can be spread by conversational contacts, which show very clear age patterns (see \cref{SEIR:fig:age-contacts}).
@@
 \section{Observation model} \label{SEIR:sec:observation}
-\todo[inline]{Be clear this section is to fit to CIS data}
 
 A variety of different approaches can be taken to link a transmission model to observations.
+The method to do so is known as the observation model.
+The formulation of this model is heavily dependent on the data available.
+This means that, unlike for transmission models, formulating a single mathematical framework is hard or impossible.
+Therefore, this section  explains the model I propose for fitting to the CIS prevalence data.
+
 % Here, I extend the system of ODEs to include compartments representing PCR positive individuals.
@@
 One approach would be to convolve the incidence based on the transmission model with the PCR duration distribution.
-\todo{tidy up this notation}
-Essentially, using the notation of \cref{E-inc-prev}, calculate the expected proportion of positive tests on day $t$ as $P_t = \sum_{j=1}^{\dmax} Z_{t-j} S(j)$.
-$Z_{t-j}$ can be related to the ODEs because $Z_t / \Npop = s(t) - s(t-1)$, as $s(t)$.
-However, this approach is computationally expensive.
+Here incidence is calculated from the ODE solution and then related to the data as explained in \cref{E-inc-prev}.
+However, this approach is computationally expensive because of the need to calculate the convolution in \cref{E-inc-prev:eq:EPt }.
 
 A much more efficient option is to approximate this convolution by adding compartments to the ODEs.
-This approach also allows the infected, pre-positive compartment to be represented.
-I will denote this new compartment as $p_0$.
+This approach also allows the infected, pre-positive compartment, which I denote $p_0$, to be represented.
 $p_0$ increases at the same rate as individuals flow into $e_1$.
@@
 
+Deciding $l$ and the length of time each individual spends in each compartment is the most challenging part of formulating the observation model.
 The class of distributions which can be represented as a series of compartments are known as \emph{phase-type} distributions.
@@
 % A $l$-phase EC distribution is the sum of a $(l-2)$-phase gamma distribution and a two-phase acyclic, arbitrary phase-type distribution.
-An EC distribution $P$ can be defined as the sum of three random variables $K$, $X_1$, and $X_2$.
+An EC distribution $P$ can be defined as the sum of three random variables: $K$, $X_1$, and $X_2$.
 It is shown schematically in \cref{SEIR:fig:EC}.
@@
     \item The parameters of $P$ are available in closed form, and the expression for these is derived.
-    \item The number of phases used by the $P$ calculated by these expressions is at most one more than the minimum for any such $P'$.
+    \item The number of phases used by the $P$ calculated by these expressions is at most one more than the minimum across all phase-type distributions that well-match $G$.
 \end{itemize}
@@
 
-\todo[inline]{maybe put something about the likelihood somewhere in this section}
+The ultimate aim of this chapter is to infer the transmission dynamics from the CIS prevalence data.
+This section considers the practicalities of achieving this aim via mechanistic models.
+First, I deal with the inference procedure including the MCMC algorithm (\cref{SEIR:sec:MCMC}), parameterising the model (\cref{SEIR:sec:parameterisation}), and the priors for these parameters (\cref{SEIR:sec:priors}).
+This leads to a fully characterised model, explained in \cref{SEIR:sec:full-model}.
+Finally, \cref{SEIR:sec:solving-ODEs} deals with solving the ODE equations numerically.
 
@@
 
-For each region, I ran 4 chains for 1,000,000 iterations.
+I ran 4 chains for 1,000,000 iterations.
 The first 500,000 were discarded as burn-in, with the remaining chains thinned to every 100th iteration.
@@
 
-\subsection{Parameterisation}
+\subsection{Parameterisation} \label{SEIR:sec:parameterisation}
 
@@
 
-During the pandemic, behaviour clearly changed over time.
-This was in response to government policy, an individual's perceived risk, and a variety of other factors.
-I follow \textcite{birrellRealtime}, using a combination of Google mobility data and a time-varying transmission intensity to include these changes.
-I follow \textcite{birrellRealtimea}'s parameterisation of $\matr{\beta}$, which includes introducing a time-dependency.
+A parsimonious model is needed to parameterise the WAIFW matrix elements with (see \cref{SEIR:sec:structured-populations}).
+The strata will represent age groups.
 
-Between-age-group contacts rates have previously been inferred from a combination of contact surveys, time use data, and mobility data~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}.
-I use the outputs from this analysis as $\matr{K}$ in equation\todo{ref where this contact matrix fits in}.
-To summarise, baseline rates and contexts of contacts between age groups are taken from the contact survey POLYMOD~\autocite{mossongSocial}.
-How these contacts relate to time use is taken from the UK Time Use Survey\todo{cite time use survey}.
-How the population spent their time each week during the pandemic is taken from the Google mobility data\todo{cite mobility data}, which measured the locations of Android users over the period of interest.
-\todo[inline]{add a bit more detail to the above}
+I use the following age-based strata, which align closely with the strata previously used for analysis of the CIS data\todo{cite analyses using these strata}.
+\begin{enumerate}
+    \item $[0, 11)$: children of pre-school and primary school age.
+    \item $[11, 16)$: children of compulsory secondary-school age.
+    \item $[16, 25)$: adolescents and young adults who may remain in formal education.
+    \item $[25, 50)$: younger working age adults.
+    \item $[50, 70)$: older working age adults.
+    \item $[70, \infty)$: elderly.
+\end{enumerate}
 
-I use previously developed estimates of how contact matrices change over time.
-These matrices were produced for \textcite{birrellRealtime}.
-These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
-These use data from the POLYMOD study~\autocite{mossongSocial} and UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
+I will use the parameterisation of \textcite{birrellRealtime}, with the decomposition $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'}$ to separate out contact patterns, $\kappa_{aa'}$, which can vary over time; transmission probabilities $q_{aa'}$, assumed constant; and a random effect for time $\beta(t)$.
+
+I take values for the contact rates, $\kappa_{aa'}(t)$, from other data sources.
+I use previously produced estimates of these rates based on time use data, and mobility data~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}; these estimates have been used for previous modelling of the COVID-19 pandemic~\autocites{vanleeuwenTime}{birrellRealtime}.
+% These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
+Baseline (pre-pandemic) estimates of contacts in various settings are taken from the POLYMOD study~\autocite{mossongSocial} and UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
 POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on age and location (home, work, school, leisure, transport, or other) of these contacts.
@@
 Google mobility data measures the location of Android users over time, relative to a pre-pandemic baseline.
-These data sources are then used to down or upweight the number of contacts in each setting, as estimated by \textcite{vanleeuwenAugmenting}.
+These data sources are then used to down or upweight the number of contacts in each setting.
+These estimates of contacts by setting and then recombined across settings to from $\kappa_{aa'}$
 For mathematical details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.
@@
 However, these are not captured by the contact survey.
-To allow for these changes, $\beta$ is allowed to vary over time.
-In particular, $\beta$ follows a random walk in log space on the weekly timescale:
+To allow for these changes and other model misspecficiation, I follow \textcite{birrellRealtime} in placing a random walk prior on $\beta(t)$ in log space on the weekly timescale:
 \begin{align}
@@
 where $w(t)$ is the week number of time $t$, with week 1 being the first week modelled.
-Two parameters are unspecified here: the initial random of the random walk, $\beta_1$, and the standard deviation of the random walk, $\sigma_\epsilon$.
-See \cref{perf-test:sec:parameters-priors} for the implementation and parameterisation of these.
+The remaining parameters $\beta_1$, the initial random of the random walk, and $\sigma_\epsilon$, the standard deviation of the random walk, will have priors placed on them, with their values inferred in the usual way.
 
@@
 For any stratum $i$, $\beta^\text{susc}_i$ is known as the \emph{relative susceptibility} of stratum $i$.
-Informally, it is how vulnerable to infection individuals in stratum $i$ are.
-Formally, define an $a_0$ and $a'_0$ and enforce $\beta^\text{susc}_{a_0}=1$ for identifiability.
+
+$q_{aa'}$ remains to be specified.
+I decompose it into two independent components such that $q_{aa'} = q^\text{susc}_{a} q^\text{inf}_{a'}$.
+$q^\text{susc}_a$ is the \emph{susceptibility} of stratum $a$, \ie how vulnerable to infection individuals in stratum $a$ are.
+$q^\text{inf}_a$ is the \emph{transmissibility} of stratum $a$, \ie how likely individuals in stratum $a$ are to pass on an infection.
 Then $\beta^\text{susc}_i$ is the relative risk of an individual in stratum $i$ being infected, relative to an individual in stratum $a_0$.
-\todo{clarify notation and explaantion here of reference class}
-it is the probability that a susceptible individual in stratum $i$ will be infected upon contact with an infectious individual, relative to a individual in the reference stratum.
-Similarly, $\beta^\text{inf}_i$ is known as the \emph{relative infectiousness} of stratum $i$.
-Informally, it is how likely individuals in stratum $i$ are to pass on the infection.
-Formally, it is the relative risk of an infectious individual in stratum $i$ infecting a susceptible individual upon contact, relative to an individual in the reference stratum.
-These measures could be biological, \eg due to differences in immune systems, or behavioural, \eg if infants are less hygenic than adults.
+These measures include both biological differences, \eg due to differences in immune systems, or behavioural, \eg if infants are less hygenic than adults.
 
@@
 
-\todo[inline]{move the below into a context-specific section}
-Based on prior work\todo{cite}, I assume that all individuals have the same \emph{infectiousness} but that there is a difference in \emph{susceptibility}.
-Therefore, $\beta^\text{inf}_{j} = 1$ for all $j$.
-The prior literature suggests that children are less susceptible than adults, but otherwise there is little difference in susceptibility.
-Therefore, I set $\beta^\text{susc}_i$ to 1 for adult-aged strata and $\beta^\text{susc}_i = \beta_c$ for child-aged strata.
+Previous studies for pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults\todo{cite where age difference in transmission comes from}.
+Therefore, I set $\beta^\text{inf}_{a} = 1$ for all $a$, $\beta^\text{susc}_{a} = q_c$ for the youngest stratum, and $\beta^\text{susc}_{a} = 1$ for the remaining strata.
+
 \subsubsection{Identifiability}
@@
 
-For any short period of time when $\vec{s}$ is approximately constant, the system of ODEs can be solved analytically to produce exponential growth.
-In fact, all the $\vec{e}$, $\vec{i}$ and $\vec{p}$ compartments grow at the same initial exponential growth rate, $\psi$.
-\todo{move R definition for each model into where they are introduced}
-If $\vec{s} = 1$ then the basic reproduction number can be linked to the growth rate of this epidemic as follows~\autocites{birrellRealtimea}{wearingAppropriate}:
-\begin{align}
+When $\vec{s}$ is approximately constant, and hence the epidemic grows exponentially (see \cref{SEIR:sec:SIR}), the $\vec{e}$, $\vec{i}$ and $\vec{p}$ compartments all grow at the same initial exponential growth rate, $\psi$.
+If this is the initial exponential phase, when $\vec{s} \approx 1$, then the basic reproduction number for a SEIR model can be linked to this growth rate by~\autocites{birrellBayesian}{wearingAppropriate}:\begin{align}
     \R = \frac{\psi \left( \frac{\psi}{2\sigma} + 1 \right)^2}{\gamma \left(1 - \frac{1}{\left(\frac{\psi}{2 \gamma} + 1 \right)^2} \right)} \label{SEIR:eq:rtoR}.
@@
 
-$\R$ can be linked to $\beta_1$ because $\R$ is also equal to the dominant eigenvalue of the next-generation matrix, $\matr{G} = \matr\beta \circ \matr{M}$~\autocites{diekmannDefinition}{birrellRealtimea}.
-Therefore, the following procedure finds $\beta_1$ such that the growth rate $\psi$ is achieved.
-First calculate $\R$ from the model parameters using \cref{SEIR:eq:rtoR}.
-Denote this $\hat{R}$.
-Then calculate $R^* = D(\matr{G})$ when $\beta_1=1$, where $D(\cdot)$ gives the dominant eigenvalue.
+Furthermore, $\R$ is linked to $\beta_1$ via $\matr{\beta}$ (see \cref{SEIR:sec:structured-populations}).
+This implies that the model can be parameterised by using $\psi$ in place of $\beta_1$.
+The growth rate $\psi$ is easily identifiable from the data, and therefore this parameterisation should be more efficient.
+The following procedure gives $\beta_1$ in terms of $\psi$.
+First, calculate $\R$ from $\psi$ using \cref{SEIR:eq:rtoR}; denote this $\hat{R}$.
+Then calculate $R^* = \rho(\matr{\beta}) / \gamma$ when $\beta_1=1$ (recalling that $\rho(\matr{\beta})$ denotes the dominant eigenvalues of $\matr{\beta}$).
 Multiplying a matrix by a scalar multiplies its dominant eigenvalue by that same scalar, therefore, setting $\beta_1 = \hat{R} / R^*$ gives the desired $\R$.
 
-To form a parsimonious parameterisation of the initial state, first focus on the initial proportion of the population that is in infected compartments.
-In steady-state exponential growth, the relative proportion of each strata that is infected is proportional to the dominant eigenvector of $\matr{G}$, which I denote $\vec{d}$.
+To form a parsimonious parameterisation of the initial state, first focus on the initial proportion of the population that is in the infected compartments.
+Recall that in the exponential growth phase, the relative proportion of each strata that is infected is proportional to the dominant eigenvector of $\matr{\beta}$, which I denote $\vec{d}$.
 Let the proportion of the population in the $i$ states at time 0 in the strata with the highest proportion infected, $i^+$, be a model parameter.
 Then, $\vec{i_0} = \vec{i_1} + \vec{i_2} = \frac{i^+}{\max \vec{d}} \vec{d}$.
-I introduce an additional vector of parameters, $\vec{\pi}$, where $\pi_i$ is the proportion of strata $i$ without immunity at time 0.
+Now introduce an additional vector of parameters, $\vec{\pi}$, where $\pi_i$ is the proportion of strata $i$ without immunity at time 0.
 That is, at time 0, $\vec{r} = 1 - \vec{\pi}$.
 
-The system of ODEs and the result that all the comparments are growing at the same rate give the following for the initial state of the compartments:
+From the system of ODEs\todo{might need to reorder this because I've not presented these yet..
+} and the knowledge that all the compartments are growing exponentially at rate $\psi$, one can derive the following for the initial state of the compartments:
 \begin{align}
@@
 
+\subsection{Priors} \label{SEIR:sec:priors}
+Priors for the parameters are given in \cref{SEIR:table:priors}.
+The parameters governing the duration of positivity ($\kappa$s) were explained previously (see \cref{SEIR:table:ec-params}).
+The initial conditions and transition rate parameters are unindentifiable in a SEIR model~\autocite{dankwaStructural}, therefore I used assumed values or strongly informative priors on them.
+All other parameters had weakly informative priors.
+
+\begin{landscape}
+\begin{table}
+\begin{tabular}{l c l l}
+    Parameter description & Symbol & Distribution & Comment \\
+    \hline \\
+    Mean latent period & $1/\sigma$ & 3.5 days (fixed) & \Textcite{zhaoEstimating} \\
+    Mean infectious period & $1/\gamma$ & 4 days (fixed) & \Textcite{zhaoEstimating} \\
+    Initial proportion without immunity & $\vec\pi$ & See \cref{SEIR:table:immunity-prior} & \\
+    Initial proportion infected & $i^+$ & $\BetaDist(0.5, 1000)$ & Weakly informative \\
+    Relative susceptibility of children & $\beta_c$ & $\LN(-0.4325, 0.1174)$ &  \\
+    Initial growth rate & $\psi$ & $\N(0.06, 0.04)$ & Weakly informative \\
+    Standard deviation of random walk steps on log scale & $\sigma_\epsilon$ & $\Exponential(80)$ & Weakly informative \\ %penalised model complexity prior~\autocite{simpsonPenalising} \\
+    Overdispersion of observations & $\rho$ & $\Exponential(2 \times 10^5)$ & Weakly informative
+\end{tabular}
+\todo[inline]{Dani/Paul: I think I took the prior for $\beta_c$ from RTM, but can't find the reference now - any ideas?}
+\caption[SEIR model priors]{Priors for each parameter. For details of the distributions and their parameterisations see \cref{E-distributions}.}
+\label{SEIR:table:priors}
+\end{table}
+
+\begin{table}
+\centering
+\begin{tabular}{l|lllll}
+         & $[0,16)$ & $[16,25)$ & $[25,50)$ & $[50,70)$ & $[70,\infty)$ \\
+        \hline
+        East Midlands & $\BetaDist(8.5, 301)$ & $\BetaDist(18, 256)$ & $\BetaDist(19, 475)$ & $\BetaDist(20, 493)$ & $\BetaDist(5, 337)$ \\
+        East of England & $\BetaDist(7.9, 236)$ & $\BetaDist(21, 248)$ & $\BetaDist(35, 677)$ & $\BetaDist(26, 613)$ & $\BetaDist(5.6, 332)$ \\
+        London & $\BetaDist(9.7, 131)$ & $\BetaDist(50, 328)$ & $\BetaDist(154, 1331)$ & $\BetaDist(106, 990)$ & $\BetaDist(7.6, 204)$ \\
+        North East & $\BetaDist(6.1, 180)$ & $\BetaDist(12, 145)$ & $\BetaDist(14, 274)$ & $\BetaDist(14, 295)$ & $\BetaDist(4.2, 240)$ \\
+        North West & $\BetaDist(12, 252)$ & $\BetaDist(38, 353)$ & $\BetaDist(72, 885)$ & $\BetaDist(44, 695)$ & $\BetaDist(6.3, 264)$ \\
+        South East & $\BetaDist(8.9, 404)$ & $\BetaDist(17, 319)$ & $\BetaDist(20, 577)$ & $\BetaDist(19, 586)$ & $\BetaDist(4.4, 356)$ \\
+        South West & $\BetaDist(6.9, 415)$ & $\BetaDist(11, 277)$ & $\BetaDist(12, 556)$ & $\BetaDist(11, 559)$ & $\BetaDist(4, 492)$ \\
+        West Midlands & $\BetaDist(8, 167)$ & $\BetaDist(19, 151)$ & $\BetaDist(32, 475)$ & $\BetaDist(33, 504)$ & $\BetaDist(5.8, 241)$ \\
+        Yorkshire and The Humber & $\BetaDist(9.2, 314)$ & $\BetaDist(19, 242)$ & $\BetaDist(27, 589)$ & $\BetaDist(18, 474)$ & $\BetaDist(4.9, 316)$ \\
+    \end{tabular}
+\caption{Priors for $\vec{\pi}$, the proportion of the population without prior immunity, in each age group and region combination.
+Based on seroprevalence in blood donors~\autocite{amirthalingamSeroprevalence} and a representative sample of children~\autocite{ratcliffeCommunity} in August 2020.}
+\label{SEIR:table:immunity-prior}
+\end{table}
+\end{landscape}
+
 \subsection{Full model} \label{SEIR:sec:full-model}
+
+\todo[inline]{check likelihood is mentioned somewhere in this section}
 
@@
 \end{align}
-Recall that each region is run independently (see \cref{SEIR:sec:geography}), $\matr{beta}(t)_{ij} = \beta(t) \beta^\text{susc}_i$ with $\beta^\text{susc}_i = \beta_c$ for child-aged strata and $\beta^\text{susc}_i = 1$ for adult-aged strata.
 
-\subsection{Solving the systems of ODEs}
-I solve the system of ODEs using Euler's method, discretised at timesteps of $\delta t = 0.5$days.
-That is to find the state of the system at time $t+1/2$ given the state at time $t$, I add half the derivative to the state at time $t$.
-\todo[inline]{add these equations}
-
-The prevalence on day $t$, used for the likelihood, is the mean of the solution on the two corresponding days.
-
-\subsection{Priors}
-Priors for the parameters are given in \cref{SEIR:table:priors}.
-The parameters governing the duration of positivity ($\kappa$s) were explained previously (see \cref{SEIR:table:ec-params}).
-The initial conditions and transition rate parameters are unindentifiable in a SEIR model~\autocite{dankwaStructural}, therefore I used assumed values or strongly informative priors on them.
-All other parameters had weakly informative priors.
-
-\begin{landscape}
-\begin{table}
-\begin{tabular}{l c l l}
-    Parameter description & Symbol & Distribution & Comment \\
-    \hline \\
-    Mean latent period & $1/\sigma$ & 3.5 days (fixed) & \Textcite{zhaoEstimating} \\
-    Mean infectious period & $1/\gamma$ & 4 days (fixed) & \Textcite{zhaoEstimating} \\
-    Initial proportion without immunity & $\vec\pi$ & See \cref{SEIR:table:immunity-prior} & \\
-    Initial proportion infected & $i^+$ & $\BetaDist(0.5, 1000)$ & Weakly informative \\
-    Relative susceptibility of children & $\beta_c$ & $\LN(-0.4325, 0.1174)$ &  \\
-    Initial growth rate & $\psi$ & $\N(0.06, 0.04)$ & Weakly informative \\
-    Standard deviation of random walk steps on log scale & $\sigma_\epsilon$ & $\Exponential(80)$ & Weakly informative \\ %penalised model complexity prior~\autocite{simpsonPenalising} \\
-    Overdispersion of observations & $\rho$ & $\Exponential(2 \times 10^5)$ & Weakly informative
-\end{tabular}
-\todo[inline]{Dani/Paul: I think I took the prior for $\beta_c$ from RTM, but can't find the reference now - any ideas?}
-\caption[SEIR model priors]{Priors for each parameter. For details of the distributions and their parameterisations see \cref{E-distributions}.}
-\label{SEIR:table:priors}
-\end{table}
-
-\begin{table}
-\centering
-\begin{tabular}{l|lllll}
-         & $[0,16)$ & $[16,25)$ & $[25,50)$ & $[50,70)$ & $[70,\infty)$ \\
-        \hline
-        East Midlands & $\BetaDist(8.5, 301)$ & $\BetaDist(18, 256)$ & $\BetaDist(19, 475)$ & $\BetaDist(20, 493)$ & $\BetaDist(5, 337)$ \\
-        East of England & $\BetaDist(7.9, 236)$ & $\BetaDist(21, 248)$ & $\BetaDist(35, 677)$ & $\BetaDist(26, 613)$ & $\BetaDist(5.6, 332)$ \\
-        London & $\BetaDist(9.7, 131)$ & $\BetaDist(50, 328)$ & $\BetaDist(154, 1331)$ & $\BetaDist(106, 990)$ & $\BetaDist(7.6, 204)$ \\
-        North East & $\BetaDist(6.1, 180)$ & $\BetaDist(12, 145)$ & $\BetaDist(14, 274)$ & $\BetaDist(14, 295)$ & $\BetaDist(4.2, 240)$ \\
-        North West & $\BetaDist(12, 252)$ & $\BetaDist(38, 353)$ & $\BetaDist(72, 885)$ & $\BetaDist(44, 695)$ & $\BetaDist(6.3, 264)$ \\
-        South East & $\BetaDist(8.9, 404)$ & $\BetaDist(17, 319)$ & $\BetaDist(20, 577)$ & $\BetaDist(19, 586)$ & $\BetaDist(4.4, 356)$ \\
-        South West & $\BetaDist(6.9, 415)$ & $\BetaDist(11, 277)$ & $\BetaDist(12, 556)$ & $\BetaDist(11, 559)$ & $\BetaDist(4, 492)$ \\
-        West Midlands & $\BetaDist(8, 167)$ & $\BetaDist(19, 151)$ & $\BetaDist(32, 475)$ & $\BetaDist(33, 504)$ & $\BetaDist(5.8, 241)$ \\
-        Yorkshire and The Humber & $\BetaDist(9.2, 314)$ & $\BetaDist(19, 242)$ & $\BetaDist(27, 589)$ & $\BetaDist(18, 474)$ & $\BetaDist(4.9, 316)$ \\
-    \end{tabular}
-\caption{Priors for $\vec{\pi}$, the proportion of the population without prior immunity, in each age group and region combination.
-Based on seroprevalence in blood donors~\autocite{amirthalingamSeroprevalence} and a representative sample of children~\autocite{ratcliffeCommunity} in August 2020.}
-\label{SEIR:table:immunity-prior}
-\end{table}
-\end{landscape}
-
-\subsection{Data}
+\todo[inline]{following paragraph needs rewording having reordered this chapter}
+Recall that each region is run independently (see \cref{SEIR:sec:geography}), $\matr{\beta}(t)_{ij} = \beta(t) \beta^\text{susc}_i$ with $\beta^\text{susc}_i = \beta_c$ for child-aged strata and $\beta^\text{susc}_i = 1$ for adult-aged strata.
 
@@
 
-By having indpendent models for each region, I assume that the epidemics in each region do not interact.
+By having independent models for each region, I assume that the epidemics in each region do not interact.
 Clearly, this is not true as travel between regions of England is common and unrestricted.
@@
 The test results are aggregated to age group and region on each day.
+
+\subsection{Solving the systems of ODEs} \label{SEIR:sec:solving-ODEs}
+I solve the system of ODEs using Euler's method, discretised at timesteps of $\delta t = 0.5$days.
+That is to find the state of the system at time $t+1/2$ given the state at time $t$, I add half the derivative to the state at time $t$.
+\todo[inline]{add these equations}
+
+The prevalence on day $t$, used for the likelihood, is the mean of the solution on the two corresponding days.
 

### 2023-11-21 incidence-prevalence.tex
Add ref to BUGS book
editing

@@
 Here, I adapt previous work to the SARS-CoV-2 context.
-In particular, recoveries need considering, which was not the case for HIV/AIDS.
+In particular, recovering from infection needs to be considering, which was not the case for HIV/AIDS.
 The observation process (\cref{inc-prev:sec:observation-process}) describes the relationship between prevalence and the observed data, ultimately allowing inference of incidence based on this data.
@@
 
-\emph{Prevalence} of a disease on day $t$, is the number of people with the disease on that day~\autocite[90]{lashModern}.
+\emph{Prevalence} of a disease on day $t$, is the proportion of people with the disease on that day~\autocite[90]{lashModern}.
 What it means to have a disease can vary by context (see\todo{ref section discussing what it means to have covid}).
@@
 
+\todo[inline]{Add some refs about duration and survival}
 Duration, or time-to-event, distributions can be expressed in several equivalent ways.
@@
 $D_i = E_i - B_i + 1$ is their \emph{duration of positivity}, the number of days for which they are prevalent.
-I model each $D_i$ as a discrete independent and identically distributed (iid) random variable; extending the methodology to account for covariates is left to further work.
+I model the $D_i$s as a discrete independent and identically distributed (iid) random variable; extending the methodology to account for covariates is left to further work.
 The duration is a positive integer, \ie $D_i \in \{1, 2, \dots, \dmax \}$, where $\dmax$ is the maximum duration.
@@
 The \emph{hazard} on day $d$ is the probability of recovering at time $d$ conditional on having survived to time $d$: $\lambda(d) = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
-Throughout this thesis, the survival function will be the primary quantity of interest, and it will normally be parameterised in terms of the hazard; these two quantities are related by $S(d) = \prod_{i=1}^{d-1} (1 - \lambda(i))$.
+Throughout this thesis, the survival function will be the primary quantity of interest, and it will normally be expressed in terms of the hazard; these two quantities are related by $S(d) = \prod_{i=1}^{d-1} (1 - \lambda(i))$.
 
@@
 A commonly-used point process for infectious disease epidemiology is the time-inhomogenous Poisson process~\autocites{brookmeyerMethod}{paganoHIV}{rosenbergBackcalculation}{brookmeyerBackcalculation}.
-A Poisson process is a point process with independent increments, and the property that $K(s) - K(t) \dist \Poi\left(\int^s_t \nu(r) dr \right)$ where $\nu(r) \geq 0$ is the \emph{intensity} of the epidemic at time $r$~\autocites[244]{yanDistribution}.
+A Poisson process is a point process with independent increments, and the property that $K(s) - K(t) \dist \Poi\left(\int^s_t \nu(r) dr \right)$ where $\nu(r) \geq 0$ is the \emph{intensity} of the epidemic process at time $r$~\autocites[244]{yanDistribution}.
 A time-homogenous Poisson process is the special case where $\nu(r)$ is constant.
@@
 
-Two practical issues arise when applying MCMC: \emph{convergence} and \emph{stability}\todo{cite BUGS course}.
+Two practical issues arise when applying MCMC: \emph{convergence} and \emph{stability}~\autocite[72]{lunnBUGS}.
 

### 2023-11-21 incidence-prevalence.tex
correct some notation/terminology

--- 2023-11-13 incidence-prevalence.tex
@@
 The duration is a positive integer, \ie $D_i \in \{1, 2, \dots, \dmax \}$, where $\dmax$ is the maximum duration.
-The first way of expressing the distribution of $D_i$ is as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
+The first way of expressing the distribution of $D_i$ is as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{k=1}^d f_{D}(k)$.
 The \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$ (a common alternative convention for survival analysis with discrete variables is to define the survival function as $\prob(D >d)$).
-The \emph{hazard} on day $d$ is the probability of recovering at time $d$ conditional on having survived to time $d$: $\lambda(d) = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
+The \emph{hazard} on day $d$ is the probability of recovering at time $d$ conditional on still being positive on day $d$: $\lambda(d) = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
 Throughout this thesis, the survival function will be the primary quantity of interest, and it will normally be parameterised in terms of the hazard; these two quantities are related by $S(d) = \prod_{i=1}^{d-1} (1 - \lambda(i))$.
 
-Since $D_i$ is a discrete random variable, $f_D(d) \geq 0$ for all $d$ and $\sum_{i=1}^{\dmax} f_D(i) = 1$.
+Since $D_i$ is a discrete random variable, $f_D(d) \geq 0$ for all $d$ and $\sum_{k=1}^{\dmax} f_D(k) = 1$.
 This implies that $F_D(d)$ is monotonically increasing.
@@
 \section{Infection process} \label{inc-prev:sec:infection-process}
-The infection process is the mathematical description of incidence.
+The infection process is the mathematical description of how infections occur.
 It is a \emph{point} or \emph{counting} process, where each infection is an event.
@@
 These additional infections will propagate the epidemic further, increasing the intensity at future times.
-Point processes can be extended to \emph{branching processes} to account for this feedback~\autocite[246]{yanDistribution}, although these are beyond the scope of this thesis.
+Point processes can be generalised to \emph{branching processes} to account for this feedback~\autocite[246]{yanDistribution}, although these are beyond the scope of this thesis.
 

### 2023-11-21 SEIR.tex

@@
 \Cref{SEIR:sec:inference-implementation} describes how I will apply the model to the CIS prevalence data, including the inference process, parameterisation, and priors for this model.
+This transmission model is the same as that in \textcite{birrellRealtime}, but the observation model is novel.
 \Cref{SEIR:sec:sim-study} checks this recovers the parameters successfully using a simulation study.
@@
 Common assumptions include assuming symmetry, \ie $\beta_{aa'} = \beta_{a'a}$, or that the direction of transmission is irrelevant, \ie $\beta_{aa'} = \beta_{a} \beta_{a'}$.
-Alternatively, as in the scalar case, $\beta_{aa'}$ can be decomposed into the contact rate between strata $a$ and $a'$, $\kappa_{aa'}$ and the probability of transmission upon contact, $q_{aa'}$, and then assumptions made about the components.
+% Alternatively, as in the scalar case, it may be easier to iden $\beta_{aa'}$ can be decomposed into the contact rate between strata $a$ and $a'$, $\kappa_{aa'}$ and the probability of transmission upon contact, $q_{aa'}$, and then assumptions made about the components.
 The choice of assumptions depends on the context, including the disease being studied and the data available.
@@
 \begin{align}
-    \prod_{i,t} \BB (y_{it} \mid n_{it}, \sum_{k=1}^l p_{ki}(t), \rho)
+    \prod_{a,t} \BB (y_{at} \mid n_{at}, \sum_{k=1}^l \bar{p}_{kat}, \rho)
+    \label{SEIR:eq:likelihood}
 \end{align}
-using the mean-dispersion parametrisation of the beta-binomial distribution (see \cref{E-distributions}).
-$\rho$ is a nuisance parameter controlling the overdispersion of the beta-binomial distribution.
-At $\rho=0$, the beta-binomial coincides with the binomial distribution.
+where $\bar{p}_{kat} = \int_{t}^{t+1} p_{ka}(s) ds$ is the mean occupancy in compartment $p_k$ and strata $a$ over day $t$ and $\rho$ is a nuisance parameter controlling the overdispersion of the beta-binomial distribution.
+In this parameterisation (see \cref{E-distributions}) at $\rho=0$ the beta-binomial coincides with the binomial distribution and as $\rho \to \infty$, the distribution tends towards the uniform distribution on $[0, 1]$.
 
@@
 
-I will use the parameterisation of \textcite{birrellRealtime}, with the decomposition $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'}$ to separate out contact patterns, $\kappa_{aa'}$, which can vary over time; transmission probabilities $q_{aa'}$, assumed constant; and a random effect for time $\beta(t)$.
+I follow the WAIFW matrix parameterisation of \textcite{birrellRealtime}, with the decomposition $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'} N_{a'}$ to separate out contact patterns, $\kappa_{aa'}$, which can vary over time; transmission probabilities $q_{aa'}$, assumed constant; a global effect for time $\beta(t)$; and the population of the stratum, $N_{a'}$.
 
-I take values for the contact rates, $\kappa_{aa'}(t)$, from other data sources.
-I use previously produced estimates of these rates based on time use data, and mobility data~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}; these estimates have been used for previous modelling of the COVID-19 pandemic~\autocites{vanleeuwenTime}{birrellRealtime}.
+This parameterisation takes values for the contact rates, $\kappa_{aa'}(t)$, from other data sources.
+These values are produced based on several data sources~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}.
+%; these estimates have been used for previous modelling of the COVID-19 pandemic~\autocites{vanleeuwenTime}{birrellRealtime}.
 % These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
-Baseline (pre-pandemic) estimates of contacts in various settings are taken from the POLYMOD study~\autocite{mossongSocial} and UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
-POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on age and location (home, work, school, leisure, transport, or other) of these contacts.
+First, take a baseline from the POLYMOD study's~\autocite{mossongSocial} baseline (pre-pandemic) estimates of contacts in various settings and the UK Time Use Survey (UKTUS)'s~\autocite{UKTUS} estimates of time use.
+POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on the contact's age and location (home, work, school, leisure, transport, or other).
 UKTUS asked a representative sample of the UK population about how much time they spent doing various activities each day.
@@
 \Textcite{birrellRealtime} then modify these contact rates based on measured changes in time use during the pandemic.
-These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
+The changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
 Google mobility data measures the location of Android users over time, relative to a pre-pandemic baseline.
 These data sources are then used to down or upweight the number of contacts in each setting.
-These estimates of contacts by setting and then recombined across settings to from $\kappa_{aa'}$
-For mathematical details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.
+These estimates of contacts by setting are then recombined across settings to from $\kappa_{aa'}$
+For further details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.
 
@@
 However, these are not captured by the contact survey.
-To allow for these changes and other model misspecficiation, I follow \textcite{birrellRealtime} in placing a random walk prior on $\beta(t)$ in log space on the weekly timescale:
+To allow for these changes and uncertainty produced by the procedure above, I follow \textcite{birrellRealtime} in placing a random walk prior on $\beta(t)$ in log space on the weekly timescale:
 \begin{align}
@@
 where $w(t)$ is the week number of time $t$, with week 1 being the first week modelled.
-The remaining parameters $\beta_1$, the initial random of the random walk, and $\sigma_\epsilon$, the standard deviation of the random walk, will have priors placed on them, with their values inferred in the usual way.
+The remaining parameters $\beta_1$, the initial random of the random walk, and $\sigma_\epsilon$, the standard deviation of the random walk, will have priors placed on them (see \cref{SEIR:sec:priors}), with their values inferred in the usual way.
 
-$K_{ij}$ and $\beta_{ij}$ appear only as a product and therefore are not individually identifiable.
-Normally, $K_{ij}$ is assumed to be known, for example from a previous contact survey~\autocite[such as][]{mossongSocial}.
-Ideally, all the $\beta_{ij}$ terms would be estimated, but there is normally inadequate data to do so, requiring assumptions to be made to reduce the number of parameters~\autocite[65]{keelingModeling}.
-% $\beta_{ij}$ is then estimated, although often with a parsimonious model\todo{ref / give examples / delete on how age-specfic susceptibility / transmissibility happens}.
-A common assumption is that $\beta_{ij}$ can be decomposed into two independent components such that $\beta_{ij} = \beta \beta^\text{susc}_{i} \beta^\text{inf}_j$.
-For any stratum $i$, $\beta^\text{susc}_i$ is known as the \emph{relative susceptibility} of stratum $i$.
+% $K_{ij}$ and $\beta_{ij}$ appear only as a product and therefore are not individually identifiable.
+% Normally, $K_{ij}$ is assumed to be known, for example from a previous contact survey~\autocite[such as][]{mossongSocial}.
+% Ideally, all the $\beta_{ij}$ terms would be estimated, but there is normally inadequate data to do so, requiring assumptions to be made to reduce the number of parameters~\autocite[65]{keelingModeling}.
+% % $\beta_{ij}$ is then estimated, although often with a parsimonious model\todo{ref / give examples / delete on how age-specfic susceptibility / transmissibility happens}.
+% A common assumption is that $\beta_{ij}$ can be decomposed into two independent components such that $\beta_{ij} = \beta \beta^\text{susc}_{i} \beta^\text{inf}_j$.
+% For any stratum $i$, $\beta^\text{susc}_i$ is known as the \emph{relative susceptibility} of stratum $i$.
 
 $q_{aa'}$ remains to be specified.
-I decompose it into two independent components such that $q_{aa'} = q^\text{susc}_{a} q^\text{inf}_{a'}$.
+I again follow the approach of \textcite{birrellRealtime}, decomposing it into two independent components such that $q_{aa'} = q^\text{susc}_{a} q^\text{inf}_{a'}$.
 $q^\text{susc}_a$ is the \emph{susceptibility} of stratum $a$, \ie how vulnerable to infection individuals in stratum $a$ are.
-$q^\text{inf}_a$ is the \emph{transmissibility} of stratum $a$, \ie how likely individuals in stratum $a$ are to pass on an infection.
-Then $\beta^\text{susc}_i$ is the relative risk of an individual in stratum $i$ being infected, relative to an individual in stratum $a_0$.
+$q^\text{inf}_{a'}$ is the \emph{transmissibility} of stratum $a'$, \ie how likely individuals in stratum $a$ are to pass on an infection.
+% Then $\beta^\text{susc}_i$ is the relative risk of an individual in stratum $i$ being infected, relative to an individual in stratum $a_0$.
 These measures include both biological differences, \eg due to differences in immune systems, or behavioural, \eg if infants are less hygenic than adults.
@@
 Previous studies for pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults\todo{cite where age difference in transmission comes from}.
-Therefore, I set $\beta^\text{inf}_{a} = 1$ for all $a$, $\beta^\text{susc}_{a} = q_c$ for the youngest stratum, and $\beta^\text{susc}_{a} = 1$ for the remaining strata.
+Therefore, I set $q^\text{inf}_{a'} = 1$ for all $a'$, $q^\text{susc}_{a} = q_c$ for the youngest stratum, and $q^\text{susc}_{a} = 1$ for the remaining strata.
 
@@
 In addition, I use a parsimonious parameterisation of the initial conditions using the initial proportion of the population that are infected and the initial proportion that are susceptible only.
-Both of these rely on an assumption that the epidemic is in the steady-state exponential growth phase at time 0.
+Both of these rely on an assumption that the epidemic is in a steady-state exponential growth phase at time 0 (see \cref{SEIR:sec:SIR}).
 
-When $\vec{s}$ is approximately constant, and hence the epidemic grows exponentially (see \cref{SEIR:sec:SIR}), the $\vec{e}$, $\vec{i}$ and $\vec{p}$ compartments all grow at the same initial exponential growth rate, $\psi$.
-If this is the initial exponential phase, when $\vec{s} \approx 1$, then the basic reproduction number for a SEIR model can be linked to this growth rate by~\autocites{birrellBayesian}{wearingAppropriate}:\begin{align}
+During such an exponential growth phase, all components of the $\vec{e}$, $\vec{i}$ and $\vec{p}$ compartments grow at the same growth rate, $\psi$.
+This follows directly from the ODEs where $\vec{s}$ is constant. 
+The growth rate $\psi$ is easily identifiable from the data, and therefore a parameterisation in terms of $\phi$ should be more computationally efficient.
+If this is the initial exponential phase, when $\vec{s} \approx 1$, then the basic reproduction number for a SEIR model is linked to this growth rate by~\autocites{birrellBayesian}{wearingAppropriate}:
+\begin{align}
     \R = \frac{\psi \left( \frac{\psi}{2\sigma} + 1 \right)^2}{\gamma \left(1 - \frac{1}{\left(\frac{\psi}{2 \gamma} + 1 \right)^2} \right)} \label{SEIR:eq:rtoR}.
@@
 
-Furthermore, $\R$ is linked to $\beta_1$ via $\matr{\beta}$ (see \cref{SEIR:sec:structured-populations}).
-This implies that the model can be parameterised by using $\psi$ in place of $\beta_1$.
-The growth rate $\psi$ is easily identifiable from the data, and therefore this parameterisation should be more efficient.
+Furthermore, $\R$ is a function of $\matr{\beta}$, and hence $\beta_1$ (as explained in \cref{SEIR:sec:structured-populations}).
+Combining these expressions, the model can be parameterised by using $\psi$ in place of $\beta_1$.
 The following procedure gives $\beta_1$ in terms of $\psi$.
 First, calculate $\R$ from $\psi$ using \cref{SEIR:eq:rtoR}; denote this $\hat{R}$.
-Then calculate $R^* = \rho(\matr{\beta}) / \gamma$ when $\beta_1=1$ (recalling that $\rho(\matr{\beta})$ denotes the dominant eigenvalues of $\matr{\beta}$).
-Multiplying a matrix by a scalar multiplies its dominant eigenvalue by that same scalar, therefore, setting $\beta_1 = \hat{R} / R^*$ gives the desired $\R$.
+Then calculate $R^* = \rho(\matr{\beta}) / \gamma$ when $\beta_1=1$ (recall that $\rho(\matr{\beta})$ denotes the dominant eigenvalue of $\matr{\beta}$).
+Multiplying a matrix by a scalar multiplies its dominant eigenvalue by that same scalar, therefore, setting $\beta_1 = \hat{R} / R^*$ gives the desired means that $\R$ at time 0 is $\hat{R}$, as desired.
 
@@
 Recall that in the exponential growth phase, the relative proportion of each strata that is infected is proportional to the dominant eigenvector of $\matr{\beta}$, which I denote $\vec{d}$.
-Let the proportion of the population in the $i$ states at time 0 in the strata with the highest proportion infected, $i^+$, be a model parameter.
+Introduce a model parameter $i^+$, the proportion of the population in the $i$ states at time 0 in the strata with the highest proportion infected.
 Then, $\vec{i_0} = \vec{i_1} + \vec{i_2} = \frac{i^+}{\max \vec{d}} \vec{d}$.
-Now introduce an additional vector of parameters, $\vec{\pi}$, where $\pi_i$ is the proportion of strata $i$ without immunity at time 0.
-That is, at time 0, $\vec{r} = 1 - \vec{\pi}$.
+Now introduce an additional vector of parameters, $\vec{\pi}$, where $\pi_a$ is the proportion of strata $i$ without immunity at time 0, hence, $\vec{r} = 1 - \vec{\pi}$.
 
@@
 
-\todo[inline]{check likelihood is mentioned somewhere in this section}
+\todo[inline]{work out where to present full model equations and then edit/rearrange this section to reflect this}
 
@@
 
-As in \cref{E-backcalc}, I use CIS data to estimate the model for each region independently.
+As in \cref{E-backcalc}, I use CIS data to estimate the model for each region independently, and over the same period as in that chapter.
 As well as the motivation given previously, this allows heterogeneity in the parameters between regions.
@@
 \subsection{Solving the systems of ODEs} \label{SEIR:sec:solving-ODEs}
-I solve the system of ODEs using Euler's method, discretised at timesteps of $\delta t = 0.5$days.
-That is to find the state of the system at time $t+1/2$ given the state at time $t$, I add half the derivative to the state at time $t$.
-\todo[inline]{add these equations}
+I solve the system of ODEs using Euler's method, discretised at timesteps of $\delta t = 1/2$ days.
+That is, if $\vec{x}_t$ is the system state at time $t$ for $t \in \{ 0, 0.5, 1, \dots \}$ then $x_{t+\frac{1}{2}} = x_t + \frac{d\vec{x}(t)}{dt} \delta t$
 
-The prevalence on day $t$, used for the likelihood, is the mean of the solution on the two corresponding days.
+The mean prevalence on day $t$, used for the likelihood (see \cref{SEIR:eq:likelihood}), can be calculated as $\bar{p}_{kat} = \frac{1}{2} ( p_{kat} + p_{ka,t+1/2} )$.
 
 \section{Simulation study} \label{SEIR:sec:sim-study}
-I ran a simulation study of the East of England region.
-This checked that the inference procedure (described in \cref{SEIR:sec:inference-implementation}) was working correctly and that the model was identifiable.
+I ran a simulation study to check that the inference procedure (described in \cref{SEIR:sec:inference-implementation}) was working correctly and that the model was identifiable.
 I ran 100 simulations, with parameters drawn from the priors (see \cref{SEIR:table:priors}) except that the prior on $\psi$ was $\LN(0.048, 0.0035)$.
@@
 Simulating from the prior then attempting to recover the parameters is the basis of simulation-based calibration~\autocites{cookValidation}{taltsValidating}.
-The coverage of the posterior credible intervals should be their nominal value (\eg 95\% of the 95\% credible intervals should contain the true paramter value).
+The coverage of the posterior credible intervals should be their nominal value (\eg 95\% of the 95\% credible intervals should contain the true parameter value).
 Furthermore, the combination of the posteriors across all simulations should recover the prior.
@@
 
+\todo[inline]{add simulation study results}
+
 \section{Application to CIS data} \label{SEIR:sec:application}
 
-All regions' MCMC chains converged satisfactorily.
-Two chains had to be restarted with new seeds to achieve this as they converged to locations which were clearly incorrect (not shown).
-Following this, the Rhat for all parameters except $\sigma_\epsilon$ had Rhat < 1.01.
-All $\sigma_\epsilon$ parameters had Rhat < 1.05.
-All parameters had ESS > 500, except $\sigma_\epsilon$ in London which had ESS of 182.
-Since $\sigma_\epsilon$ is not a parameter of interest, and all other parameters had sufficient ESS and Rhat, I considered the chains to have converged.
-
 Overall the model fits the data well (see \cref{SEIR:fig:prev-young,SEIR:fig:prev-old}).
-There are a few features of the data which are not captured by the model though.
+There are a few features of the data which are not captured by the model.
 In particular, prevalence during October is higher in the 16--25s than the model predicts and lower in the 11--16s.
@@
     \captionsetup{width=0.8\paperwidth}
-    \captionof{figure}[Posterior predictive prevalence (younger ages)]{%
-        Posterior predictive prevalence by region and age for the younger three age groups (see \cref{SEIR:fig:prev-old} for the others).
+    \captionof{figure}[SEIR prevalence goodness-of-fit (younger ages)]{%
+        Goodness-of-fit to prevalence data by region and age for the younger three age groups (see \cref{SEIR:fig:prev-old} for the others).
         The central ribbon and line shows the posterior median and 95\% CrI estimate for the proportion of the relevant strata PCR positive on each day.
@@
     \captionsetup{width=0.8\paperwidth}
-    \captionof{figure}[Posterior predictive prevalence (older ages)]{%
-        Posterior predictive prevalence for the older three age groups (see \cref{SEIR:fig:prev-young} for the other age groups and details).
+    \captionof{figure}[SEIR prevalence goodness-of-fit (older ages)]{%
+        Goodness-of-fit prevalence for the older three age groups (see \cref{SEIR:fig:prev-young} for the other age groups and details).
     }
@@
 Incidence is smooth within each week but shows discontinuities between weeks.
-This is because the parameters within $\vec{\lambda}(t)$ are constant within each week.
-A constant $\vec{\lambda}(t)$ leads to exponential growth over short time periods, once the initial dynamics have died out.
+This is because the parameters within $\matr{\beta}(t)$ are constant within each week.
+A constant $\matr{\beta}(t)$ leads to exponential growth over short time periods, once the initial dynamics have died out.
 When the peak in incidence is estimated to occur varied by region, but due to this reason, is almost always on a Friday (see \cref{SEIR:fig:peak-incidence}).
+\todo[inline]{add peak incidence date plot}
 
@@
 Each region is estimated independently, however correlation is clear across the regions.
+
+\todo[inline]{add Rt results}
+
+\todo[inline]{add attack rates and compare with an external data source}
 \begin{figure}
@@
 
+All regions' MCMC chains converged satisfactorily.
+Two chains had to be restarted with new seeds to achieve this as they converged to locations which were clearly incorrect (not shown).
+Following this, the Rhat for all parameters except $\sigma_\epsilon$ had Rhat < 1.01.
+All $\sigma_\epsilon$ parameters had Rhat < 1.05.
+All parameters had ESS > 500, except $\sigma_\epsilon$ in London which had ESS of 182.
+Since $\sigma_\epsilon$ is not a parameter of interest, and all other parameters had sufficient ESS and Rhat, I considered the chains to have converged.
+
+\todo[inline]{add comparison to backcalculation once I have those results}
+
 \section{Discussion} \label{SEIR:sec:discussion}
 
-Parameterisation of this model is limited by the inference techniques.
-The algorithm used here tends to struggle with a large number of parameters.
-Alternative proposal schemes, such as blocking\todo{cite what RTM does?}, could help, or a more advanced algorithm such as NUTS (used elsewhere in this thesis).
+The results in this chapter show the feasibility of fitting a mechanistic model using only data on prevalence.
+If the distributions for length-of-stay in each compartment are known, then all other model parameters are able to be estimated.
+This includes estimating the age distribution of the infections and the susceptibility of children.
 
-The current model is not very flexible with regards to the age distribution of infections.
-The lack of flexibility is visible in \cref{SEIR:fig:prev-young} where how well the model reproduces the data varies by age group.
-In some regions, there appears to be a trade-off between the 11--16s and 16--25s, with the model misfitting in opposite directions.
-The age distribution of infections determined by $\matr{\beta}$, and to a lesser extent $\vec{\pi}$, the former of which only contains one inferred parameter, $\beta_c$.
-More parameters would help to capture these dynamics more accurately.
-Increasing the number of parameters inferred in the contact matrices would allow for more flexibility in the age distribution of infections, either by introducing additional susceptibility or infectiousness parameters, or by allowing these parameters to vary over time.
-For example, the introduction and relaxation of control measures within schools over time is a plausible mechanism changing the age distribution of infections that cannot be captured in the current model because it only uses school attendance data with no allowance for the changing risk of within-school contacts.
+Unlike most prior work, this modelling is based on a data source that is representative of the general population.
+Therefore, the results are have fewer biases and informed by data to a greater extent than this work.
+There is little-to-no ascertainment bias in the CIS, unlike sources relying on testing behaviour (in the UK, pillar 2 testing).
+The CIS allows data-informed estimates of transmission in children, unlike sources relying on severe events that occur almost entirely in adults\todo{cite something about mildness in children or age distribution of severe events}.
+
+The parameters $\beta_t$ absorb all the time-varying factors not captured by Google mobility and school attendance data.
+\todo{discuss the results and implications for policy or understanding transmission}
+Many factors contribute to the pattern in the $\beta_t$ values, which have a minima near the start of December.
+While this analysis cannot establish causation, several hypotheses can be identified that are worth further investigation.
+The decreases in $\beta_t$ over the autumn could be due to increased caution as cases and deaths increased (\eg masking or stricter adherence to social distancing)\todo{cite something about caution and reaction to rising cases, or survey data showing this happened - CoMix?}, or the impact of local measures in the highest transmission areas\todo{cite local measures}.
+The increases in $\beta_t$ in December coincide with the rise of the Alpha variant~\autoref{lythgoeLineage}, known to be more transmissible~\auocite{daviesEstimated}.
+A multiple variant model might be able to capture this rise more accurately.
+Such a model would stratify positive tests in the CIS by pre-Alpha or Alpha variant.
+Then, each variant would be explicitly included in the model to fit to this data.
+
+These factors, and many others, are likely to be correlated across the country.
+This can explain that similar patterns emerge for $\beta_t$ in different regions (see \cref{SEIR:fig:beta-walk}), despite these not being estimated independently.
+These correlations could be exploited by linking their values across regions.
+This could improve the precision of the estimates, but would require a more complex model.
+
+\todo[inline]{discuss comparisons with the backcalculation}
+
+The current model could be more flexible with regards to the age distribution of infections.
+This would better exploit one of the strengths of the CIS: that there is data on all age groups.
+In some regions, there appears to be a trade-off between the 11--16s and 16--25s, with the model misfitting in opposite directions (see \cref{SEIR:fig:prev-young}).
+This implies that the data would allow identification of a more flexible age-based model.
+Currently, the age distribution of infections is determined by $\matr{\beta}$, and to a lesser extent $\vec{\pi}$, the former of which only contains one inferred parameter that affects the age distribution, $\beta_c$.
+More parameters could capture the age distribution more accurately.
+There are various ways to increase the model's flexibility, including introducing additional susceptibility or infectiousness parameters, or by allowing these parameters to vary over time.
+It is plausible that these distributions changed over time in a way not captured by $\matr{\kappa}$, for example due to the introduction and relaxation of control measures within schools.
+$\matr{\kappa}$ only captures changes in in-school transmission through changes in school attendance, not incorporating the probability of transmission, which could be modelled as changes in $q_{aa}$ for $a$ of school age.
+Similarly, there is evidence that $q_c$ is different for Alpha and pre-Alpha variants\todo{cite a systematic review}, which could again motivate these changing over time.
 
@@
 
-The parameters $\beta_t$ absorb all the time-varying factors not captured by Google mobility and school attendance data.
-These factors are likely to be correlated across the country, giving the similar patterns for different regions (see \cref{SEIR:fig:beta-walk}).
-Exploiting these correlations by linking their values across regions could improve the precision of the estimates.
-This would require a more complex model without the independence of parameters between regions.
-
-Various factors could explain the pattern in the $\beta_t$ values, which have a minima near the start of December.
-Some hypothesis follow, although this analysis is unable to distinguish causality.
-Increased caution as cases and deaths increased could explain the decrease (\eg masking or stricter adherence to social distancing).
-The increase coincides with the rise of the Alpha variant, known to be more transmissible.
-This is a likely cause of the increase.
-A multiple variant model might be able to capture this more accurately.
-This would stratify positive tests in CIS by pre-Alpha or Alpha variant.
-Then, each variant would be explicitly included in the model to fit to this data.
+This discussion makes clear that the results are limited by the number of parameters, currently limited by the inference algorithm.
+The algorithm used here tends to struggle with a large number of parameters.
+Alternative proposal schemes, such as updating the parameters in blocks\todo{cite what RTM does?}, could help, or a more advanced algorithm such as NUTS (used elsewhere in this thesis).
 
 Extending the estimates here to the rest of the pandemic would clearly be of interest.
-However, this requires a more complex model structure.
 I chose the end of the modelling period to coincide with the vaccination rollout.

### 2023-11-21 incidence-prevalence.tex

--- 2023-11-21 incidence-prevalence.tex
@@
 The duration is a positive integer, \ie $D_i \in \{1, 2, \dots, \dmax \}$, where $\dmax$ is the maximum duration.
-The first way of expressing the distribution of $D_i$ is as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
+The first way of expressing the distribution of $D_i$ is as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{k=1}^d f_{D}(k)$.
 The \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$ (a common alternative convention for survival analysis with discrete variables is to define the survival function as $\prob(D >d)$).
-The \emph{hazard} on day $d$ is the probability of recovering at time $d$ conditional on having survived to time $d$: $\lambda(d) = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
+The \emph{hazard} on day $d$ is the probability of recovering on day $d$ conditional on still being positive on day $d$: $\lambda(d) = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
 Throughout this thesis, the survival function will be the primary quantity of interest, and it will normally be expressed in terms of the hazard; these two quantities are related by $S(d) = \prod_{i=1}^{d-1} (1 - \lambda(i))$.
 
-Since $D_i$ is a discrete random variable, $f_D(d) \geq 0$ for all $d$ and $\sum_{i=1}^{\dmax} f_D(i) = 1$.
+Since $D_i$ is a discrete random variable, $f_D(d) \geq 0$ for all $d$ and $\sum_{k=1}^{\dmax} f_D(k) = 1$.
 This implies that $F_D(d)$ is monotonically increasing.

### 2023-11-21 SEIR.tex
fix typo

@@
 The decreases in $\beta_t$ over the autumn could be due to increased caution as cases and deaths increased (\eg masking or stricter adherence to social distancing)\todo{cite something about caution and reaction to rising cases, or survey data showing this happened - CoMix?}, or the impact of local measures in the highest transmission areas\todo{cite local measures}.
-The increases in $\beta_t$ in December coincide with the rise of the Alpha variant~\autoref{lythgoeLineage}, known to be more transmissible~\auocite{daviesEstimated}.
+The increases in $\beta_t$ in December coincide with the rise of the Alpha variant~\autoref{lythgoeLineage}, known to be more transmissible~\autocite{daviesEstimated}.
 A multiple variant model might be able to capture this rise more accurately.

### 2023-11-21 ATACCC.tex
fix maths formatting

@@
 $v_{\max,i}$ is the maximum viral load, which occurs at time $\tau_{\max,i}$.
-As $|\tau - tau_\max|$ becomes large, the viral load trajectory tends towards exponential growth (before the peak) or decline (after the peak) in viral load at rates $a_i$ and $b_i$ respectively.
+As $\lvert \tau - \tau_{\max} \rvert$ becomes large, the viral load trajectory tends towards exponential growth (before the peak) or decline (after the peak) in viral load at rates $a_i$ and $b_i$ respectively.
 Before the peak, this is because $b(\tau-\tau_{\max,i})$ is small and hence $a\exp(b(\tau-\tau_{\max,i}))$ is negligible.

### 2023-11-21 incidence-prevalence.tex

@@
 The duration is a positive integer, \ie $D_i \in \{1, 2, \dots, \dmax \}$, where $\dmax$ is the maximum duration.
-The first way of expressing the distribution of $D_i$ is as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{k=1}^d f_{D}(k)$.
-The \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$ (a common alternative convention for survival analysis with discrete variables is to define the survival function as $\prob(D >d)$).
+The first way of expressing the distribution of $D_i$ is as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_work out where to present full model equations and then edit/rearrange this section to reflect this{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
+The \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$.
+%(a common alternative convention for survival analysis with discrete variables is to define the survival function as $\prob(D >d)$).
 The \emph{hazard} on day $d$ is the probability of recovering on day $d$ conditional on still being positive on day $d$: $\lambda(d) = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
@@
 
-Since $D_i$ is a discrete random variable, $f_D(d) \geq 0$ for all $d$ and $\sum_{k=1}^{\dmax} f_D(k) = 1$.
-This implies that $F_D(d)$ is monotonically increasing.
-Therefore, $S(d)$ is monotonically decreasing.
-Both $F_D$ and $S$ are in the interval $[0, 1]$.
-$\lambda(d)$ is in the interval $[0, 1]$, and can otherwise vary arbitrarily (\ie there is no monotonic or sum constraint).
+Since $D_i$ is a discrete random variable, $0 \leq f_D(d) \leq 1$ for all $d$ and $\sum_{k=1}^{\dmax} f_D(k) = 1$.
+This implies that $F_D(d) \in [0, 1]$ and increases monotonically.
+Therefore, $S(d) \in [0, 1]$ and decreases monotonically.
+Similarly, $\lambda(d) \in [0,1]$, but can otherwise vary arbitrarily (\ie there is no monotonic or sum constraint).
 
@@
 
-A commonly-used point process for infectious disease epidemiology is the time-inhomogenous Poisson process~\autocites{brookmeyerMethod}{paganoHIV}{rosenbergBackcalculation}{brookmeyerBackcalculation}.
-A Poisson process is a point process with independent increments, and the property that $K(s) - K(t) \dist \Poi\left(\int^s_t \nu(r) dr \right)$ where $\nu(r) \geq 0$ is the \emph{intensity} of the epidemic process at time $r$~\autocites[244]{yanDistribution}.
+A commonly-used point process for infectious disease epidemiology is the time-inhomogenous Poisson process~\autocites[e.g., in the context of HIV:][]{brookmeyerMethod}{paganoHIV}{rosenbergBackcalculation}{brookmeyerBackcalculation}.
+A Poisson process is a point process with independent increments, and the property that $K(s) - K(t) \dist \Poi\left(\int^s_t \nu(r) dr \right)$ where $\nu(r) \geq 0$ is the \emph{intensity} of the infection process at time $r$~\autocites[244]{yanDistribution}.
 A time-homogenous Poisson process is the special case where $\nu(r)$ is constant.

### 2023-11-21 SEIR.tex
Start on SEIR to dos

--- 2023-11-21 SEIR.tex
@@
 
-\todo[inline]{I think this intro needs references but I'm struggling to find suitable ones}
 This section introduces the class of mechanistic transmission models known as \emph{compartmental} models.
@@
 For an introduction to stochastic models, see \textcite[chapter 6]{keelingModeling} or \textcite{birrellEvidence}.
-% \todo{Maybe say something about still having a stochastic observation model, if not clear enough from intro}
+
+These classifications only describe the transmission model.
+The observation model, which links $\vec{x}(t)$ to data, is generally includes some stochasticity regardless~\autocite{birrellEvidence}.
 
@@
 The SIR model is named because each individual in the population is divided into one of three states: susceptible, infectious, or recovered.
-This model has several assumptions which make it inappropriate to model SARS-CoV-2; therefore, the remainder of this sections relaxes these assumptions.
-First, \cref{SEIR:sec:non-exponential} relaxes the assumption that the infectious period is exponentially-distributed to allow gamma distributions.
-Then, \cref{SEIR:sec:SEIR} relaxes the assumption that individuals are infectious immediately after they are infected; it does this by extending the SIR model to the SEIR model with the additional E signifying an \emph{exposed} compartment.
-Finally, \cref{SEIR:sec:structured-populations} allows heterogeneity in the population, introducing stratification by important characteristic(s).
+This model has several assumptions which make it inappropriate to model SARS-CoV-2; therefore, in the remainder of this section I introduce standard extensions to the SIR model which relax these assumptions.
+First, in \cref{SEIR:sec:non-exponential} I relax the assumption that the infectious period is exponentially-distributed to allow gamma distributions.
+Then, in \cref{SEIR:sec:SEIR} I relax the assumption that individuals are infectious immediately after they are infected; it does this by extending the SIR model to the SEIR model with the additional E signifying an \emph{exposed} compartment.
+Finally, in \cref{SEIR:sec:structured-populations} I include heterogeneity in the population, stratifying by important characteristic(s).
 % Finally, \cref{SEIR:sec:time-varying-foi} relaxes the assumption of a constant transmission rate to allow for time-varying behavioural changes.
@@
 $\beta$ is the transmission rate, absorbing several terms as described below.
-%\todo{check what beta is called: effective transmission rate or transmission rate or something else?}, or the number of individuals an infected individual infects per day.
 $\gamma$ is the recovery rate, or the proportion of infected individuals that recover per day; hence, $1/\gamma$ is the mean infectious period~\autocite[367]{keelingModeling}.
@@
 \begin{itemize}
-    \item For any short period of time when $s(t)$ is approximately constant, the system of ODEs can be solved analytically to produce exponential growth\todo{cite exponential growth}.
+    \item For any short period of time when $s(t)$ is approximately constant, the system of ODEs can be solved analytically to produce exponential growth~\autocite[section 1.2]{diekmannMathematical}.
     A notable example is the early part of an epidemic when $s(t) \approx 1$.
@@
 Multiplying by the number of infected individuals in strata $a'$, $N_{a'} i_{a'}(t)$, gives $\kappa_{aa'} q_{aa'} s_a(t) N_{a'} i_{a'}(t)$.
-The constant terms here are absorbed into the WAIFW matrix, giving $\beta_{aa'} = \kappa_{aa'} q_{aa'} N_{a'}$.
+The constant terms here are absorbed into the WAIFW matrix, giving $\beta_{aa'} = \kappa_{aa'} q_{aa'} N_{a'}$~\autocite{section 9.2}{diekmannMathematical}.
 The total force of infection on strata $a$ from all strata is, therefore, $\lambda_a(t) = \sum_{a' \in \set{A}} \beta_{aa'} i_{a'}(t)$.
@@
 This distribution is proportional to the \emph{dominant eigenvector} (the eigenvector corresponding to the largest eigenvalue) of $\matr{\beta}$.
-Denote by $\rho(\matr{\beta})$ the \emph{dominant eigenvalue} (the largest eigenvalue).
-The basic reproduction number is then defined as $R_0 = \rho(\matr{\beta}) / \gamma$.
+Denote by $D(\matr{\beta})$ the \emph{dominant eigenvalue} (the largest eigenvalue).
+The basic reproduction number is then defined as $R_0 = D(\matr{\beta}) / \gamma$.
 This $R_0$ keeps the same interpretation as in the homogeneous case, \ie the average number of secondary infections caused by a single infection in a fully susceptible population, where the average is taken over the distribution implied by the dominant eigenvector.
@@
 
-\todo[inline]{define the strata in this section, and edit the below text to make sense in this new context}
-
 A parsimonious model is needed to parameterise the WAIFW matrix elements with (see \cref{SEIR:sec:structured-populations}).
@@
 
-I use the following age-based strata, which align closely with the strata previously used for analysis of the CIS data\todo{cite analyses using these strata}.
+I use the following age-based strata, which align closely with the strata previously used for analysis of the CIS data~\autocites[e.g.:][]{houseInferring}{walkerTracking}.
 \begin{enumerate}
@@
 \end{enumerate}
+The CIS only recruited individuals aged 2 or older, however, I assume tests are representative of the whole $[0, 11)$ age group.
 
@@
 % Ideally, all the $\beta_{ij}$ terms would be estimated, but there is normally inadequate data to do so, requiring assumptions to be made to reduce the number of parameters~\autocite[65]{keelingModeling}.
-% % $\beta_{ij}$ is then estimated, although often with a parsimonious model\todo{ref / give examples / delete on how age-specfic susceptibility / transmissibility happens}.
+% % $\beta_{ij}$ is then estimated, although often with a parsimonious model
 % A common assumption is that $\beta_{ij}$ can be decomposed into two independent components such that $\beta_{ij} = \beta \beta^\text{susc}_{i} \beta^\text{inf}_j$.
@@
 % Then $\beta^\text{susc}_i$ is the relative risk of an individual in stratum $i$ being infected, relative to an individual in stratum $a_0$.
-These measures include both biological differences, \eg due to differences in immune systems, or behavioural, \eg if infants are less hygenic than adults.
+These measures include both biological differences, \eg due to differences in immune systems, or behavioural, \eg if infants are less hygienic than adults.
 
@@
 
-Previous studies for pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults\todo{cite where age difference in transmission comes from}.
+Previous studies for pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults~\autocites{chenRole}{vinerTransmission}.
 Therefore, I set $q^\text{inf}_{a'} = 1$ for all $a'$, $q^\text{susc}_{a} = q_c$ for the youngest stratum, and $q^\text{susc}_{a} = 1$ for the remaining strata.
+The interpretation of $q_c$ is that children have a susceptibility, relative to adults, of $q_c$.
 
@@
 First, calculate $\R$ from $\psi$ using \cref{SEIR:eq:rtoR}; denote this $\hat{R}$.
-Then calculate $R^* = \rho(\matr{\beta}) / \gamma$ when $\beta_1=1$ (recall that $\rho(\matr{\beta})$ denotes the dominant eigenvalue of $\matr{\beta}$).
+Then calculate $R^* = D(\matr{\beta}) / \gamma$ when $\beta_1=1$ (recall that $D(\matr{\beta})$ denotes the dominant eigenvalue of $\matr{\beta}$).
 Multiplying a matrix by a scalar multiplies its dominant eigenvalue by that same scalar, therefore, setting $\beta_1 = \hat{R} / R^*$ gives the desired means that $\R$ at time 0 is $\hat{R}$, as desired.
@@
 The initial conditions and transition rate parameters are unindentifiable in a SEIR model~\autocite{dankwaStructural}, therefore I used assumed values or strongly informative priors on them.
-All other parameters had weakly informative priors.
+$\sigma_\epsilon$ and $\rho$ are given priors that penalise model complexity.
+The prior on $\sigma_\epsilon$ is a penalised complexity prior, favouring a null model of constant values for $\beta(t)$~\autocite{simpsonPenalising}.
+The prior on $\rho$ favours a null model of a binomial likelihood.
+The remaining priors had weakly informative priors.
 
@@
     \hline \\
-    Mean latent period & $1/\sigma$ & 3.5 days (fixed) & \Textcite{zhaoEstimating} \\
-    Mean infectious period & $1/\gamma$ & 4 days (fixed) & \Textcite{zhaoEstimating} \\
+    Mean latent period & $1/\sigma$ & 3.5 days (fixed) & Based on \textcite{zhaoEstimating} \\
+    Mean infectious period & $1/\gamma$ & 4 days (fixed) & Based on \textcite{zhaoEstimating} \\
     Initial proportion without immunity & $\vec\pi$ & See \cref{SEIR:table:immunity-prior} & \\
     Initial proportion infected & $i^+$ & $\BetaDist(0.5, 1000)$ & Weakly informative \\
-    Relative susceptibility of children & $\beta_c$ & $\LN(-0.4325, 0.1174)$ &  \\
+    Susceptibility of children & $q_c$ & $\LN(-0.4325, 0.1174)$ & Based on \textcite{vinerTransmission}  \\
     Initial growth rate & $\psi$ & $\N(0.06, 0.04)$ & Weakly informative \\
-    Standard deviation of random walk steps on log scale & $\sigma_\epsilon$ & $\Exponential(80)$ & Weakly informative \\ %penalised model complexity prior~\autocite{simpsonPenalising} \\
-    Overdispersion of observations & $\rho$ & $\Exponential(2 \times 10^5)$ & Weakly informative
+    Standard deviation of random walk steps on log scale & $\sigma_\epsilon$ & $\Exponential(80)$ & Penalising towards constant \\
+    Overdispersion of observations & $\rho$ & $\Exponential(2 \times 10^5)$ & Penalising towards binomial
 \end{tabular}
-\todo[inline]{Dani/Paul: I think I took the prior for $\beta_c$ from RTM, but can't find the reference now - any ideas?}
 \caption[SEIR model priors]{Priors for each parameter. For details of the distributions and their parameterisations see \cref{E-distributions}.}
@@
 
+\todo[inline]{sort region names in plots}
+
 Overall the model fits the data well (see \cref{SEIR:fig:prev-young,SEIR:fig:prev-old}).
@@
 When the peak in incidence is estimated to occur varied by region, but due to this reason, is almost always on a Friday (see \cref{SEIR:fig:peak-incidence}).
-\todo[inline]{add peak incidence date plot}
-
 \begin{figure}
@@
     \label{SEIR:fig:incidence}
+\end{figure}
+
+\begin{figure}
+    \includegraphics{SEIR/CIS/p_peak}
+    \caption[Posterior estimates of peak incidence timing]{%
+        Posterior estimates (95\% CrI) of incidence by region and age.
+        See main text for discussion.
+    }
+    \label{SEIR:fig:incidence}
+    \todo[inline]{write peak incidence caption}
 \end{figure}
@@
 The decreases in $\beta_t$ over the autumn could be due to increased caution as cases and deaths increased (\eg masking or stricter adherence to social distancing)\todo{cite something about caution and reaction to rising cases, or survey data showing this happened - CoMix?}, or the impact of local measures in the highest transmission areas\todo{cite local measures}.
-The increases in $\beta_t$ in December coincide with the rise of the Alpha variant~\autoref{lythgoeLineage}, known to be more transmissible~\auocite{daviesEstimated}.
+The increases in $\beta_t$ in December coincide with the rise of the Alpha variant~\autocite{lythgoeLineage}, known to be more transmissible~\autocite{daviesEstimated}.
 A multiple variant model might be able to capture this rise more accurately.

### 2023-11-21 SEIR/CIS/posterior_predictive.R

@@ -156,12 +156,19 @@ p_peak = peak_days |>
     filter(p_peak > 0.02) |>
     ggplot(aes(peak_incidence, p_peak, colour = age)) +
     geom_point() +
-    facet_wrap(~region, ncol = 2) 
+    facet_wrap(~region, nrow = 2) +
+    standard_plot_theming() +
+    theme(legend.position = "bottom") +
+    labs(
+        x = "Day",
+        y = "Posterior probability",
+        colour = "Age group"
+    )
 ggsave(
     filename = here::here("SEIR/CIS/p_peak.png"),
     plot = p_peak,
-    width = 10,
-    height = 20,
+    width = 15,
+    height = 10,
     units = "cm",
     dpi = 300
 )

### 2023-11-21 incidence-prevalence.tex
Updates from Overleaf

@@
 The duration is a positive integer, \ie $D_i \in \{1, 2, \dots, \dmax \}$, where $\dmax$ is the maximum duration.
-The first way of expressing the distribution of $D_i$ is as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_work out where to present full model equations and then edit/rearrange this section to reflect this{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
+The first way of expressing the distribution of $D_i$ is as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
 The \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$.
 %(a common alternative convention for survival analysis with discrete variables is to define the survival function as $\prob(D >d)$).
-The \emph{hazard} on day $d$ is the probability of recovering on day $d$ conditional on still being positive on day $d$: $\lambda(d) = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
+The \emph{hazard} on day $d$ is the probability of recovering at time $d$ conditional on having survived to time $d$: $\lambda(d) = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
 Throughout this thesis, the survival function will be the primary quantity of interest, and it will normally be expressed in terms of the hazard; these two quantities are related by $S(d) = \prod_{i=1}^{d-1} (1 - \lambda(i))$.
 
-Since $D_i$ is a discrete random variable, $0 \leq f_D(d) \leq 1$ for all $d$ and $\sum_{k=1}^{\dmax} f_D(k) = 1$.
+Since $D_i$ is a discrete random variable, $0 \leq f_D(d) \leq 1$ for all $d$ and $\sum_{i=1}^{\dmax} f_D(i) = 1$.
 This implies that $F_D(d) \in [0, 1]$ and increases monotonically.
@@
 \section{Infection process} \label{inc-prev:sec:infection-process}
-The infection process is the mathematical description of how infections occur.
+The infection process is the mathematical description of incidence.
 It is a \emph{point} or \emph{counting} process, where each infection is an event.
@@
 These additional infections will propagate the epidemic further, increasing the intensity at future times.
-Point processes can be generalised to \emph{branching processes} to account for this feedback~\autocite[246]{yanDistribution}, although these are beyond the scope of this thesis.
+Point processes can be extended to \emph{branching processes} to account for this feedback~\autocite[246]{yanDistribution}, although these are beyond the scope of this thesis.
 

### 2023-11-21 ATACCC.tex
comments on ATACCC chapter

@@
 \Cref{E-inc-prev} showed that the duration of PCR positivity is a key parameter for estimating the incidence of SARS-CoV-2 infections.
-This chapter estimates the duration of PCR positivity using the viral load data from the ATACCC study (described in\todo{ref section describing ATACCC}).
-I model a biomarker, the log viral load, over the course of an infection.
+The aim of this chapter is to estimate the duration of PCR positivity using the viral load data from the ATACCC study (described in\todo{ref section describing ATACCC}).
+I model a biomarker, the log viral load, over the course of an infection.\todo{make sure log viral load and Ct values are introduced in chapter 2}
 From this model, I derive the distribution of the duration of PCR positivity.
 
-\Cref{ATACCC:sec:background} explains how biomarkers are typically modelled, both for SARS-CoV-2 and historically in the biostatistical literature.
-Next, \cref{ATACCC:sec:hakki} explains \textcite{hakkiOnset}'s application of these methods to the ATACCC data and in Section \cref{ATACCC:sec:observation-modification} the observation model of Hakki is modified  to improve convergence.
-This modification is based on an earlier analysis I did with a slightly different model and a preliminary version of the dataset, described in \cref{ATACCC:sec:original-analysis}.
-I convert the model and parameter estimates of the biomarker to a duration of PCR positivity in \cref{ATACCC:sec:duration}.
+First, I will explain how biomarkers are typically modelled, both for SARS-CoV-2 and historically in the biostatistical literature (\cref{ATACCC:sec:background}).
+Next, I describe the model of \textcite{hakkiOnset} in detail, which applies these methods to the ATACCC data (\cref{ATACCC:sec:hakki}) and before I modifying its observation model (\cref{ATACCC:sec:observation-modification})\todo{think if some justification is needed here, and if so how to do it - remember Hakki is a published and peer-reviewed analysis}.
+This modification is based on an earlier analysis using a preliminary version of the dataset, described in \cref{ATACCC:sec:original-analysis}.
+Finally, I convert the model and parameter estimates of the biomarker's trajectory to a duration of PCR positivity (\cref{ATACCC:sec:duration}).
 I will refine this estimate in future chapters before its use in estimating transmission.
@@
 There are either using ordinary differential equations (ODEs) with a biological interpretation or using mathematical functions that are flexible enough to capture the dynamics.
-Both of these approaches have been used with other viruses and biomarkers.
-
-ODE-based approaches have parameters represent biological properties of the virus and host, such as the immune response ~\autocites[e.g.:][]{ejimaEstimation,keVivo,kimQuantitative,goncalvesTiming,perelsonMechanistic}.
+The ODE-based approach uses parameters representing biological properties of the virus and host~\autocites[e.g.:][]{ejimaEstimation,keVivo,kimQuantitative,goncalvesTiming,perelsonMechanistic}.
 Such models are rarely analytically-tractable and the system's evolution is computed using numeric methods.
-
-The second approach is to approximate the trajectories from these models using general mathematical functions.
-In these models, the function's parameters are clearly related to features of the biomarker's trajectory, such as the time from first detection to peak viral load, and the viral load at the peak.
-Most common is a piecewise model, where the logarithm of viral load increases linearly, peaks instantaneously, and then decreases linearly~\autocites{clearyUsing,kisslerViral,larremoreTest}; although more flexible functions have also been suggested~\autocites{quiltyQuarantine}.
-
-In both approaches, random effects are used to represent variation between hosts within the population, capturing the non-independence of longitudinal data from the same host.
-This chapter uses the model of \textcite{hakkiOnset}.
-This model is approximately piecewise log-linear, however, a slight adjustment is made which means that the model is continuous (see \cref{ATACCC:sec:hakki} for details).
-
-Random effects are used to capture the variation between individuals and the correlation between measurements within the same individual.
-In this context, a random effect means that parameter values are specific to each individual, rather than being a population average.
-The parameter's distribution across the population is assumed to be some known distribution, normally a normal or log-normal distribution~\autocite[chapter 24]{lashModern}.
-Specifically, if $\phi_j$ is the value of some parameter for individual $j$, $\mu_\phi$ is the population mean of $\phi$, and $\sigma_\phi$ is the population standard deviation, then we model $\phi_j$ as: $\phi_j \sim \N(\mu_\phi, \sigma_\phi^2)$ or $\phi_j \sim \LN(\mu_\phi, \sigma_\phi^2)$.
-$\mu_\phi$ and $\sigma_\phi$ are parameters; in a Bayesian setting, a prior is placed on them in the normal way.
-This generalises to a vector of parameters per individual, $\vec{\phi}_j$, in the obvious way.
-In this chapter, $\mu_\phi$ and $\sigma_\phi$ are the parameters of interest, because we want to generalise our model across the population.
-The $\phi_j$s, which tells us about the parameter only for individuals in our sample, are considered nuisance parameters.
-
-Following much of the literature referenced in this section, I will take a Bayesian approach in this chapter.
-In the Bayesian context, random effect models are often referred to as hierarchical or multilevel models because they can be interpreted as placing a prior at two levels of a parameter hierarchy~\autocite[chapter 5]{gelmanBDA}.
-$\phi_j$ would be referred to as the bottom level of the hierarchy, with $\mu$ and $\sigma$ being the top level.
-More complex models could have more levels, for instance if the individuals are grouped into subpopulations.
+% Both of these approaches have been used with other viruses and biomarkers.
+The approach that is the focus of this chapter is using a pieceiwse linear model for the biomaker trajectory and random effects to allow for within-individual correlation.
 
@@
 
+In these models, the function's parameters are clearly related to features of the biomarker's trajectory, such as the time from first detection to peak viral load, and the viral load at the peak.
+The piecewise linear model is the most common in the context of SARS-CoV-2 viral load~\autocites{clearyUsing,kisslerViral,larremoreTest}
+Here, the logarithm of viral load increases linearly, peaks instantaneously, and then decreases linearly; more flexible functions have also been suggested~\autocites{quiltyQuarantine}.
+This model for the trajectory I adopt in this chapter, from \textcite{hakkiOnset}, is based on the piecewise log-linear model, however, they approximate the model so that the trajectory is continuous (see \cref{ATACCC:sec:hakki} for details).
+
+Random effects represent the within-individual correlation within the population while borrowing information between individuals~\autocite[chapter 24]{lashModern}.
+Measurements of viral load from the same individual are correlated.
+A random effect means that these measurements share parameter(s), providing correlation.
+However, the random effect(s) for each individual is drawn towards a common mean, allowing information on their value to be shared between individuals.
+Following much of the literature referenced in this section, I will take a Bayesian approach in this chapter.
+In the Bayesian context, random effect models are often referred to as hierarchical or multilevel models because they can be interpreted as placing a prior at two levels of a parameter hierarchy~\autocite[chapter 5]{gelmanBDA}.
+% $\phi_j$ would be referred to as the bottom level of the hierarchy, with $\mu$ and $\sigma$ being the top level.
+% More complex models could have more levels, for instance if the individuals are grouped into subpopulations.
+
+% Random effects are used to capture the variation between individuals and the correlation between measurements within the same individual.
+% In this context, a random effect means that parameter values are specific to each individual, rather than being a population average.
+% The parameter's distribution across the population is assumed to be some known distribution, normally a normal or log-normal distribution~\autocite[chapter 24]{lashModern}.
+% Specifically, if $\phi_j$ is the value of some parameter for individual $j$, $\mu_\phi$ is the population mean of $\phi$, and $\sigma_\phi$ is the population standard deviation, then we model $\phi_j$ as: $\phi_j \sim \N(\mu_\phi, \sigma_\phi^2)$ or $\phi_j \sim \LN(\mu_\phi, \sigma_\phi^2)$.
+% $\mu_\phi$ and $\sigma_\phi$ are parameters; in a Bayesian setting, a prior is placed on them in the normal way.
+% This generalises to a vector of parameters per individual, $\vec{\phi}_j$, in the obvious way.
+% In this chapter, $\mu_\phi$ and $\sigma_\phi$ are the parameters of interest, because we want to generalise our model across the population.
+% The $\phi_j$s, which tells us about the parameter only for individuals in our sample, are considered nuisance parameters.
+
 \section{\texorpdfstring{\textcite{hakkiOnset}}{Hakki \etal (2022)}} \label{ATACCC:sec:hakki}
@@
 The model of \textcite{hakkiOnset} has two components: a latent viral load model and an observation model.
-The viral load model produces a deterministic trajectory for the viral load of an individual, based on the model parameters.
-The observation model then describes how the PCR testing process produces noisy observations of this viral load, including the possibility of false positive and negatives, and the limit of detection of the PCR test.
+The viral load model produces a deterministic trajectory for the viral load of an individual, a function of the model parameters.
+The observation model then describes how the PCR testing process produces noisy observations of this viral load.
+The observation model includes the possibility of false positive and negatives, and the limit of detection of the PCR test.
 \todo{Make sure limit of detection is introduced previously when I first talk about PCR}
@@
 
-Viral load exhibits exponential growth followed by exponential decline.
-\Textcite{hakkiOnset} use a model which is approximately piecewise linear but continuous.
-A continuous viral load trajectory should give better performance when using gradient-based inference methods such as Hamiltonian Monte Carlo (HMC)\todo{cite paper or ref section}.
-The model for the viral load at time $t$, $\hat{v}_i(t)$, in individual $i$ is shown in \cref{ATACCC:fig:viral-load-model} and given by:
+Several features of the \textcite{hakkiOnset} model can be motivated by the desire to apply Stan, a modern and convenient MCMC implementation of the NUTS algorithm (see\todo{ref wherever I end up describing NUTS}).
+The first of these is using a continuous viral load trajectory rather than the more common piecewise linear model.
+As NUTS uses gradient information, it is more efficient when the model is differentiable.
+This leads to the following model for the viral load at time $t$, $v_i(t)$, in individual $i$ is shown in \cref{ATACCC:fig:viral-load-model} and given by:
 \begin{align}
@@
 \end{align}
-Time is continuous and relative to the time that individual $i$ was first tested.
+Time, $\tau$, is continuous and defined uniquely for each individual such that $\tau = 0$ at the time that individual $i$ was first tested.
 $v_{\max,i}$ is the maximum viral load, which occurs at time $\tau_{\max,i}$.
 As $\lvert \tau - \tau_{\max} \rvert$ becomes large, the viral load trajectory tends towards exponential growth (before the peak) or decline (after the peak) in viral load at rates $a_i$ and $b_i$ respectively.
-Before the peak, this is because $b(\tau-\tau_{\max,i})$ is small and hence $a\exp(b(\tau-\tau_{\max,i}))$ is negligible.
-After the peak, this is because $b\exp(-a(\tau-\tau_{\max,i}))$ is negligible.
+% Before the peak, this is because $b(\tau-\tau_{\max,i})$ is small and hence $a\exp(b(\tau-\tau_{\max,i}))$ is negligible.
+% After the peak, this is because $b\exp(-a(\tau-\tau_{\max,i}))$ is negligible.
 This happens quickly for realistic values of the parameters.
@@
 
-The noisy observations are modelled by assuming each measurement of log viral load is a realisation of a iid Normally-distributed random variable error with mean 0 and standard deviation $\sigma_v$.
+The noisy observations are modelled by assuming each measurement of log viral load is a realisation of an iid Normally-distributed random variable error with mean 0 and standard deviation $\sigma_v$.
+\todo{Paul/Dani: is this phrasing "a realisation of [a] random variable" OK?}
 The measurements are left-censored at the limit of detection, $\vlod$, meaning that any observation which would give a log viral load of less than $\vlod$ is instead observed as negative.
@@
 p(y_{i,t} \mid \cdot) &= \begin{cases}
-    \rho f_N(y_{i,t} \mid x_0, \sigma_0^2) + (1 - \rho) f_N(y_{i,t} \mid \log \hat{v}_i(t), \sigma_v^2) & \text{if $y_{i,t} > 0$}\\
-    \rho F_N(v_\text{lod} \mid x_0, \sigma_0^2) + (1 - \rho) F_N(\vlod \mid \log \hat{v}_i(t), \sigma_v^2) & \text{if $y_{i,t} = 0$} \label{ATACCC:eq:hakki-likelihood}
+    \rho f_N(y_{i,t} \mid x_0, \sigma_0^2) + (1 - \rho) f_N(y_{i,t} \mid \log v_i(t), \sigma_v^2) & \text{if $y_{i,t} > v_\text{lod}$}\\
+    \rho F_N(v_\text{lod} \mid x_0, \sigma_0^2) + (1 - \rho) F_N(\vlod \mid \log v_i(t), \sigma_v^2) & \text{if $y_{i,t} = v_\text{lod}$} \label{ATACCC:eq:hakki-likelihood}
 \end{cases}
@@
 
-Conditional on all $\hat{v}_i$s, the observations are assumed independent.
-Because $\hat{v}_i$ is a deterministic transform of $\vec{\phi}_i$ and $\tau_i$, the observations are independent conditional on the model parameters, and the full likelihood is simply the product of the likelihoods of each observation.
+Conditional on all $v_i$s, the observations are assumed independent.
+Because $v_i$ is a deterministic transform of $\vec{\phi}_i$ and $\tau_i$, the observations are independent conditional on the model parameters, and the full likelihood is simply the product of the likelihoods of each observation.
 
@@
 where $\vec{\sigma}_\phi$ is a vector of the standard deviations of the parameters, $\matr{C}_\phi$ is a correlation matrix, and $\circ$ is element-wise multiplication.
-Combining $\matr{C}$ and $\vec{\sigma}_\phi$ allows recovering $\matr\Sigma_\phi$ because $(\matr{\Sigma}_\phi)_{ij} = (\vec{\sigma}_\phi)_i (\vec{\sigma}_\phi)_j (\matr{C}_\phi)_{ij}$.
+These are defined such that, by construction, $(\matr{\Sigma}_\phi)_{ij} = (\vec{\sigma}_\phi)_i (\vec{\sigma}_\phi)_j (\matr{C}_\phi)_{ij}$.
 Correlation matrices have diagonal elements all 1, therefore this reparameterisation has the same number of parameters as the original model.
@@
 The inference was performed using RStan~2.32.3~\autocite{RStan-2-32-3} and Stan~2.32.2~\autocite{Stan-2-32-2}.
-Inference settings were the same as \textcite{hakkiOnset}.
+Inference settings were the same as \textcite{hakkiOnset}, which included 8 chains with a total of 8000 iterations each with the first 3000 used as warm-up.
 \begin{table}
@@
 
-Based on my previous analysis (see \cref{ATACCC:sec:original-analysis}) and the fit to the data, I hypothesised that the cause of this was overcomplicating the model for test failures.
-Specifically, false positives do not appear to be present in the dataset.
+Based on the fit to the data achieved when implementing an earlier version of the model (see \cref{ATACCC:sec:original-analysis}), I hypothesised that the cause of this was overcomplicating the model for test failures.
+Specifically, false positives are not needed to be explain the observations (the results in \cref{fig:paper:duration,ATACCC:fig:goodness-of-fits} do not allow false negatives but fit the data well).
+\todo{this sentence is a bit clumsy...}
 Therefore, the model for test failures can be simplified to simply have a constant probability of a false negative.
@@
 p(y_{i,t} \mid \cdot) &= \begin{cases}
-    (1 - \rho) f_N(y_{i,t} \mid \log \hat{v}_i(t), \sigma_v^2) & \text{if $y_{i,t} > 0$}\\
-    \rho + (1 - \rho) F_N(\vlod \mid \log \hat{v}_i(t), \sigma_v^2) & \text{if $y_{i,t} = 0$.}
+    (1 - \rho) f_N(y_{i,t} \mid \log v_i(t), \sigma_v^2) & \text{if $y_{i,t} > v_\text{lod}$}\\
+    \rho + (1 - \rho) F_N(\vlod \mid \log v_i(t), \sigma_v^2) & \text{if $y_{i,t} = v_\text{lod}$.}
 \end{cases}
@@
 The parameters $x_0$ and $\sigma_0$ are no longer used in the model.
-For all other parameters, I kept the priors the same as \textcite{hakkiOnset}, including the prior on $\rho$, and performed inference with the same settings.
+For all other parameters, I kept the priors the same as \textcite{hakkiOnset}, including the prior on $\rho$, and performed inference with the same settings (see end of \cref{ATACCC:sec:hakki}).
 
@@
 So far, only the viral load trajectories have been considered.
-The time that individual $i$ are first and last positive are the two values of $\tau$ such that $\log \hat{v}_i(\tau) = 0$.
+The time that individual $i$ are first and last positive are the two values of $\tau$ such that $\log v_i(\tau) = v_\text{lod}$.
 Denote these times as $\tau_{0,i}$ and $\tau_{1,i}$.
@@
 At $\tau_{0,i}$ and $\tau_{1,i}$, the approximation as exponential growth (described in \cref{ATACCC:sec:hakki}) is good.
-Therefore, set $\log \hat{v}(\tau)$ and one of the exponential terms to 0 in \cref{ATACCC:eq:viral-load}.
-This gives the following expressions:
+Therefore, set $\log v(\tau)$ and one of the exponential terms to 0 in \cref{ATACCC:eq:viral-load}.
+When $v_\text{lod} = 0$, as it does here, this gives the following expressions:
 $$
@@
 $$
-Note that $d_i$ does not depend on $\tau_{\max,i}$, only the random effects.
+Note that $d_i$ does not depend on $\tau_{\max,i}$, only $\phi_i$.
 
+\todo[inline]{insert a better explanation of $F_D$'s posterior}
 For later use, a discretised version is preferred (as described in \cref{E-inc-prev}).
@@
 
-The original \textcite{hakkiOnset} model has convergence issues (Rhat > 1.07 and ESS < 100) for the parameters of the test failure distribution.
-These issues do not occur in the modified model introduced in \cref{ATACCC:sec:observation-modification}, although both model have convergence issues (Rhat > 1.4 and ESS < 20) for some individual-level parameters; with the simplified model this is only individual 55.
-
-Individual 55 has an extremely noisy set of observations.
-This means that two viral load trajectories fit the data well (shown in \cref{ATACCC:fig:individual-55}).
-The first, found by two chains, increases very quickly, peaks 4.0 days after the first test, then declines slowly.
-The second, found by the other six chains, is more symmetrical, peaking 7.2 days after the first test.
-Because the individual trajectories are not of interest for this analysis, and the population-level parameters agree between the two sets of chains, I consider this lack of convergence acceptable.
-\begin{figure}
-  \centering \includegraphics{ATACCC/fit_individual_55}
-  \caption[Goodness-of-fit for individual 55]{Fit to individual 55's data coloured by MCMC chain.
-Two chains favour the quickly increasing viral load trajectory, peaking at day 4.0, the other six favour the slower trajectory, peaking at day 7.2.
-\label{ATACCC:fig:individual-55}}
-\end{figure}
+The original \textcite{hakkiOnset} model has convergence issues (Rhat > 1.07 and ESS < 100) for the parameters of the test failure distribution, $\sigma_0$ and $\rho$.
+The simplified model I proposed in \cref{ATACCC:sec:observation-modification} solved this issue, meaning that this analysis converged.
 
@@
 Almost all test results are captured in the 95\% credible intervals.
-Individual 55 has some fitting issues, as explored above.
+Individual 55 has some fitting issues, explored further below.
 Individual 57 appears to have far noisier measurements than the other individuals, suggesting that the observation noise might vary between individuals.
@@
 However, it would complicate the model, be unlikely to change the estimates of the parameters of interest, and may not be identifiable.
-
 \begin{figure}
@@
 \label{ATACCC:fig:goodness-of-fits}}
+\end{figure}
+
+While the population-level parameters converged in the modified model, both model had convergence issues (Rhat > 1.4 and ESS < 20) for some individual-level parameters.
+However, this was greatly improved with the simplified model, limited to a single individual, number 55.
+Individual 55 has an extremely noisy set of observations.
+This means that two viral load trajectories fit the data well (shown in \cref{ATACCC:fig:individual-55}).
+The first, found by two chains, increases very quickly, peaks 4.0 days after the first test, then declines slowly.
+The second, found by the other six chains, is more symmetrical, peaking 7.2 days after the first test.
+The individual trajectories are not of interest for this analysis, and the population-level parameters agree between the two sets of chains, therefore this should not affect any estiamtes of interest.
+\todo{clunky phrasing...}
+\begin{figure}
+  \centering \includegraphics{ATACCC/fit_individual_55}
+  \caption[Goodness-of-fit for individual 55]{Fit to individual 55's data coloured by MCMC chain.
+Two chains favour the quickly increasing viral load trajectory, peaking at day 4.0, the other six favour the slower trajectory, peaking at day 7.2.
+\label{ATACCC:fig:individual-55}}
 \end{figure}

### 2023-11-22 SEIR.tex
continue edits

@@
 
-\todo[inline]{add structured pop refs}
 To relax the assumption of a homogeneous population, the population can be stratified.
@@
 Therefore, the expected rate of infections in strata $a$ caused by this infected individual is $\kappa_{aa'} s_a(t) q_{aa'}$.
-Multiplying by the number of infected individuals in strata $a'$, $N_{a'} i_{a'}(t)$, gives $\kappa_{aa'} q_{aa'} s_a(t) N_{a'} i_{a'}(t)$.
-The constant terms here are absorbed into the WAIFW matrix, giving $\beta_{aa'} = \kappa_{aa'} q_{aa'} N_{a'}$~\autocite{section 9.2}{diekmannMathematical}.
+Multiplying by the number of infected individuals in strata $a'$, $N_{a'} i_{a'}(t)$, gives $\kappa_{aa'} q_{aa'} s_a(t) N_{a'} i_{a'}(t)$ where $i_{a'}(t) = \sum_{j=1}^n i_{j,a'}$ is the total proportion of stratum $a'$ that is infectious, regardless of which compartment they are in.
+The constant terms here are absorbed into the WAIFW matrix, giving $\beta_{aa'} = \kappa_{aa'} q_{aa'} N_{a'}$~\autocite[section 9.2]{diekmannMathematical}.
 The total force of infection on strata $a$ from all strata is, therefore, $\lambda_a(t) = \sum_{a' \in \set{A}} \beta_{aa'} i_{a'}(t)$.
@@
 
-In practical situations, estimating all of the $\beta_{aa'}$ parameters is infeasible, so some assumptions must be made to reduce the number of parameters~\autocite[176]{andersonInfectious}.
+In practical situations, estimating all the $\beta_{aa'}$ parameters is infeasible, so some assumptions must be made to reduce the number of parameters~\autocite[176]{andersonInfectious}.
 It is infeasible because the progression of the epidemic depends on $\vec{\lambda}(t)$, which has dimension $A$, while $\matr{\beta}$ has dimension $A^2$.
@@
 where $\bar{p}_{kat} = \int_{t}^{t+1} p_{ka}(s) ds$ is the mean occupancy in compartment $p_k$ and strata $a$ over day $t$ and $\rho$ is a nuisance parameter controlling the overdispersion of the beta-binomial distribution.
-In this parameterisation (see \cref{E-distributions}) at $\rho=0$ the beta-binomial coincides with the binomial distribution and as $\rho \to \infty$, the distribution tends towards the uniform distribution on $[0, 1]$.
+In this parameterisation (detailed in \cref{E-distributions}) at $\rho=0$ the beta-binomial coincides with the binomial distribution and as $\rho \to \infty$, the distribution tends towards the uniform distribution on $[0, 1]$~\autocite{hughesUsing}.
 
@@
 
-\todo[inline]{sort region names in plots}
+\todo[inline]{format region names in plots}
 
@@
     \caption[Posterior estimates of peak incidence timing]{%
-        Posterior estimates (95\% CrI) of incidence by region and age.
-        See main text for discussion.
+        Posterior probability of the peak incidence for each region and age combination occurring on each day.
     }
-    \label{SEIR:fig:incidence}
-    \todo[inline]{write peak incidence caption}
+    \label{SEIR:fig:peak-incidence}
 \end{figure}

### 2023-11-22 cis-imperfect-testing.tex
Fix chapter numbering
Stop using "truncation"

@@
 \begin{document}
+
+\ifSubfilesClassLoaded{
+  \setcounter{chapter}{5}
+}
 
@@
 A single positive episode is one in which the episode contains exactly one positive test, intuitively these cause short estimates because there is no lower bound on the length of time the episode lasts.
-Furthermore, due to the truncation, there is little or no ability to upper bound the number of very short episodes because these would almost all be missed.
+Furthermore, due to missing short episodes, there is little or no ability to upper bound the number of very short episodes because these would almost all be missed.
 \Textcite{shenNonparametrica} (building on \textcite{panNote}) showed, in a related setting, that the maximum likelihood (frequentist) estimator is inconsistent and can lead to severe underestimation (bias) of the survival function.
@@
   The duration of the episode is less than $t_{ib_i}^N$. These first
-  two mechanisms are how truncation previously occurred, with
+  two mechanisms are why episodes were missed in the previous chapter, with combined
   probability $p_{it}$.

### 2023-11-22 biology-data.tex
Intermittent negatives discussion to new chapter 2

--- 2023-07-12 thesis.tex
@@
-\documentclass{cam-thesis}
-\usepackage{setspace}
-\onehalfspacing
-\usepackage{graphicx}
-\usepackage[utf8]{inputenc}
+\documentclass[thesis.tex]{subfiles}
 
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{bm}
-\usepackage[nameinlink]{cleveref}
-\usepackage[style=british]{csquotes}
-\usepackage[T1]{fontenc}
-\usepackage{hyperref}
-\usepackage{textcomp} % provide symbols
-\usepackage{todonotes}
-\usepackage{xr}
-% \usepackage{microtype}
-
-% Macros for common abbreviations to get the spacing right
-% See: https://stackoverflow.com/questions/3282319/correct-way-to-define-macros-etc-ie-in-latex
-\usepackage{xspace}
-\makeatletter
-\DeclareRobustCommand\onedot{\futurelet\@let@token\@onedot}
-\def\@onedot{\ifx\@let@token.\else.\null\fi\xspace}
-\def\eg{e.g\onedot} \def\Eg{\emph{E.g}\onedot}
-\def\ie{i.e\onedot} \def\Ie{\emph{I.e}\onedot}
-\def\cf{c.f\onedot} \def\Cf{\emph{C.f}\onedot}
-\def\etc{etc\onedot} \def\vs{\emph{vs}\onedot}
-\def\wrt{w.r.t\onedot} \def\dof{d.o.f\onedot}
-\def\etal{et al\onedot}
-\makeatother
-
-% Generic maths commands
-\def\reals{\mathbb{R}}
-\def\nats{\mathbb{N}}
-\def\sampSpace{\mathcal{X}}
-\def\dist{\sim}
-\DeclareMathOperator{\E}{\mathbb{E}}
-\DeclareMathOperator{\V}{\mathbb{V}}
-\DeclareMathOperator{\prob}{\mathbb{P}}
-\DeclareMathOperator{\var}{\mathbb{V}}
-\DeclareMathOperator{\indicator}{\mathbb{I}}
-\DeclareMathOperator{\cov}{Cov}
-\DeclareMathOperator{\logit}{logit}
-\DeclareMathOperator{\Ber}{Bernoulli}
-\DeclareMathOperator{\Bin}{Binomial}
-\DeclareMathOperator{\Poi}{Poisson}
-\DeclareMathOperator{\Beta}{Beta}
-\DeclareMathOperator{\NegBin}{NegBin}
-\DeclareMathOperator{\GamDist}{Gamma}
-\def\matr{\bm}
-\def\set{\mathcal}
-\renewcommand{\vec}{\bm}
-
-% Thesis-specific maths commands
-\newcommand{\psens}{p_\text{sens}}
-
-\newcommand\citePersonalComms[1]{(#1, personal communication)}
-
-\title{Estimating the incidence of SARS-CoV-2 from prevalence data}
+\title{Estimating duration in the presence of misclassification}
 \author{Joshua Blake}
 \date{\today}
-\subjectline{MRC Biostatistics Unit}
-\college{Jesus College}
-\collegeshield{CollegeShields/jesus}
-% \keywords{one two three}
-
-%% Submission date [optional]:
-% \submissiondate{November, 2042}
-
-%% You can redefine the submission notice [optional]:
-% \submissionnotice{A badass thesis submitted on time for the Degree of PhD}
-
-\usepackage[backend=biber,style=authoryear,uniquename=false]{biblatex}
-
-\usepackage{subfiles} % Best loaded last in the preamble
-\addbibresource{references.bib}
-\externaldocument[E-]{\subfix{thesis}}
-
-\abstract{%
-    TODO
-}
-
-\acknowledgements{%
-  TODO
-}
-
-\usepackage{geometry}
-\setlength{\marginparwidth}{3cm}
-\geometry{margin=3.8cm}
 
@@
 
-\frontmatter{}
+\ifSubfilesClassLoaded{
+  \setcounter{chapter}{1}
+}
 
-\listoffigures
-
-\chapter{Introduction} \label{intro}
-
-\section{The COVID-19 pandemic and ongoing burden}
-
-\section{SARS-CoV-2 and COVID-19}
-
-\section{Thesis outline}
-
-\chapter{Incidence, prevalence, and duration}
-We wish to estimate the distribution of time for which an individual is positive for.
-An individual could have multiple infections, and hence I refer to \emph{infection episodes}.
-
-I define an individual as \emph{detectable} if they would test positive if tested, in the absence of misclassification error (which is dealt with in \cref{E-imperf-test}).
-Episode $i$'s \emph{duration of positivity}, $D_i$ is the number of days that they are detectable for; I use time discretised to days, therefore, the duration is at least 1 day.
-Note that the relationship between positivity (\ie the result of a PCR test) and infectiousness (\ie an individual's ability to pass on the virus) is complex~\autocites{lascolaViral}{singanayagamDuration}, and beyond the scope of this thesis.
-I define an episode as starting at time $B_i$ and ends at $E_i$, hence $D_i = E_i - B_i + 1$.
-
-An episode is \emph{detected} when they test positive at least once, if this occurs; episode $i$ is detected at time $t$ if and only if: $i$ is detectable at $t$, $i$ is tested at $t$, and the test returns positive (a false negative does not occur).
-
-The survival time function, parameterised by $\theta$, is the probability that an episode lasts at least $t$ days: $S_\theta(t) = \prob(D_i \geq t \mid B_i = b, \theta)$, where $D_i$ and $B_i$ are the duration and starting time of episode $i$ respectively.
-I make the standard assumption that $D_i$ and $B_i$ are independent and hence: $S_\theta(t) = \prob(D_i \geq t \mid \theta)$.I utilise survival analysis for analysing this data.
-Survival analysis is the area of statistics concerned with estimating the distribution of the times between events.
-It is broadly applicable to many domains, such as the time-to-failure for a mechanical system or the effect of a treatment on time spent in hospital.
-In the context of this thesis, the distribution to estimate is that of duration of positivity.
-The \emph{initiating event} is when the duration starts, \ie becoming detectable.
-The \emph{terminating event} is when the duration ends, \ie becoming no longer detectable.
-
-\chapter{Studies used in this thesis} \label{intro:sec:studies}
+\chapter{SARS-CoV-2: biology and data} \label{intro:sec:studies}
 
@@
 The CIS (Coronavirus Infection Survey) is a longitudinal study on a representative sample of households.
-A full description of the study can be found in the supplementary materials of \textcite{pouwelsCommunity} or the study protocol~\autocite{cisProtocol}.
+A full description of the study can be found in \textcite[][supplementary materials]{pouwelsCommunity} or the study protocol~\autocite{cisProtocol}.
 In brief, households are invited to the study from databases held by the ONS.
@@
 
-\chapter{Duration from biomarkers} \label{ATACCC}
+\emph{Intermittent negatives} are the clearest example of a false negative.
+An intermittent negative is when an individual tests negative but tested positive previously and subsequently.
+Intermittent negatives are stripped out when creating the dataset used for all duration analyses in this chapter (as described in \cref{E-episode-def}), but demonstrate that false negatives do occur.
 
-\section{Background}
+Whether a future positive is part of the same episode of a reinfection is not always trivial; I rely on a process developed previously by Sarah Walker, see \cref{E-episode-def}.
+However, up until the emergence of the Omicron variant, reinfections are rare, especially in a short time frame, and hence the tricky cases are rare until this time (late 2021).
 
-\section{Original analysis}
-
-\section{Comparison to \texorpdfstring{\textcite{hakkiOnset}}{Hakki \etal (2022)}}
-
-\section{Final analysis}
-
-\section{Producing duration from parameter estimates}
-
-The ATACCC analysis produces a posterior distribution for some population-level parameters, $\phi$.
-Taking a posterior sample for this, $\phi^{(i)}$, we can then draw the parameters which determine the viral
-load trajectory for a random individual $j$, $\theta_j$, from which that individual's duration $d_j^{(i)}$ can be calculated.
-By drawing $\theta_1, \theta_2, \dots, \theta_n$ from the distribution $p(\theta_j \mid\phi^{(i)})$, the density of the duration distribution, $f_A(t)$, can be
-calculated for the posterior sample $i$ as:
-$$
-f^{(i)}_A(t) = \frac{1}{n}\sum_{j=1}^n \mathbb{I} \left( t-0.5 < d_j^{(i)} < t+0.5 \right)
-$$
-Given $f_A(t)$, the cumulative density and hazard functions can be calculated.
-Repeating this for $N$ posterior samples of $\phi$ gives the posterior distribution.
-I have used $N = 1000$ and $n = 100,000$, 
-Since implementing this solution, I have found that Monte Carlo integration to calculate durations given $\theta_i$ is more accurate by avoiding discretisation errors.
-However, the results here do not use this for consistency.
-
-\subfile{cis-perfect-testing}
-
-\subfile{cis-imperfect-testing}
-
-\chapter{Statistical backcalculation}
-
-\chapter{Mechanistic modelling}
-
-\chapter{Conclusion}
-
+\ifSubfilesClassLoaded{
   \listoftodos
-
-\printbibliography
-
-\appendix
-
-\subfile{appendix-cis-perfect-testing}
-
-\chapter{CIS episode definitions} \label{episode-def}
-
-\todo[inline]{Get this from SARAH}
+}{}
 

### 2023-11-22 cis-imperfect-testing.tex
Intermittent negatives discussion to new chapter 2

@@
 
-\todo[inline]{These next two paragraphs are kinda disjointed but not sure how to bring them in}
-
-\emph{Intermittent negatives} are the clearest example of a false negative.
-An intermittent negative is when an individual tests negative but tested positive previously and subsequently.
-Intermittent negatives are stripped out when creating the dataset used for all duration analyses in this chapter (as described in \cref{E-episode-def}), but demonstrate that false negatives do occur.
-
-Whether a future positive is part of the same episode of a reinfection is not always trivial; I rely on a process developed previously by Sarah Walker, see \cref{E-episode-def}.
-However, up until the emergence of the Omicron variant, reinfections are rare, especially in a short time frame, and hence the tricky cases are rare until this time (late 2021).
+\todo[inline]{maybe mention intermittent negatives, referring to its introduction in chapter 2}
 

### 2023-11-22 thesis.tex
Bulk of ATACCC chapter (no results yet)
First full draft of ATACCC chapter
Arrange chapters correctly
Various edits
restructure SEIR chapter
Intermittent negatives discussion to new chapter 2

@@
 \usepackage{afterpage}
+\usepackage{caption}
 % \usepackage{microtype}
+\usepackage[UKenglish]{babel}
 
@@
 \DeclareMathOperator{\V}{\mathbb{V}}
+\DeclareMathOperator{\I}{\mathbb{I}}
 \DeclareMathOperator{\prob}{\mathbb{P}}
@@
 \DeclareMathOperator{\Exponential}{Exponential}
-\DeclareMathOperator{\NB}{NegBin}
+\DeclareMathOperator{\NBr}{NegBin}
+\newcommand{\NBc}{\NBr_{c}}
+\newcommand{\NBs}{\NBr_{s}}
 \DeclareMathOperator{\BB}{BetaBin}
@@
 \DeclareMathOperator{\LN}{LN}
+\DeclareMathOperator{\LKJ}{LKJ}
 \DeclareMathOperator{\expit}{expit}
@@
 \newcommand{\R}{\mathcal{R}_{0}}
+\newcommand{\vlod}{v_\text{lod}}
 
@@
 
+\subfile{biology-data}
+
 \subfile{incidence-prevalence}
@@
 
-\chapter{Studies used in this thesis} \label{intro:sec:studies}
-
-\section{PCR testing} \label{intro:sec:PCR}
-
-\section{Assessment of Transmission and Contagiousness of COVID-19 in Contacts}
-
-\section{Coronavirus (COVID-19) Infection Survey} \label{intro:sec:cis}
-
-The CIS (Coronavirus (COVID-19) Infection Survey) continually enrols individuals and tests them on an ongoing basis, following a
-specified testing protocol. The protocol specifies that the initial five
-tests are spaced every 7 days, then the spacing is every 28 days.
-However, real-world considerations often lead to variations in test
-times and missed tests. Our analysis focuses solely on the binary result
-(positive or negative) of these tests, assuming no misclassification
-bias.
-
-The CIS (Coronavirus Infection Survey) is a longitudinal study on a representative sample of households.
-A full description of the study can be found in the supplementary materials of \textcite{pouwelsCommunity} or the study protocol~\autocite{cisProtocol}.
-In brief, households are invited to the study from databases held by the ONS.
-The number invited was to satisfy various target number of individuals swabbing per fortnight; relevant to the discussion in this document is that the target increased from 2,500 by the end of May 2020 to 150,000 in October 2020.
-Within a household, all individuals aged 2 and over were invited to participate.
-Once invited, an enrolment swab would be taken followed by 4 further weekly swabs (giving a total of 5 swabs on days 0, 7, 14, 21, 28 relative to enrolment) after which monthly swabs are taken.
-
-During the period we consider, the total cohort size expands due to the continuous recruitment into the study.
-Following this, the number then decreases and stabilises as those recruited to meet the October target transition from weekly onto monthly testing.
-See figure \@ref(fig:test-schedules) for further detail.
-
-\todo[inline]{Add some descriptive figures e.g. those in \url{~/COVID/ons-incidence/duration_estimation/reproducible-sims/reports/simulation-studies.pdf}}
-
-Consider the CIS data.
-It consists of a series of tests on the same individuals (longitudinal data), where some individuals have a series of positive tests.
-
 \subfile{ATACCC}
@@
 
-\chapter{Statistical backcalculation}
+\chapter{Estimating transmission using backcalculation} \label{backcalc}
 
@@
 
+\subfile{ATACCC-appendix-original-analysis}
+
 \chapter{CIS episode definitions} \label{episode-def}

### 2023-11-23 SEIR/CIS/params.R
Improve plots + typos

@@ -35,41 +35,47 @@ posterior = readr::read_csv(here::here("SEIR/CIS/params.csv")) |>
         )
     )
 
-# posterior |>
-#     filter(.iteration >= 500e3) |>
-#     group_nest(region, parameter, i) |>
-#     mutate(
-#         Rhat = map_dbl(
-#             data,
-#             ~.x |>
-#                 select(!.draw) |>
-#                 pivot_wider(
-#                     names_from = .chain,
-#                 ) |>
-#                 select(!.iteration) |>
-#                 as.matrix() |>
-#                 rstan::Rhat()
-#         )
-#     ) |>
-#     arrange(desc(Rhat))
+Rhat = posterior |>
+    filter(.iteration >= 500e3) |>
+    group_nest(region, parameter, i) |>
+    mutate(
+        Rhat = map_dbl(
+            data,
+            ~.x |>
+                select(!.draw) |>
+                pivot_wider(
+                    names_from = .chain,
+                ) |>
+                select(!.iteration) |>
+                as.matrix() |>
+                rstan::Rhat()
+        )
+    ) |>
+    arrange(desc(Rhat))
+readr::write_csv(Rhat, here::here("SEIR/CIS/rhat.csv"))
 
-# posterior |>
-#     filter(.iteration >= 500e3) |>
-#     group_nest(region, parameter, i) |>
-#     mutate(
-#         ESS = map_dbl(
-#             data,
-#             ~.x |>
-#                 select(!.draw) |>
-#                 pivot_wider(
-#                     names_from = .chain,
-#                 ) |>
-#                 select(!.iteration) |>
-#                 as.matrix() |>
-#                 rstan::ess_bulk()
-#         )
-#     ) |>
-#     arrange(ESS)
+Rhat |>
+    group_by(region) |>
+    summarise(n_bad = sum(Rhat > 1.1))
+
+posterior |>
+    filter(.iteration >= 500e3) |>
+    group_nest(region, parameter, i) |>
+    mutate(
+        ESS = map_dbl(
+            data,
+            ~.x |>
+                select(!.draw) |>
+                pivot_wider(
+                    names_from = .chain,
+                ) |>
+                select(!.iteration) |>
+                as.matrix() |>
+                rstan::ess_bulk()
+        )
+    ) |>
+    arrange(ESS) |>
+    readr::write_csv(here::here("SEIR/CIS/ess.csv"))
 
[…]

### 2023-11-23 SEIR/sim/coverage.R

@@ -24,13 +24,19 @@ source(here::here("utils.R"))
 posterior_samples = readr::read_csv(
     here::here("SEIR/sim/posteriors_combined.csv")
 )
+posterior_samples |>
+    filter(parameter == "psir", iteration > 300e3) |>
+    ggplot(aes(iteration, value)) +
+    geom_line() +
+    facet_wrap(~ sim, scales = "free_y")
 posterior_intervals = posterior_samples |>
-    filter(iteration > 30e3) |>
+    filter(iteration > 300e3) |>
     group_by(sim, parameter) |>
     median_qi(value, .width = c(0.5, 0.75, 0.95))
 true_params =  readr::read_csv(
     here::here("SEIR/sim/true_vals.csv")
-)
+) |>
+    rename(parameter = param)
 coverage = posterior_intervals |>
     left_join(true_params, by = c("sim", "parameter" = "param")) |>
     mutate(
@@ -64,7 +70,8 @@ p_coverage = ggplot(coverage, aes(colour = factor(.width))) +
         x = "Coverage",
         y = "Parameter",
         colour = "CrI width"
-    )
+    ) +
+    theme(legend.position = "bottom")
 ggsave(
     filename = here::here("SEIR/sim/coverage.png"),
     plot = p_coverage,
@@ -72,22 +79,38 @@ ggsave(
     height = 10,
     unit = "cm"
 )
-posterior_intervals |>
-    # grab any width: only care about point estimates here
-    filter(.width == 0.95, !str_detect(parameter, "beta")) |>
-    # calculate bias by joining with true values
-    left_join(true_params, by = c("sim", "parameter" = "param")) |>
-    mutate(
-        bias = value - true,
-        rel_bias = bias / true,
-    )
-posterior_intervals |>
+
+p_true_vs_posterior = posterior_intervals |>
     # grab any width: only care about point estimates here
     filter(.width == 0.95, !str_detect(parameter, "beta"), sim != 59) |>
     # calculate bias by joining with true values
-    left_join(true_params, by = c("sim", "parameter" = "param")) |>
-    ggplot(aes(value, true)) +
+    left_join(true_params, by = c("sim", "parameter")) |>
+    ggplot(aes(true, value)) +
     facet_wrap(~ parameter, scales = "free") +
+    geom_abline(slope = 1, intercept = 0) +
[…]

### 2023-11-24 ATACCC.tex
Improve plots + typos
add todo

@@
 
-Let the observed log viral load for individual $i$ at time $t$ be $y_{i,t}$, with $y_{i,t}=0$ if the test is negative.
+Let the observed log viral load for individual $i$ at time $t$ be $y_{i,t}$, with negative tests signified by $y_{i,t}=v_\text{lod}$.
 Then, its likelihood is:
@@
 
+\todo[inline]{Probably important to show estimated test sensitivity here to refer to later}
+
 The original \textcite{hakkiOnset} model has convergence issues (Rhat > 1.07 and ESS < 100) for the parameters of the test failure distribution, $\sigma_0$ and $\rho$.

### 2023-11-24 SEIR.tex
Restructure implmentation section
[TMP] Update figures

@@
 
-% \section{A transmission model for SARS-Cov-2} \label{SEIR:sec:transmission-specific}
-
-% I follow \textcite{birrellRealtimea} in parameterising $\matr{\beta}$, which includes introducing a time-dependency.
-
 \section{Observation model} \label{SEIR:sec:observation}
@@
 
-\section{Inference and implementation} \label{SEIR:sec:inference-implementation}
+\section{Implementation} \label{SEIR:sec:inference-implementation}
+
+\todo[inline]{Redo intro paragraph to "implementation" section}
 
@@
 
-\subsection{MCMC} \label{SEIR:sec:MCMC}
+\subsection{WAIFW matrix}
 
-I use an adaptive random-walk Metropolis-Hastings algorithm (as introduced in \cref{E-inc-prev:sec:MCMC}).
-The proposal distribution is a multivariate normal that adapts to the covariance of the posterior.
-The adaption algorithm is based on \textcite[algorithm 4]{andrieuTutorial}, as implemented in \textcite{ghoshApproximate}.
-Proposals are accepted or rejected using the standard Metropolis-Hastings (MH) acceptance probability.
-See \cref{SEIR:MCMC-algorithm} for full details.
-\begin{algorithm}
- set $\vec{X_0}$ to an initial value of the parameter vector \;
- $\vec{\mu_0} = \vec{X_0}$ \;
- $\matr{\Sigma_0} = \text{diag}(\vec{\mu_0})$ \;
- $\lambda_0 = 1$ \;
- \For{$i = 1, \dots, M$}{
-  sample $\vec{Y_}i \sim \N(\vec{\mu_{i-1}}, \lambda_{i-1}\matr{\Sigma_{i-1}})$\;
-  set $\vec{X_i}$ to $\vec{Y_i}$ or $\vec{X_{i-1}}$ using a MH acceptance step\;
-  update the proportion of proposals accepted so far, $\alpha$ \;
-  \eIf(\tcc{No adaptation for 200 iterations}){$i \leq 200$}{
-    $\lambda_i = \lambda_{i-1}$ \;
-    $\vec{\mu_i} = \vec{\mu_{i-1}}$ \;
-    $\matr{\Sigma_i} = \matr{\Sigma_{i-1}}$ \;
-   }{
-    $\gamma_i = (i - 200)^{-0.6}$ \;
-    $\log \lambda_i = \log \lambda_{i-1} + \gamma_i(\alpha - 0.234)$ \;
-    $\vec{\mu_i} = (1 - \gamma_i) \vec{\mu_{i-1}} + \gamma_i \vec{X_i}$ \;
-    $\matr{\Sigma_i} = (1 - \gamma_i) \matr{\Sigma_{i-1}} + \gamma_i (\vec{X_i} - \vec{\mu_i})(\vec{X_i} - \vec{\mu_i})^T$ \;
-  }
- }
- \caption{Algorithm for adaptive random-walk Metropolis-Hastings.
-$\vec{\mu_i}$ and $\matr{\Sigma_i}$ are an estimate of the mean and covariance of the posterior distribution using information up to iteration $i$.
-$\text{diag}(\vec{\mu_0})$ is the diagonal matrix with diagonal entries equal to $\vec{\mu_0}$.
-$\lambda_i$ is the scale parameter of the proposal distribution at iteration $i$, tuned to try and ensure an optimal proportion of proposals are accepted (23.4\%).
-$\gamma_i$ is the learning rate, which determines how much adaptation occurs.
-$\gamma_i \to 0$ as $i \to \infty$ so the rate of adaptation is \emph{vanishing}.
-Vanishing adaptation is required for the algorithm to converge to the target distribution~\autocite[section 3]{andrieuTutorial}.}
- \label{SEIR:MCMC-algorithm}
-\end{algorithm}
-
-I ran 4 chains for 1,000,000 iterations.
-The first 500,000 were discarded as burn-in, with the remaining chains thinned to every 100th iteration.
-This gives 20,000 posterior samples per region.
-This was further thinned to 2,000 samples when calculating posterior predictive quantities (incidence and prevalence) due to their computational cost.
-
-\subsection{Parameterisation} \label{SEIR:sec:parameterisation}
-
-\subsubsection{WAIFW matrix}
-
-A parsimonious model is needed to parameterise the WAIFW matrix elements with (see \cref{SEIR:sec:structured-populations}).
-The strata will represent age groups.
-
-I use the following age-based strata, which align closely with the strata previously used for analysis of the CIS data~\autocites[e.g.:][]{houseInferring}{walkerTracking}.
+A parsimonious model is needed to parameterise the WAIFW matrix (see \cref{SEIR:sec:structured-populations}).
+I use the following age-based strata, which align closely with the age groupings previously used for analysis of the CIS data~\autocites[e.g.:][]{houseInferring}{walkerTracking}.
+\todo{Contrast with age groupings used in previous chapter}
 \begin{enumerate}
@@
 \end{enumerate}
-The CIS only recruited individuals aged 2 or older, however, I assume tests are representative of the whole $[0, 11)$ age group.
 
-I follow the WAIFW matrix parameterisation of \textcite{birrellRealtime}, with the decomposition $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'} N_{a'}$ to separate out contact patterns, $\kappa_{aa'}$, which can vary over time; transmission probabilities $q_{aa'}$, assumed constant; a global effect for time $\beta(t)$; and the population of the stratum, $N_{a'}$.
+I follow the WAIFW matrix parameterisation of \textcite{birrellRealtime}, with the decomposition $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'} N_{a'}$.
+This separates contact patterns, $\kappa_{aa'}$, which can vary over time; transmission probabilities $q_{aa'}$, assumed constant; a region-wide effect for time $\beta(t)$; and the population of the stratum, $N_{a'}$.
 
-This parameterisation takes values for the contact rates, $\kappa_{aa'}(t)$, from other data sources.
-These values are produced based on several data sources~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}.
+This parameterisation takes values for the contact rates, $\kappa_{aa'}(t)$, from several other data sources, using pre-existing methodology~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}.
 %; these estimates have been used for previous modelling of the COVID-19 pandemic~\autocites{vanleeuwenTime}{birrellRealtime}.
 % These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
-First, take a baseline from the POLYMOD study's~\autocite{mossongSocial} baseline (pre-pandemic) estimates of contacts in various settings and the UK Time Use Survey (UKTUS)'s~\autocite{UKTUS} estimates of time use.
+First, take pre-pandemic baseline estimates of contat rates, from the POLYMOD study~\autocite{mossongSocial}, and time use, from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
 POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on the contact's age and location (home, work, school, leisure, transport, or other).
@@
 Therefore, I set $q^\text{inf}_{a'} = 1$ for all $a'$, $q^\text{susc}_{a} = q_c$ for the youngest stratum, and $q^\text{susc}_{a} = 1$ for the remaining strata.
-The interpretation of $q_c$ is that children have a susceptibility, relative to adults, of $q_c$.
 
-\subsubsection{Identifiability}
+\subsection{Data}
 
-Following \textcite{birrellBayesian}, I reparameterise the model to improve identifiability.
-Rather than directly inferring $\beta_1$, I infer the initial growth rate of the epidemic.
-In addition, I use a parsimonious parameterisation of the initial conditions using the initial proportion of the population that are infected and the initial proportion that are susceptible only.
-Both of these rely on an assumption that the epidemic is in a steady-state exponential growth phase at time 0 (see \cref{SEIR:sec:SIR}).
+I use the same CIS data as in \cref{E-backcalc}, again considering each region independently.
+This model is stratified only by region and age group, so the data are aggregated to these levels.
+Only individuals aged two or older are eligible to participate in the CIS; I assume that test results in all participants aged under 11 are still representative of the whole $[0, 11)$ age group.
 
-During such an exponential growth phase, all components of the $\vec{e}$, $\vec{i}$ and $\vec{p}$ compartments grow at the same growth rate, $\psi$.
-This follows directly from the ODEs where $\vec{s}$ is constant. 
-The growth rate $\psi$ is easily identifiable from the data, and therefore a parameterisation in terms of $\phi$ should be more computationally efficient.
-If this is the initial exponential phase, when $\vec{s} \approx 1$, then the basic reproduction number for a SEIR model is linked to this growth rate by~\autocites{birrellBayesian}{wearingAppropriate}:
+As well as the motivation given previously, independent regions allows heterogeneity in the parameters between regions; this heterogeneity could arise from a range of socioeconomic factors.
+Furthermore, the assumption of homogeneous mixing within age groups is more plausible for regions than the whole of England.
+By modelling each region independently, I assume that the epidemics in each region do not interact.
+Clearly, this is not true as travel between regions of England is common and unrestricted.
+However, for a large epidemic, the non-interaction is a good approximation~\autocite{birrellRealtimea}.
+
+\subsection{Full model} \label{SEIR:sec:full-model}
+
+\begin{figure}
+\makebox[\textwidth][c]{
+\begin{tikzpicture}[
+    node distance = 2.5cm,
+    on grid,
+    auto,
+    ->,>=stealth',
+    every state/.style={draw,rectangle},
+    ]
+
+    \node[state] (S) {$s$};
+    \node[state, right=of S] (E1) {$e_1$};
+    \node[state, right=of E1] (E2) {$e_2$};
+    \node[state, right=of E2] (I1) {$i_1$};
+    \node[state, right=of I1] (I2) {$i_2$};
+    \node[state, right=of I2] (R) {$r$};
+
+    \path (S) edge node {$\lambda$} (E1)
+          (E1) edge node {$\sigma$} (E2)
+          (E2) edge node {$\sigma$} (I1)
+          (I1) edge node {$\gamma$} (I2)
+          (I2) edge node {$\gamma$} (R);
+
+    \node[state, below=3cm of E1] (P0) {$p_0$};
+    \node[state, right=of P0] (P1) {$p_1$};
+    \node[state, right=of P1] (P2) {$p_2$};
+    \node[state, right=of P2] (P3) {$p_3$};
+    \node[state, right=of P3, draw=none] (P4) {};
+
+    \path  (S) edge node {$\lambda$} (P0)
+          (P0) edge node {$\kappa_0$} (P1)
+          (P1) edge node {$\kappa_Y$} (P2)
+          (P2) edge node {$p_x \kappa_{X1}$} (P3)
+               edge [out=-45,in=-135] node[below] {$(1 - p_x) \kappa_{X1}$} (P4)
+          (P3) edge node {$\kappa_{X2}$} (P4);
+
+    \node[draw, thick, inner sep=0.3cm, fit=(P1) (P3), fill=blue!10,opacity=0.2] {};
+\end{tikzpicture}
+}
+  \caption[SEIR model]{SEIR model used for the analysis in this chapter. The shaded box represents PCR-positive individuals. Shown for a single strata.}
+  \label{SEIR:fig:full-model}
+\end{figure}
+
+The full model combines the elements discussed in this chapter up until now.
+The transmission model follows \textcite{birrellRealtime}.
+It is a SEIR model with two latent and two infectious states.
+The population is stratified by age, using a time-varying matrix $\matr{\beta}(t)$ to describe the transmission between strata and how these change over time.
+The population in the model is that of one region of England.
+To form whole of England estimates, each region is modelled independently with the results, as in \cref{E-backcalc}.
+
+The observation model is novel, relating the transmission to the CIS data.
+It uses the phase-type model to represent the duration distribution of PCR positivity explained in \cref{SEIR:sec:observation}.
+
+A diagram of the model for a single stratum is shown in figure \cref{SEIR:fig:full-model}.
+The system of differential equations for each region is below.
 \begin{align}
-    \R = \frac{\psi \left( \frac{\psi}{2\sigma} + 1 \right)^2}{\gamma \left(1 - \frac{1}{\left(\frac{\psi}{2 \gamma} + 1 \right)^2} \right)} \label{SEIR:eq:rtoR}.
+    \label{SEIR:eq:fullODEs}
+    \vec{\lambda}(t) &= \beta(t) (\matr{\kappa}(t) \circ \matr{q}) (\vec{i_1}(t) + \vec{i_2}(t)) \\
+    \frac{d\vec{s}(t)}{dt} &= -\vec{\lambda}(t) \circ \vec{s}(t) \\
+    \frac{d\vec{e_1}(t)}{dt} &= \vec{\lambda}(t) \circ \vec{s}(t) - 2\sigma \vec{e_1}(t) \\
+    \frac{d\vec{e_2}(t)}{dt} &= 2\sigma \vec{e_1}(t) - 2\sigma \vec{e_2}(t) \\
+    \frac{d\vec{i_1}(t)}{dt} &= 2\sigma \vec{e_2}(t) - 2\gamma \vec{i_1}(t) \\
+    \frac{d\vec{i_2}(t)}{dt} &= 2\gamma \vec{i_1}(t) - 2\gamma \vec{i_2}(t) \\
+    \frac{d\vec{r}(t)}{dt} &= 2\gamma \vec{i_2}(t) \\
+    \frac{d\vec{p_0}(t)}{dt} &= \vec{\lambda} \circ \vec{s}(t) - \kappa_0 \vec{p_0}(t) \\
+    \frac{d\vec{p_1}(t)}{dt} &= \kappa_0 \vec{p_0}(t) - \kappa_Y \vec{p_1}(t) \\
+    \frac{d\vec{p_2}(t)}{dt} &= \kappa_Y \vec{p_1}(t) - \kappa_{X1} \vec{p_2}(t) \\
+    \frac{d\vec{p_3}(t)}{dt} &= p_x \kappa_{X1} \vec{p_2}(t) - \kappa_{X2} \vec{p_3}(t).
 \end{align}
 
-Furthermore, $\R$ is a function of $\matr{\beta}$, and hence $\beta_1$ (as explained in \cref{SEIR:sec:structured-populations}).
-Combining these expressions, the model can be parameterised by using $\psi$ in place of $\beta_1$.
-The following procedure gives $\beta_1$ in terms of $\psi$.
-First, calculate $\R$ from $\psi$ using \cref{SEIR:eq:rtoR}; denote this $\hat{R}$.
-Then calculate $R^* = D(\matr{\beta}) / \gamma$ when $\beta_1=1$ (recall that $D(\matr{\beta})$ denotes the dominant eigenvalue of $\matr{\beta}$).
-Multiplying a matrix by a scalar multiplies its dominant eigenvalue by that same scalar, therefore, setting $\beta_1 = \hat{R} / R^*$ gives the desired means that $\R$ at time 0 is $\hat{R}$, as desired.
+\subsection{Solving the systems of ODEs} \label{SEIR:sec:solving-ODEs}
+I solve the system of ODEs using Euler's method, discretised at timesteps of $\delta t = 1/2$ days.
+That is, if $\vec{x}_t$ is the system state at time $t$ for $t \in \{ 0, 0.5, 1, \dots \}$ then $x_{t+\frac{1}{2}} = x_t + (d\vec{x}(t)/dt) \delta t$
 
-To form a parsimonious parameterisation of the initial state, first focus on the initial proportion of the population that is in the infected compartments.
-Recall that in the exponential growth phase, the relative proportion of each strata that is infected is proportional to the dominant eigenvector of $\matr{\beta}$, which I denote $\vec{d}$.
-Introduce a model parameter $i^+$, the proportion of the population in the $i$ states at time 0 in the strata with the highest proportion infected.
-Then, $\vec{i_0} = \vec{i_1} + \vec{i_2} = \frac{i^+}{\max \vec{d}} \vec{d}$.
-Now introduce an additional vector of parameters, $\vec{\pi}$, where $\pi_a$ is the proportion of strata $i$ without immunity at time 0, hence, $\vec{r} = 1 - \vec{\pi}$.
-
-From the system of ODEs\todo{might need to reorder this because I've not presented these yet..
-} and the knowledge that all the compartments are growing exponentially at rate $\psi$, one can derive the following for the initial state of the compartments:
-\begin{align}
-    \vec{i_1} &= \vec{i_0} \left(1 + \frac{2\gamma}{\psi + 2\gamma} \right)^{-1} \\
-    \vec{i_2} &= \vec{i_0} - \vec{i_1} \\
-    \vec{e_2} &= \vec{i_1} \frac{\psi + 2\gamma}{2\sigma} \\
-    \vec{e_1} &= \vec{e_2} \frac{\psi + 2\sigma}{2\sigma} \\
-    \vec{s} &= \vec{\pi} - \vec{i_0} - \vec{e_1} - \vec{e_2} \\
-    \vec{p_0} &= \vec{e_1} \frac{\psi + 2\sigma}{\psi + \kappa_0} \\
-    \vec{p_1} &= \vec{p_0} \frac{\kappa_0}{\psi + \kappa_Y} \\
-    \vec{p_2} &= \vec{p_1} \frac{\kappa_Y}{\psi + \kappa_{X1}} \\
-    \vec{p_3} &= \vec{p_2} \frac{p_x \kappa_{X1}}{\psi + \kappa_{X2}}.
-\end{align}
+The mean prevalence on day $t$, used for the likelihood (see \cref{SEIR:eq:likelihood}), can be calculated as $\bar{p}_{kat} = \frac{1}{2} ( p_{kat} + p_{ka,t+1/2} )$.
 
 \subsection{Priors} \label{SEIR:sec:priors}
+
 Priors for the parameters are given in \cref{SEIR:table:priors}.
@@
 
-\subsection{Full model} \label{SEIR:sec:full-model}
+\subsection{Inference} \label{SEIR:sec:MCMC}
 
-\todo[inline]{work out where to present full model equations and then edit/rearrange this section to reflect this}
+I use an adaptive random-walk Metropolis-Hastings algorithm (previously discussed in \cref{E-inc-prev:sec:MCMC}).
+The proposal distribution is a multivariate normal that adapts to the covariance of the posterior.
+The adaption algorithm is based on \textcite[algorithm 4]{andrieuTutorial}, with the implementation based on \textcite{ghoshApproximate}'s work.
+Proposals are accepted or rejected using the standard Metropolis-Hastings (MH) acceptance probability.
+See \cref{SEIR:MCMC-algorithm} for full details.
+\begin{algorithm}
+ set $\vec{X_0}$ to an initial value of the parameter vector \;
+ $\vec{\mu_0} = \vec{X_0}$ \;
+ $\matr{\Sigma_0} = \text{diag}(\vec{\mu_0})$ \;
+ $\lambda_0 = 1$ \;
+ \For{$i = 1, \dots, M$}{
+  sample $\vec{Y_}i \sim \N(\vec{\mu_{i-1}}, \lambda_{i-1}\matr{\Sigma_{i-1}})$\;
+  set $\vec{X_i}$ to $\vec{Y_i}$ or $\vec{X_{i-1}}$ using a MH acceptance step\;
+  update the proportion of proposals accepted so far, $\alpha$ \;
+  \eIf(\tcc{No adaptation for 200 iterations}){$i \leq 200$}{
+    $\lambda_i = \lambda_{i-1}$ \;
+    $\vec{\mu_i} = \vec{\mu_{i-1}}$ \;
+    $\matr{\Sigma_i} = \matr{\Sigma_{i-1}}$ \;
+   }{
+    $\gamma_i = (i - 200)^{-0.6}$ \;
+    $\log \lambda_i = \log \lambda_{i-1} + \gamma_i(\alpha - 0.234)$ \;
+    $\vec{\mu_i} = (1 - \gamma_i) \vec{\mu_{i-1}} + \gamma_i \vec{X_i}$ \;
+    $\matr{\Sigma_i} = (1 - \gamma_i) \matr{\Sigma_{i-1}} + \gamma_i (\vec{X_i} - \vec{\mu_i})(\vec{X_i} - \vec{\mu_i})^T$ \;
+  }
+ }
+ \caption{Algorithm for adaptive random-walk Metropolis-Hastings.
+$\vec{\mu_i}$ and $\matr{\Sigma_i}$ are an estimate of the mean and covariance of the posterior distribution using information up to iteration $i$.
+$\text{diag}(\vec{\mu_0})$ is the diagonal matrix with diagonal entries equal to $\vec{\mu_0}$.
+$\lambda_i$ is the scale parameter of the proposal distribution at iteration $i$, tuned to try and ensure an optimal proportion of proposals are accepted (23.4\%).
+$\gamma_i$ is the learning rate, which determines how much adaptation occurs.
+$\gamma_i \to 0$ as $i \to \infty$ so the rate of adaptation is \emph{vanishing}.
+Vanishing adaptation is required for the algorithm to converge to the target distribution~\autocite[section 3]{andrieuTutorial}.}
+ \label{SEIR:MCMC-algorithm}
+\end{algorithm}
 
-\begin{figure}
-\makebox[\textwidth][c]{
-\begin{tikzpicture}[
-    node distance = 2.5cm,
-    on grid,
-    auto,
-    ->,>=stealth',
-    every state/.style={draw,rectangle},
-    ]
+I run 4 chains for 1,000,000 iterations each.
+The first 500,000 were discarded as burn-in, with the remaining chains thinned to every 100th iteration.
+This gives 20,000 posterior samples per region.
+This was further thinned to 2,000 samples when calculating posterior predictive quantities (incidence and prevalence) due to their computational cost.
 
-    \node[state] (S) {$s$};
-    \node[state, right=of S] (E1) {$e_1$};
-    \node[state, right=of E1] (E2) {$e_2$};
-    \node[state, right=of E2] (I1) {$i_1$};
-    \node[state, right=of I1] (I2) {$i_2$};
-    \node[state, right=of I2] (R) {$r$};
+Following \textcite{birrellBayesian}, I reparameterise the model to improve identifiability.
+Rather than directly inferring $\beta_1$, I infer the initial growth rate of the epidemic.
+In addition, I use a parsimonious parameterisation of the initial conditions using the initial proportion of the population that are infected and the initial proportion that are susceptible only.
+Both of these rely on an assumption that the epidemic is in a steady-state exponential growth phase at time 0 (see \cref{SEIR:sec:SIR}).
 
-    \path (S) edge node {$\vec{\lambda} \circ \vec{s}$} (E1)
-          (E1) edge node {$\sigma$} (E2)
-          (E2) edge node {$\sigma$} (I1)
-          (I1) edge node {$\gamma$} (I2)
-          (I2) edge node {$\gamma$} (R);
-
-    \node[state, below=3cm of E1] (P0) {$p_0$};
-    \node[state, right=of P0] (P1) {$p_1$};
-    \node[state, right=of P1] (P2) {$p_2$};
-    \node[state, right=of P2] (P3) {$p_3$};
-    \node[state, right=of P3, draw=none] (P4) {};
-
-    \path  (S) edge node {$\vec{\lambda} \circ \vec{s}$} (P0)
-          (P0) edge node {$\kappa_0$} (P1)
-          (P1) edge node {$\kappa_Y$} (P2)
-          (P2) edge node {$p_x \kappa_{X1}$} (P3)
-               edge [out=-45,in=-135] node[below] {$(1 - p_x) \kappa_{X1}$} (P4)
-          (P3) edge node {$\kappa_{X2}$} (P4);
-
-    \node[draw, thick, inner sep=0.3cm, fit=(P1) (P3), fill=blue!10,opacity=0.2] {};
-\end{tikzpicture}
-}
-  \caption[SEIR model]{SEIR model used for the analysis in this chapter. The shaded box represents PCR-positive individuals. Shown for a single strata $i$.}
-  \label{SEIR:fig:full-model}
-\end{figure}
-
-The full model combines the elements discussed in the previous sections.
-The transmission model follows \textcite{birrellRealtime}.
-It is a SEIR model with two latent and two infectious states (see \cref{SEIR:sec:non-exponential}), producing a gamma distribution.
-The population is stratified by age, using contact matrices to  (see \cref{SEIR:sec:structured-populations}).
-The force-of-infection and contact matrices changes over time as described in \cref{SEIR:sec:time-varying-foi}.
-The population in the model is that of one region of England.
-To form whole of England estimates, each region is modelled independently with the results then combined (see \cref{SEIR:sec:geography}).
-
-The observation model is novel, to relate the transmission to CIS data.
-It uses a phase-type model to represent the duration distribution of PCR positivity as explained in \cref{SEIR:sec:observation}.
-
-A diagram of the model for a single stratum is shown in figure \cref{SEIR:fig:full-model}.
-The system of differential equations for each region is below.
+During such an exponential growth phase, all components of the $\vec{e}$, $\vec{i}$ and $\vec{p}$ compartments grow at the same growth rate, $\psi$.
+This follows directly from the ODEs where $\vec{s}$ is constant. 
+The growth rate $\psi$ is easily identifiable from the data, and therefore a parameterisation in terms of $\phi$ should be more computationally efficient.
+If this is the initial exponential phase, when $\vec{s} \approx 1$, then the basic reproduction number for a SEIR model is linked to this growth rate by~\autocites{birrellBayesian}{wearingAppropriate}:
 \begin{align}
-    \label{SEIR:eq:fullODEs}
-    \frac{d\vec{s}(t)}{dt} &= -\vec{\lambda}(t) \circ \vec{s}(t) \\\\
-    \frac{d\vec{e_1}(t)}{dt} &= \vec{\lambda}(t) \circ \vec{s}(t) - 2\sigma \vec{e_1}(t) \\
-    \frac{d\vec{e_2}(t)}{dt} &= 2\sigma \vec{e_1}(t) - 2\sigma \vec{e_2}(t) \\
-    \frac{d\vec{i_1}(t)}{dt} &= 2\sigma \vec{e_2}(t) - 2\gamma \vec{i_1}(t) \\
-    \frac{d\vec{i_2}(t)}{dt} &= 2\gamma \vec{i_1}(t) - 2\gamma \vec{i_2}(t) \\
-    \frac{d\vec{r}(t)}{dt} &= 2\gamma \vec{i_2}(t) \\
-    \frac{d\vec{p_0}(t)}{dt} &= \vec{\lambda} \circ \vec{s}(t) - \kappa_0 \vec{p_0}(t) \\
-    \frac{d\vec{p_1}(t)}{dt} &= \kappa_0 \vec{p_0}(t) - \kappa_Y \vec{p_1}(t) \\
-    \frac{d\vec{p_2}(t)}{dt} &= \kappa_Y \vec{p_1}(t) - \kappa_{X1} \vec{p_2}(t) \\
-    \frac{d\vec{p_3}(t)}{dt} &= p_x \kappa_{X1} \vec{p_2}(t) - \kappa_{X2} \vec{p_3}(t) \\
-    \vec{\lambda}(t) &= (\matr{\beta}(t) \circ \matr{\kappa}(t)) (\vec{i_1}(t) + \vec{i_2}(t))
+    \R = \frac{\psi \left( \frac{\psi}{2\sigma} + 1 \right)^2}{\gamma \left(1 - \frac{1}{\left(\frac{\psi}{2 \gamma} + 1 \right)^2} \right)} \label{SEIR:eq:rtoR}.
 \end{align}
 
-\todo[inline]{following paragraph needs rewording having reordered this chapter}
-Recall that each region is run independently (see \cref{SEIR:sec:geography}), $\matr{\beta}(t)_{ij} = \beta(t) \beta^\text{susc}_i$ with $\beta^\text{susc}_i = \beta_c$ for child-aged strata and $\beta^\text{susc}_i = 1$ for adult-aged strata.
+Furthermore, $\R$ is a function of $\matr{\beta}$, and hence $\beta_1$ (as explained in \cref{SEIR:sec:structured-populations}).
+Combining these expressions, the model can be parameterised by using $\psi$ in place of $\beta_1$.
+The following procedure gives $\beta_1$ in terms of $\psi$.
+First, calculate $\R$ from $\psi$ using \cref{SEIR:eq:rtoR}; denote this $\hat{R}$.
+Then calculate $R^* = D(\matr{\beta}) / \gamma$ when $\beta_1=1$ (recall that $D(\matr{\beta})$ denotes the dominant eigenvalue of $\matr{\beta}$).
+Multiplying a matrix by a scalar multiplies its dominant eigenvalue by that same scalar, therefore, setting $\beta_1 = \hat{R} / R^*$ gives the desired means that $\R$ at time 0 is $\hat{R}$, as desired.
 
-As in \cref{E-backcalc}, I use CIS data to estimate the model for each region independently, and over the same period as in that chapter.
-As well as the motivation given previously, this allows heterogeneity in the parameters between regions.
-This heterogeneity could arise from a range of socioeconomic factors.
-Furthermore, the assumption of homogeneous mixing within age groups is more plausible for regions than the whole of England.
+To form a parsimonious parameterisation of the initial state, first focus on the initial proportion of the population that is in the infected compartments.
+Recall that in the exponential growth phase, the relative proportion of each strata that is infected is proportional to the dominant eigenvector of $\matr{\beta}$, which I denote $\vec{d}$.
+Introduce a model parameter $i^+$, the proportion of the population in the $i$ states at time 0 in the strata with the highest proportion infected.
+Then, $\vec{i_0} = \vec{i_1} + \vec{i_2} = \frac{i^+}{\max \vec{d}} \vec{d}$.
+Now introduce an additional vector of parameters, $\vec{\pi}$, where $\pi_a$ is the proportion of strata $i$ without immunity at time 0, hence, $\vec{r} = 1 - \vec{\pi}$.
 
-By having independent models for each region, I assume that the epidemics in each region do not interact.
-Clearly, this is not true as travel between regions of England is common and unrestricted.
-However, for a large epidemic, the non-interaction is a good approximation~\autocite{birrellRealtimea}.
-
-The test results are aggregated to age group and region on each day.
-
-\subsection{Solving the systems of ODEs} \label{SEIR:sec:solving-ODEs}
-I solve the system of ODEs using Euler's method, discretised at timesteps of $\delta t = 1/2$ days.
-That is, if $\vec{x}_t$ is the system state at time $t$ for $t \in \{ 0, 0.5, 1, \dots \}$ then $x_{t+\frac{1}{2}} = x_t + \frac{d\vec{x}(t)}{dt} \delta t$
-
-The mean prevalence on day $t$, used for the likelihood (see \cref{SEIR:eq:likelihood}), can be calculated as $\bar{p}_{kat} = \frac{1}{2} ( p_{kat} + p_{ka,t+1/2} )$.
+From the system of ODEs\todo{might need to reorder this because I've not presented these yet..
+} and the knowledge that all the compartments are growing exponentially at rate $\psi$, one can derive the following for the initial state of the compartments:
+\begin{align}
+    \vec{i_1} &= \vec{i_0} \left(1 + \frac{2\gamma}{\psi + 2\gamma} \right)^{-1} \\
+    \vec{i_2} &= \vec{i_0} - \vec{i_1} \\
+    \vec{e_2} &= \vec{i_1} \frac{\psi + 2\gamma}{2\sigma} \\
+    \vec{e_1} &= \vec{e_2} \frac{\psi + 2\sigma}{2\sigma} \\
+    \vec{s} &= \vec{\pi} - \vec{i_0} - \vec{e_1} - \vec{e_2} \\
+    \vec{p_0} &= \vec{e_1} \frac{\psi + 2\sigma}{\psi + \kappa_0} \\
+    \vec{p_1} &= \vec{p_0} \frac{\kappa_0}{\psi + \kappa_Y} \\
+    \vec{p_2} &= \vec{p_1} \frac{\kappa_Y}{\psi + \kappa_{X1}} \\
+    \vec{p_3} &= \vec{p_2} \frac{p_x \kappa_{X1}}{\psi + \kappa_{X2}}.
+\end{align}
 
 \section{Simulation study} \label{SEIR:sec:sim-study}
+
 I ran a simulation study to check that the inference procedure (described in \cref{SEIR:sec:inference-implementation}) was working correctly and that the model was identifiable.
@@
 Priors and population sizes were matched to the East of England region, over the same time period as the main analysis.
+
+All fixed values, including $\matr{\kappa}$, used the values from the main analysis.
+This means that the simulations created realistic incidence and prevalence curves (see \cref{SEIR:fig:sim-data}).
+In particular, the October half-term school holiday is visible, especially in the school-age groups and the November lockdown is visible to some extent in all age groups.
+\begin{figure}
+    \thisfloatpagestyle{empty}
+    \vspace{-3cm}
+    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/sim/data}}
+    \caption[Simulated data]{%
+        Incidence (top) and prevalence (bottom) for the 100 simulated data sets as a proportion of each stratum.
+        Prevalence shown excludes 
+    }
+    \label{SEIR:fig:sim-data}
+\end{figure}
 
@@
 
-All regions' MCMC chains converged satisfactorily.
-Two chains had to be restarted with new seeds to achieve this as they converged to locations which were clearly incorrect (not shown).
-Following this, the Rhat for all parameters except $\sigma_\epsilon$ had Rhat < 1.01.
-All $\sigma_\epsilon$ parameters had Rhat < 1.05.
-All parameters had ESS > 500, except $\sigma_\epsilon$ in London which had ESS of 182.
-Since $\sigma_\epsilon$ is not a parameter of interest, and all other parameters had sufficient ESS and Rhat, I considered the chains to have converged.
+All regions' MCMC chains converged, assessed using the Rhat statistic and ESS.
+Rhat for all parameters is < 1.02, and for all parameters except $\sigma_\epsilon$ (a nuisance parameter), less than 1.01.
+All parameters had ESS > 500, except $\sigma_\epsilon$ in the East Midlands, which had ESS of 334.
 
@@
 While this analysis cannot establish causation, several hypotheses can be identified that are worth further investigation.
-The decreases in $\beta_t$ over the autumn could be due to increased caution as cases and deaths increased (\eg masking or stricter adherence to social distancing)\todo{cite something about caution and reaction to rising cases, or survey data showing this happened - CoMix?}, or the impact of local measures in the highest transmission areas\todo{cite local measures}.
+The decreases in $\beta_t$ over the autumn could be due to increased caution as cases and deaths increased (\eg masking or stricter adherence to social distancing),\todo{cite something about caution and reaction to rising cases, or survey data showing this happened - CoMix?} or the impact of local measures in the highest transmission areas\todo{cite local measures}.
 The increases in $\beta_t$ in December coincide with the rise of the Alpha variant~\autocite{lythgoeLineage}, known to be more transmissible~\autocite{daviesEstimated}.

### 2023-11-28 ATACCC.tex
Add all comments as todos
ATACCC chapter edits

@@
 \label{ATACCC:fig:viral-load-model}}
+  \todo[inline]{Add a few lines to example plot showing the effect of different parameters, move parameters to key}
 \end{figure}
@@
 Based on the fit to the data achieved when implementing an earlier version of the model (see \cref{ATACCC:sec:original-analysis}), I hypothesised that the cause of this was overcomplicating the model for test failures.
-Specifically, false positives are not needed to be explain the observations (the results in \cref{fig:paper:duration,ATACCC:fig:goodness-of-fits} do not allow false negatives but fit the data well).
-\todo{this sentence is a bit clumsy...}
-Therefore, the model for test failures can be simplified to simply have a constant probability of a false negative.
+Specifically, the model excluding false positives explains the data well (the results in \cref{fig:paper:duration,ATACCC:fig:goodness-of-fits} do not allow false negatives but fit the data well).
+Therefore, the model for test failures can be simplified to only false negatives.
 % These false negatives could represent a variety of scenarios, such as poor swabbing technique.
@@
 \section{Deriving duration from the parameters} \label{ATACCC:sec:duration}
+
 The ultimate aim of this chapter is to produce an estimate of the distribution for the duration of PCR positivity.
-So far, only the viral load trajectories have been considered.
+So far, only the viral load trajectories have been considered; \textcite{hakkiOnset} does not consider this duration either.
 The time that individual $i$ are first and last positive are the two values of $\tau$ such that $\log v_i(\tau) = v_\text{lod}$.
@@
 
-\todo[inline]{Probably important to show estimated test sensitivity here to refer to later}
-
 The original \textcite{hakkiOnset} model has convergence issues (Rhat > 1.07 and ESS < 100) for the parameters of the test failure distribution, $\sigma_0$ and $\rho$.
 The simplified model I proposed in \cref{ATACCC:sec:observation-modification} solved this issue, meaning that this analysis converged.
+\todo[inline]{Add traceplots}
 
@@
 
-While the population-level parameters converged in the modified model, both model had convergence issues (Rhat > 1.4 and ESS < 20) for some individual-level parameters.
-However, this was greatly improved with the simplified model, limited to a single individual, number 55.
-Individual 55 has an extremely noisy set of observations.
-This means that two viral load trajectories fit the data well (shown in \cref{ATACCC:fig:individual-55}).
+For individual-level parameters, both models had convergence issues in places (Rhat > 1.4 and ESS < 20).
+However, this was greatly improved with the simplified model, occurring only for parameters relating to a single individual, number 55.
+Individual 55 has an extremely noisy set of observations, fit well by two different viral load trajectories (shown in \cref{ATACCC:fig:individual-55}).
 The first, found by two chains, increases very quickly, peaks 4.0 days after the first test, then declines slowly.
 The second, found by the other six chains, is more symmetrical, peaking 7.2 days after the first test.
-The individual trajectories are not of interest for this analysis, and the population-level parameters agree between the two sets of chains, therefore this should not affect any estiamtes of interest.
-\todo{clunky phrasing...}
+The population-level parameter estimates agree between the two sets of chains, therefore the issue should not affect any estiamtes of interest.
 \begin{figure}
@@
 
+\todo[inline]{Probably important to show estimated test sensitivity here to refer to later}
+
 \section{Discussion}
@@
 A strength of the ATACCC data is that the sample is drawn from the general population.
-This is unlike most estimates (see\todo{ref review of lit}) which are from hospitalised patients or healthcare workers.
+This is unlike most estimates (see\todo{ref section reviewing lit on duration}) which are from hospitalised patients or healthcare workers.
 This means that, notwithstanding the issues highlighted in the previous paragraph, the estimates are likely to be generalisable to the general population.
@@
 \ifSubfilesClassLoaded{
+  \listoftodos
   \appendix

### 2023-11-28 ATACCC.tex

@@
 The ODE-based approach uses parameters representing biological properties of the virus and host~\autocites[e.g.:][]{ejimaEstimation,keVivo,kimQuantitative,goncalvesTiming,perelsonMechanistic}.
-Such models are rarely analytically-tractable and the system's evolution is computed using numeric methods.
+Such models are rarely analytically-tractable and the system's evolution is computed using numerical methods.
 % Both of these approaches have been used with other viruses and biomarkers.
-The approach that is the focus of this chapter is using a pieceiwse linear model for the biomaker trajectory and random effects to allow for within-individual correlation.
+The approach that is the focus of this chapter is using a piecewise linear model for the biomaker trajectory and random effects to allow for within-individual correlation.
 
@@
 Test failures are modelled as producing a log viral load reading drawn from $\N(x_0, \sigma_0^2)$ where $x_0$ and $\sigma_0^2$ are parameters to be estimated.
-These readings are censored is the same way as true readings.
+These readings are censored in the same way as true readings.
 A censored test failure represents a false negative, otherwise the test failure represents a false positive.
@@
 where $f_N(y \mid \mu, \sigma^2)$ and $F_N(y \mid \mu, \sigma^2)$ are the pdf and cdf at $y$ of a Normal distribution with mean $\mu$ and variance $\sigma^2$ respectively.
-The first term on each line is the likelihood contribution from test failures, either a false positive (top line) of false negative (bottom line).
+The first term on each line is the likelihood contribution from test failures, either a false positive (top line) or false negative (bottom line).
 The second term on the first line is the likelihood of the observed log viral load from non-failed test results.
@@
 However, this was greatly improved with the simplified model, occurring only for parameters relating to a single individual, number 55.
-Individual 55 has an extremely noisy set of observations, fit well by two different viral load trajectories (shown in \cref{ATACCC:fig:individual-55}).
+Individual 55 has an extremely noisy set of observations.
+This means that two viral load trajectories fit the data well (shown in \cref{ATACCC:fig:individual-55}).
 The first, found by two chains, increases very quickly, peaks 4.0 days after the first test, then declines slowly.
 The second, found by the other six chains, is more symmetrical, peaking 7.2 days after the first test.
-The population-level parameter estimates agree between the two sets of chains, therefore the issue should not affect any estiamtes of interest.
+The population-level parameter estimates agree between the two sets of chains, therefore the issue should not affect any estimaates of interest.
 \begin{figure}

### 2023-11-28 biology-data.tex

@@
 
-\section{PCR testing} \label{intro:sec:PCR}
+\section{PCR testing} \label{biology-data:sec:PCR}
+
+False negatives could occur for a variety of reasons, most commonly that an individual swabs themselves poorly and hence the concentration of virus on the swab is too low to be detected by the PCR process.
+Plausibly, a variety of other reasons could also lead to a false negative such as logistical issues or mislabelling of samples.
 

### 2023-11-28 cis-imperfect-testing.tex

@@
 
-Applying the method of \cref{E-perf-test} to CIS data gave implausible estimates of the duration distribution (see \cref{imperf-test:fig:problem-cis-estimates}).
-In particular, I estimate a very high probability of the episode lasting less than five days, in conflict with the results of \cref{E-ATACCC} and other prior work~\autocite[e.g.:][]{cevikShedding}.
+Applying the method of \cref{E-perf-test} to CIS data gives implausible estimates of the duration distribution (see \cref{imperf-test:fig:problem-cis-estimates}).
+In particular, it estimates a very high probability of the episode lasting less than five days, in conflict with the results of \cref{E-ATACCC} and other prior work~\autocite[e.g.:][]{cevikShedding}.
 
@@
 
-Comparing the simulated and CIS data suggested that this is due to the high frequency of single positive episodes in the CIS data.
-A single positive episode is one in which the episode contains exactly one positive test, intuitively these cause short estimates because there is no lower bound on the length of time the episode lasts.
+Comparing the simulated and CIS data suggests that this is due to the high frequency of single positive episodes in the CIS data.
+A single positive episode is an episode containing exactly one positive test; intuitively, these cause short estimates because there is no lower bound on the length of time the episode lasts.
 Furthermore, due to missing short episodes, there is little or no ability to upper bound the number of very short episodes because these would almost all be missed.
-\Textcite{shenNonparametrica} (building on \textcite{panNote}) showed, in a related setting, that the maximum likelihood (frequentist) estimator is inconsistent and can lead to severe underestimation (bias) of the survival function.
-While the result is not directly applicable, and can not simply be extended, to the more complex scenario within CIS, it does still build an intuition that these single positive episodes could be causing this issue.
-Furthermore, simulations building in even a few excess single positive episodes led to problematic estimation even within the CIS setting.
-\todo[inline]{My previous application and discussion of \textcite{shenNonparametrica} in this setting would need quite significant reworking to include; is this needed or is this slightly hand-wavy discussion enough?}
+\Textcite{shenNonparametrica} (extending \textcite{panNote}) showed that, in a similar setting to here\todo{note why it is different}, the maximum likelihood (frequentist) estimator is inconsistent and can lead to severe underestimation (bias) of the survival function.
+While the result cannot simply be extended to the more complex study design of the CIS, it reinforces the intuition that single positive episodes are likely to be problematic.
+Furthermore, simulations (not shown) building in even a few excess single positive episodes led to problematic estimation even within the CIS setting.
 
-Discussion and interrogation of the data led me to believe the most likely reason for an unexpectedly high number of single positive episodes was the presence of false negatives, meaning a false negative as a negative test result returned by a detectable individual.
-This can occur for a variety of reasons, most commonly that an individual swabs themselves poorly and hence the concentration of virus on the swab is too low to be detected by the PCR process (see \cref{E-intro:sec:PCR}).
-Plausibly, a variety of other reasons could also lead to a false negative such as logistical issues or mislabelling of samples.
-Consideration of these reasons are beyond the scope of this thesis, and I group all of these possible reasons together.
-
-There are two strands of evidence that support this hypothesis.
-First, it is well known that PCR testing can return false negatives \todo{refs for sensitivity} and I needed to explicitly include these in the viral load model of \cref{E-ATACCC}; they can be directly observed within CIS as intermittent negatives (see below).
+% Discussion and interrogation of the data led me to
+I hypothesised that the reason for the unexpectedly high number of single positive episodes (compared to simulation) was the presence of false negatives results.
+There are two strands of evidence supporting this hypothesis.
+First, it is well known that PCR testing can return false negatives  (see \cref{E-biology-data:sec:PCR}); they were explicitly include these in the viral load model of \cref{E-ATACCC}.
+Intermittent negatives (described in  \cref{E-biology-data:sec:PCR}) show that they occur within the CIS.
 Second, adjusting the simulation to include false negatives can reproduce both the CIS data more faithfully and exhibits the same issues when estimating the duration distribution (see \cref{imperf-test:sec:simulate}).
-
-\todo[inline]{maybe mention intermittent negatives, referring to its introduction in chapter 2}
 
@@
 The simplest model for false negatives is a constant probability of returning a negative test even when detectable.
-The probability of testing positive given being detectable is known as the test sensitivity.
-The impact of a test sensitivity, $\psens$, less than perfect (\ie less than 100\%) is formalised by replacing the deterministic $y_i(t)$ with a random binary variable such that:
+The probability of testing positive given being detectable is known as the \emph{test sensitivity}, $\psens$.
+$\psens = 1$ means that there are no false negatives, the situation previously considered.
+False negatives are formalised by replacing the deterministic $y_i(t)$ with a random binary variable such that:
 \begin{equation}
@@
 with $Y_i(t)$ still only defined for $t \in t_i$.
-However, an implausibly low value for $\psens$ is required to reproduce the rate of single positive episodes seen in CIS.
+\todo{check all this notation: is $y_t(t)$ and $t_i$ defined?}
+However, an implausibly low value for $\psens$ is required to reproduce the rate of single positive episodes seen in the CIS.
 \todo[inline]{Figure demonstrating this and references to what reasonable test sensitivities are, including the estimate from the ATACCC chapter}
@@
 False negatives are more likely to occur when the viral load of an individual is low, because there is less virus in their body to be sampled.
-Following the short, initial period of growth, viral load largely declines monotonically with time (see \cref{E-ATACCC}).
+In \cref{E-ATACCC} this mechanism is incorporated by the left-censored observations.
 This suggests that a model with a declining test sensitivity as a function of time since infection might be more suitable.
@@
 \todo[inline]{Next paragraph needs editing for clarity}
-Examination of the CIS data also provides evidence for a declining test sensitivity.
-We can approximately bound the test sensitivity by assuming that all intermittent negatives are false negatives, that the negative following the last positive in an episode could be either a true or false negative, and all other negatives are true negatives.
-The point estimate for the test sensitivity is the true positives / (true positives + false negatives).
-In \cref{imperf-test:fig:bounding-cis-sensitivity} we consider the values produced by assuming the negative following the last positive in an episode could be either a true or false negative as a function of time since the individual was detected (\ie: the first positive test in the episode).
-While the bounds are broad, they clearly suggest a declining test sensitivity over the course of an episode.
+The CIS data provides evidence for a test sensitivity declining over the course of an episode.
+To do this, I classify all test results into three classes, making the following assumptions.
+\begin{enumerate}
+    \item All positive results are true positives.
+    \item All intermittent negatives are false negatives.
+    \item The negative at the end of an episode is a possible false negative.
+    \item All other negatives are true negatives.
+\end{enumerate}
+Here, I am assuming that the probability of two false negative tests at the end of an episode is negligible.
+Then I calculate the test sensitivity as a function of days since the first positive in an episode, bounding this by assuming either all or none of those in group 3 are false negatives.
+% The test sensitivity is the true positives / (true positives + false negatives).
+% In \cref{imperf-test:fig:bounding-cis-sensitivity} we consider the values produced by assuming the negative following the last positive in an episode could be either a true or false negative as a function of time since the individual was detected (\ie: the first positive test in the episode).
+This generates broad bounds, although suggest a declining test sensitivity over the course of an episode (see \cref{imperf-test:fig:bounding-cis-sensitivity}).
 \begin{figure}
-  \todo[inline]{Include the analysis of CIS data with bounding test sensitivity}
+  \todo[inline]{Extract figure from SRS}
   \caption[Bounding test sensitivity using CIS data]{Bounding the test sensitivity using CIS data as a function of time since the infection was detected (see main text for explanation of how the bounds are formed).\label{imperf-test:fig:bounding-cis-sensitivity}}
@@
 
-I propose the following model for test sensitivity compatible with \cref{imperf-test:fig:bounding-cis-sensitivity}, consisting of a linearly declining period then a constant period:
+Based on \cref{imperf-test:fig:bounding-cis-sensitivity}, I propose the following model for test sensitivity, consisting of a linearly declining period then a constant period:
 \begin{equation}
@@
 
-\todo[inline]{Add the evidence that false negatives do create the ssame issue in inference in real and simulated data}
+\todo[inline]{Add the evidence that false negatives do create the same issue in inference in real and simulated data}
 
@@
 A constant test sensitivity is assumed throughout this section.
-Further work is required to introduce the more realistic varying test sensitivity (discussed in \cref{imperf-test:sec:discussion}).
+Further work is required to introduce the more realistic assumption of declining test sensitivity over the course of an episode.
 
@@
 This assumption is reasonable because the test sensitivity is high early in an infection, therefore $l_i^{(b)}$ is unlikely to be a false negative.
-\todo{Are these bounds correct?}
+\todo{Are these bounds correct - check notation from prev chapter?}
 We will consider only the tests between $r_i^{(b)}$ and $l_i^{(e)}$ inclusive (the positive tests providing a lower bound on the length of the episode).
@@
 
+\todo[inline]{notes on sim results into prose}
+
 \begin{itemize}
@@
 \subsection{Data}
+
+\todo[inline]{Check if I've talked about the data before - previous chapter?}
 
@@
 
-Think about what I need to include here.
+\todo[inline]{Add additional info on methods. Prior on number missed; hyperparameters}
 Might include hyperparameters etc or assumptions made?
@@
 
+\todo[inline]{notes on real results into prose}
+
+\begin{itemize}
+    \item not sensitivite to survival prior
+    \item sensitive to test sensitivity assumptions and prior on number missed
+    \item comparison with ATACCC
+    \item mean and median summaries
+    \item 
+\end{itemize}
+
 \section{Discussion} \label{imperf-test:sec:discussion}
+
+\todo[inline]{discussion on real results into prose}
 
@@
 
+\todo[inline]{write conclusion}
+
 \ifSubfilesClassLoaded{

### 2023-11-28 ATACCC-appendix-original-analysis.tex
Updates from Overleaf

@@
 
-This chapter describes an early analysis I did on an initial version of the ATACCC dataset.
+\todo[inline]{Check for "REF" in this appendix}
+
+This appendix describes an early analysis I did on an initial version of the ATACCC dataset.
 This dataset is not publicly available.
@@
 The parameter $\vec\theta_i$ introduces dependence between observations of the same individuals.
-\autoref{sec:paper:hierarchy} explains the hierarchical structure which borrows information between individuals.
+\Cref{sec:paper:hierarchy} explains the hierarchical structure which borrows information between individuals.
 
@@
 We assume that $\log t_{i,\text{peak}}$, $C_{i,\text{peak}}$, and $\log \beta_i$ are drawn independently from univariate normal distributions (details below).
-Other parameters are not hierarchical and have prior distributions placed on them in the normal way; \autoref{fig:paper:model_DAG} shows these relationships.
+Other parameters are not hierarchical and have prior distributions placed on them in the normal way; \cref{fig:paper:model_DAG} shows these relationships.
 As in \textcite{fogartyBayesian}, a log transform is applied to the parameters which must be positive for biological plausibility.
@@
 
-The model generally shows a good fit to the observed data (see \autoref{fig:paper:goodness_of_fit}).
+The model generally shows a good fit to the observed data (see \cref{fig:paper:goodness_of_fit}).
 For individuals with few or no observations before their peak viral load, information is borrowed from other individuals to infer when this is likely to have occurred.
@@
 
-We estimate that the time from the time from when an individual has a 50\% chance of detection to peak viral load (equivalently: when $\hat{C} = C_\text{lod}$) is 2.4 (posterior median; with 95\% central CrI 1.5--3.3) days with a population standard deviation on the log scale of 0.69 (0.42--1.06).
+We estimate that the time from when an individual has a 50\% chance of detection to peak viral load (equivalently: when $\hat{C} = C_\text{lod}$) is 2.4 (posterior median; with 95\% central CrI 1.5--3.3) days with a population standard deviation on the log scale of 0.69 (0.42--1.06).
 The mean peak viral corresponds to 17.8 (17.1--18.6) Ct units, with a standard deviation of 2.2 (1.5--3.2) Ct units, after which the decrease in viral corresponds to Ct values increasing by 1.6 (1.5--1.8) Ct / day, with a standard deviation on the log scale of 0.50 (0.40--0.61).
-Predictive distributions for these parameters for a new individual are shown in \autoref{fig:paper:indiv_predict}.
+Predictive distributions for these parameters for a new individual are shown in \cref{fig:paper:indiv_predict}.
 We estimate a 6.8 (5.0--8.8) \% probability of a false negative independent of the viral load of an individual.
@@
 This gives a mean duration estimated at 18.4 (16.5--20.9) days but the distribution is skewed, as seen by the median of 16.6 (14.8--18.5) days.
-The full distribution is in \autoref{fig:paper:duration}.
+The full distribution is in \cref{fig:paper:duration}.
 \begin{figure}
@@
 This study presents estimates of viral load dynamics based on a general population sample with daily swabbing.
-Our results show a peak Ct value around 5 Ct lower than \textcite{kisslerViral}, that is a higher viral load, although occurring at a similar time (see \autoref{tab:paper:compare}) and a longer time from first detection to peak viral load than \textcite{jonesEstimating}.
+Our results show a peak Ct value around 5 Ct lower than \textcite{kisslerViral}, that is a higher viral load, although occurring at a similar time (see \cref{tab:paper:compare}) and a longer time from first detection to peak viral load than \textcite{jonesEstimating}.
 This could be because of the different populations in each study.

### 2023-11-28 ATACCC.tex
Fix typo
Updates from Overleaf

--- 2023-11-24 ATACCC.tex
@@
 
-Let the observed log viral load for individual $i$ at time $t$ be $y_{i,t}$, with negative tests signified by $y_{i,t}=v_\text{lod}$.
+Let the observed log viral load for individual $i$ at time $t$ be $y_{i,t}$, with $y_{i,t}=0$ if the test is negative.
 Then, its likelihood is:
@@
 \todo{this sentence is a bit clumsy...}
-Therefore, the model for test failures can be simplified to simply have a constant probability of a false negative.
+Therefore, the model for test failures can be simplified to have a constant probability of a false negative.
 % These false negatives could represent a variety of scenarios, such as poor swabbing technique.
@@
 
-\todo[inline]{Probably important to show estimated test sensitivity here to refer to later}
-
 The original \textcite{hakkiOnset} model has convergence issues (Rhat > 1.07 and ESS < 100) for the parameters of the test failure distribution, $\sigma_0$ and $\rho$.
@@
 
-While the population-level parameters converged in the modified model, both model had convergence issues (Rhat > 1.4 and ESS < 20) for some individual-level parameters.
+While the population-level parameters converged in the modified model, both models had convergence issues (Rhat > 1.4 and ESS < 20) for some individual-level parameters.
 However, this was greatly improved with the simplified model, limited to a single individual, number 55.
@@
 The second, found by the other six chains, is more symmetrical, peaking 7.2 days after the first test.
-The individual trajectories are not of interest for this analysis, and the population-level parameters agree between the two sets of chains, therefore this should not affect any estiamtes of interest.
+The individual trajectories are not of interest for this analysis, and the population-level parameters agree between the two sets of chains, therefore this should not affect any estimates of interest.
 \todo{clunky phrasing...}

### 2023-11-29 SEIR.tex
Updates from Overleaf
Figures and edits

@@
 Mechanistic models explicitly represent the infection process driving the spread of infectious disease, giving model's state and parameters have a biological and/or epidemiological interpretation.
-Of particular interest is that the basic reproduction number, $R_0$, and the effective reproduction number, $R_e(t)$, which can both be calculated as functions of the model parameters; using \cref{E-backcalc}'s approach, additional modelling would be required to estimate these numbers.
+Of particular interest is that the basic reproduction number, $\R$, and the effective reproduction number, $R_e(t)$, which can both be calculated as functions of the model parameters; using \cref{E-backcalc}'s approach, additional modelling would be required to estimate these numbers.
 Furthermore, mechanistic models can be used for scenario-based modelling, which simulates the effect of interventions.
@@
     It is in this phase that the basic reproduction number provides a good approximation of the epidemic dynamics.
-    \item The basic reproduction number for this model is $R_0 = \beta / \gamma$~\autocite[20]{keelingModeling}.
+    \item The basic reproduction number for this model is $\R = \beta / \gamma$~\autocite[20]{keelingModeling}.
     Intuitively, this can be thought of as the average number of infections caused by an infectious individual per day, $\beta$, multiplied by the average number of days they are infectious for, $1/\gamma$.
-    \item The effective reproduction number for this model is $R_e(t) = \beta / \gamma s(t) = R_0 s(t)$~\autocite{pellisEstimation}.
+    \item The effective reproduction number for this model is $R_e(t) = \beta / \gamma s(t) = \R s(t)$~\autocite{pellisEstimation}.
     This is the basic reproduction number reduced because some infections are avoided due to individuals being immune.
@@
 
-The reproduction number in terms of the parameters is unchanged by this extension, \ie $R_0 = \beta / \gamma$.
+The reproduction number in terms of the parameters is unchanged by this extension, \ie $\R = \beta / \gamma$.
 
@@
 
-The reproduction number in terms of the parameters is unchanged by this extension, \ie $R_0 = \beta / \gamma$, however, the growth rate in the exponential phase will be slower than before~\autocite[41]{keelingModeling}.
+The reproduction number in terms of the parameters is unchanged by this extension, \ie $\R = \beta / \gamma$, however, the growth rate in the exponential phase will be slower than before~\autocite[41]{keelingModeling}.
 
@@
 The standard solution to this was proposed by \textcite{diekmannDefinition} (\textcite[chapter 7]{diekmannMathematical} provides a more gentle, but still rigorous, explanation).
-First, they show that the distribution of infections across the strata during the initial period of exponential growth (when $\vec{s} \approx \vec{0}$) relies only on $\matr{\beta}$.
+First, they show that the distribution of infections across the strata during the initial period of exponential growth (when $\vec{s} \approx \vec{1}$) relies only on $\matr{\beta}$.
 This distribution is proportional to the \emph{dominant eigenvector} (the eigenvector corresponding to the largest eigenvalue) of $\matr{\beta}$.
 Denote by $D(\matr{\beta})$ the \emph{dominant eigenvalue} (the largest eigenvalue).
-The basic reproduction number is then defined as $R_0 = D(\matr{\beta}) / \gamma$.
-This $R_0$ keeps the same interpretation as in the homogeneous case, \ie the average number of secondary infections caused by a single infection in a fully susceptible population, where the average is taken over the distribution implied by the dominant eigenvector.
-The important threshold that exponential growth occurs (in expectation) iff $R_0 > 1$ remains.
+The basic reproduction number is then defined as $\R = D(\matr{\beta}) / \gamma$.
+This $\R$ keeps the same interpretation as in the homogeneous case, \ie the average number of secondary infections caused by a single infection in a fully susceptible population, where the average is taken over the distribution implied by the dominant eigenvector.
+The important threshold that exponential growth occurs iff $\R > 1$ remains.
 
@@
 
-From the system of ODEs\todo{might need to reorder this because I've not presented these yet..
-} and the knowledge that all the compartments are growing exponentially at rate $\psi$, one can derive the following for the initial state of the compartments:
+From the system of ODEs and the knowledge that all the compartments are growing exponentially at rate $\psi$, one can derive the following for the initial state of the compartments:
 \begin{align}
@@
         Incidence (top) and prevalence (bottom) for the 100 simulated data sets as a proportion of each stratum.
-        Prevalence shown excludes 
+        Prevalence shown is $\bar{p}_{kat}$, the true population prevalence on each day, excluding sampling noise.
+        Each line represents one simulation.
     }
@@
 On some occasions, the proposal distribution's covariance matrix became singular (or close to singular).
+\todo{Update with fitting issues if they reoccur on final runs}
 This means it is not possible to propose a new value.
 In these cases, I restarted the chain with a new seed, discarding all information from the original chain; this was required at most once per chain.
+
+To check that the inference was working correctly, I calculated the coverage of the 50\%, 75\%, and 95\% central credible intervals for each parameter and simulation.
+I define coverage as the proportion of simulations where the true value of the parameter in that simulation was within the credible interval.
+The coverage for each parameter should equal the nominal probability (\eg the coverage should be 50\% for the 50\% credible intervals).
+I constructed a confidence interval for the coverage using the Pearson-Klopper method implemented in the binom package~\autocite{binom1-1}.
+
+Overall, coverage was\todo{comment on sim study coverage}
+\begin{figure}
+    \includegraphics{SEIR/sim/coverage}
+    \caption[Coverage of simulation study]{%
+        Coverage of the 50\%, 75\%, and 95\% central credible intervals for each parameter.
+        The lines shows the nominal coverage.
+    }
+    \label{SEIR:fig:sim-coverage}
+\end{figure}
 
@@
 There is little-to-no ascertainment bias in the CIS, unlike sources relying on testing behaviour (in the UK, pillar 2 testing).
-The CIS allows data-informed estimates of transmission in children, unlike sources relying on severe events that occur almost entirely in adults\todo{cite something about mildness in children or age distribution of severe events}.
+The CIS allows data-informed estimates of transmission in children, unlike sources relying on severe events that occur almost entirely in adults~\autocite{bhopalChildren}.
 
@@
 While this analysis cannot establish causation, several hypotheses can be identified that are worth further investigation.
-The decreases in $\beta_t$ over the autumn could be due to increased caution as cases and deaths increased (\eg masking or stricter adherence to social distancing),\todo{cite something about caution and reaction to rising cases, or survey data showing this happened - CoMix?} or the impact of local measures in the highest transmission areas\todo{cite local measures}.
+The decreases in $\beta_t$ over the autumn could be due to increased caution(\eg masking or stricter adherence to social distancing) and/or interventions short of a lockdown as cases and deaths increased~\autocite{jarvisEffect}, or the impact of local measures in the highest transmission areas\todo{cite local measures}.
 The increases in $\beta_t$ in December coincide with the rise of the Alpha variant~\autocite{lythgoeLineage}, known to be more transmissible~\autocite{daviesEstimated}.
@@
 
+\todo[inline]{write conclusion and add to discussion once comparison with backcalculation is made}
+
 \ifSubfilesClassLoaded{

### 2023-11-29 SEIR/CIS/params.R

@@ -77,15 +77,15 @@ posterior |>
     arrange(ESS) |>
     readr::write_csv(here::here("SEIR/CIS/ess.csv"))
 
-posterior |>
-    filter(
-        !parameter %in% c("beta", "pi", "theta")
-    ) |>
-    ggplot(aes(x = value)) +
-    geom_density(aes(colour = "Posterior")) +
-    geom_density(aes(colour = "Prior"), data = prior_samples) +
-    facet_grid(region~parameter, scales = "free") #+
-    coord_cartesian(ylim = c(0, 1))
+# posterior |>
+#     filter(
+#         !parameter %in% c("beta", "pi", "theta")
+#     ) |>
+#     ggplot(aes(x = value)) +
+#     geom_density(aes(colour = "Posterior")) +
+#     geom_density(aes(colour = "Prior"), data = prior_samples) +
+#     facet_grid(region~parameter, scales = "free") #+
+#     coord_cartesian(ylim = c(0, 1))
 
 n_prior_samples = 10e3
 prior_samples = tibble(
@@ -101,24 +101,24 @@ prior_samples = tibble(
         values_to = "value"
     )
 
+# posterior |>
+#     filter(region == "South_East_England") |>
+#     ggplot(aes(.iteration, value, colour = .chain)) +
+#     geom_line() +
+#     facet_wrap(~parameter+i, scale = "free_y") 
 posterior |>
-    filter(region == "South_East_England") |>
+    filter(region == "North_East_England") |>
     ggplot(aes(.iteration, value, colour = .chain)) +
     geom_line() +
     facet_wrap(~parameter+i, scale = "free_y") 
-posterior |>
-    filter(region == "North_West_England") |>
-    ggplot(aes(.iteration, value, colour = .chain)) +
-    geom_line() +
-    facet_wrap(~parameter+i, scale = "free_y") 
-posterior |>
-    filter(region == "West_Midlands") |>
-    ggplot(aes(.iteration, value, colour = .chain)) +
-    geom_line() +
-    facet_wrap(~parameter+i, scale = "free_y") 
+# posterior |>
+#     filter(region == "West_Midlands") |>
+#     ggplot(aes(.iteration, value, colour = .chain)) +
+#     geom_line() +
+#     facet_wrap(~parameter+i, scale = "free_y") 
 
-posterior |>
-    filter(region == "West_Midlands", parameter == "theta") |>
[…]

### 2023-11-29 SEIR/CIS/posterior_predictive.R

@@ -7,6 +7,7 @@ library(stringr)
 library(tidybayes)
 library(tidyr)
 source(here::here("utils.R"))
+source(here::here("SEIR/utils.R"))
 
 # Read posterior predictives
 predictions = readr::read_csv(here::here("SEIR/CIS/predictive.csv")) |>
@@ -14,7 +15,8 @@ predictions = readr::read_csv(here::here("SEIR/CIS/predictive.csv")) |>
         .chain = factor(chain + 1),
         .iteration = iteration + 1,
         .draw = max(.iteration) * (as.integer(.chain) - 1) + .iteration,
-        age = str_replace(age, fixed("+"), "Inf")
+        age = str_replace(age, fixed("+"), "Inf"),
+        region = region_labels(region)
     ) |>
     select(!c(chain, iteration))
 
@@ -23,7 +25,8 @@ data = readr::read_csv(here::here("SEIR/CIS/data.csv")) |>
     mutate(
         obs_prevalence = obs_positives / num_tests,
         age = str_replace(age, fixed("("), "[") |>
-            str_replace(fixed("]"), ")")
+            str_replace(fixed("]"), ")"),
+        region = region_labels(region),
     ) |>
     rename(day = date) |>
     filter(
@@ -40,7 +43,7 @@ prediction_intervals = predictions |>
 thetas = readr::read_csv(here::here("SEIR/CIS/params.csv")) |>
     filter(parameter == "theta") |>
     transmute(
-        region,
+        region = region_labels(region),
         .chain = factor(chain + 1),
         .iteration = iteration + 1,
         theta = exp(value)
@@ -79,12 +82,13 @@ create_prev_plot = function(age_groups, label) {
     plot = weekly_predictions |>
         filter(age %in% age_groups) |>
         ggplot(aes(week)) +
-        geom_point(aes(y = p)) +
+        geom_point(aes(y = p), size = 0.5) +
         geom_ribbon(aes(ymin = pred_p.lower, ymax = pred_p.upper), alpha = 0.1) +
         geom_lineribbon(
             aes(day, prevalence, ymin = prevalence.lower, ymax = prevalence.upper),
             data = filter(prediction_intervals, age %in% age_groups),
-            alpha = 0.4
+            alpha = 0.4,
+            size = 0.2
         ) +
         facet_grid(region ~ age) +
         standard_plot_theming() +
@@ -92,7 +96,10 @@ create_prev_plot = function(age_groups, label) {
             x = "Date",
             y = "Prevalence"
[…]

### 2023-11-29 SEIR/utils.R

@@ -1,0 +1,19 @@
+# Plot labels
+# From: https://chat.openai.com/c/7e15bbc2-473a-4f5b-b54c-7bb9cef8c1f5
+parameter_labels <- function(label) {
+    labels <- dplyr::case_match(
+        label,
+        "beta[0]" ~ "sigma[epsilon]",
+        "psir" ~ "psi",
+        "theta" ~ "rho",
+        "matrix_modifiers" ~ "q[c]",
+        "i0" ~ "i^`+`",
+        .default = label
+    )
+    parse(text = labels)
+}
+
+region_labels = function(label) {
+    stringr::str_replace_all(label, "_", " ") |>
+        stringr::str_replace_all(" England", "")
+}

### 2023-11-29 utils.R

@@ -1,6 +1,6 @@
 standard_plot_theming = function() {
     rlang::list2(
-        theme_minimal()
+        theme_minimal(),
     )
 }
 

### 2023-11-29 SEIR.tex
Continue edits

@@
 
+I transformed parameters with restricted ranges for sampling, to improve efficiency.
+Parameters which are constrained to the positive reals ($\sigma_\epsilon$, $\rho$, and $q_c$) were sampled on the log scale.
+Parameters which are constrained to the unit interval ($i^+$ and $\vec{\pi}$) were sampled on the logit scale.
+
 I run 4 chains for 1,000,000 iterations each.
@@
 
+This approach provides additional understanding, compared to the approach in \cref{E-backcalc}.
+For example, decomposing the change in transmission into changing contacts, $\matr{\kappa}(t)$, changing susceptibility $\vec{s}(t)$, and other unmeasured factors, $\beta(t)$.
+By understanding which of these components is changing, policies can be better targeted.
+If it is shown that $\beta(t)$ decreases, this implies that for a given amount of activity in the population, there is less transmission.
+Such decreases might be beneficial as it allows society to operate in a more normal manner without increasing transmission.
+Breaking the trade-off between transmission and activity is desirable because it reduces the disease burden without burdesome restrictions on individuals day-to-day lives.
+
 The parameters $\beta_t$ absorb all the time-varying factors not captured by Google mobility and school attendance data.

### 2023-11-29 ATACCC.tex
ATACCC todos

--- 2023-11-28 ATACCC.tex
@@
 First, I will explain how biomarkers are typically modelled, both for SARS-CoV-2 and historically in the biostatistical literature (\cref{ATACCC:sec:background}).
-Next, I describe the model of \textcite{hakkiOnset} in detail, which applies these methods to the ATACCC data (\cref{ATACCC:sec:hakki}) and before I modifying its observation model (\cref{ATACCC:sec:observation-modification})\todo{think if some justification is needed here, and if so how to do it - remember Hakki is a published and peer-reviewed analysis}.
+Next, I describe the model of \textcite{hakkiOnset} in detail, which applies these methods to the ATACCC data (\cref{ATACCC:sec:hakki}) and before modifying its observation model to improve the inference (\cref{ATACCC:sec:observation-modification}).
 This modification is based on an earlier analysis using a preliminary version of the dataset, described in \cref{ATACCC:sec:original-analysis}.
@@
 
-The noisy observations are modelled by assuming each measurement of log viral load is a realisation of an iid Normally-distributed random variable error with mean 0 and standard deviation $\sigma_v$.
-\todo{Paul/Dani: is this phrasing "a realisation of [a] random variable" OK?}
+The noisy observations are modelled by assuming each measurement of log viral load has an iid error distributed $\N(0, \sigma_v)$.
 The measurements are left-censored at the limit of detection, $\vlod$, meaning that any observation which would give a log viral load of less than $\vlod$ is instead observed as negative.
@@
 The simplified model I proposed in \cref{ATACCC:sec:observation-modification} solved this issue, meaning that this analysis converged.
-\todo[inline]{Add traceplots}
 
@@
 The second, found by the other six chains, is more symmetrical, peaking 7.2 days after the first test.
-The population-level parameter estimates agree between the two sets of chains, therefore the issue should not affect any estimaates of interest.
+The population-level parameter estimates agree between the two sets of chains, therefore the issue should not affect any estimates of interest.
 \begin{figure}
@@
 
+The negative rate, $\rho$, is estimated to be quite low, 5.3\% (95\% CrI: 3.7--7.3\%).
+
 The $F_D$ estimates for the two groups are in \cref{ATACCC:fig:duration}.
@@
 The median duration of PCR positivity for the vaccinated population is 13 days (95\% CrI: 11--16 days), but it is 16 days (95\% CrI: 14--19 days) for the unvaccinated population.
-
-\todo[inline]{Probably important to show estimated test sensitivity here to refer to later}
 

### 2023-11-29 ATACCC/false_pos.R

@@ -1,0 +1,19 @@
+library(dplyr)
+library(ggplot2)
+library(patchwork)
+library(purrr)
+library(tidybayes)
+library(tidyr)
+source(here::here("utils.R"))
+
+fit = readRDS(here::here("ATACCC/fit.rds"))
+
+false_pos = spread_draws(fit, l_fp) |>
+    mutate(fp = exp(-l_fp))
+
+false_pos |>
+    median_qi(fp)
+
+false_pos |>
+    ggplot(aes(exp(-l_fp))) +
+    geom_density()

### 2023-11-29 ATACCC/goodness_fit.R

@@ -8,7 +8,7 @@ library(tidyverse)
 source(here::here("ATACCC/trajectory_functions.R"))
 source(here::here("utils.R"))
 
-fit = here::here("ATACCC/fit.rds") |>
+fit = here::here("ATACCC/fit2.rds") |>
     readRDS()
 
 tbl_data = here::here("ATACCC/data.rds") |>

### 2023-11-29 ATACCC/traceplots.R

@@ -1,0 +1,29 @@
+
+library(dplyr)
+library(ggplot2)
+library(purrr)
+library(rstan)
+library(patchwork)
+library(tidybayes)
+library(tidyr)
+source(here::here("utils.R"))
+
+hakki_fit = here::here("ATACCC/fit.rds") |>
+    readRDS()
+simplified_fit = here::here("ATACCC/fit2.rds") |>
+    readRDS()
+
+summary(hakki_fit)$summary |>
+    as_tibble(rownames = "Param") |>
+    filter(!stringr::str_detect(Param, "\\[")) |>
+    arrange(desc(Rhat)) |>
+    select(Param, Rhat) |>
+    print(n = 100)
+
+spread_draws(hakki_fit, l_fp) |>
+    ggplot(aes(.iteration, l_fp, colour = factor(.chain))) +
+    geom_line()
+
+spread_draws(simplified_fit, l_fp) |>
+    ggplot(aes(.iteration, l_fp, colour = factor(.chain))) +
+    geom_line()

### 2023-11-29 ATACCC/trajectories.R

@@ -2,6 +2,44 @@ library(ggplot2)
 source(here::here("ATACCC/trajectory_functions.R"))
 source(here::here("utils.R"))
 
+# min_x = -5
+# max_x = 15
+# x = seq(min_x, max_x, length.out = 1000)
+
+# param_vals = tibble::tribble(
+#   ~tau_max, ~a_i, ~b_i,
+#   exp(15), 5, 1.4,
+#   exp(12.5), 7, 2,
+# )
+
+# trajectories = purrr::pmap_dfr(
+#   param_vals,
+#   function(tau_max, a_i, b_i) {
+#     label_text = bquote(tau[max,i] == .(tau_max) * "," ~ a[i] == .(a_i) * "," ~ b[i] == .(b_i))
+#     tibble::tibble(
+#       # label = parse(
+#       #   text = paste0("tau[max,i] = ", tau_max, "; a[i] = ", a_i, "; b[i] = ", b_i)
+#       # ),
+#       label = as.character(label_text),
+#       x = x,
+#       y = logVL(x, c(tau_max, a_i, b_i)),
+#     )
+#   }
+# )
+
+# ggplot(trajectories, aes(x, y, colour = label)) +
+#   geom_line() +
+#   coord_cartesian(ylim = c(0, NA)) +
+#   labs(
+#     x = "Time from peak viral load",
+#     y = "Log viral load"
+#   ) +
+#   theme_minimal() +  # Replace with your actual theming function
+#   scale_colour_manual(
+#     values = c("blue", "red"),  # Replace with your desired colors
+#     labels = unique(trajectories$label)
+#   )
+
 p_typical_trajectory = ggplot(NULL) +
   stat_function(fun = logVL, args = list(c(exp(15), 5, 1.4)), xlim = c(-5, 15), n = 1000) +
   coord_cartesian(ylim = c(0, NA)) +

### 2023-12-04 SEIR.tex
Finish sim results first draft

@@
 Mechanistic models explicitly represent the infection process driving the spread of infectious disease, giving model's state and parameters have a biological and/or epidemiological interpretation.
-Of particular interest is that the basic reproduction number, $\R$, and the effective reproduction number, $R_e(t)$, which can both be calculated as functions of the model parameters; using \cref{E-backcalc}'s approach, additional modelling would be required to estimate these numbers.
+% Of particular interest is that the basic reproduction number, $\R$, and the effective reproduction number, $R_e(t)$, which can both be calculated as functions of the model parameters; using \cref{E-backcalc}'s approach, additional modelling would be required to estimate these numbers.
+This allow changes in transmission to be decomposed into the components of the model, \eg the number of contacts per day or the probability of transmission upon contact.
 Furthermore, mechanistic models can be used for scenario-based modelling, which simulates the effect of interventions.
@@
 
-\Cref{SEIR:sec:transmission-generic} introduces a generic framework for mechanistic modelling of infectious diseases.
+I start this chapter by introducing a generic framework for mechanistic modelling of infectious diseases (\cref{SEIR:sec:transmission-generic}).
 % This framework is then applied to SARS-CoV-2 in \cref{SEIR:sec:transmission-application}.
-It describes the \emph{transmission model}, corresponding to the infection process (from \cref{E-inc-prev:sec:infection-process}).
+It describes the \emph{transmission model}, corresponding to the infection process.
 The transmission model needs to be linked to observations; in this chapter, that will be the CIS prevalence data.
-The \emph{observation model}, introduced in \cref{SEIR:sec:observation}, is this link; it incorporates both the prevalence process (from \cref{E-inc-prev:sec:prevalence-process}) and the observation process (from \cref{E-inc-prev:sec:observation-process}).
-\Cref{SEIR:sec:inference-implementation} describes how I will apply the model to the CIS prevalence data, including the inference process, parameterisation, and priors for this model.
-This transmission model is the same as that in \textcite{birrellRealtime}, but the observation model is novel.
-\Cref{SEIR:sec:sim-study} checks this recovers the parameters successfully using a simulation study.
-\Cref{SEIR:sec:application} then describes the results with the CIS data, including making comparisons to the results in \cref{E-backcalc}.
-Finally, \cref{SEIR:sec:discussion} discusses the results, limitations, and extensions of this work.
-
-The literature on mechanistic models is vast, having been developed for over a century; the model in \cref{SEIR:sec:SIR} was first formulated in the 1920s~\autocite{kermackContribution}.
-\Cref{SEIR:sec:transmission-generic} explains only the parts relevant to the remainder of the chapter.
-For further details I recommend the tutorial paper \textcite{kretzschmarMathematical} or the textbook \textcite{keelingModeling}.
+The \emph{observation model}, introduced in \cref{SEIR:sec:observation}, is this link; it incorporates both the prevalence process and the observation process.
+These two components form the mathematical model, which I then fit to data, requiring specification of an inference process, parameterisation, and priors (\cref{SEIR:sec:inference-implementation}).
+This final formulation uses the transmission model from \textcite{birrellRealtime}, but a novel observation model.
+I then use a simulation study to check this recovers the parameters successfully (\cref{SEIR:sec:sim-study}).
+Having shown that the model can recover the parameters, I then apply it to the CIS data (\cref{SEIR:sec:application}).
+Finally, I discuss the results, limitations, and extensions of this work (\cref{SEIR:sec:discussion}).
 
@@
 %There remains randomness in the observation model (described in \cref{SEIR:sec:observation}).
+
 Stochastic models are most commonly specified in discrete time, using a Markov assumption.
-That is, they specify a probability distribution $p(\vec{x}(t) \mid \vec{x}(t-1), \vec{\theta})$ for integer $t$ and assume that, having conditioned on $\vec{x}(t-1)$, $\vec{x}(t)$ is independent of $\vec{x}(t')$ for all $t' < t - 1$~\autocite{birrellEvidence}.
+They specify a probability distribution $p(\vec{x}(t) \mid \vec{x}(t-1), \vec{\theta})$ for integer $t$ and assume that, having conditioned on $\vec{x}(t-1)$, $\vec{x}(t)$ is independent of $\vec{x}(t')$ for all $t' < t - 1$~\autocite{birrellEvidence}.
 This relaxes the assumption of independent increments in the counting process formulation of the infection process (see \cref{E-inc-prev:sec:infection-process}).
@@
 This second justification is of the same form as the one made for neglecting the stochasticity of the prevalence process in \cref{E-inc-prev:sec:observation-process}.
-The remainder of this chapter uses only deterministic models, because throughout the time period this thesis considers there were enough infectious individuals.
+
+The remainder of this chapter uses only deterministic models.
+Throughout the time period this thesis considers, there were many infectious individuals.
 For an introduction to stochastic models, see \textcite[chapter 6]{keelingModeling} or \textcite{birrellEvidence}.
 
-These classifications only describe the transmission model.
-The observation model, which links $\vec{x}(t)$ to data, is generally includes some stochasticity regardless~\autocite{birrellEvidence}.
-
 This chapter discusses deterministic models whose evolution is specified by a system of ODEs, $d\vec{x}/dt$.
-Therefore, in an ODE model, time is necessarily continuous, although discretised when solving (see \cref{SEIR:sec:inference-implementation}).
+In an ODE model, time is necessarily continuous, although discretised when solving (see \cref{SEIR:sec:inference-implementation}).
 For concreteness, I use units of days throughout, although the model can use any unit of time.
@@
 This model has several assumptions which make it inappropriate to model SARS-CoV-2; therefore, in the remainder of this section I introduce standard extensions to the SIR model which relax these assumptions.
-First, in \cref{SEIR:sec:non-exponential} I relax the assumption that the infectious period is exponentially-distributed to allow gamma distributions.
-Then, in \cref{SEIR:sec:SEIR} I relax the assumption that individuals are infectious immediately after they are infected; it does this by extending the SIR model to the SEIR model with the additional E signifying an \emph{exposed} compartment.
-Finally, in \cref{SEIR:sec:structured-populations} I include heterogeneity in the population, stratifying by important characteristic(s).
+First, in \cref{SEIR:sec:non-exponential}, I relax the assumption that the infectious period is exponentially-distributed to allow gamma distributions.
+Then, in \cref{SEIR:sec:SEIR}, I relax the assumption that individuals are infectious immediately after they are infected; it does this by extending the SIR model to the SEIR model with the additional E signifying an \emph{exposed} compartment.
+Finally, in \cref{SEIR:sec:structured-populations}, I include heterogeneity in the population, stratifying by important characteristic(s).
 % Finally, \cref{SEIR:sec:time-varying-foi} relaxes the assumption of a constant transmission rate to allow for time-varying behavioural changes.
+
+The literature on mechanistic models is vast, having been developed for over a century; the model in \cref{SEIR:sec:SIR} was first formulated by \textcite{kermackContribution}.
+In this section, I will explain only the parts relevant to the remainder of the chapter.
+For further details I recommend the tutorial paper \textcite{kretzschmarMathematical} or the textbook \textcite{keelingModeling}.
 
@@
 
-The new compartments are considered biologically identical here.
+The new compartments are biologically identical, without an interpretation differentiating the compartments.
 They should be considered as a mathematical trick so that the ODE system produces a gamma distribution rather than an interpretable quantity.
@@
 If the strata have differing transmission patterns, this can change the dynamics of the epidemic.
-The different patterns could be due to behavioural factors (\eg high-numbers of sexual partners in a sexually-transmitted disease~\autocite[69]{keelingModeling} or younger age groups making many more contacts for respiratory diseases~\autocite[176]{andersonInfectious}), or biological factors (\eg genetics reducing the susceptibility of some individuals~\autocite[208]{andersonInfectious}).
+The different patterns could be due to behavioural factors (\eg high-numbers of sexual partners in a sexually-transmitted disease~\autocite[69]{keelingModeling} or younger age groups making many more contacts for respiratory diseases~\autocite[176]{andersonInfectious}), or biological factors such as genetics~\autocite[208]{andersonInfectious}.
 I will assume that the disease progression (\ie length of the exposed and infectious periods) is the same for all strata, although this assumption can be relaxed.
@@
 This section considers the practicalities of achieving this aim via mechanistic models.
-First, I deal with the inference procedure including the MCMC algorithm (\cref{SEIR:sec:MCMC}), parameterising the model (\cref{SEIR:sec:parameterisation}), and the priors for these parameters (\cref{SEIR:sec:priors}).
-This leads to a fully characterised model, explained in \cref{SEIR:sec:full-model}.
-Finally, \cref{SEIR:sec:solving-ODEs} deals with solving the ODE equations numerically.
 
@@
 The first 500,000 were discarded as burn-in, with the remaining chains thinned to every 100th iteration.
-This gives 20,000 posterior samples per region.
+This gives 5,000 posterior samples per region.
 This was further thinned to 2,000 samples when calculating posterior predictive quantities (incidence and prevalence) due to their computational cost.
@@
 I ran 100 simulations, with parameters drawn from the priors (see \cref{SEIR:table:priors}) except that the prior on $\psi$ was $\LN(0.048, 0.0035)$.
-This is because the prior used for the main analysis was too weak on this prior to create realistic simulations.
+I used the more informative prior for the simulation study to ensure that the simulations were realistic.
+The weakly informative prior, used for the main analysis, allows a wide range of epidemics including some that are shrinking at the start of the time period.
 Priors and population sizes were matched to the East of England region, over the same time period as the main analysis.
@@
 
-Simulating from the prior then attempting to recover the parameters is the basis of simulation-based calibration~\autocites{cookValidation}{taltsValidating}.
-The coverage of the posterior credible intervals should be their nominal value (\eg 95\% of the 95\% credible intervals should contain the true parameter value).
-Furthermore, the combination of the posteriors across all simulations should recover the prior.
+% Furthermore, the combination of the posteriors across all simulations should recover the prior.
 
-To estimate the posterior, I ran a single MCMC chain per simulated data set for 500,000 iterations, discarding the first 300,000 iterations as burn-in, and thinning every 100 iterations.
-Therefore, posterior calculations are based on 2,000 samples per data set.
-On some occasions, the proposal distribution's covariance matrix became singular (or close to singular).
-\todo{Update with fitting issues if they reoccur on final runs}
-This means it is not possible to propose a new value.
-In these cases, I restarted the chain with a new seed, discarding all information from the original chain; this was required at most once per chain.
-
-To check that the inference was working correctly, I calculated the coverage of the 50\%, 75\%, and 95\% central credible intervals for each parameter and simulation.
+To check the performance of the inference procedure, I calculated the coverage of the 50\%, 75\%, and 95\% central credible intervals for each parameter and simulation.
 I define coverage as the proportion of simulations where the true value of the parameter in that simulation was within the credible interval.
+This procedure is a simplified version of simulation-based calibration~\autocites{cookValidation}{taltsValidating}.
 The coverage for each parameter should equal the nominal probability (\eg the coverage should be 50\% for the 50\% credible intervals).
@@
 
-Overall, coverage was\todo{comment on sim study coverage}
+The majority of the confidence intervals contain the nominal probability (see \cref{SEIR:fig:sim-coverage}).
+There is a tendency for the coverage to be slightly lower than the nominal probability, suggesting that the credible intervals are slightly too narrow; this is a common problem when using random-walk MCMC.
+\todo{cite issues with random-walk MCMC exploring space}
+This is similar when considering coverage of the derived quantities incidence and prevalence (see \cref{SEIR:fig:sim-inc-prev}).
 \begin{figure}
@@
     \caption[Coverage of simulation study]{%
-        Coverage of the 50\%, 75\%, and 95\% central credible intervals for each parameter.
-        The lines shows the nominal coverage.
+        Point estimate and 95\% confidence interval for coverage of the 50\%, 75\%, and 95\% central credible intervals for each parameter; see text for full definition.
+        Vertical lines show the nominal coverage.
     }
@@
 \end{figure}
+\begin{figure}
+    \includegraphics[width=0.9\paperwidth]{SEIR/sim/predictive_coverage}
+    \caption[Coverage of simulation study (derived quantities)]{%
+        Point estimate and 95\% confidence interval for coverage of the 50\%, 75\%, and 95\% central credible intervals for incidence (top) and prevalence (bottom); see text for full definition.
+        Horizontal lines show the nominal coverage.
+    }
+    \label{SEIR:fig:sim-inc-prev}
+\end{figure}
 
-\todo[inline]{add simulation study results}
+In this setup, the parameters $i^+$, $q_c$, and $\psi$ are fully recovered; the other parameters are only partially so: their posterior has moved away from the prior but is still regularised by it (see \cref{SEIR:fig:true-vs-posterior}).
+The ability to identify $q_c$ is an advantage of using data which includes measurements on children, unlike data on hospitalisations or deaths.
+This suggests that these parameters are not practically identifiable in this setting.
+The posteriors align very closely with the priors (not shown) which reinforces this finding.
+The lack of identifiability for $\vec{\pi}$ aligns with previous work showing this parameter is structurally unidentifiable, \ie not identifiable even in the limit of infinite data~\autocite{dankwaStructural}.
+\begin{figure}
+    \includegraphics{SEIR/sim/true_vs_posterior}
+    \caption[True vs posterior parameter values]{%
+        True vs posterior median parameter values for each parameter in each simulation.
+        The black line shows the line $y = x$ which the dots would lie on if the parameter is fully recovered.
+        The blue line is a linear regression line.
+        For $i^+$, $q_c$, and $\psi$, the posterior recovers the true values.
+        For $\vec{\pi}$ and $\rho$, the posterior median is very similar regardless of the true value.
+    }
+    \label{SEIR:fig:true-vs-posterior}
+\end{figure}
 
@@
 
-\todo[inline]{add Rt results}
-
-\todo[inline]{add attack rates and compare with an external data source}
 \begin{figure}
@@
         Posterior estimates $\beta_t$ for each week.
-        Shown are "violins" where the width is proportional to the kernel-density estimated posterior density at that point.
+        Shown are ``violins'' where the width is proportional to the kernel-density estimated posterior density at that point.
         Values trend downwards until early December when an increase starts.
@@
 
-All regions' MCMC chains converged, assessed using the Rhat statistic and ESS.
-Rhat for all parameters is < 1.02, and for all parameters except $\sigma_\epsilon$ (a nuisance parameter), less than 1.01.
-All parameters had ESS > 500, except $\sigma_\epsilon$ in the East Midlands, which had ESS of 334.
+One MCMC chain for the North West region appeared to get stuck, with very few moves after the first 100,000 iterations.
+I reran this chain with a new seed.
+Following this, all regions' MCMC chains converged, assessed using the Rhat statistic and ESS.
+Rhat for all parameters is < 1.01, except $\sigma_\epsilon$ in London which had a Rhat of 1.018.
+All parameters had ESS > 1000, except $\sigma_\epsilon$ in London and the East, which had ESS of 197 and 757 respectively.
+As all the parameters of primary interest had high Rhat and ESS, and no parameters had large issues, the inference procedure is considered to have converged.
+\begin{figure}
+    % \includegraphics{SEIR/CIS/attack_rates}
+    \caption{Posterior estimates of the attack rate in each region and age group.}
+    \todo[inline]{Add and write about attack rates}
+\end{figure}
 
@@
 The parameters $\beta_t$ absorb all the time-varying factors not captured by Google mobility and school attendance data.
-\todo{discuss the results and implications for policy or understanding transmission}
 Many factors contribute to the pattern in the $\beta_t$ values, which have a minima near the start of December.
 While this analysis cannot establish causation, several hypotheses can be identified that are worth further investigation.
-The decreases in $\beta_t$ over the autumn could be due to increased caution(\eg masking or stricter adherence to social distancing) and/or interventions short of a lockdown as cases and deaths increased~\autocite{jarvisEffect}, or the impact of local measures in the highest transmission areas\todo{cite local measures}.
+The decreases in $\beta_t$ over the autumn could be due to increased caution(\eg masking or stricter adherence to social distancing) and/or interventions short of a lockdown as cases and deaths increased~\autocite{jarvisEffect}, or the impact of local measures in the highest transmission areas~\autocite{scottCovid19}.
+Local measures are not well captured by $\matr{\kappa}$ because the matrix was for the whole of England.
+Allowing each region to have its own $\matr{\kappa}$ may capture the effect of local measures more accurately.
 The increases in $\beta_t$ in December coincide with the rise of the Alpha variant~\autocite{lythgoeLineage}, known to be more transmissible~\autocite{daviesEstimated}.
@@
 
-These factors, and many others, are likely to be correlated across the country.
-This can explain that similar patterns emerge for $\beta_t$ in different regions (see \cref{SEIR:fig:beta-walk}), despite these not being estimated independently.
-These correlations could be exploited by linking their values across regions.
-This could improve the precision of the estimates, but would require a more complex model.
-
-\todo[inline]{discuss comparisons with the backcalculation}
+These factors are correlated across the country.
+This can explain that similar patterns emerge for $\beta_t$ in different regions (see \cref{SEIR:fig:beta-walk}), despite these being estimated independently (\ie there is no model structure which favours correlated values).
+These correlations could be exploited to improve the precision of estimates with a more complex model which borrows information across regions.
 
@@
 $\matr{\kappa}$ only captures changes in in-school transmission through changes in school attendance, not incorporating the probability of transmission, which could be modelled as changes in $q_{aa}$ for $a$ of school age.
-Similarly, there is evidence that $q_c$ is different for Alpha and pre-Alpha variants\todo{cite a systematic review}, which could again motivate these changing over time.
+Similarly, there is evidence that $q_c$ is different for Alpha and pre-Alpha variants~\autocite{zhuRole}, which could again motivate these changing over time.
 
@@
 
-\todo[inline]{move the following discussion to conclusion chapter as relevant to whole thesis}
-Household structure is not included in this model.
-The household-based design of the survey is likely to induce clustering in results, increasing the variance of the measured prevalence.
-The beta-binomial likelihood structure should capture some of this variance within each strata\todo{ref beta-binomial as a model for clustering}, but not between strata.
+\todo[inline]{discuss comparisons with the backcalculation}
 

### 2023-12-04 thesis.tex
Mostly fix bibliography
Updates from Overleaf
Finish sim results first draft

@@
 \makeatother
-\usepackage[backend=biber,style=authoryear,uniquename=false,sortcites=true]{biblatex}
+\usepackage[backend=biber,style=authoryear,uniquename=false,sortcites=true,maxcitenames=2,sorting=ynt,style=apa]{biblatex}
 \addbibresource{references.bib}
+% Display a linked DOI
+\DeclareFieldFormat{doi}{%
+  \mkbibacro{DOI}\addcolon\space
+  \ifhyperref
+    {\href{https://doi.org/#1}{\nolinkurl{#1}}}
+    {\nolinkurl{#1}}%
+}
 
@@
 \abstract{%
-    TODO
+    \todo[inline]{Write Abstract}
 }
@@
 \acknowledgements{%
-  TODO
+    \todo[inline]{Write acknowledgements}
 }
@@
 \chapter{Conclusion}
+Household structure is not included in this model.
+The household-based design of the survey is likely to induce clustering in results, increasing the variance of the measured prevalence.
+The beta-binomial likelihood structure should capture some of this variance within each strata\todo{ref beta-binomial as a model for clustering}, but not between strata.
 

### 2023-12-04 SEIR/CIS/random_walk.R

@@ -10,7 +10,7 @@ source(here::here("utils.R"))
 start_date = ymd("2020-08-31")
 
 posterior = readr::read_csv(here::here("SEIR/CIS/params.csv")) |>
-    filter(stringr::str_starts(parameter, "beta")) |>
+    filter(stringr::str_starts(parameter, "beta"), iteration >= 500e3) |>
     extract(
         parameter,
         into = c("parameter", "i"), 
@@ -35,7 +35,7 @@ posterior = readr::read_csv(here::here("SEIR/CIS/params.csv")) |>
         value = if_else(
             i == 0,
             0,
-            value * value[i == 0]
+            value * exp(value[i == 0])
         ),
     )
 
@@ -56,34 +56,41 @@ walk_summary = walk |>
 walk_summary |>
     mutate(date = start_date + i * 7) |>
     ggplot(aes(date, beta, color = region)) +
+    geom_point() +
     geom_errorbar(aes(ymin = beta.lower, ymax = beta.upper), alpha = 0.5) +
     facet_wrap(~region) +
     standard_plot_theming() +
+    geom_hline(yintercept = 1, linetype = "dashed", alpha = 0.5) +
     scale_y_log10(
-        breaks = c(0.01, 0.1, 1, 10, 100)
+        breaks = c(0.5, 1, 2, 0.67, 1.5),
+        limits = c(0.5, NA),
+        minor_breaks = numeric()
     )
 
 
 # A violin plot of the posterior distribution of beta for each region and week.
-# The x-axis is "continuous" so that ggplot2 does something sensible with the
-# tick labels but "discrete" for the violins so that there is a separate violin
-# per week.
 p_walk = walk |>
     mutate(date = start_date + i * 7) |>
     ggplot(aes(factor(date), beta)) +
-    # geom_violin(aes(factor(date)))
-    # stat_eye(point_size = 0.1) +
     stat_slab(side = "both") +
     facet_wrap(~region, ncol = 2) +
     standard_plot_theming() +
-    # scale_y_log10() +
+    scale_y_log10(
+        breaks = c(0.5, 1, 2, 0.67, 1.5),
+        limits = c(0.5, NA),
+        minor_breaks = numeric()
+    ) +
     # tick labels every 3 weeks
     scale_x_discrete(
         breaks = function(x) x[c(TRUE, rep(FALSE, 3))]
     ) +
     geom_hline(yintercept = 1, linetype = "dashed", alpha = 0.5) +
[…]

### 2023-12-04 SEIR/sim/coverage.R

@@ -90,16 +90,17 @@ p_true_vs_posterior = posterior_intervals |>
     filter(.width == 0.95, !str_detect(parameter, "beta")) |>
     # calculate bias by joining with true values
     left_join(true_params, by = c("sim", "parameter")) |>
-    ggplot(aes(true, value)) +
-    facet_wrap(~ parameter, scales = "free") +
+    ggplot(aes(value, true)) +
+    # facet_wrap(~ parameter, scales = "free", labeller = label_parsed) +
+    facet_wrap(~ parameter, scales = "free", labeller = parameter_labeller) +
     geom_abline(slope = 1, intercept = 0) +
     geom_smooth(method = "lm", formula = y ~ x) +
     standard_plot_theming() +
     geom_point() +
-    geom_errorbar(aes(ymin = .lower, ymax = .upper), width = 0) +
+    # geom_errorbar(aes(ymin = .lower, ymax = .upper), width = 0) +
     labs(
-        y = "Posterior median",
-        x = "True value"
+        x = "Posterior median",
+        y = "True value"
     )
 ggsave(
     filename = here::here("SEIR/sim/true_vs_posterior.png"),
@@ -110,30 +111,30 @@ ggsave(
     dpi = 300
 )
 
-# posterior_samples |>
-#     filter(!str_detect(parameter, "beta"), sim != 59, iteration > 500e3) |>
-#     # calculate bias by joining with true values
-#     ggplot() +
-#     geom_histogram(aes(value, after_stat(density), colour = "Combined posterior", fill = "Combined posterior"), bins = 100) +
-#     geom_density(aes(true, colour = "Prior"), data = filter(true_params, !str_detect(parameter, "beta"))) +
-#     facet_wrap(~parameter, scales = "free")
-#     geom_abline(slope = 1, intercept = 0) +
-#     geom_smooth(method = "lm", formula = y ~ x) +
-#     standard_plot_theming() +
-#     geom_point() +
-#     labs(
-#         x = "Posterior median",
-#         y = "True value"
-#     )
+# Figure for combined posterior
+posterior_samples |>
+    filter(!stringr::str_detect(parameter, "beta"), iteration > 500e3) |>
+    # calculate bias by joining with true values
+    ggplot() +
+    geom_histogram(aes(value, after_stat(density), colour = "Combined posterior", fill = "Combined posterior"), bins = 100) +
[…]

### 2023-12-04 SEIR/sim/predictive_check.R

@@ -1,0 +1,76 @@
+library(dplyr)
+library(ggplot2)
+library(lubridate)
+library(patchwork)
+library(tidybayes)
+library(tidyr)
+source(here::here("utils.R"))
+
+truth = readr::read_csv(here::here("SEIR/sim/sim_output.csv"))
+posterior_predictive = readr::read_csv(here::here("SEIR/sim/posteriors_predictive.csv"))
+
+make_plot = function(x) {
+    x |>
+        ggplot(aes(day, mean, colour = factor(.width), fill = factor(.width))) +
+        geom_lineribbon(aes(ymin = lower, ymax = upper), alpha = 0.5) +
+        geom_hline(aes(yintercept = .width)) +
+        standard_plot_theming() +
+        facet_wrap(~age) +
+        labs(
+            x = "Day",
+            y = "Coverage",
+            fill = "Interval width",
+            colour = "Interval width"
+        )
+}
+
+posterior_intervals = posterior_predictive |>
+    rename(.iteration = iteration) |>
+    group_by(sim, day, age) |>
+    median_qi(.width = c(0.5, 0.75, 0.95)) |>
+    ungroup() |>
+    left_join(
+        truth, by = c("sim", "day" = "date", "age"),
+        suffix = c(".pred", ".true")
+    )
+
+incidence_coverage = posterior_intervals |>
+    mutate(
+        contains = incidence.lower <= incidence.true
+            & incidence.true <= incidence.upper,
+    ) |>
+    summarise(
+        x = sum(contains),
+        n = n(),
+        .by = c(day, age, .width)
+    ) |>
+    mutate(binom::binom.confint(x, n, methods = "exact"))
+
+prevalence_coverage = posterior_intervals |>
+    mutate(
+        contains = prevalence.lower <= prevalence.true
+            & prevalence.true <= prevalence.upper,
+    ) |>
+    summarise(
+        x = sum(contains),
+        n = n(),
+        .by = c(day, age, .width)
+    ) |>
+    mutate(binom::binom.confint(x, n, methods = "exact"))
+
+p_incidence = incidence_coverage |>
+    make_plot()
+p_prevalence = prevalence_coverage |>
+    make_plot()
+p_combined = p_incidence / p_prevalence +
+    # Combine legends and put at bottom of plot
+    plot_layout(guides = "collect") &
+    theme(
+        legend.position = "bottom",
+    )
+ggsave(
[…]

### 2023-12-04 ATACCC-appendix-original-analysis.tex
Updates from Overleaf

@@
     \makebox[\textwidth][c]{\includegraphics[scale=0.8]{ATACCC/appendix/Bayesian_piecewise_DAG}}
-    \caption{Directed acyclic graph showing the relationship between parameters in the model.
+    \caption[DAG showing hiearchical structure]{Directed acyclic graph showing the relationship between parameters in the model.
 Dotted lines represent deterministic (functional) relationships and solid lines stochastic (distributional) relationships.
@@
 \end{tabular}
-\caption{Priors used, columns 3 to 5 gives quantiles of the distributions specified. See main text for description of the parameters.}
+\caption[Priors]{Priors used, columns 3 to 5 gives quantiles of the distributions specified. See main text for description of the parameters.}
 \label{tab:paper:priors}
@@
     \makebox[\textwidth][c]{\includegraphics[scale=0.9]{ATACCC/appendix/goodness_of_fit}}
-    \caption{Predictive median (solid line), 50\% and 95\% predictive intervals (light and dark shading respectively) for test results (ignoring false negatives). Points show actual observations.}
+    \caption[Goodness-of-fit]{Predictive median (solid line), 50\% and 95\% predictive intervals (light and dark shading respectively) for test results (ignoring false negatives).
+Points show actual observations.}
     \label{fig:paper:goodness_of_fit}
@@
     \makebox[\textwidth][c]{\includegraphics{ATACCC/appendix/duration}}
-    \caption{Proportion of people with $\hat{C} > C_\text{lod}$ a given number of days after this first occurred. Shading indicates central 95\% credible interval.}
+    \caption[Proportion of population positive]{Proportion of people with $\hat{C} > C_\text{lod}$ a given number of days after this first occurred. Shading indicates central 95\% credible interval.}
     \label{fig:paper:duration}
@@
     \end{tabular}}
-    \caption{Summary of results in comparison to previous studies. \Textcite{jonesEstimating} do not give values in terms of Ct.}
+    \caption[Comparison to previous studies.]{Summary of results in comparison to previous studies. \Textcite{jonesEstimating} do not give values in terms of Ct.}
     \label{tab:paper:compare}

### 2023-12-04 thesis.tex
Updates from Overleaf

@@
 \chapter{Conclusion}
-Household structure is not included in this model.
-The household-based design of the survey is likely to induce clustering in results, increasing the variance of the measured prevalence.
-The beta-binomial likelihood structure should capture some of this variance within each strata\todo{ref beta-binomial as a model for clustering}, but not between strata.
 
@@
 
+\chapter{CIS episode definitions} \label{episode-def}
+
 \subfile{ATACCC-appendix-original-analysis}
-
-\chapter{CIS episode definitions} \label{episode-def}
 

### 2023-12-04 SEIR.tex
Attack rates figure

@@
 As all the parameters of primary interest had high Rhat and ESS, and no parameters had large issues, the inference procedure is considered to have converged.
+
+The \emph{attack rate} is the proportion of the population that have been infected at least once\todo{cite attack rate}.
+In this model it is $1 - \vec{s}$.
+The attack rate is highest in the $[11, 16)$ age group in the North West at 50\% (95\% CrI: 48\%--52\%) (see \cref{SEIR:fig:attack-rates}).
 \begin{figure}
-    % \includegraphics{SEIR/CIS/attack_rates}
-    \caption{Posterior estimates of the attack rate in each region and age group.}
-    \todo[inline]{Add and write about attack rates}
+    \includegraphics{SEIR/CIS/attack_rates}
+    \caption{Posterior median and 95\% CrI of the attack rate in each region and age group.}
+    \label{SEIR:fig:attack-rates}
 \end{figure}

### 2023-12-04 SEIR.tex
Updates from Overleaf

--- 2023-12-04 SEIR.tex
@@
 \begin{figure}
-    \includegraphics[width=0.9\paperwidth]{SEIR/sim/predictive_coverage}
+    \thisfloatpagestyle{empty}
+    \vspace{-3cm}
+    \makebox[\textwidth][c]{\includegraphics[width=0.9\paperwidth]{SEIR/sim/predictive_coverage}}
     \caption[Coverage of simulation study (derived quantities)]{%
@@
 A constant $\matr{\beta}(t)$ leads to exponential growth over short time periods, once the initial dynamics have died out.
-When the peak in incidence is estimated to occur varied by region, but due to this reason, is almost always on a Friday (see \cref{SEIR:fig:peak-incidence}).
+When the peak in incidence is estimated to occur varied by region, but due to this reason, is almost always on a Sunday (see \cref{SEIR:fig:peak-incidence}).
+For Yorkshire, the peak is estimated to occur in late October.
+For all other regions, it is estimated to occur in late December or Early January, the period during which severe social distancing measures were applied.
 \begin{figure}
@@
 
-\begin{figure}
-    \includegraphics{SEIR/CIS/p_peak}
-    \caption[Posterior estimates of peak incidence timing]{%
-        Posterior probability of the peak incidence for each region and age combination occurring on each day.
-    }
-    \label{SEIR:fig:peak-incidence}
-\end{figure}
+% \begin{figure}
+%     \includegraphics{SEIR/CIS/p_peak}
+%     \caption[Posterior estimates of peak incidence timing]{%
+%         Posterior probability of the peak incidence for each region and age combination occurring on each day.
+%     }
+%     \label{SEIR:fig:peak-incidence}
+% \end{figure}
 

### 2023-12-04 SEIR/CIS/random_walk.R
Fix region names

@@ -6,6 +6,7 @@ library(lubridate)
 library(tidybayes)
 library(tidyr)
 source(here::here("utils.R"))
+source(here::here("SEIR/utils.R"))
 
 start_date = ymd("2020-08-31")
 
@@ -17,7 +18,7 @@ posterior = readr::read_csv(here::here("SEIR/CIS/params.csv")) |>
         regex = "^([^\\[]+)(\\[[0-9]+\\])?$"
     ) |>
     transmute(
-        region,
+        region = region_labels(region),
         .chain = factor(chain + 1),
         .iteration = iteration + 1,
         .draw = max(.iteration) * (as.integer(.chain) - 1) + .iteration,
@@ -93,7 +94,7 @@ p_walk = walk |>
     )
 ggsave(
     plot = p_walk,
-    filename = here::here("SEIR/CIS/beta_walk.png"),
+    filename = here::here("SEIR/CIS/beta_walk.pdf"),
     width = 15,
     height = 20,
     units = "cm",

### 2023-12-04 ATACCC-appendix-original-analysis.tex
Concrete ATACCC analyses

@@
 \chapter{Original ATACCC analysis} \label{ATACCC:sec:original-analysis}
-
-\todo[inline]{Check for "REF" in this appendix}
 
@@
 
-Inference was performed by Hamiltonian Monte Carlo (HMC) implemented in Stan 2.26.1~[REF].
-To aid with mixing, the hierarchical parameters were implemented using the non-centred parameterisation~[REF].
+Inference was performed by Hamiltonian Monte Carlo (HMC) implemented in Stan 2.26.1~\autocite{Stan-2-32-2}.
+To aid with mixing, the hierarchical parameters were implemented using the non-centred parameterisation~\autocites{papaspiliopoulosGeneral,stanReparameterization}.
 In addition, $t_{i,0}$ was reparameterised in terms of the absolute time of peak.

### 2023-12-04 ATACCC.tex
Concrete ATACCC analyses

@@
 
-Several features of the \textcite{hakkiOnset} model can be motivated by the desire to apply Stan, a modern and convenient MCMC implementation of the NUTS algorithm (see\todo{ref wherever I end up describing NUTS}).
+Several features of the \textcite{hakkiOnset} model can be motivated by the desire to apply Stan, a modern and convenient MCMC implementation of the NUTS algorithm (see \cref{E-backcalc:sec:MCMC}).
 The first of these is using a continuous viral load trajectory rather than the more common piecewise linear model.
@@
 A positive PCR test has a Ct value.
-Ct values are linearly related to viral load (see\todo{ref intro section}).
+Ct values are linearly related to viral load (see\todo{make sure relationship between Ct values and viral load is introduced previously}).
 A negative PCR test signals either the viral load is too low to be detected, \ie below the limit of detection, or the test is a false negative.
@@
 Here, $D_i$ is drawn conditional on the population-level parameters $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$.
-Equivalent forms of $F_D$, such as the survival and hazard functions, can be calculated using the relations in\todo{ref section defining survival functions}.
+Equivalent forms of $F_D$, such as the survival and hazard functions, can be calculated using the relations in \cref{E-backcalc:sec:definitions}.
 I found 100,000 draws produced stable estimates of the tail of $F_D$.
@@
 \ifSubfilesClassLoaded{
-  \listoftodos
   \appendix

### 2023-12-05 cis-imperfect-testing.tex
Minor edits

@@
 Furthermore, due to missing short episodes, there is little or no ability to upper bound the number of very short episodes because these would almost all be missed.
-\Textcite{shenNonparametrica} (extending \textcite{panNote}) showed that, in a similar setting to here\todo{note why it is different}, the maximum likelihood (frequentist) estimator is inconsistent and can lead to severe underestimation (bias) of the survival function.
+\Textcite{shenNonparametrica}, extending \textcite{panNote}, showed that the maximum likelihood estimator is inconsistent and can lead to a severely biased underestimation of the survival function, albeit when only the terminating event is interval censored\todo{double check the setting of Shen}.
 While the result cannot simply be extended to the more complex study design of the CIS, it reinforces the intuition that single positive episodes are likely to be problematic.
-Furthermore, simulations (not shown) building in even a few excess single positive episodes led to problematic estimation even within the CIS setting.
+Furthermore, simulations building in even a few excess single positive episodes led to problematic estimation even within the CIS setting (not shown).
 
@@
 There are two strands of evidence supporting this hypothesis.
-First, it is well known that PCR testing can return false negatives  (see \cref{E-biology-data:sec:PCR}); they were explicitly include these in the viral load model of \cref{E-ATACCC}.
+First, it is well known that PCR testing can return false negatives  (see \cref{E-biology-data:sec:PCR}); I explicitly include them in the viral load model of \cref{E-ATACCC}.
 Intermittent negatives (described in  \cref{E-biology-data:sec:PCR}) show that they occur within the CIS.
@@
 False negatives are more likely to occur when the viral load of an individual is low, because there is less virus in their body to be sampled.
-In \cref{E-ATACCC} this mechanism is incorporated by the left-censored observations.
+In \cref{E-ATACCC}, this mechanism is incorporated by the left-censored observations.
 This suggests that a model with a declining test sensitivity as a function of time since infection might be more suitable.
 
-\todo[inline]{Next paragraph needs editing for clarity}
 The CIS data provides evidence for a test sensitivity declining over the course of an episode.
-To do this, I classify all test results into three classes, making the following assumptions.
+\Cref{imperf-test:fig:bounding-cis-sensitivity} demonstrates this by classifying all test results as follows.
 \begin{enumerate}
@@
 % In \cref{imperf-test:fig:bounding-cis-sensitivity} we consider the values produced by assuming the negative following the last positive in an episode could be either a true or false negative as a function of time since the individual was detected (\ie: the first positive test in the episode).
-This generates broad bounds, although suggest a declining test sensitivity over the course of an episode (see \cref{imperf-test:fig:bounding-cis-sensitivity}).
+This generates broad bounds, although suggest a declining test sensitivity over the course of an episode.
 \begin{figure}
@@
 
-With false negatives, an episode with can be undetected in the following
-ways:
-
+With false negatives, an episode with can be undetected in the following four ways; the first two of these are the same as before, but the second two are new.
+\todo[inline]{Check all this notation with the previous chapter's}
 \begin{enumerate}
 \item
-  The episode begins before $\min(t_i)$ (assuming there is a
-  negligible probability of a false negative test at this point).
+  The episode begins before $\min(t_i)$.
+  I assume there is a negligible probability of a false negative test at this point because viral load is high.
 \item
-  The duration of the episode is less than $t_{ib_i}^N$. These first
-  two mechanisms are why episodes were missed in the previous chapter, with combined
-  probability $p_{it}$.
+  The duration of the episode is less than $t_{ib_i}^N$.
 \item
-  The duration of the episode is at least as long as $t_{it}^N$ but
-  less than the time to the second test after $t$, $t_{it}^{2N}$,
-  and a false negative episode occurred at the first test after $t$.
+  The duration of the episode is at least as long as $t_{it}^N$ but less than the time to the second test after $t$, $t_{it}^{2N}$, and a false negative episode occurred at the first test after $t$.
   This occurs with probability:
@@
 \item
-  The duration of the episode is at least as long as $t_{it}^{2N}$,
-  and all the tests within the episode (at least two) are false
-  negatives. We assume this occurs with negligible probability.
+  The duration of the episode is at least as long as $t_{it}^{2N}$, and all the tests within the episode (at least two) are false
+negatives.
+  I assume this occurs with negligible probability.
 \end{enumerate}
@@
 &= 1 - p_{it} - (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(t_i)}^T \left( S_\theta(t_{ib}^N + 1) - S_\theta(t_{ib}^{2N} + 1)\right) \\
-&= \frac{1}{T} \sum_{b=\min(t_i)}^T \left( p_\text{sens} S_\theta(t_{ib}^N + 1) + (1 - p_\text{sens}) S_\theta(t_{ib}^{2N} + 1)\right) \\
+&= \frac{1}{T} \sum_{b=\min(t_i)}^T \left( p_\text{sens} S_\theta(t_{ib}^N + 1) + (1 - p_\text{sens}) S_\theta(t_{ib}^{2N} + 1)\right).
 \end{align}
+
+$p_{it}'$ and $p_{ia}'$ are substituted for $p_{it}$ and $p_{ia}$ respectively in \cref{E-perf-test:eq:full-posterior}.
+The posterior density is otherwise unchanged.
 
@@
 
-\todo[inline]{notes on sim results into prose}
+\todo[inline]{Turn notes on sim results into prose}
 

### 2023-12-06 cis-imperfect-testing.tex
Most simulation results for imperfect testing

@@
 &= \left( \prod_{t \in t_i'} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right)\sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - p_\text{sens} S_\theta(r_i^{(e)} - b_i - 1) \right) p(b_i \mid p_\text{sens}, \theta).
+\label{imperf-test:eq:pia-prime}
 \end{align}
@@
 &= \frac{1}{T} \sum_{b=\min(t_i)}^T \left( p_\text{sens} S_\theta(t_{ib}^N + 1) + (1 - p_\text{sens}) S_\theta(t_{ib}^{2N} + 1)\right).
+\label{imperf-test:eq:pit-prime}
 \end{align}
@@
 
-\section{Simulation study results} \label{imperf-test:sec:sim-study-results}
+\section{Simulation study} \label{imperf-test:sec:sim-study-results}
 
-\todo[inline]{Turn notes on sim results into prose}
+I repeated the simulation study in \cref{E-perf-test:sec:sim-study}, except I used the model of \cref{imperf-test:sec:modelling} with either a constant test sensitivity or the more realistic varying sensitivity.
+I focus on the independent (vague) and model combination priors for the survival (described in \cref{E-perf-test:sec:parameters-priors}.
+These were shown to be the best performing priors in \cref{E-perf-test:sec:results}
 
-\begin{itemize}
-  \item When sensitivity is too bad (0.6), cannot recover the true distribution. The assumptions in the model, that the first test is a true positive and no missed infections prior, cannot hold.
-  \item Assuming too low a test sensitivity (bottom left triangle) means that number of missed infections will be overestimated and the duration distribution will be overestimated (\ie episodes are too long on average).
-\end{itemize}
+With a constant test sensitivity of 0.8, the model recovers the true survival time well.
+However, with a lower test sensitivity of 0.6, this is no longer the case (see \cref{imperf-test:fig:constant-test-sensitivity}).
+The simplifications to the model made in \cref{imperf-test:sec:modelling} cause too large a misspecification at this point.
+The informative prior helps to overcome the misspecification, helping to recover the true survival time (in comparison to the vague prior).
+\begin{figure}
+  % \makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{cis-imperfect-testing/sim-constant-sensitivity}}
+  \includegraphics[width=\textwidth]{cis-imperfect-testing/sim-constant-sensitivity}
+  \caption[Simulation study results with constant test sensitivity]{%
+    Posterior (median and 95\% credible interval) of the survival time for the simulation study with a constant test sensitivity.
+    True survival time shown in black.
+    Left panel shows the results with a test sensitivity of 0.6, right panel shows the results with a test sensitivity of 0.8.
+  }
+  \label{imperf-test:fig:constant-test-sensitivity}
+\end{figure}
+
+If the test sensitivity is misspecified, that is the assumed value for $p_\text{sens}$ in \cref{imperf-test:eq:pit-prime,imperf-test:eq:pit-prime} is different to the value used in the simulation, then the survival time will be biased (see \cref{imperf-test:fig:misspecified-test-sensitivity}).
+If the test sensitivity is assumed to be too low, then the number of episodes inferred to have truly ended by the first negative will be too high, and the survival time will be underestimated.
+This affect dominates over the opposing bias of overestimating the number of missed episodes.
+The opposite occurs if the test sensitivity is assumed to be too high.
+\begin{figure}
+  % \makebox[\textwidth][c]{
+    \includegraphics[width=\textwidth]{cis-imperfect-testing/sim-misspecified-sensitivity}
+  \caption[Simulation study results with misspecified test sensitivity]{%
+    Posterior (median and 95\% credible interval) of the survival time for the simulation study with a misspecified test sensitivity.
+    True survival time shown in black.
+    All simulations use a constant test sensitivity of 0.8, but the model assumes different values.
+    All inference uses the combination prior, due to it performing better than the vague prior in \cref{imperf-test:fig:constant-test-sensitivity}.
+    The correctly specified model (middle pane) has a posterior median closely following the true value, with the credible intervals almost always including the true value.
+    Assuming a too low test sensitivity (left pane) starts by following the true value but then separates, with the posterior median being too high.
+    Assuming a too high (perfect) test sensitivity (right pane) underestimates the survival time almost throughout.
+  }
+  \label{imperf-test:fig:misspecified-test-sensitivity}
+\end{figure}
 
@@
 
-\subsection{Data}
+\subsection{Methods}
 
@@
 
-\todo[inline]{Descriptive analyses of the data here}
-
-\subsection{Methods}
-
-\todo[inline]{Add additional info on methods. Prior on number missed; hyperparameters}
-Might include hyperparameters etc or assumptions made?
+I form a prior for $\ntot$ from an extension of the model of \textcite{birrellRealtime}, used to produce incidence estimates.
+This gives a posterior mean of \numprint{4136368} cumulative infections in England in the time period I consider, with a posterior standard deviation of \numprint{27932}~\citePersonalComms{Paul Birrell}.
+Approximating this distribution as a negative binomial and scaling the mean to the size of the CIS cohort gives the prior $\ntot \sim \NBc(\numprint{25132}, \numprint{22047})$.
 
@@
 \begin{itemize}
-    \item not sensitivite to survival prior
+    \item not sensitive to survival prior
     \item sensitive to test sensitivity assumptions and prior on number missed
@@
   \item Survival prior matters for quantitative details but not qualitative shape. 
-  \item Number of missed infections prior is more important.
+  \item Number of missed infections prior is more important, but not the case in simulation (not shown).
   \item Unclear why this is the case, probably model misspecification meaning that fewer missed infections than expected.

### 2023-12-06 cis-imperfect-testing/CIS_survival.R

@@ -1,0 +1,20 @@
+suppressMessages(library(dplyr))
+library(ggplot2)
+library(patchwork)
+library(purrr)
+library(tidybayes)
+library(tidyr)
+source(here::here("utils.R"))
+
+posterior_draws = readRDS(here::here("cis-imperfect-testing/STATS17701/draws.rds"))
+
+posterior_draws |>
+    # filter(time == 1, .draw == 1) |>
+    # count(
+    #     survival_prior,
+    #     sensitivity,
+    #     missed_model,
+    #     r
+    # )
+    filter(survival_prior == "Informative") |>
+

### 2023-12-06 cis-imperfect-testing/sim_survival.R

@@ -1,0 +1,67 @@
+suppressMessages(library(dplyr))
+library(ggplot2)
+library(patchwork)
+library(purrr)
+library(tidybayes)
+library(tidyr)
+source(here::here("utils.R"))
+
+tbl_posteriors = readRDS(here::here("cisRuns-output/all_posteriors.rds")) |>
+    filter(
+        .width == 0.95, # Credible intervals are narrow anyway so only show 95%
+        is.na(type),
+        survival_prior %in% c(
+            "ATACCC-new",
+            "vague"
+        ),
+        # Use the reference prior on the number of missed infections as this doesn't matter
+        missed_prior == "vague", 
+        missing_model == "total"
+    ) |>
+    mutate(
+        survival_prior = case_match(
+            survival_prior,
+            "ATACCC-new" ~ "Combination",
+            "vague" ~ "Independent (vague)"
+        ),
+    )
+truth = readRDS(here::here("cisRuns-output/input_curves.rds")) |>
+  filter(source == "Combined")
+
+p_constant_sensitivity = tbl_posteriors |>
+    filter(sensitivity.simulation == sensitivity.model, sensitivity.simulation < 1) |>
+    ggplot() +
+    ggdist::geom_lineribbon(aes(time, S, ymin = S.lower, ymax = S.upper, fill = survival_prior, colour = survival_prior), linewidth = 1, alpha = 0.3) +
+    facet_wrap(~ sensitivity.model) +
+    geom_line(aes(time, S), data = truth) +
+    theme_survival_time_series()
+ggsave(
+    filename = here::here("cis-imperfect-testing/sim-constant-sensitivity.png"),
+    plot = p_constant_sensitivity,
+    width = 15,
+    height = 9,
+    units = "cm",
+    dpi = 300
+)
+
+p_misspecified_sensitivity = tbl_posteriors |>
+    filter(sensitivity.simulation == 0.8, survival_prior == "Combination") |>
+    mutate(sensitivity.model = factor(sensitivity.model)) |>
+    ggplot() +
+    ggdist::geom_lineribbon(aes(time, S, ymin = S.lower, ymax = S.upper, fill = sensitivity.model, colour = sensitivity.model), linewidth = 1, alpha = 0.3) +
+    facet_wrap(~ sensitivity.model) +
+    geom_line(aes(time, S), data = truth, alpha = 0.5) +
+    theme_survival_time_series() +
+    labs(
[…]

### 2023-12-07 cis-imperfect-testing.tex
First version of full text on imperfect testing

@@
 Applying the method of \cref{E-perf-test} to CIS data gives implausible estimates of the duration distribution (see \cref{imperf-test:fig:problem-cis-estimates}).
-In particular, it estimates a very high probability of the episode lasting less than five days, in conflict with the results of \cref{E-ATACCC} and other prior work~\autocite[e.g.:][]{cevikShedding}.
+In particular, it estimated a much shorter duration of positivity than in \cref{E-ATACCC}, despite those estimates being much more reliable over the first 2--3 weeks.
 
@@
   \end{cases}
+  \label{imperf-test:eq:variable-test-sensitivity}
 \end{equation}
@@
 
-I repeated the simulation study in \cref{E-perf-test:sec:sim-study}, except I used the model of \cref{imperf-test:sec:modelling} with either a constant test sensitivity or the more realistic varying sensitivity.
+I repeated the simulation study in \cref{E-perf-test:sec:simulation-study}, except I used the model of \cref{imperf-test:sec:modelling} with either a constant test sensitivity or the more realistic varying sensitivity.
 I focus on the independent (vague) and model combination priors for the survival (described in \cref{E-perf-test:sec:parameters-priors}.
@@
 
+The results when a varying test sensitivity (\cref{imperf-test:eq:variable-test-sensitivity}) is used for the simulation are similar to a constant 0.8 test sensitivity (see \cref{imperf-test:fig:variable-test-sensitivity}).
+This suggests that the simplified model, with constant test sensitivity, is sufficient for recovering the true survival time.
+Therefore, I will apply this model to the real CIS data in the next section.
+\begin{figure}
+  % \makebox[\textwidth][c]{
+    \includegraphics[width=\textwidth]{cis-imperfect-testing/sim-variable-sensitivity}
+  \caption[Simulation study results with varying test sensitivity]{%
+    Posterior (median and 95\% credible interval) of the survival time for the simulation study with a variable test sensitivity.
+    All simulations use a constant test sensitivity of 0.8 and the combination prior (as in \cref{imperf-test:fig:misspecified-test-sensitivity}), but the model assumes different values.
+    The posterior estimate using a model with a constant test sensitivity of 0.8 (middle panel) is similar to the true value (black line).
+  }
+  \label{imperf-test:fig:variable-test-sensitivity}
+\end{figure}
+
 \section{Application to CIS data} \label{imperf-test:sec:application}
 
-\subsection{Methods}
+This section applies the model described in this chapter to the CIS episodes dataset I described in \cref{E-perf-test:sec:approach}.
 
-\todo[inline]{Check if I've talked about the data before - previous chapter?}
+Unlike in the simulation studies, an uninformative prior on $\ntot$ led to implausible estimates of the duration distribution.
+The uninformative prior led to high posterior estimates, and hence an implausibly large number of episodes with durations of less than five days.
+Therefore, I based an informative prior for $\ntot$ on pre-existing estimates of the total number of infections over the period with posterior mean \numprint{4136368} and standard deviation \numprint{27932}~\autocite{birrellRTM2}.
+% This model gives a posterior mean of \numprint{4136368} cumulative infections in England in the time period I consider, with a posterior standard deviation of \numprint{27932}~\citePersonalComms{Paul Birrell}.
+\todo{Insert correct citation for RTM paper 2 once available}
+Approximating this distribution as a negative binomial and scaling the mean to the size f the CIS cohort gives the prior $\ntot \sim \NBc(\numprint{25132}, \numprint{22047})$.
 
-I use the CIS data described in \cref{E-intro:sec:cis}, with positive tests heuristically divided into episodes.
-Prior work categorised a series of positive tests within the same individual CIS were heuristically categorised into episodes (see \cref{E-episode-def}).
+With this prior, the model produces plausible estimates of the duration distribution (see \cref{imperf-test:fig:cis-estimates}).
+This estimate has more long episodes than the estimate in \cref{E-ATACCC}.
+The qualitative increase in long episodes is robust to the choice of prior for $\ntot$, the assumed value for $\psens$ (see \cref{imperf-test:fig:cis-sensitivity}), and the choice of prior for $\lambda$.
+However, the survival proportion over the first 4 weeks is sensitive to these choices.
+The estimate using a test sensitivity of 0.8 and $\NBc(\numprint{25132}, \numprint{22047})$ give a median survival time most similar to that of \cref{E-ATACCC}.
+\todo[inline]{Insert fig showing final estimate and comparison to ATACCC}
+\todo[inline]{
+  Insert figs showing, for different assumption:
+  proportion of individuals still positive at day 30 or 50 (tail),
+  overall survival curve, and
+  median survival time (in comparison to ATACCC).
+}
 
-I filter the episodes as follows.
-\begin{itemize}
-\item
-  The first positive in the episode occurred between 16th Oct 2020 and
-  5th Dec 2020 inclusive.
-\item
-  The individual had recorded a negative test prior to their episode
-  beginning (this could have happened at any time, including prior to
-  16th Oct).
-\item
-  A negative test to end the episode (\ie: following the final positive in the episode) has been recorded.
-\end{itemize}
-
-I form a prior for $\ntot$ from an extension of the model of \textcite{birrellRealtime}, used to produce incidence estimates.
-This gives a posterior mean of \numprint{4136368} cumulative infections in England in the time period I consider, with a posterior standard deviation of \numprint{27932}~\citePersonalComms{Paul Birrell}.
-Approximating this distribution as a negative binomial and scaling the mean to the size of the CIS cohort gives the prior $\ntot \sim \NBc(\numprint{25132}, \numprint{22047})$.
-
-\subsection{Results}
-
-\todo[inline]{notes on real results into prose}
-
-\begin{itemize}
-    \item not sensitive to survival prior
-    \item sensitive to test sensitivity assumptions and prior on number missed
-    \item comparison with ATACCC
-    \item mean and median summaries
-    \item 
-\end{itemize}
+The assumption that the estimates are most sensitive to is the strength of the prior on $\ntot$.
+A very weak, almost uninformative, prior on this quantity causes the posterior estimate to be much higher than the estimate from \textcite{birrellRTM2}.
+When increasing the prior's strength, the posterior estimate moves towards the prior smoothly, as expected (see\todo{figure showing $r$ vs $\ntot$}).
+Taking a prior from \textcite{birrellRTM2} means that the posterior survival is close to the \cref{E-ATACCC} estimate in the region where that estimate is reliable.
+This justifies the use of this prior.
 
 \section{Discussion} \label{imperf-test:sec:discussion}
+
+This chapter shows that a simple model of false negatives can be incorporated into the model of \cref{E-perf-test}.
+This model allows estimation of the duration distribution tails without the strong model assumptions used in \cref{E-ATACCC}.
+The qualitative features of the estimated distribution are robust to the assumed value for $\psens$, and the choice of prior for $\lambda$, although the quantative details are not.
+The estimates of the tail, the primary purpose of this chapter, are in addition robust to the choice of prior for $\ntot$.
+However, the rest of the distribution is sensitive to this choice.
 
@@
 
+An informative prior on $\ntot$ is required is for the bulk of the distribution to be plausible.
+In the simulation study, the uninformative prior on $\ntot$ performed well, even when the test sensitivity was misspecified.
+One hypothesis is that the test sensitivity is significantly different to the function in \cref{imperf-test:eq:variable-test-sensitivity} causing an unintuitive pattern of results.
+In particular, the test sensitivity for long episodes may become much lower than 50\%.
+This allows many more episodes to be missed than the model assumes, and hence a higher $\ntot$ is required to explain the data.
+Furthermore, it violates assumptions made in \cref{imperf-test:sec:modelling}; this is similar to when the simulating using a too low test sensitivity.
+Violating these assumptions would lead to unpredictable inference results, perhaps those seen here.
+
+Ideally, the test sensitivity would be estimated from the data.
+However, this would require a model for time-varying test sensitivity to be incorporated into the model.
+Furthermore, it is unclear that such a model would be identifiable because it would not be clear whether to attribute negative results to false negatives or to recoveries\todo{cite something on this}.
+
+The assumption, made in \cref{E-perf-test}, was that the episode start times are independent.
+However, because infections are the result of a counting process with intensity shared between individuals (see \cref{E-inc-prev:sec:infection-process}), this assumption is violated.
+Prevalence was fairly constant over the period I chose, suggesting the same is true of the incidence.
+Additionally, some simulations exploring the results of changing incidence suggested this would not have a large effect on the results (not shown).
+
+% However, comparing the CIS and simulated datasets do not show any differences which suggest what in the data is causing these differences.
+% Then, the additional missed infections implied by a 
+% This would lead to more missed infections than the model assumes, and hence a higher $\ntot$ is required to explain the data.
+
+The estimates I give in this chapter are the first estimates of the duration of PCR-positivity, including the distribution tails, estimated from general population data.
+This feature of the distribution is important for understanding the proportion of prevalence attributable to long infections and interpreting PCR test results.
+Producing these estimates required applying previous survival analysis frameworks in the novel context of the CIS.
+The framework needed further modification to account for the presence of false negatives, done in this chapter.
+
 \section{Conclusion} \label{imperf-test:sec:conclusion}
 
-\todo[inline]{write conclusion}
+This chapter has produced estimates of the duration distribution of PCR-positivity using information from two studies: ATACCC and CIS.
+Assuming a test sensitivity of 0.8 and an informative prior on the number of infections, provided the preferred estimates.
+The posterior distribution for the survival is concentrated, producing narrow credible intervals.
+In the following chapters, I will take the posterior mean for the survival time at each point, neglecting this uncertainty, as the basis for estimating transmission.
+First, I will take a backcalculation approach (\cref{E-backcalc}), based on the framework in \cref{E-inc-prev}.
+Then, I will take a mechanistic approach (\cref{E-SEIR}), modelling the transmission process explicitly.
 

### 2023-12-07 cis-imperfect-testing/CIS_survival.R

@@ -9,12 +9,16 @@ source(here::here("utils.R"))
 posterior_draws = readRDS(here::here("cis-imperfect-testing/STATS17701/draws.rds"))
 
 posterior_draws |>
-    # filter(time == 1, .draw == 1) |>
-    # count(
-    #     survival_prior,
-    #     sensitivity,
-    #     missed_model,
-    #     r
-    # )
+    filter(sensitivity == 1)
+
+
+posterior_draws |>
+    filter(time == 1, .draw == 1, missed_model == "total") |>
+    count(
+        survival_prior,
+        sensitivity,
+        missed_model,
+        r
+    )
     filter(survival_prior == "Informative") |>
 

### 2023-12-07 cis-imperfect-testing/sim_survival.R

@@ -55,8 +55,7 @@ p_misspecified_sensitivity = tbl_posteriors |>
     labs(
         colour = expression(p[sens]),
         fill = expression(p[sens])
-    ) #+
-    scale_y_log10()
+    )
 ggsave(
     filename = here::here("cis-imperfect-testing/sim-misspecified-sensitivity.png"),
     plot = p_misspecified_sensitivity,
@@ -65,3 +64,52 @@ ggsave(
     units = "cm",
     dpi = 300
 )
+
+p_variable_sensitivity = tbl_posteriors |>
+    filter(is.na(sensitivity.simulation), survival_prior == "Combination") |>
+    mutate(sensitivity.model = factor(sensitivity.model)) |>
+    ggplot() +
+    ggdist::geom_lineribbon(aes(time, S, ymin = S.lower, ymax = S.upper, fill = sensitivity.model, colour = sensitivity.model), linewidth = 1, alpha = 0.3) +
+    facet_wrap(~ sensitivity.model) +
+    geom_line(aes(time, S), data = truth, alpha = 0.5) +
+    theme_survival_time_series() +
+    labs(
+        colour = expression(p[sens]),
+        fill = expression(p[sens])
+    )
+ggsave(
+    filename = here::here("cis-imperfect-testing/sim-variable-sensitivity.png"),
+    plot = p_variable_sensitivity,
+    width = 15,
+    height = 9,
+    units = "cm",
+    dpi = 300
+)
+
+# tbl_posteriors |>
+#     ggplot() +
+#     ggdist::geom_lineribbon(aes(time, S, ymin = S.lower, ymax = S.upper), linewidth = 1, alpha = 0.3) +
+#     facet_grid(model_name~sim_name) +
+#     geom_line(aes(time, S), data = truth, alpha = 0.5) +
+#     theme_survival_time_series() +
+#     labs(
+#         colour = expression(p[sens]),
+#         fill = expression(p[sens])
+#     )
+
+# tbl_posteriors |>
+#     count(sim_name, model_name, time) |>
+#     filter(n > 1)
+
+# ggsave(
+#     filename = here::here("cis-imperfect-testing/sim-all.png"),
+#     width = 15,
+#     height = 50,
+#     units = "cm",
+#     dpi = 300
+# )
+
+# tbl_posteriors |>
+#     filter(time == 5, S < 0.8, !stringr::str_detect(model_name, "inform-total")) |>
+#     arrange(S) |>
+#     select(model_name, sim_name) 

### 2023-12-08 cis-imperfect-testing.tex
Full first draft of imperfect chapter

@@
 I show in simulation that this model will recover the duration distribution (\cref{imperf-test:sec:sim-study-results}) and apply it to the CIS data (\cref{imperf-test:sec:application}).
-Plausible estimates of the duration distribution are produced, however, they require assuming that the test sensitivity (the probability of testing positive given being detectable) is known; I discuss what would be required to relax this assumption in \cref{imperf-test:sec:discussion}.
+Finally, I discuss the results and further work in \cref{imperf-test:sec:discussion}.
 
@@
 A single positive episode is an episode containing exactly one positive test; intuitively, these cause short estimates because there is no lower bound on the length of time the episode lasts.
-Furthermore, due to missing short episodes, there is little or no ability to upper bound the number of very short episodes because these would almost all be missed.
-\Textcite{shenNonparametrica}, extending \textcite{panNote}, showed that the maximum likelihood estimator is inconsistent and can lead to a severely biased underestimation of the survival function, albeit when only the terminating event is interval censored\todo{double check the setting of Shen}.
+Furthermore, short episodes are very likely to be missed.
+Therefore, there is little or no ability to upper bound the number of very short episodes.
+\Textcite{shenNonparametrica}, extending \textcite{panNote}, showed that the maximum likelihood estimator is inconsistent and can lead to a severely biased underestimation of the survival function, albeit when only the terminating event is interval censored and the initiating event has simple left truncation.
 While the result cannot simply be extended to the more complex study design of the CIS, it reinforces the intuition that single positive episodes are likely to be problematic.
-Furthermore, simulations building in even a few excess single positive episodes led to problematic estimation even within the CIS setting (not shown).
+I found that simulations building in even a few excess single positive episodes led to problematic estimation within the CIS setting (not shown).
 
@@
 $\psens = 1$ means that there are no false negatives, the situation previously considered.
-False negatives are formalised by replacing the deterministic $y_i(t)$ with a random binary variable such that:
+False negatives are formalised by defining the outcome of a test as a random binary variable such that:
 \begin{equation}
@@
 \end{equation}
-with $Y_i(t)$ still only defined for $t \in t_i$.
-\todo{check all this notation: is $y_t(t)$ and $t_i$ defined?}
+where $Y_i(t)$ is only defined for $t \in \sched_i$.
 However, an implausibly low value for $\psens$ is required to reproduce the rate of single positive episodes seen in the CIS.
-\todo[inline]{Figure demonstrating this and references to what reasonable test sensitivities are, including the estimate from the ATACCC chapter}
 
@@
 \begin{figure}
-  \todo[inline]{Extract figure from SRS}
-  \caption[Bounding test sensitivity using CIS data]{Bounding the test sensitivity using CIS data as a function of time since the infection was detected (see main text for explanation of how the bounds are formed).\label{imperf-test:fig:bounding-cis-sensitivity}}
+  \includegraphics{cis-imperfect-testing/test-sens-bound}
+  \caption[Bounding test sensitivity using CIS data]{
+    Bounding the test sensitivity using CIS data as a function of time since the infection was detected (see main text for explanation of how the bounds are formed).
+    Small numbers of tests mean that the series is noisy near the start and the end.
+  }
+  \label{imperf-test:fig:bounding-cis-sensitivity}
 \end{figure}
@@
 Simulated data using this test sensitivity is much more similar to the data, while being more plausible than very low probabilities of false negatives.
-\todo{Figure comparing the variable test sensitivity to fixed and CIS data}
-
-\todo[inline]{Add the evidence that false negatives do create the same issue in inference in real and simulated data}
 
@@
 The simple model ensures the likelihood remains tractable.
-This requires modifying both $p_{ia}$ to allow for the episode possibly being longer than observed, and $p_{it}$ to allow for additional episodes being missed.
+This requires modifying both $p_{ia}$ to allow for the episode possibly being longer than observed, and $p_{iu}$ to allow for additional episodes being missed.
 
@@
 
-For tractability, assume that the negative test bounding the start of the infection, at $l_i^{(b)}$, is a true negative.
-This assumption is reasonable because the test sensitivity is high early in an infection, therefore $l_i^{(b)}$ is unlikely to be a false negative.
-\todo{Are these bounds correct - check notation from prev chapter?}
-We will consider only the tests between $r_i^{(b)}$ and $l_i^{(e)}$ inclusive (the positive tests providing a lower bound on the length of the episode).
-Denote these tests by $t'_i = \{ t \in t_i : r_i^{(b)} \leq t \leq l_i^{(e)} \}$ and their results by $y_i'$.
-We know that the test results at times in $t_i'$ are either true positives or false negatives. 
+For tractability, assume that the negative test bounding the start of the infection, at $l_i^{(b)}-1$, is a true negative.
+This assumption is reasonable because the test sensitivity is high early in an infection, therefore this test is unlikely to be a false negative.
+Consider only the tests between $r_i^{(b)}$ and $l_i^{(e)}$ inclusive (the positive tests providing a lower bound on the length of the episode).
+Denote these tests by $\sched'_i = \{ t \in \sched_i : r_i^{(b)} \leq t \leq l_i^{(e)} \}$ and their results by $y_i'$.
+The test results at times in $\sched_i'$ are either true positives or false negatives. 
 By definition, the test at $r_i^{(e)}$ is a false negative if and only if $e_i > r_i^{(e)}$.
-We proceed by considering the two cases.
+I proceed by considering whether this is the case.
 
 First, if $e_i \leq r_i^{(e)}$.
-In this case, the test at $r_i^{(e)}$ is a true negative, as are all other tests not in $t_i'$, and these occur with probability 1.
+In this case, the test at $r_i^{(e)}$ is a true negative, as are all other tests not in $\sched'_i$, and these occur with probability 1.
 \begin{align}
@@
 &= p(y_i' \mid l_i^{(e)} \leq e_i \leq r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
-&= \left( \prod_{t \in t_i'} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) \right)
+&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) \right)
 \end{align}
@@
 &= p(y_i \mid e_i > r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(e_i > r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
-&= \left( \prod_{t \in t_i'} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) (1 - p_\text{sens}) S_\theta(r_i^{(e)} - b_i - 1)
+&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) (1 - p_\text{sens}) S_\theta(r_i^{(e)} - b_i - 1)
 \end{align}
@@
 Combining the above, the replacement for $p_{ia}$ is:
-\todo[inline]{Fix the length of the lines of these equations}
 \begin{align}
@@
 &= \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( p(y_i', b_i, e_i \leq r_i^{(e)} \mid p_\text{sens}, \theta) p(y_i', b_i e_i > r_i^{(e)} \mid p_\text{sens}, \theta) \right) p(b_i \mid p_\text{sens}, \theta) \\
-&= \left( \prod_{t \in t_i'} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \\ & \ \times \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) + (1 - p_\text{sens}) S_\theta(r_i^{(e)} - b_i - 1) \right) p(b_i \mid p_\text{sens}, \theta) \\
-&= \left( \prod_{t \in t_i'} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right)\sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - p_\text{sens} S_\theta(r_i^{(e)} - b_i - 1) \right) p(b_i \mid p_\text{sens}, \theta).
+&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \\ & \ \times \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) + (1 - p_\text{sens}) S_\theta(r_i^{(e)} - b_i - 1) \right) \\ & \ \times p(b_i \mid p_\text{sens}, \theta) \\
+&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \\ & \ \times \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - p_\text{sens} S_\theta(r_i^{(e)} - b_i - 1) \right) p(b_i \mid p_\text{sens}, \theta).
 \label{imperf-test:eq:pia-prime}
@@
 
-\subsection{Modifying \texorpdfstring{$p_{it}$}{pit}} \label{modifying-p_it}
+\subsection{Modifying \texorpdfstring{$p_{iu}$}{piu}} \label{modifying-p_it}
 
 With false negatives, an episode with can be undetected in the following four ways; the first two of these are the same as before, but the second two are new.
-\todo[inline]{Check all this notation with the previous chapter's}
+Note that each of these is mutually exclusive.
 \begin{enumerate}
 \item
-  The episode begins before $\min(t_i)$.
-  I assume there is a negligible probability of a false negative test at this point because viral load is high.
+  The episode begins before the individual's first test: $b_i \leq \min \sched_i$.
+  % I assume there is a negligible probability of a false negative test at this point because viral load is high.
 \item
-  The duration of the episode is less than $t_{ib_i}^N$.
+  The episode begins after the individual's first test, but ends before the next test after beginning: $b_i \leq \min \sched_i$ and $e_i < b_i + \tau_{\sched_i}(b_i)$.
 \item
-  The duration of the episode is at least as long as $t_{it}^N$ but less than the time to the second test after $t$, $t_{it}^{2N}$, and a false negative episode occurred at the first test after $t$.
-  This occurs with probability:
-  \begin{math}
-    (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(t_i)}^T \left( S_\theta(t_{ib}^N + 1) - S_\theta(t_{ib}^{2N} + 1)\right).
-  \end{math}
+  The episode begins after the individual's first test, ends after the first test at or following $b_i$, that test is a false negative, but the episode ends before the second test at or following $b_i$.
+  Then, $D_i$ of the episode is greater than $\tau_{\sched_i}(b_i)$ less than  $\tau^2_{\sched_i}(b_i) \stackrel{\text{def}}{=} \tau_{\sched_i}(\tau_{\sched_i}(b_i) + 1)$, and a false negative episode occurred at the first test after $t$.
+  This probability, conditional on $b_i$ is:
+  \begin{align}
+    &\prob \left( b_i + \tau_{\sched_i}(b_i) \leq E_i < b_i + \tau^2_{\sched_i}(b_i), Y_i(b_i + \tau_{\sched_i}(b_i) = 0) \mid b_i, \theta \right) \\
+    &= \prob \left(\tau_{\sched_i}(b_i) + 1 \leq D_i < \tau^2_{\sched_i}(b_i) + 1 \mid b_i, \theta \right) (1 - \psens) \\
+    &= \left( S_\theta(\tau_{\sched_i}(b_i) + 1) - S_\theta(\tau^2_{\sched_i}(b_i) + 1) \right) (1 - \psens).
+  \end{align}
+  Integrating over $b_i$ for $b_i > \min \sched_i$, in the same way as done for \cref{E-perf-test:eq:piu}, gives:
+  \begin{align}
+    (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_i)}^T \left( S_\theta(\tau_{\sched_i}(b) + 1) - S_\theta(\tau^2_{\sched_i}(b) + 1) \right).
+  \end{align}
 \item
-  The duration of the episode is at least as long as $t_{it}^{2N}$, and all the tests within the episode (at least two) are false
-negatives.
+  The episode begins after the individual's first test, ends after the second test at or following $b_i$, and all tests within the episode (at least two) are false negatives.
   I assume this occurs with negligible probability.
@@
 
-Hence, the replacement for $1 - p_{it}$ is:
+Therefore, $p_{iu}$ is replaced by $p_{iu}'$, the sum of: $p_{iu}$ (number 1 and 2; derived in \cref{E-perf-test:eq:piu}), and number 3.
+For the posterior density (\cref{E-perf-test:eq:full-posterior}), $1 - p_{iu}'$ is the required quantity:
 \begin{align}
-1 - p_{it}'
-&= 1 - p_{it} - (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(t_i)}^T \left( S_\theta(t_{ib}^N + 1) - S_\theta(t_{ib}^{2N} + 1)\right) \\
-&= \frac{1}{T} \sum_{b=\min(t_i)}^T \left( p_\text{sens} S_\theta(t_{ib}^N + 1) + (1 - p_\text{sens}) S_\theta(t_{ib}^{2N} + 1)\right).
+1 - p_{iu}'
+&= 1 - p_{iu} - (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(t_i)}^T \left( S_\theta(\tau_{\sched_i}(b) + 1) - S_\theta(\tau^2_{\sched_i}(b) + 1) \right) \\
+&= \frac{1}{T} \sum_{b=\min(t_i)}^T \left( p_\text{sens} S_\theta(\tau_{\sched_i}(b) + 1) + (1 - p_\text{sens}) S_\theta(\tau^2_{\sched_i}(b) + 1)\right).
 \label{imperf-test:eq:pit-prime}
@@
 
-$p_{it}'$ and $p_{ia}'$ are substituted for $p_{it}$ and $p_{ia}$ respectively in \cref{E-perf-test:eq:full-posterior}.
+$p_{iu}'$ and $p_{ia}'$ are substituted for $p_{iu}$ and $p_{ia}$ respectively in \cref{E-perf-test:eq:full-posterior}.
 The posterior density is otherwise unchanged.
@@
     Posterior (median and 95\% credible interval) of the survival time for the simulation study with a variable test sensitivity.
-    All simulations use a constant test sensitivity of 0.8 and the combination prior (as in \cref{imperf-test:fig:misspecified-test-sensitivity}), but the model assumes different values.
+    Each panel shows the results of performing inference with a different assumed value for the test sensitivity.
     The posterior estimate using a model with a constant test sensitivity of 0.8 (middle panel) is similar to the true value (black line).
@@
 
+% \begin{itemize}
+%   \item Survival prior matters for quantitative details but not qualitative shape. 
+%   \item Number of missed infections prior is more important, but not the case in simulation (not shown).
+%   \item Unclear why this is the case, probably model misspecification meaning that fewer missed infections than expected.
+%   \item Two possible misspecifications: test sensitivity (not constant) and constant incidence.
+% \end{itemize}
+
 This chapter shows that a simple model of false negatives can be incorporated into the model of \cref{E-perf-test}.
@@
 However, the rest of the distribution is sensitive to this choice.
-
-\todo[inline]{discussion on real results into prose}
-
-\begin{itemize}
-  \item Survival prior matters for quantitative details but not qualitative shape. 
-  \item Number of missed infections prior is more important, but not the case in simulation (not shown).
-  \item Unclear why this is the case, probably model misspecification meaning that fewer missed infections than expected.
-  \item Two possible misspecifications: test sensitivity (not constant) and constant incidence.
-\end{itemize}
 

### 2023-12-08 cis-imperfect-testing/sim_single_positive_episodes.R

@@ -1,0 +1,37 @@
+suppressMessages(library(dplyr))
+library(ggplot2)
+library(patchwork)
+library(purrr)
+library(tidybayes)
+library(tidyr)
+source(here::here("utils.R"))
+
+sims = readRDS(here::here("cisRuns-output/sim_reps.rds")) |>
+    as_tibble()
+    
+p_single_pos = sims |>
+    summarise(
+        .by = c(.draw, scenario_name),
+        num_single = sum(n_pos == 1)
+    ) |>
+    mutate(
+        sensitivity = stringr::str_split_fixed(scenario_name, "-", 3)[,3] |>
+            stringr::str_to_sentence()
+    ) |>
+    ggplot() +
+    geom_bar(aes(num_single, fill = sensitivity)) +
+    standard_plot_theming() +
+    labs(
+        x = "Number of single positive episodes",
+        y = "Number of simulations",
+        fill = "Test sensitivity"
+    )
+
+ggsave(
+    filename = here::here("cis-imperfect-testing/sim-single-positive-episodes.png"),
+    plot = p_single_pos,
+    width = 15,
+    height = 9,
+    units = "cm",
+    dpi = 300
+)

### 2023-12-08 cis-imperfect-testing/test_sens_bound.R

@@ -1,0 +1,49 @@
+library(dplyr)
+library(ggplot2)
+library(tidyr)
+source(here::here("utils.R"))
+
+bound_data = readr::read_csv(
+    here::here("cis-imperfect-testing/STATS18596/changing-sensitivity/results_by_day_of_episode.csv")
+)
+
+p_bound = bound_data |>
+    filter(result != "result") |>
+    pivot_wider(
+        names_from = result,
+        values_from = n,
+        id_cols = days_since_first_pos,
+        values_fill = 0
+    ) |>
+    transmute(
+        days_since_first_pos,
+        true_pos = Positive,
+        min_false_neg = `Intermittent neg`,
+        max_false_neg = min_false_neg + `First neg`,
+        min_test_sens = true_pos / (true_pos + max_false_neg),
+        max_test_sens = true_pos / (true_pos + min_false_neg),
+        proposed = if_else(
+            days_since_first_pos <= 50,
+            0.9 - (0.9 - 0.5)/50 * days_since_first_pos,
+            0.5
+        )
+    ) |>
+    ggplot() +
+    geom_line(aes(days_since_first_pos, min_test_sens, colour = "Min")) +
+    geom_line(aes(days_since_first_pos, max_test_sens, colour = "Max")) +
+    # geom_line(aes(days_since_first_pos, proposed, colour = "Proposed")) +
+    standard_plot_theming() +
+    labs(
+        x = "Days since first positive test",
+        y = "Test sensitivity",
+        colour = NULL
+    )
+
+ggsave(
+    filename = here::here("cis-imperfect-testing/test-sens-bound.png"),
+    plot = p_bound,
+    width = 15,
+    height = 9,
+    units = "cm",
+    dpi = 300
+)

### 2023-12-11 cis-imperfect-testing.tex
Insert figures for draft (where available)

@@
 In particular, it estimated a much shorter duration of positivity than in \cref{E-ATACCC}, despite those estimates being much more reliable over the first 2--3 weeks.
-
 \begin{figure}
@@
 Intermittent negatives (described in  \cref{E-biology-data:sec:PCR}) show that they occur within the CIS.
-Second, adjusting the simulation to include false negatives can reproduce both the CIS data more faithfully and exhibits the same issues when estimating the duration distribution (see \cref{imperf-test:sec:simulate}).
+Second, adjusting the simulation to include false negatives can reproduce both the CIS data more faithfully and exhibits the same issues when estimating the duration distribution (shown in \cref{imperf-test:fig:sim-single-pos} with method explained in \cref{imperf-test:sec:simulate}).
+\begin{figure}
+  \centering \includegraphics{cis-imperfect-testing/sim-single-positive-episodes}
+  \todo[inline]{Figure showing estimates without false negatives}
+  \caption[Estimating survival using CIS data assuming perfect testing]{Estimating the survival function using CIS data and the model of \cref{E-perf-test} gives implausibly short estimates.}
+  \label{imperf-test:fig:sim-single-pos}
+\end{figure}
 
@@
 The estimate using a test sensitivity of 0.8 and $\NBc(\numprint{25132}, \numprint{22047})$ give a median survival time most similar to that of \cref{E-ATACCC}.
+\begin{figure}
   \todo[inline]{Insert fig showing final estimate and comparison to ATACCC}
+  \caption{Duration estimates using CIS and ATACCC data}
+  \label{imperf-test:fig:cis-estimates}
+\end{figure}
+\begin{figure}
   \todo[inline]{
@@
   }
+  \caption{Sensitivity of CIS estimates to assumptions}
+  \label{imperf-test:fig:cis-sensitivity}
+\end{figure}
 

### 2023-12-11 cis-imperfect-testing/sim_single_positive_episodes.R

@@ -20,6 +20,7 @@ p_single_pos = sims |>
     ) |>
     ggplot() +
     geom_bar(aes(num_single, fill = sensitivity)) +
+    geom_vline(aes(xintercept = 3730/4837, colour = "Observed"), size = 1) +
     standard_plot_theming() +
     labs(
         x = "Number of single positive episodes",

### 2023-12-11 SEIR.tex
Fix region names

@@
 This chapter uses mechanistic models for estimating the transmission of SARS-CoV-2, an alternative to the statistical approach in \cref{E-backcalc}.
-Mechanistic models explicitly represent the infection process driving the spread of infectious disease, giving model's state and parameters have a biological and/or epidemiological interpretation.
+Mechanistic models explicitly represent the infection process driving the spread of infectious disease, giving model's state and parameters a biological and/or epidemiological interpretation.
 % Of particular interest is that the basic reproduction number, $\R$, and the effective reproduction number, $R_e(t)$, which can both be calculated as functions of the model parameters; using \cref{E-backcalc}'s approach, additional modelling would be required to estimate these numbers.
-This allow changes in transmission to be decomposed into the components of the model, \eg the number of contacts per day or the probability of transmission upon contact.
+In particular, this allows changes in transmission to be decomposed into the components of the model, \eg the number of contacts per day or the probability of transmission upon contact.
 Furthermore, mechanistic models can be used for scenario-based modelling, which simulates the effect of interventions.
@@
 The parameters $\vec{\theta}$ determine, along with the model structure, how the state vector changes over time.
-Inference about $\vec{x}(t)$ and $\vec{\theta}$ is through an observation model, $p(\vec{y}(t) \mid \vec{x}(t), \vec{\theta})$ which links $\vec{x}(t)$ to some observation $\vec{y}(t)$~\autocite{birrellEvidence}; I discuss observation models further in \cref{SEIR:sec:observation}.
+Inference about $\vec{x}(t)$ and $\vec{\theta}$ is through an observation model, $p(\vec{y}(t) \mid \vec{x}(t), \vec{\theta})$ which links $\vec{x}(t)$ to some observation $\vec{y}(t)$~\autocite{birrellEvidence} (see\cref{SEIR:sec:observation}).
 
@@
 
-Deterministic models are also known as \emph{mean-field} models because they can are the behaviour corresponding to a stochastic model's mean.
+Deterministic models are also known as \emph{mean-field} models because they represent the mean behaviour of the corresponding to a stochastic version?.
 When the number of infected individuals is large, the law of large numbers means that the mean-field behaviour is a good approximation of the fully stochastic model~\autocite[20]{diekmannMathematical}.
-An alternative justification, when making noisy observations of the system, is that the additional noise from the stochastic model is negligible compared to this observation noise.
+An alternative justification for this approximation , when making noisy observations of the system, is that the additional noise from the stochastic model is negligible compared to this observation noise.
 This second justification is of the same form as the one made for neglecting the stochasticity of the prevalence process in \cref{E-inc-prev:sec:observation-process}.
@@
 
-\Cref{SEIR:sec:SIR} introduces the \emph{susceptible-infected-recovered} (SIR) model, a simple compartmental model for a disease where individuals are immune to reinfection following their recovery.
-The SIR model is named because each individual in the population is divided into one of three states: susceptible, infectious, or recovered.
+\Cref{SEIR:sec:SIR} introduces the \emph{susceptible-infected-recovered} (SIR) model, the simplest compartmental model for a disease where individuals are immune to reinfection following their recovery.
+The term SIR arises from the population being divided into one of three states: susceptible, infectious, or recovered.
 This model has several assumptions which make it inappropriate to model SARS-CoV-2; therefore, in the remainder of this section I introduce standard extensions to the SIR model which relax these assumptions.
@@
 The simplest compartmental model for a disease which confers immunity is the (SIR) model\todo{ref for SIR being simplest}.
-The name refers to the three compartments in the model, displayed in \cref{SEIR:fig:SIR}.
-Individuals in the susceptible compartment are those that could be infected (they have no immunity).
+The name refers to the three compartments in the model, displayed in 
+In the SIR model (see \cref{SEIR:fig:SIR).
+individuals in the susceptible compartment are those that could be infected (they have no immunity).
 The proportion of the population that is susceptible at time $t$ is $s(t)$.
@@
 where $\lambda(t) = \beta i(t)$, is the \emph{force of infection}: the risk of infection for a susceptible individual at time $t$~\autocite[17]{keelingModeling}.
-Two parameters are present in this model, $\vec{\theta} = (\beta, \gamma)^T$.
+The dynamics of the model are described by parameters $\vec{\theta} = (\beta, \gamma)^T$.
 $\beta$ is the transmission rate, absorbing several terms as described below.
@@
     It is in this phase that the basic reproduction number provides a good approximation of the epidemic dynamics.
-    \item The basic reproduction number for this model is $\R = \beta / \gamma$~\autocite[20]{keelingModeling}.
+    \item The basic reproduction number for the SIR model is $\R = \beta / \gamma$~\autocite[20]{keelingModeling}.
     Intuitively, this can be thought of as the average number of infections caused by an infectious individual per day, $\beta$, multiplied by the average number of days they are infectious for, $1/\gamma$.
@@
     \item In addition to the parameters, the state of the system at some time (normally the initial state, the state at time $t=0$) is required to solve the system.
-    \item Immunity is assumed to last until the end of the modelled period~\autocite[61]{andersonInfectious}.
-    In many contexts, including here, this assumption is justified because the modelled period is shorter than the length of immunity conferred~\autocite{milneImmunity}.
-    For modelling longer time periods, waning immunity can be incorporated by allowing individuals to move from the recovered compartment back to the susceptible compartment~\autocite[40]{keelingModeling}.
+    \item Immunity is assumed to last until the end of the  period considered ~\autocite[61]{andersonInfectious}.
+    In many contexts, including here, this assumption is justified because the period is shorter than the length of immunity conferred~\autocite{milneImmunity}.
+    For longer time periods, waning immunity can be incorporated by allowing individuals to move from the recovered compartment back to the susceptible compartment~\autocite[40]{keelingModeling}.
 \end{itemize}
@@
 The basic SIR model implies that the distribution of the time spent in the infected compartment is exponential~\autocite[96]{keelingModeling}.
-This is an unrealistic assumption for most diseases and leads to an underestimation of the reproduction number if the model is fit to data~\autocites{lloydRealistic}{wearingAppropriate}.
+This is an unrealistic assumption for most diseases and leads to an underestimation of the reproduction number ~\autocites{lloydRealistic}{wearingAppropriate}.
 A standard extension to the SIR model is to allow the infectious period to follow a gamma distribution~\autocite[94]{keelingModeling}.
@@
 I will assume that the disease progression (\ie length of the exposed and infectious periods) is the same for all strata, although this assumption can be relaxed.
-Stratification requires two major changes model changes: extension of the state vector and describing the between-strata interactions.
+Stratification requires two major  model changes: extension of the state vector and description of the between-strata interactions.
 
@@
 For example, in the basic SIR model, $\vec{x}(t) = (\vec{s}(t), \vec{i}(t), \vec{r}(t))^T$.
-The $a$th element of each vector is the proportion of strata $a$ in that compartment.
-For example, $\vec{s}(t) = (s_1(t), \dots, s_A(t))^T$ where $s_a(t)$ is the proportion of the $a$th strata that are susceptible at time $t$.
+The $a$th $(a=1 \ldots A)$?element of each vector is the proportion of strata $a$ in that compartment.
+In teh case of the susceptible compartment $\vec{s}(t) = (s_1(t), \dots, s_A(t))^T$ where $s_a(t)$ is the proportion of the $a$th strata that are susceptible at time $t$.
 Hence, $N_a s_a(t)$ is the number of susceptible people in stratum $a$, where $N_a$ is the number of individuals in stratum $a$.
-These vectors replace the scalars in the previous equations.
+These vectors replace the scalars in previous equations.
 Various other definitions are possible and equivalent, for example defining $s_a$ as the proportion of the total population that are both in strata $a$ and susceptible or parameterised using numbers of individuals~\autocite[57]{keelingModeling}.
 
-The interactions between strata also need describing.
-This is done by replacing the scalar $\beta$ with a \emph{Who Acquires Infection From Whom} (WAIFW) matrix, $\matr{\beta}$~\autocite[58]{keelingModeling}; its interpretation is described below.
-The force of infection is now a vector, $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i_j}(t)$.
-Its interpretation can be arrived at by considering a single infected individual in strata $a'$, analogous to the previous explanation in the single stratum case.
-Assume any such individual comes into contact with individuals in strata $a$ at a rate of $\kappa_{aa'}$ in units of contacts per day.
+The interactions between strata is done by replacing the scalar $\beta$ with a \emph{Who Acquires Infection From Whom} (WAIFW) matrix, $\matr{\beta}$~\autocite[58]{keelingModeling}; its interpretation is described below.
+So the force of infection is now a vector, $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i_j}(t)$.
+Its interpretation can be arrived at by considering a single infected individual in strata $a'$, analogously to the single stratum case.
+Assume any such individual comes into contact with individuals in strata $a$ at a rate $\kappa_{aa'}$ in units of contacts per day.
 % Further, assume that each contact is with an individual chosen uniformly at random from within the strata.
@@
 
-\todo[inline]{format region names in plots}
-
 Overall the model fits the data well (see \cref{SEIR:fig:prev-young,SEIR:fig:prev-old}).

### 2023-12-11 cis-imperfect-testing.tex
Improve and describe figure

@@
   \centering \includegraphics{cis-imperfect-testing/sim-single-positive-episodes}
-  \todo[inline]{Figure showing estimates without false negatives}
-  \caption[Estimating survival using CIS data assuming perfect testing]{Estimating the survival function using CIS data and the model of \cref{E-perf-test} gives implausibly short estimates.}
+  \caption[Single positive episodes in CIS simulation]{%
+    Proportion of detected episodes which have a single positive episode.
+    y-axis is the number of simulations, out of 1000, with the indicated proportion.
+    The CIS data is shown as a vertical line.
+    Simulation methodology is explained in \cref{E-perf-test:sec:simulation-study}, with false negatives included as described in \cref{imperf-test:sec:simulate}.
+  }
   \label{imperf-test:fig:sim-single-pos}

### 2023-12-11 cis-imperfect-testing/sim_single_positive_episodes.R

@@ -12,7 +12,7 @@ sims = readRDS(here::here("cisRuns-output/sim_reps.rds")) |>
 p_single_pos = sims |>
     summarise(
         .by = c(.draw, scenario_name),
-        num_single = sum(n_pos == 1)
+        num_single = mean(n_pos == 1)
     ) |>
     mutate(
         sensitivity = stringr::str_split_fixed(scenario_name, "-", 3)[,3] |>
@@ -23,9 +23,10 @@ p_single_pos = sims |>
     geom_vline(aes(xintercept = 3730/4837, colour = "Observed"), size = 1) +
     standard_plot_theming() +
     labs(
-        x = "Number of single positive episodes",
+        x = "Proportion of episodes",
         y = "Number of simulations",
-        fill = "Test sensitivity"
+        fill = "Test sensitivity",
+        colour = ""
     )
 
 ggsave(

### 2023-12-11 cis-imperfect-testing.tex
Editing

--- 2023-12-11 cis-imperfect-testing.tex
@@
 There exists little prior work on the incorporation of misclassification bias within similar survival analysis frameworks.
-\todo[inline]{Check exactly what has been done and verify exactly what I can claim here}
+\todo[inline]{Review literature properly, here or elsewhere}
 In this chapter, I modify the previous model to include false negatives (\cref{imperf-test:sec:modelling}).
@@
 Therefore, there is little or no ability to upper bound the number of very short episodes.
-\Textcite{shenNonparametrica}, extending \textcite{panNote}, showed that the maximum likelihood estimator is inconsistent and can lead to a severely biased underestimation of the survival function, albeit when only the terminating event is interval censored and the initiating event has simple left truncation.
+I found that simulations building in even a few excess single positive episodes led to problematic estimation within the CIS setting (not shown).
+
+\Textcite{shenNonparametrica}, extending \textcite{panNote}, studied the situation when only the terminating event is interval censored and the initiating event has simple left truncation.
+This is related, although simpler, setting than the CIS; the CIS has double interval censoring and a more complex pattern of missed episodes (see \cref{E-perf-test:sec:problem}).
+\Textcite{shenNonparametrica} showed that, in their situation, the maximum likelihood estimator is inconsistent and can lead to a severely biased underestimation of the survival function.
 While the result cannot simply be extended to the more complex study design of the CIS, it reinforces the intuition that single positive episodes are likely to be problematic.
-I found that simulations building in even a few excess single positive episodes led to problematic estimation within the CIS setting (not shown).
 
@@
 Intermittent negatives (described in  \cref{E-biology-data:sec:PCR}) show that they occur within the CIS.
-Second, adjusting the simulation to include false negatives can reproduce both the CIS data more faithfully and exhibits the same issues when estimating the duration distribution (shown in \cref{imperf-test:fig:sim-single-pos} with method explained in \cref{imperf-test:sec:simulate}).
+Second, adjusting the simulation to include false negatives can reproduce both the CIS data more faithfully and exhibits the same issues when estimating the duration distribution (see \cref{imperf-test:fig:sim-single-pos}).
 \begin{figure}
@@
 
-\section{Simulating false negatives} \label{imperf-test:sec:simulate}
+\section{Modelling false negatives} \label{imperf-test:sec:simulate}
 
@@
 where $t$ is the time in days since infection.
-Simulated data using this test sensitivity is much more similar to the data, while being more plausible than very low probabilities of false negatives.
+Simulated data using this test sensitivity is much more similar to the data (see \cref{imperf-test:fig:sim-single-pos}), while being more plausible than very low probabilities of false negatives.
 
@@
 In this section, I introduce a simple model of false negatives into the model of \cref{E-perf-test:sec:model}.
-The simple model ensures the likelihood remains tractable.
-This requires modifying both $p_{ia}$ to allow for the episode possibly being longer than observed, and $p_{iu}$ to allow for additional episodes being missed.
+A simple model means that the likelihood remains tractable.
+I modify both $p_{ia}$ to allow for the episode possibly being longer than observed (\cref{imperf-test:sec:modifying-p_ia}), and $p_{iu}$ (\cref{imperf-test:sec:modifying-p_iu}) to allow for additional episodes being missed.
 
@@
 
-\subsection{Modifying \texorpdfstring{$p_{ia}$}{pia}} \label{modifying-p_ia}
+\subsection{Modifying \texorpdfstring{$p_{ia}$}{pia}} \label{imperf-test:sec:modifying-p_ia}
 
@@
 First, if $e_i \leq r_i^{(e)}$.
-In this case, the test at $r_i^{(e)}$ is a true negative, as are all other tests not in $\sched'_i$, and these occur with probability 1.
+In this case, the test at $r_i^{(e)}$ is a true negative, as are all other tests not in $\sched'_i$.
+These occur with probability 1, as I continue to assume no false positives.
 \begin{align}
@@
 Second, if $e_i > r_i^{(e)}$.
-In this case, the test at $r_i^{(e)}$ is a false negative, occurring with probability $(1 - p_\text{sens})$, and we have no upper bound on the end of the infection.
+In this case, the test at $r_i^{(e)}$ is a false negative, occurring with probability $(1 - p_\text{sens})$, and the end of the infection is right-censored.
 \begin{align}
@@
 
-\subsection{Modifying \texorpdfstring{$p_{iu}$}{piu}} \label{modifying-p_it}
+\subsection{Modifying \texorpdfstring{$p_{iu}$}{piu}} \label{imperf-test:sec:modifying-p_iu}
 
@@
   \end{align}
-  Integrating over $b_i$ for $b_i > \min \sched_i$, in the same way as done for \cref{E-perf-test:eq:piu}, gives:
+  Integrating over $b_i$ for $b_i > \min \sched_i$, in the same way as \cref{E-perf-test:eq:piu}, gives:
   \begin{align}
@@
 
-Therefore, $p_{iu}$ is replaced by $p_{iu}'$, the sum of: $p_{iu}$ (number 1 and 2; derived in \cref{E-perf-test:eq:piu}), and number 3.
+Therefore, $p_{iu}$ is replaced by $p_{iu}'$, the sum of: $p_{iu}$ (for number 1 and 2; derived in \cref{E-perf-test:eq:piu}), and the probability in number 3.
 For the posterior density (\cref{E-perf-test:eq:full-posterior}), $1 - p_{iu}'$ is the required quantity:
@@
 I focus on the independent (vague) and model combination priors for the survival (described in \cref{E-perf-test:sec:parameters-priors}.
-These were shown to be the best performing priors in \cref{E-perf-test:sec:results}
+These were shown to be the best performing priors in \cref{E-perf-test:sec:results}.
 
@@
 Unlike in the simulation studies, an uninformative prior on $\ntot$ led to implausible estimates of the duration distribution.
-The uninformative prior led to high posterior estimates, and hence an implausibly large number of episodes with durations of less than five days.
+The uninformative prior led to high posterior estimates of $\ntot$, and hence an implausibly large number of episodes with durations of less than five days.
 Therefore, I based an informative prior for $\ntot$ on pre-existing estimates of the total number of infections over the period with posterior mean \numprint{4136368} and standard deviation \numprint{27932}~\autocite{birrellRTM2}.
@@
 This model allows estimation of the duration distribution tails without the strong model assumptions used in \cref{E-ATACCC}.
-The qualitative features of the estimated distribution are robust to the assumed value for $\psens$, and the choice of prior for $\lambda$, although the quantative details are not.
+The qualitative features of the estimated distribution are robust to the assumed value for $\psens$, and the choice of prior for $\lambda$, although the quantitative details are not.
 The estimates of the tail, the primary purpose of this chapter, are in addition robust to the choice of prior for $\ntot$.
@@
 
-The estimates I give in this chapter are the first estimates of the duration of PCR-positivity, including the distribution tails, estimated from general population data.
+\section{Conclusion} \label{imperf-test:sec:conclusion}
+
+This chapter has produced estimates of the duration distribution of PCR-positivity using information from two studies: ATACCC and CIS.
+Assuming a test sensitivity of 0.8 and an informative prior on the number of infections, provided the preferred estimates.
+
+These estimates are the first estimates of the duration of PCR-positivity, including the distribution tails, estimated from general population data.
 This feature of the distribution is important for understanding the proportion of prevalence attributable to long infections and interpreting PCR test results.
@@
 
-\section{Conclusion} \label{imperf-test:sec:conclusion}
-
-This chapter has produced estimates of the duration distribution of PCR-positivity using information from two studies: ATACCC and CIS.
-Assuming a test sensitivity of 0.8 and an informative prior on the number of infections, provided the preferred estimates.
 The posterior distribution for the survival is concentrated, producing narrow credible intervals.

### 2023-12-12 SEIR.tex
Comments on SEIR

@@
 However, they make stronger assumptions about the disease and population behaviour than statistical models.
+
+The literature on mechanistic models is vast, having been developed for over a century; the model in \cref{SEIR:sec:SIR} was first formulated by \textcite{kermackContribution}.
+In this chapter, I will discuss only the background relevant to the application.
+For further details I recommend the tutorial paper \textcite{kretzschmarMathematical} or the textbook \textcite{keelingModeling}.
 
@@
 A compartmental model consists of a state vector $\vec{x}(t)$ and a set of parameters $\vec{\theta}$~\autocite{birrellEvidence}.
-The elements of $\vec{x}(t)$ are the proportion of the population (or number of individuals) in each compartment at time $t$.
+The elements of $\vec{x}(t)$ are the latent proportion of the population (or number of individuals) in each compartment at time $t$.
 The parameters $\vec{\theta}$ determine, along with the model structure, how the state vector changes over time.
-Inference about $\vec{x}(t)$ and $\vec{\theta}$ is through an observation model, $p(\vec{y}(t) \mid \vec{x}(t), \vec{\theta})$ which links $\vec{x}(t)$ to some observation $\vec{y}(t)$~\autocite{birrellEvidence} (see\cref{SEIR:sec:observation}).
+Inference about $\vec{x}(t)$ and $\vec{\theta}$ is requires linking to data, $\vec{y}$, via an observation model, $p(\vec{y}(t) \mid \vec{x}, \vec{\theta})$ where $\vec{x}$ denotes conditioning on $\vec{x}(t)$ for all $t$~\autocite{birrellEvidence} (see \cref{SEIR:sec:observation}).
+That is, the observation model specifies the likelihood of the data given the model state.
 
 Compartmental models can be \emph{deterministic} or \emph{stochastic}.
-Deterministic models are defined by the state vector $\vec{x}(t)$ being a deterministic function of the parameters $\vec{\theta}$ and the initial state $\vec{x}(0)$; equivalently, $\var(\vec{x}(t) \mid \theta) = 0$~\autocite{birrellEvidence}.
-In most situations, the function is not analytically-tractable and requires numeric methods to solve.
+Deterministic models are defined by the state vector $\vec{x}(t)$ being a deterministic function of the parameters $\vec{\theta}$ and the initial state $\vec{x}(0)$; equivalently, $\var(\vec{x}(t) \mid \vec{x}(0), \theta) = 0$~\autocite{birrellEvidence}.
+% In most situations, the function is not analytically-tractable and requires numeric methods to solve.
 %The transmission model corresponds to the infection process (see \cref{E-inc-prev:sec:infection-process}).
@@
 
+Stochastic models specify a probability distribution for the current state $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta})$ where $\set{H}_t = \{ \vec{x}(t') \ssep t' < t \}$ is the epidemic's history and $\vec{\theta}$ are model parameters\todo{REF stochastic models}.
 Stochastic models are most commonly specified in discrete time, using a Markov assumption.
-They specify a probability distribution $p(\vec{x}(t) \mid \vec{x}(t-1), \vec{\theta})$ for integer $t$ and assume that, having conditioned on $\vec{x}(t-1)$, $\vec{x}(t)$ is independent of $\vec{x}(t')$ for all $t' < t - 1$~\autocite{birrellEvidence}.
+A common simplification is to use discrete time and a Markov assumption; \ie $\vec{x}(t)$ is only defined for integer $t$ and $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta}) = p(\vec{x}(t) \mid \vec{x}(t-1), \vec{\theta})$~\autocite{birrellEvidence}.
 This relaxes the assumption of independent increments in the counting process formulation of the infection process (see \cref{E-inc-prev:sec:infection-process}).
 
-Deterministic models are also known as \emph{mean-field} models because they represent the mean behaviour of the corresponding to a stochastic version?.
-When the number of infected individuals is large, the law of large numbers means that the mean-field behaviour is a good approximation of the fully stochastic model~\autocite[20]{diekmannMathematical}.
-An alternative justification for this approximation , when making noisy observations of the system, is that the additional noise from the stochastic model is negligible compared to this observation noise.
-This second justification is of the same form as the one made for neglecting the stochasticity of the prevalence process in \cref{E-inc-prev:sec:observation-process}.
+Deterministic models are also known as \emph{mean-field} models because they represent the mean behaviour of a corresponding stochastic model.
+When the number of infectious individuals is large, the law of large numbers implies that the mean-field behaviour is a good approximation of the fully stochastic model~\autocite[20]{diekmannMathematical}.
+This is especially true when making noisy observations of the system, and that the stochastic differences between the stochastic and its mean are negligible compared to the observation noise.
+A formalisation of this argument underpins the mean-field approximation of the prevalence process made in \cref{E-inc-prev:sec:observation-process}.
 
 The remainder of this chapter uses only deterministic models.
-Throughout the time period this thesis considers, there were many infectious individuals.
+This choice is appropriate for the the context this thesis considers because there were many individuals infectious with SARS-CoV-2 in this time period.
 For an introduction to stochastic models, see \textcite[chapter 6]{keelingModeling} or \textcite{birrellEvidence}.
@@
 This chapter discusses deterministic models whose evolution is specified by a system of ODEs, $d\vec{x}/dt$.
-In an ODE model, time is necessarily continuous, although discretised when solving (see \cref{SEIR:sec:inference-implementation}).
+In an ODE model, time is necessarily continuous.%, although discretised when solving (see \cref{SEIR:sec:inference-implementation}).
 For concreteness, I use units of days throughout, although the model can use any unit of time.
-The number or proportion of individuals in each compartment is also continuous, which is a reasonable approximation when the population is large.
+The number or proportion of individuals in each compartment is also continuous; clearly, this is not true in reality but is a good approximation when the population is large.
 
-\Cref{SEIR:sec:SIR} introduces the \emph{susceptible-infected-recovered} (SIR) model, the simplest compartmental model for a disease where individuals are immune to reinfection following their recovery.
-The term SIR arises from the population being divided into one of three states: susceptible, infectious, or recovered.
-This model has several assumptions which make it inappropriate to model SARS-CoV-2; therefore, in the remainder of this section I introduce standard extensions to the SIR model which relax these assumptions.
-First, in \cref{SEIR:sec:non-exponential}, I relax the assumption that the infectious period is exponentially-distributed to allow gamma distributions.
-Then, in \cref{SEIR:sec:SEIR}, I relax the assumption that individuals are infectious immediately after they are infected; it does this by extending the SIR model to the SEIR model with the additional E signifying an \emph{exposed} compartment.
-Finally, in \cref{SEIR:sec:structured-populations}, I include heterogeneity in the population, stratifying by important characteristic(s).
+In this section, I introduce the \emph{susceptible-infectious-recovered} (SIR) model and the extensions to it I require to appropriatly model SARS-CoV-2.
+
+% \Cref{SEIR:sec:SIR} introduces the \emph{susceptible-infectious-recovered} (SIR) model, the simplest compartmental model for a disease where individuals are immune to reinfection following their recovery.
+% The name SIR derives from the structure of the model.
+% In particular, the model divides the population into one of three states: susceptible, infectious, or recovered.
+
+% This model has several assumptions which make it inappropriate to model SARS-CoV-2; therefore, in the remainder of this section I introduce standard extensions to the SIR model which relax these assumptions.
+% First, in \cref{SEIR:sec:non-exponential}, I relax the assumption that the infectious period is exponentially-distributed to allow gamma distributions.
+% Then, in \cref{SEIR:sec:SEIR}, I relax the assumption that individuals are infectious immediately after they are infected; it does this by extending the SIR model to the SEIR model with the additional E signifying an \emph{exposed} compartment.
+% Finally, in \cref{SEIR:sec:structured-populations}, I include heterogeneity in the population, stratifying by important characteristic(s).
 % Finally, \cref{SEIR:sec:time-varying-foi} relaxes the assumption of a constant transmission rate to allow for time-varying behavioural changes.
-
-The literature on mechanistic models is vast, having been developed for over a century; the model in \cref{SEIR:sec:SIR} was first formulated by \textcite{kermackContribution}.
-In this section, I will explain only the parts relevant to the remainder of the chapter.
-For further details I recommend the tutorial paper \textcite{kretzschmarMathematical} or the textbook \textcite{keelingModeling}.
 
@@
 
-The simplest compartmental model for a disease which confers immunity is the (SIR) model\todo{ref for SIR being simplest}.
-The name refers to the three compartments in the model, displayed in 
-In the SIR model (see \cref{SEIR:fig:SIR).
-individuals in the susceptible compartment are those that could be infected (they have no immunity).
+The simplest compartmental model for a disease which confers immunity is the (SIR) model\todo{ref for SIR being simplest} (depicted in \cref{SEIR:fig:SIR}).
+The name SIR derives from the structure of the model.
+In particular, the model divides the population (assumed closed, an assumption discussed later) into one of three compartments: susceptible, infectious, or recovered.
+% The state of the system is the proportion of the population in each compartment.
+In the SIR model, individuals are assumed to be homogeneous except for their disease state, \ie they are otherwise identical.
+
+Individuals in the susceptible compartment are those that could be infected (they have no immunity).
 The proportion of the population that is susceptible at time $t$ is $s(t)$.
@@
 The model's state is $\vec{x}(t) = (s(t), i(t), r(t))^T$.
-In the SIR model, individuals are assumed to be homogeneous, \ie they are all identical.
 
@@
 where $\lambda(t) = \beta i(t)$, is the \emph{force of infection}: the risk of infection for a susceptible individual at time $t$~\autocite[17]{keelingModeling}.
-The dynamics of the model are described by parameters $\vec{\theta} = (\beta, \gamma)^T$.
+The dynamics of the model are described by the parameters $\vec{\theta} = (\beta, \gamma)^T$.
 $\beta$ is the transmission rate, absorbing several terms as described below.
-$\gamma$ is the recovery rate, or the proportion of infected individuals that recover per day; hence, $1/\gamma$ is the mean infectious period~\autocite[367]{keelingModeling}.
+$\gamma$ is the recovery rate, or the rate that infectious individuals that recover per day; hence, $1/\gamma$ is the mean infectious period~\autocite[367]{keelingModeling}.
 These parameters are assumed to be constant.
@@
 % Susceptible individuals move from the $s(t)$ to $i(t)$ when they are infected.
-The expression for the force of infection, $\lambda(t)$ can be arrived at by considering a single infected individual~\autocite[214]{kretzschmarMathematical}.
-Assume this individual, on average, comes into contact with other individuals in the population at a rate of $\kappa$ per day.
+The expression for the force of infection, $\lambda(t)$ can be arrived at by considering a single infectious individual~\autocite[214]{kretzschmarMathematical}.
+Assume this individual, on average, comes into contact with other individuals in the population at a rate of $\kappa$.
 Further, assume the population is \emph{well-mixed}, meaning that each contact is with an individual chosen uniformly at random from the population; this is implied by assuming a homogeneous population.
 For such a randomly selected individual, the probability that they are susceptible is $s(t)$.
-% Each of the infected individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
-Therefore, in expectation, the infected individual makes contact with susceptible individuals at a rate of $\kappa s(t)$.
-Assume that there is a constant probability of transmission upon contact between an infected and susceptible individual, $q$.
-Combined with the above, the expected rate of infections caused by this infected individual is $\kappa s(t) q$.
+% Each of the infectious individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
+Therefore, in expectation, the infectious individual makes contact with susceptible individuals at a rate of $\kappa s(t)$ per day.
+Assume that there is a constant probability of transmission upon contact between an infectious and susceptible individual, $q$.
+Combined with the above, the expected rate of infections caused by this infectious individual is $\kappa s(t) q$ per day.
 The two constant terms here (number of contacts and probability of transmission) are absorbed into the transmission rate, $\beta = \kappa q$.
-Therefore, the rate of new infections generated by a single infectious individual is $\beta s(t)$.
-Multiplying by the number of infected individuals and dividing by the population size gives the rate at which susceptible individuals are infected as a proportion of the population, $\beta s(t) i(t) = \lambda(t) s(t)$.
+Therefore, the rate of new infections generated by a single infectious individual is $\beta s(t)$ per day.
+Multiplying by the number of infectious individuals and dividing by the population size gives the rate per day at which susceptible individuals are infected as a proportion of the population, $\beta s(t) i(t) = \lambda(t) s(t)$.
 
-% The term $\gamma i(t)$, the rate at which infected individuals (as a proportion of the total population size) recover, follows directly from the definitions of $\gamma$ and $i(t)$.
+% The term $\gamma i(t)$, the rate at which infectious individuals (as a proportion of the total population size) recover, follows directly from the definitions of $\gamma$ and $i(t)$.
 
@@
 \begin{itemize}
-    \item For any short period of time when $s(t)$ is approximately constant, the system of ODEs can be solved analytically to produce exponential growth~\autocite[section 1.2]{diekmannMathematical}.
+    \item For any short period of time when $s(t)$ is approximately constant, the solution to the system of ODEs includes exponential growth in the number of infectious individuals: $i(t) = i_0 e^{\psi t}$ for some constants $i_0$ and $\psi$~\autocite[section 1.2]{diekmannMathematical}.
     A notable example is the early part of an epidemic when $s(t) \approx 1$.
@@
     \item The basic reproduction number for the SIR model is $\R = \beta / \gamma$~\autocite[20]{keelingModeling}.
-    Intuitively, this can be thought of as the average number of infections caused by an infectious individual per day, $\beta$, multiplied by the average number of days they are infectious for, $1/\gamma$.
+    Intuitively, this can be thought of as the average number of infections caused by an infectious individual per day, $\beta$, multiplied by the average number of days for which they are infectious, $1/\gamma$.
     \item The effective reproduction number for this model is $R_e(t) = \beta / \gamma s(t) = \R s(t)$~\autocite{pellisEstimation}.
     This is the basic reproduction number reduced because some infections are avoided due to individuals being immune.
-    \item The population is closed.
-    That is, the total population size is constant, and no infections are imported from outside the population.
-    Births and deaths of individuals are also excluded, although they can easily be incorporated in situations where they are relevant~\autocites[26]{keelingModeling}[214]{kretzschmarMathematical}.
+    \item The population is assumed to be closed.
+    This assumption implies several attributes.
+    First, the total population size is constant.
+    Second, only a negligble number of infections are imported from outside the population.
+    Third, births and deaths of individuals are negligble over the length of time considered; simple extensions allow including them~\autocites[26]{keelingModeling}[214]{kretzschmarMathematical}.
     \item There is no analytical solution to the system of ODEs: numerical methods are required to solve the system~\autocite[25]{keelingModeling}.
-    \item In addition to the parameters, the state of the system at some time (normally the initial state, the state at time $t=0$) is required to solve the system.
+    \item As notes previously, the parameters $\vec{\theta}$ and the intial state of the system, $\vec{x}(0)$ is required to solve the system.
     \item Immunity is assumed to last until the end of the  period considered ~\autocite[61]{andersonInfectious}.
@@
 
-The basic SIR model implies that the distribution of the time spent in the infected compartment is exponential~\autocite[96]{keelingModeling}.
-This is an unrealistic assumption for most diseases and leads to an underestimation of the reproduction number ~\autocites{lloydRealistic}{wearingAppropriate}.
+The basic SIR model implies that the distribution of the time spent in the infectious compartment is exponential~\autocite[96]{keelingModeling}.
+This is an unrealistic assumption for most diseases and would lead to underestimating of the reproduction number~\autocites{lloydRealistic}{wearingAppropriate}.
 A standard extension to the SIR model is to allow the infectious period to follow a gamma distribution~\autocite[94]{keelingModeling}.
-This is done by adding multiple infected compartments, shown in \cref{SEIR:fig:SIR-gamma}.
+This is done by adding multiple infectious compartments, shown in \cref{SEIR:fig:SIR-gamma}.
 The system can now be written as follows:
@@
 \end{align}
-where $n$ is the number of infected compartments, and redefining $\lambda(t) = \beta \sum_{j=1}^n i_j(t)$.
+where $n$ is the number of infectious compartments, and redefining $\lambda(t) = \beta \sum_{j=1}^n i_j(t)$.
 In this model the infectious period is distributed $\GamDist(n, n\gamma)$, a realistic model for a variety of diseases~\autocite{wearingAppropriate}.
@@
 The SIR model can be extended to include a latent period by adding a latent compartment~\autocite[41]{keelingModeling}, also known as an \emph{exposed} compartment.
-This forms the susceptible-exposed-infected-recovered (SEIR) model, shown in \cref{SEIR:fig:SEIR}.
+This forms the susceptible-exposed-infectious-recovered (SEIR) model, shown in \cref{SEIR:fig:SEIR}.
 
@@
 So the force of infection is now a vector, $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i_j}(t)$.
-Its interpretation can be arrived at by considering a single infected individual in strata $a'$, analogously to the single stratum case.
+Its interpretation can be arrived at by considering a single infectious individual in strata $a'$, analogously to the single stratum case.
 Assume any such individual comes into contact with individuals in strata $a$ at a rate $\kappa_{aa'}$ in units of contacts per day.
@@
 % For such a randomly selected individual, the probability that they are susceptible is $s_a(t)$.
-% Each of the infected individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
-Following the same argument as in \cref{SEIR:sec:SIR} finds that, in expectation, the infected individual makes contact with susceptible individuals in stratum $a$ at a rate of $\kappa_{aa'} s_a(t)$.
-Assume that there is a constant probability of transmission upon contact between an infected in strata $a'$ and susceptible in strata $a$ individual, $q_{aa'}$.
-Therefore, the expected rate of infections in strata $a$ caused by this infected individual is $\kappa_{aa'} s_a(t) q_{aa'}$.
-Multiplying by the number of infected individuals in strata $a'$, $N_{a'} i_{a'}(t)$, gives $\kappa_{aa'} q_{aa'} s_a(t) N_{a'} i_{a'}(t)$ where $i_{a'}(t) = \sum_{j=1}^n i_{j,a'}$ is the total proportion of stratum $a'$ that is infectious, regardless of which compartment they are in.
+% Each of the infectious individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
+Following the same argument as in \cref{SEIR:sec:SIR} finds that, in expectation, the infectious individual makes contact with susceptible individuals in stratum $a$ at a rate of $\kappa_{aa'} s_a(t)$.
+Assume that there is a constant probability of transmission upon contact between an infectious individual in strata $a'$ and susceptible in strata $a$ individual, $q_{aa'}$.
+Therefore, the expected rate of infections in strata $a$ caused by this infectious individual is $\kappa_{aa'} s_a(t) q_{aa'}$.
+Multiplying by the number of infectious individuals in strata $a'$, $N_{a'} i_{a'}(t)$, gives $\kappa_{aa'} q_{aa'} s_a(t) N_{a'} i_{a'}(t)$ where $i_{a'}(t) = \sum_{j=1}^n i_{j,a'}$ is the total proportion of stratum $a'$ that is infectious, regardless of which compartment they are in.
 The constant terms here are absorbed into the WAIFW matrix, giving $\beta_{aa'} = \kappa_{aa'} q_{aa'} N_{a'}$~\autocite[section 9.2]{diekmannMathematical}.
@@
     Initial proportion without immunity & $\vec\pi$ & See \cref{SEIR:table:immunity-prior} & \\
-    Initial proportion infected & $i^+$ & $\BetaDist(0.5, 1000)$ & Weakly informative \\
+    Initial proportion infectious & $i^+$ & $\BetaDist(0.5, 1000)$ & Weakly informative \\
     Susceptibility of children & $q_c$ & $\LN(-0.4325, 0.1174)$ & Based on \textcite{vinerTransmission}  \\
@@
 Rather than directly inferring $\beta_1$, I infer the initial growth rate of the epidemic.
-In addition, I use a parsimonious parameterisation of the initial conditions using the initial proportion of the population that are infected and the initial proportion that are susceptible only.
+In addition, I use a parsimonious parameterisation of the initial conditions using the initial proportion of the population that are infectious and the initial proportion that are susceptible only.
 Both of these rely on an assumption that the epidemic is in a steady-state exponential growth phase at time 0 (see \cref{SEIR:sec:SIR}).
@@
 
-To form a parsimonious parameterisation of the initial state, first focus on the initial proportion of the population that is in the infected compartments.
-Recall that in the exponential growth phase, the relative proportion of each strata that is infected is proportional to the dominant eigenvector of $\matr{\beta}$, which I denote $\vec{d}$.
-Introduce a model parameter $i^+$, the proportion of the population in the $i$ states at time 0 in the strata with the highest proportion infected.
+To form a parsimonious parameterisation of the initial state, first focus on the initial proportion of the population that is in the infectious compartments.
+Recall that in the exponential growth phase, the relative proportion of each strata that is infectious is proportional to the dominant eigenvector of $\matr{\beta}$, which I denote $\vec{d}$.
+Introduce a model parameter $i^+$, the proportion of the population in the $i$ states at time 0 in the strata with the highest proportion infectious.
 Then, $\vec{i_0} = \vec{i_1} + \vec{i_2} = \frac{i^+}{\max \vec{d}} \vec{d}$.

### 2023-12-14 backcalc.tex
Add backcalc chapter

\documentclass[thesis.tex]{subfiles}

\title{Backcalculation}
\author{Joshua Blake}
\date{\today}

\begin{document}

\ifSubfilesClassLoaded{
  \setcounter{chapter}{6}
}

\chapter{Estimating incidence using backcalculation} \label{backcalc}

Now that I have estimated the duration of PCR positivity, I can estimate the incidence of SARS-CoV-2.
In this chapter, I will apply the theory introduced in \cref{E-incidence-prev} to the CIS data.
I use the period from the expansion of the CIS, starting 31st August 2020, until there was significant vaccination rollout\todo{insert date}.
First, I explain the method I will apply (\cref{})

\section{Methods}

\section{Results}

\section{Discussion}

\section{Conclusion}

\ifSubfilesClassLoaded{
  \listoftodos
}{}

\end{document}

### 2023-12-14 SEIR.tex
Compare incidence estimates notes

@@
 
+\begin{itemize}
+    \item Results here disagree with previous chapter's results.
+    \item The model in previous chapter is more flexible than here, showing the limitations discussed above. Previous chapter fits better.
+    \item The previous chapter's modelling benefits from being able to use flexible approximate inference methods that scale to many parameters, including random effects.
+Here, I am limited by the inference procedure.
+MCMC was unable to scale to the full model used in the previous chapter, which has a computationally cheaper likelihood as it is a GLM, not requiring solving an ODE.
+    \item Shows that there is a lot of information in the data which could be exploited by a suitably flexible mechanistic model.
+Such a model would be useful for policy-decisions because it could infer where transmission is occurring (\eg in schools or workplaces) and how this changes over time.
+    \item The narrower credible intervals in this chapter are a realisation of the standard bias-variance tradeoff.
+By making stronger assumptions regarding how transmission occurs, the variance is reduced.
+However, if these assumptions are incorrect (which is almost certainly the case), bias occurs.
+\end{itemize}
+
 \section{Conclusion}

### 2023-12-14 backcalc.tex
Start writing about backcalculation

@@
 
+\todo[inline]{
+    There is an asymmetry with this chapter and the following one on SEIR modelling.
+    In this chapter, the theory/background is explained in an introductory/background chapter, with a short chapter with the application-specifics and results here.
+    In the following chapter, the theory/background is explained within the chapter which means it is much longer.
+    I'm not sure if this is an issue? But this chapter seems quite short currently and the next one quite long.
+    I could move the SEIR theory to the background chapter, and possibly combine this chapter with the next one into a single chapter on incidence and transmission estimation?
+    This would make sense for allowing a comparison between the results into its own section as well, rather than currently it being placed slightly awkwardly in the results and discussion sections of the SEIR chapter.
+}
+
 Now that I have estimated the duration of PCR positivity, I can estimate the incidence of SARS-CoV-2.
@@
 I use the period from the expansion of the CIS, starting 31st August 2020, until there was significant vaccination rollout\todo{insert date}.
-First, I explain the method I will apply (\cref{})
+First, I explain the method I will apply (\cref{backcalc:sec:methods}) and then present the results (\cref{backcalc:sec:results}).
+I discuss the results in \cref{backcalc:sec:discussion} and conclude in \cref{backcalc:sec:conclustion}.
 
-\section{Methods}
+\todo[inline]{
+    I should discuss the literature on estimating incidence based on SARS-CoV-2 prevalence data.
+    In particular: \textcites{mccabeCISincidence}{abbottCISincidence}.
+    However, I think this goes better in the introduction, framing the whole thesis because it is important in understanding the motivation for the whole work.
+    At the same place, I could review the literature on mechanistic models using REACT and CIS.
+}
 
-\section{Results}
+\section{Methods} \label{backcalc:sec:methods}
 
-\section{Discussion}
+As discussed in\todo{ref section}, a major issue with the CIS is non-response bias.
+However, a deterministic relation between true prevalence and incidence is a good approximation (see \cref{E-incidence-prev:sec:prevalence-process}).
+I extend the methodology previously used to estimate prevalence from the CIS data, corrected for non-response bias, to estimate incidence\todo{cite Koen's stuff}.
+The previous method was multilevel regression and poststratification (MRP)\todo{cite MRP}.
+I fit a similar model to estimate prevalence, then transform the posterior distribution of prevalence in each stratum to a posterior distribution of incidence.
+Finally, I poststratify the posterior distribution of incidence to the CIS target population and subpopulations of interest.
 
-\section{Conclusion}
+The advantage of fitting a multilevel regression model on prevalence, rather than incidence, is that prevalence can naturally be expressed as a generalised linear model (GLM).
+A natural formulation for either incidence proportion or prevalence in a stratum $i$ is $\logit(\pi_i) = g(t)$, where $\pi_i$ is the incidence proportion or prevalence in stratum $i$ and $g$ is a function of time, penalised towards being linear.
+This is a GLM with a logit link function.
+The logit transformation serves two purposes.
+First, it ensures that the incidence proportion or prevalence is between 0 and 1.
+Second, at low values of incidence or prevalence, linear growth is approximately exponential which is the natural assumption for an epidemic and has been shown to improve estimates of prevalence\todo{cite REACT paper on this, and check the exact meaning of "improve"}.
+However, if the incidence proportion is expressed as a GLM in this way, the prevalence is not a GLM because a non-linear transformation is required to obtain prevalence from the logit of incidence proportion (the relationship on the natural scale is linear, as derived in \cref{E-incidence-prev:sec:prevalence-process}).
+Using a GLM is computationally useful due to the large number of pre-existing packages available for computing the posterior distribution, or its approximation, efficiently.
+Previously, when fitting to prevalence over long periods and including all appropriate covariates to correct for non-response bias, the model was too computationally expensive to fit using MCMC; instead, an implementation of the model using the Integrated Nested Laplace Approximation (INLA) algorithm was used~\citePersonalComms{Koen Pouwels}.
+
+I modified a pre-existing version of a MRP model for CIS~\autocite{pouwelsMRPvaccination}, as recommended by that model's author for this situation~\citePersonalComms{Koen Pouwels}.
+The model included fixed effects for age group\todo{check the model I used}.
+The likelihood is a beta-binomial distribution, a generalisation of the binomial distribution allowing for overdispersion (see \cref{E-distributions}).
+The overdispersion allows for household clustering\todo{ref beta-binomial as a model for clustering} and any other overdispersion in the data\todo{ref that epi data is often overdispersed}. 
+Finally, the model was fit independently for each region, implicitly adding interactions between region and all other variables.
+This step assumes that the estimates for each region can be considered independent, which is reasonable because the regions are large with many tests conducted within each.
+
+I used age groupings:\todo{insert age groupings}.
+These align with previous work on the CIS\todo{cite CIS reports}.
+They are based on school ages (primary, compulsory secondary, and post-compulsory secondary) for children to capture that school opening and closing may have a large effect on transmission.
+
+Having fit the prevalence model, I draw 10,000\todo{check number} prevalence trajectories in each stratum from the approximate posterior distribution of the model.
+For each stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
+Considering the relationship between incidence and prevalence to be deterministic, as \cref{E-inc-prev:sec:prevalence-process} justifies, equation implies that, for any stratum $\vec{P} = \matr{S} \vec{Z}$ where $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum, $\matr{S}$ is a matrix as defined below, and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
+The matrix $\matr{S}$ needs to be defined based on the survival function to preserve the relationship between incidence and prevalence.
+It is lower-triangular of the following form:
+\begin{align}
+    (\matr{S})_{i,j} = \begin{cases}
+        S(i - j + 1) & i - \dmax - 1 \leq j \leq i \\
+        0 & \text{otherwise.}
+    \end{cases}
+\end{align}
+Since the matrix is lower-triangular, the solution for $\vec{P}$ can be found efficiently using forward substitution\todo{cite forward substitution}.
+Forward substitution is numerically stable unless the diagonal elements are close to 0.
+Here, the diagonal is $S(1) = 1$ and hence the matrix is well-conditioned.
+\todo{I explained this more clearly in my first-year report, look that up!}
+
+The final step is poststratifying the incidence proportion to the CIS target population and subpopulations of interest.
+For each posterior draw and day\todo{introduce discrete time in days earlier}, the incidence proportion in the (sub)population is the weighted average of the incidence proportion in each stratum, where the weights are the proportion of the population in each stratum.
+The population numbers for each stratum are provided by the Office for National Statistics.
+
+\section{Results} \label{backcalc:sec:results}
+
+\section{Discussion} \label{backcalc:sec:discussion}
+
+\section{Conclusion} \label{backcalc:sec:conclustion}
 

### 2023-12-15 SEIR.tex
Note copyright of copied figure

@@
 
-This chapter uses mechanistic models for estimating the transmission of SARS-CoV-2, an alternative to the statistical approach in \cref{E-backcalc}.
-Mechanistic models explicitly represent the infection process driving the spread of infectious disease, giving model's state and parameters a biological and/or epidemiological interpretation.
+In this chapter, I introduce mechanistic models for estimating the transmission of SARS-CoV-2, an alternative to the statistical approach in \cref{E-backcalc}.
+Mechanistic models explicitly represent the infection process driving the spread of infectious disease, giving the model's state and parameters a biological and/or epidemiological interpretation.
 % Of particular interest is that the basic reproduction number, $\R$, and the effective reproduction number, $R_e(t)$, which can both be calculated as functions of the model parameters; using \cref{E-backcalc}'s approach, additional modelling would be required to estimate these numbers.
@@
 This choice is appropriate for the the context this thesis considers because there were many individuals infectious with SARS-CoV-2 in this time period.
-For an introduction to stochastic models, see \textcite[chapter 6]{keelingModeling} or \textcite{birrellEvidence}.
+There are a wealth of introductory texts on stochastic models, I recommend \textcite[chapter 6]{keelingModeling} or \textcite{birrellEvidence}.
 
@@
 
-In this section, I introduce the \emph{susceptible-infectious-recovered} (SIR) model and the extensions to it I require to appropriatly model SARS-CoV-2.
+In this section, I introduce the \emph{susceptible-infectious-recovered} (SIR) model and the extensions to it I require to appropriately model SARS-CoV-2.
 
@@
     \item The effective reproduction number for this model is $R_e(t) = \beta / \gamma s(t) = \R s(t)$~\autocite{pellisEstimation}.
-    This is the basic reproduction number reduced because some infections are avoided due to individuals being immune.
+    $1-s(t)$ is the fraction of infections that are avoided because of the presence of immune individuals in the population.
     \item The population is assumed to be closed.
@@
 In this model the infectious period is distributed $\GamDist(n, n\gamma)$, a realistic model for a variety of diseases~\autocite{wearingAppropriate}.
-This parameterisation is chosen to give $1/\gamma$ as the mean infectious period (the gamma distribution's definition and its expectation are in \cref{E-distributions}).
+This parameterisation is chosen to give $1/\gamma$ as the mean infectious period (further details on the Gamma distribution are in \cref{E-distributions}).
 A gamma distribution with the first parameter equal to 1 is an exponential distribution, in which case this model collapses to the basic SIR model in \cref{SEIR:sec:SIR}.
@@
 I will assume that the disease progression (\ie length of the exposed and infectious periods) is the same for all strata, although this assumption can be relaxed.
-Stratification requires two major  model changes: extension of the state vector and description of the between-strata interactions.
+Stratification requires two major  model changes: extension of the state vector and a description of the between-strata interactions.
 
@@
 For example, in the basic SIR model, $\vec{x}(t) = (\vec{s}(t), \vec{i}(t), \vec{r}(t))^T$.
-The $a$th $(a=1 \ldots A)$?element of each vector is the proportion of strata $a$ in that compartment.
-In teh case of the susceptible compartment $\vec{s}(t) = (s_1(t), \dots, s_A(t))^T$ where $s_a(t)$ is the proportion of the $a$th strata that are susceptible at time $t$.
+Element $a \in \set{A}$ of each vector is the proportion of strata $a$ in that compartment.
+For example, $\vec{s}(t) = (s_1(t), \dots, s_A(t))^T$ where $s_a(t)$ is the proportion of the $a$th strata that are susceptible at time $t$.
 Hence, $N_a s_a(t)$ is the number of susceptible people in stratum $a$, where $N_a$ is the number of individuals in stratum $a$.
-These vectors replace the scalars in previous equations.
-Various other definitions are possible and equivalent, for example defining $s_a$ as the proportion of the total population that are both in strata $a$ and susceptible or parameterised using numbers of individuals~\autocite[57]{keelingModeling}.
-
-The interactions between strata is done by replacing the scalar $\beta$ with a \emph{Who Acquires Infection From Whom} (WAIFW) matrix, $\matr{\beta}$~\autocite[58]{keelingModeling}; its interpretation is described below.
-So the force of infection is now a vector, $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i_j}(t)$.
-Its interpretation can be arrived at by considering a single infectious individual in strata $a'$, analogously to the single stratum case.
-Assume any such individual comes into contact with individuals in strata $a$ at a rate $\kappa_{aa'}$ in units of contacts per day.
+These vectors replace the scalars in the previous equations.
+Other, equivalent formulations of this model exist within the literature.
+For example defining $s_a$ as the proportion of the total population that are both in strata $a$ and susceptible or parameterised using numbers of individuals~\autocite[57]{keelingModeling}, rather than as the proportion of strata $a$ as I have done here.
+
+The interactions between strata are included by replacing the scalar $\beta$ with a \emph{Who Acquires Infection From Whom} (WAIFW) matrix, $\matr{\beta}$~\autocite[58]{keelingModeling}.
+The force of infection is now a vector, defined as $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i_j}(t)$.
+Their interpretation can be arrived at by considering a single infectious individual in strata $a'$, analogous to the single stratum case.
+Assume any such individual comes into contact with individuals in strata $a$ at a rate $\kappa_{aa'}$ per day.
 % Further, assume that each contact is with an individual chosen uniformly at random from within the strata.
@@
 % Each of the infectious individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
-Following the same argument as in \cref{SEIR:sec:SIR} finds that, in expectation, the infectious individual makes contact with susceptible individuals in stratum $a$ at a rate of $\kappa_{aa'} s_a(t)$.
-Assume that there is a constant probability of transmission upon contact between an infectious individual in strata $a'$ and susceptible in strata $a$ individual, $q_{aa'}$.
-Therefore, the expected rate of infections in strata $a$ caused by this infectious individual is $\kappa_{aa'} s_a(t) q_{aa'}$.
+Following the same argument as in \cref{SEIR:sec:SIR} finds that, in expectation, the infectious individual makes contact with susceptible individuals in stratum $a$ at a rate of $\kappa_{aa'} s_a(t)$ per day.
+Now assume that there is a constant probability of transmission upon contact between an infectious individual in strata $a'$ and susceptible in strata $a$ individual, $q_{aa'}$.
+Therefore, the expected rate of infections in strata $a$ caused by this infectious individual is $\kappa_{aa'} s_a(t) q_{aa'}$ per day.
 Multiplying by the number of infectious individuals in strata $a'$, $N_{a'} i_{a'}(t)$, gives $\kappa_{aa'} q_{aa'} s_a(t) N_{a'} i_{a'}(t)$ where $i_{a'}(t) = \sum_{j=1}^n i_{j,a'}$ is the total proportion of stratum $a'$ that is infectious, regardless of which compartment they are in.
-The constant terms here are absorbed into the WAIFW matrix, giving $\beta_{aa'} = \kappa_{aa'} q_{aa'} N_{a'}$~\autocite[section 9.2]{diekmannMathematical}.
+The constant terms here are absorbed into the WAIFW matrix, hence $\beta_{aa'} = \kappa_{aa'} q_{aa'} N_{a'}$~\autocite[section 9.2]{diekmannMathematical}.
 The total force of infection on strata $a$ from all strata is, therefore, $\lambda_a(t) = \sum_{a' \in \set{A}} \beta_{aa'} i_{a'}(t)$.
@@
 If all $\beta_{aa'}$ are equal, then the model is equivalent to the homogeneous model (\ie without any structure).
-The rate that individuals move between the susceptible and exposed compartments is now $\vec{s}(t) \circ \vec\lambda(t)$, where $\circ$ denotes element-wise multiplication.
+The rate at which individuals move between the susceptible and exposed compartments is now $\vec{s}(t) \circ \vec\lambda(t)$, where $\circ$ denotes element-wise multiplication.
 
@@
 
-Reproduction numbers in models with structured populations need to be careful to consider the distribution of infections across the strata.
-The WAIFM allows calculation of the number of secondary infections generated by a single infection in the $a$th strata; this follows directly from its definition as $\frac{1}{\gamma} \sum_{a' \in \set{A}} \beta_{aa'}$.
+Reproduction numbers in models with structured populations need to consider the distribution of infections across the strata.
+The WAIFM allows calculation of the number of secondary infections generated by a single infection in the $a$th strata,  this follows directly from its definition as $\frac{1}{\gamma} \sum_{a' \in \set{A}} \beta_{aa'}$.
 However, how to average these contributions to calculate the basic reproduction number is not obvious.
-The standard solution to this was proposed by \textcite{diekmannDefinition} (\textcite[chapter 7]{diekmannMathematical} provides a more gentle, but still rigorous, explanation).
+The standard solution to this was proposed by \textcite{diekmannDefinition} (see \textcite[chapter 7]{diekmannMathematical} for a more gentle, but still rigorous, explanation).
 First, they show that the distribution of infections across the strata during the initial period of exponential growth (when $\vec{s} \approx \vec{1}$) relies only on $\matr{\beta}$.
@@
   \centering \includegraphics[width=\textwidth]{SEIR/contact_matrices}
-  \caption[Age-based contact matrices]{Number of contacts reported between age groups in two contact surveys.
-Left shows the number of contacts reported in the POLYMOD survey~\autocite{mossongSocial}.
-Right shows the number of contacts reported in the BBC Pandemic survey~\autocite{klepacContacts}, which only collected data on those aged over 12.
+  \caption[Age-based contact matrices]{Number of contacts reported between age groups in two contact surveys: (A) number of contacts reported in the POLYMOD survey~\autocite{mossongSocial}; (B) number of contacts reported in the BBC Pandemic survey~\autocite{klepacContacts}, which only collected data on those aged over 12.
 A clear diagonal, showing associative mixing is visible, and children have more contacts.
 Parallel diagonals off the main diagonal show mixing between parents and their children.
-Figure adapted from \textcite{klepacContacts}.}
+Figure adapted from \textcite{klepacContacts}, licensed under CC BY-NC-ND 4.0.}
   \label{SEIR:fig:age-contacts}
@@
 The formulation of this model is heavily dependent on the data available.
-This means that, unlike for transmission models, formulating a single mathematical framework is hard or impossible.
+This means that there is a wide variety of models used.
 Therefore, this section  explains the model I propose for fitting to the CIS prevalence data.
@@
 
-\todo[inline]{discuss comparisons with the backcalculation}
+\todo[inline]{discuss comparisons with the backcalculation, based on notes below (using preliminary results)}
 

### 2023-12-15 cis-perfect-testing.tex
Most simulation results for imperfect testing
First version of full text on imperfect testing
Full first draft of imperfect chapter

@@
 
-\subsection{Approach}
+\subsection{Approach} \label{perf-test:sec:approach}
 
@@
   Similarly, define $l_i^{(e)}$ and $r_i^{(e)}$ as the first and last days $e_i$ could have occurred.
-  If $i \in \set{D}$ then $n_{ia} =1$ admissible episodes occurred in individuals with test schedule $\sched{i}$, otherwise $n_{ia} = 0$.
+  If $i \in \set{D}$ then $n_{ia} =1$ admissible episodes occurred in individuals with test schedule $\sched_{i}$, otherwise $n_{ia} = 0$.
   Conditional on the episode occurring in individuals with test schedule $\sched_i$, it is admissible with probability $p_{ia} = \prob((b, e) \in \alpha_i \mid \theta)$.
@@
 By definition (see~\cpageref{perf-test:def:admissible}), we have:
+\todo{Make sure "detectable" is defined: It's really a latent quantity, which could possibly be defined in terms of true viral load.
+I think "positive probability of testing positive" under the assumption of no false positives is what I will go with, or by reference to a multi-state model type view of infection.}
 \begin{align}
@@
 &= 1 - \prob( \tau_{\sched_i}(B_i) + B_i > E_i \vee B_i \leq \min(\sched_i) ) \\
-&= 1 - \prob( \tau_{\sched_i}(B_i) > D_i , B_i > \min(\sched_i) )  - \prob( B_i \leq \min(\sched_i) ) \\
-&= 1 - \sum_{b={\min\sched_i} + 1}^T \prob( \tau_{\sched_i}(b) > D_i) \prob(B_i = b) - \sum_{t=1}^{\min\sched_i} \prob(B_i = b)\\
-&= 1 - \frac{1}{T} \sum_{b=\min(t_i)+1}^T (1 - S_\theta(\tau_{\sched_i}(b + 1))) - \frac{1}{T} \min\sched_i \\
-&= \frac{1}{T} \left(T - \sum_{b=\min\sched_i+1}^T (1 - S_\theta(\tau_{\sched_i}(b + 1))) - \min\sched_i \right) \\
+&= 1 - \prob( \tau_{\sched_i}(B_i) + 1 > D_i , B_i > \min(\sched_i) )  - \prob( B_i \leq \min(\sched_i) ) \\
+&= 1 - \sum_{b={\min\sched_i} + 1}^T \prob( \tau_{\sched_i}(b) + 1> D_i) \prob(B_i = b) - \sum_{t=1}^{\min\sched_i} \prob(B_i = b)\\
+&= 1 - \frac{1}{T} \sum_{b=\min(\sched_i)+1}^T (1 - S_\theta(\tau_{\sched_i}(b) + 1)) - \frac{1}{T} \min\sched_i \\
+&= \frac{1}{T} \left(T - \sum_{b=\min\sched_i+1}^T (1 - S_\theta(\tau_{\sched_i}(b) + 1)) - \min\sched_i \right) \\
 &= \frac{1}{T} \left(T - (T - \min\sched_i) + \sum_{b=\min\sched_i+1}^T S_\theta(\tau_{\sched_i}(b + 1)) - \min\sched_i \right) \\
-&= \frac{1}{T} \sum_{b=\min(t_i)+1}^T S_\theta(\tau_{\sched_i}(b + 1)).
+&= \frac{1}{T} \sum_{b=\min(\sched_i)+1}^T S_\theta(\tau_{\sched_i}(b) + 1). \label{perf-test:eq:piu}
 \end{align}
@@
 
-\subsection{Results}
+\subsection{Results} \label{perf-test:sec:results}
 

### 2023-12-15 biology-data.tex
bullet point introduction

@@
 Plausibly, a variety of other reasons could also lead to a false negative such as logistical issues or mislabelling of samples.
+
+\section{SARS-CoV-2 and COVID-19}
+
+Viral loads are infrequently measured exactly.
+The most common form of data obtained is cycle threshold (Ct) values from real-time quantitative  reverse-transcriptase polymerase chain reaction (PCR) tests.
+The Ct value is proportional to the negative log of the viral load which is convenient when modelling log viral load as a piecewise linear function, since Ct values themselves are piecewise linear under this assumption.
+When a host's viral load is too low (below the limit of detection of the PCR test), no Ct value is available and hence viral load can be considered censored at the limit of detection.
+The relationship between Ct and viral load is noisy and can vary based on many factors including quality of swab obtained, primers used (even between batches), and the PCR system set-up~\autocites{dahdouhCt,hanRTPCR}.
 

### 2023-12-15 introduction.tex
bullet point introduction

@@
-\section{Backcalculation}
+\documentclass[thesis.tex]{subfiles}
+
+\begin{document}
+
+\chapter{Introduction} \label{intro}
+
+\begin{itemize}
+    \item Disease burden of pandemic and ongoing burden of COVID
+    \item To mitigate its effects, need to understand incidence and transmission but cannot observe directly
+    \item Importance of surveillance
+    \item Expected future pandemics, methods should be reusable
+\end{itemize}
+
+\section{SARS-CoV-2 real-time surveillance}
+
+\begin{itemize}
+    \item Most commonly used types of data: severe events (deaths/hospitalisations), widespread testing (pillar 2), 
+    \item Advantages of designed surveys (REACT, CIS)
+    \item Use of cohorts (ATACCC)
+    \item Statistical modelling performed during the pandemic
+\end{itemize}
+
+\section{Historical context}
+
+\begin{itemize}
+    \item Infectious disease epidemiology and biostatistics heavily developed in context of HIV/AIDS epidemic
+    \item Use much methodology and concepts from this field
+\end{itemize}
+
+\subsection{Backcalculation}
 
@@
 In this context, backcalculation estimates the incidence of HIV infection from: data on the number of AIDS diagnoses, and the incubation distribution (the time between infection and AIDS diagnosis)~\autocites{brookmeyerBackcalculation}{brookmeyerMeasuring}.
-\todo{maybe move some of this to intro}
 AIDS is incurable, therefore, AIDS prevalence is equivalent to cumulative incidence.
@@
 
-\section{Durations} \label{inc-prev:sec:duration}
+\begin{itemize}
+    \item Duration is important
+    \item Mathematical details in \cref{E-inc-prev}
+\end{itemize}
+
+\subsection{Duration} \label{inc-prev:sec:duration}
 
@@
 Modelling a biomarker uses more information per observation, but requires stronger assumptions.
+
+\section{Thesis contribution}
+
+\begin{itemize}
+    \item Overall question: how can prevalence surveys best be used to understand incidence and transmission?
+    \item Inform future study design
+    \item Provide methodology
+\end{itemize}
+
+\subsection{Epidemiological contributions}
+
+\begin{itemize}
+    \item Estimate of duration of PCR-positivity: of interest for rest of thesis but also more broadly for public health (\eg interpreting PCR test results)
+    \item Use in interpreting test results
+    \item Incidence estimates from only CIS
+    \item Transmission estimates from only CIS
+\end{itemize}
+
+\subsection{Methodological contributions}
+
+\begin{itemize}
+    \item Double interval censoring with arbitrary truncation and false negatives
+    \item Combining ATACCC and CIS duration estimates
+    \item SEIR model using only prevalence surveys
+    \item Application within TRE
+\end{itemize}
+
+\subsection{Structure}
+
+Outline the chapters and how they fit together
+
+\end{document}

### 2023-12-15 biology-data.tex
Add concepts needed from todo list

@@
 
-\section{SARS-CoV-2 and COVID-19}
+Make sure I cover:
+\begin{itemize}
+  \item Relationship between Ct and viral load (and infectiousness?)
+  \item Limit of detection
+  \item The idea of being "detectable"? Check if/where/how I've used this concept in current version of text.
+  \item 
+\end{itemize}
 
@@
 
-\section{Assessment of Transmission and Contagiousness of COVID-19 in Contacts}
+\section{Studies used in this thesis}
 
-\section{Coronavirus (COVID-19) Infection Survey} \label{intro:sec:cis}
+\subsection{Assessment of Transmission and Contagiousness of COVID-19 in Contacts}
+
+\subsection{Coronavirus (COVID-19) Infection Survey} \label{intro:sec:cis}
 

### 2023-12-15 introduction.tex
Add concepts needed from todo list

@@
     \item To mitigate its effects, need to understand incidence and transmission but cannot observe directly
-    \item Importance of surveillance
+    \item Importance of surveillance objectives and concepts: incidence, transmission, basic reproduction number, effective reproduction number
     \item Expected future pandemics, methods should be reusable
@@
     \item Use of cohorts (ATACCC)
-    \item Statistical modelling performed during the pandemic
+    \item Statistical modelling performed during the pandemic. Discuss broadly but especially focus on those using CIS or REACT. Discuss my contribution in this context.
 \end{itemize}

### 2023-12-15 SEIR.tex
deal with comments

@@
 Parallel diagonals off the main diagonal show mixing between parents and their children.
-Figure adapted from \textcite{klepacContacts}, licensed under CC BY-NC-ND 4.0.}
+Figure adapted from \textcite{klepacContacts}.}
   \label{SEIR:fig:age-contacts}
@@
 
-\todo[inline]{discuss comparisons with the backcalculation, based on notes below (using preliminary results)}
-
-\begin{itemize}
-    \item Results here disagree with previous chapter's results.
-    \item The model in previous chapter is more flexible than here, showing the limitations discussed above. Previous chapter fits better.
-    \item The previous chapter's modelling benefits from being able to use flexible approximate inference methods that scale to many parameters, including random effects.
-Here, I am limited by the inference procedure.
-MCMC was unable to scale to the full model used in the previous chapter, which has a computationally cheaper likelihood as it is a GLM, not requiring solving an ODE.
-    \item Shows that there is a lot of information in the data which could be exploited by a suitably flexible mechanistic model.
-Such a model would be useful for policy-decisions because it could infer where transmission is occurring (\eg in schools or workplaces) and how this changes over time.
-    \item The narrower credible intervals in this chapter are a realisation of the standard bias-variance tradeoff.
-By making stronger assumptions regarding how transmission occurs, the variance is reduced.
-However, if these assumptions are incorrect (which is almost certainly the case), bias occurs.
-\end{itemize}
+\todo[inline]{discuss comparisons with the backcalculation}
 

### 2023-12-15 backcalc.tex
Fix section refs

@@
 Now that I have estimated the duration of PCR positivity, I can estimate the incidence of SARS-CoV-2.
-In this chapter, I will apply the theory introduced in \cref{E-incidence-prev} to the CIS data.
+In this chapter, I will apply the theory introduced in \cref{E-inc-prev} to the CIS data.
 I use the period from the expansion of the CIS, starting 31st August 2020, until there was significant vaccination rollout\todo{insert date}.
@@
 As discussed in\todo{ref section}, a major issue with the CIS is non-response bias.
-However, a deterministic relation between true prevalence and incidence is a good approximation (see \cref{E-incidence-prev:sec:prevalence-process}).
+However, a deterministic relation between true prevalence and incidence is a good approximation (see \cref{E-inc-prev:sec:prevalence-process}).
 I extend the methodology previously used to estimate prevalence from the CIS data, corrected for non-response bias, to estimate incidence\todo{cite Koen's stuff}.
@@
 Second, at low values of incidence or prevalence, linear growth is approximately exponential which is the natural assumption for an epidemic and has been shown to improve estimates of prevalence\todo{cite REACT paper on this, and check the exact meaning of "improve"}.
-However, if the incidence proportion is expressed as a GLM in this way, the prevalence is not a GLM because a non-linear transformation is required to obtain prevalence from the logit of incidence proportion (the relationship on the natural scale is linear, as derived in \cref{E-incidence-prev:sec:prevalence-process}).
+However, if the incidence proportion is expressed as a GLM in this way, the prevalence is not a GLM because a non-linear transformation is required to obtain prevalence from the logit of incidence proportion (the relationship on the natural scale is linear, as derived in \cref{E-inc-prev:sec:prevalence-process}).
 Using a GLM is computationally useful due to the large number of pre-existing packages available for computing the posterior distribution, or its approximation, efficiently.

### 2023-12-15 introduction.tex
Intro structure

--- 2023-11-06 introduction.tex
@@
+\documentclass[thesis.tex]{subfiles}
+
+\begin{document}
+
+\chapter{Introduction} \label{intro}
+
+\section{The COVID-19 pandemic and ongoing burden}
+
+\section{SARS-CoV-2 and COVID-19}
+
+Viral loads are infrequently measured exactly.
+The most common form of data obtained is cycle threshold (Ct) values from real-time quantitative  reverse-transcriptase polymerase chain reaction (PCR) tests.
+The Ct value is proportional to the negative log of the viral load which is convenient when modelling log viral load as a piecewise linear function, since Ct values themselves are piecewise linear under this assumption.
+When a host's viral load is too low (below the limit of detection of the PCR test), no Ct value is available and hence viral load can be considered censored at the limit of detection.
+The relationship between Ct and viral load is noisy and can vary based on many factors including quality of swab obtained, primers used (even between batches), and the PCR system set-up~\autocites{dahdouhCt,hanRTPCR}.
+
 \section{Backcalculation}
@@
 
-\section{Durations} \label{inc-prev:sec:duration}
+\section{Duration} \label{inc-prev:sec:duration}
 
@@
 Modelling a biomarker uses more information per observation, but requires stronger assumptions.
+
+\section{Thesis outline}
+
+\end{document}

### 2023-12-15 thesis.tex
Most simulation results for imperfect testing
First version of full text on imperfect testing
Full first draft of imperfect chapter
Good looking bibliography
Add backcalc chapter
Fix todo location
deal with comments
Intro structure

--- 2023-12-04 thesis.tex
@@
 \usepackage{caption}
+\usepackage{numprint}
 % \usepackage{microtype}
@@
 \makeatother
-\usepackage[backend=biber,style=authoryear,uniquename=false,sortcites=true,maxcitenames=2,sorting=ynt,style=apa]{biblatex}
+\usepackage[backend=biber,uniquename=false,sortcites=true,maxcitenames=2,style=apa]{biblatex}
 \addbibresource{references.bib}
@@
 \addbibresource{references.bib}
+\addbibresource{refs-custom.bib}
 \externaldocument[E-]{\subfix{thesis}}
+\externaldocument[E-]{\subfix{output}}
 
@@
 
+%% REMOVE THESE FOR FINAL VERSION WITHOUT TODOS
 \usepackage{geometry}
@@
 
-\chapter{Introduction} \label{intro}
-
-\section{The COVID-19 pandemic and ongoing burden}
-
-\section{SARS-CoV-2 and COVID-19}
-
-Viral loads are infrequently measured exactly.
-The most common form of data obtained is cycle threshold (Ct) values from real-time quantitative  reverse-transcriptase polymerase chain reaction (PCR) tests.
-The Ct value is proportional to the negative log of the viral load which is convenient when modelling log viral load as a piecewise linear function, since Ct values themselves are piecewise linear under this assumption.
-When a host's viral load is too low (below the limit of detection of the PCR test), no Ct value is available and hence viral load can be considered censored at the limit of detection.
-The relationship between Ct and viral load is noisy and can vary based on many factors including quality of swab obtained, primers used (even between batches), and the PCR system set-up~\autocites{dahdouhCt,hanRTPCR}.
-
-\section{Thesis outline}
+\subfile{introduction}
 
@@
 
-\chapter{Estimating transmission using backcalculation} \label{backcalc}
+\subfile{backcalc}
 
@@
 
+\chapter{CIS episode definitions} \label{episode-def}
+\todo[inline]{Get this from SARAH}
+
 \subfile{ATACCC-appendix-original-analysis}
 
-\chapter{CIS episode definitions} \label{episode-def}
-
-\todo[inline]{Get this from SARAH}
-
+\cref{perf-test:eq:piu}
 \end{document}

### 2023-12-20 SEIR.tex
Introduction to SEIR section

--- 2023-12-15 SEIR.tex
@@
 In this chapter, I introduce mechanistic models for estimating the transmission of SARS-CoV-2, an alternative to the statistical approach in \cref{E-backcalc}.
-Mechanistic models explicitly represent the infection process driving the spread of infectious disease, giving the model's state and parameters a biological and/or epidemiological interpretation.
+Mechanistic models have a structure which make assumptions on the biological and epidemiological mechanisms behind the infection process (see \cref{E-back-calc:sec:infection-process}), and hence the disease dynamics~\autocite{lesslerMechanistic}.
+Individuals are giving the model's state and parameters a biological and/or epidemiological interpretation.
 % Of particular interest is that the basic reproduction number, $\R$, and the effective reproduction number, $R_e(t)$, which can both be calculated as functions of the model parameters; using \cref{E-backcalc}'s approach, additional modelling would be required to estimate these numbers.
-In particular, this allows changes in transmission to be decomposed into the components of the model, \eg the number of contacts per day or the probability of transmission upon contact.
+In particular, this allows changes in transmission to be decomposed into the contributing factors, \eg the number of contacts per day or the probability of transmission upon contact.
+The ability to estimate the contribution of each of these factors can inform public health policy.
 Furthermore, mechanistic models can be used for scenario-based modelling, which simulates the effect of interventions.
+A scenario here is the effect of a proposed intervention (or lack thereof), introduced into the model by modifying the relevant parameters.
+The mechanistic model can then be simulated forward in time to understand the intervention's effects.
 All of these properties make mechanistic models a useful tool for understanding and controlling infectious diseases.
@@
 
-The literature on mechanistic models is vast, having been developed for over a century; the model in \cref{SEIR:sec:SIR} was first formulated by \textcite{kermackContribution}.
+The literature on mechanistic models is vast, having been developed for over a century.
+The model in \cref{SEIR:sec:SIR} was first formulated by \textcite{kermackContribution}.
+However, the earliest mechanistic model to be formulated mathematically was probably \textcite{rossMalariaA}'s model of malaria transmission~\autocite{lesslerMechanistic}.
+These early contributions led to important insights for public health.
+\Textcite{rossMalariaA} showed that, if mosquitoes were controlled but not eliminated, malaria could be controlled.
+\Textcite{kermackContribution} explained the observation that epidemics ended before all individuals were infected due to a population-level build-up of immunity known as \emph{herd immunity}.
+
+A full review of this literature is beyond the scope of this thesis.
 In this chapter, I will discuss only the background relevant to the application.
@@
 
-\section{Transmission models} \label{SEIR:sec:transmission-generic}
+\section{Compartmental models} \label{SEIR:sec:transmission-generic}
 
@@
 A compartmental model consists of a state vector $\vec{x}(t)$ and a set of parameters $\vec{\theta}$~\autocite{birrellEvidence}.
+\todo{Cite other papers: see Alice's thesis (chapter 3, especially 3.1), refs within Birrell, and \href{https://www.sciencedirect.com/science/article/pii/S2468042716300495}{Allen 2017}}
 The elements of $\vec{x}(t)$ are the latent proportion of the population (or number of individuals) in each compartment at time $t$.
@@
 The basic SIR model implies that the distribution of the time spent in the infectious compartment is exponential~\autocite[96]{keelingModeling}.
+\todo{Expand on what ``exponential" means here, possibly as a mean-field approximation. See papers by Lloyd.}
 This is an unrealistic assumption for most diseases and would lead to underestimating of the reproduction number~\autocites{lloydRealistic}{wearingAppropriate}.
@@
 If the strata have differing transmission patterns, this can change the dynamics of the epidemic.
-The different patterns could be due to behavioural factors (\eg high-numbers of sexual partners in a sexually-transmitted disease~\autocite[69]{keelingModeling} or younger age groups making many more contacts for respiratory diseases~\autocite[176]{andersonInfectious}), or biological factors such as genetics~\autocite[208]{andersonInfectious}.
+The different patterns could be due to behavioural factors (\eg high-numbers of sexual partners in a sexually-transmitted disease~\autocite[69]{keelingModeling} or younger age groups making many more contacts for respiratory diseases~\autocite[176]{andersonInfectious}), or biological factors such as genetics~\autocite[208]{andersonInfectious}..
+I denote the set of strata, as $\set{A}$, with the set's size as $A$.
 I will assume that the disease progression (\ie length of the exposed and infectious periods) is the same for all strata, although this assumption can be relaxed.
@@
 
-The state vector is now a vector of vectors, one for each of the compartments.
-I denote the set of strata, as $\set{A}$, with the set's size as $A$.
+The state vector is now a vector of vectors, one for each of the compartments
 For example, in the basic SIR model, $\vec{x}(t) = (\vec{s}(t), \vec{i}(t), \vec{r}(t))^T$.
@@
 \begin{figure}
+  \todo[inline]{From Dani: increase the size, particularly the title. Add (A) BBC to left (B) POLYMOD to right and refer to these in the  caption. Why "all"?}
   \centering \includegraphics[width=\textwidth]{SEIR/contact_matrices}
@@
 Therefore, this section  explains the model I propose for fitting to the CIS prevalence data.
-
+{\color{red} After introducing the observational model in generic terms above (section 8.2) I would see a reorganisation of the section 8.3 as follows:   
+\begin{itemize}
+    \item 8.3 (currently  titled Implementation) CIS-based transmission model (see comments on lines 484-486) 
+\item 8.3.1 WAIFW matrix
+\item 8.3.2 Full model (perhaps another title..model structure?) 
+\item 8.3.3 Observational model (currently section 8.2) include here the "Solving the system of ODEs" (currently in section 8.3.4)
+\item 8.3.4 Priors 
+\item 8.3.5 Inference 
+\item 8.4 Simulation study
+\end{itemize}
+}
 % Here, I extend the system of ODEs to include compartments representing PCR positive individuals.
@@
 
-This parameterisation takes values for the contact rates, $\kappa_{aa'}(t)$, from several other data sources, using pre-existing methodology~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}.
+This parameterisation takes information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using pre-existing methodology~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}, which exploits
 %; these estimates have been used for previous modelling of the COVID-19 pandemic~\autocites{vanleeuwenTime}{birrellRealtime}.
 % These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
-First, take pre-pandemic baseline estimates of contat rates, from the POLYMOD study~\autocite{mossongSocial}, and time use, from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
-POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on the contact's age and location (home, work, school, leisure, transport, or other).
+pre-pandemic baseline estimates from the POLYMOD study~\autocite{mossongSocial}, and time use, from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
+POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on the contacts' age and location (home, work, school, leisure, transport, or other).
 UKTUS asked a representative sample of the UK population about how much time they spent doing various activities each day.
@@
 \Textcite{birrellRealtime} then modify these contact rates based on measured changes in time use during the pandemic.
-The changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
-Google mobility data measures the location of Android users over time, relative to a pre-pandemic baseline.
-These data sources are then used to down or upweight the number of contacts in each setting.
-These estimates of contacts by setting are then recombined across settings to from $\kappa_{aa'}$
+These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
+Google mobility data provides the location of Android users over time, relative to a pre-pandemic baseline.
+These data sources are then used to down or up-weight the number of contacts in each setting.
+Resulting estimates of contacts by setting are then recombined across settings to form $\kappa_{aa'}$
 For further details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.
@@
 However, these are not captured by the contact survey.
-To allow for these changes and uncertainty produced by the procedure above, I follow \textcite{birrellRealtime} in placing a random walk prior on $\beta(t)$ in log space on the weekly timescale:
+To allow for these changes as well as uncertainty produced by the procedure above, I follow \textcite{birrellRealtime} in placing a random walk prior on $log\beta(t)$  on the weekly timescale:
 \begin{align}
@@
 \end{align}
-where $w(t)$ is the week number of time $t$, with week 1 being the first week modelled.
-The remaining parameters $\beta_1$, the initial random of the random walk, and $\sigma_\epsilon$, the standard deviation of the random walk, will have priors placed on them (see \cref{SEIR:sec:priors}), with their values inferred in the usual way.
+where $w(t)$ is the week number of time $t$, with week 1 being the first week considered.
+The remaining parameters: $\beta_1$, the initial value of the random walk; and $\sigma_\epsilon$, the standard deviation of the random walk, will have priors placed on them (see \cref{SEIR:sec:priors}), with their values inferred in the usual way.
 
@@
 
-$q_{aa'}$ remains to be specified.
-I again follow the approach of \textcite{birrellRealtime}, decomposing it into two independent components such that $q_{aa'} = q^\text{susc}_{a} q^\text{inf}_{a'}$.
-$q^\text{susc}_a$ is the \emph{susceptibility} of stratum $a$, \ie how vulnerable to infection individuals in stratum $a$ are.
+Also $q_{aa'}$ remains to be specified.
+I again follow the approach of \textcite{birrellRealtime}, decomposing it into two independent components such that $q_{aa'} = q^\text{susc}_{a} q^\text{inf}_{a'}$, where
+$q^\text{susc}_a$ is the \emph{susceptibility} of stratum $a$, \ie how vulnerable to infection individuals in stratum $a$ are; and
 $q^\text{inf}_{a'}$ is the \emph{transmissibility} of stratum $a'$, \ie how likely individuals in stratum $a$ are to pass on an infection.
@@
 
-As well as the motivation given previously, independent regions allows heterogeneity in the parameters between regions; this heterogeneity could arise from a range of socioeconomic factors.
+As well as the motivation given previously, independent regions allows heterogeneity in the parameters between regions, which could arise from a range of socioeconomic factors.
 Furthermore, the assumption of homogeneous mixing within age groups is more plausible for regions than the whole of England.
@@
 }
-  \caption[SEIR model]{SEIR model used for the analysis in this chapter. The shaded box represents PCR-positive individuals. Shown for a single strata.}
+  \caption[SEIR model]{SEIR model used for the analysis in this chapter. The shaded box represents PCR-positive individuals. Shown for a single stratum.}
   \label{SEIR:fig:full-model}
@@
 The population is stratified by age, using a time-varying matrix $\matr{\beta}(t)$ to describe the transmission between strata and how these change over time.
-The population in the model is that of one region of England.
+
 To form whole of England estimates, each region is modelled independently with the results, as in \cref{E-backcalc}.
 
-The observation model is novel, relating the transmission to the CIS data.
+The observation model is novel, relating the latent transmission described by the model to the CIS data.
 It uses the phase-type model to represent the duration distribution of PCR positivity explained in \cref{SEIR:sec:observation}.
@@
 
-I ran a simulation study to check that the inference procedure (described in \cref{SEIR:sec:inference-implementation}) was working correctly and that the model was identifiable.
+I ran a simulation study to check that the inference procedure (\cref{SEIR:sec:inference-implementation}) was working correctly and that the model was identifiable.
 I ran 100 simulations, with parameters drawn from the priors (see \cref{SEIR:table:priors}) except that the prior on $\psi$ was $\LN(0.048, 0.0035)$.
@@
 The results in this chapter show the feasibility of fitting a mechanistic model using only data on prevalence.
-If the distributions for length-of-stay in each compartment are known, then all other model parameters are able to be estimated.
-This includes estimating the age distribution of the infections and the susceptibility of children.
+If the distributions for length-of-stay in each compartment are known, then all other model parameters can  be estimated.
+This includes the age distribution of the infections and the susceptibility of children.
 
 Unlike most prior work, this modelling is based on a data source that is representative of the general population.
-Therefore, the results are have fewer biases and informed by data to a greater extent than this work.
+Therefore, the results have fewer biases and informed by data to a greater extent than this work.
 There is little-to-no ascertainment bias in the CIS, unlike sources relying on testing behaviour (in the UK, pillar 2 testing).
-The CIS allows data-informed estimates of transmission in children, unlike sources relying on severe events that occur almost entirely in adults~\autocite{bhopalChildren}.
+The CIS allows estimates of transmission in children, unlike sources relying on severe events that occur almost entirely in adults~\autocite{bhopalChildren}.
 
@@
 The current model could be more flexible with regards to the age distribution of infections.
-This would better exploit one of the strengths of the CIS: that there is data on all age groups.
+This would better exploit one of the strengths of the CIS: the availability of data  on all age groups.
 In some regions, there appears to be a trade-off between the 11--16s and 16--25s, with the model misfitting in opposite directions (see \cref{SEIR:fig:prev-young}).
 This implies that the data would allow identification of a more flexible age-based model.
-Currently, the age distribution of infections is determined by $\matr{\beta}$, and to a lesser extent $\vec{\pi}$, the former of which only contains one inferred parameter that affects the age distribution, $\beta_c$.
+Currently, the age distribution of infections is determined by $\matr{\beta}$, and to a lesser extent $\vec{\pi}$, the former of which only contains one parameter that affects the age distribution, $\beta_c$.
 More parameters could capture the age distribution more accurately.
@@
 
-Extending the estimates here to the rest of the pandemic would clearly be of interest.
-I chose the end of the modelling period to coincide with the vaccination rollout.
+Extending the analyses carried out here to the rest of the pandemic would clearly be of interest.
+I chose the end of the period to coincide with the vaccination rollout.
 Vaccination dramatically affects the transmission dynamics by introducing some additional immunity into the population.
-Mechanistic models require explicit representation of this immunity\todo{cite something about vaccination}.
+Mechanistic models require explicit representation of this immunity\todo{cite something about vaccination}. (Birrell et al, 2023)
 

### 2023-12-20 incidence-prevalence.tex
Fix notation

--- 2023-11-21 incidence-prevalence.tex
@@
 
-This section showed that, when using a prevalence survey such as CIS, the likelihood $p(\vec{x} \mid \vec{Z})$ is very well approximated by $\prob(\vec{X} = \vec{x} \mid \vec{P} = \E(\vec{P} \mid \vec{Z}))$.
+This section showed that, when using a prevalence survey such as CIS, the likelihood $p(\vec{x} \mid \vec{z})$ is very well approximated by $p(\vec{x} \mid \vec{P} = \E(\vec{P} \mid \vec{z}))$.
 

### 2023-12-20 biology-data.tex
Outline intro

@@
 
+\section{Natural history of the disease}
+
+Possible summaries
+\begin{enumerate}
+  \item \url{https://www.nature.com/articles/s41579-022-00822-w/figures/2}
+  \item \url{https://www.nature.com/articles/s41576-021-00360-w/figures/1}
+  \item \url{https://academic.oup.com/view-large/figure/305617142/ciaa1442_fig2.jpg}
+\end{enumerate}
+
+\begin{itemize}
+  \item Phases: pre-infection positive, infectious positive, post-infectious positive, negative
+  \item Relationship between different means of infected: infectious, PCR positive, symptomatic/diseased
+\end{itemize}
+
 \section{PCR testing} \label{biology-data:sec:PCR}
+
+\emph{Describe how PCR testing works}
 
@@
 
+\subsubsection{Episodes}
+
+\emph{Explain definition in ``SARS-CoV-2 reinfection definition'' from \textcite{weiRisk}.}
+
 \emph{Intermittent negatives} are the clearest example of a false negative.

### 2023-12-20 introduction.tex
Outline intro

--- 2023-12-15 introduction.tex
@@
 \end{itemize}
+
+- Natural history of SARS-CoV-2 infection
+	- Stylised graphic of phases: infection, pre-infection positive, infectious positive, post-infectious positive, negative
+	- How it spreads
+- Why is incidence and prevalence estimation challenging?
+	- Asymptomatic infection
+	- Ascertainment bias
+- How do we estimate prevalence
+	- CIS and REACT
+- Issues with current incidence estimation
 
@@
 \begin{itemize}
+    \item Theory justifying the use of deterministic backcalculation in this setting.
+Already used in practice but without understanding when the approximation is good.
+In particular: all that is required is that the proportion of the population sampled is small, not that the prevalence or incidence is sufficiently large.
     \item Double interval censoring with arbitrary truncation and false negatives

### 2023-12-20 SEIR.tex
Rearrange text

@@
 
-\chapter{Estimating transmission using mechanistic modelling} \label{SEIR}
+\chapter{Estimating transmission} \label{transmission}
 
-In this chapter, I introduce mechanistic models for estimating the transmission of SARS-CoV-2, an alternative to the statistical approach in \cref{E-backcalc}.
-Mechanistic models have a structure which make assumptions on the biological and epidemiological mechanisms behind the infection process (see \cref{E-back-calc:sec:infection-process}), and hence the disease dynamics~\autocite{lesslerMechanistic}.
-Individuals are giving the model's state and parameters a biological and/or epidemiological interpretation.
-% Of particular interest is that the basic reproduction number, $\R$, and the effective reproduction number, $R_e(t)$, which can both be calculated as functions of the model parameters; using \cref{E-backcalc}'s approach, additional modelling would be required to estimate these numbers.
-In particular, this allows changes in transmission to be decomposed into the contributing factors, \eg the number of contacts per day or the probability of transmission upon contact.
-The ability to estimate the contribution of each of these factors can inform public health policy.
-Furthermore, mechanistic models can be used for scenario-based modelling, which simulates the effect of interventions.
-A scenario here is the effect of a proposed intervention (or lack thereof), introduced into the model by modifying the relevant parameters.
-The mechanistic model can then be simulated forward in time to understand the intervention's effects.
-All of these properties make mechanistic models a useful tool for understanding and controlling infectious diseases.
-However, they make stronger assumptions about the disease and population behaviour than statistical models.
+\section{Backcalculation} \label{backcalc}
 
-The literature on mechanistic models is vast, having been developed for over a century.
-The model in \cref{SEIR:sec:SIR} was first formulated by \textcite{kermackContribution}.
-However, the earliest mechanistic model to be formulated mathematically was probably \textcite{rossMalariaA}'s model of malaria transmission~\autocite{lesslerMechanistic}.
-These early contributions led to important insights for public health.
-\Textcite{rossMalariaA} showed that, if mosquitoes were controlled but not eliminated, malaria could be controlled.
-\Textcite{kermackContribution} explained the observation that epidemics ended before all individuals were infected due to a population-level build-up of immunity known as \emph{herd immunity}.
+Now that I have estimated the duration of PCR positivity, I can estimate the incidence of SARS-CoV-2.
+In this chapter, I will apply the theory introduced in \cref{E-inc-prev} to the CIS data.
+I use the period from the expansion of the CIS, starting 31st August 2020, until there was significant vaccination rollout\todo{insert date}.
+First, I explain the method I will apply (\cref{backcalc:sec:methods}) and then present the results (\cref{backcalc:sec:results}).
+I discuss the results in \cref{backcalc:sec:discussion} and conclude in \cref{backcalc:sec:conclustion}.
 
-A full review of this literature is beyond the scope of this thesis.
-In this chapter, I will discuss only the background relevant to the application.
-For further details I recommend the tutorial paper \textcite{kretzschmarMathematical} or the textbook \textcite{keelingModeling}.
+\todo[inline]{
+    I should discuss the literature on estimating incidence based on SARS-CoV-2 prevalence data.
+    In particular: \textcites{mccabeCISincidence}{abbottCISincidence}.
+    However, I think this goes better in the introduction, framing the whole thesis because it is important in understanding the motivation for the whole work.
+    At the same place, I could review the literature on mechanistic models using REACT and CIS.
+}
 
-I start this chapter by introducing a generic framework for mechanistic modelling of infectious diseases (\cref{SEIR:sec:transmission-generic}).
-% This framework is then applied to SARS-CoV-2 in \cref{SEIR:sec:transmission-application}.
-It describes the \emph{transmission model}, corresponding to the infection process.
-The transmission model needs to be linked to observations; in this chapter, that will be the CIS prevalence data.
-The \emph{observation model}, introduced in \cref{SEIR:sec:observation}, is this link; it incorporates both the prevalence process and the observation process.
-These two components form the mathematical model, which I then fit to data, requiring specification of an inference process, parameterisation, and priors (\cref{SEIR:sec:inference-implementation}).
-This final formulation uses the transmission model from \textcite{birrellRealtime}, but a novel observation model.
-I then use a simulation study to check this recovers the parameters successfully (\cref{SEIR:sec:sim-study}).
-Having shown that the model can recover the parameters, I then apply it to the CIS data (\cref{SEIR:sec:application}).
-Finally, I discuss the results, limitations, and extensions of this work (\cref{SEIR:sec:discussion}).
+\subsection{Methods} \label{backcalc:sec:methods}
 
-Notation in this chapter differs from the rest of this thesis.
-I adopt conventions and notation familiar to the infectious disease modelling field, although the field's notation is not consistent between sources.
-In particular, upper or lower case no longer signifies whether a variable is random or not, but this is clear from the context.
+As discussed in\todo{ref section}, a major issue with the CIS is non-response bias.
+However, a deterministic relation between true prevalence and incidence is a good approximation (see \cref{E-inc-prev:sec:prevalence-process}).
+I extend the methodology previously used to estimate prevalence from the CIS data, corrected for non-response bias, to estimate incidence\todo{cite Koen's stuff}.
+The previous method was multilevel regression and poststratification (MRP)\todo{cite MRP}.
+I fit a similar model to estimate prevalence, then transform the posterior distribution of prevalence in each stratum to a posterior distribution of incidence.
+Finally, I poststratify the posterior distribution of incidence to the CIS target population and subpopulations of interest.
 
-\section{Compartmental models} \label{SEIR:sec:transmission-generic}
+The advantage of fitting a multilevel regression model on prevalence, rather than incidence, is that prevalence can naturally be expressed as a generalised linear model (GLM).
+A natural formulation for either incidence proportion or prevalence in a stratum $i$ is $\logit(\pi_i) = g(t)$, where $\pi_i$ is the incidence proportion or prevalence in stratum $i$ and $g$ is a function of time, penalised towards being linear.
+This is a GLM with a logit link function.
+The logit transformation serves two purposes.
+First, it ensures that the incidence proportion or prevalence is between 0 and 1.
+Second, at low values of incidence or prevalence, linear growth is approximately exponential which is the natural assumption for an epidemic and has been shown to improve estimates of prevalence\todo{cite REACT paper on this, and check the exact meaning of "improve"}.
+However, if the incidence proportion is expressed as a GLM in this way, the prevalence is not a GLM because a non-linear transformation is required to obtain prevalence from the logit of incidence proportion (the relationship on the natural scale is linear, as derived in \cref{E-inc-prev:sec:prevalence-process}).
+Using a GLM is computationally useful due to the large number of pre-existing packages available for computing the posterior distribution, or its approximation, efficiently.
+Previously, when fitting to prevalence over long periods and including all appropriate covariates to correct for non-response bias, the model was too computationally expensive to fit using MCMC; instead, an implementation of the model using the Integrated Nested Laplace Approximation (INLA) algorithm was used~\citePersonalComms{Koen Pouwels}.
 
-This section introduces the class of mechanistic transmission models known as \emph{compartmental} models.
-Compartmental models are the most widely-used class of mechanistic infectious disease model.
-A compartmental model is named because they place every member of the population under study into a compartment; each compartment representing a disease state (\eg susceptible).
+I modified a pre-existing version of a MRP model for CIS~\autocite{pouwelsMRPvaccination}, as recommended by that model's author for this situation~\citePersonalComms{Koen Pouwels}.
+The model included fixed effects for age group\todo{check the model I used}.
+The likelihood is a beta-binomial distribution, a generalisation of the binomial distribution allowing for overdispersion (see \cref{E-distributions}).
+The overdispersion allows for household clustering\todo{ref beta-binomial as a model for clustering} and any other overdispersion in the data\todo{ref that epi data is often overdispersed}. 
+Finally, the model was fit independently for each region, implicitly adding interactions between region and all other variables.
+This step assumes that the estimates for each region can be considered independent, which is reasonable because the regions are large with many tests conducted within each.
 
-A compartmental model consists of a state vector $\vec{x}(t)$ and a set of parameters $\vec{\theta}$~\autocite{birrellEvidence}.
-\todo{Cite other papers: see Alice's thesis (chapter 3, especially 3.1), refs within Birrell, and \href{https://www.sciencedirect.com/science/article/pii/S2468042716300495}{Allen 2017}}
-The elements of $\vec{x}(t)$ are the latent proportion of the population (or number of individuals) in each compartment at time $t$.
-The parameters $\vec{\theta}$ determine, along with the model structure, how the state vector changes over time.
-Inference about $\vec{x}(t)$ and $\vec{\theta}$ is requires linking to data, $\vec{y}$, via an observation model, $p(\vec{y}(t) \mid \vec{x}, \vec{\theta})$ where $\vec{x}$ denotes conditioning on $\vec{x}(t)$ for all $t$~\autocite{birrellEvidence} (see \cref{SEIR:sec:observation}).
-That is, the observation model specifies the likelihood of the data given the model state.
+I used age groupings:\todo{insert age groupings}.
+These align with previous work on the CIS\todo{cite CIS reports}.
+They are based on school ages (primary, compulsory secondary, and post-compulsory secondary) for children to capture that school opening and closing may have a large effect on transmission.
 
-Compartmental models can be \emph{deterministic} or \emph{stochastic}.
-Deterministic models are defined by the state vector $\vec{x}(t)$ being a deterministic function of the parameters $\vec{\theta}$ and the initial state $\vec{x}(0)$; equivalently, $\var(\vec{x}(t) \mid \vec{x}(0), \theta) = 0$~\autocite{birrellEvidence}.
-% In most situations, the function is not analytically-tractable and requires numeric methods to solve.
-%The transmission model corresponds to the infection process (see \cref{E-inc-prev:sec:infection-process}).
-%There remains randomness in the observation model (described in \cref{SEIR:sec:observation}).
+Having fit the prevalence model, I draw 10,000\todo{check number} prevalence trajectories in each stratum from the approximate posterior distribution of the model.
+For each stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
+Considering the relationship between incidence and prevalence to be deterministic, as \cref{E-inc-prev:sec:prevalence-process} justifies, equation implies that, for any stratum $\vec{P} = \matr{S} \vec{Z}$ where $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum, $\matr{S}$ is a matrix as defined below, and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
+The matrix $\matr{S}$ needs to be defined based on the survival function to preserve the relationship between incidence and prevalence.
+It is lower-triangular of the following form:
+\begin{align}
+    (\matr{S})_{i,j} = \begin{cases}
+        S(i - j + 1) & i - \dmax - 1 \leq j \leq i \\
+        0 & \text{otherwise.}
+    \end{cases}
+\end{align}
+Since the matrix is lower-triangular, the solution for $\vec{P}$ can be found efficiently using forward substitution\todo{cite forward substitution}.
+Forward substitution is numerically stable unless the diagonal elements are close to 0.
+Here, the diagonal is $S(1) = 1$ and hence the matrix is well-conditioned.
+\todo{I explained this more clearly in my first-year report, look that up!}
 
-Stochastic models specify a probability distribution for the current state $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta})$ where $\set{H}_t = \{ \vec{x}(t') \ssep t' < t \}$ is the epidemic's history and $\vec{\theta}$ are model parameters\todo{REF stochastic models}.
-Stochastic models are most commonly specified in discrete time, using a Markov assumption.
-A common simplification is to use discrete time and a Markov assumption; \ie $\vec{x}(t)$ is only defined for integer $t$ and $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta}) = p(\vec{x}(t) \mid \vec{x}(t-1), \vec{\theta})$~\autocite{birrellEvidence}.
-This relaxes the assumption of independent increments in the counting process formulation of the infection process (see \cref{E-inc-prev:sec:infection-process}).
+The final step is poststratifying the incidence proportion to the CIS target population and subpopulations of interest.
+For each posterior draw and day\todo{introduce discrete time in days earlier}, the incidence proportion in the (sub)population is the weighted average of the incidence proportion in each stratum, where the weights are the proportion of the population in each stratum.
+The population numbers for each stratum are provided by the Office for National Statistics.
 
-Deterministic models are also known as \emph{mean-field} models because they represent the mean behaviour of a corresponding stochastic model.
-When the number of infectious individuals is large, the law of large numbers implies that the mean-field behaviour is a good approximation of the fully stochastic model~\autocite[20]{diekmannMathematical}.
-This is especially true when making noisy observations of the system, and that the stochastic differences between the stochastic and its mean are negligible compared to the observation noise.
-A formalisation of this argument underpins the mean-field approximation of the prevalence process made in \cref{E-inc-prev:sec:observation-process}.
+\subsection{Results} \label{backcalc:sec:results}
 
-The remainder of this chapter uses only deterministic models.
-This choice is appropriate for the the context this thesis considers because there were many individuals infectious with SARS-CoV-2 in this time period.
-There are a wealth of introductory texts on stochastic models, I recommend \textcite[chapter 6]{keelingModeling} or \textcite{birrellEvidence}.
+\subsection{Discussion} \label{backcalc:sec:discussion}
 
-This chapter discusses deterministic models whose evolution is specified by a system of ODEs, $d\vec{x}/dt$.
-In an ODE model, time is necessarily continuous.%, although discretised when solving (see \cref{SEIR:sec:inference-implementation}).
-For concreteness, I use units of days throughout, although the model can use any unit of time.
-The number or proportion of individuals in each compartment is also continuous; clearly, this is not true in reality but is a good approximation when the population is large.
+\section{Mechanistic modelling}
 
-In this section, I introduce the \emph{susceptible-infectious-recovered} (SIR) model and the extensions to it I require to appropriately model SARS-CoV-2.
-
-% \Cref{SEIR:sec:SIR} introduces the \emph{susceptible-infectious-recovered} (SIR) model, the simplest compartmental model for a disease where individuals are immune to reinfection following their recovery.
-% The name SIR derives from the structure of the model.
-% In particular, the model divides the population into one of three states: susceptible, infectious, or recovered.
-
-% This model has several assumptions which make it inappropriate to model SARS-CoV-2; therefore, in the remainder of this section I introduce standard extensions to the SIR model which relax these assumptions.
-% First, in \cref{SEIR:sec:non-exponential}, I relax the assumption that the infectious period is exponentially-distributed to allow gamma distributions.
-% Then, in \cref{SEIR:sec:SEIR}, I relax the assumption that individuals are infectious immediately after they are infected; it does this by extending the SIR model to the SEIR model with the additional E signifying an \emph{exposed} compartment.
-% Finally, in \cref{SEIR:sec:structured-populations}, I include heterogeneity in the population, stratifying by important characteristic(s).
-% Finally, \cref{SEIR:sec:time-varying-foi} relaxes the assumption of a constant transmission rate to allow for time-varying behavioural changes.
-
-\subsection{The SIR model} \label{SEIR:sec:SIR}
-
-\begin{figure}[h]
-\makebox[\textwidth][c]{
-\begin{tikzpicture}[
-    node distance = 2.5cm,
-    on grid,
-    auto,
-    ->,>=stealth',
-    every state/.style={draw,rectangle},
-    ]
-
-    \node[state] (S) {$s$};
-    \node[state, right=of S] (I) {$i$};
-    \node[state, right=of I] (R) {$r$};
-
-    \path (S) edge node {$\lambda(t)$} (I)
-          (I) edge node {$\gamma$} (R);
-\end{tikzpicture}
-}
-  \caption[The SIR model]{Schematic of the basic SIR model. Arrows are labelled with the rates at which individuals move between compartments.}
-  \label{SEIR:fig:SIR}
-\end{figure}
-
-The simplest compartmental model for a disease which confers immunity is the (SIR) model\todo{ref for SIR being simplest} (depicted in \cref{SEIR:fig:SIR}).
-The name SIR derives from the structure of the model.
-In particular, the model divides the population (assumed closed, an assumption discussed later) into one of three compartments: susceptible, infectious, or recovered.
-% The state of the system is the proportion of the population in each compartment.
-In the SIR model, individuals are assumed to be homogeneous except for their disease state, \ie they are otherwise identical.
-
-Individuals in the susceptible compartment are those that could be infected (they have no immunity).
-The proportion of the population that is susceptible at time $t$ is $s(t)$.
-Individuals in the infectious compartment can infect others, and eventually recover.
-The proportion of the population that is infectious at time $t$ is $i(t)$.
-Individuals in the recovered compartment are immune to the disease, and cannot be infected again.
-The proportion of the population that have recovered by time $t$ is $r(t)$.
-The model's state is $\vec{x}(t) = (s(t), i(t), r(t))^T$.
-
-The SIR model is described by the following system of ODEs~\autocite[19]{keelingModeling}:
-\begin{align}
-\frac{ds(t)}{dt} &= -\lambda(t) s(t) \\
-\frac{di(t)}{dt} &= \lambda(t) s(t) - \gamma i(t) \\
-\frac{dr(t)}{dt} &= \gamma i(t)
-\end{align}
-where $\lambda(t) = \beta i(t)$, is the \emph{force of infection}: the risk of infection for a susceptible individual at time $t$~\autocite[17]{keelingModeling}.
-The dynamics of the model are described by the parameters $\vec{\theta} = (\beta, \gamma)^T$.
-$\beta$ is the transmission rate, absorbing several terms as described below.
-$\gamma$ is the recovery rate, or the rate that infectious individuals that recover per day; hence, $1/\gamma$ is the mean infectious period~\autocite[367]{keelingModeling}.
-These parameters are assumed to be constant.
-
-% The term $\beta s(t)i(t)$ is the rate at which susceptible individuals (as a proportion of the total population size) are infected~\autocite[18]{keelingModeling}.
-% Susceptible individuals move from the $s(t)$ to $i(t)$ when they are infected.
-The expression for the force of infection, $\lambda(t)$ can be arrived at by considering a single infectious individual~\autocite[214]{kretzschmarMathematical}.
-Assume this individual, on average, comes into contact with other individuals in the population at a rate of $\kappa$.
-Further, assume the population is \emph{well-mixed}, meaning that each contact is with an individual chosen uniformly at random from the population; this is implied by assuming a homogeneous population.
-For such a randomly selected individual, the probability that they are susceptible is $s(t)$.
-% Each of the infectious individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
-Therefore, in expectation, the infectious individual makes contact with susceptible individuals at a rate of $\kappa s(t)$ per day.
-Assume that there is a constant probability of transmission upon contact between an infectious and susceptible individual, $q$.
-Combined with the above, the expected rate of infections caused by this infectious individual is $\kappa s(t) q$ per day.
-The two constant terms here (number of contacts and probability of transmission) are absorbed into the transmission rate, $\beta = \kappa q$.
-Therefore, the rate of new infections generated by a single infectious individual is $\beta s(t)$ per day.
-Multiplying by the number of infectious individuals and dividing by the population size gives the rate per day at which susceptible individuals are infected as a proportion of the population, $\beta s(t) i(t) = \lambda(t) s(t)$.
-
-% The term $\gamma i(t)$, the rate at which infectious individuals (as a proportion of the total population size) recover, follows directly from the definitions of $\gamma$ and $i(t)$.
-
-A few properties and assumptions of the SIR model are worth noting.
-\begin{itemize}
-    \item For any short period of time when $s(t)$ is approximately constant, the solution to the system of ODEs includes exponential growth in the number of infectious individuals: $i(t) = i_0 e^{\psi t}$ for some constants $i_0$ and $\psi$~\autocite[section 1.2]{diekmannMathematical}.
-    A notable example is the early part of an epidemic when $s(t) \approx 1$.
-    This is the phase of initial exponential growth of an epidemic, which produces the characteristic early epidemic curves.
-    It is in this phase that the basic reproduction number provides a good approximation of the epidemic dynamics.
-    \item The basic reproduction number for the SIR model is $\R = \beta / \gamma$~\autocite[20]{keelingModeling}.
-    Intuitively, this can be thought of as the average number of infections caused by an infectious individual per day, $\beta$, multiplied by the average number of days for which they are infectious, $1/\gamma$.
-    \item The effective reproduction number for this model is $R_e(t) = \beta / \gamma s(t) = \R s(t)$~\autocite{pellisEstimation}.
-    $1-s(t)$ is the fraction of infections that are avoided because of the presence of immune individuals in the population.
-    \item The population is assumed to be closed.
-    This assumption implies several attributes.
-    First, the total population size is constant.
-    Second, only a negligble number of infections are imported from outside the population.
-    Third, births and deaths of individuals are negligble over the length of time considered; simple extensions allow including them~\autocites[26]{keelingModeling}[214]{kretzschmarMathematical}.
-    \item There is no analytical solution to the system of ODEs: numerical methods are required to solve the system~\autocite[25]{keelingModeling}.
-    \item As notes previously, the parameters $\vec{\theta}$ and the intial state of the system, $\vec{x}(0)$ is required to solve the system.
-    \item Immunity is assumed to last until the end of the  period considered ~\autocite[61]{andersonInfectious}.
-    In many contexts, including here, this assumption is justified because the period is shorter than the length of immunity conferred~\autocite{milneImmunity}.
-    For longer time periods, waning immunity can be incorporated by allowing individuals to move from the recovered compartment back to the susceptible compartment~\autocite[40]{keelingModeling}.
-\end{itemize}
-
-\subsection{Non-exponential waiting times} \label{SEIR:sec:non-exponential}
-\begin{figure}[h]
-\makebox[\textwidth][c]{
-\begin{tikzpicture}[
-    node distance = 2.5cm,
-    on grid,
-    auto,
-    ->,>=stealth',
-    every state/.style={draw,rectangle},
-    ]
-
-    \node[state] (S) {$s$};
-    \node[state, right=of S] (I1) {$i_1$};
-    \node[state, right=of I1] (I2) {$i_2$};
-    \node[state, right=of I2, draw=none] (I3) {$\cdots$};
-    \node[state, right=of I3] (I4) {$i_n$};
-    \node[state, right=of I4] (R) {$r$};
-
-    \path (S) edge node {$\beta i$} (I1)
-          (I1) edge node {$n\gamma$} (I2)
-          (I2) edge node {$n\gamma$} (I3)
-          (I3) edge node {$n\gamma$} (I4)
-          (I4) edge node {$n\gamma$} (R);
-\end{tikzpicture}
-}
-  \caption[The SIR model with non-exponential infectious period]{Schematic of the SIR model with non-exponential infectious period. The infectious period is modified by having multiple I states.}
-  \label{SEIR:fig:SIR-gamma}
-\end{figure}
-
-The basic SIR model implies that the distribution of the time spent in the infectious compartment is exponential~\autocite[96]{keelingModeling}.
-\todo{Expand on what ``exponential" means here, possibly as a mean-field approximation. See papers by Lloyd.}
-This is an unrealistic assumption for most diseases and would lead to underestimating of the reproduction number~\autocites{lloydRealistic}{wearingAppropriate}.
-A standard extension to the SIR model is to allow the infectious period to follow a gamma distribution~\autocite[94]{keelingModeling}.
-This is done by adding multiple infectious compartments, shown in \cref{SEIR:fig:SIR-gamma}.
-The system can now be written as follows:
-\begin{align}
-\frac{ds(t)}{dt} &= -\lambda(t) s(t)\\
-\frac{di_1(t)}{dt} &= \lambda(t) s(t) - n\gamma i_1(t) \\
-\frac{di_2(t)}{dt} &= n\gamma i_1(t) - n \gamma i_2(t) \\
-&\vdots \nonumber \\
-\frac{di_n(t)}{dt} &= n\gamma i_{n-1}(t) - n \gamma i_n(t) \\
-\frac{dr(t)}{dt} &= n\gamma i_n(t)
-\end{align}
-where $n$ is the number of infectious compartments, and redefining $\lambda(t) = \beta \sum_{j=1}^n i_j(t)$.
-In this model the infectious period is distributed $\GamDist(n, n\gamma)$, a realistic model for a variety of diseases~\autocite{wearingAppropriate}.
-This parameterisation is chosen to give $1/\gamma$ as the mean infectious period (further details on the Gamma distribution are in \cref{E-distributions}).
-A gamma distribution with the first parameter equal to 1 is an exponential distribution, in which case this model collapses to the basic SIR model in \cref{SEIR:sec:SIR}.
-As $n$ increases, the between-individual variability in infectious period length reduces, tending to a constant as $n \to \infty$~\autocite{lloydRealistic}.
-
-The new compartments are biologically identical, without an interpretation differentiating the compartments.
-They should be considered as a mathematical trick so that the ODE system produces a gamma distribution rather than an interpretable quantity.
-
-% (it is slightly unclear what it means for a variable to have a distribution in a deterministic model, the clearest explanation is that this is the distribution which the ).
-
-The reproduction number in terms of the parameters is unchanged by this extension, \ie $\R = \beta / \gamma$.
-
-\subsection{The SEIR model} \label{SEIR:sec:SEIR}
-
-\begin{figure}[h]
-\makebox[\textwidth][c]{
-\begin{tikzpicture}[
-    node distance = 2cm,
-    on grid,
-    auto,
-    ->,>=stealth',
-    every state/.style={draw,rectangle},
-    % remember picture, overlay
-    ]
-
-    \node[state, xshift=-2cm] (S) {$s$};
-    \node[state, right=of S] (E1) {$e_1$};
-    \node[state, right=of E1] (E2) {$e_2$};
-    \node[state, right=of E2, draw=none] (E3) {$\cdots$};
-    \node[state, right=of E3] (E4) {$e_m$};
-    % \node[state, right=of E4] (I1) {$i_1$};
-    \node[state, below=of E1] (I1) {$i_1$};
-    \node[state, right=of I1] (I2) {$i_2$};
-    \node[state, right=of I2, draw=none] (I3) {$\cdots$};
-    \node[state, right=of I3] (I4) {$i_n$};
-    \node[state, right=of I4] (R) {$r$};
-
-    \path (S) edge node {$\beta i$} (E1)
-          (E1) edge node {$m\sigma$} (E2)
-          (E2) edge node {$m\sigma$} (E3)
-          (E3) edge node {$m\sigma$} (E4)
-          (I1) edge node {$n\gamma$} (I2)
-          (I2) edge node {$n\gamma$} (I3)
-          (I3) edge node {$n\gamma$} (I4)
-          (I4) edge node {$n\gamma$} (R);
-    \draw[->] (E4.east) -- ++(1,0) |- ([yshift=-1cm]E4.east) node [near start, right] {$m\sigma$} -|  (I1.north);
-\end{tikzpicture}
-}
-  \caption[The SEIR model]{Schematic of the SEIR model.}
-  \label{SEIR:fig:SEIR}
-\end{figure}
-
-The SIR model assumes that individuals are immediately infectious upon infection.
-This is unrealistic for most diseases, as there is a period between an individual being infected and them being infectious, known as the \emph{latent period}.
-\todo{I will probably end up defining the latent period in the intro, so should just ref that}
-Ignoring the latent period will again overestimate the reproduction number~\autocite{wearingAppropriate}.
-
-The SIR model can be extended to include a latent period by adding a latent compartment~\autocite[41]{keelingModeling}, also known as an \emph{exposed} compartment.
-This forms the susceptible-exposed-infectious-recovered (SEIR) model, shown in \cref{SEIR:fig:SEIR}.
-
-Similar to the infectious period, the latent period can be modelled as a gamma distribution by adding multiple latent compartments.
-\begin{align}
-\frac{ds}{dt} &= -\beta si \\
-\frac{de_1}{dt} &= \beta si - m\sigma e_1 \\
-\frac{de_2}{dt} &= m\sigma e_1 - m \sigma e_2 \\
-&\vdots \nonumber \\
-\frac{de_m}{dt} &= m\sigma e_{m-1} - m \sigma e_m \\
-\frac{di_1}{dt} &= m\sigma e_m - n\gamma i_1 \\
-\frac{di_2}{dt} &= n\gamma i_1 - n \gamma i_2 \\
-&\vdots \nonumber \\
-\frac{di_n}{dt} &= n\gamma i_{n-1} - n \gamma i_n \\
-\frac{dr}{dt} &= n\gamma i_n
-\end{align}
-where $m$ is the number of latent compartments and $1/\sigma$ is the mean duration of the latent period.
-The latent period is distributed $\GamDist(m, m\sigma)$ and the infectious period is still distributed $\GamDist(n, n\gamma)$.
-
-The reproduction number in terms of the parameters is unchanged by this extension, \ie $\R = \beta / \gamma$, however, the growth rate in the exponential phase will be slower than before~\autocite[41]{keelingModeling}.
-
-\subsection{Structured populations} \label{SEIR:sec:structured-populations}
-
-To relax the assumption of a homogeneous population, the population can be stratified.
-If the strata have differing transmission patterns, this can change the dynamics of the epidemic.
-The different patterns could be due to behavioural factors (\eg high-numbers of sexual partners in a sexually-transmitted disease~\autocite[69]{keelingModeling} or younger age groups making many more contacts for respiratory diseases~\autocite[176]{andersonInfectious}), or biological factors such as genetics~\autocite[208]{andersonInfectious}..
-I denote the set of strata, as $\set{A}$, with the set's size as $A$.
-I will assume that the disease progression (\ie length of the exposed and infectious periods) is the same for all strata, although this assumption can be relaxed.
-Stratification requires two major  model changes: extension of the state vector and a description of the between-strata interactions.
-
-The state vector is now a vector of vectors, one for each of the compartments
-For example, in the basic SIR model, $\vec{x}(t) = (\vec{s}(t), \vec{i}(t), \vec{r}(t))^T$.
-Element $a \in \set{A}$ of each vector is the proportion of strata $a$ in that compartment.
-For example, $\vec{s}(t) = (s_1(t), \dots, s_A(t))^T$ where $s_a(t)$ is the proportion of the $a$th strata that are susceptible at time $t$.
-Hence, $N_a s_a(t)$ is the number of susceptible people in stratum $a$, where $N_a$ is the number of individuals in stratum $a$.
-These vectors replace the scalars in the previous equations.
-Other, equivalent formulations of this model exist within the literature.
-For example defining $s_a$ as the proportion of the total population that are both in strata $a$ and susceptible or parameterised using numbers of individuals~\autocite[57]{keelingModeling}, rather than as the proportion of strata $a$ as I have done here.
-
-The interactions between strata are included by replacing the scalar $\beta$ with a \emph{Who Acquires Infection From Whom} (WAIFW) matrix, $\matr{\beta}$~\autocite[58]{keelingModeling}.
-The force of infection is now a vector, defined as $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i_j}(t)$.
-Their interpretation can be arrived at by considering a single infectious individual in strata $a'$, analogous to the single stratum case.
-Assume any such individual comes into contact with individuals in strata $a$ at a rate $\kappa_{aa'}$ per day.
-% Further, assume that each contact is with an individual chosen uniformly at random from within the strata.
-% For such a randomly selected individual, the probability that they are susceptible is $s_a(t)$.
-% Each of the infectious individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
-Following the same argument as in \cref{SEIR:sec:SIR} finds that, in expectation, the infectious individual makes contact with susceptible individuals in stratum $a$ at a rate of $\kappa_{aa'} s_a(t)$ per day.
-Now assume that there is a constant probability of transmission upon contact between an infectious individual in strata $a'$ and susceptible in strata $a$ individual, $q_{aa'}$.
-Therefore, the expected rate of infections in strata $a$ caused by this infectious individual is $\kappa_{aa'} s_a(t) q_{aa'}$ per day.
-Multiplying by the number of infectious individuals in strata $a'$, $N_{a'} i_{a'}(t)$, gives $\kappa_{aa'} q_{aa'} s_a(t) N_{a'} i_{a'}(t)$ where $i_{a'}(t) = \sum_{j=1}^n i_{j,a'}$ is the total proportion of stratum $a'$ that is infectious, regardless of which compartment they are in.
-The constant terms here are absorbed into the WAIFW matrix, hence $\beta_{aa'} = \kappa_{aa'} q_{aa'} N_{a'}$~\autocite[section 9.2]{diekmannMathematical}.
-The total force of infection on strata $a$ from all strata is, therefore, $\lambda_a(t) = \sum_{a' \in \set{A}} \beta_{aa'} i_{a'}(t)$.
-Equivalently, $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i_j}(t)$.
-If all $\beta_{aa'}$ are equal, then the model is equivalent to the homogeneous model (\ie without any structure).
-The rate at which individuals move between the susceptible and exposed compartments is now $\vec{s}(t) \circ \vec\lambda(t)$, where $\circ$ denotes element-wise multiplication.
-
-In practical situations, estimating all the $\beta_{aa'}$ parameters is infeasible, so some assumptions must be made to reduce the number of parameters~\autocite[176]{andersonInfectious}.
-It is infeasible because the progression of the epidemic depends on $\vec{\lambda}(t)$, which has dimension $A$, while $\matr{\beta}$ has dimension $A^2$.
-Common assumptions include assuming symmetry, \ie $\beta_{aa'} = \beta_{a'a}$, or that the direction of transmission is irrelevant, \ie $\beta_{aa'} = \beta_{a} \beta_{a'}$.
-% Alternatively, as in the scalar case, it may be easier to iden $\beta_{aa'}$ can be decomposed into the contact rate between strata $a$ and $a'$, $\kappa_{aa'}$ and the probability of transmission upon contact, $q_{aa'}$, and then assumptions made about the components.
-The choice of assumptions depends on the context, including the disease being studied and the data available.
-
-Reproduction numbers in models with structured populations need to consider the distribution of infections across the strata.
-The WAIFM allows calculation of the number of secondary infections generated by a single infection in the $a$th strata,  this follows directly from its definition as $\frac{1}{\gamma} \sum_{a' \in \set{A}} \beta_{aa'}$.
-However, how to average these contributions to calculate the basic reproduction number is not obvious.
-The standard solution to this was proposed by \textcite{diekmannDefinition} (see \textcite[chapter 7]{diekmannMathematical} for a more gentle, but still rigorous, explanation).
-First, they show that the distribution of infections across the strata during the initial period of exponential growth (when $\vec{s} \approx \vec{1}$) relies only on $\matr{\beta}$.
-This distribution is proportional to the \emph{dominant eigenvector} (the eigenvector corresponding to the largest eigenvalue) of $\matr{\beta}$.
-Denote by $D(\matr{\beta})$ the \emph{dominant eigenvalue} (the largest eigenvalue).
-The basic reproduction number is then defined as $\R = D(\matr{\beta}) / \gamma$.
-This $\R$ keeps the same interpretation as in the homogeneous case, \ie the average number of secondary infections caused by a single infection in a fully susceptible population, where the average is taken over the distribution implied by the dominant eigenvector.
-The important threshold that exponential growth occurs iff $\R > 1$ remains.
-
-Age-based contact patterns are important for the transmission dynamics of respiratory diseases~\autocites{birrellBayesian}{jacksonEffects}.
-Respiratory diseases can be spread by conversational contacts, which show very clear age patterns (see \cref{SEIR:fig:age-contacts}).
-In particular, the contacts are \emph{associative} (contacts are more likely between individuals of similar ages) and younger age groups have more contacts; this will tend to concentrate the epidemic in these younger age groups~\autocite[67]{keelingModeling}.
-\begin{figure}
-  \todo[inline]{From Dani: increase the size, particularly the title. Add (A) BBC to left (B) POLYMOD to right and refer to these in the  caption. Why "all"?}
-  \centering \includegraphics[width=\textwidth]{SEIR/contact_matrices}
-  \caption[Age-based contact matrices]{Number of contacts reported between age groups in two contact surveys: (A) number of contacts reported in the POLYMOD survey~\autocite{mossongSocial}; (B) number of contacts reported in the BBC Pandemic survey~\autocite{klepacContacts}, which only collected data on those aged over 12.
-A clear diagonal, showing associative mixing is visible, and children have more contacts.
-Parallel diagonals off the main diagonal show mixing between parents and their children.
-Figure adapted from \textcite{klepacContacts}, licensed under CC BY-NC-ND 4.0.}
-  \label{SEIR:fig:age-contacts}
-\end{figure}
-
-\section{Observation model} \label{SEIR:sec:observation}
-
-A variety of different approaches can be taken to link a transmission model to observations.
-The method to do so is known as the observation model.
-The formulation of this model is heavily dependent on the data available.
-This means that there is a wide variety of models used.
-Therefore, this section  explains the model I propose for fitting to the CIS prevalence data.
 {\color{red} After introducing the observational model in generic terms above (section 8.2) I would see a reorganisation of the section 8.3 as follows:   
@@
 }
+
+\todo[inline]{Redo this intro, with some high-level comments taken from the full model section (eg following Birrell et al.)}
+
+\subsection{Methods}
+
+I use the same CIS data as in \cref{E-backcalc}, again considering each region independently.
+The model will be stratified only by region and age group, so the data are aggregated to these levels.
+Only individuals aged two or older are eligible to participate in the CIS; I assume that test results in all participants aged under 11 are still representative of the whole $[0, 11)$ age group.
+
+As well as the motivation given in \cref{E-backcalc}, independent regions allows heterogeneity in the parameters between regions, which could arise from a range of socioeconomic factors.
+Furthermore, the assumption of homogeneous mixing within age groups is more plausible for regions than the whole of England.
+By modelling each region independently, I assume that the epidemics in each region do not interact.
+Clearly, this is not realistic as travel between regions of England is common and unrestricted.
+However, for a large epidemic, the non-interaction is a good approximation~\autocite{birrellRealtimea}.
+
+\subsubsection{WAIFW matrix}
+
+A parsimonious model is needed to parameterise the WAIFW matrix (see \cref{SEIR:sec:structured-populations}).
+I use the following age-based strata, which align closely with the age groupings previously used for analysis of the CIS data~\autocites[e.g.:][]{houseInferring}{walkerTracking}.
+\todo{Contrast with age groupings used in previous chapter}
+\begin{enumerate}
+    \item $[0, 11)$: children of pre-school and primary school age.
+    \item $[11, 16)$: children of compulsory secondary-school age.
+    \item $[16, 25)$: adolescents and young adults who may remain in formal education.
+    \item $[25, 50)$: younger working age adults.
+    \item $[50, 70)$: older working age adults.
+    \item $[70, \infty)$: elderly.
+\end{enumerate}
+
+I follow the WAIFW matrix parameterisation of \textcite{birrellRealtime}, with the decomposition $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'} N_{a'}$.
+This separates contact patterns, $\kappa_{aa'}$, which can vary over time; transmission probabilities $q_{aa'}$, assumed constant; a region-wide effect for time $\beta(t)$; and the population of the stratum, $N_{a'}$.
+
+This parameterisation takes information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using pre-existing methodology~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}, which exploits
+%; these estimates have been used for previous modelling of the COVID-19 pandemic~\autocites{vanleeuwenTime}{birrellRealtime}.
+% These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
+pre-pandemic baseline estimates from the POLYMOD study~\autocite{mossongSocial}, and time use, from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
+POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on the contacts' age and location (home, work, school, leisure, transport, or other).
+UKTUS asked a representative sample of the UK population about how much time they spent doing various activities each day.
+\Textcite{vanleeuwenAugmenting} combined these two datasets to estimate the rate of contacts between age groups during each activity.
+\Textcite{birrellRealtime} then modify these contact rates based on measured changes in time use during the pandemic.
+These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
+Google mobility data provides the location of Android users over time, relative to a pre-pandemic baseline.
+These data sources are then used to down or up-weight the number of contacts in each setting.
+Resulting estimates of contacts by setting are then recombined across settings to form $\kappa_{aa'}$
+For further details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.
+
+Some behavioural changes (\eg mask wearing) impact transmission.
+However, these are not captured by the contact survey.
+To allow for these changes as well as uncertainty produced by the procedure above, I follow \textcite{birrellRealtime} in placing a random walk prior on $\log\beta(t)$  on the weekly timescale:
+\begin{align}
+    \beta(t) &= \beta_{\tilde{w}(t)} \\
+    \log\beta_{w+1} &= \log\beta_w + \sigma_\epsilon \epsilon_w \\
+    \epsilon_w &\sim \N(0, 1)
+\end{align}
+where $\tilde{w}(t)$ is the week number of time $t$, with week 1 being the first week considered.
+Two hyperparameters will be inferred: $\beta_1$, the initial value of the random walk; and $\sigma_\epsilon$, the standard deviation of the random walk, will have priors placed on them (see \cref{SEIR:sec:priors}).
+
+% $K_{ij}$ and $\beta_{ij}$ appear only as a product and therefore are not individually identifiable.
+% Normally, $K_{ij}$ is assumed to be known, for example from a previous contact survey~\autocite[such as][]{mossongSocial}.
+% Ideally, all the $\beta_{ij}$ terms would be estimated, but there is normally inadequate data to do so, requiring assumptions to be made to reduce the number of parameters~\autocite[65]{keelingModeling}.
+% % $\beta_{ij}$ is then estimated, although often with a parsimonious model
+% A common assumption is that $\beta_{ij}$ can be decomposed into two independent components such that $\beta_{ij} = \beta \beta^\text{susc}_{i} \beta^\text{inf}_j$.
+% For any stratum $i$, $\beta^\text{susc}_i$ is known as the \emph{relative susceptibility} of stratum $i$.
+
+Also $q_{aa'}$ remains to be specified.
+I again follow the approach of \textcite{birrellRealtime}, decomposing it into two independent components such that $q_{aa'} = q^\text{susc}_{a} q^\text{inf}_{a'}$, where
+$q^\text{susc}_a$ is the \emph{susceptibility} of stratum $a$, \ie how vulnerable to infection individuals in stratum $a$ are; and
+$q^\text{inf}_{a'}$ is the \emph{transmissibility} of stratum $a'$, \ie how likely individuals in stratum $a$ are to pass on an infection.
+% Then $\beta^\text{susc}_i$ is the relative risk of an individual in stratum $i$ being infected, relative to an individual in stratum $a_0$.
+These measures include both biological differences, \eg due to differences in immune systems, or behavioural, \eg if infants are less hygienic than adults.
+
+%Now, we assume that $\beta_{ij} = \beta(t) \beta_{ij}$ where $\beta(t)$ is a non-strata-specific, time-varying component representing precautions taken by the population, for example masking; and $\beta_{ij}$ represents a constant probability of transmission from an infectious individual in strata $j$ to a susceptible individual in strata $i$ given contact, which could, for instance, be affected by differential susceptibility to infection for different age-groups.
+%We arbitrarily choose one $\beta_{ij}$ to be equal to 1 for identifiability and hence the other $\beta_{ij}$ values are relative to this reference class.
+
+Previous studies for pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults~\autocites{chenRole}{vinerTransmission}.
+Therefore, I set $q^\text{inf}_{a'} = 1$ for all $a'$, $q^\text{susc}_{a} = q_c$ for the youngest stratum, and $q^\text{susc}_{a} = 1$ for the remaining strata.
+$q_c$ is then estimated, although using an informative prior (see \cref{SEIR:sec:priors}).
+
+\subsubsection{Prevalence and observation models} \label{SEIR:sec:observation}
+
+The transmission model represents the infection process (introduced in \cref{E-inc-prev:sec:infection-process}.
+A variety of different approaches can be taken to link a transmission model to observations.
+In the context of CIS, this requires representing the prevalence and observation processes (introduced in \cref{E-inc-prev:sec:prevalence-process,E-inc-prev:sec:observation-process} respectively)
+The method to do so is known as the observation model.
+The formulation of this model is heavily dependent on the data available.
+This means that there is a wide variety of models used.
+Therefore, this section  explains the model I propose for fitting to the CIS prevalence data.
+
 % Here, I extend the system of ODEs to include compartments representing PCR positive individuals.
@@
 
-\section{Implementation} \label{SEIR:sec:inference-implementation}
-
-\todo[inline]{Redo intro paragraph to "implementation" section}
-
-The ultimate aim of this chapter is to infer the transmission dynamics from the CIS prevalence data.
-This section considers the practicalities of achieving this aim via mechanistic models.
-
-\subsection{WAIFW matrix}
-
-A parsimonious model is needed to parameterise the WAIFW matrix (see \cref{SEIR:sec:structured-populations}).
-I use the following age-based strata, which align closely with the age groupings previously used for analysis of the CIS data~\autocites[e.g.:][]{houseInferring}{walkerTracking}.
-\todo{Contrast with age groupings used in previous chapter}
-\begin{enumerate}
-    \item $[0, 11)$: children of pre-school and primary school age.
-    \item $[11, 16)$: children of compulsory secondary-school age.
-    \item $[16, 25)$: adolescents and young adults who may remain in formal education.
-    \item $[25, 50)$: younger working age adults.
-    \item $[50, 70)$: older working age adults.
-    \item $[70, \infty)$: elderly.
-\end{enumerate}
-
-I follow the WAIFW matrix parameterisation of \textcite{birrellRealtime}, with the decomposition $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'} N_{a'}$.
-This separates contact patterns, $\kappa_{aa'}$, which can vary over time; transmission probabilities $q_{aa'}$, assumed constant; a region-wide effect for time $\beta(t)$; and the population of the stratum, $N_{a'}$.
-
-This parameterisation takes information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using pre-existing methodology~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}, which exploits
-%; these estimates have been used for previous modelling of the COVID-19 pandemic~\autocites{vanleeuwenTime}{birrellRealtime}.
-% These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
-pre-pandemic baseline estimates from the POLYMOD study~\autocite{mossongSocial}, and time use, from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
-POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on the contacts' age and location (home, work, school, leisure, transport, or other).
-UKTUS asked a representative sample of the UK population about how much time they spent doing various activities each day.
-\Textcite{vanleeuwenAugmenting} combined these two datasets to estimate the rate of contacts between age groups during each activity.
-\Textcite{birrellRealtime} then modify these contact rates based on measured changes in time use during the pandemic.
-These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
-Google mobility data provides the location of Android users over time, relative to a pre-pandemic baseline.
-These data sources are then used to down or up-weight the number of contacts in each setting.
-Resulting estimates of contacts by setting are then recombined across settings to form $\kappa_{aa'}$
-For further details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.
-
-Some behavioural changes (\eg mask wearing) impact transmission.
-However, these are not captured by the contact survey.
-To allow for these changes as well as uncertainty produced by the procedure above, I follow \textcite{birrellRealtime} in placing a random walk prior on $log\beta(t)$  on the weekly timescale:
-\begin{align}
-    \beta(t) &= \beta_{w(t)} \\
-    \log\beta_{w+1} &= \log\beta_w + \sigma_\epsilon \epsilon_w \\
-    \epsilon_w &\sim \N(0, 1)
-\end{align}
-where $w(t)$ is the week number of time $t$, with week 1 being the first week considered.
-The remaining parameters: $\beta_1$, the initial value of the random walk; and $\sigma_\epsilon$, the standard deviation of the random walk, will have priors placed on them (see \cref{SEIR:sec:priors}), with their values inferred in the usual way.
-
-% $K_{ij}$ and $\beta_{ij}$ appear only as a product and therefore are not individually identifiable.
-% Normally, $K_{ij}$ is assumed to be known, for example from a previous contact survey~\autocite[such as][]{mossongSocial}.
-% Ideally, all the $\beta_{ij}$ terms would be estimated, but there is normally inadequate data to do so, requiring assumptions to be made to reduce the number of parameters~\autocite[65]{keelingModeling}.
-% % $\beta_{ij}$ is then estimated, although often with a parsimonious model
-% A common assumption is that $\beta_{ij}$ can be decomposed into two independent components such that $\beta_{ij} = \beta \beta^\text{susc}_{i} \beta^\text{inf}_j$.
-% For any stratum $i$, $\beta^\text{susc}_i$ is known as the \emph{relative susceptibility} of stratum $i$.
-
-Also $q_{aa'}$ remains to be specified.
-I again follow the approach of \textcite{birrellRealtime}, decomposing it into two independent components such that $q_{aa'} = q^\text{susc}_{a} q^\text{inf}_{a'}$, where
-$q^\text{susc}_a$ is the \emph{susceptibility} of stratum $a$, \ie how vulnerable to infection individuals in stratum $a$ are; and
-$q^\text{inf}_{a'}$ is the \emph{transmissibility} of stratum $a'$, \ie how likely individuals in stratum $a$ are to pass on an infection.
-% Then $\beta^\text{susc}_i$ is the relative risk of an individual in stratum $i$ being infected, relative to an individual in stratum $a_0$.
-These measures include both biological differences, \eg due to differences in immune systems, or behavioural, \eg if infants are less hygienic than adults.
-
-%Now, we assume that $\beta_{ij} = \beta(t) \beta_{ij}$ where $\beta(t)$ is a non-strata-specific, time-varying component representing precautions taken by the population, for example masking; and $\beta_{ij}$ represents a constant probability of transmission from an infectious individual in strata $j$ to a susceptible individual in strata $i$ given contact, which could, for instance, be affected by differential susceptibility to infection for different age-groups.
-%We arbitrarily choose one $\beta_{ij}$ to be equal to 1 for identifiability and hence the other $\beta_{ij}$ values are relative to this reference class.
-
-Previous studies for pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults~\autocites{chenRole}{vinerTransmission}.
-Therefore, I set $q^\text{inf}_{a'} = 1$ for all $a'$, $q^\text{susc}_{a} = q_c$ for the youngest stratum, and $q^\text{susc}_{a} = 1$ for the remaining strata.
-
-\subsection{Data}
-
-I use the same CIS data as in \cref{E-backcalc}, again considering each region independently.
-This model is stratified only by region and age group, so the data are aggregated to these levels.
-Only individuals aged two or older are eligible to participate in the CIS; I assume that test results in all participants aged under 11 are still representative of the whole $[0, 11)$ age group.
-
-As well as the motivation given previously, independent regions allows heterogeneity in the parameters between regions, which could arise from a range of socioeconomic factors.
-Furthermore, the assumption of homogeneous mixing within age groups is more plausible for regions than the whole of England.
-By modelling each region independently, I assume that the epidemics in each region do not interact.
-Clearly, this is not true as travel between regions of England is common and unrestricted.
-However, for a large epidemic, the non-interaction is a good approximation~\autocite{birrellRealtimea}.
-
-\subsection{Full model} \label{SEIR:sec:full-model}
+\subsubsection{Model equations and solution} \label{SEIR:sec:full-model}
 
@@
 
-To form whole of England estimates, each region is modelled independently with the results, as in \cref{E-backcalc}.
+To form whole of England estimates, each region is modelled independently with the results aggregated, as in \cref{E-backcalc}.
 
@@
 
-\subsection{Solving the systems of ODEs} \label{SEIR:sec:solving-ODEs}
+\subsubsection{Solving the systems of ODEs} \label{SEIR:sec:solving-ODEs}
 I solve the system of ODEs using Euler's method, discretised at timesteps of $\delta t = 1/2$ days.
@@
 
-\subsection{Priors} \label{SEIR:sec:priors}
+\subsubsection{Priors} \label{SEIR:sec:priors}
 
@@
 
-\subsection{Inference} \label{SEIR:sec:MCMC}
+\subsubsection{Inference} \label{SEIR:sec:MCMC}
 
@@
 
-\section{Simulation study} \label{SEIR:sec:sim-study}
+\subsection{Simulation study} \label{SEIR:sec:sim-study}
 
@@
 
-All fixed values, including $\matr{\kappa}$, used the values from the main analysis.
+All fixed values, including $\matr{\kappa}$, used the previously described in this section.
 This means that the simulations created realistic incidence and prevalence curves (see \cref{SEIR:fig:sim-data}).
@@
 
-\section{Application to CIS data} \label{SEIR:sec:application}
+\subsection{Application to CIS data} \label{SEIR:sec:application}
 
@@
 
-\todo[inline]{add comparison to backcalculation once I have those results}
-
-\section{Discussion} \label{SEIR:sec:discussion}
+\subsection{Discussion} \label{SEIR:sec:discussion}
 
@@
 
+\section{Comparison of results}
+
+\todo[inline]{add comparison to backcalculation once I have those results}
+
 \section{Conclusion}

### 2023-12-20 backcalc.tex
Rearrange text

--- 2023-12-14 backcalc.tex
@@
 
-Now that I have estimated the duration of PCR positivity, I can estimate the incidence of SARS-CoV-2.
-In this chapter, I will apply the theory introduced in \cref{E-incidence-prev} to the CIS data.
-I use the period from the expansion of the CIS, starting 31st August 2020, until there was significant vaccination rollout\todo{insert date}.
-First, I explain the method I will apply (\cref{})
+\todo[inline]{
+    There is an asymmetry with this chapter and the following one on SEIR modelling.
+    In this chapter, the theory/background is explained in an introductory/background chapter, with a short chapter with the application-specifics and results here.
+    In the following chapter, the theory/background is explained within the chapter which means it is much longer.
+    I'm not sure if this is an issue? But this chapter seems quite short currently and the next one quite long.
+    I could move the SEIR theory to the background chapter, and possibly combine this chapter with the next one into a single chapter on incidence and transmission estimation?
+    This would make sense for allowing a comparison between the results into its own section as well, rather than currently it being placed slightly awkwardly in the results and discussion sections of the SEIR chapter.
+}
 
-\section{Methods}
-
-\section{Results}
-
-\section{Discussion}
-
-\section{Conclusion}
+\section{Conclusion} \label{backcalc:sec:conclustion}
 

### 2023-12-20 incidence-prevalence.tex
Updates from Overleaf
Rearrange text

@@
 
+\todo[inline]{Make sure this section is coherent: a bunch of stuff from the SEIR chapter has been copied and pasted in}
+
+\subsection{Poisson processes}
+
 A commonly-used point process for infectious disease epidemiology is the time-inhomogenous Poisson process~\autocites[e.g., in the context of HIV:][]{brookmeyerMethod}{paganoHIV}{rosenbergBackcalculation}{brookmeyerBackcalculation}.
@@
 % This is justified because the variance due to sampling is much larger than the variance due to the true prevalence.
+
+\subsection{Compartmental models} \label{SEIR:sec:transmission-generic}
+
+This section introduces the class of mechanistic transmission models known as \emph{compartmental} models.
+Compartmental models are the most widely-used class of mechanistic infectious disease model.
+A compartmental model is named because they place every member of the population under study into a compartment; each compartment representing a disease state (\eg susceptible).
+
+A compartmental model consists of a state vector $\vec{x}(t)$ and a set of parameters $\vec{\theta}$~\autocite{birrellEvidence}.
+\todo{Cite other papers: see Alice's thesis (chapter 3, especially 3.1), refs within Birrell, and \href{https://www.sciencedirect.com/science/article/pii/S2468042716300495}{Allen 2017}}
+The elements of $\vec{x}(t)$ are the latent proportion of the population (or number of individuals) in each compartment at time $t$.
+The parameters $\vec{\theta}$ determine, along with the model structure, how the state vector changes over time.
+Inference about $\vec{x}(t)$ and $\vec{\theta}$ is requires linking to data, $\vec{y}$, via an observation model, $p(\vec{y}(t) \mid \vec{x}, \vec{\theta})$ where $\vec{x}$ denotes conditioning on $\vec{x}(t)$ for all $t$~\autocite{birrellEvidence} (see \cref{SEIR:sec:observation}).
+That is, the observation model specifies the likelihood of the data given the model state.
+
+Compartmental models can be \emph{deterministic} or \emph{stochastic}.
+Deterministic models are defined by the state vector $\vec{x}(t)$ being a deterministic function of the parameters $\vec{\theta}$ and the initial state $\vec{x}(0)$; equivalently, $\var(\vec{x}(t) \mid \vec{x}(0), \theta) = 0$~\autocite{birrellEvidence}.
+% In most situations, the function is not analytically-tractable and requires numeric methods to solve.
+%The transmission model corresponds to the infection process (see \cref{E-inc-prev:sec:infection-process}).
+%There remains randomness in the observation model (described in \cref{SEIR:sec:observation}).
+
+Stochastic models specify a probability distribution for the current state $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta})$ where $\set{H}_t = \{ \vec{x}(t') \ssep t' < t \}$ is the epidemic's history and $\vec{\theta}$ are model parameters\todo{REF stochastic models}.
+Stochastic models are most commonly specified in discrete time, using a Markov assumption.
+A common simplification is to use discrete time and a Markov assumption; \ie $\vec{x}(t)$ is only defined for integer $t$ and $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta}) = p(\vec{x}(t) \mid \vec{x}(t-1), \vec{\theta})$~\autocite{birrellEvidence}.
+This relaxes the assumption of independent increments in the counting process formulation of the infection process (see \cref{E-inc-prev:sec:infection-process}).
+
+Deterministic models are also known as \emph{mean-field} models because they represent the mean behaviour of a corresponding stochastic model.
+When the number of infectious individuals is large, the law of large numbers implies that the mean-field behaviour is a good approximation of the fully stochastic model~\autocite[20]{diekmannMathematical}.
+This is especially true when making noisy observations of the system, and that the stochastic differences between the stochastic and its mean are negligible compared to the observation noise.
+A formalisation of this argument underpins the mean-field approximation of the prevalence process made in \cref{E-inc-prev:sec:observation-process}.
+
+The remainder of this chapter uses only deterministic models.
+This choice is appropriate for the the context this thesis considers because there were many individuals infectious with SARS-CoV-2 in this time period.
+There are a wealth of introductory texts on stochastic models, I recommend \textcite[chapter 6]{keelingModeling} or \textcite{birrellEvidence}.
+
+This chapter discusses deterministic models whose evolution is specified by a system of ODEs, $d\vec{x}/dt$.
+In an ODE model, time is necessarily continuous.%, although discretised when solving (see \cref{SEIR:sec:inference-implementation}).
+For concreteness, I use units of days throughout, although the model can use any unit of time.
+The number or proportion of individuals in each compartment is also continuous; clearly, this is not true in reality but is a good approximation when the population is large.
+
+In this section, I introduce the \emph{susceptible-infectious-recovered} (SIR) model and the extensions to it I require to appropriately model SARS-CoV-2.
+
+% \Cref{SEIR:sec:SIR} introduces the \emph{susceptible-infectious-recovered} (SIR) model, the simplest compartmental model for a disease where individuals are immune to reinfection following their recovery.
+% The name SIR derives from the structure of the model.
+% In particular, the model divides the population into one of three states: susceptible, infectious, or recovered.
+
+% This model has several assumptions which make it inappropriate to model SARS-CoV-2; therefore, in the remainder of this section I introduce standard extensions to the SIR model which relax these assumptions.
+% First, in \cref{SEIR:sec:non-exponential}, I relax the assumption that the infectious period is exponentially-distributed to allow gamma distributions.
+% Then, in \cref{SEIR:sec:SEIR}, I relax the assumption that individuals are infectious immediately after they are infected; it does this by extending the SIR model to the SEIR model with the additional E signifying an \emph{exposed} compartment.
+% Finally, in \cref{SEIR:sec:structured-populations}, I include heterogeneity in the population, stratifying by important characteristic(s).
+% Finally, \cref{SEIR:sec:time-varying-foi} relaxes the assumption of a constant transmission rate to allow for time-varying behavioural changes.
+
+\subsubsection{The SIR model} \label{SEIR:sec:SIR}
+
+\begin{figure}[h]
+\makebox[\textwidth][c]{
+\begin{tikzpicture}[
+    node distance = 2.5cm,
+    on grid,
+    auto,
+    ->,>=stealth',
+    every state/.style={draw,rectangle},
+    ]
+
+    \node[state] (S) {$s$};
+    \node[state, right=of S] (I) {$i$};
+    \node[state, right=of I] (R) {$r$};
+
+    \path (S) edge node {$\lambda(t)$} (I)
+          (I) edge node {$\gamma$} (R);
+\end{tikzpicture}
+}
+  \caption[The SIR model]{Schematic of the basic SIR model. Arrows are labelled with the rates at which individuals move between compartments.}
+  \label{SEIR:fig:SIR}
+\end{figure}
+
+The simplest compartmental model for a disease which confers immunity is the (SIR) model\todo{ref for SIR being simplest} (depicted in \cref{SEIR:fig:SIR}).
+The name SIR derives from the structure of the model.
+In particular, the model divides the population (assumed closed, an assumption discussed later) into one of three compartments: susceptible, infectious, or recovered.
+% The state of the system is the proportion of the population in each compartment.
+In the SIR model, individuals are assumed to be homogeneous except for their disease state, \ie they are otherwise identical.
+
+Individuals in the susceptible compartment are those that could be infected (they have no immunity).
+The proportion of the population that is susceptible at time $t$ is $s(t)$.
+Individuals in the infectious compartment can infect others, and eventually recover.
+The proportion of the population that is infectious at time $t$ is $i(t)$.
+Individuals in the recovered compartment are immune to the disease, and cannot be infected again.
+The proportion of the population that have recovered by time $t$ is $r(t)$.
+The model's state is $\vec{x}(t) = (s(t), i(t), r(t))^T$.
+
+The SIR model is described by the following system of ODEs~\autocite[19]{keelingModeling}:
+\begin{align}
+\frac{ds(t)}{dt} &= -\lambda(t) s(t) \\
+\frac{di(t)}{dt} &= \lambda(t) s(t) - \gamma i(t) \\
+\frac{dr(t)}{dt} &= \gamma i(t)
+\end{align}
+where $\lambda(t) = \beta i(t)$, is the \emph{force of infection}: the risk of infection for a susceptible individual at time $t$~\autocite[17]{keelingModeling}.
+The dynamics of the model are described by the parameters $\vec{\theta} = (\beta, \gamma)^T$.
+$\beta$ is the transmission rate, absorbing several terms as described below.
+$\gamma$ is the recovery rate, or the rate that infectious individuals that recover per day; hence, $1/\gamma$ is the mean infectious period~\autocite[367]{keelingModeling}.
+These parameters are assumed to be constant.
+
+% The term $\beta s(t)i(t)$ is the rate at which susceptible individuals (as a proportion of the total population size) are infected~\autocite[18]{keelingModeling}.
+% Susceptible individuals move from the $s(t)$ to $i(t)$ when they are infected.
+The expression for the force of infection, $\lambda(t)$ can be arrived at by considering a single infectious individual~\autocite[214]{kretzschmarMathematical}.
+Assume this individual, on average, comes into contact with other individuals in the population at a rate of $\kappa$.
+Further, assume the population is \emph{well-mixed}, meaning that each contact is with an individual chosen uniformly at random from the population; this is implied by assuming a homogeneous population.
+For such a randomly selected individual, the probability that they are susceptible is $s(t)$.
+% Each of the infectious individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
+Therefore, in expectation, the infectious individual makes contact with susceptible individuals at a rate of $\kappa s(t)$ per day.
+Assume that there is a constant probability of transmission upon contact between an infectious and susceptible individual, $q$.
+Combined with the above, the expected rate of infections caused by this infectious individual is $\kappa s(t) q$ per day.
+The two constant terms here (number of contacts and probability of transmission) are absorbed into the transmission rate, $\beta = \kappa q$.
+Therefore, the rate of new infections generated by a single infectious individual is $\beta s(t)$ per day.
+Multiplying by the number of infectious individuals and dividing by the population size gives the rate per day at which susceptible individuals are infected as a proportion of the population, $\beta s(t) i(t) = \lambda(t) s(t)$.
+
+% The term $\gamma i(t)$, the rate at which infectious individuals (as a proportion of the total population size) recover, follows directly from the definitions of $\gamma$ and $i(t)$.
+
+A few properties and assumptions of the SIR model are worth noting.
+\begin{itemize}
+    \item For any short period of time when $s(t)$ is approximately constant, the solution to the system of ODEs includes exponential growth in the number of infectious individuals: $i(t) = i_0 e^{\psi t}$ for some constants $i_0$ and $\psi$~\autocite[section 1.2]{diekmannMathematical}.
+    A notable example is the early part of an epidemic when $s(t) \approx 1$.
+    This is the phase of initial exponential growth of an epidemic, which produces the characteristic early epidemic curves.
+    It is in this phase that the basic reproduction number provides a good approximation of the epidemic dynamics.
+    \item The basic reproduction number for the SIR model is $\R = \beta / \gamma$~\autocite[20]{keelingModeling}.
+    Intuitively, this can be thought of as the average number of infections caused by an infectious individual per day, $\beta$, multiplied by the average number of days for which they are infectious, $1/\gamma$.
+    \item The effective reproduction number for this model is $R_e(t) = \beta / \gamma s(t) = \R s(t)$~\autocite{pellisEstimation}.
+    $1-s(t)$ is the fraction of infections that are avoided because of the presence of immune individuals in the population.
+    \item The population is assumed to be closed.
+    This assumption implies several attributes.
+    First, the total population size is constant.
+    Second, only a negligble number of infections are imported from outside the population.
+    Third, births and deaths of individuals are negligble over the length of time considered; simple extensions allow including them~\autocites[26]{keelingModeling}[214]{kretzschmarMathematical}.
+    \item There is no analytical solution to the system of ODEs: numerical methods are required to solve the system~\autocite[25]{keelingModeling}.
+    \item As notes previously, the parameters $\vec{\theta}$ and the intial state of the system, $\vec{x}(0)$ is required to solve the system.
+    \item Immunity is assumed to last until the end of the  period considered ~\autocite[61]{andersonInfectious}.
+    In many contexts, including here, this assumption is justified because the period is shorter than the length of immunity conferred~\autocite{milneImmunity}.
+    For longer time periods, waning immunity can be incorporated by allowing individuals to move from the recovered compartment back to the susceptible compartment~\autocite[40]{keelingModeling}.
+\end{itemize}
+
+\subsubsection{Non-exponential waiting times} \label{SEIR:sec:non-exponential}
+\begin{figure}[h]
+\makebox[\textwidth][c]{
+\begin{tikzpicture}[
+    node distance = 2.5cm,
+    on grid,
+    auto,
+    ->,>=stealth',
+    every state/.style={draw,rectangle},
+    ]
+
+    \node[state] (S) {$s$};
+    \node[state, right=of S] (I1) {$i_1$};
+    \node[state, right=of I1] (I2) {$i_2$};
+    \node[state, right=of I2, draw=none] (I3) {$\cdots$};
+    \node[state, right=of I3] (I4) {$i_n$};
+    \node[state, right=of I4] (R) {$r$};
+
+    \path (S) edge node {$\beta i$} (I1)
+          (I1) edge node {$n\gamma$} (I2)
+          (I2) edge node {$n\gamma$} (I3)
+          (I3) edge node {$n\gamma$} (I4)
+          (I4) edge node {$n\gamma$} (R);
+\end{tikzpicture}
+}
+  \caption[The SIR model with non-exponential infectious period]{Schematic of the SIR model with non-exponential infectious period. The infectious period is modified by having multiple I states.}
+  \label{SEIR:fig:SIR-gamma}
+\end{figure}
+
+The basic SIR model implies that the distribution of the time spent in the infectious compartment is exponential~\autocite[96]{keelingModeling}.
+\todo{Expand on what ``exponential" means here, possibly as a mean-field approximation. See papers by Lloyd.}
+This is an unrealistic assumption for most diseases and would lead to underestimating of the reproduction number~\autocites{lloydRealistic}{wearingAppropriate}.
+A standard extension to the SIR model is to allow the infectious period to follow a gamma distribution~\autocite[94]{keelingModeling}.
+This is done by adding multiple infectious compartments, shown in \cref{SEIR:fig:SIR-gamma}.
+The system can now be written as follows:
+\begin{align}
+\frac{ds(t)}{dt} &= -\lambda(t) s(t)\\
+\frac{di_1(t)}{dt} &= \lambda(t) s(t) - n\gamma i_1(t) \\
+\frac{di_2(t)}{dt} &= n\gamma i_1(t) - n \gamma i_2(t) \\
+&\vdots \nonumber \\
+\frac{di_n(t)}{dt} &= n\gamma i_{n-1}(t) - n \gamma i_n(t) \\
+\frac{dr(t)}{dt} &= n\gamma i_n(t)
+\end{align}
+where $n$ is the number of infectious compartments, and redefining $\lambda(t) = \beta \sum_{j=1}^n i_j(t)$.
+In this model the infectious period is distributed $\GamDist(n, n\gamma)$, a realistic model for a variety of diseases~\autocite{wearingAppropriate}.
+This parameterisation is chosen to give $1/\gamma$ as the mean infectious period (further details on the Gamma distribution are in \cref{E-distributions}).
+A gamma distribution with the first parameter equal to 1 is an exponential distribution, in which case this model collapses to the basic SIR model in \cref{SEIR:sec:SIR}.
+As $n$ increases, the between-individual variability in infectious period length reduces, tending to a constant as $n \to \infty$~\autocite{lloydRealistic}.
+
+The new compartments are biologically identical, without an interpretation differentiating the compartments.
+They should be considered as a mathematical trick so that the ODE system produces a gamma distribution rather than an interpretable quantity.
+
+% (it is slightly unclear what it means for a variable to have a distribution in a deterministic model, the clearest explanation is that this is the distribution which the ).
+
+The reproduction number in terms of the parameters is unchanged by this extension, \ie $\R = \beta / \gamma$.
+
+\subsubsection{The SEIR model} \label{SEIR:sec:SEIR}
+
+\begin{figure}[h]
+\makebox[\textwidth][c]{
+\begin{tikzpicture}[
+    node distance = 2cm,
+    on grid,
+    auto,
+    ->,>=stealth',
+    every state/.style={draw,rectangle},
+    % remember picture, overlay
+    ]
+
+    \node[state, xshift=-2cm] (S) {$s$};
+    \node[state, right=of S] (E1) {$e_1$};
+    \node[state, right=of E1] (E2) {$e_2$};
+    \node[state, right=of E2, draw=none] (E3) {$\cdots$};
+    \node[state, right=of E3] (E4) {$e_m$};
+    % \node[state, right=of E4] (I1) {$i_1$};
+    \node[state, below=of E1] (I1) {$i_1$};
+    \node[state, right=of I1] (I2) {$i_2$};
+    \node[state, right=of I2, draw=none] (I3) {$\cdots$};
+    \node[state, right=of I3] (I4) {$i_n$};
+    \node[state, right=of I4] (R) {$r$};
+
+    \path (S) edge node {$\beta i$} (E1)
+          (E1) edge node {$m\sigma$} (E2)
+          (E2) edge node {$m\sigma$} (E3)
+          (E3) edge node {$m\sigma$} (E4)
+          (I1) edge node {$n\gamma$} (I2)
+          (I2) edge node {$n\gamma$} (I3)
+          (I3) edge node {$n\gamma$} (I4)
+          (I4) edge node {$n\gamma$} (R);
+    \draw[->] (E4.east) -- ++(1,0) |- ([yshift=-1cm]E4.east) node [near start, right] {$m\sigma$} -|  (I1.north);
+\end{tikzpicture}
+}
+  \caption[The SEIR model]{Schematic of the SEIR model.}
+  \label{SEIR:fig:SEIR}
+\end{figure}
+
+The SIR model assumes that individuals are immediately infectious upon infection.
+This is unrealistic for most diseases, as there is a period between an individual being infected and them being infectious, known as the \emph{latent period}.
+\todo{I will probably end up defining the latent period in the intro, so should just ref that}
+Ignoring the latent period will again overestimate the reproduction number~\autocite{wearingAppropriate}.
+
+The SIR model can be extended to include a latent period by adding a latent compartment~\autocite[41]{keelingModeling}, also known as an \emph{exposed} compartment.
+This forms the susceptible-exposed-infectious-recovered (SEIR) model, shown in \cref{SEIR:fig:SEIR}.
+
+Similar to the infectious period, the latent period can be modelled as a gamma distribution by adding multiple latent compartments.
+\begin{align}
+\frac{ds}{dt} &= -\beta si \\
+\frac{de_1}{dt} &= \beta si - m\sigma e_1 \\
+\frac{de_2}{dt} &= m\sigma e_1 - m \sigma e_2 \\
+&\vdots \nonumber \\
+\frac{de_m}{dt} &= m\sigma e_{m-1} - m \sigma e_m \\
+\frac{di_1}{dt} &= m\sigma e_m - n\gamma i_1 \\
+\frac{di_2}{dt} &= n\gamma i_1 - n \gamma i_2 \\
+&\vdots \nonumber \\
+\frac{di_n}{dt} &= n\gamma i_{n-1} - n \gamma i_n \\
+\frac{dr}{dt} &= n\gamma i_n
+\end{align}
+where $m$ is the number of latent compartments and $1/\sigma$ is the mean duration of the latent period.
+The latent period is distributed $\GamDist(m, m\sigma)$ and the infectious period is still distributed $\GamDist(n, n\gamma)$.
+
+The reproduction number in terms of the parameters is unchanged by this extension, \ie $\R = \beta / \gamma$, however, the growth rate in the exponential phase will be slower than before~\autocite[41]{keelingModeling}.
+
+\subsubsection{Structured populations} \label{SEIR:sec:structured-populations}
+
+To relax the assumption of a homogeneous population, the population can be stratified.
+If the strata have differing transmission patterns, this can change the dynamics of the epidemic.
+The different patterns could be due to behavioural factors (\eg high-numbers of sexual partners in a sexually-transmitted disease~\autocite[69]{keelingModeling} or younger age groups making many more contacts for respiratory diseases~\autocite[176]{andersonInfectious}), or biological factors such as genetics~\autocite[208]{andersonInfectious}..
+I denote the set of strata, as $\set{A}$, with the set's size as $A$.
+I will assume that the disease progression (\ie length of the exposed and infectious periods) is the same for all strata, although this assumption can be relaxed.
+Stratification requires two major  model changes: extension of the state vector and a description of the between-strata interactions.
+
+The state vector is now a vector of vectors, one for each of the compartments
+For example, in the basic SIR model, $\vec{x}(t) = (\vec{s}(t), \vec{i}(t), \vec{r}(t))^T$.
+Element $a \in \set{A}$ of each vector is the proportion of strata $a$ in that compartment.
+For example, $\vec{s}(t) = (s_1(t), \dots, s_A(t))^T$ where $s_a(t)$ is the proportion of the $a$th strata that are susceptible at time $t$.
+Hence, $N_a s_a(t)$ is the number of susceptible people in stratum $a$, where $N_a$ is the number of individuals in stratum $a$.
+These vectors replace the scalars in the previous equations.
+Other, equivalent formulations of this model exist within the literature.
+For example defining $s_a$ as the proportion of the total population that are both in strata $a$ and susceptible or parameterised using numbers of individuals~\autocite[57]{keelingModeling}, rather than as the proportion of strata $a$ as I have done here.
+
+The interactions between strata are included by replacing the scalar $\beta$ with a \emph{Who Acquires Infection From Whom} (WAIFW) matrix, $\matr{\beta}$~\autocite[58]{keelingModeling}.
+The force of infection is now a vector, defined as $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i_j}(t)$.
+Their interpretation can be arrived at by considering a single infectious individual in strata $a'$, analogous to the single stratum case.
+Assume any such individual comes into contact with individuals in strata $a$ at a rate $\kappa_{aa'}$ per day.
+% Further, assume that each contact is with an individual chosen uniformly at random from within the strata.
+% For such a randomly selected individual, the probability that they are susceptible is $s_a(t)$.
+% Each of the infectious individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
+Following the same argument as in \cref{SEIR:sec:SIR} finds that, in expectation, the infectious individual makes contact with susceptible individuals in stratum $a$ at a rate of $\kappa_{aa'} s_a(t)$ per day.
+Now assume that there is a constant probability of transmission upon contact between an infectious individual in strata $a'$ and susceptible in strata $a$ individual, $q_{aa'}$.
+Therefore, the expected rate of infections in strata $a$ caused by this infectious individual is $\kappa_{aa'} s_a(t) q_{aa'}$ per day.
+Multiplying by the number of infectious individuals in strata $a'$, $N_{a'} i_{a'}(t)$, gives $\kappa_{aa'} q_{aa'} s_a(t) N_{a'} i_{a'}(t)$ where $i_{a'}(t) = \sum_{j=1}^n i_{j,a'}$ is the total proportion of stratum $a'$ that is infectious, regardless of which compartment they are in.
+The constant terms here are absorbed into the WAIFW matrix, hence $\beta_{aa'} = \kappa_{aa'} q_{aa'} N_{a'}$~\autocite[section 9.2]{diekmannMathematical}.
+The total force of infection on strata $a$ from all strata is, therefore, $\lambda_a(t) = \sum_{a' \in \set{A}} \beta_{aa'} i_{a'}(t)$.
+Equivalently, $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i_j}(t)$.
+If all $\beta_{aa'}$ are equal, then the model is equivalent to the homogeneous model (\ie without any structure).
+The rate at which individuals move between the susceptible and exposed compartments is now $\vec{s}(t) \circ \vec\lambda(t)$, where $\circ$ denotes element-wise multiplication.
+
+In practical situations, estimating all the $\beta_{aa'}$ parameters is infeasible, so some assumptions must be made to reduce the number of parameters~\autocite[176]{andersonInfectious}.
+It is infeasible because the progression of the epidemic depends on $\vec{\lambda}(t)$, which has dimension $A$, while $\matr{\beta}$ has dimension $A^2$.
+Common assumptions include assuming symmetry, \ie $\beta_{aa'} = \beta_{a'a}$, or that the direction of transmission is irrelevant, \ie $\beta_{aa'} = \beta_{a} \beta_{a'}$.
+% Alternatively, as in the scalar case, it may be easier to iden $\beta_{aa'}$ can be decomposed into the contact rate between strata $a$ and $a'$, $\kappa_{aa'}$ and the probability of transmission upon contact, $q_{aa'}$, and then assumptions made about the components.
+The choice of assumptions depends on the context, including the disease being studied and the data available.
+
+Reproduction numbers in models with structured populations need to consider the distribution of infections across the strata.
+The WAIFM allows calculation of the number of secondary infections generated by a single infection in the $a$th strata,  this follows directly from its definition as $\frac{1}{\gamma} \sum_{a' \in \set{A}} \beta_{aa'}$.
+However, how to average these contributions to calculate the basic reproduction number is not obvious.
+The standard solution to this was proposed by \textcite{diekmannDefinition} (see \textcite[chapter 7]{diekmannMathematical} for a more gentle, but still rigorous, explanation).
+First, they show that the distribution of infections across the strata during the initial period of exponential growth (when $\vec{s} \approx \vec{1}$) relies only on $\matr{\beta}$.
+This distribution is proportional to the \emph{dominant eigenvector} (the eigenvector corresponding to the largest eigenvalue) of $\matr{\beta}$.
+Denote by $D(\matr{\beta})$ the \emph{dominant eigenvalue} (the largest eigenvalue).
+The basic reproduction number is then defined as $\R = D(\matr{\beta}) / \gamma$.
+This $\R$ keeps the same interpretation as in the homogeneous case, \ie the average number of secondary infections caused by a single infection in a fully susceptible population, where the average is taken over the distribution implied by the dominant eigenvector.
+The important threshold that exponential growth occurs iff $\R > 1$ remains.
+
+Age-based contact patterns are important for the transmission dynamics of respiratory diseases~\autocites{birrellBayesian}{jacksonEffects}.
+Respiratory diseases can be spread by conversational contacts, which show very clear age patterns (see \cref{SEIR:fig:age-contacts}).
+In particular, the contacts are \emph{associative} (contacts are more likely between individuals of similar ages) and younger age groups have more contacts; this will tend to concentrate the epidemic in these younger age groups~\autocite[67]{keelingModeling}.
+\begin{figure}
+  \todo[inline]{From Dani: increase the size, particularly the title. Add (A) BBC to left (B) POLYMOD to right and refer to these in the  caption. Why "all"?}
+  \centering \includegraphics[width=\textwidth]{SEIR/contact_matrices}
+  \caption[Age-based contact matrices]{Number of contacts reported between age groups in two contact surveys: (A) number of contacts reported in the POLYMOD survey~\autocite{mossongSocial}; (B) number of contacts reported in the BBC Pandemic survey~\autocite{klepacContacts}, which only collected data on those aged over 12.
+A clear diagonal, showing associative mixing is visible, and children have more contacts.
+Parallel diagonals off the main diagonal show mixing between parents and their children.
+Figure adapted from \textcite{klepacContacts}, licensed under CC BY-NC-ND 4.0.}
+  \label{SEIR:fig:age-contacts}
+\end{figure}
 

### 2023-12-20 introduction.tex
Rearrange text

@@
 
+\subsection{Mechanistic models}
+
+In this chapter, I introduce mechanistic models for estimating the transmission of SARS-CoV-2, an alternative to the statistical approach in \cref{E-backcalc}.
+Mechanistic models have a structure which make assumptions on the biological and epidemiological mechanisms behind the infection process (see \cref{E-back-calc:sec:infection-process}), and hence the disease dynamics~\autocite{lesslerMechanistic}.
+Individuals are giving the model's state and parameters a biological and/or epidemiological interpretation.
+% Of particular interest is that the basic reproduction number, $\R$, and the effective reproduction number, $R_e(t)$, which can both be calculated as functions of the model parameters; using \cref{E-backcalc}'s approach, additional modelling would be required to estimate these numbers.
+In particular, this allows changes in transmission to be decomposed into the contributing factors, \eg the number of contacts per day or the probability of transmission upon contact.
+The ability to estimate the contribution of each of these factors can inform public health policy.
+Furthermore, mechanistic models can be used for scenario-based modelling, which simulates the effect of interventions.
+A scenario here is the effect of a proposed intervention (or lack thereof), introduced into the model by modifying the relevant parameters.
+The mechanistic model can then be simulated forward in time to understand the intervention's effects.
+All of these properties make mechanistic models a useful tool for understanding and controlling infectious diseases.
+However, they make stronger assumptions about the disease and population behaviour than statistical models.
+
+The literature on mechanistic models is vast, having been developed for over a century.
+The model in \cref{SEIR:sec:SIR} was first formulated by \textcite{kermackContribution}.
+However, the earliest mechanistic model to be formulated mathematically was probably \textcite{rossMalariaA}'s model of malaria transmission~\autocite{lesslerMechanistic}.
+These early contributions led to important insights for public health.
+\Textcite{rossMalariaA} showed that, if mosquitoes were controlled but not eliminated, malaria could be controlled.
+\Textcite{kermackContribution} explained the observation that epidemics ended before all individuals were infected due to a population-level build-up of immunity known as \emph{herd immunity}.
+
+A full review of this literature is beyond the scope of this thesis.
+In this chapter, I will discuss only the background relevant to the application.
+For further details I recommend the tutorial paper \textcite{kretzschmarMathematical} or the textbook \textcite{keelingModeling}.
+
+I start this chapter by introducing a generic framework for mechanistic modelling of infectious diseases (\cref{SEIR:sec:transmission-generic}).
+% This framework is then applied to SARS-CoV-2 in \cref{SEIR:sec:transmission-application}.
+It describes the \emph{transmission model}, corresponding to the infection process.
+The transmission model needs to be linked to observations; in this chapter, that will be the CIS prevalence data.
+The \emph{observation model}, introduced in \cref{SEIR:sec:observation}, is this link; it incorporates both the prevalence process and the observation process.
+These two components form the mathematical model, which I then fit to data, requiring specification of an inference process, parameterisation, and priors (\cref{SEIR:sec:inference-implementation}).
+This final formulation uses the transmission model from \textcite{birrellRealtime}, but a novel observation model.
+I then use a simulation study to check this recovers the parameters successfully (\cref{SEIR:sec:sim-study}).
+Having shown that the model can recover the parameters, I then apply it to the CIS data (\cref{SEIR:sec:application}).
+Finally, I discuss the results, limitations, and extensions of this work (\cref{SEIR:sec:discussion}).
+
+Notation in this chapter differs from the rest of this thesis.
+I adopt conventions and notation familiar to the infectious disease modelling field, although the field's notation is not consistent between sources.
+In particular, upper or lower case no longer signifies whether a variable is random or not, but this is clear from the context.
+
 \subsection{Duration} \label{inc-prev:sec:duration}

### 2023-12-20 introduction.tex
Fix file names, build, refs

@@
 The sampling is described as the observation process.
-The observation process noise is much larger than the noise of the prevalence process (see \cref{inc-prev:sec:observation-process}).
+The observation process noise is much larger than the noise of the prevalence process (see \cref{E-inc-prev:sec:observation-process}).
 Therefore, the latter is negligible and again two processes need to be considered: infection and observation.
@@
 In this chapter, I introduce mechanistic models for estimating the transmission of SARS-CoV-2, an alternative to the statistical approach in \cref{E-backcalc}.
-Mechanistic models have a structure which make assumptions on the biological and epidemiological mechanisms behind the infection process (see \cref{E-back-calc:sec:infection-process}), and hence the disease dynamics~\autocite{lesslerMechanistic}.
+Mechanistic models have a structure which make assumptions on the biological and epidemiological mechanisms behind the infection process (see \cref{E-in-prev:sec:infection-process}), and hence the disease dynamics~\autocite{lesslerMechanistic}.
 Individuals are giving the model's state and parameters a biological and/or epidemiological interpretation.
@@
 The literature on mechanistic models is vast, having been developed for over a century.
-The model in \cref{SEIR:sec:SIR} was first formulated by \textcite{kermackContribution}.
+The model in \cref{E-SEIR:sec:SIR} was first formulated by \textcite{kermackContribution}.
 However, the earliest mechanistic model to be formulated mathematically was probably \textcite{rossMalariaA}'s model of malaria transmission~\autocite{lesslerMechanistic}.
@@
 
-I start this chapter by introducing a generic framework for mechanistic modelling of infectious diseases (\cref{SEIR:sec:transmission-generic}).
-% This framework is then applied to SARS-CoV-2 in \cref{SEIR:sec:transmission-application}.
+I start this chapter by introducing a generic framework for mechanistic modelling of infectious diseases (\cref{E-SEIR:sec:transmission-generic}).
+% This framework is then applied to SARS-CoV-2 in \cref{E-SEIR:sec:transmission-application}.
 It describes the \emph{transmission model}, corresponding to the infection process.
 The transmission model needs to be linked to observations; in this chapter, that will be the CIS prevalence data.
-The \emph{observation model}, introduced in \cref{SEIR:sec:observation}, is this link; it incorporates both the prevalence process and the observation process.
-These two components form the mathematical model, which I then fit to data, requiring specification of an inference process, parameterisation, and priors (\cref{SEIR:sec:inference-implementation}).
+The \emph{observation model}, introduced in \cref{E-SEIR:sec:observation}, is this link; it incorporates both the prevalence process and the observation process.
+These two components form the mathematical model, which I then fit to data, requiring specification of an inference process, parameterisation, and priors (\cref{E-SEIR:sec:inference-implementation}).
 This final formulation uses the transmission model from \textcite{birrellRealtime}, but a novel observation model.
-I then use a simulation study to check this recovers the parameters successfully (\cref{SEIR:sec:sim-study}).
-Having shown that the model can recover the parameters, I then apply it to the CIS data (\cref{SEIR:sec:application}).
-Finally, I discuss the results, limitations, and extensions of this work (\cref{SEIR:sec:discussion}).
+I then use a simulation study to check this recovers the parameters successfully (\cref{E-SEIR:sec:sim-study}).
+Having shown that the model can recover the parameters, I then apply it to the CIS data (\cref{E-SEIR:sec:application}).
+Finally, I discuss the results, limitations, and extensions of this work (\cref{E-SEIR:sec:discussion}).
 

### 2023-12-20 transmission.tex
Fix file names, build, refs

--- 2023-12-20 SEIR.tex
@@
 \ifSubfilesClassLoaded{
-    \setcounter{chapter}{7}
+    \setcounter{chapter}{6}
 }
@@
 
-This parameterisation takes information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using pre-existing methodology~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}, which exploits
+This parameterisation takes information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using pre-existing methodology~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}.
 %; these estimates have been used for previous modelling of the COVID-19 pandemic~\autocites{vanleeuwenTime}{birrellRealtime}.
 % These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
-pre-pandemic baseline estimates from the POLYMOD study~\autocite{mossongSocial}, and time use, from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
-POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on the contacts' age and location (home, work, school, leisure, transport, or other).
-UKTUS asked a representative sample of the UK population about how much time they spent doing various activities each day.
+This method takes pre-pandemic baseline estimates of: $\kappa_{aa'}$ from the POLYMOD study~\autocite{mossongSocial}, and time use from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
+POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on each contact's age and location (home, work, school, leisure, transport, or other).
+UKTUS asked a representative sample of the UK population about how much time they spent doing various activities (\eg shopping and spending time at home) each day.
 \Textcite{vanleeuwenAugmenting} combined these two datasets to estimate the rate of contacts between age groups during each activity.
@@
 These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
-Google mobility data provides the location of Android users over time, relative to a pre-pandemic baseline.
-These data sources are then used to down or up-weight the number of contacts in each setting.
-Resulting estimates of contacts by setting are then recombined across settings to form $\kappa_{aa'}$
+Google mobility data is based on the location of Android users over time, relative to a pre-pandemic baseline.
+The mobility and school data are then used to down or up-weight the number of contacts in each setting, and then recombined across settings to form $\kappa_{aa'}$
 For further details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.
@@
 However, these are not captured by the contact survey.
-To allow for these changes as well as uncertainty produced by the procedure above, I follow \textcite{birrellRealtime} in placing a random walk prior on $\log\beta(t)$  on the weekly timescale:
+To allow for these changes and the uncertainty produced in formulating $\matr{\kaapa}(t)$, I follow \textcite{birrellRealtime} in placing a random walk prior on $\log\beta(t)$  on the weekly timescale:
 \begin{align}
@@
 
-Also $q_{aa'}$ remains to be specified.
-I again follow the approach of \textcite{birrellRealtime}, decomposing it into two independent components such that $q_{aa'} = q^\text{susc}_{a} q^\text{inf}_{a'}$, where
-$q^\text{susc}_a$ is the \emph{susceptibility} of stratum $a$, \ie how vulnerable to infection individuals in stratum $a$ are; and
-$q^\text{inf}_{a'}$ is the \emph{transmissibility} of stratum $a'$, \ie how likely individuals in stratum $a$ are to pass on an infection.
+Finally,$q_{aa'}$ needs to be specified.
+I again follow the approach of \textcite{birrellRealtime}, decomposing it into two independent factors such that $q_{aa'} = q^\text{susc}_{a} q^\text{inf}_{a'}$, where $q^\text{susc}_a$ is the \emph{susceptibility} of stratum $a$, \ie how vulnerable to infection individuals in stratum $a$ are; and $q^\text{inf}_{a'}$ is the \emph{transmissibility} of stratum $a'$, \ie how likely individuals in stratum $a$ are to pass on an infection.
 % Then $\beta^\text{susc}_i$ is the relative risk of an individual in stratum $i$ being infected, relative to an individual in stratum $a_0$.
@@
 
-Overall the model fits the data well (see \cref{SEIR:fig:prev-young,SEIR:fig:prev-old}).
-There are a few features of the data which are not captured by the model.
-In particular, prevalence during October is higher in the 16--25s than the model predicts and lower in the 11--16s.
-\begin{figure}
-    \thisfloatpagestyle{empty}
-    \vspace{-3cm}
-    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_young}}
-    \captionsetup{width=0.8\paperwidth}
-    \captionof{figure}[SEIR prevalence goodness-of-fit (younger ages)]{%
-        Goodness-of-fit to prevalence data by region and age for the younger three age groups (see \cref{SEIR:fig:prev-old} for the others).
-        The central ribbon and line shows the posterior median and 95\% CrI estimate for the proportion of the relevant strata PCR positive on each day.
-        The lighter outer ribbon is the 95\% CrI including the noise due to the sampling (\ie observation process), aggregated by week.
-        The dots show the observed prevalence in CIS, aggregated by week.
-        For a well-calibrated model, 95\% of the dots would be within the outer ribbon.
-    }
-    \label{SEIR:fig:prev-young}
-\end{figure}
-\begin{figure}
-    \thisfloatpagestyle{empty}
-    \vspace{-3cm}
-    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_old}}
-    \captionsetup{width=0.8\paperwidth}
-    \captionof{figure}[SEIR prevalence goodness-of-fit (older ages)]{%
-        Goodness-of-fit prevalence for the older three age groups (see \cref{SEIR:fig:prev-young} for the other age groups and details).
-    }
-    \label{SEIR:fig:prev-old}
-\end{figure}
-
 The 11--16 age group (roughly corresponding to compulsory secondary school education) is estimated to have the highest incidence (see \cref{SEIR:fig:incidence}).
@@
 
-The parameters $\beta_t$ absorb all the time-varying changes in transmission not captured by Google mobility and school attendance data.
+The parameters $\beta_w$ absorb all the time-varying changes in transmission not captured by Google mobility and school attendance data.
 They trend downwards up until early December when they start increasing again.
@@
     \includegraphics{SEIR/CIS/beta_walk}
-    \caption[Posterior estimates of $\beta_t$]{%
-        Posterior estimates $\beta_t$ for each week.
+    \caption[Posterior estimates of $\beta_w$]{%
+        Posterior estimates $\beta_w$ for each week.
         Shown are ``violins'' where the width is proportional to the kernel-density estimated posterior density at that point.
@@
 \end{figure}
-
-One MCMC chain for the North West region appeared to get stuck, with very few moves after the first 100,000 iterations.
-I reran this chain with a new seed.
-Following this, all regions' MCMC chains converged, assessed using the Rhat statistic and ESS.
-Rhat for all parameters is < 1.01, except $\sigma_\epsilon$ in London which had a Rhat of 1.018.
-All parameters had ESS > 1000, except $\sigma_\epsilon$ in London and the East, which had ESS of 197 and 757 respectively.
-As all the parameters of primary interest had high Rhat and ESS, and no parameters had large issues, the inference procedure is considered to have converged.
 
@@
 
+Overall the model fits the data well (see \cref{SEIR:fig:prev-young,SEIR:fig:prev-old}).
+There are a few features of the data which are not captured by the model.
+In particular, prevalence during October is higher in the 16--25s than the model predicts and lower in the 11--16s.
+\begin{figure}
+    \thisfloatpagestyle{empty}
+    \vspace{-3cm}
+    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_young}}
+    \captionsetup{width=0.8\paperwidth}
+    \captionof{figure}[SEIR prevalence goodness-of-fit (younger ages)]{%
+        Goodness-of-fit to prevalence data by region and age for the younger three age groups (see \cref{SEIR:fig:prev-old} for the others).
+        The central ribbon and line shows the posterior median and 95\% CrI estimate for the proportion of the relevant strata PCR positive on each day.
+        The lighter outer ribbon is the 95\% posterior predictive interval for the proportion of test results that are positive each week.
+        The dots show the observed prevalence in CIS, aggregated by week.
+        For a well-calibrated model, 95\% of the dots would be within the outer ribbon.
+    }
+    \label{SEIR:fig:prev-young}
+\end{figure}
+\begin{figure}
+    \thisfloatpagestyle{empty}
+    \vspace{-3cm}
+    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_old}}
+    \captionsetup{width=0.8\paperwidth}
+    \captionof{figure}[SEIR prevalence goodness-of-fit (older ages)]{%
+        Goodness-of-fit prevalence for the older three age groups (see \cref{SEIR:fig:prev-young} for the other age groups and details).
+    }
+    \label{SEIR:fig:prev-old}
+\end{figure}
+
+One MCMC chain for the North West region appeared to get stuck, with very few moves after the first 100,000 iterations (not shown).
+I reran this chain with a new seed, with the rerun results presented.
+Following this, all regions' MCMC chains converged, assessed using the Rhat statistic and ESS.
+Rhat for all parameters is < 1.01, except $\sigma_\epsilon$ in London which had a Rhat of 1.018.
+All parameters had ESS > 1000, except $\sigma_\epsilon$ in London and the East, which had ESS of 197 and 757 respectively.
+As all the parameters of primary interest had high Rhat and ESS, and no parameters had large issues, the inference procedure is considered to have converged.
+
 \subsection{Discussion} \label{SEIR:sec:discussion}
@@
 
-Unlike most prior work, this modelling is based on a data source that is representative of the general population.
-Therefore, the results have fewer biases and informed by data to a greater extent than this work.
-There is little-to-no ascertainment bias in the CIS, unlike sources relying on testing behaviour (in the UK, pillar 2 testing).
-The CIS allows estimates of transmission in children, unlike sources relying on severe events that occur almost entirely in adults~\autocite{bhopalChildren}.
+% Unlike most prior work (see\todo{ref section with lit review}), this modelling is based on a data source that is representative of the general population.
+% % Therefore, the results have fewer biases and informed by data to a greater extent than this work.
+% There is little-to-no ascertainment bias in the CIS, unlike sources relying on testing behaviour (in the UK, pillar 2 testing).
+% The CIS allows estimates of transmission in children, unlike sources relying on severe events that occur almost entirely in adults~\autocite{bhopalChildren}.
 
-This approach provides additional understanding, compared to the approach in \cref{E-backcalc}.
-For example, decomposing the change in transmission into changing contacts, $\matr{\kappa}(t)$, changing susceptibility $\vec{s}(t)$, and other unmeasured factors, $\beta(t)$.
-By understanding which of these components is changing, policies can be better targeted.
-If it is shown that $\beta(t)$ decreases, this implies that for a given amount of activity in the population, there is less transmission.
-Such decreases might be beneficial as it allows society to operate in a more normal manner without increasing transmission.
-Breaking the trade-off between transmission and activity is desirable because it reduces the disease burden without burdesome restrictions on individuals day-to-day lives.
-
-The parameters $\beta_t$ absorb all the time-varying factors not captured by Google mobility and school attendance data.
-Many factors contribute to the pattern in the $\beta_t$ values, which have a minima near the start of December.
+The parameters $\beta_w$ absorb all the time-varying factors not captured by Google mobility and school attendance data.
+Many factors contribute to the pattern in the $\beta_w$ values, which have a minima near the start of December.
 While this analysis cannot establish causation, several hypotheses can be identified that are worth further investigation.
-The decreases in $\beta_t$ over the autumn could be due to increased caution(\eg masking or stricter adherence to social distancing) and/or interventions short of a lockdown as cases and deaths increased~\autocite{jarvisEffect}, or the impact of local measures in the highest transmission areas~\autocite{scottCovid19}.
+The decreases in $\beta_w$ over the autumn could be due to increased caution(\eg masking or stricter adherence to social distancing) and/or interventions short of a lockdown as cases and deaths increased~\autocite{jarvisEffect}, or the impact of local measures in the highest transmission areas~\autocite{scottCovid19}.
 Local measures are not well captured by $\matr{\kappa}$ because the matrix was for the whole of England.
 Allowing each region to have its own $\matr{\kappa}$ may capture the effect of local measures more accurately.
-The increases in $\beta_t$ in December coincide with the rise of the Alpha variant~\autocite{lythgoeLineage}, known to be more transmissible~\autocite{daviesEstimated}.
+The increases in $\beta_w$ in December coincide with the rise of the Alpha variant~\autocite{lythgoeLineage}, known to be more transmissible~\autocite{daviesEstimated}.
 A multiple variant model might be able to capture this rise more accurately.
@@
 
-These factors are correlated across the country.
-This can explain that similar patterns emerge for $\beta_t$ in different regions (see \cref{SEIR:fig:beta-walk}), despite these being estimated independently (\ie there is no model structure which favours correlated values).
+These hypotheses for the mechanism causing changes in $\beta_w$ are correlated across the country.
+This can explain that similar patterns emerge for $\beta_w$ in different regions (see \cref{SEIR:fig:beta-walk}), despite these being estimated independently (\ie there is no model structure which favours correlated values).
 These correlations could be exploited to improve the precision of estimates with a more complex model which borrows information across regions.
@@
 This discussion makes clear that the results are limited by the number of parameters, currently limited by the inference algorithm.
+\todo{Rephrase this sentence}
 The algorithm used here tends to struggle with a large number of parameters.
@@
 Vaccination dramatically affects the transmission dynamics by introducing some additional immunity into the population.
-Mechanistic models require explicit representation of this immunity\todo{cite something about vaccination}. (Birrell et al, 2023)
+Mechanistic models require explicit representation of this immunity\todo{cite something about vaccination, including new RTM paper}.
 
@@
 
+Using a mechanistic model provides additional understanding, compared to the approach in \cref{E-backcalc}.
+For example, decomposing the change in transmission into changing contacts, $\matr{\kappa}(t)$, changing susceptibility $\vec{s}(t)$, and other unmeasured factors, $\beta(t)$.
+By understanding which of these components is changing, policies can be better targeted.
+If it is shown that $\beta(t)$ decreases, this implies that for a given contact rate in the population, there is less transmission.
+Such decreases might be beneficial as it allows society to operate in a more normal manner without increasing transmission.
+Breaking the trade-off between transmission and activity is desirable because it reduces the disease burden without burdesome restrictions on individuals day-to-day lives.
+
 \section{Conclusion}

### 2023-12-20 transmission.tex
continue restructure

@@
 \chapter{Estimating transmission} \label{transmission}
+
+In this chapter, I use the estimate of the duration of PCR positivity from \cref{E-imperf-test} to estimate the transmission of SARS-CoV-2 in England.
+This estimation is based on the CIS data (see \cref{transmission:sec:data} for further details).
+Estimating transmission is the ultimate aim of this thesis.
+
+I take two approaches.
+The theoretical advantages of each approach were discussed in \cref{E-inc-prev:sec:infection-process}.
+First, in \cref{backcalc}, I use a backcalculation approach, based on \cref{E-inc-prev:eq:EPt}.
+Then, in \cref{SEIR}, I use a mechanistic compartmental model, extending the model introduced in \cref{E-SEIR:sec:mechanistic-models}.
+Finally, in \cref{transmission:sec:comparison}, I compare the results of the two approaches.
+
+\section{Data} \label{transmission:sec:data}
+
+Throughout this chapter, the CIS data will be stratified into the regions of England and the following age groups.
+\begin{enumerate}
+    \item $[0, 11)$: children of pre-school and primary school age.
+    \item $[11, 16)$: children of compulsory secondary-school age.
+    \item $[16, 25)$: adolescents and young adults who may remain in formal education.
+    \item $[25, 50)$: younger working age adults.
+    \item $[50, 70)$: older working age adults.
+    \item $[70, \infty)$: elderly.
+\end{enumerate}
+These groupings align closely with those previously used for analysis of the CIS data~\autocites[e.g.:][]{houseInferring}{walkerTracking}.
+\todo{Contrast with age groupings used in previous chapter}
 
@@
 The advantage of fitting a multilevel regression model on prevalence, rather than incidence, is that prevalence can naturally be expressed as a generalised linear model (GLM).
-A natural formulation for either incidence proportion or prevalence in a stratum $i$ is $\logit(\pi_i) = g(t)$, where $\pi_i$ is the incidence proportion or prevalence in stratum $i$ and $g$ is a function of time, penalised towards being linear.
+A natural formulation for either incidence proportion or prevalence in a stratum $i$ is $\logit(\pi_i) = g_i(t)$, where $\pi_i$ is the incidence proportion or prevalence in stratum $i$ and $g$ is a function of time, penalised towards being linear.
 This is a GLM with a logit link function.
@@
 First, it ensures that the incidence proportion or prevalence is between 0 and 1.
-Second, at low values of incidence or prevalence, linear growth is approximately exponential which is the natural assumption for an epidemic and has been shown to improve estimates of prevalence\todo{cite REACT paper on this, and check the exact meaning of "improve"}.
+Second, at low values of incidence or prevalence, linear growth is approximately exponential which is what mechanistic models predict, and has been shown to improve estimates of prevalence\todo{cite REACT paper on this, and check the exact meaning of "improve"}.
 However, if the incidence proportion is expressed as a GLM in this way, the prevalence is not a GLM because a non-linear transformation is required to obtain prevalence from the logit of incidence proportion (the relationship on the natural scale is linear, as derived in \cref{E-inc-prev:sec:prevalence-process}).
@@
 This step assumes that the estimates for each region can be considered independent, which is reasonable because the regions are large with many tests conducted within each.
-
-I used age groupings:\todo{insert age groupings}.
-These align with previous work on the CIS\todo{cite CIS reports}.
-They are based on school ages (primary, compulsory secondary, and post-compulsory secondary) for children to capture that school opening and closing may have a large effect on transmission.
 
@@
 
-\section{Mechanistic modelling}
+\section{Mechanistic modelling} \label{SEIR}
 
-{\color{red} After introducing the observational model in generic terms above (section 8.2) I would see a reorganisation of the section 8.3 as follows:   
-\begin{itemize}
-    \item 8.3 (currently  titled Implementation) CIS-based transmission model (see comments on lines 484-486) 
-\item 8.3.1 WAIFW matrix
-\item 8.3.2 Full model (perhaps another title..model structure?) 
-\item 8.3.3 Observational model (currently section 8.2) include here the "Solving the system of ODEs" (currently in section 8.3.4)
-\item 8.3.4 Priors 
-\item 8.3.5 Inference 
-\item 8.4 Simulation study
-\end{itemize}
-}
+As discussed in \cref{E-inc-prev:sec:infection-process}, there are various advantages to using a mechanistic model to estimate transmission.
+This section uses a mechanistic model (introduced in \cref{E-SEIR:sec:mechanistic-models}) to estimate the transmission of SARS-CoV-2 in England.
 
-\todo[inline]{Redo this intro, with some high-level comments taken from the full model section (eg following Birrell et al.)}
+The transmission part of the model follows \textcite{birrellRealtime}.
+A novel approach (\cref{SEIR:sec:observation}) links the transmission to the data.
+In the mechanistic modelling literature, this part of the model is referred to as the observation model, however, it includes both the prevalence process and the observation process (as defined in \cref{E-inc-prev}).
 
@@
 
-A parsimonious model is needed to parameterise the WAIFW matrix (see \cref{SEIR:sec:structured-populations}).
-I use the following age-based strata, which align closely with the age groupings previously used for analysis of the CIS data~\autocites[e.g.:][]{houseInferring}{walkerTracking}.
-\todo{Contrast with age groupings used in previous chapter}
-\begin{enumerate}
-    \item $[0, 11)$: children of pre-school and primary school age.
-    \item $[11, 16)$: children of compulsory secondary-school age.
-    \item $[16, 25)$: adolescents and young adults who may remain in formal education.
-    \item $[25, 50)$: younger working age adults.
-    \item $[50, 70)$: older working age adults.
-    \item $[70, \infty)$: elderly.
-\end{enumerate}
-
+A parsimonious model is needed to parameterise the WAIFW matrix (see \cref{E-SEIR:sec:structured-populations}).
 I follow the WAIFW matrix parameterisation of \textcite{birrellRealtime}, with the decomposition $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'} N_{a'}$.
@@
 Here incidence is calculated from the ODE solution and then related to the data as explained in \cref{E-inc-prev}.
-However, this approach is computationally expensive because of the need to calculate the convolution in \cref{E-inc-prev:eq:EPt }.
+However, this approach is computationally expensive because of the need to calculate the convolution in \cref{E-inc-prev:eq:EPt}.
 
@@
 These compartments do not have a biological interpretation, unlike the other compartments in the system, but are purely a mathematical tool so that the time spent in these compartments matches the duration distribution.
-This is the same idea behind having multiple exposed or latent compartments (introduced in \cref{SEIR:sec:non-exponential}).
+This is the same idea behind having multiple exposed or latent compartments (introduced in \cref{E-SEIR:sec:non-exponential}).
 
@@
 
-The full model combines the elements discussed in this chapter up until now.
-The transmission model follows \textcite{birrellRealtime}.
-It is a SEIR model with two latent and two infectious states.
-The population is stratified by age, using a time-varying matrix $\matr{\beta}(t)$ to describe the transmission between strata and how these change over time.
-
+The remaining model structure follow \textcite{birrellRealtime}.
+The transmission model is a SEIR model with two latent and two infectious states.
 To form whole of England estimates, each region is modelled independently with the results aggregated, as in \cref{E-backcalc}.
-
-The observation model is novel, relating the latent transmission described by the model to the CIS data.
-It uses the phase-type model to represent the duration distribution of PCR positivity explained in \cref{SEIR:sec:observation}.
 
@@
 
-\subsubsection{Solving the systems of ODEs} \label{SEIR:sec:solving-ODEs}
 I solve the system of ODEs using Euler's method, discretised at timesteps of $\delta t = 1/2$ days.
@@
 
-\subsubsection{Inference} \label{SEIR:sec:MCMC}
+\subsubsection{Inference} \label{SEIR:sec:inference}
 
@@
 In addition, I use a parsimonious parameterisation of the initial conditions using the initial proportion of the population that are infectious and the initial proportion that are susceptible only.
-Both of these rely on an assumption that the epidemic is in a steady-state exponential growth phase at time 0 (see \cref{SEIR:sec:SIR}).
+Both of these rely on an assumption that the epidemic is in a steady-state exponential growth phase at time 0 (see \cref{E-SEIR:sec:SIR}).
 
@@
 
-Furthermore, $\R$ is a function of $\matr{\beta}$, and hence $\beta_1$ (as explained in \cref{SEIR:sec:structured-populations}).
+Furthermore, $\R$ is a function of $\matr{\beta}$, and hence $\beta_1$ (as explained in \cref{E-SEIR:sec:structured-populations}).
 Combining these expressions, the model can be parameterised by using $\psi$ in place of $\beta_1$.
@@
 
-I ran a simulation study to check that the inference procedure (\cref{SEIR:sec:inference-implementation}) was working correctly and that the model was identifiable.
+I ran a simulation study to check that the inference procedure (\cref{SEIR:sec:inference}) was working correctly and that the model was identifiable.
 I ran 100 simulations, with parameters drawn from the priors (see \cref{SEIR:table:priors}) except that the prior on $\psi$ was $\LN(0.048, 0.0035)$.
@@
 
-All fixed values, including $\matr{\kappa}$, used the previously described in this section.
+All other aspects of the simulation model, including $\matr{\kappa}$, are as previously described in this section.
 This means that the simulations created realistic incidence and prevalence curves (see \cref{SEIR:fig:sim-data}).
@@
 
+\todo[inline]{What happened to the following paragraph? It doesn't make sense currently!}
 In this setup, the parameters $i^+$, $q_c$, and $\psi$ are fully recovered; the other parameters are only partially so: their posterior has moved away from the prior but is still regularised by it (see \cref{SEIR:fig:true-vs-posterior}).
@@
 
-\section{Comparison of results}
+\section{Comparison of results} \label{transmission:sec:comparison}
 

### 2023-12-21 transmission.tex
minor edits

@@
 These groupings align closely with those previously used for analysis of the CIS data~\autocites[e.g.:][]{houseInferring}{walkerTracking}.
-\todo{Contrast with age groupings used in previous chapter}
 
@@
 
+This section uses the same notation as \cref{E-SEIR:sec:mechanistic-models}, following the infectious disease modelling literature rather than the statistical conventions used elsewhere in this thesis.
+
 \subsection{Methods}
@@
 However, these are not captured by the contact survey.
-To allow for these changes and the uncertainty produced in formulating $\matr{\kaapa}(t)$, I follow \textcite{birrellRealtime} in placing a random walk prior on $\log\beta(t)$  on the weekly timescale:
+To allow for these changes and the uncertainty produced in formulating $\matr{\kappa}(t)$, I follow \textcite{birrellRealtime} in placing a random walk prior on $\log\beta(t)$  on the weekly timescale:
 \begin{align}
@@
 The majority of the confidence intervals contain the nominal probability (see \cref{SEIR:fig:sim-coverage}).
-There is a tendency for the coverage to be slightly lower than the nominal probability, suggesting that the credible intervals are slightly too narrow; this is a common problem when using random-walk MCMC.
+There is a tendency for the coverage to be slightly lower than the nominal probability, suggesting that the credible intervals are slightly too narrow.
+It is well-known that random walk MCMC, as used here, can poorly explore the tails of the posterior in finite time~\autocite[e.g.][]{turitsynIrreversible}.
 \todo{cite issues with random-walk MCMC exploring space}
@@
 This suggests that these parameters are not practically identifiable in this setting.
-The posteriors align very closely with the priors (not shown) which reinforces this finding.
+The posteriors align very closely with the priors (not shown\todo{consider showing prior/posterior plots}) which reinforces this finding.
 The lack of identifiability for $\vec{\pi}$ aligns with previous work showing this parameter is structurally unidentifiable, \ie not identifiable even in the limit of infinite data~\autocite{dankwaStructural}.
 \begin{figure}
-    \includegraphics{SEIR/sim/true_vs_posterior}
+    \makebox[\textwidth][c]{\includegraphics{SEIR/sim/true_vs_posterior}}
     \caption[True vs posterior parameter values]{%

### 2023-12-21 introduction.tex
Add section on sode

@@
 In this chapter, I introduce mechanistic models for estimating the transmission of SARS-CoV-2, an alternative to the statistical approach in \cref{E-backcalc}.
-Mechanistic models have a structure which make assumptions on the biological and epidemiological mechanisms behind the infection process (see \cref{E-in-prev:sec:infection-process}), and hence the disease dynamics~\autocite{lesslerMechanistic}.
+Mechanistic models have a structure which make assumptions on the biological and epidemiological mechanisms behind the infection process (see \cref{E-inc-prev:sec:infection-process}), and hence the disease dynamics~\autocite{lesslerMechanistic}.
 Individuals are giving the model's state and parameters a biological and/or epidemiological interpretation.
@@
 
+The code for the analyses in this thesis is available online.
+An index to the various repositories is available at\todo{insert GitHub link}.
+Almost all code, excluding dependencies, was written from scratch for these analyses.
+The exception is the Markov chain Monte Carlo sampler used in \cref{E-SEIR}, which was created by Sanmitra Ghosh for the article \textcite{ghoshApproximate}.
+
+The ATACCC data used is found in the same location.
+CIS data can only be accessed by accredited researchers within the Office for National Statistic's Secure Research Service\todo{insert details of applying}.
+Where possible, aggregated, non-disclosive versions of the dataset are available alongside the code.
+
 \subsection{Epidemiological contributions}

### 2023-12-21 incidence-prevalence.tex
Fix file names, build, refs
continue restructure
Make into a coherent chapter

@@
 I then move on to explaining the three processes that are involved: the \emph{infection process}, the \emph{prevalence process} and the \emph{observation process}.
+
 The infection process (\cref{inc-prev:sec:infection-process}) models how infections arise.
 The concepts here are well-established within infectious disease epidemiology (see\todo{ref intro chapter}).
+
 The prevalence process (\cref{inc-prev:sec:prevalence-process}) describes the relationship between incidence and prevalence.
@@
 In particular, recovering from infection needs to be considering, which was not the case for HIV/AIDS.
+
 The observation process (\cref{inc-prev:sec:observation-process}) describes the relationship between prevalence and the observed data, ultimately allowing inference of incidence based on this data.
@@
 Therefore, the model is specific to this context, although much would be applicable to other, similarly designed prevalence surveys.
+
 Finally, \cref{inc-prev:sec:MCMC} will introduce Markov chain Monte Carlo (MCMC), an inference technique I will make extensive use of in this thesis.
+
 \Cref{inc-prev:sec:conclusion} concludes the chapter.
@@
 
-Time will be considered discrete in this chapter.
+Time will be considered discrete in this chapter, except in \cref{SEIR:sec:mechanistic-models}.
 I use days, however, the results generalise in the obvious way to other time units.
@@
 \todo[inline]{Add some refs about duration and survival}
-Duration, or time-to-event, distributions can be expressed in several equivalent ways.
+\emph{Duration}, or \emph{time-to-event}, distributions can be expressed in several equivalent ways.
 The duration distribution I will be concerned with is the distribution of the number of days for which an individual is prevalent.
@@
 The duration is a positive integer, \ie $D_i \in \{1, 2, \dots, \dmax \}$, where $\dmax$ is the maximum duration.
-The first way of expressing the distribution of $D_i$ is as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_work out where to present full model equations and then edit/rearrange this section to reflect this{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
+The first way of expressing the distribution of $D_i$ is as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
 The \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$.
@@
 Similarly, $\lambda(d) \in [0,1]$, but can otherwise vary arbitrarily (\ie there is no monotonic or sum constraint).
+The lack of monotonicity or sum constraint on the hazard makes it an attractive parameterisation for inference.
 
@@
 In the context of a disease, the point process can be discretised into the incidence on day $t$ as $Z_t = K(t) - K(t-1)$, where $t$ is an integer.
+I will explain two approaches to modelling the infection process.
 
-A counting process has \emph{independent increments} if the number of events in disjoint intervals are independent conditional on the intensity~\autocite[244]{yanDistribution}.
-That is, for any $t_1 < t_2 < t_3 < t_4$, $K(t_2) - K(t_1)$ and $K(t_4) - K(t_3)$ are independent.
+First, in \cref{inc-prev:sec:poisson-process}, as a \emph{Poisson process}.
+A Poisson process is a widely-used type of counting process both within and outside of epidemiology.
+It makes no direct assumptions on the underlying biological or epidemiological process, and is therefore a flexible model.
 
+Afterwards, in \cref{SEIR}, I introduce \emph{mechanistic models}.
+Mechanistic models explicitly describe the population and transmission between individuals.
+They vary greatly in their details and realism~\autocite{murilloMultiscale}.
 \todo[inline]{Make sure this section is coherent: a bunch of stuff from the SEIR chapter has been copied and pasted in}
 
-\subsection{Poisson processes}
+\subsection{Poisson processes} \label{inc-prev:sec:poisson-process}
 
-A commonly-used point process for infectious disease epidemiology is the time-inhomogenous Poisson process~\autocites[e.g., in the context of HIV:][]{brookmeyerMethod}{paganoHIV}{rosenbergBackcalculation}{brookmeyerBackcalculation}.
-A Poisson process is a point process with independent increments, and the property that $K(s) - K(t) \dist \Poi\left(\int^s_t \nu(r) dr \right)$ where $\nu(r) \geq 0$ is the \emph{intensity} of the infection process at time $r$~\autocites[244]{yanDistribution}.
-A time-homogenous Poisson process is the special case where $\nu(r)$ is constant.
-The independent increment property says that, conditional on $\nu(r)$, the number of infections in disjoint intervals are independent.
-This implies $Z_t$ and $Z_{t'}$ are independent for $t \neq t'$.
-Extensions to this model can incorporate covariates, such as geography or age~\autocite[e.g.][]{diggleModeling}.
+A commonly-used point process for infectious disease epidemiology is the time-inhomogenous Poisson process~\autocites[e.g., in the context of HIV,][]{brookmeyerMethod}{paganoHIV}{rosenbergBackcalculation}{brookmeyerBackcalculation}.
+% A time-homogenous Poisson process is the special case where $\nu(r)$ is constant.
+Extensions to this model, which I do not consider here, have incorporated covariates, such as geography or age~\autocite[e.g.][]{diggleModeling}.
+
+The first defining property of a Poisson process is that the counts are Poisson distributed.
+Formally, $K(s) - K(t) \dist \Poi\left(\int^s_t \nu(r) dr \right)$ where $\nu(r) \geq 0$ is the \emph{intensity} of the infection process at time $r$~\autocites[244]{yanDistribution}.
+
+The other defining property of a Poisson process is that it has \emph{independent increments}.
+A counting process has independent increments if the number of events in disjoint intervals are independent conditional on the intensity~\autocite[244]{yanDistribution}.
+That is, for any $t_1 < t_2 < t_3 < t_4$, $K(t_2) - K(t_1)$ and $K(t_4) - K(t_3)$ are independent.
+\todo{check if this should be $<$ or $\leq$}
+In the context of a Poisson process, the independent increment property implies that, conditional on $\nu(r)$, the number of infections in disjoint intervals are independent.
+Hence, $Z_t$ and $Z_{t'}$ are independent for $t \neq t'$.
 
@@
 
-A common cause of overdispersion is super-spreading events~\autocite{lloyd-smithSuperspreading}.
-A super-spreading event is an event where many more people than average are infected.
-That is, $Z_t >> \E(Z_t)$ for some $t$ during which a super-spreading event occurs.
-Coronaviruses, such as SARS-CoV-2, have particularly high overdispersion~\autocites{endoEstimating}{adamClustering}{mccloskeySARS}.
-An infamous example of a SARS-CoV-2 super-spreading event was the Skagit County choir practice in March 2020~\autocite{hamnerHigh}.
-One infected individual is thought to have infected at least 32 further individuals during a single choir practice.
-For comparison, the average individual infected around three others in the early stages of the pandemic~\autocite{pellisChallenges}. 
-Overdispersion can be incorporated by replacing the assumption that $K(s) - K(t) \dist \Poi(\int_{t}^{s} \nu(r) dr)$ with $K(s) - K(t) \dist \NBc(\int_{t}^{s} \nu(r) dr, k)$ where $k$ is the overdispersion parameter, normally assumed to be constant absent interventions.
-Note that the expected number of events has not changed, but the variance has increased (see \cref{E-distributions} for details of parameterisations and moments).
-%The negative binomial tends to Poisson as $k\to\infty$.
-Generally, $k<1$ is considered as significant levels of overdispersion.
-Estimates for SARS-CoV-2 indicate $k \approx 0.1$~\autocite{endoEstimating}, and may be decreased by some interventions including lockdowns~\autocites{quiltyReconstructing}{quiltyUnderstanding}.
-If $K(s) - K(t)$ has a negative binomial distribution, then its standard deviation scales on the order of the mean.
-If it is Poisson distributed, then the standard deviation is on the order of the square root of the mean.
-A particular implication of this is that the coefficient of variation (the standard deviation divided by the mean) tends to zero for a Poisson process, but tends to a constant for a negative binomial process.
-The disadvantage of the negative binomial distribution is that it loses much of the mathematical convenience of a Poisson distribution.
+% A common cause of overdispersion is super-spreading events~\autocite{lloyd-smithSuperspreading}.
+% A super-spreading event is an event where many more people than average are infected.
+% That is, $Z_t >> \E(Z_t)$ for some $t$ during which a super-spreading event occurs.
+% Coronaviruses, such as SARS-CoV-2, have particularly high overdispersion~\autocites{endoEstimating}{adamClustering}{mccloskeySARS}.
+% An infamous example of a SARS-CoV-2 super-spreading event was the Skagit County choir practice in March 2020~\autocite{hamnerHigh}.
+% One infected individual is thought to have infected at least 32 further individuals during a single choir practice.
+% For comparison, the average individual infected around three others in the early stages of the pandemic~\autocite{pellisChallenges}. 
+% Overdispersion can be incorporated by replacing the assumption that $K(s) - K(t) \dist \Poi(\int_{t}^{s} \nu(r) dr)$ with $K(s) - K(t) \dist \NBc(\int_{t}^{s} \nu(r) dr, k)$ where $k$ is the overdispersion parameter, normally assumed to be constant absent interventions.
+% Note that the expected number of events has not changed, but the variance has increased (see \cref{E-distributions} for details of parameterisations and moments).
+% %The negative binomial tends to Poisson as $k\to\infty$.
+% Generally, $k<1$ is considered as significant levels of overdispersion.
+% Estimates for SARS-CoV-2 indicate $k \approx 0.1$~\autocite{endoEstimating}, and may be decreased by some interventions including lockdowns~\autocites{quiltyReconstructing}{quiltyUnderstanding}.
+% If $K(s) - K(t)$ has a negative binomial distribution, then its standard deviation scales on the order of the mean.
+% If it is Poisson distributed, then the standard deviation is on the order of the square root of the mean.
+% A particular implication of this is that the coefficient of variation (the standard deviation divided by the mean) tends to zero for a Poisson process, but tends to a constant for a negative binomial process.
+% The disadvantage of the negative binomial distribution is that it loses much of the mathematical convenience of a Poisson distribution.
 
 Feedback loops mean that stochastically low or high numbers of infections affect the future number of infections.
-That is, $\nu(t+\Delta t)$ is dependent on $K(t)$, even after conditioning on $\nu(t)$, for $\Delta t > 0$.
+That is, $\nu(t+\Delta t)$, for $\Delta t > 0$, is dependent on $K(t)$, even after conditioning on $\nu(t)$.
 This violates the assumption of independent increments.
-The reasons infectious diseases generate feedback loops is because stochastically high or low incidence feeds back to the process intensity at future times.
-This can be seen most obviously in the event of a super-spreading event.
-If a super-spreading event occurs, then many more individuals than suggested by the process intensity will be infected.
-These additional infections will propagate the epidemic further, increasing the intensity at future times.
+Infectious diseases generate feedback loops because stochastically high or low incidence feeds back to the process intensity at future times.
+% This can be seen most obviously in the event of a super-spreading event, when many .
+% If a super-spreading event occurs, then many more individuals than suggested by the process intensity will be infected.
+% These additional infections will propagate the epidemic further, increasing the intensity at future times.
 Point processes can be generalised to \emph{branching processes} to account for this feedback~\autocite[246]{yanDistribution}, although these are beyond the scope of this thesis.
@@
 Despite these issues, consistent inference on the incidence can still be produced using the Poisson process model, although the uncertainty in the estimates may be understated~\autocite{beckerDependent}.
-
-An alternative to explicit use of counting processes is using \emph{mechanistic models}.
-Mechanistic models explicitly describe the population and transmission between individuals.
-They vary greatly in their details and realism~\autocite{murilloMultiscale}.
-Mechanistic models are discussed further in \cref{E-SEIR}.
 
@@
 
-\subsection{Compartmental models} \label{SEIR:sec:transmission-generic}
+\subsection{Mechanistic models} \label{SEIR:sec:mechanistic-models}
 
-This section introduces the class of mechanistic transmission models known as \emph{compartmental} models.
+Now I introduce mechanistic models, in particular the class of mechanistic transmission models known as \emph{compartmental} models.
 Compartmental models are the most widely-used class of mechanistic infectious disease model.
 A compartmental model is named because they place every member of the population under study into a compartment; each compartment representing a disease state (\eg susceptible).
+I will describe these models using the notation of the field, which differs in conventions from the rest of this thesis.
+In particular, upper case or lower case will no longer be used to distinguish between random variables and their realisations, which instead is clear by context.
 
@@
 The parameters $\vec{\theta}$ determine, along with the model structure, how the state vector changes over time.
-Inference about $\vec{x}(t)$ and $\vec{\theta}$ is requires linking to data, $\vec{y}$, via an observation model, $p(\vec{y}(t) \mid \vec{x}, \vec{\theta})$ where $\vec{x}$ denotes conditioning on $\vec{x}(t)$ for all $t$~\autocite{birrellEvidence} (see \cref{SEIR:sec:observation}).
-That is, the observation model specifies the likelihood of the data given the model state.
+% Inference about $\vec{x}(t)$ and $\vec{\theta}$ is requires linking to data, $\vec{y}$, via an observation model, $p(\vec{y}(t) \mid \vec{x}, \vec{\theta})$ where $\vec{x}$ denotes conditioning on $\vec{x}(t)$ for all $t$~\autocite{birrellEvidence} (see \cref{SEIR:sec:observation}).
+% That is, the observation model specifies the likelihood of the data given the model state.
 
@@
 
-The remainder of this chapter uses only deterministic models.
-This choice is appropriate for the the context this thesis considers because there were many individuals infectious with SARS-CoV-2 in this time period.
-There are a wealth of introductory texts on stochastic models, I recommend \textcite[chapter 6]{keelingModeling} or \textcite{birrellEvidence}.
+The remainder of this section uses only deterministic models.
+This choice is appropriate for the context of this thesis because there were many individuals infectious with SARS-CoV-2 in this time period.
+There are a wealth of introductory texts on stochastic models, I recommend \textcite[chapter 6]{keelingModeling}.
 
-This chapter discusses deterministic models whose evolution is specified by a system of ODEs, $d\vec{x}/dt$.
-In an ODE model, time is necessarily continuous.%, although discretised when solving (see \cref{SEIR:sec:inference-implementation}).
-For concreteness, I use units of days throughout, although the model can use any unit of time.
-The number or proportion of individuals in each compartment is also continuous; clearly, this is not true in reality but is a good approximation when the population is large.
+% This section discusses deterministic models whose evolution is specified by a system of ODEs, $d\vec{x}/dt$.
+% In an ODE model, time is necessarily continuous.%, although discretised when solving (see \cref{SEIR:sec:inference-implementation}).
+% For concreteness, I use units of days throughout, although the model can use any unit of time.
+% The number or proportion of individuals in each compartment is also continuous; clearly, this is not true in reality but is a good approximation when the population is large.
 
-In this section, I introduce the \emph{susceptible-infectious-recovered} (SIR) model and the extensions to it I require to appropriately model SARS-CoV-2.
+% In this section, I introduce the \emph{susceptible-infectious-recovered} (SIR) model and the extensions to it I require to appropriately model SARS-CoV-2.
 
@@
     For longer time periods, waning immunity can be incorporated by allowing individuals to move from the recovered compartment back to the susceptible compartment~\autocite[40]{keelingModeling}.
+    \item The incidence proportion on day $t$ can be defined as $Z_t / \Npop = s(t) - s(t+1)$.
+However, this will not in general give an integer value to $Z_t$.
+In practice, this is not an issue since a continuous approximation for the incidence will suffice.
 \end{itemize}
@@
 \begin{figure}
-  \todo[inline]{From Dani: increase the size, particularly the title. Add (A) BBC to left (B) POLYMOD to right and refer to these in the  caption. Why "all"?}
-  \centering \includegraphics[width=\textwidth]{SEIR/contact_matrices}
+  \makebox[\textwidth][c]{\includegraphics[width=0.9\paperwidth]{inc-prev/contact_matrices}}
   \caption[Age-based contact matrices]{Number of contacts reported between age groups in two contact surveys: (A) number of contacts reported in the POLYMOD survey~\autocite{mossongSocial}; (B) number of contacts reported in the BBC Pandemic survey~\autocite{klepacContacts}, which only collected data on those aged over 12.
@@
 Parallel diagonals off the main diagonal show mixing between parents and their children.
-Figure adapted from \textcite{klepacContacts}, licensed under CC BY-NC-ND 4.0.}
+Figure adapted from \textcite{klepacContacts}, licensed under \href{https://creativecommons.org/licenses/by-nc-nd/4.0/}{CC BY-NC-ND 4.0.}}
   \label{SEIR:fig:age-contacts}

### 2023-12-21 ATACCC.tex
Fix references

@@
 
-Several features of the \textcite{hakkiOnset} model can be motivated by the desire to apply Stan, a modern and convenient MCMC implementation of the NUTS algorithm (see \cref{E-backcalc:sec:MCMC}).
+Several features of the \textcite{hakkiOnset} model can be motivated by the desire to apply Stan, a modern and convenient MCMC implementation of the NUTS algorithm (see \cref{E-inc-prev:sec:MCMC}).
 The first of these is using a continuous viral load trajectory rather than the more common piecewise linear model.
@@
 Here, $D_i$ is drawn conditional on the population-level parameters $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$.
-Equivalent forms of $F_D$, such as the survival and hazard functions, can be calculated using the relations in \cref{E-backcalc:sec:definitions}.
+Equivalent forms of $F_D$, such as the survival and hazard functions, can be calculated using the relations in \cref{E-inc-prev:sec:definitions}.
 I found 100,000 draws produced stable estimates of the tail of $F_D$.
 
-\section{Results} \label{ATACCC:sec:results-discussion}
+\section{Results} \label{ATACCC:sec:results}
 
@@
 The $F_D$ estimates for the two groups are in \cref{ATACCC:fig:duration}.
-In \cref{ATACCC:sec:results-discussion}, the two models gave very similar results.
+In \cref{ATACCC:sec:results}, the two models gave very similar results.
 Therefore, I show results only from the modified model (which had better convergence).

### 2023-12-21 biology-data.tex
Fix references

@@
 An intermittent negative is when an individual tests negative but tested positive previously and subsequently.
-Intermittent negatives are stripped out when creating the dataset used for all duration analyses in this chapter (as described in \cref{E-episode-def}), but demonstrate that false negatives do occur.
+Intermittent negatives are stripped out when creating the dataset used for all duration analyses in this chapter, but demonstrate that false negatives do occur.
 
-Whether a future positive is part of the same episode of a reinfection is not always trivial; I rely on a process developed previously by Sarah Walker, see \cref{E-episode-def}.
+% Whether a future positive is part of the same episode of a reinfection is not always trivial; I rely on a process developed previously by Sarah Walker, see \cref{E-episode-def}.
 However, up until the emergence of the Omicron variant, reinfections are rare, especially in a short time frame, and hence the tricky cases are rare until this time (late 2021).

### 2023-12-21 cis-perfect-testing.tex
Fix references

@@
 Inference was implemented in Stan via RStan~2.21.8~\autocite{rstan2-21-8} using default settings.
-Convergence was assessed using Rhat and ESS as described in \cref{E-ATACCC:sec:convergence}\todo{add ref to place explaining diagnostics}, and all runs checked for divergent transitions.
+Convergence was assessed using Rhat and ESS as described in \cref{E-inc-prev:sec:MCMC}, and all runs checked for divergent transitions.
 
@@
 I have assumed constant and independent incidence (a priori).
-Prevalence in the CIS data was approximately constant over the period of interest (see \cref{E-intro:fig:prevalence}).
+Prevalence in the CIS data was approximately constant over the period of interest (see\todo{ref prev constant over period}).
 A sensitivity analysis, where an epidemic in exponential growth or decline was simulated (not shown), showed minimal impact on survival estimates.

### 2023-12-21 incidence-prevalence.tex
Fix references

@@
 The ideal proposal distribution is the posterior of interest, however, this is impossible because that is what MCMC is trying to estimate~\autocite[296]{gelmanBDA}.
-One alternative is \emph{adaptive} algorithms which tune the proposal distribution based on previous draws in an attempt to approximate the posterior; I use such a scheme in \cref{E-SEIR:sec:MCMC}.
+One alternative is \emph{adaptive} algorithms which tune the proposal distribution based on previous draws in an attempt to approximate the posterior; I use such a scheme in \cref{E-SEIR:sec:inference}.
 

### 2023-12-21 introduction.tex
Fix references

@@
 
-I start this chapter by introducing a generic framework for mechanistic modelling of infectious diseases (\cref{E-SEIR:sec:transmission-generic}).
-% This framework is then applied to SARS-CoV-2 in \cref{E-SEIR:sec:transmission-application}.
-It describes the \emph{transmission model}, corresponding to the infection process.
-The transmission model needs to be linked to observations; in this chapter, that will be the CIS prevalence data.
-The \emph{observation model}, introduced in \cref{E-SEIR:sec:observation}, is this link; it incorporates both the prevalence process and the observation process.
-These two components form the mathematical model, which I then fit to data, requiring specification of an inference process, parameterisation, and priors (\cref{E-SEIR:sec:inference-implementation}).
-This final formulation uses the transmission model from \textcite{birrellRealtime}, but a novel observation model.
-I then use a simulation study to check this recovers the parameters successfully (\cref{E-SEIR:sec:sim-study}).
-Having shown that the model can recover the parameters, I then apply it to the CIS data (\cref{E-SEIR:sec:application}).
-Finally, I discuss the results, limitations, and extensions of this work (\cref{E-SEIR:sec:discussion}).
-
-Notation in this chapter differs from the rest of this thesis.
-I adopt conventions and notation familiar to the infectious disease modelling field, although the field's notation is not consistent between sources.
-In particular, upper or lower case no longer signifies whether a variable is random or not, but this is clear from the context.
+% I start this chapter by introducing a generic framework for mechanistic modelling of infectious diseases (\cref{E-SEIR:sec:transmission-generic}).
+% % This framework is then applied to SARS-CoV-2 in \cref{E-SEIR:sec:transmission-application}.
+% It describes the \emph{transmission model}, corresponding to the infection process.
+% The transmission model needs to be linked to observations; in this chapter, that will be the CIS prevalence data.
+% The \emph{observation model}, introduced in \cref{E-SEIR:sec:observation}, is this link; it incorporates both the prevalence process and the observation process.
+% These two components form the mathematical model, which I then fit to data, requiring specification of an inference process, parameterisation, and priors (\cref{E-SEIR:sec:inference-implementation}).
+% This final formulation uses the transmission model from \textcite{birrellRealtime}, but a novel observation model.
+% I then use a simulation study to check this recovers the parameters successfully (\cref{E-SEIR:sec:sim-study}).
+% Having shown that the model can recover the parameters, I then apply it to the CIS data (\cref{E-SEIR:sec:application}).
+% Finally, I discuss the results, limitations, and extensions of this work (\cref{E-SEIR:sec:discussion}).
 

### 2023-12-21 ATACCC.tex
easy edits

@@
 The aim of this chapter is to estimate the duration of PCR positivity using the viral load data from the ATACCC study (described in\todo{ref section describing ATACCC}).
-I model a biomarker, the log viral load, over the course of an infection.\todo{make sure log viral load and Ct values are introduced in chapter 2}
+I model a biomarker, the log viral load, over the course of an infection.
 From this model, I derive the distribution of the duration of PCR positivity.
@@
 The observation model includes the possibility of false positive and negatives, and the limit of detection of the PCR test.
-\todo{Make sure limit of detection is introduced previously when I first talk about PCR}
 % The limit of detection means that viral loads below a certain threshold are not detected by the PCR test.
@@
 A positive PCR test has a Ct value.
-Ct values are linearly related to viral load (see\todo{make sure relationship between Ct values and viral load is introduced previously}).
+Ct values are linearly related to viral load (see \cref{E-biology-data:sec:PCR})
 A negative PCR test signals either the viral load is too low to be detected, \ie below the limit of detection, or the test is a false negative.

### 2023-12-21 cis-imperfect-testing.tex
easy edits

@@
 However, this would require a model for time-varying test sensitivity to be incorporated into the model.
-Furthermore, it is unclear that such a model would be identifiable because it would not be clear whether to attribute negative results to false negatives or to recoveries\todo{cite something on this}.
+Furthermore, it is unclear that such a model would be identifiable because it would not be clear whether to attribute negative results to false negatives or to recoveries\todo{cite identifiability issues with varying sensitivity}.
 

### 2023-12-21 incidence-prevalence.tex
easy edits

@@
 They vary greatly in their details and realism~\autocite{murilloMultiscale}.
-\todo[inline]{Make sure this section is coherent: a bunch of stuff from the SEIR chapter has been copied and pasted in}
 
@@
 That is, for any $t_1 < t_2 < t_3 < t_4$, $K(t_2) - K(t_1)$ and $K(t_4) - K(t_3)$ are independent.
-\todo{check if this should be $<$ or $\leq$}
 In the context of a Poisson process, the independent increment property implies that, conditional on $\nu(r)$, the number of infections in disjoint intervals are independent.
@@
 
-Stochastic models specify a probability distribution for the current state $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta})$ where $\set{H}_t = \{ \vec{x}(t') \ssep t' < t \}$ is the epidemic's history and $\vec{\theta}$ are model parameters\todo{REF stochastic models}.
+Stochastic models specify a probability distribution for the current state $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta})$ where $\set{H}_t = \{ \vec{x}(t') \ssep t' < t \}$ is the epidemic's history and $\vec{\theta}$ are model parameters\todo{Cite stochastic models}.
 Stochastic models are most commonly specified in discrete time, using a Markov assumption.
@@
 
-The simplest compartmental model for a disease which confers immunity is the (SIR) model\todo{ref for SIR being simplest} (depicted in \cref{SEIR:fig:SIR}).
+The simplest compartmental model for a disease which confers immunity is the (SIR) model\todo{cite SIR being simplest} (depicted in \cref{SEIR:fig:SIR}).
 The name SIR derives from the structure of the model.
@@
 
-The SIR model assumes that individuals are immediately infectious upon infection.
-This is unrealistic for most diseases, as there is a period between an individual being infected and them being infectious, known as the \emph{latent period}.
-\todo{I will probably end up defining the latent period in the intro, so should just ref that}
+The SIR model assumes that individuals are immediately infectious upon infection, neglecting the latent period (see \cref{E-biology-data:sec:natural-history}).
 Ignoring the latent period will again overestimate the reproduction number~\autocite{wearingAppropriate}.
-
-The SIR model can be extended to include a latent period by adding a latent compartment~\autocite[41]{keelingModeling}, also known as an \emph{exposed} compartment.
+The SIR model can be extended by adding a latent compartment~\autocite[41]{keelingModeling}, also known as an \emph{exposed} compartment.
 This forms the susceptible-exposed-infectious-recovered (SEIR) model, shown in \cref{SEIR:fig:SEIR}.
@@
 MCMC using a Normal proposal distribution can be inefficient.
-In particular, it often fails to propose moves into the tails of the posterior distribution\todo{any suggested refs for this (Paul?) - we've discussed this lots but I've never read it anywhere}.
+In particular, it often poorly explores the tails of the posterior in finite time~\autocite[e.g.][]{turitsynIrreversible}.
 \emph{Hamiltonian Monte Carlo} (HMC)~\autocites{nealMCMC}{nealImproved}{duaneHybrid} improves the proposals by using information from the gradient of the posterior density (taken with respect to the parameters).
@@
 I will make use of the implementation in the package \emph{Sampling through adaptive neighborhoods} (Stan)~\autocite{Stan-2-32-2}.
-Stan implements \emph{automatic differentiation}\todo{ref automatic diff} to calculate the gradient of the posterior density without any user input.
+Stan implements \emph{automatic differentiation}\todo{cite automatic diff} to calculate the gradient of the posterior density without any user input.
 Automatic differentiation uses the chain rule to calculate the gradient of a function from its code by breaking it down into a series of elementary operations, for which the gradient is known.

### 2023-12-21 transmission.tex
Fix references
easy edits

@@
 There is a tendency for the coverage to be slightly lower than the nominal probability, suggesting that the credible intervals are slightly too narrow.
-It is well-known that random walk MCMC, as used here, can poorly explore the tails of the posterior in finite time~\autocite[e.g.][]{turitsynIrreversible}.
-\todo{cite issues with random-walk MCMC exploring space}
+This is a common issue with random walk MCMC (see \cref{E-inc-prev:sec:MCMC}).
 This is similar when considering coverage of the derived quantities incidence and prevalence (see \cref{SEIR:fig:sim-inc-prev}).
@@
 \end{figure}
-
-% \begin{figure}
-%     \includegraphics{SEIR/CIS/p_peak}
-%     \caption[Posterior estimates of peak incidence timing]{%
-%         Posterior probability of the peak incidence for each region and age combination occurring on each day.
-%     }
-%     \label{SEIR:fig:peak-incidence}
-% \end{figure}
+\begin{figure}
+    \includegraphics{SEIR/CIS/p_peak}
+    \caption[Posterior estimates of peak incidence timing]{%
+        Posterior probability of the peak incidence for each region and age combination occurring on each day.
+    }
+    \label{SEIR:fig:peak-incidence}
+\end{figure}
 

### 2023-12-21 incidence-prevalence.tex
easy edits

@@
 I will make use of the implementation in the package \emph{Sampling through adaptive neighborhoods} (Stan)~\autocite{Stan-2-32-2}.
-Stan implements \emph{automatic differentiation}\todo{cite automatic diff} to calculate the gradient of the posterior density without any user input.
-Automatic differentiation uses the chain rule to calculate the gradient of a function from its code by breaking it down into a series of elementary operations, for which the gradient is known.
+Stan implements \emph{automatic differentiation} to calculate the gradient of the posterior density without any user input.
+Automatic differentiation uses the chain rule to calculate the gradient of a function from its code by breaking it down into a series of elementary operations, for which the gradient is known~\autocite{autodiff}.
 For a more detailed introduction to HMC, NUTS, and Stan see \textcite[chapter 12]{gelmanBDA}.

### 2023-12-21 introduction.tex
easy edits

@@
 The ATACCC data used is found in the same location.
-CIS data can only be accessed by accredited researchers within the Office for National Statistic's Secure Research Service\todo{insert details of applying}.
+CIS data can only be accessed by accredited researchers within the Office for National Statistic's Secure Research Service\todo{insert details of applying to SRS}.
 Where possible, aggregated, non-disclosive versions of the dataset are available alongside the code.

### 2024-01-03 ATACCC.tex
Show multiple parameter values in fig

@@
 $\tau_{\max,i}$ is inferred independently for each individual because the time from first test to peak is not a biological concept shared between individuals.
-\begin{figure}
+\begin{figure}[tbhp]
   \centering \includegraphics{ATACCC/typical_trajectory}
-  \caption[Example viral load model trajectory.]{Example viral load trajectory produced by \textcite{hakkiOnset}'s viral load model, using realistic parameter values.
-The increase in log viral load is approximately linear with a gradient of $a_i$, here 5.
-The decrease is also linear, with gradient $-b_i$, here -1.4.
-The peak is at time $\tau_{\max,i}$, here 0, and log viral load $\log v_{\max,i}$, here = 15.
-The smoothing (lack of linearity) near the peak is visible.
-\label{ATACCC:fig:viral-load-model}}
-  \todo[inline]{Add a few lines to example plot showing the effect of different parameters, move parameters to key}
+  \caption[Example viral load model trajectory.]{%
+    Example viral load trajectory produced by \textcite{hakkiOnset}'s viral load model, using realistic parameter values.
+    The increase and decrease in log viral load are approximately linear with a gradient of $a_i$ and $b_i$ respectively.
+    The peak is at time $\tau_{\max,i} = 0$ and log viral load $\log v_{\max,i}$.
+    The subscripts are omitted in the key for clarity.
+    The smoothing (lack of linearity) near the peak is visible. \label{ATACCC:fig:viral-load-model}
+  }
 \end{figure}

### 2024-01-03 ATACCC/trajectories.R

@@ -2,54 +2,44 @@ library(ggplot2)
 source(here::here("ATACCC/trajectory_functions.R"))
 source(here::here("utils.R"))
 
-# min_x = -5
-# max_x = 15
-# x = seq(min_x, max_x, length.out = 1000)
+min_x = -5
+max_x = 15
+x = seq(min_x, max_x, length.out = 1000)
 
-# param_vals = tibble::tribble(
-#   ~tau_max, ~a_i, ~b_i,
-#   exp(15), 5, 1.4,
-#   exp(12.5), 7, 2,
-# )
+param_vals = tibble::tribble(
+  ~tau_max, ~a_i, ~b_i,
+  exp(15), 5, 1.4,
+  exp(12.5), 3, 2,
+)
 
-# trajectories = purrr::pmap_dfr(
-#   param_vals,
-#   function(tau_max, a_i, b_i) {
-#     label_text = bquote(tau[max,i] == .(tau_max) * "," ~ a[i] == .(a_i) * "," ~ b[i] == .(b_i))
-#     tibble::tibble(
-#       # label = parse(
-#       #   text = paste0("tau[max,i] = ", tau_max, "; a[i] = ", a_i, "; b[i] = ", b_i)
-#       # ),
-#       label = as.character(label_text),
-#       x = x,
-#       y = logVL(x, c(tau_max, a_i, b_i)),
-#     )
-#   }
-# )
+trajectories = purrr::pmap_dfr(
+  param_vals,
+  function(tau_max, a_i, b_i) {
+    label_text = bquote(tau[max,i] == .(tau_max) * "," ~ a[i] == .(a_i) * "," ~ b[i] == .(b_i))
+    tibble::tibble(
+      label = paste0("log v = ", log(tau_max), "; a = ", a_i, "; b = ", b_i),
+      #  parse(
+      #   text = paste0("tau[max,i] = ", tau_max, "; a[i] = ", a_i, "; b[i] = ", b_i)
+      # ),
+      # label = as.character(label_text),
+      x = x,
+      y = logVL(x, c(tau_max, a_i, b_i)),
+    )
+  }
+)
 
-# ggplot(trajectories, aes(x, y, colour = label)) +
-#   geom_line() +
-#   coord_cartesian(ylim = c(0, NA)) +
-#   labs(
-#     x = "Time from peak viral load",
-#     y = "Log viral load"
-#   ) +
-#   theme_minimal() +  # Replace with your actual theming function
-#   scale_colour_manual(
[…]

### 2024-01-03 incidence-prevalence.tex
Note about possibility of merging sections

@@
 \subsubsection{The SEIR model} \label{SEIR:sec:SEIR}
+
+\todo[inline]{Maybe merge SEIR intro into the general intro, not introducing SIR first}
 
@@
 I will make use of the implementation in the package \emph{Sampling through adaptive neighborhoods} (Stan)~\autocite{Stan-2-32-2}.
-Stan implements \emph{automatic differentiation} to calculate the gradient of the posterior density without any user input.
+Stan implements \emph{automatic differentiation}\todo{cite automatic diff} to calculate the gradient of the posterior density without any user input.
 Automatic differentiation uses the chain rule to calculate the gradient of a function from its code by breaking it down into a series of elementary operations, for which the gradient is known~\autocite{autodiff}.

### 2024-01-04 cis-imperfect-testing.tex
Figures with imperfect testing results

@@
 \begin{figure}
-  \centering %\includegraphic{}
-  \todo[inline]{Figure showing estimates without false negatives}
-  \caption[Estimating survival using CIS data assuming perfect testing]{Estimating the survival function using CIS data and the model of \cref{E-perf-test} gives implausibly short estimates.
+  \centering \includegraphics{cis-imperfect-testing/CIS_perfect}
+  \caption[Estimating survival using CIS data assuming perfect testing]{Estimates of the survival function using CIS data alongside the ATACCC-based estimates of \cref{E-perf-test}.
+CIS-based is implausibly short over the first 14 days.
 \label{imperf-test:fig:problem-cis-estimates}}
@@
 \begin{figure}
-  \todo[inline]{Insert fig showing final estimate and comparison to ATACCC}
+  \centering \includegraphics{cis-imperfect-testing/CIS_final}
   \caption{Duration estimates using CIS and ATACCC data}
@@
 \begin{figure}
-  \todo[inline]{
-    Insert figs showing, for different assumption:
-    proportion of individuals still positive at day 30 or 50 (tail),
-    overall survival curve, and
-    median survival time (in comparison to ATACCC).
+  \vspace{-2.5cm}
+  \makebox[\textwidth][c]{\includegraphics{cis-imperfect-testing/CIS_vary}}
+  \caption{%
+    Sensitivity of CIS estimates to: (A-C) the value of $r$ in the prior for $\ntot$ and (D-F) $\psens$.
+    A and D: median survival time, in comparison to the ATACCC-based estimate of \cref{E-ATACCC}.
+    B and E: survival at day 50, $S_\theta(50)$.
+    C and F: full survival curves out to day 100.
   }
-  \caption{Sensitivity of CIS estimates to assumptions}
   \label{imperf-test:fig:cis-sensitivity}

### 2024-01-04 cis-imperfect-testing/CIS_survival.R

@@ -6,13 +6,36 @@ library(tidybayes)
 library(tidyr)
 source(here::here("utils.R"))
 
-posterior_draws = readRDS(here::here("cis-imperfect-testing/STATS17701/draws.rds"))
+#############################################################################################
+### LOAD THE ESTIMATES TO PLOT
+#############################################################################################
 
-posterior_draws |>
-    filter(sensitivity == 1)
+posterior_draws = readRDS(here::here("cis-imperfect-testing/STATS18744/draws.rds"))
+posterior_draws = bind_rows(
+    posterior_draws,
+    readRDS(here::here("cis-imperfect-testing/STATS17701/draws.rds")) |>
+        filter(!name %in% posterior_draws$name)
+)
+median_survival = posterior_draws |>
+    summarise(
+        median = min(time[S <= 0.5]),
+        .by = c(
+            .draw,
+            survival_prior,
+            sensitivity,
+            missed_model,
+            r
+        )
+    )
 
+ataccc_duration = readRDS(here::here("ATACCC-distributions/posterior_samples.rds")) |>
+    group_by(time) |>
+    median_qi()
+ataccc_medians = readRDS(here::here("ATACCC-distributions/posterior_samples.rds"))  |>
+    group_by(.draw) |>
+    summarise(median = min(time[S <= 0.5]))
 
-posterior_draws |>
+check_counts = posterior_draws |>
     filter(time == 1, .draw == 1, missed_model == "total") |>
     count(
         survival_prior,
@@ -20,5 +43,147 @@ posterior_draws |>
         missed_model,
         r
     )
-    filter(survival_prior == "Informative") |>
+stopifnot(check_counts$n == 1)
+rm(check_counts)
 
+#############################################################################################
+### FIGURE ASSUMING NO FALSE NEGATIVES
+#############################################################################################
+
+p_perfect = posterior_draws |>
+    filter(sensitivity == 1, survival_prior == "Informative", r == 22047) |>
+    ggplot() +
+    stat_lineribbon(
+        aes(time, S, fill = "CIS-based", colour = "CIS-based"),
+        alpha = 0.5,
+        .width = 0.95
+    ) +
+    geom_lineribbon(
+        aes(time, S, ymin = S.lower, ymax = S.upper, fill = "ATACCC-based", colour = "ATACCC-based"),
[…]

### 2024-01-04 cis-imperfect-testing.tex
Figure with ntot posterior sensitivity

@@
 A very weak, almost uninformative, prior on this quantity causes the posterior estimate to be much higher than the estimate from \textcite{birrellRTM2}.
-When increasing the prior's strength, the posterior estimate moves towards the prior smoothly, as expected (see\todo{figure showing $r$ vs $\ntot$}).
+When increasing the prior's strength, the posterior estimate moves towards the prior smoothly, as expected (see \cref{imperf-test:fig:ntot}).
 Taking a prior from \textcite{birrellRTM2} means that the posterior survival is close to the \cref{E-ATACCC} estimate in the region where that estimate is reliable.
 This justifies the use of this prior.
+\begin{figure}
+  \centering \includegraphics{cis-imperfect-testing/CIS_ntot}
+  \caption{How the posterior estimate of $\ntot$ changes with the value of $r$ in the prior on $\ntot$.}
+  \label{imperf-test:fig:ntot}
+\end{figure}
 

### 2024-01-04 cis-imperfect-testing/CIS_ntot.R

@@ -1,0 +1,34 @@
+suppressMessages(library(dplyr))
+library(ggplot2)
+library(patchwork)
+library(purrr)
+library(tidybayes)
+library(tidyr)
+source(here::here("utils.R"))
+
+#############################################################################################
+### LOAD THE ESTIMATES TO PLOT
+#############################################################################################
+posterior = readRDS(here::here("cis-imperfect-testing/STATS18744/means.rds"))
+
+#############################################################################################
+### CREATE FIGURE
+#############################################################################################
+fig = posterior |>
+    filter(sensitivity == 0.8, survival_prior == "Informative", missed_model == "total") |>
+    ggplot() +
+    stat_pointinterval(
+        aes(r, n_tot),
+        .width = 0.95
+    ) +
+    standard_plot_theming() +
+    scale_x_log10() +
+    labs(y = expression(n[tot]))
+ggsave(
+    filename = here::here("cis-imperfect-testing/CIS_ntot.pdf"),
+    plot = fig,
+    width = 11,
+    height = 7,
+    units = "cm",
+    dpi = 300
+)

### 2024-01-04 ATACCC.tex
explain calculation of $F_D$ ATACCC posterior

@@
 When $v_\text{lod} = 0$, as it does here, this gives the following expressions:
-$$
-\begin{aligned}
+\begin{align}
 \tau_{0,i}
@@
 &= \frac{1}{a_i b_i} \left( (a_i + b_i) (\log v_{\max,i} + \log(a_i + b_i)) - a_i \log a_i - b_i \log b_i \right).
-\end{aligned}
-$$
-Note that $d_i$ does not depend on $\tau_{\max,i}$, only $\phi_i$.
+\label{ATACCC:eq:di}
+\end{align}
 
-\todo[inline]{insert a better explanation of $F_D$'s posterior}
-For later use, a discretised version is preferred (as described in \cref{E-inc-prev}).
-To calculate the discrete version of this distribution, $F_D$, a Monte Carlo integration is used with the following relation.
+From \cref{ATACCC:eq:di}, we can see that $d_i$ depends only on $i$'s random effects vector $\vec{\phi}_i$.
+Therefore, the distribution of $d_i$ across the population is determined by the $\vec{phi}_i$'s mean, $\vec{\mu}_\phi$, and covariance matrix, $\matr{\Sigma}_\phi$.
+Hence, the uncertainty in the posterior distribution of the discretised duration cumulative distribution function, $F_D(d) = \prob(D_i \leq d + 0.5)$ originate from the uncertainty in $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$.
+
+For a known $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$, $F_D$ is equal to the following expression.
 \begin{align}
@@
   &= \E_{D_i}(\I(D_i \leq d + 0.5))
+  \label{ATACCC:eq:FDintegral}
 \end{align}
 where $\I$ is the indicator function.
-Here, $D_i$ is drawn conditional on the population-level parameters $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$.
+The integral in \cref{ATACCC:eq:FDintegral} can be approximated using Monte Carlo integration (see \cref{ATACCC:alg:MonteCarlo}).
 Equivalent forms of $F_D$, such as the survival and hazard functions, can be calculated using the relations in \cref{E-inc-prev:sec:definitions}.
-I found 100,000 draws produced stable estimates of the tail of $F_D$.
+
+\begin{algorithm}
+\For{$i = 1, \dots, M$}{
+  \For{$j = 1, \dots, M$}{
+    Sample an individual-level random effects vector $\vec{\phi}_{i,j} \dist \LN(\vec{\mu}_{\phi,g}, \matr{\Sigma}_\phi)$ \;
+    Calculate $d_{i,j}$ using \cref{ATACCC:eq:di} \;
+  }
+  \For{$d = 1, \dots, \dmax$}{
+    Set $F_D^{(i)}(d) = \frac{1}{M} \sum_{j=1}^M \I(d_{i,j} \leq d + 0.5)$ \;
+  }
+}
+  \caption{%
+  Algorithm to calculate the Monte Carlo integral in \cref{ATACCC:eq:FDintegral}, producing the posterior distribution of $F_D(d)$ for group $g$.
+  The posterior distribution is represented as a sample $F_D^{(1)}, \dots, F_D^{(N)}$.
+  $\vec{\mu}_\phi^{(i)}$ and $\matr{\Sigma}_\phi^{(i)}$ denote the $i$th posterior sample of $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$ respectively, taken from the MCMC output.
+  $M$ is the number of individuals to draw, I found $M = 100,000$ produced stable estimates of the tail of $F_D(d)$ if $\dmax = 45$.
+ }
+ \label{ATACCC:alg:MonteCarlo}
+\end{algorithm}
 

### 2024-01-05 incidence-prevalence.tex
talk about distributions of time infectious

@@
 
-The basic SIR model implies that the distribution of the time spent in the infectious compartment is exponential~\autocite[96]{keelingModeling}.
-\todo{Expand on what ``exponential" means here, possibly as a mean-field approximation. See papers by Lloyd.}
+The basic SIR model is the deterministic approximation of a stochastic model where the duration of time in the infectious compartment is exponentially distributed~\autocite[96]{keelingModeling}.
 This is an unrealistic assumption for most diseases and would lead to underestimating of the reproduction number~\autocites{lloydRealistic}{wearingAppropriate}.
-A standard extension to the SIR model is to allow the infectious period to follow a gamma distribution~\autocite[94]{keelingModeling}.
+A standard extension to the SIR model is to instead assume a gamma distribution~\autocite[94]{keelingModeling}{andersonSpread}.
 This is done by adding multiple infectious compartments, shown in \cref{SEIR:fig:SIR-gamma}.
@@
 where $n$ is the number of infectious compartments, and redefining $\lambda(t) = \beta \sum_{j=1}^n i_j(t)$.
+The additional compartments added do not, in general, correspond to any biological feature or have a biological interpretation; they are purely a mathematical device to modify the distribution of time between becoming and ceasing to be infectious~\autocite{lloydRealistic}.
 In this model the infectious period is distributed $\GamDist(n, n\gamma)$, a realistic model for a variety of diseases~\autocite{wearingAppropriate}.

### 2024-01-05 transmission.tex
Clarify the procedure and justification for inference check

@@
 To check the performance of the inference procedure, I calculated the coverage of the 50\%, 75\%, and 95\% central credible intervals for each parameter and simulation.
-I define coverage as the proportion of simulations where the true value of the parameter in that simulation was within the credible interval.
-This procedure is a simplified version of simulation-based calibration~\autocites{cookValidation}{taltsValidating}.
-The coverage for each parameter should equal the nominal probability (\eg the coverage should be 50\% for the 50\% credible intervals).
+Define coverage as the proportion of simulations where the true value of the parameter in that simulation was within the credible interval.
+As the parameters are drawn from the prior, if the inference is performed correctly then the coverage for each parameter should equal the nominal probability (\eg the coverage should be 50\% for the 50\% credible intervals)~\autocite{cookValidation}.
+This check is related to, although simpler than, simulation-based calibration~\autocite{taltsValidating}.
 I constructed a confidence interval for the coverage using the Pearson-Klopper method implemented in the binom package~\autocite{binom1-1}.

### 2024-01-05 transmission.tex
Improve explanation of backcalc model

@@
 
+I use the period from the expansion of the CIS, starting 31st August 2020, until there was significant vaccination rollout\todo{insert date}.
+
 \section{Backcalculation} \label{backcalc}
 
-Now that I have estimated the duration of PCR positivity, I can estimate the incidence of SARS-CoV-2.
-In this chapter, I will apply the theory introduced in \cref{E-inc-prev} to the CIS data.
-I use the period from the expansion of the CIS, starting 31st August 2020, until there was significant vaccination rollout\todo{insert date}.
-First, I explain the method I will apply (\cref{backcalc:sec:methods}) and then present the results (\cref{backcalc:sec:results}).
-I discuss the results in \cref{backcalc:sec:discussion} and conclude in \cref{backcalc:sec:conclustion}.
+In this section, I will apply the theory introduced in \cref{E-inc-prev} to estimate incidence.
+% First, I explain the method I will apply (\cref{backcalc:sec:methods}) and then present the results (\cref{backcalc:sec:results}).
+% I discuss the results in \cref{backcalc:sec:discussion} and conclude in \cref{backcalc:sec:conclustion}.
 
@@
 However, a deterministic relation between true prevalence and incidence is a good approximation (see \cref{E-inc-prev:sec:prevalence-process}).
-I extend the methodology previously used to estimate prevalence from the CIS data, corrected for non-response bias, to estimate incidence\todo{cite Koen's stuff}.
+
+Therefore, I extend the methodology previously used to estimate prevalence from the CIS data, corrected for non-response bias, to estimate incidence\todo{cite Koen's stuff}.
 The previous method was multilevel regression and poststratification (MRP)\todo{cite MRP}.
-I fit a similar model to estimate prevalence, then transform the posterior distribution of prevalence in each stratum to a posterior distribution of incidence.
-Finally, I poststratify the posterior distribution of incidence to the CIS target population and subpopulations of interest.
+Here, a Bayesian multilevel logistic regression model is first fit to predict PCR positivity from demographic covariates.
+Then, the posterior distribution is used to predict the prevalence in each stratum.
+Finally, the prevalence is poststratified to the CIS target population and subpopulations of interest.
+I follow this same process, except I transform the posterior distribution of prevalence in each stratum to a posterior distribution of incidence before poststratifying.
+% I fit a modification model to estimate prevalence, then transform the posterior distribution of prevalence in each stratum to a posterior distribution of incidence.
+% Finally, I poststratify the posterior distribution of incidence to the CIS target population and subpopulations of interest.
 
-The advantage of fitting a multilevel regression model on prevalence, rather than incidence, is that prevalence can naturally be expressed as a generalised linear model (GLM).
-A natural formulation for either incidence proportion or prevalence in a stratum $i$ is $\logit(\pi_i) = g_i(t)$, where $\pi_i$ is the incidence proportion or prevalence in stratum $i$ and $g$ is a function of time, penalised towards being linear.
-This is a GLM with a logit link function.
+The advantage of fitting a multilevel regression model on prevalence, rather than incidence, is that prevalence with appropriate prior (or penalisation) can be expressed as a generalised linear model (GLM).
+A simple formulation\todo{I don't like the wording "simple formulation", maybe "default"?} for either incidence proportion or prevalence in a stratum $i$ is $\logit(\pi_i) = g(i, t)$, where $\pi_i$ is the incidence proportion or prevalence in stratum $i$ and $g$ is a function of the stratum and time, penalised towards being linear in time.
+If $g$ is chosen appropriately, this is a GLM with a logit link function.
+
 The logit transformation serves two purposes.
 First, it ensures that the incidence proportion or prevalence is between 0 and 1.
-Second, at low values of incidence or prevalence, linear growth is approximately exponential which is what mechanistic models predict, and has been shown to improve estimates of prevalence\todo{cite REACT paper on this, and check the exact meaning of "improve"}.
-However, if the incidence proportion is expressed as a GLM in this way, the prevalence is not a GLM because a non-linear transformation is required to obtain prevalence from the logit of incidence proportion (the relationship on the natural scale is linear, as derived in \cref{E-inc-prev:sec:prevalence-process}).
+Second, at low values of incidence or prevalence, linear growth is approximately exponential~\autocite[see discussion in][and references therein]{ealesAppropriately}.
+Approximately exponential growth is what mechanistic models predict, and penalising towards exponential growth has previously been shown to improve estimates of prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
+
+However, if the incidence proportion is expressed as a GLM in this way, the prevalence is not a GLM.
+The relationship between incidence and prevalence is linear, as derived in \cref{E-inc-prev:sec:prevalence-process}.
+Therefore, the relationship between the logit of incidence proportion and prevalence is non-linear.
+
 Using a GLM is computationally useful due to the large number of pre-existing packages available for computing the posterior distribution, or its approximation, efficiently.
-Previously, when fitting to prevalence over long periods and including all appropriate covariates to correct for non-response bias, the model was too computationally expensive to fit using MCMC; instead, an implementation of the model using the Integrated Nested Laplace Approximation (INLA) algorithm was used~\citePersonalComms{Koen Pouwels}.
+Previously, when fitting to prevalence over long periods and including all appropriate covariates to correct for non-response bias, the model was too computationally expensive to fit using MCMC~\citePersonalComms{Koen Pouwels}.
+Instead, an implementation of the model using the Integrated Nested Laplace Approximation (INLA) algorithm was used.
 
 I modified a pre-existing version of a MRP model for CIS~\autocite{pouwelsMRPvaccination}, as recommended by that model's author for this situation~\citePersonalComms{Koen Pouwels}.
-The model included fixed effects for age group\todo{check the model I used}.
-The likelihood is a beta-binomial distribution, a generalisation of the binomial distribution allowing for overdispersion (see \cref{E-distributions}).
-The overdispersion allows for household clustering\todo{ref beta-binomial as a model for clustering} and any other overdispersion in the data\todo{ref that epi data is often overdispersed}. 
+\todo[inline]{Describe the model precisely, see eg RTM paper 2}
+% The model included fixed effects for age group\todo{check the model I used}.
+% The likelihood is a beta-binomial distribution, a generalisation of the binomial distribution allowing for overdispersion (see \cref{E-distributions}).
+% The overdispersion allows for household clustering\todo{ref beta-binomial as a model for clustering} and any other overdispersion in the data\todo{ref that epi data is often overdispersed}. 
 Finally, the model was fit independently for each region, implicitly adding interactions between region and all other variables.
@@
 For each stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
-Considering the relationship between incidence and prevalence to be deterministic, as \cref{E-inc-prev:sec:prevalence-process} justifies, equation implies that, for any stratum $\vec{P} = \matr{S} \vec{Z}$ where $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum, $\matr{S}$ is a matrix as defined below, and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
-The matrix $\matr{S}$ needs to be defined based on the survival function to preserve the relationship between incidence and prevalence.
-It is lower-triangular of the following form:
+I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:prevalence-process} justifies.
+Therefore, \cref{E-inc-prev:eq:EPt} implies that $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix of the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
+The matrix $\matr{S}$ preserves the relationship between incidence and prevalence.
+This gives a lower-triangular of the following form:
 \begin{align}

### 2024-01-16 transmission.tex
Updates from Overleaf

--- 2023-12-21 transmission.tex
@@
 However, a deterministic relation between true prevalence and incidence is a good approximation (see \cref{E-inc-prev:sec:prevalence-process}).
-I extend the methodology previously used to estimate prevalence from the CIS data, corrected for non-response bias, to estimate incidence\todo{cite Koen's stuff}.
+I extend the methodology used to estimate prevalence from the CIS data, corrected for non-response bias, to estimate incidence\todo{cite Koen's stuff}.
 The previous method was multilevel regression and poststratification (MRP)\todo{cite MRP}.
 I fit a similar model to estimate prevalence, then transform the posterior distribution of prevalence in each stratum to a posterior distribution of incidence.
-Finally, I poststratify the posterior distribution of incidence to the CIS target population and subpopulations of interest.
+Finally, I poststratify the posterior distributions of incidence to the CIS target population and subpopulations of interest.
 
@@
 Considering the relationship between incidence and prevalence to be deterministic, as \cref{E-inc-prev:sec:prevalence-process} justifies, equation implies that, for any stratum $\vec{P} = \matr{S} \vec{Z}$ where $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum, $\matr{S}$ is a matrix as defined below, and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
-The matrix $\matr{S}$ needs to be defined based on the survival function to preserve the relationship between incidence and prevalence.
+The matrix $\matr{S}$ is defined from the survival function to preserve the relationship between incidence and prevalence.
 It is lower-triangular of the following form:
@@
 The transmission part of the model follows \textcite{birrellRealtime}.
-A novel approach (\cref{SEIR:sec:observation}) links the transmission to the data.
+A novel approach (\cref{SEIR:sec:observation}) links the transmission model to the data.
 In the mechanistic modelling literature, this part of the model is referred to as the observation model, however, it includes both the prevalence process and the observation process (as defined in \cref{E-inc-prev}).
@@
 \subsection{Methods}
-
+\todo[inline]{refer to where quantities are defined}
 I use the same CIS data as in \cref{E-backcalc}, again considering each region independently.
@@
 
-Previous studies for pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults~\autocites{chenRole}{vinerTransmission}.
+Studies for pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults~\autocites{chenRole}{vinerTransmission}.
 Therefore, I set $q^\text{inf}_{a'} = 1$ for all $a'$, $q^\text{susc}_{a} = q_c$ for the youngest stratum, and $q^\text{susc}_{a} = 1$ for the remaining strata.
@@
 
-The transmission model represents the infection process (introduced in \cref{E-inc-prev:sec:infection-process}.
+The transmission model represents the infection process (introduced in \cref{E-inc-prev:sec:infection-process}).
 A variety of different approaches can be taken to link a transmission model to observations.
@@
 This means that there is a wide variety of models used.
-Therefore, this section  explains the model I propose for fitting to the CIS prevalence data.
+This section  explains the model I propose for fitting to the CIS prevalence data.
 
@@
 $p_0$ increases at the same rate as individuals flow into $e_1$.
-These individuals then transition into a set of compartment with their transition rates chosen to match the duration of positivity estimates in \cref{E-imperf-test}, $p_1, \dots, p_l$.
+These individuals then transition into a set of compartment with their transition rates, $p_1, \dots, p_l$, chosen to match the duration of positivity estimates in \cref{E-imperf-test}.
 These compartments do not have a biological interpretation, unlike the other compartments in the system, but are purely a mathematical tool so that the time spent in these compartments matches the duration distribution.
@@
 
-The modelled proportion of individuals that are PCR positive in the can then be linked to the data using a Beta-Binomial likelihood (see\todo{ref section on clustering} for the issues with using a binomial likelihood).
+The modelled proportion of individuals that are PCR positive in the are linked to the data using a Beta-Binomial likelihood (see\todo{ref section on clustering} for the issues with using a binomial likelihood).
 If on each day $t$, each strata $i$ is observed to have $y_{it}$ positives out of $n_{it}$ tests, then the likelihood is:
@@
 The parameters governing the duration of positivity ($\kappa$s) were explained previously (see \cref{SEIR:table:ec-params}).
-The initial conditions and transition rate parameters are unindentifiable in a SEIR model~\autocite{dankwaStructural}, therefore I used assumed values or strongly informative priors on them.
+The initial conditions and transition rate parameters are unindentifiable in a SEIR model~\autocite{dankwaStructural}, therefore strongly informative priors are used for them.
 $\sigma_\epsilon$ and $\rho$ are given priors that penalise model complexity.

### 2024-01-16 cis-imperfect-testing.tex

@@
 
-Comparing the simulated and CIS data suggests that this is due to the high frequency of single positive episodes in the CIS data.
-A single positive episode is an episode containing exactly one positive test; intuitively, these cause short estimates because there is no lower bound on the length of time the episode lasts.
+Comparing the simulated data in the previous chapter and CIS data suggests that this is due to the high frequency of single positive episodes in the CIS data.
+A single positive episode is an episode containing exactly one positive test; intuitively, these cause short estimates because there is no non-zero lower bound on the length of time the episode lasts.
 Furthermore, short episodes are very likely to be missed.
 Therefore, there is little or no ability to upper bound the number of very short episodes.
-I found that simulations building in even a few excess single positive episodes led to problematic estimation within the CIS setting (not shown).
+I found that simulations that included even a few excess single positive episodes led to problematic estimation within the CIS setting (not shown).
 
 \Textcite{shenNonparametrica}, extending \textcite{panNote}, studied the situation when only the terminating event is interval censored and the initiating event has simple left truncation.
-This is related, although simpler, setting than the CIS; the CIS has double interval censoring and a more complex pattern of missed episodes (see \cref{E-perf-test:sec:problem}).
-\Textcite{shenNonparametrica} showed that, in their situation, the maximum likelihood estimator is inconsistent and can lead to a severely biased underestimation of the survival function.
+This is a related, although simpler, setting than the CIS.
+The CIS has double interval censoring and a more complex pattern of missed episodes (see \cref{E-perf-test:sec:problem}).
+\Textcite{shenNonparametrica} showed that, in their situation, the maximum likelihood estimator is inconsistent if there is at least one single positive episode (check the precise statement) and can lead to a severely biased underestimation of the survival function.
 While the result cannot simply be extended to the more complex study design of the CIS, it reinforces the intuition that single positive episodes are likely to be problematic.
@@
 % Discussion and interrogation of the data led me to
-I hypothesised that the reason for the unexpectedly high number of single positive episodes (compared to simulation) was the presence of false negatives results.
+I hypothesised that the reason for the unexpectedly high number of single positive episodes (compared to simulation) was the presence of false negative results.
 There are two strands of evidence supporting this hypothesis.
@@
 Second, adjusting the simulation to include false negatives can reproduce both the CIS data more faithfully and exhibits the same issues when estimating the duration distribution (see \cref{imperf-test:fig:sim-single-pos}).
+Simulation methodology is explained in \cref{E-perf-test:sec:simulation-study}, with false negatives included as described in \cref{imperf-test:sec:simulate}.
 \begin{figure}
@@
     The CIS data is shown as a vertical line.
-    Simulation methodology is explained in \cref{E-perf-test:sec:simulation-study}, with false negatives included as described in \cref{imperf-test:sec:simulate}.
   }
@@
 
-\section{Modelling false negatives} \label{imperf-test:sec:simulate}
+\section{Generative model for false negatives} \label{imperf-test:sec:simulate}
 
+In this section, I introduce a probabilistic model for the generation of false negatives.
 The simplest model for false negatives is a constant probability of returning a negative test even when detectable.
@@
 $\psens = 1$ means that there are no false negatives, the situation previously considered.
-False negatives are formalised by defining the outcome of a test as a random binary variable such that:
-\begin{equation}
-  \prob(Y_i(t) = 1) \begin{cases}
-      \psens &b_i \leq t \leq e_i \\
-      0 &\text{otherwise}
-  \end{cases} 
-\end{equation}
-where $Y_i(t)$ is only defined for $t \in \sched_i$.
-However, an implausibly low value for $\psens$ is required to reproduce the rate of single positive episodes seen in the CIS.
+\todo{check if detectable and test sensitivity defined somewhere, and in a consistent way}
+In \cref{E-perf-test}, the outcome of a test was deterministic, positive if the individual being tested had an ongoing infection episode.
+Now, the outcome of such a test result is a Bernoulli random variable with a probability $\psens$ of producing a positive.
+If $\psens = 1$, this recovers the previous model without false negatives.
+As previously, tests in individuals without ongoing infection episodes are always negative.
+
+I modified the simulation study from \cref{E-perf-test:sec:simulation-study} to include the possibility of false negatives.
+More single positive episodes are seen at lower values of $\psens$ (see \cref{imperf-test:fig:sim-single-pos}).
+A value between 60\% and 80\% for $\psens$ reproduces the number of single positive episodes observed within the CIS data.
+In \cref{E-ATACCC}, I estimated that the rate of false negatives is\todo{add what the ATACCC false negative rate is}.
+The daily testing in the ATACCC study, used for the \cref{E-ATACCC} estimates, should provide a reliable estimate of the false negative rate during the period for which the individuals are followed-up\todo{rephrase this sentence, currently unclear}.
+The requirement for a much lower value of $\psens$ suggests this model of false negatives is inadequate.
 
@@
 False negatives are more likely to occur when the viral load of an individual is low, because there is less virus in their body to be sampled.
-In \cref{E-ATACCC}, this mechanism is incorporated by the left-censored observations.
-This suggests that a model with a declining test sensitivity as a function of time since infection might be more suitable.
+In \cref{E-ATACCC}, this mechanism is incorporated by treating negative test results as possible left-censored observations.
+This observation suggests that a model with a declining test sensitivity as a function of time since infection might be more suitable.
 
@@
 Then I calculate the test sensitivity as a function of days since the first positive in an episode, bounding this by assuming either all or none of those in group 3 are false negatives.
-% The test sensitivity is the true positives / (true positives + false negatives).
+The test sensitivity is the true positives / (true positives + false negatives).
 % In \cref{imperf-test:fig:bounding-cis-sensitivity} we consider the values produced by assuming the negative following the last positive in an episode could be either a true or false negative as a function of time since the individual was detected (\ie: the first positive test in the episode).
 This generates broad bounds, although suggest a declining test sensitivity over the course of an episode.
+Small numbers of tests mean that the series is noisy near the start and the end, such as.
 \begin{figure}
@@
   \caption[Bounding test sensitivity using CIS data]{
-    Bounding the test sensitivity using CIS data as a function of time since the infection was detected (see main text for explanation of how the bounds are formed).
-    Small numbers of tests mean that the series is noisy near the start and the end.
+    Bounding the test sensitivity using CIS data as a function of time since the infection was detected.
   }
@@
 where $t$ is the time in days since infection.
-Simulated data using this test sensitivity is much more similar to the data (see \cref{imperf-test:fig:sim-single-pos}), while being more plausible than very low probabilities of false negatives.
+Adapting the simulation study to incorporate this test sensitivity is much more similar to the data (see purple histogram in \cref{imperf-test:fig:sim-single-pos}), while being more plausible than very low probabilities of false negatives.
 
-\section{Modelling} \label{imperf-test:sec:modelling}
+\section{Likelihood modification for false negatives} \label{imperf-test:sec:modelling}
 
-In this section, I introduce a simple model of false negatives into the model of \cref{E-perf-test:sec:model}.
-A simple model means that the likelihood remains tractable.
-I modify both $p_{ia}$ to allow for the episode possibly being longer than observed (\cref{imperf-test:sec:modifying-p_ia}), and $p_{iu}$ (\cref{imperf-test:sec:modifying-p_iu}) to allow for additional episodes being missed.
+In this section, I introduce a simple model of false negatives (\ie allowing $\psens < 1$) into the likelihood derived in \cref{E-perf-test:sec:model}.
+The simple model, including a constant test sensitivity, means the likelihood remains tractable.
 
-A constant test sensitivity is assumed throughout this section.
-Further work is required to introduce the more realistic assumption of declining test sensitivity over the course of an episode.
+First, in \cref{imperf-test:sec:modifying-p_ia}, I modify $p_{ia}$, the probability of an episode being in the admissible, to allow for the episode possibly being longer than observed.
+Then, in \cref{imperf-test:sec:modifying-p_iu}, I modify $p_{iu}$, the probability of an episode being undetected, to allow for additional episodes being missed.
 
@@
 Consider only the tests between $r_i^{(b)}$ and $l_i^{(e)}$ inclusive (the positive tests providing a lower bound on the length of the episode).
-Denote these tests by $\sched'_i = \{ t \in \sched_i : r_i^{(b)} \leq t \leq l_i^{(e)} \}$ and their results by $y_i'$.
+Denote these test times by $\sched'_i = \{ t \in \sched_i : r_i^{(b)} \leq t \leq l_i^{(e)} \}$ and their results by $y_i'$.
 The test results at times in $\sched_i'$ are either true positives or false negatives. 
@@
 
-With false negatives, an episode with can be undetected in the following four ways; the first two of these are the same as before, but the second two are new.
-Note that each of these is mutually exclusive.
+With false negatives, an episode can be undetected in the following four mutually exclusive ways; the first two of these are the same as in \cref{E-pref-test}, but the second two are new.
 \begin{enumerate}
@@
 
-I repeated the simulation study in \cref{E-perf-test:sec:simulation-study}, except I used the model of \cref{imperf-test:sec:modelling} with either a constant test sensitivity or the more realistic varying sensitivity.
-I focus on the independent (vague) and model combination priors for the survival (described in \cref{E-perf-test:sec:parameters-priors}.
-These were shown to be the best performing priors in \cref{E-perf-test:sec:results}.
+Here, I modify the simulation study in \cref{E-perf-test:sec:simulation-study} to assess several features of the statistical model I proposed in the previous section.
+First, the ability to identify the survival function even with false negatives.
+Second, the impact of misspecifying $\psens$, including violating the assumption that is constant.
+Finally, the impact of the simplying approximations made in \cref{imperf-test:sec:modelling} for tractability.
+These approximations are assuming that the previous negative is a true negative and that there is a negligible probability of missing an episode due to two false negatives.
 
-With a constant test sensitivity of 0.8, the model recovers the true survival time well.
-However, with a lower test sensitivity of 0.6, this is no longer the case (see \cref{imperf-test:fig:constant-test-sensitivity}).
-The simplifications to the model made in \cref{imperf-test:sec:modelling} cause too large a misspecification at this point.
-The informative prior helps to overcome the misspecification, helping to recover the true survival time (in comparison to the vague prior).
+I simulate datasets under four conditions: a constant $\psens$ of 0.6, 0.8, or 1.0 (the latter being the same as the simulated data in \cref{E-perf-test:sec:simulation-study}), or the varying test sensitivity model proposed at the end of \cref{imperf-test:sec:simulate}.
+The number of single positive episodes in each of these conditions, compared to the number observed in the CIS data, is shown in \cref{imperf-test:fig:sim-single-pos}.
+
+For each simulated dataset, I estimate the survival function using the model proposed in \cref{imperf-test:sec:modelling} and assuming $\psens$ equals 0.6, 0.8, or 1.0.
+If the dataset was simulated using a constant $\psens$, and that value of $\psens$ matches the value assumed during inference, I refer to $\psens$ as being correctly specified.
+However, there is still some misspecification of the model to false negatives owing to the simplifying assumptions made.
+The amount that the simplifying assumptions are violated increases as $\psens$ decreases.
+
+I use either the independent (vague) and model combination priors for the survival (described in \cref{E-perf-test:sec:parameters-priors}).
+These were shown to be the best performing priors in \cref{E-perf-test:sec:results}).
+Therefore, there are a total of 24 possible scenarios (each combination of: 4 simulated datasets, 3 values of $\psens$ in inference, and 2 priors).
+
+The results with correctly specified $\psens$ are in \cref{imperf-test:fig:constant-test-sensitivity}.
+When $\psens = 0.8$, the model recovers the true survival time well.
+The informative prior, in comparison to the vague prior, helps to overcome the misspecification due to the simplyifying assumptions, moving the estimated survival function closer to its true survival time.
+However, when $\psens = 0.6$, this is no longer the case (see \cref{imperf-test:fig:constant-test-sensitivity}).
+This is likely caused by too large a violation of the simplifying assumptions made in \cref{imperf-test:sec:modelling}.
 \begin{figure}
@@
     True survival time shown in black.
-    Left panel shows the results with a test sensitivity of 0.6, right panel shows the results with a test sensitivity of 0.8.
+    (A) shows the results with a test sensitivity of 0.6, (B) shows the results with a test sensitivity of 0.8.
   }
@@
 
-This section applies the model described in this chapter to the CIS episodes dataset I described in \cref{E-perf-test:sec:approach}.
+In this section I apply the model described in this chapter to the CIS episodes dataset I described in \cref{E-perf-test:sec:approach}.
 
@@
 \todo{Insert correct citation for RTM paper 2 once available}
-Approximating this distribution as a negative binomial and scaling the mean to the size f the CIS cohort gives the prior $\ntot \sim \NBc(\numprint{25132}, \numprint{22047})$.
+Approximating this distribution as a negative binomial and scaling the mean to the size of the CIS cohort gives the prior $\ntot \sim \NBc(\numprint{25132}, \numprint{22047})$ (distribution defined in).
 
@@
 This estimate has more long episodes than the estimate in \cref{E-ATACCC}.
-The qualitative increase in long episodes is robust to the choice of prior for $\ntot$, the assumed value for $\psens$ (see \cref{imperf-test:fig:cis-sensitivity}), and the choice of prior for $\lambda$.
+The qualitative increase in long episodes is robust to the choice of prior for $\ntot$, the assumed value for $\psens$ (see \cref{imperf-test:fig:cis-sensitivity}), and the choice of prior for the hazard (defined in REF), $\lambda$.
 However, the survival proportion over the first 4 weeks is sensitive to these choices.
@@
   \caption{%
-    Sensitivity of CIS estimates to: (A-C) the value of $r$ in the prior for $\ntot$ and (D-F) $\psens$.
+    Sensitivity of CIS estimates to: (A-C) the value of $r$, the dispersion prior, in the prior for $\ntot$ and (D-F) $\psens$.
     A and D: median survival time, in comparison to the ATACCC-based estimate of \cref{E-ATACCC}.

### 2024-01-16 transmission.tex

--- 2024-01-05 transmission.tex
@@
 
+\todo[inline]{make sure to proofread this section due to merge conflicts}
 As discussed in\todo{ref section}, a major issue with the CIS is non-response bias.
 However, a deterministic relation between true prevalence and incidence is a good approximation (see \cref{E-inc-prev:sec:prevalence-process}).
-
 Therefore, I extend the methodology previously used to estimate prevalence from the CIS data, corrected for non-response bias, to estimate incidence\todo{cite Koen's stuff}.

### 2024-01-16 transmission.tex
Edit SEIR

@@
 This is because the parameters within $\matr{\beta}(t)$ are constant within each week.
-A constant $\matr{\beta}(t)$ leads to exponential growth over short time periods, once the initial dynamics have died out.
-When the peak in incidence is estimated to occur varied by region, but due to this reason, is almost always on a Sunday (see \cref{SEIR:fig:peak-incidence}).
+A constant $\matr{\beta}(t)$ would lead to exponential growth following a period of transitory dynamics, although the weekly switching may mean that these transitory dynamics never end, especially when incidence is high~\autocite{rhodesConvergence}.
+When the peak in incidence is estimated to occur varied by region, but due to the switching, is almost always on a Sunday (see \cref{SEIR:fig:peak-incidence}).
 For Yorkshire, the peak is estimated to occur in late October.
@@
 
-The parameters $\beta_w$ absorb all the time-varying changes in transmission not captured by Google mobility and school attendance data.
-They trend downwards up until early December when they start increasing again.
-Each region is estimated independently, however correlation is clear across the regions.
+The $\beta_w$ parameters absorb all the time-varying changes in transmission not captured by Google mobility and school attendance data.
+They trend downwards up until early December, then they start increasing (see \cref{SEIR:fig:beta-walk}).
+Correlation is clear across the regions even though each region is estimated independently, suggesting underlying biological or epidemiological changes across regions as opposed to modelling assumptions.
 
@@
     \caption{Posterior median and 95\% CrI of the attack rate in each region and age group.}
+    \todo[inline]{Maybe compare this to blood donors, as used in the prior.}
     \label{SEIR:fig:attack-rates}
@@
 
-\todo[inline]{discuss comparisons with the backcalculation, based on notes below (using preliminary results)}
+\section{Comparison of results} \label{transmission:sec:comparison}
+
+\todo[inline]{add comparison to backcalculation once I have those results}
+
+Using a mechanistic model provides additional understanding, compared to the approach in \cref{E-backcalc}.
+For example, decomposing the change in transmission into changing contacts, $\matr{\kappa}(t)$, changing susceptibility $\vec{s}(t)$, and other unmeasured factors, $\beta(t)$.
+By understanding which of these components is changing, policies can be better targeted.
+If it is shown that $\beta(t)$ decreases, this implies that for a given contact rate in the population, there is less transmission.
+Such decreases might be beneficial as it allows society to operate in a more normal manner without increasing transmission.
+Breaking the trade-off between transmission and activity is desirable because it reduces the disease burden without burdesome restrictions on individuals day-to-day lives.
 
@@
 
-\section{Comparison of results} \label{transmission:sec:comparison}
-
-\todo[inline]{add comparison to backcalculation once I have those results}
-
-Using a mechanistic model provides additional understanding, compared to the approach in \cref{E-backcalc}.
-For example, decomposing the change in transmission into changing contacts, $\matr{\kappa}(t)$, changing susceptibility $\vec{s}(t)$, and other unmeasured factors, $\beta(t)$.
-By understanding which of these components is changing, policies can be better targeted.
-If it is shown that $\beta(t)$ decreases, this implies that for a given contact rate in the population, there is less transmission.
-Such decreases might be beneficial as it allows society to operate in a more normal manner without increasing transmission.
-Breaking the trade-off between transmission and activity is desirable because it reduces the disease burden without burdesome restrictions on individuals day-to-day lives.
-
 \section{Conclusion}

### 2024-01-16 transmission.tex
Start adding bits about backcalc

@@
 
-In this chapter, I use the estimate of the duration of PCR positivity from \cref{E-imperf-test} to estimate the transmission of SARS-CoV-2 in England.
+In this chapter, I use the preferred estimate of the duration of PCR positivity (see \cref{E-imperf-test:sec:conclusion}) to estimate the transmission of SARS-CoV-2 in England.
 This estimation is based on the CIS data (see \cref{transmission:sec:data} for further details).
@@
 
-I take two approaches.
-The theoretical advantages of each approach were discussed in \cref{E-inc-prev:sec:infection-process}.
+I take two approaches; the theoretical advantages of each were discussed in \cref{E-inc-prev:sec:infection-process}.
 First, in \cref{backcalc}, I use a backcalculation approach, based on \cref{E-inc-prev:eq:EPt}.
-Then, in \cref{SEIR}, I use a mechanistic compartmental model, extending the model introduced in \cref{E-SEIR:sec:mechanistic-models}.
+Then, in \cref{SEIR}, I use a mechanistic compartmental model, an implementation of the model introduced in \cref{E-SEIR:sec:mechanistic-models}.
 Finally, in \cref{transmission:sec:comparison}, I compare the results of the two approaches.
@@
 
-\todo[inline]{make sure to proofread this section due to merge conflicts}
 As discussed in\todo{ref section}, a major issue with the CIS is non-response bias.
@@
 I modified a pre-existing version of a MRP model for CIS~\autocite{pouwelsMRPvaccination}, as recommended by that model's author for this situation~\citePersonalComms{Koen Pouwels}.
-\todo[inline]{Describe the model precisely, see eg RTM paper 2}
-% The model included fixed effects for age group\todo{check the model I used}.
-% The likelihood is a beta-binomial distribution, a generalisation of the binomial distribution allowing for overdispersion (see \cref{E-distributions}).
-% The overdispersion allows for household clustering\todo{ref beta-binomial as a model for clustering} and any other overdispersion in the data\todo{ref that epi data is often overdispersed}. 
-Finally, the model was fit independently for each region, implicitly adding interactions between region and all other variables.
+First, a Gaussian Markov Random Field (GMRF) logistic regression model is used to estimate the prevalence in each stratum.
+For a day $t$, age group $a$, ethnicity $e$, and sex $s$ assume that:
+\begin{align}
+    \logit(p_{taes}) &= \beta + \beta_s + \beta_{w(e)} + u_a + v_e + w_t + l_{a,t} \\
+    y_{taes} &\dist \BB(n_{taes}, p_{taes}, \rho)
+\end{align}
+The terms are as follows.
+\begin{itemize}
+    \item $\beta$ is a global intercept term with the prior $\beta \dist \N(0, 1/0.001^2)$.
+    \item $\beta_s$ is a fixed effect for sex, with $\beta_s = 0$ for males and estimated for females.
+$\beta_{w(e)}$ is a fixed effect for white or non-white ethnicity.
+$\beta_{w(e)} = 0$ if $e$ corresponds to white ethnicity and $\beta_{w(e)} = \beta_{n}$ otherwise (\ie non-white ethnic groups).% $\beta_n$ and $\beta_f$ have the default INLA prior for fixed effects\todo{check what this is}.
+    \item $u_a$ and $v_e$ are independent Gaussian random effects for age and ethnic group respectively.
+Their standard deviation have a penalised complexity prior\todo{cite PC prior} with a 10\% probability of being greater than 1.
+    \item $w_t$ is a second-order random walk, that is $w_t = 2w_{t-1} - w_{t-2} + \epsilon_t$ where $\epsilon_t \dist \N(0, \sigma_w^2)$.
+    \item $l_{a,t}$ is an age-group specific second-order random walk. That is $l_{a,t} = 2l_{a,t-1} - l_{a,t-2} + \epsilon_{a,t}$ where $\epsilon_{at} \dist \N(0, \sigma_l^2)$.
+    \item The likelihood is a beta-binomial model (defined in \cref{E-distributions}), an overdispersed binomial.
+Overdispersion allows for household clustering\todo{cite beta-binomial as a model for clustering} and other overdispersion in the data, which is common in epidemiological data\todo{cite that epi data is often overdispersed}.
+\end{itemize}
+All parameters have priors using the default specified in INLA, except where stated above.
+For these parameters, the large amount of data means that the prior has little influence on the posterior distribution.
+
+The model was fit separately and independently for each region, implicitly adding interactions between region and all other variables.
 This step assumes that the estimates for each region can be considered independent, which is reasonable because the regions are large with many tests conducted within each.
@@
 Therefore, \cref{E-inc-prev:eq:EPt} implies that $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix of the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
-The matrix $\matr{S}$ preserves the relationship between incidence and prevalence.
+The matrix $\matr{S}$ is defined to maintain the relationship between incidence and prevalence in \cref{E-inc-prev:eq:EPt}.
 This gives a lower-triangular of the following form:
@@
 
+\todo{Insert constant incidence assumption methodology}
+
+Model misspecification means that the procedure described could produce negative incidence.
+For any stratum and day $t$, this occurs if the difference between $P_t$ and $P_{t+1}$ is too large based on the number of recoveries that occur on day $t$ based on the incidence up to day $t$ and the duration distribution.
+
+There are two places where this misspecification could occur.
+The first is the survival matrix, $\matr{S}$.
+This is assumed known but \cref{E-imperf-test} showed that it is sensitive to assumptions made.
+The second is the model used to estimate prevalence.
+
+A prior constraining incidence to be non-negative and then fit directly to the data could be used but, as previously discussed, this would no longer be a GLM and could not be fit within the R-INLA framework.
+Hence, an alternative inference methodology would be required.
+MCMC is likely too computationally expensive.
+Other approaches, such as a Laplace approximation or variational inference, could be considered.
+For this analysis, I replaced any value of incidence estimated to be negative with 0.
+
 The final step is poststratifying the incidence proportion to the CIS target population and subpopulations of interest.
-For each posterior draw and day\todo{introduce discrete time in days earlier}, the incidence proportion in the (sub)population is the weighted average of the incidence proportion in each stratum, where the weights are the proportion of the population in each stratum.
+For each posterior draw and day, the incidence proportion in the (sub)population is the weighted average of the incidence proportion in each stratum, where the weights are the proportion of the population in each stratum.
 The population numbers for each stratum are provided by the Office for National Statistics.
 
-\subsection{Results} \label{backcalc:sec:results}
+\subsection{Results and discussion} \label{backcalc:sec:results}
 
-\subsection{Discussion} \label{backcalc:sec:discussion}
+A slow increase in incidence occurs over the autumn 2020 period, peaking in late November in most regions (see \cref{transmission:fig:backcalc-regions}).
+An exception is the North East and East of England regions (see \cref{transmission:fig:backcalc-regions}(C) and (D) respectively); these regions show no consistent increase or decrease in incidence during October.
+All regions then decrease during November, coinciding with the national lockdown.
+A rise then occurs, although starting at different times in different regions.
+This could be due to the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country\todo{cite Alpha variant}.
+This leads to a second peak in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
+\begin{figure}
+    \todo[inline]{Four-panel regions figure}
+    \capton[Incidence estimated using backcalculation by region]{%
+        Incidence proportion for all of England (A) and its component regions.
+        Regions are split into broad groupings of three regions each: Northern (B), Midlands (C), and Southern (D).
+        Yorkshire and the Humber is labelled as Yorkshire for brevity.
+    }
+    \label{transmission:fig:backcalc-regions}
+\end{figure}
+
+The incidence proportion was highest among the 11--15 and 16--25 age groups (see \cref{transmission:fig:backcalc-ages}).
+Except the 16--25 age group, all age groups show similar trends (\ie parallel lines on a log scale).
+This is the pattern predicted by an age-stratified transmission model (see \cref{E-SEIR:sec:structured-populations}), if behavioural changes are independent of age group, and there is negligible impact of immunity.
+The 16--25 age group decreases faster during November.
+There are many plausible explanations for this effect.
+The higher incidence in this age group prior to November means that this could be an immunity effect.
+Alternatively, it could be a consequence of schools remaining open in the November lockdown, which has a small effect on 16--25 as they interact less with school-age children than other age groups.
+Many other explanations are possible, and further work is required to determine causality.
+\begin{figure}
+    \todo[inline]{age groups figure}
+    \capton[Incidence estimated using backcalculation by age group]{%
+        Incidence proportion for each age group, on both natural and log scales.
+    }
+    \label{transmission:fig:backcalc-ages}
+\end{figure}
+
+An edge effect where a rapid rise is seen between day 1 and 2 is visible (see \cref{transmission:fig:start-effect}).
+This is due to the incorrect assumption that incidence is constant in the past.
+Changing this assumption (\eg to exponential growth) modifies the estimated value on the first day, but the rest of the series is insensitive to this change (not shown).
+Therefore, the first day's estimate is unreliable, and has been excluded from all other plots.
+However, the rest of the series is robust to this assumption and hence reliable.
+\begin{figure}
+    \todo[inline]{start effect figure}
+    \capton[Edge effects in backcalculation method]{%
+        First seven days of incidence proportion estimate for the whole of England, highlighting the edge effect.
+    }
+    \label{transmission:fig:start-effect}
+\end{figure}
 
@@
 
-The modelled proportion of individuals that are PCR positive in the are linked to the data using a Beta-Binomial likelihood (see\todo{ref section on clustering} for the issues with using a binomial likelihood).
+The modelled proportion of individuals that are PCR positive in the are linked to the data using a beta-binomial likelihood (see\todo{ref section on clustering} for the issues with using a binomial likelihood).
 If on each day $t$, each strata $i$ is observed to have $y_{it}$ positives out of $n_{it}$ tests, then the likelihood is:

### 2024-01-16 transmission/incidence.R

@@ -1,0 +1,113 @@
+suppressMessages(library(dplyr))
+library(ggplot2)
+library(tidybayes)
+library(tidyr)
+
+base_data_dir = here::here("transmission/outputs")
+load_prevalence = function() {
+    readRDS(file.path(base_data_dir, "predict.rds")) |>
+    left_join(
+        readr::read_csv(
+            file.path(base_data_dir, "groups.csv"),
+            show_col_types = FALSE
+        ),
+        by = ".group"
+    ) |>
+    mutate(prev_predict = expit(.predict))
+}
+
+load_incidence = function() {
+    readRDS(file.path(base_data_dir, "deconv.rds"))
+}
+
+group_by_strata = function(x, ...) {
+  all_cols = rlang::exprs(region, daynr, sex, age_group, ethnicityg)
+  exclude = rlang::ensyms(...)
+  stopifnot(all(exclude %in% all_cols))
+  include = setdiff(all_cols, exclude)
+  return(group_by(x, !!!include))
+}
+
+load_poststrat_table = function() {
+    readr::read_csv(file.path(base_data_dir, "poststrat.csv"), show_col_types = FALSE) |>
+        select(!.groups) |>
+        mutate(
+            region = case_match(
+                Region_Name,
+                "North_East_England"  ~ "1_NE",
+                "North_West_England"  ~ "2_NW",
+                "Yorkshire"  ~ "3_YH",
+                "East_Midlands"  ~ "4_EM",
+                "West_Midlands"  ~ "5_WM",
+                "East_England"  ~ "6_EE",
+                "London"  ~ "7_LD",
+                "South_East_England"  ~ "8_SE",
+                "South_West_England"  ~ "9_SW",
+                .default = NA_character_
+            )
+        ) |>
+        assertr::verify(
+            !is.na(region)
+            | Region_Name %in% c("Wales", "Scotland", "Northern_Ireland")
+        ) |>
+        filter(!is.na(region))
+}
+
+poststratify = function(data, postrat_table, col, ...) {
+    left_join(
+        data,
+        postrat_table,
+        by = c("region", "age_group", "ethnicityg" = "ethnicity", "sex")
+    ) |>
+        assertr::assert(assertr::not_na, pop) |>
+        mutate(n = {{ col }} * pop) |>
+        group_by(daynr, .draw, !!!ensyms(...)) |>
+        summarise(
+            N = sum(pop),
+            val = sum(n) / N,
[…]

### 2024-01-16 cis-imperfect-testing.tex
Label sub-plots

@@
 
+\todo[inline]{The first paragraph needs restructuring. Dani will send some ideas.}
 In \cref{E-perf-test}, a framework for analysing the CIS to produce estimates of the duration of positivity was developed.
@@
 Applying the method of \cref{E-perf-test} to CIS data gives implausible estimates of the duration distribution (see \cref{imperf-test:fig:problem-cis-estimates}).
-In particular, it estimated a much shorter duration of positivity than in \cref{E-ATACCC}, despite those estimates being much more reliable over the first 2--3 weeks.
+In particular, it estimated a much shorter duration of positivity than in \cref{E-ATACCC}.
+The \cref{E-ATACCC} estimates are more reliable over the first 2--3 weeks due to them having dense data collection here.
+This suggests an issue in the estimation using CIS data.
 \begin{figure}
@@
   \caption[Estimating survival using CIS data assuming perfect testing]{Estimates of the survival function using CIS data alongside the ATACCC-based estimates of \cref{E-perf-test}.
-CIS-based is implausibly short over the first 14 days.
+CIS-based has more episodes ending in the first 14 days.
 \label{imperf-test:fig:problem-cis-estimates}}
@@
 
-Comparing the simulated data in the previous chapter and CIS data suggests that this is due to the high frequency of single positive episodes in the CIS data.
-A single positive episode is an episode containing exactly one positive test; intuitively, these cause short estimates because there is no non-zero lower bound on the length of time the episode lasts.
-Furthermore, short episodes are very likely to be missed.
-Therefore, there is little or no ability to upper bound the number of very short episodes.
-I found that simulations that included even a few excess single positive episodes led to problematic estimation within the CIS setting (not shown).
+Comparing repeated data simulations using the setup of \cref{E-perf-test:sec:simulation-study} and the observed CIS data suggests that this is due to the high frequency of single positive episodes in the CIS data (compare the histogram with $\psens=1$ and the vertical line in \cref{imperf-test:fig:sim-single-pos}).
+A single positive episode is an episode containing exactly one positive test; intuitively, these cause short estimates because the lower bound on the length of time the episode is 1 day, the shortest possible.
+Specifically, if $i$ is a single positive episode then $i$'s admissible region (as defined in \cref{perf-test:sec:model}) includes $b_i = e_i = r_i^{(b)} = l_i^{(e)}$, the beginning and end day of the episode both being equal to the day of this one positive test.
+If this is true, then the duration $d_i = e_i - b_i + 1 = 1$.
+Additionally, short episodes are very likely to be missed; therefore, the data is compatible with a large number of missed short episodes.
+Therefore, single positive episodes suggest 
+I confirmed this hypothesis by estimating the survival function in one of these simulated datasets with the addition of even a few single positives.
+This produced the same pattern, that is estimating too many short episodes (not shown).
 
@@
 Intermittent negatives (described in  \cref{E-biology-data:sec:PCR}) show that they occur within the CIS.
-Second, adjusting the simulation to include false negatives can reproduce both the CIS data more faithfully and exhibits the same issues when estimating the duration distribution (see \cref{imperf-test:fig:sim-single-pos}).
-Simulation methodology is explained in \cref{E-perf-test:sec:simulation-study}, with false negatives included as described in \cref{imperf-test:sec:simulate}.
+Second, adjusting the simulation to include false negatives can reproduce both the CIS data more faithfully, and exhibits the same issues when estimating the duration distribution (see \cref{imperf-test:fig:sim-single-pos}).
+The simulation setup expands the method of \cref{E-perf-test:sec:simulation-study}, including false negatives (see \cref{imperf-test:sec:simulate} for more details).
 \begin{figure}
@@
     y-axis is the number of simulations, out of 1000, with the indicated proportion.
+    Test sensitivity is $\psens$.
     The CIS data is shown as a vertical line.
@@
 False negatives are more likely to occur when the viral load of an individual is low, because there is less virus in their body to be sampled.
-In \cref{E-ATACCC}, this mechanism is incorporated by treating negative test results as possible left-censored observations.
+In \cref{E-ATACCC}, this mechanism is incorporated by viewing negative tests as a left-truncation of the observation noise (see \cref{E-ATACCC:sec:observation-modification}).
 This observation suggests that a model with a declining test sensitivity as a function of time since infection might be more suitable.
@@
 Here, I am assuming that the probability of two false negative tests at the end of an episode is negligible.
-Then I calculate the test sensitivity as a function of days since the first positive in an episode, bounding this by assuming either all or none of those in group 3 are false negatives.
-The test sensitivity is the true positives / (true positives + false negatives).
+As test sensitivity is the proportion of individuals who should test positive that do so, it can be empirically calculated as number of true positives / (number of true positives + number of false negatives).
+Therefore, I calculate the test sensitivity for each day, considering the first positive result in an episode as day 0.
+As the latent start day of the infection episode is before the first positive result, day 0 reflects some unknown number of days into the infection episode.
+I bound the test sensitivity on each day by assuming either all or none of the possible false negatives (group 3) are in fact false negatives.
 % In \cref{imperf-test:fig:bounding-cis-sensitivity} we consider the values produced by assuming the negative following the last positive in an episode could be either a true or false negative as a function of time since the individual was detected (\ie: the first positive test in the episode).
 This generates broad bounds, although suggest a declining test sensitivity over the course of an episode.
-Small numbers of tests mean that the series is noisy near the start and the end, such as.
+Small numbers of tests mean that the series is noisy near the start and the end, such as both bounds being 0 after day 40.
+However, the general trend is clear.
 \begin{figure}
@@
 
-Based on \cref{imperf-test:fig:bounding-cis-sensitivity}, I propose the following model for test sensitivity, consisting of a linearly declining period then a constant period:
+To check sensitivity of the inference, it will be useful to simulate from a model with a varying test sensitivity as a function of days since the start of the infection episode.
+Through consideration of \cref{imperf-test:fig:bounding-cis-sensitivity}, and the interaction of false negatives due to viral load in \cref{E-ATACCC}'s results, I propose the following model.
 \begin{equation}
@@
 \end{equation}
-where $t$ is the time in days since infection.
-Adapting the simulation study to incorporate this test sensitivity is much more similar to the data (see purple histogram in \cref{imperf-test:fig:sim-single-pos}), while being more plausible than very low probabilities of false negatives.
+where $t$ is the number of days since infection.
+Therefore, this model consists of a linear decline in test sensitivity, before a constant minimum point.
+Adapting the simulation study to incorporate this test sensitivity is much more similar to the data (see purple histogram in \cref{imperf-test:fig:sim-single-pos}).
+This model does not require low test sensitivities early in the infection, maintaining compatibility with the results in \cref{E-ATACCC}.
 
@@
 
-I modify $p_{ia}$ to allow the negative test following the last positive to be a false negative.
-If it is a false negative, then we consider the episode's length right-censored.
-However, if it is in fact a true negative, then we are in the same case as before with a bound on the length of the episode.
-A mixture of these scenarios then forms the episode's likelihood contribution, with the mixture probability determined by the test sensitivity.
+\todo[inline]{These definitions (up until next orange box) will move to chapter 2 where episodes are introduced}
+The beginning of episode $i$ is known to occur in the interval $[l_i^{(b)}, r_i^{(b)}]$, and similarly for the end of the infection in $[l_i^{(e)}, r_i^{(e)}]$.
 
-For tractability, assume that the negative test bounding the start of the infection, at $l_i^{(b)}-1$, is a true negative.
-This assumption is reasonable because the test sensitivity is high early in an infection, therefore this test is unlikely to be a false negative.
-Consider only the tests between $r_i^{(b)}$ and $l_i^{(e)}$ inclusive (the positive tests providing a lower bound on the length of the episode).
-Denote these test times by $\sched'_i = \{ t \in \sched_i : r_i^{(b)} \leq t \leq l_i^{(e)} \}$ and their results by $y_i'$.
-The test results at times in $\sched_i'$ are either true positives or false negatives. 
-By definition, the test at $r_i^{(e)}$ is a false negative if and only if $e_i > r_i^{(e)}$.
+\todo[inline]{From here, this is the section's content}
+
+I will modify $p_{ia}$ to allow the negative test following the last positive to be a false negative.
+If it is a false negative, then I will consider the episode's end right-censored.
+However, if it is a true negative, then the episode's end is interval-censored as previously.
+A mixture of these scenarios forms the episode's likelihood contribution, with the mixture probability determined by the test sensitivity.
+
+I start by finding the tests that need consideration.
+For tractability, assume that the negative test bounding the start of the infection, on day $l_i^{(b)}-1$, is a true negative.
+This assumption is reasonable because the test sensitivity is high early in an infection; therefore, this test is unlikely to be a false negative.
+For tractability and simplicity, I consider only tests between the positive tests providing a lower bound on the length of the episode.
+By definition (see\todo{insert where these are defined in chapter 2}), these are the tests between $r_i^{(b)}$ and $l_i^{(e)}$ inclusive.
+They are a subset of $\sched_i$, the set of test times for the individual in which episode $i$ occurs.
+Denote this subset as $\sched'_i = \{ t \in \sched_i \ssep r_i^{(b)} \leq t \leq l_i^{(e)} \}$, and their results by $\vec{y_i'} = \{ y_i(t) \ssep t \in \sched'_i \}$ where $y_i(t) = 1$ if the test on day $t$ is positive and 0 otherwise.
+As I assume that there are no false positives, the infection episode must span at least this period starting and ending with a positive test.
+Therefore, the test results at times in $\sched_i'$ are either true positives or false negatives.
+
+Now consider the negative test at $r_i^{(e)}+1$, the first negative after the start of the episode which may be a false negative.
+It is a false negative if and only if the episode ends after the test, \ie $e_i > r_i^{(e)}$.
 I proceed by considering whether this is the case.
 
-First, if $e_i \leq r_i^{(e)}$.
+First, if $e_i \leq r_i^{(e)}$, meaning that the test at $r_i^{(e)}+1$ is a true negative and the end of the episode is interval-censored as in the previous chapter.
 In this case, the test at $r_i^{(e)}$ is a true negative, as are all other tests not in $\sched'_i$.
 These occur with probability 1, as I continue to assume no false positives.
+The tests in $\sched_i'$ are either true positives or false negatives.
 \begin{align}
-&p(y_i', e_i \leq r_i^{(e)} | b_i, p_\text{sens}, \theta) \\
-&= p(y_i', l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\ % &\text{as no false positives}
-&= p(y_i' \mid l_i^{(e)} \leq e_i \leq r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
+&p(\vec{y_i'}, e_i \leq r_i^{(e)} | b_i, p_\text{sens}, \theta) \\
+&= p(\vec{y_i'}, l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\ % &\text{as no false positives}
+&= p(\vec{y_i'} \mid l_i^{(e)} \leq e_i \leq r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
 &= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) \right)
@@
 Second, if $e_i > r_i^{(e)}$.
-In this case, the test at $r_i^{(e)}$ is a false negative, occurring with probability $(1 - p_\text{sens})$, and the end of the infection is right-censored.
+In this case, the test at $r_i^{(e)}$ is a false negative, occurring with probability $(1 - p_\text{sens})$.
+As I only consider the tests in $\sched_i'$, the end of the episode is right-censored.
+Taking the same approach as before:
 \begin{align}
-&p(y_i', e_i > r_i^{(e)} | b_i, p_\text{sens}, \theta) \\
+&p(\vec{y_i'}, e_i > r_i^{(e)} | b_i, p_\text{sens}, \theta) \\
 &= p(y_i \mid e_i > r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(e_i > r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
@@
 p_{ia}'
-&= p(y_i' \mid p_\text{sens}, \theta) \\
-&= \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( p(y_i', b_i, e_i \leq r_i^{(e)} \mid p_\text{sens}, \theta) p(y_i', b_i e_i > r_i^{(e)} \mid p_\text{sens}, \theta) \right) p(b_i \mid p_\text{sens}, \theta) \\
+&= p(\vec{y_i'} \mid p_\text{sens}, \theta) \\
+&= \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( p(\vec{y_i'}, b_i, e_i \leq r_i^{(e)} \mid p_\text{sens}, \theta) p(\vec{y_i'}, b_i e_i > r_i^{(e)} \mid p_\text{sens}, \theta) \right) p(b_i \mid p_\text{sens}, \theta) \\
 &= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \\ & \ \times \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) + (1 - p_\text{sens}) S_\theta(r_i^{(e)} - b_i - 1) \right) \\ & \ \times p(b_i \mid p_\text{sens}, \theta) \\
@@
 
-With false negatives, an episode can be undetected in the following four mutually exclusive ways; the first two of these are the same as in \cref{E-pref-test}, but the second two are new.
+False negatives add additional mechanisms by which episode $i$ could be undetected, in addition to those considered when deriving $p_{iu}$ in \cref{E-perf-test:eq:piu}.
+Specifically, episode $i$ could be undetected if the first test after the $b_i$, episode $i$'s start day, is a false negative and then there are no subsequent positive tests.
+
+This false negative would occur on day $b_i + \tau_{\sched_i}(b_i)$ (see\todo{ref where this was derived}).
+A false negative occurring requires that the episode has not yet ended but a negative still occurs.
+The episode has not yet ended at the time of the test if $e_i = b_i + d_i - 1$ is after the test, that is $d_i > \tau_{\sched_i}(b_i)$.
+Conditional on the episode having not yet ended, the test result is negative with probability $1 - \psens$.
+
+For there to be no subsequent positive tests, all tests up until day $e_i$ are false negatives.
+I assume there is a negligible probability of two false negatives.
+Therefore, this can only occur if the episode ends before another test.
+Denote the number of days between $b_i$ and the test following the false negative as $\tau^2_{\sched_i}(b_i) \stackrel{\text{def}}{=} \tau_{\sched_i}(\tau_{\sched_i}(b_i) + 1)$.
+Then, the episode ends before this test if $d_i < \tau^2_{\sched_i}(b_i)$.
+
+Therefore, this mechanism causes episode $i$ to be undetected if all of the following conditions hold.
 \begin{enumerate}
-\item
-  The episode begins before the individual's first test: $b_i \leq \min \sched_i$.
-  % I assume there is a negligible probability of a false negative test at this point because viral load is high.
-\item
-  The episode begins after the individual's first test, but ends before the next test after beginning: $b_i \leq \min \sched_i$ and $e_i < b_i + \tau_{\sched_i}(b_i)$.
-\item
-  The episode begins after the individual's first test, ends after the first test at or following $b_i$, that test is a false negative, but the episode ends before the second test at or following $b_i$.
-  Then, $D_i$ of the episode is greater than $\tau_{\sched_i}(b_i)$ less than  $\tau^2_{\sched_i}(b_i) \stackrel{\text{def}}{=} \tau_{\sched_i}(\tau_{\sched_i}(b_i) + 1)$, and a false negative episode occurred at the first test after $t$.
-  This probability, conditional on $b_i$ is:
+    \item The episode would have been detected considering only the mechanisms in \cref{E-perf-test:eq:piu}. That is $b_i > \min(\sched_i)$ and $\tau_{\sched_i}(b_i) + b_i \leq e_i$.
+    \item The episode ends in the interval $[\tau_{\sched_i}(b_i) + b_i, \tau^2_{\sched_i}(b_i) + b_i]$. Equivalently, $\tau_{\sched_i}(b_i) + 1 \leq d_i < \tau^2_{\sched_i}(b_i) + 1$.
+    \item A false negative occurs on day $\tau_{\sched_i}(b_i) + b_i$. Conditional on the previous condition, this occurs with probability $1 - \psens$.
+\end{enumerate}
+
+The probability of this occurring, conditional on $B_i = b_i > \min(\sched_i)$ is:
 \begin{align}
-    &\prob \left( b_i + \tau_{\sched_i}(b_i) \leq E_i < b_i + \tau^2_{\sched_i}(b_i), Y_i(b_i + \tau_{\sched_i}(b_i) = 0) \mid b_i, \theta \right) \\
-    &= \prob \left(\tau_{\sched_i}(b_i) + 1 \leq D_i < \tau^2_{\sched_i}(b_i) + 1 \mid b_i, \theta \right) (1 - \psens) \\
+&\prob \left(
+    \tau_{\sched_i}(b_i) + 1 \leq D_i < \tau^2_{\sched_i}(b_i) + 1
+    \mid b_i,
+\theta \right) (1 - \psens) \\
 &= \left( S_\theta(\tau_{\sched_i}(b_i) + 1) - S_\theta(\tau^2_{\sched_i}(b_i) + 1) \right) (1 - \psens).
@@
 \begin{align}
-    (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_i)}^T \left( S_\theta(\tau_{\sched_i}(b) + 1) - S_\theta(\tau^2_{\sched_i}(b) + 1) \right).
+\eta = (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_i)}^T \left( S_\theta(\tau_{\sched_i}(b) + 1) - S_\theta(\tau^2_{\sched_i}(b) + 1) \right).
 \end{align}
-\item
-  The episode begins after the individual's first test, ends after the second test at or following $b_i$, and all tests within the episode (at least two) are false negatives.
-  I assume this occurs with negligible probability.
-\end{enumerate}
 
-Therefore, $p_{iu}$ is replaced by $p_{iu}'$, the sum of: $p_{iu}$ (for number 1 and 2; derived in \cref{E-perf-test:eq:piu}), and the probability in number 3.
-For the posterior density (\cref{E-perf-test:eq:full-posterior}), $1 - p_{iu}'$ is the required quantity:
+Denote the replacement for $p_{iu}$ as $p_{iu}'$.
+Since the new mechanism is mutually exclusive with the previous one, $p_{iu}' = p_{iu} + \eta$ is the sum of $p_{iu}$.
+For the posterior density (\cref{E-perf-test:eq:full-posterior}), $1 - p_{iu}'$ is the required quantity.
 \begin{align}
@@
 
-The results with correctly specified $\psens$ are in \cref{imperf-test:fig:constant-test-sensitivity}.
-When $\psens = 0.8$, the model recovers the true survival time well.
+When $\psens = 0.8$ and is correctly specified, the model recovers the true survival time well (see \cref{imperf-test:fig:constant-test-sensitivity}(B)).
 The informative prior, in comparison to the vague prior, helps to overcome the misspecification due to the simplyifying assumptions, moving the estimated survival function closer to its true survival time.
-However, when $\psens = 0.6$, this is no longer the case (see \cref{imperf-test:fig:constant-test-sensitivity}).
+However, when $\psens = 0.6$, this is no longer the case (see \cref{imperf-test:fig:constant-test-sensitivity}(A)).
 This is likely caused by too large a violation of the simplifying assumptions made in \cref{imperf-test:sec:modelling}.
@@
   \caption[Simulation study results with constant test sensitivity]{%
-    Posterior (median and 95\% credible interval) of the survival time for the simulation study with a constant test sensitivity.
+    Posterior (median and 95\% credible interval) of the survival time for the simulation study with a correctly specified test sensitivity.
     True survival time shown in black.
-    (A) shows the results with a test sensitivity of 0.6, (B) shows the results with a test sensitivity of 0.8.
   }
@@
 If the test sensitivity is misspecified, that is the assumed value for $p_\text{sens}$ in \cref{imperf-test:eq:pit-prime,imperf-test:eq:pit-prime} is different to the value used in the simulation, then the survival time will be biased (see \cref{imperf-test:fig:misspecified-test-sensitivity}).
-If the test sensitivity is assumed to be too low, then the number of episodes inferred to have truly ended by the first negative will be too high, and the survival time will be underestimated.
+If $\psens$ is misspecified and too low (see \cref{imperf-test:fig:misspecified-test-sensitivity}(A)), then the posterior estimate initially follows the true value but then separates,
+The number of episodes inferred to have truly ended by the first negative is too high, and hence the survival time is underestimated.
 This affect dominates over the opposing bias of overestimating the number of missed episodes.
-The opposite occurs if the test sensitivity is assumed to be too high.
+The opposite occurs if the test sensitivity is assumed to be too high, although the posterior moves away from the truth earlier (see \cref{imperf-test:fig:misspecified-test-sensitivity}(C)).
 \begin{figure}
@@
   \caption[Simulation study results with misspecified test sensitivity]{%
-    Posterior (median and 95\% credible interval) of the survival time for the simulation study with a misspecified test sensitivity.
+    Posterior (median and 95\% credible interval) of the survival time for the simulation study with a possibly misspecified test sensitivity.
     True survival time shown in black.
-    All simulations use a constant test sensitivity of 0.8, but the model assumes different values.
-    All inference uses the combination prior, due to it performing better than the vague prior in \cref{imperf-test:fig:constant-test-sensitivity}.
-    The correctly specified model (middle pane) has a posterior median closely following the true value, with the credible intervals almost always including the true value.
-    Assuming a too low test sensitivity (left pane) starts by following the true value but then separates, with the posterior median being too high.
-    Assuming a too high (perfect) test sensitivity (right pane) underestimates the survival time almost throughout.
+    All simulations use a constant test sensitivity of 0.8, but the inference procedure assumes different values, as per the key.
+    Hence (B) is correctly specified, and the same as in \cref{imperf-test:fig:constant-test-sensitivity}, but (A) and (C) have misspecified $\psens$.
+    All results uses the combination prior, due to it performing better than the vague prior in \cref{imperf-test:fig:constant-test-sensitivity}.
   }
@@
 In this section I apply the model described in this chapter to the CIS episodes dataset I described in \cref{E-perf-test:sec:approach}.
+This is the infection episodes with start dates between and\todo{check dataset inclusion criteria}.
 
@@
 \todo{Insert correct citation for RTM paper 2 once available}
-Approximating this distribution as a negative binomial and scaling the mean to the size of the CIS cohort gives the prior $\ntot \sim \NBc(\numprint{25132}, \numprint{22047})$ (distribution defined in).
+Approximating this distribution as a negative binomial and scaling the mean to the size of the CIS cohort gives the prior $\ntot \sim \NBc(\numprint{25132}, \numprint{22047})$ (distribution defined in \cref{E-distributions}).
 
@@
 This estimate has more long episodes than the estimate in \cref{E-ATACCC}.
-The qualitative increase in long episodes is robust to the choice of prior for $\ntot$, the assumed value for $\psens$ (see \cref{imperf-test:fig:cis-sensitivity}), and the choice of prior for the hazard (defined in REF), $\lambda$.
+The qualitative increase in long episodes is robust to the choice of prior for $\ntot$, the assumed value for $\psens$ (see \cref{imperf-test:fig:cis-sensitivity}), and the choice of prior for the hazard (see \cref{perf-test:sec:parameters-priors}), $\lambda$.
 However, the survival proportion over the first 4 weeks is sensitive to these choices.
@@
   \caption{%
-    Sensitivity of CIS estimates to: (A-C) the value of $r$, the dispersion prior, in the prior for $\ntot$ and (D-F) $\psens$.
+    Sensitivity of CIS estimates to: (A-C) the value of $r$, the dispersion of the prior for $\ntot$, and (D-F) $\psens$.
     A and D: median survival time, in comparison to the ATACCC-based estimate of \cref{E-ATACCC}.
@@
 Ideally, the test sensitivity would be estimated from the data.
-However, this would require a model for time-varying test sensitivity to be incorporated into the model.
+However, this would require incorporating time-varying test sensitivity into the likelihood.
 Furthermore, it is unclear that such a model would be identifiable because it would not be clear whether to attribute negative results to false negatives or to recoveries\todo{cite identifiability issues with varying sensitivity}.

### 2024-01-16 cis-imperfect-testing/sim_survival.R

@@ -6,6 +6,36 @@ library(tidybayes)
 library(tidyr)
 source(here::here("utils.R"))
 
+base_plot = function(df, colour_curves_by, colour_key = NULL) {
+    p = df |>
+        mutate(
+            sensitivity.model = factor(sensitivity.model),
+            facet_label = latex2exp::TeX(glue::glue(
+                "({LETTERS[dense_rank(sensitivity.model)]})\\ $p_{{sens}} = {sensitivity.model}"
+            ), output = "character")
+        ) |>
+        ggplot() +
+        ggdist::geom_lineribbon(
+            aes(
+                time, S, ymin = S.lower, ymax = S.upper,
+                fill = {{ colour_curves_by }}, colour = {{ colour_curves_by }}
+            ),
+            linewidth = 1, alpha = 0.3
+        ) +
+        facet_wrap(~facet_label, labeller = label_parsed) +
+        geom_line(aes(time, S), data = truth, alpha = 0.5) +
+        theme_survival_time_series()
+
+    if (!is.null(colour_key)) {
+        p = p +
+            labs(
+                colour = colour_key,
+                fill = colour_key
+            )
+    }
+    return(p)
+}
+
 tbl_posteriors = readRDS(here::here("cisRuns-output/all_posteriors.rds")) |>
     filter(
         .width == 0.95, # Credible intervals are narrow anyway so only show 95%
@@ -30,11 +60,7 @@ truth = readRDS(here::here("cisRuns-output/input_curves.rds")) |>
 
 p_constant_sensitivity = tbl_posteriors |>
     filter(sensitivity.simulation == sensitivity.model, sensitivity.simulation < 1) |>
-    ggplot() +
-    ggdist::geom_lineribbon(aes(time, S, ymin = S.lower, ymax = S.upper, fill = survival_prior, colour = survival_prior), linewidth = 1, alpha = 0.3) +
-    facet_wrap(~ sensitivity.model) +
-    geom_line(aes(time, S), data = truth) +
-    theme_survival_time_series()
+    base_plot(survival_prior)
 ggsave(
     filename = here::here("cis-imperfect-testing/sim-constant-sensitivity.pdf"),
     plot = p_constant_sensitivity,
@@ -46,16 +72,7 @@ ggsave(
 
 p_misspecified_sensitivity = tbl_posteriors |>
[…]

### 2024-01-16 cis-imperfect-testing.tex
include false negative estimates

@@
 I modified the simulation study from \cref{E-perf-test:sec:simulation-study} to include the possibility of false negatives.
-More single positive episodes are seen at lower values of $\psens$ (see \cref{imperf-test:fig:sim-single-pos}).
-A value between 60\% and 80\% for $\psens$ reproduces the number of single positive episodes observed within the CIS data.
-In \cref{E-ATACCC}, I estimated that the rate of false negatives is\todo{add what the ATACCC false negative rate is}.
-The daily testing in the ATACCC study, used for the \cref{E-ATACCC} estimates, should provide a reliable estimate of the false negative rate during the period for which the individuals are followed-up\todo{rephrase this sentence, currently unclear}.
-The requirement for a much lower value of $\psens$ suggests this model of false negatives is inadequate.
+More single positive episodes are seen at lower values of $\psens$, with a value between 60\% and 80\% reproducing the number of single positive episodes observed in the CIS data (see \cref{imperf-test:fig:sim-single-pos}).
+In \cref{E-ATACCC}, I estimated that $\psens$ ($1-\rho$ in \cref{E-ATACCC}'s parameterisation) is 95\% (95\% CrI: 96\%--93\%).
+The daily testing in the ATACCC study, used for the \cref{E-ATACCC} estimates, means this is a reliable estimate during the period for which the individuals are followed-up.
+The requirement for a much lower value of $\psens$ in the simulation study suggests this model of false negatives is inadequate.
 

### 2024-01-16 cis-imperfect-testing.tex
Fix small issues

@@
 More single positive episodes are seen at lower values of $\psens$, with a value between 60\% and 80\% reproducing the number of single positive episodes observed in the CIS data (see \cref{imperf-test:fig:sim-single-pos}).
-In \cref{E-ATACCC}, I estimated that $\psens$ ($1-\rho$ in \cref{E-ATACCC}'s parameterisation) is 95\% (95\% CrI: 96\%--93\%).
+In \cref{E-ATACCC}, I estimated that $\psens$ ($1-\rho$ in \cref{E-ATACCC}'s parameterisation) is 95\% (95\% CrI: 93--96\%).
 The daily testing in the ATACCC study, used for the \cref{E-ATACCC} estimates, means this is a reliable estimate during the period for which the individuals are followed-up.
@@
 
-This false negative would occur on day $b_i + \tau_{\sched_i}(b_i)$ (see\todo{ref where this was derived}).
+This false negative would occur on day $b_i + \tau_{\sched_i}(b_i)$ (see \cref{E-perf-test:sec:model}).
 A false negative occurring requires that the episode has not yet ended but a negative still occurs.
@@
 In this section I apply the model described in this chapter to the CIS episodes dataset I described in \cref{E-perf-test:sec:approach}.
-This is the infection episodes with start dates between and\todo{check dataset inclusion criteria}.
+This is the 4,800 infection episodes that were first detected between 10th Oct 2020 and 6th Dec 2020 inclusive, with negatives bounding the start and end time of the episode.
 

### 2024-01-17 cis-imperfect-testing.tex
proof-read

@@
 
-Comparing repeated data simulations using the setup of \cref{E-perf-test:sec:simulation-study} and the observed CIS data suggests that this is due to the high frequency of single positive episodes in the CIS data (compare the histogram with $\psens=1$ and the vertical line in \cref{imperf-test:fig:sim-single-pos}).
+Comparing repeated data simulations using the setup of \cref{E-perf-test:sec:simulation-study} to the observed CIS data, suggests that this is due to the high frequency of single positive episodes in the CIS data (compare the histogram with $\psens=1$ and the vertical line in \cref{imperf-test:fig:sim-single-pos}).
 A single positive episode is an episode containing exactly one positive test; intuitively, these cause short estimates because the lower bound on the length of time the episode is 1 day, the shortest possible.
-Specifically, if $i$ is a single positive episode then $i$'s admissible region (as defined in \cref{perf-test:sec:model}) includes $b_i = e_i = r_i^{(b)} = l_i^{(e)}$, the beginning and end day of the episode both being equal to the day of this one positive test.
-If this is true, then the duration $d_i = e_i - b_i + 1 = 1$.
-Additionally, short episodes are very likely to be missed; therefore, the data is compatible with a large number of missed short episodes.
-Therefore, single positive episodes suggest 
-I confirmed this hypothesis by estimating the survival function in one of these simulated datasets with the addition of even a few single positives.
+Specifically, if $i$ is a single positive episode then $i$'s admissible region (as defined in \cref{E-perf-test:sec:model}) includes $b_i = e_i = r_i^{(b)} = l_i^{(e)}$.
+In this case, the beginning and end day of the episode are both equal to the day of this one positive test.
+If this is true, then the duration is $d_i = e_i - b_i + 1 = 1$.
+Additionally, short episodes are very likely to be missed, which means that the design of the CIS cannot rule out many of them occurring.
+\todo{possibly a bit unclear phrasing here? struggling with it}
 This produced the same pattern, that is estimating too many short episodes (not shown).
@@
 \Textcite{shenNonparametrica}, extending \textcite{panNote}, studied the situation when only the terminating event is interval censored and the initiating event has simple left truncation.
-This is a related, although simpler, setting than the CIS.
-The CIS has double interval censoring and a more complex pattern of missed episodes (see \cref{E-perf-test:sec:problem}).
-\Textcite{shenNonparametrica} showed that, in their situation, the maximum likelihood estimator is inconsistent if there is at least one single positive episode (check the precise statement) and can lead to a severely biased underestimation of the survival function.
-While the result cannot simply be extended to the more complex study design of the CIS, it reinforces the intuition that single positive episodes are likely to be problematic.
+They showed that the maximum likelihood estimator is inconsistent if there is at least one single positive episode\todo{check the precise results and condition} and can lead to a severely biased underestimation of the survival function.
+The CIS is a more complex setting, including double interval censoring and a complex pattern of missed episodes (see \cref{E-perf-test:sec:problem}).
+This means the result cannot simply be extended to the CIS, but still reinforces the intuition that single positive episodes are likely to be problematic.
 
@@
 Intermittent negatives (described in  \cref{E-biology-data:sec:PCR}) show that they occur within the CIS.
-Second, adjusting the simulation to include false negatives can reproduce both the CIS data more faithfully, and exhibits the same issues when estimating the duration distribution (see \cref{imperf-test:fig:sim-single-pos}).
-The simulation setup expands the method of \cref{E-perf-test:sec:simulation-study}, including false negatives (see \cref{imperf-test:sec:simulate} for more details).
+Second, I expanded the simulation study of \cref{E-perf-test:sec:simulation-study} to include false negatives (see \cref{imperf-test:sec:simulate} for more details).
+This reproduced the CIS data more faithfully and exhibited the same issues when estimating the duration distribution (see \cref{imperf-test:fig:sim-single-pos}).
 \begin{figure}
@@
 \todo{check if detectable and test sensitivity defined somewhere, and in a consistent way}
-In \cref{E-perf-test}, the outcome of a test was deterministic, positive if the individual being tested had an ongoing infection episode.
+In \cref{E-perf-test}, the outcome of a test was deterministic: always positive if the individual being tested had an ongoing infection episode.
 Now, the outcome of such a test result is a Bernoulli random variable with a probability $\psens$ of producing a positive.
@@
 As previously, tests in individuals without ongoing infection episodes are always negative.
+I modified the simulation study from \cref{E-perf-test:sec:simulation-study} to generate test results in this way.
 
-I modified the simulation study from \cref{E-perf-test:sec:simulation-study} to include the possibility of false negatives.
-More single positive episodes are seen at lower values of $\psens$, with a value between 60\% and 80\% reproducing the number of single positive episodes observed in the CIS data (see \cref{imperf-test:fig:sim-single-pos}).
+Following the modification, more single positive episodes are seen at lower values of $\psens$, with a value between 60\% and 80\% reproducing the number of single positive episodes observed in the CIS data (see \cref{imperf-test:fig:sim-single-pos}).
+This is expected: a lower $\psens$ means that multiple positive tests are less likely, and hence the probability of a single positive episode is higher.
+
 In \cref{E-ATACCC}, I estimated that $\psens$ ($1-\rho$ in \cref{E-ATACCC}'s parameterisation) is 95\% (95\% CrI: 93--96\%).
 The daily testing in the ATACCC study, used for the \cref{E-ATACCC} estimates, means this is a reliable estimate during the period for which the individuals are followed-up.
-The requirement for a much lower value of $\psens$ in the simulation study suggests this model of false negatives is inadequate.
+The requirement for a much lower value of $\psens$ in the simulation study here suggests this model of false negatives is inadequate.
 
-More plausibly, the rate of false negatives varies over the course of an episode.
+A better model would be to allow the rate of false negatives to vary over the course of an episode.
 False negatives are more likely to occur when the viral load of an individual is low, because there is less virus in their body to be sampled.
@@
 % In \cref{imperf-test:fig:bounding-cis-sensitivity} we consider the values produced by assuming the negative following the last positive in an episode could be either a true or false negative as a function of time since the individual was detected (\ie: the first positive test in the episode).
-This generates broad bounds, although suggest a declining test sensitivity over the course of an episode.
+This generates broad bounds, although suggest a declining test sensitivity over the course of an episode (see \cref{imperf-test:fig:bounding-cis-sensitivity}).
 Small numbers of tests mean that the series is noisy near the start and the end, such as both bounds being 0 after day 40.
@@
 
-To check sensitivity of the inference, it will be useful to simulate from a model with a varying test sensitivity as a function of days since the start of the infection episode.
+To assess the sensitivity of the procedure to a time-varying test sensitivity, it will be useful to simulate from a model with a varying test sensitivity as a function of days since the start of the infection episode.
 Through consideration of \cref{imperf-test:fig:bounding-cis-sensitivity}, and the interaction of false negatives due to viral load in \cref{E-ATACCC}'s results, I propose the following model.
@@
 
-First, in \cref{imperf-test:sec:modifying-p_ia}, I modify $p_{ia}$, the probability of an episode being in the admissible, to allow for the episode possibly being longer than observed.
-Then, in \cref{imperf-test:sec:modifying-p_iu}, I modify $p_{iu}$, the probability of an episode being undetected, to allow for additional episodes being missed.
+First, in \cref{imperf-test:sec:modifying-p_ia}, I modify $p_{ia}$, the probability episode $i$ being admissible, to allow for the episode possibly being longer than observed.
+Then, in \cref{imperf-test:sec:modifying-p_iu}, I modify $p_{iu}$, the probability episode $i$ being undetected, to allow for additional episodes being missed.
 
@@
 If it is a false negative, then I will consider the episode's end right-censored.
-However, if it is a true negative, then the episode's end is interval-censored as previously.
+However, if it is a true negative, then the episode's end is interval-censored, as previously.
 A mixture of these scenarios forms the episode's likelihood contribution, with the mixture probability determined by the test sensitivity.
@@
 Denote this subset as $\sched'_i = \{ t \in \sched_i \ssep r_i^{(b)} \leq t \leq l_i^{(e)} \}$, and their results by $\vec{y_i'} = \{ y_i(t) \ssep t \in \sched'_i \}$ where $y_i(t) = 1$ if the test on day $t$ is positive and 0 otherwise.
-As I assume that there are no false positives, the infection episode must span at least this period starting and ending with a positive test.
+As I assume that there are no false positives, the infection episode must span at least this period because it starts and ends with a positive test.
 Therefore, the test results at times in $\sched_i'$ are either true positives or false negatives.
@@
 In this case, the test at $r_i^{(e)}$ is a true negative, as are all other tests not in $\sched'_i$.
-These occur with probability 1, as I continue to assume no false positives.
+These occur with probability 1, by the assumption of no false positives.
 The tests in $\sched_i'$ are either true positives or false negatives.
@@
 
-False negatives add additional mechanisms by which episode $i$ could be undetected, in addition to those considered when deriving $p_{iu}$ in \cref{E-perf-test:eq:piu}.
+Several mechanisms for episodes being undetected were considered when deriving $p_{iu}$ in \cref{E-perf-test:eq:piu}.
+I now consider the additional mechanisms arising due to false negatives.
 Specifically, episode $i$ could be undetected if the first test after the $b_i$, episode $i$'s start day, is a false negative and then there are no subsequent positive tests.
@@
 
-Therefore, this mechanism causes episode $i$ to be undetected if all of the following conditions hold.
+Therefore, this mechanism causes episode $i$ to be undetected if all the following conditions hold.
 \begin{enumerate}
     \item The episode would have been detected considering only the mechanisms in \cref{E-perf-test:eq:piu}. That is $b_i > \min(\sched_i)$ and $\tau_{\sched_i}(b_i) + b_i \leq e_i$.
-    \item The episode ends in the interval $[\tau_{\sched_i}(b_i) + b_i, \tau^2_{\sched_i}(b_i) + b_i]$. Equivalently, $\tau_{\sched_i}(b_i) + 1 \leq d_i < \tau^2_{\sched_i}(b_i) + 1$.
+    \item The episode ends in the interval $[\tau_{\sched_i}(b_i) + b_i, \tau^2_{\sched_i}(b_i) + b_i]$.
+      Note that the lower bound here is exactly the bound on $e_i$ in the previous condition.
+      Equivalently, $\tau_{\sched_i}(b_i) + 1 \leq d_i \leq \tau^2_{\sched_i}(b_i) + 1$.
     \item A false negative occurs on day $\tau_{\sched_i}(b_i) + b_i$. Conditional on the previous condition, this occurs with probability $1 - \psens$.
@@
 &\prob \left(
-    \tau_{\sched_i}(b_i) + 1 \leq D_i < \tau^2_{\sched_i}(b_i) + 1
+    \tau_{\sched_i}(b_i) + 1 \leq D_i \leq \tau^2_{\sched_i}(b_i) + 1
     \mid b_i,
@@
 Denote the replacement for $p_{iu}$ as $p_{iu}'$.
-Since the new mechanism is mutually exclusive with the previous one, $p_{iu}' = p_{iu} + \eta$ is the sum of $p_{iu}$.
-For the posterior density (\cref{E-perf-test:eq:full-posterior}), $1 - p_{iu}'$ is the required quantity.
+Since the new mechanism is mutually exclusive with the previous one, $p_{iu}' = p_{iu} + \eta$.
+For the posterior density, \cref{E-perf-test:eq:full-posterior}, $1 - p_{iu}'$ is the required quantity.
 \begin{align}
@@
 Second, the impact of misspecifying $\psens$, including violating the assumption that is constant.
-Finally, the impact of the simplying approximations made in \cref{imperf-test:sec:modelling} for tractability.
-These approximations are assuming that the previous negative is a true negative and that there is a negligible probability of missing an episode due to two false negatives.
+Finally, the impact of the simplifying assumptions made in \cref{imperf-test:sec:modelling} for tractability.
+These assumptions are that the previous negative is a true negative and that there is a negligible probability of missing an episode due to two false negatives.
 
 I simulate datasets under four conditions: a constant $\psens$ of 0.6, 0.8, or 1.0 (the latter being the same as the simulated data in \cref{E-perf-test:sec:simulation-study}), or the varying test sensitivity model proposed at the end of \cref{imperf-test:sec:simulate}.
+The simulation is otherwise unchanged from that in \cref{E-perf-test:sec:simulation-study}.
 The number of single positive episodes in each of these conditions, compared to the number observed in the CIS data, is shown in \cref{imperf-test:fig:sim-single-pos}.
@@
 For each simulated dataset, I estimate the survival function using the model proposed in \cref{imperf-test:sec:modelling} and assuming $\psens$ equals 0.6, 0.8, or 1.0.
-If the dataset was simulated using a constant $\psens$, and that value of $\psens$ matches the value assumed during inference, I refer to $\psens$ as being correctly specified.
+If the dataset was simulated using a constant $\psens$, and that value of $\psens$ matches the value assumed during inference, I refer to $\psens$ as being correctly specified; otherwise I refer to it as misspecified.
 However, there is still some misspecification of the model to false negatives owing to the simplifying assumptions made.
@@
 
-I use either the independent (vague) and model combination priors for the survival (described in \cref{E-perf-test:sec:parameters-priors}).
-These were shown to be the best performing priors in \cref{E-perf-test:sec:results}).
-Therefore, there are a total of 24 possible scenarios (each combination of: 4 simulated datasets, 3 values of $\psens$ in inference, and 2 priors).
+I use either the independent (vague) or model combination priors for the survival (described in \cref{E-perf-test:sec:parameters-priors}).
+These were shown to be the best performing priors in \cref{E-perf-test:sec:results}.
+Therefore, there are a total of 24 possible scenarios (each combination of: 4 simulated datasets, 3 values of $\psens$ in inference, and 2 priors), although not all are shown.
 
 When $\psens = 0.8$ and is correctly specified, the model recovers the true survival time well (see \cref{imperf-test:fig:constant-test-sensitivity}(B)).
-The informative prior, in comparison to the vague prior, helps to overcome the misspecification due to the simplyifying assumptions, moving the estimated survival function closer to its true survival time.
+The informative prior, in comparison to the vague prior, helps to overcome the misspecification due to the simplifying assumptions, moving the estimated survival function closer to its true survival time.
 However, when $\psens = 0.6$, this is no longer the case (see \cref{imperf-test:fig:constant-test-sensitivity}(A)).
@@
 If the test sensitivity is misspecified, that is the assumed value for $p_\text{sens}$ in \cref{imperf-test:eq:pit-prime,imperf-test:eq:pit-prime} is different to the value used in the simulation, then the survival time will be biased (see \cref{imperf-test:fig:misspecified-test-sensitivity}).
-If $\psens$ is misspecified and too low (see \cref{imperf-test:fig:misspecified-test-sensitivity}(A)), then the posterior estimate initially follows the true value but then separates,
+If $\psens$ is misspecified and too low (see \cref{imperf-test:fig:misspecified-test-sensitivity}(A)), then the posterior estimate initially follows the true value but then separates.
 The number of episodes inferred to have truly ended by the first negative is too high, and hence the survival time is underestimated.
@@
     All simulations use a constant test sensitivity of 0.8, but the inference procedure assumes different values, as per the key.
-    Hence (B) is correctly specified, and the same as in \cref{imperf-test:fig:constant-test-sensitivity}, but (A) and (C) have misspecified $\psens$.
-    All results uses the combination prior, due to it performing better than the vague prior in \cref{imperf-test:fig:constant-test-sensitivity}.
+    Hence, (B) is correctly specified, and the same as in \cref{imperf-test:fig:constant-test-sensitivity}, but (A) and (C) have misspecified $\psens$.
+    All results use the combination prior, due to it performing better than the vague prior in \cref{imperf-test:fig:constant-test-sensitivity}.
   }
@@
 
-In this section I apply the model described in this chapter to the CIS episodes dataset I described in \cref{E-perf-test:sec:approach}.
-This is the 4,800 infection episodes that were first detected between 10th Oct 2020 and 6th Dec 2020 inclusive, with negatives bounding the start and end time of the episode.
+In this section I apply the approach described in this chapter to the CIS episodes dataset I described in \cref{E-perf-test:sec:approach}.
+This is the 4,800 CIS infection episodes first detected between 10th Oct 2020 and 6th Dec 2020 inclusive and have negatives bounding the start and end time of the episode.
 
@@
 This estimate has more long episodes than the estimate in \cref{E-ATACCC}.
-The qualitative increase in long episodes is robust to the choice of prior for $\ntot$, the assumed value for $\psens$ (see \cref{imperf-test:fig:cis-sensitivity}), and the choice of prior for the hazard (see \cref{perf-test:sec:parameters-priors}), $\lambda$.
+The qualitative increase in long episodes is robust to the choice of prior for $\ntot$, the assumed value for $\psens$ (see \cref{imperf-test:fig:cis-sensitivity}), and the choice of prior for the hazard (see \cref{E-perf-test:sec:parameters-priors}), $\lambda$.
 However, the survival proportion over the first 4 weeks is sensitive to these choices.
@@
   \caption{%
-    Sensitivity of CIS estimates to: (A-C) the value of $r$, the dispersion of the prior for $\ntot$, and (D-F) $\psens$.
+    Sensitivity of CIS estimates to: (A-C) the value of $r$, the dispersion of the prior for $\ntot$ (a larger $r$ means a more informative prior), and (D-F) $\psens$.
     A and D: median survival time, in comparison to the ATACCC-based estimate of \cref{E-ATACCC}.
@@
 
-% \begin{itemize}
-%   \item Survival prior matters for quantitative details but not qualitative shape. 
-%   \item Number of missed infections prior is more important, but not the case in simulation (not shown).
-%   \item Unclear why this is the case, probably model misspecification meaning that fewer missed infections than expected.
-%   \item Two possible misspecifications: test sensitivity (not constant) and constant incidence.
-% \end{itemize}
-
-This chapter shows that a simple model of false negatives can be incorporated into the model of \cref{E-perf-test}.
-This model allows estimation of the duration distribution tails without the strong model assumptions used in \cref{E-ATACCC}.
+In this chapter, I developed a novel method to incorporate false negatives into the survival analysis of \cref{E-perf-test}.
+I then applied this method to the CIS data, estimating the tail of the duration distribution without the strong model assumptions and extrapolation in \cref{E-ATACCC}.
 The qualitative features of the estimated distribution are robust to the assumed value for $\psens$, and the choice of prior for $\lambda$, although the quantitative details are not.
-The estimates of the tail, the primary purpose of this chapter, are in addition robust to the choice of prior for $\ntot$.
+The estimates of the tail, the primary purpose of this chapter, are in addition somewhat robust to the choice of prior for $\ntot$.
 However, the rest of the distribution is sensitive to this choice.
 
-An informative prior on $\ntot$ is required is for the bulk of the distribution to be plausible.
+An informative prior on $\ntot$ is required is for the bulk of the distribution to agree with those in \cref{E-ATACCC} and the wider literature.
 In the simulation study, the uninformative prior on $\ntot$ performed well, even when the test sensitivity was misspecified.
-One hypothesis is that the test sensitivity is significantly different to the function in \cref{imperf-test:eq:variable-test-sensitivity} causing an unintuitive pattern of results.
+A possible explanation is that the true variation in the test sensitivity is significantly different to the function in \cref{imperf-test:eq:variable-test-sensitivity}.
 In particular, the test sensitivity for long episodes may become much lower than 50\%.
@@
 
-This chapter has produced estimates of the duration distribution of PCR-positivity using information from two studies: ATACCC and CIS.
+In this chapter, I developed and applied novel methodology to estimate the duration distribution of PCR-positivity using information from two studies: ATACCC and CIS.
 Assuming a test sensitivity of 0.8 and an informative prior on the number of infections, provided the preferred estimates.
@@
 This feature of the distribution is important for understanding the proportion of prevalence attributable to long infections and interpreting PCR test results.
+\todo{possibly cut some of this text?}
 Producing these estimates required applying previous survival analysis frameworks in the novel context of the CIS.
@@
 
-The posterior distribution for the survival is concentrated, producing narrow credible intervals.
+The posterior distribution for the survival is concentrated, producing narrow credible intervals, likely due to the large sample size in the CIS.
 In the following chapters, I will take the posterior mean for the survival time at each point, neglecting this uncertainty, as the basis for estimating transmission.
-First, I will take a backcalculation approach (\cref{E-backcalc}), based on the framework in \cref{E-inc-prev}.
-Then, I will take a mechanistic approach (\cref{E-SEIR}), modelling the transmission process explicitly.
+% First, I will take a backcalculation approach (\cref{E-backcalc}), based on the framework in \cref{E-inc-prev}.
+% Then, I will take a mechanistic approach (\cref{E-SEIR}), modelling the transmission process explicitly.
 

### 2024-01-17 cis-imperfect-testing.tex
deal with (almost all) comments or turn into todos

--- 2024-01-16 cis-imperfect-testing.tex
@@
 I modified the simulation study from \cref{E-perf-test:sec:simulation-study} to include the possibility of false negatives.
-More single positive episodes are seen at lower values of $\psens$, with a value between 60\% and 80\% reproducing the number of single positive episodes observed in the CIS data (see \cref{imperf-test:fig:sim-single-pos}).
-In \cref{E-ATACCC}, I estimated that $\psens$ ($1-\rho$ in \cref{E-ATACCC}'s parameterisation) is 95\% (95\% CrI: 96\%--93\%).
-The daily testing in the ATACCC study, used for the \cref{E-ATACCC} estimates, means this is a reliable estimate during the period for which the individuals are followed-up.
-The requirement for a much lower value of $\psens$ in the simulation study suggests this model of false negatives is inadequate.
+More single positive episodes are seen at lower values of $\psens$ (see \cref{imperf-test:fig:sim-single-pos}).
+A value between 60\% and 80\% for $\psens$ reproduces the number of single positive episodes observed within the CIS data.
+In \cref{E-ATACCC}, I estimated that the rate of false negatives is\todo{add what the ATACCC false negative rate is}.
+The daily testing in the ATACCC study, used for the \cref{E-ATACCC} estimates, should provide a reliable estimate of the false negative rate during the period for which the individuals are followed-up\todo{rephrase this sentence, currently unclear}.
+The requirement for a much lower value of $\psens$ suggests this model of false negatives is inadequate.
 
@@
   }
+  \todo[inline]{Label 0.6 sub-plot as (A) and 0.8 as (B). Include psens in the facet labelling.}
   \label{imperf-test:fig:constant-test-sensitivity}
@@
   }
+   \todo[inline]{Label sub-plots (A--C). Include psens in the facet labelling.}
   \label{imperf-test:fig:misspecified-test-sensitivity}

### 2024-01-23 transmission.tex
section/chapter consistent usage

@@
 }
-  \caption[SEIR model]{SEIR model used for the analysis in this chapter. The shaded box represents PCR-positive individuals. Shown for a single stratum.}
+  \caption[SEIR model]{SEIR model used for the analysis in this section. The shaded box represents PCR-positive individuals. Shown for a single stratum.}
   \label{SEIR:fig:full-model}
@@
 
-The results in this chapter show the feasibility of fitting a mechanistic model using only data on prevalence.
+The results in this section show the feasibility of fitting a mechanistic model using only data on prevalence.
 If the distributions for length-of-stay in each compartment are known, then all other model parameters can  be estimated.

### 2024-01-23 transmission.tex
backcalc plots

@@
 \begin{figure}
-    \todo[inline]{Four-panel regions figure}
-    \capton[Incidence estimated using backcalculation by region]{%
+    \includegraphics{transmission/backcalc-regions}
+    \caption[Incidence estimated using backcalculation by region]{%
         Incidence proportion for all of England (A) and its component regions.
@@
 \begin{figure}
-    \todo[inline]{age groups figure}
-    \capton[Incidence estimated using backcalculation by age group]{%
+    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{transmission/backcalc-ages}}
+    \caption[Incidence estimated using backcalculation by age group]{%
         Incidence proportion for each age group, on both natural and log scales.
@@
 \begin{figure}
-    \todo[inline]{start effect figure}
-    \capton[Edge effects in backcalculation method]{%
+    \centering \includegraphics{transmission/backcalc-start-effect}
+    \caption[Edge effects in backcalculation method]{%
         First seven days of incidence proportion estimate for the whole of England, highlighting the edge effect.
     }
-    \label{transmission:fig:start-effect}
+    \label{transmission:fig:backcalc-start-effect}
 \end{figure}

### 2024-01-23 transmission/backcalc-ages.R

@@ -1,0 +1,31 @@
+suppressMessages(library(dplyr))
+library(ggplot2)
+library(patchwork)
+library(tidybayes)
+library(tidyr)
+source(here::here("utils.R"))
+
+output_dir = here::here("transmission/outputs")
+p_age_base = readRDS(file.path(output_dir, "age_incidence.rds")) |>
+    filter(daynr > 1) |>
+    ggplot(aes(date, val, colour = age_group, fill = age_group)) +
+    stat_lineribbon(alpha = 0.3, .width = 0.95) +
+    labs(
+        x = "Date",
+        y = "Incidence proportion",
+        fill = "Age group",
+        colour = "Age group"
+    ) +
+    standard_plot_theming()
+p_age = (p_age_base + scale_y_continuous(labels = scales::label_percent())) +
+        (p_age_base + scale_y_log10(labels = scales::label_percent()) + theme(axis.title.y = element_blank())) +
+        # collect legends and move to bottom of plot
+        plot_layout(guides = "collect") &
+        theme(legend.position = "bottom")
+
+ggsave(
+    filename = here::here("transmission", "backcalc-ages.pdf"),
+    plot = p_age,
+    width = 6.5,
+    height = 3.6
+)

### 2024-01-23 transmission/backcalc-regions.R

@@ -1,0 +1,46 @@
+suppressMessages(library(dplyr))
+library(ggplot2)
+library(tidybayes)
+library(tidyr)
+source(here::here("utils.R"))
+
+output_dir = here::here("transmission/outputs")
+label_letters = as_labeller(function(labels, base_labeller = label_value) {
+    glue::glue("({LETTERS[seq_along(labels)]}) {base_labeller(labels)}")
+})
+
+p_region = readRDS(file.path(output_dir, "region_incidence.rds")) |>
+    filter(daynr > 1) |>
+    mutate(
+        panel = case_match(
+            region,
+            "England" ~ "England",
+            "North East" ~ "North",
+            "North West" ~ "North",
+            "Yorkshire" ~ "North",
+            "East Midlands" ~ "Midlands",
+            "West Midlands" ~ "Midlands",
+            "East of England" ~ "Midlands",
+            "London" ~ "South",
+            "South East" ~ "South",
+            "South West" ~ "South",
+        ),
+    ) |>
+    ggplot(aes(date, val, colour = region, fill = region)) +
+    stat_lineribbon(alpha = 0.3, .width = 0.95) +
+    facet_wrap(~panel, labeller = label_letters) +
+    scale_y_continuous(labels = scales::label_percent()) +
+    labs(
+        x = "Date",
+        y = "Incidence proportion",
+        fill = "Region",
+        colour = "Region"
+    ) +
+    standard_plot_theming() +
+    theme(legend.position = "bottom")
+ggsave(
+    filename = here::here("transmission", "backcalc-regions.pdf"),
+    plot = p_region,
+    width = 6,
+    height = 5.5
+)

### 2024-01-23 transmission/backcalc-start-effect.R

@@ -1,0 +1,25 @@
+suppressMessages(library(dplyr))
+library(ggplot2)
+library(tidybayes)
+library(tidyr)
+source(here::here("utils.R"))
+
+output_dir = here::here("transmission/outputs")
+p_start_effect = readRDS(file.path(output_dir, "region_incidence.rds")) |>
+    filter(daynr <= 20, region == "England") |>
+    ggplot(aes(date, val)) +
+    stat_lineribbon(alpha = 0.8, .width = 0.95) +
+    scale_y_continuous(labels = scales::label_percent()) +
+    scale_fill_brewer() +
+    labs(
+        x = "Date",
+        y = "Incidence proportion",
+    ) +
+    standard_plot_theming() +
+    theme(legend.position = "none")
+ggsave(
+    filename = here::here("transmission", "backcalc-start-effect.pdf"),
+    plot = p_start_effect,
+    width = 4,
+    height = 3.8
+)

### 2024-01-23 transmission/incidence.R

@@ -1,113 +1,0 @@
-suppressMessages(library(dplyr))
-library(ggplot2)
-library(tidybayes)
-library(tidyr)
-
-base_data_dir = here::here("transmission/outputs")
-load_prevalence = function() {
-    readRDS(file.path(base_data_dir, "predict.rds")) |>
-    left_join(
-        readr::read_csv(
-            file.path(base_data_dir, "groups.csv"),
-            show_col_types = FALSE
-        ),
-        by = ".group"
-    ) |>
-    mutate(prev_predict = expit(.predict))
-}
-
-load_incidence = function() {
-    readRDS(file.path(base_data_dir, "deconv.rds"))
-}
-
-group_by_strata = function(x, ...) {
-  all_cols = rlang::exprs(region, daynr, sex, age_group, ethnicityg)
-  exclude = rlang::ensyms(...)
-  stopifnot(all(exclude %in% all_cols))
-  include = setdiff(all_cols, exclude)
-  return(group_by(x, !!!include))
-}
-
-load_poststrat_table = function() {
-    readr::read_csv(file.path(base_data_dir, "poststrat.csv"), show_col_types = FALSE) |>
-        select(!.groups) |>
-        mutate(
-            region = case_match(
-                Region_Name,
-                "North_East_England"  ~ "1_NE",
-                "North_West_England"  ~ "2_NW",
-                "Yorkshire"  ~ "3_YH",
-                "East_Midlands"  ~ "4_EM",
-                "West_Midlands"  ~ "5_WM",
-                "East_England"  ~ "6_EE",
-                "London"  ~ "7_LD",
-                "South_East_England"  ~ "8_SE",
-                "South_West_England"  ~ "9_SW",
-                .default = NA_character_
-            )
-        ) |>
-        assertr::verify(
-            !is.na(region)
-            | Region_Name %in% c("Wales", "Scotland", "Northern_Ireland")
-        ) |>
-        filter(!is.na(region))
-}
-
-poststratify = function(data, postrat_table, col, ...) {
-    left_join(
-        data,
-        postrat_table,
-        by = c("region", "age_group", "ethnicityg" = "ethnicity", "sex")
-    ) |>
-        assertr::assert(assertr::not_na, pop) |>
-        mutate(n = {{ col }} * pop) |>
-        group_by(daynr, .draw, !!!ensyms(...)) |>
-        summarise(
-            N = sum(pop),
-            val = sum(n) / N,
[…]

### 2024-01-25 biology-data.tex
easy edits
Figure of SARS-CoV-2 natural history

@@
 
-\section{Natural history of the disease}
+\section{Natural history of the disease} \label{biology-data:sec:natural-history}
+
+\begin{figure}
+  \centering\includegraphics[width=\textwidth]{biology-data/natural-history}
+  \caption[Natural history of SARS-CoV-2.]{Natural history of SARS-CoV-2. Reproduced from \textcite{puhachSARSCoV2} with permission.}
+  \label{biology-data:fig:natural-history}
+\end{figure}
 
@@
 The CIS (Coronavirus Infection Survey) is a longitudinal study on a representative sample of households.
+Households are invited to the study from databases held by the ONS.
+The number invited was to satisfy various target number of individuals swabbing per fortnight.
+Within a selected household, all individuals aged 2 and over were invited to participate.
+Once invited, an enrolment swab would be taken followed by 4 further weekly swabs (giving a total of 5 swabs on days 0, 7, 14, 21, 28 relative to enrolment) after which monthly swabs are taken.
 A full description of the study can be found in \textcite[][supplementary materials]{pouwelsCommunity} or the study protocol~\autocite{cisProtocol}.
-In brief, households are invited to the study from databases held by the ONS.
-The number invited was to satisfy various target number of individuals swabbing per fortnight; relevant to the discussion in this document is that the target increased from 2,500 by the end of May 2020 to 150,000 in October 2020.
-Within a household, all individuals aged 2 and over were invited to participate.
-Once invited, an enrolment swab would be taken followed by 4 further weekly swabs (giving a total of 5 swabs on days 0, 7, 14, 21, 28 relative to enrolment) after which monthly swabs are taken.
 
@@
 Following this, the number then decreases and stabilises as those recruited to meet the October target transition from weekly onto monthly testing.
-See figure \@ref(fig:test-schedules) for further detail.
 

### 2024-01-25 transmission.tex
Compare model outputs

@@
 
+\begin{figure}
+    \centering \includegraphics{transmission/compare-regions}
+    \caption[Comparing each models estimate of regional incidence.]{%
+        Incidence proportion in England and each region estimated by the two models discussed in this chapter.
+        Posterior median and 95\% CrI shown.
+        For the SEIR model, newly PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the backcalculation results.
+    }
+    \label{transmission:fig:compare-regions}
+\end{figure}
+
+\begin{figure}
+    \centering \includegraphics{transmission/compare-NE}
+    \caption[Comparing each models estimate of North East incidence by age.]{%
+        Incidence proportion in each age group in the North East estimated by the two models discussed in this chapter.
+        Posterior median and 95\% CrI shown.
+        For the SEIR model, newly PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the backcalculation results.
+    }
+    \label{transmission:fig:compare-NE}
+\end{figure}
 \section{Conclusion}

### 2024-01-25 transmission/compare_NE.R

@@ -1,0 +1,54 @@
+suppressMessages(library(dplyr))
+library(lubridate)
+library(ggplot2)
+library(patchwork)
+library(purrr)
+library(stringr)
+library(tidybayes)
+library(tidyr)
+source(here::here("utils.R"))
+source(here::here("SEIR/utils.R"))
+source(here::here("transmission/utils.R"))
+
+
+backcalc_prevalence_summary = readRDS(here::here("transmission", "outputs", "region_age.rds")) |>
+    filter(region == "North East") |>
+    group_by(date, age_group) |>
+    median_qi(incidence, prevalence, .simple_names = FALSE) |>
+    ungroup() |>
+    mutate(Model = "Backcalculation")
+
+seir_incidence_summary = load_seir_predictive() |>
+    filter(region == "North East") |>
+    rename(age_group = age) |>
+    poststratify_SEIR(load_poststrat_table(), new_pcr_pos, age_group) |>
+    rename(incidence = val) |>
+    group_by(day, age_group) |>
+    median_qi(incidence, .simple_names = FALSE) |>
+    ungroup() |>
+    mutate(Model = "SEIR")
+
+p_compare = bind_rows(
+    seir_incidence_summary |>
+        rename(
+            date = day,
+        ),
+    backcalc_prevalence_summary,
+) |>
+    mutate(age_group = normalise_age_groups(age_group)) |>
+    ggplot(aes(date, incidence, ymin = incidence.lower, ymax = incidence.upper, colour = Model, fill = Model)) +
+    geom_lineribbon(alpha = 0.4, linewidth = 0.2) +
+    facet_wrap(~age_group) +
+    standard_plot_theming() +
+    theme(legend.position = "bottom") +
+    scale_y_continuous(labels = scales::label_percent()) +
+    labs(
+        x = "Date",
+        y = "Incidence proportion"
+    )
+ggsave(
+    filename = here::here("transmission", "compare-NE.pdf"),
+    plot = p_compare,
+    width = 6,
+    height = 4
+)

### 2024-01-25 transmission/compare_regions.R

@@ -1,0 +1,53 @@
+suppressMessages(library(dplyr))
+library(lubridate)
+library(ggplot2)
+library(patchwork)
+library(purrr)
+library(stringr)
+library(tidybayes)
+library(tidyr)
+source(here::here("utils.R"))
+source(here::here("SEIR/utils.R"))
+source(here::here("transmission/utils.R"))
+
+seir_predictive = load_seir_predictive()
+seir_incidence_summary = seir_predictive |>
+    rename(age_group = age) |>
+    poststratify_SEIR(load_poststrat_table(), new_pcr_pos, region) |>
+    rename(incidence = val) |>
+    group_by(day, region) |>
+    median_qi(incidence, .simple_names = FALSE) |>
+    ungroup() |>
+    mutate(Model = "SEIR")
+
+backcalc_prevalence_summary = readRDS(here::here("transmission", "outputs", "region.rds")) |>
+    group_by(date, region) |>
+    median_qi(incidence, prevalence, .simple_names = FALSE) |>
+    ungroup() |>
+    mutate(Model = "Backcalculation")
+
+p_compare = bind_rows(
+    seir_incidence_summary |>
+        rename(
+            date = day,
+        ),
+    backcalc_prevalence_summary,
+) |>
+    mutate(region = normalise_region_names(region)) |>
+    filter(region != "England") |>
+    ggplot(aes(date, incidence, ymin = incidence.lower, ymax = incidence.upper, colour = Model, fill = Model)) +
+    geom_lineribbon(alpha = 0.4, linewidth = 0.2) +
+    facet_wrap(~region) +
+    standard_plot_theming() +
+    theme(legend.position = "bottom") +
+    scale_y_continuous(labels = scales::label_percent()) +
+    labs(
+        x = "Date",
+        y = "Incidence proportion"
+    )
+ggsave(
+    filename = here::here("transmission", "compare-regions.pdf"),
+    plot = p_compare,
+    width = 6,
+    height = 5.5
+)

### 2024-01-25 transmission.tex
Start notes on comparison

@@
 
+Comparing the regional incidence estimates between the two models in this chapter.
+
+Example of bias---variance trade-off.
+
+Demonstrated by fitting to the North East.
+Quality of fit in SEIR model is variable between different age groups.
+This is because the model is not flexible enough to capture the true age distribution of infections.
+
 \begin{itemize}

### 2024-01-25 transmission.tex
Edits

@@
 
-\todo[inline]{What happened to the following paragraph? It doesn't make sense currently!}
 In this setup, the parameters $i^+$, $q_c$, and $\psi$ are fully recovered; the other parameters are only partially so: their posterior has moved away from the prior but is still regularised by it (see \cref{SEIR:fig:true-vs-posterior}).
 The ability to identify $q_c$ is an advantage of using data which includes measurements on children, unlike data on hospitalisations or deaths.
+
+For $\vec{\pi}$ and $\rho$, the posterior median is very similar regardless of the true value (see \cref{SEIR:fig:true-vs-posterior}).
 This suggests that these parameters are not practically identifiable in this setting.
@@
 A constant $\matr{\beta}(t)$ would lead to exponential growth following a period of transitory dynamics, although the weekly switching may mean that these transitory dynamics never end, especially when incidence is high~\autocite{rhodesConvergence}.
-When the peak in incidence is estimated to occur varied by region, but due to the switching, is almost always on a Sunday (see \cref{SEIR:fig:peak-incidence}).
-For Yorkshire, the peak is estimated to occur in late October.
-For all other regions, it is estimated to occur in late December or Early January, the period during which severe social distancing measures were applied.
+When the peak in incidence is estimated to occur varied by region, with the most common time being early January 2021 (see \cref{SEIR:fig:peak-incidence}).
 \begin{figure}
@@
 The \emph{attack rate} is the proportion of the population that have been infected at least once\todo{cite attack rate}.
-In this model it is $1 - \vec{s}$.
-The attack rate is highest in the $[11, 16)$ age group in the North West at 50\% (95\% CrI: 48\%--52\%) (see \cref{SEIR:fig:attack-rates}).
+In this model it is $1 - \vec{s}$, the proportion susceptible.
+The attack rate is highest in the $[11, 16)$ age group in the North West at 68\% (95\% CrI: 66\%--70\%) (see \cref{SEIR:fig:attack-rates}).
 \begin{figure}
@@
 Overall the model fits the data well (see \cref{SEIR:fig:prev-young,SEIR:fig:prev-old}).
-There are a few features of the data which are not captured by the model.
-In particular, prevalence during October is higher in the 16--25s than the model predicts and lower in the 11--16s.
+There are a few features of the data which are not captured by the model and the posterior estimate appears to be a compromise with the prevalence too low in one age group and too high in another.
+For example, in the North East region, the 11--16 age group has its prevalence overestimated in October but the 16--25 age group has its prevalence underestimated.
 \begin{figure}
@@
 
-One MCMC chain for the North West region appeared to get stuck, with very few moves after the first 100,000 iterations (not shown).
-I reran this chain with a new seed, with the rerun results presented.
-Following this, all regions' MCMC chains converged, assessed using the Rhat statistic and ESS.
-Rhat for all parameters is < 1.01, except $\sigma_\epsilon$ in London which had a Rhat of 1.018.
-All parameters had ESS > 1000, except $\sigma_\epsilon$ in London and the East, which had ESS of 197 and 757 respectively.
-As all the parameters of primary interest had high Rhat and ESS, and no parameters had large issues, the inference procedure is considered to have converged.
+All regions' MCMC chains converged, assessed using the Rhat statistic and ESS (as defined in \cref{E-inc-prev:sec:MCMC}).
+The lowest Rhat was 1.01 for $\sigma_\epsilon$ in London.
+All parameters had ESS > 1000, except $\sigma_\epsilon$ and $\psi$ in London, which had ESS of 450 and 950 respectively.
 

### 2024-01-25 transmission.tex
almost all chapter text written

@@
 The prior on $\rho$ favours a null model of a binomial likelihood.
-The remaining priors had weakly informative priors.
+The remaining parameters had weakly informative priors.
 
@@
 This suggests that these parameters are not practically identifiable in this setting.
-The posteriors align very closely with the priors (not shown\todo{consider showing prior/posterior plots}) which reinforces this finding.
+The posteriors align very closely with the priors (not shown) which reinforces this finding.
 The lack of identifiability for $\vec{\pi}$ aligns with previous work showing this parameter is structurally unidentifiable, \ie not identifiable even in the limit of infinite data~\autocite{dankwaStructural}.
@@
 While this analysis cannot establish causation, several hypotheses can be identified that are worth further investigation.
-The decreases in $\beta_w$ over the autumn could be due to increased caution(\eg masking or stricter adherence to social distancing) and/or interventions short of a lockdown as cases and deaths increased~\autocite{jarvisEffect}, or the impact of local measures in the highest transmission areas~\autocite{scottCovid19}.
-Local measures are not well captured by $\matr{\kappa}$ because the matrix was for the whole of England.
-Allowing each region to have its own $\matr{\kappa}$ may capture the effect of local measures more accurately.
-The increases in $\beta_w$ in December coincide with the rise of the Alpha variant~\autocite{lythgoeLineage}, known to be more transmissible~\autocite{daviesEstimated}.
+
+The decreases in $\beta_w$ over the autumn could be due to increased caution (\eg masking or stricter adherence to social distancing) as cases and deaths increased~\autocite{jarvisEffect} and/or interventions not captured by $\matr{\kappa}$ such as the impact of local measures in the highest transmission areas~\autocite{scottCovid19}.
+Local measures are not well captured by $\matr{\kappa}$ because the matrix averages behaviour across the whole of England.
+Allowing each region to have its own $\matr{\kappa}$ may capture the effect of local measures more accurately, although their targeted nature may require more granular methods.
+
+The increases in $\beta_w$ in December coincide with the rise of the Alpha variant~\autocite{lythgoeLineage}, known to be more transmissible~\autocite[e.g.][]{daviesEstimated}.
 A multiple variant model might be able to capture this rise more accurately.
@@
 This implies that the data would allow identification of a more flexible age-based model.
-Currently, the age distribution of infections is determined by $\matr{\beta}$, and to a lesser extent $\vec{\pi}$, the former of which only contains one parameter that affects the age distribution, $\beta_c$.
-More parameters could capture the age distribution more accurately.
-There are various ways to increase the model's flexibility, including introducing additional susceptibility or infectiousness parameters, or by allowing these parameters to vary over time.
-It is plausible that these distributions changed over time in a way not captured by $\matr{\kappa}$, for example due to the introduction and relaxation of control measures within schools.
+
+Currently, the age distribution of infections is determined by $\matr{\beta}$, and to a lesser extent $\vec{\pi}$, the former of which only contains one parameter that affects the age distribution, $q_c$.
+Increasing the number of parameters in $\matr{\beta}$ would be the simplest approach to increase the model's flexibility.
+
+The only time-changing part of $\matr{\beta}$ that varies between ages is $\matr{\kappa}$.
+It is plausible that this would not capture all behavioural changes.
+For example, the introduction and relaxation of control measures within schools.
 $\matr{\kappa}$ only captures changes in in-school transmission through changes in school attendance, not incorporating the probability of transmission, which could be modelled as changes in $q_{aa}$ for $a$ of school age.
@@
 
-The weekly patterns in incidence are another limitation of the current parameterisation.
+The weekly patterns in incidence are a limitation of the current parameterisation.
 The random walk and contact matrices both vary only on the weekly timescale.
@@
 However, this would greatly increase the number of parameters.
-An alternative would be a smoothly-varying function or an approximation of a random walk such as that proposed by \textcite{ghoshApproximate}.
+An alternative would be a smoothly-varying function or an approximation of a random walk, such as that proposed by \textcite{ghoshApproximate}.
 
@@
 
-Comparing the regional incidence estimates between the two models in this chapter.
-
-Example of bias---variance trade-off.
-
-Demonstrated by fitting to the North East.
-Quality of fit in SEIR model is variable between different age groups.
-This is because the model is not flexible enough to capture the true age distribution of infections.
-
-\begin{itemize}
-    \item Results here disagree with previous chapter's results.
-    \item The model in previous chapter is more flexible than here, showing the limitations discussed above. Previous chapter fits better.
-    \item The previous chapter's modelling benefits from being able to use flexible approximate inference methods that scale to many parameters, including random effects.
-Here, I am limited by the inference procedure.
-MCMC was unable to scale to the full model used in the previous chapter, which has a computationally cheaper likelihood as it is a GLM, not requiring solving an ODE.
-    \item Shows that there is a lot of information in the data which could be exploited by a suitably flexible mechanistic model.
-Such a model would be useful for policy-decisions because it could infer where transmission is occurring (\eg in schools or workplaces) and how this changes over time.
-    \item The narrower credible intervals in this chapter are a realisation of the standard bias-variance tradeoff.
-By making stronger assumptions regarding how transmission occurs, the variance is reduced.
-However, if these assumptions are incorrect (which is almost certainly the case), bias occurs.
-\end{itemize}
-
+\Cref{transmission:fig:compare-regions} compares the regional incidence estimates between the two models in this chapter.
+Generally the estimates agree, with the SEIR model having much narrower credible intervals.
+The largest discrepancy is that the backcalculation model estimates a second peak in mid January in several regions (most clearly in London, North West, and the East of England).
+However, even here, the credible intervals overlap except in London.
+This is a specific case of a larger pattern: the SEIR model produces smoother estimates of incidence than the backcalculation model.
 \begin{figure}
@@
 
+These patterns can be interpreted as an instance of the common bias---variance trade-off.
+The SEIR model imposes stronger assumptions about the epidemic dynamics.
+The assumptions reduces the variance of the estimates, including making them smoother, but if the assumptions are incorrect, bias is introduced.
+This trade-off based on the strength of assumptions is well-known to apply in the backcalulcation context~\autocite[e.g.][section 8.3]{brookmeyerBackcalculation}.
+The SEIR model can be viewed as a backcalulation method where the transmisson part of the model is a smoothing method.
+
+The backcalculation method benefits from being able to be implemented in a fast, approximate inference package (R-INLA) that easily scales to many parameters, including random effects.
+Therefore, it is far more flexible and can easily include time-varying effects, ethnicity, etc.
+The SEIR model is limited by the inference procedure and the requirement to sovle an ODE to evaluate the likelihood.
+To incorporate a more flexible model, and reduce its bias, a more efficient inference procedure would be required.
+
+Many of these issues are clearly demonstrated by the estimates for the North East (see \cref{transmission:fig:compare-NE}).
+I previously observed the quality of the SEIR model's fit is variable between different age groups (see \cref{SEIR:fig:prev-young,SEIR:fig:prev-old}).
+This is because the model is not flexible enough to capture the true age distribution of infections.
+The overestimation of prevalence in the 11--16 age group and underestimation in the 16--25 age group is reflected when comparing incidence estimates to those from the backcalculation model, which have reduced bias.
 \begin{figure}
@@
 \end{figure}
+
 \section{Conclusion}

### 2024-01-26 transmission.tex
Editing

@@
 \chapter{Estimating transmission} \label{transmission}
+
+\todo[inline]{Rewrite the intro: eg see Dani's Overleaf comments}
 
@@
 \section{Data} \label{transmission:sec:data}
+
+\todo[inline]{Consider moving all of this section into chapter 2, where CIS is introduced}
 
@@
 
-As discussed in\todo{ref section}, a major issue with the CIS is non-response bias.
-However, a deterministic relation between true prevalence and incidence is a good approximation (see \cref{E-inc-prev:sec:prevalence-process}).
-Therefore, I extend the methodology previously used to estimate prevalence from the CIS data, corrected for non-response bias, to estimate incidence\todo{cite Koen's stuff}.
-The previous method was multilevel regression and poststratification (MRP)\todo{cite MRP}.
-Here, a Bayesian multilevel logistic regression model is first fit to predict PCR positivity from demographic covariates.
+A deterministic relationship between true prevalence an incidence is a good approximation (see \cref{E-inc-prev:sec:prevalence-process}).
+A well-established methodology, using multilevel regression and poststratification (MRP)\todo{cite MRP}, has already been developed for estimating prevalence from the CIS data\todo{cite Koen's stuff}.
+Therefore, I extend the methodology previously used to estimate prevalence to estimate incidence.
+
+In MRP, a Bayesian multilevel logistic regression model is first fit to predict PCR positivity from demographic covariates.
 Then, the posterior distribution is used to predict the prevalence in each stratum.
 Finally, the prevalence is poststratified to the CIS target population and subpopulations of interest.
-I follow this same process, except I transform the posterior distribution of prevalence in each stratum to a posterior distribution of incidence before poststratifying.
+I modify this process by transforming the posterior distribution of prevalence in each stratum to a posterior distribution of incidence before poststratifying.
 % I fit a modification model to estimate prevalence, then transform the posterior distribution of prevalence in each stratum to a posterior distribution of incidence.
@@
 
-The advantage of fitting a multilevel regression model on prevalence, rather than incidence, is that prevalence with appropriate prior (or penalisation) can be expressed as a generalised linear model (GLM).
-A simple formulation\todo{I don't like the wording "simple formulation", maybe "default"?} for either incidence proportion or prevalence in a stratum $i$ is $\logit(\pi_i) = g(i, t)$, where $\pi_i$ is the incidence proportion or prevalence in stratum $i$ and $g$ is a function of the stratum and time, penalised towards being linear in time.
+\todo[inline]{Think about the framing of this paragraph, maybe talk about being linear in its parameters and pre-existing software packages.}
+The advantage of fitting a multilevel regression model on prevalence, rather than incidence, is that prevalence with appropriate prior (or penalisation) can be expressed as a generalised linear model (GLM)\todo{maybe cite backcalculation as a GLM}.
+Either incidence proportion or prevalence, $pi_i$ in a stratum $i$ can be expressed as $\logit(\pi_i) = g(i, t)$, where is a function of the stratum and time, penalised towards being linear in time.
 If $g$ is chosen appropriately, this is a GLM with a logit link function.
@@
 First, it ensures that the incidence proportion or prevalence is between 0 and 1.
-Second, at low values of incidence or prevalence, linear growth is approximately exponential~\autocite[see discussion in][and references therein]{ealesAppropriately}.
+Second, at low values of incidence or prevalence, linear growth of $\logit(\pi_i)$ implies approximately exponential in $\pi_i$~\autocite[see discussion in][and references therein]{ealesAppropriately}.
 Approximately exponential growth is what mechanistic models predict, and penalising towards exponential growth has previously been shown to improve estimates of prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
@@
 Previously, when fitting to prevalence over long periods and including all appropriate covariates to correct for non-response bias, the model was too computationally expensive to fit using MCMC~\citePersonalComms{Koen Pouwels}.
-Instead, an implementation of the model using the Integrated Nested Laplace Approximation (INLA) algorithm was used.
+Instead, an implementation of the model using the Integrated Nested Laplace Approximation (INLA) algorithm was used\todo{cite INLA}.
 
@@
     \item $\beta_s$ is a fixed effect for sex, with $\beta_s = 0$ for males and estimated for females.
-$\beta_{w(e)}$ is a fixed effect for white or non-white ethnicity.
+    \item $\beta_{w(e)}$ is a fixed effect for white or non-white ethnicity.
 $\beta_{w(e)} = 0$ if $e$ corresponds to white ethnicity and $\beta_{w(e)} = \beta_{n}$ otherwise (\ie non-white ethnic groups).% $\beta_n$ and $\beta_f$ have the default INLA prior for fixed effects\todo{check what this is}.
@@
 \end{itemize}
-All parameters have priors using the default specified in INLA, except where stated above.
+All parameters have the default R-INLA priors, except where stated above.
 For these parameters, the large amount of data means that the prior has little influence on the posterior distribution.
 
-The model was fit separately and independently for each region, implicitly adding interactions between region and all other variables.
+The model was fit separately and independently for each region, implicitly adding all interactions between region and the other variables.
 This step assumes that the estimates for each region can be considered independent, which is reasonable because the regions are large with many tests conducted within each.
@@
 For each stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
-I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:prevalence-process} justifies.
-Therefore, \cref{E-inc-prev:eq:EPt} implies that $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix of the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
-The matrix $\matr{S}$ is defined to maintain the relationship between incidence and prevalence in \cref{E-inc-prev:eq:EPt}.
+I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:prevalence-process} justifies, that is $P_t = \E(P_t \mid Z)$.
+\Cref{E-inc-prev:eq:EPt}, considered across different values of $t$, can be written as $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix formed from the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
+The matrix $\matr{S}$ is defined to maintain the relationship between incidence and prevalence.
 This gives a lower-triangular of the following form:
@@
 Model misspecification means that the procedure described could produce negative incidence.
-For any stratum and day $t$, this occurs if the difference between $P_t$ and $P_{t+1}$ is too large based on the number of recoveries that occur on day $t$ based on the incidence up to day $t$ and the duration distribution.
+For any stratum and day $t$, this occurs if the if the prevalence reduces between day $t$ and $t+1$ more than the number of recoveries on day $t$, based on the incidence up until day $t$ and the survival function.
 
@@
 The first is the survival matrix, $\matr{S}$.
-This is assumed known but \cref{E-imperf-test} showed that it is sensitive to assumptions made.
-The second is the model used to estimate prevalence.
+For example, \cref{E-imperf-test} showed that it is sensitive to assumptions made.
+The second is the model used to estimate prevalence, for example the form of the smoothing used.
 
-A prior constraining incidence to be non-negative and then fit directly to the data could be used but, as previously discussed, this would no longer be a GLM and could not be fit within the R-INLA framework.
-Hence, an alternative inference methodology would be required.
-MCMC is likely too computationally expensive.
-Other approaches, such as a Laplace approximation or variational inference, could be considered.
+A prior constraining incidence to be non-negative and then fit directly to the data could be used, but this could no longer be fit within the R-INLA framework.
+Hence, an alternative inference methodology, or simpler model (\eg containing fewer random effects), would be required.
+% MCMC is likely too computationally expensive.
+% Other approaches, such as a Laplace approximation or variational inference, could be considered.
 For this analysis, I replaced any value of incidence estimated to be negative with 0.
@@
 The final step is poststratifying the incidence proportion to the CIS target population and subpopulations of interest.
+I draw 500\todo{check number of samples} samples from the approximate posterior distribution estimated.
 For each posterior draw and day, the incidence proportion in the (sub)population is the weighted average of the incidence proportion in each stratum, where the weights are the proportion of the population in each stratum.
@@
 
-A slow increase in incidence occurs over the autumn 2020 period, peaking in late November in most regions (see \cref{transmission:fig:backcalc-regions}).
+A slow increase in incidence is estimated to occur over the autumn 2020 period, peaking in late November in most regions (see \cref{transmission:fig:backcalc-regions}).
 An exception is the North East and East of England regions (see \cref{transmission:fig:backcalc-regions}(C) and (D) respectively); these regions show no consistent increase or decrease in incidence during October.
 All regions then decrease during November, coinciding with the national lockdown.
-A rise then occurs, although starting at different times in different regions.
+
+During November, incidence starts rising, although starting at different times in different regions.
 This could be due to the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country\todo{cite Alpha variant}.
-This leads to a second peak in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
+A second peak occurs in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
 \begin{figure}
@@
 
-The incidence proportion was highest among the 11--15 and 16--25 age groups (see \cref{transmission:fig:backcalc-ages}).
+The incidence proportion was highest among the 11--15 and 16--25 age groups throughout the period (see \cref{transmission:fig:backcalc-ages}).
 Except the 16--25 age group, all age groups show similar trends (\ie parallel lines on a log scale).
 This is the pattern predicted by an age-stratified transmission model (see \cref{E-SEIR:sec:structured-populations}), if behavioural changes are independent of age group, and there is negligible impact of immunity.
-The 16--25 age group decreases faster during November.
+The incidence proportion in the 16--25 age group decreases faster than other age groups during November.
 There are many plausible explanations for this effect.
-The higher incidence in this age group prior to November means that this could be an immunity effect.
-Alternatively, it could be a consequence of schools remaining open in the November lockdown, which has a small effect on 16--25 as they interact less with school-age children than other age groups.
+The higher incidence in this age group prior to November means it could be an immunity effect.
+Alternatively, it could be a consequence of schools remaining open in the November lockdown; 20--24 year olds rarely interact with school-aged children (see \cref{E-SEIR:fig:age-contacts})
 Many other explanations are possible, and further work is required to determine causality.
@@
 
-An edge effect where a rapid rise is seen between day 1 and 2 is visible (see \cref{transmission:fig:start-effect}).
+An edge effect where a rapid rise is seen between day 1 and 2 is visible (see \cref{transmission:fig:backcalc-start-effect}).
 This is due to the incorrect assumption that incidence is constant in the past.
@@
 
-As discussed in \cref{E-inc-prev:sec:infection-process}, there are various advantages to using a mechanistic model to estimate transmission.
+There are various advantages to using a mechanistic model to estimate transmission (see \cref{E-inc-prev:sec:infection-process}).
 This section uses a mechanistic model (introduced in \cref{E-SEIR:sec:mechanistic-models}) to estimate the transmission of SARS-CoV-2 in England.
@@
 A novel approach (\cref{SEIR:sec:observation}) links the transmission model to the data.
-In the mechanistic modelling literature, this part of the model is referred to as the observation model, however, it includes both the prevalence process and the observation process (as defined in \cref{E-inc-prev}).
+In the mechanistic modelling literature, this part of the model is referred to as the observation model; however, in the terminology of \cref{E-inc-prev}, it includes both the prevalence process and the observation process.
 
@@
 \subsection{Methods}
-\todo[inline]{refer to where quantities are defined}
-I use the same CIS data as in \cref{E-backcalc}, again considering each region independently.
+\todo[inline]{add a bit more intro in the context of chapter 3}
 The model will be stratified only by region and age group, so the data are aggregated to these levels.
@@
 A parsimonious model is needed to parameterise the WAIFW matrix (see \cref{E-SEIR:sec:structured-populations}).
-I follow the WAIFW matrix parameterisation of \textcite{birrellRealtime}, with the decomposition $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'} N_{a'}$.
-This separates contact patterns, $\kappa_{aa'}$, which can vary over time; transmission probabilities $q_{aa'}$, assumed constant; a region-wide effect for time $\beta(t)$; and the population of the stratum, $N_{a'}$.
+I follow the WAIFW matrix parameterisation of \textcite{birrellRealtime}, decomposition $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'} N_{a'}$.\todo{remind readers what these quantities are here}
+This separates contact patterns, $\kappa_{aa'}$, which can vary over time; transmission probability upon contact between age groups $a$ and $a'$, $q_{aa'}$, assumed constant; a region-wide effect for time, $\beta(t)$; and the population of the stratum, $N_{a'}$.
 
-This parameterisation takes information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using pre-existing methodology~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}.
+I take information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using pre-existing methodology~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}.
 %; these estimates have been used for previous modelling of the COVID-19 pandemic~\autocites{vanleeuwenTime}{birrellRealtime}.
@@
 Google mobility data is based on the location of Android users over time, relative to a pre-pandemic baseline.
-The mobility and school data are then used to down or up-weight the number of contacts in each setting, and then recombined across settings to form $\kappa_{aa'}$
+The mobility and school data are averaged over a week then used to down or up-weight the number of contacts in each setting, before combining contacts across settings to form $\kappa_{aa'}(t)$.
+$\kappa_{aa'}(t)$ is assumed to be piecewise constant, changing only at the start of each week.
 For further details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.
@@
 where $\tilde{w}(t)$ is the week number of time $t$, with week 1 being the first week considered.
-Two hyperparameters will be inferred: $\beta_1$, the initial value of the random walk; and $\sigma_\epsilon$, the standard deviation of the random walk, will have priors placed on them (see \cref{SEIR:sec:priors}).
+Two hyperparameters will be inferred: $\beta_1$, the initial value of the random walk; and $\sigma_\epsilon$, the standard deviation of the random walk (see \cref{SEIR:sec:priors}).
 
@@
 
-Finally,$q_{aa'}$ needs to be specified.
-I again follow the approach of \textcite{birrellRealtime}, decomposing it into two independent factors such that $q_{aa'} = q^\text{susc}_{a} q^\text{inf}_{a'}$, where $q^\text{susc}_a$ is the \emph{susceptibility} of stratum $a$, \ie how vulnerable to infection individuals in stratum $a$ are; and $q^\text{inf}_{a'}$ is the \emph{transmissibility} of stratum $a'$, \ie how likely individuals in stratum $a$ are to pass on an infection.
+Finally, $q_{aa'}$ needs to be specified.
+I follow a common approach\todo{cite anderson and may}, decomposing it into two independent factors such that $q_{aa'} = q^\text{susc}_{a} q^\text{inf}_{a'}$, where $q^\text{susc}_a$ is the \emph{susceptibility} of stratum $a$, \ie how vulnerable to infection individuals in stratum $a$ are; and $q^\text{inf}_{a'}$ is the \emph{transmissibility} of stratum $a'$, \ie how likely individuals in stratum $a$ are to pass on an infection.
 % Then $\beta^\text{susc}_i$ is the relative risk of an individual in stratum $i$ being infected, relative to an individual in stratum $a_0$.
@@
 Studies for pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults~\autocites{chenRole}{vinerTransmission}.
-Therefore, I set $q^\text{inf}_{a'} = 1$ for all $a'$, $q^\text{susc}_{a} = q_c$ for the youngest stratum, and $q^\text{susc}_{a} = 1$ for the remaining strata.
+Therefore, I set $q^\text{inf}_{a'} = 1$ for all $a'$, $q^\text{susc}_{a} = q_c$ for 0--11 year olds, and $q^\text{susc}_{a} = 1$ for the remaining strata.
 $q_c$ is then estimated, although using an informative prior (see \cref{SEIR:sec:priors}).
@@
 
+\todo[inline]{consider if this paragraph is too repetitive}
+
 The transmission model represents the infection process (introduced in \cref{E-inc-prev:sec:infection-process}).
 A variety of different approaches can be taken to link a transmission model to observations.
-In the context of CIS, this requires representing the prevalence and observation processes (introduced in \cref{E-inc-prev:sec:prevalence-process,E-inc-prev:sec:observation-process} respectively)
+In the context of CIS, this requires representing the prevalence and observation processes (introduced in \cref{E-inc-prev:sec:prevalence-process,E-inc-prev:sec:observation-process} respectively).
 The method to do so is known as the observation model.
@@
 One approach would be to convolve the incidence based on the transmission model with the PCR duration distribution.
-Here incidence is calculated from the ODE solution and then related to the data as explained in \cref{E-inc-prev}.
+Here, the daily incidence proportion for each stratum is by solving the system of ODEs governing the transmission model, and then prevalence calculated as in \cref{backcalc}.
 However, this approach is computationally expensive because of the need to calculate the convolution in \cref{E-inc-prev:eq:EPt}.
 
-A much more efficient option is to approximate this convolution by adding compartments to the ODEs.
-This approach also allows the infected, pre-positive compartment, which I denote $p_0$, to be represented.
+A computationally cheaper option is to approximate this convolution by adding compartments to the ODEs.
+This approach allows a infected, pre-positive compartment, which I denote $p_0$, to be included simply.
 $p_0$ increases at the same rate as individuals flow into $e_1$.
 These individuals then transition into a set of compartment with their transition rates, $p_1, \dots, p_l$, chosen to match the duration of positivity estimates in \cref{E-imperf-test}.
-These compartments do not have a biological interpretation, unlike the other compartments in the system, but are purely a mathematical tool so that the time spent in these compartments matches the duration distribution.
-This is the same idea behind having multiple exposed or latent compartments (introduced in \cref{E-SEIR:sec:non-exponential}).
+These compartments do not have a biological interpretation; the same idea is behind having multiple exposed or latent compartments (introduced in \cref{E-SEIR:sec:non-exponential}).
 
@@
 % The time between entering the 1st state and the $l+1$th state, the absorbing state, is the duration.
+These distributions contain five parameters.
+Three are positive reals, $\kappa_Y$, $\kappa_{X1}$, and $\kappa_{X2}$; one is a probability, $p_x$, in the interval $[0, 1]$; and one is a positive integer, $l$.
+
+The name Erlang-Coxian comes from the two distributions that make up the EC distribution.
 Gamma distributions with integer first parameters (as $K$ has) are also known as Erlang distributions.
@@
     \item If there exists a phase-type distribution $P'$ that well-matches $G$ then there exists an EC distribution $P$ that well-matches $G$.
-    \item The parameters of $P$ are available in closed form, and the expression for these is derived.
-    \item The number of phases used by the $P$ calculated by these expressions is at most one more than the minimum across all phase-type distributions that well-match $G$.
+    \item The parameters of $P$ are available in closed form, which \textcite{osogamiClosed} give.
+    \item The number of phases used by the $P$ calculated by these expressions is at most one more than the minimum across all well-matching $P'$.
 \end{itemize}
@@
 I use the mapfit package~\autocite{mapfit} to compute $P$ for the posterior mean of the distribution estimated in \cref{E-imperf-test}.
-This is the same duration distribution used in \cref{E-backcalc}.
-The parameter estimates are in \cref{SEIR:table:ec-params}.
+The resulting parameters for the EC distribution are in \cref{SEIR:table:ec-params}.
 \begin{table}
@@
 
-The modelled proportion of individuals that are PCR positive in the are linked to the data using a beta-binomial likelihood (see\todo{ref section on clustering} for the issues with using a binomial likelihood).
+I link the modelled proportion of individuals that are PCR positive to the data using a beta-binomial likelihood (see\todo{ref section on clustering} for the issues with using a binomial likelihood).
 If on each day $t$, each strata $i$ is observed to have $y_{it}$ positives out of $n_{it}$ tests, then the likelihood is:
@@
 where $\bar{p}_{kat} = \int_{t}^{t+1} p_{ka}(s) ds$ is the mean occupancy in compartment $p_k$ and strata $a$ over day $t$ and $\rho$ is a nuisance parameter controlling the overdispersion of the beta-binomial distribution.
-In this parameterisation (detailed in \cref{E-distributions}) at $\rho=0$ the beta-binomial coincides with the binomial distribution and as $\rho \to \infty$, the distribution tends towards the uniform distribution on $[0, 1]$~\autocite{hughesUsing}.
+In this parameterisation (detailed in \cref{E-distributions}), $\rho=0$ means the beta-binomial coincides with the binomial distribution.
+As $\rho \to \infty$, the distribution tends towards the uniform distribution on $[0, 1]$~\autocite{hughesUsing}.
 
@@
 
-The remaining model structure follow \textcite{birrellRealtime}.
+The remaining details of the transmission model structure follow \textcite{birrellRealtime}.
 The transmission model is a SEIR model with two latent and two infectious states.
@@
 A diagram of the model for a single stratum is shown in figure \cref{SEIR:fig:full-model}.
-The system of differential equations for each region is below.
+The full system of differential equations for each region is below.
 \begin{align}
@@
 The prior on $\rho$ favours a null model of a binomial likelihood.
-The remaining parameters had weakly informative priors.
+The remaining parameters have weakly informative priors.
 
@@
 
-I use an adaptive random-walk Metropolis-Hastings algorithm (previously discussed in \cref{E-inc-prev:sec:MCMC}).
+I compute the posterior distribution using MCMC.
+Specifically, I use an adaptive random-walk Metropolis-Hastings algorithm (previously discussed in \cref{E-inc-prev:sec:MCMC}).
 The proposal distribution is a multivariate normal that adapts to the covariance of the posterior.
@@
 During such an exponential growth phase, all components of the $\vec{e}$, $\vec{i}$ and $\vec{p}$ compartments grow at the same growth rate, $\psi$.
-This follows directly from the ODEs where $\vec{s}$ is constant. 
+This follows directly from the ODEs if $\vec{s}$ is constant. 
 The growth rate $\psi$ is easily identifiable from the data, and therefore a parameterisation in terms of $\phi$ should be more computationally efficient.
@@
 
-Furthermore, $\R$ is a function of $\matr{\beta}$, and hence $\beta_1$ (as explained in \cref{E-SEIR:sec:structured-populations}).
+Furthermore, $\R$ is a function of $\matr{\beta}(0)$, and hence $\beta_1$ (as explained in \cref{E-SEIR:sec:structured-populations}).
 Combining these expressions, the model can be parameterised by using $\psi$ in place of $\beta_1$.
@@
 First, calculate $\R$ from $\psi$ using \cref{SEIR:eq:rtoR}; denote this $\hat{R}$.
-Then calculate $R^* = D(\matr{\beta}) / \gamma$ when $\beta_1=1$ (recall that $D(\matr{\beta})$ denotes the dominant eigenvalue of $\matr{\beta}$).
-Multiplying a matrix by a scalar multiplies its dominant eigenvalue by that same scalar, therefore, setting $\beta_1 = \hat{R} / R^*$ gives the desired means that $\R$ at time 0 is $\hat{R}$, as desired.
+Then calculate $R^* = D(\matr{\beta}(0)) / \gamma$ when $\beta_1=1$ (recall that $D(\matr{\beta})$ denotes the dominant eigenvalue of $\matr{\beta}$).
+Multiplying a matrix by a scalar multiplies its dominant eigenvalue by that same scalar, therefore, setting $\beta_1 = \hat{R} / R^*$ gives means that $\R$ at time 0 is $\hat{R}$, as desired.
 
@@
 If the distributions for length-of-stay in each compartment are known, then all other model parameters can  be estimated.
-This includes the age distribution of the infections and the susceptibility of children.
+This includes the age distribution of the infections and the components of transmission.
 
@@
 % The CIS allows estimates of transmission in children, unlike sources relying on severe events that occur almost entirely in adults~\autocite{bhopalChildren}.
+Using a mechanistic model provides additional understanding, compared to the approach in \cref{E-backcalc}.
+For example, decomposing the change in transmission into changing contacts, $\matr{\kappa}(t)$, changing susceptibility $\vec{s}(t)$, and other unmeasured factors, $\beta(t)$.
+By understanding which of these components is changing, policies can be better targeted.
+If it is shown that $\beta(t)$ decreases, this implies that for a given contact rate in the population, there is less transmission.
+Such decreases might be beneficial as it allows society to operate in a more normal manner without increasing transmission.
+% Increasing activity without increasing transmission is desirable because it enables reductions the disease burden without burdesome restrictions on individuals day-to-day lives.
 
@@
 Many factors contribute to the pattern in the $\beta_w$ values, which have a minima near the start of December.
-While this analysis cannot establish causation, several hypotheses can be identified that are worth further investigation.
+While this analysis cannot establish causation, several hypotheses that are worth considering.
 
-The decreases in $\beta_w$ over the autumn could be due to increased caution (\eg masking or stricter adherence to social distancing) as cases and deaths increased~\autocite{jarvisEffect} and/or interventions not captured by $\matr{\kappa}$ such as the impact of local measures in the highest transmission areas~\autocite{scottCovid19}.
-Local measures are not well captured by $\matr{\kappa}$ because the matrix averages behaviour across the whole of England.
-Allowing each region to have its own $\matr{\kappa}$ may capture the effect of local measures more accurately, although their targeted nature may require more granular methods.
+The decreases in $\beta_w$ over the autumn could be due to increased caution (\eg masking or stricter adherence to social distancing) as cases and deaths increased~\autocite{jarvisEffect} and/or not captured by $\matr{\kappa}(t)$, such as the impact of local measures in the highest transmission areas~\autocite{scottCovid19}.
+Local measures are not well captured by $\matr{\kappa}(t)$ because the matrix averages behaviour across the whole of England.
+Allowing each region to have its own $\matr{\kappa}(t)$ may capture the effect of local measures more accurately, although their targeted nature may require more granular methods.
 
@@
 
-\todo[inline]{add comparison to backcalculation once I have those results}
-
-Using a mechanistic model provides additional understanding, compared to the approach in \cref{E-backcalc}.
-For example, decomposing the change in transmission into changing contacts, $\matr{\kappa}(t)$, changing susceptibility $\vec{s}(t)$, and other unmeasured factors, $\beta(t)$.
-By understanding which of these components is changing, policies can be better targeted.
-If it is shown that $\beta(t)$ decreases, this implies that for a given contact rate in the population, there is less transmission.
-Such decreases might be beneficial as it allows society to operate in a more normal manner without increasing transmission.
-Breaking the trade-off between transmission and activity is desirable because it reduces the disease burden without burdesome restrictions on individuals day-to-day lives.
-
 \Cref{transmission:fig:compare-regions} compares the regional incidence estimates between the two models in this chapter.
@@
 The largest discrepancy is that the backcalculation model estimates a second peak in mid January in several regions (most clearly in London, North West, and the East of England).
-However, even here, the credible intervals overlap except in London.
+However, except in London, the credible intervals still overlap.
 This is a specific case of a larger pattern: the SEIR model produces smoother estimates of incidence than the backcalculation model.
@@
 The SEIR model imposes stronger assumptions about the epidemic dynamics.
-The assumptions reduces the variance of the estimates, including making them smoother, but if the assumptions are incorrect, bias is introduced.
-This trade-off based on the strength of assumptions is well-known to apply in the backcalulcation context~\autocite[e.g.][section 8.3]{brookmeyerBackcalculation}.
+The assumptions reduces the variance of the estimates, including by making them smoother, but if the assumptions are incorrect, bias is introduced.
+This trade-off based on the strength of assumptions has long been known to apply in the backcalulcation context~\autocite[e.g.][section 8.3]{brookmeyerBackcalculation}.
 The SEIR model can be viewed as a backcalulation method where the transmisson part of the model is a smoothing method.

### 2024-01-26 cis-imperfect-testing.tex

--- 2024-01-17 cis-imperfect-testing.tex
@@
 
+{\color{red}{NEW BIT}}
+In \cref{E-perf-test}, a framework for analysing the CIS to produce estimates of the duration of positivity was developed.
+However, initial application of this framework produced implausibly short estimate.
+In this chapter I identify this issue to be the result of false negatives, a form of misclassification bias, within the CIS (\cref{imperf-test:sec:problem}).
+In \cref{imperf-test:sec:modelling} I develop the model in \cref{E-perf-test} to include the possibility of observing false negatives.
+There exists little prior work on the incorporation of misclassification bias within similar survival analysis frameworks.
+\todo[inline]{Review literature properly, here or elsewhere}. Building on this literature, model adaptation to include false negatives involves.....GIVE DETAILS OF WHAT YOU DID
+After showing through simulation (\cref{imperf-test:sec:sim-study-results}) that the new model recovers the true duration distribution (\cref{imperf-test:sec:sim-study-results}),  and I apply it to the CIS data (\cref{imperf-test:sec:application}).
+Finally, I discuss the results and further work in \cref{imperf-test:sec:discussion}.
+
 \section{The problem} \label{imperf-test:sec:problem}
 
-Applying the method of \cref{E-perf-test} to CIS data gives implausible estimates of the duration distribution (see \cref{imperf-test:fig:problem-cis-estimates}).
-In particular, it estimated a much shorter duration of positivity than in \cref{E-ATACCC}.
+Applying the method of \cref{E-perf-test} to CIS data estimates fewer individuals of a duration less than three weeks, compared to \cref{E-ATACCC} (see \cref{imperf-test:fig:problem-cis-estimates}).
 The \cref{E-ATACCC} estimates are more reliable over the first 2--3 weeks due to them having dense data collection here.
@@
 Comparing repeated data simulations using the setup of \cref{E-perf-test:sec:simulation-study} to the observed CIS data, suggests that this is due to the high frequency of single positive episodes in the CIS data (compare the histogram with $\psens=1$ and the vertical line in \cref{imperf-test:fig:sim-single-pos}).
-A single positive episode is an episode containing exactly one positive test; intuitively, these cause short estimates because the lower bound on the length of time the episode is 1 day, the shortest possible.
+A single positive episode is an episode containing exactly one positive test; intuitively, these cause estimates to be shorter because the lower bound on the length of time the episode is 1 day, the shortest possible.
 Specifically, if $i$ is a single positive episode then $i$'s admissible region (as defined in \cref{E-perf-test:sec:model}) includes $b_i = e_i = r_i^{(b)} = l_i^{(e)}$.
@@
 Additionally, short episodes are very likely to be missed, which means that the design of the CIS cannot rule out many of them occurring.
-\todo{possibly a bit unclear phrasing here? struggling with it}
-This produced the same pattern, that is estimating too many short episodes (not shown).
+% This produced the same pattern, that is estimating too many short episodes (not shown).
 
@@
 I hypothesised that the reason for the unexpectedly high number of single positive episodes (compared to simulation) was the presence of false negative results.
-There are two strands of evidence supporting this hypothesis.
+There are several strands of evidence supporting this hypothesis.
 First, it is well known that PCR testing can return false negatives  (see \cref{E-biology-data:sec:PCR}); I explicitly include them in the viral load model of \cref{E-ATACCC}.
 Intermittent negatives (described in  \cref{E-biology-data:sec:PCR}) show that they occur within the CIS.
-Second, I expanded the simulation study of \cref{E-perf-test:sec:simulation-study} to include false negatives (see \cref{imperf-test:sec:simulate} for more details).
+Furthermore, in \cref{imperf-test:sec:simulate}, I expanded the simulation study of \cref{E-perf-test:sec:simulation-study} to include false negatives (see  for more details).
 This reproduced the CIS data more faithfully and exhibited the same issues when estimating the duration distribution (see \cref{imperf-test:fig:sim-single-pos}).
@@
 A better model would be to allow the rate of false negatives to vary over the course of an episode.
-False negatives are more likely to occur when the viral load of an individual is low, because there is less virus in their body to be sampled.
+False negatives are more likely to occur when the viral load of an individual is low, because there is less virus in their body to sample.
 In \cref{E-ATACCC}, this mechanism is incorporated by viewing negative tests as a left-truncation of the observation noise (see \cref{E-ATACCC:sec:observation-modification}).
@@
 It is a false negative if and only if the episode ends after the test, \ie $e_i > r_i^{(e)}$.
-I proceed by considering whether this is the case.
+I proceed by considering whether this is the case, and considering the case with $b_i$ known; $b_i$ will later be integrated out.
 
@@
 &= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) \right)
+\label{imperf-test:eq:ll-ei-lt-ri}
 \end{align}
@@
 In this case, the test at $r_i^{(e)}$ is a false negative, occurring with probability $(1 - p_\text{sens})$.
-As I only consider the tests in $\sched_i'$, the end of the episode is right-censored.
+To avoid having to consider tests after $r_i^{(e)}$, which could greatly complicate the likelihood, I model this case as the episode being right-censored at $r_i^{(e)}$.
+\todo{check previous notes - I think I investigated this assumption a bit more and maybe did some sensitivity analysis}
 Taking the same approach as before:
@@
 &= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) (1 - p_\text{sens}) S_\theta(r_i^{(e)} - b_i - 1)
+\label{imperf-test:eq:ll-ei-gt-ri}
 \end{align}
 
-Combining the above, the replacement for $p_{ia}$ is:
+These expressions can now be used to form the full replacement for $p_{ia}$, $p'_{ia}$, the likelihood of the data $\vec{y_i'}$.
+First, augment the data with $b_i$, and split into the cases just discussed:
 \begin{align}
@@
 &= p(\vec{y_i'} \mid p_\text{sens}, \theta) \\
-&= \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( p(\vec{y_i'}, b_i, e_i \leq r_i^{(e)} \mid p_\text{sens}, \theta) p(\vec{y_i'}, b_i e_i > r_i^{(e)} \mid p_\text{sens}, \theta) \right) p(b_i \mid p_\text{sens}, \theta) \\
-&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \\ & \ \times \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) + (1 - p_\text{sens}) S_\theta(r_i^{(e)} - b_i - 1) \right) \\ & \ \times p(b_i \mid p_\text{sens}, \theta) \\
+&= \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( p(\vec{y_i'}, e_i \leq r_i^{(e)} \mid b_i, p_\text{sens}, \theta) + p(\vec{y_i'}, e_i > r_i^{(e)} \mid b_i, p_\text{sens}, \theta) \right) p(b_i \mid p_\text{sens}, \theta).
+\intertext{Now, substitute in \cref{imperf-test:eq:ll-ei-lt-ri,imperf-test:eq:ll-ei-gt-ri}, then take out the common factor:}
+&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \\ & \ \times \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) + (1 - p_\text{sens}) S_\theta(r_i^{(e)} - b_i - 1) \right) \\ & \ \times p(b_i \mid p_\text{sens}, \theta).
+\intertext{This simplifies to:}
 &= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \\ & \ \times \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - p_\text{sens} S_\theta(r_i^{(e)} - b_i - 1) \right) p(b_i \mid p_\text{sens}, \theta).
@@
 
-This false negative would occur on day $b_i + \tau_{\sched_i}(b_i)$ (see \cref{E-perf-test:sec:model}).
+This false negative would occur at the first test after the infection episode begins, on day $b_i + \tau_{\sched_i}(b_i)$ (see \cref{E-perf-test:sec:model}).
 A false negative occurring requires that the episode has not yet ended but a negative still occurs.
-The episode has not yet ended at the time of the test if $e_i = b_i + d_i - 1$ is after the test, that is $d_i > \tau_{\sched_i}(b_i)$.
+The episode has not yet ended at the time of the test if $e_i = b_i + d_i - 1$ is after the test, that is the duration of the infection $d_i > \tau_{\sched_i}(b_i)$.
 Conditional on the episode having not yet ended, the test result is negative with probability $1 - \psens$.
@@
 \end{align}
-Integrating over $b_i$ for $b_i > \min \sched_i$, in the same way as \cref{E-perf-test:eq:piu}, gives:
+Integrating over $b_i$ for $\min \sched_i < b_i \leq T$ ($T$ being the last day of the period considered), in the same way as \cref{E-perf-test:eq:piu}, gives:
 \begin{align}
-\eta = (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_i)}^T \left( S_\theta(\tau_{\sched_i}(b) + 1) - S_\theta(\tau^2_{\sched_i}(b) + 1) \right).
+\eta = (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_i) + 1}^T \left( S_\theta(\tau_{\sched_i}(b) + 1) - S_\theta(\tau^2_{\sched_i}(b) + 1) \right).
 \end{align}
@@
 Denote the replacement for $p_{iu}$ as $p_{iu}'$.
-Since the new mechanism is mutually exclusive with the previous one, $p_{iu}' = p_{iu} + \eta$.
+$p_{iu}'$ is the probability of episode $i$ being missed, considering both the mechanisms from \cref{E-perf-test} and the new mechanism here.
+These mechanisms are mutually exclusive.
+Hence, $p_{iu}'$ is the sum of these, $p_{iu}' = p_{iu} + \eta$.
 For the posterior density, \cref{E-perf-test:eq:full-posterior}, $1 - p_{iu}'$ is the required quantity.
@@
 1 - p_{iu}'
-&= 1 - p_{iu} - (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(t_i)}^T \left( S_\theta(\tau_{\sched_i}(b) + 1) - S_\theta(\tau^2_{\sched_i}(b) + 1) \right) \\
-&= \frac{1}{T} \sum_{b=\min(t_i)}^T \left( p_\text{sens} S_\theta(\tau_{\sched_i}(b) + 1) + (1 - p_\text{sens}) S_\theta(\tau^2_{\sched_i}(b) + 1)\right).
+&= 1 - p_{iu} - (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_i) + 1}^T \left( S_\theta(\tau_{\sched_i}(b) + 1) - S_\theta(\tau^2_{\sched_i}(b) + 1) \right) \\
+&= \frac{1}{T} \sum_{b=\min(\sched_i) + 1}^T \left( p_\text{sens} S_\theta(\tau_{\sched_i}(b) + 1) + (1 - p_\text{sens}) S_\theta(\tau^2_{\sched_i}(b) + 1)\right).
 \label{imperf-test:eq:pit-prime}

### 2024-01-26 incidence-prevalence.tex

@@
 \emph{Incidence} on day $t$, $Z_t$, is the number of people in a given population who become infected on day $t$; when expressed as a proportion of the population it is the \emph{incidence proportion}~\autocite[89]{lashModern}, $Z_t/\Npop$, where $\Npop$ is the number of people in the population.
+\todo{add discussion of proportions vs rates}
 As discussed in\todo{ref intro}, the timing of infection events is very rarely directly observable and must be inferred from other data.

### 2024-01-26 transmission.tex

@@
     I should discuss the literature on estimating incidence based on SARS-CoV-2 prevalence data.
-    In particular: \textcites{mccabeCISincidence}{abbottCISincidence}.
+    In particular: \textcites{mccabeCISincidence}{abbottCISincidence}. Also Toffe.
     However, I think this goes better in the introduction, framing the whole thesis because it is important in understanding the motivation for the whole work.
@@
 
-Studies for pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults~\autocites{chenRole}{vinerTransmission}.
+Studies of pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults~\autocites{chenRole}{vinerTransmission}.
 Therefore, I set $q^\text{inf}_{a'} = 1$ for all $a'$, $q^\text{susc}_{a} = q_c$ for 0--11 year olds, and $q^\text{susc}_{a} = 1$ for the remaining strata.
+$q_c$ can be interpreted as the relative probability that a child is infected (compared to an adult), given contact with an infectious individual.
 $q_c$ is then estimated, although using an informative prior (see \cref{SEIR:sec:priors}).

### 2024-01-26 transmission.tex
Overleaf editing

--- 2024-01-25 transmission.tex
@@
 
-In this chapter, I use the preferred estimate of the duration of PCR positivity (see \cref{E-imperf-test:sec:conclusion}) to estimate the transmission of SARS-CoV-2 in England.
+In this chapter, I use the estimate of the duration of PCR positivity from \cref{E-imperf-test} to estimate the transmission of SARS-CoV-2 in England.
 This estimation is based on the CIS data (see \cref{transmission:sec:data} for further details).
@@
 
-I take two approaches; the theoretical advantages of each were discussed in \cref{E-inc-prev:sec:infection-process}.
-First, in \cref{backcalc}, I use a backcalculation approach, based on \cref{E-inc-prev:eq:EPt}.
-Then, in \cref{SEIR}, I use a mechanistic compartmental model, an implementation of the model introduced in \cref{E-SEIR:sec:mechanistic-models}.
+I take two approaches.
+The theoretical advantages of each approach were discussed in \cref{E-inc-prev:sec:infection-process}.
+First, in \cref{backcalc}, I use a back-calculation approach, based on \cref{E-inc-prev:eq:EPt}.
+Then, in \cref{SEIR}, I use a mechanistic compartmental model, extending the model introduced in \cref{E-SEIR:sec:mechanistic-models}.
 Finally, in \cref{transmission:sec:comparison}, I compare the results of the two approaches.
@@
 
+\todo[inline]{make sure to proofread this section due to merge conflicts}
 As discussed in\todo{ref section}, a major issue with the CIS is non-response bias.
 However, a deterministic relation between true prevalence and incidence is a good approximation (see \cref{E-inc-prev:sec:prevalence-process}).
-Therefore, I extend the methodology previously used to estimate prevalence from the CIS data, corrected for non-response bias, to estimate incidence\todo{cite Koen's stuff}.
+Therefore, I extend the methodology previously previously used to estimate prevalence from the CIS data, corrected for non-response bias, to estimate incidence\todo{cite Koen's stuff}.
 The previous method was multilevel regression and poststratification (MRP)\todo{cite MRP}.
@@
 Finally, the prevalence is poststratified to the CIS target population and subpopulations of interest.
-I follow this same process, except I transform the posterior distribution of prevalence in each stratum to a posterior distribution of incidence before poststratifying.
+I follow this same process, except I transform the posterior distribution of prevalence in each stratum to a posterior distribution of incidence before post-stratifying.
 % I fit a modification model to estimate prevalence, then transform the posterior distribution of prevalence in each stratum to a posterior distribution of incidence.
@@
 I modified a pre-existing version of a MRP model for CIS~\autocite{pouwelsMRPvaccination}, as recommended by that model's author for this situation~\citePersonalComms{Koen Pouwels}.
-First, a Gaussian Markov Random Field (GMRF) logistic regression model is used to estimate the prevalence in each stratum.
-For a day $t$, age group $a$, ethnicity $e$, and sex $s$ assume that:
-\begin{align}
-    \logit(p_{taes}) &= \beta + \beta_s + \beta_{w(e)} + u_a + v_e + w_t + l_{a,t} \\
-    y_{taes} &\dist \BB(n_{taes}, p_{taes}, \rho)
-\end{align}
-The terms are as follows.
-\begin{itemize}
-    \item $\beta$ is a global intercept term with the prior $\beta \dist \N(0, 1/0.001^2)$.
-    \item $\beta_s$ is a fixed effect for sex, with $\beta_s = 0$ for males and estimated for females.
-$\beta_{w(e)}$ is a fixed effect for white or non-white ethnicity.
-$\beta_{w(e)} = 0$ if $e$ corresponds to white ethnicity and $\beta_{w(e)} = \beta_{n}$ otherwise (\ie non-white ethnic groups).% $\beta_n$ and $\beta_f$ have the default INLA prior for fixed effects\todo{check what this is}.
-    \item $u_a$ and $v_e$ are independent Gaussian random effects for age and ethnic group respectively.
-Their standard deviation have a penalised complexity prior\todo{cite PC prior} with a 10\% probability of being greater than 1.
-    \item $w_t$ is a second-order random walk, that is $w_t = 2w_{t-1} - w_{t-2} + \epsilon_t$ where $\epsilon_t \dist \N(0, \sigma_w^2)$.
-    \item $l_{a,t}$ is an age-group specific second-order random walk. That is $l_{a,t} = 2l_{a,t-1} - l_{a,t-2} + \epsilon_{a,t}$ where $\epsilon_{at} \dist \N(0, \sigma_l^2)$.
-    \item The likelihood is a beta-binomial model (defined in \cref{E-distributions}), an overdispersed binomial.
-Overdispersion allows for household clustering\todo{cite beta-binomial as a model for clustering} and other overdispersion in the data, which is common in epidemiological data\todo{cite that epi data is often overdispersed}.
-\end{itemize}
-All parameters have priors using the default specified in INLA, except where stated above.
-For these parameters, the large amount of data means that the prior has little influence on the posterior distribution.
-
-The model was fit separately and independently for each region, implicitly adding interactions between region and all other variables.
+\todo[inline]{Describe the model precisely, see eg RTM paper 2}
+% The model included fixed effects for age group\todo{check the model I used}.
+% The likelihood is a beta-binomial distribution, a generalisation of the binomial distribution allowing for overdispersion (see \cref{E-distributions}).
+% The overdispersion allows for household clustering\todo{ref beta-binomial as a model for clustering} and any other overdispersion in the data\todo{ref that epi data is often overdispersed}. 
+Finally, the model was fit independently for each region, implicitly adding interactions between region and all other variables.
 This step assumes that the estimates for each region can be considered independent, which is reasonable because the regions are large with many tests conducted within each.
@@
 Therefore, \cref{E-inc-prev:eq:EPt} implies that $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix of the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
-The matrix $\matr{S}$ is defined to maintain the relationship between incidence and prevalence in \cref{E-inc-prev:eq:EPt}.
+The matrix $\matr{S}$needs to bebased on preserves the relationship between incidence and prevalence.
 This gives a lower-triangular of the following form:
@@
 
-\todo{Insert constant incidence assumption methodology}
-
-Model misspecification means that the procedure described could produce negative incidence.
-For any stratum and day $t$, this occurs if the difference between $P_t$ and $P_{t+1}$ is too large based on the number of recoveries that occur on day $t$ based on the incidence up to day $t$ and the duration distribution.
-
-There are two places where this misspecification could occur.
-The first is the survival matrix, $\matr{S}$.
-This is assumed known but \cref{E-imperf-test} showed that it is sensitive to assumptions made.
-The second is the model used to estimate prevalence.
-
-A prior constraining incidence to be non-negative and then fit directly to the data could be used but, as previously discussed, this would no longer be a GLM and could not be fit within the R-INLA framework.
-Hence, an alternative inference methodology would be required.
-MCMC is likely too computationally expensive.
-Other approaches, such as a Laplace approximation or variational inference, could be considered.
-For this analysis, I replaced any value of incidence estimated to be negative with 0.
-
 The final step is poststratifying the incidence proportion to the CIS target population and subpopulations of interest.
-For each posterior draw and day, the incidence proportion in the (sub)population is the weighted average of the incidence proportion in each stratum, where the weights are the proportion of the population in each stratum.
+For each posterior draw and day\todo{introduce discrete time in days earlier}, the incidence proportion in the (sub)population is the weighted average of the incidence proportion in each stratum, where the weights are the proportion of the population in each stratum.
 The population numbers for each stratum are provided by the Office for National Statistics.
 
-\subsection{Results and discussion} \label{backcalc:sec:results}
+\subsection{Results} \label{backcalc:sec:results}
 
-A slow increase in incidence occurs over the autumn 2020 period, peaking in late November in most regions (see \cref{transmission:fig:backcalc-regions}).
-An exception is the North East and East of England regions (see \cref{transmission:fig:backcalc-regions}(C) and (D) respectively); these regions show no consistent increase or decrease in incidence during October.
-All regions then decrease during November, coinciding with the national lockdown.
-A rise then occurs, although starting at different times in different regions.
-This could be due to the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country\todo{cite Alpha variant}.
-This leads to a second peak in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
-\begin{figure}
-    \includegraphics{transmission/backcalc-regions}
-    \caption[Incidence estimated using backcalculation by region]{%
-        Incidence proportion for all of England (A) and its component regions.
-        Regions are split into broad groupings of three regions each: Northern (B), Midlands (C), and Southern (D).
-        Yorkshire and the Humber is labelled as Yorkshire for brevity.
-    }
-    \label{transmission:fig:backcalc-regions}
-\end{figure}
-
-The incidence proportion was highest among the 11--15 and 16--25 age groups (see \cref{transmission:fig:backcalc-ages}).
-Except the 16--25 age group, all age groups show similar trends (\ie parallel lines on a log scale).
-This is the pattern predicted by an age-stratified transmission model (see \cref{E-SEIR:sec:structured-populations}), if behavioural changes are independent of age group, and there is negligible impact of immunity.
-The 16--25 age group decreases faster during November.
-There are many plausible explanations for this effect.
-The higher incidence in this age group prior to November means that this could be an immunity effect.
-Alternatively, it could be a consequence of schools remaining open in the November lockdown, which has a small effect on 16--25 as they interact less with school-age children than other age groups.
-Many other explanations are possible, and further work is required to determine causality.
-\begin{figure}
-    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{transmission/backcalc-ages}}
-    \caption[Incidence estimated using backcalculation by age group]{%
-        Incidence proportion for each age group, on both natural and log scales.
-    }
-    \label{transmission:fig:backcalc-ages}
-\end{figure}
-
-An edge effect where a rapid rise is seen between day 1 and 2 is visible (see \cref{transmission:fig:start-effect}).
-This is due to the incorrect assumption that incidence is constant in the past.
-Changing this assumption (\eg to exponential growth) modifies the estimated value on the first day, but the rest of the series is insensitive to this change (not shown).
-Therefore, the first day's estimate is unreliable, and has been excluded from all other plots.
-However, the rest of the series is robust to this assumption and hence reliable.
-\begin{figure}
-    \centering \includegraphics{transmission/backcalc-start-effect}
-    \caption[Edge effects in backcalculation method]{%
-        First seven days of incidence proportion estimate for the whole of England, highlighting the edge effect.
-    }
-    \label{transmission:fig:backcalc-start-effect}
-\end{figure}
+\subsection{Discussion} \label{backcalc:sec:discussion}
 
@@
 
-Studies for pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults~\autocites{chenRole}{vinerTransmission}.
+Studies of pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults~\autocites{chenRole}{vinerTransmission}.
 Therefore, I set $q^\text{inf}_{a'} = 1$ for all $a'$, $q^\text{susc}_{a} = q_c$ for the youngest stratum, and $q^\text{susc}_{a} = 1$ for the remaining strata.
@@
 
-The modelled proportion of individuals that are PCR positive in the are linked to the data using a beta-binomial likelihood (see\todo{ref section on clustering} for the issues with using a binomial likelihood).
+The modelled proportion of individuals that are PCR positive in the are linked to the data using a Beta-Binomial likelihood (see\todo{ref section on clustering} for the issues with using a binomial likelihood).
 If on each day $t$, each strata $i$ is observed to have $y_{it}$ positives out of $n_{it}$ tests, then the likelihood is:
@@
 }
-  \caption[SEIR model]{SEIR model used for the analysis in this section. The shaded box represents PCR-positive individuals. Shown for a single stratum.}
+  \caption[SEIR model]{SEIR model used for the analysis in this chapter. The shaded box represents PCR-positive individuals. Shown for a single stratum.}
   \label{SEIR:fig:full-model}
@@
 The prior on $\rho$ favours a null model of a binomial likelihood.
-The remaining parameters had weakly informative priors.
+The remaining priors had weakly informative priors.
 
@@
 
+\todo[inline]{What happened to the following paragraph? It doesn't make sense currently!}
 In this setup, the parameters $i^+$, $q_c$, and $\psi$ are fully recovered; the other parameters are only partially so: their posterior has moved away from the prior but is still regularised by it (see \cref{SEIR:fig:true-vs-posterior}).
 The ability to identify $q_c$ is an advantage of using data which includes measurements on children, unlike data on hospitalisations or deaths.
-
-For $\vec{\pi}$ and $\rho$, the posterior median is very similar regardless of the true value (see \cref{SEIR:fig:true-vs-posterior}).
 This suggests that these parameters are not practically identifiable in this setting.
-The posteriors align very closely with the priors (not shown) which reinforces this finding.
+The posteriors align very closely with the priors (not shown\todo{consider showing prior/posterior plots}) which reinforces this finding.
 The lack of identifiability for $\vec{\pi}$ aligns with previous work showing this parameter is structurally unidentifiable, \ie not identifiable even in the limit of infinite data~\autocite{dankwaStructural}.
@@
 A constant $\matr{\beta}(t)$ would lead to exponential growth following a period of transitory dynamics, although the weekly switching may mean that these transitory dynamics never end, especially when incidence is high~\autocite{rhodesConvergence}.
-When the peak in incidence is estimated to occur varied by region, with the most common time being early January 2021 (see \cref{SEIR:fig:peak-incidence}).
+When the peak in incidence is estimated to occur varied by region, but due to the switching, is almost always on a Sunday (see \cref{SEIR:fig:peak-incidence}).
+For Yorkshire, the peak is estimated to occur in late October.
+For all other regions, it is estimated to occur in late December or Early January, the period during which severe social distancing measures were applied.
 \begin{figure}
@@
 The \emph{attack rate} is the proportion of the population that have been infected at least once\todo{cite attack rate}.
-In this model it is $1 - \vec{s}$, the proportion susceptible.
-The attack rate is highest in the $[11, 16)$ age group in the North West at 68\% (95\% CrI: 66\%--70\%) (see \cref{SEIR:fig:attack-rates}).
+In this model it is $1 - \vec{s}$.
+The attack rate is highest in the $[11, 16)$ age group in the North West at 50\% (95\% CrI: 48\%--52\%) (see \cref{SEIR:fig:attack-rates}).
 \begin{figure}
@@
 Overall the model fits the data well (see \cref{SEIR:fig:prev-young,SEIR:fig:prev-old}).
-There are a few features of the data which are not captured by the model and the posterior estimate appears to be a compromise with the prevalence too low in one age group and too high in another.
-For example, in the North East region, the 11--16 age group has its prevalence overestimated in October but the 16--25 age group has its prevalence underestimated.
+There are a few features of the data which are not captured by the model.
+In particular, prevalence during October is higher in the 16--25s than the model predicts and lower in the 11--16s.
 \begin{figure}
@@
 
-All regions' MCMC chains converged, assessed using the Rhat statistic and ESS (as defined in \cref{E-inc-prev:sec:MCMC}).
-The lowest Rhat was 1.01 for $\sigma_\epsilon$ in London.
-All parameters had ESS > 1000, except $\sigma_\epsilon$ and $\psi$ in London, which had ESS of 450 and 950 respectively.
+One MCMC chain for the North West region appeared to get stuck, with very few moves after the first 100,000 iterations (not shown).
+I reran this chain with a new seed, with the rerun results presented.
+Following this, all regions' MCMC chains converged, assessed using the Rhat statistic and ESS.
+Rhat for all parameters is < 1.01, except $\sigma_\epsilon$ in London which had a Rhat of 1.018.
+All parameters had ESS > 1000, except $\sigma_\epsilon$ in London and the East, which had ESS of 197 and 757 respectively.
+As all the parameters of primary interest had high Rhat and ESS, and no parameters had large issues, the inference procedure is considered to have converged.
 
@@
 
-The results in this section show the feasibility of fitting a mechanistic model using only data on prevalence.
+The results in this chapter show the feasibility of fitting a mechanistic model using only data on prevalence.
 If the distributions for length-of-stay in each compartment are known, then all other model parameters can  be estimated.
@@
 While this analysis cannot establish causation, several hypotheses can be identified that are worth further investigation.
-
-The decreases in $\beta_w$ over the autumn could be due to increased caution (\eg masking or stricter adherence to social distancing) as cases and deaths increased~\autocite{jarvisEffect} and/or interventions not captured by $\matr{\kappa}$ such as the impact of local measures in the highest transmission areas~\autocite{scottCovid19}.
-Local measures are not well captured by $\matr{\kappa}$ because the matrix averages behaviour across the whole of England.
-Allowing each region to have its own $\matr{\kappa}$ may capture the effect of local measures more accurately, although their targeted nature may require more granular methods.
-
-The increases in $\beta_w$ in December coincide with the rise of the Alpha variant~\autocite{lythgoeLineage}, known to be more transmissible~\autocite[e.g.][]{daviesEstimated}.
+The decreases in $\beta_w$ over the autumn could be due to increased caution(\eg masking or stricter adherence to social distancing) and/or interventions short of a lockdown as cases and deaths increased~\autocite{jarvisEffect}, or the impact of local measures in the highest transmission areas~\autocite{scottCovid19}.
+Local measures are not well captured by $\matr{\kappa}$ because the matrix was for the whole of England.
+Allowing each region to have its own $\matr{\kappa}$ may capture the effect of local measures more accurately.
+The increases in $\beta_w$ in December coincide with the rise of the Alpha variant~\autocite{lythgoeLineage}, known to be more transmissible~\autocite{daviesEstimated}.
 A multiple variant model might be able to capture this rise more accurately.
@@
 This implies that the data would allow identification of a more flexible age-based model.
-
-Currently, the age distribution of infections is determined by $\matr{\beta}$, and to a lesser extent $\vec{\pi}$, the former of which only contains one parameter that affects the age distribution, $q_c$.
-Increasing the number of parameters in $\matr{\beta}$ would be the simplest approach to increase the model's flexibility.
-
-The only time-changing part of $\matr{\beta}$ that varies between ages is $\matr{\kappa}$.
-It is plausible that this would not capture all behavioural changes.
-For example, the introduction and relaxation of control measures within schools.
+Currently, the age distribution of infections is determined by $\matr{\beta}$, and to a lesser extent $\vec{\pi}$, the former of which only contains one parameter that affects the age distribution, $\beta_c$.
+More parameters could capture the age distribution more accurately.
+There are various ways to increase the model's flexibility, including introducing additional susceptibility or infectiousness parameters, or by allowing these parameters to vary over time.
+It is plausible that these distributions changed over time in a way not captured by $\matr{\kappa}$, for example due to the introduction and relaxation of control measures within schools.
 $\matr{\kappa}$ only captures changes in in-school transmission through changes in school attendance, not incorporating the probability of transmission, which could be modelled as changes in $q_{aa}$ for $a$ of school age.
@@
 
-The weekly patterns in incidence are a limitation of the current parameterisation.
+The weekly patterns in incidence are another limitation of the current parameterisation.
 The random walk and contact matrices both vary only on the weekly timescale.
@@
 However, this would greatly increase the number of parameters.
-An alternative would be a smoothly-varying function or an approximation of a random walk, such as that proposed by \textcite{ghoshApproximate}.
+An alternative would be a smoothly-varying function or an approximation of a random walk such as that proposed by \textcite{ghoshApproximate}.
 
@@
 
-\Cref{transmission:fig:compare-regions} compares the regional incidence estimates between the two models in this chapter.
-Generally the estimates agree, with the SEIR model having much narrower credible intervals.
-The largest discrepancy is that the backcalculation model estimates a second peak in mid January in several regions (most clearly in London, North West, and the East of England).
-However, even here, the credible intervals overlap except in London.
-This is a specific case of a larger pattern: the SEIR model produces smoother estimates of incidence than the backcalculation model.
-\begin{figure}
-    \centering \includegraphics{transmission/compare-regions}
-    \caption[Comparing each models estimate of regional incidence.]{%
-        Incidence proportion in England and each region estimated by the two models discussed in this chapter.
-        Posterior median and 95\% CrI shown.
-        For the SEIR model, newly PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the backcalculation results.
-    }
-    \label{transmission:fig:compare-regions}
-\end{figure}
-
-These patterns can be interpreted as an instance of the common bias---variance trade-off.
-The SEIR model imposes stronger assumptions about the epidemic dynamics.
-The assumptions reduces the variance of the estimates, including making them smoother, but if the assumptions are incorrect, bias is introduced.
-This trade-off based on the strength of assumptions is well-known to apply in the backcalulcation context~\autocite[e.g.][section 8.3]{brookmeyerBackcalculation}.
-The SEIR model can be viewed as a backcalulation method where the transmisson part of the model is a smoothing method.
-
-The backcalculation method benefits from being able to be implemented in a fast, approximate inference package (R-INLA) that easily scales to many parameters, including random effects.
-Therefore, it is far more flexible and can easily include time-varying effects, ethnicity, etc.
-The SEIR model is limited by the inference procedure and the requirement to sovle an ODE to evaluate the likelihood.
-To incorporate a more flexible model, and reduce its bias, a more efficient inference procedure would be required.
-
-Many of these issues are clearly demonstrated by the estimates for the North East (see \cref{transmission:fig:compare-NE}).
-I previously observed the quality of the SEIR model's fit is variable between different age groups (see \cref{SEIR:fig:prev-young,SEIR:fig:prev-old}).
-This is because the model is not flexible enough to capture the true age distribution of infections.
-The overestimation of prevalence in the 11--16 age group and underestimation in the 16--25 age group is reflected when comparing incidence estimates to those from the backcalculation model, which have reduced bias.
-\begin{figure}
-    \centering \includegraphics{transmission/compare-NE}
-    \caption[Comparing each models estimate of North East incidence by age.]{%
-        Incidence proportion in each age group in the North East estimated by the two models discussed in this chapter.
-        Posterior median and 95\% CrI shown.
-        For the SEIR model, newly PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the backcalculation results.
-    }
-    \label{transmission:fig:compare-NE}
-\end{figure}
+\begin{itemize}
+    \item Results here disagree with previous chapter's results.
+    \item The model in previous chapter is more flexible than here, showing the limitations discussed above. Previous chapter fits better.
+    \item The previous chapter's modelling benefits from being able to use flexible approximate inference methods that scale to many parameters, including random effects.
+Here, I am limited by the inference procedure.
+MCMC was unable to scale to the full model used in the previous chapter, which has a computationally cheaper likelihood as it is a GLM, not requiring solving an ODE.
+    \item Shows that there is a lot of information in the data which could be exploited by a suitably flexible mechanistic model.
+Such a model would be useful for policy-decisions because it could infer where transmission is occurring (\eg in schools or workplaces) and how this changes over time.
+    \item The narrower credible intervals in this chapter are a realisation of the standard bias-variance tradeoff.
+By making stronger assumptions regarding how transmission occurs, the variance is reduced.
+However, if these assumptions are incorrect (which is almost certainly the case), bias occurs.
+\end{itemize}
 

### 2024-01-28 transmission.tex
to dos

--- 2024-01-26 transmission.tex
@@
 
+\todo[inline]{
+    Lit review to go in introduction chapter, framing the whole thesis because it is important in understanding the motivation for the whole work.
+    Would include \textcites{mccabeCISincidence}{abbottCISincidence} on backcalculation. Also Toffe?
+    As well as the literature on mechanistic models using REACT and CIS I have previously circulated.
+}
 \section{Data} \label{transmission:sec:data}
@@
 
-I use the period from the expansion of the CIS, starting 31st August 2020, until there was significant vaccination rollout\todo{insert date}.
+I use the period from the expansion of the CIS, starting Mon 31st August 2020 until Sun 24th Jan 2021 (inclusive).
+This period is chosen because there were very few positives in CIS in summer 2020 due to low prevalence and a smaller sample size, and ends when vaccination crossed 10\% of England's population.
 
@@
 % I discuss the results in \cref{backcalc:sec:discussion} and conclude in \cref{backcalc:sec:conclustion}.
-
-\todo[inline]{
-    I should discuss the literature on estimating incidence based on SARS-CoV-2 prevalence data.
-    In particular: \textcites{mccabeCISincidence}{abbottCISincidence}. Also Toffe.
-    However, I think this goes better in the introduction, framing the whole thesis because it is important in understanding the motivation for the whole work.
-    At the same place, I could review the literature on mechanistic models using REACT and CIS.
-}
 
@@
 Second, at low values of incidence or prevalence, linear growth of $\logit(\pi_i)$ implies approximately exponential in $\pi_i$~\autocite[see discussion in][and references therein]{ealesAppropriately}.
-Approximately exponential growth is what mechanistic models predict, and penalising towards exponential growth has previously been shown to improve estimates of prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
+Penalising towards exponential growth is a natural formulation because it is what mechanistic models predict over short periods of time.
+Furthermore, it has previously been shown to improve estimates of prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
 
@@
 
-\todo{Insert constant incidence assumption methodology}
+For $\vec{P} = \matr{S} \vec{I}$ to have a unique solution, $\vec{I}$ and $\vec{P}$ need the same dimension.
+However, I am using a period of prevalence which starts after the first infections.
+These prior infections will still contribute to the prevalence.
+I assume that the incidence up to and including the first day of prevalence data, at $t = 1$, is constant at $I_0$.
+Therefore, $P_1 = I_0 \sum_{i=1}^{\dmax} S(i)$ and $I_0 = P_1 / \sum_{i=1}^{\dmax} S(i)$.
+For day $t$, the contribution of the incidence for $t < 1$ can be calculated as $P'_t = I_0 \sum_{i=t+1}^{\dmax} S(i)$.
+Subtracting $\vec{P'}$ from $\vec{P}$ before applying the forward solving procedure accounts for the contribution of the incidence prior to the first day of prevalence data.
 
@@
 The final step is poststratifying the incidence proportion to the CIS target population and subpopulations of interest.
-I draw 500\todo{check number of samples} samples from the approximate posterior distribution estimated.
+For this, I first draw 500 samples from the approximate posterior distribution estimated.
 For each posterior draw and day, the incidence proportion in the (sub)population is the weighted average of the incidence proportion in each stratum, where the weights are the proportion of the population in each stratum.
@@
 
-I take information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using pre-existing methodology~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}.
+I take information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using the methodology of \textcites{vanleeuwenTime}{vanleeuwenAugmenting}.
 %; these estimates have been used for previous modelling of the COVID-19 pandemic~\autocites{vanleeuwenTime}{birrellRealtime}.
@@
 This means that there is a wide variety of models used.
-This section  explains the model I propose for fitting to the CIS prevalence data.
+This section  explains the novel modelling approach I propose for fitting to the CIS prevalence data.
 
@@
 
-\Textcite{osogamiClosed} define $P$ as well-matching an arbitrary distribution $G$ if $P$ and $G$ have the same first three moments.
-They then prove that EC distributions have the following properties:
+\Textcite{osogamiClosed} define a phase-type distribution $P$ as well representing an arbitrary distribution $G$ if $P$ and $G$ have the same first three moments.
+They then prove that EC distributions have the following properties.
 \begin{itemize}
-    \item If there exists a phase-type distribution $P'$ that well-matches $G$ then there exists an EC distribution $P$ that well-matches $G$.
-    \item The parameters of $P$ are available in closed form, which \textcite{osogamiClosed} give.
-    \item The number of phases used by the $P$ calculated by these expressions is at most one more than the minimum across all well-matching $P'$.
+    \item Let $\set{W}$ be the set of all phase-types distributions that well represent $G$ and $w$ be the minimum number of phases of any member of $\set{W}$.
+If $\set{W}$ is non-empty, then there exists a $l$-phase EC distribution $P' \in \set{W}$.
+    \item Furthermore, at least one such $P'$ has $l \leq w + 1$. Fewer phases is generally beneficial as it reduces the time and space complexity of working with the distribution.
+    \item The parameters of a $P'$ with $l \leq w + 1$ can be found using a closed-form expression (given by \textcite{osogamiClosed}).
+This is beneficial because finding arbitrary members of $\set{W}$ is a hard problem.
 \end{itemize}

### 2024-01-28 incidence-prevalence.tex
editing

@@
 If the strata have differing transmission patterns, this can change the dynamics of the epidemic.
-The different patterns could be due to behavioural factors (\eg high-numbers of sexual partners in a sexually-transmitted disease~\autocite[69]{keelingModeling} or younger age groups making many more contacts for respiratory diseases~\autocite[176]{andersonInfectious}), or biological factors such as genetics~\autocite[208]{andersonInfectious}..
+The different patterns could be due to behavioural factors (\eg high-numbers of sexual partners in a sexually-transmitted disease~\autocite[69]{keelingModeling} or younger age groups making many more contacts for respiratory diseases~\autocite[176]{andersonInfectious}), or biological factors such as genetics~\autocite[208]{andersonInfectious}.
 I denote the set of strata, as $\set{A}$, with the set's size as $A$.

### 2024-01-28 transmission.tex
editing

@@
 The advantage of fitting a multilevel regression model on prevalence, rather than incidence, is that prevalence with appropriate prior (or penalisation) can be expressed as a generalised linear model (GLM)\todo{maybe cite backcalculation as a GLM}.
-Either incidence proportion or prevalence, $pi_i$ in a stratum $i$ can be expressed as $\logit(\pi_i) = g(i, t)$, where is a function of the stratum and time, penalised towards being linear in time.
+Either incidence proportion or prevalence, $\pi_i$ in a stratum $i$ can be expressed as $\logit(\pi_i) = g(i, t)$, where is a function of the stratum and time, penalised towards being linear in time.
 If $g$ is chosen appropriately, this is a GLM with a logit link function.
@@
 
-Having fit the prevalence model, I draw 10,000\todo{check number} prevalence trajectories in each stratum from the approximate posterior distribution of the model.
+Having fit the prevalence model, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution.
 For each stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
@@
 \end{align}
-Since the matrix is lower-triangular, the solution for $\vec{P}$ can be found efficiently using forward substitution\todo{cite forward substitution}.
-Forward substitution is numerically stable unless the diagonal elements are close to 0.
-Here, the diagonal is $S(1) = 1$ and hence the matrix is well-conditioned.
-\todo{I explained this more clearly in my first-year report, look that up!}
+$\matr{S}$ is lower-triangular, hence to solve for $\vec{P}$ I use forward-substitution~\autocite{cormenMatrix}.
+Forward substitution is numerically stable unless the diagonal elements, here $S(1)$, are small relative to the off-diagonal elements, here $S(i)$ for $i > 0$~\autocite[section 8.2]{highamAccuracy}.
+Here, the diagonal is $S(1) = 1$ and hence the matrix the solution is numerically stable.
 
@@
 The final step is poststratifying the incidence proportion to the CIS target population and subpopulations of interest.
-For this, I first draw 500 samples from the approximate posterior distribution estimated.
+For this, I start with the 500 samples of the approximate posterior distribution of the incidence.
 For each posterior draw and day, the incidence proportion in the (sub)population is the weighted average of the incidence proportion in each stratum, where the weights are the proportion of the population in each stratum.
@@
 During November, incidence starts rising, although starting at different times in different regions.
-This could be due to the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country\todo{cite Alpha variant}.
+This could be due to the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocites{walkerTracking}{daviesEstimated}{lythgoeLineage}.
 A second peak occurs in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
@@
 \subsection{Methods}
-\todo[inline]{add a bit more intro in the context of chapter 3}
-The model will be stratified only by region and age group, so the data are aggregated to these levels.
-Only individuals aged two or older are eligible to participate in the CIS; I assume that test results in all participants aged under 11 are still representative of the whole $[0, 11)$ age group.
 
+% The model will be stratified only by region and age group, so the data are aggregated to these levels.
+% Only individuals aged two or older are eligible to participate in the CIS; I assume that test results in all participants aged under 11 are still representative of the whole $[0, 11)$ age group.
+
+I model each region fully independently, \ie with no shared parameters or transmission between regions.
 As well as the motivation given in \cref{E-backcalc}, independent regions allows heterogeneity in the parameters between regions, which could arise from a range of socioeconomic factors.
@@
 A parsimonious model is needed to parameterise the WAIFW matrix (see \cref{E-SEIR:sec:structured-populations}).
-I follow the WAIFW matrix parameterisation of \textcite{birrellRealtime}, decomposition $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'} N_{a'}$.\todo{remind readers what these quantities are here}
+I follow the WAIFW matrix parameterisation of \textcite{birrellRealtime}, decomposing the rate of transmission from individuals in age group $a'$ to susceptible individuals in age group $a$ as $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'} N_{a'}$.
 This separates contact patterns, $\kappa_{aa'}$, which can vary over time; transmission probability upon contact between age groups $a$ and $a'$, $q_{aa'}$, assumed constant; a region-wide effect for time, $\beta(t)$; and the population of the stratum, $N_{a'}$.
@@
 However, except in London, the credible intervals still overlap.
-This is a specific case of a larger pattern: the SEIR model produces smoother estimates of incidence than the backcalculation model.
+This is a specific case of a larger pattern: the SEIR model produces smoother estimates of incidence than the backcalculation model.;
 \begin{figure}
@@
 
-\todo[inline]{write conclusion and add to discussion once comparison with backcalculation is made}
+The major contribution in this chapter is showing that transmission and incidence can be estimated using only a
 

### 2024-01-29 transmission.tex
Updates from Overleaf

--- 2024-01-28 transmission.tex
@@
 In this chapter, I use the preferred estimate of the duration of PCR positivity (see \cref{E-imperf-test:sec:conclusion}) to estimate the transmission of SARS-CoV-2 in England.
-This estimation is based on the CIS data (see \cref{transmission:sec:data} for further details).
-Estimating transmission is the ultimate aim of this thesis.
+This estimation is based only on the CIS data (see \cref{transmission:sec:data} for further details).
+I will show that this data is sufficient to estimate the parameters of the transmission process, suggesting surveys of this type are sufficient for surveillance of an epidemic.
 
@@
 First, in \cref{backcalc}, I use a backcalculation approach, based on \cref{E-inc-prev:eq:EPt}.
+This approach enables estimation of incidence within (sub)populations of England.
 Then, in \cref{SEIR}, I use a mechanistic compartmental model, an implementation of the model introduced in \cref{E-SEIR:sec:mechanistic-models}.
+In addition to estimating incidence, this approach also estimates the components of transmission, \eg whether changes can be explained by changes in contact rates.
 Finally, in \cref{transmission:sec:comparison}, I compare the results of the two approaches.
 
-\todo[inline]{
-    Lit review to go in introduction chapter, framing the whole thesis because it is important in understanding the motivation for the whole work.
-    Would include \textcites{mccabeCISincidence}{abbottCISincidence} on backcalculation. Also Toffe?
-    As well as the literature on mechanistic models using REACT and CIS I have previously circulated.
-}
 \section{Data} \label{transmission:sec:data}
 
-\todo[inline]{Consider moving all of this section into chapter 2, where CIS is introduced}
+\todo[inline]{Consider moving all of this section into chapter 2, where CIS is introduced, and spell it out a bit more in that context}
 
@@
 
-I use the period from the expansion of the CIS, starting Mon 31st August 2020 until Sun 24th Jan 2021 (inclusive).
-This period is chosen because there were very few positives in CIS in summer 2020 due to low prevalence and a smaller sample size, and ends when vaccination crossed 10\% of England's population.
+I use the period from the expansion of the CIS, starting 31st August 2020, until there was significant vaccination rollout\todo{insert date}.
 
@@
 % I discuss the results in \cref{backcalc:sec:discussion} and conclude in \cref{backcalc:sec:conclustion}.
+
+\todo[inline]{
+    I should discuss the literature on estimating incidence based on SARS-CoV-2 prevalence data.
+    In particular: \textcites{mccabeCISincidence}{abbottCISincidence}. Also Toffe.
+    However, I think this goes better in the introduction, framing the whole thesis because it is important in understanding the motivation for the whole work.
+    At the same place, I could review the literature on mechanistic models using REACT and CIS.
+}
 
@@
 Then, the posterior distribution is used to predict the prevalence in each stratum.
-Finally, the prevalence is poststratified to the CIS target population and subpopulations of interest.
+Finally, the prevalence is poststratified (as explained later) to the CIS target population and subpopulations of interest.
 I modify this process by transforming the posterior distribution of prevalence in each stratum to a posterior distribution of incidence before poststratifying.
@@
 Second, at low values of incidence or prevalence, linear growth of $\logit(\pi_i)$ implies approximately exponential in $\pi_i$~\autocite[see discussion in][and references therein]{ealesAppropriately}.
-Penalising towards exponential growth is a natural formulation because it is what mechanistic models predict over short periods of time.
-Furthermore, it has previously been shown to improve estimates of prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
+Approximately exponential growth is what mechanistic models predict, and penalising towards exponential growth has previously been shown to improve estimates of prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
 
@@
 
-For $\vec{P} = \matr{S} \vec{I}$ to have a unique solution, $\vec{I}$ and $\vec{P}$ need the same dimension.
-However, I am using a period of prevalence which starts after the first infections.
-These prior infections will still contribute to the prevalence.
-I assume that the incidence up to and including the first day of prevalence data, at $t = 1$, is constant at $I_0$.
-Therefore, $P_1 = I_0 \sum_{i=1}^{\dmax} S(i)$ and $I_0 = P_1 / \sum_{i=1}^{\dmax} S(i)$.
-For day $t$, the contribution of the incidence for $t < 1$ can be calculated as $P'_t = I_0 \sum_{i=t+1}^{\dmax} S(i)$.
-Subtracting $\vec{P'}$ from $\vec{P}$ before applying the forward solving procedure accounts for the contribution of the incidence prior to the first day of prevalence data.
+\todo{Insert constant incidence assumption methodology}
 
@@
 The final step is poststratifying the incidence proportion to the CIS target population and subpopulations of interest.
-For this, I first draw 500 samples from the approximate posterior distribution estimated.
+I draw 500\todo{check number of samples} samples from the approximate posterior distribution estimated.
 For each posterior draw and day, the incidence proportion in the (sub)population is the weighted average of the incidence proportion in each stratum, where the weights are the proportion of the population in each stratum.
@@
 
-I take information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using the methodology of \textcites{vanleeuwenTime}{vanleeuwenAugmenting}.
+I take information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using pre-existing methodology~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}.
 %; these estimates have been used for previous modelling of the COVID-19 pandemic~\autocites{vanleeuwenTime}{birrellRealtime}.
@@
 This means that there is a wide variety of models used.
-This section  explains the novel modelling approach I propose for fitting to the CIS prevalence data.
+This section  explains the model I propose for fitting to the CIS prevalence data.
 
@@
 
-\Textcite{osogamiClosed} define a phase-type distribution $P$ as well representing an arbitrary distribution $G$ if $P$ and $G$ have the same first three moments.
-They then prove that EC distributions have the following properties.
+\Textcite{osogamiClosed} define $P$ as well-matching an arbitrary distribution $G$ if $P$ and $G$ have the same first three moments.
+They then prove that EC distributions have the following properties:
 \begin{itemize}
-    \item Let $\set{W}$ be the set of all phase-types distributions that well represent $G$ and $w$ be the minimum number of phases of any member of $\set{W}$.
-If $\set{W}$ is non-empty, then there exists a $l$-phase EC distribution $P' \in \set{W}$.
-    \item Furthermore, at least one such $P'$ has $l \leq w + 1$. Fewer phases is generally beneficial as it reduces the time and space complexity of working with the distribution.
-    \item The parameters of a $P'$ with $l \leq w + 1$ can be found using a closed-form expression (given by \textcite{osogamiClosed}).
-This is beneficial because finding arbitrary members of $\set{W}$ is a hard problem.
+    \item If there exists a phase-type distribution $P'$ that well-matches $G$ then there exists an EC distribution $P$ that well-matches $G$.
+    \item The parameters of $P$ are available in closed form, which \textcite{osogamiClosed} give.
+    \item The number of phases used by the $P$ calculated by these expressions is at most one more than the minimum across all well-matching $P'$.
 \end{itemize}

### 2024-01-29 transmission.tex

--- 2024-01-28 transmission.tex
@@
 First, in \cref{backcalc}, I use a backcalculation approach, based on \cref{E-inc-prev:eq:EPt}.
+This approach enables estimation of incidence within (sub)populations of England.
 Then, in \cref{SEIR}, I use a mechanistic compartmental model, an implementation of the model introduced in \cref{E-SEIR:sec:mechanistic-models}.
+In addition to estimating incidence, this approach also estimates the components of transmission, \eg whether changes can be explained by changes in contact rates.
 Finally, in \cref{transmission:sec:comparison}, I compare the results of the two approaches.
@@
 
-\todo[inline]{Consider moving all of this section into chapter 2, where CIS is introduced}
+\todo[inline]{Consider moving all of this section into chapter 2, where CIS is introduced, and spell it out a bit more in that context}
 

### 2024-01-30 transmission.tex
Start conclusion

--- 2024-01-28 transmission.tex
@@
 
-The major contribution in this chapter is showing that transmission and incidence can be estimated using only a
+The major contribution in this chapter is showing that transmission and incidence can be estimated using only data from the CIS.
+A similar survey could, therefore, be considered as the primary form of surveillance for a future pandemic.
+The advantage of such a setup is the unbiased information across the entire population, including children.
+
+The transmission model presented in this chapter could be extended to fully extract the information in the CIS (see \cref{SEIR:sec:discussion}).
+This would require a more flexible model, which would likely require a more efficient inference procedure, such as NUTS.
 

### 2024-01-30 transmission.tex
Proof-reading

@@
 
+I will model the test results as overdispersed relative to a binomial model.
+Overdispersion allows for household clustering\todo{cite beta-binomial as a model for clustering} and other overdispersion in the data, which is common in epidemiological data\todo{cite that epi data is often overdispersed}.
+The parameterisation of the beta-binomial that I use (detailed in \cref{E-distributions}) has an overdispersion parameter $\rho$, regarded as a nuisance parameter in this context.
+$\rho=0$ means the beta-binomial coincides with the binomial distribution.
+As $\rho \to \infty$, the distribution tends towards the uniform distribution on $[0, 1]$~\autocite{hughesUsing}.
+
 \section{Backcalculation} \label{backcalc}
@@
 
-\todo[inline]{Think about the framing of this paragraph, maybe talk about being linear in its parameters and pre-existing software packages.}
-The advantage of fitting a multilevel regression model on prevalence, rather than incidence, is that prevalence with appropriate prior (or penalisation) can be expressed as a generalised linear model (GLM)\todo{maybe cite backcalculation as a GLM}.
-Either incidence proportion or prevalence, $\pi_i$ in a stratum $i$ can be expressed as $\logit(\pi_i) = g(i, t)$, where is a function of the stratum and time, penalised towards being linear in time.
+\todo[inline]{I am struggling the framing of these paragraphs. Maybe talk about being linear in its parameters and pre-existing software packages and not mention GLMs?}
+The advantage of fitting a multilevel regression model on prevalence, rather than incidence, is that prevalence with appropriate prior (or penalisation) can be expressed as a generalised linear model (GLM)\todo{maybe cite backcalculation as a GLM} or Gaussian Markov Random Field (GMRF).
+Standard, efficient, and widely-used software can then be applied for inference.
+
+To use a GLM or GMRF, express incidence proportion or prevalence in a stratum $i$ on day $t$, $\pi_{it}$, as $\logit(\pi_{it}) = g(i, t)$, where is a function of the stratum and time, penalised towards being linear in time.
 If $g$ is chosen appropriately, this is a GLM with a logit link function.
@@
 First, it ensures that the incidence proportion or prevalence is between 0 and 1.
-Second, at low values of incidence or prevalence, linear growth of $\logit(\pi_i)$ implies approximately exponential in $\pi_i$~\autocite[see discussion in][and references therein]{ealesAppropriately}.
-Penalising towards exponential growth is a natural formulation because it is what mechanistic models predict over short periods of time.
-Furthermore, it has previously been shown to improve estimates of prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
+Second, at low values of incidence or prevalence, linear growth of $\logit(\pi_{it})$ over time implies approximately exponential growth of $\pi_{it}$~\autocite[see discussion in][and references therein]{ealesAppropriately}.
+Penalising towards exponential growth is a natural formulation because mechanistic models predict expoential growth over short periods of time.
+Furthermore, such penalisation has previously been shown to improve estimates of prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
 
-However, if the incidence proportion is expressed as a GLM in this way, the prevalence is not a GLM.
+However, if the incidence proportion is expressed as a GLM in this way, the prevalence is not expressible as a GLM.
+Specifically, the prevalence (or an element-wise transformation of it) is no longer linear in the model parameters.
 The relationship between incidence and prevalence is linear, as derived in \cref{E-inc-prev:sec:prevalence-process}.
@@
     \item $l_{a,t}$ is an age-group specific second-order random walk. That is $l_{a,t} = 2l_{a,t-1} - l_{a,t-2} + \epsilon_{a,t}$ where $\epsilon_{at} \dist \N(0, \sigma_l^2)$.
-    \item The likelihood is a beta-binomial model (defined in \cref{E-distributions}), an overdispersed binomial.
-Overdispersion allows for household clustering\todo{cite beta-binomial as a model for clustering} and other overdispersion in the data, which is common in epidemiological data\todo{cite that epi data is often overdispersed}.
+    \item $n_{taes}$ and $y_{taes}$ are the number of individuals tested and the number of positive test results on day $t$ in individuals of age group $a$, ethnicity $e$, and sex $s$.
+The likelihood is a beta-binomial model (see discussion in \cref{transmission:sec:data}).
+an overdispersed binomial.
 \end{itemize}
@@
 For each stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
-I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:prevalence-process} justifies, that is $P_t = \E(P_t \mid Z)$.
+I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:observation-process} justifies, that is $P_t = \E(P_t \mid \vec{Z})$.
 \Cref{E-inc-prev:eq:EPt}, considered across different values of $t$, can be written as $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix formed from the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
+Dividing each side by the total number of individuals in each stratum transforms number of prevalent individuals to prevalence and incidence to incidence proportion.
+
 The matrix $\matr{S}$ is defined to maintain the relationship between incidence and prevalence.
@@
 $\matr{S}$ is lower-triangular, hence to solve for $\vec{P}$ I use forward-substitution~\autocite{cormenMatrix}.
-Forward substitution is numerically stable unless the diagonal elements, here $S(1)$, are small relative to the off-diagonal elements, here $S(i)$ for $i > 0$~\autocite[section 8.2]{highamAccuracy}.
-Here, the diagonal is $S(1) = 1$ and hence the matrix the solution is numerically stable.
+% Forward substitution is numerically stable unless the diagonal elements, here $S(1)$, are small relative to the off-diagonal elements, here $S(i)$ for $i > 0$~\autocite[section 8.2]{highamAccuracy}.
+% Here, the diagonal is $S(1) = 1$ and hence the matrix the solution is numerically stable.
 
@@
 Model misspecification means that the procedure described could produce negative incidence.
-For any stratum and day $t$, this occurs if the if the prevalence reduces between day $t$ and $t+1$ more than the number of recoveries on day $t$, based on the incidence up until day $t$ and the survival function.
+\todo{This paragraph might be too terse.}
+For any stratum and day $t$, this occurs if the prevalence reduces between day $t$ and $t+1$ more than the number of recoveries on day $t$, based on the incidence up until day $t$ and the survival function.
 
@@
 For example, \cref{E-imperf-test} showed that it is sensitive to assumptions made.
-The second is the model used to estimate prevalence, for example the form of the smoothing used.
+The second is the specification of the statistical model used to estimate prevalence, for example the form of the smoothing used.
 
+\todo[inline]{Maybe relate following paragraph to the GLM discussion at the start of the section - its basically the same point}
 A prior constraining incidence to be non-negative and then fit directly to the data could be used, but this could no longer be fit within the R-INLA framework.
@@
 An exception is the North East and East of England regions (see \cref{transmission:fig:backcalc-regions}(C) and (D) respectively); these regions show no consistent increase or decrease in incidence during October.
-All regions then decrease during November, coinciding with the national lockdown.
+All regions then decrease during November; this coincides with the national lockdown which came into force in England on 5th Nov 2020~\autocite{ifgTimeline}.
 
 During November, incidence starts rising, although starting at different times in different regions.
-This could be due to the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocites{walkerTracking}{daviesEstimated}{lythgoeLineage}.
+The most explored explanation for this is the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocites{walkerTracking}{daviesEstimated}{lythgoeLineage}.
 A second peak occurs in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
@@
 Except the 16--25 age group, all age groups show similar trends (\ie parallel lines on a log scale).
-This is the pattern predicted by an age-stratified transmission model (see \cref{E-SEIR:sec:structured-populations}), if behavioural changes are independent of age group, and there is negligible impact of immunity.
+This is the pattern predicted by an age-stratified, mechanistic transmission model (such as the one introduced in \cref{E-SEIR:sec:structured-populations}), if behavioural changes are independent of age group, and there is negligible impact of immunity.
 The incidence proportion in the 16--25 age group decreases faster than other age groups during November.
@@
 There are various advantages to using a mechanistic model to estimate transmission (see \cref{E-inc-prev:sec:infection-process}).
-This section uses a mechanistic model (introduced in \cref{E-SEIR:sec:mechanistic-models}) to estimate the transmission of SARS-CoV-2 in England.
+This section applies the mechanistic modelling framework introduced in \cref{E-SEIR:sec:mechanistic-models} to estimate the transmission of SARS-CoV-2 in England.
 
@@
 % These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
-This method takes pre-pandemic baseline estimates of: $\kappa_{aa'}$ from the POLYMOD study~\autocite{mossongSocial}, and time use from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
+This method starts with pre-pandemic baseline estimates of: $\kappa_{aa'}$ from the POLYMOD study~\autocite{mossongSocial}, and time use from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
 POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on each contact's age and location (home, work, school, leisure, transport, or other).
@@
 One approach would be to convolve the incidence based on the transmission model with the PCR duration distribution.
-Here, the daily incidence proportion for each stratum is by solving the system of ODEs governing the transmission model, and then prevalence calculated as in \cref{backcalc}.
-However, this approach is computationally expensive because of the need to calculate the convolution in \cref{E-inc-prev:eq:EPt}.
+Here, the daily incidence proportion for each stratum is by solving the system of ODEs governing the transmission model, and then prevalence calculated as using \cref{E-inc-prev:eq:EPt}.
+However, this approach is computationally expensive, likely prohibitively so if performed within a MCMC loop.
 
-A computationally cheaper option is to approximate this convolution by adding compartments to the ODEs.
+A computationally cheaper option, which I will take, is to approximate this convolution by adding compartments to the ODEs.
 This approach allows a infected, pre-positive compartment, which I denote $p_0$, to be included simply.
@@
 It is shown schematically in \cref{SEIR:fig:EC}.
-Here, $K \dist \GamDist(l-2, \kappa_Y)$, $X_1 \dist \Exponential(\kappa_{X1})$ and $X_2$ is a mixture distribution taking the value 0 with probability $1-p_x$ or otherwise (\ie with probability $p_x$) distributed $\Exponential(\kappa_{X2})$.
+Here, $K \dist \GamDist(l-2, \omega_Y)$, $X_1 \dist \Exponential(\omega_{X1})$ and $X_2$ is a mixture distribution taking the value 0 with probability $1-p_x$ or otherwise (\ie with probability $p_x$) distributed $\Exponential(\omega_{X2})$.
 % The time between entering the 1st state and the $l+1$th state, the absorbing state, is the duration.
 These distributions contain five parameters.
-Three are positive reals, $\kappa_Y$, $\kappa_{X1}$, and $\kappa_{X2}$; one is a probability, $p_x$, in the interval $[0, 1]$; and one is a positive integer, $l$.
+Three are positive reals, $\omega_Y$, $\omega_{X1}$, and $\omega_{X2}$; one is a probability, $p_x$, in the interval $[0, 1]$; and one is a positive integer, $l$.
 
@@
 
-    \path (1) edge node {$\kappa_Y$} (2)
-          (2) edge node {$\kappa_Y$} (dots)
-          (dots) edge node {$\kappa_Y$} (lm2)
-          (lm2) edge node {$\kappa_Y$} (lm1)
-          (lm1) edge node {$p_x \kappa_{X1}$} (l)
-                edge [out=-45,in=-135] node[below] {$(1 - p_x) \kappa_{X1}$} (absorbing)
-          (l) edge node {$\kappa_{X2}$} (absorbing);
+    \path (1) edge node {$\omega_Y$} (2)
+          (2) edge node {$\omega_Y$} (dots)
+          (dots) edge node {$\omega_Y$} (lm2)
+          (lm2) edge node {$\omega_Y$} (lm1)
+          (lm1) edge node {$p_x \omega_{X1}$} (l)
+                edge [out=-45,in=-135] node[below] {$(1 - p_x) \omega_{X1}$} (absorbing)
+          (l) edge node {$\omega_{X2}$} (absorbing);
 \end{tikzpicture}
@@
 The distribution is the time from entering state 1 and entering the absorbing state.
-The time between entering state 1 and state $l-2$ is distributed $K \dist \GamDist(l-2, \kappa_Y)$, also known as an Erlang distribution.
+The time between entering state 1 and state $l-2$ is distributed $K \dist \GamDist(l-2, \omega_Y)$, also known as an Erlang distribution.
 The time between entering state $l-2$ and the absorbing state has a Coxian distribution (see the main text for details).}
@@
     \item $l$, the number of phases.
-    \item $\kappa_Y$, the rate parameter of the Erlang distribution.
-    \item $\kappa_{X1}$, the rate parameter of the first state after the Erlang distribution.
-    \item $\kappa_{X2}$, the rate parameter of the second state after the Erlang distribution.
+    \item $\omega_Y$, the rate parameter of the Erlang distribution.
+    \item $\omega_{X1}$, the rate parameter of the first state after the Erlang distribution.
+    \item $\omega_{X2}$, the rate parameter of the second state after the Erlang distribution.
     \item $p_x$, the probability of moving from the end state of the Erlang distribution to the second (as opposed to directly to the absorbing state).
@@
 
-I use the mapfit package~\autocite{mapfit} to compute $P$ for the posterior mean of the distribution estimated in \cref{E-imperf-test}.
-The resulting parameters for the EC distribution are in \cref{SEIR:table:ec-params}.
+The mapfit package~\autocite{mapfit} implements the expressions to calculate the EC distribution parameters given a distribution's first three moments.
+Applying the package to the posterior mean of the distribution estimated in \cref{E-imperf-test} gives the parameter estimates in \cref{SEIR:table:ec-params}.
 \begin{table}
@@
     \begin{tabular}{c c c c c}
-        $l$ & $\kappa_Y$ & $\kappa_{X1}$ & $\kappa_{X2}$ & $p_x$ \\
+        $l$ & $\omega_Y$ & $\omega_{X1}$ & $\omega_{X2}$ & $p_x$ \\
         3 & 0.0820 & 0.126 & 0.0223 & 0.00973  \\
@@
 
-I link the modelled proportion of individuals that are PCR positive to the data using a beta-binomial likelihood (see\todo{ref section on clustering} for the issues with using a binomial likelihood).
+I link the modelled proportion of individuals that are PCR positive to the data using a beta-binomial likelihood (as discussed in \cref{transmission:sec:data}).
 If on each day $t$, each strata $i$ is observed to have $y_{it}$ positives out of $n_{it}$ tests, then the likelihood is:
@@
 where $\bar{p}_{kat} = \int_{t}^{t+1} p_{ka}(s) ds$ is the mean occupancy in compartment $p_k$ and strata $a$ over day $t$ and $\rho$ is a nuisance parameter controlling the overdispersion of the beta-binomial distribution.
-In this parameterisation (detailed in \cref{E-distributions}), $\rho=0$ means the beta-binomial coincides with the binomial distribution.
-As $\rho \to \infty$, the distribution tends towards the uniform distribution on $[0, 1]$~\autocite{hughesUsing}.
 
@@
     \path  (S) edge node {$\lambda$} (P0)
-          (P0) edge node {$\kappa_0$} (P1)
-          (P1) edge node {$\kappa_Y$} (P2)
-          (P2) edge node {$p_x \kappa_{X1}$} (P3)
-               edge [out=-45,in=-135] node[below] {$(1 - p_x) \kappa_{X1}$} (P4)
-          (P3) edge node {$\kappa_{X2}$} (P4);
+          (P0) edge node {$\omega_0$} (P1)
+          (P1) edge node {$\omega_Y$} (P2)
+          (P2) edge node {$p_x \omega_{X1}$} (P3)
+               edge [out=-45,in=-135] node[below] {$(1 - p_x) \omega_{X1}$} (P4)
+          (P3) edge node {$\omega_{X2}$} (P4);
 
@@
 The transmission model is a SEIR model with two latent and two infectious states.
-To form whole of England estimates, each region is modelled independently with the results aggregated, as in \cref{E-backcalc}.
+To form whole of England estimates, each region is modelled independently with the results aggregated, using the poststratification methodology explained in \cref{E-backcalc}.
 
@@
     \frac{d\vec{r}(t)}{dt} &= 2\gamma \vec{i_2}(t) \\
-    \frac{d\vec{p_0}(t)}{dt} &= \vec{\lambda} \circ \vec{s}(t) - \kappa_0 \vec{p_0}(t) \\
-    \frac{d\vec{p_1}(t)}{dt} &= \kappa_0 \vec{p_0}(t) - \kappa_Y \vec{p_1}(t) \\
-    \frac{d\vec{p_2}(t)}{dt} &= \kappa_Y \vec{p_1}(t) - \kappa_{X1} \vec{p_2}(t) \\
-    \frac{d\vec{p_3}(t)}{dt} &= p_x \kappa_{X1} \vec{p_2}(t) - \kappa_{X2} \vec{p_3}(t).
+    \frac{d\vec{p_0}(t)}{dt} &= \vec{\lambda} \circ \vec{s}(t) - \omega_0 \vec{p_0}(t) \\
+    \frac{d\vec{p_1}(t)}{dt} &= \omega_0 \vec{p_0}(t) - \omega_Y \vec{p_1}(t) \\
+    \frac{d\vec{p_2}(t)}{dt} &= \omega_Y \vec{p_1}(t) - \omega_{X1} \vec{p_2}(t) \\
+    \frac{d\vec{p_3}(t)}{dt} &= p_x \omega_{X1} \vec{p_2}(t) - \omega_{X2} \vec{p_3}(t).
 \end{align}
@@
 Priors for the parameters are given in \cref{SEIR:table:priors}.
-The parameters governing the duration of positivity ($\kappa$s) were explained previously (see \cref{SEIR:table:ec-params}).
+The parameters governing the duration of positivity ($\omega$s) were explained previously (see \cref{SEIR:table:ec-params}).
 The initial conditions and transition rate parameters are unindentifiable in a SEIR model~\autocite{dankwaStructural}, therefore strongly informative priors are used for them.
@@
 
-Furthermore, $\R$ is a function of $\matr{\beta}(0)$, and hence $\beta_1$ (as explained in \cref{E-SEIR:sec:structured-populations}).
-Combining these expressions, the model can be parameterised by using $\psi$ in place of $\beta_1$.
-The following procedure gives $\beta_1$ in terms of $\psi$.
-First, calculate $\R$ from $\psi$ using \cref{SEIR:eq:rtoR}; denote this $\hat{R}$.
-Then calculate $R^* = D(\matr{\beta}(0)) / \gamma$ when $\beta_1=1$ (recall that $D(\matr{\beta})$ denotes the dominant eigenvalue of $\matr{\beta}$).
-Multiplying a matrix by a scalar multiplies its dominant eigenvalue by that same scalar, therefore, setting $\beta_1 = \hat{R} / R^*$ gives means that $\R$ at time 0 is $\hat{R}$, as desired.
+$\R$ can also be expressed a function of $\matr{\beta}(0)$, and hence $\beta_1$ (as explained in \cref{E-SEIR:sec:structured-populations}).
+Combining these expressions, the model can be parameterised by using $\psi$ in place of $\beta_1$ by using the following procedure.
+Start by calculating $\R$ from $\psi$ using \cref{SEIR:eq:rtoR}; denote this $\hat{R}$.
+$\hat{R}$ is the $\R$ that the current value of $\psi$ implies.
+Hence, $\beta_1$ should be set such that $\matr{\beta}(0)$ implies $\R = \hat{R}$.
+To do so, define $R^*$ as $\R$ $\beta_1 = 1$.
+Using the results of \cref{E-SEIR:sec:structured-populations}, $R^* = D(\matr{\beta}(0)) / \gamma$, computed using $\beta_1=1$ (recall that $D(\matr{\beta})$ denotes the dominant eigenvalue of $\matr{\beta}$).
+Finally, set $\beta_1 = \hat{R} / R^*$; as multiplying a matrix by a scalar also multiplies its dominant eigenvalue by that same scalar, this means that $\R$ at time 0 is $\hat{R}$, as desired.
 
 To form a parsimonious parameterisation of the initial state, first focus on the initial proportion of the population that is in the infectious compartments.
-Recall that in the exponential growth phase, the relative proportion of each strata that is infectious is proportional to the dominant eigenvector of $\matr{\beta}$, which I denote $\vec{d}$.
+Recall, from \cref{E-SEIR:sec:structured-populations}, that in the exponential growth phase, the relative proportion of each stratum that is infectious is proportional to the dominant eigenvector of $\matr{\beta}$, which I denote $\vec{d}$.
 Introduce a model parameter $i^+$, the proportion of the population in the $i$ states at time 0 in the strata with the highest proportion infectious.
-Then, $\vec{i_0} = \vec{i_1} + \vec{i_2} = \frac{i^+}{\max \vec{d}} \vec{d}$.
-Now introduce an additional vector of parameters, $\vec{\pi}$, where $\pi_a$ is the proportion of strata $i$ without immunity at time 0, hence, $\vec{r} = 1 - \vec{\pi}$.
+Define the total number of infectious individuals at time 0 as $\vec{i_0} = \vec{i_1} + \vec{i_2} = \frac{i^+}{\max \vec{d}} \vec{d}$.
 
-From the system of ODEs and the knowledge that all the compartments are growing exponentially at rate $\psi$, one can derive the following for the initial state of the compartments:
+From the system of ODEs, the knowledge that all the compartments are growing exponentially at rate $\psi$, and the parameters $\vec{\pi}$ (where $\pi_a$ is the proportion of strata $i$ without immunity at time 0, hence, $\vec{r} = 1 - \vec{\pi}$) one can derive the following for the initial state of the compartments:
 \begin{align}
@@
     \vec{s} &= \vec{\pi} - \vec{i_0} - \vec{e_1} - \vec{e_2} \\
-    \vec{p_0} &= \vec{e_1} \frac{\psi + 2\sigma}{\psi + \kappa_0} \\
-    \vec{p_1} &= \vec{p_0} \frac{\kappa_0}{\psi + \kappa_Y} \\
-    \vec{p_2} &= \vec{p_1} \frac{\kappa_Y}{\psi + \kappa_{X1}} \\
-    \vec{p_3} &= \vec{p_2} \frac{p_x \kappa_{X1}}{\psi + \kappa_{X2}}.
+    \vec{p_0} &= \vec{e_1} \frac{\psi + 2\sigma}{\psi + \omega_0} \\
+    \vec{p_1} &= \vec{p_0} \frac{\omega_0}{\psi + \omega_Y} \\
+    \vec{p_2} &= \vec{p_1} \frac{\omega_Y}{\psi + \omega_{X1}} \\
+    \vec{p_3} &= \vec{p_2} \frac{p_x \omega_{X1}}{\psi + \omega_{X2}}.
 \end{align}

### 2024-02-01 biology-data.tex
most of bio text

@@
 
-\chapter{SARS-CoV-2: biology and data} \label{intro:sec:studies}
+\chapter{SARS-CoV-2: biology and data} \label{biology-data}
+
+SARS-CoV-2 is a beta coronavirus that emerged in late 2019.
+It is closely related to previous coronaviruses that have caused outbreaks in humans, SARS-CoV and MERS-CoV.
+\todo{Cite and check how SARS-CoV-2 is related to SARS-CoV and MERS-CoV, and other human coronaviruses.}
+
+Coronaviruses are RNA\todo{spell out the acronym RNA} viruses.
+RNA viruses use RNA, rather than DNA, as their genetic material.
+\todo{why is being a RNA virus important?}
 
@@
   \caption[Natural history of SARS-CoV-2.]{Natural history of SARS-CoV-2. Reproduced from \textcite{puhachSARSCoV2} with permission.}
+  \todo[inline]{Define the acronyms and terminology in the figure.}
   \label{biology-data:fig:natural-history}
@@
 
-Possible summaries
-\begin{enumerate}
-  \item \url{https://www.nature.com/articles/s41579-022-00822-w/figures/2}
-  \item \url{https://www.nature.com/articles/s41576-021-00360-w/figures/1}
-  \item \url{https://academic.oup.com/view-large/figure/305617142/ciaa1442_fig2.jpg}
-\end{enumerate}
+An individual infected with SARS-CoV-2 goes through several stages (see \cref{biology-data:fig:natural-history}).
+\todo{Find citations for disease stages.}
+First, they are exposed to the virus and become infected.
+At this initial stage they are not infectious (cannot spread the disease to others), do not have symptoms, and will not test positive on a PCR test.
+After a few days, they will start testing positive on a PCR test; this is normally prior to being infectious or having symptoms.
+Next, the individual becomes infectious; normally, this occurs prior to symptom onset for those who experience symptoms.
+Recovery from symptoms occurs after a few days, around individual is no longer infectious.
+Finally, the individual will test negative on a PCR test.
+\todo{Talk about PCR testing detecting RNA material even with no viable virus.}
 
-\begin{itemize}
-  \item Phases: pre-infection positive, infectious positive, post-infectious positive, negative
-  \item Relationship between different means of infected: infectious, PCR positive, symptomatic/diseased
-\end{itemize}
+The time for each of these stages depends on various features of host and virus characteristics.
+In this thesis, I focus on pre-Alpha variants of SARS-CoV-2 in unvaccinated individuals.
+The time from infection to infectiousness is known as the \emph{latent period} and is typically around\todo{cite latent period}.
+The time from infection to experiencing symptoms is known as the \emph{incubation period} and is typically around\todo{cite incubation period}.
+The latent period is undefined for asymptomatic individuals, around\todo{cite proportion asymptomatic} of those infected.
+
+The above description gives three possible ways to define an individual who is infected with SARS-CoV-2: based on their PCR test result, whether they are infectious, or whether they are symptomatic.
+In this thesis, I adopt a definition of infection based on PCR test results.
+This is because it coincides with the data collected on the studies I use (see \cref{biology-data:sec:studies}).
+
+\todo[inline]{Add something about viral load being the process underlying these processes.}
+
+% Possible summaries
+% \begin{enumerate}
+%   \item \url{https://www.nature.com/articles/s41579-022-00822-w/figures/2}
+%   \item \url{https://www.nature.com/articles/s41576-021-00360-w/figures/1}
+%   \item \url{https://academic.oup.com/view-large/figure/305617142/ciaa1442_fig2.jpg}
+% \end{enumerate}
 
@@
 
-\emph{Describe how PCR testing works}
+\todo[inline]{Consider finding or making a figure to explain PCR testing.}
 
+PCR testing consists of a series of cycles, each cycle containing a three-step process.
+PCR testing is used to detect the prence of a specific sequence of DNA in a sample.
+However, SARS-CoV-2 is an RNA virus and hence the first step is to convert the RNA to DNA.
+This step is known as reverse transcription, with the whole process being known as reverse-transcription polymerase chain reaction (RT-PCR).
+
+Reverse transcription is the process of converting RNA to DNA.
+In reverse transcription, a naturally-occurring enzyme known as reverse transcriptase generates the complementary DNA (cDNA) from the RNA template.
+\todo{Maybe explain what it means for RNA to have cDNA and what a template is.}
+The cDNA can then be used in the standard PCR process.
+
+The first step of the PCR cycle is \emph{denaturation}.
+In denaturation, the DNA is heated to separate the two strands of DNA.
+The second step is \emph{annealing}.
+In annealing, the temperature is lowered to allow primers to bind to the DNA.
+Primers are short sequences of DNA that are complementary to the DNA sequence of interest.
+By choosing the primer that matches the cDNA sequence that we want to detect the PCR test can be made specific to the virus of interest.
+The final step is \emph{extension}.
+In extension, the temperature is raised to allow the enzyme DNA polymerase to extend the primers and hence replicate the DNA sequence of interest.
+\todo{Check all these temperatures and maybe remove specifics?}
+
+Over the course of the cycle, the amount of DNA doubles (although there is some stochastic variation).
+Therefore, the number of DNA copies at the end of each cycle increases geometrically.
+Once enough DNA has been produced, the DNA can be detected by various methods.
+The number of cycles required to reach detection is known as the \emph{cycle threshold} (Ct) value.
+
+The Ct value increases linearly as the logarithm of the amount of RNA present in the original sample.
+That is, smaller Ct values correspond to larger amounts of RNA in the original sample.
+If there is more RNA present, more cDNA is created, and hence fewer cycles are required to reach detection.
+The logarithm is because the process is geometric, and hence the number of DNA copies increases linearly on the log-scale.
+\todo{Clearer explanation}
+
+Like any diagnostic test, PCR tests are not perfect.
+False positives and false negatives can occur.
+The \emph{specificity} of a test is the probability of a negative test result given that the individual is not infected.
+The \emph{sensitivity} of a test is the probability of a positive test result given that the individual is infected.
+
+False positives are very rare for PCR tests.
+In summer 2020, the CIS performed X PCR tests and found only Y positives, lower-bounding the test positivity at.
+\todo{Insert actual numbers of CIS bounding specificity.}
+Due to the very high specificity, I ignore false positives in this thesis.
+
+False negatives are more common.
+Early estimates of the sensitivity of PCR tests were around X\%; I produce estimates in \cref{E-ATACCC}.
 False negatives could occur for a variety of reasons, most commonly that an individual swabs themselves poorly and hence the concentration of virus on the swab is too low to be detected by the PCR process.
 Plausibly, a variety of other reasons could also lead to a false negative such as logistical issues or mislabelling of samples.
+\todo{Cite reasons false negative results occur.}
 
-Make sure I cover:
-\begin{itemize}
-  \item Relationship between Ct and viral load (and infectiousness?)
-  \item Limit of detection
-  \item The idea of being "detectable"? Check if/where/how I've used this concept in current version of text.
-  \item 
-\end{itemize}
+The minimum amount of virus required to be detected by a PCR test is known as the \emph{limit of detection}.
+\todo{Cite limit of detection and explain a bit more about it.}
 
@@
 
-\section{Studies used in this thesis}
+\todo[inline]{Include the idea of being "detectable"? Check if/where/how I've used this concept in current version of text.}
+
+\section{Studies used in this thesis} \label{biology-data:sec:studies}
 
@@
 
-The CIS (Coronavirus (COVID-19) Infection Survey) continually enrols individuals and tests them on an ongoing basis, following a
-specified testing protocol. The protocol specifies that the initial five
-tests are spaced every 7 days, then the spacing is every 28 days.
-However, real-world considerations often lead to variations in test
-times and missed tests. Our analysis focuses solely on the binary result
-(positive or negative) of these tests, assuming no misclassification
-bias.
+The CIS (Coronavirus (COVID-19) Infection Survey) continually enrols individuals and tests them on an ongoing basis, following a specified testing protocol.
+The protocol specifies that the initial five tests are spaced every 7 days, then the spacing is every 28 days.
+However, real-world considerations often lead to variations in test times and missed tests.
 
@@
 \todo[inline]{Add some descriptive figures e.g. those in \url{~/COVID/ons-incidence/duration_estimation/reproducible-sims/reports/simulation-studies.pdf}}
+
+\begin{figure}
+  \centering \includegraphics[width=\textwidth]{biology-data/CIS-positivity}
+  \caption[CIS prevalence]{%
+    CIS prevalence in the period considered in this thesis (Sep 2020 until Jan 2021).
+    Dots show raw proportion of non-void tests that are positive on each day (days with fewer than 100 tests excluded for readability).
+    Ribbon shows modelled 95\% CrI for the prevalence, corrected for non-response bias and smoothed over time (using the methodology described in \cref{E-backcalc:sec:methods}).
+    The correction tends to increase the estimated prevalence because underrepresented groups tend to have higher prevalence.
+  }
+\end{figure}
+\begin{figure}
+  \centering \includegraphics[width=\textwidth]{biology-data/CIS-num-tests}
+  \caption[Number of CIS tests]{%
+    Daily number of CIS tests conducted in the period considered in this thesis (Sep 2020 until Jan 2021).
+  }
+\end{figure}
 

### 2024-02-01 ATACCC.tex
include false negative estimates
Use RT-PCR (not PCR)

@@
 
-\Cref{E-inc-prev} showed that the duration of PCR positivity is a key parameter for estimating the incidence of SARS-CoV-2 infections.
-The aim of this chapter is to estimate the duration of PCR positivity using the viral load data from the ATACCC study (described in\todo{ref section describing ATACCC}).
+\Cref{E-inc-prev} showed that the duration of RT-PCR positivity is a key parameter for estimating the incidence of SARS-CoV-2 infections.
+The aim of this chapter is to estimate the duration of RT-PCR positivity using the viral load data from the ATACCC study (described in\todo{ref section describing ATACCC}).
 I model a biomarker, the log viral load, over the course of an infection.
-From this model, I derive the distribution of the duration of PCR positivity.
+From this model, I derive the distribution of the duration of RT-PCR positivity.
 
@@
 This modification is based on an earlier analysis using a preliminary version of the dataset, described in \cref{ATACCC:sec:original-analysis}.
-Finally, I convert the model and parameter estimates of the biomarker's trajectory to a duration of PCR positivity (\cref{ATACCC:sec:duration}).
+Finally, I convert the model and parameter estimates of the biomarker's trajectory to a duration of RT-PCR positivity (\cref{ATACCC:sec:duration}).
 I will refine this estimate in future chapters before its use in estimating transmission.
@@
 The viral load model produces a deterministic trajectory for the viral load of an individual, a function of the model parameters.
-The observation model then describes how the PCR testing process produces noisy observations of this viral load.
-The observation model includes the possibility of false positive and negatives, and the limit of detection of the PCR test.
-% The limit of detection means that viral loads below a certain threshold are not detected by the PCR test.
+The observation model then describes how the RT-PCR testing process produces noisy observations of this viral load.
+The observation model includes the possibility of false positive and negatives, and the limit of detection of the RT-PCR test.
+% The limit of detection means that viral loads below a certain threshold are not detected by the RT-PCR test.
 % They can be represented as a censoring of the viral load at the limit of detection.
@@
 
-PCR tests are noisy observations of the viral load.
-A positive PCR test has a Ct value.
+RT-PCR tests are noisy observations of the viral load.
+A positive RT-PCR test has a Ct value.
 Ct values are linearly related to viral load (see \cref{E-biology-data:sec:PCR})
-A negative PCR test signals either the viral load is too low to be detected, \ie below the limit of detection, or the test is a false negative.
-Positive PCR tests, especially those with low viral loads, could be false positives, for example caused by contamination.
+A negative RT-PCR test signals either the viral load is too low to be detected, \ie below the limit of detection, or the test is a false negative.
+Positive RT-PCR tests, especially those with low viral loads, could be false positives, for example caused by contamination.
 
@@
 
-The ultimate aim of this chapter is to produce an estimate of the distribution for the duration of PCR positivity.
+The ultimate aim of this chapter is to produce an estimate of the distribution for the duration of RT-PCR positivity.
 So far, only the viral load trajectories have been considered; \textcite{hakkiOnset} does not consider this duration either.
@@
 Denote these times as $\tau_{0,i}$ and $\tau_{1,i}$.
-Hence, their duration of PCR positivity, in continuous time, is $d_i = \tau_{1,i} - \tau_{0,i}$.
+Hence, their duration of RT-PCR positivity, in continuous time, is $d_i = \tau_{1,i} - \tau_{0,i}$.
 
@@
 Posterior estimates are conditional on the result not being a false negative.
-Negative PCR tests are plotted at the limit of detection, \ie a log viral load of 0.
+Negative RT-PCR tests are plotted at the limit of detection, \ie a log viral load of 0.
 Each plot is labelled with the individual's position in the dataset, an arbitrary value.
@@
 The rates of viral load increase at similar rates in the two groups, the vaccinated group have higher peak viral loads and clear the virus faster.
-This leads to a shorter duration of PCR positivity in the vaccinated group.
+This leads to a shorter duration of RT-PCR positivity in the vaccinated group.
 \begin{figure}
@@
 
-The negative rate, $\rho$, is estimated to be quite low, 5.3\% (95\% CrI: 3.7--7.3\%).
+The false negative rate, $\rho$, is estimated to be quite low, 5.3\% (95\% CrI: 3.7--7.3\%).
 
@@
   \centering \includegraphics{ATACCC/duration}
-  \caption[Duration of PCR positivity.]{Estimated distribution (survival function) of duration of PCR positivity for the two populations.
-The vaccinated population have shorter durations of PCR positivity.
+  \caption[Duration of RT-PCR positivity.]{Estimated distribution (survival function) of duration of RT-PCR positivity for the two populations.
+The vaccinated population have shorter durations of RT-PCR positivity.
 \label{ATACCC:fig:duration}}
@@
 
-The vaccinated population have shorter durations of PCR positivity than the unvaccinated population (see \cref{ATACCC:fig:duration}).
-The median duration of PCR positivity for the vaccinated population is 13 days (95\% CrI: 11--16 days), but it is 16 days (95\% CrI: 14--19 days) for the unvaccinated population.
+The vaccinated population have shorter durations of RT-PCR positivity than the unvaccinated population (see \cref{ATACCC:fig:duration}).
+The median duration of RT-PCR positivity for the vaccinated population is 13 days (95\% CrI: 11--16 days), but it is 16 days (95\% CrI: 14--19 days) for the unvaccinated population.
 
@@
 
-This chapter has estimated the distribution of PCR positive duration based on the viral load data from the ATACCC study.
+This chapter has estimated the distribution of RT-PCR positive duration based on the viral load data from the ATACCC study.
 

### 2024-02-01 cis-imperfect-testing.tex
Use RT-PCR (not PCR)

@@
 There are several strands of evidence supporting this hypothesis.
-First, it is well known that PCR testing can return false negatives  (see \cref{E-biology-data:sec:PCR}); I explicitly include them in the viral load model of \cref{E-ATACCC}.
+First, it is well known that RT-PCR testing can return false negatives  (see \cref{E-biology-data:sec:PCR}); I explicitly include them in the viral load model of \cref{E-ATACCC}.
 Intermittent negatives (described in  \cref{E-biology-data:sec:PCR}) show that they occur within the CIS.
@@
 
-In this chapter, I developed and applied novel methodology to estimate the duration distribution of PCR-positivity using information from two studies: ATACCC and CIS.
+In this chapter, I developed and applied novel methodology to estimate the duration distribution of RT-PCR-positivity using information from two studies: ATACCC and CIS.
 Assuming a test sensitivity of 0.8 and an informative prior on the number of infections, provided the preferred estimates.
 
-These estimates are the first estimates of the duration of PCR-positivity, including the distribution tails, estimated from general population data.
-This feature of the distribution is important for understanding the proportion of prevalence attributable to long infections and interpreting PCR test results.
+These estimates are the first estimates of the duration of RT-PCR-positivity, including the distribution tails, estimated from general population data.
+This feature of the distribution is important for understanding the proportion of prevalence attributable to long infections and interpreting RT-PCR test results.
 \todo{possibly cut some of this text?}

### 2024-02-01 incidence-prevalence.tex
Use RT-PCR (not PCR)

@@
 This chapter defines incidence and prevalence, and the relationship between these quantities.
-In the course of explaining this relationship, it will become clear that the duration of PCR positivity, and its distribution in the population, is the vital parameter which will allow estimation of incidence from prevalence, a major aim of this thesis.
+In the course of explaining this relationship, it will become clear that the duration of RT-PCR positivity, and its distribution in the population, is the vital parameter which will allow estimation of incidence from prevalence, a major aim of this thesis.
 
@@
 What it means to have a disease can vary by context (see\todo{ref section discussing what it means to have covid}).
-Here, an individual is defined to have SARS-CoV-2 on day $t$ if, had they been swabbed on day $t$, that swab would return a positive PCR test.
+Here, an individual is defined to have SARS-CoV-2 on day $t$ if, had they been swabbed on day $t$, that swab would return a positive RT-PCR test.
 I refer to these individuals as \emph{prevalent} individuals.
@@
 Since, by definition, $S(1)=1$ here, this is not an issue in this context.
-%However, if instead of estimating the incidence of PCR positives, the aim was to estimate the incidence of infections, then $S(1)$ would be small and this would be an issue.
-%While this quantity is more epidemiologically relevant, the time from infection to PCR positivity is short and, for now, I will neglect the difference.
+%However, if instead of estimating the incidence of RT-PCR positives, the aim was to estimate the incidence of infections, then $S(1)$ would be small and this would be an issue.
+%While this quantity is more epidemiologically relevant, the time from infection to RT-PCR positivity is short and, for now, I will neglect the difference.
 
@@
 Robust, unbiased prevalence estimates already exist (see\todo{ref prevalence estimates}), and incidence estimates are the objective of this thesis.
-This chapter has shown that the key quantity relating these quantities is the distribution of the duration of PCR positivity.
+This chapter has shown that the key quantity relating these quantities is the distribution of the duration of RT-PCR positivity.
 However, no estimates of this quantity in the general population exist.
-Therefore, I now turn to estimating the duration of PCR positivity.
+Therefore, I now turn to estimating the duration of RT-PCR positivity.
 

### 2024-02-01 introduction.tex
Use RT-PCR (not PCR)

@@
 Two approaches are commonly used to estimate durations: either modelling an underlying biomarker or modelling the duration directly~\autocite{sweetingEstimating}.
-For SARS-CoV-2, both approaches would be based on the results of PCR testing.
+For SARS-CoV-2, both approaches would be based on the results of RT-PCR testing.
 Modelling a biomarker would mean modelling the viral load, measured as a Ct value (see \cref{E-ATACCC}).
@@
 \begin{itemize}
-    \item Estimate of duration of PCR-positivity: of interest for rest of thesis but also more broadly for public health (\eg interpreting PCR test results)
+    \item Estimate of duration of RT-PCR-positivity: of interest for rest of thesis but also more broadly for public health (\eg interpreting RT-PCR test results)
     \item Use in interpreting test results

### 2024-02-01 transmission.tex
Use RT-PCR (not PCR)

@@
 
-In this chapter, I use the preferred estimate of the duration of PCR positivity (see \cref{E-imperf-test:sec:conclusion}) to estimate the transmission of SARS-CoV-2 in England.
+In this chapter, I use the preferred estimate of the duration of RT-PCR positivity (see \cref{E-imperf-test:sec:conclusion}) to estimate the transmission of SARS-CoV-2 in England.
 This estimation is based only on the CIS data (see \cref{transmission:sec:data} for further details).
@@
 
-In MRP, a Bayesian multilevel logistic regression model is first fit to predict PCR positivity from demographic covariates.
+In MRP, a Bayesian multilevel logistic regression model is first fit to predict RT-PCR positivity from demographic covariates.
 Then, the posterior distribution is used to predict the prevalence in each stratum.
@@
 
-% Here, I extend the system of ODEs to include compartments representing PCR positive individuals.
-% The proportion of PCR positive individuals in each strata is then linked to the CIS using a Beta-Binomial likelihood, with the modelled proportion that are PCR positive as the mean of the Beta distribution.
-One approach would be to convolve the incidence based on the transmission model with the PCR duration distribution.
+% Here, I extend the system of ODEs to include compartments representing RT-PCR positive individuals.
+% The proportion of RT-PCR positive individuals in each strata is then linked to the CIS using a Beta-Binomial likelihood, with the modelled proportion that are RT-PCR positive as the mean of the Beta distribution.
+One approach would be to convolve the incidence based on the transmission model with the RT-PCR duration distribution.
 Here, the daily incidence proportion for each stratum is by solving the system of ODEs governing the transmission model, and then prevalence calculated as using \cref{E-inc-prev:eq:EPt}.
@@
 
-I link the modelled proportion of individuals that are PCR positive to the data using a beta-binomial likelihood (as discussed in \cref{transmission:sec:data}).
+I link the modelled proportion of individuals that are RT-PCR positive to the data using a beta-binomial likelihood (as discussed in \cref{transmission:sec:data}).
 If on each day $t$, each strata $i$ is observed to have $y_{it}$ positives out of $n_{it}$ tests, then the likelihood is:
@@
 }
-  \caption[SEIR model]{SEIR model used for the analysis in this section. The shaded box represents PCR-positive individuals. Shown for a single stratum.}
+  \caption[SEIR model]{SEIR model used for the analysis in this section. The shaded box represents RT-PCR-positive individuals. Shown for a single stratum.}
   \label{SEIR:fig:full-model}
@@
         Goodness-of-fit to prevalence data by region and age for the younger three age groups (see \cref{SEIR:fig:prev-old} for the others).
-        The central ribbon and line shows the posterior median and 95\% CrI estimate for the proportion of the relevant strata PCR positive on each day.
+        The central ribbon and line shows the posterior median and 95\% CrI estimate for the proportion of the relevant strata RT-PCR positive on each day.
         The lighter outer ribbon is the 95\% posterior predictive interval for the proportion of test results that are positive each week.
@@
         Posterior median and 95\% CrI shown.
-        For the SEIR model, newly PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the backcalculation results.
+        For the SEIR model, newly RT-PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the backcalculation results.
     }

### 2024-02-01 biology-data.tex
more chapter 2 text

@@
 
+\todo[inline]{Rewrite chapter 2 intro after writing thesis intro because it will be clearer how to divide content.}
+
 SARS-CoV-2 is a beta coronavirus that emerged in late 2019.
@@
 
-Coronaviruses are RNA\todo{spell out the acronym RNA} viruses.
+Coronaviruses are RNA (Ribonucleic acid) viruses.
 RNA viruses use RNA, rather than DNA, as their genetic material.
@@
 
-PCR testing consists of a series of cycles, each cycle containing a three-step process.
-PCR testing is used to detect the prence of a specific sequence of DNA in a sample.
-However, SARS-CoV-2 is an RNA virus and hence the first step is to convert the RNA to DNA.
-This step is known as reverse transcription, with the whole process being known as reverse-transcription polymerase chain reaction (RT-PCR).
+PCR (Polymerase Chain Reaction) is a procedure that produces millions to billions of copies of a specific segment of DNA (Deoxyribonucleic acid) from a single copy~\autocites{smithPCR}{garibyanPCR}.
+This is known as \emph{amplifying} the DNA.
+PCR has many uses within the biological sciences; Mark R.\ Hughes, deputy director of the Human Genome Project, has claimed it is the most important scientific technology of the last hundred years~\autocite{powledgePCR}.
+It is central to the work in this thesis because it can detect the presence of SARS-CoV-2 in a sample taken from a human, most commonly a nose and/or throat swab.
 
-Reverse transcription is the process of converting RNA to DNA.
-In reverse transcription, a naturally-occurring enzyme known as reverse transcriptase generates the complementary DNA (cDNA) from the RNA template.
-\todo{Maybe explain what it means for RNA to have cDNA and what a template is.}
+SARS-CoV-2 is an RNA virus, but PCR amplifies DNA.
+Therefore, before the PCR process can be used to detect SARS-CoV-2, the RNA must be converted to DNA.
+This process is known as \emph{reverse transcription}.
+In reverse transcription, a naturally-occurring enzyme known as reverse transcriptase generates cDNA (complementary DNA) from the RNA in the sample~\autocites{valasekPower}.
+Any sequence of cDNA has a one-to-one correspondence with a sequence of RNA.
 The cDNA can then be used in the standard PCR process.
 
+PCR consists of a series of cycles, each cycle containing a three-step process~\autocites{powledgePCR}{garibyanPCR}.
 The first step of the PCR cycle is \emph{denaturation}.
@@
 Primers are short sequences of DNA that are complementary to the DNA sequence of interest.
-By choosing the primer that matches the cDNA sequence that we want to detect the PCR test can be made specific to the virus of interest.
+By choosing the primers, the PCR process targets a portion of DNA starting and ending with a specified sequence.
 The final step is \emph{extension}.
 In extension, the temperature is raised to allow the enzyme DNA polymerase to extend the primers and hence replicate the DNA sequence of interest.
-\todo{Check all these temperatures and maybe remove specifics?}
+In the course of such a cycle, the amount of DNA doubles.
+Therefore, the number of DNA copies increases geometrically with the number of cycles completed.
 
-Over the course of the cycle, the amount of DNA doubles (although there is some stochastic variation).
-Therefore, the number of DNA copies at the end of each cycle increases geometrically.
-Once enough DNA has been produced, the DNA can be detected by various methods.
+Quantitative real-time PCR (or RT-PCR) gives a measurement of the amount of the target DNA sequence present in the original sample~\autocite{yangPCRdiagnostics}.
+After each PCR cycle, the amount of target DNA is measured and compared to a pre-derived standard.
+Once the measured amount of DNA exceeds the standard, the target DNA is said to be detected.
 The number of cycles required to reach detection is known as the \emph{cycle threshold} (Ct) value.
+Quantitative real-time PCR have a number of other practical advantages over previous PCR techniques (\eg speed); see \textcites{yangPCRdiagnostics}{valasekPower} for more details.
+All RT-PCR tests I consider in this thesis are quantitative real-time RT-PCR tests, and produce Ct values.
 
-The Ct value increases linearly as the logarithm of the amount of RNA present in the original sample.
-That is, smaller Ct values correspond to larger amounts of RNA in the original sample.
-If there is more RNA present, more cDNA is created, and hence fewer cycles are required to reach detection.
-The logarithm is because the process is geometric, and hence the number of DNA copies increases linearly on the log-scale.
-\todo{Clearer explanation}
+The Ct value decreases as the amount of DNA present in the original sample increases.
+If there is more RNA present, more DNA is created, and hence fewer cycles are required to reach detection.
+The logarithm is because the process is geometric, and hence the number of DNA copies increases linearly on the log-scale each cycle.
+Therefore, the Ct value is proportional to the negative log of the amount of DNA present in the original sample.
+The amount of cDNA produced by reverse transcription is proportional to the amount of RNA present in the original sample.
+Therefore, when testing for SARS-CoV-2, the Ct value is proportional to the negative log of the amount of SARS-CoV-2 RNA present in the original sample.
 
@@
 False positives are very rare for PCR tests.
-In summer 2020, the CIS performed X PCR tests and found only Y positives, lower-bounding the test positivity at.
+In summer 2020, the CIS performed X PCR tests and found only Y positives, lower-bounding the test specificity at.
 \todo{Insert actual numbers of CIS bounding specificity.}
-Due to the very high specificity, I ignore false positives in this thesis.
 
@@
 
+RT-PCR enables the detection of SARS-CoV-2 viral RNA in a sample.
+However, this does not necessarily mean that the virus is viable (\ie able to reproduce)~\autocite{puhachSARSCoV2}.
+High viral loads (low Ct values) are more likely to be viable, but more sophisticated tests are required to determine viability~\autocite{singanayagamDuration,puhachSARSCoV2,hakkiOnset}.
+
 \todo[inline]{Include the idea of being "detectable"? Check if/where/how I've used this concept in current version of text.}
@@
 
-The CIS (Coronavirus (COVID-19) Infection Survey) continually enrols individuals and tests them on an ongoing basis, following a specified testing protocol.
-The protocol specifies that the initial five tests are spaced every 7 days, then the spacing is every 28 days.
-However, real-world considerations often lead to variations in test times and missed tests.
+The CIS was setup in April 2020 to provide gold-standard measurements of the prevalence of SARS-CoV-2 in the community.
+It is a longitudinal study with a representative sample of private households, this excludes settings such as care homes and student halls.
+Initially, it was limited to England, but expanded to cover the whole of the UK in September 2020.
+Enrolment was continuous until 31st Jan 2022, with data collected until 13th Mar 2023~\autocite{weiRisk}.
 
-The CIS (Coronavirus Infection Survey) is a longitudinal study on a representative sample of households.
-Households are invited to the study from databases held by the ONS.
-The number invited was to satisfy various target number of individuals swabbing per fortnight.
-Within a selected household, all individuals aged 2 and over were invited to participate.
-Once invited, an enrolment swab would be taken followed by 4 further weekly swabs (giving a total of 5 swabs on days 0, 7, 14, 21, 28 relative to enrolment) after which monthly swabs are taken.
-A full description of the study can be found in \textcite[][supplementary materials]{pouwelsCommunity} or the study protocol~\autocite{cisProtocol}.
+The CIS had a household-based design, meaning that households were invited to participate in the study.
+From households that participated, all individuals aged 2 and over were invited to participate.
+Once invited, an enrolment swab would be taken at the first visit followed by 4 further weekly visits (giving a total of 5 swabs on days 0, 7, 14, 21, 28 relative to enrolment) after which visits were monthly.
+As well as a swab to test for SARS-CoV-2, participants were asked to complete a questionnaire, and some participants were asked to provide blood samples.
+Real-world issues meant that visits were often not on this precise schedule, and occasionally missed.
+A full description of the study can be found in the study protocol~\autocite{cisProtocol}.
 
-During the period we consider, the total cohort size expands due to the continuous recruitment into the study.
-Following this, the number then decreases and stabilises as those recruited to meet the October target transition from weekly onto monthly testing.
+Up until July 2020, households were selected from those that had previously participated in an ONS survey~\autocite{CIStechData}.
+Around 50\% of households agreed to participate, with around 90\% of eligible individuals (96,113 in total) in these households contributing multiple swabs to the survey.
+After July 2020, addresses were also randomly sampled from a database of all UK addresses held by the ONS.
+The response rate was lower, with only 12\% of households agreeing to participate, from which 86\% of eligible individuals (383,732 in total) contributed multiple swabs to the survey.
 
-\todo[inline]{Add some descriptive figures e.g. those in \url{~/COVID/ons-incidence/duration_estimation/reproducible-sims/reports/simulation-studies.pdf}}
+In this thesis, I focus on the period from August 2020 until January 2021.
+During the start of this period, the recruitment rate of CIS was expanded (see \cref{biology-data:fig:CIS-recruitment}(A)).
+Following this, the recruitment then decreases and stabilises, as does the rate of swabs taken (see \cref{biology-data:fig:CIS-recruitment}(B)).
+The recruitment pattern was chosen to hit targets regarding the fortnightly number of swabs.
+Because the testing becomes less frequent after the first 4 weeks, some continuous recruitment is needed to maintain the number of swabs, as well as to replace participants who drop out.
+
+\todo[inline]{Discuss non-response bias}
+
+\begin{figure}
+  \centering \includegraphics{biology-data/CIS-recruitment}
+  \caption[CIS swab numbers]{%
+    Number of swabs taken by the CIS in the period considered in this thesis (Sep 2020 until Jan 2021).
+    (A) Only enrolment (\ie first) swabs for each individual.
+    (B) Total swabs.
+    Note the differing y-axis scales.
+    Data from \textcite{CIStechData}.
+  }
+  \label{biology-data:fig:CIS-recruitment}
+\end{figure}
 
@@
   }
+  \label{biology-data:fig:CIS-positivity}
 \end{figure}
@@
   }
+  \label{biology-data:fig:CIS-num-tests}
+  \todo[inline]{Cut this repetition of number of tests? Maybe highlight period of interest in previous plot}
 \end{figure}
-
-Consider the CIS data.
-It consists of a series of tests on the same individuals (longitudinal data), where some individuals have a series of positive tests.
 
@@
 
-\emph{Explain definition in ``SARS-CoV-2 reinfection definition'' from \textcite{weiRisk}.}
+For the work in \cref{E-cis-perf-testing,E-cis-imperf-testing}, positive test results need to be grouped into \emph{infection episodes}.
+An infection episode is a series of positive tests that are part of the same infection.
+I use a method developed throughout the pandemic, and published within \textcite{weiRisk}; see that publication for full justification of the criteria.
+Denote the time of a positive test by $t$, the number of consecutive negative tests prior to $t$ by $n$, and the time of the first positive test in the same individual by $t_p$ ($t_p = -\infty$ if there is no previous positive test in the same individual).
+The test at $t$, in the pre-Omicron era (prior to December 2021), is defined as the start of a new infection episode if any of the following apply.
+\begin{enumerate}
+  \item $t_p = -\infty$.
+  \item $n \geq 1$ and $t - t_p > 120$.
+  \item $n \geq 2$ and $t - t_p > 90$.
+  \item $n \geq 3$ and $t - t_p > 60$.
+  \item $n \geq 4$.
+  \item A complex heuristic based on the likely variant of the positive tests (see \textcite{weiRisk} for details).
+This heuristic rarely applied in the pre-Omicron era~\citePersonalComms{Sarah Walker}.
+\end{enumerate}
 
-\emph{Intermittent negatives} are the clearest example of a false negative.
+A negative between two positive tests, when those positives are classed as part of the same infection episode, is known as an \emph{intermittent negative}.
+Intermittent negatives are the clearest example of a false negative.
 An intermittent negative is when an individual tests negative but tested positive previously and subsequently.
@@
 % Whether a future positive is part of the same episode of a reinfection is not always trivial; I rely on a process developed previously by Sarah Walker, see \cref{E-episode-def}.
-However, up until the emergence of the Omicron variant, reinfections are rare, especially in a short time frame, and hence the tricky cases are rare until this time (late 2021).
 

### 2024-02-01 biology-data/CIS-positivity.R

@@ -34,23 +34,23 @@ corrected_model = readRDS(
     group_by(date) |>
     mean_qi(prevalence)
 
+counts_to_plot = raw_counts |>
+    filter(
+        num_tests > 100,
+        between(date, min(corrected_model$date), max(corrected_model$date))
+    )
 
 plot_positivity = ggplot() +
     geom_point(
         aes(date, mean),
-        data = raw_counts |> filter(num_tests > 100)
+        data = counts_to_plot
     ) +
     geom_ribbon(
         aes(date, prevalence, ymin = .lower, ymax = .upper),
         data = corrected_model,
         alpha = 0.5
     ) +
-    scale_y_continuous(
-        labels = scales::percent_format(accuracy = 1),
-    ) +
-    coord_cartesian(
-        c(min(corrected_model$date), max(corrected_model$date))
-    ) +
+    scale_y_continuous(labels = scales::label_percent()) +
     labs(
         x = "Date",
         y = "Prevalence"

### 2024-02-01 biology-data.tex
preferable citation sorting

@@
 
-PCR (Polymerase Chain Reaction) is a procedure that produces millions to billions of copies of a specific segment of DNA (Deoxyribonucleic acid) from a single copy~\autocites{smithPCR}{garibyanPCR}.
+PCR (Polymerase Chain Reaction) is a procedure that produces millions to billions of copies of a specific segment of DNA (Deoxyribonucleic acid) from a single copy~\autocite{smithPCR,garibyanPCR}.
 This is known as \emph{amplifying} the DNA.
@@
 
-PCR consists of a series of cycles, each cycle containing a three-step process~\autocites{powledgePCR}{garibyanPCR}.
+PCR consists of a series of cycles, each cycle containing a three-step process~\autocite{powledgePCR,garibyanPCR}.
 The first step of the PCR cycle is \emph{denaturation}.

### 2024-02-01 cis-perfect-testing.tex
easy edits
preferable citation sorting

@@
 This is especially true within the human biostatistical literature.
-The literature does include theoretical frameworks which include the double censored and truncated case~\autocites{turnbullEmpirical}{dempsterMaximum}.
+The literature does include theoretical frameworks which include the double censored and truncated case~\autocite{turnbullEmpirical,dempsterMaximum}.
 The frameworks have only been applied when the terminating event was either uncensored or right censored~\autocite{sunEmpirical,bacchettiNonparametric}.
@@
 
-Bayesian methods been used for nest survival before~\autocites{heBayesiana}{heBayesian}{caoModeling}, augmenting the data with the unobserved times of infection.
+Bayesian methods been used for nest survival before~\autocite{heBayesiana,heBayesian,caoModeling}, augmenting the data with the unobserved times of infection.
 The augmentation creates a parameter per detected infection.
@@
 This approach extends the framework of \textcite{heiseyModelling} to include individuals without detected episodes.
-This framework is itself based long-standing methodology~\autocites{dempsterMaximum}{turnbullEmpirical}.
+This framework is itself based long-standing methodology~\autocite{dempsterMaximum,turnbullEmpirical}.
 I augment the data set with the full number of episodes, including those that are undetected.
@@
 
-\todo{Not sure if the followin paragraph is understandable}
 In forming the probability $p_{ia}$, the data is used.
@@
 Having approximated the estimate as a multivariate normal, I add additional uncertainty using a discrete Beta process.
-The discrete Beta process prior generalises the form of prior used in \cref{perf-test:sec:independent-priors} by allowing the central estimate of the hazard to vary over time~\autocites{ibrahimBayesian}{sunStatisticala}.
+The discrete Beta process prior generalises the form of prior used in \cref{perf-test:sec:independent-priors} by allowing the central estimate of the hazard to vary over time~\autocite{ibrahimBayesian,sunStatisticala}.
 It is:
@@
 Most importantly, covariates (\eg age, immunity from vaccination and/or previous infection, or variant) could be included.
-Immunity in particular affects the duration substantially, although age and variant may also make a contribution~\autocites{hakkiOnset}{russellWithinhost}.
+Immunity in particular affects the duration substantially, although age and variant may also make a contribution~\autocite{hakkiOnset,russellWithinhost}.
 Covariates can be included with a variety of models, the simplest being a proportional hazards model.

### 2024-02-01 incidence-prevalence.tex
preferable citation sorting

@@
 The target of inference can be either the process's intensity or the counts of the specific realisation that occurred.
-The realised counts are normally of greater interest because we care about the realised epidemic, not what may happen if there were future realisations of the same epidemic~\autocites{beckerDependent}{brookmeyerMethod}.
+The realised counts are normally of greater interest because we care about the realised epidemic, not what may happen if there were future realisations of the same epidemic~\autocite{beckerDependent,brookmeyerMethod}.
 
@@
 % That is, $Z_t >> \E(Z_t)$ for some $t$ during which a super-spreading event occurs.
-% Coronaviruses, such as SARS-CoV-2, have particularly high overdispersion~\autocites{endoEstimating}{adamClustering}{mccloskeySARS}.
+% Coronaviruses, such as SARS-CoV-2, have particularly high overdispersion~\autocite{endoEstimating,adamClustering}{mccloskeySARS}.
 % An infamous example of a SARS-CoV-2 super-spreading event was the Skagit County choir practice in March 2020~\autocite{hamnerHigh}.
@@
 % Generally, $k<1$ is considered as significant levels of overdispersion.
-% Estimates for SARS-CoV-2 indicate $k \approx 0.1$~\autocite{endoEstimating}, and may be decreased by some interventions including lockdowns~\autocites{quiltyReconstructing}{quiltyUnderstanding}.
+% Estimates for SARS-CoV-2 indicate $k \approx 0.1$~\autocite{endoEstimating}, and may be decreased by some interventions including lockdowns~\autocite{quiltyReconstructing,quiltyUnderstanding}.
 % If $K(s) - K(t)$ has a negative binomial distribution, then its standard deviation scales on the order of the mean.
@@
 The basic SIR model is the deterministic approximation of a stochastic model where the duration of time in the infectious compartment is exponentially distributed~\autocite[96]{keelingModeling}.
-This is an unrealistic assumption for most diseases and would lead to underestimating of the reproduction number~\autocites{lloydRealistic}{wearingAppropriate}.
+This is an unrealistic assumption for most diseases and would lead to underestimating of the reproduction number~\autocite{lloydRealistic,wearingAppropriate}.
 A standard extension to the SIR model is to instead assume a gamma distribution~\autocite[94]{keelingModeling}{andersonSpread}.
@@
 
-Age-based contact patterns are important for the transmission dynamics of respiratory diseases~\autocites{birrellBayesian}{jacksonEffects}.
+Age-based contact patterns are important for the transmission dynamics of respiratory diseases~\autocite{birrellBayesian,jacksonEffects}.
 Respiratory diseases can be spread by conversational contacts, which show very clear age patterns (see \cref{SEIR:fig:age-contacts}).
@@
 In particular, it often poorly explores the tails of the posterior in finite time~\autocite[e.g.][]{turitsynIrreversible}.
-\emph{Hamiltonian Monte Carlo} (HMC)~\autocites{nealMCMC}{nealImproved}{duaneHybrid} improves the proposals by using information from the gradient of the posterior density (taken with respect to the parameters).
+\emph{Hamiltonian Monte Carlo} (HMC)~\autocite{nealMCMC,nealImproved,duaneHybrid} improves the proposals by using information from the gradient of the posterior density (taken with respect to the parameters).
 HMC simulates a physical system travelling through the posterior space with momentum.

### 2024-02-01 introduction.tex
preferable citation sorting

@@
 It has killed 40 million people, peaking at 2 million annual deaths~\autocite{unaids2023}.
-In this context, backcalculation estimates the incidence of HIV infection from: data on the number of AIDS diagnoses, and the incubation distribution (the time between infection and AIDS diagnosis)~\autocites{brookmeyerBackcalculation}{brookmeyerMeasuring}.
+In this context, backcalculation estimates the incidence of HIV infection from: data on the number of AIDS diagnoses, and the incubation distribution (the time between infection and AIDS diagnosis)~\autocite{brookmeyerBackcalculation,brookmeyerMeasuring}.
 AIDS is incurable, therefore, AIDS prevalence is equivalent to cumulative incidence.

### 2024-02-01 transmission.tex
preferable citation sorting

@@
 During November, incidence starts rising, although starting at different times in different regions.
-The most explored explanation for this is the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocites{walkerTracking}{daviesEstimated}{lythgoeLineage}.
+The most explored explanation for this is the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocite{walkerTracking,daviesEstimated,lythgoeLineage}.
 A second peak occurs in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
@@
 I take information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using the methodology of \textcites{vanleeuwenTime}{vanleeuwenAugmenting}.
-%; these estimates have been used for previous modelling of the COVID-19 pandemic~\autocites{vanleeuwenTime}{birrellRealtime}.
+%; these estimates have been used for previous modelling of the COVID-19 pandemic~\autocite{vanleeuwenTime,birrellRealtime}.
 % These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
@@
 
-Studies of pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults~\autocites{chenRole}{vinerTransmission}.
+Studies of pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults~\autocite{chenRole,vinerTransmission}.
 Therefore, I set $q^\text{inf}_{a'} = 1$ for all $a'$, $q^\text{susc}_{a} = q_c$ for 0--11 year olds, and $q^\text{susc}_{a} = 1$ for the remaining strata.
@@
 The growth rate $\psi$ is easily identifiable from the data, and therefore a parameterisation in terms of $\phi$ should be more computationally efficient.
-If this is the initial exponential phase, when $\vec{s} \approx 1$, then the basic reproduction number for a SEIR model is linked to this growth rate by~\autocites{birrellBayesian}{wearingAppropriate}:
+If this is the initial exponential phase, when $\vec{s} \approx 1$, then the basic reproduction number for a SEIR model is linked to this growth rate by~\autocite{birrellBayesian,wearingAppropriate}:
 \begin{align}

### 2024-02-02 transmission.tex
format textcite

@@
     Lit review to go in introduction chapter, framing the whole thesis because it is important in understanding the motivation for the whole work.
-    Would include \textcites{mccabeCISincidence}{abbottCISincidence} on backcalculation. Also Toffe?
+    Would include \textcite{mccabeCISincidence,abbottCISincidence} on backcalculation. Also Toffe?
     As well as the literature on mechanistic models using REACT and CIS I have previously circulated.
@@
 
-I take information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using the methodology of \textcites{vanleeuwenTime}{vanleeuwenAugmenting}.
+I take information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using the methodology of \textcite{vanleeuwenTime,vanleeuwenAugmenting}.
 %; these estimates have been used for previous modelling of the COVID-19 pandemic~\autocite{vanleeuwenTime,birrellRealtime}.

### 2024-02-02 biology-data.tex
more chapter 2 writing

@@
 
+\todo[inline]{I should maybe swap 2.1 and 2.2 order.
+I like the current order because natural history feels more basic and introductory, but it does reference a bunch of stuff about PCR testing so it might be easier to follow the other way round.}
+
 \todo[inline]{Rewrite chapter 2 intro after writing thesis intro because it will be clearer how to divide content.}
@@
 \begin{figure}
+  \makebox[\textwidth][c]{
+  \begin{tikzpicture}[
+      node distance = 4cm,
+      on grid,
+      auto,
+      ->,>=stealth',
+      every state/.style={draw,rectangle,align=center},
+      ]
+
+      \node[state] (pre) {PCR negative \\ Not infectious};
+      \node[state, right of=pre] (pos) {PCR positive \\ Not infectious};
+      \node[state, right of=pos] (infectious) {PCR positive \\ Infectious};
+      \node[state, right of=infectious] (not-infectious) {PCR positive \\ Not infectious};
+      \node[state, right of=not-infectious] (neg) {PCR negative \\ Not infectious};
+
+      \node[draw=none, below=2cm of pre] (infected) {} ;
+      \node[draw=none, below=2cm of neg] (recovered) {} ;
+
+      \node[above=2cm of pre] (titleA) {(A)};
+      \node[below=1cm of infected] (titleB) {(B)};
+
+      % \draw (0,-2) -- node {Infected} (pre);
+      % \draw (not-infectious) -- (0,-2);
+
+      \path (infected) edge node {Infected} (pre)
+          (pre) edge (pos)
+            (pos) edge (infectious)
+            (infectious) edge (not-infectious)
+            (not-infectious) edge (neg)
+            (neg) edge (recovered);
+
+    % Detectable
+    \node[below = 1cm of infectious] (detect-label) {Detectable};
+    \node[draw, thick, inner sep=0.3cm, fit=(pos) (infectious) (not-infectious) (detect-label), fill=blue!10,opacity=0.2] (detect-box) {};
+  \end{tikzpicture}
+  }
   \centering\includegraphics[width=\textwidth]{biology-data/natural-history}
-  \caption[Natural history of SARS-CoV-2.]{Natural history of SARS-CoV-2. Reproduced from \textcite{puhachSARSCoV2} with permission.}
-  \todo[inline]{Define the acronyms and terminology in the figure.}
+  \caption[Natural history of SARS-CoV-2.]{%
+    Natural history of SARS-CoV-2.
+    (A) A simplified state-based model, which I generally use in this thesis.
+    Shading represents states where the individual is detectable by a PCR test.
+    (B) A full model of the natural history, including continuous processes (Reproduced from \textcite{puhachSARSCoV2} with permission; durations are based on averages, there is a large amount of individual variation).
+    PCR positive refers to the main diagnostic test used in this thesis, and detect RNA genomes (see \cref{biology-data:sec:PCR} for more details).
+    Ag-RDT (antigen rapid diagnostic test) tests are lateral flow tests, which detect infectious virus rather than RNA, and are not used in this thesis.
+  }
   \label{biology-data:fig:natural-history}
@@
 Next, the individual becomes infectious; normally, this occurs prior to symptom onset for those who experience symptoms.
-Recovery from symptoms occurs after a few days, around individual is no longer infectious.
+Recovery from symptoms occurs after a few days, around the time the individual is no longer infectious.
 Finally, the individual will test negative on a PCR test.
-\todo{Talk about PCR testing detecting RNA material even with no viable virus.}
+
+\todo[inline]{Maybe move the following paragraph because it is referencing forward a bunch of concepts not yet introduced properly, or if sections 2.1 and 2.2 are swapped it starts making sense.}
+I will refer to individuals between the time they start testing positive on a PCR test and the time they stop testing positive on a PCR test as \emph{detectable}.
+It is not possible to directly observe whether an individual is detectable or not due to noise in the PCR test (see \cref{biology-data:sec:PCR}).
+However, viewing the number of RNA genome copies in an individual as a latent process this can then be defined as the time when the number of RNA genome copies is above the limit of detection of the PCR test.
 
@@
 
-\todo[inline]{Add something about viral load being the process underlying these processes.}
+A host's \emph{viral load}, the quantity of viral particles in a host, is the underlying biological process determining many of these changes~\autocite{puhachSARSCoV2}.
+Any host's viral load is determined by the interaction of many biological processes, most prominently the virus's replication increasing the viral load and the host's immune response reducing the viral load~\autocite{keVivo}.
+Viral replication requires the virus to infect a host cell and then use the host cell's machinery to produce more virus; as more host cells are infected this process slows because there are fewer remaining host cells, known as \emph{target cells}, to infect~\autocite{keVivo}.
+The host's immune system can slow this process by various mechanisms; the most relevant for SARS-CoV-2 is by making some target cells less susceptible to infection.
+A host's immune response is triggered by the viral load, and becomes stronger as the viral load increases.
+A high viral load is associated with being more infectious, more likely to test positive on a PCR test, and the start of symptoms~\autocite{puhachSARSCoV2,keVivo}.
 
@@
 The number of cycles required to reach detection is known as the \emph{cycle threshold} (Ct) value.
-Quantitative real-time PCR have a number of other practical advantages over previous PCR techniques (\eg speed); see \textcites{yangPCRdiagnostics}{valasekPower} for more details.
+Quantitative real-time PCR have a number of other practical advantages over previous PCR techniques (\eg speed); see \textcite{yangPCRdiagnostics,valasekPower} for more details.
 All RT-PCR tests I consider in this thesis are quantitative real-time RT-PCR tests, and produce Ct values.
@@
 When a host's viral load is too low (below the limit of detection of the PCR test), no Ct value is available and hence viral load can be considered censored at the limit of detection.
-The relationship between Ct and viral load is noisy and can vary based on many factors including quality of swab obtained, primers used (even between batches), and the PCR system set-up~\autocites{dahdouhCt,hanRTPCR}.
+
+Ct values cannot be transformed into viral loads (expressed in units such as the number of viral RNA copies per swab) without additional information~\autocite{dahdouhCt,puhachSARSCoV2}.
+Ct values are test-specific, and can even vary between different batches of the same test.
+Testing a series of samples with known viral loads allows the relationship between Ct and viral load to be estimated for a specific setup.
 
@@
 However, this does not necessarily mean that the virus is viable (\ie able to reproduce)~\autocite{puhachSARSCoV2}.
-High viral loads (low Ct values) are more likely to be viable, but more sophisticated tests are required to determine viability~\autocite{singanayagamDuration,puhachSARSCoV2,hakkiOnset}.
-
-\todo[inline]{Include the idea of being "detectable"? Check if/where/how I've used this concept in current version of text.}
+High viral loads (low Ct values) are more likely to be viable, but more sophisticated tests are required to determine viability.
+The gold standard procedure is attempting to grow and recover viral material from a sample, a process known as \emph{cell culturing}~\autocite{singanayagamDuration,puhachSARSCoV2,hakkiOnset}.
 
@@
 
+In this thesis, I make use of data from two studies which recruited participants from the general population in the UK.
+Unlike many studies of SARS-CoV-2, these studies are not focused on specific groups such as healthcare workers or hospital patients.
+Therefore, the results from these studies are more generalisable to the entire UK population.
+The first study, ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts) performed daily testing of a small number of individuals with limited follow-up.
+The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability, for but the results are limited because the small sample size and short follow-up means that the tails of these distributions cannot be estimated.
+The second study, the CIS (Coronavirus (COVID-19) Infection Survey), performed weekly or monthly testing of a large number of individuals with long follow-up.
+The large sample size and long follow-up allows for reliable estimates of the tails of the distributions of the quantities of interest, but the results are limited because the sparse sampling means that the viral load dynamics cannot be estimated.
+In particular, the sparse sampling means that the vast majority of detected infections had only one or two positive tests, and hence the viral load trajectory of these infections cannot be estimated.
+
 \subsection{Assessment of Transmission and Contagiousness of COVID-19 in Contacts}
 
+The ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts) study was a longitudinal study of contacts reported to the NHS Test and Trace contact tracing service~\autocite{hakkiOnset}.
+Contacts were invited to participate if they were 5 years or older and  could be reached within 5 days of the index case's symptom onset date.
+ATACCC was formed of two waves: ATACCC1 enrolled 393 contacts between 13th Sep 2020 and 31st Mar 2021 (when pre-Alpha and Alpha variants were circulating) and ATACCC2 enrolled 345 contacts between 24th May 2021 and 228th Oct 2021 (when the Delta variant was circulating).
+Contacts were swabbed daily for up to 20 days, with a variety of tests performed on the swabs; relevant for this thesis is the RT-PCR test for SARS-CoV-2.
+For all positive test results, the viral load was measured using the Ct value and then converted to a viral load using a standard curve.
+Other data such as demography and symptoms was also collected.
+For further details see \textcite{singanayagamDuration,hakkiOnset}.
+
+\Textcite{hakkiOnset} analyse the viral load of the 57 contacts who tested positive for SARS-CoV-2 and had the growth phase in their viral load measured.
+For 115 contacts with at least one positive RT-PCR test, the growth phase of viral load was not captured; these contacts were excluded from this analysis to ensure all individuals viral load trajectories could be identified.
+More details on this analysis and my extensions to it can be found in \cref{E-ATACCC}.
+
 \subsection{Coronavirus (COVID-19) Infection Survey} \label{intro:sec:cis}
+
+\todo[inline]{Do I need to distinguish between estimates of positivity and prevalence?}
 
@@
 
+These rates of response raise the possibility of non-response bias.
+In particular, some demographic groups, such as young adults and individuals of non-white ethnicity, were underrepresented among those who responded~\autocite{pouwelsCommunity}.
+To correct for this, various methods have been used to adjust the estimates of prevalence.
+I explain the preferred method, MRP (multilevel regression and post-stratification), in \cref{E-backcalc:sec:methods}.
+
 In this thesis, I focus on the period from August 2020 until January 2021.
@@
 Because the testing becomes less frequent after the first 4 weeks, some continuous recruitment is needed to maintain the number of swabs, as well as to replace participants who drop out.
-
-\todo[inline]{Discuss non-response bias}
 

### 2024-02-02 distributions.tex
Cite LKJ distribution
fix mixing constant

@@
 Exponential & $\Exponential(\lambda)$ & $\lambda$ (rate) & $\frac{1}{\lambda}$ & $\frac{1}{\lambda^2}$ & $\lambda e^{-\lambda x}$ \\
-LKJ* & $\LKJ(\eta)$ & $\eta > 0$ (shape) & $\matr{I}$ (identity matrix) & N/A & $\det(\matr{x})^{\eta-1}$ \\
+LKJ* & $\LKJ(\eta)$ & $\eta > 0$ (shape) & $\matr{I}$ (identity matrix) & N/A & $C \det(\matr{x})^{\eta-1}$ \\
 \end{tabular}
 
-* The LKJ distribution has support over positive definite symmetric matrices $\matr{x}$.
+* The LKJ distribution~\autocite{lewandowskiGenerating} has support over positive definite symmetric matrices $\matr{x}$.
 These matrices are exactly the matrices that are valid correlation matrices.

### 2024-02-02 biology-data.tex
editing etc

@@
 It is closely related to previous coronaviruses that have caused outbreaks in humans, SARS-CoV and MERS-CoV.
-\todo{Cite and check how SARS-CoV-2 is related to SARS-CoV and MERS-CoV, and other human coronaviruses.}
 
@@
 RNA viruses use RNA, rather than DNA, as their genetic material.
-\todo{why is being a RNA virus important?}
 
@@
 
-An individual infected with SARS-CoV-2 goes through several stages (see \cref{biology-data:fig:natural-history}).
+An individual infected with SARS-CoV-2 goes through several disease stages (see \cref{biology-data:fig:natural-history}).
 \todo{Find citations for disease stages.}
@@
 However, viewing the number of RNA genome copies in an individual as a latent process this can then be defined as the time when the number of RNA genome copies is above the limit of detection of the PCR test.
+While this definition is not perfectly precise, it is clear enough for the purposes it used in this thesis.
 
@@
 The time from infection to experiencing symptoms is known as the \emph{incubation period} and is typically around\todo{cite incubation period}.
-The latent period is undefined for asymptomatic individuals, around\todo{cite proportion asymptomatic} of those infected.
+The incubation period is undefined for asymptomatic individuals, around\todo{cite proportion asymptomatic} of those infected.
 
@@
 Any host's viral load is determined by the interaction of many biological processes, most prominently the virus's replication increasing the viral load and the host's immune response reducing the viral load~\autocite{keVivo}.
-Viral replication requires the virus to infect a host cell and then use the host cell's machinery to produce more virus; as more host cells are infected this process slows because there are fewer remaining host cells, known as \emph{target cells}, to infect~\autocite{keVivo}.
+Viral replication requires the virus to infect a host cell and then use the host cell's machinery to produce more virus; as more host cells are infected this process slows because there are fewer remaining host cells, known as \emph{target cells}, to infect.
 The host's immune system can slow this process by various mechanisms; the most relevant for SARS-CoV-2 is by making some target cells less susceptible to infection.
@@
 PCR has many uses within the biological sciences; Mark R.\ Hughes, deputy director of the Human Genome Project, has claimed it is the most important scientific technology of the last hundred years~\autocite{powledgePCR}.
-It is central to the work in this thesis because it can detect the presence of SARS-CoV-2 in a sample taken from a human, most commonly a nose and/or throat swab.
+It is central to the work in this thesis because it can be used to detect the presence of SARS-CoV-2 in a sample taken from a human, most commonly a nose and/or throat swab.
 
@@
 
+\todo[inline]{%
+  Rewrite this section on imperfect testing.
+  Including analytical sensitivity: the minimum number of copies of the target RNA material in the sample that the test can detect, measured using the 
+}
 Like any diagnostic test, PCR tests are not perfect.
 False positives and false negatives can occur.
-The \emph{specificity} of a test is the probability of a negative test result given that the individual is not infected.
-The \emph{sensitivity} of a test is the probability of a positive test result given that the individual is infected.
+These can be quantified in a number of ways.
 
-False positives are very rare for PCR tests.
+The \emph{analytical sensitivity} of the test is the minimum number of copies of the target RNA material in the sample that the test can detect, typically expressed as the \emph{limit of detection}~\autocite{bustinMIQE}.
+The limit of detection is the lowest concentration of the target RNA material that can be reliably detected using the procedure.
+This is different to the \emph{clinical sensitivity} of the test, the ability of the test to detect a detectable individual (being detectable is defined in \cref{biology-data:sec:natural-history}).
+Clinical sensitivity can be affected by a large variety of preanalytical factors, such as sample collection (\ie the swabbing process) and sample transportation~\autocite{lippiPotential}.
+Clinical sensitivity is of primary interest in this thesis.
+I will quantify it as the \emph{test sensitivity} (strictly the clinical test sensitivity), the probability of a positive test result for a detectable individual.
+A negative result for a detectable individual is a \emph{false negative}.
+
+The \emph{analytical specificity} of the test is the ability of the test to distinguish between the intended target rather than other, non-specific RNA (or DNA) sequences present in the sample~\autocite{bustinMIQE}.
+As with sensitivity, this is different to the \emph{clinical specificity} of the test, the ability of the test to distinguish between detectable and undetectable individuals.
+A positive result for an undetectable individual is a \emph{false positive}.
+False positives are very rare for SARS-CoV-2 RT-PCR tests.
 In summer 2020, the CIS performed X PCR tests and found only Y positives, lower-bounding the test specificity at.
 \todo{Insert actual numbers of CIS bounding specificity.}
-
-False negatives are more common.
-Early estimates of the sensitivity of PCR tests were around X\%; I produce estimates in \cref{E-ATACCC}.
-False negatives could occur for a variety of reasons, most commonly that an individual swabs themselves poorly and hence the concentration of virus on the swab is too low to be detected by the PCR process.
-Plausibly, a variety of other reasons could also lead to a false negative such as logistical issues or mislabelling of samples.
-\todo{Cite reasons false negative results occur.}
-
-The minimum amount of virus required to be detected by a PCR test is known as the \emph{limit of detection}.
-\todo{Cite limit of detection and explain a bit more about it.}
-
-Viral loads are infrequently measured exactly.
-The most common form of data obtained is cycle threshold (Ct) values from real-time quantitative  reverse-transcriptase polymerase chain reaction (PCR) tests.
-The Ct value is proportional to the negative log of the viral load which is convenient when modelling log viral load as a piecewise linear function, since Ct values themselves are piecewise linear under this assumption.
-When a host's viral load is too low (below the limit of detection of the PCR test), no Ct value is available and hence viral load can be considered censored at the limit of detection.
-
-Ct values cannot be transformed into viral loads (expressed in units such as the number of viral RNA copies per swab) without additional information~\autocite{dahdouhCt,puhachSARSCoV2}.
-Ct values are test-specific, and can even vary between different batches of the same test.
-Testing a series of samples with known viral loads allows the relationship between Ct and viral load to be estimated for a specific setup.
 
@@
 
-The ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts) study was a longitudinal study of contacts reported to the NHS Test and Trace contact tracing service~\autocite{hakkiOnset}.
+The ATACCC study was a longitudinal study of contacts reported to the NHS Test and Trace contact tracing service~\autocite{hakkiOnset}.
 Contacts were invited to participate if they were 5 years or older and  could be reached within 5 days of the index case's symptom onset date.
@@
 To correct for this, various methods have been used to adjust the estimates of prevalence.
+These methods typically also employed some borrowing of information across time because of the low number of daily positives, especially among any specific subpopulation (\eg age group or region).
 I explain the preferred method, MRP (multilevel regression and post-stratification), in \cref{E-backcalc:sec:methods}.

### 2024-02-05 cis-imperfect-testing.tex
edits/comments from Overleaf

--- 2024-01-26 cis-imperfect-testing.tex
@@
 Therefore, this model consists of a linear decline in test sensitivity, before a constant minimum point.
+
+This could be incorporated into the simulation study be generating test results at ...
 Adapting the simulation study to incorporate this test sensitivity is much more similar to the data (see purple histogram in \cref{imperf-test:fig:sim-single-pos}).
@@
 First, if $e_i \leq r_i^{(e)}$, meaning that the test at $r_i^{(e)}+1$ is a true negative and the end of the episode is interval-censored as in the previous chapter.
-In this case, the test at $r_i^{(e)}$ is a true negative, as are all other tests not in $\sched'_i$.
+In this case, the test at $r_i^{(e)} + 1$ is a true negative, as are all other tests not in $\sched'_i$.
 These occur with probability 1, by the assumption of no false positives.
@@
 Note that if $p_\text{sens} = 1$ then $p_{ia}' = p_{ia}$.
+
+Constant terms => proprtional to simpler thing
 
@@
 
-For each simulated dataset, I estimate the survival function using the model proposed in \cref{imperf-test:sec:modelling} and assuming $\psens$ equals 0.6, 0.8, or 1.0.
+For each simulated dataset, I estimate the survival function using the likelihood proposed in \cref{imperf-test:sec:modelling} and assuming $\psens$ equals 0.6, 0.8, or 1.0.
 If the dataset was simulated using a constant $\psens$, and that value of $\psens$ matches the value assumed during inference, I refer to $\psens$ as being correctly specified; otherwise I refer to it as misspecified.
@@
     Each panel shows the results of performing inference with a different assumed value for the test sensitivity.
-    The posterior estimate using a model with a constant test sensitivity of 0.8 (middle panel) is similar to the true value (black line).
+    The posterior estimate using a model with a constant test sensitivity of 0.8 (B) is similar to the true value (black line).
   }

### 2024-02-05 transmission.tex
edits/comments from Overleaf

--- 2024-02-01 transmission.tex
@@
 
-\todo[inline]{Rewrite the intro: eg see Dani's Overleaf comments}
+\todo[inline]{Intro probably needs rewriting, with novelty claims and a bit more context. Revisit once intro chapters written.}
 
-In this chapter, I use the preferred estimate of the duration of RT-PCR positivity (see \cref{E-imperf-test:sec:conclusion}) to estimate the transmission of SARS-CoV-2 in England.
+In this chapter, I use the preferred estimate of the duration of PCR positivity (see \cref{E-imperf-test:sec:conclusion}) to estimate the transmission of SARS-CoV-2 in England.
 This estimation is based only on the CIS data (see \cref{transmission:sec:data} for further details).
@@
 
-\todo[inline]{Consider moving all of this section into chapter 2, where CIS is introduced, and spell it out a bit more in that context}
+\todo[inline]{This is now mostly in chapter 2, double-check there's nothing missed
 
@@
 
-In MRP, a Bayesian multilevel logistic regression model is first fit to predict RT-PCR positivity from demographic covariates.
+In MRP, a Bayesian multilevel logistic regression model is first fit to predict PCR positivity from demographic covariates.
 Then, the posterior distribution is used to predict the prevalence in each stratum.
@@
 % Finally, I poststratify the posterior distribution of incidence to the CIS target population and subpopulations of interest.
+
+The deterministic relationship between incidence proportion and prevalence means that either can be inferred, with the other then calculated.
+Denote the quantity to be inferred at time $t$ is strata $i$ as $\pi_{it}$.
+Note that for any $i$ and $t$, the number of positives will be small, often fewer than 5; therefore some form of borrowing between time and strata is required.
+The statistical model used for inference, including the borrowing of information, should have the following properties.
+\begin{itemize}
+    \item The smoothing model should penalise towards exponential growth.
+        Penalising towards exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time.
+        Furthermore, such penalisation has previously been shown to improve estimates of prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
+\end{itemize}
 
@@
 
-The model was fit separately and independently for each region, implicitly adding all interactions between region and the other variables.
+The model was fitted separately and independently for each region, implicitly adding all interactions between region and the other variables.
 This step assumes that the estimates for each region can be considered independent, which is reasonable because the regions are large with many tests conducted within each.
@@
 The first is the survival matrix, $\matr{S}$.
-For example, \cref{E-imperf-test} showed that it is sensitive to assumptions made.
+For example, \cref{E-imperf-test} showed that it is sensitive to assumptions made, such as the assumed $\psens$ and ....
 The second is the specification of the statistical model used to estimate prevalence, for example the form of the smoothing used.
@@
 Except the 16--25 age group, all age groups show similar trends (\ie parallel lines on a log scale).
-This is the pattern predicted by an age-stratified, mechanistic transmission model (such as the one introduced in \cref{E-SEIR:sec:structured-populations}), if behavioural changes are independent of age group, and there is negligible impact of immunity.
+The distribution of infections among age groups remaining constant over short periods of time is the pattern predicted by standard age-stratified, mechanistic transmission model (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) in equilibrium, if behavioural changes are independent of age group, and there is negligible impact of immunity.
 The incidence proportion in the 16--25 age group decreases faster than other age groups during November.
@@
 The transmission part of the model follows \textcite{birrellRealtime}.
-A novel approach (\cref{SEIR:sec:observation}) links the transmission model to the data.
+A novel approach (\cref{SEIR:sec:observation}) links the transmission model to prevalence data (here from the CIS, but the approach is generalisable).
 In the mechanistic modelling literature, this part of the model is referred to as the observation model; however, in the terminology of \cref{E-inc-prev}, it includes both the prevalence process and the observation process.
@@
 
-% Here, I extend the system of ODEs to include compartments representing RT-PCR positive individuals.
-% The proportion of RT-PCR positive individuals in each strata is then linked to the CIS using a Beta-Binomial likelihood, with the modelled proportion that are RT-PCR positive as the mean of the Beta distribution.
-One approach would be to convolve the incidence based on the transmission model with the RT-PCR duration distribution.
+% Here, I extend the system of ODEs to include compartments representing PCR positive individuals.
+% The proportion of PCR positive individuals in each strata is then linked to the CIS using a Beta-Binomial likelihood, with the modelled proportion that are PCR positive as the mean of the Beta distribution.
+One approach would be to convolve the incidence based on the transmission model with the PCR duration distribution.
 Here, the daily incidence proportion for each stratum is by solving the system of ODEs governing the transmission model, and then prevalence calculated as using \cref{E-inc-prev:eq:EPt}.
@@
 
-I link the modelled proportion of individuals that are RT-PCR positive to the data using a beta-binomial likelihood (as discussed in \cref{transmission:sec:data}).
+I link the modelled proportion of individuals that are PCR positive to the data using a beta-binomial likelihood (as discussed in \cref{transmission:sec:data}).
 If on each day $t$, each strata $i$ is observed to have $y_{it}$ positives out of $n_{it}$ tests, then the likelihood is:
@@
 }
-  \caption[SEIR model]{SEIR model used for the analysis in this section. The shaded box represents RT-PCR-positive individuals. Shown for a single stratum.}
+  \caption[SEIR model]{SEIR model used for the analysis in this section. The shaded box represents PCR-positive individuals. Shown for a single stratum.}
   \label{SEIR:fig:full-model}
@@
 This follows directly from the ODEs if $\vec{s}$ is constant. 
-The growth rate $\psi$ is easily identifiable from the data, and therefore a parameterisation in terms of $\phi$ should be more computationally efficient.
+The growth rate $\psi$ is easily identifiable from the data, and therefore a parameterisation in terms of $\psi$ should be more computationally efficient.
 If this is the initial exponential phase, when $\vec{s} \approx 1$, then the basic reproduction number for a SEIR model is linked to this growth rate by~\autocites{birrellBayesian}{wearingAppropriate}:
@@
     \vspace{-3cm}
+    \todo[inline]{Add (A) incidence to the top panel and (B) prevalence to the bottom panel}
     \makebox[\textwidth][c]{\includegraphics[width=0.9\paperwidth]{SEIR/sim/predictive_coverage}}
@@
         Goodness-of-fit to prevalence data by region and age for the younger three age groups (see \cref{SEIR:fig:prev-old} for the others).
-        The central ribbon and line shows the posterior median and 95\% CrI estimate for the proportion of the relevant strata RT-PCR positive on each day.
+        The central ribbon and line shows the posterior median and 95\% CrI estimate for the proportion of the relevant strata PCR positive on each day.
         The lighter outer ribbon is the 95\% posterior predictive interval for the proportion of test results that are positive each week.
@@
 This can explain that similar patterns emerge for $\beta_w$ in different regions (see \cref{SEIR:fig:beta-walk}), despite these being estimated independently (\ie there is no model structure which favours correlated values).
-These correlations could be exploited to improve the precision of estimates with a more complex model which borrows information across regions.
+These correlations could be exploited to improve the precision of estimates with a more complex model which borrows information across regions, \eg a hierarchical model.
 
@@
         Posterior median and 95\% CrI shown.
-        For the SEIR model, newly RT-PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the backcalculation results.
+        For the SEIR model, newly PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the backcalculation results.
     }

### 2024-02-05 biology-data.tex
finish initial write of chapter 2

@@
 
-\todo[inline]{I should maybe swap 2.1 and 2.2 order.
-I like the current order because natural history feels more basic and introductory, but it does reference a bunch of stuff about PCR testing so it might be easier to follow the other way round.}
-
 \todo[inline]{Rewrite chapter 2 intro after writing thesis intro because it will be clearer how to divide content.}
@@
 RNA viruses use RNA, rather than DNA, as their genetic material.
-
-\section{Natural history of the disease} \label{biology-data:sec:natural-history}
-
-\begin{figure}
-  \makebox[\textwidth][c]{
-  \begin{tikzpicture}[
-      node distance = 4cm,
-      on grid,
-      auto,
-      ->,>=stealth',
-      every state/.style={draw,rectangle,align=center},
-      ]
-
-      \node[state] (pre) {PCR negative \\ Not infectious};
-      \node[state, right of=pre] (pos) {PCR positive \\ Not infectious};
-      \node[state, right of=pos] (infectious) {PCR positive \\ Infectious};
-      \node[state, right of=infectious] (not-infectious) {PCR positive \\ Not infectious};
-      \node[state, right of=not-infectious] (neg) {PCR negative \\ Not infectious};
-
-      \node[draw=none, below=2cm of pre] (infected) {} ;
-      \node[draw=none, below=2cm of neg] (recovered) {} ;
-
-      \node[above=2cm of pre] (titleA) {(A)};
-      \node[below=1cm of infected] (titleB) {(B)};
-
-      % \draw (0,-2) -- node {Infected} (pre);
-      % \draw (not-infectious) -- (0,-2);
-
-      \path (infected) edge node {Infected} (pre)
-          (pre) edge (pos)
-            (pos) edge (infectious)
-            (infectious) edge (not-infectious)
-            (not-infectious) edge (neg)
-            (neg) edge (recovered);
-
-    % Detectable
-    \node[below = 1cm of infectious] (detect-label) {Detectable};
-    \node[draw, thick, inner sep=0.3cm, fit=(pos) (infectious) (not-infectious) (detect-label), fill=blue!10,opacity=0.2] (detect-box) {};
-  \end{tikzpicture}
-  }
-  \centering\includegraphics[width=\textwidth]{biology-data/natural-history}
-  \caption[Natural history of SARS-CoV-2.]{%
-    Natural history of SARS-CoV-2.
-    (A) A simplified state-based model, which I generally use in this thesis.
-    Shading represents states where the individual is detectable by a PCR test.
-    (B) A full model of the natural history, including continuous processes (Reproduced from \textcite{puhachSARSCoV2} with permission; durations are based on averages, there is a large amount of individual variation).
-    PCR positive refers to the main diagnostic test used in this thesis, and detect RNA genomes (see \cref{biology-data:sec:PCR} for more details).
-    Ag-RDT (antigen rapid diagnostic test) tests are lateral flow tests, which detect infectious virus rather than RNA, and are not used in this thesis.
-  }
-  \label{biology-data:fig:natural-history}
-\end{figure}
-
-An individual infected with SARS-CoV-2 goes through several disease stages (see \cref{biology-data:fig:natural-history}).
-\todo{Find citations for disease stages.}
-First, they are exposed to the virus and become infected.
-At this initial stage they are not infectious (cannot spread the disease to others), do not have symptoms, and will not test positive on a PCR test.
-After a few days, they will start testing positive on a PCR test; this is normally prior to being infectious or having symptoms.
-Next, the individual becomes infectious; normally, this occurs prior to symptom onset for those who experience symptoms.
-Recovery from symptoms occurs after a few days, around the time the individual is no longer infectious.
-Finally, the individual will test negative on a PCR test.
-
-\todo[inline]{Maybe move the following paragraph because it is referencing forward a bunch of concepts not yet introduced properly, or if sections 2.1 and 2.2 are swapped it starts making sense.}
-I will refer to individuals between the time they start testing positive on a PCR test and the time they stop testing positive on a PCR test as \emph{detectable}.
-It is not possible to directly observe whether an individual is detectable or not due to noise in the PCR test (see \cref{biology-data:sec:PCR}).
-However, viewing the number of RNA genome copies in an individual as a latent process this can then be defined as the time when the number of RNA genome copies is above the limit of detection of the PCR test.
-While this definition is not perfectly precise, it is clear enough for the purposes it used in this thesis.
-
-The time for each of these stages depends on various features of host and virus characteristics.
-In this thesis, I focus on pre-Alpha variants of SARS-CoV-2 in unvaccinated individuals.
-The time from infection to infectiousness is known as the \emph{latent period} and is typically around\todo{cite latent period}.
-The time from infection to experiencing symptoms is known as the \emph{incubation period} and is typically around\todo{cite incubation period}.
-The incubation period is undefined for asymptomatic individuals, around\todo{cite proportion asymptomatic} of those infected.
-
-The above description gives three possible ways to define an individual who is infected with SARS-CoV-2: based on their PCR test result, whether they are infectious, or whether they are symptomatic.
-In this thesis, I adopt a definition of infection based on PCR test results.
-This is because it coincides with the data collected on the studies I use (see \cref{biology-data:sec:studies}).
-
-A host's \emph{viral load}, the quantity of viral particles in a host, is the underlying biological process determining many of these changes~\autocite{puhachSARSCoV2}.
-Any host's viral load is determined by the interaction of many biological processes, most prominently the virus's replication increasing the viral load and the host's immune response reducing the viral load~\autocite{keVivo}.
-Viral replication requires the virus to infect a host cell and then use the host cell's machinery to produce more virus; as more host cells are infected this process slows because there are fewer remaining host cells, known as \emph{target cells}, to infect.
-The host's immune system can slow this process by various mechanisms; the most relevant for SARS-CoV-2 is by making some target cells less susceptible to infection.
-A host's immune response is triggered by the viral load, and becomes stronger as the viral load increases.
-A high viral load is associated with being more infectious, more likely to test positive on a PCR test, and the start of symptoms~\autocite{puhachSARSCoV2,keVivo}.
-
-% Possible summaries
-% \begin{enumerate}
-%   \item \url{https://www.nature.com/articles/s41579-022-00822-w/figures/2}
-%   \item \url{https://www.nature.com/articles/s41576-021-00360-w/figures/1}
-%   \item \url{https://academic.oup.com/view-large/figure/305617142/ciaa1442_fig2.jpg}
-% \end{enumerate}
 
@@
 
-Quantitative real-time PCR (or RT-PCR) gives a measurement of the amount of the target DNA sequence present in the original sample~\autocite{yangPCRdiagnostics}.
+Quantitative real-time (RT-)PCR gives a measurement of the amount of the target DNA sequence present in the original sample~\autocite{yangPCRdiagnostics}.
 After each PCR cycle, the amount of target DNA is measured and compared to a pre-derived standard.
@@
 The Ct value decreases as the amount of DNA present in the original sample increases.
+\todo{cite interpretation of Ct values}
 If there is more RNA present, more DNA is created, and hence fewer cycles are required to reach detection.
@@
 
-\todo[inline]{%
-  Rewrite this section on imperfect testing.
-  Including analytical sensitivity: the minimum number of copies of the target RNA material in the sample that the test can detect, measured using the 
-}
 Like any diagnostic test, PCR tests are not perfect.
@@
 False positives are very rare for SARS-CoV-2 RT-PCR tests.
-In summer 2020, the CIS performed X PCR tests and found only Y positives, lower-bounding the test specificity at.
-\todo{Insert actual numbers of CIS bounding specificity.}
+In summer 2020, the CIS performed 208,730 PCR tests and found only 159 positives, lower-bounding the test specificity (assuming all the positives are false positives) at 99.9\%~\autocite[section 5]{cisMethodsONS}.
 
@@
 The gold standard procedure is attempting to grow and recover viral material from a sample, a process known as \emph{cell culturing}~\autocite{singanayagamDuration,puhachSARSCoV2,hakkiOnset}.
+Further details on cell culturing are beyond the scope of this thesis.
 
@@
 The first study, ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts) performed daily testing of a small number of individuals with limited follow-up.
-The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability, for but the results are limited because the small sample size and short follow-up means that the tails of these distributions cannot be estimated.
+The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability, but the results are limited because the small sample size and short follow-up means that the tails of these distributions cannot be estimated.
 The second study, the CIS (Coronavirus (COVID-19) Infection Survey), performed weekly or monthly testing of a large number of individuals with long follow-up.
@@
 In particular, the sparse sampling means that the vast majority of detected infections had only one or two positive tests, and hence the viral load trajectory of these infections cannot be estimated.
+
+\section{Natural history of the disease} \label{biology-data:sec:natural-history}
+
+\begin{figure}
+  \makebox[\textwidth][c]{
+  \begin{tikzpicture}[
+      node distance = 4cm,
+      on grid,
+      auto,
+      ->,>=stealth',
+      every state/.style={draw,rectangle,align=center},
+      ]
+
+      \node[state] (pre) {RT-PCR negative \\ Not infectious};
+      \node[state, right of=pre] (pos) {RT-PCR positive \\ Not infectious};
+      \node[state, right of=pos] (infectious) {RT-PCR positive \\ Infectious};
+      \node[state, right of=infectious] (not-infectious) {RT-PCR positive \\ Not infectious};
+      \node[state, right of=not-infectious] (neg) {RT-PCR negative \\ Not infectious};
+
+      \node[draw=none, below=2cm of pre] (infected) {} ;
+      \node[draw=none, below=2cm of neg] (recovered) {} ;
+
+      \node[above=2cm of pre] (titleA) {(A)};
+      \node[below=1cm of infected] (titleB) {(B)};
+
+      % \draw (0,-2) -- node {Infected} (pre);
+      % \draw (not-infectious) -- (0,-2);
+
+      \path (infected) edge node {Infected} (pre)
+          (pre) edge (pos)
+            (pos) edge (infectious)
+            (infectious) edge (not-infectious)
+            (not-infectious) edge (neg)
+            (neg) edge (recovered);
+
+    % Detectable
+    \node[below = 1cm of infectious] (detect-label) {Detectable};
+    \node[draw, thick, inner sep=0.3cm, fit=(pos) (infectious) (not-infectious) (detect-label), fill=blue!10,opacity=0.2] (detect-box) {};
+  \end{tikzpicture}
+  }
+  \centering\includegraphics[width=\textwidth]{biology-data/natural-history}
+  \caption[Natural history of SARS-CoV-2.]{%
+    Natural history of SARS-CoV-2.
+    (A) A simplified state-based model, which I generally use in this thesis.
+    Shading represents states where the individual is detectable by a RT-PCR test.
+    (B) A full model of the natural history, including continuous processes (Reproduced from \textcite{puhachSARSCoV2} with permission; durations are based on averages, there is a large amount of individual variation).
+    PCR (strictly RT-PCR) positive refers to the main diagnostic test used in this thesis, and detect RNA genomes (see \cref{biology-data:sec:PCR} for more details).
+    Ag-RDT (antigen rapid diagnostic test) tests are lateral flow tests, which detect infectious virus rather than RNA, and are not used in this thesis.
+  }
+  \label{biology-data:fig:natural-history}
+\end{figure}
+
+An individual infected with SARS-CoV-2 goes through several disease stages, depicted in \cref{biology-data:fig:natural-history}~\autocite{puhachSARSCoV2}.
+First, they are exposed to the virus and become infected.
+At this initial stage they are not infectious (cannot spread the disease to others), do not have symptoms, and will not test positive on a RT-PCR test.
+After a few days, they will start testing positive on a RT-PCR test; this is normally prior to being infectious or having symptoms.
+Next, the individual becomes infectious; normally, this occurs prior to symptom onset for those who experience symptoms.
+Recovery from symptoms occurs after a few days, around the time the individual is no longer infectious.
+Finally, the individual will test negative on a RT-PCR test.
+
+The time for each of these stages depends on various features of host and virus characteristics.
+In this thesis, I focus on pre-Alpha variants of SARS-CoV-2 in unvaccinated individuals.
+The time from infection to infectiousness is known as the \emph{latent period}.
+The length of the latent period is hard to estimate because neither the time of infection or the start of infectiousness can be observed; estimates for the mean length are generally 3--4 days~\autocite{zhaoEstimating}.
+The time from infection to experiencing symptoms is known as the \emph{incubation period} and is estimated at have a mean length of around 6 days, with some sensitivity to the sampling and analysis technique used~\autocite{wuIncubation,quesadaIncubation,aleneSerial}.
+The incubation period is undefined for asymptomatic individuals of those infected.
+
+The above description gives three possible ways to define an individual who is infected with SARS-CoV-2: based on their RT-PCR test result, whether they are infectious, or whether they are symptomatic.
+In this thesis, I adopt a definition of infection based on RT-PCR test results.
+This is because it coincides with the data collected on the studies I use (see \cref{biology-data:sec:studies}).
+
+A host's \emph{viral load}, the quantity of viral particles in a host, is the underlying biological process determining many of these changes~\autocite{puhachSARSCoV2}.
+Any host's viral load is determined by the interaction of many biological processes, most prominently the virus's replication increasing the viral load and the host's immune response reducing the viral load~\autocite{keVivo}.
+Viral replication requires the virus to infect a host cell and then use the host cell's machinery to produce more virus; as more host cells are infected this process slows because there are fewer remaining potential host cells, known as \emph{target cells}, to infect.
+The host's immune system can slow this process by various mechanisms; the most relevant for SARS-CoV-2 is by making some target cells less susceptible to infection.
+A host's immune response is triggered by the viral load, and becomes stronger as the viral load increases.
+A high viral load is associated with being more infectious, more likely to test positive on a RT-PCR test, and the start of symptoms~\autocite{puhachSARSCoV2,keVivo}.
+
+% Possible summaries
+% \begin{enumerate}
+%   \item \url{https://www.nature.com/articles/s41579-022-00822-w/figures/2}
+%   \item \url{https://www.nature.com/articles/s41576-021-00360-w/figures/1}
+%   \item \url{https://academic.oup.com/view-large/figure/305617142/ciaa1442_fig2.jpg}
+% \end{enumerate}
+
+\todo[inline]{Not sure if this is the best location for this important paragraph defining the concept of ``detectability''}
+I will refer to individuals between the time they start testing positive on a RT-PCR test and the time they stop testing positive on a RT-PCR test as \emph{detectable}.
+It is not possible to directly observe whether an individual is detectable or not due to noise in the RT-PCR test (see \cref{biology-data:sec:PCR}).
+However, viewing the number of RNA genome copies in an individual as a latent process this can then be defined as the time when the number of RNA genome copies is above the limit of detection of the RT-PCR test.
+While this definition is not perfectly precise, it is clear enough for the purposes it used in this thesis.
 
@@
 For 115 contacts with at least one positive RT-PCR test, the growth phase of viral load was not captured; these contacts were excluded from this analysis to ensure all individuals viral load trajectories could be identified.
-More details on this analysis and my extensions to it can be found in \cref{E-ATACCC}.
+I will revisit and extend this analysis in \cref{E-ATACCC}.
 
@@
 
-\todo[inline]{Do I need to distinguish between estimates of positivity and prevalence?}
-
 The CIS was setup in April 2020 to provide gold-standard measurements of the prevalence of SARS-CoV-2 in the community.
 It is a longitudinal study with a representative sample of private households, this excludes settings such as care homes and student halls.
+It is globally unique in providing a representative, longitudinal, and large-scale study.
 Initially, it was limited to England, but expanded to cover the whole of the UK in September 2020.
@@
 In particular, some demographic groups, such as young adults and individuals of non-white ethnicity, were underrepresented among those who responded~\autocite{pouwelsCommunity}.
-To correct for this, various methods have been used to adjust the estimates of prevalence.
-These methods typically also employed some borrowing of information across time because of the low number of daily positives, especially among any specific subpopulation (\eg age group or region).
-I explain the preferred method, MRP (multilevel regression and post-stratification), in \cref{E-backcalc:sec:methods}.
 
-In this thesis, I focus on the period from August 2020 until January 2021.
+In this thesis, I focus on the period from Mon 31st Aug 2020 until Sun 24th Jan 2021 (inclusive).
+This period is chosen because there were very few positives in CIS in summer 2020 due to low prevalence and a smaller sample size, and ends when vaccination crossed 10\% of England's population.
+Vaccination complicates the dynamics and immune response, which would require more complex models which are more specific to SARS-CoV-2 beyond the scope of this thesis, which primarily intends to show the potential of a study such as CIS in general.
+\todo{make sure I justify the period choice in the intro when introducing the thesis's scope and aims}
+
 During the start of this period, the recruitment rate of CIS was expanded (see \cref{biology-data:fig:CIS-recruitment}(A)).
@@
 \end{figure}
+% \begin{figure}
+%   \centering \includegraphics[width=\textwidth]{biology-data/CIS-num-tests}
+%   \caption[Number of CIS tests]{%
+%     Daily number of CIS tests conducted in the period considered in this thesis (Sep 2020 until Jan 2021).
+%   }
+%   \label{biology-data:fig:CIS-num-tests}
+%   \todo[inline]{Cut this repetition of number of tests? Maybe highlight period of interest in previous plot}
+% \end{figure}
+
+\subsubsection{Estimating positivity rate}
+
+The \emph{positivity rate} is the proportion of people in a given population that (if they were tested) would test positive for SARS-CoV-2 at a defined point in time; in this context test means an RT-PCR test~\autocite{cisMethodsONS}.
+A notable feature of this definition is that it explicitly does not correct for false positive or negative results.
+Generally, the point in time of interest is a specific day, although it has also been applied to a week~\autocite{cisMethodsONS,pouwelsMRPvaccination,pouwelsCommunity}.
+Larger time periods (\ie weeks instead of days) allow more granular estimates on other variables, most notably geographic area.
+
+If the CIS was truly a random sample, an unbiased estimator of the positivity rate would be the proportion of positive tests in the sample; I refer to this as the uncorrected positivity rate.
+However, in practice, two issues arise with this.
+First, the non-response bias discussed previously.
+Second, the number of positives on any day is low, especially among any specific subpopulation (\eg age group or region).
+Statistical methods are used to correct for these issues.
+
+The statistical preferred method for estimating the positivity rate is MRP (multi-level regression and poststratification)~\autocite{cisMethodsONS,pouwelsCommunity}.
+The poststratification step is used to correct for non-response bias.
+To address the low number of positives, MRP uses a hierarchical Bayesian model to borrow information across time and other covariates.
+I explain the details of this model, and how I use it for estimating incidence in \cref{E-backcalc:sec:methods}
+
+The outputs of this model compared to the raw positivity rate are shown in \cref{biology-data:fig:CIS-positivity}.
+Both show similar trends
+The MRP estimate tends to increase the estimated positivity rate, compared to uncorrected positivity rate, because underrepresented groups tend to have higher positivity.
 
@@
   \centering \includegraphics[width=\textwidth]{biology-data/CIS-positivity}
-  \caption[CIS prevalence]{%
-    CIS prevalence in the period considered in this thesis (Sep 2020 until Jan 2021).
+  \caption[CIS positivity]{%
+    CIS positivity rate in the period considered in this thesis (Sep 2020 until Jan 2021).
     Dots show raw proportion of non-void tests that are positive on each day (days with fewer than 100 tests excluded for readability).
-    Ribbon shows modelled 95\% CrI for the prevalence, corrected for non-response bias and smoothed over time (using the methodology described in \cref{E-backcalc:sec:methods}).
-    The correction tends to increase the estimated prevalence because underrepresented groups tend to have higher prevalence.
+    Ribbon shows modelled 95\% CrI for the positivity, corrected for non-response bias and smoothed over time (using the methodology described in \cref{E-backcalc:sec:methods}).
   }
@@
 \end{figure}
-\begin{figure}
-  \centering \includegraphics[width=\textwidth]{biology-data/CIS-num-tests}
-  \caption[Number of CIS tests]{%
-    Daily number of CIS tests conducted in the period considered in this thesis (Sep 2020 until Jan 2021).
-  }
-  \label{biology-data:fig:CIS-num-tests}
-  \todo[inline]{Cut this repetition of number of tests? Maybe highlight period of interest in previous plot}
-\end{figure}
+
+Interpreting positivity, or prevalence, directly is difficult because it is a delayed and smoothed measure of incidence (see \cref{inc-prev:sec:prevalence-process}).
+The major aim of this thesis is the ability to produce estimates of incidence, which can be interpreted in terms of population behaviour.
+
+As well as estimating the positivity rate for the whole of England (and the devolved nations), the CIS also estimates the positivity rate for various subpopulations, in particular geographic areas and age groups.
+The exact specification of these have varied over time.
+
+The geographic areas I will adopt are the nine standard regions of England (International Territorial Level 1, previously Nomenclature of Territorial Units for Statistics 1); these are: North East, North West, Yorkshire and the Humber (commonly abbreviated to Yorkshire), East Midlands, West Midlands, East, London, South East, and South West~\autoref{onsRegions}.
+These are the geographic areas used throughout the major CIS analyses in England, although sometimes more granular areas were defined~\autocite[e.g.][]{pouwelsMRPvaccination,pouwelsCommunity,cisMethodsONS,houseInferring,walkerTracking}.
+
+The age groups used have varied slightly for different CIS analyses~\autocite[e.g.][]{pouwelsMRPvaccination,pouwelsCommunity,cisMethodsONS,houseInferring,walkerTracking}.
+I will use the age groups: 0--10 (pre-school and primary school), 11--15 (compulsory secondary school), 16--24 (adolescents and young adults who may remain in formal education), 25--49 (younger working age adults), 50--69 (older working age adults), and 70+ (elderly).
+
+The household-based design of the CIS means that the data is clustered.
+In particular, individuals in the same household are likely to transmit to each other and hence have correlated test results.
+Other behavioural or genetic factors could also contribute to individuals in the same household being more similar than randomly selected individuals in the population.
+The positivity analyses use a beta-binomial likelihood (defined in \cref{E-distributions}) to account for this.
+
+The beta-binomial distribution is overdispersed relative to the binomial distribution.
+Overdispersion allows for household clustering\todo{cite beta-binomial as a model for clustering} and other overdispersion in the data, which is common in epidemiological data\todo{cite that epi data is often overdispersed}.
+The parameterisation of the beta-binomial that I use (detailed in \cref{E-distributions}) has an overdispersion parameter $\rho$, regarded as a nuisance parameter in this context.
+$\rho=0$ means the beta-binomial coincides with the binomial distribution.
+As $\rho \to \infty$, the distribution tends towards the uniform distribution on $[0, 1]$~\autocite{hughesUsing}.
 
@@
 
-For the work in \cref{E-cis-perf-testing,E-cis-imperf-testing}, positive test results need to be grouped into \emph{infection episodes}.
-An infection episode is a series of positive tests that are part of the same infection.
+Positive test results need to be grouped into \emph{infection episodes}.
+An infection episode is a series of positive tests that originate from a single transmission event.
 I use a method developed throughout the pandemic, and published within \textcite{weiRisk}; see that publication for full justification of the criteria.
@@
 
-% Whether a future positive is part of the same episode of a reinfection is not always trivial; I rely on a process developed previously by Sarah Walker, see \cref{E-episode-def}.
+A sample of 500 randomly selected infection episodes (selected from the episodes used in the analyses in this thesis) have their test results.
+As can be seen, the number of tests within an episode, their length, and other characteristics vary greatly.
+\begin{figure}
+  \thisfloatpagestyle{empty}
+  \vspace{-5cm}
+  \makebox[\textwidth][c]{\includegraphics[height=.9\paperheight]{biology-data/STATS13734/data_viz}}
+  \caption[CIS infection episodes]{%
+    500 randomly selected infection episodes.
+    Shown are the negative tests before and after the outermost positive tests associated with the episode.
+    Red dots are negative tests, green dots are positive tests, and blue dots are void tests (where the result could not be determined; excluded from other analyses).
+
+    This figure has been provided as management information for operational planning purposes.
+They are provisional estimates circulated before public release.
+This management information should not be shared widely and should only be used for operational planning purposes.
+This information is not to be used publicly and ahead of being released publicly by ONS.
+  }
+\end{figure}
+  \todo{Work out what to do with this figure....}
 

### 2024-02-05 transmission.tex

--- 2024-02-02 transmission.tex
@@
 
-\todo[inline]{Rewrite the intro: eg see Dani's Overleaf comments}
+\todo[inline]{Intro probably needs rewriting, with novelty claims and a bit more context. Revisit once intro chapters written.}
 
@@
 This follows directly from the ODEs if $\vec{s}$ is constant. 
-The growth rate $\psi$ is easily identifiable from the data, and therefore a parameterisation in terms of $\phi$ should be more computationally efficient.
+The growth rate $\psi$ is easily identifiable from the data, and therefore a parameterisation in terms of $\psi$ should be more computationally efficient.
 If this is the initial exponential phase, when $\vec{s} \approx 1$, then the basic reproduction number for a SEIR model is linked to this growth rate by~\autocite{birrellBayesian,wearingAppropriate}:

### 2024-02-05 biology-data.tex
proof read

@@
 Further details on cell culturing are beyond the scope of this thesis.
-
-\section{Studies used in this thesis} \label{biology-data:sec:studies}
-
-In this thesis, I make use of data from two studies which recruited participants from the general population in the UK.
-Unlike many studies of SARS-CoV-2, these studies are not focused on specific groups such as healthcare workers or hospital patients.
-Therefore, the results from these studies are more generalisable to the entire UK population.
-The first study, ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts) performed daily testing of a small number of individuals with limited follow-up.
-The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability, but the results are limited because the small sample size and short follow-up means that the tails of these distributions cannot be estimated.
-The second study, the CIS (Coronavirus (COVID-19) Infection Survey), performed weekly or monthly testing of a large number of individuals with long follow-up.
-The large sample size and long follow-up allows for reliable estimates of the tails of the distributions of the quantities of interest, but the results are limited because the sparse sampling means that the viral load dynamics cannot be estimated.
-In particular, the sparse sampling means that the vast majority of detected infections had only one or two positive tests, and hence the viral load trajectory of these infections cannot be estimated.
 
@@
 
+\section{Studies used in this thesis} \label{biology-data:sec:studies}
+
+In this thesis, I use data from two studies which recruited participants from the general population in the UK.
+Unlike many studies of SARS-CoV-2, these studies are not focused on specific groups such as healthcare workers or hospital patients.
+Therefore, the results from these studies are more generalisable to the entire UK population.
+
+\todo[inline]{Some of this might be moved into the introductory chapter motivating each part of the thesis.}
+
+The first study, ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts), performed daily testing of a small number of individuals with short follow-up.
+The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability, but the results are limited because the small sample size and short follow-up means that the tails of these distributions cannot be estimated.
+
+The second study, the CIS (Coronavirus (COVID-19) Infection Survey), performed weekly or monthly testing of a large number of individuals with long follow-up.
+The large sample size and long follow-up allows for reliable estimates of the tails of the distributions of the quantities of interest.
+However, the results are limited because the sparse sampling which means that the viral load dynamics cannot be estimated.
+In particular, the vast majority of detected infections had only one or two positive tests, and hence the viral load trajectory of these infections cannot be estimated.
+
 \subsection{Assessment of Transmission and Contagiousness of COVID-19 in Contacts}
@@
 Generally, the point in time of interest is a specific day, although it has also been applied to a week~\autocite{cisMethodsONS,pouwelsMRPvaccination,pouwelsCommunity}.
-Larger time periods (\ie weeks instead of days) allow more granular estimates on other variables, most notably geographic area.
+Larger time periods (\ie weeks instead of days) allow more granular estimates in other dimensions, most commonly geography.
 
@@
 The outputs of this model compared to the raw positivity rate are shown in \cref{biology-data:fig:CIS-positivity}.
-Both show similar trends
-The MRP estimate tends to increase the estimated positivity rate, compared to uncorrected positivity rate, because underrepresented groups tend to have higher positivity.
+Both show similar trends.
+The MRP estimate most often increases the estimated positivity rate, compared to uncorrected positivity rate, because the underrepresented demographic groups tended to have higher positivity rates.
 
@@
 
-Interpreting positivity, or prevalence, directly is difficult because it is a delayed and smoothed measure of incidence (see \cref{inc-prev:sec:prevalence-process}).
+Interpreting positivity, or prevalence, directly is difficult because it is a delayed and smoothed measure of incidence (see \cref{E-inc-prev:sec:prevalence-process}).
 The major aim of this thesis is the ability to produce estimates of incidence, which can be interpreted in terms of population behaviour.
@@
 An infection episode is a series of positive tests that originate from a single transmission event.
-I use a method developed throughout the pandemic, and published within \textcite{weiRisk}; see that publication for full justification of the criteria.
-Denote the time of a positive test by $t$, the number of consecutive negative tests prior to $t$ by $n$, and the time of the first positive test in the same individual by $t_p$ ($t_p = -\infty$ if there is no previous positive test in the same individual).
-The test at $t$, in the pre-Omicron era (prior to December 2021), is defined as the start of a new infection episode if any of the following apply.
+I use a heuristic-based algorithm developed throughout the pandemic, and published within \textcite{weiRisk}; see that publication for full justification of the criteria.
+
+The algorithm determines whether a new positive tests, at time $t$, is part of the previous infection episode in the same individual, where that previous infection episode had its first positive test at $t_p$.
+If the individual has no previous positive tests, the new positive test is always the start of a new infection episode.
+Otherwise, the test at $t$, in the pre-Omicron era (prior to December 2021), is defined as the start of a new infection episode if any of the following apply.
 \begin{enumerate}
-  \item $t_p = -\infty$.
   \item $n \geq 1$ and $t - t_p > 120$.
@@
 Intermittent negatives are the clearest example of a false negative.
-An intermittent negative is when an individual tests negative but tested positive previously and subsequently.
-Intermittent negatives are stripped out when creating the dataset used for all duration analyses in this chapter, but demonstrate that false negatives do occur.
+I will generally consider that intermittent negatives are uninformative on the duration of detectability, and hence exclude them from several analyses.
 
-A sample of 500 randomly selected infection episodes (selected from the episodes used in the analyses in this thesis) have their test results.
+A sample of 500 randomly selected infection episodes (selected from the episodes used in the analyses in this thesis) have their test results shown in \cref{biology-data:fig:episodes}.
 As can be seen, the number of tests within an episode, their length, and other characteristics vary greatly.
@@
   }
+  \label{biology-data:fig:episodes}
 \end{figure}

### 2024-02-05 biology-data.tex
Improve chapter 2 figure

@@
 
-The CIS was setup in April 2020 to provide gold-standard measurements of the prevalence of SARS-CoV-2 in the community.
+The CIS~\autocite{CIS} was setup in April 2020 to provide gold-standard measurements of the prevalence of SARS-CoV-2 in the community.
 It is a longitudinal study with a representative sample of private households, this excludes settings such as care homes and student halls.
@@
   \vspace{-5cm}
-  \makebox[\textwidth][c]{\includegraphics[height=.9\paperheight]{biology-data/STATS13734/data_viz}}
+  \makebox[\textwidth][c]{\includegraphics[height=.7\paperheight]{biology-data/STATS13734/data_viz}}
   \caption[CIS infection episodes]{%

### 2024-02-05 cis-imperfect-testing.tex

@@
 
-This could be incorporated into the simulation study be generating test results at ...
+This could be incorporated into the simulation study be generating test results at ...\todo{finish sentence}
 Adapting the simulation study to incorporate this test sensitivity is much more similar to the data (see purple histogram in \cref{imperf-test:fig:sim-single-pos}).

### 2024-02-07 transmission.tex
chapter 6 comment

@@
 In this chapter, I use the preferred estimate of the duration of RT-PCR positivity (see \cref{E-imperf-test:sec:conclusion}) to estimate the transmission of SARS-CoV-2 in England.
-This estimation is based only on the CIS data (see \cref{transmission:sec:data} for further details).
+This estimation is based only on the CIS data (as shown in \cref{E-biology-data:sec:positivity} for further details).
 I will show that this data is sufficient to estimate the parameters of the transmission process, suggesting surveys of this type are sufficient for surveillance of an epidemic.
@@
 
-\todo[inline]{This is now mostly in chapter 2, double-check there's nothing missed
+\todo[inline]{This is now in \cref{E-biology-data:sec:posivity})}
 
@@
 Note that for any $i$ and $t$, the number of positives will be small, often fewer than 5; therefore some form of borrowing between time and strata is required.
-The statistical model used for inference, including the borrowing of information, should have the following properties.
+The statistical model and inference procedure, including the borrowing of information, should have the following properties.
 \begin{itemize}
@@
         Furthermore, such penalisation has previously been shown to improve estimates of prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
+    \item Limit estimates to epidemiologically and biologically plausible values.
+        That is, incidence proportion and prevalence should be in the interval $[0, 1]$.
+        While incidence proportion can theoretically be greater than 1, this would imply that some individuals were infected multiple times in a single time unit (day) which is not biologically plausible.
+    \item Have sufficient parameters to incorporate all necessary covariates to correct for non-response bias (see \cref{E-biology-data:sec:positivity}).
+    \item Computationally efficient.
+        The model will be fit to granular data and therefore needs to be run within the SRS, with its computational limitations (see\todo{ref section on SRS computational limitations}).
+        In particular, MCMC should be avoided because incorporating sufficient covariates to account for all groups led to a model that was too computationally expensive~\citePersonalComms{Koen Pouwels}.
 \end{itemize}

### 2024-02-07 biology-data.tex

@@
 
-\subsubsection{Estimating positivity rate}
+\subsubsection{Estimating positivity rate} \label{biology-data:sec:posivity}
 
@@
 \end{figure}
-  \todo{Work out what to do with this figure....}
+
+\todo[inline]{%
+    From chapter 7 add info on: how this is aggregated, SRS limitations, definitions for each episode (sketched below).
+}
+
+The beginning of episode $i$ is known to occur in the interval $[l_i^{(b)}, r_i^{(b)}]$, and similarly for the end of the infection in $[l_i^{(e)}, r_i^{(e)}]$.
 

### 2024-02-07 cis-imperfect-testing.tex

--- 2024-02-05 cis-imperfect-testing.tex
@@
 
-\todo[inline]{The first paragraph needs restructuring. Dani will send some ideas.}
-In \cref{E-perf-test}, a framework for analysing the CIS to produce estimates of the duration of positivity was developed.
-However, initial application of this framework produced implausibly short estimates; I identified the issue as being due to the presence of false negatives, a form of misclassification bias, within the CIS (\cref{imperf-test:sec:problem}).
-There exists little prior work on the incorporation of misclassification bias within similar survival analysis frameworks.
-\todo[inline]{Review literature properly, here or elsewhere}
-In this chapter, I modify the previous model to include false negatives (\cref{imperf-test:sec:modelling}).
-I show in simulation that this model will recover the duration distribution (\cref{imperf-test:sec:sim-study-results}) and apply it to the CIS data (\cref{imperf-test:sec:application}).
-Finally, I discuss the results and further work in \cref{imperf-test:sec:discussion}.
-
-{\color{red}{NEW BIT}}
 In \cref{E-perf-test}, a framework for analysing the CIS to produce estimates of the duration of positivity was developed.
@@
 In this chapter I identify this issue to be the result of false negatives, a form of misclassification bias, within the CIS (\cref{imperf-test:sec:problem}).
-In \cref{imperf-test:sec:modelling} I develop the model in \cref{E-perf-test} to include the possibility of observing false negatives.
-There exists little prior work on the incorporation of misclassification bias within similar survival analysis frameworks.
-\todo[inline]{Review literature properly, here or elsewhere}. Building on this literature, model adaptation to include false negatives involves.....GIVE DETAILS OF WHAT YOU DID
-After showing through simulation (\cref{imperf-test:sec:sim-study-results}) that the new model recovers the true duration distribution (\cref{imperf-test:sec:sim-study-results}),  and I apply it to the CIS data (\cref{imperf-test:sec:application}).
+
+The incorporation of false negatives into survival analysis is an area of much interest, especially in the application of tests for HIV infants~\autocite[e.g.]{brownBayesian,balasubramanianEstimation}.
+\Textcite{piresIntervalMisclassify} provide a comprehensive review of approaches.
+However, false negatives with either doubly censored or truncated data has not been addressed.
+The situation here requires both.
+
+In \cref{imperf-test:sec:modelling}, I extend the model of \cref{E-perf-test} to incorporate false negatives.
+After showing through simulation that the new model recovers the true duration distribution (\cref{imperf-test:sec:sim-study-results}), I apply it to the CIS data (\cref{imperf-test:sec:application}).
 Finally, I discuss the results and further work in \cref{imperf-test:sec:discussion}.
@@
 
-\Textcite{shenNonparametrica}, extending \textcite{panNote}, studied the situation when only the terminating event is interval censored and the initiating event has simple left truncation.
-They showed that the maximum likelihood estimator is inconsistent if there is at least one single positive episode\todo{check the precise results and condition} and can lead to a severely biased underestimation of the survival function.
+\Textcite{shenNonparametrica}, generalising \textcite{panNote}, studied the situation when only the terminating event is interval censored (\ie single censored data) and left truncated.
+They showed that the non-parameteric maximum likelihood estimator is inconsistent if single positive episodes are possible, and the survival function is severely biased downwards.
+
 The CIS is a more complex setting, including double interval censoring and a complex pattern of missed episodes (see \cref{E-perf-test:sec:problem}).
@@
 $\psens = 1$ means that there are no false negatives, the situation previously considered.
-\todo{check if detectable and test sensitivity defined somewhere, and in a consistent way}
 In \cref{E-perf-test}, the outcome of a test was deterministic: always positive if the individual being tested had an ongoing infection episode.
@@
 The CIS data provides evidence for a test sensitivity declining over the course of an episode.
-\Cref{imperf-test:fig:bounding-cis-sensitivity} demonstrates this by classifying all test results as follows.
+\Cref{imperf-test:fig:bounding-cis-sensitivity} demonstrates this by using the following classification for test results.
 \begin{enumerate}
@@
 
-This could be incorporated into the simulation study be generating test results at ...
+I generate results using this model by, for each test in a detectable individual, by calculating $\psens(t-b)$, where $t$ is the day of the test and $b$ is the day the episode began.
+The result of the test is then a Bernoulli random variable, independent of all other considerations.
 Adapting the simulation study to incorporate this test sensitivity is much more similar to the data (see purple histogram in \cref{imperf-test:fig:sim-single-pos}).
@@
 
+Similar likelihoods have previously appeared in the literature~\autocite[e.g.][eq.\ (2)]{piresIntervalMisclassify}.
+However, this prior work was for singly interval censored data; incorporating the double interval censored nature of the CIS data will involve summing over the possible episode start times.
+
 I start by finding the tests that need consideration.
-For tractability, assume that the negative test bounding the start of the infection, on day $l_i^{(b)}-1$, is a true negative.
-This assumption is reasonable because the test sensitivity is high early in an infection; therefore, this test is unlikely to be a false negative.
-For tractability and simplicity, I consider only tests between the positive tests providing a lower bound on the length of the episode.
-By definition (see\todo{insert where these are defined in chapter 2}), these are the tests between $r_i^{(b)}$ and $l_i^{(e)}$ inclusive.
+For tractability, assume that the negative test bounding the start of the episode, on day $l_i^{(b)}-1$, is a true negative.
+This assumption is reasonable because, since a positive test follows at $r_i^{(b)}$, the negative at $l_i^{(b)}-1$ is likely early in the infection and the test sensitivity is high early in an infection.
+Therefore, this is a test in a period with high test sensitivity \ie unlikely to be a false negative.
+For tractability and simplicity, I consider only tests between the negative tests providing an upper bound on the length of the episode.
+By definition (see\todo{insert where these are defined in chapter 2}), these are the tests between $l_i^{(b)} - 1$ and $r_i^{(e)} + 1$ inclusive.
 They are a subset of $\sched_i$, the set of test times for the individual in which episode $i$ occurs.
-Denote this subset as $\sched'_i = \{ t \in \sched_i \ssep r_i^{(b)} \leq t \leq l_i^{(e)} \}$, and their results by $\vec{y_i'} = \{ y_i(t) \ssep t \in \sched'_i \}$ where $y_i(t) = 1$ if the test on day $t$ is positive and 0 otherwise.
-As I assume that there are no false positives, the infection episode must span at least this period because it starts and ends with a positive test.
-Therefore, the test results at times in $\sched_i'$ are either true positives or false negatives.
+The test at $l_i^{(b)} - 1$ is assumed to be a true negative, occurring with probability 1, and hence does not contribute to the likelihood.
+Denote the remaining tests considered as $\sched'_i = \{ t \in \sched_i \ssep r_i^{(b)} \leq t \leq r_i^{(e)} + 1 \}$, and their results by $\vec{y_i} = \{ y_i(t) \ssep t \in \sched'_i \}$ where $y_i(t) = 1$ if the test on day $t$ is positive and 0 otherwise.
+As I assume that there are no false positives, the infection episode must span at least the period $[r^{(b)}_i, l^{(e)}_i]$ because it starts and ends with a positive test.
+This includes all $t \in \sched'_i$ except $t = r_i^{(e)}+1$.
+Therefore, the test results at times in $\sched_i'$, except the test at $r_i^{(e)}+1$, are either true positives or false negatives.
 
@@
 First, if $e_i \leq r_i^{(e)}$, meaning that the test at $r_i^{(e)}+1$ is a true negative and the end of the episode is interval-censored as in the previous chapter.
-In this case, the test at $r_i^{(e)} + 1$ is a true negative, as are all other tests not in $\sched'_i$.
-These occur with probability 1, by the assumption of no false positives.
+% In this case, the test at $r_i^{(e)} + 1$ is a true negative, as are all other tests not in $\sched'_i$.
+The true negative occurs with probability 1, by the assumption of no false positives.
 The tests in $\sched_i'$ are either true positives or false negatives.
 \begin{align}
-&p(\vec{y_i'}, e_i \leq r_i^{(e)} | b_i, p_\text{sens}, \theta) \\
-&= p(\vec{y_i'}, l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\ % &\text{as no false positives}
-&= p(\vec{y_i'} \mid l_i^{(e)} \leq e_i \leq r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
+&p(\vec{y_i}, e_i \leq r_i^{(e)} | b_i, p_\text{sens}, \theta) \\
+&= p(\vec{y_i}, l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\ % &\text{as no false positives}
+&= p(\vec{y_i} \mid l_i^{(e)} \leq e_i \leq r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
 &= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) \right)
@@
 \begin{align}
-&p(\vec{y_i'}, e_i > r_i^{(e)} | b_i, p_\text{sens}, \theta) \\
+&p(\vec{y_i}, e_i > r_i^{(e)} | b_i, p_\text{sens}, \theta) \\
 &= p(y_i \mid e_i > r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(e_i > r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
@@
 
-These expressions can now be used to form the full replacement for $p_{ia}$, $p'_{ia}$, the likelihood of the data $\vec{y_i'}$.
+These expressions can now be used to form the full replacement for $p_{ia}$, $p'_{ia}$, the likelihood of the data $\vec{y_i}$.
 First, augment the data with $b_i$, and split into the cases just discussed:
@@
 p_{ia}'
-&= p(\vec{y_i'} \mid p_\text{sens}, \theta) \\
-&= \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( p(\vec{y_i'}, e_i \leq r_i^{(e)} \mid b_i, p_\text{sens}, \theta) + p(\vec{y_i'}, e_i > r_i^{(e)} \mid b_i, p_\text{sens}, \theta) \right) p(b_i \mid p_\text{sens}, \theta).
+&= p(\vec{y_i} \mid p_\text{sens}, \theta) \\
+&= \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( p(\vec{y_i}, e_i \leq r_i^{(e)} \mid b_i, p_\text{sens}, \theta) + p(\vec{y_i}, e_i > r_i^{(e)} \mid b_i, p_\text{sens}, \theta) \right) p(b_i \mid p_\text{sens}, \theta).
 \intertext{Now, substitute in \cref{imperf-test:eq:ll-ei-lt-ri,imperf-test:eq:ll-ei-gt-ri}, then take out the common factor:}
@@
 
-Constant terms => proprtional to simpler thing
+If $\psens$ is fixed (\ie has a point prior) and $p(b_i \mid \psens, \theta) \propto 1$ (as assumed in \cref{E-perf-test}) then:
+\begin{align}
+p_{ia}'
+&\propto \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} S_\theta(l_i^{(e)} - b_i - 1) - p_\text{sens} S_\theta(r_i^{(e)} - b_i - 1).
+\label{imperf-test:eq:pia-prime-constant}
+\end{align}
 
@@
 Finally, the impact of the simplifying assumptions made in \cref{imperf-test:sec:modelling} for tractability.
-These assumptions are that the previous negative is a true negative and that there is a negligible probability of missing an episode due to two false negatives.
+These assumptions are that the negative at $l^{(b)_i} - 1$ is a true negative and that there is a negligible probability of missing an episode due to two false negatives.
 
@@
 If $\psens$ is misspecified and too low (see \cref{imperf-test:fig:misspecified-test-sensitivity}(A)), then the posterior estimate initially follows the true value but then separates.
-The number of episodes inferred to have truly ended by the first negative is too high, and hence the survival time is underestimated.
+The number of episodes inferred to have truly ended by the first negative is too high, and hence the survival time is overestimated.
 This affect dominates over the opposing bias of overestimating the number of missed episodes.
@@
 This suggests that the simplified model, with constant test sensitivity, is sufficient for recovering the true survival time.
+Estimating the test sensitivity is not possible without a more complex model, as discussed in \cref{imperf-test:sec:discussion}.
 Therefore, I will apply this model to the real CIS data in the next section.
@@
 The uninformative prior led to high posterior estimates of $\ntot$, and hence an implausibly large number of episodes with durations of less than five days.
-Therefore, I based an informative prior for $\ntot$ on pre-existing estimates of the total number of infections over the period with posterior mean \numprint{4136368} and standard deviation \numprint{27932}~\autocite{birrellRTM2}.
+Therefore, I based an informative prior for $\ntot$, $\ntot \sim \NBc(\mu, r)$ (as in \cref{E-perf-test:sec:model}, see \cref{E-distributions} for definition), using pre-existing estimates of the total number of infections to give $\mu\inform$ and $r\inform$.
+\Textcite{birrellRTM2} estimated the total number of infections in England over the time period I consider, with posterior mean \numprint{4136368} and standard deviation \numprint{27932}.
 % This model gives a posterior mean of \numprint{4136368} cumulative infections in England in the time period I consider, with a posterior standard deviation of \numprint{27932}~\citePersonalComms{Paul Birrell}.
 \todo{Insert correct citation for RTM paper 2 once available}
-Approximating this distribution as a negative binomial and scaling the mean to the size of the CIS cohort gives the prior $\ntot \sim \NBc(\numprint{25132}, \numprint{22047})$ (distribution defined in \cref{E-distributions}).
+Approximating this distribution as a negative binomial and scaling the mean to the size of the CIS cohort gives the prior $\mu\inform = 25132$ and $r\inform = 22047)$.
 
 With this prior, the model produces plausible estimates of the duration distribution (see \cref{imperf-test:fig:cis-estimates}).
-This estimate has more long episodes than the estimate in \cref{E-ATACCC}.
+This estimate (blue) has more long episodes than the estimate in \cref{E-ATACCC} (red).
 The qualitative increase in long episodes is robust to the choice of prior for $\ntot$, the assumed value for $\psens$ (see \cref{imperf-test:fig:cis-sensitivity}), and the choice of prior for the hazard (see \cref{E-perf-test:sec:parameters-priors}), $\lambda$.
 However, the survival proportion over the first 4 weeks is sensitive to these choices.
-The estimate using a test sensitivity of 0.8 and $\NBc(\numprint{25132}, \numprint{22047})$ give a median survival time most similar to that of \cref{E-ATACCC}.
+The estimate using a test sensitivity of 0.8 and $\NBc(\mu\inform, r\inform)$ give a median survival time most similar to that of \cref{E-ATACCC}.
 \begin{figure}
@@
   \caption{%
-    Sensitivity of CIS estimates to: (A-C) the value of $r$, the dispersion of the prior for $\ntot$ (a larger $r$ means a more informative prior), and (D-F) $\psens$.
-    A and D: median survival time, in comparison to the ATACCC-based estimate of \cref{E-ATACCC}.
+    (A-C) Sensitivity of CIS estimates to the value of $r$, the dispersion of the prior for $\ntot$ (a larger $r$ means a more informative prior) when $\psens = 0.8$.
+    (D-F) Sensitivity of CIS estimates to the value of $\psens$ when $r = r\inform$.
+    A and D: median survival time, in comparison to the ATACCC-based estimate of \cref{E-ATACCC} (shown in red).
     B and E: survival at day 50, $S_\theta(50)$.
@@
   \label{imperf-test:fig:cis-sensitivity}
+  \todo[inline]{Optional improvement: Could highlight the "preferred" model which exists in both rows and is the one used subsequently, and/or include ATACCC in the far-right hand column}
 \end{figure}
 
-The assumption that the estimates are most sensitive to is the strength of the prior on $\ntot$.
+The assumption that the estimates are most sensitive to is $r$, the strength of the prior on $\ntot$.
 A very weak, almost uninformative, prior on this quantity causes the posterior estimate to be much higher than the estimate from \textcite{birrellRTM2}.
 When increasing the prior's strength, the posterior estimate moves towards the prior smoothly, as expected (see \cref{imperf-test:fig:ntot}).
-Taking a prior from \textcite{birrellRTM2} means that the posterior survival is close to the \cref{E-ATACCC} estimate in the region where that estimate is reliable.
-This justifies the use of this prior.
+As discussed previously, the \cref{E-ATACCC} estimates are reliable for the first 2--3 weeks, notably including the median time.
+Therefore, I choose a value of $r$ to match closely \cref{E-ATACCC}'s median estimate.
+The value $r\inform$ achieves this and is a principled choice because it is based directly on the prior work \textcite{birrellRTM2}.
 \begin{figure}
@@
 However, this would require incorporating time-varying test sensitivity into the likelihood.
-Furthermore, it is unclear that such a model would be identifiable because it would not be clear whether to attribute negative results to false negatives or to recoveries\todo{cite identifiability issues with varying sensitivity}.
+If the current model, with a constant $\psens$, is used then the estimate of $\psens$ would be heavily informed be intermittent negatives (the first, constant term in \cref{imperf-test:eq:pia-prime}).
+These negatives will, in general, be further from the end of an episode than a randomly selected test.
+Therefore, considering the viral dynamics from \cref{E-ATACCC}, the viral load will be higher and false negatives rarer.
+
+Estimating the test sensitivity excluding intermittent negatives is not possible because \cref{imperf-test:eq:pia-prime-constant,imperf-test:eq:pit-prime} are both monotonically decreasing in $\psens$; therefore, the likelihood always favours $\psens = 0$ (\ie no true positives).
+This aligns with the situation with singly interval-censored, untruncated data, when stopping at the first observed time of the terminating event (which may be misclassified) means that the test sensitivity cannot be estimated~\autocite[e.g.]{titmanMisclassify}.
+
+Estimating a time-varying $\psens$ with $S_\theta$ jointly may cause identifiability issues.
+These can be avoided by including external information (such as a prior giving the magnitude) on $\psens$, or test results from later follow-up~\autocite[and references therein]{piresIntervalMisclassify}.
+However, whether these methods are sufficient for the model to be identifiable with a time-varying $\psens$ needs further investigation.
+A simple parameteric form of the test sensitivity, such as that proposed by \textcite{brownBayesian}, may be sufficient to allow identifiability.
+In any case, the likelihood, especially $p_{iu}'$, would be substantially complicated by such an addition which may greatly increase the computational cost of inference.
 
@@
 
-The posterior distribution for the survival is concentrated, producing narrow credible intervals, likely due to the large sample size in the CIS.
+The posterior distribution for the survival has narrow credible intervals, likely due to the large sample size in the CIS.
 In the following chapters, I will take the posterior mean for the survival time at each point, neglecting this uncertainty, as the basis for estimating transmission.

### 2024-02-08 cis-imperfect-testing.tex

@@
 
-The incorporation of false negatives into survival analysis is an area of much interest, especially in the application of tests for HIV infants~\autocite[e.g.]{brownBayesian,balasubramanianEstimation}.
-\Textcite{piresIntervalMisclassify} provide a comprehensive review of approaches.
-However, false negatives with either doubly censored or truncated data has not been addressed.
+The inclusion of false negatives into survival analysis is an area of much interest, especially in the application of tests for HIV in infants~\autocite[e.g.][]{brownBayesian,balasubramanianEstimation};
+\textcite{piresIntervalMisclassify} provide a comprehensive review of approaches.
+However, including false negatives with either doubly censored or truncated data has not been addressed.
 The situation here requires both.

### 2024-02-08 transmission.tex

@@
 
-In this chapter, I use the preferred estimate of the duration of RT-PCR positivity (see \cref{E-imperf-test:sec:conclusion}) to estimate the transmission of SARS-CoV-2 in England.
-This estimation is based only on the CIS data (as shown in \cref{E-biology-data:sec:positivity} for further details).
-I will show that this data is sufficient to estimate the parameters of the transmission process, suggesting surveys of this type are sufficient for surveillance of an epidemic.
+In this chapter, I use the preferred estimate of the duration of RT-PCR positivity (see \cref{E-imperf-test:sec:conclusion}) to investigate estimation of incidence and transmission of SARS-CoV-2 in England, solely on the basis of .
+CIS data (as shown in \cref{E-biology-data:sec:positivity} for further details).
+I will show that the CIS data is sufficient to derive such estimates, suggesting surveys of this type would be sufficient for surveillance of an epidemic.
 
-I take two approaches; the theoretical advantages of each were discussed in \cref{E-inc-prev:sec:infection-process}.
-First, in \cref{backcalc}, I use a backcalculation approach, based on \cref{E-inc-prev:eq:EPt}.
+I take two approaches, the theoretical advantages of each were discussed in \cref{E-inc-prev:sec:infection-process}.
+First, in \cref{backcalc}, I use a back-calculation approach, based on \cref{E-inc-prev:eq:EPt}.
 This approach enables estimation of incidence within (sub)populations of England.
-Then, in \cref{SEIR}, I use a mechanistic compartmental model, an implementation of the model introduced in \cref{E-SEIR:sec:mechanistic-models}.
-In addition to estimating incidence, this approach also estimates the components of transmission, \eg whether changes can be explained by changes in contact rates.
+Then, in \cref{SEIR}, I use a mechanistic compartmental model, a variation of the model introduced in \cref{E-SEIR:sec:mechanistic-models}, to estimate both incidence and components of transmission, \eg and whether changes in transmission can be explained by changes in contact rates.
 Finally, in \cref{transmission:sec:comparison}, I compare the results of the two approaches.
@@
     Lit review to go in introduction chapter, framing the whole thesis because it is important in understanding the motivation for the whole work.
-    Would include \textcite{mccabeCISincidence,abbottCISincidence} on backcalculation. Also Toffe?
+    Would include \textcite{mccabeCISincidence,abbottCISincidence} on backcalculation. Also Taffe?
     As well as the literature on mechanistic models using REACT and CIS I have previously circulated.
@@
 
-\todo[inline]{This is now in \cref{E-biology-data:sec:posivity})}
+\todo[inline]{This is now in the final section of chapter 2.}
 
-Throughout this chapter, the CIS data will be stratified into the regions of England and the following age groups.
-\begin{enumerate}
-    \item $[0, 11)$: children of pre-school and primary school age.
-    \item $[11, 16)$: children of compulsory secondary-school age.
-    \item $[16, 25)$: adolescents and young adults who may remain in formal education.
-    \item $[25, 50)$: younger working age adults.
-    \item $[50, 70)$: older working age adults.
-    \item $[70, \infty)$: elderly.
-\end{enumerate}
-These groupings align closely with those previously used for analysis of the CIS data~\autocites[e.g.:][]{houseInferring}{walkerTracking}.
+% Throughout this chapter, the CIS data will be stratified into the regions of England and the following age groups.
+% \begin{enumerate}
+%     \item $[0, 11)$: children of pre-school and primary school age.
+%     \item $[11, 16)$: children of compulsory secondary-school age.
+%     \item $[16, 25)$: adolescents and young adults who may remain in formal education.
+%     \item $[25, 50)$: younger working age adults.
+%     \item $[50, 70)$: older working age adults.
+%     \item $[70, \infty)$: elderly.
+% \end{enumerate}
+% These groupings align closely with those previously used for analysis of the CIS data~\autocites[e.g.:][]{houseInferring}{walkerTracking}.
 
-I use the period from the expansion of the CIS, starting Mon 31st August 2020 until Sun 24th Jan 2021 (inclusive).
-This period is chosen because there were very few positives in CIS in summer 2020 due to low prevalence and a smaller sample size, and ends when vaccination crossed 10\% of England's population.
+% I use the period from the expansion of the CIS, starting Mon 31st August 2020 until Sun 24th Jan 2021 (inclusive).
+% This period is chosen because there were very few positives in CIS in summer 2020 due to low prevalence and a smaller sample size, and ends when vaccination crossed 10\% of England's population.
 
-I will model the test results as overdispersed relative to a binomial model.
-Overdispersion allows for household clustering\todo{cite beta-binomial as a model for clustering} and other overdispersion in the data, which is common in epidemiological data\todo{cite that epi data is often overdispersed}.
-The parameterisation of the beta-binomial that I use (detailed in \cref{E-distributions}) has an overdispersion parameter $\rho$, regarded as a nuisance parameter in this context.
-$\rho=0$ means the beta-binomial coincides with the binomial distribution.
-As $\rho \to \infty$, the distribution tends towards the uniform distribution on $[0, 1]$~\autocite{hughesUsing}.
+% I will model the test results as overdispersed relative to a binomial model.
+% Overdispersion allows for household clustering\todo{cite beta-binomial as a model for clustering} and other overdispersion in the data, which is common in epidemiological data\todo{cite that epi data is often overdispersed}.
+% The parameterisation of the beta-binomial that I use (detailed in \cref{E-distributions}) has an overdispersion parameter $\rho$, regarded as a nuisance parameter in this context.
+% $\rho=0$ means the beta-binomial coincides with the binomial distribution.
+% As $\rho \to \infty$, the distribution tends towards the uniform distribution on $[0, 1]$~\autocite{hughesUsing}.
 
@@
 
-A deterministic relationship between true prevalence an incidence is a good approximation (see \cref{E-inc-prev:sec:prevalence-process}).
-A well-established methodology, using multilevel regression and poststratification (MRP)\todo{cite MRP}, has already been developed for estimating prevalence from the CIS data\todo{cite Koen's stuff}.
-Therefore, I extend the methodology previously used to estimate prevalence to estimate incidence.
+In \cref{E-inc-prev:sec:observation-process}, I showed that a deterministic approximation to the relationship between true prevalence and incidence is a good approximation for estimating incidence.
+The deterministic relationship between incidence proportion and prevalence means that either can be inferred, with the other then calculated.
+Estimation of prevalence from the CIS data has been carried out through a well-established methodology, used throughout the pandemic~\autocite{pouwelsCommunity,cisMethodsONS}.
+The method is based on multilevel regression and post-stratification (MRP)\todo{cite MRP}.
+In this section, I will extend this methodology to also estimate incidence.
 
-In MRP, a Bayesian multilevel logistic regression model is first fit to predict RT-PCR positivity from demographic covariates.
-Then, the posterior distribution is used to predict the prevalence in each stratum.
-Finally, the prevalence is poststratified (as explained later) to the CIS target population and subpopulations of interest.
-I modify this process by transforming the posterior distribution of prevalence in each stratum to a posterior distribution of incidence before poststratifying.
-% I fit a modification model to estimate prevalence, then transform the posterior distribution of prevalence in each stratum to a posterior distribution of incidence.
-% Finally, I poststratify the posterior distribution of incidence to the CIS target population and subpopulations of interest.
-
-The deterministic relationship between incidence proportion and prevalence means that either can be inferred, with the other then calculated.
-Denote the quantity to be inferred at time $t$ is strata $i$ as $\pi_{it}$.
-Note that for any $i$ and $t$, the number of positives will be small, often fewer than 5; therefore some form of borrowing between time and strata is required.
-The statistical model and inference procedure, including the borrowing of information, should have the following properties.
-\begin{itemize}
-    \item The smoothing model should penalise towards exponential growth.
+First, I will motivate this approach, as opposed to direct estimation of incidence.
+Denote the quantity to be inferred (incidence proportion or prevalence) at time $t$ in strata $i$ as $\pi_{it}$.
+For any $i$ and $t$, the number of positives will be small, often fewer than 5; therefore some form of borrowing between time and strata is required.
+The statistical model and inference procedure should have the following properties.
+\begin{enumerate}
+    \item The smoothing model over time should penalise towards exponential growth.
 Penalising towards exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time.
-        Furthermore, such penalisation has previously been shown to improve estimates of prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
+Furthermore, such penalisation has previously been shown to improve estimates of SARS-CoV-2 prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
     \item Limit estimates to epidemiologically and biologically plausible values.
         That is, incidence proportion and prevalence should be in the interval $[0, 1]$.
-        While incidence proportion can theoretically be greater than 1, this would imply that some individuals were infected multiple times in a single time unit (day) which is not biologically plausible.
+        While incidence proportion can theoretically be greater than 1, this would imply that some individuals were infected multiple times in a day which is not biologically plausible.
     \item Have sufficient parameters to incorporate all necessary covariates to correct for non-response bias (see \cref{E-biology-data:sec:positivity}).
-    \item Computationally efficient.
+    \item Borrow information across the different covariates.
+        The borrowing should penalise towards the distribution of incidence/prevalence across strata in the same geographic region being constant.
+        Standard stratified, mechanistic transmission model (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) predict this, if behavioural changes are independent of strata and there is negligible impact of immunity.
+    \item Be computationally efficient.
         The model will be fit to granular data and therefore needs to be run within the SRS, with its computational limitations (see\todo{ref section on SRS computational limitations}).
-        In particular, MCMC should be avoided because incorporating sufficient covariates to account for all groups led to a model that was too computationally expensive~\citePersonalComms{Koen Pouwels}.
-\end{itemize}
+\end{enumerate}
 
-\todo[inline]{I am struggling the framing of these paragraphs. Maybe talk about being linear in its parameters and pre-existing software packages and not mention GLMs?}
-The advantage of fitting a multilevel regression model on prevalence, rather than incidence, is that prevalence with appropriate prior (or penalisation) can be expressed as a generalised linear model (GLM)\todo{maybe cite backcalculation as a GLM} or Gaussian Markov Random Field (GMRF).
-Standard, efficient, and widely-used software can then be applied for inference.
+Properties 1 and 2 suggest a model of the form $\logit(\pi_{it}) = g(i, t)$, where $g$ is a function penalised towards being linear in $t$.
+The $\logit$ link function both limits $\pi_{it}$ to the interval $[0, 1]$, and logit-linear growth approximates exponential growth for small $\pi_{it}$.
+Combined with property 4, the form of $g$ should be penalised towards $g(i, t) = a_0 + a_t t + f(i)$ where $a_0$ and $a_t$ are model parameters and $f$ is a function to be specified.
+However, it should be possible for sufficient information in the data to move the posterior estimate of $\pi_{it}$ away from this model.
 
-To use a GLM or GMRF, express incidence proportion or prevalence in a stratum $i$ on day $t$, $\pi_{it}$, as $\logit(\pi_{it}) = g(i, t)$, where is a function of the stratum and time, penalised towards being linear in time.
-If $g$ is chosen appropriately, this is a GLM with a logit link function.
+Properties 3 and 4 suggest the extension of the previously-developed MRP method.
+% This method has previously been used successfully to estimate prevalence, corrected for non-response bias, from the CIS data~\autocite{cisMethodsONS,pouwelsCommunity,pouwelsMRPvaccination}.
+In MRP, a Bayesian multilevel logistic regression model is used to predict RT-PCR test results based on demographic covariates.
+Then, the posterior distribution is used to predict the prevalence in each stratum.
+Finally, the prevalence is post-stratified to the CIS target population and sub-populations of interest.
+I propose this process by transforming the posterior distribution of prevalence in each stratum to a posterior distribution of incidence before post-stratifying.
+Further details on this methodology follow.
 
-The logit transformation serves two purposes.
-First, it ensures that the incidence proportion or prevalence is between 0 and 1.
-Second, at low values of incidence or prevalence, linear growth of $\logit(\pi_{it})$ over time implies approximately exponential growth of $\pi_{it}$~\autocite[see discussion in][and references therein]{ealesAppropriately}.
-Penalising towards exponential growth is a natural formulation because mechanistic models predict expoential growth over short periods of time.
-Furthermore, such penalisation has previously been shown to improve estimates of prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
+Properties 3 and 5 suggest the use of an approximate method for computing the posterior distribution.
+In particular, MCMC should be avoided.
+Incorporating sufficient covariates to adjust for the non-response bias over a period of more than 6--8 weeks leads to a model that was too computationally expensive~\citePersonalComms{Koen Pouwels}.
+Instead, previous MRP implementations have used the Integrated Nested Laplace Approximation (INLA)\todo{cite INLA}, implemented in R-INLA\todo{cite INLA}.
 
-However, if the incidence proportion is expressed as a GLM in this way, the prevalence is not expressible as a GLM.
-Specifically, the prevalence (or an element-wise transformation of it) is no longer linear in the model parameters.
-The relationship between incidence and prevalence is linear, as derived in \cref{E-inc-prev:sec:prevalence-process}.
-Therefore, the relationship between the logit of incidence proportion and prevalence is non-linear.
-
-Using a GLM is computationally useful due to the large number of pre-existing packages available for computing the posterior distribution, or its approximation, efficiently.
-Previously, when fitting to prevalence over long periods and including all appropriate covariates to correct for non-response bias, the model was too computationally expensive to fit using MCMC~\citePersonalComms{Koen Pouwels}.
-Instead, an implementation of the model using the Integrated Nested Laplace Approximation (INLA) algorithm was used\todo{cite INLA}.
+The use of R-INLA requires that $\pi_{i,t}$ represent incidence proportion rather than prevalence.
+If $\vec{\pi}$ represented incidence proportion, then $\logit \vec{pi}$ could be linear in the model parameters, as explained previously, while satisfying the other criteria.
+However, the expected prevalence at time $t$ is linear in $\vec{\pi}$ (see \cref{E-inc-prev:eq:EPt}), and hence non-linear in $\logit \vec{pi}$.
+The expected proportion of positive tests (\ie the expectation of the data likelihood) is equal to the prevalence.
+Applying a non-linear transformation then a linear transformation in this way to form the data likelihood's expectation cannot be implemented within R-INLA's framework.
 
@@
     \item $w_t$ is a second-order random walk, that is $w_t = 2w_{t-1} - w_{t-2} + \epsilon_t$ where $\epsilon_t \dist \N(0, \sigma_w^2)$.
-    \item $l_{a,t}$ is an age-group specific second-order random walk. That is $l_{a,t} = 2l_{a,t-1} - l_{a,t-2} + \epsilon_{a,t}$ where $\epsilon_{at} \dist \N(0, \sigma_l^2)$.
+    \item $l_{a,t}$ is an age-group specific second-order random walk. That is $l_{a,t} = 2l_{a,t-1} - l_{a,t-2} + \epsilon_{a,t}$ where $\epsilon_{a,t} \dist \N(0, \sigma_l^2)$.
     \item $n_{taes}$ and $y_{taes}$ are the number of individuals tested and the number of positive test results on day $t$ in individuals of age group $a$, ethnicity $e$, and sex $s$.
@@
 
-Having fit the prevalence model, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution.
+Having fitted the prevalence model, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution.
 For each stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
-I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:observation-process} justifies, that is $P_t = \E(P_t \mid \vec{Z})$.
+I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:observation-process} justifies, that is the likelihood $\vec{P} \mid \vec{Z}$ is well-approximated by only considering its mean.
 \Cref{E-inc-prev:eq:EPt}, considered across different values of $t$, can be written as $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix formed from the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
@@
 $\matr{S}$ is lower-triangular, hence to solve for $\vec{P}$ I use forward-substitution~\autocite{cormenMatrix}.
+Forward-substitution sequentially solves the $t$th element of $\vec{Z}$ ($t = 1, \dots, T$) by calculating:
+\begin{align}
+Z_t
+&= \frac{P_t - \sum_{i=1}^{t-1} S_{ti} Z_i}{S_{tt}} \\
+&= P_t - \sum_{i=1}^{t-1} S(t - i + 1) Z_i.
+\label{transmission:eq:forward-substitute}
+\end{align}
+Unlike an inference procedure which infers the parameters simultaneously, there is no feedback loop: the value of $Z_{t}$ does not affect $Z_{t'}$ for $t' < t$.
+The second term here is the contribution of the incidence prior to day $t$ to day $t$'s prevalence.
+The remainder of the prevalence on day $t$ is therefore due to incidence on day $t$.
 % Forward substitution is numerically stable unless the diagonal elements, here $S(1)$, are small relative to the off-diagonal elements, here $S(i)$ for $i > 0$~\autocite[section 8.2]{highamAccuracy}.
 % Here, the diagonal is $S(1) = 1$ and hence the matrix the solution is numerically stable.
+
+Violation of assumptions in various parts of this procedure can lead to the second term in \cref{transmission:eq:forward-substitute} to be greater than the first, giving negative incidence.
+In this case, the incidence prior to day $t$ imply that the prevalence on day $t$ is higher than that predicted by the fitted regression model for this strata/day combination.
+There are two sets of assumptions which are most likely to be violated.
+The first set of assumptions is in constructing survival matrix, $\matr{S}$, which is assumed to be known with no uncertainty in this procedure.
+However, \cref{E-imperf-test} showed that the estimate of $S(t)$ is sensitive to assumptions made, such as the assumed $\psens$ and strength of priors used.
+The second set of assumptions are in specifying the statistical model used to estimate prevalence, for example the form of the smoothing used.
+
+A prior constraining incidence to be non-negative and then fit directly to the data could be used, but this could no longer be fit within the R-INLA framework.
+\todo{Maybe move this paragraph to the discussion?}
+Hence, an alternative inference methodology, or simpler model (\eg containing fewer random effects), would be required.
+% MCMC is likely too computationally expensive.
+% Other approaches, such as a Laplace approximation or variational inference, could be considered.
+For this analysis, I replaced any value of incidence estimated to be negative with 0.
 
@@
 Subtracting $\vec{P'}$ from $\vec{P}$ before applying the forward solving procedure accounts for the contribution of the incidence prior to the first day of prevalence data.
-
-Model misspecification means that the procedure described could produce negative incidence.
-\todo{This paragraph might be too terse.}
-For any stratum and day $t$, this occurs if the prevalence reduces between day $t$ and $t+1$ more than the number of recoveries on day $t$, based on the incidence up until day $t$ and the survival function.
-
-There are two places where this misspecification could occur.
-The first is the survival matrix, $\matr{S}$.
-For example, \cref{E-imperf-test} showed that it is sensitive to assumptions made, such as the assumed $\psens$ and ....
-The second is the specification of the statistical model used to estimate prevalence, for example the form of the smoothing used.
-
-\todo[inline]{Maybe relate following paragraph to the GLM discussion at the start of the section - its basically the same point}
-A prior constraining incidence to be non-negative and then fit directly to the data could be used, but this could no longer be fit within the R-INLA framework.
-Hence, an alternative inference methodology, or simpler model (\eg containing fewer random effects), would be required.
-% MCMC is likely too computationally expensive.
-% Other approaches, such as a Laplace approximation or variational inference, could be considered.
-For this analysis, I replaced any value of incidence estimated to be negative with 0.
 
@@
 
-During November, incidence starts rising, although starting at different times in different regions.
+During November, incidence starts rising, with regional heterogeneity in the timing of when this occurs.
 The most explored explanation for this is the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocite{walkerTracking,daviesEstimated,lythgoeLineage}.
@@
 The incidence proportion was highest among the 11--15 and 16--25 age groups throughout the period (see \cref{transmission:fig:backcalc-ages}).
-Except the 16--25 age group, all age groups show similar trends (\ie parallel lines on a log scale).
-The distribution of infections among age groups remaining constant over short periods of time is the pattern predicted by standard age-stratified, mechanistic transmission model (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) in equilibrium, if behavioural changes are independent of age group, and there is negligible impact of immunity.
+Except the 16--25 age group, all age groups show similar trends (\ie parallel lines on a log scale), meaning there was insufficient information for the posterior to overcome the prior penalisation towards this situation.
 The incidence proportion in the 16--25 age group decreases faster than other age groups during November.
@@
 I follow the WAIFW matrix parameterisation of \textcite{birrellRealtime}, decomposing the rate of transmission from individuals in age group $a'$ to susceptible individuals in age group $a$ as $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'} N_{a'}$.
-This separates contact patterns, $\kappa_{aa'}$, which can vary over time; transmission probability upon contact between age groups $a$ and $a'$, $q_{aa'}$, assumed constant; a region-wide effect for time, $\beta(t)$; and the population of the stratum, $N_{a'}$.
+This separates contact patterns, $\kappa_{aa'}$, which can vary over time; probabilities of transmission upon contact between age groups $a$ and $a'$, $q_{aa'}$, assumed constant; a region-wide effect for time, $\beta(t)$; and the population of the stratum, $N_{a'}$.
 
@@
 These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
-Google mobility data is based on the location of Android users over time, relative to a pre-pandemic baseline.
+Google mobility data are based on the location of Android users over time, relative to a pre-pandemic baseline.
 The mobility and school data are averaged over a week then used to down or up-weight the number of contacts in each setting, before combining contacts across settings to form $\kappa_{aa'}(t)$.
@@
 One approach would be to convolve the incidence based on the transmission model with the RT-PCR duration distribution.
-Here, the daily incidence proportion for each stratum is by solving the system of ODEs governing the transmission model, and then prevalence calculated as using \cref{E-inc-prev:eq:EPt}.
+Here, the daily incidence proportion for each stratum is by solving the system of ODEs governing the transmission model, and then prevalence calculated using \cref{E-inc-prev:eq:EPt}.
 However, this approach is computationally expensive, likely prohibitively so if performed within a MCMC loop.
@@
 $p_0$ increases at the same rate as individuals flow into $e_1$.
-These individuals then transition into a set of compartment with their transition rates, $p_1, \dots, p_l$, chosen to match the duration of positivity estimates in \cref{E-imperf-test}.
-These compartments do not have a biological interpretation; the same idea is behind having multiple exposed or latent compartments (introduced in \cref{E-SEIR:sec:non-exponential}).
+These individuals then transition into a set of compartments with transition rates, $p_1, \dots, p_l$, chosen to match the duration of positivity estimates in \cref{E-imperf-test}.
+The new compartments do not have a biological interpretation; the same idea is behind having multiple exposed or latent compartments (introduced in \cref{E-SEIR:sec:non-exponential}).
 

### 2024-02-08 ATACCC.tex
more chapter 2 text
proof read ATACCC

@@
 First, I will explain how biomarkers are typically modelled, both for SARS-CoV-2 and historically in the biostatistical literature (\cref{ATACCC:sec:background}).
-Next, I describe the model of \textcite{hakkiOnset} in detail, which applies these methods to the ATACCC data (\cref{ATACCC:sec:hakki}) and before modifying its observation model to improve the inference (\cref{ATACCC:sec:observation-modification}).
+Next, I describe the model of \textcite{hakkiOnset} in detail, which applies these methods to the ATACCC data (\cref{ATACCC:sec:hakki}) and before modifying its observation model to improve the robustness of the inference (\cref{ATACCC:sec:observation-modification}).
 This modification is based on an earlier analysis using a preliminary version of the dataset, described in \cref{ATACCC:sec:original-analysis}.
 Finally, I convert the model and parameter estimates of the biomarker's trajectory to a duration of RT-PCR positivity (\cref{ATACCC:sec:duration}).
-I will refine this estimate in future chapters before its use in estimating transmission.
+I will refine this estimate in the following chapters before its use in estimating transmission in \cref{E-transmission}.
 
@@
 
-Two major approaches have been taken to model SARS-CoV-2 viral load trajectories.
-There are either using ordinary differential equations (ODEs) with a biological interpretation or using mathematical functions that are flexible enough to capture the dynamics.
-The ODE-based approach uses parameters representing biological properties of the virus and host~\autocites[e.g.:][]{ejimaEstimation,keVivo,kimQuantitative,goncalvesTiming,perelsonMechanistic}.
+Two major approaches have been taken to model SARS-CoV-2 viral load trajectories, both based on well-well-established methods within biostatisics.
+The first approach is an ODE-based approach uses parameters representing biological properties of the virus and host~\autocite[e.g.:][]{ejimaEstimation,keVivo,kimQuantitative,goncalvesTiming,perelsonMechanistic}.
+The model here is an application of a standard model for viral dynamics, used for many diseases\todo{cite history of viral load ODE models}.
 Such models are rarely analytically-tractable and the system's evolution is computed using numerical methods.
 % Both of these approaches have been used with other viruses and biomarkers.
-The approach that is the focus of this chapter is using a piecewise linear model for the biomaker trajectory and random effects to allow for within-individual correlation.
+The second approach, and the focus of this chapter, is using a piecewise linear model for the biomaker trajectory and random effects to allow for within-individual correlation.
 
-Piecewise linear, random effect models have a long history of use for longitudinal data in biostatistics~\autocites{slateStatistical}.
+Piecewise linear, random effect models have a long history of use for longitudinal data in biostatistics (its history is reviewed by \textcite{slateStatistical}).
 Often, as here, they are used to model the temporal evolution of biomarkers.
-A few example applications are: CD4 counts following HIV infection~\autocites{langeHierarchical,lynchPredicting}, parasite density in blood samples following malaria treatment~\autocites{fogartyBayesian}, and measurements of cognitive decline due to Alzheimer's disease~\autocites{bealisleBayesian}.
+A few example applications are: CD4 counts following HIV infection~\autocite{langeHierarchical,lynchPredicting}, parasite density in blood samples following malaria treatment~\autocite{fogartyBayesian}, and measurements of cognitive decline due to Alzheimer's disease~\autocite{bealisleBayesian}.
 
 In these models, the function's parameters are clearly related to features of the biomarker's trajectory, such as the time from first detection to peak viral load, and the viral load at the peak.
-The piecewise linear model is the most common in the context of SARS-CoV-2 viral load~\autocites{clearyUsing,kisslerViral,larremoreTest}
-Here, the logarithm of viral load increases linearly, peaks instantaneously, and then decreases linearly; more flexible functions have also been suggested~\autocites{quiltyQuarantine}.
-This model for the trajectory I adopt in this chapter, from \textcite{hakkiOnset}, is based on the piecewise log-linear model, however, they approximate the model so that the trajectory is continuous (see \cref{ATACCC:sec:hakki} for details).
+The piecewise linear model is the most common in the context of SARS-CoV-2 viral load~\autocite{clearyUsing,kisslerViral,larremoreTest}
+In this context, the piecewise linear model is applied to the log of the viral load.
+Hence, the log of the viral load increases linearly, peaks instantaneously, and then decreases linearly; more flexible functions have also been suggested~\autocite{quiltyQuarantine}.
 
-Random effects represent the within-individual correlation within the population while borrowing information between individuals~\autocite[chapter 24]{lashModern}.
+The model for the trajectory I adopt in this chapter, from \textcite{hakkiOnset}, is based on the piecewise log-linear model.
+However, they approximate the model so that the function is continuous (see \cref{ATACCC:sec:hakki} for details).
+
+Random effects represent the within-individual correlation while borrowing information between individuals~\autocite[chapter 24]{lashModern}.
 Measurements of viral load from the same individual are correlated.
@@
 However, the random effect(s) for each individual is drawn towards a common mean, allowing information on their value to be shared between individuals.
-Following much of the literature referenced in this section, I will take a Bayesian approach in this chapter.
+Following much of the literature mentioned previously, I will take a Bayesian approach in this chapter.
 In the Bayesian context, random effect models are often referred to as hierarchical or multilevel models because they can be interpreted as placing a prior at two levels of a parameter hierarchy~\autocite[chapter 5]{gelmanBDA}.
@@
 The model of \textcite{hakkiOnset} has two components: a latent viral load model and an observation model.
-The viral load model produces a deterministic trajectory for the viral load of an individual, a function of the model parameters.
-The observation model then describes how the RT-PCR testing process produces noisy observations of this viral load.
-The observation model includes the possibility of false positive and negatives, and the limit of detection of the RT-PCR test.
+The viral load model is a deterministic function of the model parameters, producing an individual's viral load trajectory.
+The observation model mathematically formalises how the RT-PCR testing process produces noisy observations of the latent viral load.
+The observation model includes the possibility of false positive and negatives, and the limit of detection of the RT-PCR test (see \cref{E-biology-data:sec:PCR}).
 % The limit of detection means that viral loads below a certain threshold are not detected by the RT-PCR test.
@@
 As NUTS uses gradient information, it is more efficient when the model is differentiable.
-This leads to the following model for the viral load at time $t$, $v_i(t)$, in individual $i$ is shown in \cref{ATACCC:fig:viral-load-model} and given by:
+This leads to the following model for the viral load (examples are shown in \cref{ATACCC:fig:viral-load-model}):
 \begin{align}
-v_i(\tau) = v_{\max,i} \frac{a_i+b_i}{b_ie^{-a_i(\tau-\tau_{\max,i})} + a_ie^{b_i(\tau-\tau_{max,i})}}. \label{ATACCC:eq:viral-load}
+v_i(t) = v_{\max,i} \frac{a_i+b_i}{b_ie^{-a_i(t-\tau_{\max,i})} + a_ie^{b_i(t-\tau_{max,i})}}. \label{ATACCC:eq:viral-load}
 \end{align}
-Time, $\tau$, is continuous and defined uniquely for each individual such that $\tau = 0$ at the time that individual $i$ was first tested.
+Time, $t$, is continuous and defined uniquely for each individual such that $t = 0$ at the time that individual $i$ was first tested.
 $v_{\max,i}$ is the maximum viral load, which occurs at time $\tau_{\max,i}$.
-As $\lvert \tau - \tau_{\max} \rvert$ becomes large, the viral load trajectory tends towards exponential growth (before the peak) or decline (after the peak) in viral load at rates $a_i$ and $b_i$ respectively.
-% Before the peak, this is because $b(\tau-\tau_{\max,i})$ is small and hence $a\exp(b(\tau-\tau_{\max,i}))$ is negligible.
-% After the peak, this is because $b\exp(-a(\tau-\tau_{\max,i}))$ is negligible.
+As $\lvert t - \tau_{\max} \rvert$ becomes large, the viral load trajectory tends towards exponential growth (before the peak) or decline (after the peak) in viral load at rates $a_i$ and $b_i$ respectively.
+% Before the peak, this is because $b(t-\tau_{\max,i})$ is small and hence $a\exp(b(t-\tau_{\max,i}))$ is negligible.
+% After the peak, this is because $b\exp(-a(t-\tau_{\max,i}))$ is negligible.
 This happens quickly for realistic values of the parameters.
-Therefore, the log viral load is approximately piecewise linear (see \cref{ATACCC:fig:viral-load-model} for a representative viral load trajectory).
+Therefore, the log viral load is approximately piecewise linear (see \cref{ATACCC:fig:viral-load-model} for representative viral load trajectories).
 $\vec{\phi}_i = (v_{\max,i}, a_i, b_i)^T$ are the random effects for individual $i$.
-$\tau_{\max,i}$ is inferred independently for each individual because the time from first test to peak is not a biological concept shared between individuals.
+
+$\tau_{\max,i}$ is inferred independently (\ie without a shared mean) for each individual because time is defined for each individual based on when they are first tested.
+This time is based on logistical and behavioural factors such as delays in the NHS Test and Trace system and the study design.
 \begin{figure}[tbhp]
@@
 
-RT-PCR tests are noisy observations of the viral load.
-A positive RT-PCR test has a Ct value.
-Ct values are linearly related to viral load (see \cref{E-biology-data:sec:PCR})
-A negative RT-PCR test signals either the viral load is too low to be detected, \ie below the limit of detection, or the test is a false negative.
-Positive RT-PCR tests, especially those with low viral loads, could be false positives, for example caused by contamination.
-
-The noisy observations are modelled by assuming each measurement of log viral load has an iid error distributed $\N(0, \sigma_v)$.
+RT-PCR tests are noisy observations of the viral load (see \cref{E-biology-data:sec:PCR}).
+The noisy observations are modelled by assuming each measurement of log viral load has an iid error distributed $\N(0, \sigma_v^2)$.
 The measurements are left-censored at the limit of detection, $\vlod$, meaning that any observation which would give a log viral load of less than $\vlod$ is instead observed as negative.
@@
 Let the observed log viral load for individual $i$ at time $t$ be $y_{i,t}$, with negative tests signified by $y_{i,t}=v_\text{lod}$.
-Then, its likelihood is:
+Then, the likelihood contribution of this observation is:
 \begin{align}
@@
 
-Conditional on all $v_i$s, the observations are assumed independent.
-Because $v_i$ is a deterministic transform of $\vec{\phi}_i$ and $\tau_i$, the observations are independent conditional on the model parameters, and the full likelihood is simply the product of the likelihoods of each observation.
+Conditional on all parameters, the observations are assumed independent.
+Because $v_i$ is a deterministic transform of $\vec{\phi}_i$ and $\tau_{\max,i}$, the observations are independent conditional on the model parameters, and the full likelihood is the product of the likelihoods of each observation.
 
@@
 The second group is vaccinated and all were infected by the Delta variant.
-My focus is on the first group since they best correspond to the infections in the time period used throughout this thesis, Autumn 2020.
+My focus is on the first group since they best correspond to the infections in the time period used throughout this thesis.
 
 Biological plausibility requires constraining $\vec{\phi}_i$ to positive real values.
-Each $\phi_i$ is modelled as a log-normal random effects by using the hierarchical prior $\phi_i \dist \LN(\vec{\mu}_{\phi,g(i)}, \matr{\Sigma}_\phi)$ where $g(i)$ is the group of individual $i$.
+Each $\vec{\phi}_i$ is modelled as a log-normal random effects by using the hierarchical prior $\phi_i \dist \LN(\vec{\mu}_{\phi,g(i)}, \matr{\Sigma}_\phi)$ where $g(i)$ is the group of individual $i$.
 $\vec{\mu}_{\phi,g} = (\mu_{v_{\max},g}, \mu_{a,g}, \mu_{b,g})^T$ is a vector of the population means of the log-parameters.
@@
 For example, an individual with a high peak viral load may also clear the virus slowly because both have an underlying cause of a weak immune system.
-To improve sampling efficiency, the multivariate version of the non-centred parameterisation~\autocites{papaspiliopoulosGeneral,stanReparameterization} is used.
+To improve sampling efficiency, the multivariate version of the non-centred parameterisation~\autocite{papaspiliopoulosGeneral,stanReparameterization} is used.
 This can be expressed as follows.
 \begin{align}
-  \vec{\eta}_i &\dist \N(0, \matr{C}_\phi) \\
+  \vec{\eta}_i &\dist \N(\vec{0}, \matr{C}_\phi) \\
   \log \vec{\phi}_i &= \vec{\mu}_\phi + \vec{\sigma}_\phi \circ \vec{\eta}_i
@@
 These are defined such that, by construction, $(\matr{\Sigma}_\phi)_{ij} = (\vec{\sigma}_\phi)_i (\vec{\sigma}_\phi)_j (\matr{C}_\phi)_{ij}$.
-Correlation matrices have diagonal elements all 1, therefore this reparameterisation has the same number of parameters as the original model.
+% Correlation matrices have diagonal elements all 1, therefore this reparameterisation has the same number of parameters as the original model.
 
@@
 \label{ATACCC:table:hakki-priors}}
+\todo[inline]{Check if the $\mu$ parameter priors are on the mean of log parameters or mean of parameters}
 \end{table}
@@
 The model of \textcite{hakkiOnset} did not converge fully (see \cref{ATACCC:sec:results}).
-In particular, the Rhat values, traceplots, and effective sample size for the parameters of the observation model were poor.
+In particular, the Rhat values, traceplots, and effective sample size for the parameters of the observation model ($x_0$, $\sigma_0$, and $\rho$) were poor.
 
@@
 The parameters $x_0$ and $\sigma_0$ are no longer used in the model.
-For all other parameters, I kept the priors the same as \textcite{hakkiOnset}, including the prior on $\rho$, and performed inference with the same settings (see end of \cref{ATACCC:sec:hakki}).
+
+I reanalyse the data using this modified model.
+The priors are the same as \textcite{hakkiOnset}, including the prior on $\rho$, and performed inference with the same settings (described at the end of \cref{ATACCC:sec:hakki}).
 
@@
 The ultimate aim of this chapter is to produce an estimate of the distribution for the duration of RT-PCR positivity.
-So far, only the viral load trajectories have been considered; \textcite{hakkiOnset} does not consider this duration either.
-The time that individual $i$ are first and last positive are the two values of $\tau$ such that $\log v_i(\tau) = v_\text{lod}$.
+So far, only the viral load trajectories have been considered; \textcite{hakkiOnset} does not consider duration.
+The time that individual $i$ is first and last positive are the two values of $t$ such that $\log v_i(t) = v_\text{lod}$.
 Denote these times as $\tau_{0,i}$ and $\tau_{1,i}$.
-Hence, their duration of RT-PCR positivity, in continuous time, is $d_i = \tau_{1,i} - \tau_{0,i}$.
+Hence, individual $i$'s duration of RT-PCR positivity, in continuous time, is $d_i = \tau_{1,i} - \tau_{0,i}$.
 
-At $\tau_{0,i}$ and $\tau_{1,i}$, the approximation as exponential growth (described in \cref{ATACCC:sec:hakki}) is good.
-Therefore, set $\log v(\tau)$ and one of the exponential terms to 0 in \cref{ATACCC:eq:viral-load}.
-When $v_\text{lod} = 0$, as it does here, this gives the following expressions:
+At $\tau_{0,i}$ and $\tau_{1,i}$, the approximation as exponential growth (described in \cref{ATACCC:sec:hakki}) is good, meaning that one of the exponential terms in the denominator of \cref{ATACCC:eq:viral-load} is 0.
+Substituting in this, and $\log v_i(\tau_{0,i}) = \log v_i(\tau_{1,i}) = v_\text{lod} = 0$, gives the following expressions for $\tau_{0,i}$ and $\tau_{1,i}$.
 \begin{align}
@@
 From \cref{ATACCC:eq:di}, we can see that $d_i$ depends only on $i$'s random effects vector $\vec{\phi}_i$.
-Therefore, the distribution of $d_i$ across the population is determined by the $\vec{phi}_i$'s mean, $\vec{\mu}_\phi$, and covariance matrix, $\matr{\Sigma}_\phi$.
-Hence, the uncertainty in the posterior distribution of the discretised duration cumulative distribution function, $F_D(d) = \prob(D_i \leq d + 0.5)$ originate from the uncertainty in $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$.
+Therefore, the distribution of $d_i$ across the population is determined by $\vec{\phi}_i$'s mean, $\vec{\mu}_\phi$, and covariance matrix, $\matr{\Sigma}_\phi$.
+Hence, the uncertainty in the posterior distribution of the discretised duration cumulative distribution function for group $g$, $F_{D,g}(d) = \prob(D_i \leq d + 0.5)$ originate from the uncertainty in $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$.
 
-For a known $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$, $F_D$ is equal to the following expression.
+For a known $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$, $F_{D,g}$ is found as:
 \begin{align}
-  F_D(d)
+  F_{D,g}(d)
   &= \prob(D_i \leq d + 0.5) \\
@@
 The integral in \cref{ATACCC:eq:FDintegral} can be approximated using Monte Carlo integration (see \cref{ATACCC:alg:MonteCarlo}).
-Equivalent forms of $F_D$, such as the survival and hazard functions, can be calculated using the relations in \cref{E-inc-prev:sec:definitions}.
+Equivalent forms of $F_{D,g}$, such as the survival and hazard functions, can be calculated using the relations in \cref{E-inc-prev:sec:definitions}.
 
@@
   \For{$d = 1, \dots, \dmax$}{
-    Set $F_D^{(i)}(d) = \frac{1}{M} \sum_{j=1}^M \I(d_{i,j} \leq d + 0.5)$ \;
+    Set $F_{D,g}^{(i)}(d) = \frac{1}{M} \sum_{j=1}^M \I(d_{i,j} \leq d + 0.5)$ \;
   }
@@
   \caption{%
-  Algorithm to calculate the Monte Carlo integral in \cref{ATACCC:eq:FDintegral}, producing the posterior distribution of $F_D(d)$ for group $g$.
-  The posterior distribution is represented as a sample $F_D^{(1)}, \dots, F_D^{(N)}$.
+  Algorithm to calculate the Monte Carlo integral in \cref{ATACCC:eq:FDintegral}, producing the posterior distribution of $F_{D,g}(d)$.
+  The posterior distribution is represented as a sample $F_{D,g}^{(1)}, \dots, F_{D,g}^{(N)}$.
   $\vec{\mu}_\phi^{(i)}$ and $\matr{\Sigma}_\phi^{(i)}$ denote the $i$th posterior sample of $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$ respectively, taken from the MCMC output.
-  $M$ is the number of individuals to draw, I found $M = 100,000$ produced stable estimates of the tail of $F_D(d)$ if $\dmax = 45$.
+  $M$ is the number of individuals to draw, I found $M = 100,000$ produced stable estimates of the tail of $F_{D,g}(d)$ if $\dmax = 45$.
  }
@@
 The simplified model I proposed in \cref{ATACCC:sec:observation-modification} solved this issue, meaning that this analysis converged.
-
-The modified model produced good fits to the data (see \cref{ATACCC:fig:goodness-of-fits}).
-Almost all test results are captured in the 95\% credible intervals.
-Individual 55 has some fitting issues, explored further below.
-Individual 57 appears to have far noisier measurements than the other individuals, suggesting that the observation noise might vary between individuals.
-Including a random effect on $\sigma_v$ could capture this affect.
-However, it would complicate the model, be unlikely to change the estimates of the parameters of interest, and may not be identifiable.
-\begin{figure}
-  \thisfloatpagestyle{empty}
-  \vspace{-3cm}
-  \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{ATACCC/fits}}
-  \captionsetup{width=0.8\paperwidth}
-  \captionof{figure}[Posterior predictive viral load fits]{Posterior predictive observed viral loads (median and 95\% CrI, lines and ribbon) for each individual, overlaid with their observed test results (dots).
-Posterior estimates are conditional on the result not being a false negative.
-Negative RT-PCR tests are plotted at the limit of detection, \ie a log viral load of 0.
-Each plot is labelled with the individual's position in the dataset, an arbitrary value.
-\label{ATACCC:fig:goodness-of-fits}}
-\end{figure}
-
-For individual-level parameters, both models had convergence issues in places (Rhat > 1.4 and ESS < 20).
-However, this was greatly improved with the simplified model, occurring only for parameters relating to a single individual, number 55.
-Individual 55 has an extremely noisy set of observations.
-This means that two viral load trajectories fit the data well (shown in \cref{ATACCC:fig:individual-55}).
-The first, found by two chains, increases very quickly, peaks 4.0 days after the first test, then declines slowly.
-The second, found by the other six chains, is more symmetrical, peaking 7.2 days after the first test.
-The population-level parameter estimates agree between the two sets of chains, therefore the issue should not affect any estimates of interest.
-\begin{figure}
-  \centering \includegraphics{ATACCC/fit_individual_55}
-  \caption[Goodness-of-fit for individual 55]{Fit to individual 55's data coloured by MCMC chain.
-Two chains favour the quickly increasing viral load trajectory, peaking at day 4.0, the other six favour the slower trajectory, peaking at day 7.2.
-\label{ATACCC:fig:individual-55}}
-\end{figure}
 
@@
 The false negative rate, $\rho$, is estimated to be quite low, 5.3\% (95\% CrI: 3.7--7.3\%).
+The test sensitivity, which will be relevant for \cref{E-imperf-test}, is the probability that a detectable individual tests positive, $\psens = 1 - \rho$.
 
-The $F_D$ estimates for the two groups are in \cref{ATACCC:fig:duration}.
-In \cref{ATACCC:sec:results}, the two models gave very similar results.
-Therefore, I show results only from the modified model (which had better convergence).
+The $F_{D,g}$ estimates for the two groups are in \cref{ATACCC:fig:duration}.
+The vaccinated population have shorter durations of RT-PCR positivity\todo{check posterior probability that vaccinated population have shorter durations} than the unvaccinated population (see \cref{ATACCC:fig:duration}).
+The median duration of RT-PCR positivity for the vaccinated population is 13 days (95\% CrI: 11--16 days), but it is 16 days (95\% CrI: 14--19 days) for the unvaccinated population.
 \begin{figure}
@@
 
-The vaccinated population have shorter durations of RT-PCR positivity than the unvaccinated population (see \cref{ATACCC:fig:duration}).
-The median duration of RT-PCR positivity for the vaccinated population is 13 days (95\% CrI: 11--16 days), but it is 16 days (95\% CrI: 14--19 days) for the unvaccinated population.
+The modified model produced good fits to the data (see \cref{ATACCC:fig:goodness-of-fits}).
+Almost all test results are captured in the 95\% credible intervals.
+Individual 55 has some fitting issues, explored further below.
+\begin{figure}
+  \thisfloatpagestyle{empty}
+  \vspace{-3cm}
+  \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{ATACCC/fits}}
+  \captionsetup{width=0.8\paperwidth}
+  \captionof{figure}[Posterior predictive viral load fits]{Posterior predictive observed viral loads (median and 95\% CrI, lines and ribbon) for each individual, overlaid with their observed test results (dots).
+Posterior estimates are conditional on the result not being a false negative.
+Negative RT-PCR tests are plotted at the limit of detection, \ie a log viral load of 0.
+Each plot is labelled with the individual's position in the dataset, an arbitrary value.
+\label{ATACCC:fig:goodness-of-fits}}
+\end{figure}
+
+For individual-level parameters, both models had convergence issues in places (Rhat > 1.4 and ESS < 20).
+However, this was greatly improved with the simplified model, occurring only for parameters relating to a single individual, number 55.
+Individual 55 has an extremely noisy set of observations.
+This means that two viral load trajectories fit the data well (shown in \cref{ATACCC:fig:individual-55}).
+The first, found by two chains, increases very quickly, peaks 4.0 days after the first test, then declines slowly.
+The second, found by the other six chains, is more symmetrical, peaking 7.2 days after the first test.
+The population-level parameter estimates agree between the two sets of chains, therefore the issue should not affect any estimates of interest.
+\begin{figure}
+  \centering \includegraphics{ATACCC/fit_individual_55}
+  \caption[Goodness-of-fit for individual 55]{Fit to individual 55's data coloured by MCMC chain.
+Two chains favour the quickly increasing viral load trajectory, peaking at day 4.0, the other six favour the slower trajectory, peaking at day 7.2.
+\label{ATACCC:fig:individual-55}}
+\end{figure}
 
 \section{Discussion}
-
-This chapter has estimated the distribution of RT-PCR positive duration based on the viral load data from the ATACCC study.
 
@@
 All these factors could change the viral dynamics~\autocite{russellWithinhost} and hence the duration.
+\todo{Do the impact of these factors agree with other estiamtes? Does immunity matter more than other factors?}
 Therefore, the relationships shown here cannot be considered causal, or even generalisable to contexts where these other confounders are not present.
@@
 
+In the ATACCC study, individuals were tested for at most 20 days.
+Yet, the estimates here sill have 33\% (95\% CrI: 21--46\%) of individuals positive at 20 days following infection.
+The remainder of the estimate for $f_D$, such as that almost all (>99\%) of individuals recover before day 50, is driven by the assumed form of the viral load trajectory.
+
+Individuals 55 and 57 appear to have far noisier measurements than the other individuals (see \cref{ATACCC:fig:goodness-of-fits}), suggesting that the observation noise might vary between individuals.
+Including a random effect on $\sigma_v$ could capture this affect.
+However, it would complicate the model, be unlikely to change the estimates of the parameters of interest, and may not be identifiable.
+
+\section{Conclusion}
+
+This chapter has estimated the distribution of RT-PCR positive duration based on the viral load data from the ATACCC study.
 In the following chapters, I will use the estimates from the unvaccinated individuals.
@@
 
-A major issue with the estimates in this chapter is that they are extrapolated.
-In ATACCC, individuals were tested for at most 20 days.
-Yet, the estimates here sill have 33\% (95\% CrI: 21--46\%) of individuals positive at 20 days following infection and less than 1\% positive at 50 days.
-The remainder of the estimate for $f_D$ is driven by the assumed form of the viral load trajectory.
-However, the CIS has individuals testing positive for very long periods of time.
+A major issue with the estimates of duration presented here is that the number of individuals who test positive for a long period of time is based on extrapolation of the viral load model.
 In periods of low incidence following a large peak, the number of individuals who are currently positive is dominated by those who have been positive for a long time.
@@
 The next two chapters will incorporate CIS data to improve the estimates of these tail probabilities.
+Therefore, I will now incorporate the CIS data to improve the estimates of the tail probabilities of the duration of RT-PCR positivity.
 

### 2024-02-08 biology-data.tex
Overleaf comments/edits
proof read ATACCC

@@
 This process is known as \emph{reverse transcription}.
-In reverse transcription, a naturally-occurring enzyme known as reverse transcriptase generates cDNA (complementary DNA) from the RNA in the sample~\autocites{valasekPower}.
+In reverse transcription, a naturally-occurring enzyme known as reverse transcriptase generates cDNA (complementary DNA) from the RNA in the sample~\autocite{valasekPower}.
 Any sequence of cDNA has a one-to-one correspondence with a sequence of RNA.
@@
 Therefore, when testing for SARS-CoV-2, the Ct value is proportional to the negative log of the amount of SARS-CoV-2 RNA present in the original sample.
+
+\todo[inline]{Discuss noise in Ct values (batch effects, swabbing process, etc.) see commented-out text below}
+% The noise arises from a variety of sources, including the sampling/swabbing process and batch effects in the RT-PCR process~\autocite{hanRTPCR}.
+% A positive RT-PCR test has a Ct value.
+% Ct values are linearly related to viral load (see \cref{E-biology-data:sec:PCR})
+% A negative RT-PCR test signals either the viral load is too low to be detected, \ie below the limit of detection, or the test is a false negative.
+% Positive RT-PCR tests, especially those with low viral loads, could be false positives, for example caused by contamination.
 
@@
 For all positive test results, the viral load was measured using the Ct value and then converted to a viral load using a standard curve.
-Other data such as demography and symptoms was also collected.
+Other data such as demography and symptoms was also collected, although I do not use this information.
 For further details see \textcite{singanayagamDuration,hakkiOnset}.
@@
 
-\subsubsection{Estimating positivity rate} \label{biology-data:sec:posivity}
+\subsubsection{Estimating positivity rate} \label{biology-data:sec:positivity}
 

### 2024-02-08 incidence-prevalence.tex
proof read ATACCC

@@
 $K(t)$ must take non-negative integer values and be monotonically increasing in $t$.
-The interpretation of $K(t)$ is the number of events, here infections, that have occurred in the interval $(-\infty, t]$~\autocites[244]{yanDistribution}.
+The interpretation of $K(t)$ is the number of events, here infections, that have occurred in the interval $(-\infty, t]$~\autocite[244]{yanDistribution}.
 In the context of a disease, the point process can be discretised into the incidence on day $t$ as $Z_t = K(t) - K(t-1)$, where $t$ is an integer.
@@
 
-A commonly-used point process for infectious disease epidemiology is the time-inhomogenous Poisson process~\autocites[e.g., in the context of HIV,][]{brookmeyerMethod}{paganoHIV}{rosenbergBackcalculation}{brookmeyerBackcalculation}.
+A commonly-used point process for infectious disease epidemiology is the time-inhomogenous Poisson process~\autocite[e.g., in the context of HIV,][]{brookmeyerMethod}{paganoHIV}{rosenbergBackcalculation}{brookmeyerBackcalculation}.
 % A time-homogenous Poisson process is the special case where $\nu(r)$ is constant.
@@
 The first defining property of a Poisson process is that the counts are Poisson distributed.
-Formally, $K(s) - K(t) \dist \Poi\left(\int^s_t \nu(r) dr \right)$ where $\nu(r) \geq 0$ is the \emph{intensity} of the infection process at time $r$~\autocites[244]{yanDistribution}.
+Formally, $K(s) - K(t) \dist \Poi\left(\int^s_t \nu(r) dr \right)$ where $\nu(r) \geq 0$ is the \emph{intensity} of the infection process at time $r$~\autocite[244]{yanDistribution}.
 

### 2024-02-08 ATACCC.tex
clarify priors are on the log scale

@@
     \hline \\
-    Mean max viral load & $\mu_{v_\text{max}, g}$ & $\N(15, 15^2)$ \\
+    Mean max log viral load & $\mu_{v_\text{max}, g}$ & $\N(15, 15^2)$ \\
     Standard deviation of max log viral load & $\sigma_{v_\text{max}}$ & $\N_+(0, 10^2)$ \\
-    Mean rate of viral load increase & $\mu_{a, g}$ & $\N(1.25, 0.75^2)$ \\
-    Standard deviation of rate of viral load increase & $\sigma_{a}$ & $\N_+(0, 1)$ \\
-    Mean rate of viral load decrease & $\mu_{b, g}$ & $\N(0.5, 1.4^2)$ \\
-    Standard deviation of rate of viral load decrease & $\sigma_{b}$ & $\N_+(0, 1)$ \\
+    Mean rate of log viral load increase & $\mu_{a, g}$ & $\N(1.25, 0.75^2)$ \\
+    Standard deviation of rate of log viral load increase & $\sigma_{a}$ & $\N_+(0, 1)$ \\
+    Mean rate of log viral load decrease & $\mu_{b, g}$ & $\N(0.5, 1.4^2)$ \\
+    Standard deviation of rate of log viral load decrease & $\sigma_{b}$ & $\N_+(0, 1)$ \\
     Correlation of viral load parameters & $\matr{C}_\phi$ & $\LKJ(1)$ \\
@@
 \label{ATACCC:table:hakki-priors}}
-\todo[inline]{Check if the $\mu$ parameter priors are on the mean of log parameters or mean of parameters}
 \end{table}

### 2024-02-08 ATACCC.tex
add posterior probability that ATACCC shorter

@@
 The $F_{D,g}$ estimates for the two groups are in \cref{ATACCC:fig:duration}.
-The vaccinated population have shorter durations of RT-PCR positivity\todo{check posterior probability that vaccinated population have shorter durations} than the unvaccinated population (see \cref{ATACCC:fig:duration}).
+The vaccinated population very likely has a shorter duration of RT-PCR positive (93\% posterior probability of a shorter median duration) than the unvaccinated population (see \cref{ATACCC:fig:duration}).
 The median duration of RT-PCR positivity for the vaccinated population is 13 days (95\% CrI: 11--16 days), but it is 16 days (95\% CrI: 14--19 days) for the unvaccinated population.

### 2024-02-08 ATACCC/duration.R

@@ -43,3 +43,21 @@ ggsave(
     units = "cm",
     dpi = 300
 )
+
+bind_rows(
+    posterior_draws |>
+        mutate(Group = "Unvaccinated"),
+    posterior_draws2 |>
+        mutate(Group = "Vaccinated")
+) |>
+    summarise(
+        median = time[min(which(F >= 0.5))],
+        .by = c(.draw, Group),
+    ) |>
+    pivot_wider(
+        names_from = Group,
+        values_from = median,
+        id_cols = .draw
+    ) |>
+    mutate(vaccinated_shorter = Vaccinated < Unvaccinated) |>
+    summarise(p_vaccinated_shorter_median = mean(vaccinated_shorter))

### 2024-02-08 ATACCC.tex
previous viral load ODE models
impact of factors on duration

@@
 The first approach is an ODE-based approach uses parameters representing biological properties of the virus and host~\autocite[e.g.:][]{ejimaEstimation,keVivo,kimQuantitative,goncalvesTiming,perelsonMechanistic}.
-The model here is an application of a standard model for viral dynamics, used for many diseases\todo{cite history of viral load ODE models}.
+The model here is an application of a standard model for viral dynamics, used for many diseases~\autocite{perelsonMechanistic,hernandez-vargasInhost}.
 Such models are rarely analytically-tractable and the system's evolution is computed using numerical methods.
@@
 They were also infected by different variants because the vaccinated individuals were infected later.
-All these factors could change the viral dynamics~\autocite{russellWithinhost} and hence the duration.
-\todo{Do the impact of these factors agree with other estiamtes? Does immunity matter more than other factors?}
+All these factors could change the viral dynamics~\autocite{russellWithinhost,kisslerDynamicsVaccines} and hence the duration.
 Therefore, the relationships shown here cannot be considered causal, or even generalisable to contexts where these other confounders are not present.
+Despite these confounders, the result is in agreement with previous work showing that the Alpha and Delta variants of concern (compared to pre-Alpha variants) and being vaccinated are all associated with shorter durations of RT-PCR positivity, although the vaccination affect is larger~\autocite{kisslerDynamicsVaccines}.
 

### 2024-02-08 transmission.tex
proof read ATACCC
better justify only local transmission

@@
 % \end{enumerate}
-% These groupings align closely with those previously used for analysis of the CIS data~\autocites[e.g.:][]{houseInferring}{walkerTracking}.
+% These groupings align closely with those previously used for analysis of the CIS data~\autocite[e.g.:][]{houseInferring}{walkerTracking}.
 
@@
 Clearly, this is not realistic as travel between regions of England is common and unrestricted.
-However, for a large epidemic, the non-interaction is a good approximation~\autocite{birrellRealtimea}.
+However, except in an epidemic's early stages, the interaction between large regions has a negligible impact on the epidemics progression~\autocite[e.g.][]{birrellRealtimea,gogSpatial,eggoSpatial}.
 

### 2024-02-09 transmission.tex
Updates from Overleaf

--- 2024-02-08 transmission.tex
@@
 
-There are various advantages to using a mechanistic model to estimate transmission (see \cref{E-inc-prev:sec:infection-process}).
-This section applies the mechanistic modelling framework introduced in \cref{E-SEIR:sec:mechanistic-models} to estimate the transmission of SARS-CoV-2 in England.
+In this section, I apply the general mechanistic transmission modelling framework from \cref{E-SEIR} to the CIS prevalence data.
+The application of this framework follows \textcite{birrellRealtime}, with the most substantial context-specific work in parameterising the WAIFW matrix (\cref{SEIR:sec:WAIFW-application}).
+A novel approach (\cref{SEIR:sec:observation}) links this transmission model to the CIS prevalence data; the approach is generalisable to other studies collecting representative prevalence data, \eg REACT.
+In the mechanistic modelling literature, this part of the model is referred to as the observation model, however, in the terminology of \cref{E-inc-prev}, it includes both the prevalence process and the observation process.
+Having specified the full model (\cref{SEIR:sec:methods-application}), I conducted a simulation study to validate the inference process and identifiability (\cref{SEIR:sec:sim-study}), before applying the model to the CIS data (\cref{SEIR:sec:application}).
+In \cref{SEIR:sec:discussion}, I discuss the results; in \cref{transmission:sec:comparison} I compare the results to those in the previous section.
 
-The transmission part of the model follows \textcite{birrellRealtime}.
-A novel approach (\cref{SEIR:sec:observation}) links the transmission model to prevalence data (here from the CIS, but the approach is generalisable).
-In the mechanistic modelling literature, this part of the model is referred to as the observation model; however, in the terminology of \cref{E-inc-prev}, it includes both the prevalence process and the observation process.
+In common with \cref{E-SEIR:sec:mechanistic-models}, the notation in this section differs slightly from the rest of the thesis.
 
-This section uses the same notation as \cref{E-SEIR:sec:mechanistic-models}, following the infectious disease modelling literature rather than the statistical conventions used elsewhere in this thesis.
-
-\subsection{Methods}
+\subsection{Methods} \label{SEIR:sec:methods-application}
 
@@
 
-\subsubsection{WAIFW matrix}
+\subsubsection{WAIFW matrix} \label{SEIR:sec:WAIFW-application}
 
 A parsimonious model is needed to parameterise the WAIFW matrix (see \cref{E-SEIR:sec:structured-populations}).
-I follow the WAIFW matrix parameterisation of \textcite{birrellRealtime}, decomposing the rate of transmission from individuals in age group $a'$ to susceptible individuals in age group $a$ as $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'} N_{a'}$.
-This separates contact patterns, $\kappa_{aa'}$, which can vary over time; probabilities of transmission upon contact between age groups $a$ and $a'$, $q_{aa'}$, assumed constant; a region-wide effect for time, $\beta(t)$; and the population of the stratum, $N_{a'}$.
+I follow the WAIFW matrix parameterisation of \textcite{birrellRealtime}, decomposing the rate of transmission from infectious individuals in age group $a'$ to susceptible individuals in age group $a$ as $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'} N_{a'}$.
+This separates contact patterns, $\kappa_{aa'}(t)$, which can vary over time; probabilities of transmission upon contact between age groups $a$ and $a'$, $q_{aa'}$, assumed constant; a region-wide effect for time, $\beta(t)$; and the population of the stratum, $N_{a'}$.
 
@@
 % These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
-This method starts with pre-pandemic baseline estimates of: $\kappa_{aa'}$ from the POLYMOD study~\autocite{mossongSocial}, and time use from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
+This method starts with pre-pandemic baseline estimates of $\kappa_{aa'}$ from the POLYMOD study~\autocite{mossongSocial}, and time use from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
 POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on each contact's age and location (home, work, school, leisure, transport, or other).
@@
 $q_c$ can be interpreted as the relative probability that a child is infected (compared to an adult), given contact with an infectious individual.
-$q_c$ is then estimated, although using an informative prior (see \cref{SEIR:sec:priors}).
+%$q_c$ is then estimated, although using an informative prior (see \cref{SEIR:sec:priors}).
 
@@
 
-\todo[inline]{consider if this paragraph is too repetitive}
-
-The transmission model represents the infection process (introduced in \cref{E-inc-prev:sec:infection-process}).
-A variety of different approaches can be taken to link a transmission model to observations.
-In the context of CIS, this requires representing the prevalence and observation processes (introduced in \cref{E-inc-prev:sec:prevalence-process,E-inc-prev:sec:observation-process} respectively).
-The method to do so is known as the observation model.
-The formulation of this model is heavily dependent on the data available.
-This means that there is a wide variety of models used.
-This section  explains the novel modelling approach I propose for fitting to the CIS prevalence data.
+This section  explains the novel observation model approach I propose for fitting to the CIS prevalence data.
+This will link the model in the previous section to the CIS prevalence data.
 
@@
 One approach would be to convolve the incidence based on the transmission model with the RT-PCR duration distribution.
-Here, the daily incidence proportion for each stratum is by solving the system of ODEs governing the transmission model, and then prevalence calculated using \cref{E-inc-prev:eq:EPt}.
+Here, the daily incidence proportion for each stratum derived by solving the system of ODEs governing the transmission model, and then prevalence calculated using \cref{E-inc-prev:eq:EPt}.
 However, this approach is computationally expensive, likely prohibitively so if performed within a MCMC loop.
@@
 A computationally cheaper option, which I will take, is to approximate this convolution by adding compartments to the ODEs.
-This approach allows a infected, pre-positive compartment, which I denote $p_0$, to be included simply.
+This approach allows an infected, pre-positive compartment, which I denote $p_0$, to be included.
 $p_0$ increases at the same rate as individuals flow into $e_1$.
 These individuals then transition into a set of compartments with transition rates, $p_1, \dots, p_l$, chosen to match the duration of positivity estimates in \cref{E-imperf-test}.
+This will lead to the model in FIGURE - with transition rates etc etc
 The new compartments do not have a biological interpretation; the same idea is behind having multiple exposed or latent compartments (introduced in \cref{E-SEIR:sec:non-exponential}).
@@
 % A $l$-phase EC distribution is the sum of a $(l-2)$-phase gamma distribution and a two-phase acyclic, arbitrary phase-type distribution.
-An EC distribution $P$ can be defined as the sum of three random variables: $K$, $X_1$, and $X_2$.
+An $l$-phase EC distribution $P$ can be defined as the sum of three random variables: $K$, $X_1$, and $X_2$.
 It is shown schematically in \cref{SEIR:fig:EC}.
@@
 
-The name Erlang-Coxian comes from the two distributions that make up the EC distribution.
-Gamma distributions with integer first parameters (as $K$ has) are also known as Erlang distributions.
-The distribution $X_1 + X_2$ is a special case of the Coxian distribution.
-Hence, the overall distribution is known as a Erlang-Coxian distribution.
+% The name Erlang-Coxian comes from the two distributions that make up the EC distribution.
+% Gamma distributions with integer first parameters (as $K$ has) are also known as Erlang distributions.
+% The distribution $X_1 + X_2$ is a special case of the Coxian distribution.
+% Hence, the overall distribution is known as a Erlang-Coxian distribution.
 

### 2024-02-09 transmission.tex
Overleaf comments

@@
 
-A parsimonious model is needed to parameterise the WAIFW matrix (see \cref{E-SEIR:sec:structured-populations}).
-I follow the WAIFW matrix parameterisation of \textcite{birrellRealtime}, decomposing the rate of transmission from infectious individuals in age group $a'$ to susceptible individuals in age group $a$ as $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'} N_{a'}$.
-This separates contact patterns, $\kappa_{aa'}(t)$, which can vary over time; probabilities of transmission upon contact between age groups $a$ and $a'$, $q_{aa'}$, assumed constant; a region-wide effect for time, $\beta(t)$; and the population of the stratum, $N_{a'}$.
+A parsimonious model is needed to parameterise the WAIFW matrix.
+In \cref{E-SEIR:sec:structured-populations}, I explained how the elements of the WAIFW matrix, $\matr{\beta}$ can be decomposed as $\beta_{aa'} = \kappa_{aa'} q_{aa'} N_{a'}$ where\todo{should I repeat these definitions?
+they already exist in ch 3}: $\kappa_{aa'}$ is the mean rate at which an individual in stratum $a'$ makes contact with individuals in stratum $a$; $q_{aa'}$ is the probability of transmission upon contact between an infectious individual in stratum $a'$ and a susceptible individual in stratum $a$; and $N_{a'}$ is the total number of individuals in stratum $a'$.
+For many epidemics, where the population behaviour is fairly constant over time, a constant $\matr{\beta}$ is a good approximation.
+However, in SARS-CoV-2, population behaviour changed dramatically, both spontaneously (\eg due to fear) and in response to government-imposed restrictions.
+Therefore, following \textcite{birrellRealtime}, I introduce a time-dependence giving: $\beta_{aa'}(t) = \kappa_{aa'}(t) q_{aa'}(t) N_{a'}$; $N_{a'}$ is constant because the change in the total population is negligible over the length of time considered here.
 
@@
 
-Some behavioural changes (\eg mask wearing) impact transmission.
-However, these are not captured by the contact survey.
-To allow for these changes and the uncertainty produced in formulating $\matr{\kappa}(t)$, I follow \textcite{birrellRealtime} in placing a random walk prior on $\log\beta(t)$  on the weekly timescale:
+The probability of transmission, $q_{aa'}(t)$, needs further decomposition to be identifiable, for the same issue of dimensionality preventing each element of $\matr{\beta}$ being estimated independently (see \cref{SEIR:sec:structured-populations} for a full explanation).
+I assume that $q_{aa'}(t) = \beta(t) q^\text{susc}_{a} q^\text{inf}_{a'}$ where $\beta(t)$ is a time-varying reference probability of transmission; $q^\text{susc}_a$ is the relative \emph{susceptibility} of stratum $a$, \ie how vulnerable to infection individuals in stratum $a$ are; and $q^\text{inf}_{a'}$ is the relative \emph{transmissibility} of stratum $a'$, \ie how likely individuals in stratum $a$ are to pass on an infection.
+Separating relative transmissibility and susceptibility into two independent factors is a common approach\todo{cite anderson and may}.
+These measures include both biological differences, \eg due to differences in immune systems, or behavioural, \eg if infants are less hygienic than adults.
+
+The time-varying component, $\beta(t)$, allows for behavioural changes that impact transmission but are not captured by the contact matrices (\eg mask wearing).
+I follow prior work~\autocite[e.g.][]{birrellRealtime}\todo{find other references for random walk on transmission potential} in using a random walk prior on $\log\beta(t)$  on the weekly timescale.
+This prior encodes that behaviour changes smoothly over time.
 \begin{align}
@@
 Two hyperparameters will be inferred: $\beta_1$, the initial value of the random walk; and $\sigma_\epsilon$, the standard deviation of the random walk (see \cref{SEIR:sec:priors}).
+$\beta(t)$ is the only time-varying part of the transmission model that is not fixed a priori.
+Therefore, in practice, it will also absorb other effects such as any error in $\matr{\kappa}$.
 
@@
 % For any stratum $i$, $\beta^\text{susc}_i$ is known as the \emph{relative susceptibility} of stratum $i$.
-
-Finally, $q_{aa'}$ needs to be specified.
-I follow a common approach\todo{cite anderson and may}, decomposing it into two independent factors such that $q_{aa'} = q^\text{susc}_{a} q^\text{inf}_{a'}$, where $q^\text{susc}_a$ is the \emph{susceptibility} of stratum $a$, \ie how vulnerable to infection individuals in stratum $a$ are; and $q^\text{inf}_{a'}$ is the \emph{transmissibility} of stratum $a'$, \ie how likely individuals in stratum $a$ are to pass on an infection.
-% Then $\beta^\text{susc}_i$ is the relative risk of an individual in stratum $i$ being infected, relative to an individual in stratum $a_0$.
-These measures include both biological differences, \eg due to differences in immune systems, or behavioural, \eg if infants are less hygienic than adults.
 
@@
 This approach allows an infected, pre-positive compartment, which I denote $p_0$, to be included.
+Up until now, this stage of the disease (shown in \cref{E-biology-data:fig:natural-history}) has been neglected, assumed to be of negligble duration.
+An advantage of this modelling approach is that it can be included without complications (\eg issues of numerical stability).
+
 $p_0$ increases at the same rate as individuals flow into $e_1$.
-These individuals then transition into a set of compartments with transition rates, $p_1, \dots, p_l$, chosen to match the duration of positivity estimates in \cref{E-imperf-test}.
-This will lead to the model in FIGURE - with transition rates etc etc
-The new compartments do not have a biological interpretation; the same idea is behind having multiple exposed or latent compartments (introduced in \cref{E-SEIR:sec:non-exponential}).
+These individuals then transition into a set of compartments, $p_1, \dots, p_l$, each with their own transition rate.
+I will choose $l$ and the transition rates such that combined length of stay (\ie time from entry in $p_1$ until exit from $p_l$) approximates the duration of positivity distribution estimated in \cref{E-imperf-test}.
+This idea extends the technique of having multiple exposed or latent compartments (introduced in \cref{E-SEIR:sec:non-exponential}), and is commonly used to enforce correct durations in compartmental models\todo{cite phase-type distributions in compartmental models}.
 
-Deciding $l$ and the length of time each individual spends in each compartment is the most challenging part of formulating the observation model.
-The class of distributions which can be represented as a series of compartments are known as \emph{phase-type} distributions.
-Discussions of general phase-type distributions are beyond the scope of this thesis; a review in a biological context can be found in \textcite{hobolthPhasetype}.
-I will restrict myself to the subset of phase-type distributions defined by \textcite{osogamiClosed} as Erlang-Coxian (EC) distributions.
-EC distributions are convenient because they are flexible enough to approximate almost any distribution, and the parameters of the approximation are given by closed-form functions of the distribution's moments.
+\Textcite{osogamiClosed} provide a method to construct the compartments, implemented in the mapfit R package~\autocite{mapfit} implements the expressions to calculate the EC distribution parameters given a distribution's first three moments.
+Applying the package to the posterior mean of the distribution estimated in \cref{E-imperf-test} gives the model in \cref{SEIR:fig:full-model} with the parameter estimates in \cref{SEIR:table:ec-params} (see \cref{transmission:sec:phase-type} for details).
+\begin{table}
+    \centering
+    \begin{tabular}{c c c c c}
+        $l$ & $\omega_Y$ & $\omega_{X1}$ & $\omega_{X2}$ & $\pi$ \\
+        3 & 0.0820 & 0.126 & 0.0223 & 0.00973  \\
+    \end{tabular}
+    \caption{Parameter estimates for the EC distribution approximating the posterior mean of the duration distribution estimated in \cref{E-imperf-test}.}
+    \label{SEIR:table:ec-params}
+\end{table}
 
-% Formally, a phase-type distribution as the distribution of \emph{absorption time} in a \emph{continuous-time Markov process} with a single absorbing state~\autocite{hobolthPhasetype}.
-% Given a probability distribution over starting states, a rate at which 
-% The process can be defined by its size, $l$, a $l \times l$ \emph{transition matrix}, $\matr{M} \in \reals^{l \times l}$, an \emph{exit rate} vector, $\vec{\alpha}$ 
-% The process's state at time $t$, $W_t$, is a scalar $W_t \in \{ 1, \dots, l+1 \}$.
-% Its transition matrix is defined, for $j \neq k$, as $\prob(W_{t+\delta t} = k \mid ) = M_{jk} \delta t$ for small, positive $\delta t$.
-% The diagonal is defined as $M_{jj} = - \sum_{k \neq j} M_{jk}$ so that the matrix $\matr{M}$ has row sums 0.
-% The state $l+1$ is an absorbing state, that is $\prob(W_{t+t'} = l+1 \mid W_t = l_1) = 1$ for any non-negative $t'$ (once the process enters it, it cannot leave).
-% The exit rate gives the transition rates into the absorbing state: $\prob(W_{t+\delta t} = l + 1 \mid W_t = j) = \alpha_j \delta t$ for small $\delta t$.
-% The process's absorption time is when it first arrives at the absorbing state, $\tau = \inf \{ t \ssep W_t = l+1 \}$.
-% That is, $P$ is a phase-type distribution if it can be represented as a process 
-% They are additions of arbitrary exponential distributions, and can (if there is no limit on the number of compartments used) represent any distribution.
-% \todo{check/cite these claims}
-% I use the method of \textcite{osogamiClosed} to determine $l$ and the transition rates between the compartments.
-
-% A $l$-phase EC distribution is the sum of a $(l-2)$-phase gamma distribution and a two-phase acyclic, arbitrary phase-type distribution.
-An $l$-phase EC distribution $P$ can be defined as the sum of three random variables: $K$, $X_1$, and $X_2$.
-It is shown schematically in \cref{SEIR:fig:EC}.
-Here, $K \dist \GamDist(l-2, \omega_Y)$, $X_1 \dist \Exponential(\omega_{X1})$ and $X_2$ is a mixture distribution taking the value 0 with probability $1-p_x$ or otherwise (\ie with probability $p_x$) distributed $\Exponential(\omega_{X2})$.
-% The time between entering the 1st state and the $l+1$th state, the absorbing state, is the duration.
-These distributions contain five parameters.
-Three are positive reals, $\omega_Y$, $\omega_{X1}$, and $\omega_{X2}$; one is a probability, $p_x$, in the interval $[0, 1]$; and one is a positive integer, $l$.
-
-% The name Erlang-Coxian comes from the two distributions that make up the EC distribution.
-% Gamma distributions with integer first parameters (as $K$ has) are also known as Erlang distributions.
-% The distribution $X_1 + X_2$ is a special case of the Coxian distribution.
-% Hence, the overall distribution is known as a Erlang-Coxian distribution.
-
-\begin{figure}
-\makebox[\textwidth][c]{
-\begin{tikzpicture}[
-    node distance = 2.5cm,
-    on grid,
-    auto,
-    ->,>=stealth',
-    every state/.style={draw,rectangle},
-    ]
-
-    \node[state] (1) {1};
-    \node[state, right of=1] (2) {2};
-    \node[state, right of=2, draw=none] (dots) {$\dots$};
-    \node[state, right of=dots] (lm2) {$l-2$};
-    \node[state, right of=lm2] (lm1) {$l-1$};
-    \node[state, right of=lm1] (l) {$l$};
-    \node[state, right of=l, align=center] (absorbing) {absorbing\\state};
-
-    \path (1) edge node {$\omega_Y$} (2)
-          (2) edge node {$\omega_Y$} (dots)
-          (dots) edge node {$\omega_Y$} (lm2)
-          (lm2) edge node {$\omega_Y$} (lm1)
-          (lm1) edge node {$p_x \omega_{X1}$} (l)
-                edge [out=-45,in=-135] node[below] {$(1 - p_x) \omega_{X1}$} (absorbing)
-          (l) edge node {$\omega_{X2}$} (absorbing);
-\end{tikzpicture}
-}
-\caption[A $l$ phase EC distribution.]{Representation of an EC distribution as a series of compartments.
-The distribution is the time from entering state 1 and entering the absorbing state.
-The time between entering state 1 and state $l-2$ is distributed $K \dist \GamDist(l-2, \omega_Y)$, also known as an Erlang distribution.
-The time between entering state $l-2$ and the absorbing state has a Coxian distribution (see the main text for details).}
-\label{SEIR:fig:EC}
-\end{figure}
-
-\Textcite{osogamiClosed} define a phase-type distribution $P$ as well representing an arbitrary distribution $G$ if $P$ and $G$ have the same first three moments.
-They then prove that EC distributions have the following properties.
-\begin{itemize}
-    \item Let $\set{W}$ be the set of all phase-types distributions that well represent $G$ and $w$ be the minimum number of phases of any member of $\set{W}$.
-If $\set{W}$ is non-empty, then there exists a $l$-phase EC distribution $P' \in \set{W}$.
-    \item Furthermore, at least one such $P'$ has $l \leq w + 1$. Fewer phases is generally beneficial as it reduces the time and space complexity of working with the distribution.
-    \item The parameters of a $P'$ with $l \leq w + 1$ can be found using a closed-form expression (given by \textcite{osogamiClosed}).
-This is beneficial because finding arbitrary members of $\set{W}$ is a hard problem.
-\end{itemize}
-
-A total of five parameters are required to specify a $l$-phase EC distribution.
-\begin{itemize}
-    \item $l$, the number of phases.
-    \item $\omega_Y$, the rate parameter of the Erlang distribution.
-    \item $\omega_{X1}$, the rate parameter of the first state after the Erlang distribution.
-    \item $\omega_{X2}$, the rate parameter of the second state after the Erlang distribution.
-    \item $p_x$, the probability of moving from the end state of the Erlang distribution to the second (as opposed to directly to the absorbing state).
-\end{itemize}
-
-The mapfit package~\autocite{mapfit} implements the expressions to calculate the EC distribution parameters given a distribution's first three moments.
+The mapfit R package~\autocite{mapfit} implements the expressions to calculate the EC distribution parameters given a distribution's first three moments.
 Applying the package to the posterior mean of the distribution estimated in \cref{E-imperf-test} gives the parameter estimates in \cref{SEIR:table:ec-params}.
@@
     \begin{tabular}{c c c c c}
-        $l$ & $\omega_Y$ & $\omega_{X1}$ & $\omega_{X2}$ & $p_x$ \\
+        $l$ & $\omega_Y$ & $\omega_{X1}$ & $\omega_{X2}$ & $\pi$ \\
         3 & 0.0820 & 0.126 & 0.0223 & 0.00973  \\
@@
 I link the modelled proportion of individuals that are RT-PCR positive to the data using a beta-binomial likelihood (as discussed in \cref{transmission:sec:data}).
-If on each day $t$, each strata $i$ is observed to have $y_{it}$ positives out of $n_{it}$ tests, then the likelihood is:
+If on each day $t$, each strata $i$ is observed to have $y_{at}$ positives out of $n_{at}$ tests, then the likelihood is:
 \begin{align}
-    \prod_{a,t} \BB (y_{at} \mid n_{at}, \sum_{k=1}^l \bar{p}_{kat}, \rho)
+    \prod_{a,t} \BB (y_{at} \mid n_{at}, \bar{p}_{at}, \rho)
     \label{SEIR:eq:likelihood}
 \end{align}
-where $\bar{p}_{kat} = \int_{t}^{t+1} p_{ka}(s) ds$ is the mean occupancy in compartment $p_k$ and strata $a$ over day $t$ and $\rho$ is a nuisance parameter controlling the overdispersion of the beta-binomial distribution.
+where: $\bar{p}_{at}$ is the mean proportion of individuals in strata $a$ that are RT-PCR positive on day $t$; and $\rho$ is a nuisance parameter controlling the overdispersion of the beta-binomial distribution.
+Denoting the proportion of stratum $a$ in state $p_k$ as $p_{ka}(\tau)$, then $\bar{p}_{at} = \int_{t}^{t+1} \sum_{k=1}^l p_{ka}(\tau) d\tau$.
 
@@
           (P1) edge node {$\omega_Y$} (P2)
-          (P2) edge node {$p_x \omega_{X1}$} (P3)
-               edge [out=-45,in=-135] node[below] {$(1 - p_x) \omega_{X1}$} (P4)
+          (P2) edge node {$\pi \omega_{X1}$} (P3)
+               edge [out=-45,in=-135] node[below] {$(1 - \pi) \omega_{X1}$} (P4)
           (P3) edge node {$\omega_{X2}$} (P4);
@@
     \label{SEIR:eq:fullODEs}
-    \vec{\lambda}(t) &= \beta(t) (\matr{\kappa}(t) \circ \matr{q}) (\vec{i_1}(t) + \vec{i_2}(t)) \\
+    \vec{\lambda}(t) &= \beta(t) (\matr{\kappa}(t) \circ \matr{q}) ((\vec{i_1}(t) + \vec{i_2}(t)) \circ \vec{N}) \\
     \frac{d\vec{s}(t)}{dt} &= -\vec{\lambda}(t) \circ \vec{s}(t) \\
@@
     \frac{d\vec{p_2}(t)}{dt} &= \omega_Y \vec{p_1}(t) - \omega_{X1} \vec{p_2}(t) \\
-    \frac{d\vec{p_3}(t)}{dt} &= p_x \omega_{X1} \vec{p_2}(t) - \omega_{X2} \vec{p_3}(t).
+    \frac{d\vec{p_3}(t)}{dt} &= \pi \omega_{X1} \vec{p_2}(t) - \omega_{X2} \vec{p_3}(t)
 \end{align}
+where $\circ$ indicates element-wise multiplication.
 
@@
 
-The mean prevalence on day $t$, used for the likelihood (see \cref{SEIR:eq:likelihood}), can be calculated as $\bar{p}_{kat} = \frac{1}{2} ( p_{kat} + p_{ka,t+1/2} )$.
+The mean prevalence on day $t$, used for the likelihood (see \cref{SEIR:eq:likelihood}), can be calculated as $\bar{p}_{at} = \frac{1}{2} \sum_{k=1}^l ( p_{kat} + p_{ka,t+1/2} )$.
 
@@
     \vec{p_2} &= \vec{p_1} \frac{\omega_Y}{\psi + \omega_{X1}} \\
-    \vec{p_3} &= \vec{p_2} \frac{p_x \omega_{X1}}{\psi + \omega_{X2}}.
+    \vec{p_3} &= \vec{p_2} \frac{\pi \omega_{X1}}{\psi + \omega_{X2}}.
 \end{align}
@@
 \ifSubfilesClassLoaded{
+  \appendix
+  \subfile{transmission-appendix}
   \listoftodos

### 2024-02-09 transmission.tex
finish split into appendix

@@
 
-\Textcite{osogamiClosed} provide a method to construct the compartments, implemented in the mapfit R package~\autocite{mapfit} implements the expressions to calculate the EC distribution parameters given a distribution's first three moments.
+\Textcite{osogamiClosed} provide a method to construct the compartments, implemented in the mapfit R package~\autocite{mapfit}. 
+The resulting compartmental approximation matches the original distribution up to the first three moments.
 Applying the package to the posterior mean of the distribution estimated in \cref{E-imperf-test} gives the model in \cref{SEIR:fig:full-model} with the parameter estimates in \cref{SEIR:table:ec-params} (see \cref{transmission:sec:phase-type} for details).
@@
     \begin{tabular}{c c c c c}
-        $l$ & $\omega_Y$ & $\omega_{X1}$ & $\omega_{X2}$ & $\pi$ \\
-        3 & 0.0820 & 0.126 & 0.0223 & 0.00973  \\
-    \end{tabular}
-    \caption{Parameter estimates for the EC distribution approximating the posterior mean of the duration distribution estimated in \cref{E-imperf-test}.}
-    \label{SEIR:table:ec-params}
-\end{table}
-
-The mapfit R package~\autocite{mapfit} implements the expressions to calculate the EC distribution parameters given a distribution's first three moments.
-Applying the package to the posterior mean of the distribution estimated in \cref{E-imperf-test} gives the parameter estimates in \cref{SEIR:table:ec-params}.
-\begin{table}
-    \centering
-    \begin{tabular}{c c c c c}
-        $l$ & $\omega_Y$ & $\omega_{X1}$ & $\omega_{X2}$ & $\pi$ \\
+        $l$ & $\omega_1$ & $\omega_{2}$ & $\omega_{3}$ & $\pi$ \\
         3 & 0.0820 & 0.126 & 0.0223 & 0.00973  \\
@@
 
-I link the modelled proportion of individuals that are RT-PCR positive to the data using a beta-binomial likelihood (as discussed in \cref{transmission:sec:data}).
+I link the modelled proportion of individuals that are RT-PCR positive to the data using a beta-binomial likelihood (as discussed in \cref{E-biology-data:sec:positivity}).
 If on each day $t$, each strata $i$ is observed to have $y_{at}$ positives out of $n_{at}$ tests, then the likelihood is:
@@
           (P0) edge node {$\omega_0$} (P1)
-          (P1) edge node {$\omega_Y$} (P2)
-          (P2) edge node {$\pi \omega_{X1}$} (P3)
-               edge [out=-45,in=-135] node[below] {$(1 - \pi) \omega_{X1}$} (P4)
-          (P3) edge node {$\omega_{X2}$} (P4);
+          (P1) edge node {$\omega_1$} (P2)
+          (P2) edge node {$\pi \omega_{2}$} (P3)
+               edge [out=-45,in=-135] node[below] {$(1 - \pi) \omega_{2}$} (P4)
+          (P3) edge node {$\omega_{3}$} (P4);
 
@@
     \frac{d\vec{p_0}(t)}{dt} &= \vec{\lambda} \circ \vec{s}(t) - \omega_0 \vec{p_0}(t) \\
-    \frac{d\vec{p_1}(t)}{dt} &= \omega_0 \vec{p_0}(t) - \omega_Y \vec{p_1}(t) \\
-    \frac{d\vec{p_2}(t)}{dt} &= \omega_Y \vec{p_1}(t) - \omega_{X1} \vec{p_2}(t) \\
-    \frac{d\vec{p_3}(t)}{dt} &= \pi \omega_{X1} \vec{p_2}(t) - \omega_{X2} \vec{p_3}(t)
+    \frac{d\vec{p_1}(t)}{dt} &= \omega_0 \vec{p_0}(t) - \omega_1 \vec{p_1}(t) \\
+    \frac{d\vec{p_2}(t)}{dt} &= \omega_1 \vec{p_1}(t) - \omega_{2} \vec{p_2}(t) \\
+    \frac{d\vec{p_3}(t)}{dt} &= \pi \omega_{2} \vec{p_2}(t) - \omega_{3} \vec{p_3}(t)
 \end{align}
@@
     \vec{p_0} &= \vec{e_1} \frac{\psi + 2\sigma}{\psi + \omega_0} \\
-    \vec{p_1} &= \vec{p_0} \frac{\omega_0}{\psi + \omega_Y} \\
-    \vec{p_2} &= \vec{p_1} \frac{\omega_Y}{\psi + \omega_{X1}} \\
-    \vec{p_3} &= \vec{p_2} \frac{\pi \omega_{X1}}{\psi + \omega_{X2}}.
+    \vec{p_1} &= \vec{p_0} \frac{\omega_0}{\psi + \omega_1} \\
+    \vec{p_2} &= \vec{p_1} \frac{\omega_1}{\psi + \omega_{2}} \\
+    \vec{p_3} &= \vec{p_2} \frac{\pi \omega_{2}}{\psi + \omega_{3}}.
 \end{align}

### 2024-02-13 transmission.tex
reminder to comment on lack of non-response bias adj in SEIR
Updates from Overleaf

@@
 In \cref{E-inc-prev:sec:observation-process}, I showed that a deterministic approximation to the relationship between true prevalence and incidence is a good approximation for estimating incidence.
-The deterministic relationship between incidence proportion and prevalence means that either can be inferred, with the other then calculated.
+This relationship between incidence proportion and prevalence means that once one is inferred, the other is calculated via a deterministic function.
+Therefore, a statistical methodology for estimating one of these quantities is required.
+
 Estimation of prevalence from the CIS data has been carried out through a well-established methodology, used throughout the pandemic~\autocite{pouwelsCommunity,cisMethodsONS}.
-The method is based on multilevel regression and post-stratification (MRP)\todo{cite MRP}.
-In this section, I will extend this methodology to also estimate incidence.
+The method is based on MRP, a statistical methodology developed to correct political opinion polling suffering from response bias\todo{cite MRP}.
+MRP is a two-step process.
+In this context, the first step, multilevel regression, consists of using a Bayesian multilevel logistic regression model is to predict RT-PCR test results in the sample based on demographic covariates.
+The second step, poststratification, is using the posterior distribution of the parameters is used to predict the population prevalence in each stratum.
+These are combined by taking a weighted average across the strata, where a stratum's weight is equal to its (known) population size.
+In this section, I propose extending this process by transforming the posterior distribution of prevalence in each stratum to a posterior distribution of incidence before post-stratifying.
 
-First, I will motivate this approach, as opposed to direct estimation of incidence.
-Denote the quantity to be inferred (incidence proportion or prevalence) at time $t$ in strata $i$ as $\pi_{it}$.
-For any $i$ and $t$, the number of positives will be small, often fewer than 5; therefore some form of borrowing between time and strata is required.
+As in prior work, I parameterise the regression model in terms of prevalence, not incidence proportion.
+This is required to fulfil several desirable features of the model.
+Denote the quantity to be inferred (which could be either incidence proportion or prevalence) at time $t$ in strata $i$ as $\pi_{it}$.
+The strata will represent at least age group and region, although including ethnicity and sex is also due to non-response in the survey (see \cref{E-biology-data:sec:positivity}).
+\todo{make sure ch 2 mentions strata explicitly}
+For many combinations of stratum $i$ and day $t$, the number of positives will be small, often fewer than 5; therefore some form of borrowing between time and strata is required.
 The statistical model and inference procedure should have the following properties.
 \begin{enumerate}
-    \item The smoothing model over time should penalise towards exponential growth.
-Penalising towards exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time.
+    \item The smoothing model over time should penalise changes from continuous exponential growth.
+        Penalising changes from continuous exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time.
         Furthermore, such penalisation has previously been shown to improve estimates of SARS-CoV-2 prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
@@
         That is, incidence proportion and prevalence should be in the interval $[0, 1]$.
-        While incidence proportion can theoretically be greater than 1, this would imply that some individuals were infected multiple times in a day which is not biologically plausible.
+        % While incidence proportion can theoretically be greater than 1, this would imply that some individuals were infected multiple times in a day which is not biologically plausible.
     \item Have sufficient parameters to incorporate all necessary covariates to correct for non-response bias (see \cref{E-biology-data:sec:positivity}).
@@
 
-Properties 1 and 2 suggest a model of the form $\logit(\pi_{it}) = g(i, t)$, where $g$ is a function penalised towards being linear in $t$.
-The $\logit$ link function both limits $\pi_{it}$ to the interval $[0, 1]$, and logit-linear growth approximates exponential growth for small $\pi_{it}$.
-Combined with property 4, the form of $g$ should be penalised towards $g(i, t) = a_0 + a_t t + f(i)$ where $a_0$ and $a_t$ are model parameters and $f$ is a function to be specified.
-However, it should be possible for sufficient information in the data to move the posterior estimate of $\pi_{it}$ away from this model.
+Properties 1 and 2 suggest a model of the form $\logit(\pi_{it}) = g(i, t)$, where $g$ is a function with changes in the first derivative penalised (\ie changes away from linear are penalised).
+The $\logit$ link function both limits $\pi_{it}$ to the interval $[0, 1]$, and logit-linear growth approximates exponential growth for small $\pi_{it}$~\autocite{ealesAppropriately}.
+Combined with property 4, the form of $g$ the null model is $g(i, t) = a_0 + a_t t + f(i)$ where $a_0$ and $a_t$ are model parameters and $f$ is a function to be specified; that is, changes away from this model should be penalised.
+To add additional flexibility,
 
-Properties 3 and 4 suggest the extension of the previously-developed MRP method.
+% Properties 3 and 4 suggest the extension of the previously-developed MRP method, used to already estimate prevalence, which already satisfies these properties.
 % This method has previously been used successfully to estimate prevalence, corrected for non-response bias, from the CIS data~\autocite{cisMethodsONS,pouwelsCommunity,pouwelsMRPvaccination}.
-In MRP, a Bayesian multilevel logistic regression model is used to predict RT-PCR test results based on demographic covariates.
-Then, the posterior distribution is used to predict the prevalence in each stratum.
-Finally, the prevalence is post-stratified to the CIS target population and sub-populations of interest.
-I propose this process by transforming the posterior distribution of prevalence in each stratum to a posterior distribution of incidence before post-stratifying.
-Further details on this methodology follow.
 
 Properties 3 and 5 suggest the use of an approximate method for computing the posterior distribution.
-In particular, MCMC should be avoided.
-Incorporating sufficient covariates to adjust for the non-response bias over a period of more than 6--8 weeks leads to a model that was too computationally expensive~\citePersonalComms{Koen Pouwels}.
-Instead, previous MRP implementations have used the Integrated Nested Laplace Approximation (INLA)\todo{cite INLA}, implemented in R-INLA\todo{cite INLA}.
+In particular, using MCMC, sufficient covariates to adjust for the non-response bias, and a period of more than 6--8 weeks leads to a model that is computationally prohibitive ~\citePersonalComms{Koen Pouwels}.
+\todo{make sure SRS computational issues are mentioned in ch 1 or 2}
+Instead, previous MRP implementations have used the Integrated Nested Laplace Approximation (INLA)~\autocite{rueINLA}, implemented in R-INLA~\autocite{RINLA}\todo{add an appendix explaining INLA}.
 
-The use of R-INLA requires that $\pi_{i,t}$ represent incidence proportion rather than prevalence.
-If $\vec{\pi}$ represented incidence proportion, then $\logit \vec{pi}$ could be linear in the model parameters, as explained previously, while satisfying the other criteria.
-However, the expected prevalence at time $t$ is linear in $\vec{\pi}$ (see \cref{E-inc-prev:eq:EPt}), and hence non-linear in $\logit \vec{pi}$.
-The expected proportion of positive tests (\ie the expectation of the data likelihood) is equal to the prevalence.
-Applying a non-linear transformation then a linear transformation in this way to form the data likelihood's expectation cannot be implemented within R-INLA's framework.
+INLA-based inference can only be applied to LGMs (\emph{Latent Gaussian Models})~\autocite{rueINLAseminar,rueINLA,martinoINLAtutorial}.
+\todo{check notation here doesn't clash with other notation}
+An LGM is defined in terms of a vector of hyperparameters, $\vec{\theta}$; a deterministic transformation of the hyperparameters into covariance matrix, $\matr{\Sigma}(\vec{\theta})$; a vector of latent variables, $\vec{x}$; a vector of $n$ data points, $\vec{y} = [y_1, \dots, y_n]^T$; and $n$ linear predictors $\vec{\eta} = [\eta_1, \dots, \eta_n]^T$.
+To be an LGM, $\vec{x}$ must be a latent Gaussian field, that is $\vec{x} \mid \vec{\theta} \dist \N(\vec{0}, \matr{\Sigma}(\vec{\theta}))$; and the likelihood must have the conditional independence property $p(\vec{y} \mid \vec{\theta}, \vec{x}) = \prod_j p(y_j \mid \eta_j(\vec{x}), \vec{\theta})$.
 
-I modified a pre-existing version of a MRP model for CIS~\autocite{pouwelsMRPvaccination}, as recommended by that model's author for this situation~\citePersonalComms{Koen Pouwels}.
-First, a Gaussian Markov Random Field (GMRF) logistic regression model is used to estimate the prevalence in each stratum.
+In this context, each element $\vec{y}$ corresponds to the observed RT-PCR results in one stratum on one day.
+Hence, $\eta_{j}$ must determine the distribution of $y_{j}$ (up to some hyperparameters); where $j$ can be mapped to a mapping of a stratum $i$ and day $t$.
+In practice, this means that $\eta_j$ is a one-to-one transformation of the population prevalence in stratum $i$ on day $t$.
+If $\eta_j = \logit(\pi_{it})$, where $\pi$ is prevalence, then a model of $\eta_j = g(i, t)$, as before, is a LGM.
+However, if $\pi$ is instead representing incidence proportion, then no such linear relationship between $\eta_j$ and $g$ exists.
+In particular, if $\eta_j$ is the prevalence, then $\eta_{it} = \sum_{k=-\infty}^0 \pi_{ik} S(k) = \sum_{k=-\infty}^0 \expit(g(i, k)) S(k)$.
+As $g(i, k)$ is linear in the model parameters, $\vec{x}$, then $\eta_{it}$ is not, violating the requirements of a LGM.
+Therefore, the application of the INLA approximation, while satisfying the other desirable properties, requires that the regression model predicts RT-PCR prevalence, not incidence proportion.
+
+Next, I explain the modification of the previous logistic regression model, and then how to extend the procedure to calculate incidence.
+Specifically, I modified the regression compomenet of a pre-existing version of a MRP model for CIS~\autocite{pouwelsMRPvaccination}, as recommended by that model's author for this situation~\citePersonalComms{Koen Pouwels}.
 For a day $t$, age group $a$, ethnicity $e$, and sex $s$ assume that:
@@
 \begin{itemize}
+    \item $n_{taes}$ and $y_{taes}$ are the number of individuals tested and the number of positive test results on day $t$ in individuals of age group $a$, ethnicity $e$, and sex $s$.
+I use a beta-binomial distribution, as discussed in \cref{transmission:sec:data}.
     \item $\beta$ is a global intercept term with the prior $\beta \dist \N(0, 1/0.001^2)$.
@@
     \item $l_{a,t}$ is an age-group specific second-order random walk. That is $l_{a,t} = 2l_{a,t-1} - l_{a,t-2} + \epsilon_{a,t}$ where $\epsilon_{a,t} \dist \N(0, \sigma_l^2)$.
-    \item $n_{taes}$ and $y_{taes}$ are the number of individuals tested and the number of positive test results on day $t$ in individuals of age group $a$, ethnicity $e$, and sex $s$.
-The likelihood is a beta-binomial model (see discussion in \cref{transmission:sec:data}).
-an overdispersed binomial.
 \end{itemize}
 All parameters have the default R-INLA priors, except where stated above.
+\todo{include what the default priors are (maybe in the appendix)}
 For these parameters, the large amount of data means that the prior has little influence on the posterior distribution.
 
-The model was fitted separately and independently for each region, implicitly adding all interactions between region and the other variables.
+I fitted the model separately and independently for each region, implicitly adding all interactions between region and the other variables.
 This step assumes that the estimates for each region can be considered independent, which is reasonable because the regions are large with many tests conducted within each.
@@
 Having fitted the prevalence model, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution.
+\todo{include how R-INLA samples from the posterior in an appendix}
 For each stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
-I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:observation-process} justifies, that is the likelihood $\vec{P} \mid \vec{Z}$ is well-approximated by only considering its mean.
-\Cref{E-inc-prev:eq:EPt}, considered across different values of $t$, can be written as $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix formed from the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
+I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:observation-process} justifies, that is the estimate of the incidence is negligibly affected by approximating the distribution $\vec{P} \mid \vec{Z}$ by its mean.
+\Cref{E-inc-prev:eq:EPt}\todo{consider writing the matrix version where this is introduced in ch 3}, considered across different values of $t$, can be written as $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix formed from the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
 Dividing each side by the total number of individuals in each stratum transforms number of prevalent individuals to prevalence and incidence to incidence proportion.
@@
 \end{figure}
+
+\todo[inline]{check discussing non-INLA approximations}
 
@@
 I solve the system of ODEs using Euler's method, discretised at timesteps of $\delta t = 1/2$ days.
-That is, if $\vec{x}_t$ is the system state at time $t$ for $t \in \{ 0, 0.5, 1, \dots \}$ then $x_{t+\frac{1}{2}} = x_t + (d\vec{x}(t)/dt) \delta t$
+That is, if $\vec{x}_t$ is the system state at time $t$ for $t \in \{ 0, 0.5, 1, \dots \}$ then $x_{t+\frac{1}{2}} = x_t + (d\vec{x}(t)/dt) \delta t$.
 
@@
 The parameters governing the duration of positivity ($\omega$s) were explained previously (see \cref{SEIR:table:ec-params}).
-The initial conditions and transition rate parameters are unindentifiable in a SEIR model~\autocite{dankwaStructural}, therefore strongly informative priors are used for them.
-$\sigma_\epsilon$ and $\rho$ are given priors that penalise model complexity.
+The initial conditions and transition rate parameters are unindentifiable in a SEIR model~\autocite{dankwaStructural}, therefore strongly informative priors or fixed values are used for them.
+$\sigma_\epsilon$ and $\rho$ are given priors that favour a simpler model.
 The prior on $\sigma_\epsilon$ is a penalised complexity prior, favouring a null model of constant values for $\beta(t)$~\autocite{simpsonPenalising}.
@@
 \end{table}
+
+\todo[inline]{insert where the seroprevalence priors come from}
 
@@
 \begin{figure}
+    \todo[inline]{Add (A) incidence to the top panel and (B) prevalence to the bottom panel}
     \thisfloatpagestyle{empty}
@@
 
-In this setup, the parameters $i^+$, $q_c$, and $\psi$ are fully recovered; the other parameters are only partially so: their posterior has moved away from the prior but is still regularised by it (see \cref{SEIR:fig:true-vs-posterior}).
+In this setup, the parameters $i^+$, $q_c$, and $\psi$ are fully identified; $\pi_0$ is not identified at all; the other parameters are partially identified, their posterior has moved away from the prior but is still regularised by it (see \cref{SEIR:fig:true-vs-posterior}).
 The ability to identify $q_c$ is an advantage of using data which includes measurements on children, unlike data on hospitalisations or deaths.
+\todo{is $\pi_0$ unindentified because of $q_c$ - check correlation plot}
 
@@
 A constant $\matr{\beta}(t)$ would lead to exponential growth following a period of transitory dynamics, although the weekly switching may mean that these transitory dynamics never end, especially when incidence is high~\autocite{rhodesConvergence}.
-When the peak in incidence is estimated to occur varied by region, with the most common time being early January 2021 (see \cref{SEIR:fig:peak-incidence}).
+The estimated timing of the peak in incidence varies by region, with the most common time being early January 2021 (see \cref{SEIR:fig:peak-incidence}).
 \begin{figure}
@@
 The attack rate is highest in the $[11, 16)$ age group in the North West at 68\% (95\% CrI: 66\%--70\%) (see \cref{SEIR:fig:attack-rates}).
+\todo{insert increase in attack rate over period}
 \begin{figure}
@@
 Overall the model fits the data well (see \cref{SEIR:fig:prev-young,SEIR:fig:prev-old}).
+\todo{move one or some of the goodness-of-fit plots into an appendix - maybe by region?}
 There are a few features of the data which are not captured by the model and the posterior estimate appears to be a compromise with the prevalence too low in one age group and too high in another.

### 2024-02-13 transmission.tex

@@
 
-The probability of transmission, $q_{aa'}(t)$, needs further decomposition to be identifiable, for the same issue of dimensionality preventing each element of $\matr{\beta}$ being estimated independently (see \cref{SEIR:sec:structured-populations} for a full explanation).
+The probability of transmission, $q_{aa'}(t)$, needs further decomposition to be identifiable, for the same issue of dimensionality preventing each element of $\matr{\beta}$ being estimated independently (see \cref{E-SEIR:sec:structured-populations} for a full explanation).
 I assume that $q_{aa'}(t) = \beta(t) q^\text{susc}_{a} q^\text{inf}_{a'}$ where $\beta(t)$ is a time-varying reference probability of transmission; $q^\text{susc}_a$ is the relative \emph{susceptibility} of stratum $a$, \ie how vulnerable to infection individuals in stratum $a$ are; and $q^\text{inf}_{a'}$ is the relative \emph{transmissibility} of stratum $a'$, \ie how likely individuals in stratum $a$ are to pass on an infection.
@@
 These patterns can be interpreted as an instance of the common bias---variance trade-off.
+\todo{Comment on SEIR not fully adjusted for non-response bias}
 The SEIR model imposes stronger assumptions about the epidemic dynamics.

### 2024-02-13 transmission.tex
alpha vs incidence increase in backcalc

--- 2024-02-13 transmission.tex
@@
 
-During November, incidence starts rising, with regional heterogeneity in the timing of when this occurs.
+During November, incidence starts rising, with regional heterogeneity in the timing of when this occurs (see \cref{transmission:fig:backcalc-regions}).
 The most explored explanation for this is the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocite{walkerTracking,daviesEstimated,lythgoeLineage}.
+There is weak evidence that when incidence starts increasing is correlated with when the Alpha variant's spread (see \cref{transmission:fig:backcalc-alpha}).
 A second peak occurs in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
@@
     \label{transmission:fig:backcalc-regions}
+\end{figure}
+\begin{figure}
+    \includegraphics{transmission/backcalc-alpha}
+    \caption[Correlation of the emergence of Alpha and minimum incidence.]{%
+        x-axis: time that S-gene target failure, a marker of the Alpha variant, is estimated to start increasing from baseline by region among CIS test results; from \textcite[table S3]{walkerTracking}.
+        y-axis: posterior median and 95\% CrI of the date of minimum incidence in November or early December in each region.
+        Linear regression line shown.
+    }
 \end{figure}

### 2024-02-13 incidence-prevalence.tex
ch 3 proofing

@@
 This model is guided by the specific study design of the CIS.
-Therefore, the model is specific to this context, although much would be applicable to other, similarly designed prevalence surveys.
 
@@
 \emph{Incidence} on day $t$, $Z_t$, is the number of people in a given population who become infected on day $t$; when expressed as a proportion of the population it is the \emph{incidence proportion}~\autocite[89]{lashModern}, $Z_t/\Npop$, where $\Npop$ is the number of people in the population.
-\todo{add discussion of proportions vs rates}
 As discussed in\todo{ref intro}, the timing of infection events is very rarely directly observable and must be inferred from other data.
@@
 \emph{Prevalence} of a disease on day $t$, is the proportion of people with the disease on that day~\autocite[90]{lashModern}.
-What it means to have a disease can vary by context (see\todo{ref section discussing what it means to have covid}).
-Here, an individual is defined to have SARS-CoV-2 on day $t$ if, had they been swabbed on day $t$, that swab would return a positive RT-PCR test.
+Within this thesis, I define an individual as having SARS-CoV-2 on day $t$ if, had they been swabbed on day $t$, that swab would return a positive RT-PCR test.
 I refer to these individuals as \emph{prevalent} individuals.
@@
 Consider an individual $i$ infected on day $B_i$ that is prevalent during the interval $[B_i, E_i]$.
+\todo{check $B_i$ and $E_i$ are not defined elsewhere}
 I define this period as an \emph{infection episode}, beginning on day $B_i$ and ending on day $E_i$.
@@
 
-Afterwards, in \cref{SEIR}, I introduce \emph{mechanistic models}.
+Afterwards, in \cref{SEIR:sec:mechanistic-models}, I introduce \emph{mechanistic models}.
 Mechanistic models explicitly describe the population and transmission between individuals.
@@
 
-A commonly-used point process for infectious disease epidemiology is the time-inhomogenous Poisson process~\autocite[e.g., in the context of HIV,][]{brookmeyerMethod}{paganoHIV}{rosenbergBackcalculation}{brookmeyerBackcalculation}.
+A commonly-used point process for infectious disease epidemiology is the time-inhomogenous Poisson process~\autocite[e.g., in the context of HIV,][]{brookmeyerMethod,paganoHIV,rosenbergBackcalculation,brookmeyerBackcalculation}.
 % A time-homogenous Poisson process is the special case where $\nu(r)$ is constant.
@@
 Therefore, an overdispersed epidemic modelled as a Poisson process will be less variable than the realised epidemic.
+Overdispersion is common in infectious diseases, especially for coronaviruses such as SARS-CoV-2~\autocite{endoEstimating,adamClustering,mccloskeySARS}.
 
@@
 This is especially true when making noisy observations of the system, and that the stochastic differences between the stochastic and its mean are negligible compared to the observation noise.
-A formalisation of this argument underpins the mean-field approximation of the prevalence process made in \cref{E-inc-prev:sec:observation-process}.
+I make a formalisation of this argument to justify a mean-field approximation of the prevalence process in \cref{E-inc-prev:sec:observation-process}.
 
@@
 This choice is appropriate for the context of this thesis because there were many individuals infectious with SARS-CoV-2 in this time period.
-There are a wealth of introductory texts on stochastic models, I recommend \textcite[chapter 6]{keelingModeling}.
+Hence, further details on stochastic models are not outside the scope this thesis; a wealth of introductory texts with further details on stochastic models are available, I recommend \textcite[chapter 6]{keelingModeling}.
 
@@
 
-The simplest compartmental model for a disease which confers immunity is the (SIR) model\todo{cite SIR being simplest} (depicted in \cref{SEIR:fig:SIR}).
+The simplest compartmental model for a disease which confers immunity is the SIR (susceptible-infectious-recovered) model\todo{cite SIR being simplest} (depicted in \cref{SEIR:fig:SIR}).
 The name SIR derives from the structure of the model.
@@
 The expression for the force of infection, $\lambda(t)$ can be arrived at by considering a single infectious individual~\autocite[214]{kretzschmarMathematical}.
-Assume this individual, on average, comes into contact with other individuals in the population at a rate of $\kappa$.
+Assume this individual comes into contact with other individuals in the population at a mean rate of $\kappa$.
 Further, assume the population is \emph{well-mixed}, meaning that each contact is with an individual chosen uniformly at random from the population; this is implied by assuming a homogeneous population.
@@
 % Each of the infectious individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
-Therefore, in expectation, the infectious individual makes contact with susceptible individuals at a rate of $\kappa s(t)$ per day.
+Therefore the infectious individual makes contact with susceptible individuals at a mean rate of $\kappa s(t)$ per day.
 Assume that there is a constant probability of transmission upon contact between an infectious and susceptible individual, $q$.
-Combined with the above, the expected rate of infections caused by this infectious individual is $\kappa s(t) q$ per day.
-The two constant terms here (number of contacts and probability of transmission) are absorbed into the transmission rate, $\beta = \kappa q$.
+Hence, the mean rate of infections caused by this infectious individual is $\kappa s(t) q$ per day.
+The two constant terms are absorbed into the transmission rate, $\beta = \kappa q$.
 Therefore, the rate of new infections generated by a single infectious individual is $\beta s(t)$ per day.
@@
 \begin{itemize}
-    \item For any short period of time when $s(t)$ is approximately constant, the solution to the system of ODEs includes exponential growth in the number of infectious individuals: $i(t) = i_0 e^{\psi t}$ for some constants $i_0$ and $\psi$~\autocite[section 1.2]{diekmannMathematical}.
+    \item For any short period of time when the change in $s(t)$ is negligible, the solution to the system of ODEs includes exponential growth in the number of infectious individuals: $i(t) = i_0 e^{\psi t}$ for some constants $i_0$ and $\psi$~\autocite[section 1.2]{diekmannMathematical}.
     A notable example is the early part of an epidemic when $s(t) \approx 1$.
-    This is the phase of initial exponential growth of an epidemic, which produces the characteristic early epidemic curves.
+    This is the phase of initial exponential growth of an epidemic, producing the characteristic early epidemic curves.
     It is in this phase that the basic reproduction number provides a good approximation of the epidemic dynamics.
     \item The basic reproduction number for the SIR model is $\R = \beta / \gamma$~\autocite[20]{keelingModeling}.
-    Intuitively, this can be thought of as the average number of infections caused by an infectious individual per day, $\beta$, multiplied by the average number of days for which they are infectious, $1/\gamma$.
+    Intuitively, this can be thought of as the mean number of infections caused by an infectious individual per day, $\beta$, multiplied by the mean number of days for which they are infectious, $1/\gamma$.
     \item The effective reproduction number for this model is $R_e(t) = \beta / \gamma s(t) = \R s(t)$~\autocite{pellisEstimation}.
@@
     First, the total population size is constant.
-    Second, only a negligble number of infections are imported from outside the population.
-    Third, births and deaths of individuals are negligble over the length of time considered; simple extensions allow including them~\autocites[26]{keelingModeling}[214]{kretzschmarMathematical}.
+    Second, only a negligible number of infections are imported from outside the population.
+    Third, births and deaths of individuals are negligble over the length of time considered; model extensions allow these factors in contexts where they are relevant~\autocites[26]{keelingModeling}[214]{kretzschmarMathematical}.
     \item There is no analytical solution to the system of ODEs: numerical methods are required to solve the system~\autocite[25]{keelingModeling}.
-    \item As notes previously, the parameters $\vec{\theta}$ and the intial state of the system, $\vec{x}(0)$ is required to solve the system.
+    \item As noted previously, the parameters $\vec{\theta}$ and the intial state of the system, $\vec{x}(0)$ are required to solve the system.
     \item Immunity is assumed to last until the end of the  period considered ~\autocite[61]{andersonInfectious}.
@@
 This is an unrealistic assumption for most diseases and would lead to underestimating of the reproduction number~\autocite{lloydRealistic,wearingAppropriate}.
-A standard extension to the SIR model is to instead assume a gamma distribution~\autocite[94]{keelingModeling}{andersonSpread}.
+A standard extension to the SIR model is to instead assume a gamma distribution~\autocite[94]{keelingModeling,andersonSpread}.
 This is done by adding multiple infectious compartments, shown in \cref{SEIR:fig:SIR-gamma}.
-The system can now be written as follows:
+The ODEs are now:
 \begin{align}
@@
 As $n$ increases, the between-individual variability in infectious period length reduces, tending to a constant as $n \to \infty$~\autocite{lloydRealistic}.
-
-The new compartments are biologically identical, without an interpretation differentiating the compartments.
-They should be considered as a mathematical trick so that the ODE system produces a gamma distribution rather than an interpretable quantity.
-
-% (it is slightly unclear what it means for a variable to have a distribution in a deterministic model, the clearest explanation is that this is the distribution which the ).
 
@@
 
-Similar to the infectious period, the latent period can be modelled as a gamma distribution by adding multiple latent compartments.
+Similar to the infectious period, the latent period can be modelled as a gamma distribution using multiple latent compartments, $e_1, dots, e_m$.
 \begin{align}
@@
 \end{align}
-where $m$ is the number of latent compartments and $1/\sigma$ is the mean duration of the latent period.
+where $1/\sigma$ is the mean duration of the latent period.
 The latent period is distributed $\GamDist(m, m\sigma)$ and the infectious period is still distributed $\GamDist(n, n\gamma)$.
@@
 For example, in the basic SIR model, $\vec{x}(t) = (\vec{s}(t), \vec{i}(t), \vec{r}(t))^T$.
-Element $a \in \set{A}$ of each vector is the proportion of strata $a$ in that compartment.
+The $a$th element of each vector ($a \in \set{A}$) is the proportion of strata $a$ in that compartment.
 For example, $\vec{s}(t) = (s_1(t), \dots, s_A(t))^T$ where $s_a(t)$ is the proportion of the $a$th strata that are susceptible at time $t$.
@@
 
-The interactions between strata are included by replacing the scalar $\beta$ with a \emph{Who Acquires Infection From Whom} (WAIFW) matrix, $\matr{\beta}$~\autocite[58]{keelingModeling}.
+The interactions between strata are included by replacing the scalar $\beta$ with a WAIFW (\emph{Who Acquires Infection From Whom}) matrix, $\matr{\beta}$~\autocite[58]{keelingModeling}.
 The force of infection is now a vector, defined as $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i_j}(t)$.
@@
 The constant terms here are absorbed into the WAIFW matrix, hence $\beta_{aa'} = \kappa_{aa'} q_{aa'} N_{a'}$~\autocite[section 9.2]{diekmannMathematical}.
-The total force of infection on strata $a$ from all strata is, therefore, $\lambda_a(t) = \sum_{a' \in \set{A}} \beta_{aa'} i_{a'}(t)$.
-Equivalently, $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i_j}(t)$.
+This gives the total force of infection on strata $a$ from all strata is, therefore, $\lambda_a(t) = \sum_{a' \in \set{A}} \beta_{aa'} i_{a'}(t)$ as above.
 If all $\beta_{aa'}$ are equal, then the model is equivalent to the homogeneous model (\ie without any structure).
 The rate at which individuals move between the susceptible and exposed compartments is now $\vec{s}(t) \circ \vec\lambda(t)$, where $\circ$ denotes element-wise multiplication.
+\todo{Is a division by $N_a$ necessary here? Check previous notes on Overleaf including on what Edwin provides.}
 
@@
 It is infeasible because the progression of the epidemic depends on $\vec{\lambda}(t)$, which has dimension $A$, while $\matr{\beta}$ has dimension $A^2$.
-Common assumptions include assuming symmetry, \ie $\beta_{aa'} = \beta_{a'a}$, or that the direction of transmission is irrelevant, \ie $\beta_{aa'} = \beta_{a} \beta_{a'}$.
+One common assumptions is symmetry, \ie $\beta_{aa'} = \beta_{a'a}$.
 % Alternatively, as in the scalar case, it may be easier to iden $\beta_{aa'}$ can be decomposed into the contact rate between strata $a$ and $a'$, $\kappa_{aa'}$ and the probability of transmission upon contact, $q_{aa'}$, and then assumptions made about the components.
@@
 
-Reproduction numbers in models with structured populations need to consider the distribution of infections across the strata.
+Reproduction numbers in models with structured populations are more complex than the homogeneous case.
 The WAIFM allows calculation of the number of secondary infections generated by a single infection in the $a$th strata,  this follows directly from its definition as $\frac{1}{\gamma} \sum_{a' \in \set{A}} \beta_{aa'}$.
@@
 The standard solution to this was proposed by \textcite{diekmannDefinition} (see \textcite[chapter 7]{diekmannMathematical} for a more gentle, but still rigorous, explanation).
-First, they show that the distribution of infections across the strata during the initial period of exponential growth (when $\vec{s} \approx \vec{1}$) relies only on $\matr{\beta}$.
+First, they show that the distribution of infections across the strata during the initial period of exponential growth (when $\vec{s} \approx \vec{1}$) relies only on $\matr{\beta}$, and not the starting conditions.
 This distribution is proportional to the \emph{dominant eigenvector} (the eigenvector corresponding to the largest eigenvalue) of $\matr{\beta}$.
-Denote by $D(\matr{\beta})$ the \emph{dominant eigenvalue} (the largest eigenvalue).
+Denote by $D(\matr{\beta})$ the \emph{dominant eigenvalue} (the largest eigenvalue) of $\matr{\beta}$.
 The basic reproduction number is then defined as $\R = D(\matr{\beta}) / \gamma$.
@@
 
+\todo[inline]{Does something need to link these sections?}
+
 \section{Prevalence process} \label{inc-prev:sec:prevalence-process}
@@
 Denote by $R_{t,t'}$ the number of infection episodes with $B_i=t$ and $E_i=t'$.
-For each infection beginning at $t$, the probability that it ends at $t'$ is $\prob(E_i = t' \mid B_i = t) = \prob(D_i = t' - t + 1 \mid B_i = t) = f_D(t' - t + 1)$.
+For any infection beginning at $t$, the probability that it ends at $t'$ is $\prob(E_i = t' \mid B_i = t) = \prob(D_i = t' - t + 1 \mid B_i = t) = f_D(t' - t + 1)$.
 Since these durations are assumed independent, the number of episodes of each duration is multinomially distributed~\autocite{paganoHIV}.
-Therefore:
+Specifically:
 \begin{align}
@@
 This reduces the stochasticity to only that attributable to the prevalence process.
-In particular, because the duration of individuals is independent, $R_{t_z,t_p}$ and $R_{t_z',t_p'}$ will be independent if $t_z \neq t_z'$; importantly, this holds even if $t_p = t_p'$.
+In particular, because the duration of infection episodes is independent, $R_{t_z,t_p}$ and $R_{t_z',t_p'}$ will be independent if $t_z \neq t_z'$; importantly, this holds even if $t_p = t_p'$.
 Therefore, $P_{t_z,t_p}$ and $P_{t_z',t_p'}$ will be independent if $t_z \neq t_z'$.
@@
 In \cref{inc-prev:sec:observation-process} I will show that only $\E(P_t \mid \vec{Z})$ and bounds on $\V(P_t \mid \vec{Z})$ will be relevant for inferring incidence.
-These can be derived from their constituent parts.
+I know derive these from their constituent parts.
 
@@
 \end{align}
-where $F_D(t) = \sum_{i=1}^t f_D(t) = \prob(D \leq t)$ is the CDF of $D$.
 Therefore:
@@
 
-The first moment follows.
+The first moment of $P_t$ follows.
 \begin{align}
@@
   &= \sum_{i=0}^{d_\text{max}-1} Z_{t-i} (1 - F_D(i)) \\
-  &= \sum_{i=0}^{d_\text{max}-1} Z_{t-i} S(i+1) \label{inc-prev:eq:EPt}
+  &= \sum_{i=0}^{d_\text{max}-1} Z_{t-i} S(i+1) \label{inc-prev:eq:EPt}.
 \end{align}
-where $S(i+1) = 1-F_D(i) = \prob(D \geq i+1)$ is the \emph{survival function} of $D$.
 
@@
 
+\todo[inline]{Relate the following paragarph to the similar discussion in ch 7}
 The expectation in \cref{inc-prev:eq:EPt} is a discrete convolution equation.
-The equation is of the same as the equation used for HIV/AIDS backcalculation, which itself is often found in a variety of applications~\autocite[and references therein]{brookmeyerBackcalculation}.
-The system can be written in the form $\E(\vec{P} \mid \vec{Z}) = \matr{L} \vec{Z}$ where $\matr{L}$ is a lower-triangular matrix.
+The equation is of the same form as the equation used for HIV/AIDS backcalculation, as well as being found in many other fields~\autocite[and references therein]{brookmeyerBackcalculation}.
+The system can be written in the form $\E(\vec{P} \mid \vec{Z}) = \matr{S} \vec{Z}$ where $\matr{S}$ is a lower-triangular matrix formed from $S$.
 Numerical methods for solving systems of this form, that is solving for $\vec{Z}$ (non-random in this expression) for a known $\E (\vec{P} \mid \vec{Z} )$, are well-established~\autocite[e.g.:][section 8.2]{highamAccuracy}.
@@
 The sampling is described mathematically by the observation process.
-The survey design considered here reflects the CIS, although is applicable to many randomly sampled prevalence survey with samples for all $t$.
+The survey design considered here reflects the CIS, although it would be applicable to many prevalence surveys.
 
@@
   &<< \frac{n_t n_{t'} \lvert \cov(P_t, P_{t'}) \rvert}{\Npop^2 \sqrt{\var(P_t) \var(P_{t'})}} &\text{as $\V(X_t) >> \V(P_t)$} \\
-  &= \frac{n_t n_{t'}}{\Npop^2} \lvert \cor(P_t, P_{t'}) \rvert \\
-  &\leq \frac{n_t n_{t'}}{\Npop^2}.
+  &= \frac{n_t n_{t'}}{\Npop^2} \lvert \cor(P_t, P_{t'}) \rvert &\text{by def of $\cor$}\\
+  &\leq \frac{n_t n_{t'}}{\Npop^2} &\text{as $\lvert \cor(\cdot) \rvert \leq 1$}.
 \end{align}
@@
 A common way of constructing a Markov chain with $p(\psi \mid y)$ as its stationary distribution is using the Metropolis-Hastings algorithm~\autocite{hastingsMCMC}.
-The algorithm requires a proposal distribution $q(\psi' \mid \psi)$, which proposes that the chain moves to a new state $\psi'$ from its current state $\psi$, and a calculation of $p(\psi \mid y)$ up to some constant, $\pi(\psi)$.
+The algorithm requires a proposal distribution $q(\psi' \mid \psi)$, which proposes that the chain moves to a new state $\psi'$ from its current state $\psi$; $\pi(\psi)$, a calculation of $p(\psi \mid y)$ up to some constant.
 The proposal is then either accepted or rejected based on the \emph{acceptance ratio}, which is computed using $\pi$.
@@
 The principle here is that if all chains have converged then they are sampling from the same distribution.
-This means that the within-chain and between-chain variances are the same~\autocite[82]{brooksMCMCNotes}.
-In particular, I will make use of the improved Rhat statistic~\autocite{vehtariRhat}.
+Therefore, the within-chain and between-chain variances are the same~\autocite[82]{brooksMCMCNotes}.
+Various metrics quantify thiVarious metrics quantify this I will make use of the improved Rhat statistic~\autocite{vehtariRhat}.
 %The Rhat statistic is defined for a scalar parameter $\psi$.
@@
 
-The stability of estimates considers the error arising because only a finite sample of the distribution of interest is available.
-Furthermore, the sample will tend to be autocorrelated meaning that estimates using the example have higher variance than if it were an iid sample from the posterior~\autocite[286]{gelmanBDA}.
+The stability of estimates considers the error arising from only using a finite sample of the distribution and the samples being autocorrelated.
+Autocorrelation means that estimates using the sample have higher variance than if it were an iid sample from the posterior~\autocite[286]{gelmanBDA}.
 The \emph{effective sample size} corrects for the autocorrelation.
@@
 This chapter has shown that the key quantity relating these quantities is the distribution of the duration of RT-PCR positivity.
-However, no estimates of this quantity in the general population exist.
+However, no estimates of this quantity in the general population exist\todo{ref lit review of duration}.
 Therefore, I now turn to estimating the duration of RT-PCR positivity.

### 2024-02-13 introduction.tex
proof read ATACCC
ch 3 proofing

@@
     \item Use much methodology and concepts from this field
+    \item Surveillance in general (incidence)
+\end{itemize}
+
+\subsection{Surveillance}
+
+\begin{itemize}
+    \item Reproduction numbers
+    \item Incidence: almost never observed directly
 \end{itemize}
@@
     \item Mathematical details in \cref{E-inc-prev}
+    \item Infection process well established, what needs adapting for here (recoveries, sampling)? See first-year report.
 \end{itemize}
@@
 
+Expand on this biomarker vs survival analysis comparison
+\begin{itemize}
+    \item Why do I apply each method?
+    \item Biomarker: stronger assumptions but more information used.
+    \item Literature review on estimates of length of PCR positivity (none in general pop)
+\end{itemize}
+
+\section{Notational conventions}
+
+\begin{itemize}
+    \item Lower-case variables are realisations of the respective upper-case random variables.
+    \item Matrices and vectors are bold and have elements with the same unbolded letter, \eg $\vec{x} = [x_1, x_2, \dots]^T$
+\end{itemize}
+
 \section{Thesis contribution}

### 2024-02-13 symbols.tex
ch 3 proofing

\documentclass[thesis.tex]{subfiles}

\begin{document}

\chapter{Tables of distributions} \label{distributions}

Currently split by chapter.
Ordered Roman letters alphabetically then Greek.
Excluding transmission model for now.

From chapter 3

\begin{itemize}
    \item $B_i$ day infection episode $i$ begins
    \item $D_i = E_i - B_i + 1$ duration of infection episode $i$. $D_i \in \{1, \dots, \dmax \}$
    \item $\dmax$ maximum duration of infection episode
    \item $E_i$ day infection episode $i$ ends
    \item $f_D(d)$ probability density function of $D_i$
    \item $F_D(d)$ cumulative density function of $D_i$
    \item $\Npop$ population size
    \item $n_t$ number of samples on day $t$
    \item $P_t$ prevalent individuals, $P_t / \Npop$ prevalence
    \item $P_{t_z,t_p}$ number of prevalent individuals at time $t_p$ who were infected at time $t_z$
    \item $R_{t,t'}$ number of infection episodes that began at time $t$ and ended at time $t'$
    \item $S(d)$ survival function of $D_i$
    \item $T$ the final day of the observation period used, the first day is day 1
    \item $X_t$ the random variable representing the number of positive tests on day $t$
    \item $\vec{x} = (x_1, \dots, x_T)$ the vector of counts of observed positive tests on each day of the survey
    \item $Z_t$ incidence, $Z_t / \Npop$ incidence proportion
    \item $\lambda(d)$ hazard function of $D_i$
\end{itemize}

\end{document}

### 2024-02-13 transmission.tex
ch 3 proofing

@@
 One approach would be to convolve the incidence based on the transmission model with the RT-PCR duration distribution.
-Here, the daily incidence proportion for each stratum derived by solving the system of ODEs governing the transmission model, and then prevalence calculated using \cref{E-inc-prev:eq:EPt}.
+Here, the daily incidence proportion for each stratum derived by solving the system of ODEs governing the transmission model (see \cref{E-SEIR:sec:SIR}), and then prevalence calculated using \cref{E-inc-prev:eq:EPt}.
 However, this approach is computationally expensive, likely prohibitively so if performed within a MCMC loop.

### 2024-02-13 ATACCC.tex
proofing ch 4

@@
 \Cref{E-inc-prev} showed that the duration of RT-PCR positivity is a key parameter for estimating the incidence of SARS-CoV-2 infections.
-The aim of this chapter is to estimate the duration of RT-PCR positivity using the viral load data from the ATACCC study (described in\todo{ref section describing ATACCC}).
+The aim of this chapter is to estimate the duration of RT-PCR positivity using the viral load data from the ATACCC study (introduced in \cref{E-biology-data:sec:ataccc}).
 I model a biomarker, the log viral load, over the course of an infection.
@@
 The piecewise linear model is the most common in the context of SARS-CoV-2 viral load~\autocite{clearyUsing,kisslerViral,larremoreTest}
+\todo{work out why Kissler citation is displaying weird}
 In this context, the piecewise linear model is applied to the log of the viral load.
@@
     The subscripts are omitted in the key for clarity.
-    The smoothing (lack of linearity) near the peak is visible. \label{ATACCC:fig:viral-load-model}
   }
+  \label{ATACCC:fig:viral-load-model}
 \end{figure}
 
-RT-PCR tests are noisy observations of the viral load (see \cref{E-biology-data:sec:PCR}).
-The noisy observations are modelled by assuming each measurement of log viral load has an iid error distributed $\N(0, \sigma_v^2)$.
+The noisy observations of viral log load from RT-PCR testing are modelled by assuming each measurement of log viral load has an iid error distributed $\N(0, \sigma_v^2)$.
 The measurements are left-censored at the limit of detection, $\vlod$, meaning that any observation which would give a log viral load of less than $\vlod$ is instead observed as negative.
 False positives and negatives are modelled by allowing a test failure with probability $\rho$.
-Test failures are modelled as producing a log viral load reading drawn from $\N(x_0, \sigma_0^2)$ where $x_0$ and $\sigma_0^2$ are parameters to be estimated.
+Test failures are modelled as producing a log viral load reading drawn from $\N(x_0, \sigma_0^2)$.
 These readings are censored in the same way as true readings.
@@
 Biological plausibility requires constraining $\vec{\phi}_i$ to positive real values.
-Each $\vec{\phi}_i$ is modelled as a log-normal random effects by using the hierarchical prior $\phi_i \dist \LN(\vec{\mu}_{\phi,g(i)}, \matr{\Sigma}_\phi)$ where $g(i)$ is the group of individual $i$.
+Each $\vec{\phi}_i$ is modelled as a log-normal random effects by using the hierarchical prior $\vec{\phi}_i \dist \LN(\vec{\mu}_{\phi,g(i)}, \matr{\Sigma}_\phi)$ where $g(i)$ is the group of individual $i$.
 $\vec{\mu}_{\phi,g} = (\mu_{v_{\max},g}, \mu_{a,g}, \mu_{b,g})^T$ is a vector of the population means of the log-parameters.
@@
 \end{align}
-where $\vec{\sigma}_\phi$ is a vector of the standard deviations of the parameters, $\matr{C}_\phi$ is a correlation matrix, and $\circ$ is element-wise multiplication.
+where $\vec{\sigma}_\phi$ is a vector of the standard deviations of the parameters, and $\matr{C}_\phi$ is a correlation matrix.
 These are defined such that, by construction, $(\matr{\Sigma}_\phi)_{ij} = (\vec{\sigma}_\phi)_i (\vec{\sigma}_\phi)_j (\matr{C}_\phi)_{ij}$.
@@
 Specifically, the model excluding false positives explains the data well (the results in \cref{fig:paper:duration,ATACCC:fig:goodness-of-fits} do not allow false negatives but fit the data well).
-Therefore, the model for test failures can be simplified to only false negatives.
+Therefore, I simplified the model for test failures to only false negatives.
 % These false negatives could represent a variety of scenarios, such as poor swabbing technique.
-Formally, the model can be expressed by replacing \cref{ATACCC:eq:hakki-likelihood} with the following.
+Formally, I replaced \cref{ATACCC:eq:hakki-likelihood} with the following.
 \begin{align}
@@
 So far, only the viral load trajectories have been considered; \textcite{hakkiOnset} does not consider duration.
+In this section, I derive an expression for an infection episode's duration, $d_i$, in terms of $\vec{\phi}_i$.
+
 The time that individual $i$ is first and last positive are the two values of $t$ such that $\log v_i(t) = v_\text{lod}$.
@@
 These are the main parameters of interest (as explained in \cref{ATACCC:sec:duration}), suggesting that the results are robust to the choice of model.
-As the estimates are similar and the modified model converges better, I use the modified model for the remainder of this chapter.
+As the estimates are similar and the modified model converges better, I use only the results from the modified model from now on.
 \begin{figure}
@@
 All these factors could change the viral dynamics~\autocite{russellWithinhost,kisslerDynamicsVaccines} and hence the duration.
+\todo{work out why Kissler citation is displaying weird}
 Therefore, the relationships shown here cannot be considered causal, or even generalisable to contexts where these other confounders are not present.
-Despite these confounders, the result is in agreement with previous work showing that the Alpha and Delta variants of concern (compared to pre-Alpha variants) and being vaccinated are all associated with shorter durations of RT-PCR positivity, although the vaccination affect is larger~\autocite{kisslerDynamicsVaccines}.
+Despite these confounders, the result is in agreement with previous work showing that the Alpha and Delta variants of concern (compared to pre-Alpha variants) and being vaccinated are all associated with shorter durations of RT-PCR positivity, although the vaccination effect is larger~\autocite{kisslerDynamicsVaccines}.
 
@@
 
+Individuals 55 and 57 appear to have far noisier measurements than the other individuals (see \cref{ATACCC:fig:goodness-of-fits}), suggesting that the observation noise might vary between individuals.
+Including a random effect on $\sigma_v$ could capture this affect.
+However, it would complicate the model, be unlikely to change the estimates of the parameters of interest, and may not be identifiable.
+
 In the ATACCC study, individuals were tested for at most 20 days.
@@
 The remainder of the estimate for $f_D$, such as that almost all (>99\%) of individuals recover before day 50, is driven by the assumed form of the viral load trajectory.
-
-Individuals 55 and 57 appear to have far noisier measurements than the other individuals (see \cref{ATACCC:fig:goodness-of-fits}), suggesting that the observation noise might vary between individuals.
-Including a random effect on $\sigma_v$ could capture this affect.
-However, it would complicate the model, be unlikely to change the estimates of the parameters of interest, and may not be identifiable.
 

### 2024-02-13 symbols.tex
proofing ch 4

@@
 
+From chapter 4
+\begin{itemize}
+    \item $a_i$ rate of growth of viral load of infection episode $i$
+    \item $b_i$ rate of decline of viral load of infection episode $i$
+    \item $g(i)$ the group of infection episode $i$
+    \item $\vlod$ the limit of detection of the RT-PCR test
+    \item $v_{\max,i}$ maximum viral load of infection episode $i$
+    \item $x_0$ the mean log viral load for test failures
+    \item $y_{i,t}$ the observed log viral load for infection episode $i$ at time $t$
+    \item $\sigma_0$ the standard deviation of the log viral load for test failures
+    \item $\matr{\Sigma}_\phi$ the covariance matrix of the random effects
+    \item $\sigma_v$ standard deviation of the measurement error in viral load
+    \item $\rho$ the probability of a test failure
+    \item $\tau_{0,i}$ and $\tau_{1,i}$ the times of the first and last time infection episode $i$ is detectable
+    \item $\tau_{\max,i}$ time of the maximum viral load of infection episode $i$
+    \item $\vec{\phi}_i = [v_{\max,i}, a_i, b_i]^T$ the vector of random effects for infection episode $i$
+\end{itemize}
+
 \end{document}

### 2024-02-13 ATACCC/goodness_fit.R

@@ -1,10 +1,9 @@
-library(dplyr)
+suppressMessages(library(dplyr))
 library(ggplot2)
 library(purrr)
-library(rstan)
+suppressMessages(library(rstan))
 library(patchwork)
 library(tidybayes)
-library(tidyverse)
 source(here::here("ATACCC/trajectory_functions.R"))
 source(here::here("utils.R"))
 
@@ -91,7 +90,7 @@ p_fits = posterior_summary |>
     aes(day, yrep, ymin = .lower, ymax = .upper),
     show.legend = FALSE
   ) +
-  geom_point(aes(day, y, colour = result), data = tbl_data) +
+  geom_point(aes(day, y, colour = result), size = 0.7, data = tbl_data) +
   facet_wrap(~i, ncol = 6) +
   theme_trajectory()
 ggsave(

### 2024-02-14 transmission.tex
resolve all Overleaf comments

@@
 
-\todo[inline]{Intro probably needs rewriting, with novelty claims and a bit more context. Revisit once intro chapters written.}
+\todo[inline]{%
+    Intro probably needs rewriting, with novelty claims and a bit more context.
+    Revisit once intro chapters written.
+    Question: is CIS (or similar) enough to estimate incidence / transmission?
 
-In this chapter, I use the preferred estimate of the duration of RT-PCR positivity (see \cref{E-imperf-test:sec:conclusion}) to investigate estimation of incidence and transmission of SARS-CoV-2 in England, solely on the basis of .
-CIS data (as shown in \cref{E-biology-data:sec:positivity} for further details).
-I will show that the CIS data is sufficient to derive such estimates, suggesting surveys of this type would be sufficient for surveillance of an epidemic.
+}
 
-I take two approaches, the theoretical advantages of each were discussed in \cref{E-inc-prev:sec:infection-process}.
+In this chapter, I use the estimate of the duration of RT-PCR positivity from combining information from the ATACCC and CIS studies (see \cref{E-imperf-test:sec:conclusion}) to investigate estimation of incidence and transmission of SARS-CoV-2 in England.
+This estimation is based only on the CIS data (as shown in \cref{E-biology-data:sec:positivity} for further details).
+I will show that the CIS data is sufficient; demonstrating the value of surveys of this type for pandemic surveillance.
+
+I take two approaches, backcalculation and a mechanistic model, the theoretical advantages of each were discussed in \cref{E-inc-prev:sec:infection-process}.
+\todo{what are the advantages of each approach specifically in this context?}
+
 First, in \cref{backcalc}, I use a back-calculation approach, based on \cref{E-inc-prev:eq:EPt}.
 This approach enables estimation of incidence within (sub)populations of England.
-Then, in \cref{SEIR}, I use a mechanistic compartmental model, a variation of the model introduced in \cref{E-SEIR:sec:mechanistic-models}, to estimate both incidence and components of transmission, \eg and whether changes in transmission can be explained by changes in contact rates.
+
+In \cref{SEIR}, I use a mechanistic compartmental model, within the modelling framework introduced in \cref{E-SEIR:sec:mechanistic-models}.
+This approach enables estimating both incidence and components of transmission.
+Estimating transmission in addition to incidence answers important epidemiological questions, \eg whether changes in transmission can be explained by changes in contact rates.
 Finally, in \cref{transmission:sec:comparison}, I compare the results of the two approaches.
 
-\todo[inline]{
-    Lit review to go in introduction chapter, framing the whole thesis because it is important in understanding the motivation for the whole work.
-    Would include \textcite{mccabeCISincidence,abbottCISincidence} on backcalculation. Also Taffe?
-    As well as the literature on mechanistic models using REACT and CIS I have previously circulated.
-}
 \section{Data} \label{transmission:sec:data}
@@
 
-During November, incidence starts rising, with regional heterogeneity in the timing of when this occurs (see \cref{transmission:fig:backcalc-regions}).
+During November, incidence starts rising, with regional heterogeneity in the timing of when this occurs.
 The most explored explanation for this is the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocite{walkerTracking,daviesEstimated,lythgoeLineage}.
-There is weak evidence that when incidence starts increasing is correlated with when the Alpha variant's spread (see \cref{transmission:fig:backcalc-alpha}).
 A second peak occurs in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
@@
 \end{figure}
-\begin{figure}
-    \includegraphics{transmission/backcalc-alpha}
-    \caption[Correlation of the emergence of Alpha and minimum incidence.]{%
-        x-axis: time that S-gene target failure, a marker of the Alpha variant, is estimated to start increasing from baseline by region among CIS test results; from \textcite[table S3]{walkerTracking}.
-        y-axis: posterior median and 95\% CrI of the date of minimum incidence in November or early December in each region.
-        Linear regression line shown.
-    }
-\end{figure}
 
 The incidence proportion was highest among the 11--15 and 16--25 age groups throughout the period (see \cref{transmission:fig:backcalc-ages}).
-Except the 16--25 age group, all age groups show similar trends (\ie parallel lines on a log scale), meaning there was insufficient information for the posterior to overcome the prior penalisation towards this situation.
+Except the 16--25 age group, all age groups show similar trends (\ie parallel lines on a log scale), meaning there was insufficient information in the data to overcome the prior which favoured this situation.
 The incidence proportion in the 16--25 age group decreases faster than other age groups during November.
@@
 In this section, I apply the general mechanistic transmission modelling framework from \cref{E-SEIR} to the CIS prevalence data.
-The application of this framework follows \textcite{birrellRealtime}, with the most substantial context-specific work in parameterising the WAIFW matrix (\cref{SEIR:sec:WAIFW-application}).
+The application of this framework follows \textcite{birrellRealtime}, with the most substantial context-specific work\todo{maybe change ``work" to something else: I want a succinct way to say that this was the major part that needed changing from the generic framework in chapter 3} in parameterising the WAIFW matrix (\cref{SEIR:sec:WAIFW-application}).
 A novel approach (\cref{SEIR:sec:observation}) links this transmission model to the CIS prevalence data; the approach is generalisable to other studies collecting representative prevalence data, \eg REACT.
@@
 
+\todo[inline]{Rewrite this intro paragraph}
 This section  explains the novel observation model approach I propose for fitting to the CIS prevalence data.
@@
 One approach would be to convolve the incidence based on the transmission model with the RT-PCR duration distribution.
-Here, the daily incidence proportion for each stratum derived by solving the system of ODEs governing the transmission model (see \cref{E-SEIR:sec:SIR}), and then prevalence calculated using \cref{E-inc-prev:eq:EPt}.
+Here, the daily incidence proportion for each stratum derived by solving the system of ODEs governing the transmission model, and then prevalence calculated using \cref{E-inc-prev:eq:EPt}.
 However, this approach is computationally expensive, likely prohibitively so if performed within a MCMC loop.

### 2024-02-14 biology-data.tex
proofing ch 4
ch 2 now cleared by SRS

@@
 
-\subsection{Assessment of Transmission and Contagiousness of COVID-19 in Contacts}
+\subsection{Assessment of Transmission and Contagiousness of COVID-19 in Contacts} \label{biology-data:sec:ataccc}
 
@@
     Red dots are negative tests, green dots are positive tests, and blue dots are void tests (where the result could not be determined; excluded from other analyses).
-
-    This figure has been provided as management information for operational planning purposes.
-They are provisional estimates circulated before public release.
-This management information should not be shared widely and should only be used for operational planning purposes.
-This information is not to be used publicly and ahead of being released publicly by ONS.
   }
-  \label{biology-data:fig:episodes}
 \end{figure}

### 2024-02-14 transmission-appendix-INLA.tex
add INLA appendix

--- 2024-02-13 symbols.tex
@@
 
-\chapter{Tables of distributions} \label{distributions}
-
-Currently split by chapter.
-Ordered Roman letters alphabetically then Greek.
-Excluding transmission model for now.
-
-From chapter 3
-
-\begin{itemize}
-    \item $B_i$ day infection episode $i$ begins
-    \item $D_i = E_i - B_i + 1$ duration of infection episode $i$. $D_i \in \{1, \dots, \dmax \}$
-    \item $\dmax$ maximum duration of infection episode
-    \item $E_i$ day infection episode $i$ ends
-    \item $f_D(d)$ probability density function of $D_i$
-    \item $F_D(d)$ cumulative density function of $D_i$
-    \item $\Npop$ population size
-    \item $n_t$ number of samples on day $t$
-    \item $P_t$ prevalent individuals, $P_t / \Npop$ prevalence
-    \item $P_{t_z,t_p}$ number of prevalent individuals at time $t_p$ who were infected at time $t_z$
-    \item $R_{t,t'}$ number of infection episodes that began at time $t$ and ended at time $t'$
-    \item $S(d)$ survival function of $D_i$
-    \item $T$ the final day of the observation period used, the first day is day 1
-    \item $X_t$ the random variable representing the number of positive tests on day $t$
-    \item $\vec{x} = (x_1, \dots, x_T)$ the vector of counts of observed positive tests on each day of the survey
-    \item $Z_t$ incidence, $Z_t / \Npop$ incidence proportion
-    \item $\lambda(d)$ hazard function of $D_i$
-\end{itemize}
-
 \end{document}

### 2024-02-14 transmission-appendix-phase-type.tex
add INLA appendix

--- 2023-12-20 SEIR.tex
@@
 \begin{document}
-\ifSubfilesClassLoaded{
-    \setcounter{chapter}{7}
-}
 
-\chapter{Estimating transmission} \label{transmission}
-
-\section{Backcalculation} \label{backcalc}
-
-Now that I have estimated the duration of PCR positivity, I can estimate the incidence of SARS-CoV-2.
-In this chapter, I will apply the theory introduced in \cref{E-inc-prev} to the CIS data.
-I use the period from the expansion of the CIS, starting 31st August 2020, until there was significant vaccination rollout\todo{insert date}.
-First, I explain the method I will apply (\cref{backcalc:sec:methods}) and then present the results (\cref{backcalc:sec:results}).
-I discuss the results in \cref{backcalc:sec:discussion} and conclude in \cref{backcalc:sec:conclustion}.
-
-\todo[inline]{
-    I should discuss the literature on estimating incidence based on SARS-CoV-2 prevalence data.
-    In particular: \textcites{mccabeCISincidence}{abbottCISincidence}.
-    However, I think this goes better in the introduction, framing the whole thesis because it is important in understanding the motivation for the whole work.
-    At the same place, I could review the literature on mechanistic models using REACT and CIS.
-}
-
-\subsection{Methods} \label{backcalc:sec:methods}
-
-As discussed in\todo{ref section}, a major issue with the CIS is non-response bias.
-However, a deterministic relation between true prevalence and incidence is a good approximation (see \cref{E-inc-prev:sec:prevalence-process}).
-I extend the methodology previously used to estimate prevalence from the CIS data, corrected for non-response bias, to estimate incidence\todo{cite Koen's stuff}.
-The previous method was multilevel regression and poststratification (MRP)\todo{cite MRP}.
-I fit a similar model to estimate prevalence, then transform the posterior distribution of prevalence in each stratum to a posterior distribution of incidence.
-Finally, I poststratify the posterior distribution of incidence to the CIS target population and subpopulations of interest.
-
-The advantage of fitting a multilevel regression model on prevalence, rather than incidence, is that prevalence can naturally be expressed as a generalised linear model (GLM).
-A natural formulation for either incidence proportion or prevalence in a stratum $i$ is $\logit(\pi_i) = g(t)$, where $\pi_i$ is the incidence proportion or prevalence in stratum $i$ and $g$ is a function of time, penalised towards being linear.
-This is a GLM with a logit link function.
-The logit transformation serves two purposes.
-First, it ensures that the incidence proportion or prevalence is between 0 and 1.
-Second, at low values of incidence or prevalence, linear growth is approximately exponential which is the natural assumption for an epidemic and has been shown to improve estimates of prevalence\todo{cite REACT paper on this, and check the exact meaning of "improve"}.
-However, if the incidence proportion is expressed as a GLM in this way, the prevalence is not a GLM because a non-linear transformation is required to obtain prevalence from the logit of incidence proportion (the relationship on the natural scale is linear, as derived in \cref{E-inc-prev:sec:prevalence-process}).
-Using a GLM is computationally useful due to the large number of pre-existing packages available for computing the posterior distribution, or its approximation, efficiently.
-Previously, when fitting to prevalence over long periods and including all appropriate covariates to correct for non-response bias, the model was too computationally expensive to fit using MCMC; instead, an implementation of the model using the Integrated Nested Laplace Approximation (INLA) algorithm was used~\citePersonalComms{Koen Pouwels}.
-
-I modified a pre-existing version of a MRP model for CIS~\autocite{pouwelsMRPvaccination}, as recommended by that model's author for this situation~\citePersonalComms{Koen Pouwels}.
-The model included fixed effects for age group\todo{check the model I used}.
-The likelihood is a beta-binomial distribution, a generalisation of the binomial distribution allowing for overdispersion (see \cref{E-distributions}).
-The overdispersion allows for household clustering\todo{ref beta-binomial as a model for clustering} and any other overdispersion in the data\todo{ref that epi data is often overdispersed}. 
-Finally, the model was fit independently for each region, implicitly adding interactions between region and all other variables.
-This step assumes that the estimates for each region can be considered independent, which is reasonable because the regions are large with many tests conducted within each.
-
-I used age groupings:\todo{insert age groupings}.
-These align with previous work on the CIS\todo{cite CIS reports}.
-They are based on school ages (primary, compulsory secondary, and post-compulsory secondary) for children to capture that school opening and closing may have a large effect on transmission.
-
-Having fit the prevalence model, I draw 10,000\todo{check number} prevalence trajectories in each stratum from the approximate posterior distribution of the model.
-For each stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
-Considering the relationship between incidence and prevalence to be deterministic, as \cref{E-inc-prev:sec:prevalence-process} justifies, equation implies that, for any stratum $\vec{P} = \matr{S} \vec{Z}$ where $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum, $\matr{S}$ is a matrix as defined below, and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
-The matrix $\matr{S}$ needs to be defined based on the survival function to preserve the relationship between incidence and prevalence.
-It is lower-triangular of the following form:
-\begin{align}
-    (\matr{S})_{i,j} = \begin{cases}
-        S(i - j + 1) & i - \dmax - 1 \leq j \leq i \\
-        0 & \text{otherwise.}
-    \end{cases}
-\end{align}
-Since the matrix is lower-triangular, the solution for $\vec{P}$ can be found efficiently using forward substitution\todo{cite forward substitution}.
-Forward substitution is numerically stable unless the diagonal elements are close to 0.
-Here, the diagonal is $S(1) = 1$ and hence the matrix is well-conditioned.
-\todo{I explained this more clearly in my first-year report, look that up!}
-
-The final step is poststratifying the incidence proportion to the CIS target population and subpopulations of interest.
-For each posterior draw and day\todo{introduce discrete time in days earlier}, the incidence proportion in the (sub)population is the weighted average of the incidence proportion in each stratum, where the weights are the proportion of the population in each stratum.
-The population numbers for each stratum are provided by the Office for National Statistics.
-
-\subsection{Results} \label{backcalc:sec:results}
-
-\subsection{Discussion} \label{backcalc:sec:discussion}
-
-\section{Mechanistic modelling}
-
-{\color{red} After introducing the observational model in generic terms above (section 8.2) I would see a reorganisation of the section 8.3 as follows:   
-\begin{itemize}
-    \item 8.3 (currently  titled Implementation) CIS-based transmission model (see comments on lines 484-486) 
-\item 8.3.1 WAIFW matrix
-\item 8.3.2 Full model (perhaps another title..model structure?) 
-\item 8.3.3 Observational model (currently section 8.2) include here the "Solving the system of ODEs" (currently in section 8.3.4)
-\item 8.3.4 Priors 
-\item 8.3.5 Inference 
-\item 8.4 Simulation study
-\end{itemize}
-}
-
-\todo[inline]{Redo this intro, with some high-level comments taken from the full model section (eg following Birrell et al.)}
-
-\subsection{Methods}
-
-I use the same CIS data as in \cref{E-backcalc}, again considering each region independently.
-The model will be stratified only by region and age group, so the data are aggregated to these levels.
-Only individuals aged two or older are eligible to participate in the CIS; I assume that test results in all participants aged under 11 are still representative of the whole $[0, 11)$ age group.
-
-As well as the motivation given in \cref{E-backcalc}, independent regions allows heterogeneity in the parameters between regions, which could arise from a range of socioeconomic factors.
-Furthermore, the assumption of homogeneous mixing within age groups is more plausible for regions than the whole of England.
-By modelling each region independently, I assume that the epidemics in each region do not interact.
-Clearly, this is not realistic as travel between regions of England is common and unrestricted.
-However, for a large epidemic, the non-interaction is a good approximation~\autocite{birrellRealtimea}.
-
-\subsubsection{WAIFW matrix}
-
-A parsimonious model is needed to parameterise the WAIFW matrix (see \cref{SEIR:sec:structured-populations}).
-I use the following age-based strata, which align closely with the age groupings previously used for analysis of the CIS data~\autocites[e.g.:][]{houseInferring}{walkerTracking}.
-\todo{Contrast with age groupings used in previous chapter}
-\begin{enumerate}
-    \item $[0, 11)$: children of pre-school and primary school age.
-    \item $[11, 16)$: children of compulsory secondary-school age.
-    \item $[16, 25)$: adolescents and young adults who may remain in formal education.
-    \item $[25, 50)$: younger working age adults.
-    \item $[50, 70)$: older working age adults.
-    \item $[70, \infty)$: elderly.
-\end{enumerate}
-
-I follow the WAIFW matrix parameterisation of \textcite{birrellRealtime}, with the decomposition $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'} N_{a'}$.
-This separates contact patterns, $\kappa_{aa'}$, which can vary over time; transmission probabilities $q_{aa'}$, assumed constant; a region-wide effect for time $\beta(t)$; and the population of the stratum, $N_{a'}$.
-
-This parameterisation takes information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using pre-existing methodology~\autocites{vanleeuwenTime}{vanleeuwenAugmenting}, which exploits
-%; these estimates have been used for previous modelling of the COVID-19 pandemic~\autocites{vanleeuwenTime}{birrellRealtime}.
-% These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
-pre-pandemic baseline estimates from the POLYMOD study~\autocite{mossongSocial}, and time use, from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
-POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on the contacts' age and location (home, work, school, leisure, transport, or other).
-UKTUS asked a representative sample of the UK population about how much time they spent doing various activities each day.
-\Textcite{vanleeuwenAugmenting} combined these two datasets to estimate the rate of contacts between age groups during each activity.
-\Textcite{birrellRealtime} then modify these contact rates based on measured changes in time use during the pandemic.
-These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
-Google mobility data provides the location of Android users over time, relative to a pre-pandemic baseline.
-These data sources are then used to down or up-weight the number of contacts in each setting.
-Resulting estimates of contacts by setting are then recombined across settings to form $\kappa_{aa'}$
-For further details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.
-
-Some behavioural changes (\eg mask wearing) impact transmission.
-However, these are not captured by the contact survey.
-To allow for these changes as well as uncertainty produced by the procedure above, I follow \textcite{birrellRealtime} in placing a random walk prior on $\log\beta(t)$  on the weekly timescale:
-\begin{align}
-    \beta(t) &= \beta_{\tilde{w}(t)} \\
-    \log\beta_{w+1} &= \log\beta_w + \sigma_\epsilon \epsilon_w \\
-    \epsilon_w &\sim \N(0, 1)
-\end{align}
-where $\tilde{w}(t)$ is the week number of time $t$, with week 1 being the first week considered.
-Two hyperparameters will be inferred: $\beta_1$, the initial value of the random walk; and $\sigma_\epsilon$, the standard deviation of the random walk, will have priors placed on them (see \cref{SEIR:sec:priors}).
-
-% $K_{ij}$ and $\beta_{ij}$ appear only as a product and therefore are not individually identifiable.
-% Normally, $K_{ij}$ is assumed to be known, for example from a previous contact survey~\autocite[such as][]{mossongSocial}.
-% Ideally, all the $\beta_{ij}$ terms would be estimated, but there is normally inadequate data to do so, requiring assumptions to be made to reduce the number of parameters~\autocite[65]{keelingModeling}.
-% % $\beta_{ij}$ is then estimated, although often with a parsimonious model
-% A common assumption is that $\beta_{ij}$ can be decomposed into two independent components such that $\beta_{ij} = \beta \beta^\text{susc}_{i} \beta^\text{inf}_j$.
-% For any stratum $i$, $\beta^\text{susc}_i$ is known as the \emph{relative susceptibility} of stratum $i$.
-
-Also $q_{aa'}$ remains to be specified.
-I again follow the approach of \textcite{birrellRealtime}, decomposing it into two independent components such that $q_{aa'} = q^\text{susc}_{a} q^\text{inf}_{a'}$, where
-$q^\text{susc}_a$ is the \emph{susceptibility} of stratum $a$, \ie how vulnerable to infection individuals in stratum $a$ are; and
-$q^\text{inf}_{a'}$ is the \emph{transmissibility} of stratum $a'$, \ie how likely individuals in stratum $a$ are to pass on an infection.
-% Then $\beta^\text{susc}_i$ is the relative risk of an individual in stratum $i$ being infected, relative to an individual in stratum $a_0$.
-These measures include both biological differences, \eg due to differences in immune systems, or behavioural, \eg if infants are less hygienic than adults.
-
-%Now, we assume that $\beta_{ij} = \beta(t) \beta_{ij}$ where $\beta(t)$ is a non-strata-specific, time-varying component representing precautions taken by the population, for example masking; and $\beta_{ij}$ represents a constant probability of transmission from an infectious individual in strata $j$ to a susceptible individual in strata $i$ given contact, which could, for instance, be affected by differential susceptibility to infection for different age-groups.
-%We arbitrarily choose one $\beta_{ij}$ to be equal to 1 for identifiability and hence the other $\beta_{ij}$ values are relative to this reference class.
-
-Previous studies for pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults~\autocites{chenRole}{vinerTransmission}.
-Therefore, I set $q^\text{inf}_{a'} = 1$ for all $a'$, $q^\text{susc}_{a} = q_c$ for the youngest stratum, and $q^\text{susc}_{a} = 1$ for the remaining strata.
-$q_c$ is then estimated, although using an informative prior (see \cref{SEIR:sec:priors}).
-
-\subsubsection{Prevalence and observation models} \label{SEIR:sec:observation}
-
-The transmission model represents the infection process (introduced in \cref{E-inc-prev:sec:infection-process}.
-A variety of different approaches can be taken to link a transmission model to observations.
-In the context of CIS, this requires representing the prevalence and observation processes (introduced in \cref{E-inc-prev:sec:prevalence-process,E-inc-prev:sec:observation-process} respectively)
-The method to do so is known as the observation model.
-The formulation of this model is heavily dependent on the data available.
-This means that there is a wide variety of models used.
-Therefore, this section  explains the model I propose for fitting to the CIS prevalence data.
-
-% Here, I extend the system of ODEs to include compartments representing PCR positive individuals.
-% The proportion of PCR positive individuals in each strata is then linked to the CIS using a Beta-Binomial likelihood, with the modelled proportion that are PCR positive as the mean of the Beta distribution.
-One approach would be to convolve the incidence based on the transmission model with the PCR duration distribution.
-Here incidence is calculated from the ODE solution and then related to the data as explained in \cref{E-inc-prev}.
-However, this approach is computationally expensive because of the need to calculate the convolution in \cref{E-inc-prev:eq:EPt }.
-
-A much more efficient option is to approximate this convolution by adding compartments to the ODEs.
-This approach also allows the infected, pre-positive compartment, which I denote $p_0$, to be represented.
-$p_0$ increases at the same rate as individuals flow into $e_1$.
-These individuals then transition into a set of compartment with their transition rates chosen to match the duration of positivity estimates in \cref{E-imperf-test}, $p_1, \dots, p_l$.
-These compartments do not have a biological interpretation, unlike the other compartments in the system, but are purely a mathematical tool so that the time spent in these compartments matches the duration distribution.
-This is the same idea behind having multiple exposed or latent compartments (introduced in \cref{SEIR:sec:non-exponential}).
+\chapter{Creating RT-PCR positive states} \label{transmission:sec:phase-type}
 
@@
 % A $l$-phase EC distribution is the sum of a $(l-2)$-phase gamma distribution and a two-phase acyclic, arbitrary phase-type distribution.
-An EC distribution $P$ can be defined as the sum of three random variables: $K$, $X_1$, and $X_2$.
+An $l$-phase EC distribution $P$ can be defined as the sum of three random variables: $K$, $X_1$, and $X_2$.
 It is shown schematically in \cref{SEIR:fig:EC}.
-Here, $K \dist \GamDist(l-2, \kappa_Y)$, $X_1 \dist \Exponential(\kappa_{X1})$ and $X_2$ is a mixture distribution taking the value 0 with probability $1-p_x$ or otherwise (\ie with probability $p_x$) distributed $\Exponential(\kappa_{X2})$.
+Here, $K \dist \GamDist(l-2, \omega_Y)$, $X_1 \dist \Exponential(\omega_{X1})$ and $X_2$ is a mixture distribution taking the value 0 with probability $1-\pi$ or otherwise (\ie with probability $\pi$) distributed $\Exponential(\omega_{X2})$.
 % The time between entering the 1st state and the $l+1$th state, the absorbing state, is the duration.
+These distributions contain five parameters.
+Three are positive reals, $\omega_Y$, $\omega_{X1}$, and $\omega_{X2}$; one is a probability, $\pi$, in the interval $[0, 1]$; and one is a positive integer, $l$.
+
+The name Erlang-Coxian comes from the two distributions that make up the EC distribution.
 Gamma distributions with integer first parameters (as $K$ has) are also known as Erlang distributions.
@@
 
-    \path (1) edge node {$\kappa_Y$} (2)
-          (2) edge node {$\kappa_Y$} (dots)
-          (dots) edge node {$\kappa_Y$} (lm2)
-          (lm2) edge node {$\kappa_Y$} (lm1)
-          (lm1) edge node {$p_x \kappa_{X1}$} (l)
-                edge [out=-45,in=-135] node[below] {$(1 - p_x) \kappa_{X1}$} (absorbing)
-          (l) edge node {$\kappa_{X2}$} (absorbing);
+    \path (1) edge node {$\omega_Y$} (2)
+          (2) edge node {$\omega_Y$} (dots)
+          (dots) edge node {$\omega_Y$} (lm2)
+          (lm2) edge node {$\omega_Y$} (lm1)
+          (lm1) edge node {$\pi \omega_{X1}$} (l)
+                edge [out=-45,in=-135] node[below] {$(1 - \pi) \omega_{X1}$} (absorbing)
+          (l) edge node {$\omega_{X2}$} (absorbing);
 \end{tikzpicture}
 }
-\caption[A $l$ phase EC distribution.]{Representation of an EC distribution as a series of compartments.
-The distribution is the time from entering state 1 and entering the absorbing state.
-The time between entering state 1 and state $l-2$ is distributed $K \dist \GamDist(l-2, \kappa_Y)$, also known as an Erlang distribution.
-The time between entering state $l-2$ and the absorbing state has a Coxian distribution (see the main text for details).}
+\caption[A $l$ phase EC distribution.]{Representation of an EC distribution as a progression through a series of compartments.
+The distribution is the time from entering state 1 and entering the absorbing state.}
 \label{SEIR:fig:EC}
@@
 
-\Textcite{osogamiClosed} define $P$ as well-matching an arbitrary distribution $G$ if $P$ and $G$ have the same first three moments.
-They then prove that EC distributions have the following properties:
+\Textcite{osogamiClosed} define a phase-type distribution $P$ as \emph{well representing} an arbitrary distribution $G$ if $P$ and $G$ have the same first three moments.
+They then prove that EC distributions have the following properties.
 \begin{itemize}
-    \item If there exists a phase-type distribution $P'$ that well-matches $G$ then there exists an EC distribution $P$ that well-matches $G$.
-    \item The parameters of $P$ are available in closed form, and the expression for these is derived.
-    \item The number of phases used by the $P$ calculated by these expressions is at most one more than the minimum across all phase-type distributions that well-match $G$.
+    \item Let $\set{W}$ be the set of all phase-types distributions that well represent $G$.
+        Let $\set{W}' \subseteq \set{W}$ be the set of EC distributions that well represent $G$.
+        If $\set{W}$ is non-empty (which is true for most real-life distributions), then $\set{W'}$ is also non-empty; the following properties assume that $\set{W}$ is non-empty.
+    \item Let $w$ and $w'$ be the minimum number of phases of any member of $\set{W}$ and $\set{W'}$ respectively.
+        Then $w' = w$ or $w' = w + 1$.
+        That is, the minimum number of phases for an EC distribution to well represent $G$ is at most one more than the minimum number of phases for any phase-type distribution to well represent $G$.
+        Fewer phases is generally beneficial as it reduces the time and space complexity of working with the distribution.
+    \item A $l$-phase EC distribution $P' \in \set{W'}$ with $l \leq w + 1$ can be found using a closed-form expression (given by \textcite{osogamiClosed}).
+        This is beneficial because finding arbitrary members of $\set{W}$ is challenging.
 \end{itemize}
@@
     \item $l$, the number of phases.
-    \item $\kappa_Y$, the rate parameter of the Erlang distribution.
-    \item $\kappa_{X1}$, the rate parameter of the first state after the Erlang distribution.
-    \item $\kappa_{X2}$, the rate parameter of the second state after the Erlang distribution.
-    \item $p_x$, the probability of moving from the end state of the Erlang distribution to the second (as opposed to directly to the absorbing state).
+    \item $\omega_Y$, the rate parameter of the Erlang distribution.
+    \item $\omega_{X1}$, the rate parameter of the first state after the Erlang distribution.
+    \item $\omega_{X2}$, the rate parameter of the second state after the Erlang distribution.
+    \item $\pi$, the probability of moving from the end state of the Erlang distribution to the second (as opposed to directly to the absorbing state).
 \end{itemize}
 
-I use the mapfit package~\autocite{mapfit} to compute $P$ for the posterior mean of the distribution estimated in \cref{E-imperf-test}.
-This is the same duration distribution used in \cref{E-backcalc}.
-The parameter estimates are in \cref{SEIR:table:ec-params}.
-\begin{table}
-    \centering
-    \begin{tabular}{c c c c c}
-        $l$ & $\kappa_Y$ & $\kappa_{X1}$ & $\kappa_{X2}$ & $p_x$ \\
-        3 & 0.0820 & 0.126 & 0.0223 & 0.00973  \\
-    \end{tabular}
-    \caption{Parameter estimates for the EC distribution approximating the posterior mean of the duration distribution estimated in \cref{E-imperf-test}.}
-    \label{SEIR:table:ec-params}
-\end{table}
-% end table of ec params
-
-The modelled proportion of individuals that are PCR positive in the can then be linked to the data using a Beta-Binomial likelihood (see\todo{ref section on clustering} for the issues with using a binomial likelihood).
-If on each day $t$, each strata $i$ is observed to have $y_{it}$ positives out of $n_{it}$ tests, then the likelihood is:
-\begin{align}
-    \prod_{a,t} \BB (y_{at} \mid n_{at}, \sum_{k=1}^l \bar{p}_{kat}, \rho)
-    \label{SEIR:eq:likelihood}
-\end{align}
-where $\bar{p}_{kat} = \int_{t}^{t+1} p_{ka}(s) ds$ is the mean occupancy in compartment $p_k$ and strata $a$ over day $t$ and $\rho$ is a nuisance parameter controlling the overdispersion of the beta-binomial distribution.
-In this parameterisation (detailed in \cref{E-distributions}) at $\rho=0$ the beta-binomial coincides with the binomial distribution and as $\rho \to \infty$, the distribution tends towards the uniform distribution on $[0, 1]$~\autocite{hughesUsing}.
-
-\subsubsection{Model equations and solution} \label{SEIR:sec:full-model}
-
-\begin{figure}
-\makebox[\textwidth][c]{
-\begin{tikzpicture}[
-    node distance = 2.5cm,
-    on grid,
-    auto,
-    ->,>=stealth',
-    every state/.style={draw,rectangle},
-    ]
-
-    \node[state] (S) {$s$};
-    \node[state, right=of S] (E1) {$e_1$};
-    \node[state, right=of E1] (E2) {$e_2$};
-    \node[state, right=of E2] (I1) {$i_1$};
-    \node[state, right=of I1] (I2) {$i_2$};
-    \node[state, right=of I2] (R) {$r$};
-
-    \path (S) edge node {$\lambda$} (E1)
-          (E1) edge node {$\sigma$} (E2)
-          (E2) edge node {$\sigma$} (I1)
-          (I1) edge node {$\gamma$} (I2)
-          (I2) edge node {$\gamma$} (R);
-
-    \node[state, below=3cm of E1] (P0) {$p_0$};
-    \node[state, right=of P0] (P1) {$p_1$};
-    \node[state, right=of P1] (P2) {$p_2$};
-    \node[state, right=of P2] (P3) {$p_3$};
-    \node[state, right=of P3, draw=none] (P4) {};
-
-    \path  (S) edge node {$\lambda$} (P0)
-          (P0) edge node {$\kappa_0$} (P1)
-          (P1) edge node {$\kappa_Y$} (P2)
-          (P2) edge node {$p_x \kappa_{X1}$} (P3)
-               edge [out=-45,in=-135] node[below] {$(1 - p_x) \kappa_{X1}$} (P4)
-          (P3) edge node {$\kappa_{X2}$} (P4);
-
-    \node[draw, thick, inner sep=0.3cm, fit=(P1) (P3), fill=blue!10,opacity=0.2] {};
-\end{tikzpicture}
-}
-  \caption[SEIR model]{SEIR model used for the analysis in this chapter. The shaded box represents PCR-positive individuals. Shown for a single stratum.}
-  \label{SEIR:fig:full-model}
-\end{figure}
-
-The full model combines the elements discussed in this chapter up until now.
-The transmission model follows \textcite{birrellRealtime}.
-It is a SEIR model with two latent and two infectious states.
-The population is stratified by age, using a time-varying matrix $\matr{\beta}(t)$ to describe the transmission between strata and how these change over time.
-
-To form whole of England estimates, each region is modelled independently with the results aggregated, as in \cref{E-backcalc}.
-
-The observation model is novel, relating the latent transmission described by the model to the CIS data.
-It uses the phase-type model to represent the duration distribution of PCR positivity explained in \cref{SEIR:sec:observation}.
-
-A diagram of the model for a single stratum is shown in figure \cref{SEIR:fig:full-model}.
-The system of differential equations for each region is below.
-\begin{align}
-    \label{SEIR:eq:fullODEs}
-    \vec{\lambda}(t) &= \beta(t) (\matr{\kappa}(t) \circ \matr{q}) (\vec{i_1}(t) + \vec{i_2}(t)) \\
-    \frac{d\vec{s}(t)}{dt} &= -\vec{\lambda}(t) \circ \vec{s}(t) \\
-    \frac{d\vec{e_1}(t)}{dt} &= \vec{\lambda}(t) \circ \vec{s}(t) - 2\sigma \vec{e_1}(t) \\
-    \frac{d\vec{e_2}(t)}{dt} &= 2\sigma \vec{e_1}(t) - 2\sigma \vec{e_2}(t) \\
-    \frac{d\vec{i_1}(t)}{dt} &= 2\sigma \vec{e_2}(t) - 2\gamma \vec{i_1}(t) \\
-    \frac{d\vec{i_2}(t)}{dt} &= 2\gamma \vec{i_1}(t) - 2\gamma \vec{i_2}(t) \\
-    \frac{d\vec{r}(t)}{dt} &= 2\gamma \vec{i_2}(t) \\
-    \frac{d\vec{p_0}(t)}{dt} &= \vec{\lambda} \circ \vec{s}(t) - \kappa_0 \vec{p_0}(t) \\
-    \frac{d\vec{p_1}(t)}{dt} &= \kappa_0 \vec{p_0}(t) - \kappa_Y \vec{p_1}(t) \\
-    \frac{d\vec{p_2}(t)}{dt} &= \kappa_Y \vec{p_1}(t) - \kappa_{X1} \vec{p_2}(t) \\
-    \frac{d\vec{p_3}(t)}{dt} &= p_x \kappa_{X1} \vec{p_2}(t) - \kappa_{X2} \vec{p_3}(t).
-\end{align}
-
-\subsubsection{Solving the systems of ODEs} \label{SEIR:sec:solving-ODEs}
-I solve the system of ODEs using Euler's method, discretised at timesteps of $\delta t = 1/2$ days.
-That is, if $\vec{x}_t$ is the system state at time $t$ for $t \in \{ 0, 0.5, 1, \dots \}$ then $x_{t+\frac{1}{2}} = x_t + (d\vec{x}(t)/dt) \delta t$
-
-The mean prevalence on day $t$, used for the likelihood (see \cref{SEIR:eq:likelihood}), can be calculated as $\bar{p}_{kat} = \frac{1}{2} ( p_{kat} + p_{ka,t+1/2} )$.
-
-\subsubsection{Priors} \label{SEIR:sec:priors}
-
-Priors for the parameters are given in \cref{SEIR:table:priors}.
-The parameters governing the duration of positivity ($\kappa$s) were explained previously (see \cref{SEIR:table:ec-params}).
-The initial conditions and transition rate parameters are unindentifiable in a SEIR model~\autocite{dankwaStructural}, therefore I used assumed values or strongly informative priors on them.
-$\sigma_\epsilon$ and $\rho$ are given priors that penalise model complexity.
-The prior on $\sigma_\epsilon$ is a penalised complexity prior, favouring a null model of constant values for $\beta(t)$~\autocite{simpsonPenalising}.
-The prior on $\rho$ favours a null model of a binomial likelihood.
-The remaining priors had weakly informative priors.
-
-\begin{landscape}
-\begin{table}
-\begin{tabular}{l c l l}
-    Parameter description & Symbol & Distribution & Comment \\
-    \hline \\
-    Mean latent period & $1/\sigma$ & 3.5 days (fixed) & Based on \textcite{zhaoEstimating} \\
-    Mean infectious period & $1/\gamma$ & 4 days (fixed) & Based on \textcite{zhaoEstimating} \\
-    Initial proportion without immunity & $\vec\pi$ & See \cref{SEIR:table:immunity-prior} & \\
-    Initial proportion infectious & $i^+$ & $\BetaDist(0.5, 1000)$ & Weakly informative \\
-    Susceptibility of children & $q_c$ & $\LN(-0.4325, 0.1174)$ & Based on \textcite{vinerTransmission}  \\
-    Initial growth rate & $\psi$ & $\N(0.06, 0.04)$ & Weakly informative \\
-    Standard deviation of random walk steps on log scale & $\sigma_\epsilon$ & $\Exponential(80)$ & Penalising towards constant \\
-    Overdispersion of observations & $\rho$ & $\Exponential(2 \times 10^5)$ & Penalising towards binomial
-\end{tabular}
-\caption[SEIR model priors]{Priors for each parameter. For details of the distributions and their parameterisations see \cref{E-distributions}.}
-\label{SEIR:table:priors}
-\end{table}
-
-\begin{table}
-\centering
-\begin{tabular}{l|lllll}
-         & $[0,16)$ & $[16,25)$ & $[25,50)$ & $[50,70)$ & $[70,\infty)$ \\
-        \hline
-        East Midlands & $\BetaDist(8.5, 301)$ & $\BetaDist(18, 256)$ & $\BetaDist(19, 475)$ & $\BetaDist(20, 493)$ & $\BetaDist(5, 337)$ \\
-        East of England & $\BetaDist(7.9, 236)$ & $\BetaDist(21, 248)$ & $\BetaDist(35, 677)$ & $\BetaDist(26, 613)$ & $\BetaDist(5.6, 332)$ \\
-        London & $\BetaDist(9.7, 131)$ & $\BetaDist(50, 328)$ & $\BetaDist(154, 1331)$ & $\BetaDist(106, 990)$ & $\BetaDist(7.6, 204)$ \\
-        North East & $\BetaDist(6.1, 180)$ & $\BetaDist(12, 145)$ & $\BetaDist(14, 274)$ & $\BetaDist(14, 295)$ & $\BetaDist(4.2, 240)$ \\
-        North West & $\BetaDist(12, 252)$ & $\BetaDist(38, 353)$ & $\BetaDist(72, 885)$ & $\BetaDist(44, 695)$ & $\BetaDist(6.3, 264)$ \\
-        South East & $\BetaDist(8.9, 404)$ & $\BetaDist(17, 319)$ & $\BetaDist(20, 577)$ & $\BetaDist(19, 586)$ & $\BetaDist(4.4, 356)$ \\
-        South West & $\BetaDist(6.9, 415)$ & $\BetaDist(11, 277)$ & $\BetaDist(12, 556)$ & $\BetaDist(11, 559)$ & $\BetaDist(4, 492)$ \\
-        West Midlands & $\BetaDist(8, 167)$ & $\BetaDist(19, 151)$ & $\BetaDist(32, 475)$ & $\BetaDist(33, 504)$ & $\BetaDist(5.8, 241)$ \\
-        Yorkshire and The Humber & $\BetaDist(9.2, 314)$ & $\BetaDist(19, 242)$ & $\BetaDist(27, 589)$ & $\BetaDist(18, 474)$ & $\BetaDist(4.9, 316)$ \\
-    \end{tabular}
-\caption{Priors for $\vec{\pi}$, the proportion of the population without prior immunity, in each age group and region combination.
-Based on seroprevalence in blood donors~\autocite{amirthalingamSeroprevalence} and a representative sample of children~\autocite{ratcliffeCommunity} in August 2020.}
-\label{SEIR:table:immunity-prior}
-\end{table}
-\end{landscape}
-
-\subsubsection{Inference} \label{SEIR:sec:MCMC}
-
-I use an adaptive random-walk Metropolis-Hastings algorithm (previously discussed in \cref{E-inc-prev:sec:MCMC}).
-The proposal distribution is a multivariate normal that adapts to the covariance of the posterior.
-The adaption algorithm is based on \textcite[algorithm 4]{andrieuTutorial}, with the implementation based on \textcite{ghoshApproximate}'s work.
-Proposals are accepted or rejected using the standard Metropolis-Hastings (MH) acceptance probability.
-See \cref{SEIR:MCMC-algorithm} for full details.
-\begin{algorithm}
- set $\vec{X_0}$ to an initial value of the parameter vector \;
- $\vec{\mu_0} = \vec{X_0}$ \;
- $\matr{\Sigma_0} = \text{diag}(\vec{\mu_0})$ \;
- $\lambda_0 = 1$ \;
- \For{$i = 1, \dots, M$}{
-  sample $\vec{Y_}i \sim \N(\vec{\mu_{i-1}}, \lambda_{i-1}\matr{\Sigma_{i-1}})$\;
-  set $\vec{X_i}$ to $\vec{Y_i}$ or $\vec{X_{i-1}}$ using a MH acceptance step\;
-  update the proportion of proposals accepted so far, $\alpha$ \;
-  \eIf(\tcc{No adaptation for 200 iterations}){$i \leq 200$}{
-    $\lambda_i = \lambda_{i-1}$ \;
-    $\vec{\mu_i} = \vec{\mu_{i-1}}$ \;
-    $\matr{\Sigma_i} = \matr{\Sigma_{i-1}}$ \;
-   }{
-    $\gamma_i = (i - 200)^{-0.6}$ \;
-    $\log \lambda_i = \log \lambda_{i-1} + \gamma_i(\alpha - 0.234)$ \;
-    $\vec{\mu_i} = (1 - \gamma_i) \vec{\mu_{i-1}} + \gamma_i \vec{X_i}$ \;
-    $\matr{\Sigma_i} = (1 - \gamma_i) \matr{\Sigma_{i-1}} + \gamma_i (\vec{X_i} - \vec{\mu_i})(\vec{X_i} - \vec{\mu_i})^T$ \;
-  }
- }
- \caption{Algorithm for adaptive random-walk Metropolis-Hastings.
-$\vec{\mu_i}$ and $\matr{\Sigma_i}$ are an estimate of the mean and covariance of the posterior distribution using information up to iteration $i$.
-$\text{diag}(\vec{\mu_0})$ is the diagonal matrix with diagonal entries equal to $\vec{\mu_0}$.
-$\lambda_i$ is the scale parameter of the proposal distribution at iteration $i$, tuned to try and ensure an optimal proportion of proposals are accepted (23.4\%).
-$\gamma_i$ is the learning rate, which determines how much adaptation occurs.
-$\gamma_i \to 0$ as $i \to \infty$ so the rate of adaptation is \emph{vanishing}.
-Vanishing adaptation is required for the algorithm to converge to the target distribution~\autocite[section 3]{andrieuTutorial}.}
- \label{SEIR:MCMC-algorithm}
-\end{algorithm}
-
-I transformed parameters with restricted ranges for sampling, to improve efficiency.
-Parameters which are constrained to the positive reals ($\sigma_\epsilon$, $\rho$, and $q_c$) were sampled on the log scale.
-Parameters which are constrained to the unit interval ($i^+$ and $\vec{\pi}$) were sampled on the logit scale.
-
-I run 4 chains for 1,000,000 iterations each.
-The first 500,000 were discarded as burn-in, with the remaining chains thinned to every 100th iteration.
-This gives 5,000 posterior samples per region.
-This was further thinned to 2,000 samples when calculating posterior predictive quantities (incidence and prevalence) due to their computational cost.
-
-Following \textcite{birrellBayesian}, I reparameterise the model to improve identifiability.
-Rather than directly inferring $\beta_1$, I infer the initial growth rate of the epidemic.
-In addition, I use a parsimonious parameterisation of the initial conditions using the initial proportion of the population that are infectious and the initial proportion that are susceptible only.
-Both of these rely on an assumption that the epidemic is in a steady-state exponential growth phase at time 0 (see \cref{SEIR:sec:SIR}).
-
-During such an exponential growth phase, all components of the $\vec{e}$, $\vec{i}$ and $\vec{p}$ compartments grow at the same growth rate, $\psi$.
-This follows directly from the ODEs where $\vec{s}$ is constant. 
-The growth rate $\psi$ is easily identifiable from the data, and therefore a parameterisation in terms of $\phi$ should be more computationally efficient.
-If this is the initial exponential phase, when $\vec{s} \approx 1$, then the basic reproduction number for a SEIR model is linked to this growth rate by~\autocites{birrellBayesian}{wearingAppropriate}:
-\begin{align}
-    \R = \frac{\psi \left( \frac{\psi}{2\sigma} + 1 \right)^2}{\gamma \left(1 - \frac{1}{\left(\frac{\psi}{2 \gamma} + 1 \right)^2} \right)} \label{SEIR:eq:rtoR}.
-\end{align}
-
-Furthermore, $\R$ is a function of $\matr{\beta}$, and hence $\beta_1$ (as explained in \cref{SEIR:sec:structured-populations}).
-Combining these expressions, the model can be parameterised by using $\psi$ in place of $\beta_1$.
-The following procedure gives $\beta_1$ in terms of $\psi$.
-First, calculate $\R$ from $\psi$ using \cref{SEIR:eq:rtoR}; denote this $\hat{R}$.
-Then calculate $R^* = D(\matr{\beta}) / \gamma$ when $\beta_1=1$ (recall that $D(\matr{\beta})$ denotes the dominant eigenvalue of $\matr{\beta}$).
-Multiplying a matrix by a scalar multiplies its dominant eigenvalue by that same scalar, therefore, setting $\beta_1 = \hat{R} / R^*$ gives the desired means that $\R$ at time 0 is $\hat{R}$, as desired.
-
-To form a parsimonious parameterisation of the initial state, first focus on the initial proportion of the population that is in the infectious compartments.
-Recall that in the exponential growth phase, the relative proportion of each strata that is infectious is proportional to the dominant eigenvector of $\matr{\beta}$, which I denote $\vec{d}$.
-Introduce a model parameter $i^+$, the proportion of the population in the $i$ states at time 0 in the strata with the highest proportion infectious.
-Then, $\vec{i_0} = \vec{i_1} + \vec{i_2} = \frac{i^+}{\max \vec{d}} \vec{d}$.
-Now introduce an additional vector of parameters, $\vec{\pi}$, where $\pi_a$ is the proportion of strata $i$ without immunity at time 0, hence, $\vec{r} = 1 - \vec{\pi}$.
-
-From the system of ODEs and the knowledge that all the compartments are growing exponentially at rate $\psi$, one can derive the following for the initial state of the compartments:
-\begin{align}
-    \vec{i_1} &= \vec{i_0} \left(1 + \frac{2\gamma}{\psi + 2\gamma} \right)^{-1} \\
-    \vec{i_2} &= \vec{i_0} - \vec{i_1} \\
-    \vec{e_2} &= \vec{i_1} \frac{\psi + 2\gamma}{2\sigma} \\
-    \vec{e_1} &= \vec{e_2} \frac{\psi + 2\sigma}{2\sigma} \\
-    \vec{s} &= \vec{\pi} - \vec{i_0} - \vec{e_1} - \vec{e_2} \\
-    \vec{p_0} &= \vec{e_1} \frac{\psi + 2\sigma}{\psi + \kappa_0} \\
-    \vec{p_1} &= \vec{p_0} \frac{\kappa_0}{\psi + \kappa_Y} \\
-    \vec{p_2} &= \vec{p_1} \frac{\kappa_Y}{\psi + \kappa_{X1}} \\
-    \vec{p_3} &= \vec{p_2} \frac{p_x \kappa_{X1}}{\psi + \kappa_{X2}}.
-\end{align}
-
-\subsection{Simulation study} \label{SEIR:sec:sim-study}
-
-I ran a simulation study to check that the inference procedure (\cref{SEIR:sec:inference-implementation}) was working correctly and that the model was identifiable.
-I ran 100 simulations, with parameters drawn from the priors (see \cref{SEIR:table:priors}) except that the prior on $\psi$ was $\LN(0.048, 0.0035)$.
-I used the more informative prior for the simulation study to ensure that the simulations were realistic.
-The weakly informative prior, used for the main analysis, allows a wide range of epidemics including some that are shrinking at the start of the time period.
-Priors and population sizes were matched to the East of England region, over the same time period as the main analysis.
-
-All fixed values, including $\matr{\kappa}$, used the previously described in this section.
-This means that the simulations created realistic incidence and prevalence curves (see \cref{SEIR:fig:sim-data}).
-In particular, the October half-term school holiday is visible, especially in the school-age groups and the November lockdown is visible to some extent in all age groups.
-\begin{figure}
-    \thisfloatpagestyle{empty}
-    \vspace{-3cm}
-    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/sim/data}}
-    \caption[Simulated data]{%
-        Incidence (top) and prevalence (bottom) for the 100 simulated data sets as a proportion of each stratum.
-        Prevalence shown is $\bar{p}_{kat}$, the true population prevalence on each day, excluding sampling noise.
-        Each line represents one simulation.
-    }
-    \label{SEIR:fig:sim-data}
-\end{figure}
-
-% Furthermore, the combination of the posteriors across all simulations should recover the prior.
-
-To check the performance of the inference procedure, I calculated the coverage of the 50\%, 75\%, and 95\% central credible intervals for each parameter and simulation.
-I define coverage as the proportion of simulations where the true value of the parameter in that simulation was within the credible interval.
-This procedure is a simplified version of simulation-based calibration~\autocites{cookValidation}{taltsValidating}.
-The coverage for each parameter should equal the nominal probability (\eg the coverage should be 50\% for the 50\% credible intervals).
-I constructed a confidence interval for the coverage using the Pearson-Klopper method implemented in the binom package~\autocite{binom1-1}.
-
-The majority of the confidence intervals contain the nominal probability (see \cref{SEIR:fig:sim-coverage}).
-There is a tendency for the coverage to be slightly lower than the nominal probability, suggesting that the credible intervals are slightly too narrow; this is a common problem when using random-walk MCMC.
-\todo{cite issues with random-walk MCMC exploring space}
-This is similar when considering coverage of the derived quantities incidence and prevalence (see \cref{SEIR:fig:sim-inc-prev}).
-\begin{figure}
-    \includegraphics{SEIR/sim/coverage}
-    \caption[Coverage of simulation study]{%
-        Point estimate and 95\% confidence interval for coverage of the 50\%, 75\%, and 95\% central credible intervals for each parameter; see text for full definition.
-        Vertical lines show the nominal coverage.
-    }
-    \label{SEIR:fig:sim-coverage}
-\end{figure}
-\begin{figure}
-    \thisfloatpagestyle{empty}
-    \vspace{-3cm}
-    \makebox[\textwidth][c]{\includegraphics[width=0.9\paperwidth]{SEIR/sim/predictive_coverage}}
-    \caption[Coverage of simulation study (derived quantities)]{%
-        Point estimate and 95\% confidence interval for coverage of the 50\%, 75\%, and 95\% central credible intervals for incidence (top) and prevalence (bottom); see text for full definition.
-        Horizontal lines show the nominal coverage.
-    }
-    \label{SEIR:fig:sim-inc-prev}
-\end{figure}
-
-In this setup, the parameters $i^+$, $q_c$, and $\psi$ are fully recovered; the other parameters are only partially so: their posterior has moved away from the prior but is still regularised by it (see \cref{SEIR:fig:true-vs-posterior}).
-The ability to identify $q_c$ is an advantage of using data which includes measurements on children, unlike data on hospitalisations or deaths.
-This suggests that these parameters are not practically identifiable in this setting.
-The posteriors align very closely with the priors (not shown) which reinforces this finding.
-The lack of identifiability for $\vec{\pi}$ aligns with previous work showing this parameter is structurally unidentifiable, \ie not identifiable even in the limit of infinite data~\autocite{dankwaStructural}.
-\begin{figure}
-    \includegraphics{SEIR/sim/true_vs_posterior}
-    \caption[True vs posterior parameter values]{%
-        True vs posterior median parameter values for each parameter in each simulation.
-        The black line shows the line $y = x$ which the dots would lie on if the parameter is fully recovered.
-        The blue line is a linear regression line.
-        For $i^+$, $q_c$, and $\psi$, the posterior recovers the true values.
-        For $\vec{\pi}$ and $\rho$, the posterior median is very similar regardless of the true value.
-    }
-    \label{SEIR:fig:true-vs-posterior}
-\end{figure}
-
-\subsection{Application to CIS data} \label{SEIR:sec:application}
-
-Overall the model fits the data well (see \cref{SEIR:fig:prev-young,SEIR:fig:prev-old}).
-There are a few features of the data which are not captured by the model.
-In particular, prevalence during October is higher in the 16--25s than the model predicts and lower in the 11--16s.
-\begin{figure}
-    \thisfloatpagestyle{empty}
-    \vspace{-3cm}
-    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_young}}
-    \captionsetup{width=0.8\paperwidth}
-    \captionof{figure}[SEIR prevalence goodness-of-fit (younger ages)]{%
-        Goodness-of-fit to prevalence data by region and age for the younger three age groups (see \cref{SEIR:fig:prev-old} for the others).
-        The central ribbon and line shows the posterior median and 95\% CrI estimate for the proportion of the relevant strata PCR positive on each day.
-        The lighter outer ribbon is the 95\% CrI including the noise due to the sampling (\ie observation process), aggregated by week.
-        The dots show the observed prevalence in CIS, aggregated by week.
-        For a well-calibrated model, 95\% of the dots would be within the outer ribbon.
-    }
-    \label{SEIR:fig:prev-young}
-\end{figure}
-\begin{figure}
-    \thisfloatpagestyle{empty}
-    \vspace{-3cm}
-    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_old}}
-    \captionsetup{width=0.8\paperwidth}
-    \captionof{figure}[SEIR prevalence goodness-of-fit (older ages)]{%
-        Goodness-of-fit prevalence for the older three age groups (see \cref{SEIR:fig:prev-young} for the other age groups and details).
-    }
-    \label{SEIR:fig:prev-old}
-\end{figure}
-
-The 11--16 age group (roughly corresponding to compulsory secondary school education) is estimated to have the highest incidence (see \cref{SEIR:fig:incidence}).
-Incidence is smooth within each week but shows discontinuities between weeks.
-This is because the parameters within $\matr{\beta}(t)$ are constant within each week.
-A constant $\matr{\beta}(t)$ leads to exponential growth over short time periods, once the initial dynamics have died out.
-When the peak in incidence is estimated to occur varied by region, but due to this reason, is almost always on a Sunday (see \cref{SEIR:fig:peak-incidence}).
-For Yorkshire, the peak is estimated to occur in late October.
-For all other regions, it is estimated to occur in late December or Early January, the period during which severe social distancing measures were applied.
-\begin{figure}
-    \includegraphics{SEIR/CIS/incidence}
-    \caption[Posterior estimates of incidence]{%
-        Posterior estimates (95\% CrI) of incidence by region and age.
-        See main text for discussion.
-    }
-    \label{SEIR:fig:incidence}
-\end{figure}
-
-% \begin{figure}
-%     \includegraphics{SEIR/CIS/p_peak}
-%     \caption[Posterior estimates of peak incidence timing]{%
-%         Posterior probability of the peak incidence for each region and age combination occurring on each day.
-%     }
-%     \label{SEIR:fig:peak-incidence}
-% \end{figure}
-
-The parameters $\beta_t$ absorb all the time-varying changes in transmission not captured by Google mobility and school attendance data.
-They trend downwards up until early December when they start increasing again.
-Each region is estimated independently, however correlation is clear across the regions.
-
-\begin{figure}
-    \includegraphics{SEIR/CIS/beta_walk}
-    \caption[Posterior estimates of $\beta_t$]{%
-        Posterior estimates $\beta_t$ for each week.
-        Shown are ``violins'' where the width is proportional to the kernel-density estimated posterior density at that point.
-        Values trend downwards until early December when an increase starts.
-    }
-    \label{SEIR:fig:beta-walk}
-\end{figure}
-
-One MCMC chain for the North West region appeared to get stuck, with very few moves after the first 100,000 iterations.
-I reran this chain with a new seed.
-Following this, all regions' MCMC chains converged, assessed using the Rhat statistic and ESS.
-Rhat for all parameters is < 1.01, except $\sigma_\epsilon$ in London which had a Rhat of 1.018.
-All parameters had ESS > 1000, except $\sigma_\epsilon$ in London and the East, which had ESS of 197 and 757 respectively.
-As all the parameters of primary interest had high Rhat and ESS, and no parameters had large issues, the inference procedure is considered to have converged.
-
-The \emph{attack rate} is the proportion of the population that have been infected at least once\todo{cite attack rate}.
-In this model it is $1 - \vec{s}$.
-The attack rate is highest in the $[11, 16)$ age group in the North West at 50\% (95\% CrI: 48\%--52\%) (see \cref{SEIR:fig:attack-rates}).
-\begin{figure}
-    \includegraphics{SEIR/CIS/attack_rates}
-    \caption{Posterior median and 95\% CrI of the attack rate in each region and age group.}
-    \label{SEIR:fig:attack-rates}
-\end{figure}
-
-\subsection{Discussion} \label{SEIR:sec:discussion}
-
-The results in this chapter show the feasibility of fitting a mechanistic model using only data on prevalence.
-If the distributions for length-of-stay in each compartment are known, then all other model parameters can  be estimated.
-This includes the age distribution of the infections and the susceptibility of children.
-
-Unlike most prior work, this modelling is based on a data source that is representative of the general population.
-Therefore, the results have fewer biases and informed by data to a greater extent than this work.
-There is little-to-no ascertainment bias in the CIS, unlike sources relying on testing behaviour (in the UK, pillar 2 testing).
-The CIS allows estimates of transmission in children, unlike sources relying on severe events that occur almost entirely in adults~\autocite{bhopalChildren}.
-
-This approach provides additional understanding, compared to the approach in \cref{E-backcalc}.
-For example, decomposing the change in transmission into changing contacts, $\matr{\kappa}(t)$, changing susceptibility $\vec{s}(t)$, and other unmeasured factors, $\beta(t)$.
-By understanding which of these components is changing, policies can be better targeted.
-If it is shown that $\beta(t)$ decreases, this implies that for a given amount of activity in the population, there is less transmission.
-Such decreases might be beneficial as it allows society to operate in a more normal manner without increasing transmission.
-Breaking the trade-off between transmission and activity is desirable because it reduces the disease burden without burdesome restrictions on individuals day-to-day lives.
-
-The parameters $\beta_t$ absorb all the time-varying factors not captured by Google mobility and school attendance data.
-Many factors contribute to the pattern in the $\beta_t$ values, which have a minima near the start of December.
-While this analysis cannot establish causation, several hypotheses can be identified that are worth further investigation.
-The decreases in $\beta_t$ over the autumn could be due to increased caution(\eg masking or stricter adherence to social distancing) and/or interventions short of a lockdown as cases and deaths increased~\autocite{jarvisEffect}, or the impact of local measures in the highest transmission areas~\autocite{scottCovid19}.
-Local measures are not well captured by $\matr{\kappa}$ because the matrix was for the whole of England.
-Allowing each region to have its own $\matr{\kappa}$ may capture the effect of local measures more accurately.
-The increases in $\beta_t$ in December coincide with the rise of the Alpha variant~\autocite{lythgoeLineage}, known to be more transmissible~\autocite{daviesEstimated}.
-A multiple variant model might be able to capture this rise more accurately.
-Such a model would stratify positive tests in the CIS by pre-Alpha or Alpha variant.
-Then, each variant would be explicitly included in the model to fit to this data.
-
-These factors are correlated across the country.
-This can explain that similar patterns emerge for $\beta_t$ in different regions (see \cref{SEIR:fig:beta-walk}), despite these being estimated independently (\ie there is no model structure which favours correlated values).
-These correlations could be exploited to improve the precision of estimates with a more complex model which borrows information across regions.
-
-The current model could be more flexible with regards to the age distribution of infections.
-This would better exploit one of the strengths of the CIS: the availability of data  on all age groups.
-In some regions, there appears to be a trade-off between the 11--16s and 16--25s, with the model misfitting in opposite directions (see \cref{SEIR:fig:prev-young}).
-This implies that the data would allow identification of a more flexible age-based model.
-Currently, the age distribution of infections is determined by $\matr{\beta}$, and to a lesser extent $\vec{\pi}$, the former of which only contains one parameter that affects the age distribution, $\beta_c$.
-More parameters could capture the age distribution more accurately.
-There are various ways to increase the model's flexibility, including introducing additional susceptibility or infectiousness parameters, or by allowing these parameters to vary over time.
-It is plausible that these distributions changed over time in a way not captured by $\matr{\kappa}$, for example due to the introduction and relaxation of control measures within schools.
-$\matr{\kappa}$ only captures changes in in-school transmission through changes in school attendance, not incorporating the probability of transmission, which could be modelled as changes in $q_{aa}$ for $a$ of school age.
-Similarly, there is evidence that $q_c$ is different for Alpha and pre-Alpha variants~\autocite{zhuRole}, which could again motivate these changing over time.
-
-The weekly patterns in incidence are another limitation of the current parameterisation.
-The random walk and contact matrices both vary only on the weekly timescale.
-A daily random walk could provide more flexibility here.
-However, this would greatly increase the number of parameters.
-An alternative would be a smoothly-varying function or an approximation of a random walk such as that proposed by \textcite{ghoshApproximate}.
-
-This discussion makes clear that the results are limited by the number of parameters, currently limited by the inference algorithm.
-The algorithm used here tends to struggle with a large number of parameters.
-Alternative proposal schemes, such as updating the parameters in blocks\todo{cite what RTM does?}, could help, or a more advanced algorithm such as NUTS (used elsewhere in this thesis).
-
-Extending the analyses carried out here to the rest of the pandemic would clearly be of interest.
-I chose the end of the period to coincide with the vaccination rollout.
-Vaccination dramatically affects the transmission dynamics by introducing some additional immunity into the population.
-Mechanistic models require explicit representation of this immunity\todo{cite something about vaccination}. (Birrell et al, 2023)
-
-\todo[inline]{discuss comparisons with the backcalculation, based on notes below (using preliminary results)}
-
-\begin{itemize}
-    \item Results here disagree with previous chapter's results.
-    \item The model in previous chapter is more flexible than here, showing the limitations discussed above. Previous chapter fits better.
-    \item The previous chapter's modelling benefits from being able to use flexible approximate inference methods that scale to many parameters, including random effects.
-Here, I am limited by the inference procedure.
-MCMC was unable to scale to the full model used in the previous chapter, which has a computationally cheaper likelihood as it is a GLM, not requiring solving an ODE.
-    \item Shows that there is a lot of information in the data which could be exploited by a suitably flexible mechanistic model.
-Such a model would be useful for policy-decisions because it could infer where transmission is occurring (\eg in schools or workplaces) and how this changes over time.
-    \item The narrower credible intervals in this chapter are a realisation of the standard bias-variance tradeoff.
-By making stronger assumptions regarding how transmission occurs, the variance is reduced.
-However, if these assumptions are incorrect (which is almost certainly the case), bias occurs.
-\end{itemize}
-
-\section{Comparison of results}
-
-\todo[inline]{add comparison to backcalculation once I have those results}
-
-\section{Conclusion}
-
-\todo[inline]{write conclusion and add to discussion once comparison with backcalculation is made}
-
-\ifSubfilesClassLoaded{
-  \listoftodos
-}{}
-
 \end{document}

### 2024-02-14 transmission-appendix-INLA.tex
start working through to do list in ch 7

@@
 
+\chapter{INLA} \label{transmission:sec:INLA}
+
+INLA (Integrated Nested Laplace Approximation) is a method for approximate Bayesian inference.
+It was introduced by \textcite{rueINLA}.
+This appendix is based on that paper, as well as the seminar \textcite{rueINLAseminar} and the tutorial paper \textcite{martinoINLAtutorial}.
+
+\section{Inference procedure} \label{transmission:sec:INLA:inference}
+
+\section{Drawing from the posterior} \label{transmission:sec:INLA:posterior}
+
+\section{Default priors} \label{transmission:sec:INLA:priors}
+
 \end{document}

### 2024-02-14 transmission.tex
add INLA appendix
start working through to do list in ch 7

@@
 \todo[inline]{%
-    Intro probably needs rewriting, with novelty claims and a bit more context.
-    Revisit once intro chapters written.
+    Revisit intro.
     Question: is CIS (or similar) enough to estimate incidence / transmission?
-
 }
@@
 I take two approaches, backcalculation and a mechanistic model, the theoretical advantages of each were discussed in \cref{E-inc-prev:sec:infection-process}.
-\todo{what are the advantages of each approach specifically in this context?}
 
@@
 This approach enables estimation of incidence within (sub)populations of England.
+There are several attractive features about this approach in the context here.
+The backcalculation approach makes no assumptions about the transmission process, which have the potential to introduce bias.
+It can also easily incorporate arbitrary covariates to adjust for non-response bias.
+The scale of the CIS and desire to correct for non-response bias presents challenges for this methodology, especially within the limited computation of the SRS environment.
 
@@
 Finally, in \cref{transmission:sec:comparison}, I compare the results of the two approaches.
-
-\section{Data} \label{transmission:sec:data}
-
-\todo[inline]{This is now in the final section of chapter 2.}
-
-% Throughout this chapter, the CIS data will be stratified into the regions of England and the following age groups.
-% \begin{enumerate}
-%     \item $[0, 11)$: children of pre-school and primary school age.
-%     \item $[11, 16)$: children of compulsory secondary-school age.
-%     \item $[16, 25)$: adolescents and young adults who may remain in formal education.
-%     \item $[25, 50)$: younger working age adults.
-%     \item $[50, 70)$: older working age adults.
-%     \item $[70, \infty)$: elderly.
-% \end{enumerate}
-% These groupings align closely with those previously used for analysis of the CIS data~\autocite[e.g.:][]{houseInferring}{walkerTracking}.
-
-% I use the period from the expansion of the CIS, starting Mon 31st August 2020 until Sun 24th Jan 2021 (inclusive).
-% This period is chosen because there were very few positives in CIS in summer 2020 due to low prevalence and a smaller sample size, and ends when vaccination crossed 10\% of England's population.
-
-% I will model the test results as overdispersed relative to a binomial model.
-% Overdispersion allows for household clustering\todo{cite beta-binomial as a model for clustering} and other overdispersion in the data, which is common in epidemiological data\todo{cite that epi data is often overdispersed}.
-% The parameterisation of the beta-binomial that I use (detailed in \cref{E-distributions}) has an overdispersion parameter $\rho$, regarded as a nuisance parameter in this context.
-% $\rho=0$ means the beta-binomial coincides with the binomial distribution.
-% As $\rho \to \infty$, the distribution tends towards the uniform distribution on $[0, 1]$~\autocite{hughesUsing}.
+I develop a novel observation model to allow the transmission model to be fit to the CIS data.
+Furthermore, fitting to just prevalence data has not previously been done in the literature (see\todo{ref lit review in intro})\todo{is using prevalence data novel or only in SARS-CoV-2 context?}.
 
@@
 Estimation of prevalence from the CIS data has been carried out through a well-established methodology, used throughout the pandemic~\autocite{pouwelsCommunity,cisMethodsONS}.
-The method is based on MRP, a statistical methodology developed to correct political opinion polling suffering from response bias\todo{cite MRP}.
+The method is based on MRP, a statistical methodology developed to correct political opinion polling suffering from response bias~\autocite{gelmanMRT,gelmanPoststratication,parkMRP}.
 MRP is a two-step process.
@@
 The strata will represent at least age group and region, although including ethnicity and sex is also due to non-response in the survey (see \cref{E-biology-data:sec:positivity}).
-\todo{make sure ch 2 mentions strata explicitly}
 For many combinations of stratum $i$ and day $t$, the number of positives will be small, often fewer than 5; therefore some form of borrowing between time and strata is required.
@@
 In particular, using MCMC, sufficient covariates to adjust for the non-response bias, and a period of more than 6--8 weeks leads to a model that is computationally prohibitive ~\citePersonalComms{Koen Pouwels}.
-\todo{make sure SRS computational issues are mentioned in ch 1 or 2}
-Instead, previous MRP implementations have used the Integrated Nested Laplace Approximation (INLA)~\autocite{rueINLA}, implemented in R-INLA~\autocite{RINLA}\todo{add an appendix explaining INLA}.
+Instead, previous MRP implementations have used the Integrated Nested Laplace Approximation (INLA)~\autocite{rueINLA}, implemented in R-INLA~\autocite{RINLA} (see \cref{transmission:sec:INLA}).
 
@@
     \item $n_{taes}$ and $y_{taes}$ are the number of individuals tested and the number of positive test results on day $t$ in individuals of age group $a$, ethnicity $e$, and sex $s$.
-I use a beta-binomial distribution, as discussed in \cref{transmission:sec:data}.
+I use a beta-binomial distribution, as discussed in \cref{E-biology-data:sec:positivity}.
     \item $\beta$ is a global intercept term with the prior $\beta \dist \N(0, 1/0.001^2)$.
@@
     \item $u_a$ and $v_e$ are independent Gaussian random effects for age and ethnic group respectively.
-Their standard deviation have a penalised complexity prior\todo{cite PC prior} with a 10\% probability of being greater than 1.
+Their standard deviation have a penalised complexity prior~\autocite{simpsonPenalising} with a 10\% probability of being greater than 1.
     \item $w_t$ is a second-order random walk, that is $w_t = 2w_{t-1} - w_{t-2} + \epsilon_t$ where $\epsilon_t \dist \N(0, \sigma_w^2)$.
@@
 \end{itemize}
-All parameters have the default R-INLA priors, except where stated above.
-\todo{include what the default priors are (maybe in the appendix)}
+All parameters have the default R-INLA priors (see \cref{transmission:sec:INLA:priors}), except where stated above.
 For these parameters, the large amount of data means that the prior has little influence on the posterior distribution.
@@
 
-Having fitted the prevalence model, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution.
-\todo{include how R-INLA samples from the posterior in an appendix}
+Having fitted the prevalence model, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution (using the procedure described in \cref{transmission:sec:INLA:posterior}).
 For each stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
 I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:observation-process} justifies, that is the estimate of the incidence is negligibly affected by approximating the distribution $\vec{P} \mid \vec{Z}$ by its mean.
-\Cref{E-inc-prev:eq:EPt}\todo{consider writing the matrix version where this is introduced in ch 3}, considered across different values of $t$, can be written as $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix formed from the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
+\Cref{E-inc-prev:eq:EPt}, considered across different values of $t$, can be written as $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix formed from the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
 Dividing each side by the total number of individuals in each stratum transforms number of prevalent individuals to prevalence and incidence to incidence proportion.
@@
 
-A prior constraining incidence to be non-negative and then fit directly to the data could be used, but this could no longer be fit within the R-INLA framework.
-\todo{Maybe move this paragraph to the discussion?}
-Hence, an alternative inference methodology, or simpler model (\eg containing fewer random effects), would be required.
-% MCMC is likely too computationally expensive.
-% Other approaches, such as a Laplace approximation or variational inference, could be considered.
-For this analysis, I replaced any value of incidence estimated to be negative with 0.
-
 For $\vec{P} = \matr{S} \vec{I}$ to have a unique solution, $\vec{I}$ and $\vec{P}$ need the same dimension.
@@
 The most explored explanation for this is the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocite{walkerTracking,daviesEstimated,lythgoeLineage}.
-There is weak evidence that when incidence starts increasing is correlated with when the Alpha variant's spread (see \cref{transmission:fig:backcalc-alpha}).
+There is evidence that when incidence starts increasing is correlated with when the Alpha variant's spread (see \cref{transmission:fig:backcalc-alpha}).
 A second peak occurs in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
@@
     }
+    \label{transmission:fig:backcalc-alpha}
 \end{figure}
@@
 
-\todo[inline]{check discussing non-INLA approximations}
+\todo[inline]{check discussing non-INLA approximations, including how this would prevent negative incidence}
+A prior constraining incidence to be non-negative and then fit directly to the data could be used, but this could no longer be fit within the R-INLA framework.
+Hence, an alternative inference methodology, or simpler model (\eg containing fewer random effects), would be required.
+% MCMC is likely too computationally expensive.
+% Other approaches, such as a Laplace approximation or variational inference, could be considered.
+For this analysis, I replaced any value of incidence estimated to be negative with 0.
 
@@
 I assume that $q_{aa'}(t) = \beta(t) q^\text{susc}_{a} q^\text{inf}_{a'}$ where $\beta(t)$ is a time-varying reference probability of transmission; $q^\text{susc}_a$ is the relative \emph{susceptibility} of stratum $a$, \ie how vulnerable to infection individuals in stratum $a$ are; and $q^\text{inf}_{a'}$ is the relative \emph{transmissibility} of stratum $a'$, \ie how likely individuals in stratum $a$ are to pass on an infection.
-Separating relative transmissibility and susceptibility into two independent factors is a common approach\todo{cite anderson and may}.
 These measures include both biological differences, \eg due to differences in immune systems, or behavioural, \eg if infants are less hygienic than adults.
@@
 I will choose $l$ and the transition rates such that combined length of stay (\ie time from entry in $p_1$ until exit from $p_l$) approximates the duration of positivity distribution estimated in \cref{E-imperf-test}.
-This idea extends the technique of having multiple exposed or latent compartments (introduced in \cref{E-SEIR:sec:non-exponential}), and is commonly used to enforce correct durations in compartmental models\todo{cite phase-type distributions in compartmental models}.
+This idea extends the technique of having multiple exposed or latent compartments (introduced in \cref{E-SEIR:sec:non-exponential}).
+While the use of exponential or Erlang (gamma distributions with integer first parameter) is common in the observation model of compartmental models~\autocite[e.g.][]{overtonEpiBeds}, the use of more flexible phase-type distributions, as here, has rarely, if ever, been used in practice~\autocite{hurtadoGLCT}.
+This broader class of distributions allows the model to capture the tail of the distribution, which is important when incidence suddenly decreases.
 
@@
 The remaining parameters have weakly informative priors.
-
 \begin{landscape}
@@
     Initial growth rate & $\psi$ & $\N(0.06, 0.04)$ & Weakly informative \\
-    Standard deviation of random walk steps on log scale & $\sigma_\epsilon$ & $\Exponential(80)$ & Penalising towards constant \\
-    Overdispersion of observations & $\rho$ & $\Exponential(2 \times 10^5)$ & Penalising towards binomial
+    Standard deviation of random walk steps on log scale & $\sigma_\epsilon$ & $\Exponential(80)$ & Penalised complexity prior \\
+    Overdispersion of observations & $\rho$ & $\Exponential(2 \times 10^5)$ & Penalising deviations from binomial
 \end{tabular}
@@
 
-\todo[inline]{insert where the seroprevalence priors come from}
-
+To form the prior for $\vec{\pi}$, I used the analysis framework of \textcite{amirthalingamSeroprevalence}, applied to seroprevalence data from blood donors~\autocite{amirthalingamSeroprevalence} and a representative sample of children~\autocite{ratcliffeCommunity} in August 2020.
+This analysis uses a Bayesian framework to correct the seroprevalence data for the sensitivity and specificity of the serological tests, using an informative prior.
+\Textcite{amirthalingamSeroprevalence} formed the prior for the tests by evaluating the tests using samples which are known to be negative (taken before the pandemic) or positive (taken from an individual following a positive RT-PCR test).
+A regression model predicts the proportion of the population with prior immunity in each age group and region.
+I approximated the posterior of this proportion with a beta distribution to form the prior for the analysis in this chapter (see \cref{SEIR:table:immunity-prior}).
 \begin{table}
@@
     \end{tabular}
-\caption{Priors for $\vec{\pi}$, the proportion of the population without prior immunity, in each age group and region combination.
-Based on seroprevalence in blood donors~\autocite{amirthalingamSeroprevalence} and a representative sample of children~\autocite{ratcliffeCommunity} in August 2020.}
+\caption{Priors for $\vec{\pi}$, the proportion of the population without prior immunity, in each age group and region combination.}
 \label{SEIR:table:immunity-prior}
@@
 \begin{figure}
-    \todo[inline]{Add (A) incidence to the top panel and (B) prevalence to the bottom panel}
     \thisfloatpagestyle{empty}
@@
     \caption[Simulated data]{%
-        Incidence (top) and prevalence (bottom) for the 100 simulated data sets as a proportion of each stratum.
+        (A) Incidence and (B) prevalence for the 100 simulated data sets as a proportion of each stratum.
         Prevalence shown is $\bar{p}_{kat}$, the true population prevalence on each day, excluding sampling noise.
@@
     \vspace{-3cm}
-    \todo[inline]{Add (A) incidence to the top panel and (B) prevalence to the bottom panel}
     \makebox[\textwidth][c]{\includegraphics[width=0.9\paperwidth]{SEIR/sim/predictive_coverage}}
     \caption[Coverage of simulation study (derived quantities)]{%
-        Point estimate and 95\% confidence interval for coverage of the 50\%, 75\%, and 95\% central credible intervals for incidence (top) and prevalence (bottom); see text for full definition.
+        Point estimate and 95\% confidence interval for coverage of the 50\%, 75\%, and 95\% central credible intervals for (A) incidence and (B) prevalence; see main text for full definition.
         Horizontal lines show the nominal coverage.
@@
 The ability to identify $q_c$ is an advantage of using data which includes measurements on children, unlike data on hospitalisations or deaths.
-\todo{is $\pi_0$ unindentified because of $q_c$ - check correlation plot}
 
@@
 
-The \emph{attack rate} is the proportion of the population that have been infected at least once\todo{cite attack rate}.
+The \emph{attack rate} is the proportion of the population that have been infected at least once.
 In this model it is $1 - \vec{s}$, the proportion susceptible.
-The attack rate is highest in the $[11, 16)$ age group in the North West at 68\% (95\% CrI: 66\%--70\%) (see \cref{SEIR:fig:attack-rates}).
-\todo{insert increase in attack rate over period}
+The attack rate is highest in the $[11, 16)$ age group in the North West at 68\% (95\% CrI: 66--70\%) (see \cref{SEIR:fig:attack-rates}).
+Some of these infections occurred prior to the period covered by this study; during this period 23\% (95\% CrI: 21--24\%) of the population in the North West aged 11--16 were infected.
 \begin{figure}
@@
 
-This discussion makes clear that the results are limited by the number of parameters, currently limited by the inference algorithm.
-\todo{Rephrase this sentence}
-The algorithm used here tends to struggle with a large number of parameters.
-Alternative proposal schemes, such as updating the parameters in blocks\todo{cite what RTM does?}, could help, or a more advanced algorithm such as NUTS (used elsewhere in this thesis).
+This discussion makes clear that the results are limited by the number of parameters, which is currently limited by the inference algorithm.
+The MCMC algorithm used tends to struggle with a large number of parameters.
+Alternative proposal schemes, such as updating the parameters in blocks could help, or a more advanced algorithm such as NUTS (used elsewhere in this thesis).
 
@@
   \appendix
-  \subfile{transmission-appendix}
+  \subfile{transmission-appendix-INLA}
+  \subfile{transmission-appendix-phase-type}
   \listoftodos

### 2024-02-14 SEIR/CIS/attack_rates.R

@@ -6,7 +6,7 @@ library(tidyr)
 source(here::here("utils.R"))
 source(here::here("SEIR/utils.R"))
 
-final_states = readr::read_csv(here::here("SEIR/CIS/final_state.csv")) |>
+final_states = readr::read_csv(here::here("SEIR/CIS/final_state.csv"), show_col_types = FALSE) |>
     mutate(
         .chain = factor(chain + 1),
         .iteration = iteration + 1,
@@ -16,12 +16,38 @@ final_states = readr::read_csv(here::here("SEIR/CIS/final_state.csv")) |>
     select(!c(chain, iteration)) |>
     filter(.iteration >= 500e3)
 
+raw_params = readr::read_csv(here::here("SEIR/CIS/params.csv"), show_col_types = FALSE) |>
+    filter(stringr::str_starts(parameter, "pi"))#
+initial_attack = raw_params |>
+    separate_wider_position(parameter, c(3, "age" = 1, 1)) |>
+    transmute(
+        .chain = factor(chain + 1),
+        .iteration = iteration + 1,
+        .draw = max(.iteration) * (as.integer(.chain) - 1) + .iteration,
+        region = normalise_region_names(region),
+        age_int = as.integer(age) + 1,
+        initial_attack = 1 - expit(value),
+    )
+
 attack_rate = final_states |>
     filter(state == "S") |>
-    mutate(occupancy = 1 - occupancy) |>
+    mutate(
+        occupancy = 1 - occupancy,
+        age = forcats::as_factor(age),
+        age_int = as.integer(age),
+    ) |>
+    # filter(region=="North East",age_int==1,.chain==1,)#.iteration==1)
+    left_join(
+        initial_attack,
+        by = c(
+            "region", "age_int",
+            ".chain", ".iteration"
+        )
+    ) |>
     select(!state) |>
+    mutate(increase = occupancy - initial_attack) |>
     group_by(region, age) |>
-    median_qi(.simple_names = FALSE)
+    median_qi(occupancy, increase, initial_attack, .simple_names = FALSE)
 readr::write_csv(attack_rate, here::here("SEIR/CIS/attack_rates.csv"))
 
 p_attack_rate = ggplot(attack_rate, aes(x = age, y = occupancy, colour = age)) +

### 2024-02-14 SEIR/sim/data.R

@@ -6,7 +6,7 @@ library(stringr)
 library(tidyr)
 source(here::here("utils.R"))
 
-inputs = readr::read_csv(here::here("SEIR/sim/sim_output.csv"))
+inputs = readr::read_csv(here::here("SEIR/sim/sim_output.csv"), show_col_types = FALSE)
 
 do_plot = function(base_gg, y_axis_label) {
     return(
@@ -21,11 +21,13 @@ do_plot = function(base_gg, y_axis_label) {
 
 p_incidence = inputs |>
     ggplot(aes(y = incidence)) |>
-    do_plot("Incidence")
+    do_plot("Incidence") +
+    ggtitle("(A) Incidence")
 
 p_prevalence = inputs |>
     ggplot(aes(y = prevalence)) |>
-    do_plot("Prevalence")
+    do_plot("Prevalence") +
+    ggtitle("(B) Prevalence")
 
 ggsave(
     filename = here::here("SEIR/sim/data.pdf"),

### 2024-02-14 transmission.tex
transmission chapter almost there

@@
 I develop a novel observation model to allow the transmission model to be fit to the CIS data.
-Furthermore, fitting to just prevalence data has not previously been done in the literature (see\todo{ref lit review in intro})\todo{is using prevalence data novel or only in SARS-CoV-2 context?}.
+Furthermore, fitting to just prevalence data has not previously been done in the literature (see\todo{ref lit review in intro and copy its novelty claims}).
 
@@
 Instead, previous MRP implementations have used the Integrated Nested Laplace Approximation (INLA)~\autocite{rueINLA}, implemented in R-INLA~\autocite{RINLA} (see \cref{transmission:sec:INLA}).
+theta
 
+\todo[inline]{Is it OK that $x$ and $y$ in the following two paragraphs don't align with the rest of the thesis chapter?
+Should I use alterantive notation - but then different to standard in the INLA papers.}
 INLA-based inference can only be applied to LGMs (\emph{Latent Gaussian Models})~\autocite{rueINLAseminar,rueINLA,martinoINLAtutorial}.
-\todo{check notation here doesn't clash with other notation}
-An LGM is defined in terms of a vector of hyperparameters, $\vec{\theta}$; a deterministic transformation of the hyperparameters into covariance matrix, $\matr{\Sigma}(\vec{\theta})$; a vector of latent variables, $\vec{x}$; a vector of $n$ data points, $\vec{y} = [y_1, \dots, y_n]^T$; and $n$ linear predictors $\vec{\eta} = [\eta_1, \dots, \eta_n]^T$.
-To be an LGM, $\vec{x}$ must be a latent Gaussian field, that is $\vec{x} \mid \vec{\theta} \dist \N(\vec{0}, \matr{\Sigma}(\vec{\theta}))$; and the likelihood must have the conditional independence property $p(\vec{y} \mid \vec{\theta}, \vec{x}) = \prod_j p(y_j \mid \eta_j(\vec{x}), \vec{\theta})$.
+An LGM is defined in terms of a vector of hyperparameters, $\vec{\varphi}$; a deterministic transformation of the hyperparameters into covariance matrix, $\matr{\Sigma}(\vec{\varphi})$; a vector of latent variables, $\vec{x}$; a vector of $n$ data points, $\vec{y} = [y_1, \dots, y_n]^T$; and $n$ linear predictors $\vec{\eta} = [\eta_1, \dots, \eta_n]^T$.
+To be an LGM, $\vec{x}$ must be a latent Gaussian field, that is $\vec{x} \mid \vec{\varphi} \dist \N(\vec{0}, \matr{\Sigma}(\vec{\varphi}))$; and the likelihood must have the conditional independence property $p(\vec{y} \mid \vec{\varphi}, \vec{x}) = \prod_j p(y_j \mid \eta_j(\vec{x}), \vec{\varphi})$.
 
@@
 
-\todo[inline]{check discussing non-INLA approximations, including how this would prevent negative incidence}
-A prior constraining incidence to be non-negative and then fit directly to the data could be used, but this could no longer be fit within the R-INLA framework.
-Hence, an alternative inference methodology, or simpler model (\eg containing fewer random effects), would be required.
-% MCMC is likely too computationally expensive.
-% Other approaches, such as a Laplace approximation or variational inference, could be considered.
-For this analysis, I replaced any value of incidence estimated to be negative with 0.
+Alternative inference techniques could allow more powerful modelling approaches, if they allowed a broader set of models than LGMs.
+In particular, the incidence proportion could be directly modelled rather than the prevalence.
+This could allow more interpretable priors to be used because the infection process is the process that corresponds to biological and behavioural processes.
+Furthermore, the incidence proportion could be restricted to be non-negative, preventing the negative incidence seen in some strata.
+
+Various alternative inference techniques could be considered.
+If the infection process is considered a Poisson process (see \cref{E-inc-prev:sec:poisson-process}), then, conditional on the process's intensity, there is independence between individuals' infection status.
+Therefore, the incidence proportion and prevalence can both be viewed as the mean of many Bernoulli variables, meaning that the Central Limit Theorem suggests they are approximately normally distributed.
+Exploiting this property could enable efficient approximations to the posterior.
 
@@
 The time-varying component, $\beta(t)$, allows for behavioural changes that impact transmission but are not captured by the contact matrices (\eg mask wearing).
-I follow prior work~\autocite[e.g.][]{birrellRealtime}\todo{find other references for random walk on transmission potential} in using a random walk prior on $\log\beta(t)$  on the weekly timescale.
+I impose a random walk prior on $\log\beta(t)$ on the weekly timescale, smoothing its value over time.
 This prior encodes that behaviour changes smoothly over time.
@@
 Vaccination dramatically affects the transmission dynamics by introducing some additional immunity into the population.
-Mechanistic models require explicit representation of this immunity\todo{cite something about vaccination, including new RTM paper}.
+Mechanistic models require explicit representation of this immunity.
 
@@
 These patterns can be interpreted as an instance of the common bias---variance trade-off.
-\todo{Comment on SEIR not fully adjusted for non-response bias}
 The SEIR model imposes stronger assumptions about the epidemic dynamics.
 The assumptions reduces the variance of the estimates, including by making them smoother, but if the assumptions are incorrect, bias is introduced.
+Furthermore, the SEIR model does not fully adjust for the non-response bias.
+In particular, it makes no adjustment for the under-representation of ethnic minority groups in the CIS.
+Adjusting for more covariates would require further stratifying the population, and finding an appropriate parameterisation of the contact matrix.
 This trade-off based on the strength of assumptions has long been known to apply in the backcalulcation context~\autocite[e.g.][section 8.3]{brookmeyerBackcalculation}.

### 2024-02-16 incidence-prevalence.tex
proofing ch 4
ch 3 to dos

@@
 
-\todo[inline]{Add some refs about duration and survival}
-\emph{Duration}, or \emph{time-to-event}, distributions can be expressed in several equivalent ways.
+\emph{Duration}, or \emph{time-to-event}, distributions can be expressed in several equivalent ways~\autocite[17]{sunStatisticala}.
 The duration distribution I will be concerned with is the distribution of the number of days for which an individual is prevalent.
@@
 
-A compartmental model consists of a state vector $\vec{x}(t)$ and a set of parameters $\vec{\theta}$~\autocite{birrellEvidence}.
-\todo{Cite other papers: see Alice's thesis (chapter 3, especially 3.1), refs within Birrell, and \href{https://www.sciencedirect.com/science/article/pii/S2468042716300495}{Allen 2017}}
+A compartmental model consists of a state vector $\vec{x}(t)$ and a set of parameters $\vec{\theta}$~\autocite{birrellEvidence,dukicTracking,corbellaThesis}.
 The elements of $\vec{x}(t)$ are the latent proportion of the population (or number of individuals) in each compartment at time $t$.
@@
 
-Stochastic models specify a probability distribution for the current state $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta})$ where $\set{H}_t = \{ \vec{x}(t') \ssep t' < t \}$ is the epidemic's history and $\vec{\theta}$ are model parameters\todo{Cite stochastic models}.
+Stochastic models specify a probability distribution for the current state $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta})$ where $\set{H}_t = \{ \vec{x}(t') \ssep t' < t \}$ is the epidemic's history and $\vec{\theta}$ are model parameters~\autocite{birrellEvidence,dukicTracking,corbellaThesis,keelingModeling}.
 Stochastic models are most commonly specified in discrete time, using a Markov assumption.
@@
 
-The simplest compartmental model for a disease which confers immunity is the SIR (susceptible-infectious-recovered) model\todo{cite SIR being simplest} (depicted in \cref{SEIR:fig:SIR}).
+The simplest compartmental model for a disease which confers immunity is the SIR (susceptible-infectious-recovered) model~\autocite[15]{keelingModeling}, depicted in \cref{SEIR:fig:SIR}.
 The name SIR derives from the structure of the model.
@@
 
-The SIR model is described by the following system of ODEs~\autocite[19]{keelingModeling}:
+The SIR model is described by the following system of ODEs (ordinary differential equations)~\autocite[19]{keelingModeling}:
 \begin{align}
@@
 % Each of the infectious individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
-Following the same argument as in \cref{SEIR:sec:SIR} finds that, in expectation, the infectious individual makes contact with susceptible individuals in stratum $a$ at a rate of $\kappa_{aa'} s_a(t)$ per day.
+Following the same argument as in \cref{SEIR:sec:SIR} finds that the infectious individual makes contact with susceptible individuals in stratum $a$ at a mean rate of $\kappa_{aa'} s_a(t)$ per day.
 Now assume that there is a constant probability of transmission upon contact between an infectious individual in strata $a'$ and susceptible in strata $a$ individual, $q_{aa'}$.
-Therefore, the expected rate of infections in strata $a$ caused by this infectious individual is $\kappa_{aa'} s_a(t) q_{aa'}$ per day.
-Multiplying by the number of infectious individuals in strata $a'$, $N_{a'} i_{a'}(t)$, gives $\kappa_{aa'} q_{aa'} s_a(t) N_{a'} i_{a'}(t)$ where $i_{a'}(t) = \sum_{j=1}^n i_{j,a'}$ is the total proportion of stratum $a'$ that is infectious, regardless of which compartment they are in.
-The constant terms here are absorbed into the WAIFW matrix, hence $\beta_{aa'} = \kappa_{aa'} q_{aa'} N_{a'}$~\autocite[section 9.2]{diekmannMathematical}.
+Therefore, the mean rate of infections in strata $a$ per day caused by this infectious individual is $\kappa_{aa'} s_a(t) q_{aa'}$ per day.
+Multiplying by the number of infectious individuals in strata $a'$, $N_{a'} i_{a'}(t)$, and dividing by the size of strata $a$, $N_a$, gives $\kappa_{aa'} q_{aa'} s_a(t) \frac{1}{N_a} N_{a'} i_{a'}(t)$ where $i_{a'}(t) = \sum_{j=1}^n i_{j,a'}$ is the total proportion of stratum $a'$ that is infectious, regardless of which compartment they are in.
+The constant terms here are absorbed into the WAIFW matrix, hence $\beta_{aa'} = \kappa_{aa'} q_{aa'} \frac{1}{N_a} N_{a'}$~\autocite[section 9.2]{diekmannMathematical}.
 This gives the total force of infection on strata $a$ from all strata is, therefore, $\lambda_a(t) = \sum_{a' \in \set{A}} \beta_{aa'} i_{a'}(t)$ as above.
@@
 The rate at which individuals move between the susceptible and exposed compartments is now $\vec{s}(t) \circ \vec\lambda(t)$, where $\circ$ denotes element-wise multiplication.
-\todo{Is a division by $N_a$ necessary here? Check previous notes on Overleaf including on what Edwin provides.}
 
@@
 
-\todo[inline]{Relate the following paragarph to the similar discussion in ch 7}
 The expectation in \cref{inc-prev:eq:EPt} is a discrete convolution equation.
 The equation is of the same form as the equation used for HIV/AIDS backcalculation, as well as being found in many other fields~\autocite[and references therein]{brookmeyerBackcalculation}.
-The system can be written in the form $\E(\vec{P} \mid \vec{Z}) = \matr{S} \vec{Z}$ where $\matr{S}$ is a lower-triangular matrix formed from $S$.
-Numerical methods for solving systems of this form, that is solving for $\vec{Z}$ (non-random in this expression) for a known $\E (\vec{P} \mid \vec{Z} )$, are well-established~\autocite[e.g.:][section 8.2]{highamAccuracy}.
-However, these algorithms can be unstable, meaning that numerical errors are large.
-This is a particular issue if $S(1)$ is small.
-Since, by definition, $S(1)=1$ here, this is not an issue in this context.
+
+Considering multiple values of $t$ makes \cref{inc-prev:eq:EPt} into a system of simultaneous equations.
+The system can be written in the form $\E(\vec{P} \mid \vec{Z}) = \matr{S} \vec{Z}$ where $\matr{S}$ is a lower-triangular matrix formed from $S$:
+\begin{align}
+    (\matr{S})_{i,j} = \begin{cases}
+        S(i - j + 1) & i - \dmax - 1 \leq j \leq i \\
+        0 & \text{otherwise.}
+    \end{cases}
+\end{align}
+In \cref{E-backcalc}, I will need to solve for this system for $\vec{Z}$ given a known $\hat{P} = \E(\vec{P} \mid {Z})$.
+As $\matr{S}$ is lower-triangular, I use forward-substitution~\autocite{cormenMatrix}.
+Forward-substitution sequentially solves the $t$th element of $\vec{Z}$ ($t = 1, \dots, T$) by calculating:
+\begin{align}
+Z_t
+&= \frac{\hat{P}_t - \sum_{i=1}^{t-1} S_{ti} Z_i}{S_{tt}} \\
+&= \hat{P}_t - \sum_{i=1}^{t-1} S(t - i + 1) Z_i.
+\label{inc-prev:eq:forward-substitute}
+\end{align}
+
+% Numerical methods for solving systems of this form, that is solving for $\vec{Z}$ (non-random in this expression) for a known $\E (\vec{P} \mid \vec{Z} )$, are well-established~\autocite[e.g.:][section 8.2]{highamAccuracy}.
+% However, these algorithms can be unstable, meaning that numerical errors are large.
+% This is a particular issue if $S(1)$ is small.
+% Since, by definition, $S(1)=1$ here, this is not an issue in this context.
 %However, if instead of estimating the incidence of RT-PCR positives, the aim was to estimate the incidence of infections, then $S(1)$ would be small and this would be an issue.
 %While this quantity is more epidemiologically relevant, the time from infection to RT-PCR positivity is short and, for now, I will neglect the difference.
+
+Unlike an inference procedure which infers the parameters simultaneously, there is no feedback loop: the value of $Z_{t}$ does not affect $Z_{t'}$ for $t' < t$.
+The second term here is the contribution of the incidence prior to day $t$ to day $t$'s prevalence.
+The remainder of the prevalence on day $t$ is therefore due to incidence on day $t$.
+% Forward substitution is numerically stable unless the diagonal elements, here $S(1)$, are small relative to the off-diagonal elements, here $S(i)$ for $i > 0$~\autocite[section 8.2]{highamAccuracy}.
+% Here, the diagonal is $S(1) = 1$ and hence the matrix the solution is numerically stable.
 
@@
 The momentum increases in the direction the posterior density increases, using the gradient.
-\todo{add something about leapfrog steps}
 This means that the sampler tends to propose moves with higher posterior density.
@@
 
-The \emph{No U-Turn Sampler} (NUTS)~\autocite{hoffmanNUTS} is a variant of HMC that automatically tunes the number of leapfrog steps to take in the Hamiltonian dynamics.
+The \emph{No U-Turn Sampler} (NUTS)~\autocite{hoffmanNUTS} is a variant of HMC that automatically tunes the Hamiltonian dynamics hyperparameters.
 NUTS has been implemented in a variety of software packages making it easy to apply to arbitrary posterior distributions.
 I will make use of the implementation in the package \emph{Sampling through adaptive neighborhoods} (Stan)~\autocite{Stan-2-32-2}.
-Stan implements \emph{automatic differentiation}\todo{cite automatic diff} to calculate the gradient of the posterior density without any user input.
+Stan implements \emph{automatic differentiation}~\autocite{griewankAutoDiff} to calculate the gradient of the posterior density without any user input.
 Automatic differentiation uses the chain rule to calculate the gradient of a function from its code by breaking it down into a series of elementary operations, for which the gradient is known~\autocite{autodiff}.
@@
 
-Robust, unbiased prevalence estimates already exist (see\todo{ref prevalence estimates}), and incidence estimates are the objective of this thesis.
+Robust, unbiased prevalence estimates already exist (\cref{E-biology-data:sec:positivity}), and incidence estimates are the objective of this thesis.
 This chapter has shown that the key quantity relating these quantities is the distribution of the duration of RT-PCR positivity.

### 2024-02-16 thesis.tex
Fix file names, build, refs
Fix references
explain calculation of $F_D$ ATACCC posterior
Use RT-PCR (not PCR)
tidy thesis code
preferable citation sorting
chapter 6 comment
Overleaf comments
include bib in toc
add INLA appendix
ch 3 to dos

@@
 \usepackage{bm}
-\usepackage{algorithm2e}
+\usepackage[algochapter]{algorithm2e}
 \usepackage[nameinlink]{cleveref}
@@
 \newcommand{\vlod}{v_\text{lod}}
+\newcommand{\inform}{{_{\text{inform}}}}
 
@@
 \makeatother
-\usepackage[backend=biber,uniquename=false,sortcites=true,maxcitenames=2,style=apa]{biblatex}
+
+%% Bibliography
+\usepackage[backend=biber,uniquename=false,sortcites=true,sorting=ynt,maxcitenames=2,style=apa]{biblatex}
 \addbibresource{references.bib}
+\addbibresource{refs-custom.bib}
 % Display a linked DOI
@@
 \usepackage{subfiles} % Best loaded last in the preamble
-\addbibresource{references.bib}
-\addbibresource{refs-custom.bib}
-\externaldocument[E-]{\subfix{thesis}}
-\externaldocument[E-]{\subfix{output}}
+
+%----Helper code for dealing with external references----
+% (by cyberSingularity at http://tex.stackexchange.com/a/69832/226)
+
+\usepackage{xr}
+\makeatletter
+
+\newcommand*{\addFileDependency}[1]{% argument=file name and extension
+\typeout{(#1)}% latexmk will find this if $recorder=0
+% however, in that case, it will ignore #1 if it is a .aux or 
+% .pdf file etc and it exists! If it doesn't exist, it will appear 
+% in the list of dependents regardless)
+%
+% Write the following if you want it to appear in \listfiles 
+% --- although not really necessary and latexmk doesn't use this
+%
+\@addtofilelist{#1}
+%
+% latexmk will find this message if #1 doesn't exist (yet)
+\IfFileExists{#1}{}{\typeout{No file #1.}}
+}\makeatother
+
+\newcommand*{\myexternaldocument}[1]{%
+\externaldocument[E-]{\subfix{#1}}%
+\addFileDependency{#1.aux}%
+\addFileDependency{#1.tex}%
+}
+%------------End of helper code--------------
+
+% put all the external documents here!
+\myexternaldocument{thesis}
 
@@
 \subfile{introduction}
-
 \subfile{biology-data}
-
 \subfile{incidence-prevalence}
-
-% We wish to estimate the distribution of time for which an individual is positive for.
-% An individual could have multiple infections, and hence I refer to \emph{infection episodes}.
-
-% I define an individual as \emph{detectable} if they would test positive if tested, in the absence of misclassification error (which is dealt with in \cref{E-imperf-test}).
-% Episode $i$'s \emph{duration of positivity}, $D_i$ is the number of days that they are detectable for; I use time discretised to days, therefore, the duration is at least 1 day.
-% Note that the relationship between positivity (\ie the result of a PCR test) and infectiousness (\ie an individual's ability to pass on the virus) is complex~\autocites{lascolaViral}{singanayagamDuration}, and beyond the scope of this thesis.
-% I define an episode as starting at time $B_i$ and ends at $E_i$, hence $D_i = E_i - B_i + 1$.
-
-% An episode is \emph{detected} when they test positive at least once, if this occurs; episode $i$ is detected at time $t$ if and only if: $i$ is detectable at $t$, $i$ is tested at $t$, and the test returns positive (a false negative does not occur).
-
-% The survival time function, parameterised by $\theta$, is the probability that an episode lasts at least $t$ days: $S_\theta(t) = \prob(D_i \geq t \mid B_i = b, \theta)$, where $D_i$ and $B_i$ are the duration and starting time of episode $i$ respectively.
-% I make the standard assumption that $D_i$ and $B_i$ are independent and hence: $S_\theta(t) = \prob(D_i \geq t \mid \theta)$.I utilise survival analysis for analysing this data.
-% Survival analysis is the area of statistics concerned with estimating the distribution of the times between events.
-% It is broadly applicable to many domains, such as the time-to-failure for a mechanical system or the effect of a treatment on time spent in hospital.
-% In the context of this thesis, the distribution to estimate is that of duration of positivity.
-% The \emph{initiating event} is when the duration starts, \ie becoming detectable.
-% The \emph{terminating event} is when the duration ends, \ie becoming no longer detectable.
-
 \subfile{ATACCC}
-
 \subfile{cis-perfect-testing}
-
 \subfile{cis-imperfect-testing}
-
-\subfile{backcalc}
-
-\subfile{SEIR}
+\subfile{transmission}
 
@@
 
-\printbibliography
+\newrefcontext[sorting=nyt]
+\printbibliography[heading=bibintoc]
 
@@
 \subfile{distributions}
+\subfile{ATACCC-appendix-original-analysis}
+\subfile{transmission-appendix-INLA}
+\subfile{transmission-appendix-phase-type}
 
-\chapter{CIS episode definitions} \label{episode-def}
-\todo[inline]{Get this from SARAH}
-
-\subfile{ATACCC-appendix-original-analysis}
-
+\todo[inline]{Delete below lines - only here to make compiling individual chapters work}
 \cref{perf-test:eq:piu}
+\cref{inc-prev:eq:forward-substitute}
 \end{document}

### 2024-02-16 transmission.tex
ch 3 to dos

@@
 I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:observation-process} justifies, that is the estimate of the incidence is negligibly affected by approximating the distribution $\vec{P} \mid \vec{Z}$ by its mean.
-\Cref{E-inc-prev:eq:EPt}, considered across different values of $t$, can be written as $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix formed from the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
-Dividing each side by the total number of individuals in each stratum transforms number of prevalent individuals to prevalence and incidence to incidence proportion.
-
-The matrix $\matr{S}$ is defined to maintain the relationship between incidence and prevalence.
-This gives a lower-triangular of the following form:
-\begin{align}
-    (\matr{S})_{i,j} = \begin{cases}
-        S(i - j + 1) & i - \dmax - 1 \leq j \leq i \\
-        0 & \text{otherwise.}
-    \end{cases}
-\end{align}
-$\matr{S}$ is lower-triangular, hence to solve for $\vec{P}$ I use forward-substitution~\autocite{cormenMatrix}.
-Forward-substitution sequentially solves the $t$th element of $\vec{Z}$ ($t = 1, \dots, T$) by calculating:
+I solve this system using forward substitution, calculating the value of $Z_t$ in \cref{E-inc-prev:eq:forward-substitute} sequentially for $t = 1, \dots, T$.
+\Cref{E-inc-prev:eq:forward-substitute} is repeated here for convenience:
 \begin{align}
@@
 \end{align}
-Unlike an inference procedure which infers the parameters simultaneously, there is no feedback loop: the value of $Z_{t}$ does not affect $Z_{t'}$ for $t' < t$.
-The second term here is the contribution of the incidence prior to day $t$ to day $t$'s prevalence.
-The remainder of the prevalence on day $t$ is therefore due to incidence on day $t$.
-% Forward substitution is numerically stable unless the diagonal elements, here $S(1)$, are small relative to the off-diagonal elements, here $S(i)$ for $i > 0$~\autocite[section 8.2]{highamAccuracy}.
-% Here, the diagonal is $S(1) = 1$ and hence the matrix the solution is numerically stable.
+% \Cref{E-inc-prev:eq:EPt}, considered across different values of $t$, can be written as $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix formed from the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
+Dividing each side by the total number of individuals in each stratum transforms number of prevalent individuals to prevalence and incidence to incidence proportion.
 
@@
 A parsimonious model is needed to parameterise the WAIFW matrix.
-In \cref{E-SEIR:sec:structured-populations}, I explained how the elements of the WAIFW matrix, $\matr{\beta}$ can be decomposed as $\beta_{aa'} = \kappa_{aa'} q_{aa'} N_{a'}$ where\todo{should I repeat these definitions?
-they already exist in ch 3}: $\kappa_{aa'}$ is the mean rate at which an individual in stratum $a'$ makes contact with individuals in stratum $a$; $q_{aa'}$ is the probability of transmission upon contact between an infectious individual in stratum $a'$ and a susceptible individual in stratum $a$; and $N_{a'}$ is the total number of individuals in stratum $a'$.
+In \cref{E-SEIR:sec:structured-populations}, I explained how the elements of the WAIFW matrix, $\matr{\beta}$ can be decomposed as $\beta_{aa'} = \kappa_{aa'} q_{aa'} N_{a'} / N_{a}$ where\todo{should I repeat these definitions?
+they already exist in ch 3}: $\kappa_{aa'}$ is the mean rate at which an individual in stratum $a'$ makes contact with individuals in stratum $a$; $q_{aa'}$ is the probability of transmission upon contact between an infectious individual in stratum $a'$ and a susceptible individual in stratum $a$; and $N_{a}$ is the total number of individuals in stratum $a$.
 For many epidemics, where the population behaviour is fairly constant over time, a constant $\matr{\beta}$ is a good approximation.
 However, in SARS-CoV-2, population behaviour changed dramatically, both spontaneously (\eg due to fear) and in response to government-imposed restrictions.
-Therefore, following \textcite{birrellRealtime}, I introduce a time-dependence giving: $\beta_{aa'}(t) = \kappa_{aa'}(t) q_{aa'}(t) N_{a'}$; $N_{a'}$ is constant because the change in the total population is negligible over the length of time considered here.
+Therefore, following \textcite{birrellRealtime}, I introduce a time-dependence giving: $\beta_{aa'}(t) = \kappa_{aa'}(t) q_{aa'}(t) N_{a'} / N_a$; $N_a$ and $N_{a'}$ are constant because the change in the total population is negligible over the length of time considered here.
 
@@
     \frac{d\vec{r}(t)}{dt} &= 2\gamma \vec{i_2}(t) \\
-    \frac{d\vec{p_0}(t)}{dt} &= \vec{\lambda} \circ \vec{s}(t) - \omega_0 \vec{p_0}(t) \\
+    \frac{d\vec{p_0}(t)}{dt} &= \vec{\lambda} \circ \vec{s}(t) \circ \vec{N} - \omega_0 \vec{p_0}(t) \\
     \frac{d\vec{p_1}(t)}{dt} &= \omega_0 \vec{p_0}(t) - \omega_1 \vec{p_1}(t) \\

### 2024-02-19 transmission.tex
Ch 7 intros

@@
 
-\todo[inline]{%
-    Revisit intro.
-    Question: is CIS (or similar) enough to estimate incidence / transmission?
-}
-
-In this chapter, I use the estimate of the duration of RT-PCR positivity from combining information from the ATACCC and CIS studies (see \cref{E-imperf-test:sec:conclusion}) to investigate estimation of incidence and transmission of SARS-CoV-2 in England.
-This estimation is based only on the CIS data (as shown in \cref{E-biology-data:sec:positivity} for further details).
+In this chapter, I use the estimate of the duration of RT-PCR positivity from combining information from the ATACCC and CIS studies (see \cref{E-imperf-test:sec:conclusion}) to investigate the feasibility of estimating incidence and transmission of SARS-CoV-2 from only the CIS data (described in \cref{E-biology-data:sec:positivity}).
 I will show that the CIS data is sufficient; demonstrating the value of surveys of this type for pandemic surveillance.
 
-I take two approaches, backcalculation and a mechanistic model, the theoretical advantages of each were discussed in \cref{E-inc-prev:sec:infection-process}.
+I take two approaches, backcalculation and a mechanistic model, both previously discussed in \cref{E-inc-prev:sec:infection-process}.
 
@@
 Estimating transmission in addition to incidence answers important epidemiological questions, \eg whether changes in transmission can be explained by changes in contact rates.
+I develop a novel observation model to allow the transmission model to be fit to the CIS data.
+
 Finally, in \cref{transmission:sec:comparison}, I compare the results of the two approaches.
-I develop a novel observation model to allow the transmission model to be fit to the CIS data.
-Furthermore, fitting to just prevalence data has not previously been done in the literature (see\todo{ref lit review in intro and copy its novelty claims}).
+I show they give similar values, providing reassurance of the robustness of the results to modelling assumptions.
+Overall, there is a bias-variance trade-off between the models.
+Fitting to just prevalence data has not previously been done in the literature (see\todo{ref lit review in intro and copy its novelty claims}).
 
@@
 
-\todo[inline]{Rewrite this intro paragraph}
 This section  explains the novel observation model approach I propose for fitting to the CIS prevalence data.

### 2024-02-19 transmission-appendix-INLA.tex
transmission chapter almost there
start writing INLA appendix

@@
 
+\section{Latent Gaussian model}
+
+The INLA method requires a LGM (latent Gaussian model).
+An LGM is defined in terms of a vector of hyperparameters, $\vec{\varphi}$; a deterministic transformation of the hyperparameters into covariance matrix, $\matr{\Sigma}(\vec{\varphi})$; a vector of latent variables, $\vec{x}$; a vector of $n$ data points, $\vec{y} = [y_1, \dots, y_n]^T$; and $n$ linear predictors $\vec{\eta} = [\eta_1, \dots, \eta_n]^T$.
+To be an LGM, $\vec{x}$ must be a latent Gaussian field, that is $\vec{x} \mid \vec{\varphi} \dist \N(\vec{0}, \matr{\Sigma}(\vec{\varphi}))$; and the likelihood must have the conditional independence property $p(\vec{y} \mid \vec{\varphi}, \vec{x}) = \prod_j p(y_j \mid \eta_j(\vec{x}), \vec{\varphi})$.
+
+The Gaussian structure allows the use of the approximations in \cref{transmission:sec:INLA:inference}.
+The conditional independence structure means that the precision matrix, $\matr{Q} = \matr{\Sigma}^{-1}$, is sparse.
+The sparsity of $\matr{Q}$ allows for efficient computation of the procedure.
+
 \section{Inference procedure} \label{transmission:sec:INLA:inference}
+
+\subsection{Gaussian likelihood}
+
+Start by the case in which the likelihood, $\vec{y} \mid \vec{x}, \vec{\theta}$, is Gaussian.
+In this case, as $\vec{x} \mid \vec{\theta}$ is also Gaussian, the distributions of $\vec{x}, \vec{y} \mid \vec{\theta}$ and $\vec{x} \mid \vec{y}, \vec{\theta}$ are both Gaussian too.
+
+The posterior density of $\vec{\theta}$, $\vec{\theta }\mid \vec{y}$, is not Gaussian.
+Its density (up to a normalising constant) is:
+\begin{align}
+p(\vec{\theta} \mid \vec{y}) \propto \frac{p(\vec{x}, \vec{y} \mid \vec{\theta}) p(\vec{\theta})}{p(\vec{x} \mid \vec{y}, \vec{\theta})}.
+\end{align}
+All of the densities on the right-hand side can easily be found analytically.
+
+Assume that $\vec{\theta} \mid \vec{y}$ is unimodal, which is very likely for reasonable priors on $\vec{\theta}$.
+Find its mode via optimisation and explore its area, until the density is negligible.
+By integrating over this area, the normalising constant can be found.
+Hence, the posterior density of the hyperparameters is known.
+
+Finally, the marginal posterior density of $\vec{x}$ is found by integrating over $\vec{\theta}$:
+\begin{align}
+p(\vec{x} \mid \vec{y}) = \int p(\vec{x} \mid \vec{\theta}) p(\vec{\theta} \mid \vec{y}) d\vec{\theta}.
+\end{align}
+which can be computed using a Monte Carlo integration over $p(\vec{\theta} \mid \vec{y})$, previously calculated.
+
+\subsection{Non-Gaussian likelihood}
+
+Now, the INLA approach needs extending to non-Gaussian, but known, likelihoods for $\vec{y} \mid \vec{x}, \vec{\theta}$.
+The density $p(\vec{x}, \vec{y} \mid \vec{\theta})$ is not Gaussian, but still known.
+However, $\vec{x} \mid \vec{y}, \vec{\theta}$ is not Gaussian, and its density is not known.
+
+INLA assumes that $\vec{x} \mid \vec{y}, \vec{\theta}$ is approximately Gaussian, and therefore takes a Laplace approximation to the distribution.
+The greater the smoothing, the better the approximation.
+This is because, if $\vec{x} \mid \vec{y}, \vec{\theta}$ is then dominated by the prior $\vec{x} \mid \vec{\theta}$ and the prior is Gaussian.
+\todo[inline]{Add justification for the Laplace approximation}
 

### 2024-02-19 biology-data.tex
start working through to do list in ch 7

@@
 
-\todo[inline]{Consider finding or making a figure to explain PCR testing.}
-
 PCR (Polymerase Chain Reaction) is a procedure that produces millions to billions of copies of a specific segment of DNA (Deoxyribonucleic acid) from a single copy~\autocite{smithPCR,garibyanPCR}.
@@
 PCR has many uses within the biological sciences; Mark R.\ Hughes, deputy director of the Human Genome Project, has claimed it is the most important scientific technology of the last hundred years~\autocite{powledgePCR}.
-It is central to the work in this thesis because it can be used to detect the presence of SARS-CoV-2 in a sample taken from a human, most commonly a nose and/or throat swab.
+It is central to the work in this thesis because it is used to detect the presence of SARS-CoV-2 in a sample taken from a human, most commonly a nose and/or throat swab.
+
+\subsection{Reverse transcription}
 
@@
 
+\subsection{The PCR process} \label{biology-data:sec:PCR-process}
+
+\todo[inline]{Consider finding or making a figure to explain PCR testing.}
+
 PCR consists of a series of cycles, each cycle containing a three-step process~\autocite{powledgePCR,garibyanPCR}.
-The first step of the PCR cycle is \emph{denaturation}.
-In denaturation, the DNA is heated to separate the two strands of DNA.
-The second step is \emph{annealing}.
-In annealing, the temperature is lowered to allow primers to bind to the DNA.
+In the first step, \emph{denaturation}, the DNA is heated to separate the two strands of DNA.
+In the second step, \emph{annealing}, the temperature is lowered to allow primers to bind to the DNA.
 Primers are short sequences of DNA that are complementary to the DNA sequence of interest.
 By choosing the primers, the PCR process targets a portion of DNA starting and ending with a specified sequence.
-The final step is \emph{extension}.
-In extension, the temperature is raised to allow the enzyme DNA polymerase to extend the primers and hence replicate the DNA sequence of interest.
+In the final step, \emph{extension}, the temperature is raised to allow the enzyme DNA polymerase to extend the primers and hence replicate the DNA sequence of interest.
 In the course of such a cycle, the amount of DNA doubles.
@@
 
-The Ct value decreases as the amount of DNA present in the original sample increases.
-\todo{cite interpretation of Ct values}
+The Ct value decreases as the amount of DNA present in the original sample increases~\autocite{hanRTPCR}.
 If there is more RNA present, more DNA is created, and hence fewer cycles are required to reach detection.
-The logarithm is because the process is geometric, and hence the number of DNA copies increases linearly on the log-scale each cycle.
-Therefore, the Ct value is proportional to the negative log of the amount of DNA present in the original sample.
-The amount of cDNA produced by reverse transcription is proportional to the amount of RNA present in the original sample.
-Therefore, when testing for SARS-CoV-2, the Ct value is proportional to the negative log of the amount of SARS-CoV-2 RNA present in the original sample.
+As the number of DNA copies doubles each cycle, the Ct value decreases increases linearly on the log-scale each cycle (see \cref{biology-data:fig:ct-calibration}).
+A \emph{standard curve} translates the Ct value to viral load for a given test.
+They are produced by measuring the Ct value using reference samples with known viral load.
+As seen in \cref{biology-data:fig:ct-calibration}, they vary between studies.
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{biology-data/ct-calibration}
+    \caption[Example standard Ct curves]{%
+        Example standard curves, which relate the logarithm of the viral copy number (the number of copies of the target DNA or RNA in the sample) to the measured Ct value.
+        Reproduced from \textcite{hanRTPCR} with permission.
+    }
+    \label{biology-data:fig:ct-calibration}
+\end{figure}
 
-\todo[inline]{Discuss noise in Ct values (batch effects, swabbing process, etc.) see commented-out text below}
-% The noise arises from a variety of sources, including the sampling/swabbing process and batch effects in the RT-PCR process~\autocite{hanRTPCR}.
-% A positive RT-PCR test has a Ct value.
-% Ct values are linearly related to viral load (see \cref{E-biology-data:sec:PCR})
-% A negative RT-PCR test signals either the viral load is too low to be detected, \ie below the limit of detection, or the test is a false negative.
-% Positive RT-PCR tests, especially those with low viral loads, could be false positives, for example caused by contamination.
+\subsection{Detectable individuals} \label{biology-data:sec:detectable}
+
+An important concept throughout this thesis will be that of \emph{detectable} individuals.
+An individual is defined as detectable between the time they start testing positive on a RT-PCR test and the time they stop testing positive on a RT-PCR test.
+It is not possible to directly observe whether an individual is detectable or not due to noise in the RT-PCR test (see \cref{biology-data:sec:PCR}).
+However, viewing the number of RNA genome copies in an individual as a latent process this can then be defined as the time when the number of RNA genome copies is above the limit of detection of the RT-PCR test (defined in \cref{biology-data:sec:measurement-error}).
+While this definition is not perfectly precise, it is clear enough for the purposes it used in this thesis.
+
+\subsection{Measurement error} \label{biology-data:sec:measurement-error}
 
 Like any diagnostic test, PCR tests are not perfect.
-False positives and false negatives can occur.
-These can be quantified in a number of ways.
+The quantitative result (the Ct value) is noisy, and the qualitative result (positive or negative) can be wrong.
+Errors arise from a variety of sources, including the sampling/swabbing process and batch effects in the RT-PCR process~\autocite{hanRTPCR}.
 
@@
 The limit of detection is the lowest concentration of the target RNA material that can be reliably detected using the procedure.
-This is different to the \emph{clinical sensitivity} of the test, the ability of the test to detect a detectable individual (being detectable is defined in \cref{biology-data:sec:natural-history}).
+This is different to the \emph{clinical sensitivity} of the test, the ability of the test to detect a detectable individual.
 Clinical sensitivity can be affected by a large variety of preanalytical factors, such as sample collection (\ie the swabbing process) and sample transportation~\autocite{lippiPotential}.
@@
 A positive result for an undetectable individual is a \emph{false positive}.
+
 False positives are very rare for SARS-CoV-2 RT-PCR tests.
-In summer 2020, the CIS performed 208,730 PCR tests and found only 159 positives, lower-bounding the test specificity (assuming all the positives are false positives) at 99.9\%~\autocite[section 5]{cisMethodsONS}.
+In summer 2020, the CIS performed 208,730 PCR tests and found only 159 positives~\autocite[section 5]{cisMethodsONS}.
+If these were all false positives, then the test specificity would be 99.9\%.
+In reality, it is almost certain that some of these are true positives, and hence this is a lower-bound on the true test senstivity.
 
@@
     \node[below = 1cm of infectious] (detect-label) {Detectable};
-    \node[draw, thick, inner sep=0.3cm, fit=(pos) (infectious) (not-infectious) (detect-label), fill=blue!10,opacity=0.2] (detect-box) {};
+    \node[draw, thick, inner sep=0.3cm, fit=(pos) (infectious) (not-infectious) (detect-label), fill=red!10,opacity=0.3] (detect-box) {};
   \end{tikzpicture}
@@
     Natural history of SARS-CoV-2.
-    (A) A simplified state-based model, which I generally use in this thesis.
+    (A) A simplified state-based representation of the natural history.
     Shading represents states where the individual is detectable by a RT-PCR test.
-    (B) A full model of the natural history, including continuous processes (Reproduced from \textcite{puhachSARSCoV2} with permission; durations are based on averages, there is a large amount of individual variation).
-    PCR (strictly RT-PCR) positive refers to the main diagnostic test used in this thesis, and detect RNA genomes (see \cref{biology-data:sec:PCR} for more details).
+    (B) A fuller representation of the natural history, including the latent processes of the quantity of RNA genome copied and infectious virus within and individual (reproduced from \textcite{puhachSARSCoV2} with permission).
     Ag-RDT (antigen rapid diagnostic test) tests are lateral flow tests, which detect infectious virus rather than RNA, and are not used in this thesis.
@@
 
-The time for each of these stages depends on various features of host and virus characteristics.
+The time for each of these stages depends on various features of host and virus characteristics, and has a large amount of variation.
 In this thesis, I focus on pre-Alpha variants of SARS-CoV-2 in unvaccinated individuals.
 The time from infection to infectiousness is known as the \emph{latent period}.
-The length of the latent period is hard to estimate because neither the time of infection or the start of infectiousness can be observed; estimates for the mean length are generally 3--4 days~\autocite{zhaoEstimating}.
-The time from infection to experiencing symptoms is known as the \emph{incubation period} and is estimated at have a mean length of around 6 days, with some sensitivity to the sampling and analysis technique used~\autocite{wuIncubation,quesadaIncubation,aleneSerial}.
-The incubation period is undefined for asymptomatic individuals of those infected.
+The length of the latent period is hard to estimate because neither the time of infection or the start of infectiousness can be observed.
+The length has been estimated based on careful contact tracing, where transmission times are thought to be known with a reasonable degree of certainty.
+These estimates vary, but suggest a mean length of 3--4 days for pre-Alpha variants~\autocite[and references therein]{zhaoEstimating}.
+The time from infection to experiencing symptoms is known as the \emph{incubation period}; it is undefined for aymptomatic individuals.
+As the symptom onset time (when symptoms start) can be observed directly, this is easier to estimate than the latent period.
+However, the unknown infection time still poses challenges, again requiring contact tracing data.
+Therefore, estimates are sensitive to the choice of data and statistical model.
+The mean length has been estimated as around 6 days~\autocite{wuIncubation,quesadaIncubation,aleneSerial}.
 
-The above description gives three possible ways to define an individual who is infected with SARS-CoV-2: based on their RT-PCR test result, whether they are infectious, or whether they are symptomatic.
-In this thesis, I adopt a definition of infection based on RT-PCR test results.
-This is because it coincides with the data collected on the studies I use (see \cref{biology-data:sec:studies}).
-
-A host's \emph{viral load}, the quantity of viral particles in a host, is the underlying biological process determining many of these changes~\autocite{puhachSARSCoV2}.
+A host's \emph{viral load}, the quantity of viral particles in a host, is the underlying biological process driving the different periods and characteristics of an infection~\autocite{puhachSARSCoV2}.
 Any host's viral load is determined by the interaction of many biological processes, most prominently the virus's replication increasing the viral load and the host's immune response reducing the viral load~\autocite{keVivo}.
@@
 
-\todo[inline]{Not sure if this is the best location for this important paragraph defining the concept of ``detectability''}
-I will refer to individuals between the time they start testing positive on a RT-PCR test and the time they stop testing positive on a RT-PCR test as \emph{detectable}.
-It is not possible to directly observe whether an individual is detectable or not due to noise in the RT-PCR test (see \cref{biology-data:sec:PCR}).
-However, viewing the number of RNA genome copies in an individual as a latent process this can then be defined as the time when the number of RNA genome copies is above the limit of detection of the RT-PCR test.
-While this definition is not perfectly precise, it is clear enough for the purposes it used in this thesis.
-
 \section{Studies used in this thesis} \label{biology-data:sec:studies}
 
-In this thesis, I use data from two studies which recruited participants from the general population in the UK.
+In this section, I expand on the major two studies I use in this thesis (ATACCC and CIS), first introduced in\todo{ref where introducing studies}.
 Unlike many studies of SARS-CoV-2, these studies are not focused on specific groups such as healthcare workers or hospital patients.
 Therefore, the results from these studies are more generalisable to the entire UK population.
-
-\todo[inline]{Some of this might be moved into the introductory chapter motivating each part of the thesis.}
-
-The first study, ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts), performed daily testing of a small number of individuals with short follow-up.
-The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability, but the results are limited because the small sample size and short follow-up means that the tails of these distributions cannot be estimated.
-
-The second study, the CIS (Coronavirus (COVID-19) Infection Survey), performed weekly or monthly testing of a large number of individuals with long follow-up.
-The large sample size and long follow-up allows for reliable estimates of the tails of the distributions of the quantities of interest.
-However, the results are limited because the sparse sampling which means that the viral load dynamics cannot be estimated.
-In particular, the vast majority of detected infections had only one or two positive tests, and hence the viral load trajectory of these infections cannot be estimated.
 
@@
 Contacts were invited to participate if they were 5 years or older and  could be reached within 5 days of the index case's symptom onset date.
-ATACCC was formed of two waves: ATACCC1 enrolled 393 contacts between 13th Sep 2020 and 31st Mar 2021 (when pre-Alpha and Alpha variants were circulating) and ATACCC2 enrolled 345 contacts between 24th May 2021 and 228th Oct 2021 (when the Delta variant was circulating).
+ATACCC enrolled a total of 738 contacts across two waves.
+The first wave enrolled 393 contacts between 13th Sep 2020 and 31st Mar 2021, when pre-Alpha and Alpha variants were circulating.
+The second wave enrolled 345 contacts between 24th May 2021 and 228th Oct 2021, when the Delta variant was circulating.
 Contacts were swabbed daily for up to 20 days, with a variety of tests performed on the swabs; relevant for this thesis is the RT-PCR test for SARS-CoV-2.
-For all positive test results, the viral load was measured using the Ct value and then converted to a viral load using a standard curve.
+For all positive test results, the viral load was measured using the Ct value and then converted to a viral load using a standard curve (see \cref{biology-data:sec:PCR-process}).
 Other data such as demography and symptoms was also collected, although I do not use this information.
@@
 
-\Textcite{hakkiOnset} analyse the viral load of the 57 contacts who tested positive for SARS-CoV-2 and had the growth phase in their viral load measured.
-For 115 contacts with at least one positive RT-PCR test, the growth phase of viral load was not captured; these contacts were excluded from this analysis to ensure all individuals viral load trajectories could be identified.
+Of the 738 contacts enrolled, 172 (23\%) had at least one positive test
+\Textcite{hakkiOnset} analysed the 57 contacts who tested positive for SARS-CoV-2 and had the growth phase in their viral load measured.
+They excluded the remaining 115 contacts with at least one positive RT-PCR test, ATACCC did not capture the phase of the infection with increasing viral load.
+This exclusion was to ensure viral load trajectories could be identified.
 I will revisit and extend this analysis in \cref{E-ATACCC}.
@@
 The CIS~\autocite{CIS} was setup in April 2020 to provide gold-standard measurements of the prevalence of SARS-CoV-2 in the community.
-It is a longitudinal study with a representative sample of private households, this excludes settings such as care homes and student halls.
-It is globally unique in providing a representative, longitudinal, and large-scale study.
+It was a longitudinal study with a representative sample of private households, \ie excluding settings such as care homes and student halls.
+The dataset is globally unique in providing a representative, longitudinal, and large-scale study across the pandemic.
 Initially, it was limited to England, but expanded to cover the whole of the UK in September 2020.
@@
 As well as a swab to test for SARS-CoV-2, participants were asked to complete a questionnaire, and some participants were asked to provide blood samples.
-Real-world issues meant that visits were often not on this precise schedule, and occasionally missed.
+In reality, visits were often not on this precise schedule, and occasionally missed.
 A full description of the study can be found in the study protocol~\autocite{cisProtocol}.
@@
 In this thesis, I focus on the period from Mon 31st Aug 2020 until Sun 24th Jan 2021 (inclusive).
-This period is chosen because there were very few positives in CIS in summer 2020 due to low prevalence and a smaller sample size, and ends when vaccination crossed 10\% of England's population.
-Vaccination complicates the dynamics and immune response, which would require more complex models which are more specific to SARS-CoV-2 beyond the scope of this thesis, which primarily intends to show the potential of a study such as CIS in general.
-\todo{make sure I justify the period choice in the intro when introducing the thesis's scope and aims}
+\todo{choice of period might be better in intro}
+This period starts when CIS was expanded, and ends when vaccination crossed 10\% of England's population.
+Vaccination complicates the dynamics of transmission and viral load.
+Analysis of data including vaccinated individuals requires more complex models, with specific adaptations to SARS-CoV-2.
+These are beyond the scope of this thesis, which primarily intends to show the potential of a study such as CIS.
 
@@
 
-\subsubsection{Estimating positivity rate} \label{biology-data:sec:positivity}
+\subsubsection{Positivity} \label{biology-data:sec:positivity}
 
-The \emph{positivity rate} is the proportion of people in a given population that (if they were tested) would test positive for SARS-CoV-2 at a defined point in time; in this context test means an RT-PCR test~\autocite{cisMethodsONS}.
+The main result reported by the study was what they defined as the \emph{positivity rate}, the proportion of people in a given (sub)population that (if they were tested using RT-PCR) would test positive for SARS-CoV-2 at a defined point in time~\autocite{cisMethodsONS}.
 A notable feature of this definition is that it explicitly does not correct for false positive or negative results.
@@
 
-If the CIS was truly a random sample, an unbiased estimator of the positivity rate would be the proportion of positive tests in the sample; I refer to this as the uncorrected positivity rate.
+\subsubsection{Strata}
+
+The geographic areas I will adopt are the nine standard regions of England (International Territorial Level 1, previously Nomenclature of Territorial Units for Statistics 1); these are: North East, North West, Yorkshire and the Humber (commonly abbreviated to Yorkshire), East Midlands, West Midlands, East, London, South East, and South West~\autoref{onsRegions}.
+These are the geographic areas used throughout the major CIS analyses in England, although sometimes more granular areas were defined~\autocite[e.g.][]{pouwelsMRPvaccination,pouwelsCommunity,cisMethodsONS,houseInferring,walkerTracking}.
+
+The age groups used have varied slightly for different CIS analyses~\autocite[e.g.][]{pouwelsMRPvaccination,pouwelsCommunity,cisMethodsONS,houseInferring,walkerTracking}.
+I will use the age groups: 0--10 (pre-school and primary school), 11--15 (compulsory secondary school), 16--24 (adolescents and young adults who may remain in formal education), 25--49 (younger working age adults), 50--69 (older working age adults), and 70+ (elderly).
+
+\todo[inline]{Add ethnicity and other strata used in ch 7}
+
+\subsubsection{Adjustment for non-response and data sparsity}
+
+If the CIS was truly a random sample, an unbiased estimator of the positivity rate would be the proportion of positive tests in the sample; I refer to this estimator as the unadjusted positivity rate.
 However, in practice, two issues arise with this.
@@
 
-The statistical preferred method for estimating the positivity rate is MRP (multi-level regression and poststratification)~\autocite{cisMethodsONS,pouwelsCommunity}.
+The preferred statistical method for estimating the positivity rate is MRP (multi-level regression and poststratification)~\autocite{cisMethodsONS,pouwelsCommunity}.
 The poststratification step is used to correct for non-response bias.
@@
 
-The outputs of this model compared to the raw positivity rate are shown in \cref{biology-data:fig:CIS-positivity}.
+The outputs of this model compared to the unadjusted positivity rate are shown in \cref{biology-data:fig:CIS-positivity}.
 Both show similar trends.
@@
     CIS positivity rate in the period considered in this thesis (Sep 2020 until Jan 2021).
-    Dots show raw proportion of non-void tests that are positive on each day (days with fewer than 100 tests excluded for readability).
+    Dots show the unadjusted positivity rate (days with fewer than 100 tests are excluded for readability).
     Ribbon shows modelled 95\% CrI for the positivity, corrected for non-response bias and smoothed over time (using the methodology described in \cref{E-backcalc:sec:methods}).
@@
 
-Interpreting positivity, or prevalence, directly is difficult because it is a delayed and smoothed measure of incidence (see \cref{E-inc-prev:sec:prevalence-process}).
-The major aim of this thesis is the ability to produce estimates of incidence, which can be interpreted in terms of population behaviour.
-
 As well as estimating the positivity rate for the whole of England (and the devolved nations), the CIS also estimates the positivity rate for various subpopulations, in particular geographic areas and age groups.
@@
 
-The geographic areas I will adopt are the nine standard regions of England (International Territorial Level 1, previously Nomenclature of Territorial Units for Statistics 1); these are: North East, North West, Yorkshire and the Humber (commonly abbreviated to Yorkshire), East Midlands, West Midlands, East, London, South East, and South West~\autoref{onsRegions}.
-These are the geographic areas used throughout the major CIS analyses in England, although sometimes more granular areas were defined~\autocite[e.g.][]{pouwelsMRPvaccination,pouwelsCommunity,cisMethodsONS,houseInferring,walkerTracking}.
-
-The age groups used have varied slightly for different CIS analyses~\autocite[e.g.][]{pouwelsMRPvaccination,pouwelsCommunity,cisMethodsONS,houseInferring,walkerTracking}.
-I will use the age groups: 0--10 (pre-school and primary school), 11--15 (compulsory secondary school), 16--24 (adolescents and young adults who may remain in formal education), 25--49 (younger working age adults), 50--69 (older working age adults), and 70+ (elderly).
+\subsubsection{Clustering}
 
@@
 
-The algorithm determines whether a new positive tests, at time $t$, is part of the previous infection episode in the same individual, where that previous infection episode had its first positive test at $t_p$.
+The algorithm determines whether a new positive test, at time $t$, is part of the previous infection episode in the same individual.
 If the individual has no previous positive tests, the new positive test is always the start of a new infection episode.
-Otherwise, the test at $t$, in the pre-Omicron era (prior to December 2021), is defined as the start of a new infection episode if any of the following apply.
+Otherwise, denote the first positive test associated with that previous infection episode as $t_p$, and the number of negative tests immediately prior to the test at $t$ as $n$.
+The test at $t$, in the pre-Omicron era (prior to December 2021), is defined as the start of a new infection episode if any of the following apply.
 \begin{enumerate}
@@
 \todo[inline]{%
-    From chapter 7 add info on: how this is aggregated, SRS limitations, definitions for each episode (sketched below).
+    From chapter 7 add info on: how this is aggregated/stratified/etc, SRS computational limitations, definitions for each episode (sketched below).
 }

### 2024-02-19 introduction.tex
proofing ch 4
resolve all Overleaf comments

@@
 
-- Natural history of SARS-CoV-2 infection
-	- Stylised graphic of phases: infection, pre-infection positive, infectious positive, post-infectious positive, negative
-	- How it spreads
-- Why is incidence and prevalence estimation challenging?
-	- Asymptomatic infection
-	- Ascertainment bias
-- How do we estimate prevalence
-	- CIS and REACT
-- Issues with current incidence estimation
+% - Natural history of SARS-CoV-2 infection
+% 	- Stylised graphic of phases: infection, pre-infection positive, infectious positive, post-infectious positive, negative
+% 	- How it spreads
+% - Why is incidence and prevalence estimation challenging?
+% 	- Asymptomatic infection
+% 	- Ascertainment bias
+% - How do we estimate prevalence
+% 	- CIS and REACT
+% - Issues with current incidence estimation
 
@@
 \end{itemize}
+
+\subsection{Modelling}
+
+\begin{itemize}
+    \item SPI-M
+    \item Literature on mechanistic models using REACT and CIS I have previously circulated.
+    \item \textcite{mccabeCISincidence,abbottCISincidence} on backcalculation.
+    \item Also Taffe?
+\end{itemize}
+
+\subsection{Surveillance data}
+
+The first study, ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts), performed daily testing of a small number of individuals with short follow-up.
+The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability, but the results are limited because the small sample size and short follow-up means that the tails of these distributions cannot be estimated.
+
+The second study, the CIS (Coronavirus (COVID-19) Infection Survey), performed weekly or monthly testing of a large number of individuals with long follow-up.
+The large sample size and long follow-up allows for reliable estimates of the tails of the distributions of the quantities of interest.
+However, the results are limited because the sparse sampling which means that the viral load dynamics cannot be estimated.
+In particular, the vast majority of detected infections had only one or two positive tests, and hence the viral load trajectory of these infections cannot be estimated.
 
@@
     \item Matrices and vectors are bold and have elements with the same unbolded letter, \eg $\vec{x} = [x_1, x_2, \dots]^T$
+    \item $\cdot$ denotes element-wise multiplication
 \end{itemize}

### 2024-02-19 incidence-prevalence.tex
Ch 2 Overleaf comments, except ch 7 ones

@@
 \emph{Prevalence} of a disease on day $t$, is the proportion of people with the disease on that day~\autocite[90]{lashModern}.
-Within this thesis, I define an individual as having SARS-CoV-2 on day $t$ if, had they been swabbed on day $t$, that swab would return a positive RT-PCR test.
-I refer to these individuals as \emph{prevalent} individuals.
+I refer to individuals with the disease as \emph{prevalent} individuals.
+Whether an individual is prevalent is often expensive or error-prone to measure, therefore, it will often be regarded as a latent quantity.
+Furthermore, what is meant by ``having'' a disease can depend on context; for example, it could be the presence of symptoms, the ability to transmit, or detectable status (the difference between these was discussed in \cref{E-biology-data:sec:natural-history})
+Within this thesis, I define an individual as prevalent on day $t$ if they are detectable (detectable is defined in \cref{E-biology-data:sec:detectable}).
 I assume that an individual's prevalent status cannot change over the course of a day.

### 2024-02-19 biology-data.tex
move MRP intro to ch 2

@@
 These rates of response raise the possibility of non-response bias.
-In particular, some demographic groups, such as young adults and individuals of non-white ethnicity, were underrepresented among those who responded~\autocite{pouwelsCommunity}.
+In particular, some demographic groups, such as young adults and individuals of non-White ethnicity, were underrepresented among those who responded~\autocite{pouwelsCommunity}.
 
@@
 
-\subsubsection{Positivity} \label{biology-data:sec:positivity}
-
 The main result reported by the study was what they defined as the \emph{positivity rate}, the proportion of people in a given (sub)population that (if they were tested using RT-PCR) would test positive for SARS-CoV-2 at a defined point in time~\autocite{cisMethodsONS}.
@@
 
-The geographic areas I will adopt are the nine standard regions of England (International Territorial Level 1, previously Nomenclature of Territorial Units for Statistics 1); these are: North East, North West, Yorkshire and the Humber (commonly abbreviated to Yorkshire), East Midlands, West Midlands, East, London, South East, and South West~\autoref{onsRegions}.
-These are the geographic areas used throughout the major CIS analyses in England, although sometimes more granular areas were defined~\autocite[e.g.][]{pouwelsMRPvaccination,pouwelsCommunity,cisMethodsONS,houseInferring,walkerTracking}.
+In estimating positivity, I will divide the population into strata based on age group, sex, ethnicity, and geographic area.
+I will report results aggregated to geographic regions and age groups, as has been done previously.
+For the purposes of both fitting and reporting results, test results are aggregated to daily counts for each stratum.
 
@@
 
-\todo[inline]{Add ethnicity and other strata used in ch 7}
+I use two levels of ethnic grouping.
+Broad groupings (White and non-White), because the White population makes up a large majority of the UK.
+I also use more granular ethnicities (Asian, Black, Mixed, White, and Other).
 
-\subsubsection{Adjustment for non-response and data sparsity}
+The geographic areas I will adopt are the nine standard regions of England (International Territorial Level 1, previously Nomenclature of Territorial Units for Statistics 1); these are: North East, North West, Yorkshire and the Humber (commonly abbreviated to Yorkshire), East Midlands, West Midlands, East, London, South East, and South West~\autocite{onsRegions}.
+These are the geographic areas used throughout the major CIS analyses in England, although sometimes more granular areas were defined~\autocite[e.g.][]{pouwelsMRPvaccination,pouwelsCommunity,cisMethodsONS,houseInferring,walkerTracking}.
+
+\subsubsection{Adjustment for non-response and data sparsity} \label{biology-data:sec:MRP}
 
@@
 The preferred statistical method for estimating the positivity rate is MRP (multi-level regression and poststratification)~\autocite{cisMethodsONS,pouwelsCommunity}.
+MRP is a statistical methodology developed to correct political opinion polling suffering from response bias~\autocite{gelmanMRT,gelmanPoststratication,parkMRP}.
+MRP is a two-step process.
+In this context, the first step, multilevel regression, consists of using a Bayesian multilevel logistic regression model to predict RT-PCR test results in the sample based on demographic covariates.
+To address the low number of positives, MRP uses a hierarchical Bayesian model to borrow information across time and other covariates.
+The second step, poststratification, is using the posterior distribution of the parameters is used to predict the population prevalence in each stratum.
+These are combined by taking a weighted average across the strata, where a stratum's weight is equal to its (known) population size.
 The poststratification step is used to correct for non-response bias.
-To address the low number of positives, MRP uses a hierarchical Bayesian model to borrow information across time and other covariates.
-I explain the details of this model, and how I use it for estimating incidence in \cref{E-backcalc:sec:methods}
 
-The outputs of this model compared to the unadjusted positivity rate are shown in \cref{biology-data:fig:CIS-positivity}.
+I modified the regression component of a pre-existing version of a MRP model for CIS~\autocite{pouwelsMRPvaccination}, as recommended by that model's author for the specific context of this thesis~\citePersonalComms{Koen Pouwels}.
+As in prior work~\autocite{pouwelsMRPvaccination}, I used INLA for inference, implemented in R-INLA (see \cref{E-transmission:sec:INLA}).
+For a day $t$, age group $a$, ethnicity $e$, and sex $s$ the regression model is:
+\begin{align}
+    \logit(p_{taes}) &= \beta + \beta_s + \beta_{w(e)} + u_a + v_e + w_t + l_{a,t} \\
+    y_{taes} &\dist \BB(n_{taes}, p_{taes}, \rho).
+\end{align}
+The terms are as follows.
+\begin{itemize}
+    \item $n_{taes}$ and $y_{taes}$ are the number of individuals tested and the number of positive test results on day $t$ in individuals of age group $a$, ethnicity $e$, and sex $s$.
+I use a beta-binomial distribution, as discussed in \cref{biology-data:sec:clustering}.
+    \item $\beta$ is a global intercept term with the prior $\beta \dist \N(0, 1/0.001^2)$.
+    \item $\beta_s$ is a fixed effect for sex, with $\beta_s = 0$ for males and estimated for females.
+    \item $\beta_{w(e)}$ is a fixed effect for White or non-White ethnicity.
+$\beta_{w(e)} = 0$ if $e$ corresponds to White ethnicity and $\beta_{w(e)} = \beta_{n}$ otherwise (\ie non-White ethnic groups).% $\beta_n$ and $\beta_f$ have the default INLA prior for fixed effects\todo{check what this is}.
+    \item $u_a$ and $v_e$ are independent Gaussian random effects for age and ethnic group respectively.
+Their standard deviation have a penalised complexity prior~\autocite{simpsonPenalising} with a 10\% probability of being greater than 1.
+    \item $w_t$ is a second-order random walk, that is $w_t = 2w_{t-1} - w_{t-2} + \epsilon_t$ where $\epsilon_t \dist \N(0, \sigma_w^2)$.
+    \item $l_{a,t}$ is an age-group specific second-order random walk. That is $l_{a,t} = 2l_{a,t-1} - l_{a,t-2} + \epsilon_{a,t}$ where $\epsilon_{a,t} \dist \N(0, \sigma_l^2)$.
+\end{itemize}
+All parameters have the default R-INLA priors (see \cref{E-transmission:sec:INLA:priors}), except where stated above.
+For these parameters, the large amount of data means that the prior has little influence on the posterior distribution.
+
+I fitted the model separately and independently for each region, implicitly adding all interactions between region and the other variables.
+This step assumes that the estimates for each region can be considered independent, which is reasonable because the regions are large with many tests conducted within each.
+
+The final step is poststratifying the incidence proportion to the CIS target population and subpopulations of interest.
+To start, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution (using the procedure described in \cref{transmission:sec:INLA:posterior}).
+For each posterior draw and day, the incidence proportion in the (sub)population is the weighted average of the incidence proportion in each stratum, where the weights are the number of individuals in that stratum.
+The population numbers for each stratum are provided by the Office for National Statistics.
+
+\subsubsection{Clustering} \label{biology-data:sec:clustering}
+
+The household-based design of the CIS means that the data is clustered.
+In particular, individuals in the same household are likely to transmit to each other and hence have correlated test results.
+Other behavioural or genetic factors could also contribute to individuals in the same household being more similar than randomly selected individuals in the population.
+The positivity analyses use a beta-binomial likelihood (defined in \cref{E-distributions}) to account for this.
+
+The beta-binomial distribution is overdispersed relative to the binomial distribution.
+Overdispersion allows for household clustering and other overdispersion in the data, which is common in epidemiological data~\autocite{griffithsBBD}.
+The parameterisation of the beta-binomial that I use (detailed in \cref{E-distributions}) has an overdispersion parameter $\rho$, regarded as a nuisance parameter in this context.
+$\rho=0$ means the beta-binomial coincides with the binomial distribution.
+As $\rho \to \infty$, the distribution tends towards the uniform distribution on $[0, 1]$~\autocite{hughesUsing}.
+
+\subsubsection{Positivity results}
+
+The estimates from this model compared to the unadjusted positivity rate are shown in \cref{biology-data:fig:CIS-positivity}.
 Both show similar trends.
@@
 \end{figure}
-
-As well as estimating the positivity rate for the whole of England (and the devolved nations), the CIS also estimates the positivity rate for various subpopulations, in particular geographic areas and age groups.
-The exact specification of these have varied over time.
-
-\subsubsection{Clustering}
-
-The household-based design of the CIS means that the data is clustered.
-In particular, individuals in the same household are likely to transmit to each other and hence have correlated test results.
-Other behavioural or genetic factors could also contribute to individuals in the same household being more similar than randomly selected individuals in the population.
-The positivity analyses use a beta-binomial likelihood (defined in \cref{E-distributions}) to account for this.
-
-The beta-binomial distribution is overdispersed relative to the binomial distribution.
-Overdispersion allows for household clustering\todo{cite beta-binomial as a model for clustering} and other overdispersion in the data, which is common in epidemiological data\todo{cite that epi data is often overdispersed}.
-The parameterisation of the beta-binomial that I use (detailed in \cref{E-distributions}) has an overdispersion parameter $\rho$, regarded as a nuisance parameter in this context.
-$\rho=0$ means the beta-binomial coincides with the binomial distribution.
-As $\rho \to \infty$, the distribution tends towards the uniform distribution on $[0, 1]$~\autocite{hughesUsing}.
 

### 2024-02-19 transmission.tex
move MRP intro to ch 2

@@
 
-Estimation of prevalence from the CIS data has been carried out through a well-established methodology, used throughout the pandemic~\autocite{pouwelsCommunity,cisMethodsONS}.
-The method is based on MRP, a statistical methodology developed to correct political opinion polling suffering from response bias~\autocite{gelmanMRT,gelmanPoststratication,parkMRP}.
-MRP is a two-step process.
-In this context, the first step, multilevel regression, consists of using a Bayesian multilevel logistic regression model is to predict RT-PCR test results in the sample based on demographic covariates.
-The second step, poststratification, is using the posterior distribution of the parameters is used to predict the population prevalence in each stratum.
-These are combined by taking a weighted average across the strata, where a stratum's weight is equal to its (known) population size.
+Estimation of prevalence from the CIS data was carried out using MRP throughout the pandemic (see \cref{E-biology-data:sec:MRP}).
 In this section, I propose extending this process by transforming the posterior distribution of prevalence in each stratum to a posterior distribution of incidence before post-stratifying.
 
-As in prior work, I parameterise the regression model in terms of prevalence, not incidence proportion.
+I keep the same logistic regression model in terms of prevalence, transforming its posterior distribution of prevalence to a posterior distribution of incidence.
+This results in a two-step process: first, the same inference process described in \cref{E-biology-data:sec:MRP} is used to estimate prevalence; second, the prevalence is transformed to incidence.
+An alternative would be an inference procedure that estimates incidence proportion, fit directly to the RT-PCR test results.
 This is required to fulfil several desirable features of the model.
@@
 
-Next, I explain the modification of the previous logistic regression model, and then how to extend the procedure to calculate incidence.
-Specifically, I modified the regression compomenet of a pre-existing version of a MRP model for CIS~\autocite{pouwelsMRPvaccination}, as recommended by that model's author for this situation~\citePersonalComms{Koen Pouwels}.
-For a day $t$, age group $a$, ethnicity $e$, and sex $s$ assume that:
-\begin{align}
-    \logit(p_{taes}) &= \beta + \beta_s + \beta_{w(e)} + u_a + v_e + w_t + l_{a,t} \\
-    y_{taes} &\dist \BB(n_{taes}, p_{taes}, \rho)
-\end{align}
-The terms are as follows.
-\begin{itemize}
-    \item $n_{taes}$ and $y_{taes}$ are the number of individuals tested and the number of positive test results on day $t$ in individuals of age group $a$, ethnicity $e$, and sex $s$.
-I use a beta-binomial distribution, as discussed in \cref{E-biology-data:sec:positivity}.
-    \item $\beta$ is a global intercept term with the prior $\beta \dist \N(0, 1/0.001^2)$.
-    \item $\beta_s$ is a fixed effect for sex, with $\beta_s = 0$ for males and estimated for females.
-    \item $\beta_{w(e)}$ is a fixed effect for white or non-white ethnicity.
-$\beta_{w(e)} = 0$ if $e$ corresponds to white ethnicity and $\beta_{w(e)} = \beta_{n}$ otherwise (\ie non-white ethnic groups).% $\beta_n$ and $\beta_f$ have the default INLA prior for fixed effects\todo{check what this is}.
-    \item $u_a$ and $v_e$ are independent Gaussian random effects for age and ethnic group respectively.
-Their standard deviation have a penalised complexity prior~\autocite{simpsonPenalising} with a 10\% probability of being greater than 1.
-    \item $w_t$ is a second-order random walk, that is $w_t = 2w_{t-1} - w_{t-2} + \epsilon_t$ where $\epsilon_t \dist \N(0, \sigma_w^2)$.
-    \item $l_{a,t}$ is an age-group specific second-order random walk. That is $l_{a,t} = 2l_{a,t-1} - l_{a,t-2} + \epsilon_{a,t}$ where $\epsilon_{a,t} \dist \N(0, \sigma_l^2)$.
-\end{itemize}
-All parameters have the default R-INLA priors (see \cref{transmission:sec:INLA:priors}), except where stated above.
-For these parameters, the large amount of data means that the prior has little influence on the posterior distribution.
-
-I fitted the model separately and independently for each region, implicitly adding all interactions between region and the other variables.
-This step assumes that the estimates for each region can be considered independent, which is reasonable because the regions are large with many tests conducted within each.
-
-Having fitted the prevalence model, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution (using the procedure described in \cref{transmission:sec:INLA:posterior}).
+Next, I explain how I extended the MRP model introduced in \cref{E-biology-data:sec:MRP} to calculate incidence.
+I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution (using the procedure described in \cref{transmission:sec:INLA:posterior}).
 For each stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
@@
 
-The final step is poststratifying the incidence proportion to the CIS target population and subpopulations of interest.
-For this, I start with the 500 samples of the approximate posterior distribution of the incidence.
-For each posterior draw and day, the incidence proportion in the (sub)population is the weighted average of the incidence proportion in each stratum, where the weights are the proportion of the population in each stratum.
-The population numbers for each stratum are provided by the Office for National Statistics.
+Once the incidence proportion in each stratum has been calculated using this procedure, I poststratify it to the CIS target population and subpopulations of interest.
+This is done using the methodology explained in \cref{E-biology-data:sec:MRP}.
 

### 2024-02-19 ATACCC.tex
proof ch 4

@@
 The observation model mathematically formalises how the RT-PCR testing process produces noisy observations of the latent viral load.
-The observation model includes the possibility of false positive and negatives, and the limit of detection of the RT-PCR test (see \cref{E-biology-data:sec:PCR}).
+The observation model includes the possibility of false positive and negatives, and the limit of detection of the RT-PCR test (\cref{E-biology-data:sec:PCR}).
 % The limit of detection means that viral loads below a certain threshold are not detected by the RT-PCR test.
@@
 
-The noisy observations of viral log load from RT-PCR testing are modelled by assuming each measurement of log viral load has an iid error distributed $\N(0, \sigma_v^2)$.
+The noisy RT-PCR observations of viral log load are modelled by assuming each measurement of log viral load has an iid error distributed $\N(0, \sigma_v^2)$.
 The measurements are left-censored at the limit of detection, $\vlod$, meaning that any observation which would give a log viral load of less than $\vlod$ is instead observed as negative.
@@
 The second group is vaccinated and all were infected by the Delta variant.
-My focus is on the first group since they best correspond to the infections in the time period used throughout this thesis.
+My focus is on the first group since they most closely correspond to the infections in the time period used throughout this thesis.
 
@@
 
-\textcite{hakkiOnset} uses weakly informative priors, similar to those previously used~\autocite{singanayagamCommunity}.
-These are given in \cref{ATACCC:table:hakki-priors}.
+\textcite{hakkiOnset} uses weakly informative priors, similar to those previously used~\autocite{singanayagamCommunity}, see \cref{ATACCC:table:hakki-priors}.
 I reran \textcite{hakkiOnset}'s analysis using R~4.3.1~\autocite{R-4-3-1}.

### 2024-02-19 ATACCC.tex
ch5 proof

@@
 Two major approaches have been taken to model SARS-CoV-2 viral load trajectories, both based on well-well-established methods within biostatisics.
-The first approach is an ODE-based approach uses parameters representing biological properties of the virus and host~\autocite[e.g.:][]{ejimaEstimation,keVivo,kimQuantitative,goncalvesTiming,perelsonMechanistic}.
+The first approach is an ODE-based approach uses parameters representing biological properties of the virus and host~\autocite[e.g.][]{ejimaEstimation,keVivo,kimQuantitative,goncalvesTiming,perelsonMechanistic}.
 The model here is an application of a standard model for viral dynamics, used for many diseases~\autocite{perelsonMechanistic,hernandez-vargasInhost}.

### 2024-02-19 cis-perfect-testing.tex
ch5 proof

@@
 
-\section{Definitions}
-\todo[inline]{These will be defined in the introductory chapters, but are here so that this draft is understandable}
-\begin{itemize}
-\item
-  We assume that, for all episodes $i$, $D_i$ is iid and independent
-  of the time of the infection. Define the survival function
-  $\prob(D_i \geq t \mid B_i = b, \theta) = \prob(D_i \geq t \mid \theta) = S_\theta(t)$,
-  where $\theta$ are the parameters controlling the survival
-  distribution (the discussion in this document is valid regardless of
-  the model specified for $S_\theta$ and hence we consider $\theta$
-  as an arbitrary vector of parameters).
-\item
-  The beginning of episode $i$ is known to occur in the interval
-  $[l_i^{(b)}, r_i^{(b)}]$, and similarly for the end of the infection
-  in $[l_i^{(e)}, r_i^{(e)}]$.
-\end{itemize}
+% \section{Definitions}
+% \todo[inline]{These will be defined in the introductory chapters, but are here so that this draft is understandable}
+% \begin{itemize}
+% \item
+%   We assume that, for all episodes $i$, $D_i$ is iid and independent
+%   of the time of the infection. Define the survival function
+%   $\prob(D_i \geq t \mid B_i = b, \theta) = \prob(D_i \geq t \mid \theta) = S_\theta(t)$,
+%   where $\theta$ are the parameters controlling the survival
+%   distribution (the discussion in this document is valid regardless of
+%   the model specified for $S_\theta$ and hence we consider $\theta$
+%   as an arbitrary vector of parameters).
+% \item
+%   The beginning of episode $i$ is known to occur in the interval
+%   $[l_i^{(b)}, r_i^{(b)}]$, and similarly for the end of the infection
+%   in $[l_i^{(e)}, r_i^{(e)}]$.
+% \end{itemize}
 
@@
 Therefore, estimates of the mean are sensitive to the tail of the duration distribution.
+
 In \cref{E-ATACCC} I estimated the duration distribution from the ATACCC dataset.
-However, this study suffered from a small sample size and only 20 days of follow-up.
-Therefore, model assumptions drive the size of the estimated tails.
+However, this study suffered from a small sample size and only 20 days of follow-up, enabling estimation of the tail.
+Therefore, model assumptions drive the size of the estimated tail.
+
 In this chapter, I develop a method to estimate the duration distribution using CIS data.
@@
 
-I take a Bayesian survival analysis approach to model the coarse sampling of the CIS (see \cref{perf-test:sec:problem}).
-I use a simulation study (see \cref{perf-test:sec:simulation-study}) to evaluate a range of priors for the survival distribution (described in \cref{perf-test:sec:parameters-priors}).
+I take a Bayesian survival analysis approach to model the coarse sampling of the CIS (\cref{perf-test:sec:problem}).
+I use a simulation study (\cref{perf-test:sec:simulation-study}) to evaluate a range of priors for the survival distribution (\cref{perf-test:sec:parameters-priors}).
 These include a prior to combine the information in the CIS with the results of \cref{E-ATACCC}.
@@
 However, the method must be able to account for false negatives, discussed in \cref{E-imperf-test}.
+
+\todo[inline]{make sure following paragraph makes sense in context (copy/pasted here)}
+In the rest of this and the following chapter, I use survival analysis rather than the biomarker approach taken in \cref{E-ATACCC}.
+This choice is motivated by computational efficiency and limitations of the CIS's study design.
+The biomarker approach is computational challenging because the number of random effects scales with the number of detected infection episodes.
+There are many detected episodes in CIS; in the period I will consider, around 50 times more than in ATACCC.
+Therefore, the biomarker approach would be computationally challenging.
+The computational challenge is compounded by the data being stored in a TRE, which has limited computational power and software.
+The issue with random effects is made worse by the CIS's study design.
+Infrequent testing means that most detected episodes have only a single positive test.
+Hence, there is no information on the viral load trajectory and the random effects are unidentifiable.
 
@@
 CIS has many undetected infections.
-This is because the interval between tests in the same individual (up to 28 days, see\todo{ref section on the survey design}) is longer than most infections (median around 15 days, see\todo{ref ATACCC chapter}).
+This is because the interval between tests in the same individual (up to 28 days, see\todo{ref section on the survey design}) is longer than most infections (median 16 days, see \cref{E-ATACCC:sec:results}).
 \begin{figure}
@@
 \subsection{Double interval censoring}
+\todo{using double interval censored before its defined}
 
@@
 
-In the rest of this and the following chapter, I use survival analysis rather than the biomarker approach taken in \cref{E-ATACCC}.
-This choice is motivated by computational efficiency and limitations of the CIS's study design.
-The biomarker approach is computational challenging because the number of random effects scales with the number of detected infection episodes.
-There are many detected episodes in CIS; in the period I will consider, around 50 times more than in ATACCC.
-Therefore, the biomarker approach would be computationally challenging.
-The computational challenge is compounded by the data being stored in a TRE, which has limited computational power and software.
-The issue with random effects is made worse by the CIS's study design.
-Infrequent testing means that most detected episodes have only a single positive test.
-Hence, there is no information on the viral load trajectory and the random effects are unidentifiable.
-
 \section{Survival analysis}
@@
 In standard left truncation, nothing is known about the missed events.
-I incorporate the additional information into the analysis.
-I refer to these events as ``undetected'' to emphasise this difference.
+Here, I will incorporate the additional information into the analysis.
+I refer to these events as ``undetected'', rather than truncated, to emphasise this difference.
 
@@
 This is especially true within the human biostatistical literature.
-The literature does include theoretical frameworks which include the double censored and truncated case~\autocite{turnbullEmpirical,dempsterMaximum}.
-The frameworks have only been applied when the terminating event was either uncensored or right censored~\autocite{sunEmpirical,bacchettiNonparametric}.
+The literature does include theoretical frameworks which include the double censored and truncated case~\autocite{turnbullEmpirical,dempsterMaximum}, yet these frameworks have only been applied when the terminating event was either uncensored or right censored~\autocite{sunEmpirical,bacchettiNonparametric}.
 These studies \autocite[and elsewhere, e.g.:][]{shenNonparametric} take a conditional likelihood approach.
@@
 In CIS, infection episodes are more likely to be detected when an individual is in the weekly testing phase.
-Therefore, the probability of an infection episode being detected conditional on being infected is higher than the overall probability of being detected.\todo{this is a bit unclear, try and rephrase?}
+Therefore, the probability of an infection episode being detected conditional on being infected is higher than the overall probability of being detected.
 
@@
 I use Bayesian inference for two reasons.
-First, frequentist methods for estimating the parameter's variance and covariance are poorly developed.
-Commonly used methods are based on the Fisher information matrix or resampling.
-Neither have theoretical justification~\autocite[230]{sunAnalysis}\todo{Try and understand why this is the case}.
+First, frequentist estimators are poorly developed and their properties (\eg consistency and convergence rate) are poorly understood~\autocite{sunAnalysis,dengNonparametric}.
 Second, a prior can include information from previous studies, such as from \cref{E-ATACCC}.
@@
 
-Bayesian methods been used for nest survival before~\autocite{heBayesiana,heBayesian,caoModeling}, augmenting the data with the unobserved times of infection.
+Bayesian methods have been used for nest survival before~\autocite{heBayesiana,heBayesian,caoModeling} by augmenting the data with the unobserved times of infection.
 The augmentation creates a parameter per detected infection.
-This increases the dimensionality of the problem greatly, making it computationally infeasible here.
+This increases the dimensionality of the problem greatly, making it computationally infeasible within the SRS.
 I develop a novel Bayesian framework without explicit data augmentation.
@@
 
-Integrating over the number of missed episodes is analytically tractable.
+Integrating over the number of missed episodes is possible analytically.
 This makes inference tractable within modern, high-performance Bayesian software such as Stan.
@@
   This space is empty for $i \notin \set{D}$.
-  Otherwise, define $l_i^{(b)}$ and $r_i^{(b)}$ as the first and last days $b_i$ could have occurred.
-  $l_i^{(b)}$ is the day after the last negative test prior to the first positive test in the episode.
-  $r_i^{(b)}$ is the day of the first positive test.
-  Similarly, define $l_i^{(e)}$ and $r_i^{(e)}$ as the first and last days $e_i$ could have occurred.
-  If $i \in \set{D}$ then $n_{ia} =1$ admissible episodes occurred in individuals with test schedule $\sched_{i}$, otherwise $n_{ia} = 0$.
+  If $i \in \set{D}$ then $n_{ia} =1$ admissible episodes occurred in an individual with test schedule $\sched_{i}$, otherwise $n_{ia} = 0$.
   Conditional on the episode occurring in individuals with test schedule $\sched_i$, it is admissible with probability $p_{ia} = \prob((b, e) \in \alpha_i \mid \theta)$.
@@
 
-The detected region, $\Omega_i = \alpha_i \cup \beta_i$, for test schedule $\sched_i$ is the region where we could have observed episodes. 
-The notation is for consistency with \textcite{heiseyModelling}.
-$\Omega_i$ is the complement of $\Omega^C_i$.
+% The detected region, $\Omega_i = \alpha_i \cup \beta_i$, for test schedule $\sched_i$ is the region where we could have observed episodes. 
+% The notation is for consistency with \textcite{heiseyModelling}.
+% $\Omega_i$ is the complement of $\Omega^C_i$.
 
@@
 These three classes span all possible episodes and are mutually exclusive.
-Therefore, these definitions are consistent with the previous definition of $\ntot$.
-That is, $\ntot = \ndet + \nnodet + n_{i\bar{a}} = \sum_{i=1}^{\ndet} n_{ia} + \sum_{i=1}^{\ntot} n_{iu} + 0$.
+Hence, $\ntot = \ndet + \nnodet + n_{i\bar{a}} = \sum_{i=1}^{\ndet} n_{ia} + \sum_{i=1}^{\ntot} n_{iu} + 0$.
 Furthermore, for all $i$, $p_{ia} + p_{iu} + p_{i\bar{a}} = 1$.
@@
 In forming the probability $p_{ia}$, the data is used.
-This is not quite correct because forming a likelihood should not depend on the data.
+This is not quite correct because the cell probabilities in the likelihood should not depend on the data.
 However, an alternative where each combination of infection and recovery times is considered its own class is possible.
@@
 
-The posterior we are interested in is $p(\theta \mid \na)$.
-The posterior density can be written as follows.
+The posterior density we are interested in is $p(\theta \mid \na)$.
+It can be written as follows.
 \begin{align}
@@
 
-Now I derive an analytical solution to $\sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under the prior $\ntot \dist \NBc(\mu, r)$.
-I assume that $\theta$, which are only the parameters of the survival distribution, is a priori independent of $\ntot$, the number of infections that occurred in the cohort.
+To start, I derive an analytical solution to $\sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under the prior $\ntot \dist \NBc(\mu, r)$.
+I assume that $\theta$, which are only the parameters of the survival distribution, has a prior independent of $\ntot$, the number of infections that occurred in the cohort.
 Therefore, $p(\ntot \mid \theta) = p(\ntot)$.
@@
 
-$\ntot$ is generally a nuisance parameter, however, its posterior distribution can be useful for diagnostic purposes (as in X\todo{ref where I use this in the next chapter}).
-Its posterior can be reconstructed using a posterior sample of $\theta$.
-For each posterior sample of $\theta$, sample from the full conditional of $\ntot$ to give the joint posterior of $\ntot$ and $\theta$.
-The full conditional of $\ntot$ is given by:
-\begin{align}
-p(n_\text{tot} \mid \ncis, \theta)
-&\propto p(\ntot \mid \theta) p(\ncis \mid \theta, \ntot) \\
-&\propto p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} &\text{by \cref{perf-test:eq:augmented-likelihood}} \\
-&\propto \frac{\Gamma(r + \ntot)}{\ntot!} \left( \frac{\mu}{r + \mu} \right)^{\ntot} \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot} \\
-&= \frac{\Gamma(r + \ntot)}{(\ntot-\ndet)!} \left( \frac{\mu \pnodet}{r + \mu} \right)^{\ntot}  \\
-&= \frac{\Gamma((r + \ndet) + (\ntot- \ndet))}{(\ntot-\ndet)!} \left( \frac{\mu \pnodet}{r + \mu} \right)^{\ntot-\ndet}.
-\end{align}
-Comparing this expression to the pmf of a negative binomial, we find that $\nnodet = \ntot - \ndet$ is distributed negative binomial with size parameter $r+\ndet$ and probability parameter $(r + \mu(1 - \pnodet))/(r + \mu)$.
+% $\ntot$ is generally a nuisance parameter, however, its posterior distribution can be useful for diagnostic purposes (as in X\todo{ref where I use this in the next chapter}).
+% Its posterior can be reconstructed using a posterior sample of $\theta$.
+% For each posterior sample of $\theta$, sample from the full conditional of $\ntot$ to give the joint posterior of $\ntot$ and $\theta$.
+% The full conditional of $\ntot$ is given by:
+% \begin{align}
+% p(n_\text{tot} \mid \ncis, \theta)
+% &\propto p(\ntot \mid \theta) p(\ncis \mid \theta, \ntot) \\
+% &\propto p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} &\text{by \cref{perf-test:eq:augmented-likelihood}} \\
+% &\propto \frac{\Gamma(r + \ntot)}{\ntot!} \left( \frac{\mu}{r + \mu} \right)^{\ntot} \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot} \\
+% &= \frac{\Gamma(r + \ntot)}{(\ntot-\ndet)!} \left( \frac{\mu \pnodet}{r + \mu} \right)^{\ntot}  \\
+% &= \frac{\Gamma((r + \ndet) + (\ntot- \ndet))}{(\ntot-\ndet)!} \left( \frac{\mu \pnodet}{r + \mu} \right)^{\ntot-\ndet}.
+% \end{align}
+% Comparing this expression to the pmf of a negative binomial, we find that $\nnodet = \ntot - \ndet$ is distributed negative binomial with size parameter $r+\ndet$ and probability parameter $(r + \mu(1 - \pnodet))/(r + \mu)$.
 %The mean of this distribution is $(r+\ndet)\mu \pnodet/(r+\mu(1-\pnodet))$.
@@
 By definition (see~\cpageref{perf-test:def:admissible}), we have:
-\todo{Make sure "detectable" is defined: It's really a latent quantity, which could possibly be defined in terms of true viral load.
-I think "positive probability of testing positive" under the assumption of no false positives is what I will go with, or by reference to a multi-state model type view of infection.}
 \begin{align}
@@
 
-Next, I derive $p_{iu} = \prob((B_i, E_i) \in \Omega^C_i \mid \theta)$, the probability that episode $i$ was undetected.
+The next step is to derive $1 - p_u = 1 - \frac{1}{N} \sum_{i=1}^{\Ncis} p_{iu} = \frac{1}{N} \sum_{i=1}^{\Ncis} (1 - p_{iu})$.
+Therefore, the crucial component is $1 - p_{iu} = 1 - \prob((B_i, E_i) \in \Omega^C_i \mid \theta)$, one minus the probability that episode $i$ was undetected.
 An episode is undetected if and only if no tests are performed during the episode or if there was no negative test prior to the episode.
@@
 \end{align}
-Only $1 - p_u = 1 - \frac{1}{N} \sum_i^{\Ncis} p_{iu} = \frac{1}{N} \sum_i^{\Ncis} (1 - p_{iu})$ is required (see \cref{perf-test:eq:full-posterior-simplified}).
-Therefore, I derive $1 - p_{iu}$, again suppressing the conditioning on $\theta$:
+Therefore, suppressing the conditioning on $\theta$:
 \begin{align}
@@
 
-The form of $S_\theta(t)$ is yet to be specified.
-As frequently observed in the literature~\autocite[e.g.:][]{heBayesian}, it is more convenient to operate with the hazard than the survival or probability mass function directly.
-The mass function must sum to one.
-The survival must be monotonically decreasing
-But, there are no constraints across multiple hazards.
-This would complicate the inference process in a way that cannot be removed through a simple transformation.
-The only constraint on the hazard is to be in the interval $[0, 1]$, since they are probabilities.
-The [0, 1] interval is easily be mapped to the unconstrained reals using a logit transformation (where $\logit(x) = \log(x/(1-x))$).
-
+In this section I specify the form of $S_\theta(t)$.
+% As frequently observed in the literature~\autocite[e.g.][]{heBayesian}, it is more convenient to operate with the hazard than the survival or probability mass function directly.
+% The mass function must sum to one.
+% The survival must be monotonically decreasing
+% But, there are no constraints across multiple hazards.
+% This would complicate the inference process in a way that cannot be removed through a simple transformation.
+% The only constraint on the hazard is to be in the interval $[0, 1]$, since they are probabilities.
+% The [0, 1] interval is easily be mapped to the unconstrained reals using a logit transformation (where $\logit(x) = \log(x/(1-x))$).
 Priors cannot be vague on both the hazard and the survival, the implications of which I consider in what follows.
@@
 Specifically, they tend to favour shorter survival times (see \cref{perf-test:fig:flat-prior}).
-An intuitive explanation of this can be drawn from the fact that these types of priors have an expected value of 0.5.
+An intuitive explanation of this can be drawn from the fact that priors of this form have an expected value of 0.5.
 The expected survival time, $\E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right)$, is therefore equal to $0.5^{t-1}$, a quantity that declines rapidly.
@@
 Smoothing priors encode the biological consideration that the hazard should not vary significantly from day-to-day.
-It is the Bayesian equivalent of a penalised likelihood, used frequently in the non-parametric frequentist setting~\autocite[e.g.:][]{bacchettiNonparametric}.
+It is the Bayesian equivalent of a penalised likelihood, used frequently in the non-parametric frequentist setting~\autocite[e.g.][]{bacchettiNonparametric}.
 Many possible forms of smoothing priors exist, including splines, Gaussian Processes, and random walks.
-I use a second-order random walk (RW2) because it is simple and produces plausible prior predictive distributions.
+I use a second-order random walk (RW2) because it is simple and produces sensible prior predictive distributions.
 
@@
 I base this assumption on the estimate from \cref{E-ATACCC} with an inflated tail to represent what is seen within CIS.
-The tail is modified based on an unpublished analysis of CIS data by Sarah Walker.
-Sarah uses survival analysis to estimate the duration from CIS data.
+The tail is modified based on an unpublished analysis of CIS data~\citePersonalComms{A.\ Sarah Walker}.
+Walker uses survival analysis to estimate the duration from CIS data.
 The initiating event is assumed known as the time the episode was detected.
@@
 I'M NOT DOING THIS ANYMORE OH NO!!
-Denote by $f_A(t)$ the distribution function estimated from ATACCC and $f_C(t)$ that from these CIS-based estimates.
+Denote by $f_A(t)$ the distribution function estimated in \cref{E-ATACCC} and $f_C(t)$ that from the CIS-based estimates just derived.
 Then define:
@@
 This meant that the prior was very diffuse.
-
-A framework for performing CIS-like simulations is available at \url{https://github.com/joshuablake/cisSimulation}.
-A framework for fitting the class of models described here and in \cref{E-imperf-test} is available at \url{https://github.com/joshuablake/cisDurationModel}.
-Code utilising these frameworks to perform the analysis described in this chapter is available at \url{https://github.com/joshuablake/cisRuns}.
 
@@
 
-I have assumed constant and independent incidence (a priori).
-Prevalence in the CIS data was approximately constant over the period of interest (see\todo{ref prev constant over period}).
+I have assumed a prior of constant incidence and that incidence is independent between individuals.
+Prevalence in the CIS data was approximately constant over the period of interest (see \cref{E-biology-data:sec:positivity-results}).
 A sensitivity analysis, where an epidemic in exponential growth or decline was simulated (not shown), showed minimal impact on survival estimates.

### 2024-02-19 introduction.tex
ch5 proof

@@
 
+HIGHLIGHT THE FLEXIBLE FRAMEWORK AS A CONTRIBUTION.
+A framework for performing CIS-like simulations is available at \url{https://github.com/joshuablake/cisSimulation}.
+A framework for fitting the class of models described here and in \cref{E-imperf-test} is available at \url{https://github.com/joshuablake/cisDurationModel}.
+Code utilising these frameworks to perform the analysis described in this chapter is available at \url{https://github.com/joshuablake/cisRuns}.
+
 The ATACCC data used is found in the same location.

### 2024-02-19 symbols.tex
ch5 proof

@@
 
+From chapter 5
+\begin{itemize}
+    \item $b_i$ the day infection episode $i$ begins
+    \item $\set{D}$ the set of test schedules with a detected infection episode
+    \item $d_i$ the duration of infection episode $i$
+    \item $e_i$ the day infection episode $i$ ends
+    \item $f_A(t)$ the pdf of the duration estimate from chapter 4
+    \item $f_C(t)$ the pdf of the duration estimate from Walker's analysis
+    \item $k_t$ the belief in the ATACCC estimate at time $t$
+    \item $h_t$ the estimate of the hazard at time $t$ according to ATACCC
+    \item $p_{ia}$ the probability of an admissible infection episode occurring in an individual with test schedule $\sched_i$
+    \item $p_{i\bar{a}}$ the probability of an inadmissible infection episode occurring in an individual with test schedule $\sched_i$
+    \item $p_{iu}$ the probability of an undetected infection episode occurring in an individual with test schedule $\sched_i$
+    \item $\Ncis$ the number of individuals in CIS
+    \item $\ndet$ the number of detected infection episodes
+    \item $n_{ia}$ the number of admissible infection episodes in individuals with test schedule $\sched_i$
+    \item $n_{i\bar{a}}$ the number of inadmissible infection episodes in individuals with test schedule $\sched_i$
+    \item $n_{iu}$ the number of undetectable infection episodes in individuals with test schedule $\sched_i$
+    \item $n_\sched$ the number of unique test schedules in CIS
+    \item $\ntot$ the total number of infection episodes
+    \item $\nnodet$ the number of undetected infection episodes
+    \item $\alpha_0$ the prior in the absence of ATACCC data
+    \item $\alpha_i$ the admissible space for test schedule $i$
+    \item $\beta_0$ the prior in the absence of ATACCC data
+    \item $\beta_i$ the inadmissible space for test schedule $i$
+    \item $\epsilon_t$ the random part of the RW2 model for $\lambda_{t+1}$
+    \item $\eta_i$ the number of individuals with test schedule $i$
+    \item $\theta$ the survival function parameters
+    \item $\lambda_t$ the hazard at time $t$
+    \item $\mu_{\epsilon_1}$ the mean of the prior on $\epsilon_1$
+    \item $\mu_{\lambda_1}$ the mean of the prior on $\lambda_1$
+    \item $\mu_\sigma$ the mean of the prior on $\sigma$
+    \item $\sigma$ the standard deviation of the RW2 prior on $\lambda_t$
+    \item $\sigma_{\epsilon_1}$ the standard deviation of the prior on $\epsilon_1$
+    \item $\sigma_{\lambda_1}$ the standard deviation of the prior on $\lambda_1$
+    \item $\tau_{\sched_i}(t)$ the time between $t$ and the next test for test schedule $i$
+    \item $\Omega_i^C$ the undetected space for test schedule $i$
+\end{itemize}
+
 \end{document}

### 2024-02-20 transmission.tex
fix links

@@
 
-In this chapter, I use the estimate of the duration of RT-PCR positivity from combining information from the ATACCC and CIS studies (see \cref{E-imperf-test:sec:conclusion}) to investigate the feasibility of estimating incidence and transmission of SARS-CoV-2 from only the CIS data (described in \cref{E-biology-data:sec:positivity}).
+In this chapter, I use the estimate of the duration of RT-PCR positivity from combining information from the ATACCC and CIS studies (see \cref{E-imperf-test:sec:conclusion}) to investigate the feasibility of estimating incidence and transmission of SARS-CoV-2 from only the CIS data (described in \cref{E-intro:sec:cis}).
 I will show that the CIS data is sufficient; demonstrating the value of surveys of this type for pandemic surveillance.
@@
 Denote the quantity to be inferred (which could be either incidence proportion or prevalence) at time $t$ in strata $i$ as $\pi_{it}$.
-The strata will represent at least age group and region, although including ethnicity and sex is also due to non-response in the survey (see \cref{E-biology-data:sec:positivity}).
+The strata will represent at least age group and region, although including ethnicity and sex is also due to non-response in the survey (see \cref{E-intro:sec:cis-strata}).
 For many combinations of stratum $i$ and day $t$, the number of positives will be small, often fewer than 5; therefore some form of borrowing between time and strata is required.
@@
         % While incidence proportion can theoretically be greater than 1, this would imply that some individuals were infected multiple times in a day which is not biologically plausible.
-    \item Have sufficient parameters to incorporate all necessary covariates to correct for non-response bias (see \cref{E-biology-data:sec:positivity}).
+    \item Have sufficient parameters to incorporate all necessary covariates to correct for non-response bias (see \cref{E-biology-data:sec:cis}).
     \item Borrow information across the different covariates.
@@
 
-I link the modelled proportion of individuals that are RT-PCR positive to the data using a beta-binomial likelihood (as discussed in \cref{E-biology-data:sec:positivity}).
+I link the modelled proportion of individuals that are RT-PCR positive to the data using a beta-binomial likelihood (as discussed in \cref{E-biology-data:sec:clustering}).
 If on each day $t$, each strata $i$ is observed to have $y_{at}$ positives out of $n_{at}$ tests, then the likelihood is:

### 2024-02-20 biology-data.tex
ch5 proof
fix links
add all intro stuff needed

@@
 
-\subsubsection{Strata}
+\subsubsection{Strata} \label{biology-data:sec:cis-strata}
 
@@
 
-\subsubsection{Positivity results}
+\subsubsection{Positivity results} \label{biology-data:sec:positivity-results}
 
@@
 
-\subsubsection{Episodes}
+\subsubsection{Episodes} \label{biology-data:sec:cis-episodes}
 
@@
 
-\todo[inline]{%
-    From chapter 7 add info on: how this is aggregated/stratified/etc, SRS computational limitations, definitions for each episode (sketched below).
-}
+The crucial features of an infection episode are the start and end times of the infection.
+Ignoring misclassified test results, latent start time of infection episode $i$, $B_i$, can be bounded as between the day after the last negative test, $l_i^{(b)}$, and the day of the first positive test, $r_i^{(b)}$.
+Similarly, the latent end time of infection episode $i$, $E_i$, can be bounded as between the day of the last positive test, $l_i^{(e)}$, and the day before the following negative test, $r_i^{(e)}$.
 
-The beginning of episode $i$ is known to occur in the interval $[l_i^{(b)}, r_i^{(b)}]$, and similarly for the end of the infection in $[l_i^{(e)}, r_i^{(e)}]$.
+\subsubsection{Computational limitations and privacy protection}
+
+For privacy reasons, the CIS data is stored in the ONS's SRS (Secure Research Service).
+The SRS is a Trusted Research Environment (TRE) that allows researchers to access sensitive data in a secure environment.
+
+Statistical disclosure controls must be applied to any data or results to be extracted from the SRS.
+Statistical disclosure control ensures that no individual can be identified from the data.
+This means that data must be aggregated to counts of at least 5, and no individual-data included.
+
+Analyses requiring more fine-grained stratification (such as that in \cref{biology-data:sec:MRP}) or linking data from the same individual over time (such as that in \cref{E-perf-test,E-imperf-test}) must be performed within the SRS.
+Therefore, computationally efficient approaches must be taken to this work.
+
+\section{Conclusion}
+
+In this chapter, I have introduced RT-PCR testing, the natural history of SARS-CoV-2, and the two studies used in this thesis.
+The next chapter will introduce the data generating processes behind infection, prevalence, and our observations of these processes.
+Then, I will apply these concept to the two studies.
 

### 2024-02-20 incidence-prevalence.tex
ch5 proof
fix links
add all intro stuff needed

@@
 The duration distribution I will be concerned with is the distribution of the number of days for which an individual is prevalent.
-Consider an individual $i$ infected on day $B_i$ that is prevalent during the interval $[B_i, E_i]$.
-\todo{check $B_i$ and $E_i$ are not defined elsewhere}
+Consider an individual $i$ infected on day $B_i$ that is prevalent during the interval $[B_i, E_i]$ (as in \cref{E-biology-data:sec:cis-episodes}).
 I define this period as an \emph{infection episode}, beginning on day $B_i$ and ending on day $E_i$.
@@
 
-% Numerical methods for solving systems of this form, that is solving for $\vec{Z}$ (non-random in this expression) for a known $\E (\vec{P} \mid \vec{Z} )$, are well-established~\autocite[e.g.:][section 8.2]{highamAccuracy}.
+% Numerical methods for solving systems of this form, that is solving for $\vec{Z}$ (non-random in this expression) for a known $\E (\vec{P} \mid \vec{Z} )$, are well-established~\autocite[e.g.][section 8.2]{highamAccuracy}.
 % However, these algorithms can be unstable, meaning that numerical errors are large.
@@
 
-Robust, unbiased prevalence estimates already exist (\cref{E-biology-data:sec:positivity}), and incidence estimates are the objective of this thesis.
+Robust, unbiased prevalence estimates already exist (\cref{E-biology-data:sec:cis}), and incidence estimates are the objective of this thesis.
 This chapter has shown that the key quantity relating these quantities is the distribution of the duration of RT-PCR positivity.

### 2024-02-20 ATACCC.tex
fix long list of names in citation

@@
 In these models, the function's parameters are clearly related to features of the biomarker's trajectory, such as the time from first detection to peak viral load, and the viral load at the peak.
-The piecewise linear model is the most common in the context of SARS-CoV-2 viral load~\autocite{clearyUsing,kisslerViral,larremoreTest}
-\todo{work out why Kissler citation is displaying weird}
+The piecewise linear model is the most common in the context of SARS-CoV-2 viral load~\autocite[e.g.][]{clearyUsing,kisslerViral,larremoreTest}
 In this context, the piecewise linear model is applied to the log of the viral load.
@@
 All these factors could change the viral dynamics~\autocite{russellWithinhost,kisslerDynamicsVaccines} and hence the duration.
-\todo{work out why Kissler citation is displaying weird}
 Therefore, the relationships shown here cannot be considered causal, or even generalisable to contexts where these other confounders are not present.

### 2024-02-20 cis-perfect-testing.tex
ch5 clarity

@@
 
-\todo[inline]{make sure following paragraph makes sense in context (copy/pasted here)}
+\todo[inline]{Are the following paragraphs better here or in the intro chapter when motivating each chapter and the overall narrative of the thesis?}
 In the rest of this and the following chapter, I use survival analysis rather than the biomarker approach taken in \cref{E-ATACCC}.
-This choice is motivated by computational efficiency and limitations of the CIS's study design.
+This choice is motivated by the computational limitations of the SRS and the CIS's study design (see \cref{E-intro:sec:cis}).
 The biomarker approach is computational challenging because the number of random effects scales with the number of detected infection episodes.
+
 There are many detected episodes in CIS; in the period I will consider, around 50 times more than in ATACCC.
@@
 
-The CIS tests each individual infrequently, either weekly or monthly (see\todo{refer to section with CIS intro}).
+The CIS tests each individual infrequently, either weekly or monthly (see \cref{E-intro:sec:cis}).
 The infrequent testing complicates analysis of the data in two ways.
@@
 CIS has many undetected infections.
-This is because the interval between tests in the same individual (up to 28 days, see\todo{ref section on the survey design}) is longer than most infections (median 16 days, see \cref{E-ATACCC:sec:results}).
+This is because the interval between tests in the same individual (up to 28 days, see \cref{E-intro:sec:cis}) is longer than most infections (median 16 days, see \cref{E-ATACCC:sec:results}).
 \begin{figure}
@@
 
-\subsection{Double interval censoring}
-\todo{using double interval censored before its defined}
+\subsection{Unobserved episode beginning and end times}
 
@@
 Secondly, the uncertainty in the estimates from \cref{E-ATACCC} should be increased for this analysis.
-The additional uncertainty is because the \cref{E-ATACCC} analysis extrapolates beyond the data using strong model assumptions (see\todo{reference discussion on extrapolation in ATACCC chapter}).
+The additional uncertainty is because the \cref{E-ATACCC} analysis extrapolates beyond the data using strong model assumptions (see \cref{E-ATACCC:sec:discussion}).
 %Furthermore, there are differences in the study design and laboratories used between the two studies (see \cref{E-intro:sec:studies}) which may mean that results do not generalise between the studies.

### 2024-02-20 cis-imperfect-testing.tex
proof ch 6

@@
 In \cref{E-perf-test}, a framework for analysing the CIS to produce estimates of the duration of positivity was developed.
-However, initial application of this framework produced implausibly short estimate.
+However, initial application of this framework produced an implausibly short estimate.
 In this chapter I identify this issue to be the result of false negatives, a form of misclassification bias, within the CIS (\cref{imperf-test:sec:problem}).
+I then propose several models of false negatives (\cref{imperf-test:sec:simulate}) and extend the model of \cref{E-perf-test} to incorporate one of them (\cref{imperf-test:sec:modelling})
+After showing through simulation that the new model recovers the true duration distribution (\cref{imperf-test:sec:sim-study-results}), I apply it to the CIS data (\cref{imperf-test:sec:application}).
+Finally, I discuss the results and further work in \cref{imperf-test:sec:discussion}.
 
@@
 
-In \cref{imperf-test:sec:modelling}, I extend the model of \cref{E-perf-test} to incorporate false negatives.
-After showing through simulation that the new model recovers the true duration distribution (\cref{imperf-test:sec:sim-study-results}), I apply it to the CIS data (\cref{imperf-test:sec:application}).
-Finally, I discuss the results and further work in \cref{imperf-test:sec:discussion}.
-
 \section{The problem} \label{imperf-test:sec:problem}
 
-Applying the method of \cref{E-perf-test} to CIS data estimates fewer individuals of a duration less than three weeks, compared to \cref{E-ATACCC} (see \cref{imperf-test:fig:problem-cis-estimates}).
-The \cref{E-ATACCC} estimates are more reliable over the first 2--3 weeks due to them having dense data collection here.
+Applying the method of \cref{E-perf-test} to CIS data estimates fewer episodes with a duration of less than three weeks, compared to \cref{E-ATACCC} (see \cref{imperf-test:fig:problem-cis-estimates}).
+The \cref{E-ATACCC} estimates are more reliable over the first 2--3 weeks due to them having frequent data collection here (see \cref{E-biology-data:sec:ataccc,E-ATACCC:sec:discussion}).
 This suggests an issue in the estimation using CIS data.
@@
 
-Comparing repeated data simulations using the setup of \cref{E-perf-test:sec:simulation-study} to the observed CIS data, suggests that this is due to the high frequency of single positive episodes in the CIS data (compare the histogram with $\psens=1$ and the vertical line in \cref{imperf-test:fig:sim-single-pos}).
+Comparing repeated data simulations, using the setup of \cref{E-perf-test:sec:simulation-study}, to the observed CIS data, suggests that this is due to the high frequency of single positive episodes in the CIS data (compare the histogram with $\psens=1$ and the vertical line in \cref{imperf-test:fig:sim-single-pos}).
 A single positive episode is an episode containing exactly one positive test; intuitively, these cause estimates to be shorter because the lower bound on the length of time the episode is 1 day, the shortest possible.
@@
 
-\Textcite{shenNonparametrica}, generalising \textcite{panNote}, studied the situation when only the terminating event is interval censored (\ie single censored data) and left truncated.
-They showed that the non-parameteric maximum likelihood estimator is inconsistent if single positive episodes are possible, and the survival function is severely biased downwards.
+\Textcite{shenNonparametrica}, generalising \textcite{panNote}, studied the situation when only the terminating event is interval censored (\ie singly censored data) and left truncated.
+They showed that the non-parameteric maximum likelihood estimator is inconsistent if single positive episodes occur, and the survival function is severely biased downwards.
 
@@
 In this section, I introduce a probabilistic model for the generation of false negatives.
-The simplest model for false negatives is a constant probability of returning a negative test even when detectable.
-The probability of testing positive given being detectable is known as the \emph{test sensitivity}, $\psens$.
-$\psens = 1$ means that there are no false negatives, the situation previously considered.
-In \cref{E-perf-test}, the outcome of a test was deterministic: always positive if the individual being tested had an ongoing infection episode.
-Now, the outcome of such a test result is a Bernoulli random variable with a probability $\psens$ of producing a positive.
-If $\psens = 1$, this recovers the previous model without false negatives.
+The simplest model for false negatives is a constant, non-zero probability of returning a negative test when detectable.
+That is the test sensitivity, $\psens$, is less than 1.
+
+Under this model, each test result in a detectable individual is an iid Bernoulli random variable with a probability $\psens$ of producing a positive.
 As previously, tests in individuals without ongoing infection episodes are always negative.
@@
 
-In \cref{E-ATACCC}, I estimated that $\psens$ ($1-\rho$ in \cref{E-ATACCC}'s parameterisation) is 95\% (95\% CrI: 93--96\%).
-The daily testing in the ATACCC study, used for the \cref{E-ATACCC} estimates, means this is a reliable estimate during the period for which the individuals are followed-up.
+In \cref{E-ATACCC}, I estimated that $\psens$ is 95\% (95\% CrI: 93--96\%).
+The daily testing this is based on means this is a reliable estimate during the period for which the individuals are followed-up.
 The requirement for a much lower value of $\psens$ in the simulation study here suggests this model of false negatives is inadequate.
@@
 This generates broad bounds, although suggest a declining test sensitivity over the course of an episode (see \cref{imperf-test:fig:bounding-cis-sensitivity}).
-Small numbers of tests mean that the series is noisy near the start and the end, such as both bounds being 0 after day 40.
+Small numbers of tests mean that the series is noisy near the start and the end, such as both bounds being 0 on some days after day 40.
 However, the general trend is clear.
@@
 
-I generate results using this model by, for each test in a detectable individual, by calculating $\psens(t-b)$, where $t$ is the day of the test and $b$ is the day the episode began.
+I simulate data under this model by, for each test in a detectable individual, calculating $\psens(t-b)$, where $t$ is the day of the test and $b$ is the day the episode began.
 The result of the test is then a Bernoulli random variable, independent of all other considerations.
@@
 
-\todo[inline]{These definitions (up until next orange box) will move to chapter 2 where episodes are introduced}
-The beginning of episode $i$ is known to occur in the interval $[l_i^{(b)}, r_i^{(b)}]$, and similarly for the end of the infection in $[l_i^{(e)}, r_i^{(e)}]$.
-
-\todo[inline]{From here, this is the section's content}
-
 I will modify $p_{ia}$ to allow the negative test following the last positive to be a false negative.
@@
 
+\todo[inline]{Proof-read the following paragraphs (up until cases). Also check indices: is $i$ indexing individuals, episodes, or schedules?}
 I start by finding the tests that need consideration.
-For tractability, assume that the negative test bounding the start of the episode, on day $l_i^{(b)}-1$, is a true negative.
-This assumption is reasonable because, since a positive test follows at $r_i^{(b)}$, the negative at $l_i^{(b)}-1$ is likely early in the infection and the test sensitivity is high early in an infection.
-Therefore, this is a test in a period with high test sensitivity \ie unlikely to be a false negative.
 For tractability and simplicity, I consider only tests between the negative tests providing an upper bound on the length of the episode.
@@
 They are a subset of $\sched_i$, the set of test times for the individual in which episode $i$ occurs.
-The test at $l_i^{(b)} - 1$ is assumed to be a true negative, occurring with probability 1, and hence does not contribute to the likelihood.
+
+For tractability, assume that the negative test bounding the start of the episode, on day $l_i^{(b)}-1$, is a true negative.
+This assumption is reasonable because, since a positive test follows at $r_i^{(b)}$, the negative at $l_i^{(b)}-1$ is likely early in the infection and the test sensitivity is high early in an infection.
+Therefore, this is unlikely to be a false negative because, if the individual was detectable at this time, they are in a period with high test sensitivity.
+This true negative, occurs with probability 1, and hence does not contribute to the likelihood.
 Denote the remaining tests considered as $\sched'_i = \{ t \in \sched_i \ssep r_i^{(b)} \leq t \leq r_i^{(e)} + 1 \}$, and their results by $\vec{y_i} = \{ y_i(t) \ssep t \in \sched'_i \}$ where $y_i(t) = 1$ if the test on day $t$ is positive and 0 otherwise.
+
 As I assume that there are no false positives, the infection episode must span at least the period $[r^{(b)}_i, l^{(e)}_i]$ because it starts and ends with a positive test.
@@
 
-Now consider the negative test at $r_i^{(e)}+1$, the first negative after the start of the episode which may be a false negative.
+Consider the negative test at $r_i^{(e)}+1$, the first negative after the start of the episode which may be a false negative.
 It is a false negative if and only if the episode ends after the test, \ie $e_i > r_i^{(e)}$.
@@
 The true negative occurs with probability 1, by the assumption of no false positives.
-The tests in $\sched_i'$ are either true positives or false negatives.
 \begin{align}
@@
 To avoid having to consider tests after $r_i^{(e)}$, which could greatly complicate the likelihood, I model this case as the episode being right-censored at $r_i^{(e)}$.
-\todo{check previous notes - I think I investigated this assumption a bit more and maybe did some sensitivity analysis}
 Taking the same approach as before:
@@
 &p(\vec{y_i}, e_i > r_i^{(e)} | b_i, p_\text{sens}, \theta) \\
-&= p(y_i \mid e_i > r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(e_i > r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
+&= p(\vec{y_i} \mid e_i > r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(e_i > r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
 &= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) (1 - p_\text{sens}) S_\theta(r_i^{(e)} - b_i - 1)
@@
 \intertext{Now, substitute in \cref{imperf-test:eq:ll-ei-lt-ri,imperf-test:eq:ll-ei-gt-ri}, then take out the common factor:}
-&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \\ & \ \times \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) + (1 - p_\text{sens}) S_\theta(r_i^{(e)} - b_i - 1) \right) \\ & \ \times p(b_i \mid p_\text{sens}, \theta).
-\intertext{This simplifies to:}
+&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \\ & \ \times \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) + (1 - p_\text{sens}) S_\theta(r_i^{(e)} - b_i - 1) \right) \\ & \ \times p(b_i \mid p_\text{sens}, \theta) \\
 &= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \\ & \ \times \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - p_\text{sens} S_\theta(r_i^{(e)} - b_i - 1) \right) p(b_i \mid p_\text{sens}, \theta).
@@
 First, the ability to identify the survival function even with false negatives.
-Second, the impact of misspecifying $\psens$, including violating the assumption that is constant.
+Second, the impact of misspecifying $\psens$, such as the assumption that it is constant.
 Finally, the impact of the simplifying assumptions made in \cref{imperf-test:sec:modelling} for tractability.
-These assumptions are that the negative at $l^{(b)_i} - 1$ is a true negative and that there is a negligible probability of missing an episode due to two false negatives.
+These assumptions are that the negative at $l^{(b)}_i - 1$ is a true negative and that there is a negligible probability of missing an episode due to two false negatives.
 
@@
 These were shown to be the best performing priors in \cref{E-perf-test:sec:results}.
-Therefore, there are a total of 24 possible scenarios (each combination of: 4 simulated datasets, 3 values of $\psens$ in inference, and 2 priors), although not all are shown.
+Therefore, there are a total of 24 possible scenarios, although not all are shown.
+These are each combination of: 4 simulated datasets, with varying $\psens$; 3 values of $\psens$ in inference; and 2 priors for the hazard.
 
@@
 
+Next, I considered the consequence of $\psens$ being misspecified, using the vague prior; the better performing prior in the correctly specified case.
 If the test sensitivity is misspecified, that is the assumed value for $p_\text{sens}$ in \cref{imperf-test:eq:pit-prime,imperf-test:eq:pit-prime} is different to the value used in the simulation, then the survival time will be biased (see \cref{imperf-test:fig:misspecified-test-sensitivity}).
@@
 The number of episodes inferred to have truly ended by the first negative is too high, and hence the survival time is overestimated.
-This affect dominates over the opposing bias of overestimating the number of missed episodes.
+This effect dominates over the opposing bias of overestimating the number of missed episodes.
 The opposite occurs if the test sensitivity is assumed to be too high, although the posterior moves away from the truth earlier (see \cref{imperf-test:fig:misspecified-test-sensitivity}(C)).
@@
     All simulations use a constant test sensitivity of 0.8, but the inference procedure assumes different values, as per the key.
-    Hence, (B) is correctly specified, and the same as in \cref{imperf-test:fig:constant-test-sensitivity}, but (A) and (C) have misspecified $\psens$.
-    All results use the combination prior, due to it performing better than the vague prior in \cref{imperf-test:fig:constant-test-sensitivity}.
+    Hence, (B) is correctly specified but (A) and (C) have misspecified $\psens$.
   }
@@
 This estimate (blue) has more long episodes than the estimate in \cref{E-ATACCC} (red).
-The qualitative increase in long episodes is robust to the choice of prior for $\ntot$, the assumed value for $\psens$ (see \cref{imperf-test:fig:cis-sensitivity}), and the choice of prior for the hazard (see \cref{E-perf-test:sec:parameters-priors}), $\lambda$.
+
+The qualitative increase in long episodes is robust to the choice of prior for $\ntot$, the assumed value for $\psens$ (see \cref{imperf-test:fig:cis-sensitivity}), and the choice of prior for the hazards, $\lambda_t$ (see \cref{E-perf-test:sec:parameters-priors}).
 However, the survival proportion over the first 4 weeks is sensitive to these choices.
@@
 
-The assumption that the estimates are most sensitive to is $r$, the strength of the prior on $\ntot$.
+The the estimates are most sensitive to the choice of $r$, the strength of the prior on $\ntot$.
 A very weak, almost uninformative, prior on this quantity causes the posterior estimate to be much higher than the estimate from \textcite{birrellRTM2}.
@@
 As discussed previously, the \cref{E-ATACCC} estimates are reliable for the first 2--3 weeks, notably including the median time.
-Therefore, I choose a value of $r$ to match closely \cref{E-ATACCC}'s median estimate.
-The value $r\inform$ achieves this and is a principled choice because it is based directly on the prior work \textcite{birrellRTM2}.
+The median using $r\inform$ matched \cref{E-ATACCC}'s median estimate and is a principled choice because it is based directly on the prior work \textcite{birrellRTM2}.
+Therefore, I use this estimate in the application to the CIS data.
 \begin{figure}
@@
 In the simulation study, the uninformative prior on $\ntot$ performed well, even when the test sensitivity was misspecified.
-A possible explanation is that the true variation in the test sensitivity is significantly different to the function in \cref{imperf-test:eq:variable-test-sensitivity}.
+A possible explanation is that the true changes in the test sensitivity is significantly different to the function in \cref{imperf-test:eq:variable-test-sensitivity}.
 In particular, the test sensitivity for long episodes may become much lower than 50\%.
@@
 Estimating a time-varying $\psens$ with $S_\theta$ jointly may cause identifiability issues.
-These can be avoided by including external information (such as a prior giving the magnitude) on $\psens$, or test results from later follow-up~\autocite[and references therein]{piresIntervalMisclassify}.
-However, whether these methods are sufficient for the model to be identifiable with a time-varying $\psens$ needs further investigation.
+Previous studies have avoided issues by including external information (such as a prior giving the magnitude) on $\psens$, or test results from later follow-up~\autocite[and references therein]{piresIntervalMisclassify}.
+However, these studies use a constant $\psens$.
+Whether these methods are sufficient for the model to be identifiable with a time-varying $\psens$ needs further investigation.
 A simple parameteric form of the test sensitivity, such as that proposed by \textcite{brownBayesian}, may be sufficient to allow identifiability.
@@
 
-In this chapter, I developed and applied novel methodology to estimate the duration distribution of RT-PCR-positivity using information from two studies: ATACCC and CIS.
+In this chapter, I developed and applied novel methodology to estimate the duration distribution of RT-PCR positivity using information from two studies: ATACCC and CIS.
 Assuming a test sensitivity of 0.8 and an informative prior on the number of infections, provided the preferred estimates.
 
-These estimates are the first estimates of the duration of RT-PCR-positivity, including the distribution tails, estimated from general population data.
+These estimates are the first estimates of the duration of RT-PCR positivity, including the distribution tails, estimated from general population data.
 This feature of the distribution is important for understanding the proportion of prevalence attributable to long infections and interpreting RT-PCR test results.
-\todo{possibly cut some of this text?}
+
 Producing these estimates required applying previous survival analysis frameworks in the novel context of the CIS.

### 2024-02-20 symbols.tex
proof ch 6

@@
     \item $\nnodet$ the number of undetected infection episodes
+    \item $r$ the dispersion of the prior on $\ntot$
+    \item $\sched_i$ the $i$th unique test schedule
     \item $\alpha_0$ the prior in the absence of ATACCC data
@@
     \item $\lambda_t$ the hazard at time $t$
+    \item $\mu$ the mean of the prior on $\ntot$
     \item $\mu_{\epsilon_1}$ the mean of the prior on $\epsilon_1$
@@
 
+From chapter 6
+\begin{itemize}
+    \item $p_{ia}'$ the replacement for $p_{ia}$ including false negatives
+    \item $p_{iu}'$ the replacement for $p_{iu}$ including false negatives
+    \item $\psens$ test sensitivity
+    \item $r\inform$ the dispersion of the informative prior on $\ntot$
+    \item $y_i(t)$ the test result for infection episode $i$ at time $t$
+    \item $\vec{y_i}$ the test results associated with infection episode $i$
+    \item $\eta$ the additional probability of missing an infection episode due to false negatives
+    \item $\mu\inform$ the mean of the informative prior on $\ntot$
+    \item $\tau_{\sched_i}(t)$ the time between $t$ and the next but one test for test schedule $i$
+\end{itemize}
+
 \end{document}

### 2024-02-22 introduction.tex
start outling intro

@@
 
-\begin{itemize}
-    \item Disease burden of pandemic and ongoing burden of COVID
-    \item To mitigate its effects, need to understand incidence and transmission but cannot observe directly
-    \item Importance of surveillance objectives and concepts: incidence, transmission, basic reproduction number, effective reproduction number
-    \item Expected future pandemics, methods should be reusable
-\end{itemize}
+COVID-19 (Coronavirus Disease 2019), the disease caused by the virus SARS-CoV-2 (Severe Acute Respiratory Syndrome Coronavirus 2), is estimated to have killed around\todo{insert excess deaths estimates by Economist and/or WHO}.
+While it is tempting to consider this a rare or extreme event, history suggests otherwise (see \cref{intro:fig:pandemic-timeline}).
+\begin{figure}
+    \includegraphics[width=\textwidth]{introduction/pandemic-timeline}
+    \caption[Timeline of pandemics.]{%
+        Timeline of pandemics.
+        See \url{https://ourworldindata.org/historical-pandemics} for details and sources.
+        Licensed under \href{https://creativecommons.org/licenses/by/4.0/}{CC-BY} by Saloni Dattani, Klara Auerbach, Marwa Boukarim, and Max Roser.
+    }
+    \label{intro:fig:pandemic-timeline}
+\end{figure}
 
-% - Natural history of SARS-CoV-2 infection
-% 	- Stylised graphic of phases: infection, pre-infection positive, infectious positive, post-infectious positive, negative
-% 	- How it spreads
+Fully quantifying the burden of historical pandemics is challenging due to the absence of systematic records.
+However, our best records suggests that, over the last 400 years, they have (on average) occurred every four years with a mean death toll of 5.6 million\todo{cite Marani et al.}.
+Modelling of our current situation suggests that we should expect X deaths per year from pandemics\todo{get figures from and cite Madhov et al.}.
+
+\section{Pandemic and epidemic surveillance}
+
+\todo[inline]{Importance of pandemic surveillance, and epidemic v pandemic}
+
+No single quantity summarises the full state of an epidemic, and no single data source fully informs all quantities of interest.
+Therefore, a range of metrics and surveillance systems are required to understand the state of an epidemic and the danger of a pathogen (a pathogen is any biological agent that can cause disease).
+
+\subsection{Metrics}
+
+Various forms of \emph{reproduction number} were prominent throughout the COVID-19 pandemic\todo{cite Pellis et al. discussion paper? or maybe something they cite}.
+\todo[inline]{Define different reproduction numbers}
+
+\todo[inline]{Talk about growth rates}
+
+Growth rates and reproduction numbers only provide information on how an epidemic is changing.
+\emph{Incidence} is the primary quantity of interest for the current (or historical) magnitude of a pandemic.
+
+\todo[inline]{Insert discussion on points in following comment}
 % - Why is incidence and prevalence estimation challenging?
@@
 
-\section{SARS-CoV-2 real-time surveillance}
+\todo[inline]{Define transmission}
 
-\begin{itemize}
-    \item Most commonly used types of data: severe events (deaths/hospitalisations), widespread testing (pillar 2), 
-    \item Advantages of designed surveys (REACT, CIS)
-    \item Use of cohorts (ATACCC)
-    \item Statistical modelling performed during the pandemic. Discuss broadly but especially focus on those using CIS or REACT. Discuss my contribution in this context.
-\end{itemize}
+To translate any of these quantities into the burden of the disease, a measure of severity is required.
+For example, the IFR (infection fatality rate), the proportion of infected individuals who die.
+Severity does not feature in this thesis, but it is important to note that it is a key component of understanding the burden of a pandemic.\todo{maybe reword sentence}
 
-\subsection{Modelling}
+\subsection{Data sources}
 
-\begin{itemize}
-    \item SPI-M
-    \item Literature on mechanistic models using REACT and CIS I have previously circulated.
-    \item \textcite{mccabeCISincidence,abbottCISincidence} on backcalculation.
-    \item Also Taffe?
-\end{itemize}
+A wide variety of designed studies, administrative data, and other data sources have been used for pandemic and epidemic surveillance.
+In the UK SARS-CoV-2 pandemic, the most prominent were: community testing (pillar 2), the number of severe events (\eg hospital admissions or deaths), seroprevalence testing (seroprevalence is the proportion of people with antibodies to a disease, a marker of prior infection\todo{maybe define antibodies and seroprevalence / make this more succint}), and prevalence surveys (\eg REACT and CIS).\todo{add citations for data sources and spell out REACT and CIS}
 
-\subsection{Surveillance data}
-
-The first study, ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts), performed daily testing of a small number of individuals with short follow-up.
-The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability, but the results are limited because the small sample size and short follow-up means that the tails of these distributions cannot be estimated.
+Each of these data sources have strengths and weaknesses.
+\todo[inline]{compare different data sources and why I focus on CIS}
 
@@
 
-\section{Historical context}
+In addition to population-level surveillance, cohort studies help understand other aspects of the pathogen.
+For example, contact tracing data can inform how likely it is for transmission to occur in different settings and/or between different groups of people.
+The ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts) study is an example of a cohort study in the UK SARS-CoV-2 pandemic\todo{add citation}.
+
+The first study, ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts), performed daily testing of a small number of individuals with short follow-up.
+The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability, but the results are limited because the small sample size and short follow-up means that the tails of these distributions cannot be estimated.
+
+\begin{itemize}
+    \item Most commonly used types of data: severe events (deaths/hospitalisations), widespread testing (pillar 2), 
+    \item Advantages of designed surveys (REACT, CIS)
+    \item Use of cohorts (ATACCC)
+    \item 
+\end{itemize}
+
+\subsection{Estimating incidence and transmission}
+
+\subsubsection{Prior to SARS-CoV-2}
 
@@
 \end{itemize}
-
-\subsection{Surveillance}
-
-\begin{itemize}
-    \item Reproduction numbers
-    \item Incidence: almost never observed directly
-\end{itemize}
-
-\subsection{Backcalculation}
 
@@
 \end{itemize}
-
-\subsection{Mechanistic models}
 
@@
 
-\subsection{Duration} \label{inc-prev:sec:duration}
+\subsubsection{For SARS-CoV-2}
+
+\todo[inline]{Statistical modelling performed during the pandemic. Discuss broadly but especially focus on those using CIS or REACT. Discuss my contribution in this context.}
+
+\begin{itemize}
+    \item SPI-M
+    \item Literature on mechanistic models using REACT and CIS I have previously circulated.
+    \item \textcite{mccabeCISincidence,abbottCISincidence} on backcalculation.
+    \item Also Taffe?
+\end{itemize}
+
+\subsection{Estimating duration}
 
@@
 
+\section{Bayesian inference}
+
 \section{Notational conventions}
@@
 \section{Thesis contribution}
+
+\todo[inline]{Insert contributions listed in Google Doc}
 
@@
 
-The ATACCC data used is found in the same location.
-CIS data can only be accessed by accredited researchers within the Office for National Statistic's Secure Research Service\todo{insert details of applying to SRS}.
 Where possible, aggregated, non-disclosive versions of the dataset are available alongside the code.
+The original ATACCC data used is found in the same location.
+The original CIS data can only be accessed by accredited researchers within the Office for National Statistic's Secure Research Service\todo{insert details of applying to SRS}.
 

### 2024-02-26 introduction.tex
first attempt in outlining intro

@@
 
+\todo[inline]{Maybe talk about pandemic v epidemic v outbreak terminology?}
+
 \section{Pandemic and epidemic surveillance}
@@
 
-No single quantity summarises the full state of an epidemic, and no single data source fully informs all quantities of interest.
+No single metric summarises the full state of an epidemic, and no single data source fully informs all metrics of interest.
 Therefore, a range of metrics and surveillance systems are required to understand the state of an epidemic and the danger of a pathogen (a pathogen is any biological agent that can cause disease).
@@
 Growth rates and reproduction numbers only provide information on how an epidemic is changing.
-\emph{Incidence} is the primary quantity of interest for the current (or historical) magnitude of a pandemic.
+\emph{Incidence} is the primary metric for the magnitude of a pandemic.
+See \cref{E-inc-prev:sec:definitions} for a formal definition of incidence and prevalence.
 
@@
 
-To translate any of these quantities into the burden of the disease, a measure of severity is required.
+To translate incidence into the burden of the disease, a measure of severity is required.
 For example, the IFR (infection fatality rate), the proportion of infected individuals who die.
-Severity does not feature in this thesis, but it is important to note that it is a key component of understanding the burden of a pandemic.\todo{maybe reword sentence}
+% Severity does not feature in this thesis, but it is important to note that it is a key component of understanding the burden of a pandemic.\todo{maybe reword sentence}
 

### 2024-02-28 ATACCC.tex
ch5 clarity
write most of intro

@@
 
-Several features of the \textcite{hakkiOnset} model can be motivated by the desire to apply Stan, a modern and convenient MCMC implementation of the NUTS algorithm (see \cref{E-inc-prev:sec:MCMC}).
+Several features of the \textcite{hakkiOnset} model can be motivated by the desire to apply Stan, a modern and convenient MCMC implementation of the NUTS algorithm (see \cref{E-MCMC}).
 The first of these is using a continuous viral load trajectory rather than the more common piecewise linear model.
@@
 
-\section{Discussion}
+\section{Discussion} \label{ATACCC:sec:discussion}
 

### 2024-02-28 MCMC-appendix.tex
write most of intro

--- 2024-02-16 incidence-prevalence.tex
@@
 \begin{document}
-\ifSubfilesClassLoaded{
-  \setcounter{chapter}{2}
-}
 
-\chapter{Statistical and epidemiological background} \label{inc-prev}
+\chapter{Markov chain Monte Carlo} \label{MCMC}
 
-%\todo[inline]{This chapter is pretty short. Not sure if this matters or it needs to merged with another chapter? For example, one background chapter split into epi/biology and stats.}
+This appendix outlines the basic theory and implementation of Markov chain Monte Carlo (MCMC) methods.
+Any reader interested in more technical details or proofs of the results here should consult the references within4.
 
-This chapter defines incidence and prevalence, and the relationship between these quantities.
-In the course of explaining this relationship, it will become clear that the duration of RT-PCR positivity, and its distribution in the population, is the vital parameter which will allow estimation of incidence from prevalence, a major aim of this thesis.
-
-I start by defining these terms in \cref{inc-prev:sec:definitions}.
-I then move on to explaining the three processes that are involved: the \emph{infection process}, the \emph{prevalence process} and the \emph{observation process}.
-
-The infection process (\cref{inc-prev:sec:infection-process}) models how infections arise.
-The concepts here are well-established within infectious disease epidemiology (see\todo{ref intro chapter}).
-
-The prevalence process (\cref{inc-prev:sec:prevalence-process}) describes the relationship between incidence and prevalence.
-Here, I adapt previous work to the SARS-CoV-2 context.
-In particular, recovering from infection needs to be considering, which was not the case for HIV/AIDS.
-
-The observation process (\cref{inc-prev:sec:observation-process}) describes the relationship between prevalence and the observed data, ultimately allowing inference of incidence based on this data.
-Unlike in HIV/AIDS backcalculation, the CIS has a sampling mechanism that needs to be modelled.
-This model is guided by the specific study design of the CIS.
-
-Finally, \cref{inc-prev:sec:MCMC} will introduce Markov chain Monte Carlo (MCMC), an inference technique I will make extensive use of in this thesis.
-
-\Cref{inc-prev:sec:conclusion} concludes the chapter.
-
-\section{Definitions} \label{inc-prev:sec:definitions}
-
-Time will be considered discrete in this chapter, except in \cref{SEIR:sec:mechanistic-models}.
-I use days, however, the results generalise in the obvious way to other time units.
-I choose discrete time for pragmatic reasons.
-Most epidemiological data is only available at daily granularity or, when more granular data exists, is often unreliable.
-Furthermore, within-day patterns produce variations in transmission (\eg due to sleep) and data (\eg due to logistical considerations around collection); these variations are not of interest in most contexts.
-
-\emph{Incidence} on day $t$, $Z_t$, is the number of people in a given population who become infected on day $t$; when expressed as a proportion of the population it is the \emph{incidence proportion}~\autocite[89]{lashModern}, $Z_t/\Npop$, where $\Npop$ is the number of people in the population.
-As discussed in\todo{ref intro}, the timing of infection events is very rarely directly observable and must be inferred from other data.
-Despite this issue, it remains the single most important quantity for informing the response to an epidemic or pandemic.
-
-\emph{Prevalence} of a disease on day $t$, is the proportion of people with the disease on that day~\autocite[90]{lashModern}.
-Within this thesis, I define an individual as having SARS-CoV-2 on day $t$ if, had they been swabbed on day $t$, that swab would return a positive RT-PCR test.
-I refer to these individuals as \emph{prevalent} individuals.
-I assume that an individual's prevalent status cannot change over the course of a day.
-Therefore, prevalence is constant over any single day $t$.
-I denote the number of prevalent individuals on day $t$ as $P_t$.
-Therefore, the prevalence is $P_t/\Npop$.
-
-\emph{Duration}, or \emph{time-to-event}, distributions can be expressed in several equivalent ways~\autocite[17]{sunStatisticala}.
-The duration distribution I will be concerned with is the distribution of the number of days for which an individual is prevalent.
-Consider an individual $i$ infected on day $B_i$ that is prevalent during the interval $[B_i, E_i]$.
-\todo{check $B_i$ and $E_i$ are not defined elsewhere}
-I define this period as an \emph{infection episode}, beginning on day $B_i$ and ending on day $E_i$.
-$D_i = E_i - B_i + 1$ is their \emph{duration of positivity}, the number of days for which they are prevalent.
-I model the $D_i$s as a discrete independent and identically distributed (iid) random variable; extending the methodology to account for covariates is left to further work.
-The duration is a positive integer, \ie $D_i \in \{1, 2, \dots, \dmax \}$, where $\dmax$ is the maximum duration.
-The first way of expressing the distribution of $D_i$ is as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
-The \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$.
-%(a common alternative convention for survival analysis with discrete variables is to define the survival function as $\prob(D >d)$).
-The \emph{hazard} on day $d$ is the probability of recovering on day $d$ conditional on still being positive on day $d$: $\lambda(d) = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
-Throughout this thesis, the survival function will be the primary quantity of interest, and it will normally be expressed in terms of the hazard; these two quantities are related by $S(d) = \prod_{i=1}^{d-1} (1 - \lambda(i))$.
-
-Since $D_i$ is a discrete random variable, $0 \leq f_D(d) \leq 1$ for all $d$ and $\sum_{k=1}^{\dmax} f_D(k) = 1$.
-This implies that $F_D(d) \in [0, 1]$ and increases monotonically.
-Therefore, $S(d) \in [0, 1]$ and decreases monotonically.
-Similarly, $\lambda(d) \in [0,1]$, but can otherwise vary arbitrarily (\ie there is no monotonic or sum constraint).
-The lack of monotonicity or sum constraint on the hazard makes it an attractive parameterisation for inference.
-
-\section{Infection process} \label{inc-prev:sec:infection-process}
-The infection process is the mathematical description of how infections occur.
-It is a \emph{point} or \emph{counting} process, where each infection is an event.
-
-A point process is a random function of continuous time $K(t)$.
-$K(t)$ must take non-negative integer values and be monotonically increasing in $t$.
-The interpretation of $K(t)$ is the number of events, here infections, that have occurred in the interval $(-\infty, t]$~\autocite[244]{yanDistribution}.
-In the context of a disease, the point process can be discretised into the incidence on day $t$ as $Z_t = K(t) - K(t-1)$, where $t$ is an integer.
-I will explain two approaches to modelling the infection process.
-
-First, in \cref{inc-prev:sec:poisson-process}, as a \emph{Poisson process}.
-A Poisson process is a widely-used type of counting process both within and outside of epidemiology.
-It makes no direct assumptions on the underlying biological or epidemiological process, and is therefore a flexible model.
-
-Afterwards, in \cref{SEIR:sec:mechanistic-models}, I introduce \emph{mechanistic models}.
-Mechanistic models explicitly describe the population and transmission between individuals.
-They vary greatly in their details and realism~\autocite{murilloMultiscale}.
-
-\subsection{Poisson processes} \label{inc-prev:sec:poisson-process}
-
-A commonly-used point process for infectious disease epidemiology is the time-inhomogenous Poisson process~\autocite[e.g., in the context of HIV,][]{brookmeyerMethod,paganoHIV,rosenbergBackcalculation,brookmeyerBackcalculation}.
-% A time-homogenous Poisson process is the special case where $\nu(r)$ is constant.
-Extensions to this model, which I do not consider here, have incorporated covariates, such as geography or age~\autocite[e.g.][]{diggleModeling}.
-
-The first defining property of a Poisson process is that the counts are Poisson distributed.
-Formally, $K(s) - K(t) \dist \Poi\left(\int^s_t \nu(r) dr \right)$ where $\nu(r) \geq 0$ is the \emph{intensity} of the infection process at time $r$~\autocite[244]{yanDistribution}.
-
-The other defining property of a Poisson process is that it has \emph{independent increments}.
-A counting process has independent increments if the number of events in disjoint intervals are independent conditional on the intensity~\autocite[244]{yanDistribution}.
-That is, for any $t_1 < t_2 < t_3 < t_4$, $K(t_2) - K(t_1)$ and $K(t_4) - K(t_3)$ are independent.
-In the context of a Poisson process, the independent increment property implies that, conditional on $\nu(r)$, the number of infections in disjoint intervals are independent.
-Hence, $Z_t$ and $Z_{t'}$ are independent for $t \neq t'$.
-
-The target of inference can be either the process's intensity or the counts of the specific realisation that occurred.
-The realised counts are normally of greater interest because we care about the realised epidemic, not what may happen if there were future realisations of the same epidemic~\autocite{beckerDependent,brookmeyerMethod}.
-
-Two features of a Poisson process are violated for epidemics.
-Epidemics commonly exhibit \emph{overdispersion} and have feedback loops~\autocite{beckerDependent}.
-
-Overdispersion means that $\V(Z_t) > \E(Z_t)$.
-In a Poisson process $Z_t$ has a Poisson distribution, hence, $\V(Z_t) = \E(Z_t)$.
-Therefore, an overdispersed epidemic modelled as a Poisson process will be less variable than the realised epidemic.
-Overdispersion is common in infectious diseases, especially for coronaviruses such as SARS-CoV-2~\autocite{endoEstimating,adamClustering,mccloskeySARS}.
-
-% A common cause of overdispersion is super-spreading events~\autocite{lloyd-smithSuperspreading}.
-% A super-spreading event is an event where many more people than average are infected.
-% That is, $Z_t >> \E(Z_t)$ for some $t$ during which a super-spreading event occurs.
-% Coronaviruses, such as SARS-CoV-2, have particularly high overdispersion~\autocite{endoEstimating,adamClustering}{mccloskeySARS}.
-% An infamous example of a SARS-CoV-2 super-spreading event was the Skagit County choir practice in March 2020~\autocite{hamnerHigh}.
-% One infected individual is thought to have infected at least 32 further individuals during a single choir practice.
-% For comparison, the average individual infected around three others in the early stages of the pandemic~\autocite{pellisChallenges}. 
-% Overdispersion can be incorporated by replacing the assumption that $K(s) - K(t) \dist \Poi(\int_{t}^{s} \nu(r) dr)$ with $K(s) - K(t) \dist \NBc(\int_{t}^{s} \nu(r) dr, k)$ where $k$ is the overdispersion parameter, normally assumed to be constant absent interventions.
-% Note that the expected number of events has not changed, but the variance has increased (see \cref{E-distributions} for details of parameterisations and moments).
-% %The negative binomial tends to Poisson as $k\to\infty$.
-% Generally, $k<1$ is considered as significant levels of overdispersion.
-% Estimates for SARS-CoV-2 indicate $k \approx 0.1$~\autocite{endoEstimating}, and may be decreased by some interventions including lockdowns~\autocite{quiltyReconstructing,quiltyUnderstanding}.
-% If $K(s) - K(t)$ has a negative binomial distribution, then its standard deviation scales on the order of the mean.
-% If it is Poisson distributed, then the standard deviation is on the order of the square root of the mean.
-% A particular implication of this is that the coefficient of variation (the standard deviation divided by the mean) tends to zero for a Poisson process, but tends to a constant for a negative binomial process.
-% The disadvantage of the negative binomial distribution is that it loses much of the mathematical convenience of a Poisson distribution.
-
-Feedback loops mean that stochastically low or high numbers of infections affect the future number of infections.
-That is, $\nu(t+\Delta t)$, for $\Delta t > 0$, is dependent on $K(t)$, even after conditioning on $\nu(t)$.
-This violates the assumption of independent increments.
-Infectious diseases generate feedback loops because stochastically high or low incidence feeds back to the process intensity at future times.
-% This can be seen most obviously in the event of a super-spreading event, when many .
-% If a super-spreading event occurs, then many more individuals than suggested by the process intensity will be infected.
-% These additional infections will propagate the epidemic further, increasing the intensity at future times.
-Point processes can be generalised to \emph{branching processes} to account for this feedback~\autocite[246]{yanDistribution}, although these are beyond the scope of this thesis.
-
-Both overdispersion and feedback loops violate the properties of a Poisson process.
-Despite these issues, consistent inference on the incidence can still be produced using the Poisson process model, although the uncertainty in the estimates may be understated~\autocite{beckerDependent}.
-
-% In \emph{deterministic backcalculation}, the observed prevalence, $x_t/n_t$, is assumed to be equal to $P_t/\Npop$.
-% However, this is often a poor approximation.
-% \emph{Statistical backcalculation} retains the sampling distribution of $x_t$ (or an approximation of it, such as a Poisson).
-
-% Define the first day of an infection episode in individual $i$ as $B_i$, and assume that the probability of $i$ being infected multiple times within the period of interest is negligible.
-% The time between being infected and first being detectable is short (see\todo{ref relevant part}), and therefore I often assume that $B_i$ is the same as the time of infection.
-% I keep this assumption for the remainder of this chapter.
-
-% Backcalculation makes use of this relationship to estimate incidence from prevalence, assuming that the duration is known.
-
-% Deterministic backcalculation assumes that the variance in the population is negligible, and therefore that the prevalence is equal to the mean prevalence.
-% This is justified because the variance due to sampling is much larger than the variance due to the true prevalence.
-
-\subsection{Mechanistic models} \label{SEIR:sec:mechanistic-models}
-
-Now I introduce mechanistic models, in particular the class of mechanistic transmission models known as \emph{compartmental} models.
-Compartmental models are the most widely-used class of mechanistic infectious disease model.
-A compartmental model is named because they place every member of the population under study into a compartment; each compartment representing a disease state (\eg susceptible).
-I will describe these models using the notation of the field, which differs in conventions from the rest of this thesis.
-In particular, upper case or lower case will no longer be used to distinguish between random variables and their realisations, which instead is clear by context.
-
-A compartmental model consists of a state vector $\vec{x}(t)$ and a set of parameters $\vec{\theta}$~\autocite{birrellEvidence,dukicTracking,corbellaThesis}.
-The elements of $\vec{x}(t)$ are the latent proportion of the population (or number of individuals) in each compartment at time $t$.
-The parameters $\vec{\theta}$ determine, along with the model structure, how the state vector changes over time.
-% Inference about $\vec{x}(t)$ and $\vec{\theta}$ is requires linking to data, $\vec{y}$, via an observation model, $p(\vec{y}(t) \mid \vec{x}, \vec{\theta})$ where $\vec{x}$ denotes conditioning on $\vec{x}(t)$ for all $t$~\autocite{birrellEvidence} (see \cref{SEIR:sec:observation}).
-% That is, the observation model specifies the likelihood of the data given the model state.
-
-Compartmental models can be \emph{deterministic} or \emph{stochastic}.
-Deterministic models are defined by the state vector $\vec{x}(t)$ being a deterministic function of the parameters $\vec{\theta}$ and the initial state $\vec{x}(0)$; equivalently, $\var(\vec{x}(t) \mid \vec{x}(0), \theta) = 0$~\autocite{birrellEvidence}.
-% In most situations, the function is not analytically-tractable and requires numeric methods to solve.
-%The transmission model corresponds to the infection process (see \cref{E-inc-prev:sec:infection-process}).
-%There remains randomness in the observation model (described in \cref{SEIR:sec:observation}).
-
-Stochastic models specify a probability distribution for the current state $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta})$ where $\set{H}_t = \{ \vec{x}(t') \ssep t' < t \}$ is the epidemic's history and $\vec{\theta}$ are model parameters~\autocite{birrellEvidence,dukicTracking,corbellaThesis,keelingModeling}.
-Stochastic models are most commonly specified in discrete time, using a Markov assumption.
-A common simplification is to use discrete time and a Markov assumption; \ie $\vec{x}(t)$ is only defined for integer $t$ and $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta}) = p(\vec{x}(t) \mid \vec{x}(t-1), \vec{\theta})$~\autocite{birrellEvidence}.
-This relaxes the assumption of independent increments in the counting process formulation of the infection process (see \cref{E-inc-prev:sec:infection-process}).
-
-Deterministic models are also known as \emph{mean-field} models because they represent the mean behaviour of a corresponding stochastic model.
-When the number of infectious individuals is large, the law of large numbers implies that the mean-field behaviour is a good approximation of the fully stochastic model~\autocite[20]{diekmannMathematical}.
-This is especially true when making noisy observations of the system, and that the stochastic differences between the stochastic and its mean are negligible compared to the observation noise.
-I make a formalisation of this argument to justify a mean-field approximation of the prevalence process in \cref{E-inc-prev:sec:observation-process}.
-
-The remainder of this section uses only deterministic models.
-This choice is appropriate for the context of this thesis because there were many individuals infectious with SARS-CoV-2 in this time period.
-Hence, further details on stochastic models are not outside the scope this thesis; a wealth of introductory texts with further details on stochastic models are available, I recommend \textcite[chapter 6]{keelingModeling}.
-
-% This section discusses deterministic models whose evolution is specified by a system of ODEs, $d\vec{x}/dt$.
-% In an ODE model, time is necessarily continuous.%, although discretised when solving (see \cref{SEIR:sec:inference-implementation}).
-% For concreteness, I use units of days throughout, although the model can use any unit of time.
-% The number or proportion of individuals in each compartment is also continuous; clearly, this is not true in reality but is a good approximation when the population is large.
-
-% In this section, I introduce the \emph{susceptible-infectious-recovered} (SIR) model and the extensions to it I require to appropriately model SARS-CoV-2.
-
-% \Cref{SEIR:sec:SIR} introduces the \emph{susceptible-infectious-recovered} (SIR) model, the simplest compartmental model for a disease where individuals are immune to reinfection following their recovery.
-% The name SIR derives from the structure of the model.
-% In particular, the model divides the population into one of three states: susceptible, infectious, or recovered.
-
-% This model has several assumptions which make it inappropriate to model SARS-CoV-2; therefore, in the remainder of this section I introduce standard extensions to the SIR model which relax these assumptions.
-% First, in \cref{SEIR:sec:non-exponential}, I relax the assumption that the infectious period is exponentially-distributed to allow gamma distributions.
-% Then, in \cref{SEIR:sec:SEIR}, I relax the assumption that individuals are infectious immediately after they are infected; it does this by extending the SIR model to the SEIR model with the additional E signifying an \emph{exposed} compartment.
-% Finally, in \cref{SEIR:sec:structured-populations}, I include heterogeneity in the population, stratifying by important characteristic(s).
-% Finally, \cref{SEIR:sec:time-varying-foi} relaxes the assumption of a constant transmission rate to allow for time-varying behavioural changes.
-
-\subsubsection{The SIR model} \label{SEIR:sec:SIR}
-
-\begin{figure}[h]
-\makebox[\textwidth][c]{
-\begin{tikzpicture}[
-    node distance = 2.5cm,
-    on grid,
-    auto,
-    ->,>=stealth',
-    every state/.style={draw,rectangle},
-    ]
-
-    \node[state] (S) {$s$};
-    \node[state, right=of S] (I) {$i$};
-    \node[state, right=of I] (R) {$r$};
-
-    \path (S) edge node {$\lambda(t)$} (I)
-          (I) edge node {$\gamma$} (R);
-\end{tikzpicture}
-}
-  \caption[The SIR model]{Schematic of the basic SIR model. Arrows are labelled with the rates at which individuals move between compartments.}
-  \label{SEIR:fig:SIR}
-\end{figure}
-
-The simplest compartmental model for a disease which confers immunity is the SIR (susceptible-infectious-recovered) model~\autocite[15]{keelingModeling}, depicted in \cref{SEIR:fig:SIR}.
-The name SIR derives from the structure of the model.
-In particular, the model divides the population (assumed closed, an assumption discussed later) into one of three compartments: susceptible, infectious, or recovered.
-% The state of the system is the proportion of the population in each compartment.
-In the SIR model, individuals are assumed to be homogeneous except for their disease state, \ie they are otherwise identical.
-
-Individuals in the susceptible compartment are those that could be infected (they have no immunity).
-The proportion of the population that is susceptible at time $t$ is $s(t)$.
-Individuals in the infectious compartment can infect others, and eventually recover.
-The proportion of the population that is infectious at time $t$ is $i(t)$.
-Individuals in the recovered compartment are immune to the disease, and cannot be infected again.
-The proportion of the population that have recovered by time $t$ is $r(t)$.
-The model's state is $\vec{x}(t) = (s(t), i(t), r(t))^T$.
-
-The SIR model is described by the following system of ODEs (ordinary differential equations)~\autocite[19]{keelingModeling}:
-\begin{align}
-\frac{ds(t)}{dt} &= -\lambda(t) s(t) \\
-\frac{di(t)}{dt} &= \lambda(t) s(t) - \gamma i(t) \\
-\frac{dr(t)}{dt} &= \gamma i(t)
-\end{align}
-where $\lambda(t) = \beta i(t)$, is the \emph{force of infection}: the risk of infection for a susceptible individual at time $t$~\autocite[17]{keelingModeling}.
-The dynamics of the model are described by the parameters $\vec{\theta} = (\beta, \gamma)^T$.
-$\beta$ is the transmission rate, absorbing several terms as described below.
-$\gamma$ is the recovery rate, or the rate that infectious individuals that recover per day; hence, $1/\gamma$ is the mean infectious period~\autocite[367]{keelingModeling}.
-These parameters are assumed to be constant.
-
-% The term $\beta s(t)i(t)$ is the rate at which susceptible individuals (as a proportion of the total population size) are infected~\autocite[18]{keelingModeling}.
-% Susceptible individuals move from the $s(t)$ to $i(t)$ when they are infected.
-The expression for the force of infection, $\lambda(t)$ can be arrived at by considering a single infectious individual~\autocite[214]{kretzschmarMathematical}.
-Assume this individual comes into contact with other individuals in the population at a mean rate of $\kappa$.
-Further, assume the population is \emph{well-mixed}, meaning that each contact is with an individual chosen uniformly at random from the population; this is implied by assuming a homogeneous population.
-For such a randomly selected individual, the probability that they are susceptible is $s(t)$.
-% Each of the infectious individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
-Therefore the infectious individual makes contact with susceptible individuals at a mean rate of $\kappa s(t)$ per day.
-Assume that there is a constant probability of transmission upon contact between an infectious and susceptible individual, $q$.
-Hence, the mean rate of infections caused by this infectious individual is $\kappa s(t) q$ per day.
-The two constant terms are absorbed into the transmission rate, $\beta = \kappa q$.
-Therefore, the rate of new infections generated by a single infectious individual is $\beta s(t)$ per day.
-Multiplying by the number of infectious individuals and dividing by the population size gives the rate per day at which susceptible individuals are infected as a proportion of the population, $\beta s(t) i(t) = \lambda(t) s(t)$.
-
-% The term $\gamma i(t)$, the rate at which infectious individuals (as a proportion of the total population size) recover, follows directly from the definitions of $\gamma$ and $i(t)$.
-
-A few properties and assumptions of the SIR model are worth noting.
-\begin{itemize}
-    \item For any short period of time when the change in $s(t)$ is negligible, the solution to the system of ODEs includes exponential growth in the number of infectious individuals: $i(t) = i_0 e^{\psi t}$ for some constants $i_0$ and $\psi$~\autocite[section 1.2]{diekmannMathematical}.
-    A notable example is the early part of an epidemic when $s(t) \approx 1$.
-    This is the phase of initial exponential growth of an epidemic, producing the characteristic early epidemic curves.
-    It is in this phase that the basic reproduction number provides a good approximation of the epidemic dynamics.
-    \item The basic reproduction number for the SIR model is $\R = \beta / \gamma$~\autocite[20]{keelingModeling}.
-    Intuitively, this can be thought of as the mean number of infections caused by an infectious individual per day, $\beta$, multiplied by the mean number of days for which they are infectious, $1/\gamma$.
-    \item The effective reproduction number for this model is $R_e(t) = \beta / \gamma s(t) = \R s(t)$~\autocite{pellisEstimation}.
-    $1-s(t)$ is the fraction of infections that are avoided because of the presence of immune individuals in the population.
-    \item The population is assumed to be closed.
-    This assumption implies several attributes.
-    First, the total population size is constant.
-    Second, only a negligible number of infections are imported from outside the population.
-    Third, births and deaths of individuals are negligble over the length of time considered; model extensions allow these factors in contexts where they are relevant~\autocites[26]{keelingModeling}[214]{kretzschmarMathematical}.
-    \item There is no analytical solution to the system of ODEs: numerical methods are required to solve the system~\autocite[25]{keelingModeling}.
-    \item As noted previously, the parameters $\vec{\theta}$ and the intial state of the system, $\vec{x}(0)$ are required to solve the system.
-    \item Immunity is assumed to last until the end of the  period considered ~\autocite[61]{andersonInfectious}.
-    In many contexts, including here, this assumption is justified because the period is shorter than the length of immunity conferred~\autocite{milneImmunity}.
-    For longer time periods, waning immunity can be incorporated by allowing individuals to move from the recovered compartment back to the susceptible compartment~\autocite[40]{keelingModeling}.
-    \item The incidence proportion on day $t$ can be defined as $Z_t / \Npop = s(t) - s(t+1)$.
-However, this will not in general give an integer value to $Z_t$.
-In practice, this is not an issue since a continuous approximation for the incidence will suffice.
-\end{itemize}
-
-\subsubsection{Non-exponential waiting times} \label{SEIR:sec:non-exponential}
-\begin{figure}[h]
-\makebox[\textwidth][c]{
-\begin{tikzpicture}[
-    node distance = 2.5cm,
-    on grid,
-    auto,
-    ->,>=stealth',
-    every state/.style={draw,rectangle},
-    ]
-
-    \node[state] (S) {$s$};
-    \node[state, right=of S] (I1) {$i_1$};
-    \node[state, right=of I1] (I2) {$i_2$};
-    \node[state, right=of I2, draw=none] (I3) {$\cdots$};
-    \node[state, right=of I3] (I4) {$i_n$};
-    \node[state, right=of I4] (R) {$r$};
-
-    \path (S) edge node {$\beta i$} (I1)
-          (I1) edge node {$n\gamma$} (I2)
-          (I2) edge node {$n\gamma$} (I3)
-          (I3) edge node {$n\gamma$} (I4)
-          (I4) edge node {$n\gamma$} (R);
-\end{tikzpicture}
-}
-  \caption[The SIR model with non-exponential infectious period]{Schematic of the SIR model with non-exponential infectious period. The infectious period is modified by having multiple I states.}
-  \label{SEIR:fig:SIR-gamma}
-\end{figure}
-
-The basic SIR model is the deterministic approximation of a stochastic model where the duration of time in the infectious compartment is exponentially distributed~\autocite[96]{keelingModeling}.
-This is an unrealistic assumption for most diseases and would lead to underestimating of the reproduction number~\autocite{lloydRealistic,wearingAppropriate}.
-A standard extension to the SIR model is to instead assume a gamma distribution~\autocite[94]{keelingModeling,andersonSpread}.
-This is done by adding multiple infectious compartments, shown in \cref{SEIR:fig:SIR-gamma}.
-The ODEs are now:
-\begin{align}
-\frac{ds(t)}{dt} &= -\lambda(t) s(t)\\
-\frac{di_1(t)}{dt} &= \lambda(t) s(t) - n\gamma i_1(t) \\
-\frac{di_2(t)}{dt} &= n\gamma i_1(t) - n \gamma i_2(t) \\
-&\vdots \nonumber \\
-\frac{di_n(t)}{dt} &= n\gamma i_{n-1}(t) - n \gamma i_n(t) \\
-\frac{dr(t)}{dt} &= n\gamma i_n(t)
-\end{align}
-where $n$ is the number of infectious compartments, and redefining $\lambda(t) = \beta \sum_{j=1}^n i_j(t)$.
-The additional compartments added do not, in general, correspond to any biological feature or have a biological interpretation; they are purely a mathematical device to modify the distribution of time between becoming and ceasing to be infectious~\autocite{lloydRealistic}.
-In this model the infectious period is distributed $\GamDist(n, n\gamma)$, a realistic model for a variety of diseases~\autocite{wearingAppropriate}.
-This parameterisation is chosen to give $1/\gamma$ as the mean infectious period (further details on the Gamma distribution are in \cref{E-distributions}).
-A gamma distribution with the first parameter equal to 1 is an exponential distribution, in which case this model collapses to the basic SIR model in \cref{SEIR:sec:SIR}.
-As $n$ increases, the between-individual variability in infectious period length reduces, tending to a constant as $n \to \infty$~\autocite{lloydRealistic}.
-
-The reproduction number in terms of the parameters is unchanged by this extension, \ie $\R = \beta / \gamma$.
-
-\subsubsection{The SEIR model} \label{SEIR:sec:SEIR}
-
-\todo[inline]{Maybe merge SEIR intro into the general intro, not introducing SIR first}
-
-\begin{figure}[h]
-\makebox[\textwidth][c]{
-\begin{tikzpicture}[
-    node distance = 2cm,
-    on grid,
-    auto,
-    ->,>=stealth',
-    every state/.style={draw,rectangle},
-    % remember picture, overlay
-    ]
-
-    \node[state, xshift=-2cm] (S) {$s$};
-    \node[state, right=of S] (E1) {$e_1$};
-    \node[state, right=of E1] (E2) {$e_2$};
-    \node[state, right=of E2, draw=none] (E3) {$\cdots$};
-    \node[state, right=of E3] (E4) {$e_m$};
-    % \node[state, right=of E4] (I1) {$i_1$};
-    \node[state, below=of E1] (I1) {$i_1$};
-    \node[state, right=of I1] (I2) {$i_2$};
-    \node[state, right=of I2, draw=none] (I3) {$\cdots$};
-    \node[state, right=of I3] (I4) {$i_n$};
-    \node[state, right=of I4] (R) {$r$};
-
-    \path (S) edge node {$\beta i$} (E1)
-          (E1) edge node {$m\sigma$} (E2)
-          (E2) edge node {$m\sigma$} (E3)
-          (E3) edge node {$m\sigma$} (E4)
-          (I1) edge node {$n\gamma$} (I2)
-          (I2) edge node {$n\gamma$} (I3)
-          (I3) edge node {$n\gamma$} (I4)
-          (I4) edge node {$n\gamma$} (R);
-    \draw[->] (E4.east) -- ++(1,0) |- ([yshift=-1cm]E4.east) node [near start, right] {$m\sigma$} -|  (I1.north);
-\end{tikzpicture}
-}
-  \caption[The SEIR model]{Schematic of the SEIR model.}
-  \label{SEIR:fig:SEIR}
-\end{figure}
-
-The SIR model assumes that individuals are immediately infectious upon infection, neglecting the latent period (see \cref{E-biology-data:sec:natural-history}).
-Ignoring the latent period will again overestimate the reproduction number~\autocite{wearingAppropriate}.
-The SIR model can be extended by adding a latent compartment~\autocite[41]{keelingModeling}, also known as an \emph{exposed} compartment.
-This forms the susceptible-exposed-infectious-recovered (SEIR) model, shown in \cref{SEIR:fig:SEIR}.
-
-Similar to the infectious period, the latent period can be modelled as a gamma distribution using multiple latent compartments, $e_1, dots, e_m$.
-\begin{align}
-\frac{ds}{dt} &= -\beta si \\
-\frac{de_1}{dt} &= \beta si - m\sigma e_1 \\
-\frac{de_2}{dt} &= m\sigma e_1 - m \sigma e_2 \\
-&\vdots \nonumber \\
-\frac{de_m}{dt} &= m\sigma e_{m-1} - m \sigma e_m \\
-\frac{di_1}{dt} &= m\sigma e_m - n\gamma i_1 \\
-\frac{di_2}{dt} &= n\gamma i_1 - n \gamma i_2 \\
-&\vdots \nonumber \\
-\frac{di_n}{dt} &= n\gamma i_{n-1} - n \gamma i_n \\
-\frac{dr}{dt} &= n\gamma i_n
-\end{align}
-where $1/\sigma$ is the mean duration of the latent period.
-The latent period is distributed $\GamDist(m, m\sigma)$ and the infectious period is still distributed $\GamDist(n, n\gamma)$.
-
-The reproduction number in terms of the parameters is unchanged by this extension, \ie $\R = \beta / \gamma$, however, the growth rate in the exponential phase will be slower than before~\autocite[41]{keelingModeling}.
-
-\subsubsection{Structured populations} \label{SEIR:sec:structured-populations}
-
-To relax the assumption of a homogeneous population, the population can be stratified.
-If the strata have differing transmission patterns, this can change the dynamics of the epidemic.
-The different patterns could be due to behavioural factors (\eg high-numbers of sexual partners in a sexually-transmitted disease~\autocite[69]{keelingModeling} or younger age groups making many more contacts for respiratory diseases~\autocite[176]{andersonInfectious}), or biological factors such as genetics~\autocite[208]{andersonInfectious}.
-I denote the set of strata, as $\set{A}$, with the set's size as $A$.
-I will assume that the disease progression (\ie length of the exposed and infectious periods) is the same for all strata, although this assumption can be relaxed.
-Stratification requires two major  model changes: extension of the state vector and a description of the between-strata interactions.
-
-The state vector is now a vector of vectors, one for each of the compartments
-For example, in the basic SIR model, $\vec{x}(t) = (\vec{s}(t), \vec{i}(t), \vec{r}(t))^T$.
-The $a$th element of each vector ($a \in \set{A}$) is the proportion of strata $a$ in that compartment.
-For example, $\vec{s}(t) = (s_1(t), \dots, s_A(t))^T$ where $s_a(t)$ is the proportion of the $a$th strata that are susceptible at time $t$.
-Hence, $N_a s_a(t)$ is the number of susceptible people in stratum $a$, where $N_a$ is the number of individuals in stratum $a$.
-These vectors replace the scalars in the previous equations.
-Other, equivalent formulations of this model exist within the literature.
-For example defining $s_a$ as the proportion of the total population that are both in strata $a$ and susceptible or parameterised using numbers of individuals~\autocite[57]{keelingModeling}, rather than as the proportion of strata $a$ as I have done here.
-
-The interactions between strata are included by replacing the scalar $\beta$ with a WAIFW (\emph{Who Acquires Infection From Whom}) matrix, $\matr{\beta}$~\autocite[58]{keelingModeling}.
-The force of infection is now a vector, defined as $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i_j}(t)$.
-Their interpretation can be arrived at by considering a single infectious individual in strata $a'$, analogous to the single stratum case.
-Assume any such individual comes into contact with individuals in strata $a$ at a rate $\kappa_{aa'}$ per day.
-% Further, assume that each contact is with an individual chosen uniformly at random from within the strata.
-% For such a randomly selected individual, the probability that they are susceptible is $s_a(t)$.
-% Each of the infectious individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
-Following the same argument as in \cref{SEIR:sec:SIR} finds that the infectious individual makes contact with susceptible individuals in stratum $a$ at a mean rate of $\kappa_{aa'} s_a(t)$ per day.
-Now assume that there is a constant probability of transmission upon contact between an infectious individual in strata $a'$ and susceptible in strata $a$ individual, $q_{aa'}$.
-Therefore, the mean rate of infections in strata $a$ per day caused by this infectious individual is $\kappa_{aa'} s_a(t) q_{aa'}$ per day.
-Multiplying by the number of infectious individuals in strata $a'$, $N_{a'} i_{a'}(t)$, and dividing by the size of strata $a$, $N_a$, gives $\kappa_{aa'} q_{aa'} s_a(t) \frac{1}{N_a} N_{a'} i_{a'}(t)$ where $i_{a'}(t) = \sum_{j=1}^n i_{j,a'}$ is the total proportion of stratum $a'$ that is infectious, regardless of which compartment they are in.
-The constant terms here are absorbed into the WAIFW matrix, hence $\beta_{aa'} = \kappa_{aa'} q_{aa'} \frac{1}{N_a} N_{a'}$~\autocite[section 9.2]{diekmannMathematical}.
-This gives the total force of infection on strata $a$ from all strata is, therefore, $\lambda_a(t) = \sum_{a' \in \set{A}} \beta_{aa'} i_{a'}(t)$ as above.
-If all $\beta_{aa'}$ are equal, then the model is equivalent to the homogeneous model (\ie without any structure).
-The rate at which individuals move between the susceptible and exposed compartments is now $\vec{s}(t) \circ \vec\lambda(t)$, where $\circ$ denotes element-wise multiplication.
-
-In practical situations, estimating all the $\beta_{aa'}$ parameters is infeasible, so some assumptions must be made to reduce the number of parameters~\autocite[176]{andersonInfectious}.
-It is infeasible because the progression of the epidemic depends on $\vec{\lambda}(t)$, which has dimension $A$, while $\matr{\beta}$ has dimension $A^2$.
-One common assumptions is symmetry, \ie $\beta_{aa'} = \beta_{a'a}$.
-% Alternatively, as in the scalar case, it may be easier to iden $\beta_{aa'}$ can be decomposed into the contact rate between strata $a$ and $a'$, $\kappa_{aa'}$ and the probability of transmission upon contact, $q_{aa'}$, and then assumptions made about the components.
-The choice of assumptions depends on the context, including the disease being studied and the data available.
-
-Reproduction numbers in models with structured populations are more complex than the homogeneous case.
-The WAIFM allows calculation of the number of secondary infections generated by a single infection in the $a$th strata,  this follows directly from its definition as $\frac{1}{\gamma} \sum_{a' \in \set{A}} \beta_{aa'}$.
-However, how to average these contributions to calculate the basic reproduction number is not obvious.
-The standard solution to this was proposed by \textcite{diekmannDefinition} (see \textcite[chapter 7]{diekmannMathematical} for a more gentle, but still rigorous, explanation).
-First, they show that the distribution of infections across the strata during the initial period of exponential growth (when $\vec{s} \approx \vec{1}$) relies only on $\matr{\beta}$, and not the starting conditions.
-This distribution is proportional to the \emph{dominant eigenvector} (the eigenvector corresponding to the largest eigenvalue) of $\matr{\beta}$.
-Denote by $D(\matr{\beta})$ the \emph{dominant eigenvalue} (the largest eigenvalue) of $\matr{\beta}$.
-The basic reproduction number is then defined as $\R = D(\matr{\beta}) / \gamma$.
-This $\R$ keeps the same interpretation as in the homogeneous case, \ie the average number of secondary infections caused by a single infection in a fully susceptible population, where the average is taken over the distribution implied by the dominant eigenvector.
-The important threshold that exponential growth occurs iff $\R > 1$ remains.
-
-Age-based contact patterns are important for the transmission dynamics of respiratory diseases~\autocite{birrellBayesian,jacksonEffects}.
-Respiratory diseases can be spread by conversational contacts, which show very clear age patterns (see \cref{SEIR:fig:age-contacts}).
-In particular, the contacts are \emph{associative} (contacts are more likely between individuals of similar ages) and younger age groups have more contacts; this will tend to concentrate the epidemic in these younger age groups~\autocite[67]{keelingModeling}.
-\begin{figure}
-  \makebox[\textwidth][c]{\includegraphics[width=0.9\paperwidth]{inc-prev/contact_matrices}}
-  \caption[Age-based contact matrices]{Number of contacts reported between age groups in two contact surveys: (A) number of contacts reported in the POLYMOD survey~\autocite{mossongSocial}; (B) number of contacts reported in the BBC Pandemic survey~\autocite{klepacContacts}, which only collected data on those aged over 12.
-A clear diagonal, showing associative mixing is visible, and children have more contacts.
-Parallel diagonals off the main diagonal show mixing between parents and their children.
-Figure adapted from \textcite{klepacContacts}, licensed under \href{https://creativecommons.org/licenses/by-nc-nd/4.0/}{CC BY-NC-ND 4.0.}}
-  \label{SEIR:fig:age-contacts}
-\end{figure}
-
-\todo[inline]{Does something need to link these sections?}
-
-\section{Prevalence process} \label{inc-prev:sec:prevalence-process}
-
-The relationship between incidence and prevalence can be derived by considering the probability that an individual is prevalent at any time following an infection.
-%I define the period the individual is prevalent for as their \emph{infection episode}.
-%I index infection episodes with $i$.
-%Denote the time that the infection episode begins as $B_i$ and the time it ends as $E_i \geq B_i$.
-%The individual is prevalent during the interval $[B_i, E_i]$ (\ie including $B_i$ and $E_i$).
-%The duration of the infection episode, the number of days for which they are prevalent, is $D_i = E_i - B_i + 1$.
-%I assume that the $D_i$s are iid, with a discrete distribution defined by $\prob(D_i = d) = f_D(d)$ for $d = 1, \dots, d_\text{max}$, for some maximum episode length $d_\text{max}$, and $\prob(D_i = d) = 0$ otherwise.
-%Denote by $F_D$ the cdf of $D_i$, that is $F_D(d) = \prob(D_i \leq d)$.
-From this, we can derive the relationship between incidence and prevalence.
-
-Denote by $R_{t,t'}$ the number of infection episodes with $B_i=t$ and $E_i=t'$.
-For any infection beginning at $t$, the probability that it ends at $t'$ is $\prob(E_i = t' \mid B_i = t) = \prob(D_i = t' - t + 1 \mid B_i = t) = f_D(t' - t + 1)$.
-Since these durations are assumed independent, the number of episodes of each duration is multinomially distributed~\autocite{paganoHIV}.
-Specifically:
-\begin{align}
-\begin{bmatrix}
-  R_{t,t} \\ R_{t,t+1} \\ \vdots \\ R_{t,t+d_\text{max}-1}
-\end{bmatrix} \mid Z_t
-\sim \MN \left(
-  Z_t, 
-  \begin{bmatrix}
-    f_D(1) \\ f_D(2) \\ \vdots \\ f_D(d_\text{max})
-  \end{bmatrix}
-\right).
-\end{align}
-Now, denote by $P_{t_z,t_p}$ the number of individuals infected at $t_z$ and prevalent at $t_p$.
-Formally, $P_{t_z,t_p}$ is the infection episodes with $B_i = t_z$ and $E_i \geq t_p$.
-$P_{t_z,t_p}$ is $Z_{t_z}$ minus those that recover before $t_p$ (\ie at or up to $t_p - 1$).
-Hence:
-\begin{align}
-    P_{t_z,t_p} = \begin{cases}
-      0 &t_p < t_z\\
-      Z_{t_z} - \sum_{i=0}^{t_p-t_z-1} R_{t_z,t_z+i} &t_z \leq t_p < t_z + d_\text{max}\\
-      0 &t_p \geq t_z + d_\text{max}.
-  \end{cases} \label{inc-prev:eq:Ptt-to-R}
-\end{align}
-Finally, the $P_t$ can be expressed as the total number of people prevalent at time $t$, regardless of when they were infected.
-The limits of the sum are limited by the zeroes in \cref{inc-prev:eq:Ptt-to-R}.
-\begin{align}
-  P_t
-  &= \sum_{t_z=-\infty}^\infty P_{t_z,t} \\
-  &= \sum_{i=0}^{\dmax-1} P_{t-i,t} \label{inc-prev:eq:Pt-to-Ptt} \\
-  &= \sum_{i=0}^{\dmax-1} \left(Z_{t_z} - \sum_{j=0}^{i-1} R_{t-i,t-i+j} \right) &\text{by \cref{inc-prev:eq:Ptt-to-R}}\label{inc-prev:eq:Pt-to-Rtt}.
-\end{align}
-
-In general, all the random variables in this section are dependent.
-This is because they are linked by the $Z_t$s, which are themselves dependent (as explained in \cref{inc-prev:sec:infection-process}).
-However, if we condition on the vector of incidence, $\vec{Z} = (Z_1, Z_2, \dots)^T$, then we will induce some independence.
-This reduces the stochasticity to only that attributable to the prevalence process.
-In particular, because the duration of infection episodes is independent, $R_{t_z,t_p}$ and $R_{t_z',t_p'}$ will be independent if $t_z \neq t_z'$; importantly, this holds even if $t_p = t_p'$.
-Therefore, $P_{t_z,t_p}$ and $P_{t_z',t_p'}$ will be independent if $t_z \neq t_z'$.
-
-In \cref{inc-prev:sec:observation-process} I will show that only $\E(P_t \mid \vec{Z})$ and bounds on $\V(P_t \mid \vec{Z})$ will be relevant for inferring incidence.
-I know derive these from their constituent parts.
-
-Start by considering the recoveries.
-The sum of recoveries of infections that occurred on the same day $t$ is the sum of multinomial cell probabilities, and hence binomially distributed~\autocite{alamAnalysis}.
-Specifically:
-\begin{align}
-  \sum_{i=0}^{t'} R_{t,t+i} \mid \vec{Z} &\sim \text{Binomial}(Z_t, F_D(t'+1)). \label{inc-prev:eq:binomialRt}
-\end{align}
-Therefore:
-\begin{align}
-  \E \left( \sum_{i=0}^{t'} R_{t,t+i} \mid \vec{Z} \right) &= Z_t F_D(t'+1) \label{inc-prev:eq:EsumRt} \\
-  \V \left( \sum_{i=0}^{t'} R_{t,t+i} \mid \vec{Z} \right) &= Z_t F_D(t'+1) (1 - F_D(t'+1)) \label{inc-prev:eq:VsumRt}
-\end{align}
-by the properties of the binomial distribution (see \cref{E-distributions}).
-
-The first moment of $P_t$ follows.
-\begin{align}
-\E(P_t \mid \vec{Z})
-  &= \E\left(\sum_{i=0}^{d_\text{max}-1} \left( Z_{t-i} - \sum_{j=0}^{i-1} R_{t-i,t-i+j} \right) \mid \vec{Z} \right) &\text{by \cref{inc-prev:eq:Pt-to-Rtt}}\\
-  &= \sum_{i=0}^{d_\text{max}-1} \left( Z_{t-i} - \sum_{j=0}^{i-1} \E( R_{t-i,t-i+j} \mid \vec{Z}) \right) \\
-  &= \sum_{i=0}^{d_\text{max}-1} \left( Z_{t-i} - Z_{t-i} F_D(i) \right) &\text{by \cref{inc-prev:eq:EsumRt}}\\
-  &= \sum_{i=0}^{d_\text{max}-1} Z_{t-i} (1 - F_D(i)) \\
-  &= \sum_{i=0}^{d_\text{max}-1} Z_{t-i} S(i+1) \label{inc-prev:eq:EPt}.
-\end{align}
-
-The second moment is simplified by the independence.
-\begin{align}
-\V(P_t \mid \vec{Z})
-  &= \V\left(\sum_{i=1}^{d_\text{max}-1} P_{t-i,t} \mid \vec{Z} \right) &\text{by \cref{inc-prev:eq:Pt-to-Ptt}} \\
-  &= \sum_{i=1}^{d_\text{max}-1} \V\left(P_{t-i,t} \mid \vec{Z} \right) &\text{by the conditional independence} \\
-  &= \sum_{i=1}^{d_\text{max}-1} \V\left(\sum_{j=0}^{i-1} R_{t-i,t-i+j} \mid \vec{Z} \right) &\text{by~\cref{inc-prev:eq:Ptt-to-R}}\\
-  &= \sum_{i=1}^{d_\text{max}-1} Z_{t-i} F_D(i) (1 - F_D(i)) &\text{by~\cref{inc-prev:eq:VsumRt}} \\
-  &\leq \sum_{i=1}^{d_\text{max}-1} Z_{t-i} (1 - F_D(i)) &\text{as $F_D \leq 1$}\\
-  &= \E(P_t \mid \vec{Z}). \label{inc-prev:eq:boundVPt}
-\end{align}
-
-% The probability of being positive can be modelled in different ways depending on what assumptions are most reasonable.
-% The most important assumption from a statistical perspective is the whether the probability of testing positive on a given day is independent conditional on the infection time.
-
-% \todo[inline]{Is the discussion of this model just a distraction?}
-% Assuming that the probability of testing positive depends only on the infection time implies no individual variation.
-% This model is formalised as $Pr(Z_i(t) = 1 \mid B_i, Z_i(1), Z_i(2) \dots) = Pr(Z_i(t) \mid B_i) = p_{t-B_i}$, that is the probability of individual $i$ testing positive $t$ days after infection depends only on the infection start time and no other quantities.
-% A slight relaxation of this model is introducing covariates, allowing some variation based on an individual's characteristics.
-
-% The model with the most dependence between times, while remaining realistic, is a multi-state model.
-% Therefore, if an individual has tested negative before the present time $t$ but after $B_i$ then they will test negative at time $t$.
-% The complexity of the transitions between positive and negative can be very complex.
-% I consider a slight extension to allow for the possibility of false negatives, meaning a negative test between $B_i$ and $E_i$, and that this probability may change.
-% This leads to the following model.
-% \begin{align}
-%   \prob(Z_i(t) = 1 \mid B_i, \leq E_i) &= \begin{cases}
-%     \psens(t - B_i) &B_i \leq t \leq E_i\\
-%     0 &\text{otherwise}
-%   \end{cases}\\
-%   \prob(E_i = t \mid B_i) = f_i(B_i)
-% \end{align}
-% Conditional on both $B_i$ and $E_i$, then $Z_i(t)$ and $Z_i(t')$ are independent for $t \neq t'$.
-% However, unlike the previous model, conditioning only on $B_i$ is not sufficient for $Z_i(t)$ and $Z_i(t')$ to be independent because each $Z_i(t)$ provides information on $E_i(t)$ (in particular, $Z_i(t) = 1$ implies $E_i > t$).
-% For SARS-CoV-2, the latter model is more appropriate, although, as we shall see, the assumption that $\psens$ is independent of $t$ is violated.
-
-The expectation in \cref{inc-prev:eq:EPt} is a discrete convolution equation.
-The equation is of the same form as the equation used for HIV/AIDS backcalculation, as well as being found in many other fields~\autocite[and references therein]{brookmeyerBackcalculation}.
-
-Considering multiple values of $t$ makes \cref{inc-prev:eq:EPt} into a system of simultaneous equations.
-The system can be written in the form $\E(\vec{P} \mid \vec{Z}) = \matr{S} \vec{Z}$ where $\matr{S}$ is a lower-triangular matrix formed from $S$:
-\begin{align}
-    (\matr{S})_{i,j} = \begin{cases}
-        S(i - j + 1) & i - \dmax - 1 \leq j \leq i \\
-        0 & \text{otherwise.}
-    \end{cases}
-\end{align}
-In \cref{E-backcalc}, I will need to solve for this system for $\vec{Z}$ given a known $\hat{P} = \E(\vec{P} \mid {Z})$.
-As $\matr{S}$ is lower-triangular, I use forward-substitution~\autocite{cormenMatrix}.
-Forward-substitution sequentially solves the $t$th element of $\vec{Z}$ ($t = 1, \dots, T$) by calculating:
-\begin{align}
-Z_t
-&= \frac{\hat{P}_t - \sum_{i=1}^{t-1} S_{ti} Z_i}{S_{tt}} \\
-&= \hat{P}_t - \sum_{i=1}^{t-1} S(t - i + 1) Z_i.
-\label{inc-prev:eq:forward-substitute}
-\end{align}
-
-% Numerical methods for solving systems of this form, that is solving for $\vec{Z}$ (non-random in this expression) for a known $\E (\vec{P} \mid \vec{Z} )$, are well-established~\autocite[e.g.:][section 8.2]{highamAccuracy}.
-% However, these algorithms can be unstable, meaning that numerical errors are large.
-% This is a particular issue if $S(1)$ is small.
-% Since, by definition, $S(1)=1$ here, this is not an issue in this context.
-%However, if instead of estimating the incidence of RT-PCR positives, the aim was to estimate the incidence of infections, then $S(1)$ would be small and this would be an issue.
-%While this quantity is more epidemiologically relevant, the time from infection to RT-PCR positivity is short and, for now, I will neglect the difference.
-
-Unlike an inference procedure which infers the parameters simultaneously, there is no feedback loop: the value of $Z_{t}$ does not affect $Z_{t'}$ for $t' < t$.
-The second term here is the contribution of the incidence prior to day $t$ to day $t$'s prevalence.
-The remainder of the prevalence on day $t$ is therefore due to incidence on day $t$.
-% Forward substitution is numerically stable unless the diagonal elements, here $S(1)$, are small relative to the off-diagonal elements, here $S(i)$ for $i > 0$~\autocite[section 8.2]{highamAccuracy}.
-% Here, the diagonal is $S(1) = 1$ and hence the matrix the solution is numerically stable.
-
-The variance in \cref{inc-prev:eq:boundVPt} is upper bounded by \cref{inc-prev:eq:EPt}, a fact that will prove useful in \cref{inc-prev:sec:observation-process}.
-\Cref{inc-prev:eq:EPt} shows that knowledge of the survival function, $S$, is crucial to modelling the prevalence process.
-
-\section{Observation process} \label{inc-prev:sec:observation-process}
-
-% A \emph{prevalence survey} is a survey that tests a sample of the population for the presence of an infection.
-% Series of prevalence surveys are the form of measuring prevalence that this thesis is concerned with.
-% This thesis aims to infer the incidence rate from prevalence surveys.
-% I use Bayesian inference, therefore the data contributes through the likelihood.
-The previous sections have explained the processes leading to the population prevalence.
-However, we do not directly observe the population prevalence but a sample of it.
-The sampling is described mathematically by the observation process.
-The survey design considered here reflects the CIS, although it would be applicable to many prevalence surveys.
-
-On day $t$ (for $t = 1, \dots, T$), a prevalence survey such as the CIS samples $n_t$ individuals at random from the population of interest and tests if they are prevalent.
-Therefore, I model the number of positive tests, $X_t$, as $X_t \mid P_t, n_t, \Npop \dist \Bin(n_t, P_t/\Npop)$.
-The conditioning on $n_t$ and $\Npop$ is implicit in what follows.
-These quantities are assumed to be fixed by the study design and population choice, and independent of $P_t$ and $Z_t$.
-
-The $X_t$s are conditionally independent, given the population prevalences at each time, $\vec{P} = (P_1, \dots, P_t)^T$. 
-%For now, I assume that the sample is chosen uniformly at random from the population of interest.
-%Assume that the surveys are independent of each other and $\vec{Z}$, conditional on the population prevalence at that time.
-%The population prevalence is a latent quantity.
-Prevalence surveys are used in this thesis to infer incidence.
-I base this inference on the likelihood, $p(\vec{x} \mid \vec{Z})$.% = \int p(\vec{x} \mid \vec{P}) p(\vec{P} \mid \vec{Z}) d\vec{P} = \int \prod_t p(x_t \mid P_t) p(\vec{P} \mid \vec{Z}) d\vec{P}$, where $\vec{x} = (x_1, x_2, \dots)^T$.
-Here $\vec{x} = (x_1, \dots, x_T)$ is the vector of observed number of positive tests in the prevalence survey on each day and $\vec{Z} = (Z_{-\dmax}, Z_{-\dmax+1} \dots, Z_T)$ is the vector of incidence on each day.
-The incidence back to day $-\dmax$ will affect $P_1$ and therefore need to be included in $\vec{Z}$.
-
-This section argues that $p(\vec{x} \mid \vec{Z}) \approx \prob(\vec{X} = \vec{x} \mid \vec{P} = \E(\vec{P} \mid \vec{Z}))$.
-As the $x_t$s are independent conditional on $\vec{P}$, it follows that $p(\vec{x} \mid \vec{Z}) \approx \prod_t \prob(X_t = x_t \mid P_t = \E(P_t \mid \vec{Z}))$.
-This relationship will make inference for $\vec{Z}$ much simpler because the latent vector $\vec{P}$ and the dependence of the $x_t$s can be ignored.
-
-The basis of the argument is that the first two moments of $\vec{x} \mid \vec{Z}$ and $\vec{X} \mid \vec{P} = \E(\vec{P} \mid \vec{Z})$ have negligible difference.
-The first moment follows directly from the tower property of expectations.
-The majority of the argument is to show that the variance-covariance matrix $\V(\vec{X} \mid \vec{Z}) \approx \V(\vec{X} \mid \vec{P} = \E(P \mid Z))$.
-There are two steps to this argument.
-I start with the diagonal elements, the variances.
-I show that $\V(X_t \mid \vec{Z}) \approx \V(X_t \mid \vec{P} = \E(\vec{P} = \vec{Z}))$.
-The conditional independence of the $x_t$s implies that the conditional covariances are 0.
-Therefore, I show that the correlation between $X_t$ and $X_{t'}$ conditional only on $\vec{Z}$ is negligible.
-
-Starting with the variances, by the law of total variance:
-\begin{align}
-  &\var\left( X_t \mid \vec{Z} \right) \\
-    &= \var\left[\E\left( X_t \mid \vec{Z}, \vec{P} \right) \mid \vec{Z} \right] + \E\left[\var\left( X_t \mid \vec{Z}, \vec{P} \right) \mid \vec{Z} \right] \\
-    &= \var\left( n_t \frac{P_t}{\Npop} \mid \vec{Z} \right) + \E\left( n_t \frac{P_t}{\Npop} \left(1 - \frac{P_t}{\Npop} \right) \mid \vec{Z} \right) \\
-    &= \left( \frac{n_t}{\Npop} \right)^2 \var\left(P_t \mid \vec{Z} \right) + \frac{n_t}{\Npop} \E\left(P_t \mid \vec{Z} \right)  - \frac{n_t}{\Npop^2} \E\left(P_t^2 \mid \vec{Z} \right) \\
-    &= \left( \frac{n_t}{\Npop} \right)^2 \var\left(P_t \mid \vec{Z} \right) + \frac{n_t}{\Npop} \E\left(P_t \mid \vec{Z} \right)  - \frac{n_t}{\Npop^2} \left(\E\left(P_t \mid \vec{Z} \right) ^ 2 + \V(P_t \mid \vec{X})\right) \\
-    &= \frac{n_t(n_t - 1)}{\Npop^2} \var\left(P_t \mid \vec{Z} \right) + n_t \frac{\E\left(P_t \mid \vec{Z} \right)}{\Npop}\left(1 - \frac{\E\left(P_t \mid \vec{Z} \right)}{\Npop} \right) \\
-    &= \frac{n_t(n_t - 1)}{\Npop^2} \var\left(P_t \mid \vec{Z} \right) + \V(X_t \mid P_t = \E(P_t \mid \vec{Z})).
-\end{align}
-The first term of this expression is negligible in the contexts I consider.
-These contexts have $n_t << \Npop$.
-$\Npop$ is the population of England (around 56 million) while $n_t$ is the daily sample size (10s of thousands).
-Further $\V(P_t \mid \vec{Z}) \leq \E(P_t \mid \vec{Z})$ (\cref{inc-prev:eq:boundVPt}).
-Therefore, $\var\left( X_t \mid \vec{Z} \right) \approx \V(X_t \mid P_t = \E(P_t \mid \vec{Z}))$.
-
-A similar argument will show that the correlation between $x_t$ and $x_{t'}$ ($t \neq t'$) is negligible.
-The basic structure of this argument is as the previous one.
-The result follows because the observation noise dominates, and this is uncorrelated.
-In the following, all are implicitly conditioned on $\vec{Z}$.
-However, for clarity this is omitted from the notation.
-\begin{align}
-  &\lvert \cor(X_t, X_{t'}) \rvert \\
-  &= \frac{\lvert\cov(X_t, X_{t'})\rvert}{\sqrt{\var(X_t) \var(X_{t'})}} &\text{by definition}\\
-  &= \frac{\lvert \E(\cov(X_t, X_{t'} \mid \vec{P})) + \cov(\E(X_t \mid \vec{P}), \E(X_{t'} \mid \vec{P})) \rvert}{\sqrt{\var(X_t) \var(X_{t'})}} &\text{by law of total covariance} \\
-  &= \frac{\lvert 0 + \cov(n_t P_t / \Npop, n_{t'} P_{t'} / \Npop) \rvert}{\sqrt{\var(X_t) \var(X_{t'})}} \\
-  &= \frac{n_t n_{t'} \lvert \cov(P_t, P_{t'}) \rvert}{\Npop^2 \sqrt{\var(X_t) \var(X_{t'})}}  \\
-  &<< \frac{n_t n_{t'} \lvert \cov(P_t, P_{t'}) \rvert}{\Npop^2 \sqrt{\var(P_t) \var(P_{t'})}} &\text{as $\V(X_t) >> \V(P_t)$} \\
-  &= \frac{n_t n_{t'}}{\Npop^2} \lvert \cor(P_t, P_{t'}) \rvert &\text{by def of $\cor$}\\
-  &\leq \frac{n_t n_{t'}}{\Npop^2} &\text{as $\lvert \cor(\cdot) \rvert \leq 1$}.
-\end{align}
-$n_t n_{t'} / \Npop^2$ is very small, for the same reason that $n_t / \Npop$ is small.
-The argument above shows that $\lvert \cor(X_t, X_{t'}) \rvert$ is much smaller than this.
-Therefore, neglecting the correlation between $X_t$ and $X_{t'}$ is a good approximation.
-
-This section showed that, when using a prevalence survey such as CIS, the likelihood $p(\vec{x} \mid \vec{z})$ is very well approximated by $p(\vec{x} \mid \vec{P} = \E(\vec{P} \mid \vec{z}))$.
-
-\section{Markov chain Monte Carlo} \label{inc-prev:sec:MCMC}
-
-MCMC is a class of algorithms for sampling from a distribution where the normalising constant for the distribution is unknown.
-I will use it to sample a set of parameters, say $\psi$ (which may be a scalar or vector), from a posterior distribution $p(\psi \mid y)$ where $y$ is the data (again, scalar or vector).
-MCMC algorithms produce a sample $\psi^1, \psi^2, \dots, \psi^M$ from $p(\psi \mid y)$ by constructing a Markov chain with $p(\psi \mid y)$ as its stationary distribution~\autocite[275]{gelmanBDA}.
-It can be proved that as $M\to\infty$, the sample tends towards a sample from the stationary distribution, \ie the target posterior.
-This section outlines some basic principles of MCMC and considerations when using it in practice.
-Any reader interested in more technical details or proofs of the results here should consult the references within this section.
+\section{Metropolis-Hastings}
 
@@
 
+\section{Improving MCMC efficiency}
+
 MCMC using a Normal proposal distribution can be inefficient.
@@
 
-Two practical issues arise when applying MCMC: \emph{convergence} and \emph{stability}~\autocite[72]{lunnBUGS}.
+\section{Convergence}
 
-Convergence is how quickly the Markov chain approaches the stationary distribution.
-Early draws from the Markov chain reflect the choice of the starting point, $\psi^0$, rather than the stationary distribution that the chain will eventually converge to~\autocite[282]{gelmanBDA}.
 The simplest technique for assessing convergence is graphically, using traceplots~\autocite[81]{brooksMCMCNotes}.
@@
 A more complex measure of convergence when running multiple MCMC chains is comparing the within-chain variance to the between-chains variance.
+This is the basis of the Rhat statistic used throughout this thesis (see \cref{E-intro:sec:MCMC}).
 The principle here is that if all chains have converged then they are sampling from the same distribution.
 Therefore, the within-chain and between-chain variances are the same~\autocite[82]{brooksMCMCNotes}.
-Various metrics quantify thiVarious metrics quantify this I will make use of the improved Rhat statistic~\autocite{vehtariRhat}.
-%The Rhat statistic is defined for a scalar parameter $\psi$.
-%Let $\psi^{(ij)}$ denote the $i$th (of $M$) sample from the $j$th chain (of $N$).
-%Let $\bar{\psi}^j$ denote the mean value of the samples of $\psi$ from the $j$th chain, $\bar{\psi}^j = \frac{1}{M} \sum_{i=1}^M \psi^{(ij)}$.
-%Let $\bar{\psi}$ denote the mean value of the samples of $\psi$ from all chains, $\bar{\psi} = \frac{1}{N} \sum_{j=1}^N \bar{\psi}^j$.
-%Then the between-chain variance is $B = \frac{M}{N-1} \sum_{j=1}^N (\bar{\psi}^j - \bar{\psi})^2$.
-%The within-chain variance is $W = \frac{1}{N} \sum_{j=1}^N \frac{1}{M-1} \sum_{i=1}^M (\psi^{(ij)} - \bar{\psi}^j)^2$.
-A Rhat of 1 means that the within-chain and between-chain variance are equal, informally that convergence is perfect, as far as can be measured by this technique.
-Various rules-of-thumb exist for what should be considered an acceptable Rhat.
-Historically, 1.1 has commonly been recommended, although more recent work suggests that 1.01 or 1.05 is more appropriate~\autocite{vehtariRhat}.
-
-The stability of estimates considers the error arising from only using a finite sample of the distribution and the samples being autocorrelated.
-Autocorrelation means that estimates using the sample have higher variance than if it were an iid sample from the posterior~\autocite[286]{gelmanBDA}.
-The \emph{effective sample size} corrects for the autocorrelation.
-The effective sample size can be interpreted as the number of iid samples that would have the same variance as the MCMC sample~\autocites[286]{gelmanBDA}{vehtariRhat}.
-Various techniques have been proposed for computing the effective sample size, I will follow the approach of \textcite{vehtariRhat}, implemented in RStan~\autocite{RStan-2-32-3}.
-
-\section{Conclusion} \label{inc-prev:sec:conclusion}
-
-Robust, unbiased prevalence estimates already exist (\cref{E-biology-data:sec:positivity}), and incidence estimates are the objective of this thesis.
-This chapter has shown that the key quantity relating these quantities is the distribution of the duration of RT-PCR positivity.
-However, no estimates of this quantity in the general population exist\todo{ref lit review of duration}.
-Therefore, I now turn to estimating the duration of RT-PCR positivity.
-
-\ifSubfilesClassLoaded{
-  \listoftodos
-}{}
 

### 2024-02-28 incidence-prevalence.tex
start outling intro
write most of intro

@@
 \chapter{Statistical and epidemiological background} \label{inc-prev}
+
+\todo[inline]{%
+  Chapter 3 needs a better title: this is not all the statistical and epidemiological background and some of it is mathematical modelling / disease dynamics.
+  Perhaps: ``Data generating procsses and disease dynamics''.
+  Would need to define disease dynamics.
+}
 
@@
 This model is guided by the specific study design of the CIS.
-
-Finally, \cref{inc-prev:sec:MCMC} will introduce Markov chain Monte Carlo (MCMC), an inference technique I will make extensive use of in this thesis.
 
@@
     Intuitively, this can be thought of as the mean number of infections caused by an infectious individual per day, $\beta$, multiplied by the mean number of days for which they are infectious, $1/\gamma$.
-    \item The effective reproduction number for this model is $R_e(t) = \beta / \gamma s(t) = \R s(t)$~\autocite{pellisEstimation}.
+    \item The effective reproduction number for this model is $\Re = \beta / \gamma s(t) = \R s(t)$~\autocite{pellisEstimation}.
     $1-s(t)$ is the fraction of infections that are avoided because of the presence of immune individuals in the population.
@@
 
-\section{Markov chain Monte Carlo} \label{inc-prev:sec:MCMC}
-
-MCMC is a class of algorithms for sampling from a distribution where the normalising constant for the distribution is unknown.
-I will use it to sample a set of parameters, say $\psi$ (which may be a scalar or vector), from a posterior distribution $p(\psi \mid y)$ where $y$ is the data (again, scalar or vector).
-MCMC algorithms produce a sample $\psi^1, \psi^2, \dots, \psi^M$ from $p(\psi \mid y)$ by constructing a Markov chain with $p(\psi \mid y)$ as its stationary distribution~\autocite[275]{gelmanBDA}.
-It can be proved that as $M\to\infty$, the sample tends towards a sample from the stationary distribution, \ie the target posterior.
-This section outlines some basic principles of MCMC and considerations when using it in practice.
-Any reader interested in more technical details or proofs of the results here should consult the references within this section.
-
-A common way of constructing a Markov chain with $p(\psi \mid y)$ as its stationary distribution is using the Metropolis-Hastings algorithm~\autocite{hastingsMCMC}.
-The algorithm requires a proposal distribution $q(\psi' \mid \psi)$, which proposes that the chain moves to a new state $\psi'$ from its current state $\psi$; $\pi(\psi)$, a calculation of $p(\psi \mid y)$ up to some constant.
-The proposal is then either accepted or rejected based on the \emph{acceptance ratio}, which is computed using $\pi$.
-The full algorithm is given in \cref{inc-prev:algorithm:MH}, with further details in available in a wide-range of sources~\autocites[e.g.][]{brooksMCMCNotes}[chapter 11]{gelmanBDA}.
-\begin{algorithm}
- set $\psi^0$ to some initial value\;
- \For{$i = 1, \dots, M$}{
-  sample a proposal $\psi' \sim q(\psi' \mid \psi)$ \;
-  calculate the acceptance probability $\alpha(\psi, \psi') = \min \left( 1, \frac{\pi(\psi') q(\psi \mid \psi')}{\pi(\psi) q(\psi' \mid \psi)} \right)$ \;
-  with probability $\alpha(\psi, \psi')$ set $\psi^i$ to $\psi'$, otherwise set $\psi^i$ to $\psi^{i-1}$ \;
- }
- \caption{The Metropolis-Hastings algorithm.}
- \label{inc-prev:algorithm:MH}
-\end{algorithm}
-
-The proposal distribution $q$ is most commonly a (possibly multivariate) Normal distribution centred on $\psi$; this scheme is used in \cref{E-SEIR}.
-Tuning the proposal distribution to propose ``good'' moves is a key part of MCMC.
-The ideal proposal distribution is the posterior of interest, however, this is impossible because that is what MCMC is trying to estimate~\autocite[296]{gelmanBDA}.
-One alternative is \emph{adaptive} algorithms which tune the proposal distribution based on previous draws in an attempt to approximate the posterior; I use such a scheme in \cref{E-SEIR:sec:inference}.
-
-MCMC using a Normal proposal distribution can be inefficient.
-In particular, it often poorly explores the tails of the posterior in finite time~\autocite[e.g.][]{turitsynIrreversible}.
-\emph{Hamiltonian Monte Carlo} (HMC)~\autocite{nealMCMC,nealImproved,duaneHybrid} improves the proposals by using information from the gradient of the posterior density (taken with respect to the parameters).
-HMC simulates a physical system travelling through the posterior space with momentum.
-The momentum increases in the direction the posterior density increases, using the gradient.
-This means that the sampler tends to propose moves with higher posterior density.
-The Metropolis-Hastings algorithm is then used to accept or reject the proposed move.
-
-The \emph{No U-Turn Sampler} (NUTS)~\autocite{hoffmanNUTS} is a variant of HMC that automatically tunes the Hamiltonian dynamics hyperparameters.
-NUTS has been implemented in a variety of software packages making it easy to apply to arbitrary posterior distributions.
-I will make use of the implementation in the package \emph{Sampling through adaptive neighborhoods} (Stan)~\autocite{Stan-2-32-2}.
-Stan implements \emph{automatic differentiation}~\autocite{griewankAutoDiff} to calculate the gradient of the posterior density without any user input.
-Automatic differentiation uses the chain rule to calculate the gradient of a function from its code by breaking it down into a series of elementary operations, for which the gradient is known~\autocite{autodiff}.
-For a more detailed introduction to HMC, NUTS, and Stan see \textcite[chapter 12]{gelmanBDA}.
-
-Two practical issues arise when applying MCMC: \emph{convergence} and \emph{stability}~\autocite[72]{lunnBUGS}.
-
-Convergence is how quickly the Markov chain approaches the stationary distribution.
-Early draws from the Markov chain reflect the choice of the starting point, $\psi^0$, rather than the stationary distribution that the chain will eventually converge to~\autocite[282]{gelmanBDA}.
-The simplest technique for assessing convergence is graphically, using traceplots~\autocite[81]{brooksMCMCNotes}.
-Traceplots show the value of $\psi^i$ against $i$.
-If the chain has converged, then the traceplot will show a random walk around the stationary distribution.
-This can be improved by using multiple chains with different starting points~\autocite[283]{gelmanBDA}.
-If all chains have converged, then they will be in the same location.
-
-A more complex measure of convergence when running multiple MCMC chains is comparing the within-chain variance to the between-chains variance.
-The principle here is that if all chains have converged then they are sampling from the same distribution.
-Therefore, the within-chain and between-chain variances are the same~\autocite[82]{brooksMCMCNotes}.
-Various metrics quantify thiVarious metrics quantify this I will make use of the improved Rhat statistic~\autocite{vehtariRhat}.
-%The Rhat statistic is defined for a scalar parameter $\psi$.
-%Let $\psi^{(ij)}$ denote the $i$th (of $M$) sample from the $j$th chain (of $N$).
-%Let $\bar{\psi}^j$ denote the mean value of the samples of $\psi$ from the $j$th chain, $\bar{\psi}^j = \frac{1}{M} \sum_{i=1}^M \psi^{(ij)}$.
-%Let $\bar{\psi}$ denote the mean value of the samples of $\psi$ from all chains, $\bar{\psi} = \frac{1}{N} \sum_{j=1}^N \bar{\psi}^j$.
-%Then the between-chain variance is $B = \frac{M}{N-1} \sum_{j=1}^N (\bar{\psi}^j - \bar{\psi})^2$.
-%The within-chain variance is $W = \frac{1}{N} \sum_{j=1}^N \frac{1}{M-1} \sum_{i=1}^M (\psi^{(ij)} - \bar{\psi}^j)^2$.
-A Rhat of 1 means that the within-chain and between-chain variance are equal, informally that convergence is perfect, as far as can be measured by this technique.
-Various rules-of-thumb exist for what should be considered an acceptable Rhat.
-Historically, 1.1 has commonly been recommended, although more recent work suggests that 1.01 or 1.05 is more appropriate~\autocite{vehtariRhat}.
-
-The stability of estimates considers the error arising from only using a finite sample of the distribution and the samples being autocorrelated.
-Autocorrelation means that estimates using the sample have higher variance than if it were an iid sample from the posterior~\autocite[286]{gelmanBDA}.
-The \emph{effective sample size} corrects for the autocorrelation.
-The effective sample size can be interpreted as the number of iid samples that would have the same variance as the MCMC sample~\autocites[286]{gelmanBDA}{vehtariRhat}.
-Various techniques have been proposed for computing the effective sample size, I will follow the approach of \textcite{vehtariRhat}, implemented in RStan~\autocite{RStan-2-32-3}.
-
 \section{Conclusion} \label{inc-prev:sec:conclusion}

### 2024-02-28 introduction.tex
write most of intro

--- 2024-02-19 introduction.tex
@@
 
-\begin{itemize}
-    \item Disease burden of pandemic and ongoing burden of COVID
-    \item To mitigate its effects, need to understand incidence and transmission but cannot observe directly
-    \item Importance of surveillance objectives and concepts: incidence, transmission, basic reproduction number, effective reproduction number
-    \item Expected future pandemics, methods should be reusable
-\end{itemize}
+COVID-19 (coronavirus disease 2019), the disease caused by the virus SARS-CoV-2 (severe acute respiratory syndrome coronavirus 2), is estimated to have killed approximately 15 million in the acutest phase (to the end of 2021) and 30 million globally~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
+While it is tempting to consider this a rare or extreme event, history suggests otherwise (see \cref{intro:fig:pandemic-timeline}).
+\begin{figure}
+    \includegraphics[width=\textwidth]{introduction/pandemic-timeline}
+    \caption[Timeline of pandemics.]{%
+        Timeline of pandemics.
+        See \url{https://ourworldindata.org/historical-pandemics} for details and sources.
+        Licensed under \href{https://creativecommons.org/licenses/by/4.0/}{CC-BY} by Saloni Dattani, Klara Auerbach, Marwa Boukarim, and Max Roser.
+    }
+    \label{intro:fig:pandemic-timeline}
+\end{figure}
 
-% - Natural history of SARS-CoV-2 infection
-% 	- Stylised graphic of phases: infection, pre-infection positive, infectious positive, post-infectious positive, negative
-% 	- How it spreads
-% - Why is incidence and prevalence estimation challenging?
-% 	- Asymptomatic infection
-% 	- Ascertainment bias
-% - How do we estimate prevalence
-% 	- CIS and REACT
-% - Issues with current incidence estimation
+Fully quantifying the burden of historical pandemics is challenging due to the absence of systematic records.
+However, our best records suggests that, over the last 400 years, they have (on average) occurred every four years with a mean death toll of 5.6 million~\autocite{maraniNovelEpidemics}.
+Modern estimates suggest that we should expect over 2.5 million deaths per year from pandemics~\autocite{madhavPandemicMortality}.
 
-\section{SARS-CoV-2 real-time surveillance}
+\todo[inline]{Maybe talk about pandemic v epidemic v outbreak terminology?}
+\todo[inline]{Need to motivate the importance of surveillance}
 
-\begin{itemize}
-    \item Most commonly used types of data: severe events (deaths/hospitalisations), widespread testing (pillar 2), 
-    \item Advantages of designed surveys (REACT, CIS)
-    \item Use of cohorts (ATACCC)
-    \item Statistical modelling performed during the pandemic. Discuss broadly but especially focus on those using CIS or REACT. Discuss my contribution in this context.
-\end{itemize}
+\section{Surveillance metrics}
 
-\subsection{Modelling}
+No single metric summarises the full state of an epidemic, and no single data source fully informs all metrics of interest~\autocite{royalSocietyRnumber,pellisEstimation,paragGrowthRates}.
+Therefore, a range of metrics and surveillance systems are required to understand the state of an epidemic and the danger of a pathogen (a pathogen is any biological agent that can cause disease).
+
+\subsection{Growth of an epidemic}
+
+Various forms of \emph{reproduction number} were prominent throughout the COVID-19 pandemic~\autocite{pellisEstimation}.
+Broadly, a reproduction number describes the number of secondary infections generated by a single infection.
+Here, following \textcite{pellisEstimation} (who themselves are building on over a century of infectious disease dynamics literature), I define three that are broadly useful.
+
+The most famous reproduction number is the \emph{basic reproduction number}, $\R$.
+$\R$ is the mean number of secondary infections generated by a single infection in a \emph{fully susceptible} population with \emph{standard behaviour patterns}.
+A fully susceptible population means none of its members have any immunity to the disease.
+Note that this definition is a constant for a given pathogen and population.
+In the early stages of an epidemic caused by a novel pathogen, the number of individuals with immunity is negligible, and hence $\R$ describes this part of the epidemic well.
+$\R$ has a threshold value of 1: if $\R < 1$ then the epidemic will die out.
+
+The \emph{effective reproduction number}, $\Re$, is the mean number of secondary infections generated by a single infection at time $t$.
+In particular, it uses the behaviour and immunity status at time $t$.
+With constant behaviour, $\Re$ declines smoothly as individuals in the population become immune due to infection.
+$\Re$ is useful because it describes the current dynamics of the epidemic: $\Re > 1$ means the epidemic is growing, $\Re < 1$ means it is shrinking, and $\Re = 1$ means it is stable.
+In the simplest epidemic models (see \cref{E-SEIR:sec:SIR}), $\Re$ is proportional to the product of the rate of contacts and the proportion of the population that is susceptible.
+Therefore, it informs the strength of intervention (either reducing contacts or amount of immunisation) required to control the epidemic.
+
+The \emph{control reproduction number}, $\Rc$, is the mean number of secondary infections generated by a single infection with the behaviour at time $t$ but in a fully susceptible population.
+$\Rc$ is also known as the \emph{reproduction number excluding immunity}.
+It describes the effect on the epidemic of interventions (except vaccination) that have been put into place.
+$\Rc$ is useful when comparing the effects of interventions at different points in time, when the immunity status of the population has changed.
+For example, when comparing the English lockdown in March 2020 and Janurary 2021, there are two effects in play: that there are more individuals with immunity in January 2021, and that the lockdown had different measures in place.
+$\Rc$ focuses on the latter effect.
+
+These definitions ignore many complexities that arise when theoretically defining or practically estimating any of the reproduction numbers~\autocite{pellisEstimation}.
+For example, effects due to changing weather, what a ``mean infection'' means in a heterogeneous population, and the lack of a ground truth for these numbers in almost all practical situations.
+However, their intuitive simplicity and convenient properties in simple epidemic models (see \cref{E-SEIR:sec:mechanistic-models}) mean that they are useful concepts for guiding policy.
+
+Reproduction numbers are strongly connected to \emph{growth rates}.
+Over short periods of time, epidemics tend to grow exponentially, and hence have an exponential growth rate, normally denoted $r$.
+For a given disease and population, an analogous growth rate can be defined for each reproduction number~\autocite{pellisEstimation,paragGrowthRates,wallingaGI}.
+The advantage of using the growth rate is that they are easier to estimate and capture how quickly spread happens in calendar time; however, they do not capture the strength of intervention required to control the epidemic~\autocite{royalSocietyRnumber}.
+
+\subsection{Size of an epidemic}
+
+Growth rates and reproduction numbers only provide information on how an epidemic is changing.
+Yet the size of the epidemic is at least as important as how it is changing~\autocite{pellisEstimation}.
+\emph{Incidence} (the number of new infections in a given time period) is the primary metric for the magnitude of a pandemic.
+Incidence informs the burden of the disease, for example, the future number of severe events (hospitalisations or deaths).
+
+\todo[inline]{What is a good definition of ``transmission'' as a concept?}
+The \emph{transmission} of an infectious disease generates new infections.
+The components of transmission (\eg number of contacts per day and age-stratified rates) inform how interventions can be targeted.
+They are also important for forecasting future incidence.
+
+However, incidence and transmission are almost never directly observed.
+Therefore, they need to be estimated from other data sources.
+
+\emph{Prevalence} is the proportion of the population with the disease at a given time.
+Prevalence can be observed, most commonly with a prevalence survey.
+
+See \cref{E-inc-prev:sec:definitions} for a formal definition of incidence and prevalence.
+Prevalence and incidence are strongly related, with the relationship determined by how long the disease lasts (see \cref{E-inc-prev}).
+
+To translate incidence into the burden of the disease, a measure of severity is required.
+For example, the IFR (infection fatality rate), the proportion of infected individuals who die.
+Estimation of severity is beyond the scope of this thesis.
+% Severity does not feature in this thesis, but it is important to note that it is a key component of understanding the burden of a pandemic.\todo{maybe reword sentence}
+
+\section{Data sources}
+
+A wide variety of designed studies, administrative data, and other data sources have been used for pandemic and epidemic surveillance.
+In the context of the SARS-CoV-2 pandemic in England, the most prominent were: community testing (pillar 2), the number of severe events (\eg hospital admissions or deaths), seroprevalence testing (seroprevalence is the proportion of people with antibodies to a disease, a marker of prior infection), and prevalence surveys.
+Each of these data sources have strengths and weaknesses, see \textcite{royalSocietyRnumber} for further details.
+
+Community testing is broadly available, but does not test a random sample of the population.
+The number of tests conducted varies over time and space, for a variety of known and unknown reasons.
+These include: policy on who is eligible for testing, the availability of tests, and the willingness of individuals to be tested.
+Therefore, fluctuations in the number of positive tests may not reflect changes in the true incidence of the disease.
+For large parts of the pandemic, community testing was widely available in the UK meaning that this is a large dataset, allowing detailed analyses to be performed, including stratification by small geographic areas or age groups.
+
+Seroprevalence testing is a marker of recent prior infection.
+Prior infection is more common than current infection, hence seroprevalence estimation requires a smaller sample size than current prevalence.
+However, there is a delay between infection and the development of antibodies, hence seroprevalence is not a real-time measure of the burden of the disease.
+
+Severe events are generally a reliable measure of the burden of the disease.
+The criteria for being a severe event are well-defined and generally well-recorded in the UK context.
+Therefore, the data is subject to fewer biases than community testing.
+However, SARS-CoV-2 rarely leads to severe events in younger individuals, meaning severe events only capture transmission to a subset of the population.
+
+Prevalence surveys in representative samples of the population are the most reliable measure of the burden of the disease.
+They are not subject to the biases of community testing and observe the whole population, unlike severe events.
+However, they are expensive to power to inform detailed analyses.
+Often, some modelling is required to borrow information across strata or time to provide stable estimates (see \cref{E-intro:sec:cis}).
+
+The use of prevalence surveys for estimating incidence and transmission is the focus of this thesis.
+In particular, I focus on the CIS (Coronavirus (COVID-19) Infection Survey), run by the Office for National Statistics (ONS)~\autocite{CIS,cisMethodsONS}.
+The CIS was a longitudinal study, testing a large number of indidividuals weekly or monthly for SARS-CoV-2; it is described in detail in \cref{E-intro:sec:cis}.
+Early in the pandemic, a methodology was developed for using CIS data to estimate prevalence (see \cref{E-intro:sec:cis}).
+The primary aim of this thesis is to estimate incidence and transmission from this data.
+
+\section{Estimating incidence and transmission} \cref{intro:sec:estimating-incidence}
+
+As noted above, incidence and transmission are extremely important quantities for informing the response to a pandemic but cannot be observed directly.
+Therefore, they must be estimated using some form of modelling.
+
+Many methods for estimating incidence and transmission were developed in the context of the HIV/AIDS pandemic.
+% HIV is transmitted sexually or through blood and causes the disease AIDS.
+This ongoing pandemic has killed 40 million people, peaking at 2 million annual deaths~\autocite{unaids2023}.
+The large burden of the disease has motivated many advances in infectious disease epidemiology and biostatistics, including several which form the methodological basis of this thesis.
+
+\subsection{Backcalculation}
+
+\emph{Backcalculation} is a statistical procedure for estimating incidence based on a measurement of prevalence.
+It was proposed by \textcite{brookmeyerMethod} in the context of HIV/AIDS.
+In this context, backcalculation estimates the incidence of HIV infection from data on the number of AIDS diagnoses, and the time between infection and AIDS diagnosis~\autocite{brookmeyerBackcalculation,brookmeyerMeasuring}.
+Two processes needed to be considered in the context of HIV/AIDS: one generating infections and one relating infections to prevalence.
+% AIDS is incurable, therefore, AIDS prevalence is equivalent to cumulative incidence.
+
+The statistical framework developed in the context of HIV transmission is applicable to SARS-CoV-2.
+This infection process can be described as a Poisson process (see \cref{E-inc-prev:sec:infection-process}).
+
+Unlike HIV/AIDS, recovery from SARS-CoV-2 happens in the vast majority of infections.
+Therefore, recoveries need to be considered in the backcalculation.
+This introduces a dependency structure in the prevalence data (see \cref{E-inc-prev:sec:prevalence-process}).
+The development of this framework makes clear that the duration of positivity (how long an individual tests positive for SARS-CoV-2) is the key quantity relating prevalence to incidence.
+Therefore, estimation of duration is central to this thesis (see \cref{intro:sec:duration}).
+
+Furthermore, prevalence surveys are noisy because they are based on population sampling.
+This noise needs to be considered in the application of backcalculation (see \cref{E-inc-prev:sec:observation-process}).
+
+The original formulations of backcalculation neglected the observation process.
+AIDS is almost always diagnosed and recorded in developed countries.
+That is, the observations of AIDS diagnoses are the same as the true incidence of AIDS.
+An exception is that the reporting delay may be important~\autocite{paganoHIV}; but reporting delay is not important for this thesis's work.
+Since its development, backcalculation has been further developed for a wide-variety of contexts.
+\todo[inline]{probably need to cite some review or developments of backcalculation or give a brief lit review? not sure how much details is worthwhile as its not directly relevant}
+% Reporting delay is not relevant to the studies in this thesis, and I ignore it in this and further discussion.
+
+\Textcite{abbottCISincidence} applied backcalculation to CIS data to estimate the incidence of SARS-CoV-2 in the UK.
+They used a Gaussian process prior to smooth the incidence, applying the prior to either smooth the log-incidence or its derivative; smoothing of the incidence is required to stabalise the estimates~\autocite{brookmeyerBackcalculation}.
+\Textcite{abbottCISincidence} used \textcite{hellewellPCRSensitivity}'s estimates of the sensitivity of testing to inform the backcalculation, which are based on a longitudinal study in healthcare workers (see \cref{intro:sec:duration-SARS-CoV-2}).
+They fit to summary statistics of the CIS data, using a daily, independent Gaussian likelihood with parameters estimated from the posterior point estimate and 95\% credible interval published by the ONS.
+
+The estimates I present in \cref{E-backcalc} are methodologically similar.
+However, I fully propagate the uncertainty from the prevalence estimation procedure and use estimates of the duration of positivity from a general population study.
+\Textcite{abbottCISincidence} include the short period of low but increasing test sensitivity at the start of an infection (see \cref{E-biology-data:sec:natural-history}).
+
+\subsection{Mechanistic models}
+
+\subsubsection{Background}
+
+Mechanistic models make assumptions on the biological and epidemiological mechanisms driving transmission~\autocite{lesslerMechanistic}.
+Therefore, the model's state and parameters have a biological and/or epidemiological interpretation.
+% Of particular interest is that the basic reproduction number, $\R$, and the effective reproduction number, $R_e(t)$, which can both be calculated as functions of the model parameters; using \cref{E-backcalc}'s approach, additional modelling would be required to estimate these numbers.
+In particular, this allows changes in transmission to be decomposed into the contributing factors, \eg the number of contacts per day or the probability of transmission upon contact (see \cref{E-SEIR:sec:mechanistic-models}).
+The ability to estimate the contribution of each of these factors can inform public health policy.
+Furthermore, mechanistic models can be used for scenario-based modelling, which simulates the effect of interventions.
+A scenario here is the effect of a proposed intervention (or lack thereof), introduced into the model by modifying the relevant parameters.
+% The mechanistic model can then be simulated forward in time to understand the intervention's effects.
+% All of these properties make mechanistic models a useful tool for understanding and controlling infectious diseases.
+% However, they make stronger assumptions about the disease and population behaviour than statistical models.
+
+The literature on mechanistic models is vast, having been developed for over a century.
+The model I introduce in \cref{E-SEIR:sec:SIR} was first formulated by \textcite{kermackContribution}.
+However, the earliest mechanistic model to be formulated mathematically was probably \textcite{rossMalariaA}'s model of malaria transmission~\autocite{lesslerMechanistic}.
+This early contribution led to an important insight for public health, showing that malaria could be controlled even if mosquitoes were not fully eliminated.
+
+\Textcite{kermackContribution} was a seminal paper for the field, explaining the observation that epidemics ended before all individuals were infected due to a population-level build-up of immunity known as \emph{herd immunity}.
+Their model has inspired much of the subsequent literature, a full review of which is beyond the scope of this thesis.
+The relevant background information is explained in \cref{E-SEIR:sec:mechanistic-models}.
+
+\subsubsection{Application to SARS-CoV-2}
+
+Many mechanistic modelling studies were conducted during the pandemic, with at least 100 articles published in the first six months of 2020 alone~\autocite{shankarSystematic}.
 
@@
     \item Literature on mechanistic models using REACT and CIS I have previously circulated.
-    \item \textcite{mccabeCISincidence,abbottCISincidence} on backcalculation.
-    \item Also Taffe?
 \end{itemize}
 
-\subsection{Surveillance data}
+\subsection{Other approaches}
 
-The first study, ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts), performed daily testing of a small number of individuals with short follow-up.
-The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability, but the results are limited because the small sample size and short follow-up means that the tails of these distributions cannot be estimated.
+Other approaches have been taken to estimate incidence and/or transmission from the CIS data.
 
-The second study, the CIS (Coronavirus (COVID-19) Infection Survey), performed weekly or monthly testing of a large number of individuals with long follow-up.
-The large sample size and long follow-up allows for reliable estimates of the tails of the distributions of the quantities of interest.
-However, the results are limited because the sparse sampling which means that the viral load dynamics cannot be estimated.
-In particular, the vast majority of detected infections had only one or two positive tests, and hence the viral load trajectory of these infections cannot be estimated.
+\Textcite{mccabeCISincidence}, applying the methodology of \textcite{ealesAppropriately}, estimate $\Re$ and the epidemic's growth rate from the CIS data.
+They estimate a posterior distribution of the parameters of a spline, penalising deviations from a constant first derivative.
+The growth rate is then calculated from the derivative of the spline, an analytical transformation of the parameters.
+Since, under reasonable model assumptions, there is a one-to-one correspondence between reproduction numbers and growth rates~\autocite{wallingaGI}, this also estimates $\Re$.
 
-\section{Historical context}
+\section{Estimating duration} \label{intro:sec:duration}
 
-\begin{itemize}
-    \item Infectious disease epidemiology and biostatistics heavily developed in context of HIV/AIDS epidemic
-    \item Use much methodology and concepts from this field
-    \item Surveillance in general (incidence)
-\end{itemize}
+\subsection{Methodology}
 
-\subsection{Surveillance}
+Two approaches are commonly used to estimate the duration of a disease: either modelling an underlying biomarker or modelling the duration directly~\autocite{sweetingEstimating}.
+For SARS-CoV-2, both approaches would be based on the results of RT-PCR testing (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}).
 
-\begin{itemize}
-    \item Reproduction numbers
-    \item Incidence: almost never observed directly
-\end{itemize}
+Modelling a biomarker, as I do in \cref{E-ATACCC}, models how the amount of virus in an individual varies over time.
+The RT-PCR tests commonly used give a proxy for viral load for positive tests (see \cref{E-biology-data:sec:PCR-process}).
+Modelling a biomarker uses more information per observation, but requires stronger assumptions.
+For example, a functional form of the behaviour over time of the biomarker is required.
+In addition, if individual-level variation wants to be included, which is generally important, a distributional form of random effects is required.
+This requires both sufficient observations per individual and an assumption on the form of the distribution.
 
-\subsection{Backcalculation}
-
-\emph{Backcalculation} is the procedure for estimating incidence based on a measurement of prevalence.
-\Textcite{brookmeyerMethod} developed it in the context of HIV (human immunodeficiency virus) and AIDS (acquired immune deficiency syndrome).
-HIV is transmitted sexually or through blood and causes the disease AIDS.
-It has killed 40 million people, peaking at 2 million annual deaths~\autocite{unaids2023}.
-In this context, backcalculation estimates the incidence of HIV infection from: data on the number of AIDS diagnoses, and the incubation distribution (the time between infection and AIDS diagnosis)~\autocite{brookmeyerBackcalculation,brookmeyerMeasuring}.
-AIDS is incurable, therefore, AIDS prevalence is equivalent to cumulative incidence.
-
-The original formulations of backcalculation neglected the observation process.
-AIDS is almost always diagnosed and recorded in developed countries.
-That is, the observations of AIDS diagnoses are the same as the true incidence of AIDS.
-An exception is that the reporting delay may be important~\autocite{paganoHIV}.
-Reporting delay is not relevant to the studies in this thesis, and I ignore it in this and further discussion.
-Therefore, two processes needed to be considered in the context of HIV/AIDS: infection and prevalence.
-
-For SARS-CoV-2, measurements of prevalence are noisy because they are based on population sampling.
-The sampling is described as the observation process.
-The observation process noise is much larger than the noise of the prevalence process (see \cref{E-inc-prev:sec:observation-process}).
-Therefore, the latter is negligible and again two processes need to be considered: infection and observation.
-
-\begin{itemize}
-    \item Duration is important
-    \item Mathematical details in \cref{E-inc-prev}
-    \item Infection process well established, what needs adapting for here (recoveries, sampling)? See first-year report.
-\end{itemize}
-
-\subsection{Mechanistic models}
-
-In this chapter, I introduce mechanistic models for estimating the transmission of SARS-CoV-2, an alternative to the statistical approach in \cref{E-backcalc}.
-Mechanistic models have a structure which make assumptions on the biological and epidemiological mechanisms behind the infection process (see \cref{E-inc-prev:sec:infection-process}), and hence the disease dynamics~\autocite{lesslerMechanistic}.
-Individuals are giving the model's state and parameters a biological and/or epidemiological interpretation.
-% Of particular interest is that the basic reproduction number, $\R$, and the effective reproduction number, $R_e(t)$, which can both be calculated as functions of the model parameters; using \cref{E-backcalc}'s approach, additional modelling would be required to estimate these numbers.
-In particular, this allows changes in transmission to be decomposed into the contributing factors, \eg the number of contacts per day or the probability of transmission upon contact.
-The ability to estimate the contribution of each of these factors can inform public health policy.
-Furthermore, mechanistic models can be used for scenario-based modelling, which simulates the effect of interventions.
-A scenario here is the effect of a proposed intervention (or lack thereof), introduced into the model by modifying the relevant parameters.
-The mechanistic model can then be simulated forward in time to understand the intervention's effects.
-All of these properties make mechanistic models a useful tool for understanding and controlling infectious diseases.
-However, they make stronger assumptions about the disease and population behaviour than statistical models.
-
-The literature on mechanistic models is vast, having been developed for over a century.
-The model in \cref{E-SEIR:sec:SIR} was first formulated by \textcite{kermackContribution}.
-However, the earliest mechanistic model to be formulated mathematically was probably \textcite{rossMalariaA}'s model of malaria transmission~\autocite{lesslerMechanistic}.
-These early contributions led to important insights for public health.
-\Textcite{rossMalariaA} showed that, if mosquitoes were controlled but not eliminated, malaria could be controlled.
-\Textcite{kermackContribution} explained the observation that epidemics ended before all individuals were infected due to a population-level build-up of immunity known as \emph{herd immunity}.
-
-A full review of this literature is beyond the scope of this thesis.
-In this chapter, I will discuss only the background relevant to the application.
-For further details I recommend the tutorial paper \textcite{kretzschmarMathematical} or the textbook \textcite{keelingModeling}.
-
-% I start this chapter by introducing a generic framework for mechanistic modelling of infectious diseases (\cref{E-SEIR:sec:transmission-generic}).
-% % This framework is then applied to SARS-CoV-2 in \cref{E-SEIR:sec:transmission-application}.
-% It describes the \emph{transmission model}, corresponding to the infection process.
-% The transmission model needs to be linked to observations; in this chapter, that will be the CIS prevalence data.
-% The \emph{observation model}, introduced in \cref{E-SEIR:sec:observation}, is this link; it incorporates both the prevalence process and the observation process.
-% These two components form the mathematical model, which I then fit to data, requiring specification of an inference process, parameterisation, and priors (\cref{E-SEIR:sec:inference-implementation}).
-% This final formulation uses the transmission model from \textcite{birrellRealtime}, but a novel observation model.
-% I then use a simulation study to check this recovers the parameters successfully (\cref{E-SEIR:sec:sim-study}).
-% Having shown that the model can recover the parameters, I then apply it to the CIS data (\cref{E-SEIR:sec:application}).
-% Finally, I discuss the results, limitations, and extensions of this work (\cref{E-SEIR:sec:discussion}).
-
-\subsection{Duration} \label{inc-prev:sec:duration}
-
-Two approaches are commonly used to estimate durations: either modelling an underlying biomarker or modelling the duration directly~\autocite{sweetingEstimating}.
-For SARS-CoV-2, both approaches would be based on the results of RT-PCR testing.
-Modelling a biomarker would mean modelling the viral load, measured as a Ct value (see \cref{E-ATACCC}).
-Modelling the duration would consider only the binary result at each test (positive or negative); in this thesis a \emph{survival analysis} framework is adopted (see \cref{E-perf-test,E-imperf-test}).
+Modelling the duration directly, as I do in \cref{E-perf-test,E-imperf-test} considers only the binary result at each test (positive or negative); in this thesis a \emph{survival analysis} framework is adopted.
+This approach generally requires fewer parameters, and no random effect, reducing the computational burden.
+It is also easier to specify non-parameteric forms of the distribution, which is important for SARS-CoV-2 as the tail behaviour may be important.
+A more detailed review of the relevant literature is in the relevant chapters.
 % Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
 Other related approaches, such as a multi-state model~\autocite{jacksonMSM}, could also be considered (multi-state models can be viewed as a generalisation of survival analysis).
-Therefore, the biomarker approach considers more information per observation, however, a model for the biomarker is required which generally requires stronger assumptions.
-Modelling a biomarker uses more information per observation, but requires stronger assumptions.
 
-Expand on this biomarker vs survival analysis comparison
-\begin{itemize}
-    \item Why do I apply each method?
-    \item Biomarker: stronger assumptions but more information used.
-    \item Literature review on estimates of length of PCR positivity (none in general pop)
-\end{itemize}
+In both cases, longitudinal data is required.
+Therefore, cohort studies, where the same group of indivdiuals are followed over time, are the most appropriate study design.
 
-\section{Notational conventions}
+\subsection{SARS-CoV-2} \label{intro:sec:duration-SARS-CoV-2}
 
-\begin{itemize}
-    \item Lower-case variables are realisations of the respective upper-case random variables.
-    \item Matrices and vectors are bold and have elements with the same unbolded letter, \eg $\vec{x} = [x_1, x_2, \dots]^T$
-    \item $\cdot$ denotes element-wise multiplication
-\end{itemize}
+\subsubsection{Cohort studies}
 
-\section{Thesis contribution}
+In this thesis, I make use of two cohort studies with differing designs.
+These designs have complementary strengths, and I combine information from the studies to make use of these.
+Both of the studies recruited individuals from the general population, aiming to observe a cohort of infections that are representative of the infections in the UK community.
 
-\begin{itemize}
-    \item Overall question: how can prevalence surveys best be used to understand incidence and transmission?
-    \item Inform future study design
-    \item Provide methodology
-\end{itemize}
+% In addition to population-level surveillance, cohort studies help understand other aspects of the pathogen~\autocite{royalSocietyRnumber}.
+% For example, contact tracing data can inform how likely it is for transmission to occur in different settings and/or between different groups of people.
+The first study I use, ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts)~\autocite{singanayagamCommunity,hakkiOnset}, performed daily testing of a small number of individuals with short follow-up.
+The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability, but the results are limited because the small sample size and short follow-up means that the tails of these distributions cannot be estimated.
+I describe ATACCC in detail in \cref{E-biology-data:sec:ataccc} and analyse it in \cref{E-ATACCC}.
 
-The code for the analyses in this thesis is available online.
+The second study is the CIS, which I previously mentioned.
+In addition to estimating prevalence, the CIS followed the same individuals over a long period of time (over a year in many cases), collecting longitudinal data.
+The large sample size and long follow-up mean that the tails of the distribution of the duration of positivity can be estimated.
+However, the lack of frequent testing (at most weekly) means that the viral load dynamics cannot be estimated.
+In particular, the vast majority of detected infections had only one or two positive tests, and hence the viral load trajectory of these infections cannot be estimated.
+I describe the CIS in detail in \cref{E-intro:sec:cis} and analyse it in \cref{E-perf-test,E-imperf-test}.
+
+\subsubsection{Available estimates}
+
+An early meta-analysis estimated that individuals test positive for SARS-CoV-2 for 14.6 days (95\% CI: 9.3--20.0 days)~\autocite{cevikShedding}.
+The maximum observed duration was 83 days, although beyond day 9 no live virus was detected, indicating that the individuals testing positive for longer than this were not infectious.
+However, the majority of the studies included in the meta-analysis involved hospitalised patients or had unclear inclusion criteria, and hence may not be representative of the general population.
+These summary estimates are insufficient for backcalculation, which requires a distribution of the duration of positivity.
+
+\Textcite{hellewellPCRSensitivity} estimated the probability of an individual testing positive at each day after infection.
+This measure is highly related to the duration of positivity.
+The cohort in their study was drawn from healthcare workers, and hence was focused on working-age adults.
+
+\Textcite{binnySensitivity} applied a slight variant of \textcite{hellewellPCRSensitivity}'s methodology to data from New Zealand.
+The situation in New Zealand, who were successfully pursuing a zero-COVID strategy at this time, means that the individuals in the study were unusual.
+For example, a third were travelers detected during their quarantine period on arrival to New Zealand, but would have had negative RT-PCR tests before their departure.
+
+\section{Bayesian inference} \label{intro:sec:Bayes}
+
+All inference in this thesis is performed within the Bayesian framework.
+Bayesian inference has several advantages relevant for this work.
+Prior information allows identifiability in situations where this would not otherwise be possible, allows for combining knowledge from multiple sources, and much prior work I build on has been performed within the Bayesian framework.
+
+Analytically deriving the posterior distribution is impossible for the models in this thesis.
+Therefore, I use computational methods to sample from the posterior distribution.
+The method I use for the majority of the analyses is MCMC (Markov chain Monte Carlo).
+
+\subsection{Markov chain Monte Carlo} \label{intro:sec:MCMC}
+
+MCMC is a class of algorithms for sampling from a distribution where the normalising constant for the distribution is unknown.
+I will use it to sample a set of parameters, say $\psi$ (which may be a scalar or vector), from a posterior distribution $p(\psi \mid y)$ where $y$ is the data (again, scalar or vector).
+MCMC algorithms produce a sample $\psi^1, \psi^2, \dots, \psi^M$ from $p(\psi \mid y)$ by constructing a Markov chain with $p(\psi \mid y)$ as its stationary distribution~\autocite[275]{gelmanBDA}.
+It can be proved that as $M\to\infty$, the sample tends towards a sample from the stationary distribution, \ie the target posterior.
+Further details of MCMC are given in \cref{E-MCMC}.
+Two practical issues arise when applying MCMC: \emph{convergence} and \emph{stability}~\autocite[72]{lunnBUGS}.
+
+Convergence is how quickly the Markov chain approaches the stationary distribution.
+Early draws from the Markov chain reflect the choice of the starting point, $\psi^0$, rather than the stationary distribution that the chain will eventually converge to~\autocite[282]{gelmanBDA}.
+Various metrics quantify this, I will make use of the improved Rhat statistic~\autocite{vehtariRhat}.
+%The Rhat statistic is defined for a scalar parameter $\psi$.
+%Let $\psi^{(ij)}$ denote the $i$th (of $M$) sample from the $j$th chain (of $N$).
+%Let $\bar{\psi}^j$ denote the mean value of the samples of $\psi$ from the $j$th chain, $\bar{\psi}^j = \frac{1}{M} \sum_{i=1}^M \psi^{(ij)}$.
+%Let $\bar{\psi}$ denote the mean value of the samples of $\psi$ from all chains, $\bar{\psi} = \frac{1}{N} \sum_{j=1}^N \bar{\psi}^j$.
+%Then the between-chain variance is $B = \frac{M}{N-1} \sum_{j=1}^N (\bar{\psi}^j - \bar{\psi})^2$.
+%The within-chain variance is $W = \frac{1}{N} \sum_{j=1}^N \frac{1}{M-1} \sum_{i=1}^M (\psi^{(ij)} - \bar{\psi}^j)^2$.
+A Rhat of 1 means that the within-chain and between-chain variance are equal, informally that convergence is perfect, as far as can be measured by this technique.
+Various rules-of-thumb exist for what should be considered an acceptable Rhat.
+Historically, 1.1 has commonly been recommended, although more recent work suggests that 1.01 or 1.05 is more appropriate~\autocite{vehtariRhat}.
+
+The stability of estimates considers the error arising from only using a finite sample of the distribution and the samples being autocorrelated.
+Autocorrelation means that estimates using the sample have higher variance than if it were an iid sample from the posterior~\autocite[286]{gelmanBDA}.
+The \emph{effective sample size} corrects for the autocorrelation.
+The effective sample size can be interpreted as the number of iid samples that would have the same variance as the MCMC sample~\autocites[286]{gelmanBDA}{vehtariRhat}.
+Various techniques have been proposed for computing the effective sample size, I will follow the approach of \textcite{vehtariRhat}, implemented in RStan~\autocite{RStan-2-32-3}.
+
+% \section{Notational conventions}
+
+% \begin{itemize}
+%     \item Lower-case variables are realisations of the respective upper-case random variables.
+%     \item Matrices and vectors are bold and have elements with the same unbolded letter, \eg $\vec{x} = [x_1, x_2, \dots]^T$
+%     \item $\cdot$ denotes element-wise multiplication
+%     \item Distributions used defined in \cref{E-distributions}
+% \end{itemize}
+
+\section{Thesis aims and contribution}
+
+The overall question that I investigate in this thesis is: how can prevalence surveys best be used to understand incidence and transmission?
+Prevalence surveys provide data that is far more representative of the general population than other data sources (see \cref{intro:sec:data-sources}), therefore, conclusions based on them should be far more robust.
+There are two components to this question.
+First, how can prevalence surveys be designed to help estimate these quantities.
+Second, what statistical methodology can be used to estimate these quantities from prevalence surveys.
+This thesis will focus on the second component.
+However, natural extensions of the work I undertake, will inform the first component.
+
+In order to achieve these aims, I use the SARS-CoV-2 epidemic in England from Mon 31st Aug 2020 until Sun 25th Jan 2021 as a case study.
+This case study has several features that make it a useful case study for the aims of this thesis.
+First, the CIS was of sufficient size and incidence sufficiently high that there is enough data to perform the analyses.
+Second, incidence begins rising over this period, which is the period of most interest.
+In particular, an epidemic which will cause issues starts by increasing, therefore if the methods are to be applied early in a future epidemic or pandemic, they will start with an increasing period.
+The CIS began in Aril 2020, when incidence was falling due to the first lockdown in the UK.
+Incidence remained low and fairly flat in England over summer 2020, until it began rising in mid or late August.
+\todo{cite the phases of the UK epidemic}
+Finally, the period ends before the roll-out of the vaccine, which likely affects many important aspects of the analysis.
+In particular, the duration of positivity and the susceptibility of the population.
+Including both of these elements would require more complex modelling but include idiosyncratic features of the SARS-CoV-2 pandemic that are probably less generalisable to future diseases.
+
+A secondary aim of this thesis is to improve our understanding of the epidemic in England over this period, and our understanding of the SARS-CoV-2 virus more broadly.
+The results of the analyses I perform in pursuit of the primary aim will naturally contribute to this aim.
+
+\subsection{Software and data availability}
+
+In order to best fulfill this thesis's primary aim, the methodology developed should be easy for others to use.
+Therefore, the code for the analyses in this thesis is available online.
 An index to the various repositories is available at\todo{insert GitHub link}.
+Two elements of the code have been structured as R packages to enable their application to future contexts.
+
+First, the framework to simulation the CIS (used in \cref{E-imperf-test:sec:simulate,E-perf-test:sec:simulation-study}) is structured as a R package.
+The package enables customisation of various features of the study, for example the testing regime of the participants, the sensitivity of the tests, and the duration of positivity.
+This package would be particularly useful in answering questions regarding the design of future prevalence surveys, for example, how frequently should individuals be tested, or how many individuals should be tested.
+
+Second, the survival analysis framework used to infer the duration of positivity in \cref{E-perf-test,E-imperf-test}.
+The framework allows for all the different analyses performed in these chapters to be specified by the user in a simple interface.
+Furthermore, adding new analyses, for example varying the priors or the inclusion of covariates, is straightforward.
+
+Where possible, aggregated, non-disclosive versions of the datasets are available alongside the code.
+The original ATACCC data used is available alongside the code.
+The original CIS data can only be accessed by accredited researchers within the ONS's SRS (Secure Research Service); see \url{https://www.ons.gov.uk/aboutus/whatwedo/statistics/requestingstatistics/secureresearchservice} for details.
+
 Almost all code, excluding dependencies, was written from scratch for these analyses.
-The exception is the Markov chain Monte Carlo sampler used in \cref{E-SEIR}, which was created by Sanmitra Ghosh for the article \textcite{ghoshApproximate}.
+The exception is the Markov chain Monte Carlo sampler used in \cref{E-SEIR}, which was written by Sanmitra Ghosh for the article \textcite{ghoshApproximate}.
 
-HIGHLIGHT THE FLEXIBLE FRAMEWORK AS A CONTRIBUTION.
-A framework for performing CIS-like simulations is available at \url{https://github.com/joshuablake/cisSimulation}.
-A framework for fitting the class of models described here and in \cref{E-imperf-test} is available at \url{https://github.com/joshuablake/cisDurationModel}.
-Code utilising these frameworks to perform the analysis described in this chapter is available at \url{https://github.com/joshuablake/cisRuns}.
+\subsection{Methodological contributions}
 
-The ATACCC data used is found in the same location.
-CIS data can only be accessed by accredited researchers within the Office for National Statistic's Secure Research Service\todo{insert details of applying to SRS}.
-Where possible, aggregated, non-disclosive versions of the dataset are available alongside the code.
+As its primary aim, this thesis makes several methodological contributions.
+
+First, in \cref{E-inc-prev:sec:observation-process}, I provide a theoretic justification for the use of deterministic backcalculation in this setting.
+A deterministic backcalculation approach was used by \textcite{abbottCISincidence} and I will use it in \cref{E-backcalc}.
+However, it assumes that the prevalence at any time is a random variable, instead modelling it as a deterministic function of the incidence.
+I show that, for all that is required for this assumption to hold is that the proportion of the population sampled is small.
+In particular, I show that there is no requirement for the prevalence or incidence to be large.
+
+Next, in \cref{E-perf-test}, I estimate the duration of RT-PCR-positivity from double censored data with arbitrary periods where infections could be missed (a generalisation of truncation).
+This is based on the CIS data.
+While statistical frameworks have been previously developed which included this case, they had only been applied in special cases.
+In \cref{E-imperf-test}, I extend this framework to include false negatives.
+
+Within these analyses, I include information derived from the ATACCC study.
+In order to do so, I propose a novel prior which appropriately discounts the information from analysing ATACCC study, which has inappropriate model assumptions when applied to the larger sample size in the CIS data.
+
+A particular challenge with these analyses is that they must be performed within the SRS because they require individual-level data.
+The SRS is a form of TRE (Trusted Research Environment), with limited computing power.
+Analyses with long run times (\ie days) and/or requiring large amount of parallelisation (\eg a high-performance cluster) are not feasible.
+
+Finally, in \cref{E-SEIR}, I estimate the incidence of SARS-CoV-2 in England from only the CIS data.
+This includes the development of a novel observation model to link the mechanistic transmission model to the data.
+The observation model allows the inclusion of the long tail in the duration of positivity, without being computationally infeasible.
 
@@
 
-\begin{itemize}
-    \item Estimate of duration of RT-PCR-positivity: of interest for rest of thesis but also more broadly for public health (\eg interpreting RT-PCR test results)
-    \item Use in interpreting test results
-    \item Incidence estimates from only CIS
-    \item Transmission estimates from only CIS
-\end{itemize}
+The secondary aim of this thesis is providing new estimates to improve our understanding of SARS-CoV-2 in general, and the epidemic in England over the period being studied.
 
-\subsection{Methodological contributions}
+In \cref{E-imperf-test}, I present estimates of the duration of RT-PCR-positivity in the general population.
+These estimates could be used to inform public health policy, for example, in interpreting RT-PCR test results or formulating quarantine policies.
 
-\begin{itemize}
-    \item Theory justifying the use of deterministic backcalculation in this setting.
-Already used in practice but without understanding when the approximation is good.
-In particular: all that is required is that the proportion of the population sampled is small, not that the prevalence or incidence is sufficiently large.
-    \item Double interval censoring with arbitrary truncation and false negatives
-    \item Combining ATACCC and CIS duration estimates
-    \item SEIR model using only prevalence surveys
-    \item Application within TRE
-\end{itemize}
+In \cref{E-transmission}, I present estimates of the incidence and transmission of SARS-CoV-2 in England over the period being studied.
+These estimates contribute to the literature reconstructing the SARS-CoV-2 epidemic in England.
+Understanding the dynamics of the past infection is important for informing the response to future pandemics, and the future burden of SARS-CoV-2.
 
@@
 
-Outline the chapters and how they fit together
+In \cref{E-biology-data}, I provide background information on the natural history of SARS-CoV-2, the RT-PCR testing process used to produce the data used throughout this thesis, and describe the two cohort studies used in this thesis.
+Then, \cref{E-inc-prev}, I formally describe the stochastic processes which generate infections, prevalence, and the observations of prevalence.
+A key insight in this chapter is that the distribution of the duration of positivity is the key quantity linking incidence and prevalence.
 
+Therefore, I then turn to estimating this distribution.
+I start, in \cref{E-ATACCC}, by estimating the duration of positivity from the ATACCC study.
+This provides a reliable estimate of the bulk of the distribution (out to around 20 days) but shows that there is a significant tail.
+To estimate the tail, a large sample size and long follow-up is required.
+Therefore, I turn to the CIS data in \cref{E-perf-test,E-imperf-test}.
+In \cref{E-perf-test} I provide a framework for estimating the duration of positivity from the CIS data, assuming perfect tests.
+However, the assumption of perfect tests is not realistic; in \cref{E-imperf-test} I extend the framework to include false negatives.
+
+Having estimated the duration of positivity, I then turn to estimating incidence and transmission from the CIS data in \cref{E-transmission}, thus fulfilling the primary aim of this thesis.
+I use both of the approaches discussed in \cref{intro:sec:estimating-incidence}, contrasting their results and discussing the advantages and disadvantages of each in this context.
+Finally, I conclude in \cref{E-conclusion}.
+
+Definitions of the probability distributions used throughout this thesis, are given in \cref{E-distributions}.
+Other appendices provide additional details.
+
+\ifSubfilesClassLoaded{
+  \appendix
+  \subfile{MCMC-appendix}
+  \listoftodos
+}{}
 \end{document}

### 2024-02-28 thesis.tex
reduce num headers in TOC
fix long list of names in citation
todolist at end for ease of finding
write most of intro

@@
 \newcommand{\R}{\mathcal{R}_{0}}
+\renewcommand{\Re}{\mathcal{R}_{e}(t)}
+\newcommand{\Rc}{\mathcal{R}_{c}(t)}
 \newcommand{\vlod}{v_\text{lod}}
@@
 %% Bibliography
-\usepackage[backend=biber,uniquename=false,sortcites=true,sorting=ynt,maxcitenames=2,style=apa]{biblatex}
+\usepackage[backend=biber,uniquename=init,sortcites=true,sorting=ynt,maxcitenames=2,style=apa,uniquelist=false]{biblatex}
 \addbibresource{references.bib}
@@
 
+\setcounter{tocdepth}{1}
+
 %% REMOVE THESE FOR FINAL VERSION WITHOUT TODOS
-\usepackage{geometry}
-\setlength{\marginparwidth}{3cm}
-\geometry{margin=3.8cm}
+% \usepackage{geometry}
+% \setlength{\marginparwidth}{3cm}
+% \geometry{margin=3.8cm}
 
@@
 
-\chapter{Conclusion}
+\chapter{Conclusion} \label{conclusion}
 Household structure is not included in this model.
@@
 The beta-binomial likelihood structure should capture some of this variance within each strata\todo{ref beta-binomial as a model for clustering}, but not between strata.
-
-\listoftodos
 
@@
 \subfile{distributions}
+\subfile{MCMC-appendix}
 \subfile{ATACCC-appendix-original-analysis}
@@
 \cref{inc-prev:eq:forward-substitute}
+\listoftodos
+
 \end{document}

### 2024-02-28 transmission.tex
write most of intro

@@
 I compute the posterior distribution using MCMC.
-Specifically, I use an adaptive random-walk Metropolis-Hastings algorithm (previously discussed in \cref{E-inc-prev:sec:MCMC}).
+Specifically, I use an adaptive random-walk Metropolis-Hastings algorithm (discussed in \cref{E-MCMC}).
 The proposal distribution is a multivariate normal that adapts to the covariance of the posterior.
@@
 
-All regions' MCMC chains converged, assessed using the Rhat statistic and ESS (as defined in \cref{E-inc-prev:sec:MCMC}).
+All regions' MCMC chains converged, assessed using the Rhat statistic and ESS (as defined in \cref{E-intro:sec:MCMC}).
 The lowest Rhat was 1.01 for $\sigma_\epsilon$ in London.

### 2024-02-29 incidence-prevalence.tex
sort to dos, add a little more text

@@
 The infection process (\cref{inc-prev:sec:infection-process}) models how infections arise.
-The concepts here are well-established within infectious disease epidemiology (see\todo{ref intro chapter}).
+The concepts here are well-established within infectious disease epidemiology.
 
@@
 \emph{Incidence} on day $t$, $Z_t$, is the number of people in a given population who become infected on day $t$; when expressed as a proportion of the population it is the \emph{incidence proportion}~\autocite[89]{lashModern}, $Z_t/\Npop$, where $\Npop$ is the number of people in the population.
-As discussed in\todo{ref intro}, the timing of infection events is very rarely directly observable and must be inferred from other data.
+The timing of infection events is very rarely directly observable and must be inferred from other data.
 Despite this issue, it remains the single most important quantity for informing the response to an epidemic or pandemic.

### 2024-02-29 introduction.tex
sort to dos, add a little more text

@@
 
+\todo[inline]{Rewrite with more clarity backcalculation background: do I need the equation? What is it in practice? What did I write in my first-year report, and what comments did it get?}
+
 \emph{Backcalculation} is a statistical procedure for estimating incidence based on a measurement of prevalence.
@@
 Many mechanistic modelling studies were conducted during the pandemic, with at least 100 articles published in the first six months of 2020 alone~\autocite{shankarSystematic}.
+In the UK, the most influential were the outputs from SPI-M\todo{cite SPI-M and its outputs - Medley paper?}.
+For example, the stages of lifting the lockdown in 2021 each had a set of modelling studies informing the government's decision\todo{cite roadmap 2021 papers}.
+
+Each week, the members of SPI-M would produce a set of estimates of the current state of the epidemic (\eg ther value of $\Re$), known as \emph{nowcasting}.
+These estimates, once any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value.\todo{cite SPI-M combination}
+The consensus value was published on \url{gov.uk} and informed the government's response to the epidemic.\todo{cite gov.uk R number}
+
+The academic groups contributing estimates used a wide variety of methodologies
+
+\todo[inline]{Finish lit review of SPI-M methods}
 
@@
 Since, under reasonable model assumptions, there is a one-to-one correspondence between reproduction numbers and growth rates~\autocite{wallingaGI}, this also estimates $\Re$.
+
+\Textcite{colmanAscertainment} used CIS prevalence estiamtes to estimate the \emph{ascertainment rate} of community testing (the proportion of all infections that community testing detected).
+Their methodology was similar to \textcite{abbottCISincidence}.
+\Textcite{colmanAscertainment} do not require a Gaussian process (or other prior) to stabilise incidence estimates because they only estimate the ascertainment rate.
+\todo[inline]{Finish writing about \textcite{colmanAscertainment}}
 

### 2024-02-29 transmission.tex
edit backcalc methods

@@
 
-I keep the same logistic regression model in terms of prevalence, transforming its posterior distribution of prevalence to a posterior distribution of incidence.
+% I keep the same logistic regression model for predicting prevalence, then transform its posterior distribution of prevalence to a posterior distribution of incidence.
 This results in a two-step process: first, the same inference process described in \cref{E-biology-data:sec:MRP} is used to estimate prevalence; second, the prevalence is transformed to incidence.
+This is required to fulfil several desirable features of the model.
+
 An alternative would be an inference procedure that estimates incidence proportion, fit directly to the RT-PCR test results.
-This is required to fulfil several desirable features of the model.
 Denote the quantity to be inferred (which could be either incidence proportion or prevalence) at time $t$ in strata $i$ as $\pi_{it}$.
@@
     \item Borrow information across the different covariates.
-        The borrowing should penalise towards the distribution of incidence/prevalence across strata in the same geographic region being constant.
+        The borrowing should penalise away from changes in the distribution of incidence/prevalence across strata in the same geographic region being constant.
         Standard stratified, mechanistic transmission model (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) predict this, if behavioural changes are independent of strata and there is negligible impact of immunity.
     \item Be computationally efficient.
-        The model will be fit to granular data and therefore needs to be run within the SRS, with its computational limitations (see\todo{ref section on SRS computational limitations}).
+        The model will be fit to granular data and therefore needs to be run within the SRS, with its computational limitations (see \cref{E-biology-data:sec:SRS}).
 \end{enumerate}
@@
 Combined with property 4, the form of $g$ the null model is $g(i, t) = a_0 + a_t t + f(i)$ where $a_0$ and $a_t$ are model parameters and $f$ is a function to be specified; that is, changes away from this model should be penalised.
-To add additional flexibility,
 
@@
 In particular, using MCMC, sufficient covariates to adjust for the non-response bias, and a period of more than 6--8 weeks leads to a model that is computationally prohibitive ~\citePersonalComms{Koen Pouwels}.
-Instead, previous MRP implementations have used the Integrated Nested Laplace Approximation (INLA)~\autocite{rueINLA}, implemented in R-INLA~\autocite{RINLA} (see \cref{transmission:sec:INLA}).
-theta
+Instead, previous MRP implementations used INLA (see \cref{E-biology-data:sec:MRP}).
 
@@
 Should I use alterantive notation - but then different to standard in the INLA papers.}
-INLA-based inference can only be applied to LGMs (\emph{Latent Gaussian Models})~\autocite{rueINLAseminar,rueINLA,martinoINLAtutorial}.
+INLA-based inference can only be applied to LGMs (\emph{Latent Gaussian Models})~\autocite{rueINLAseminar,rueINLA,martinoINLAtutorial}; see \cref{E-transmission:sec:INLA} for further details on INLA.
 An LGM is defined in terms of a vector of hyperparameters, $\vec{\varphi}$; a deterministic transformation of the hyperparameters into covariance matrix, $\matr{\Sigma}(\vec{\varphi})$; a vector of latent variables, $\vec{x}$; a vector of $n$ data points, $\vec{y} = [y_1, \dots, y_n]^T$; and $n$ linear predictors $\vec{\eta} = [\eta_1, \dots, \eta_n]^T$.
@@
 
-In this context, each element $\vec{y}$ corresponds to the observed RT-PCR results in one stratum on one day.
-Hence, $\eta_{j}$ must determine the distribution of $y_{j}$ (up to some hyperparameters); where $j$ can be mapped to a mapping of a stratum $i$ and day $t$.
+In this context, each $y_j$ corresponds to the observed RT-PCR results in one stratum on one day.
+Hence, $\eta_{j}$ must determine the distribution of $y_{j}$ (up to some hyperparameters); where $j$ uniquely identifies a stratum $i$ and day $t$.
 In practice, this means that $\eta_j$ is a one-to-one transformation of the population prevalence in stratum $i$ on day $t$.

### 2024-02-29 introduction.tex
fix labels
remove notes

@@
 
-\section{Data sources}
+\section{Data sources} \label{intro:sec:data-sources}
 
@@
 
-\section{Estimating incidence and transmission} \cref{intro:sec:estimating-incidence}
+\section{Estimating incidence and transmission} \label{intro:sec:estimating-incidence}
 
@@
 \subsection{Backcalculation}
-
-\todo[inline]{Rewrite with more clarity backcalculation background: do I need the equation? What is it in practice? What did I write in my first-year report, and what comments did it get?}
 
@@
 Since its development, backcalculation has been further developed for a wide-variety of contexts.
-\todo[inline]{probably need to cite some review or developments of backcalculation or give a brief lit review? not sure how much details is worthwhile as its not directly relevant}
 % Reporting delay is not relevant to the studies in this thesis, and I ignore it in this and further discussion.

### 2024-02-29 transmission.tex

@@
 See \cref{SEIR:MCMC-algorithm} for full details.
+
 \begin{algorithm}
- set $\vec{X_0}$ to an initial value of the parameter vector \;
+ Set $\vec{X_0}$ to an initial value of the parameter vector \;
  $\vec{\mu_0} = \vec{X_0}$ \;
@@
 Both of these rely on an assumption that the epidemic is in a steady-state exponential growth phase at time 0 (see \cref{E-SEIR:sec:SIR}).
-
 During such an exponential growth phase, all components of the $\vec{e}$, $\vec{i}$ and $\vec{p}$ compartments grow at the same growth rate, $\psi$.

### 2024-02-29 transmission.tex
Updates from Overleaf

--- 2024-02-28 transmission.tex
@@
 I compute the posterior distribution using MCMC.
-Specifically, I use an adaptive random-walk Metropolis-Hastings algorithm (discussed in \cref{E-MCMC}).
+Specifically, I use an adaptive random-walk Metropolis-Hastings algorithm (previously discussed in \cref{E-inc-prev:sec:MCMC}).
 The proposal distribution is a multivariate normal that adapts to the covariance of the posterior.
@@
 
-All regions' MCMC chains converged, assessed using the Rhat statistic and ESS (as defined in \cref{E-intro:sec:MCMC}).
+All regions' MCMC chains converged, assessed using the Rhat statistic and ESS (as defined in \cref{E-inc-prev:sec:MCMC}).
 The lowest Rhat was 1.01 for $\sigma_\epsilon$ in London.

### 2024-03-04 introduction.tex
proof intro

@@
 
-COVID-19 (coronavirus disease 2019), the disease caused by the virus SARS-CoV-2 (severe acute respiratory syndrome coronavirus 2), is estimated to have killed approximately 15 million in the acutest phase (to the end of 2021) and 30 million globally~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
-While it is tempting to consider this a rare or extreme event, history suggests otherwise (see \cref{intro:fig:pandemic-timeline}).
+COVID-19 (coronavirus disease 2019), the disease caused by the virus SARS-CoV-2 (severe acute respiratory syndrome coronavirus 2), is estimated to have killed approximately 15 million in the acutest phase (to the end of 2021) and 30 million in total~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
+While would be reassuring if this was a rare or extreme event, but history suggests otherwise (see \cref{intro:fig:pandemic-timeline}).
 \begin{figure}
@@
 
-The statistical framework developed in the context of HIV transmission is applicable to SARS-CoV-2.
+The statistical models used to describe HIV transmission are applicable to SARS-CoV-2.
 This infection process can be described as a Poisson process (see \cref{E-inc-prev:sec:infection-process}).
@@
 The ability to estimate the contribution of each of these factors can inform public health policy.
-Furthermore, mechanistic models can be used for scenario-based modelling, which simulates the effect of interventions.
-A scenario here is the effect of a proposed intervention (or lack thereof), introduced into the model by modifying the relevant parameters.
+Furthermore, mechanistic models can be used for scenario modelling, which simulates the effect of interventions.
+A scenario is the effect of a proposed intervention (or lack thereof), simulated by modifying the relevant parameters.
 % The mechanistic model can then be simulated forward in time to understand the intervention's effects.
@@
 Each week, the members of SPI-M would produce a set of estimates of the current state of the epidemic (\eg ther value of $\Re$), known as \emph{nowcasting}.
+The nowcasting estimates were based on mechanistic models.
 These estimates, once any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value.\todo{cite SPI-M combination}
@@
 
-The academic groups contributing estimates used a wide variety of methodologies
+The academic groups contributing estimates used a wide variety of methodologies~\autocite{royalSocietyRnumber}.
+These can be split into two types: those based on renewal equations and those based on compartmental models.
+Renewal
 
@@
 
+\todo[inline]{Introduce estimating duration}
+
 \subsection{Methodology}
@@
 % Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
-Other related approaches, such as a multi-state model~\autocite{jacksonMSM}, could also be considered (multi-state models can be viewed as a generalisation of survival analysis).
+Other related approaches, such as a multi-state model~\autocite{jacksonMSM}\todo{find best reference for MSMs}, could also be considered (multi-state models can be viewed as a generalisation of survival analysis).
 
@@
 The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability, but the results are limited because the small sample size and short follow-up means that the tails of these distributions cannot be estimated.
-I describe ATACCC in detail in \cref{E-biology-data:sec:ataccc} and analyse it in \cref{E-ATACCC}.
+I describe ATACCC in detail in \cref{E-biology-data:sec:ataccc} and use it to estimate duration in \cref{E-ATACCC}.
 
@@
 In particular, the vast majority of detected infections had only one or two positive tests, and hence the viral load trajectory of these infections cannot be estimated.
-I describe the CIS in detail in \cref{E-intro:sec:cis} and analyse it in \cref{E-perf-test,E-imperf-test}.
+I describe the CIS in detail in \cref{E-intro:sec:cis} and use it to estimate duration in \cref{E-perf-test,E-imperf-test}.
 
@@
 
+For the purposes of this thesis, estimates applicable to the general population are required.
+Therefore, I estimate the duration of positivity from the ATACCC and CIS studies, as described above.
+
 \section{Bayesian inference} \label{intro:sec:Bayes}
 
-All inference in this thesis is performed within the Bayesian framework.
-Bayesian inference has several advantages relevant for this work.
-Prior information allows identifiability in situations where this would not otherwise be possible, allows for combining knowledge from multiple sources, and much prior work I build on has been performed within the Bayesian framework.
+All inference in this thesis is Bayesian.
+Bayesian inference has several relevant advantages.
+Prior information allows identifiability in situations where this would not otherwise be possible, allows for combining knowledge from multiple sources, and many previous studies I build used Bayesian inference.
 
@@
 In particular, an epidemic which will cause issues starts by increasing, therefore if the methods are to be applied early in a future epidemic or pandemic, they will start with an increasing period.
-The CIS began in Aril 2020, when incidence was falling due to the first lockdown in the UK.
+\todo{Rephrase this sentence}
+The CIS began in April 2020, when incidence was falling due to the first lockdown in the UK.
 Incidence remained low and fairly flat in England over summer 2020, until it began rising in mid or late August.
@@
 
-First, the framework to simulation the CIS (used in \cref{E-imperf-test:sec:simulate,E-perf-test:sec:simulation-study}) is structured as a R package.
+First, the framework to simulation the CIS (used in \cref{E-imperf-test:sec:simulate,E-perf-test:sec:simulation-study}).
 The package enables customisation of various features of the study, for example the testing regime of the participants, the sensitivity of the tests, and the duration of positivity.
@@
 However, it assumes that the prevalence at any time is a random variable, instead modelling it as a deterministic function of the incidence.
+\todo{Rephrase this sentence}
 I show that, for all that is required for this assumption to hold is that the proportion of the population sampled is small.

### 2024-03-05 introduction.tex
expand lit review on SPI-M models

@@
 Many mechanistic modelling studies were conducted during the pandemic, with at least 100 articles published in the first six months of 2020 alone~\autocite{shankarSystematic}.
-In the UK, the most influential were the outputs from SPI-M\todo{cite SPI-M and its outputs - Medley paper?}.
-For example, the stages of lifting the lockdown in 2021 each had a set of modelling studies informing the government's decision\todo{cite roadmap 2021 papers}.
+In the UK, SPI-M-O (Scientific Pandemic Influenza Group on Modelling, Operational sub-group) was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), generating evidence based on modelling for policy development~\autocite{medleySPIM,govSPIMO}.
+For example, the stages of lifting the lockdown in 2021 each had a set of modelling studies informing the government's decision~\autocite{sageEvidence}.
 
-Each week, the members of SPI-M would produce a set of estimates of the current state of the epidemic (\eg ther value of $\Re$), known as \emph{nowcasting}.
+Each week, the members of SPI-M would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
 The nowcasting estimates were based on mechanistic models.
-These estimates, once any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value.\todo{cite SPI-M combination}
-The consensus value was published on \url{gov.uk} and informed the government's response to the epidemic.\todo{cite gov.uk R number}
+These estimates, once any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value~\autocite{parkCombining}.
+The consensus value was published on \url{gov.uk} and informed the government's response to the epidemic~\autocite{govRnumber}.
+The academic groups contributing estimates used a variety of models, mostly based on either \emph{renewal equations} or \emph{compartmental models}~\autocite{royalSocietyRnumber}.
 
-The academic groups contributing estimates used a wide variety of methodologies~\autocite{royalSocietyRnumber}.
-These can be split into two types: those based on renewal equations and those based on compartmental models.
-Renewal
+The renewal equation~\autocite{fraserRenewal} for an infectious disease, in the most general form, is:
+\begin{align}
+    I(t) &= \int_{0}^{t} \beta(t,\tau) I(t-\tau) d \tau
+\end{align}
+where $I(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the instantaneous probability of an individual infected $\tau$ time units ago infects an individual at time $t$.
 
@@
 
-\Textcite{colmanAscertainment} used CIS prevalence estiamtes to estimate the \emph{ascertainment rate} of community testing (the proportion of all infections that community testing detected).
+\Textcite{colmanAscertainment} used CIS prevalence estimates to estimate the \emph{ascertainment rate} of community testing (the proportion of all infections that community testing detected).
 Their methodology was similar to \textcite{abbottCISincidence}.

### 2024-03-05 introduction.tex
lit review

@@
 
-\todo[inline]{Maybe talk about pandemic v epidemic v outbreak terminology?}
-\todo[inline]{Need to motivate the importance of surveillance}
+All pandemics start as an \emph{outbreak}, a sudden and unexpected rise in the number of people with a disease in a local area; if this spreads over a larger geographic area (\eg a country) it becomes an \emph{epidemic}; if it spreads globally then it is a \emph{pandemic}~\autocite{grennanPandemic}.
+The precise meaning of these terms is contested, determined by usage rather than formal definitions~\autocite{morensPandemic,doshiElusive}.
+I will refer to the SARS-CoV-2 epidemic within England (or the UK), and the global impact as the pandemic.
+
+In this context, \emph{surveillance} is the ability to detect and monitor the spread of an outbreak, epidemic, or pandemic.
+Surveillance is important at every stage: early on, for detecting a new outbreak quickly and assessing its likely impact and later for monitoring the ongoing burden of the disease and the effectiveness of interventions~\autocite{whoFluSurveillance}.
+In all cases, surveillance is required to inform public health policy.
 
@@
 Early in the pandemic, a methodology was developed for using CIS data to estimate prevalence (see \cref{E-intro:sec:cis}).
-The primary aim of this thesis is to estimate incidence and transmission from this data.
+
+A second prevalence survey ran in the UK, known as REACT (REal-time Assessment of Community Transmission)~\autocite{rileyResurgence,rileyREACT}.
+REACT was a series of cross-sectional surveys, run across 19 rounds between May 2020 and March 2022.
+Each round consisted of at least 100,000 randomly-sampled individuals from the UK contributing swabs for RT-PCR testing and questionnaire data~\autocite{elliottTwin}.
+
+A major advantage of the CIS over REACT is that the CIS has daily testing of individuals, while REACT has no test results between rounds.
+This means that REACT cannot be used to estimate transmission or incidence between rounds, instead relying on model assumptions at these times.
+REACT missed the peak in late December 2020, which complicated analysis of the survey data at that time~\autocite{rileyREACTround8}.
+
+The primary aim of this thesis is to estimate incidence and transmission from the CIS data.
 
@@
 The consensus value was published on \url{gov.uk} and informed the government's response to the epidemic~\autocite{govRnumber}.
-The academic groups contributing estimates used a variety of models, mostly based on either \emph{renewal equations} or \emph{compartmental models}~\autocite{royalSocietyRnumber}.
+The academic groups contributing estimates used a variety of models, mostly based on either \emph{renewal equations} or \emph{compartmental models}.
+The methods were reviewed by \textcite{royalSocietyRnumber}; see appendix 2 of that review for details of the specific implementations.
 
@@
 where $I(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the instantaneous probability of an individual infected $\tau$ time units ago infects an individual at time $t$.
+A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
+The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
+Renewal equations do not feature in this thesis, for further information, including practical considerations, see \textcite{nashEstimating,gosticPractical}.
 
-\todo[inline]{Finish lit review of SPI-M methods}
+Compartmental models split the population into compartments, each representing a different state of the disease.
+The \emph{dynamics} of the epidemic, how it changes over time, are described by the transitions between compartments.
+Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
+Unlike renewal equations, there are not widely-used implementations of compartmental models.
+Most of the implementations used by SPI-M-O members were in the SEIR (susceptible, exposed, infectious, recovered) framework.
+The SEIR framework, has four components (susceptible, exposed, infectious, recovered), with the population divided into these components.
+The SEIR framework is described in detail in \cref{E-SEIR:sec:mechanistic-models}.
+Compartmental models can be fit to almost any data source, and commonly incorporate multiple data sources, known as \emph{data synthesis}, a technique reviewed by \textcite{birrellEvidence}.
+Most compartmental models used by SPI-M-O were age-stratified, as differences between age groups are important for both the transmission and severity of SARS-CoV-2 (see \cref{E-SEIR:sec:structured-populations}).
 
-\begin{itemize}
-    \item SPI-M
-    \item Literature on mechanistic models using REACT and CIS I have previously circulated.
-\end{itemize}
+Several compartmental models incorporate data from SARS-CoV-2 prevalence surveys, alongside other data sources~\autocite{daviesAssociation,ironsEstimating,knockKey,nicholsonImproving,pooleyEstimation,birrellRTM2}.
+These models incorporate prevalence survey data by adding additional compartments representing RT-PCR individuals, who may not be infectious (this is common, see \cref{E-biology-data}); except \textcite{nicholsonImproving} who have a more sophisticated approach (below) and \textcite{ironsEstimating}, who assume RT-PCR positivity coincides with infectiousness.
+The length of stay in these compartments (the duration of time positive) is assumed to follow a gamma distribution, with the parameters estimated from external data or having at most one free parameter.
+The proportion of the population in these compartments are then fit to the prevalence survey data using a binomial likelihood where possible, although some approaches take a Gaussian approximation.
+These approaches are similar to how these models are fit to other data sources, such as hospital admissions or deaths.
+
+\Textcite{nicholsonImproving} have the most sophisticated use of prevalence survey data.
+They allow an arbitrary distribution of the duration of positivity, based on the estimates of \textcite{hellewellPCRSensitivity} (see \cref{intro:sec:duration-SARS-CoV-2} for discussion of the \textcite{hellewellPCRSensitivity} study).
+Furthermore, they incorporate the sensitivity and specificity of the RT-PCR test into their model.
+% However, they do not incorporate age-stratification into their transmission model, and primarily aim to estimate the bias of community testing in the UK rather than incidence and/or transmission.
+
+In \cref{E-SEIR}, I describe a compartmental model for SARS-CoV-2 transmission, which I fit to CIS data.
+I allow a more flexible distribution of the duration of positivity than most of the papers discussed above, incorporating the skew of the distribution.
+This distribution is based on data from a representative sample of the UK population, rather than convenience samples (see \cref{intro:sec:duration-SARS-CoV-2}).
+Unlike \cref{nicholsonImproving}, I include age-stratification in the transmission model, allowing estimation of how transmission varies by age.
+I use only data from the CIS, and show this is sufficient for estimating incidence and transmission.
+Finally, I fit directly to counts from the CIS, rather than an approximation of this data, which has not previously be done.
 

### 2024-03-05 introduction.tex
to do list

@@
 This distribution is based on data from a representative sample of the UK population, rather than convenience samples (see \cref{intro:sec:duration-SARS-CoV-2}).
-Unlike \cref{nicholsonImproving}, I include age-stratification in the transmission model, allowing estimation of how transmission varies by age.
+Unlike \textcite{nicholsonImproving}, I include age-stratification in the transmission model, allowing estimation of how transmission varies by age.
 I use only data from the CIS, and show this is sufficient for estimating incidence and transmission.
@@
 
+\todo[inline]{Discuss \textcite{ealesCharacterising}}
+
 \section{Bayesian inference} \label{intro:sec:Bayes}
@@
 
-In order to achieve these aims, I use the SARS-CoV-2 epidemic in England from Mon 31st Aug 2020 until Sun 25th Jan 2021 as a case study.
-This case study has several features that make it a useful case study for the aims of this thesis.
-First, the CIS was of sufficient size and incidence sufficiently high that there is enough data to perform the analyses.
-Second, incidence begins rising over this period, which is the period of most interest.
-In particular, an epidemic which will cause issues starts by increasing, therefore if the methods are to be applied early in a future epidemic or pandemic, they will start with an increasing period.
-\todo{Rephrase this sentence}
+In order to achieve these aims, I use a case study, specifically the SARS-CoV-2 epidemic in England from Mon 31st Aug 2020 until Sun 25th Jan 2021.
+This case study has several features that make it useful for the aims of this thesis.
+These include that the CIS was of sufficient size and incidence sufficiently high that there is enough data to perform the analyses.
+Additionally, incidence begins rising over this period, which is the period when a surveillance system is needed to provide a warning of upcoming issues.
 The CIS began in April 2020, when incidence was falling due to the first lockdown in the UK.
 Incidence remained low and fairly flat in England over summer 2020, until it began rising in mid or late August.
-\todo{cite the phases of the UK epidemic}
 Finally, the period ends before the roll-out of the vaccine, which likely affects many important aspects of the analysis.
@@
 A deterministic backcalculation approach was used by \textcite{abbottCISincidence} and I will use it in \cref{E-backcalc}.
-However, it assumes that the prevalence at any time is a random variable, instead modelling it as a deterministic function of the incidence.
-\todo{Rephrase this sentence}
-I show that, for all that is required for this assumption to hold is that the proportion of the population sampled is small.
-In particular, I show that there is no requirement for the prevalence or incidence to be large.
+This approach models the prevalence at any time as a deterministic function of the incidence, rather than as a random variable.
+I show that, this is a good approximation as long as the proportion of the population sampled is small.
+In particular, it does not require the prevalence or incidence to be large.
 

### 2024-03-07 biology-data.tex
write most of intro
edit backcalc methods
Resolve to dos

@@
 
-In this section, I expand on the major two studies I use in this thesis (ATACCC and CIS), first introduced in\todo{ref where introducing studies}.
+In this section, I expand on the major two studies I use in this thesis (ATACCC and CIS).
 Unlike many studies of SARS-CoV-2, these studies are not focused on specific groups such as healthcare workers or hospital patients.
@@
 
-In this thesis, I focus on the period from Mon 31st Aug 2020 until Sun 24th Jan 2021 (inclusive).
-\todo{choice of period might be better in intro}
-This period starts when CIS was expanded, and ends when vaccination crossed 10\% of England's population.
-Vaccination complicates the dynamics of transmission and viral load.
-Analysis of data including vaccinated individuals requires more complex models, with specific adaptations to SARS-CoV-2.
-These are beyond the scope of this thesis, which primarily intends to show the potential of a study such as CIS.
-
+In this thesis, I focus on the period from Mon 31st Aug 2020 until Sun 24th Jan 2021 (see \cref{E-intro:sec:aims}).
 During the start of this period, the recruitment rate of CIS was expanded (see \cref{biology-data:fig:CIS-recruitment}(A)).
@@
 A notable feature of this definition is that it explicitly does not correct for false positive or negative results.
+However, for many purposes, including this thesis, the difference between the positivity rate and prevalence is negligible
+
 Generally, the point in time of interest is a specific day, although it has also been applied to a week~\autocite{cisMethodsONS,pouwelsMRPvaccination,pouwelsCommunity}.
@@
 
-\subsubsection{Computational limitations and privacy protection}
+\subsubsection{Computational limitations and privacy protection} \label{biology-data:sec:SRS}
 

### 2024-03-07 cis-perfect-testing.tex
write most of intro
Resolve to dos

@@
 
-\todo[inline]{Are the following paragraphs better here or in the intro chapter when motivating each chapter and the overall narrative of the thesis?}
-In the rest of this and the following chapter, I use survival analysis rather than the biomarker approach taken in \cref{E-ATACCC}.
-This choice is motivated by the computational limitations of the SRS and the CIS's study design (see \cref{E-intro:sec:cis}).
-The biomarker approach is computational challenging because the number of random effects scales with the number of detected infection episodes.
-
-There are many detected episodes in CIS; in the period I will consider, around 50 times more than in ATACCC.
-Therefore, the biomarker approach would be computationally challenging.
-The computational challenge is compounded by the data being stored in a TRE, which has limited computational power and software.
-The issue with random effects is made worse by the CIS's study design.
-Infrequent testing means that most detected episodes have only a single positive test.
-Hence, there is no information on the viral load trajectory and the random effects are unidentifiable.
-
 \section{Problem description} \label{perf-test:sec:problem}
@@
 
-\subsection{Unobserved episode beginning and end times}
+\subsection{Unobserved episode beginning and end times} \label{perf-test:sec:interval-censoring}
 
@@
 
-Few studies look at the combination of both double interval censoring and undetected (or truncated) events
+Few studies look at the combination of both double interval censoring and undetected (or truncated) events.
 This is especially true within the human biostatistical literature.

### 2024-03-07 incidence-prevalence.tex
Resolve to dos

@@
 
-\chapter{Statistical and epidemiological background} \label{inc-prev}
-
-\todo[inline]{%
-  Chapter 3 needs a better title: this is not all the statistical and epidemiological background and some of it is mathematical modelling / disease dynamics.
-  Perhaps: ``Data generating procsses and disease dynamics''.
-  Would need to define disease dynamics.
-}
-
-%\todo[inline]{This chapter is pretty short. Not sure if this matters or it needs to merged with another chapter? For example, one background chapter split into epi/biology and stats.}
+\chapter{Data generating process} \label{inc-prev}
 
@@
 
-\todo[inline]{Does something need to link these sections?}
-
 \section{Prevalence process} \label{inc-prev:sec:prevalence-process}
@@
 This chapter has shown that the key quantity relating these quantities is the distribution of the duration of RT-PCR positivity.
-However, no estimates of this quantity in the general population exist\todo{ref lit review of duration}.
+However, no estimates of this quantity in the general population exist (see \cref{E-intro:sec:duration-SARS-CoV-2}).
 Therefore, I now turn to estimating the duration of RT-PCR positivity.

### 2024-03-07 introduction.tex
Resolve to dos
Proofing/to do

@@
 Fully quantifying the burden of historical pandemics is challenging due to the absence of systematic records.
-However, our best records suggests that, over the last 400 years, they have (on average) occurred every four years with a mean death toll of 5.6 million~\autocite{maraniNovelEpidemics}.
+However, our best records suggests that, since 1600, they have (on average) occurred every four years with a mean death toll of 5.6 million~\autocite{maraniNovelEpidemics}.
 Modern estimates suggest that we should expect over 2.5 million deaths per year from pandemics~\autocite{madhavPandemicMortality}.
@@
 All pandemics start as an \emph{outbreak}, a sudden and unexpected rise in the number of people with a disease in a local area; if this spreads over a larger geographic area (\eg a country) it becomes an \emph{epidemic}; if it spreads globally then it is a \emph{pandemic}~\autocite{grennanPandemic}.
-The precise meaning of these terms is contested, determined by usage rather than formal definitions~\autocite{morensPandemic,doshiElusive}.
+The precise meaning of these terms is contested, determined implicitly by their usage rather than formal definitions~\autocite{morensPandemic,doshiElusive}.
 I will refer to the SARS-CoV-2 epidemic within England (or the UK), and the global impact as the pandemic.
 
-In this context, \emph{surveillance} is the ability to detect and monitor the spread of an outbreak, epidemic, or pandemic.
+In this context, \emph{surveillance} is the ability to detect and monitor the spread of an outbreak.
 Surveillance is important at every stage: early on, for detecting a new outbreak quickly and assessing its likely impact and later for monitoring the ongoing burden of the disease and the effectiveness of interventions~\autocite{whoFluSurveillance}.
@@
 
-\subsection{Backcalculation}
+\subsection{Backcalculation} \label{intro:sec:backcalc}
 
@@
 
-\subsubsection{Application to SARS-CoV-2}
-
-Many mechanistic modelling studies were conducted during the pandemic, with at least 100 articles published in the first six months of 2020 alone~\autocite{shankarSystematic}.
-In the UK, SPI-M-O (Scientific Pandemic Influenza Group on Modelling, Operational sub-group) was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), generating evidence based on modelling for policy development~\autocite{medleySPIM,govSPIMO}.
-For example, the stages of lifting the lockdown in 2021 each had a set of modelling studies informing the government's decision~\autocite{sageEvidence}.
-
-Each week, the members of SPI-M would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
-The nowcasting estimates were based on mechanistic models.
-These estimates, once any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value~\autocite{parkCombining}.
-The consensus value was published on \url{gov.uk} and informed the government's response to the epidemic~\autocite{govRnumber}.
-The academic groups contributing estimates used a variety of models, mostly based on either \emph{renewal equations} or \emph{compartmental models}.
-The methods were reviewed by \textcite{royalSocietyRnumber}; see appendix 2 of that review for details of the specific implementations.
+Two types of mechanistic models were prominent during the SARS-CoV-2 pandemic: those based on a \emph{renewal equation} and those with a \emph{compartmental} structure.
 
@@
 
-Several compartmental models incorporate data from SARS-CoV-2 prevalence surveys, alongside other data sources~\autocite{daviesAssociation,ironsEstimating,knockKey,nicholsonImproving,pooleyEstimation,birrellRTM2}.
+\subsubsection{Application to SARS-CoV-2}
+
+Many mechanistic modelling studies were conducted during the pandemic, with at least 100 articles published in the first six months of 2020 alone~\autocite{shankarSystematic}.
+In the UK, SPI-M-O (Scientific Pandemic Influenza Group on Modelling, Operational sub-group) was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), generating evidence based on modelling for policy development~\autocite{medleySPIM,govSPIMO}.
+For example, the stages of lifting the lockdown in 2021 each had a set of modelling studies informing the government's decision~\autocite{sageEvidence}.
+
+Each week, the members of SPI-M would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
+The nowcasting estimates were based on mechanistic models.
+These estimates, once any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value~\autocite{parkCombining}.
+The consensus value was published on \url{gov.uk} and informed the government's response to the epidemic~\autocite{govRnumber}.
+The academic groups contributing estimates used a variety of models, reviewed by \textcite{royalSocietyRnumber}; see appendix 2 of that review for details of the specific implementations.
+
+In this thesis, I focus on the use of data from prevalence surveys.
+Several compartmental models incorporated data from SARS-CoV-2 prevalence surveys, alongside other data sources~\autocite{daviesAssociation,ironsEstimating,knockKey,nicholsonImproving,pooleyEstimation,birrellRTM2}.
 These models incorporate prevalence survey data by adding additional compartments representing RT-PCR individuals, who may not be infectious (this is common, see \cref{E-biology-data}); except \textcite{nicholsonImproving} who have a more sophisticated approach (below) and \textcite{ironsEstimating}, who assume RT-PCR positivity coincides with infectiousness.
@@
 
-\Textcite{mccabeCISincidence}, applying the methodology of \textcite{ealesAppropriately}, estimate $\Re$ and the epidemic's growth rate from the CIS data.
+\Textcite{mccabeCISincidence}, adapting the methodology \textcite{ealesAppropriately} developed for REACT, estimate $\Re$ and the growth rate from the CIS data.
 They estimate a posterior distribution of the parameters of a spline, penalising deviations from a constant first derivative.
 The growth rate is then calculated from the derivative of the spline, an analytical transformation of the parameters.
-Since, under reasonable model assumptions, there is a one-to-one correspondence between reproduction numbers and growth rates~\autocite{wallingaGI}, this also estimates $\Re$.
+Since, under reasonable model assumptions, there is a one-to-one correspondence between reproduction numbers and growth rates~\autocite{wallingaGI}, and they assume that the growth rate in incidence and prevalence is equal, this also estimates $\Re$.
 
 \Textcite{colmanAscertainment} used CIS prevalence estimates to estimate the \emph{ascertainment rate} of community testing (the proportion of all infections that community testing detected).
-Their methodology was similar to \textcite{abbottCISincidence}.
-\Textcite{colmanAscertainment} do not require a Gaussian process (or other prior) to stabilise incidence estimates because they only estimate the ascertainment rate.
-\todo[inline]{Finish writing about \textcite{colmanAscertainment}}
+Their methodology was similar to \textcite{abbottCISincidence} (see \cref{intro:sec:backcalc}), including using estimates from \textcite{hellewellPCRSensitivity} for how the sensitivity of RT-PCR testing varies over an infection.
+Rather than using a Gaussian process, as \textcite{abbottCISincidence} do, \textcite{colmanAscertainment} provide stability by assuming the ascertainment rate is assumed to be piecewise constant.
+They do not fully propagate the uncertainty from the CIS prevalence estimates, considering only the bounds of the published 95\% confidence intervals.
 
@@
 
-\todo[inline]{Introduce estimating duration}
+The distribution of the duration of positivity is central to the relationship between incidence and prevalence (see \cref{E-inc-prev:sec:prevalence-process}).
+Therefore, a reliable estimate of this quantity is required to achieve this thesis's aims.
 
@@
 Modelling the duration directly, as I do in \cref{E-perf-test,E-imperf-test} considers only the binary result at each test (positive or negative); in this thesis a \emph{survival analysis} framework is adopted.
-This approach generally requires fewer parameters, and no random effect, reducing the computational burden.
+This approach generally requires fewer parameters, and no random effect(s), reducing the computational burden.
 It is also easier to specify non-parameteric forms of the distribution, which is important for SARS-CoV-2 as the tail behaviour may be important.
@@
 % Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
-Other related approaches, such as a multi-state model~\autocite{jacksonMSM}\todo{find best reference for MSMs}, could also be considered (multi-state models can be viewed as a generalisation of survival analysis).
+Other related approaches, such as a multi-state model (reviewed by \textcite{jacksonMSM}), could also be considered; multi-state models can be viewed as a generalisation of survival analysis.
 
@@
 
+\Textcite{ealesCharacterising} estimated the duration of positivity from REACT data.
+Specifically, individuals who tested positive in round 8 of REACT (6th--22nd Jan 2021) were invited to take two additional swab tests.
+Defining the time of the initial positive test at time 0, \textcite{ealesCharacterising} estimated the probability of remaining positive at each day after the initial positive test.
+They made a strong parametric assumption that the probability decreases exponentially, with a sensitivity analysis assuming a plateau before the decay.
+This assumption is necessary because they had at most 17 days of follow-up, and hence could not estimate the tail of the distribution.
+Overall, they estimate a median duration of 14.0 days (95\% CI: 12.9--15.4 days) but that this varied substantially by the viral load of an individual at the time of the initial positive test.
+Individuals with a higher viral load were estimated to have a longer duration of positivity; the highest viral load individuals were estimated to have a median duration of 19.6 days (16.1--25.8 days).
+This could be due to the higher viral load leading to a longer duration of positivity, or that individuals with a higher viral load were more likely to have been recently infected at the time of the initial positive test.
+As they do not consider the latent true time of infection, they cannot distinguish between these two possibilities and underestimate the duration of positivity.
+
 For the purposes of this thesis, estimates applicable to the general population are required.
 Therefore, I estimate the duration of positivity from the ATACCC and CIS studies, as described above.
-
-\todo[inline]{Discuss \textcite{ealesCharacterising}}
 
@@
 
-\section{Thesis aims and contribution}
+\section{Thesis aims and contribution} \label{intro:sec:aims}
 

### 2024-03-07 biology-data.tex
Overleaf comments

--- 2024-02-20 biology-data.tex
@@
 A \emph{standard curve} translates the Ct value to viral load for a given test.
-They are produced by measuring the Ct value using reference samples with known viral load.
-As seen in \cref{biology-data:fig:ct-calibration}, they vary between studies.
+They are produced by measuring the Ct value using reference samples with known viral load, and, as seen in \cref{biology-data:fig:ct-calibration}, they vary between studies.
 \begin{figure}

### 2024-03-07 introduction.tex
Overleaf comments

--- 2024-03-05 introduction.tex
@@
 
-COVID-19 (coronavirus disease 2019), the disease caused by the virus SARS-CoV-2 (severe acute respiratory syndrome coronavirus 2), is estimated to have killed approximately 15 million in the acutest phase (to the end of 2021) and 30 million in total~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
-While would be reassuring if this was a rare or extreme event, but history suggests otherwise (see \cref{intro:fig:pandemic-timeline}).
+The coronavirus disease 2019 (COVID-19) caused by the Ssevere acute respiratory syndrome coronavirus 2 (SARS-CoV-2), is estimated to have killed approximately 15 million in the acutest phase (to the end of 2021) and 30 million globally~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
+While it is tempting to consider this a rare or extreme event, history suggests otherwise (see \cref{intro:fig:pandemic-timeline}).
 \begin{figure}
@@
 
-All pandemics start as an \emph{outbreak}, a sudden and unexpected rise in the number of people with a disease in a local area; if this spreads over a larger geographic area (\eg a country) it becomes an \emph{epidemic}; if it spreads globally then it is a \emph{pandemic}~\autocite{grennanPandemic}.
-The precise meaning of these terms is contested, determined by usage rather than formal definitions~\autocite{morensPandemic,doshiElusive}.
-I will refer to the SARS-CoV-2 epidemic within England (or the UK), and the global impact as the pandemic.
-
-In this context, \emph{surveillance} is the ability to detect and monitor the spread of an outbreak, epidemic, or pandemic.
-Surveillance is important at every stage: early on, for detecting a new outbreak quickly and assessing its likely impact and later for monitoring the ongoing burden of the disease and the effectiveness of interventions~\autocite{whoFluSurveillance}.
-In all cases, surveillance is required to inform public health policy.
+\todo[inline]{Maybe talk about pandemic v epidemic v outbreak terminology?}
+\todo[inline]{Need to motivate the importance of surveillance}
 
@@
 Broadly, a reproduction number describes the number of secondary infections generated by a single infection.
-Here, following \textcite{pellisEstimation} (who themselves are building on over a century of infectious disease dynamics literature), I define three that are broadly useful.
+Here, following \textcite{pellisEstimation},  building on over a century of infectious disease dynamics literature, I define three that are broadly useful.
 
@@
 With constant behaviour, $\Re$ declines smoothly as individuals in the population become immune due to infection.
-$\Re$ is useful because it describes the current dynamics of the epidemic: $\Re > 1$ means the epidemic is growing, $\Re < 1$ means it is shrinking, and $\Re = 1$ means it is stable.
+$\Re$ describes the current dynamics of the epidemic: $\Re > 1$ means the epidemic is growing, $\Re < 1$ means it is shrinking, and $\Re = 1$ means it is stable.
 In the simplest epidemic models (see \cref{E-SEIR:sec:SIR}), $\Re$ is proportional to the product of the rate of contacts and the proportion of the population that is susceptible.
@@
 $\Rc$ is useful when comparing the effects of interventions at different points in time, when the immunity status of the population has changed.
-For example, when comparing the English lockdown in March 2020 and Janurary 2021, there are two effects in play: that there are more individuals with immunity in January 2021, and that the lockdown had different measures in place.
+For example, when comparing the English lockdown in March 2020 and January 2021, there are two effects in play: that there are more individuals with immunity in January 2021, and that the lockdown had different measures in place.
 $\Rc$ focuses on the latter effect.
@@
 To translate incidence into the burden of the disease, a measure of severity is required.
-For example, the IFR (infection fatality rate), the proportion of infected individuals who die.
+For example, the infection fatality rate (IFR), the proportion of infected individuals who die.
 Estimation of severity is beyond the scope of this thesis.
@@
 Therefore, fluctuations in the number of positive tests may not reflect changes in the true incidence of the disease.
-For large parts of the pandemic, community testing was widely available in the UK meaning that this is a large dataset, allowing detailed analyses to be performed, including stratification by small geographic areas or age groups.
+For large parts of the pandemic, community testing was widely available in the UK meaning resulting in a rich dataset, allowing detailed analyses to be performed, including stratification by small geographic areas or age groups.
 
-Seroprevalence testing is a marker of recent prior infection.
+Seroprevalence testing provides information on a marker of recent prior infection, by measuring ..????
 Prior infection is more common than current infection, hence seroprevalence estimation requires a smaller sample size than current prevalence.
@@
 Severe events are generally a reliable measure of the burden of the disease.
-The criteria for being a severe event are well-defined and generally well-recorded in the UK context.
+The criteria for being a severe event are well-defined and generally well-recorded in the UK context (reference?).
 Therefore, the data is subject to fewer biases than community testing.
-However, SARS-CoV-2 rarely leads to severe events in younger individuals, meaning severe events only capture transmission to a subset of the population.
+However, SARS-CoV-2 rarely leads to severe events in younger individuals, meaning severe events only capture sequelae of transmission in a subset of the population.
 
@@
 The use of prevalence surveys for estimating incidence and transmission is the focus of this thesis.
-In particular, I focus on the CIS (Coronavirus (COVID-19) Infection Survey), run by the Office for National Statistics (ONS)~\autocite{CIS,cisMethodsONS}.
+In particular, I focus on the Coronavirus (COVID-19) Infection Survey (CIS ), run by the Office for National Statistics (ONS)~\autocite{CIS,cisMethodsONS}.
 The CIS was a longitudinal study, testing a large number of indidividuals weekly or monthly for SARS-CoV-2; it is described in detail in \cref{E-intro:sec:cis}.
 Early in the pandemic, a methodology was developed for using CIS data to estimate prevalence (see \cref{E-intro:sec:cis}).
-
-A second prevalence survey ran in the UK, known as REACT (REal-time Assessment of Community Transmission)~\autocite{rileyResurgence,rileyREACT}.
-REACT was a series of cross-sectional surveys, run across 19 rounds between May 2020 and March 2022.
-Each round consisted of at least 100,000 randomly-sampled individuals from the UK contributing swabs for RT-PCR testing and questionnaire data~\autocite{elliottTwin}.
-
-A major advantage of the CIS over REACT is that the CIS has daily testing of individuals, while REACT has no test results between rounds.
-This means that REACT cannot be used to estimate transmission or incidence between rounds, instead relying on model assumptions at these times.
-REACT missed the peak in late December 2020, which complicated analysis of the survey data at that time~\autocite{rileyREACTround8}.
-
-The primary aim of this thesis is to estimate incidence and transmission from the CIS data.
+The primary aim of this thesis is to estimate incidence and transmission from this data.
 
@@
 As noted above, incidence and transmission are extremely important quantities for informing the response to a pandemic but cannot be observed directly.
-Therefore, they must be estimated using some form of modelling.
+Therefore, they must be estimated.
 
@@
 In this context, backcalculation estimates the incidence of HIV infection from data on the number of AIDS diagnoses, and the time between infection and AIDS diagnosis~\autocite{brookmeyerBackcalculation,brookmeyerMeasuring}.
-Two processes needed to be considered in the context of HIV/AIDS: one generating infections and one relating infections to prevalence.
+Two processes needed to be considered: one generating infections and one relating infections to prevalence.
 % AIDS is incurable, therefore, AIDS prevalence is equivalent to cumulative incidence.
 
-The statistical models used to describe HIV transmission are applicable to SARS-CoV-2.
-This infection process can be described as a Poisson process (see \cref{E-inc-prev:sec:infection-process}).
+The statistical framework developed for HIV is applicable to SARS-CoV-2.
+(see \cref{E-inc-prev:sec:infection-process}).
 
-Unlike HIV/AIDS, recovery from SARS-CoV-2 happens in the vast majority of infections.
-Therefore, recoveries need to be considered in the backcalculation.
+However, Unlike HIV/AIDS, recovery from SARS-CoV-2 infection is possible, so recoveries need to be considered in the backcalculation.
 This introduces a dependency structure in the prevalence data (see \cref{E-inc-prev:sec:prevalence-process}).
@@
 Furthermore, prevalence surveys are noisy because they are based on population sampling.
-This noise needs to be considered in the application of backcalculation (see \cref{E-inc-prev:sec:observation-process}).
+This noise needs to be considered in the development of the backcalculation approach (see \cref{E-inc-prev:sec:observation-process}).
 
@@
 The ability to estimate the contribution of each of these factors can inform public health policy.
-Furthermore, mechanistic models can be used for scenario modelling, which simulates the effect of interventions.
-A scenario is the effect of a proposed intervention (or lack thereof), simulated by modifying the relevant parameters.
+Furthermore, mechanistic models can be used for scenario-based modelling, which simulates the effect of interventions.
+A scenario here is the effect of a proposed intervention (or lack thereof), introduced into the model by modifying the relevant parameters.
 % The mechanistic model can then be simulated forward in time to understand the intervention's effects.
@@
 Many mechanistic modelling studies were conducted during the pandemic, with at least 100 articles published in the first six months of 2020 alone~\autocite{shankarSystematic}.
-In the UK, SPI-M-O (Scientific Pandemic Influenza Group on Modelling, Operational sub-group) was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), generating evidence based on modelling for policy development~\autocite{medleySPIM,govSPIMO}.
-For example, the stages of lifting the lockdown in 2021 each had a set of modelling studies informing the government's decision~\autocite{sageEvidence}.
+In the UK, the most influential were the outputs from SPI-M\todo{cite SPI-M and its outputs - Medley paper?}.
+For example, the stages of lifting the lockdown in 2021 each had a set of modelling studies informing the government's decision\todo{cite roadmap 2021 papers}.
 
-Each week, the members of SPI-M would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
-The nowcasting estimates were based on mechanistic models.
-These estimates, once any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value~\autocite{parkCombining}.
-The consensus value was published on \url{gov.uk} and informed the government's response to the epidemic~\autocite{govRnumber}.
-The academic groups contributing estimates used a variety of models, mostly based on either \emph{renewal equations} or \emph{compartmental models}.
-The methods were reviewed by \textcite{royalSocietyRnumber}; see appendix 2 of that review for details of the specific implementations.
+Each week, the members of SPI-M would produce a set of estimates of the current state of the epidemic (\eg ther value of $\Re$), known as \emph{nowcasting}.
+These estimates, once any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value.\todo{cite SPI-M combination}
+The consensus value was published on \url{gov.uk} and informed the government's response to the epidemic.\todo{cite gov.uk R number}
 
-The renewal equation~\autocite{fraserRenewal} for an infectious disease, in the most general form, is:
-\begin{align}
-    I(t) &= \int_{0}^{t} \beta(t,\tau) I(t-\tau) d \tau
-\end{align}
-where $I(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the instantaneous probability of an individual infected $\tau$ time units ago infects an individual at time $t$.
-A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
-The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
-Renewal equations do not feature in this thesis, for further information, including practical considerations, see \textcite{nashEstimating,gosticPractical}.
+The academic groups contributing estimates used a wide variety of methodologies
 
-Compartmental models split the population into compartments, each representing a different state of the disease.
-The \emph{dynamics} of the epidemic, how it changes over time, are described by the transitions between compartments.
-Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
-Unlike renewal equations, there are not widely-used implementations of compartmental models.
-Most of the implementations used by SPI-M-O members were in the SEIR (susceptible, exposed, infectious, recovered) framework.
-The SEIR framework, has four components (susceptible, exposed, infectious, recovered), with the population divided into these components.
-The SEIR framework is described in detail in \cref{E-SEIR:sec:mechanistic-models}.
-Compartmental models can be fit to almost any data source, and commonly incorporate multiple data sources, known as \emph{data synthesis}, a technique reviewed by \textcite{birrellEvidence}.
-Most compartmental models used by SPI-M-O were age-stratified, as differences between age groups are important for both the transmission and severity of SARS-CoV-2 (see \cref{E-SEIR:sec:structured-populations}).
+\todo[inline]{Finish lit review of SPI-M methods}
 
-Several compartmental models incorporate data from SARS-CoV-2 prevalence surveys, alongside other data sources~\autocite{daviesAssociation,ironsEstimating,knockKey,nicholsonImproving,pooleyEstimation,birrellRTM2}.
-These models incorporate prevalence survey data by adding additional compartments representing RT-PCR individuals, who may not be infectious (this is common, see \cref{E-biology-data}); except \textcite{nicholsonImproving} who have a more sophisticated approach (below) and \textcite{ironsEstimating}, who assume RT-PCR positivity coincides with infectiousness.
-The length of stay in these compartments (the duration of time positive) is assumed to follow a gamma distribution, with the parameters estimated from external data or having at most one free parameter.
-The proportion of the population in these compartments are then fit to the prevalence survey data using a binomial likelihood where possible, although some approaches take a Gaussian approximation.
-These approaches are similar to how these models are fit to other data sources, such as hospital admissions or deaths.
-
-\Textcite{nicholsonImproving} have the most sophisticated use of prevalence survey data.
-They allow an arbitrary distribution of the duration of positivity, based on the estimates of \textcite{hellewellPCRSensitivity} (see \cref{intro:sec:duration-SARS-CoV-2} for discussion of the \textcite{hellewellPCRSensitivity} study).
-Furthermore, they incorporate the sensitivity and specificity of the RT-PCR test into their model.
-% However, they do not incorporate age-stratification into their transmission model, and primarily aim to estimate the bias of community testing in the UK rather than incidence and/or transmission.
-
-In \cref{E-SEIR}, I describe a compartmental model for SARS-CoV-2 transmission, which I fit to CIS data.
-I allow a more flexible distribution of the duration of positivity than most of the papers discussed above, incorporating the skew of the distribution.
-This distribution is based on data from a representative sample of the UK population, rather than convenience samples (see \cref{intro:sec:duration-SARS-CoV-2}).
-Unlike \cref{nicholsonImproving}, I include age-stratification in the transmission model, allowing estimation of how transmission varies by age.
-I use only data from the CIS, and show this is sufficient for estimating incidence and transmission.
-Finally, I fit directly to counts from the CIS, rather than an approximation of this data, which has not previously be done.
+\begin{itemize}
+    \item SPI-M
+    \item Literature on mechanistic models using REACT and CIS I have previously circulated.
+\end{itemize}
 
@@
 \section{Estimating duration} \label{intro:sec:duration}
-
-\todo[inline]{Introduce estimating duration}
 
@@
 % Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
-Other related approaches, such as a multi-state model~\autocite{jacksonMSM}\todo{find best reference for MSMs}, could also be considered (multi-state models can be viewed as a generalisation of survival analysis).
+Other related approaches, such as a multi-state model~\autocite{jacksonMSM}, could also be considered (multi-state models can be viewed as a generalisation of survival analysis).
 
@@
 The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability, but the results are limited because the small sample size and short follow-up means that the tails of these distributions cannot be estimated.
-I describe ATACCC in detail in \cref{E-biology-data:sec:ataccc} and use it to estimate duration in \cref{E-ATACCC}.
+I describe ATACCC in detail in \cref{E-biology-data:sec:ataccc} and analyse it in \cref{E-ATACCC}.
 
@@
 In particular, the vast majority of detected infections had only one or two positive tests, and hence the viral load trajectory of these infections cannot be estimated.
-I describe the CIS in detail in \cref{E-intro:sec:cis} and use it to estimate duration in \cref{E-perf-test,E-imperf-test}.
+I describe the CIS in detail in \cref{E-intro:sec:cis} and analyse it in \cref{E-perf-test,E-imperf-test}.
 
@@
 
-For the purposes of this thesis, estimates applicable to the general population are required.
-Therefore, I estimate the duration of positivity from the ATACCC and CIS studies, as described above.
-
 \section{Bayesian inference} \label{intro:sec:Bayes}
 
-All inference in this thesis is Bayesian.
-Bayesian inference has several relevant advantages.
-Prior information allows identifiability in situations where this would not otherwise be possible, allows for combining knowledge from multiple sources, and many previous studies I build used Bayesian inference.
+All inference in this thesis is performed within the Bayesian framework.
+Bayesian inference has several advantages relevant for this work.
+Prior information allows identifiability in situations where this would not otherwise be possible, allows for combining knowledge from multiple sources, and much prior work I build on has been performed within the Bayesian framework.
 
@@
 In particular, an epidemic which will cause issues starts by increasing, therefore if the methods are to be applied early in a future epidemic or pandemic, they will start with an increasing period.
-\todo{Rephrase this sentence}
-The CIS began in April 2020, when incidence was falling due to the first lockdown in the UK.
+The CIS began in Aril 2020, when incidence was falling due to the first lockdown in the UK.
 Incidence remained low and fairly flat in England over summer 2020, until it began rising in mid or late August.
@@
 
-First, the framework to simulation the CIS (used in \cref{E-imperf-test:sec:simulate,E-perf-test:sec:simulation-study}).
+First, the framework to simulation the CIS (used in \cref{E-imperf-test:sec:simulate,E-perf-test:sec:simulation-study}) is structured as a R package.
 The package enables customisation of various features of the study, for example the testing regime of the participants, the sensitivity of the tests, and the duration of positivity.
@@
 However, it assumes that the prevalence at any time is a random variable, instead modelling it as a deterministic function of the incidence.
-\todo{Rephrase this sentence}
 I show that, for all that is required for this assumption to hold is that the proportion of the population sampled is small.

### 2024-03-07 transmission.tex
Resolve to dos
Overleaf comments

--- 2024-02-29 transmission.tex
@@
 However, in SARS-CoV-2, population behaviour changed dramatically, both spontaneously (\eg due to fear) and in response to government-imposed restrictions.
-Therefore, following \textcite{birrellRealtime}, I introduce a time-dependence giving: $\beta_{aa'}(t) = \kappa_{aa'}(t) q_{aa'}(t) N_{a'} / N_a$; $N_a$ and $N_{a'}$ are constant because the change in the total population is negligible over the length of time considered here.
+Therefore, following \textcite{birrellRealtime}, I introduce a time-dependence giving: $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'}(t) N_{a'} / N_a$; $N_a$ and $N_{a'}$ are constant because the change in the total population is negligible over the length of time considered here.
 
@@
 The full system of differential equations for each region is below.
-\begin{align}
-    \label{SEIR:eq:fullODEs}
-    \vec{\lambda}(t) &= \beta(t) (\matr{\kappa}(t) \circ \matr{q}) ((\vec{i_1}(t) + \vec{i_2}(t)) \circ \vec{N}) \\
+\begin{equation}
+\begin{aligned}
+    \vec{\lambda}(t) &= \beta(t) (\matr{\kappa}(t) \circ \matr{q} \circ \matr{N}) ((\vec{i_1}(t) + \vec{i_2}(t))) \\
     \frac{d\vec{s}(t)}{dt} &= -\vec{\lambda}(t) \circ \vec{s}(t) \\
@@
     \frac{d\vec{r}(t)}{dt} &= 2\gamma \vec{i_2}(t) \\
-    \frac{d\vec{p_0}(t)}{dt} &= \vec{\lambda} \circ \vec{s}(t) \circ \vec{N} - \omega_0 \vec{p_0}(t) \\
+    \frac{d\vec{p_0}(t)}{dt} &= \vec{\lambda} \circ \vec{s}(t) - \omega_0 \vec{p_0}(t) \\
     \frac{d\vec{p_1}(t)}{dt} &= \omega_0 \vec{p_0}(t) - \omega_1 \vec{p_1}(t) \\
@@
     \frac{d\vec{p_3}(t)}{dt} &= \pi \omega_{2} \vec{p_2}(t) - \omega_{3} \vec{p_3}(t)
-\end{align}
-where $\circ$ indicates element-wise multiplication.
+\end{aligned}
+\label{SEIR:eq:fullODEs}
+\end{equation}
+where $\circ$ indicates element-wise multiplication and $\matr{N}$ is a matrix with $(\matr{N})_{aa'} = N_{a'} / N_{a}$.
+See \cref{SEIR:eq:fullODEs}
 
@@
 The remaining parameters have weakly informative priors.
+
+To form the prior for $\vec{\pi}$, I used the analysis framework of \textcite{amirthalingamSeroprevalence}, applied to seroprevalence data from blood donors~\autocite{amirthalingamSeroprevalence} and a representative sample of children~\autocite{ratcliffeCommunity} in August 2020.
+This analysis uses a Bayesian framework to correct the seroprevalence data for the sensitivity and specificity of the serological tests, using an informative prior.
+\Textcite{amirthalingamSeroprevalence} formed the prior for the tests by evaluating the tests using samples which are known to be negative (taken before the pandemic) or positive (taken from an individual following a positive RT-PCR test).
+A regression model predicts the proportion of the population with prior immunity in each age group and region.
+I approximated the posterior of this proportion with a beta distribution to form the prior for the analysis in this chapter (see \cref{SEIR:table:immunity-prior}).
+
 \begin{landscape}
@@
 \end{table}
+\begin{table}
 
-To form the prior for $\vec{\pi}$, I used the analysis framework of \textcite{amirthalingamSeroprevalence}, applied to seroprevalence data from blood donors~\autocite{amirthalingamSeroprevalence} and a representative sample of children~\autocite{ratcliffeCommunity} in August 2020.
-This analysis uses a Bayesian framework to correct the seroprevalence data for the sensitivity and specificity of the serological tests, using an informative prior.
-\Textcite{amirthalingamSeroprevalence} formed the prior for the tests by evaluating the tests using samples which are known to be negative (taken before the pandemic) or positive (taken from an individual following a positive RT-PCR test).
-A regression model predicts the proportion of the population with prior immunity in each age group and region.
-I approximated the posterior of this proportion with a beta distribution to form the prior for the analysis in this chapter (see \cref{SEIR:table:immunity-prior}).
-\begin{table}
 \centering
@@
 
-I transformed parameters with restricted ranges for sampling, to improve efficiency.
+I transformed parameters with bounded supports to the real line to improve sampling efficiency.
 Parameters which are constrained to the positive reals ($\sigma_\epsilon$, $\rho$, and $q_c$) were sampled on the log scale.
@@
 
-I run 4 chains for 1,000,000 iterations each.
+All results are based on four Markov chains each of length 1,000,000 iterations.
 The first 500,000 were discarded as burn-in, with the remaining chains thinned to every 100th iteration.
@@
 Following \textcite{birrellBayesian}, I reparameterise the model to improve identifiability.
-Rather than directly inferring $\beta_1$, I infer the initial growth rate of the epidemic.
-In addition, I use a parsimonious parameterisation of the initial conditions using the initial proportion of the population that are infectious and the initial proportion that are susceptible only.
-Both of these rely on an assumption that the epidemic is in a steady-state exponential growth phase at time 0 (see \cref{E-SEIR:sec:SIR}).
+I parameterise $\beta_1$ in terms of the initial growth of the epidemic.
+I also use a parsimonious parameterisation of the initial conditions using the initial proportion of the population that are infectious and the initial proportion that are susceptible only.
+
+Both reparameterisations rely on an assumption that the epidemic is in a steady-state exponential growth phase at time 0 (see \cref{E-SEIR:sec:SIR}).
 During such an exponential growth phase, all components of the $\vec{e}$, $\vec{i}$ and $\vec{p}$ compartments grow at the same growth rate, $\psi$.
+That is:
+\begin{equation}
+\begin{aligned}
+\frac{d\vec{e_1}(t)}{dt} = \psi \vec{e_2}(t) \\
+\frac{d\vec{e_2}(t)}{dt} = \psi \vec{e_2}(t) \\
+\frac{d\vec{i_1}(t)}{dt} = \psi \vec{i_2}(t) \\
+\frac{d\vec{i_2}(t)}{dt} = \psi \vec{i_2}(t) \\
+\frac{d\vec{p_0}(t)}{dt} = \psi \vec{p_0}(t) \\
+\frac{d\vec{p_1}(t)}{dt} = \psi \vec{p_1}(t) \\
+\frac{d\vec{p_2}(t)}{dt} = \psi \vec{p_2}(t) \\
+\frac{d\vec{p_3}(t)}{dt} = \psi \vec{p_3}(t).
+\end{aligned}
+\label{SEIR:eq:exponential-growth}
+\end{equation}
 This follows directly from the ODEs if $\vec{s}$ is constant.
+
 The growth rate $\psi$ is easily identifiable from the data, and therefore a parameterisation in terms of $\psi$ should be more computationally efficient.
-If this is the initial exponential phase, when $\vec{s} \approx 1$, then the basic reproduction number for a SEIR model is linked to this growth rate by~\autocite{birrellBayesian,wearingAppropriate}:
+If this is the initial exponential phase, when $\vec{s} \approx 1$, then the basic reproduction number $\R$ for a SEIR model is linked to this growth rate by~\autocite{birrellBayesian,wearingAppropriate}:
 \begin{align}
-    \R = \frac{\psi \left( \frac{\psi}{2\sigma} + 1 \right)^2}{\gamma \left(1 - \frac{1}{\left(\frac{\psi}{2 \gamma} + 1 \right)^2} \right)} \label{SEIR:eq:rtoR}.
+    R_{(\psi)} = \frac{\psi \left( \frac{\psi}{2\sigma} + 1 \right)^2}{\gamma \left(1 - \frac{1}{\left(\frac{\psi}{2 \gamma} + 1 \right)^2} \right)} \label{SEIR:eq:rtoR}.
 \end{align}
 
-$\R$ can also be expressed a function of $\matr{\beta}(0)$, and hence $\beta_1$ (as explained in \cref{E-SEIR:sec:structured-populations}).
-Combining these expressions, the model can be parameterised by using $\psi$ in place of $\beta_1$ by using the following procedure.
-Start by calculating $\R$ from $\psi$ using \cref{SEIR:eq:rtoR}; denote this $\hat{R}$.
-$\hat{R}$ is the $\R$ that the current value of $\psi$ implies.
-Hence, $\beta_1$ should be set such that $\matr{\beta}(0)$ implies $\R = \hat{R}$.
-To do so, define $R^*$ as $\R$ $\beta_1 = 1$.
-Using the results of \cref{E-SEIR:sec:structured-populations}, $R^* = D(\matr{\beta}(0)) / \gamma$, computed using $\beta_1=1$ (recall that $D(\matr{\beta})$ denotes the dominant eigenvalue of $\matr{\beta}$).
-Finally, set $\beta_1 = \hat{R} / R^*$; as multiplying a matrix by a scalar also multiplies its dominant eigenvalue by that same scalar, this means that $\R$ at time 0 is $\hat{R}$, as desired.
+Using the results of \cref{E-SEIR:sec:structured-populations}, $\R$ can also be written as $R_{(\beta)} = \beta_1 D(\matr{\kappa}(t) \circ \matr{q} \circ \matr{N})$ (recall that $D(\matr{\beta})$ denotes the dominant eigenvalue of $\matr{\beta}$).
+Hence setting $\beta_1 = R_\psi / D(\matr{\kappa}(t) \circ \matr{q} \circ \matr{N})$ gives $R_\psi = R_\beta$.
+Using this expression, $\beta_1$ has been reparameterised in terms of $\psi$, $q_c$, and fixed quantities.
 
@@
 Recall, from \cref{E-SEIR:sec:structured-populations}, that in the exponential growth phase, the relative proportion of each stratum that is infectious is proportional to the dominant eigenvector of $\matr{\beta}$, which I denote $\vec{d}$.
-Introduce a model parameter $i^+$, the proportion of the population in the $i$ states at time 0 in the strata with the highest proportion infectious.
-Define the total number of infectious individuals at time 0 as $\vec{i_0} = \vec{i_1} + \vec{i_2} = \frac{i^+}{\max \vec{d}} \vec{d}$.
+Introduce a model parameter $i^+$, the proportion of the population in either of the $i$ states at time 0 in the strata with the highest proportion infectious.
+Define the total number of infectious individuals at time 0 as $\vec{i_0}$.
+Then:
+\begin{align}
+\vec{i_0} = \vec{i_1}(0) + \vec{i_2}(0) = \frac{i^+}{\max \vec{d}} \vec{d}.
+\label{SEIR:eq:i0}
+\end{align}
 
-From the system of ODEs, the knowledge that all the compartments are growing exponentially at rate $\psi$, and the parameters $\vec{\pi}$ (where $\pi_a$ is the proportion of strata $i$ without immunity at time 0, hence, $\vec{r} = 1 - \vec{\pi}$) one can derive the following for the initial state of the compartments:
+Equating the expressions for $d\vec{i_2}/dt$ in \cref{SEIR:eq:fullODEs} and \cref{SEIR:eq:exponential-growth} and then setting $t=0$ to allow substituting in \cref{SEIR:eq:i0} gives:
 \begin{align}
-    \vec{i_1} &= \vec{i_0} \left(1 + \frac{2\gamma}{\psi + 2\gamma} \right)^{-1} \\
-    \vec{i_2} &= \vec{i_0} - \vec{i_1} \\
-    \vec{e_2} &= \vec{i_1} \frac{\psi + 2\gamma}{2\sigma} \\
-    \vec{e_1} &= \vec{e_2} \frac{\psi + 2\sigma}{2\sigma} \\
-    \vec{s} &= \vec{\pi} - \vec{i_0} - \vec{e_1} - \vec{e_2} \\
-    \vec{p_0} &= \vec{e_1} \frac{\psi + 2\sigma}{\psi + \omega_0} \\
-    \vec{p_1} &= \vec{p_0} \frac{\omega_0}{\psi + \omega_1} \\
-    \vec{p_2} &= \vec{p_1} \frac{\omega_1}{\psi + \omega_{2}} \\
-    \vec{p_3} &= \vec{p_2} \frac{\pi \omega_{2}}{\psi + \omega_{3}}.
+    \vec{i_1}(0) &= \vec{i_0} \left(1 + \frac{2\gamma}{\psi + 2\gamma} \right)^{-1}.
+\intertext{By definition of $\vec{i_0}$:}
+    \vec{i_2}(0) &= \vec{i_0} - \vec{i_1}(0).
 \end{align}
+
+A similar process, equating the expressions in \cref{SEIR:eq:fullODEs} and \cref{SEIR:eq:exponential-growth}, gives the following solution for the initial state of these compartments.
+\begin{align}
+    \frac{d\vec{e_1}(t)}{dt} &= \vec{\lambda}(t) \circ \vec{s}(t) - 2\sigma \vec{e_1}(t) \\
+    \frac{d\vec{e_2}(t)}{dt} &= 2\sigma \vec{e_1}(t) - 2\sigma \vec{e_2}(t) \\
+    \frac{d\vec{p_0}(t)}{dt} &= \vec{\lambda} \circ \vec{s}(t) - \omega_0 \vec{p_0}(t) \\
+    \frac{d\vec{p_1}(t)}{dt} &= \omega_0 \vec{p_0}(t) - \omega_1 \vec{p_1}(t) \\
+    \frac{d\vec{p_2}(t)}{dt} &= \omega_1 \vec{p_1}(t) - \omega_{2} \vec{p_2}(t) \\
+    \frac{d\vec{p_3}(t)}{dt} &= \pi \omega_{2} \vec{p_2}(t) - \omega_{3} \vec{p_3}(t).
+\end{align}
+
+The definition of $\vec{\pi}$ as the proportion of the population without initial immunity gives $\vec{r}(0) = 1 - \vec{\pi}$.
+Finally, to ensure the population proportions sum to one, $\vec{s}(0) = \vec{\pi} - \vec{i_0} - \vec{e_1} - \vec{e_2}$.
 
@@
 I ran a simulation study to check that the inference procedure (\cref{SEIR:sec:inference}) was working correctly and that the model was identifiable.
-I ran 100 simulations, with parameters drawn from the priors (see \cref{SEIR:table:priors}) except that the prior on $\psi$ was $\LN(0.048, 0.0035)$.
-I used the more informative prior for the simulation study to ensure that the simulations were realistic.
+I generated 100 simulated datasets representing the East of England using the following procedure.
+
+First, I drew 100 parameter sets drawn from the priors (see \cref{SEIR:table:priors}), except that the prior on $\psi$ was $\LN(0.048, 0.0035)$.
+I used this more informative prior for the simulation study to restrict the simulations to initially-growing epidemics.
 The weakly informative prior, used for the main analysis, allows a wide range of epidemics including some that are shrinking at the start of the time period.
-Priors and population sizes were matched to the East of England region, over the same time period as the main analysis.
+The prior for $\vec{\pi}$ was the same as the East of England.
 
-All other aspects of the simulation model, including $\matr{\kappa}$, are as previously described in this section.
+Next, I solved the ODE system in \cref{SEIR:eq:fullODEs} for each parameter set.
+All other aspects of the ODEs, including the data-informed contact matrices, $\matr{\kappa}$, are as previously described in this section.
+Population sizes were those of the East of England, and the time period the same as the main analysis.
 This means that the simulations created realistic incidence and prevalence curves (see \cref{SEIR:fig:sim-data}).
@@
 
-% Furthermore, the combination of the posteriors across all simulations should recover the prior.
+Finally, I generated one dataset of test results for each parameter set.
+These were generated according to the likelihood, \cref{SEIR:eq:likelihood}.
+The number of daily tests was set to the mean of what CIS performed in the region over this period (1630), distributed across the population in proportion to its size.
 
-To check the performance of the inference procedure, I calculated the coverage of the 50\%, 75\%, and 95\% central credible intervals for each parameter and simulation.
-Define coverage as the proportion of simulations where the true value of the parameter in that simulation was within the credible interval.
-As the parameters are drawn from the prior, if the inference is performed correctly then the coverage for each parameter should equal the nominal probability (\eg the coverage should be 50\% for the 50\% credible intervals)~\autocite{cookValidation}.
+Having generated 100 datasets, I used the inference procedure in \cref{SEIR:sec:inference} to find the posterior distribution conditional on each dataset.
+Hence, I ended up with 100 parameter sets distributed according to the priors (the ``generative values"), with an associated dataset and posterior distribution for it.
+
+To check the performance of the inference procedure, I calculated the coverage of the 50\%, 75\%, and 95\% central credible intervals for each posterior distribution.
+Define coverage for a parameter as the proportion of credible intervals for that parameter containing the generative value.
+As the parameters are drawn from the prior, if the posterior distribution is inferred correctly then the coverage for each parameter should equal the nominal probability (\eg the coverage should be 50\% for the 50\% credible intervals)~\autocite{cookValidation}.
 This check is related to, although simpler than, simulation-based calibration~\autocite{taltsValidating}.
@@
 This is a common issue with random walk MCMC (see \cref{E-inc-prev:sec:MCMC}).
-This is similar when considering coverage of the derived quantities incidence and prevalence (see \cref{SEIR:fig:sim-inc-prev}).
+The same issue arises with the coverage for derived quantities incidence and prevalence (see \cref{SEIR:fig:sim-inc-prev}).
 \begin{figure}
@@
 
-In this setup, the parameters $i^+$, $q_c$, and $\psi$ are fully identified; $\pi_0$ is not identified at all; the other parameters are partially identified, their posterior has moved away from the prior but is still regularised by it (see \cref{SEIR:fig:true-vs-posterior}).
+In this setup, the parameters $i^+$, $q_c$, and $\psi$ are fully identified; $\pi_0$ is not identified at all; the other parameters are partially identified, their posterior has moved away from the prior but there is insufficient information in the data to fully overcome the prior (see \cref{SEIR:fig:true-vs-posterior}).
 The ability to identify $q_c$ is an advantage of using data which includes measurements on children, unlike data on hospitalisations or deaths.
@@
 A constant $\matr{\beta}(t)$ would lead to exponential growth following a period of transitory dynamics, although the weekly switching may mean that these transitory dynamics never end, especially when incidence is high~\autocite{rhodesConvergence}.
-The estimated timing of the peak in incidence varies by region, with the most common time being early January 2021 (see \cref{SEIR:fig:peak-incidence}).
+The estimated timing of the peak in incidence varies by region, with the most common time being late December 2020 (see \cref{SEIR:fig:peak-incidence}).
+In Yorkshire and the Humber, the October peak is estimated to be greater than the winter peak; for North West there is significant probability on either being larger.
 \begin{figure}
@@
 The \emph{attack rate} is the proportion of the population that have been infected at least once.
-In this model it is $1 - \vec{s}$, the proportion susceptible.
+In this model it is $1 - \vec{s}$, the proportion susceptible at the time the study ends.
 The attack rate is highest in the $[11, 16)$ age group in the North West at 68\% (95\% CrI: 66--70\%) (see \cref{SEIR:fig:attack-rates}).
-Some of these infections occurred prior to the period covered by this study; during this period 23\% (95\% CrI: 21--24\%) of the population in the North West aged 11--16 were infected.
+Some of these infections occurred prior to the period covered by this study; during the period of this study 23\% (95\% CrI: 21--24\%) of the population in the North West aged 11--16 were infected.
 \begin{figure}
@@
     \label{SEIR:fig:prev-old}
+    \todo[inline]{Remake/reposition this plot so that it fits on a page.}
 \end{figure}
@@
 All regions' MCMC chains converged, assessed using the Rhat statistic and ESS (as defined in \cref{E-intro:sec:MCMC}).
-The lowest Rhat was 1.01 for $\sigma_\epsilon$ in London.
+The highest Rhat was 1.01 for $\sigma_\epsilon$ in London.
 All parameters had ESS > 1000, except $\sigma_\epsilon$ and $\psi$ in London, which had ESS of 450 and 950 respectively.
@@
 By understanding which of these components is changing, policies can be better targeted.
-If it is shown that $\beta(t)$ decreases, this implies that for a given contact rate in the population, there is less transmission.
-Such decreases might be beneficial as it allows society to operate in a more normal manner without increasing transmission.
+If it is shown that $\beta(t)$ decreases, this implies that there is less transmission for the same number of contacts.
+% Such decreases might be beneficial as it allows society to operate in a more normal manner without increasing transmission.
 % Increasing activity without increasing transmission is desirable because it enables reductions the disease burden without burdesome restrictions on individuals day-to-day lives.
@@
 Many factors contribute to the pattern in the $\beta_w$ values, which have a minima near the start of December.
-While this analysis cannot establish causation, several hypotheses that are worth considering.
+While this analysis cannot establish causation, several hypotheses are worth considering.
 
@@
 
-The only time-changing part of $\matr{\beta}$ that varies between ages is $\matr{\kappa}$.
+The only component of $\matr{\beta}$ that allows age-dependent heterogeneity in the temporal changes is $\matr{\kappa}$.
 It is plausible that this would not capture all behavioural changes.

### 2024-03-07 introduction.tex
proof intro

--- 2024-03-07 introduction.tex
@@
 COVID-19 (coronavirus disease 2019), the disease caused by the virus SARS-CoV-2 (severe acute respiratory syndrome coronavirus 2), is estimated to have killed approximately 15 million in the acutest phase (to the end of 2021) and 30 million in total~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
-While would be reassuring if this was a rare or extreme event, but history suggests otherwise (see \cref{intro:fig:pandemic-timeline}).
+History suggests severe pandemics are not rare (see \cref{intro:fig:pandemic-timeline}).
 \begin{figure}
@@
 However, our best records suggests that, since 1600, they have (on average) occurred every four years with a mean death toll of 5.6 million~\autocite{maraniNovelEpidemics}.
-Modern estimates suggest that we should expect over 2.5 million deaths per year from pandemics~\autocite{madhavPandemicMortality}.
+Recent estimates suggest that the current mean impact of pandemics is over 2.5 million deaths per year~\autocite{madhavPandemicMortality}.
 
@@
 In particular, it uses the behaviour and immunity status at time $t$.
-With constant behaviour, $\Re$ declines smoothly as individuals in the population become immune due to infection.
+With constant behaviour and a disease that grants immunity following infection, $\Re$ declines smoothly.
 $\Re$ is useful because it describes the current dynamics of the epidemic: $\Re > 1$ means the epidemic is growing, $\Re < 1$ means it is shrinking, and $\Re = 1$ means it is stable.
-In the simplest epidemic models (see \cref{E-SEIR:sec:SIR}), $\Re$ is proportional to the product of the rate of contacts and the proportion of the population that is susceptible.
-Therefore, it informs the strength of intervention (either reducing contacts or amount of immunisation) required to control the epidemic.
+In simple epidemic models (see \cref{E-SEIR:sec:SIR}), $\Re$ is proportional to the product of the rate of contacts, the riskiness of those contacts, and the proportion of the population that is susceptible.
+Therefore, it informs the strength of intervention, \eg social distancing or vaccination, required to control the epidemic.
 
@@
 Therefore, fluctuations in the number of positive tests may not reflect changes in the true incidence of the disease.
-For large parts of the pandemic, community testing was widely available in the UK meaning that this is a large dataset, allowing detailed analyses to be performed, including stratification by small geographic areas or age groups.
+For large parts of the pandemic, community testing was widely available in the UK.
+Therefore, this is a large dataset, allowing detailed analyses to be performed, including stratification by small geographic areas or age groups.
 
@@
 However, there is a delay between infection and the development of antibodies, hence seroprevalence is not a real-time measure of the burden of the disease.
+Furthermore, seroprevalence data is normally based on convenience samples (\eg blood donors) who are not representative of the general population.
 
 Severe events are generally a reliable measure of the burden of the disease.
-The criteria for being a severe event are well-defined and generally well-recorded in the UK context.
+The criteria for being a severe event are normally well-defined and well-recorded in the UK context.
 Therefore, the data is subject to fewer biases than community testing.
@@
 This means that REACT cannot be used to estimate transmission or incidence between rounds, instead relying on model assumptions at these times.
-REACT missed the peak in late December 2020, which complicated analysis of the survey data at that time~\autocite{rileyREACTround8}.
+A particular instance of this is that a peak in late December 2020 occurred between rounds of the REACT survey, complicating analysis of the survey data at that time~\autocite{rileyREACTround8}.
 
@@
 Furthermore, prevalence surveys are noisy because they are based on population sampling.
-This noise needs to be considered in the application of backcalculation (see \cref{E-inc-prev:sec:observation-process}).
+This observation noise needs to be considered in the application of backcalculation (see \cref{E-inc-prev:sec:observation-process}).
 
-The original formulations of backcalculation neglected the observation process.
+The original formulations of backcalculation neglected any observation noise.
 AIDS is almost always diagnosed and recorded in developed countries.
@@
 Compartmental models split the population into compartments, each representing a different state of the disease.
-The \emph{dynamics} of the epidemic, how it changes over time, are described by the transitions between compartments.
+The \emph{dynamics} of the epidemic, how it changes over time, are represented by the transitions between compartments.
 Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
 Unlike renewal equations, there are not widely-used implementations of compartmental models.
-Most of the implementations used by SPI-M-O members were in the SEIR (susceptible, exposed, infectious, recovered) framework.
-The SEIR framework, has four components (susceptible, exposed, infectious, recovered), with the population divided into these components.
-The SEIR framework is described in detail in \cref{E-SEIR:sec:mechanistic-models}.
+A commonly used frameork is the SEIR (susceptible, exposed, infectious, recovered) model.
+The SEIR model, has four components (susceptible, exposed, infectious, recovered), with the population divided into these components; it is described in detail in \cref{E-SEIR:sec:mechanistic-models}.
+
 Compartmental models can be fit to almost any data source, and commonly incorporate multiple data sources, known as \emph{data synthesis}, a technique reviewed by \textcite{birrellEvidence}.
-Most compartmental models used by SPI-M-O were age-stratified, as differences between age groups are important for both the transmission and severity of SARS-CoV-2 (see \cref{E-SEIR:sec:structured-populations}).
 
@@
 The academic groups contributing estimates used a variety of models, reviewed by \textcite{royalSocietyRnumber}; see appendix 2 of that review for details of the specific implementations.
+Most compartmental models used by SPI-M-O were age-stratified, as differences between age groups are important for both the transmission and severity of SARS-CoV-2 (see \cref{E-SEIR:sec:structured-populations}).
 
@@
 Two approaches are commonly used to estimate the duration of a disease: either modelling an underlying biomarker or modelling the duration directly~\autocite{sweetingEstimating}.
-For SARS-CoV-2, both approaches would be based on the results of RT-PCR testing (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}).
+For SARS-CoV-2, both approaches would be based on the results of RT-PCR testing (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}\todo{move to where RT-PCR / PCR is first mentioned}).
 
-Modelling a biomarker, as I do in \cref{E-ATACCC}, models how the amount of virus in an individual varies over time.
+The biomarker approach, which I take in \cref{E-ATACCC}, models how the amount of virus in an individual varies over time.
 The RT-PCR tests commonly used give a proxy for viral load for positive tests (see \cref{E-biology-data:sec:PCR-process}).
@@
 
-An early meta-analysis estimated that individuals test positive for SARS-CoV-2 for 14.6 days (95\% CI: 9.3--20.0 days)~\autocite{cevikShedding}.
+An early meta-analysis estimated that individuals test positive for SARS-CoV-2 for a mean of 14.6 days (95\% CI: 9.3--20.0 days)~\autocite{cevikShedding}.
 The maximum observed duration was 83 days, although beyond day 9 no live virus was detected, indicating that the individuals testing positive for longer than this were not infectious.
@@
 For the purposes of this thesis, estimates applicable to the general population are required.
+There are no reliable estimates of the full distribution of the duration of positivity in a general population cohort available in the literature.
 Therefore, I estimate the duration of positivity from the ATACCC and CIS studies, as described above.
@@
 All inference in this thesis is Bayesian.
-Bayesian inference has several relevant advantages.
-Prior information allows identifiability in situations where this would not otherwise be possible, allows for combining knowledge from multiple sources, and many previous studies I build used Bayesian inference.
+Including prior information allows identifiability in situations where this would not otherwise be possible and allows for combining knowledge from multiple sources.
+Many previous studies I build on used Bayesian inference.
 
@@
 This thesis will focus on the second component.
-However, natural extensions of the work I undertake, will inform the first component.
+However, natural extensions of the work I undertake would inform the first component.
 
@@
 This is based on the CIS data.
-While statistical frameworks have been previously developed which included this case, they had only been applied in special cases.
+While statistical frameworks have been previously developed which included this case, they had not been used in practice.
 In \cref{E-imperf-test}, I extend this framework to include false negatives.
@@
 
+\todo[inline]{Check what methodological claims I want to make about ch 7}
 Finally, in \cref{E-SEIR}, I estimate the incidence of SARS-CoV-2 in England from only the CIS data.
@@
 In \cref{E-biology-data}, I provide background information on the natural history of SARS-CoV-2, the RT-PCR testing process used to produce the data used throughout this thesis, and describe the two cohort studies used in this thesis.
-Then, \cref{E-inc-prev}, I formally describe the stochastic processes which generate infections, prevalence, and the observations of prevalence.
+Then, in \cref{E-inc-prev}, I formally describe the stochastic processes which generate infections, prevalence, and the observations of prevalence.
 A key insight in this chapter is that the distribution of the duration of positivity is the key quantity linking incidence and prevalence.
@@
 Therefore, I turn to the CIS data in \cref{E-perf-test,E-imperf-test}.
-In \cref{E-perf-test} I provide a framework for estimating the duration of positivity from the CIS data, assuming perfect tests.
-However, the assumption of perfect tests is not realistic; in \cref{E-imperf-test} I extend the framework to include false negatives.
+In \cref{E-perf-test}, I provide a framework for estimating the duration of positivity from the CIS data, assuming perfect tests.
+However, the assumption of perfect tests is not realistic; in \cref{E-imperf-test}, I extend the framework to include false negatives.
 

### 2024-03-07 introduction.tex
Overleaf comments

--- 2024-03-07 introduction.tex
@@
 
-The coronavirus disease 2019 (COVID-19) caused by the Ssevere acute respiratory syndrome coronavirus 2 (SARS-CoV-2), is estimated to have killed approximately 15 million in the acutest phase (to the end of 2021) and 30 million globally~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
+COVID-19 (coronavirus disease 2019), the disease caused by the virus SARS-CoV-2 (severe acute respiratory syndrome coronavirus 2), is estimated to have killed approximately 15 million in the acutest phase (to the end of 2021) and 30 million globally~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
 While it is tempting to consider this a rare or extreme event, history suggests otherwise (see \cref{intro:fig:pandemic-timeline}).
@@
 To translate incidence into the burden of the disease, a measure of severity is required.
-For example, the infection fatality rate (IFR), the proportion of infected individuals who die.
+For example, the IFR (infection fatality rate), the proportion of infected individuals who die.
 Estimation of severity is beyond the scope of this thesis.
@@
 Therefore, fluctuations in the number of positive tests may not reflect changes in the true incidence of the disease.
-For large parts of the pandemic, community testing was widely available in the UK meaning resulting in a rich dataset, allowing detailed analyses to be performed, including stratification by small geographic areas or age groups.
+For large parts of the pandemic, community testing was widely available in the UK meaning that this is a large dataset, allowing detailed analyses to be performed, including stratification by small geographic areas or age groups.
 
@@
 The use of prevalence surveys for estimating incidence and transmission is the focus of this thesis.
-In particular, I focus on the Coronavirus (COVID-19) Infection Survey (CIS ), run by the Office for National Statistics (ONS)~\autocite{CIS,cisMethodsONS}.
+In particular, I focus on the CIS (Coronavirus (COVID-19) Infection Survey), run by the Office for National Statistics (ONS)~\autocite{CIS,cisMethodsONS}.
 The CIS was a longitudinal study, testing a large number of indidividuals weekly or monthly for SARS-CoV-2; it is described in detail in \cref{E-intro:sec:cis}.

### 2024-03-07 introduction.tex

--- 2024-03-07 introduction.tex
@@
 With constant behaviour and a disease that grants immunity following infection, $\Re$ declines smoothly.
-$\Re$ is useful because it describes the current dynamics of the epidemic: $\Re > 1$ means the epidemic is growing, $\Re < 1$ means it is shrinking, and $\Re = 1$ means it is stable.
+$\Re$ describes the current dynamics of the epidemic: $\Re > 1$ means the epidemic is growing, $\Re < 1$ means it is shrinking, and $\Re = 1$ means it is stable.
 In simple epidemic models (see \cref{E-SEIR:sec:SIR}), $\Re$ is proportional to the product of the rate of contacts, the riskiness of those contacts, and the proportion of the population that is susceptible.
@@
 
-Seroprevalence testing is a marker of recent prior infection.
+Seroprevalence testing provides information on a marker of recent prior infection, by measuring ..????
 Prior infection is more common than current infection, hence seroprevalence estimation requires a smaller sample size than current prevalence.
@@
 Severe events are generally a reliable measure of the burden of the disease.
-The criteria for being a severe event are normally well-defined and well-recorded in the UK context.
+The criteria for being a severe event are normally well-defined and generally well-recorded in the UK context (reference?).
 Therefore, the data is subject to fewer biases than community testing.
-However, SARS-CoV-2 rarely leads to severe events in younger individuals, meaning severe events only capture transmission to a subset of the population.
+However, SARS-CoV-2 rarely leads to severe events in younger individuals, meaning severe events only capture sequelae of transmission in a subset of the population.
 
@@
 Furthermore, prevalence surveys are noisy because they are based on population sampling.
-This observation noise needs to be considered in the application of backcalculation (see \cref{E-inc-prev:sec:observation-process}).
+This observation noise needs to be considered in the development of the backcalculation approach (see \cref{E-inc-prev:sec:observation-process}).
 

### 2024-03-08 transmission.tex
to do list

@@
 Overall, there is a bias-variance trade-off between the models.
-Fitting to just prevalence data has not previously been done in the literature (see\todo{ref lit review in intro and copy its novelty claims}).
+Incidence and transmission have not previously been estimated from only SARS-CoV-2 prevalence surveys before.
 
@@
 In this section, I apply the general mechanistic transmission modelling framework from \cref{E-SEIR} to the CIS prevalence data.
-The application of this framework follows \textcite{birrellRealtime}, with the most substantial context-specific work\todo{maybe change ``work" to something else: I want a succinct way to say that this was the major part that needed changing from the generic framework in chapter 3} in parameterising the WAIFW matrix (\cref{SEIR:sec:WAIFW-application}).
+The application of this framework follows \textcite{birrellRealtime}.
+The largest modification from the generic framework presented in \cref{E-SEIR} is the parameterisation of the WAIFW matrix (\cref{SEIR:sec:WAIFW-application}).
 A novel approach (\cref{SEIR:sec:observation}) links this transmission model to the CIS prevalence data; the approach is generalisable to other studies collecting representative prevalence data, \eg REACT.
@@
     \label{SEIR:fig:true-vs-posterior}
+    \todo[inline]{Fix x-axis labels overlapping / being unreadable}
 \end{figure}

### 2024-03-08 introduction.tex
explain ch 7 contribution

@@
 
-\todo[inline]{Check what methodological claims I want to make about ch 7}
-Finally, in \cref{E-SEIR}, I estimate the incidence of SARS-CoV-2 in England from only the CIS data.
+Finally, in \cref{E-SEIR}, I estimate the transmission of SARS-CoV-2 in England using a mechanistic model, from only the CIS data.
 This includes the development of a novel observation model to link the mechanistic transmission model to the data.
-The observation model allows the inclusion of the long tail in the duration of positivity, without being computationally infeasible.
+The observation model allows the inclusion of the skew in the duration of positivity, without being computationally infeasible.
 

### 2024-03-08 MCMC-appendix.tex
proof

@@
 This appendix outlines the basic theory and implementation of Markov chain Monte Carlo (MCMC) methods.
-Any reader interested in more technical details or proofs of the results here should consult the references within4.
+Any reader interested in more technical details or proofs of the results here should consult the references within.
 
@@
 The algorithm requires a proposal distribution $q(\psi' \mid \psi)$, which proposes that the chain moves to a new state $\psi'$ from its current state $\psi$; $\pi(\psi)$, a calculation of $p(\psi \mid y)$ up to some constant.
-The proposal is then either accepted or rejected based on the \emph{acceptance ratio}, which is computed using $\pi$.
+At each iteration, a proposal is draw from $q$ and then accepted or rejected based on the \emph{acceptance ratio}, computed using $\pi$.
 The full algorithm is given in \cref{inc-prev:algorithm:MH}, with further details in available in a wide-range of sources~\autocites[e.g.][]{brooksMCMCNotes}[chapter 11]{gelmanBDA}.

### 2024-03-08 introduction.tex
proof

@@
 \end{figure}
+\todo{Investigate why there's an extra linebreak here}
 
@@
 In all cases, surveillance is required to inform public health policy.
+In this thesis, I consider methods to improve surveillance, using the SARS-SARS-CoV-2 epidemic in England as a case study.
 
@@
 
-Various forms of \emph{reproduction number} were prominent throughout the COVID-19 pandemic~\autocite{pellisEstimation}.
+Various forms of \emph{reproduction number} were prominent throughout the SARS-CoV-2 pandemic~\autocite{pellisEstimation}.
 Broadly, a reproduction number describes the number of secondary infections generated by a single infection.
@@
 The \emph{effective reproduction number}, $\Re$, is the mean number of secondary infections generated by a single infection at time $t$.
-In particular, it uses the behaviour and immunity status at time $t$.
+In particular, it uses the behaviour and immunity status of the population at time $t$.
 With constant behaviour and a disease that grants immunity following infection, $\Re$ declines smoothly.
@@
 \emph{Prevalence} is the proportion of the population with the disease at a given time.
-Prevalence can be observed, most commonly with a prevalence survey.
+Whether an individual is prevalent (has the disease) is normally measurable.
+Therefore, the prevalence of the population can be estimated with a prevalence survey.
 
@@
 A wide variety of designed studies, administrative data, and other data sources have been used for pandemic and epidemic surveillance.
-In the context of the SARS-CoV-2 pandemic in England, the most prominent were: community testing (pillar 2), the number of severe events (\eg hospital admissions or deaths), seroprevalence testing (seroprevalence is the proportion of people with antibodies to a disease, a marker of prior infection), and prevalence surveys.
+In the context of the SARS-CoV-2 pandemic in England, the most prominent were: community testing (known as pillar 2 testing), the number of severe events (\eg hospital admissions or deaths), seroprevalence testing (seroprevalence is the proportion of people with antibodies to a disease, a marker of prior infection), and prevalence surveys.
 Each of these data sources have strengths and weaknesses, see \textcite{royalSocietyRnumber} for further details.
@@
 For large parts of the pandemic, community testing was widely available in the UK.
-Therefore, this is a large dataset, allowing detailed analyses to be performed, including stratification by small geographic areas or age groups.
+Therefore, there are many test results, allowing detailed analyses to be performed, including stratification by small geographic areas or age groups.
 
-Seroprevalence testing provides information on a marker of recent prior infection, by measuring ..????
+Seroprevalence testing provides information on a marker of recent prior infection.
 Prior infection is more common than current infection, hence seroprevalence estimation requires a smaller sample size than current prevalence.
@@
 
-Severe events are generally a reliable measure of the burden of the disease.
-The criteria for being a severe event are normally well-defined and generally well-recorded in the UK context (reference?).
+Severe events counts are generally a reliable measure of the burden of the disease.
+The criteria for being a severe event are normally well-defined and generally well-recorded in the UK context\todo{ref severe events criteria}.
 Therefore, the data is subject to fewer biases than community testing.
-However, SARS-CoV-2 rarely leads to severe events in younger individuals, meaning severe events only capture sequelae of transmission in a subset of the population.
+However, SARS-CoV-2 rarely leads to severe events in younger individuals, meaning severe event counts are insensitive to changes in transmission only occurring in younger individuals.
 
 Prevalence surveys in representative samples of the population are the most reliable measure of the burden of the disease.
-They are not subject to the biases of community testing and observe the whole population, unlike severe events.
+They are not subject to the biases of community testing and observe the whole population, unlike severe events, although they can suffer from non-response bias.
 However, they are expensive to power to inform detailed analyses.
-Often, some modelling is required to borrow information across strata or time to provide stable estimates (see \cref{E-intro:sec:cis}).
+Often, some modelling is required to borrow information across strata or time to provide stable estimates, as well as to correct for non-response bias (see \cref{E-intro:sec:cis}).
 
 The use of prevalence surveys for estimating incidence and transmission is the focus of this thesis.
-In particular, I focus on the CIS (Coronavirus (COVID-19) Infection Survey), run by the Office for National Statistics (ONS)~\autocite{CIS,cisMethodsONS}.
+In particular, I focus on the CIS (Coronavirus (COVID-19) Infection Survey), run by the ONS (Office for National Statistics)~\autocite{CIS,cisMethodsONS}.
 The CIS was a longitudinal study, testing a large number of indidividuals weekly or monthly for SARS-CoV-2; it is described in detail in \cref{E-intro:sec:cis}.
@@
 REACT was a series of cross-sectional surveys, run across 19 rounds between May 2020 and March 2022.
-Each round consisted of at least 100,000 randomly-sampled individuals from the UK contributing swabs for RT-PCR testing and questionnaire data~\autocite{elliottTwin}.
+Each round consisted of at least 100,000 randomly-sampled individuals from the UK contributing swabs for testing and questionnaire data~\autocite{elliottTwin}.
 
@@
 As noted above, incidence and transmission are extremely important quantities for informing the response to a pandemic but cannot be observed directly.
-Therefore, they must be estimated.
+Therefore, they must be inferred from observable quantities.
 
@@
 This ongoing pandemic has killed 40 million people, peaking at 2 million annual deaths~\autocite{unaids2023}.
-The large burden of the disease has motivated many advances in infectious disease epidemiology and biostatistics, including several which form the methodological basis of this thesis.
+The large burden of the disease has motivated many advances in infectious disease epidemiology and biostatistics, including several which form the methodological background of this thesis.
 
@@
 (see \cref{E-inc-prev:sec:infection-process}).
-
-However, Unlike HIV/AIDS, recovery from SARS-CoV-2 infection is possible, so recoveries need to be considered in the backcalculation.
+However, unlike with HIV/AIDS, recovery from SARS-CoV-2 infection is possible, so recoveries need to be considered in the backcalculation.
 This introduces a dependency structure in the prevalence data (see \cref{E-inc-prev:sec:prevalence-process}).
@@
 Furthermore, prevalence surveys are noisy because they are based on population sampling.
-This observation noise needs to be considered in the development of the backcalculation approach (see \cref{E-inc-prev:sec:observation-process}).
-
 The original formulations of backcalculation neglected any observation noise.
@@
 Since its development, backcalculation has been further developed for a wide-variety of contexts.
+I incorporate observation noise into the backcalculation approach in \cref{inc-prev:sec:observation-process}.
 % Reporting delay is not relevant to the studies in this thesis, and I ignore it in this and further discussion.
@@
 
-\Textcite{kermackContribution} was a seminal paper for the field, explaining the observation that epidemics ended before all individuals were infected due to a population-level build-up of immunity known as \emph{herd immunity}.
+\Textcite{kermackContribution} was a seminal paper for the field, explaining the observation that epidemics ended before all individuals were infected.
+They developed a model showing that a population-level build-up of immunity known as \emph{herd immunity} is sufficient to explain this observation.
 Their model has inspired much of the subsequent literature, a full review of which is beyond the scope of this thesis.
@@
 The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
-Renewal equations do not feature in this thesis, for further information, including practical considerations, see \textcite{nashEstimating,gosticPractical}.
+Renewal equations do not feature in this thesis, for further information, including practical considerations when using them as the basis of a statistical model, see \textcite{nashEstimating,gosticPractical}.
 
@@
 
-Each week, the members of SPI-M would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
+Each week, the members of SPI-M-O would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
 The nowcasting estimates were based on mechanistic models.
-These estimates, once any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value~\autocite{parkCombining}.
+These estimates, after removing any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value~\autocite{parkCombining}.
 The consensus value was published on \url{gov.uk} and informed the government's response to the epidemic~\autocite{govRnumber}.
@@
 In this thesis, I focus on the use of data from prevalence surveys.
+Specifically, surveys which collect swabs from individuals for testing by RT-PCR (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}).
 Several compartmental models incorporated data from SARS-CoV-2 prevalence surveys, alongside other data sources~\autocite{daviesAssociation,ironsEstimating,knockKey,nicholsonImproving,pooleyEstimation,birrellRTM2}.
-These models incorporate prevalence survey data by adding additional compartments representing RT-PCR individuals, who may not be infectious (this is common, see \cref{E-biology-data}); except \textcite{nicholsonImproving} who have a more sophisticated approach (below) and \textcite{ironsEstimating}, who assume RT-PCR positivity coincides with infectiousness.
+These models incorporate prevalence survey data by adding additional compartments representing RT-PCR positive individuals, who may not be infectious (this is common, see \cref{E-biology-data}); except \textcite{nicholsonImproving} who have a more sophisticated approach (below) and \textcite{ironsEstimating}, who assume RT-PCR positivity coincides with infectiousness.
 The length of stay in these compartments (the duration of time positive) is assumed to follow a gamma distribution, with the parameters estimated from external data or having at most one free parameter.
@@
 
+\todo[inline]{Maybe the following paragraph should be in the contribution section?}
 In \cref{E-SEIR}, I describe a compartmental model for SARS-CoV-2 transmission, which I fit to CIS data.
-I allow a more flexible distribution of the duration of positivity than most of the papers discussed above, incorporating the skew of the distribution.
+I allow a more flexible distribution of the duration of positivity than any of the papers discussed above, except \textcite{nicholsonImproving}, incorporating the skew of the distribution.
 This distribution is based on data from a representative sample of the UK population, rather than convenience samples (see \cref{intro:sec:duration-SARS-CoV-2}).
@@
 \Textcite{mccabeCISincidence}, adapting the methodology \textcite{ealesAppropriately} developed for REACT, estimate $\Re$ and the growth rate from the CIS data.
-They estimate a posterior distribution of the parameters of a spline, penalising deviations from a constant first derivative.
+They estimate a posterior distribution of the parameters of a spline by fitting to the prevalence results via a binomial distribution.
+Their setup penalises the spline deviating from a constant growth rate.
 The growth rate is then calculated from the derivative of the spline, an analytical transformation of the parameters.
@@
 Rather than using a Gaussian process, as \textcite{abbottCISincidence} do, \textcite{colmanAscertainment} provide stability by assuming the ascertainment rate is assumed to be piecewise constant.
-They do not fully propagate the uncertainty from the CIS prevalence estimates, considering only the bounds of the published 95\% confidence intervals.
+They do not fully propagate the uncertainty from the CIS prevalence estimates, considering only the bounds of the 95\% confidence intervals the ONS publishes for prevalence.
 
@@
 Two approaches are commonly used to estimate the duration of a disease: either modelling an underlying biomarker or modelling the duration directly~\autocite{sweetingEstimating}.
-For SARS-CoV-2, both approaches would be based on the results of RT-PCR testing (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}\todo{move to where RT-PCR / PCR is first mentioned}).
+For SARS-CoV-2, both approaches would be based on the results of RT-PCR testing.
 
@@
 It is also easier to specify non-parameteric forms of the distribution, which is important for SARS-CoV-2 as the tail behaviour may be important.
-A more detailed review of the relevant literature is in the relevant chapters.
+A more detailed review of the literature is in the relevant chapters.
 % Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
@@
 
-\subsubsection{Available estimates}
+\subsubsection{Previous estimates}
 
 An early meta-analysis estimated that individuals test positive for SARS-CoV-2 for a mean of 14.6 days (95\% CI: 9.3--20.0 days)~\autocite{cevikShedding}.
-The maximum observed duration was 83 days, although beyond day 9 no live virus was detected, indicating that the individuals testing positive for longer than this were not infectious.
-However, the majority of the studies included in the meta-analysis involved hospitalised patients or had unclear inclusion criteria, and hence may not be representative of the general population.
+The longest reported duration was observed at 83 days, although beyond day 9 no live virus was detected, indicating that the individuals testing positive for longer than this were not infectious.
+However, most studies included in the meta-analysis involved hospitalised patients or had unclear inclusion criteria, and hence may not be representative of the general population.
 These summary estimates are insufficient for backcalculation, which requires a distribution of the duration of positivity.
@@
 This could be due to the higher viral load leading to a longer duration of positivity, or that individuals with a higher viral load were more likely to have been recently infected at the time of the initial positive test.
-As they do not consider the latent true time of infection, they cannot distinguish between these two possibilities and underestimate the duration of positivity.
+They cannot estimate how long before the initial positive test the individual was infected; therefore, they cannot distinguish between these two possibilities.
+This also means they underestimate the duration of positivity.
 
@@
 
-Convergence is how quickly the Markov chain approaches the stationary distribution.
+Convergence is how close a MCMC chain's samples are to its stationary distribution.
 Early draws from the Markov chain reflect the choice of the starting point, $\psi^0$, rather than the stationary distribution that the chain will eventually converge to~\autocite[282]{gelmanBDA}.
@@
 
-First, the framework to simulation the CIS (used in \cref{E-imperf-test:sec:simulate,E-perf-test:sec:simulation-study}).
+First, the framework to simulate the CIS (used in \cref{E-imperf-test:sec:simulate,E-perf-test:sec:simulation-study}).
 The package enables customisation of various features of the study, for example the testing regime of the participants, the sensitivity of the tests, and the duration of positivity.

### 2024-03-08 introduction.tex
Updates from Overleaf

--- 2024-03-07 introduction.tex
@@
 \end{figure}
-
 Fully quantifying the burden of historical pandemics is challenging due to the absence of systematic records.
@@
 
-Compartmental models can be fit to almost any data source, and commonly incorporate multiple data sources, known as \emph{data synthesis}, a technique reviewed by \textcite{birrellEvidence}.
+Compartmental models can be fitted to almost any data source, and commonly incorporate multiple data sources, known as \emph{data synthesis}, a technique reviewed by \textcite{birrellEvidence}.
 
 \subsubsection{Application to SARS-CoV-2}
+{\color{red} you are missing a big opportunity here.
+You could elaborate a bit more the SPIMO part to explain what kind of data people used and introduce the question of whether the same evidence could be provided just by using data from a CIS type model.
+This is what motivates your research.
+Other people have used (only?) prevalence data (you give a lot of references below)..but have they recosntructed incidence and transmission only using prevalence data?}
 
-Many mechanistic modelling studies were conducted during the pandemic, with at least 100 articles published in the first six months of 2020 alone~\autocite{shankarSystematic}.
-In the UK, SPI-M-O (Scientific Pandemic Influenza Group on Modelling, Operational sub-group) was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), generating evidence based on modelling for policy development~\autocite{medleySPIM,govSPIMO}.
-For example, the stages of lifting the lockdown in 2021 each had a set of modelling studies informing the government's decision~\autocite{sageEvidence}.
+Many studies involving mechanistic models were conducted during the pandemic, with at least 100 articles published in the first six months of 2020 alone~\autocite{shankarSystematic}.
+In the UK, the Scientific Pandemic Influenza Group on Modelling, Operational sub-group (SPI-M-O) was a sub-committee of the Scientific Advisory Group for Emergencies (SAGE), using this type of models to generate evidence for policy development~\autocite{medleySPIM,govSPIMO}.
+For example, each of the stages of lifting the lockdown in 2021 was supported by  modelling studies informing the government's decision~\autocite{sageEvidence}.
 
 Each week, the members of SPI-M would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
-The nowcasting estimates were based on mechanistic models.
+The nowcasting estimates were mainly based on mechanistic models.
 These estimates, once any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value~\autocite{parkCombining}.
 The consensus value was published on \url{gov.uk} and informed the government's response to the epidemic~\autocite{govRnumber}.
-The academic groups contributing estimates used a variety of models, reviewed by \textcite{royalSocietyRnumber}; see appendix 2 of that review for details of the specific implementations.
+The academic groups contributing estimates used a variety of models, reviewed by \textcite{royalSocietyRnumber} (see appendix 2 of that review for details of the specific implementations).
 Most compartmental models used by SPI-M-O were age-stratified, as differences between age groups are important for both the transmission and severity of SARS-CoV-2 (see \cref{E-SEIR:sec:structured-populations}).

### 2024-03-08 introduction.tex

--- 2024-03-08 introduction.tex
@@
 \end{figure}
-\todo{Investigate why there's an extra linebreak here}
-
 Fully quantifying the burden of historical pandemics is challenging due to the absence of systematic records.
@@
 
-Compartmental models can be fit to almost any data source, and commonly incorporate multiple data sources, known as \emph{data synthesis}, a technique reviewed by \textcite{birrellEvidence}.
+Compartmental models can be fitted to almost any data source, and commonly incorporate multiple data sources, known as \emph{data synthesis}, a technique reviewed by \textcite{birrellEvidence}.
 
 \subsubsection{Application to SARS-CoV-2}
+{\color{red} you are missing a big opportunity here.
+You could elaborate a bit more the SPIMO part to explain what kind of data people used and introduce the question of whether the same evidence could be provided just by using data from a CIS type model.
+This is what motivates your research.
+Other people have used (only?) prevalence data (you give a lot of references below)..but have they recosntructed incidence and transmission only using prevalence data?}
 
-Many mechanistic modelling studies were conducted during the pandemic, with at least 100 articles published in the first six months of 2020 alone~\autocite{shankarSystematic}.
-In the UK, SPI-M-O (Scientific Pandemic Influenza Group on Modelling, Operational sub-group) was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), generating evidence based on modelling for policy development~\autocite{medleySPIM,govSPIMO}.
-For example, the stages of lifting the lockdown in 2021 each had a set of modelling studies informing the government's decision~\autocite{sageEvidence}.
+Many studies involving mechanistic models were conducted during the pandemic, with at least 100 articles published in the first six months of 2020 alone~\autocite{shankarSystematic}.
+In the UK, the Scientific Pandemic Influenza Group on Modelling, Operational sub-group (SPI-M-O) was a sub-committee of the Scientific Advisory Group for Emergencies (SAGE), using this type of models to generate evidence for policy development~\autocite{medleySPIM,govSPIMO}.
+For example, each of the stages of lifting the lockdown in 2021 was supported by  modelling studies informing the government's decision~\autocite{sageEvidence}.
 
-Each week, the members of SPI-M-O would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
-The nowcasting estimates were based on mechanistic models.
-These estimates, after removing any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value~\autocite{parkCombining}.
+Each week, the members of SPI-M would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
+The nowcasting estimates were mainly based on mechanistic models.
+These estimates, once any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value~\autocite{parkCombining}.
 The consensus value was published on \url{gov.uk} and informed the government's response to the epidemic~\autocite{govRnumber}.
-The academic groups contributing estimates used a variety of models, reviewed by \textcite{royalSocietyRnumber}; see appendix 2 of that review for details of the specific implementations.
+The academic groups contributing estimates used a variety of models, reviewed by \textcite{royalSocietyRnumber} (see appendix 2 of that review for details of the specific implementations).
 Most compartmental models used by SPI-M-O were age-stratified, as differences between age groups are important for both the transmission and severity of SARS-CoV-2 (see \cref{E-SEIR:sec:structured-populations}).

### 2024-03-08 introduction.tex

@@
 Many studies involving mechanistic models were conducted during the pandemic, with at least 100 articles published in the first six months of 2020 alone~\autocite{shankarSystematic}.
-In the UK, the Scientific Pandemic Influenza Group on Modelling, Operational sub-group (SPI-M-O) was a sub-committee of the Scientific Advisory Group for Emergencies (SAGE), using this type of models to generate evidence for policy development~\autocite{medleySPIM,govSPIMO}.
+In the UK, SPI-M-O (Scientific Pandemic Influenza Group on Modelling, Operational sub-group) was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), using this type of models to generate evidence for policy development~\autocite{medleySPIM,govSPIMO}.
 For example, each of the stages of lifting the lockdown in 2021 was supported by  modelling studies informing the government's decision~\autocite{sageEvidence}.

### 2024-03-08 biology-data.tex
Overleaf comments / two parts intro

--- 2024-03-07 biology-data.tex
@@
 A \emph{standard curve} translates the Ct value to viral load for a given test.
-They are produced by measuring the Ct value using reference samples with known viral load.
-As seen in \cref{biology-data:fig:ct-calibration}, they vary between studies.
+A standard curve is produced by measuring the Ct value using reference samples with known viral load, and, as seen in \cref{biology-data:fig:ct-calibration}, they vary between studies.
 \begin{figure}
@@
 An important concept throughout this thesis will be that of \emph{detectable} individuals.
-An individual is defined as detectable between the time they start testing positive on a RT-PCR test and the time they stop testing positive on a RT-PCR test.
-It is not possible to directly observe whether an individual is detectable or not due to noise in the RT-PCR test (see \cref{biology-data:sec:PCR}).
+An individual is defined as detectable between the time they would first test positive on a perfect RT-PCR test (if tested) and the time they would stop doing so.
+It is not possible to conclusively determine whether an individual is detectable because perfect RT-PCR tests do not exist (see \cref{biology-data:sec:PCR}).
 However, viewing the number of RNA genome copies in an individual as a latent process this can then be defined as the time when the number of RNA genome copies is above the limit of detection of the RT-PCR test (defined in \cref{biology-data:sec:measurement-error}).
@@
 If these were all false positives, then the test specificity would be 99.9\%.
-In reality, it is almost certain that some of these are true positives, and hence this is a lower-bound on the true test senstivity.
+In reality, it is almost certain that some of these are true positives, and hence this is a lower-bound on the true test specificity.
 
 RT-PCR enables the detection of SARS-CoV-2 viral RNA in a sample.
-However, this does not necessarily mean that the virus is viable (\ie able to reproduce)~\autocite{puhachSARSCoV2}.
+However, this does not necessarily mean that the virus is viable (\ie able to reproduce, a necessary condition for transmission to occur)~\autocite{puhachSARSCoV2}.
 High viral loads (low Ct values) are more likely to be viable, but more sophisticated tests are required to determine viability.

### 2024-03-08 introduction-comments.tex
Overleaf comments / two parts intro

--- 2024-03-08 introduction.tex
@@
 A fully susceptible population means none of its members have any immunity to the disease.
-Note that this definition is a constant for a given pathogen and population.
+Note that this definition implies $\R$ is constant for a given pathogen and population.
 In the early stages of an epidemic caused by a novel pathogen, the number of individuals with immunity is negligible, and hence $\R$ describes this part of the epidemic well.
@@
 These definitions ignore many complexities that arise when theoretically defining or practically estimating any of the reproduction numbers~\autocite{pellisEstimation}.
-For example, effects due to changing weather, what a ``mean infection'' means in a heterogeneous population, and the lack of a ground truth for these numbers in almost all practical situations.
+For example, effects due to seasonality, what a ``mean infection'' means in a heterogeneous population, and the lack of a ground truth for these numbers in almost all practical situations.
 However, their intuitive simplicity and convenient properties in simple epidemic models (see \cref{E-SEIR:sec:mechanistic-models}) mean that they are useful concepts for guiding policy.
@@
 Reproduction numbers are strongly connected to \emph{growth rates}.
-Over short periods of time, epidemics tend to grow exponentially, and hence have an exponential growth rate, normally denoted $r$.
+Over short periods of time, epidemics tend to grow (or shrink) exponentially, and hence have an exponential growth rate, normally denoted $r$.
 For a given disease and population, an analogous growth rate can be defined for each reproduction number~\autocite{pellisEstimation,paragGrowthRates,wallingaGI}.
@@
 
-However, incidence and transmission are almost never directly observed.
-Therefore, they need to be estimated from other data sources.
-
 \emph{Prevalence} is the proportion of the population with the disease at a given time.
-Whether an individual is prevalent (has the disease) is normally measurable.
+Whether an individual is \emph{prevalent} (has the disease) is normally measurable.
 Therefore, the prevalence of the population can be estimated with a prevalence survey.
-
-See \cref{E-inc-prev:sec:definitions} for a formal definition of incidence and prevalence.
-Prevalence and incidence are strongly related, with the relationship determined by how long the disease lasts (see \cref{E-inc-prev}).
+However, these are very expensive to power appropriately; SARS-CoV-2 is unusual in that its impact is so large that the UK government did this.
+\cref{E-inc-prev:sec:definitions} provides a formal definition of incidence and prevalence.
+Prevalence and incidence are strongly related, with the relationship determined by by the duration of an infection (see \cref{E-inc-prev}).
 
@@
 
-\section{Data sources} \label{intro:sec:data-sources}
+However, incidence and transmission and prevalence, are almost never directly observed.
+Therefore, they need to be estimated from other data sources.
+
+Introduce next section eg incidence / transmission not observed directly, need infer from data such as next section.
+
+\section{Data sources in SARS-CoV-2} \label{intro:sec:data-sources}
 
@@
 The criteria for being a severe event are normally well-defined and generally well-recorded in the UK context\todo{ref severe events criteria}.
-Therefore, the data is subject to fewer biases than community testing.
+Therefore, these data are subject to fewer biases than community testing.
 However, SARS-CoV-2 rarely leads to severe events in younger individuals, meaning severe event counts are insensitive to changes in transmission only occurring in younger individuals.
@@
 \Textcite{abbottCISincidence} applied backcalculation to CIS data to estimate the incidence of SARS-CoV-2 in the UK.
-They used a Gaussian process prior to smooth the incidence, applying the prior to either smooth the log-incidence or its derivative; smoothing of the incidence is required to stabalise the estimates~\autocite{brookmeyerBackcalculation}.
+They used a Gaussian process prior to smooth the incidence, applying the prior to either smooth the log-incidence or its derivative; smoothing of the incidence is required to stabilise the estimates~\autocite{brookmeyerBackcalculation}.
 \Textcite{abbottCISincidence} used \textcite{hellewellPCRSensitivity}'s estimates of the sensitivity of testing to inform the backcalculation, which are based on a longitudinal study in healthcare workers (see \cref{intro:sec:duration-SARS-CoV-2}).
@@
 
-Each week, the members of SPI-M would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
+Each week, the members of SPI-M-O would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
 The nowcasting estimates were mainly based on mechanistic models.
@@
 
-\subsection{Methodology}
+\subsection{Approaches}
 
@@
 
-The overall question that I investigate in this thesis is: how can prevalence surveys best be used to understand incidence and transmission?
+The overall question that I investigate in this thesis is: can prevalence surveys alone be used to infer incidence and transmission, and if so how?
 Prevalence surveys provide data that is far more representative of the general population than other data sources (see \cref{intro:sec:data-sources}), therefore, conclusions based on them should be far more robust.
@@
 First, in \cref{E-inc-prev:sec:observation-process}, I provide a theoretic justification for the use of deterministic backcalculation in this setting.
-A deterministic backcalculation approach was used by \textcite{abbottCISincidence} and I will use it in \cref{E-backcalc}.
-This approach models the prevalence at any time as a deterministic function of the incidence, rather than as a random variable.
+A \emph{deterministic} backcalculation approach was used by \textcite{abbottCISincidence} and I will use it in \cref{E-backcalc}.
+The deterministic approach models the prevalence at any time as a deterministic function of the incidence, rather than as a random variable.
 I show that, this is a good approximation as long as the proportion of the population sampled is small.

### 2024-03-08 transmission.tex
Overleaf comments / two parts intro

@@
 
-An alternative would be an inference procedure that estimates incidence proportion, fit directly to the RT-PCR test results.
+An alternative approach would be to estimate incidence proportion directly from the RT-PCR test results.
+
 Denote the quantity to be inferred (which could be either incidence proportion or prevalence) at time $t$ in strata $i$ as $\pi_{it}$.
-The strata will represent at least age group and region, although including ethnicity and sex is also due to non-response in the survey (see \cref{E-intro:sec:cis-strata}).
+The strata will represent at least age group and region, although including ethnicity and sex is also included due to non-response in the survey (see \cref{E-intro:sec:cis-strata}).
 For many combinations of stratum $i$ and day $t$, the number of positives will be small, often fewer than 5; therefore some form of borrowing between time and strata is required.
@@
 I take information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using the methodology of \textcite{vanleeuwenTime,vanleeuwenAugmenting}.
-%; these estimates have been used for previous modelling of the COVID-19 pandemic~\autocite{vanleeuwenTime,birrellRealtime}.
+%; these estimates have been used for previous modelling of the SARS-CoV-2 pandemic~\autocite{vanleeuwenTime,birrellRealtime}.
 % These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.

### 2024-03-09 introduction.tex
restructure intro

--- 2024-03-08 introduction.tex
@@
 
-\section{Data sources} \label{intro:sec:data-sources}
+However, incidence and transmission and prevalence, are almost never directly observed.
+Therefore, they need to be estimated from other data sources.
 
-A wide variety of designed studies, administrative data, and other data sources have been used for pandemic and epidemic surveillance.
-In the context of the SARS-CoV-2 pandemic in England, the most prominent were: community testing (known as pillar 2 testing), the number of severe events (\eg hospital admissions or deaths), seroprevalence testing (seroprevalence is the proportion of people with antibodies to a disease, a marker of prior infection), and prevalence surveys.
+Introduce next section eg incidence / transmission not observed directly, need infer from data such as next section.
+
+\section{Data sources}
+
+The data sources used for outbreak surveillance are often observational in nature.
+Therefore, they suffer from biases that can be hard to understand and quantify.
+For example, community testing which relies on individuals presenting for a test, hence biased by the willingness of individuals to be tested.
+
+Well-designed studies can provide direct estimates of incidence and/or prevalence in specific cohorts.
+However, recruiting a representative sample of the population is challenging, and the results of these studies are often not generalisable to the whole population.
+Furthermore, for these studies to be appropriately powered, they often require a large cohort, and hence are expensive to run.
+A unique feature of the COVID-19 pandemic is that its impact was so large that the UK Government were willing to fund such studies.
+The availability of this data motivates the work in this thesis.
+
+\section{Transmission model}
+
+The first component required to estimate incidence and transmission is a \emph{transmission model}, a mathematical description of how incidence arises.
+A spectrum of models exist, categorised by the strenght of assumptions they make about the transmission process.
+Stronger assumptions allow more insight into the transmission process and generally lower the variance of the estimates, but can introduce bias and are often computationally challenging to fit.
+
+The models with the weakest assumptions are statistical models.
+These make almost no assumptions about the transmission process, viewing incidence as a generic counting process.
+The most assumption is that incidence follows a Poisson process (defined in \cref{E-inc-prev:sec:poisson-process}), and the intensity of the process is the target of incidence.
+% The use of Poisson processes to model incidence of an infectious disease was very popular in the context of HIV/AIDS\todo{maybe a ref for Poisson in HIV/AIDS?}.
+
+Next, are semi-mechanistic models.
+These have components of the transmission process in them but do not fully represent the process.
+The most common type of semi-mechanistic model are those based on \emph{renewal equation}.
+The renewal equation~\autocite{fraserRenewal} for an infectious disease, in the most general form, is:
+\begin{align}
+    I(t) &= \int_{0}^{t} \beta(t,\tau) I(t-\tau) d \tau
+\end{align}
+where $I(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the instantaneous probability of an individual infected $\tau$ time units ago infects an individual at time $t$.
+Therefore, the renewal equation relates incidence in the past to incidence in the future.
+The effective reproduction number can be constructed as a function of $\beta(t, \tau)$.
+% A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
+% The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
+Renewal equations do not feature futher in this thesis, for details, including practical considerations when using them as the basis of a statistical model, see \textcite{nashEstimating,gosticPractical}.
+
+Finally, mechanistic models are the most complex and aim to fully represent the transmission process.
+\emph{Compartmental models} are the most common type of mechanistic model.
+Compartmental models split the population into compartments, each representing a different state of the disease.
+The \emph{dynamics} of the epidemic, how it changes over time, are represented by the transitions between compartments.
+Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
+Unlike renewal equations, there are not widely-used implementations of compartmental models.
+A commonly used frameork is the SEIR (susceptible, exposed, infectious, recovered) model.
+The SEIR model, has four components (susceptible, exposed, infectious, recovered), with the population divided into these components; it is described in detail in \cref{E-SEIR:sec:mechanistic-models}.
+
+\section{Observation models}
+
+The \emph{observation model} links the transmission model to the data.
+The choice of observation model can, in theory, be independent of the transmission model.
+However, pragmatic considerations often mean that the observation model is chosen to be in a similar form to the transmission model.
+
+Typically, the core of an observation model is a \emph{convolution equation} of an approximation of it.
+This equation links an observation at time $t$, $y_t$, to the incidence prior to time $t$.
+In discrete time, it is:
+\begin{align}
+    y_t &= \zeta \sum_{i=0}^\infty I(t-i) f(i)
+\end{align}
+where $I(t)$ is the incidence at time $t$; $\zeta$ is the proportion of infections that are ever observed; and $f(i)$ is the probability that an infection at time $t-i$ is observed at time $t$, conditional on it ever being observed.
+
+% The choice of an observation model must be heavily informed by the data that will be used.
+% Two types of data commonly arise in the context of infectious disease: incidence measures and prevalence measures.
+
+% An incidence measure is characterised by each infection only appearing in the dataset at most once.
+% Common examples are hospital admissions and deaths.
+% The basic form of an observation model for an incidence measure is a scaling and delay of the incidence.
+% The scaling is the proportion of infections that are observed (\eg the proportion of infections that are fatal), and the delay is the time between an infection occurring and it being observed (\eg the time between infection and death).
+% In discrete time this is:
+% \begin{align}
+%     M(t) = \zeta \sum_{\tau=0}^\infty I(t-tau) f(\tau)
+% \end{align}
+% where $M(t)$ is the observation of the incidence measure at time $t$; $I(t)$ is the incidence at time $t$; $\zeta$ is the scaling; and $f(\tau)$ is the delay, formalised as the probability that an infection is observed $\tau$ days after it occurs, given that it will be observed.
+
+% A prevalence measure is characterised by each infection possibly appearing in the dataset multiple times.
+% Common examples are the results of a prevalence survey or the number of individuals in hospital with a disease.
+% \todo[inline]{Finish writing about prevalence measures}
+
+Solving a convolution equation for $I$ is an ill-posed problem when $f(0)$ is small.
+Either an approach which gives a large $f(0)$, for example by assuming that the delay between incidence and being prevelant is 0, or smoothing on $I(t)$ is required.
+
+For a prevalence survey, the delay distribution is the \emph{duration of positivity}, the distribution of the time between an individual becoming infected and the time they are no longer infected.
+
+\section{Backcalculation}
+
+\emph{Backcalculation} was a modelling framework developed in the context of HIV/AIDS but since applied more broadly.
+The aim of backcalculation is to estimate the incidence of a disease from the observation of a prevalence measure.
+It combines a statistical transmission model with
+
+\section{Estimating the duration of positivity}
+
+Two approaches are commonly used to estimate the duration of a disease: either modelling an underlying biomarker or modelling the duration directly~\autocite[e.g.][]{sweetingEstimating}.
+Modelling a biomarker uses more information per observation, but requires stronger assumptions.
+
+The biomarker approach models the trajectory of a biomarker over time in a cohort of individuals.
+An individual is considered to have the disease while the biomarker is above or below some threshold.
+Parametric assumptions are commonly required for this approach.
+For example, a functional form of the behaviour over time of the biomarker is required.
+In addition, if individual-level variation wants to be included, which is generally important, a distributional form of random effects is required.
+This requires both sufficient observations per individual and an assumption on the form of the distribution.
+
+Modelling the duration directly considers only the binary result at each test (positive or negative); for instance in a \emph{survival analysis} framework.
+This approach generally requires fewer parameters, and no random effect(s), reducing the computational burden.
+It is also easier to specify non-parameteric forms of the distribution, which is important for SARS-CoV-2 as the tail behaviour may be important.
+% A more detailed review of the literature is in the relevant chapters.
+% Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
+% Other related approaches, such as a multi-state model (reviewed by \textcite{jacksonMSM}), could also be considered; multi-state models can be viewed as a generalisation of survival analysis.
+
+In both cases, longitudinal data is required.
+Therefore, cohort studies, where the same group of individuals are followed over time, are the most appropriate study design.
+
+\section{Application to SARS-CoV-2 in England}
+
+The application I will consider in this thesis is the SARS-CoV-2 epidemic in England.
+
+\subsection{Data sources}
+
+\subsubsection{For estimating incidence and transmission}
+
+The main data sources informing the response to the SARS-Cov-2 epidemic in England were: community testing (known as pillar 2 testing), the number of severe events (\eg hospital admissions or deaths), seroprevalence surveys (estimating the proportion of the population with antibodies to a disease, a marker of prior infection), and prevalence surveys (estimating the proportion of the population currently with the disease).
 Each of these data sources have strengths and weaknesses, see \textcite{royalSocietyRnumber} for further details.
 
-Community testing is broadly available, but does not test a random sample of the population.
-The number of tests conducted varies over time and space, for a variety of known and unknown reasons.
+For large parts of the pandemic, community testing was widely available in the UK.
+The number of tests conducted varied over time and space, for a variety of known and unknown reasons.
 These include: policy on who is eligible for testing, the availability of tests, and the willingness of individuals to be tested.
 Therefore, fluctuations in the number of positive tests may not reflect changes in the true incidence of the disease.
-For large parts of the pandemic, community testing was widely available in the UK.
-Therefore, there are many test results, allowing detailed analyses to be performed, including stratification by small geographic areas or age groups.
+However, there were many test results, allowing detailed analyses to be performed, including stratification by small geographic areas or age groups.
 
-Seroprevalence testing provides information on a marker of recent prior infection.
+Seroprevalence surveys indicate the proportion of the population that have been infected with the disease.
 Prior infection is more common than current infection, hence seroprevalence estimation requires a smaller sample size than current prevalence.
@@
 The use of prevalence surveys for estimating incidence and transmission is the focus of this thesis.
-In particular, I focus on the CIS (Coronavirus (COVID-19) Infection Survey), run by the ONS (Office for National Statistics)~\autocite{CIS,cisMethodsONS}.
+These surveys use RT-PCT (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
+I focus on the CIS (Coronavirus (COVID-19) Infection Survey), run by the ONS (Office for National Statistics)~\autocite{CIS,cisMethodsONS}.
 The CIS was a longitudinal study, testing a large number of indidividuals weekly or monthly for SARS-CoV-2; it is described in detail in \cref{E-intro:sec:cis}.
@@
 A particular instance of this is that a peak in late December 2020 occurred between rounds of the REACT survey, complicating analysis of the survey data at that time~\autocite{rileyREACTround8}.
-
-The primary aim of this thesis is to estimate incidence and transmission from the CIS data.
-
-\section{Estimating incidence and transmission} \label{intro:sec:estimating-incidence}
-
-As noted above, incidence and transmission are extremely important quantities for informing the response to a pandemic but cannot be observed directly.
-Therefore, they must be inferred from observable quantities.
-
-Many methods for estimating incidence and transmission were developed in the context of the HIV/AIDS pandemic.
-% HIV is transmitted sexually or through blood and causes the disease AIDS.
-This ongoing pandemic has killed 40 million people, peaking at 2 million annual deaths~\autocite{unaids2023}.
-The large burden of the disease has motivated many advances in infectious disease epidemiology and biostatistics, including several which form the methodological background of this thesis.
-
-\subsection{Backcalculation} \label{intro:sec:backcalc}
-
-\emph{Backcalculation} is a statistical procedure for estimating incidence based on a measurement of prevalence.
-It was proposed by \textcite{brookmeyerMethod} in the context of HIV/AIDS.
-In this context, backcalculation estimates the incidence of HIV infection from data on the number of AIDS diagnoses, and the time between infection and AIDS diagnosis~\autocite{brookmeyerBackcalculation,brookmeyerMeasuring}.
-Two processes needed to be considered: one generating infections and one relating infections to prevalence.
-% AIDS is incurable, therefore, AIDS prevalence is equivalent to cumulative incidence.
-
-The statistical framework developed for HIV is applicable to SARS-CoV-2.
-(see \cref{E-inc-prev:sec:infection-process}).
-However, unlike with HIV/AIDS, recovery from SARS-CoV-2 infection is possible, so recoveries need to be considered in the backcalculation.
-This introduces a dependency structure in the prevalence data (see \cref{E-inc-prev:sec:prevalence-process}).
-The development of this framework makes clear that the duration of positivity (how long an individual tests positive for SARS-CoV-2) is the key quantity relating prevalence to incidence.
-Therefore, estimation of duration is central to this thesis (see \cref{intro:sec:duration}).
-
-Furthermore, prevalence surveys are noisy because they are based on population sampling.
-The original formulations of backcalculation neglected any observation noise.
-AIDS is almost always diagnosed and recorded in developed countries.
-That is, the observations of AIDS diagnoses are the same as the true incidence of AIDS.
-An exception is that the reporting delay may be important~\autocite{paganoHIV}; but reporting delay is not important for this thesis's work.
-Since its development, backcalculation has been further developed for a wide-variety of contexts.
-I incorporate observation noise into the backcalculation approach in \cref{inc-prev:sec:observation-process}.
-% Reporting delay is not relevant to the studies in this thesis, and I ignore it in this and further discussion.
-
-\Textcite{abbottCISincidence} applied backcalculation to CIS data to estimate the incidence of SARS-CoV-2 in the UK.
-They used a Gaussian process prior to smooth the incidence, applying the prior to either smooth the log-incidence or its derivative; smoothing of the incidence is required to stabalise the estimates~\autocite{brookmeyerBackcalculation}.
-\Textcite{abbottCISincidence} used \textcite{hellewellPCRSensitivity}'s estimates of the sensitivity of testing to inform the backcalculation, which are based on a longitudinal study in healthcare workers (see \cref{intro:sec:duration-SARS-CoV-2}).
-They fit to summary statistics of the CIS data, using a daily, independent Gaussian likelihood with parameters estimated from the posterior point estimate and 95\% credible interval published by the ONS.
-
-The estimates I present in \cref{E-backcalc} are methodologically similar.
-However, I fully propagate the uncertainty from the prevalence estimation procedure and use estimates of the duration of positivity from a general population study.
-\Textcite{abbottCISincidence} include the short period of low but increasing test sensitivity at the start of an infection (see \cref{E-biology-data:sec:natural-history}).
-
-\subsection{Mechanistic models}
-
-\subsubsection{Background}
-
-Mechanistic models make assumptions on the biological and epidemiological mechanisms driving transmission~\autocite{lesslerMechanistic}.
-Therefore, the model's state and parameters have a biological and/or epidemiological interpretation.
-% Of particular interest is that the basic reproduction number, $\R$, and the effective reproduction number, $R_e(t)$, which can both be calculated as functions of the model parameters; using \cref{E-backcalc}'s approach, additional modelling would be required to estimate these numbers.
-In particular, this allows changes in transmission to be decomposed into the contributing factors, \eg the number of contacts per day or the probability of transmission upon contact (see \cref{E-SEIR:sec:mechanistic-models}).
-The ability to estimate the contribution of each of these factors can inform public health policy.
-Furthermore, mechanistic models can be used for scenario modelling, which simulates the effect of interventions.
-A scenario is the effect of a proposed intervention (or lack thereof), simulated by modifying the relevant parameters.
-% The mechanistic model can then be simulated forward in time to understand the intervention's effects.
-% All of these properties make mechanistic models a useful tool for understanding and controlling infectious diseases.
-% However, they make stronger assumptions about the disease and population behaviour than statistical models.
-
-The literature on mechanistic models is vast, having been developed for over a century.
-The model I introduce in \cref{E-SEIR:sec:SIR} was first formulated by \textcite{kermackContribution}.
-However, the earliest mechanistic model to be formulated mathematically was probably \textcite{rossMalariaA}'s model of malaria transmission~\autocite{lesslerMechanistic}.
-This early contribution led to an important insight for public health, showing that malaria could be controlled even if mosquitoes were not fully eliminated.
-
-\Textcite{kermackContribution} was a seminal paper for the field, explaining the observation that epidemics ended before all individuals were infected.
-They developed a model showing that a population-level build-up of immunity known as \emph{herd immunity} is sufficient to explain this observation.
-Their model has inspired much of the subsequent literature, a full review of which is beyond the scope of this thesis.
-The relevant background information is explained in \cref{E-SEIR:sec:mechanistic-models}.
-
-Two types of mechanistic models were prominent during the SARS-CoV-2 pandemic: those based on a \emph{renewal equation} and those with a \emph{compartmental} structure.
-
-The renewal equation~\autocite{fraserRenewal} for an infectious disease, in the most general form, is:
-\begin{align}
-    I(t) &= \int_{0}^{t} \beta(t,\tau) I(t-\tau) d \tau
-\end{align}
-where $I(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the instantaneous probability of an individual infected $\tau$ time units ago infects an individual at time $t$.
-A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
-The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
-Renewal equations do not feature in this thesis, for further information, including practical considerations when using them as the basis of a statistical model, see \textcite{nashEstimating,gosticPractical}.
-
-Compartmental models split the population into compartments, each representing a different state of the disease.
-The \emph{dynamics} of the epidemic, how it changes over time, are represented by the transitions between compartments.
-Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
-Unlike renewal equations, there are not widely-used implementations of compartmental models.
-A commonly used frameork is the SEIR (susceptible, exposed, infectious, recovered) model.
-The SEIR model, has four components (susceptible, exposed, infectious, recovered), with the population divided into these components; it is described in detail in \cref{E-SEIR:sec:mechanistic-models}.
-
-Compartmental models can be fitted to almost any data source, and commonly incorporate multiple data sources, known as \emph{data synthesis}, a technique reviewed by \textcite{birrellEvidence}.
-
-\subsubsection{Application to SARS-CoV-2}
-{\color{red} you are missing a big opportunity here.
-You could elaborate a bit more the SPIMO part to explain what kind of data people used and introduce the question of whether the same evidence could be provided just by using data from a CIS type model.
-This is what motivates your research.
-Other people have used (only?) prevalence data (you give a lot of references below)..but have they recosntructed incidence and transmission only using prevalence data?}
-
-Many studies involving mechanistic models were conducted during the pandemic, with at least 100 articles published in the first six months of 2020 alone~\autocite{shankarSystematic}.
-In the UK, SPI-M-O (Scientific Pandemic Influenza Group on Modelling, Operational sub-group) was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), using this type of models to generate evidence for policy development~\autocite{medleySPIM,govSPIMO}.
-For example, each of the stages of lifting the lockdown in 2021 was supported by  modelling studies informing the government's decision~\autocite{sageEvidence}.
-
-Each week, the members of SPI-M would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
-The nowcasting estimates were mainly based on mechanistic models.
-These estimates, once any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value~\autocite{parkCombining}.
-The consensus value was published on \url{gov.uk} and informed the government's response to the epidemic~\autocite{govRnumber}.
-The academic groups contributing estimates used a variety of models, reviewed by \textcite{royalSocietyRnumber} (see appendix 2 of that review for details of the specific implementations).
-Most compartmental models used by SPI-M-O were age-stratified, as differences between age groups are important for both the transmission and severity of SARS-CoV-2 (see \cref{E-SEIR:sec:structured-populations}).
-
-In this thesis, I focus on the use of data from prevalence surveys.
-Specifically, surveys which collect swabs from individuals for testing by RT-PCR (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}).
-Several compartmental models incorporated data from SARS-CoV-2 prevalence surveys, alongside other data sources~\autocite{daviesAssociation,ironsEstimating,knockKey,nicholsonImproving,pooleyEstimation,birrellRTM2}.
-These models incorporate prevalence survey data by adding additional compartments representing RT-PCR positive individuals, who may not be infectious (this is common, see \cref{E-biology-data}); except \textcite{nicholsonImproving} who have a more sophisticated approach (below) and \textcite{ironsEstimating}, who assume RT-PCR positivity coincides with infectiousness.
-The length of stay in these compartments (the duration of time positive) is assumed to follow a gamma distribution, with the parameters estimated from external data or having at most one free parameter.
-The proportion of the population in these compartments are then fit to the prevalence survey data using a binomial likelihood where possible, although some approaches take a Gaussian approximation.
-These approaches are similar to how these models are fit to other data sources, such as hospital admissions or deaths.
-
-\Textcite{nicholsonImproving} have the most sophisticated use of prevalence survey data.
-They allow an arbitrary distribution of the duration of positivity, based on the estimates of \textcite{hellewellPCRSensitivity} (see \cref{intro:sec:duration-SARS-CoV-2} for discussion of the \textcite{hellewellPCRSensitivity} study).
-Furthermore, they incorporate the sensitivity and specificity of the RT-PCR test into their model.
-% However, they do not incorporate age-stratification into their transmission model, and primarily aim to estimate the bias of community testing in the UK rather than incidence and/or transmission.
-
-\todo[inline]{Maybe the following paragraph should be in the contribution section?}
-In \cref{E-SEIR}, I describe a compartmental model for SARS-CoV-2 transmission, which I fit to CIS data.
-I allow a more flexible distribution of the duration of positivity than any of the papers discussed above, except \textcite{nicholsonImproving}, incorporating the skew of the distribution.
-This distribution is based on data from a representative sample of the UK population, rather than convenience samples (see \cref{intro:sec:duration-SARS-CoV-2}).
-Unlike \textcite{nicholsonImproving}, I include age-stratification in the transmission model, allowing estimation of how transmission varies by age.
-I use only data from the CIS, and show this is sufficient for estimating incidence and transmission.
-Finally, I fit directly to counts from the CIS, rather than an approximation of this data, which has not previously be done.
-
-\subsection{Other approaches}
-
-Other approaches have been taken to estimate incidence and/or transmission from the CIS data.
-
-\Textcite{mccabeCISincidence}, adapting the methodology \textcite{ealesAppropriately} developed for REACT, estimate $\Re$ and the growth rate from the CIS data.
-They estimate a posterior distribution of the parameters of a spline by fitting to the prevalence results via a binomial distribution.
-Their setup penalises the spline deviating from a constant growth rate.
-The growth rate is then calculated from the derivative of the spline, an analytical transformation of the parameters.
-Since, under reasonable model assumptions, there is a one-to-one correspondence between reproduction numbers and growth rates~\autocite{wallingaGI}, and they assume that the growth rate in incidence and prevalence is equal, this also estimates $\Re$.
-
-\Textcite{colmanAscertainment} used CIS prevalence estimates to estimate the \emph{ascertainment rate} of community testing (the proportion of all infections that community testing detected).
-Their methodology was similar to \textcite{abbottCISincidence} (see \cref{intro:sec:backcalc}), including using estimates from \textcite{hellewellPCRSensitivity} for how the sensitivity of RT-PCR testing varies over an infection.
-Rather than using a Gaussian process, as \textcite{abbottCISincidence} do, \textcite{colmanAscertainment} provide stability by assuming the ascertainment rate is assumed to be piecewise constant.
-They do not fully propagate the uncertainty from the CIS prevalence estimates, considering only the bounds of the 95\% confidence intervals the ONS publishes for prevalence.
-
-    \section{Estimating duration} \label{intro:sec:duration}
-
-The distribution of the duration of positivity is central to the relationship between incidence and prevalence (see \cref{E-inc-prev:sec:prevalence-process}).
-Therefore, a reliable estimate of this quantity is required to achieve this thesis's aims.
-
-\subsection{Methodology}
-
-Two approaches are commonly used to estimate the duration of a disease: either modelling an underlying biomarker or modelling the duration directly~\autocite{sweetingEstimating}.
-For SARS-CoV-2, both approaches would be based on the results of RT-PCR testing.
-
-The biomarker approach, which I take in \cref{E-ATACCC}, models how the amount of virus in an individual varies over time.
-The RT-PCR tests commonly used give a proxy for viral load for positive tests (see \cref{E-biology-data:sec:PCR-process}).
-Modelling a biomarker uses more information per observation, but requires stronger assumptions.
-For example, a functional form of the behaviour over time of the biomarker is required.
-In addition, if individual-level variation wants to be included, which is generally important, a distributional form of random effects is required.
-This requires both sufficient observations per individual and an assumption on the form of the distribution.
-
-Modelling the duration directly, as I do in \cref{E-perf-test,E-imperf-test} considers only the binary result at each test (positive or negative); in this thesis a \emph{survival analysis} framework is adopted.
-This approach generally requires fewer parameters, and no random effect(s), reducing the computational burden.
-It is also easier to specify non-parameteric forms of the distribution, which is important for SARS-CoV-2 as the tail behaviour may be important.
-A more detailed review of the literature is in the relevant chapters.
-% Survival analysis is the area of statistics concerned with estimating the distribution of the times between two events.
-Other related approaches, such as a multi-state model (reviewed by \textcite{jacksonMSM}), could also be considered; multi-state models can be viewed as a generalisation of survival analysis.
-
-In both cases, longitudinal data is required.
-Therefore, cohort studies, where the same group of indivdiuals are followed over time, are the most appropriate study design.
-
-\subsection{SARS-CoV-2} \label{intro:sec:duration-SARS-CoV-2}
 
@@
 
-\subsubsection{Previous estimates}
+\subsection{SPI-M-O}
+
+In the UK, SPI-M-O (Scientific Pandemic Influenza Group on Modelling, Operational sub-group) was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), using this type of models to generate evidence for policy development~\autocite{medleySPIM,govSPIMO}.
+For example, each of the stages of lifting the lockdown in 2021 was supported by  modelling studies informing the government's decision~\autocite{sageEvidence}.
+
+Each week, the members of SPI-M-O would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
+These estimates, once any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value~\autocite{parkCombining}.
+The consensus value was published on the UK Government's website and informed its response to the epidemic~\autocite{govRnumber}.
+
+The nowcasting estimates were mainly based on mechanistic and semi-mechanistic models with a wide variety of data sources.
+The mechanistic models commonly incorporated multiple data sources, known as \emph{data synthesis}, a technique reviewed by \textcite{birrellEvidence}.
+The models used were reviewed by \textcite{royalSocietyRnumber} (see appendix 2 of that review for details of the specific implementations).
+
+I will investigate whether estimates relying on only one data source, the CIS, could replace this process.
+Such estimates should be more robust, due to the representative nature of CIS.
+
+\subsection{Use of prevalence surveys}
+
+Several previous studies have used SARS-CoV-2 prevalence surveys to estimating incidence and/or transmission.
+
+The \emph{backcalculation} framework~\autocite[e.g.][]{brookmeyerMeasuring,brookmeyerMethod,brookmeyerBackcalculation}, developed in the context of HIV/AIDS
+\todo{this is a bad place for an intro to backcalculation: where does it go in this structure?}
+
+\Textcite{abbottCISincidence} applied backcalculation to CIS data to estimate the incidence of SARS-CoV-2 in the UK.
+They used a Gaussian process prior to smooth the incidence, applying the prior to either smooth the log-incidence or its derivative; smoothing of the incidence is required to stabilise the estimates~\autocite{brookmeyerBackcalculation}.
+\Textcite{abbottCISincidence} used \textcite{hellewellPCRSensitivity}'s estimates of the sensitivity of testing to inform the backcalculation, which are based on a longitudinal study in healthcare workers (see \cref{intro:sec:duration-SARS-CoV-2}).
+They fit to summary statistics of the CIS data, using a daily, independent Gaussian likelihood with parameters estimated from the posterior point estimate and 95\% credible interval published by the ONS.
+
+The estimates I present in \cref{E-backcalc} are methodologically similar.
+However, I fully propagate the uncertainty from the prevalence estimation procedure and use estimates of the duration of positivity from a general population study.
+\Textcite{abbottCISincidence} include the short period of low but increasing test sensitivity at the start of an infection (see \cref{E-biology-data:sec:natural-history}).
+
+\Textcite{colmanAscertainment} used CIS prevalence estimates to estimate the \emph{ascertainment rate} of community testing (the proportion of all infections that community testing detected).
+Their methodology was similar to \textcite{abbottCISincidence} (see \cref{intro:sec:backcalc}), including using estimates from \textcite{hellewellPCRSensitivity} for how the sensitivity of RT-PCR testing varies over an infection.
+Rather than using a Gaussian process, as \textcite{abbottCISincidence} do, \textcite{colmanAscertainment} provide stability by assuming the ascertainment rate is assumed to be piecewise constant.
+They do not fully propagate the uncertainty from the CIS prevalence estimates, considering only the bounds of the 95\% confidence intervals the ONS publishes for prevalence.
+
+Several compartmental models incorporated data from SARS-CoV-2 prevalence surveys, alongside other data sources~\autocite{daviesAssociation,ironsEstimating,knockKey,nicholsonImproving,pooleyEstimation,birrellRTM2}.
+These models incorporate prevalence survey data by adding additional compartments representing RT-PCR positive individuals, who may not be infectious (this is common, see \cref{E-biology-data}); except \textcite{nicholsonImproving} who have a more sophisticated approach (below) and \textcite{ironsEstimating}, who assume RT-PCR positivity coincides with infectiousness.
+The length of stay in these compartments (the duration of time positive) is assumed to follow a gamma distribution, with the parameters estimated from external data or having at most one free parameter.
+The proportion of the population in these compartments are then fit to the prevalence survey data using a binomial likelihood where possible, although some approaches take a Gaussian approximation.
+These approaches are similar to how these models are fit to other data sources, such as hospital admissions or deaths.
+
+\Textcite{nicholsonImproving} have the most sophisticated use of prevalence survey data.
+They allow an arbitrary distribution of the duration of positivity, based on the estimates of \textcite{hellewellPCRSensitivity} (see \cref{intro:sec:duration-SARS-CoV-2} for discussion of the \textcite{hellewellPCRSensitivity} study).
+Furthermore, they incorporate the sensitivity and specificity of the RT-PCR test into their model.
+
+\Textcite{mccabeCISincidence}, adapting the methodology \textcite{ealesAppropriately} developed for REACT, estimate $\Re$ and the growth rate from the CIS data.
+They estimate a posterior distribution of the parameters of a spline by fitting to the prevalence results via a binomial distribution.
+Their setup penalises the spline deviating from a constant growth rate.
+The growth rate is then calculated from the derivative of the spline, an analytical transformation of the parameters.
+Since, under reasonable model assumptions, there is a one-to-one correspondence between reproduction numbers and growth rates~\autocite{wallingaGI}, and they assume that the growth rate in incidence and prevalence is equal, this also estimates $\Re$.
+
+\subsection{Previous duration estimates}
 
@@
 
-\subsection{Markov chain Monte Carlo} \label{intro:sec:MCMC}
-
-MCMC is a class of algorithms for sampling from a distribution where the normalising constant for the distribution is unknown.
-I will use it to sample a set of parameters, say $\psi$ (which may be a scalar or vector), from a posterior distribution $p(\psi \mid y)$ where $y$ is the data (again, scalar or vector).
-MCMC algorithms produce a sample $\psi^1, \psi^2, \dots, \psi^M$ from $p(\psi \mid y)$ by constructing a Markov chain with $p(\psi \mid y)$ as its stationary distribution~\autocite[275]{gelmanBDA}.
-It can be proved that as $M\to\infty$, the sample tends towards a sample from the stationary distribution, \ie the target posterior.
-Further details of MCMC are given in \cref{E-MCMC}.
-Two practical issues arise when applying MCMC: \emph{convergence} and \emph{stability}~\autocite[72]{lunnBUGS}.
-
-Convergence is how close a MCMC chain's samples are to its stationary distribution.
-Early draws from the Markov chain reflect the choice of the starting point, $\psi^0$, rather than the stationary distribution that the chain will eventually converge to~\autocite[282]{gelmanBDA}.
-Various metrics quantify this, I will make use of the improved Rhat statistic~\autocite{vehtariRhat}.
-%The Rhat statistic is defined for a scalar parameter $\psi$.
-%Let $\psi^{(ij)}$ denote the $i$th (of $M$) sample from the $j$th chain (of $N$).
-%Let $\bar{\psi}^j$ denote the mean value of the samples of $\psi$ from the $j$th chain, $\bar{\psi}^j = \frac{1}{M} \sum_{i=1}^M \psi^{(ij)}$.
-%Let $\bar{\psi}$ denote the mean value of the samples of $\psi$ from all chains, $\bar{\psi} = \frac{1}{N} \sum_{j=1}^N \bar{\psi}^j$.
-%Then the between-chain variance is $B = \frac{M}{N-1} \sum_{j=1}^N (\bar{\psi}^j - \bar{\psi})^2$.
-%The within-chain variance is $W = \frac{1}{N} \sum_{j=1}^N \frac{1}{M-1} \sum_{i=1}^M (\psi^{(ij)} - \bar{\psi}^j)^2$.
-A Rhat of 1 means that the within-chain and between-chain variance are equal, informally that convergence is perfect, as far as can be measured by this technique.
-Various rules-of-thumb exist for what should be considered an acceptable Rhat.
-Historically, 1.1 has commonly been recommended, although more recent work suggests that 1.01 or 1.05 is more appropriate~\autocite{vehtariRhat}.
-
-The stability of estimates considers the error arising from only using a finite sample of the distribution and the samples being autocorrelated.
-Autocorrelation means that estimates using the sample have higher variance than if it were an iid sample from the posterior~\autocite[286]{gelmanBDA}.
-The \emph{effective sample size} corrects for the autocorrelation.
-The effective sample size can be interpreted as the number of iid samples that would have the same variance as the MCMC sample~\autocites[286]{gelmanBDA}{vehtariRhat}.
-Various techniques have been proposed for computing the effective sample size, I will follow the approach of \textcite{vehtariRhat}, implemented in RStan~\autocite{RStan-2-32-3}.
-
-% \section{Notational conventions}
-
-% \begin{itemize}
-%     \item Lower-case variables are realisations of the respective upper-case random variables.
-%     \item Matrices and vectors are bold and have elements with the same unbolded letter, \eg $\vec{x} = [x_1, x_2, \dots]^T$
-%     \item $\cdot$ denotes element-wise multiplication
-%     \item Distributions used defined in \cref{E-distributions}
-% \end{itemize}
-
 \section{Thesis aims and contribution} \label{intro:sec:aims}
 
-The overall question that I investigate in this thesis is: how can prevalence surveys best be used to understand incidence and transmission?
+The overall question that I investigate in this thesis is: can prevalence surveys alone be used to infer incidence and transmission, and if so how?
 Prevalence surveys provide data that is far more representative of the general population than other data sources (see \cref{intro:sec:data-sources}), therefore, conclusions based on them should be far more robust.
@@
 First, in \cref{E-inc-prev:sec:observation-process}, I provide a theoretic justification for the use of deterministic backcalculation in this setting.
-A deterministic backcalculation approach was used by \textcite{abbottCISincidence} and I will use it in \cref{E-backcalc}.
-This approach models the prevalence at any time as a deterministic function of the incidence, rather than as a random variable.
+A \emph{deterministic} backcalculation approach was used by \textcite{abbottCISincidence} and I will use it in \cref{E-backcalc}.
+The deterministic approach models the prevalence at any time as a deterministic function of the incidence, rather than as a random variable.
 I show that, this is a good approximation as long as the proportion of the population sampled is small.
 In particular, it does not require the prevalence or incidence to be large.
+The estimates I produce in \cref{E-backcalc} fully propagate the uncertainty from the prevalence estimation procedure and use estimates of the duration of positivity from a general population study, unlike prior work.
 
@@
 
-Finally, in \cref{E-SEIR}, I estimate the transmission of SARS-CoV-2 in England using a mechanistic model, from only the CIS data.
-This includes the development of a novel observation model to link the mechanistic transmission model to the data.
-The observation model allows the inclusion of the skew in the duration of positivity, without being computationally infeasible.
+Finally, in \cref{E-SEIR}, I describe a compartmental model for SARS-CoV-2 transmission, which I fit to CIS data.
+I allow a more flexible distribution of the duration of positivity than any of the papers discussed above, except \textcite{nicholsonImproving}, incorporating the skew of the distribution.
+This distribution is based on data from a representative sample of the UK population, rather than convenience samples (see \cref{intro:sec:duration-SARS-CoV-2}).
+Unlike \textcite{nicholsonImproving}, I include age-stratification in the transmission model, allowing estimation of how transmission varies by age.
+I use only data from the CIS, and show this is sufficient for estimating incidence and transmission.
 
@@
   \appendix
-  \subfile{MCMC-appendix}
+%   \subfile{MCMC-appendix}
   \listoftodos

### 2024-03-09 introduction-comments.tex
Overleaf comments

@@
 No single metric summarises the full state of an epidemic, and no single data source fully informs all metrics of interest~\autocite{royalSocietyRnumber,pellisEstimation,paragGrowthRates}.
-Therefore, a range of metrics and surveillance systems are required to understand the state of an epidemic and the danger of a pathogen (a pathogen is any biological agent that can cause disease).
+Therefore, a range of metrics and surveillance systems are required to understand the state of an epidemic and the risk it poses.
 
@@
 
-The most famous reproduction number is the \emph{basic reproduction number}, $\R$.
+The most common reproduction number is the \emph{basic reproduction number}, $\R$.
 $\R$ is the mean number of secondary infections generated by a single infection in a \emph{fully susceptible} population with \emph{standard behaviour patterns}.
@@
 In the early stages of an epidemic caused by a novel pathogen, the number of individuals with immunity is negligible, and hence $\R$ describes this part of the epidemic well.
-$\R$ has a threshold value of 1: if $\R < 1$ then the epidemic will die out.
+$\R$ has a threshold value of 1: an outbreak has the potential to cause an epidemic if and only if $\R < 1$~\autocite[76]{diekmannMathematical}.
 
@@
 Over short periods of time, epidemics tend to grow (or shrink) exponentially, and hence have an exponential growth rate, normally denoted $r$.
-For a given disease and population, an analogous growth rate can be defined for each reproduction number~\autocite{pellisEstimation,paragGrowthRates,wallingaGI}.
-The advantage of using the growth rate is that they are easier to estimate and capture how quickly spread happens in calendar time; however, they do not capture the strength of intervention required to control the epidemic~\autocite{royalSocietyRnumber}.
+For a given disease and population, an analogous growth rate can be defined for each definition of the reproduction number~\autocite{pellisEstimation,paragGrowthRates,wallingaGI}.
+The advantage of using the growth rate is that they are easier to estimate and capture how quickly spread happens in calendar time; however, they do not directly give the strength of intervention required to control the epidemic~\autocite{royalSocietyRnumber}.
 
@@
 \emph{Incidence} (the number of new infections in a given time period) is the primary metric for the magnitude of a pandemic.
-Incidence informs the burden of the disease, for example, the future number of severe events (hospitalisations or deaths).
+Incidence informs the short-term future burden of the disease, for example, the future number of severe events (hospitalisations or deaths).
+To translate incidence into the burden of the disease, a measure of severity is required.
+For example, the IFR (infection fatality rate), the proportion of infected individuals who die.
+Estimation of severity is beyond the scope of this thesis.
 
-\todo[inline]{What is a good definition of ``transmission'' as a concept?}
-The \emph{transmission} of an infectious disease generates new infections.
+\emph{Transmission} refers to the process by which an infectious disease generates new infections.
+Estimating transmission refers to estimating the parameters of this process, including its components.
 The components of transmission (\eg number of contacts per day and age-stratified rates) inform how interventions can be targeted.
@@
 \emph{Prevalence} is the proportion of the population with the disease at a given time.
-Whether an individual is \emph{prevalent} (has the disease) is normally measurable.
+Whether a given individual is \emph{prevalent} (has the disease) is often measurable through a diagnostic test.
 Therefore, the prevalence of the population can be estimated with a prevalence survey.
 However, these are very expensive to power appropriately; SARS-CoV-2 is unusual in that its impact is so large that the UK government did this.
-\cref{E-inc-prev:sec:definitions} provides a formal definition of incidence and prevalence.
 Prevalence and incidence are strongly related, with the relationship determined by by the duration of an infection (see \cref{E-inc-prev}).
 
-To translate incidence into the burden of the disease, a measure of severity is required.
-For example, the IFR (infection fatality rate), the proportion of infected individuals who die.
-Estimation of severity is beyond the scope of this thesis.
 % Severity does not feature in this thesis, but it is important to note that it is a key component of understanding the burden of a pandemic.\todo{maybe reword sentence}
@@
 However, incidence and transmission and prevalence, are almost never directly observed.
-Therefore, they need to be estimated from other data sources.
-
-Introduce next section eg incidence / transmission not observed directly, need infer from data such as next section.
+Therefore, they need to be inferred from other data sources.
 
@@
 
-A wide variety of designed studies, administrative data, and other data sources have been used for pandemic and epidemic surveillance.
 In the context of the SARS-CoV-2 pandemic in England, the most prominent were: community testing (known as pillar 2 testing), the number of severe events (\eg hospital admissions or deaths), seroprevalence testing (seroprevalence is the proportion of people with antibodies to a disease, a marker of prior infection), and prevalence surveys.
@@
 
-Community testing is broadly available, but does not test a random sample of the population.
+Community testing was broadly available in England for much of the SARS-CoV-2 epidemic.
+However, it does not test a random sample of the population.
 The number of tests conducted varies over time and space, for a variety of known and unknown reasons.
@@
 Therefore, fluctuations in the number of positive tests may not reflect changes in the true incidence of the disease.
-For large parts of the pandemic, community testing was widely available in the UK.
-Therefore, there are many test results, allowing detailed analyses to be performed, including stratification by small geographic areas or age groups.
+The large number of tests conducted allowed detailed analyses to be performed, including stratification by small geographic areas or age groups, although this does not solve the issue of the biases present.
 
-Seroprevalence testing provides information on a marker of recent prior infection.
+Seroprevalence surveys estimate the proportion of the population previously infected.
 Prior infection is more common than current infection, hence seroprevalence estimation requires a smaller sample size than current prevalence.
 However, there is a delay between infection and the development of antibodies, hence seroprevalence is not a real-time measure of the burden of the disease.
-Furthermore, seroprevalence data is normally based on convenience samples (\eg blood donors) who are not representative of the general population.
+Furthermore, seroprevalence data is normally based on convenience samples (\eg blood donors) whose representativenss is unclear.
 
@@
 Since its development, backcalculation has been further developed for a wide-variety of contexts.
-I incorporate observation noise into the backcalculation approach in \cref{inc-prev:sec:observation-process}.
+% I incorporate observation noise into the backcalculation approach in \cref{inc-prev:sec:observation-process}.
 % Reporting delay is not relevant to the studies in this thesis, and I ignore it in this and further discussion.
@@
 Many studies involving mechanistic models were conducted during the pandemic, with at least 100 articles published in the first six months of 2020 alone~\autocite{shankarSystematic}.
-In the UK, SPI-M-O (Scientific Pandemic Influenza Group on Modelling, Operational sub-group) was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), using this type of models to generate evidence for policy development~\autocite{medleySPIM,govSPIMO}.
-For example, each of the stages of lifting the lockdown in 2021 was supported by  modelling studies informing the government's decision~\autocite{sageEvidence}.
+In the UK, SPI-M-O (Scientific Pandemic Influenza Group on Modelling, Operational sub-group) was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), generating evidence based on infectious disease modelling for policy development~\autocite{medleySPIM,govSPIMO}.
+For example, the decision at each stage of lifting the lockdown in 2021 was informed by a set of modelling studies informing the government's decision~\autocite{sageEvidence}.
 
 Each week, the members of SPI-M-O would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
-The nowcasting estimates were mainly based on mechanistic models.
+The nowcasting estimates were based on mechanistic models.
 These estimates, once any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value~\autocite{parkCombining}.

### 2024-03-09 introduction.tex
Move Overleaf comments on prev version across

@@
 For large parts of the pandemic, community testing was widely available in the UK.
+However, it does not test a random sample of the population.
 The number of tests conducted varied over time and space, for a variety of known and unknown reasons.
@@
 Therefore, fluctuations in the number of positive tests may not reflect changes in the true incidence of the disease.
-However, there were many test results, allowing detailed analyses to be performed, including stratification by small geographic areas or age groups.
+The large number of tests conducted allowed detailed analyses to be performed, including stratification by small geographic areas or age groups, although this does not solve the issue of the biases present.
 
-Seroprevalence surveys indicate the proportion of the population that have been infected with the disease.
+Seroprevalence surveys estimate the proportion of the population previously infected.
 Prior infection is more common than current infection, hence seroprevalence estimation requires a smaller sample size than current prevalence.
 However, there is a delay between infection and the development of antibodies, hence seroprevalence is not a real-time measure of the burden of the disease.
-Furthermore, seroprevalence data is normally based on convenience samples (\eg blood donors) who are not representative of the general population.
+Furthermore, seroprevalence data is normally based on convenience samples (\eg blood donors) who are not representativeness of the wider population is unclear.
 
@@
 
-In the UK, SPI-M-O (Scientific Pandemic Influenza Group on Modelling, Operational sub-group) was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), using this type of models to generate evidence for policy development~\autocite{medleySPIM,govSPIMO}.
+In the UK, SPI-M-O (Scientific Pandemic Influenza Group on Modelling, Operational sub-group) was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), generating evidence based on infectious disease modelling for policy development~\autocite{medleySPIM,govSPIMO}.
 For example, each of the stages of lifting the lockdown in 2021 was supported by  modelling studies informing the government's decision~\autocite{sageEvidence}.
@@
 
-The nowcasting estimates were mainly based on mechanistic and semi-mechanistic models with a wide variety of data sources.
+The nowcasting estimates were based on mechanistic and semi-mechanistic models with a wide variety of data sources.
 The mechanistic models commonly incorporated multiple data sources, known as \emph{data synthesis}, a technique reviewed by \textcite{birrellEvidence}.

### 2024-03-09 introduction.tex
edit new structure

@@
 In all cases, surveillance is required to inform public health policy.
-In this thesis, I consider methods to improve surveillance, using the SARS-SARS-CoV-2 epidemic in England as a case study.
+In this thesis, I consider methods to improve surveillance, using the SARS-CoV-2 epidemic in England as a case study.
 
@@
 Growth rates and reproduction numbers only provide information on how an epidemic is changing.
-Yet the size of the epidemic is at least as important as how it is changing~\autocite{pellisEstimation}.
+Yet the size of the epidemic is just as, if not more, important~\autocite{pellisEstimation}.
 \emph{Incidence} (the number of new infections in a given time period) is the primary metric for the magnitude of a pandemic.
@@
 \emph{Transmission} refers to the process by which an infectious disease generates new infections.
-Estimating transmission refers to estimating the parameters of this process, including its components.
+Estimating transmission refers to estimating the parameters of this process, the infection process, including its components.
 The components of transmission (\eg number of contacts per day and age-stratified rates) inform how interventions can be targeted.
@@
 
+\todo[inline]{Not sure this section (data sources) needs to exist, particularly in this location}
+\todo[inline]{If this section kept, probably want to cite some things on generic data for pandemic surveillance}
+
 The data sources used for outbreak surveillance are often observational in nature.
@@
 
-\section{Transmission model}
+\section{Infection process model}
 
-The first component required to estimate incidence and transmission is a \emph{transmission model}, a mathematical description of how incidence arises.
-A spectrum of models exist, categorised by the strenght of assumptions they make about the transmission process.
+The first component required to estimate incidence and transmission is a model of the infection process, a mathematical description of how incidence arises.
+A spectrum of models exists, categorised by the strength of assumptions they make about the transmission process.
 Stronger assumptions allow more insight into the transmission process and generally lower the variance of the estimates, but can introduce bias and are often computationally challenging to fit.
@@
 The models with the weakest assumptions are statistical models.
-These make almost no assumptions about the transmission process, viewing incidence as a generic counting process.
-The most assumption is that incidence follows a Poisson process (defined in \cref{E-inc-prev:sec:poisson-process}), and the intensity of the process is the target of incidence.
+These make almost no assumptions about the transmission process, viewing incidence as a generic counting process (introduced in full in \cref{E-inc-prev:sec:infection-process}).
+The most common model is that incidence follows a Poisson process, and the intensity of the process is the target of incidence.
 % The use of Poisson processes to model incidence of an infectious disease was very popular in the context of HIV/AIDS\todo{maybe a ref for Poisson in HIV/AIDS?}.
@@
 % The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
-Renewal equations do not feature futher in this thesis, for details, including practical considerations when using them as the basis of a statistical model, see \textcite{nashEstimating,gosticPractical}.
+Renewal equations do not feature futher in this thesis, for details, including practical considerations when using them as the basis for inference, see \textcite{nashEstimating,gosticPractical}.
 
@@
 \emph{Compartmental models} are the most common type of mechanistic model.
-Compartmental models split the population into compartments, each representing a different state of the disease.
+Compartmental models split the population into compartments, each representing a different state of the disease (described in full in \cref{E-SEIR:sec:mechanistic-models}).
 The \emph{dynamics} of the epidemic, how it changes over time, are represented by the transitions between compartments.
-Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
-Unlike renewal equations, there are not widely-used implementations of compartmental models.
-A commonly used frameork is the SEIR (susceptible, exposed, infectious, recovered) model.
-The SEIR model, has four components (susceptible, exposed, infectious, recovered), with the population divided into these components; it is described in detail in \cref{E-SEIR:sec:mechanistic-models}.
+Simple mechanistic models use a homogeneous population, with individuals differing only due to their disease state, and exponential distributions for the time in each disease state.
+More complex models relax these assumptions, in particular the age structure of the population is important for respiratory diseases.
+% Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
+% Unlike renewal equations, there are not widely-used implementations of compartmental models.
+% A commonly used frameork is the SEIR (susceptible, exposed, infectious, recovered) model.
+% The SEIR model, has four components (susceptible, exposed, infectious, recovered), with the population divided into these components; it is described in detail in \cref{E-SEIR:sec:mechanistic-models}.
 
 \section{Observation models}
+
+\todo[inline]{Needs some generic citations about observation models}
 
@@
 
-Typically, the core of an observation model is a \emph{convolution equation} of an approximation of it.
+Typically, the core of an observation model is a \emph{convolution equation} or an approximation of it.
 This equation links an observation at time $t$, $y_t$, to the incidence prior to time $t$.
@@
 
+\todo[inline]{Expand slightly on convolution equations and why solving them is challenging, including some citations}
 Solving a convolution equation for $I$ is an ill-posed problem when $f(0)$ is small.
@@
 
-For a prevalence survey, the delay distribution is the \emph{duration of positivity}, the distribution of the time between an individual becoming infected and the time they are no longer infected.
+\section{Backcalculation}
 
-\section{Backcalculation}
+\todo[inline]{
+  Decide whether to keep this section.
+  I think backcalculation probably does need to be called out explicitly, but it feels a little out-of-place.
+  IF kept, needs expanding a bit, flow a bit better, and citations added.
+}
 
@@
 The aim of backcalculation is to estimate the incidence of a disease from the observation of a prevalence measure.
-It combines a statistical transmission model with
+It combines a statistical transmission model with an observation model for prevalence data.
+
+Here, $\zeta = 1$ and $f(t)$ represents the probability of having the disease $t$ days after infection.
+In this way, $f$ represents a duration.
+
+\todo[inline]{Following paragraph seems a bit specific but is needed to motivate the following section}
+For a prevalence survey, the delay distribution is the \emph{duration of positivity}, the distribution of the time between an individual becoming infected and the time they are no longer infected.
 
@@
 
-The application I will consider in this thesis is the SARS-CoV-2 epidemic in England.
-
-\subsection{Data sources}
-
-\subsubsection{For estimating incidence and transmission}
+\subsection{Data for estimating incidence and transmission} \label{intro:sec:data-for-estimating}
 
@@
 
-\subsubsection{Cohort studies}
+\subsection{Cohort studies}
 
-In this thesis, I make use of two cohort studies with differing designs.
-These designs have complementary strengths, and I combine information from the studies to make use of these.
+In this thesis, I make use of two cohort studies with differing designs to estimate duration.
+These designs have complementary strengths, and I combine information from the studies.
 Both of the studies recruited individuals from the general population, aiming to observe a cohort of infections that are representative of the infections in the UK community.
@@
 
-The second study is the CIS, which I previously mentioned.
+The second study is the CIS (introduced in \cref{intro:sec:data-for-estimating}).
 In addition to estimating prevalence, the CIS followed the same individuals over a long period of time (over a year in many cases), collecting longitudinal data.
@@
 
-The \emph{backcalculation} framework~\autocite[e.g.][]{brookmeyerMeasuring,brookmeyerMethod,brookmeyerBackcalculation}, developed in the context of HIV/AIDS
-\todo{this is a bad place for an intro to backcalculation: where does it go in this structure?}
-
 \Textcite{abbottCISincidence} applied backcalculation to CIS data to estimate the incidence of SARS-CoV-2 in the UK.
-They used a Gaussian process prior to smooth the incidence, applying the prior to either smooth the log-incidence or its derivative; smoothing of the incidence is required to stabilise the estimates~\autocite{brookmeyerBackcalculation}.
-\Textcite{abbottCISincidence} used \textcite{hellewellPCRSensitivity}'s estimates of the sensitivity of testing to inform the backcalculation, which are based on a longitudinal study in healthcare workers (see \cref{intro:sec:duration-SARS-CoV-2}).
+They used a Gaussian process prior to smooth the incidence and stabilise the estimates.
+\Textcite{abbottCISincidence} used \textcite{hellewellPCRSensitivity}'s estimates of the sensitivity of testing to inform the observation model, which are based on a longitudinal study in healthcare workers (see \cref{intro:sec:duration-SARS-CoV-2}).
 They fit to summary statistics of the CIS data, using a daily, independent Gaussian likelihood with parameters estimated from the posterior point estimate and 95\% credible interval published by the ONS.
 
-The estimates I present in \cref{E-backcalc} are methodologically similar.
-However, I fully propagate the uncertainty from the prevalence estimation procedure and use estimates of the duration of positivity from a general population study.
-\Textcite{abbottCISincidence} include the short period of low but increasing test sensitivity at the start of an infection (see \cref{E-biology-data:sec:natural-history}).
-
 \Textcite{colmanAscertainment} used CIS prevalence estimates to estimate the \emph{ascertainment rate} of community testing (the proportion of all infections that community testing detected).
-Their methodology was similar to \textcite{abbottCISincidence} (see \cref{intro:sec:backcalc}), including using estimates from \textcite{hellewellPCRSensitivity} for how the sensitivity of RT-PCR testing varies over an infection.
+Their methodology was similar to \textcite{abbottCISincidence}, including using estimates from \textcite{hellewellPCRSensitivity} for how the sensitivity of RT-PCR testing varies over an infection.
 Rather than using a Gaussian process, as \textcite{abbottCISincidence} do, \textcite{colmanAscertainment} provide stability by assuming the ascertainment rate is assumed to be piecewise constant.
@@
 Several compartmental models incorporated data from SARS-CoV-2 prevalence surveys, alongside other data sources~\autocite{daviesAssociation,ironsEstimating,knockKey,nicholsonImproving,pooleyEstimation,birrellRTM2}.
-These models incorporate prevalence survey data by adding additional compartments representing RT-PCR positive individuals, who may not be infectious (this is common, see \cref{E-biology-data}); except \textcite{nicholsonImproving} who have a more sophisticated approach (below) and \textcite{ironsEstimating}, who assume RT-PCR positivity coincides with infectiousness.
-The length of stay in these compartments (the duration of time positive) is assumed to follow a gamma distribution, with the parameters estimated from external data or having at most one free parameter.
-The proportion of the population in these compartments are then fit to the prevalence survey data using a binomial likelihood where possible, although some approaches take a Gaussian approximation.
-These approaches are similar to how these models are fit to other data sources, such as hospital admissions or deaths.
+% These models incorporate prevalence survey data by adding additional compartments representing RT-PCR positive individuals, who may not be infectious (this is common, see \cref{E-biology-data}); except \textcite{nicholsonImproving} who have a more sophisticated approach (below) and \textcite{ironsEstimating}, who assume RT-PCR positivity coincides with infectiousness.
+% The length of stay in these compartments (the duration of time positive) is assumed to follow a gamma distribution, with the parameters estimated from external data or having at most one free parameter.
+The duration of positivity in these models is limited by the model structure.
+The models are fit to the prevalence survey data using a binomial likelihood where possible, although some approaches take a Gaussian approximation.
+% These approaches are similar to how these models are fit to other data sources, such as hospital admissions or deaths.
 
-\Textcite{nicholsonImproving} have the most sophisticated use of prevalence survey data.
+\Textcite{nicholsonImproving} fit a simple mechanistic model, notably without age structure, but have the most sophisticated observation model.
 They allow an arbitrary distribution of the duration of positivity, based on the estimates of \textcite{hellewellPCRSensitivity} (see \cref{intro:sec:duration-SARS-CoV-2} for discussion of the \textcite{hellewellPCRSensitivity} study).

### 2024-03-11 biology-data.tex
fix subscripts for survival analysis

@@
 The crucial features of an infection episode are the start and end times of the infection.
-Ignoring misclassified test results, latent start time of infection episode $i$, $B_i$, can be bounded as between the day after the last negative test, $l_i^{(b)}$, and the day of the first positive test, $r_i^{(b)}$.
-Similarly, the latent end time of infection episode $i$, $E_i$, can be bounded as between the day of the last positive test, $l_i^{(e)}$, and the day before the following negative test, $r_i^{(e)}$.
+Ignoring misclassified test results, latent start time of infection episode $j$, $B_j$, can be bounded as between the day after the last negative test, $l_j^{(b)}$, and the day of the first positive test, $r_j^{(b)}$.
+Similarly, the latent end time of infection episode $i$, $E_j$, can be bounded as between the day of the last positive test, $l_j^{(e)}$, and the day before the following negative test, $r_j^{(e)}$.
 

### 2024-03-11 cis-imperfect-testing.tex
fix ref
fix subscripts for survival analysis

@@
 A single positive episode is an episode containing exactly one positive test; intuitively, these cause estimates to be shorter because the lower bound on the length of time the episode is 1 day, the shortest possible.
-Specifically, if $i$ is a single positive episode then $i$'s admissible region (as defined in \cref{E-perf-test:sec:model}) includes $b_i = e_i = r_i^{(b)} = l_i^{(e)}$.
+Specifically, if $j$ is a single positive episode then $j$'s admissible region (as defined in \cref{E-perf-test:sec:model}) includes $b_j = e_j = r_j^{(b)} = l_j^{(e)}$.
 In this case, the beginning and end day of the episode are both equal to the day of this one positive test.
-If this is true, then the duration is $d_i = e_i - b_i + 1 = 1$.
+If this is true, then the duration is $d_j = e_j - b_j + 1 = 1$.
 Additionally, short episodes are very likely to be missed, which means that the design of the CIS cannot rule out many of them occurring.
@@
 
-First, in \cref{imperf-test:sec:modifying-p_ia}, I modify $p_{ia}$, the probability episode $i$ being admissible, to allow for the episode possibly being longer than observed.
-Then, in \cref{imperf-test:sec:modifying-p_iu}, I modify $p_{iu}$, the probability episode $i$ being undetected, to allow for additional episodes being missed.
+First, in \cref{imperf-test:sec:modifying-p_ia}, I modify $p_{ia}$, the probability of an episode with test schedule $i$ being admissible, to allow for the episode possibly being longer than observed.
+Then, in \cref{imperf-test:sec:modifying-p_iu}, I modify $p_{iu}$, the probability of an episode with test schedule $i$ being undetected, to allow for additional episodes being missed.
 
@@
 
-\todo[inline]{Proof-read the following paragraphs (up until cases). Also check indices: is $i$ indexing individuals, episodes, or schedules?}
 I start by finding the tests that need consideration.
 For tractability and simplicity, I consider only tests between the negative tests providing an upper bound on the length of the episode.
-By definition (see\todo{insert where these are defined in chapter 2}), these are the tests between $l_i^{(b)} - 1$ and $r_i^{(e)} + 1$ inclusive.
+By definition (see \cref{E-biology-data:sec:cis-episodes}), these are the tests between $l_{j(i)}^{(b)} - 1$ and $r_{j(i)}^{(e)} + 1$ inclusive.
 They are a subset of $\sched_i$, the set of test times for the individual in which episode $i$ occurs.
 
-For tractability, assume that the negative test bounding the start of the episode, on day $l_i^{(b)}-1$, is a true negative.
-This assumption is reasonable because, since a positive test follows at $r_i^{(b)}$, the negative at $l_i^{(b)}-1$ is likely early in the infection and the test sensitivity is high early in an infection.
+For tractability, assume that the negative test bounding the start of the episode, on day $l_{j(i)}^{(b)}-1$, is a true negative.
+This assumption is reasonable because, since a positive test follows at $r_{j(i)}^{(b)}$, the negative at $l_{j(i)}^{(b)}-1$ is likely early in the infection and the test sensitivity is high early in an infection.
 Therefore, this is unlikely to be a false negative because, if the individual was detectable at this time, they are in a period with high test sensitivity.
 This true negative, occurs with probability 1, and hence does not contribute to the likelihood.
-Denote the remaining tests considered as $\sched'_i = \{ t \in \sched_i \ssep r_i^{(b)} \leq t \leq r_i^{(e)} + 1 \}$, and their results by $\vec{y_i} = \{ y_i(t) \ssep t \in \sched'_i \}$ where $y_i(t) = 1$ if the test on day $t$ is positive and 0 otherwise.
+Denote the remaining tests considered as $\sched'_i = \{ t \in \sched_i \ssep r_{j(i)}^{(b)} \leq t \leq r_{j(i)}^{(e)} + 1 \}$, and their results by $\vec{y_{j(i)}} = \{ y_{j(i)}(t) \ssep t \in \sched'_i \}$ where $y_{j(i)}(t) = 1$ if the test on day $t$ is positive and 0 otherwise.
 
-As I assume that there are no false positives, the infection episode must span at least the period $[r^{(b)}_i, l^{(e)}_i]$ because it starts and ends with a positive test.
-This includes all $t \in \sched'_i$ except $t = r_i^{(e)}+1$.
-Therefore, the test results at times in $\sched_i'$, except the test at $r_i^{(e)}+1$, are either true positives or false negatives.
+As I assume that there are no false positives, the infection episode must span at least the period $[r^{(b)}_{j(i)}, l^{(e)}_{j(i)}]$ because it starts and ends with a positive test.
+This includes all $t \in \sched'_i$ except $t = r_{j(i)}^{(e)}+1$.
+Therefore, the test results at times in $\sched_i'$, except the test at $r_{j(i)}^{(e)}+1$, are either true positives or false negatives.
 
-Consider the negative test at $r_i^{(e)}+1$, the first negative after the start of the episode which may be a false negative.
-It is a false negative if and only if the episode ends after the test, \ie $e_i > r_i^{(e)}$.
-I proceed by considering whether this is the case, and considering the case with $b_i$ known; $b_i$ will later be integrated out.
+Consider the negative test at $r_{j(i)}^{(e)}+1$, the first negative after the start of the episode which may be a false negative.
+It is a false negative if and only if the episode ends after the test, \ie $e_{j(i)} > r_{j(i)}^{(e)}$.
+I proceed by considering whether this is the case, and considering the case with $b_{j(i)}$ known; $b_{j(i)}$ will later be integrated out.
 
-First, if $e_i \leq r_i^{(e)}$, meaning that the test at $r_i^{(e)}+1$ is a true negative and the end of the episode is interval-censored as in the previous chapter.
-% In this case, the test at $r_i^{(e)} + 1$ is a true negative, as are all other tests not in $\sched'_i$.
+First, if $e_{j(i)} \leq r_{j(i)}^{(e)}$, meaning that the test at $r_{j(i)}^{(e)}+1$ is a true negative and the end of the episode is interval-censored as in the previous chapter.
+% In this case, the test at $r_{j(i)}^{(e)} + 1$ is a true negative, as are all other tests not in $\sched'_i$.
 The true negative occurs with probability 1, by the assumption of no false positives.
 \begin{align}
-&p(\vec{y_i}, e_i \leq r_i^{(e)} | b_i, p_\text{sens}, \theta) \\
-&= p(\vec{y_i}, l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\ % &\text{as no false positives}
-&= p(\vec{y_i} \mid l_i^{(e)} \leq e_i \leq r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(l_i^{(e)} \leq e_i \leq r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
-&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) \right)
+&p(\vec{y_{j(i)}}, e_{j(i)} \leq r_{j(i)}^{(e)} | b_{j(i)}, p_\text{sens}, \theta) \\
+&= p(\vec{y_{j(i)}}, l_{j(i)}^{(e)} \leq e_{j(i)} \leq r_{j(i)}^{(e)} | t_i, b_{j(i)}, p_\text{sens}, \theta) \\ % &\text{as no false positives}
+&= p(\vec{y_{j(i)}} \mid l_{j(i)}^{(e)} \leq e_{j(i)} \leq r_{j(i)}^{(e)}, t_i, b_{j(i)}, p_\text{sens}, \theta) p(l_{j(i)}^{(e)} \leq e_{j(i)} \leq r_{j(i)}^{(e)} | t_i, b_{j(i)}, p_\text{sens}, \theta) \\
+&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_{j(i)}(t)} (1 - p_\text{sens})^{(1 - y_{j(i)}(t))} \right) \left( S_\theta(l_{j(i)}^{(e)} - b_{j(i)} - 1) - S_\theta(r_{j(i)}^{(e)} - b_{j(i)} - 1) \right)
 \label{imperf-test:eq:ll-ei-lt-ri}
@@
 
-Second, if $e_i > r_i^{(e)}$.
-In this case, the test at $r_i^{(e)}$ is a false negative, occurring with probability $(1 - p_\text{sens})$.
-To avoid having to consider tests after $r_i^{(e)}$, which could greatly complicate the likelihood, I model this case as the episode being right-censored at $r_i^{(e)}$.
+Second, if $e_{j(i)} > r_{j(i)}^{(e)}$.
+In this case, the test at $r_{j(i)}^{(e)}$ is a false negative, occurring with probability $(1 - p_\text{sens})$.
+To avoid having to consider tests after $r_{j(i)}^{(e)}$, which could greatly complicate the likelihood, I model this case as the episode being right-censored at $r_{j(i)}^{(e)}$.
 Taking the same approach as before:
 \begin{align}
-&p(\vec{y_i}, e_i > r_i^{(e)} | b_i, p_\text{sens}, \theta) \\
-&= p(\vec{y_i} \mid e_i > r_i^{(e)}, t_i, b_i, p_\text{sens}, \theta) p(e_i > r_i^{(e)} | t_i, b_i, p_\text{sens}, \theta) \\
-&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) (1 - p_\text{sens}) S_\theta(r_i^{(e)} - b_i - 1)
+&p(\vec{y_{j(i)}}, e_{j(i)} > r_{j(i)}^{(e)} | b_{j(i)}, p_\text{sens}, \theta) \\
+&= p(\vec{y_{j(i)}} \mid e_{j(i)} > r_{j(i)}^{(e)}, t_i, b_{j(i)}, p_\text{sens}, \theta) p(e_{j(i)} > r_{j(i)}^{(e)} | t_i, b_{j(i)}, p_\text{sens}, \theta) \\
+&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_{j(i)}(t)} (1 - p_\text{sens})^{(1 - y_{j(i)}(t))} \right) (1 - p_\text{sens}) S_\theta(r_{j(i)}^{(e)} - b_{j(i)} - 1)
 \label{imperf-test:eq:ll-ei-gt-ri}
@@
 
-These expressions can now be used to form the full replacement for $p_{ia}$, $p'_{ia}$, the likelihood of the data $\vec{y_i}$.
-First, augment the data with $b_i$, and split into the cases just discussed:
+These expressions can now be used to form the full replacement for $p_{ia}$, $p'_{ia}$, the likelihood of the data $\vec{y_{j(i)}}$.
+First, augment the data with $b_{j(i)}$, and split into the cases just discussed:
 \begin{align}
 p_{ia}'
-&= p(\vec{y_i} \mid p_\text{sens}, \theta) \\
-&= \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( p(\vec{y_i}, e_i \leq r_i^{(e)} \mid b_i, p_\text{sens}, \theta) + p(\vec{y_i}, e_i > r_i^{(e)} \mid b_i, p_\text{sens}, \theta) \right) p(b_i \mid p_\text{sens}, \theta).
+&= p(\vec{y_{j(i)}} \mid p_\text{sens}, \theta) \\
+&= \sum_{b_{j(i)} = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( p(\vec{y_{j(i)}}, e_{j(i)} \leq r_{j(i)}^{(e)} \mid b_{j(i)}, p_\text{sens}, \theta) + p(\vec{y_{j(i)}}, e_{j(i)} > r_{j(i)}^{(e)} \mid b_{j(i)}, p_\text{sens}, \theta) \right) p(b_{j(i)} \mid p_\text{sens}, \theta).
 \intertext{Now, substitute in \cref{imperf-test:eq:ll-ei-lt-ri,imperf-test:eq:ll-ei-gt-ri}, then take out the common factor:}
-&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \\ & \ \times \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - S_\theta(r_i^{(e)} - b_i - 1) + (1 - p_\text{sens}) S_\theta(r_i^{(e)} - b_i - 1) \right) \\ & \ \times p(b_i \mid p_\text{sens}, \theta) \\
-&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_i(t)} (1 - p_\text{sens})^{(1 - y_i(t))} \right) \\ & \ \times \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b_i - 1) - p_\text{sens} S_\theta(r_i^{(e)} - b_i - 1) \right) p(b_i \mid p_\text{sens}, \theta).
+&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_{j(i)}(t)} (1 - p_\text{sens})^{(1 - y_{j(i)}(t))} \right) \\ & \ \times \sum_{b_{j(i)} = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_\theta(l_{j(i)}^{(e)} - b_{j(i)} - 1) - S_\theta(r_{j(i)}^{(e)} - b_{j(i)} - 1) + (1 - p_\text{sens}) S_\theta(r_{j(i)}^{(e)} - b_{j(i)} - 1) \right) \\ & \ \times p(b_{j(i)} \mid p_\text{sens}, \theta) \\
+&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_{j(i)}(t)} (1 - p_\text{sens})^{(1 - y_{j(i)}(t))} \right) \\ & \ \times \sum_{b_{j(i)} = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_\theta(l_{j(i)}^{(e)} - b_{j(i)} - 1) - p_\text{sens} S_\theta(r_{j(i)}^{(e)} - b_{j(i)} - 1) \right) p(b_{j(i)} \mid p_\text{sens}, \theta).
 \label{imperf-test:eq:pia-prime}
@@
 
-If $\psens$ is fixed (\ie has a point prior) and $p(b_i \mid \psens, \theta) \propto 1$ (as assumed in \cref{E-perf-test}) then:
+If $\psens$ is fixed (\ie has a point prior) and $p(b_{j(i)} \mid \psens, \theta) \propto 1$ (as assumed in \cref{E-perf-test}) then:
 \begin{align}
 p_{ia}'
-&\propto \sum_{b_i = l_i^{(b)}}^{r_i^{(b)}} S_\theta(l_i^{(e)} - b_i - 1) - p_\text{sens} S_\theta(r_i^{(e)} - b_i - 1).
+&\propto \sum_{b_{j(i)} = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} S_\theta(l_{j(i)}^{(e)} - b_{j(i)} - 1) - p_\text{sens} S_\theta(r_{j(i)}^{(e)} - b_{j(i)} - 1).
 \label{imperf-test:eq:pia-prime-constant}
@@
 I now consider the additional mechanisms arising due to false negatives.
-Specifically, episode $i$ could be undetected if the first test after the $b_i$, episode $i$'s start day, is a false negative and then there are no subsequent positive tests.
+Specifically, episode $i$ could be undetected if the first test after the $b_{j(i)}$, episode $i$'s start day, is a false negative and then there are no subsequent positive tests.
 
-This false negative would occur at the first test after the infection episode begins, on day $b_i + \tau_{\sched_i}(b_i)$ (see \cref{E-perf-test:sec:model}).
+This false negative would occur at the first test after the infection episode begins, on day $b_{j(i)} + \tau_{\sched_i}(b_{j(i)})$ (see \cref{E-perf-test:sec:model}).
 A false negative occurring requires that the episode has not yet ended but a negative still occurs.
-The episode has not yet ended at the time of the test if $e_i = b_i + d_i - 1$ is after the test, that is the duration of the infection $d_i > \tau_{\sched_i}(b_i)$.
+The episode has not yet ended at the time of the test if $e_{j(i)} = b_{j(i)} + d_i - 1$ is after the test, that is the duration of the infection $d_i > \tau_{\sched_i}(b_{j(i)})$.
 Conditional on the episode having not yet ended, the test result is negative with probability $1 - \psens$.
 
-For there to be no subsequent positive tests, all tests up until day $e_i$ are false negatives.
+For there to be no subsequent positive tests, all tests up until day $e_{j(i)}$ are false negatives.
 I assume there is a negligible probability of two false negatives.
 Therefore, this can only occur if the episode ends before another test.
-Denote the number of days between $b_i$ and the test following the false negative as $\tau^2_{\sched_i}(b_i) \stackrel{\text{def}}{=} \tau_{\sched_i}(\tau_{\sched_i}(b_i) + 1)$.
-Then, the episode ends before this test if $d_i < \tau^2_{\sched_i}(b_i)$.
+Denote the number of days between $b_{j(i)}$ and the test following the false negative as $\tau^2_{\sched_i}(b_{j(i)}) \stackrel{\text{def}}{=} \tau_{\sched_i}(\tau_{\sched_i}(b_{j(i)}) + 1)$.
+Then, the episode ends before this test if $d_i < \tau^2_{\sched_i}(b_{j(i)})$.
 
@@
 \begin{enumerate}
-    \item The episode would have been detected considering only the mechanisms in \cref{E-perf-test:eq:piu}. That is $b_i > \min(\sched_i)$ and $\tau_{\sched_i}(b_i) + b_i \leq e_i$.
-    \item The episode ends in the interval $[\tau_{\sched_i}(b_i) + b_i, \tau^2_{\sched_i}(b_i) + b_i]$.
-      Note that the lower bound here is exactly the bound on $e_i$ in the previous condition.
-      Equivalently, $\tau_{\sched_i}(b_i) + 1 \leq d_i \leq \tau^2_{\sched_i}(b_i) + 1$.
-    \item A false negative occurs on day $\tau_{\sched_i}(b_i) + b_i$. Conditional on the previous condition, this occurs with probability $1 - \psens$.
+    \item The episode would have been detected considering only the mechanisms in \cref{E-perf-test:eq:piu}.
+That is $b_{j(i)} > \min(\sched_i)$ and $\tau_{\sched_i}(b_{j(i)}) + b_{j(i)} \leq e_{j(i)}$.
+    \item The episode ends in the interval $[\tau_{\sched_i}(b_{j(i)}) + b_{j(i)}, \tau^2_{\sched_i}(b_{j(i)}) + b_{j(i)}]$.
+      Note that the lower bound here is exactly the bound on $e_{j(i)}$ in the previous condition.
+      Equivalently, $\tau_{\sched_i}(b_{j(i)}) + 1 \leq d_i \leq \tau^2_{\sched_i}(b_{j(i)}) + 1$.
+    \item A false negative occurs on day $\tau_{\sched_i}(b_{j(i)}) + b_{j(i)}$. Conditional on the previous condition, this occurs with probability $1 - \psens$.
 \end{enumerate}
 
-The probability of this occurring, conditional on $B_i = b_i > \min(\sched_i)$ is:
+The probability of this occurring, conditional on $B_{j(i)} = b_{j(i)} > \min(\sched_i)$ is:
 \begin{align}
 &\prob \left(
-    \tau_{\sched_i}(b_i) + 1 \leq D_i \leq \tau^2_{\sched_i}(b_i) + 1
-    \mid b_i,
+    \tau_{\sched_i}(b_{j(i)}) + 1 \leq D_i \leq \tau^2_{\sched_i}(b_{j(i)}) + 1
+    \mid b_{j(i)},
 \theta \right) (1 - \psens) \\
-&= \left( S_\theta(\tau_{\sched_i}(b_i) + 1) - S_\theta(\tau^2_{\sched_i}(b_i) + 1) \right) (1 - \psens).
+&= \left( S_\theta(\tau_{\sched_i}(b_{j(i)}) + 1) - S_\theta(\tau^2_{\sched_i}(b_{j(i)}) + 1) \right) (1 - \psens).
 \end{align}
-Integrating over $b_i$ for $\min \sched_i < b_i \leq T$ ($T$ being the last day of the period considered), in the same way as \cref{E-perf-test:eq:piu}, gives:
+Integrating over $b_{j(i)}$ for $\min \sched_i < b_{j(i)} \leq T$ ($T$ being the last day of the period considered), in the same way as \cref{E-perf-test:eq:piu}, gives:
 \begin{align}

### 2024-03-11 cis-perfect-testing.tex
fix subscripts for survival analysis

@@
 %   of the time of the infection. Define the survival function
-%   $\prob(D_i \geq t \mid B_i = b, \theta) = \prob(D_i \geq t \mid \theta) = S_\theta(t)$,
+%   $\prob(D_i \geq t \mid B_{j(i)} = b, \theta) = \prob(D_i \geq t \mid \theta) = S_\theta(t)$,
 %   where $\theta$ are the parameters controlling the survival
@@
 %   The beginning of episode $i$ is known to occur in the interval
-%   $[l_i^{(b)}, r_i^{(b)}]$, and similarly for the end of the infection
-%   in $[l_i^{(e)}, r_i^{(e)}]$.
+%   $[l_{j(i)}^{(b)}, r_{j(i)}^{(b)}]$, and similarly for the end of the infection
+%   in $[l_{j(i)}^{(e)}, r_{j(i)}^{(e)}]$.
 % \end{itemize}
@@
 The second complication is that the exact beginning and end times of detected infection episodes are unknown.
-The imprecision in the beginning of the infection episode arises because the only observation that an episode has started is that an individual was negative at one test, on day $l_i^{(b)}-1$ (the minus one is because the first day that the infection could have begun is the day after the negative test), and positive at the next test, on day $r_i^{(b)}$.
-Similarly, the end of the infection episode is bounded by the final positive test on day $l_i^{(e)}$ and the following negative test on day $r_i^{(e)}+1$ (the plus one is because the last day that the infection could have ended is the day before the negative test).
+The imprecision in the beginning of the infection episode arises because the only observation that episode $j$ has started is that an individual was negative at one test, on day $l_j^{(b)}-1$ (the minus one is because the first day that the infection could have begun is the day after the negative test), and positive at the next test, on day $r_j^{(b)}$.
+Similarly, the end of the infection episode is bounded by the final positive test on day $l_j^{(e)}$ and the following negative test on day $r_j^{(e)}+1$ (the plus one is because the last day that the infection could have ended is the day before the negative test).
 \Cref{perf-test:fig:double-interval-censor} shows this issue graphically.
@@
 Let the total number of infection episodes in the cohort be $\ntot = \ndet + \nnodet$ where $\ndet$ is the number of detected infections and $\nnodet$ is the number of undetected infections.
-Let $\Ncis$ be the number of individuals in the cohort and $n_\sched$ be the number of test schedules.
+Let $\Ncis$ be the number of individuals in the cohort and $\nsched$ be the number of test schedules.
 %Without loss of generality, let the individuals with a detected episode be indexed by $i = 1, \dots, \ndet$ and the individuals without a detected episode be indexed by $i = \ndet + 1, \dots, \Ncis$.
-Let $i = 1, \dots \ntot$ index infection episodes.
+Let $i = 1, \dots, \nsched$ index infection schedules and $j = 1, \dots \ntot$ index infection episodes.
 
-All the relevant information about episode $i$ can be fully characterised by the (unobserved) triplet $(b_i, e_i, \sched_i)$ where $b_i$ and $e_i$ are the times the episode being and end respectively and $\sched_i$ is the set of test times for the individual in which the episode occurs.
-The episode's duration, the number of days the individual in which the episode occurred, was positive for is $d_i = e_i - b_i + 1$.
+All the relevant information about episode $i$ can be fully characterised by the (unobserved) triplet $(b_j, e_j, \sched_j)$ where $b_j$ and $e_j$ are the times the episode being and end respectively and $\sched_j$ is the set of test times for the individual in which the episode occurs.
+The episode's duration is $d_j = e_j - b_j + 1$.
+%, the number of days the individual in which the episode occurred, was positive for 
 This triplet is assumed iid.
-If $\sched_i$ contains at least one time between $b_i$ and $e_i$ (inclusive), one time prior to $b_i$, and one time after $e_i$, then the episode is detected, otherwise it is undetected.
-$b_i$ and $e_i$ span possible infection and recovery times.
-$\sched_i$ is one of the observed test schedules, with any other test schedule assumed to occur with probability zero.
+
+$b_j$ and $e_j$ span possible infection and recovery times.
+$\sched_j$ is one of the observed test schedules, with any other test schedule assumed to occur with probability zero.
 Over the short period of time considered, each unique test schedule has zero or one detected episodes.
-Let $\set{D}$ be the set of test schedules with a detected infection in the cohort.
 
-To relate the state space to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie: are strict subsets of $T \times T$), for each test schedule.
+Let $i \in \set{D}$ indicate $\sched_i$ is associated with a detected episode.
+That is, $i \in \set{D}$ if and only if there exists a $j(i)$ such that $b_{j(i)}$ and $e_{j(i)}$ such that $(b_{j(i)}, e_{j(i)}, \sched_i)$ is a detected episode.
+$(b_{j(i)}, e_{j(i)}, \sched_i)$ is a detected episode when the following three conditions are met.
+First, $\exists t \in \sched_i$ such that $b_{j(i)} \leq t \leq e_{j(i)}$; this condition is equivalent to having at least one positive test for the episode.
+Second, $\exists t \in \sched_i$ such that $t < b_{j(i)}$, to lower bound the start of the episode.
+Finally, $\exists t \in \sched_i$ such that $t > e_{j(i)}$, to upper bound the end of the episode; this condition will be met for any episode sufficiently far in the past in individuals not lost to follow-up (almost always the case in this context).
+For a new context where recent infections need to be considered, this condition can be relaxed by considering these episodes as right-censored, at little cost except a more complicated likelihood expression.
+
+To relate the state space to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie: are strict subsets of $T \times T$), for a test schedule $\sched_i$.
 Figure \ref{perf-test:fig:partitionSpace} shows this graphically.
@@
   Admissible episodes, $\alpha_i$, which have an infection and recovery time consistent with the data observed.
-  This space is empty for $i \notin \set{D}$.
-  If $i \in \set{D}$ then $n_{ia} =1$ admissible episodes occurred in an individual with test schedule $\sched_{i}$, otherwise $n_{ia} = 0$.
+  This space is empty for $i \notin \set{D}$; in this case, $n_{ia} = 0$ admissible episodes occurred in individuals with test schedule $\sched_i$.
+  Otherwise, $i \in \set{D}$ then $n_{ia} =1$ admissible episode occurred in an individual with test schedule $\sched_{i}$.
   Conditional on the episode occurring in individuals with test schedule $\sched_i$, it is admissible with probability $p_{ia} = \prob((b, e) \in \alpha_i \mid \theta)$.
@@
 These three classes span all possible episodes and are mutually exclusive.
-Hence, $\ntot = \ndet + \nnodet + n_{i\bar{a}} = \sum_{i=1}^{\ndet} n_{ia} + \sum_{i=1}^{\ntot} n_{iu} + 0$.
+Hence, $\ntot = \ndet + \nnodet + n_{i\bar{a}} = \sum_{i=1}^{\nsched} n_{ia} + \sum_{i=1}^{\nsched} n_{iu} + 0$.
 Furthermore, for all $i$, $p_{ia} + p_{iu} + p_{i\bar{a}} = 1$.
 
-Any episode $i$ independently belongs to one of these classes.
+Any episode $j$ independently belongs to one of these classes.
 I assume the individual in which each episode occurs is iid uniform across the cohort.
@@
 Because the counts have a joint multinomial distribution, this is equivalent to only considering the sum of the counts.
-The overall probability of an episode being unobserved is $\pnodet = \frac{1}{\Ncis} \sum_{i=1}^{n_\sched} \eta_i p_{iu}$.
+The overall probability of an episode being unobserved is $\pnodet = \frac{1}{\Ncis} \sum_{i=1}^{\nsched} \eta_i p_{iu}$.
 Furthermore, the inadmissible classes are guaranteed to have zero episodes.
@@
 
-Let $\na = (n_{1a}, \dots, n_{n_\sched a})^T$ be the vector of counts of episodes in the admissible classes.
+Let $\na = (n_{1a}, \dots, n_{\nsched a})^T$ be the vector of counts of episodes in the admissible classes.
 The counts $\na$ and $\nnodet$ are jointly distributed multinomially.
-The support of the distribution requires $\nnodet = \ntot - \sum_{i=1}^{n_\sched} n_{ia} = \ntot - \ndet$ and $\ntot \geq \ndet$.
+The support of the distribution requires $\nnodet = \ntot - \sum_{i=1}^{\nsched} n_{ia} = \ntot - \ndet$ and $\ntot \geq \ndet$.
 Hence:
@@
 &= p(\nnodet, \na\mid \ntot, \theta) \\
-&= \frac{\ntot!}{\nnodet! \prod_{i=1}^{n_\sched} n_{ia}!}  \pnodet^{\nnodet}\prod_{i=1}^{n_\sched} \left(\frac{1}{n_\sched}p_{ia}\right)^{n_{ia}} \\
+&= \frac{\ntot!}{\nnodet! \prod_{i=1}^{\nsched} n_{ia}!}  \pnodet^{\nnodet}\prod_{i=1}^{\nsched} \left(\frac{1}{\nsched}p_{ia}\right)^{n_{ia}} \\
 &\propto \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i \in \set{D}} p_{ia} \label{perf-test:eq:augmented-likelihood}
@@
 &= p(\theta) \sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) p(\na \mid \theta, \ntot) \\
-&\propto p(\theta) \sum_{\ntot=\ndet}^\infty \left( p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i=\ndet}^{\ndet} p_{ia} \right) \\
-&= p(\theta) \left( \prod_{i=\ndet}^{\ndet} p_{ia} \right) \left( \sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \right) \label{perf-test:eq:full-posterior}.
+&\propto p(\theta) \sum_{\ntot=\ndet}^\infty \left( p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i \in \set{D}} p_{ia} \right) \\
+&= p(\theta) \left( \prod_{i \in \set{D}} p_{ia} \right) \left( \sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \right) \label{perf-test:eq:full-posterior}.
 \end{align}
@@
 p(\theta \mid \na)
-&\propto p(\theta) \left( \prod_{i=\ndet}^{\ndet} p_{ia} \right) (r + \mu (1- \pnodet))^{-(r+\ndet)} \label{perf-test:eq:full-posterior-simplified}.
+&\propto p(\theta) \left( \prod_{i \in \set{D}} p_{ia} \right) (r + \mu (1- \pnodet))^{-(r+\ndet)} \label{perf-test:eq:full-posterior-simplified}.
 \end{align}
@@
 Next, I derive $p_{ia}$.
-By definition (see~\cpageref{perf-test:def:admissible}), we have:
+By definition (see~\cpageref{perf-test:def:admissible}), we have, for any $i \in \set{D}$:
 \begin{align}
 p_{ia} &= \prob((b, e) \in \alpha_i \mid \theta) \\
-\alpha_i &= \{ (b, e) : l_i^{(b)} \leq b \leq r_i^{(b)} \wedge l_i^{(e)} \leq e \leq r_i^{(e)}\}.
+\alpha_i &= \{ (b, e) : l_{j(i)}^{(b)} \leq b \leq r_{j(i)}^{(b)} \wedge l_{j(i)}^{(e)} \leq e \leq r_{j(i)}^{(e)}\}.
 \intertext{Suppressing the conditioning on $\theta$, this gives:}
 p_{ia}
-&= \prob \left( l_i^{(b)} \leq B_i \leq r_i^{(b)}, l_i^{(e)} \leq E_i \leq r_i^{(e)} \right) \\
-&= \prob \left( l_i^{(e)} \leq E_i \leq r_i^{(e)} \mid l_i^{(b)} \leq B_i \leq r_i^{(b)} \right) \prob \left( l_i^{(b)} \leq B_i \leq r_i^{(b)} \right) \\
-&=\sum_{b = l_i^{(l)}}^{r_i^{(b)}} \prob \left( l_i^{(e)} \leq E_i \leq r_i^{(e)} \mid B_i = b \right) \prob \left(B_i = b \right) \\
-&=\sum_{b = l_i^{(b)}}^{r_i^{(b)}} \prob \left( l_i^{(e)} - b + 1 \leq D_i \leq r_i^{(e)} - b + 1 \right) \prob \left(B_i = b \right) &\text{by def of $D_i$} \\
-&=\sum_{b = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b + 1) - S_\theta(r_i^{(e)} - b + 1) \right) \prob \left(B_i = b \right) &\text{by def of $S_\theta$} \\
-&\propto \sum_{b = l_i^{(b)}}^{r_i^{(b)}} \left( S_\theta(l_i^{(e)} - b + 1) - S_\theta(r_i^{(e)} - b + 1) \right)
+&= \prob \left( l_{j(i)}^{(b)} \leq B_{j(i)} \leq r_{j(i)}^{(b)}, l_{j(i)}^{(e)} \leq E_{j(i)} \leq r_{j(i)}^{(e)} \right) \\
+&= \prob \left( l_{j(i)}^{(e)} \leq E_{j(i)} \leq r_{j(i)}^{(e)} \mid l_{j(i)}^{(b)} \leq B_{j(i)} \leq r_{j(i)}^{(b)} \right) \prob \left( l_{j(i)}^{(b)} \leq B_{j(i)} \leq r_{j(i)}^{(b)} \right) \\
+&=\sum_{b = l_{j(i)}^{(l)}}^{r_{j(i)}^{(b)}} \prob \left( l_{j(i)}^{(e)} \leq E_{j(i)} \leq r_{j(i)}^{(e)} \mid B_{j(i)} = b \right) \prob \left(B_{j(i)} = b \right) \\
+&=\sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \prob \left( l_{j(i)}^{(e)} - b + 1 \leq D_{j(i)} \leq r_{j(i)}^{(e)} - b + 1 \right) \prob \left(B_{j(i)} = b \right) &\text{by def of $D_{j(i)}$} \\
+&=\sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_\theta(l_{j(i)}^{(e)} - b + 1) - S_\theta(r_{j(i)}^{(e)} - b + 1) \right) \prob \left(B_{j(i)} = b \right) &\text{by def of $S_\theta$} \\
+&\propto \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_\theta(l_{j(i)}^{(e)} - b + 1) - S_\theta(r_{j(i)}^{(e)} - b + 1) \right)
 \end{align}
@@
 
-The next step is to derive $1 - p_u = 1 - \frac{1}{N} \sum_{i=1}^{\Ncis} p_{iu} = \frac{1}{N} \sum_{i=1}^{\Ncis} (1 - p_{iu})$.
-Therefore, the crucial component is $1 - p_{iu} = 1 - \prob((B_i, E_i) \in \Omega^C_i \mid \theta)$, one minus the probability that episode $i$ was undetected.
+The next step is to derive $1 - p_u = 1 - \frac{1}{N} \sum_{i=1}^{\nsched} p_{iu} = \frac{1}{N} \sum_{i=1}^{\nsched} (1 - p_{iu})$.
+Therefore, the crucial component is $1 - p_{iu} = 1 - \prob((B_{j(i)}, E_{j(i)}) \in \Omega^C_i \mid \theta)$, one minus the probability that episode $i$ was undetected.
 An episode is undetected if and only if no tests are performed during the episode or if there was no negative test prior to the episode.
@@
 1 - p_{iu}
-&= 1 - \prob((B_i, E_i) \in \Omega^C_i) \\
-&= 1 - \prob( \tau_{\sched_i}(B_i) + B_i > E_i \vee B_i \leq \min(\sched_i) ) \\
-&= 1 - \prob( \tau_{\sched_i}(B_i) + 1 > D_i , B_i > \min(\sched_i) )  - \prob( B_i \leq \min(\sched_i) ) \\
-&= 1 - \sum_{b={\min\sched_i} + 1}^T \prob( \tau_{\sched_i}(b) + 1> D_i) \prob(B_i = b) - \sum_{t=1}^{\min\sched_i} \prob(B_i = b)\\
+&= 1 - \prob((B_{j(i)}, E_{j(i)}) \in \Omega^C_i) \\
+&= 1 - \prob( \tau_{\sched_i}(B_{j(i)}) + B_{j(i)} > E_{j(i)} \vee B_{j(i)} \leq \min(\sched_i) ) \\
+&= 1 - \prob( \tau_{\sched_i}(B_{j(i)}) + 1 > D_i , B_{j(i)} > \min(\sched_i) )  - \prob( B_{j(i)} \leq \min(\sched_i) ) \\
+&= 1 - \sum_{b={\min\sched_i} + 1}^T \prob( \tau_{\sched_i}(b) + 1> D_i) \prob(B_{j(i)} = b) - \sum_{t=1}^{\min\sched_i} \prob(B_{j(i)} = b)\\
 &= 1 - \frac{1}{T} \sum_{b=\min(\sched_i)+1}^T (1 - S_\theta(\tau_{\sched_i}(b) + 1)) - \frac{1}{T} \min\sched_i \\
@@
     \item Extract the test schedules for each individual who had at least one test during the period of interest.
-    \item Draw an episode start time, $b_i$ for each individual uniformly at random between 2nd July 2020 (100 days before the period where a detected episode would be included) and 6th December 2020 (the end of this period).
-    \item Draw a duration of episode for each individual, $d_i$, based on the distribution described below. Then calculate the end of their infection episode, $e_i = b_i + d_i - 1$.
-    \item Simulate the test results based on the test schedule, $b_i$, and $e_i$. Tests between $b_i$ and $e_i$ (inclusive) are positive, all other tests are negative.
+    \item Draw an episode start time, $b_{j(i)}$ for each individual uniformly at random between 2nd July 2020 (100 days before the period where a detected episode would be included) and 6th December 2020 (the end of this period).
+    \item Draw a duration of episode for each individual, $d_i$, based on the distribution described below. Then calculate the end of their infection episode, $e_{j(i)} = b_{j(i)} + d_i - 1$.
+    \item Simulate the test results based on the test schedule, $b_{j(i)}$, and $e_{j(i)}$. Tests between $b_{j(i)}$ and $e_{j(i)}$ (inclusive) are positive, all other tests are negative.
     \item Discard episodes where there are no positive tests (\ie undetected episodes) and then apply the same inclusions/exclusions as in \cref{perf-test:sec:problem}.
@@
 This is needed because in step 2 the entire cohort was infected, while in the real study only a (unknown) portion is infected.
-    \item For this final set of episodes, calculate $(l_i^{(b)}, r_i^{b}, l_i^{(e)}, r_i^{(e)})$ by taking the day after the last negative prior to any positives, the first positive, the last positive, and the day before the negative following the last positive respectively.
+    \item For this final set of episodes, calculate $(l_j^{(b)}, r_j^{b}, l_j^{(e)}, r_j^{(e)})$ by taking the day after the last negative prior to any positives, the first positive, the last positive, and the day before the negative following the last positive respectively.
 \end{enumerate}

### 2024-03-11 symbols.tex
fix subscripts for survival analysis

@@
     \item $r\inform$ the dispersion of the informative prior on $\ntot$
-    \item $y_i(t)$ the test result for infection episode $i$ at time $t$
-    \item $\vec{y_i}$ the test results associated with infection episode $i$
+    \item $y_{j(i)}(t)$ the test result for infection episode $i$ at time $t$
+    \item $\vec{y_{j(i)}}$ the test results associated with infection episode $j(i)$ with test schedule $\sched_i$
     \item $\eta$ the additional probability of missing an infection episode due to false negatives

### 2024-03-12 transmission.tex
Overleaf comments

@@
 Estimation of prevalence from the CIS data was carried out using MRP throughout the pandemic (see \cref{E-biology-data:sec:MRP}).
-In this section, I propose extending this process by transforming the posterior distribution of prevalence in each stratum to a posterior distribution of incidence before post-stratifying.
+In this section, I propose extending this process by transforming the posterior distribution of prevalence in each stratum to a posterior distribution of incidence before the poststratification step.
 
@@
 
-An alternative approach would be to estimate incidence proportion directly from the RT-PCR test results.
-
+An alternative would be an inference procedure that estimates incidence proportion, fit directly to the RT-PCR test results.
 Denote the quantity to be inferred (which could be either incidence proportion or prevalence) at time $t$ in strata $i$ as $\pi_{it}$.
-The strata will represent at least age group and region, although including ethnicity and sex is also included due to non-response in the survey (see \cref{E-intro:sec:cis-strata}).
+The strata are defined by at least age group, region, ethnicity, and sex due to differential non-response in the survey (see \cref{E-intro:sec:cis-strata}).
 For many combinations of stratum $i$ and day $t$, the number of positives will be small, often fewer than 5; therefore some form of borrowing between time and strata is required.
@@
     \item Borrow information across the different covariates.
-        The borrowing should penalise away from changes in the distribution of incidence/prevalence across strata in the same geographic region being constant.
+        The borrowing should penalise changes in the distribution of incidence/prevalence across strata in the same geographic region being constant.
         Standard stratified, mechanistic transmission model (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) predict this, if behavioural changes are independent of strata and there is negligible impact of immunity.

### 2024-03-12 introduction.tex

@@
 Typically, the core of an observation model is a \emph{convolution equation} or an approximation of it.
-This equation links an observation at time $t$, $y_t$, to the incidence prior to time $t$.
+This equation links the expectation of an observation at time $t$, $y_t$, conditional on the incidence prior to time $t$.
 In discrete time, it is:
@@
 
-\subsection{Data for estimating incidence and transmission} \label{intro:sec:data-for-estimating}
+In this case we needed...
+
+Urgency of situtation (cf HIV/AIDS) - rate of harm .
+
+Needed rapid use of ideas
+
+\subsection{Surveillance data} \label{intro:sec:data-for-estimating}
 
@@
 
-\subsection{Software and data availability}
-
-In order to best fulfill this thesis's primary aim, the methodology developed should be easy for others to use.
-Therefore, the code for the analyses in this thesis is available online.
-An index to the various repositories is available at\todo{insert GitHub link}.
-Two elements of the code have been structured as R packages to enable their application to future contexts.
-
-First, the framework to simulate the CIS (used in \cref{E-imperf-test:sec:simulate,E-perf-test:sec:simulation-study}).
-The package enables customisation of various features of the study, for example the testing regime of the participants, the sensitivity of the tests, and the duration of positivity.
-This package would be particularly useful in answering questions regarding the design of future prevalence surveys, for example, how frequently should individuals be tested, or how many individuals should be tested.
-
-Second, the survival analysis framework used to infer the duration of positivity in \cref{E-perf-test,E-imperf-test}.
-The framework allows for all the different analyses performed in these chapters to be specified by the user in a simple interface.
-Furthermore, adding new analyses, for example varying the priors or the inclusion of covariates, is straightforward.
-
-Where possible, aggregated, non-disclosive versions of the datasets are available alongside the code.
-The original ATACCC data used is available alongside the code.
-The original CIS data can only be accessed by accredited researchers within the ONS's SRS (Secure Research Service); see \url{https://www.ons.gov.uk/aboutus/whatwedo/statistics/requestingstatistics/secureresearchservice} for details.
-
-Almost all code, excluding dependencies, was written from scratch for these analyses.
-The exception is the Markov chain Monte Carlo sampler used in \cref{E-SEIR}, which was written by Sanmitra Ghosh for the article \textcite{ghoshApproximate}.
-
 \subsection{Methodological contributions}
@@
 
+\subsection{Software and data availability}
+
+In order to best fulfill this thesis's primary aim, the methodology developed should be easy for others to use.
+Therefore, the code for the analyses in this thesis is available online.
+An index to the various repositories is available at\todo{insert GitHub link}.
+Two elements of the code have been structured as R packages to enable their application to future contexts.
+
+First, the framework to simulate the CIS (used in \cref{E-imperf-test:sec:simulate,E-perf-test:sec:simulation-study}).
+The package enables customisation of various features of the study, for example the testing regime of the participants, the sensitivity of the tests, and the duration of positivity.
+This package would be particularly useful in answering questions regarding the design of future prevalence surveys, for example, how frequently should individuals be tested, or how many individuals should be tested.
+
+Second, the survival analysis framework used to infer the duration of positivity in \cref{E-perf-test,E-imperf-test}.
+The framework allows for all the different analyses performed in these chapters to be specified by the user in a simple interface.
+Furthermore, adding new analyses, for example varying the priors or the inclusion of covariates, is straightforward.
+
+Where possible, aggregated, non-disclosive versions of the datasets are available alongside the code.
+The original ATACCC data used is available alongside the code.
+The original CIS data can only be accessed by accredited researchers within the ONS's SRS (Secure Research Service); see \url{https://www.ons.gov.uk/aboutus/whatwedo/statistics/requestingstatistics/secureresearchservice} for details.
+
+Almost all code, excluding dependencies, was written from scratch for these analyses.
+The exception is the Markov chain Monte Carlo sampler used in \cref{E-SEIR}, which was written by Sanmitra Ghosh for the article \textcite{ghoshApproximate}.
+
 \subsection{Structure}

### 2024-03-12 MCMC-appendix.tex
Add bits previously in intro to MCMC appendix

@@
 
+MCMC (Markov chain Monte Carlo) is a class of algorithms for sampling from a distribution where the normalising constant for the distribution is unknown.
+I use it to sample a set of parameters, say $\psi$ (which may be a scalar or vector), from a posterior distribution $p(\psi \mid y)$ where $y$ is the data (again, scalar or vector).
+MCMC algorithms produce a sample $\psi^1, \psi^2, \dots, \psi^M$ from $p(\psi \mid y)$ by constructing a Markov chain with $p(\psi \mid y)$ as its stationary distribution~\autocite[275]{gelmanBDA}.
+It can be proved that as $M\to\infty$, the sample tends towards a sample from the stationary distribution, \ie the target posterior.
 This appendix outlines the basic theory and implementation of Markov chain Monte Carlo (MCMC) methods.
@@
 
-\section{Improving MCMC efficiency}
+\section{Practical issues}
+
+Two practical issues arise when applying MCMC: \emph{convergence} and \emph{stability}~\autocite[72]{lunnBUGS}.
+
+\subsection{Convergence} \label{MCMC:sec:convergence}
+
+Convergence is how close a MCMC chain's samples are to its stationary distribution.
+Early draws from the Markov chain reflect the choice of the starting point, $\psi^0$, rather than the stationary distribution that the chain will eventually converge to~\autocite[282]{gelmanBDA}.
+%The Rhat statistic is defined for a scalar parameter $\psi$.
+%Let $\psi^{(ij)}$ denote the $i$th (of $M$) sample from the $j$th chain (of $N$).
+%Let $\bar{\psi}^j$ denote the mean value of the samples of $\psi$ from the $j$th chain, $\bar{\psi}^j = \frac{1}{M} \sum_{i=1}^M \psi^{(ij)}$.
+%Let $\bar{\psi}$ denote the mean value of the samples of $\psi$ from all chains, $\bar{\psi} = \frac{1}{N} \sum_{j=1}^N \bar{\psi}^j$.
+%Then the between-chain variance is $B = \frac{M}{N-1} \sum_{j=1}^N (\bar{\psi}^j - \bar{\psi})^2$.
+%The within-chain variance is $W = \frac{1}{N} \sum_{j=1}^N \frac{1}{M-1} \sum_{i=1}^M (\psi^{(ij)} - \bar{\psi}^j)^2$.
+
+The simplest technique for assessing convergence is graphically, using traceplots~\autocite[81]{brooksMCMCNotes}.
+Traceplots show the value of $\psi^i$ against $i$.
+If the chain has converged, then the traceplot will show a random walk around the stationary distribution.
+This can be improved by using multiple chains with different starting points~\autocite[283]{gelmanBDA}.
+If all chains have converged, then they will be in the same location.
+
+A more complex measure of convergence when running multiple MCMC chains is comparing the within-chain variance to the between-chains variance.
+The principle here is that if all chains have converged then they are sampling from the same distribution.
+Therefore, the within-chain and between-chain variances are the same~\autocite[82]{brooksMCMCNotes}.
+
+Various metrics quantify this, I make use of the improved Rhat statistic~\autocite{vehtariRhat}.
+A Rhat of 1 means that the within-chain and between-chain variance are equal, informally that convergence is perfect, as far as can be measured by this technique.
+Various rules-of-thumb exist for what should be considered an acceptable Rhat.
+Historically, 1.1 has commonly been recommended, although more recent work suggests that 1.01 or 1.05 is more appropriate~\autocite{vehtariRhat}.
+
+\subsection{Stability}
+
+The stability of estimates considers the error arising from only using a finite sample of the distribution and the samples being autocorrelated.
+Autocorrelation means that estimates using the sample have higher variance than if it were an iid sample from the posterior~\autocite[286]{gelmanBDA}.
+The \emph{effective sample size} corrects for the autocorrelation.
+The effective sample size can be interpreted as the number of iid samples that would have the same variance as the MCMC sample~\autocites[286]{gelmanBDA}{vehtariRhat}.
+Various techniques have been proposed for computing the effective sample size, I will follow the approach of \textcite{vehtariRhat}, implemented in RStan~\autocite{RStan-2-32-3}.
+
+\section{Improving MCMC efficiency} \label{MCMC:sec:improving}
 
@@
 
-\section{Convergence}
-
-The simplest technique for assessing convergence is graphically, using traceplots~\autocite[81]{brooksMCMCNotes}.
-Traceplots show the value of $\psi^i$ against $i$.
-If the chain has converged, then the traceplot will show a random walk around the stationary distribution.
-This can be improved by using multiple chains with different starting points~\autocite[283]{gelmanBDA}.
-If all chains have converged, then they will be in the same location.
-
-A more complex measure of convergence when running multiple MCMC chains is comparing the within-chain variance to the between-chains variance.
-This is the basis of the Rhat statistic used throughout this thesis (see \cref{E-intro:sec:MCMC}).
-The principle here is that if all chains have converged then they are sampling from the same distribution.
-Therefore, the within-chain and between-chain variances are the same~\autocite[82]{brooksMCMCNotes}.
-
 \end{document}

### 2024-03-12 incidence-prevalence.tex
restructure sec 7.1

@@
 Considering multiple values of $t$ makes \cref{inc-prev:eq:EPt} into a system of simultaneous equations.
-The system can be written in the form $\E(\vec{P} \mid \vec{Z}) = \matr{S} \vec{Z}$ where $\matr{S}$ is a lower-triangular matrix formed from $S$:
+Therefore, the system can be written in matrix form:
 \begin{align}
-    (\matr{S})_{i,j} = \begin{cases}
+  \E(\vec{P} \mid \vec{Z}) = \matr{S} \vec{Z}. \label{inc-prev:eq:EPt-matrix} \\
+\intertext{$\matr{S}$ is a lower-triangular matrix formed from $S$:}
+    (\matr{S})_{i,j} &= \begin{cases}
         S(i - j + 1) & i - \dmax - 1 \leq j \leq i \\
         0 & \text{otherwise.}
-    \end{cases}
+    \end{cases} \label{inc-prev:eq:Smatrix}
 \end{align}

### 2024-03-12 transmission.tex
restructure sec 7.1

@@
 In this section, I will apply the theory introduced in \cref{E-inc-prev} to estimate incidence.
-% First, I explain the method I will apply (\cref{backcalc:sec:methods}) and then present the results (\cref{backcalc:sec:results}).
-% I discuss the results in \cref{backcalc:sec:discussion} and conclude in \cref{backcalc:sec:conclustion}.
+I view the CIS as a series of daily, cross-sectional prevalence surveys.
+This is the same approach as previously used to estimate prevalence from the CIS (see \cref{E-biology-data:sec:MRP}).
 
-\subsection{Methods} \label{backcalc:sec:methods}
+An important consideration in this section is the non-response bias.
+I adopt the poststratification method introduced in \cref{E-biology-data:sec:MRP} to correct for this bias.
+Therefore, incidence must be estimated in each stratum before being poststratified to the target population and subpopulations of interest.
+The strata are defined by at least age group, region, ethnicity, and sex due to differential non-response in the survey (see \cref{E-intro:sec:cis}).
+
+\subsection{Approach} \label{backcalc:sec:approach}
 
@@
 This relationship between incidence proportion and prevalence means that once one is inferred, the other is calculated via a deterministic function.
-Therefore, a statistical methodology for estimating one of these quantities is required.
+This gives two possible approaches for estimating incidence from prevalence data.
 
-Estimation of prevalence from the CIS data was carried out using MRP throughout the pandemic (see \cref{E-biology-data:sec:MRP}).
-In this section, I propose extending this process by transforming the posterior distribution of prevalence in each stratum to a posterior distribution of incidence before the poststratification step.
+\subsubsection{Possible approaches}
 
-% I keep the same logistic regression model for predicting prevalence, then transform its posterior distribution of prevalence to a posterior distribution of incidence.
-This results in a two-step process: first, the same inference process described in \cref{E-biology-data:sec:MRP} is used to estimate prevalence; second, the prevalence is transformed to incidence.
-This is required to fulfil several desirable features of the model.
+The first approach, which I call the \emph{incidence-inferred} approach, is a single statistical model where incidence is inferred, with prevalence as a derived quantity.
+The basis of the statistical model, in a Bayesian framework, would be as follows.
+\begin{align}
+    \vec{Z} \mid \vec{\phi} &\dist g_Z(\vec{\phi}) \\
+    P_{i,t} &= \matr{S} \vec{Z}_i \\
+    Y_{i,t} \mid \vec{P} &\dist \BB(n_{i,t}, P_{i,t})
+\end{align}
+where $\vec{Z}$ is a vector of incidence proportion for each strata and day defined by some model $g$ with parameters $\vec{\phi}$; $P_{i,t}$ is the prevalence in stratum $i$ on day $t$; $\matr{S}$ is the survival matrix defined in \cref{E-inc-prev:eq:Smatrix}; $Y_{i,t}$ is the number of positive tests in stratum $i$ on day $t$; and $n_{i,t}$ is the number of tests in stratum $i$ on day $t$.
+The model $g_Z$ would be specified to share information across time and/or strata.
+This both helps stabilise the estimation (see\todo{ref wherever I end up explaining backcalculation stability issues}) and allows borrowing of information across strata.
+Borrowing information is required because some strata and time combinations have few positives (fewer than five per day).
+The relationship between $P$ and $Z$ is from \cref{E-inc-prev:eq:EPt-matrix}, although dividing both sides by $\Npop$ to have the values in terms of incidence proportion and prevalence rather than incidence and number of prevalent individuals.
 
-An alternative would be an inference procedure that estimates incidence proportion, fit directly to the RT-PCR test results.
-Denote the quantity to be inferred (which could be either incidence proportion or prevalence) at time $t$ in strata $i$ as $\pi_{it}$.
-The strata are defined by at least age group, region, ethnicity, and sex due to differential non-response in the survey (see \cref{E-intro:sec:cis-strata}).
-For many combinations of stratum $i$ and day $t$, the number of positives will be small, often fewer than 5; therefore some form of borrowing between time and strata is required.
-The statistical model and inference procedure should have the following properties.
+The second approach, which I call the \emph{prevalence-inferred} approach takes a two-step approach.
+First, it uses the statistical model from \cref{E-biology-data:sec:MRP} to estimate prevalence.
+Then, it transforms the posterior distribution of prevalence to a posterior distribution of incidence using the forward substitution as in \cref{E-inc-prev:eq:forward-substitute}.
+At a high-level, the statistical model is:
+\begin{align}
+    \vec{P} \sim \vec{\phi} &\dist g_P(\vec{\phi}) \\
+    Y_{i,t} &\dist \BB(n_{i,t}, P_{i,t})
+\end{align}
+where $g_P$ is analogous to $g_Z$ in providing a joint prior for the prevalence in each strata and day combination.
+
+\subsubsection{Desirable properties}
+
+The incidence-inferred approach is more parsimonious in being a single step procedure.
+However, it poses several pragmatic challenges which means I favour the prevalence-inferred approach.
+Here I explain that further developments, beyond the scope of this thesis, would be required to use the incidence-inferred approach with a suitable $\pi_Z$ within the computational constraints of the SRS.
+
+Denote the quantity to be inferred at time $t$ in strata $i$ as $\pi_{i,t}$.
+That is, under the incidence-inferred approach, $\pi_{i,t}$ is the incidence proportion, and under the prevalence-inferred approach, $\pi_{i,t}$ is the prevalence.
+
+The chosen approach should have the following properties.
 \begin{enumerate}
@@
         % While incidence proportion can theoretically be greater than 1, this would imply that some individuals were infected multiple times in a day which is not biologically plausible.
-    \item Have sufficient parameters to incorporate all necessary covariates to correct for non-response bias (see \cref{E-biology-data:sec:cis}).
+    \item Have sufficient parameters to incorporate all necessary covariates to correct for non-response bias.
     \item Borrow information across the different covariates.
         The borrowing should penalise changes in the distribution of incidence/prevalence across strata in the same geographic region being constant.
-        Standard stratified, mechanistic transmission model (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) predict this, if behavioural changes are independent of strata and there is negligible impact of immunity.
-    \item Be computationally efficient.
+        A constant distribution is predicted by the theoretical study of epidemics, for example the dynamics of stratified, mechanistic transmission models (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) over short periods of time with negligible behavioural and immunity changes.
+    \item Be computationally efficient
         The model will be fit to granular data and therefore needs to be run within the SRS, with its computational limitations (see \cref{E-biology-data:sec:SRS}).
@@
 
-Properties 1 and 2 suggest a model of the form $\logit(\pi_{it}) = g(i, t)$, where $g$ is a function with changes in the first derivative penalised (\ie changes away from linear are penalised).
-The $\logit$ link function both limits $\pi_{it}$ to the interval $[0, 1]$, and logit-linear growth approximates exponential growth for small $\pi_{it}$~\autocite{ealesAppropriately}.
+Properties 1 and 2 suggest a model of the form $\logit(\pi_{i,t}) = g(i, t)$, where $g$ is a function with changes in the first derivative penalised (\ie changes away from linear are penalised).
+The $\logit$ link function both limits $\pi_{i,t}$ to the interval $[0, 1]$, and logit-linear growth approximates exponential growth for small $\pi_{i,t}$~\autocite{ealesAppropriately}.
 Combined with property 4, the form of $g$ the null model is $g(i, t) = a_0 + a_t t + f(i)$ where $a_0$ and $a_t$ are model parameters and $f$ is a function to be specified; that is, changes away from this model should be penalised.
@@
 
-\todo[inline]{Is it OK that $x$ and $y$ in the following two paragraphs don't align with the rest of the thesis chapter?
-Should I use alterantive notation - but then different to standard in the INLA papers.}
 INLA-based inference can only be applied to LGMs (\emph{Latent Gaussian Models})~\autocite{rueINLAseminar,rueINLA,martinoINLAtutorial}; see \cref{E-transmission:sec:INLA} for further details on INLA.
-An LGM is defined in terms of a vector of hyperparameters, $\vec{\varphi}$; a deterministic transformation of the hyperparameters into covariance matrix, $\matr{\Sigma}(\vec{\varphi})$; a vector of latent variables, $\vec{x}$; a vector of $n$ data points, $\vec{y} = [y_1, \dots, y_n]^T$; and $n$ linear predictors $\vec{\eta} = [\eta_1, \dots, \eta_n]^T$.
-To be an LGM, $\vec{x}$ must be a latent Gaussian field, that is $\vec{x} \mid \vec{\varphi} \dist \N(\vec{0}, \matr{\Sigma}(\vec{\varphi}))$; and the likelihood must have the conditional independence property $p(\vec{y} \mid \vec{\varphi}, \vec{x}) = \prod_j p(y_j \mid \eta_j(\vec{x}), \vec{\varphi})$.
 
-In this context, each $y_j$ corresponds to the observed RT-PCR results in one stratum on one day.
-Hence, $\eta_{j}$ must determine the distribution of $y_{j}$ (up to some hyperparameters); where $j$ uniquely identifies a stratum $i$ and day $t$.
+An LGM is defined in terms of a vector of hyperparameters, $\vec{\varphi}$; a deterministic transformation of the hyperparameters into covariance matrix, $\matr{\Sigma}(\vec{\varphi})$; a vector of latent variables, $\vec{z}$; a vector of $n$ data points, $\vec{w} = [w_1, \dots, w_n]^T$; and $n$ linear predictors $\vec{\eta} = [\eta_1, \dots, \eta_n]^T$.
+To be an LGM, $\vec{z}$ must be a latent Gaussian field, that is $\vec{z} \mid \vec{\varphi} \dist \N(\vec{0}, \matr{\Sigma}(\vec{\varphi}))$; and the likelihood must have the conditional independence property $p(\vec{w} \mid \vec{\varphi}, \vec{z}) = \prod_j p(w_j \mid \eta_j(\vec{z}), \vec{\varphi})$.
+
+In this context, each $w_j$ corresponds to the observed RT-PCR results in one stratum on one day.
+Hence, $\eta_{j}$ must determine the distribution of $w_{j}$ (up to some hyperparameters); where $j$ uniquely identifies a stratum $i$ and day $t$.
+
 In practice, this means that $\eta_j$ is a one-to-one transformation of the population prevalence in stratum $i$ on day $t$.
-If $\eta_j = \logit(\pi_{it})$, where $\pi$ is prevalence, then a model of $\eta_j = g(i, t)$, as before, is a LGM.
+If $\eta_j = \logit(\pi_{i,t})$, where $\pi$ is prevalence, then a model of $\eta_j = g(i, t)$, as before, is a LGM.
 However, if $\pi$ is instead representing incidence proportion, then no such linear relationship between $\eta_j$ and $g$ exists.
-In particular, if $\eta_j$ is the prevalence, then $\eta_{it} = \sum_{k=-\infty}^0 \pi_{ik} S(k) = \sum_{k=-\infty}^0 \expit(g(i, k)) S(k)$.
-As $g(i, k)$ is linear in the model parameters, $\vec{x}$, then $\eta_{it}$ is not, violating the requirements of a LGM.
-Therefore, the application of the INLA approximation, while satisfying the other desirable properties, requires that the regression model predicts RT-PCR prevalence, not incidence proportion.
+In particular, if $\eta_j$ is the prevalence, then:
+\begin{align}
+    \eta_{i,t}
+    &= \sum_{k=-\infty}^0 \pi_{ik} S(k) \\
+    &= \sum_{k=-\infty}^0 \expit(g(i, k)) S(k).
+\end{align}
+As $g(i, k)$ is linear in the model parameters, $\vec{z}$, then $\eta_{i,t}$ is not, violating the requirements of a LGM.
 
-Next, I explain how I extended the MRP model introduced in \cref{E-biology-data:sec:MRP} to calculate incidence.
-I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution (using the procedure described in \cref{transmission:sec:INLA:posterior}).
+Therefore, using the INLA approximation while satisfying the other desirable properties requires that $g$ predicts prevalence, not incidence proportion.
+Hence, I adopt the prevalence-inferred approach.
+
+\subsection{Methods} \label{backcalc:sec:methods}
+
+Now I explain how I extended the MRP model introduced in \cref{E-biology-data:sec:MRP} to implement the prevalence-inferred approach.
+
+First, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution (using the procedure described in \cref{transmission:sec:INLA:posterior}).
 For each stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
@@
 The second set of assumptions are in specifying the statistical model used to estimate prevalence, for example the form of the smoothing used.
+In this context, that means the prior $\pi_P$ places non-zero mass on prevalence values which imply negative incidence under the other model assumptions.
 
@@
 Alternative inference techniques could allow more powerful modelling approaches, if they allowed a broader set of models than LGMs.
-In particular, the incidence proportion could be directly modelled rather than the prevalence.
+In particular, the incidence-inferred approach could be adopted.
 This could allow more interpretable priors to be used because the infection process is the process that corresponds to biological and behavioural processes.
@@
 Various alternative inference techniques could be considered.
-If the infection process is considered a Poisson process (see \cref{E-inc-prev:sec:poisson-process}), then, conditional on the process's intensity, there is independence between individuals' infection status.
-Therefore, the incidence proportion and prevalence can both be viewed as the mean of many Bernoulli variables, meaning that the Central Limit Theorem suggests they are approximately normally distributed.
+If the infection process is considered a Poisson process (defined in \cref{E-inc-prev:sec:poisson-process}), then, conditional on the process's intensity, there is independence between individuals' infection status.
+The conditional independence implies the incidence proportion and prevalence are the mean of many independent Bernoulli variables.
+Therefore, the Central Limit Theorem implies they are approximately normally distributed.
 Exploiting this property could enable efficient approximations to the posterior.
+For example, a Laplace approximation or variational inference.
+Further work, to check the quality of these approximations, is required.
 
@@
 There is a tendency for the coverage to be slightly lower than the nominal probability, suggesting that the credible intervals are slightly too narrow.
-This is a common issue with random walk MCMC (see \cref{E-inc-prev:sec:MCMC}).
+This is a common issue with random walk MCMC (see \cref{E-MCMC:sec:improving}).
 The same issue arises with the coverage for derived quantities incidence and prevalence (see \cref{SEIR:fig:sim-inc-prev}).
@@
 
-All regions' MCMC chains converged, assessed using the Rhat statistic and ESS (as defined in \cref{E-intro:sec:MCMC}).
+All regions' MCMC chains converged, assessed using the Rhat statistic and ESS (as defined in \cref{E-MCMC:sec:convergence}).
 The highest Rhat was 1.01 for $\sigma_\epsilon$ in London.

### 2024-03-12 cis-imperfect-testing.tex
fix mistake in text

@@
 
-Next, I considered the consequence of $\psens$ being misspecified, using the vague prior; the better performing prior in the correctly specified case.
+Next, I considered the consequence of $\psens$ being misspecified, using the model combination prior; the better performing prior in the correctly specified case.
 If the test sensitivity is misspecified, that is the assumed value for $p_\text{sens}$ in \cref{imperf-test:eq:pit-prime,imperf-test:eq:pit-prime} is different to the value used in the simulation, then the survival time will be biased (see \cref{imperf-test:fig:misspecified-test-sensitivity}).

### 2024-03-12 paper-surv-analysis.tex
initial go at paper on survival analysis

--- 2024-03-11 cis-perfect-testing.tex
@@
-\documentclass[thesis.tex]{subfiles}
+\documentclass[12pt, letterpaper]{article} % Sets the font size to 12pt and paper size to letter
+
+\usepackage[margin=1in]{geometry} % Sets the margins to 1 inch all around
+\usepackage{setspace} % Allows for double spacing and setting line spacing
+\setlength{\marginparwidth}{2cm}
+
+\usepackage{graphicx}
+\usepackage[utf8]{inputenc}
+
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{floatpag}
+\usepackage{bm}
+\usepackage{algorithm2e}
+\usepackage{csquotes}
+\usepackage[T1]{fontenc}
+\usepackage{hyperref}
+\usepackage{textcomp} % provide symbols
+\usepackage{todonotes}
+\usepackage{xr}
+\usepackage{pdflscape}
+\usepackage{afterpage}
+\usepackage{caption}
+\usepackage{numprint}
+% \usepackage{microtype}
+\usepackage[USenglish]{babel}
+\usepackage[nameinlink]{cleveref}
+\usepackage[backend=biber,style=apa,natbib=true]{biblatex} % Load biblatex with the desired settings
+\addbibresource{references.bib}
+\addbibresource{refs-custom.bib}
+% no URLs
+\AtEveryBibitem{
+  \clearfield{url}
+  \clearfield{doi}
+}
+
+\captionsetup{font=footnotesize} % Adjust the caption font size as necessary for readability
+
+% \usepackage{lineno} % Include the package for line numbering
+% \linenumbers % Activates line numbering for the document
+
+% Diagrams
+\usepackage{tikz}
+\usetikzlibrary{arrows,positioning,automata,fit}
+
+% Macros for common abbreviations to get the spacing right
+% See: https://stackoverflow.com/questions/3282319/correct-way-to-define-macros-etc-ie-in-latex
+\usepackage{xspace}
+\makeatletter
+\DeclareRobustCommand\onedot{\futurelet\@let@token\@onedot}
+\def\@onedot{\ifx\@let@token.\else.\null\fi\xspace}
+\def\eg{e.g\onedot} \def\Eg{\emph{E.g}\onedot}
+\def\ie{i.e\onedot} \def\Ie{\emph{I.e}\onedot}
+\def\cf{c.f\onedot} \def\Cf{\emph{C.f}\onedot}
+\def\etc{etc\onedot} \def\vs{\emph{vs}\onedot}
+\def\wrt{w.r.t\onedot} \def\dof{d.o.f\onedot}
+\def\etal{et al\onedot}
+\makeatother
+
+% Generic maths commands
+\def\reals{\mathbb{R}}
+\def\nats{\mathbb{N}}
+\def\sampSpace{\mathcal{X}}
+\def\dist{\sim}
+\DeclareMathOperator{\E}{\mathbb{E}}
+\DeclareMathOperator{\V}{\mathbb{V}}
+\DeclareMathOperator{\I}{\mathbb{I}}
+\DeclareMathOperator{\prob}{\mathbb{P}}
+\DeclareMathOperator{\p}{\pi}
+\DeclareMathOperator{\var}{\mathbb{V}}
+\DeclareMathOperator{\indicator}{\mathbb{I}}
+\DeclareMathOperator{\cov}{Cov}
+\DeclareMathOperator{\cor}{Cor}
+\DeclareMathOperator{\logit}{logit}
+\DeclareMathOperator{\Ber}{Bernoulli}
+\DeclareMathOperator{\Bin}{Binomial}
+\DeclareMathOperator{\Poi}{Poisson}
+\DeclareMathOperator{\BetaDist}{Beta}
+\DeclareMathOperator{\Exponential}{Exponential}
+\DeclareMathOperator{\NBr}{NegBin}
+\newcommand{\NBc}{\NBr_{c}}
+\newcommand{\NBs}{\NBr_{s}}
+\DeclareMathOperator{\BB}{BetaBin}
+\DeclareMathOperator{\GamDist}{Gamma}
+\DeclareMathOperator{\MN}{Multinomial}
+\DeclareMathOperator{\N}{N}
+\DeclareMathOperator{\LN}{LN}
+\DeclareMathOperator{\LKJ}{LKJ}
+\DeclareMathOperator{\expit}{expit}
+\newcommand\matr{\bm}
+\newcommand\set{\mathcal}
+\renewcommand{\vec}{\bm}
+\newcommand{\ssep}{:}
+
+% Thesis-specific maths commands
+\newcommand{\dmax}{d_\text{max}}
+\newcommand{\psens}{p_\text{sens}}
+\newcommand{\ntot}{n_\text{tot}}
+\newcommand{\ndet}{n_\text{d}}
+\newcommand{\nnodet}{n_\text{u}}
+\newcommand{\pnodet}{p_\text{u}}
+\newcommand{\Npop}{N_\text{pop}}
+\newcommand{\Ncis}{N_\text{CIS}}
+\newcommand{\ncis}{\vec{n_\text{CIS}}}
+\newcommand{\na}{\vec{n_\text{a}}}
+\newcommand{\pcis}{\vec{p_\text{CIS}}}
+\newcommand{\sched}{\mathbb{T}}
+\newcommand{\nsched}{n_{\sched}}
+\newcommand{\rate}{\mathfrak{R}}
+\newcommand{\R}{\mathcal{R}_{0}}
+\renewcommand{\Re}{\mathcal{R}_{e}(t)}
+\newcommand{\Rc}{\mathcal{R}_{c}(t)}
+\newcommand{\vlod}{v_\text{lod}}
+\newcommand{\inform}{{_{\text{inform}}}}
 
 \begin{document}
-\ifSubfilesClassLoaded{
 
-% \section{Definitions}
-% \todo[inline]{These will be defined in the introductory chapters, but are here so that this draft is understandable}
-% \begin{itemize}
-% \item
-%   We assume that, for all episodes $i$, $D_i$ is iid and independent
-%   of the time of the infection. Define the survival function
-%   $\prob(D_i \geq t \mid B_{j(i)} = b, \theta) = \prob(D_i \geq t \mid \theta) = S_\theta(t)$,
-%   where $\theta$ are the parameters controlling the survival
-%   distribution (the discussion in this document is valid regardless of
-%   the model specified for $S_\theta$ and hence we consider $\theta$
-%   as an arbitrary vector of parameters).
-% \item
-%   The beginning of episode $i$ is known to occur in the interval
-%   $[l_{j(i)}^{(b)}, r_{j(i)}^{(b)}]$, and similarly for the end of the infection
-%   in $[l_{j(i)}^{(e)}, r_{j(i)}^{(e)}]$.
-% \end{itemize}
+\title{Estimating length of SARS-CoV-2 from doubly interval censored data with missed infections}
+\author{Blake, TBC, Birrell, De Angelis}
+% \date{\today}
+\maketitle
 
-\setcounter{chapter}{4}
-}
+\begin{abstract}
+Abstract goes here.
+\end{abstract}
 
-\chapter{Estimating duration with infrequent testing} \label{perf-test}
+The format requirements from JASA are at \url{https://www.tandfonline.com/journals/uasa20}.
+The most important parts are reproduced here.
+\begin{itemize}
+    \item Should be written with the following elements in the following order: title page; author footnote; abstract; keywords; article text (table(s); figures); acknowledgments; appendices; references
+    \item Should be no more than 35 pages, including the abstract, figures, tables, and references.
+Appendices, proofs, and other supporting material should be placed in a separate supplement file (anonymized for review)
+    \item Should contain an unstructured abstract of 200 words.
+    \item Should contain between 3 and 5 keywords. Read making your article more discoverable, including information on choosing a title and search engine optimization.
+    \item JASA requires that all papers be formatted for 8 1/2 x 11-inch paper, one side only.
+    \item 12 point, double-spaced font (which we define as 26 lines of text per page)
+    \item the average JASA paper is about 30 manuscript pages (with the above page format and including all appendices, references, tables, and figures), and it is very uncommon for papers much longer than about 35 pages to appear in JASA.
+\end{itemize}
 
-When estimating incidence from prevalence, the mean duration, not its median, is the most important feature of the duration distribution~\autocite{freemanPrevalence}.
-Duration distributions tend to be right-skewed.
-Therefore, estimates of the mean are sensitive to the tail of the duration distribution.
+Notes / thoughts on current version
+\begin{itemize}
+    \item Some of the derivations can be moved to a supplement (cut 2--3 pages).
+    \item Need to rearranage the ATACCC stuff --- add some details of what I did (thesis ch 4) and move stuff from prior info section into supplement.
+    \item Probably don't need some of the figures (maybe 5--8)?
+    \item Should the sensitivity analyses (figs 13/14 be in the main text)?
+    \item The above changes make this about the right length.
+\end{itemize}
 
-In \cref{E-ATACCC} I estimated the duration distribution from the ATACCC dataset.
-However, this study suffered from a small sample size and only 20 days of follow-up, enabling estimation of the tail.
-Therefore, model assumptions drive the size of the estimated tail.
+Table of contents, cut for final version but useful for now to compare lengths of sections.
 
-In this chapter, I develop a method to estimate the duration distribution using CIS data.
-The CIS sampling is less frequent, but with a large sample size and long follow-up.
+\tableofcontents
 
-I take a Bayesian survival analysis approach to model the coarse sampling of the CIS (\cref{perf-test:sec:problem}).
-I use a simulation study (\cref{perf-test:sec:simulation-study}) to evaluate a range of priors for the survival distribution (\cref{perf-test:sec:parameters-priors}).
-These include a prior to combine the information in the CIS with the results of \cref{E-ATACCC}.
-Reasonable priors successfully recover the true distribution.
-However, the method must be able to account for false negatives, discussed in \cref{E-imperf-test}.
+\newpage
+\setstretch{1.7} % Fully double-spaced font. Adjust the value if needed to achieve 26 lines per page
 
-\section{Problem description} \label{perf-test:sec:problem}
+\section{Introduction}
 
-The CIS tests each individual infrequently, either weekly or monthly (see \cref{E-intro:sec:cis}).
-The infrequent testing complicates analysis of the data in two ways.
+\todo[inline]{Rough outline of this part, which motivates the problem. }
+
+The distribution of duration an individual tests RT-PCR positive with SARS-CoV-2 is important for blah blah reasons.
+
+Investigations into this distribution have been predominantly conducted in highly selected population, \eg healthcare works.
+Cite and expand.
+
+The ideal study design is frequent, long-term, repeated testing of a large, representative sample of the population.
+No study exists with this design.
+
+Here, we use external information from ATACCC to inform estimates with CIS.
+We estimate the distribution of duration an individual tests RT-PCR positive with SARS-CoV-2.
+
+The analysis of CIS is challenging because it is doubly interval censored, misses infections, and contains false negative results.
+We develop a Bayesian survival analysis model to estimate the distribution of duration, accounting for all these issues.
+
+\section{Study design and data}
+
+The CIS~\autocite{CIS} was setup in April 2020.
+The dataset is globally unique in providing a representative, longitudinal, and large-scale study across the pandemic.
+% Initially, it was limited to England, but expanded to cover the whole of the UK in September 2020.
+Enrolment was continuous until 31st Jan 2022, with data collected until 13th Mar 2023~\autocite{weiRisk}.
+
+The CIS had a household-based design inviting all individuals aged 2 and over from selected households.
+Once invited, an enrolment swab would be taken at the first visit followed by 4 further weekly visits (giving a total of 5 swabs on days 0, 7, 14, 21, 28 relative to enrolment) after which visits were monthly.
+As well as a swab to test for SARS-CoV-2, participants were asked to complete a questionnaire, and some participants were asked to provide blood samples.
+In reality, visits were often not on this precise schedule, and occasionally missed.
+A full description of the study can be found in the study protocol~\autocite{cisProtocol}.
+
+Positive tests are grouped into \emph{infection episodes} which are assumed to originate from the same infection (procedure described in \cref{sec:episodes}).
+I base my analysis on the 4,800 episodes that were first detected between 10th Oct 2020 and 6th Dec 2020 inclusive with negatives bounding the start and end time of the episode.
+That is, the individual tested negative before and after the episode.
+Without a negative before the episode there is no lower bound on the episode's start time.
+It could be arbitrarily long and there is little information on its length.
+A negative test follows each episode unless the individual was lost to follow-up before their episode ended.
+This is rare.
+Episodes removed in this step are incorporated in the adjustment for undetected episodes.
+
+For privacy reasons, the CIS data is stored in the ONS's SRS (Secure Research Service).
+The SRS is a Trusted Research Environment (TRE) that allows researchers to access sensitive data in a secure environment.
+The SRS must be used to perform analyses using individual-level data.
+Therefore, computationally efficient approaches must be taken to this work.
+
+\section{Problem description}
+
+The infrequent testing used by the CIS complicates analysis of the data in two ways.
 First, infection episodes can go undetected.
 Second, the infection episode start and end times are known imprecisely.
-This section explains these issues. The rest of the chapter develops a method to tackle  them.
+This section explains these issues.
 
@@
 Therefore, there is no information directly observed about the episode; even the number that occur is unobserved.
+
+Undetected events where there is no information on these events is known as \emph{truncation}.
+A situation that is particularly relevant is \emph{left truncation}, when only events lasting a minimum amount of time are observed.
+There is a subtlety that makes the CIS different from standard left truncation.
+In CIS, individuals in which undetected infections occurred are enrolled in the study.
+Therefore, some data has been collected about them.
+In standard left truncation, nothing is known about the missed events.
+Here, I will incorporate the additional information into the analysis.
+I refer to these events as ``undetected'', rather than truncated, to emphasise this difference.
+
 They occur when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
@@
 
-\subsection{Unobserved episode beginning and end times} \label{perf-test:sec:interval-censoring}
+\subsection{Double interval censoring} \label{perf-test:sec:interval-censoring}
 
-The second complication is that the exact beginning and end times of detected infection episodes are unknown.
+The second complication is that the infections are double interval censored.
 The imprecision in the beginning of the infection episode arises because the only observation that episode $j$ has started is that an individual was negative at one test, on day $l_j^{(b)}-1$ (the minus one is because the first day that the infection could have begun is the day after the negative test), and positive at the next test, on day $r_j^{(b)}$.
@@
 
-\subsection{Approach} \label{perf-test:sec:approach}
+\subsection{False negatives}
 
-I base my analysis on the 4,800 episodes that were first detected between 10th Oct 2020 and 6th Dec 2020 inclusive with negatives bounding the start and end time of the episode.
-That is, the individual tested negative before and after the episode.
-Without a negative before the episode there is no lower bound on the episode's start time.
-It could be arbitrarily long and there is little information on its length.
-A negative test follows each episode unless the individual was lost to follow-up before their episode ended.
-This is rare.
-Episodes removed in this step are incorporated in the adjustment for undetected episodes.
+Write something about false negatives being common.
 
-\section{Survival analysis}
-
-In survival analysis, the event at the start of the time of interest is the \emph{initiating event}, and the event at the end of the time of interest is the \emph{terminating event}.
-It is used in many areas, including how long a machine lasts before breaking down or how long a patient spends in hospital after a treatment.
-Here, I am estimating the distribution of the duration of positivity.
-The initiating event is the episode's start; the terminating event is the episode's end.
-
-Undetected events where there is no information on these events is known as \emph{truncation}.
-A situation that is particularly relevant is \emph{left truncation}, when only events lasting a minimum amount of time are observed.
-There is a subtlety that makes the CIS different from standard left truncation.
-In CIS, individuals in which undetected infections occurred are enrolled in the study.
-Therefore, some data has been collected about them.
-In standard left truncation, nothing is known about the missed events.
-Here, I will incorporate the additional information into the analysis.
-I refer to these events as ``undetected'', rather than truncated, to emphasise this difference.
-
-The time of the initiating and/or terminating events are not known exactly.
-This is known as \emph{censored} in the survival analysis literature.
-Censored events are present in almost all survival analyses.
-An event is \emph{left censored} if it is known to occur before a certain time.
-An event is \emph{right censored} if it is known to occur after a certain time.
-An event is \emph{interval censored} if it is known to occur within an interval.
-%Left and right censoring can be viewed as special cases of interval censoring where the lower bound is negative infinity or the upper bound is positive infinity respectively.
-As explained in \cref{perf-test:sec:problem}, the initiating and terminating events are only bounded, and hence interval censored.
-The data is \emph{doubly censored} because both the initiating and terminating events are censored.
-To handle doubly censored data, the times need to be jointly modelled~\autocite[and references therein]{liSemiparametric}.
-In CIS data both the initiating and the terminating events are interval censored.
-Therefore, the CIS data is \emph{double interval censored}.
-\Textcite{sunAnalysis,bogaertsSurvival} review methods for double interval censored data.
+\section{Related work}
 
@@
 These studies \autocite[and elsewhere, e.g.:][]{shenNonparametric} take a conditional likelihood approach.
-They condition on the truncation (a selection effect) and the period where the initiating event occurs.
-They condition on the initiating event time because it provides negligible information on the distribution of interest~\citePersonalComms{Nick Jewell}.
+They condition on the truncation (a selection effect) and the period where the initiating event occurs, which provides negligible information on the distribution of interest.
 This is not the case for the CIS.
@@
 I adapt this framework for CIS, when each individual has their own visit schedule (see section \cref{perf-test:sec:model}).
+This framework is itself based long-standing methodology~\autocite{dempsterMaximum,turnbullEmpirical}.
 
-I use Bayesian inference for two reasons.
-First, frequentist estimators are poorly developed and their properties (\eg consistency and convergence rate) are poorly understood~\autocite{sunAnalysis,dengNonparametric}.
-Second, a prior can include information from previous studies, such as from \cref{E-ATACCC}.
-Alternatively, the prior can favour a smoothly changing hazard.
-I explore priors in section 1.4.
-These prior structures overcome issues caused by the undetected episodes ~\autocite{caoBias}.
+The inclusion of false negatives into survival analysis is an area of much interest, especially in the application of tests for HIV in infants~\autocite[e.g.][]{brownBayesian,balasubramanianEstimation};
+\textcite{piresIntervalMisclassify} provide a comprehensive review of approaches.
+However, including false negatives with either doubly censored or truncated data has not been addressed.
+The situation here requires both.
 
-Bayesian methods have been used for nest survival before~\autocite{heBayesiana,heBayesian,caoModeling} by augmenting the data with the unobserved times of infection.
-The augmentation creates a parameter per detected infection.
-This increases the dimensionality of the problem greatly, making it computationally infeasible within the SRS.
-I develop a novel Bayesian framework without explicit data augmentation.
-This method scales to the large number of episodes present in CIS.
+\section{Methodology}
 
-\section{Modelling the duration}\label{perf-test:sec:model}
+\subsection{Survival analysis framework}
 
-In this section, I develop a model accounting for double interval censoring and undetected episodes.
+First, we consider the situtation without false negatives.
 This approach extends the framework of \textcite{heiseyModelling} to include individuals without detected episodes.
-This framework is itself based long-standing methodology~\autocite{dempsterMaximum,turnbullEmpirical}.
 I augment the data set with the full number of episodes, including those that are undetected.
@@
 
-% $\ntot$ is generally a nuisance parameter, however, its posterior distribution can be useful for diagnostic purposes (as in X\todo{ref where I use this in the next chapter}).
-% Its posterior can be reconstructed using a posterior sample of $\theta$.
-% For each posterior sample of $\theta$, sample from the full conditional of $\ntot$ to give the joint posterior of $\ntot$ and $\theta$.
-% The full conditional of $\ntot$ is given by:
-% \begin{align}
-% p(n_\text{tot} \mid \ncis, \theta)
-% &\propto p(\ntot \mid \theta) p(\ncis \mid \theta, \ntot) \\
-% &\propto p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} &\text{by \cref{perf-test:eq:augmented-likelihood}} \\
-% &\propto \frac{\Gamma(r + \ntot)}{\ntot!} \left( \frac{\mu}{r + \mu} \right)^{\ntot} \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot} \\
-% &= \frac{\Gamma(r + \ntot)}{(\ntot-\ndet)!} \left( \frac{\mu \pnodet}{r + \mu} \right)^{\ntot}  \\
-% &= \frac{\Gamma((r + \ndet) + (\ntot- \ndet))}{(\ntot-\ndet)!} \left( \frac{\mu \pnodet}{r + \mu} \right)^{\ntot-\ndet}.
-% \end{align}
-% Comparing this expression to the pmf of a negative binomial, we find that $\nnodet = \ntot - \ndet$ is distributed negative binomial with size parameter $r+\ndet$ and probability parameter $(r + \mu(1 - \pnodet))/(r + \mu)$.
-%The mean of this distribution is $(r+\ndet)\mu \pnodet/(r+\mu(1-\pnodet))$.
-
 Next, I derive $p_{ia}$.
@@
 
-\section{Parameterisation and priors for the survival function} \label{perf-test:sec:parameters-priors}
+\subsection{False negatives}
 
-In this section I specify the form of $S_\theta(t)$.
-% As frequently observed in the literature~\autocite[e.g.][]{heBayesian}, it is more convenient to operate with the hazard than the survival or probability mass function directly.
-% The mass function must sum to one.
-% The survival must be monotonically decreasing
-% But, there are no constraints across multiple hazards.
-% This would complicate the inference process in a way that cannot be removed through a simple transformation.
-% The only constraint on the hazard is to be in the interval $[0, 1]$, since they are probabilities.
-% The [0, 1] interval is easily be mapped to the unconstrained reals using a logit transformation (where $\logit(x) = \log(x/(1-x))$).
-Priors cannot be vague on both the hazard and the survival, the implications of which I consider in what follows.
-Informative priors are also attractive to allow the incorporation of the estimates from \cref{E-ATACCC}.
+\subsubsection{Generative model for false negatives} \label{imperf-test:sec:simulate}
 
-\subsection{Independent priors} \label{perf-test:sec:independent-priors}
-Take the standard assumption of independent priors on each parameter.
-Each hazard, $\lambda_t$, is a (conditional) probability, and hence a Beta distribution is a natural choice.
+\todo[inline]{Probably the wrong place for this subsection, and can be edited for brevity a lot}
 
-Uninformative Beta priors on the hazard are problematic in this context.
-An uninformative prior would be of the form $\lambda_t \dist \BetaDist(\alpha, \alpha)$.
-Commonly, $\alpha$ is chosen to be $0.5$ (Jeffreys' Prior) or $1$ (a uniform prior).
-Even though these priors are uninformative on the hazard, they become highly informative on the survival. 
-Specifically, they tend to favour shorter survival times (see \cref{perf-test:fig:flat-prior}).
-An intuitive explanation of this can be drawn from the fact that priors of this form have an expected value of 0.5.
-The expected survival time, $\E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right)$, is therefore equal to $0.5^{t-1}$, a quantity that declines rapidly.
-This prior expresses extreme scepticism in estimates from previous studies, such as \cref{E-ATACCC} and the meta-analysis of \textcite{cevikShedding}.
+In this section, I introduce a probabilistic model for the generation of false negatives.
+The simplest model for false negatives is a constant, non-zero probability of returning a negative test when detectable.
+That is the test sensitivity, $\psens$, is less than 1.
+
+Under this model, each test result in a detectable individual is an iid Bernoulli random variable with a probability $\psens$ of producing a positive.
+As previously, tests in individuals without ongoing infection episodes are always negative.
+I modified the simulation study from \cref{E-perf-test:sec:simulation-study} to generate test results in this way.
+
+Following the modification, more single positive episodes are seen at lower values of $\psens$, with a value between 60\% and 80\% reproducing the number of single positive episodes observed in the CIS data (see \cref{imperf-test:fig:sim-single-pos}).
+This is expected: a lower $\psens$ means that multiple positive tests are less likely, and hence the probability of a single positive episode is higher.
+
+External information (see \cref{sec:ATACCC-analysis}) gives that $\psens$ is 95\% (95\% CrI: 93--96\%).
+The requirement for a much lower value of $\psens$ in the simulation study here suggests this model of false negatives is inadequate.
+
+A better model would be to allow the rate of false negatives to vary over the course of an episode.
+False negatives are more likely to occur when the viral load of an individual is low, because there is less virus in their body to sample.
+In \cref{E-ATACCC}, this mechanism is incorporated by viewing negative tests as a left-truncation of the observation noise (see \cref{E-ATACCC:sec:observation-modification}).
+This observation suggests that a model with a declining test sensitivity as a function of time since infection might be more suitable.
+
+The CIS data provides evidence for a test sensitivity declining over the course of an episode.
+\Cref{imperf-test:fig:bounding-cis-sensitivity} demonstrates this by using the following classification for test results.
+\begin{enumerate}
+    \item All positive results are true positives.
+    \item All intermittent negatives are false negatives.
+    \item The negative at the end of an episode is a possible false negative.
+    \item All other negatives are true negatives.
+\end{enumerate}
+Here, I am assuming that the probability of two false negative tests at the end of an episode is negligible.
+As test sensitivity is the proportion of individuals who should test positive that do so, it can be empirically calculated as number of true positives / (number of true positives + number of false negatives).
+Therefore, I calculate the test sensitivity for each day, considering the first positive result in an episode as day 0.
+As the latent start day of the infection episode is before the first positive result, day 0 reflects some unknown number of days into the infection episode.
+I bound the test sensitivity on each day by assuming either all or none of the possible false negatives (group 3) are in fact false negatives.
+% In \cref{imperf-test:fig:bounding-cis-sensitivity} we consider the values produced by assuming the negative following the last positive in an episode could be either a true or false negative as a function of time since the individual was detected (\ie: the first positive test in the episode).
+This generates broad bounds, although suggest a declining test sensitivity over the course of an episode (see \cref{imperf-test:fig:bounding-cis-sensitivity}).
+Small numbers of tests mean that the series is noisy near the start and the end, such as both bounds being 0 on some days after day 40.
+However, the general trend is clear.
 \begin{figure}
-  \centering \includegraphics{cis-perfect-testing/flat-prior}
-  \caption[Uninformative priors for the hazard]{The density of Beta(1, 1) flat prior and Beta(0.5, 0.5) uninformative priors (left) with the prior predictive survival time (right).
-The implied prior on the survival time is very short when using either prior on the hazard at each time.
-\label{perf-test:fig:flat-prior}}
+  \includegraphics{cis-imperfect-testing/test-sens-bound}
+  \caption[Bounding test sensitivity using CIS data]{
+    Bounding the test sensitivity using CIS data as a function of time since the infection was detected.
+  }
+  \label{imperf-test:fig:bounding-cis-sensitivity}
 \end{figure}
 
-Instead, I propose a weakly informative (or vague) prior Beta(0.1, 1.9), which has mean 0.05 and minimal information.
-The amount of information in a Beta distribution is related to the sum of its parameters.
-Here, the sum equals 2, the same as the flat prior case.
-The central 95\% probability mass of Beta(0.1, 1.9) is 0.00--0.47.
-The central estimate, of 0.05, is in line with previous estimates that the median duration is in the range 15--20 days~\autocite{cevikShedding}.
-This prior gives a very vague prior predictive distribution on the survival time (see \cref{perf-test:fig:vague-prior}).
-\begin{figure}
-  \centering \includegraphics{cis-perfect-testing/vague-prior}
-  \caption[Vague prior for the hazard]{The density of Beta(0.1, 1.9) vague prior (left) and the prior predictive survival time (right).
-The prior predictive survival time is very vague using a Beta(0.1, 1.9) prior.
-Note the different x-axis to \cref{perf-test:fig:flat-prior}.
-\label{perf-test:fig:vague-prior}}
-\end{figure}
+To assess the sensitivity of the procedure to a time-varying test sensitivity, it will be useful to simulate from a model with a varying test sensitivity as a function of days since the start of the infection episode.
+Through consideration of \cref{imperf-test:fig:bounding-cis-sensitivity}, and the interaction of false negatives due to viral load in \cref{E-ATACCC}'s results, I propose the following model.
+\begin{equation}
+  p_\text{sens}(t) = \begin{cases}
+    0.9 - \frac{0.9-0.5}{50}t &t \leq 50 \\
+    0.5 &t > 50
+  \end{cases}
+  \label{imperf-test:eq:variable-test-sensitivity}
+\end{equation}
+where $t$ is the number of days since infection.
+Therefore, this model consists of a linear decline in test sensitivity, before a constant minimum point.
 
-\subsection{Smoothing priors}
+I simulate data under this model by, for each test in a detectable individual, calculating $\psens(t-b)$, where $t$ is the day of the test and $b$ is the day the episode began.
+The result of the test is then a Bernoulli random variable, independent of all other considerations.
+Adapting the simulation study to incorporate this test sensitivity is much more similar to the data (see purple histogram in \cref{imperf-test:fig:sim-single-pos}).
+This model does not require low test sensitivities early in the infection, maintaining compatibility with the results in \cref{E-ATACCC}.
 
-Smoothing priors encode the biological consideration that the hazard should not vary significantly from day-to-day.
-It is the Bayesian equivalent of a penalised likelihood, used frequently in the non-parametric frequentist setting~\autocite[e.g.][]{bacchettiNonparametric}.
-Many possible forms of smoothing priors exist, including splines, Gaussian Processes, and random walks.
-I use a second-order random walk (RW2) because it is simple and produces sensible prior predictive distributions.
+\subsubsection{Incorporation into the survival analysis framework}
 
-RW2 prior encodes that the hazard should be linearly changing with some random changes at each time step.
-Specifically, the difference on the logit scale between $\lambda_t$ to $\lambda_{t+1}$ is the difference between $\lambda_{t-1}$ and $\lambda_t$ plus some Normally-distributed noise.
-Formally:
+Now we modify the survival framework to incorporate false negatives.
+We use the model with a constant test sensitivity to ensure the likelihood remains tractable.
+
+First, in \cref{imperf-test:sec:modifying-p_ia}, I modify $p_{ia}$, the probability of an episode with test schedule $i$ being admissible, to allow for the episode possibly being longer than observed.
+Then, in \cref{imperf-test:sec:modifying-p_iu}, I modify $p_{iu}$, the probability of an episode with test schedule $i$ being undetected, to allow for additional episodes being missed.
+
+\subsection{Modifying \texorpdfstring{$p_{ia}$}{pia}} \label{imperf-test:sec:modifying-p_ia}
+
+I will modify $p_{ia}$ to allow the negative test following the last positive to be a false negative.
+If it is a false negative, then I will consider the episode's end right-censored.
+However, if it is a true negative, then the episode's end is interval-censored, as previously.
+A mixture of these scenarios forms the episode's likelihood contribution, with the mixture probability determined by the test sensitivity.
+
+Similar likelihoods have previously appeared in the literature~\autocite[e.g.][eq.\ (2)]{piresIntervalMisclassify}.
+However, this prior work was for singly interval censored data; incorporating the double interval censored nature of the CIS data will involve summing over the possible episode start times.
+
+I start by finding the tests that need consideration.
+For tractability and simplicity, I consider only tests between the negative tests providing an upper bound on the length of the episode.
+By definition (see \cref{E-biology-data:sec:cis-episodes}), these are the tests between $l_{j(i)}^{(b)} - 1$ and $r_{j(i)}^{(e)} + 1$ inclusive.
+They are a subset of $\sched_i$, the set of test times for the individual in which episode $i$ occurs.
+
+For tractability, assume that the negative test bounding the start of the episode, on day $l_{j(i)}^{(b)}-1$, is a true negative.
+This assumption is reasonable because, since a positive test follows at $r_{j(i)}^{(b)}$, the negative at $l_{j(i)}^{(b)}-1$ is likely early in the infection and the test sensitivity is high early in an infection.
+Therefore, this is unlikely to be a false negative because, if the individual was detectable at this time, they are in a period with high test sensitivity.
+This true negative, occurs with probability 1, and hence does not contribute to the likelihood.
+Denote the remaining tests considered as $\sched'_i = \{ t \in \sched_i \ssep r_{j(i)}^{(b)} \leq t \leq r_{j(i)}^{(e)} + 1 \}$, and their results by $\vec{y_{j(i)}} = \{ y_{j(i)}(t) \ssep t \in \sched'_i \}$ where $y_{j(i)}(t) = 1$ if the test on day $t$ is positive and 0 otherwise.
+
+As I assume that there are no false positives, the infection episode must span at least the period $[r^{(b)}_{j(i)}, l^{(e)}_{j(i)}]$ because it starts and ends with a positive test.
+This includes all $t \in \sched'_i$ except $t = r_{j(i)}^{(e)}+1$.
+Therefore, the test results at times in $\sched_i'$, except the test at $r_{j(i)}^{(e)}+1$, are either true positives or false negatives.
+
+Consider the negative test at $r_{j(i)}^{(e)}+1$, the first negative after the start of the episode which may be a false negative.
+It is a false negative if and only if the episode ends after the test, \ie $e_{j(i)} > r_{j(i)}^{(e)}$.
+I proceed by considering whether this is the case, and considering the case with $b_{j(i)}$ known; $b_{j(i)}$ will later be integrated out.
+
+First, if $e_{j(i)} \leq r_{j(i)}^{(e)}$, meaning that the test at $r_{j(i)}^{(e)}+1$ is a true negative and the end of the episode is interval-censored as in the previous chapter.
+% In this case, the test at $r_{j(i)}^{(e)} + 1$ is a true negative, as are all other tests not in $\sched'_i$.
+The true negative occurs with probability 1, by the assumption of no false positives.
 \begin{align}
-  \logit\lambda_{t+1}
-  &= \logit\lambda_t + (\logit\lambda_t - \logit\lambda_{t-1}) + \sigma \epsilon_t &\text{for $t \geq 2$} \\
-  &= 2\logit\lambda_t - \logit\lambda_{t-1} + \sigma \epsilon_t \\
-  \epsilon_t &\dist \N(0, 1) &\text{for $t \geq 2$}  \\
-  \logit\lambda_2 &= \logit\lambda_1 + \epsilon_1 \\\
-  \epsilon_1 &\dist \N(\mu_{\epsilon_1}, \sigma_{\epsilon_1}^2) \\
-  \logit \lambda_1 &\dist \N(\mu_{\lambda_1}, \sigma_{\lambda_1}^2) \\
-  \sigma &\dist \Exponential(1/\mu_\sigma).
+&p(\vec{y_{j(i)}}, e_{j(i)} \leq r_{j(i)}^{(e)} | b_{j(i)}, p_\text{sens}, \theta) \\
+&= p(\vec{y_{j(i)}}, l_{j(i)}^{(e)} \leq e_{j(i)} \leq r_{j(i)}^{(e)} | t_i, b_{j(i)}, p_\text{sens}, \theta) \\ % &\text{as no false positives}
+&= p(\vec{y_{j(i)}} \mid l_{j(i)}^{(e)} \leq e_{j(i)} \leq r_{j(i)}^{(e)}, t_i, b_{j(i)}, p_\text{sens}, \theta) p(l_{j(i)}^{(e)} \leq e_{j(i)} \leq r_{j(i)}^{(e)} | t_i, b_{j(i)}, p_\text{sens}, \theta) \\
+&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_{j(i)}(t)} (1 - p_\text{sens})^{(1 - y_{j(i)}(t))} \right) \left( S_\theta(l_{j(i)}^{(e)} - b_{j(i)} - 1) - S_\theta(r_{j(i)}^{(e)} - b_{j(i)} - 1) \right)
+\label{imperf-test:eq:ll-ei-lt-ri}
 \end{align}
-The hyperparameters $\mu_{\lambda_1}$ and $\sigma_{\lambda_1}$ specify the prior on the hazard at time 1; $\mu_{\epsilon_1}$ and $\sigma_{\epsilon_1}$ the prior on the initial gradient; and $\mu_\sigma$ the smoothness of the random walk.
-I use $\mu_{\lambda_1} = -17.5$, $\sigma_{\lambda_1} = 6$, $\mu_\epsilon = 1.09$, $\sigma_\epsilon = 0.03$, and $\mu_\epsilon = 0.1$ based on the analysis in \cref{E-ATACCC} (see \cref{perf-test:fig:rw2-prior}).
-\begin{figure}
-  \centering \includegraphics{cis-perfect-testing/rw2-prior}
-  \caption[RW2 prior for the hazard]{The of prior predictive distribution (mean and 95\% central interval) for the RW2 prior specified in the main text.
-The posterior mean estimate using ATACCC data in \cref{E-ATACCC} are shown for comparison.
-\label{perf-test:fig:rw2-prior}}
-\end{figure}
 
-\subsection{Model combination priors} \label{perf-test:sec:informative-priors}
+Second, if $e_{j(i)} > r_{j(i)}^{(e)}$.
+In this case, the test at $r_{j(i)}^{(e)}$ is a false negative, occurring with probability $(1 - p_\text{sens})$.
+To avoid having to consider tests after $r_{j(i)}^{(e)}$, which could greatly complicate the likelihood, I model this case as the episode being right-censored at $r_{j(i)}^{(e)}$.
+Taking the same approach as before:
+\begin{align}
+&p(\vec{y_{j(i)}}, e_{j(i)} > r_{j(i)}^{(e)} | b_{j(i)}, p_\text{sens}, \theta) \\
+&= p(\vec{y_{j(i)}} \mid e_{j(i)} > r_{j(i)}^{(e)}, t_i, b_{j(i)}, p_\text{sens}, \theta) p(e_{j(i)} > r_{j(i)}^{(e)} | t_i, b_{j(i)}, p_\text{sens}, \theta) \\
+&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_{j(i)}(t)} (1 - p_\text{sens})^{(1 - y_{j(i)}(t))} \right) (1 - p_\text{sens}) S_\theta(r_{j(i)}^{(e)} - b_{j(i)} - 1)
+\label{imperf-test:eq:ll-ei-gt-ri}
+\end{align}
+
+These expressions can now be used to form the full replacement for $p_{ia}$, $p'_{ia}$, the likelihood of the data $\vec{y_{j(i)}}$.
+First, augment the data with $b_{j(i)}$, and split into the cases just discussed:
+\begin{align}
+p_{ia}'
+&= p(\vec{y_{j(i)}} \mid p_\text{sens}, \theta) \\
+&= \sum_{b_{j(i)} = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( p(\vec{y_{j(i)}}, e_{j(i)} \leq r_{j(i)}^{(e)} \mid b_{j(i)}, p_\text{sens}, \theta) + p(\vec{y_{j(i)}}, e_{j(i)} > r_{j(i)}^{(e)} \mid b_{j(i)}, p_\text{sens}, \theta) \right) p(b_{j(i)} \mid p_\text{sens}, \theta).
+\intertext{Now, substitute in \cref{imperf-test:eq:ll-ei-lt-ri,imperf-test:eq:ll-ei-gt-ri}, then take out the common factor:}
+&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_{j(i)}(t)} (1 - p_\text{sens})^{(1 - y_{j(i)}(t))} \right) \\ & \ \times \sum_{b_{j(i)} = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_\theta(l_{j(i)}^{(e)} - b_{j(i)} - 1) - S_\theta(r_{j(i)}^{(e)} - b_{j(i)} - 1) + (1 - p_\text{sens}) S_\theta(r_{j(i)}^{(e)} - b_{j(i)} - 1) \right) \\ & \ \times p(b_{j(i)} \mid p_\text{sens}, \theta) \\
+&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_{j(i)}(t)} (1 - p_\text{sens})^{(1 - y_{j(i)}(t))} \right) \\ & \ \times \sum_{b_{j(i)} = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_\theta(l_{j(i)}^{(e)} - b_{j(i)} - 1) - p_\text{sens} S_\theta(r_{j(i)}^{(e)} - b_{j(i)} - 1) \right) p(b_{j(i)} \mid p_\text{sens}, \theta).
+\label{imperf-test:eq:pia-prime}
+\end{align}
+Note that if $p_\text{sens} = 1$ then $p_{ia}' = p_{ia}$.
+
+If $\psens$ is fixed (\ie has a point prior) and $p(b_{j(i)} \mid \psens, \theta) \propto 1$ (as assumed in \cref{E-perf-test}) then:
+\begin{align}
+p_{ia}'
+&\propto \sum_{b_{j(i)} = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} S_\theta(l_{j(i)}^{(e)} - b_{j(i)} - 1) - p_\text{sens} S_\theta(r_{j(i)}^{(e)} - b_{j(i)} - 1).
+\label{imperf-test:eq:pia-prime-constant}
+\end{align}
+
+\subsection{Modifying \texorpdfstring{$p_{iu}$}{piu}} \label{imperf-test:sec:modifying-p_iu}
+
+Several mechanisms for episodes being undetected were considered when deriving $p_{iu}$ in \cref{E-perf-test:eq:piu}.
+I now consider the additional mechanisms arising due to false negatives.
+Specifically, episode $i$ could be undetected if the first test after the $b_{j(i)}$, episode $i$'s start day, is a false negative and then there are no subsequent positive tests.
+
+This false negative would occur at the first test after the infection episode begins, on day $b_{j(i)} + \tau_{\sched_i}(b_{j(i)})$ (see \cref{E-perf-test:sec:model}).
+A false negative occurring requires that the episode has not yet ended but a negative still occurs.
+The episode has not yet ended at the time of the test if $e_{j(i)} = b_{j(i)} + d_i - 1$ is after the test, that is the duration of the infection $d_i > \tau_{\sched_i}(b_{j(i)})$.
+Conditional on the episode having not yet ended, the test result is negative with probability $1 - \psens$.
+
+For there to be no subsequent positive tests, all tests up until day $e_{j(i)}$ are false negatives.
+I assume there is a negligible probability of two false negatives.
+Therefore, this can only occur if the episode ends before another test.
+Denote the number of days between $b_{j(i)}$ and the test following the false negative as $\tau^2_{\sched_i}(b_{j(i)}) \stackrel{\text{def}}{=} \tau_{\sched_i}(\tau_{\sched_i}(b_{j(i)}) + 1)$.
+Then, the episode ends before this test if $d_i < \tau^2_{\sched_i}(b_{j(i)})$.
+
+Therefore, this mechanism causes episode $i$ to be undetected if all the following conditions hold.
+\begin{enumerate}
+    \item The episode would have been detected considering only the mechanisms in \cref{E-perf-test:eq:piu}.
+That is $b_{j(i)} > \min(\sched_i)$ and $\tau_{\sched_i}(b_{j(i)}) + b_{j(i)} \leq e_{j(i)}$.
+    \item The episode ends in the interval $[\tau_{\sched_i}(b_{j(i)}) + b_{j(i)}, \tau^2_{\sched_i}(b_{j(i)}) + b_{j(i)}]$.
+      Note that the lower bound here is exactly the bound on $e_{j(i)}$ in the previous condition.
+      Equivalently, $\tau_{\sched_i}(b_{j(i)}) + 1 \leq d_i \leq \tau^2_{\sched_i}(b_{j(i)}) + 1$.
+    \item A false negative occurs on day $\tau_{\sched_i}(b_{j(i)}) + b_{j(i)}$. Conditional on the previous condition, this occurs with probability $1 - \psens$.
+\end{enumerate}
+
+The probability of this occurring, conditional on $B_{j(i)} = b_{j(i)} > \min(\sched_i)$ is:
+\begin{align}
+&\prob \left(
+    \tau_{\sched_i}(b_{j(i)}) + 1 \leq D_i \leq \tau^2_{\sched_i}(b_{j(i)}) + 1
+    \mid b_{j(i)},
+\theta \right) (1 - \psens) \\
+&= \left( S_\theta(\tau_{\sched_i}(b_{j(i)}) + 1) - S_\theta(\tau^2_{\sched_i}(b_{j(i)}) + 1) \right) (1 - \psens).
+\end{align}
+Integrating over $b_{j(i)}$ for $\min \sched_i < b_{j(i)} \leq T$ ($T$ being the last day of the period considered), in the same way as \cref{E-perf-test:eq:piu}, gives:
+\begin{align}
+\eta = (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_i) + 1}^T \left( S_\theta(\tau_{\sched_i}(b) + 1) - S_\theta(\tau^2_{\sched_i}(b) + 1) \right).
+\end{align}
+
+Denote the replacement for $p_{iu}$ as $p_{iu}'$.
+$p_{iu}'$ is the probability of episode $i$ being missed, considering both the mechanisms from \cref{E-perf-test} and the new mechanism here.
+These mechanisms are mutually exclusive.
+Hence, $p_{iu}'$ is the sum of these, $p_{iu}' = p_{iu} + \eta$.
+For the posterior density, \cref{E-perf-test:eq:full-posterior}, $1 - p_{iu}'$ is the required quantity.
+\begin{align}
+1 - p_{iu}'
+&= 1 - p_{iu} - (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_i) + 1}^T \left( S_\theta(\tau_{\sched_i}(b) + 1) - S_\theta(\tau^2_{\sched_i}(b) + 1) \right) \\
+&= \frac{1}{T} \sum_{b=\min(\sched_i) + 1}^T \left( p_\text{sens} S_\theta(\tau_{\sched_i}(b) + 1) + (1 - p_\text{sens}) S_\theta(\tau^2_{\sched_i}(b) + 1)\right).
+\label{imperf-test:eq:pit-prime}
+\end{align}
+
+$p_{iu}'$ and $p_{ia}'$ are substituted for $p_{iu}$ and $p_{ia}$ respectively in \cref{E-perf-test:eq:full-posterior}.
+The posterior density is otherwise unchanged.
+
+\subsection{Incorporating prior information}
+
+\todo[inline]{A bunch of this can probably be cut / moved to supp material. Supp material should also contain the ATACCC analysis.}
 
@@
 
-\section{Simulation study} \label{perf-test:sec:simulation-study}
+\section{Simulation study}
 
@@
     \item Draw an episode start time, $b_{j(i)}$ for each individual uniformly at random between 2nd July 2020 (100 days before the period where a detected episode would be included) and 6th December 2020 (the end of this period).
-    \item Draw a duration of episode for each individual, $d_i$, based on the distribution described below. Then calculate the end of their infection episode, $e_{j(i)} = b_{j(i)} + d_i - 1$.
-    \item Simulate the test results based on the test schedule, $b_{j(i)}$, and $e_{j(i)}$. Tests between $b_{j(i)}$ and $e_{j(i)}$ (inclusive) are positive, all other tests are negative.
+    \item Draw a duration of episode for each individual, $d_i$, based on a modification of previous estimates (see \cref{duration-ground-truth}).
+Then calculate the end of their infection episode, $e_{j(i)} = b_{j(i)} + d_i - 1$.
+    \item Simulate the test results based on the test schedule, $b_{j(i)}$, and $e_{j(i)}$.
+Tests between $b_{j(i)}$ and $e_{j(i)}$ (inclusive) are positive with positive probability (see below).
+All tests outside this interval are negative.
     \item Discard episodes where there are no positive tests (\ie undetected episodes) and then apply the same inclusions/exclusions as in \cref{perf-test:sec:problem}.
@@
 
+I simulate data under this model by, for each test in a detectable individual, calculating $\psens(t-b)$, where $t$ is the day of the test and $b$ is the day the episode began.
+The result of the test is then a Bernoulli random variable, independent of all other considerations.
+Adapting the simulation study to incorporate this test sensitivity is much more similar to the data (see purple histogram in \cref{imperf-test:fig:sim-single-pos}).
+This model does not require low test sensitivities early in the infection, maintaining compatibility with the results in \cref{E-ATACCC}.
+
+I simulate datasets under four conditions: a constant $\psens$ of 0.6, 0.8, or 1.0, or the varying test sensitivity model proposed at the end of \cref{imperf-test:sec:simulate}.
+The simulation is otherwise unchanged from that in \cref{E-perf-test:sec:simulation-study}.
+The number of single positive episodes in each of these conditions, compared to the number observed in the CIS data, is shown in \cref{imperf-test:fig:sim-single-pos}.
+
+For each simulated dataset, I estimate the survival function using the likelihood proposed in \cref{imperf-test:sec:modelling} and assuming $\psens$ equals 0.6, 0.8, or 1.0.
+If the dataset was simulated using a constant $\psens$, and that value of $\psens$ matches the value assumed during inference, I refer to $\psens$ as being correctly specified; otherwise I refer to it as misspecified.
+However, there is still some misspecification of the model to false negatives owing to the simplifying assumptions made.
+The amount that the simplifying assumptions are violated increases as $\psens$ decreases.
+
+\subsection{Implementation}
+
+Simulation was performed in R~4.2.0~\autocite{R-4-2-0} using tidyverse~2.0.0~\autocite{tidyverse} and targets~1.1.3~\autocite{targetsPackage}.
+
+Inference was implemented in Stan via RStan~2.21.8~\autocite{rstan2-21-8} using default settings.
+Convergence was assessed using Rhat and ESS as described in \cref{E-inc-prev:sec:MCMC}, and all runs checked for divergent transitions.
+
+The prior on the total number of infections was centred on the true value with a dispersion parameter of 1.
+This meant that the prior was very diffuse.
+
+\subsection{Results}
+
+When $\psens = 0.8$ and is correctly specified, the model recovers the true survival time well (see \cref{imperf-test:fig:constant-test-sensitivity}(B)).
+The informative prior, in comparison to the vague prior, helps to overcome the misspecification due to the simplifying assumptions, moving the estimated survival function closer to its true survival time.
+However, when $\psens = 0.6$, this is no longer the case (see \cref{imperf-test:fig:constant-test-sensitivity}(A)).
+This is likely caused by too large a violation of the simplifying assumptions made in \cref{imperf-test:sec:modelling}.
+\begin{figure}
+  % \makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{cis-imperfect-testing/sim-constant-sensitivity}}
+  \includegraphics[width=\textwidth]{cis-imperfect-testing/sim-constant-sensitivity}
+  \caption[Simulation study results with constant test sensitivity]{%
+    Posterior (median and 95\% credible interval) of the survival time for the simulation study with a correctly specified test sensitivity.
+    True survival time shown in black.
+  }
+  \label{imperf-test:fig:constant-test-sensitivity}
+\end{figure}
+
+Next, I considered the consequence of $\psens$ being misspecified, using the model combination prior; the better performing prior in the correctly specified case.
+If the test sensitivity is misspecified, that is the assumed value for $p_\text{sens}$ in \cref{imperf-test:eq:pit-prime,imperf-test:eq:pit-prime} is different to the value used in the simulation, then the survival time will be biased (see \cref{imperf-test:fig:misspecified-test-sensitivity}).
+If $\psens$ is misspecified and too low (see \cref{imperf-test:fig:misspecified-test-sensitivity}(A)), then the posterior estimate initially follows the true value but then separates.
+The number of episodes inferred to have truly ended by the first negative is too high, and hence the survival time is overestimated.
+This effect dominates over the opposing bias of overestimating the number of missed episodes.
+The opposite occurs if the test sensitivity is assumed to be too high, although the posterior moves away from the truth earlier (see \cref{imperf-test:fig:misspecified-test-sensitivity}(C)).
+\begin{figure}
+  % \makebox[\textwidth][c]{
+    \includegraphics[width=\textwidth]{cis-imperfect-testing/sim-misspecified-sensitivity}
+  \caption[Simulation study results with misspecified test sensitivity]{%
+    Posterior (median and 95\% credible interval) of the survival time for the simulation study with a possibly misspecified test sensitivity.
+    True survival time shown in black.
+    All simulations use a constant test sensitivity of 0.8, but the inference procedure assumes different values, as per the key.
+    Hence, (B) is correctly specified but (A) and (C) have misspecified $\psens$.
+  }
+  \label{imperf-test:fig:misspecified-test-sensitivity}
+\end{figure}
+
+The results when a varying test sensitivity (\cref{imperf-test:eq:variable-test-sensitivity}) is used for the simulation are similar to a constant 0.8 test sensitivity (see \cref{imperf-test:fig:variable-test-sensitivity}).
+This suggests that the simplified model, with constant test sensitivity, is sufficient for recovering the true survival time.
+Estimating the test sensitivity is not possible without a more complex model, as discussed in \cref{imperf-test:sec:discussion}.
+Therefore, I will apply this model to the real CIS data in the next section.
+\begin{figure}
+  % \makebox[\textwidth][c]{
+    \includegraphics[width=\textwidth]{cis-imperfect-testing/sim-variable-sensitivity}
+  \caption[Simulation study results with varying test sensitivity]{%
+    Posterior (median and 95\% credible interval) of the survival time for the simulation study with a variable test sensitivity.
+    Each panel shows the results of performing inference with a different assumed value for the test sensitivity.
+    The posterior estimate using a model with a constant test sensitivity of 0.8 (B) is similar to the true value (black line).
+  }
+  \label{imperf-test:fig:variable-test-sensitivity}
+\end{figure}
+
+\section{Application}
+
+In this section I apply the approach described in this chapter to the CIS episodes dataset I described in \cref{E-perf-test:sec:approach}.
+This is the 4,800 CIS infection episodes first detected between 10th Oct 2020 and 6th Dec 2020 inclusive and have negatives bounding the start and end time of the episode.
+
+Unlike in the simulation studies, an uninformative prior on $\ntot$ led to implausible estimates of the duration distribution.
+The uninformative prior led to high posterior estimates of $\ntot$, and hence an implausibly large number of episodes with durations of less than five days.
+Therefore, I based an informative prior for $\ntot$, $\ntot \sim \NBc(\mu, r)$ (as in \cref{E-perf-test:sec:model}, see \cref{E-distributions} for definition), using pre-existing estimates of the total number of infections to give $\mu\inform$ and $r\inform$.
+\Textcite{birrellRTM2} estimated the total number of infections in England over the time period I consider, with posterior mean \numprint{4136368} and standard deviation \numprint{27932}.
+% This model gives a posterior mean of \numprint{4136368} cumulative infections in England in the time period I consider, with a posterior standard deviation of \numprint{27932}~\citePersonalComms{Paul Birrell}.
+\todo{Insert correct citation for RTM paper 2 once available}
+Approximating this distribution as a negative binomial and scaling the mean to the size of the CIS cohort gives the prior $\mu\inform = 25132$ and $r\inform = 22047)$.
+
+With this prior, the model produces plausible estimates of the duration distribution (see \cref{imperf-test:fig:cis-estimates}).
+This estimate (blue) has more long episodes than the estimate in \cref{E-ATACCC} (red).
+
+The qualitative increase in long episodes is robust to the choice of prior for $\ntot$, the assumed value for $\psens$ (see \cref{imperf-test:fig:cis-sensitivity}), and the choice of prior for the hazards, $\lambda_t$ (see \cref{E-perf-test:sec:parameters-priors}).
+However, the survival proportion over the first 4 weeks is sensitive to these choices.
+The estimate using a test sensitivity of 0.8 and $\NBc(\mu\inform, r\inform)$ give a median survival time most similar to that of \cref{E-ATACCC}.
+\begin{figure}
+  \centering \includegraphics{cis-imperfect-testing/CIS_final}
+  \caption{Duration estimates using CIS and ATACCC data}
+  \label{imperf-test:fig:cis-estimates}
+\end{figure}
+\begin{figure}
+  \vspace{-2.5cm}
+  \makebox[\textwidth][c]{\includegraphics{cis-imperfect-testing/CIS_vary}}
+  \caption{%
+    (A-C) Sensitivity of CIS estimates to the value of $r$, the dispersion of the prior for $\ntot$ (a larger $r$ means a more informative prior) when $\psens = 0.8$.
+    (D-F) Sensitivity of CIS estimates to the value of $\psens$ when $r = r\inform$.
+    A and D: median survival time, in comparison to the ATACCC-based estimate of \cref{E-ATACCC} (shown in red).
+    B and E: survival at day 50, $S_\theta(50)$.
+    C and F: full survival curves out to day 100.
+  }
+  \label{imperf-test:fig:cis-sensitivity}
+  \todo[inline]{Optional improvement: Could highlight the "preferred" model which exists in both rows and is the one used subsequently, and/or include ATACCC in the far-right hand column}
+\end{figure}
+
+The the estimates are most sensitive to the choice of $r$, the strength of the prior on $\ntot$.
+A very weak, almost uninformative, prior on this quantity causes the posterior estimate to be much higher than the estimate from \textcite{birrellRTM2}.
+When increasing the prior's strength, the posterior estimate moves towards the prior smoothly, as expected (see \cref{imperf-test:fig:ntot}).
+As discussed previously, the \cref{E-ATACCC} estimates are reliable for the first 2--3 weeks, notably including the median time.
+The median using $r\inform$ matched \cref{E-ATACCC}'s median estimate and is a principled choice because it is based directly on the prior work \textcite{birrellRTM2}.
+Therefore, I use this estimate in the application to the CIS data.
+\begin{figure}
+  \centering \includegraphics{cis-imperfect-testing/CIS_ntot}
+  \caption{How the posterior estimate of $\ntot$ changes with the value of $r$ in the prior on $\ntot$.}
+  \label{imperf-test:fig:ntot}
+\end{figure}
+
+\section{Discussion}
+
+In this chapter, I developed a novel method to incorporate false negatives into the survival analysis of \cref{E-perf-test}.
+I then applied this method to the CIS data, estimating the tail of the duration distribution without the strong model assumptions and extrapolation in \cref{E-ATACCC}.
+The qualitative features of the estimated distribution are robust to the assumed value for $\psens$, and the choice of prior for $\lambda$, although the quantitative details are not.
+The estimates of the tail, the primary purpose of this chapter, are in addition somewhat robust to the choice of prior for $\ntot$.
+However, the rest of the distribution is sensitive to this choice.
+
+An informative prior on $\ntot$ is required is for the bulk of the distribution to agree with those in \cref{E-ATACCC} and the wider literature.
+In the simulation study, the uninformative prior on $\ntot$ performed well, even when the test sensitivity was misspecified.
+A possible explanation is that the true changes in the test sensitivity is significantly different to the function in \cref{imperf-test:eq:variable-test-sensitivity}.
+In particular, the test sensitivity for long episodes may become much lower than 50\%.
+This allows many more episodes to be missed than the model assumes, and hence a higher $\ntot$ is required to explain the data.
+Furthermore, it violates assumptions made in \cref{imperf-test:sec:modelling}; this is similar to when the simulating using a too low test sensitivity.
+Violating these assumptions would lead to unpredictable inference results, perhaps those seen here.
+
+Ideally, the test sensitivity would be estimated from the data.
+However, this would require incorporating time-varying test sensitivity into the likelihood.
+If the current model, with a constant $\psens$, is used then the estimate of $\psens$ would be heavily informed be intermittent negatives (the first, constant term in \cref{imperf-test:eq:pia-prime}).
+These negatives will, in general, be further from the end of an episode than a randomly selected test.
+Therefore, considering the viral dynamics from \cref{E-ATACCC}, the viral load will be higher and false negatives rarer.
+
+Estimating the test sensitivity excluding intermittent negatives is not possible because \cref{imperf-test:eq:pia-prime-constant,imperf-test:eq:pit-prime} are both monotonically decreasing in $\psens$; therefore, the likelihood always favours $\psens = 0$ (\ie no true positives).
+This aligns with the situation with singly interval-censored, untruncated data, when stopping at the first observed time of the terminating event (which may be misclassified) means that the test sensitivity cannot be estimated~\autocite[e.g.]{titmanMisclassify}.
+
+Estimating a time-varying $\psens$ with $S_\theta$ jointly may cause identifiability issues.
+Previous studies have avoided issues by including external information (such as a prior giving the magnitude) on $\psens$, or test results from later follow-up~\autocite[and references therein]{piresIntervalMisclassify}.
+However, these studies use a constant $\psens$.
+Whether these methods are sufficient for the model to be identifiable with a time-varying $\psens$ needs further investigation.
+A simple parameteric form of the test sensitivity, such as that proposed by \textcite{brownBayesian}, may be sufficient to allow identifiability.
+In any case, the likelihood, especially $p_{iu}'$, would be substantially complicated by such an addition which may greatly increase the computational cost of inference.
+
+The assumption, made in \cref{E-perf-test}, was that the episode start times are independent.
+However, because infections are the result of a counting process with intensity shared between individuals (see \cref{E-inc-prev:sec:infection-process}), this assumption is violated.
+Prevalence was fairly constant over the period I chose, suggesting the same is true of the incidence.
+Additionally, some simulations exploring the results of changing incidence suggested this would not have a large effect on the results (not shown).
+\printbibliography[heading=bibintoc]
+
+\appendix
+
+\section{Defining episodes} \label{episodes}
+
+\todo[inline]{This could be cut for a reference to \textcite{weiRisk}}
+
+Positive test results need to be grouped into \emph{infection episodes}.
+An infection episode is a series of positive tests that originate from a single transmission event.
+I use a heuristic-based algorithm developed throughout the pandemic, and published within \textcite{weiRisk}; see that publication for full justification of the criteria.
+
+The algorithm determines whether a new positive test, at time $t$, is part of the previous infection episode in the same individual.
+If the individual has no previous positive tests, the new positive test is always the start of a new infection episode.
+Otherwise, denote the first positive test associated with that previous infection episode as $t_p$, and the number of negative tests immediately prior to the test at $t$ as $n$.
+The test at $t$, in the pre-Omicron era (prior to December 2021), is defined as the start of a new infection episode if any of the following apply.
+\begin{enumerate}
+  \item $n \geq 1$ and $t - t_p > 120$.
+  \item $n \geq 2$ and $t - t_p > 90$.
+  \item $n \geq 3$ and $t - t_p > 60$.
+  \item $n \geq 4$.
+  \item A complex heuristic based on the likely variant of the positive tests (see \textcite{weiRisk} for details). This heuristic rarely applied in the pre-Omicron era.
+\end{enumerate}
+
+A negative between two positive tests, when those positives are classed as part of the same infection episode, is known as an \emph{intermittent negative}.
+Intermittent negatives are the clearest example of a false negative.
+I will generally consider that intermittent negatives are uninformative on the duration of detectability, and hence exclude them from several analyses.
+
+A sample of 500 randomly selected infection episodes (selected from the episodes used in the analyses in this thesis) have their test results shown in \cref{biology-data:fig:episodes}.
+As can be seen, the number of tests within an episode, their length, and other characteristics vary greatly.
+\begin{figure}
+  \thisfloatpagestyle{empty}
+  \vspace{-5cm}
+  \makebox[\textwidth][c]{\includegraphics[height=.7\paperheight]{biology-data/STATS13734/data_viz}}
+  \caption[CIS infection episodes]{%
+    500 randomly selected infection episodes.
+    Shown are the negative tests before and after the outermost positive tests associated with the episode.
+    Red dots are negative tests, green dots are positive tests, and blue dots are void tests (where the result could not be determined; excluded from other analyses).
+  }
+  \label{biology-data:fig:episodes}
+\end{figure}
+
+The crucial features of an infection episode are the start and end times of the infection.
+Ignoring misclassified test results, latent start time of infection episode $j$, $B_j$, can be bounded as between the day after the last negative test, $l_j^{(b)}$, and the day of the first positive test, $r_j^{(b)}$.
+Similarly, the latent end time of infection episode $i$, $E_j$, can be bounded as between the day of the last positive test, $l_j^{(e)}$, and the day before the following negative test, $r_j^{(e)}$.
+
+\section{Duration distribution for simulation} \label{duration-ground-truth}
+
 To generate the durations, an assumption is required for the distribution of the duration of positivity.
 I base this assumption on the estimate from \cref{E-ATACCC} with an inflated tail to represent what is seen within CIS.
-The tail is modified based on an unpublished analysis of CIS data~\citePersonalComms{A.\ Sarah Walker}.
+The tail is modified based on a previously unpublished analysis of CIS data, without accounting for several of the issues previously described.
 Walker uses survival analysis to estimate the duration from CIS data.
@@
 The first 30 days of the distribution is proportional to the ATACCC estimate, with the rest proportional to this CIS-based estimates.
-% The CIS-based estimates are shifted 3 days in order to make this smooth; this counteracts the above bias of missing the start of the infection (for times above 30 days, the other bias, missing short infections, is negligible).
-ERR...
-I'M NOT DOING THIS ANYMORE OH NO!!
 Denote by $f_A(t)$ the distribution function estimated in \cref{E-ATACCC} and $f_C(t)$ that from the CIS-based estimates just derived.
@@
 
-\subsection{Implementation}
-
-Simulation was performed in R~4.2.0~\autocite{R-4-2-0} using tidyverse~2.0.0~\autocite{tidyverse} and targets~1.1.3~\autocite{targetsPackage}.
-
-Inference was implemented in Stan via RStan~2.21.8~\autocite{rstan2-21-8} using default settings.
-Convergence was assessed using Rhat and ESS as described in \cref{E-inc-prev:sec:MCMC}, and all runs checked for divergent transitions.
-
-The prior on the total number of infections was centred on the true value with a dispersion parameter of 1.
-This meant that the prior was very diffuse.
-
-\subsection{Results} \label{perf-test:sec:results}
-
-Repeating the simulation showed negligible variation in the results (not shown).
-This is caused by the large sample size (4800 detected infections).
-Therefore, only results from a single simulation are shown and discussed.
-
-All analyses converged.
-Rhat were below 1.01, all parameters had an effective sample size of 1000, and there were no divergent transitions.
-
-The true value of the survival function was generally well recovered (see \cref{perf-test:fig:survival-results}).
-This was especially true for the first 30 days of the distribution.
-
-The second-order random walk prior performed worst.
-It showed oversmoothing, especially in the later parts of the distribution.
-It smooths towards a linearly changing hazard; here, linearly decreasing (see \cref{perf-test:fig:hazard-results}).
-This means that the survival function decreases slower than an exponential distribution in the tail.
-
-The informative and vague independent priors performed similarly.
-The informative prior had tighter credible intervals, as expected.
-This prior was centred on the correct value for the initial part of the distribution.
-Therefore, the narrow credible intervals are unsurprising.
-For hazards after day 40, the informative and vague priors are the same.
-Therefore, the posterior distributions are almost identical.
-There remains a difference because the hazard estimates are partially dependent on earlier estimates.
-
-\begin{figure}
-  \centering \includegraphics{cis-perfect-testing/survival-results}
-  \caption[Comparison of survival function estimates under different priors]{Comparison of the posterior estimate of the survival distribution (pointwise median and 95\% CrI) when using different priors on the hazard.
-Results using each of the three different priors discussed in \cref{perf-test:sec:parameters-priors} are shown.
-y-axis has either natural (left) or log (right) scale.
-Only showing first 100 days of the distribution.
-\label{perf-test:fig:survival-results}}
-\end{figure}
-
-\begin{figure}
-  \centering \includegraphics{cis-perfect-testing/hazard-results}
-  \caption[Comparison of hazard estimates under different priors]{Comparison of the posterior estimate of the hazard (pointwise median and 95\% CrI) when using different priors on the hazard.
-Results using each of the three different priors discussed in \cref{perf-test:sec:parameters-priors} are shown.
-\label{perf-test:fig:hazard-results}}
-\end{figure}
-
-\section{Discussion} \label{perf-test:sec:discussion}
-
-This chapter shows that the model I propose can recover the true survival function well.
-However, the problem becomes substantially harder if false negatives are present, as we shall see in the next chapter.
-Unfortunately, false negatives do occur in the CIS data.
-
-Reasonable choices of the prior for the hazard are vague (although not flat) or informative.
-A smoothing prior performed poorly, although that may be influenced by the choice of hyperparameters.
-Reducing the amount of smoothing or type of smoothing may help.
-More flexible smoothers, such as a Gaussian Process for additional flexibility and smoothing and different time-scales~\autocite{saulGaussian}.
-
-Various extensions could be worth exploring.
-Most importantly, covariates (\eg age, immunity from vaccination and/or previous infection, or variant) could be included.
-Immunity in particular affects the duration substantially, although age and variant may also make a contribution~\autocite{hakkiOnset,russellWithinhost}.
-Covariates can be included with a variety of models, the simplest being a proportional hazards model.
-A major challenge here is that the calculation of $p_t$ is greatly complicated, because it is no longer linear in the survival function.
-
-The prior would likely be more important if the number of detected infections was smaller.
-Therefore, exploring its impact could be important for future surveys.
-A particular application would be estimation in the early stages of an epidemic of a novel pathogen.
-Here, little would be known about the pathogen, but gathering information with robust estimates of uncertainty is important to inform the response.
-
-I have assumed a prior of constant incidence and that incidence is independent between individuals.
-Prevalence in the CIS data was approximately constant over the period of interest (see \cref{E-biology-data:sec:positivity-results}).
-A sensitivity analysis, where an epidemic in exponential growth or decline was simulated (not shown), showed minimal impact on survival estimates.
-An incorrect assumption of constant incidence can lead to biased estimates~\autocite{degruttolaAnalysis}.
-Therefore, the assumption could be important in other contexts, especially if the incidence is changing rapidly.
-
-I have developed a flexible simulation and inference framework for this type of analysis.
-This framework could be used to simulate alternate study designs.
-The design of CIS was created on a very short timescale in March 2020, in response to the pandemic's rapid spread.
-Therefore, it is likely that there are more efficient designs.
-These could be more cost-effective.
-Improved cost-effectiveness could allow more rapid response to potential pandemics because the threshold for policy-makers to approve the study would be lower.
-A long-term, preparatory effort to develop a more efficient design in preparation of a future pandemic would be worthwhile.
-
-Other useful preparatory work would be developing reasonable priors.
-These could be based on seasonal viruses that are of the same family as those likely to cause future pandemics.
-For example, seasonal influenzas and coronaviruses.
-These would allow analyses such as those performed in these chapters to be performed more rapidly in a future pandemic scenario.
-
-\section{Conclusion} \label{perf-test:sec:conclusion}
-This chapter has shown the feasibility of estimating the survival function from a study with a CIS-like testing schedule.
-While the infrequent testing schedule is not ideal, appropriate modelling can recover the true survival function well.
-
-\ifSubfilesClassLoaded{
-  \listoftodos
-}{}
 \end{document}

### 2024-03-12 paper-surv-analysis.tex
rearrange bits of paper

@@
 
-The format requirements from JASA are at \url{https://www.tandfonline.com/journals/uasa20}.
+\textbf{Notes / thoughts on current version}
+\begin{itemize}
+    \item Some of the derivations can be moved to a supplement (cut 2--3 pages).
+    \item Need to rearranage the ATACCC stuff --- add some details of what I did (thesis ch 4) and move stuff from prior info section into supplement.
+    \item Probably don't need some of the figures (maybe 5--8)?
+    \item Should the sensitivity analyses (figs 13/14 be in the main text)?
+    \item The above changes make this about the right length.
+\end{itemize}
+
+The \textbf{format requirements from JASA} are at \url{https://www.tandfonline.com/journals/uasa20}.
 The most important parts are reproduced here.
@@
 
-Notes / thoughts on current version
-\begin{itemize}
-    \item Some of the derivations can be moved to a supplement (cut 2--3 pages).
-    \item Need to rearranage the ATACCC stuff --- add some details of what I did (thesis ch 4) and move stuff from prior info section into supplement.
-    \item Probably don't need some of the figures (maybe 5--8)?
-    \item Should the sensitivity analyses (figs 13/14 be in the main text)?
-    \item The above changes make this about the right length.
-\end{itemize}
+\tableofcontents
 
-Table of contents, cut for final version but useful for now to compare lengths of sections.
-
-\tableofcontents
+\todo[inline]{Cut table of contents for final version but useful for now to compare lengths of sections.}
 
@@
 The CIS had a household-based design inviting all individuals aged 2 and over from selected households.
-Once invited, an enrolment swab would be taken at the first visit followed by 4 further weekly visits (giving a total of 5 swabs on days 0, 7, 14, 21, 28 relative to enrolment) after which visits were monthly.
-As well as a swab to test for SARS-CoV-2, participants were asked to complete a questionnaire, and some participants were asked to provide blood samples.
+Once invited, an enrollment swab would be taken at the first visit followed by 4 further weekly visits (giving a total of 5 swabs on days 0, 7, 14, 21, 28 relative to enrollment) after which visits were monthly.
+Swabs were tested for the presence of SARS-CoV-2 using RT-PCR.
 In reality, visits were often not on this precise schedule, and occasionally missed.
@@
 
-Positive tests are grouped into \emph{infection episodes} which are assumed to originate from the same infection (procedure described in \cref{sec:episodes}).
+Positive tests are grouped into \emph{infection episodes} which are assumed to originate from the same infection (using the procedure described in \cref{sec:episodes}).
 I base my analysis on the 4,800 episodes that were first detected between 10th Oct 2020 and 6th Dec 2020 inclusive with negatives bounding the start and end time of the episode.
@@
 Episodes removed in this step are incorporated in the adjustment for undetected episodes.
+
+The crucial features of an infection episode are the start and end times of the infection.
+Ignoring misclassified test results, latent start time of infection episode $j$, $B_j$, can be bounded as between the day after the last negative test, $l_j^{(b)}$, and the day of the first positive test, $r_j^{(b)}$.
+Similarly, the latent end time of infection episode $i$, $E_j$, can be bounded as between the day of the last positive test, $l_j^{(e)}$, and the day before the following negative test, $r_j^{(e)}$.
 
@@
 Undetected events where there is no information on these events is known as \emph{truncation}.
-A situation that is particularly relevant is \emph{left truncation}, when only events lasting a minimum amount of time are observed.
+% A situation that is particularly relevant is \emph{left truncation}, when only events lasting a minimum amount of time are observed.
 There is a subtlety that makes the CIS different from standard left truncation.
@@
 
+In this section, I introduce a probabilistic model for the generation of false negatives.
+The simplest model for false negatives is a constant, non-zero probability of returning a negative test when detectable.
+That is the test sensitivity, $\psens$, is less than 1.
+
+A better model would be to allow the rate of false negatives to vary over the course of an episode.
+False negatives are more likely to occur when the viral load of an individual is low, because there is less virus in their body to sample.
+In \cref{E-ATACCC}, this mechanism is incorporated by viewing negative tests as a left-truncation of the observation noise (see \cref{E-ATACCC:sec:observation-modification}).
+This observation suggests that a model with a declining test sensitivity as a function of time since infection might be more suitable.
+
+The CIS data provides evidence for a test sensitivity declining over the course of an episode.
+\Cref{imperf-test:fig:bounding-cis-sensitivity} demonstrates this by using the following classification for test results.
+\begin{enumerate}
+    \item All positive results are true positives.
+    \item All intermittent negatives are false negatives.
+    \item The negative at the end of an episode is a possible false negative.
+    \item All other negatives are true negatives.
+\end{enumerate}
+Here, I am assuming that the probability of two false negative tests at the end of an episode is negligible.
+As test sensitivity is the proportion of individuals who should test positive that do so, it can be empirically calculated as number of true positives / (number of true positives + number of false negatives).
+Therefore, I calculate the test sensitivity for each day, considering the first positive result in an episode as day 0.
+As the latent start day of the infection episode is before the first positive result, day 0 reflects some unknown number of days into the infection episode.
+I bound the test sensitivity on each day by assuming either all or none of the possible false negatives (group 3) are in fact false negatives.
+% In \cref{imperf-test:fig:bounding-cis-sensitivity} we consider the values produced by assuming the negative following the last positive in an episode could be either a true or false negative as a function of time since the individual was detected (\ie: the first positive test in the episode).
+This generates broad bounds, although suggest a declining test sensitivity over the course of an episode (see \cref{imperf-test:fig:bounding-cis-sensitivity}).
+Small numbers of tests mean that the series is noisy near the start and the end, such as both bounds being 0 on some days after day 40.
+However, the general trend is clear.
+\begin{figure}
+  \includegraphics{cis-imperfect-testing/test-sens-bound}
+  \caption[Bounding test sensitivity using CIS data]{
+    Bounding the test sensitivity using CIS data as a function of time since the infection was detected.
+  }
+  \label{imperf-test:fig:bounding-cis-sensitivity}
+\end{figure}
+
+To assess the sensitivity of the procedure to a time-varying test sensitivity, it will be useful to simulate from a model with a varying test sensitivity as a function of days since the start of the infection episode.
+Through consideration of \cref{imperf-test:fig:bounding-cis-sensitivity}, and the interaction of false negatives due to viral load in \cref{E-ATACCC}'s results, I propose the following model.
+\begin{equation}
+  p_\text{sens}(t) = \begin{cases}
+    0.9 - \frac{0.9-0.5}{50}t &t \leq 50 \\
+    0.5 &t > 50
+  \end{cases}
+  \label{imperf-test:eq:variable-test-sensitivity}
+\end{equation}
+where $t$ is the number of days since infection.
+Therefore, this model consists of a linear decline in test sensitivity, before a constant minimum point.
+
+Under this model, each test result in a detectable individual is an iid Bernoulli random variable with a probability $\psens$ of producing a positive.
+As previously, tests in individuals without ongoing infection episodes are always negative.
+I modified the simulation study from \cref{E-perf-test:sec:simulation-study} to generate test results in this way.
+
+Following the modification, more single positive episodes are seen at lower values of $\psens$, with a value between 60\% and 80\% reproducing the number of single positive episodes observed in the CIS data (see \cref{imperf-test:fig:sim-single-pos}).
+This is expected: a lower $\psens$ means that multiple positive tests are less likely, and hence the probability of a single positive episode is higher.
+
+External information (see \cref{sec:ATACCC-analysis}) gives that $\psens$ is 95\% (95\% CrI: 93--96\%).
+The requirement for a much lower value of $\psens$ in the simulation study here suggests this model of false negatives is inadequate.
+
 \section{Related work}
@@
 The situation here requires both.
+
+Similar likelihoods have previously appeared in the literature~\autocite[e.g.][eq.\ (2)]{piresIntervalMisclassify}.
+However, this prior work was for singly interval censored data; incorporating the double interval censored nature of the CIS data will involve summing over the possible episode start times.
 
@@
 \begin{align}
-&p(\na \mid \ntot, \theta) \\
+p(\na \mid \ntot, \theta)
 &= p(\nnodet, \na\mid \ntot, \theta) \\
@@
 \begin{align}
-&p(\theta \mid \na) \\
+p(\theta \mid \na)
 &\propto p(\theta) p(\na \mid \theta) \\
@@
 
-\subsubsection{Generative model for false negatives} \label{imperf-test:sec:simulate}
-
-\todo[inline]{Probably the wrong place for this subsection, and can be edited for brevity a lot}
-
-In this section, I introduce a probabilistic model for the generation of false negatives.
-The simplest model for false negatives is a constant, non-zero probability of returning a negative test when detectable.
-That is the test sensitivity, $\psens$, is less than 1.
-
-Under this model, each test result in a detectable individual is an iid Bernoulli random variable with a probability $\psens$ of producing a positive.
-As previously, tests in individuals without ongoing infection episodes are always negative.
-I modified the simulation study from \cref{E-perf-test:sec:simulation-study} to generate test results in this way.
-
-Following the modification, more single positive episodes are seen at lower values of $\psens$, with a value between 60\% and 80\% reproducing the number of single positive episodes observed in the CIS data (see \cref{imperf-test:fig:sim-single-pos}).
-This is expected: a lower $\psens$ means that multiple positive tests are less likely, and hence the probability of a single positive episode is higher.
-
-External information (see \cref{sec:ATACCC-analysis}) gives that $\psens$ is 95\% (95\% CrI: 93--96\%).
-The requirement for a much lower value of $\psens$ in the simulation study here suggests this model of false negatives is inadequate.
-
-A better model would be to allow the rate of false negatives to vary over the course of an episode.
-False negatives are more likely to occur when the viral load of an individual is low, because there is less virus in their body to sample.
-In \cref{E-ATACCC}, this mechanism is incorporated by viewing negative tests as a left-truncation of the observation noise (see \cref{E-ATACCC:sec:observation-modification}).
-This observation suggests that a model with a declining test sensitivity as a function of time since infection might be more suitable.
-
-The CIS data provides evidence for a test sensitivity declining over the course of an episode.
-\Cref{imperf-test:fig:bounding-cis-sensitivity} demonstrates this by using the following classification for test results.
-\begin{enumerate}
-    \item All positive results are true positives.
-    \item All intermittent negatives are false negatives.
-    \item The negative at the end of an episode is a possible false negative.
-    \item All other negatives are true negatives.
-\end{enumerate}
-Here, I am assuming that the probability of two false negative tests at the end of an episode is negligible.
-As test sensitivity is the proportion of individuals who should test positive that do so, it can be empirically calculated as number of true positives / (number of true positives + number of false negatives).
-Therefore, I calculate the test sensitivity for each day, considering the first positive result in an episode as day 0.
-As the latent start day of the infection episode is before the first positive result, day 0 reflects some unknown number of days into the infection episode.
-I bound the test sensitivity on each day by assuming either all or none of the possible false negatives (group 3) are in fact false negatives.
-% In \cref{imperf-test:fig:bounding-cis-sensitivity} we consider the values produced by assuming the negative following the last positive in an episode could be either a true or false negative as a function of time since the individual was detected (\ie: the first positive test in the episode).
-This generates broad bounds, although suggest a declining test sensitivity over the course of an episode (see \cref{imperf-test:fig:bounding-cis-sensitivity}).
-Small numbers of tests mean that the series is noisy near the start and the end, such as both bounds being 0 on some days after day 40.
-However, the general trend is clear.
-\begin{figure}
-  \includegraphics{cis-imperfect-testing/test-sens-bound}
-  \caption[Bounding test sensitivity using CIS data]{
-    Bounding the test sensitivity using CIS data as a function of time since the infection was detected.
-  }
-  \label{imperf-test:fig:bounding-cis-sensitivity}
-\end{figure}
-
-To assess the sensitivity of the procedure to a time-varying test sensitivity, it will be useful to simulate from a model with a varying test sensitivity as a function of days since the start of the infection episode.
-Through consideration of \cref{imperf-test:fig:bounding-cis-sensitivity}, and the interaction of false negatives due to viral load in \cref{E-ATACCC}'s results, I propose the following model.
-\begin{equation}
-  p_\text{sens}(t) = \begin{cases}
-    0.9 - \frac{0.9-0.5}{50}t &t \leq 50 \\
-    0.5 &t > 50
-  \end{cases}
-  \label{imperf-test:eq:variable-test-sensitivity}
-\end{equation}
-where $t$ is the number of days since infection.
-Therefore, this model consists of a linear decline in test sensitivity, before a constant minimum point.
-
-I simulate data under this model by, for each test in a detectable individual, calculating $\psens(t-b)$, where $t$ is the day of the test and $b$ is the day the episode began.
-The result of the test is then a Bernoulli random variable, independent of all other considerations.
-Adapting the simulation study to incorporate this test sensitivity is much more similar to the data (see purple histogram in \cref{imperf-test:fig:sim-single-pos}).
-This model does not require low test sensitivities early in the infection, maintaining compatibility with the results in \cref{E-ATACCC}.
-
-\subsubsection{Incorporation into the survival analysis framework}
-
 Now we modify the survival framework to incorporate false negatives.
@@
 
-First, in \cref{imperf-test:sec:modifying-p_ia}, I modify $p_{ia}$, the probability of an episode with test schedule $i$ being admissible, to allow for the episode possibly being longer than observed.
-Then, in \cref{imperf-test:sec:modifying-p_iu}, I modify $p_{iu}$, the probability of an episode with test schedule $i$ being undetected, to allow for additional episodes being missed.
-
-\subsection{Modifying \texorpdfstring{$p_{ia}$}{pia}} \label{imperf-test:sec:modifying-p_ia}
+\subsubsection{Modifying \texorpdfstring{$p_{ia}$}{pia}} \label{imperf-test:sec:modifying-p_ia}
 
@@
 A mixture of these scenarios forms the episode's likelihood contribution, with the mixture probability determined by the test sensitivity.
-
-Similar likelihoods have previously appeared in the literature~\autocite[e.g.][eq.\ (2)]{piresIntervalMisclassify}.
-However, this prior work was for singly interval censored data; incorporating the double interval censored nature of the CIS data will involve summing over the possible episode start times.
 
@@
 
-\subsection{Modifying \texorpdfstring{$p_{iu}$}{piu}} \label{imperf-test:sec:modifying-p_iu}
+\subsubsection{Modifying \texorpdfstring{$p_{iu}$}{piu}} \label{imperf-test:sec:modifying-p_iu}
 

### 2024-03-12 cis-perfect-testing.tex
format equations better

@@
 \begin{align}
-p_{ia} &= \prob((b, e) \in \alpha_i \mid \theta) \\
-\alpha_i &= \{ (b, e) : l_{j(i)}^{(b)} \leq b \leq r_{j(i)}^{(b)} \wedge l_{j(i)}^{(e)} \leq e \leq r_{j(i)}^{(e)}\}.
+p_{ia} =& \prob((b, e) \in \alpha_i \mid \theta) \\
+\alpha_i =& \{ (b, e) : l_{j(i)}^{(b)} \leq b \leq r_{j(i)}^{(b)} \wedge l_{j(i)}^{(e)} \leq e \leq r_{j(i)}^{(e)}\}.
 \intertext{Suppressing the conditioning on $\theta$, this gives:}
 p_{ia}
-&= \prob \left( l_{j(i)}^{(b)} \leq B_{j(i)} \leq r_{j(i)}^{(b)}, l_{j(i)}^{(e)} \leq E_{j(i)} \leq r_{j(i)}^{(e)} \right) \\
-&= \prob \left( l_{j(i)}^{(e)} \leq E_{j(i)} \leq r_{j(i)}^{(e)} \mid l_{j(i)}^{(b)} \leq B_{j(i)} \leq r_{j(i)}^{(b)} \right) \prob \left( l_{j(i)}^{(b)} \leq B_{j(i)} \leq r_{j(i)}^{(b)} \right) \\
-&=\sum_{b = l_{j(i)}^{(l)}}^{r_{j(i)}^{(b)}} \prob \left( l_{j(i)}^{(e)} \leq E_{j(i)} \leq r_{j(i)}^{(e)} \mid B_{j(i)} = b \right) \prob \left(B_{j(i)} = b \right) \\
-&=\sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \prob \left( l_{j(i)}^{(e)} - b + 1 \leq D_{j(i)} \leq r_{j(i)}^{(e)} - b + 1 \right) \prob \left(B_{j(i)} = b \right) &\text{by def of $D_{j(i)}$} \\
-&=\sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_\theta(l_{j(i)}^{(e)} - b + 1) - S_\theta(r_{j(i)}^{(e)} - b + 1) \right) \prob \left(B_{j(i)} = b \right) &\text{by def of $S_\theta$} \\
-&\propto \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_\theta(l_{j(i)}^{(e)} - b + 1) - S_\theta(r_{j(i)}^{(e)} - b + 1) \right)
+=& \prob \left( l_{j(i)}^{(b)} \leq B_{j(i)} \leq r_{j(i)}^{(b)}, l_{j(i)}^{(e)} \leq E_{j(i)} \leq r_{j(i)}^{(e)} \right) \\
+=& \prob \left( l_{j(i)}^{(e)} \leq E_{j(i)} \leq r_{j(i)}^{(e)} \mid l_{j(i)}^{(b)} \leq B_{j(i)} \leq r_{j(i)}^{(b)} \right) \\
+   &\times\prob \left( l_{j(i)}^{(b)} \leq B_{j(i)} \leq r_{j(i)}^{(b)} \right) \\
+=& \sum_{b = l_{j(i)}^{(l)}}^{r_{j(i)}^{(b)}} \prob \left( l_{j(i)}^{(e)} \leq E_{j(i)} \leq r_{j(i)}^{(e)} \mid B_{j(i)} = b \right) \prob \left(B_{j(i)} = b \right) \\
+=& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \prob \left( l_{j(i)}^{(e)} - b + 1 \leq D_{j(i)} \leq r_{j(i)}^{(e)} - b + 1 \right) \prob \left(B_{j(i)} = b \right) &\text{by def of $D_{j(i)}$} \\
+=& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_\theta(l_{j(i)}^{(e)} - b + 1) - S_\theta(r_{j(i)}^{(e)} - b + 1) \right) \prob \left(B_{j(i)} = b \right) &\text{by def of $S_\theta$} \\
+\propto& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_\theta(l_{j(i)}^{(e)} - b + 1) - S_\theta(r_{j(i)}^{(e)} - b + 1) \right)
 \end{align}

### 2024-03-13 conclusion.tex
start writing conclusion

\documentclass[thesis.tex]{subfiles}

\begin{document}
\ifSubfilesClassLoaded{
    \setcounter{chapter}{7}
}

\chapter{Conclusions} \label{conclusion}

Pandemics will continue to pose a significant threat to public health for the foreseeable future.
Therefore, there is a public health need to respond to pandemics.
Such a response needs to be informed by monitoring.
In this thesis, I have investigated one method of monitoring pandemics: prevalence surveys.

\section{Main findings}

In \cref{E-intro:sec:aims}, I posed the central question for this thesis: can prevalence surveys alone be used to infer incidence and transmission, and if so how?

In \cref{E-transmission}, I showed that this it is possible by appying two models.
The differing models, one statistical in nature and one mechanistic, each had their own strengths and weaknesses.
The statistical approach, based on the backcalculation framework, estimates incidence and makes fewer assumptions about the infection process.
However, the estimates are very uncertain.
The mechanistic approach, within the compartmental modelling framework, reduces the uncertainty and allows estimation of the components of transmission by making stronger assumptions about the infection process.

These results highlighted the heterogeneity across regions and age groups in England.

These models relied on the theoretical work of \cref{E-inc-prev}.
This work developed the backcalculation framework for the context of prevalence surveys for SARS-CoV-2.
The major insight was the justification of the deterministic model framework and the importance of the distribution of the duration of positivity.

In \cref{E-perf-test,E-imperf-test}, I developed methodology to estimate the distribution of the duration of positivity from the CIS prevalence survey.
Therefore, showing that this important parameter can be estimated from prevalence surveys.
While I did make use of external estimates, from \cref{E-ATACCC}, this was to address weaknesses of the CIS study design that could be addressed in future studies (see \cref{conclusion:sec:future-work}).

Both approaches assumed that the distribution of the duration of positivity was perfectly known.

Estimating this distribution was the focus of the majority of this thesis.

\section{Future work} \label{conclusion:sec:future-work}

The feasibility of jointly modelling duration and incidence was not investigated in this thesis.
The model for duration in \cref{E-perf-test,E-imperf-test} estimated the total number of infections in the cohort but assumed these occurred uniformly over the duration of the survey.
Relaxing the uniformity assumption, would give incidence estimation.
The issue of including covariates to control for non-response bias, encountered in \cref{E-transmission}, as well as the parameterisation of the shape of the incidence curve would be major challenges in this extension.
Modelling frameworks could be adopted from the existing literature~\autocite[e.g.][]{taffeJoint,haySerosolver} used approximations (\eg empirical Bayes), smaller sample sizes, or high-performance computing to address these challenges.
However, the CIS is larger and incidence estimation desired at a more granular level than previously encountered.
Furthermore, high-performance computing cannot be used within the SRS.

A common theme in this thesis was computational challenges.
The challenges could be addressed by more efficient MCMC algorithms, approximate methods, or alternate computing methods (\eg the use of GPUs).
Alternate computing methods would require appropriate hardware within the SRS, which should be considered when choosing the design of future TREs.

The CIS was a household-based survey, but the household structure was not considered in this thesis.
The household-based design is likely to induce clustering in results.
Therefore, the uncertainty in my results might be underestimated.
While the beta-binomial likelihood used in \cref{E-transmission} should capture this, it does not consider the clustering between different strata.
Other chapters did not consider clustering at all.
Similarly, the treatment of a longitudinal survey as a cross-sectional survey in \cref{E-perf-test} may have underestimated the uncertainty in the results.
More sophisticated incorporation of clustering, for instance using random effects for households and/or individuals, could be considered in future work.

The challenges in estimating the distribution of the duration of positivity originated in the design of the studies used.
Exploring the design of future studies to better estimate the distribution of the duration of positivity would be beneficial.
A particularly promising direction would be intensive follow-up of individuals following a positive test (\eg daily testing) until they have two or more negative tests.
Intensive follow-up would increase the precision by which the end of the infection episode is known.
Furthermore, it reduces the impact of false negatives because they are likely to be followed by a true positive.
I have developed software that could be used to perform simulation studies of alternative designs.

Other design questions are also posed by the work in this thesis.
For example, whether a representative sample of the population is optimal, or if certain groups should be oversampled.
Certain groups, such as younger individuals, were more likely to be infected.
It is plausible that sampling in proportion to incidence would be beneficial.
This is especially true if policy questions (\eg closing of schools) are of importance.
These questions could be investigated through simulation studies.

\ifSubfilesClassLoaded{
  \listoftodos
}{}

\end{document}

### 2024-03-14 conclusion.tex
first draft of conclusion

@@
 
-In \cref{E-transmission}, I showed that this it is possible by appying two models.
-The differing models, one statistical in nature and one mechanistic, each had their own strengths and weaknesses.
-The statistical approach, based on the backcalculation framework, estimates incidence and makes fewer assumptions about the infection process.
-However, the estimates are very uncertain.
+In \cref{E-transmission}, I found that this it is possible, developing two models that can infer incidence and transmission from prevalence surveys.
+One model was statistical in nature and the other mechanistic.
+The statistical approach, based on the backcalculation framework, estimates incidence and makes fewer assumptions about the infection process; however, the estimates are very uncertain.
 The mechanistic approach, within the compartmental modelling framework, reduces the uncertainty and allows estimation of the components of transmission by making stronger assumptions about the infection process.
-
 These results highlighted the heterogeneity across regions and age groups in England.
 
-These models relied on the theoretical work of \cref{E-inc-prev}.
+The work in \cref{E-transmission} relied on the theoretical work of \cref{E-inc-prev}.
 This work developed the backcalculation framework for the context of prevalence surveys for SARS-CoV-2.
@@
 
-In \cref{E-perf-test,E-imperf-test}, I developed methodology to estimate the distribution of the duration of positivity from the CIS prevalence survey.
-Therefore, showing that this important parameter can be estimated from prevalence surveys.
-While I did make use of external estimates, from \cref{E-ATACCC}, this was to address weaknesses of the CIS study design that could be addressed in future studies (see \cref{conclusion:sec:future-work}).
+These models both required knowledge of the distribution of the duration of positivity.
+Estimating this distribution was the focus of the majority of this thesis.
+In \cref{E-perf-test,E-imperf-test}, I showed how this quantity could be estimated the CIS prevalence survey.
+This did require external information from the ATACCC study (the results of \cref{E-ATACCC}), demonstrating that the CIS's design is not sufficient for estimating transmission.
+However, future surveys may be able to avoid this issue (see \cref{conclusion:sec:future-work}).
 
-Both approaches assumed that the distribution of the duration of positivity was perfectly known.
+Prior to this thesis, the duration of positivity in the general population was not well understood (see \cref{E-intro:sec:previous-duration-estimates}).
+A particularly understudied area was quantifying the shape and size of the tail of the distribution, due to limited follow-up of individuals who test positive.
 
-Estimating this distribution was the focus of the majority of this thesis.
+Overall, I have found that representative prevalence surveys are a promising approach for pandemic surveillance.
+A well-designed survey is sufficient to estimate all of the quantities of interest (as described in \cref{E-intro:sec:metrics}), except severity.
 
@@
 The model for duration in \cref{E-perf-test,E-imperf-test} estimated the total number of infections in the cohort but assumed these occurred uniformly over the duration of the survey.
-Relaxing the uniformity assumption, would give incidence estimation.
+If, rather than assuming uniformity, the time of incidence was estimated, this would be a form of incidence estimation.
 The issue of including covariates to control for non-response bias, encountered in \cref{E-transmission}, as well as the parameterisation of the shape of the incidence curve would be major challenges in this extension.
-Modelling frameworks could be adopted from the existing literature~\autocite[e.g.][]{taffeJoint,haySerosolver} used approximations (\eg empirical Bayes), smaller sample sizes, or high-performance computing to address these challenges.
+The basis of such a model could be existing modelling frameworks~\autocite[e.g.][]{taffeJoint,haySerosolver}.
+These frameworks addressed the challenges by: taking approximations (\eg empirical Bayes), having smaller sample sizes, or using high-performance computing.
 However, the CIS is larger and incidence estimation desired at a more granular level than previously encountered.
@@
 These questions could be investigated through simulation studies.
+In constructing such studies, it will be important to consider how these considerations might be different for different pathogens (\eg influenza, the most common source of respiratory pandemics) and tests (\eg rapid antigen tests, which have a shorter duration of positivity).
+
+This thesis did not consider detection or early surveillance of emerging pandemics.
+Initial uncertainty about local transmission in the UK may have led to suboptimal decisions at the start of 2020~\autocite{pellisChallenges}.
+Community testing was not widely available at this stage of the pandemic, limiting the knowledge of the extent of local (as opposed to imported) transmission~\autocite{whittyTechReportCOVID}.
+A representative survey of the population could have provided this information.
+However, challenges in constructing a survey with the power to detect low levels of transmission, as well as the time required to set-up such a survey, would need to be addressed.
+
+A final consideration is the estimation of the severity of a pathogen.
+Severity of a pathogen is crucial for risk assessment, and can change over time~\autocite[e.g.][]{kirwanSeverityTrends}.
+Two approaches could be taken to using a prevalence surveys for this purpose.
+First, a cohort approach where the outcomes of individuals who test positive are followed up.
+The low number of positive cases in the CIS would make this challenging.
+In addition, the data would need to be linked to relevant records.
+Second, a transmission model, such as the one in \cref{E-SEIR}, could jointly model severe events and prevalence.
+Such an approach has previously been used~\autocite{daviesAssociation,ironsEstimating,knockKey,nicholsonImproving,pooleyEstimation,birrellRTM2} but without the sophistication of the observation model developed in \cref{E-SEIR:sec:observation}.
 

### 2024-03-14 thesis.tex
Resolve to dos
fix subscripts for survival analysis
restructure sec 7.1
more detail in toc / navigation
first draft of conclusion

@@
 \newcommand{\sched}{\mathbb{T}}
+\newcommand{\nsched}{n_{\sched}}
 \newcommand{\rate}{\mathfrak{R}}
@@
 
-\setcounter{tocdepth}{1}
+\setcounter{tocdepth}{2}
 
@@
 \frontmatter{}
+
+\todo[inline]{Mention use of early version of this work in ONS reports}
 
@@
 
-\chapter{Conclusion} \label{conclusion}
-Household structure is not included in this model.
-The household-based design of the survey is likely to induce clustering in results, increasing the variance of the measured prevalence.
-The beta-binomial likelihood structure should capture some of this variance within each strata\todo{ref beta-binomial as a model for clustering}, but not between strata.
+\subfile{conclusion}
 
@@
 \cref{inc-prev:eq:forward-substitute}
+\cref{inc-prev:eq:EPt-matrix}
+\cref{inc-prev:eq:Smatrix}
 \listoftodos

### 2024-03-15 biology-data.tex
finish info on INLA

@@
 The terms are as follows.
+\todo[inline]{Check I've written the priors correctly}
 \begin{itemize}
@@
 I use a beta-binomial distribution, as discussed in \cref{biology-data:sec:clustering}.
-    \item $\beta$ is a global intercept term with the prior $\beta \dist \N(0, 1/0.001^2)$.
-    \item $\beta_s$ is a fixed effect for sex, with $\beta_s = 0$ for males and estimated for females.
+    \item $\beta$ is a global intercept term with the vague prior $\beta \dist \N(0, 1/0.001^2)$.
+    \item $\beta_s$ is a fixed effect for sex, with $\beta_s = 0$ for males and estimated for females. The prior for $\beta_s$ is vague, $\beta_s \dist \N(0, 1/0.001^2)$.
     \item $\beta_{w(e)}$ is a fixed effect for White or non-White ethnicity.
-$\beta_{w(e)} = 0$ if $e$ corresponds to White ethnicity and $\beta_{w(e)} = \beta_{n}$ otherwise (\ie non-White ethnic groups).% $\beta_n$ and $\beta_f$ have the default INLA prior for fixed effects\todo{check what this is}.
+$\beta_{w(e)} = 0$ if $e$ corresponds to White ethnicity and $\beta_{w(e)} = \beta_{n}$ otherwise (\ie non-White ethnic groups).
+      The prior for $\beta_n$ is vague, $\beta_n \dist \N(0, 1/0.001^2)$.
+      % $\beta_n$ and $\beta_f$ have the default INLA prior for fixed effects\todo{check what this is}.
     \item $u_a$ and $v_e$ are independent Gaussian random effects for age and ethnic group respectively.
@@
     \item $w_t$ is a second-order random walk, that is $w_t = 2w_{t-1} - w_{t-2} + \epsilon_t$ where $\epsilon_t \dist \N(0, \sigma_w^2)$.
-    \item $l_{a,t}$ is an age-group specific second-order random walk. That is $l_{a,t} = 2l_{a,t-1} - l_{a,t-2} + \epsilon_{a,t}$ where $\epsilon_{a,t} \dist \N(0, \sigma_l^2)$.
+      $\sigma_w$ has a vague prior, $1/\sigma_w \dist \GamDist(1, 5 \times 10^{-5})$.
+    \item $l_{a,t}$ is an age-group specific second-order random walk.
+      That is $l_{a,t} = 2l_{a,t-1} - l_{a,t-2} + \epsilon_{a,t}$ where $\epsilon_{a,t} \dist \N(0, \sigma_l^2)$.
+      $\sigma_l$ has a vague prior, $1/\sigma_l \dist \GamDist(1, 5 \times 10^{-5})$.
 \end{itemize}
-All parameters have the default R-INLA priors (see \cref{E-transmission:sec:INLA:priors}), except where stated above.
-For these parameters, the large amount of data means that the prior has little influence on the posterior distribution.
+For parameters given vague priors, the large amount of data means that the prior has little influence on the posterior distribution.
 

### 2024-03-15 transmission-appendix-INLA.tex
notes on joint INLA posterior
finish info on INLA

--- 2024-02-14 transmission-appendix-INLA.tex
@@
 
+The standard implementation of INLA is the R-INLA package~\autocite{RINLA}.
+
+\section{Latent Gaussian model}
+
+INLA requires a LGM (latent Gaussian model).
+An LGM is defined in terms of a vector of hyperparameters, $\vec{\varphi}$; a deterministic transformation of the hyperparameters into covariance matrix, $\matr{\Sigma}(\vec{\varphi})$; a vector of latent variables, $\vec{x}$; a vector of $n$ data points, $\vec{y} = [y_1, \dots, y_n]^T$; and $n$ linear predictors $\vec{\eta} = [\eta_1, \dots, \eta_n]^T$.
+To be an LGM, $\vec{x}$ must be a latent Gaussian field, that is $\vec{x} \mid \vec{\varphi} \dist \N(\vec{0}, \matr{\Sigma}(\vec{\varphi}))$; and the likelihood must have the conditional independence property $\pi(\vec{y} \mid \vec{\varphi}, \vec{x}) = \prod_j \pi(y_j \mid \eta_j(\vec{x}), \vec{\varphi})$.
+% The Gaussian structure allows the use of the approximations in \cref{transmission:sec:INLA:inference}.
+% The conditional independence structure means that the precision matrix, $\matr{Q} = \matr{\Sigma}^{-1}$, is sparse.
+
+For INLA to perform well, \ie be compuutationally efficient and an accurate approximation, $\vec{\theta}$ should be of low dimension, \textcite{martinoINLAtutorial} recommend $< 15$; and the precision matrix, $\matr{Q}$, should be sparse.
+The low dimensionality of $\vec{\theta}$ is because the method requires numerical integration over it.
+The sparsity of $\matr{Q}$ allows for efficient computation.
+Sparsity is introduced by conditional independence structures of the latent field that are present in many models.
+For example, in the classic random effect setting, all random effects are independent conditional on the hyperparameters.
+
 \section{Inference procedure} \label{transmission:sec:INLA:inference}
+
+The core of INLA is an approximation for the posterior distribution of the hyperparameters $\pi(\vec{\theta} \mid \vec{y})$ and the marginal posterior distribution of the latent field, $\pi(x_i \mid \vec{y})$.
+
+\subsection{Gaussian likelihood}
+
+Start by the case in which the likelihood, $\vec{y} \mid \vec{x}, \vec{\theta}$, is Gaussian.
+In this case, as $\vec{x} \mid \vec{\theta}$ is also Gaussian, the distributions of $\vec{x} \mid \vec{y}, \vec{\theta}$ is Gaussian too.
+The Gaussian form of $\vec{x} \mid \vec{y}, \vec{\theta}$ follows from the conjugate property of the Gaussian distribution.
+This distribution can easily be found analytically.
+
+The posterior density of $\vec{\theta}$, $\vec{\theta }\mid \vec{y}$, is not Gaussian.
+Approximate it by the distribution $\tilde\pi(\vec{\theta} \mid \vec{y})$ defined as:
+\begin{align}
+\tilde\pi(\vec{\theta} \mid \vec{y})
+&\propto \frac{\pi(\vec{x^*}, \vec{y} \mid \vec{\theta}) \pi(\vec{\theta})}{\pi(\vec{x^*} \mid \vec{y}, \vec{\theta})}
+\end{align}
+where $\vec{x^*} = \argmax_{\vec{x}} \pi(\vec{x} \mid \vec{y}, \vec{\theta})$.
+All of the densities on the right-hand side can easily be found analytically.
+
+Assume that $\vec{\theta} \mid \vec{y}$ is unimodal, which is very likely for reasonable priors on $\vec{\theta}$.
+Find its mode via optimisation and explore its area, until the density is negligible.
+By integrating over this area, the normalising constant can be found.
+Hence, the approximate posterior density of the hyperparameters is known.
+
+Finally, the marginal posterior density of $x_i$ is found by integrating over $\vec{\theta}$:
+\begin{align}
+\pi(x_i \mid \vec{y})
+&\approx \int \pi(x_i \mid \vec{\theta}, \vec{y}) \tilde\pi(\vec{\theta} \mid \vec{y}) d\vec{\theta}.
+\end{align}
+which can be computed using numerical integration over $\tilde\pi(\vec{\theta} \mid \vec{y})$, previously calculated.
+
+\subsection{Non-Gaussian likelihood}
+
+Now, the INLA needs extending to non-Gaussian, but known, likelihoods for $\vec{y} \mid \vec{x}, \vec{\theta}$.
+The density $\pi(\vec{x}, \vec{y} \mid \vec{\theta})$ is not Gaussian, but still known.
+However, $\vec{x} \mid \vec{y}, \vec{\theta}$ is not Gaussian, and its density is not known.
+
+INLA assumes that $\vec{x} \mid \vec{y}, \vec{\theta}$ is approximately Gaussian, and therefore takes a Laplace approximation to the distribution.
+The greater the smoothing, the better the approximation.
+This is because, if $\vec{x} \mid \vec{y}, \vec{\theta}$ is then dominated by the prior $\vec{x} \mid \vec{\theta}$ and the prior is Gaussian.
+Exploiting the structure of the problem, in particular the conditional independence structure, the Laplace approximation can be computed efficiently.
 
@@
 
-\section{Default priors} \label{transmission:sec:INLA:priors}
+Various methods exist to sample from the joint posterior of the hyperparameters and the latent variables, based on the INLA approximation.
+The joint posterior is not approximated by default.
+I used the one implemented in the R-INLA function \texttt{inla.posterior.sample}, described in \textcite[section 4]{chiuchioloJoint}.
+
+This method incorporates the first three moments of each marginal distribution, and uses a skew Gaussian copula to approximate the joint distribution.
 

### 2024-03-18 biology-data.tex
Overleaf comments

--- 2024-03-11 biology-data.tex
@@
 
-PCR consists of a series of cycles, each cycle containing a three-step process~\autocite{powledgePCR,garibyanPCR}.
+At the start of the PCR process, four substances are required: template DNA, nucleotides, primers, and DNA polymerase~\autocite{garibyanPCR}.
+The \emph{template DNA}, in this context, is the sample which is being tested~\autocite{caseTemplate}.
+\emph{Nucleotides} or \emph{bases} are the building blocks of DNA.
+There are four nucleotides in DNA (A, C, G, T) each of which forms a pair, known as a \emph{base pair}, with its \emph{complementary} nucleotide (A with T, and G with C)~\autocite{batesBase}.
+The choice of \emph{primers} specifies the exact DNA which will be amplified; they are the complementary DNA sequence to the start and end of the DNA sequence to be amplified~\autocite{garibyanPCR}.
+DNA \emph{polymerase} is an enzyme required to link individual nucleotides together~\autocite{garibyanPCR}.
+
+The process is a series of cycles, each cycle containing three steps~\autocite{powledgePCR,garibyanPCR}.
 In the first step, \emph{denaturation}, the DNA is heated to separate the two strands of DNA.
-In the second step, \emph{annealing}, the temperature is lowered to allow primers to bind to the DNA.
-Primers are short sequences of DNA that are complementary to the DNA sequence of interest.
-By choosing the primers, the PCR process targets a portion of DNA starting and ending with a specified sequence.
-In the final step, \emph{extension}, the temperature is raised to allow the enzyme DNA polymerase to extend the primers and hence replicate the DNA sequence of interest.
+In the second step, \emph{annealing}, the temperature is lowered to allow the primers to bind to the DNA.
+In the final step, \emph{extension}, the temperature is raised, causing the DNA polymerase to replicate the DNA sequence between the primers.
 In the course of such a cycle, the amount of DNA doubles.
@@
 
-Quantitative real-time (RT-)PCR gives a measurement of the amount of the target DNA sequence present in the original sample~\autocite{yangPCRdiagnostics}.
+\emph{Quantitative real-time (RT-)PCR} gives a measurement of the amount of the target DNA sequence present in the original sample~\autocite{yangPCRdiagnostics}.
 After each PCR cycle, the amount of target DNA is measured and compared to a pre-derived standard.
@@
 The number of cycles required to reach detection is known as the \emph{cycle threshold} (Ct) value.
+Originally, the amount of target DNA was only measured after the completion of all cycles, hence the name real-time.
 Quantitative real-time PCR have a number of other practical advantages over previous PCR techniques (\eg speed); see \textcite{yangPCRdiagnostics,valasekPower} for more details.
@@
 
-The Ct value decreases as the amount of DNA present in the original sample increases~\autocite{hanRTPCR}.
-If there is more RNA present, more DNA is created, and hence fewer cycles are required to reach detection.
-As the number of DNA copies doubles each cycle, the Ct value decreases increases linearly on the log-scale each cycle (see \cref{biology-data:fig:ct-calibration}).
+% The Ct value decreases as the amount of DNA present in the original sample increases~\autocite{hanRTPCR}.
+If there is more RNA present, proportionally more DNA is created in reverse transcription, and hence fewer cycles are required to reach detection; that is, the Ct value is lower~\autocite{hanRTPCR}.
+The number of DNA copies doubles each cycle, therefore the log Ct value decreases linearly with the quanitiy of RNA or DNA in the original sample (see \cref{biology-data:fig:ct-calibration}).
 A \emph{standard curve} translates the Ct value to viral load for a given test.
@@
 However, viewing the number of RNA genome copies in an individual as a latent process this can then be defined as the time when the number of RNA genome copies is above the limit of detection of the RT-PCR test (defined in \cref{biology-data:sec:measurement-error}).
-While this definition is not perfectly precise, it is clear enough for the purposes it used in this thesis.
+% While this definition is not perfectly precise, it is clear enough for the purposes it used in this thesis.
 
@@
 The \emph{analytical sensitivity} of the test is the minimum number of copies of the target RNA material in the sample that the test can detect, typically expressed as the \emph{limit of detection}~\autocite{bustinMIQE}.
-The limit of detection is the lowest concentration of the target RNA material that can be reliably detected using the procedure.
+The limit of detection is the lowest concentration of the target RNA material that can be reliably detected using the procedure; below this level there is a substantial (generally, >5\%) chance that errors in the process cause the amplification procedure to fail~\autocite{forootanLOD}.
 This is different to the \emph{clinical sensitivity} of the test, the ability of the test to detect a detectable individual.

### 2024-03-18 conclusion.tex
Overleaf comments

@@
 One model was statistical in nature and the other mechanistic.
-The statistical approach, based on the backcalculation framework, estimates incidence and makes fewer assumptions about the infection process; however, the estimates are very uncertain.
-The mechanistic approach, within the compartmental modelling framework, reduces the uncertainty and allows estimation of the components of transmission by making stronger assumptions about the infection process.
+The statistical approach, based on the backcalculation framework, provide estimates of incidence and makes fewer assumptions about the infection process; however, the estimates are very uncertain.
+The mechanistic approach, based on a SIR-type compartmental model, reduces the uncertainty and allows estimation of the components of transmission by making stronger assumptions about the infection process.
 These results highlighted the heterogeneity across regions and age groups in England.
@@
 The feasibility of jointly modelling duration and incidence was not investigated in this thesis.
-The model for duration in \cref{E-perf-test,E-imperf-test} estimated the total number of infections in the cohort but assumed these occurred uniformly over the duration of the survey.
+The model for duration in \cref{E-perf-test,E-imperf-test} allowed estimation of the total number of infections in the cohort but assumed these occurred uniformly over the duration of the survey.
 If, rather than assuming uniformity, the time of incidence was estimated, this would be a form of incidence estimation.

### 2024-03-18 incidence-prevalence.tex
add todo from warning

@@
 
-This chapter defines incidence and prevalence, and the relationship between these quantities.
+This chapter formally defines incidence and prevalence, and the relationship between these quantities.
 In the course of explaining this relationship, it will become clear that the duration of RT-PCR positivity, and its distribution in the population, is the vital parameter which will allow estimation of incidence from prevalence, a major aim of this thesis.
@@
 I then move on to explaining the three processes that are involved: the \emph{infection process}, the \emph{prevalence process} and the \emph{observation process}.
+The mathematical description of are an infection model, disease model, and observation model respectively.
 
@@
 A standard extension to the SIR model is to instead assume a gamma distribution~\autocite[94]{keelingModeling,andersonSpread}.
+\todo{check this citation is formatted correctly}
 This is done by adding multiple infectious compartments, shown in \cref{SEIR:fig:SIR-gamma}.

### 2024-03-18 introduction.tex
first draft of conclusion
Updates from Overleaf

--- 2024-03-09 introduction.tex
@@
 
-COVID-19 (coronavirus disease 2019), the disease caused by the virus SARS-CoV-2 (severe acute respiratory syndrome coronavirus 2), is estimated to have killed approximately 15 million in the acutest phase (to the end of 2021) and 30 million in total~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
+COVID-19 (coronavirus disease 2019), the disease caused by the virus SARS-CoV-2 (severe acute respiratory syndrome coronavirus 2), is estimated to have killed approximately 15 million in the most acute phase (to the end of 2021) and 30 million in total~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
 History suggests severe pandemics are not rare (see \cref{intro:fig:pandemic-timeline}).
@@
 
-\section{Surveillance metrics}
+\section{Surveillance metrics} \label{intro:sec:metrics}
 
@@
 In the early stages of an epidemic caused by a novel pathogen, the number of individuals with immunity is negligible, and hence $\R$ describes this part of the epidemic well.
-$\R$ has a threshold value of 1: an outbreak has the potential to cause an epidemic if and only if $\R < 1$~\autocite[76]{diekmannMathematical}.
+$\R$ has a threshold value of 1: an outbreak has the potential to cause an epidemic if and only if $\R > 1$~\autocite[76]{diekmannMathematical}.
 
@@
 Incidence informs the short-term future burden of the disease, for example, the future number of severe events (hospitalisations or deaths).
-To translate incidence into the burden of the disease, a measure of severity is required.
-For example, the IFR (infection fatality rate), the proportion of infected individuals who die.
+To translate incidence into the burden of the disease, a measure of severity is required; for example, the IFR (infection fatality rate), the proportion of infected individuals who die.
 Estimation of severity is beyond the scope of this thesis.
@@
 \emph{Transmission} refers to the process by which an infectious disease generates new infections.
-Estimating transmission refers to estimating the parameters of this process, the infection process, including its components.
+Estimating transmission refers to inferring the parameters of this process, including its components.
 The components of transmission (\eg number of contacts per day and age-stratified rates) inform how interventions can be targeted.
@@
 Whether a given individual is \emph{prevalent} (has the disease) is often measurable through a diagnostic test.
-Therefore, the prevalence of the population can be estimated with a prevalence survey.
-However, these are very expensive to power appropriately; SARS-CoV-2 is unusual in that its impact is so large that the UK government did this.
-Prevalence and incidence are strongly related, with the relationship determined by by the duration of an infection (see \cref{E-inc-prev}).
+Therefore, the prevalence of the population can be estimated with a prevalence survey, although these are typically very expensive to power appropriately.
+Prevalence and incidence are strongly related, via the duration of an infection (see \cref{E-inc-prev}).
 
@@
 
-However, incidence and transmission and prevalence, are almost never directly observed.
-Therefore, they need to be inferred from other data sources.
+Collecting data which directly informs inference of the quantities in this section is expensive or impossible.
+Hence, they are typically inferred from other data sources through modelling.
 
@@
 
-\todo[inline]{Not sure this section (data sources) needs to exist, particularly in this location}
-\todo[inline]{If this section kept, probably want to cite some things on generic data for pandemic surveillance}
-
 The data sources used for outbreak surveillance are often observational in nature.
 Therefore, they suffer from biases that can be hard to understand and quantify.
-For example, community testing which relies on individuals presenting for a test, hence biased by the willingness of individuals to be tested.
+Here I discuss three types of data which are useful for surveillance.
+Incidence-type and prevalence-type data can directly inform estimation of incidence and transmission, however, this estimation almost always relies on prior knowledge of epidemiological and/or biological quantities.
+These quantities can be estimated from other studies, \eg cohort studies.
 
-Well-designed studies can provide direct estimates of incidence and/or prevalence in specific cohorts.
+\subsection{Incidence-type data}
+
+Incidence-type data originate from observations of the consequences of a disease, \eg the development of symptoms or admission to hospital.
+This type of data is typically affected by \emph{ascertainment bias}: the proportion of infections which are counted can vary over time or within different subgroups of the population~\autocite[chapter 9]{lashModern}.
+
+An example of incidence-type data is community testing.
+This relies on individuals presenting for a test, hence is biased by individuals' ability and willingness to be tested.
+
+More severe symptoms are typically less affected by ascertainment bias.
+An extreme example is the development of AIDS following infection by HIV.
+Here, the disease is so severe that it is extremely unlikely for an individual to develop the disease without seeking healthcare~\autocite{evansCompleteness}.
+
+However, severe symptoms can be rare in some population subgroups (\eg young, otherwise healthy individuals infected with SARS-CoV-2).
+This means that the observations do not provide much information on incidence in those subgroups.
+
+\subsection{Prevalence-type data}
+
+Prevalence-type data originate from observations of the number or proportion of a population in a specific disease state at a specific time.
+For example, the number of hospital beds occupied with individuals with the disease or the proportion of the population with a marker of prior infection.
+The issues associated with incidence-type data also apply to prevalence-type data.
+
+Prevalence surveys are one form of prevalence-type data that can be unbiased.
+Here, a representative proportion of the population (or a subpopulation of interest) are tested for a marker of current infection.
+For many diseases, prevalence is sufficiently low that this is prohibitively expensive to power.
+
+However, \emph{seroprevalence} surveys are more feasible.
+Seroprevalence is the proportion of individuals in the population which have antibodies to the disease.
+The presence of antibodies indicates that the individual was previously infected.
+Many more individuals have ever been infected than are currently infected, therefore, seroprevalence surveys can be powered much more cheaply.
+However, seroprevalence is a delayed measure because there is a lag between an individual being infected and them developing antibodies.
+
+\subsection{Cohort studies}
+
+Other epidemiological and/or biological quantities are normally required to estimate transmssion.
+These depend on the model type, but could include the \emph{generation interval} (the time between an individual being infected and going on to infect further individuals) or duration of different disease stages.
+Cohort studies, where a group of individuals are followed over time, are a study design which can estimate many of these quantities.
+
+Well-designed cohort studies can provide direct estimates of incidence and/or prevalence in the cohort.
 However, recruiting a representative sample of the population is challenging, and the results of these studies are often not generalisable to the whole population.
 Furthermore, for these studies to be appropriately powered, they often require a large cohort, and hence are expensive to run.
-A unique feature of the COVID-19 pandemic is that its impact was so large that the UK Government were willing to fund such studies.
-The availability of this data motivates the work in this thesis.
 
-\section{Infection process model}
+\section{Infection models}
 
-The first component required to estimate incidence and transmission is a model of the infection process, a mathematical description of how incidence arises.
+The first component required to estimate incidence and transmission is a model of the infection process, a mathematical description of the process through which incidence arises.
 A spectrum of models exists, categorised by the strength of assumptions they make about the transmission process.
@@
 These make almost no assumptions about the transmission process, viewing incidence as a generic counting process (introduced in full in \cref{E-inc-prev:sec:infection-process}).
-The most common model is that incidence follows a Poisson process, and the intensity of the process is the target of incidence.
+The most common model is that incidence follows a Poisson process, and the intensity of the process is the target of inference.
 % The use of Poisson processes to model incidence of an infectious disease was very popular in the context of HIV/AIDS\todo{maybe a ref for Poisson in HIV/AIDS?}.
@@
 These have components of the transmission process in them but do not fully represent the process.
-The most common type of semi-mechanistic model are those based on \emph{renewal equation}.
+The most common type of semi-mechanistic model are those based on a \emph{renewal equation}.
 The renewal equation~\autocite{fraserRenewal} for an infectious disease, in the most general form, is:
 \begin{align}
-    I(t) &= \int_{0}^{t} \beta(t,\tau) I(t-\tau) d \tau
+    Z(t) &= \int_{0}^{t} \beta(t,\tau) Z(t-\tau) d \tau
 \end{align}
-where $I(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the instantaneous probability of an individual infected $\tau$ time units ago infects an individual at time $t$.
+where $Z(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the instantaneous probability of an individual infected $\tau$ time units ago infects an individual at time $t$.
 Therefore, the renewal equation relates incidence in the past to incidence in the future.
-The effective reproduction number can be constructed as a function of $\beta(t, \tau)$.
+The effective reproduction number can be constructed as a function of $\beta(t, \tau)$ and the generation interval.
 % A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
 % The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
-Renewal equations do not feature futher in this thesis, for details, including practical considerations when using them as the basis for inference, see \textcite{nashEstimating,gosticPractical}.
+Renewal equations do not feature further in this thesis; for details, including practical considerations when using them as the basis for inference, see \textcite{nashEstimating,gosticPractical}.
 
-Finally, mechanistic models are the most complex and aim to fully represent the transmission process.
-\emph{Compartmental models} are the most common type of mechanistic model.
-Compartmental models split the population into compartments, each representing a different state of the disease (described in full in \cref{E-SEIR:sec:mechanistic-models}).
-The \emph{dynamics} of the epidemic, how it changes over time, are represented by the transitions between compartments.
+Finally, mechanistic models are the most complex, representing the transmission process in varying levels of detail.
+\emph{Compartmental} models, the most common type of mechanistic model, split the population into compartments, each representing a different state of the disease.
+The \emph{dynamics} of the epidemic, how it changes over time, are represented by the transitions between compartments (for more details see \cref{E-SEIR:sec:mechanistic-models}).
 Simple mechanistic models use a homogeneous population, with individuals differing only due to their disease state, and exponential distributions for the time in each disease state.
-More complex models relax these assumptions, in particular the age structure of the population is important for respiratory diseases.
+More complex models relax these assumptions, in particular incorporating heterogeneity; the age structure of the population is important for respiratory diseases.
 % Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
@@
 
-\section{Observation models}
+\section{Disease models}
 
-\todo[inline]{Needs some generic citations about observation models}
+The \emph{disease model} links the transmission model to a population-level quantity which we will make observations of.
+Examples are the incidence or prevalence of symptomatic individuals and admissions to hospital.
 
-The \emph{observation model} links the transmission model to the data.
 The choice of observation model can, in theory, be independent of the transmission model.
@@
 Typically, the core of an observation model is a \emph{convolution equation} or an approximation of it.
-This equation links an observation at time $t$, $y_t$, to the incidence prior to time $t$.
+This equation describes the expectation of the quantity at time $t$, $W_t$, conditional on the incidence prior to time $t$.
 In discrete time, it is:
 \begin{align}
-    y_t &= \zeta \sum_{i=0}^\infty I(t-i) f(i)
+    \E(W_t \mid Z_1, Z_2, \dots) &= \zeta \sum_{i=0}^\infty Z_{t-i} f(i)
+    \label{intro:eq:disease-model}
 \end{align}
-where $I(t)$ is the incidence at time $t$; $\zeta$ is the proportion of infections that are ever observed; and $f(i)$ is the probability that an infection at time $t-i$ is observed at time $t$, conditional on it ever being observed.
+where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that are ever observed; and $f(i)$ is the probability that an infection at time $t-i$ is observed at time $t$, conditional on it ever being observed.
+$f$ is known as a \emph{delay distribution}.
 
-% The choice of an observation model must be heavily informed by the data that will be used.
-% Two types of data commonly arise in the context of infectious disease: incidence measures and prevalence measures.
+Solving \cref{intro:eq:disease-model} for $Z$ is an ill-posed problem when $f(0)$ is small.
+The forwards error of the solution is large when the ratio of the off-diagonal to diagonal elements is large~\autocite[chapter 8.2]{highamAccuracy}; for this situation, that is when the value of $\max_{i > 0} f(i) / f(0)$ is large.
+A large forwards error means that the computed value is far from the true solution~\autocite[6--9]{highamAccuracy}.
+Either an approach which gives a large $f(0)$ or smoothing on $Z_t$ is required.
 
-% An incidence measure is characterised by each infection only appearing in the dataset at most once.
-% Common examples are hospital admissions and deaths.
-% The basic form of an observation model for an incidence measure is a scaling and delay of the incidence.
-% The scaling is the proportion of infections that are observed (\eg the proportion of infections that are fatal), and the delay is the time between an infection occurring and it being observed (\eg the time between infection and death).
-% In discrete time this is:
-% \begin{align}
-%     M(t) = \zeta \sum_{\tau=0}^\infty I(t-tau) f(\tau)
-% \end{align}
-% where $M(t)$ is the observation of the incidence measure at time $t$; $I(t)$ is the incidence at time $t$; $\zeta$ is the scaling; and $f(\tau)$ is the delay, formalised as the probability that an infection is observed $\tau$ days after it occurs, given that it will be observed.
+\section{Observation models}
 
-% A prevalence measure is characterised by each infection possibly appearing in the dataset multiple times.
-% Common examples are the results of a prevalence survey or the number of individuals in hospital with a disease.
-% \todo[inline]{Finish writing about prevalence measures}
+An observation model links $W_t$ to a possibly biased and/or noisy observation of it at time $t$.
+Biases can arise in a variety of ways, \eg healthcare-seeking behaviour or tests which are less than 100\% accurate.
+Bias can be modelled through additional parameters or data quantifying these issues.
+Noise could come from a variety of sources, \eg sampling error, clustering, or unexplained overdispersion.
+Noise is typically modelled by a parameteric distribution \eg the negative binomial distribution allowing for overdispersion compared to a Poisson process assumption.
 
-\todo[inline]{Expand slightly on convolution equations and why solving them is challenging, including some citations}
-Solving a convolution equation for $I$ is an ill-posed problem when $f(0)$ is small.
-Either an approach which gives a large $f(0)$, for example by assuming that the delay between incidence and being prevelant is 0, or smoothing on $I(t)$ is required.
+\section{Delay distributions}
 
-\section{Backcalculation}
+Knowledge of a delay distribution, $f$ in \cref{intro:eq:disease-model}, is required for many surveillance procedures.
+Typically, this is either the time until an event or the duration of a disease stage.
 
-\todo[inline]{
-  Decide whether to keep this section.
-  I think backcalculation probably does need to be called out explicitly, but it feels a little out-of-place.
-  IF kept, needs expanding a bit, flow a bit better, and citations added.
-}
-
-\emph{Backcalculation} was a modelling framework developed in the context of HIV/AIDS but since applied more broadly.
-The aim of backcalculation is to estimate the incidence of a disease from the observation of a prevalence measure.
-It combines a statistical transmission model with an observation model for prevalence data.
-
-Here, $\zeta = 1$ and $f(t)$ represents the probability of having the disease $t$ days after infection.
-In this way, $f$ represents a duration.
-
-\todo[inline]{Following paragraph seems a bit specific but is needed to motivate the following section}
-For a prevalence survey, the delay distribution is the \emph{duration of positivity}, the distribution of the time between an individual becoming infected and the time they are no longer infected.
-
-\section{Estimating the duration of positivity}
-
-Two approaches are commonly used to estimate the duration of a disease: either modelling an underlying biomarker or modelling the duration directly~\autocite[e.g.][]{sweetingEstimating}.
+Two approaches are commonly used to estimate the duration of a disease stage: either modelling an underlying biomarker or modelling the duration directly~\autocite[e.g.][]{sweetingEstimating}.
 Modelling a biomarker uses more information per observation, but requires stronger assumptions.
@@
 This approach generally requires fewer parameters, and no random effect(s), reducing the computational burden.
-It is also easier to specify non-parameteric forms of the distribution, which is important for SARS-CoV-2 as the tail behaviour may be important.
+It is also easier to specify non-parametric forms of the distribution, which is important for SARS-CoV-2 as the tail behaviour may be important.
 % A more detailed review of the literature is in the relevant chapters.
@@
 
-\section{Application to SARS-CoV-2 in England}
+\section{SARS-CoV-2 in England}
 
-\subsection{Data for estimating incidence and transmission} \label{intro:sec:data-for-estimating}
+The surveillance systems used in response to SARS-CoV-2 are historically unique.
+Modern technology allowed large amounts of data to be collected, distributed, and analysed.
+While the HIV/AIDS pandemic had inspired a large amount of statistical methodology to be developed, the pandemic was far slower moving.
+Much analysis of SARS-CoV-2 data was done at great haste, requiring trade-offs to be made between timeliness and rigour of results presented.
+
+Deep understanding of each data source's potential was not possible.
+In this thesis, I will investigate the potential of the prevalence surveys in greater detail.
+
+\subsection{Data sources} \label{intro:sec:data-for-estimating}
 
@@
 
+\subsubsection{Incidence-type data}
+
 For large parts of the pandemic, community testing was widely available in the UK.
-However, it does not test a random sample of the population.
+However, it did not test a random or even representative sample of the population.
 The number of tests conducted varied over time and space, for a variety of known and unknown reasons.
@@
 The large number of tests conducted allowed detailed analyses to be performed, including stratification by small geographic areas or age groups, although this does not solve the issue of the biases present.
-
-Seroprevalence surveys estimate the proportion of the population previously infected.
-Prior infection is more common than current infection, hence seroprevalence estimation requires a smaller sample size than current prevalence.
-However, there is a delay between infection and the development of antibodies, hence seroprevalence is not a real-time measure of the burden of the disease.
-Furthermore, seroprevalence data is normally based on convenience samples (\eg blood donors) who are not representativeness of the wider population is unclear.
 
@@
 
-Prevalence surveys in representative samples of the population are the most reliable measure of the burden of the disease.
-They are not subject to the biases of community testing and observe the whole population, unlike severe events, although they can suffer from non-response bias.
-However, they are expensive to power to inform detailed analyses.
-Often, some modelling is required to borrow information across strata or time to provide stable estimates, as well as to correct for non-response bias (see \cref{E-intro:sec:cis}).
+\subsubsection{Prevalence-type data}
+
+Seroprevalence surveys were conducted throughout the epidemic, primarly in blood donors.
+In addition to the normal problem of the delay between infection and antibody response, blood donors are a  convenience samples whose representativeness of the wider population is unclear.
+
+The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys in representative samples of the general population.
+These were not subject to the biases of community testing and observed the whole population, unlike severe events, although suffered from non-response bias.
+Therefore, some modelling is required to borrow information across strata or time to provide stable estimates, as well as to correct for non-response bias (see \cref{E-intro:sec:cis}).
+Despite this issue, they were considered the gold-standard metric.
 
 The use of prevalence surveys for estimating incidence and transmission is the focus of this thesis.
-These surveys use RT-PCT (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
+These surveys use RT-PCR (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
 I focus on the CIS (Coronavirus (COVID-19) Infection Survey), run by the ONS (Office for National Statistics)~\autocite{CIS,cisMethodsONS}.
-The CIS was a longitudinal study, testing a large number of indidividuals weekly or monthly for SARS-CoV-2; it is described in detail in \cref{E-intro:sec:cis}.
+The CIS was a longitudinal study, testing a large number of individuals weekly or monthly for SARS-CoV-2; it is described in detail in \cref{E-intro:sec:cis}.
 Early in the pandemic, a methodology was developed for using CIS data to estimate prevalence (see \cref{E-intro:sec:cis}).
@@
 
-\subsection{Cohort studies}
+\subsubsection{Cohort studies}
 
-In this thesis, I make use of two cohort studies with differing designs to estimate duration.
+Many cohort studies to estiamte a wide variety of quantities ran throughout the pandemic.
+In this thesis, I make use of two cohort studies with differing designs to estimate the duration or PCR positivity.
 These designs have complementary strengths, and I combine information from the studies.
@@
 The second study is the CIS (introduced in \cref{intro:sec:data-for-estimating}).
-In addition to estimating prevalence, the CIS followed the same individuals over a long period of time (over a year in many cases), collecting longitudinal data.
+In addition to its role as a prevalence survey, the CIS followed the same individuals over a long period of time (over a year in many cases), collecting longitudinal data.
 The large sample size and long follow-up mean that the tails of the distribution of the duration of positivity can be estimated.
@@
 
-\subsection{SPI-M-O}
+Early on in the epidemic, incidence could be estimated directly from the longitudinal survey data by considering the proportion of the cohort that tested negative then positive at consecutive tests~\autocite{onsIncidenceOld}.
+However, the method was abandoned in autumn 2020 when the testing was too infrequent to ensure that all infections were captured (\ie tested positive at least once).
 
-In the UK, SPI-M-O (Scientific Pandemic Influenza Group on Modelling, Operational sub-group) was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), generating evidence based on infectious disease modelling for policy development~\autocite{medleySPIM,govSPIMO}.
-For example, each of the stages of lifting the lockdown in 2021 was supported by  modelling studies informing the government's decision~\autocite{sageEvidence}.
+\subsection{Models}
+
+In the UK, the most modelling was produced by SPI-M-O (the Scientific Pandemic Influenza Group on Modelling, Operational sub-group).
+SPI-M-O was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), generating evidence based on infectious disease modelling for policy development~\autocite{medleySPIM,govSPIMO}.
+This used a wide range of  modelling approaches depending on the data available, which fluctuated throughout the epidemic, and the question being addressed.
+Some were fit to data while others were purely used for simulation based on plausible parameter values.
 
@@
 
-I will investigate whether estimates relying on only one data source, the CIS, could replace this process.
-Such estimates should be more robust, due to the representative nature of CIS.
+Some of these models were used to explore policy scenarios.
+For example, each of the stages of lifting the lockdown in 2021 was supported by modelling studies informing the government's decision~\autocite{sageEvidence}.
+These aided decision-makers in understanding plausible impacts of their actions.
 
-\subsection{Use of prevalence surveys}
+\section{Thesis context}
+
+In this thesis, I consider how prevalence surveys could be used for epidemic surveillance.
+These surveys have well-understood designs which means that estimates derived from them are more robust.
+A crucial quantity in their use will be the distribution of the duration of RT-PCR positivity.
+
+There is a large amount of related work in this area, from both prior to the pandemic and during its course.
+
+\subsection{Backcalculation}
+
+\emph{Backcalculation} is a modelling framework developed in the context of HIV/AIDS~\autocite{brookmeyerMethod}, but applied very broadly.
+It specifies a particular combination of infection, disease, and observation model to relate times of HIV infection with times of AIDS diagnosis via the incubation period.
+Further details are in the recent review \textcite{sunModeling}.
+The theory behind the framework underpins much of the work in this thesis.
+
+The original formulation used a Poisson process as the infection model; a disease model where all infections are eventually observed ($\zeta = 1$) and the delay distribution, $f$, is the incubation period; and an observation model where all diagnoses are observed, and hence has no influence on the data.
+
+To apply backcalculation to SARS-CoV-2 prevalence surveys, adaptations to the disease and observation models are needed (elaborated on in \cref{E-inc-prev}).
+The disease model must consider that this is now prevalence-type, rather than incidence-type data.
+This difference induces a dependency structure between the observations on each day.
+The observation model must consider that this is now a survey, with sampling error, rather than assuming all infections are observed.
+
+\subsection{Existing duration estimates} \label{intro:sec:previous-duration-estimates}
+
+An early meta-analysis estimated that individuals test positive for SARS-CoV-2 for a mean of 14.6 days (95\% CI: 9.3--20.0 days)~\autocite{cevikShedding}.
+The longest reported duration was observed at 83 days, although beyond day 9 no live virus was detected, indicating that the individuals testing positive for longer than this were not infectious.
+However, most studies included in the meta-analysis involved hospitalised patients or had unclear inclusion criteria, and hence may not be representative of the general population.
+These summary estimates are insufficient for backcalculation, which requires a distribution of the duration of positivity.
+
+\Textcite{hellewellPCRSensitivity} estimated the probability of an individual testing positive at each day after infection.
+This measure is highly related to the duration of positivity.
+The cohort in their study was drawn from healthcare workers, and hence was focused on working-age adults.
+
+\Textcite{binnySensitivity} applied a slight variant of \textcite{hellewellPCRSensitivity}'s methodology to data from New Zealand.
+The situation in New Zealand, who were successfully pursuing a zero-COVID strategy at this time, means that the individuals in the study were unusual.
+For example, a third were travelers detected during their quarantine period on arrival to New Zealand, but would have had negative RT-PCR tests before their departure.
+
+\Textcite{ealesCharacterising} estimated the duration of positivity from REACT data.
+Specifically, individuals who tested positive in round 8 of REACT (6th--22nd Jan 2021) were invited to take two additional swab tests.
+Defining the time of the initial positive test at time 0, \textcite{ealesCharacterising} estimated the probability of remaining positive at each day after the initial positive test.
+They made a strong parametric assumption that the probability decreases exponentially, with a sensitivity analysis assuming a plateau before the decay.
+This assumption is necessary because they had at most 17 days of follow-up, and hence could not estimate the tail of the distribution.
+Overall, they estimate a median duration of 14.0 days (95\% CI: 12.9--15.4 days) but that this varied substantially by the viral load of an individual at the time of the initial positive test.
+Individuals with a higher viral load were estimated to have a longer duration of positivity; the highest viral load individuals were estimated to have a median duration of 19.6 days (16.1--25.8 days).
+This could be due to the higher viral load leading to a longer duration of positivity, or that individuals with a higher viral load were more likely to have been recently infected at the time of the initial positive test.
+They cannot estimate how long before the initial positive test the individual was infected; therefore, they cannot distinguish between these two possibilities.
+This also means they underestimate the duration of positivity.
+
+For the purposes of this thesis, estimates applicable to the general population are required.
+There are no reliable estimates of the full distribution of the duration of positivity in a general population cohort available in the literature.
+Therefore, I estimate the duration of positivity from the ATACCC and CIS studies, as described above.
+
+\subsection{Existing estimation from prevalence surveys}
 
@@
 
-\subsection{Previous duration estimates}
+\subsection{Secure research service}
 
-An early meta-analysis estimated that individuals test positive for SARS-CoV-2 for a mean of 14.6 days (95\% CI: 9.3--20.0 days)~\autocite{cevikShedding}.
-The longest reported duration was observed at 83 days, although beyond day 9 no live virus was detected, indicating that the individuals testing positive for longer than this were not infectious.
-However, most studies included in the meta-analysis involved hospitalised patients or had unclear inclusion criteria, and hence may not be representative of the general population.
-These summary estimates are insufficient for backcalculation, which requires a distribution of the duration of positivity.
+The data from the CIS is stored in the SRS (Secure Research Service)~\autocite{onsSRS}.
+The SRS gives accredited researchers working on projects in the public interest access to granular data which cannot be legally shared publicly.
+Only non-disclosive data can be extracted from the SRS.
+In particular, counts normally need to be greater than 5.
 
-\Textcite{hellewellPCRSensitivity} estimated the probability of an individual testing positive at each day after infection.
-This measure is highly related to the duration of positivity.
-The cohort in their study was drawn from healthcare workers, and hence was focused on working-age adults.
-
-\Textcite{binnySensitivity} applied a slight variant of \textcite{hellewellPCRSensitivity}'s methodology to data from New Zealand.
-The situation in New Zealand, who were successfully pursuing a zero-COVID strategy at this time, means that the individuals in the study were unusual.
-For example, a third were travelers detected during their quarantine period on arrival to New Zealand, but would have had negative RT-PCR tests before their departure.
-
-\Textcite{ealesCharacterising} estimated the duration of positivity from REACT data.
-Specifically, individuals who tested positive in round 8 of REACT (6th--22nd Jan 2021) were invited to take two additional swab tests.
-Defining the time of the initial positive test at time 0, \textcite{ealesCharacterising} estimated the probability of remaining positive at each day after the initial positive test.
-They made a strong parametric assumption that the probability decreases exponentially, with a sensitivity analysis assuming a plateau before the decay.
-This assumption is necessary because they had at most 17 days of follow-up, and hence could not estimate the tail of the distribution.
-Overall, they estimate a median duration of 14.0 days (95\% CI: 12.9--15.4 days) but that this varied substantially by the viral load of an individual at the time of the initial positive test.
-Individuals with a higher viral load were estimated to have a longer duration of positivity; the highest viral load individuals were estimated to have a median duration of 19.6 days (16.1--25.8 days).
-This could be due to the higher viral load leading to a longer duration of positivity, or that individuals with a higher viral load were more likely to have been recently infected at the time of the initial positive test.
-They cannot estimate how long before the initial positive test the individual was infected; therefore, they cannot distinguish between these two possibilities.
-This also means they underestimate the duration of positivity.
-
-For the purposes of this thesis, estimates applicable to the general population are required.
-There are no reliable estimates of the full distribution of the duration of positivity in a general population cohort available in the literature.
-Therefore, I estimate the duration of positivity from the ATACCC and CIS studies, as described above.
+The SRS contains limited computational power.
+It has\todo{insert SRS computational power}.
+However, these are shared computational resources.
+Due consideration must be given to other users.
+Processes which run for multiple days are liable to be cancelled.
 
@@
 
+\subsection{Methodological contributions}
+
+As its primary aim, this thesis makes several methodological contributions.
+
+First, in \cref{E-inc-prev:sec:observation-process}, I provide a theoretic justification for the use of deterministic backcalculation in this setting.
+A \emph{deterministic} backcalculation approach was used by \textcite{abbottCISincidence} and I will use it in \cref{E-backcalc}.
+The deterministic approach models the prevalence at any time as a deterministic function of the incidence, rather than as a random variable.
+I show that, this is a good approximation as long as the proportion of the population sampled is small.
+In particular, it does not require the prevalence or incidence to be large.
+The estimates I produce in \cref{E-backcalc} fully propagate the uncertainty from the prevalence estimation procedure and use estimates of the duration of positivity from a general population study, unlike prior work.
+
+Next, in \cref{E-perf-test}, I estimate the duration of RT-PCR-positivity from double censored data with arbitrary periods where infections could be missed (a generalisation of truncation).
+This is based on the CIS data.
+While statistical frameworks have been previously developed which included this case, they had not been used in practice.
+In \cref{E-imperf-test}, I extend this framework to include false negatives.
+
+Within these analyses, I include information derived from the ATACCC study.
+In order to do so, I propose a novel prior which appropriately discounts the information from analysing ATACCC study, which has inappropriate model assumptions when applied to the larger sample size in the CIS data.
+
+A particular challenge with these analyses is that they must be performed within the SRS because they require individual-level data.
+
+Finally, in \cref{E-SEIR}, I describe a compartmental model for SARS-CoV-2 transmission, which I fit to CIS data.
+I allow a more flexible distribution of the duration of positivity than any of the papers discussed above, except \textcite{nicholsonImproving}, incorporating the skew of the distribution.
+This distribution is based on data from a representative sample of the UK population, rather than convenience samples (see \cref{intro:sec:duration-SARS-CoV-2}).
+Unlike \textcite{nicholsonImproving}, I include age-stratification in the transmission model, allowing estimation of how transmission varies by age.
+I use only data from the CIS, and show this is sufficient for estimating incidence and transmission.
+
+\subsection{Epidemiological contributions}
+
+The secondary aim of this thesis is providing new estimates to improve our understanding of SARS-CoV-2 in general, and the epidemic in England over the period being studied.
+
+In \cref{E-imperf-test}, I present estimates of the duration of RT-PCR-positivity in the general population.
+These estimates could be used to inform public health policy, for example, in interpreting RT-PCR test results or formulating quarantine policies.
+
+In \cref{E-transmission}, I present estimates of the incidence and transmission of SARS-CoV-2 in England over the period being studied.
+These estimates contribute to the literature reconstructing the SARS-CoV-2 epidemic in England.
+Understanding the dynamics of the past infection is important for informing the response to future pandemics, and the future burden of SARS-CoV-2.
+
 \subsection{Software and data availability}
@@
 The exception is the Markov chain Monte Carlo sampler used in \cref{E-SEIR}, which was written by Sanmitra Ghosh for the article \textcite{ghoshApproximate}.
-
-\subsection{Methodological contributions}
-
-As its primary aim, this thesis makes several methodological contributions.
-
-First, in \cref{E-inc-prev:sec:observation-process}, I provide a theoretic justification for the use of deterministic backcalculation in this setting.
-A \emph{deterministic} backcalculation approach was used by \textcite{abbottCISincidence} and I will use it in \cref{E-backcalc}.
-The deterministic approach models the prevalence at any time as a deterministic function of the incidence, rather than as a random variable.
-I show that, this is a good approximation as long as the proportion of the population sampled is small.
-In particular, it does not require the prevalence or incidence to be large.
-The estimates I produce in \cref{E-backcalc} fully propagate the uncertainty from the prevalence estimation procedure and use estimates of the duration of positivity from a general population study, unlike prior work.
-
-Next, in \cref{E-perf-test}, I estimate the duration of RT-PCR-positivity from double censored data with arbitrary periods where infections could be missed (a generalisation of truncation).
-This is based on the CIS data.
-While statistical frameworks have been previously developed which included this case, they had not been used in practice.
-In \cref{E-imperf-test}, I extend this framework to include false negatives.
-
-Within these analyses, I include information derived from the ATACCC study.
-In order to do so, I propose a novel prior which appropriately discounts the information from analysing ATACCC study, which has inappropriate model assumptions when applied to the larger sample size in the CIS data.
-
-A particular challenge with these analyses is that they must be performed within the SRS because they require individual-level data.
-The SRS is a form of TRE (Trusted Research Environment), with limited computing power.
-Analyses with long run times (\ie days) and/or requiring large amount of parallelisation (\eg a high-performance cluster) are not feasible.
-
-Finally, in \cref{E-SEIR}, I describe a compartmental model for SARS-CoV-2 transmission, which I fit to CIS data.
-I allow a more flexible distribution of the duration of positivity than any of the papers discussed above, except \textcite{nicholsonImproving}, incorporating the skew of the distribution.
-This distribution is based on data from a representative sample of the UK population, rather than convenience samples (see \cref{intro:sec:duration-SARS-CoV-2}).
-Unlike \textcite{nicholsonImproving}, I include age-stratification in the transmission model, allowing estimation of how transmission varies by age.
-I use only data from the CIS, and show this is sufficient for estimating incidence and transmission.
-
-\subsection{Epidemiological contributions}
-
-The secondary aim of this thesis is providing new estimates to improve our understanding of SARS-CoV-2 in general, and the epidemic in England over the period being studied.
-
-In \cref{E-imperf-test}, I present estimates of the duration of RT-PCR-positivity in the general population.
-These estimates could be used to inform public health policy, for example, in interpreting RT-PCR test results or formulating quarantine policies.
-
-In \cref{E-transmission}, I present estimates of the incidence and transmission of SARS-CoV-2 in England over the period being studied.
-These estimates contribute to the literature reconstructing the SARS-CoV-2 epidemic in England.
-Understanding the dynamics of the past infection is important for informing the response to future pandemics, and the future burden of SARS-CoV-2.
 

### 2024-03-18 introduction.tex
fixing missing cites and refs

@@
 For example, the number of hospital beds occupied with individuals with the disease or the proportion of the population with a marker of prior infection.
-The issues associated with incidence-type data also apply to prevalence-type data.
+The issues associated with incidence-type data often also apply to prevalence-type data.
 
@@
 They used a Gaussian process prior to smooth the incidence and stabilise the estimates.
-\Textcite{abbottCISincidence} used \textcite{hellewellPCRSensitivity}'s estimates of the sensitivity of testing to inform the observation model, which are based on a longitudinal study in healthcare workers (see \cref{intro:sec:duration-SARS-CoV-2}).
+\Textcite{abbottCISincidence} used \textcite{hellewellPCRSensitivity}'s estimates of the sensitivity of testing to inform the observation model, which are based on a longitudinal study in healthcare workers (see \cref{intro:sec:previous-duration-estimates}).
 They fit to summary statistics of the CIS data, using a daily, independent Gaussian likelihood with parameters estimated from the posterior point estimate and 95\% credible interval published by the ONS.
@@
 \Textcite{nicholsonImproving} fit a simple mechanistic model, notably without age structure, but have the most sophisticated observation model.
-They allow an arbitrary distribution of the duration of positivity, based on the estimates of \textcite{hellewellPCRSensitivity} (see \cref{intro:sec:duration-SARS-CoV-2} for discussion of the \textcite{hellewellPCRSensitivity} study).
+They allow an arbitrary distribution of the duration of positivity, based on the estimates of \textcite{hellewellPCRSensitivity} (see \cref{intro:sec:previous-duration-estimates} for discussion of the \textcite{hellewellPCRSensitivity} study).
 Furthermore, they incorporate the sensitivity and specificity of the RT-PCR test into their model.
@@
 The overall question that I investigate in this thesis is: can prevalence surveys alone be used to infer incidence and transmission, and if so how?
-Prevalence surveys provide data that is far more representative of the general population than other data sources (see \cref{intro:sec:data-sources}), therefore, conclusions based on them should be far more robust.
+Prevalence surveys provide data that is far more representative of the general population than other data sources, therefore, conclusions based on them should be far more robust.
 There are two components to this question.
@@
 I allow a more flexible distribution of the duration of positivity than any of the papers discussed above, except \textcite{nicholsonImproving}, incorporating the skew of the distribution.
-This distribution is based on data from a representative sample of the UK population, rather than convenience samples (see \cref{intro:sec:duration-SARS-CoV-2}).
+This distribution is based on data from a representative sample of the UK population, rather than convenience samples (see \cref{intro:sec:previous-duration-estimates}).
 Unlike \textcite{nicholsonImproving}, I include age-stratification in the transmission model, allowing estimation of how transmission varies by age.
@@
 Having estimated the duration of positivity, I then turn to estimating incidence and transmission from the CIS data in \cref{E-transmission}, thus fulfilling the primary aim of this thesis.
-I use both of the approaches discussed in \cref{intro:sec:estimating-incidence}, contrasting their results and discussing the advantages and disadvantages of each in this context.
+I use both a statistical and mechanistic transmission model, contrasting their results and discussing the advantages and disadvantages of each in this context.
 Finally, I conclude in \cref{E-conclusion}.

### 2024-03-18 conclusion.tex
Statistical => phenomonological

@@
 In \cref{E-transmission}, I found that this it is possible, developing two models that can infer incidence and transmission from prevalence surveys.
-One model was statistical in nature and the other mechanistic.
-The statistical approach, based on the backcalculation framework, provide estimates of incidence and makes fewer assumptions about the infection process; however, the estimates are very uncertain.
+One model was phenomenological in nature and the other mechanistic.
+The phenomenological approach, based on the backcalculation framework, provide estimates of incidence and makes fewer assumptions about the infection process; however, the estimates are very uncertain.
 The mechanistic approach, based on a SIR-type compartmental model, reduces the uncertainty and allows estimation of the components of transmission by making stronger assumptions about the infection process.

### 2024-03-18 introduction.tex
Statistical => phenomonological

@@
 
-The models with the weakest assumptions are statistical models.
+The models with the weakest assumptions are \emph{phenomenological} models.
 These make almost no assumptions about the transmission process, viewing incidence as a generic counting process (introduced in full in \cref{E-inc-prev:sec:infection-process}).
@@
 Having estimated the duration of positivity, I then turn to estimating incidence and transmission from the CIS data in \cref{E-transmission}, thus fulfilling the primary aim of this thesis.
-I use both a statistical and mechanistic transmission model, contrasting their results and discussing the advantages and disadvantages of each in this context.
+I use both a phenomenological and mechanistic transmission model, contrasting their results and discussing the advantages and disadvantages of each in this context.
 Finally, I conclude in \cref{E-conclusion}.

### 2024-03-19 MCMC-appendix.tex
spelling / grammar check

@@
 
-\section{Metropolis-Hastings}
+\section{Metropolis--Hastings}
 
-A common way of constructing a Markov chain with $p(\psi \mid y)$ as its stationary distribution is using the Metropolis-Hastings algorithm~\autocite{hastingsMCMC}.
+A common way of constructing a Markov chain with $p(\psi \mid y)$ as its stationary distribution is using the Metropolis--Hastings algorithm~\autocite{hastingsMCMC}.
 The algorithm requires a proposal distribution $q(\psi' \mid \psi)$, which proposes that the chain moves to a new state $\psi'$ from its current state $\psi$; $\pi(\psi)$, a calculation of $p(\psi \mid y)$ up to some constant.
@@
  }
- \caption{The Metropolis-Hastings algorithm.}
+ \caption{The Metropolis--Hastings algorithm.}
  \label{inc-prev:algorithm:MH}
@@
 
-The proposal distribution $q$ is most commonly a (possibly multivariate) Normal distribution centred on $\psi$; this scheme is used in \cref{E-SEIR}.
+The proposal distribution $q$ is most commonly a (possibly multivariate) normal distribution centred on $\psi$; this scheme is used in \cref{E-SEIR}.
 Tuning the proposal distribution to propose ``good'' moves is a key part of MCMC.
@@
 
-MCMC using a Normal proposal distribution can be inefficient.
+MCMC using a normal proposal distribution can be inefficient.
 In particular, it often poorly explores the tails of the posterior in finite time~\autocite[e.g.][]{turitsynIrreversible}.
@@
 This means that the sampler tends to propose moves with higher posterior density.
-The Metropolis-Hastings algorithm is then used to accept or reject the proposed move.
+The Metropolis--Hastings algorithm is then used to accept or reject the proposed move.
 

### 2024-03-19 cis-imperfect-testing.tex
spelling / grammar check

--- 2024-03-11 cis-imperfect-testing.tex
@@
 In \cref{E-perf-test}, a framework for analysing the CIS to produce estimates of the duration of positivity was developed.
-However, initial application of this framework produced an implausibly short estimate.
+However, the initial application of this framework produced an implausibly short estimate.
 In this chapter I identify this issue to be the result of false negatives, a form of misclassification bias, within the CIS (\cref{imperf-test:sec:problem}).
@@
 
-Comparing repeated data simulations, using the setup of \cref{E-perf-test:sec:simulation-study}, to the observed CIS data, suggests that this is due to the high frequency of single positive episodes in the CIS data (compare the histogram with $\psens=1$ and the vertical line in \cref{imperf-test:fig:sim-single-pos}).
+Comparing repeated data simulations, using the set-up of \cref{E-perf-test:sec:simulation-study}, to the observed CIS data, suggests that this is due to the high frequency of single positive episodes in the CIS data (compare the histogram with $\psens=1$ and the vertical line in \cref{imperf-test:fig:sim-single-pos}).
 A single positive episode is an episode containing exactly one positive test; intuitively, these cause estimates to be shorter because the lower bound on the length of time the episode is 1 day, the shortest possible.
@@
 \Textcite{shenNonparametrica}, generalising \textcite{panNote}, studied the situation when only the terminating event is interval censored (\ie singly censored data) and left truncated.
-They showed that the non-parameteric maximum likelihood estimator is inconsistent if single positive episodes occur, and the survival function is severely biased downwards.
+They showed that the non-parametric maximum likelihood estimator is inconsistent if single positive episodes occur, and the survival function is severely biased downwards.
 
@@
 where $t$ is the number of days since infection.
-Therefore, this model consists of a linear decline in test sensitivity, before a constant minimum point.
+Therefore, this model consists of a linear decline in test sensitivity before a constant minimum point.
 
@@
 I will modify $p_{ia}$ to allow the negative test following the last positive to be a false negative.
-If it is a false negative, then I will consider the episode's end right-censored.
-However, if it is a true negative, then the episode's end is interval-censored, as previously.
+If it is a false negative, then I will consider the episode's end right censored.
+However, if it is a true negative, then the episode's end is interval censored, as previously.
 A mixture of these scenarios forms the episode's likelihood contribution, with the mixture probability determined by the test sensitivity.
@@
 Therefore, this is unlikely to be a false negative because, if the individual was detectable at this time, they are in a period with high test sensitivity.
-This true negative, occurs with probability 1, and hence does not contribute to the likelihood.
+This true negative occurs with probability 1, and hence does not contribute to the likelihood.
 Denote the remaining tests considered as $\sched'_i = \{ t \in \sched_i \ssep r_{j(i)}^{(b)} \leq t \leq r_{j(i)}^{(e)} + 1 \}$, and their results by $\vec{y_{j(i)}} = \{ y_{j(i)}(t) \ssep t \in \sched'_i \}$ where $y_{j(i)}(t) = 1$ if the test on day $t$ is positive and 0 otherwise.
@@
 It is a false negative if and only if the episode ends after the test, \ie $e_{j(i)} > r_{j(i)}^{(e)}$.
-I proceed by considering whether this is the case, and considering the case with $b_{j(i)}$ known; $b_{j(i)}$ will later be integrated out.
+I proceed by considering whether this is the case and considering the case with $b_{j(i)}$ known; $b_{j(i)}$ will later be integrated out.
 
-First, if $e_{j(i)} \leq r_{j(i)}^{(e)}$, meaning that the test at $r_{j(i)}^{(e)}+1$ is a true negative and the end of the episode is interval-censored as in the previous chapter.
+First, if $e_{j(i)} \leq r_{j(i)}^{(e)}$, meaning that the test at $r_{j(i)}^{(e)}+1$ is a true negative and the end of the episode is interval censored as in the previous chapter.
 % In this case, the test at $r_{j(i)}^{(e)} + 1$ is a true negative, as are all other tests not in $\sched'_i$.
@@
 In this case, the test at $r_{j(i)}^{(e)}$ is a false negative, occurring with probability $(1 - p_\text{sens})$.
-To avoid having to consider tests after $r_{j(i)}^{(e)}$, which could greatly complicate the likelihood, I model this case as the episode being right-censored at $r_{j(i)}^{(e)}$.
+To avoid having to consider tests after $r_{j(i)}^{(e)}$, which could greatly complicate the likelihood, I model this case as the episode being right censored at $r_{j(i)}^{(e)}$.
 Taking the same approach as before:
@@
 For each simulated dataset, I estimate the survival function using the likelihood proposed in \cref{imperf-test:sec:modelling} and assuming $\psens$ equals 0.6, 0.8, or 1.0.
-If the dataset was simulated using a constant $\psens$, and that value of $\psens$ matches the value assumed during inference, I refer to $\psens$ as being correctly specified; otherwise I refer to it as misspecified.
+If the dataset was simulated using a constant $\psens$, and that value of $\psens$ matches the value assumed during inference, I refer to $\psens$ as being correctly specified; otherwise, I refer to it as misspecified.
 However, there is still some misspecification of the model to false negatives owing to the simplifying assumptions made.
@@
     Posterior (median and 95\% credible interval) of the survival time for the simulation study with a correctly specified test sensitivity.
-    True survival time shown in black.
+    The true survival time is shown in black
   }
@@
 
-Next, I considered the consequence of $\psens$ being misspecified, using the vague prior; the better performing prior in the correctly specified case.
+Next, I considered the consequence of $\psens$ being misspecified, I use the model combination prior because it is the better performing prior in the correctly specified case.
 If the test sensitivity is misspecified, that is the assumed value for $p_\text{sens}$ in \cref{imperf-test:eq:pit-prime,imperf-test:eq:pit-prime} is different to the value used in the simulation, then the survival time will be biased (see \cref{imperf-test:fig:misspecified-test-sensitivity}).
@@
     Posterior (median and 95\% credible interval) of the survival time for the simulation study with a possibly misspecified test sensitivity.
-    True survival time shown in black.
+    The true survival time is shown in black
     All simulations use a constant test sensitivity of 0.8, but the inference procedure assumes different values, as per the key.
@@
   \centering \includegraphics{cis-imperfect-testing/CIS_final}
-  \caption{Duration estimates using CIS and ATACCC data}
+  \caption{Duration estimates using CIS and ATACCC data.}
   \label{imperf-test:fig:cis-estimates}
@@
 
-The the estimates are most sensitive to the choice of $r$, the strength of the prior on $\ntot$.
+The estimates are most sensitive to the choice of $r$, the strength of the prior on $\ntot$.
 A very weak, almost uninformative, prior on this quantity causes the posterior estimate to be much higher than the estimate from \textcite{birrellRTM2}.
@@
 Estimating the test sensitivity excluding intermittent negatives is not possible because \cref{imperf-test:eq:pia-prime-constant,imperf-test:eq:pit-prime} are both monotonically decreasing in $\psens$; therefore, the likelihood always favours $\psens = 0$ (\ie no true positives).
-This aligns with the situation with singly interval-censored, untruncated data, when stopping at the first observed time of the terminating event (which may be misclassified) means that the test sensitivity cannot be estimated~\autocite[e.g.]{titmanMisclassify}.
+This aligns with the situation with singly interval censored, untruncated data, when stopping at the first observed time of the terminating event (which may be misclassified) means that the test sensitivity cannot be estimated~\autocite[e.g.]{titmanMisclassify}.
 
@@
 Whether these methods are sufficient for the model to be identifiable with a time-varying $\psens$ needs further investigation.
-A simple parameteric form of the test sensitivity, such as that proposed by \textcite{brownBayesian}, may be sufficient to allow identifiability.
+A simple parametric form of the test sensitivity, such as that proposed by \textcite{brownBayesian}, may be sufficient to allow identifiability.
 In any case, the likelihood, especially $p_{iu}'$, would be substantially complicated by such an addition which may greatly increase the computational cost of inference.

### 2024-03-19 incidence-prevalence.tex
spelling / grammar check

@@
 I start by defining these terms in \cref{inc-prev:sec:definitions}.
-I then move on to explaining the three processes that are involved: the \emph{infection process}, the \emph{prevalence process} and the \emph{observation process}.
+I then move on to explaining the three processes that are involved: the \emph{infection process}, the \emph{prevalence process}, and the \emph{observation process}.
 The mathematical description of are an infection model, disease model, and observation model respectively.
@@
 Here, I adapt previous work to the SARS-CoV-2 context.
-In particular, recovering from infection needs to be considering, which was not the case for HIV/AIDS.
+In particular, recovering from infection needs to be considered, which was not the case for HIV/AIDS.
 
@@
 
-Time will be considered discrete in this chapter, except in \cref{SEIR:sec:mechanistic-models}.
-I use days, however, the results generalise in the obvious way to other time units.
+I use discrete time in this chapter, except in \cref{SEIR:sec:mechanistic-models}.
+I use days; however, the results generalise in the obvious way to other time units.
 I choose discrete time for pragmatic reasons.
@@
 \emph{Incidence} on day $t$, $Z_t$, is the number of people in a given population who become infected on day $t$; when expressed as a proportion of the population it is the \emph{incidence proportion}~\autocite[89]{lashModern}, $Z_t/\Npop$, where $\Npop$ is the number of people in the population.
-The timing of infection events is very rarely directly observable and must be inferred from other data.
+The timing of infection events is very rarely directly observable and instead is inferred from other data.
 Despite this issue, it remains the single most important quantity for informing the response to an epidemic or pandemic.
@@
 I refer to individuals with the disease as \emph{prevalent} individuals.
-Whether an individual is prevalent is often expensive or error-prone to measure, therefore, it will often be regarded as a latent quantity.
+Whether an individual is prevalent is often expensive or error-prone to measure and hence is a latent quantity.
 Furthermore, what is meant by ``having'' a disease can depend on context; for example, it could be the presence of symptoms, the ability to transmit, or detectable status (the difference between these was discussed in \cref{E-biology-data:sec:natural-history})
@@
 
-\emph{Duration}, or \emph{time-to-event}, distributions can be expressed in several equivalent ways~\autocite[17]{sunStatisticala}.
+\emph{Duration}, or \emph{time-to-event}, distributions are expressed in several equivalent ways~\autocite[17]{sunStatisticala}.
 The duration distribution I will be concerned with is the distribution of the number of days for which an individual is prevalent.
@@
 The \emph{hazard} on day $d$ is the probability of recovering on day $d$ conditional on still being positive on day $d$: $\lambda(d) = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
-Throughout this thesis, the survival function will be the primary quantity of interest, and it will normally be expressed in terms of the hazard; these two quantities are related by $S(d) = \prod_{i=1}^{d-1} (1 - \lambda(i))$.
+Throughout this thesis, the survival function will be the primary quantity of interest, and normally parameterized in terms of the hazard; these two quantities are related by $S(d) = \prod_{i=1}^{d-1} (1 - \lambda(i))$.
 
@@
 Similarly, $\lambda(d) \in [0,1]$, but can otherwise vary arbitrarily (\ie there is no monotonic or sum constraint).
-The lack of monotonicity or sum constraint on the hazard makes it an attractive parameterisation for inference.
+The lack of monotonicity or sum constraint on the hazard makes it an attractive parameterization for inference.
 
@@
 $K(t)$ must take non-negative integer values and be monotonically increasing in $t$.
-The interpretation of $K(t)$ is the number of events, here infections, that have occurred in the interval $(-\infty, t]$~\autocite[244]{yanDistribution}.
+The interpretation of $K(t)$ is the number of events, here infections, which have occurred in the interval $(-\infty, t]$~\autocite[244]{yanDistribution}.
 In the context of a disease, the point process can be discretised into the incidence on day $t$ as $Z_t = K(t) - K(t-1)$, where $t$ is an integer.
@@
 First, in \cref{inc-prev:sec:poisson-process}, as a \emph{Poisson process}.
-A Poisson process is a widely-used type of counting process both within and outside of epidemiology.
-It makes no direct assumptions on the underlying biological or epidemiological process, and is therefore a flexible model.
+A Poisson process is a widely used type of counting process both within and outside of epidemiology.
+It makes no direct assumptions on the underlying biological or epidemiological process and is therefore a flexible model.
 
@@
 
-A commonly-used point process for infectious disease epidemiology is the time-inhomogenous Poisson process~\autocite[e.g., in the context of HIV,][]{brookmeyerMethod,paganoHIV,rosenbergBackcalculation,brookmeyerBackcalculation}.
+A commonly used point process for infectious disease epidemiology is the time inhomogeneous Poisson process~\autocite[e.g., in the context of HIV,][]{brookmeyerMethod,paganoHIV,rosenbergBackcalculation,brookmeyerBackcalculation}.
 % A time-homogenous Poisson process is the special case where $\nu(r)$ is constant.
@@
 % Overdispersion can be incorporated by replacing the assumption that $K(s) - K(t) \dist \Poi(\int_{t}^{s} \nu(r) dr)$ with $K(s) - K(t) \dist \NBc(\int_{t}^{s} \nu(r) dr, k)$ where $k$ is the overdispersion parameter, normally assumed to be constant absent interventions.
-% Note that the expected number of events has not changed, but the variance has increased (see \cref{E-distributions} for details of parameterisations and moments).
+% Note that the expected number of events has not changed, but the variance has increased (see \cref{E-distributions} for details of parameterizations and moments).
 % %The negative binomial tends to Poisson as $k\to\infty$.
@@
 Both overdispersion and feedback loops violate the properties of a Poisson process.
-Despite these issues, consistent inference on the incidence can still be produced using the Poisson process model, although the uncertainty in the estimates may be understated~\autocite{beckerDependent}.
+Despite these issues, consistent inference on the incidence is still produced using the Poisson process model, although the uncertainty in the estimates may be understated~\autocite{beckerDependent}.
 
@@
 
-Now I introduce mechanistic models, in particular the class of mechanistic transmission models known as \emph{compartmental} models.
-Compartmental models are the most widely-used class of mechanistic infectious disease model.
+Now I will introduce mechanistic models, in particular the class of mechanistic transmission models known as \emph{compartmental} models.
+Compartmental models are the most widely used class of mechanistic infectious disease models.
 A compartmental model is named because they place every member of the population under study into a compartment; each compartment representing a disease state (\eg susceptible).
@@
 Stochastic models are most commonly specified in discrete time, using a Markov assumption.
-A common simplification is to use discrete time and a Markov assumption; \ie $\vec{x}(t)$ is only defined for integer $t$ and $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta}) = p(\vec{x}(t) \mid \vec{x}(t-1), \vec{\theta})$~\autocite{birrellEvidence}.
+A common simplification is to use discrete time and a Markov assumption, \ie $\vec{x}(t)$ is only defined for integer $t$ and $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta}) = p(\vec{x}(t) \mid \vec{x}(t-1), \vec{\theta})$~\autocite{birrellEvidence}.
 This relaxes the assumption of independent increments in the counting process formulation of the infection process (see \cref{E-inc-prev:sec:infection-process}).
@@
 This choice is appropriate for the context of this thesis because there were many individuals infectious with SARS-CoV-2 in this time period.
-Hence, further details on stochastic models are not outside the scope this thesis; a wealth of introductory texts with further details on stochastic models are available, I recommend \textcite[chapter 6]{keelingModeling}.
+Hence, further details on stochastic models are outside the scope of this thesis; a wealth of introductory texts with further details on stochastic models are available, I recommend \textcite[chapter 6]{keelingModeling}.
 
@@
 The proportion of the population that is infectious at time $t$ is $i(t)$.
-Individuals in the recovered compartment are immune to the disease, and cannot be infected again.
+Individuals in the recovered compartment are immune to the disease and cannot be infected again.
 The proportion of the population that have recovered by time $t$ is $r(t)$.
@@
 Assume this individual comes into contact with other individuals in the population at a mean rate of $\kappa$.
-Further, assume the population is \emph{well-mixed}, meaning that each contact is with an individual chosen uniformly at random from the population; this is implied by assuming a homogeneous population.
+Further, assume the population is \emph{well-mixed}, meaning that each contact is with an individual chosen uniformly at random from the population, implied by assuming a homogeneous population.
 For such a randomly selected individual, the probability that they are susceptible is $s(t)$.
 % Each of the infectious individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
-Therefore the infectious individual makes contact with susceptible individuals at a mean rate of $\kappa s(t)$ per day.
+Therefore, the infectious individual makes contact with susceptible individuals at a mean rate of $\kappa s(t)$ per day.
 Assume that there is a constant probability of transmission upon contact between an infectious and susceptible individual, $q$.
 Hence, the mean rate of infections caused by this infectious individual is $\kappa s(t) q$ per day.
-The two constant terms are absorbed into the transmission rate, $\beta = \kappa q$.
-Therefore, the rate of new infections generated by a single infectious individual is $\beta s(t)$ per day.
-Multiplying by the number of infectious individuals and dividing by the population size gives the rate per day at which susceptible individuals are infected as a proportion of the population, $\beta s(t) i(t) = \lambda(t) s(t)$.
+The transmission rate $\beta = \kappa q$ absorbs the two constant terms.
+Therefore, $\beta s(t)$ is the rate of new infections generated by a single infectious individual per day.
+Multiplying by the number of infectious individuals and dividing by the population size gives the daily rate of infection of susceptible individuals as a proportion of the population, $\beta s(t) i(t) = \lambda(t) s(t)$.
 
@@
     \item The basic reproduction number for the SIR model is $\R = \beta / \gamma$~\autocite[20]{keelingModeling}.
-    Intuitively, this can be thought of as the mean number of infections caused by an infectious individual per day, $\beta$, multiplied by the mean number of days for which they are infectious, $1/\gamma$.
+    This is the mean number of infections caused by an infectious individual per day, $\beta$, multiplied by the mean number of days for which they are infectious, $1/\gamma$.
     \item The effective reproduction number for this model is $\Re = \beta / \gamma s(t) = \R s(t)$~\autocite{pellisEstimation}.
-    $1-s(t)$ is the fraction of infections that are avoided because of the presence of immune individuals in the population.
-    \item The population is assumed to be closed.
+    $1-s(t)$ is the fraction of infections that are avoided due to the presence of immune individuals in the population.
+    \item The model assumes a closed population.
     This assumption implies several attributes.
     First, the total population size is constant.
-    Second, only a negligible number of infections are imported from outside the population.
-    Third, births and deaths of individuals are negligble over the length of time considered; model extensions allow these factors in contexts where they are relevant~\autocites[26]{keelingModeling}[214]{kretzschmarMathematical}.
-    \item There is no analytical solution to the system of ODEs: numerical methods are required to solve the system~\autocite[25]{keelingModeling}.
-    \item As noted previously, the parameters $\vec{\theta}$ and the intial state of the system, $\vec{x}(0)$ are required to solve the system.
-    \item Immunity is assumed to last until the end of the  period considered ~\autocite[61]{andersonInfectious}.
+    Second, imported infections are negligible. 
+    Third, births and deaths of individuals are negligible over the length of time considered; model extensions allow these factors in contexts where they are relevant~\autocites[26]{keelingModeling}[214]{kretzschmarMathematical}.
+    \item There is no analytical solution to the system of ODEs.
+      Solving the system requires numerical methods~\autocite[25]{keelingModeling}.
+    \item As noted previously, solving the system of ODEs requires the parameters $\vec{\theta}$ and the initial state of the system, $\vec{x}(0)$.
+    \item The model assumes immunity lasts until the end of the period considered~\autocite[61]{andersonInfectious}.
     In many contexts, including here, this assumption is justified because the period is shorter than the length of immunity conferred~\autocite{milneImmunity}.
-    For longer time periods, waning immunity can be incorporated by allowing individuals to move from the recovered compartment back to the susceptible compartment~\autocite[40]{keelingModeling}.
-    \item The incidence proportion on day $t$ can be defined as $Z_t / \Npop = s(t) - s(t+1)$.
+    For longer time periods, the model can be expanded to include waning immunity~\autocite[40]{keelingModeling}.
+    \item The incidence proportion on day $t$ is defined as $Z_t / \Npop = s(t) - s(t+1)$.
 However, this will not in general give an integer value to $Z_t$.
@@
 \todo{check this citation is formatted correctly}
-This is done by adding multiple infectious compartments, shown in \cref{SEIR:fig:SIR-gamma}.
+The extension adds adding multiple infectious compartments, shown in \cref{SEIR:fig:SIR-gamma}.
 The ODEs are now:
@@
 In this model the infectious period is distributed $\GamDist(n, n\gamma)$, a realistic model for a variety of diseases~\autocite{wearingAppropriate}.
-This parameterisation is chosen to give $1/\gamma$ as the mean infectious period (further details on the Gamma distribution are in \cref{E-distributions}).
+This parameterization is chosen to give $1/\gamma$ as the mean infectious period (further details on the Gamma distribution are in \cref{E-distributions}).
 A gamma distribution with the first parameter equal to 1 is an exponential distribution, in which case this model collapses to the basic SIR model in \cref{SEIR:sec:SIR}.
@@
 If the strata have differing transmission patterns, this can change the dynamics of the epidemic.
-The different patterns could be due to behavioural factors (\eg high-numbers of sexual partners in a sexually-transmitted disease~\autocite[69]{keelingModeling} or younger age groups making many more contacts for respiratory diseases~\autocite[176]{andersonInfectious}), or biological factors such as genetics~\autocite[208]{andersonInfectious}.
+The different patterns could be due to behavioural factors (\eg high numbers of sexual partners in a sexually transmitted disease~\autocite[69]{keelingModeling} or younger age groups making many more contacts for respiratory diseases~\autocite[176]{andersonInfectious}), or biological factors such as genetics~\autocite[208]{andersonInfectious}.
 I denote the set of strata, as $\set{A}$, with the set's size as $A$.
@@
 
-The state vector is now a vector of vectors, one for each of the compartments
+The state vector is now a vector of vectors, one for each of the compartments.
 For example, in the basic SIR model, $\vec{x}(t) = (\vec{s}(t), \vec{i}(t), \vec{r}(t))^T$.
@@
 Other, equivalent formulations of this model exist within the literature.
-For example defining $s_a$ as the proportion of the total population that are both in strata $a$ and susceptible or parameterised using numbers of individuals~\autocite[57]{keelingModeling}, rather than as the proportion of strata $a$ as I have done here.
+For example, defining $s_a$ as the proportion of the total population that are both in strata $a$ and susceptible or parameterized using numbers of individuals~\autocite[57]{keelingModeling}, rather than as the proportion of strata $a$ as I have done here.
 
@@
 It is infeasible because the progression of the epidemic depends on $\vec{\lambda}(t)$, which has dimension $A$, while $\matr{\beta}$ has dimension $A^2$.
-One common assumptions is symmetry, \ie $\beta_{aa'} = \beta_{a'a}$.
+A common assumption is symmetry, \ie $\beta_{aa'} = \beta_{a'a}$.
 % Alternatively, as in the scalar case, it may be easier to iden $\beta_{aa'}$ can be decomposed into the contact rate between strata $a$ and $a'$, $\kappa_{aa'}$ and the probability of transmission upon contact, $q_{aa'}$, and then assumptions made about the components.
@@
 However, how to average these contributions to calculate the basic reproduction number is not obvious.
-The standard solution to this was proposed by \textcite{diekmannDefinition} (see \textcite[chapter 7]{diekmannMathematical} for a more gentle, but still rigorous, explanation).
+The standard solution to this was proposed by \textcite{diekmannDefinition} (see \textcite[chapter 7]{diekmannMathematical} for a gentler, but still rigorous, explanation).
 First, they show that the distribution of infections across the strata during the initial period of exponential growth (when $\vec{s} \approx \vec{1}$) relies only on $\matr{\beta}$, and not the starting conditions.
@@
 For any infection beginning at $t$, the probability that it ends at $t'$ is $\prob(E_i = t' \mid B_i = t) = \prob(D_i = t' - t + 1 \mid B_i = t) = f_D(t' - t + 1)$.
-Since these durations are assumed independent, the number of episodes of each duration is multinomially distributed~\autocite{paganoHIV}.
+Since I assume these durations are independent, the number of episodes of each duration is multinomially distributed~\autocite{paganoHIV}.
 Specifically:
@@
 \end{align}
-Finally, the $P_t$ can be expressed as the total number of people prevalent at time $t$, regardless of when they were infected.
-The limits of the sum are limited by the zeroes in \cref{inc-prev:eq:Ptt-to-R}.
+Finally, express $P_t$ as the total number of people prevalent at time $t$, regardless of when they were infected.
+The zeroes in \cref{inc-prev:eq:Ptt-to-R} determine the sum's limits.
 \begin{align}
@@
 
-The second moment is simplified by the independence.
+The independence structure simplifies the second moment.
 \begin{align}
@@
 
-The expectation in \cref{inc-prev:eq:EPt} is a discrete convolution equation.
-The equation is of the same form as the equation used for HIV/AIDS backcalculation, as well as being found in many other fields~\autocite[and references therein]{brookmeyerBackcalculation}.
+The expectation in \cref{inc-prev:eq:EPt} is a discrete convolution equation in the same form as \cref{intro:eq:disease-model} with $\zeta = 1$.
+% The equation is of the same form as the equation used for HIV/AIDS backcalculation, as well as being found in many other fields~\autocite[and references therein]{brookmeyerBackcalculation}.
 
@@
 The conditioning on $n_t$ and $\Npop$ is implicit in what follows.
-These quantities are assumed to be fixed by the study design and population choice, and independent of $P_t$ and $Z_t$.
+These quantities are fixed by the study design and population choice, and hence I assume independent of independent of $P_t$ and $Z_t$.
 
@@
 %The population prevalence is a latent quantity.
-Prevalence surveys are used in this thesis to infer incidence.
+I will use prevalence surveys to infer incidence.
 I base this inference on the likelihood, $p(\vec{x} \mid \vec{Z})$.% = \int p(\vec{x} \mid \vec{P}) p(\vec{P} \mid \vec{Z}) d\vec{P} = \int \prod_t p(x_t \mid P_t) p(\vec{P} \mid \vec{Z}) d\vec{P}$, where $\vec{x} = (x_1, x_2, \dots)^T$.
@@
 In the following, all are implicitly conditioned on $\vec{Z}$.
-However, for clarity this is omitted from the notation.
+However, for clarity I omit the conditioning from the notation.
 \begin{align}

### 2024-03-19 introduction.tex
proof intro
spelling / grammar check

@@
 All pandemics start as an \emph{outbreak}, a sudden and unexpected rise in the number of people with a disease in a local area; if this spreads over a larger geographic area (\eg a country) it becomes an \emph{epidemic}; if it spreads globally then it is a \emph{pandemic}~\autocite{grennanPandemic}.
-The precise meaning of these terms is contested, determined implicitly by their usage rather than formal definitions~\autocite{morensPandemic,doshiElusive}.
+The precise meaning of these terms varies, they are defined implicitly by their usage rather than formal definitions~\autocite{morensPandemic,doshiElusive}.
 I will refer to the SARS-CoV-2 epidemic within England (or the UK), and the global impact as the pandemic.
@@
 Surveillance is important at every stage: early on, for detecting a new outbreak quickly and assessing its likely impact and later for monitoring the ongoing burden of the disease and the effectiveness of interventions~\autocite{whoFluSurveillance}.
-In all cases, surveillance is required to inform public health policy.
+In all cases, surveillance informs public health policy.
 In this thesis, I consider methods to improve surveillance, using the SARS-CoV-2 epidemic in England as a case study.
@@
 Broadly, a reproduction number describes the number of secondary infections generated by a single infection.
-Here, following \textcite{pellisEstimation},  building on over a century of infectious disease dynamics literature, I define three that are broadly useful.
+Here, following \textcite{pellisEstimation}, and building on over a century of infectious disease dynamics literature, I define three that are broadly useful.
 
@@
 $\Rc$ is also known as the \emph{reproduction number excluding immunity}.
-It describes the effect on the epidemic of interventions (except vaccination) that have been put into place.
+It describes the effect on the epidemic of interventions (except vaccination) that are in place at time $t$.
 $\Rc$ is useful when comparing the effects of interventions at different points in time, when the immunity status of the population has changed.
@@
 For a given disease and population, an analogous growth rate can be defined for each definition of the reproduction number~\autocite{pellisEstimation,paragGrowthRates,wallingaGI}.
-The advantage of using the growth rate is that they are easier to estimate and capture how quickly spread happens in calendar time; however, they do not directly give the strength of intervention required to control the epidemic~\autocite{royalSocietyRnumber}.
+The advantage of using the growth rate it is easier to estimate and captures how quickly spread happens in calendar time; however, they do not directly give the strength of intervention required to control the epidemic~\autocite{royalSocietyRnumber}.
 
@@
 Incidence informs the short-term future burden of the disease, for example, the future number of severe events (hospitalisations or deaths).
-To translate incidence into the burden of the disease, a measure of severity is required; for example, the IFR (infection fatality rate), the proportion of infected individuals who die.
+Incidence is translated into disease burden using a measure of severity is required; for example, the IFR (infection fatality rate), the proportion of infected individuals who die.
 Estimation of severity is beyond the scope of this thesis.
@@
 Estimating transmission refers to inferring the parameters of this process, including its components.
-The components of transmission (\eg number of contacts per day and age-stratified rates) inform how interventions can be targeted.
+The components of transmission (\eg the number of contacts each individual makes per day) allow interventions to be targeted.
 They are also important for forecasting future incidence.
@@
 Here I discuss three types of data which are useful for surveillance.
-Incidence-type and prevalence-type data can directly inform estimation of incidence and transmission, however, this estimation almost always relies on prior knowledge of epidemiological and/or biological quantities.
+Incidence-type and prevalence-type data can directly inform estimation of incidence and transmission; however, this estimation almost always relies on prior knowledge of epidemiological and/or biological quantities.
 These quantities can be estimated from other studies, \eg cohort studies.
@@
 
-Incidence-type data originate from observations of the consequences of a disease, \eg the development of symptoms or admission to hospital.
+Incidence-type data originates from observations of the consequences of a disease, \eg the development of symptoms or admission to hospital.
 This type of data is typically affected by \emph{ascertainment bias}: the proportion of infections which are counted can vary over time or within different subgroups of the population~\autocite[chapter 9]{lashModern}.
@@
 
-Prevalence-type data originate from observations of the number or proportion of a population in a specific disease state at a specific time.
-For example, the number of hospital beds occupied with individuals with the disease or the proportion of the population with a marker of prior infection.
+Prevalence-type data originates from observations of the number or proportion of a population in a specific disease state at a specific time.
+For example, the number of hospital beds occupied by individuals with the disease or the proportion of the population with a marker of prior infection.
 The issues associated with incidence-type data often also apply to prevalence-type data.
@@
 
-Other epidemiological and/or biological quantities are normally required to estimate transmssion.
-These depend on the model type, but could include the \emph{generation interval} (the time between an individual being infected and going on to infect further individuals) or duration of different disease stages.
-Cohort studies, where a group of individuals are followed over time, are a study design which can estimate many of these quantities.
+Other epidemiological and/or biological quantities are normally required to estimate transmission.
+These depend on the model type but could include the \emph{generation interval} (the time between an individual being infected and going on to infect further individuals) or duration of different disease stages.
+Cohort studies (following a group of individuals over time), are a study design which can estimate many of these quantities.
 
@@
 However, recruiting a representative sample of the population is challenging, and the results of these studies are often not generalisable to the whole population.
-Furthermore, for these studies to be appropriately powered, they often require a large cohort, and hence are expensive to run.
+Furthermore, these studies often require a large cohort to be appropriately power.
+Large cohort studies are expensive to run.
 
@@
 A spectrum of models exists, categorised by the strength of assumptions they make about the transmission process.
-Stronger assumptions allow more insight into the transmission process and generally lower the variance of the estimates, but can introduce bias and are often computationally challenging to fit.
+Stronger assumptions allow more insight into the transmission process and generally lower the variance of the estimates but can introduce bias and are often computationally challenging to fit.
 
@@
 These have components of the transmission process in them but do not fully represent the process.
-The most common type of semi-mechanistic model are those based on a \emph{renewal equation}.
+The most common type of semi-mechanistic model is those based on a \emph{renewal equation}.
 The renewal equation~\autocite{fraserRenewal} for an infectious disease, in the most general form, is:
@@
 Therefore, the renewal equation relates incidence in the past to incidence in the future.
-The effective reproduction number can be constructed as a function of $\beta(t, \tau)$ and the generation interval.
+The effective reproduction number can be expressed as a function of $\beta(t, \tau)$.
 % A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
@@
 
-\section{Disease models}
+\section{Disease models} \label{intro:sec:disease-model}
 
-The \emph{disease model} links the transmission model to a population-level quantity which we will make observations of.
+The \emph{disease model} links the transmission model to a quantity, $W_t$, observable albeit possibly in a noisy and/or biased way (see next section).
 Examples are the incidence or prevalence of symptomatic individuals and admissions to hospital.
@@
 The choice of observation model can, in theory, be independent of the transmission model.
-However, pragmatic considerations often mean that the observation model is chosen to be in a similar form to the transmission model.
+However, pragmatic considerations often mean choosing the observation model to be in a similar form to the transmission model.
 
 Typically, the core of an observation model is a \emph{convolution equation} or an approximation of it.
-This equation describes the expectation of the quantity at time $t$, $W_t$, conditional on the incidence prior to time $t$.
+This equation describes the expectation of $W_t$, conditional on the incidence prior to time $t$.
 In discrete time, it is:
@@
 \end{align}
-where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that are ever observed; and $f(i)$ is the probability that an infection at time $t-i$ is observed at time $t$, conditional on it ever being observed.
+where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that ever appear in $W_t$; and $f(i)$ is the probability that an infection at time $t-i$ is included in count $W_t$, conditional on it ever being in one of the counts.
 $f$ is known as a \emph{delay distribution}.
@@
 Solving \cref{intro:eq:disease-model} for $Z$ is an ill-posed problem when $f(0)$ is small.
-The forwards error of the solution is large when the ratio of the off-diagonal to diagonal elements is large~\autocite[chapter 8.2]{highamAccuracy}; for this situation, that is when the value of $\max_{i > 0} f(i) / f(0)$ is large.
-A large forwards error means that the computed value is far from the true solution~\autocite[6--9]{highamAccuracy}.
-Either an approach which gives a large $f(0)$ or smoothing on $Z_t$ is required.
+The forwards error of the solution is large when the ratio of the off-diagonal to diagonal elements is large~\autocite[chapter 8.2]{highamAccuracy}; for this situation, which is when the value of $\max_{i > 0} f(i) / f(0)$ is large.
+A large forwards error means that the computed value is far from the true solution~\autocite[6--9]{highamAccuracy}; this is avoided by either using an approach which gives a large $f(0)$ or smoothing $Z_t$.
 
@@
 An observation model links $W_t$ to a possibly biased and/or noisy observation of it at time $t$.
+
 Biases can arise in a variety of ways, \eg healthcare-seeking behaviour or tests which are less than 100\% accurate.
 Bias can be modelled through additional parameters or data quantifying these issues.
+
 Noise could come from a variety of sources, \eg sampling error, clustering, or unexplained overdispersion.
-Noise is typically modelled by a parameteric distribution \eg the negative binomial distribution allowing for overdispersion compared to a Poisson process assumption.
+Noise is typically modelled by a parametric distribution \eg the negative binomial distribution allowing for overdispersion compared to a Poisson process assumption.
 
@@
 Two approaches are commonly used to estimate the duration of a disease stage: either modelling an underlying biomarker or modelling the duration directly~\autocite[e.g.][]{sweetingEstimating}.
-Modelling a biomarker uses more information per observation, but requires stronger assumptions.
+Modelling a biomarker uses more information per observation but requires stronger assumptions.
 
 The biomarker approach models the trajectory of a biomarker over time in a cohort of individuals.
-An individual is considered to have the disease while the biomarker is above or below some threshold.
-Parametric assumptions are commonly required for this approach.
-For example, a functional form of the behaviour over time of the biomarker is required.
+An individual is in the disease state while the biomarker is above or below some threshold.
+This approach often requires parametric assumption(s).
+For example, the functional form of the biomarker's trajectory.
 In addition, if individual-level variation wants to be included, which is generally important, a distributional form of random effects is required.
@@
 
-Modelling the duration directly considers only the binary result at each test (positive or negative); for instance in a \emph{survival analysis} framework.
+Modelling the duration directly considers only the binary result at each test (positive or negative), for instance in a \emph{survival analysis} framework.
 This approach generally requires fewer parameters, and no random effect(s), reducing the computational burden.
@@
 
-In both cases, longitudinal data is required.
-Therefore, cohort studies, where the same group of individuals are followed over time, are the most appropriate study design.
+Both cases require longitudinal data, suggesting that cohort studies are the most appropriate study design.
 
@@
 Modern technology allowed large amounts of data to be collected, distributed, and analysed.
-While the HIV/AIDS pandemic had inspired a large amount of statistical methodology to be developed, the pandemic was far slower moving.
+While the HIV/AIDS pandemic inspired the development of a large amount of statistical methodology, the pandemic was far slower moving.
 Much analysis of SARS-CoV-2 data was done at great haste, requiring trade-offs to be made between timeliness and rigour of results presented.
 
-Deep understanding of each data source's potential was not possible.
+A deep understanding of each data source's potential was not possible.
 In this thesis, I will investigate the potential of the prevalence surveys in greater detail.
+The disease model will be a prevalence model (defined further in \cref{inc-prev:sec:prevalence-process}) with the delay distribution being the duration of PCR positivity.
 
@@
 
-Seroprevalence surveys were conducted throughout the epidemic, primarly in blood donors.
-In addition to the normal problem of the delay between infection and antibody response, blood donors are a  convenience samples whose representativeness of the wider population is unclear.
+Seroprevalence surveys were conducted throughout the epidemic, primarily in blood donors.
+In addition to the normal problem of the delay between infection and antibody response, blood donors are a convenience sample whose representativeness of the wider population is unclear.
 
@@
 Therefore, some modelling is required to borrow information across strata or time to provide stable estimates, as well as to correct for non-response bias (see \cref{E-intro:sec:cis}).
-Despite this issue, they were considered the gold-standard metric.
+Despite this issue, they are considered the gold-standard metric.
 
@@
 REACT was a series of cross-sectional surveys, run across 19 rounds between May 2020 and March 2022.
-Each round consisted of at least 100,000 randomly-sampled individuals from the UK contributing swabs for testing and questionnaire data~\autocite{elliottTwin}.
+Each round consisted of at least 100,000 randomly sampled individuals from the UK contributing swabs for testing and questionnaire data~\autocite{elliottTwin}.
 
@@
 
-Many cohort studies to estiamte a wide variety of quantities ran throughout the pandemic.
+Many cohort studies to estimate a wide variety of quantities ran throughout the pandemic.
 In this thesis, I make use of two cohort studies with differing designs to estimate the duration or PCR positivity.
@@
 % For example, contact tracing data can inform how likely it is for transmission to occur in different settings and/or between different groups of people.
-The first study I use, ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts)~\autocite{singanayagamCommunity,hakkiOnset}, performed daily testing of a small number of individuals with short follow-up.
-The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability, but the results are limited because the small sample size and short follow-up means that the tails of these distributions cannot be estimated.
+The first study I used, ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts)~\autocite{singanayagamCommunity,hakkiOnset}, performed daily testing of a small number of individuals with short follow-up.
+The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability, but the results are limited because the small sample size and short follow-up prohibit estimation of the tails of these distributions.
 I describe ATACCC in detail in \cref{E-biology-data:sec:ataccc} and use it to estimate duration in \cref{E-ATACCC}.
@@
 In addition to its role as a prevalence survey, the CIS followed the same individuals over a long period of time (over a year in many cases), collecting longitudinal data.
-The large sample size and long follow-up mean that the tails of the distribution of the duration of positivity can be estimated.
-However, the lack of frequent testing (at most weekly) means that the viral load dynamics cannot be estimated.
+The large sample size and long follow-up allows estimation of the tails of the distribution of the duration of positivity.
+However, the lack of frequent testing (at most weekly) prohibits estimation of the viral load dynamics.
 In particular, the vast majority of detected infections had only one or two positive tests, and hence the viral load trajectory of these infections cannot be estimated.
@@
 
-In the UK, the most modelling was produced by SPI-M-O (the Scientific Pandemic Influenza Group on Modelling, Operational sub-group).
+In the UK, SPI-M-O (the Scientific Pandemic Influenza Group on Modelling, Operational sub-group) produced the most prominent modelling
 SPI-M-O was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), generating evidence based on infectious disease modelling for policy development~\autocite{medleySPIM,govSPIMO}.
@@
 Each week, the members of SPI-M-O would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
-These estimates, once any estimates unduly affected by issues (\eg data quality issues), were statistically combined into a consensus value~\autocite{parkCombining}.
+The committee reviewed these estimates, removing any estimates unduly affected by issues (e.g..
+data quality issues), and then formed a consensus value using a statistical combination of the estimates~\autocite{parkCombining}.
 The consensus value was published on the UK Government's website and informed its response to the epidemic~\autocite{govRnumber}.
@@
 Some of these models were used to explore policy scenarios.
-For example, each of the stages of lifting the lockdown in 2021 was supported by modelling studies informing the government's decision~\autocite{sageEvidence}.
+For example, the decision at each of stage of lifting the lockdown in 2021 was informed by modelling studies~\autocite{sageEvidence}.
 These aided decision-makers in understanding plausible impacts of their actions.
@@
 
-\emph{Backcalculation} is a modelling framework developed in the context of HIV/AIDS~\autocite{brookmeyerMethod}, but applied very broadly.
+\emph{Backcalculation} is a modelling framework developed in the context of HIV/AIDS~\autocite{brookmeyerMethod} but applied very broadly.
 It specifies a particular combination of infection, disease, and observation model to relate times of HIV infection with times of AIDS diagnosis via the incubation period.
@@
 
-To apply backcalculation to SARS-CoV-2 prevalence surveys, adaptations to the disease and observation models are needed (elaborated on in \cref{E-inc-prev}).
+The disease and observation models must be adapted to apply backcalculation to SARS-CoV-2 prevalence surveys (elaborated on in \cref{E-inc-prev}).
 The disease model must consider that this is now prevalence-type, rather than incidence-type data.
 This difference induces a dependency structure between the observations on each day.
-The observation model must consider that this is now a survey, with sampling error, rather than assuming all infections are observed.
+The observation model must consider that this is now a survey, with sampling error.
 
@@
 
-\Textcite{hellewellPCRSensitivity} estimated the probability of an individual testing positive at each day after infection.
+\Textcite{hellewellPCRSensitivity} estimated the probability of an individual testing positive each day after infection.
 This measure is highly related to the duration of positivity.
-The cohort in their study was drawn from healthcare workers, and hence was focused on working-age adults.
+The study's cohort consisted of healthcare workers, \ie working-age adults.
 
@@
 The situation in New Zealand, who were successfully pursuing a zero-COVID strategy at this time, means that the individuals in the study were unusual.
-For example, a third were travelers detected during their quarantine period on arrival to New Zealand, but would have had negative RT-PCR tests before their departure.
+For example, a third were travellers detected during their quarantine period on arrival to New Zealand but would have had negative RT-PCR tests before their departure.
 
@@
 This assumption is necessary because they had at most 17 days of follow-up, and hence could not estimate the tail of the distribution.
-Overall, they estimate a median duration of 14.0 days (95\% CI: 12.9--15.4 days) but that this varied substantially by the viral load of an individual at the time of the initial positive test.
-Individuals with a higher viral load were estimated to have a longer duration of positivity; the highest viral load individuals were estimated to have a median duration of 19.6 days (16.1--25.8 days).
-This could be due to the higher viral load leading to a longer duration of positivity, or that individuals with a higher viral load were more likely to have been recently infected at the time of the initial positive test.
-They cannot estimate how long before the initial positive test the individual was infected; therefore, they cannot distinguish between these two possibilities.
-This also means they underestimate the duration of positivity.
+Overall, they estimate a median duration of 14.0 days (95\% CI: 12.9--15.4 days) but with substantial positive correlation with each individual’s viral load at the time of the initial positive test (\ie a higher initial viral load was associated with a longer duration of positivity)
+The individuals with the highest viral load individuals were estimated to have a median duration of 19.6 days (16.1--25.8 days).
+The association could be due to individuals with a higher initial viral load being (on average) more recently infected at the time of the initial positive test.
+The study design prohibits estimating how long before the initial positive test the infection occurred.
+The same issue means the study underestimates the duration of positivity.
 
-For the purposes of this thesis, estimates applicable to the general population are required.
+For the purposes of this thesis, I require estimates applicable to the general population.
 There are no reliable estimates of the full distribution of the duration of positivity in a general population cohort available in the literature.
-Therefore, I estimate the duration of positivity from the ATACCC and CIS studies, as described above.
+Therefore, I estimate the duration of positivity from the ATACCC and CIS studies.
 
@@
 
-Several previous studies have used SARS-CoV-2 prevalence surveys to estimating incidence and/or transmission.
+Several previous studies have used SARS-CoV-2 prevalence surveys to estimate incidence and/or transmission.
 
@@
 \Textcite{abbottCISincidence} used \textcite{hellewellPCRSensitivity}'s estimates of the sensitivity of testing to inform the observation model, which are based on a longitudinal study in healthcare workers (see \cref{intro:sec:previous-duration-estimates}).
-They fit to summary statistics of the CIS data, using a daily, independent Gaussian likelihood with parameters estimated from the posterior point estimate and 95\% credible interval published by the ONS.
+They fit to summary statistics of the CIS data, using a daily, independent normal likelihood with parameters estimated from the posterior point estimate and 95\% credible interval published by the ONS.
 
@@
 Their methodology was similar to \textcite{abbottCISincidence}, including using estimates from \textcite{hellewellPCRSensitivity} for how the sensitivity of RT-PCR testing varies over an infection.
-Rather than using a Gaussian process, as \textcite{abbottCISincidence} do, \textcite{colmanAscertainment} provide stability by assuming the ascertainment rate is assumed to be piecewise constant.
+Rather than using a normal process to provide stability, as \textcite{abbottCISincidence} do, \textcite{colmanAscertainment} assume the ascertainment rate is piecewise constant.
 They do not fully propagate the uncertainty from the CIS prevalence estimates, considering only the bounds of the 95\% confidence intervals the ONS publishes for prevalence.
@@
 % The length of stay in these compartments (the duration of time positive) is assumed to follow a gamma distribution, with the parameters estimated from external data or having at most one free parameter.
-The duration of positivity in these models is limited by the model structure.
-The models are fit to the prevalence survey data using a binomial likelihood where possible, although some approaches take a Gaussian approximation.
+Their model structure limits the form of the duration of positivity. 
+They fit the models to prevalence survey data using a binomial likelihood where possible, although some approaches take a normal approximation.
 % These approaches are similar to how these models are fit to other data sources, such as hospital admissions or deaths.
@@
 They estimate a posterior distribution of the parameters of a spline by fitting to the prevalence results via a binomial distribution.
-Their setup penalises the spline deviating from a constant growth rate.
-The growth rate is then calculated from the derivative of the spline, an analytical transformation of the parameters.
+Their set-up penalizes the spline deviating from a constant growth rate.
+They calculate the growth rate from the derivative of the spline, an analytical transformation of the parameters. 
 Since, under reasonable model assumptions, there is a one-to-one correspondence between reproduction numbers and growth rates~\autocite{wallingaGI}, and they assume that the growth rate in incidence and prevalence is equal, this also estimates $\Re$.
@@
 There are two components to this question.
-First, how can prevalence surveys be designed to help estimate these quantities.
-Second, what statistical methodology can be used to estimate these quantities from prevalence surveys.
+First, how can prevalence surveys be designed to help estimate these quantities?
+Second, what statistical methodology can be used to estimate these quantities from prevalence surveys?
 This thesis will focus on the second component.
@@
 These include that the CIS was of sufficient size and incidence sufficiently high that there is enough data to perform the analyses.
-Additionally, incidence begins rising over this period, which is the period when a surveillance system is needed to provide a warning of upcoming issues.
+Additionally, incidence begins rising over this period, which is the period when a surveillance system needs to provide a warning of upcoming issues.
 The CIS began in April 2020, when incidence was falling due to the first lockdown in the UK.
@@
 The deterministic approach models the prevalence at any time as a deterministic function of the incidence, rather than as a random variable.
-I show that, this is a good approximation as long as the proportion of the population sampled is small.
+I show that this is a good approximation as long as the proportion of the population sampled is small.
 In particular, it does not require the prevalence or incidence to be large.
@@
 Unlike \textcite{nicholsonImproving}, I include age-stratification in the transmission model, allowing estimation of how transmission varies by age.
-I use only data from the CIS, and show this is sufficient for estimating incidence and transmission.
+I use only data from the CIS and show this is sufficient for estimating incidence and transmission.
 
@@
 
-The secondary aim of this thesis is providing new estimates to improve our understanding of SARS-CoV-2 in general, and the epidemic in England over the period being studied.
+The secondary aim of this thesis is providing new estimates to improve our understanding of SARS-CoV-2 in general, and the epidemic in England over the period I study.
 
 In \cref{E-imperf-test}, I present estimates of the duration of RT-PCR-positivity in the general population.
-These estimates could be used to inform public health policy, for example, in interpreting RT-PCR test results or formulating quarantine policies.
+These estimates could inform public health policy, for example, in interpreting RT-PCR test results or formulating quarantine policies.
 
-In \cref{E-transmission}, I present estimates of the incidence and transmission of SARS-CoV-2 in England over the period being studied.
+In \cref{E-transmission}, I present estimates of the incidence and transmission of SARS-CoV-2 in England over the period.
 These estimates contribute to the literature reconstructing the SARS-CoV-2 epidemic in England.
@@
 
-In order to best fulfill this thesis's primary aim, the methodology developed should be easy for others to use.
+In order to best fulfil this thesis's primary aim, the methodology developed should be easy for others to use.
 Therefore, the code for the analyses in this thesis is available online.
 An index to the various repositories is available at\todo{insert GitHub link}.
-Two elements of the code have been structured as R packages to enable their application to future contexts.
+I structured two elements of the code as R packages to enable their application to future contexts.
 
@@
 The package enables customisation of various features of the study, for example the testing regime of the participants, the sensitivity of the tests, and the duration of positivity.
-This package would be particularly useful in answering questions regarding the design of future prevalence surveys, for example, how frequently should individuals be tested, or how many individuals should be tested.
+This package would be particularly useful in answering questions regarding the design of future prevalence surveys, for example, how frequently individuals should be tested, or how many individuals should be tested.
 
 Second, the survival analysis framework used to infer the duration of positivity in \cref{E-perf-test,E-imperf-test}.
-The framework allows for all the different analyses performed in these chapters to be specified by the user in a simple interface.
+The framework allows the user to specify all the different analyses performed in these chapters in a simple interface. 
 Furthermore, adding new analyses, for example varying the priors or the inclusion of covariates, is straightforward.
@@
 Almost all code, excluding dependencies, was written from scratch for these analyses.
-The exception is the Markov chain Monte Carlo sampler used in \cref{E-SEIR}, which was written by Sanmitra Ghosh for the article \textcite{ghoshApproximate}.
+The exception is the Markov chain Monte Carlo sampler used in \cref{E-SEIR}, written by Sanmitra Ghosh for the article \textcite{ghoshApproximate}.
 
@@
 This provides a reliable estimate of the bulk of the distribution (out to around 20 days) but shows that there is a significant tail.
-To estimate the tail, a large sample size and long follow-up is required.
+Estimating the tail requires a large sample size and long follow-up. 
 Therefore, I turn to the CIS data in \cref{E-perf-test,E-imperf-test}.
@@
 
-Definitions of the probability distributions used throughout this thesis, are given in \cref{E-distributions}.
+\Cref{E-distributions} provides definitions of the probability distributions used in this thesis.
 Other appendices provide additional details.

### 2024-03-19 paper-surv-analysis.tex
spelling / grammar check

@@
 Finally, $\exists t \in \sched_i$ such that $t > e_{j(i)}$, to upper bound the end of the episode; this condition will be met for any episode sufficiently far in the past in individuals not lost to follow-up (almost always the case in this context).
-For a new context where recent infections need to be considered, this condition can be relaxed by considering these episodes as right-censored, at little cost except a more complicated likelihood expression.
+For a new context where recent infections need to be considered, this condition can be relaxed by considering these episodes as rightcensored, at little cost except a more complicated likelihood expression.
 
@@
 If it is a false negative, then I will consider the episode's end right-censored.
-However, if it is a true negative, then the episode's end is interval-censored, as previously.
+However, if it is a true negative, then the episode's end is interval censored, as previously.
 A mixture of these scenarios forms the episode's likelihood contribution, with the mixture probability determined by the test sensitivity.
@@
 
-First, if $e_{j(i)} \leq r_{j(i)}^{(e)}$, meaning that the test at $r_{j(i)}^{(e)}+1$ is a true negative and the end of the episode is interval-censored as in the previous chapter.
+First, if $e_{j(i)} \leq r_{j(i)}^{(e)}$, meaning that the test at $r_{j(i)}^{(e)}+1$ is a true negative and the end of the episode is interval censored as in the previous chapter.
 % In this case, the test at $r_{j(i)}^{(e)} + 1$ is a true negative, as are all other tests not in $\sched'_i$.
@@
 The initiating event is assumed known as the time the episode was detected.
-The final event is assumed interval-censored between the time of the final positive test and the subsequent negative test, or right-censored if a negative test has not yet been observed.
+The final event is assumed interval censored between the time of the final positive test and the subsequent negative test, or right-censored if a negative test has not yet been observed.
 A flexible, spline-based form is assumed for the baseline survival function~\autocite{roystonSTPM,roystonFlexible} with covariates introduced via proportional odds.

### 2024-03-19 transmission.tex
spelling / grammar check

@@
 In this chapter, I use the estimate of the duration of RT-PCR positivity from combining information from the ATACCC and CIS studies (see \cref{E-imperf-test:sec:conclusion}) to investigate the feasibility of estimating incidence and transmission of SARS-CoV-2 from only the CIS data (described in \cref{E-intro:sec:cis}).
-I will show that the CIS data is sufficient; demonstrating the value of surveys of this type for pandemic surveillance.
+I will show that the CIS data is sufficient, demonstrating the value of surveys of this type for pandemic surveillance.
 
@@
 There are several attractive features about this approach in the context here.
-The backcalculation approach makes no assumptions about the transmission process, which have the potential to introduce bias.
+The backcalculation approach makes no assumptions about the transmission process; the assumptions have the potential to introduce bias.
 It can also easily incorporate arbitrary covariates to adjust for non-response bias.
@@
 The model $g_Z$ would be specified to share information across time and/or strata.
-This both helps stabilise the estimation (see\todo{ref wherever I end up explaining backcalculation stability issues}) and allows borrowing of information across strata.
+This both helps stabilise the estimation (see \cref{intro:sec:disease-model}) and allows borrowing of information across strata.
 Borrowing information is required because some strata and time combinations have few positives (fewer than five per day).
@@
 \end{align}
-where $g_P$ is analogous to $g_Z$ in providing a joint prior for the prevalence in each strata and day combination.
+where $g_P$ is analogous to $g_Z$ in providing a joint prior for the prevalence in each stratum and day combination.
 
@@
 \begin{enumerate}
-    \item The smoothing model over time should penalise changes from continuous exponential growth.
-        Penalising changes from continuous exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time.
-        Furthermore, such penalisation has previously been shown to improve estimates of SARS-CoV-2 prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
+    \item The smoothing model over time should penalize changes from continuous exponential growth.
+        penalizing changes from continuous exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time.
+        Furthermore, such penalization has previously been shown to improve estimates of SARS-CoV-2 prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
     \item Limit estimates to epidemiologically and biologically plausible values.
@@
     \item Borrow information across the different covariates.
-        The borrowing should penalise changes in the distribution of incidence/prevalence across strata in the same geographic region being constant.
+        The borrowing should penalize changes in the distribution of incidence/prevalence across strata in the same geographic region being constant.
         A constant distribution is predicted by the theoretical study of epidemics, for example the dynamics of stratified, mechanistic transmission models (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) over short periods of time with negligible behavioural and immunity changes.
-    \item Be computationally efficient
+    \item Be computationally efficient.
         The model will be fit to granular data and therefore needs to be run within the SRS, with its computational limitations (see \cref{E-biology-data:sec:SRS}).
@@
 
-Properties 1 and 2 suggest a model of the form $\logit(\pi_{i,t}) = g(i, t)$, where $g$ is a function with changes in the first derivative penalised (\ie changes away from linear are penalised).
+Properties 1 and 2 suggest a model of the form $\logit(\pi_{i,t}) = g(i, t)$, where $g$ is a function with changes in the first derivative penalized (\ie changes away from linear are penalized).
 The $\logit$ link function both limits $\pi_{i,t}$ to the interval $[0, 1]$, and logit-linear growth approximates exponential growth for small $\pi_{i,t}$~\autocite{ealesAppropriately}.
-Combined with property 4, the form of $g$ the null model is $g(i, t) = a_0 + a_t t + f(i)$ where $a_0$ and $a_t$ are model parameters and $f$ is a function to be specified; that is, changes away from this model should be penalised.
+Combined with property 4, the form of $g$ the null model is $g(i, t) = a_0 + a_t t + f(i)$ where $a_0$ and $a_t$ are model parameters and $f$ is a function to be specified; that is, changes away from this model should be penalized.
 
@@
 In practice, this means that $\eta_j$ is a one-to-one transformation of the population prevalence in stratum $i$ on day $t$.
-If $\eta_j = \logit(\pi_{i,t})$, where $\pi$ is prevalence, then a model of $\eta_j = g(i, t)$, as before, is a LGM.
+If $\eta_j = \logit(\pi_{i,t})$, where $\pi$ is prevalence, then a model of $\eta_j = g(i, t)$, as before, is an LGM.
 However, if $\pi$ is instead representing incidence proportion, then no such linear relationship between $\eta_j$ and $g$ exists.
@@
 \end{align}
-As $g(i, k)$ is linear in the model parameters, $\vec{z}$, then $\eta_{i,t}$ is not, violating the requirements of a LGM.
+As $g(i, k)$ is linear in the model parameters, $\vec{z}$, then $\eta_{i,t}$ is not, violating the requirements of an LGM.
 
@@
 % \Cref{E-inc-prev:eq:EPt}, considered across different values of $t$, can be written as $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix formed from the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
-Dividing each side by the total number of individuals in each stratum transforms number of prevalent individuals to prevalence and incidence to incidence proportion.
+Dividing each side by the total number of individuals in each stratum transforms the number of prevalent individuals to prevalence and incidence to incidence proportion.
 
@@
 The second set of assumptions are in specifying the statistical model used to estimate prevalence, for example the form of the smoothing used.
-In this context, that means the prior $\pi_P$ places non-zero mass on prevalence values which imply negative incidence under the other model assumptions.
+In particular, the prior for $\pi_P$ places non-zero mass on prevalence values which imply negative incidence under the other model assumptions.
 
@@
 
-Once the incidence proportion in each stratum has been calculated using this procedure, I poststratify it to the CIS target population and subpopulations of interest.
+Once the incidence proportion in each stratum have been calculated using this procedure, I poststratify it to the CIS target population and subpopulations of interest.
 This is done using the methodology explained in \cref{E-biology-data:sec:MRP}.
@@
 For example, a Laplace approximation or variational inference.
-Further work, to check the quality of these approximations, is required.
+Further work to check the quality of these approximations, is required.
 
@@
 The application of this framework follows \textcite{birrellRealtime}.
-The largest modification from the generic framework presented in \cref{E-SEIR} is the parameterisation of the WAIFW matrix (\cref{SEIR:sec:WAIFW-application}).
+The largest modification from the generic framework presented in \cref{E-SEIR} is the parameterization of the WAIFW matrix (\cref{SEIR:sec:WAIFW-application}).
 A novel approach (\cref{SEIR:sec:observation}) links this transmission model to the CIS prevalence data; the approach is generalisable to other studies collecting representative prevalence data, \eg REACT.
@@
 I model each region fully independently, \ie with no shared parameters or transmission between regions.
-As well as the motivation given in \cref{E-backcalc}, independent regions allows heterogeneity in the parameters between regions, which could arise from a range of socioeconomic factors.
+As well as the motivation given in \cref{E-backcalc}, independent regions allow heterogeneity in the parameters between regions, which could arise from a range of socioeconomic factors.
 Furthermore, the assumption of homogeneous mixing within age groups is more plausible for regions than the whole of England.
@@
 Clearly, this is not realistic as travel between regions of England is common and unrestricted.
-However, except in an epidemic's early stages, the interaction between large regions has a negligible impact on the epidemics progression~\autocite[e.g.][]{birrellRealtimea,gogSpatial,eggoSpatial}.
+However, except in an epidemic's early stages, the interaction between large regions has a negligible impact on the epidemic's progression~\autocite[e.g.][]{birrellRealtimea,gogSpatial,eggoSpatial}.
 
@@
 
-A parsimonious model is needed to parameterise the WAIFW matrix.
+A parsimonious model is needed to parameterize the WAIFW matrix.
 In \cref{E-SEIR:sec:structured-populations}, I explained how the elements of the WAIFW matrix, $\matr{\beta}$ can be decomposed as $\beta_{aa'} = \kappa_{aa'} q_{aa'} N_{a'} / N_{a}$ where\todo{should I repeat these definitions?
@@
 Here, the daily incidence proportion for each stratum derived by solving the system of ODEs governing the transmission model (see \cref{E-SEIR:sec:SIR}), and then prevalence calculated using \cref{E-inc-prev:eq:EPt}.
-However, this approach is computationally expensive, likely prohibitively so if performed within a MCMC loop.
+However, this approach is computationally expensive, likely prohibitively expensive if performed within a MCMC loop.
 
@@
 This approach allows an infected, pre-positive compartment, which I denote $p_0$, to be included.
-Up until now, this stage of the disease (shown in \cref{E-biology-data:fig:natural-history}) has been neglected, assumed to be of negligble duration.
+Up until now, this stage of the disease (shown in \cref{E-biology-data:fig:natural-history}) has been neglected, assumed to be of negligible duration.
 An advantage of this modelling approach is that it can be included without complications (\eg issues of numerical stability).
@@
 I link the modelled proportion of individuals that are RT-PCR positive to the data using a beta-binomial likelihood (as discussed in \cref{E-biology-data:sec:clustering}).
-If on each day $t$, each strata $i$ is observed to have $y_{at}$ positives out of $n_{at}$ tests, then the likelihood is:
+If on each day $t$, each stratum $i$ is observed to have $y_{at}$ positives out of $n_{at}$ tests, then the likelihood is:
 \begin{align}
@@
 The remaining details of the transmission model structure follow \textcite{birrellRealtime}.
-The transmission model is a SEIR model with two latent and two infectious states.
+The transmission model is an SEIR model with two latent and two infectious states.
 To form whole of England estimates, each region is modelled independently with the results aggregated, using the poststratification methodology explained in \cref{E-backcalc}.
@@
 The parameters governing the duration of positivity ($\omega$s) were explained previously (see \cref{SEIR:table:ec-params}).
-The initial conditions and transition rate parameters are unindentifiable in a SEIR model~\autocite{dankwaStructural}, therefore strongly informative priors or fixed values are used for them.
+The initial conditions and transition rate parameters are unidentifiable in an SEIR model~\autocite{dankwaStructural}, therefore strongly informative priors or fixed values are used for them.
 $\sigma_\epsilon$ and $\rho$ are given priors that favour a simpler model.
-The prior on $\sigma_\epsilon$ is a penalised complexity prior, favouring a null model of constant values for $\beta(t)$~\autocite{simpsonPenalising}.
+The prior on $\sigma_\epsilon$ is a penalized complexity prior, favouring a null model of constant values for $\beta(t)$~\autocite{simpsonPenalising}.
 The prior on $\rho$ favours a null model of a binomial likelihood.
@@
     Initial growth rate & $\psi$ & $\N(0.06, 0.04)$ & Weakly informative \\
-    Standard deviation of random walk steps on log scale & $\sigma_\epsilon$ & $\Exponential(80)$ & Penalised complexity prior \\
-    Overdispersion of observations & $\rho$ & $\Exponential(2 \times 10^5)$ & Penalising deviations from binomial
+    Standard deviation of random walk steps on log scale & $\sigma_\epsilon$ & $\Exponential(80)$ & penalized complexity prior \\
+    Overdispersion of observations & $\rho$ & $\Exponential(2 \times 10^5)$ & penalizing deviations from binomial
 \end{tabular}
-\caption[SEIR model priors]{Priors for each parameter. For details of the distributions and their parameterisations see \cref{E-distributions}.}
+\caption[SEIR model priors]{Priors for each parameter. For details of the distributions and their parameterizations see \cref{E-distributions}.}
 \label{SEIR:table:priors}
@@
 I compute the posterior distribution using MCMC.
-Specifically, I use an adaptive random-walk Metropolis-Hastings algorithm (discussed in \cref{E-MCMC}).
+Specifically, I use an adaptive random-walk Metropolis--Hastings algorithm (discussed in \cref{E-MCMC}).
 The proposal distribution is a multivariate normal that adapts to the covariance of the posterior.
-The adaption algorithm is based on \textcite[algorithm 4]{andrieuTutorial}, with the implementation based on \textcite{ghoshApproximate}'s work.
-Proposals are accepted or rejected using the standard Metropolis-Hastings (MH) acceptance probability.
+The adaptive algorithm is based on \textcite[algorithm 4]{andrieuTutorial}, with the implementation based on \textcite{ghoshApproximate}'s work.
+Proposals are accepted or rejected using the standard Metropolis--Hastings (MH) acceptance probability.
 See \cref{SEIR:MCMC-algorithm} for full details.
@@
  }
- \caption{Algorithm for adaptive random-walk Metropolis-Hastings.
+ \caption{Algorithm for adaptive random-walk Metropolis--Hastings.
 $\vec{\mu_i}$ and $\matr{\Sigma_i}$ are an estimate of the mean and covariance of the posterior distribution using information up to iteration $i$.
@@
 
-Following \textcite{birrellBayesian}, I reparameterise the model to improve identifiability.
-I parameterise $\beta_1$ in terms of the initial growth of the epidemic.
-I also use a parsimonious parameterisation of the initial conditions using the initial proportion of the population that are infectious and the initial proportion that are susceptible only.
+Following \textcite{birrellBayesian}, I reparameterize the model to improve identifiability.
+I parameterize $\beta_1$ in terms of the initial growth of the epidemic.
+I also use a parsimonious parameterization of the initial conditions using the initial proportion of the population that are infectious and the initial proportion that are susceptible only.
 
-Both reparameterisations rely on an assumption that the epidemic is in a steady-state exponential growth phase at time 0 (see \cref{E-SEIR:sec:SIR}).
+Both reparameterizations rely on an assumption that the epidemic is in a steady-state exponential growth phase at time 0 (see \cref{E-SEIR:sec:SIR}).
 During such an exponential growth phase, all components of the $\vec{e}$, $\vec{i}$ and $\vec{p}$ compartments grow at the same growth rate, $\psi$.
@@
 
-The growth rate $\psi$ is easily identifiable from the data, and therefore a parameterisation in terms of $\psi$ should be more computationally efficient.
-If this is the initial exponential phase, when $\vec{s} \approx 1$, then the basic reproduction number $\R$ for a SEIR model is linked to this growth rate by~\autocite{birrellBayesian,wearingAppropriate}:
+The growth rate $\psi$ is easily identifiable from the data, and therefore a parameterization in terms of $\psi$ should be more computationally efficient.
+If this is the initial exponential phase, when $\vec{s} \approx 1$, then the basic reproduction number $\R$ for an SEIR model is linked to this growth rate by~\autocite{birrellBayesian,wearingAppropriate}:
 \begin{align}
@@
 Hence setting $\beta_1 = R_\psi / D(\matr{\kappa}(t) \circ \matr{q} \circ \matr{N})$ gives $R_\psi = R_\beta$.
-Using this expression, $\beta_1$ has been reparameterised in terms of $\psi$, $q_c$, and fixed quantities.
+Using this expression, $\beta_1$ has been reparameterized in terms of $\psi$, $q_c$, and fixed quantities.
 
-To form a parsimonious parameterisation of the initial state, first focus on the initial proportion of the population that is in the infectious compartments.
+To form a parsimonious parameterization of the initial state, first focus on the initial proportion of the population that is in the infectious compartments.
 Recall, from \cref{E-SEIR:sec:structured-populations}, that in the exponential growth phase, the relative proportion of each stratum that is infectious is proportional to the dominant eigenvector of $\matr{\beta}$, which I denote $\vec{d}$.
@@
 First, I drew 100 parameter sets drawn from the priors (see \cref{SEIR:table:priors}), except that the prior on $\psi$ was $\LN(0.048, 0.0035)$.
-I used this more informative prior for the simulation study to restrict the simulations to initially-growing epidemics.
+I used this more informative prior for the simulation study to restrict the simulations to epidemics growing at $t=0$.
 The weakly informative prior, used for the main analysis, allows a wide range of epidemics including some that are shrinking at the start of the time period.
@@
 
-In this setup, the parameters $i^+$, $q_c$, and $\psi$ are fully identified; $\pi_0$ is not identified at all; the other parameters are partially identified, their posterior has moved away from the prior but there is insufficient information in the data to fully overcome the prior (see \cref{SEIR:fig:true-vs-posterior}).
+In this set-up, the parameters $i^+$, $q_c$, and $\psi$ are fully identified; $\pi_0$ is not identified at all; the other parameters are partially identified, their posterior has moved away from the prior but there is insufficient information in the data to fully overcome the prior (see \cref{SEIR:fig:true-vs-posterior}).
 The ability to identify $q_c$ is an advantage of using data which includes measurements on children, unlike data on hospitalisations or deaths.
 
-For $\vec{\pi}$ and $\rho$, the posterior median is very similar regardless of the true value (see \cref{SEIR:fig:true-vs-posterior}).
+For $\vec{\pi}$ and $\rho$, the correlation between the posterior median and the generative value is weak (see \cref{SEIR:fig:true-vs-posterior}).
 This suggests that these parameters are not practically identifiable in this setting.
@@
 The estimated timing of the peak in incidence varies by region, with the most common time being late December 2020 (see \cref{SEIR:fig:peak-incidence}).
-In Yorkshire and the Humber, the October peak is estimated to be greater than the winter peak; for North West there is significant probability on either being larger.
+In Yorkshire and the Humber, the October peak is estimated to be greater than the winter peak; for North West there is a significant probability on either being larger.
 \begin{figure}
@@
 The attack rate is highest in the $[11, 16)$ age group in the North West at 68\% (95\% CrI: 66--70\%) (see \cref{SEIR:fig:attack-rates}).
-Some of these infections occurred prior to the period covered by this study; during the period of this study 23\% (95\% CrI: 21--24\%) of the population in the North West aged 11--16 were infected.
+Some of these infections occurred prior to the period covered by this study; during the period of this study 23\% (95\% CrI: 21--24\%) of the population in the North West aged 11--16 was infected.
 \begin{figure}
@@
 
-Overall the model fits the data well (see \cref{SEIR:fig:prev-young,SEIR:fig:prev-old}).
+Overall, the model fits the data well (see \cref{SEIR:fig:prev-young,SEIR:fig:prev-old}).
 \todo{move one or some of the goodness-of-fit plots into an appendix - maybe by region?}
@@
         Goodness-of-fit to prevalence data by region and age for the younger three age groups (see \cref{SEIR:fig:prev-old} for the others).
-        The central ribbon and line shows the posterior median and 95\% CrI estimate for the proportion of the relevant strata RT-PCR positive on each day.
+        The central ribbon and line show the posterior median and 95\% CrI estimate for the proportion of the relevant stratum RT-PCR positive on each day.
         The lighter outer ribbon is the 95\% posterior predictive interval for the proportion of test results that are positive each week.
@@
 The parameters $\beta_w$ absorb all the time-varying factors not captured by Google mobility and school attendance data.
-Many factors contribute to the pattern in the $\beta_w$ values, which have a minima near the start of December.
+Many factors contribute to the pattern in the $\beta_w$ values, which have minima near the start of December.
 While this analysis cannot establish causation, several hypotheses are worth considering.
@@
 
-The weekly patterns in incidence are a limitation of the current parameterisation.
+The weekly patterns in incidence are a limitation of the current parameterization.
 The random walk and contact matrices both vary only on the weekly timescale.
@@
 However, this would greatly increase the number of parameters.
-An alternative would be a smoothly-varying function or an approximation of a random walk, such as that proposed by \textcite{ghoshApproximate}.
+An alternative would be a smoothly varying function or an approximation of a random walk, such as that proposed by \textcite{ghoshApproximate}.
 
@@
 \Cref{transmission:fig:compare-regions} compares the regional incidence estimates between the two models in this chapter.
-Generally the estimates agree, with the SEIR model having much narrower credible intervals.
-The largest discrepancy is that the backcalculation model estimates a second peak in mid January in several regions (most clearly in London, North West, and the East of England).
+Generally, the estimates agree with the SEIR model having much narrower credible intervals.
+The largest discrepancy is that the backcalculation model estimates a second peak in mid-January in several regions (most clearly in London, North West, and the East of England).
 However, except in London, the credible intervals still overlap.
-This is a specific case of a larger pattern: the SEIR model produces smoother estimates of incidence than the backcalculation model.;
+This is a specific case of a larger pattern: the SEIR model produces smoother estimates of incidence than the backcalculation model.
 \begin{figure}
@@
 The SEIR model imposes stronger assumptions about the epidemic dynamics.
-The assumptions reduces the variance of the estimates, including by making them smoother, but if the assumptions are incorrect, bias is introduced.
+The assumptions reduce the variance of the estimates, including by making them smoother, but if the assumptions are incorrect, bias is introduced.
 Furthermore, the SEIR model does not fully adjust for the non-response bias.
 In particular, it makes no adjustment for the under-representation of ethnic minority groups in the CIS.
-Adjusting for more covariates would require further stratifying the population, and finding an appropriate parameterisation of the contact matrix.
-This trade-off based on the strength of assumptions has long been known to apply in the backcalulcation context~\autocite[e.g.][section 8.3]{brookmeyerBackcalculation}.
-The SEIR model can be viewed as a backcalulation method where the transmisson part of the model is a smoothing method.
+Adjusting for more covariates would require further stratifying the population and finding an appropriate parameterization of the contact matrix.
+This trade-off based on the strength of assumptions has long been known to apply in the backcalculation context~\autocite[e.g.][section 8.3]{brookmeyerBackcalculation}.
+The SEIR model can be viewed as a backcalculation method where the transmission part of the model is a smoothing method.
 
@@
 Therefore, it is far more flexible and can easily include time-varying effects, ethnicity, etc.
-The SEIR model is limited by the inference procedure and the requirement to sovle an ODE to evaluate the likelihood.
+The SEIR model is limited by the inference procedure and the requirement to solve an ODE to evaluate the likelihood.
 To incorporate a more flexible model, and reduce its bias, a more efficient inference procedure would be required.
@@
 A similar survey could, therefore, be considered as the primary form of surveillance for a future pandemic.
-The advantage of such a setup is the unbiased information across the entire population, including children.
+The advantage of such a set-up is the unbiased information across the entire population, including children.
 

### 2024-03-19 introduction.tex
Overleaf comments

--- 2024-03-18 introduction.tex
@@
 \emph{Transmission} refers to the process by which an infectious disease generates new infections.
-Estimating transmission refers to inferring the parameters of this process, including its components.
-The components of transmission (\eg number of contacts per day and age-stratified rates) inform how interventions can be targeted.
+Estimating transmission refers to inferring the parameters of this process.
+The components of transmission (\eg the number of contacts each individual makes per day) allow interventions to be targeted.
 They are also important for forecasting future incidence.
@@
 Incidence-type and prevalence-type data can directly inform estimation of incidence and transmission, however, this estimation almost always relies on prior knowledge of epidemiological and/or biological quantities.
+For example, the time between infection and symptom onset.
 These quantities can be estimated from other studies, \eg cohort studies.
@@
 However, severe symptoms can be rare in some population subgroups (\eg young, otherwise healthy individuals infected with SARS-CoV-2).
-This means that the observations do not provide much information on incidence in those subgroups.
+This means that the observations do not provide much information on incidence in those subgroups.Furthermore, the delay between infection and the severe event occurring means that recent infections cannot be estimated.
 
@@
 \end{align}
-where $Z(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the instantaneous probability of an individual infected $\tau$ time units ago infects an individual at time $t$.
+where $Z(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the mean rate an individual infected $\tau$ time units ago infects individuals at time $t$.
 Therefore, the renewal equation relates incidence in the past to incidence in the future.
+This partially describes the transmission process, hence the name semi-mechanistic.
 The effective reproduction number can be constructed as a function of $\beta(t, \tau)$ and the generation interval.
@@
 Simple mechanistic models use a homogeneous population, with individuals differing only due to their disease state, and exponential distributions for the time in each disease state.
-More complex models relax these assumptions, in particular incorporating heterogeneity; the age structure of the population is important for respiratory diseases.
+More complex models relax these assumptions, in particular to incorporate heterogeneity.
+The age structure of the population is important for respiratory diseases because individuals meet individuals of their own age more frequently.
 % Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
@@
 
-The \emph{disease model} links the transmission model to a population-level quantity which we will make observations of.
-Examples are the incidence or prevalence of symptomatic individuals and admissions to hospital.
+\emph{Disease models} link transmission models to population-level quantities on which we will make observation.
+For example, the incidence or prevalence of symptomatic individuals and admissions to hospital.
+They will then be recorded through some system (\eg a testing programme) described by the observation model (see next section).
 
-The choice of observation model can, in theory, be independent of the transmission model.
-However, pragmatic considerations often mean that the observation model is chosen to be in a similar form to the transmission model.
+% The choice of observation model can, in theory, be independent of the transmission model.
+% However, pragmatic considerations often mean that the observation model is chosen to be in a similar form to the transmission model.
 
-Typically, the core of an observation model is a \emph{convolution equation} or an approximation of it.
+At the core of many observation models is a \emph{convolution equation} or an approximation of it.
 This equation describes the expectation of the quantity at time $t$, $W_t$, conditional on the incidence prior to time $t$.

### 2024-03-19 transmission.tex
Overleaf comments

--- 2024-03-12 transmission.tex
@@
 
-First, in \cref{backcalc}, I use a back-calculation approach, based on \cref{E-inc-prev:eq:EPt}.
+First, in \cref{backcalc}, I use a backcalculation approach, based on \cref{E-inc-prev:eq:EPt}.
 This approach enables estimation of incidence within (sub)populations of England.
 There are several attractive features about this approach in the context here.
-The backcalculation approach makes no assumptions about the transmission process, which have the potential to introduce bias.
+The backcalculation approach makes no assumptions about how transmission occurs, which have the potential to introduce bias.
 It can also easily incorporate arbitrary covariates to adjust for non-response bias.
@@
 
-\subsection{Approach} \label{backcalc:sec:approach}
+\subsection{Approaches} \label{backcalc:sec:approach}
 
@@
 
-\subsubsection{Possible approaches}
+% \subsubsection{Possible approaches}
 
@@
     \vec{Z} \mid \vec{\phi} &\dist g_Z(\vec{\phi}) \\
-    P_{i,t} &= \matr{S} \vec{Z}_i \\
+    \vec{P_{i}} &= \matr{S} \vec{Z}_i \\
     Y_{i,t} \mid \vec{P} &\dist \BB(n_{i,t}, P_{i,t})
@@
 \begin{align}
-    \vec{P} \sim \vec{\phi} &\dist g_P(\vec{\phi}) \\
-    Y_{i,t} &\dist \BB(n_{i,t}, P_{i,t})
+    \vec{P} \mid \vec{\phi} &\dist g_P(\vec{\phi}) \\
+    Y_{i,t} \mid \vec{P} &\dist \BB(n_{i,t}, P_{i,t})
 \end{align}
 where $g_P$ is analogous to $g_Z$ in providing a joint prior for the prevalence in each strata and day combination.
+Step 2 is...
 
-\subsubsection{Desirable properties}
+\subsection{Desirable properties}
 
@@
 However, it poses several pragmatic challenges which means I favour the prevalence-inferred approach.
-Here I explain that further developments, beyond the scope of this thesis, would be required to use the incidence-inferred approach with a suitable $\pi_Z$ within the computational constraints of the SRS.
+Here I explain that further developments, beyond the scope of this thesis, would be required to use the incidence-inferred approach with a suitable $g_Z$ within the computational constraints of the SRS.
 
@@
 \begin{enumerate}
-    \item The smoothing model over time should penalise changes from continuous exponential growth.
+    \item The smoothing model, $g_Z$ or $g_P$, over time should penalise changes from continuous exponential growth of $\pi$ over time.
         Penalising changes from continuous exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time.
         Furthermore, such penalisation has previously been shown to improve estimates of SARS-CoV-2 prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
+Used by CIS (section 3.?).
     \item Limit estimates to epidemiologically and biologically plausible values.
@@
         A constant distribution is predicted by the theoretical study of epidemics, for example the dynamics of stratified, mechanistic transmission models (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) over short periods of time with negligible behavioural and immunity changes.
-    \item Be computationally efficient
+    \item Be computationally efficient.
         The model will be fit to granular data and therefore needs to be run within the SRS, with its computational limitations (see \cref{E-biology-data:sec:SRS}).
@@
 
-Properties 1 and 2 suggest a model of the form $\logit(\pi_{i,t}) = g(i, t)$, where $g$ is a function with changes in the first derivative penalised (\ie changes away from linear are penalised).
+Properties 1 and 2 suggest a model of the form $\logit(\pi_{i,t}) = g(i, t)$, where $g$ is a function with changes in the first derivative penalised w.r.t.\ $t$ (\ie changes away from linear are penalised).
 The $\logit$ link function both limits $\pi_{i,t}$ to the interval $[0, 1]$, and logit-linear growth approximates exponential growth for small $\pi_{i,t}$~\autocite{ealesAppropriately}.
@@
 
-For $\vec{\pi}$ and $\rho$, the posterior median is very similar regardless of the true value (see \cref{SEIR:fig:true-vs-posterior}).
+For $\vec{\pi}$ and $\rho$, the correlation between the posterior median and the generative value is weak (see \cref{SEIR:fig:true-vs-posterior}).
 This suggests that these parameters are not practically identifiable in this setting.

### 2024-03-19 ATACCC-appendix-original-analysis.tex

@@
 
-This parameterisation leads to the following Ct value trajectory for individual $i$:
+This parameterization leads to the following Ct value trajectory for individual $i$:
 $$
@@
 
-Observed Ct values are assumed to be drawn from a Normal distribution with mean $\hat{C}_i(t)$ and standard deviation $\sigma_\text{obs}$.
+Observed Ct values are assumed to be drawn from a normal distribution with mean $\hat{C}_i(t)$ and standard deviation $\sigma_\text{obs}$.
 censored at the limit of detection.
@@
 test results, and $f_N(y \mid \mu, \sigma^2)$ and $F_N(y \mid \mu, \sigma^2)$ are the pdf
-and cdf (respectively) of a Normal distribution with mean $\mu$ and variance $\sigma^2$.
+and cdf (respectively) of a normal distribution with mean $\mu$ and variance $\sigma^2$.
 
@@
 Inference was performed by Hamiltonian Monte Carlo (HMC) implemented in Stan 2.26.1~\autocite{Stan-2-32-2}.
-To aid with mixing, the hierarchical parameters were implemented using the non-centred parameterisation~\autocites{papaspiliopoulosGeneral,stanReparameterization}.
-In addition, $t_{i,0}$ was reparameterised in terms of the absolute time of peak.
+To aid with mixing, the hierarchical parameters were implemented using the non-centred parameterization~\autocites{papaspiliopoulosGeneral,stanReparameterization}.
+In addition, $t_{i,0}$ was reparameterized in terms of the absolute time of peak.
 Convergence was assessed by checking R-hat < 1.1, and visual inspection of traceplots, and effective sample size.

### 2024-03-19 ATACCC.tex
Resolve to dos

@@
 
-Two major approaches have been taken to model SARS-CoV-2 viral load trajectories, both based on well-well-established methods within biostatisics.
-The first approach is an ODE-based approach uses parameters representing biological properties of the virus and host~\autocite[e.g.][]{ejimaEstimation,keVivo,kimQuantitative,goncalvesTiming,perelsonMechanistic}.
+Two major approaches have been taken to model SARS-CoV-2 viral load trajectories, both based on well-well-established methods within biostatistics.
+The first approach is an ODE-based approach that uses parameters representing biological properties of the virus and host~\autocite[e.g.][]{ejimaEstimation,keVivo,kimQuantitative,goncalvesTiming,perelsonMechanistic}.
 The model here is an application of a standard model for viral dynamics, used for many diseases~\autocite{perelsonMechanistic,hernandez-vargasInhost}.
-Such models are rarely analytically-tractable and the system's evolution is computed using numerical methods.
+Such models are rarely analytically tractable and the system's evolution is computed using numerical methods.
 % Both of these approaches have been used with other viruses and biomarkers.
-The second approach, and the focus of this chapter, is using a piecewise linear model for the biomaker trajectory and random effects to allow for within-individual correlation.
+The second approach, and the focus of this chapter, is using a piecewise linear model for the biomarker trajectory and random effects to allow for within-individual correlation.
 
@@
 Following much of the literature mentioned previously, I will take a Bayesian approach in this chapter.
-In the Bayesian context, random effect models are often referred to as hierarchical or multilevel models because they can be interpreted as placing a prior at two levels of a parameter hierarchy~\autocite[chapter 5]{gelmanBDA}.
+Bayesians often refer to random effect models as hierarchical or multilevel models because they can be interpreted as placing a prior at two levels of a parameter hierarchy~\autocite[chapter 5]{gelmanBDA}.
 % $\phi_j$ would be referred to as the bottom level of the hierarchy, with $\mu$ and $\sigma$ being the top level.
@@
 As NUTS uses gradient information, it is more efficient when the model is differentiable.
-This leads to the following model for the viral load (examples are shown in \cref{ATACCC:fig:viral-load-model}):
+This leads to the following model for the viral load (see \cref{ATACCC:fig:viral-load-model} for examples):
 \begin{align}
@@
     The peak is at time $\tau_{\max,i} = 0$ and log viral load $\log v_{\max,i}$.
-    The subscripts are omitted in the key for clarity.
+    The key omits subscripts for brevity.
   }
@@
 The noisy RT-PCR observations of viral log load are modelled by assuming each measurement of log viral load has an iid error distributed $\N(0, \sigma_v^2)$.
-The measurements are left-censored at the limit of detection, $\vlod$, meaning that any observation which would give a log viral load of less than $\vlod$ is instead observed as negative.
+Measurements are left censored at the limit of detection, $\vlod$, meaning that any observation which would give a log viral load of less than $\vlod$ is instead observed as negative.
 False positives and negatives are modelled by allowing a test failure with probability $\rho$.
@@
 \end{align}
-where $f_N(y \mid \mu, \sigma^2)$ and $F_N(y \mid \mu, \sigma^2)$ are the pdf and cdf at $y$ of a Normal distribution with mean $\mu$ and variance $\sigma^2$ respectively.
+where $f_N(y \mid \mu, \sigma^2)$ and $F_N(y \mid \mu, \sigma^2)$ are the pdf and cdf at $y$ of a normal distribution with mean $\mu$ and variance $\sigma^2$ respectively.
 The first term on each line is the likelihood contribution from test failures, either a false positive (top line) or false negative (bottom line).
@@
 The first group is unvaccinated and infected with a variety of pre-Alpha, Alpha, and Delta variants.
-The second group is vaccinated and all were infected by the Delta variant.
+The second group is vaccinated, and all were infected by the Delta variant.
 My focus is on the first group since they most closely correspond to the infections in the time period used throughout this thesis.
@@
 For example, an individual with a high peak viral load may also clear the virus slowly because both have an underlying cause of a weak immune system.
-To improve sampling efficiency, the multivariate version of the non-centred parameterisation~\autocite{papaspiliopoulosGeneral,stanReparameterization} is used.
+To improve sampling efficiency, the multivariate version of the non-centred parameterization~\autocite{papaspiliopoulosGeneral,stanReparameterization} is used.
 This can be expressed as follows.
@@
 These are defined such that, by construction, $(\matr{\Sigma}_\phi)_{ij} = (\vec{\sigma}_\phi)_i (\vec{\sigma}_\phi)_j (\matr{C}_\phi)_{ij}$.
-% Correlation matrices have diagonal elements all 1, therefore this reparameterisation has the same number of parameters as the original model.
+% Correlation matrices have diagonal elements all 1, therefore this reparameterization has the same number of parameters as the original model.
 
@@
 $\N_+$ indicates a normal distribution truncated to the non-negative reals.
-Details of the distributions, including their parameterisations, are in \cref{E-distributions}.
+Details of the distributions, including their parameterizations, are in \cref{E-distributions}.
 The elements of $\vec{\mu}_{\phi,g}$ and $\vec{\sigma}_\phi$ have independent priors placed on them, as specified in the first lines of the table.
@@
 These are the main parameters of interest (as explained in \cref{ATACCC:sec:duration}), suggesting that the results are robust to the choice of model.
-As the estimates are similar and the modified model converges better, I use only the results from the modified model from now on.
+As the estimates are similar and the modified model converges better, I will use only the results from the modified model.
 \begin{figure}
@@
 The second, found by the other six chains, is more symmetrical, peaking 7.2 days after the first test.
-The population-level parameter estimates agree between the two sets of chains, therefore the issue should not affect any estimates of interest.
+The population-level parameter estimates agree between the two sets of chains; therefore, the issue should not affect any estimates of interest.
 \begin{figure}
@@
 For example, the vaccinated individuals are likely to be older and/or have co-morbidities because the roll-out of the vaccine in the UK prioritised these groups~\autocite{naoCovidVaccination}.
-They were also infected by different variants because the vaccinated individuals were infected later.
+They were also infected by different variants because the infections in vaccinated individuals occurred later.
 All these factors could change the viral dynamics~\autocite{russellWithinhost,kisslerDynamicsVaccines} and hence the duration.
-Therefore, the relationships shown here cannot be considered causal, or even generalisable to contexts where these other confounders are not present.
+Therefore, the relationships shown here are not causal, or even generalisable to contexts where these other confounders are not present.
 Despite these confounders, the result is in agreement with previous work showing that the Alpha and Delta variants of concern (compared to pre-Alpha variants) and being vaccinated are all associated with shorter durations of RT-PCR positivity, although the vaccination effect is larger~\autocite{kisslerDynamicsVaccines}.
 
-A strength of the ATACCC data is that the sample is drawn from the general population.
-This is unlike most estimates (see\todo{ref section reviewing lit on duration}) which are from hospitalised patients or healthcare workers.
+A strength of the ATACCC data is its general population sample.
+This is unlike most estimates (see \cref{E-intro:sec:duration-SARS-CoV-2}) which are from hospitalised patients or healthcare workers.
 This means that, notwithstanding the issues highlighted in the previous paragraph, the estimates are likely to be generalisable to the general population.
@@
 Including a random effect on $\sigma_v$ could capture this affect.
-However, it would complicate the model, be unlikely to change the estimates of the parameters of interest, and may not be identifiable.
+However, it would complicate the model, be unlikely to change the estimates of the parameters of interest and may not be identifiable.
 
 In the ATACCC study, individuals were tested for at most 20 days.
-Yet, the estimates here sill have 33\% (95\% CrI: 21--46\%) of individuals positive at 20 days following infection.
+Yet, the estimates here still have 33\% (95\% CrI: 21--46\%) of individuals positive at 20 days following infection.
 The remainder of the estimate for $f_D$, such as that almost all (>99\%) of individuals recover before day 50, is driven by the assumed form of the viral load trajectory.
@@
 In the following chapters, I will use the estimates from the unvaccinated individuals.
-The CIS data I use is from autumn 2020 and early 2021 when few individuals were vaccinated and the majority of infections were by pre-Alpha or Alpha variants.
+The CIS data I use is from autumn 2020 and early 2021 when few individuals were vaccinated, and the majority of infections were by pre-Alpha or Alpha variants.
 Therefore, this group of people should be fairly representative of those within CIS.
@@
 A major issue with the estimates of duration presented here is that the number of individuals who test positive for a long period of time is based on extrapolation of the viral load model.
-In periods of low incidence following a large peak, the number of individuals who are currently positive is dominated by those who have been positive for a long time.
+In periods of low incidence following a large peak, prevalent individuals could be mostly infected a long time ago.
 Therefore, understanding if these probabilities are correct is important for understanding transmission in the period after a peak.

### 2024-03-19 biology-data.tex

@@
 PCR has many uses within the biological sciences; Mark R.\ Hughes, deputy director of the Human Genome Project, has claimed it is the most important scientific technology of the last hundred years~\autocite{powledgePCR}.
-It is central to the work in this thesis because it is used to detect the presence of SARS-CoV-2 in a sample taken from a human, most commonly a nose and/or throat swab.
+It is central to the work in this thesis because it can detect the presence of SARS-CoV-2 in a sample taken from a human, most commonly a nose and/or throat swab.
 
@@
 SARS-CoV-2 is an RNA virus, but PCR amplifies DNA.
-Therefore, before the PCR process can be used to detect SARS-CoV-2, the RNA must be converted to DNA.
+Therefore, before using PCR process to detect SARS-CoV-2, the RNA must be converted to DNA.
 This process is known as \emph{reverse transcription}.
-In reverse transcription, a naturally-occurring enzyme known as reverse transcriptase generates cDNA (complementary DNA) from the RNA in the sample~\autocite{valasekPower}.
+In reverse transcription, a naturally occurring enzyme known as reverse transcriptase generates cDNA (complementary DNA) from the RNA in the sample~\autocite{valasekPower}.
 Any sequence of cDNA has a one-to-one correspondence with a sequence of RNA.
@@
 
-At the start of the PCR process, four substances are required: template DNA, nucleotides, primers, and DNA polymerase~\autocite{garibyanPCR}.
+The PCR process requires four substances at its start: template DNA, nucleotides, primers, and DNA polymerase~\autocite{garibyanPCR}.
 The \emph{template DNA}, in this context, is the sample which is being tested~\autocite{caseTemplate}.
@@
 There are four nucleotides in DNA (A, C, G, T) each of which forms a pair, known as a \emph{base pair}, with its \emph{complementary} nucleotide (A with T, and G with C)~\autocite{batesBase}.
-The choice of \emph{primers} specifies the exact DNA which will be amplified; they are the complementary DNA sequence to the start and end of the DNA sequence to be amplified~\autocite{garibyanPCR}.
+The choice of \emph{primers} specifies the exact DNA to amplify; they are the complementary DNA sequence to the start and end of the DNA sequence to be amplified~\autocite{garibyanPCR}.
 DNA \emph{polymerase} is an enzyme required to link individual nucleotides together~\autocite{garibyanPCR}.
@@
 The process is a series of cycles, each cycle containing three steps~\autocite{powledgePCR,garibyanPCR}.
-In the first step, \emph{denaturation}, the DNA is heated to separate the two strands of DNA.
-In the second step, \emph{annealing}, the temperature is lowered to allow the primers to bind to the DNA.
-In the final step, \emph{extension}, the temperature is raised, causing the DNA polymerase to replicate the DNA sequence between the primers.
+The first step, \emph{denaturation}, heats the DNA causing its two strands to separate.
+The second step, \emph{annealing}, lowers the temperature, causing the primers to bind to the DNA.
+The final step, \emph{extension}, raises the temperature, causing the DNA polymerase to replicate the DNA sequence between the primers. 
 In the course of such a cycle, the amount of DNA doubles.
@@
 After each PCR cycle, the amount of target DNA is measured and compared to a pre-derived standard.
-Once the measured amount of DNA exceeds the standard, the target DNA is said to be detected.
+Once the measurement exceeds the standard, the target DNA is declared detected. 
 The number of cycles required to reach detection is known as the \emph{cycle threshold} (Ct) value.
-Originally, the amount of target DNA was only measured after the completion of all cycles, hence the name real-time.
-Quantitative real-time PCR have a number of other practical advantages over previous PCR techniques (\eg speed); see \textcite{yangPCRdiagnostics,valasekPower} for more details.
-All RT-PCR tests I consider in this thesis are quantitative real-time RT-PCR tests, and produce Ct values.
+Originally, measurement occurred only after the completion of all cycles, hence the name real-time.
+Quantitative real-time PCR has other practical advantages (\eg speed); see \textcite{yangPCRdiagnostics,valasekPower} for more details.
+All RT-PCR tests I consider in this thesis are quantitative real-time RT-PCR tests and produce Ct values.
 
 % The Ct value decreases as the amount of DNA present in the original sample increases~\autocite{hanRTPCR}.
-If there is more RNA present, proportionally more DNA is created in reverse transcription, and hence fewer cycles are required to reach detection; that is, the Ct value is lower~\autocite{hanRTPCR}.
-The number of DNA copies doubles each cycle, therefore the log Ct value decreases linearly with the quanitiy of RNA or DNA in the original sample (see \cref{biology-data:fig:ct-calibration}).
+If there is more RNA present, reverse transcription creates proportionally more DNA.
+More initial DNA means fewer cycles until detection; that is, the Ct value is lower~\autocite{hanRTPCR}.
+The number of DNA copies doubles each cycle, therefore the log Ct value decreases linearly with the quantity of RNA or DNA in the original sample (see \cref{biology-data:fig:ct-calibration}).
 A \emph{standard curve} translates the Ct value to viral load for a given test.
@@
 This is different to the \emph{clinical sensitivity} of the test, the ability of the test to detect a detectable individual.
-Clinical sensitivity can be affected by a large variety of preanalytical factors, such as sample collection (\ie the swabbing process) and sample transportation~\autocite{lippiPotential}.
+Clinical sensitivity is affected by a large variety of preanalytical factors, \eg sample collection (\ie the swabbing process) and sample transportation~\autocite{lippiPotential}.
 Clinical sensitivity is of primary interest in this thesis.
@@
 However, this does not necessarily mean that the virus is viable (\ie able to reproduce, a necessary condition for transmission to occur)~\autocite{puhachSARSCoV2}.
-High viral loads (low Ct values) are more likely to be viable, but more sophisticated tests are required to determine viability.
-The gold standard procedure is attempting to grow and recover viral material from a sample, a process known as \emph{cell culturing}~\autocite{singanayagamDuration,puhachSARSCoV2,hakkiOnset}.
+Determining viability requires more sophisticated tests, although high viral loads (low Ct values) are positively correlated with viability.
+The gold standard procedure for assessing viability is attempting to grow and recover viral material from a sample, a process known as \emph{cell culturing}~\autocite{singanayagamDuration,puhachSARSCoV2,hakkiOnset}.
 Further details on cell culturing are beyond the scope of this thesis.
@@
     (A) A simplified state-based representation of the natural history.
-    Shading represents states where the individual is detectable by a RT-PCR test.
+    Shading represents states where the individual is detectable by an RT-PCR test.
     (B) A fuller representation of the natural history, including the latent processes of the quantity of RNA genome copied and infectious virus within and individual (reproduced from \textcite{puhachSARSCoV2} with permission).
@@
 An individual infected with SARS-CoV-2 goes through several disease stages, depicted in \cref{biology-data:fig:natural-history}~\autocite{puhachSARSCoV2}.
-First, they are exposed to the virus and become infected.
-At this initial stage they are not infectious (cannot spread the disease to others), do not have symptoms, and will not test positive on a RT-PCR test.
-After a few days, they will start testing positive on a RT-PCR test; this is normally prior to being infectious or having symptoms.
+First, the individual is exposed to the virus and becomes infected.
+At this initial stage they are not infectious (cannot spread the disease to others), do not have symptoms, and will not test positive on an RT-PCR test.
+After a few days, they will start testing positive on an RT-PCR test; this is normally prior to being infectious or having symptoms.
 Next, the individual becomes infectious; normally, this occurs prior to symptom onset for those who experience symptoms.
 Recovery from symptoms occurs after a few days, around the time the individual is no longer infectious.
-Finally, the individual will test negative on a RT-PCR test.
+Finally, the individual will test negative on an RT-PCR test.
 
-The time for each of these stages depends on various features of host and virus characteristics, and has a large amount of variation.
+The time for each of these stages depends on various features of host and virus characteristics and has a large amount of variation.
 In this thesis, I focus on pre-Alpha variants of SARS-CoV-2 in unvaccinated individuals.
 The time from infection to infectiousness is known as the \emph{latent period}.
-The length of the latent period is hard to estimate because neither the time of infection or the start of infectiousness can be observed.
-The length has been estimated based on careful contact tracing, where transmission times are thought to be known with a reasonable degree of certainty.
-These estimates vary, but suggest a mean length of 3--4 days for pre-Alpha variants~\autocite[and references therein]{zhaoEstimating}.
-The time from infection to experiencing symptoms is known as the \emph{incubation period}; it is undefined for aymptomatic individuals.
-As the symptom onset time (when symptoms start) can be observed directly, this is easier to estimate than the latent period.
+The length of the latent period is hard to estimate because we rarely observe either the time of infection or the start of infectiousness.
+Estimates of its length are based on careful contact tracing, where transmission times are thought to be known with a reasonable degree of certainty.
+These estimates vary but suggest a mean length of 3--4 days for pre-Alpha variants~\autocite[and references therein]{zhaoEstimating}.
+The time from infection to experiencing symptoms is known as the \emph{incubation period}; it is undefined for asymptomatic individuals.
+As the symptom onset time (when symptoms start) is often known, this is easier to estimate than the latent period.
 However, the unknown infection time still poses challenges, again requiring contact tracing data.
 Therefore, estimates are sensitive to the choice of data and statistical model.
-The mean length has been estimated as around 6 days~\autocite{wuIncubation,quesadaIncubation,aleneSerial}.
+Its mean length is around 6 days~\autocite{wuIncubation,quesadaIncubation,aleneSerial}.
 
@@
 Viral replication requires the virus to infect a host cell and then use the host cell's machinery to produce more virus; as more host cells are infected this process slows because there are fewer remaining potential host cells, known as \emph{target cells}, to infect.
+Host cells which can potentially be infected are known as \emph{target cells}.
+The rate at which host cells are infected slows over time because there are fewer remaining target cells to infect.
 The host's immune system can slow this process by various mechanisms; the most relevant for SARS-CoV-2 is by making some target cells less susceptible to infection.
-A host's immune response is triggered by the viral load, and becomes stronger as the viral load increases.
-A high viral load is associated with being more infectious, more likely to test positive on a RT-PCR test, and the start of symptoms~\autocite{puhachSARSCoV2,keVivo}.
+A host's immune response is triggered by the viral load and becomes stronger as the viral load increases.
+A high viral load is associated with being more infectious, more likely to test positive on an RT-PCR test, and the start of symptoms~\autocite{puhachSARSCoV2,keVivo}.
 
@@
 For all positive test results, the viral load was measured using the Ct value and then converted to a viral load using a standard curve (see \cref{biology-data:sec:PCR-process}).
-Other data such as demography and symptoms was also collected, although I do not use this information.
+Other data such as demography and symptoms were also collected, although I do not use this information.
 For further details see \textcite{singanayagamDuration,hakkiOnset}.
@@
 They excluded the remaining 115 contacts with at least one positive RT-PCR test, ATACCC did not capture the phase of the infection with increasing viral load.
-This exclusion was to ensure viral load trajectories could be identified.
+This exclusion was to ensure viral load trajectories are identifiable.
 I will revisit and extend this analysis in \cref{E-ATACCC}.
@@
 
-The CIS~\autocite{CIS} was setup in April 2020 to provide gold-standard measurements of the prevalence of SARS-CoV-2 in the community.
+The CIS~\autocite{CIS} was set up in April 2020 to provide gold-standard measurements of the prevalence of SARS-CoV-2 in the community.
 It was a longitudinal study with a representative sample of private households, \ie excluding settings such as care homes and student halls.
@@
 The recruitment pattern was chosen to hit targets regarding the fortnightly number of swabs.
-Because the testing becomes less frequent after the first 4 weeks, some continuous recruitment is needed to maintain the number of swabs, as well as to replace participants who drop out.
+Because testing becomes less frequent after the first 4 weeks, some continuous recruitment is needed to maintain the number of swabs, as well as to replace participants who drop out.
 
@@
 A notable feature of this definition is that it explicitly does not correct for false positive or negative results.
-However, for many purposes, including this thesis, the difference between the positivity rate and prevalence is negligible
+However, for many purposes, including this thesis, the difference between the positivity rate and prevalence is negligible.
 
@@
 In estimating positivity, I will divide the population into strata based on age group, sex, ethnicity, and geographic area.
-I will report results aggregated to geographic regions and age groups, as has been done previously.
-For the purposes of both fitting and reporting results, test results are aggregated to daily counts for each stratum.
+I will report results aggregated to geographic regions and age groups, as done previously.
+For the purposes of both fitting and reporting results, test results are aggregated into daily counts for each stratum.
 
@@
 To address the low number of positives, MRP uses a hierarchical Bayesian model to borrow information across time and other covariates.
-The second step, poststratification, is using the posterior distribution of the parameters is used to predict the population prevalence in each stratum.
+The second step, poststratification, uses the posterior distribution of the parameters predict the population prevalence in each stratum.
 These are combined by taking a weighted average across the strata, where a stratum's weight is equal to its (known) population size.
-The poststratification step is used to correct for non-response bias.
+The poststratification step corrects for non-response bias.
 
@@
       % $\beta_n$ and $\beta_f$ have the default INLA prior for fixed effects\todo{check what this is}.
-    \item $u_a$ and $v_e$ are independent Gaussian random effects for age and ethnic group respectively.
-Their standard deviation have a penalised complexity prior~\autocite{simpsonPenalising} with a 10\% probability of being greater than 1.
+    \item $u_a$ and $v_e$ are independent normal random effects for age and ethnic group respectively.
+Their standard deviation has a penalized complexity prior~\autocite{simpsonPenalising} with a 10\% probability of being greater than 1.
     \item $w_t$ is a second-order random walk, that is $w_t = 2w_{t-1} - w_{t-2} + \epsilon_t$ where $\epsilon_t \dist \N(0, \sigma_w^2)$.
@@
 Overdispersion allows for household clustering and other overdispersion in the data, which is common in epidemiological data~\autocite{griffithsBBD}.
-The parameterisation of the beta-binomial that I use (detailed in \cref{E-distributions}) has an overdispersion parameter $\rho$, regarded as a nuisance parameter in this context.
+The parameterization of the beta-binomial that I use (detailed in \cref{E-distributions}) has an overdispersion parameter $\rho$, regarded as a nuisance parameter in this context.
 $\rho=0$ means the beta-binomial coincides with the binomial distribution.
@@
 
-The estimates from this model compared to the unadjusted positivity rate are shown in \cref{biology-data:fig:CIS-positivity}.
+\Cref{biology-data:fig:CIS-positivity} shows estimates from this model compared to the unadjusted positivity rate.
 Both show similar trends.
@@
 An infection episode is a series of positive tests that originate from a single transmission event.
-I use a heuristic-based algorithm developed throughout the pandemic, and published within \textcite{weiRisk}; see that publication for full justification of the criteria.
+I use a heuristic-based algorithm developed throughout the pandemic and published within \textcite{weiRisk}; see that publication for full justification of the criteria.
 
@@
 
-A negative between two positive tests, when those positives are classed as part of the same infection episode, is known as an \emph{intermittent negative}.
+A negative between two positive tests, when those positives are part of the same infection episode, is known as an \emph{intermittent negative}.
 Intermittent negatives are the clearest example of a false negative.
@@
 A sample of 500 randomly selected infection episodes (selected from the episodes used in the analyses in this thesis) have their test results shown in \cref{biology-data:fig:episodes}.
-As can be seen, the number of tests within an episode, their length, and other characteristics vary greatly.
+The number of tests within an episode, their length, and other characteristics vary greatly.
 \begin{figure}
@@
 Statistical disclosure control ensures that no individual can be identified from the data.
-This means that data must be aggregated to counts of at least 5, and no individual-data included.
+This means that data must be aggregated to counts of at least 5, and no individual-level data included.
 
@@
 The next chapter will introduce the data generating processes behind infection, prevalence, and our observations of these processes.
-Then, I will apply these concept to the two studies.
+Then, I will apply these concepts to the two studies.
 

### 2024-03-19 cis-perfect-testing.tex

@@
 The CIS tests each individual infrequently, either weekly or monthly (see \cref{E-intro:sec:cis}).
-The infrequent testing complicates analysis of the data in two ways.
+Infrequent testing complicates analysis of the data in two ways.
 First, infection episodes can go undetected.
@@
 
-I base my analysis on the 4,800 episodes that were first detected between 10th Oct 2020 and 6th Dec 2020 inclusive with negatives bounding the start and end time of the episode.
+I base my analysis on the 4,800 episodes first detected between 10th Oct 2020 and 6th Dec 2020 inclusive with negatives bounding the start and end time of the episode.
 That is, the individual tested negative before and after the episode.
@@
 The framework is general, allowing for arbitrary censoring patterns and double interval censoring.
-They apply it to a situation the visit times are global (all nests could be detected on the same days), simplifying the likelihood.
+They apply it to a situation where the visit times (possible detection times) are global, simplifying the likelihood.
 I adapt this framework for CIS, when each individual has their own visit schedule (see section \cref{perf-test:sec:model}).
@@
 This increases the dimensionality of the problem greatly, making it computationally infeasible within the SRS.
-I develop a novel Bayesian framework without explicit data augmentation.
+I developed a novel Bayesian framework without explicit data augmentation.
 This method scales to the large number of episodes present in CIS.
@@
 This approach extends the framework of \textcite{heiseyModelling} to include individuals without detected episodes.
-This framework is itself based long-standing methodology~\autocite{dempsterMaximum,turnbullEmpirical}.
+This framework is itself based on long-standing methodology~\autocite{dempsterMaximum,turnbullEmpirical}.
 I augment the data set with the full number of episodes, including those that are undetected.
@@
 
-All the relevant information about episode $i$ can be fully characterised by the (unobserved) triplet $(b_j, e_j, \sched_j)$ where $b_j$ and $e_j$ are the times the episode being and end respectively and $\sched_j$ is the set of test times for the individual in which the episode occurs.
+All the relevant information about episode $i$ can be fully characterized by the (unobserved) triplet $(b_j, e_j, \sched_j)$ where $b_j$ and $e_j$ are the times the episode being and end respectively and $\sched_j$ is the set of test times for the individual in which the episode occurs.
 The episode's duration is $d_j = e_j - b_j + 1$.
@@
 Finally, $\exists t \in \sched_i$ such that $t > e_{j(i)}$, to upper bound the end of the episode; this condition will be met for any episode sufficiently far in the past in individuals not lost to follow-up (almost always the case in this context).
-For a new context where recent infections need to be considered, this condition can be relaxed by considering these episodes as right-censored, at little cost except a more complicated likelihood expression.
+For a new context, including recent infections, this condition can be relaxed by considering these episodes as right censored, at little cost except a more complicated likelihood expression.
 
@@
   detected.
-  This is all remaining episodes not in the previous sets.
+  These are all remaining episodes not in the previous sets.
   $n_{i\bar{a}} = 0$ of these occurred in any individuals with test schedule $\sched_i$.
@@
 negative tests at times 7, 14, and 56, and positive tests at times 21
-and 28. Inadmissible region is unshaded. The black (impossible) region is because it does not make sense for recovery to occur prior to infection. \label{perf-test:fig:partitionSpace}}
+and 28. Inadmissible regions are unshaded. The black (impossible) region is because it does not make sense for recovery to occur prior to infection. \label{perf-test:fig:partitionSpace}}
 \end{figure}
@@
 The above probabilities ($p_{iu}$, $p_{ia}$, and $p_{i\bar{a}}$) are conditional on the episode occurring with test schedule $\sched_i$.
-These probabilities are multiplied by $\eta_i/\Ncis$, where $\eta_i$ is the number of individuals with test schedule $\sched_i$, to give the probability that an episode belongs to each class, without conditioning on a specific individual.
+I multiply these probabilities by $\eta_i/\Ncis$, where $\eta_i$ is the number of individuals with test schedule $\sched_i$, to give the probability that an episode belongs to each class, without conditioning on a specific individual.
 
@@
 
-In forming the probability $p_{ia}$, the data is used.
-This is not quite correct because the cell probabilities in the likelihood should not depend on the data.
-However, an alternative where each combination of infection and recovery times is considered its own class is possible.
-Then, the counts for all the classes that previously contained undetected episodes would be zero.
-The resulting likelihood would be the same.
+In forming the probability $p_{ia}$, I use the test outcomes.
+This is not strictly correct because the cell probabilities in the likelihood should not depend on the data.
+However, an alternative where each combination of infection and recovery times is considered its own cell is possible.
+The new cells would align with the inadmissible region.
+Since each of these cells have an observed count of 0, they can be combined to give the same likelihood.
+That is, the approach is mathematically equivalent but greatly complicates the formulae and notation.
 Therefore, I present this version as it is clearer.
@@
 Specifically, they tend to favour shorter survival times (see \cref{perf-test:fig:flat-prior}).
-An intuitive explanation of this can be drawn from the fact that priors of this form have an expected value of 0.5.
+An intuitive explanation of this follows from the fact that priors of this form have an expected value of 0.5.
 The expected survival time, $\E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right)$, is therefore equal to $0.5^{t-1}$, a quantity that declines rapidly.
@@
 The central 95\% probability mass of Beta(0.1, 1.9) is 0.00--0.47.
-The central estimate, of 0.05, is in line with previous estimates that the median duration is in the range 15--20 days~\autocite{cevikShedding}.
+The central estimate, of 0.05, is in line with previous estimates that the median duration i in the range 15--20 days~\autocite{cevikShedding}.
 This prior gives a very vague prior predictive distribution on the survival time (see \cref{perf-test:fig:vague-prior}).
@@
 Smoothing priors encode the biological consideration that the hazard should not vary significantly from day-to-day.
-It is the Bayesian equivalent of a penalised likelihood, used frequently in the non-parametric frequentist setting~\autocite[e.g.][]{bacchettiNonparametric}.
+It is the Bayesian equivalent of a penalized likelihood, used frequently in the non-parametric frequentist setting~\autocite[e.g.][]{bacchettiNonparametric}.
 Many possible forms of smoothing priors exist, including splines, Gaussian Processes, and random walks.
@@
 RW2 prior encodes that the hazard should be linearly changing with some random changes at each time step.
-Specifically, the difference on the logit scale between $\lambda_t$ to $\lambda_{t+1}$ is the difference between $\lambda_{t-1}$ and $\lambda_t$ plus some Normally-distributed noise.
+Specifically, the difference on the logit scale between $\lambda_t$ to $\lambda_{t+1}$ is the difference between $\lambda_{t-1}$ and $\lambda_t$ plus some normally distributed noise.
 Formally:
@@
   \centering \includegraphics{cis-perfect-testing/rw2-prior}
-  \caption[RW2 prior for the hazard]{The of prior predictive distribution (mean and 95\% central interval) for the RW2 prior specified in the main text.
+  \caption[RW2 prior for the hazard]{The prior predictive distribution (mean and 95\% central interval) for the RW2 prior specified in the main text.
 The posterior mean estimate using ATACCC data in \cref{E-ATACCC} are shown for comparison.
@@
 That is, the prior used in this analysis cannot be independent across the hazards.
-Secondly, the uncertainty in the estimates from \cref{E-ATACCC} should be increased for this analysis.
+Secondly, the uncertainty in the estimates from \cref{E-ATACCC} is underestimated.
 The additional uncertainty is because the \cref{E-ATACCC} analysis extrapolates beyond the data using strong model assumptions (see \cref{E-ATACCC:sec:discussion}).
@@
 This can be viewed as an approximation of Markov melding~\autocite{goudieJoining}.
-Using a multivariate normal, as opposed to multiple univariate distribution, ensures that the correlation between the hazards is preserved.
+Using a multivariate normal, as opposed to multiple univariate distribution preserves the correlation between the hazards.
 The logit transformation maps from the $[0, 1]$ interval to the full real line.
@@
 where $k_t$, $\alpha_0$, and $\beta_0$ are hyperparameters; and $h_t$ is a point estimate of the hazard at time $t$ from ATACCC.
-An intuition for what this distribution represents can be gained from a conjugate model for $\lambda_t$ with a beta prior and a binomial likelihood.
+An intuition for what this distribution represents derives from a conjugate model for $\lambda_t$ with a beta prior and a binomial likelihood.
 If $\lambda_t$ is given the prior distribution $\text{Beta}(\alpha_0, \beta_0)$, and we then have $k_t$ observations with $k_t h_t$ successes, then the posterior distribution for $\lambda_t$ is $\text{Beta}(\alpha_t, \beta_t)$ (as defined above).
@@
 
-I simulate a dataset of detected episodes that has the same characteristics as that in the CIS by the following procedure.
+I simulate a dataset of detected episodes with the same characteristics as that in the CIS by the following procedure.
 \begin{enumerate}
@@
     \item Of these remaining episodes, sample 4,800 to match the sample size of the true dataset.
-This is needed because in step 2 the entire cohort was infected, while in the real study only a (unknown) portion is infected.
+This is needed because in step 2 the entire cohort was infected, while in the real study only an (unknown) portion is infected.
     \item For this final set of episodes, calculate $(l_j^{(b)}, r_j^{b}, l_j^{(e)}, r_j^{(e)})$ by taking the day after the last negative prior to any positives, the first positive, the last positive, and the day before the negative following the last positive respectively.
@@
 
-To generate the durations, an assumption is required for the distribution of the duration of positivity.
-I base this assumption on the estimate from \cref{E-ATACCC} with an inflated tail to represent what is seen within CIS.
-The tail is modified based on an unpublished analysis of CIS data~\citePersonalComms{A.\ Sarah Walker}.
+Step 3 requires a distribution of the duration of positivity.
+I base this assumption on the estimate from \cref{E-ATACCC} with an inflated tail to be consistent with the CIS.
+I modify the tail based on an unpublished analysis of CIS data~\citePersonalComms{A.\ Sarah Walker}.
 Walker uses survival analysis to estimate the duration from CIS data.
-The initiating event is assumed known as the time the episode was detected.
-The final event is assumed interval-censored between the time of the final positive test and the subsequent negative test, or right-censored if a negative test has not yet been observed.
-A flexible, spline-based form is assumed for the baseline survival function~\autocite{roystonSTPM,roystonFlexible} with covariates introduced via proportional odds.
+This analysis assumes the initiating event is known, and equal to the episode’s detection time. 
+It assumes the final event is interval censored between the time of the final positive test and the subsequent negative test, or right censored if a negative test has not yet been observed.
+A flexible, spline-based form is used for the baseline survival function~\autocite{roystonSTPM,roystonFlexible} with covariates introduced via proportional odds.
 By not accounting for either the undetected infections or the interval censoring of the initiating event, this analysis has competing biases which makes them hard to interpret~\autocite{cisMethodsONS}.
@@
 To form the duration distribution used in the simulation, we combine the two estimates.
-The first 30 days of the distribution is proportional to the ATACCC estimate, with the rest proportional to this CIS-based estimates.
+The pdf over the first 30 days is proportional to the ATACCC estimate, with the rest proportional to this CIS-based estimates.
 % The CIS-based estimates are shifted 3 days in order to make this smooth; this counteracts the above bias of missing the start of the infection (for times above 30 days, the other bias, missing short infections, is negligible).
@@
 
-Simulation was performed in R~4.2.0~\autocite{R-4-2-0} using tidyverse~2.0.0~\autocite{tidyverse} and targets~1.1.3~\autocite{targetsPackage}.
+I implemented the simulation in R~4.2.0~\autocite{R-4-2-0} using tidyverse~2.0.0~\autocite{tidyverse} and targets~1.1.3~\autocite{targetsPackage}.
 
-Inference was implemented in Stan via RStan~2.21.8~\autocite{rstan2-21-8} using default settings.
+I implemented the inference in Stan via RStan~2.21.8~\autocite{rstan2-21-8} using default settings.
 Convergence was assessed using Rhat and ESS as described in \cref{E-inc-prev:sec:MCMC}, and all runs checked for divergent transitions.
@@
 
-Repeating the simulation showed negligible variation in the results (not shown).
-This is caused by the large sample size (4800 detected infections).
-Therefore, only results from a single simulation are shown and discussed.
+Repeating the simulation showed negligible variation in the results (not shown) due to the large sample size (4800 detected infections).
+Therefore, I show and discuss results from only a single simulation.
 
@@
 
-The informative and vague independent priors performed similarly.
-The informative prior had tighter credible intervals, as expected.
-This prior was centred on the correct value for the initial part of the distribution.
-Therefore, the narrow credible intervals are unsurprising.
+The informative and vague independent priors performed similarly, with the informative prior having tighter credible intervals.
+% The informative prior was centred on the correct value for the initial part of the distribution.
+% Therefore, the narrow credible intervals are unsurprising.
 For hazards after day 40, the informative and vague priors are the same.
@@
 y-axis has either natural (left) or log (right) scale.
-Only showing first 100 days of the distribution.
+I show only the first 100 days of the distribution for readability.
 \label{perf-test:fig:survival-results}}
@@
 Reducing the amount of smoothing or type of smoothing may help.
-More flexible smoothers, such as a Gaussian Process for additional flexibility and smoothing and different time-scales~\autocite{saulGaussian}.
+More flexible smoothers, such as a Gaussian Process for additional flexibility and smoothing and different timescales~\autocite{saulGaussian}.
 
 Various extensions could be worth exploring.
-Most importantly, covariates (\eg age, immunity from vaccination and/or previous infection, or variant) could be included.
+Most importantly, including covariates (\eg age, immunity from vaccination and/or previous infection, or variant).
 Immunity in particular affects the duration substantially, although age and variant may also make a contribution~\autocite{hakkiOnset,russellWithinhost}.
 Covariates can be included with a variety of models, the simplest being a proportional hazards model.
-A major challenge here is that the calculation of $p_t$ is greatly complicated, because it is no longer linear in the survival function.
+A major challenge here is that the calculation of $p_t$ is greatly complicated because it is no longer linear in the survival function.
 
@@
 These could be more cost-effective.
-Improved cost-effectiveness could allow more rapid response to potential pandemics because the threshold for policy-makers to approve the study would be lower.
-A long-term, preparatory effort to develop a more efficient design in preparation of a future pandemic would be worthwhile.
+Improved cost-effectiveness could allow more rapid response to potential pandemics because the threshold for policymakers to approve the study would be lower.
+A long-term, preparatory effort to develop a more efficient design in preparation for a future pandemic would be worthwhile.
 

### 2024-03-19 conclusion.tex

@@
 One model was phenomenological in nature and the other mechanistic.
-The phenomenological approach, based on the backcalculation framework, provide estimates of incidence and makes fewer assumptions about the infection process; however, the estimates are very uncertain.
-The mechanistic approach, based on a SIR-type compartmental model, reduces the uncertainty and allows estimation of the components of transmission by making stronger assumptions about the infection process.
+The phenomenological approach, based on the backcalculation framework, provides estimates of incidence and makes fewer assumptions about the infection process; however, the estimates are very uncertain.
+The mechanistic approach, based on an SIR-type compartmental model, reduces the uncertainty and allows estimation of the components of transmission by making stronger assumptions about the infection process.
 These results highlighted the heterogeneity across regions and age groups in England.
@@
 These models both required knowledge of the distribution of the duration of positivity.
-Estimating this distribution was the focus of the majority of this thesis.
+Estimating this distribution was the focus of most of this thesis.
 In \cref{E-perf-test,E-imperf-test}, I showed how this quantity could be estimated the CIS prevalence survey.
@@
 Overall, I have found that representative prevalence surveys are a promising approach for pandemic surveillance.
-A well-designed survey is sufficient to estimate all of the quantities of interest (as described in \cref{E-intro:sec:metrics}), except severity.
+A well-designed survey is sufficient to estimate all the quantities of interest (as described in \cref{E-intro:sec:metrics}), except severity.
 
@@
 If, rather than assuming uniformity, the time of incidence was estimated, this would be a form of incidence estimation.
-The issue of including covariates to control for non-response bias, encountered in \cref{E-transmission}, as well as the parameterisation of the shape of the incidence curve would be major challenges in this extension.
+The issue of including covariates to control for non-response bias, encountered in \cref{E-transmission}, as well as the parameterization of the shape of the incidence curve would be major challenges in this extension.
 The basis of such a model could be existing modelling frameworks~\autocite[e.g.][]{taffeJoint,haySerosolver}.
@@
 A representative survey of the population could have provided this information.
-However, challenges in constructing a survey with the power to detect low levels of transmission, as well as the time required to set-up such a survey, would need to be addressed.
+However, challenges in constructing a survey with the power to detect low levels of transmission, as well as the time required to set up such a survey, would need to be addressed.
 
 A final consideration is the estimation of the severity of a pathogen.
-Severity of a pathogen is crucial for risk assessment, and can change over time~\autocite[e.g.][]{kirwanSeverityTrends}.
+Severity of a pathogen is crucial for risk assessment and can change over time~\autocite[e.g.][]{kirwanSeverityTrends}.
 Two approaches could be taken to using a prevalence surveys for this purpose.

### 2024-03-19 introduction-comments.tex
remove appendix from chapter-by-chapter compile

@@
 They estimate a posterior distribution of the parameters of a spline by fitting to the prevalence results via a binomial distribution.
-Their setup penalises the spline deviating from a constant growth rate.
+Their setup penalizes the spline deviating from a constant growth rate.
 The growth rate is then calculated from the derivative of the spline, an analytical transformation of the parameters.
@@
 This approach generally requires fewer parameters, and no random effect(s), reducing the computational burden.
-It is also easier to specify non-parameteric forms of the distribution, which is important for SARS-CoV-2 as the tail behaviour may be important.
+It is also easier to specify non-parametric forms of the distribution, which is important for SARS-CoV-2 as the tail behaviour may be important.
 A more detailed review of the literature is in the relevant chapters.

### 2024-03-19 introduction.tex

--- 2024-03-19 introduction.tex
@@
 \emph{Transmission} refers to the process by which an infectious disease generates new infections.
-Estimating transmission refers to inferring the parameters of this process, including its components.
+Estimating transmission refers to inferring the parameters of this process.
 The components of transmission (\eg the number of contacts each individual makes per day) allow interventions to be targeted.
@@
 Incidence-type and prevalence-type data can directly inform estimation of incidence and transmission; however, this estimation almost always relies on prior knowledge of epidemiological and/or biological quantities.
+For example, the time between infection and symptom onset.
 These quantities can be estimated from other studies, \eg cohort studies.
@@
 \end{align}
-where $Z(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the instantaneous probability of an individual infected $\tau$ time units ago infects an individual at time $t$.
+where $Z(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the mean rate an individual infected $\tau$ time units ago infects individuals at time $t$.
 Therefore, the renewal equation relates incidence in the past to incidence in the future.
+This partially describes the transmission process, hence the name semi-mechanistic.
 The effective reproduction number can be expressed as a function of $\beta(t, \tau)$.
@@
 
-The \emph{disease model} links the transmission model to a quantity, $W_t$, observable albeit possibly in a noisy and/or biased way (see next section).
-Examples are the incidence or prevalence of symptomatic individuals and admissions to hospital.
+\emph{Disease models} link the transmission model to a latent quantity, $W_t$.
+$W_t$ will then be recorded through some system (\eg a testing programme) described by the observation model (see next section).
+For example, $W_t$ could be the incidence or prevalence of symptomatic individuals on day $t$.
 
-The choice of observation model can, in theory, be independent of the transmission model.
-However, pragmatic considerations often mean choosing the observation model to be in a similar form to the transmission model.
-
-Typically, the core of an observation model is a \emph{convolution equation} or an approximation of it.
+At the core of many observation models is a \emph{convolution equation} or an approximation of it.
 This equation describes the expectation of $W_t$, conditional on the incidence prior to time $t$.

### 2024-03-19 transmission.tex

--- 2024-03-19 transmission.tex
@@
 
-First, in \cref{backcalc}, I use a back-calculation approach, based on \cref{E-inc-prev:eq:EPt}.
+First, in \cref{backcalc}, I use a backcalculation approach, based on \cref{E-inc-prev:eq:EPt}.
 This approach enables estimation of incidence within (sub)populations of England.
 There are several attractive features about this approach in the context here.
-The backcalculation approach makes no assumptions about the transmission process; the assumptions have the potential to introduce bias.
+The backcalculation approach makes no assumptions about how transmission occurs; the assumptions have the potential to introduce bias.
 It can also easily incorporate arbitrary covariates to adjust for non-response bias.
@@
 \begin{align}
-    \vec{P} \sim \vec{\phi} &\dist g_P(\vec{\phi}) \\
-    Y_{i,t} &\dist \BB(n_{i,t}, P_{i,t})
+    \vec{P} \mid \vec{\phi} &\dist g_P(\vec{\phi}) \\
+    Y_{i,t} \mid \vec{P} &\dist \BB(n_{i,t}, P_{i,t})
 \end{align}
 where $g_P$ is analogous to $g_Z$ in providing a joint prior for the prevalence in each stratum and day combination.
+Step 2 is...
 
-\subsubsection{Desirable properties}
+\subsection{Desirable properties}
 
@@
 \begin{enumerate}
-    \item The smoothing model over time should penalize changes from continuous exponential growth.
-        penalizing changes from continuous exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time.
+    \item The smoothing model, $g_Z$ or $g_P$, over time should penalise changes from continuous exponential growth of $\pi$ over time.
+        Penalizing changes from continuous exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time.
         Furthermore, such penalization has previously been shown to improve estimates of SARS-CoV-2 prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
+Used by CIS (section 3.?).
     \item Limit estimates to epidemiologically and biologically plausible values.
@@
 
-Properties 1 and 2 suggest a model of the form $\logit(\pi_{i,t}) = g(i, t)$, where $g$ is a function with changes in the first derivative penalized (\ie changes away from linear are penalized).
+Properties 1 and 2 suggest a model of the form $\logit(\pi_{i,t}) = g(i, t)$, where $g$ is a function with changes in the first derivative penalized w.r.t.\ $t$ (\ie changes away from linear are penalised).
 The $\logit$ link function both limits $\pi_{i,t}$ to the interval $[0, 1]$, and logit-linear growth approximates exponential growth for small $\pi_{i,t}$~\autocite{ealesAppropriately}.

### 2024-03-19 conclusion.tex
Overleaf comments and rewrite 7.1

@@
 The phenomenological approach, based on the backcalculation framework, provides estimates of incidence and makes fewer assumptions about the infection process; however, the estimates are very uncertain.
-The mechanistic approach, based on an SIR-type compartmental model, reduces the uncertainty and allows estimation of the components of transmission by making stronger assumptions about the infection process.
-These results highlighted the heterogeneity across regions and age groups in England.
+The mechanistic approach, based on an SEIR-type compartmental model, reduces the uncertainty and allows estimation of the components of transmission by making stronger assumptions about the infection process.
+Both sets of results highlighted the heterogeneity across regions and age groups in England.
+The choice of modelling approach should depend on the situation.
+The phenomenological approach is more robust when there are high levels of uncertainty; for example, early in an outbreak or in the presence of new variants the infectious period duration may not be known.
+However, the mechanistic approach provides more information.
+Where possible, both methods should be implemented; discrepancies between results may highlight deficiencies in modelling assumptions.
 
@@
 
-The feasibility of jointly modelling duration and incidence was not investigated in this thesis.
+A natural extension of the work in this thesis would be a joint model of duration and incidence.
+Such a model poses both identifiability and computational challenges.
 The model for duration in \cref{E-perf-test,E-imperf-test} allowed estimation of the total number of infections in the cohort but assumed these occurred uniformly over the duration of the survey.
-If, rather than assuming uniformity, the time of incidence was estimated, this would be a form of incidence estimation.
+If, rather than assuming uniformity, a distribution for the infection episode start times was estimated, this would be a form of incidence estimation.
+Previous studies~\autocite[e.g.][]{bacchettiNonparametric} have suggested that a joint model would be unidentifiable; however, CIS data contains information on the number of individuals that were never infected which may be sufficient to resolve this issue.
 The issue of including covariates to control for non-response bias, encountered in \cref{E-transmission}, as well as the parameterization of the shape of the incidence curve would be major challenges in this extension.

### 2024-03-19 transmission.tex
Overleaf comments and rewrite 7.1

@@
 \begin{align}
-    \vec{Z} \mid \vec{\phi} &\dist g_Z(\vec{\phi}) \\
-    \vec{P_{i}} &= \matr{S} \vec{Z}_i \\
-    Y_{i,t} \mid \vec{P} &\dist \BB(n_{i,t}, P_{i,t})
+    \vec{Z} &\dist F_Z \\
+    P_{i,t}  &= \sum_{k=0}^{d_\text{max}-1} Z_{i,t-k} S(k+1) \\
+    \rho &\dist F_\rho \\
+    Y_{i,t} \mid \vec{P}, \rho &\dist \BB(n_{i,t}, P_{i,t}, \rho)
 \end{align}
-where $\vec{Z}$ is a vector of incidence proportion for each strata and day defined by some model $g$ with parameters $\vec{\phi}$; $P_{i,t}$ is the prevalence in stratum $i$ on day $t$; $\matr{S}$ is the survival matrix defined in \cref{E-inc-prev:eq:Smatrix}; $Y_{i,t}$ is the number of positive tests in stratum $i$ on day $t$; and $n_{i,t}$ is the number of tests in stratum $i$ on day $t$.
-The model $g_Z$ would be specified to share information across time and/or strata.
-This both helps stabilise the estimation (see \cref{intro:sec:disease-model}) and allows borrowing of information across strata.
-Borrowing information is required because some strata and time combinations have few positives (fewer than five per day).
-The relationship between $P$ and $Z$ is from \cref{E-inc-prev:eq:EPt-matrix}, although dividing both sides by $\Npop$ to have the values in terms of incidence proportion and prevalence rather than incidence and number of prevalent individuals.
+where $\vec{Z}$ is a vector such that $Z_{i,t}$ is the incidence proportion in stratum $i$ on day $t$, and has prior $F_Z$; $P_{i,t}$ is the prevalence in stratum $i$ on day $t$; $\rho$ is the overdispersion parameter of the beta-binomial distribution with prior $F_\rho$; $Y_{i,t}$ is the number of positive tests in stratum $i$ on day $t$; and $n_{i,t}$ is the number of tests in stratum $i$ on day $t$.
+Fitting this model to a prevalence survey such as the CIS (\eg using MCMC) gives a posterior distribution of $\vec{Z}$, the target of inference.
+
+The relationship between $P$ and $Z$ is from \cref{E-inc-prev:eq:EPt}, although dividing both sides by $\Npop$ to have the values in terms of incidence proportion and prevalence rather than incidence and number of prevalent individuals.
+
+Specifying $F_Z$ would depend on the application of the approach, but should consider the need to provide structure to stabilise the problem (see \cref{E-intro:sec:disease-model}).
+In the context of the CIS, it needs to allow borrowing information across time and/or strata; this is required because some stratum and time combinations have few positives (fewer than five per day).
 
@@
 \begin{align}
-    \vec{P} \mid \vec{\phi} &\dist g_P(\vec{\phi}) \\
-    Y_{i,t} \mid \vec{P} &\dist \BB(n_{i,t}, P_{i,t})
+    \vec{P}  &\dist F_P \\
+    \rho &\dist F_\rho \\
+    Y_{i,t} \mid \vec{P}, \rho &\dist \BB(n_{i,t}, P_{i,t}, \rho)
 \end{align}
-where $g_P$ is analogous to $g_Z$ in providing a joint prior for the prevalence in each stratum and day combination.
-Step 2 is...
+where $F_P$ is analogous to $F_Z$ in providing a joint prior for the prevalence in each stratum and day combination, and all other notation is as in the incidence-inferred approach.
+The first step is to fit this model, giving a posterior distribution for $\vec{P}$.
+
+The second step is to transform the posterior distribution for $\vec{P}$ to a posterior distribution for $\vec{Z}$.
+A posterior sample for $\vec{P}$ can be transformed to a posterior sample for $\vec{Z}$ based on \cref{E-inc-prev:eq:forward-substitute}, see \cref{backcalc:sec:methods} for details.
 
@@
 However, it poses several pragmatic challenges which means I favour the prevalence-inferred approach.
-Here I explain that further developments, beyond the scope of this thesis, would be required to use the incidence-inferred approach with a suitable $g_Z$ within the computational constraints of the SRS.
+Here I explain that further developments, beyond the scope of this thesis, would be required to use the incidence-inferred approach within the computational constraints of the SRS.
 
-Denote the quantity to be inferred at time $t$ in strata $i$ as $\pi_{i,t}$.
-That is, under the incidence-inferred approach, $\pi_{i,t}$ is the incidence proportion, and under the prevalence-inferred approach, $\pi_{i,t}$ is the prevalence.
+It will be convenient to define a variable equating to the posterior initially estimated.
+Let $\pi_{i,t} = Z_{i,t}$ under the inference-inferred approach and $\pi_{i,t} = P_{i,t}$ is the prevalence.
 
@@
 \begin{enumerate}
-    \item The smoothing model, $g_Z$ or $g_P$, over time should penalise changes from continuous exponential growth of $\pi$ over time.
-        Penalizing changes from continuous exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time.
-        Furthermore, such penalization has previously been shown to improve estimates of SARS-CoV-2 prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
-Used by CIS (section 3.?).
+    \item The prior should favour $\pi_{i,t}$ changing exponentially in $t$, \ie penalize changes from exponential growth.
+        % Penalizing changes from continuous exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time.
+        Such penalization has previously been shown to improve estimates of SARS-CoV-2 prevalence by balancing over and under-fitting~\autocite{ealesAppropriately} and is the approach used for the CIS prevalence model (see \cref{E-biology-data:sec:MRP}).
     \item Limit estimates to epidemiologically and biologically plausible values.
@@
 
-Properties 1 and 2 suggest a model of the form $\logit(\pi_{i,t}) = g(i, t)$, where $g$ is a function with changes in the first derivative penalized w.r.t.\ $t$ (\ie changes away from linear are penalised).
+A parsimonious model satisfying properties 1 and 2 is to specify $\logit(\pi_{i,t}) = g_\varphi(i, t)$, where $g$ is a function parameterised by $\varphi$ and the prior on $\varphi$ is specified to favour $d^2g/dt^2 = 0$, \ie favour linear growth in $\logit(\pi_{i,t}$.
+The prior $F_Z$ or $F_P$ is now a prior on $\varphi$.
 The $\logit$ link function both limits $\pi_{i,t}$ to the interval $[0, 1]$, and logit-linear growth approximates exponential growth for small $\pi_{i,t}$~\autocite{ealesAppropriately}.
-Combined with property 4, the form of $g$ the null model is $g(i, t) = a_0 + a_t t + f(i)$ where $a_0$ and $a_t$ are model parameters and $f$ is a function to be specified; that is, changes away from this model should be penalized.
+
+The logit transformation also allows property 4 to be conveniently satisfied using an additive model for $g_\varphi$.
+An additive model on the logit scale is multiplicative on the natural scale for the small values of $\pi_{i,t}$ which will be common.
+
+This form of model includes the MRP prevalence model in \cref{E-biology-data:sec:MRP}.
+
+% Combined with property 4, the form of $g$ the null model is $g_\varphi(i, t) = \varphi_0 + \varphi_1 t + f(i)$ where $a_0$ and $a_t$ are model parameters and $f$ is a function to be specified; that is, changes away from this model should be penalized.
 
@@
 
-INLA-based inference can only be applied to LGMs (\emph{Latent Gaussian Models})~\autocite{rueINLAseminar,rueINLA,martinoINLAtutorial}; see \cref{E-transmission:sec:INLA} for further details on INLA.
+INLA-based inference can only be applied to LGMs (\emph{Latent Gaussian Models}); see \cref{E-transmission:sec:INLA} for further details on INLA and LGMs.
+The LGM framework requires that a transformation of $P_{i,t}$ is linear in $g_\varphi = \logit(\phi_{i,t})$.
+
+If $\pi_{i,t} = P_{i,t}$ then this is trivially satisfied.
+However, if $\pi_{i,t} = Z_{i,t}$, then no such linear relationship exists.
+In particular:
+\begin{align}
+    P_{i,t}
+    &= \sum_{k=0}^{\dmax} \pi_{i,t-k} S(k+1) \\
+    &= \sum_{k=0}^{\dmax} \expit(g_\varphi(i, t-k)) S(k+1).
+\end{align}
+
+Therefore, satisfying all the desirable properties requires the prevalence-inferred approach, which I adopt.
+
+\todo[inline]{Here to the end of the section moved to Appendix.}
 
@@
 
-Therefore, using the INLA approximation while satisfying the other desirable properties requires that $g$ predicts prevalence, not incidence proportion.
-Hence, I adopt the prevalence-inferred approach.
-
 \subsection{Methods} \label{backcalc:sec:methods}
@@
 Now I explain how I extended the MRP model introduced in \cref{E-biology-data:sec:MRP} to implement the prevalence-inferred approach.
+I start with the prevalence posterior from \cref{E-biology-data:sec:positivity-results}, denoting the times the estimates are for as $t = 1, \dots, T$.
+I draw 500 prevalence time series in each stratum from the approximate posterior distribution (using the procedure described in \cref{transmission:sec:INLA:posterior}).
 
-First, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution (using the procedure described in \cref{transmission:sec:INLA:posterior}).
-For each stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
-I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:observation-process} justifies, that is the estimate of the incidence is negligibly affected by approximating the distribution $\vec{P} \mid \vec{Z}$ by its mean.
-I solve this system using forward substitution, calculating the value of $Z_t$ in \cref{E-inc-prev:eq:forward-substitute} sequentially for $t = 1, \dots, T$.
-\Cref{E-inc-prev:eq:forward-substitute} is repeated here for convenience:
+Then, I transform this posterior sample using \cref{transmission:alg:calculate-Z}.
+The algorithm is based on \cref{E-inc-prev:eq:forward-substitute}, which is repeated here for convenience:
 \begin{align}
 Z_t
-&= \frac{P_t - \sum_{i=1}^{t-1} S_{ti} Z_i}{S_{tt}} \\
+% &= \frac{P_t - \sum_{i=1}^{t-1} S_{ti} Z_i}{S_{tt}} \\
 &= P_t - \sum_{i=1}^{t-1} S(t - i + 1) Z_i.
@@
 \end{align}
-% \Cref{E-inc-prev:eq:EPt}, considered across different values of $t$, can be written as $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix formed from the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
-Dividing each side by the total number of individuals in each stratum transforms the number of prevalent individuals to prevalence and incidence to incidence proportion.
+Dividing each side of \cref{transmission:eq:forward-substitute} by the total number of individuals in each stratum transforms the number of prevalent individuals to prevalence and incidence to incidence proportion.
+The algorithm uses a deterministic relationship between incidence and prevalence within a stratum, as \cref{E-inc-prev:sec:observation-process} justifies; \ie, the estimate of the incidence is negligibly affected by approximating the distribution $\vec{P} \mid \vec{Z}$ by its mean.
+
+\begin{algorithm}
+    \For{$j = 1, \dots, 500$}{
+      Set $\vec{P}^{(j)}$ to a posterior sample of $\vec{P}$ \;
+      \For{each stratum $i$}{
+        \For{$t = -\dmax, \dots, 0$}{
+            $Z^{(j)}_{i,t} = P^{(j)}_{i,1} / \sum_{k=1}^{\dmax} S(k)$
+        }
+        \For{$t = 1, \dots, T$}{
+            Set $Z^{(j)}_{i,t} = P^{(j)}_{i,t} - \sum_{k=-\dmax}^{t-1} S(t - k + 1) Z^{(j)}_{i,k}$ \;
+        }
+      }
+    }
+    \caption{%
+      Algorithm to transform 500 posterior samples posterior samples of $\vec{P}$ to posterior samples of $\vec{Z}$.
+      Assumes constant incidence for $t \leq 1$ where $t = 1$ is the first day of prevalence data.
+      Calculation of $Z^{(j)}_{i,t}$ is from \cref{transmission:eq:forward-substitute}.
+    }
+    \label{transmission:alg:calculate-Z}
+\end{algorithm}
+
+\Cref{transmission:eq:forward-substitute} is derived from \cref{E-inc-prev:eq:EPt-matrix}: $\vec{P} = \matr{S} \vec{I}$.
+For this to have a unique solution, $\vec{I}$ and $\vec{P}$ need the same dimension.
+However, I am using a period of prevalence which starts after the first infections.
+These prior infections will still contribute to the prevalence.
+I assume that the incidence for $t \leq 1$ is constant.
+Therefore, $P_1 = I_0 \sum_{i=1}^{\dmax} S(i)$ and $I_t = P_1 / \sum_{i=1}^{\dmax} S(i)$ for $t \leq 1$.
+
+% First, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution (using the procedure described in \cref{transmission:sec:INLA:posterior}).
+% For each sample and stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
+% I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:observation-process} justifies, that is the estimate of the incidence is negligibly affected by approximating the distribution $\vec{P} \mid \vec{Z}$ by its mean.
+% I solve this system using forward substitution, calculating the value of $Z_t$ in \cref{E-inc-prev:eq:forward-substitute} sequentially for $t = 1, \dots, T$.
+% $t = 1$ is defined as the first day where prevalence data is used.
+% \Cref{E-inc-prev:eq:forward-substitute} is repeated here for convenience:
+% \begin{align}
+% Z_t
+% &= \frac{P_t - \sum_{i=1}^{t-1} S_{ti} Z_i}{S_{tt}} \\
+% &= P_t - \sum_{i=1}^{t-1} S(t - i + 1) Z_i.
+% \label{transmission:eq:forward-substitute}
+% \end{align}
+% % \Cref{E-inc-prev:eq:EPt}, considered across different values of $t$, can be written as $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix formed from the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
+% Dividing each side by the total number of individuals in each stratum transforms the number of prevalent individuals to prevalence and incidence to incidence proportion.
 
@@
 There are two sets of assumptions which are most likely to be violated.
-The first set of assumptions is in constructing survival matrix, $\matr{S}$, which is assumed to be known with no uncertainty in this procedure.
+
+The first set of assumptions is in estimating the survival function, $S(t)$, which is assumed to be known with no uncertainty in this procedure.
 However, \cref{E-imperf-test} showed that the estimate of $S(t)$ is sensitive to assumptions made, such as the assumed $\psens$ and strength of priors used.
-The second set of assumptions are in specifying the statistical model used to estimate prevalence, for example the form of the smoothing used.
-In particular, the prior for $\pi_P$ places non-zero mass on prevalence values which imply negative incidence under the other model assumptions.
 
-For $\vec{P} = \matr{S} \vec{I}$ to have a unique solution, $\vec{I}$ and $\vec{P}$ need the same dimension.
-However, I am using a period of prevalence which starts after the first infections.
-These prior infections will still contribute to the prevalence.
-I assume that the incidence up to and including the first day of prevalence data, at $t = 1$, is constant at $I_0$.
-Therefore, $P_1 = I_0 \sum_{i=1}^{\dmax} S(i)$ and $I_0 = P_1 / \sum_{i=1}^{\dmax} S(i)$.
-For day $t$, the contribution of the incidence for $t < 1$ can be calculated as $P'_t = I_0 \sum_{i=t+1}^{\dmax} S(i)$.
-Subtracting $\vec{P'}$ from $\vec{P}$ before applying the forward solving procedure accounts for the contribution of the incidence prior to the first day of prevalence data.
+The second set of assumptions are in specifying the prior for $\vec{P}$.
+In particular, the prior places non-zero mass on prevalence values which imply negative incidence under the other model assumptions.
+Rectifying this assumption is challenging without switching to the inference-inferred approach, discussed further in \cref{backcalc:sec:results}.
 
-Once the incidence proportion in each stratum have been calculated using this procedure, I poststratify it to the CIS target population and subpopulations of interest.
-This is done using the methodology explained in \cref{E-biology-data:sec:MRP}.
+For each posterior sample of $\vec{Z}$, I poststratify it to the CIS target population and subpopulations of interest.
+This is done for each posterior sample using the methodology explained in \cref{E-biology-data:sec:MRP}.
+The poststratified posterior samples are now posterior samples from the posterior of interest.
 
@@
 \begin{figure}
+    \todo[inline]{Make figure backcalc-regions easier to read}
     \includegraphics{transmission/backcalc-regions}
@@
 Exploiting this property could enable efficient approximations to the posterior.
-For example, a Laplace approximation or variational inference.
-Further work to check the quality of these approximations, is required.
+For example, a Laplace approximation or variational inference~\autocite{bleiVariational}.
+These methods require further work to check the quality of the approximations in this context.
 

### 2024-03-19 biology-data.tex
Overleaf comments and rewrite 7.1
SRS computational resources

@@
 For parameters given vague priors, the large amount of data means that the prior has little influence on the posterior distribution.
+The second-order random walk priors favours linear growth in $\logit(p_{taes})$ over time.
+For the small values of $p_{taes}$ likely to be encountered, this approximates exponential growth.
 
@@
 
-For privacy reasons, the CIS data is stored in the ONS's SRS (Secure Research Service).
+For privacy reasons, the CIS data is stored in the ONS's SRS (Secure Research Service)~\autocite{onsSRS}.
 The SRS is a Trusted Research Environment (TRE) that allows researchers to access sensitive data in a secure environment.
+The SRS gives accredited researchers working on projects in the public interest access to granular data which cannot be legally shared publicly.
 
@@
 Therefore, computationally efficient approaches must be taken to this work.
+The SRS contains limited computational power.
+It has\todo{insert SRS computational power}.
+However, these are shared computational resources and due consideration must be given to other users.
+Processes which run for multiple days are liable to be cancelled.
+
+This computational limitation will feature in several places in this thesis.
 

### 2024-03-19 introduction.tex
Overleaf comments and rewrite 7.1
SRS computational resources

@@
 
-\subsection{Secure research service}
-
-The data from the CIS is stored in the SRS (Secure Research Service)~\autocite{onsSRS}.
-The SRS gives accredited researchers working on projects in the public interest access to granular data which cannot be legally shared publicly.
-Only non-disclosive data can be extracted from the SRS.
-In particular, counts normally need to be greater than 5.
-
-The SRS contains limited computational power.
-It has\todo{insert SRS computational power}.
-However, these are shared computational resources.
-Due consideration must be given to other users.
-Processes which run for multiple days are liable to be cancelled.
-
 \section{Bayesian inference} \label{intro:sec:Bayes}

### 2024-03-19 biology-data.tex
ch 2 intro

@@
 
-\todo[inline]{Rewrite chapter 2 intro after writing thesis intro because it will be clearer how to divide content.}
-
-SARS-CoV-2 is a beta coronavirus that emerged in late 2019.
-It is closely related to previous coronaviruses that have caused outbreaks in humans, SARS-CoV and MERS-CoV.
-
+SARS-CoV-2 is closely related to previous coronaviruses that have caused outbreaks in humans, SARS-CoV and MERS-CoV.
 Coronaviruses are RNA (Ribonucleic acid) viruses.
 RNA viruses use RNA, rather than DNA, as their genetic material.
+RNA viruses, which include influenza, mutate rapidly and are the most likely class of viruses to cause pandemics~\autocite{woolhouseRNA}.
+
+The initial and most sensitive diagnostic test for SARS-CoV-2 uses RT-PCR, which I explain in \cref{biology-data:sec:PCR}.
+SARS-CoV-2 causes the disease COVID-19; I discuss the \emph{natural history} of the disease, the course it takes from beginning to end~\autocite[193]{portaEpiDictionary}, in \cref{biology-data:sec:natural-history}.
+
+I then move on to discussing the two studies I use in this thesis.
+These are ATACCC (\cref{biology-data:sec:ataccc}) and the CIS (\cref{intro:sec:cis}), both briefly discussed in \cref{E-intro}.
+Unlike many studies of SARS-CoV-2, these studies are not focused on specific groups such as healthcare workers or hospital patients.
+Therefore, the results from these studies are more generalizable to the entire UK population.
 
@@
 
-\section{Studies used in this thesis} \label{biology-data:sec:studies}
+% \section{Studies used ie this thesis} \label{biology-data:sec:studies}
 
-In this section, I expand on the major two studies I use in this thesis (ATACCC and CIS).
-Unlike many studies of SARS-CoV-2, these studies are not focused on specific groups such as healthcare workers or hospital patients.
-Therefore, the results from these studies are more generalisable to the entire UK population.
+% In this section, I expand on the major two studies I use in this thesis (ATACCC and CIS).
+% Unlike many studies of SARS-CoV-2, these studies are not focused on specific groups such as healthcare workers or hospital patients.
+% Therefore, the results from these studies are more generalisable to the entire UK population.
 
-\subsection{Assessment of Transmission and Contagiousness of COVID-19 in Contacts} \label{biology-data:sec:ataccc}
+\section{Assessment of Transmission and Contagiousness of COVID-19 in Contacts} \label{biology-data:sec:ataccc}
 
@@
 
-\subsection{Coronavirus (COVID-19) Infection Survey} \label{intro:sec:cis}
+\section{Coronavirus (COVID-19) Infection Survey} \label{intro:sec:cis}
 
@@
 
-\subsubsection{Strata} \label{biology-data:sec:cis-strata}
+\subsection{Strata} \label{biology-data:sec:cis-strata}
 
@@
 
-\subsubsection{Adjustment for non-response and data sparsity} \label{biology-data:sec:MRP}
+\subsection{Adjustment for non-response and data sparsity} \label{biology-data:sec:MRP}
 
@@
 
-\subsubsection{Clustering} \label{biology-data:sec:clustering}
+\subsection{Clustering} \label{biology-data:sec:clustering}
 
@@
 
-\subsubsection{Positivity results} \label{biology-data:sec:positivity-results}
+\subsection{Positivity results} \label{biology-data:sec:positivity-results}
 
@@
 
-\subsubsection{Episodes} \label{biology-data:sec:cis-episodes}
+\subsection{Episodes} \label{biology-data:sec:cis-episodes}
 
@@
 
-\subsubsection{Computational limitations and privacy protection} \label{biology-data:sec:SRS}
+\subsection{Computational limitations and privacy protection} \label{biology-data:sec:SRS}
 

### 2024-03-20 introduction.tex
add to do for citations
intro citations

@@
 The data sources used for outbreak surveillance are often observational in nature.
-Therefore, they suffer from biases that can be hard to understand and quantify.
+Therefore, they suffer from biases that can be hard to understand and quantify~\autocite{shadboltChallenges}.
 Here I discuss three types of data which are useful for surveillance.
@@
 Incidence-type data originates from observations of the consequences of a disease, \eg the development of symptoms or admission to hospital.
-This type of data is typically affected by \emph{ascertainment bias}: the proportion of infections which are counted can vary over time or within different subgroups of the population~\autocite[chapter 9]{lashModern}.
+This type of data typically suffers from reporting biases, particularly that the proportion of infections which are counted can vary over time or within different subgroups of the population~\autocites[chapter 9]{lashModern}{shadboltChallenges}.
 
@@
 
-However, severe symptoms can be rare in some population subgroups (\eg young, otherwise healthy individuals infected with SARS-CoV-2).
-This means that the observations do not provide much information on incidence in those subgroups.Furthermore, the delay between infection and the severe event occurring means that recent infections cannot be estimated.
+However, severe symptoms can be rare in some population subgroups (\eg young, otherwise healthy individuals infected with SARS-CoV-2~\autocite{wardSero,bhopalChildren}).
+This means that the observations do not provide much information on incidence in those subgroups.
+Furthermore, the delay between infection and the severe event occurring means that recent infections cannot be estimated~\autocite{swallow2022challenges}.
 
@@
 However, \emph{seroprevalence} surveys are more feasible.
-Seroprevalence is the proportion of individuals in the population which have antibodies to the disease.
-The presence of antibodies indicates that the individual was previously infected.
-Many more individuals have ever been infected than are currently infected, therefore, seroprevalence surveys can be powered much more cheaply.
+Seroprevalence is the proportion of individuals in the population which have antibodies to the disease, indicating that the individual was previously infected~\autocite{cdcSeroprevalence}.
+Many more individuals have ever been infected than are currently infected, therefore, seroprevalence surveys can be powered much more cheaply~\autocite{wuSeroprevSimulation}.
 However, seroprevalence is a delayed measure because there is a lag between an individual being infected and them developing antibodies.
@@
 Other epidemiological and/or biological quantities are normally required to estimate transmission.
-These depend on the model type but could include the \emph{generation interval} (the time between an individual being infected and going on to infect further individuals) or duration of different disease stages.
+These depend on the model type, often including the duration of different disease stages or the distribution of the time between infection events~\autocite{wallingaGI,dankwaStructural}.
 Cohort studies (following a group of individuals over time), are a study design which can estimate many of these quantities.
 
-Well-designed cohort studies can provide direct estimates of incidence and/or prevalence in the cohort.
+Well-designed cohort studies can provide direct estimates of incidence and/or prevalence in the cohort~\autocite[chapter 7]{lashModern}.
 However, recruiting a representative sample of the population is challenging, and the results of these studies are often not generalisable to the whole population.
-Furthermore, these studies often require a large cohort to be appropriately power.
+Furthermore, these studies often require a large cohort to be appropriately powered.
 Large cohort studies are expensive to run.
@@
 The first component required to estimate incidence and transmission is a model of the infection process, a mathematical description of the process through which incidence arises.
-A spectrum of models exists, categorised by the strength of assumptions they make about the transmission process.
+A spectrum of models exists, categorised by the strength of assumptions they make about the transmission process~\autocite{beckerCOVIDmodels}.
 Stronger assumptions allow more insight into the transmission process and generally lower the variance of the estimates but can introduce bias and are often computationally challenging to fit.
@@
 These have components of the transmission process in them but do not fully represent the process.
-The most common type of semi-mechanistic model is those based on a \emph{renewal equation}.
+The most common type of semi-mechanistic model is those based on a \emph{renewal equation}~\autocite{bhattSemimechanistic}.
 The renewal equation~\autocite{fraserRenewal} for an infectious disease, in the most general form, is:
@@
     Z(t) &= \int_{0}^{t} \beta(t,\tau) Z(t-\tau) d \tau
+    \label{intro:eq:renewal}
 \end{align}
@@
 This partially describes the transmission process, hence the name semi-mechanistic.
-The effective reproduction number can be expressed as a function of $\beta(t, \tau)$.
+The effective reproduction number is $\Re = \int_0^\infty \beta(t, \tau) d\tau$.
 % A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
 % The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
+Extensions of \cref{intro:eq:renewal} to track the number of susceptible individuals become full mechanistic models, equivalent to the compartmental models introduced next~\autocite{champredonEquivalence}.
 Renewal equations do not feature further in this thesis; for details, including practical considerations when using them as the basis for inference, see \textcite{nashEstimating,gosticPractical}.
@@
 
-At the core of many observation models is a \emph{convolution equation} or an approximation of it.
+At the core of many observation models is a \emph{convolution equation} or an approximation of it~\autocite[e.g.][]{brookmeyerBackcalculation,abbottEstimating,birrellRealtimea}.
 This equation describes the expectation of $W_t$, conditional on the incidence prior to time $t$.
@@
 where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that ever appear in $W_t$; and $f(i)$ is the probability that an infection at time $t-i$ is included in count $W_t$, conditional on it ever being in one of the counts.
-$f$ is known as a \emph{delay distribution}.
+$f$ is known as a \emph{delay function}.
+$f$ is commonly the pdf of a distribution, or a transformation of it.
 
@@
 Biases can arise in a variety of ways, \eg healthcare-seeking behaviour or tests which are less than 100\% accurate.
-Bias can be modelled through additional parameters or data quantifying these issues.
+Bias can be modelled through additional parameters or data quantifying these issues~\autocite[e.g.][]{sherrattExploring,birrellRealtimea,nicholsonImproving,swallow2022challenges}.
 
 Noise could come from a variety of sources, \eg sampling error, clustering, or unexplained overdispersion.
-Noise is typically modelled by a parametric distribution \eg the negative binomial distribution allowing for overdispersion compared to a Poisson process assumption.
+Noise is typically modelled by a parametric distribution \eg the negative binomial distribution allowing for overdispersion compared to a Poisson process assumption~\autocite[e.g.][]{birrellRealtimea,frassoBayesian}.
 
-\section{Delay distributions}
+\section{Delay functions}
 
-Knowledge of a delay distribution, $f$ in \cref{intro:eq:disease-model}, is required for many surveillance procedures.
+Applying \cref{intro:eq:disease-model} requires the delay function, $f$.
 Typically, this is either the time until an event or the duration of a disease stage.
+Time series data on $W_t$ cannot normally identify $f$, instead external information needs incorporating~\autocite{swallow2022challenges}.
 
@@
 This requires both sufficient observations per individual and an assumption on the form of the distribution.
+More details are in \cref{E-ATACCC}.
 
@@
 % Other related approaches, such as a multi-state model (reviewed by \textcite{jacksonMSM}), could also be considered; multi-state models can be viewed as a generalisation of survival analysis.
+More details are in \cref{E-perf-test}.
 
@@
 In this thesis, I will investigate the potential of the prevalence surveys in greater detail.
-The disease model will be a prevalence model (defined further in \cref{inc-prev:sec:prevalence-process}) with the delay distribution being the duration of PCR positivity.
+The disease model will be a prevalence model (defined further in \cref{E-inc-prev:sec:prevalence-process}) with the delay function being the duration of PCR positivity.
 
@@
 The main data sources informing the response to the SARS-Cov-2 epidemic in England were: community testing (known as pillar 2 testing), the number of severe events (\eg hospital admissions or deaths), seroprevalence surveys (estimating the proportion of the population with antibodies to a disease, a marker of prior infection), and prevalence surveys (estimating the proportion of the population currently with the disease).
-Each of these data sources have strengths and weaknesses, see \textcite{royalSocietyRnumber} for further details.
+Each of these data sources have strengths and weaknesses, see \textcite{royalSocietyRnumber} and references therein for further details on each of these data sources.
 
@@
 Severe events counts are generally a reliable measure of the burden of the disease.
-The criteria for being a severe event are normally well-defined and generally well-recorded in the UK context\todo{ref severe events criteria}.
+The criteria for being a severe event are normally well-defined and generally well-recorded in the UK context.
 Therefore, these data are subject to fewer biases than community testing.
@@
 
-The original formulation used a Poisson process as the infection model; a disease model where all infections are eventually observed ($\zeta = 1$) and the delay distribution, $f$, is the incubation period; and an observation model where all diagnoses are observed, and hence has no influence on the data.
+The original formulation used a Poisson process as the infection model; a disease model where all infections are eventually observed ($\zeta = 1$) and the delay function, $f$, is the incubation period distribution's pdf; and an observation model where all diagnoses are observed, and hence has no influence on the data.
 

### 2024-03-20 thesis.tex
finish info on INLA
adjust preface
mvoe hack into a to do

@@
 \newcommand{\ssep}{:}
+\DeclareMathOperator*{\argmax}{arg\,max}
 
@@
 \title{Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys}
-\author{Joshua Blake}
+\author{Joshua Benjamin Blake}
 \date{\today}
@@
 
-%% Submission date [optional]:
-% \submissiondate{November, 2042}
+%% Submission date :
+\submissiondate{March 2024}
 
 %% You can redefine the submission notice [optional]:
-% \submissionnotice{A badass thesis submitted on time for the Degree of PhD}
+\submissionnotice{%
+    This thesis is submitted for the degree of \\ \emph{Doctor of Philosophy} \\
+    March 2024
+}
 
@@
 
+\preface{%
+    \todo[inline]{Write preface}
+    \begin{itemize}
+        \item Early version of this work used in ONS reports
+        \item Chapters in JASA paper
+        \item Chapter for PNAS (or similar) journal
+    \end{itemize}
+}
+
 \setcounter{tocdepth}{2}
@@
 \frontmatter{}
-
-\todo[inline]{Mention use of early version of this work in ONS reports}
 
@@
 
-\todo[inline]{Delete below lines - only here to make compiling individual chapters work}
+\todo[inline]{
+    Equation numbering hack
     \cref{perf-test:eq:piu}
@@
     \cref{inc-prev:eq:Smatrix}
+}
 \listoftodos

### 2024-03-20 incidence-prevalence.tex
minor to dos

@@
 A standard extension to the SIR model is to instead assume a gamma distribution~\autocite[94]{keelingModeling,andersonSpread}.
-\todo{check this citation is formatted correctly}
 The extension adds adding multiple infectious compartments, shown in \cref{SEIR:fig:SIR-gamma}.
@@
 \subsubsection{The SEIR model} \label{SEIR:sec:SEIR}
-
-\todo[inline]{Maybe merge SEIR intro into the general intro, not introducing SIR first}
 

### 2024-03-20 transmission.tex
More readable fig 7.1

@@
 \begin{figure}
-    \todo[inline]{Make figure backcalc-regions easier to read}
     \includegraphics{transmission/backcalc-regions}
     \caption[Incidence estimated using backcalculation by region]{%
-        Incidence proportion for all of England (A) and its component regions.
-        Regions are split into broad groupings of three regions each: Northern (B), Midlands (C), and Southern (D).
-        Yorkshire and the Humber is labelled as Yorkshire for brevity.
+        Incidence proportion for (A) all of England and (B) its component regions.
+        Yorkshire and the Humber is labelled Yorkshire for brevity.
     }

### 2024-03-20 transmission/backcalc-regions.R

@@ -1,43 +1,11 @@
 suppressMessages(library(dplyr))
 library(ggplot2)
+library(patchwork)
 library(tidybayes)
 library(tidyr)
 source(here::here("utils.R"))
 
-decide_panel = function(region) {
-    case_match(
-        region,
-        "England" ~ "England",
-        "North East" ~ "North",
-        "North West" ~ "North",
-        "Yorkshire" ~ "North",
-        "East Midlands" ~ "Midlands",
-        "West Midlands" ~ "Midlands",
-        "East of England" ~ "Midlands",
-        "London" ~ "South",
-        "South East" ~ "South",
-        "South West" ~ "South",
-    )
-}
-
 output_dir = here::here("transmission/outputs")
-label_letters = as_labeller(function(labels, base_labeller = label_value) {
-    glue::glue("({LETTERS[seq_along(labels)]}) {base_labeller(labels)}")
-})
-
-# alpha_dates = tribble(
-#     ~region, ~date,
-#     "North East", lubridate::dmy("09/12/2020"),
-#     "North West", lubridate::dmy("17/12/2020"),
-#     "Yorkshire", lubridate::dmy("08/12/2020"),
-#     "East Midlands", lubridate::dmy("24/11/2020"),
-#     "West Midlands", lubridate::dmy("27/11/2020"),
-#     "East of England", lubridate::dmy("27/11/2020"),
-#     "London", lubridate::dmy("21/11/2020"),
-#     "South East", lubridate::dmy("03/12/2020"),
-#     "South West", lubridate::dmy("21/12/2020"),
-# ) |>
-    #  mutate(panel = decide_panel(region))
 
 alpha_dates = tribble(
     ~region, ~date,
@@ -50,40 +18,44 @@ alpha_dates = tribble(
     "London", lubridate::dmy("21/11/2020"),
     "South East", lubridate::dmy("03/12/2020"),
     # "South West", lubridate::dmy("21/12/2020"),
-) |>
-    mutate(panel = decide_panel(region))
+)
 
-p_region = readRDS(file.path(output_dir, "region.rds")) |>
-    filter(daynr > 1) |>
-    mutate(
-        panel = decide_panel(region),
-    ) |>
-    ggplot(aes(date, incidence, colour = region, fill = region)) +
-    stat_lineribbon(alpha = 0.3, .width = 0.95) +
-    facet_wrap(~panel, labeller = label_letters) +
-    scale_y_continuous(labels = scales::label_percent()) +
-    labs(
-        x = "Date",
[…]

### 2024-03-20 SEIR/sim/coverage.R
make true vs posterior plot clearer

@@ -99,22 +99,21 @@ p_true_vs_posterior = posterior_intervals |>
         )
     )) |>
     ggplot(aes(value, true)) +
-    # facet_wrap(~ parameter, scales = "free", labeller = label_parsed) +
-    facet_wrap(~ parameter, scales = "free", labeller = parameter_labeller) +
+    facet_wrap(~ parameter, scales = "free", labeller = parameter_labeller, ncol = 3) +
     geom_abline(slope = 1, intercept = 0) +
     geom_smooth(method = "lm", formula = y ~ x) +
     standard_plot_theming() +
     geom_point() +
-    # geom_errorbar(aes(ymin = .lower, ymax = .upper), width = 0) +
     labs(
         x = "Posterior median",
         y = "True value"
-    )
+    ) +
+    theme(axis.text.x = element_text(angle = 45, hjust = 1))
 ggsave(
     filename = here::here("SEIR/sim/true_vs_posterior.pdf"),
     plot = p_true_vs_posterior,
     width = 20,
-    height = 15,
+    height = 22,
     units = "cm",
     dpi = 300
 )

### 2024-03-20 transmission-appendix-plots.tex
move half of plot to appendix

--- 2024-03-19 transmission.tex
@@
 \begin{document}
-\ifSubfilesClassLoaded{
-    \setcounter{chapter}{6}
-}
 
-\chapter{Estimating transmission} \label{transmission}
+\chapter{Additional plots for \cref{E-transmission}}
 
-In this chapter, I use the estimate of the duration of RT-PCR positivity from combining information from the ATACCC and CIS studies (see \cref{E-imperf-test:sec:conclusion}) to investigate the feasibility of estimating incidence and transmission of SARS-CoV-2 from only the CIS data (described in \cref{E-intro:sec:cis}).
-I will show that the CIS data is sufficient, demonstrating the value of surveys of this type for pandemic surveillance.
-
-I take two approaches, backcalculation and a mechanistic model, both previously discussed in \cref{E-inc-prev:sec:infection-process}.
-
-First, in \cref{backcalc}, I use a backcalculation approach, based on \cref{E-inc-prev:eq:EPt}.
-This approach enables estimation of incidence within (sub)populations of England.
-There are several attractive features about this approach in the context here.
-The backcalculation approach makes no assumptions about how transmission occurs; the assumptions have the potential to introduce bias.
-It can also easily incorporate arbitrary covariates to adjust for non-response bias.
-The scale of the CIS and desire to correct for non-response bias presents challenges for this methodology, especially within the limited computation of the SRS environment.
-
-In \cref{SEIR}, I use a mechanistic compartmental model, within the modelling framework introduced in \cref{E-SEIR:sec:mechanistic-models}.
-This approach enables estimating both incidence and components of transmission.
-Estimating transmission in addition to incidence answers important epidemiological questions, \eg whether changes in transmission can be explained by changes in contact rates.
-I develop a novel observation model to allow the transmission model to be fit to the CIS data.
-
-Finally, in \cref{transmission:sec:comparison}, I compare the results of the two approaches.
-I show they give similar values, providing reassurance of the robustness of the results to modelling assumptions.
-Overall, there is a bias-variance trade-off between the models.
-Incidence and transmission have not previously been estimated from only SARS-CoV-2 prevalence surveys before.
-
-\section{Backcalculation} \label{backcalc}
-
-In this section, I will apply the theory introduced in \cref{E-inc-prev} to estimate incidence.
-I view the CIS as a series of daily, cross-sectional prevalence surveys.
-This is the same approach as previously used to estimate prevalence from the CIS (see \cref{E-biology-data:sec:MRP}).
-
-An important consideration in this section is the non-response bias.
-I adopt the poststratification method introduced in \cref{E-biology-data:sec:MRP} to correct for this bias.
-Therefore, incidence must be estimated in each stratum before being poststratified to the target population and subpopulations of interest.
-The strata are defined by at least age group, region, ethnicity, and sex due to differential non-response in the survey (see \cref{E-intro:sec:cis}).
-
-\subsection{Approaches} \label{backcalc:sec:approach}
-
-In \cref{E-inc-prev:sec:observation-process}, I showed that a deterministic approximation to the relationship between true prevalence and incidence is a good approximation for estimating incidence.
-This relationship between incidence proportion and prevalence means that once one is inferred, the other is calculated via a deterministic function.
-This gives two possible approaches for estimating incidence from prevalence data.
-
-% \subsubsection{Possible approaches}
-
-The first approach, which I call the \emph{incidence-inferred} approach, is a single statistical model where incidence is inferred, with prevalence as a derived quantity.
-The basis of the statistical model, in a Bayesian framework, would be as follows.
-\begin{align}
-    \vec{Z} \mid \vec{\phi} &\dist g_Z(\vec{\phi}) \\
-    \vec{P_{i}} &= \matr{S} \vec{Z}_i \\
-    Y_{i,t} \mid \vec{P} &\dist \BB(n_{i,t}, P_{i,t})
-\end{align}
-where $\vec{Z}$ is a vector of incidence proportion for each strata and day defined by some model $g$ with parameters $\vec{\phi}$; $P_{i,t}$ is the prevalence in stratum $i$ on day $t$; $\matr{S}$ is the survival matrix defined in \cref{E-inc-prev:eq:Smatrix}; $Y_{i,t}$ is the number of positive tests in stratum $i$ on day $t$; and $n_{i,t}$ is the number of tests in stratum $i$ on day $t$.
-The model $g_Z$ would be specified to share information across time and/or strata.
-This both helps stabilise the estimation (see \cref{intro:sec:disease-model}) and allows borrowing of information across strata.
-Borrowing information is required because some strata and time combinations have few positives (fewer than five per day).
-The relationship between $P$ and $Z$ is from \cref{E-inc-prev:eq:EPt-matrix}, although dividing both sides by $\Npop$ to have the values in terms of incidence proportion and prevalence rather than incidence and number of prevalent individuals.
-
-The second approach, which I call the \emph{prevalence-inferred} approach takes a two-step approach.
-First, it uses the statistical model from \cref{E-biology-data:sec:MRP} to estimate prevalence.
-Then, it transforms the posterior distribution of prevalence to a posterior distribution of incidence using the forward substitution as in \cref{E-inc-prev:eq:forward-substitute}.
-At a high-level, the statistical model is:
-\begin{align}
-    \vec{P} \mid \vec{\phi} &\dist g_P(\vec{\phi}) \\
-    Y_{i,t} \mid \vec{P} &\dist \BB(n_{i,t}, P_{i,t})
-\end{align}
-where $g_P$ is analogous to $g_Z$ in providing a joint prior for the prevalence in each stratum and day combination.
-Step 2 is...
-
-\subsection{Desirable properties}
-
-The incidence-inferred approach is more parsimonious in being a single step procedure.
-However, it poses several pragmatic challenges which means I favour the prevalence-inferred approach.
-Here I explain that further developments, beyond the scope of this thesis, would be required to use the incidence-inferred approach with a suitable $g_Z$ within the computational constraints of the SRS.
-
-Denote the quantity to be inferred at time $t$ in strata $i$ as $\pi_{i,t}$.
-That is, under the incidence-inferred approach, $\pi_{i,t}$ is the incidence proportion, and under the prevalence-inferred approach, $\pi_{i,t}$ is the prevalence.
-
-The chosen approach should have the following properties.
-\begin{enumerate}
-    \item The smoothing model, $g_Z$ or $g_P$, over time should penalise changes from continuous exponential growth of $\pi$ over time.
-        Penalizing changes from continuous exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time.
-        Furthermore, such penalization has previously been shown to improve estimates of SARS-CoV-2 prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
-Used by CIS (section 3.?).
-    \item Limit estimates to epidemiologically and biologically plausible values.
-        That is, incidence proportion and prevalence should be in the interval $[0, 1]$.
-        % While incidence proportion can theoretically be greater than 1, this would imply that some individuals were infected multiple times in a day which is not biologically plausible.
-    \item Have sufficient parameters to incorporate all necessary covariates to correct for non-response bias.
-    \item Borrow information across the different covariates.
-        The borrowing should penalize changes in the distribution of incidence/prevalence across strata in the same geographic region being constant.
-        A constant distribution is predicted by the theoretical study of epidemics, for example the dynamics of stratified, mechanistic transmission models (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) over short periods of time with negligible behavioural and immunity changes.
-    \item Be computationally efficient.
-        The model will be fit to granular data and therefore needs to be run within the SRS, with its computational limitations (see \cref{E-biology-data:sec:SRS}).
-\end{enumerate}
-
-Properties 1 and 2 suggest a model of the form $\logit(\pi_{i,t}) = g(i, t)$, where $g$ is a function with changes in the first derivative penalized w.r.t.\ $t$ (\ie changes away from linear are penalised).
-The $\logit$ link function both limits $\pi_{i,t}$ to the interval $[0, 1]$, and logit-linear growth approximates exponential growth for small $\pi_{i,t}$~\autocite{ealesAppropriately}.
-Combined with property 4, the form of $g$ the null model is $g(i, t) = a_0 + a_t t + f(i)$ where $a_0$ and $a_t$ are model parameters and $f$ is a function to be specified; that is, changes away from this model should be penalized.
-
-% Properties 3 and 4 suggest the extension of the previously-developed MRP method, used to already estimate prevalence, which already satisfies these properties.
-% This method has previously been used successfully to estimate prevalence, corrected for non-response bias, from the CIS data~\autocite{cisMethodsONS,pouwelsCommunity,pouwelsMRPvaccination}.
-
-Properties 3 and 5 suggest the use of an approximate method for computing the posterior distribution.
-In particular, using MCMC, sufficient covariates to adjust for the non-response bias, and a period of more than 6--8 weeks leads to a model that is computationally prohibitive ~\citePersonalComms{Koen Pouwels}.
-Instead, previous MRP implementations used INLA (see \cref{E-biology-data:sec:MRP}).
-
-INLA-based inference can only be applied to LGMs (\emph{Latent Gaussian Models})~\autocite{rueINLAseminar,rueINLA,martinoINLAtutorial}; see \cref{E-transmission:sec:INLA} for further details on INLA.
-
-An LGM is defined in terms of a vector of hyperparameters, $\vec{\varphi}$; a deterministic transformation of the hyperparameters into covariance matrix, $\matr{\Sigma}(\vec{\varphi})$; a vector of latent variables, $\vec{z}$; a vector of $n$ data points, $\vec{w} = [w_1, \dots, w_n]^T$; and $n$ linear predictors $\vec{\eta} = [\eta_1, \dots, \eta_n]^T$.
-To be an LGM, $\vec{z}$ must be a latent Gaussian field, that is $\vec{z} \mid \vec{\varphi} \dist \N(\vec{0}, \matr{\Sigma}(\vec{\varphi}))$; and the likelihood must have the conditional independence property $p(\vec{w} \mid \vec{\varphi}, \vec{z}) = \prod_j p(w_j \mid \eta_j(\vec{z}), \vec{\varphi})$.
-
-In this context, each $w_j$ corresponds to the observed RT-PCR results in one stratum on one day.
-Hence, $\eta_{j}$ must determine the distribution of $w_{j}$ (up to some hyperparameters); where $j$ uniquely identifies a stratum $i$ and day $t$.
-
-In practice, this means that $\eta_j$ is a one-to-one transformation of the population prevalence in stratum $i$ on day $t$.
-If $\eta_j = \logit(\pi_{i,t})$, where $\pi$ is prevalence, then a model of $\eta_j = g(i, t)$, as before, is an LGM.
-However, if $\pi$ is instead representing incidence proportion, then no such linear relationship between $\eta_j$ and $g$ exists.
-In particular, if $\eta_j$ is the prevalence, then:
-\begin{align}
-    \eta_{i,t}
-    &= \sum_{k=-\infty}^0 \pi_{ik} S(k) \\
-    &= \sum_{k=-\infty}^0 \expit(g(i, k)) S(k).
-\end{align}
-As $g(i, k)$ is linear in the model parameters, $\vec{z}$, then $\eta_{i,t}$ is not, violating the requirements of an LGM.
-
-Therefore, using the INLA approximation while satisfying the other desirable properties requires that $g$ predicts prevalence, not incidence proportion.
-Hence, I adopt the prevalence-inferred approach.
-
-\subsection{Methods} \label{backcalc:sec:methods}
-
-Now I explain how I extended the MRP model introduced in \cref{E-biology-data:sec:MRP} to implement the prevalence-inferred approach.
-
-First, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution (using the procedure described in \cref{transmission:sec:INLA:posterior}).
-For each stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
-I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:observation-process} justifies, that is the estimate of the incidence is negligibly affected by approximating the distribution $\vec{P} \mid \vec{Z}$ by its mean.
-I solve this system using forward substitution, calculating the value of $Z_t$ in \cref{E-inc-prev:eq:forward-substitute} sequentially for $t = 1, \dots, T$.
-\Cref{E-inc-prev:eq:forward-substitute} is repeated here for convenience:
-\begin{align}
-Z_t
-&= \frac{P_t - \sum_{i=1}^{t-1} S_{ti} Z_i}{S_{tt}} \\
-&= P_t - \sum_{i=1}^{t-1} S(t - i + 1) Z_i.
-\label{transmission:eq:forward-substitute}
-\end{align}
-% \Cref{E-inc-prev:eq:EPt}, considered across different values of $t$, can be written as $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix formed from the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
-Dividing each side by the total number of individuals in each stratum transforms the number of prevalent individuals to prevalence and incidence to incidence proportion.
-
-Violation of assumptions in various parts of this procedure can lead to the second term in \cref{transmission:eq:forward-substitute} to be greater than the first, giving negative incidence.
-In this case, the incidence prior to day $t$ imply that the prevalence on day $t$ is higher than that predicted by the fitted regression model for this strata/day combination.
-There are two sets of assumptions which are most likely to be violated.
-The first set of assumptions is in constructing survival matrix, $\matr{S}$, which is assumed to be known with no uncertainty in this procedure.
-However, \cref{E-imperf-test} showed that the estimate of $S(t)$ is sensitive to assumptions made, such as the assumed $\psens$ and strength of priors used.
-The second set of assumptions are in specifying the statistical model used to estimate prevalence, for example the form of the smoothing used.
-In particular, the prior for $\pi_P$ places non-zero mass on prevalence values which imply negative incidence under the other model assumptions.
-
-For $\vec{P} = \matr{S} \vec{I}$ to have a unique solution, $\vec{I}$ and $\vec{P}$ need the same dimension.
-However, I am using a period of prevalence which starts after the first infections.
-These prior infections will still contribute to the prevalence.
-I assume that the incidence up to and including the first day of prevalence data, at $t = 1$, is constant at $I_0$.
-Therefore, $P_1 = I_0 \sum_{i=1}^{\dmax} S(i)$ and $I_0 = P_1 / \sum_{i=1}^{\dmax} S(i)$.
-For day $t$, the contribution of the incidence for $t < 1$ can be calculated as $P'_t = I_0 \sum_{i=t+1}^{\dmax} S(i)$.
-Subtracting $\vec{P'}$ from $\vec{P}$ before applying the forward solving procedure accounts for the contribution of the incidence prior to the first day of prevalence data.
-
-Once the incidence proportion in each stratum have been calculated using this procedure, I poststratify it to the CIS target population and subpopulations of interest.
-This is done using the methodology explained in \cref{E-biology-data:sec:MRP}.
-
-\subsection{Results and discussion} \label{backcalc:sec:results}
-
-A slow increase in incidence is estimated to occur over the autumn 2020 period, peaking in late November in most regions (see \cref{transmission:fig:backcalc-regions}).
-An exception is the North East and East of England regions (see \cref{transmission:fig:backcalc-regions}(C) and (D) respectively); these regions show no consistent increase or decrease in incidence during October.
-All regions then decrease during November; this coincides with the national lockdown which came into force in England on 5th Nov 2020~\autocite{ifgTimeline}.
-
-During November, incidence starts rising, with regional heterogeneity in the timing of when this occurs (see \cref{transmission:fig:backcalc-regions}).
-The most explored explanation for this is the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocite{walkerTracking,daviesEstimated,lythgoeLineage}.
-There is evidence that when incidence starts increasing is correlated with when the Alpha variant's spread (see \cref{transmission:fig:backcalc-alpha}).
-A second peak occurs in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
-\begin{figure}
-    \includegraphics{transmission/backcalc-regions}
-    \caption[Incidence estimated using backcalculation by region]{%
-        Incidence proportion for all of England (A) and its component regions.
-        Regions are split into broad groupings of three regions each: Northern (B), Midlands (C), and Southern (D).
-        Yorkshire and the Humber is labelled as Yorkshire for brevity.
-    }
-    \label{transmission:fig:backcalc-regions}
-\end{figure}
-\begin{figure}
-    \includegraphics{transmission/backcalc-alpha}
-    \caption[Correlation of the emergence of Alpha and minimum incidence.]{%
-        x-axis: time that S-gene target failure, a marker of the Alpha variant, is estimated to start increasing from baseline by region among CIS test results; from \textcite[table S3]{walkerTracking}.
-        y-axis: posterior median and 95\% CrI of the date of minimum incidence in November or early December in each region.
-        Linear regression line shown.
-    }
-    \label{transmission:fig:backcalc-alpha}
-\end{figure}
-
-The incidence proportion was highest among the 11--15 and 16--25 age groups throughout the period (see \cref{transmission:fig:backcalc-ages}).
-Except the 16--25 age group, all age groups show similar trends (\ie parallel lines on a log scale), meaning there was insufficient information in the data to overcome the prior which favoured this situation.
-The incidence proportion in the 16--25 age group decreases faster than other age groups during November.
-There are many plausible explanations for this effect.
-The higher incidence in this age group prior to November means it could be an immunity effect.
-Alternatively, it could be a consequence of schools remaining open in the November lockdown; 20--24 year olds rarely interact with school-aged children (see \cref{E-SEIR:fig:age-contacts})
-Many other explanations are possible, and further work is required to determine causality.
-\begin{figure}
-    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{transmission/backcalc-ages}}
-    \caption[Incidence estimated using backcalculation by age group]{%
-        Incidence proportion for each age group, on both natural and log scales.
-    }
-    \label{transmission:fig:backcalc-ages}
-\end{figure}
-
-An edge effect where a rapid rise is seen between day 1 and 2 is visible (see \cref{transmission:fig:backcalc-start-effect}).
-This is due to the incorrect assumption that incidence is constant in the past.
-Changing this assumption (\eg to exponential growth) modifies the estimated value on the first day, but the rest of the series is insensitive to this change (not shown).
-Therefore, the first day's estimate is unreliable, and has been excluded from all other plots.
-However, the rest of the series is robust to this assumption and hence reliable.
-\begin{figure}
-    \centering \includegraphics{transmission/backcalc-start-effect}
-    \caption[Edge effects in backcalculation method]{%
-        First seven days of incidence proportion estimate for the whole of England, highlighting the edge effect.
-    }
-    \label{transmission:fig:backcalc-start-effect}
-\end{figure}
-
-Alternative inference techniques could allow more powerful modelling approaches, if they allowed a broader set of models than LGMs.
-In particular, the incidence-inferred approach could be adopted.
-This could allow more interpretable priors to be used because the infection process is the process that corresponds to biological and behavioural processes.
-Furthermore, the incidence proportion could be restricted to be non-negative, preventing the negative incidence seen in some strata.
-
-Various alternative inference techniques could be considered.
-If the infection process is considered a Poisson process (defined in \cref{E-inc-prev:sec:poisson-process}), then, conditional on the process's intensity, there is independence between individuals' infection status.
-The conditional independence implies the incidence proportion and prevalence are the mean of many independent Bernoulli variables.
-Therefore, the Central Limit Theorem implies they are approximately normally distributed.
-Exploiting this property could enable efficient approximations to the posterior.
-For example, a Laplace approximation or variational inference.
-Further work to check the quality of these approximations, is required.
-
-\section{Mechanistic modelling} \label{SEIR}
-
-In this section, I apply the general mechanistic transmission modelling framework from \cref{E-SEIR} to the CIS prevalence data.
-The application of this framework follows \textcite{birrellRealtime}.
-The largest modification from the generic framework presented in \cref{E-SEIR} is the parameterization of the WAIFW matrix (\cref{SEIR:sec:WAIFW-application}).
-A novel approach (\cref{SEIR:sec:observation}) links this transmission model to the CIS prevalence data; the approach is generalisable to other studies collecting representative prevalence data, \eg REACT.
-In the mechanistic modelling literature, this part of the model is referred to as the observation model, however, in the terminology of \cref{E-inc-prev}, it includes both the prevalence process and the observation process.
-Having specified the full model (\cref{SEIR:sec:methods-application}), I conducted a simulation study to validate the inference process and identifiability (\cref{SEIR:sec:sim-study}), before applying the model to the CIS data (\cref{SEIR:sec:application}).
-In \cref{SEIR:sec:discussion}, I discuss the results; in \cref{transmission:sec:comparison} I compare the results to those in the previous section.
-
-In common with \cref{E-SEIR:sec:mechanistic-models}, the notation in this section differs slightly from the rest of the thesis.
-
-\subsection{Methods} \label{SEIR:sec:methods-application}
-
-% The model will be stratified only by region and age group, so the data are aggregated to these levels.
-% Only individuals aged two or older are eligible to participate in the CIS; I assume that test results in all participants aged under 11 are still representative of the whole $[0, 11)$ age group.
-
-I model each region fully independently, \ie with no shared parameters or transmission between regions.
-As well as the motivation given in \cref{E-backcalc}, independent regions allow heterogeneity in the parameters between regions, which could arise from a range of socioeconomic factors.
-Furthermore, the assumption of homogeneous mixing within age groups is more plausible for regions than the whole of England.
-By modelling each region independently, I assume that the epidemics in each region do not interact.
-Clearly, this is not realistic as travel between regions of England is common and unrestricted.
-However, except in an epidemic's early stages, the interaction between large regions has a negligible impact on the epidemic's progression~\autocite[e.g.][]{birrellRealtimea,gogSpatial,eggoSpatial}.
-
-\subsubsection{WAIFW matrix} \label{SEIR:sec:WAIFW-application}
-
-A parsimonious model is needed to parameterize the WAIFW matrix.
-In \cref{E-SEIR:sec:structured-populations}, I explained how the elements of the WAIFW matrix, $\matr{\beta}$ can be decomposed as $\beta_{aa'} = \kappa_{aa'} q_{aa'} N_{a'} / N_{a}$ where\todo{should I repeat these definitions?
-they already exist in ch 3}: $\kappa_{aa'}$ is the mean rate at which an individual in stratum $a'$ makes contact with individuals in stratum $a$; $q_{aa'}$ is the probability of transmission upon contact between an infectious individual in stratum $a'$ and a susceptible individual in stratum $a$; and $N_{a}$ is the total number of individuals in stratum $a$.
-For many epidemics, where the population behaviour is fairly constant over time, a constant $\matr{\beta}$ is a good approximation.
-However, in SARS-CoV-2, population behaviour changed dramatically, both spontaneously (\eg due to fear) and in response to government-imposed restrictions.
-Therefore, following \textcite{birrellRealtime}, I introduce a time-dependence giving: $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'}(t) N_{a'} / N_a$; $N_a$ and $N_{a'}$ are constant because the change in the total population is negligible over the length of time considered here.
-
-I take information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using the methodology of \textcite{vanleeuwenTime,vanleeuwenAugmenting}.
-%; these estimates have been used for previous modelling of the SARS-CoV-2 pandemic~\autocite{vanleeuwenTime,birrellRealtime}.
-% These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
-This method starts with pre-pandemic baseline estimates of $\kappa_{aa'}$ from the POLYMOD study~\autocite{mossongSocial}, and time use from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
-POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on each contact's age and location (home, work, school, leisure, transport, or other).
-UKTUS asked a representative sample of the UK population about how much time they spent doing various activities (\eg shopping and spending time at home) each day.
-\Textcite{vanleeuwenAugmenting} combined these two datasets to estimate the rate of contacts between age groups during each activity.
-\Textcite{birrellRealtime} then modify these contact rates based on measured changes in time use during the pandemic.
-These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
-Google mobility data are based on the location of Android users over time, relative to a pre-pandemic baseline.
-The mobility and school data are averaged over a week then used to down or up-weight the number of contacts in each setting, before combining contacts across settings to form $\kappa_{aa'}(t)$.
-$\kappa_{aa'}(t)$ is assumed to be piecewise constant, changing only at the start of each week.
-For further details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.
-
-The probability of transmission, $q_{aa'}(t)$, needs further decomposition to be identifiable, for the same issue of dimensionality preventing each element of $\matr{\beta}$ being estimated independently (see \cref{E-SEIR:sec:structured-populations} for a full explanation).
-I assume that $q_{aa'}(t) = \beta(t) q^\text{susc}_{a} q^\text{inf}_{a'}$ where $\beta(t)$ is a time-varying reference probability of transmission; $q^\text{susc}_a$ is the relative \emph{susceptibility} of stratum $a$, \ie how vulnerable to infection individuals in stratum $a$ are; and $q^\text{inf}_{a'}$ is the relative \emph{transmissibility} of stratum $a'$, \ie how likely individuals in stratum $a$ are to pass on an infection.
-These measures include both biological differences, \eg due to differences in immune systems, or behavioural, \eg if infants are less hygienic than adults.
-
-The time-varying component, $\beta(t)$, allows for behavioural changes that impact transmission but are not captured by the contact matrices (\eg mask wearing).
-I impose a random walk prior on $\log\beta(t)$ on the weekly timescale, smoothing its value over time.
-This prior encodes that behaviour changes smoothly over time.
-\begin{align}
-    \beta(t) &= \beta_{\tilde{w}(t)} \\
-    \log\beta_{w+1} &= \log\beta_w + \sigma_\epsilon \epsilon_w \\
-    \epsilon_w &\sim \N(0, 1)
-\end{align}
-where $\tilde{w}(t)$ is the week number of time $t$, with week 1 being the first week considered.
-Two hyperparameters will be inferred: $\beta_1$, the initial value of the random walk; and $\sigma_\epsilon$, the standard deviation of the random walk (see \cref{SEIR:sec:priors}).
-$\beta(t)$ is the only time-varying part of the transmission model that is not fixed a priori.
-Therefore, in practice, it will also absorb other effects such as any error in $\matr{\kappa}$.
-
-% $K_{ij}$ and $\beta_{ij}$ appear only as a product and therefore are not individually identifiable.
-% Normally, $K_{ij}$ is assumed to be known, for example from a previous contact survey~\autocite[such as][]{mossongSocial}.
-% Ideally, all the $\beta_{ij}$ terms would be estimated, but there is normally inadequate data to do so, requiring assumptions to be made to reduce the number of parameters~\autocite[65]{keelingModeling}.
-% % $\beta_{ij}$ is then estimated, although often with a parsimonious model
-% A common assumption is that $\beta_{ij}$ can be decomposed into two independent components such that $\beta_{ij} = \beta \beta^\text{susc}_{i} \beta^\text{inf}_j$.
-% For any stratum $i$, $\beta^\text{susc}_i$ is known as the \emph{relative susceptibility} of stratum $i$.
-
-%Now, we assume that $\beta_{ij} = \beta(t) \beta_{ij}$ where $\beta(t)$ is a non-strata-specific, time-varying component representing precautions taken by the population, for example masking; and $\beta_{ij}$ represents a constant probability of transmission from an infectious individual in strata $j$ to a susceptible individual in strata $i$ given contact, which could, for instance, be affected by differential susceptibility to infection for different age-groups.
-%We arbitrarily choose one $\beta_{ij}$ to be equal to 1 for identifiability and hence the other $\beta_{ij}$ values are relative to this reference class.
-
-Studies of pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults~\autocite{chenRole,vinerTransmission}.
-Therefore, I set $q^\text{inf}_{a'} = 1$ for all $a'$, $q^\text{susc}_{a} = q_c$ for 0--11 year olds, and $q^\text{susc}_{a} = 1$ for the remaining strata.
-$q_c$ can be interpreted as the relative probability that a child is infected (compared to an adult), given contact with an infectious individual.
-%$q_c$ is then estimated, although using an informative prior (see \cref{SEIR:sec:priors}).
-
-\subsubsection{Prevalence and observation models} \label{SEIR:sec:observation}
-
-This section  explains the novel observation model approach I propose for fitting to the CIS prevalence data.
-This will link the model in the previous section to the CIS prevalence data.
-
-% Here, I extend the system of ODEs to include compartments representing RT-PCR positive individuals.
-% The proportion of RT-PCR positive individuals in each strata is then linked to the CIS using a Beta-Binomial likelihood, with the modelled proportion that are RT-PCR positive as the mean of the Beta distribution.
-One approach would be to convolve the incidence based on the transmission model with the RT-PCR duration distribution.
-Here, the daily incidence proportion for each stratum derived by solving the system of ODEs governing the transmission model (see \cref{E-SEIR:sec:SIR}), and then prevalence calculated using \cref{E-inc-prev:eq:EPt}.
-However, this approach is computationally expensive, likely prohibitively expensive if performed within a MCMC loop.
-
-A computationally cheaper option, which I will take, is to approximate this convolution by adding compartments to the ODEs.
-This approach allows an infected, pre-positive compartment, which I denote $p_0$, to be included.
-Up until now, this stage of the disease (shown in \cref{E-biology-data:fig:natural-history}) has been neglected, assumed to be of negligible duration.
-An advantage of this modelling approach is that it can be included without complications (\eg issues of numerical stability).
-
-$p_0$ increases at the same rate as individuals flow into $e_1$.
-These individuals then transition into a set of compartments, $p_1, \dots, p_l$, each with their own transition rate.
-I will choose $l$ and the transition rates such that combined length of stay (\ie time from entry in $p_1$ until exit from $p_l$) approximates the duration of positivity distribution estimated in \cref{E-imperf-test}.
-This idea extends the technique of having multiple exposed or latent compartments (introduced in \cref{E-SEIR:sec:non-exponential}).
-While the use of exponential or Erlang (gamma distributions with integer first parameter) is common in the observation model of compartmental models~\autocite[e.g.][]{overtonEpiBeds}, the use of more flexible phase-type distributions, as here, has rarely, if ever, been used in practice~\autocite{hurtadoGLCT}.
-This broader class of distributions allows the model to capture the tail of the distribution, which is important when incidence suddenly decreases.
-
-\Textcite{osogamiClosed} provide a method to construct the compartments, implemented in the mapfit R package~\autocite{mapfit}. 
-The resulting compartmental approximation matches the original distribution up to the first three moments.
-Applying the package to the posterior mean of the distribution estimated in \cref{E-imperf-test} gives the model in \cref{SEIR:fig:full-model} with the parameter estimates in \cref{SEIR:table:ec-params} (see \cref{transmission:sec:phase-type} for details).
-\begin{table}
-    \centering
-    \begin{tabular}{c c c c c}
-        $l$ & $\omega_1$ & $\omega_{2}$ & $\omega_{3}$ & $\pi$ \\
-        3 & 0.0820 & 0.126 & 0.0223 & 0.00973  \\
-    \end{tabular}
-    \caption{Parameter estimates for the EC distribution approximating the posterior mean of the duration distribution estimated in \cref{E-imperf-test}.}
-    \label{SEIR:table:ec-params}
-\end{table}
-% end table of ec params
-
-I link the modelled proportion of individuals that are RT-PCR positive to the data using a beta-binomial likelihood (as discussed in \cref{E-biology-data:sec:clustering}).
-If on each day $t$, each stratum $i$ is observed to have $y_{at}$ positives out of $n_{at}$ tests, then the likelihood is:
-\begin{align}
-    \prod_{a,t} \BB (y_{at} \mid n_{at}, \bar{p}_{at}, \rho)
-    \label{SEIR:eq:likelihood}
-\end{align}
-where: $\bar{p}_{at}$ is the mean proportion of individuals in strata $a$ that are RT-PCR positive on day $t$; and $\rho$ is a nuisance parameter controlling the overdispersion of the beta-binomial distribution.
-Denoting the proportion of stratum $a$ in state $p_k$ as $p_{ka}(\tau)$, then $\bar{p}_{at} = \int_{t}^{t+1} \sum_{k=1}^l p_{ka}(\tau) d\tau$.
-
-\subsubsection{Model equations and solution} \label{SEIR:sec:full-model}
-
-\begin{figure}
-\makebox[\textwidth][c]{
-\begin{tikzpicture}[
-    node distance = 2.5cm,
-    on grid,
-    auto,
-    ->,>=stealth',
-    every state/.style={draw,rectangle},
-    ]
-
-    \node[state] (S) {$s$};
-    \node[state, right=of S] (E1) {$e_1$};
-    \node[state, right=of E1] (E2) {$e_2$};
-    \node[state, right=of E2] (I1) {$i_1$};
-    \node[state, right=of I1] (I2) {$i_2$};
-    \node[state, right=of I2] (R) {$r$};
-
-    \path (S) edge node {$\lambda$} (E1)
-          (E1) edge node {$\sigma$} (E2)
-          (E2) edge node {$\sigma$} (I1)
-          (I1) edge node {$\gamma$} (I2)
-          (I2) edge node {$\gamma$} (R);
-
-    \node[state, below=3cm of E1] (P0) {$p_0$};
-    \node[state, right=of P0] (P1) {$p_1$};
-    \node[state, right=of P1] (P2) {$p_2$};
-    \node[state, right=of P2] (P3) {$p_3$};
-    \node[state, right=of P3, draw=none] (P4) {};
-
-    \path  (S) edge node {$\lambda$} (P0)
-          (P0) edge node {$\omega_0$} (P1)
-          (P1) edge node {$\omega_1$} (P2)
-          (P2) edge node {$\pi \omega_{2}$} (P3)
-               edge [out=-45,in=-135] node[below] {$(1 - \pi) \omega_{2}$} (P4)
-          (P3) edge node {$\omega_{3}$} (P4);
-
-    \node[draw, thick, inner sep=0.3cm, fit=(P1) (P3), fill=blue!10,opacity=0.2] {};
-\end{tikzpicture}
-}
-  \caption[SEIR model]{SEIR model used for the analysis in this section. The shaded box represents RT-PCR-positive individuals. Shown for a single stratum.}
-  \label{SEIR:fig:full-model}
-\end{figure}
-
-The remaining details of the transmission model structure follow \textcite{birrellRealtime}.
-The transmission model is an SEIR model with two latent and two infectious states.
-To form whole of England estimates, each region is modelled independently with the results aggregated, using the poststratification methodology explained in \cref{E-backcalc}.
-
-A diagram of the model for a single stratum is shown in figure \cref{SEIR:fig:full-model}.
-The full system of differential equations for each region is below.
-\begin{equation}
-\begin{aligned}
-    \vec{\lambda}(t) &= \beta(t) (\matr{\kappa}(t) \circ \matr{q} \circ \matr{N}) ((\vec{i_1}(t) + \vec{i_2}(t))) \\
-    \frac{d\vec{s}(t)}{dt} &= -\vec{\lambda}(t) \circ \vec{s}(t) \\
-    \frac{d\vec{e_1}(t)}{dt} &= \vec{\lambda}(t) \circ \vec{s}(t) - 2\sigma \vec{e_1}(t) \\
-    \frac{d\vec{e_2}(t)}{dt} &= 2\sigma \vec{e_1}(t) - 2\sigma \vec{e_2}(t) \\
-    \frac{d\vec{i_1}(t)}{dt} &= 2\sigma \vec{e_2}(t) - 2\gamma \vec{i_1}(t) \\
-    \frac{d\vec{i_2}(t)}{dt} &= 2\gamma \vec{i_1}(t) - 2\gamma \vec{i_2}(t) \\
-    \frac{d\vec{r}(t)}{dt} &= 2\gamma \vec{i_2}(t) \\
-    \frac{d\vec{p_0}(t)}{dt} &= \vec{\lambda} \circ \vec{s}(t) - \omega_0 \vec{p_0}(t) \\
-    \frac{d\vec{p_1}(t)}{dt} &= \omega_0 \vec{p_0}(t) - \omega_1 \vec{p_1}(t) \\
-    \frac{d\vec{p_2}(t)}{dt} &= \omega_1 \vec{p_1}(t) - \omega_{2} \vec{p_2}(t) \\
-    \frac{d\vec{p_3}(t)}{dt} &= \pi \omega_{2} \vec{p_2}(t) - \omega_{3} \vec{p_3}(t)
-\end{aligned}
-\label{SEIR:eq:fullODEs}
-\end{equation}
-where $\circ$ indicates element-wise multiplication and $\matr{N}$ is a matrix with $(\matr{N})_{aa'} = N_{a'} / N_{a}$.
-See \cref{SEIR:eq:fullODEs}
-
-I solve the system of ODEs using Euler's method, discretised at timesteps of $\delta t = 1/2$ days.
-That is, if $\vec{x}_t$ is the system state at time $t$ for $t \in \{ 0, 0.5, 1, \dots \}$ then $x_{t+\frac{1}{2}} = x_t + (d\vec{x}(t)/dt) \delta t$.
-
-The mean prevalence on day $t$, used for the likelihood (see \cref{SEIR:eq:likelihood}), can be calculated as $\bar{p}_{at} = \frac{1}{2} \sum_{k=1}^l ( p_{kat} + p_{ka,t+1/2} )$.
-
-\subsubsection{Priors} \label{SEIR:sec:priors}
-
-Priors for the parameters are given in \cref{SEIR:table:priors}.
-The parameters governing the duration of positivity ($\omega$s) were explained previously (see \cref{SEIR:table:ec-params}).
-The initial conditions and transition rate parameters are unidentifiable in an SEIR model~\autocite{dankwaStructural}, therefore strongly informative priors or fixed values are used for them.
-$\sigma_\epsilon$ and $\rho$ are given priors that favour a simpler model.
-The prior on $\sigma_\epsilon$ is a penalized complexity prior, favouring a null model of constant values for $\beta(t)$~\autocite{simpsonPenalising}.
-The prior on $\rho$ favours a null model of a binomial likelihood.
-The remaining parameters have weakly informative priors.
-
-To form the prior for $\vec{\pi}$, I used the analysis framework of \textcite{amirthalingamSeroprevalence}, applied to seroprevalence data from blood donors~\autocite{amirthalingamSeroprevalence} and a representative sample of children~\autocite{ratcliffeCommunity} in August 2020.
-This analysis uses a Bayesian framework to correct the seroprevalence data for the sensitivity and specificity of the serological tests, using an informative prior.
-\Textcite{amirthalingamSeroprevalence} formed the prior for the tests by evaluating the tests using samples which are known to be negative (taken before the pandemic) or positive (taken from an individual following a positive RT-PCR test).
-A regression model predicts the proportion of the population with prior immunity in each age group and region.
-I approximated the posterior of this proportion with a beta distribution to form the prior for the analysis in this chapter (see \cref{SEIR:table:immunity-prior}).
-
-\begin{landscape}
-\begin{table}
-\begin{tabular}{l c l l}
-    Parameter description & Symbol & Distribution & Comment \\
-    \hline \\
-    Mean latent period & $1/\sigma$ & 3.5 days (fixed) & Based on \textcite{zhaoEstimating} \\
-    Mean infectious period & $1/\gamma$ & 4 days (fixed) & Based on \textcite{zhaoEstimating} \\
-    Initial proportion without immunity & $\vec\pi$ & See \cref{SEIR:table:immunity-prior} & \\
-    Initial proportion infectious & $i^+$ & $\BetaDist(0.5, 1000)$ & Weakly informative \\
-    Susceptibility of children & $q_c$ & $\LN(-0.4325, 0.1174)$ & Based on \textcite{vinerTransmission}  \\
-    Initial growth rate & $\psi$ & $\N(0.06, 0.04)$ & Weakly informative \\
-    Standard deviation of random walk steps on log scale & $\sigma_\epsilon$ & $\Exponential(80)$ & penalized complexity prior \\
-    Overdispersion of observations & $\rho$ & $\Exponential(2 \times 10^5)$ & penalizing deviations from binomial
-\end{tabular}
-\caption[SEIR model priors]{Priors for each parameter. For details of the distributions and their parameterizations see \cref{E-distributions}.}
-\label{SEIR:table:priors}
-\end{table}
-\begin{table}
-
-\centering
-\begin{tabular}{l|lllll}
-         & $[0,16)$ & $[16,25)$ & $[25,50)$ & $[50,70)$ & $[70,\infty)$ \\
-        \hline
-        East Midlands & $\BetaDist(8.5, 301)$ & $\BetaDist(18, 256)$ & $\BetaDist(19, 475)$ & $\BetaDist(20, 493)$ & $\BetaDist(5, 337)$ \\
-        East of England & $\BetaDist(7.9, 236)$ & $\BetaDist(21, 248)$ & $\BetaDist(35, 677)$ & $\BetaDist(26, 613)$ & $\BetaDist(5.6, 332)$ \\
-        London & $\BetaDist(9.7, 131)$ & $\BetaDist(50, 328)$ & $\BetaDist(154, 1331)$ & $\BetaDist(106, 990)$ & $\BetaDist(7.6, 204)$ \\
-        North East & $\BetaDist(6.1, 180)$ & $\BetaDist(12, 145)$ & $\BetaDist(14, 274)$ & $\BetaDist(14, 295)$ & $\BetaDist(4.2, 240)$ \\
-        North West & $\BetaDist(12, 252)$ & $\BetaDist(38, 353)$ & $\BetaDist(72, 885)$ & $\BetaDist(44, 695)$ & $\BetaDist(6.3, 264)$ \\
-        South East & $\BetaDist(8.9, 404)$ & $\BetaDist(17, 319)$ & $\BetaDist(20, 577)$ & $\BetaDist(19, 586)$ & $\BetaDist(4.4, 356)$ \\
-        South West & $\BetaDist(6.9, 415)$ & $\BetaDist(11, 277)$ & $\BetaDist(12, 556)$ & $\BetaDist(11, 559)$ & $\BetaDist(4, 492)$ \\
-        West Midlands & $\BetaDist(8, 167)$ & $\BetaDist(19, 151)$ & $\BetaDist(32, 475)$ & $\BetaDist(33, 504)$ & $\BetaDist(5.8, 241)$ \\
-        Yorkshire and The Humber & $\BetaDist(9.2, 314)$ & $\BetaDist(19, 242)$ & $\BetaDist(27, 589)$ & $\BetaDist(18, 474)$ & $\BetaDist(4.9, 316)$ \\
-    \end{tabular}
-\caption{Priors for $\vec{\pi}$, the proportion of the population without prior immunity, in each age group and region combination.}
-\label{SEIR:table:immunity-prior}
-\end{table}
-\end{landscape}
-
-\subsubsection{Inference} \label{SEIR:sec:inference}
-
-I compute the posterior distribution using MCMC.
-Specifically, I use an adaptive random-walk Metropolis--Hastings algorithm (discussed in \cref{E-MCMC}).
-The proposal distribution is a multivariate normal that adapts to the covariance of the posterior.
-The adaptive algorithm is based on \textcite[algorithm 4]{andrieuTutorial}, with the implementation based on \textcite{ghoshApproximate}'s work.
-Proposals are accepted or rejected using the standard Metropolis--Hastings (MH) acceptance probability.
-See \cref{SEIR:MCMC-algorithm} for full details.
-
-\begin{algorithm}
- Set $\vec{X_0}$ to an initial value of the parameter vector \;
- $\vec{\mu_0} = \vec{X_0}$ \;
- $\matr{\Sigma_0} = \text{diag}(\vec{\mu_0})$ \;
- $\lambda_0 = 1$ \;
- \For{$i = 1, \dots, M$}{
-  sample $\vec{Y_}i \sim \N(\vec{\mu_{i-1}}, \lambda_{i-1}\matr{\Sigma_{i-1}})$\;
-  set $\vec{X_i}$ to $\vec{Y_i}$ or $\vec{X_{i-1}}$ using a MH acceptance step\;
-  update the proportion of proposals accepted so far, $\alpha$ \;
-  \eIf(\tcc{No adaptation for 200 iterations}){$i \leq 200$}{
-    $\lambda_i = \lambda_{i-1}$ \;
-    $\vec{\mu_i} = \vec{\mu_{i-1}}$ \;
-    $\matr{\Sigma_i} = \matr{\Sigma_{i-1}}$ \;
-   }{
-    $\gamma_i = (i - 200)^{-0.6}$ \;
-    $\log \lambda_i = \log \lambda_{i-1} + \gamma_i(\alpha - 0.234)$ \;
-    $\vec{\mu_i} = (1 - \gamma_i) \vec{\mu_{i-1}} + \gamma_i \vec{X_i}$ \;
-    $\matr{\Sigma_i} = (1 - \gamma_i) \matr{\Sigma_{i-1}} + \gamma_i (\vec{X_i} - \vec{\mu_i})(\vec{X_i} - \vec{\mu_i})^T$ \;
-  }
- }
- \caption{Algorithm for adaptive random-walk Metropolis--Hastings.
-$\vec{\mu_i}$ and $\matr{\Sigma_i}$ are an estimate of the mean and covariance of the posterior distribution using information up to iteration $i$.
-$\text{diag}(\vec{\mu_0})$ is the diagonal matrix with diagonal entries equal to $\vec{\mu_0}$.
-$\lambda_i$ is the scale parameter of the proposal distribution at iteration $i$, tuned to try and ensure an optimal proportion of proposals are accepted (23.4\%).
-$\gamma_i$ is the learning rate, which determines how much adaptation occurs.
-$\gamma_i \to 0$ as $i \to \infty$ so the rate of adaptation is \emph{vanishing}.
-Vanishing adaptation is required for the algorithm to converge to the target distribution~\autocite[section 3]{andrieuTutorial}.}
- \label{SEIR:MCMC-algorithm}
-\end{algorithm}
-
-I transformed parameters with bounded supports to the real line to improve sampling efficiency.
-Parameters which are constrained to the positive reals ($\sigma_\epsilon$, $\rho$, and $q_c$) were sampled on the log scale.
-Parameters which are constrained to the unit interval ($i^+$ and $\vec{\pi}$) were sampled on the logit scale.
-
-All results are based on four Markov chains each of length 1,000,000 iterations.
-The first 500,000 were discarded as burn-in, with the remaining chains thinned to every 100th iteration.
-This gives 5,000 posterior samples per region.
-This was further thinned to 2,000 samples when calculating posterior predictive quantities (incidence and prevalence) due to their computational cost.
-
-Following \textcite{birrellBayesian}, I reparameterize the model to improve identifiability.
-I parameterize $\beta_1$ in terms of the initial growth of the epidemic.
-I also use a parsimonious parameterization of the initial conditions using the initial proportion of the population that are infectious and the initial proportion that are susceptible only.
-
-Both reparameterizations rely on an assumption that the epidemic is in a steady-state exponential growth phase at time 0 (see \cref{E-SEIR:sec:SIR}).
-During such an exponential growth phase, all components of the $\vec{e}$, $\vec{i}$ and $\vec{p}$ compartments grow at the same growth rate, $\psi$.
-That is:
-\begin{equation}
-\begin{aligned}
-\frac{d\vec{e_1}(t)}{dt} = \psi \vec{e_2}(t) \\
-\frac{d\vec{e_2}(t)}{dt} = \psi \vec{e_2}(t) \\
-\frac{d\vec{i_1}(t)}{dt} = \psi \vec{i_2}(t) \\
-\frac{d\vec{i_2}(t)}{dt} = \psi \vec{i_2}(t) \\
-\frac{d\vec{p_0}(t)}{dt} = \psi \vec{p_0}(t) \\
-\frac{d\vec{p_1}(t)}{dt} = \psi \vec{p_1}(t) \\
-\frac{d\vec{p_2}(t)}{dt} = \psi \vec{p_2}(t) \\
-\frac{d\vec{p_3}(t)}{dt} = \psi \vec{p_3}(t).
-\end{aligned}
-\label{SEIR:eq:exponential-growth}
-\end{equation}
-This follows directly from the ODEs if $\vec{s}$ is constant.
-
-The growth rate $\psi$ is easily identifiable from the data, and therefore a parameterization in terms of $\psi$ should be more computationally efficient.
-If this is the initial exponential phase, when $\vec{s} \approx 1$, then the basic reproduction number $\R$ for an SEIR model is linked to this growth rate by~\autocite{birrellBayesian,wearingAppropriate}:
-\begin{align}
-    R_{(\psi)} = \frac{\psi \left( \frac{\psi}{2\sigma} + 1 \right)^2}{\gamma \left(1 - \frac{1}{\left(\frac{\psi}{2 \gamma} + 1 \right)^2} \right)} \label{SEIR:eq:rtoR}.
-\end{align}
-
-Using the results of \cref{E-SEIR:sec:structured-populations}, $\R$ can also be written as $R_{(\beta)} = \beta_1 D(\matr{\kappa}(t) \circ \matr{q} \circ \matr{N})$ (recall that $D(\matr{\beta})$ denotes the dominant eigenvalue of $\matr{\beta}$).
-Hence setting $\beta_1 = R_\psi / D(\matr{\kappa}(t) \circ \matr{q} \circ \matr{N})$ gives $R_\psi = R_\beta$.
-Using this expression, $\beta_1$ has been reparameterized in terms of $\psi$, $q_c$, and fixed quantities.
-
-To form a parsimonious parameterization of the initial state, first focus on the initial proportion of the population that is in the infectious compartments.
-Recall, from \cref{E-SEIR:sec:structured-populations}, that in the exponential growth phase, the relative proportion of each stratum that is infectious is proportional to the dominant eigenvector of $\matr{\beta}$, which I denote $\vec{d}$.
-Introduce a model parameter $i^+$, the proportion of the population in either of the $i$ states at time 0 in the strata with the highest proportion infectious.
-Define the total number of infectious individuals at time 0 as $\vec{i_0}$.
-Then:
-\begin{align}
-\vec{i_0} = \vec{i_1}(0) + \vec{i_2}(0) = \frac{i^+}{\max \vec{d}} \vec{d}.
-\label{SEIR:eq:i0}
-\end{align}
-
-Equating the expressions for $d\vec{i_2}/dt$ in \cref{SEIR:eq:fullODEs} and \cref{SEIR:eq:exponential-growth} and then setting $t=0$ to allow substituting in \cref{SEIR:eq:i0} gives:
-\begin{align}
-    \vec{i_1}(0) &= \vec{i_0} \left(1 + \frac{2\gamma}{\psi + 2\gamma} \right)^{-1}.
-\intertext{By definition of $\vec{i_0}$:}
-    \vec{i_2}(0) &= \vec{i_0} - \vec{i_1}(0).
-\end{align}
-
-A similar process, equating the expressions in \cref{SEIR:eq:fullODEs} and \cref{SEIR:eq:exponential-growth}, gives the following solution for the initial state of these compartments.
-\begin{align}
-    \frac{d\vec{e_1}(t)}{dt} &= \vec{\lambda}(t) \circ \vec{s}(t) - 2\sigma \vec{e_1}(t) \\
-    \frac{d\vec{e_2}(t)}{dt} &= 2\sigma \vec{e_1}(t) - 2\sigma \vec{e_2}(t) \\
-    \frac{d\vec{p_0}(t)}{dt} &= \vec{\lambda} \circ \vec{s}(t) - \omega_0 \vec{p_0}(t) \\
-    \frac{d\vec{p_1}(t)}{dt} &= \omega_0 \vec{p_0}(t) - \omega_1 \vec{p_1}(t) \\
-    \frac{d\vec{p_2}(t)}{dt} &= \omega_1 \vec{p_1}(t) - \omega_{2} \vec{p_2}(t) \\
-    \frac{d\vec{p_3}(t)}{dt} &= \pi \omega_{2} \vec{p_2}(t) - \omega_{3} \vec{p_3}(t).
-\end{align}
-
-The definition of $\vec{\pi}$ as the proportion of the population without initial immunity gives $\vec{r}(0) = 1 - \vec{\pi}$.
-Finally, to ensure the population proportions sum to one, $\vec{s}(0) = \vec{\pi} - \vec{i_0} - \vec{e_1} - \vec{e_2}$.
-
-\subsection{Simulation study} \label{SEIR:sec:sim-study}
-
-I ran a simulation study to check that the inference procedure (\cref{SEIR:sec:inference}) was working correctly and that the model was identifiable.
-I generated 100 simulated datasets representing the East of England using the following procedure.
-
-First, I drew 100 parameter sets drawn from the priors (see \cref{SEIR:table:priors}), except that the prior on $\psi$ was $\LN(0.048, 0.0035)$.
-I used this more informative prior for the simulation study to restrict the simulations to epidemics growing at $t=0$.
-The weakly informative prior, used for the main analysis, allows a wide range of epidemics including some that are shrinking at the start of the time period.
-The prior for $\vec{\pi}$ was the same as the East of England.
-
-Next, I solved the ODE system in \cref{SEIR:eq:fullODEs} for each parameter set.
-All other aspects of the ODEs, including the data-informed contact matrices, $\matr{\kappa}$, are as previously described in this section.
-Population sizes were those of the East of England, and the time period the same as the main analysis.
-This means that the simulations created realistic incidence and prevalence curves (see \cref{SEIR:fig:sim-data}).
-In particular, the October half-term school holiday is visible, especially in the school-age groups and the November lockdown is visible to some extent in all age groups.
 \begin{figure}
@@
     \vspace{-3cm}
-    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/sim/data}}
-    \caption[Simulated data]{%
-        (A) Incidence and (B) prevalence for the 100 simulated data sets as a proportion of each stratum.
-        Prevalence shown is $\bar{p}_{kat}$, the true population prevalence on each day, excluding sampling noise.
-        Each line represents one simulation.
-    }
-    \label{SEIR:fig:sim-data}
-\end{figure}
-
-Finally, I generated one dataset of test results for each parameter set.
-These were generated according to the likelihood, \cref{SEIR:eq:likelihood}.
-The number of daily tests was set to the mean of what CIS performed in the region over this period (1630), distributed across the population in proportion to its size.
-
-Having generated 100 datasets, I used the inference procedure in \cref{SEIR:sec:inference} to find the posterior distribution conditional on each dataset.
-Hence, I ended up with 100 parameter sets distributed according to the priors (the ``generative values"), with an associated dataset and posterior distribution for it.
-
-To check the performance of the inference procedure, I calculated the coverage of the 50\%, 75\%, and 95\% central credible intervals for each posterior distribution.
-Define coverage for a parameter as the proportion of credible intervals for that parameter containing the generative value.
-As the parameters are drawn from the prior, if the posterior distribution is inferred correctly then the coverage for each parameter should equal the nominal probability (\eg the coverage should be 50\% for the 50\% credible intervals)~\autocite{cookValidation}.
-This check is related to, although simpler than, simulation-based calibration~\autocite{taltsValidating}.
-I constructed a confidence interval for the coverage using the Pearson-Klopper method implemented in the binom package~\autocite{binom1-1}.
-
-The majority of the confidence intervals contain the nominal probability (see \cref{SEIR:fig:sim-coverage}).
-There is a tendency for the coverage to be slightly lower than the nominal probability, suggesting that the credible intervals are slightly too narrow.
-This is a common issue with random walk MCMC (see \cref{E-MCMC:sec:improving}).
-The same issue arises with the coverage for derived quantities incidence and prevalence (see \cref{SEIR:fig:sim-inc-prev}).
-\begin{figure}
-    \includegraphics{SEIR/sim/coverage}
-    \caption[Coverage of simulation study]{%
-        Point estimate and 95\% confidence interval for coverage of the 50\%, 75\%, and 95\% central credible intervals for each parameter; see text for full definition.
-        Vertical lines show the nominal coverage.
-    }
-    \label{SEIR:fig:sim-coverage}
-\end{figure}
-\begin{figure}
-    \thisfloatpagestyle{empty}
-    \vspace{-3cm}
-    \makebox[\textwidth][c]{\includegraphics[width=0.9\paperwidth]{SEIR/sim/predictive_coverage}}
-    \caption[Coverage of simulation study (derived quantities)]{%
-        Point estimate and 95\% confidence interval for coverage of the 50\%, 75\%, and 95\% central credible intervals for (A) incidence and (B) prevalence; see main text for full definition.
-        Horizontal lines show the nominal coverage.
-    }
-    \label{SEIR:fig:sim-inc-prev}
-\end{figure}
-
-In this set-up, the parameters $i^+$, $q_c$, and $\psi$ are fully identified; $\pi_0$ is not identified at all; the other parameters are partially identified, their posterior has moved away from the prior but there is insufficient information in the data to fully overcome the prior (see \cref{SEIR:fig:true-vs-posterior}).
-The ability to identify $q_c$ is an advantage of using data which includes measurements on children, unlike data on hospitalisations or deaths.
-
-For $\vec{\pi}$ and $\rho$, the correlation between the posterior median and the generative value is weak (see \cref{SEIR:fig:true-vs-posterior}).
-This suggests that these parameters are not practically identifiable in this setting.
-The posteriors align very closely with the priors (not shown) which reinforces this finding.
-The lack of identifiability for $\vec{\pi}$ aligns with previous work showing this parameter is structurally unidentifiable, \ie not identifiable even in the limit of infinite data~\autocite{dankwaStructural}.
-\begin{figure}
-    \makebox[\textwidth][c]{\includegraphics{SEIR/sim/true_vs_posterior}}
-    \caption[True vs posterior parameter values]{%
-        True vs posterior median parameter values for each parameter in each simulation.
-        The black line shows the line $y = x$ which the dots would lie on if the parameter is fully recovered.
-        The blue line is a linear regression line.
-        For $i^+$, $q_c$, and $\psi$, the posterior recovers the true values.
-        For $\vec{\pi}$ and $\rho$, the posterior median is very similar regardless of the true value.
-    }
-    \label{SEIR:fig:true-vs-posterior}
-    \todo[inline]{Fix x-axis labels overlapping / being unreadable}
-\end{figure}
-
-\subsection{Application to CIS data} \label{SEIR:sec:application}
-
-The 11--16 age group (roughly corresponding to compulsory secondary school education) is estimated to have the highest incidence (see \cref{SEIR:fig:incidence}).
-Incidence is smooth within each week but shows discontinuities between weeks.
-This is because the parameters within $\matr{\beta}(t)$ are constant within each week.
-A constant $\matr{\beta}(t)$ would lead to exponential growth following a period of transitory dynamics, although the weekly switching may mean that these transitory dynamics never end, especially when incidence is high~\autocite{rhodesConvergence}.
-The estimated timing of the peak in incidence varies by region, with the most common time being late December 2020 (see \cref{SEIR:fig:peak-incidence}).
-In Yorkshire and the Humber, the October peak is estimated to be greater than the winter peak; for North West there is a significant probability on either being larger.
-\begin{figure}
-    \includegraphics{SEIR/CIS/incidence}
-    \caption[Posterior estimates of incidence]{%
-        Posterior estimates (95\% CrI) of incidence by region and age.
-        See main text for discussion.
-    }
-    \label{SEIR:fig:incidence}
-\end{figure}
-\begin{figure}
-    \includegraphics{SEIR/CIS/p_peak}
-    \caption[Posterior estimates of peak incidence timing]{%
-        Posterior probability of the peak incidence for each region and age combination occurring on each day.
-    }
-    \label{SEIR:fig:peak-incidence}
-\end{figure}
-
-The $\beta_w$ parameters absorb all the time-varying changes in transmission not captured by Google mobility and school attendance data.
-They trend downwards up until early December, then they start increasing (see \cref{SEIR:fig:beta-walk}).
-Correlation is clear across the regions even though each region is estimated independently, suggesting underlying biological or epidemiological changes across regions as opposed to modelling assumptions.
-
-\begin{figure}
-    \includegraphics{SEIR/CIS/beta_walk}
-    \caption[Posterior estimates of $\beta_w$]{%
-        Posterior estimates $\beta_w$ for each week.
-        Shown are ``violins'' where the width is proportional to the kernel-density estimated posterior density at that point.
-        Values trend downwards until early December when an increase starts.
-    }
-    \label{SEIR:fig:beta-walk}
-\end{figure}
-
-The \emph{attack rate} is the proportion of the population that have been infected at least once.
-In this model it is $1 - \vec{s}$, the proportion susceptible at the time the study ends.
-The attack rate is highest in the $[11, 16)$ age group in the North West at 68\% (95\% CrI: 66--70\%) (see \cref{SEIR:fig:attack-rates}).
-Some of these infections occurred prior to the period covered by this study; during the period of this study 23\% (95\% CrI: 21--24\%) of the population in the North West aged 11--16 was infected.
-\begin{figure}
-    \includegraphics{SEIR/CIS/attack_rates}
-    \caption{Posterior median and 95\% CrI of the attack rate in each region and age group.}
-    \todo[inline]{Maybe compare this to blood donors, as used in the prior.}
-    \label{SEIR:fig:attack-rates}
-\end{figure}
-
-Overall, the model fits the data well (see \cref{SEIR:fig:prev-young,SEIR:fig:prev-old}).
-\todo{move one or some of the goodness-of-fit plots into an appendix - maybe by region?}
-There are a few features of the data which are not captured by the model and the posterior estimate appears to be a compromise with the prevalence too low in one age group and too high in another.
-For example, in the North East region, the 11--16 age group has its prevalence overestimated in October but the 16--25 age group has its prevalence underestimated.
-\begin{figure}
-    \thisfloatpagestyle{empty}
-    \vspace{-3cm}
-    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_young}}
+    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_appendix}}
     \captionsetup{width=0.8\paperwidth}
-    \captionof{figure}[SEIR prevalence goodness-of-fit (younger ages)]{%
-        Goodness-of-fit to prevalence data by region and age for the younger three age groups (see \cref{SEIR:fig:prev-old} for the others).
+    \captionof{figure}[Other SEIR prevalence goodness-of-fit]{%
+        Goodness-of-fit to prevalence data by region and age for selected regions (see \cref{SEIR:fig:prev} for others).
         The central ribbon and line show the posterior median and 95\% CrI estimate for the proportion of the relevant stratum RT-PCR positive on each day.
@@
     }
-    \label{SEIR:fig:prev-young}
-\end{figure}
-\begin{figure}
-    \thisfloatpagestyle{empty}
-    \vspace{-3cm}
-    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_old}}
-    \captionsetup{width=0.8\paperwidth}
-    \captionof{figure}[SEIR prevalence goodness-of-fit (older ages)]{%
-        Goodness-of-fit prevalence for the older three age groups (see \cref{SEIR:fig:prev-young} for the other age groups and details).
-    }
-    \label{SEIR:fig:prev-old}
-    \todo[inline]{Remake/reposition this plot so that it fits on a page.}
+    \label{SEIR:fig:prev-appendix}
 \end{figure}
 
-All regions' MCMC chains converged, assessed using the Rhat statistic and ESS (as defined in \cref{E-MCMC:sec:convergence}).
-The highest Rhat was 1.01 for $\sigma_\epsilon$ in London.
-All parameters had ESS > 1000, except $\sigma_\epsilon$ and $\psi$ in London, which had ESS of 450 and 950 respectively.
-
-\subsection{Discussion} \label{SEIR:sec:discussion}
-
-The results in this section show the feasibility of fitting a mechanistic model using only data on prevalence.
-If the distributions for length-of-stay in each compartment are known, then all other model parameters can  be estimated.
-This includes the age distribution of the infections and the components of transmission.
-
-% Unlike most prior work (see\todo{ref section with lit review}), this modelling is based on a data source that is representative of the general population.
-% % Therefore, the results have fewer biases and informed by data to a greater extent than this work.
-% There is little-to-no ascertainment bias in the CIS, unlike sources relying on testing behaviour (in the UK, pillar 2 testing).
-% The CIS allows estimates of transmission in children, unlike sources relying on severe events that occur almost entirely in adults~\autocite{bhopalChildren}.
-Using a mechanistic model provides additional understanding, compared to the approach in \cref{E-backcalc}.
-For example, decomposing the change in transmission into changing contacts, $\matr{\kappa}(t)$, changing susceptibility $\vec{s}(t)$, and other unmeasured factors, $\beta(t)$.
-By understanding which of these components is changing, policies can be better targeted.
-If it is shown that $\beta(t)$ decreases, this implies that there is less transmission for the same number of contacts.
-% Such decreases might be beneficial as it allows society to operate in a more normal manner without increasing transmission.
-% Increasing activity without increasing transmission is desirable because it enables reductions the disease burden without burdesome restrictions on individuals day-to-day lives.
-
-The parameters $\beta_w$ absorb all the time-varying factors not captured by Google mobility and school attendance data.
-Many factors contribute to the pattern in the $\beta_w$ values, which have minima near the start of December.
-While this analysis cannot establish causation, several hypotheses are worth considering.
-
-The decreases in $\beta_w$ over the autumn could be due to increased caution (\eg masking or stricter adherence to social distancing) as cases and deaths increased~\autocite{jarvisEffect} and/or not captured by $\matr{\kappa}(t)$, such as the impact of local measures in the highest transmission areas~\autocite{scottCovid19}.
-Local measures are not well captured by $\matr{\kappa}(t)$ because the matrix averages behaviour across the whole of England.
-Allowing each region to have its own $\matr{\kappa}(t)$ may capture the effect of local measures more accurately, although their targeted nature may require more granular methods.
-
-The increases in $\beta_w$ in December coincide with the rise of the Alpha variant~\autocite{lythgoeLineage}, known to be more transmissible~\autocite[e.g.][]{daviesEstimated}.
-A multiple variant model might be able to capture this rise more accurately.
-Such a model would stratify positive tests in the CIS by pre-Alpha or Alpha variant.
-Then, each variant would be explicitly included in the model to fit to this data.
-
-These hypotheses for the mechanism causing changes in $\beta_w$ are correlated across the country.
-This can explain that similar patterns emerge for $\beta_w$ in different regions (see \cref{SEIR:fig:beta-walk}), despite these being estimated independently (\ie there is no model structure which favours correlated values).
-These correlations could be exploited to improve the precision of estimates with a more complex model which borrows information across regions, \eg a hierarchical model.
-
-The current model could be more flexible with regards to the age distribution of infections.
-This would better exploit one of the strengths of the CIS: the availability of data  on all age groups.
-In some regions, there appears to be a trade-off between the 11--16s and 16--25s, with the model misfitting in opposite directions (see \cref{SEIR:fig:prev-young}).
-This implies that the data would allow identification of a more flexible age-based model.
-
-Currently, the age distribution of infections is determined by $\matr{\beta}$, and to a lesser extent $\vec{\pi}$, the former of which only contains one parameter that affects the age distribution, $q_c$.
-Increasing the number of parameters in $\matr{\beta}$ would be the simplest approach to increase the model's flexibility.
-
-The only component of $\matr{\beta}$ that allows age-dependent heterogeneity in the temporal changes is $\matr{\kappa}$.
-It is plausible that this would not capture all behavioural changes.
-For example, the introduction and relaxation of control measures within schools.
-$\matr{\kappa}$ only captures changes in in-school transmission through changes in school attendance, not incorporating the probability of transmission, which could be modelled as changes in $q_{aa}$ for $a$ of school age.
-Similarly, there is evidence that $q_c$ is different for Alpha and pre-Alpha variants~\autocite{zhuRole}, which could again motivate these changing over time.
-
-The weekly patterns in incidence are a limitation of the current parameterization.
-The random walk and contact matrices both vary only on the weekly timescale.
-A daily random walk could provide more flexibility here.
-However, this would greatly increase the number of parameters.
-An alternative would be a smoothly varying function or an approximation of a random walk, such as that proposed by \textcite{ghoshApproximate}.
-
-This discussion makes clear that the results are limited by the number of parameters, which is currently limited by the inference algorithm.
-The MCMC algorithm used tends to struggle with a large number of parameters.
-Alternative proposal schemes, such as updating the parameters in blocks could help, or a more advanced algorithm such as NUTS (used elsewhere in this thesis).
-
-Extending the analyses carried out here to the rest of the pandemic would clearly be of interest.
-I chose the end of the period to coincide with the vaccination rollout.
-Vaccination dramatically affects the transmission dynamics by introducing some additional immunity into the population.
-Mechanistic models require explicit representation of this immunity.
-
-\section{Comparison of results} \label{transmission:sec:comparison}
-
-\Cref{transmission:fig:compare-regions} compares the regional incidence estimates between the two models in this chapter.
-Generally, the estimates agree with the SEIR model having much narrower credible intervals.
-The largest discrepancy is that the backcalculation model estimates a second peak in mid-January in several regions (most clearly in London, North West, and the East of England).
-However, except in London, the credible intervals still overlap.
-This is a specific case of a larger pattern: the SEIR model produces smoother estimates of incidence than the backcalculation model.
-\begin{figure}
-    \centering \includegraphics{transmission/compare-regions}
-    \caption[Comparing each models estimate of regional incidence.]{%
-        Incidence proportion in England and each region estimated by the two models discussed in this chapter.
-        Posterior median and 95\% CrI shown.
-        For the SEIR model, newly RT-PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the backcalculation results.
-    }
-    \label{transmission:fig:compare-regions}
-\end{figure}
-
-These patterns can be interpreted as an instance of the common bias---variance trade-off.
-The SEIR model imposes stronger assumptions about the epidemic dynamics.
-The assumptions reduce the variance of the estimates, including by making them smoother, but if the assumptions are incorrect, bias is introduced.
-Furthermore, the SEIR model does not fully adjust for the non-response bias.
-In particular, it makes no adjustment for the under-representation of ethnic minority groups in the CIS.
-Adjusting for more covariates would require further stratifying the population and finding an appropriate parameterization of the contact matrix.
-This trade-off based on the strength of assumptions has long been known to apply in the backcalculation context~\autocite[e.g.][section 8.3]{brookmeyerBackcalculation}.
-The SEIR model can be viewed as a backcalculation method where the transmission part of the model is a smoothing method.
-
-The backcalculation method benefits from being able to be implemented in a fast, approximate inference package (R-INLA) that easily scales to many parameters, including random effects.
-Therefore, it is far more flexible and can easily include time-varying effects, ethnicity, etc.
-The SEIR model is limited by the inference procedure and the requirement to solve an ODE to evaluate the likelihood.
-To incorporate a more flexible model, and reduce its bias, a more efficient inference procedure would be required.
-
-Many of these issues are clearly demonstrated by the estimates for the North East (see \cref{transmission:fig:compare-NE}).
-I previously observed the quality of the SEIR model's fit is variable between different age groups (see \cref{SEIR:fig:prev-young,SEIR:fig:prev-old}).
-This is because the model is not flexible enough to capture the true age distribution of infections.
-The overestimation of prevalence in the 11--16 age group and underestimation in the 16--25 age group is reflected when comparing incidence estimates to those from the backcalculation model, which have reduced bias.
-\begin{figure}
-    \centering \includegraphics{transmission/compare-NE}
-    \caption[Comparing each models estimate of North East incidence by age.]{%
-        Incidence proportion in each age group in the North East estimated by the two models discussed in this chapter.
-        Posterior median and 95\% CrI shown.
-        For the SEIR model, newly RT-PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the backcalculation results.
-    }
-    \label{transmission:fig:compare-NE}
-\end{figure}
-
-\section{Conclusion}
-
-The major contribution in this chapter is showing that transmission and incidence can be estimated using only data from the CIS.
-A similar survey could, therefore, be considered as the primary form of surveillance for a future pandemic.
-The advantage of such a set-up is the unbiased information across the entire population, including children.
-
-The transmission model presented in this chapter could be extended to fully extract the information in the CIS (see \cref{SEIR:sec:discussion}).
-This would require a more flexible model, which would likely require a more efficient inference procedure, such as NUTS.
-
-\ifSubfilesClassLoaded{
-  \appendix
-  \subfile{transmission-appendix-INLA}
-  \subfile{transmission-appendix-phase-type}
-  \listoftodos
-}{}
-
 \end{document}

### 2024-03-20 transmission.tex
make true vs posterior plot clearer
move half of plot to appendix

@@
         Incidence proportion for (A) all of England and (B) its component regions.
-        Yorkshire and the Humber is labelled Yorkshire for brevity.
     }
@@
     \label{SEIR:fig:true-vs-posterior}
-    \todo[inline]{Fix x-axis labels overlapping / being unreadable}
 \end{figure}
@@
 
-Overall, the model fits the data well (see \cref{SEIR:fig:prev-young,SEIR:fig:prev-old}).
-\todo{move one or some of the goodness-of-fit plots into an appendix - maybe by region?}
+Overall, the model fits the data well (see \cref{SEIR:fig:prev,SEIR:fig:prev-appendix}).
 There are a few features of the data which are not captured by the model and the posterior estimate appears to be a compromise with the prevalence too low in one age group and too high in another.
@@
     \vspace{-3cm}
-    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_young}}
+    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_main}}
     \captionsetup{width=0.8\paperwidth}
-    \captionof{figure}[SEIR prevalence goodness-of-fit (younger ages)]{%
-        Goodness-of-fit to prevalence data by region and age for the younger three age groups (see \cref{SEIR:fig:prev-old} for the others).
+    \captionof{figure}[SEIR prevalence goodness-of-fit]{%
+        Goodness-of-fit to prevalence data by region and age for selected regions (see \cref{SEIR:fig:prev-appendix} for others).
         The central ribbon and line show the posterior median and 95\% CrI estimate for the proportion of the relevant stratum RT-PCR positive on each day.
@@
     }
-    \label{SEIR:fig:prev-young}
-\end{figure}
-\begin{figure}
-    \thisfloatpagestyle{empty}
-    \vspace{-3cm}
-    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_old}}
-    \captionsetup{width=0.8\paperwidth}
-    \captionof{figure}[SEIR prevalence goodness-of-fit (older ages)]{%
-        Goodness-of-fit prevalence for the older three age groups (see \cref{SEIR:fig:prev-young} for the other age groups and details).
-    }
-    \label{SEIR:fig:prev-old}
-    \todo[inline]{Remake/reposition this plot so that it fits on a page.}
+    \label{SEIR:fig:prev}
 \end{figure}
@@
 This would better exploit one of the strengths of the CIS: the availability of data  on all age groups.
-In some regions, there appears to be a trade-off between the 11--16s and 16--25s, with the model misfitting in opposite directions (see \cref{SEIR:fig:prev-young}).
+In some regions, there appears to be a trade-off between under and overfitting in different age groups (see \cref{SEIR:fig:prev}).
+For example, in the North East, most of the data points in the $[11, 16)$ age group are below the median posterior estimate, while most of the data points in the $[16, 25)$ age group are above the median posterior estimate.
 This implies that the data would allow identification of a more flexible age-based model.
@@
 Many of these issues are clearly demonstrated by the estimates for the North East (see \cref{transmission:fig:compare-NE}).
-I previously observed the quality of the SEIR model's fit is variable between different age groups (see \cref{SEIR:fig:prev-young,SEIR:fig:prev-old}).
+I previously observed the quality of the SEIR model's fit is variable between different age groups (see \cref{SEIR:fig:prev,SEIR:fig:prev-appendix}).
 This is because the model is not flexible enough to capture the true age distribution of infections.
@@
   \subfile{transmission-appendix-phase-type}
+  \subfile{transmission-appendix-plots}
   \listoftodos

### 2024-03-20 SEIR/CIS/posterior_predictive.R

@@ -79,19 +79,19 @@ weekly_predictions = noisy_predict |>
     group_by(region, age, week) |>
     median_qi()
 
-create_prev_plot = function(age_groups, label) {
+create_prev_plot = function(regions, label) {
     plot = weekly_predictions |>
-        filter(age %in% age_groups) |>
+        filter(region %in% regions) |>
         ggplot(aes(week)) +
         geom_point(aes(y = p), size = 0.5) +
         geom_ribbon(aes(ymin = pred_p.lower, ymax = pred_p.upper), alpha = 0.1) +
         geom_lineribbon(
             aes(day, prevalence, ymin = prevalence.lower, ymax = prevalence.upper),
-            data = filter(prediction_intervals, age %in% age_groups),
+            data = filter(prediction_intervals, region %in% regions),
             alpha = 0.4,
             size = 0.2
         ) +
-        facet_grid(region ~ age) +
+        facet_grid(age ~ region) +
         standard_plot_theming() +
         labs(
             x = "Date",
@@ -99,7 +99,8 @@ create_prev_plot = function(age_groups, label) {
         ) +
         theme(
             legend.position = "none",
-            strip.text = element_text(size = 7)
+            strip.text = element_text(size = 7),
+            axis.text.x = element_text(angle = 45, hjust = 1)
         )
 
     ggsave(
@@ -108,17 +109,23 @@ create_prev_plot = function(age_groups, label) {
         ),
         plot = plot,
         width = 15,
-        height = 20,
+        height = 15,
         units = "cm",
         dpi = 300
     )
 }
 
 # Generate goodness-of-fit plots, separately for age groups due to size
-young_ages = unique(weekly_predictions$age)[1:3]
-old_ages = unique(weekly_predictions$age)[4:6]
-create_prev_plot(young_ages, "young")
-create_prev_plot(old_ages, "old")
+all_regions = unique(weekly_predictions$region)
+regions_in_main = c(
+    "East of England",
+    "North East",
+    "London",
+    "Yorkshire"
+)
+regions_in_appendix = all_regions[!all_regions %in% regions_in_main]
+create_prev_plot(regions_in_main, "main")
+create_prev_plot(regions_in_appendix, "appendix")
 
 # Generate posterior incidence plots
 p_incidence = prediction_intervals |>

### 2024-03-20 ATACCC.tex
fix display in nav of Hakki et al.
fix tex warnings

@@
 
-\section{\texorpdfstring{\textcite{hakkiOnset}}{Hakki \etal (2022)}} \label{ATACCC:sec:hakki}
+\section{\texorpdfstring{\textcite{hakkiOnset}}{Hakki et al. (2022)}} \label{ATACCC:sec:hakki}
 
@@
 
-\textcite{hakkiOnset} uses weakly informative priors, similar to those previously used~\autocite{singanayagamCommunity}, see \cref{ATACCC:table:hakki-priors}.
+\textcite{hakkiOnset} uses weakly informative priors, similar to \textcite{singanayagamCommunity}, see \cref{ATACCC:table:hakki-priors}.
 I reran \textcite{hakkiOnset}'s analysis using R~4.3.1~\autocite{R-4-3-1}.
@@
 A strength of the ATACCC data is its general population sample.
-This is unlike most estimates (see \cref{E-intro:sec:duration-SARS-CoV-2}) which are from hospitalised patients or healthcare workers.
+This is unlike most estimates (see \cref{E-intro:sec:previous-duration-estimates}) which are from hospitalised patients or healthcare workers.
 This means that, notwithstanding the issues highlighted in the previous paragraph, the estimates are likely to be generalisable to the general population.

### 2024-03-20 incidence-prevalence.tex
fix tex warnings

@@
 
-Robust, unbiased prevalence estimates already exist (\cref{E-biology-data:sec:cis}), and incidence estimates are the objective of this thesis.
+Robust, unbiased prevalence estimates already exist (\cref{E-intro:sec:cis}), and incidence estimates are the objective of this thesis.
 This chapter has shown that the key quantity relating these quantities is the distribution of the duration of RT-PCR positivity.
-However, no estimates of this quantity in the general population exist (see \cref{E-intro:sec:duration-SARS-CoV-2}).
+However, no estimates of this quantity in the general population exist (see \cref{E-intro:sec:previous-duration-estimates}).
 Therefore, I now turn to estimating the duration of RT-PCR positivity.

### 2024-03-20 introduction-comments.tex
fix tex warnings

@@
 They used a Gaussian process prior to smooth the incidence, applying the prior to either smooth the log-incidence or its derivative; smoothing of the incidence is required to stabilise the estimates~\autocite{brookmeyerBackcalculation}.
-\Textcite{abbottCISincidence} used \textcite{hellewellPCRSensitivity}'s estimates of the sensitivity of testing to inform the backcalculation, which are based on a longitudinal study in healthcare workers (see \cref{intro:sec:duration-SARS-CoV-2}).
+\Textcite{abbottCISincidence} used \textcite{hellewellPCRSensitivity}'s estimates of the sensitivity of testing to inform the backcalculation, which are based on a longitudinal study in healthcare workers (see \cref{intro:sec:previous-duration-estimates}).
 They fit to summary statistics of the CIS data, using a daily, independent Gaussian likelihood with parameters estimated from the posterior point estimate and 95\% credible interval published by the ONS.
@@
 \Textcite{nicholsonImproving} have the most sophisticated use of prevalence survey data.
-They allow an arbitrary distribution of the duration of positivity, based on the estimates of \textcite{hellewellPCRSensitivity} (see \cref{intro:sec:duration-SARS-CoV-2} for discussion of the \textcite{hellewellPCRSensitivity} study).
+They allow an arbitrary distribution of the duration of positivity, based on the estimates of \textcite{hellewellPCRSensitivity} (see \cref{intro:sec:previous-duration-estimates} for discussion of the \textcite{hellewellPCRSensitivity} study).
 Furthermore, they incorporate the sensitivity and specificity of the RT-PCR test into their model.
@@
 
-\subsection{SARS-CoV-2} \label{intro:sec:duration-SARS-CoV-2}
+\subsection{SARS-CoV-2} \label{intro:sec:previous-duration-estimates}
 

### 2024-03-20 ATACCC.tex
improve ATACCC plot

@@
 \begin{figure}
-  \thisfloatpagestyle{empty}
-  \vspace{-3cm}
   \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{ATACCC/fits}}
-  \captionsetup{width=0.8\paperwidth}
-  \captionof{figure}[Posterior predictive viral load fits]{Posterior predictive observed viral loads (median and 95\% CrI, lines and ribbon) for each individual, overlaid with their observed test results (dots).
+  \caption[Posterior predictive viral load fits]{Posterior predictive observed viral loads (median and 95\% CrI, lines and ribbon) for selected individuals, overlaid with their observed test results (dots).
 Posterior estimates are conditional on the result not being a false negative.

### 2024-03-20 ATACCC/goodness_fit.R

@@ -84,19 +84,27 @@ ggsave(
 posterior_summary = posterior_predictive |>
   group_by(i, day) |>
   median_qi(yrep) 
+all_individuals = unique(posterior_summary$i)
+selected_individuals = all_individuals[all_individuals > 8 & all_individuals %% 2 == 1]
 p_fits = posterior_summary |>
+  filter(i %in% selected_individuals) |>
   ggplot() +
   geom_lineribbon(
     aes(day, yrep, ymin = .lower, ymax = .upper),
     show.legend = FALSE
   ) +
-  geom_point(aes(day, y, colour = result), size = 0.7, data = tbl_data) +
-  facet_wrap(~i, ncol = 6) +
+  geom_point(
+    aes(day, y, colour = result),
+    size = 0.7,
+    data = tbl_data |>
+      filter(i %in% selected_individuals)
+  ) +
+  facet_wrap(~i, ncol = 5) +
   theme_trajectory()
 ggsave(
     filename = here::here("ATACCC/fits.pdf"),
     plot = p_fits,
-    height = 20,
+    height = 15,
     width = 15,
     units = "cm",
     dpi = 300

### 2024-03-20 biology-data.tex
fix tex warnings
clearer episodes fig

@@
 
+The crucial features of an infection episode are the start and end times of the infection.
+Ignoring misclassified test results, latent start time of infection episode $j$, $B_j$, can be bounded as between the day after the last negative test, $l_j^{(b)}$, and the day of the first positive test, $r_j^{(b)}$.
+Similarly, the latent end time of infection episode $i$, $E_j$, can be bounded as between the day of the last positive test, $l_j^{(e)}$, and the day before the following negative test, $r_j^{(e)}$.
+
 A sample of 500 randomly selected infection episodes (selected from the episodes used in the analyses in this thesis) have their test results shown in \cref{biology-data:fig:episodes}.
@@
   \thisfloatpagestyle{empty}
-  \vspace{-5cm}
-  \makebox[\textwidth][c]{\includegraphics[height=.7\paperheight]{biology-data/STATS13734/data_viz}}
+  \vspace{-1cm}
+  \makebox[\textwidth][c]{\includegraphics[height=.9\paperheight]{biology-data/STATS13734/data_viz}}
   \caption[CIS infection episodes]{%
     500 randomly selected infection episodes.
-    Shown are the negative tests before and after the outermost positive tests associated with the episode.
-    Red dots are negative tests, green dots are positive tests, and blue dots are void tests (where the result could not be determined; excluded from other analyses).
+    Includes all tests between $l_j^{(b)}$ (day 0) and $r_j^{(e)}$.
+    Colours give test result (red is negative, green is positive, and blue is void).
   }
+  \label{biology-data:fig:episodes}
 \end{figure}
-
-The crucial features of an infection episode are the start and end times of the infection.
-Ignoring misclassified test results, latent start time of infection episode $j$, $B_j$, can be bounded as between the day after the last negative test, $l_j^{(b)}$, and the day of the first positive test, $r_j^{(b)}$.
-Similarly, the latent end time of infection episode $i$, $E_j$, can be bounded as between the day of the last positive test, $l_j^{(e)}$, and the day before the following negative test, $r_j^{(e)}$.
 

### 2024-03-20 ATACCC.tex
clarify caption

@@
   \centering \includegraphics{ATACCC/mean_trajectories}
-  \caption[Median viral load trajectories]{Trajectory (posterior median and 95\% CrI) using the median parameter values, $\exp \vec{\mu}_{\phi,g}$, excluding the within-population variation.
+  \caption[Median viral load trajectories]{Trajectory (posterior median and 95\% CrI) using the population median parameter values, $\exp \vec{\mu}_{\phi,g}$, excluding the within-population variation.
 These assume $\tau_{\max,i}=0$, however modifying this simply translates the trajectories left or right.

### 2024-03-20 cis-perfect-testing.tex
Perf test figures

@@
   \centering \includegraphics{cis-perfect-testing/flat-prior}
-  \caption[Uninformative priors for the hazard]{The density of Beta(1, 1) flat prior and Beta(0.5, 0.5) uninformative priors (left) with the prior predictive survival time (right).
+  \caption[Uninformative priors for the hazard]{
+    (A) Density of Beta(1, 1) flat prior and Beta(0.5, 0.5) uninformative priors.
+    (B) Prior predictive survival time (mean and 95\% CrI).
     The implied prior on the survival time is very short when using either prior on the hazard at each time.
-\label{perf-test:fig:flat-prior}}
+  }
+ \label{perf-test:fig:flat-prior}
 \end{figure}
@@
   \centering \includegraphics{cis-perfect-testing/vague-prior}
-  \caption[Vague prior for the hazard]{The density of Beta(0.1, 1.9) vague prior (left) and the prior predictive survival time (right).
-The prior predictive survival time is very vague using a Beta(0.1, 1.9) prior.
+  \caption[Vague prior for the hazard]{
+    (A) Density of Beta(0.1, 1.9) vague prior.
+    (B) Prior predictive survival time (mean and 95\% CrI).
     Note the different x-axis to \cref{perf-test:fig:flat-prior}.
-\label{perf-test:fig:vague-prior}}
+  }
+  \label{perf-test:fig:vague-prior}
 \end{figure}
@@
   \centering \includegraphics{cis-perfect-testing/rw2-prior}
-  \caption[RW2 prior for the hazard]{The prior predictive distribution (mean and 95\% central interval) for the RW2 prior specified in the main text.
+  \caption[RW2 prior for the hazard]{The prior predictive survival time (mean and 95\% central interval) for the RW2 prior specified in the main text.
 The posterior mean estimate using ATACCC data in \cref{E-ATACCC} are shown for comparison.
@@
   \thisfloatpagestyle{empty}
-  \vspace{-2.5cm}
+  % \vspace{-1cm}
   \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{cis-perfect-testing/ataccc-approximation-hazard}}
-  \caption[Approximating the ATACCC posterior hazard as a logit-normal]{Comparison of the posterior estimate of the hazard from \cref{E-ATACCC} and its approximation with a logit-normal distribution (kernel density estimate smoothed) for the first 15 hazards.
+  \caption[Approximating the ATACCC posterior hazard as a logit-normal]{Comparison of the posterior estimate of the hazard from \cref{E-ATACCC} and its approximation with a logit-normal distribution (kernel density estimate smoothed) for hazards between day 4 and 15.
+The density for smaller hazards is concentrated very close to 0.
 \label{perf-test:fig:approximate-ATACCC-hazard}}
@@
 ATACCC is the posterior mean from \cref{E-ATACCC}'s analysis, CIS is Sarah Walker's analysis, and Combined is the combination of these used for the simulation.
-See main text for details of each analysis.
 \label{perf-test:fig:duration-dist}}

### 2024-03-20 cis-perfect-testing/priors.R

@@ -87,8 +87,10 @@ ataccc_posterior = readRDS(here::here("cisRuns-output/input_curves.rds")) |>
   filter(source == "ATACCC")
 
 p_flat_dens = beta_plot_dens(flat_params, flat_params) +
-  theme(legend.position = "none")
-p_flat_surv = beta_plot_surv(flat_params, flat_params, prior_len = 15)
+  theme(legend.position = "none") +
+  labs(subtitle = ("(A) Hazard density"))
+p_flat_surv = beta_plot_surv(flat_params, flat_params, prior_len = 15) +
+  labs(subtitle = ("(B) Survival"))
 p_flat = p_flat_dens + p_flat_surv
 ggsave(
   filename = "cis-perfect-testing/flat-prior.pdf",
@@ -100,9 +102,11 @@ ggsave(
 )
 
 p_vague_dens = beta_plot_dens(0.1, 1.9) +
-  theme(legend.position = "none")
+  theme(legend.position = "none") +
+  labs(subtitle = ("(A) Hazard density"))
 p_vague_surv = beta_plot_surv(0.1, 1.9, prior_len = 30) +
-  theme(legend.position = "none")
+  theme(legend.position = "none") +
+  labs(subtitle = ("(B) survival"))
 p_vague = p_vague_dens + p_vague_surv
 ggsave(
   filename = "cis-perfect-testing/vague-prior.pdf",

### 2024-03-20 cis-imperfect-testing.tex
Tweak imperf test figs / captions

@@
   \caption[Estimating survival using CIS data assuming perfect testing]{Estimates of the survival function using CIS data alongside the ATACCC-based estimates of \cref{E-perf-test}.
-CIS-based has more episodes ending in the first 14 days.
 \label{imperf-test:fig:problem-cis-estimates}}
@@
 \begin{figure}
-  \vspace{-2.5cm}
+  \thisfloatpagestyle{empty}
   \makebox[\textwidth][c]{\includegraphics{cis-imperfect-testing/CIS_vary}}
-  \caption{%
+  \caption[Sensitivity of CIS estimates to the prior.]{%
     (A-C) Sensitivity of CIS estimates to the value of $r$, the dispersion of the prior for $\ntot$ (a larger $r$ means a more informative prior) when $\psens = 0.8$.
@@
   \centering \includegraphics{cis-imperfect-testing/CIS_ntot}
-  \caption{How the posterior estimate of $\ntot$ changes with the value of $r$ in the prior on $\ntot$.}
+  \caption[Sensitivity of estimates of $\ntot$ to the prior.]{How the posterior estimate of $\ntot$ changes with the value of $r$ in the prior on $\ntot$.}
   \label{imperf-test:fig:ntot}

### 2024-03-20 SEIR/CIS/posterior_predictive.R
Tweak final plots / captions

@@ -109,7 +109,7 @@ create_prev_plot = function(regions, label) {
         ),
         plot = plot,
         width = 15,
-        height = 15,
+        height = 16,
         units = "cm",
         dpi = 300
     )

### 2024-03-20 cis-imperfect-testing.tex
Fix refs

@@
 
-Several mechanisms for episodes being undetected were considered when deriving $p_{iu}$ in \cref{E-perf-test:eq:piu}.
+Several mechanisms for episodes being undetected were considered when deriving $p_{iu}$ in \cref{perf-test:eq:piu}.
 I now consider the additional mechanisms arising due to false negatives.
@@
 \begin{enumerate}
-    \item The episode would have been detected considering only the mechanisms in \cref{E-perf-test:eq:piu}.
-That is $b_{j(i)} > \min(\sched_i)$ and $\tau_{\sched_i}(b_{j(i)}) + b_{j(i)} \leq e_{j(i)}$.
+    \item The episode would have been detected considering only the mechanisms in \cref{perf-test:eq:piu}. That is $b_{j(i)} > \min(\sched_i)$ and $\tau_{\sched_i}(b_{j(i)}) + b_{j(i)} \leq e_{j(i)}$.
     \item The episode ends in the interval $[\tau_{\sched_i}(b_{j(i)}) + b_{j(i)}, \tau^2_{\sched_i}(b_{j(i)}) + b_{j(i)}]$.
@@
 \end{align}
-Integrating over $b_{j(i)}$ for $\min \sched_i < b_{j(i)} \leq T$ ($T$ being the last day of the period considered), in the same way as \cref{E-perf-test:eq:piu}, gives:
+Integrating over $b_{j(i)}$ for $\min \sched_i < b_{j(i)} \leq T$ ($T$ being the last day of the period considered), in the same way as \cref{perf-test:eq:piu}, gives:
 \begin{align}

### 2024-03-20 paper-surv-analysis.tex
Fix refs

@@
 Inference was implemented in Stan via RStan~2.21.8~\autocite{rstan2-21-8} using default settings.
-Convergence was assessed using Rhat and ESS as described in \cref{E-inc-prev:sec:MCMC}, and all runs checked for divergent transitions.
+Convergence was assessed using Rhat and ESS as described in \cref{E-MCMC:sec:convergence}, and all runs checked for divergent transitions.
 

### 2024-03-20 transmission.tex
Tweak final plots / captions
Fix refs

@@
 First, it uses the statistical model from \cref{E-biology-data:sec:MRP} to estimate prevalence.
-Then, it transforms the posterior distribution of prevalence to a posterior distribution of incidence using the forward substitution as in \cref{E-inc-prev:eq:forward-substitute}.
+Then, it transforms the posterior distribution of prevalence to a posterior distribution of incidence using the forward substitution as in \cref{inc-prev:eq:forward-substitute}.
 At a high-level, the statistical model is:
@@
 The second step is to transform the posterior distribution for $\vec{P}$ to a posterior distribution for $\vec{Z}$.
-A posterior sample for $\vec{P}$ can be transformed to a posterior sample for $\vec{Z}$ based on \cref{E-inc-prev:eq:forward-substitute}, see \cref{backcalc:sec:methods} for details.
+A posterior sample for $\vec{P}$ can be transformed to a posterior sample for $\vec{Z}$ based on \cref{inc-prev:eq:forward-substitute}, see \cref{backcalc:sec:methods} for details.
 
@@
 Then, I transform this posterior sample using \cref{transmission:alg:calculate-Z}.
-The algorithm is based on \cref{E-inc-prev:eq:forward-substitute}, which is repeated here for convenience:
+The algorithm is based on \cref{inc-prev:eq:forward-substitute}, which is repeated here for convenience:
 \begin{align}
@@
 
-\Cref{transmission:eq:forward-substitute} is derived from \cref{E-inc-prev:eq:EPt-matrix}: $\vec{P} = \matr{S} \vec{I}$.
+\Cref{transmission:eq:forward-substitute} is derived from \cref{inc-prev:eq:EPt-matrix}: $\vec{P} = \matr{S} \vec{I}$.
 For this to have a unique solution, $\vec{I}$ and $\vec{P}$ need the same dimension.
@@
 % I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:observation-process} justifies, that is the estimate of the incidence is negligibly affected by approximating the distribution $\vec{P} \mid \vec{Z}$ by its mean.
-% I solve this system using forward substitution, calculating the value of $Z_t$ in \cref{E-inc-prev:eq:forward-substitute} sequentially for $t = 1, \dots, T$.
+% I solve this system using forward substitution, calculating the value of $Z_t$ in \cref{inc-prev:eq:forward-substitute} sequentially for $t = 1, \dots, T$.
 % $t = 1$ is defined as the first day where prevalence data is used.
-% \Cref{E-inc-prev:eq:forward-substitute} is repeated here for convenience:
+% \Cref{inc-prev:eq:forward-substitute} is repeated here for convenience:
 % \begin{align}
@@
 \begin{figure}
-    \includegraphics{SEIR/CIS/attack_rates}
+    \makebox[\textwidth][c]{\includegraphics{SEIR/CIS/attack_rates}}
     \caption{Posterior median and 95\% CrI of the attack rate in each region and age group.}
@@
 \begin{figure}
-    \thisfloatpagestyle{empty}
-    \vspace{-3cm}
+    % \thisfloatpagestyle{empty}
+    % \vspace{-3cm}
     \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_main}}

### 2024-03-21 thesis.tex
move half of plot to appendix
remove display of to dos
add toc and lof to bookmarks
Fix dates
add abstract

@@
 \usepackage{textcomp} % provide symbols
-\usepackage{todonotes}
+\usepackage[disable]{todonotes}
 \usepackage{xr}
@@
 %% Submission date :
-\submissiondate{March 2024}
+\newcommand{\textdate}{March 2024}
+\submissiondate{\textdate}
 
 %% You can redefine the submission notice [optional]:
+% For some reason the submission notice breaks if I just use \submissiondate
 \submissionnotice{%
     This thesis is submitted for the degree of \\ \emph{Doctor of Philosophy} \\
-    March 2024
+    \textdate
 }
@@
 \let\blx@noerroretextools\empty
+
+% add contents and list of figures to PDF bookmarks
+% See: https://tex.stackexchange.com/a/1821/95193
+\pretocmd{\tableofcontents}{%
+  \if@openright\cleardoublepage\else\clearpage\fi
+  \pdfbookmark[0]{\contentsname}{toc}%
+}{}{}%
+\pretocmd{\listoffigures}{%
+  \if@openright\cleardoublepage\else\clearpage\fi
+  \pdfbookmark[0]{\listfigurename}{lof}%
+}{}{}%
 \makeatother
@@
 \abstract{%
-    \todo[inline]{Write Abstract}
+    Surveillance is a crucial component to a pandemic response, informing public health and wider policy.
+    Unfortunately, it is often subject to reporting and behavioural biases, reducing its robustness.
+    % During the SARS-CoV-2 epidemic, the UK government funded two large-scale prevalence surveys with samples recruited randomly from the general population.
+    % One of these, the CIS (Coronavirus Infection Survey), was a longitudinal survey run from April 2020 to March 2023.
+    The CIS (Coronavirus Infection Survey) was a large-scale prevalence survey with participants recruited randomly from the general population, therefore avoiding these issues.
+
+    In this thesis, I consider how to best use the CIS data for surveillance.
+    %, can be used for surveillance to perform more robust inference.
+    In particular, I consider the estimation of incidence (the number of new infections) and transmission (where infections are occurring) in England.
+
+    This relies on knowledge of the duration of positivity (how long individuals test positive for).
+    Therefore, I first consider how to estimate the distribution of this duration in the general population.
+    I extend a pre-existing model of how an individual's viral load changes over time to estimate the duration of positivity.
+    % I take two complementary approaches using different datasets.
+    % This model is fitted to data from a longitudinal study of contacts of individuals with confirmed SARS-CoV-2 infection.
+    % In \cref{ATACCC}, I extend a pre-existing model of how an individual's viral load changes over time to estimate the duration of positivity.
+    % This model is fitted to data from the ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts) study, a longitudinal study of viral load in contacts of individuals with confirmed SARS-CoV-2 infection.
+    I use this detailed model to exploit frequent (daily) sampling of individuals in a cohort of contact traced individuals.
+    However, the cohort was small and follow-up short meaning that the tail of the duration distribution is poorly estimated.
+
+    % Therefore, in \cref{perf-test,imperf-test}, I combine the information from ATACCC with the larger CIS dataset.
+    % I analyse the CIS in a survival analysis framework.
+    Therefore, I combine the information with the larger and longer CIS dataset.
+    The CIS data poses several challenges because it is double interval censored and some individuals in the cohort have infections that are never detected.
+    I apply a survival analysis framework to analyse this data, which had previously only been in simpler settings.
+    In addition, I develop novel methodology to incorporate false negatives into the framework.
+    % The results showed that false negatives in the CIS data needed to be included in the model.
+    % In \cref{imperf-test} I develop a novel methodology to do so.
+    This results in an estimate of the full duration distribution in the general population, not previously available.
+
+    Finally, I use this estimate to estimate the incidence and transmission of SARS-CoV-2 in England.
+    I fully propagate the sampling error, accurately include the shape of the duration distribution, and estimate the quantities by age; previous work lacked at least one of these.
+    Furthermore, I fit to only the CIS data, showing that such a survey is adequate to estimate these quantities.
+    % without requiring an evidence synthesis approach previously used.
 }
@@
 \subfile{transmission-appendix-phase-type}
+\subfile{transmission-appendix-plots}
 

### 2024-03-21 biology-data.tex
add SRS details

@@
     \item $n_{taes}$ and $y_{taes}$ are the number of individuals tested and the number of positive test results on day $t$ in individuals of age group $a$, ethnicity $e$, and sex $s$.
-I use a beta-binomial distribution, as discussed in \cref{biology-data:sec:clustering}.
+I use a beta-binomial distribution for the likelihood, as discussed in \cref{biology-data:sec:clustering}.
     \item $\beta$ is a global intercept term with the vague prior $\beta \dist \N(0, 1/0.001^2)$.
@@
 \end{itemize}
-For parameters given vague priors, the large amount of data means that the prior has little influence on the posterior distribution.
+% For parameters given vague priors, the large amount of data means that the prior has little influence on the posterior distribution.
 The second-order random walk priors favours linear growth in $\logit(p_{taes})$ over time.
@@
 
+An approximation of the posterior distribution is computed using INLA (the Integrated Nested Laplace Approximation), see \cref{E-transmission:sec:INLA} for details.
 I fitted the model separately and independently for each region, implicitly adding all interactions between region and the other variables.
@@
 The final step is poststratifying the incidence proportion to the CIS target population and subpopulations of interest.
-To start, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution (using the procedure described in \cref{transmission:sec:INLA:posterior}).
+To start, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution.
 For each posterior draw and day, the incidence proportion in the (sub)population is the weighted average of the incidence proportion in each stratum, where the weights are the number of individuals in that stratum.
@@
 Analyses requiring more fine-grained stratification (such as that in \cref{biology-data:sec:MRP}) or linking data from the same individual over time (such as that in \cref{E-perf-test,E-imperf-test}) must be performed within the SRS.
-Therefore, computationally efficient approaches must be taken to this work.
-The SRS contains limited computational power.
-It has\todo{insert SRS computational power}.
-However, these are shared computational resources and due consideration must be given to other users.
+The SRS contains five 2.40\,GHz dual-core Intel Xeon E5-4650 processors and 164\,GB RAM.
+These computational resources are shared between all users of the CIS SRS project and due consideration must be given to other users.
 Processes which run for multiple days are liable to be cancelled.
 
+Therefore, computational approaches requiring high-performance computing, graphical processing units, or long runtimes are not feasible within the SRS.
 This computational limitation will feature in several places in this thesis.

### 2024-03-22 cis-perfect-testing.tex
Fix refs
Ch 5 Overleaf comments

@@
 Duration distributions tend to be right-skewed.
-Therefore, estimates of the mean are sensitive to the tail of the duration distribution.
+Therefore, estimates of the mean would be sensitive to the tail of the duration distribution.
 
 In \cref{E-ATACCC} I estimated the duration distribution from the ATACCC dataset.
-However, this study suffered from a small sample size and only 20 days of follow-up, enabling estimation of the tail.
+However, this study suffered from a small sample size and only 20 days of follow-up, preventing accurate estimation of the tail.
 Therefore, model assumptions drive the size of the estimated tail.
@@
 In this chapter, I develop a method to estimate the duration distribution using CIS data.
-The CIS sampling is less frequent, but with a large sample size and long follow-up.
+The CIS sampling is less frequent, but with has a large sample size (\numprint{4800} detected infection episodes) and a long follow-up.
 
-I take a Bayesian survival analysis approach to model the coarse sampling of the CIS (\cref{perf-test:sec:problem}).
-I use a simulation study (\cref{perf-test:sec:simulation-study}) to evaluate a range of priors for the survival distribution (\cref{perf-test:sec:parameters-priors}).
-These include a prior to combine the information in the CIS with the results of \cref{E-ATACCC}.
-Reasonable priors successfully recover the true distribution.
-However, the method must be able to account for false negatives, discussed in \cref{E-imperf-test}.
+In \cref{perf-test:sec:problem}, I describe the challenges analysing the data presents.
+This motivates using a survival analysis framework which I introduce in \cref{perf-test:sec:surv-analysis}, before deriving the statistical model for the CIS setting in \cref{perf-test:sec:model}.
+Next, in \cref{perf-test:sec:parameters-priors}, I consider how to parameterise the survival function and choose priors; these include vague and smoothing priors as well as including information from \cref{E-ATACCC}.
+I evaluate the model and priors using a simulation study (\cref{perf-test:sec:simulation-study}), showing that they perform well.
+Finally, I discuss the results in \cref{perf-test:sec:discussion}.
 
@@
 
-The CIS tests each individual infrequently, either weekly or monthly (see \cref{E-intro:sec:cis}).
-Infrequent testing complicates analysis of the data in two ways.
+The CIS tests each individual infrequently, either weekly or monthly (see \cref{E-intro:sec:cis}), complicating the data in two ways.
 First, infection episodes can go undetected.
 Second, the infection episode start and end times are known imprecisely.
-This section explains these issues. The rest of the chapter develops a method to tackle  them.
+I explain these issues below, and devote the rest of the chapter to the development of a method to tackle  them.
 
-\subsection{Undetected infections}
+\subsection{Undetected infections} \label{perf-test:sec:undetected}
 
@@
 Undetected infection episodes are episodes with no associated positive tests.
-Therefore, there is no information directly observed about the episode; even the number that occur is unobserved.
+Therefore, there is no information directly observed about the episode; even the number that occur is latent.
 They occur when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
-Undetected episodes are important because they are shorter than detected episodes.
-To see why this is the case, consider the undetected infection in the bottom of \cref{perf-test:fig:truncation}.
-The infection episode depicted begins at time 33 and has a 10-day duration.
-If the duration had been 24 days or more, the individual would still be detectable at their next test and detected then.
-CIS has many undetected infections.
+Undetected episodes are important because they are, on average, shorter than detected episodes; therefore, neglecting them would lead to a biased estimate of the survival distribution.
+To see why this is the case, consider the infection episode in figure \cref{perf-test:fig:truncation}.
+Here two individuals are shown, with identical data collected for each of them (a series of negative
+tests).
+The fist individual was never infected whereas the second individual was; the latter is an undetected infection. 
+The undetected infection episode begins at time 33 and has a 10-day duration.
+The individual would have been detected at their next test if their duration had instead been 24 days or more.
+The gap between tests in the CIS (up to 28 days, see \cref{E-intro:sec:cis}) is longer than most infections (median 16 days, see \cref{E-ATACCC:sec:results}).
+Therefore, it is likely that many infection in the CIS cohort were undetected.
 This is because the interval between tests in the same individual (up to 28 days, see \cref{E-intro:sec:cis}) is longer than most infections (median 16 days, see \cref{E-ATACCC:sec:results}).
 \begin{figure}
-  \makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{cis-perfect-testing/truncation}}
-  \caption[Undetected episodes in CIS data]{An unknown number of infection episodes in CIS are undetected.
-Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was.
-See the main text for why this is an issue.
-\label{perf-test:fig:truncation}}
+  \makebox[\textwidth][c]{\includegraphics[width=\textwidth]{cis-perfect-testing/truncation}}
+  \caption[Undetected episodes in CIS data]{Graphical illustration of an undetected infection episode, represented by the shaded area. \label{perf-test:fig:truncation}}
 \end{figure}
 
-\subsection{Unobserved episode beginning and end times} \label{perf-test:sec:interval-censoring}
+\subsection{Unobserved times for each episode's beginning and end} \label{perf-test:sec:interval-censoring}
 
-The second complication is that the exact beginning and end times of detected infection episodes are unknown.
-The imprecision in the beginning of the infection episode arises because the only observation that episode $j$ has started is that an individual was negative at one test, on day $l_j^{(b)}-1$ (the minus one is because the first day that the infection could have begun is the day after the negative test), and positive at the next test, on day $r_j^{(b)}$.
-Similarly, the end of the infection episode is bounded by the final positive test on day $l_j^{(e)}$ and the following negative test on day $r_j^{(e)}+1$ (the plus one is because the last day that the infection could have ended is the day before the negative test).
+The second complication is that the beginning and end times of detected infection episodes is latent.
+Recall, from \cref{E-biology-data:sec:cis-episodes}, that the time episode $j$ starts, $b_j$, is only known to lie in the interval $[l_j^{(b)}, r_j^{(b)}]$ where $l_j^{(b)}$ is the day after the negative test prior to the start of episode $j$ and $r_j^{(b)}$ is the first positive test in episode $j$.
+Similarly, for the end of the infection episode, $e_j \in [l_j^{(e)}, r_j^{(e)}]$.
 \Cref{perf-test:fig:double-interval-censor} shows this issue graphically.
 \begin{figure}
-  \centering \includegraphics{cis-perfect-testing/double-interval-censor}
-  \caption[Double-interval censoring in CIS data]{Episodes data in CIS is double interval censored, meaning that both the start and end of the episode are only known up to an interval.
-Demonstrated here by a participant who is recorded negative at time 7 to positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode.
-\label{perf-test:fig:double-interval-censor}}
+  \makebox[\textwidth][c]{\includegraphics[width=\textwidth]{cis-perfect-testing/double-interval-censor}}
+  \caption[Double-interval censoring in CIS data]{%
+        The beginning and end of each episode observed in the CIS data is only known up to an interval.
+        Demonstrated here by a participant who is recorded negative at time 7 and positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode.
+    }
+    \label{perf-test:fig:double-interval-censor}
 \end{figure}
 
-\subsection{Approach} \label{perf-test:sec:approach}
+\section{Survival analysis} \label{perf-test:sec:surv-analysis}
 
-I base my analysis on the 4,800 episodes first detected between 10th Oct 2020 and 6th Dec 2020 inclusive with negatives bounding the start and end time of the episode.
-That is, the individual tested negative before and after the episode.
-Without a negative before the episode there is no lower bound on the episode's start time.
-It could be arbitrarily long and there is little information on its length.
-A negative test follows each episode unless the individual was lost to follow-up before their episode ended.
-This is rare.
-Episodes removed in this step are incorporated in the adjustment for undetected episodes.
+In survival analysis, the interest is in estimating the distribution of the interval between an \emph{initiating event} and a \emph{terminating event}.
+In this context, the initiating event is the episode's start; the terminating event is the episode's end.
+I start by introducing the relevant, standard concepts in survival analysis; a wide range of texts cover them including \textcite[chapter 1]{bogaertsSurvival}.
 
-\section{Survival analysis}
-
-In survival analysis, the event at the start of the time of interest is the \emph{initiating event}, and the event at the end of the time of interest is the \emph{terminating event}.
-It is used in many areas, including how long a machine lasts before breaking down or how long a patient spends in hospital after a treatment.
-Here, I am estimating the distribution of the duration of positivity.
-The initiating event is the episode's start; the terminating event is the episode's end.
-
-Undetected events where there is no information on these events is known as \emph{truncation}.
-A situation that is particularly relevant is \emph{left truncation}, when only events lasting a minimum amount of time are observed.
-There is a subtlety that makes the CIS different from standard left truncation.
-In CIS, individuals in which undetected infections occurred are enrolled in the study.
-Therefore, some data has been collected about them.
-In standard left truncation, nothing is known about the missed events.
-Here, I will incorporate the additional information into the analysis.
-I refer to these events as ``undetected'', rather than truncated, to emphasise this difference.
-
-The time of the initiating and/or terminating events are not known exactly.
-This is known as \emph{censored} in the survival analysis literature.
-Censored events are present in almost all survival analyses.
+A common issue is that the initiating and/or terminating events are not known exactly.
+This is known as the event being \emph{censored}.
 An event is \emph{left censored} if it is known to occur before a certain time.
@@
 An event is \emph{interval censored} if it is known to occur within an interval.
-%Left and right censoring can be viewed as special cases of interval censoring where the lower bound is negative infinity or the upper bound is positive infinity respectively.
-As explained in \cref{perf-test:sec:problem}, the initiating and terminating events are only bounded, and hence interval censored.
-The data is \emph{doubly censored} because both the initiating and terminating events are censored.
+If both the initiating and terminating event are censored, the data are \emph{doubly censored}.
 To handle doubly censored data, the times need to be jointly modelled~\autocite[and references therein]{liSemiparametric}.
-In CIS data both the initiating and the terminating events are interval censored.
-Therefore, the CIS data is \emph{double interval censored}.
-\Textcite{sunAnalysis,bogaertsSurvival} review methods for double interval censored data.
+If both events are interval censored then the data are \emph{doubly interval censored}.
 
-Few studies look at the combination of both double interval censoring and undetected (or truncated) events.
-This is especially true within the human biostatistical literature.
-The literature does include theoretical frameworks which include the double censored and truncated case~\autocite{turnbullEmpirical,dempsterMaximum}, yet these frameworks have only been applied when the terminating event was either uncensored or right censored~\autocite{sunEmpirical,bacchettiNonparametric}.
-These studies \autocite[and elsewhere, e.g.:][]{shenNonparametric} take a conditional likelihood approach.
-They condition on the truncation (a selection effect) and the period where the initiating event occurs.
-They condition on the initiating event time because it provides negligible information on the distribution of interest~\citePersonalComms{Nick Jewell}.
-This is not the case for the CIS.
-In CIS, infection episodes are more likely to be detected when an individual is in the weekly testing phase.
-Therefore, the probability of an infection episode being detected conditional on being infected is higher than the overall probability of being detected.
+Sometimes, only intervals satisfying some property are recorded.
+In this case, data are \emph{truncated}.
+\emph{Left truncation} is when only events lasting longer than a minimum amount of time are observed.
+\emph{Right truncation} is when only events lasting shorter than a maximum amount of time are observed.
+Without correcting for the truncation, inference will be biased.
 
-Censored and truncated data are normal for estimation of the survival time of bird and insect nests~\autocite{heiseyABCs}.
-\textcite{heiseyModelling} developed a framework for handling this situation.
-The framework is general, allowing for arbitrary censoring patterns and double interval censoring.
-They apply it to a situation where the visit times (possible detection times) are global, simplifying the likelihood.
+In the CIS data, the initiating and terminating events are only bounded (see \cref{perf-test:sec:interval-censoring}).
+Therefore, the data are doubly interval censored.
+
+There is also undetected infections in the CIS (see \cref{perf-test:sec:undetected}).
+However, the situation is subtly different from the standard truncation situation encountered in the literature.
+In the standard situation, nothing is known about the unrecorded intervals.
+But in CIS, individuals in which undetected infections occurred are enrolled in the study.
+Therefore, we know the times that they were tested.
+This provides information on how long their infections may have been.
+I refer to these events as ``undetected'', rather than truncated, to emphasise this difference.
+
+\Textcite{sunAnalysis,bogaertsSurvival} review methods for doubly interval censored data.
+However, few studies look at data both doubly interval censored and with undetected (or truncated) events.
+% This is especially true within the biostatistical literature on human diseases.
+General theoretical frameworks, which include this case, have been developed~\autocite{turnbullEmpirical,dempsterMaximum}.
+Yet only the special case where the terminating event is either uncensored or right censored has been applied in practice~\autocite[e.g.][]{sunEmpirical,bacchettiNonparametric,shenNonparametric}.
+These studies base their inference on a likelihood contribution for each detected infection episode conditional on $b_j \in [l_j^{(b)}, r_j^{(b)}]$.
+
+For this conditional likelihood to be valid, they assume this conditioning provides negligible information on the distribution of interest~\citePersonalComms{Nick Jewell}.
+This assumption performs well in these settings, when the length of the intervals is fairly constant.
+In particular, when the interval lengths are near-constant, $p(\text{episode $j$ detected} \mid b_j \in [l_j^{(b)}, r_j^{(b)}]) \approx p(\text{episode $j$ detected})$.
+However, the interval lengths change from one to four weeks in the CIS (see \cref{E-intro:sec:cis}).
+
+Infection episodes in an individual's one week testing period are over-represented in the detected infections.
+This is because more frequent testing will detect a higher proportion of infections.
+Therefore, for the average detected episode, $p(\text{episode $j$ detected} \mid b_j \in [l_j^{(b)}, r_j^{(b)}]) > p(\text{episode $j$ detected})$.
+Therefore, conditioning on $b_j \in [l_j^{(b)}, r_j^{(b)}]$ is inappropriate.
+
+Doubly interval censored and truncated data are common when assessing the survival time of bird and insect nests~\autocite{heiseyABCs}.
+\textcite{heiseyModelling}, building on on long-standing methodology~\autocite{dempsterMaximum,turnbullEmpirical}, develops a likelihood-based inference framework, allowing for arbitrary truncation patterns and double interval censoring.
+They apply it to a situation where the visit times (possible detection times) are common to all the individuals, simplifying the likelihood.
 I adapt this framework for CIS, when each individual has their own visit schedule (see section \cref{perf-test:sec:model}).
@@
 I use Bayesian inference for two reasons.
-First, frequentist estimators are poorly developed and their properties (\eg consistency and convergence rate) are poorly understood~\autocite{sunAnalysis,dengNonparametric}.
+First, the frequentist properties (\eg consistency and convergence rate) of proposed estimators are poorly understood~\autocite{sunAnalysis,dengNonparametric}.
 Second, a prior can include information from previous studies, such as from \cref{E-ATACCC}.
-Alternatively, the prior can favour a smoothly changing hazard.
-I explore priors in section 1.4.
-These prior structures overcome issues caused by the undetected episodes ~\autocite{caoBias}.
+Alternatively, the prior can favour a smoothly changing hazard, which is biologically sensible, and can overcome issues caused by the undetected episodes ~\autocite{caoBias}.
 
 Bayesian methods have been used for nest survival before~\autocite{heBayesiana,heBayesian,caoModeling} by augmenting the data with the unobserved times of infection.
-The augmentation creates a parameter per detected infection.
-This increases the dimensionality of the problem greatly, making it computationally infeasible within the SRS.
-I developed a novel Bayesian framework without explicit data augmentation.
-This method scales to the large number of episodes present in CIS.
+The data augmentation approach creates a parameter for each detected infection.
+This increases the dimensionality of the problem greatly, making it computationally infeasible with the large sample size of the CIS within the SRS.
 
@@
 
-In this section, I develop a model accounting for double interval censoring and undetected episodes.
-This approach extends the framework of \textcite{heiseyModelling} to include individuals without detected episodes.
-This framework is itself based on long-standing methodology~\autocite{dempsterMaximum,turnbullEmpirical}.
-I augment the data set with the full number of episodes, including those that are undetected.
+In this section, I develop a statistical model accounting for double interval censoring and undetected episodes.
+This approach extends the framework of \textcite{heiseyModelling} to include all individuals in the CIS, including those without detected episodes.
 
-Integrating over the number of missed episodes is possible analytically.
-This makes inference tractable within modern, high-performance Bayesian software such as Stan.
+The number of missed episodes is now a latent quantity that needs to be considered.
+Marginalising over the number of missed episodes is possible analytically.
+This makes inference tractable using the large sample size and the computing resources available within the SRS (see \cref{E-biology-data:sec:SRS}).
+In particular, the posterior can be expressed within modern, high-performance Bayesian software such as Stan.
+
+The model will be used to analyse on the 4,800 episodes first detected between 10th Oct 2020 and 6th Dec 2020 inclusive, with negatives bounding the start and end time of the episode.
+A negative test before the start of the episode provides a lower bound on the episode's start time; otherwise, there is little information on its length.
+Similar information comes from the negative test after the end of the episode.
+This is typically observed, unless the individual is lost to follow-up before their episode is ended, which happens rarely.
+I will incorporate episodes excluded in this step in the model for undetected infections.
+
+\subsection{Types of episodes}
+
 Let the total number of infection episodes in the cohort be $\ntot = \ndet + \nnodet$ where $\ndet$ is the number of detected infections and $\nnodet$ is the number of undetected infections.
-Let $\Ncis$ be the number of individuals in the cohort and $\nsched$ be the number of test schedules.
+Let $\Ncis$ be the number of individuals in the cohort and $\nsched$ be the number of distinct test schedules.
+A test schedule, $\sched_i$, is the set of days an individual in the cohort is tested at.
+Each individual has exactly one test schedule but a test schedule could have multiple individuals associated with it.
+A test schedule, $\sched_i$ is the set of all times 
 %Without loss of generality, let the individuals with a detected episode be indexed by $i = 1, \dots, \ndet$ and the individuals without a detected episode be indexed by $i = \ndet + 1, \dots, \Ncis$.
@@
 
-All the relevant information about episode $i$ can be fully characterized by the (unobserved) triplet $(b_j, e_j, \sched_j)$ where $b_j$ and $e_j$ are the times the episode being and end respectively and $\sched_j$ is the set of test times for the individual in which the episode occurs.
-The episode's duration is $d_j = e_j - b_j + 1$.
+All the relevant information about the infection episode $j$ can be fully characterized by the (unobserved) triplet $(b_j, e_j, \sched_j)$ where $b_j$ and $e_j$ are the times the episode begin and end respectively and $\sched_j$ is the set of test times for the individual in which the episode occurs.
+The episode's duration is $d_j = e_j - b_j + 1$ (see sec 3.?).
 %, the number of days the individual in which the episode occurred, was positive for 
@@
 
-$b_j$ and $e_j$ span possible infection and recovery times.
+The times $b_j$ and $e_j$ span possible infection and recovery times.
 $\sched_j$ is one of the observed test schedules, with any other test schedule assumed to occur with probability zero.
-Over the short period of time considered, each unique test schedule has zero or one detected episodes.
 
+Over the short calendar period considered, each individual has zero or one detected episodes.
 Let $i \in \set{D}$ indicate $\sched_i$ is associated with a detected episode.
 That is, $i \in \set{D}$ if and only if there exists a $j(i)$ such that $b_{j(i)}$ and $e_{j(i)}$ such that $(b_{j(i)}, e_{j(i)}, \sched_i)$ is a detected episode. 
-$(b_{j(i)}, e_{j(i)}, \sched_i)$ is a detected episode when the following three conditions are met.
-First, $\exists t \in \sched_i$ such that $b_{j(i)} \leq t \leq e_{j(i)}$; this condition is equivalent to having at least one positive test for the episode.
-Second, $\exists t \in \sched_i$ such that $t < b_{j(i)}$, to lower bound the start of the episode.
-Finally, $\exists t \in \sched_i$ such that $t > e_{j(i)}$, to upper bound the end of the episode; this condition will be met for any episode sufficiently far in the past in individuals not lost to follow-up (almost always the case in this context).
+This happens
+$(b_{j(i)}, e_{j(i)}, \sched_i)$ is a detected episode when all the following conditions are met.
+\begin{enumerate}
+    \item $\exists t \in \sched_i$ such that $b_{j(i)} \leq t \leq e_{j(i)}$; this condition is equivalent to having at least one positive test for the episode.
+    \item $\exists t \in \sched_i$ such that $t < b_{j(i)}$, lower bounding the start of the episode.
+    \item $\exists t \in \sched_i$ such that $t > e_{j(i)}$, upper bounding the end of the episode.
+\end{enumerate}
 For a new context, including recent infections, this condition can be relaxed by considering these episodes as right censored, at little cost except a more complicated likelihood expression.
 
-To relate the state space to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie: are strict subsets of $T \times T$), for a test schedule $\sched_i$.
+To relate the possible value for $b_j$ and $e_j$, given a test schedule $\sched_j$, to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie: are strict subsets of $T \times T$).
 Figure \ref{perf-test:fig:partitionSpace} shows this graphically.
-Each episode, whether detected or not, must fall into one of these three classes with a probability that is a function of the model parameters, $\theta$.
+Each episode, whether detected or not, must fall into one of these three classes with a probability that is a function of the parameters of the survival function, $\theta$, discussed in ??.
 
@@
   This space is empty for $i \notin \set{D}$; in this case, $n_{ia} = 0$ admissible episodes occurred in individuals with test schedule $\sched_i$.
-  Otherwise, $i \in \set{D}$ then $n_{ia} =1$ admissible episode occurred in an individual with test schedule $\sched_{i}$.
-  Conditional on the episode occurring in individuals with test schedule $\sched_i$, it is admissible with probability $p_{ia} = \prob((b, e) \in \alpha_i \mid \theta)$.
+  Otherwise, if $i \in \set{D}$ then $n_{ia} =1$ admissible episode occurred in an individual with test schedule $\sched_{i}$.
+  Conditional on the episode occurring in individuals with test schedule $\sched_i$, the episode is admissible with probability $p_{ia} = \prob((b, e) \in \alpha_i \mid \theta)$.
   \label{perf-test:def:admissible}
 \item
-  Undetected episodes, $\Omega_i^C$, which have an infection and recovery time such that they would not have tested positive.
-  An unknown number, $n_{iu}$ of these observed in individuals with test schedule $\sched_i$.
-  Conditional on the episode occurring in individuals with test schedule $\sched_i$, it is undetected with probability $p_{iu} = \prob((b, e) \in \Omega^C_i \mid \theta)$.
+  Undetected episodes, $\Omega_i$, which have an infection and recovery time such that they would not have tested positive.
+  An unknown number, $n_{iu}$ of these occur in individuals with test schedule $\sched_i$.
+  Conditional on the episode occurring in individuals with test schedule $\sched_i$, the episode is undetected with probability $p_{iu} = \prob((b, e) \in \Omega_i \mid \theta)$.
 \item
@@
   detected.
-  These are all remaining episodes not in the previous sets.
-  $n_{i\bar{a}} = 0$ of these occurred in any individuals with test schedule $\sched_i$.
-  Conditional on the episode occurring in these individuals, it is inadmissible with probability $p_{i\bar{a}} = \prob((b, e) \in \beta_i \mid \theta)$.
+  These are all remaining episodes not in the previous sets, with $n_{i\bar{a}} = 0$ of these occurred in any individuals with test schedule $\sched_i$.
+  Conditional on the episode occurring in these individuals, the episode inadmissible with probability $p_{i\bar{a}} = \prob((b, e) \in \beta_i \mid \theta)$.
 \end{itemize}
-
-% The detected region, $\Omega_i = \alpha_i \cup \beta_i$, for test schedule $\sched_i$ is the region where we could have observed episodes. 
-% The notation is for consistency with \textcite{heiseyModelling}.
-% $\Omega_i$ is the complement of $\Omega^C_i$.
 
@@
 negative tests at times 7, 14, and 56, and positive tests at times 21
-and 28. Inadmissible regions are unshaded. The black (impossible) region is because it does not make sense for recovery to occur prior to infection. \label{perf-test:fig:partitionSpace}}
+and 28. Inadmissible regions are unshaded. The black (impossible) region covers the cases in which recovery occurs prior to infection. \label{perf-test:fig:partitionSpace}}
 \end{figure}
@@
 Therefore, conditional on $\ntot$, the number of episodes in each class, $(n_{ia}, n_{iu}, n_{i\bar{a}})^T$, are distributed multinomially.
-The above probabilities ($p_{iu}$, $p_{ia}$, and $p_{i\bar{a}}$) are conditional on the episode occurring with test schedule $\sched_i$.
-I multiply these probabilities by $\eta_i/\Ncis$, where $\eta_i$ is the number of individuals with test schedule $\sched_i$, to give the probability that an episode belongs to each class, without conditioning on a specific individual.
+with probabilities ($p_{iu}$, $p_{ia}$, and $p_{i\bar{a}}$), which are conditional on these episode occurring with test schedule $\sched_i$.
+By multiplying these probabilities by $\eta_i/\Ncis$, where $\eta_i$ is the number of individuals with test schedule $\sched_i$, the probability that an episode belongs to each class, without conditioning on a specific individual can be derived.
 
+{\color{red} Would this be the place to introduce a new subsection? on how to calculate the likelihood? Or on line 242?}
 The undetected classes for each individual cannot be distinguished by any observable data.
@@
 Furthermore, the inadmissible classes are guaranteed to have zero episodes.
-Counts of zero do not contribute to a multinomial pmf (they multiply by 1).
+Counts of zero do not contribute to the multinomial probability mass function (they multiply by 1).
 Therefore, I combine the undetected classes into a single class and ignore the inadmissible classes.
 
+{\color{red} subsection here instead of line 233?}
 Let $\na = (n_{1a}, \dots, n_{\nsched a})^T$ be the vector of counts of episodes in the admissible classes.
@@
 \alpha_i =& \{ (b, e) : l_{j(i)}^{(b)} \leq b \leq r_{j(i)}^{(b)} \wedge l_{j(i)}^{(e)} \leq e \leq r_{j(i)}^{(e)}\}.
-\intertext{Suppressing the conditioning on $\theta$, this gives:}
+\intertext{Omitting the conditioning on $\theta$, this gives:}
 p_{ia}
@@
 under the assumption of uniform probability of infection time.
+This is the standard form of the likelihood for doubly interval censored data with truncation.
 
 The next step is to derive $1 - p_u = 1 - \frac{1}{N} \sum_{i=1}^{\nsched} p_{iu} = \frac{1}{N} \sum_{i=1}^{\nsched} (1 - p_{iu})$.
-Therefore, the crucial component is $1 - p_{iu} = 1 - \prob((B_{j(i)}, E_{j(i)}) \in \Omega^C_i \mid \theta)$, one minus the probability that episode $i$ was undetected.
+Therefore, the crucial component is $1 - p_{iu} = 1 - \prob((B_{j(i)}, E_{j(i)}) \in \Omega_i \mid \theta)$, one minus the probability that episode $i$ was undetected.
 An episode is undetected if and only if no tests are performed during the episode or if there was no negative test prior to the episode.
@@
 \end{align}
-Then $\Omega^C_i$ can be written as:
+Then $\Omega_i$ can be written as:
 \begin{align}
-\Omega_i^C = \{ (b, e) \ssep \tau_{\sched_i}(b) + b > e \vee b \leq \min(\sched_i) \}.
+\Omega_i = \{ (b, e) \ssep \tau_{\sched_i}(b) + b > e \vee b \leq \min(\sched_i) \}.
 \end{align}
@@
 1 - p_{iu}
-&= 1 - \prob((B_{j(i)}, E_{j(i)}) \in \Omega^C_i) \\
+&= 1 - \prob((B_{j(i)}, E_{j(i)}) \in \Omega_i) \\
 &= 1 - \prob( \tau_{\sched_i}(B_{j(i)}) + B_{j(i)} > E_{j(i)} \vee B_{j(i)} \leq \min(\sched_i) ) \\
@@
 I implemented the inference in Stan via RStan~2.21.8~\autocite{rstan2-21-8} using default settings.
-Convergence was assessed using Rhat and ESS as described in \cref{E-inc-prev:sec:MCMC}, and all runs checked for divergent transitions.
+Convergence was assessed using Rhat and ESS as described in \cref{E-MCMC:sec:convergence}, and all runs checked for divergent transitions.
 

### 2024-03-22 cis-perfect-testing.tex
typo fix

@@
 Doubly interval censored and truncated data are common when assessing the survival time of bird and insect nests~\autocite{heiseyABCs}.
-\textcite{heiseyModelling}, building on on long-standing methodology~\autocite{dempsterMaximum,turnbullEmpirical}, develops a likelihood-based inference framework, allowing for arbitrary truncation patterns and double interval censoring.
+\textcite{heiseyModelling}, building on long-standing methodology~\autocite{dempsterMaximum,turnbullEmpirical}, develops a likelihood-based inference framework, allowing for arbitrary truncation patterns and double interval censoring.
 They apply it to a situation where the visit times (possible detection times) are common to all the individuals, simplifying the likelihood.

### 2024-03-22 cis-imperfect-testing.tex
Vector for theta

@@
 \begin{align}
-&p(\vec{y_{j(i)}}, e_{j(i)} \leq r_{j(i)}^{(e)} | b_{j(i)}, p_\text{sens}, \theta) \\
-&= p(\vec{y_{j(i)}}, l_{j(i)}^{(e)} \leq e_{j(i)} \leq r_{j(i)}^{(e)} | t_i, b_{j(i)}, p_\text{sens}, \theta) \\ % &\text{as no false positives}
-&= p(\vec{y_{j(i)}} \mid l_{j(i)}^{(e)} \leq e_{j(i)} \leq r_{j(i)}^{(e)}, t_i, b_{j(i)}, p_\text{sens}, \theta) p(l_{j(i)}^{(e)} \leq e_{j(i)} \leq r_{j(i)}^{(e)} | t_i, b_{j(i)}, p_\text{sens}, \theta) \\
-&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_{j(i)}(t)} (1 - p_\text{sens})^{(1 - y_{j(i)}(t))} \right) \left( S_\theta(l_{j(i)}^{(e)} - b_{j(i)} - 1) - S_\theta(r_{j(i)}^{(e)} - b_{j(i)} - 1) \right)
+&p(\vec{y_{j(i)}}, e_{j(i)} \leq r_{j(i)}^{(e)} | b_{j(i)}, p_\text{sens}, \vec{\theta}) \\
+&= p(\vec{y_{j(i)}}, l_{j(i)}^{(e)} \leq e_{j(i)} \leq r_{j(i)}^{(e)} | t_i, b_{j(i)}, p_\text{sens}, \vec{\theta}) \\ % &\text{as no false positives}
+&= p(\vec{y_{j(i)}} \mid l_{j(i)}^{(e)} \leq e_{j(i)} \leq r_{j(i)}^{(e)}, t_i, b_{j(i)}, p_\text{sens}, \vec{\theta}) p(l_{j(i)}^{(e)} \leq e_{j(i)} \leq r_{j(i)}^{(e)} | t_i, b_{j(i)}, p_\text{sens}, \vec{\theta}) \\
+&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_{j(i)}(t)} (1 - p_\text{sens})^{(1 - y_{j(i)}(t))} \right) \left( S_{\vec{\theta}}(l_{j(i)}^{(e)} - b_{j(i)} - 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b_{j(i)} - 1) \right)
 \label{imperf-test:eq:ll-ei-lt-ri}
@@
 \begin{align}
-&p(\vec{y_{j(i)}}, e_{j(i)} > r_{j(i)}^{(e)} | b_{j(i)}, p_\text{sens}, \theta) \\
-&= p(\vec{y_{j(i)}} \mid e_{j(i)} > r_{j(i)}^{(e)}, t_i, b_{j(i)}, p_\text{sens}, \theta) p(e_{j(i)} > r_{j(i)}^{(e)} | t_i, b_{j(i)}, p_\text{sens}, \theta) \\
-&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_{j(i)}(t)} (1 - p_\text{sens})^{(1 - y_{j(i)}(t))} \right) (1 - p_\text{sens}) S_\theta(r_{j(i)}^{(e)} - b_{j(i)} - 1)
+&p(\vec{y_{j(i)}}, e_{j(i)} > r_{j(i)}^{(e)} | b_{j(i)}, p_\text{sens}, \vec{\theta}) \\
+&= p(\vec{y_{j(i)}} \mid e_{j(i)} > r_{j(i)}^{(e)}, t_i, b_{j(i)}, p_\text{sens}, \vec{\theta}) p(e_{j(i)} > r_{j(i)}^{(e)} | t_i, b_{j(i)}, p_\text{sens}, \vec{\theta}) \\
+&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_{j(i)}(t)} (1 - p_\text{sens})^{(1 - y_{j(i)}(t))} \right) (1 - p_\text{sens}) S_{\vec{\theta}}(r_{j(i)}^{(e)} - b_{j(i)} - 1)
 \label{imperf-test:eq:ll-ei-gt-ri}
@@
 p_{ia}'
-&= p(\vec{y_{j(i)}} \mid p_\text{sens}, \theta) \\
-&= \sum_{b_{j(i)} = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( p(\vec{y_{j(i)}}, e_{j(i)} \leq r_{j(i)}^{(e)} \mid b_{j(i)}, p_\text{sens}, \theta) + p(\vec{y_{j(i)}}, e_{j(i)} > r_{j(i)}^{(e)} \mid b_{j(i)}, p_\text{sens}, \theta) \right) p(b_{j(i)} \mid p_\text{sens}, \theta).
+&= p(\vec{y_{j(i)}} \mid p_\text{sens}, \vec{\theta}) \\
+&= \sum_{b_{j(i)} = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( p(\vec{y_{j(i)}}, e_{j(i)} \leq r_{j(i)}^{(e)} \mid b_{j(i)}, p_\text{sens}, \vec{\theta}) + p(\vec{y_{j(i)}}, e_{j(i)} > r_{j(i)}^{(e)} \mid b_{j(i)}, p_\text{sens}, \vec{\theta}) \right) p(b_{j(i)} \mid p_\text{sens}, \vec{\theta}).
 \intertext{Now, substitute in \cref{imperf-test:eq:ll-ei-lt-ri,imperf-test:eq:ll-ei-gt-ri}, then take out the common factor:}
-&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_{j(i)}(t)} (1 - p_\text{sens})^{(1 - y_{j(i)}(t))} \right) \\ & \ \times \sum_{b_{j(i)} = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_\theta(l_{j(i)}^{(e)} - b_{j(i)} - 1) - S_\theta(r_{j(i)}^{(e)} - b_{j(i)} - 1) + (1 - p_\text{sens}) S_\theta(r_{j(i)}^{(e)} - b_{j(i)} - 1) \right) \\ & \ \times p(b_{j(i)} \mid p_\text{sens}, \theta) \\
-&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_{j(i)}(t)} (1 - p_\text{sens})^{(1 - y_{j(i)}(t))} \right) \\ & \ \times \sum_{b_{j(i)} = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_\theta(l_{j(i)}^{(e)} - b_{j(i)} - 1) - p_\text{sens} S_\theta(r_{j(i)}^{(e)} - b_{j(i)} - 1) \right) p(b_{j(i)} \mid p_\text{sens}, \theta).
+&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_{j(i)}(t)} (1 - p_\text{sens})^{(1 - y_{j(i)}(t))} \right) \\ & \ \times \sum_{b_{j(i)} = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_{j(i)}^{(e)} - b_{j(i)} - 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b_{j(i)} - 1) + (1 - p_\text{sens}) S_{\vec{\theta}}(r_{j(i)}^{(e)} - b_{j(i)} - 1) \right) \\ & \ \times p(b_{j(i)} \mid p_\text{sens}, \vec{\theta}) \\
+&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_{j(i)}(t)} (1 - p_\text{sens})^{(1 - y_{j(i)}(t))} \right) \\ & \ \times \sum_{b_{j(i)} = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_{j(i)}^{(e)} - b_{j(i)} - 1) - p_\text{sens} S_{\vec{\theta}}(r_{j(i)}^{(e)} - b_{j(i)} - 1) \right) p(b_{j(i)} \mid p_\text{sens}, \vec{\theta}).
 \label{imperf-test:eq:pia-prime}
@@
 
-If $\psens$ is fixed (\ie has a point prior) and $p(b_{j(i)} \mid \psens, \theta) \propto 1$ (as assumed in \cref{E-perf-test}) then:
+If $\psens$ is fixed (\ie has a point prior) and $p(b_{j(i)} \mid \psens, \vec{\theta}) \propto 1$ (as assumed in \cref{E-perf-test}) then:
 \begin{align}
 p_{ia}'
-&\propto \sum_{b_{j(i)} = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} S_\theta(l_{j(i)}^{(e)} - b_{j(i)} - 1) - p_\text{sens} S_\theta(r_{j(i)}^{(e)} - b_{j(i)} - 1).
+&\propto \sum_{b_{j(i)} = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} S_{\vec{\theta}}(l_{j(i)}^{(e)} - b_{j(i)} - 1) - p_\text{sens} S_{\vec{\theta}}(r_{j(i)}^{(e)} - b_{j(i)} - 1).
 \label{imperf-test:eq:pia-prime-constant}
@@
     \mid b_{j(i)},
-\theta \right) (1 - \psens) \\
-&= \left( S_\theta(\tau_{\sched_i}(b_{j(i)}) + 1) - S_\theta(\tau^2_{\sched_i}(b_{j(i)}) + 1) \right) (1 - \psens).
+\vec{\theta} \right) (1 - \psens) \\
+&= \left( S_{\vec{\theta}}(\tau_{\sched_i}(b_{j(i)}) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_i}(b_{j(i)}) + 1) \right) (1 - \psens).
 \end{align}
@@
 \begin{align}
-\eta = (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_i) + 1}^T \left( S_\theta(\tau_{\sched_i}(b) + 1) - S_\theta(\tau^2_{\sched_i}(b) + 1) \right).
+\eta = (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_i) + 1}^T \left( S_{\vec{\theta}}(\tau_{\sched_i}(b) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_i}(b) + 1) \right).
 \end{align}
@@
 1 - p_{iu}'
-&= 1 - p_{iu} - (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_i) + 1}^T \left( S_\theta(\tau_{\sched_i}(b) + 1) - S_\theta(\tau^2_{\sched_i}(b) + 1) \right) \\
-&= \frac{1}{T} \sum_{b=\min(\sched_i) + 1}^T \left( p_\text{sens} S_\theta(\tau_{\sched_i}(b) + 1) + (1 - p_\text{sens}) S_\theta(\tau^2_{\sched_i}(b) + 1)\right).
+&= 1 - p_{iu} - (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_i) + 1}^T \left( S_{\vec{\theta}}(\tau_{\sched_i}(b) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_i}(b) + 1) \right) \\
+&= \frac{1}{T} \sum_{b=\min(\sched_i) + 1}^T \left( p_\text{sens} S_{\vec{\theta}}(\tau_{\sched_i}(b) + 1) + (1 - p_\text{sens}) S_{\vec{\theta}}(\tau^2_{\sched_i}(b) + 1)\right).
 \label{imperf-test:eq:pit-prime}
@@
     A and D: median survival time, in comparison to the ATACCC-based estimate of \cref{E-ATACCC} (shown in red).
-    B and E: survival at day 50, $S_\theta(50)$.
+    B and E: survival at day 50, $S_{\vec{\theta}}(50)$.
     C and F: full survival curves out to day 100.
@@
 
-Estimating a time-varying $\psens$ with $S_\theta$ jointly may cause identifiability issues.
+Estimating a time-varying $\psens$ with $S_{\vec{\theta}}$ jointly may cause identifiability issues.
 Previous studies have avoided issues by including external information (such as a prior giving the magnitude) on $\psens$, or test results from later follow-up~\autocite[and references therein]{piresIntervalMisclassify}.

### 2024-03-22 cis-perfect-testing.tex
Vector for theta

@@
 %   of the time of the infection. Define the survival function
-%   $\prob(D_i \geq t \mid B_{j(i)} = b, \theta) = \prob(D_i \geq t \mid \theta) = S_\theta(t)$,
-%   where $\theta$ are the parameters controlling the survival
+%   $\prob(D_i \geq t \mid B_{j(i)} = b, \vec{\theta}) = \prob(D_i \geq t \mid \vec{\theta}) = S_{\vec{\theta}}(t)$,
+%   where $\vec{\theta}$ are the parameters controlling the survival
 %   distribution (the discussion in this document is valid regardless of
-%   the model specified for $S_\theta$ and hence we consider $\theta$
+%   the model specified for $S_{\vec{\theta}}$ and hence we consider $\vec{\theta}$
 %   as an arbitrary vector of parameters).
@@
   Otherwise, if $i \in \set{D}$ then $n_{ia} =1$ admissible episode occurred in an individual with test schedule $\sched_{i}$.
-  Conditional on the episode occurring in individuals with test schedule $\sched_i$, the episode is admissible with probability $p_{ia} = \prob((b, e) \in \alpha_i \mid \theta)$.
+  Conditional on the episode occurring in individual $i$, the episode is admissible with probability $p_{ia} = \prob((b, e) \in \alpha_i \mid \vec{\theta})$.
   \label{perf-test:def:admissible}
@@
   An unknown number, $n_{iu}$ of these occur in individuals with test schedule $\sched_i$.
-  Conditional on the episode occurring in individuals with test schedule $\sched_i$, the episode is undetected with probability $p_{iu} = \prob((b, e) \in \Omega_i \mid \theta)$.
+  Conditional on the episode occurring in individuals with test schedule $\sched_i$, the episode is undetected with probability $p_{iu} = \prob((b, e) \in \Omega_i \mid \vec{\theta})$.
 \item
@@
   These are all remaining episodes not in the previous sets, with $n_{i\bar{a}} = 0$ of these occurred in any individuals with test schedule $\sched_i$.
-  Conditional on the episode occurring in these individuals, the episode inadmissible with probability $p_{i\bar{a}} = \prob((b, e) \in \beta_i \mid \theta)$.
+  Conditional on the episode occurring in these individuals, the episode inadmissible with probability $p_{i\bar{a}} = \prob((b, e) \in \beta_i \mid \vec{\theta})$.
 \end{itemize}
@@
 \begin{align}
-p(\theta \mid \na)
-&\propto p(\theta) \left( \prod_{i \in \set{D}} p_{ia} \right) (r + \mu (1- \pnodet))^{-(r+\ndet)} \label{perf-test:eq:full-posterior-simplified}.
+p(\vec{\theta} \mid \na)
+&\propto p(\vec{\theta}) \left( \prod_{i \in \set{D}} p_{ia} \right) (r + \mu (1- \pnodet))^{-(r+\ndet)} \label{perf-test:eq:full-posterior-simplified}.
 \end{align}
@@
 % $\ntot$ is generally a nuisance parameter, however, its posterior distribution can be useful for diagnostic purposes (as in X\todo{ref where I use this in the next chapter}).
-% Its posterior can be reconstructed using a posterior sample of $\theta$.
-% For each posterior sample of $\theta$, sample from the full conditional of $\ntot$ to give the joint posterior of $\ntot$ and $\theta$.
+% Its posterior can be reconstructed using a posterior sample of $\vec{\theta}$.
+% For each posterior sample of $\vec{\theta}$, sample from the full conditional of $\ntot$ to give the joint posterior of $\ntot$ and $\vec{\theta}$.
 % The full conditional of $\ntot$ is given by:
 % \begin{align}
-% p(n_\text{tot} \mid \ncis, \theta)
-% &\propto p(\ntot \mid \theta) p(\ncis \mid \theta, \ntot) \\
-% &\propto p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} &\text{by \cref{perf-test:eq:augmented-likelihood}} \\
+% p(n_\text{tot} \mid \ncis, \vec{\theta})
+% &\propto p(\ntot \mid \vec{\theta}) p(\ncis \mid \vec{\theta}, \ntot) \\
+% &\propto p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} &\text{by \cref{perf-test:eq:augmented-likelihood}} \\
 % &\propto \frac{\Gamma(r + \ntot)}{\ntot!} \left( \frac{\mu}{r + \mu} \right)^{\ntot} \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot} \\
@@
 \begin{align}
-p_{ia} =& \prob((b, e) \in \alpha_i \mid \theta) \\
+p_{ia} =& \prob((b, e) \in \alpha_i \mid \vec{\theta}) \\
 \alpha_i =& \{ (b, e) : l_{j(i)}^{(b)} \leq b \leq r_{j(i)}^{(b)} \wedge l_{j(i)}^{(e)} \leq e \leq r_{j(i)}^{(e)}\}.
-\intertext{Omitting the conditioning on $\theta$, this gives:}
+\intertext{Omitting the conditioning on $\vec{\theta}$, this gives:}
 p_{ia}
@@
 =& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \prob \left( l_{j(i)}^{(e)} - b + 1 \leq D_{j(i)} \leq r_{j(i)}^{(e)} - b + 1 \right) \prob \left(B_{j(i)} = b \right) &\text{by def of $D_{j(i)}$} \\
-=& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_\theta(l_{j(i)}^{(e)} - b + 1) - S_\theta(r_{j(i)}^{(e)} - b + 1) \right) \prob \left(B_{j(i)} = b \right) &\text{by def of $S_\theta$} \\
-\propto& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_\theta(l_{j(i)}^{(e)} - b + 1) - S_\theta(r_{j(i)}^{(e)} - b + 1) \right)
+=& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_{j(i)}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 1) \right) \prob \left(B_{j(i)} = b \right) &\text{by def of $S_{\vec{\theta}}$} \\
+\propto& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_{j(i)}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 1) \right)
 \end{align}
@@
 The next step is to derive $1 - p_u = 1 - \frac{1}{N} \sum_{i=1}^{\nsched} p_{iu} = \frac{1}{N} \sum_{i=1}^{\nsched} (1 - p_{iu})$.
-Therefore, the crucial component is $1 - p_{iu} = 1 - \prob((B_{j(i)}, E_{j(i)}) \in \Omega_i \mid \theta)$, one minus the probability that episode $i$ was undetected.
+Therefore, the crucial component is $1 - p_{iu} = 1 - \prob((B_{j(i)}, E_{j(i)}) \in \Omega_i \mid \vec{\theta})$, one minus the probability that episode $i$ was undetected.
 An episode is undetected if and only if no tests are performed during the episode or if there was no negative test prior to the episode.
@@
 \end{align}
-Therefore, suppressing the conditioning on $\theta$:
+Therefore, suppressing the conditioning on $\vec{\theta}$:
 \begin{align}
@@
 &= 1 - \sum_{b={\min\sched_i} + 1}^T \prob( \tau_{\sched_i}(b) + 1> D_i) \prob(B_{j(i)} = b) - \sum_{t=1}^{\min\sched_i} \prob(B_{j(i)} = b)\\
-&= 1 - \frac{1}{T} \sum_{b=\min(\sched_i)+1}^T (1 - S_\theta(\tau_{\sched_i}(b) + 1)) - \frac{1}{T} \min\sched_i \\
-&= \frac{1}{T} \left(T - \sum_{b=\min\sched_i+1}^T (1 - S_\theta(\tau_{\sched_i}(b) + 1)) - \min\sched_i \right) \\
-&= \frac{1}{T} \left(T - (T - \min\sched_i) + \sum_{b=\min\sched_i+1}^T S_\theta(\tau_{\sched_i}(b + 1)) - \min\sched_i \right) \\
-&= \frac{1}{T} \sum_{b=\min(\sched_i)+1}^T S_\theta(\tau_{\sched_i}(b) + 1). \label{perf-test:eq:piu}
+&= 1 - \frac{1}{T} \sum_{b=\min(\sched_i)+1}^T (1 - S_{\vec{\theta}}(\tau_{\sched_i}(b) + 1)) - \frac{1}{T} \min\sched_i \\
+&= \frac{1}{T} \left(T - \sum_{b=\min\sched_i+1}^T (1 - S_{\vec{\theta}}(\tau_{\sched_i}(b) + 1)) - \min\sched_i \right) \\
+&= \frac{1}{T} \left(T - (T - \min\sched_i) + \sum_{b=\min\sched_i+1}^T S_{\vec{\theta}}(\tau_{\sched_i}(b + 1)) - \min\sched_i \right) \\
+&= \frac{1}{T} \sum_{b=\min(\sched_i)+1}^T S_{\vec{\theta}}(\tau_{\sched_i}(b) + 1). \label{perf-test:eq:piu}
 \end{align}
@@
 
-In this section I specify the form of $S_\theta(t)$.
+In this section I specify the form of $S_{\vec{\theta}}(t)$.
 % As frequently observed in the literature~\autocite[e.g.][]{heBayesian}, it is more convenient to operate with the hazard than the survival or probability mass function directly.

### 2024-03-22 cis-perfect-testing.tex
restructure chapter

@@
 %   of the time of the infection. Define the survival function
-%   $\prob(D_i \geq t \mid B_{j(i)} = b, \vec{\theta}) = \prob(D_i \geq t \mid \vec{\theta}) = S_{\vec{\theta}}(t)$,
+%   $\prob(D_i \geq t \mid B_{j} = b, \vec{\theta}) = \prob(D_i \geq t \mid \vec{\theta}) = S_{\vec{\theta}}(t)$,
 %   where $\vec{\theta}$ are the parameters controlling the survival
@@
 %   The beginning of episode $i$ is known to occur in the interval
-%   $[l_{j(i)}^{(b)}, r_{j(i)}^{(b)}]$, and similarly for the end of the infection
-%   in $[l_{j(i)}^{(e)}, r_{j(i)}^{(e)}]$.
+%   $[l_{j}^{(b)}, r_{j}^{(b)}]$, and similarly for the end of the infection
+%   in $[l_{j}^{(e)}, r_{j}^{(e)}]$.
 % \end{itemize}
@@
 
-In this chapter, I develop a method to estimate the duration distribution using CIS data.
 The CIS sampling is less frequent, but with has a large sample size (\numprint{4800} detected infection episodes) and a long follow-up.
+Therefore, it is suitable for estimating the tail of the duration distribution.
+In this chapter, I develop a method to estimate the duration distribution using CIS data, assuming no false negatives or positives.
 
@@
 
-The model will be used to analyse on the 4,800 episodes first detected between 10th Oct 2020 and 6th Dec 2020 inclusive, with negatives bounding the start and end time of the episode.
+The model will be used to analyse the 4,800 episodes first detected between 10th Oct 2020 and 6th Dec 2020 inclusive, with negatives bounding the start and end time of the episode.
+This period is denoted as $t = 1, \dots, T$.
 A negative test before the start of the episode provides a lower bound on the episode's start time; otherwise, there is little information on its length.
@@
 This is typically observed, unless the individual is lost to follow-up before their episode is ended, which happens rarely.
-I will incorporate episodes excluded in this step in the model for undetected infections.
 
-\subsection{Types of episodes}
+\subsection{Types of episodes} \label{perf-test:sec:episode-types}
 
 Let the total number of infection episodes in the cohort be $\ntot = \ndet + \nnodet$ where $\ndet$ is the number of detected infections and $\nnodet$ is the number of undetected infections.
-Let $\Ncis$ be the number of individuals in the cohort and $\nsched$ be the number of distinct test schedules.
-A test schedule, $\sched_i$, is the set of days an individual in the cohort is tested at.
-Each individual has exactly one test schedule but a test schedule could have multiple individuals associated with it.
-A test schedule, $\sched_i$ is the set of all times 
+Let $\Ncis$ be the number of individuals in the cohort.
 %Without loss of generality, let the individuals with a detected episode be indexed by $i = 1, \dots, \ndet$ and the individuals without a detected episode be indexed by $i = \ndet + 1, \dots, \Ncis$.
-Let $i = 1, \dots, \nsched$ index infection schedules and $j = 1, \dots \ntot$ index infection episodes.
+Let $i = 1, \dots, \Ncis$ index individuals and $j = 1, \dots \ntot$ index infection episodes.
+The test schedule $\sched_i$ is the set of times individual $i$ is tested at, from their last test prior to time 1 (if it exists) up until time $T$.
+This is all the times the infection episode could be detected at and be included, plus one prior test.
+For example, for the individual in \cref{perf-test:sec:undetected}, $\sched_i = \{ 0, 7, 14, 21, 28, 56 \}$.
+Each individual has exactly one test schedule, but a test schedule could have multiple individuals associated with it.
 
-All the relevant information about the infection episode $j$ can be fully characterized by the (unobserved) triplet $(b_j, e_j, \sched_j)$ where $b_j$ and $e_j$ are the times the episode begin and end respectively and $\sched_j$ is the set of test times for the individual in which the episode occurs.
-The episode's duration is $d_j = e_j - b_j + 1$ (see sec 3.?).
+All the relevant information about the infection episode $j$ can be fully characterized by the (unobserved) triplet $(b_j, e_j, i(j))$ where $b_j$ and $e_j$ are the times the episode begin and end respectively and $i(j)$ denotes the individual the episode occurred in.
+The episode's duration is $d_j = e_j - b_j + 1$ (see \cref{E-inc-prev:sec:definitions}).
 %, the number of days the individual in which the episode occurred, was positive for 
-This triplet is assumed iid.
+% This triplet is assumed iid.
 
-The times $b_j$ and $e_j$ span possible infection and recovery times.
-$\sched_j$ is one of the observed test schedules, with any other test schedule assumed to occur with probability zero.
+% The times $b_j$ and $e_j$ span possible infection and recovery times.
+% $\sched_j$ is one of the observed test schedules, with any other test schedule assumed to occur with probability zero.
 
 Over the short calendar period considered, each individual has zero or one detected episodes.
-Let $i \in \set{D}$ indicate $\sched_i$ is associated with a detected episode.
-That is, $i \in \set{D}$ if and only if there exists a $j(i)$ such that $b_{j(i)}$ and $e_{j(i)}$ such that $(b_{j(i)}, e_{j(i)}, \sched_i)$ is a detected episode. 
-This happens
-$(b_{j(i)}, e_{j(i)}, \sched_i)$ is a detected episode when all the following conditions are met.
-\begin{enumerate}
-    \item $\exists t \in \sched_i$ such that $b_{j(i)} \leq t \leq e_{j(i)}$; this condition is equivalent to having at least one positive test for the episode.
-    \item $\exists t \in \sched_i$ such that $t < b_{j(i)}$, lower bounding the start of the episode.
-    \item $\exists t \in \sched_i$ such that $t > e_{j(i)}$, upper bounding the end of the episode.
-\end{enumerate}
-For a new context, including recent infections, this condition can be relaxed by considering these episodes as right censored, at little cost except a more complicated likelihood expression.
+Let $i \in \set{D}$ indicate individual $i$ has a detected episode.
+I define the precise conditions for this to occur in \cref{perf-test:sec:prob-undetected}.
+% That is, $i \in \set{D}$ if and only if there exists a $j$ such that $b_{j}$ and $e_{j}$ such that $(b_{j}, e_{j}, \sched_i)$ is a detected episode.
 
-To relate the possible value for $b_j$ and $e_j$, given a test schedule $\sched_j$, to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie: are strict subsets of $T \times T$).
+To relate the possible value for $b_j$ and $e_j$, given a test schedule $\sched_{i(j)}$, to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie: are strict subsets of $T \times T$).
 Figure \ref{perf-test:fig:partitionSpace} shows this graphically.
-Each episode, whether detected or not, must fall into one of these three classes with a probability that is a function of the parameters of the survival function, $\theta$, discussed in ??.
+Each episode, whether detected or not, must fall into one of these three classes with a probability that is a function of the parameters of the survival function, $\vec{\vec{\theta}}$, discussed in \cref{perf-test:sec:parameters-priors}.
 
@@
   Admissible episodes, $\alpha_i$, which have an infection and recovery time consistent with the data observed.
-  This space is empty for $i \notin \set{D}$; in this case, $n_{ia} = 0$ admissible episodes occurred in individuals with test schedule $\sched_i$.
-  Otherwise, if $i \in \set{D}$ then $n_{ia} =1$ admissible episode occurred in an individual with test schedule $\sched_{i}$.
-  Conditional on the episode occurring in individual $i$, the episode is admissible with probability $p_{ia} = \prob((b, e) \in \alpha_i \mid \vec{\theta})$.
+  This space is empty for $i \notin \set{D}$; in this case individual $i$ had $n_{ia} = 0$ admissible episodes.
+  Otherwise, if $i \in \set{D}$, then individual $i$ had $n_{ia} =1$ admissible episode.
+  Conditional on the episode occurring in individual $i$, the episode is admissible with probability $p_{ia} = \prob((B_j, E_j) \in \alpha_i \mid \vec{\theta}, i(j) = i)$.
   \label{perf-test:def:admissible}
@@
   An unknown number, $n_{iu}$ of these occur in individuals with test schedule $\sched_i$.
-  Conditional on the episode occurring in individuals with test schedule $\sched_i$, the episode is undetected with probability $p_{iu} = \prob((b, e) \in \Omega_i \mid \vec{\theta})$.
+  Conditional on the episode occurring in individuals with test schedule $\sched_i$, the episode is undetected with probability $p_{iu} = \prob((B_j, E_j) \in \Omega_i \mid \vec{\theta}, i(j) = i)$.
 \item
@@
   These are all remaining episodes not in the previous sets, with $n_{i\bar{a}} = 0$ of these occurred in any individuals with test schedule $\sched_i$.
-  Conditional on the episode occurring in these individuals, the episode inadmissible with probability $p_{i\bar{a}} = \prob((b, e) \in \beta_i \mid \vec{\theta})$.
+  Conditional on the episode occurring in these individuals, the episode inadmissible with probability $p_{i\bar{a}} = \prob((B_j, E_j) \in \beta_i \mid \vec{\theta}, i(j) = i)$.
 \end{itemize}
@@
 These three classes span all possible episodes and are mutually exclusive.
-Hence, $\ntot = \ndet + \nnodet + n_{i\bar{a}} = \sum_{i=1}^{\nsched} n_{ia} + \sum_{i=1}^{\nsched} n_{iu} + 0$.
+Hence, $\ntot = \ndet + \nnodet + n_{i\bar{a}} = \sum_{i=1}^{\Ncis} n_{ia} + \sum_{i=1}^{\Ncis} n_{iu} + 0$.
 Furthermore, for all $i$, $p_{ia} + p_{iu} + p_{i\bar{a}} = 1$.
 
-Any episode $j$ independently belongs to one of these classes.
+% Any episode $j$ independently belongs to one of these classes.
+The probabilities ($p_{iu}$, $p_{ia}$, and $p_{i\bar{a}}$), are conditional on the episode occurring in individual $i$.
 I assume the individual in which each episode occurs is iid uniform across the cohort.
-Therefore, conditional on $\ntot$, the number of episodes in each class, $(n_{ia}, n_{iu}, n_{i\bar{a}})^T$, are distributed multinomially.
-with probabilities ($p_{iu}$, $p_{ia}$, and $p_{i\bar{a}}$), which are conditional on these episode occurring with test schedule $\sched_i$.
-By multiplying these probabilities by $\eta_i/\Ncis$, where $\eta_i$ is the number of individuals with test schedule $\sched_i$, the probability that an episode belongs to each class, without conditioning on a specific individual can be derived.
+Therefore, the unconditional probabilities of an infection belonging to a class need to be multiplied by $1/\Ncis$.
+For example, the probability of an episode being in class $\alpha_i$ is $p_{ia} / \Ncis$.
 
-{\color{red} Would this be the place to introduce a new subsection? on how to calculate the likelihood? Or on line 242?}
-The undetected classes for each individual cannot be distinguished by any observable data.
-Each class's count needs to be marginalised over to calculate the likelihood.
-Because the counts have a joint multinomial distribution, this is equivalent to only considering the sum of the counts.
-The overall probability of an episode being unobserved is $\pnodet = \frac{1}{\Ncis} \sum_{i=1}^{\nsched} \eta_i p_{iu}$.
-Furthermore, the inadmissible classes are guaranteed to have zero episodes.
-Counts of zero do not contribute to the multinomial probability mass function (they multiply by 1).
-Therefore, I combine the undetected classes into a single class and ignore the inadmissible classes.
+Each episode occurs in one of the classes independently and with the same probability.
+Therefore, conditional on $\ntot$, the number of episodes in each class is multinomial.
+Specifically:
+\begin{align}
+  \begin{bmatrix}
+    n_{1a} \\ n_{1u} \\ n_{1\bar{a}} \\ \vdots \\ n_{\Ncis,a} \\ n_{\Ncis,u} \\ n_{\Ncis,\bar{a}}
+  \end{bmatrix}
+  \mid \ntot
+  =
+  \MN\left(
+    \ntot,
+    \frac{1}{\Ncis}
+    \begin{bmatrix}
+      p_{1a} \\ p_{1u} \\ p_{1\bar{a}} \\ \vdots \\ p_{\Ncis,a} \\ p_{\Ncis,u} \\ p_{\Ncis,\bar{a}}
+    \end{bmatrix}
+  \right)
+\end{align}
 
-{\color{red} subsection here instead of line 233?}
-Let $\na = (n_{1a}, \dots, n_{\nsched a})^T$ be the vector of counts of episodes in the admissible classes.
-The counts $\na$ and $\nnodet$ are jointly distributed multinomially.
-The support of the distribution requires $\nnodet = \ntot - \sum_{i=1}^{\nsched} n_{ia} = \ntot - \ndet$ and $\ntot \geq \ndet$.
-Hence:
-\begin{align}
-p(\na \mid \ntot, \theta)
-&= p(\nnodet, \na\mid \ntot, \theta) \\
-&= \frac{\ntot!}{\nnodet! \prod_{i=1}^{\nsched} n_{ia}!}  \pnodet^{\nnodet}\prod_{i=1}^{\nsched} \left(\frac{1}{\nsched}p_{ia}\right)^{n_{ia}} \\
-&\propto \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i \in \set{D}} p_{ia} \label{perf-test:eq:augmented-likelihood}
-\end{align}
-the final line follows because $\nnodet = \ntot - \ndet$, $n_{ia} = 1$ for $i \in \set{D}$, and $n_{ia} = 0$ otherwise.
-
-In forming the probability $p_{ia}$, I use the test outcomes.
+In forming the region $\alpha_i$, I use the test outcomes.
 This is not strictly correct because the cell probabilities in the likelihood should not depend on the data.
@@
 
-The posterior density we are interested in is $p(\theta \mid \na)$.
+The undetected classes cannot be distinguished.
+Therefore, these classes can be combined into a single class.
+The combined class has a count $\nnodet = \sum_{i=1}^{\Ncis} n_{iu}$ and probability $\pnodet = \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} p_{iu}$.
+
+\subsection{Posterior density}
+
+Let $\na = (n_{1a}, \dots, n_{\Ncis a})^T$ be the vector of counts of episodes in the admissible classes.
+This represents all our observations.
+Therefore, the posterior density we are interested in is $p(\vec{\theta} \mid \na)$.
 It can be written as follows.
 \begin{align}
-p(\theta \mid \na)
-&\propto p(\theta) p(\na \mid \theta) \\
-&= p(\theta) \sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) p(\na \mid \theta, \ntot) \\
-&\propto p(\theta) \sum_{\ntot=\ndet}^\infty \left( p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i \in \set{D}} p_{ia} \right) \\
-&= p(\theta) \left( \prod_{i \in \set{D}} p_{ia} \right) \left( \sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \right) \label{perf-test:eq:full-posterior}.
+p(\vec{\theta} \mid \na)
+&\propto p(\vec{\theta}) p(\na \mid \vec{\theta}) \\
+&= p(\vec{\theta}) \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) p(\na \mid \ntot, \vec{\theta})
+\label{perf-test:eq:posterior1}
+\end{align}
+where the last line augments the likelihood with the total number of episodes, $\ntot$.
+
+From \cref{perf-test:sec:episode-types}, counts $\na$ and $\nnodet$ conditional on $\ntot$ are jointly distributed multinomially.
+All $n_{i\bar{a}} = 0$, therefore these classes can be ignored; counts of zero do not contribute to the multinomial probability mass function (they multiply by 1).
+The support of the distribution requires $\nnodet = \ntot - \sum_{i=1}^{\Ncis} n_{ia} = \ntot - \ndet$ and $\ntot \geq \ndet$.
+Hence:
+\begin{align}
+p(\na \mid \ntot, \vec{\theta})
+&= p(\nnodet  = \ntot - \na, \na\mid \ntot, \vec{\theta}) \\
+&= \frac{\ntot!}{(\ntot - \ndet)! \prod_{i=1}^{\Ncis} n_{ia}!}  \pnodet^{\ntot - \ndet}\prod_{i=1}^{\Ncis} \left(\frac{1}{\Ncis}p_{ia}\right)^{n_{ia}} \\
+&\propto \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i \in \set{D}} p_{ia} \label{perf-test:eq:augmented-likelihood}
+\end{align}
+the final line follows because $n_{ia} = 1$ for $i \in \set{D}$ and $n_{ia} = 0$ otherwise.
+
+Substituting \cref{perf-test:eq:augmented-likelihood} into \cref{perf-test:eq:posterior1}:
+\begin{align}
+&\propto p(\vec{\theta}) \sum_{\ntot=\ndet}^\infty \left( p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i \in \set{D}} p_{ia} \right) \\
+&= p(\vec{\theta}) \left( \prod_{i \in \set{D}} p_{ia} \right) \left( \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \right) \label{perf-test:eq:posterior2}.
 \end{align}
@@
 
-To start, I derive an analytical solution to $\sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under the prior $\ntot \dist \NBc(\mu, r)$.
-I assume that $\theta$, which are only the parameters of the survival distribution, has a prior independent of $\ntot$, the number of infections that occurred in the cohort.
-Therefore, $p(\ntot \mid \theta) = p(\ntot)$.
+\subsection{Expression for the sum}
+
+To start, I derive an analytical solution to $\sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under the prior $\ntot \dist \NBc(\mu, r)$.
+I assume that $\vec{\theta}$, which are only the parameters of the survival distribution, has a prior independent of $\ntot$, the number of infections that occurred in the cohort.
+Therefore, $p(\ntot \mid \vec{\theta}) = p(\ntot)$.
 Putting a negative binomial prior on $\ntot$ is equivalent to the following gamma-Poisson composite; its use simplifies the derivation.
@@
 
-Substituting this into \cref{perf-test:eq:full-posterior} gives:
+Substituting this into \cref{perf-test:eq:posterior2} gives:
 \begin{align}
 p(\vec{\theta} \mid \na)
-&\propto p(\vec{\theta}) \left( \prod_{i \in \set{D}} p_{ia} \right) (r + \mu (1- \pnodet))^{-(r+\ndet)} \label{perf-test:eq:full-posterior-simplified}.
+&\propto p(\vec{\theta}) \left( \prod_{i \in \set{D}} p_{ia} \right) (r + \mu (1- \pnodet))^{-(r+\ndet)} \label{perf-test:eq:full-posterior}.
 \end{align}
@@
 
+\subsection{Admissible probabilities}
+
 Next, I derive $p_{ia}$.
-By definition (see~\cpageref{perf-test:def:admissible}), we have, for any $i \in \set{D}$:
+From \cref{perf-test:eq:full-posterior}, this is only needed for $i \in \set{D}$, that is for individuals with detected episodes.
+For such an individual, let $j$ denote their detected episode.
+
+By definition (see \cpageref{perf-test:def:admissible}), we have, for any $i \in \set{D}$:
 \begin{align}
-p_{ia} =& \prob((b, e) \in \alpha_i \mid \vec{\theta}) \\
-\alpha_i =& \{ (b, e) : l_{j(i)}^{(b)} \leq b \leq r_{j(i)}^{(b)} \wedge l_{j(i)}^{(e)} \leq e \leq r_{j(i)}^{(e)}\}.
-\intertext{Omitting the conditioning on $\vec{\theta}$, this gives:}
+p_{ia} =& \prob((B_j, E_j) \in \alpha_i \mid \vec{\theta}, i(j) = i) \\
+\alpha_i =& \{ (b, e) : l_{j}^{(b)} \leq b \leq r_{j}^{(b)} \wedge l_{j}^{(e)} \leq e \leq r_{j}^{(e)}\}.
+\intertext{Omitting the conditioning on $\vec{\theta}$ and $i(j) = i$, this gives:}
 p_{ia}
-=& \prob \left( l_{j(i)}^{(b)} \leq B_{j(i)} \leq r_{j(i)}^{(b)}, l_{j(i)}^{(e)} \leq E_{j(i)} \leq r_{j(i)}^{(e)} \right) \\
-=& \prob \left( l_{j(i)}^{(e)} \leq E_{j(i)} \leq r_{j(i)}^{(e)} \mid l_{j(i)}^{(b)} \leq B_{j(i)} \leq r_{j(i)}^{(b)} \right) \\
-   &\times\prob \left( l_{j(i)}^{(b)} \leq B_{j(i)} \leq r_{j(i)}^{(b)} \right) \\
-=& \sum_{b = l_{j(i)}^{(l)}}^{r_{j(i)}^{(b)}} \prob \left( l_{j(i)}^{(e)} \leq E_{j(i)} \leq r_{j(i)}^{(e)} \mid B_{j(i)} = b \right) \prob \left(B_{j(i)} = b \right) \\
-=& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \prob \left( l_{j(i)}^{(e)} - b + 1 \leq D_{j(i)} \leq r_{j(i)}^{(e)} - b + 1 \right) \prob \left(B_{j(i)} = b \right) &\text{by def of $D_{j(i)}$} \\
-=& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_{j(i)}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 1) \right) \prob \left(B_{j(i)} = b \right) &\text{by def of $S_{\vec{\theta}}$} \\
-\propto& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_{j(i)}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 1) \right)
+=& \prob \left( l_{j}^{(b)} \leq B_{j} \leq r_{j}^{(b)}, l_{j}^{(e)} \leq E_{j} \leq r_{j}^{(e)} \right) \\
+=& \prob \left( l_{j}^{(e)} \leq E_{j} \leq r_{j}^{(e)} \mid l_{j}^{(b)} \leq B_{j} \leq r_{j}^{(b)} \right) \\
+   &\times\prob \left( l_{j}^{(b)} \leq B_{j} \leq r_{j}^{(b)} \right) \\
+=& \sum_{b = l_{j}^{(l)}}^{r_{j}^{(b)}} \prob \left( l_{j}^{(e)} \leq E_{j} \leq r_{j}^{(e)} \mid B_{j} = b \right) \prob \left(B_{j} = b \right) \\
+=& \sum_{b = l_{j}^{(b)}}^{r_{j}^{(b)}} \prob \left( l_{j}^{(e)} - b + 1 \leq D_{j} \leq r_{j}^{(e)} - b + 1 \right) \prob \left(B_{j} = b \right) &\text{by def of $D_{j}$} \\
+=& \sum_{b = l_{j}^{(b)}}^{r_{j}^{(b)}} \left( S_{\vec{\theta}}(l_{j}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j}^{(e)} - b + 1) \right) \prob \left(B_{j} = b \right) &\text{by def of $S_{\vec{\theta}}$} \\
+\propto& \sum_{b = l_{j}^{(b)}}^{r_{j}^{(b)}} \left( S_{\vec{\theta}}(l_{j}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j}^{(e)} - b + 1) \right)
 \end{align}
 under the assumption of uniform probability of infection time.
-This is the standard form of the likelihood for doubly interval censored data with truncation.
+This is the standard form of the likelihood for doubly interval censored data without truncation.
 
-The next step is to derive $1 - p_u = 1 - \frac{1}{N} \sum_{i=1}^{\nsched} p_{iu} = \frac{1}{N} \sum_{i=1}^{\nsched} (1 - p_{iu})$.
-Therefore, the crucial component is $1 - p_{iu} = 1 - \prob((B_{j(i)}, E_{j(i)}) \in \Omega_i \mid \vec{\theta})$, one minus the probability that episode $i$ was undetected.
-An episode is undetected if and only if no tests are performed during the episode or if there was no negative test prior to the episode.
-Equivalently, that the first test at or after $b$ is after $e$, or that there is no negative test prior to $b$.
+\subsection{Undetected probability} \label{perf-test:sec:prob-undetected}
+
+The final component of \cref{perf-test:eq:full-posterior} required is $1 - p_u$.
+From \cref{perf-test:sec:episode-types}:
+\begin{align}
+  1 - p_u
+  &= 1 - \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} p_{iu} \\
+  &= \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} (1 - p_{iu}).
+  \label{perf-test:eq:pu}
+\end{align}
+
+Therefore, the crucial component is $1 - p_{iu} = 1 - \prob((B_j, E_j) \in \Omega_i \mid \vec{\theta}, i(j) = i)$.
+This is one minus the probability that an episode in individual $i$ was undetected, \ie the probability that episode such an episode is detected.
+
+An episode $j$ in individual $i(j)$ is detected if and only if all the following conditions are met.
+\begin{enumerate}
+    \item $\exists t \in \sched_{i(j)}$ such that $b_j \leq t \leq e_j$; this condition is equivalent to having at least one positive test for the episode.
+    \item $\exists t \in \sched_{i(j)}$ such that $t < b_j$.
+      For individuals enrolled during the period considered ($\min \sched_{i(j)} > 0$), this ensures that the beginning of the episode is lower bounded.
+      For individuals enrolled prior to the period considered ($\min \sched_{i(j)} \leq 0$), this means that the episode was not detected prior to time 1.
+    \item $\exists t \in \sched_{i(j)}$ such that $t > e_j$, upper bounding the end of the episode.
+      For episodes detected in the period I consider, this happens a negligible number of times, only in individuals lost to follow-up; therefore, I neglect this possibility.
+      For a new context, including recent infections, this condition could be relaxed by considering episodes that do not meet this criteria as right censored.
+\end{enumerate}
+
+% An episode is undetected if and only if no tests are performed during the episode or if there was no negative test prior to the episode.
+% Equivalently, that the first test at or after $b$ is after $e$, or that there is no negative test prior to $b$.
 To formalise this, first define $\tau_{\sched_i}(t)$ as the time until the next test at or after time $t$ in the schedule $\sched_i$:
@@
 \end{align}
-Then $\Omega_i$ can be written as:
-\begin{align}
-\Omega_i = \{ (b, e) \ssep \tau_{\sched_i}(b) + b > e \vee b \leq \min(\sched_i) \}.
-\end{align}
-Therefore, suppressing the conditioning on $\vec{\theta}$:
+The condition that $\exists t \in \sched_{i(j)}$ such that $b_j \leq t \leq e_j$ can now be expressed as $e_j \geq b_j + \tau_{\sched_{i(j)}}(b_j)$.
+Equivalently, $d_j \geq \tau_{\sched_{i(j)}}(b_j) + 1$.
+
+% Then $\Omega_i$ can be written as:
+% \begin{align}
+% \Omega_i = \{ (b, e) \ssep \tau_{\sched_i}(b) + b > e \vee b \leq \min(\sched_i) \}.
+% \end{align}
+Therefore, suppressing the conditioning on $\vec{\theta}$ and $i(j) = i$:
 \begin{align}
 1 - p_{iu}
-&= 1 - \prob((B_{j(i)}, E_{j(i)}) \in \Omega_i) \\
-&= 1 - \prob( \tau_{\sched_i}(B_{j(i)}) + B_{j(i)} > E_{j(i)} \vee B_{j(i)} \leq \min(\sched_i) ) \\
-&= 1 - \prob( \tau_{\sched_i}(B_{j(i)}) + 1 > D_i , B_{j(i)} > \min(\sched_i) )  - \prob( B_{j(i)} \leq \min(\sched_i) ) \\
-&= 1 - \sum_{b={\min\sched_i} + 1}^T \prob( \tau_{\sched_i}(b) + 1> D_i) \prob(B_{j(i)} = b) - \sum_{t=1}^{\min\sched_i} \prob(B_{j(i)} = b)\\
-&= 1 - \frac{1}{T} \sum_{b=\min(\sched_i)+1}^T (1 - S_{\vec{\theta}}(\tau_{\sched_i}(b) + 1)) - \frac{1}{T} \min\sched_i \\
-&= \frac{1}{T} \left(T - \sum_{b=\min\sched_i+1}^T (1 - S_{\vec{\theta}}(\tau_{\sched_i}(b) + 1)) - \min\sched_i \right) \\
-&= \frac{1}{T} \left(T - (T - \min\sched_i) + \sum_{b=\min\sched_i+1}^T S_{\vec{\theta}}(\tau_{\sched_i}(b + 1)) - \min\sched_i \right) \\
-&= \frac{1}{T} \sum_{b=\min(\sched_i)+1}^T S_{\vec{\theta}}(\tau_{\sched_i}(b) + 1). \label{perf-test:eq:piu}
+&= 1 - \prob((B_{j}, E_{j}) \in \Omega_i) \\
+&= \prob(D_j \geq \tau_{\sched_{i}}(B_j)+ 1, B_j > \min \sched_{j}) \\
+&= \sum_{b = \min \sched_{i} + 1}^T \prob(D_j \geq \tau_{\sched_{i}}(b) + 1 \mid B_j = b) \prob(B_j = b)\\
+&= \frac{1}{T} \sum_{b = \min \sched_{i} + 1}^T S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1).
+\intertext{Substituting into \cref{perf-test:eq:pu}:}
+1 - p_u
+& \propto \sum_{i=1}^{\Ncis} \sum_{b = \min \sched_{i} + 1}^T S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1).
 \end{align}
@@
     \item Extract the test schedules for each individual who had at least one test during the period of interest.
-    \item Draw an episode start time, $b_{j(i)}$ for each individual uniformly at random between 2nd July 2020 (100 days before the period where a detected episode would be included) and 6th December 2020 (the end of this period).
-    \item Draw a duration of episode for each individual, $d_i$, based on the distribution described below. Then calculate the end of their infection episode, $e_{j(i)} = b_{j(i)} + d_i - 1$.
-    \item Simulate the test results based on the test schedule, $b_{j(i)}$, and $e_{j(i)}$. Tests between $b_{j(i)}$ and $e_{j(i)}$ (inclusive) are positive, all other tests are negative.
+    \item Draw an episode start time, $b_{j}$ for each individual uniformly at random between 2nd July 2020 (100 days before the period where a detected episode would be included) and 6th December 2020 (the end of this period).
+    \item Draw a duration of episode for each individual, $d_i$, based on the distribution described below. Then calculate the end of their infection episode, $e_{j} = b_{j} + d_i - 1$.
+    \item Simulate the test results based on the test schedule, $b_{j}$, and $e_{j}$. Tests between $b_{j}$ and $e_{j}$ (inclusive) are positive, all other tests are negative.
     \item Discard episodes where there are no positive tests (\ie undetected episodes) and then apply the same inclusions/exclusions as in \cref{perf-test:sec:problem}.

### 2024-03-22 cis-perfect-testing.tex
finish 5.3 rewrite

@@
 In this chapter, I develop a method to estimate the duration distribution using CIS data, assuming no false negatives or positives.
+Formally, I aim to estimate the survival function $S$ defined in \cref{inc-prev:sec:definitions}.
 
@@
 This motivates using a survival analysis framework which I introduce in \cref{perf-test:sec:surv-analysis}, before deriving the statistical model for the CIS setting in \cref{perf-test:sec:model}.
-Next, in \cref{perf-test:sec:parameters-priors}, I consider how to parameterise the survival function and choose priors; these include vague and smoothing priors as well as including information from \cref{E-ATACCC}.
+Next, in \cref{perf-test:sec:parameters-priors}, I consider how to parameterize the survival function and choose priors; these include vague and smoothing priors as well as including information from \cref{E-ATACCC}.
 I evaluate the model and priors using a simulation study (\cref{perf-test:sec:simulation-study}), showing that they perform well.
@@
 
-Infection episodes in an individual's one week testing period are over-represented in the detected infections.
+Infection episodes in an individual's one-week testing period are over-represented in the detected infections.
 This is because more frequent testing will detect a higher proportion of infections.
@@
 In this section, I develop a statistical model accounting for double interval censoring and undetected episodes.
-This approach extends the framework of \textcite{heiseyModelling} to include all individuals in the CIS, including those without detected episodes.
+In \cref{perf-test:sec:data-notation}, I introduce the data and notation, defining which quantities are observed and which are latent.
+Then, in \cref{perf-test:sec:episode-types}, I apply the framework of \textcite{heiseyModelling} to classify possible episode beginning and end types into \emph{admissible} episodes, which match the doubly interval censored pattern of the observations; \emph{undetected} episodes, which may have occurred but would have been undetected; and \emph{inadmissible} episodes, which the data rules out having occurred.
+The number of such episodes in each class is a random variable, with a multinomial distribution, conditional on the latent total number of infections (including undetected infections).
 
-The number of missed episodes is now a latent quantity that needs to be considered.
-Marginalising over the number of missed episodes is possible analytically.
+In \cref{perf-test:sec:posterior-density}, I outline the form of the posterior density.
+The following subsections then derive its various components.
+Unlike in \textcite{heiseyModelling}, I must consider the probability of undetected episodes for each individual seperately because they have differing visit times.
+In addition, I include the information from the individuals without a detected episode.
+
+Unlike prior work, I marginalise the likelihood over the number of undetected episodes.
 This makes inference tractable using the large sample size and the computing resources available within the SRS (see \cref{E-biology-data:sec:SRS}).
@@
 
-The model will be used to analyse the 4,800 episodes first detected between 10th Oct 2020 and 6th Dec 2020 inclusive, with negatives bounding the start and end time of the episode.
+\subsection{Data and notation} \label{perf-test:sec:data-notation}
+
+I analyse the 4,800 episodes first detected between 10th Oct 2020 and 6th Dec 2020 inclusive, with negatives bounding the start and end time of the episode.
 This period is denoted as $t = 1, \dots, T$.
@@
 Similar information comes from the negative test after the end of the episode.
-This is typically observed, unless the individual is lost to follow-up before their episode is ended, which happens rarely.
+A negligible number of episodes have a negative test after the end of the episode.
+I refer to these episodes as \emph{detected}; in \cref{perf-test:sec:prob-undetected} I formalise the conditions for an episode to be detected.
+Let $\ndet$ denote the total number of detected infections.
+
+The cohort of individuals considered is all those in the CIS who could have had a detected test.
+This requires that they had at least three tests before time $T$ (a negative prior to the episode, a positive, and a negative following the episode).
+Index these individuals by $i = 1, \dots, \Ncis$ where $\Ncis$ is the total number of such individuals.
+
+An individual $i$ has a \emph{test schedule} $\sched_i$.
+The test schedule $\sched_i$ is the set of times individual $i$ is tested at, from their last test prior to time 1 (if it exists) up until time $T$.
+This is all the times the infection episode could be detected at and be included, plus one prior test.
+For example, for the individual in \cref{perf-test:sec:undetected}, $\sched_i = \{ 0, 7, 14, 21, 28, 56 \}$.
+Each individual has exactly one test schedule, but it is possible that $\sched_i = \sched_{i'}$ for $i \neq i'$.
+I assume that the test schedules are uninformative on all quantities of interest because they are part of the study design.
+Therefore, I condition on them implicitly in all probabilities that follow.
+
+For a detected episode $j$, five pieces of data are observed (see \cref{E-biology-data:sec:cis-episodes} for how these are constructed).
+The individual in which the episode occurs, $i(j)$; the earliest and latest time the episode could have begun, $l_j^{(b)}$ and $r_j^{(b)}$ respectively; and the earliest and latest time the episode could have ended, $l_j^{(e)}$ and $r_j^{(e)}$ respectively.
+Over the short calendar period considered, each individual has zero or one detected episodes.
+Let $i \in \set{D}$ indicate individual $i$ has a detected episode.
+
+In addition, the individuals in the cohort experienced an unknown number of undetected infections, $\nnodet$.
+The total number of infections in the cohort is $\ntot = \ndet + \nnodet$ and I index them with $j = 1, \dots, \ntot$.
+
+Any infection episode, whether detected or not, can be fully characterised by the triplet $(b_j, e_j, i(j))$ where $b_j$ and $e_j$ are the times the episode begin and end respectively, and $i(j)$ denotes the individual the episode occurred in.
+The episode's duration is $d_j = e_j - b_j + 1$ (see \cref{E-inc-prev:sec:definitions}).
+For all $j$, the exact values $b_j$ and $e_j$ are latent.
+If $j$ is undetected then $i(j)$ is also latent and nothing is known about $b_j$ or $e_j$.
+If $j$ is detected, then $i(j)$ is known; $b_j \in [l_j^{(b)}, r_j^{(b)}]$; and $e_j \in [l_j^{(e)}, r_j^{(e)}]$.
+These latter two properties are what makes the data doubly interval censored.
 
@@
 
-Let the total number of infection episodes in the cohort be $\ntot = \ndet + \nnodet$ where $\ndet$ is the number of detected infections and $\nnodet$ is the number of undetected infections.
-Let $\Ncis$ be the number of individuals in the cohort.
-%Without loss of generality, let the individuals with a detected episode be indexed by $i = 1, \dots, \ndet$ and the individuals without a detected episode be indexed by $i = \ndet + 1, \dots, \Ncis$.
-Let $i = 1, \dots, \Ncis$ index individuals and $j = 1, \dots \ntot$ index infection episodes.
-The test schedule $\sched_i$ is the set of times individual $i$ is tested at, from their last test prior to time 1 (if it exists) up until time $T$.
-This is all the times the infection episode could be detected at and be included, plus one prior test.
-For example, for the individual in \cref{perf-test:sec:undetected}, $\sched_i = \{ 0, 7, 14, 21, 28, 56 \}$.
-Each individual has exactly one test schedule, but a test schedule could have multiple individuals associated with it.
-
-All the relevant information about the infection episode $j$ can be fully characterized by the (unobserved) triplet $(b_j, e_j, i(j))$ where $b_j$ and $e_j$ are the times the episode begin and end respectively and $i(j)$ denotes the individual the episode occurred in.
-The episode's duration is $d_j = e_j - b_j + 1$ (see \cref{E-inc-prev:sec:definitions}).
-%, the number of days the individual in which the episode occurred, was positive for 
-% This triplet is assumed iid.
-
-% The times $b_j$ and $e_j$ span possible infection and recovery times.
-% $\sched_j$ is one of the observed test schedules, with any other test schedule assumed to occur with probability zero.
-
-Over the short calendar period considered, each individual has zero or one detected episodes.
-Let $i \in \set{D}$ indicate individual $i$ has a detected episode.
-I define the precise conditions for this to occur in \cref{perf-test:sec:prob-undetected}.
-% That is, $i \in \set{D}$ if and only if there exists a $j$ such that $b_{j}$ and $e_{j}$ such that $(b_{j}, e_{j}, \sched_i)$ is a detected episode.
-
 To relate the possible value for $b_j$ and $e_j$, given a test schedule $\sched_{i(j)}$, to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie: are strict subsets of $T \times T$).
 Figure \ref{perf-test:fig:partitionSpace} shows this graphically.
-Each episode, whether detected or not, must fall into one of these three classes with a probability that is a function of the parameters of the survival function, $\vec{\vec{\theta}}$, discussed in \cref{perf-test:sec:parameters-priors}.
+Each episode, whether detected or not, must fall into one of these three classes with a probability that is a function of the parameters of the survival function, $\vec{\vec{\theta}}$, specified in \cref{perf-test:sec:parameters-priors}.
 
@@
   Otherwise, if $i \in \set{D}$, then individual $i$ had $n_{ia} =1$ admissible episode.
+  The admissible region is defined by the intersection of the intervals $[l_j^{(b)}, r_j^{(b)}]$ and $[l_j^{(e)}, r_j^{(e)}]$.
+  This is the purple area in \cref{perf-test:fig:partitionSpace}; the boundaries in the $x$ direction of the area are $l_j^{(b)}$ and $r_j^{(b)}$ and in the $y$ direction are $l_j^{(e)}$ and $r_j^{(e)}$.
+  The shape of this region accounts for the doubly interval censored nature of the data.
   Conditional on the episode occurring in individual $i$, the episode is admissible with probability $p_{ia} = \prob((B_j, E_j) \in \alpha_i \mid \vec{\theta}, i(j) = i)$.
@@
   Undetected episodes, $\Omega_i$, which have an infection and recovery time such that they would not have tested positive.
-  An unknown number, $n_{iu}$ of these occur in individuals with test schedule $\sched_i$.
-  Conditional on the episode occurring in individuals with test schedule $\sched_i$, the episode is undetected with probability $p_{iu} = \prob((B_j, E_j) \in \Omega_i \mid \vec{\theta}, i(j) = i)$.
+  An unknown number, $n_{iu}$ of these occur in individual $i$.
+  The undetected region is the red region in \cref{perf-test:fig:partitionSpace}, which are when the individual is infected then recovers before their next test, plus when the individual is infected before their enrolment.
+  The shape of this region accounts for the pattern of possible undetected infections.
+  Conditional on the episode occurring in individual $i$, the episode is undetected with probability $p_{iu} = \prob((B_j, E_j) \in \Omega_i \mid \vec{\theta}, i(j) = i)$.
 \item
-  Inadmissible episodes, $\beta_i$, which are not consistent with the data observed but would have been
-  detected.
-  These are all remaining episodes not in the previous sets, with $n_{i\bar{a}} = 0$ of these occurred in any individuals with test schedule $\sched_i$.
+  Inadmissible episodes, $\beta_i$, which would be detected (if they occur) but were not.
+  Therefore, we know $n_{i\bar{a}} = 0$ of these occurred in individual $i$.
+  These are all remaining episodes not in the previous sets.
   Conditional on the episode occurring in these individuals, the episode inadmissible with probability $p_{i\bar{a}} = \prob((B_j, E_j) \in \beta_i \mid \vec{\theta}, i(j) = i)$.
@@
 \includegraphics[width=\textwidth]{cis-perfect-testing/regions_diag}
-\caption[Admissible, inadmissible, and undetected infections]{The regions (as defined in the main text) for an individual with test schedule $\sched_i$ whose first test was at time 0 and negative, with subsequent
-negative tests at times 7, 14, and 56, and positive tests at times 21
-and 28. Inadmissible regions are unshaded. The black (impossible) region covers the cases in which recovery occurs prior to infection. \label{perf-test:fig:partitionSpace}}
+\caption[Admissible, inadmissible, and undetected regions]{%
+  The regions for individual $i$.
+  $i$'s first test was at time 0 and negative, with subsequent negative tests at times 7, 14, and 56, and positive tests at times 21 and 28.
+  Inadmissible regions are unshaded.
+  The black (impossible) region covers the cases in which recovery occurs prior to infection.
+  \label{perf-test:fig:partitionSpace}
+}
 \end{figure}
@@
   \right)
+  \label{perf-test:eq:multinomial}
 \end{align}
@@
 
-\subsection{Posterior density}
+\subsection{Posterior density} \label{perf-test:sec:posterior-density}
 
 Let $\na = (n_{1a}, \dots, n_{\Ncis a})^T$ be the vector of counts of episodes in the admissible classes.
-This represents all our observations.
+This represents all the observations used in \cref{perf-test:eq:multinomial}.
 Therefore, the posterior density we are interested in is $p(\vec{\theta} \mid \na)$.
@@
 
-From \cref{perf-test:sec:episode-types}, counts $\na$ and $\nnodet$ conditional on $\ntot$ are jointly distributed multinomially.
+First, I derive the form of $p(\na \mid \ntot, \vec{\theta})$.
+From \cref{perf-test:sec:episode-types}, the counts $\na$ and $\nnodet$'s joint distribution, conditional on $\ntot$, is multinomial.
 All $n_{i\bar{a}} = 0$, therefore these classes can be ignored; counts of zero do not contribute to the multinomial probability mass function (they multiply by 1).
@@
 \end{align}
-The rest of this section derives expressions for each $p_{ia}$, $p_{u}$ and the sum.
+For convenience, define the summation term as:
+\begin{align}
+  \eta =  \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}.
+  \label{perf-test:eq:eta}
+\end{align}
+The rest of this section derives expressions for each of $p_{ia}$, $p_{u}$ and $\eta$.
 
-\subsection{Expression for the sum}
+\subsection{Expression for $\eta$}
 
-To start, I derive an analytical solution to $\sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under the prior $\ntot \dist \NBc(\mu, r)$.
+I derive an analytical solution to $\eta$ under the prior $\ntot \dist \NBc(\mu, r)$.
 I assume that $\vec{\theta}$, which are only the parameters of the survival distribution, has a prior independent of $\ntot$, the number of infections that occurred in the cohort.
 Therefore, $p(\ntot \mid \vec{\theta}) = p(\ntot)$.
+
 Putting a negative binomial prior on $\ntot$ is equivalent to the following gamma-Poisson composite; its use simplifies the derivation.
@@
 \begin{align}
-&\sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} p(\ntot) \\
+\eta
 &= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} p(\ntot \mid \lambda) p(\lambda) d\lambda &\text{$\lambda$ explicit}\\
@@
     \item $\exists t \in \sched_{i(j)}$ such that $t < b_j$.
-      For individuals enrolled during the period considered ($\min \sched_{i(j)} > 0$), this ensures that the beginning of the episode is lower bounded.
+      For individuals enrolled during the period considered ($\min \sched_{i(j)} > 0$), this ensures that the beginning of the episode is lower bounded; thereby adjusting for those who enrolled during the period being less likely to have a detected infection.
       For individuals enrolled prior to the period considered ($\min \sched_{i(j)} \leq 0$), this means that the episode was not detected prior to time 1.
+    \item $\exists t \in \sched_{i(j)}$ such that $t \geq b_j$.
+      This means that the test is detected before $T$.
     \item $\exists t \in \sched_{i(j)}$ such that $t > e_j$, upper bounding the end of the episode.
@@
 \begin{align}
-\tau_{\sched_i}(t) &= \min \{ t' \in \sched_i : t' \geq t \} - t.
+\tau_{\sched_i}(t) &= \min \{ t' \in \sched_i : t' \geq t \} - t
 \end{align}
+defining $\min \emptyset = \infty$; \ie $\tau_{\sched_i}(t) = \infty$ if there is no $t' \in \sched_i$ such that $t' \geq t$.
 The condition that $\exists t \in \sched_{i(j)}$ such that $b_j \leq t \leq e_j$ can now be expressed as $e_j \geq b_j + \tau_{\sched_{i(j)}}(b_j)$.
@@
 % \end{align}
-Therefore, suppressing the conditioning on $\vec{\theta}$ and $i(j) = i$:
+Therefore, omitting the conditioning on $\vec{\theta}$ and $i(j) = i$:
 \begin{align}
@@
 &= \sum_{b = \min \sched_{i} + 1}^T \prob(D_j \geq \tau_{\sched_{i}}(b) + 1 \mid B_j = b) \prob(B_j = b)\\
-&= \frac{1}{T} \sum_{b = \min \sched_{i} + 1}^T S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1).
-\intertext{Substituting into \cref{perf-test:eq:pu}:}
+&= \frac{1}{T} \sum_{b = \min \sched_{i} + 1}^T S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1)
+\end{align}
+defining $S_{\vec{\theta}}(\infty) = 0$.
+
+Substituting into \cref{perf-test:eq:pu}:
+\begin{align}
 1 - p_u
-& \propto \sum_{i=1}^{\Ncis} \sum_{b = \min \sched_{i} + 1}^T S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1).
+& \propto \sum_{i=1}^{\Ncis} \sum_{b = \min \sched_{i} + 1}^T S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1)
 \end{align}

### 2024-03-22 cis-perfect-testing/regions_diag.R

@@ -40,7 +40,7 @@ plot = ggplot() +
         ),
         labels = c(
             "Admissible" = "Admissible, α",
-            "Undetected" = bquote("Undetected, Ω"^C),
+            "Undetected" = "Undetected, Ω",
             "Impossible" = "Impossible",
             "Inadmissible" = "Inadmissible, β"
         )

### 2024-03-22 cis-perfect-testing.tex
Updates from Overleaf

--- 2024-03-22 cis-perfect-testing.tex
@@
 %   of the time of the infection. Define the survival function
-%   $\prob(D_i \geq t \mid B_{j(i)} = b, \vec{\theta}) = \prob(D_i \geq t \mid \vec{\theta}) = S_{\vec{\theta}}(t)$,
-%   where $\vec{\theta}$ are the parameters controlling the survival
+%   $\prob(D_i \geq t \mid B_{j(i)} = b, \theta) = \prob(D_i \geq t \mid \theta) = S_\theta(t)$,
+%   where $\theta$ are the parameters controlling the survival
 %   distribution (the discussion in this document is valid regardless of
-%   the model specified for $S_{\vec{\theta}}$ and hence we consider $\vec{\theta}$
+%   the model specified for $S_\theta$ and hence we consider $\theta$
 %   as an arbitrary vector of parameters).
@@
 Doubly interval censored and truncated data are common when assessing the survival time of bird and insect nests~\autocite{heiseyABCs}.
-\textcite{heiseyModelling}, building on long-standing methodology~\autocite{dempsterMaximum,turnbullEmpirical}, develops a likelihood-based inference framework, allowing for arbitrary truncation patterns and double interval censoring.
+\textcite{heiseyModelling}, building on on long-standing methodology~\autocite{dempsterMaximum,turnbullEmpirical}, develops a likelihood-based inference framework, allowing for arbitrary truncation patterns and double interval censoring.
 They apply it to a situation where the visit times (possible detection times) are common to all the individuals, simplifying the likelihood.
@@
 I will incorporate episodes excluded in this step in the model for undetected infections.
+{\color{red} introduce here what is being done in teh next subsections to give some ideas of why you do what you do}
 
@@
   Otherwise, if $i \in \set{D}$ then $n_{ia} =1$ admissible episode occurred in an individual with test schedule $\sched_{i}$.
-  Conditional on the episode occurring in individual $i$, the episode is admissible with probability $p_{ia} = \prob((b, e) \in \alpha_i \mid \vec{\theta})$.
+  Conditional on the episode occurring in individuals with test schedule $\sched_i$, the episode is admissible with probability $p_{ia} = \prob((b, e) \in \alpha_i \mid \theta)$.
   \label{perf-test:def:admissible}
@@
   An unknown number, $n_{iu}$ of these occur in individuals with test schedule $\sched_i$.
-  Conditional on the episode occurring in individuals with test schedule $\sched_i$, the episode is undetected with probability $p_{iu} = \prob((b, e) \in \Omega_i \mid \vec{\theta})$.
+  Conditional on the episode occurring in individuals with test schedule $\sched_i$, the episode is undetected with probability $p_{iu} = \prob((b, e) \in \Omega_i \mid \theta)$.
 \item
@@
   These are all remaining episodes not in the previous sets, with $n_{i\bar{a}} = 0$ of these occurred in any individuals with test schedule $\sched_i$.
-  Conditional on the episode occurring in these individuals, the episode inadmissible with probability $p_{i\bar{a}} = \prob((b, e) \in \beta_i \mid \vec{\theta})$.
+  Conditional on the episode occurring in these individuals, the episode inadmissible with probability $p_{i\bar{a}} = \prob((b, e) \in \beta_i \mid \theta)$.
 \end{itemize}
@@
 
-{\color{red} Would this be the place to introduce a new subsection? on how to calculate the likelihood? Or on line 242?}
+{\color{red} Would this be the place to introduce a new subsection? on how to calculate the likelihood? }
 The undetected classes for each individual cannot be distinguished by any observable data.
@@
 
-{\color{red} subsection here instead of line 233?}
+{\color{red} subsection on likelihood calculation here instead of above?}
 Let $\na = (n_{1a}, \dots, n_{\nsched a})^T$ be the vector of counts of episodes in the admissible classes.
@@
 
-To start, I derive an analytical solution to $\sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under the prior $\ntot \dist \NBc(\mu, r)$.
+\paragraph{name of the sum} To start, I derive an analytical solution to $\sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under the prior $\ntot \dist \NBc(\mu, r)$.
 I assume that $\theta$, which are only the parameters of the survival distribution, has a prior independent of $\ntot$, the number of infections that occurred in the cohort.
@@
 \begin{align}
-p(\vec{\theta} \mid \na)
-&\propto p(\vec{\theta}) \left( \prod_{i \in \set{D}} p_{ia} \right) (r + \mu (1- \pnodet))^{-(r+\ndet)} \label{perf-test:eq:full-posterior-simplified}.
+p(\theta \mid \na)
+&\propto p(\theta) \left( \prod_{i \in \set{D}} p_{ia} \right) (r + \mu (1- \pnodet))^{-(r+\ndet)} \label{perf-test:eq:full-posterior-simplified}.
 \end{align}
@@
 % $\ntot$ is generally a nuisance parameter, however, its posterior distribution can be useful for diagnostic purposes (as in X\todo{ref where I use this in the next chapter}).
-% Its posterior can be reconstructed using a posterior sample of $\vec{\theta}$.
-% For each posterior sample of $\vec{\theta}$, sample from the full conditional of $\ntot$ to give the joint posterior of $\ntot$ and $\vec{\theta}$.
+% Its posterior can be reconstructed using a posterior sample of $\theta$.
+% For each posterior sample of $\theta$, sample from the full conditional of $\ntot$ to give the joint posterior of $\ntot$ and $\theta$.
 % The full conditional of $\ntot$ is given by:
 % \begin{align}
-% p(n_\text{tot} \mid \ncis, \vec{\theta})
-% &\propto p(\ntot \mid \vec{\theta}) p(\ncis \mid \vec{\theta}, \ntot) \\
-% &\propto p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} &\text{by \cref{perf-test:eq:augmented-likelihood}} \\
+% p(n_\text{tot} \mid \ncis, \theta)
+% &\propto p(\ntot \mid \theta) p(\ncis \mid \theta, \ntot) \\
+% &\propto p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} &\text{by \cref{perf-test:eq:augmented-likelihood}} \\
 % &\propto \frac{\Gamma(r + \ntot)}{\ntot!} \left( \frac{\mu}{r + \mu} \right)^{\ntot} \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot} \\
@@
 
-Next, I derive $p_{ia}$.
+\Paragraph{$p_{ia}$}
 By definition (see~\cpageref{perf-test:def:admissible}), we have, for any $i \in \set{D}$:
 \begin{align}
-p_{ia} =& \prob((b, e) \in \alpha_i \mid \vec{\theta}) \\
+p_{ia} =& \prob((b, e) \in \alpha_i \mid \theta) \\
 \alpha_i =& \{ (b, e) : l_{j(i)}^{(b)} \leq b \leq r_{j(i)}^{(b)} \wedge l_{j(i)}^{(e)} \leq e \leq r_{j(i)}^{(e)}\}.
-\intertext{Omitting the conditioning on $\vec{\theta}$, this gives:}
+\intertext{Omitting the conditioning on $\theta$, this gives:}
 p_{ia}
@@
 =& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \prob \left( l_{j(i)}^{(e)} - b + 1 \leq D_{j(i)} \leq r_{j(i)}^{(e)} - b + 1 \right) \prob \left(B_{j(i)} = b \right) &\text{by def of $D_{j(i)}$} \\
-=& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_{j(i)}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 1) \right) \prob \left(B_{j(i)} = b \right) &\text{by def of $S_{\vec{\theta}}$} \\
-\propto& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_{j(i)}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 1) \right)
+=& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_\theta(l_{j(i)}^{(e)} - b + 1) - S_\theta(r_{j(i)}^{(e)} - b + 1) \right) \prob \left(B_{j(i)} = b \right) &\text{by def of $S_\theta$} \\
+\propto& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_\theta(l_{j(i)}^{(e)} - b + 1) - S_\theta(r_{j(i)}^{(e)} - b + 1) \right)
 \end{align}
@@
 
-The next step is to derive $1 - p_u = 1 - \frac{1}{N} \sum_{i=1}^{\nsched} p_{iu} = \frac{1}{N} \sum_{i=1}^{\nsched} (1 - p_{iu})$.
-Therefore, the crucial component is $1 - p_{iu} = 1 - \prob((B_{j(i)}, E_{j(i)}) \in \Omega_i \mid \vec{\theta})$, one minus the probability that episode $i$ was undetected.
+%The next step is to derive 
+\paragraph{$1 - p_u$}
+Note that $1 - p_u= 1 - \frac{1}{N} \sum_{i=1}^{\nsched} p_{iu} = \frac{1}{N} \sum_{i=1}^{\nsched} (1 - p_{iu})$.
+Therefore, the crucial component is $1 - p_{iu} = 1 - \prob((B_{j(i)}, E_{j(i)}) \in \Omega_i \mid \theta)$, one minus the probability that episode $i$ was undetected.
 An episode is undetected if and only if no tests are performed during the episode or if there was no negative test prior to the episode.
@@
 \end{align}
-Therefore, suppressing the conditioning on $\vec{\theta}$:
+Therefore, suppressing the conditioning on $\theta$:
 \begin{align}
@@
 &= 1 - \sum_{b={\min\sched_i} + 1}^T \prob( \tau_{\sched_i}(b) + 1> D_i) \prob(B_{j(i)} = b) - \sum_{t=1}^{\min\sched_i} \prob(B_{j(i)} = b)\\
-&= 1 - \frac{1}{T} \sum_{b=\min(\sched_i)+1}^T (1 - S_{\vec{\theta}}(\tau_{\sched_i}(b) + 1)) - \frac{1}{T} \min\sched_i \\
-&= \frac{1}{T} \left(T - \sum_{b=\min\sched_i+1}^T (1 - S_{\vec{\theta}}(\tau_{\sched_i}(b) + 1)) - \min\sched_i \right) \\
-&= \frac{1}{T} \left(T - (T - \min\sched_i) + \sum_{b=\min\sched_i+1}^T S_{\vec{\theta}}(\tau_{\sched_i}(b + 1)) - \min\sched_i \right) \\
-&= \frac{1}{T} \sum_{b=\min(\sched_i)+1}^T S_{\vec{\theta}}(\tau_{\sched_i}(b) + 1). \label{perf-test:eq:piu}
+&= 1 - \frac{1}{T} \sum_{b=\min(\sched_i)+1}^T (1 - S_\theta(\tau_{\sched_i}(b) + 1)) - \frac{1}{T} \min\sched_i \\
+&= \frac{1}{T} \left(T - \sum_{b=\min\sched_i+1}^T (1 - S_\theta(\tau_{\sched_i}(b) + 1)) - \min\sched_i \right) \\
+&= \frac{1}{T} \left(T - (T - \min\sched_i) + \sum_{b=\min\sched_i+1}^T S_\theta(\tau_{\sched_i}(b + 1)) - \min\sched_i \right) \\
+&= \frac{1}{T} \sum_{b=\min(\sched_i)+1}^T S_\theta(\tau_{\sched_i}(b) + 1). \label{perf-test:eq:piu}
 \end{align}
@@
 
-In this section I specify the form of $S_{\vec{\theta}}(t)$.
-% As frequently observed in the literature~\autocite[e.g.][]{heBayesian}, it is more convenient to operate with the hazard than the survival or probability mass function directly.
-% The mass function must sum to one.
-% The survival must be monotonically decreasing
-% But, there are no constraints across multiple hazards.
-% This would complicate the inference process in a way that cannot be removed through a simple transformation.
-% The only constraint on the hazard is to be in the interval $[0, 1]$, since they are probabilities.
-% The [0, 1] interval is easily be mapped to the unconstrained reals using a logit transformation (where $\logit(x) = \log(x/(1-x))$).
+In this section I specify the form of $S_\theta(t)$.
+I parameterise $S$ in terms of the hazard $S_{\vec{\theta}}(t) = \prod_{i=1}^{t-1} (1 - \lambda(i))$ (see \cref{E-inc-prev:sec:definitions}).
+Therefore, $\vec{\theta} = [\lambda_1, \dots, \lambda_{\dmax-1}]^T$.
+
 Priors cannot be vague on both the hazard and the survival, the implications of which I consider in what follows.
@@
 The central 95\% probability mass of Beta(0.1, 1.9) is 0.00--0.47.
-The central estimate, of 0.05, is in line with previous estimates that the median duration i in the range 15--20 days~\autocite{cevikShedding}.
+The central estimate, of 0.05, is in line with previous estimates that the median duration in the range 15--20 days~\autocite{cevikShedding}.
 This prior gives a very vague prior predictive distribution on the survival time (see \cref{perf-test:fig:vague-prior}).
@@
 Smoothing priors encode the biological consideration that the hazard should not vary significantly from day-to-day.
-It is the Bayesian equivalent of a penalized likelihood, used frequently in the non-parametric frequentist setting~\autocite[e.g.][]{bacchettiNonparametric}.
+It is the Bayesian equivalent of a penalized likelihood, which is used frequently in the non-parametric frequentist setting~\autocite[e.g.][]{bacchettiNonparametric}.
 Many possible forms of smoothing priors exist, including splines, Gaussian Processes, and random walks.
@@
 \end{align}
-The hyperparameters $\mu_{\lambda_1}$ and $\sigma_{\lambda_1}$ specify the prior on the hazard at time 1; $\mu_{\epsilon_1}$ and $\sigma_{\epsilon_1}$ the prior on the initial gradient; and $\mu_\sigma$ the smoothness of the random walk.
+The hyperparameters $\mu_{\lambda_1}$ and $\sigma_{\lambda_1}$ specify the prior on the hazard at time 1; $\mu_{\epsilon_1}$ and $\sigma_{\epsilon_1}$ the prior on the initial gradient; and $\mu_\sigma$ represents the smoothness of the random walk.
 I use $\mu_{\lambda_1} = -17.5$, $\sigma_{\lambda_1} = 6$, $\mu_\epsilon = 1.09$, $\sigma_\epsilon = 0.03$, and $\mu_\epsilon = 0.1$ based on the analysis in \cref{E-ATACCC} (see \cref{perf-test:fig:rw2-prior}).
@@
   \caption[RW2 prior for the hazard]{The prior predictive survival time (mean and 95\% central interval) for the RW2 prior specified in the main text.
-The posterior mean estimate using ATACCC data in \cref{E-ATACCC} are shown for comparison.
+The posterior mean estimate using ATACCC data in \cref{E-ATACCC} is shown for comparison.
 \label{perf-test:fig:rw2-prior}}

### 2024-03-22 cis-imperfect-testing.tex

@@
 If this is true, then the duration is $d_j = e_j - b_j + 1 = 1$.
-Additionally, short episodes are very likely to be missed, which means that the design of the CIS cannot rule out many of them occurring.
+Additionally, short episodes are very likely to be undetected, which means that the design of the CIS cannot rule out many of them occurring.
 % This produced the same pattern, that is estimating too many short episodes (not shown).
@@
 
-The CIS is a more complex setting, including double interval censoring and a complex pattern of missed episodes (see \cref{E-perf-test:sec:problem}).
+The CIS is a more complex setting, including double interval censoring and a complex pattern of undetected episodes (see \cref{E-perf-test:sec:problem}).
 This means the result cannot simply be extended to the CIS, but still reinforces the intuition that single positive episodes are likely to be problematic.
@@
 First, in \cref{imperf-test:sec:modifying-p_ia}, I modify $p_{ia}$, the probability of an episode with test schedule $i$ being admissible, to allow for the episode possibly being longer than observed.
-Then, in \cref{imperf-test:sec:modifying-p_iu}, I modify $p_{iu}$, the probability of an episode with test schedule $i$ being undetected, to allow for additional episodes being missed.
+Then, in \cref{imperf-test:sec:modifying-p_iu}, I modify $p_{iu}$, the probability of an episode with test schedule $i$ being undetected, to allow for additional episodes being undetected.
 
@@
 Denote the replacement for $p_{iu}$ as $p_{iu}'$.
-$p_{iu}'$ is the probability of episode $i$ being missed, considering both the mechanisms from \cref{E-perf-test} and the new mechanism here.
+$p_{iu}'$ is the probability of episode $i$ being undetected, considering both the mechanisms from \cref{E-perf-test} and the new mechanism here.
 These mechanisms are mutually exclusive.
@@
 The number of episodes inferred to have truly ended by the first negative is too high, and hence the survival time is overestimated.
-This effect dominates over the opposing bias of overestimating the number of missed episodes.
+This effect dominates over the opposing bias of overestimating the number of undetected episodes.
 The opposite occurs if the test sensitivity is assumed to be too high, although the posterior moves away from the truth earlier (see \cref{imperf-test:fig:misspecified-test-sensitivity}(C)).
@@
 In particular, the test sensitivity for long episodes may become much lower than 50\%.
-This allows many more episodes to be missed than the model assumes, and hence a higher $\ntot$ is required to explain the data.
+This allows many more episodes to be undetected than the model assumes, and hence a higher $\ntot$ is required to explain the data.
 Furthermore, it violates assumptions made in \cref{imperf-test:sec:modelling}; this is similar to when the simulating using a too low test sensitivity.
@@
 % However, comparing the CIS and simulated datasets do not show any differences which suggest what in the data is causing these differences.
-% Then, the additional missed infections implied by a 
-% This would lead to more missed infections than the model assumes, and hence a higher $\ntot$ is required to explain the data.
+% Then, the additional undetected infections implied by a 
+% This would lead to more undetected infections than the model assumes, and hence a higher $\ntot$ is required to explain the data.
 

### 2024-03-22 cis-perfect-testing.tex

--- 2024-03-22 cis-perfect-testing.tex
@@
 In this section I specify the form of $S_{\vec{\theta}}(t)$.
-% As frequently observed in the literature~\autocite[e.g.][]{heBayesian}, it is more convenient to operate with the hazard than the survival or probability mass function directly.
-% The mass function must sum to one.
-% The survival must be monotonically decreasing
-% But, there are no constraints across multiple hazards.
-% This would complicate the inference process in a way that cannot be removed through a simple transformation.
-% The only constraint on the hazard is to be in the interval $[0, 1]$, since they are probabilities.
-% The [0, 1] interval is easily be mapped to the unconstrained reals using a logit transformation (where $\logit(x) = \log(x/(1-x))$).
+I parameterize $S$ in terms of the hazard $S_{\vec{\theta}}(t) = \prod_{i=1}^{t-1} (1 - \lambda(i))$ (see \cref{E-inc-prev:sec:definitions}).
+Therefore, $\vec{\theta} = [\lambda_1, \dots, \lambda_{\dmax-1}]^T$.
+
 Priors cannot be vague on both the hazard and the survival, the implications of which I consider in what follows.

### 2024-03-22 introduction.tex
minor to dos

@@
 
-Next, in \cref{E-perf-test}, I estimate the duration of RT-PCR-positivity from double censored data with arbitrary periods where infections could be missed (a generalisation of truncation).
+Next, in \cref{E-perf-test}, I estimate the duration of RT-PCR-positivity from double censored data with arbitrary periods where infections could be undetected (a generalisation of truncation).
 This is based on the CIS data.
@@
 Therefore, the code for the analyses in this thesis is available online.
-An index to the various repositories is available at\todo{insert GitHub link}.
+An index to the code and, where possible, data is available at \url{https://github.com/joshuablake/thesis-index}.
 I structured two elements of the code as R packages to enable their application to future contexts.

### 2024-03-22 transmission-appendix-INLA.tex

@@
 
-INLA (Integrated Nested Laplace Approximation) is a method for approximate Bayesian inference.
+INLA (Integrated Nested Laplace Approximation) is a method for approximate Bayesian inference for LGMs (latent Gaussian models).
+LGMs are a broad class of models, including all generalised linear mixed models and generalised linear additive mixed models, as well as many more complex models.
+
 It was introduced by \textcite{rueINLA}.
@@
 
-INLA requires a LGM (latent Gaussian model).
-An LGM is defined in terms of a vector of hyperparameters, $\vec{\varphi}$; a deterministic transformation of the hyperparameters into covariance matrix, $\matr{\Sigma}(\vec{\varphi})$; a vector of latent variables, $\vec{x}$; a vector of $n$ data points, $\vec{y} = [y_1, \dots, y_n]^T$; and $n$ linear predictors $\vec{\eta} = [\eta_1, \dots, \eta_n]^T$.
-To be an LGM, $\vec{x}$ must be a latent Gaussian field, that is $\vec{x} \mid \vec{\varphi} \dist \N(\vec{0}, \matr{\Sigma}(\vec{\varphi}))$; and the likelihood must have the conditional independence property $\pi(\vec{y} \mid \vec{\varphi}, \vec{x}) = \prod_j \pi(y_j \mid \eta_j(\vec{x}), \vec{\varphi})$.
+An LGM is a model for a series of observations $\vec{y} = [y_1, \dots, y_n]^T$.
+Each observation has an associated linear predictor $\vec{\eta} = [\eta_1, \dots, \eta_n]^T$.
+The linear predictor and, optionally, some hyperparameters $\vec{\theta}$ give the conditional independence structure:
+\begin{align}
+    \pi(\vec{y} \mid \vec{\eta}, \vec{\theta}) = \prod_{i=1}^n \pi(y_i \mid \eta_i, \vec{\theta}).
+\end{align}
+Such a structure is familiar from linear regression models which have $\vec{\eta} = \matr{X} \vec{\beta}$, where $\matr{X}$ is the design matrix and $\vec{\beta}$ are the coefficients; $\vec{\theta}$ is one-dimensional; and $\pi(y_i \mid \eta_i, \vec{\theta})$ is the normal distribution with mean $\eta_i$ and standard deviation $\theta$.
+
+An LGM introduces a latent field $\vec{x}$ with a multivariate Gaussian distribution (hence the name):
+\begin{align}
+    \vec{x} \mid \vec{\theta} \dist \N(\vec{0}, \matr{\Sigma}(\vec{\theta}))
+\end{align}
+where $\Sigma(\vec{\theta})$ is the covariance matrix formed as a deterministic transformation of the hyperparameters.
+The linear predictors must be able to be expressed as a linear transformation of the latent field, \ie $\vec{\eta} = \matr{A} \vec{x}$ for some pre-specified matrix $\matr{A}$.
+For the linear regression model, $\matr{A} = \matr{X}$, $\vec{x} = \vec{\beta}$, and $\matr{\Sigma}(\vec{\theta})$ would be a fixed diagonal matrix specifying the prior on the regression coefficients.
+In the (generalised) linear model case $\vec{\theta}$ is fixed; however, LGMs allow an arbitrary prior distribution on $\vec{\theta}$ which means that some elements of $\vec{x}$ can be random effects (or more complex structures).
+
+In summary, the full specification of an LGM is:
+\begin{align}
+    \vec{\theta} &\dist \pi(\vec{\theta}) \\
+    \vec{x} \mid \vec{\theta} &\dist \N(\vec{0}, \matr{\Sigma}(\vec{\theta})) \\
+    \vec{\eta} &= \matr{A} \vec{x} \\
+    \pi(\vec{y} \mid \vec{\eta}, \vec{\theta}) &= \prod_{i=1}^n \pi(y_i \mid \eta_i, \vec{\theta})
+\end{align}
+where $\pi(\vec{\theta})$ is an arbitrary prior distribution on the hyperparameters.
+
+% An LGM is defined in terms of a vector of hyperparameters, $\vec{\varphi}$; a deterministic transformation of the hyperparameters into covariance matrix, $\matr{\Sigma}(\vec{\varphi})$; a vector of latent variables, $\vec{x}$; a vector of $n$ data points, $\vec{y} = [y_1, \dots, y_n]^T$; and $n$ linear predictors $\vec{\eta} = [\eta_1, \dots, \eta_n]^T$.
+% To be an LGM, $\vec{x}$ must be a latent Gaussian field, that is $\vec{x} \mid \vec{\varphi} \dist \N(\vec{0}, \matr{\Sigma}(\vec{\varphi}))$; and the likelihood must have the conditional independence property $\pi(\vec{y} \mid \vec{\varphi}, \vec{x}) = \prod_j \pi(y_j \mid \eta_j(\vec{x}), \vec{\varphi})$.
 % The Gaussian structure allows the use of the approximations in \cref{transmission:sec:INLA:inference}.
@@
 
-For INLA to perform well, \ie be compuutationally efficient and an accurate approximation, $\vec{\theta}$ should be of low dimension, \textcite{martinoINLAtutorial} recommend $< 15$; and the precision matrix, $\matr{Q}$, should be sparse.
+For INLA to perform well, \ie be compuutationally efficient and an accurate approximation, $\vec{\theta}$ should be of low dimension (\textcite{martinoINLAtutorial} recommend $< 15$); and the precision matrix, $\matr{\Sigma}^{-1}$, should be sparse.
+However, the dimension of $\vec{x}$ is unrestricted, allowing a large number of random effects, or more complex structures.
 The low dimensionality of $\vec{\theta}$ is because the method requires numerical integration over it.
-The sparsity of $\matr{Q}$ allows for efficient computation.
+The sparsity of $\matr{\Sigma}^{-1}$ allows for efficient computation.
 Sparsity is introduced by conditional independence structures of the latent field that are present in many models.
 For example, in the classic random effect setting, all random effects are independent conditional on the hyperparameters.
+
+% \subsection{Application}
+
+% Next, I explain why the inference-inferred approach proposed in \cref{backcalc:sec:approach} cannot be expressed as an LGM.
+% In this context, each $i$ refers to a specific stratum and day combination.
+% The number of positive test results in this stratum and day is $y_i$.
+% $\eta_i$ is the population prevalence in this stratum on this day.
+% Hence, $y_i \mid \eta_i, \rho \dist \BB(n_i, \eta_i, \rho)$ where $\rho$ is a hyperparameter (in $\vec{\theta}$) determining the overdispersion of the beta-binomial distribution and $n_i$ is the number of .
+
+% In this context, each $w_j$ corresponds to the observed RT-PCR results in one stratum on one day.
+% Hence, $\eta_{j}$ must determine the distribution of $w_{j}$ (up to some hyperparameters); where $j$ uniquely identifies a stratum $i$ and day $t$.
+
+% In practice, this means that $\eta_j$ is a one-to-one transformation of the population prevalence in stratum $i$ on day $t$.
+% If $\eta_j = \logit(\pi_{i,t})$, where $\pi$ is prevalence, then a model of $\eta_j = g(i, t)$, as before, is an LGM.
+% However, if $\pi$ is instead representing incidence proportion, then no such linear relationship between $\eta_j$ and $g$ exists.
+% In particular, if $\eta_j$ is the prevalence, then:
+% \begin{align}
+%     \eta_{i,t}
+%     &= \sum_{k=-\infty}^0 \pi_{ik} S(k) \\
+%     &= \sum_{k=-\infty}^0 \expit(g(i, k)) S(k).
+% \end{align}
+% As $g(i, k)$ is linear in the model parameters, $\vec{z}$, then $\eta_{i,t}$ is not, violating the requirements of an LGM.
 
@@
 \pi(x_i \mid \vec{y})
-&\approx \int \pi(x_i \mid \vec{\theta}, \vec{y}) \tilde\pi(\vec{\theta} \mid \vec{y}) d\vec{\theta}.
+&\approx \int \pi(x_i \mid \vec{\theta}, \vec{y}) \tilde\pi(\vec{\theta} \mid \vec{y}) d\vec{\theta}
 \end{align}
-which can be computed using numerical integration over $\tilde\pi(\vec{\theta} \mid \vec{y})$, previously calculated.
+computed using numerical integration.
 
@@
 INLA assumes that $\vec{x} \mid \vec{y}, \vec{\theta}$ is approximately Gaussian, and therefore takes a Laplace approximation to the distribution.
-The greater the smoothing, the better the approximation.
+The approximation then proceeds as before.
+Exploiting the structure of the problem, in particular the conditional independence structure, the Laplace approximation can be computed efficiently.
+
+The greater the smoothing, the better the Laplace approximation.
 This is because, if $\vec{x} \mid \vec{y}, \vec{\theta}$ is then dominated by the prior $\vec{x} \mid \vec{\theta}$ and the prior is Gaussian.
-Exploiting the structure of the problem, in particular the conditional independence structure, the Laplace approximation can be computed efficiently.
 

### 2024-03-22 transmission.tex

@@
 Therefore, satisfying all the desirable properties requires the prevalence-inferred approach, which I adopt.
-
-\todo[inline]{Here to the end of the section moved to Appendix.}
-
-An LGM is defined in terms of a vector of hyperparameters, $\vec{\varphi}$; a deterministic transformation of the hyperparameters into covariance matrix, $\matr{\Sigma}(\vec{\varphi})$; a vector of latent variables, $\vec{z}$; a vector of $n$ data points, $\vec{w} = [w_1, \dots, w_n]^T$; and $n$ linear predictors $\vec{\eta} = [\eta_1, \dots, \eta_n]^T$.
-To be an LGM, $\vec{z}$ must be a latent Gaussian field, that is $\vec{z} \mid \vec{\varphi} \dist \N(\vec{0}, \matr{\Sigma}(\vec{\varphi}))$; and the likelihood must have the conditional independence property $p(\vec{w} \mid \vec{\varphi}, \vec{z}) = \prod_j p(w_j \mid \eta_j(\vec{z}), \vec{\varphi})$.
-
-In this context, each $w_j$ corresponds to the observed RT-PCR results in one stratum on one day.
-Hence, $\eta_{j}$ must determine the distribution of $w_{j}$ (up to some hyperparameters); where $j$ uniquely identifies a stratum $i$ and day $t$.
-
-In practice, this means that $\eta_j$ is a one-to-one transformation of the population prevalence in stratum $i$ on day $t$.
-If $\eta_j = \logit(\pi_{i,t})$, where $\pi$ is prevalence, then a model of $\eta_j = g(i, t)$, as before, is an LGM.
-However, if $\pi$ is instead representing incidence proportion, then no such linear relationship between $\eta_j$ and $g$ exists.
-In particular, if $\eta_j$ is the prevalence, then:
-\begin{align}
-    \eta_{i,t}
-    &= \sum_{k=-\infty}^0 \pi_{ik} S(k) \\
-    &= \sum_{k=-\infty}^0 \expit(g(i, k)) S(k).
-\end{align}
-As $g(i, k)$ is linear in the model parameters, $\vec{z}$, then $\eta_{i,t}$ is not, violating the requirements of an LGM.
 

### 2024-03-22 cis-perfect-testing.tex
Updates from Overleaf

@@
 under the assumption of uniform probability of infection time.
-This is the standard form of the likelihood for doubly interval censored data without truncation.
+This is the standard form of the likelihood for doubly interval censored data without truncation~\autocite[e.g.][]{sunEmpirical}.
 
@@
 
-Smoothing priors encode the biological consideration that the hazard should not vary significantly from day-to-day.
-It is the Bayesian equivalent of a penalized likelihood, which is used frequently in the non-parametric frequentist setting~\autocite[e.g.][]{bacchettiNonparametric}.
+Smoothing priors~\autocite{gerschSmoothness} encode the biological consideration that the hazard should not vary significantly from day-to-day.
+They serve the same purpose as a penalized likelihood, which is used frequently in the non-parametric maximum-likelihood setting~\autocite[e.g.][]{bacchettiNonparametric}.
 Many possible forms of smoothing priors exist, including splines, Gaussian Processes, and random walks.
@@
 \end{align}
-The hyperparameters $\mu_{\lambda_1}$ and $\sigma_{\lambda_1}$ specify the prior on the hazard at time 1; $\mu_{\epsilon_1}$ and $\sigma_{\epsilon_1}$ the prior on the initial gradient; and $\mu_\sigma$ represents the smoothness of the random walk.
+The hyperparameters $\mu_{\lambda_1}$ and $\sigma_{\lambda_1}$ specify the prior on the hazard at time 1; $\mu_{\epsilon_1}$ and $\sigma_{\epsilon_1}$ specify the prior on the initial gradient, $\lambda_2 - \lambda_1$; and $\mu_\sigma$ specifies the smoothness of the random walk.
 I use $\mu_{\lambda_1} = -17.5$, $\sigma_{\lambda_1} = 6$, $\mu_\epsilon = 1.09$, $\sigma_\epsilon = 0.03$, and $\mu_\epsilon = 0.1$ based on the analysis in \cref{E-ATACCC} (see \cref{perf-test:fig:rw2-prior}).
@@
 
-I first approximate the \cref{E-ATACCC} posterior estimate of the hazard as a multivariate normal on the logit scale.
-This can be viewed as an approximation of Markov melding~\autocite{goudieJoining}.
-Using a multivariate normal, as opposed to multiple univariate distribution preserves the correlation between the hazards.
+I first approximate the \cref{E-ATACCC} posterior estimate of the hazard as $\logit{\vec{h}} \dist \MNorm(\vec{\mu}_A, \matr{\Sigma}_A)$ where $\vec{h}$ is the posterior distribution of the hazards, and $\vec{\mu}_A$ and $\matr{\Sigma}_A$ are the mean and covariance matrix computed from the samples of $\logit{\vec{h}}$ derived using the method of \cref{ATACCC:sec:duration}.
+Using a multivariate normal, as opposed to multiple univariate distribution for each $h_t$, preserves the correlation between the hazards.
 The logit transformation maps from the $[0, 1]$ interval to the full real line.
-To estimate the parameters of the multivariate normal, I use the method of moments.
 The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC-hazard} and \cref{perf-test:fig:approximate-ATACCC-survival}).
@@
 \begin{align}
-  \lambda_t &\sim \text{Beta}(\alpha_t, \beta_t) &t = 1, 2, \dots \\
+  \logit \vec{h} &\dist \MNorm(\vec{\mu}_A, \matr{\Sigma}_A) \\
+  \lambda_t &\dist \text{Beta}(\alpha_t, \beta_t) &t = 1, 2, \dots \\
   \alpha_t &= k_t h_t + \alpha_0 \\
@@
 \end{align}
-where $k_t$, $\alpha_0$, and $\beta_0$ are hyperparameters; and $h_t$ is a point estimate of the hazard at time $t$ from ATACCC.
+where $k_t$, $\alpha_0$, and $\beta_0$ are hyperparameters; and $h_t$ is an estimate of the hazard at time $t$ from ATACCC.
 An intuition for what this distribution represents derives from a conjugate model for $\lambda_t$ with a beta prior and a binomial likelihood.

### 2024-03-22 cis-imperfect-testing.tex
rewrite ch 6

@@
 
-I start by finding the tests that need consideration.
-For tractability and simplicity, I consider only tests between the negative tests providing an upper bound on the length of the episode.
-By definition (see \cref{E-biology-data:sec:cis-episodes}), these are the tests between $l_{j(i)}^{(b)} - 1$ and $r_{j(i)}^{(e)} + 1$ inclusive.
-They are a subset of $\sched_i$, the set of test times for the individual in which episode $i$ occurs.
+I start by finding the tests that need consideration for episode $j$.
+For tractability and simplicity, I consider only tests between the negative tests providing an upper bound on the length of episode $j$.
+By definition (see \cref{E-biology-data:sec:cis-episodes}), these are the tests in individual $i(j)$ between $l_j^{(b)} - 1$ and $r_j^{(e)} + 1$ inclusive.
 
-For tractability, assume that the negative test bounding the start of the episode, on day $l_{j(i)}^{(b)}-1$, is a true negative.
-This assumption is reasonable because, since a positive test follows at $r_{j(i)}^{(b)}$, the negative at $l_{j(i)}^{(b)}-1$ is likely early in the infection and the test sensitivity is high early in an infection.
+For tractability, assume that the negative test bounding the start of the episode, on day $l_j^{(b)}-1$, is a true negative.
+This assumption is reasonable because, since a positive test follows at $r_j^{(b)}$, the negative at $l_j^{(b)}-1$ is likely early in the infection and the test sensitivity is high early in an infection.
 Therefore, this is unlikely to be a false negative because, if the individual was detectable at this time, they are in a period with high test sensitivity.
-This true negative occurs with probability 1, and hence does not contribute to the likelihood.
-Denote the remaining tests considered as $\sched'_i = \{ t \in \sched_i \ssep r_{j(i)}^{(b)} \leq t \leq r_{j(i)}^{(e)} + 1 \}$, and their results by $\vec{y_{j(i)}} = \{ y_{j(i)}(t) \ssep t \in \sched'_i \}$ where $y_{j(i)}(t) = 1$ if the test on day $t$ is positive and 0 otherwise.
+True negatives occur with probability 1, and hence this test does not contribute to the likelihood.
 
-As I assume that there are no false positives, the infection episode must span at least the period $[r^{(b)}_{j(i)}, l^{(e)}_{j(i)}]$ because it starts and ends with a positive test.
-This includes all $t \in \sched'_i$ except $t = r_{j(i)}^{(e)}+1$.
-Therefore, the test results at times in $\sched_i'$, except the test at $r_{j(i)}^{(e)}+1$, are either true positives or false negatives.
+Denote the set of the remaining test times to consider as $\sched'_{i(j)}$.
+$\sched'_{i(j)}$ is the set of times individual $i(j)$ was tested between $l_j^{(b)}$ and $r_j^{(e)} + 1$ inclusive.
+Denote their results by $\vec{y}_j = \{ y_{i(j)}(t) \ssep t \in \sched'_{i(j)} \}$ where $y_{i(j)}(t) = 1$ if the test on day $t$ in individual $i(j)$ is positive and 0 otherwise.
 
-Consider the negative test at $r_{j(i)}^{(e)}+1$, the first negative after the start of the episode which may be a false negative.
-It is a false negative if and only if the episode ends after the test, \ie $e_{j(i)} > r_{j(i)}^{(e)}$.
-I proceed by considering whether this is the case and considering the case with $b_{j(i)}$ known; $b_{j(i)}$ will later be integrated out.
+As I assume that there are no false positives, the infection episode must span at least the period $[r^{(b)}_j, l^{(e)}_j]$, because this period starts and ends with a positive test.
+This includes all $t \in \sched'_{i(j)}$ except $t = r_j^{(e)}+1$.
+Therefore, the test results $\vec{y}_j$, except the test at $r_j^{(e)}+1$, are either true positives or false negatives.
 
-First, if $e_{j(i)} \leq r_{j(i)}^{(e)}$, meaning that the test at $r_{j(i)}^{(e)}+1$ is a true negative and the end of the episode is interval censored as in the previous chapter.
-% In this case, the test at $r_{j(i)}^{(e)} + 1$ is a true negative, as are all other tests not in $\sched'_i$.
+Consider the negative test at $r_j^{(e)}+1$, the first negative after the start of the episode which may be a false negative.
+It is a false negative if and only if the episode ends after the test, \ie $E_j > r_j^{(e)}$.
+I proceed by considering whether this is the case and with $B_j = b$ known.
+
+First, if $E_j \leq r_j^{(e)}$, meaning that the test at $r_j^{(e)}+1$ is a true negative and the end of the episode is interval censored as in the previous chapter.
+% In this case, the test at $r_j^{(e)} + 1$ is a true negative, as are all other tests not in $\sched'_{i(j)}$.
 The true negative occurs with probability 1, by the assumption of no false positives.
 \begin{align}
-&p(\vec{y_{j(i)}}, e_{j(i)} \leq r_{j(i)}^{(e)} | b_{j(i)}, p_\text{sens}, \vec{\theta}) \\
-&= p(\vec{y_{j(i)}}, l_{j(i)}^{(e)} \leq e_{j(i)} \leq r_{j(i)}^{(e)} | t_i, b_{j(i)}, p_\text{sens}, \vec{\theta}) \\ % &\text{as no false positives}
-&= p(\vec{y_{j(i)}} \mid l_{j(i)}^{(e)} \leq e_{j(i)} \leq r_{j(i)}^{(e)}, t_i, b_{j(i)}, p_\text{sens}, \vec{\theta}) p(l_{j(i)}^{(e)} \leq e_{j(i)} \leq r_{j(i)}^{(e)} | t_i, b_{j(i)}, p_\text{sens}, \vec{\theta}) \\
-&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_{j(i)}(t)} (1 - p_\text{sens})^{(1 - y_{j(i)}(t))} \right) \left( S_{\vec{\theta}}(l_{j(i)}^{(e)} - b_{j(i)} - 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b_{j(i)} - 1) \right)
+&\prob(\vec{y_j}, E_j \leq r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
+&= \prob(\vec{y_j}, l_j^{(e)} \leq E_j \leq r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\ % &\text{as no false positives}
+&= \prob(\vec{y_j} \mid l_j^{(e)} \leq E_j \leq r_j^{(e)}, B_j = b, p_\text{sens}, \vec{\theta}) \prob(l_j^{(e)} \leq E_j \leq r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
+&= \left( \prod_{t \in \sched'_{i(j)}} p_\text{sens}^{y_{i(j)}(t)} (1 - p_\text{sens})^{(1 - y_{i(j)}(t))} \right) \left( S_{\vec{\theta}}(l_j^{(e)} - b + 1) - S_{\vec{\theta}}(r_j^{(e)} - b + 2) \right)
 \label{imperf-test:eq:ll-ei-lt-ri}
@@
 
-Second, if $e_{j(i)} > r_{j(i)}^{(e)}$.
-In this case, the test at $r_{j(i)}^{(e)}$ is a false negative, occurring with probability $(1 - p_\text{sens})$.
-To avoid having to consider tests after $r_{j(i)}^{(e)}$, which could greatly complicate the likelihood, I model this case as the episode being right censored at $r_{j(i)}^{(e)}$.
+Second, if $E_j > r_j^{(e)}$.
+In this case, the test at $r_j^{(e)}$ is a false negative, occurring with probability $(1 - p_\text{sens})$.
+To avoid having to consider tests after $r_j^{(e)}$, which could greatly complicate the likelihood, I model this case as the episode being right censored at $r_j^{(e)}$.
 Taking the same approach as before:
 \begin{align}
-&p(\vec{y_{j(i)}}, e_{j(i)} > r_{j(i)}^{(e)} | b_{j(i)}, p_\text{sens}, \vec{\theta}) \\
-&= p(\vec{y_{j(i)}} \mid e_{j(i)} > r_{j(i)}^{(e)}, t_i, b_{j(i)}, p_\text{sens}, \vec{\theta}) p(e_{j(i)} > r_{j(i)}^{(e)} | t_i, b_{j(i)}, p_\text{sens}, \vec{\theta}) \\
-&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_{j(i)}(t)} (1 - p_\text{sens})^{(1 - y_{j(i)}(t))} \right) (1 - p_\text{sens}) S_{\vec{\theta}}(r_{j(i)}^{(e)} - b_{j(i)} - 1)
+&\prob(\vec{y_j}, E_j > r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
+&= \prob(\vec{y_j} \mid E_j > r_j^{(e)}, B_j = b, p_\text{sens}, \vec{\theta}) \prob(E_j > r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
+&= \left( \prod_{t \in \sched'_{i(j)}} p_\text{sens}^{y_{i(j)}(t)} (1 - p_\text{sens})^{(1 - y_{i(j)}(t))} \right) (1 - p_\text{sens}) S_{\vec{\theta}}(r_j^{(e)} - b_j + 2)
 \label{imperf-test:eq:ll-ei-gt-ri}
@@
 
-These expressions can now be used to form the full replacement for $p_{ia}$, $p'_{ia}$, the likelihood of the data $\vec{y_{j(i)}}$.
-First, augment the data with $b_{j(i)}$, and split into the cases just discussed:
+These expressions can now be used to form the full replacement for $p_{ia}$, $p'_{ia}$, the likelihood of the data $\vec{y_j}$.
+First, augment the data with $b_j$, and split into the cases just discussed:
 \begin{align}
 p_{ia}'
-&= p(\vec{y_{j(i)}} \mid p_\text{sens}, \vec{\theta}) \\
-&= \sum_{b_{j(i)} = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( p(\vec{y_{j(i)}}, e_{j(i)} \leq r_{j(i)}^{(e)} \mid b_{j(i)}, p_\text{sens}, \vec{\theta}) + p(\vec{y_{j(i)}}, e_{j(i)} > r_{j(i)}^{(e)} \mid b_{j(i)}, p_\text{sens}, \vec{\theta}) \right) p(b_{j(i)} \mid p_\text{sens}, \vec{\theta}).
-\intertext{Now, substitute in \cref{imperf-test:eq:ll-ei-lt-ri,imperf-test:eq:ll-ei-gt-ri}, then take out the common factor:}
-&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_{j(i)}(t)} (1 - p_\text{sens})^{(1 - y_{j(i)}(t))} \right) \\ & \ \times \sum_{b_{j(i)} = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_{j(i)}^{(e)} - b_{j(i)} - 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b_{j(i)} - 1) + (1 - p_\text{sens}) S_{\vec{\theta}}(r_{j(i)}^{(e)} - b_{j(i)} - 1) \right) \\ & \ \times p(b_{j(i)} \mid p_\text{sens}, \vec{\theta}) \\
-&= \left( \prod_{t \in \sched'_i} p_\text{sens}^{y_{j(i)}(t)} (1 - p_\text{sens})^{(1 - y_{j(i)}(t))} \right) \\ & \ \times \sum_{b_{j(i)} = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_{j(i)}^{(e)} - b_{j(i)} - 1) - p_\text{sens} S_{\vec{\theta}}(r_{j(i)}^{(e)} - b_{j(i)} - 1) \right) p(b_{j(i)} \mid p_\text{sens}, \vec{\theta}).
+=& \prob(\vec{y_j} \mid p_\text{sens}, \vec{\theta}) \\
+=& \sum_{b = l_j^{(b)}}^{r_j^{(b)}} \left( \prob(\vec{y_j}, E_j \leq r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) + \prob(\vec{y_j}, E_j > r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \right) \\
+  &\times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}).
+\intertext{Now, substitute in \cref{imperf-test:eq:ll-ei-lt-ri,imperf-test:eq:ll-ei-gt-ri} and take out the common factor:}
+=& \left( \prod_{t \in \sched'_{i(j)}} p_\text{sens}^{y_{i(j)}(t)} (1 - p_\text{sens})^{(1 - y_{i(j)}(t))} \right) \\ & \ \times \sum_{b = l_j^{(b)}}^{r_j^{(b)}} \left( S_{\vec{\theta}}(l_j^{(e)} - b + 1) - S_{\vec{\theta}}(r_j^{(e)} - b + 2) + (1 - p_\text{sens}) S_{\vec{\theta}}(r_j^{(e)} - b + 2) \right) \\
+  & \times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}) \\
+=& \left( \prod_{t \in \sched'_{i(j)}} p_\text{sens}^{y_{i(j)}(t)} (1 - p_\text{sens})^{(1 - y_{i(j)}(t))} \right) \\
+  & \times \sum_{b_j = l_j^{(b)}}^{r_j^{(b)}} \left( S_{\vec{\theta}}(l_j^{(e)} - b_j + 1) - p_\text{sens} S_{\vec{\theta}}(r_j^{(e)} - b_j + 2) \right) \\
+  & \times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}).
 \label{imperf-test:eq:pia-prime}
@@
 
-If $\psens$ is fixed (\ie has a point prior) and $p(b_{j(i)} \mid \psens, \vec{\theta}) \propto 1$ (as assumed in \cref{E-perf-test}) then:
+If $\psens$ is fixed (\ie has a point prior) and $p(b_j \mid \psens, \vec{\theta}) \propto 1$ (as assumed in \cref{E-perf-test}) then:
 \begin{align}
 p_{ia}'
-&\propto \sum_{b_{j(i)} = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} S_{\vec{\theta}}(l_{j(i)}^{(e)} - b_{j(i)} - 1) - p_\text{sens} S_{\vec{\theta}}(r_{j(i)}^{(e)} - b_{j(i)} - 1).
+&\propto \sum_{b = l_j^{(b)}}^{r_j^{(b)}} S_{\vec{\theta}}(l_j^{(e)} - b_j + 1) - p_\text{sens} S_{\vec{\theta}}(r_j^{(e)} - b + 2).
 \label{imperf-test:eq:pia-prime-constant}
@@
 
-Several mechanisms for episodes being undetected were considered when deriving $p_{iu}$ in \cref{perf-test:eq:piu}.
+Several mechanisms for episodes being undetected were considered when deriving $p_{iu}$ in \cref{E-perf-test:sec:prob-undetected}.
 I now consider the additional mechanisms arising due to false negatives.
-Specifically, episode $i$ could be undetected if the first test after the $b_{j(i)}$, episode $i$'s start day, is a false negative and then there are no subsequent positive tests.
+Specifically, episode $j$ could be undetected if the first test after $b_j$ is a false negative and then there are no subsequent positive tests.
 
-This false negative would occur at the first test after the infection episode begins, on day $b_{j(i)} + \tau_{\sched_i}(b_{j(i)})$ (see \cref{E-perf-test:sec:model}).
+This false negative would occur at the first test after the infection episode begins, on day $b_j + \tau_{\sched_{i(j)}}(b_j)$ (see \cref{E-perf-test:sec:prob-undetected}).
 A false negative occurring requires that the episode has not yet ended but a negative still occurs.
-The episode has not yet ended at the time of the test if $e_{j(i)} = b_{j(i)} + d_i - 1$ is after the test, that is the duration of the infection $d_i > \tau_{\sched_i}(b_{j(i)})$.
+The episode has not yet ended at the time of the test if $e_j = b_j + d_j - 1$ is after the test, that is the duration of the infection $d_j > \tau_{\sched_{i(j)}}(b_j)$.
 Conditional on the episode having not yet ended, the test result is negative with probability $1 - \psens$.
 
-For there to be no subsequent positive tests, all tests up until day $e_{j(i)}$ are false negatives.
+For there to be no subsequent positive tests, all tests up until day $e_j$ are false negatives.
 I assume there is a negligible probability of two false negatives.
 Therefore, this can only occur if the episode ends before another test.
-Denote the number of days between $b_{j(i)}$ and the test following the false negative as $\tau^2_{\sched_i}(b_{j(i)}) \stackrel{\text{def}}{=} \tau_{\sched_i}(\tau_{\sched_i}(b_{j(i)}) + 1)$.
-Then, the episode ends before this test if $d_i < \tau^2_{\sched_i}(b_{j(i)})$.
+Denote the number of days between $b_j$ and the test following the false negative as $\tau^2_{\sched_{i(j)}}(b_j) \stackrel{\text{def}}{=} \tau_{\sched_{i(j)}}(\tau_{\sched_{i(j)}}(b_j) + 1)$.
+Then, the episode ends before this test if $d_j < \tau^2_{\sched_{i(j)}}(b_j)$.
 
-Therefore, this mechanism causes episode $i$ to be undetected if all the following conditions hold.
+Therefore, this mechanism causes episode $j$ to be undetected if all the following conditions hold.
 \begin{enumerate}
-    \item The episode would have been detected considering only the mechanisms in \cref{perf-test:eq:piu}. That is $b_{j(i)} > \min(\sched_i)$ and $\tau_{\sched_i}(b_{j(i)}) + b_{j(i)} \leq e_{j(i)}$.
-    \item The episode ends in the interval $[\tau_{\sched_i}(b_{j(i)}) + b_{j(i)}, \tau^2_{\sched_i}(b_{j(i)}) + b_{j(i)}]$.
-      Note that the lower bound here is exactly the bound on $e_{j(i)}$ in the previous condition.
-      Equivalently, $\tau_{\sched_i}(b_{j(i)}) + 1 \leq d_i \leq \tau^2_{\sched_i}(b_{j(i)}) + 1$.
-    \item A false negative occurs on day $\tau_{\sched_i}(b_{j(i)}) + b_{j(i)}$. Conditional on the previous condition, this occurs with probability $1 - \psens$.
+    \item The episode would have been detected considering only the mechanisms in \cref{E-perf-test:sec:prob-undetected}.
+That is $b_j > \min(\sched_{i(j)})$ and $e_j \geq \tau_{\sched_{i(j)}}(b_j) + b_j$.
+    \item The episode ends in the interval $[\tau_{\sched_{i(j)}}(b_j) + b_j, \tau^2_{\sched_{i(j)}}(b_j) + b_j]$.
+      Note that the lower bound here is exactly the bound on $e_j$ in the previous condition.
+      Equivalently, $\tau_{\sched_{i(j)}}(b_j) + 1 \leq d_j \leq \tau^2_{\sched_{i(j)}}(b_j) + 1$.
+    \item A false negative occurs on day $\tau_{\sched_{i(j)}}(b_j) + b_j$. Conditional on the previous condition, this occurs with probability $1 - \psens$.
 \end{enumerate}
 
-The probability of this occurring, conditional on $B_{j(i)} = b_{j(i)} > \min(\sched_i)$ is:
+The probability of this occurring, conditional on $B_j = b_j > \min(\sched_{i(j)})$ is:
 \begin{align}
 &\prob \left(
-    \tau_{\sched_i}(b_{j(i)}) + 1 \leq D_i \leq \tau^2_{\sched_i}(b_{j(i)}) + 1
-    \mid b_{j(i)},
-\vec{\theta} \right) (1 - \psens) \\
-&= \left( S_{\vec{\theta}}(\tau_{\sched_i}(b_{j(i)}) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_i}(b_{j(i)}) + 1) \right) (1 - \psens).
+    \tau_{\sched_{i(j)}}(b_j) + 1 \leq D_j \leq \tau^2_{\sched_{i(j)}}(b_j) + 1
+    \mid B_j = b, \vec{\theta} \right) (1 - \psens) \\
+&= \left( S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b_j) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b_j) + 1) \right) (1 - \psens).
 \end{align}
-Integrating over $b_{j(i)}$ for $\min \sched_i < b_{j(i)} \leq T$ ($T$ being the last day of the period considered), in the same way as \cref{perf-test:eq:piu}, gives:
+Summing over $b_j$ for $\min \sched_{i(j)} < b_j \leq T$ ($T$ being the last day of the period considered), in the same way as \cref{perf-test:eq:piu}, gives:
 \begin{align}
-\eta = (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_i) + 1}^T \left( S_{\vec{\theta}}(\tau_{\sched_i}(b) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_i}(b) + 1) \right).
+\zeta = (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_{i(j)}) + 1}^T \left( S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b) + 1) \right).
 \end{align}
@@
 These mechanisms are mutually exclusive.
-Hence, $p_{iu}'$ is the sum of these, $p_{iu}' = p_{iu} + \eta$.
+Hence, $p_{iu}'$ is the sum of these, $p_{iu}' = p_{iu} + \zeta$.
 For the posterior density, \cref{E-perf-test:eq:full-posterior}, $1 - p_{iu}'$ is the required quantity.
@@
 1 - p_{iu}'
-&= 1 - p_{iu} - (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_i) + 1}^T \left( S_{\vec{\theta}}(\tau_{\sched_i}(b) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_i}(b) + 1) \right) \\
-&= \frac{1}{T} \sum_{b=\min(\sched_i) + 1}^T \left( p_\text{sens} S_{\vec{\theta}}(\tau_{\sched_i}(b) + 1) + (1 - p_\text{sens}) S_{\vec{\theta}}(\tau^2_{\sched_i}(b) + 1)\right).
+&= 1 - p_{iu} - \zeta \\
+&= \frac{1}{T} \sum_{b=\min(\sched_{i(j)}) + 1}^T \left( p_\text{sens} S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b) + 1) + (1 - p_\text{sens}) S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b) + 1)\right).
 \label{imperf-test:eq:pit-prime}
@@
 Finally, the impact of the simplifying assumptions made in \cref{imperf-test:sec:modelling} for tractability.
-These assumptions are that the negative at $l^{(b)}_i - 1$ is a true negative and that there is a negligible probability of missing an episode due to two false negatives.
+These assumptions are that the negative at $l^{(b)}_j - 1$ is a true negative and that there is a negligible probability of missing an episode due to two false negatives.
 

### 2024-03-22 cis-perfect-testing.tex
rewrite ch 6

--- 2024-03-22 cis-perfect-testing.tex
@@
 =& \sum_{b = l_{j}^{(b)}}^{r_{j}^{(b)}} \prob \left( l_{j}^{(e)} - b + 1 \leq D_{j} \leq r_{j}^{(e)} - b + 1 \right) \prob \left(B_{j} = b \right) &\text{by def of $D_{j}$} \\
-=& \sum_{b = l_{j}^{(b)}}^{r_{j}^{(b)}} \left( S_{\vec{\theta}}(l_{j}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j}^{(e)} - b + 1) \right) \prob \left(B_{j} = b \right) &\text{by def of $S_{\vec{\theta}}$} \\
-\propto& \sum_{b = l_{j}^{(b)}}^{r_{j}^{(b)}} \left( S_{\vec{\theta}}(l_{j}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j}^{(e)} - b + 1) \right)
+=& \sum_{b = l_{j}^{(b)}}^{r_{j}^{(b)}} \left( S_{\vec{\theta}}(l_{j}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j}^{(e)} - b + 2) \right) \prob \left(B_{j} = b \right) &\text{by def of $S_{\vec{\theta}}$} \\
+\propto& \sum_{b = l_{j}^{(b)}}^{r_{j}^{(b)}} \left( S_{\vec{\theta}}(l_{j}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j}^{(e)} - b + 2) \right)
 \end{align}

### 2024-03-22 distributions.tex

@@
 Normal & $\N(\mu, \sigma^2)$ & $\mu$ (mean)\newline $\sigma^2 > 0$ (variance) & $\mu$ & $\sigma^2$ & $\frac{1}{\sqrt{2\pi\sigma^2}}e^{-\frac{(x-\mu)^2}{2\sigma^2}}$ \\
+\makecell{Multivariate \\ normal} & $\MNorm(\vec{\mu}, \matr{\Sigma})$ & $\mu \in \reals^k$ (mean vector)\newline $\Sigma \in \reals^{k \times k}$ (covariance matrix, positive semi-definite) & $\vec{\mu}$ & $\matr{\Sigma}$ & $(2 \pi)^{-k / 2} \operatorname{det}(\matr{\Sigma})^{-1 / 2} \exp \left(-\frac{1}{2}(\vec{x}-\vec{\mu})^T \matr{\Sigma}^{-1}(\vec{x}-\vec{\mu})\right)$ \\
 Lognormal & $\LN(\mu, \sigma^2)$ & $\mu$ (mean of the logarithm)\newline $\sigma^2 > 0$ (variance of the logarithm) & $e^{\mu + \frac{\sigma^2}{2}}$ & $(e^{\sigma^2} - 1)e^{2\mu + \sigma^2}$ & $\frac{1}{x\sigma\sqrt{2\pi}}e^{-\frac{(\ln(x)-\mu)^2}{2\sigma^2}}$ \\

### 2024-03-22 cis-perfect-testing.tex
Updates from Overleaf

--- 2024-03-22 cis-perfect-testing.tex
@@
 The information from \cref{E-ATACCC} can be incorporated through the prior.
-This is desirable because the ATACCC study frequently sampled individuals early in the infection.
-Therefore, \cref{E-ATACCC}'s estimate of the survival distribution in the early infection period should be reliable.
-In the later infection period, the ATACCC study has fewer observations and hence the estimates are less reliable.
-CIS can inform the posterior in this region.
+This is desirable because the ATACCC study frequently sampled individuals for around the first 20 days of their infection.
+Therefore, the first 20 days of \cref{E-ATACCC}'s estimate of the survival distribution is reliable.
+The CIS data can then inform the posterior after this point.
 
@@
 That is, the prior used in this analysis cannot be independent across the hazards.
-Secondly, the uncertainty in the estimates from \cref{E-ATACCC} is underestimated.
-The additional uncertainty is because the \cref{E-ATACCC} analysis extrapolates beyond the data using strong model assumptions (see \cref{E-ATACCC:sec:discussion}).
+Secondly, the uncertainty in the estimates from \cref{E-ATACCC} is underestimated because it extrapolates beyond the data using strong model assumptions (see \cref{E-ATACCC:sec:discussion}).
 %Furthermore, there are differences in the study design and laboratories used between the two studies (see \cref{E-intro:sec:studies}) which may mean that results do not generalise between the studies.
-I form the prior for the combination in two steps.
+I form the prior to include this prior information into the CIS analysis two steps.
 

### 2024-03-23 thesis.tex
Vector for theta
restructure chapter
fix heading with eta in
preface material

@@
 \usepackage[T1]{fontenc}
-\usepackage{hyperref}
+\usepackage[unicode,psdextra]{hyperref}
 \usepackage{textcomp} % provide symbols
@@
 \usepackage[UKenglish]{babel}
+\usepackage{makecell}
 
@@
 \DeclareMathOperator{\N}{N}
+\DeclareMathOperator{\MNorm}{N}
 \DeclareMathOperator{\LN}{LN}
@@
 \newcommand\set{\mathcal}
-\renewcommand{\vec}{\bm}
+\renewcommand{\vec}[1]{\bm{#1}}
 \newcommand{\ssep}{:}
@@
 \newcommand{\ncis}{\vec{n_\text{CIS}}}
-\newcommand{\na}{\vec{n_\text{a}}}
+\newcommand{\na}{\vec{n}_\text{a}}
 \newcommand{\pcis}{\vec{p_\text{CIS}}}
@@
 }
+% Declare the format for the pubstate field
+\DeclareFieldFormat{pubstate}{\mkbibparens{#1}}
+% Append pubstate to the end of the entry
+\renewbibmacro*{addendum+pubstate}{%
+  \printfield{addendum}%
+  \newunit\newblock
+  \printfield{pubstate}%
+}
 
@@
     % One of these, the CIS (Coronavirus Infection Survey), was a longitudinal survey run from April 2020 to March 2023.
-    The CIS (Coronavirus Infection Survey) was a large-scale prevalence survey with participants recruited randomly from the general population, therefore avoiding these issues.
+    The CIS (Coronavirus Infection Survey) was a large-scale prevalence survey, with participants recruited randomly from the general population, therefore avoiding these issues.
 
@@
     % I analyse the CIS in a survival analysis framework.
-    Therefore, I combine the information with the larger and longer CIS dataset.
+    To overcome this issue, I combine the information with the larger and longer CIS dataset.
     The CIS data poses several challenges because it is double interval censored and some individuals in the cohort have infections that are never detected.
-    I apply a survival analysis framework to analyse this data, which had previously only been in simpler settings.
+    I apply a survival analysis framework to analyse this data, which had previously only been used in simpler settings.
     In addition, I develop novel methodology to incorporate false negatives into the framework.
@@
 \acknowledgements{%
-    \todo[inline]{Write acknowledgements}
+    It takes a village to raise a child, and a city to write a PhD.
+    In other words, naming everyone who has helped in this process would not only be an impossible test of my memory, but also require a thesis-length acknowledgements section.
+    However, there are of course several individuals who stand out for their contribution.
+
+    First, I would like to thank my supervisors, Professor Daniela De Angelis and Dr Paul Birrell for their advice and unwavering support throughout my PhD.
+    Dani, for your experience and your constant drive to improve the rigour of my work.
+    Paul, for your constant ideas on new ways to attack a problem.
+    Our long discussions, whether online or in front of a whiteboard, were a necessary part of this thesis.
+    Without your guidance, I would have been lost and given up a long time ago.
+
+    I am extremely fortunate in that much of this work was supervised by a much larger group of people: Professor A.\ Sarah Walker, Dr Koen Pouwels, Professor Thomas House, and Professor Theodore Kypraios.
+    Sarah, your encyclopaedic knowledge of the CIS data was invaluable and your ability to get things done inspirational.
+    Koen introduced me to several ideas and methods that I would never have considered.
+    Thomas provided constant ambition and seemingly infinite ideas.
+    Theo gave insight into the black art of MCMC.
+    Thank you all for all the time you put into helping me throughout this process.
+
+    The MRC Biostatistics Unit has been a great source of support.
+    Many discussions, especially those with Drs Shaun Seaman and Brian Tom, contributed to progressing my work.
+    % I would especially like to thank Drs Shaun Seaman and Brian Tom for their help with survival analysis.
+    The admin team is a great asset to the Unit and were outstandingly supportive through the disrupted the pandemic caused.
+    My fellow PhD students have created a great environment, having peers at the same stage in my journey has been invaluable.
+    The Unit also arranged my funding, through the Bayes 4 Health grant, which I gratefully acknowledge.
+
+    % This work would not have been possible without the data from the Coronavirus Infection Survey, run by the Office for National Statistics and funded by the Department of Health and Social Care, and the Assessment of Transmission and Contagiousness of COVID-19 in Contacts study.
+    % Thank you to everyone organising, funding, and participating in these studies; none of whom necessarily endorse views expressed in this thesis.
+
+    In addition to the supervision and advice mentioned, several people contributed code or results to this work.
+    Professor A.\ Sarah Walker created the episodes dataset in \cref{biology-data:sec:cis-episodes} and performed the initial survival analysis I use for a simulation study \cref{perf-test:sec:simulation-study}.
+    Dr Koen Pouwels provided code used as the basis of the positivity model in \cref{biology-data:sec:MRP}.
+    Dr Edwin van Leeuwen created the contact matrices and Dr Sanmitra Ghosh provided code for the MCMC sampler used in \cref{E-SEIR}.
+    Thank you all for your contributions.
+
+    Of course, the support outside of work has been just as, if not more, important.
+    The many people I have met in Cambridge and have become lifelong friends have kept me sane.
+    Many of you I have met through Jesus College Boat Club or the Meridian co-working space, both inspiring groups of people.
+    I could write several pages naming all the individuals who have kept me going, given me an escape from work, and kept my spirits high.
+    You all know who you are.
+
+    There is of course one individual in this group who I must name: Hena McGhee.
+    You were always there from the very worst moments when I thought nothing would ever work, and this thesis would never be written, to the best moments of the breakthroughs.
+    I am forever grateful for your support, words of encouragement, and most of all love; without you I simply wouldn't have kept going.
+
+    Finally, I must thank my family, especially my parents, Laurence and Tracy Blake.
+    Throughout my life, no ask has ever been too much.
+    Your love and support are infinite and unconditional, your gentle nudges towards the right path irreplaceable.
+    Thank you for everything.
+
+    Having written this, I realise how fortunate I am to have been supported by such great people.
+    I can never thank you enough.
 }
@@
 \preface{%
-    \todo[inline]{Write preface}
-    \begin{itemize}
-        \item Early version of this work used in ONS reports
-        \item Chapters in JASA paper
-        \item Chapter for PNAS (or similar) journal
-    \end{itemize}
+    In October 2019 I began a PhD project entitled ``Advancing Bayesian Statistical Methods for Real-Time Epidemic Monitoring''.
+    So began a journey of reading, and an expectation of few deadlines, a lot of deep thinking, and complex statistical theory.
+    Like much of the world, the events of early 2020 changed that.
+    Following a brief interlude to provide rapid support to the response, the origins of this thesis emerged.
+
+    Several aspects of my work were done to support the ongoing response.
+    In particular, an earlier version of the backcalculation work was adopted by the Office for National Statistics to estimate incidence, published in their weekly online reports.
+    This method was also used for the analysis of several respiratory diseases over winter 2022/3~\autocite{dietzSARSCoV2final}.
+
+    I plan two publications based on the work in this thesis.
+    The analysis in \cref{perf-test,imperf-test} will be submitted to a statistical or biostatistical journal.
+    The results in \cref{transmission} will be submitted to a multidisciplinary scientific journal.
+
+    During this period, I have been involved in several other pieces of work and discussion contributing to the response, not directly used in this thesis.
+    These have led to several publications, listed in \cref{publications} for completeness.
+
+    This work contains statistical data from ONS which is Crown Copyright.
+    The use of the ONS statistical data in this work does not imply the endorsement of the ONS in relation to the interpretation or analysis of the statistical data.
+    This work uses research datasets which may not exactly reproduce National Statistics aggregates. 
 }
@@
 \subfile{distributions}
+
+\chapter{List of publications} \label{publications}
+\begin{refsection}
+\newrefcontext[sorting=ynt]
+\nocite{birrellRealtime, funkShortterm, pellisEstimation, swallow2022challenges, birrellRTM2}
+\printbibliography[heading=none]
+\end{refsection}
+
 \subfile{MCMC-appendix}

### 2024-03-23 cis-perfect-testing.tex
spell / grammar check

@@
 First, infection episodes can go undetected.
-Second, the infection episode start and end times are known imprecisely.
-I explain these issues below, and devote the rest of the chapter to the development of a method to tackle  them.
+Second, the infection episodes' start and end times are known imprecisely.
+I explain these issues below and devote the rest of the chapter to the development of a method to tackle  them.
 
@@
 tests).
-The fist individual was never infected whereas the second individual was; the latter is an undetected infection. 
+The first individual was never infected whereas the second individual has an undetected infection. 
 The undetected infection episode begins at time 33 and has a 10-day duration.
@@
 The gap between tests in the CIS (up to 28 days, see \cref{E-intro:sec:cis}) is longer than most infections (median 16 days, see \cref{E-ATACCC:sec:results}).
-Therefore, it is likely that many infection in the CIS cohort were undetected.
+Therefore, it is likely that many infections in the CIS cohort were undetected.
 This is because the interval between tests in the same individual (up to 28 days, see \cref{E-intro:sec:cis}) is longer than most infections (median 16 days, see \cref{E-ATACCC:sec:results}).
@@
 
-The second complication is that the beginning and end times of detected infection episodes is latent.
+The second complication is that the beginning and end times of detected infection episodes are latent.
 Recall, from \cref{E-biology-data:sec:cis-episodes}, that the time episode $j$ starts, $b_j$, is only known to lie in the interval $[l_j^{(b)}, r_j^{(b)}]$ where $l_j^{(b)}$ is the day after the negative test prior to the start of episode $j$ and $r_j^{(b)}$ is the first positive test in episode $j$.
@@
 To handle doubly censored data, the times need to be jointly modelled~\autocite[and references therein]{liSemiparametric}.
-If both events are interval censored then the data are \emph{doubly interval censored}.
+If both events are interval censored, then the data are \emph{doubly interval censored}.
 
@@
 
-There is also undetected infections in the CIS (see \cref{perf-test:sec:undetected}).
+There are also undetected infections in the CIS (see \cref{perf-test:sec:undetected}).
 However, the situation is subtly different from the standard truncation situation encountered in the literature.
@@
 The test schedule $\sched_i$ is the set of times individual $i$ is tested at, from their last test prior to time 1 (if it exists) up until time $T$.
-This is all the times the infection episode could be detected at and be included, plus one prior test.
+These are all the times an infection episode can be detected at, plus one prior test
 For example, for the individual in \cref{perf-test:sec:undetected}, $\sched_i = \{ 0, 7, 14, 21, 28, 56 \}$.
@@
 If $j$ is detected, then $i(j)$ is known; $b_j \in [l_j^{(b)}, r_j^{(b)}]$; and $e_j \in [l_j^{(e)}, r_j^{(e)}]$.
-These latter two properties are what makes the data doubly interval censored.
+These latter two properties are what make the data doubly interval censored.
 
@@
   Otherwise, if $i \in \set{D}$, then individual $i$ had $n_{ia} =1$ admissible episode.
-  The admissible region is defined by the intersection of the intervals $[l_j^{(b)}, r_j^{(b)}]$ and $[l_j^{(e)}, r_j^{(e)}]$.
+  The admissible region is the intersection of the intervals $[l_j^{(b)}, r_j^{(b)}]$ and $[l_j^{(e)}, r_j^{(e)}]$.
   This is the purple area in \cref{perf-test:fig:partitionSpace}; the boundaries in the $x$ direction of the area are $l_j^{(b)}$ and $r_j^{(b)}$ and in the $y$ direction are $l_j^{(e)}$ and $r_j^{(e)}$.
@@
 I assume the individual in which each episode occurs is iid uniform across the cohort.
-Therefore, the unconditional probabilities of an infection belonging to a class need to be multiplied by $1/\Ncis$.
+Therefore, these probabilities are multiplied by $1/\Ncis$ to form the unconditional probabilities.
 For example, the probability of an episode being in class $\alpha_i$ is $p_{ia} / \Ncis$.
@@
 This is not strictly correct because the cell probabilities in the likelihood should not depend on the data.
-However, an alternative where each combination of infection and recovery times is considered its own cell is possible.
+However, an alternative considering each combination of infection and recovery times as its own cell is possible.
 The new cells would align with the inadmissible region.
-Since each of these cells have an observed count of 0, they can be combined to give the same likelihood.
+Since each of these cells have an observed count of zero, they can be combined to give the same likelihood.
 That is, the approach is mathematically equivalent but greatly complicates the formulae and notation.
@@
 The undetected classes cannot be distinguished.
-Therefore, these classes can be combined into a single class.
+Therefore, I combine these classes into a single class.
 The combined class has a count $\nnodet = \sum_{i=1}^{\Ncis} n_{iu}$ and probability $\pnodet = \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} p_{iu}$.
@@
 Therefore, the crucial component is $1 - p_{iu} = 1 - \prob((B_j, E_j) \in \Omega_i \mid \vec{\theta}, i(j) = i)$.
-This is one minus the probability that an episode in individual $i$ was undetected, \ie the probability that episode such an episode is detected.
+This is one minus the probability that an episode in individual $i$ was undetected, \ie the probability of detecting an episode.
 
@@
       For episodes detected in the period I consider, this happens a negligible number of times, only in individuals lost to follow-up; therefore, I neglect this possibility.
-      For a new context, including recent infections, this condition could be relaxed by considering episodes that do not meet this criteria as right censored.
+      For a new context, including recent infections, this condition could be relaxed by considering episodes that do not meet this criterion as right censored.
 \end{enumerate}
@@
     (B) Prior predictive survival time (mean and 95\% CrI).
-    The implied prior on the survival time is very short when using either prior on the hazard at each time.
+    The implied prior on the survival time is short when using either prior on the hazard at each time.
   }
@@
 Smoothing priors~\autocite{gerschSmoothness} encode the biological consideration that the hazard should not vary significantly from day-to-day.
-They serve the same purpose as a penalized likelihood, which is used frequently in the non-parametric maximum-likelihood setting~\autocite[e.g.][]{bacchettiNonparametric}.
+They serve the same purpose as a penalized likelihood, which is often used in the non-parametric maximum-likelihood setting~\autocite[e.g.][]{bacchettiNonparametric}.
 Many possible forms of smoothing priors exist, including splines, Gaussian Processes, and random walks.
@@
 To form the duration distribution used in the simulation, we combine the two estimates.
-The pdf over the first 30 days is proportional to the ATACCC estimate, with the rest proportional to this CIS-based estimates.
-% The CIS-based estimates are shifted 3 days in order to make this smooth; this counteracts the above bias of missing the start of the infection (for times above 30 days, the other bias, missing short infections, is negligible).
-ERR...
-I'M NOT DOING THIS ANYMORE OH NO!!
+The pdf over the first 30 days is proportional to the ATACCC estimate, with the rest proportional to this CIS-based estimate.
 Denote by $f_A(t)$ the distribution function estimated in \cref{E-ATACCC} and $f_C(t)$ that from the CIS-based estimates just derived.

### 2024-03-23 biology-data.tex
Standardise age group names

@@
 As well as a swab to test for SARS-CoV-2, participants were asked to complete a questionnaire, and some participants were asked to provide blood samples.
-In reality, visits were often not on this precise schedule, and occasionally missed.
+In reality, visits were often not on this precise schedule, and occasionally missed altogether.
 A full description of the study can be found in the study protocol~\autocite{cisProtocol}.
@@
 The age groups used have varied slightly for different CIS analyses~\autocite[e.g.][]{pouwelsMRPvaccination,pouwelsCommunity,cisMethodsONS,houseInferring,walkerTracking}.
-I will use the age groups: 0--10 (pre-school and primary school), 11--15 (compulsory secondary school), 16--24 (adolescents and young adults who may remain in formal education), 25--49 (younger working age adults), 50--69 (older working age adults), and 70+ (elderly).
+I will use the age groups: 0--11 (pre-school and primary school), 12--16 (compulsory secondary school), 17--25 (adolescents and young adults who may remain in formal education), 26--50 (younger working age adults), 51--70 (older working age adults), and 71+ (elderly).
 

### 2024-03-23 transmission.tex
Standardise age group names

@@
 
-The incidence proportion was highest among the 11--15 and 16--25 age groups throughout the period (see \cref{transmission:fig:backcalc-ages}).
-Except the 16--25 age group, all age groups show similar trends (\ie parallel lines on a log scale), meaning there was insufficient information in the data to overcome the prior which favoured this situation.
-The incidence proportion in the 16--25 age group decreases faster than other age groups during November.
+The incidence proportion was highest among the 12--16 and 17--25 age groups throughout the period (see \cref{transmission:fig:backcalc-ages}).
+Except the 17--25 age group, all age groups show similar trends (\ie parallel lines on a log scale), meaning there was insufficient information in the data to overcome the prior which favoured this situation.
+The incidence proportion in the 17--25 age group decreases faster than other age groups during November.
 There are many plausible explanations for this effect.
@@
 % The model will be stratified only by region and age group, so the data are aggregated to these levels.
-% Only individuals aged two or older are eligible to participate in the CIS; I assume that test results in all participants aged under 11 are still representative of the whole $[0, 11)$ age group.
+% Only individuals aged two or older are eligible to participate in the CIS; I assume that test results in all participants aged under 11 are still representative of the whole 0--11 age group.
 
@@
 \begin{tabular}{l|lllll}
-         & $[0,16)$ & $[16,25)$ & $[25,50)$ & $[50,70)$ & $[70,\infty)$ \\
+         & 0--11 & 12--16 & 17--25 & 26--50 & 71+ \\
         \hline
@@
 
-The 11--16 age group (roughly corresponding to compulsory secondary school education) is estimated to have the highest incidence (see \cref{SEIR:fig:incidence}).
+The 12--16 age group (roughly corresponding to compulsory secondary school education) is estimated to have the highest incidence (see \cref{SEIR:fig:incidence}).
 Incidence is smooth within each week but shows discontinuities between weeks.
@@
 In this model it is $1 - \vec{s}$, the proportion susceptible at the time the study ends.
-The attack rate is highest in the $[11, 16)$ age group in the North West at 68\% (95\% CrI: 66--70\%) (see \cref{SEIR:fig:attack-rates}).
+The attack rate is highest in the 12--16 age group in the North West at 68\% (95\% CrI: 66--70\%) (see \cref{SEIR:fig:attack-rates}).
 Some of these infections occurred prior to the period covered by this study; during the period of this study 23\% (95\% CrI: 21--24\%) of the population in the North West aged 11--16 was infected.
@@
 There are a few features of the data which are not captured by the model and the posterior estimate appears to be a compromise with the prevalence too low in one age group and too high in another.
-For example, in the North East region, the 11--16 age group has its prevalence overestimated in October but the 16--25 age group has its prevalence underestimated.
+For example, in the North East region, the 12--16 age group has its prevalence overestimated in October but the 17--25 age group has its prevalence underestimated.
 \begin{figure}
@@
 In some regions, there appears to be a trade-off between under and overfitting in different age groups (see \cref{SEIR:fig:prev}).
-For example, in the North East, most of the data points in the $[11, 16)$ age group are below the median posterior estimate, while most of the data points in the $[16, 25)$ age group are above the median posterior estimate.
+For example, in the North East, most of the data points in the 12--16 age group are below the median posterior estimate, while most of the data points in the 17--25 age group are above the median posterior estimate.
 This implies that the data would allow identification of a more flexible age-based model.
@@
 This is because the model is not flexible enough to capture the true age distribution of infections.
-The overestimation of prevalence in the 11--16 age group and underestimation in the 16--25 age group is reflected when comparing incidence estimates to those from the backcalculation model, which have reduced bias.
+The overestimation of prevalence in the 12--16 age group and underestimation in the 17--25 age group is reflected when comparing incidence estimates to those from the backcalculation model, which have reduced bias.
 \begin{figure}

### 2024-03-23 transmission/backcalc-ages.R

@@ -4,10 +4,12 @@ library(patchwork)
 library(tidybayes)
 library(tidyr)
 source(here::here("utils.R"))
+source(here::here("transmission/utils.R"))
 
 output_dir = here::here("transmission/outputs")
 p_age_base = readRDS(file.path(output_dir, "age.rds")) |>
     filter(daynr > 1) |>
+    mutate(age_group = normalise_age_groups(age_group)) |>
     ggplot(aes(date, incidence, colour = age_group, fill = age_group)) +
     stat_lineribbon(alpha = 0.3, .width = 0.95) +
     labs(
@@ -25,6 +27,7 @@ p_age = (p_age_base + scale_y_continuous(labels = scales::label_percent())) +
 
 ggsave(
     filename = here::here("transmission", "backcalc-ages.pdf"),
+    device = cairo_pdf,
     plot = p_age,
     width = 6.5,
     height = 3.6

### 2024-03-23 transmission/utils.R

@@ -19,9 +19,17 @@ normalise_region_names = function(in_names) {
 
 normalise_age_groups = function(in_names) {
     paste0("[", stringr::str_sub(in_names, 2, -2), ")") |>
-        stringr::str_replace(stringr::fixed("[1,"), "[0,") |>
         stringr::str_replace(stringr::fixed("+"), "Inf") |>
-        stringr::str_replace(stringr::fixed(", "), ",")
+        stringr::str_replace(stringr::fixed(", "), ",") |>
+        stringr::str_replace(stringr::fixed("[1,"), "[0,") |>
+        dplyr::case_match(
+            "[0,11)" ~ "0–11",
+            "[11,16)" ~ "12–16",
+            "[16,25)" ~ "17–25",
+            "[25,50)" ~ "26–50",
+            "[50,70)" ~ "51–70",
+            "[70,Inf)" ~ "71+",
+        )
 }
 
 load_poststrat_table = function() {

### 2024-03-25 ATACCC.tex
proof and format numbers

@@
 The inference was performed using RStan~2.32.3~\autocite{RStan-2-32-3} and Stan~2.32.2~\autocite{Stan-2-32-2}.
-Inference settings were the same as \textcite{hakkiOnset}, which included 8 chains with a total of 8000 iterations each with the first 3000 used as warm-up.
+Inference settings were the same as \textcite{hakkiOnset}, which included 8 chains with a total of \numprint{8000} iterations each with the first \numprint{3000} used as warm-up.
 \begin{table}
@@
   $\vec{\mu}_\phi^{(i)}$ and $\matr{\Sigma}_\phi^{(i)}$ denote the $i$th posterior sample of $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$ respectively, taken from the MCMC output.
-  $M$ is the number of individuals to draw, I found $M = 100,000$ produced stable estimates of the tail of $F_{D,g}(d)$ if $\dmax = 45$.
+  $M$ is the number of individuals to draw, I found $M =\numprint{100000}$ produced stable estimates of the tail of $F_{D,g}(d)$ if $\dmax = 45$.
  }

### 2024-03-25 biology-data.tex
proof and format numbers

@@
 False positives are very rare for SARS-CoV-2 RT-PCR tests.
-In summer 2020, the CIS performed 208,730 PCR tests and found only 159 positives~\autocite[section 5]{cisMethodsONS}.
+In summer 2020, the CIS performed\numprint{208730} PCR tests and found only 159 positives~\autocite[section 5]{cisMethodsONS}.
 If these were all false positives, then the test specificity would be 99.9\%.
@@
 
-% Possible summaries
-% \begin{enumerate}
-%   \item \url{https://www.nature.com/articles/s41579-022-00822-w/figures/2}
-%   \item \url{https://www.nature.com/articles/s41576-021-00360-w/figures/1}
-%   \item \url{https://academic.oup.com/view-large/figure/305617142/ciaa1442_fig2.jpg}
-% \end{enumerate}
-
-% \section{Studies used ie this thesis} \label{biology-data:sec:studies}
-
-% In this section, I expand on the major two studies I use in this thesis (ATACCC and CIS).
-% Unlike many studies of SARS-CoV-2, these studies are not focused on specific groups such as healthcare workers or hospital patients.
-% Therefore, the results from these studies are more generalisable to the entire UK population.
-
 \section{Assessment of Transmission and Contagiousness of COVID-19 in Contacts} \label{biology-data:sec:ataccc}
@@
 ATACCC enrolled a total of 738 contacts across two waves.
-The first wave enrolled 393 contacts between 13th Sep 2020 and 31st Mar 2021, when pre-Alpha and Alpha variants were circulating.
-The second wave enrolled 345 contacts between 24th May 2021 and 228th Oct 2021, when the Delta variant was circulating.
+The first wave enrolled 393 contacts between 13 Sep 2020 and 31 Mar 2021, when pre-Alpha and Alpha variants were circulating.
+The second wave enrolled 345 contacts between 24 May 2021 and 28 Oct 2021, when the Delta variant was circulating.
 Contacts were swabbed daily for up to 20 days, with a variety of tests performed on the swabs; relevant for this thesis is the RT-PCR test for SARS-CoV-2.
@@
 Initially, it was limited to England, but expanded to cover the whole of the UK in September 2020.
-Enrolment was continuous until 31st Jan 2022, with data collected until 13th Mar 2023~\autocite{weiRisk}.
+Enrolment was continuous until 31 Jan 2022, with data collected until 13 Mar 2023~\autocite{weiRisk}.
 
@@
 Up until July 2020, households were selected from those that had previously participated in an ONS survey~\autocite{CIStechData}.
-Around 50\% of households agreed to participate, with around 90\% of eligible individuals (96,113 in total) in these households contributing multiple swabs to the survey.
+Around 50\% of households agreed to participate, with around 90\% of eligible individuals (\numprint{96113} in total) in these households contributing multiple swabs to the survey.
 After July 2020, addresses were also randomly sampled from a database of all UK addresses held by the ONS.
-The response rate was lower, with only 12\% of households agreeing to participate, from which 86\% of eligible individuals (383,732 in total) contributed multiple swabs to the survey.
+The response rate was lower, with only 12\% of households agreeing to participate, from which 86\% of eligible individuals (\numprint{383732} in total) contributed multiple swabs to the survey.
 
@@
 
-In this thesis, I focus on the period from Mon 31st Aug 2020 until Sun 24th Jan 2021 (see \cref{E-intro:sec:aims}).
+In this thesis, I focus on the period from Mon 31 Aug 2020 until Sun 24 Jan 2021 (see \cref{E-intro:sec:aims}).
 During the start of this period, the recruitment rate of CIS was expanded (see \cref{biology-data:fig:CIS-recruitment}(A)).
@@
 Analyses requiring more fine-grained stratification (such as that in \cref{biology-data:sec:MRP}) or linking data from the same individual over time (such as that in \cref{E-perf-test,E-imperf-test}) must be performed within the SRS.
-The SRS contains five 2.40\,GHz dual-core Intel Xeon E5-4650 processors and 164\,GB RAM.
+The SRS contains five \numprint[GHz]{2.40} dual-core Intel Xeon E5-4650 processors and \numprint[GB]{164} RAM.
 These computational resources are shared between all users of the CIS SRS project and due consideration must be given to other users.

### 2024-03-25 cis-imperfect-testing.tex
fix refs
proof and format numbers

@@
     Proportion of detected episodes which have a single positive episode.
-    y-axis is the number of simulations, out of 1000, with the indicated proportion.
+    y-axis is the number of simulations, out of \numprint{1000}, with the indicated proportion.
     Test sensitivity is $\psens$.
@@
 
-In this section I apply the approach described in this chapter to the CIS episodes dataset I described in \cref{E-perf-test:sec:approach}.
-This is the 4,800 CIS infection episodes first detected between 10th Oct 2020 and 6th Dec 2020 inclusive and have negatives bounding the start and end time of the episode.
+In this section I apply the approach described in this chapter to the CIS episodes dataset I described in \cref{perf-test:sec:data-notation}.
+This is the \numprint{4800} CIS infection episodes first detected between 10 Oct 2020 and 6 Dec 2020 inclusive and have negatives bounding the start and end time of the episode.
 
@@
 \todo{Insert correct citation for RTM paper 2 once available}
-Approximating this distribution as a negative binomial and scaling the mean to the size of the CIS cohort gives the prior $\mu\inform = 25132$ and $r\inform = 22047)$.
+Approximating this distribution as a negative binomial and scaling the mean to the size of the CIS cohort gives the prior $\mu\inform = \numprint{25132}$ and $r\inform = \numprint{22047}$.
 

### 2024-03-25 cis-perfect-testing.tex
fix refs
proof and format numbers

@@
 Duration distributions tend to be right-skewed.
-Therefore, estimates of the mean would be sensitive to the tail of the duration distribution.
+Therefore, estimates of the mean can be sensitive to the tail of the duration distribution.
 
@@
 They occur when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
-Undetected episodes are important because they are, on average, shorter than detected episodes; therefore, neglecting them would lead to a biased estimate of the survival distribution.
-To see why this is the case, consider the infection episode in figure \cref{perf-test:fig:truncation}.
+Undetected episodes are, on average, shorter than detected episodes; therefore, neglecting them would lead to a biased estimate of the survival distribution.
+To see why they are shorter, consider the infection episode in \cref{perf-test:fig:truncation}.
 Here two individuals are shown, with identical data collected for each of them (a series of negative
@@
 Therefore, it is likely that many infections in the CIS cohort were undetected.
-This is because the interval between tests in the same individual (up to 28 days, see \cref{E-intro:sec:cis}) is longer than most infections (median 16 days, see \cref{E-ATACCC:sec:results}).
 \begin{figure}
   \makebox[\textwidth][c]{\includegraphics[width=\textwidth]{cis-perfect-testing/truncation}}
-  \caption[Undetected episodes in CIS data]{Graphical illustration of an undetected infection episode, represented by the shaded area. \label{perf-test:fig:truncation}}
+  \caption[Undetected episodes in CIS data]{%
+    Identical data for two individuals, with a series of negative tests.
+    Top individual remains uninfected, bottom individual has an undetected infection.
+    The shaded area represents when the individual is detectable.
+  }
+  \label{perf-test:fig:truncation}
 \end{figure}
@@
 This assumption performs well in these settings, when the length of the intervals is fairly constant.
-In particular, when the interval lengths are near-constant, $p(\text{episode $j$ detected} \mid b_j \in [l_j^{(b)}, r_j^{(b)}]) \approx p(\text{episode $j$ detected})$.
+A necessary condition for it to perform well, satisfied when the interval lengths are near-constant, is $p(\text{episode $j$ detected} \mid b_j \in [l_j^{(b)}, r_j^{(b)}]) \approx p(\text{episode $j$ detected})$.
 However, the interval lengths change from one to four weeks in the CIS (see \cref{E-intro:sec:cis}).
@@
 This is because more frequent testing will detect a higher proportion of infections.
-Therefore, for the average detected episode, $p(\text{episode $j$ detected} \mid b_j \in [l_j^{(b)}, r_j^{(b)}]) > p(\text{episode $j$ detected})$.
-Therefore, conditioning on $b_j \in [l_j^{(b)}, r_j^{(b)}]$ is inappropriate.
+Therefore, for the average detected episode, $p(\text{episode $j$ detected} \mid b_j \in [l_j^{(b)}, r_j^{(b)}]) > p(\text{episode $j$ detected})$ and conditioning on $b_j \in [l_j^{(b)}, r_j^{(b)}]$ is inappropriate.
 
@@
 They apply it to a situation where the visit times (possible detection times) are common to all the individuals, simplifying the likelihood.
-I adapt this framework for CIS, when each individual has their own visit schedule (see section \cref{perf-test:sec:model}).
+I adapt this framework for CIS, when each individual has their own visit schedule (see \cref{perf-test:sec:model}).
 
@@
 First, the frequentist properties (\eg consistency and convergence rate) of proposed estimators are poorly understood~\autocite{sunAnalysis,dengNonparametric}.
-Second, a prior can include information from previous studies, such as from \cref{E-ATACCC}.
+Second, a prior can include information from previous studies, such as \cref{E-ATACCC}.
 Alternatively, the prior can favour a smoothly changing hazard, which is biologically sensible, and can overcome issues caused by the undetected episodes ~\autocite{caoBias}.
@@
 
-In this section, I develop a statistical model accounting for double interval censoring and undetected episodes.
+In this section, I develop a computationally feasible statistical model accounting for double interval censoring and undetected episodes.
 In \cref{perf-test:sec:data-notation}, I introduce the data and notation, defining which quantities are observed and which are latent.
-Then, in \cref{perf-test:sec:episode-types}, I apply the framework of \textcite{heiseyModelling} to classify possible episode beginning and end types into \emph{admissible} episodes, which match the doubly interval censored pattern of the observations; \emph{undetected} episodes, which may have occurred but would have been undetected; and \emph{inadmissible} episodes, which the data rules out having occurred.
-The number of such episodes in each class is a random variable, with a multinomial distribution, conditional on the latent total number of infections (including undetected infections).
+Then, in \cref{perf-test:sec:episode-types}, I apply the framework of \textcite{heiseyModelling} which classifies possible episode beginning and end time into \emph{admissible} episodes, which match the doubly interval censored pattern of the observations; \emph{undetected} episodes, which may have occurred and been undetected; and \emph{inadmissible} episodes, which the data rules out having occurred.
+The number of such episodes in each class has a multinomial distribution conditional on the latent total number of infections (including undetected infections).
 
@@
 The following subsections then derive its various components.
-Unlike in \textcite{heiseyModelling}, I must consider the probability of undetected episodes for each individual seperately because they have differing visit times.
+Unlike in \textcite{heiseyModelling}, I must consider the individual's probability of undetected episodes separately because they have differing visit times.
 In addition, I include the information from the individuals without a detected episode.
 
-Unlike prior work, I marginalise the likelihood over the number of undetected episodes.
+Unlike prior work, I marginalise the posterior density over the number of undetected episodes.
 This makes inference tractable using the large sample size and the computing resources available within the SRS (see \cref{E-biology-data:sec:SRS}).
@@
 
-I analyse the 4,800 episodes first detected between 10th Oct 2020 and 6th Dec 2020 inclusive, with negatives bounding the start and end time of the episode.
+I analyse the \numprint{4800} episodes first detected between 10 Oct 2020 and 6 Dec 2020 inclusive, with negatives bounding the start and end time of the episode.
 This period is denoted as $t = 1, \dots, T$.
@@
 The test schedule $\sched_i$ is the set of times individual $i$ is tested at, from their last test prior to time 1 (if it exists) up until time $T$.
-These are all the times an infection episode can be detected at, plus one prior test
+These are all the times an infection episode can be detected at, plus one prior test.
 For example, for the individual in \cref{perf-test:sec:undetected}, $\sched_i = \{ 0, 7, 14, 21, 28, 56 \}$.
@@
 
-To relate the possible value for $b_j$ and $e_j$, given a test schedule $\sched_{i(j)}$, to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie: are strict subsets of $T \times T$).
+To relate the possible value for $b_j$ and $e_j$, given a test schedule $\sched_{i(j)}$, to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie are strict subsets of $T \times T$).
 Figure \ref{perf-test:fig:partitionSpace} shows this graphically.
@@
 &= \frac{1}{T} \sum_{b = \min \sched_{i} + 1}^T S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1)
+\label{perf-test:eq:piu}
 \end{align}
@@
     \item Extract the test schedules for each individual who had at least one test during the period of interest.
-    \item Draw an episode start time, $b_{j}$ for each individual uniformly at random between 2nd July 2020 (100 days before the period where a detected episode would be included) and 6th December 2020 (the end of this period).
+    \item Draw an episode start time, $b_{j}$ for each individual uniformly at random between 2 July 2020 (100 days before the period where a detected episode would be included) and 6 December 2020 (the end of this period).
     \item Draw a duration of episode for each individual, $d_i$, based on the distribution described below. Then calculate the end of their infection episode, $e_{j} = b_{j} + d_i - 1$.
@@
     \item Discard episodes where there are no positive tests (\ie undetected episodes) and then apply the same inclusions/exclusions as in \cref{perf-test:sec:problem}.
-    \item Of these remaining episodes, sample 4,800 to match the sample size of the true dataset.
+    \item Of these remaining episodes, sample \numprint{4800} to match the sample size of the true dataset.
 This is needed because in step 2 the entire cohort was infected, while in the real study only an (unknown) portion is infected.
@@
 
-Repeating the simulation showed negligible variation in the results (not shown) due to the large sample size (4800 detected infections).
+Repeating the simulation showed negligible variation in the results (not shown) due to the large sample size (\numprint{4800} detected infections).
 Therefore, I show and discuss results from only a single simulation.
@@
 All analyses converged.
-Rhat were below 1.01, all parameters had an effective sample size of 1000, and there were no divergent transitions.
+Rhat were below 1.01, all parameters had an effective sample size of \numprint{1000}, and there were no divergent transitions.
 

### 2024-03-25 introduction.tex
proof and format numbers

@@
 REACT was a series of cross-sectional surveys, run across 19 rounds between May 2020 and March 2022.
-Each round consisted of at least 100,000 randomly sampled individuals from the UK contributing swabs for testing and questionnaire data~\autocite{elliottTwin}.
+Each round consisted of at least \numprint{100000} randomly sampled individuals from the UK contributing swabs for testing and questionnaire data~\autocite{elliottTwin}.
 
@@
 \Textcite{ealesCharacterising} estimated the duration of positivity from REACT data.
-Specifically, individuals who tested positive in round 8 of REACT (6th--22nd Jan 2021) were invited to take two additional swab tests.
+Specifically, individuals who tested positive in round 8 of REACT (6th--22 Jan 2021) were invited to take two additional swab tests.
 Defining the time of the initial positive test at time 0, \textcite{ealesCharacterising} estimated the probability of remaining positive at each day after the initial positive test.
@@
 
-In order to achieve these aims, I use a case study, specifically the SARS-CoV-2 epidemic in England from Mon 31st Aug 2020 until Sun 25th Jan 2021.
+In order to achieve these aims, I use a case study, specifically the SARS-CoV-2 epidemic in England from Mon 31 Aug 2020 until Sun 25 Jan 2021.
 This case study has several features that make it useful for the aims of this thesis.

### 2024-03-25 thesis.tex
proof and format numbers

@@
 \usepackage{numprint}
+\npfourdigitnosep
+\npdecimalsign{.}
 % \usepackage{microtype}
@@
 \preface{%
-    In October 2019 I began a PhD project entitled ``Advancing Bayesian Statistical Methods for Real-Time Epidemic Monitoring''.
+    In October 2019 I began a PhD project entitled ``Advancing Bayesian Statistical Methods for Real-Time Epidemic Monitoring'', aiming to develop methods in preparation for a future influenza pandemic.
     So began a journey of reading, and an expectation of few deadlines, a lot of deep thinking, and complex statistical theory.
-    Like much of the world, the events of early 2020 changed that.
+    Like much of the world, the events of early 2020 rapidly changed that.
     Following a brief interlude to provide rapid support to the response, the origins of this thesis emerged.
@@
     Several aspects of my work were done to support the ongoing response.
-    In particular, an earlier version of the backcalculation work was adopted by the Office for National Statistics to estimate incidence, published in their weekly online reports.
+    In particular, an earlier version of the backcalculation work was adopted by the Office for National Statistics to estimate incidence, published in their weekly online reports~\autocite{cisMethodsONS}.
     This method was also used for the analysis of several respiratory diseases over winter 2022/3~\autocite{dietzSARSCoV2final}.

### 2024-03-25 transmission.tex
proof and format numbers

@@
 An exception is the North East and East of England regions (see \cref{transmission:fig:backcalc-regions}(C) and (D) respectively); these regions show no consistent increase or decrease in incidence during October.
-All regions then decrease during November; this coincides with the national lockdown which came into force in England on 5th Nov 2020~\autocite{ifgTimeline}.
+All regions then decrease during November; this coincides with the national lockdown which came into force in England on 5 Nov 2020~\autocite{ifgTimeline}.
 
@@
 
-All results are based on four Markov chains each of length 1,000,000 iterations.
-The first 500,000 were discarded as burn-in, with the remaining chains thinned to every 100th iteration.
-This gives 5,000 posterior samples per region.
-This was further thinned to 2,000 samples when calculating posterior predictive quantities (incidence and prevalence) due to their computational cost.
+All results are based on four Markov chains each of length \numprint{1000000} iterations.
+The first \numprint{500000} were discarded as burn-in, with the remaining chains thinned to every 100th iteration.
+This gives \numprint{5000} posterior samples per region.
+This was further thinned to \numprint{2000} samples when calculating posterior predictive quantities (incidence and prevalence) due to their computational cost.
 
@@
 These were generated according to the likelihood, \cref{SEIR:eq:likelihood}.
-The number of daily tests was set to the mean of what CIS performed in the region over this period (1630), distributed across the population in proportion to its size.
+The number of daily tests was set to the mean of what CIS performed in the region over this period (\numprint{1630}), distributed across the population in proportion to its size.
 
@@
 The highest Rhat was 1.01 for $\sigma_\epsilon$ in London.
-All parameters had ESS > 1000, except $\sigma_\epsilon$ and $\psi$ in London, which had ESS of 450 and 950 respectively.
+All parameters had ESS > \numprint{1000}, except $\sigma_\epsilon$ and $\psi$ in London, which had ESS of 450 and 950 respectively.
 

### 2024-03-25 cis-perfect-testing.tex
proof ch 5

@@
 
-I analyse the \numprint{4800} episodes first detected between 10 Oct 2020 and 6 Dec 2020 inclusive, with negatives bounding the start and end time of the episode.
+I analyse the $\ndet = \numprint{4800}$ episodes with their first positive test between 10 Oct 2020 and 6 Dec 2020 inclusive, with negatives bounding the start and end time of the episode; I refer to these episodes as \emph{detected}.
+In \cref{perf-test:sec:prob-undetected} I formalise the conditions for an episode to be detected.
 This period is denoted as $t = 1, \dots, T$.
@@
 Similar information comes from the negative test after the end of the episode.
-A negligible number of episodes have a negative test after the end of the episode.
-I refer to these episodes as \emph{detected}; in \cref{perf-test:sec:prob-undetected} I formalise the conditions for an episode to be detected.
-Let $\ndet$ denote the total number of detected infections.
+A negligible number of episodes are missing a negative test after the end of the episode.
 
-The cohort of individuals considered is all those in the CIS who could have had a detected test.
-This requires that they had at least three tests before time $T$ (a negative prior to the episode, a positive, and a negative following the episode).
-Index these individuals by $i = 1, \dots, \Ncis$ where $\Ncis$ is the total number of such individuals.
+I consider the cohort of $\Ncis$ individuals who could have had a detected episode; this is a subset of the whole CIS cohort.
+To be considered, an individual must have had at least three tests before time $T$ (a negative prior to the episode, a positive, and a negative following the episode).
+I index these individuals by $i = 1, \dots, \Ncis$.
 
@@
 These are all the times an infection episode can be detected at, plus one prior test.
-For example, for the individual in \cref{perf-test:sec:undetected}, $\sched_i = \{ 0, 7, 14, 21, 28, 56 \}$.
+For example, for the individual in \cref{perf-test:fig:double-interval-censor}, $\sched_i = \{ 0, 7, 14, 21, 28, 56 \}$.
 Each individual has exactly one test schedule, but it is possible that $\sched_i = \sched_{i'}$ for $i \neq i'$.
@@
 
-Any infection episode, whether detected or not, can be fully characterised by the triplet $(b_j, e_j, i(j))$ where $b_j$ and $e_j$ are the times the episode begin and end respectively, and $i(j)$ denotes the individual the episode occurred in.
+Any infection episode, whether detected or not, is fully characterised for my purposes by the triplet $(b_j, e_j, i(j))$ where $b_j$ and $e_j$ are the times the episode begin and end respectively, and $i(j)$ denotes the individual the episode occurred in.
 The episode's duration is $d_j = e_j - b_j + 1$ (see \cref{E-inc-prev:sec:definitions}).
@@
 
-To relate the possible value for $b_j$ and $e_j$, given a test schedule $\sched_{i(j)}$, to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie are strict subsets of $T \times T$).
-Figure \ref{perf-test:fig:partitionSpace} shows this graphically.
+To relate the possible value for $b_j$ and $e_j$, given a test schedule $\sched_{i(j)}$, to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie are strict subsets of $T \times T$), shown in \cref{perf-test:fig:partitionSpace}.
 Each episode, whether detected or not, must fall into one of these three classes with a probability that is a function of the parameters of the survival function, $\vec{\vec{\theta}}$, specified in \cref{perf-test:sec:parameters-priors}.
@@
   An unknown number, $n_{iu}$ of these occur in individual $i$.
-  The undetected region is the red region in \cref{perf-test:fig:partitionSpace}, which are when the individual is infected then recovers before their next test, plus when the individual is infected before their enrolment.
+  The undetected region is the red region in \cref{perf-test:fig:partitionSpace}, indicating combinations of episode beginning and end times which do not span a testing time plus episodes beginning before the individual's enrolment.
   The shape of this region accounts for the pattern of possible undetected infections.
@@
 Each episode occurs in one of the classes independently and with the same probability.
-Therefore, conditional on $\ntot$, the number of episodes in each class is multinomial.
-Specifically:
+Therefore:
 \begin{align}
@@
 That is, the approach is mathematically equivalent but greatly complicates the formulae and notation.
-Therefore, I present this version as it is clearer.
+Therefore, I present this version for clarity.
 
@@
 Therefore, I combine these classes into a single class.
-The combined class has a count $\nnodet = \sum_{i=1}^{\Ncis} n_{iu}$ and probability $\pnodet = \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} p_{iu}$.
+The combined class has a count $\nnodet = \sum_{i=1}^{\Ncis} n_{iu}$ and probability parameter $\pnodet = \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} p_{iu}$.
 
@@
 \begin{align}
+p(\vec{\theta} \mid \na)
 &\propto p(\vec{\theta}) \sum_{\ntot=\ndet}^\infty \left( p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i \in \set{D}} p_{ia} \right) \\
@@
 &= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) e^{\lambda \pnodet} d\lambda &\text{Maclaurin series of $e$} \\
-&= \int \lambda^{\ndet} e^{-\lambda(1 - \pnodet)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda &\text{Gamma pdf}\\
+&= \int \lambda^{\ndet} e^{-\lambda(1 - \pnodet)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda &\lambda \dist \GamDist\\
 &= \int \frac{b^a}{\Gamma(a)} \lambda^{a+\ndet-1} e^{-(b+1-\pnodet)\lambda} \lambda d\lambda \\
@@
 From \cref{perf-test:eq:full-posterior}, this is only needed for $i \in \set{D}$, that is for individuals with detected episodes.
-For such an individual, let $j$ denote their detected episode.
+For such an individual, let $j(i)$ denote their detected episode.
 
@@
 p_{ia} =& \prob((B_j, E_j) \in \alpha_i \mid \vec{\theta}, i(j) = i) \\
-\alpha_i =& \{ (b, e) : l_{j}^{(b)} \leq b \leq r_{j}^{(b)} \wedge l_{j}^{(e)} \leq e \leq r_{j}^{(e)}\}.
+\alpha_i =& \{ (b, e) : l_{j(i)}^{(b)} \leq b \leq r_{j(i)}^{(b)} \wedge l_{j(i)}^{(e)} \leq e \leq r_{j(i)}^{(e)}\}.
 \intertext{Omitting the conditioning on $\vec{\theta}$ and $i(j) = i$, this gives:}
 p_{ia}
-=& \prob \left( l_{j}^{(b)} \leq B_{j} \leq r_{j}^{(b)}, l_{j}^{(e)} \leq E_{j} \leq r_{j}^{(e)} \right) \\
-=& \prob \left( l_{j}^{(e)} \leq E_{j} \leq r_{j}^{(e)} \mid l_{j}^{(b)} \leq B_{j} \leq r_{j}^{(b)} \right) \\
-   &\times\prob \left( l_{j}^{(b)} \leq B_{j} \leq r_{j}^{(b)} \right) \\
-=& \sum_{b = l_{j}^{(l)}}^{r_{j}^{(b)}} \prob \left( l_{j}^{(e)} \leq E_{j} \leq r_{j}^{(e)} \mid B_{j} = b \right) \prob \left(B_{j} = b \right) \\
-=& \sum_{b = l_{j}^{(b)}}^{r_{j}^{(b)}} \prob \left( l_{j}^{(e)} - b + 1 \leq D_{j} \leq r_{j}^{(e)} - b + 1 \right) \prob \left(B_{j} = b \right) &\text{by def of $D_{j}$} \\
-=& \sum_{b = l_{j}^{(b)}}^{r_{j}^{(b)}} \left( S_{\vec{\theta}}(l_{j}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j}^{(e)} - b + 2) \right) \prob \left(B_{j} = b \right) &\text{by def of $S_{\vec{\theta}}$} \\
-\propto& \sum_{b = l_{j}^{(b)}}^{r_{j}^{(b)}} \left( S_{\vec{\theta}}(l_{j}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j}^{(e)} - b + 2) \right)
+=& \prob \left( l_{j(i)}^{(b)} \leq B_{j} \leq r_{j(i)}^{(b)}, l_{j(i)}^{(e)} \leq E_{j} \leq r_{j(i)}^{(e)} \right) \\
+=& \prob \left( l_{j(i)}^{(e)} \leq E_{j} \leq r_{j(i)}^{(e)} \mid l_{j(i)}^{(b)} \leq B_{j} \leq r_{j(i)}^{(b)} \right) \\
+   &\times\prob \left( l_{j(i)}^{(b)} \leq B_{j} \leq r_{j(i)}^{(b)} \right) \\
+=& \sum_{b = l_{j(i)}^{(l)}}^{r_{j(i)}^{(b)}} \prob \left( l_{j(i)}^{(e)} \leq E_{j} \leq r_{j(i)}^{(e)} \mid B_{j} = b \right) \prob \left(B_{j} = b \right) \\
+=& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \prob \left( l_{j(i)}^{(e)} - b + 1 \leq D_{j} \leq r_{j(i)}^{(e)} - b + 1 \right) \prob \left(B_{j} = b \right) &\text{by def of $D_{j}$} \\
+=& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_{j(i)}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 2) \right) \prob \left(B_{j} = b \right) &\text{by def of $S_{\vec{\theta}}$} \\
+\propto& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_{j(i)}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 2) \right)
 \end{align}
@@
       This means that the test is detected before $T$.
-    \item $\exists t \in \sched_{i(j)}$ such that $t > e_j$, upper bounding the end of the episode.
+    \item Individual $i(j)$ is tested at a time $t > e_j$, upper bounding the end of the episode.
+      Note that this could be after $T$, and hence it is not necessarily the case that $t \in \sched_{i(j)}$.
       For episodes detected in the period I consider, this happens a negligible number of times, only in individuals lost to follow-up; therefore, I neglect this possibility.
-      For a new context, including recent infections, this condition could be relaxed by considering episodes that do not meet this criterion as right censored.
+      % For a new context, including recent infections, this condition could be relaxed by considering episodes that do not meet this criterion as right censored.
 \end{enumerate}
@@
 defining $\min \emptyset = \infty$; \ie $\tau_{\sched_i}(t) = \infty$ if there is no $t' \in \sched_i$ such that $t' \geq t$.
-The condition that $\exists t \in \sched_{i(j)}$ such that $b_j \leq t \leq e_j$ can now be expressed as $e_j \geq b_j + \tau_{\sched_{i(j)}}(b_j)$.
+The first condition can now be expressed as $e_j \geq b_j + \tau_{\sched_{i(j)}}(b_j)$.
 Equivalently, $d_j \geq \tau_{\sched_{i(j)}}(b_j) + 1$.
@@
 Firstly, the model structure from \cref{E-ATACCC} leads to positive correlation in the posterior estimates of the hazard, which should be propagated into this analysis.
-That is, the prior used in this analysis cannot be independent across the hazards.
+That is, the prior used in this analysis should have correlation between $\lambda_t$ and $\lambda_{t'}$.
 Secondly, the uncertainty in the estimates from \cref{E-ATACCC} is underestimated because it extrapolates beyond the data using strong model assumptions (see \cref{E-ATACCC:sec:discussion}).
@@
 
-I first approximate the \cref{E-ATACCC} posterior estimate of the hazard as $\logit{\vec{h}} \dist \MNorm(\vec{\mu}_A, \matr{\Sigma}_A)$ where $\vec{h}$ is the posterior distribution of the hazards, and $\vec{\mu}_A$ and $\matr{\Sigma}_A$ are the mean and covariance matrix computed from the samples of $\logit{\vec{h}}$ derived using the method of \cref{ATACCC:sec:duration}.
+I first approximate the \cref{E-ATACCC} posterior estimate of the hazard as $\logit{\vec{h}} \dist \MNorm(\vec{\mu}_A, \matr{\Sigma}_A)$ where $\vec{h}$ is the hazard, and $\vec{\mu}_A$ and $\matr{\Sigma}_A$ are the mean and covariance matrix computed from the samples of $\logit{\vec{h}}$ derived using the method of \cref{ATACCC:sec:duration}.
 Using a multivariate normal, as opposed to multiple univariate distribution for each $h_t$, preserves the correlation between the hazards.
-The logit transformation maps from the $[0, 1]$ interval to the full real line.
+% The logit transformation maps from the $[0, 1]$ interval to the full real line.
 The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC-hazard} and \cref{perf-test:fig:approximate-ATACCC-survival}).
@@
 Having approximated the estimate as a multivariate normal, I add additional uncertainty using a discrete Beta process.
-The discrete Beta process prior generalises the form of prior used in \cref{perf-test:sec:independent-priors} by allowing the central estimate of the hazard to vary over time~\autocite{ibrahimBayesian,sunStatisticala}.
+The discrete Beta process prior~\autocite{ibrahimBayesian,sunStatisticala} generalises the form of prior used in \cref{perf-test:sec:independent-priors} by allowing the central estimate of the hazard to vary over time.
 It is:
@@
 \end{align}
-where $k_t$, $\alpha_0$, and $\beta_0$ are hyperparameters; and $h_t$ is an estimate of the hazard at time $t$ from ATACCC.
+where $k_t$, $\alpha_0$, and $\beta_0$ are hyperparameters; and, as before, $h_t$ is an estimate of the hazard at time $t$ from \cref{E-ATACCC}.
 An intuition for what this distribution represents derives from a conjugate model for $\lambda_t$ with a beta prior and a binomial likelihood.
@@
   0 &t > 39
-\end{cases}
+\end{cases}.
 \end{align}
-where $\expit(x) = 1 / (1 + \exp(-x))$.
 
@@
 Reducing the amount of smoothing or type of smoothing may help.
-More flexible smoothers, such as a Gaussian Process for additional flexibility and smoothing and different timescales~\autocite{saulGaussian}.
+More flexible smoothing priors could be used, such as a Gaussian Process for additional flexibility and smoothing and different timescales~\autocite{saulGaussian}.
 

### 2024-03-25 paper-surv-analysis.tex
proof ch 5

@@
 The CIS had a household-based design inviting all individuals aged 2 and over from selected households.
-Once invited, an enrollment swab would be taken at the first visit followed by 4 further weekly visits (giving a total of 5 swabs on days 0, 7, 14, 21, 28 relative to enrollment) after which visits were monthly.
+Once invited, an enrolment swab would be taken at the first visit followed by 4 further weekly visits (giving a total of 5 swabs on days 0, 7, 14, 21, 28 relative to enrolment) after which visits were monthly.
 Swabs were tested for the presence of SARS-CoV-2 using RT-PCR.

### 2024-03-25 introduction.tex
proof intro

@@
 $\R$ is the mean number of secondary infections generated by a single infection in a \emph{fully susceptible} population with \emph{standard behaviour patterns}.
-A fully susceptible population means none of its members have any immunity to the disease.
+A fully susceptible population means that none of its members have any immunity to the disease.
 Note that this definition implies $\R$ is constant for a given pathogen and population.
@@
 $\Rc$ is useful when comparing the effects of interventions at different points in time, when the immunity status of the population has changed.
-For example, when comparing the English lockdown in March 2020 and January 2021, there are two effects in play: that there are more individuals with immunity in January 2021, and that the lockdown had different measures in place.
+For example, when comparing the English lockdown in March 2020 and January 2021, there are two effects in play: there are more individuals with immunity in January 2021, and the lockdowns had different measures in place.
 $\Rc$ focuses on the latter effect.
@@
 These definitions ignore many complexities that arise when theoretically defining or practically estimating any of the reproduction numbers~\autocite{pellisEstimation}.
-For example, effects due to seasonality, what a ``mean infection'' means in a heterogeneous population, and the lack of a ground truth for these numbers in almost all practical situations.
-However, their intuitive simplicity and convenient properties in simple epidemic models (see \cref{E-SEIR:sec:mechanistic-models}) mean that they are useful concepts for guiding policy.
+For example, effects due to seasonality, what a ``mean infection'' means in a heterogeneous population, and the lack of a ground truth for these numbers in almost all situations.
+However, their intuitive simplicity and convenient properties in simple epidemic models mean that they are useful concepts for guiding policy.
 
@@
 For a given disease and population, an analogous growth rate can be defined for each definition of the reproduction number~\autocite{pellisEstimation,paragGrowthRates,wallingaGI}.
-The advantage of using the growth rate it is easier to estimate and captures how quickly spread happens in calendar time; however, they do not directly give the strength of intervention required to control the epidemic~\autocite{royalSocietyRnumber}.
+The advantage of using the growth rate is that it isn easier to estimate and captures how quickly spread happens in calendar time; however, they do not directly give the strength of intervention required to control the epidemic~\autocite{royalSocietyRnumber}.
 
@@
 Incidence informs the short-term future burden of the disease, for example, the future number of severe events (hospitalisations or deaths).
-Incidence is translated into disease burden using a measure of severity is required; for example, the IFR (infection fatality rate), the proportion of infected individuals who die.
+Incidence is translated into disease burden using a measure of severity; for example, the IFR (infection fatality rate), the proportion of infected individuals who die.
 Estimation of severity is beyond the scope of this thesis.
@@
 Therefore, they suffer from biases that can be hard to understand and quantify~\autocite{shadboltChallenges}.
-Here I discuss three types of data which are useful for surveillance.
-Incidence-type and prevalence-type data can directly inform estimation of incidence and transmission; however, this estimation almost always relies on prior knowledge of epidemiological and/or biological quantities.
+Here I discuss three types of data which are useful for surveillance: incidence-type, prevalence-type, and data from cohort studies.
+\emph{Incidence-type} and \emph{prevalence-type} data can directly inform estimation of incidence and transmission; however, this estimation almost always relies on prior knowledge of epidemiological and/or biological quantities.
 For example, the time between infection and symptom onset.
@@
 
-More severe symptoms are typically less affected by ascertainment bias.
+Data on more severe symptoms are typically less affected by these biases.
 An extreme example is the development of AIDS following infection by HIV.
@@
 A spectrum of models exists, categorised by the strength of assumptions they make about the transmission process~\autocite{beckerCOVIDmodels}.
-Stronger assumptions allow more insight into the transmission process and generally lower the variance of the estimates but can introduce bias and are often computationally challenging to fit.
+Stronger assumptions allow more insight into the transmission process and lower the variance of the estimates but can introduce bias and be computationally challenging to fit.
 
@@
 The most common type of semi-mechanistic model is those based on a \emph{renewal equation}~\autocite{bhattSemimechanistic}.
-The renewal equation~\autocite{fraserRenewal} for an infectious disease, in the most general form, is:
+The renewal equation for an infectious disease~\autocite{fraserRenewal}, in the most general form, is:
 \begin{align}
@@
 \end{align}
-where $Z(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the mean rate an individual infected $\tau$ time units ago infects individuals at time $t$.
+where $Z(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the mean rate an individual infected $\tau$ time units ago infects individuals at calendar time $t$.
 Therefore, the renewal equation relates incidence in the past to incidence in the future.
@@
 Extensions of \cref{intro:eq:renewal} to track the number of susceptible individuals become full mechanistic models, equivalent to the compartmental models introduced next~\autocite{champredonEquivalence}.
-Renewal equations do not feature further in this thesis; for details, including practical considerations when using them as the basis for inference, see \textcite{nashEstimating,gosticPractical}.
+Renewal equations do not feature further in this thesis; for further details, including practical considerations when using them as the basis for inference, see \textcite{nashEstimating,gosticPractical}.
 
@@
 
-Solving \cref{intro:eq:disease-model} for $Z$ is an ill-posed problem when $f(0)$ is small.
+Solving \cref{intro:eq:disease-model} for $Z$ is an ill-posed problem.
 The forwards error of the solution is large when the ratio of the off-diagonal to diagonal elements is large~\autocite[chapter 8.2]{highamAccuracy}; for this situation, which is when the value of $\max_{i > 0} f(i) / f(0)$ is large.
@@
 Modelling the duration directly considers only the binary result at each test (positive or negative), for instance in a \emph{survival analysis} framework.
-This approach generally requires fewer parameters, and no random effect(s), reducing the computational burden.
+This approach generally requires fewer parameters reducing the computational burden.
 It is also easier to specify non-parametric forms of the distribution, which is important for SARS-CoV-2 as the tail behaviour may be important.
@@
 
-A deep understanding of each data source's potential was not possible.
+A deep understanding of each data source's potential was not possible in real-time.
 In this thesis, I will investigate the potential of the prevalence surveys in greater detail.
@@
 
-The main data sources informing the response to the SARS-Cov-2 epidemic in England were: community testing (known as pillar 2 testing), the number of severe events (\eg hospital admissions or deaths), seroprevalence surveys (estimating the proportion of the population with antibodies to a disease, a marker of prior infection), and prevalence surveys (estimating the proportion of the population currently with the disease).
+The main data sources informing the response to the SARS-Cov-2 epidemic in England were: community testing (known as pillar 2 testing), the number of severe events, seroprevalence surveys, and prevalence surveys.
 Each of these data sources have strengths and weaknesses, see \textcite{royalSocietyRnumber} and references therein for further details on each of these data sources.
@@
 The number of tests conducted varied over time and space, for a variety of known and unknown reasons.
-These include: policy on who is eligible for testing, the availability of tests, and the willingness of individuals to be tested.
+These include: the policy on who is eligible for testing, the availability of tests, and the willingness of individuals to be tested.
 Therefore, fluctuations in the number of positive tests may not reflect changes in the true incidence of the disease.
@@
 Therefore, some modelling is required to borrow information across strata or time to provide stable estimates, as well as to correct for non-response bias (see \cref{E-intro:sec:cis}).
-Despite this issue, they are considered the gold-standard metric.
+Despite this, they are considered the gold-standard metric.
 
@@
 The CIS was a longitudinal study, testing a large number of individuals weekly or monthly for SARS-CoV-2; it is described in detail in \cref{E-intro:sec:cis}.
-Early in the pandemic, a methodology was developed for using CIS data to estimate prevalence (see \cref{E-intro:sec:cis}).
 
@@
 
-Many cohort studies to estimate a wide variety of quantities ran throughout the pandemic.
-In this thesis, I make use of two cohort studies with differing designs to estimate the duration or PCR positivity.
+Many cohort studies, to estimate a wide variety of quantities, ran throughout the pandemic.
+In this thesis, I make use of two cohort studies with differing designs to estimate the duration of PCR positivity.
 These designs have complementary strengths, and I combine information from the studies.
@@
 
-In the UK, SPI-M-O (the Scientific Pandemic Influenza Group on Modelling, Operational sub-group) produced the most prominent modelling
+In the UK, SPI-M-O (the Scientific Pandemic Influenza Group on Modelling, Operational sub-group) produced the most prominent modelling.
 SPI-M-O was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), generating evidence based on infectious disease modelling for policy development~\autocite{medleySPIM,govSPIMO}.
@@
 This assumption is necessary because they had at most 17 days of follow-up, and hence could not estimate the tail of the distribution.
-Overall, they estimate a median duration of 14.0 days (95\% CI: 12.9--15.4 days) but with substantial positive correlation with each individual’s viral load at the time of the initial positive test (\ie a higher initial viral load was associated with a longer duration of positivity)
+Overall, they estimate a median duration of 14.0 days (95\% CI: 12.9--15.4 days) with substantial positive correlation with each individual’s viral load at the time of the initial positive test (\ie a higher initial viral load was associated with a longer duration of positivity).
 The individuals with the highest viral load individuals were estimated to have a median duration of 19.6 days (16.1--25.8 days).
@@
 Their methodology was similar to \textcite{abbottCISincidence}, including using estimates from \textcite{hellewellPCRSensitivity} for how the sensitivity of RT-PCR testing varies over an infection.
-Rather than using a normal process to provide stability, as \textcite{abbottCISincidence} do, \textcite{colmanAscertainment} assume the ascertainment rate is piecewise constant.
+Rather than using a Gaussian process to provide stability, as \textcite{abbottCISincidence} do, \textcite{colmanAscertainment} assume the ascertainment rate is piecewise constant.
 They do not fully propagate the uncertainty from the CIS prevalence estimates, considering only the bounds of the 95\% confidence intervals the ONS publishes for prevalence.
@@
 Therefore, I use computational methods to sample from the posterior distribution.
-The method I use for the majority of the analyses is MCMC (Markov chain Monte Carlo).
+The method I use for the majority of the analyses is MCMC (Markov chain Monte Carlo) (see \cref{E-MCMC} for details).
 
@@
 In particular, the duration of positivity and the susceptibility of the population.
-Including both of these elements would require more complex modelling but include idiosyncratic features of the SARS-CoV-2 pandemic that are probably less generalisable to future diseases.
+Including both of these elements would require more complex modelling and including idiosyncratic features of the SARS-CoV-2 pandemic that are less generalisable to other diseases.
 
@@
 
-Next, in \cref{E-perf-test}, I estimate the duration of RT-PCR-positivity from double censored data with arbitrary periods where infections could be undetected (a generalisation of truncation).
+In \cref{E-perf-test} I estimate the duration of RT-PCR-positivity from double censored data with arbitrary periods where infections could be undetected (a generalisation of truncation).
 This is based on the CIS data.
@@
 
+A particular challenge with these analyses is that they must be performed within a trusted research environment with limited computational resources.
+
 Within these analyses, I include information derived from the ATACCC study.
-In order to do so, I propose a novel prior which appropriately discounts the information from analysing ATACCC study, which has inappropriate model assumptions when applied to the larger sample size in the CIS data.
-
-A particular challenge with these analyses is that they must be performed within the SRS because they require individual-level data.
+In order to do so, I propose a prior which appropriately discounts the information from analysing the ATACCC study, which has inappropriate model assumptions when applied to the larger sample size in the CIS data.
 
@@
 Almost all code, excluding dependencies, was written from scratch for these analyses.
-The exception is the Markov chain Monte Carlo sampler used in \cref{E-SEIR}, written by Sanmitra Ghosh for the article \textcite{ghoshApproximate}.
+The exceptions are the Markov chain Monte Carlo sampler used in \cref{E-SEIR}, the model for prevlaence used in \cref{E-backcalc}, and the original viral load model used in \cref{ATACCC:sec:hakki}.
 
@@
 
-Therefore, I then turn to estimating this distribution.
+I then turn to estimating this distribution.
 I start, in \cref{E-ATACCC}, by estimating the duration of positivity from the ATACCC study.
@@
 
-Having estimated the duration of positivity, I then turn to estimating incidence and transmission from the CIS data in \cref{E-transmission}, thus fulfilling the primary aim of this thesis.
+Having estimated the duration of positivity, next I estimate incidence and transmission from the CIS data in \cref{E-transmission}, thus fulfilling the primary aim of this thesis.
 I use both a phenomenological and mechanistic transmission model, contrasting their results and discussing the advantages and disadvantages of each in this context.

### 2024-03-25 introduction.tex
plural data

@@
 
-An example of incidence-type data is community testing.
+An example of incidence-type data are community testing.
 This relies on individuals presenting for a test, hence is biased by individuals' ability and willingness to be tested.
@@
 Therefore, the code for the analyses in this thesis is available online.
-An index to the code and, where possible, data is available at \url{https://github.com/joshuablake/thesis-index}.
+An index to the code and, where possible, data are available at \url{https://github.com/joshuablake/thesis-index}.
 I structured two elements of the code as R packages to enable their application to future contexts.

### 2024-03-25 cis-perfect-testing.tex
plural data
Updates from Overleaf

--- 2024-03-23 cis-perfect-testing.tex
@@
 In \cref{E-ATACCC} I estimated the duration distribution from the ATACCC dataset.
-However, this study suffered from a small sample size and only 20 days of follow-up, preventing accurate estimation of the tail.
-Therefore, model assumptions drive the size of the estimated tail.
+However, this study had a small sample size and only 20 days of follow-up, which prevented accurate estimation of the tail.
+Therefore, model assumptions drive the estimated size of the tail.
 
-The CIS sampling is less frequent, but with has a large sample size (\numprint{4800} detected infection episodes) and a long follow-up.
+The CIS samples individuals less frequently, but has a large sample size (\numprint{4800} detected infection episodes) and up to years of follow-up.
 Therefore, it is suitable for estimating the tail of the duration distribution.
-In this chapter, I develop a method to estimate the duration distribution using CIS data, assuming no false negatives or positives.
-Formally, I aim to estimate the survival function $S$ defined in \cref{inc-prev:sec:definitions}.
+In this chapter, I develop a method to estimate the duration distribution using CIS data, assuming the test ???
+has perfect sensitivity and specificity, {\it i.e.} there are no false negatives or positives.
+Formally, I aim to estimate the survival function $S(t)$ defined in \cref{inc-prev:sec:definitions}.
 
@@
 This motivates using a survival analysis framework which I introduce in \cref{perf-test:sec:surv-analysis}, before deriving the statistical model for the CIS setting in \cref{perf-test:sec:model}.
-Next, in \cref{perf-test:sec:parameters-priors}, I consider how to parameterize the survival function and choose priors; these include vague and smoothing priors as well as including information from \cref{E-ATACCC}.
+In \cref{perf-test:sec:parameters-priors}, I consider how to parameterize the survival function and choose its priors, including vague and smoothing priors as well as a prior constructed to include the information from \cref{E-ATACCC}.
 I evaluate the model and priors using a simulation study (\cref{perf-test:sec:simulation-study}), showing that they perform well.
@@
 
-The CIS tests each individual infrequently, either weekly or monthly (see \cref{E-intro:sec:cis}), complicating the data in two ways.
+In the CIS, individuals are tested infrequently, either weekly or monthly (see \cref{E-intro:sec:cis}), complicating the data in two ways.
 First, infection episodes can go undetected.
-Second, the infection episodes' start and end times are known imprecisely.
-I explain these issues below and devote the rest of the chapter to the development of a method to tackle  them.
+Second, the infection episode start and end times are known imprecisely.
+I explain these issues below, and devote the rest of the chapter to the development of a method to tackle  them.
 
@@
 tests).
-The first individual was never infected whereas the second individual has an undetected infection. 
+The fist individual was never infected whereas the second individual was; but their infection episode was undetected.
 The undetected infection episode begins at time 33 and has a 10-day duration.
 The individual would have been detected at their next test if their duration had instead been 24 days or more.
-The gap between tests in the CIS (up to 28 days, see \cref{E-intro:sec:cis}) is longer than most infections (median 16 days, see \cref{E-ATACCC:sec:results}).
-Therefore, it is likely that many infections in the CIS cohort were undetected.
+The gap between tests in the CIS (up to 28 days, see \cref{E-intro:sec:cis}) is longer than the duration of most infections (median 16 days, see \cref{E-ATACCC:sec:results}).
+Therefore, it is likely that many infection in the CIS cohort were undetected.
 This is because the interval between tests in the same individual (up to 28 days, see \cref{E-intro:sec:cis}) is longer than most infections (median 16 days, see \cref{E-ATACCC:sec:results}).
@@
 In survival analysis, the interest is in estimating the distribution of the interval between an \emph{initiating event} and a \emph{terminating event}.
-In this context, the initiating event is the episode's start; the terminating event is the episode's end.
+In the current context, the initiating event is the episode's start; the terminating event is the episode's end.
 I start by introducing the relevant, standard concepts in survival analysis; a wide range of texts cover them including \textcite[chapter 1]{bogaertsSurvival}.
@@
 This is known as the event being \emph{censored}.
-An event is \emph{left censored} if it is known to occur before a certain time.
-An event is \emph{right censored} if it is known to occur after a certain time.
-An event is \emph{interval censored} if it is known to occur within an interval.
-If both the initiating and terminating event are censored, the data are \emph{doubly censored}.
-To handle doubly censored data, the times need to be jointly modelled~\autocite[and references therein]{liSemiparametric}.
-If both events are interval censored, then the data are \emph{doubly interval censored}.
+An event is \emph{left censored} if it is known to have occurred before a certain time.
+An event is \emph{right censored} if it is known to have occurred after a certain time.
+An event is \emph{interval censored} if it is known to have occurred within an interval.
+If both the initiating and terminating event are interval censored, the data are \emph{doubly interval censored}.
+To handle doubly interval censored data, the times need to be jointly modelled~\autocite[and references therein]{liSemiparametric}.
 
+Another important concepts typical of survival data is truncation, which refers to ...?
 Sometimes, only intervals satisfying some property are recorded.
@@
 \emph{Right truncation} is when only events lasting shorter than a maximum amount of time are observed.
-Without correcting for the truncation, inference will be biased.
+Without correcting for truncation, inference on the distribution of a duration will be biased.
 
@@
 
-There are also undetected infections in the CIS (see \cref{perf-test:sec:undetected}).
+There is also undetected infections in the CIS (see \cref{perf-test:sec:undetected}).
 However, the situation is subtly different from the standard truncation situation encountered in the literature.
@@
 Therefore, we know the times that they were tested.
-This provides information on how long their infections may have been.
+This provides information on how long their infections may have occurred.
 I refer to these events as ``undetected'', rather than truncated, to emphasise this difference.
@@
 In particular, when the interval lengths are near-constant, $p(\text{episode $j$ detected} \mid b_j \in [l_j^{(b)}, r_j^{(b)}]) \approx p(\text{episode $j$ detected})$.
-However, the interval lengths change from one to four weeks in the CIS (see \cref{E-intro:sec:cis}).
+However, in the CIS (see \cref{E-intro:sec:cis}) the interval lengths change from one to four weeks.
 
@@
 The test schedule $\sched_i$ is the set of times individual $i$ is tested at, from their last test prior to time 1 (if it exists) up until time $T$.
-These are all the times an infection episode can be detected at, plus one prior test
+This is all the times the infection episode could be detected at and be included, plus one prior test.
 For example, for the individual in \cref{perf-test:sec:undetected}, $\sched_i = \{ 0, 7, 14, 21, 28, 56 \}$.
@@
 If $j$ is detected, then $i(j)$ is known; $b_j \in [l_j^{(b)}, r_j^{(b)}]$; and $e_j \in [l_j^{(e)}, r_j^{(e)}]$.
-These latter two properties are what make the data doubly interval censored.
+These latter two properties are what makes the data doubly interval censored.
 
@@
   Otherwise, if $i \in \set{D}$, then individual $i$ had $n_{ia} =1$ admissible episode.
-  The admissible region is the intersection of the intervals $[l_j^{(b)}, r_j^{(b)}]$ and $[l_j^{(e)}, r_j^{(e)}]$.
+  The admissible region is defined by the intersection of the intervals $[l_j^{(b)}, r_j^{(b)}]$ and $[l_j^{(e)}, r_j^{(e)}]$.
   This is the purple area in \cref{perf-test:fig:partitionSpace}; the boundaries in the $x$ direction of the area are $l_j^{(b)}$ and $r_j^{(b)}$ and in the $y$ direction are $l_j^{(e)}$ and $r_j^{(e)}$.
@@
 I assume the individual in which each episode occurs is iid uniform across the cohort.
-Therefore, these probabilities are multiplied by $1/\Ncis$ to form the unconditional probabilities.
+Therefore, the unconditional probabilities of an infection belonging to a class need to be multiplied by $1/\Ncis$.
 For example, the probability of an episode being in class $\alpha_i$ is $p_{ia} / \Ncis$.
@@
 This is not strictly correct because the cell probabilities in the likelihood should not depend on the data.
-However, an alternative considering each combination of infection and recovery times as its own cell is possible.
+However, an alternative where each combination of infection and recovery times is considered its own cell is possible.
 The new cells would align with the inadmissible region.
-Since each of these cells have an observed count of zero, they can be combined to give the same likelihood.
+Since each of these cells have an observed count of 0, they can be combined to give the same likelihood.
 That is, the approach is mathematically equivalent but greatly complicates the formulae and notation.
@@
 The undetected classes cannot be distinguished.
-Therefore, I combine these classes into a single class.
+Therefore, these classes can be combined into a single class.
 The combined class has a count $\nnodet = \sum_{i=1}^{\Ncis} n_{iu}$ and probability $\pnodet = \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} p_{iu}$.
@@
 Therefore, the crucial component is $1 - p_{iu} = 1 - \prob((B_j, E_j) \in \Omega_i \mid \vec{\theta}, i(j) = i)$.
-This is one minus the probability that an episode in individual $i$ was undetected, \ie the probability of detecting an episode.
+This is one minus the probability that an episode in individual $i$ was undetected, \ie the probability that episode such an episode is detected.
 
@@
       For episodes detected in the period I consider, this happens a negligible number of times, only in individuals lost to follow-up; therefore, I neglect this possibility.
-      For a new context, including recent infections, this condition could be relaxed by considering episodes that do not meet this criterion as right censored.
+      For a new context, including recent infections, this condition could be relaxed by considering episodes that do not meet this criteria as right censored.
 \end{enumerate}
@@
     (B) Prior predictive survival time (mean and 95\% CrI).
-    The implied prior on the survival time is short when using either prior on the hazard at each time.
+    The implied prior on the survival time is very short when using either prior on the hazard at each time.
   }
@@
 Smoothing priors~\autocite{gerschSmoothness} encode the biological consideration that the hazard should not vary significantly from day-to-day.
-They serve the same purpose as a penalized likelihood, which is often used in the non-parametric maximum-likelihood setting~\autocite[e.g.][]{bacchettiNonparametric}.
+They serve the same purpose as a penalized likelihood, which is used frequently in the non-parametric maximum-likelihood setting~\autocite[e.g.][]{bacchettiNonparametric}.
 Many possible forms of smoothing priors exist, including splines, Gaussian Processes, and random walks.
@@
 To form the duration distribution used in the simulation, we combine the two estimates.
-The pdf over the first 30 days is proportional to the ATACCC estimate, with the rest proportional to this CIS-based estimate.
+The pdf over the first 30 days is proportional to the ATACCC estimate, with the rest proportional to this CIS-based estimates.
+% The CIS-based estimates are shifted 3 days in order to make this smooth; this counteracts the above bias of missing the start of the infection (for times above 30 days, the other bias, missing short infections, is negligible).
+ERR...
+I'M NOT DOING THIS ANYMORE OH NO!!
 Denote by $f_A(t)$ the distribution function estimated in \cref{E-ATACCC} and $f_C(t)$ that from the CIS-based estimates just derived.

### 2024-03-25 introduction.tex
Updates from Overleaf

--- 2024-03-25 introduction.tex
@@
 COVID-19 (coronavirus disease 2019), the disease caused by the virus SARS-CoV-2 (severe acute respiratory syndrome coronavirus 2), is estimated to have killed approximately 15 million in the most acute phase (to the end of 2021) and 30 million in total~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
+
 History suggests severe pandemics are not rare (see \cref{intro:fig:pandemic-timeline}).
@@
 For a given disease and population, an analogous growth rate can be defined for each definition of the reproduction number~\autocite{pellisEstimation,paragGrowthRates,wallingaGI}.
-The advantage of using the growth rate it is easier to estimate and captures how quickly spread happens in calendar time; however, they do not directly give the strength of intervention required to control the epidemic~\autocite{royalSocietyRnumber}.
+The advantages of using the growth rates are that they are easier to estimate and capture how quickly spread happens in calendar time; however, they do not directly give the strength of intervention required to control the epidemic~\autocite{royalSocietyRnumber}.
 
@@
 \emph{Prevalence} is the proportion of the population with the disease at a given time.
-Whether a given individual is \emph{prevalent} (has the disease) is often measurable through a diagnostic test.
+Whether a given individual is \emph{prevalent} (has the disease) is often measurable through diagnostic testing.
 Therefore, the prevalence of the population can be estimated with a prevalence survey, although these are typically very expensive to power appropriately.
@@
 
-Collecting data which directly informs inference of the quantities in this section is expensive or impossible.
+Collecting data that directly inform inference of the quantities in this section is expensive or impossible.
 Hence, they are typically inferred from other data sources through modelling.
@@
 
-The data sources used for outbreak surveillance are often observational in nature.
+Data sources used for outbreak surveillance are often observational in nature.
 Therefore, they suffer from biases that can be hard to understand and quantify~\autocite{shadboltChallenges}.
 Here I discuss three types of data which are useful for surveillance.
-Incidence-type and prevalence-type data can directly inform estimation of incidence and transmission; however, this estimation almost always relies on prior knowledge of epidemiological and/or biological quantities.
+Incidence-type and prevalence-type data can inform estimation of incidence and transmission; however, this estimation almost always relies on prior knowledge of epidemiological and/or biological quantities.
 For example, the time between infection and symptom onset.
@@
 
-Incidence-type data originates from observations of the consequences of a disease, \eg the development of symptoms or admission to hospital.
+Incidence-type data originate from observations of the consequences of a disease, \eg the development of symptoms or admission to hospital.
 This type of data typically suffers from reporting biases, particularly that the proportion of infections which are counted can vary over time or within different subgroups of the population~\autocites[chapter 9]{lashModern}{shadboltChallenges}.
@@
 This means that the observations do not provide much information on incidence in those subgroups.
-Furthermore, the delay between infection and the severe event occurring means that recent infections cannot be estimated~\autocite{swallow2022challenges}.
+Furthermore, the delay between infection and the severe event occurring means that recent infections cannot be estimated with much precision~\autocite{swallow2022challenges}.
 
@@
 
-Prevalence-type data originates from observations of the number or proportion of a population in a specific disease state at a specific time.
+Prevalence-type data originate from observations of the number or proportion of a population in a specific disease state at a specific time.
 For example, the number of hospital beds occupied by individuals with the disease or the proportion of the population with a marker of prior infection.
@@
 These have components of the transmission process in them but do not fully represent the process.
-The most common type of semi-mechanistic model is those based on a \emph{renewal equation}~\autocite{bhattSemimechanistic}.
+A common type of semi-mechanistic model is those based on a \emph{renewal equation}~\autocite{bhattSemimechanistic}.
 The renewal equation~\autocite{fraserRenewal} for an infectious disease, in the most general form, is:
@@
 \end{align}
-where $Z(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the mean rate an individual infected $\tau$ time units ago infects individuals at time $t$.
+where $Z(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the mean rate at which an individual infected $\tau$ time units ago infects individuals at time $t$.
 Therefore, the renewal equation relates incidence in the past to incidence in the future.
@@
 % The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
-Extensions of \cref{intro:eq:renewal} to track the number of susceptible individuals become full mechanistic models, equivalent to the compartmental models introduced next~\autocite{champredonEquivalence}.
+Extensions of \cref{intro:eq:renewal} that track the number of susceptible individuals become full mechanistic models, equivalent to the compartmental models introduced next~\autocite{champredonEquivalence}.
 Renewal equations do not feature further in this thesis; for details, including practical considerations when using them as the basis for inference, see \textcite{nashEstimating,gosticPractical}.
 
-Finally, mechanistic models are the most complex, representing the transmission process in varying levels of detail.
+Finally, mechanistic models are the most complex models, representing the transmission process in varying levels of detail.
 \emph{Compartmental} models, the most common type of mechanistic model, split the population into compartments, each representing a different state of the disease.
@@
 Solving \cref{intro:eq:disease-model} for $Z$ is an ill-posed problem when $f(0)$ is small.
-The forwards error of the solution is large when the ratio of the off-diagonal to diagonal elements is large~\autocite[chapter 8.2]{highamAccuracy}; for this situation, which is when the value of $\max_{i > 0} f(i) / f(0)$ is large.
+The forwards error of the solution is large when $\max_{i > 0} f(i) / f(0)$~\autocite[chapter 8.2]{highamAccuracy}.
 A large forwards error means that the computed value is far from the true solution~\autocite[6--9]{highamAccuracy}; this is avoided by either using an approach which gives a large $f(0)$ or smoothing $Z_t$.
@@
 
-An observation model links $W_t$ to a possibly biased and/or noisy observation of it at time $t$.
+Observation models link $W_t$ to a possibly biased and/or noisy observation of it at time $t$.
 
@@
 An individual is in the disease state while the biomarker is above or below some threshold.
-This approach often requires parametric assumption(s).
-For example, the functional form of the biomarker's trajectory.
-In addition, if individual-level variation wants to be included, which is generally important, a distributional form of random effects is required.
-This requires both sufficient observations per individual and an assumption on the form of the distribution.
+This approach often requires parametric assumption(s) such as a functional form of the biomarker's trajectory.
+In addition, if individual-level variation is included, which is generally important, a distributional form of random effects is required.
+This requires both sufficient observations of the biomarker per individual and an assumption on the form of the distribution.
 More details are in \cref{E-ATACCC}.
@@
 Modern technology allowed large amounts of data to be collected, distributed, and analysed.
-While the HIV/AIDS pandemic inspired the development of a large amount of statistical methodology, the pandemic was far slower moving.
+While the HIV/AIDS pandemic inspired the development of a large amount of statistical methodology, it was far slower moving.
 Much analysis of SARS-CoV-2 data was done at great haste, requiring trade-offs to be made between timeliness and rigour of results presented.
@@
 For large parts of the pandemic, community testing was widely available in the UK.
-However, it did not test a random or even representative sample of the population.
+However, it did not test a representative sample of the population.
 The number of tests conducted varied over time and space, for a variety of known and unknown reasons.
@@
 
-Severe events counts are generally a reliable measure of the burden of the disease.
+Severe events counts are generally a more reliable measure of the burden of the disease.
 The criteria for being a severe event are normally well-defined and generally well-recorded in the UK context.
@@
 The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys in representative samples of the general population.
-These were not subject to the biases of community testing and observed the whole population, unlike severe events, although suffered from non-response bias.
+These were designed to avoid the biases of community testing and observe the whole population.
 Therefore, some modelling is required to borrow information across strata or time to provide stable estimates, as well as to correct for non-response bias (see \cref{E-intro:sec:cis}).
@@
 REACT was a series of cross-sectional surveys, run across 19 rounds between May 2020 and March 2022.
-Each round consisted of at least \numprint{100000} randomly sampled individuals from the UK contributing swabs for testing and questionnaire data~\autocite{elliottTwin}.
+Each round consisted of at least 100,000 randomly sampled individuals from the UK contributing swabs for testing and questionnaire data~\autocite{elliottTwin}.
 
@@
 \Textcite{ealesCharacterising} estimated the duration of positivity from REACT data.
-Specifically, individuals who tested positive in round 8 of REACT (6th--22 Jan 2021) were invited to take two additional swab tests.
+Specifically, individuals who tested positive in round 8 of REACT (6th--22nd Jan 2021) were invited to take two additional swab tests.
 Defining the time of the initial positive test at time 0, \textcite{ealesCharacterising} estimated the probability of remaining positive at each day after the initial positive test.
@@
 
-In order to achieve these aims, I use a case study, specifically the SARS-CoV-2 epidemic in England from Mon 31 Aug 2020 until Sun 25 Jan 2021.
+In order to achieve these aims, I use a case study, specifically the SARS-CoV-2 epidemic in England from Mon 31st Aug 2020 until Sun 25th Jan 2021.
 This case study has several features that make it useful for the aims of this thesis.

### 2024-03-25 ATACCC-appendix-original-analysis.tex
Tweak final plots / captions

@@
 \begin{figure}
-    \centering
-    \vspace*{-3cm}
+    \thisfloatpagestyle{empty}
+    \vspace*{-1cm}
     \makebox[\textwidth][c]{\includegraphics[scale=0.9]{ATACCC/appendix/goodness_of_fit}}
@@
 This could be because of the different populations in each study.
-\Textcite{kisslerViral}'s data is collected from longitudinal swabbing of individuals professionally involved in the National Basketball Association (NBA).
+\Textcite{kisslerViral}'s data are collected from longitudinal swabbing of individuals professionally involved in the National Basketball Association (NBA).
 Of the individuals who were included in their analysis, around half were in professional athletes, likely in peak physical condition, and 90\% were male.
-Our data is collected from the general population who were reported as contacts of individuals detected through surveillance open to all members of the public.
+Our data are collected from the general population who were reported as contacts of individuals detected through surveillance open to all members of the public.
 Therefore, our sample is likely to be, on average, in worse physical condition which may lead to a higher viral load (lower Ct).

### 2024-03-25 biology-data.tex

@@
 
-The household-based design of the CIS means that the data is clustered.
+The household-based design of the CIS means that the data are clustered.
 In particular, individuals in the same household are likely to transmit to each other and hence have correlated test results.
@@
 
-For privacy reasons, the CIS data is stored in the ONS's SRS (Secure Research Service)~\autocite{onsSRS}.
+For privacy reasons, the CIS data are stored in the ONS's SRS (Secure Research Service)~\autocite{onsSRS}.
 The SRS is a Trusted Research Environment (TRE) that allows researchers to access sensitive data in a secure environment.

### 2024-03-25 cis-perfect-testing.tex

--- 2024-03-25 cis-perfect-testing.tex
@@
 
-The CIS tests each individual infrequently, either weekly or monthly (see \cref{E-intro:sec:cis}), complicating the data in two ways.
+In the CIS, individuals are tested infrequently, either weekly or monthly (see \cref{E-intro:sec:cis}), complicating the data in two ways.
 First, infection episodes can go undetected.
@@
 The individual would have been detected at their next test if their duration had instead been 24 days or more.
-The gap between tests in the CIS (up to 28 days, see \cref{E-intro:sec:cis}) is longer than most infections (median 16 days, see \cref{E-ATACCC:sec:results}).
+The gap between tests in the CIS (up to 28 days, see \cref{E-intro:sec:cis}) is longer than the duration of most infections (median 16 days, see \cref{E-ATACCC:sec:results}).
 Therefore, it is likely that many infections in the CIS cohort were undetected.
@@
   \caption[Double-interval censoring in CIS data]{%
-        The beginning and end of each episode observed in the CIS data is only known up to an interval.
+        The beginning and end of each episode observed in the CIS data are only known up to an interval.
         Demonstrated here by a participant who is recorded negative at time 7 and positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode.
@@
 This assumption performs well in these settings, when the length of the intervals is fairly constant.
-A necessary condition for it to perform well, satisfied when the interval lengths are near-constant, is $p(\text{episode $j$ detected} \mid b_j \in [l_j^{(b)}, r_j^{(b)}]) \approx p(\text{episode $j$ detected})$.
+In particular, when the interval lengths are near-constant, $p(\text{episode $j$ detected} \mid b_j \in [l_j^{(b)}, r_j^{(b)}]) \approx p(\text{episode $j$ detected})$.
 However, the interval lengths change from one to four weeks in the CIS (see \cref{E-intro:sec:cis}).

### 2024-03-25 introduction.tex

--- 2024-03-25 introduction.tex
@@
 For a given disease and population, an analogous growth rate can be defined for each definition of the reproduction number~\autocite{pellisEstimation,paragGrowthRates,wallingaGI}.
-The advantage of using the growth rate is that it isn easier to estimate and captures how quickly spread happens in calendar time; however, they do not directly give the strength of intervention required to control the epidemic~\autocite{royalSocietyRnumber}.
+The advantages of using the growth rates are that they are easier to estimate and capture how quickly spread happens in calendar time; however, they do not directly give the strength of intervention required to control the epidemic~\autocite{royalSocietyRnumber}.
 
@@
 
-The data sources used for outbreak surveillance are often observational in nature.
+Data sources used for outbreak surveillance are often observational in nature.
 Therefore, they suffer from biases that can be hard to understand and quantify~\autocite{shadboltChallenges}.
 Here I discuss three types of data which are useful for surveillance: incidence-type, prevalence-type, and data from cohort studies.
-\emph{Incidence-type} and \emph{prevalence-type} data can directly inform estimation of incidence and transmission; however, this estimation almost always relies on prior knowledge of epidemiological and/or biological quantities.
+\emph{Incidence-type} and \emph{prevalence-type} data can inform estimation of incidence and transmission; however, this estimation almost always relies on prior knowledge of epidemiological and/or biological quantities.
 For example, the time between infection and symptom onset.
@@
 These have components of the transmission process in them but do not fully represent the process.
-The most common type of semi-mechanistic model is those based on a \emph{renewal equation}~\autocite{bhattSemimechanistic}.
+A common type of semi-mechanistic model is those based on a \emph{renewal equation}~\autocite{bhattSemimechanistic}.
 The renewal equation for an infectious disease~\autocite{fraserRenewal}, in the most general form, is:
@@
 \end{align}
-where $Z(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the mean rate an individual infected $\tau$ time units ago infects individuals at calendar time $t$.
+where $Z(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the mean rate at which an individual infected $\tau$ time units ago infects individuals at calendar time $t$.
 Therefore, the renewal equation relates incidence in the past to incidence in the future.
@@
 % The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
-Extensions of \cref{intro:eq:renewal} to track the number of susceptible individuals become full mechanistic models, equivalent to the compartmental models introduced next~\autocite{champredonEquivalence}.
+Extensions of \cref{intro:eq:renewal} that track the number of susceptible individuals become full mechanistic models, equivalent to the compartmental models introduced next~\autocite{champredonEquivalence}.
 Renewal equations do not feature further in this thesis; for further details, including practical considerations when using them as the basis for inference, see \textcite{nashEstimating,gosticPractical}.
 
-Finally, mechanistic models are the most complex, representing the transmission process in varying levels of detail.
+Finally, mechanistic models are the most complex models, representing the transmission process in varying levels of detail.
 \emph{Compartmental} models, the most common type of mechanistic model, split the population into compartments, each representing a different state of the disease.
@@
 Solving \cref{intro:eq:disease-model} for $Z$ is an ill-posed problem.
-The forwards error of the solution is large when the ratio of the off-diagonal to diagonal elements is large~\autocite[chapter 8.2]{highamAccuracy}; for this situation, which is when the value of $\max_{i > 0} f(i) / f(0)$ is large.
+The forwards error of the solution is large when $\max_{i > 0} f(i) / f(0)$~\autocite[chapter 8.2]{highamAccuracy}.
 A large forwards error means that the computed value is far from the true solution~\autocite[6--9]{highamAccuracy}; this is avoided by either using an approach which gives a large $f(0)$ or smoothing $Z_t$.

### 2024-03-25 thesis.tex

@@
 \abstract{%
-    Surveillance is a crucial component to a pandemic response, informing public health and wider policy.
+    Surveillance is a crucial component of a pandemic response, informing public health and wider policy.
     Unfortunately, it is often subject to reporting and behavioural biases, reducing its robustness.
@@
     % One of these, the CIS (Coronavirus Infection Survey), was a longitudinal survey run from April 2020 to March 2023.
-    The CIS (Coronavirus Infection Survey) was a large-scale prevalence survey, with participants recruited randomly from the general population, therefore avoiding these issues.
+    The CIS (Coronavirus Infection Survey) was a large-scale prevalence survey, with participants recruited randomly from the general population, thereby avoiding these issues.
 
@@
     I use this detailed model to exploit frequent (daily) sampling of individuals in a cohort of contact traced individuals.
-    However, the cohort was small and follow-up short meaning that the tail of the duration distribution is poorly estimated.
+    However, the cohort was small and the follow-up time short meaning that the tail of the duration distribution is poorly estimated.
 
@@
     To overcome this issue, I combine the information with the larger and longer CIS dataset.
-    The CIS data poses several challenges because it is double interval censored and some individuals in the cohort have infections that are never detected.
+    The CIS data poses several challenges because it is doubly interval censored and some individuals in the cohort have infections that are never detected.
     I apply a survival analysis framework to analyse this data, which had previously only been used in simpler settings.
@@
 
-    Finally, I use this estimate to estimate the incidence and transmission of SARS-CoV-2 in England.
+    Finally, I use this estimate to infer the incidence and transmission of SARS-CoV-2 in England from the CIS.
     I fully propagate the sampling error, accurately include the shape of the duration distribution, and estimate the quantities by age; previous work lacked at least one of these.
@@
     Professor A.\ Sarah Walker created the episodes dataset in \cref{biology-data:sec:cis-episodes} and performed the initial survival analysis I use for a simulation study \cref{perf-test:sec:simulation-study}.
-    Dr Koen Pouwels provided code used as the basis of the positivity model in \cref{biology-data:sec:MRP}.
-    Dr Edwin van Leeuwen created the contact matrices and Dr Sanmitra Ghosh provided code for the MCMC sampler used in \cref{E-SEIR}.
+    Dr Koen Pouwels provided code I use as the basis of the positivity model in \cref{biology-data:sec:MRP}.
+    Dr Edwin van Leeuwen created the contact matrices and Dr Sanmitra Ghosh provided code for the MCMC sampler I use in \cref{E-SEIR}.
     Thank you all for your contributions.
@@
 
-    Several aspects of my work were done to support the ongoing response.
+    Several aspects of the work in this thesis were completed in support of the ongoing response.
     In particular, an earlier version of the backcalculation work was adopted by the Office for National Statistics to estimate incidence, published in their weekly online reports~\autocite{cisMethodsONS}.

### 2024-03-25 ATACCC-appendix-original-analysis.tex
proof

@@
 For an individual $i$, their trajectory is determined by two points and one slope parameter.
-The points are the time at which the individual has a 50\% chance of testing positive, $t_{i,0}$, and the time at which their viral load
+The points are the time at which the individual has a 50\% chance of testing positive, $t_{i,0}$, and the time at which their viral load.
 peaks, $t_{i,\text{peak}}$ after $t_{i,0}$; the slope parameter, $\beta_i$ determines the rate their Ct increases (viral load decreases) following the peak.

### 2024-03-25 biology-data.tex
proof

@@
 
-SARS-CoV-2 is closely related to previous coronaviruses that have caused outbreaks in humans, SARS-CoV and MERS-CoV.
+SARS-CoV-2 is closely related to previous coronaviruses that have caused outbreaks in humans, including SARS-CoV and MERS-CoV.
 Coronaviruses are RNA (Ribonucleic acid) viruses.
-RNA viruses use RNA, rather than DNA, as their genetic material.
+RNA viruses use RNA, rather than DNA (Deoxyribonucleic acid), as their genetic material.
 RNA viruses, which include influenza, mutate rapidly and are the most likely class of viruses to cause pandemics~\autocite{woolhouseRNA}.
@@
 The initial and most sensitive diagnostic test for SARS-CoV-2 uses RT-PCR, which I explain in \cref{biology-data:sec:PCR}.
-SARS-CoV-2 causes the disease COVID-19; I discuss the \emph{natural history} of the disease, the course it takes from beginning to end~\autocite[193]{portaEpiDictionary}, in \cref{biology-data:sec:natural-history}.
+I discuss the SARS-CoV-2 infection \emph{natural history}, the course it takes from beginning to end~\autocite[193]{portaEpiDictionary}, in \cref{biology-data:sec:natural-history}.
 
@@
 
-PCR (Polymerase Chain Reaction) is a procedure that produces millions to billions of copies of a specific segment of DNA (Deoxyribonucleic acid) from a single copy~\autocite{smithPCR,garibyanPCR}.
+PCR (Polymerase Chain Reaction) is a procedure that produces millions to billions of copies of a specific segment of DNA  from a single copy~\autocite{smithPCR,garibyanPCR}.
 This is known as \emph{amplifying} the DNA.
@@
 
-The process is a series of cycles, each cycle containing three steps~\autocite{powledgePCR,garibyanPCR}.
+The PCR process is a series of cycles, each cycle containing three steps~\autocite{powledgePCR,garibyanPCR}.
 The first step, \emph{denaturation}, heats the DNA causing its two strands to separate.
-The second step, \emph{annealing}, lowers the temperature, causing the primers to bind to the DNA.
+The second step, \emph{annealing}, lowers the temperature, causing the primers to bind to one of the strands of DNA.
 The final step, \emph{extension}, raises the temperature, causing the DNA polymerase to replicate the DNA sequence between the primers. 
@@
 The number of DNA copies doubles each cycle, therefore the log Ct value decreases linearly with the quantity of RNA or DNA in the original sample (see \cref{biology-data:fig:ct-calibration}).
-A \emph{standard curve} translates the Ct value to viral load for a given test.
+A \emph{standard curve}, also known as a \emph{calibration curve}, translates the Ct value to viral load for a given test~\autocite{bustinMIQE}.
 A standard curve is produced by measuring the Ct value using reference samples with known viral load, and, as seen in \cref{biology-data:fig:ct-calibration}, they vary between studies.
@@
 An individual is defined as detectable between the time they would first test positive on a perfect RT-PCR test (if tested) and the time they would stop doing so.
-It is not possible to conclusively determine whether an individual is detectable because perfect RT-PCR tests do not exist (see \cref{biology-data:sec:PCR}).
-However, viewing the number of RNA genome copies in an individual as a latent process this can then be defined as the time when the number of RNA genome copies is above the limit of detection of the RT-PCR test (defined in \cref{biology-data:sec:measurement-error}).
+It is not possible to conclusively determine whether an individual is detectable because perfect RT-PCR tests do not exist.
+However, viewing the number of RNA genome copies in an individual as a latent process, this can then be defined as the time when the number of RNA genome copies is above the limit of detection of the RT-PCR test (defined in \cref{biology-data:sec:measurement-error}).
 % While this definition is not perfectly precise, it is clear enough for the purposes it used in this thesis.
@@
 In this thesis, I focus on pre-Alpha variants of SARS-CoV-2 in unvaccinated individuals.
-The time from infection to infectiousness is known as the \emph{latent period}.
+The time from infection to infectiousness is known as the \emph{latent period}~\autocite{goughEstimation}.
 The length of the latent period is hard to estimate because we rarely observe either the time of infection or the start of infectiousness.
@@
 These estimates vary but suggest a mean length of 3--4 days for pre-Alpha variants~\autocite[and references therein]{zhaoEstimating}.
-The time from infection to experiencing symptoms is known as the \emph{incubation period}; it is undefined for asymptomatic individuals.
+The time from infection to experiencing symptoms is known as the \emph{incubation period}; it is undefined for asymptomatic individuals~\autocite{portaEpiDictionary}.
 As the symptom onset time (when symptoms start) is often known, this is easier to estimate than the latent period.
@@
 Viral replication requires the virus to infect a host cell and then use the host cell's machinery to produce more virus; as more host cells are infected this process slows because there are fewer remaining potential host cells, known as \emph{target cells}, to infect.
-Host cells which can potentially be infected are known as \emph{target cells}.
 The rate at which host cells are infected slows over time because there are fewer remaining target cells to infect.
@@
 
-Of the 738 contacts enrolled, 172 (23\%) had at least one positive test
+Of the 738 contacts enrolled, 172 (23\%) had at least one positive test.
 \Textcite{hakkiOnset} analysed the 57 contacts who tested positive for SARS-CoV-2 and had the growth phase in their viral load measured.

### 2024-03-25 cis-imperfect-testing.tex
proof ch 5
plural data
proof

@@
     Test sensitivity is $\psens$.
-    The CIS data is shown as a vertical line.
+    The CIS data are shown as a vertical line.
   }
@@
 I bound the test sensitivity on each day by assuming either all or none of the possible false negatives (group 3) are in fact false negatives.
-% In \cref{imperf-test:fig:bounding-cis-sensitivity} we consider the values produced by assuming the negative following the last positive in an episode could be either a true or false negative as a function of time since the individual was detected (\ie: the first positive test in the episode).
+% In \cref{imperf-test:fig:bounding-cis-sensitivity} we consider the values produced by assuming the negative following the last positive in an episode could be either a true or false negative as a function of time since the individual was detected (\ie the first positive test in the episode).
 This generates broad bounds, although suggest a declining test sensitivity over the course of an episode (see \cref{imperf-test:fig:bounding-cis-sensitivity}).
@@
     Posterior (median and 95\% credible interval) of the survival time for the simulation study with a correctly specified test sensitivity.
-    The true survival time is shown in black
+    The true survival time is shown in black.
   }
@@
     Posterior (median and 95\% credible interval) of the survival time for the simulation study with a possibly misspecified test sensitivity.
-    The true survival time is shown in black
+    The true survival time is shown in black.
     All simulations use a constant test sensitivity of 0.8, but the inference procedure assumes different values, as per the key.
@@
 
-% However, comparing the CIS and simulated datasets do not show any differences which suggest what in the data is causing these differences.
+% However, comparing the CIS and simulated datasets do not show any differences which suggest what in the data are causing these differences.
 % Then, the additional undetected infections implied by a 

### 2024-03-25 cis-perfect-testing.tex
proof

@@
 Therefore, it is suitable for estimating the tail of the duration distribution.
-In this chapter, I develop a method to estimate the duration distribution using CIS data, assuming the test ???
-has perfect sensitivity and specificity, {\it i.e.} there are no false negatives or positives.
+In this chapter, I develop a method to estimate the duration distribution using CIS data, assuming the test has perfect sensitivity and specificity, {\it i.e.} there are no false negatives or positives.
 Formally, I aim to estimate the survival function $S(t)$ defined in \cref{inc-prev:sec:definitions}.
@@
 I evaluate the model and priors using a simulation study (\cref{perf-test:sec:simulation-study}), showing that they perform well.
-Finally, I discuss the results in \cref{perf-test:sec:discussion}.
+Finally, I discuss the results in \c{perf-test:sec:discussion}.
 
@@
 I use Bayesian inference for two reasons.
-First, the frequentist properties (\eg consistency and convergence rate) of proposed estimators are poorly understood~\autocite{sunAnalysis,dengNonparametric}.
+First, the frequentist properties (\eg consistency and convergence rate) of existing estimators are poorly understood~\autocite{sunAnalysis,dengNonparametric}.
 Second, a prior can include information from previous studies, such as \cref{E-ATACCC}.
@@
 
-In this section, I develop a computationally feasible statistical model accounting for double interval censoring and undetected episodes.
+In this section, I develop a computationally feasible Bayesian statistical model accounting for double interval censoring and undetected episodes.
 In \cref{perf-test:sec:data-notation}, I introduce the data and notation, defining which quantities are observed and which are latent.
-Then, in \cref{perf-test:sec:episode-types}, I apply the framework of \textcite{heiseyModelling} which classifies possible episode beginning and end time into \emph{admissible} episodes, which match the doubly interval censored pattern of the observations; \emph{undetected} episodes, which may have occurred and been undetected; and \emph{inadmissible} episodes, which the data rules out having occurred.
+Then, in \cref{perf-test:sec:episode-types}, I adapt the framework of \textcite{heiseyModelling} which classifies possible episode beginning and end time into \emph{admissible} episodes, which match the doubly interval censored pattern of the observations; \emph{undetected} episodes, which may have occurred and been undetected; and \emph{inadmissible} episodes, which the data rules out having occurred.
 The number of such episodes in each class has a multinomial distribution conditional on the latent total number of infections (including undetected infections).
@@
 
-I consider the cohort of $\Ncis$ individuals who could have had a detected episode; this is a subset of the whole CIS cohort.
+I consider the cohort of $\Ncis = \numprint{437590}$ individuals who could have had a detected episode; this is a subset of the whole CIS cohort.
 To be considered, an individual must have had at least three tests before time $T$ (a negative prior to the episode, a positive, and a negative following the episode).
@@
 For a detected episode $j$, five pieces of data are observed (see \cref{E-biology-data:sec:cis-episodes} for how these are constructed).
-The individual in which the episode occurs, $i(j)$; the earliest and latest time the episode could have begun, $l_j^{(b)}$ and $r_j^{(b)}$ respectively; and the earliest and latest time the episode could have ended, $l_j^{(e)}$ and $r_j^{(e)}$ respectively.
+The individual in which the episode occurs, denoted $i(j)$; the earliest and latest time the episode could have begun, $l_j^{(b)}$ and $r_j^{(b)}$ respectively; and the earliest and latest time the episode could have ended, $l_j^{(e)}$ and $r_j^{(e)}$ respectively.
 Over the short calendar period considered, each individual has zero or one detected episodes.
-Let $i \in \set{D}$ indicate individual $i$ has a detected episode.
+Let $i \in \set{D}$ indicate individual $i$ has a detected episode and denote $j(i)$ as the detected episode in individual $i$; $j(i)$ is only defined for $i \in \set{D}$.
 
-In addition, the individuals in the cohort experienced an unknown number of undetected infections, $\nnodet$.
-The total number of infections in the cohort is $\ntot = \ndet + \nnodet$ and I index them with $j = 1, \dots, \ntot$.
+In addition, the individuals in the cohort experienced a total of $\nnodet$ undetected infection episodes.
+$\nnodet$ is a latent quantity.
+The total number of infection episodes in the cohort is $\ntot = \ndet + \nnodet$ and I index them with $j = 1, \dots, \ntot$.
 
-Any infection episode, whether detected or not, is fully characterised for my purposes by the triplet $(b_j, e_j, i(j))$ where $b_j$ and $e_j$ are the times the episode begin and end respectively, and $i(j)$ denotes the individual the episode occurred in.
+Any infection episode, whether detected or not, is fully characterised for my purposes by the triplet $(b_j, e_j, i(j))$ where $b_j$ and $e_j$ are the times the episode begin and end respectively, and $i(j)$ is the individual the episode occurred in.
 The episode's duration is $d_j = e_j - b_j + 1$ (see \cref{E-inc-prev:sec:definitions}).
@@
 To relate the possible value for $b_j$ and $e_j$, given a test schedule $\sched_{i(j)}$, to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie are strict subsets of $T \times T$), shown in \cref{perf-test:fig:partitionSpace}.
-Each episode, whether detected or not, must fall into one of these three classes with a probability that is a function of the parameters of the survival function, $\vec{\vec{\theta}}$, specified in \cref{perf-test:sec:parameters-priors}.
+Each episode, whether detected or not, must fall into one of these three classes with a probability that depends on the parameters of the survival function, $\vec{\theta}$, specified in \cref{perf-test:sec:parameters-priors}.
+The regions are defined for each individual $i$.
 
@@
 \item
-  Admissible episodes, $\alpha_i$, which have an infection and recovery time consistent with the data observed.
-  This space is empty for $i \notin \set{D}$; in this case individual $i$ had $n_{ia} = 0$ admissible episodes.
-  Otherwise, if $i \in \set{D}$, then individual $i$ had $n_{ia} =1$ admissible episode.
-  The admissible region is the intersection of the intervals $[l_j^{(b)}, r_j^{(b)}]$ and $[l_j^{(e)}, r_j^{(e)}]$.
+  The region of admissible episodes for individual $i$, $\alpha_i$.
+  An admissible episode has an infection and recovery time consistent with the data observed.
+  Let $n_{ia}$ be the number of admissible episodes for $i$.
+  $\alpha_i$ is empty if $i$ had no detected episodes (\ie $i \notin \set{D}$); in this case the observation is $n_{ia} = 0$.
+  Otherwise, if $i \in \set{D}$, then $n_{ia} =1$.
+  The admissible region is the intersection of the intervals $[l_{j(i)}^{(b)}, r_{j(i)}^{(b)}]$ and $[l_{j(i)}^{(e)}, r_{j(i)}^{(e)}]$.
   This is the purple area in \cref{perf-test:fig:partitionSpace}; the boundaries in the $x$ direction of the area are $l_j^{(b)}$ and $r_j^{(b)}$ and in the $y$ direction are $l_j^{(e)}$ and $r_j^{(e)}$.
   The shape of this region accounts for the doubly interval censored nature of the data.
-  Conditional on the episode occurring in individual $i$, the episode is admissible with probability $p_{ia} = \prob((B_j, E_j) \in \alpha_i \mid \vec{\theta}, i(j) = i)$.
+  Let $p_{ia} = \prob((B_j, E_j) \in \alpha_i \mid \vec{\theta}, i(j) = i)$, the probability that $j \in \alpha_i$ given that $j$ occurs in individual $i$.
   \label{perf-test:def:admissible}
 \item
-  Undetected episodes, $\Omega_i$, which have an infection and recovery time such that they would not have tested positive.
-  An unknown number, $n_{iu}$ of these occur in individual $i$.
+  The region of undetected episode for individual $i$, $\Omega_i$.
+  Let $n_{iu}$ be the (latent) number of undetected episodes for $i$.
   The undetected region is the red region in \cref{perf-test:fig:partitionSpace}, indicating combinations of episode beginning and end times which do not span a testing time plus episodes beginning before the individual's enrolment.
-  The shape of this region accounts for the pattern of possible undetected infections.
-  Conditional on the episode occurring in individual $i$, the episode is undetected with probability $p_{iu} = \prob((B_j, E_j) \in \Omega_i \mid \vec{\theta}, i(j) = i)$.
+  Let $p_{iu} = \prob((B_j, E_j) \in \Omega_i \mid \vec{\theta}, i(j) = i)$, the probability that $j \in \Omega_i$ given that $j$ occurs in individual $i$.
 \item
-  Inadmissible episodes, $\beta_i$, which would be detected (if they occur) but were not.
-  Therefore, we know $n_{i\bar{a}} = 0$ of these occurred in individual $i$.
-  These are all remaining episodes not in the previous sets.
-  Conditional on the episode occurring in these individuals, the episode inadmissible with probability $p_{i\bar{a}} = \prob((B_j, E_j) \in \beta_i \mid \vec{\theta}, i(j) = i)$.
+  The region of inadmissible episodes, $\beta_i$ for individual $i$.
+  An inadmissible episode is one which would have been detected had it occurred, but the data rule out having occurred.
+  Let $n_{i\bar{a}}$ be the number of inadmissible episodes for $i$, which is observed as $n_{i\bar{a}} = 0$.
+  These are all remaining (a priori possible) episodes not in the previous sets.
+  Let $p_{i\bar{a}} = \prob((B_j, E_j) \in \beta_i \mid \vec{\theta}, i(j) = i)$, the probability that $j \in \beta_i$ given that $j$ occurs in individual $i$.
 \end{itemize}
@@
   Inadmissible regions are unshaded.
-  The black (impossible) region covers the cases in which recovery occurs prior to infection.
+  The grey (impossible) region covers the cases in which recovery occurs prior to infection.
   \label{perf-test:fig:partitionSpace}
@@
 
-These three classes span all possible episodes and are mutually exclusive.
-Hence, $\ntot = \ndet + \nnodet + n_{i\bar{a}} = \sum_{i=1}^{\Ncis} n_{ia} + \sum_{i=1}^{\Ncis} n_{iu} + 0$.
-Furthermore, for all $i$, $p_{ia} + p_{iu} + p_{i\bar{a}} = 1$.
+These three regions span all possible episodes and are mutually exclusive.
+Hence the following relations hold:
+\begin{align}
+\ntot
+&= \ndet + \nnodet + \sum_{i=1}^{\Ncis} n_{i\bar{a}} \\
+&= \sum_{i=1}^{\Ncis} n_{ia} + \sum_{i=1}^{\Ncis} n_{iu} + 0
+\intertext{and, for all $i$,}
+p_{ia} + p_{iu} + p_{i\bar{a}} &= 1.
+\end{align}
+
+I will use these regions to form the data likelihood, conditional on $\ntot$.
+The definition of the admissible region relies on the test results.
+However the likelihood should be conditional only on $\vec{\theta}$, $\ntot$, and the test schedules.
+Therefore, this is not strictly correct.
+This issue can be rectified be considering the inadmissible region as many subregions each with an observed count of zero (sketched in \cref{perf-test:sec:alt-likelihood}).
+Count zero regions do not contribute to the probability meaning the result is the same.
+A formalisation of this approach requires much more complex notation for the same result; therefore, I prefer this presentation.
 
 % Any episode $j$ independently belongs to one of these classes.
-The probabilities ($p_{iu}$, $p_{ia}$, and $p_{i\bar{a}}$), are conditional on the episode occurring in individual $i$.
-I assume the individual in which each episode occurs is iid uniform across the cohort.
-Therefore, these probabilities are multiplied by $1/\Ncis$ to form the unconditional probabilities.
+The probabilities $p_{iu}$, $p_{ia}$, and $p_{i\bar{a}}$ are conditional on the episode occurring in individual $i$.
+I assume the individual in which an episode occurs is iid uniform across the cohort, that is $\prob(i(j) = i) = 1/\Ncis$.
+Therefore, the conditional probabilities are multiplied by $1/\Ncis$ to form the unconditional probabilities.
 For example, the probability of an episode being in class $\alpha_i$ is $p_{ia} / \Ncis$.
 
-Each episode occurs in one of the classes independently and with the same probability.
+An episode occurs in one of the regions independently and with the same probability.
 Therefore:
@@
   \end{bmatrix}
-  \mid \ntot
-  =
+  \mid \ntot, \vec{\theta}
+  \dist
   \MN\left(
@@
 
-In forming the region $\alpha_i$, I use the test outcomes.
-This is not strictly correct because the cell probabilities in the likelihood should not depend on the data.
-However, an alternative considering each combination of infection and recovery times as its own cell is possible.
-The new cells would align with the inadmissible region.
-Since each of these cells have an observed count of zero, they can be combined to give the same likelihood.
-That is, the approach is mathematically equivalent but greatly complicates the formulae and notation.
-Therefore, I present this version for clarity.
-
-The undetected classes cannot be distinguished.
-Therefore, I combine these classes into a single class.
+The undetected regions cannot be distinguished.
+Therefore, I combine these regions into a single cell within the multinomial distribution.
 The combined class has a count $\nnodet = \sum_{i=1}^{\Ncis} n_{iu}$ and probability parameter $\pnodet = \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} p_{iu}$.
@@
 From \cref{perf-test:eq:full-posterior}, this is only needed for $i \in \set{D}$, that is for individuals with detected episodes.
-For such an individual, let $j(i)$ denote their detected episode.
 
@@
 When constructing the prior, two aspects need consideration.
-Firstly, the model structure from \cref{E-ATACCC} leads to positive correlation in the posterior estimates of the hazard, which should be propagated into this analysis.
-That is, the prior used in this analysis should have correlation between $\lambda_t$ and $\lambda_{t'}$.
+Firstly, the model structure from \cref{E-ATACCC} leads to the posterior distribution on the hazards having a positive correlation between $\lambda_t$ and $\lambda_{t'}$, especially when $t$ and $t'$ have a small difference.
+The prior used in this analysis should preserve this correlation.
 Secondly, the uncertainty in the estimates from \cref{E-ATACCC} is underestimated because it extrapolates beyond the data using strong model assumptions (see \cref{E-ATACCC:sec:discussion}).
@@
 \ifSubfilesClassLoaded{
+  \appendix
+  \subfile{cis-perf-new-ll}
   \listoftodos

### 2024-03-25 biology-data.tex
proof ch 2

@@
 This exclusion was to ensure viral load trajectories are identifiable.
-I will revisit and extend this analysis in \cref{E-ATACCC}.
+I will revisit and extend their analysis in \cref{E-ATACCC}.
 
@@
 The CIS had a household-based design, meaning that households were invited to participate in the study.
-From households that participated, all individuals aged 2 and over were invited to participate.
+From the households that participated, all individuals aged 2 and over were invited to participate.
 Once invited, an enrolment swab would be taken at the first visit followed by 4 further weekly visits (giving a total of 5 swabs on days 0, 7, 14, 21, 28 relative to enrolment) after which visits were monthly.
@@
 A notable feature of this definition is that it explicitly does not correct for false positive or negative results.
-However, for many purposes, including this thesis, the difference between the positivity rate and prevalence is negligible.
+However, for many purposes, including this thesis, the difference between the positivity rate and prevalence is neglected.
 
@@
 However, in practice, two issues arise with this.
-First, the non-response bias discussed previously.
-Second, the number of positives on any day is low, especially among any specific subpopulation (\eg age group or region).
+First, the non-response (discussed above) causes bias.
+Second, the number of positives on any day is low, especially among any specific subpopulation (\eg age group or region) increases variance.
 Statistical methods are used to correct for these issues.
 
-The preferred statistical method for estimating the positivity rate is MRP (multi-level regression and poststratification)~\autocite{cisMethodsONS,pouwelsCommunity}.
-MRP is a statistical methodology developed to correct political opinion polling suffering from response bias~\autocite{gelmanMRT,gelmanPoststratication,parkMRP}.
+The preferred statistical method for estimating the positivity rate is MRP (multilevel regression and poststratification)~\autocite{cisMethodsONS,pouwelsCommunity}.
+MRP is a statistical methodology developed to correct political opinion polling suffering from non-response bias~\autocite{gelmanMRT,gelmanPoststratication,parkMRP}.
 MRP is a two-step process.
@@
 To address the low number of positives, MRP uses a hierarchical Bayesian model to borrow information across time and other covariates.
-The second step, poststratification, uses the posterior distribution of the parameters predict the population prevalence in each stratum.
+The second step, poststratification, uses the posterior distribution of the parameters to predict the population prevalence in each stratum.
 These are combined by taking a weighted average across the strata, where a stratum's weight is equal to its (known) population size.
@@
 
-I modified the regression component of a pre-existing version of a MRP model for CIS~\autocite{pouwelsMRPvaccination}, as recommended by that model's author for the specific context of this thesis~\citePersonalComms{Koen Pouwels}.
+I modified the regression component of a pre-existing version of a MRP model for the CIS~\autocite{pouwelsMRPvaccination}, as recommended by that model's author for the specific context of this thesis~\citePersonalComms{Koen Pouwels}.
 As in prior work~\autocite{pouwelsMRPvaccination}, I used INLA for inference, implemented in R-INLA (see \cref{E-transmission:sec:INLA}).
@@
 \begin{itemize}
-    \item $n_{taes}$ and $y_{taes}$ are the number of individuals tested and the number of positive test results on day $t$ in individuals of age group $a$, ethnicity $e$, and sex $s$.
+    \item $n_{taes}$ and $y_{taes}$ are the number of individuals tested and the number of positive test results respectively on day $t$ in individuals of age group $a$, ethnicity $e$, and sex $s$.
 I use a beta-binomial distribution for the likelihood, as discussed in \cref{biology-data:sec:clustering}.
@@
 Both show similar trends.
-The MRP estimate most often increases the estimated positivity rate, compared to uncorrected positivity rate, because the underrepresented demographic groups tended to have higher positivity rates.
+The MRP estimate is normally greater than the uncorrected positivity rate because the underrepresented demographic groups tended to have higher positivity rates.
 
@@
   \item $n \geq 4$.
-  \item A complex heuristic based on the likely variant of the positive tests (see \textcite{weiRisk} for details).
+  \item A complex heuristic based on the most probable variant of the positive tests (see \textcite{weiRisk} for details).
 This heuristic rarely applied in the pre-Omicron era~\citePersonalComms{Sarah Walker}.
@@
 
-Therefore, computational approaches requiring high-performance computing, graphical processing units, or long runtimes are not feasible within the SRS.
+Computational approaches requiring high-performance computing, graphical processing units, or long runtimes are not feasible within the SRS.
 This computational limitation will feature in several places in this thesis.

### 2024-03-25 incidence-prevalence.tex
Proof ch 3

@@
 I then move on to explaining the three processes that are involved: the \emph{infection process}, the \emph{prevalence process}, and the \emph{observation process}.
-The mathematical description of are an infection model, disease model, and observation model respectively.
+These are the mathematical description of an infection model, disease model, and observation model respectively.
 
@@
 Here, I adapt previous work to the SARS-CoV-2 context.
-In particular, recovering from infection needs to be considered, which was not the case for HIV/AIDS.
 
 The observation process (\cref{inc-prev:sec:observation-process}) describes the relationship between prevalence and the observed data, ultimately allowing inference of incidence based on this data.
-Unlike in HIV/AIDS backcalculation, the CIS has a sampling mechanism that needs to be modelled.
-This model is guided by the specific study design of the CIS.
+In particular, I include the sampling mechanism of a population prevalence survey, guided by the design of the CIS.
 
@@
 I choose discrete time for pragmatic reasons.
-Most epidemiological data is only available at daily granularity or, when more granular data exists, is often unreliable.
+Most epidemiological data are only available at daily granularity or, when more granular data exists, is often unreliable.
 Furthermore, within-day patterns produce variations in transmission (\eg due to sleep) and data (\eg due to logistical considerations around collection); these variations are not of interest in most contexts.
@@
 \emph{Incidence} on day $t$, $Z_t$, is the number of people in a given population who become infected on day $t$; when expressed as a proportion of the population it is the \emph{incidence proportion}~\autocite[89]{lashModern}, $Z_t/\Npop$, where $\Npop$ is the number of people in the population.
-The timing of infection events is very rarely directly observable and instead is inferred from other data.
-Despite this issue, it remains the single most important quantity for informing the response to an epidemic or pandemic.
 
@@
 Therefore, prevalence is constant over any single day $t$.
-I denote the number of prevalent individuals on day $t$ as $P_t$.
-Therefore, the prevalence is $P_t/\Npop$.
+I denote the number of prevalent individuals on day $t$ as $P_t$; hence, the prevalence is $P_t/\Npop$.
 
 \emph{Duration}, or \emph{time-to-event}, distributions are expressed in several equivalent ways~\autocite[17]{sunStatisticala}.
-The duration distribution I will be concerned with is the distribution of the number of days for which an individual is prevalent.
+The duration distribution I will be concerned with is the distribution of the duration of positivity.
 Consider an individual $i$ infected on day $B_i$ that is prevalent during the interval $[B_i, E_i]$ (as in \cref{E-biology-data:sec:cis-episodes}).
-I define this period as an \emph{infection episode}, beginning on day $B_i$ and ending on day $E_i$.
-$D_i = E_i - B_i + 1$ is their \emph{duration of positivity}, the number of days for which they are prevalent.
-I model the $D_i$s as a discrete independent and identically distributed (iid) random variable; extending the methodology to account for covariates is left to further work.
+I define this period as an \emph{infection episode}, beginning on day $B_i$ and ending on day $E_i$; therefore, with a duration $D_i = E_i - B_i + 1$.
+I model the $D_i$s as a discrete iid (independent and identically distributed) random variable; extending the methodology to account for covariates is left to further work.
 The duration is a positive integer, \ie $D_i \in \{1, 2, \dots, \dmax \}$, where $\dmax$ is the maximum duration.
-The first way of expressing the distribution of $D_i$ is as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
-The \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$.
+The distribution of $D_i$ can be expressed as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
+Alternatively, the \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$.
 %(a common alternative convention for survival analysis with discrete variables is to define the survival function as $\prob(D >d)$).
@@
 Similarly, $\lambda(d) \in [0,1]$, but can otherwise vary arbitrarily (\ie there is no monotonic or sum constraint).
-The lack of monotonicity or sum constraint on the hazard makes it an attractive parameterization for inference.
+The lack of monotonicity or sum constraint on the hazard makes it an attractive parameterization for inference~\autocite{heBayesian}.
 
@@
 Both overdispersion and feedback loops violate the properties of a Poisson process.
-Despite these issues, consistent inference on the incidence is still produced using the Poisson process model, although the uncertainty in the estimates may be understated~\autocite{beckerDependent}.
-
-% In \emph{deterministic backcalculation}, the observed prevalence, $x_t/n_t$, is assumed to be equal to $P_t/\Npop$.
-% However, this is often a poor approximation.
-% \emph{Statistical backcalculation} retains the sampling distribution of $x_t$ (or an approximation of it, such as a Poisson).
-
-% Define the first day of an infection episode in individual $i$ as $B_i$, and assume that the probability of $i$ being infected multiple times within the period of interest is negligible.
-% The time between being infected and first being detectable is short (see\todo{ref relevant part}), and therefore I often assume that $B_i$ is the same as the time of infection.
-% I keep this assumption for the remainder of this chapter.
-
-% Backcalculation makes use of this relationship to estimate incidence from prevalence, assuming that the duration is known.
-
-% Deterministic backcalculation assumes that the variance in the population is negligible, and therefore that the prevalence is equal to the mean prevalence.
-% This is justified because the variance due to sampling is much larger than the variance due to the true prevalence.
+Despite these issues, near-unbiased inference on the incidence is still produced using the Poisson process model, although the uncertainty in the estimates may be understated~\autocite{beckerDependent}.
 
@@
 When the number of infectious individuals is large, the law of large numbers implies that the mean-field behaviour is a good approximation of the fully stochastic model~\autocite[20]{diekmannMathematical}.
-This is especially true when making noisy observations of the system, and that the stochastic differences between the stochastic and its mean are negligible compared to the observation noise.
-I make a formalisation of this argument to justify a mean-field approximation of the prevalence process in \cref{E-inc-prev:sec:observation-process}.
+The approximation works especially well if used to make infernce based on noisy observations of the system, when the observation noise dominates is much larger than the infection process noise; I make a formalisation of this argument to justify a mean-field approximation of the prevalence process in \cref{E-inc-prev:sec:observation-process}.
 
@@
 $\beta$ is the transmission rate, absorbing several terms as described below.
-$\gamma$ is the recovery rate, or the rate that infectious individuals that recover per day; hence, $1/\gamma$ is the mean infectious period~\autocite[367]{keelingModeling}.
+$\gamma$ is the recovery rate, the rate that infectious individuals that recover; hence, $1/\gamma$ is the mean infectious period~\autocite[367]{keelingModeling}.
 These parameters are assumed to be constant.
@@
     This is the phase of initial exponential growth of an epidemic, producing the characteristic early epidemic curves.
-    It is in this phase that the basic reproduction number provides a good approximation of the epidemic dynamics.
-    \item The basic reproduction number for the SIR model is $\R = \beta / \gamma$~\autocite[20]{keelingModeling}.
+    It is in this phase that $\R$ provides a good approximation of the epidemic dynamics.
+    \item $\R = \beta / \gamma$ in the SIR model~\autocite[20]{keelingModeling}.
     This is the mean number of infections caused by an infectious individual per day, $\beta$, multiplied by the mean number of days for which they are infectious, $1/\gamma$.
@@
     Second, imported infections are negligible. 
-    Third, births and deaths of individuals are negligible over the length of time considered; model extensions allow these factors in contexts where they are relevant~\autocites[26]{keelingModeling}[214]{kretzschmarMathematical}.
+    Third, births and deaths of individuals are negligible over the length of time considered.
+    Model extensions allow these factors in contexts where they are relevant~\autocites[26]{keelingModeling}[214]{kretzschmarMathematical}.
     \item There is no analytical solution to the system of ODEs.
@@
 \frac{di_n(t)}{dt} &= n\gamma i_{n-1}(t) - n \gamma i_n(t) \\
-\frac{dr(t)}{dt} &= n\gamma i_n(t)
+\frac{dr(t)}{dt} &= n\gamma i_n(t) \\
+\lambda(t) &= \beta \sum_{j=1}^n i_j(t)
 \end{align}
-where $n$ is the number of infectious compartments, and redefining $\lambda(t) = \beta \sum_{j=1}^n i_j(t)$.
+where $n$ is the number of infectious compartments.
 The additional compartments added do not, in general, correspond to any biological feature or have a biological interpretation; they are purely a mathematical device to modify the distribution of time between becoming and ceasing to be infectious~\autocite{lloydRealistic}.
@@
 
-The reproduction number in terms of the parameters is unchanged by this extension, \ie $\R = \beta / \gamma$.
+The reproduction numbers in terms of the parameters are unchanged by this extension.
 
@@
 
-Similar to the infectious period, the latent period can be modelled as a gamma distribution using multiple latent compartments, $e_1, dots, e_m$.
+Similar to the infectious period, the latent period can be modelled as a gamma distribution using multiple latent compartments, $e_1, \dots, e_m$.
 \begin{align}
-\frac{ds}{dt} &= -\beta si \\
-\frac{de_1}{dt} &= \beta si - m\sigma e_1 \\
-\frac{de_2}{dt} &= m\sigma e_1 - m \sigma e_2 \\
+\frac{ds(t)}{dt} &= -\lambda s(t) \\
+\frac{de_1(t)}{dt} &= \lambda s(t) - m\sigma e_1(t) \\
+\frac{de_2(t)}{dt} &= m\sigma e_1(t) - m \sigma e_2(t) \\
 &\vdots \nonumber \\
-\frac{de_m}{dt} &= m\sigma e_{m-1} - m \sigma e_m \\
-\frac{di_1}{dt} &= m\sigma e_m - n\gamma i_1 \\
-\frac{di_2}{dt} &= n\gamma i_1 - n \gamma i_2 \\
+\frac{de_m(t)}{dt} &= m\sigma e_{m-1}(t) - m \sigma e_m(t) \\
+\frac{di_1(t)}{dt} &= m\sigma e_m(t) - n\gamma i_1(t) \\
+\frac{di_2(t)}{dt} &= n\gamma i_1(t) - n \gamma i_2(t) \\
 &\vdots \nonumber \\
-\frac{di_n}{dt} &= n\gamma i_{n-1} - n \gamma i_n \\
-\frac{dr}{dt} &= n\gamma i_n
+\frac{di_n(t)}{dt} &= n\gamma i_{n-1}(t) - n \gamma i_n(t) \\
+\frac{dr(t)}{dt} &= n\gamma i_n(t) \\
+\lambda(t) &= \beta \sum_{j=1}^n i_j(t)
 \end{align}
@@
 The different patterns could be due to behavioural factors (\eg high numbers of sexual partners in a sexually transmitted disease~\autocite[69]{keelingModeling} or younger age groups making many more contacts for respiratory diseases~\autocite[176]{andersonInfectious}), or biological factors such as genetics~\autocite[208]{andersonInfectious}.
-I denote the set of strata, as $\set{A}$, with the set's size as $A$.
+I denote the set of strata as $\set{A}$, with the set's size as $A$.
 I will assume that the disease progression (\ie length of the exposed and infectious periods) is the same for all strata, although this assumption can be relaxed.
@@
 The interactions between strata are included by replacing the scalar $\beta$ with a WAIFW (\emph{Who Acquires Infection From Whom}) matrix, $\matr{\beta}$~\autocite[58]{keelingModeling}.
-The force of infection is now a vector, defined as $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i_j}(t)$.
+The force of infection is now a vector, defined as $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i}_j(t)$ where $\vec{i}_j$ is a vector such that $i_{j,a}$ is the proportion of stratum $a$ in the $j$th infectious state.
 Their interpretation can be arrived at by considering a single infectious individual in strata $a'$, analogous to the single stratum case.
@@
 This $\R$ keeps the same interpretation as in the homogeneous case, \ie the average number of secondary infections caused by a single infection in a fully susceptible population, where the average is taken over the distribution implied by the dominant eigenvector.
-The important threshold that exponential growth occurs iff $\R > 1$ remains.
+The important threshold that exponential growth occurs if and only if $\R > 1$ remains.
 
@@
 For any infection beginning at $t$, the probability that it ends at $t'$ is $\prob(E_i = t' \mid B_i = t) = \prob(D_i = t' - t + 1 \mid B_i = t) = f_D(t' - t + 1)$.
-Since I assume these durations are independent, the number of episodes of each duration is multinomially distributed~\autocite{paganoHIV}.
-Specifically:
+As I assume these durations are independent~\autocite{paganoHIV}:
 \begin{align}
@@
 In \cref{inc-prev:sec:observation-process} I will show that only $\E(P_t \mid \vec{Z})$ and bounds on $\V(P_t \mid \vec{Z})$ will be relevant for inferring incidence.
-I know derive these from their constituent parts.
+I now derive these from their constituent parts.
 
 Start by considering the recoveries.
-The sum of recoveries of infections that occurred on the same day $t$ is the sum of multinomial cell probabilities, and hence binomially distributed~\autocite{alamAnalysis}.
-Specifically:
+From the properties of a multinomial distribution~\autocite{alamAnalysis}:
 \begin{align}
@@
 
-The expectation in \cref{inc-prev:eq:EPt} is a discrete convolution equation in the same form as \cref{intro:eq:disease-model} with $\zeta = 1$.
+The expectation in \cref{inc-prev:eq:EPt} is a discrete convolution equation in the same form as \cref{E-intro:eq:disease-model} with $\zeta = 1$.
 % The equation is of the same form as the equation used for HIV/AIDS backcalculation, as well as being found in many other fields~\autocite[and references therein]{brookmeyerBackcalculation}.
@@
 On day $t$ (for $t = 1, \dots, T$), a prevalence survey such as the CIS samples $n_t$ individuals at random from the population of interest and tests if they are prevalent.
-Therefore, I model the number of positive tests, $X_t$, as $X_t \mid P_t, n_t, \Npop \dist \Bin(n_t, P_t/\Npop)$.
+Therefore, I model the number of positive tests, $Y_t$, as $Y_t \mid P_t, n_t, \Npop \dist \Bin(n_t, P_t/\Npop)$.
 The conditioning on $n_t$ and $\Npop$ is implicit in what follows.
@@
 
-The $X_t$s are conditionally independent, given the population prevalences at each time, $\vec{P} = (P_1, \dots, P_t)^T$. 
+The $Y_t$s are conditionally independent, given the population prevalences at each time, $\vec{P} = (P_1, \dots, P_t)^T$. 
 %For now, I assume that the sample is chosen uniformly at random from the population of interest.
@@
 I will use prevalence surveys to infer incidence.
-I base this inference on the likelihood, $p(\vec{x} \mid \vec{Z})$.% = \int p(\vec{x} \mid \vec{P}) p(\vec{P} \mid \vec{Z}) d\vec{P} = \int \prod_t p(x_t \mid P_t) p(\vec{P} \mid \vec{Z}) d\vec{P}$, where $\vec{x} = (x_1, x_2, \dots)^T$.
-Here $\vec{x} = (x_1, \dots, x_T)$ is the vector of observed number of positive tests in the prevalence survey on each day and $\vec{Z} = (Z_{-\dmax}, Z_{-\dmax+1} \dots, Z_T)$ is the vector of incidence on each day.
+I base this inference on the likelihood, $p(\vec{y} \mid \vec{Z})$.
+Here $\vec{y} = (y_1, \dots, y_T)$ is the vector of observed number of positive tests in the prevalence survey on each day and $\vec{Z} = (Z_{-\dmax}, Z_{-\dmax+1} \dots, Z_T)$ is the vector of incidence on each day.
 The incidence back to day $-\dmax$ will affect $P_1$ and therefore need to be included in $\vec{Z}$.
 
-This section argues that $p(\vec{x} \mid \vec{Z}) \approx \prob(\vec{X} = \vec{x} \mid \vec{P} = \E(\vec{P} \mid \vec{Z}))$.
-As the $x_t$s are independent conditional on $\vec{P}$, it follows that $p(\vec{x} \mid \vec{Z}) \approx \prod_t \prob(X_t = x_t \mid P_t = \E(P_t \mid \vec{Z}))$.
-This relationship will make inference for $\vec{Z}$ much simpler because the latent vector $\vec{P}$ and the dependence of the $x_t$s can be ignored.
+This section argues that $p(\vec{y} \mid \vec{Z}) \approx \prob(\vec{X} = \vec{y} \mid \vec{P} = \E(\vec{P} \mid \vec{Z}))$.
+As the $y_t$s are independent conditional on $\vec{P}$, it follows that $p(\vec{y} \mid \vec{Z}) \approx \prod_t \prob(Y_t = y_t \mid P_t = \E(P_t \mid \vec{Z}))$.
+This relationship will make inference for $\vec{Z}$ much simpler because the latent vector $\vec{P}$ and the dependence of the $y_t$s can be ignored.
 
-The basis of the argument is that the first two moments of $\vec{x} \mid \vec{Z}$ and $\vec{X} \mid \vec{P} = \E(\vec{P} \mid \vec{Z})$ have negligible difference.
+The basis of the argument is that the first two moments of $\vec{y} \mid \vec{Z}$ and $\vec{X} \mid \vec{P} = \E(\vec{P} \mid \vec{Z})$ have negligible difference.
 The first moment follows directly from the tower property of expectations.
@@
 I start with the diagonal elements, the variances.
-I show that $\V(X_t \mid \vec{Z}) \approx \V(X_t \mid \vec{P} = \E(\vec{P} = \vec{Z}))$.
-The conditional independence of the $x_t$s implies that the conditional covariances are 0.
-Therefore, I show that the correlation between $X_t$ and $X_{t'}$ conditional only on $\vec{Z}$ is negligible.
+I show that $\V(Y_t \mid \vec{Z}) \approx \V(Y_t \mid \vec{P} = \E(\vec{P} = \vec{Z}))$.
+Then I show that the correlation between $Y_t$ and $Y_{t'}$ conditional only on $\vec{Z}$ is negligible.
 
@@
 \begin{align}
-  &\var\left( X_t \mid \vec{Z} \right) \\
-    &= \var\left[\E\left( X_t \mid \vec{Z}, \vec{P} \right) \mid \vec{Z} \right] + \E\left[\var\left( X_t \mid \vec{Z}, \vec{P} \right) \mid \vec{Z} \right] \\
-    &= \var\left( n_t \frac{P_t}{\Npop} \mid \vec{Z} \right) + \E\left( n_t \frac{P_t}{\Npop} \left(1 - \frac{P_t}{\Npop} \right) \mid \vec{Z} \right) \\
+  &\var\left( Y_t \mid \vec{Z} \right) \\
+    &= \var\left[\E\left( Y_t \mid \vec{Z}, \vec{P} \right) \mid \vec{Z} \right] + \E\left[\var\left( Y_t \mid \vec{Z}, \vec{P} \right) \mid \vec{Z} \right] \\
+    &= \var\left( n_t \frac{P_t}{\Npop} \mid \vec{Z} \right) + \E\left( n_t \frac{P_t}{\Npop} \left(1 - \frac{P_t}{\Npop} \right) \mid \vec{Z} \right) &Y_t \mid \vec{P} \dist \Bin \\
     &= \left( \frac{n_t}{\Npop} \right)^2 \var\left(P_t \mid \vec{Z} \right) + \frac{n_t}{\Npop} \E\left(P_t \mid \vec{Z} \right)  - \frac{n_t}{\Npop^2} \E\left(P_t^2 \mid \vec{Z} \right) \\
@@
     &= \frac{n_t(n_t - 1)}{\Npop^2} \var\left(P_t \mid \vec{Z} \right) + n_t \frac{\E\left(P_t \mid \vec{Z} \right)}{\Npop}\left(1 - \frac{\E\left(P_t \mid \vec{Z} \right)}{\Npop} \right) \\
-    &= \frac{n_t(n_t - 1)}{\Npop^2} \var\left(P_t \mid \vec{Z} \right) + \V(X_t \mid P_t = \E(P_t \mid \vec{Z})).
 \end{align}
-The first term of this expression is negligible in the contexts I consider.
-These contexts have $n_t << \Npop$.
-$\Npop$ is the population of England (around 56 million) while $n_t$ is the daily sample size (10s of thousands).
-Further $\V(P_t \mid \vec{Z}) \leq \E(P_t \mid \vec{Z})$ (\cref{inc-prev:eq:boundVPt}).
-Therefore, $\var\left( X_t \mid \vec{Z} \right) \approx \V(X_t \mid P_t = \E(P_t \mid \vec{Z}))$.
+In the contexts I consider $n_t << \Npop$ because $\Npop$ is the population of England (around 56 million) while $n_t$ is the daily sample size (10s of thousands).
+Therefore:
+\begin{align}
+    \frac{n_t(n_t - 1)}{\Npop^2} \var\left(P_t \mid \vec{Z} \right)
+    &<< \frac{n_t}{\Npop} \var\left(P_t \mid \vec{Z} \right) \\
+    &\leq \frac{n_t}{\Npop} \E\left(P_t \mid \vec{Z} \right) &\text{by \cref{inc-prev:eq:boundVPt}} \\ 
+    &\leq \frac{n_t}{\Npop} \E\left(P_t \mid \vec{Z} \right) \left(1 - \frac{\E(P_t \mid \vec{Z}}{\Npop} \right) &\text{as $P_t \leq \Npop$.}
+\end{align}
+Combining these results:
+\begin{align}
+  \var\left( Y_t \mid \vec{Z} \right)
+    &\approx \frac{\E\left(P_t \mid \vec{Z} \right)}{\Npop}\left(1 - \frac{\E\left(P_t \mid \vec{Z} \right)}{\Npop} \right) \\
+    &= \V(Y_t \mid P_t = \E(P_t \mid \vec{Z}))
+\end{align}
+which is the required result.
 
-A similar argument will show that the correlation between $x_t$ and $x_{t'}$ ($t \neq t'$) is negligible.
+A similar argument will show that the correlation between $y_t$ and $y_{t'}$ ($t \neq t'$) is negligible.
 The basic structure of this argument is as the previous one.
@@
 \begin{align}
-  &\lvert \cor(X_t, X_{t'}) \rvert \\
-  &= \frac{\lvert\cov(X_t, X_{t'})\rvert}{\sqrt{\var(X_t) \var(X_{t'})}} &\text{by definition}\\
-  &= \frac{\lvert \E(\cov(X_t, X_{t'} \mid \vec{P})) + \cov(\E(X_t \mid \vec{P}), \E(X_{t'} \mid \vec{P})) \rvert}{\sqrt{\var(X_t) \var(X_{t'})}} &\text{by law of total covariance} \\
-  &= \frac{\lvert 0 + \cov(n_t P_t / \Npop, n_{t'} P_{t'} / \Npop) \rvert}{\sqrt{\var(X_t) \var(X_{t'})}} \\
-  &= \frac{n_t n_{t'} \lvert \cov(P_t, P_{t'}) \rvert}{\Npop^2 \sqrt{\var(X_t) \var(X_{t'})}}  \\
-  &<< \frac{n_t n_{t'} \lvert \cov(P_t, P_{t'}) \rvert}{\Npop^2 \sqrt{\var(P_t) \var(P_{t'})}} &\text{as $\V(X_t) >> \V(P_t)$} \\
+  &\lvert \cor(Y_t, Y_{t'}) \rvert \\
+  &= \frac{\lvert\cov(Y_t, Y_{t'})\rvert}{\sqrt{\var(Y_t) \var(Y_{t'})}} &\text{by definition}\\
+  &= \frac{\lvert \E(\cov(Y_t, Y_{t'} \mid \vec{P})) + \cov(\E(Y_t \mid \vec{P}), \E(Y_{t'} \mid \vec{P})) \rvert}{\sqrt{\var(Y_t) \var(Y_{t'})}} &\text{by law of total covariance} \\
+  &= \frac{\lvert 0 + \cov(n_t P_t / \Npop, n_{t'} P_{t'} / \Npop) \rvert}{\sqrt{\var(Y_t) \var(Y_{t'})}} \\
+  &= \frac{n_t n_{t'} \lvert \cov(P_t, P_{t'}) \rvert}{\Npop^2 \sqrt{\var(Y_t) \var(Y_{t'})}}  \\
+  &<< \frac{n_t n_{t'} \lvert \cov(P_t, P_{t'}) \rvert}{\Npop^2 \sqrt{\var(P_t) \var(P_{t'})}} &\text{as $\V(Y_t) >> \V(P_t)$} \\
   &= \frac{n_t n_{t'}}{\Npop^2} \lvert \cor(P_t, P_{t'}) \rvert &\text{by def of $\cor$}\\
@@
 $n_t n_{t'} / \Npop^2$ is very small, for the same reason that $n_t / \Npop$ is small.
-The argument above shows that $\lvert \cor(X_t, X_{t'}) \rvert$ is much smaller than this.
-Therefore, neglecting the correlation between $X_t$ and $X_{t'}$ is a good approximation.
+$\lvert \cor(Y_t, Y_{t'}) \rvert$ is even smaller than this.
+Therefore, neglecting the correlation between $Y_t$ and $Y_{t'}$ is a good approximation.
 
-This section showed that, when using a prevalence survey such as CIS, the likelihood $p(\vec{x} \mid \vec{z})$ is very well approximated by $p(\vec{x} \mid \vec{P} = \E(\vec{P} \mid \vec{z}))$.
+This section showed that, when using a prevalence survey such as CIS, the likelihood $p(\vec{y} \mid \vec{z})$ is very well approximated by $p(\vec{y} \mid \vec{P} = \E(\vec{P} \mid \vec{z}))$.
 

### 2024-03-25 introduction.tex
Proof ch 3

@@
 
-COVID-19 (coronavirus disease 2019), the disease caused by the virus SARS-CoV-2 (severe acute respiratory syndrome coronavirus 2), is estimated to have killed approximately 15 million in the most acute phase (to the end of 2021) and 30 million in total~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
+COVID-19, the disease caused by the virus SARS-CoV-2, is estimated to have killed approximately 15 million in the most acute phase (to the end of 2021) and 30 million in total~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
+Beyond the lives lost, there were massive health, economic, and societal costs.
 
@@
 $\Re$ describes the current dynamics of the epidemic: $\Re > 1$ means the epidemic is growing, $\Re < 1$ means it is shrinking, and $\Re = 1$ means it is stable.
-In simple epidemic models (see \cref{E-SEIR:sec:SIR}), $\Re$ is proportional to the product of the rate of contacts, the riskiness of those contacts, and the proportion of the population that is susceptible.
-Therefore, it informs the strength of intervention, \eg social distancing or vaccination, required to control the epidemic.
+In simple epidemic models (see \cref{E-SEIR:sec:SIR}), $\Re$ is proportional to the product of the rate of contacts, the probability of transmission upon contact, and the proportion of the population that is susceptible.
+Interventions act by reducing one of these components \eg masks reduce the probability of transmission and vaccination reduce the susceptible population.
+Therefore, $\Re$ informs the strength of intervention required to control the epidemic.
 
-The \emph{control reproduction number}, $\Rc$, is the mean number of secondary infections generated by a single infection with the behaviour at time $t$ but in a fully susceptible population.
+The \emph{control reproduction number}, $\Rc$, is the mean number of secondary infections that would be generated by a single infection at time $t$ if in a fully susceptible population.
 $\Rc$ is also known as the \emph{reproduction number excluding immunity}.
@@
 For example, when comparing the English lockdown in March 2020 and January 2021, there are two effects in play: there are more individuals with immunity in January 2021, and the lockdowns had different measures in place.
-$\Rc$ focuses on the latter effect.
+$\Rc$ quantifies the latter effect.
 
@@
 Reproduction numbers are strongly connected to \emph{growth rates}.
-Over short periods of time, epidemics tend to grow (or shrink) exponentially, and hence have an exponential growth rate, normally denoted $r$.
+If changes in behaviour and susceptibility are negligible, as they often are over short periods of time, epidemics grow (or shrink) exponentially.
+The rate of this exponential growth is its \emph{growth rate}, normally denoted $r$.
 For a given disease and population, an analogous growth rate can be defined for each definition of the reproduction number~\autocite{pellisEstimation,paragGrowthRates,wallingaGI}.
@@
 Whether a given individual is \emph{prevalent} (has the disease) is often measurable through diagnostic testing.
-Therefore, the prevalence of the population can be estimated with a prevalence survey, although these are typically very expensive to power appropriately.
+Performing diagnostic tests in a sample of the population at a given point in time is known as a \emph{prevalence survey}.
+Prevalence surveys can estimate the population prevalence but are typically very expensive to power appropriately.
 Prevalence and incidence are strongly related, via the duration of an infection (see \cref{E-inc-prev}).
@@
 Therefore, they suffer from biases that can be hard to understand and quantify~\autocite{shadboltChallenges}.
+These biases need to be accounted for when analysing the data.
+
 Here I discuss three types of data which are useful for surveillance: incidence-type, prevalence-type, and data from cohort studies.
@@
 
-An example of incidence-type data are community testing.
-This relies on individuals presenting for a test, hence is biased by individuals' ability and willingness to be tested.
+An example of incidence-type data are community testing, reliant on individuals presenting for a test.
+However, individuals' ability and willingness to be tested varies over time, which needs to be accounted for in order to perform unbiased inference.
+Unfortunately, data to estimate these changes is normally unavailable.
 
@@
 
-Prevalence-type data originate from observations of the number or proportion of a population in a specific disease state at a specific time.
+Prevalence-type data originate from observations of the number or proportion of a population in a specific \emph{disease state} at a specific time.
+Commonly used disease states are the presence or value of a biomarker or symptoms.
 For example, the number of hospital beds occupied by individuals with the disease or the proportion of the population with a marker of prior infection.
@@
 Other epidemiological and/or biological quantities are normally required to estimate transmission.
-These depend on the model type, often including the duration of different disease stages or the distribution of the time between infection events~\autocite{wallingaGI,dankwaStructural}.
+These depend on the model type, often including the duration of different disease states or the distribution of the time between infection events~\autocite{wallingaGI,dankwaStructural}.
 Cohort studies (following a group of individuals over time), are a study design which can estimate many of these quantities.
@@
 
-At the core of many observation models is a \emph{convolution equation} or an approximation of it~\autocite[e.g.][]{brookmeyerBackcalculation,abbottEstimating,birrellRealtimea}.
+At the core of many observation models is a \emph{convolution equation} or an approximation of it~\autocite[e.g.][]{brookmeyerBackcalculation,abbottEstimating,birrellBayesian}.
 This equation describes the expectation of $W_t$, conditional on the incidence prior to time $t$.
@@
 Biases can arise in a variety of ways, \eg healthcare-seeking behaviour or tests which are less than 100\% accurate.
-Bias can be modelled through additional parameters or data quantifying these issues~\autocite[e.g.][]{sherrattExploring,birrellRealtimea,nicholsonImproving,swallow2022challenges}.
+Bias can be modelled through additional parameters or data quantifying these issues~\autocite[e.g.][]{sherrattExploring,birrellBayesian,nicholsonImproving,swallow2022challenges}.
 
 Noise could come from a variety of sources, \eg sampling error, clustering, or unexplained overdispersion.
-Noise is typically modelled by a parametric distribution \eg the negative binomial distribution allowing for overdispersion compared to a Poisson process assumption~\autocite[e.g.][]{birrellRealtimea,frassoBayesian}.
+Noise is typically modelled by a parametric distribution \eg the negative binomial distribution allowing for overdispersion compared to a Poisson process assumption~\autocite[e.g.][]{birrellBayesian,frassoBayesian}.
 
@@
 
-The main data sources informing the response to the SARS-Cov-2 epidemic in England were: community testing (known as pillar 2 testing), the number of severe events, seroprevalence surveys, and prevalence surveys.
+The main data sources informing the response to the SARS-Cov-2 epidemic in England were based on: community testing (known as pillar 2 testing), the number of severe events, seroprevalence surveys, and prevalence surveys.
 Each of these data sources have strengths and weaknesses, see \textcite{royalSocietyRnumber} and references therein for further details on each of these data sources.
@@
 Therefore, these data are subject to fewer biases than community testing.
-However, SARS-CoV-2 rarely leads to severe events in younger individuals, meaning severe event counts are insensitive to changes in transmission only occurring in younger individuals.
+However, SARS-CoV-2 rarely leads to severe events in younger individuals, meaning severe event counts are insensitive to changes in transmission only affecting younger individuals.
 
@@
 
-Seroprevalence surveys were conducted throughout the epidemic, primarily in blood donors.
+Seroprevalence surveys were conducted throughout the epidemic, the largest of which was tested donated blood~\autocite{amirthalingamSeroprevalence}.
 In addition to the normal problem of the delay between infection and antibody response, blood donors are a convenience sample whose representativeness of the wider population is unclear.
@@
 
-A major advantage of the CIS over REACT is that the CIS has daily testing of individuals, while REACT has no test results between rounds.
-This means that REACT cannot be used to estimate transmission or incidence between rounds, instead relying on model assumptions at these times.
+A major advantage of the CIS over REACT is that the CIS had continual data collection, while REACT has no test results between rounds.
+The lack of data between rounds meant that using REACT data to infer transmission in these periods requires strong model assumptions.
 A particular instance of this is that a peak in late December 2020 occurred between rounds of the REACT survey, complicating analysis of the survey data at that time~\autocite{rileyREACTround8}.
@@
 
-In the UK, SPI-M-O (the Scientific Pandemic Influenza Group on Modelling, Operational sub-group) produced the most prominent modelling.
+Members of SPI-M-O (the Scientific Pandemic Influenza Group on Modelling, Operational sub-group) produced the most prominent modelling of the UK epidemic.
 SPI-M-O was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), generating evidence based on infectious disease modelling for policy development~\autocite{medleySPIM,govSPIMO}.
@@
 
-Each week, the members of SPI-M-O would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
-The committee reviewed these estimates, removing any estimates unduly affected by issues (e.g..
-data quality issues), and then formed a consensus value using a statistical combination of the estimates~\autocite{parkCombining}.
+Each week, members of SPI-M-O would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
+The committee peer-reviewed these estimates to provide quality assurance, removing any estimates unduly affected by issues (\eg data quality issues).
+They then formed a consensus value using a statistical combination of the estimates~\autocite{parkCombining}.
 The consensus value was published on the UK Government's website and informed its response to the epidemic~\autocite{govRnumber}.
@@
 
-Some of these models were used to explore policy scenarios.
+Additionally, these models were used to explore policy scenarios.
 For example, the decision at each of stage of lifting the lockdown in 2021 was informed by modelling studies~\autocite{sageEvidence}.
@@
 
-In this thesis, I consider how prevalence surveys could be used for epidemic surveillance.
+In this thesis, I consider how prevalence surveys could be used for epidemic surveillance, in particular estimates of incidence and transmission.
 These surveys have well-understood designs which means that estimates derived from them are more robust.
@@
 
-There is a large amount of related work in this area, from both prior to the pandemic and during its course.
+In the following subsection, I introduce the most relevant related work, from both prior to the pandemic and during its course.
 
@@
 \emph{Backcalculation} is a modelling framework developed in the context of HIV/AIDS~\autocite{brookmeyerMethod} but applied very broadly.
-It specifies a particular combination of infection, disease, and observation model to relate times of HIV infection with times of AIDS diagnosis via the incubation period.
+It specifies a particular combination of infection, disease, and observation model which relate times of HIV infection with times of AIDS diagnosis via the incubation period.
 Further details are in the recent review \textcite{sunModeling}.
@@
 
-The original formulation used a Poisson process as the infection model; a disease model where all infections are eventually observed ($\zeta = 1$) and the delay function, $f$, is the incubation period distribution's pdf; and an observation model where all diagnoses are observed, and hence has no influence on the data.
+The original formulations~\autocite{brookmeyerMethod,rosenbergBackcalculation} used a Poisson process as the infection model; a disease model where all infections are eventually observed ($\zeta = 1$) and the delay function, $f$, is the incubation period distribution's pdf; and an observation model where all diagnoses are observed, and hence has no influence on the data.
 
-The disease and observation models must be adapted to apply backcalculation to SARS-CoV-2 prevalence surveys (elaborated on in \cref{E-inc-prev}).
+The disease and observation models must be adapted to apply backcalculation to SARS-CoV-2 prevalence surveys (see \cref{E-inc-prev}).
 The disease model must consider that this is now prevalence-type, rather than incidence-type data.
 This difference induces a dependency structure between the observations on each day.
-The observation model must consider that this is now a survey, with sampling error.
+The observation model must now include the with error arising from using a sample of the population rather than (assumed) full recording of AIDS diagnoses.
 
@@
 This measure is highly related to the duration of positivity.
-The study's cohort consisted of healthcare workers, \ie working-age adults.
+However, the study's cohort consisted only of healthcare workers, \ie working-age adults.
 

### 2024-03-25 transmission.tex
plural data
proof
Proof ch 3

@@
 In this chapter, I use the estimate of the duration of RT-PCR positivity from combining information from the ATACCC and CIS studies (see \cref{E-imperf-test:sec:conclusion}) to investigate the feasibility of estimating incidence and transmission of SARS-CoV-2 from only the CIS data (described in \cref{E-intro:sec:cis}).
-I will show that the CIS data is sufficient, demonstrating the value of surveys of this type for pandemic surveillance.
+I will show that the CIS data are sufficient, demonstrating the value of surveys of this type for pandemic surveillance.
 
@@
 % I solve this system using forward substitution, calculating the value of $Z_t$ in \cref{inc-prev:eq:forward-substitute} sequentially for $t = 1, \dots, T$.
-% $t = 1$ is defined as the first day where prevalence data is used.
+% $t = 1$ is defined as the first day where prevalence data are used.
 % \Cref{inc-prev:eq:forward-substitute} is repeated here for convenience:
@@
 Clearly, this is not realistic as travel between regions of England is common and unrestricted.
-However, except in an epidemic's early stages, the interaction between large regions has a negligible impact on the epidemic's progression~\autocite[e.g.][]{birrellRealtimea,gogSpatial,eggoSpatial}.
+However, except in an epidemic's early stages, the interaction between large regions has a negligible impact on the epidemic's progression~\autocite[e.g.][]{birrellBayesian,gogSpatial,eggoSpatial}.
 
@@
 % Ideally, all the $\beta_{ij}$ terms would be estimated, but there is normally inadequate data to do so, requiring assumptions to be made to reduce the number of parameters~\autocite[65]{keelingModeling}.
-% % $\beta_{ij}$ is then estimated, although often with a parsimonious model
+% % $\beta_{ij}$ is then estimated, although often with a parsimonious model.
 % A common assumption is that $\beta_{ij}$ can be decomposed into two independent components such that $\beta_{ij} = \beta \beta^\text{susc}_{i} \beta^\text{inf}_j$.
@@
 I solve the system of ODEs using Euler's method, discretised at timesteps of $\delta t = 1/2$ days.
-That is, if $\vec{x}_t$ is the system state at time $t$ for $t \in \{ 0, 0.5, 1, \dots \}$ then $x_{t+\frac{1}{2}} = x_t + (d\vec{x}(t)/dt) \delta t$.
+That is, if $\vec{x}_t$ is the system state at time $t$ for $t \in \{ 0, 0.5, 1, \dots \}$ then $y_{t+\frac{1}{2}} = y_t + (d\vec{x}(t)/dt) \delta t$.
 

### 2024-03-26 cis-perf-new-ll.tex
Overleaf: mostly ch 5 corrections

\documentclass[thesis.tex]{subfiles}

\begin{document}

\chapter{Alternative derivation of \cref{E-perf-test:eq:multinomial}} \label{perf-test:sec:alt-likelihood}

Denote the characteristics of episode $j$ as $\nu_j = (B_j, E_j, i(j))$.
I now define a series of sets $\set{E}_1, \dots, \set{E}_{N_E}, \set{E}_u$ for an integer $N_E$ such that any possible $\nu_j$ is in exactly one of the classes and each set corresponds to the possible infection episdoes that lead to identical observations (conditional on the test schedules).

If $\nu_j$ would be undetected then $\nu_j \in \set{E}_u$.
Hence, $\prob(\nu_j \in \set{E}_u \mid \vec{\theta}) = p_u$.

Otherwise, the sets are defined such that $\nu_j$ and $\nu_{j'}$ are in the same class if and only if $i(j) = i(j')$ and the episodes beginning and end times are censored to the same interval.
If $i \in \set{D}$ then there is exactly one $k$ such that $\set{E}_k$ is identical to $i$'s admissible region $\alpha_i$.
% In this case, I write $\alpha_{i(k)}$ to denote this admissible region.

More formally, the sets can be defined in terms of an equivalence relation $\equiv$.
Define $\equiv$ such that $\nu_j \equiv \nu_{j'}$, where $j$ and $j'$ are arbitrary possible episodes (\ie these are episodes that, a priori, could have occurred but may or may not have done), if and only if either both $j$ and $j'$ would be undetected (as defined in \cref{E-perf-test:sec:prob-undetected}) or the following four conditions all hold.
\begin{enumerate}
    \item Both $\nu_j$ and $\nu_{j'}$ would be detected.
    \item $i(j) = i(j')$.
    \item $\tau_{\sched_{i(j)}}(B_j) = \tau_{\sched_{i(j')}}(B_{j'})$.
    \item $\tau_{\sched_{i(j)}}(E_j) = \tau_{\sched_{i(j')}}(E_{j'})$.
\end{enumerate}
Then each $\set{E}_k$ is an equivalence class under this relation and the set of all possible $\nu_j$.

Let $\xi_k$ denote $\prob(\nu_j \in \set{E}_k)$.
Assume, for tractability, that the events $\nu_j \in \set{E}_k$ and $\nu_{j'} \in \set{E}_{k'}$ are independent for $j \neq j'$.
Denote by $m_k$ then number of episodes in $\set{E}_k$.
$\vec{m} = [m_1, \dots m_{N_E}, m_u]^T$ are the observations.
Then:
\begin{align}
  \vec{m}
  \mid \ntot, \vec{\theta}
  \dist
  \MN\left(
    \ntot,
    \frac{1}{\Ncis}
    \begin{bmatrix}
        \xi_1 \\ \xi_2 \\ \vdots \\ \xi_{N_E} \\ \xi_u
    \end{bmatrix}
  \right).
  \label{perf-test:eq:alt-likelihood}
\end{align}

If the set $\set{E}_k$ aligns with some $\alpha_i$ then $m_k = 1$ and $\xi_k = p_{ia}$.
Otherwise, for $k \neq u$, $m_k = 0$.
Finally, $m_u = n_u$ and $\xi_u = p_u$.
Therefore, the probability masses of \cref{perf-test:eq:alt-likelihood} and \cref{E-perf-test:eq:multinomial} match.

\end{document}

### 2024-03-26 introduction_new.tex
Overleaf: mostly ch 5 corrections

--- 2024-03-25 introduction.tex
@@
 
-At the core of many observation models is a \emph{convolution equation} or an approximation of it~\autocite[e.g.][]{brookmeyerBackcalculation,abbottEstimating,birrellBayesian}.
+At the core of many observation models is a \emph{convolution equation} or an approximation of it~\autocite[e.g.][]{brookmeyerBackcalculation,abbottEstimating,birrellRealtimea}.
 This equation describes the expectation of $W_t$, conditional on the incidence prior to time $t$.
@@
 Biases can arise in a variety of ways, \eg healthcare-seeking behaviour or tests which are less than 100\% accurate.
-Bias can be modelled through additional parameters or data quantifying these issues~\autocite[e.g.][]{sherrattExploring,birrellBayesian,nicholsonImproving,swallow2022challenges}.
+Bias can be modelled through additional parameters or data quantifying these issues~\autocite[e.g.][]{sherrattExploring,birrellRealtimea,nicholsonImproving,swallow2022challenges}.
 
 Noise could come from a variety of sources, \eg sampling error, clustering, or unexplained overdispersion.
-Noise is typically modelled by a parametric distribution \eg the negative binomial distribution allowing for overdispersion compared to a Poisson process assumption~\autocite[e.g.][]{birrellBayesian,frassoBayesian}.
+Noise is typically modelled by a parametric distribution \eg the negative binomial distribution allowing for overdispersion compared to a Poisson process assumption~\autocite[e.g.][]{birrellRealtimea,frassoBayesian}.
 

### 2024-03-26 biology-data.tex
Clarify size of immune response

@@
 Viral replication requires the virus to infect a host cell and then use the host cell's machinery to produce more virus; as more host cells are infected this process slows because there are fewer remaining potential host cells, known as \emph{target cells}, to infect.
+The host's immune system can modify this process by various mechanisms; the most relevant for SARS-CoV-2 is by making some target cells temporarily immune from infection.
+The immune system response contribution to the reduction in target cells is several orders of magnitude larger than the reduction in target cells due to viral replication.
 The rate at which host cells are infected slows over time because there are fewer remaining target cells to infect.
-The host's immune system can slow this process by various mechanisms; the most relevant for SARS-CoV-2 is by making some target cells less susceptible to infection.
 A host's immune response is triggered by the viral load and becomes stronger as the viral load increases.

### 2024-03-26 ATACCC.tex
plural data
proof ch 4

@@
 The aim of this chapter is to estimate the duration of RT-PCR positivity using the viral load data from the ATACCC study (introduced in \cref{E-biology-data:sec:ataccc}).
-I model a biomarker, the log viral load, over the course of an infection.
+I model this biomarker over the course of an infection.
 From this model, I derive the distribution of the duration of RT-PCR positivity.
@@
 First, I will explain how biomarkers are typically modelled, both for SARS-CoV-2 and historically in the biostatistical literature (\cref{ATACCC:sec:background}).
-Next, I describe the model of \textcite{hakkiOnset} in detail, which applies these methods to the ATACCC data (\cref{ATACCC:sec:hakki}) and before modifying its observation model to improve the robustness of the inference (\cref{ATACCC:sec:observation-modification}).
+Next, I describe the model of \textcite{hakkiOnset} in detail, which applies these methods to the ATACCC data (\cref{ATACCC:sec:hakki}), before modifying its observation model to improve the robustness of the inference (\cref{ATACCC:sec:observation-modification}).
 This modification is based on an earlier analysis using a preliminary version of the dataset, described in \cref{ATACCC:sec:original-analysis}.
-Finally, I convert the model and parameter estimates of the biomarker's trajectory to a duration of RT-PCR positivity (\cref{ATACCC:sec:duration}).
+I convert the model and parameter estimates of the biomarker's trajectory to a duration of RT-PCR positivity (\cref{ATACCC:sec:duration}).
+Finally, I discuss the results (\cref{ATACCC:sec:discussion}).
 I will refine this estimate in the following chapters before its use in estimating transmission in \cref{E-transmission}.
@@
 
-Two major approaches have been taken to model SARS-CoV-2 viral load trajectories, both based on well-well-established methods within biostatistics.
+Two major approaches have been taken to model SARS-CoV-2 viral load trajectories, both based on well-established methods within biostatistics.
 The first approach is an ODE-based approach that uses parameters representing biological properties of the virus and host~\autocite[e.g.][]{ejimaEstimation,keVivo,kimQuantitative,goncalvesTiming,perelsonMechanistic}.
@@
 
-Piecewise linear, random effect models have a long history of use for longitudinal data in biostatistics (its history is reviewed by \textcite{slateStatistical}).
+Piecewise linear, random effect models have a long history of use for longitudinal data in biostatistics (reviewed by \textcite{slateStatistical}).
 Often, as here, they are used to model the temporal evolution of biomarkers.
-A few example applications are: CD4 counts following HIV infection~\autocite{langeHierarchical,lynchPredicting}, parasite density in blood samples following malaria treatment~\autocite{fogartyBayesian}, and measurements of cognitive decline due to Alzheimer's disease~\autocite{bealisleBayesian}.
+A few example applications are: CD4 counts following HIV infection~\autocite{langeHierarchical,lynchPredicting} and parasite density in blood samples following malaria treatment~\autocite{fogartyBayesian}.
 
@@
 However, the random effect(s) for each individual is drawn towards a common mean, allowing information on their value to be shared between individuals.
+
 Following much of the literature mentioned previously, I will take a Bayesian approach in this chapter.
@@
 
-$\tau_{\max,i}$ is inferred independently (\ie without a shared mean) for each individual because time is defined for each individual based on when they are first tested.
+$\tau_{\max,i}$ is a fixed effect for each individual because time is defined for each individual based on when they are first tested.
 This time is based on logistical and behavioural factors such as delays in the NHS Test and Trace system and the study design.
@@
 
-The noisy RT-PCR observations of viral log load are modelled by assuming each measurement of log viral load has an iid error distributed $\N(0, \sigma_v^2)$.
+The noise in the RT-PCR is modelled by assuming each measurement of log viral load has an iid error distributed $\N(0, \sigma_v^2)$.
 Measurements are left censored at the limit of detection, $\vlod$, meaning that any observation which would give a log viral load of less than $\vlod$ is instead observed as negative.
@@
 Based on the fit to the data achieved when implementing an earlier version of the model (see \cref{ATACCC:sec:original-analysis}), I hypothesised that the cause of this was overcomplicating the model for test failures.
-Specifically, the model excluding false positives explains the data well (the results in \cref{fig:paper:duration,ATACCC:fig:goodness-of-fits} do not allow false negatives but fit the data well).
+Specifically, excluding false positives from the model still explains the data well (the results in \cref{fig:paper:duration,ATACCC:fig:goodness-of-fits} do not allow false negatives but fit the data well).
+Furthermore, a prior that false negatives and false positives are equally likely is not supported by other information (see \cref{biology-data:sec:measurement-error}).
 Therefore, I simplified the model for test failures to only false negatives.
@@
 So far, only the viral load trajectories have been considered; \textcite{hakkiOnset} does not consider duration.
-In this section, I derive an expression for an infection episode's duration, $d_i$, in terms of $\vec{\phi}_i$.
+In this section, I derive an expression for individual $i$'s infection episode duration, $d_i$, in terms of $\vec{\phi}_i$.
 
-The time that individual $i$ is first and last positive are the two values of $t$ such that $\log v_i(t) = v_\text{lod}$.
+The time that individual $i$ is first and last detectable are the two values of $t$ such that $\log v_i(t) = v_\text{lod}$.
 Denote these times as $\tau_{0,i}$ and $\tau_{1,i}$.
-Hence, individual $i$'s duration of RT-PCR positivity, in continuous time, is $d_i = \tau_{1,i} - \tau_{0,i}$.
+Hence, $d_i = \tau_{1,i} - \tau_{0,i}$.
 
@@
 Therefore, the distribution of $d_i$ across the population is determined by $\vec{\phi}_i$'s mean, $\vec{\mu}_\phi$, and covariance matrix, $\matr{\Sigma}_\phi$.
-Hence, the uncertainty in the posterior distribution of the discretised duration cumulative distribution function for group $g$, $F_{D,g}(d) = \prob(D_i \leq d + 0.5)$ originate from the uncertainty in $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$.
+% Hence, the uncertainty in the posterior distribution of the discretised duration cumulative distribution function for group $g$, $F_{D,g}(d) = \prob(D_i \leq d + 0.5)$ originate from the uncertainty in $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$.
 
@@
 \begin{algorithm}
-\For{$i = 1, \dots, M$}{
+\For{$i = 1, \dots, N$}{
   \For{$j = 1, \dots, M$}{
@@
   The posterior distribution is represented as a sample $F_{D,g}^{(1)}, \dots, F_{D,g}^{(N)}$.
-  $\vec{\mu}_\phi^{(i)}$ and $\matr{\Sigma}_\phi^{(i)}$ denote the $i$th posterior sample of $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$ respectively, taken from the MCMC output.
+  $\vec{\mu}_\phi^{(i)}$ and $\matr{\Sigma}_\phi^{(i)}$ denote the $i$th posterior sample of $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$ respectively, taken from the MCMC output which produced $N$ posterior samples.
   $M$ is the number of individuals to draw, I found $M =\numprint{100000}$ produced stable estimates of the tail of $F_{D,g}(d)$ if $\dmax = 45$.
@@
 \end{algorithm}
+
+Applying \cref{ATACCC:alg:MonteCarlo} produces a sample of the posterior distribution of $F_{D,g}(d)$.
+One sample is produced for each iteration of the MCMC output used to estimate $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$.
 
@@
 The false negative rate, $\rho$, is estimated to be quite low, 5.3\% (95\% CrI: 3.7--7.3\%).
-The test sensitivity, which will be relevant for \cref{E-imperf-test}, is the probability that a detectable individual tests positive, $\psens = 1 - \rho$.
+The test sensitivity, which will be relevant in \cref{E-imperf-test}, is the probability that a detectable individual tests positive, $\psens = 1 - \rho$.
 
@@
 Almost all test results are captured in the 95\% credible intervals.
-Individual 55 has some fitting issues, explored further below.
 \begin{figure}
@@
 However, this was greatly improved with the simplified model, occurring only for parameters relating to a single individual, number 55.
-Individual 55 has an extremely noisy set of observations.
+
+Individual 55 has an extremely noisy set of observations (see \cref{ATACCC:fig:goodness-of-fits}).
 This means that two viral load trajectories fit the data well (shown in \cref{ATACCC:fig:individual-55}).
@@
 The second, found by the other six chains, is more symmetrical, peaking 7.2 days after the first test.
-The population-level parameter estimates agree between the two sets of chains; therefore, the issue should not affect any estimates of interest.
+The population-level parameter estimates agree between the two sets of chains (not shown); therefore, the issue should not affect any estimates of interest.
 \begin{figure}
@@
 
-A strength of the ATACCC data is its general population sample.
+A strength of the ATACCC data are its general population sample.
 This is unlike most estimates (see \cref{E-intro:sec:previous-duration-estimates}) which are from hospitalised patients or healthcare workers.

### 2024-03-26 cis-imperfect-testing.tex
proof ch 6 inc new figures

@@
 However, the initial application of this framework produced an implausibly short estimate.
-In this chapter I identify this issue to be the result of false negatives, a form of misclassification bias, within the CIS (\cref{imperf-test:sec:problem}).
+In this chapter, I identify this issue to be the result of false negatives, a form of misclassification bias, within the CIS (\cref{imperf-test:sec:problem}).
 I then propose several models of false negatives (\cref{imperf-test:sec:simulate}) and extend the model of \cref{E-perf-test} to incorporate one of them (\cref{imperf-test:sec:modelling})
@@
 
-Applying the method of \cref{E-perf-test} to CIS data estimates fewer episodes with a duration of less than three weeks, compared to \cref{E-ATACCC} (see \cref{imperf-test:fig:problem-cis-estimates}).
-The \cref{E-ATACCC} estimates are more reliable over the first 2--3 weeks due to them having frequent data collection here (see \cref{E-biology-data:sec:ataccc,E-ATACCC:sec:discussion}).
+Applying the method of \cref{E-perf-test} to CIS data estimates a median duration of less than 14 days, several days shorter than the estimates from \cref{E-ATACCC} (see \cref{imperf-test:fig:problem-cis-estimates}).
+The \cref{E-ATACCC} estimates are more reliable over the first 2--3 weeks due to ATACCC's frequent data collection (see \cref{E-biology-data:sec:ataccc,E-ATACCC:sec:discussion}).
 This suggests an issue in the estimation using CIS data.
@@
 
-Comparing repeated data simulations, using the set-up of \cref{E-perf-test:sec:simulation-study}, to the observed CIS data, suggests that this is due to the high frequency of single positive episodes in the CIS data (compare the histogram with $\psens=1$ and the vertical line in \cref{imperf-test:fig:sim-single-pos}).
-A single positive episode is an episode containing exactly one positive test; intuitively, these cause estimates to be shorter because the lower bound on the length of time the episode is 1 day, the shortest possible.
+To investigate this further, I simulated \numprint{1000} datasets using the setup of \cref{E-perf-test:sec:simulation-study}.
+The simulations produced far fewer single positive episodes than seen in the CIS data (compare the histogram with $\psens=1$ and the vertical line in \cref{imperf-test:fig:sim-single-pos}).
+A single positive episode is an episode containing exactly one positive test; intuitively, these cause shorter estimates because the lower bound on the length of time the episode is one day, the shortest possible.
 Specifically, if $j$ is a single positive episode then $j$'s admissible region (as defined in \cref{E-perf-test:sec:model}) includes $b_j = e_j = r_j^{(b)} = l_j^{(e)}$.
-In this case, the beginning and end day of the episode are both equal to the day of this one positive test.
 If this is true, then the duration is $d_j = e_j - b_j + 1 = 1$.
@@
 Intermittent negatives (described in  \cref{E-biology-data:sec:PCR}) show that they occur within the CIS.
-Furthermore, in \cref{imperf-test:sec:simulate}, I expanded the simulation study of \cref{E-perf-test:sec:simulation-study} to include false negatives (see  for more details).
+Furthermore, in \cref{imperf-test:sec:simulate}, I expanded the simulation study of \cref{E-perf-test:sec:simulation-study} to include false negatives.
 This reproduced the CIS data more faithfully and exhibited the same issues when estimating the duration distribution (see \cref{imperf-test:fig:sim-single-pos}).
@@
 In this section, I introduce a probabilistic model for the generation of false negatives.
-The simplest model for false negatives is a constant, non-zero probability of returning a negative test when detectable.
-That is the test sensitivity, $\psens$, is less than 1.
+Recall, from \cref{E-biology-data:sec:measurement-error}, that the test sensitivity, $\psens$, is the probability of a positive test result given that the individual is detectable.
+The simplest model for false negatives is a constant $\psens < 1$.
 
-Under this model, each test result in a detectable individual is an iid Bernoulli random variable with a probability $\psens$ of producing a positive.
-As previously, tests in individuals without ongoing infection episodes are always negative.
+Under this model, each test in a detectable individual generates an iid $\Ber(\psens)$ result where a one indicates a positive and zero indicates a negative.
+As previously, tests in undetectable individuals are always negative.
 I modified the simulation study from \cref{E-perf-test:sec:simulation-study} to generate test results in this way.
@@
 
-The CIS data provides evidence for a test sensitivity declining over the course of an episode.
+The CIS data provides evidence for the test sensitivity declining over the course of an episode.
 \Cref{imperf-test:fig:bounding-cis-sensitivity} demonstrates this by using the following classification for test results.
@@
 \end{enumerate}
-Here, I am assuming that the probability of two false negative tests at the end of an episode is negligible.
+The classification assumes that the probability of two false negative tests at the end of an episode is negligible.
 As test sensitivity is the proportion of individuals who should test positive that do so, it can be empirically calculated as number of true positives / (number of true positives + number of false negatives).
-Therefore, I calculate the test sensitivity for each day, considering the first positive result in an episode as day 0.
-As the latent start day of the infection episode is before the first positive result, day 0 reflects some unknown number of days into the infection episode.
-I bound the test sensitivity on each day by assuming either all or none of the possible false negatives (group 3) are in fact false negatives.
+This can be bounded by assuming either all or none of the possible false negatives (group 3) are in fact false negatives.
+I calculate the test sensitivity for each day, where the day of the first positive result in an episode is day zero.
+Day zero is an unknown number of days after the episode's latent true beginning time.
 % In \cref{imperf-test:fig:bounding-cis-sensitivity} we consider the values produced by assuming the negative following the last positive in an episode could be either a true or false negative as a function of time since the individual was detected (\ie the first positive test in the episode).
-This generates broad bounds, although suggest a declining test sensitivity over the course of an episode (see \cref{imperf-test:fig:bounding-cis-sensitivity}).
+I generate broad bounds, although suggest a declining test sensitivity over the course of an episode (see \cref{imperf-test:fig:bounding-cis-sensitivity}).
 Small numbers of tests mean that the series is noisy near the start and the end, such as both bounds being 0 on some days after day 40.
@@
 
-To assess the sensitivity of the procedure to a time-varying test sensitivity, it will be useful to simulate from a model with a varying test sensitivity as a function of days since the start of the infection episode.
+The following model allows simulating with a varying test sensitivity, allowing assessment of the inference procedure's sensitivity to this.
 Through consideration of \cref{imperf-test:fig:bounding-cis-sensitivity}, and the interaction of false negatives due to viral load in \cref{E-ATACCC}'s results, I propose the following model.
@@
 where $t$ is the number of days since infection.
-Therefore, this model consists of a linear decline in test sensitivity before a constant minimum point.
+The model consists of a linear decline in test sensitivity before a constant minimum point.
 
@@
 The result of the test is then a Bernoulli random variable, independent of all other considerations.
-Adapting the simulation study to incorporate this test sensitivity is much more similar to the data (see purple histogram in \cref{imperf-test:fig:sim-single-pos}).
-This model does not require low test sensitivities early in the infection, maintaining compatibility with the results in \cref{E-ATACCC}.
+Incorporating this test sensitivity into the simulation gives much more similar results to the CIS data (see purple histogram in \cref{imperf-test:fig:sim-single-pos}).
+Furthermore, it does not require low test sensitivities early in the infection, maintaining compatibility with the results in \cref{E-ATACCC}.
 
@@
 Similar likelihoods have previously appeared in the literature~\autocite[e.g.][eq.\ (2)]{piresIntervalMisclassify}.
-However, this prior work was for singly interval censored data; incorporating the double interval censored nature of the CIS data will involve summing over the possible episode start times.
+However, this prior work was for singly interval censored data; incorporating the double interval censored nature of the CIS data involves summing over the possible episode start times.
 
@@
 This assumption is reasonable because, since a positive test follows at $r_j^{(b)}$, the negative at $l_j^{(b)}-1$ is likely early in the infection and the test sensitivity is high early in an infection.
-Therefore, this is unlikely to be a false negative because, if the individual was detectable at this time, they are in a period with high test sensitivity.
 True negatives occur with probability 1, and hence this test does not contribute to the likelihood.
 
-Denote the set of the remaining test times to consider as $\sched'_{i(j)}$.
-$\sched'_{i(j)}$ is the set of times individual $i(j)$ was tested between $l_j^{(b)}$ and $r_j^{(e)} + 1$ inclusive.
+Denote the set of the remaining test times to consider as $\sched'_{i(j)}$, the set of times individual $i(j)$ was tested between $l_j^{(b)}$ and $r_j^{(e)} + 1$ inclusive.
 Denote their results by $\vec{y}_j = \{ y_{i(j)}(t) \ssep t \in \sched'_{i(j)} \}$ where $y_{i(j)}(t) = 1$ if the test on day $t$ in individual $i(j)$ is positive and 0 otherwise.
@@
 Therefore, the test results $\vec{y}_j$, except the test at $r_j^{(e)}+1$, are either true positives or false negatives.
+This gives $t_+ = \sum_{t \in \sched'_{i(j)}} y_{i(j)}(t)$ true positives and $f_- = \sum_{t \in \sched'_{i(j)}} (1 - y_{i(j)}(t))$ false negatives.
 
@@
 
-First, if $E_j \leq r_j^{(e)}$, meaning that the test at $r_j^{(e)}+1$ is a true negative and the end of the episode is interval censored as in the previous chapter.
+First, the case when $E_j \leq r_j^{(e)}$.
+In this case, the test at $r_j^{(e)}+1$ is a true negative and the end of the episode is interval censored as in the previous chapter.
 % In this case, the test at $r_j^{(e)} + 1$ is a true negative, as are all other tests not in $\sched'_{i(j)}$.
@@
 &= \prob(\vec{y_j} \mid l_j^{(e)} \leq E_j \leq r_j^{(e)}, B_j = b, p_\text{sens}, \vec{\theta}) \prob(l_j^{(e)} \leq E_j \leq r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
-&= \left( \prod_{t \in \sched'_{i(j)}} p_\text{sens}^{y_{i(j)}(t)} (1 - p_\text{sens})^{(1 - y_{i(j)}(t))} \right) \left( S_{\vec{\theta}}(l_j^{(e)} - b + 1) - S_{\vec{\theta}}(r_j^{(e)} - b + 2) \right)
+&= p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} \left( S_{\vec{\theta}}(l_j^{(e)} - b + 1) - S_{\vec{\theta}}(r_j^{(e)} - b + 2) \right)
 \label{imperf-test:eq:ll-ei-lt-ri}
@@
 
-Second, if $E_j > r_j^{(e)}$.
+Second, the case when $E_j > r_j^{(e)}$.
 In this case, the test at $r_j^{(e)}$ is a false negative, occurring with probability $(1 - p_\text{sens})$.
@@
 &= \prob(\vec{y_j} \mid E_j > r_j^{(e)}, B_j = b, p_\text{sens}, \vec{\theta}) \prob(E_j > r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
-&= \left( \prod_{t \in \sched'_{i(j)}} p_\text{sens}^{y_{i(j)}(t)} (1 - p_\text{sens})^{(1 - y_{i(j)}(t))} \right) (1 - p_\text{sens}) S_{\vec{\theta}}(r_j^{(e)} - b_j + 2)
+&= p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} (1 - p_\text{sens}) S_{\vec{\theta}}(r_j^{(e)} - b_j + 2)
 \label{imperf-test:eq:ll-ei-gt-ri}
@@
 
-These expressions can now be used to form the full replacement for $p_{ia}$, $p'_{ia}$, the likelihood of the data $\vec{y_j}$.
+These expressions can now be used to form the full replacement for $p_{ia}$, $p'_{ia}$, the probability of observing $\vec{y_j}$ given that an infection occurs in individual $i$.
 First, augment the data with $b_j$, and split into the cases just discussed:
@@
 =& \prob(\vec{y_j} \mid p_\text{sens}, \vec{\theta}) \\
-=& \sum_{b = l_j^{(b)}}^{r_j^{(b)}} \left( \prob(\vec{y_j}, E_j \leq r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) + \prob(\vec{y_j}, E_j > r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \right) \\
+=& \sum_{b = l_j^{(b)}}^{r_{j(i)}^{(b)}} \left( \prob(\vec{y_j}, E_j \leq r_{j(i)}^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) + \prob(\vec{y_j}, E_j > r_{j(i)}^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \right) \\
   &\times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}).
 \intertext{Now, substitute in \cref{imperf-test:eq:ll-ei-lt-ri,imperf-test:eq:ll-ei-gt-ri} and take out the common factor:}
-=& \left( \prod_{t \in \sched'_{i(j)}} p_\text{sens}^{y_{i(j)}(t)} (1 - p_\text{sens})^{(1 - y_{i(j)}(t))} \right) \\ & \ \times \sum_{b = l_j^{(b)}}^{r_j^{(b)}} \left( S_{\vec{\theta}}(l_j^{(e)} - b + 1) - S_{\vec{\theta}}(r_j^{(e)} - b + 2) + (1 - p_\text{sens}) S_{\vec{\theta}}(r_j^{(e)} - b + 2) \right) \\
+=\ &  p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} \\
+ & \times \sum_{b = l_j^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_j^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 2) + (1 - p_\text{sens}) S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 2) \right) \\
   & \times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}) \\
-=& \left( \prod_{t \in \sched'_{i(j)}} p_\text{sens}^{y_{i(j)}(t)} (1 - p_\text{sens})^{(1 - y_{i(j)}(t))} \right) \\
-  & \times \sum_{b_j = l_j^{(b)}}^{r_j^{(b)}} \left( S_{\vec{\theta}}(l_j^{(e)} - b_j + 1) - p_\text{sens} S_{\vec{\theta}}(r_j^{(e)} - b_j + 2) \right) \\
+=\ &  p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} \\
+  & \times \sum_{b_j = l_j^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_j^{(e)} - b_j + 1) - p_\text{sens} S_{\vec{\theta}}(r_{j(i)}^{(e)} - b_j + 2) \right) \\
   & \times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}).
@@
 \end{align}
-Note that if $p_\text{sens} = 1$ then $p_{ia}' = p_{ia}$.
+Note that if $p_\text{sens} = 1$ then $p_{ia}' = p_{ia}$ (see \cref{perf-test:eq:pia}).
 
@@
 p_{ia}'
-&\propto \sum_{b = l_j^{(b)}}^{r_j^{(b)}} S_{\vec{\theta}}(l_j^{(e)} - b_j + 1) - p_\text{sens} S_{\vec{\theta}}(r_j^{(e)} - b + 2).
+&\propto \sum_{b = l_j^{(b)}}^{r_{j(i)}^{(b)}} S_{\vec{\theta}}(l_j^{(e)} - b_j + 1) - p_\text{sens} S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 2).
 \label{imperf-test:eq:pia-prime-constant}
@@
     \item The episode would have been detected considering only the mechanisms in \cref{E-perf-test:sec:prob-undetected}.
-That is $b_j > \min(\sched_{i(j)})$ and $e_j \geq \tau_{\sched_{i(j)}}(b_j) + b_j$.
+That is $\min(\sched_{i(j)}) < b_j \leq \max(\sched_{i(j)})$ and $e_j \geq \tau_{\sched_{i(j)}}(b_j) + b_j$.
     \item The episode ends in the interval $[\tau_{\sched_{i(j)}}(b_j) + b_j, \tau^2_{\sched_{i(j)}}(b_j) + b_j]$.
@@
 &\prob \left(
-    \tau_{\sched_{i(j)}}(b_j) + 1 \leq D_j \leq \tau^2_{\sched_{i(j)}}(b_j) + 1
+    \tau_{\sched_{i(j)}}(b_j) + 1 \leq D_j \leq \tau^2_{\sched_{i(j)}}(b_j) + 1, b_j \leq \max(\sched_{i(j)}),
     \mid B_j = b, \vec{\theta} \right) (1 - \psens) \\
-&= \left( S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b_j) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b_j) + 1) \right) (1 - \psens).
+&= \left( S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b_j) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b_j) + 2) \right) (1 - \psens).
 \end{align}
@@
 \begin{align}
-\zeta = (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_{i(j)}) + 1}^T \left( S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b) + 1) \right).
+\zeta = (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_{i(j)}) + 1}^T \left( S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b) + 2) \right).
 \end{align}
@@
 
-I simulate datasets under four conditions: a constant $\psens$ of 0.6, 0.8, or 1.0 (the latter being the same as the simulated data in \cref{E-perf-test:sec:simulation-study}), or the varying test sensitivity model proposed at the end of \cref{imperf-test:sec:simulate}.
+Denote by $\psenss$ the test sensitivity value used when simulating the dataset, with $\psens = v$ indicating the varying test sensitivity was used.
+Denote by $\psensi$ the constant test sensitivity value assumed during inference.
+
+I use one dataset from \cref{imperf-test:sec:simulate} for each $\psenss \in \{ 0.6, 0.8, 1.0, v \}$ ($\psenss = 1$ is the simulated data in \cref{E-perf-test:sec:simulation-study}).
 The simulation is otherwise unchanged from that in \cref{E-perf-test:sec:simulation-study}.
@@
 
-For each simulated dataset, I estimate the survival function using the likelihood proposed in \cref{imperf-test:sec:modelling} and assuming $\psens$ equals 0.6, 0.8, or 1.0.
-If the dataset was simulated using a constant $\psens$, and that value of $\psens$ matches the value assumed during inference, I refer to $\psens$ as being correctly specified; otherwise, I refer to it as misspecified.
+For each simulated dataset, I estimate the survival function using the posterior density proposed in \cref{imperf-test:sec:modelling} and assuming each $\psensi \in \{0.6, 0.8, 1.0\}$.
+If $\psenss = \psensi$, I refer to $\psens$ as being correctly specified; otherwise, I refer to it as misspecified.
 However, there is still some misspecification of the model to false negatives owing to the simplifying assumptions made.
-The amount that the simplifying assumptions are violated increases as $\psens$ decreases.
+The amount that the simplifying assumptions are violated increases as $\psenss$ decreases.
 
@@
 Therefore, there are a total of 24 possible scenarios, although not all are shown.
-These are each combination of: 4 simulated datasets, with varying $\psens$; 3 values of $\psens$ in inference; and 2 priors for the hazard.
+These are each combination of: 4 simulated datasets, with the different values of $\psenss$; 3 values of $\psensi$; and 2 priors for the hazard.
 
@@
   \caption[Simulation study results with constant test sensitivity]{%
-    Posterior (median and 95\% credible interval) of the survival time for the simulation study with a correctly specified test sensitivity.
+    Posterior (median and 95\% credible interval) of the survival time for the simulation study with $\psens$ correctly specified.
     The true survival time is shown in black.
@@
 Next, I considered the consequence of $\psens$ being misspecified, I use the model combination prior because it is the better performing prior in the correctly specified case.
-If the test sensitivity is misspecified, that is the assumed value for $p_\text{sens}$ in \cref{imperf-test:eq:pit-prime,imperf-test:eq:pit-prime} is different to the value used in the simulation, then the survival time will be biased (see \cref{imperf-test:fig:misspecified-test-sensitivity}).
-If $\psens$ is misspecified and too low (see \cref{imperf-test:fig:misspecified-test-sensitivity}(A)), then the posterior estimate initially follows the true value but then separates.
-The number of episodes inferred to have truly ended by the first negative is too high, and hence the survival time is overestimated.
+If the test sensitivity is misspecified then the survival time is biased.
+If $\psensi < \psenss$, then the posterior estimate initially follows the true value but then separates (see \cref{imperf-test:fig:misspecified-test-sensitivity}(A)).
+The number of episodes inferred to have truly ended by the first negative is too low, and hence the survival time is overestimated.
 This effect dominates over the opposing bias of overestimating the number of undetected episodes.
-The opposite occurs if the test sensitivity is assumed to be too high, although the posterior moves away from the truth earlier (see \cref{imperf-test:fig:misspecified-test-sensitivity}(C)).
+The opposite occurs if $\psensi > \psenss$, although the posterior moves away from the truth earlier (see \cref{imperf-test:fig:misspecified-test-sensitivity}(C)).
 \begin{figure}
@@
     The true survival time is shown in black.
-    All simulations use a constant test sensitivity of 0.8, but the inference procedure assumes different values, as per the key.
+    All results use $\psenss = 0.8$, but $\psensi$ varies, as per panel labels.
     Hence, (B) is correctly specified but (A) and (C) have misspecified $\psens$.
@@
 
-The results when a varying test sensitivity (\cref{imperf-test:eq:variable-test-sensitivity}) is used for the simulation are similar to a constant 0.8 test sensitivity (see \cref{imperf-test:fig:variable-test-sensitivity}).
+The results when $\psenss = v$ (\ie using \cref{imperf-test:eq:variable-test-sensitivity}) are similar to $\psenss = 0.8$ (see \cref{imperf-test:fig:variable-test-sensitivity}).
 This suggests that the simplified model, with constant test sensitivity, is sufficient for recovering the true survival time.
-Estimating the test sensitivity is not possible without a more complex model, as discussed in \cref{imperf-test:sec:discussion}.
-Therefore, I will apply this model to the real CIS data in the next section.
+% Estimating the test sensitivity is not possible without a more complex model, as discussed in \cref{imperf-test:sec:discussion}.
+Therefore, I apply the constant test sensitivity model to the real CIS data in the next section.
 \begin{figure}
@@
     Each panel shows the results of performing inference with a different assumed value for the test sensitivity.
-    The posterior estimate using a model with a constant test sensitivity of 0.8 (B) is similar to the true value (black line).
+    All results use $\psenss = v$, but $\psensi$ varies, as per panel labels.
   }
@@
 
-In this section I apply the approach described in this chapter to the CIS episodes dataset I described in \cref{perf-test:sec:data-notation}.
+In this section, I apply the approach described in this chapter to the CIS episodes dataset I described in \cref{perf-test:sec:data-notation}.
 This is the \numprint{4800} CIS infection episodes first detected between 10 Oct 2020 and 6 Dec 2020 inclusive and have negatives bounding the start and end time of the episode.
@@
 The uninformative prior led to high posterior estimates of $\ntot$, and hence an implausibly large number of episodes with durations of less than five days.
-Therefore, I based an informative prior for $\ntot$, $\ntot \sim \NBc(\mu, r)$ (as in \cref{E-perf-test:sec:model}, see \cref{E-distributions} for definition), using pre-existing estimates of the total number of infections to give $\mu\inform$ and $r\inform$.
+Therefore, I formed an informative prior for $\ntot$, $\ntot \sim \NBc(\mu\inform, r\inform)$ (as in \cref{E-perf-test:sec:model}, see \cref{E-distributions} for definition), using pre-existing estimates of the total number of infections to give $\mu\inform$ and $r\inform$.
 \Textcite{birrellRTM2} estimated the total number of infections in England over the time period I consider, with posterior mean \numprint{4136368} and standard deviation \numprint{27932}.
@@
 \todo{Insert correct citation for RTM paper 2 once available}
-Approximating this distribution as a negative binomial and scaling the mean to the size of the CIS cohort gives the prior $\mu\inform = \numprint{25132}$ and $r\inform = \numprint{22047}$.
+Approximating this distribution as a negative binomial and the mean number of infections to the size of the CIS cohort gives the prior $\mu\inform = \numprint{25132}$ and $r\inform = \numprint{22047}$.
 
@@
   \centering \includegraphics{cis-imperfect-testing/CIS_final}
-  \caption{Duration estimates using CIS and ATACCC data.}
+  \caption{
+    Duration estimates using the model in this chapter (CIS) alongside those from \cref{E-ATACCC} (ATACCC).
+    Uses the informative prior on $\ntot$, the model combination prior for the hazards, and $\psensi = 0.8$.
+  }
   \label{imperf-test:fig:cis-estimates}
@@
 The estimates are most sensitive to the choice of $r$, the strength of the prior on $\ntot$.
-A very weak, almost uninformative, prior on this quantity causes the posterior estimate to be much higher than the estimate from \textcite{birrellRTM2}.
+A very weak, almost uninformative, prior on this quantity causes its posterior estimate to be much higher than the estimate from \textcite{birrellRTM2}.
 When increasing the prior's strength, the posterior estimate moves towards the prior smoothly, as expected (see \cref{imperf-test:fig:ntot}).
@@
 
+The mean survival time in this final estimate is 21.2 days (95\% CrI: 20.5--21.9), compared to \cref{E-ATACCC}'s estimate of 18.8 days (95\% CrI: 16.5--21.6).
+This is a 13\% increase (95\% CrI: 2\% decrease to 29\% increase).
+
 \section{Discussion} \label{imperf-test:sec:discussion}
@@
 The estimates of the tail, the primary purpose of this chapter, are in addition somewhat robust to the choice of prior for $\ntot$.
-However, the rest of the distribution is sensitive to this choice.
+The increased estimates in the tail lead to an increased estimate of the mean survival time.
+This will decrease estimates of incidence.
 
-An informative prior on $\ntot$ is required is for the bulk of the distribution to agree with those in \cref{E-ATACCC} and the wider literature.
+Regardless of the assumptions used, a substantial portion of infection episodes last a long time (\eg more than 50 days) (see \cref{imperf-test:fig:cis-estimates,imperf-test:fig:cis-sensitivity}).
+This has implications for surveillance because infections detected by prevalence surveys could have been present for a long time.
+If an intervention such as a lockdown is implemented, these long infections could contribute a significant portion to ongoing prevalence levels.
+Furthermore, care should be taken when designing testing policies (\eg to release individuals from quarantine or discharge them from hospital) because individuals could test RT-PCR positive for a long time, but cannot be assumed to be infectious.
+
+An informative prior on $\ntot$ is required for the bulk of the distribution to agree with those in \cref{E-ATACCC} and the wider literature.
 In the simulation study, the uninformative prior on $\ntot$ performed well, even when the test sensitivity was misspecified.
-A possible explanation is that the true changes in the test sensitivity is significantly different to the function in \cref{imperf-test:eq:variable-test-sensitivity}.
-In particular, the test sensitivity for long episodes may become much lower than 50\%.
+A possible explanation is that the true variation in the test sensitivity over an episode is significantly different to the function in \cref{imperf-test:eq:variable-test-sensitivity}.
+In particular, the test sensitivity for long episodes could be lower than 50\%.
 This allows many more episodes to be undetected than the model assumes, and hence a higher $\ntot$ is required to explain the data.
-Furthermore, it violates assumptions made in \cref{imperf-test:sec:modelling}; this is similar to when the simulating using a too low test sensitivity.
+Furthermore, it violates assumptions made in \cref{imperf-test:sec:modelling}; this is similar to the situation in simulation when $\psensi < \psenss$.
 Violating these assumptions would lead to unpredictable inference results, perhaps those seen here.
@@
 However, this would require incorporating time-varying test sensitivity into the likelihood.
-If the current model, with a constant $\psens$, is used then the estimate of $\psens$ would be heavily informed be intermittent negatives (the first, constant term in \cref{imperf-test:eq:pia-prime}).
-These negatives will, in general, be further from the end of an episode than a randomly selected test.
-Therefore, considering the viral dynamics from \cref{E-ATACCC}, the viral load will be higher and false negatives rarer.
+If the current model, with a constant $\psens$, is used then the estimate of $\psens$ would be heavily informed by intermittent negatives (the first, constant term in \cref{imperf-test:eq:pia-prime}).
+These negatives will, in general, be close to the beginning of the episode with a higher test sensitivity than average.
 
 Estimating the test sensitivity excluding intermittent negatives is not possible because \cref{imperf-test:eq:pia-prime-constant,imperf-test:eq:pit-prime} are both monotonically decreasing in $\psens$; therefore, the likelihood always favours $\psens = 0$ (\ie no true positives).
-This aligns with the situation with singly interval censored, untruncated data, when stopping at the first observed time of the terminating event (which may be misclassified) means that the test sensitivity cannot be estimated~\autocite[e.g.]{titmanMisclassify}.
+This aligns with the situation with singly interval censored, untruncated data, when stopping at the first observed time of the terminating event (which may be misclassified) means that the test sensitivity cannot be estimated~\autocite[e.g.][]{titmanMisclassify}.
 
@@
 In this chapter, I developed and applied novel methodology to estimate the duration distribution of RT-PCR positivity using information from two studies: ATACCC and CIS.
-Assuming a test sensitivity of 0.8 and an informative prior on the number of infections, provided the preferred estimates.
+The preferred estimates assume a test sensitivity of 0.8 and use an informative prior on the number of infections.
 

### 2024-03-26 distributions.tex
proof ch 6 inc new figures

@@
 \hline \\
+Bernoulli & $\Ber(p)$ & $0 \leq p \leq 1$ (probability of success) & $p$ & $p(1-p)$ & $p^x(1-p)^{1-x}$ \\
 Binomial & $\Bin(n, p)$ & $n \in \nats$ (number of trials, integer)\newline $0 \leq p \leq 1$ (probability of success) & $np$ & $np(1-p)$ & ${n \choose x}p^x(1-p)^{n-x}$ \\

### 2024-03-26 cis-imperfect-testing/sim_survival.R

@@ -6,12 +6,12 @@ library(tidybayes)
 library(tidyr)
 source(here::here("utils.R"))
 
-base_plot = function(df, colour_curves_by, colour_key = NULL) {
+base_plot = function(df, colour_curves_by, colour_key = NULL, facet_suffix = "") {
     p = df |>
         mutate(
             sensitivity.model = factor(sensitivity.model),
             facet_label = latex2exp::TeX(glue::glue(
-                "({LETTERS[dense_rank(sensitivity.model)]})\\ $p_{{sens}} = {sensitivity.model}"
+                "({LETTERS[dense_rank(sensitivity.model)]})\\ $p_{{sens}}{facet_suffix} = {sensitivity.model}"
             ), output = "character")
         ) |>
         ggplot() +
@@ -67,7 +67,7 @@ ggsave(
 
 p_misspecified_sensitivity = tbl_posteriors |>
     filter(sensitivity.simulation == 0.8, survival_prior == "Combination") |>
-    base_plot(sensitivity.model, colour_key = expression(p[sens]))
+    base_plot(sensitivity.model, colour_key = expression(p[sens]^`(i)`), facet_suffix = "^{(i)}")
 ggsave(
     filename = here::here("cis-imperfect-testing/sim-misspecified-sensitivity.pdf"),
     plot = p_misspecified_sensitivity,
@@ -79,7 +79,7 @@ ggsave(
 
 p_variable_sensitivity = tbl_posteriors |>
     filter(is.na(sensitivity.simulation), survival_prior == "Combination") |>
-    base_plot(sensitivity.model, colour_key = expression(p[sens]))
+    base_plot(sensitivity.model, colour_key = expression(p[sens]^`(i)`), facet_suffix = "^{(i)}")
 ggsave(
     filename = here::here("cis-imperfect-testing/sim-variable-sensitivity.pdf"),
     plot = p_variable_sensitivity,

### 2024-03-26 cis-perf-new-ll.tex
Updates from Overleaf

@@
 Denote the characteristics of episode $j$ as $\nu_j = (B_j, E_j, i(j))$.
-I now define a series of sets $\set{E}_1, \dots, \set{E}_{N_E}, \set{E}_u$ for an integer $N_E$ such that any possible $\nu_j$ is in exactly one of the classes and each set corresponds to the possible infection episdoes that lead to identical observations (conditional on the test schedules).
+I now define a series of sets $\set{E}_1, \dots, \set{E}_{N_E}, \set{E}_u$ and an integer $N_E$ such that any possible $\nu_j$ is in exactly one of the classes and each set corresponds to the possible infection episdoes that lead to identical observations (conditional on the test schedules).
 
-If $\nu_j$ would be undetected then $\nu_j \in \set{E}_u$.
+Define $\set{E}_u$ such that $\nu_h \in \set{E}_u$ if and only if $\nu_j$ would be undetected (if it occurred).
 Hence, $\prob(\nu_j \in \set{E}_u \mid \vec{\theta}) = p_u$.
@@
 
-More formally, the sets can be defined in terms of an equivalence relation $\equiv$.
-Define $\equiv$ such that $\nu_j \equiv \nu_{j'}$, where $j$ and $j'$ are arbitrary possible episodes (\ie these are episodes that, a priori, could have occurred but may or may not have done), if and only if either both $j$ and $j'$ would be undetected (as defined in \cref{E-perf-test:sec:prob-undetected}) or the following four conditions all hold.
+More formally, the sets can be defined in terms of an equivalence relation $\equiv$ and two arbitrary possible episodes $j$ and $j'$.
+Define $\nu_j \equiv \nu_{j'}$ to hold if and only if either both $j$ and $j'$ would be undetected (as defined in \cref{E-perf-test:sec:prob-undetected}) or the following four conditions all hold.
 \begin{enumerate}
@@
     \item $i(j) = i(j')$.
-    \item $\tau_{\sched_{i(j)}}(B_j) = \tau_{\sched_{i(j')}}(B_{j'})$.
+    \item $\tau_{\sched_{i(j)}}(B_j) = \tau_{\sched_{i(j')}}(B_{j'})$, where $\tau$ is as defined in \cref{E-perf-test:eq:tau-def}.
     \item $\tau_{\sched_{i(j)}}(E_j) = \tau_{\sched_{i(j')}}(E_{j'})$.
@@
 Then each $\set{E}_k$ is an equivalence class under this relation and the set of all possible $\nu_j$.
+That is, taking any $k = 1, \dots, N_E, u$ then $\nu_j \in \set{E}_k$ and $\nu_{j'} \in \set{E}_k$ if and only if $\nu_j \equiv \nu_{j'}$.
 

### 2024-03-26 introduction.tex
Updates from Overleaf

@@
 This measure is highly related to the duration of positivity.
-However, the study's cohort consisted only of healthcare workers, \ie working-age adults.
+However, the study's cohort consisted only of healthcare workers, \ie working-age adults, and included only 27 infections.
+Therefore, the data are not suitable for inferring the full distribution in the general population.
 
@@
 For example, a third were travellers detected during their quarantine period on arrival to New Zealand but would have had negative RT-PCR tests before their departure.
+Furthermore, asymptomatic infections were excluded due to limitations of their methodology.
+Duration of positivity may be correlated with severity of symptoms, and hence this exclusion could bias their results.
 

### 2024-03-26 transmission.tex
proof ch 7

--- 2024-03-25 transmission.tex
@@
 There are several attractive features about this approach in the context here.
-The backcalculation approach makes no assumptions about how transmission occurs; the assumptions have the potential to introduce bias.
+The backcalculation approach makes few assumptions about how transmission occurs; the assumptions have the potential to introduce bias.
 It can also easily incorporate arbitrary covariates to adjust for non-response bias.
@@
 
-In this section, I will apply the theory introduced in \cref{E-inc-prev} to estimate incidence.
+In this section, I apply the theory introduced in \cref{E-inc-prev} to estimate incidence.
 I view the CIS as a series of daily, cross-sectional prevalence surveys.
@@
 \end{align}
-where $\vec{Z}$ is a vector such that $Z_{i,t}$ is the incidence proportion in stratum $i$ on day $t$, and has prior $F_Z$; $P_{i,t}$ is the prevalence in stratum $i$ on day $t$; $\rho$ is the overdispersion parameter of the beta-binomial distribution with prior $F_\rho$; $Y_{i,t}$ is the number of positive tests in stratum $i$ on day $t$; and $n_{i,t}$ is the number of tests in stratum $i$ on day $t$.
+where $\vec{Z}$ is a vector such that $Z_{i,t}$ is the incidence proportion in stratum $i$ on day $t$, with a joint prior $F_Z$; $P_{i,t}$ is the prevalence in stratum $i$ on day $t$; $\rho$ is the overdispersion parameter of the beta-binomial distribution with prior $F_\rho$; $Y_{i,t}$ is the number of positive tests in stratum $i$ on day $t$; and $n_{i,t}$ is the number of tests in stratum $i$ on day $t$.
 Fitting this model to a prevalence survey such as the CIS (\eg using MCMC) gives a posterior distribution of $\vec{Z}$, the target of inference.
 
-The relationship between $P$ and $Z$ is from \cref{E-inc-prev:eq:EPt}, although dividing both sides by $\Npop$ to have the values in terms of incidence proportion and prevalence rather than incidence and number of prevalent individuals.
+The relationship between $P$ and $Z$ is from \cref{E-inc-prev:eq:EPt}, although dividing both sides by the stratum's population (assumed constant) to have the values in terms of incidence proportion and prevalence rather than incidence and number of prevalent individuals.
 
 Specifying $F_Z$ would depend on the application of the approach, but should consider the need to provide structure to stabilise the problem (see \cref{E-intro:sec:disease-model}).
-In the context of the CIS, it needs to allow borrowing information across time and/or strata; this is required because some stratum and time combinations have few positives (fewer than five per day).
+In the context of the CIS, it needs to allow borrowing information across time and/or strata (see \cref{E-biology-data:sec:MRP}).
 
@@
 However, it poses several pragmatic challenges which means I favour the prevalence-inferred approach.
-Here I explain that further developments, beyond the scope of this thesis, would be required to use the incidence-inferred approach within the computational constraints of the SRS.
 
-It will be convenient to define a variable equating to the posterior initially estimated.
+It will be convenient to define a variable equating to the quantity initially estimated.
 Let $\pi_{i,t} = Z_{i,t}$ under the inference-inferred approach and $\pi_{i,t} = P_{i,t}$ is the prevalence.
@@
     \item Borrow information across the different covariates.
-        The borrowing should penalize changes in the distribution of incidence/prevalence across strata in the same geographic region being constant.
+        The borrowing should penalize changes in the distribution of incidence/prevalence across strata in the same geographic region.
         A constant distribution is predicted by the theoretical study of epidemics, for example the dynamics of stratified, mechanistic transmission models (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) over short periods of time with negligible behavioural and immunity changes.
@@
 
-A parsimonious model satisfying properties 1 and 2 is to specify $\logit(\pi_{i,t}) = g_\varphi(i, t)$, where $g$ is a function parameterised by $\varphi$ and the prior on $\varphi$ is specified to favour $d^2g/dt^2 = 0$, \ie favour linear growth in $\logit(\pi_{i,t}$.
+A parsimonious model satisfying properties 1 and 2 is to specify $\logit(\pi_{i,t}) = g_\varphi(i, t)$, where $g$ is a function parameterised by $\varphi$ and the prior on $\varphi$ is specified to favour $d^2g/dt^2 = 0$, \ie favour linear growth in $\logit(\pi_{i,t})$.
 The prior $F_Z$ or $F_P$ is now a prior on $\varphi$.
@@
 The logit transformation also allows property 4 to be conveniently satisfied using an additive model for $g_\varphi$.
-An additive model on the logit scale is multiplicative on the natural scale for the small values of $\pi_{i,t}$ which will be common.
+An additive model on the logit scale is approximately multiplicative on the natural scale for small values of $\pi_{i,t}$ (which will be common).
 
@@
 Properties 3 and 5 suggest the use of an approximate method for computing the posterior distribution.
-In particular, using MCMC, sufficient covariates to adjust for the non-response bias, and a period of more than 6--8 weeks leads to a model that is computationally prohibitive ~\citePersonalComms{Koen Pouwels}.
-Instead, previous MRP implementations used INLA (see \cref{E-biology-data:sec:MRP}).
+In particular, using MCMC, sufficient covariates to adjust for the non-response bias, over a period of more than 6--8 weeks leads to a model that is computationally prohibitive ~\citePersonalComms{Koen Pouwels}.
+Instead, previous MRP implementations for the CIS used INLA (see \cref{E-biology-data:sec:MRP}).
 
 INLA-based inference can only be applied to LGMs (\emph{Latent Gaussian Models}); see \cref{E-transmission:sec:INLA} for further details on INLA and LGMs.
-The LGM framework requires that a transformation of $P_{i,t}$ is linear in $g_\varphi = \logit(\phi_{i,t})$.
+The LGM framework requires that a transformation of $P_{i,t}$ is linear in a vector of $g_{\varphi}(i, t)$s combining all $i$s and $t$s.
 
@@
 In this case, the incidence prior to day $t$ imply that the prevalence on day $t$ is higher than that predicted by the fitted regression model for this strata/day combination.
-There are two sets of assumptions which are most likely to be violated.
-
-The first set of assumptions is in estimating the survival function, $S(t)$, which is assumed to be known with no uncertainty in this procedure.
-However, \cref{E-imperf-test} showed that the estimate of $S(t)$ is sensitive to assumptions made, such as the assumed $\psens$ and strength of priors used.
-
-The second set of assumptions are in specifying the prior for $\vec{P}$.
-In particular, the prior places non-zero mass on prevalence values which imply negative incidence under the other model assumptions.
-Rectifying this assumption is challenging without switching to the inference-inferred approach, discussed further in \cref{backcalc:sec:results}.
 
@@
 
-A slow increase in incidence is estimated to occur over the autumn 2020 period, peaking in late November in most regions (see \cref{transmission:fig:backcalc-regions}).
+A slow increase in incidence is estimated to occur over the autumn 2020 period, peaking in October in most regions (see \cref{transmission:fig:backcalc-regions}).
 An exception is the North East and East of England regions (see \cref{transmission:fig:backcalc-regions}(C) and (D) respectively); these regions show no consistent increase or decrease in incidence during October.
@@
 The most explored explanation for this is the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocite{walkerTracking,daviesEstimated,lythgoeLineage}.
-There is evidence that when incidence starts increasing is correlated with when the Alpha variant's spread (see \cref{transmission:fig:backcalc-alpha}).
+There is evidence that when incidence starts increasing is correlated with the timing of the Alpha variant's spread (see \cref{transmission:fig:backcalc-alpha}).
 A second peak occurs in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
@@
 
-An edge effect where a rapid rise is seen between day 1 and 2 is visible (see \cref{transmission:fig:backcalc-start-effect}).
+An edge effect, a rapid rise in incidence between days one and two, is visible (see \cref{transmission:fig:backcalc-start-effect}).
 This is due to the incorrect assumption that incidence is constant in the past.
 Changing this assumption (\eg to exponential growth) modifies the estimated value on the first day, but the rest of the series is insensitive to this change (not shown).
-Therefore, the first day's estimate is unreliable, and has been excluded from all other plots.
+Therefore, the first day's estimate is unreliable, and has been excluded from all other results.
 However, the rest of the series is robust to this assumption and hence reliable.
@@
 
+The issues with negative incidence are likely caused by issues with modelling assumptions.
+There are two sets of assumptions which are most likely to be violated.
+
+The first set of assumptions is in estimating the survival function, $S(t)$, which is assumed to be known with no uncertainty in this procedure.
+However, \cref{E-imperf-test} showed that the estimate of $S(t)$ is sensitive to assumptions made, such as the assumed $\psens$ and strength of priors used.
+
+The second set of assumptions are in specifying the prior for $\vec{P}$.
+In particular, the prior places non-zero mass on prevalence values which imply negative incidence under the other model assumptions.
+Rectifying this assumption is challenging without switching to the inference-inferred approach, discussed further in \cref{backcalc:sec:results}.
+
 Alternative inference techniques could allow more powerful modelling approaches, if they allowed a broader set of models than LGMs.
@@
 This could allow more interpretable priors to be used because the infection process is the process that corresponds to biological and behavioural processes.
-Furthermore, the incidence proportion could be restricted to be non-negative, preventing the negative incidence seen in some strata.
+Furthermore, it would prevent the issues with negative incidence.
 
@@
 
-In common with \cref{E-SEIR:sec:mechanistic-models}, the notation in this section differs slightly from the rest of the thesis.
+In common with \cref{E-SEIR:sec:mechanistic-models}, the notational conventions in this section differs slightly from the rest of the thesis.
 
@@
 However, in SARS-CoV-2, population behaviour changed dramatically, both spontaneously (\eg due to fear) and in response to government-imposed restrictions.
-Therefore, following \textcite{birrellRealtime}, I introduce a time-dependence giving: $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'}(t) N_{a'} / N_a$; $N_a$ and $N_{a'}$ are constant because the change in the total population is negligible over the length of time considered here.
+Therefore, following \textcite{birrellRealtime}, I introduce a time-dependence giving: $\beta_{aa'}(t) = \kappa_{aa'}(t) q_{aa'}(t) N_{a'} / N_a$; $N_a$ and $N_{a'}$ are constant because the change in the total population is negligible over the length of time considered here.
 
@@
 Google mobility data are based on the location of Android users over time, relative to a pre-pandemic baseline.
-The mobility and school data are averaged over a week then used to down or up-weight the number of contacts in each setting, before combining contacts across settings to form $\kappa_{aa'}(t)$.
+The mobility and school data are averaged over a week down or up-weight the number of contacts in each setting, before combining contacts across settings to form $\kappa_{aa'}(t)$.
 $\kappa_{aa'}(t)$ is assumed to be piecewise constant, changing only at the start of each week.
@@
     \begin{tabular}{c c c c c}
-        $l$ & $\omega_1$ & $\omega_{2}$ & $\omega_{3}$ & $\pi$ \\
+        $l$ & $\omega_1$ & $\omega_{2}$ & $\omega_{3}$ & $\chi$ \\
         3 & 0.0820 & 0.126 & 0.0223 & 0.00973  \\
     \end{tabular}
-    \caption{Parameter estimates for the EC distribution approximating the posterior mean of the duration distribution estimated in \cref{E-imperf-test}.}
+    \caption{Parameter estimates for the phase-type distribution with the same first three moments as the posterior mean of the duration distribution estimated in \cref{E-imperf-test}.}
     \label{SEIR:table:ec-params}
@@
           (P1) edge node {$\omega_1$} (P2)
-          (P2) edge node {$\pi \omega_{2}$} (P3)
-               edge [out=-45,in=-135] node[below] {$(1 - \pi) \omega_{2}$} (P4)
+          (P2) edge node {$\chi \omega_{2}$} (P3)
+               edge [out=-45,in=-135] node[below] {$(1 - \chi) \omega_{2}$} (P4)
           (P3) edge node {$\omega_{3}$} (P4);
@@
     \frac{d\vec{p_2}(t)}{dt} &= \omega_1 \vec{p_1}(t) - \omega_{2} \vec{p_2}(t) \\
-    \frac{d\vec{p_3}(t)}{dt} &= \pi \omega_{2} \vec{p_2}(t) - \omega_{3} \vec{p_3}(t)
+    \frac{d\vec{p_3}(t)}{dt} &= \chi \omega_{2} \vec{p_2}(t) - \omega_{3} \vec{p_3}(t)
 \end{aligned}
@@
 
-\subsubsection{Priors} \label{SEIR:sec:priors}
+\subsubsection{Inference} \label{SEIR:sec:inference}
 
-Priors for the parameters are given in \cref{SEIR:table:priors}.
-The parameters governing the duration of positivity ($\omega$s) were explained previously (see \cref{SEIR:table:ec-params}).
-The initial conditions and transition rate parameters are unidentifiable in an SEIR model~\autocite{dankwaStructural}, therefore strongly informative priors or fixed values are used for them.
-$\sigma_\epsilon$ and $\rho$ are given priors that favour a simpler model.
-The prior on $\sigma_\epsilon$ is a penalized complexity prior, favouring a null model of constant values for $\beta(t)$~\autocite{simpsonPenalising}.
-The prior on $\rho$ favours a null model of a binomial likelihood.
-The remaining parameters have weakly informative priors.
+Following \textcite{birrellBayesian}, I reparameterize the model to improve identifiability.
+I parameterize $\beta_1$ in terms of the initial growth of the epidemic.
+I also use a parsimonious parameterization of the initial conditions using the initial proportion of the population that are infectious and the initial proportion that are susceptible only.
 
-To form the prior for $\vec{\pi}$, I used the analysis framework of \textcite{amirthalingamSeroprevalence}, applied to seroprevalence data from blood donors~\autocite{amirthalingamSeroprevalence} and a representative sample of children~\autocite{ratcliffeCommunity} in August 2020.
-This analysis uses a Bayesian framework to correct the seroprevalence data for the sensitivity and specificity of the serological tests, using an informative prior.
-\Textcite{amirthalingamSeroprevalence} formed the prior for the tests by evaluating the tests using samples which are known to be negative (taken before the pandemic) or positive (taken from an individual following a positive RT-PCR test).
-A regression model predicts the proportion of the population with prior immunity in each age group and region.
-I approximated the posterior of this proportion with a beta distribution to form the prior for the analysis in this chapter (see \cref{SEIR:table:immunity-prior}).
+Both reparameterizations rely on an assumption that the epidemic is in a steady-state exponential growth phase at time 0 (see \cref{E-SEIR:sec:SIR}).
+During such an exponential growth phase, all components of the $\vec{e}$, $\vec{i}$ and $\vec{p}$ compartments grow at the same growth rate, $\psi$.
+That is:
+\begin{equation}
+\begin{aligned}
+\frac{d\vec{e_1}(t)}{dt} = \psi \vec{e_1}(t) \\
+\frac{d\vec{e_2}(t)}{dt} = \psi \vec{e_2}(t) \\
+\frac{d\vec{i_1}(t)}{dt} = \psi \vec{i_1}(t) \\
+\frac{d\vec{i_2}(t)}{dt} = \psi \vec{i_2}(t) \\
+\frac{d\vec{p_0}(t)}{dt} = \psi \vec{p_0}(t) \\
+\frac{d\vec{p_1}(t)}{dt} = \psi \vec{p_1}(t) \\
+\frac{d\vec{p_2}(t)}{dt} = \psi \vec{p_2}(t) \\
+\frac{d\vec{p_3}(t)}{dt} = \psi \vec{p_3}(t).
+\end{aligned}
+\label{SEIR:eq:exponential-growth}
+\end{equation}
+This follows directly from the ODEs if $\vec{s}$ is constant.
 
-\begin{landscape}
-\begin{table}
-\begin{tabular}{l c l l}
-    Parameter description & Symbol & Distribution & Comment \\
-    \hline \\
-    Mean latent period & $1/\sigma$ & 3.5 days (fixed) & Based on \textcite{zhaoEstimating} \\
-    Mean infectious period & $1/\gamma$ & 4 days (fixed) & Based on \textcite{zhaoEstimating} \\
-    Initial proportion without immunity & $\vec\pi$ & See \cref{SEIR:table:immunity-prior} & \\
-    Initial proportion infectious & $i^+$ & $\BetaDist(0.5, 1000)$ & Weakly informative \\
-    Susceptibility of children & $q_c$ & $\LN(-0.4325, 0.1174)$ & Based on \textcite{vinerTransmission}  \\
-    Initial growth rate & $\psi$ & $\N(0.06, 0.04)$ & Weakly informative \\
-    Standard deviation of random walk steps on log scale & $\sigma_\epsilon$ & $\Exponential(80)$ & penalized complexity prior \\
-    Overdispersion of observations & $\rho$ & $\Exponential(2 \times 10^5)$ & penalizing deviations from binomial
-\end{tabular}
-\caption[SEIR model priors]{Priors for each parameter. For details of the distributions and their parameterizations see \cref{E-distributions}.}
-\label{SEIR:table:priors}
-\end{table}
-\begin{table}
+The growth rate $\psi$ is easily identifiable from the data, and therefore a parameterization in terms of $\psi$ should be more computationally efficient.
+If this is the initial exponential phase, when $\vec{s} \approx 1$, then the basic reproduction number $\R$ for an SEIR model is linked to this growth rate by~\autocite{birrellBayesian,wearingAppropriate}:
+\begin{align}
+    R_{\psi} = \frac{\psi \left( \frac{\psi}{2\sigma} + 1 \right)^2}{\gamma \left(1 - \frac{1}{\left(\frac{\psi}{2 \gamma} + 1 \right)^2} \right)} \label{SEIR:eq:rtoR}.
+\end{align}
 
-\centering
-\begin{tabular}{l|lllll}
-         & 0--11 & 12--16 & 17--25 & 26--50 & 71+ \\
-        \hline
-        East Midlands & $\BetaDist(8.5, 301)$ & $\BetaDist(18, 256)$ & $\BetaDist(19, 475)$ & $\BetaDist(20, 493)$ & $\BetaDist(5, 337)$ \\
-        East of England & $\BetaDist(7.9, 236)$ & $\BetaDist(21, 248)$ & $\BetaDist(35, 677)$ & $\BetaDist(26, 613)$ & $\BetaDist(5.6, 332)$ \\
-        London & $\BetaDist(9.7, 131)$ & $\BetaDist(50, 328)$ & $\BetaDist(154, 1331)$ & $\BetaDist(106, 990)$ & $\BetaDist(7.6, 204)$ \\
-        North East & $\BetaDist(6.1, 180)$ & $\BetaDist(12, 145)$ & $\BetaDist(14, 274)$ & $\BetaDist(14, 295)$ & $\BetaDist(4.2, 240)$ \\
-        North West & $\BetaDist(12, 252)$ & $\BetaDist(38, 353)$ & $\BetaDist(72, 885)$ & $\BetaDist(44, 695)$ & $\BetaDist(6.3, 264)$ \\
-        South East & $\BetaDist(8.9, 404)$ & $\BetaDist(17, 319)$ & $\BetaDist(20, 577)$ & $\BetaDist(19, 586)$ & $\BetaDist(4.4, 356)$ \\
-        South West & $\BetaDist(6.9, 415)$ & $\BetaDist(11, 277)$ & $\BetaDist(12, 556)$ & $\BetaDist(11, 559)$ & $\BetaDist(4, 492)$ \\
-        West Midlands & $\BetaDist(8, 167)$ & $\BetaDist(19, 151)$ & $\BetaDist(32, 475)$ & $\BetaDist(33, 504)$ & $\BetaDist(5.8, 241)$ \\
-        Yorkshire and The Humber & $\BetaDist(9.2, 314)$ & $\BetaDist(19, 242)$ & $\BetaDist(27, 589)$ & $\BetaDist(18, 474)$ & $\BetaDist(4.9, 316)$ \\
-    \end{tabular}
-\caption{Priors for $\vec{\pi}$, the proportion of the population without prior immunity, in each age group and region combination.}
-\label{SEIR:table:immunity-prior}
-\end{table}
-\end{landscape}
+Using the results of \cref{E-SEIR:sec:structured-populations}, $\R$ can also be written as $R_{\beta} = \beta_1 D(\matr{\kappa}(t) \circ \matr{q} \circ \matr{N})$ (recall that $D(\matr{\beta})$ denotes the dominant eigenvalue of $\matr{\beta}$).
+Hence setting $\beta_1 = R_\psi / D(\matr{\kappa}(t) \circ \matr{q} \circ \matr{N})$ gives $R_\psi = R_\beta$.
+Using this expression, $\beta_1$ has been reparameterized in terms of $\psi$, $q_c$, and fixed quantities.
 
-\subsubsection{Inference} \label{SEIR:sec:inference}
+To form a parsimonious parameterization of the initial state, first focus on the initial proportion of the population that is in the infectious compartments.
+Recall, from \cref{E-SEIR:sec:structured-populations}, that in the exponential growth phase, the relative proportion of each stratum that is infectious is proportional to the dominant eigenvector of $\matr{\beta}$, which I denote $\vec{d}$.
+Introduce a model parameter $i^+$, the proportion of the population in either of the $i$ states at time 0 in the strata with the highest proportion infectious.
+Let $\vec{i}_0 = \vec{i}_1(0) + \vec{i}_2(0)$ be the total number of infectious individuals at time zero.
+Then:
+\begin{align}
+\vec{i_0} = \frac{i^+}{\max \vec{d}} \vec{d}.
+\label{SEIR:eq:i0}
+\end{align}
+
+Equating the expressions for $d\vec{i_2}/dt$ in \cref{SEIR:eq:fullODEs} and \cref{SEIR:eq:exponential-growth} and setting $t=0$ to allow substituting in \cref{SEIR:eq:i0} gives:
+\begin{align}
+    \vec{i_1}(0) &= \vec{i_0} \left(1 + \frac{2\gamma}{\psi + 2\gamma} \right)^{-1}.
+\intertext{By definition of $\vec{i_0}$:}
+    \vec{i_2}(0) &= \vec{i_0} - \vec{i_1}(0).
+\end{align}
+
+Equating the expressions in \cref{SEIR:eq:fullODEs} and \cref{SEIR:eq:exponential-growth}, gives the following solution for the initial state of several compartments.
+\begin{align}
+    \vec{e_2} &= \vec{i_1} \frac{\psi + 2\gamma}{2\sigma} \\
+    \vec{e_1} &= \vec{e_2} \frac{\psi + 2\sigma}{2\sigma} \\
+    \vec{p_0} &= \vec{e_1} \frac{\psi + 2\sigma}{\psi + \omega_0} \\
+    \vec{p_1} &= \vec{p_0} \frac{\omega_0}{\psi + \omega_1} \\
+    \vec{p_2} &= \vec{p_1} \frac{\omega_1}{\psi + \omega_{2}} \\
+    \vec{p_3} &= \vec{p_2} \frac{\pi \omega_{2}}{\psi + \omega_{3}}.
+\end{align}
+
+Define a parameter $\vec{\pi}$ as the proportion of each stratum without initial immunity, then $\vec{r}(0) = 1 - \vec{\pi}$.
+Finally, to ensure the population proportions sum to one, $\vec{s}(0) = \vec{\pi} - \vec{i_0} - \vec{e_1} - \vec{e_2}$.
 
@@
 $\gamma_i \to 0$ as $i \to \infty$ so the rate of adaptation is \emph{vanishing}.
-Vanishing adaptation is required for the algorithm to converge to the target distribution~\autocite[section 3]{andrieuTutorial}.}
+Vanishing adaptation guarantees the algorithm converges to the target distribution~\autocite[section 3]{andrieuTutorial}.}
  \label{SEIR:MCMC-algorithm}
@@
 
-Following \textcite{birrellBayesian}, I reparameterize the model to improve identifiability.
-I parameterize $\beta_1$ in terms of the initial growth of the epidemic.
-I also use a parsimonious parameterization of the initial conditions using the initial proportion of the population that are infectious and the initial proportion that are susceptible only.
+\subsubsection{Priors} \label{SEIR:sec:priors}
 
-Both reparameterizations rely on an assumption that the epidemic is in a steady-state exponential growth phase at time 0 (see \cref{E-SEIR:sec:SIR}).
-During such an exponential growth phase, all components of the $\vec{e}$, $\vec{i}$ and $\vec{p}$ compartments grow at the same growth rate, $\psi$.
-That is:
-\begin{equation}
-\begin{aligned}
-\frac{d\vec{e_1}(t)}{dt} = \psi \vec{e_2}(t) \\
-\frac{d\vec{e_2}(t)}{dt} = \psi \vec{e_2}(t) \\
-\frac{d\vec{i_1}(t)}{dt} = \psi \vec{i_2}(t) \\
-\frac{d\vec{i_2}(t)}{dt} = \psi \vec{i_2}(t) \\
-\frac{d\vec{p_0}(t)}{dt} = \psi \vec{p_0}(t) \\
-\frac{d\vec{p_1}(t)}{dt} = \psi \vec{p_1}(t) \\
-\frac{d\vec{p_2}(t)}{dt} = \psi \vec{p_2}(t) \\
-\frac{d\vec{p_3}(t)}{dt} = \psi \vec{p_3}(t).
-\end{aligned}
-\label{SEIR:eq:exponential-growth}
-\end{equation}
-This follows directly from the ODEs if $\vec{s}$ is constant.
+Priors for the parameters are given in \cref{SEIR:table:priors}.
+The parameters governing the duration of positivity ($\omega$s) were explained previously (see \cref{SEIR:table:ec-params}).
+The initial conditions and transition rate parameters are unidentifiable in an SEIR model~\autocite{dankwaStructural}, therefore strongly informative priors or fixed values are used for them.
+$\sigma_\epsilon$ and $\rho$ are given priors that favour parsimony.
+The prior on $\sigma_\epsilon$ is a penalized complexity prior, favouring a null model of constant values for $\beta(t)$~\autocite{simpsonPenalising}.
+The prior on $\rho$ favours a null model of a binomial likelihood.
+The remaining parameters have weakly informative priors.
 
-The growth rate $\psi$ is easily identifiable from the data, and therefore a parameterization in terms of $\psi$ should be more computationally efficient.
-If this is the initial exponential phase, when $\vec{s} \approx 1$, then the basic reproduction number $\R$ for an SEIR model is linked to this growth rate by~\autocite{birrellBayesian,wearingAppropriate}:
-\begin{align}
-    R_{(\psi)} = \frac{\psi \left( \frac{\psi}{2\sigma} + 1 \right)^2}{\gamma \left(1 - \frac{1}{\left(\frac{\psi}{2 \gamma} + 1 \right)^2} \right)} \label{SEIR:eq:rtoR}.
-\end{align}
+To form the prior for $\vec{\pi}$, I used the analysis framework of \textcite{amirthalingamSeroprevalence}, applied to seroprevalence data from blood donors~\autocite{amirthalingamSeroprevalence} and a representative sample of children~\autocite{ratcliffeCommunity} in August 2020.
+This analysis uses a Bayesian framework to correct the seroprevalence data for the sensitivity and specificity of the serological tests, using an informative prior.
+\Textcite{amirthalingamSeroprevalence} formed the prior for the tests by evaluating the tests using samples which are known to be negative (taken before the pandemic) or positive (taken from an individual following a positive RT-PCR test).
+A regression model predicts the proportion of the population with prior immunity in each age group and region.
+I approximated the posterior of this proportion with a beta distribution to form the prior for the analysis in this chapter (see \cref{SEIR:table:immunity-prior}).
 
-Using the results of \cref{E-SEIR:sec:structured-populations}, $\R$ can also be written as $R_{(\beta)} = \beta_1 D(\matr{\kappa}(t) \circ \matr{q} \circ \matr{N})$ (recall that $D(\matr{\beta})$ denotes the dominant eigenvalue of $\matr{\beta}$).
-Hence setting $\beta_1 = R_\psi / D(\matr{\kappa}(t) \circ \matr{q} \circ \matr{N})$ gives $R_\psi = R_\beta$.
-Using this expression, $\beta_1$ has been reparameterized in terms of $\psi$, $q_c$, and fixed quantities.
+\begin{landscape}
+\begin{table}
+\begin{tabular}{l c l l}
+    Parameter description & Symbol & Distribution & Comment \\
+    \hline \\
+    Mean latent period & $1/\sigma$ & 3.5 days (fixed) & Based on \textcite{zhaoEstimating} \\
+    Mean infectious period & $1/\gamma$ & 4 days (fixed) & Based on \textcite{zhaoEstimating} \\
+    Initial proportion without immunity & $\vec\pi$ & See \cref{SEIR:table:immunity-prior} & \\
+    Initial proportion infectious & $i^+$ & $\BetaDist(0.5, 1000)$ & Weakly informative \\
+    Susceptibility of children & $q_c$ & $\LN(-0.4325, 0.1174)$ & Based on \textcite{vinerTransmission}  \\
+    Initial growth rate & $\psi$ & $\N(0.06, 0.04)$ & Weakly informative \\
+    Standard deviation of random walk steps on log scale & $\sigma_\epsilon$ & $\Exponential(80)$ & penalized complexity prior \\
+    Overdispersion of observations & $\rho$ & $\Exponential(2 \times 10^5)$ & penalizing deviations from binomial
+\end{tabular}
+\caption[SEIR model priors]{Priors for each parameter. For details of the distributions and their parameterizations see \cref{E-distributions}.}
+\label{SEIR:table:priors}
+\end{table}
+\begin{table}
 
-To form a parsimonious parameterization of the initial state, first focus on the initial proportion of the population that is in the infectious compartments.
-Recall, from \cref{E-SEIR:sec:structured-populations}, that in the exponential growth phase, the relative proportion of each stratum that is infectious is proportional to the dominant eigenvector of $\matr{\beta}$, which I denote $\vec{d}$.
-Introduce a model parameter $i^+$, the proportion of the population in either of the $i$ states at time 0 in the strata with the highest proportion infectious.
-Define the total number of infectious individuals at time 0 as $\vec{i_0}$.
-Then:
-\begin{align}
-\vec{i_0} = \vec{i_1}(0) + \vec{i_2}(0) = \frac{i^+}{\max \vec{d}} \vec{d}.
-\label{SEIR:eq:i0}
-\end{align}
-
-Equating the expressions for $d\vec{i_2}/dt$ in \cref{SEIR:eq:fullODEs} and \cref{SEIR:eq:exponential-growth} and then setting $t=0$ to allow substituting in \cref{SEIR:eq:i0} gives:
-\begin{align}
-    \vec{i_1}(0) &= \vec{i_0} \left(1 + \frac{2\gamma}{\psi + 2\gamma} \right)^{-1}.
-\intertext{By definition of $\vec{i_0}$:}
-    \vec{i_2}(0) &= \vec{i_0} - \vec{i_1}(0).
-\end{align}
-
-A similar process, equating the expressions in \cref{SEIR:eq:fullODEs} and \cref{SEIR:eq:exponential-growth}, gives the following solution for the initial state of these compartments.
-\begin{align}
-    \frac{d\vec{e_1}(t)}{dt} &= \vec{\lambda}(t) \circ \vec{s}(t) - 2\sigma \vec{e_1}(t) \\
-    \frac{d\vec{e_2}(t)}{dt} &= 2\sigma \vec{e_1}(t) - 2\sigma \vec{e_2}(t) \\
-    \frac{d\vec{p_0}(t)}{dt} &= \vec{\lambda} \circ \vec{s}(t) - \omega_0 \vec{p_0}(t) \\
-    \frac{d\vec{p_1}(t)}{dt} &= \omega_0 \vec{p_0}(t) - \omega_1 \vec{p_1}(t) \\
-    \frac{d\vec{p_2}(t)}{dt} &= \omega_1 \vec{p_1}(t) - \omega_{2} \vec{p_2}(t) \\
-    \frac{d\vec{p_3}(t)}{dt} &= \pi \omega_{2} \vec{p_2}(t) - \omega_{3} \vec{p_3}(t).
-\end{align}
-
-The definition of $\vec{\pi}$ as the proportion of the population without initial immunity gives $\vec{r}(0) = 1 - \vec{\pi}$.
-Finally, to ensure the population proportions sum to one, $\vec{s}(0) = \vec{\pi} - \vec{i_0} - \vec{e_1} - \vec{e_2}$.
+\centering
+\begin{tabular}{l|lllll}
+         & 0--11 & 12--16 & 17--25 & 26--50 & 71+ \\
+        \hline
+        East Midlands & $\BetaDist(301, 8.5)$ & $\BetaDist(256, 18)$ & $\BetaDist(475, 19)$ & $\BetaDist(493, 20)$ & $\BetaDist(337, 5)$ \\
+        East of England & $\BetaDist(236, 7.9)$ & $\BetaDist(248, 21)$ & $\BetaDist(677, 35)$ & $\BetaDist(613, 26)$ & $\BetaDist(332, 5.6)$ \\
+        London & $\BetaDist(131, 9.7)$ & $\BetaDist(328, 50)$ & $\BetaDist(1331, 154)$ & $\BetaDist(990, 106)$ & $\BetaDist(204, 7.6)$ \\
+        North East & $\BetaDist(180, 6.1)$ & $\BetaDist(145, 12)$ & $\BetaDist(274, 14)$ & $\BetaDist(295, 14)$ & $\BetaDist(240, 4.2)$ \\
+        North West & $\BetaDist(252, 12)$ & $\BetaDist(353, 38)$ & $\BetaDist(885, 72)$ & $\BetaDist(695, 44)$ & $\BetaDist(264, 6.3)$ \\
+        South East & $\BetaDist(404, 8.9)$ & $\BetaDist(319, 17)$ & $\BetaDist(577, 20)$ & $\BetaDist(586, 19)$ & $\BetaDist(356, 4.4)$ \\
+        South West & $\BetaDist(415, 6.9)$ & $\BetaDist(277, 11)$ & $\BetaDist(556, 12)$ & $\BetaDist(559, 11)$ & $\BetaDist(492, 4)$ \\
+        West Midlands & $\BetaDist(167, 8)$ & $\BetaDist(151, 19)$ & $\BetaDist(475, 32)$ & $\BetaDist(504, 33)$ & $\BetaDist(241, 5.8)$ \\
+        Yorkshire and The Humber & $\BetaDist(314, 9.2)$ & $\BetaDist(242, 19)$ & $\BetaDist(589, 27)$ & $\BetaDist(474, 18)$ & $\BetaDist(316, 4.9)$ \\
+    \end{tabular}
+\caption{Priors for $\vec{\pi}$, the proportion of the population without prior immunity, in each age group and region combination.}
+\label{SEIR:table:immunity-prior}
+\end{table}
+\end{landscape}
 
@@
     \thisfloatpagestyle{empty}
-    \vspace{-3cm}
     \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/sim/data}}
@@
 These were generated according to the likelihood, \cref{SEIR:eq:likelihood}.
-The number of daily tests was set to the mean of what CIS performed in the region over this period (\numprint{1630}), distributed across the population in proportion to its size.
+The daily number of tests was set to the mean of what CIS performed in the region over this period (\numprint{1630}), distributed across the population in proportion to its size.
 
@@
 \begin{figure}
-    \thisfloatpagestyle{empty}
-    \vspace{-3cm}
     \makebox[\textwidth][c]{\includegraphics[width=0.9\paperwidth]{SEIR/sim/predictive_coverage}}
@@
         The blue line is a linear regression line.
-        For $i^+$, $q_c$, and $\psi$, the posterior recovers the true values.
-        For $\vec{\pi}$ and $\rho$, the posterior median is very similar regardless of the true value.
     }
@@
     \caption[Posterior estimates of incidence]{%
-        Posterior estimates (95\% CrI) of incidence by region and age.
-        See main text for discussion.
+        Posterior estimates (95\% CrI) of incidence proportion by region and age.
     }
@@
     \caption[Posterior estimates of peak incidence timing]{%
-        Posterior probability of the peak incidence for each region and age combination occurring on each day.
+        Posterior probability of the peak incidence proportion for each region and age combination occurring on each day.
     }
@@
 They trend downwards up until early December, then they start increasing (see \cref{SEIR:fig:beta-walk}).
-Correlation is clear across the regions even though each region is estimated independently, suggesting underlying biological or epidemiological changes across regions as opposed to modelling assumptions.
+Correlation is clear across the regions even though each region is estimated independently, suggesting underlying biological or epidemiological changes across regions.
 
@@
 The results in this section show the feasibility of fitting a mechanistic model using only data on prevalence.
-If the distributions for length-of-stay in each compartment are known, then all other model parameters can  be estimated.
+If the distributions for length-of-stay in each compartment are known, then the transmission parameters can be estimated.
 This includes the age distribution of the infections and the components of transmission.
@@
 Using a mechanistic model provides additional understanding, compared to the approach in \cref{E-backcalc}.
-For example, decomposing the change in transmission into changing contacts, $\matr{\kappa}(t)$, changing susceptibility $\vec{s}(t)$, and other unmeasured factors, $\beta(t)$.
+For example, decomposing the change in transmission into changing contacts, $\matr{\kappa}(t)$, changing susceptibility $\vec{s}(t)$, and other factors, $\beta(t)$.
 By understanding which of these components is changing, policies can be better targeted.
@@
 These hypotheses for the mechanism causing changes in $\beta_w$ are correlated across the country.
-This can explain that similar patterns emerge for $\beta_w$ in different regions (see \cref{SEIR:fig:beta-walk}), despite these being estimated independently (\ie there is no model structure which favours correlated values).
+This can explain the similar patterns that emerge for $\beta_w$ in different regions (see \cref{SEIR:fig:beta-walk}), despite these being estimated independently (\ie there is no model structure which favours correlated values).
 These correlations could be exploited to improve the precision of estimates with a more complex model which borrows information across regions, \eg a hierarchical model.
@@
 For example, in the North East, most of the data points in the 12--16 age group are below the median posterior estimate, while most of the data points in the 17--25 age group are above the median posterior estimate.
-This implies that the data would allow identification of a more flexible age-based model.
 
@@
     \centering \includegraphics{transmission/compare-NE}
-    \caption[Comparing each models estimate of North East incidence by age.]{%
+    \caption[Comparing each models estimate of North East incidence proportion by age.]{%
         Incidence proportion in each age group in the North East estimated by the two models discussed in this chapter.
@@
 \end{figure}
+
+A limitation of both sets of results is neglecting false negative and false positive test results.
+Including these would increase the estimates of incidence because false negatives are more common than false positives (see \cref{biology-data:sec:measurement-error}).
+The observation model could be adapted to include these errors, but the results of \cref{imperf-test} suggest they may change over the course of an infection.
+
+An extension of either (or both) models into the rest of the epidemic would be useful.
+The main obstacle is knowledge of how the duration of positivity distribution changes, and the need to incorporate the changes into the model.
+The SEIR model would also need to include the effect of vaccination on transmission.
 

### 2024-03-26 SEIR/sim/coverage.R

@@ -82,7 +82,8 @@ ggsave(
     plot = p_coverage,
     width = 12,
     height = 10,
-    unit = "cm"
+    unit = "cm",
+    device = cairo_pdf
 )
 
 p_true_vs_posterior = posterior_intervals |>
@@ -115,7 +116,8 @@ ggsave(
     width = 20,
     height = 22,
     units = "cm",
-    dpi = 300
+    dpi = 300,
+    device = cairo_pdf
 )
 
 # Figure for combined posterior

### 2024-03-26 SEIR/sim/predictive_check.R

@@ -5,12 +5,14 @@ library(patchwork)
 library(tidybayes)
 library(tidyr)
 source(here::here("utils.R"))
+source(here::here("transmission/utils.R"))
 
 truth = readr::read_csv(here::here("SEIR/sim/sim_output.csv"), show_col_types = FALSE)
 posterior_predictive = readr::read_csv(here::here("SEIR/sim/posteriors_predictive.csv"), show_col_types = FALSE)
 
 make_plot = function(x) {
     x |>
+        mutate(age = normalise_age_groups(age)) |>
         ggplot(aes(day, mean, colour = factor(.width), fill = factor(.width))) +
         geom_lineribbon(aes(ymin = lower, ymax = upper), alpha = 0.5) +
         geom_hline(aes(yintercept = .width)) +
@@ -74,5 +76,6 @@ ggsave(
     filename = "SEIR/sim/predictive_coverage.pdf",
     plot = p_combined,
     width = 15,
-    height = 15
+    height = 15,
+    device = cairo_pdf
 )

### 2024-03-26 transmission/backcalc-ages.R

@@ -11,7 +11,7 @@ p_age_base = readRDS(file.path(output_dir, "age.rds")) |>
     filter(daynr > 1) |>
     mutate(age_group = normalise_age_groups(age_group)) |>
     ggplot(aes(date, incidence, colour = age_group, fill = age_group)) +
-    stat_lineribbon(alpha = 0.3, .width = 0.95) +
+    stat_lineribbon(alpha = 0.3, .width = 0.95, linewidth = 0.5) +
     labs(
         x = "Date",
         y = "Incidence proportion",

### 2024-03-26 transmission/backcalc-regions.R

@@ -24,7 +24,7 @@ alpha_dates = tribble(
 incidence_plot = function(df) {
     df |>
         ggplot(aes(date, incidence)) +
-        stat_lineribbon(alpha = 0.3, .width = 0.95) +
+        stat_lineribbon(alpha = 0.3, .width = 0.95, linewidth = 0.5) +
         scale_y_continuous(labels = scales::label_percent()) +
         labs(
             x = "Date",

### 2024-03-26 introduction_new.tex

@@
 
-COVID-19, the disease caused by the virus SARS-CoV-2, is estimated to have killed approximately 15 million in the most acute phase (to the end of 2021) and 30 million in total~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
+Infection with the SARS-CoV-2 virus is estimated to have killed approximately 15 million in the most acute phase (to the end of 2021) and 30 million in total~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
 Beyond the lives lost, there were massive health, economic, and societal costs.
@@
 However, our best records suggests that, since 1600, they have (on average) occurred every four years with a mean death toll of 5.6 million~\autocite{maraniNovelEpidemics}.
-Recent estimates suggest that the current mean impact of pandemics is over 2.5 million deaths per year~\autocite{madhavPandemicMortality}.
+Recent estimates suggest that the current impact of pandemics is on average of over 2.5 million deaths per year~\autocite{madhavPandemicMortality}.
 
@@
 $\Re$ describes the current dynamics of the epidemic: $\Re > 1$ means the epidemic is growing, $\Re < 1$ means it is shrinking, and $\Re = 1$ means it is stable.
-In simple epidemic models (see \cref{E-SEIR:sec:SIR}), $\Re$ is proportional to the product of the rate of contacts, the probability of transmission upon contact, and the proportion of the population that is susceptible.
+In simple epidemic models (see \cref{E-SEIR:sec:SIR}), $\Re$ is proportional to the product of the rate of contact, the probability of transmission upon contact, and the proportion of the population that is susceptible.
 Interventions act by reducing one of these components \eg masks reduce the probability of transmission and vaccination reduce the susceptible population.
@@
 $\Rc$ is useful when comparing the effects of interventions at different points in time, when the immunity status of the population has changed.
-For example, when comparing the English lockdown in March 2020 and January 2021, there are two effects in play: there are more individuals with immunity in January 2021, and the lockdowns had different measures in place.
+For example, when comparing the English lockdown in March 2020 and January 2021, there are two effects in play: the number ofindividuals with immunity, higher in 2021; and the different measures already (?) in place when the lockdown was implemented???
 $\Rc$ quantifies the latter effect.
@@
 Reproduction numbers are strongly connected to \emph{growth rates}.
-If changes in behaviour and susceptibility are negligible, as they often are over short periods of time, epidemics grow (or shrink) exponentially.
-The rate of this exponential growth is its \emph{growth rate}, normally denoted $r$.
+If changes in behaviour and susceptibility are negligible, as they often are over short periods of time, epidemics grow (or shrink) exponentially, with a \emph{growth rate} $r$.
 For a given disease and population, an analogous growth rate can be defined for each definition of the reproduction number~\autocite{pellisEstimation,paragGrowthRates,wallingaGI}.
@@
 
-Growth rates and reproduction numbers only provide information on how an epidemic is changing.
-Yet the size of the epidemic is just as, if not more, important~\autocite{pellisEstimation}.
-\emph{Incidence} (the number of new infections in a given time period) is the primary metric for the magnitude of a pandemic.
-Incidence informs the short-term future burden of the disease, for example, the future number of severe events (hospitalisations or deaths).
-Incidence is translated into disease burden using a measure of severity; for example, the IFR (infection fatality rate), the proportion of infected individuals who die.
-Estimation of severity is beyond the scope of this thesis.
+Growth rates and reproduction numbers only provide information on whether how an epidemic is changing.
+The size of the epidemic is just as, if not more, important~\autocite{pellisEstimation}.
+\emph{Incidence}, the number of new infections in a given time period  is the primary metric for the magnitude of a pandemic.
+Incidence informs the short-term future burden of the disease, for example, the future number of severe events (hospitalisations or deaths) when combined with severity measures, such as the infection fatality rate (IFR), the proportion of infected individuals who die.
 
-\emph{Transmission} refers to the process by which an infectious disease generates new infections.
-Estimating transmission refers to inferring the parameters of this process.
-The components of transmission (\eg the number of contacts each individual makes per day) allow interventions to be targeted.
-They are also important for forecasting future incidence.
+\emph{Transmission} refers to the process by which an infectious individual generates new infections.
+Estimating transmission refers to inferring this process.
+Understanding components of transmission  (\it{e.g.} the number of contacts each individual makes per day) allow interventions to be targeted.
+and to forecast future incidence.
 
-\emph{Prevalence} is the proportion of the population with the disease at a given time.
-Whether a given individual is \emph{prevalent} (has the disease) is often measurable through diagnostic testing.
+\emph{Prevalence} is the proportion of the population with the disease at a given time .
+Whether a given individual is \emph{prevalent}, {\it{i.e.}} has the disease, is often measurable through diagnostic testing.
 Performing diagnostic tests in a sample of the population at a given point in time is known as a \emph{prevalence survey}.
@@
 
-Collecting data that directly inform inference of the quantities in this section is expensive or impossible.
-Hence, they are typically inferred from other data sources through modelling.
+Collecting data that {\it directly} inform inference of the quantities in this section is challenging or prohibitively expensive .
+Hence, they are typically estimated from other data sources through appropriate inferential methodologies.{\color{red{refer to some papers examples:Phil O'neil, stats in medicine, 1990 on linking models to data; ; De Angelis, Presanis, 2018, Analysing Multiple Epidemic Data Sources.
+"Handbook of
+Infectious Disease Data Analysis", Held, L., Hens, N., O'Neill, P.D. and Wallinga, J. (Eds.).
+Chapman & Hall/CRC, 2019.}}}
 
@@
 
-Data sources used for outbreak surveillance are often observational in nature.
+Data sources to inform outbreak surveillance are often observational in nature.
 Therefore, they suffer from biases that can be hard to understand and quantify~\autocite{shadboltChallenges}.
@@
 
-Here I discuss three types of data which are useful for surveillance: incidence-type, prevalence-type, and data from cohort studies.
+In what follows I discuss three types of data: incidence-type, prevalence-type, and data from specific studies.
 \emph{Incidence-type} and \emph{prevalence-type} data can inform estimation of incidence and transmission; however, this estimation almost always relies on prior knowledge of epidemiological and/or biological quantities.
-For example, the time between infection and symptom onset.
-These quantities can be estimated from other studies, \eg cohort studies.
+For example, the time between infection and symptom onset. These quantities can be estimated from other studies.
 
@@
 
-Incidence-type data originate from observations of the consequences of a disease, \eg the development of symptoms or admission to hospital.
-This type of data typically suffers from reporting biases, particularly that the proportion of infections which are counted can vary over time or within different subgroups of the population~\autocites[chapter 9]{lashModern}{shadboltChallenges}.
+Incidence-type data originate from observations of the consequences of an infection, \eg the development of symptoms or admission to hospital.
+This type of data typically suffers from reporting biases, particularly that the proportion of infections which are detected can vary over time or within different subgroups of the population~\autocites[chapter 9]{lashModern}{shadboltChallenges}.
 
-An example of incidence-type data are community testing, reliant on individuals presenting for a test.
+An example of incidence-type data is the number of cases identified through community testing, reliant on individuals presenting for a test.
 However, individuals' ability and willingness to be tested varies over time, which needs to be accounted for in order to perform unbiased inference.
@@
 
-Data on more severe symptoms are typically less affected by these biases.
+Data on severe disease is typically less affected by these biases.
 An extreme example is the development of AIDS following infection by HIV.
@@
 
-However, severe symptoms can be rare in some population subgroups (\eg young, otherwise healthy individuals infected with SARS-CoV-2~\autocite{wardSero,bhopalChildren}).
-This means that the observations do not provide much information on incidence in those subgroups.
+However, severe disease can be rare in some population subgroups (\eg young, otherwise healthy individuals infected with SARS-CoV-2~\autocite{wardSero,bhopalChildren}) and the resulting sparse data 
+ does not provide much information on incidence in those subgroups.
 Furthermore, the delay between infection and the severe event occurring means that recent infections cannot be estimated with much precision~\autocite{swallow2022challenges}.
@@
 
-Prevalence-type data originate from observations of the number or proportion of a population in a specific \emph{disease state} at a specific time.
-Commonly used disease states are the presence or value of a biomarker or symptoms.
-For example, the number of hospital beds occupied by individuals with the disease or the proportion of the population with a marker of prior infection.
+Prevalence-type data originates from observations of the number or proportion of a population in a specific \emph{disease state} at a specific time.
+For example, the number of individuals hospitalised with the disease or the proportion of the population with a marker of prior infection.
 The issues associated with incidence-type data often also apply to prevalence-type data.
 
-Prevalence surveys are one form of prevalence-type data that can be unbiased.
-Here, a representative proportion of the population (or a subpopulation of interest) are tested for a marker of current infection.
+Prevalence surveys designed to draw random samples from the population are one form of prevalence-type data that can be unbiased.
+Here, a representative proportion of the population (or a sub-population of interest) are tested for a marker of current infection.
 For many diseases, prevalence is sufficiently low that this is prohibitively expensive to power.

### 2024-03-26 introduction_new.tex
Updates from Overleaf

@@
 
-Prevalence surveys designed to draw random samples from the population are one form of prevalence-type data that can be unbiased.
-Here, a representative proportion of the population (or a sub-population of interest) are tested for a marker of current infection.
+Prevalence surveys designed to draw random samples from the population are one source of prevalence-type data that can be unbiased.
+For instance, a representative proportion of the population (or a sub-population of interest) are tested for a marker of current infection.
 For many diseases, prevalence is sufficiently low that this is prohibitively expensive to power.

### 2024-03-27 MCMC-appendix.tex
Fix refs
finish proof reading

@@
 
-MCMC (Markov chain Monte Carlo) is a class of algorithms for sampling from a distribution where the normalising constant for the distribution is unknown.
+MCMC (Markov chain Monte Carlo) is a class of algorithms which can sample from a distribution where the normalising constant for the distribution is unknown.
 I use it to sample a set of parameters, say $\psi$ (which may be a scalar or vector), from a posterior distribution $p(\psi \mid y)$ where $y$ is the data (again, scalar or vector).
@@
 
-The proposal distribution $q$ is most commonly a (possibly multivariate) normal distribution centred on $\psi$; this scheme is used in \cref{E-SEIR}.
+The proposal distribution $q$ is often a (possibly multivariate) normal distribution centred on $\psi$; this scheme is used in \cref{E-SEIR}.
 Tuning the proposal distribution to propose ``good'' moves is a key part of MCMC.
@@
 
-\subsection{Stability}
+\subsection{Stability} \label{MCMC:sec:stability}
 

### 2024-03-27 cis-perfect-testing.tex
Overleaf: mostly ch 5 corrections
proof ch 6 inc new figures
Updates from Overleaf
finish proof reading

@@
 \propto& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_{j(i)}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 2) \right)
+\label{perf-test:eq:pia}
 \end{align}
@@
 \tau_{\sched_i}(t) &= \min \{ t' \in \sched_i : t' \geq t \} - t
+\label{perf-test:eq:tau-def}
 \end{align}
@@
 Equivalently, $d_j \geq \tau_{\sched_{i(j)}}(b_j) + 1$.
+The third condition can be expressed as $\tau_{\sched_{i(j)}}(b_j) < \infty$.
 
@@
 &= 1 - \prob((B_{j}, E_{j}) \in \Omega_i) \\
-&= \prob(D_j \geq \tau_{\sched_{i}}(B_j)+ 1, B_j > \min \sched_{j}) \\
-&= \sum_{b = \min \sched_{i} + 1}^T \prob(D_j \geq \tau_{\sched_{i}}(b) + 1 \mid B_j = b) \prob(B_j = b)\\
+&= \prob(D_j \geq \tau_{\sched_{i}}(B_j)+ 1, \tau_{\sched_{i(j)}}(b_j) < \infty, B_j > \min \sched_{j}) \\
+&= \sum_{b = \min \sched_{i} + 1}^T \prob(D_j \geq \tau_{\sched_{i}}(b) + 1, \tau_{\sched_{i(j)}}(b_j) < \infty \mid B_j = b) \prob(B_j = b)\\
 &= \frac{1}{T} \sum_{b = \min \sched_{i} + 1}^T S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1)
@@
 
-I implemented the simulation in R~4.2.0~\autocite{R-4-2-0} using tidyverse~2.0.0~\autocite{tidyverse} and targets~1.1.3~\autocite{targetsPackage}.
+I implemented the simulation in R~4.2.0~\autocite{R-4-2-0} using tidyverse~2.0.0~\autocite{tidyverse}.
 

### 2024-03-27 incidence-prevalence.tex
finish proof reading

@@
 In \cref{E-backcalc}, I will need to solve for this system for $\vec{Z}$ given a known $\hat{P} = \E(\vec{P} \mid {Z})$.
-As $\matr{S}$ is lower-triangular, I use forward-substitution~\autocite{cormenMatrix}.
+Assuming $\vec{P}$ and $\vec{Z}$ have the same dimension then $\matr{S}$ is lower-triangular with strictly positive diagonal elements; therefore, $\matr{S}$ is invertible, and the system has a unique solution~\autocite[424]{bernsteinFacts}.
+The lower-triangular structure of $\matr{S}$ makes allows solving the system through forward-substitution, an efficient method that avoids explicit matrix inversion~\autocite{cormenMatrix}.
 Forward-substitution sequentially solves the $t$th element of $\vec{Z}$ ($t = 1, \dots, T$) by calculating:

### 2024-03-27 transmission-appendix-INLA.tex
finish proof reading

@@
 Sparsity is introduced by conditional independence structures of the latent field that are present in many models.
-For example, in the classic random effect setting, all random effects are independent conditional on the hyperparameters.
+For example, in the second-order random walk from \cref{biology-data:sec:MRP}, $w_t$ is independent from the rest of the parameters conditional on $w_{t-1}$, $w_{t-2}$, $w_{t+1}$, $w_{t+2}$.
 
@@
 The joint posterior is not approximated by default.
-I used the one implemented in the R-INLA function \texttt{inla.posterior.sample}, described in \textcite[section 4]{chiuchioloJoint}.
+I use the implementation in the R-INLA function \texttt{inla.posterior.sample}, described in \textcite[section 4]{chiuchioloJoint}.
 

### 2024-03-27 transmission-appendix-phase-type.tex
finish proof reading

@@
 
-Deciding $l$ and the length of time each individual spends in each compartment is the most challenging part of formulating the observation model.
+Deciding $l$ and the length of time each individual spends in each compartment is the most challenging part of formulating the observation model in \cref{E-SEIR:sec:observation}.
 The class of distributions which can be represented as a series of compartments are known as \emph{phase-type} distributions.
@@
 
-A total of five parameters are required to specify a $l$-phase EC distribution.
-\begin{itemize}
-    \item $l$, the number of phases.
-    \item $\omega_Y$, the rate parameter of the Erlang distribution.
-    \item $\omega_{X1}$, the rate parameter of the first state after the Erlang distribution.
-    \item $\omega_{X2}$, the rate parameter of the second state after the Erlang distribution.
-    \item $\pi$, the probability of moving from the end state of the Erlang distribution to the second (as opposed to directly to the absorbing state).
-\end{itemize}
-
 \end{document}

### 2024-03-27 transmission.tex
ch 7 proofing

@@
 
-In this chapter, I use the estimate of the duration of RT-PCR positivity from combining information from the ATACCC and CIS studies (see \cref{E-imperf-test:sec:conclusion}) to investigate the feasibility of estimating incidence and transmission of SARS-CoV-2 from only the CIS data (described in \cref{E-intro:sec:cis}).
+In this chapter, I use the estimate of the distribution of the duration of RT-PCR positivity from combining information from the ATACCC and CIS studies (see \cref{E-imperf-test:sec:conclusion}) to investigate the feasibility of estimating incidence and transmission of SARS-CoV-2 from only the CIS data (described in \cref{E-intro:sec:cis}).
 I will show that the CIS data are sufficient, demonstrating the value of surveys of this type for pandemic surveillance.
@@
 
-First, in \cref{backcalc}, I use a backcalculation approach, based on \cref{E-inc-prev:eq:EPt}.
+I start with the backcalculation (\cref{backcalc}), based on \cref{E-inc-prev:eq:EPt}.
 This approach enables estimation of incidence within (sub)populations of England.
 There are several attractive features about this approach in the context here.
-The backcalculation approach makes few assumptions about how transmission occurs; the assumptions have the potential to introduce bias.
+The backcalculation approach makes few assumptions about how transmission occurs; stronger assumptions increase the potential for introduce bias.
 It can also easily incorporate arbitrary covariates to adjust for non-response bias.
@@
 
-An important consideration in this section is the non-response bias.
-I adopt the poststratification method introduced in \cref{E-biology-data:sec:MRP} to correct for this bias.
-Therefore, incidence must be estimated in each stratum before being poststratified to the target population and subpopulations of interest.
+An important consideration in this section is non-response bias.
+I correct for the non-response through the poststratification method introduced in \cref{E-biology-data:sec:MRP}.
+Poststratification requires estimates of incidence in each stratum, before combining them to the target population and subpopulations of interest.
 The strata are defined by at least age group, region, ethnicity, and sex due to differential non-response in the survey (see \cref{E-intro:sec:cis}).
@@
 This relationship between incidence proportion and prevalence means that once one is inferred, the other is calculated via a deterministic function.
-This gives two possible approaches for estimating incidence from prevalence data.
+This gives two possible approaches for estimating incidence from prevalence data; in \cref{transmission:sec:backcalc-properties} I argue the second approach is more appropriate in this context.
 
@@
 \end{align}
-where $\vec{Z}$ is a vector such that $Z_{i,t}$ is the incidence proportion in stratum $i$ on day $t$, with a joint prior $F_Z$; $P_{i,t}$ is the prevalence in stratum $i$ on day $t$; $\rho$ is the overdispersion parameter of the beta-binomial distribution with prior $F_\rho$; $Y_{i,t}$ is the number of positive tests in stratum $i$ on day $t$; and $n_{i,t}$ is the number of tests in stratum $i$ on day $t$.
+where $\vec{Z}$ is a vector such that $Z_{i,t}$ is the incidence proportion in stratum $i$ on day $t$, with a joint prior $F_Z$; $P_{i,t}$ is the prevalence in stratum $i$ on day $t$; $\rho$ is the overdispersion parameter of the beta-binomial distribution (see \cref{E-distributions}) with prior $F_\rho$; $Y_{i,t}$ is the number of positive tests in stratum $i$ on day $t$; and $n_{i,t}$ is the number of tests in stratum $i$ on day $t$.
 Fitting this model to a prevalence survey such as the CIS (\eg using MCMC) gives a posterior distribution of $\vec{Z}$, the target of inference.
 
-The relationship between $P$ and $Z$ is from \cref{E-inc-prev:eq:EPt}, although dividing both sides by the stratum's population (assumed constant) to have the values in terms of incidence proportion and prevalence rather than incidence and number of prevalent individuals.
+The relationship between $\vec{P}$ and $\vec{Z}$ is from \cref{E-inc-prev:eq:EPt}, although dividing both sides by the stratum's population (assumed constant) to have the values in terms of incidence proportion and prevalence rather than incidence and number of prevalent individuals.
 
@@
 where $F_P$ is analogous to $F_Z$ in providing a joint prior for the prevalence in each stratum and day combination, and all other notation is as in the incidence-inferred approach.
-The first step is to fit this model, giving a posterior distribution for $\vec{P}$.
+The first step in the approach is to calculate the posterior distribution for $\vec{P}$.
 
 The second step is to transform the posterior distribution for $\vec{P}$ to a posterior distribution for $\vec{Z}$.
-A posterior sample for $\vec{P}$ can be transformed to a posterior sample for $\vec{Z}$ based on \cref{inc-prev:eq:forward-substitute}, see \cref{backcalc:sec:methods} for details.
+The transformation is based on \cref{inc-prev:eq:forward-substitute}, which gives a one-to-one mapping between $\vec{P}$ and $\vec{Z}$; implementation details are in \cref{backcalc:sec:methods}.
 
-\subsection{Desirable properties}
+\subsection{Desirable properties} \label{transmission:sec:backcalc-properties}
 
 The incidence-inferred approach is more parsimonious in being a single step procedure.
-However, it poses several pragmatic challenges which means I favour the prevalence-inferred approach.
+However, computational considerations mean I favour the prevalence-inferred approach.
 
@@
         The borrowing should penalize changes in the distribution of incidence/prevalence across strata in the same geographic region.
-        A constant distribution is predicted by the theoretical study of epidemics, for example the dynamics of stratified, mechanistic transmission models (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) over short periods of time with negligible behavioural and immunity changes.
+        A constant distribution is predicted by the theoretical study of epidemics, for example the dynamics of stratified, mechanistic transmission models (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) when behavioural and immunity changes are negligible.
     \item Be computationally efficient.
@@
 
-A parsimonious model satisfying properties 1 and 2 is to specify $\logit(\pi_{i,t}) = g_\varphi(i, t)$, where $g$ is a function parameterised by $\varphi$ and the prior on $\varphi$ is specified to favour $d^2g/dt^2 = 0$, \ie favour linear growth in $\logit(\pi_{i,t})$.
-The prior $F_Z$ or $F_P$ is now a prior on $\varphi$.
+A parsimonious model satisfying properties 1 and 2 is to specify $\logit(\pi_{i,t}) = g_{\vec{\varphi}}(i, t)$, where $g$ is a function parameterised by ${\vec{\varphi}}$ and the prior on ${\vec{\varphi}}$ is specified to favour $d^2g/dt^2 = 0$, \ie favour linear growth in $\logit(\pi_{i,t})$.
+The prior $F_Z$ or $F_P$ is now a prior on ${\vec{\varphi}}$.
 The $\logit$ link function both limits $\pi_{i,t}$ to the interval $[0, 1]$, and logit-linear growth approximates exponential growth for small $\pi_{i,t}$~\autocite{ealesAppropriately}.
 
-The logit transformation also allows property 4 to be conveniently satisfied using an additive model for $g_\varphi$.
+The logit transformation also allows property 4 to be conveniently satisfied using an additive model for $g_{\vec{\varphi}}$.
 An additive model on the logit scale is approximately multiplicative on the natural scale for small values of $\pi_{i,t}$ (which will be common).
@@
 Properties 3 and 5 suggest the use of an approximate method for computing the posterior distribution.
-In particular, using MCMC, sufficient covariates to adjust for the non-response bias, over a period of more than 6--8 weeks leads to a model that is computationally prohibitive ~\citePersonalComms{Koen Pouwels}.
+A particular issue is that using MCMC, sufficient covariates to adjust for the non-response bias, and a period of more than 6--8 weeks leads to a model that is computationally prohibitive ~\citePersonalComms{Koen Pouwels}.
 Instead, previous MRP implementations for the CIS used INLA (see \cref{E-biology-data:sec:MRP}).
@@
 INLA-based inference can only be applied to LGMs (\emph{Latent Gaussian Models}); see \cref{E-transmission:sec:INLA} for further details on INLA and LGMs.
-The LGM framework requires that a transformation of $P_{i,t}$ is linear in a vector of $g_{\varphi}(i, t)$s combining all $i$s and $t$s.
+The LGM framework requires that a transformation of $P_{i,t}$ is linear in $\vec{g} = [g_{{\vec{\varphi}}}(1, 1), \dots, g_{{\vec{\varphi}}}(1, T), \dots, g_{{\vec{\varphi}}}(2, 1), \dots, g_{{\vec{\varphi}}}(I, T)$ where $I$ and $T$ are the number of strata and days respectively.
 
@@
     &= \sum_{k=0}^{\dmax} \pi_{i,t-k} S(k+1) \\
-    &= \sum_{k=0}^{\dmax} \expit(g_\varphi(i, t-k)) S(k+1).
+    &= \sum_{k=0}^{\dmax} \expit(g_{\vec{\varphi}}(i, t-k)) S(k+1).
 \end{align}
@@
 I start with the prevalence posterior from \cref{E-biology-data:sec:positivity-results}, denoting the times the estimates are for as $t = 1, \dots, T$.
-I draw 500 prevalence time series in each stratum from the approximate posterior distribution (using the procedure described in \cref{transmission:sec:INLA:posterior}).
+I draw 500 samples of $\vec{P}$ from the approximate posterior distribution (using the procedure described in \cref{transmission:sec:INLA:posterior}).
 
 Then, I transform this posterior sample using \cref{transmission:alg:calculate-Z}.
-The algorithm is based on \cref{inc-prev:eq:forward-substitute}, which is repeated here for convenience:
+The algorithm is based on the forward-substitution algorithm from \cref{inc-prev:eq:forward-substitute}, which including stratum is:
 \begin{align}
-Z_t
-% &= \frac{P_t - \sum_{i=1}^{t-1} S_{ti} Z_i}{S_{tt}} \\
-&= P_t - \sum_{i=1}^{t-1} S(t - i + 1) Z_i.
+Z_{i,t}
+&= P_{i,t} - \sum_{j=1}^{t-1} S(t - j + 1) Z_{i,j}.
 \label{transmission:eq:forward-substitute}
 \end{align}
-Dividing each side of \cref{transmission:eq:forward-substitute} by the total number of individuals in each stratum transforms the number of prevalent individuals to prevalence and incidence to incidence proportion.
+Dividing each side of \cref{inc-prev:eq:forward-substitute} by the total number of individuals in each stratum transforms the number of prevalent individuals to prevalence and incidence to incidence proportion.
 The algorithm uses a deterministic relationship between incidence and prevalence within a stratum, as \cref{E-inc-prev:sec:observation-process} justifies; \ie, the estimate of the incidence is negligibly affected by approximating the distribution $\vec{P} \mid \vec{Z}$ by its mean.
@@
 For each posterior sample of $\vec{Z}$, I poststratify it to the CIS target population and subpopulations of interest.
-This is done for each posterior sample using the methodology explained in \cref{E-biology-data:sec:MRP}.
-The poststratified posterior samples are now posterior samples from the posterior of interest.
+Specifically, define a (sub) population of interest as a set of strata $\set{S}$.
+Then the incidence proportion in the subpopulation on day $t$ is
+\begin{align}
+Z_t(\set{S})
+&= \frac{\sum_{j \in \set{S}} Z_{j,t} N_j}{\sum_{j \in \set{S}} N_j}
+\label{backcalc:eq:poststratify}
+\end{align}
+where $N_j$ is the population size of stratum $j$.
+The poststratified samples are now from the posterior of interest.
 
@@
 An exception is the North East and East of England regions (see \cref{transmission:fig:backcalc-regions}(C) and (D) respectively); these regions show no consistent increase or decrease in incidence during October.
-All regions then decrease during November; this coincides with the national lockdown which came into force in England on 5 Nov 2020~\autocite{ifgTimeline}.
+All regions then decrease during November; partially coincidng with the national lockdown in force in England between 5 Nov and 2 Dec 2020~\autocite{ifgTimeline}.
 
@@
 The most explored explanation for this is the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocite{walkerTracking,daviesEstimated,lythgoeLineage}.
-There is evidence that when incidence starts increasing is correlated with the timing of the Alpha variant's spread (see \cref{transmission:fig:backcalc-alpha}).
+There is weak evidence that when incidence starts increasing is correlated with the timing of the Alpha variant's spread (see \cref{transmission:fig:backcalc-alpha}).
 A second peak occurs in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
@@
 Therefore, the first day's estimate is unreliable, and has been excluded from all other results.
-However, the rest of the series is robust to this assumption and hence reliable.
+However, the rest of the series is robust to this assumption.
 \begin{figure}
@@
 In particular, the prior places non-zero mass on prevalence values which imply negative incidence under the other model assumptions.
-Rectifying this assumption is challenging without switching to the inference-inferred approach, discussed further in \cref{backcalc:sec:results}.
-
-Alternative inference techniques could allow more powerful modelling approaches, if they allowed a broader set of models than LGMs.
-In particular, the incidence-inferred approach could be adopted.
-This could allow more interpretable priors to be used because the infection process is the process that corresponds to biological and behavioural processes.
-Furthermore, it would prevent the issues with negative incidence.
+Rectifying this assumption is challenging without switching to the inference-inferred approach.
+The inference-inferred approach could also allow more interpretable priors to be used because the infection process is the process that corresponds to biological and behavioural processes.
+The inference-inferred approach, and other more powerful modelling approaches, require an alternative inference technique allowing a broader set of models than LGMs.
 
@@
 % These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
-This method starts with pre-pandemic baseline estimates of $\kappa_{aa'}$ from the POLYMOD study~\autocite{mossongSocial}, and time use from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
+This method starts with pre-pandemic baseline estimates of $\kappa_{aa'}$ from the POLYMOD study~\autocite{mossongSocial}, and time use from the UKTUS (UK Time Use Survey)~\autocite{UKTUS}.
 POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on each contact's age and location (home, work, school, leisure, transport, or other).
@@
 Two hyperparameters will be inferred: $\beta_1$, the initial value of the random walk; and $\sigma_\epsilon$, the standard deviation of the random walk (see \cref{SEIR:sec:priors}).
-$\beta(t)$ is the only time-varying part of the transmission model that is not fixed a priori.
+$\beta(t)$ is the only time-varying fitted parameter.
 Therefore, in practice, it will also absorb other effects such as any error in $\matr{\kappa}$.
@@
 The transmission model is an SEIR model with two latent and two infectious states.
-To form whole of England estimates, each region is modelled independently with the results aggregated, using the poststratification methodology explained in \cref{E-backcalc}.
+To form whole of England estimates, each region is modelled independently with the results aggregated using the poststratification methodology explained in \cref{backcalc:sec:methods}.
 
@@
 \begin{aligned}
-    \vec{\lambda}(t) &= \beta(t) (\matr{\kappa}(t) \circ \matr{q} \circ \matr{N}) ((\vec{i_1}(t) + \vec{i_2}(t))) \\
+    \vec{\lambda}(t) &= (\matr{\kappa}(t) \circ \matr{q}(t) \circ \matr{N}) ((\vec{i}_1(t) + \vec{i}_2(t))) \\
     \frac{d\vec{s}(t)}{dt} &= -\vec{\lambda}(t) \circ \vec{s}(t) \\
-    \frac{d\vec{e_1}(t)}{dt} &= \vec{\lambda}(t) \circ \vec{s}(t) - 2\sigma \vec{e_1}(t) \\
-    \frac{d\vec{e_2}(t)}{dt} &= 2\sigma \vec{e_1}(t) - 2\sigma \vec{e_2}(t) \\
-    \frac{d\vec{i_1}(t)}{dt} &= 2\sigma \vec{e_2}(t) - 2\gamma \vec{i_1}(t) \\
-    \frac{d\vec{i_2}(t)}{dt} &= 2\gamma \vec{i_1}(t) - 2\gamma \vec{i_2}(t) \\
-    \frac{d\vec{r}(t)}{dt} &= 2\gamma \vec{i_2}(t) \\
-    \frac{d\vec{p_0}(t)}{dt} &= \vec{\lambda} \circ \vec{s}(t) - \omega_0 \vec{p_0}(t) \\
-    \frac{d\vec{p_1}(t)}{dt} &= \omega_0 \vec{p_0}(t) - \omega_1 \vec{p_1}(t) \\
-    \frac{d\vec{p_2}(t)}{dt} &= \omega_1 \vec{p_1}(t) - \omega_{2} \vec{p_2}(t) \\
-    \frac{d\vec{p_3}(t)}{dt} &= \chi \omega_{2} \vec{p_2}(t) - \omega_{3} \vec{p_3}(t)
+    \frac{d\vec{e}_1(t)}{dt} &= \vec{\lambda}(t) \circ \vec{s}(t) - 2\sigma \vec{e}_1(t) \\
+    \frac{d\vec{e}_2(t)}{dt} &= 2\sigma \vec{e}_1(t) - 2\sigma \vec{e}_2(t) \\
+    \frac{d\vec{i}_1(t)}{dt} &= 2\sigma \vec{e}_2(t) - 2\gamma \vec{i}_1(t) \\
+    \frac{d\vec{i}_2(t)}{dt} &= 2\gamma \vec{i}_1(t) - 2\gamma \vec{i}_2(t) \\
+    \frac{d\vec{r}(t)}{dt} &= 2\gamma \vec{i}_2(t) \\
+    \frac{d\vec{p}_0(t)}{dt} &= \vec{\lambda} \circ \vec{s}(t) - \omega_0 \vec{p}_0(t) \\
+    \frac{d\vec{p}_1(t)}{dt} &= \omega_0 \vec{p}_0(t) - \omega_1 \vec{p}_1(t) \\
+    \frac{d\vec{p}_2(t)}{dt} &= \omega_1 \vec{p}_1(t) - \omega_{2} \vec{p}_2(t) \\
+    \frac{d\vec{p}_3(t)}{dt} &= \chi \omega_{2} \vec{p}_2(t) - \omega_{3} \vec{p}_3(t)
 \end{aligned}
@@
 \begin{aligned}
-\frac{d\vec{e_1}(t)}{dt} = \psi \vec{e_1}(t) \\
-\frac{d\vec{e_2}(t)}{dt} = \psi \vec{e_2}(t) \\
-\frac{d\vec{i_1}(t)}{dt} = \psi \vec{i_1}(t) \\
-\frac{d\vec{i_2}(t)}{dt} = \psi \vec{i_2}(t) \\
-\frac{d\vec{p_0}(t)}{dt} = \psi \vec{p_0}(t) \\
-\frac{d\vec{p_1}(t)}{dt} = \psi \vec{p_1}(t) \\
-\frac{d\vec{p_2}(t)}{dt} = \psi \vec{p_2}(t) \\
-\frac{d\vec{p_3}(t)}{dt} = \psi \vec{p_3}(t).
+\frac{d\vec{e}_1(t)}{dt} = \psi \vec{e}_1(t) \\
+\frac{d\vec{e}_2(t)}{dt} = \psi \vec{e}_2(t) \\
+\frac{d\vec{i}_1(t)}{dt} = \psi \vec{i}_1(t) \\
+\frac{d\vec{i}_2(t)}{dt} = \psi \vec{i}_2(t) \\
+\frac{d\vec{p}_0(t)}{dt} = \psi \vec{p}_0(t) \\
+\frac{d\vec{p}_1(t)}{dt} = \psi \vec{p}_1(t) \\
+\frac{d\vec{p}_2(t)}{dt} = \psi \vec{p}_2(t) \\
+\frac{d\vec{p}_3(t)}{dt} = \psi \vec{p}_3(t).
 \end{aligned}
@@
 \end{equation}
-This follows directly from the ODEs if $\vec{s}$ is constant.
+This follows directly from \cref{SEIR:eq:fullODEs} if $\vec{s}$ is constant.
 
@@
 Using the results of \cref{E-SEIR:sec:structured-populations}, $\R$ can also be written as $R_{\beta} = \beta_1 D(\matr{\kappa}(t) \circ \matr{q} \circ \matr{N})$ (recall that $D(\matr{\beta})$ denotes the dominant eigenvalue of $\matr{\beta}$).
-Hence setting $\beta_1 = R_\psi / D(\matr{\kappa}(t) \circ \matr{q} \circ \matr{N})$ gives $R_\psi = R_\beta$.
-Using this expression, $\beta_1$ has been reparameterized in terms of $\psi$, $q_c$, and fixed quantities.
+Hence, I set $\beta_1 = R_\psi / D(\matr{\kappa}(t) \circ \matr{q} \circ \matr{N})$ to give $R_\psi = R_\beta$.
 
@@
 \begin{align}
-\vec{i_0} = \frac{i^+}{\max \vec{d}} \vec{d}.
+\vec{i}_0 = \frac{i^+}{\max \vec{d}} \vec{d}.
 \label{SEIR:eq:i0}
@@
 
-Equating the expressions for $d\vec{i_2}/dt$ in \cref{SEIR:eq:fullODEs} and \cref{SEIR:eq:exponential-growth} and setting $t=0$ to allow substituting in \cref{SEIR:eq:i0} gives:
+Equating the expressions for $d\vec{i}_2/dt$ in \cref{SEIR:eq:fullODEs} and \cref{SEIR:eq:exponential-growth} and setting $t=0$ to allow substituting in \cref{SEIR:eq:i0} gives:
 \begin{align}
-    \vec{i_1}(0) &= \vec{i_0} \left(1 + \frac{2\gamma}{\psi + 2\gamma} \right)^{-1}.
-\intertext{By definition of $\vec{i_0}$:}
-    \vec{i_2}(0) &= \vec{i_0} - \vec{i_1}(0).
+    \vec{i}_1(0) &= \vec{i}_0 \left(1 + \frac{2\gamma}{\psi + 2\gamma} \right)^{-1}.
+\intertext{By definition of $\vec{i}_0$:}
+    \vec{i}_2(0) &= \vec{i}_0 - \vec{i}_1(0).
 \end{align}
@@
 \begin{align}
-    \vec{e_2} &= \vec{i_1} \frac{\psi + 2\gamma}{2\sigma} \\
-    \vec{e_1} &= \vec{e_2} \frac{\psi + 2\sigma}{2\sigma} \\
-    \vec{p_0} &= \vec{e_1} \frac{\psi + 2\sigma}{\psi + \omega_0} \\
-    \vec{p_1} &= \vec{p_0} \frac{\omega_0}{\psi + \omega_1} \\
-    \vec{p_2} &= \vec{p_1} \frac{\omega_1}{\psi + \omega_{2}} \\
-    \vec{p_3} &= \vec{p_2} \frac{\pi \omega_{2}}{\psi + \omega_{3}}.
+    \vec{e}_2(0) &= \vec{i}_1(0) \frac{\psi + 2\gamma}{2\sigma} \\
+    \vec{e}_1(0) &= \vec{e}_2(0) \frac{\psi + 2\sigma}{2\sigma} \\
+    \vec{p}_0(0) &= \vec{e}_1(0) \frac{\psi + 2\sigma}{\psi + \omega_0} \\
+    \vec{p}_1(0) &= \vec{p}_0(0) \frac{\omega_0}{\psi + \omega_1} \\
+    \vec{p}_2(0) &= \vec{p}_1(0) \frac{\omega_1}{\psi + \omega_{2}} \\
+    \vec{p}_3(0) &= \vec{p}_2(0) \frac{\pi \omega_{2}}{\psi + \omega_{3}}.
 \end{align}
@@
 Define a parameter $\vec{\pi}$ as the proportion of each stratum without initial immunity, then $\vec{r}(0) = 1 - \vec{\pi}$.
-Finally, to ensure the population proportions sum to one, $\vec{s}(0) = \vec{\pi} - \vec{i_0} - \vec{e_1} - \vec{e_2}$.
+Finally, to ensure the population proportions sum to one, $\vec{s}(0) = \vec{\pi} - \vec{i}_0 - \vec{e}_1 - \vec{e}_2$.
 
@@
 \begin{algorithm}
- Set $\vec{X_0}$ to an initial value of the parameter vector \;
- $\vec{\mu_0} = \vec{X_0}$ \;
+ Set $\vec{X}_0$ to an initial value of the parameter vector \;
+ $\vec{\mu_0} = \vec{X}_0$ \;
  $\matr{\Sigma_0} = \text{diag}(\vec{\mu_0})$ \;
@@
 Priors for the parameters are given in \cref{SEIR:table:priors}.
-The parameters governing the duration of positivity ($\omega$s) were explained previously (see \cref{SEIR:table:ec-params}).
+The parameters governing the duration of positivity were explained previously (see \cref{SEIR:table:ec-params}).
+
 The initial conditions and transition rate parameters are unidentifiable in an SEIR model~\autocite{dankwaStructural}, therefore strongly informative priors or fixed values are used for them.
+
 $\sigma_\epsilon$ and $\rho$ are given priors that favour parsimony.
@@
 The prior on $\rho$ favours a null model of a binomial likelihood.
-The remaining parameters have weakly informative priors.
 
-To form the prior for $\vec{\pi}$, I used the analysis framework of \textcite{amirthalingamSeroprevalence}, applied to seroprevalence data from blood donors~\autocite{amirthalingamSeroprevalence} and a representative sample of children~\autocite{ratcliffeCommunity} in August 2020.
+To form the prior for $\vec{\pi}$, I used the analysis framework of \textcite{amirthalingamSeroprevalence}, applied to Aug 2020 seroprevalence data from blood donors~\autocite{amirthalingamSeroprevalence} and a representative sample of children~\autocite{ratcliffeCommunity}.
 This analysis uses a Bayesian framework to correct the seroprevalence data for the sensitivity and specificity of the serological tests, using an informative prior.
@@
 I approximated the posterior of this proportion with a beta distribution to form the prior for the analysis in this chapter (see \cref{SEIR:table:immunity-prior}).
+
+The remaining parameters have weakly informative priors.
 
@@
 
-I ran a simulation study to check that the inference procedure (\cref{SEIR:sec:inference}) was working correctly and that the model was identifiable.
+I ran a simulation study to study the inference procedure's performance and the model's practical identifiability.
 I generated 100 simulated datasets representing the East of England using the following procedure.
@@
 This check is related to, although simpler than, simulation-based calibration~\autocite{taltsValidating}.
-I constructed a confidence interval for the coverage using the Pearson-Klopper method implemented in the binom package~\autocite{binom1-1}.
+I constructed a confidence interval for the coverage using the Pearson-Klopper method implemented in the binom R package~\autocite{binom1-1}.
 
@@
 If the distributions for length-of-stay in each compartment are known, then the transmission parameters can be estimated.
-This includes the age distribution of the infections and the components of transmission.
 
@@
 
-The decreases in $\beta_w$ over the autumn could be due to increased caution (\eg masking or stricter adherence to social distancing) as cases and deaths increased~\autocite{jarvisEffect} and/or not captured by $\matr{\kappa}(t)$, such as the impact of local measures in the highest transmission areas~\autocite{scottCovid19}.
+The decreases in $\beta_w$ over the autumn could be due to increased caution (\eg masking or stricter adherence to social distancing) as cases and deaths increased~\autocite{jarvisEffect} and/or other effects not captured by $\matr{\kappa}(t)$, such as the impact of local measures in the highest transmission areas~\autocite{scottCovid19}.
 Local measures are not well captured by $\matr{\kappa}(t)$ because the matrix averages behaviour across the whole of England.
@@
 
-Currently, the age distribution of infections is determined by $\matr{\beta}$, and to a lesser extent $\vec{\pi}$, the former of which only contains one parameter that affects the age distribution, $q_c$.
+Currently, the age distribution of infections is determined by $\matr{\beta}$, and to a lesser extent $\vec{\pi}$, the former of which only contains one fitted parameter that affects the age distribution, $q_c$.
 Increasing the number of parameters in $\matr{\beta}$ would be the simplest approach to increase the model's flexibility.
@@
 
-This discussion makes clear that the results are limited by the number of parameters, which is currently limited by the inference algorithm.
+This discussion makes clear that the results are limited by the number of parameters, which is currently limited by the inference algorithm~\citePersonalComms{Sanmitra Ghosh}.
 The MCMC algorithm used tends to struggle with a large number of parameters.

### 2024-03-27 biology-data.tex
finish proof reading
minor corrections / improvements

@@
 Therefore, estimates are sensitive to the choice of data and statistical model.
-Its mean length is around 6 days~\autocite{wuIncubation,quesadaIncubation,aleneSerial}.
+Its mean length is around 6 days~\autocite{wuIncubation,raiIncubation,aleneSerial}.
 
@@
 To start, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution.
-For each posterior draw and day, the incidence proportion in the (sub)population is the weighted average of the incidence proportion in each stratum, where the weights are the number of individuals in that stratum.
+Denote these by $p^(j)_{t,i}$ for the $j$th sample where $i$ is a stratum, \ie some value of $a$, $e$, and $s$.
+Define a (sub) population of interest as a set of strata $\set{S}$.
+Then the prevalence in the (sub)population on day $t$ is
+\begin{align}
+p^{(j)}_t(\set{S})
+&= \frac{\sum_{i \in \set{S}} p_{t,i} N_j}{\sum_{i \in \set{S}} N_i}
+\end{align}
+where $N_j$ is the population size of stratum $j$.
 The population numbers for each stratum are provided by the Office for National Statistics.

### 2024-03-27 introduction.tex
minor corrections / improvements

@@
 Additionally, these models were used to explore policy scenarios.
-For example, the decision at each of stage of lifting the lockdown in 2021 was informed by modelling studies~\autocite{sageEvidence}.
+For example, the decision at each stage of lifting the lockdown in 2021 was informed by modelling studies~\autocite{sageEvidence}.
 These aided decision-makers in understanding plausible impacts of their actions.
@@
 This difference induces a dependency structure between the observations on each day.
-The observation model must now include the with error arising from using a sample of the population rather than (assumed) full recording of AIDS diagnoses.
+The observation model must now include the error arising from using a sample of the population rather than (assumed) full recording of AIDS diagnoses.
 

### 2024-03-27 transmission.tex
minor corrections / improvements

@@
 
-For each posterior sample of $\vec{Z}$, I poststratify it to the CIS target population and subpopulations of interest.
-Specifically, define a (sub) population of interest as a set of strata $\set{S}$.
-Then the incidence proportion in the subpopulation on day $t$ is
+For each posterior sample of $\vec{Z}$, I poststratify it to the CIS target population and subpopulations of interest, using the same basic method as in \cref{E-biology-data:sec:MRP}.
+Specifically, define a (sub)population of interest as a set of strata $\set{S}$.
+Then the incidence proportion in the (sub)population on day $t$ is
 \begin{align}

### 2024-03-27 biology-data.tex
Updates from Overleaf

--- 2024-03-26 biology-data.tex
@@
 Like any diagnostic test, PCR tests are not perfect.
-The quantitative result (the Ct value) is noisy, and the qualitative result (positive or negative) can be wrong.
+The quantitative result (the Ct value) has random variation even if the sample has the same initial amount of DNA or RNA present, and the qualitative result (positive or negative) can be wrong.
 Errors arise from a variety of sources, including the sampling/swabbing process and batch effects in the RT-PCR process~\autocite{hanRTPCR}.

### 2024-03-27 cis-perfect-testing.tex
Updates from Overleaf

@@
 Therefore, it is suitable for estimating the tail of the duration distribution.
-In this chapter, I develop a method to estimate the duration distribution using CIS data, assuming the test has perfect sensitivity and specificity, {\it i.e.} there are no false negatives or positives.
+In this chapter, I develop a method to estimate the duration distribution using CIS data, assuming the test has perfect sensitivity and specificity, \ie there are no false negatives or positives.
 Formally, I aim to estimate the survival function $S(t)$ defined in \cref{inc-prev:sec:definitions}.
@@
 In \cref{perf-test:sec:parameters-priors}, I consider how to parameterize the survival function and choose its priors, including vague and smoothing priors as well as a prior constructed to include the information from \cref{E-ATACCC}.
-I evaluate the model and priors using a simulation study (\cref{perf-test:sec:simulation-study}), showing that they perform well.
+I evaluate the model and priors using a simulation study (\cref{perf-test:sec:simulation-study}), showing that they successfully recover $S(t)$.
 Finally, I discuss the results in \c{perf-test:sec:discussion}.
@@
 
-Any infection episode, whether detected or not, is fully characterised for my purposes by the triplet $(b_j, e_j, i(j))$ where $b_j$ and $e_j$ are the times the episode begin and end respectively, and $i(j)$ is the individual the episode occurred in.
+Any infection episode, whether detected or not, is fully characterised by the triplet $(b_j, e_j, i(j))$ where $b_j$ and $e_j$ are the times the episode begin and end respectively, and $i(j)$ is the individual the episode occurred in.
 The episode's duration is $d_j = e_j - b_j + 1$ (see \cref{E-inc-prev:sec:definitions}).
@@
 &= \sum_{i=1}^{\Ncis} n_{ia} + \sum_{i=1}^{\Ncis} n_{iu} + 0
-\intertext{and, for all $i$,}
-p_{ia} + p_{iu} + p_{i\bar{a}} &= 1.
 \end{align}
+and, for all $i$, $p_{ia} + p_{iu} + p_{i\bar{a}} = 1$.
 
@@
 
-An episode occurs in one of the regions independently and with the same probability.
+These probabilities apply to each episode independently.
 Therefore:
@@
 &= p(\vec{\theta}) \left( \prod_{i \in \set{D}} p_{ia} \right) \left( \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \right) \label{perf-test:eq:posterior2}.
-\end{align}
-For convenience, define the summation term as:
-\begin{align}
-  \eta =  \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}.
+\intertext{For convenience, define the summation term as:}
+\eta &= 
+\sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}.
 \label{perf-test:eq:eta}
@@
 
-I derive an analytical solution to $\eta$ under the prior $\ntot \dist \NBc(\mu, r)$.
+I derive an analytical solution to $\eta$ (defined in \cref{perf-test:eq:eta}) under the prior $\ntot \dist \NBc(\mu, r)$.
 I assume that $\vec{\theta}$, which are only the parameters of the survival distribution, has a prior independent of $\ntot$, the number of infections that occurred in the cohort.
@@
 
-I implemented the simulation in R~4.2.0~\autocite{R-4-2-0} using tidyverse~2.0.0~\autocite{tidyverse}.
+I implemented the simulation in R~4.2.0~\autocite{R-4-2-0} using tidyverse~2.0.0~\autocite{tidyverse} and targets~1.1.3~\autocite{targetsPackage}.
 

### 2024-03-27 introduction_new.tex

--- 2024-03-26 introduction_new.tex
@@
 Estimating transmission refers to inferring this process.
-Understanding components of transmission  (\it{e.g.} the number of contacts each individual makes per day) allow interventions to be targeted.
+Understanding components of transmission  {\it{e.g.}} the number of contacts each individual makes per day) allow interventions to be targeted.
 and to forecast future incidence.
@@
 Whether a given individual is \emph{prevalent}, {\it{i.e.}} has the disease, is often measurable through diagnostic testing.
-Performing diagnostic tests in a sample of the population at a given point in time is known as a \emph{prevalence survey}.
-Prevalence surveys can estimate the population prevalence but are typically very expensive to power appropriately.
+Prevalence surveys, designed  to perform diagnostic testing of a sample of the population,  can be used to estimate the population prevalence,  but are typically very expensive to power appropriately.
 Prevalence and incidence are strongly related, via the duration of an infection (see \cref{E-inc-prev}).
@@
 Collecting data that {\it directly} inform inference of the quantities in this section is challenging or prohibitively expensive .
-Hence, they are typically estimated from other data sources through appropriate inferential methodologies.{\color{red{refer to some papers examples:Phil O'neil, stats in medicine, 1990 on linking models to data; ; De Angelis, Presanis, 2018, Analysing Multiple Epidemic Data Sources.
-"Handbook of
+Hence, they are typically estimated from other data sources through appropriate inferential methodologies.
+{\color{red}{refer to some papers examples:Phil O'neil, stats in medicine, 1990 on linking models to data; De Angelis, Presanis, Analysing Multiple Epidemic Data Sources. "Handbook of
 Infectious Disease Data Analysis", Held, L., Hens, N., O'Neill, P.D. and Wallinga, J. (Eds.).
-Chapman & Hall/CRC, 2019.}}}
+Chapman & Hall/CRC, 2019.}}
 
@@
 
-Data sources to inform outbreak surveillance are often observational in nature.
+Data sources to inform outbreak surveillance are often observational in nature, {\it i.e.} they are not the output of designed studies.
 Therefore, they suffer from biases that can be hard to understand and quantify~\autocite{shadboltChallenges}.
@@
 
-Incidence-type data originate from observations of the consequences of an infection, \eg the development of symptoms or admission to hospital.
-This type of data typically suffers from reporting biases, particularly that the proportion of infections which are detected can vary over time or within different subgroups of the population~\autocites[chapter 9]{lashModern}{shadboltChallenges}.
+Incidence-type data originates from observations of the consequences of an infection, \eg the development of symptoms or admission to hospital.
+This type of data typically suffers from reporting biases, as the proportion of infections which are detected can vary over time or within different subgroups of the population~\autocites[chapter 9]{lashModern}{shadboltChallenges}.
 
-An example of incidence-type data is the number of cases identified through community testing, reliant on individuals presenting for a test.
-However, individuals' ability and willingness to be tested varies over time, which needs to be accounted for in order to perform unbiased inference.
-Unfortunately, data to estimate these changes is normally unavailable.
+An example of incidence-type data is the number of cases identified through community testing, which relies on individuals presenting for a test.
+Individuals' ability and willingness to be tested varies over time, which needs to be accounted for in order to perform unbiased inference.
 
-Data on severe disease is typically less affected by these biases.
+Data on severe disease is typically less affected by biases.
 An extreme example is the development of AIDS following infection by HIV.
@@
 
-Prevalence-type data originates from observations of the number or proportion of a population in a specific \emph{disease state} at a specific time.
-For example, the number of individuals hospitalised with the disease or the proportion of the population with a marker of prior infection.
+Prevalence-type data derives from observations of the number or proportion of a population in a specific \emph{disease state} at a specific time.
+For example, the number of individuals in hospitalwith the disease or the proportion of the population with a marker of prior infection.
 The issues associated with incidence-type data often also apply to prevalence-type data.
 
-Prevalence surveys designed to draw random samples from the population are one form of prevalence-type data that can be unbiased.
-Here, a representative proportion of the population (or a sub-population of interest) are tested for a marker of current infection.
+Prevalence surveys are one source of prevalence-type data that can be unbiased.
+For instance, a representative proportion of the population (or a sub-population of interest) might be tested for a marker of current infection.
 For many diseases, prevalence is sufficiently low that this is prohibitively expensive to power.
 
-However, \emph{seroprevalence} surveys are more feasible.
-Seroprevalence is the proportion of individuals in the population which have antibodies to the disease, indicating that the individual was previously infected~\autocite{cdcSeroprevalence}.
+On the other hand, \emph{seroprevalence} surveys are more feasible.
+Seroprevalence is the proportion of individuals in the population which have antibodies against a particular infection, indicating that the individual was previously infected~\autocite{cdcSeroprevalence}.
 Many more individuals have ever been infected than are currently infected, therefore, seroprevalence surveys can be powered much more cheaply~\autocite{wuSeroprevSimulation}.
@@
 
-\subsection{Cohort studies}
+\subsection{Other studies}
 
 Other epidemiological and/or biological quantities are normally required to estimate transmission.
-These depend on the model type, often including the duration of different disease states or the distribution of the time between infection events~\autocite{wallingaGI,dankwaStructural}.
-Cohort studies (following a group of individuals over time), are a study design which can estimate many of these quantities.
+These include the duration of different disease states or the distribution of the time between infection events~\autocite{wallingaGI,dankwaStructural}.
+Specific studies, {\it e.g.} longitudinal studies that follow  individuals over time, allow  estimation of estimate many of these quantities.
 
@@
 However, recruiting a representative sample of the population is challenging, and the results of these studies are often not generalisable to the whole population.
-Furthermore, these studies often require a large cohort to be appropriately powered.
-Large cohort studies are expensive to run.
+Furthermore, these studies often require a large cohort to be appropriately powered, which are expensive to run.
 
@@
 
-The first component required to estimate incidence and transmission is a model of the infection process, a mathematical description of the process through which incidence arises.
+The first component required to estimate incidence and transmission is a model of the infection process, a mathematical description of the process through which infections are generated.
 A spectrum of models exists, categorised by the strength of assumptions they make about the transmission process~\autocite{beckerCOVIDmodels}.
-Stronger assumptions allow more insight into the transmission process and lower the variance of the estimates but can introduce bias and be computationally challenging to fit.
 
-The models with the weakest assumptions are \emph{phenomenological} models.
-These make almost no assumptions about the transmission process, viewing incidence as a generic counting process (introduced in full in \cref{E-inc-prev:sec:infection-process}).
-The most common model is that incidence follows a Poisson process, and the intensity of the process is the target of inference.
+The models with the weakest assumptions are the so-called \emph{phenomenological} models (add a reference!! Still Becker?).
+These make almost no assumptions about the transmission process, viewing incidence as resulting from a point process (introduced in full in \cref{E-inc-prev:sec:infection-process}).
+A common model is the Poisson process, and the intensity of the process is the target of inference.
 % The use of Poisson processes to model incidence of an infectious disease was very popular in the context of HIV/AIDS\todo{maybe a ref for Poisson in HIV/AIDS?}.
@@
 Next, are semi-mechanistic models.
-These have components of the transmission process in them but do not fully represent the process.
-A common type of semi-mechanistic model is those based on a \emph{renewal equation}~\autocite{bhattSemimechanistic}.
-The renewal equation for an infectious disease~\autocite{fraserRenewal}, in the most general form, is:
+These include components of the transmission process  but do not explicitly describe the process.
+A common type of semi-mechanistic models is that based on a \emph{renewal equation}~\autocite{bhattSemimechanistic} of the most form:
 \begin{align}
@@
 \end{align}
-where $Z(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the mean rate at which an individual infected $\tau$ time units ago infects individuals at calendar time $t$.
+where $Z(t)$ is the expected incidence at time $t$ and $\beta(t, \tau)$ is the expected rate at which an individual infected $\tau$ time units ago infects individuals at calendar time $t$.
 Therefore, the renewal equation relates incidence in the past to incidence in the future.
@@
 % The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
-Extensions of \cref{intro:eq:renewal} that track the number of susceptible individuals become full mechanistic models, equivalent to the compartmental models introduced next~\autocite{champredonEquivalence}.
 Renewal equations do not feature further in this thesis; for further details, including practical considerations when using them as the basis for inference, see \textcite{nashEstimating,gosticPractical}.
@@
 \emph{Compartmental} models, the most common type of mechanistic model, split the population into compartments, each representing a different state of the disease.
-The \emph{dynamics} of the epidemic, how it changes over time, are represented by the transitions between compartments (for more details see \cref{E-SEIR:sec:mechanistic-models}).
+The \emph{dynamics} of the epidemic are represented by the transitions between compartments (for more details see \cref{E-SEIR:sec:mechanistic-models}).
 Simple mechanistic models use a homogeneous population, with individuals differing only due to their disease state, and exponential distributions for the time in each disease state.
-More complex models relax these assumptions, in particular to incorporate heterogeneity.
-The age structure of the population is important for respiratory diseases because individuals meet individuals of their own age more frequently.
+More complex models relax these assumptions to incorporate heterogeneity, for example, to account for the age structure and the interaction between age groups.
 % Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
@@
 
-\section{Disease models} \label{intro:sec:disease-model}
-
-\emph{Disease models} link the transmission model to a latent quantity, $W_t$.
-$W_t$ will then be recorded through some system (\eg a testing programme) described by the observation model (see next section).
-For example, $W_t$ could be the incidence or prevalence of symptomatic individuals on day $t$.
-
+\section{Observational models} \label{intro:sec:disease-model}
+ The infection process is hardly ever directly observed.
+However, as seen in previous sections,  indirect observations are available on the infection's sequelae, which can be linked to the latent infection process through appropriate observational models.
 At the core of many observation models is a \emph{convolution equation} or an approximation of it~\autocite[e.g.][]{brookmeyerBackcalculation,abbottEstimating,birrellRealtimea}.
-This equation describes the expectation of $W_t$, conditional on the incidence prior to time $t$.
+This equation describes the expectation of an infectious sequelae process ({\it e.g.} doctor's consultation, death) $W_t$, conditional on the incidence prior to time $t$.
 In discrete time, it is:
@@
 \end{align}
-where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that ever appear in $W_t$; and $f(i)$ is the probability that an infection at time $t-i$ is included in count $W_t$, conditional on it ever being in one of the counts.
-$f$ is known as a \emph{delay function}.
+where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that result in $W_t$; and, conditional on being observed, $f(i)$ describes the (discretised) probability distribution of the time from infection to observation.
+$f$ is an example of a \emph{delay function}, and  
 $f$ is commonly the pdf of a distribution, or a transformation of it.
@@
 
-\section{Observation models}
-
-Observation models link $W_t$ to a possibly biased and/or noisy observation of it at time $t$.
-
-Biases can arise in a variety of ways, \eg healthcare-seeking behaviour or tests which are less than 100\% accurate.
-Bias can be modelled through additional parameters or data quantifying these issues~\autocite[e.g.][]{sherrattExploring,birrellRealtimea,nicholsonImproving,swallow2022challenges}.
-
-Noise could come from a variety of sources, \eg sampling error, clustering, or unexplained overdispersion.
-Noise is typically modelled by a parametric distribution \eg the negative binomial distribution allowing for overdispersion compared to a Poisson process assumption~\autocite[e.g.][]{birrellRealtimea,frassoBayesian}.
-
 \section{Delay functions}
 
-Applying \cref{intro:eq:disease-model} requires the delay function, $f$.
-Typically, this is either the time until an event or the duration of a disease stage.
-Time series data on $W_t$ cannot normally identify $f$, instead external information needs incorporating~\autocite{swallow2022challenges}.
+Applying \cref{intro:eq:disease-model} requires knowledge of the delay function, $f$.
+Typically, this is either the time to an event or the duration of a disease stage.
+Time series data on $W_t$ cannot identify $f$, instead external information needs incorporating~\autocite{swallow2022challenges}.
+This generally comes from longitudinal studies where the occurrence of the infection's sequelea is monitored in a cohort of individuals ({\it e.g.} Collaborative Group on AIDS Incubation and HIV Survival including the CASCADE EC
+Concerted Action (2000). Time from HIV- 1
+seroconversion to AIDS and death in 13,000 individuals before the widespread use of
+highly active anti-retroviral therapy: effect of age at seroconversion and exposure
+category. Lancet 355, 1131-1137.)
 
-Two approaches are commonly used to estimate the duration of a disease stage: either modelling an underlying biomarker or modelling the duration directly~\autocite[e.g.][]{sweetingEstimating}.
-Modelling a biomarker uses more information per observation but requires stronger assumptions.
+Two approaches are commonly used to estimate the duration of a disease stage: either modelling an underlying biomarker or modelling the duration directly, for instance, in a \emph{survival analysis} framework.~\autocite[e.g.][]{sweetingEstimating}.
 
-The biomarker approach models the trajectory of a biomarker over time in a cohort of individuals.
-An individual is in the disease state while the biomarker is above or below some threshold.
+The biomarker approach models the trajectory of a biomarker over time.
+An individual has an observable/detectable infection while the biomarker is above or below some threshold.
 This approach often requires parametric assumption(s) such as a functional form of the biomarker's trajectory.
-In addition, if individual-level variation is included, which is generally important, a distributional form of random effects is required.
-This requires both sufficient observations of the biomarker per individual and an assumption on the form of the distribution.
+In addition, if individual-level variation is included, which is generally important,  sufficient observations of the biomarker per individual and an assumption on the form of the distribution are required
 More details are in \cref{E-ATACCC}.
 
-Modelling the duration directly considers only the binary result at each test (positive or negative), for instance in a \emph{survival analysis} framework.
+Modelling the duration directly considers only the binary result at each test (positive or negative).
 This approach generally requires fewer parameters reducing the computational burden.
-It is also easier to specify non-parametric forms of the distribution, which is important for SARS-CoV-2 as the tail behaviour may be important.
+It is also easier to specify non-parametric forms of the distribution, which is useful in situation where the tail behaviour may be important.
 % A more detailed review of the literature is in the relevant chapters.
@@
 
-Both cases require longitudinal data, suggesting that cohort studies are the most appropriate study design.
-
 \section{SARS-CoV-2 in England}
 
-The surveillance systems used in response to SARS-CoV-2 are historically unique.
-Modern technology allowed large amounts of data to be collected, distributed, and analysed.
-While the HIV/AIDS pandemic inspired the development of a large amount of statistical methodology, it was far slower moving.
-Much analysis of SARS-CoV-2 data was done at great haste, requiring trade-offs to be made between timeliness and rigour of results presented.
+The response to the SARS-CoV-2 pandemic in the UK has been unprecedented, both in terms of the scientific input to policy and the surveillance data made available.
+Members of the Scientific Pandemic Influenza Group on Modelling Operational sub-group (SPI-M-O ), a sub-committee of the Scientific Advisory Group for Emergencies (SAGE), generated evidence throughout the pandemic to inform policy development~\autocite{medleySPIM,govSPIMO}.
+Each week, SPI-M-O would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
+The committee peer-reviewed these estimates to provide quality assurance and formed a consensus value using a statistical combination of the estimates~\autocite{parkCombining}.
+The consensus value was published on the Government's website and informed its response to the pandemic~\autocite{govRnumber}.
 
-A deep understanding of each data source's potential was not possible in real-time.
-In this thesis, I will investigate the potential of the prevalence surveys in greater detail.
-The disease model will be a prevalence model (defined further in \cref{E-inc-prev:sec:prevalence-process}) with the delay function being the duration of PCR positivity.
+The nowcasting estimates were based on mechanistic and semi-mechanistic models with a wide variety of data sources.
+The mechanistic models, in  particular, commonly incorporated multiple data sources \textcite{birrellEvidence}.
+The models used were reviewed by \textcite{royalSocietyRnumber} (see appendix 2 of that review for details of the specific implementations).
+
+Additionally, these models were used to explore policy scenarios.
+For example, the decision at each of stage of lifting the lockdown in 2021 was informed by modelling studies~\autocite{sageEvidence}.
+These aided decision-makers in understanding plausible impacts of their actions.
 
@@
 
-The main data sources informing the response to the SARS-Cov-2 epidemic in England were based on: community testing (known as pillar 2 testing), the number of severe events, seroprevalence surveys, and prevalence surveys.
+The surveillance systems used in response to SARS-CoV-2 are historically unique. 
+The main data sources informing the response to the SARS-Cov-2 epidemic in England were based on: community testing; severe events; seroprevalence and prevalence surveys.
 Each of these data sources have strengths and weaknesses, see \textcite{royalSocietyRnumber} and references therein for further details on each of these data sources.
@@
 For large parts of the pandemic, community testing was widely available in the UK.
-However, it did not test a representative sample of the population.
-The number of tests conducted varied over time and space, for a variety of known and unknown reasons.
-These include: the policy on who is eligible for testing, the availability of tests, and the willingness of individuals to be tested.
-Therefore, fluctuations in the number of positive tests may not reflect changes in the true incidence of the disease.
-The large number of tests conducted allowed detailed analyses to be performed, including stratification by small geographic areas or age groups, although this does not solve the issue of the biases present.
+However,
+the number of tests carried out varied over time and space, for a variety of reasons, including: policy changes on who was eligible for testing;availability of tests; and willingness of individuals to be tested.
+Therefore, fluctuations in the number of positive tests did not necessarily reflect changes in disease incidence.
+The large number of tests conducted allowed detailed analyses to be performed, including stratification by small geographic areas or age groups, although the lack of interpretability remained.
 
-Severe events counts are generally a more reliable measure of the burden of the disease.
-The criteria for being a severe event are normally well-defined and generally well-recorded in the UK context.
-Therefore, these data are subject to fewer biases than community testing.
-However, SARS-CoV-2 rarely leads to severe events in younger individuals, meaning severe event counts are insensitive to changes in transmission only affecting younger individuals.
+Severe events counts were generally a more reliable measure of the burden of the disease.
+The criteria for being a severe event are normally well-defined and generally well-recorded.
+Therefore, these data were subject to fewer biases than the cases detected through community testing.
+However, SARS-CoV-2 rarely led to severe events in younger individuals, so severe event counts were insensitive to changes in transmission only affecting younger individuals.
 
@@
 
-Seroprevalence surveys were conducted throughout the epidemic, the largest of which was tested donated blood~\autocite{amirthalingamSeroprevalence}.
-In addition to the normal problem of the delay between infection and antibody response, blood donors are a convenience sample whose representativeness of the wider population is unclear.
-
-The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys in representative samples of the general population.
-These were designed to avoid the biases of community testing and observe the whole population.
-Therefore, some modelling is required to borrow information across strata or time to provide stable estimates, as well as to correct for non-response bias (see \cref{E-intro:sec:cis}).
+However, some modelling is required to borrow information across strata or time to provide stable estimates, as well as to correct for non-response bias (see \cref{E-intro:sec:cis}).
 Despite this, they are considered the gold-standard metric.
 
-The use of prevalence surveys for estimating incidence and transmission is the focus of this thesis.
+Seroprevalence surveys were conducted throughout the epidemic, the largest of which was testing donated blood~\autocite{amirthalingamSeroprevalence}.
+In addition to the problem of the delay between infection and antibody response, blood donors are a convenience sample whose representativeness of the wider population is unclear.
+The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys in representative samples of the general population.
+These were designed to avoid the biases of community testing and attempt to observe the general population.
 These surveys use RT-PCR (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
-I focus on the CIS (Coronavirus (COVID-19) Infection Survey), run by the ONS (Office for National Statistics)~\autocite{CIS,cisMethodsONS}.
-The CIS was a longitudinal study, testing a large number of individuals weekly or monthly for SARS-CoV-2; it is described in detail in \cref{E-intro:sec:cis}.
+The first survey was the  Coronavirus (COVID-19) Infection Survey (CIS), run by the Office for National Statistics (ONS)~\autocite{CIS,cisMethodsONS}.
+The CIS was a longitudinal household study, testing a large number of individuals weekly or monthly for SARS-CoV-2 infection (see \cref{E-intro:sec:cis} for a detailed description).
 
-A second prevalence survey ran in the UK, known as REACT (REal-time Assessment of Community Transmission)~\autocite{rileyResurgence,rileyREACT}.
-REACT was a series of cross-sectional surveys, run across 19 rounds between May 2020 and March 2022.
-Each round consisted of at least \numprint{100000} randomly sampled individuals from the UK contributing swabs for testing and questionnaire data~\autocite{elliottTwin}.
+A second prevalence survey, known as the REal-time Assessment of Community Transmission (REACT)~\autocite{rileyResurgence,rileyREACT},
+consisted in a series of cross-sectional surveys, run across 19 rounds between May 2020 and March 2022.
+Each round included at least \numprint{100000} randomly sampled individuals from the UK contributing swabs for testing and questionnaire data~\autocite{elliottTwin}.
 
 A major advantage of the CIS over REACT is that the CIS had continual data collection, while REACT has no test results between rounds.
-The lack of data between rounds meant that using REACT data to infer transmission in these periods requires strong model assumptions.
-A particular instance of this is that a peak in late December 2020 occurred between rounds of the REACT survey, complicating analysis of the survey data at that time~\autocite{rileyREACTround8}.
+This meant that using REACT data to infer transmission in these periods required strong model assumptions.
+A particular instance of this is that a peak in transmission in late December 2020 occurred between rounds of the REACT survey, challenging analysis of the REACT data at that time~\autocite{rileyREACTround8}.
 
-\subsubsection{Cohort studies}
+\subsubsection{Other studies}
 
-Many cohort studies, to estimate a wide variety of quantities, ran throughout the pandemic.
-In this thesis, I make use of two cohort studies with differing designs to estimate the duration of PCR positivity.
-These designs have complementary strengths, and I combine information from the studies.
-Both of the studies recruited individuals from the general population, aiming to observe a cohort of infections that are representative of the infections in the UK community.
-
+Many ad hoc studies to estimate a wide variety of epidemiological and biological quantities of the SARS-CoV2 virus were carried out throughout the pandemic.
 % In addition to population-level surveillance, cohort studies help understand other aspects of the pathogen~\autocite{royalSocietyRnumber}.
 % For example, contact tracing data can inform how likely it is for transmission to occur in different settings and/or between different groups of people.
-The first study I used, ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts)~\autocite{singanayagamCommunity,hakkiOnset}, performed daily testing of a small number of individuals with short follow-up.
-The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability, but the results are limited because the small sample size and short follow-up prohibit estimation of the tails of these distributions.
+One of the study, which  I will use in what follows, is the Assessment of Transmission and Contagiousness of COVID-19 in Contacts (ATACCC) study.
+~\autocite{singanayagamCommunity,hakkiOnset}.
+InATACCC  daily testing of a small number of individuals with short follow-up was performed.
+The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability.
+However, the results from ATACCC are of limited value because of the small sample size and, particularly, of the short follow-up, which  prohibit estimation of the tails of these distributions.
 I describe ATACCC in detail in \cref{E-biology-data:sec:ataccc} and use it to estimate duration in \cref{E-ATACCC}.
@@
 
-Members of SPI-M-O (the Scientific Pandemic Influenza Group on Modelling, Operational sub-group) produced the most prominent modelling of the UK epidemic.
-SPI-M-O was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), generating evidence based on infectious disease modelling for policy development~\autocite{medleySPIM,govSPIMO}.
-This used a wide range of  modelling approaches depending on the data available, which fluctuated throughout the epidemic, and the question being addressed.
-Some were fit to data while others were purely used for simulation based on plausible parameter values.
-
-Each week, members of SPI-M-O would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
-The committee peer-reviewed these estimates to provide quality assurance, removing any estimates unduly affected by issues (\eg data quality issues).
-They then formed a consensus value using a statistical combination of the estimates~\autocite{parkCombining}.
-The consensus value was published on the UK Government's website and informed its response to the epidemic~\autocite{govRnumber}.
-
-The nowcasting estimates were based on mechanistic and semi-mechanistic models with a wide variety of data sources.
-The mechanistic models commonly incorporated multiple data sources, known as \emph{data synthesis}, a technique reviewed by \textcite{birrellEvidence}.
-The models used were reviewed by \textcite{royalSocietyRnumber} (see appendix 2 of that review for details of the specific implementations).
-
-Additionally, these models were used to explore policy scenarios.
-For example, the decision at each of stage of lifting the lockdown in 2021 was informed by modelling studies~\autocite{sageEvidence}.
-These aided decision-makers in understanding plausible impacts of their actions.
-
 \section{Thesis context}
 
-In this thesis, I consider how prevalence surveys could be used for epidemic surveillance, in particular estimates of incidence and transmission.
-These surveys have well-understood designs which means that estimates derived from them are more robust.
-A crucial quantity in their use will be the distribution of the duration of RT-PCR positivity.
+During the SARS-CoV-2 pandemic, analysis of the unprecedently available data was carried out in haste, as results were needed at pace.
+Therefore, a deep understanding of the potential of each data source was not possible.
+In this thesis I investigate the potential of the CIS as a sinlge source of information for incidence and transmission estimation.
 
-In the following subsection, I introduce the most relevant related work, from both prior to the pandemic and during its course.
+In the following subsection, I introduce the most relevant related work, carried out  both prior to the pandemic and during its course.
 
@@
 \emph{Backcalculation} is a modelling framework developed in the context of HIV/AIDS~\autocite{brookmeyerMethod} but applied very broadly.
-It specifies a particular combination of infection, disease, and observation model which relate times of HIV infection with times of AIDS diagnosis via the incubation period.
+It specifies a particular combination of infection and observation model which relate times of HIV infection with times of AIDS diagnosis via the incubation period.
 Further details are in the recent review \textcite{sunModeling}.
@@
 
-The original formulations~\autocite{brookmeyerMethod,rosenbergBackcalculation} used a Poisson process as the infection model; a disease model where all infections are eventually observed ($\zeta = 1$) and the delay function, $f$, is the incubation period distribution's pdf; and an observation model where all diagnoses are observed, and hence has no influence on the data.
+The original formulations~\autocite{brookmeyerMethod,rosenbergBackcalculation} uses a Poisson process as the infection model and a delay function, $f$, the distribution of the time between infection and symptomatic disease (incubation period); and an observation model where all diagnoses are observed.
 
-The disease and observation models must be adapted to apply backcalculation to SARS-CoV-2 prevalence surveys (see \cref{E-inc-prev}).
-The disease model must consider that this is now prevalence-type, rather than incidence-type data.
+In (see \cref{E-inc-prev}) I will adapt this backcalculation framework to data from the CIS, which is of a prevalence-type rather than incidence-type.
 This difference induces a dependency structure between the observations on each day.
@@
 
-An early meta-analysis estimated that individuals test positive for SARS-CoV-2 for a mean of 14.6 days (95\% CI: 9.3--20.0 days)~\autocite{cevikShedding}.
-The longest reported duration was observed at 83 days, although beyond day 9 no live virus was detected, indicating that the individuals testing positive for longer than this were not infectious.
+In the application of the backcalculation to CIS data, the delay function $f$ is represented by the duration of positivity.
+Many attempts have been made to estimate its distribution.
+An initial meta-analysis estimated that individuals test positive for SARS-CoV-2 for a mean of 14.6 days (95\% CI: 9.3--20.0 days)~\autocite{cevikShedding}, with
+the longest reported duration at 83 days, also reporting lack of virus detection beyond day 9, indicating that the individuals testing positive for longer than this were not infectious.
 However, most studies included in the meta-analysis involved hospitalised patients or had unclear inclusion criteria, and hence may not be representative of the general population.
-These summary estimates are insufficient for backcalculation, which requires a distribution of the duration of positivity.
+In addition, summary estimates are insufficient for backcalculation, which requires a distribution of the duration of positivity.
 
@@
 This measure is highly related to the duration of positivity.
-However, the study's cohort consisted only of healthcare workers, \ie working-age adults.
+However, the study's cohort consisted only of healthcare workers, {it i.e.} working-age adults.
 
-\Textcite{binnySensitivity} applied a slight variant of \textcite{hellewellPCRSensitivity}'s methodology to data from New Zealand.
-The situation in New Zealand, who were successfully pursuing a zero-COVID strategy at this time, means that the individuals in the study were unusual.
+\Textcite{binnySensitivity} applied a slight variant of the methodology  in \textcite{hellewellPCRSensitivity} to data from New Zealand.
+The situation in New Zealand, who were successfully pursuing a zero-COVID strategy in the earlier part of the pandemic, means that the individuals in the study were unusual.
 For example, a third were travellers detected during their quarantine period on arrival to New Zealand but would have had negative RT-PCR tests before their departure.
 
+{\color{red} GOT HERE! STILL HAVE A FEW EDITS IN WHAT FOLLOWS}
 \Textcite{ealesCharacterising} estimated the duration of positivity from REACT data.

### 2024-03-27 appendix-plots.tex
add extra gof plots for ATACCC

--- 2024-03-20 transmission-appendix-plots.tex
@@
 
-\chapter{Additional plots for \cref{E-transmission}}
+\chapter{Additional plots}
 
@@
     \thisfloatpagestyle{empty}
-    \vspace{-3cm}
+    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{ATACCC/appendix_fits}}
+  \caption[Remaining ATACCC posterior predictive viral load fits]{
+    Posterior predictive observed viral loads (median and 95\% CrI, lines and ribbon) from the model in \cref{E-ATACCC} for individuals not shown in \cref{E-ATACCC:fig:goodness-of-fits}, overlaid with their observed test results (dots).
+Posterior estimates are conditional on the result not being a false negative.
+Negative RT-PCR tests are plotted at the limit of detection, \ie a log viral load of 0.
+Each plot is labelled with the individual's position in the dataset, an arbitrary value.
+  }
+    \label{ATACCC:fig:appendix-goodness-of-fits}
+\end{figure}
+
+\begin{figure}
+    \thisfloatpagestyle{empty}
     \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_appendix}}

### 2024-03-27 ATACCC/goodness_fit.R

@@ -81,29 +81,43 @@ ggsave(
   dpi = 300
 )
 
+produce_gof_plot = function(individuals) {
+  p_fits = posterior_summary |>
+    filter(i %in% individuals) |>
+    ggplot() +
+    geom_lineribbon(
+      aes(day, yrep, ymin = .lower, ymax = .upper),
+      show.legend = FALSE
+    ) +
+    geom_point(
+      aes(day, y, colour = result),
+      size = 0.7,
+      data = tbl_data |>
+        filter(i %in% individuals)
+    ) +
+    facet_wrap(~i, ncol = 5) +
+    theme_trajectory()
+}
+
 posterior_summary = posterior_predictive |>
   group_by(i, day) |>
   median_qi(yrep) 
 all_individuals = unique(posterior_summary$i)
 selected_individuals = all_individuals[all_individuals > 8 & all_individuals %% 2 == 1]
-p_fits = posterior_summary |>
-  filter(i %in% selected_individuals) |>
-  ggplot() +
-  geom_lineribbon(
-    aes(day, yrep, ymin = .lower, ymax = .upper),
-    show.legend = FALSE
-  ) +
-  geom_point(
-    aes(day, y, colour = result),
-    size = 0.7,
-    data = tbl_data |>
-      filter(i %in% selected_individuals)
-  ) +
-  facet_wrap(~i, ncol = 5) +
-  theme_trajectory()
+appendix_individuals = all_individuals[!all_individuals %in% selected_individuals]
+
 ggsave(
     filename = here::here("ATACCC/fits.pdf"),
-    plot = p_fits,
+    plot = produce_gof_plot(selected_individuals),
+    height = 15,
+    width = 15,
+    units = "cm",
+    dpi = 300
+)
+
+ggsave(
+    filename = here::here("ATACCC/appendix_fits.pdf"),
+    plot = produce_gof_plot(appendix_individuals),
     height = 15,
     width = 15,
     units = "cm",

### 2024-03-28 transmission.tex
add extra gof plots for ATACCC
add implementation details

@@
 The poststratified samples are now from the posterior of interest.
+
+The posterior samples of $\vec{P}$ are produced within the SRS and then extracted.
+I implement the calculation of $\vec{Z}$ and poststratification in R~4.2.0~\autocite{R-4-2-0} using the built in \texttt{forwardsolve} function, tidybayes~3.0.6~\autocite{tidybayes}, and the tidyverse set of packages~\autocite{tidyverse}.
 
@@
 
+\subsubsection{Implementation}
+
+I implemented the simulations study and inference code in Python~3.8.2~\autocite{python} using pandas~2.0.3~\autocite{pandas}, numpy~1.24.4~\autocite{numpy}, and scipy~1.10.1~\autocite{scipy}.
+The strata are larger than those in \cref{backcalc}, allowing the data to be extracted from the SRS.
+Therefore, inference was performed using the University of Cambridge high-performance cluster.
+
 \subsection{Simulation study} \label{SEIR:sec:sim-study}
@@
   \subfile{transmission-appendix-phase-type}
-  \subfile{transmission-appendix-plots}
+  \subfile{appendix-plots}
   \listoftodos

### 2024-03-28 cis-perf-new-ll.tex
ch 7 proofing
new method of likelihood derivation

@@
 \chapter{Alternative derivation of \cref{E-perf-test:eq:multinomial}} \label{perf-test:sec:alt-likelihood}
+
+\section{Attempt 1}
 
@@
     \item $i(j) = i(j')$.
-    \item $\tau_{\sched_{i(j)}}(B_j) = \tau_{\sched_{i(j')}}(B_{j'})$, where $\tau$ is as defined in \cref{E-perf-test:eq:tau-def}.
+    \item $\tau_{\sched_{i(j)}}(B_j) = \tau_{\sched_{i(j')}}(B_{j'})$, where $\tau$ is as defined in \cref{perf-test:eq:tau-def}.
     \item $\tau_{\sched_{i(j)}}(E_j) = \tau_{\sched_{i(j')}}(E_{j'})$.
@@
 
+\section{Attempt 2}
+
+\subsection{Perfect testing}
+
+The test schedule $\sched_i$ is the set of times individual $i$ is tested at, starting from their last test prior to time 1, if it exists, or their time of enrolment otherwise.
+Note that $\sched_i$ includes any tests that occur after $T$.
+
+For a detected episode $j$, we observe $O_j = [l_j^{(b)}, r_j^{(b)}, l_j^{(e)}, r_j^{e}, i(j)]^T$.
+These are: the individual in which the episode occurs, denoted $i(j)$; the earliest and latest time the episode could have begun, $l_j^{(b)}$ and $r_j^{(b)}$ respectively; and the earliest and latest time the episode could have ended, $l_j^{(e)}$ and $r_j^{(e)}$ respectively.
+
+For an undetected episode $j$, there are no observations.
+In which case, define $O(j) = \emptyset$.
+
+Define an integer $D$ and $\set{D} = \vec{\nu}_1, \dots, \vec{\nu}_D$ as the set of all possible observations of detected episodes; that is, $O(j) \in \set{D}$ if and only if $j$ is a detected infection.
+For any $\vec{\nu} = [l^{(b)}, r^{(b)}, l^{(e)}, r^{e}, i]^T$, $\vec{\nu} \in \set{D}$ if and only if the following conditions hold.
+\begin{enumerate}
+  \item $l^{(b)} - 1, r^{(b)}, l^{(e)}, r^{e} + 1 \in \sched_i$, that is individual $i$ was tested at all these times.
+  \item $r^{(b)} = \min\{t \in \sched_i \ssep t \geq l^{(b)}\}$, that is $r^{(b)}$ is the first time at or after $l^{(b)}$ that individual $i$ was tested.
+  \item $r^{(e)} = \min\{t \in \sched_i \ssep t > l^{(e)}\} - 1$, that is $r^{(e)}$ is the day before the next time after $l^{(e)}$ that individual $i$ was tested.
+  \item $1 \leq r^{(b)} \leq T$, that is the episode begins in the period of interest.
+\end{enumerate}
+Let $n_k$ denote the number of observations of $\vec{\nu}_k$, $\nnodet$ the latent number of undetected episodes, $\vec{n} = [n_1, \dots, n_D, \nnodet]^T$, and $\ntot = \sum_{i=1}^D n_i + \nnodet$.
+For $\vec{\nu}_k \in \set{D}$, let $p_k = \prob(O(j) = \vec{\nu}_k \mid \vec{\theta})$, $p_u = \prob(O(j) = \emptyset \mid \vec{\theta})$, and $\vec{p} = [p_1, \dots, p_D, p_u]^T$.
+Assume, for tractability, that the events $O(j) = \vec{\nu}_k$ and $O(j') = \vec{\nu}_{k'}$ are independent for $j \neq j'$; this assumption is discussed in \cref{perf-test:sec:discussion}.
+Then:
+\begin{align}
+  \vec{n} \mid \ntot, \vec{\theta} &\dist \MN(\ntot, \vec{p})
+\intertext{that is:}
+  p(\vec{n} \mid \ntot, \vec{\theta}) &= \frac{\ntot!}{\nnodet!\prod_{k=1}^D n_k!} p_u^{\nnodet} \prod_{k=1}^D p_k^{n_k}.
+  \label{perf-test:eq:multinomial-ll}
+\end{align}
+
+In the CIS data, each $n_k$ is either 0 or 1.
+Let $k \in \set{C}$ indicate that $n_k = 1$.
+Furthermore, note that the support of the multinomial distribution requires that $\nnodet = \ntot - \ndet$.
+Then \cref{perf-test:eq:multinomial-ll} simplifies to:
+\begin{align}
+  p(\vec{n} \mid \ntot, \vec{\theta})
+  &= \frac{\ntot!}{(\ntot - \ndet)!} p_u^{\ntot-\ndet} \prod_{k \in \set{C}} p_k.
+  \label{perf-test:eq:multinomial}
+\end{align}
+
+This is the previous likelihood where each $p_{ia}$ corresponds to a $p_k$ and $p_u$ is unchanged.
+
+\subsection{Imperfect testing}
+
+Just a sketch, probably ignore for now.
+
+\begin{itemize}
+  \item Add $\vec{y}_j$ as the observed test results for episode $j$ to $O(j)$ (as defined previously).
+  \item Modify $p_k$ and $p_u$ as done currently.
+\end{itemize}
+
 \end{document}

### 2024-03-28 cis-perf-new-ll.tex
Updates from Overleaf

--- 2024-03-26 cis-perf-new-ll.tex
@@
 Denote the characteristics of episode $j$ as $\nu_j = (B_j, E_j, i(j))$.
-I now define a series of sets $\set{E}_1, \dots, \set{E}_{N_E}, \set{E}_u$ and an integer $N_E$ such that any possible $\nu_j$ is in exactly one of the classes and each set corresponds to the possible infection episdoes that lead to identical observations (conditional on the test schedules).
+I now define an integer $N_E$ and a series of sets $\set{E}_1, \dots, \set{E}_{N_E}, \set{E}_u$ such that any possible $\nu_j$ is in exactly one of the classes and each set corresponds to the possible infection episodes that would generate identical observations (conditional on the test schedules).
 
@@
 
-Otherwise, the sets are defined such that $\nu_j$ and $\nu_{j'}$ are in the same class if and only if $i(j) = i(j')$ and the episodes beginning and end times are censored to the same interval.
+Otherwise, the sets are defined such that $\nu_j$ and $\nu_{j'}$ are in the same class if and only if $i(j) = i(j')$ and the episodes beginning and end times are censored to be in the same pair of intervals.
 If $i \in \set{D}$ then there is exactly one $k$ such that $\set{E}_k$ is identical to $i$'s admissible region $\alpha_i$.
@@
 Assume, for tractability, that the events $\nu_j \in \set{E}_k$ and $\nu_{j'} \in \set{E}_{k'}$ are independent for $j \neq j'$.
-Denote by $m_k$ then number of episodes in $\set{E}_k$.
+Denote by $m_k$ the number of episodes in $\set{E}_k$.
 $\vec{m} = [m_1, \dots m_{N_E}, m_u]^T$ are the observations.

### 2024-03-28 cis-perfect-testing.tex
Updates from Overleaf

@@
 
-Any infection episode, whether detected or not, is fully characterised by the triplet $(b_j, e_j, i(j))$ where $b_j$ and $e_j$ are the times the episode begin and end respectively, and $i(j)$ is the individual the episode occurred in.
+Any infection episode, whether detected or not, is fully characterised by the triplet $(b_j, e_j, i(j))$ where $b_j$ and $e_j$ are the times the episode begin and end respectively; as $i(j)$ is the individual the episode occurred in it also identifies the test schedule.
 The episode's duration is $d_j = e_j - b_j + 1$ (see \cref{E-inc-prev:sec:definitions}).
 For all $j$, the exact values $b_j$ and $e_j$ are latent.
-If $j$ is undetected then $i(j)$ is also latent and nothing is known about $b_j$ or $e_j$.
+If $j$ is undetected then $i(j)$ is also latent and little is known about $b_j$ or $e_j$.
 If $j$ is detected, then $i(j)$ is known; $b_j \in [l_j^{(b)}, r_j^{(b)}]$; and $e_j \in [l_j^{(e)}, r_j^{(e)}]$.
@@
 Therefore, this is not strictly correct.
-This issue can be rectified be considering the inadmissible region as many subregions each with an observed count of zero (sketched in \cref{perf-test:sec:alt-likelihood}).
-Count zero regions do not contribute to the probability meaning the result is the same.
-A formalisation of this approach requires much more complex notation for the same result; therefore, I prefer this presentation.
+This issue can be rectified be considering the inadmissible region as many subregions, which leads to the same result (sketched in \cref{perf-test:sec:alt-likelihood}).
+A full formalisation of the derivation in \cref{perf-test:sec:alt-likelihood} would require much more complex notation for the same result; therefore, I prefer the presentation to follow.
 
 % Any episode $j$ independently belongs to one of these classes.
-The probabilities $p_{iu}$, $p_{ia}$, and $p_{i\bar{a}}$ are conditional on the episode occurring in individual $i$.
+The probabilities $p_{iu}$, $p_{ia}$, and $p_{i\bar{a}}$ are calculated conditional on the episode occurring in individual $i$.
 I assume the individual in which an episode occurs is iid uniform across the cohort, that is $\prob(i(j) = i) = 1/\Ncis$.
-Therefore, the conditional probabilities are multiplied by $1/\Ncis$ to form the unconditional probabilities.
-For example, the probability of an episode being in class $\alpha_i$ is $p_{ia} / \Ncis$.
+Therefore, the conditional probabilities are multiplied by $1/\Ncis$ to form the probabilities .
+For example, $\prob((B_j, E_j) \in \alpha_i, i(j) = i \mid \vec\theta) = p_{ia} / \Ncis$.
 
@@
 &\propto p(\vec{\theta}) p(\na \mid \vec{\theta}) \\
+&= p(\vec\theta) \sum_{\ntot= \ndet}^{\infty} p(\ntot, \na \mid \vec{\theta}) \\
 &= p(\vec{\theta}) \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) p(\na \mid \ntot, \vec{\theta})
@@
 
-I derive an analytical solution to $\eta$ (defined in \cref{perf-test:eq:eta}) under the prior $\ntot \dist \NBc(\mu, r)$.
-I assume that $\vec{\theta}$, which are only the parameters of the survival distribution, has a prior independent of $\ntot$, the number of infections that occurred in the cohort.
+I derive an analytical solution to $\eta$ (defined in \cref{perf-test:eq:eta}).
+I assume the prior $\ntot \dist \NBc(\mu, r)$, independently of \vec{\theta}, the parameters of the survival distribution.
 Therefore, $p(\ntot \mid \vec{\theta}) = p(\ntot)$.
@@
 Equivalently, $d_j \geq \tau_{\sched_{i(j)}}(b_j) + 1$.
-The third condition can be expressed as $\tau_{\sched_{i(j)}}(b_j) < \infty$.
+The fourth condition can be expressed as $\tau_{\sched_{i(j)}}(b_j) < \infty$.
 

### 2024-03-28 transmission.tex
Updates from Overleaf

--- 2024-03-27 transmission.tex
@@
 
-In this chapter, I use the estimate of the distribution of the duration of RT-PCR positivity from combining information from the ATACCC and CIS studies (see \cref{E-imperf-test:sec:conclusion}) to investigate the feasibility of estimating incidence and transmission of SARS-CoV-2 from only the CIS data (described in \cref{E-intro:sec:cis}).
+In this chapter, I use the estimate of the distribution of the duration of RT-PCR positivity from the combination of information from the ATACCC and CIS studies (see \cref{E-imperf-test:sec:conclusion}) to investigate the feasibility of estimating incidence and transmission of SARS-CoV-2 from only the CIS data (described in \cref{E-intro:sec:cis}).
 I will show that the CIS data are sufficient, demonstrating the value of surveys of this type for pandemic surveillance.
@@
 
-I start with the backcalculation (\cref{backcalc}), based on \cref{E-inc-prev:eq:EPt}.
-This approach enables estimation of incidence within (sub)populations of England.
-There are several attractive features about this approach in the context here.
-The backcalculation approach makes few assumptions about how transmission occurs; stronger assumptions increase the potential for introduce bias.
-It can also easily incorporate arbitrary covariates to adjust for non-response bias.
+I start with the backcalculation (\cref{backcalc}), based on \cref{E-inc-prev:eq:EPt}, which enables estimation of incidence within (sub)populations of England.
+The approach makes few assumptions about how transmission occurs, decreasing the potential to introduce bias and can also easily incorporate arbitrary covariates to adjust for non-response bias.
 The scale of the CIS and desire to correct for non-response bias presents challenges for this methodology, especially within the limited computation of the SRS environment.
@@
 In \cref{SEIR}, I use a mechanistic compartmental model, within the modelling framework introduced in \cref{E-SEIR:sec:mechanistic-models}.
-This approach enables estimating both incidence and components of transmission.
-Estimating transmission in addition to incidence answers important epidemiological questions, \eg whether changes in transmission can be explained by changes in contact rates.
-I develop a novel observation model to allow the transmission model to be fit to the CIS data.
+I estimate both incidence and several components of transmission.
+Estimating transmission from CIS data requires developing a novel observation model.
 
@@
 I show they give similar values, providing reassurance of the robustness of the results to modelling assumptions.
-Overall, there is a bias-variance trade-off between the models.
-Incidence and transmission have not previously been estimated from only SARS-CoV-2 prevalence surveys before.
+% Overall, there is a bias-variance trade-off between the models.
+% Incidence and transmission have not previously been estimated from only SARS-CoV-2 prevalence surveys before.
 
@@
 In this section, I apply the theory introduced in \cref{E-inc-prev} to estimate incidence.
-I view the CIS as a series of daily, cross-sectional prevalence surveys.
-This is the same approach as previously used to estimate prevalence from the CIS (see \cref{E-biology-data:sec:MRP}).
+I view the CIS as a series of daily, cross-sectional prevalence surveys, as previously used to estimate prevalence (see \cref{E-biology-data:sec:MRP}).
 
-An important consideration in this section is non-response bias.
+An important consideration is non-response bias in the CIS study.
 I correct for the non-response through the poststratification method introduced in \cref{E-biology-data:sec:MRP}.
@@
 In \cref{E-inc-prev:sec:observation-process}, I showed that a deterministic approximation to the relationship between true prevalence and incidence is a good approximation for estimating incidence.
-This relationship between incidence proportion and prevalence means that once one is inferred, the other is calculated via a deterministic function.
+This means that once one of the two quantities is inferred, the other can be calculated via a deterministic function.
 This gives two possible approaches for estimating incidence from prevalence data; in \cref{transmission:sec:backcalc-properties} I argue the second approach is more appropriate in this context.
@@
 
-The first approach, which I call the \emph{incidence-inferred} approach, is a single statistical model where incidence is inferred, with prevalence as a derived quantity.
-The basis of the statistical model, in a Bayesian framework, would be as follows.
+The first approach, which I call the \emph{incidence-inferred} approach, specifies a single statistical model including both incidence and prevalence.
+The basis of this statistical model, in a Bayesian framework, would be as follows:
 \begin{align}
@@
 where $\vec{Z}$ is a vector such that $Z_{i,t}$ is the incidence proportion in stratum $i$ on day $t$, with a joint prior $F_Z$; $P_{i,t}$ is the prevalence in stratum $i$ on day $t$; $\rho$ is the overdispersion parameter of the beta-binomial distribution (see \cref{E-distributions}) with prior $F_\rho$; $Y_{i,t}$ is the number of positive tests in stratum $i$ on day $t$; and $n_{i,t}$ is the number of tests in stratum $i$ on day $t$.
-Fitting this model to a prevalence survey such as the CIS (\eg using MCMC) gives a posterior distribution of $\vec{Z}$, the target of inference.
-
+Fitting this model to a prevalence survey such as the CIS gives a posterior distribution of $\vec{Z}$, the target of inference.
 The relationship between $\vec{P}$ and $\vec{Z}$ is from \cref{E-inc-prev:eq:EPt}, although dividing both sides by the stratum's population (assumed constant) to have the values in terms of incidence proportion and prevalence rather than incidence and number of prevalent individuals.
 
-Specifying $F_Z$ would depend on the application of the approach, but should consider the need to provide structure to stabilise the problem (see \cref{E-intro:sec:disease-model}).
+Specifying $F_Z$ should consider the need to provide structure to stabilise the problem (see \cref{E-intro:sec:disease-model}).
 In the context of the CIS, it needs to allow borrowing information across time and/or strata (see \cref{E-biology-data:sec:MRP}).
@@
 The incidence-inferred approach is more parsimonious in being a single step procedure.
-However, computational considerations mean I favour the prevalence-inferred approach.
+However, computational challenges make makes mean I favour the prevalence-inferred approach.
 
 It will be convenient to define a variable equating to the quantity initially estimated.
-Let $\pi_{i,t} = Z_{i,t}$ under the inference-inferred approach and $\pi_{i,t} = P_{i,t}$ is the prevalence.
+Let $\pi_{i,t} = Z_{i,t}$ under the inference-inferred approach and $\pi_{i,t} = P_{i,t}$ under the prevalence-inferred approach.
 
@@
 \begin{enumerate}
-    \item The prior should favour $\pi_{i,t}$ changing exponentially in $t$, \ie penalize changes from exponential growth.
+    \item The prior for $\pi_{i,t}$ should favour $\pi_{i,t}$ changing exponentially in $t$, \ie penalize changes from exponential growth.
         % Penalizing changes from continuous exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time.
@@
         % While incidence proportion can theoretically be greater than 1, this would imply that some individuals were infected multiple times in a day which is not biologically plausible.
-    \item Have sufficient parameters to incorporate all necessary covariates to correct for non-response bias.
+    \item Include sufficient parameters to incorporate all necessary covariates to correct for non-response bias.
     \item Borrow information across the different covariates.
@@
     \item Be computationally efficient.
-        The model will be fit to granular data and therefore needs to be run within the SRS, with its computational limitations (see \cref{E-biology-data:sec:SRS}).
+        The model will be fitted to granular data and therefore needs to be run within the SRS, with its computational limitations (see \cref{E-biology-data:sec:SRS}).
 \end{enumerate}

### 2024-03-28 incidence-prevalence.tex
ch 7 proofing

@@
 The infection process is the mathematical description of how infections occur.
-It is a \emph{point} or \emph{counting} process, where each infection is an event.
+It is a \emph{counting} process, where each infection is an event.
 
-A point process is a random function of continuous time $K(t)$.
+A counting process is a random function of continuous time $K(t)$.
 $K(t)$ must take non-negative integer values and be monotonically increasing in $t$.
 The interpretation of $K(t)$ is the number of events, here infections, which have occurred in the interval $(-\infty, t]$~\autocite[244]{yanDistribution}.
-In the context of a disease, the point process can be discretised into the incidence on day $t$ as $Z_t = K(t) - K(t-1)$, where $t$ is an integer.
+In the context of a disease, the counting process can be discretised into the incidence on day $t$ as $Z_t = K(t) - K(t-1)$, where $t$ is an integer.
 I will explain two approaches to modelling the infection process.
@@
 
-A commonly used point process for infectious disease epidemiology is the time inhomogeneous Poisson process~\autocite[e.g., in the context of HIV,][]{brookmeyerMethod,paganoHIV,rosenbergBackcalculation,brookmeyerBackcalculation}.
+A commonly used counting process for infectious disease epidemiology is the time inhomogeneous Poisson process~\autocite[e.g., in the context of HIV,][]{brookmeyerMethod,paganoHIV,rosenbergBackcalculation,brookmeyerBackcalculation}.
 % A time-homogenous Poisson process is the special case where $\nu(r)$ is constant.
@@
 \end{align}
+where $S(t)$ for $t > \dmax$ is defined to be 0.
 

### 2024-03-28 introduction_new.tex

@@
 However, our best records suggests that, since 1600, they have (on average) occurred every four years with a mean death toll of 5.6 million~\autocite{maraniNovelEpidemics}.
-Recent estimates suggest that the current impact of pandemics is on average of over 2.5 million deaths per year~\autocite{madhavPandemicMortality}.
+Recent estimates suggest that the current impact of pandemics is, on average, over 2.5 million deaths per year~\autocite{madhavPandemicMortality}.
 
@@
 The \emph{effective reproduction number}, $\Re$, is the mean number of secondary infections generated by a single infection at time $t$.
-In particular, it uses the behaviour and immunity status of the population at time $t$.
+In particular, it includes the effects of the behaviour and immunity status of the population at time $t$.
 With constant behaviour and a disease that grants immunity following infection, $\Re$ declines smoothly.
 $\Re$ describes the current dynamics of the epidemic: $\Re > 1$ means the epidemic is growing, $\Re < 1$ means it is shrinking, and $\Re = 1$ means it is stable.
-In simple epidemic models (see \cref{E-SEIR:sec:SIR}), $\Re$ is proportional to the product of the rate of contact, the probability of transmission upon contact, and the proportion of the population that is susceptible.
+In simple epidemic models (see \cref{E-SEIR:sec:SIR}), $\Re$ is proportional to the product of the rate of contacts, the probability of transmission upon contact, and the proportion of the population that is susceptible.
 Interventions act by reducing one of these components \eg masks reduce the probability of transmission and vaccination reduce the susceptible population.
@@
 $\Rc$ is useful when comparing the effects of interventions at different points in time, when the immunity status of the population has changed.
-For example, when comparing the English lockdown in March 2020 and January 2021, there are two effects in play: the number ofindividuals with immunity, higher in 2021; and the different measures already (?) in place when the lockdown was implemented???
-$\Rc$ quantifies the latter effect.
+
+To see the differing uses of the reproduction numbers, consider the difference in the epidemic in England in March 2020, just following the start of the first lockdown, and in January 2021 just following the start of the third lockdown.
+These times have (at least) three epidemiologically relevant differences: (1) in January 2021 the Alpha variant was circulating which is more transmissible than the original variant present in March 2020; (2) the population behaved differently because the exact meaning of ``lockdown'' differed and attitudes had changed; (3) in January 2021, some individuals had immunity due to either prior infection or the start of the vaccination roll-out.
+The original and Alpha variants have different values for $\R$, isolating effect 1.
+$\Rc$ differs during the two periods, including effects 1 and 2.
+$\Re$ includes all these effects, as well as any others.
 
 These definitions ignore many complexities that arise when theoretically defining or practically estimating any of the reproduction numbers~\autocite{pellisEstimation}.
-For example, effects due to seasonality, what a ``mean infection'' means in a heterogeneous population, and the lack of a ground truth for these numbers in almost all situations.
+For example, how to average across seasonal effects and a heterogeneous population, and the lack of a ground truth for these numbers in almost all situations.
 However, their intuitive simplicity and convenient properties in simple epidemic models mean that they are useful concepts for guiding policy.
@@
 If changes in behaviour and susceptibility are negligible, as they often are over short periods of time, epidemics grow (or shrink) exponentially, with a \emph{growth rate} $r$.
-For a given disease and population, an analogous growth rate can be defined for each definition of the reproduction number~\autocite{pellisEstimation,paragGrowthRates,wallingaGI}.
+For a given disease and population, a growth rate can be defined corresponding to each definition of the reproduction number~\autocite{pellisEstimation,paragGrowthRates,wallingaGI}.
 The advantages of using the growth rates are that they are easier to estimate and capture how quickly spread happens in calendar time; however, they do not directly give the strength of intervention required to control the epidemic~\autocite{royalSocietyRnumber}.
@@
 
-Growth rates and reproduction numbers only provide information on whether how an epidemic is changing.
+Growth rates and reproduction numbers only provide information on how an epidemic is changing.
 The size of the epidemic is just as, if not more, important~\autocite{pellisEstimation}.
-\emph{Incidence}, the number of new infections in a given time period  is the primary metric for the magnitude of a pandemic.
-Incidence informs the short-term future burden of the disease, for example, the future number of severe events (hospitalisations or deaths) when combined with severity measures, such as the infection fatality rate (IFR), the proportion of infected individuals who die.
+
+\emph{Incidence} is the number of new infections in a given time period (see \cref{E-inc-prev:sec:definitions} for a formal definition).
+It is the primary metric for the magnitude of a pandemic .
+Incidence informs the short-term future burden of the disease, for example, the future number of severe events (hospitalisations or deaths); although this requires a severity measure, such as the IFR (infection fatality rate), the proportion of infected individuals who die.
 
@@
 Estimating transmission refers to inferring this process.
-Understanding components of transmission  {\it{e.g.}} the number of contacts each individual makes per day) allow interventions to be targeted.
-and to forecast future incidence.
+Understanding this process, \eg the number of contacts each individual makes per day, allows targeting interventions and forecasting future incidence.
 
-\emph{Prevalence} is the proportion of the population with the disease at a given time .
-Whether a given individual is \emph{prevalent}, {\it{i.e.}} has the disease, is often measurable through diagnostic testing.
-Prevalence surveys, designed  to perform diagnostic testing of a sample of the population,  can be used to estimate the population prevalence,  but are typically very expensive to power appropriately.
+\emph{Prevalence} is the proportion of the population with the disease at a given time (see \cref{E-inc-prev:sec:definitions} for a formal definition).
+Whether a given individual is \emph{prevalent}, \ie has the disease, is often measurable through diagnostic testing.
+Prevalence can be estimated through a \emph{prevalence survey}.
+A prevalence survey is a designed study which performs diagnostic testing of a sample of the population; they are typically very expensive to power appropriately.
 Prevalence and incidence are strongly related, via the duration of an infection (see \cref{E-inc-prev}).
@@
 
-Collecting data that {\it directly} inform inference of the quantities in this section is challenging or prohibitively expensive .
-Hence, they are typically estimated from other data sources through appropriate inferential methodologies.
+Collecting data that \emph{directly} inform inference of the quantities in this section is challenging or prohibitively expensive.
+Hence, they are typically estimated from other data sources through appropriate ???? methodologies.
 {\color{red}{refer to some papers examples:Phil O'neil, stats in medicine, 1990 on linking models to data; De Angelis, Presanis, Analysing Multiple Epidemic Data Sources. "Handbook of
@@
 
-Data sources to inform outbreak surveillance are often observational in nature, {\it i.e.} they are not the output of designed studies.
+Data sources to inform outbreak surveillance are often observational in nature, \ie they are not the output of designed studies.
 Therefore, they suffer from biases that can be hard to understand and quantify~\autocite{shadboltChallenges}.
@@
 \emph{Incidence-type} and \emph{prevalence-type} data can inform estimation of incidence and transmission; however, this estimation almost always relies on prior knowledge of epidemiological and/or biological quantities.
-For example, the time between infection and symptom onset. These quantities can be estimated from other studies.
+For example, the time between infection and symptom onset.
+These quantities can be estimated from other studies.
 
@@
 
-Incidence-type data originates from observations of the consequences of an infection, \eg the development of symptoms or admission to hospital.
-This type of data typically suffers from reporting biases, as the proportion of infections which are detected can vary over time or within different subgroups of the population~\autocites[chapter 9]{lashModern}{shadboltChallenges}.
+Incidence-type data originate from observations of the consequences of an infection, \eg the development of symptoms or admission to hospital.
+This type of data typically suffers from reporting biases, such as the proportion of infections detected varying over time or within different subgroups of the population~\autocites[chapter 9]{lashModern}{shadboltChallenges}.
 
@@
 
-Prevalence-type data derives from observations of the number or proportion of a population in a specific \emph{disease state} at a specific time.
-For example, the number of individuals in hospitalwith the disease or the proportion of the population with a marker of prior infection.
-The issues associated with incidence-type data often also apply to prevalence-type data.
+Prevalence-type data originate from observations of the number or proportion of a population in a specific \emph{disease state} at a specific time.
+For example, the number of individuals in hospital with the disease or the proportion of the population with a marker of prior infection.
+The same issues associated with incidence-type data often also apply to prevalence-type data.
 
 Prevalence surveys are one source of prevalence-type data that can be unbiased.
-For instance, a representative proportion of the population (or a sub-population of interest) might be tested for a marker of current infection.
+For instance, testing a representative proportion of the population (or a subpopulation of interest) for a marker of current infection.
 For many diseases, prevalence is sufficiently low that this is prohibitively expensive to power.
 
-On the other hand, \emph{seroprevalence} surveys are more feasible.
+However, \emph{seroprevalence} surveys are more feasible.
 Seroprevalence is the proportion of individuals in the population which have antibodies against a particular infection, indicating that the individual was previously infected~\autocite{cdcSeroprevalence}.
@@
 These include the duration of different disease states or the distribution of the time between infection events~\autocite{wallingaGI,dankwaStructural}.
-Specific studies, {\it e.g.} longitudinal studies that follow  individuals over time, allow  estimation of estimate many of these quantities.
+Specific studies, \eg longitudinal studies that follow  individuals over time, allow  estimation of many of these quantities.
 
@@
 However, recruiting a representative sample of the population is challenging, and the results of these studies are often not generalisable to the whole population.
-Furthermore, these studies often require a large cohort to be appropriately powered, which are expensive to run.
+Furthermore, these studies often require a large cohort to be appropriately powered.
+Large cohort studies are expensive to run.
 
@@
 
-The models with the weakest assumptions are the so-called \emph{phenomenological} models (add a reference!! Still Becker?).
+The models with the weakest assumptions are \emph{phenomenological} models.
 These make almost no assumptions about the transmission process, viewing incidence as resulting from a point process (introduced in full in \cref{E-inc-prev:sec:infection-process}).
@@
 Next, are semi-mechanistic models.
+These relate incidence in the past to current (and future) incidence but do not explicitly represent the susceptible population.
 These include components of the transmission process  but do not explicitly describe the process.
-A common type of semi-mechanistic models is that based on a \emph{renewal equation}~\autocite{bhattSemimechanistic} of the most form:
+A common type of semi-mechanistic models is that based on a \emph{renewal equation}~\autocite{bhattSemimechanistic,fraserRenewal} of the most form:
 \begin{align}
@@
 Simple mechanistic models use a homogeneous population, with individuals differing only due to their disease state, and exponential distributions for the time in each disease state.
-More complex models relax these assumptions to incorporate heterogeneity, for example, to account for the age structure and the interaction between age groups.
+More complex models relax these assumptions to incorporate heterogeneity, \eg through stratification.
 % Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
@@
 \section{Observational models} \label{intro:sec:disease-model}
+
 The infection process is hardly ever directly observed.
-However, as seen in previous sections,  indirect observations are available on the infection's sequelae, which can be linked to the latent infection process through appropriate observational models.
+However, as seen in previous sections, indirect observations are available on the infection's consequences.
+These can be linked to the latent infection process through appropriate observational models.
 At the core of many observation models is a \emph{convolution equation} or an approximation of it~\autocite[e.g.][]{brookmeyerBackcalculation,abbottEstimating,birrellRealtimea}.
@@
 Time series data on $W_t$ cannot identify $f$, instead external information needs incorporating~\autocite{swallow2022challenges}.
-This generally comes from longitudinal studies where the occurrence of the infection's sequelea is monitored in a cohort of individuals ({\it e.g.} Collaborative Group on AIDS Incubation and HIV Survival including the CASCADE EC
+This generally comes from longitudinal studies where the occurrence of the infection's sequelae is monitored in a cohort of individuals ({\it e.g.} Collaborative Group on AIDS Incubation and HIV Survival including the CASCADE EC
 Concerted Action (2000). Time from HIV- 1
@@
 For large parts of the pandemic, community testing was widely available in the UK.
-However,
-the number of tests carried out varied over time and space, for a variety of reasons, including: policy changes on who was eligible for testing;availability of tests; and willingness of individuals to be tested.
+However, the number of tests carried out varied over time and space, and for a variety of reasons.
+Reasons included: policy changes on who was eligible for testing; availability of tests; and willingness of individuals to be tested.
 Therefore, fluctuations in the number of positive tests did not necessarily reflect changes in disease incidence.
@@
 
-However, some modelling is required to borrow information across strata or time to provide stable estimates, as well as to correct for non-response bias (see \cref{E-intro:sec:cis}).
-Despite this, they are considered the gold-standard metric.
-
 Seroprevalence surveys were conducted throughout the epidemic, the largest of which was testing donated blood~\autocite{amirthalingamSeroprevalence}.
 In addition to the problem of the delay between infection and antibody response, blood donors are a convenience sample whose representativeness of the wider population is unclear.
+
 The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys in representative samples of the general population.
@@
 % For example, contact tracing data can inform how likely it is for transmission to occur in different settings and/or between different groups of people.
-One of the study, which  I will use in what follows, is the Assessment of Transmission and Contagiousness of COVID-19 in Contacts (ATACCC) study.
-~\autocite{singanayagamCommunity,hakkiOnset}.
-InATACCC  daily testing of a small number of individuals with short follow-up was performed.
-The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability.
-However, the results from ATACCC are of limited value because of the small sample size and, particularly, of the short follow-up, which  prohibit estimation of the tails of these distributions.
-I describe ATACCC in detail in \cref{E-biology-data:sec:ataccc} and use it to estimate duration in \cref{E-ATACCC}.
+One of the study, which  I will use in what follows, is the ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts) study~\autocite{singanayagamCommunity,hakkiOnset}.
+In ATACCC, daily testing of a small number of individuals with short follow-up was performed.
+These individuals were recruited from contacts of known infections reported to NHS Test and Trace.
+The density of this sampling allows for reliable estimates of how the viral load (the amount of virus in an individual) varies over time and other quantities of interest, most importantly the duration of detectability (how long an individual is RT-PCR positive for).
+However, the results from ATACCC are of limited value because of the small sample size and the short follow-up; prohibiting estimation of how quantities vary over long infections.
+I describe ATACCC in detail in \cref{E-biology-data:sec:ataccc} and use it to estimate the duration of detectability in \cref{E-ATACCC}.
 
@@
 In addition to its role as a prevalence survey, the CIS followed the same individuals over a long period of time (over a year in many cases), collecting longitudinal data.
-The large sample size and long follow-up allows estimation of the tails of the distribution of the duration of positivity.
+The large sample size and long follow-up allows estimation of the tails of the distribution of the duration of detectability.
 However, the lack of frequent testing (at most weekly) prohibits estimation of the viral load dynamics.
@@
 
-\subsection{Models}
-
 \section{Thesis context}
 
-During the SARS-CoV-2 pandemic, analysis of the unprecedently available data was carried out in haste, as results were needed at pace.
+The massive impact of the SARS-CoV-2 pandemic led to data collection on a scale never previously seen for an infectious disease.
+Analysis of this data was carried out rapidly, to provide the timely results needed to inform the ongoing and evolving response.
 Therefore, a deep understanding of the potential of each data source was not possible.
-In this thesis I investigate the potential of the CIS as a sinlge source of information for incidence and transmission estimation.
+In this thesis I investigate the potential of the CIS as a single source of information for incidence and transmission estimation.
 
-In the following subsection, I introduce the most relevant related work, carried out  both prior to the pandemic and during its course.
+In the following subsections, I introduce the most relevant related work, carried out both prior to the pandemic and during its course.
 
@@
 
-The original formulations~\autocite{brookmeyerMethod,rosenbergBackcalculation} uses a Poisson process as the infection model and a delay function, $f$, the distribution of the time between infection and symptomatic disease (incubation period); and an observation model where all diagnoses are observed.
+The original formulations~\autocite{brookmeyerMethod,rosenbergBackcalculation} use a Poisson process as the infection model and a delay function, $f$, representing the pdf of the time between HIV infection and AIDS diagnosis; and a trivial observation model where all infections are (eventually) detected.
 
-In (see \cref{E-inc-prev}) I will adapt this backcalculation framework to data from the CIS, which is of a prevalence-type rather than incidence-type.
-This difference induces a dependency structure between the observations on each day.
-The observation model must now include the with error arising from using a sample of the population rather than (assumed) full recording of AIDS diagnoses.
+In (see \cref{E-inc-prev}) I will adapt this backcalculation framework for use with prevalence survey data, such as the CIS.
+The use of prevalence-type data induces a dependency structure between the values of the latent process $W_1, \dots, W_t$.
+In addition, unlike in the original formulations, not all infections are detected by the CIS as in the CIS a sample of the population receive RT-PCR tests, not the entire population.
+Therefore, I modify the observation model to include both the dependency and the sampling error arising from this process.
+% The observation model must now include the with error arising from using a sample of the population rather than (assumed) full recording of AIDS diagnoses.
 
@@
 
-In the application of the backcalculation to CIS data, the delay function $f$ is represented by the duration of positivity.
-Many attempts have been made to estimate its distribution.
-An initial meta-analysis estimated that individuals test positive for SARS-CoV-2 for a mean of 14.6 days (95\% CI: 9.3--20.0 days)~\autocite{cevikShedding}, with
-the longest reported duration at 83 days, also reporting lack of virus detection beyond day 9, indicating that the individuals testing positive for longer than this were not infectious.
+In the application of the backcalculation to the CIS data, the delay function $f$ represents the duration of detectability.
+Previous work has attempted to estimate $f$.
+An initial meta-analysis estimated that individuals test positive for SARS-CoV-2 for a mean of 14.6 days (95\% CI: 9.3--20.0 days)~\autocite{cevikShedding}.
+The longest reported observed duration was 83 days, although no individuals were infectious beyond day nine.
 However, most studies included in the meta-analysis involved hospitalised patients or had unclear inclusion criteria, and hence may not be representative of the general population.
-In addition, summary estimates are insufficient for backcalculation, which requires a distribution of the duration of positivity.
+In addition, summary estimates are insufficient for backcalculation, which requires a distribution of the duration of detectability.
 
 \Textcite{hellewellPCRSensitivity} estimated the probability of an individual testing positive each day after infection.
-This measure is highly related to the duration of positivity.
-However, the study's cohort consisted only of healthcare workers, {it i.e.} working-age adults.
+This measure is highly related to the duration of detectability.
+However, the study's cohort consisted only of healthcare workers, \ie working-age adults.
 
-\Textcite{binnySensitivity} applied a slight variant of the methodology  in \textcite{hellewellPCRSensitivity} to data from New Zealand.
-The situation in New Zealand, who were successfully pursuing a zero-COVID strategy in the earlier part of the pandemic, means that the individuals in the study were unusual.
+\Textcite{binnySensitivity} applied a slight variant of \textcite{hellewellPCRSensitivity}'s methodology to data from New Zealand.
+The situation in New Zealand, which was successfully pursuing a zero-COVID strategy at the time, means that the individuals in the study were unusual.
 For example, a third were travellers detected during their quarantine period on arrival to New Zealand but would have had negative RT-PCR tests before their departure.
 
-{\color{red} GOT HERE! STILL HAVE A FEW EDITS IN WHAT FOLLOWS}
-\Textcite{ealesCharacterising} estimated the duration of positivity from REACT data.
+\Textcite{ealesCharacterising} estimated the duration of detectability from REACT data.
 Specifically, individuals who tested positive in round 8 of REACT (6th--22 Jan 2021) were invited to take two additional swab tests.
-Defining the time of the initial positive test at time 0, \textcite{ealesCharacterising} estimated the probability of remaining positive at each day after the initial positive test.
+Taking the time of the initial positive test as time 0, \textcite{ealesCharacterising} estimated the probability of remaining positive at each day after the initial positive test.
 They made a strong parametric assumption that the probability decreases exponentially, with a sensitivity analysis assuming a plateau before the decay.
 This assumption is necessary because they had at most 17 days of follow-up, and hence could not estimate the tail of the distribution.
-Overall, they estimate a median duration of 14.0 days (95\% CI: 12.9--15.4 days) with substantial positive correlation with each individual’s viral load at the time of the initial positive test (\ie a higher initial viral load was associated with a longer duration of positivity).
+Overall, they estimate a median duration of 14.0 days (95\% CI: 12.9--15.4 days) with substantial positive correlation with each individual’s viral load at the time of the initial positive test (\ie a higher initial viral load was associated with a longer duration of detectability).
 The individuals with the highest viral load individuals were estimated to have a median duration of 19.6 days (16.1--25.8 days).
 The association could be due to individuals with a higher initial viral load being (on average) more recently infected at the time of the initial positive test.
-The study design prohibits estimating how long before the initial positive test the infection occurred.
-The same issue means the study underestimates the duration of positivity.
+The study design does not provide information on how long before the initial positive test the infection occurred.
+The same issue means the study underestimates the duration of detectability, because the infection's true start time would be prior to the first positive test.
 
-For the purposes of this thesis, I require estimates applicable to the general population.
-There are no reliable estimates of the full distribution of the duration of positivity in a general population cohort available in the literature.
-Therefore, I estimate the duration of positivity from the ATACCC and CIS studies.
+For the purposes of this thesis, I require estimates of duration applicable to the general population.
+As no such estimates of the full distribution of the duration are available in the literature, a key component of this work is devoted to estimating them.
+To do so, I will combine the information from the ATACCC and CIS studies.
 
@@
 
-\Textcite{abbottCISincidence} applied backcalculation to CIS data to estimate the incidence of SARS-CoV-2 in the UK.
-They used a Gaussian process prior to smooth the incidence and stabilise the estimates.
-\Textcite{abbottCISincidence} used \textcite{hellewellPCRSensitivity}'s estimates of the sensitivity of testing to inform the observation model, which are based on a longitudinal study in healthcare workers (see \cref{intro:sec:previous-duration-estimates}).
-They fit to summary statistics of the CIS data, using a daily, independent normal likelihood with parameters estimated from the posterior point estimate and 95\% credible interval published by the ONS.
+\Textcite{abbottCISincidence} applied the backcalculation to CIS data to estimate the incidence of SARS-CoV-2 in the UK.
+They used the summary statistics of prevalence from the CIS  published regularly on the ONS website as data (CHECK), specifically the prevalence's posterior point estimate and bounds of the 95\% credible interval.
+From these summary statistics, they derive a normally-distributed likelihood, independent for each day.
+They used \textcite{hellewellPCRSensitivity}'s estimates of the sensitivity of testing as $f$ in their observation model; these estimates are based on a longitudinal study in healthcare workers (see \cref{intro:sec:previous-duration-estimates}).
+% They fit to summary statistics of the CIS data, using a daily, independent normal likelihood with parameters estimated from the posterior point estimate and 95\% credible interval published by the ONS.
 
-\Textcite{colmanAscertainment} used CIS prevalence estimates to estimate the \emph{ascertainment rate} of community testing (the proportion of all infections that community testing detected).
-Their methodology was similar to \textcite{abbottCISincidence}, including using estimates from \textcite{hellewellPCRSensitivity} for how the sensitivity of RT-PCR testing varies over an infection.
-Rather than using a Gaussian process to provide stability, as \textcite{abbottCISincidence} do, \textcite{colmanAscertainment} assume the ascertainment rate is piecewise constant.
-They do not fully propagate the uncertainty from the CIS prevalence estimates, considering only the bounds of the 95\% confidence intervals the ONS publishes for prevalence.
+\Textcite{colmanAscertainment} used similar methodology to estimate the \emph{ascertainment rate} of community testing (the proportion of all infections that community testing detected).
+% Their methodology was similar to \textcite{abbottCISincidence}, including using estimates of $f$ from \textcite{hellewellPCRSensitivity}.
+% Rather than using a smooth prior to provide stability, as \textcite{abbottCISincidence} do, \textcite{colmanAscertainment} assume the ascertainment rate is piecewise constant.
+% They do not fully propagate the uncertainty from the CIS prevalence estimates, considering only the bounds of the 95\% confidence intervals the ONS publishes for prevalence.
 
@@
 % The length of stay in these compartments (the duration of time positive) is assumed to follow a gamma distribution, with the parameters estimated from external data or having at most one free parameter.
-Their model structure limits the form of the duration of positivity. 
-They fit the models to prevalence survey data using a binomial likelihood where possible, although some approaches take a normal approximation.
+The model structure used in these studies limited the distribution of the duration of detectability to specific classes of distributions.
 % These approaches are similar to how these models are fit to other data sources, such as hospital admissions or deaths.
 
-\Textcite{nicholsonImproving} fit a simple mechanistic model, notably without age structure, but have the most sophisticated observation model.
-They allow an arbitrary distribution of the duration of positivity, based on the estimates of \textcite{hellewellPCRSensitivity} (see \cref{intro:sec:previous-duration-estimates} for discussion of the \textcite{hellewellPCRSensitivity} study).
-Furthermore, they incorporate the sensitivity and specificity of the RT-PCR test into their model.
+\Textcite{nicholsonImproving} estimate prevalence at a small geographic scale, using REACT data to remove the observational biases in community testing data.
+They use a simple mechanistic model, without age structure, but a sophisticated observation model.
+They allow an arbitrary distribution of the duration of detectability, based on the estimates of \textcite{hellewellPCRSensitivity} (see \cref{intro:sec:previous-duration-estimates} for discussion of the \textcite{hellewellPCRSensitivity} study), and incorporate the sensitivity and specificity of the RT-PCR test.
 
 \Textcite{mccabeCISincidence}, adapting the methodology \textcite{ealesAppropriately} developed for REACT, estimate $\Re$ and the growth rate from the CIS data.
-They estimate a posterior distribution of the parameters of a spline by fitting to the prevalence results via a binomial distribution.
-Their set-up penalizes the spline deviating from a constant growth rate.
-They calculate the growth rate from the derivative of the spline, an analytical transformation of the parameters. 
-Since, under reasonable model assumptions, there is a one-to-one correspondence between reproduction numbers and growth rates~\autocite{wallingaGI}, and they assume that the growth rate in incidence and prevalence is equal, this also estimates $\Re$.
+They first estimate prevalence using summary statistics from the CIS, then derive $\Re$ and $r$ using a semi-mechanistic model fitted to their prevalence estimates.
+% They then fit a semi-mechanistic model to the estimating prevalence 
+% a posterior distribution of the parameters of a spline by fitting to the prevalence results via a binomial distribution.
+% Their set-up penalizes the spline deviating from a constant growth rate.
+% They calculate the growth rate from the derivative of the spline, an analytical transformation of the parameters. 
+% Since, under reasonable model assumptions, there is a one-to-one correspondence between reproduction numbers and growth rates~\autocite{wallingaGI}, and they assume that the growth rate in incidence and prevalence is equal, this also estimates $\Re$.
 
-\section{Bayesian inference} \label{intro:sec:Bayes}
+% \section{Bayesian inference} \label{intro:sec:Bayes}
 
-All inference in this thesis is Bayesian.
-Including prior information allows identifiability in situations where this would not otherwise be possible and allows for combining knowledge from multiple sources.
-Many previous studies I build on used Bayesian inference.
+% All inference in this thesis is Bayesian.
+% Including prior information allows identifiability in situations where this would not otherwise be possible and allows for combining knowledge from multiple sources.
+% Many previous studies I build on used Bayesian inference.
 
-Analytically deriving the posterior distribution is impossible for the models in this thesis.
-Therefore, I use computational methods to sample from the posterior distribution.
-The method I use for the majority of the analyses is MCMC (Markov chain Monte Carlo) (see \cref{E-MCMC} for details).
+% Analytically deriving the posterior distribution is impossible for the models in this thesis.
+% Therefore, I use computational methods to sample from the posterior distribution.
+% The method I use for the majority of the analyses is MCMC (Markov chain Monte Carlo) (see \cref{E-MCMC} for details).
 
@@
 
-The overall question that I investigate in this thesis is: can prevalence surveys alone be used to infer incidence and transmission, and if so how?
-Prevalence surveys provide data that is far more representative of the general population than other data sources, therefore, conclusions based on them should be far more robust.
-There are two components to this question.
-First, how can prevalence surveys be designed to help estimate these quantities?
-Second, what statistical methodology can be used to estimate these quantities from prevalence surveys?
-This thesis will focus on the second component.
-However, natural extensions of the work I undertake would inform the first component.
+The overall question that I investigate in this thesis is whether a prevalence survey \emph{alone} is sufficient to infer incidence and transmission and, if so, what methodology can be used.
+% First, how can prevalence surveys be designed to help estimate these quantities?
 
-In order to achieve these aims, I use a case study, specifically the SARS-CoV-2 epidemic in England from Mon 31 Aug 2020 until Sun 25 Jan 2021.
-This case study has several features that make it useful for the aims of this thesis.
-These include that the CIS was of sufficient size and incidence sufficiently high that there is enough data to perform the analyses.
-Additionally, incidence begins rising over this period, which is the period when a surveillance system needs to provide a warning of upcoming issues.
-The CIS began in April 2020, when incidence was falling due to the first lockdown in the UK.
-Incidence remained low and fairly flat in England over summer 2020, until it began rising in mid or late August.
-Finally, the period ends before the roll-out of the vaccine, which likely affects many important aspects of the analysis.
-In particular, the duration of positivity and the susceptibility of the population.
-Including both of these elements would require more complex modelling and including idiosyncratic features of the SARS-CoV-2 pandemic that are less generalisable to other diseases.
+To address this question, I analyse data from the CIS study.
+I use, as a case study, data from the SARS-CoV-2 epidemic in England from Mon 31 Aug 2020 until Sun 25 Jan 2021.
+This period has several interesting features making it suitable for my aim.
+First, the CIS was of sufficient size and transmission was sufficiently high to provide enough data to analyse.
+Second, incidence began to rise over this period, after the
+fall due to the first lockdown.
+Finally, the period ends before the vaccine roll-out, which  affected many important aspects of the analysis; most relevantly, the duration of detectability and the susceptibility of the population.
+Incorporating vaccination effects would require more complex modelling and include idiosyncratic features of the SARS-CoV-2 pandemic that are less generalisable to other diseases.
 
 A secondary aim of this thesis is to improve our understanding of the epidemic in England over this period, and our understanding of the SARS-CoV-2 virus more broadly.
-The results of the analyses I perform in pursuit of the primary aim will naturally contribute to this aim.
+The results of the analyses I perform in pursuit of the primary aim will naturally contribute to this second aim.
 
@@
 
-As its primary aim, this thesis makes several methodological contributions.
+This thesis makes several methodological contributions.
+\begin{enumerate}
+    \item In \cref{E-inc-prev:sec:observation-process}, I provide a theoretic justification for the use of a \emph{deterministic} backcalculation for prevalence surveys.
+A deterministic approach approximates the prevalence at any time as a deterministic function of the incidence, rather than as a random variable.
+    A deterministic backcalculation has previously been used to analyse the CIS data (\eg by \textcite{abbottCISincidence}); however, no justification for the approach was provided.
+    In \cref{E-inc-prev:sec:observation-process} I show that this is a good approximation as long as the proportion of the population sampled is small.
+    In particular, the accuracy of the approximation does not require the prevalence or incidence to be large.
 
-First, in \cref{E-inc-prev:sec:observation-process}, I provide a theoretic justification for the use of deterministic backcalculation in this setting.
-A \emph{deterministic} backcalculation approach was used by \textcite{abbottCISincidence} and I will use it in \cref{E-backcalc}.
-The deterministic approach models the prevalence at any time as a deterministic function of the incidence, rather than as a random variable.
-I show that this is a good approximation as long as the proportion of the population sampled is small.
-In particular, it does not require the prevalence or incidence to be large.
-The estimates I produce in \cref{E-backcalc} fully propagate the uncertainty from the prevalence estimation procedure and use estimates of the duration of positivity from a general population study, unlike prior work.
+    \item In \cref{E-perf-test}, I investigate estimating the duration of detectability using the CIS data.
+    The CIS data are doubly interval censored with arbitrary periods where infections could be undetected.
+    While a statistical framework for this case exists, it had to be generalised to deal with the CIS design.
+    To improve estimation, I propose a novel prior to integrate information from the ATACCC study into this analysis.
 
-In \cref{E-perf-test} I estimate the duration of RT-PCR-positivity from double censored data with arbitrary periods where infections could be undetected (a generalisation of truncation).
-This is based on the CIS data.
-While statistical frameworks have been previously developed which included this case, they had not been used in practice.
-In \cref{E-imperf-test}, I extend this framework to include false negatives.
-
+    \item In \cref{E-imperf-test}, I further extend the framework above to account for false negatives.
     A particular challenge with these analyses is that they must be performed within a trusted research environment with limited computational resources.
 
-Within these analyses, I include information derived from the ATACCC study.
-In order to do so, I propose a prior which appropriately discounts the information from analysing the ATACCC study, which has inappropriate model assumptions when applied to the larger sample size in the CIS data.
+    \item In \cref{E-backcalc}, I use a deterministic backcalculation to estimate incidence with full propagation of the uncertainty from the prevalence estimation.
+    Furthermore, I use the estimate of the duration of detectability from a general population study estimated in previous chapter.
 
-Finally, in \cref{E-SEIR}, I describe a compartmental model for SARS-CoV-2 transmission, which I fit to CIS data.
-I allow a more flexible distribution of the duration of positivity than any of the papers discussed above, except \textcite{nicholsonImproving}, incorporating the skew of the distribution.
-This distribution is based on data from a representative sample of the UK population, rather than convenience samples (see \cref{intro:sec:previous-duration-estimates}).
-Unlike \textcite{nicholsonImproving}, I include age-stratification in the transmission model, allowing estimation of how transmission varies by age.
-I use only data from the CIS and show this is sufficient for estimating incidence and transmission.
+    \item In \cref{E-SEIR}, I develop a compartmental model for SARS-CoV-2 transmission, relying \emph{only} on CIS data. 
+    This uses a distribution of duration of detectability derived by translating the estimates in \cref{E-imperf-test} into a progression through model states via an Erlang-Coxian formulation.
+I include age-stratification in the transmission model, allowing estimation of age-specific transmission.
+    I show the CIS data are sufficient for estimating incidence and transmission.
+\end{enumerate}
 
@@
 
-In \cref{E-imperf-test}, I present estimates of the duration of RT-PCR-positivity in the general population.
+In \cref{E-imperf-test}, I generate estimates of the duration of RT-PCR-positivity in the general population.
 These estimates could inform public health policy, for example, in interpreting RT-PCR test results or formulating quarantine policies.
 
-In \cref{E-transmission}, I present estimates of the incidence and transmission of SARS-CoV-2 in England over the period.
+In \cref{E-transmission}, I estimate incidence and transmission of SARS-CoV-2 in England over the period WHICH.
 These estimates contribute to the literature reconstructing the SARS-CoV-2 epidemic in England.
-Understanding the dynamics of the past infection is important for informing the response to future pandemics, and the future burden of SARS-CoV-2.
+Understanding the dynamics of past infection is important for informing the response to future pandemics, and the future burden of SARS-CoV-2.
 
@@
 
-In order to best fulfil this thesis's primary aim, the methodology developed should be easy for others to use.
-Therefore, the code for the analyses in this thesis is available online.
+To carry out the work in this thesis I developed bespoke code. To ensure the methodology developed can be easily adopted by others, the code used for the analyses is available online.
 An index to the code and, where possible, data are available at \url{https://github.com/joshuablake/thesis-index}.
@@
 First, the framework to simulate the CIS (used in \cref{E-imperf-test:sec:simulate,E-perf-test:sec:simulation-study}).
-The package enables customisation of various features of the study, for example the testing regime of the participants, the sensitivity of the tests, and the duration of positivity.
+The package enables customisation of various features of the study, for example the testing regime of the participants, the sensitivity of the tests, and the duration of detectability.
 This package would be particularly useful in answering questions regarding the design of future prevalence surveys, for example, how frequently individuals should be tested, or how many individuals should be tested.
 
-Second, the survival analysis framework used to infer the duration of positivity in \cref{E-perf-test,E-imperf-test}.
+Second, the survival analysis framework used to infer the duration of detectability in \cref{E-perf-test,E-imperf-test}.
 The framework allows the user to specify all the different analyses performed in these chapters in a simple interface. 
@@
 Almost all code, excluding dependencies, was written from scratch for these analyses.
-The exceptions are the Markov chain Monte Carlo sampler used in \cref{E-SEIR}, the model for prevlaence used in \cref{E-backcalc}, and the original viral load model used in \cref{ATACCC:sec:hakki}.
+The exceptions are the Markov chain Monte Carlo sampler used in \cref{E-SEIR}, the model for prevalence used in \cref{E-backcalc}, and the original viral load model used in \cref{E-ATACCC:sec:hakki}.
 
@@
 
-In \cref{E-biology-data}, I provide background information on the natural history of SARS-CoV-2, the RT-PCR testing process used to produce the data used throughout this thesis, and describe the two cohort studies used in this thesis.
-Then, in \cref{E-inc-prev}, I formally describe the stochastic processes which generate infections, prevalence, and the observations of prevalence.
-A key insight in this chapter is that the distribution of the duration of positivity is the key quantity linking incidence and prevalence.
+In \cref{E-biology-data}, I provide background information on: the natural history of SARS-CoV-2, the RT-PCR testing process used to produce the data analysed throughout this thesis, and describe the information available in the CIS and ATACCC studies.
 
-I then turn to estimating this distribution.
-I start, in \cref{E-ATACCC}, by estimating the duration of positivity from the ATACCC study.
-This provides a reliable estimate of the bulk of the distribution (out to around 20 days) but shows that there is a significant tail.
-Estimating the tail requires a large sample size and long follow-up. 
-Therefore, I turn to the CIS data in \cref{E-perf-test,E-imperf-test}.
-In \cref{E-perf-test}, I provide a framework for estimating the duration of positivity from the CIS data, assuming perfect tests.
-However, the assumption of perfect tests is not realistic; in \cref{E-imperf-test}, I extend the framework to include false negatives.
+In \cref{E-inc-prev}, I formally describe the stochastic processes which generate infections, prevalence, and the observations of prevalence.
+I show the distribution of the duration of detectability is the key quantity linking incidence and prevalence.
 
-Having estimated the duration of positivity, next I estimate incidence and transmission from the CIS data in \cref{E-transmission}, thus fulfilling the primary aim of this thesis.
+In \cref{E-ATACCC}, I estimate the duration of detectability from the ATACCC study.
+This provides a reliable estimate of the bulk of the distribution (out to around 20 days),  but shows that estimating the tail requires a large sample size and long follow-up.
+
+In \cref{E-perf-test}, I provide a framework for estimating the duration of detectability assuming perfect tests from prevalence surveys such as the CIS.
+However, as the assumption of perfect tests is not realistic; in \cref{E-imperf-test}, I extend the framework to include false negatives.
+
+Having estimated the duration of detectability, I estimate incidence and transmission from the CIS data in \cref{E-transmission}.
 I use both a phenomenological and mechanistic transmission model, contrasting their results and discussing the advantages and disadvantages of each in this context.
-Finally, I conclude in \cref{E-conclusion}.
 
-\Cref{E-distributions} provides definitions of the probability distributions used in this thesis.
-Other appendices provide additional details.
+Finally, I conclude in \cref{E-conclusion} with a discussion of the limitations of this work and ideas for future research.
+
+Appendices provide definitions of the probability distributions used in this thesis, further details, and additional results.
 

### 2024-03-29 cis-perf-new-ll.tex
More on new version

@@
 
-For a detected episode $j$, we observe $O_j = [l_j^{(b)}, r_j^{(b)}, l_j^{(e)}, r_j^{e}, i(j)]^T$.
+For any episode $j$, define $O(j)$ as the observations of that episode.
+Test results are deterministic based on the time of the test and the individual's infection status.
+Therefore, $O(j)$ is fully determined by the triplet $(B_j, E_j, i(j))$, $j$'s beginning time, end time, and the individual in which the episode occurs respectively.
+
+For a detected episode, we have $O(j) = [l_j^{(b)}, r_j^{(b)}, l_j^{(e)}, r_j^{e}, i(j)]^T$ (see \cref{E-biology-data:sec:cis-episodes}).
 These are: the individual in which the episode occurs, denoted $i(j)$; the earliest and latest time the episode could have begun, $l_j^{(b)}$ and $r_j^{(b)}$ respectively; and the earliest and latest time the episode could have ended, $l_j^{(e)}$ and $r_j^{(e)}$ respectively.
 
-For an undetected episode $j$, there are no observations.
-In which case, define $O(j) = \emptyset$.
+For an undetected episode $j$, there are no observations; in which case, define $O(j) = \emptyset$.
 
-Define an integer $D$ and $\set{D} = \vec{\nu}_1, \dots, \vec{\nu}_D$ as the set of all possible observations of detected episodes; that is, $O(j) \in \set{D}$ if and only if $j$ is a detected infection.
-For any $\vec{\nu} = [l^{(b)}, r^{(b)}, l^{(e)}, r^{e}, i]^T$, $\vec{\nu} \in \set{D}$ if and only if the following conditions hold.
+Define an integer $N_D$ and $\set{D} = \vec{\nu}_1, \dots, \vec{\nu}_{N_D}$ as the set of all possible observations of detected episodes; that is, $O(j) \in \set{D}$ if and only if $j$ is a detected infection.
+For any $\vec{\nu} = [l^{(b)}, r^{(b)}, l^{(e)}, r^{e}, i]^T$, $\vec{\nu} \in \set{D}$ if and only if all the following conditions hold.
 \begin{enumerate}
   \item $l^{(b)} - 1, r^{(b)}, l^{(e)}, r^{e} + 1 \in \sched_i$, that is individual $i$ was tested at all these times.
+  \item $l^{(b)} \leq r^{(b)} \leq l^{(e)} \leq r^{(e)}$.
   \item $r^{(b)} = \min\{t \in \sched_i \ssep t \geq l^{(b)}\}$, that is $r^{(b)}$ is the first time at or after $l^{(b)}$ that individual $i$ was tested.
@@
 \end{enumerate}
-Let $n_k$ denote the number of observations of $\vec{\nu}_k$, $\nnodet$ the latent number of undetected episodes, $\vec{n} = [n_1, \dots, n_D, \nnodet]^T$, and $\ntot = \sum_{i=1}^D n_i + \nnodet$.
+Let $n_k$ denote the number of observations of $\vec{\nu}_k$, $\nnodet$ the latent number of undetected episodes, $\vec{n} = [n_1, \dots, n_D, \nnodet]^T$, and $\ntot = \sum_{i=1}^{N_D} n_i + \nnodet$.
 For $\vec{\nu}_k \in \set{D}$, let $p_k = \prob(O(j) = \vec{\nu}_k \mid \vec{\theta})$, $p_u = \prob(O(j) = \emptyset \mid \vec{\theta})$, and $\vec{p} = [p_1, \dots, p_D, p_u]^T$.
@@
 \intertext{that is:}
-  p(\vec{n} \mid \ntot, \vec{\theta}) &= \frac{\ntot!}{\nnodet!\prod_{k=1}^D n_k!} p_u^{\nnodet} \prod_{k=1}^D p_k^{n_k}.
+  p(\vec{n} \mid \ntot, \vec{\theta}) &= \frac{\ntot!}{\nnodet!\prod_{k=1}^{N_D} n_k!} p_u^{\nnodet} \prod_{k=1}^{N_D} p_k^{n_k}.
   \label{perf-test:eq:multinomial-ll}
@@
 
-In the CIS data, each $n_k$ is either 0 or 1.
+In the CIS data, each $n_k$ ($k \neq u$) is observed as either 0 or 1.
 Let $k \in \set{C}$ indicate that $n_k = 1$.
@@
 
+Introducing false negatives means that $O(j)$ is now a random variable because the test results between the start and the end of the episode could contain false negatives.
+Therefore, the
+
 Just a sketch, probably ignore for now.

### 2024-03-29 cis-perfect-testing.tex
More on new version
proof

@@
 I consider the cohort of $\Ncis = \numprint{437590}$ individuals who could have had a detected episode; this is a subset of the whole CIS cohort.
-To be considered, an individual must have had at least three tests before time $T$ (a negative prior to the episode, a positive, and a negative following the episode).
+To be considered, an individual must have at least one test between time 1 and $T$.
 I index these individuals by $i = 1, \dots, \Ncis$.
@@
 I derive an analytical solution to $\eta$ (defined in \cref{perf-test:eq:eta}).
-I assume the prior $\ntot \dist \NBc(\mu, r)$, independently of \vec{\theta}, the parameters of the survival distribution.
+I assume the prior $\ntot \dist \NBc(\mu, r)$ is independent of $\vec{\theta}$, the parameters of the survival distribution.
 Therefore, $p(\ntot \mid \vec{\theta}) = p(\ntot)$.
@@
 
-Step 3 requires a distribution of the duration of positivity.
+Step 3 requires a distribution of the duration of detectability.
 I base this assumption on the estimate from \cref{E-ATACCC} with an inflated tail to be consistent with the CIS.
@@
 These curves and the combined curve are compared in \cref{perf-test:fig:duration-dist}.
-Individual $i$'s duration of positivity, $D_i$, is then an independent draw from this distribution.
+Individual $i$'s duration of detectability, $D_i$, is then an independent draw from this distribution.
 \begin{figure}

### 2024-03-29 conclusion.tex
proof ch 8
proof

@@
 Pandemics will continue to pose a significant threat to public health for the foreseeable future.
-Therefore, there is a public health need to respond to pandemics.
-Such a response needs to be informed by monitoring.
-In this thesis, I have investigated one method of monitoring pandemics: prevalence surveys.
+Therefore, there is a need for a public health response to pandemics, informed by surveillance.
+In this thesis, I have investigated the potential of prevalence surveys, using the CIS as an example, to provide this information.
 
@@
 
-In \cref{E-intro:sec:aims}, I posed the central question for this thesis: can prevalence surveys alone be used to infer incidence and transmission, and if so how?
+In \cref{E-intro:sec:aims}, I posed the central question for this thesis: can a prevalence survey \emph{alone} be sufficient to infer incidence and transmission and, if so, what method(s) are optimal?
 
-In \cref{E-transmission}, I found that this it is possible, developing two models that can infer incidence and transmission from prevalence surveys.
+In \cref{E-transmission}, I found that this it is possible by developing two models that can infer incidence and transmission from prevalence surveys.
+These did both rely on knowledge of the distribution of the duration of detectability, which can partially be inferred from the CIS.
 One model was phenomenological in nature and the other mechanistic.
@@
 The mechanistic approach, based on an SEIR-type compartmental model, reduces the uncertainty and allows estimation of the components of transmission by making stronger assumptions about the infection process.
-Both sets of results highlighted the heterogeneity across regions and age groups in England.
 The choice of modelling approach should depend on the situation.
-The phenomenological approach is more robust when there are high levels of uncertainty; for example, early in an outbreak or in the presence of new variants the infectious period duration may not be known.
-However, the mechanistic approach provides more information.
+A phenomenological approach is more robust when there are high levels of uncertainty; for example, early in an outbreak or in the presence of new variants the infectious period duration may not be known.
+On the other hand, a mechanistic approach provides more information.
 Where possible, both methods should be implemented; discrepancies between results may highlight deficiencies in modelling assumptions.
 
-The work in \cref{E-transmission} relied on the theoretical work of \cref{E-inc-prev}.
-This work developed the backcalculation framework for the context of prevalence surveys for SARS-CoV-2.
-The major insight was the justification of the deterministic model framework and the importance of the distribution of the duration of positivity.
+The work in \cref{E-transmission} relied on the theoretical work of \cref{E-inc-prev}, adapting the backcalculation framework for the context of prevalence surveys for SARS-CoV-2.
+The major insight was the justification of a deterministic backcalculation and the importance of the distribution of the duration of detectability.
 
-These models both required knowledge of the distribution of the duration of positivity.
+Estimation of incidence and/or transmission required knowledge of the distribution of the duration of detectability.
 Estimating this distribution was the focus of most of this thesis.
-In \cref{E-perf-test,E-imperf-test}, I showed how this quantity could be estimated the CIS prevalence survey.
-This did require external information from the ATACCC study (the results of \cref{E-ATACCC}), demonstrating that the CIS's design is not sufficient for estimating transmission.
-However, future surveys may be able to avoid this issue (see \cref{conclusion:sec:future-work}).
+In \cref{E-perf-test,E-imperf-test}, I showed how this quantity could be estimated from the CIS prevalence survey.
+This did require external information from the ATACCC study and prior knowledge on the number of infections, demonstrating that some additional studies are required in addition to the CIS, at least with its current design.
 
-Prior to this thesis, the duration of positivity in the general population was not well understood (see \cref{E-intro:sec:previous-duration-estimates}).
+Prior to this thesis, the duration of detectability in the general population was not well understood (see \cref{E-intro:sec:previous-duration-estimates}).
 A particularly understudied area was quantifying the shape and size of the tail of the distribution, due to limited follow-up of individuals who test positive.
@@
 Overall, I have found that representative prevalence surveys are a promising approach for pandemic surveillance.
-A well-designed survey is sufficient to estimate all the quantities of interest (as described in \cref{E-intro:sec:metrics}), except severity.
+A well-designed survey is sufficient to estimate incidence and transmission, although not currently the duration of detectability.
 
 \section{Future work} \label{conclusion:sec:future-work}
+
+\subsection{Joint modelling}
 
@@
 
-A common theme in this thesis was computational challenges.
-The challenges could be addressed by more efficient MCMC algorithms, approximate methods, or alternate computing methods (\eg the use of GPUs).
-Alternate computing methods would require appropriate hardware within the SRS, which should be considered when choosing the design of future TREs.
+\subsection{Study design}
 
@@
 
-The challenges in estimating the distribution of the duration of positivity originated in the design of the studies used.
-Exploring the design of future studies to better estimate the distribution of the duration of positivity would be beneficial.
+The challenges in estimating the distribution of the duration of detectability originated in the design of the studies used.
+Exploring the design of future studies to better estimate the distribution of the duration of detectability would be beneficial.
 A particularly promising direction would be intensive follow-up of individuals following a positive test (\eg daily testing) until they have two or more negative tests.
@@
 It is plausible that sampling in proportion to incidence would be beneficial.
-This is especially true if policy questions (\eg closing of schools) are of importance.
+This is especially true if policy questions such as closing of schools are of importance.
 These questions could be investigated through simulation studies.
-In constructing such studies, it will be important to consider how these considerations might be different for different pathogens (\eg influenza, the most common source of respiratory pandemics) and tests (\eg rapid antigen tests, which have a shorter duration of positivity).
+In constructing such studies, it will be important to consider how these considerations might be different for different pathogens (\eg influenza, the most common source of respiratory pandemics) and tests (\eg rapid antigen tests, which have a shorter duration of detectability).
+
+\subsection{Other extensions}
 
@@
 
+A common theme in this thesis was computational challenges.
+The challenges could be addressed by more efficient MCMC algorithms, approximate methods, or alternate computing methods (\eg the use of GPUs).
+Alternate computing methods would require appropriate hardware within the SRS, which should be considered when choosing the design of future TREs.
+
 \ifSubfilesClassLoaded{

### 2024-03-29 introduction.tex
proof

@@
 However, most studies included in the meta-analysis involved hospitalised patients or had unclear inclusion criteria, and hence may not be representative of the general population.
-These summary estimates are insufficient for backcalculation, which requires a distribution of the duration of positivity.
+These summary estimates are insufficient for backcalculation, which requires a distribution of the duration of detectability.
 
 \Textcite{hellewellPCRSensitivity} estimated the probability of an individual testing positive each day after infection.
-This measure is highly related to the duration of positivity.
+This measure is highly related to the duration of detectability.
 However, the study's cohort consisted only of healthcare workers, \ie working-age adults, and included only 27 infections.
@@
 Furthermore, asymptomatic infections were excluded due to limitations of their methodology.
-Duration of positivity may be correlated with severity of symptoms, and hence this exclusion could bias their results.
+Duration of detectability may be correlated with severity of symptoms, and hence this exclusion could bias their results.
 
-\Textcite{ealesCharacterising} estimated the duration of positivity from REACT data.
+\Textcite{ealesCharacterising} estimated the duration of detectability from REACT data.
 Specifically, individuals who tested positive in round 8 of REACT (6th--22 Jan 2021) were invited to take two additional swab tests.
@@
 This assumption is necessary because they had at most 17 days of follow-up, and hence could not estimate the tail of the distribution.
-Overall, they estimate a median duration of 14.0 days (95\% CI: 12.9--15.4 days) with substantial positive correlation with each individual’s viral load at the time of the initial positive test (\ie a higher initial viral load was associated with a longer duration of positivity).
+Overall, they estimate a median duration of 14.0 days (95\% CI: 12.9--15.4 days) with substantial positive correlation with each individual’s viral load at the time of the initial positive test (\ie a higher initial viral load was associated with a longer duration of detectability).
 The individuals with the highest viral load individuals were estimated to have a median duration of 19.6 days (16.1--25.8 days).
@@
 The study design prohibits estimating how long before the initial positive test the infection occurred.
-The same issue means the study underestimates the duration of positivity.
+The same issue means the study underestimates the duration of detectability.
 
 For the purposes of this thesis, I require estimates applicable to the general population.
-There are no reliable estimates of the full distribution of the duration of positivity in a general population cohort available in the literature.
-Therefore, I estimate the duration of positivity from the ATACCC and CIS studies.
+There are no reliable estimates of the full distribution of the duration of detectability in a general population cohort available in the literature.
+Therefore, I estimate the duration of detectability from the ATACCC and CIS studies.
 
@@
 % The length of stay in these compartments (the duration of time positive) is assumed to follow a gamma distribution, with the parameters estimated from external data or having at most one free parameter.
-Their model structure limits the form of the duration of positivity. 
+Their model structure limits the form of the duration of detectability. 
 They fit the models to prevalence survey data using a binomial likelihood where possible, although some approaches take a normal approximation.
@@
 \Textcite{nicholsonImproving} fit a simple mechanistic model, notably without age structure, but have the most sophisticated observation model.
-They allow an arbitrary distribution of the duration of positivity, based on the estimates of \textcite{hellewellPCRSensitivity} (see \cref{intro:sec:previous-duration-estimates} for discussion of the \textcite{hellewellPCRSensitivity} study).
+They allow an arbitrary distribution of the duration of detectability, based on the estimates of \textcite{hellewellPCRSensitivity} (see \cref{intro:sec:previous-duration-estimates} for discussion of the \textcite{hellewellPCRSensitivity} study).
 Furthermore, they incorporate the sensitivity and specificity of the RT-PCR test into their model.
@@
 Finally, the period ends before the roll-out of the vaccine, which likely affects many important aspects of the analysis.
-In particular, the duration of positivity and the susceptibility of the population.
+In particular, the duration of detectability and the susceptibility of the population.
 Including both of these elements would require more complex modelling and including idiosyncratic features of the SARS-CoV-2 pandemic that are less generalisable to other diseases.
@@
 In particular, it does not require the prevalence or incidence to be large.
-The estimates I produce in \cref{E-backcalc} fully propagate the uncertainty from the prevalence estimation procedure and use estimates of the duration of positivity from a general population study, unlike prior work.
+The estimates I produce in \cref{E-backcalc} fully propagate the uncertainty from the prevalence estimation procedure and use estimates of the duration of detectability from a general population study, unlike prior work.
 
@@
 Finally, in \cref{E-SEIR}, I describe a compartmental model for SARS-CoV-2 transmission, which I fit to CIS data.
-I allow a more flexible distribution of the duration of positivity than any of the papers discussed above, except \textcite{nicholsonImproving}, incorporating the skew of the distribution.
+I allow a more flexible distribution of the duration of detectability than any of the papers discussed above, except \textcite{nicholsonImproving}, incorporating the skew of the distribution.
 This distribution is based on data from a representative sample of the UK population, rather than convenience samples (see \cref{intro:sec:previous-duration-estimates}).
@@
 Then, in \cref{E-inc-prev}, I formally describe the stochastic processes which generate infections, prevalence, and the observations of prevalence.
-A key insight in this chapter is that the distribution of the duration of positivity is the key quantity linking incidence and prevalence.
+A key insight in this chapter is that the distribution of the duration of detectability is the key quantity linking incidence and prevalence.
 
 I then turn to estimating this distribution.
-I start, in \cref{E-ATACCC}, by estimating the duration of positivity from the ATACCC study.
+I start, in \cref{E-ATACCC}, by estimating the duration of detectability from the ATACCC study.
 This provides a reliable estimate of the bulk of the distribution (out to around 20 days) but shows that there is a significant tail.
@@
 Therefore, I turn to the CIS data in \cref{E-perf-test,E-imperf-test}.
-In \cref{E-perf-test}, I provide a framework for estimating the duration of positivity from the CIS data, assuming perfect tests.
+In \cref{E-perf-test}, I provide a framework for estimating the duration of detectability from the CIS data, assuming perfect tests.
 However, the assumption of perfect tests is not realistic; in \cref{E-imperf-test}, I extend the framework to include false negatives.
 
-Having estimated the duration of positivity, next I estimate incidence and transmission from the CIS data in \cref{E-transmission}, thus fulfilling the primary aim of this thesis.
+Having estimated the duration of detectability, next I estimate incidence and transmission from the CIS data in \cref{E-transmission}, thus fulfilling the primary aim of this thesis.
 I use both a phenomenological and mechanistic transmission model, contrasting their results and discussing the advantages and disadvantages of each in this context.

### 2024-03-29 introduction_new.tex
proof

@@
 
-The overall question that I investigate in this thesis is whether a prevalence survey \emph{alone} is sufficient to infer incidence and transmission and, if so, what methodology can be used.
+The overall question that I investigate in this thesis is: can a prevalence survey \emph{alone} be sufficient to infer incidence and transmission and, if so, what method(s) are optimal?
 % First, how can prevalence surveys be designed to help estimate these quantities?

### 2024-03-29 paper-surv-analysis.tex
proof

@@
 
-For privacy reasons, the CIS data is stored in the ONS's SRS (Secure Research Service).
+For privacy reasons, the CIS data are stored in the ONS's SRS (Secure Research Service).
 The SRS is a Trusted Research Environment (TRE) that allows researchers to access sensitive data in a secure environment.
@@
 
-In forming the probability $p_{ia}$, the data is used.
+In forming the probability $p_{ia}$, the data are used.
 This is not quite correct because the cell probabilities in the likelihood should not depend on the data.
@@
 
-To generate the durations, an assumption is required for the distribution of the duration of positivity.
+To generate the durations, an assumption is required for the distribution of the duration of detectability.
 I base this assumption on the estimate from \cref{E-ATACCC} with an inflated tail to represent what is seen within CIS.

### 2024-03-29 thesis.tex
proof ch 6 inc new figures
Updates from Overleaf
finish proof reading
minor corrections / improvements
add extra gof plots for ATACCC
proof

@@
 \newcommand{\psens}{p_\text{sens}}
+\newcommand{\psenss}{p_\text{sens}^{(s)}}
+\newcommand{\psensi}{p_\text{sens}^{(i)}}
 \newcommand{\ntot}{n_\text{tot}}
@@
 \usepackage[backend=biber,uniquename=init,sortcites=true,sorting=ynt,maxcitenames=2,style=apa,uniquelist=false]{biblatex}
+\setcounter{biburllcpenalty}{3000}
 \addbibresource{references.bib}
@@
 
-    This relies on knowledge of the duration of positivity (how long individuals test positive for).
+    This relies on knowledge of the duration of detectability (how long individuals test positive for).
     Therefore, I first consider how to estimate the distribution of this duration in the general population.
-    I extend a pre-existing model of how an individual's viral load changes over time to estimate the duration of positivity.
+    I extend a pre-existing model of how an individual's viral load changes over time to estimate the duration of detectability.
     % I take two complementary approaches using different datasets.
     % This model is fitted to data from a longitudinal study of contacts of individuals with confirmed SARS-CoV-2 infection.
-    % In \cref{ATACCC}, I extend a pre-existing model of how an individual's viral load changes over time to estimate the duration of positivity.
+    % In \cref{ATACCC}, I extend a pre-existing model of how an individual's viral load changes over time to estimate the duration of detectability.
     % This model is fitted to data from the ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts) study, a longitudinal study of viral load in contacts of individuals with confirmed SARS-CoV-2 infection.
@@
     % I would especially like to thank Drs Shaun Seaman and Brian Tom for their help with survival analysis.
-    The admin team is a great asset to the Unit and were outstandingly supportive through the disrupted the pandemic caused.
+    The admin team is a great asset to the Unit and were outstandingly supportive through the disruption the pandemic caused.
     My fellow PhD students have created a great environment, having peers at the same stage in my journey has been invaluable.
@@
 \subfile{ATACCC-appendix-original-analysis}
+\subfile{appendix-plots}
+\subfile{cis-perf-new-ll}
 \subfile{transmission-appendix-INLA}
 \subfile{transmission-appendix-phase-type}
-\subfile{transmission-appendix-plots}
 

### 2024-03-29 transmission.tex
proof

@@
 These individuals then transition into a set of compartments, $p_1, \dots, p_l$, each with their own transition rate.
-I will choose $l$ and the transition rates such that combined length of stay (\ie time from entry in $p_1$ until exit from $p_l$) approximates the duration of positivity distribution estimated in \cref{E-imperf-test}.
+I will choose $l$ and the transition rates such that combined length of stay (\ie time from entry in $p_1$ until exit from $p_l$) approximates the duration of detectability distribution estimated in \cref{E-imperf-test}.
 This idea extends the technique of having multiple exposed or latent compartments (introduced in \cref{E-SEIR:sec:non-exponential}).
@@
 Priors for the parameters are given in \cref{SEIR:table:priors}.
-The parameters governing the duration of positivity were explained previously (see \cref{SEIR:table:ec-params}).
+The parameters governing the duration of detectability were explained previously (see \cref{SEIR:table:ec-params}).
 
@@
 An extension of either (or both) models into the rest of the epidemic would be useful.
-The main obstacle is knowledge of how the duration of positivity distribution changes, and the need to incorporate the changes into the model.
+The main obstacle is knowledge of how the duration of detectability distribution changes, and the need to incorporate the changes into the model.
 The SEIR model would also need to include the effect of vaccination on transmission.

### 2024-03-29 cis-perf-new-ll.tex
Start on false negs

@@
 
-Define an integer $N_D$ and $\set{D} = \vec{\nu}_1, \dots, \vec{\nu}_{N_D}$ as the set of all possible observations of detected episodes; that is, $O(j) \in \set{D}$ if and only if $j$ is a detected infection.
+Define an integer $N_D$ and $\set{D} = \{ \vec{\nu}_1, \dots, \vec{\nu}_{N_D} \}$ as the set of all possible observations of detected episodes; that is, $O(j) \in \set{D}$ if and only if $j$ is a detected infection.
 For any $\vec{\nu} = [l^{(b)}, r^{(b)}, l^{(e)}, r^{e}, i]^T$, $\vec{\nu} \in \set{D}$ if and only if all the following conditions hold.
@@
 
-Introducing false negatives means that $O(j)$ is now a random variable because the test results between the start and the end of the episode could contain false negatives.
-Therefore, the
+Introducing false negatives means that $O(j)$ is now a random, even if $B_j$ and $E_j$ are known.
+Define the relevant testing times for episode $j$ as $\sched'_j = \{ t \in \sched_{i(j)} \ssep r_j^{(b)} \leq t \leq r_j^{(e)} + 1 \}$, let $m_j$ denote the size of this set.
+Let $O'(j) = [O(j), \vec{y}_j]^T$ where $\vec{y}_j$ is a binary vector of length $m_j$, with the $l$th element of of $\vec{y}_j$ being the test result at the chronologically $l$th time $t \in \sched'_j$.
+$O'(j)$ is the observed data for episode $j$.
+Defining analogous variables to \cref{E-perf-test}, let $\set{D}' = \{ \vec{\nu}'_1, \dots, \vec{\nu}'_{N_D} \}$ where $\vec{\nu}'_j \in \set{D}'_j$ if and only if $\vec{\nu}'_j = [\vec{\nu}_j, \vec{y}_j]^T$ could be a detected observation, equivalent to satisfying all the following conditions.
+\begin{enumerate}
+  \item $\vec{\nu}_j \in \set{D}$.
+  \item $\vec{y}_j \in \{0, 1\}^{m_j}$.
+  \item The elements of $\vec{y}_j$ corresponding to the tests at times $r_j^{(b)}$ (its first element) and $l_j^{(e)}$ (its penultimate element) are 1.
+  \item The element of $\vec{y}_j$ corresponding to the test at time $r_j^{(e)} + 1$ (its last element) is 0.
+\end{enumerate}
+The final two conditions are due the construction of the intervals as positive and negative tests bounding the beginning and end times of the episode.
 

### 2024-03-29 biology-data.tex
Updates from Overleaf

@@
 
+The ATACCC and CIS studies have complementary features that make combining information for the studies to estimate the distribution of the duration of detectability an appealing approach.
+ATACCC tested individual frequently but had only 57 infections followed for a maximum of 20 days.
+The frequent testing makes estimation for these first 20 days reliable, but longer periods are rare and the data does not exist; therefore, the tails of the distribution cannot be estimated.
+The CIS had more infrequent testing but a large sample size (thousands of infections) and up to years of follow-up.
+Intuitively, information from ATACCC can inform the first two to three weeks of the distribution and information from the CIS can inform the remainder of the distribution.
+
 In this chapter, I have introduced RT-PCR testing, the natural history of SARS-CoV-2, and the two studies used in this thesis.

### 2024-03-29 incidence-prevalence.tex
proof
Updates from Overleaf

@@
 
-First, in \cref{inc-prev:sec:poisson-process}, as a \emph{Poisson process}.
+First, in \cref{inc-prev:sec:poisson-process}, I introduce a phenomenological model where the infection process is a \emph{Poisson process}.
 A Poisson process is a widely used type of counting process both within and outside of epidemiology.
@@
 Mechanistic models explicitly describe the population and transmission between individuals.
+In particular, they include the feedback loop inherent in infectious diseases, where increased infections at a time $t$ lead to more infections at times $> t$.
 They vary greatly in their details and realism~\autocite{murilloMultiscale}.
@@
 % These additional infections will propagate the epidemic further, increasing the intensity at future times.
-Point processes can be generalised to \emph{branching processes} to account for this feedback~\autocite[246]{yanDistribution}, although these are beyond the scope of this thesis.
+Counting processes can be generalised to \emph{branching processes} to account for this feedback~\autocite[246]{yanDistribution}, although these are beyond the scope of this thesis.
 

### 2024-03-29 introduction_new.tex
Updates from Overleaf

@@
 
-To see the differing uses of the reproduction numbers, consider the difference in the epidemic in England in March 2020, just following the start of the first lockdown, and in January 2021 just following the start of the third lockdown.
-These times have (at least) three epidemiologically relevant differences: (1) in January 2021 the Alpha variant was circulating which is more transmissible than the original variant present in March 2020; (2) the population behaved differently because the exact meaning of ``lockdown'' differed and attitudes had changed; (3) in January 2021, some individuals had immunity due to either prior infection or the start of the vaccination roll-out.
-The original and Alpha variants have different values for $\R$, isolating effect 1.
-$\Rc$ differs during the two periods, including effects 1 and 2.
-$\Re$ includes all these effects, as well as any others.
+% To see the differing uses of the reproduction numbers, consider the difference in the epidemic in England in March 2020, just following the start of the first lockdown, and in January 2021 just following the start of the third lockdown.
+% These times have (at least) three epidemiologically relevant differences: (1) in January 2021 the Alpha variant was circulating which is more transmissible than the original variant present in March 2020; (2) the population behaved differently because the exact meaning of ``lockdown'' differed and attitudes had changed; (3) in January 2021, some individuals had immunity due to either prior infection or the start of the vaccination roll-out.
+% The original and Alpha variants have different values for $\R$, isolating effect 1.
+% $\Rc$ differs during the two periods, including effects 1 and 2.
+% $\Re$ includes all these effects, as well as any others.
 
-These definitions ignore many complexities that arise when theoretically defining or practically estimating any of the reproduction numbers~\autocite{pellisEstimation}.
-For example, how to average across seasonal effects and a heterogeneous population, and the lack of a ground truth for these numbers in almost all situations.
-However, their intuitive simplicity and convenient properties in simple epidemic models mean that they are useful concepts for guiding policy.
+% These definitions ignore many complexities that arise when theoretically defining or practically estimating any of the reproduction numbers~\autocite{pellisEstimation}.
+% For example, how to average across seasonal effects and a heterogeneous population, and the lack of a ground truth for these numbers in almost all situations.
+% However, their intuitive simplicity and convenient properties in simple epidemic models mean that they are useful concepts for guiding policy.
 
@@
 Whether a given individual is \emph{prevalent}, \ie has the disease, is often measurable through diagnostic testing.
-Prevalence can be estimated through a \emph{prevalence survey}.
-A prevalence survey is a designed study which performs diagnostic testing of a sample of the population; they are typically very expensive to power appropriately.
 Prevalence and incidence are strongly related, via the duration of an infection (see \cref{E-inc-prev}).
@@
 Collecting data that \emph{directly} inform inference of the quantities in this section is challenging or prohibitively expensive.
-Hence, they are typically estimated from other data sources through appropriate ???? methodologies.
-{\color{red}{refer to some papers examples:Phil O'neil, stats in medicine, 1990 on linking models to data; De Angelis, Presanis, Analysing Multiple Epidemic Data Sources. "Handbook of
-Infectious Disease Data Analysis", Held, L., Hens, N., O'Neill, P.D. and Wallinga, J. (Eds.).
-Chapman & Hall/CRC, 2019.}}
+Hence, they are typically estimated from other data sources through appropriate statistical methods~\autocite{heldHandbook,oneillIntroduction}.
 
@@
 
-In what follows I discuss three types of data: incidence-type, prevalence-type, and data from specific studies.
-\emph{Incidence-type} and \emph{prevalence-type} data can inform estimation of incidence and transmission; however, this estimation almost always relies on prior knowledge of epidemiological and/or biological quantities.
+In what follows I discuss three types of data: incidence-type, prevalence-type, and data from longitudinal studies.
+\emph{Incidence-type} and \emph{prevalence-type} data can inform estimation of incidence and transmission.
+However, this estimation almost always relies on prior knowledge of epidemiological and/or biological quantities, typically estimated from longitudinal studies.
 For example, the time between infection and symptom onset.
-These quantities can be estimated from other studies.
 
@@
 
-Prevalence surveys are one source of prevalence-type data that can be unbiased.
+Prevalence can be estimated through a \emph{prevalence survey}.
+A prevalence survey is a designed study which performs diagnostic testing of a sample of the population; they are typically very expensive to power appropriately.
+On the other hand, they are a source of prevalence-type data that can be unbiased.
 For instance, testing a representative proportion of the population (or a subpopulation of interest) for a marker of current infection.
-For many diseases, prevalence is sufficiently low that this is prohibitively expensive to power.
 
@@
 
-\subsection{Other studies}
+\subsection{Longitudinal studies}
 
@@
 These include the duration of different disease states or the distribution of the time between infection events~\autocite{wallingaGI,dankwaStructural}.
-Specific studies, \eg longitudinal studies that follow  individuals over time, allow  estimation of many of these quantities.
+Longitudinal studies, that follow the same individuals over time, allow estimation of many of these quantities.
 
@@
 These relate incidence in the past to current (and future) incidence but do not explicitly represent the susceptible population.
-These include components of the transmission process  but do not explicitly describe the process.
 A common type of semi-mechanistic models is that based on a \emph{renewal equation}~\autocite{bhattSemimechanistic,fraserRenewal} of the most form:
@@
 where $Z(t)$ is the expected incidence at time $t$ and $\beta(t, \tau)$ is the expected rate at which an individual infected $\tau$ time units ago infects individuals at calendar time $t$.
-Therefore, the renewal equation relates incidence in the past to incidence in the future.
-This partially describes the transmission process, hence the name semi-mechanistic.
-The effective reproduction number is $\Re = \int_0^\infty \beta(t, \tau) d\tau$.
+The model parameters can be used to define an effective reproduction number as either $\Re = \int_0^\infty \beta(t, \tau) d\tau$ or $\Re = \int_0^\infty \beta(t+\tau, \tau) d\tau$~\autocite{gosticPractical}.
 % A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
 % The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
-Renewal equations do not feature further in this thesis; for further details, including practical considerations when using them as the basis for inference, see \textcite{nashEstimating,gosticPractical}.
+Renewal equations do not feature further in this thesis; for further details, including practical considerations when using them as the basis for inference, see \textcite{thompsonImproved,gosticPractical}.
 
@@
 Simple mechanistic models use a homogeneous population, with individuals differing only due to their disease state, and exponential distributions for the time in each disease state.
-More complex models relax these assumptions to incorporate heterogeneity, \eg through stratification.
+More complex models relax these assumptions to incorporate more flexible definitions and/or heterogeneity (see \cref{SEIR:sec:mechanistic-models}).
 % Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
@@
 Typically, this is either the time to an event or the duration of a disease stage.
-Time series data on $W_t$ cannot identify $f$, instead external information needs incorporating~\autocite{swallow2022challenges}.
-This generally comes from longitudinal studies where the occurrence of the infection's sequelae is monitored in a cohort of individuals ({\it e.g.} Collaborative Group on AIDS Incubation and HIV Survival including the CASCADE EC
-Concerted Action (2000). Time from HIV- 1
-seroconversion to AIDS and death in 13,000 individuals before the widespread use of
-highly active anti-retroviral therapy: effect of age at seroconversion and exposure
-category. Lancet 355, 1131-1137.)
+Time series data on $W_t$ cannot identify $f$, instead external information needs incorporating~\autocite{swallow2022challenges}, most commonly longitudinal studies.
 
@@
 The nowcasting estimates were based on mechanistic and semi-mechanistic models with a wide variety of data sources.
-The mechanistic models, in  particular, commonly incorporated multiple data sources \textcite{birrellEvidence}.
+The mechanistic models, in  particular, commonly incorporated multiple data sources.
 The models used were reviewed by \textcite{royalSocietyRnumber} (see appendix 2 of that review for details of the specific implementations).
@@
 These surveys use RT-PCR (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
-The first survey was the  Coronavirus (COVID-19) Infection Survey (CIS), run by the Office for National Statistics (ONS)~\autocite{CIS,cisMethodsONS}.
+The first survey was the CIS (Coronavirus (COVID-19) Infection Survey), run by the ONS (Office for National Statistics)~\autocite{CIS,cisMethodsONS}.
 The CIS was a longitudinal household study, testing a large number of individuals weekly or monthly for SARS-CoV-2 infection (see \cref{E-intro:sec:cis} for a detailed description).
@@
 
-\subsubsection{Other studies}
+\subsubsection{Longitudinal studies}
 
-Many ad hoc studies to estimate a wide variety of epidemiological and biological quantities of the SARS-CoV2 virus were carried out throughout the pandemic.
+Many longitudinal studies to estimate a wide variety of epidemiological and biological quantities of the SARS-CoV2 virus were carried out throughout the pandemic.
+These include the CIS, previously mentioned.
 % In addition to population-level surveillance, cohort studies help understand other aspects of the pathogen~\autocite{royalSocietyRnumber}.
 % For example, contact tracing data can inform how likely it is for transmission to occur in different settings and/or between different groups of people.
-One of the study, which  I will use in what follows, is the ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts) study~\autocite{singanayagamCommunity,hakkiOnset}.
+
+One of the studies, which I use in this thesis, is the ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts) study~\autocite{singanayagamCommunity,hakkiOnset}.
 In ATACCC, daily testing of a small number of individuals with short follow-up was performed.
@@
 I describe ATACCC in detail in \cref{E-biology-data:sec:ataccc} and use it to estimate the duration of detectability in \cref{E-ATACCC}.
-
-The second study is the CIS (introduced in \cref{intro:sec:data-for-estimating}).
-In addition to its role as a prevalence survey, the CIS followed the same individuals over a long period of time (over a year in many cases), collecting longitudinal data.
-The large sample size and long follow-up allows estimation of the tails of the distribution of the duration of detectability.
-However, the lack of frequent testing (at most weekly) prohibits estimation of the viral load dynamics.
-In particular, the vast majority of detected infections had only one or two positive tests, and hence the viral load trajectory of these infections cannot be estimated.
-I describe the CIS in detail in \cref{E-intro:sec:cis} and use it to estimate duration in \cref{E-perf-test,E-imperf-test}.
-
-Early on in the epidemic, incidence could be estimated directly from the longitudinal survey data by considering the proportion of the cohort that tested negative then positive at consecutive tests~\autocite{onsIncidenceOld}.
-However, the method was abandoned in autumn 2020 when the testing was too infrequent to ensure that all infections were captured (\ie tested positive at least once).
 
@@
 In this thesis I investigate the potential of the CIS as a single source of information for incidence and transmission estimation.
+
+All inference in this thesis is Bayesian.
+Including prior information allows identifiability in situations where this would not otherwise be possible and allows for combining knowledge from multiple sources.
+Additionally, many previous studies I build on used Bayesian inference.
 
@@
 \Textcite{abbottCISincidence} applied the backcalculation to CIS data to estimate the incidence of SARS-CoV-2 in the UK.
-They used the summary statistics of prevalence from the CIS  published regularly on the ONS website as data (CHECK), specifically the prevalence's posterior point estimate and bounds of the 95\% credible interval.
+They used summary statistics of prevalence estimates from the CIS, taken from the ONS website, as data; specifically, the prevalence's posterior point estimate and bounds of the 95\% credible interval.
 From these summary statistics, they derive a normally-distributed likelihood, independent for each day.
@@
 % They fit to summary statistics of the CIS data, using a daily, independent normal likelihood with parameters estimated from the posterior point estimate and 95\% credible interval published by the ONS.
-
-\Textcite{colmanAscertainment} used similar methodology to estimate the \emph{ascertainment rate} of community testing (the proportion of all infections that community testing detected).
+\Textcite{colmanAscertainment} used a similar method to estimate the \emph{ascertainment rate} of community testing (the proportion of all infections that community testing detected).
 % Their methodology was similar to \textcite{abbottCISincidence}, including using estimates of $f$ from \textcite{hellewellPCRSensitivity}.
@@
 
-The overall question that I investigate in this thesis is: can a prevalence survey \emph{alone} be sufficient to infer incidence and transmission and, if so, what method(s) are optimal?
+The overall question that I investigate in this thesis is whether a prevalence survey \emph{alone} is sufficient to infer incidence and transmission and, if so, what method(s) are optimal?
 % First, how can prevalence surveys be designed to help estimate these quantities?
@@
     A deterministic approach approximates the prevalence at any time as a deterministic function of the incidence, rather than as a random variable.
-    A deterministic backcalculation has previously been used to analyse the CIS data (\eg by \textcite{abbottCISincidence}); however, no justification for the approach was provided.
-    In \cref{E-inc-prev:sec:observation-process} I show that this is a good approximation as long as the proportion of the population sampled is small.
+    This approximation has previously been used to analyse the CIS data (\eg by \textcite{abbottCISincidence}); however, no justification for the approach was provided.
+    In \cref{E-inc-prev:sec:observation-process} I show that it is a good approximation as long as the proportion of the population sampled is small.
     In particular, the accuracy of the approximation does not require the prevalence or incidence to be large.
@@
     The CIS data are doubly interval censored with arbitrary periods where infections could be undetected.
-    While a statistical framework for this case exists, it had to be generalised to deal with the CIS design.
-    To improve estimation, I propose a novel prior to integrate information from the ATACCC study into this analysis.
+    I develop a statistical framework to account for these issues.
+    To improve estimation, I propose a novel prior to integrate information from my analysis of the ATACCC study into this analysis, accounting for model uncertainty.
 
@@
 
-    \item In \cref{E-backcalc}, I use a deterministic backcalculation to estimate incidence with full propagation of the uncertainty from the prevalence estimation.
-    Furthermore, I use the estimate of the duration of detectability from a general population study estimated in previous chapter.
+    \item In \cref{E-backcalc}, I use a deterministic backcalculation to estimate incidence, including full propagation of the uncertainty from the prevalence estimation.
+    Furthermore, I use my estimate of the duration of detectability from general population studies.
 
     \item In \cref{E-SEIR}, I develop a compartmental model for SARS-CoV-2 transmission, relying \emph{only} on CIS data. 
-    This uses a distribution of duration of detectability derived by translating the estimates in \cref{E-imperf-test} into a progression through model states via an Erlang-Coxian formulation.
+    I translate my estimate of the distribution of the duration of detectability into a progression through model states through an Erlang---Coxian distribution; this distribution is more flexible than standard approaches in the infectious disease literature.
     I include age-stratification in the transmission model, allowing estimation of age-specific transmission.
@@
 
-In \cref{E-transmission}, I estimate incidence and transmission of SARS-CoV-2 in England over the period WHICH.
+In \cref{E-transmission}, I estimate incidence and transmission of SARS-CoV-2 in England over the period.
 These estimates contribute to the literature reconstructing the SARS-CoV-2 epidemic in England.
@@
 
-To carry out the work in this thesis I developed bespoke code. To ensure the methodology developed can be easily adopted by others, the code used for the analyses is available online.
+To carry out the work in this thesis I developed bespoke code.
+To ensure the methodology developed can be easily adopted by others, this code is available online.
 An index to the code and, where possible, data are available at \url{https://github.com/joshuablake/thesis-index}.

### 2024-03-29 introduction_new.tex

@@
 
-The overall question that I investigate in this thesis is whether a prevalence survey \emph{alone} is sufficient to infer incidence and transmission and, if so, what method(s) are optimal?
+The overall question that I investigate in this thesis is: can a prevalence survey \emph{alone} be sufficient to infer incidence and transmission and, if so, what method(s) are optimal?
 % First, how can prevalence surveys be designed to help estimate these quantities?

### 2024-03-29 cis-perf-new-ll.tex
Finish new likelihood sketch

@@
 
+\subsubsection{Deriving $p_k$}
+
+Decompose $p_k$ as $p_k = \prob(O(j) = \nu_k \mid \vec{\theta}) = p_{ik} \prob(i(j) = i_k \mid \vec{\theta})$ where $p_{ik} = \prob(O(j) = \nu_k \mid i(j) = i_k, \vec{\theta})$.
+This is valid as $\prob(O(j) = \nu_k \mid i(j) \neq i_k) = 0$ due to the condition here being equivalent to equating the two vectors' final elements.
+Assume that each infection episode occurs independently and with equal probability in any individual, \ie $\prob(i(j) = i_k) = 1/\Ncis$ for all $j$ and $k$.
+
+Then the derivation for $p_{ik}$ proceeds as per the derivation for $p_{ia}$ in the original version, except replacing $j(i)$s with $k$s.
+
 \subsection{Imperfect testing}
@@
 
-Just a sketch, probably ignore for now.
-
-\begin{itemize}
-  \item Add $\vec{y}_j$ as the observed test results for episode $j$ to $O(j)$ (as defined previously).
-  \item Modify $p_k$ and $p_u$ as done currently.
-\end{itemize}
+Similarly, let $p_k'$ and $p_{ik}'$ have the same definition as $p_k$ and $p_{ik}$ except replacing $O$ with $O'$ and $\vec{\nu}_k$ with $\vec{\nu}_k'$.
 

### 2024-03-29 cis-perf-new-ll.tex
correct notation

@@
 $O'(j)$ is the observed data for episode $j$.
-Defining analogous variables to \cref{E-perf-test}, let $\set{D}' = \{ \vec{\nu}'_1, \dots, \vec{\nu}'_{N_D} \}$ where $\vec{\nu}'_j \in \set{D}'_j$ if and only if $\vec{\nu}'_j = [\vec{\nu}_j, \vec{y}_j]^T$ could be a detected observation, equivalent to satisfying all the following conditions.
+
+Defining analogous variables to \cref{E-perf-test}, let $\set{D}' = \{ \vec{\nu}'_1, \dots, \vec{\nu}'_{N_{D'}} \}$ where $\vec{\nu}'_k \in \set{D}'_k$ if and only if $\vec{\nu}'_k = [\vec{\nu}_k, \vec{y}_k]^T$ could be a detected observation, equivalent to satisfying all the following conditions.
 \begin{enumerate}
-  \item $\vec{\nu}_j \in \set{D}$.
-  \item $\vec{y}_j \in \{0, 1\}^{m_j}$.
-  \item The elements of $\vec{y}_j$ corresponding to the tests at times $r_j^{(b)}$ (its first element) and $l_j^{(e)}$ (its penultimate element) are 1.
-  \item The element of $\vec{y}_j$ corresponding to the test at time $r_j^{(e)} + 1$ (its last element) is 0.
+  \item $\vec{\nu}_k \in \set{D}$.
+  \item $\vec{y}_k \in \{0, 1\}^{m_k}$.
+  \item The elements of $\vec{y}_k$ corresponding to the tests at times $r_k^{(b)}$ (its first element) and $l_k^{(e)}$ (its penultimate element) are 1.
+  \item The element of $\vec{y}_k$ corresponding to the test at time $r_k^{(e)} + 1$ (its last element) is 0.
 \end{enumerate}

### 2024-03-29 cis-perfect-testing.tex
proof

@@
 I evaluate the model and priors using a simulation study (\cref{perf-test:sec:simulation-study}), showing that they successfully recover $S(t)$.
-Finally, I discuss the results in \c{perf-test:sec:discussion}.
+Finally, I discuss the results in \cref{perf-test:sec:discussion}.
 
@@
 
-Another important concepts typical of survival data is truncation, which refers to ...?
 Sometimes, only intervals satisfying some property are recorded.
@@
 Therefore, this is not strictly correct.
-This issue can be rectified be considering the inadmissible region as many subregions, which leads to the same result (sketched in \cref{perf-test:sec:alt-likelihood}).
+This issue can be rectified by considering the inadmissible region as many subregions, which leads to the same result (sketched in \cref{perf-test:sec:alt-likelihood}).
 A full formalisation of the derivation in \cref{perf-test:sec:alt-likelihood} would require much more complex notation for the same result; therefore, I prefer the presentation to follow.
@@
 %Furthermore, there are differences in the study design and laboratories used between the two studies (see \cref{E-intro:sec:studies}) which may mean that results do not generalise between the studies.
-I form the prior to include this prior information into the CIS analysis two steps.
+I form the prior to include this prior information into the CIS analysis in two steps.
 

### 2024-03-29 incidence-prevalence.tex
proof

@@
 \emph{Duration}, or \emph{time-to-event}, distributions are expressed in several equivalent ways~\autocite[17]{sunStatisticala}.
-The duration distribution I will be concerned with is the distribution of the duration of positivity.
+The duration distribution I will be concerned with is the distribution of the duration of detectability.
 Consider an individual $i$ infected on day $B_i$ that is prevalent during the interval $[B_i, E_i]$ (as in \cref{E-biology-data:sec:cis-episodes}).
@@
 When the number of infectious individuals is large, the law of large numbers implies that the mean-field behaviour is a good approximation of the fully stochastic model~\autocite[20]{diekmannMathematical}.
-The approximation works especially well if used to make infernce based on noisy observations of the system, when the observation noise dominates is much larger than the infection process noise; I make a formalisation of this argument to justify a mean-field approximation of the prevalence process in \cref{E-inc-prev:sec:observation-process}.
+The approximation works especially well if used to make inference based on noisy observations of the system, when the observation noise is much larger than the infection process noise; I make a formalisation of this argument to justify a mean-field approximation of the prevalence process in \cref{E-inc-prev:sec:observation-process}.
 
@@
 A standard extension to the SIR model is to instead assume a gamma distribution~\autocite[94]{keelingModeling,andersonSpread}.
-The extension adds adding multiple infectious compartments, shown in \cref{SEIR:fig:SIR-gamma}.
+The extension adds multiple infectious compartments, shown in \cref{SEIR:fig:SIR-gamma}.
 The ODEs are now:
@@
 The constant terms here are absorbed into the WAIFW matrix, hence $\beta_{aa'} = \kappa_{aa'} q_{aa'} \frac{1}{N_a} N_{a'}$~\autocite[section 9.2]{diekmannMathematical}.
-This gives the total force of infection on strata $a$ from all strata is, therefore, $\lambda_a(t) = \sum_{a' \in \set{A}} \beta_{aa'} i_{a'}(t)$ as above.
+This gives the total force of infection on strata $a$ from all strata is $\lambda_a(t) = \sum_{a' \in \set{A}} \beta_{aa'} i_{a'}(t)$ as above.
 If all $\beta_{aa'}$ are equal, then the model is equivalent to the homogeneous model (\ie without any structure).
@@
   \makebox[\textwidth][c]{\includegraphics[width=0.9\paperwidth]{inc-prev/contact_matrices}}
-  \caption[Age-based contact matrices]{Number of contacts reported between age groups in two contact surveys: (A) number of contacts reported in the POLYMOD survey~\autocite{mossongSocial}; (B) number of contacts reported in the BBC Pandemic survey~\autocite{klepacContacts}, which only collected data on those aged over 12.
+  \caption[Age-based contact matrices]{Number of contacts reported between age groups in two contact surveys: (A) number of contacts reported in the BBC Pandemic survey~\autocite{klepacContacts}, which only collected data on those aged over 12; (B) number of contacts reported in the POLYMOD survey~\autocite{mossongSocial}.
 A clear diagonal, showing associative mixing is visible, and children have more contacts.
@@
 The conditioning on $n_t$ and $\Npop$ is implicit in what follows.
-These quantities are fixed by the study design and population choice, and hence I assume independent of independent of $P_t$ and $Z_t$.
+These quantities are fixed by the study design and population choice, and hence I assume independent of $P_t$ and $Z_t$.
 

### 2024-03-29 introduction.tex
merge new intro into old

@@
 However, our best records suggests that, since 1600, they have (on average) occurred every four years with a mean death toll of 5.6 million~\autocite{maraniNovelEpidemics}.
-Recent estimates suggest that the current mean impact of pandemics is over 2.5 million deaths per year~\autocite{madhavPandemicMortality}.
+Recent estimates suggest that the current impact of pandemics is, on average, over 2.5 million deaths per year~\autocite{madhavPandemicMortality}.
 
@@
 The \emph{effective reproduction number}, $\Re$, is the mean number of secondary infections generated by a single infection at time $t$.
-In particular, it uses the behaviour and immunity status of the population at time $t$.
+In particular, it includes the effects of the behaviour and immunity status of the population at time $t$.
 With constant behaviour and a disease that grants immunity following infection, $\Re$ declines smoothly.
@@
 $\Rc$ is useful when comparing the effects of interventions at different points in time, when the immunity status of the population has changed.
-For example, when comparing the English lockdown in March 2020 and January 2021, there are two effects in play: there are more individuals with immunity in January 2021, and the lockdowns had different measures in place.
-$\Rc$ quantifies the latter effect.
 
-These definitions ignore many complexities that arise when theoretically defining or practically estimating any of the reproduction numbers~\autocite{pellisEstimation}.
-For example, effects due to seasonality, what a ``mean infection'' means in a heterogeneous population, and the lack of a ground truth for these numbers in almost all situations.
-However, their intuitive simplicity and convenient properties in simple epidemic models mean that they are useful concepts for guiding policy.
+% To see the differing uses of the reproduction numbers, consider the difference in the epidemic in England in March 2020, just following the start of the first lockdown, and in January 2021 just following the start of the third lockdown.
+% These times have (at least) three epidemiologically relevant differences: (1) in January 2021 the Alpha variant was circulating which is more transmissible than the original variant present in March 2020; (2) the population behaved differently because the exact meaning of ``lockdown'' differed and attitudes had changed; (3) in January 2021, some individuals had immunity due to either prior infection or the start of the vaccination roll-out.
+% The original and Alpha variants have different values for $\R$, isolating effect 1.
+% $\Rc$ differs during the two periods, including effects 1 and 2.
+% $\Re$ includes all these effects, as well as any others.
+
+% These definitions ignore many complexities that arise when theoretically defining or practically estimating any of the reproduction numbers~\autocite{pellisEstimation}.
+% For example, how to average across seasonal effects and a heterogeneous population, and the lack of a ground truth for these numbers in almost all situations.
+% However, their intuitive simplicity and convenient properties in simple epidemic models mean that they are useful concepts for guiding policy.
 
 Reproduction numbers are strongly connected to \emph{growth rates}.
-If changes in behaviour and susceptibility are negligible, as they often are over short periods of time, epidemics grow (or shrink) exponentially.
-The rate of this exponential growth is its \emph{growth rate}, normally denoted $r$.
-For a given disease and population, an analogous growth rate can be defined for each definition of the reproduction number~\autocite{pellisEstimation,paragGrowthRates,wallingaGI}.
+If changes in behaviour and susceptibility are negligible, as they often are over short periods of time, epidemics grow (or shrink) exponentially, with a growth rate $r$.
+For a given disease and population, a growth rate can be defined corresponding to each definition of the reproduction number~\autocite{pellisEstimation,paragGrowthRates,wallingaGI}.
 The advantages of using the growth rates are that they are easier to estimate and capture how quickly spread happens in calendar time; however, they do not directly give the strength of intervention required to control the epidemic~\autocite{royalSocietyRnumber}.
@@
 Growth rates and reproduction numbers only provide information on how an epidemic is changing.
-Yet the size of the epidemic is just as, if not more, important~\autocite{pellisEstimation}.
-\emph{Incidence} (the number of new infections in a given time period) is the primary metric for the magnitude of a pandemic.
-Incidence informs the short-term future burden of the disease, for example, the future number of severe events (hospitalisations or deaths).
-Incidence is translated into disease burden using a measure of severity; for example, the IFR (infection fatality rate), the proportion of infected individuals who die.
-Estimation of severity is beyond the scope of this thesis.
+The size of the epidemic is just as, if not more, important~\autocite{pellisEstimation}.
 
-\emph{Transmission} refers to the process by which an infectious disease generates new infections.
-Estimating transmission refers to inferring the parameters of this process.
-The components of transmission (\eg the number of contacts each individual makes per day) allow interventions to be targeted.
-They are also important for forecasting future incidence.
+\emph{Incidence} is the number of new infections in a given time period (see \cref{E-inc-prev:sec:definitions} for a formal definition).
+It is the primary metric for the magnitude of a pandemic .
+Incidence informs the short-term future burden of the disease, for example, the future number of severe events (hospitalisations or deaths); although this requires a severity measure, such as the IFR (infection fatality rate), the proportion of infected individuals who die.
 
-\emph{Prevalence} is the proportion of the population with the disease at a given time.
-Whether a given individual is \emph{prevalent} (has the disease) is often measurable through diagnostic testing.
-Performing diagnostic tests in a sample of the population at a given point in time is known as a \emph{prevalence survey}.
-Prevalence surveys can estimate the population prevalence but are typically very expensive to power appropriately.
+\emph{Transmission} refers to the process by which an infectious individual generates further infections.
+Estimating transmission refers to inferring this process.
+Understanding this process, \eg the number of contacts each individual makes per day, allows targeting interventions and forecasting future incidence.
+
+\emph{Prevalence} is the proportion of the population with the disease at a given time (see \cref{E-inc-prev:sec:definitions} for a formal definition).
+Whether a given individual is \emph{prevalent}, \ie has the disease, is often measurable through diagnostic testing.
 Prevalence and incidence are strongly related, via the duration of an infection (see \cref{E-inc-prev}).
@@
 
-Collecting data that directly inform inference of the quantities in this section is expensive or impossible.
-Hence, they are typically inferred from other data sources through modelling.
+Collecting data that \emph{directly} inform inference of the quantities in this section is challenging or prohibitively expensive.
+Hence, they are typically estimated from other data sources through appropriate statistical methods~\autocite{heldHandbook,oneillIntroduction}.
 
@@
 
-Data sources used for outbreak surveillance are often observational in nature.
+Data sources to inform outbreak surveillance are often observational in nature, \ie they are not the output of designed studies.
 Therefore, they suffer from biases that can be hard to understand and quantify~\autocite{shadboltChallenges}.
@@
 
-Here I discuss three types of data which are useful for surveillance: incidence-type, prevalence-type, and data from cohort studies.
-\emph{Incidence-type} and \emph{prevalence-type} data can inform estimation of incidence and transmission; however, this estimation almost always relies on prior knowledge of epidemiological and/or biological quantities.
+In what follows, I discuss three types of data: incidence-type, prevalence-type, and data from longitudinal studies.
+\emph{Incidence-type} and \emph{prevalence-type} data can inform estimation of incidence and transmission.
+However, this estimation almost always relies on prior knowledge of epidemiological and/or biological quantities, typically estimated from longitudinal studies.
 For example, the time between infection and symptom onset.
-These quantities can be estimated from other studies, \eg cohort studies.
 
@@
 
-Incidence-type data originate from observations of the consequences of a disease, \eg the development of symptoms or admission to hospital.
-This type of data typically suffers from reporting biases, particularly that the proportion of infections which are counted can vary over time or within different subgroups of the population~\autocites[chapter 9]{lashModern}{shadboltChallenges}.
+Incidence-type data originate from observations of the consequences of an infection, \eg the development of symptoms or admission to hospital.
+This type of data typically suffers from reporting biases, most problematically that the proportion of infections detected varying over time or within different subgroups of the population~\autocites[chapter 9]{lashModern}{shadboltChallenges}.
 
-An example of incidence-type data are community testing, reliant on individuals presenting for a test.
-However, individuals' ability and willingness to be tested varies over time, which needs to be accounted for in order to perform unbiased inference.
-Unfortunately, data to estimate these changes is normally unavailable.
+An example of incidence-type data are the number of cases identified through community testing, which relies on individuals presenting for a test.
+Individuals' ability and willingness to be tested varies over time, which needs to be accounted for in order to perform unbiased inference.
 
-Data on more severe symptoms are typically less affected by these biases.
+Data on severe disease is typically less affected by biases.
 An extreme example is the development of AIDS following infection by HIV.
@@
 
-However, severe symptoms can be rare in some population subgroups (\eg young, otherwise healthy individuals infected with SARS-CoV-2~\autocite{wardSero,bhopalChildren}).
-This means that the observations do not provide much information on incidence in those subgroups.
+However, severe disease can be rare in some population subgroups (\eg young, otherwise healthy individuals infected with SARS-CoV-2~\autocite{wardSero,bhopalChildren}), meaning they rarely appear in this data.
+Therefore, the data does not provide much information on incidence in those subgroups.
 Furthermore, the delay between infection and the severe event occurring means that recent infections cannot be estimated with much precision~\autocite{swallow2022challenges}.
@@
 Prevalence-type data originate from observations of the number or proportion of a population in a specific \emph{disease state} at a specific time.
-Commonly used disease states are the presence or value of a biomarker or symptoms.
-For example, the number of hospital beds occupied by individuals with the disease or the proportion of the population with a marker of prior infection.
-The issues associated with incidence-type data often also apply to prevalence-type data.
+For example, the number of individuals in hospital with the disease or the proportion of the population with a marker of prior infection.
+The same issues associated with incidence-type data often also apply to prevalence-type data.
 
-Prevalence surveys are one form of prevalence-type data that can be unbiased.
-Here, a representative proportion of the population (or a subpopulation of interest) are tested for a marker of current infection.
-For many diseases, prevalence is sufficiently low that this is prohibitively expensive to power.
+Prevalence can be estimated through a \emph{prevalence survey}.
+A prevalence survey is a designed study which performs diagnostic testing of a sample of the population; they are typically very expensive to power appropriately.
+On the other hand, they are a source of prevalence-type data that can be unbiased.
+For instance, testing a representative proportion of the population (or a subpopulation of interest) for a marker of current infection.
 
 However, \emph{seroprevalence} surveys are more feasible.
-Seroprevalence is the proportion of individuals in the population which have antibodies to the disease, indicating that the individual was previously infected~\autocite{cdcSeroprevalence}.
+Seroprevalence is the proportion of individuals in the population which have antibodies against a particular infection, indicating that the individual was previously infected~\autocite{cdcSeroprevalence}.
 Many more individuals have ever been infected than are currently infected, therefore, seroprevalence surveys can be powered much more cheaply~\autocite{wuSeroprevSimulation}.
@@
 
-\subsection{Cohort studies}
+\subsection{Longitudinal studies}
 
 Other epidemiological and/or biological quantities are normally required to estimate transmission.
-These depend on the model type, often including the duration of different disease states or the distribution of the time between infection events~\autocite{wallingaGI,dankwaStructural}.
-Cohort studies (following a group of individuals over time), are a study design which can estimate many of these quantities.
+These include the duration of different disease states or the distribution of the time between infection events~\autocite{wallingaGI,dankwaStructural}.
+Longitudinal studies, that follow the same individuals over time, allow estimation of many of these quantities.
 
@@
 
-The first component required to estimate incidence and transmission is a model of the infection process, a mathematical description of the process through which incidence arises.
+The first component required to estimate incidence and transmission is a model of the infection process, a mathematical description of the process through which infections are generated.
 A spectrum of models exists, categorised by the strength of assumptions they make about the transmission process~\autocite{beckerCOVIDmodels}.
-Stronger assumptions allow more insight into the transmission process and lower the variance of the estimates but can introduce bias and be computationally challenging to fit.
 
 The models with the weakest assumptions are \emph{phenomenological} models.
-These make almost no assumptions about the transmission process, viewing incidence as a generic counting process (introduced in full in \cref{E-inc-prev:sec:infection-process}).
-The most common model is that incidence follows a Poisson process, and the intensity of the process is the target of inference.
+These make almost no assumptions about the transmission process, viewing incidence as resulting from a counting process (introduced in full in \cref{E-inc-prev:sec:infection-process}).
+A common model is the Poisson process, and the intensity of the process is the target of inference.
 % The use of Poisson processes to model incidence of an infectious disease was very popular in the context of HIV/AIDS\todo{maybe a ref for Poisson in HIV/AIDS?}.
@@
 Next, are semi-mechanistic models.
-These have components of the transmission process in them but do not fully represent the process.
-A common type of semi-mechanistic model is those based on a \emph{renewal equation}~\autocite{bhattSemimechanistic}.
-The renewal equation for an infectious disease~\autocite{fraserRenewal}, in the most general form, is:
+These relate incidence in the past to current (and future) incidence but do not explicitly represent the susceptible population.
+A common type of semi-mechanistic models is that based on a \emph{renewal equation}~\autocite{bhattSemimechanistic,fraserRenewal} of the most form:
 \begin{align}
@@
 \end{align}
-where $Z(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the mean rate at which an individual infected $\tau$ time units ago infects individuals at calendar time $t$.
-Therefore, the renewal equation relates incidence in the past to incidence in the future.
-This partially describes the transmission process, hence the name semi-mechanistic.
-The effective reproduction number is $\Re = \int_0^\infty \beta(t, \tau) d\tau$.
+where $Z(t)$ is the expected incidence at time $t$ and $\beta(t, \tau)$ is the expected rate at which an individual infected $\tau$ time units ago infects individuals at calendar time $t$.
+The model parameters can be used to define an effective reproduction number as either $\Re = \int_0^\infty \beta(t, \tau) d\tau$ or $\Re = \int_0^\infty \beta(t+\tau, \tau) d\tau$~\autocite{gosticPractical}.
 % A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
 % The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
-Extensions of \cref{intro:eq:renewal} that track the number of susceptible individuals become full mechanistic models, equivalent to the compartmental models introduced next~\autocite{champredonEquivalence}.
-Renewal equations do not feature further in this thesis; for further details, including practical considerations when using them as the basis for inference, see \textcite{nashEstimating,gosticPractical}.
+Renewal equations do not feature further in this thesis; for further details, including practical considerations when using them as the basis for inference, see \textcite{thompsonImproved,gosticPractical}.
 
@@
 \emph{Compartmental} models, the most common type of mechanistic model, split the population into compartments, each representing a different state of the disease.
-The \emph{dynamics} of the epidemic, how it changes over time, are represented by the transitions between compartments (for more details see \cref{E-SEIR:sec:mechanistic-models}).
+The \emph{dynamics} of the epidemic (how it changes over time) are represented by the transitions between compartments (for more details see \cref{E-SEIR:sec:mechanistic-models}).
 Simple mechanistic models use a homogeneous population, with individuals differing only due to their disease state, and exponential distributions for the time in each disease state.
-More complex models relax these assumptions, in particular to incorporate heterogeneity.
-The age structure of the population is important for respiratory diseases because individuals meet individuals of their own age more frequently.
+More complex models relax these assumptions, incorporating more flexible definitions and/or heterogeneity (see \cref{SEIR:sec:mechanistic-models}).
 % Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
@@
 
-\emph{Disease models} link the transmission model to a latent quantity, $W_t$.
-$W_t$ will then be recorded through some system (\eg a testing programme) described by the observation model (see next section).
+The infection process is hardly ever directly observed; however, we often observe a second process.
+\emph{Disease models} link the transmission model to the value, $W_t$ of this second process at time $t$.
 For example, $W_t$ could be the incidence or prevalence of symptomatic individuals on day $t$.
+$W_t$ will then be recorded through some system (\eg a testing programme) described by the observation model (see next section), which could introduce additional complications.
 
@@
 \end{align}
-where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that ever appear in $W_t$; and $f(i)$ is the probability that an infection at time $t-i$ is included in count $W_t$, conditional on it ever being in one of the counts.
-$f$ is known as a \emph{delay function}.
-$f$ is commonly the pdf of a distribution, or a transformation of it.
+where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that result in $W_t$; and, conditional on being observed, $f(i)$ describes the (discretised) probability that an infection is included in the count $W_t$ at $i$ time units after the infection.
+$f$ is an example of a \emph{delay function}, and $f$ is commonly the pdf of a distribution, or a transformation of it.
 
@@
 
-Applying \cref{intro:eq:disease-model} requires the delay function, $f$.
-Typically, this is either the time until an event or the duration of a disease stage.
-Time series data on $W_t$ cannot normally identify $f$, instead external information needs incorporating~\autocite{swallow2022challenges}.
+Applying \cref{intro:eq:disease-model} requires knowledge of the delay function, $f$.
+Typically, this is either the time to an event or the duration of a disease stage.
+Time series data on $W_t$ cannot identify $f$, instead external information needs incorporating~\autocite{swallow2022challenges}, most commonly longitudinal studies.
 
-Two approaches are commonly used to estimate the duration of a disease stage: either modelling an underlying biomarker or modelling the duration directly~\autocite[e.g.][]{sweetingEstimating}.
-Modelling a biomarker uses more information per observation but requires stronger assumptions.
+Two approaches are commonly used to estimate the duration of a disease stage: either modelling an underlying biomarker or modelling the duration directly, for instance, in a \emph{survival analysis} framework~\autocite[e.g.][]{sweetingEstimating}.
 
-The biomarker approach models the trajectory of a biomarker over time in a cohort of individuals.
-An individual is in the disease state while the biomarker is above or below some threshold.
+The biomarker approach models the trajectory of a biomarker over time.
+Applying a threshold to the biomarker defines a disease state.
 This approach often requires parametric assumption(s) such as a functional form of the biomarker's trajectory.
-In addition, if individual-level variation is included, which is generally important, a distributional form of random effects is required.
-This requires both sufficient observations of the biomarker per individual and an assumption on the form of the distribution.
+In addition, including individual-level variation, which is generally important, requires sufficient observations of the biomarker per individual and an assumption on the form of the distribution of the variation.
 More details are in \cref{E-ATACCC}.
 
-Modelling the duration directly considers only the binary result at each test (positive or negative), for instance in a \emph{survival analysis} framework.
+Modelling the duration directly considers only the binary result at each test (positive or negative).
 This approach generally requires fewer parameters reducing the computational burden.
-It is also easier to specify non-parametric forms of the distribution, which is important for SARS-CoV-2 as the tail behaviour may be important.
+It is also easier to specify non-parametric forms of the distribution, which is useful in situation where the tail behaviour may be important.
 % A more detailed review of the literature is in the relevant chapters.
@@
 
-Both cases require longitudinal data, suggesting that cohort studies are the most appropriate study design.
-
 \section{SARS-CoV-2 in England}
 
-The surveillance systems used in response to SARS-CoV-2 are historically unique.
-Modern technology allowed large amounts of data to be collected, distributed, and analysed.
-While the HIV/AIDS pandemic inspired the development of a large amount of statistical methodology, it was far slower moving.
-Much analysis of SARS-CoV-2 data was done at great haste, requiring trade-offs to be made between timeliness and rigour of results presented.
+The response to the SARS-CoV-2 pandemic in the UK has been unprecedented, both in terms of the scientific input to policy and the surveillance data made available.
+The most prominent surveillance outputs were produced by members of SPI-M-O (the Scientific Pandemic Influenza Group on Modelling Operational sub-group), a sub-committee of SAGE (the Scientific Advisory Group for Emergencies)~\autocite{medleySPIM,govSPIMO}.
+SPI-M-O produced weekly estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
+The committee peer-reviewed these estimates to provide quality assurance and formed a consensus value using a statistical combination of the estimates~\autocite{parkCombining}.
+The consensus value was published on the Government's website and informed its response to the pandemic~\autocite{govRnumber}.
 
-A deep understanding of each data source's potential was not possible in real-time.
-In this thesis, I will investigate the potential of the prevalence surveys in greater detail.
-The disease model will be a prevalence model (defined further in \cref{E-inc-prev:sec:prevalence-process}) with the delay function being the duration of PCR positivity.
+The nowcasting estimates were based on mechanistic and semi-mechanistic models with a wide variety of data sources.
+The mechanistic models, in  particular, commonly incorporated multiple data sources.
+The models used were reviewed by \textcite{royalSocietyRnumber} (see appendix 2 of that review for details of the specific implementations).
+
+Additionally, these models were used to explore policy scenarios.
+For example, the decision at each of stage of lifting the lockdown in 2021 was informed by modelling studies~\autocite{sageEvidence}.
+These aided decision-makers in understanding plausible impacts of their actions.
 
@@
 
-The main data sources informing the response to the SARS-Cov-2 epidemic in England were based on: community testing (known as pillar 2 testing), the number of severe events, seroprevalence surveys, and prevalence surveys.
+The surveillance systems used in response to SARS-CoV-2 are unique. 
+The main data sources informing the response to the SARS-Cov-2 epidemic in England were based on: community testing; severe events; seroprevalence and prevalence surveys.
 Each of these data sources have strengths and weaknesses, see \textcite{royalSocietyRnumber} and references therein for further details on each of these data sources.
@@
 For large parts of the pandemic, community testing was widely available in the UK.
-However, it did not test a representative sample of the population.
-The number of tests conducted varied over time and space, for a variety of known and unknown reasons.
-These include: the policy on who is eligible for testing, the availability of tests, and the willingness of individuals to be tested.
-Therefore, fluctuations in the number of positive tests may not reflect changes in the true incidence of the disease.
-The large number of tests conducted allowed detailed analyses to be performed, including stratification by small geographic areas or age groups, although this does not solve the issue of the biases present.
+However, the number of tests carried out varied over time and space, and for a variety of reasons.
+Reasons included: policy changes on who was eligible for testing; availability of tests; and willingness of individuals to be tested.
+Therefore, changes in the number of positive tests did not necessarily reflect changes in disease incidence.
+The large number of tests conducted allowed detailed analyses to be performed, including stratification by small geographic areas or age groups, although the lack of interpretability remained.
 
-Severe events counts are generally a more reliable measure of the burden of the disease.
-The criteria for being a severe event are normally well-defined and generally well-recorded in the UK context.
-Therefore, these data are subject to fewer biases than community testing.
-However, SARS-CoV-2 rarely leads to severe events in younger individuals, meaning severe event counts are insensitive to changes in transmission only affecting younger individuals.
+Severe events counts were generally a more reliable measure of the burden of the disease.
+The criteria for being a severe event are normally well-defined and generally well-recorded.
+Therefore, these data were subject to fewer biases than the cases detected through community testing.
+However, SARS-CoV-2 rarely led to severe events in younger individuals, so severe event counts were insensitive to changes in transmission only affecting younger individuals.
 
@@
 
-Seroprevalence surveys were conducted throughout the epidemic, the largest of which was tested donated blood~\autocite{amirthalingamSeroprevalence}.
-In addition to the normal problem of the delay between infection and antibody response, blood donors are a convenience sample whose representativeness of the wider population is unclear.
+Seroprevalence surveys were conducted throughout the epidemic, the largest of which tested donated blood~\autocite{amirthalingamSeroprevalence}.
+In addition to the problem of the delay between infection and antibody response, blood donors are a convenience sample whose representativeness of the wider population is unclear.
 
 The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys in representative samples of the general population.
-These were designed to avoid the biases of community testing and observe the whole population.
-Therefore, some modelling is required to borrow information across strata or time to provide stable estimates, as well as to correct for non-response bias (see \cref{E-intro:sec:cis}).
-Despite this, they are considered the gold-standard metric.
+These were designed to avoid the biases of community testing and attempt to observe the general population.
+These surveys use RT-PCR (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
+The first survey was the CIS (Coronavirus (COVID-19) Infection Survey), run by the ONS (Office for National Statistics)~\autocite{CIS,cisMethodsONS}.
+The CIS was a longitudinal household study, testing a large number of individuals weekly or monthly for SARS-CoV-2 infection (see \cref{E-intro:sec:cis} for a detailed description).
 
-The use of prevalence surveys for estimating incidence and transmission is the focus of this thesis.
-These surveys use RT-PCR (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
-I focus on the CIS (Coronavirus (COVID-19) Infection Survey), run by the ONS (Office for National Statistics)~\autocite{CIS,cisMethodsONS}.
-The CIS was a longitudinal study, testing a large number of individuals weekly or monthly for SARS-CoV-2; it is described in detail in \cref{E-intro:sec:cis}.
-
-A second prevalence survey ran in the UK, known as REACT (REal-time Assessment of Community Transmission)~\autocite{rileyResurgence,rileyREACT}.
-REACT was a series of cross-sectional surveys, run across 19 rounds between May 2020 and March 2022.
-Each round consisted of at least \numprint{100000} randomly sampled individuals from the UK contributing swabs for testing and questionnaire data~\autocite{elliottTwin}.
+A second prevalence survey, known as REACT (the REal-time Assessment of Community Transmission)~\autocite{rileyResurgence,rileyREACT}, consisted in a series of cross-sectional surveys.
+REACT conducted 19 surveys between May 2020 and March 2022, each including at least \numprint{100000} randomly sampled individuals from the UK, who contributed swabs for RT-PCR testing and questionnaire data~\autocite{elliottTwin}.
 
 A major advantage of the CIS over REACT is that the CIS had continual data collection, while REACT has no test results between rounds.
-The lack of data between rounds meant that using REACT data to infer transmission in these periods requires strong model assumptions.
-A particular instance of this is that a peak in late December 2020 occurred between rounds of the REACT survey, complicating analysis of the survey data at that time~\autocite{rileyREACTround8}.
+This meant that using REACT data to infer transmission in these periods requires strong model assumptions.
+A particular instance of this is that a peak in prevalence in late December 2020 occurred between rounds of the REACT survey, complcating any analysis of REACT data over that period~\autocite{rileyREACTround8}.
 
-\subsubsection{Cohort studies}
+\subsubsection{Longitudinal studies}
 
-Many cohort studies, to estimate a wide variety of quantities, ran throughout the pandemic.
-In this thesis, I make use of two cohort studies with differing designs to estimate the duration of PCR positivity.
-These designs have complementary strengths, and I combine information from the studies.
-Both of the studies recruited individuals from the general population, aiming to observe a cohort of infections that are representative of the infections in the UK community.
-
+Many longitudinal studies were conducted throughout the pandemic.
+They aimed to estimate a wide variety of epidemiological and biological quantities of the SARS-CoV2 virus.
+These include the CIS, previously mentioned.
 % In addition to population-level surveillance, cohort studies help understand other aspects of the pathogen~\autocite{royalSocietyRnumber}.
 % For example, contact tracing data can inform how likely it is for transmission to occur in different settings and/or between different groups of people.
-The first study I used, ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts)~\autocite{singanayagamCommunity,hakkiOnset}, performed daily testing of a small number of individuals with short follow-up.
-The density of this sampling allows for reliable estimates of the viral load dynamics and other quantities of interest, most importantly the duration of detectability, but the results are limited because the small sample size and short follow-up prohibit estimation of the tails of these distributions.
-I describe ATACCC in detail in \cref{E-biology-data:sec:ataccc} and use it to estimate duration in \cref{E-ATACCC}.
 
-The second study is the CIS (introduced in \cref{intro:sec:data-for-estimating}).
-In addition to its role as a prevalence survey, the CIS followed the same individuals over a long period of time (over a year in many cases), collecting longitudinal data.
-The large sample size and long follow-up allows estimation of the tails of the distribution of the duration of detectability.
-However, the lack of frequent testing (at most weekly) prohibits estimation of the viral load dynamics.
-In particular, the vast majority of detected infections had only one or two positive tests, and hence the viral load trajectory of these infections cannot be estimated.
-I describe the CIS in detail in \cref{E-intro:sec:cis} and use it to estimate duration in \cref{E-perf-test,E-imperf-test}.
-
-Early on in the epidemic, incidence could be estimated directly from the longitudinal survey data by considering the proportion of the cohort that tested negative then positive at consecutive tests~\autocite{onsIncidenceOld}.
-However, the method was abandoned in autumn 2020 when the testing was too infrequent to ensure that all infections were captured (\ie tested positive at least once).
-
-\subsection{Models}
-
-Members of SPI-M-O (the Scientific Pandemic Influenza Group on Modelling, Operational sub-group) produced the most prominent modelling of the UK epidemic.
-SPI-M-O was a sub-committee of SAGE (the Scientific Advisory Group for Emergencies), generating evidence based on infectious disease modelling for policy development~\autocite{medleySPIM,govSPIMO}.
-This used a wide range of  modelling approaches depending on the data available, which fluctuated throughout the epidemic, and the question being addressed.
-Some were fit to data while others were purely used for simulation based on plausible parameter values.
-
-Each week, members of SPI-M-O would produce a set of estimates of the current state of the epidemic (\eg the value of $\Re$), known as \emph{nowcasting}.
-The committee peer-reviewed these estimates to provide quality assurance, removing any estimates unduly affected by issues (\eg data quality issues).
-They then formed a consensus value using a statistical combination of the estimates~\autocite{parkCombining}.
-The consensus value was published on the UK Government's website and informed its response to the epidemic~\autocite{govRnumber}.
-
-The nowcasting estimates were based on mechanistic and semi-mechanistic models with a wide variety of data sources.
-The mechanistic models commonly incorporated multiple data sources, known as \emph{data synthesis}, a technique reviewed by \textcite{birrellEvidence}.
-The models used were reviewed by \textcite{royalSocietyRnumber} (see appendix 2 of that review for details of the specific implementations).
-
-Additionally, these models were used to explore policy scenarios.
-For example, the decision at each stage of lifting the lockdown in 2021 was informed by modelling studies~\autocite{sageEvidence}.
-These aided decision-makers in understanding plausible impacts of their actions.
+One of the studies, which I use in this thesis, is the ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts) study~\autocite{singanayagamCommunity,hakkiOnset}.
+ATACCC recruited individuals from contacts of known infections reported to NHS Test and Trace, and collected data on them for 20 days.
+This included daily RT-PCR tests, which allowed quantification of their viral load (the amount of virus in an individual), a biomarker.
+The density of the sampling allows for modelling the trajectory of viral load to estimate the duration of detectability (how long an individual is RT-PCR positive for).
+However, the results from ATACCC are of limited value because of the small sample size and the short follow-up; prohibiting estimation of how quantities vary over long infections.
+I describe ATACCC in detail in \cref{E-biology-data:sec:ataccc} and use it to estimate the duration of detectability in \cref{E-ATACCC}.
 
@@
 
-In this thesis, I consider how prevalence surveys could be used for epidemic surveillance, in particular estimates of incidence and transmission.
-These surveys have well-understood designs which means that estimates derived from them are more robust.
-A crucial quantity in their use will be the distribution of the duration of RT-PCR positivity.
+The massive impact of the SARS-CoV-2 pandemic led to data collection on a scale never previously seen for an infectious disease.
+Analysis of this data was carried out rapidly, to provide the timely results needed to inform the ongoing and evolving response.
+Therefore, a deep understanding of the potential of each data source was not possible.
+In this thesis I investigate the potential of the CIS as a single source of information for incidence and transmission estimation.
 
-In the following subsection, I introduce the most relevant related work, from both prior to the pandemic and during its course.
+I use Bayesian inference throughout this thesis.
+Including prior information allows identifiability in situations where this would not otherwise be possible and allows for combining knowledge from multiple sources.
+Additionally, many previous studies I build on used Bayesian inference.
+
+In the following subsections, I introduce the most relevant related work, carried out both prior to the pandemic and during its course.
 
@@
 
-The original formulations~\autocite{brookmeyerMethod,rosenbergBackcalculation} used a Poisson process as the infection model; a disease model where all infections are eventually observed ($\zeta = 1$) and the delay function, $f$, is the incubation period distribution's pdf; and an observation model where all diagnoses are observed, and hence has no influence on the data.
+The original formulations~\autocite{brookmeyerMethod,rosenbergBackcalculation} use a Poisson process as the infection model; a disease model with a delay function, $f$, representing the pdf of the time between HIV infection and AIDS diagnosis and assumed $\zeta=$; and a trivial observation model where all infections are (eventually) detected.
 
-The disease and observation models must be adapted to apply backcalculation to SARS-CoV-2 prevalence surveys (see \cref{E-inc-prev}).
-The disease model must consider that this is now prevalence-type, rather than incidence-type data.
-This difference induces a dependency structure between the observations on each day.
-The observation model must now include the error arising from using a sample of the population rather than (assumed) full recording of AIDS diagnoses.
+In (see \cref{E-inc-prev}) I will adapt this backcalculation framework for use with prevalence survey data, such as the CIS.
+The use of prevalence-type data induces a dependency structure between the values of the latent process $W_1, \dots, W_t$; I include the dependency by modifying the disease model.
+In addition, unlike in the original formulations, not all infections are detected by the CIS.
+In the CIS, a sample of the population receive RT-PCR tests, not the entire population; I include the sampling error arising from this in the observation model.
 
@@
 
-An early meta-analysis estimated that individuals test positive for SARS-CoV-2 for a mean of 14.6 days (95\% CI: 9.3--20.0 days)~\autocite{cevikShedding}.
-The longest reported duration was observed at 83 days, although beyond day 9 no live virus was detected, indicating that the individuals testing positive for longer than this were not infectious.
+In the application of the backcalculation to the CIS data, the delay function $f$ represents the duration of detectability.
+Previous work has attempted to estimate $f$.
+An initial meta-analysis estimated that individuals test positive for SARS-CoV-2 for a mean of 14.6 days (95\% CI: 9.3--20.0 days)~\autocite{cevikShedding}.
+The longest reported observed duration was 83 days, although no individuals were infectious beyond day nine.
 However, most studies included in the meta-analysis involved hospitalised patients or had unclear inclusion criteria, and hence may not be representative of the general population.
-These summary estimates are insufficient for backcalculation, which requires a distribution of the duration of detectability.
+In addition, summary estimates are insufficient for backcalculation, which requires a distribution of the duration of detectability.
 
@@
 This measure is highly related to the duration of detectability.
-However, the study's cohort consisted only of healthcare workers, \ie working-age adults, and included only 27 infections.
-Therefore, the data are not suitable for inferring the full distribution in the general population.
+However, the study's cohort consisted only of healthcare workers, \ie working-age adults.
 
 \Textcite{binnySensitivity} applied a slight variant of \textcite{hellewellPCRSensitivity}'s methodology to data from New Zealand.
-The situation in New Zealand, who were successfully pursuing a zero-COVID strategy at this time, means that the individuals in the study were unusual.
+New Zealand was successfully pursuing a zero-COVID strategy at the time, meaning that the individuals in the study were unusual.
 For example, a third were travellers detected during their quarantine period on arrival to New Zealand but would have had negative RT-PCR tests before their departure.
-Furthermore, asymptomatic infections were excluded due to limitations of their methodology.
-Duration of detectability may be correlated with severity of symptoms, and hence this exclusion could bias their results.
 
@@
 Specifically, individuals who tested positive in round 8 of REACT (6th--22 Jan 2021) were invited to take two additional swab tests.
-Defining the time of the initial positive test at time 0, \textcite{ealesCharacterising} estimated the probability of remaining positive at each day after the initial positive test.
+Taking the time of the initial positive test as time 0, \textcite{ealesCharacterising} estimated the probability of remaining positive at each day after the initial positive test.
 They made a strong parametric assumption that the probability decreases exponentially, with a sensitivity analysis assuming a plateau before the decay.
@@
 The association could be due to individuals with a higher initial viral load being (on average) more recently infected at the time of the initial positive test.
-The study design prohibits estimating how long before the initial positive test the infection occurred.
-The same issue means the study underestimates the duration of detectability.
+The study design does not provide any information on how long before the initial positive test the infection occurred.
+This issue also means the study underestimates the duration of detectability, because the infection's true start time would be prior to the first positive test.
 
-For the purposes of this thesis, I require estimates applicable to the general population.
-There are no reliable estimates of the full distribution of the duration of detectability in a general population cohort available in the literature.
-Therefore, I estimate the duration of detectability from the ATACCC and CIS studies.
+For the purposes of this thesis, I require estimates of the duration of detectability applicable to the general population.
+As no such estimates of the full distribution of the duration are available in the literature, a key component of this work is devoted to estimating them.
+To do so, I will combine the information from the ATACCC and CIS studies.
 
@@
 
-\Textcite{abbottCISincidence} applied backcalculation to CIS data to estimate the incidence of SARS-CoV-2 in the UK.
-They used a Gaussian process prior to smooth the incidence and stabilise the estimates.
-\Textcite{abbottCISincidence} used \textcite{hellewellPCRSensitivity}'s estimates of the sensitivity of testing to inform the observation model, which are based on a longitudinal study in healthcare workers (see \cref{intro:sec:previous-duration-estimates}).
-They fit to summary statistics of the CIS data, using a daily, independent normal likelihood with parameters estimated from the posterior point estimate and 95\% credible interval published by the ONS.
-
-\Textcite{colmanAscertainment} used CIS prevalence estimates to estimate the \emph{ascertainment rate} of community testing (the proportion of all infections that community testing detected).
-Their methodology was similar to \textcite{abbottCISincidence}, including using estimates from \textcite{hellewellPCRSensitivity} for how the sensitivity of RT-PCR testing varies over an infection.
-Rather than using a Gaussian process to provide stability, as \textcite{abbottCISincidence} do, \textcite{colmanAscertainment} assume the ascertainment rate is piecewise constant.
-They do not fully propagate the uncertainty from the CIS prevalence estimates, considering only the bounds of the 95\% confidence intervals the ONS publishes for prevalence.
+\Textcite{abbottCISincidence} applied the backcalculation to CIS data to estimate the incidence of SARS-CoV-2 in the UK.
+They used summary statistics of prevalence estimates from the CIS, taken from the ONS website, as data; specifically, the prevalence's posterior point estimate and bounds of the 95\% credible interval.
+From these summary statistics, they derive a normally-distributed likelihood, independent for each day.
+They used \textcite{hellewellPCRSensitivity}'s estimates of the sensitivity of testing as $f$ in their observation model; these estimates are based on a longitudinal study in healthcare workers (see \cref{intro:sec:previous-duration-estimates}).
+% They fit to summary statistics of the CIS data, using a daily, independent normal likelihood with parameters estimated from the posterior point estimate and 95\% credible interval published by the ONS.
+\Textcite{colmanAscertainment} used a similar method to estimate the \emph{ascertainment rate} of community testing (the proportion of all infections that community testing detected).
+% Their methodology was similar to \textcite{abbottCISincidence}, including using estimates of $f$ from \textcite{hellewellPCRSensitivity}.
+% Rather than using a smooth prior to provide stability, as \textcite{abbottCISincidence} do, \textcite{colmanAscertainment} assume the ascertainment rate is piecewise constant.
+% They do not fully propagate the uncertainty from the CIS prevalence estimates, considering only the bounds of the 95\% confidence intervals the ONS publishes for prevalence.
 
@@
 % The length of stay in these compartments (the duration of time positive) is assumed to follow a gamma distribution, with the parameters estimated from external data or having at most one free parameter.
-Their model structure limits the form of the duration of detectability. 
-They fit the models to prevalence survey data using a binomial likelihood where possible, although some approaches take a normal approximation.
+The model structure used in these studies limited the distribution of the duration of detectability to specific classes of distributions.
 % These approaches are similar to how these models are fit to other data sources, such as hospital admissions or deaths.
 
-\Textcite{nicholsonImproving} fit a simple mechanistic model, notably without age structure, but have the most sophisticated observation model.
-They allow an arbitrary distribution of the duration of detectability, based on the estimates of \textcite{hellewellPCRSensitivity} (see \cref{intro:sec:previous-duration-estimates} for discussion of the \textcite{hellewellPCRSensitivity} study).
-Furthermore, they incorporate the sensitivity and specificity of the RT-PCR test into their model.
+\Textcite{nicholsonImproving} estimate prevalence at a small geographic scale, using REACT data to remove the observational biases in community testing data.
+They use a simple mechanistic model, without age structure, but a sophisticated observation model.
+They allow an arbitrary distribution of the duration of detectability, based on the estimates of \textcite{hellewellPCRSensitivity}, and incorporate the sensitivity and specificity of the RT-PCR test.
 
 \Textcite{mccabeCISincidence}, adapting the methodology \textcite{ealesAppropriately} developed for REACT, estimate $\Re$ and the growth rate from the CIS data.
-They estimate a posterior distribution of the parameters of a spline by fitting to the prevalence results via a binomial distribution.
-Their set-up penalizes the spline deviating from a constant growth rate.
-They calculate the growth rate from the derivative of the spline, an analytical transformation of the parameters. 
-Since, under reasonable model assumptions, there is a one-to-one correspondence between reproduction numbers and growth rates~\autocite{wallingaGI}, and they assume that the growth rate in incidence and prevalence is equal, this also estimates $\Re$.
+They first estimate prevalence using summary statistics from the CIS, then derive $\Re$ and $r$ using a semi-mechanistic model fitted to their prevalence estimates.
+% They then fit a semi-mechanistic model to the estimating prevalence 
+% a posterior distribution of the parameters of a spline by fitting to the prevalence results via a binomial distribution.
+% Their set-up penalizes the spline deviating from a constant growth rate.
+% They calculate the growth rate from the derivative of the spline, an analytical transformation of the parameters. 
+% Since, under reasonable model assumptions, there is a one-to-one correspondence between reproduction numbers and growth rates~\autocite{wallingaGI}, and they assume that the growth rate in incidence and prevalence is equal, this also estimates $\Re$.
 
-\section{Bayesian inference} \label{intro:sec:Bayes}
+% \section{Bayesian inference} \label{intro:sec:Bayes}
 
-All inference in this thesis is Bayesian.
-Including prior information allows identifiability in situations where this would not otherwise be possible and allows for combining knowledge from multiple sources.
-Many previous studies I build on used Bayesian inference.
+% All inference in this thesis is Bayesian.
+% Including prior information allows identifiability in situations where this would not otherwise be possible and allows for combining knowledge from multiple sources.
+% Many previous studies I build on used Bayesian inference.
 
-Analytically deriving the posterior distribution is impossible for the models in this thesis.
-Therefore, I use computational methods to sample from the posterior distribution.
-The method I use for the majority of the analyses is MCMC (Markov chain Monte Carlo) (see \cref{E-MCMC} for details).
+% Analytically deriving the posterior distribution is impossible for the models in this thesis.
+% Therefore, I use computational methods to sample from the posterior distribution.
+% The method I use for the majority of the analyses is MCMC (Markov chain Monte Carlo) (see \cref{E-MCMC} for details).
 
@@
 
-The overall question that I investigate in this thesis is: can prevalence surveys alone be used to infer incidence and transmission, and if so how?
-Prevalence surveys provide data that is far more representative of the general population than other data sources, therefore, conclusions based on them should be far more robust.
-There are two components to this question.
-First, how can prevalence surveys be designed to help estimate these quantities?
-Second, what statistical methodology can be used to estimate these quantities from prevalence surveys?
-This thesis will focus on the second component.
-However, natural extensions of the work I undertake would inform the first component.
+The overall question that I investigate in this thesis is: can a prevalence survey \emph{alone} be sufficient to infer incidence and transmission and, if so, what method(s) are optimal?
+% First, how can prevalence surveys be designed to help estimate these quantities?
 
-In order to achieve these aims, I use a case study, specifically the SARS-CoV-2 epidemic in England from Mon 31 Aug 2020 until Sun 25 Jan 2021.
-This case study has several features that make it useful for the aims of this thesis.
-These include that the CIS was of sufficient size and incidence sufficiently high that there is enough data to perform the analyses.
-Additionally, incidence begins rising over this period, which is the period when a surveillance system needs to provide a warning of upcoming issues.
-The CIS began in April 2020, when incidence was falling due to the first lockdown in the UK.
-Incidence remained low and fairly flat in England over summer 2020, until it began rising in mid or late August.
-Finally, the period ends before the roll-out of the vaccine, which likely affects many important aspects of the analysis.
-In particular, the duration of detectability and the susceptibility of the population.
-Including both of these elements would require more complex modelling and including idiosyncratic features of the SARS-CoV-2 pandemic that are less generalisable to other diseases.
+To address this question, I analyse data from the CIS study.
+I use, as a case study, data from the SARS-CoV-2 epidemic in England from Mon 31 Aug 2020 until Sun 25 Jan 2021.
+This period has several interesting features making it suitable for my aim.
+First, the CIS was of sufficient size and transmission was sufficiently high to provide enough data to analyse.
+Second, incidence began to rise over this period, after the fall due to the first lockdown.
+Finally, the period ends before the vaccine roll-out, which  affected many important aspects of the analysis; most relevantly, the duration of detectability and the susceptibility of the population.
+Incorporating vaccination effects would require more complex modelling and include idiosyncratic features of the SARS-CoV-2 pandemic that are less generalisable to other diseases.
 
 A secondary aim of this thesis is to improve our understanding of the epidemic in England over this period, and our understanding of the SARS-CoV-2 virus more broadly.
-The results of the analyses I perform in pursuit of the primary aim will naturally contribute to this aim.
+The results of the analyses I perform in pursuit of the primary aim will naturally contribute to this second aim.
 
@@
 
-As its primary aim, this thesis makes several methodological contributions.
+This thesis makes several methodological contributions.
+\begin{enumerate}
+    \item In \cref{E-inc-prev:sec:observation-process}, I provide a theoretic justification for the use of a \emph{deterministic} backcalculation for prevalence surveys.
+    A deterministic approach approximates the prevalence at any time as a deterministic function of the incidence, rather than as a random variable.
+    This approximation has previously been used to analyse the CIS data (\eg by \textcite{abbottCISincidence}); however, no justification for the approach was provided.
+    In \cref{E-inc-prev:sec:observation-process} I show that it is a good approximation as long as the proportion of the population sampled is small.
+    In particular, the accuracy of the approximation does not require the prevalence or incidence to be large.
 
-First, in \cref{E-inc-prev:sec:observation-process}, I provide a theoretic justification for the use of deterministic backcalculation in this setting.
-A \emph{deterministic} backcalculation approach was used by \textcite{abbottCISincidence} and I will use it in \cref{E-backcalc}.
-The deterministic approach models the prevalence at any time as a deterministic function of the incidence, rather than as a random variable.
-I show that this is a good approximation as long as the proportion of the population sampled is small.
-In particular, it does not require the prevalence or incidence to be large.
-The estimates I produce in \cref{E-backcalc} fully propagate the uncertainty from the prevalence estimation procedure and use estimates of the duration of detectability from a general population study, unlike prior work.
+    \item In \cref{E-perf-test}, I develop a novel statistical framework to estimate the duration of detectability using the CIS data.
+    The framework accounts for the CIS data being doubly interval censored and having arbitrary periods where infections could be undetected.
+    To improve inference within this framework, I propose a novel prior to integrate information from my analysis of the ATACCC study into this analysis, accounting for model uncertainty.
 
-In \cref{E-perf-test} I estimate the duration of RT-PCR-positivity from double censored data with arbitrary periods where infections could be undetected (a generalisation of truncation).
-This is based on the CIS data.
-While statistical frameworks have been previously developed which included this case, they had not been used in practice.
-In \cref{E-imperf-test}, I extend this framework to include false negatives.
-
+    \item In \cref{E-imperf-test}, I extend the framework above to account for false negatives.
     A particular challenge with these analyses is that they must be performed within a trusted research environment with limited computational resources.
 
-Within these analyses, I include information derived from the ATACCC study.
-In order to do so, I propose a prior which appropriately discounts the information from analysing the ATACCC study, which has inappropriate model assumptions when applied to the larger sample size in the CIS data.
+    \item In \cref{E-backcalc}, I use a deterministic backcalculation to estimate incidence, including full propagation of the uncertainty from the prevalence estimation.
+    Furthermore, I use my estimate of the duration of detectability from general population studies.
 
-Finally, in \cref{E-SEIR}, I describe a compartmental model for SARS-CoV-2 transmission, which I fit to CIS data.
-I allow a more flexible distribution of the duration of detectability than any of the papers discussed above, except \textcite{nicholsonImproving}, incorporating the skew of the distribution.
-This distribution is based on data from a representative sample of the UK population, rather than convenience samples (see \cref{intro:sec:previous-duration-estimates}).
-Unlike \textcite{nicholsonImproving}, I include age-stratification in the transmission model, allowing estimation of how transmission varies by age.
-I use only data from the CIS and show this is sufficient for estimating incidence and transmission.
+    \item In \cref{E-SEIR}, I develop a compartmental model for SARS-CoV-2 transmission, relying \emph{only} on CIS data. 
+    I translate my estimate of the distribution of the duration of detectability into a progression through model states through an Erlang---Coxian distribution; this distribution is more flexible than standard approaches in the infectious disease literature.
+    I include age-stratification in the transmission model, allowing estimation of age-specific transmission.
+    I show the CIS data are sufficient for estimating incidence and transmission.
+\end{enumerate}
 
@@
 
-In \cref{E-imperf-test}, I present estimates of the duration of RT-PCR-positivity in the general population.
+In \cref{E-imperf-test}, I generate estimates of the duration of RT-PCR-positivity in the general population.
 These estimates could inform public health policy, for example, in interpreting RT-PCR test results or formulating quarantine policies.
 
-In \cref{E-transmission}, I present estimates of the incidence and transmission of SARS-CoV-2 in England over the period.
+In \cref{E-transmission}, I estimate incidence and transmission of SARS-CoV-2 in England over the period.
 These estimates contribute to the literature reconstructing the SARS-CoV-2 epidemic in England.
-Understanding the dynamics of the past infection is important for informing the response to future pandemics, and the future burden of SARS-CoV-2.
+Understanding the dynamics of past infection is important for informing the response to future pandemics, and the future burden of SARS-CoV-2.
 
@@
 
-In order to best fulfil this thesis's primary aim, the methodology developed should be easy for others to use.
-Therefore, the code for the analyses in this thesis is available online.
+To carry out the work in this thesis I developed bespoke code.
+To ensure the methodology developed can be easily adopted by others, this code is available online.
 An index to the code and, where possible, data are available at \url{https://github.com/joshuablake/thesis-index}.
@@
 
-In \cref{E-biology-data}, I provide background information on the natural history of SARS-CoV-2, the RT-PCR testing process used to produce the data used throughout this thesis, and describe the two cohort studies used in this thesis.
-Then, in \cref{E-inc-prev}, I formally describe the stochastic processes which generate infections, prevalence, and the observations of prevalence.
-A key insight in this chapter is that the distribution of the duration of detectability is the key quantity linking incidence and prevalence.
+In \cref{E-biology-data}, I provide background information on: the natural history of SARS-CoV-2, the RT-PCR testing process used to produce the data analysed throughout this thesis, and describe the information available in the CIS and ATACCC studies.
 
-I then turn to estimating this distribution.
-I start, in \cref{E-ATACCC}, by estimating the duration of detectability from the ATACCC study.
-This provides a reliable estimate of the bulk of the distribution (out to around 20 days) but shows that there is a significant tail.
-Estimating the tail requires a large sample size and long follow-up. 
-Therefore, I turn to the CIS data in \cref{E-perf-test,E-imperf-test}.
-In \cref{E-perf-test}, I provide a framework for estimating the duration of detectability from the CIS data, assuming perfect tests.
-However, the assumption of perfect tests is not realistic; in \cref{E-imperf-test}, I extend the framework to include false negatives.
+In \cref{E-inc-prev}, I formally describe the stochastic processes which generate infections, prevalence, and the observations of prevalence.
+I show the distribution of the duration of detectability is the key quantity linking incidence and prevalence.
 
-Having estimated the duration of detectability, next I estimate incidence and transmission from the CIS data in \cref{E-transmission}, thus fulfilling the primary aim of this thesis.
+In \cref{E-ATACCC}, I estimate the duration of detectability from the ATACCC study.
+This provides a reliable estimate of the bulk of the distribution (out to around 20 days),  but shows that estimating the tail requires a large sample size and long follow-up. 
+These features are present in the CIS, in \cref{E-perf-test} I provide a framework for estimating the duration of detectability assuming perfect tests from prevalence surveys such as the CIS.
+However, as the assumption of perfect tests is not realistic; in \cref{E-imperf-test}, I extend the framework to include false negatives.
+
+Having estimated the duration of detectability, I estimate incidence and transmission from the CIS data in \cref{E-transmission}.
 I use both a phenomenological and mechanistic transmission model, contrasting their results and discussing the advantages and disadvantages of each in this context.
-Finally, I conclude in \cref{E-conclusion}.
 
-\Cref{E-distributions} provides definitions of the probability distributions used in this thesis.
-Other appendices provide additional details.
+Finally, I conclude in \cref{E-conclusion} with a discussion of the limitations of this work and ideas for future research.
+
+Appendices provide definitions of the probability distributions used in this thesis, further details, and additional results.
 

### 2024-03-29 cis-perf-new-ll.tex

@@
 Define an integer $N_D$ and $\set{D} = \{ \vec{\nu}_1, \dots, \vec{\nu}_{N_D} \}$ as the set of all possible observations of detected episodes; that is, $O(j) \in \set{D}$ if and only if $j$ is a detected infection.
-For any $\vec{\nu} = [l^{(b)}, r^{(b)}, l^{(e)}, r^{e}, i]^T$, $\vec{\nu} \in \set{D}$ if and only if all the following conditions hold.
+REF FIG 5.3 HERE
+Let $\vec{\nu}_k = [l^{(b)}_k, r^{(b)}_k, l^{(e)}_k, r^{e}_k, i_k]^T$ be an arbitrary vector if possible observations, $\vec{\nu}_k \in \set{D}$ if and only if all the following conditions hold.
 \begin{enumerate}
-  \item $l^{(b)} - 1, r^{(b)}, l^{(e)}, r^{e} + 1 \in \sched_i$, that is individual $i$ was tested at all these times.
-  \item $l^{(b)} \leq r^{(b)} \leq l^{(e)} \leq r^{(e)}$.
-  \item $r^{(b)} = \min\{t \in \sched_i \ssep t \geq l^{(b)}\}$, that is $r^{(b)}$ is the first time at or after $l^{(b)}$ that individual $i$ was tested.
-  \item $r^{(e)} = \min\{t \in \sched_i \ssep t > l^{(e)}\} - 1$, that is $r^{(e)}$ is the day before the next time after $l^{(e)}$ that individual $i$ was tested.
-  \item $1 \leq r^{(b)} \leq T$, that is the episode begins in the period of interest.
+  \item $l^{(b)}_k - 1, r^{(b)}_k, l^{(e)}_k, r^{e}_k + 1 \in \sched_{i_k}$, that is individual ${i_k}$ was tested at all these times.
+  \item $l^{(b)}_k \leq r^{(b)}_k \leq l^{(e)}_k \leq r^{(e)}_k$.
+  \item $r^{(b)}_k = \min\{t \in \sched_{i_k} \ssep t \geq l^{(b)}_k\}$, that is $r^{(b)}_k$ is the first time at or after $l^{(b)}_k$ that individual ${i_k}$ was tested.
+  \item $r^{(e)}_k = \min\{t \in \sched_{i_k} \ssep t > l^{(e)}_k\} - 1$, that is $r^{(e)}_k$ is the day before the next time after $l^{(e)}_k$ that individual ${i_k}$ was tested.
+  \item $1 \leq r^{(b)}_k \leq T$, that is the episode begins in the period of interest.
 \end{enumerate}
@@
 In the CIS data, each $n_k$ ($k \neq u$) is observed as either 0 or 1.
-Let $k \in \set{C}$ indicate that $n_k = 1$.
+Define $C = \{ k \ssep n_k = 1 \}$.
 Furthermore, note that the support of the multinomial distribution requires that $\nnodet = \ntot - \ndet$.
@@
 
-Introducing false negatives means that $O(j)$ is now a random, even if $B_j$ and $E_j$ are known.
+Introducing false negatives means that $O(j)$ is now random, even if $B_j$ and $E_j$ are known.
 Define the relevant testing times for episode $j$ as $\sched'_j = \{ t \in \sched_{i(j)} \ssep r_j^{(b)} \leq t \leq r_j^{(e)} + 1 \}$, let $m_j$ denote the size of this set.
-Let $O'(j) = [O(j), \vec{y}_j]^T$ where $\vec{y}_j$ is a binary vector of length $m_j$, with the $l$th element of of $\vec{y}_j$ being the test result at the chronologically $l$th time $t \in \sched'_j$.
+Let $t_{j,1} < \dots < t_{j,m_j}$ be the elements of $sched'_j$.
+Let $O'(j) = [O(j), \vec{y}_j]^T$ where $\vec{y}_j$ is a binary vector of length $m_j$, with $y_{j,l}$ being the test result at time $t_{j,l}$.
 $O'(j)$ is the observed data for episode $j$.

### 2024-03-29 cis-perf-new-ll.tex
new likelihood in main text

@@
 
-\chapter{Alternative derivation of \cref{E-perf-test:eq:multinomial}} \label{perf-test:sec:alt-likelihood}
+\chapter{Additional details of derivations in \cref{E-perf-test} and \cref{E-imperf-test}}
 
-\section{Attempt 1}
+\section{Defining possible observations} \label{perf-test:sec:conditions-nu-E}
 
-Denote the characteristics of episode $j$ as $\nu_j = (B_j, E_j, i(j))$.
-I now define an integer $N_E$ and a series of sets $\set{E}_1, \dots, \set{E}_{N_E}, \set{E}_u$ such that any possible $\nu_j$ is in exactly one of the classes and each set corresponds to the possible infection episodes that would generate identical observations (conditional on the test schedules).
-
-Define $\set{E}_u$ such that $\nu_h \in \set{E}_u$ if and only if $\nu_j$ would be undetected (if it occurred).
-Hence, $\prob(\nu_j \in \set{E}_u \mid \vec{\theta}) = p_u$.
-
-Otherwise, the sets are defined such that $\nu_j$ and $\nu_{j'}$ are in the same class if and only if $i(j) = i(j')$ and the episodes beginning and end times are censored to be in the same pair of intervals.
-If $i \in \set{D}$ then there is exactly one $k$ such that $\set{E}_k$ is identical to $i$'s admissible region $\alpha_i$.
-% In this case, I write $\alpha_{i(k)}$ to denote this admissible region.
-
-More formally, the sets can be defined in terms of an equivalence relation $\equiv$ and two arbitrary possible episodes $j$ and $j'$.
-Define $\nu_j \equiv \nu_{j'}$ to hold if and only if either both $j$ and $j'$ would be undetected (as defined in \cref{E-perf-test:sec:prob-undetected}) or the following four conditions all hold.
+Let $\vec{\nu} = [l^{(b)}, r^{(b)}, l^{(e)}, r^{e}, i]^T$ be an arbitrary vector of possible observations, $\vec{\nu} \in \set{E}$ if and only if all the following conditions hold.
 \begin{enumerate}
-    \item Both $\nu_j$ and $\nu_{j'}$ would be detected.
-    \item $i(j) = i(j')$.
-    \item $\tau_{\sched_{i(j)}}(B_j) = \tau_{\sched_{i(j')}}(B_{j'})$, where $\tau$ is as defined in \cref{perf-test:eq:tau-def}.
-    \item $\tau_{\sched_{i(j)}}(E_j) = \tau_{\sched_{i(j')}}(E_{j'})$.
+  \item $l^{(b)} 1, r^{(b)}, l^{(e)}, r^{e} + 1 \in \sched_{i}$, that is individual ${i}$ was tested at all these times.
+  \item $l^{(b)} \leq r^{(b)} \leq l^{(e)} \leq r^{(e)}$.
+  \item $r^{(b)} = \min\{t \in \sched_{i} \ssep t \geq l^{(b)}\}$, that is $r^{(b)}$ is the first time at or after $l^{(b)}$ that individual ${i}$ was tested.
+  \item $r^{(e)} = \min\{t \in \sched_{i} \ssep t > l^{(e)}\} - 1$, that is $r^{(e)}$ is the day before the next time after $l^{(e)}$ that individual ${i}$ was tested.
+  \item $1 \leq r^{(b)} \leq T$, that is the episode begins in the period of interest.
 \end{enumerate}
-Then each $\set{E}_k$ is an equivalence class under this relation and the set of all possible $\nu_j$.
-That is, taking any $k = 1, \dots, N_E, u$ then $\nu_j \in \set{E}_k$ and $\nu_{j'} \in \set{E}_k$ if and only if $\nu_j \equiv \nu_{j'}$.
-
-Let $\xi_k$ denote $\prob(\nu_j \in \set{E}_k)$.
-Assume, for tractability, that the events $\nu_j \in \set{E}_k$ and $\nu_{j'} \in \set{E}_{k'}$ are independent for $j \neq j'$.
-Denote by $m_k$ the number of episodes in $\set{E}_k$.
-$\vec{m} = [m_1, \dots m_{N_E}, m_u]^T$ are the observations.
-Then:
-\begin{align}
-  \vec{m}
-  \mid \ntot, \vec{\theta}
-  \dist
-  \MN\left(
-    \ntot,
-    \frac{1}{\Ncis}
-    \begin{bmatrix}
-        \xi_1 \\ \xi_2 \\ \vdots \\ \xi_{N_E} \\ \xi_u
-    \end{bmatrix}
-  \right).
-  \label{perf-test:eq:alt-likelihood}
-\end{align}
-
-If the set $\set{E}_k$ aligns with some $\alpha_i$ then $m_k = 1$ and $\xi_k = p_{ia}$.
-Otherwise, for $k \neq u$, $m_k = 0$.
-Finally, $m_u = n_u$ and $\xi_u = p_u$.
-Therefore, the probability masses of \cref{perf-test:eq:alt-likelihood} and \cref{E-perf-test:eq:multinomial} match.
-
-\section{Attempt 2}
-
-\subsection{Perfect testing}
-
-The test schedule $\sched_i$ is the set of times individual $i$ is tested at, starting from their last test prior to time 1, if it exists, or their time of enrolment otherwise.
-Note that $\sched_i$ includes any tests that occur after $T$.
-
-For any episode $j$, define $O(j)$ as the observations of that episode.
-Test results are deterministic based on the time of the test and the individual's infection status.
-Therefore, $O(j)$ is fully determined by the triplet $(B_j, E_j, i(j))$, $j$'s beginning time, end time, and the individual in which the episode occurs respectively.
-
-For a detected episode, we have $O(j) = [l_j^{(b)}, r_j^{(b)}, l_j^{(e)}, r_j^{e}, i(j)]^T$ (see \cref{E-biology-data:sec:cis-episodes}).
-These are: the individual in which the episode occurs, denoted $i(j)$; the earliest and latest time the episode could have begun, $l_j^{(b)}$ and $r_j^{(b)}$ respectively; and the earliest and latest time the episode could have ended, $l_j^{(e)}$ and $r_j^{(e)}$ respectively.
-
-For an undetected episode $j$, there are no observations; in which case, define $O(j) = \emptyset$.
-
-Define an integer $N_D$ and $\set{D} = \{ \vec{\nu}_1, \dots, \vec{\nu}_{N_D} \}$ as the set of all possible observations of detected episodes; that is, $O(j) \in \set{D}$ if and only if $j$ is a detected infection.
-REF FIG 5.3 HERE
-Let $\vec{\nu}_k = [l^{(b)}_k, r^{(b)}_k, l^{(e)}_k, r^{e}_k, i_k]^T$ be an arbitrary vector if possible observations, $\vec{\nu}_k \in \set{D}$ if and only if all the following conditions hold.
-\begin{enumerate}
-  \item $l^{(b)}_k - 1, r^{(b)}_k, l^{(e)}_k, r^{e}_k + 1 \in \sched_{i_k}$, that is individual ${i_k}$ was tested at all these times.
-  \item $l^{(b)}_k \leq r^{(b)}_k \leq l^{(e)}_k \leq r^{(e)}_k$.
-  \item $r^{(b)}_k = \min\{t \in \sched_{i_k} \ssep t \geq l^{(b)}_k\}$, that is $r^{(b)}_k$ is the first time at or after $l^{(b)}_k$ that individual ${i_k}$ was tested.
-  \item $r^{(e)}_k = \min\{t \in \sched_{i_k} \ssep t > l^{(e)}_k\} - 1$, that is $r^{(e)}_k$ is the day before the next time after $l^{(e)}_k$ that individual ${i_k}$ was tested.
-  \item $1 \leq r^{(b)}_k \leq T$, that is the episode begins in the period of interest.
-\end{enumerate}
-Let $n_k$ denote the number of observations of $\vec{\nu}_k$, $\nnodet$ the latent number of undetected episodes, $\vec{n} = [n_1, \dots, n_D, \nnodet]^T$, and $\ntot = \sum_{i=1}^{N_D} n_i + \nnodet$.
-For $\vec{\nu}_k \in \set{D}$, let $p_k = \prob(O(j) = \vec{\nu}_k \mid \vec{\theta})$, $p_u = \prob(O(j) = \emptyset \mid \vec{\theta})$, and $\vec{p} = [p_1, \dots, p_D, p_u]^T$.
-Assume, for tractability, that the events $O(j) = \vec{\nu}_k$ and $O(j') = \vec{\nu}_{k'}$ are independent for $j \neq j'$; this assumption is discussed in \cref{perf-test:sec:discussion}.
-Then:
-\begin{align}
-  \vec{n} \mid \ntot, \vec{\theta} &\dist \MN(\ntot, \vec{p})
-\intertext{that is:}
-  p(\vec{n} \mid \ntot, \vec{\theta}) &= \frac{\ntot!}{\nnodet!\prod_{k=1}^{N_D} n_k!} p_u^{\nnodet} \prod_{k=1}^{N_D} p_k^{n_k}.
-  \label{perf-test:eq:multinomial-ll}
-\end{align}
-
-In the CIS data, each $n_k$ ($k \neq u$) is observed as either 0 or 1.
-Define $C = \{ k \ssep n_k = 1 \}$.
-Furthermore, note that the support of the multinomial distribution requires that $\nnodet = \ntot - \ndet$.
-Then \cref{perf-test:eq:multinomial-ll} simplifies to:
-\begin{align}
-  p(\vec{n} \mid \ntot, \vec{\theta})
-  &= \frac{\ntot!}{(\ntot - \ndet)!} p_u^{\ntot-\ndet} \prod_{k \in \set{C}} p_k.
-  \label{perf-test:eq:multinomial}
-\end{align}
-
-This is the previous likelihood where each $p_{ia}$ corresponds to a $p_k$ and $p_u$ is unchanged.
 
 \subsubsection{Deriving $p_k$}
-
-Decompose $p_k$ as $p_k = \prob(O(j) = \nu_k \mid \vec{\theta}) = p_{ik} \prob(i(j) = i_k \mid \vec{\theta})$ where $p_{ik} = \prob(O(j) = \nu_k \mid i(j) = i_k, \vec{\theta})$.
-This is valid as $\prob(O(j) = \nu_k \mid i(j) \neq i_k) = 0$ due to the condition here being equivalent to equating the two vectors' final elements.
-Assume that each infection episode occurs independently and with equal probability in any individual, \ie $\prob(i(j) = i_k) = 1/\Ncis$ for all $j$ and $k$.
 

### 2024-03-29 cis-perfect-testing.tex
new likelihood in main text

@@
 In this section, I develop a computationally feasible Bayesian statistical model accounting for double interval censoring and undetected episodes.
-In \cref{perf-test:sec:data-notation}, I introduce the data and notation, defining which quantities are observed and which are latent.
-Then, in \cref{perf-test:sec:episode-types}, I adapt the framework of \textcite{heiseyModelling} which classifies possible episode beginning and end time into \emph{admissible} episodes, which match the doubly interval censored pattern of the observations; \emph{undetected} episodes, which may have occurred and been undetected; and \emph{inadmissible} episodes, which the data rules out having occurred.
-The number of such episodes in each class has a multinomial distribution conditional on the latent total number of infections (including undetected infections).
+In \cref{perf-test:sec:data}, I introduce the data, defining which quantities are observed and which are latent.
+Then, in \cref{perf-test:sec:episode-types}, I build a framework, using the concept of \textcite{heiseyModelling}, which classifies possible episode beginning and end time into \emph{admissible} regions, each of which match a possible pattern of the doubly interval censored pattern of the observations; and an \emph{undetected} region, which lead to the episode being undetected.
+The number of such episodes in each region has a multinomial distribution conditional on the latent total number of infections (including undetected infections).
 
-In \cref{perf-test:sec:posterior-density}, I outline the form of the posterior density.
+In \cref{perf-test:sec:posterior}, I outline the form of the posterior density.
 The following subsections then derive its various components.
@@
 
-\subsection{Data and notation} \label{perf-test:sec:data-notation}
+\subsection{Data} \label{perf-test:sec:data}
 
@@
 An individual $i$ has a \emph{test schedule} $\sched_i$.
-The test schedule $\sched_i$ is the set of times individual $i$ is tested at, from their last test prior to time 1 (if it exists) up until time $T$.
-These are all the times an infection episode can be detected at, plus one prior test.
+The test schedule $\sched_i$ is the set of times individual $i$ is tested at, starting from their last test prior to time 1, if it exists, or their time of enrolment otherwise.
+Note that $\sched_i$ includes any tests that occur after $T$.
 For example, for the individual in \cref{perf-test:fig:double-interval-censor}, $\sched_i = \{ 0, 7, 14, 21, 28, 56 \}$.
@@
 
-For a detected episode $j$, five pieces of data are observed (see \cref{E-biology-data:sec:cis-episodes} for how these are constructed).
-The individual in which the episode occurs, denoted $i(j)$; the earliest and latest time the episode could have begun, $l_j^{(b)}$ and $r_j^{(b)}$ respectively; and the earliest and latest time the episode could have ended, $l_j^{(e)}$ and $r_j^{(e)}$ respectively.
-Over the short calendar period considered, each individual has zero or one detected episodes.
-Let $i \in \set{D}$ indicate individual $i$ has a detected episode and denote $j(i)$ as the detected episode in individual $i$; $j(i)$ is only defined for $i \in \set{D}$.
+For any episode $j$, define $O(j)$ as the observations of that episode.
+Test results are deterministic based on the time of the test and the individual's infection status.
+Therefore, $O(j)$ is fully determined by the triplet $(B_j, E_j, i(j))$ consisting of $j$'s beginning time, end time, and the individual in which the episode occurs respectively.
 
-In addition, the individuals in the cohort experienced a total of $\nnodet$ undetected infection episodes.
-$\nnodet$ is a latent quantity.
-The total number of infection episodes in the cohort is $\ntot = \ndet + \nnodet$ and I index them with $j = 1, \dots, \ntot$.
+For a detected episode, we have $O(j) = [l_j^{(b)}, r_j^{(b)}, l_j^{(e)}, r_j^{e}, i(j)]^T$.
+These are: the individual in which the episode occurs, denoted $i(j)$; the earliest and latest time the episode could have begun, $l_j^{(b)}$ and $r_j^{(b)}$ respectively; and the earliest and latest time the episode could have ended, $l_j^{(e)}$ and $r_j^{(e)}$ respectively.
+How the episode data are constructed is detailed in \cref{E-biology-data:sec:cis-episodes}.
+
+For an undetected episode $j$, there are no observations; in which case, define $O(j) = \emptyset$.
+Let $\nnodet$ denote the (latent) number of undetected episodes experienced across all $\Ncis$ individuals.
+Therefore, the total number of infection episodes in the cohort is $\ntot = \ndet + \nnodet$ and I index them with $j = 1, \dots, \ntot$.
+
+\subsection{Posterior density} \label{perf-test:sec:posterior}
 
@@
 
-\subsection{Types of episodes} \label{perf-test:sec:episode-types}
+Define an integer $N_E$ and $\set{E} = \{ \vec{\nu}_1, \dots, \vec{\nu}_{N_E} \}$ as the set of all possible observations of detected episodes; that is, $O(j) \in \set{E}$ if and only if $j$ is a detected infection.
+Let $\vec{\nu}_k = [l^{(b)}_k, r^{(b)}_k, l^{(e)}_k, r^{e}_k, i_k]^T$ be an arbitrary vector of possible observations, visualised in \cref{perf-test:fig:partitionSpace}; the conditions for $\vec{\nu}_k \in \set{E}$ are in \cref{perf-test:sec:conditions-nu-E}.
 
-To relate the possible value for $b_j$ and $e_j$, given a test schedule $\sched_{i(j)}$, to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie are strict subsets of $T \times T$), shown in \cref{perf-test:fig:partitionSpace}.
-Each episode, whether detected or not, must fall into one of these three classes with a probability that depends on the parameters of the survival function, $\vec{\theta}$, specified in \cref{perf-test:sec:parameters-priors}.
-The regions are defined for each individual $i$.
+The target of inference is $\vec{\theta}$, the parameters of the survival function $S_{\vec{\theta}}(t)$.
+The contents of $\vec{\theta}$ are discussed in \cref{perf-test:sec:parameters-priors}.
 
-\begin{itemize}
-\item
-  The region of admissible episodes for individual $i$, $\alpha_i$.
-  An admissible episode has an infection and recovery time consistent with the data observed.
-  Let $n_{ia}$ be the number of admissible episodes for $i$.
-  $\alpha_i$ is empty if $i$ had no detected episodes (\ie $i \notin \set{D}$); in this case the observation is $n_{ia} = 0$.
-  Otherwise, if $i \in \set{D}$, then $n_{ia} =1$.
-  The admissible region is the intersection of the intervals $[l_{j(i)}^{(b)}, r_{j(i)}^{(b)}]$ and $[l_{j(i)}^{(e)}, r_{j(i)}^{(e)}]$.
-  This is the purple area in \cref{perf-test:fig:partitionSpace}; the boundaries in the $x$ direction of the area are $l_j^{(b)}$ and $r_j^{(b)}$ and in the $y$ direction are $l_j^{(e)}$ and $r_j^{(e)}$.
-  The shape of this region accounts for the doubly interval censored nature of the data.
-  Let $p_{ia} = \prob((B_j, E_j) \in \alpha_i \mid \vec{\theta}, i(j) = i)$, the probability that $j \in \alpha_i$ given that $j$ occurs in individual $i$.
-  \label{perf-test:def:admissible}
-\item
-  The region of undetected episode for individual $i$, $\Omega_i$.
-  Let $n_{iu}$ be the (latent) number of undetected episodes for $i$.
-  The undetected region is the red region in \cref{perf-test:fig:partitionSpace}, indicating combinations of episode beginning and end times which do not span a testing time plus episodes beginning before the individual's enrolment.
-  Let $p_{iu} = \prob((B_j, E_j) \in \Omega_i \mid \vec{\theta}, i(j) = i)$, the probability that $j \in \Omega_i$ given that $j$ occurs in individual $i$.
-\item
-  The region of inadmissible episodes, $\beta_i$ for individual $i$.
-  An inadmissible episode is one which would have been detected had it occurred, but the data rule out having occurred.
-  Let $n_{i\bar{a}}$ be the number of inadmissible episodes for $i$, which is observed as $n_{i\bar{a}} = 0$.
-  These are all remaining (a priori possible) episodes not in the previous sets.
-  Let $p_{i\bar{a}} = \prob((B_j, E_j) \in \beta_i \mid \vec{\theta}, i(j) = i)$, the probability that $j \in \beta_i$ given that $j$ occurs in individual $i$.
-\end{itemize}
-
+Let $n_k$ denote the number of observations of $\vec{\nu}_k$; $\nnodet$ denote the latent number of undetected episodes; and $\vec{n} = [n_1, \dots, n_{N_D}, \nnodet]^T$.
+Hence, $\ntot = \sum_{i=1}^{N_D} n_i + \nnodet$.
+For $\vec{\nu}_k \in \set{E}$, let $p_k = \prob(O(j) = \vec{\nu}_k \mid \vec{\theta})$; $p_u = \prob(O(j) = \emptyset \mid \vec{\theta})$; and $\vec{p} = [p_1, \dots, p_D, p_u]^T$.
+Assume, for tractability, that the events $O(j) = \vec{\nu}_k$ and $O(j') = \vec{\nu}_{k'}$ are independent for $j \neq j'$; this assumption is discussed in \cref{perf-test:sec:discussion}.
+Then:
+\begin{align}
+  \vec{n} \mid \ntot, \vec{\theta} &\dist \MN(\ntot, \vec{p})
+\intertext{that is:}
+  p(\vec{n} \mid \ntot, \vec{\theta}) &= \frac{\ntot!}{\nnodet!\prod_{k=1}^{N_D} n_k!} p_u^{\nnodet} \prod_{k=1}^{N_D} p_k^{n_k}.
+  \label{perf-test:eq:multinomial-ll}
+\end{align}
 \begin{figure}
@@
 
-These three regions span all possible episodes and are mutually exclusive.
-Hence the following relations hold:
+In the CIS data, each $n_k$ ($k \neq u$) is observed as either 0 or 1.
+Define $D = \{ k \ssep n_k = 1 \}$.
+Furthermore, note that the support of the multinomial distribution requires that $\nnodet = \ntot - \ndet$.
+Let $\na = [n_{1}, \dots, n_{N_E}]^T$ be the observed portion of $\vec{n}$.
+Then \cref{perf-test:eq:multinomial-ll} simplifies to:
 \begin{align}
-\ntot
-&= \ndet + \nnodet + \sum_{i=1}^{\Ncis} n_{i\bar{a}} \\
-&= \sum_{i=1}^{\Ncis} n_{ia} + \sum_{i=1}^{\Ncis} n_{iu} + 0
-\end{align}
-and, for all $i$, $p_{ia} + p_{iu} + p_{i\bar{a}} = 1$.
-
-I will use these regions to form the data likelihood, conditional on $\ntot$.
-The definition of the admissible region relies on the test results.
-However the likelihood should be conditional only on $\vec{\theta}$, $\ntot$, and the test schedules.
-Therefore, this is not strictly correct.
-This issue can be rectified by considering the inadmissible region as many subregions, which leads to the same result (sketched in \cref{perf-test:sec:alt-likelihood}).
-A full formalisation of the derivation in \cref{perf-test:sec:alt-likelihood} would require much more complex notation for the same result; therefore, I prefer the presentation to follow.
-
-% Any episode $j$ independently belongs to one of these classes.
-The probabilities $p_{iu}$, $p_{ia}$, and $p_{i\bar{a}}$ are calculated conditional on the episode occurring in individual $i$.
-I assume the individual in which an episode occurs is iid uniform across the cohort, that is $\prob(i(j) = i) = 1/\Ncis$.
-Therefore, the conditional probabilities are multiplied by $1/\Ncis$ to form the probabilities .
-For example, $\prob((B_j, E_j) \in \alpha_i, i(j) = i \mid \vec\theta) = p_{ia} / \Ncis$.
-
-These probabilities apply to each episode independently.
-Therefore:
-\begin{align}
-  \begin{bmatrix}
-    n_{1a} \\ n_{1u} \\ n_{1\bar{a}} \\ \vdots \\ n_{\Ncis,a} \\ n_{\Ncis,u} \\ n_{\Ncis,\bar{a}}
-  \end{bmatrix}
-  \mid \ntot, \vec{\theta}
-  \dist
-  \MN\left(
-    \ntot,
-    \frac{1}{\Ncis}
-    \begin{bmatrix}
-      p_{1a} \\ p_{1u} \\ p_{1\bar{a}} \\ \vdots \\ p_{\Ncis,a} \\ p_{\Ncis,u} \\ p_{\Ncis,\bar{a}}
-    \end{bmatrix}
-  \right)
+  p(\vec{n} \mid \ntot, \vec{\theta})
+  &= p(\na \mid \ntot, \vec{\theta}) \\
+  &= \frac{\ntot!}{(\ntot - \ndet)!} p_u^{\ntot-\ndet} \prod_{k \in \set{D}} p_k.
   \label{perf-test:eq:multinomial}
@@
 
-The undetected regions cannot be distinguished.
-Therefore, I combine these regions into a single cell within the multinomial distribution.
-The combined class has a count $\nnodet = \sum_{i=1}^{\Ncis} n_{iu}$ and probability parameter $\pnodet = \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} p_{iu}$.
-
-\subsection{Posterior density} \label{perf-test:sec:posterior-density}
-
-Let $\na = (n_{1a}, \dots, n_{\Ncis a})^T$ be the vector of counts of episodes in the admissible classes.
-This represents all the observations used in \cref{perf-test:eq:multinomial}.
-Therefore, the posterior density we are interested in is $p(\vec{\theta} \mid \na)$.
-It can be written as follows.
+The relevant information from the CIS data is fully contained in the vector $\na$.
+Therefore, the posterior of interest is:
 \begin{align}
@@
 &= p(\vec\theta) \sum_{\ntot= \ndet}^{\infty} p(\ntot, \na \mid \vec{\theta}) \\
-&= p(\vec{\theta}) \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) p(\na \mid \ntot, \vec{\theta})
+&= p(\vec{\theta}) \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) p(\na \mid \ntot, \vec{\theta}) \\
+&= p(\vec{\theta}) \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{k \in \set{D}} p_k &\text{by \cref{perf-test:eq:multinomial}} \\
+&= p(\vec{\theta}) \left( \prod_{k \in \set{D}} p_k \right) \left( \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \right)
 \label{perf-test:eq:posterior1}
-\end{align}
-where the last line augments the likelihood with the total number of episodes, $\ntot$.
-
-First, I derive the form of $p(\na \mid \ntot, \vec{\theta})$.
-From \cref{perf-test:sec:episode-types}, the counts $\na$ and $\nnodet$'s joint distribution, conditional on $\ntot$, is multinomial.
-All $n_{i\bar{a}} = 0$, therefore these classes can be ignored; counts of zero do not contribute to the multinomial probability mass function (they multiply by 1).
-The support of the distribution requires $\nnodet = \ntot - \sum_{i=1}^{\Ncis} n_{ia} = \ntot - \ndet$ and $\ntot \geq \ndet$.
-Hence:
-\begin{align}
-p(\na \mid \ntot, \vec{\theta})
-&= p(\nnodet  = \ntot - \na, \na\mid \ntot, \vec{\theta}) \\
-&= \frac{\ntot!}{(\ntot - \ndet)! \prod_{i=1}^{\Ncis} n_{ia}!}  \pnodet^{\ntot - \ndet}\prod_{i=1}^{\Ncis} \left(\frac{1}{\Ncis}p_{ia}\right)^{n_{ia}} \\
-&\propto \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i \in \set{D}} p_{ia} \label{perf-test:eq:augmented-likelihood}
-\end{align}
-the final line follows because $n_{ia} = 1$ for $i \in \set{D}$ and $n_{ia} = 0$ otherwise.
-
-Substituting \cref{perf-test:eq:augmented-likelihood} into \cref{perf-test:eq:posterior1}:
-\begin{align}
-p(\vec{\theta} \mid \na)
-&\propto p(\vec{\theta}) \sum_{\ntot=\ndet}^\infty \left( p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i \in \set{D}} p_{ia} \right) \\
-&= p(\vec{\theta}) \left( \prod_{i \in \set{D}} p_{ia} \right) \left( \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \right) \label{perf-test:eq:posterior2}.
 \intertext{For convenience, define the summation term as:}
@@
 \end{align}
-The rest of this section derives expressions for each of $p_{ia}$, $p_{u}$ and $\eta$.
+The rest of this section derives expressions for each of $p_{k}$, $p_{u}$ and $\eta$.
 
@@
 
-Substituting this into \cref{perf-test:eq:posterior2} gives:
+Substituting this into \cref{perf-test:eq:posterior1} gives:
 \begin{align}
 p(\vec{\theta} \mid \na)
-&\propto p(\vec{\theta}) \left( \prod_{i \in \set{D}} p_{ia} \right) (r + \mu (1- \pnodet))^{-(r+\ndet)} \label{perf-test:eq:full-posterior}.
+&\propto p(\vec{\theta}) \left( \prod_{i \in \set{D}} p_k \right) (r + \mu (1- \pnodet))^{-(r+\ndet)} \label{perf-test:eq:full-posterior}.
 \end{align}
@@
 
-\subsection{Admissible probabilities}
+\subsection{Deriving $p_k$}
 
-Next, I derive $p_{ia}$.
-From \cref{perf-test:eq:full-posterior}, this is only needed for $i \in \set{D}$, that is for individuals with detected episodes.
+Recall the definitions $p_k = \prob(O(j) = \vec{\nu}_k \mid \vec{\theta})$ where $O(j)$ are the observations of episode $j$ and $\vec{\nu}_k = [l^{(b)}_k, r^{(b)}_k, l^{(e)}_k, r^{e}_k, i_k]^T$.
 
-By definition (see \cpageref{perf-test:def:admissible}), we have, for any $i \in \set{D}$:
+Decompose $p_k$ as $p_k = p_{ik} \prob(i(j) = i_k \mid \vec{\theta})$
+where $p_{ik} = \prob(O(j) = \nu_k \mid i(j) = i_k, \vec{\theta})$.
+% This is valid as $\prob(O(j) = \nu_k \mid i(j) \neq i_k) = 0$ due to the condition here being equivalent to equating the two vectors' final elements.
+Assume that each infection episode occurs independently and with equal probability in any individual, \ie $\prob(i(j) = i_k) = 1/\Ncis$ for all $j$ and $k$.
+
+If $i(j) = i_k$ then the event $O(j) = \vec{\nu}_k$ occurs if and only if the episode starts in the interval $[l^{(b)}_k, r^{(b)}_k]$ and ends in the interval $[l^{(e)}_k, r^{e}_k]$.
+Omitting the conditioning on $\vec{\theta}$ and $i(j) = i$, this gives:
 \begin{align}
-p_{ia} =& \prob((B_j, E_j) \in \alpha_i \mid \vec{\theta}, i(j) = i) \\
-\alpha_i =& \{ (b, e) : l_{j(i)}^{(b)} \leq b \leq r_{j(i)}^{(b)} \wedge l_{j(i)}^{(e)} \leq e \leq r_{j(i)}^{(e)}\}.
-\intertext{Omitting the conditioning on $\vec{\theta}$ and $i(j) = i$, this gives:}
-p_{ia}
-=& \prob \left( l_{j(i)}^{(b)} \leq B_{j} \leq r_{j(i)}^{(b)}, l_{j(i)}^{(e)} \leq E_{j} \leq r_{j(i)}^{(e)} \right) \\
-=& \prob \left( l_{j(i)}^{(e)} \leq E_{j} \leq r_{j(i)}^{(e)} \mid l_{j(i)}^{(b)} \leq B_{j} \leq r_{j(i)}^{(b)} \right) \\
-   &\times\prob \left( l_{j(i)}^{(b)} \leq B_{j} \leq r_{j(i)}^{(b)} \right) \\
-=& \sum_{b = l_{j(i)}^{(l)}}^{r_{j(i)}^{(b)}} \prob \left( l_{j(i)}^{(e)} \leq E_{j} \leq r_{j(i)}^{(e)} \mid B_{j} = b \right) \prob \left(B_{j} = b \right) \\
-=& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \prob \left( l_{j(i)}^{(e)} - b + 1 \leq D_{j} \leq r_{j(i)}^{(e)} - b + 1 \right) \prob \left(B_{j} = b \right) &\text{by def of $D_{j}$} \\
-=& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_{j(i)}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 2) \right) \prob \left(B_{j} = b \right) &\text{by def of $S_{\vec{\theta}}$} \\
-\propto& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_{j(i)}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 2) \right)
+p_{ik}
+=& \prob \left( l_k^{(b)} \leq B_{j} \leq r_k^{(b)}, l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \right) \\
+=& \prob \left( l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \mid l_k^{(b)} \leq B_{j} \leq r_k^{(b)} \right) \times\prob \left( l_k^{(b)} \leq B_{j} \leq r_k^{(b)} \right) \\
+=& \sum_{b = l_k^{(l)}}^{r_k^{(b)}} \prob \left( l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \mid B_{j} = b \right) \prob \left(B_{j} = b \right) \\
+=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \prob \left( l_k^{(e)} - b + 1 \leq D_{j} \leq r_k^{(e)} - b + 1 \right) \prob \left(B_{j} = b \right) &\text{by def of $D_{j}$} \\
+=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right) \prob \left(B_{j} = b \right) &\text{by def of $S_{\vec{\theta}}$} \\
+\propto& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right)
 \label{perf-test:eq:pia}
@@
 
-\subsection{Undetected probability} \label{perf-test:sec:prob-undetected}
+\subsection{Deriving $1 - p_u$} \label{perf-test:sec:prob-undetected}
 
 The final component of \cref{perf-test:eq:full-posterior} required is $1 - p_u$.
-From \cref{perf-test:sec:episode-types}:
+Recall the definition $p_u = \prob(O(j) = \emptyset \mid \vec{\theta})$.
+Therefore:
 \begin{align}
   1 - p_u
-  &= 1 - \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} p_{iu} \\
-  &= \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} (1 - p_{iu}).
+  &= 1 - \sum_{i=1}^{\Ncis} \prob(O(j) = \emptyset, i(j) = i \mid \vec{\theta}) \\
+  &= 1 - \sum_{i=1}^{\Ncis} \prob(O(j) = \emptyset \mid i(j) = i \mid \vec{\theta}) P(i(j) = i \mid \vec{\theta}) \\
+  &= 1 - \frac{1}{\Ncis}\sum_{i=1}^{\Ncis} \prob(O(j) = \emptyset \mid i(j) = i \mid \vec{\theta}) \\
+  &= \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} (1 - \prob(O(j) = \emptyset \mid i(j) = i \mid \vec{\theta}))
   \label{perf-test:eq:pu}
 \end{align}
-
-Therefore, the crucial component is $1 - p_{iu} = 1 - \prob((B_j, E_j) \in \Omega_i \mid \vec{\theta}, i(j) = i)$.
-This is one minus the probability that an episode in individual $i$ was undetected, \ie the probability of detecting an episode.
+assuming $P(i(j) = i \mid \vec{\theta})$ as before.
+Let $p_{iu} = \prob(O(j) = \emptyset \mid i(j) = i \mid \vec{\theta})$.
+Therefore, the crucial component is $1 - p_{iu}$.
+This is one minus the probability that an episode in individual $i$ was undetected, \ie the probability of the episode being detected.
 
@@
     \item $\exists t \in \sched_{i(j)}$ such that $b_j \leq t \leq e_j$; this condition is equivalent to having at least one positive test for the episode.
-    \item $\exists t \in \sched_{i(j)}$ such that $t < b_j$.
+    \item $b_j > \min \sched_{i(j)}$.
       For individuals enrolled during the period considered ($\min \sched_{i(j)} > 0$), this ensures that the beginning of the episode is lower bounded; thereby adjusting for those who enrolled during the period being less likely to have a detected infection.
       For individuals enrolled prior to the period considered ($\min \sched_{i(j)} \leq 0$), this means that the episode was not detected prior to time 1.
-    \item $\exists t \in \sched_{i(j)}$ such that $t \geq b_j$.
-      This means that the test is detected before $T$.
+    \item $b_j \leq T_{i(j)}$ where $T_{i(j)} = \max \{ t \in \sched_{i(j)} \ssep t \leq T \}$ is the last time that $i(j)$ is tested in the period, meaning that the test is detected within the period.
     \item Individual $i(j)$ is tested at a time $t > e_j$, upper bounding the end of the episode.
@@
 \end{align}
-defining $\min \emptyset = \infty$; \ie $\tau_{\sched_i}(t) = \infty$ if there is no $t' \in \sched_i$ such that $t' \geq t$.
+% defining $\min \emptyset = \infty$; \ie $\tau_{\sched_i}(t) = \infty$ if there is no $t' \in \sched_i$ such that $t' \geq t$.
 The first condition can now be expressed as $e_j \geq b_j + \tau_{\sched_{i(j)}}(b_j)$.
 Equivalently, $d_j \geq \tau_{\sched_{i(j)}}(b_j) + 1$.
-The fourth condition can be expressed as $\tau_{\sched_{i(j)}}(b_j) < \infty$.
+% The fourth condition can be expressed as $\tau_{\sched_{i(j)}}(b_j) < \infty$.
 
@@
 1 - p_{iu}
-&= 1 - \prob((B_{j}, E_{j}) \in \Omega_i) \\
-&= \prob(D_j \geq \tau_{\sched_{i}}(B_j)+ 1, \tau_{\sched_{i(j)}}(b_j) < \infty, B_j > \min \sched_{j}) \\
-&= \sum_{b = \min \sched_{i} + 1}^T \prob(D_j \geq \tau_{\sched_{i}}(b) + 1, \tau_{\sched_{i(j)}}(b_j) < \infty \mid B_j = b) \prob(B_j = b)\\
-&= \frac{1}{T} \sum_{b = \min \sched_{i} + 1}^T S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1)
+% &= 1 - \prob((B_{j}, E_{j}) \in \Omega_i) \\
+&= \prob(D_j \geq \tau_{\sched_{i}}(B_j)+ 1, \min \sched_{i(j)} < B_j \leq T_{i(j)}) \\
+&= \sum_{b = \min \sched_{i} + 1}^{T_{i(j)}} \prob(D_j \geq \tau_{\sched_{i}}(b) + 1, \tau_{\sched_{i(j)}}(b_j) < \infty \mid B_j = b) \prob(B_j = b)\\
+&= \frac{1}{T} \sum_{b = \min \sched_{i} + 1}^{T_{i(j)}} S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1).
 \label{perf-test:eq:piu}
 \end{align}
-defining $S_{\vec{\theta}}(\infty) = 0$.
 
@@
 1 - p_u
-& \propto \sum_{i=1}^{\Ncis} \sum_{b = \min \sched_{i} + 1}^T S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1)
+& \propto \sum_{i=1}^{\Ncis} \sum_{b = \min \sched_{i} + 1}^{T_{i(j)}} S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1).
 \end{align}

### 2024-03-29 cis-imperfect-testing.tex
proof
new likelihood in ch 6

@@
 
-In \cref{E-perf-test}, a framework for analysing the CIS to produce estimates of the duration of positivity was developed.
+In \cref{E-perf-test}, a framework for analysing the CIS to produce estimates of the duration of detectability was developed.
 However, the initial application of this framework produced an implausibly short estimate.
@@
 A single positive episode is an episode containing exactly one positive test; intuitively, these cause shorter estimates because the lower bound on the length of time the episode is one day, the shortest possible.
-Specifically, if $j$ is a single positive episode then $j$'s admissible region (as defined in \cref{E-perf-test:sec:model}) includes $b_j = e_j = r_j^{(b)} = l_j^{(e)}$.
+Specifically, if $j$ is a single positive episode then $O(j)$ allows $b_j = e_j = r_j^{(b)} = l_j^{(e)}$.
 If this is true, then the duration is $d_j = e_j - b_j + 1 = 1$.
@@
 
-First, in \cref{imperf-test:sec:modifying-p_ia}, I modify $p_{ia}$, the probability of an episode with test schedule $i$ being admissible, to allow for the episode possibly being longer than observed.
-Then, in \cref{imperf-test:sec:modifying-p_iu}, I modify $p_{iu}$, the probability of an episode with test schedule $i$ being undetected, to allow for additional episodes being undetected.
+I start, in \cref{imperf-test:sec:notation}, by introducing some additional notation needed.
+I then modify the components of the likelihood in \cref{imperf-test:sec:modifying-p_ia} and \cref{imperf-test:sec:modifying-p_iu} to incorporate false negatives.
 
-\subsection{Modifying \texorpdfstring{$p_{ia}$}{pia}} \label{imperf-test:sec:modifying-p_ia}
+\subsection{Notation} \label{imperf-test:sec:notation}
 
-I will modify $p_{ia}$ to allow the negative test following the last positive to be a false negative.
+Introducing false negatives means that $O(j)$, the observation vector of episode $j$, is now random, even if $B_j$ and $E_j$ are known.
+Therefore, the results between the tests that bound episode $j$'s beginning and end times should now enter into the likelihood.
+For tractability, I consider only tests between the negative tests providing an upper bound on the length of episode $j$.
+
+To formalise this, define the relevant testing times for episode $j$ as $\sched'_j = \{ t \in \sched_{i(j)} \ssep r_j^{(b)} \leq t \leq r_j^{(e)} + 1 \}$ and let $m_j$ denote the size of this set.
+Denote the elements of $\sched'_j$ by $t_{j,1} < \dots < t_{j,m_j}$.
+The new observation vector for $j$ is $O'(j) = [O(j), \vec{y}_j]^T$ where $\vec{y}_j$ is a binary vector of length $m_j$, with $y_{j,l}$ being the test result at time $t_{j,l}$.
+
+Defining analogous variables to \cref{E-perf-test}, let $\set{D}' = \{ \vec{\nu}'_1, \dots, \vec{\nu}'_{N_{D'}} \}$ where $\vec{\nu}'_k \in \set{D}'_k$ if and only if $\vec{\nu}'_k = [\vec{\nu}_k, \vec{y}_k]^T$ would be a detected observation, equivalent to satisfying all the following conditions.
+\begin{enumerate}
+  \item $\vec{\nu}_k \in \set{D}$.
+  \item $\vec{y}_k \in \{0, 1\}^{m_k}$.
+  \item The elements of $\vec{y}_k$ corresponding to the tests at times $r_k^{(b)}$ (its first element) and $l_k^{(e)}$ (its penultimate element) are 1.
+  \item The element of $\vec{y}_k$ corresponding to the test at time $r_k^{(e)} + 1$ (its last element) is 0.
+\end{enumerate}
+The final two conditions are due the construction of the intervals as positive and negative tests bounding the beginning and end times of the episode.
+
+Similarly, let $p_k'$, $p_{ik}'$, $p_u$, and $p_{iu}$ have the same definition as $p_k$, $p_{ik}$, $p_u$, and $p_{iu}$ respectively except replacing $O$ with $O'$ and $\vec{\nu}_k$ with $\vec{\nu}_k'$, and conditioning on $\psens$.
+As before, $p'_k = 1/\Ncis p'_{i_kk}$ and $p'_u = 1/\Ncis \sum_{i=1}^{\Ncis} p'{iu}$.
+
+\subsection{Deriving $p'_{ik}$} \label{imperf-test:sec:modifying-p_ia}
+
+I will modify $p_{ik}$ to form $p_{ik}' = \prob(O'(j) = \nu'_k \mid i(j) = i_k, \vec{\theta})$, which allows the negative test following the last positive to be a false negative.
 If it is a false negative, then I will consider the episode's end right censored.
@@
 
-I start by finding the tests that need consideration for episode $j$.
-For tractability and simplicity, I consider only tests between the negative tests providing an upper bound on the length of episode $j$.
-By definition (see \cref{E-biology-data:sec:cis-episodes}), these are the tests in individual $i(j)$ between $l_j^{(b)} - 1$ and $r_j^{(e)} + 1$ inclusive.
+% I proceed by assuming that we have observed an arbitrary episode:
+% \begin{align}
+%   O'(j) = [[l_k^{(b)}, r_k^{(b)}, l_k^{(e)}, r_k^{(e)}, i(j)]^T, \vec{y}_k]^T.
+% \end{align}
+% I derive $p_k$ by considering the generating process that could have produced this episode.
 
-For tractability, assume that the negative test bounding the start of the episode, on day $l_j^{(b)}-1$, is a true negative.
-This assumption is reasonable because, since a positive test follows at $r_j^{(b)}$, the negative at $l_j^{(b)}-1$ is likely early in the infection and the test sensitivity is high early in an infection.
+For tractability, assume that the negative test bounding the start of the episode, on day $l_k^{(b)}-1$, is a true negative.
+This assumption is reasonable because, since a positive test follows at $r_k^{(b)}$, the negative at $l_k^{(b)}-1$ is likely early in the infection and the test sensitivity is high early in an infection.
 True negatives occur with probability 1, and hence this test does not contribute to the likelihood.
 
-Denote the set of the remaining test times to consider as $\sched'_{i(j)}$, the set of times individual $i(j)$ was tested between $l_j^{(b)}$ and $r_j^{(e)} + 1$ inclusive.
-Denote their results by $\vec{y}_j = \{ y_{i(j)}(t) \ssep t \in \sched'_{i(j)} \}$ where $y_{i(j)}(t) = 1$ if the test on day $t$ in individual $i(j)$ is positive and 0 otherwise.
+As I assume that there are no false positives, the infection episode must span at least the period $[r^{(b)}_k, l^{(e)}_k]$, a period starting and ending with a positive test.
+This includes all $t \in \sched'_{i(j)}$ except $t = r_k^{(e)}+1$.
+Therefore, the test results $\vec{y}_k$, except the test at $r_k^{(e)}+1$, are either true positives or false negatives.
+This gives $t_+ = \sum_{l=1}^{m_k} y_{k,l}(t)$ true positives and $f_- = m_k - t_{k,+}$ false negatives.
 
-As I assume that there are no false positives, the infection episode must span at least the period $[r^{(b)}_j, l^{(e)}_j]$, because this period starts and ends with a positive test.
-This includes all $t \in \sched'_{i(j)}$ except $t = r_j^{(e)}+1$.
-Therefore, the test results $\vec{y}_j$, except the test at $r_j^{(e)}+1$, are either true positives or false negatives.
-This gives $t_+ = \sum_{t \in \sched'_{i(j)}} y_{i(j)}(t)$ true positives and $f_- = \sum_{t \in \sched'_{i(j)}} (1 - y_{i(j)}(t))$ false negatives.
-
-Consider the negative test at $r_j^{(e)}+1$, the first negative after the start of the episode which may be a false negative.
-It is a false negative if and only if the episode ends after the test, \ie $E_j > r_j^{(e)}$.
+Consider the negative test at $r_k^{(e)}+1$, the first negative after the start of the episode which may be a false negative.
+It is a false negative if and only if the episode ends after the test, \ie $E_j > r_k^{(e)}$.
 I proceed by considering whether this is the case and with $B_j = b$ known.
 
-First, the case when $E_j \leq r_j^{(e)}$.
-In this case, the test at $r_j^{(e)}+1$ is a true negative and the end of the episode is interval censored as in the previous chapter.
-% In this case, the test at $r_j^{(e)} + 1$ is a true negative, as are all other tests not in $\sched'_{i(j)}$.
+First, the case when $E_j \leq r_k^{(e)}$.
+In this case, the test at $r_k^{(e)}+1$ is a true negative and the end of the episode is interval censored as in the previous chapter.
+% In this case, the test at $r_k^{(e)} + 1$ is a true negative, as are all other tests not in $\sched'_{i(j)}$.
 The true negative occurs with probability 1, by the assumption of no false positives.
 \begin{align}
-&\prob(\vec{y_j}, E_j \leq r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
-&= \prob(\vec{y_j}, l_j^{(e)} \leq E_j \leq r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\ % &\text{as no false positives}
-&= \prob(\vec{y_j} \mid l_j^{(e)} \leq E_j \leq r_j^{(e)}, B_j = b, p_\text{sens}, \vec{\theta}) \prob(l_j^{(e)} \leq E_j \leq r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
-&= p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} \left( S_{\vec{\theta}}(l_j^{(e)} - b + 1) - S_{\vec{\theta}}(r_j^{(e)} - b + 2) \right)
+&\prob(\vec{y}_k, E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
+&= \prob(\vec{y}_k, l_k^{(e)} \leq E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\ % &\text{as no false positives}
+&= \prob(\vec{y}_k \mid l_k^{(e)} \leq E_j \leq r_k^{(e)}, B_j = b, p_\text{sens}, \vec{\theta}) \prob(l_k^{(e)} \leq E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
+&= p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right)
 \label{imperf-test:eq:ll-ei-lt-ri}
@@
 
-Second, the case when $E_j > r_j^{(e)}$.
-In this case, the test at $r_j^{(e)}$ is a false negative, occurring with probability $(1 - p_\text{sens})$.
-To avoid having to consider tests after $r_j^{(e)}$, which could greatly complicate the likelihood, I model this case as the episode being right censored at $r_j^{(e)}$.
+Second, the case when $E_j > r_k^{(e)}$.
+In this case, the test at $r_k^{(e)}$ is a false negative, occurring with probability $(1 - p_\text{sens})$.
+To avoid having to consider tests after $r_k^{(e)}$, which could greatly complicate the likelihood, I model this case as the episode being right censored at $r_k^{(e)}$.
 Taking the same approach as before:
 \begin{align}
-&\prob(\vec{y_j}, E_j > r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
-&= \prob(\vec{y_j} \mid E_j > r_j^{(e)}, B_j = b, p_\text{sens}, \vec{\theta}) \prob(E_j > r_j^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
-&= p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} (1 - p_\text{sens}) S_{\vec{\theta}}(r_j^{(e)} - b_j + 2)
+&\prob(\vec{y_k}, E_j > r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
+&= \prob(\vec{y_k} \mid E_j > r_k^{(e)}, B_j = b, p_\text{sens}, \vec{\theta}) \prob(E_j > r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
+&= p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} (1 - p_\text{sens}) S_{\vec{\theta}}(r_k^{(e)} - b_j + 2)
 \label{imperf-test:eq:ll-ei-gt-ri}
@@
 
-These expressions can now be used to form the full replacement for $p_{ia}$, $p'_{ia}$, the probability of observing $\vec{y_j}$ given that an infection occurs in individual $i$.
+These expressions can now be used to form the full replacement for $p_{ia}$, $p'_{ia}$, the probability of observing $\vec{y_k}$ given that an infection occurs in individual $i$.
 First, augment the data with $b_j$, and split into the cases just discussed:
 \begin{align}
-p_{ia}'
-=& \prob(\vec{y_j} \mid p_\text{sens}, \vec{\theta}) \\
-=& \sum_{b = l_j^{(b)}}^{r_{j(i)}^{(b)}} \left( \prob(\vec{y_j}, E_j \leq r_{j(i)}^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) + \prob(\vec{y_j}, E_j > r_{j(i)}^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \right) \\
+p_{ik}'
+=& \prob(\vec{y_k} \mid p_\text{sens}, \vec{\theta}) \\
+=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( \prob(\vec{y_k}, E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) + \prob(\vec{y_k}, E_j > r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \right) \\
   &\times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}).
@@
 =\ &  p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} \\
- & \times \sum_{b = l_j^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_j^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 2) + (1 - p_\text{sens}) S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 2) \right) \\
+ & \times \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) + (1 - p_\text{sens}) S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right) \\ 
   & \times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}) \\
 =\ &  p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} \\
-  & \times \sum_{b_j = l_j^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_j^{(e)} - b_j + 1) - p_\text{sens} S_{\vec{\theta}}(r_{j(i)}^{(e)} - b_j + 2) \right) \\
+  & \times \sum_{b_j = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b_j + 1) - p_\text{sens} S_{\vec{\theta}}(r_k^{(e)} - b_j + 2) \right) \\
   & \times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}).
@@
 \begin{align}
-p_{ia}'
-&\propto \sum_{b = l_j^{(b)}}^{r_{j(i)}^{(b)}} S_{\vec{\theta}}(l_j^{(e)} - b_j + 1) - p_\text{sens} S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 2).
+p_{ik}'
+&\propto \sum_{b = l_k^{(b)}}^{r_k^{(b)}} S_{\vec{\theta}}(l_k^{(e)} - b_j + 1) - p_\text{sens} S_{\vec{\theta}}(r_k^{(e)} - b + 2).
 \label{imperf-test:eq:pia-prime-constant}
@@
 
-\subsection{Modifying \texorpdfstring{$p_{iu}$}{piu}} \label{imperf-test:sec:modifying-p_iu}
+\subsection{Deriving $p'_{iu}$} \label{imperf-test:sec:modifying-p_iu}
 
+I now modify $p_{iu}$ to form $p_{iu}' = \prob(O'(j) = \emptyset \mid i(j) = i_k, \vec{\theta})$, which allows the negative test following the last positive to be a false negative.
 Several mechanisms for episodes being undetected were considered when deriving $p_{iu}$ in \cref{E-perf-test:sec:prob-undetected}.
@@
     \item The episode would have been detected considering only the mechanisms in \cref{E-perf-test:sec:prob-undetected}.
-That is $\min(\sched_{i(j)}) < b_j \leq \max(\sched_{i(j)})$ and $e_j \geq \tau_{\sched_{i(j)}}(b_j) + b_j$.
+That is $\min(\sched_{i(j)}) < b_j \leq T_{i(j)}$ and $e_j \geq \tau_{\sched_{i(j)}}(b_j) + b_j$.
     \item The episode ends in the interval $[\tau_{\sched_{i(j)}}(b_j) + b_j, \tau^2_{\sched_{i(j)}}(b_j) + b_j]$.
@@
 
-The probability of this occurring, conditional on $B_j = b_j > \min(\sched_{i(j)})$ is:
+The probability of this occurring, conditional on $B_j = b$ where $\min(\sched_{i(j)}) < b \leq T_{i(j)}$ is:
 \begin{align}
 &\prob \left(
-    \tau_{\sched_{i(j)}}(b_j) + 1 \leq D_j \leq \tau^2_{\sched_{i(j)}}(b_j) + 1, b_j \leq \max(\sched_{i(j)}),
+    \tau_{\sched_{i(j)}}(b) + 1 \leq D_j \leq \tau^2_{\sched_{i(j)}}(b) + 1
     \mid B_j = b, \vec{\theta} \right) (1 - \psens) \\
-&= \left( S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b_j) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b_j) + 2) \right) (1 - \psens).
+&= \left( S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b) + 2) \right) (1 - \psens).
 \end{align}
-Summing over $b_j$ for $\min \sched_{i(j)} < b_j \leq T$ ($T$ being the last day of the period considered), in the same way as \cref{perf-test:eq:piu}, gives:
+Summing over $b$, in the same way as \cref{perf-test:eq:piu}, gives:
 \begin{align}
-\zeta = (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_{i(j)}) + 1}^T \left( S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b) + 2) \right).
+\zeta = (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_{i(j)}) + 1}^{T_{i(j)}} \left( S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b) + 2) \right).
+\label{imperf-test:eq:zeta}
 \end{align}
 
-Denote the replacement for $p_{iu}$ as $p_{iu}'$.
 $p_{iu}'$ is the probability of episode $i$ being undetected, considering both the mechanisms from \cref{E-perf-test} and the new mechanism here.
@@
 Hence, $p_{iu}'$ is the sum of these, $p_{iu}' = p_{iu} + \zeta$.
-For the posterior density, \cref{E-perf-test:eq:full-posterior}, $1 - p_{iu}'$ is the required quantity.
+As in \cref{E-perf-test:sec:prob-undetected}, $1 - p_{iu}'$ is the required quantity.
 \begin{align}
 1 - p_{iu}'
-&= 1 - p_{iu} - \zeta \\
-&= \frac{1}{T} \sum_{b=\min(\sched_{i(j)}) + 1}^T \left( p_\text{sens} S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b) + 1) + (1 - p_\text{sens}) S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b) + 1)\right).
+=& 1 - p_{iu} - \zeta \\
+=& \frac{1}{T} \sum_{b = \min \sched_{i} + 1}^{T_{i(j)}} S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1) \\
+&- (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_{i(j)}) + 1}^{T_{i(j)}} \left( S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b) + 2) \right) \\
+=& \frac{1}{T} \sum_{b=\min(\sched_{i(j)}) + 1}^{T_{i(j)}} \left( p_\text{sens} S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b) + 1) + (1 - p_\text{sens}) S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b) + 1)\right).
 \label{imperf-test:eq:pit-prime}
 \end{align}
-
-$p_{iu}'$ and $p_{ia}'$ are substituted for $p_{iu}$ and $p_{ia}$ respectively in \cref{E-perf-test:eq:full-posterior}.
-The posterior density is otherwise unchanged.
 

### 2024-03-29 cis-perf-new-ll.tex
new likelihood in ch 6

@@
 
-\subsubsection{Deriving $p_k$}
-
-Then the derivation for $p_{ik}$ proceeds as per the derivation for $p_{ia}$ in the original version, except replacing $j(i)$s with $k$s.
-
 \subsection{Imperfect testing}
 
-Introducing false negatives means that $O(j)$ is now random, even if $B_j$ and $E_j$ are known.
-Define the relevant testing times for episode $j$ as $\sched'_j = \{ t \in \sched_{i(j)} \ssep r_j^{(b)} \leq t \leq r_j^{(e)} + 1 \}$, let $m_j$ denote the size of this set.
-Let $t_{j,1} < \dots < t_{j,m_j}$ be the elements of $sched'_j$.
-Let $O'(j) = [O(j), \vec{y}_j]^T$ where $\vec{y}_j$ is a binary vector of length $m_j$, with $y_{j,l}$ being the test result at time $t_{j,l}$.
-$O'(j)$ is the observed data for episode $j$.
-
-Defining analogous variables to \cref{E-perf-test}, let $\set{D}' = \{ \vec{\nu}'_1, \dots, \vec{\nu}'_{N_{D'}} \}$ where $\vec{\nu}'_k \in \set{D}'_k$ if and only if $\vec{\nu}'_k = [\vec{\nu}_k, \vec{y}_k]^T$ could be a detected observation, equivalent to satisfying all the following conditions.
-\begin{enumerate}
-  \item $\vec{\nu}_k \in \set{D}$.
-  \item $\vec{y}_k \in \{0, 1\}^{m_k}$.
-  \item The elements of $\vec{y}_k$ corresponding to the tests at times $r_k^{(b)}$ (its first element) and $l_k^{(e)}$ (its penultimate element) are 1.
-  \item The element of $\vec{y}_k$ corresponding to the test at time $r_k^{(e)} + 1$ (its last element) is 0.
-\end{enumerate}
-The final two conditions are due the construction of the intervals as positive and negative tests bounding the beginning and end times of the episode.
-
-Similarly, let $p_k'$ and $p_{ik}'$ have the same definition as $p_k$ and $p_{ik}$ except replacing $O$ with $O'$ and $\vec{\nu}_k$ with $\vec{\nu}_k'$.
-
 \end{document}

### 2024-03-29 biology-data.tex
proof

@@
 SARS-CoV-2 is closely related to previous coronaviruses that have caused outbreaks in humans, including SARS-CoV and MERS-CoV.
-Coronaviruses are RNA (Ribonucleic acid) viruses.
-RNA viruses use RNA, rather than DNA (Deoxyribonucleic acid), as their genetic material.
+Coronaviruses are RNA (Ribonucleic acid) viruses which means they use RNA, rather than DNA (Deoxyribonucleic acid), as their genetic material.
 RNA viruses, which include influenza, mutate rapidly and are the most likely class of viruses to cause pandemics~\autocite{woolhouseRNA}.
@@
 The final step, \emph{extension}, raises the temperature, causing the DNA polymerase to replicate the DNA sequence between the primers. 
-In the course of such a cycle, the amount of DNA doubles.
+In the course of such a cycle, the amount of target DNA doubles.
 Therefore, the number of DNA copies increases geometrically with the number of cycles completed.
@@
 The quantitative result (the Ct value) has random variation even if the sample has the same initial amount of DNA or RNA present, and the qualitative result (positive or negative) can be wrong.
-Errors arise from a variety of sources, including the sampling/swabbing process and batch effects in the RT-PCR process~\autocite{hanRTPCR}.
+Random variation arises from a variety of sources, including the sampling/swabbing process and batch effects in the RT-PCR process~\autocite{hanRTPCR}.
 
@@
 False positives are very rare for SARS-CoV-2 RT-PCR tests.
-In summer 2020, the CIS performed\numprint{208730} PCR tests and found only 159 positives~\autocite[section 5]{cisMethodsONS}.
+In summer 2020, the CIS performed \numprint{208730} PCR tests and found only 159 positives~\autocite[section 5]{cisMethodsONS}.
 If these were all false positives, then the test specificity would be 99.9\%.
@@
 It was a longitudinal study with a representative sample of private households, \ie excluding settings such as care homes and student halls.
-The dataset is globally unique in providing a representative, longitudinal, and large-scale study across the pandemic.
 Initially, it was limited to England, but expanded to cover the whole of the UK in September 2020.
 Enrolment was continuous until 31 Jan 2022, with data collected until 13 Mar 2023~\autocite{weiRisk}. 
+The dataset is globally unique in providing a representative, longitudinal, and large-scale study across the pandemic.
 
@@
 Up until July 2020, households were selected from those that had previously participated in an ONS survey~\autocite{CIStechData}.
-Around 50\% of households agreed to participate, with around 90\% of eligible individuals (\numprint{96113} in total) in these households contributing multiple swabs to the survey.
+In this initial phase, around 50\% of households agreed to participate, with around 90\% of eligible individuals (\numprint{96113} in total) in these households contributing multiple swabs to the survey.
 After July 2020, addresses were also randomly sampled from a database of all UK addresses held by the ONS.
-The response rate was lower, with only 12\% of households agreeing to participate, from which 86\% of eligible individuals (\numprint{383732} in total) contributed multiple swabs to the survey.
+In this phase, response rate was lower, with only 12\% of households agreeing to participate, from which 86\% of eligible individuals (\numprint{383732} in total) contributed multiple swabs to the survey.
 
@@
 
-If the CIS was truly a random sample, an unbiased estimator of the positivity rate would be the proportion of positive tests in the sample; I refer to this estimator as the unadjusted positivity rate.
+If the CIS was truly a random sample, an unbiased estimator of the population positivity rate would be the proportion of positive tests in the sample; I refer to this estimator as the unadjusted positivity rate.
 However, in practice, two issues arise with this.
@@
 I modified the regression component of a pre-existing version of a MRP model for the CIS~\autocite{pouwelsMRPvaccination}, as recommended by that model's author for the specific context of this thesis~\citePersonalComms{Koen Pouwels}.
-As in prior work~\autocite{pouwelsMRPvaccination}, I used INLA for inference, implemented in R-INLA (see \cref{E-transmission:sec:INLA}).
 For a day $t$, age group $a$, ethnicity $e$, and sex $s$ the regression model is:
@@
 To start, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution.
-Denote these by $p^(j)_{t,i}$ for the $j$th sample where $i$ is a stratum, \ie some value of $a$, $e$, and $s$.
+Denote these by $p^{(j)}_{t,i}$ for the $j$th sample where $i$ is a stratum, \ie some value of $a$, $e$, and $s$.
 Define a (sub) population of interest as a set of strata $\set{S}$.
@@
 p^{(j)}_t(\set{S})
-&= \frac{\sum_{i \in \set{S}} p_{t,i} N_j}{\sum_{i \in \set{S}} N_i}
+&= \frac{\sum_{i \in \set{S}} p^{(j)}_{t,i} N_i}{\sum_{i \in \set{S}} N_i}
 \end{align}
-where $N_j$ is the population size of stratum $j$.
-The population numbers for each stratum are provided by the Office for National Statistics.
+where $N_i$ is the population size of stratum $i$.
+The population numbers for each stratum are provided by the ONS.
 
@@
 Intermittent negatives are the clearest example of a false negative.
-I will generally consider that intermittent negatives are uninformative on the duration of detectability, and hence exclude them from several analyses.
 
 The crucial features of an infection episode are the start and end times of the infection.
-Ignoring misclassified test results, latent start time of infection episode $j$, $B_j$, can be bounded as between the day after the last negative test, $l_j^{(b)}$, and the day of the first positive test, $r_j^{(b)}$.
-Similarly, the latent end time of infection episode $i$, $E_j$, can be bounded as between the day of the last positive test, $l_j^{(e)}$, and the day before the following negative test, $r_j^{(e)}$.
+Ignoring misclassified test results, the latent start time of infection episode $j$, $B_j$, can be bounded as between the day after the last negative test, $l_j^{(b)}$, and the day of the first positive test, $r_j^{(b)}$.
+Similarly, the latent end time of infection episode $j$, $E_j$, can be bounded as between the day of the last positive test, $l_j^{(e)}$, and the day before the following negative test, $r_j^{(e)}$.
 
@@
 For privacy reasons, the CIS data are stored in the ONS's SRS (Secure Research Service)~\autocite{onsSRS}.
-The SRS is a Trusted Research Environment (TRE) that allows researchers to access sensitive data in a secure environment.
+The SRS is a TRE (Trusted Research Environment) that allows researchers to access sensitive data in a secure environment.
 The SRS gives accredited researchers working on projects in the public interest access to granular data which cannot be legally shared publicly.

### 2024-03-29 cis-perfect-testing.tex
add figure for new likelihood

@@
 \begin{figure}
-\includegraphics[width=\textwidth]{cis-perfect-testing/regions_diag}
-\caption[Admissible, inadmissible, and undetected regions]{%
-  The regions for individual $i$.
-  $i$'s first test was at time 0 and negative, with subsequent negative tests at times 7, 14, and 56, and positive tests at times 21 and 28.
-  Inadmissible regions are unshaded.
-  The grey (impossible) region covers the cases in which recovery occurs prior to infection.
+\makebox[\textwidth][c]{\includegraphics[width=0.9\paperwidth]{cis-perfect-testing/regions_diag}}
+\thisfloatpagestyle{empty}
+\caption[Episode regions]{%
+  Each dot is a combination of $b_j$ and $e_j$ for an arbitrary individual $i$.
+  Each box, bounded by dashed lines, are combinations giving $O(j) = \nu_k$, $k$ corresponding to the numeric label; labels are arbitrary but unique across all individuals.
+
+  $i$ had negative tests at times 0, 7, 14, 56, and 84 (not shown) and positive tests at times 21 and 28.
+
+  The purple region corresponds to the doubly interval censored episode in this individual.
+  That is, $n_8 = 1$ and $n_k = 0$ for $k = 1, \dots, 7, 9, \dots, 15$.
+
+  The red region corresponds to combinations giving $O(j) = \emptyset$.
+
+  Impossible region violates $b_j \leq e_j$.
+}
 \label{perf-test:fig:partitionSpace}
-}
 \end{figure}

### 2024-03-29 cis-perfect-testing/regions_diag.R

@@ -1,3 +1,4 @@
+library(dplyr)
 library(ggplot2)
 
 neg_test_dates = c(0, 7, 14, 56)
@@ -13,21 +14,49 @@ test_lines = function(values, result) {
         colour = "Green"
     }
     list(
-        geom_vline(aes(colour = label, xintercept = !!values), linetype = "dotted"),
-        geom_hline(aes(colour = label, yintercept = !!values), linetype = "dotted")
+        geom_vline(aes(colour = label, xintercept = !!values)),
+        geom_hline(aes(colour = label, yintercept = !!values))
     )
 }
+
+tbl_labels = inner_join(
+    tibble::tibble(x1 = all_test_dates, x2 = lead(all_test_dates)),
+    tibble::tibble(y1 = all_test_dates, y2 = lead(all_test_dates)),
+    by = join_by(x1 < y1)
+) |>
+    tidyr::replace_na(list(y2 = 60)) |>
+    arrange(desc(y1), x1) |>
+    tibble::rowid_to_column("label") |>
+    mutate(
+        x = (x1 + x2) / 2,
+        y = (y1 + y2) / 2,
+    )
 
 shade_alpha = 0.3
 plot = ggplot() +
     test_lines(pos_test_dates, TRUE) +
     test_lines(neg_test_dates, FALSE) +
+    geom_hline(aes(yintercept = all_test_dates - 0.5), linetype = "dashed") +
+    geom_vline(aes(xintercept = all_test_dates + 0.5), linetype = "dashed") +
+    # add labels
+    geom_text(
+        data = tbl_labels,
+        aes(x, y, label = label),
+        size = 10
+    ) +
     theme_minimal() +
     scale_x_continuous(breaks = all_test_dates, minor_breaks = -100:100, limits = c(-10, 60), expand = c(0, 0)) +
-    scale_y_continuous(breaks = all_test_dates, minor_breaks = -100:100, limits = c(0, 60), expand = c(0, 0)) +
+    scale_y_continuous(breaks = all_test_dates, minor_breaks = -100:100, limits = c(0, 59.5), expand = c(0, 0)) +
+    geom_point(
+        aes(x, y),
+        tidyr::expand_grid(x = -100:100, y = -100:100),
+        size = 0.1,
+        alpha = 0.5,
+        colour = "grey"
+    ) +
     labs(
-        x = "Infection time",
-        y = "Recovery time",
+        x = expression(b[j]),
+        y = expression(e[j]),
         colour = "Test results",
[…]

### 2024-03-29 cis-perfect-testing.tex
Updates from Overleaf

--- 2024-03-29 cis-perfect-testing.tex
@@
 In \cref{perf-test:sec:data}, I introduce the data, defining which quantities are observed and which are latent.
-Then, in \cref{perf-test:sec:episode-types}, I build a framework, using the concept of \textcite{heiseyModelling}, which classifies possible episode beginning and end time into \emph{admissible} regions, each of which match a possible pattern of the doubly interval censored pattern of the observations; and an \emph{undetected} region, which lead to the episode being undetected.
+Then, in \cref{perf-test:sec:posterior}, I build a framework, using the concept of \textcite{heiseyModelling}.
+I classify episode based on the possible patterns of the doubly interval censored pattern of the observations; and an \emph{undetected} region, which lead to the episode being undetected.
 The number of such episodes in each region has a multinomial distribution conditional on the latent total number of infections (including undetected infections).
@@
 In \cref{perf-test:sec:prob-undetected} I formalise the conditions for an episode to be detected.
-This period is denoted as $t = 1, \dots, T$.
+Each day in this period is denoted as $t = 1, \dots, T$.
 A negative test before the start of the episode provides a lower bound on the episode's start time; otherwise, there is little information on its length.
@@
 
-\subsection{Posterior density} \label{perf-test:sec:posterior}
-
-Any infection episode, whether detected or not, is fully characterised by the triplet $(b_j, e_j, i(j))$ where $b_j$ and $e_j$ are the times the episode begin and end respectively; as $i(j)$ is the individual the episode occurred in it also identifies the test schedule.
+Any infection episode, whether detected or not, is fully characterised by the triplet $(b_j, e_j, i(j))$ where $b_j$ and $e_j$ are the times the episode begin and end respectively; and $i(j)$ is the individual the episode occurred in.
+$i(j)$ identifies the test schedule.
 The episode's duration is $d_j = e_j - b_j + 1$ (see \cref{E-inc-prev:sec:definitions}).
@@
 These latter two properties are what make the data doubly interval censored.
+
+\subsection{Posterior density} \label{perf-test:sec:posterior}
 
@@
 In the CIS data, each $n_k$ ($k \neq u$) is observed as either 0 or 1.
-Define $D = \{ k \ssep n_k = 1 \}$.
+Define $\set{D} = \{ k \ssep n_k = 1 \}$, the detected episodes.
 Furthermore, note that the support of the multinomial distribution requires that $\nnodet = \ntot - \ndet$.

### 2024-03-29 cis-imperfect-testing.tex

@@
 In this section, I introduce a simple model of false negatives (\ie allowing $\psens < 1$) into the likelihood derived in \cref{E-perf-test:sec:model}.
-The simple model, including a constant test sensitivity, means the likelihood remains tractable.
+Using a simple model, in particular a constant test sensitivity, means the likelihood remains tractable.
 
@@
 Similarly, let $p_k'$, $p_{ik}'$, $p_u$, and $p_{iu}$ have the same definition as $p_k$, $p_{ik}$, $p_u$, and $p_{iu}$ respectively except replacing $O$ with $O'$ and $\vec{\nu}_k$ with $\vec{\nu}_k'$, and conditioning on $\psens$.
-As before, $p'_k = 1/\Ncis p'_{i_kk}$ and $p'_u = 1/\Ncis \sum_{i=1}^{\Ncis} p'{iu}$.
+As before, $p'_k = 1/\Ncis p'_{ik}$ and $p'_u = 1/\Ncis \sum_{i=1}^{\Ncis} p'_{iu}$.
 
@@
 Therefore, the test results $\vec{y}_k$, except the test at $r_k^{(e)}+1$, are either true positives or false negatives.
-This gives $t_+ = \sum_{l=1}^{m_k} y_{k,l}(t)$ true positives and $f_- = m_k - t_{k,+}$ false negatives.
+This gives $t_+ = \sum_{l=1}^{m_k-1} y_{k,l}(t)$ true positives and $f_- = m_k - 1 - t_{k,+}$ false negatives.
 

### 2024-03-29 cis-imperfect-testing.tex
overleaf changes

@@
 Denote the elements of $\sched'_j$ by $t_{j,1} < \dots < t_{j,m_j}$.
-The new observation vector for $j$ is $O'(j) = [O(j), \vec{y}_j]^T$ where $\vec{y}_j$ is a binary vector of length $m_j$, with $y_{j,l}$ being the test result at time $t_{j,l}$.
+The new observation vector for $j$ is $O'(j) = [O(j), \vec{y}_j]^T$ where $\vec{y}_j \in \{ 0, 1 \}^{m_j}$, with $y_{j,l}$ being the test result at time $t_{j,l}$.
 
@@
   \item $\vec{\nu}_k \in \set{D}$.
-  \item $\vec{y}_k \in \{0, 1\}^{m_k}$.
   \item The elements of $\vec{y}_k$ corresponding to the tests at times $r_k^{(b)}$ (its first element) and $l_k^{(e)}$ (its penultimate element) are 1.
@@
 
-I will modify $p_{ik}$ to form $p_{ik}' = \prob(O'(j) = \nu'_k \mid i(j) = i_k, \vec{\theta})$, which allows the negative test following the last positive to be a false negative.
-If it is a false negative, then I will consider the episode's end right censored.
-However, if it is a true negative, then the episode's end is interval censored, as previously.
-A mixture of these scenarios forms the episode's likelihood contribution, with the mixture probability determined by the test sensitivity.
+I will modify $p_{ik}$ to form $p_{ik}' = \prob(O'(j) = \nu'_k \mid i(j) = i_k, \vec{\theta})$, taking into account false negatives.
+I consider a mixture of two scenarios, whether the final test in $\sched'_{i(j)}$ is a false positive, with the mixture probability determined by the test sensitivity.
 
@@
 Therefore, the test results $\vec{y}_k$, except the test at $r_k^{(e)}+1$, are either true positives or false negatives.
-This gives $t_+ = \sum_{l=1}^{m_k-1} y_{k,l}(t)$ true positives and $f_- = m_k - 1 - t_{k,+}$ false negatives.
+This gives $t_+ = \sum_{l=1}^{m_k-1} y_{k,l}(t)$ true positives and $f_- = \sum_{l=1}^{m_k-1} (1 - y_{k,l}(t))$ false negatives.
 
@@
 
-I now modify $p_{iu}$ to form $p_{iu}' = \prob(O'(j) = \emptyset \mid i(j) = i_k, \vec{\theta})$, which allows the negative test following the last positive to be a false negative.
+I now modify $p_{iu}$ to form $p_{iu}' = \prob(O'(j) = \emptyset \mid i(j) = i_k, \vec{\theta})$ to take into account false negatives.
 Several mechanisms for episodes being undetected were considered when deriving $p_{iu}$ in \cref{E-perf-test:sec:prob-undetected}.

### 2024-03-29 cis-perfect-testing.tex
overleaf changes

@@
 In \cref{perf-test:sec:data}, I introduce the data, defining which quantities are observed and which are latent.
-Then, in \cref{perf-test:sec:posterior}, I build a framework, using the concept of \textcite{heiseyModelling}.
-I classify episode based on the possible patterns of the doubly interval censored pattern of the observations; and an \emph{undetected} region, which lead to the episode being undetected.
-The number of such episodes in each region has a multinomial distribution conditional on the latent total number of infections (including undetected infections).
+In the remainder of the section, I derive the posterior density, adapting the general approach of \textcite{heiseyModelling}.
+I classify episodes based on how they would be observed (\ie the intervals that their beginning and end times would be censored to).
+These potential classification can then be compared to the actual observations.
 
-In \cref{perf-test:sec:posterior}, I outline the form of the posterior density.
-The following subsections then derive its various components.
-Unlike in \textcite{heiseyModelling}, I must consider the individual's probability of undetected episodes separately because they have differing visit times.
+Unlike in \textcite{heiseyModelling}, I must consider each individual's probability of undetected episodes separately because they are tested at different times.
 In addition, I include the information from the individuals without a detected episode.
-
-Unlike prior work, I marginalise the posterior density over the number of undetected episodes.
+Furthermore, I marginalise the posterior density over the number of undetected episodes.
 This makes inference tractable using the large sample size and the computing resources available within the SRS (see \cref{E-biology-data:sec:SRS}).
-In particular, the posterior can be expressed within modern, high-performance Bayesian software such as Stan.
+% In particular, the posterior can be expressed within modern, high-performance Bayesian software such as Stan.
 
@@
 
-I analyse the $\ndet = \numprint{4800}$ episodes with their first positive test between 10 Oct 2020 and 6 Dec 2020 inclusive, with negatives bounding the start and end time of the episode; I refer to these episodes as \emph{detected}.
-In \cref{perf-test:sec:prob-undetected} I formalise the conditions for an episode to be detected.
+I analyse the $\ndet = \numprint{4800}$ episodes with their first positive test between 10 Oct 2020 and 6 Dec 2020 inclusive, with negatives bounding the start and end time of the episode; I refer to these episodes as \emph{detected}; in \cref{perf-test:sec:prob-undetected} I formalise the conditions for an episode to be detected.
 Each day in this period is denoted as $t = 1, \dots, T$.
-A negative test before the start of the episode provides a lower bound on the episode's start time; otherwise, there is little information on its length.
-Similar information comes from the negative test after the end of the episode.
-A negligible number of episodes are missing a negative test after the end of the episode.
+I use these episodes because a negative test before the start of the episode provides a lower bound on the episode's start time; otherwise, there is little information on its length.
+Similar information comes from the negative test after the end of the episode, although a negligible number of episodes are missing a negative test after the end of the episode.
+How the episode data are constructed is detailed in \cref{E-biology-data:sec:cis-episodes}.
 
-I consider the cohort of $\Ncis = \numprint{437590}$ individuals who could have had a detected episode; this is a subset of the whole CIS cohort.
-To be considered, an individual must have at least one test between time 1 and $T$.
+I consider the cohort of $\Ncis = \numprint{437590}$ individuals who could have had a detected episode.
+This is the subset of the whole CIS cohort with at least one test between time 1 and $T$.
 I index these individuals by $i = 1, \dots, \Ncis$.
@@
 Test results are deterministic based on the time of the test and the individual's infection status.
-Therefore, $O(j)$ is fully determined by the triplet $(B_j, E_j, i(j))$ consisting of $j$'s beginning time, end time, and the individual in which the episode occurs respectively.
+Therefore, $O(j)$ is fully determined by the triplet $(b_j, e_j, i(j))$ consisting of $j$'s beginning time, end time, and the individual in which the episode occurs respectively.
+The episode's duration is $d_j = e_j - b_j + 1$ (see \cref{E-inc-prev:sec:definitions}).
 
-For a detected episode, we have $O(j) = [l_j^{(b)}, r_j^{(b)}, l_j^{(e)}, r_j^{e}, i(j)]^T$.
-These are: the individual in which the episode occurs, denoted $i(j)$; the earliest and latest time the episode could have begun, $l_j^{(b)}$ and $r_j^{(b)}$ respectively; and the earliest and latest time the episode could have ended, $l_j^{(e)}$ and $r_j^{(e)}$ respectively.
-How the episode data are constructed is detailed in \cref{E-biology-data:sec:cis-episodes}.
+If $j$ is detected, then $O(j) = [l_j^{(b)}, r_j^{(b)}, l_j^{(e)}, r_j^{e}, i(j)]^T$, where $i(j)$ is the individual in which the episode occurs; $l_j^{(b)}$ and $r_j^{(b)}$ are the earliest and latest time the episode could have begun respectively; and $l_j^{(e)}$ and $r_j^{(e)}$ are the earliest and latest time the episode could have ended respectively.
 
-For an undetected episode $j$, there are no observations; in which case, define $O(j) = \emptyset$.
+If $j$ is undetected, there are no observations; in which case, define $O(j) = \emptyset$.
 Let $\nnodet$ denote the (latent) number of undetected episodes experienced across all $\Ncis$ individuals.
@@
 
-Any infection episode, whether detected or not, is fully characterised by the triplet $(b_j, e_j, i(j))$ where $b_j$ and $e_j$ are the times the episode begin and end respectively; and $i(j)$ is the individual the episode occurred in.
-$i(j)$ identifies the test schedule.
-The episode's duration is $d_j = e_j - b_j + 1$ (see \cref{E-inc-prev:sec:definitions}).
-For all $j$, the exact values $b_j$ and $e_j$ are latent.
-If $j$ is undetected then $i(j)$ is also latent and little is known about $b_j$ or $e_j$.
-If $j$ is detected, then $i(j)$ is known; $b_j \in [l_j^{(b)}, r_j^{(b)}]$; and $e_j \in [l_j^{(e)}, r_j^{(e)}]$.
-These latter two properties are what make the data doubly interval censored.
+% Any infection episode, whether detected or not, is fully characterised by the triplet $(b_j, e_j, i(j))$ where $b_j$ and $e_j$ are the times the episode begin and end respectively; and $i(j)$ is the individual the episode occurred in.
+% $i(j)$ identifies the test schedule.
+% For all $j$, the exact values $b_j$ and $e_j$ are latent.
+% If $j$ is undetected then $i(j)$ is also latent and little is known about $b_j$ or $e_j$.
+% If $j$ is detected, then $i(j)$ is known; $b_j \in [l_j^{(b)}, r_j^{(b)}]$; and $e_j \in [l_j^{(e)}, r_j^{(e)}]$.
+% These latter two properties are what make the data doubly interval censored.
 
@@
 Hence, $\ntot = \sum_{i=1}^{N_D} n_i + \nnodet$.
-For $\vec{\nu}_k \in \set{E}$, let $p_k = \prob(O(j) = \vec{\nu}_k \mid \vec{\theta})$; $p_u = \prob(O(j) = \emptyset \mid \vec{\theta})$; and $\vec{p} = [p_1, \dots, p_D, p_u]^T$.
+
+Now define the probabilities of each possible outcome of $O(j)$.
+For detected outcomes, $\vec{\nu}_k \in \set{E}$, let $p_k = \prob(O(j) = \vec{\nu}_k \mid \vec{\theta})$.
+For undetected outcomes, let $p_u = \prob(O(j) = \emptyset \mid \vec{\theta})$.
+Together, $\vec{p} = [p_1, \dots, p_D, p_u]^T$.
 Assume, for tractability, that the events $O(j) = \vec{\nu}_k$ and $O(j') = \vec{\nu}_{k'}$ are independent for $j \neq j'$; this assumption is discussed in \cref{perf-test:sec:discussion}.
@@
   Each box, bounded by dashed lines, are combinations giving $O(j) = \nu_k$, $k$ corresponding to the numeric label; labels are arbitrary but unique across all individuals.
-
   $i$ had negative tests at times 0, 7, 14, 56, and 84 (not shown) and positive tests at times 21 and 28.
-
   The purple region corresponds to the doubly interval censored episode in this individual.
   That is, $n_8 = 1$ and $n_k = 0$ for $k = 1, \dots, 7, 9, \dots, 15$.
-
   The red region corresponds to combinations giving $O(j) = \emptyset$.
-
   Impossible region violates $b_j \leq e_j$.
@@
 \end{align}
-assuming $P(i(j) = i \mid \vec{\theta})$ as before.
+assuming $P(i(j) = i \mid \vec{\theta}) = 1/\Ncis$ as before.
 Let $p_{iu} = \prob(O(j) = \emptyset \mid i(j) = i \mid \vec{\theta})$.
@@
     \item $b_j \leq T_{i(j)}$ where $T_{i(j)} = \max \{ t \in \sched_{i(j)} \ssep t \leq T \}$ is the last time that $i(j)$ is tested in the period, meaning that the test is detected within the period.
-    \item Individual $i(j)$ is tested at a time $t > e_j$, upper bounding the end of the episode.
-      Note that this could be after $T$, and hence it is not necessarily the case that $t \in \sched_{i(j)}$.
-      For episodes detected in the period I consider, this happens a negligible number of times, only in individuals lost to follow-up; therefore, I neglect this possibility.
+    \item $\exists t \in \sched_{i(j)}$ such that $t > e_j$, upper bounding the end of the episode.
+      For episodes detected in the period I consider, a negligible number of episodes are excluded due to this critera.
+      It only occurs in individuals lost to follow-up; therefore, I neglect this possibility.
       % For a new context, including recent infections, this condition could be relaxed by considering episodes that do not meet this criterion as right censored.
@@
 &= \prob(D_j \geq \tau_{\sched_{i}}(B_j)+ 1, \min \sched_{i(j)} < B_j \leq T_{i(j)}) \\
-&= \sum_{b = \min \sched_{i} + 1}^{T_{i(j)}} \prob(D_j \geq \tau_{\sched_{i}}(b) + 1, \tau_{\sched_{i(j)}}(b_j) < \infty \mid B_j = b) \prob(B_j = b)\\
+&= \sum_{b = \min \sched_{i} + 1}^{T_{i(j)}} \prob(D_j \geq \tau_{\sched_{i}}(b) + 1 \mid B_j = b) \prob(B_j = b)\\
 &= \frac{1}{T} \sum_{b = \min \sched_{i} + 1}^{T_{i(j)}} S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1).
@@
 
+The derivation in \cref{perf-test:sec:model} relied on an assumption that $O(j)$ and $O(j')$ occur independently for $j \neq j'$.
+This is not the case, as immunity means that an individual previously infected is less likely to be infected in the future.
+Additionally, the same individual cannot have multiple concurrent infections, due to the way that the episodes dataset is constructed (see \cref{E-biology-data:sec:cis-episodes}) limiting certain combinations from occurring.
+In practice, this means that $n_k > 1$ cannot occur.
+The simulation study violated this assumption to a large extent: each individual could only have one infection.
+Despite this, the method performed well, recovering the true survival function.
+Therefore, it is unlikely that this assumption matters, possibly due to the large number of individuals without detected episodes.
+Further work could explore different assumptions, \eg a ``full immunity'' assumption meaning each individual has zero or one episode in the period.
+
 Various extensions could be worth exploring.

### 2024-03-30 incidence-prevalence.tex
Dani comments

@@
 
-This chapter formally defines incidence and prevalence, and the relationship between these quantities.
+This chapter formally defines incidence and prevalence, and the relationship between them and prevalence survey data.
 In the course of explaining this relationship, it will become clear that the duration of RT-PCR positivity, and its distribution in the population, is the vital parameter which will allow estimation of incidence from prevalence, a major aim of this thesis.
@@
 I start by defining these terms in \cref{inc-prev:sec:definitions}.
-I then move on to explaining the three processes that are involved: the \emph{infection process}, the \emph{prevalence process}, and the \emph{observation process}.
-These are the mathematical description of an infection model, disease model, and observation model respectively.
+I then move on to explaining the three processes that link the incidence, prevalence, and the data: the \emph{infection process}, the \emph{prevalence process}, and the \emph{observation process}.
 
-The infection process (\cref{inc-prev:sec:infection-process}) models how infections arise.
-The concepts here are well-established within infectious disease epidemiology.
-
-The prevalence process (\cref{inc-prev:sec:prevalence-process}) describes the relationship between incidence and prevalence.
-Here, I adapt previous work to the SARS-CoV-2 context.
-
-The observation process (\cref{inc-prev:sec:observation-process}) describes the relationship between prevalence and the observed data, ultimately allowing inference of incidence based on this data.
-In particular, I include the sampling mechanism of a population prevalence survey, guided by the design of the CIS.
+In (\cref{inc-prev:sec:infection-process}) I introduce the infection process, describing the infections' generating mechanism; in (\cref{inc-prev:sec:prevalence-process}) the prevalence process, which generates prevalence from the incidence; and in (\cref{inc-prev:sec:observation-process}), the observation process, which generates the data from the prevalence through the sampling mechanism of a population prevalence survey, guided by the design of the CIS.
 
@@
 
-I use discrete time in this chapter, except in \cref{SEIR:sec:mechanistic-models}.
-I use days; however, the results generalise in the obvious way to other time units.
+I use discrete time in this chapter, except in \cref{SEIR:sec:mechanistic-models}, with the time unit of a day. However, results generalise in the obvious way to other time units.
 I choose discrete time for pragmatic reasons.
 Most epidemiological data are only available at daily granularity or, when more granular data exists, is often unreliable.
-Furthermore, within-day patterns produce variations in transmission (\eg due to sleep) and data (\eg due to logistical considerations around collection); these variations are not of interest in most contexts.
+Furthermore, within-day patterns produce variations in transmission (\eg due to sleep) and data (\eg due to logistical considerations around collection), but these variations are not of interest in most contexts.
 
-\emph{Incidence} on day $t$, $Z_t$, is the number of people in a given population who become infected on day $t$; when expressed as a proportion of the population it is the \emph{incidence proportion}~\autocite[89]{lashModern}, $Z_t/\Npop$, where $\Npop$ is the number of people in the population.
+\emph{Incidence} $Z_t$, is the number of people in a given population who become infected on day $t$.
+When expressed as a proportion of the population it is the \emph{incidence proportion}~\autocite[89]{lashModern}, $Z_t/\Npop$, where $\Npop$ is the size of the population.
 
@@
 I define this period as an \emph{infection episode}, beginning on day $B_i$ and ending on day $E_i$; therefore, with a duration $D_i = E_i - B_i + 1$.
-I model the $D_i$s as a discrete iid (independent and identically distributed) random variable; extending the methodology to account for covariates is left to further work.
+I model the $D_i$s as a discrete iid (independent and identically distributed) random variables; extending the methodology to account for covariates is left to further work.
 The duration is a positive integer, \ie $D_i \in \{1, 2, \dots, \dmax \}$, where $\dmax$ is the maximum duration.
-The distribution of $D_i$ can be expressed as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
+The distribution of $D_i$ can be expressed through the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
 Alternatively, the \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$.
@@
 Since $D_i$ is a discrete random variable, $0 \leq f_D(d) \leq 1$ for all $d$ and $\sum_{k=1}^{\dmax} f_D(k) = 1$.
-This implies that $F_D(d) \in [0, 1]$ and increases monotonically.
+By definition, $F_D(d) \in [0, 1]$ and increases monotonically.
 Therefore, $S(d) \in [0, 1]$ and decreases monotonically.
@@
 \section{Infection process} \label{inc-prev:sec:infection-process}
-The infection process is the mathematical description of how infections occur.
-It is a \emph{counting} process, where each infection is an event.
 
-A counting process is a random function of continuous time $K(t)$.
+The infection process describes how infections occur.
+It is typically represented as a \emph{counting} process, where each infection is an event.
+
+A counting process is a stochastic process represented by a random function of continuous time $K(t)$.
 $K(t)$ must take non-negative integer values and be monotonically increasing in $t$.
@@
 
-First, in \cref{inc-prev:sec:poisson-process}, I introduce a phenomenological model where the infection process is a \emph{Poisson process}.
+First, in \cref{inc-prev:sec:poisson-process}, I introduce phenomenological models, where the infection process is a \emph{Poisson process}.
 A Poisson process is a widely used type of counting process both within and outside of epidemiology.
-It makes no direct assumptions on the underlying biological or epidemiological process and is therefore a flexible model.
+It makes no explicit assumptions on the underlying biological or epidemiological mechanism and can be a flexible model.
 
@@
 Extensions to this model, which I do not consider here, have incorporated covariates, such as geography or age~\autocite[e.g.][]{diggleModeling}.
+The Poisson process has two key properties.
 
-The first defining property of a Poisson process is that the counts are Poisson distributed.
+First, the Poisson process has counts that are Poisson distributed.
 Formally, $K(s) - K(t) \dist \Poi\left(\int^s_t \nu(r) dr \right)$ where $\nu(r) \geq 0$ is the \emph{intensity} of the infection process at time $r$~\autocite[244]{yanDistribution}.
 
-The other defining property of a Poisson process is that it has \emph{independent increments}.
+Second, the Poisson process has \emph{independent increments}.
 A counting process has independent increments if the number of events in disjoint intervals are independent conditional on the intensity~\autocite[244]{yanDistribution}.
@@
 
-The target of inference can be either the process's intensity or the counts of the specific realisation that occurred.
+In an epidemic, there are two features of a Poisson process: the underlying process intensity, $\nu(r)$, and the single realisation that occurred, $K(t)$; either of these can be the target of inference.
 The realised counts are normally of greater interest because we care about the realised epidemic, not what may happen if there were future realisations of the same epidemic~\autocite{beckerDependent,brookmeyerMethod}.
@@
 Therefore, an overdispersed epidemic modelled as a Poisson process will be less variable than the realised epidemic.
-Overdispersion is common in infectious diseases, especially for coronaviruses such as SARS-CoV-2~\autocite{endoEstimating,adamClustering,mccloskeySARS}.
+Overdispersion is common in infectious diseases, especially for coronaviruses such as SARS-CoV-2, where some individuals infect many more than others~\autocite{endoEstimating,adamClustering,mccloskeySARS}.
 
@@
 
-Now I will introduce mechanistic models, in particular the class of mechanistic transmission models known as \emph{compartmental} models.
+Now I will introduce mechanistic models, in particular the class of mechanistic transmission models known as \emph{compartmental} models~\autocite{andersonInfectious,keelingModeling}.
 Compartmental models are the most widely used class of mechanistic infectious disease models.
-A compartmental model is named because they place every member of the population under study into a compartment; each compartment representing a disease state (\eg susceptible).
+A compartmental model categorises every member of the population under study into a compartment; each compartment representing a disease state (\eg susceptible).
 I will describe these models using the notation of the field, which differs in conventions from the rest of this thesis.
@@
 The parameters $\vec{\theta}$ determine, along with the model structure, how the state vector changes over time.
-% Inference about $\vec{x}(t)$ and $\vec{\theta}$ is requires linking to data, $\vec{y}$, via an observation model, $p(\vec{y}(t) \mid \vec{x}, \vec{\theta})$ where $\vec{x}$ denotes conditioning on $\vec{x}(t)$ for all $t$~\autocite{birrellEvidence} (see \cref{SEIR:sec:observation}).
-% That is, the observation model specifies the likelihood of the data given the model state.
+Compartmental models can be \emph{deterministic} or \emph{stochastic}.
 
-Compartmental models can be \emph{deterministic} or \emph{stochastic}.
-Deterministic models are defined by the state vector $\vec{x}(t)$ being a deterministic function of the parameters $\vec{\theta}$ and the initial state $\vec{x}(0)$; equivalently, $\var(\vec{x}(t) \mid \vec{x}(0), \theta) = 0$~\autocite{birrellEvidence}.
-% In most situations, the function is not analytically-tractable and requires numeric methods to solve.
-%The transmission model corresponds to the infection process (see \cref{E-inc-prev:sec:infection-process}).
-%There remains randomness in the observation model (described in \cref{SEIR:sec:observation}).
+Stochastic models are specified by a probability distribution for the current state $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta})$, where $\set{H}_t = \{ \vec{x}(t') \ssep t' < t \}$ is the epidemic's history and $\vec{\theta}$ are model parameters~\autocite{birrellEvidence,dukicTracking,corbellaThesis,keelingModeling}.
+A common simplification is to use discrete time and a Markov assumption, \ie $\vec{x}(t)$ is only defined for integer $t$ and $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta}) = p(\vec{x}(t) \mid \vec{x}(t-1), \vec{\theta})$ respectively; the Markov assumption means that the state at time $t$ only depends on the state at time $t-1$~\autocite{birrellEvidence}.
+$x(t)$ and $x(t-1$ are now dependent, relaxing the assumption of independent increments in the Poisson process formulation of the infection process (see \cref{E-inc-prev:sec:infection-process}).
 
-Stochastic models specify a probability distribution for the current state $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta})$ where $\set{H}_t = \{ \vec{x}(t') \ssep t' < t \}$ is the epidemic's history and $\vec{\theta}$ are model parameters~\autocite{birrellEvidence,dukicTracking,corbellaThesis,keelingModeling}.
-Stochastic models are most commonly specified in discrete time, using a Markov assumption.
-A common simplification is to use discrete time and a Markov assumption, \ie $\vec{x}(t)$ is only defined for integer $t$ and $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta}) = p(\vec{x}(t) \mid \vec{x}(t-1), \vec{\theta})$~\autocite{birrellEvidence}.
-This relaxes the assumption of independent increments in the counting process formulation of the infection process (see \cref{E-inc-prev:sec:infection-process}).
-
+Deterministic models are defined by the state vector $\vec{x}(t)$ being a deterministic function of the parameters $\vec{\theta}$ and the initial state $\vec{x}(0)$; equivalently, $\var(\vec{x}(t) \mid \vec{x}(0), \theta) = \matr{0}$~\autocite{birrellEvidence}.
 Deterministic models are also known as \emph{mean-field} models because they represent the mean behaviour of a corresponding stochastic model.
@@
 The remainder of this section uses only deterministic models.
-This choice is appropriate for the context of this thesis because there were many individuals infectious with SARS-CoV-2 in this time period.
+This choice is appropriate for the context of this thesis because there were many individuals infectious with SARS-CoV-2 in the time period considered in this thesis.
 Hence, further details on stochastic models are outside the scope of this thesis; a wealth of introductory texts with further details on stochastic models are available, I recommend \textcite[chapter 6]{keelingModeling}.
@@
 The name SIR derives from the structure of the model.
-In particular, the model divides the population (assumed closed, an assumption discussed later) into one of three compartments: susceptible, infectious, or recovered.
+In particular, the model divides the population (assumed closed, an assumption discussed later) into one of three compartments: Susceptible, Infectious, or Recovered.
 % The state of the system is the proportion of the population in each compartment.
@@
 where $\lambda(t) = \beta i(t)$, is the \emph{force of infection}: the risk of infection for a susceptible individual at time $t$~\autocite[17]{keelingModeling}.
-The dynamics of the model are described by the parameters $\vec{\theta} = (\beta, \gamma)^T$.
-$\beta$ is the transmission rate, absorbing several terms as described below.
-$\gamma$ is the recovery rate, the rate that infectious individuals that recover; hence, $1/\gamma$ is the mean infectious period~\autocite[367]{keelingModeling}.
-These parameters are assumed to be constant.
+The dynamics of the model are described by the parameters $\vec{\theta} = (\beta, \gamma)^T$, where
+$\beta$ is the transmission rate, absorbing several terms as described below;
+$\gamma$ is the recovery rate, the rate that infectious individuals that recover.
+Hence, $1/\gamma$ is the mean infectious period~\autocite[367]{keelingModeling}.
+These parameters are assumed to be constant over time.
 
@@
 The expression for the force of infection, $\lambda(t)$ can be arrived at by considering a single infectious individual~\autocite[214]{kretzschmarMathematical}.
-Assume this individual comes into contact with other individuals in the population at a mean rate of $\kappa$.
+Assume this individual comes into contact with other individuals in the population at a mean rate of $\kappa$ per day.
 Further, assume the population is \emph{well-mixed}, meaning that each contact is with an individual chosen uniformly at random from the population, implied by assuming a homogeneous population.
@@
     \item There is no analytical solution to the system of ODEs.
-      Solving the system requires numerical methods~\autocite[25]{keelingModeling}.
-    \item As noted previously, solving the system of ODEs requires the parameters $\vec{\theta}$ and the initial state of the system, $\vec{x}(0)$.
+      Solving the system requires numerical methods~\autocite[25]{keelingModeling}, and uses the parameters, $\vec{\theta}$, and initial state of the system, $\vec{x}(0)$.
     \item The model assumes immunity lasts until the end of the period considered~\autocite[61]{andersonInfectious}.
-    In many contexts, including here, this assumption is justified because the period is shorter than the length of immunity conferred~\autocite{milneImmunity}.
+    In many contexts, including this thesis, this assumption is justified because the period is shorter than the length of immunity conferred~\autocite{milneImmunity}.
     For longer time periods, the model can be expanded to include waning immunity~\autocite[40]{keelingModeling}.
-    \item The incidence proportion on day $t$ is defined as $Z_t / \Npop = s(t) - s(t+1)$.
+    \item A discretised incidence proportion for day $t$ can be defined as $Z_t / \Npop = s(t) - s(t+1)$.
         However, this will not in general give an integer value to $Z_t$.
@@
 }
-  \caption[The SIR model with non-exponential infectious period]{Schematic of the SIR model with non-exponential infectious period. The infectious period is modified by having multiple I states.}
+  \caption[The SIR model with non-exponential infectious period]{Schematic of the SIR model with non-exponential infectious period. The infectious period is modified by having $n$ I states.}
   \label{SEIR:fig:SIR-gamma}

### 2024-03-30 transmission.tex
Dani comments

@@
 
-In this chapter, I use the estimate of the distribution of the duration of RT-PCR positivity from the combination of information from the ATACCC and CIS studies (see \cref{E-imperf-test:sec:conclusion}) to investigate the feasibility of estimating incidence and transmission of SARS-CoV-2 from only the CIS data (described in \cref{E-intro:sec:cis}).
+In this chapter, I use the estimate of the distribution of the duration of RT-PCR positivity based on combining information from the ATACCC and CIS studies (see \cref{E-imperf-test:sec:conclusion}) to investigate the feasibility of estimating incidence and transmission of SARS-CoV-2 from only the CIS data (described in \cref{E-intro:sec:cis}).
 I will show that the CIS data are sufficient, demonstrating the value of surveys of this type for pandemic surveillance.
 
-I take two approaches, backcalculation and a mechanistic model, both previously discussed in \cref{E-inc-prev:sec:infection-process}.
+I take two approaches, one phenomenological and one mechanistic, both previously discussed in \cref{E-inc-prev:sec:infection-process}.
 
-I start with the backcalculation (\cref{backcalc}), based on \cref{E-inc-prev:eq:EPt}, which enables estimation of incidence within (sub)populations of England.
+I start with the phenomenological approach (\cref{backcalc}), based on \cref{E-inc-prev:eq:EPt}, which enables estimation of incidence within (sub)populations of England.
 The approach makes few assumptions about how transmission occurs, decreasing the potential to introduce bias and can also easily incorporate arbitrary covariates to adjust for non-response bias.
@@
 
-\section{Backcalculation} \label{backcalc}
+\section{Phenomenological approach} \label{backcalc}
 
@@
 Specifying $F_Z$ should consider the need to provide structure to stabilise the problem (see \cref{E-intro:sec:disease-model}).
-In the context of the CIS, it needs to allow borrowing information across time and/or strata (see \cref{E-biology-data:sec:MRP}).
+In the context of the CIS, it needs to allow borrowing information across time and/or strata (see \cref{E-biology-data:sec:MRP}), which is required because for some stratum/time combinations the data become very sparse.
 
 The second approach, which I call the \emph{prevalence-inferred} approach takes a two-step approach.
-First, it uses the statistical model from \cref{E-biology-data:sec:MRP} to estimate prevalence.
+First, it estimates prevalence using the statistical model from \cref{E-biology-data:sec:MRP}.
 Then, it transforms the posterior distribution of prevalence to a posterior distribution of incidence using the forward substitution as in \cref{inc-prev:eq:forward-substitute}.
@@
 The incidence-inferred approach is more parsimonious in being a single step procedure.
-However, computational challenges make makes mean I favour the prevalence-inferred approach.
+However, computational challenges make the prevalence-inferred approach more appealing.
 
-It will be convenient to define a variable equating to the quantity initially estimated.
-Let $\pi_{i,t} = Z_{i,t}$ under the inference-inferred approach and $\pi_{i,t} = P_{i,t}$ under the prevalence-inferred approach.
+It will be convenient to define a variable equating to the quantity initially estimated: let $\pi_{i,t} = Z_{i,t}$ under the inference-inferred approach and $\pi_{i,t} = P_{i,t}$ under the prevalence-inferred approach.
 
@@
 The logit transformation also allows property 4 to be conveniently satisfied using an additive model for $g_{\vec{\varphi}}$.
-An additive model on the logit scale is approximately multiplicative on the natural scale for small values of $\pi_{i,t}$ (which will be common).
+An additive model on the logit scale is approximately multiplicative on the natural scale for small values of $\pi_{i,t}$, which will be common.
 
-This form of model includes the MRP prevalence model in \cref{E-biology-data:sec:MRP}.
+This form of the model includes the MRP prevalence model in \cref{E-biology-data:sec:MRP}.
 
@@
     \includegraphics{transmission/backcalc-regions}
-    \caption[Incidence estimated using backcalculation by region]{%
+    \caption[Incidence estimated using the phenomenological approach by region]{%
         Incidence proportion for (A) all of England and (B) its component regions.
@@
     \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{transmission/backcalc-ages}}
-    \caption[Incidence estimated using backcalculation by age group]{%
+    \caption[Incidence estimated using the phenomenological approach by age group]{%
         Incidence proportion for each age group, on both natural and log scales.
@@
     \centering \includegraphics{transmission/backcalc-start-effect}
-    \caption[Edge effects in backcalculation method]{%
+    \caption[Edge effects in the phenomenological approach method]{%
         First seven days of incidence proportion estimate for the whole of England, highlighting the edge effect.
@@
 Generally, the estimates agree with the SEIR model having much narrower credible intervals.
-The largest discrepancy is that the backcalculation model estimates a second peak in mid-January in several regions (most clearly in London, North West, and the East of England).
+The largest discrepancy is that the phenomenological approach model estimates a second peak in mid-January in several regions (most clearly in London, North West, and the East of England).
 However, except in London, the credible intervals still overlap.
-This is a specific case of a larger pattern: the SEIR model produces smoother estimates of incidence than the backcalculation model.
+This is a specific case of a larger pattern: the SEIR model produces smoother estimates of incidence than the phenomenological model.
 \begin{figure}
@@
         Posterior median and 95\% CrI shown.
-        For the SEIR model, newly RT-PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the backcalculation results.
+        For the SEIR model, newly RT-PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the phenomenological results.
     }
@@
 
-The backcalculation method benefits from being able to be implemented in a fast, approximate inference package (R-INLA) that easily scales to many parameters, including random effects.
+The phenomenological method benefits from being able to be implemented in a fast, approximate inference package (R-INLA) that easily scales to many parameters, including random effects.
 Therefore, it is far more flexible and can easily include time-varying effects, ethnicity, etc.
@@
 This is because the model is not flexible enough to capture the true age distribution of infections.
-The overestimation of prevalence in the 12--16 age group and underestimation in the 17--25 age group is reflected when comparing incidence estimates to those from the backcalculation model, which have reduced bias.
+The overestimation of prevalence in the 12--16 age group and underestimation in the 17--25 age group is reflected when comparing incidence estimates to those from the phenomenological model, which have reduced bias.
 \begin{figure}

### 2024-03-30 introduction.tex
Use phenomenological consistently

@@
 
-    \item In \cref{E-backcalc}, I use a deterministic backcalculation to estimate incidence, including full propagation of the uncertainty from the prevalence estimation.
+    \item In \cref{E-backcalc}, I use a deterministic, phenomenological backcalculation to estimate incidence, including full propagation of the uncertainty from the prevalence estimation.
     Furthermore, I use my estimate of the duration of detectability from general population studies.

### 2024-03-30 thesis.tex
add figure for new likelihood
overleaf changes
Use phenomenological consistently

@@
 \def\@onedot{\ifx\@let@token.\else.\null\fi\xspace}
-\def\eg{e.g\onedot} \def\Eg{\emph{E.g}\onedot}
-\def\ie{i.e\onedot} \def\Ie{\emph{I.e}\onedot}
-\def\cf{c.f\onedot} \def\Cf{\emph{C.f}\onedot}
-\def\etc{etc\onedot} \def\vs{\emph{vs}\onedot}
+\def\eg{e.g\onedot} \def\Eg{{E.g}\onedot}
+\def\ie{i.e\onedot} \def\Ie{{I.e}\onedot}
+\def\cf{c.f\onedot} \def\Cf{{C.f}\onedot}
+\def\etc{etc\onedot} \def\vs{{vs}\onedot}
 \def\wrt{w.r.t\onedot} \def\dof{d.o.f\onedot}
@@
 \newcommand{\ncis}{\vec{n_\text{CIS}}}
-\newcommand{\na}{\vec{n}_\text{a}}
+\newcommand{\na}{\vec{n}_\text{obs}}
 \newcommand{\pcis}{\vec{p_\text{CIS}}}
@@
     Several aspects of the work in this thesis were completed in support of the ongoing response.
-    In particular, an earlier version of the backcalculation work was adopted by the Office for National Statistics to estimate incidence, published in their weekly online reports~\autocite{cisMethodsONS}.
+    In particular, an earlier version of the phenomenological backcalculation was adopted by the Office for National Statistics to estimate incidence, published in their weekly online reports~\autocite{cisMethodsONS}.
     This method was also used for the analysis of several respiratory diseases over winter 2022/3~\autocite{dietzSARSCoV2final}.

### 2024-03-30 transmission/compare_NE.R

@@ -16,7 +16,7 @@ backcalc_prevalence_summary = readRDS(here::here("transmission", "outputs", "reg
     group_by(date, age_group) |>
     median_qi(incidence, prevalence, .simple_names = FALSE) |>
     ungroup() |>
-    mutate(Model = "Backcalculation", age_group = normalise_age_groups(age_group))
+    mutate(Model = "Phenomenological", age_group = normalise_age_groups(age_group))
 
 seir_incidence_summary = load_seir_predictive() |>
     filter(region == "North East") |>

### 2024-03-30 transmission/compare_regions.R

@@ -25,7 +25,7 @@ backcalc_prevalence_summary = readRDS(here::here("transmission", "outputs", "reg
     group_by(date, region) |>
     median_qi(incidence, prevalence, .simple_names = FALSE) |>
     ungroup() |>
-    mutate(Model = "Backcalculation")
+    mutate(Model = "Phenomenological")
 
 p_compare = bind_rows(
     seir_incidence_summary |>

### 2024-03-30 incidence-prevalence.tex
Fix derivation of approximation

@@
     &= \left( \frac{n_t}{\Npop} \right)^2 \var\left(P_t \mid \vec{Z} \right) + \frac{n_t}{\Npop} \E\left(P_t \mid \vec{Z} \right)  - \frac{n_t}{\Npop^2} \left(\E\left(P_t \mid \vec{Z} \right) ^ 2 + \V(P_t \mid \vec{X})\right) \\
-    &= \frac{n_t(n_t - 1)}{\Npop^2} \var\left(P_t \mid \vec{Z} \right) + n_t \frac{\E\left(P_t \mid \vec{Z} \right)}{\Npop}\left(1 - \frac{\E\left(P_t \mid \vec{Z} \right)}{\Npop} \right) \\
+    &= \frac{n_t(n_t-1)}{\Npop} \var\left(P_t \mid \vec{Z} \right) + \frac{n_t}{\Npop} \E\left(P_t \mid \vec{Z} \right)  - \frac{n_t}{\Npop^2} \E\left(P_t \mid \vec{Z} \right) ^ 2.
+    \label{inc-prev:eq:VY1}
 \end{align}
-In the contexts I consider $n_t << \Npop$ because $\Npop$ is the population of England (around 56 million) while $n_t$ is the daily sample size (10s of thousands).
+In the CIS $n_t << \Npop$ because $\Npop$ is the population of England (around 56 million) while $n_t$ is the daily sample size (10s of thousands).
 Therefore:
@@
     &<< \frac{n_t}{\Npop} \var\left(P_t \mid \vec{Z} \right) \\
-    &\leq \frac{n_t}{\Npop} \E\left(P_t \mid \vec{Z} \right) &\text{by \cref{inc-prev:eq:boundVPt}} \\ 
-    &\leq \frac{n_t}{\Npop} \E\left(P_t \mid \vec{Z} \right) \left(1 - \frac{\E(P_t \mid \vec{Z}}{\Npop} \right) &\text{as $P_t \leq \Npop$.}
-\end{align}
-Combining these results:
-\begin{align}
+    &\leq \frac{n_t}{\Npop} \E\left(P_t \mid \vec{Z} \right) &\text{by \cref{inc-prev:eq:boundVPt}}.
+\intertext{Therefore:}
+\frac{n_t}{\Npop} \E\left(P_t \mid \vec{Z} \right) &\approx\frac{n_t(n_t-1)}{\Npop} \var\left(P_t \mid \vec{Z} \right) + \frac{n_t}{\Npop} \E\left(P_t \mid \vec{Z} \right).
+\intertext{Substitute into \cref{inc-prev:eq:VY1}:}
   \var\left( Y_t \mid \vec{Z} \right)
-    &\approx \frac{\E\left(P_t \mid \vec{Z} \right)}{\Npop}\left(1 - \frac{\E\left(P_t \mid \vec{Z} \right)}{\Npop} \right) \\
+    &\approx \frac{n_t}{\Npop} \E\left(P_t \mid \vec{Z} \right)  - \frac{n_t}{\Npop^2} \E\left(P_t \mid \vec{Z} \right) ^ 2 \\
+    &= n_t \frac{\E\left(P_t \mid \vec{Z} \right)}{\Npop} \left(1 - \frac{\E\left(P_t \mid \vec{Z} \right)}{\Npop} \right) \\
     &= \V(Y_t \mid P_t = \E(P_t \mid \vec{Z}))

### 2024-03-30 cis-imperfect-testing.tex
fix ref
Modify notation definitions

@@
 
-To formalise this, define the relevant testing times for episode $j$ as $\sched'_j = \{ t \in \sched_{i(j)} \ssep r_j^{(b)} \leq t \leq r_j^{(e)} + 1 \}$ and let $m_j$ denote the size of this set.
-Denote the elements of $\sched'_j$ by $t_{j,1} < \dots < t_{j,m_j}$.
-The new observation vector for $j$ is $O'(j) = [O(j), \vec{y}_j]^T$ where $\vec{y}_j \in \{ 0, 1 \}^{m_j}$, with $y_{j,l}$ being the test result at time $t_{j,l}$.
+To formalise this, I define analogous variables to those in \cref{E-perf-test:sec:posterior}.
+To start with, define the relevant testing times for episodes with observations $\vec{\nu}_k$ as $\sched'_k = \{ t \in \sched_{i_k} \ssep r_k^{(b)} \leq t \leq r_k^{(e)} + 1 \}$ and let $m_k$ denote the size of this set.
+Denote the elements of $\sched'_k$ by $t_{k,1} < \dots < t_{k,m_k}$.
+Let $\set{E}' = \{ \vec{\nu}_1', \dots, \vec{\nu}_{N_E'}' \}$ be a new set of possible observations of detected episodes, including the pattern of positive and negative tests.
+Represent the observations such that for any $\vec{\nu}_k' \in \set{E}'$, $\vec{\nu}_k' = [\vec{\nu}_{k}, \vec{y}_k]^T$, where $\vec{y}_j \in \{ 0, 1 \}^{m_j}$ with $y_{j,l}$ being the test result at time $t_{j,l}$.
 
-Defining analogous variables to \cref{E-perf-test}, let $\set{D}' = \{ \vec{\nu}'_1, \dots, \vec{\nu}'_{N_{D'}} \}$ where $\vec{\nu}'_k \in \set{D}'_k$ if and only if $\vec{\nu}'_k = [\vec{\nu}_k, \vec{y}_k]^T$ would be a detected observation, equivalent to satisfying all the following conditions.
+$\vec{\nu}'_k \in \set{E}'$ occurs if and only if all the following conditions are met.
 \begin{enumerate}
-  \item $\vec{\nu}_k \in \set{D}$.
-  \item The elements of $\vec{y}_k$ corresponding to the tests at times $r_k^{(b)}$ (its first element) and $l_k^{(e)}$ (its penultimate element) are 1.
-  \item The element of $\vec{y}_k$ corresponding to the test at time $r_k^{(e)} + 1$ (its last element) is 0.
+  \item $\vec{\nu}_k \in \set{E}$.
+  \item The elements of $\vec{y}_k$ corresponding to the tests at times $r_k^{(b)}$ and $l_k^{(e)}$ are positive, \ie $y_{k,1} = y_{k,m_k-1} = 1$.
+  \item The element of $\vec{y}_k$ corresponding to the test at time $r_k^{(e)} + 1$ is negative, \ie $y_{k,m_k} = 0$.
 \end{enumerate}
@@
 
-Similarly, let $p_k'$, $p_{ik}'$, $p_u$, and $p_{iu}$ have the same definition as $p_k$, $p_{ik}$, $p_u$, and $p_{iu}$ respectively except replacing $O$ with $O'$ and $\vec{\nu}_k$ with $\vec{\nu}_k'$, and conditioning on $\psens$.
+Let $O'(j)$ be the observations generated by episode $j$ including the new vector of results.
+As before, $O'(j) \in \set{E}'$ indicates that episode $j$ was detected; otherwise, $O'(j) = \emptyset$.
+
+Similarly, let $p_k'$, $p_{ik}'$, $p_u'$, and $p_{iu}'$ have the same definition as $p_k$, $p_{ik}$, $p_u$, and $p_{iu}$ respectively except replacing $O$ with $O'$ and $\vec{\nu}_k$ with $\vec{\nu}_k'$, and conditioning on $\psens$.
 As before, $p'_k = 1/\Ncis p'_{ik}$ and $p'_u = 1/\Ncis \sum_{i=1}^{\Ncis} p'_{iu}$.
@@
 
-I will modify $p_{ik}$ to form $p_{ik}' = \prob(O'(j) = \nu'_k \mid i(j) = i_k, \vec{\theta})$, taking into account false negatives.
+I will modify $p_{ik}$ to form $p_{ik}' = \prob(O'(j) = \vec{\nu}'_k \mid i(j) = i_k, \vec{\theta})$, taking into account false negatives.
 I consider a mixture of two scenarios, whether the final test in $\sched'_{i(j)}$ is a false positive, with the mixture probability determined by the test sensitivity.

### 2024-03-30 incidence-prevalence.tex
Modify notation definitions

--- 2024-03-30 incidence-prevalence.tex
@@
 
-\chapter{Data generating process} \label{inc-prev}
+\chapter{Data generating processes} \label{inc-prev}
 
@@
 Similarly, $\lambda(d) \in [0,1]$, but can otherwise vary arbitrarily (\ie there is no monotonic or sum constraint).
-The lack of monotonicity or sum constraint on the hazard makes it an attractive parameterization for inference~\autocite{heBayesian}.
+The lack of monotonicity or sum constraint on the hazard makes it an attractive parameterization for inference because these constraints are challenging to implement in many inference frameworks~\autocite{heBayesian}.
 

### 2024-03-30 cis-imperfect-testing.tex
Better notation in 6.3

@@
 
-\section{Likelihood modification for false negatives} \label{imperf-test:sec:modelling}
+\section{Posterior density modification for false negatives} \label{imperf-test:sec:modelling}
 
-In this section, I introduce a simple model of false negatives (\ie allowing $\psens < 1$) into the likelihood derived in \cref{E-perf-test:sec:model}.
-Using a simple model, in particular a constant test sensitivity, means the likelihood remains tractable.
+In this section, I introduce a simple model of false negatives (\ie allowing $\psens < 1$) into the posterior density derived in \cref{E-perf-test:sec:model}.
+Using a simple model, in particular a constant test sensitivity, means calculating likelihood remains tractable.
 
@@
 I will modify $p_{ik}$ to form $p_{ik}' = \prob(O'(j) = \vec{\nu}'_k \mid i(j) = i_k, \vec{\theta})$, taking into account false negatives.
-I consider a mixture of two scenarios, whether the final test in $\sched'_{i(j)}$ is a false positive, with the mixture probability determined by the test sensitivity.
+I consider a mixture of two scenarios, whether the final test in $\sched'_{i(j)}$ is a false negative, with the mixture probability determined by the test sensitivity.
 
@@
 \begin{align}
-&\prob(\vec{y}_k, E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
-&= \prob(\vec{y}_k, l_k^{(e)} \leq E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\ % &\text{as no false positives}
-&= \prob(\vec{y}_k \mid l_k^{(e)} \leq E_j \leq r_k^{(e)}, B_j = b, p_\text{sens}, \vec{\theta}) \prob(l_k^{(e)} \leq E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
+&\prob(\vec{\nu}_k, E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
+&= \prob(\vec{\nu}_k, l_k^{(e)} \leq E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\ % &\text{as no false positives}
+&= \prob(\vec{\nu}_k \mid l_k^{(e)} \leq E_j \leq r_k^{(e)}, B_j = b, p_\text{sens}, \vec{\theta}) \prob(l_k^{(e)} \leq E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
 &= p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right)
@@
 \begin{align}
-&\prob(\vec{y_k}, E_j > r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
-&= \prob(\vec{y_k} \mid E_j > r_k^{(e)}, B_j = b, p_\text{sens}, \vec{\theta}) \prob(E_j > r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
+&\prob(\vec{\nu}_k, E_j > r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
+&= \prob(\vec{\nu}_k \mid E_j > r_k^{(e)}, B_j = b, p_\text{sens}, \vec{\theta}) \prob(E_j > r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
 &= p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} (1 - p_\text{sens}) S_{\vec{\theta}}(r_k^{(e)} - b_j + 2)
@@
 
-These expressions can now be used to form the full replacement for $p_{ia}$, $p'_{ia}$, the probability of observing $\vec{y_k}$ given that an infection occurs in individual $i$.
+These expressions can now be used to form the full replacement for $p_{ik}$, $p'_{ik} = \prob(O'(j ) = \vec{\nu}_k' \mid i(j) = i_k)$.
 First, augment the data with $b_j$, and split into the cases just discussed:
@@
 \end{align}
-Note that if $p_\text{sens} = 1$ then $p_{ia}' = p_{ia}$ (see \cref{perf-test:eq:pia}).
+Note that if $p_\text{sens} = 1$ then $p_{ik}' = p_{ik}$ (see \cref{perf-test:eq:pia}).
 
-If $\psens$ is fixed (\ie has a point prior) and $p(b_j \mid \psens, \vec{\theta}) \propto 1$ (as assumed in \cref{E-perf-test}) then:
+If $\psens$ is fixed (\ie has a point prior) and $\prob(B_j = b \mid \psens, \vec{\theta}) \propto 1$ (as assumed in \cref{E-perf-test}) then:
 \begin{align}
@@
 
-I now modify $p_{iu}$ to form $p_{iu}' = \prob(O'(j) = \emptyset \mid i(j) = i_k, \vec{\theta})$ to take into account false negatives.
+I now modify $p_{iu}$ to form $p_{iu}' = \prob(O'(j) = \emptyset \mid i(j) = i, \vec{\theta})$ to take into account false negatives.
 Several mechanisms for episodes being undetected were considered when deriving $p_{iu}$ in \cref{E-perf-test:sec:prob-undetected}.
@@
 A false negative occurring requires that the episode has not yet ended but a negative still occurs.
-The episode has not yet ended at the time of the test if $e_j = b_j + d_j - 1$ is after the test, that is the duration of the infection $d_j > \tau_{\sched_{i(j)}}(b_j)$.
+The episode has not yet ended at the time of the test if $e_j = b_j + d_j - 1 \geq b_j + \tau_{\sched_{i(j)}}(b_j) + 1$, that is the duration of the infection $d_j \geq \tau_{\sched_{i(j)}}(b_j) + 1$.
 Conditional on the episode having not yet ended, the test result is negative with probability $1 - \psens$.
@@
 Denote the number of days between $b_j$ and the test following the false negative as $\tau^2_{\sched_{i(j)}}(b_j) \stackrel{\text{def}}{=} \tau_{\sched_{i(j)}}(\tau_{\sched_{i(j)}}(b_j) + 1)$.
-Then, the episode ends before this test if $d_j < \tau^2_{\sched_{i(j)}}(b_j)$.
+Then, the episode ends before this test if $d_j \leq \tau^2_{\sched_{i(j)}}(b_j) + 1$.
 

### 2024-03-30 transmission.tex
Overleaf

@@
 Specifying $F_Z$ should consider the need to provide structure to stabilise the problem (see \cref{E-intro:sec:disease-model}).
-In the context of the CIS, it needs to allow borrowing information across time and/or strata (see \cref{E-biology-data:sec:MRP}), which is required because for some stratum/time combinations the data become very sparse.
+In the context of the CIS, it needs to allow borrowing information across time and/or strata (see \cref{E-biology-data:sec:MRP}).
 

### 2024-03-30 incidence-prevalence.tex
comment

--- 2024-03-30 incidence-prevalence.tex
@@
 This relationship will make inference for $\vec{Z}$ much simpler because the latent vector $\vec{P}$ and the dependence of the $y_t$s can be ignored.
+Intuitively, this means that the variance due to the observation process is negligible compared to the variance due to the prevalence process.
 

### 2024-03-31 ATACCC.tex
add extra gof plots for ATACCC
proof

@@
 Finally, I discuss the results (\cref{ATACCC:sec:discussion}).
-I will refine this estimate in the following chapters before its use in estimating transmission in \cref{E-transmission}.
 
@@
 However, the random effect(s) for each individual is drawn towards a common mean, allowing information on their value to be shared between individuals.
-
-Following much of the literature mentioned previously, I will take a Bayesian approach in this chapter.
 Bayesians often refer to random effect models as hierarchical or multilevel models because they can be interpreted as placing a prior at two levels of a parameter hierarchy~\autocite[chapter 5]{gelmanBDA}.
@@
 
-Several features of the \textcite{hakkiOnset} model can be motivated by the desire to apply Stan, a modern and convenient MCMC implementation of the NUTS algorithm (see \cref{E-MCMC}).
+Several features of the \textcite{hakkiOnset} model can be motivated by the desire to apply Stan, a modern and convenient implementation of the NUTS algorithm (see \cref{E-MCMC}).
 The first of these is using a continuous viral load trajectory rather than the more common piecewise linear model.
 As NUTS uses gradient information, it is more efficient when the model is differentiable.
-This leads to the following model for the viral load (see \cref{ATACCC:fig:viral-load-model} for examples):
+This leads to the following model for the viral load (see \cref{ATACCC:fig:viral-load-model}):
 \begin{align}
@@
 \end{align}
-Time, $t$, is continuous and defined uniquely for each individual such that $t = 0$ at the time that individual $i$ was first tested.
+Time, $t$, is continuous and defined for individual $i$ such that $i$s first test is at $t = 0$.
 $v_{\max,i}$ is the maximum viral load, which occurs at time $\tau_{\max,i}$.
@@
 
-The noise in the RT-PCR is modelled by assuming each measurement of log viral load has an iid error distributed $\N(0, \sigma_v^2)$.
+The noise in the RT-PCR test result is modelled by assuming each measurement of log viral load has an iid error distributed $\N(0, \sigma_v^2)$.
 Measurements are left censored at the limit of detection, $\vlod$, meaning that any observation which would give a log viral load of less than $\vlod$ is instead observed as negative.
@@
 The first term on each line is the likelihood contribution from test failures, either a false positive (top line) or false negative (bottom line).
-The second term on the first line is the likelihood of the observed log viral load from non-failed test results.
-The second term on the bottom line is the likelihood of the negative (censored) tests from non-failed test results.
+The second term is the likelihood contribution from non-failed test results, either measurements of viral load (top line) or negative results (bottom line).
 
@@
 Each $\vec{\phi}_i$ is modelled as a log-normal random effects by using the hierarchical prior $\vec{\phi}_i \dist \LN(\vec{\mu}_{\phi,g(i)}, \matr{\Sigma}_\phi)$ where $g(i)$ is the group of individual $i$.
-$\vec{\mu}_{\phi,g} = (\mu_{v_{\max},g}, \mu_{a,g}, \mu_{b,g})^T$ is a vector of the population means of the log-parameters.
+$\vec{\mu}_{\phi,g} = [\mu_{v_{\max},g}, \mu_{a,g}, \mu_{b,g}]^T$ is a vector of the population means of the log-parameters.
 As the random effects have a log-normal, rather than normal, distribution $\exp(\vec{\mu}_{\phi,g})$ is the population median of the parameter value but not the mean.
@@
 Based on the fit to the data achieved when implementing an earlier version of the model (see \cref{ATACCC:sec:original-analysis}), I hypothesised that the cause of this was overcomplicating the model for test failures.
-Specifically, excluding false positives from the model still explains the data well (the results in \cref{fig:paper:duration,ATACCC:fig:goodness-of-fits} do not allow false negatives but fit the data well).
+Specifically, excluding false positives from the model still explains the data well (see \cref{ATACCC:fig:goodness-of-fits}).
 Furthermore, a prior that false negatives and false positives are equally likely is not supported by other information (see \cref{biology-data:sec:measurement-error}).
@@
 
-The modified model produced good fits to the data (see \cref{ATACCC:fig:goodness-of-fits}).
+The modified model produced good fits to the data (see \cref{ATACCC:fig:goodness-of-fits,ATACCC:fig:appendix-goodness-of-fits}).
 Almost all test results are captured in the 95\% credible intervals.
@@
 Each plot is labelled with the individual's position in the dataset, an arbitrary value.
+  Remaining individuals shown in \cref{ATACCC:fig:appendix-goodness-of-fits}.
   \label{ATACCC:fig:goodness-of-fits}}

### 2024-03-31 MCMC-appendix.tex
proof

@@
 The effective sample size can be interpreted as the number of iid samples that would have the same variance as the MCMC sample~\autocites[286]{gelmanBDA}{vehtariRhat}.
-Various techniques have been proposed for computing the effective sample size, I will follow the approach of \textcite{vehtariRhat}, implemented in RStan~\autocite{RStan-2-32-3}.
+Various techniques have been proposed for computing the effective sample size, I follow the approach of \textcite{vehtariRhat}, implemented in RStan~\autocite{RStan-2-32-3}.
 
@@
 NUTS has been implemented in a variety of software packages making it easy to apply to arbitrary posterior distributions.
-I will make use of the implementation in the package \emph{Sampling through adaptive neighborhoods} (Stan)~\autocite{Stan-2-32-2}.
+I make use of the implementation in the package \emph{Sampling through adaptive neighborhoods} (Stan)~\autocite{Stan-2-32-2}.
 Stan implements \emph{automatic differentiation}~\autocite{griewankAutoDiff} to calculate the gradient of the posterior density without any user input.

### 2024-03-31 biology-data.tex
proof

@@
 SARS-CoV-2 is closely related to previous coronaviruses that have caused outbreaks in humans, including SARS-CoV and MERS-CoV.
-Coronaviruses are RNA (Ribonucleic acid) viruses which means they use RNA, rather than DNA (Deoxyribonucleic acid), as their genetic material.
+Coronaviruses are RNA (Ribonucleic acid) viruses, which means they use RNA, rather than DNA (Deoxyribonucleic acid), as their genetic material.
 RNA viruses, which include influenza, mutate rapidly and are the most likely class of viruses to cause pandemics~\autocite{woolhouseRNA}.
@@
 SARS-CoV-2 is an RNA virus, but PCR amplifies DNA.
-Therefore, before using PCR process to detect SARS-CoV-2, the RNA must be converted to DNA.
+Therefore, before using the PCR process to detect SARS-CoV-2, the RNA must be converted to DNA.
 This process is known as \emph{reverse transcription}.
-In reverse transcription, a naturally occurring enzyme known as reverse transcriptase generates cDNA (complementary DNA) from the RNA in the sample~\autocite{valasekPower}.
+In reverse transcription, a naturally occurring enzyme known as \emph{reverse transcriptase} generates cDNA (complementary DNA) from the RNA in the sample~\autocite{valasekPower}.
 Any sequence of cDNA has a one-to-one correspondence with a sequence of RNA.
@@
 The immune system response contribution to the reduction in target cells is several orders of magnitude larger than the reduction in target cells due to viral replication.
+The response is triggered by the viral load and becomes stronger as the viral load increases.
 The rate at which host cells are infected slows over time because there are fewer remaining target cells to infect.
-A host's immune response is triggered by the viral load and becomes stronger as the viral load increases.
 A high viral load is associated with being more infectious, more likely to test positive on an RT-PCR test, and the start of symptoms~\autocite{puhachSARSCoV2,keVivo}.
@@
 For all positive test results, the viral load was measured using the Ct value and then converted to a viral load using a standard curve (see \cref{biology-data:sec:PCR-process}).
-Other data such as demography and symptoms were also collected, although I do not use this information.
+Other data such as demography and symptoms were also collected.
 For further details see \textcite{singanayagamDuration,hakkiOnset}.
@@
   \caption[CIS swab numbers]{%
-    Number of swabs taken by the CIS in the period considered in this thesis (Sep 2020 until Jan 2021).
+    Number of swabs taken by the CIS up until Jan 2021.
     (A) Only enrolment (\ie first) swabs for each individual.
@@
 I use two levels of ethnic grouping.
-Broad groupings (White and non-White), because the White population makes up a large majority of the UK.
+Broad groupings (White and non-White), because the White population makes up over 80\% of the UK.
 I also use more granular ethnicities (Asian, Black, Mixed, White, and Other).
@@
 
-If the CIS was truly a random sample, an unbiased estimator of the population positivity rate would be the proportion of positive tests in the sample; I refer to this estimator as the unadjusted positivity rate.
+If the CIS was truly a sample representative of the general population, an unbiased estimator of the population positivity rate would be the proportion of positive tests in the sample; I refer to this estimator as the unadjusted positivity rate.
 However, in practice, two issues arise with this.
@@
 Define a (sub) population of interest as a set of strata $\set{S}$.
-Then the prevalence in the (sub)population on day $t$ is
+Then the poststratified prevalence in the (sub)population on day $t$ is
 \begin{align}
@@
 Both show similar trends.
-The MRP estimate is normally greater than the uncorrected positivity rate because the underrepresented demographic groups tended to have higher positivity rates.
+The MRP estimate is normally greater than the unadjusted positivity rate because the underrepresented demographic groups tended to have higher positivity rates.
 

### 2024-03-31 cis-imperfect-testing.tex
proof

@@
 Furthermore, in \cref{imperf-test:sec:simulate}, I expanded the simulation study of \cref{E-perf-test:sec:simulation-study} to include false negatives.
-This reproduced the CIS data more faithfully and exhibited the same issues when estimating the duration distribution (see \cref{imperf-test:fig:sim-single-pos}).
+This reproduced the CIS data more faithfully (see \cref{imperf-test:fig:sim-single-pos}) and exhibited the same issues when estimating the duration distribution (not shown).
 \begin{figure}
@@
 % In \cref{imperf-test:fig:bounding-cis-sensitivity} we consider the values produced by assuming the negative following the last positive in an episode could be either a true or false negative as a function of time since the individual was detected (\ie the first positive test in the episode).
-I generate broad bounds, although suggest a declining test sensitivity over the course of an episode (see \cref{imperf-test:fig:bounding-cis-sensitivity}).
+I generate broad bounds, suggestive of a declining test sensitivity over the course of an episode (see \cref{imperf-test:fig:bounding-cis-sensitivity}).
 Small numbers of tests mean that the series is noisy near the start and the end, such as both bounds being 0 on some days after day 40.
@@
 As I assume that there are no false positives, the infection episode must span at least the period $[r^{(b)}_k, l^{(e)}_k]$, a period starting and ending with a positive test.
-This includes all $t \in \sched'_{i(j)}$ except $t = r_k^{(e)}+1$.
+This includes all $t \in \sched'_k$ except $t = r_k^{(e)}+1$.
 Therefore, the test results $\vec{y}_k$, except the test at $r_k^{(e)}+1$, are either true positives or false negatives.
@@
 Consider the negative test at $r_k^{(e)}+1$, the first negative after the start of the episode which may be a false negative.
-It is a false negative if and only if the episode ends after the test, \ie $E_j > r_k^{(e)}$.
+It is a false negative if and only if the episode ends at or after the test, \ie $E_j > r_k^{(e)}$.
 I proceed by considering whether this is the case and with $B_j = b$ known.
@@
 p_{ik}'
-=& \prob(\vec{y_k} \mid p_\text{sens}, \vec{\theta}) \\
-=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( \prob(\vec{y_k}, E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) + \prob(\vec{y_k}, E_j > r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \right) \\
+=& \prob(\vec{\nu}_k \mid p_\text{sens}, \vec{\theta}) \\
+=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( \prob(\vec{\nu}_k, E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) + \prob(\vec{\nu}_k, E_j > r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \right) \\
   &\times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}).
@@
 
-Denote by $\psenss$ the test sensitivity value used when simulating the dataset, with $\psens = v$ indicating the varying test sensitivity was used.
+Denote by $\psenss$ the test sensitivity value used when simulating the dataset, with $\psens = v$ indicating the varying test sensitivity, from \cref{imperf-test:eq:variable-test-sensitivity}, was used.
 Denote by $\psensi$ the constant test sensitivity value assumed during inference.
 
-I use one dataset from \cref{imperf-test:sec:simulate} for each $\psenss \in \{ 0.6, 0.8, 1.0, v \}$ ($\psenss = 1$ is the simulated data in \cref{E-perf-test:sec:simulation-study}).
+I use one dataset from \cref{imperf-test:sec:simulate} for each $\psenss \in \{ 0.6, 0.8, 1.0, v \}$.
 The simulation is otherwise unchanged from that in \cref{E-perf-test:sec:simulation-study}.
@@
 
-The results when $\psenss = v$ (\ie using \cref{imperf-test:eq:variable-test-sensitivity}) are similar to $\psenss = 0.8$ (see \cref{imperf-test:fig:variable-test-sensitivity}).
+The results when $\psenss = v$ are similar to $\psenss = 0.8$ (see \cref{imperf-test:fig:variable-test-sensitivity}).
 This suggests that the simplified model, with constant test sensitivity, is sufficient for recovering the true survival time.
@@
 
-In this section, I apply the approach described in this chapter to the CIS episodes dataset I described in \cref{perf-test:sec:data-notation}.
+In this section, I apply the approach described in this chapter to the CIS episodes dataset I described in \cref{perf-test:sec:data}.
 This is the \numprint{4800} CIS infection episodes first detected between 10 Oct 2020 and 6 Dec 2020 inclusive and have negatives bounding the start and end time of the episode.
@@
 The uninformative prior led to high posterior estimates of $\ntot$, and hence an implausibly large number of episodes with durations of less than five days.
-Therefore, I formed an informative prior for $\ntot$, $\ntot \sim \NBc(\mu\inform, r\inform)$ (as in \cref{E-perf-test:sec:model}, see \cref{E-distributions} for definition), using pre-existing estimates of the total number of infections to give $\mu\inform$ and $r\inform$.
+Therefore, I formed an informative prior for $\ntot$, $\ntot \sim \NBc(\mu\inform, r\inform)$, using pre-existing estimates of the total number of infections to give $\mu\inform$ and $r\inform$.
 \Textcite{birrellRTM2} estimated the total number of infections in England over the time period I consider, with posterior mean \numprint{4136368} and standard deviation \numprint{27932}.
@@
 \todo{Insert correct citation for RTM paper 2 once available}
-Approximating this distribution as a negative binomial and the mean number of infections to the size of the CIS cohort gives the prior $\mu\inform = \numprint{25132}$ and $r\inform = \numprint{22047}$.
+Scaling to the size of the CIS and approximating as a negative binomial gives $\mu\inform = \numprint{25132}$ and $r\inform = \numprint{22047}$.
 
@@
   \centering \includegraphics{cis-imperfect-testing/CIS_final}
-  \caption{
+  \caption[Comparing estimates of the duration distribution.]{
     Duration estimates using the model in this chapter (CIS) alongside those from \cref{E-ATACCC} (ATACCC).
@@
 
-The estimates are most sensitive to the choice of $r$, the strength of the prior on $\ntot$.
-A very weak, almost uninformative, prior on this quantity causes its posterior estimate to be much higher than the estimate from \textcite{birrellRTM2}.
+The estimates are sensitive to the choice of $r$, the strength of the prior on $\ntot$.
+A low value for $r$, giving a very weak, almost uninformative, prior on $\ntot$ causes its posterior estimate to be much higher than the estimate from \textcite{birrellRTM2}.
 When increasing the prior's strength, the posterior estimate moves towards the prior smoothly, as expected (see \cref{imperf-test:fig:ntot}).
@@
   \centering \includegraphics{cis-imperfect-testing/CIS_ntot}
-  \caption[Sensitivity of estimates of $\ntot$ to the prior.]{How the posterior estimate of $\ntot$ changes with the value of $r$ in the prior on $\ntot$.}
+  \caption[Sensitivity of $\ntot$'s posterior to its prior.]{How the posterior estimate of $\ntot$ changes with the value of $r$ in the prior on $\ntot$.}
   \label{imperf-test:fig:ntot}
@@
 Regardless of the assumptions used, a substantial portion of infection episodes last a long time (\eg more than 50 days) (see \cref{imperf-test:fig:cis-estimates,imperf-test:fig:cis-sensitivity}).
-This has implications for surveillance because infections detected by prevalence surveys could have been present for a long time.
-If an intervention such as a lockdown is implemented, these long infections could contribute a significant portion to ongoing prevalence levels.
+This has implications for surveillance because episodes detected by prevalence surveys could have occurred a long time ago.
+If an intervention such as a lockdown is implemented, these long episodes could contribute a significant portion to ongoing prevalence levels.
 Furthermore, care should be taken when designing testing policies (\eg to release individuals from quarantine or discharge them from hospital) because individuals could test RT-PCR positive for a long time, but cannot be assumed to be infectious.
@@
 The posterior distribution for the survival has narrow credible intervals, likely due to the large sample size in the CIS.
-In the following chapters, I will take the posterior mean for the survival time at each point, neglecting this uncertainty, as the basis for estimating transmission.
+In the following chapter, I will take the posterior mean for the survival time at each point, neglecting this uncertainty, as the basis for estimating transmission.
 % First, I will take a backcalculation approach (\cref{E-backcalc}), based on the framework in \cref{E-inc-prev}.

### 2024-03-31 cis-perfect-testing.tex
proof

@@
 In the current context, the initiating event is the episode's start; the terminating event is the episode's end.
-I start by introducing the relevant, standard concepts in survival analysis; a wide range of texts cover them including \textcite[chapter 1]{bogaertsSurvival}.
+I start by introducing the relevant, standard concepts in survival analysis; a wide range of texts cover these concepts including \textcite[chapter 1]{bogaertsSurvival}.
 
@@
 Therefore, we know the times that they were tested.
-This provides information on how long their infections may have occurred.
+This provides information on how long their infections may have been.
 I refer to these events as ``undetected'', rather than truncated, to emphasise this difference.
@@
 Furthermore, I marginalise the posterior density over the number of undetected episodes.
-This makes inference tractable using the large sample size and the computing resources available within the SRS (see \cref{E-biology-data:sec:SRS}).
+This makes inference tractable using the large sample size and limited computing resources available within the SRS (see \cref{E-biology-data:sec:SRS}).
 % In particular, the posterior can be expressed within modern, high-performance Bayesian software such as Stan.
@@
 An individual $i$ has a \emph{test schedule} $\sched_i$.
-The test schedule $\sched_i$ is the set of times individual $i$ is tested at, starting from their last test prior to time 1, if it exists, or their time of enrolment otherwise.
-Note that $\sched_i$ includes any tests that occur after $T$.
+The test schedule $\sched_i$ is the set of times individual $i$ is tested at, starting from their last test prior to time 1, if it exists, or their time of enrolment otherwise; note that $\sched_i$ includes any tests that occur after $T$.
 For example, for the individual in \cref{perf-test:fig:double-interval-censor}, $\sched_i = \{ 0, 7, 14, 21, 28, 56 \}$.
@@
 Now define the probabilities of each possible outcome of $O(j)$.
-For detected outcomes, $\vec{\nu}_k \in \set{E}$, let $p_k = \prob(O(j) = \vec{\nu}_k \mid \vec{\theta})$.
+For detected outcomes, \ie $\vec{\nu}_k \in \set{E}$, let $p_k = \prob(O(j) = \vec{\nu}_k \mid \vec{\theta})$.
 For undetected outcomes, let $p_u = \prob(O(j) = \emptyset \mid \vec{\theta})$.
@@
 Assume, for tractability, that the events $O(j) = \vec{\nu}_k$ and $O(j') = \vec{\nu}_{k'}$ are independent for $j \neq j'$; this assumption is discussed in \cref{perf-test:sec:discussion}.
+
 Then:
@@
 Decompose $p_k$ as $p_k = p_{ik} \prob(i(j) = i_k \mid \vec{\theta})$
-where $p_{ik} = \prob(O(j) = \nu_k \mid i(j) = i_k, \vec{\theta})$.
+where $p_{ik} = \prob(O(j) = \vec{\nu}_k \mid i(j) = i_k, \vec{\theta})$.
 % This is valid as $\prob(O(j) = \nu_k \mid i(j) \neq i_k) = 0$ due to the condition here being equivalent to equating the two vectors' final elements.
@@
 % Equivalently, that the first test at or after $b$ is after $e$, or that there is no negative test prior to $b$.
-To formalise this, first define $\tau_{\sched_i}(t)$ as the time until the next test at or after time $t$ in the schedule $\sched_i$:
+First define $\tau_{\sched_i}(t)$ as the time until the next test at or after time $t$ in the schedule $\sched_i$:
 \begin{align}
@@
 In this section I specify the form of $S_{\vec{\theta}}(t)$.
-I parameterize $S$ in terms of the hazard $S_{\vec{\theta}}(t) = \prod_{i=1}^{t-1} (1 - \lambda(i))$ (see \cref{E-inc-prev:sec:definitions}).
+I parameterize $S$ in terms of the hazard $S_{\vec{\theta}}(t) = \prod_{i=1}^{t-1} (1 - \lambda_{i})$ (see \cref{E-inc-prev:sec:definitions}).
 Therefore, $\vec{\theta} = [\lambda_1, \dots, \lambda_{\dmax-1}]^T$.
@@
 
-RW2 prior encodes that the hazard should be linearly changing with some random changes at each time step.
+An RW2 prior encodes that the hazard should be linearly changing with some random changes at each time step.
 Specifically, the difference on the logit scale between $\lambda_t$ to $\lambda_{t+1}$ is the difference between $\lambda_{t-1}$ and $\lambda_t$ plus some normally distributed noise.
@@
   \centering \includegraphics{cis-perfect-testing/ataccc-approximation-survival}
-  \caption[Approximating the ATACCC posterior survival as a logit-normal]{Comparison of the posterior estimate of the survival from \cref{E-ATACCC} and its approximation with a logit-normal distribution (median and central 95\% interval).
+  \caption[Approximating the ATACCC posterior survival]{Comparison of the posterior estimate of the survival from \cref{E-ATACCC} and its approximation with a logit-normal distribution (median and central 95\% interval).
 The survival is a function of all the hazards up until that point, therefore the correlation between the hazards must be well-approximated to have these agree.
@@
 \end{align}
-where $k_t$, $\alpha_0$, and $\beta_0$ are hyperparameters; and, as before, $h_t$ is an estimate of the hazard at time $t$ from \cref{E-ATACCC}.
+where $k_t$, $\alpha_0$, and $\beta_0$ are hyperparameters.
 An intuition for what this distribution represents derives from a conjugate model for $\lambda_t$ with a beta prior and a binomial likelihood.
@@
     \item Extract the test schedules for each individual who had at least one test during the period of interest.
-    \item Draw an episode start time, $b_{j}$ for each individual uniformly at random between 2 July 2020 (100 days before the period where a detected episode would be included) and 6 December 2020 (the end of this period).
+    \item Draw an episode start time, $b_{j}$ for each individual uniformly at random between 2 July 2020 ($t = -100$) and 6 December 2020 ($t = T$).
     \item Draw a duration of episode for each individual, $d_i$, based on the distribution described below. Then calculate the end of their infection episode, $e_{j} = b_{j} + d_i - 1$.
@@
 I base this assumption on the estimate from \cref{E-ATACCC} with an inflated tail to be consistent with the CIS.
-I modify the tail based on an unpublished analysis of CIS data~\citePersonalComms{A.\ Sarah Walker}.
+I modify the tail based on an unpublished analysis of the CIS data~\citePersonalComms{A.\ Sarah Walker}.
 Walker uses survival analysis to estimate the duration from CIS data.
@@
 
+The prior on the total number of infections was centred on the true value with a dispersion parameter of 1.
+
 \subsection{Implementation}
@@
 Convergence was assessed using Rhat and ESS as described in \cref{E-MCMC:sec:convergence}, and all runs checked for divergent transitions.
-
-The prior on the total number of infections was centred on the true value with a dispersion parameter of 1.
 This meant that the prior was very diffuse.
@@
 A smoothing prior performed poorly, although that may be influenced by the choice of hyperparameters.
-Reducing the amount of smoothing or type of smoothing may help.
+Reducing the amount of smoothing or changing the type of smoothing may help.
 More flexible smoothing priors could be used, such as a Gaussian Process for additional flexibility and smoothing and different timescales~\autocite{saulGaussian}.
@@
 This is not the case, as immunity means that an individual previously infected is less likely to be infected in the future.
-Additionally, the same individual cannot have multiple concurrent infections, due to the way that the episodes dataset is constructed (see \cref{E-biology-data:sec:cis-episodes}) limiting certain combinations from occurring.
+Additionally, the same individual cannot have multiple detected concurrent infections, due to the way that the episodes dataset is constructed (see \cref{E-biology-data:sec:cis-episodes}); this limits certain combinations from occurring.
 In practice, this means that $n_k > 1$ cannot occur.
@@
 Covariates can be included with a variety of models, the simplest being a proportional hazards model.
-A major challenge here is that the calculation of $p_t$ is greatly complicated because it is no longer linear in the survival function.
 

### 2024-03-31 conclusion.tex
proof

@@
 On the other hand, a mechanistic approach provides more information.
-Where possible, both methods should be implemented; discrepancies between results may highlight deficiencies in modelling assumptions.
+Where possible, both methods should be implemented; discrepancies between results can highlight deficiencies in modelling assumptions.
 
@@
 In addition, the data would need to be linked to relevant records.
-Second, a transmission model, such as the one in \cref{E-SEIR}, could jointly model severe events and prevalence.
+Second, a mechanistic model, such as the one in \cref{E-SEIR}, could jointly model severe events and prevalence.
 Such an approach has previously been used~\autocite{daviesAssociation,ironsEstimating,knockKey,nicholsonImproving,pooleyEstimation,birrellRTM2} but without the sophistication of the observation model developed in \cref{E-SEIR:sec:observation}.

### 2024-03-31 incidence-prevalence.tex
proof

@@
 I start by defining these terms in \cref{inc-prev:sec:definitions}.
-I then move on to explaining the three processes that link the incidence, prevalence, and the data: the \emph{infection process}, the \emph{prevalence process}, and the \emph{observation process}.
-
-In (\cref{inc-prev:sec:infection-process}) I introduce the infection process, describing the infections' generating mechanism; in (\cref{inc-prev:sec:prevalence-process}) the prevalence process, which generates prevalence from the incidence; and in (\cref{inc-prev:sec:observation-process}), the observation process, which generates the data from the prevalence through the sampling mechanism of a population prevalence survey, guided by the design of the CIS.
+I then explain the three relevant models: one for the infection process (\cref{inc-prev:sec:infection-process}); one for the prevalence process, the relevant disease model (\cref{inc-prev:sec:prevalence-process}) linking infections to prevalence; and finally the observation process for a prevalence survey like the CIS (\cref{inc-prev:sec:observation-process}).
 
@@
 I refer to individuals with the disease as \emph{prevalent} individuals.
-Whether an individual is prevalent is often expensive or error-prone to measure and hence is a latent quantity.
-Furthermore, what is meant by ``having'' a disease can depend on context; for example, it could be the presence of symptoms, the ability to transmit, or detectable status (the difference between these was discussed in \cref{E-biology-data:sec:natural-history})
+What is meant by ``having'' a disease can depend on context; for example, it could be the presence of symptoms, the ability to transmit, or detectable status (the difference between these was discussed in \cref{E-biology-data:sec:natural-history})
 Within this thesis, I define an individual as prevalent on day $t$ if they are detectable (detectable is defined in \cref{E-biology-data:sec:detectable}).
@@
 I denote the number of prevalent individuals on day $t$ as $P_t$; hence, the prevalence is $P_t/\Npop$.
+$P_t$ corresponds with $W_t$ in \cref{E-intro:eq:disease-model}.
 
@@
 %(a common alternative convention for survival analysis with discrete variables is to define the survival function as $\prob(D >d)$).
-The \emph{hazard} on day $d$ is the probability of recovering on day $d$ conditional on still being positive on day $d$: $\lambda(d) = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
-Throughout this thesis, the survival function will be the primary quantity of interest, and normally parameterized in terms of the hazard; these two quantities are related by $S(d) = \prod_{i=1}^{d-1} (1 - \lambda(i))$.
+The \emph{hazard} on day $d$ is the probability of recovering on day $d$ conditional on still being positive on day $d$: $\lambda_{d} = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
+Throughout this thesis, the survival function will be the primary quantity of interest, and normally parameterized in terms of the hazard; these two quantities are related by $S(d) = \prod_{i=1}^{d-1} (1 - \lambda_{i})$.
 
@@
 Therefore, $S(d) \in [0, 1]$ and decreases monotonically.
-Similarly, $\lambda(d) \in [0,1]$, but can otherwise vary arbitrarily (\ie there is no monotonic or sum constraint).
+Similarly, $\lambda_{d} \in [0,1]$, but can otherwise vary arbitrarily (\ie there is no monotonic or sum constraint).
 The lack of monotonicity or sum constraint on the hazard makes it an attractive parameterization for inference because these constraints are challenging to implement in many inference frameworks~\autocite{heBayesian}.
@@
 A Poisson process is a widely used type of counting process both within and outside of epidemiology.
-It makes no explicit assumptions on the underlying biological or epidemiological mechanism and can be a flexible model.
 
 Afterwards, in \cref{SEIR:sec:mechanistic-models}, I introduce \emph{mechanistic models}.
-Mechanistic models explicitly describe the population and transmission between individuals.
-In particular, they include the feedback loop inherent in infectious diseases, where increased infections at a time $t$ lead to more infections at times $> t$.
-They vary greatly in their details and realism~\autocite{murilloMultiscale}.
+Unlike phenomenological models, they include the feedback loop inherent in infectious diseases, where increased infections at a time $t$ lead to more infections at times $> t$.
 
@@
 The proportion of the population that have recovered by time $t$ is $r(t)$.
-The model's state is $\vec{x}(t) = (s(t), i(t), r(t))^T$.
+The model's state is $\vec{x}(t) = [s(t), i(t), r(t)]^T$.
 
@@
 \frac{di(t)}{dt} &= \lambda(t) s(t) - \gamma i(t) \\
-\frac{dr(t)}{dt} &= \gamma i(t)
+\frac{dr(t)}{dt} &= \gamma i(t) \\
+\lambda(t) &= \beta i(t)
 \end{align}
-where $\lambda(t) = \beta i(t)$, is the \emph{force of infection}: the risk of infection for a susceptible individual at time $t$~\autocite[17]{keelingModeling}.
-The dynamics of the model are described by the parameters $\vec{\theta} = (\beta, \gamma)^T$, where
+where $\lambda(t)$, is the \emph{force of infection}: the risk of infection for a susceptible individual at time $t$~\autocite[17]{keelingModeling}.
+The dynamics of the model are described by the parameters $\vec{\theta} = [\beta, \gamma]^T$, where
 $\beta$ is the transmission rate, absorbing several terms as described below;
@@
     For longer time periods, the model can be expanded to include waning immunity~\autocite[40]{keelingModeling}.
-    \item A discretised incidence proportion for day $t$ can be defined as $Z_t / \Npop = s(t) - s(t+1)$.
+    \item A discretised incidence proportion for day $t$ ($t$ an integer) can be defined as $Z_t / \Npop = s(t) - s(t+1)$.
         However, this will not in general give an integer value to $Z_t$.
@@
 
-The basic SIR model is the deterministic approximation of a stochastic model where the duration of time in the infectious compartment is exponentially distributed~\autocite[96]{keelingModeling}.
+The basic SIR model is the mean-field approximation of a stochastic model where the duration of time in the infectious compartment is exponentially distributed~\autocite[96]{keelingModeling}.
 This is an unrealistic assumption for most diseases and would lead to underestimating of the reproduction number~\autocite{lloydRealistic,wearingAppropriate}.
@@
 
-The reproduction number in terms of the parameters is unchanged by this extension, \ie $\R = \beta / \gamma$, however, the growth rate in the exponential phase will be slower than before~\autocite[41]{keelingModeling}.
+The reproduction number in terms of the parameters is unchanged by this extension, \ie $\R = \beta / \gamma$, however, the growth rate in the initial exponential phase is slower than before~\autocite[41]{keelingModeling}.
 
@@
 The state vector is now a vector of vectors, one for each of the compartments.
-For example, in the basic SIR model, $\vec{x}(t) = (\vec{s}(t), \vec{i}(t), \vec{r}(t))^T$.
+For example, in the basic SIR model, $\vec{x}(t) = [\vec{s}(t), \vec{i}(t), \vec{r}(t)]^T$.
 The $a$th element of each vector ($a \in \set{A}$) is the proportion of strata $a$ in that compartment.
-For example, $\vec{s}(t) = (s_1(t), \dots, s_A(t))^T$ where $s_a(t)$ is the proportion of the $a$th strata that are susceptible at time $t$.
+For example, $\vec{s}(t) = [s_1(t), \dots, s_A(t)]^T$ where $s_a(t)$ is the proportion of the $a$th strata that are susceptible at time $t$.
 Hence, $N_a s_a(t)$ is the number of susceptible people in stratum $a$, where $N_a$ is the number of individuals in stratum $a$.
@@
 The force of infection is now a vector, defined as $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i}_j(t)$ where $\vec{i}_j$ is a vector such that $i_{j,a}$ is the proportion of stratum $a$ in the $j$th infectious state.
-Their interpretation can be arrived at by considering a single infectious individual in strata $a'$, analogous to the single stratum case.
+
+$\vec{\lambda}_t$ can be arrived at by considering a single infectious individual in strata $a'$, analogous to the single stratum case.
 Assume any such individual comes into contact with individuals in strata $a$ at a rate $\kappa_{aa'}$ per day.
@@
 Reproduction numbers in models with structured populations are more complex than the homogeneous case.
-The WAIFM allows calculation of the number of secondary infections generated by a single infection in the $a$th strata,  this follows directly from its definition as $\frac{1}{\gamma} \sum_{a' \in \set{A}} \beta_{aa'}$.
+The WAIFM allows calculation of the number of secondary infections generated by a single infection in the $a$th strata,  this follows directly from its definition as $\frac{1}{\gamma} \sum_{a' \in \set{A}} \beta_{a'a}$.
 However, how to average these contributions to calculate the basic reproduction number is not obvious.
+
 The standard solution to this was proposed by \textcite{diekmannDefinition} (see \textcite[chapter 7]{diekmannMathematical} for a gentler, but still rigorous, explanation).
@@
 This $\R$ keeps the same interpretation as in the homogeneous case, \ie the average number of secondary infections caused by a single infection in a fully susceptible population, where the average is taken over the distribution implied by the dominant eigenvector.
-The important threshold that exponential growth occurs if and only if $\R > 1$ remains.
+The important threshold that, in the mean-field approximation, exponential growth occurs if and only if $\R > 1$ remains.
 
@@
 
+The prevalence process is the disease model (see \cref{E-intro:sec:disease-model}) relating incidence and prevalence.
 The relationship between incidence and prevalence can be derived by considering the probability that an individual is prevalent at any time following an infection.
@@
 %Denote by $F_D$ the cdf of $D_i$, that is $F_D(d) = \prob(D_i \leq d)$.
-From this, we can derive the relationship between incidence and prevalence.
+In what follows, I use this to derive the relationship between incidence and prevalence.
 
@@
 This is because they are linked by the $Z_t$s, which are themselves dependent (as explained in \cref{inc-prev:sec:infection-process}).
-However, if we condition on the vector of incidence, $\vec{Z} = (Z_1, Z_2, \dots)^T$, then we will induce some independence.
+However, if we condition on the vector of incidence, $\vec{Z} = [Z_1, Z_2, \dots]^T$, then we will induce some independence.
 This reduces the stochasticity to only that attributable to the prevalence process.
@@
 
-The expectation in \cref{inc-prev:eq:EPt} is a discrete convolution equation in the same form as \cref{E-intro:eq:disease-model} with $\zeta = 1$.
+The expectation in \cref{inc-prev:eq:EPt} is a discrete convolution equation in the same form as \cref{E-intro:eq:disease-model} with $\zeta = 1$ and $f(i) = S(i+1)$.
 % The equation is of the same form as the equation used for HIV/AIDS backcalculation, as well as being found in many other fields~\autocite[and references therein]{brookmeyerBackcalculation}.
@@
 \end{align}
+where $\vec{P} = [P_1, P_2, \dots]$.
 In \cref{E-backcalc}, I will need to solve for this system for $\vec{Z}$ given a known $\hat{P} = \E(\vec{P} \mid {Z})$.
@@
 The sampling is described mathematically by the observation process.
-The survey design considered here reflects the CIS, although it would be applicable to many prevalence surveys.
+I consider the CIS's study design, ignoring the household-based element, although it would be applicable to many prevalence surveys.
 
-On day $t$ (for $t = 1, \dots, T$), a prevalence survey such as the CIS samples $n_t$ individuals at random from the population of interest and tests if they are prevalent.
-Therefore, I model the number of positive tests, $Y_t$, as $Y_t \mid P_t, n_t, \Npop \dist \Bin(n_t, P_t/\Npop)$.
-The conditioning on $n_t$ and $\Npop$ is implicit in what follows.
+On day $t$ (for $t = 1, \dots, T$), the CIS samples $n_t$ individuals at random from the population of interest and tests if they are prevalent.
+Therefore, I model the number of positive tests, $Y_t$, as $Y_t \mid P_t, n_t, \Npop \dist \Bin(n_t, P_t/\Npop)$; the conditioning on $n_t$ and $\Npop$ is implicit in what follows.
 These quantities are fixed by the study design and population choice, and hence I assume independent of $P_t$ and $Z_t$.
 
-The $Y_t$s are conditionally independent, given the population prevalences at each time, $\vec{P} = (P_1, \dots, P_t)^T$. 
+The $Y_t$s are conditionally independent, given the population prevalences at each time, $\vec{P}$.
 %For now, I assume that the sample is chosen uniformly at random from the population of interest.
@@
 I will use prevalence surveys to infer incidence.
-I base this inference on the likelihood, $p(\vec{y} \mid \vec{Z})$.
-Here $\vec{y} = (y_1, \dots, y_T)$ is the vector of observed number of positive tests in the prevalence survey on each day and $\vec{Z} = (Z_{-\dmax}, Z_{-\dmax+1} \dots, Z_T)$ is the vector of incidence on each day.
-The incidence back to day $-\dmax$ will affect $P_1$ and therefore need to be included in $\vec{Z}$.
+I base this inference on the likelihood, $p(\vec{y} \mid \vec{Z})$, where $\vec{y} = [y_1, \dots, y_T]$ is the vector of observed number of positive tests in the prevalence survey on each day.
 
-This section argues that $p(\vec{y} \mid \vec{Z}) \approx \prob(\vec{X} = \vec{y} \mid \vec{P} = \E(\vec{P} \mid \vec{Z}))$.
+This section argues that $p(\vec{y} \mid \vec{Z}) \approx \prob(\vec{Y} = \vec{y} \mid \vec{P} = \E(\vec{P} \mid \vec{Z}))$.
 As the $y_t$s are independent conditional on $\vec{P}$, it follows that $p(\vec{y} \mid \vec{Z}) \approx \prod_t \prob(Y_t = y_t \mid P_t = \E(P_t \mid \vec{Z}))$.
-This relationship will make inference for $\vec{Z}$ much simpler because the latent vector $\vec{P}$ and the dependence of the $y_t$s can be ignored.
+This relationship makes likelihood-based inference for $\vec{Z}$ much simpler because the latent vector $\vec{P}$ and the dependence of the $y_t$s can be ignored.
 Intuitively, this means that the variance due to the observation process is negligible compared to the variance due to the prevalence process.
 
-The basis of the argument is that the first two moments of $\vec{y} \mid \vec{Z}$ and $\vec{X} \mid \vec{P} = \E(\vec{P} \mid \vec{Z})$ have negligible difference.
+The basis of the argument is that the first two moments of $\vec{y} \mid \vec{Z}$ and $\vec{Y} \mid \vec{P} = \E(\vec{P} \mid \vec{Z})$ have negligible difference.
 The first moment follows directly from the tower property of expectations.
-The majority of the argument is to show that the variance-covariance matrix $\V(\vec{X} \mid \vec{Z}) \approx \V(\vec{X} \mid \vec{P} = \E(P \mid Z))$.
+The majority of the argument is to show that the variance-covariance matrix $\V(\vec{Y} \mid \vec{Z}) \approx \V(\vec{Y} \mid \vec{P} = \E(P \mid Z))$.
 There are two steps to this argument.
@@
     &= \left( \frac{n_t}{\Npop} \right)^2 \var\left(P_t \mid \vec{Z} \right) + \frac{n_t}{\Npop} \E\left(P_t \mid \vec{Z} \right)  - \frac{n_t}{\Npop^2} \E\left(P_t^2 \mid \vec{Z} \right) \\
-    &= \left( \frac{n_t}{\Npop} \right)^2 \var\left(P_t \mid \vec{Z} \right) + \frac{n_t}{\Npop} \E\left(P_t \mid \vec{Z} \right)  - \frac{n_t}{\Npop^2} \left(\E\left(P_t \mid \vec{Z} \right) ^ 2 + \V(P_t \mid \vec{X})\right) \\
+    &= \left( \frac{n_t}{\Npop} \right)^2 \var\left(P_t \mid \vec{Z} \right) + \frac{n_t}{\Npop} \E\left(P_t \mid \vec{Z} \right)  - \frac{n_t}{\Npop^2} \left(\E\left(P_t \mid \vec{Z} \right) ^ 2 + \V(P_t \mid \vec{Z})\right) \\
     &= \frac{n_t(n_t-1)}{\Npop} \var\left(P_t \mid \vec{Z} \right) + \frac{n_t}{\Npop} \E\left(P_t \mid \vec{Z} \right)  - \frac{n_t}{\Npop^2} \E\left(P_t \mid \vec{Z} \right) ^ 2.

### 2024-03-31 introduction.tex
proof

--- 2024-03-29 introduction.tex
@@
 Fully quantifying the burden of historical pandemics is challenging due to the absence of systematic records.
-However, our best records suggests that, since 1600, they have (on average) occurred every four years with a mean death toll of 5.6 million~\autocite{maraniNovelEpidemics}.
+However, our best records suggests that, since 1600, they have, on average, occurred every four years with a mean death toll of 5.6 million~\autocite{maraniNovelEpidemics}.
 Recent estimates suggest that the current impact of pandemics is, on average, over 2.5 million deaths per year~\autocite{madhavPandemicMortality}.
@@
 Broadly, a reproduction number describes the number of secondary infections generated by a single infection.
-Here, following \textcite{pellisEstimation}, and building on over a century of infectious disease dynamics literature, I define three that are broadly useful.
+Here, following \textcite{pellisEstimation}, and building on over a century of infectious disease dynamics literature, I define three below.
 
@@
 \emph{Incidence} is the number of new infections in a given time period (see \cref{E-inc-prev:sec:definitions} for a formal definition).
-It is the primary metric for the magnitude of a pandemic .
-Incidence informs the short-term future burden of the disease, for example, the future number of severe events (hospitalisations or deaths); although this requires a severity measure, such as the IFR (infection fatality rate), the proportion of infected individuals who die.
+It is the primary metric for the magnitude of a pandemic.
+Incidence informs the short-term future burden of the disease, for example, the future number of severe events (hospitalisations or deaths) through a severity measure, such as the infection fatality rate, the proportion of infected individuals who die.
 
@@
 
-Data on severe disease is typically less affected by biases.
+Data on the severe consequences of an infection typically less affected by biases.
 An extreme example is the development of AIDS following infection by HIV.
@@
 
-However, severe disease can be rare in some population subgroups (\eg young, otherwise healthy individuals infected with SARS-CoV-2~\autocite{wardSero,bhopalChildren}), meaning they rarely appear in this data.
+However, severe consequences can be rare in some population subgroups (\eg young, otherwise healthy individuals infected with SARS-CoV-2~\autocite{wardSero,bhopalChildren}), meaning they rarely appear in this data.
 Therefore, the data does not provide much information on incidence in those subgroups.
@@
 Prevalence can be estimated through a \emph{prevalence survey}.
-A prevalence survey is a designed study which performs diagnostic testing of a sample of the population; they are typically very expensive to power appropriately.
-On the other hand, they are a source of prevalence-type data that can be unbiased.
-For instance, testing a representative proportion of the population (or a subpopulation of interest) for a marker of current infection.
+A prevalence survey is a designed study which performs diagnostic testing of a sample of the population; they are typically very expensive to power appropriately but can provide unbiased prevalence-type data.
+An example of an unbiased prevalence survey is testing a representative proportion of the population (or a subpopulation of interest) for a marker of current infection.
 
@@
 However, recruiting a representative sample of the population is challenging, and the results of these studies are often not generalisable to the whole population.
-Furthermore, these studies often require a large cohort to be appropriately powered.
-Large cohort studies are expensive to run.
+Furthermore, these studies often require a large sample size to be appropriately powered, making them expensive to run.
 
@@
 The models with the weakest assumptions are \emph{phenomenological} models.
-These make almost no assumptions about the transmission process, viewing incidence as resulting from a counting process (introduced in full in \cref{E-inc-prev:sec:infection-process}).
-A common model is the Poisson process, and the intensity of the process is the target of inference.
+These make almost no assumptions about the transmission process, viewing incidence as resulting from a generic counting process (introduced in full in \cref{E-inc-prev:sec:infection-process}), such as a Poisson process.
 % The use of Poisson processes to model incidence of an infectious disease was very popular in the context of HIV/AIDS\todo{maybe a ref for Poisson in HIV/AIDS?}.
@@
 These relate incidence in the past to current (and future) incidence but do not explicitly represent the susceptible population.
-A common type of semi-mechanistic models is that based on a \emph{renewal equation}~\autocite{bhattSemimechanistic,fraserRenewal} of the most form:
+A common type of semi-mechanistic models is that based on a \emph{renewal equation}~\autocite{bhattSemimechanistic,fraserRenewal}.
+In discrete time, it takes the form:
 \begin{align}
-    Z(t) &= \int_{0}^{t} \beta(t,\tau) Z(t-\tau) d \tau
+    \E(Z_t \mid Z_0, \dots, Z_{t-1}) &= \sum_{\tau=0}^t \beta(t,\tau) Z_{t-\tau} d \tau
     \label{intro:eq:renewal}
 \end{align}
-where $Z(t)$ is the expected incidence at time $t$ and $\beta(t, \tau)$ is the expected rate at which an individual infected $\tau$ time units ago infects individuals at calendar time $t$.
-The model parameters can be used to define an effective reproduction number as either $\Re = \int_0^\infty \beta(t, \tau) d\tau$ or $\Re = \int_0^\infty \beta(t+\tau, \tau) d\tau$~\autocite{gosticPractical}.
+where $Z_t$ is the expected incidence at time $t$ and $\beta(t, \tau)$ is the expected rate at which an individual infected $\tau$ time units ago infects individuals at calendar time $t$.
+The model parameters can be used to define an effective reproduction number as either $\Re = \sum{\tau=0}^\infty \beta(t, \tau)$ or $\Re = \sum_{\tau=0}^\infty \beta(t+\tau, \tau)$~\autocite{gosticPractical}.
 % A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
@@
 
-The infection process is hardly ever directly observed; however, we often observe a second process.
-\emph{Disease models} link the transmission model to the value, $W_t$ of this second process at time $t$.
+The infection process is hardly ever directly observed; however, we often observe a consequence of an infection.
+\emph{Disease models} link the infection model to the number of these consequences, $W_t$ at time $t$.
 For example, $W_t$ could be the incidence or prevalence of symptomatic individuals on day $t$.
-$W_t$ will then be recorded through some system (\eg a testing programme) described by the observation model (see next section), which could introduce additional complications.
+$W_t$ will then be recorded through some system (\eg a testing programme) described by the observation model (see the next section).
 
@@
 
-Observation models link $W_t$ to a possibly biased and/or noisy observation of it at time $t$.
+Observation models link $W_t$ to a (possibly biased and/or noisy) observation of it at time $t$.
 
@@
 
-Modelling the duration directly considers only the binary result at each test (positive or negative).
+Modelling the duration directly considers only the binary result at each test (positive or negative), using less information than a biomarker approach.
 This approach generally requires fewer parameters reducing the computational burden.
@@
 
-The surveillance systems used in response to SARS-CoV-2 are unique. 
-The main data sources informing the response to the SARS-Cov-2 epidemic in England were based on: community testing; severe events; seroprevalence and prevalence surveys.
+The surveillance systems set-up in response to SARS-CoV-2 were on a scale never previously seen.
+The main data sources informing the response to the SARS-Cov-2 epidemic in England were based on: community testing, severe events, seroprevalence surveys, and prevalence surveys.
 Each of these data sources have strengths and weaknesses, see \textcite{royalSocietyRnumber} and references therein for further details on each of these data sources.
@@
 
-Severe events counts were generally a more reliable measure of the burden of the disease.
-The criteria for being a severe event are normally well-defined and generally well-recorded.
-Therefore, these data were subject to fewer biases than the cases detected through community testing.
+The number of severe events was generally a more reliable measure of the burden of the disease, subject to fewer changes than the number of cases detected through community testing.
 However, SARS-CoV-2 rarely led to severe events in younger individuals, so severe event counts were insensitive to changes in transmission only affecting younger individuals.
@@
 The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys in representative samples of the general population.
-These were designed to avoid the biases of community testing and attempt to observe the general population.
+These were designed to avoid the biases of community testing and observe the general population.
 These surveys use RT-PCR (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
 The first survey was the CIS (Coronavirus (COVID-19) Infection Survey), run by the ONS (Office for National Statistics)~\autocite{CIS,cisMethodsONS}.
-The CIS was a longitudinal household study, testing a large number of individuals weekly or monthly for SARS-CoV-2 infection (see \cref{E-intro:sec:cis} for a detailed description).
+The CIS was a longitudinal, household-based study, testing a large number of individuals weekly or monthly for SARS-CoV-2 infection (see \cref{E-intro:sec:cis} for a detailed description).
 
@@
 This meant that using REACT data to infer transmission in these periods requires strong model assumptions.
-A particular instance of this is that a peak in prevalence in late December 2020 occurred between rounds of the REACT survey, complcating any analysis of REACT data over that period~\autocite{rileyREACTround8}.
+A particular instance of this is that a peak in prevalence in late December 2020 occurred between rounds of the REACT survey, complicating any analysis of REACT data over that period~\autocite{rileyREACTround8}.
 
@@
 
-One of the studies, which I use in this thesis, is the ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts) study~\autocite{singanayagamCommunity,hakkiOnset}.
+Another study, which I use in this thesis, is the ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts) study~\autocite{singanayagamCommunity,hakkiOnset}.
 ATACCC recruited individuals from contacts of known infections reported to NHS Test and Trace, and collected data on them for 20 days.
@@
 
-The massive impact of the SARS-CoV-2 pandemic led to data collection on a scale never previously seen for an infectious disease.
-Analysis of this data was carried out rapidly, to provide the timely results needed to inform the ongoing and evolving response.
+Analysis of pandemic data was carried out rapidly, to provide the timely results needed to inform the ongoing and evolving response.
 Therefore, a deep understanding of the potential of each data source was not possible.
@@
 Further details are in the recent review \textcite{sunModeling}.
-The theory behind the framework underpins much of the work in this thesis.
+I build on the theory behind the framework in this thesis.
 
-The original formulations~\autocite{brookmeyerMethod,rosenbergBackcalculation} use a Poisson process as the infection model; a disease model with a delay function, $f$, representing the pdf of the time between HIV infection and AIDS diagnosis and assumed $\zeta=$; and a trivial observation model where all infections are (eventually) detected.
+The original formulations~\autocite{brookmeyerMethod,rosenbergBackcalculation} use a phenomenological Poisson process as the infection model; a disease model with a delay function, $f$, representing the pdf of the time between HIV infection and AIDS diagnosis and assumed $\zeta=1$; and a trivial observation model where all diagnoses are recorded.
 
-In (see \cref{E-inc-prev}) I will adapt this backcalculation framework for use with prevalence survey data, such as the CIS.
+In \cref{E-inc-prev}, I will adapt this backcalculation framework for use with prevalence survey data, such as the CIS.
 The use of prevalence-type data induces a dependency structure between the values of the latent process $W_1, \dots, W_t$; I include the dependency by modifying the disease model.
@@
 Previous work has attempted to estimate $f$.
+
 An initial meta-analysis estimated that individuals test positive for SARS-CoV-2 for a mean of 14.6 days (95\% CI: 9.3--20.0 days)~\autocite{cevikShedding}.
-The longest reported observed duration was 83 days, although no individuals were infectious beyond day nine.
+The longest reported observed duration was 83 days, although no individuals were infectious beyond day 9.
 However, most studies included in the meta-analysis involved hospitalised patients or had unclear inclusion criteria, and hence may not be representative of the general population.
-In addition, summary estimates are insufficient for backcalculation, which requires a distribution of the duration of detectability.
+In addition, summary estimates are insufficient for backcalculation, which requires a full distribution.
 
@@
 \Textcite{ealesCharacterising} estimated the duration of detectability from REACT data.
-Specifically, individuals who tested positive in round 8 of REACT (6th--22 Jan 2021) were invited to take two additional swab tests.
+Specifically, individuals who tested positive in round 8 of REACT (6--22 Jan 2021) were invited to take two additional swab tests.
 Taking the time of the initial positive test as time 0, \textcite{ealesCharacterising} estimated the probability of remaining positive at each day after the initial positive test.
 They made a strong parametric assumption that the probability decreases exponentially, with a sensitivity analysis assuming a plateau before the decay.
-This assumption is necessary because they had at most 17 days of follow-up, and hence could not estimate the tail of the distribution.
+Strong assumptions are necessary because they had at most 17 days of follow-up, and hence could not estimate the tail of the distribution.
 Overall, they estimate a median duration of 14.0 days (95\% CI: 12.9--15.4 days) with substantial positive correlation with each individual’s viral load at the time of the initial positive test (\ie a higher initial viral load was associated with a longer duration of detectability).
@@
 The association could be due to individuals with a higher initial viral load being (on average) more recently infected at the time of the initial positive test.
-The study design does not provide any information on how long before the initial positive test the infection occurred.
-This issue also means the study underestimates the duration of detectability, because the infection's true start time would be prior to the first positive test.
+The study design prevents assessing this hypothesis because the design does not provide any information on how long before the initial positive test the infection occurred.
+This issue also means the study underestimates the duration of detectability, because the infection's true start time would be prior to the initial positive test.
 
 For the purposes of this thesis, I require estimates of the duration of detectability applicable to the general population.
-As no such estimates of the full distribution of the duration are available in the literature, a key component of this work is devoted to estimating them.
+As no such estimates of the full distribution of the duration are available in the literature, a key component of this thesis is estimating them.
 To do so, I will combine the information from the ATACCC and CIS studies.
 
-\subsection{Existing estimation from prevalence surveys}
+\subsection{Existing use of prevalence surveys}
 
@@
 
-To address this question, I analyse data from the CIS study.
-I use, as a case study, data from the SARS-CoV-2 epidemic in England from Mon 31 Aug 2020 until Sun 25 Jan 2021.
-This period has several interesting features making it suitable for my aim.
-First, the CIS was of sufficient size and transmission was sufficiently high to provide enough data to analyse.
+To address this question, I use a case study.
+Specifically, the CIS data from the SARS-CoV-2 epidemic in England from Mon 31 Aug 2020 until Sun 25 Jan 2021.
+This case study is suitable for addressing this question for several reasons.
+First, the CIS was of sufficient size, and transmission was sufficiently high, to provide enough data for the analyses.
 Second, incidence began to rise over this period, after the fall due to the first lockdown.
@@
 
-    \item In \cref{E-backcalc}, I use a deterministic backcalculation to estimate incidence, including full propagation of the uncertainty from the prevalence estimation.
-    Furthermore, I use my estimate of the duration of detectability from general population studies.
+    \item In \cref{E-backcalc}, I use a deterministic, phenomenological backcalculation to estimate incidence, including full propagation of the uncertainty from the prevalence survey.
+    Furthermore, I use my estimate of the duration of detectability, which is from general population studies.
 
     \item In \cref{E-SEIR}, I develop a compartmental model for SARS-CoV-2 transmission, relying \emph{only} on CIS data. 
-    I translate my estimate of the distribution of the duration of detectability into a progression through model states through an Erlang---Coxian distribution; this distribution is more flexible than standard approaches in the infectious disease literature.
+    I translate my estimate of the distribution of the duration of detectability into a progression through model states through an Erlang--Coxian distribution; this distribution is more flexible than standard approaches in the infectious disease literature.
     I include age-stratification in the transmission model, allowing estimation of age-specific transmission.
@@
 Almost all code, excluding dependencies, was written from scratch for these analyses.
-The exceptions are the Markov chain Monte Carlo sampler used in \cref{E-SEIR}, the model for prevalence used in \cref{E-backcalc}, and the original viral load model used in \cref{E-ATACCC:sec:hakki}.
+The exceptions are the Markov chain Monte Carlo sampler used in \cref{E-SEIR}, the model for prevalence used in \cref{E-biology-data:sec:MRP}, and the original viral load model used in \cref{E-ATACCC:sec:hakki}.
 
@@
 
-In \cref{E-inc-prev}, I formally describe the stochastic processes which generate infections, prevalence, and the observations of prevalence.
+In \cref{E-inc-prev}, I formally describe the infection, disease, and observation models for a prevalence survey.
 I show the distribution of the duration of detectability is the key quantity linking incidence and prevalence.
@@
 This provides a reliable estimate of the bulk of the distribution (out to around 20 days),  but shows that estimating the tail requires a large sample size and long follow-up. 
-These features are present in the CIS, in \cref{E-perf-test} I provide a framework for estimating the duration of detectability assuming perfect tests from prevalence surveys such as the CIS.
+These features are present in the CIS.
+In \cref{E-perf-test}, I provide a framework for estimating the duration of detectability assuming perfect tests from prevalence surveys such as the CIS.
 However, as the assumption of perfect tests is not realistic; in \cref{E-imperf-test}, I extend the framework to include false negatives.
@@
 
-Finally, I conclude in \cref{E-conclusion} with a discussion of the limitations of this work and ideas for future research.
+Finally, I conclude in \cref{E-conclusion} with a discussion of my findings, the limitations of this work, and ideas for future research.
 

### 2024-03-31 thesis.tex
proof

@@
     % One of these, the CIS (Coronavirus Infection Survey), was a longitudinal survey run from April 2020 to March 2023.
-    The CIS (Coronavirus Infection Survey) was a large-scale prevalence survey, with participants recruited randomly from the general population, thereby avoiding these issues.
+    The CIS (Coronavirus Infection Survey) was a large-scale, longitudinal prevalence survey, with participants recruited randomly from the general population, thereby avoiding these issues.
 
@@
     Therefore, I first consider how to estimate the distribution of this duration in the general population.
-    I extend a pre-existing model of how an individual's viral load changes over time to estimate the duration of detectability.
+    I adapt a pre-existing model of how an individual's viral load changes over time for this purpose.
     % I take two complementary approaches using different datasets.
@@
     I use this detailed model to exploit frequent (daily) sampling of individuals in a cohort of contact traced individuals.
-    However, the cohort was small and the follow-up time short meaning that the tail of the duration distribution is poorly estimated.
+    The cohort was small and the follow-up time short meaning that the tail of the duration distribution is poorly estimated.
 
@@
     The CIS data poses several challenges because it is doubly interval censored and some individuals in the cohort have infections that are never detected.
-    I apply a survival analysis framework to analyse this data, which had previously only been used in simpler settings.
-    In addition, I develop novel methodology to incorporate false negatives into the framework.
+    I develop a survival analysis framework to analyse this data, including novel methodology to incorporate false negatives into the framework.
     % The results showed that false negatives in the CIS data needed to be included in the model.
@@
 
-    Finally, I use this estimate to infer the incidence and transmission of SARS-CoV-2 in England from the CIS.
+    Finally, I use this estimate to infer the incidence and transmission of SARS-CoV-2 in England, using only the CIS data.
     I fully propagate the sampling error, accurately include the shape of the duration distribution, and estimate the quantities by age; previous work lacked at least one of these.
@@
     Thank you for everything.
-
-    Having written this, I realise how fortunate I am to have been supported by such great people.
-    I can never thank you enough.
 }
@@
     Several aspects of the work in this thesis were completed in support of the ongoing response.
-    In particular, an earlier version of the phenomenological backcalculation was adopted by the Office for National Statistics to estimate incidence, published in their weekly online reports~\autocite{cisMethodsONS}.
-    This method was also used for the analysis of several respiratory diseases over winter 2022/3~\autocite{dietzSARSCoV2final}.
+    In particular, an earlier version of the phenomenological backcalculation developed in \cref{backcalc} was adopted by the Office for National Statistics to estimate incidence, published in their weekly online reports~\autocite{cisMethodsONS}.
+    This method was also used for an analysis of several respiratory diseases over winter 2022/3~\autocite{dietzSARSCoV2final}.
 
@@
 
-    During this period, I have been involved in several other pieces of work and discussion contributing to the response, not directly used in this thesis.
+    During this period, I have been involved in several other pieces of work and discussion in response to the pandemic, not directly used in this thesis.
     These have led to several publications, listed in \cref{publications} for completeness.

### 2024-03-31 transmission-appendix-INLA.tex
proof

@@
 Each observation has an associated linear predictor $\vec{\eta} = [\eta_1, \dots, \eta_n]^T$.
-The linear predictor and, optionally, some hyperparameters $\vec{\theta}$ give the conditional independence structure:
+The linear predictor and, optionally, some hyperparameters $\vec{\theta}$ have the conditional independence structure:
 \begin{align}
@@
 For the linear regression model, $\matr{A} = \matr{X}$, $\vec{x} = \vec{\beta}$, and $\matr{\Sigma}(\vec{\theta})$ would be a fixed diagonal matrix specifying the prior on the regression coefficients.
-In the (generalised) linear model case $\vec{\theta}$ is fixed; however, LGMs allow an arbitrary prior distribution on $\vec{\theta}$ which means that some elements of $\vec{x}$ can be random effects (or more complex structures).
+In the (generalised) linear model case $\matr{\Sigma(\vec{\theta})}$ is fixed; however, LGMs allow an arbitrary prior distribution on $\vec{\theta}$ which means that some elements of $\vec{x}$ can be random effects (or more complex structures).
 

### 2024-03-31 transmission.tex
proof

@@
 A particular issue is that using MCMC, sufficient covariates to adjust for the non-response bias, and a period of more than 6--8 weeks leads to a model that is computationally prohibitive ~\citePersonalComms{Koen Pouwels}.
-Instead, previous MRP implementations for the CIS used INLA (see \cref{E-biology-data:sec:MRP}).
+Instead, previous MRP implementations for the CIS used INLA (as in \cref{E-biology-data:sec:MRP}).
 
 INLA-based inference can only be applied to LGMs (\emph{Latent Gaussian Models}); see \cref{E-transmission:sec:INLA} for further details on INLA and LGMs.
-The LGM framework requires that a transformation of $P_{i,t}$ is linear in $\vec{g} = [g_{{\vec{\varphi}}}(1, 1), \dots, g_{{\vec{\varphi}}}(1, T), \dots, g_{{\vec{\varphi}}}(2, 1), \dots, g_{{\vec{\varphi}}}(I, T)$ where $I$ and $T$ are the number of strata and days respectively.
+The LGM framework requires that a transformation of $P_{i,t}$ is linear in $\vec{g} = [g_{{\vec{\varphi}}}(1, 1), \dots, g_{{\vec{\varphi}}}(1, T), \dots, g_{{\vec{\varphi}}}(2, 1), \dots, g_{{\vec{\varphi}}}(I, T)]^T$ where $I$ and $T$ are the number of strata and days respectively.
 
@@
 Dividing each side of \cref{inc-prev:eq:forward-substitute} by the total number of individuals in each stratum transforms the number of prevalent individuals to prevalence and incidence to incidence proportion.
-The algorithm uses a deterministic relationship between incidence and prevalence within a stratum, as \cref{E-inc-prev:sec:observation-process} justifies; \ie, the estimate of the incidence is negligibly affected by approximating the distribution $\vec{P} \mid \vec{Z}$ by its mean.
+The algorithm uses a mean-field approximation to the relationship between incidence and prevalence within a stratum, as \cref{E-inc-prev:sec:observation-process} justifies.
 
@@
         \For{$t = -\dmax, \dots, 0$}{
-            $Z^{(j)}_{i,t} = P^{(j)}_{i,1} / \sum_{k=1}^{\dmax} S(k)$
+            Set $Z^{(j)}_{i,t} = P^{(j)}_{i,1} / \sum_{k=1}^{\dmax} S(k)$ \;
         }
@@
 In this case, the incidence prior to day $t$ imply that the prevalence on day $t$ is higher than that predicted by the fitted regression model for this strata/day combination.
+In this case, I replace the negative value for $Z_{i,t}$ with zero.
 
@@
 Alternatively, it could be a consequence of schools remaining open in the November lockdown; 20--24 year olds rarely interact with school-aged children (see \cref{E-SEIR:fig:age-contacts})
-Many other explanations are possible, and further work is required to determine causality.
+Further work is required to determine causality.
 \begin{figure}
@@
     \centering \includegraphics{transmission/backcalc-start-effect}
-    \caption[Edge effects in the phenomenological approach method]{%
+    \caption[Edge effects in the phenomenological approach]{%
         First seven days of incidence proportion estimate for the whole of England, highlighting the edge effect.
@@
 In the mechanistic modelling literature, this part of the model is referred to as the observation model, however, in the terminology of \cref{E-inc-prev}, it includes both the prevalence process and the observation process.
+
 Having specified the full model (\cref{SEIR:sec:methods-application}), I conducted a simulation study to validate the inference process and identifiability (\cref{SEIR:sec:sim-study}), before applying the model to the CIS data (\cref{SEIR:sec:application}).
@@
 I link the modelled proportion of individuals that are RT-PCR positive to the data using a beta-binomial likelihood (as discussed in \cref{E-biology-data:sec:clustering}).
-If on each day $t$, each stratum $i$ is observed to have $y_{at}$ positives out of $n_{at}$ tests, then the likelihood is:
+If on each day $t$, each stratum $a$ is observed to have $y_{at}$ positives out of $n_{at}$ tests, then the likelihood is:
 \begin{align}
@@
 
-Equating the expressions in \cref{SEIR:eq:fullODEs} and \cref{SEIR:eq:exponential-growth}, gives the following solution for the initial state of several compartments.
+Equating the expressions in \cref{SEIR:eq:fullODEs} and \cref{SEIR:eq:exponential-growth}, gives the following solution for the initial state of most of the compartments.
 \begin{align}
@@
 The adaptive algorithm is based on \textcite[algorithm 4]{andrieuTutorial}, with the implementation based on \textcite{ghoshApproximate}'s work.
-Proposals are accepted or rejected using the standard Metropolis--Hastings (MH) acceptance probability.
+Proposals are accepted or rejected using the standard Metropolis--Hastings acceptance probability.
 See \cref{SEIR:MCMC-algorithm} for full details.
@@
 
-I ran a simulation study to study the inference procedure's performance and the model's practical identifiability.
+I ran a simulation study to investigate the inference procedure's performance and the model's practical identifiability.
 I generated 100 simulated datasets representing the East of England using the following procedure.
@@
 Next, I solved the ODE system in \cref{SEIR:eq:fullODEs} for each parameter set.
-All other aspects of the ODEs, including the data-informed contact matrices, $\matr{\kappa}$, are as previously described in this section.
+All other aspects of the ODEs, including the data-informed contact matrices, $\matr{\kappa}(t)$, are as previously described in this section.
 Population sizes were those of the East of England, and the time period the same as the main analysis.
@@
 Having generated 100 datasets, I used the inference procedure in \cref{SEIR:sec:inference} to find the posterior distribution conditional on each dataset.
-Hence, I ended up with 100 parameter sets distributed according to the priors (the ``generative values"), with an associated dataset and posterior distribution for it.
+Hence, I ended up with 100 parameter sets distributed according to the priors (the ``generative values''), with an associated dataset and posterior distribution for it.
 
@@
 
-In this set-up, the parameters $i^+$, $q_c$, and $\psi$ are fully identified; $\pi_0$ is not identified at all; the other parameters are partially identified, their posterior has moved away from the prior but there is insufficient information in the data to fully overcome the prior (see \cref{SEIR:fig:true-vs-posterior}).
+In this set-up, the parameters $i^+$, $q_c$, and $\psi$ are fully identified; $\pi_0$ and $\rho$ are not identified at all; the other parameters are partially identified, their posterior has moved away from the prior but there is insufficient information in the data to fully overcome the prior (see \cref{SEIR:fig:true-vs-posterior}).
 The ability to identify $q_c$ is an advantage of using data which includes measurements on children, unlike data on hospitalisations or deaths.
@@
     \makebox[\textwidth][c]{\includegraphics{SEIR/CIS/attack_rates}}
-    \caption{Posterior median and 95\% CrI of the attack rate in each region and age group.}
+    \caption[Posterior estimates of the attack rate.]{Posterior median and 95\% CrI of the attack rate in each region and age group.}
     \todo[inline]{Maybe compare this to blood donors, as used in the prior.}
@@
 
-These patterns can be interpreted as an instance of the common bias---variance trade-off.
+These patterns can be interpreted as an instance of the common bias--variance trade-off.
 The SEIR model imposes stronger assumptions about the epidemic dynamics.

### 2024-03-31 introduction.tex
proof ch 1

@@
 
-However, \emph{seroprevalence} surveys are more feasible.
+\emph{Seroprevalence} surveys are more feasible.
 Seroprevalence is the proportion of individuals in the population which have antibodies against a particular infection, indicating that the individual was previously infected~\autocite{cdcSeroprevalence}.
-Many more individuals have ever been infected than are currently infected, therefore, seroprevalence surveys can be powered much more cheaply~\autocite{wuSeroprevSimulation}.
+Many more individuals have ever been infected than are currently infected; therefore, seroprevalence surveys can be powered much more cheaply~\autocite{wuSeroprevSimulation}.
 However, seroprevalence is a delayed measure because there is a lag between an individual being infected and them developing antibodies.
@@
 \begin{align}
-    \E(Z_t \mid Z_0, \dots, Z_{t-1}) &= \sum_{\tau=0}^t \beta(t,\tau) Z_{t-\tau} d \tau
+    \E(Z_t \mid Z_0, \dots, Z_{t-1}) &= \sum_{\tau=1}^t \beta(t,\tau) Z_{t-\tau} d \tau
     \label{intro:eq:renewal}
@@
 where $Z_t$ is the expected incidence at time $t$ and $\beta(t, \tau)$ is the expected rate at which an individual infected $\tau$ time units ago infects individuals at calendar time $t$.
-The model parameters can be used to define an effective reproduction number as either $\Re = \sum{\tau=0}^\infty \beta(t, \tau)$ or $\Re = \sum_{\tau=0}^\infty \beta(t+\tau, \tau)$~\autocite{gosticPractical}.
+The model parameters can be used to define an effective reproduction number as either $\Re = \sum_{\tau=0}^\infty \beta(t, \tau)$ or $\Re = \sum_{\tau=0}^\infty \beta(t+\tau, \tau)$~\autocite{gosticPractical}.
 % A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
@@
 The infection process is hardly ever directly observed; however, we often observe a consequence of an infection.
-\emph{Disease models} link the infection model to the number of these consequences, $W_t$ at time $t$.
+\emph{Disease models} link the infection model to the number of these consequences, $W_t$, at time $t$.
 For example, $W_t$ could be the incidence or prevalence of symptomatic individuals on day $t$.
@@
 \end{align}
-where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that result in $W_t$; and, conditional on being observed, $f(i)$ describes the (discretised) probability that an infection is included in the count $W_t$ at $i$ time units after the infection.
+where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that result in the consequence of interest; and, conditional on being observed, $f(i)$ describes the (discretised) probability that an infection is included in the count $W_t$ at $i$ time units after the infection.
 $f$ is an example of a \emph{delay function}, and $f$ is commonly the pdf of a distribution, or a transformation of it.
@@
 
-In the application of the backcalculation to the CIS data, the delay function $f$ represents the duration of detectability.
+In the application of \cref{intro:eq:disease-model} to the CIS data, the delay function $f$ represents the duration of detectability.
 Previous work has attempted to estimate $f$.
@@
 
-\Textcite{mccabeCISincidence}, adapting the methodology \textcite{ealesAppropriately} developed for REACT, estimate $\Re$ and the growth rate from the CIS data.
+\Textcite{mccabeCISincidence}, adapting the methodology \textcite{ealesAppropriately} developed for REACT, estimate $\Re$ and $r$ from the CIS data.
 They first estimate prevalence using summary statistics from the CIS, then derive $\Re$ and $r$ using a semi-mechanistic model fitted to their prevalence estimates.
@@
 
-A secondary aim of this thesis is to improve our understanding of the epidemic in England over this period, and our understanding of the SARS-CoV-2 virus more broadly.
+The secondary aim of this thesis is to improve our understanding of the epidemic in England over this period, and our understanding of the SARS-CoV-2 virus more broadly.
 The results of the analyses I perform in pursuit of the primary aim will naturally contribute to this second aim.
@@
     \item In \cref{E-SEIR}, I develop a compartmental model for SARS-CoV-2 transmission, relying \emph{only} on CIS data. 
-    I translate my estimate of the distribution of the duration of detectability into a progression through model states through an Erlang--Coxian distribution; this distribution is more flexible than standard approaches in the infectious disease literature.
+    I translate my estimate of the distribution of the duration of detectability into a progression through model states.
+    To do this, I use an Erlang--Coxian distribution; this distribution is more flexible than standard approaches in the infectious disease literature.
     I include age-stratification in the transmission model, allowing estimation of age-specific transmission.
@@
 
-The secondary aim of this thesis is providing new estimates to improve our understanding of SARS-CoV-2 in general, and the epidemic in England over the period I study.
+The secondary aim of this thesis is providing new estimates to improve our understanding of the epidemic in England over the period I study.
 

### 2024-03-31 transmission.tex
include kappa time dependencies

@@
 $\beta(t)$ is the only time-varying fitted parameter.
-Therefore, in practice, it will also absorb other effects such as any error in $\matr{\kappa}$.
+Therefore, in practice, it will also absorb other effects such as any error in $\matr{\kappa}(t)$.
 
@@
 
-The only component of $\matr{\beta}$ that allows age-dependent heterogeneity in the temporal changes is $\matr{\kappa}$.
+The only component of $\matr{\beta}$ that allows age-dependent heterogeneity in the temporal changes is $\matr{\kappa}(t)$.
 It is plausible that this would not capture all behavioural changes.
 For example, the introduction and relaxation of control measures within schools.
-$\matr{\kappa}$ only captures changes in in-school transmission through changes in school attendance, not incorporating the probability of transmission, which could be modelled as changes in $q_{aa}$ for $a$ of school age.
+$\matr{\kappa}(t)$ only captures changes in in-school transmission through changes in school attendance, not incorporating the probability of transmission, which could be modelled as changes in $q_{aa}$ for $a$ of school age.
 Similarly, there is evidence that $q_c$ is different for Alpha and pre-Alpha variants~\autocite{zhuRole}, which could again motivate these changing over time.

### 2024-03-31 incidence-prevalence.tex
proof ch 3

@@
 This chapter formally defines incidence and prevalence, and the relationship between them and prevalence survey data.
-In the course of explaining this relationship, it will become clear that the duration of RT-PCR positivity, and its distribution in the population, is the vital parameter which will allow estimation of incidence from prevalence, a major aim of this thesis.
+In the course of explaining this relationship, it will become clear that the duration of RT-PCR positivity, and its distribution in the population, is the vital parameter which will allow estimation of incidence from prevalence, a major component of the aim of this thesis.
 
 I start by defining these terms in \cref{inc-prev:sec:definitions}.
-I then explain the three relevant models: one for the infection process (\cref{inc-prev:sec:infection-process}); one for the prevalence process, the relevant disease model (\cref{inc-prev:sec:prevalence-process}) linking infections to prevalence; and finally the observation process for a prevalence survey like the CIS (\cref{inc-prev:sec:observation-process}).
-
-\Cref{inc-prev:sec:conclusion} concludes the chapter.
+I then explain the three relevant models: one for the infection process (\cref{inc-prev:sec:infection-process}); one for the prevalence process, the relevant disease model that links infections to prevalence (\cref{inc-prev:sec:prevalence-process}); and finally the observation process for a prevalence survey like the CIS (\cref{inc-prev:sec:observation-process}).
 
@@
 
-I use discrete time in this chapter, except in \cref{SEIR:sec:mechanistic-models}, with the time unit of a day. However, results generalise in the obvious way to other time units.
+I use discrete time in this chapter, except in \cref{SEIR:sec:mechanistic-models}, with the time unit of a day; results generalise in the obvious way to other time units.
 I choose discrete time for pragmatic reasons.
-Most epidemiological data are only available at daily granularity or, when more granular data exists, is often unreliable.
+Most epidemiological data are only available at daily granularity or, when more granular data exists, are often unreliable.
 Furthermore, within-day patterns produce variations in transmission (\eg due to sleep) and data (\eg due to logistical considerations around collection), but these variations are not of interest in most contexts.
@@
 I refer to individuals with the disease as \emph{prevalent} individuals.
-What is meant by ``having'' a disease can depend on context; for example, it could be the presence of symptoms, the ability to transmit, or detectable status (the difference between these was discussed in \cref{E-biology-data:sec:natural-history})
+What is meant by ``having'' a disease can depend on context; for example, it could be the presence of symptoms, the ability to transmit, or detectable status (the difference between these was discussed in \cref{E-biology-data:sec:natural-history}).
 Within this thesis, I define an individual as prevalent on day $t$ if they are detectable (detectable is defined in \cref{E-biology-data:sec:detectable}).
@@
 I denote the number of prevalent individuals on day $t$ as $P_t$; hence, the prevalence is $P_t/\Npop$.
-$P_t$ corresponds with $W_t$ in \cref{E-intro:eq:disease-model}.
+$P_t$ corresponds to $W_t$ in \cref{E-intro:eq:disease-model}.
 
-\emph{Duration}, or \emph{time-to-event}, distributions are expressed in several equivalent ways~\autocite[17]{sunStatisticala}.
+\emph{Duration}, or \emph{time-to-event}, distributions can be expressed in several equivalent ways~\autocite[17]{sunStatisticala}.
 The duration distribution I will be concerned with is the distribution of the duration of detectability.
@@
 Similarly, $\lambda_{d} \in [0,1]$, but can otherwise vary arbitrarily (\ie there is no monotonic or sum constraint).
-The lack of monotonicity or sum constraint on the hazard makes it an attractive parameterization for inference because these constraints are challenging to implement in many inference frameworks~\autocite{heBayesian}.
+The lack of monotonicity or sum constraint on the hazard makes it an attractive parameterization for inference because these constraints are challenging to enforce in many inference frameworks~\autocite{heBayesian}.
 
@@
 
-In an epidemic, there are two features of a Poisson process: the underlying process intensity, $\nu(r)$, and the single realisation that occurred, $K(t)$; either of these can be the target of inference.
+In an epidemic, there are two relevant quantities of a Poisson process: the underlying process intensity, $\nu(r)$, and the single realisation that occurred, $K(t)$; either of these can be the target of inference.
 The realised counts are normally of greater interest because we care about the realised epidemic, not what may happen if there were future realisations of the same epidemic~\autocite{beckerDependent,brookmeyerMethod}.
@@
 % A super-spreading event is an event where many more people than average are infected.
-% That is, $Z_t >> \E(Z_t)$ for some $t$ during which a super-spreading event occurs.
+% That is, $Z_t \gg \E(Z_t)$ for some $t$ during which a super-spreading event occurs.
 % Coronaviruses, such as SARS-CoV-2, have particularly high overdispersion~\autocite{endoEstimating,adamClustering}{mccloskeySARS}.
@@
 A common simplification is to use discrete time and a Markov assumption, \ie $\vec{x}(t)$ is only defined for integer $t$ and $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta}) = p(\vec{x}(t) \mid \vec{x}(t-1), \vec{\theta})$ respectively; the Markov assumption means that the state at time $t$ only depends on the state at time $t-1$~\autocite{birrellEvidence}.
-$x(t)$ and $x(t-1$ are now dependent, relaxing the assumption of independent increments in the Poisson process formulation of the infection process (see \cref{E-inc-prev:sec:infection-process}).
+$x(t)$ and $x(t-1)$ are now dependent, relaxing the assumption of independent increments in the Poisson process formulation of the infection process (see \cref{E-inc-prev:sec:infection-process}).
 
-Deterministic models are defined by the state vector $\vec{x}(t)$ being a deterministic function of the parameters $\vec{\theta}$ and the initial state $\vec{x}(0)$; equivalently, $\var(\vec{x}(t) \mid \vec{x}(0), \theta) = \matr{0}$~\autocite{birrellEvidence}.
+Deterministic models are defined by the state vector $\vec{x}(t)$ being a deterministic function of the parameters $\vec{\theta}$ and the initial state $\vec{x}(0)$; equivalently, $\var(\vec{x}(t) \mid \vec{x}(0), \vec\theta) = \matr{0}$~\autocite{birrellEvidence}.
 Deterministic models are also known as \emph{mean-field} models because they represent the mean behaviour of a corresponding stochastic model.
 When the number of infectious individuals is large, the law of large numbers implies that the mean-field behaviour is a good approximation of the fully stochastic model~\autocite[20]{diekmannMathematical}.
-The approximation works especially well if used to make inference based on noisy observations of the system, when the observation noise is much larger than the infection process noise; I make a formalisation of this argument to justify a mean-field approximation of the prevalence process in \cref{E-inc-prev:sec:observation-process}.
+The approximation works especially well if used to make inference based on noisy observations of the system, when the observation noise is much larger than the infection process noise; I make a formalisation of this form of argument to justify a mean-field approximation of the prevalence process in \cref{E-inc-prev:sec:observation-process}.
 
@@
 
-The basis of the argument is that the first two moments of $\vec{y} \mid \vec{Z}$ and $\vec{Y} \mid \vec{P} = \E(\vec{P} \mid \vec{Z})$ have negligible difference.
+The basis of the argument is that the first two moments of $\vec{Y} \mid \vec{Z}$ and $\vec{Y} \mid \vec{P} = \E(\vec{P} \mid \vec{Z})$ have negligible difference.
 The first moment follows directly from the tower property of expectations.
@@
 \end{align}
-In the CIS $n_t << \Npop$ because $\Npop$ is the population of England (around 56 million) while $n_t$ is the daily sample size (10s of thousands).
+In the CIS $n_t \ll \Npop$ because $\Npop$ is the population of England (around 56 million) while $n_t$ is the daily sample size (10s of thousands).
 Therefore:
@@
     \frac{n_t(n_t - 1)}{\Npop^2} \var\left(P_t \mid \vec{Z} \right)
-    &<< \frac{n_t}{\Npop} \var\left(P_t \mid \vec{Z} \right) \\
+    &\ll \frac{n_t}{\Npop} \var\left(P_t \mid \vec{Z} \right) \\
     &\leq \frac{n_t}{\Npop} \E\left(P_t \mid \vec{Z} \right) &\text{by \cref{inc-prev:eq:boundVPt}}.
@@
 
-A similar argument will show that the correlation between $y_t$ and $y_{t'}$ ($t \neq t'$) is negligible.
+A similar argument will show that the correlation between $Y_t$ and $Y_{t'}$ ($t \neq t'$) is negligible.
 The basic structure of this argument is as the previous one.
@@
   &= \frac{n_t n_{t'} \lvert \cov(P_t, P_{t'}) \rvert}{\Npop^2 \sqrt{\var(Y_t) \var(Y_{t'})}}  \\
-  &<< \frac{n_t n_{t'} \lvert \cov(P_t, P_{t'}) \rvert}{\Npop^2 \sqrt{\var(P_t) \var(P_{t'})}} &\text{as $\V(Y_t) >> \V(P_t)$} \\
+  &\ll \frac{n_t n_{t'} \lvert \cov(P_t, P_{t'}) \rvert}{\Npop^2 \sqrt{\var(P_t) \var(P_{t'})}} &\text{as $\V(Y_t) \gg \V(P_t)$} \\
   &= \frac{n_t n_{t'}}{\Npop^2} \lvert \cor(P_t, P_{t'}) \rvert &\text{by def of $\cor$}\\

### 2024-03-31 cis-imperfect-testing.tex
proof ch 6

@@
 In this section, I introduce a simple model of false negatives (\ie allowing $\psens < 1$) into the posterior density derived in \cref{E-perf-test:sec:model}.
-Using a simple model, in particular a constant test sensitivity, means calculating likelihood remains tractable.
+Using a simple model, in particular a constant test sensitivity, means calculating the likelihood remains tractable.
 
@@
 To formalise this, I define analogous variables to those in \cref{E-perf-test:sec:posterior}.
-To start with, define the relevant testing times for episodes with observations $\vec{\nu}_k$ as $\sched'_k = \{ t \in \sched_{i_k} \ssep r_k^{(b)} \leq t \leq r_k^{(e)} + 1 \}$ and let $m_k$ denote the size of this set.
+To start with, define the relevant testing times for any episode $j$ with $O(j) = \vec{\nu}_k$ as $\sched'_k = \{ t \in \sched_{i_k} \ssep r_k^{(b)} \leq t \leq r_k^{(e)} + 1 \}$ and let $m_k$ denote the size of this set.
 Denote the elements of $\sched'_k$ by $t_{k,1} < \dots < t_{k,m_k}$.
 Let $\set{E}' = \{ \vec{\nu}_1', \dots, \vec{\nu}_{N_E'}' \}$ be a new set of possible observations of detected episodes, including the pattern of positive and negative tests.
-Represent the observations such that for any $\vec{\nu}_k' \in \set{E}'$, $\vec{\nu}_k' = [\vec{\nu}_{k}, \vec{y}_k]^T$, where $\vec{y}_j \in \{ 0, 1 \}^{m_j}$ with $y_{j,l}$ being the test result at time $t_{j,l}$.
+Represent the observations such that for any $\vec{\nu}_k' \in \set{E}'$, $\vec{\nu}_k' = [\vec{\nu}_{k}, \vec{y}_k]^T$, where $\vec{y}_k \in \{ 0, 1 \}^{m_k}$ with $y_{k,l}$ being the test result at time $t_{k,l}$.
 
@@
 I will modify $p_{ik}$ to form $p_{ik}' = \prob(O'(j) = \vec{\nu}'_k \mid i(j) = i_k, \vec{\theta})$, taking into account false negatives.
-I consider a mixture of two scenarios, whether the final test in $\sched'_{i(j)}$ is a false negative, with the mixture probability determined by the test sensitivity.
+I consider a mixture of two scenarios, whether the final test in $\sched'_{k}$ is a false negative, with the mixture probability determined by the test sensitivity.
 
@@
 In this case, the test at $r_k^{(e)}+1$ is a true negative and the end of the episode is interval censored as in the previous chapter.
-% In this case, the test at $r_k^{(e)} + 1$ is a true negative, as are all other tests not in $\sched'_{i(j)}$.
+% In this case, the test at $r_k^{(e)} + 1$ is a true negative, as are all other tests not in $\sched'_{k}$.
 The true negative occurs with probability 1, by the assumption of no false positives.
 \begin{align}
-&\prob(\vec{\nu}_k, E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
-&= \prob(\vec{\nu}_k, l_k^{(e)} \leq E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\ % &\text{as no false positives}
-&= \prob(\vec{\nu}_k \mid l_k^{(e)} \leq E_j \leq r_k^{(e)}, B_j = b, p_\text{sens}, \vec{\theta}) \prob(l_k^{(e)} \leq E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
+&\prob(O'(j) = \vec{\nu}_k', E_j \leq r_k^{(e)} \mid B_j = b, i(j) = i_k, \psens, \vec{\theta}) \\
+&= \prob(O'(j) = \vec{\nu}_k', l_k^{(e)} \leq E_j \leq r_k^{(e)} \mid B_j = b, i(j) = i_k, \psens, \vec{\theta}) &\text{the test at $l_k^{(e)}$ is positive} \\
+&= \prob(O'(j) = \vec{\nu}_k' \mid l_k^{(e)} \leq E_j \leq r_k^{(e)}, B_j = b, i(j) = i_k, \psens, \vec{\theta}) \\
+&\ \ \  \times \prob(l_k^{(e)} \leq E_j \leq r_k^{(e)} \mid B_j = b, i(j) = i_k, \psens, \vec{\theta}) \\
 &= p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right)
@@
 \begin{align}
-&\prob(\vec{\nu}_k, E_j > r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
-&= \prob(\vec{\nu}_k \mid E_j > r_k^{(e)}, B_j = b, p_\text{sens}, \vec{\theta}) \prob(E_j > r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \\
-&= p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} (1 - p_\text{sens}) S_{\vec{\theta}}(r_k^{(e)} - b_j + 2)
+&\prob(O'(j) = \vec{\nu}_k', E_j > r_k^{(e)} \mid B_j = b, i(j) = i_k, \psens, \vec{\theta}) \\
+&= \prob(O'(j) = \vec{\nu}_k' \mid E_j > r_k^{(e)}, B_j = b, i(j) = i_k, \psens, \vec{\theta}) \\
+  &\ \ \  \times \prob(E_j > r_k^{(e)} \mid B_j = b, i(j) = i_k, \psens, \vec{\theta}) \\
+&= p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} (1 - p_\text{sens}) S_{\vec{\theta}}(r_k^{(e)} - b + 2)
 \label{imperf-test:eq:ll-ei-gt-ri}
@@
 
-These expressions can now be used to form the full replacement for $p_{ik}$, $p'_{ik} = \prob(O'(j ) = \vec{\nu}_k' \mid i(j) = i_k)$.
-First, augment the data with $b_j$, and split into the cases just discussed:
+These expressions can now be used to derive $p'_{ik}$.
+First, augment the data with $b$, and split into the cases just discussed, omitting the conditioning on $\psens$, $\vec{\theta}$, and $i(j) = i_k$:
 \begin{align}
 p_{ik}'
-=& \prob(\vec{\nu}_k \mid p_\text{sens}, \vec{\theta}) \\
-=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( \prob(\vec{\nu}_k, E_j \leq r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) + \prob(\vec{\nu}_k, E_j > r_k^{(e)} \mid B_j = b, p_\text{sens}, \vec{\theta}) \right) \\
-  &\times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}).
+=& \prob(O'(j) = \vec{\nu}_k') \\
+=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( \prob(O'(j) = \vec{\nu}_k', E_j \leq r_k^{(e)} \mid B_j = b) + \prob(O'(j) = \vec{\nu}_k, E_j > r_k^{(e)} \mid B_j = b) \right) \prob(B_j = b). \\
 \intertext{Now, substitute in \cref{imperf-test:eq:ll-ei-lt-ri,imperf-test:eq:ll-ei-gt-ri} and take out the common factor:}
@@
 =\ &  p_\text{sens}^{t_+} (1 - p_\text{sens})^{f_-} \\
-  & \times \sum_{b_j = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b_j + 1) - p_\text{sens} S_{\vec{\theta}}(r_k^{(e)} - b_j + 2) \right) \\
+  & \times \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - p_\text{sens} S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right) \\
   & \times \prob(B_j = b \mid p_\text{sens}, \vec{\theta}).
@@
 p_{ik}'
-&\propto \sum_{b = l_k^{(b)}}^{r_k^{(b)}} S_{\vec{\theta}}(l_k^{(e)} - b_j + 1) - p_\text{sens} S_{\vec{\theta}}(r_k^{(e)} - b + 2).
+&\propto \sum_{b = l_k^{(b)}}^{r_k^{(b)}} S_{\vec{\theta}}(l_k^{(e)} - b + 1) - p_\text{sens} S_{\vec{\theta}}(r_k^{(e)} - b + 2).
 \label{imperf-test:eq:pia-prime-constant}
@@
 A false negative occurring requires that the episode has not yet ended but a negative still occurs.
-The episode has not yet ended at the time of the test if $e_j = b_j + d_j - 1 \geq b_j + \tau_{\sched_{i(j)}}(b_j) + 1$, that is the duration of the infection $d_j \geq \tau_{\sched_{i(j)}}(b_j) + 1$.
+The episode has not yet ended at the time of the test if $e_j = b_j + d_j - 1 \geq b_j + \tau_{\sched_{i(j)}}(b_j)$, that is the duration of the infection $d_j \geq \tau_{\sched_{i(j)}}(b_j) + 1$.
 Conditional on the episode having not yet ended, the test result is negative with probability $1 - \psens$.
@@
 Denote the number of days between $b_j$ and the test following the false negative as $\tau^2_{\sched_{i(j)}}(b_j) \stackrel{\text{def}}{=} \tau_{\sched_{i(j)}}(\tau_{\sched_{i(j)}}(b_j) + 1)$.
-Then, the episode ends before this test if $d_j \leq \tau^2_{\sched_{i(j)}}(b_j) + 1$.
+Then, the episode ends before this test if $d_j \leq \tau^2_{\sched_{i(j)}}(b_j)$.
 
@@
 That is $\min(\sched_{i(j)}) < b_j \leq T_{i(j)}$ and $e_j \geq \tau_{\sched_{i(j)}}(b_j) + b_j$.
-    \item The episode ends in the interval $[\tau_{\sched_{i(j)}}(b_j) + b_j, \tau^2_{\sched_{i(j)}}(b_j) + b_j]$.
+    \item The episode ends in the interval $[\tau_{\sched_{i(j)}}(b_j) + b_j, \tau^2_{\sched_{i(j)}}(b_j) + b_j - 1]$.
       Note that the lower bound here is exactly the bound on $e_j$ in the previous condition.
-      Equivalently, $\tau_{\sched_{i(j)}}(b_j) + 1 \leq d_j \leq \tau^2_{\sched_{i(j)}}(b_j) + 1$.
+      Equivalently, $\tau_{\sched_{i(j)}}(b_j) + 1 \leq d_j \leq \tau^2_{\sched_{i(j)}}(b_j)$.
     \item A false negative occurs on day $\tau_{\sched_{i(j)}}(b_j) + b_j$. Conditional on the previous condition, this occurs with probability $1 - \psens$.
@@
 &\prob \left(
-    \tau_{\sched_{i(j)}}(b) + 1 \leq D_j \leq \tau^2_{\sched_{i(j)}}(b) + 1
+    \tau_{\sched_{i(j)}}(b) + 1 \leq D_j \leq \tau^2_{\sched_{i(j)}}(b)
     \mid B_j = b, \vec{\theta} \right) (1 - \psens) \\
-&= \left( S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b) + 2) \right) (1 - \psens).
+&= \left( S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b) + 1) \right) (1 - \psens).
 \end{align}
@@
 \begin{align}
-\zeta = (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_{i(j)}) + 1}^{T_{i(j)}} \left( S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b) + 2) \right).
+\zeta = (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_{i(j)}) + 1}^{T_{i(j)}} \left( S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b) + 1) \right).
 \label{imperf-test:eq:zeta}
@@
 =& \frac{1}{T} \sum_{b = \min \sched_{i} + 1}^{T_{i(j)}} S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1) \\
-&- (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_{i(j)}) + 1}^{T_{i(j)}} \left( S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b) + 2) \right) \\
+&- (1 - p_\text{sens})\frac{1}{T} \sum_{b=\min(\sched_{i(j)}) + 1}^{T_{i(j)}} \left( S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b) + 1) - S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b) + 1) \right) \\
 =& \frac{1}{T} \sum_{b=\min(\sched_{i(j)}) + 1}^{T_{i(j)}} \left( p_\text{sens} S_{\vec{\theta}}(\tau_{\sched_{i(j)}}(b) + 1) + (1 - p_\text{sens}) S_{\vec{\theta}}(\tau^2_{\sched_{i(j)}}(b) + 1)\right).

### 2024-04-01 cis-imperfect-testing.tex
proof 5.3 and 6.3

@@
 I start, in \cref{imperf-test:sec:notation}, by introducing some additional notation needed.
-I then modify the components of the likelihood in \cref{imperf-test:sec:modifying-p_ia} and \cref{imperf-test:sec:modifying-p_iu} to incorporate false negatives.
+I then modify the components of the likelihood in \cref{imperf-test:sec:modifying-p_ia,imperf-test:sec:modifying-p_iu} to incorporate false negatives.
 
@@
 \end{enumerate}
-The final two conditions are due the construction of the intervals as positive and negative tests bounding the beginning and end times of the episode.
+The final two conditions are due to the construction of the intervals as positive and negative tests bounding the beginning and end times of the episode.
 

### 2024-04-01 cis-perf-new-ll.tex
proof
proof 5.3 and 6.3

@@
 
-\chapter{Additional details of derivations in \cref{E-perf-test} and \cref{E-imperf-test}}
+\chapter{Defining possible observations} \label{perf-test:sec:conditions-nu-E}
 
-\section{Defining possible observations} \label{perf-test:sec:conditions-nu-E}
-
-Let $\vec{\nu} = [l^{(b)}, r^{(b)}, l^{(e)}, r^{e}, i]^T$ be an arbitrary vector of possible observations, $\vec{\nu} \in \set{E}$ if and only if all the following conditions hold.
+Let $\vec{\nu} = [l^{(b)}, r^{(b)}, l^{(e)}, r^{(e)}, i]^T$ be an arbitrary vector of possible observations, $\vec{\nu} \in \set{E}$ if and only if all the following conditions hold.
 \begin{enumerate}
-  \item $l^{(b)} 1, r^{(b)}, l^{(e)}, r^{e} + 1 \in \sched_{i}$, that is individual ${i}$ was tested at all these times.
+  \item $l^{(b)} 1, r^{(b)}, l^{(e)}, r^{(e)} + 1 \in \sched_{i}$, that is individual ${i}$ was tested at all these times.
   \item $l^{(b)} \leq r^{(b)} \leq l^{(e)} \leq r^{(e)}$.
@@
 
-\subsection{Imperfect testing}
-
 \end{document}

### 2024-04-01 cis-perfect-testing.tex
proof 5.3 and 6.3

@@
 
-If $j$ is detected, then $O(j) = [l_j^{(b)}, r_j^{(b)}, l_j^{(e)}, r_j^{e}, i(j)]^T$, where $i(j)$ is the individual in which the episode occurs; $l_j^{(b)}$ and $r_j^{(b)}$ are the earliest and latest time the episode could have begun respectively; and $l_j^{(e)}$ and $r_j^{(e)}$ are the earliest and latest time the episode could have ended respectively.
+If $j$ is detected, then $O(j) = [l_j^{(b)}, r_j^{(b)}, l_j^{(e)}, r_j^{(e)}, i(j)]^T$, where $i(j)$ is the individual in which the episode occurs; $l_j^{(b)}$ and $r_j^{(b)}$ are the earliest and latest time the episode could have begun respectively; and $l_j^{(e)}$ and $r_j^{(e)}$ are the earliest and latest time the episode could have ended respectively.
 
@@
 Define an integer $N_E$ and $\set{E} = \{ \vec{\nu}_1, \dots, \vec{\nu}_{N_E} \}$ as the set of all possible observations of detected episodes; that is, $O(j) \in \set{E}$ if and only if $j$ is a detected infection.
-Let $\vec{\nu}_k = [l^{(b)}_k, r^{(b)}_k, l^{(e)}_k, r^{e}_k, i_k]^T$ be an arbitrary vector of possible observations, visualised in \cref{perf-test:fig:partitionSpace}; the conditions for $\vec{\nu}_k \in \set{E}$ are in \cref{perf-test:sec:conditions-nu-E}.
+Let $\vec{\nu}_k = [l^{(b)}_k, r^{(b)}_k, l^{(e)}_k, r^{(e)}_k, i_k]^T$ be an arbitrary vector of possible observations, visualised in \cref{perf-test:fig:partitionSpace}; the conditions for $\vec{\nu}_k \in \set{E}$ are in \cref{perf-test:sec:conditions-nu-E}.
 
@@
   Each dot is a combination of $b_j$ and $e_j$ for an arbitrary individual $i$.
-  Each box, bounded by dashed lines, are combinations giving $O(j) = \nu_k$, $k$ corresponding to the numeric label; labels are arbitrary but unique across all individuals.
+  Each box, bounded by dashed lines, are combinations giving $O(j) = \vec{\nu}_k$, $k$ corresponding to the numeric label; labels are arbitrary but unique across all individuals.
   $i$ had negative tests at times 0, 7, 14, 56, and 84 (not shown) and positive tests at times 21 and 28.
@@
 &= p(\vec{\theta}) \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{k \in \set{D}} p_k &\text{by \cref{perf-test:eq:multinomial}} \\
-&= p(\vec{\theta}) \left( \prod_{k \in \set{D}} p_k \right) \left( \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \right)
+&= p(\vec{\theta}) \left( \prod_{k \in \set{D}} p_k \right) \left( \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \right).
 \label{perf-test:eq:posterior1}
@@
 
-Recall the definitions $p_k = \prob(O(j) = \vec{\nu}_k \mid \vec{\theta})$ where $O(j)$ are the observations of episode $j$ and $\vec{\nu}_k = [l^{(b)}_k, r^{(b)}_k, l^{(e)}_k, r^{e}_k, i_k]^T$.
+Recall the definitions $p_k = \prob(O(j) = \vec{\nu}_k \mid \vec{\theta})$ where $O(j)$ are the observations of episode $j$ and $\vec{\nu}_k = [l^{(b)}_k, r^{(b)}_k, l^{(e)}_k, r^{(e)}_k, i_k]^T$.
 
@@
 
-If $i(j) = i_k$ then the event $O(j) = \vec{\nu}_k$ occurs if and only if the episode starts in the interval $[l^{(b)}_k, r^{(b)}_k]$ and ends in the interval $[l^{(e)}_k, r^{e}_k]$.
+If $i(j) = i_k$ then the event $O(j) = \vec{\nu}_k$ occurs if and only if the episode starts in the interval $[l^{(b)}_k, r^{(b)}_k]$ and ends in the interval $[l^{(e)}_k, r^{(e)}_k]$.
 Omitting the conditioning on $\vec{\theta}$ and $i(j) = i$, this gives:
@@
 =& \prob \left( l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \mid l_k^{(b)} \leq B_{j} \leq r_k^{(b)} \right) \times\prob \left( l_k^{(b)} \leq B_{j} \leq r_k^{(b)} \right) \\
-=& \sum_{b = l_k^{(l)}}^{r_k^{(b)}} \prob \left( l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \mid B_{j} = b \right) \prob \left(B_{j} = b \right) \\
+=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \prob \left( l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \mid B_{j} = b \right) \prob \left(B_{j} = b \right) \\
 =& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \prob \left( l_k^{(e)} - b + 1 \leq D_{j} \leq r_k^{(e)} - b + 1 \right) \prob \left(B_{j} = b \right) &\text{by def of $D_{j}$} \\
@@
 assuming $P(i(j) = i \mid \vec{\theta}) = 1/\Ncis$ as before.
-Let $p_{iu} = \prob(O(j) = \emptyset \mid i(j) = i \mid \vec{\theta})$.
+Let $p_{iu} = \prob(O(j) = \emptyset \mid i(j) = i, \vec{\theta})$.
 Therefore, the crucial component is $1 - p_{iu}$.
@@
 &= \sum_{b = \min \sched_{i} + 1}^{T_{i(j)}} \prob(D_j \geq \tau_{\sched_{i}}(b) + 1 \mid B_j = b) \prob(B_j = b)\\
-&= \frac{1}{T} \sum_{b = \min \sched_{i} + 1}^{T_{i(j)}} S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1).
+&\propto \sum_{b = \min \sched_{i} + 1}^{T_{i(j)}} S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1).
 \label{perf-test:eq:piu}

### 2024-07-29 biology-data.tex
Rob p41 clarification

@@
 This exclusion was to ensure viral load trajectories are identifiable.
-I will revisit and extend their analysis in \cref{E-ATACCC}.
+The 57 contacts were divided into two groups.
+The first, consisting of 32 individuals, were unvaccinated and infected with a variety of varients.
+The second, consisting of the remaining 25 individuals, were vaccinated and all infected with the Delta variant.
+I will revisit and extend \textcite{hakkiOnset}'s analysis in \cref{E-ATACCC}, using the same 57 contacts and groups.
+The first group is more relevant to the period of interest in this thesis, and I focus on this group.
 

### 2024-07-29 incidence-prevalence.tex
Rob p52 comment

@@
 The duration distribution I will be concerned with is the distribution of the duration of detectability.
-Consider an individual $i$ infected on day $B_i$ that is prevalent during the interval $[B_i, E_i]$ (as in \cref{E-biology-data:sec:cis-episodes}).
+Consider an individual $i$ infected on day $B_i$ that is prevalent during the inclusive interval $[B_i, E_i]$ (as in \cref{E-biology-data:sec:cis-episodes}); if $B_i = E_i$ then $i$ is prevalent for exactly one day during which they are detectable.
 I define this period as an \emph{infection episode}, beginning on day $B_i$ and ending on day $E_i$; therefore, with a duration $D_i = E_i - B_i + 1$.

### 2024-07-29 ATACCC.tex
ref section in another chapter correctly

@@
 Specifically, excluding false positives from the model still explains the data well (see \cref{ATACCC:fig:goodness-of-fits}).
-Furthermore, a prior that false negatives and false positives are equally likely is not supported by other information (see \cref{biology-data:sec:measurement-error}).
+Furthermore, a prior that false negatives and false positives are equally likely is not supported by other information (see \cref{E-biology-data:sec:measurement-error}).
 Therefore, I simplified the model for test failures to only false negatives.

### 2024-07-29 ATACCC.tex
Rob p74 comment

@@
   \vec{\eta}_i &\dist \N(\vec{0}, \matr{C}_\phi) \\
-  \log \vec{\phi}_i &= \vec{\mu}_\phi + \vec{\sigma}_\phi \circ \vec{\eta}_i
+  \log \vec{\phi}_i &= \vec{\mu}_{\phi,g(i)} + \vec{\sigma}_\phi \circ \vec{\eta}_i
 \end{align}
@@
 From \cref{ATACCC:eq:di}, we can see that $d_i$ depends only on $i$'s random effects vector $\vec{\phi}_i$.
-Therefore, the distribution of $d_i$ across the population is determined by $\vec{\phi}_i$'s mean, $\vec{\mu}_\phi$, and covariance matrix, $\matr{\Sigma}_\phi$.
+Therefore, the distribution of $d_i$ across the population of group $g$ is determined by $\vec{\phi}_i$'s mean, $\vec{\mu}_{\phi,g}$, and covariance matrix, $\matr{\Sigma}_\phi$.
 % Hence, the uncertainty in the posterior distribution of the discretised duration cumulative distribution function for group $g$, $F_{D,g}(d) = \prob(D_i \leq d + 0.5)$ originate from the uncertainty in $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$.
 
-For a known $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$, $F_{D,g}$ is found as:
+For a known $\vec{\mu}_{\phi,g}$ and $\matr{\Sigma}_\phi$, $F_{D,g}$ is found as:
 \begin{align}
@@
   The posterior distribution is represented as a sample $F_{D,g}^{(1)}, \dots, F_{D,g}^{(N)}$.
-  $\vec{\mu}_\phi^{(i)}$ and $\matr{\Sigma}_\phi^{(i)}$ denote the $i$th posterior sample of $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$ respectively, taken from the MCMC output which produced $N$ posterior samples.
+  $\vec{\mu}_{\phi,g}^{(i)}$ and $\matr{\Sigma}_\phi^{(i)}$ denote the $i$th posterior sample of $\vec{\mu}_{\phi,g}$ and $\matr{\Sigma}_\phi$ respectively, taken from the MCMC output which produced $N$ posterior samples.
   $M$ is the number of individuals to draw, I found $M =\numprint{100000}$ produced stable estimates of the tail of $F_{D,g}(d)$ if $\dmax = 45$.
@@
 Applying \cref{ATACCC:alg:MonteCarlo} produces a sample of the posterior distribution of $F_{D,g}(d)$.
-One sample is produced for each iteration of the MCMC output used to estimate $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$.
+One sample is produced for each iteration of the MCMC output used to estimate $\vec{\mu}_{\phi,g}$ and $\matr{\Sigma}_\phi$.
 

### 2024-07-29 ATACCC.tex
Rob first p75 comment

@@
   \For{$j = 1, \dots, M$}{
-    Sample an individual-level random effects vector $\vec{\phi}_{i,j} \dist \LN(\vec{\mu}_{\phi,g}, \matr{\Sigma}_\phi)$ \;
+    Sample an individual-level random effects vector $\vec{\phi}_{i,j} \dist \LN(\vec{\mu}_{\phi,g}^{(i)}, \matr{\Sigma}^{(i)}_\phi)$ \;
     Calculate $d_{i,j}$ using \cref{ATACCC:eq:di} \;

### 2024-07-29 cis-perfect-testing.tex
Rob p84 comment
Rob first p87 comment
Rob third p87 comment

@@
     The shaded area represents when the individual is detectable.
+    Time is relative to the individual's first CIS test.
   }
@@
         Demonstrated here by a participant who is recorded negative at time 7 and positive at time 14, bounding the start of the episode; similarly, changing from positive at time 28 to negative at time 56 bounds the end of the episode.
+      Time is relative to the individual's first CIS test.
     }
@@
 I analyse the $\ndet = \numprint{4800}$ episodes with their first positive test between 10 Oct 2020 and 6 Dec 2020 inclusive, with negatives bounding the start and end time of the episode; I refer to these episodes as \emph{detected}; in \cref{perf-test:sec:prob-undetected} I formalise the conditions for an episode to be detected.
-Each day in this period is denoted as $t = 1, \dots, T$.
+Each day in this period is denoted as $t = 1, \dots, T$; \ie day $1$ is the 10 Oct 2020 and day $T$ is the 6 Dec 2020.
 I use these episodes because a negative test before the start of the episode provides a lower bound on the episode's start time; otherwise, there is little information on its length.
@@
 
-For any episode $j$, define $O(j)$ as the observations of that episode.
+For any episode $j$, define $O(j)$ as the observations of that episode; $j = 1, \dots, \ntot$ where $\ntot$ is the total number of infection episodes in the cohort.
 Test results are deterministic based on the time of the test and the individual's infection status.
@@
 Let $\nnodet$ denote the (latent) number of undetected episodes experienced across all $\Ncis$ individuals.
-Therefore, the total number of infection episodes in the cohort is $\ntot = \ndet + \nnodet$ and I index them with $j = 1, \dots, \ntot$.
+Therefore, the total number of infection episodes in the cohort is $\ntot = \ndet + \nnodet$.
 

### 2024-07-29 cis-perfect-testing.tex
Rob second p88 comment

@@
 
-Let $n_k$ denote the number of observations of $\vec{\nu}_k$; $\nnodet$ denote the latent number of undetected episodes; and $\vec{n} = [n_1, \dots, n_{N_D}, \nnodet]^T$.
-Hence, $\ntot = \sum_{i=1}^{N_D} n_i + \nnodet$.
+Let $n_k$ denote the number of observations of $\vec{\nu}_k$; $\nnodet$ denote the latent number of undetected episodes; and $\vec{n} = [n_1, \dots, n_{N_E}, \nnodet]^T$.
+Hence, $\ntot = \sum_{i=1}^{N_E} n_i + \nnodet$.
 
@@
 For undetected outcomes, let $p_u = \prob(O(j) = \emptyset \mid \vec{\theta})$.
-Together, $\vec{p} = [p_1, \dots, p_D, p_u]^T$.
+Together, $\vec{p} = [p_1, \dots, p_{N_E}, p_u]^T$.
 Assume, for tractability, that the events $O(j) = \vec{\nu}_k$ and $O(j') = \vec{\nu}_{k'}$ are independent for $j \neq j'$; this assumption is discussed in \cref{perf-test:sec:discussion}.
@@
 \intertext{that is:}
-  p(\vec{n} \mid \ntot, \vec{\theta}) &= \frac{\ntot!}{\nnodet!\prod_{k=1}^{N_D} n_k!} p_u^{\nnodet} \prod_{k=1}^{N_D} p_k^{n_k}.
+  p(\vec{n} \mid \ntot, \vec{\theta}) &= \frac{\ntot!}{\nnodet!\prod_{k=1}^{N_E} n_k!} p_u^{\nnodet} \prod_{k=1}^{N_E} p_k^{n_k}.
   \label{perf-test:eq:multinomial-ll}

### 2024-07-29 cis-perfect-testing.tex
Rob p91 comment

@@
 &= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) e^{\lambda \pnodet} d\lambda &\text{Maclaurin series of $e$} \\
-&= \int \lambda^{\ndet} e^{-\lambda(1 - \pnodet)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda &\lambda \dist \GamDist\\
-&= \int \frac{b^a}{\Gamma(a)} \lambda^{a+\ndet-1} e^{-(b+1-\pnodet)\lambda} \lambda d\lambda \\
+&= \int \lambda^{\ndet} e^{-\lambda(1 - \pnodet)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} d\lambda &\lambda \dist \GamDist\\
+&= \int \frac{b^a}{\Gamma(a)} \lambda^{a+\ndet-1} e^{-(b+1-\pnodet)\lambda} d\lambda \\
 &= \frac{b^a}{\Gamma(a)} \frac{\Gamma(a+\ndet)}{(b+1-\pnodet)^{a+\ndet}} &\text{Gamma pdf}\\

### 2024-07-29 cis-perfect-testing.tex
Rob p97 comment

@@
 
-I first approximate the \cref{E-ATACCC} posterior estimate of the hazard as $\logit{\vec{h}} \dist \MNorm(\vec{\mu}_A, \matr{\Sigma}_A)$ where $\vec{h}$ is the hazard, and $\vec{\mu}_A$ and $\matr{\Sigma}_A$ are the mean and covariance matrix computed from the samples of $\logit{\vec{h}}$ derived using the method of \cref{ATACCC:sec:duration}.
+I first approximate the \cref{E-ATACCC} posterior estimate of the hazard as $\logit{\vec{h}} \dist \MNorm(\vec{\mu}_A, \matr{\Sigma}_A)$ where $\vec{h} = [h_1, \dots, h_{\dmax}]^T$ is the hazard, and $\vec{\mu}_A$ and $\matr{\Sigma}_A$ are the mean and covariance matrix computed from the samples of $\logit{\vec{h}}$ derived using the method of \cref{E-ATACCC:sec:duration}.
+Therefore, $\vec{h}$ represents a sample from the approximation of \cref{E-ATACCC}'s posterior estimate of the hazard before increasing the uncertainty to account for the strong model assumptions.
 Using a multivariate normal, as opposed to multiple univariate distribution for each $h_t$, preserves the correlation between the hazards.

### 2024-07-29 transmission.tex
Rob p127 comment

@@
 
-\Cref{transmission:eq:forward-substitute} is derived from \cref{inc-prev:eq:EPt-matrix}: $\vec{P} = \matr{S} \vec{I}$.
-For this to have a unique solution, $\vec{I}$ and $\vec{P}$ need the same dimension.
+\Cref{transmission:eq:forward-substitute} is derived from \cref{inc-prev:eq:EPt-matrix}: $\vec{P} = \matr{S} \vec{Z}$.
+For this to have a unique solution, $\vec{Z}$ and $\vec{P}$ need the same dimension.
 However, I am using a period of prevalence which starts after the first infections.
@@
 I assume that the incidence for $t \leq 1$ is constant.
-Therefore, $P_1 = I_0 \sum_{i=1}^{\dmax} S(i)$ and $I_t = P_1 / \sum_{i=1}^{\dmax} S(i)$ for $t \leq 1$.
+Therefore, $P_1 = Z_0 \sum_{i=1}^{\dmax} S(i)$ and $Z_t = P_1 / \sum_{i=1}^{\dmax} S(i)$ for $t \leq 1$.
 

### 2024-07-29 transmission.tex
Rob p135 comment
Rob p147 comment

@@
 These individuals then transition into a set of compartments, $p_1, \dots, p_l$, each with their own transition rate.
+Individuals in compartments $p_1, \dots, p_l$ are detectable; exiting compartment $p_l$ corresponds to the end of the detectable period.
 I will choose $l$ and the transition rates such that combined length of stay (\ie time from entry in $p_1$ until exit from $p_l$) approximates the duration of detectability distribution estimated in \cref{E-imperf-test}.
@@
 The $\beta_w$ parameters absorb all the time-varying changes in transmission not captured by Google mobility and school attendance data.
-They trend downwards up until early December, then they start increasing (see \cref{SEIR:fig:beta-walk}).
+They trend downward and are then flat until they begin rising around the start of December (see \cref{SEIR:fig:beta-walk}).
 Correlation is clear across the regions even though each region is estimated independently, suggesting underlying biological or epidemiological changes across regions.

### 2024-07-29 MCMC-appendix.tex
Rob typo comments

@@
  \For{$i = 1, \dots, M$}{
-  sample a proposal $\psi' \sim q(\psi' \mid \psi)$ \;
+  sample a proposal $\psi' \sim q(\psi' \mid \psi^{i-1})$ \;
   calculate the acceptance probability $\alpha(\psi, \psi') = \min \left( 1, \frac{\pi(\psi') q(\psi \mid \psi')}{\pi(\psi) q(\psi' \mid \psi)} \right)$ \;
@@
 
-Various metrics quantify this, I make use of the improved Rhat statistic~\autocite{vehtariRhat}.
+Various metrics quantify this; I make use of the improved Rhat statistic~\autocite{vehtariRhat}.
 A Rhat of 1 means that the within-chain and between-chain variance are equal, informally that convergence is perfect, as far as can be measured by this technique.

### 2024-07-29 cis-imperfect-testing.tex
Rob p110 comment
Rob typo comments

@@
 In this chapter, I identify this issue to be the result of false negatives, a form of misclassification bias, within the CIS (\cref{imperf-test:sec:problem}).
-I then propose several models of false negatives (\cref{imperf-test:sec:simulate}) and extend the model of \cref{E-perf-test} to incorporate one of them (\cref{imperf-test:sec:modelling})
+I then propose several models of false negatives (\cref{imperf-test:sec:simulate}) and extend the model of \cref{E-perf-test} to incorporate one of them (\cref{imperf-test:sec:modelling}).
 After showing through simulation that the new model recovers the true duration distribution (\cref{imperf-test:sec:sim-study-results}), I apply it to the CIS data (\cref{imperf-test:sec:application}).
@@
 
-$\vec{\nu}'_k \in \set{E}'$ occurs if and only if all the following conditions are met.
+The elements of $\set{E}'$ are all $\vec{\nu}_k'$ such that:
 \begin{enumerate}

### 2024-07-29 cis-perfect-testing.tex
Rob p101 comment
Rob p102 comment
Rob typo comments

@@
   &= 1 - \sum_{i=1}^{\Ncis} \prob(O(j) = \emptyset, i(j) = i \mid \vec{\theta}) \\
-  &= 1 - \sum_{i=1}^{\Ncis} \prob(O(j) = \emptyset \mid i(j) = i \mid \vec{\theta}) P(i(j) = i \mid \vec{\theta}) \\
-  &= 1 - \frac{1}{\Ncis}\sum_{i=1}^{\Ncis} \prob(O(j) = \emptyset \mid i(j) = i \mid \vec{\theta}) \\
-  &= \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} (1 - \prob(O(j) = \emptyset \mid i(j) = i \mid \vec{\theta}))
+  &= 1 - \sum_{i=1}^{\Ncis} \prob(O(j) = \emptyset \mid i(j) = i, \vec{\theta}) P(i(j) = i \mid \vec{\theta}) \\
+  &= 1 - \frac{1}{\Ncis}\sum_{i=1}^{\Ncis} \prob(O(j) = \emptyset \mid i(j) = i, \vec{\theta}) \\
+  &= \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} (1 - \prob(O(j) = \emptyset \mid i(j) = i, \vec{\theta}))
   \label{perf-test:eq:pu}
@@
 
-\subsection{Setup}
+\subsection{Setup} \label{perf-test:sec:sim-setup}
 
@@
 This is needed because in step 2 the entire cohort was infected, while in the real study only an (unknown) portion is infected.
-    \item For this final set of episodes, calculate $(l_j^{(b)}, r_j^{b}, l_j^{(e)}, r_j^{(e)})$ by taking the day after the last negative prior to any positives, the first positive, the last positive, and the day before the negative following the last positive respectively.
+    \item For this final set of episodes, calculate $(l_j^{(b)}, r_j^{(b)}, l_j^{(e)}, r_j^{(e)})$ by taking the day after the last negative prior to any positives, the first positive, the last positive, and the day before the negative following the last positive respectively.
 \end{enumerate}
@@
 The prior on the total number of infections was centred on the true value with a dispersion parameter of 1.
+This meant that the prior was very diffuse.
 
@@
 Convergence was assessed using Rhat and ESS as described in \cref{E-MCMC:sec:convergence}, and all runs checked for divergent transitions.
-This meant that the prior was very diffuse.
 
@@
 I show only the first 100 days of the distribution for readability.
+Black line shows the value used in the simulation (see \cref{perf-test:sec:sim-setup}).
 \label{perf-test:fig:survival-results}}
@@
 Results using each of the three different priors discussed in \cref{perf-test:sec:parameters-priors} are shown.
+Black line shows the value used in the simulation (see \cref{perf-test:sec:sim-setup}).
 \label{perf-test:fig:hazard-results}}

### 2024-07-29 distributions.tex
Rob typo comments

@@
 Lognormal & $\LN(\mu, \sigma^2)$ & $\mu$ (mean of the logarithm)\newline $\sigma^2 > 0$ (variance of the logarithm) & $e^{\mu + \frac{\sigma^2}{2}}$ & $(e^{\sigma^2} - 1)e^{2\mu + \sigma^2}$ & $\frac{1}{x\sigma\sqrt{2\pi}}e^{-\frac{(\ln(x)-\mu)^2}{2\sigma^2}}$ \\
-Gamma & $\GamDist(\alpha, \beta)$ & $\alpha > 0$ (shape) \newline $\beta > 0$ (rate) & $\frac{\alpha}{\beta}$ & $\frac{\alpha}{\beta^2}$ & $\frac{\beta^\alpha}{\Gamma(\alpha)\theta}x^{\alpha-1}e^{-\beta x}$ \\
+Gamma & $\GamDist(\alpha, \beta)$ & $\alpha > 0$ (shape) \newline $\beta > 0$ (rate) & $\frac{\alpha}{\beta}$ & $\frac{\alpha}{\beta^2}$ & $\frac{\beta^\alpha}{\Gamma(\alpha)}x^{\alpha-1}e^{-\beta x}$ \\
 Beta & $\BetaDist(\alpha, \beta)$ & $\alpha > 0$ (shape)\newline $\beta > 0$ (shape) & $\frac{\alpha}{\alpha+\beta}$ & $\frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}$ & $\frac{1}{B(\alpha,\beta)}x^{\alpha-1}(1-x)^{\beta-1}$ \\

### 2024-07-29 incidence-prevalence.tex
Rob typo comments

@@
 The distribution of $D_i$ can be expressed through the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
-Alternatively, the \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$.
+Alternatively, the \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(d - 1)$ with $S(1) = 1$.
 %(a common alternative convention for survival analysis with discrete variables is to define the survival function as $\prob(D >d)$).
@@
 Assuming $\vec{P}$ and $\vec{Z}$ have the same dimension then $\matr{S}$ is lower-triangular with strictly positive diagonal elements; therefore, $\matr{S}$ is invertible, and the system has a unique solution~\autocite[424]{bernsteinFacts}.
-The lower-triangular structure of $\matr{S}$ makes allows solving the system through forward-substitution, an efficient method that avoids explicit matrix inversion~\autocite{cormenMatrix}.
+The lower-triangular structure of $\matr{S}$ allows solving the system through forward-substitution, an efficient method that avoids explicit matrix inversion~\autocite{cormenMatrix}.
 Forward-substitution sequentially solves the $t$th element of $\vec{Z}$ ($t = 1, \dots, T$) by calculating:

### 2024-07-29 introduction.tex
Rob p24 comment
Rob p25 comment
Rob typo comments

@@
 Incidence-type data originate from observations of the consequences of an infection, \eg the development of symptoms or admission to hospital.
-This type of data typically suffers from reporting biases, most problematically that the proportion of infections detected varying over time or within different subgroups of the population~\autocites[chapter 9]{lashModern}{shadboltChallenges}.
+This type of data typically suffers from reporting biases, most problematically that the proportion of infections detected varies over time or within different subgroups of the population~\autocites[chapter 9]{lashModern}{shadboltChallenges}.
 
@@
 
-Data on the severe consequences of an infection typically less affected by biases.
+Data on the severe consequences of an infection are typically less affected by biases.
 An extreme example is the development of AIDS following infection by HIV.
@@
 \begin{align}
-    \E(Z_t \mid Z_0, \dots, Z_{t-1}) &= \sum_{\tau=1}^t \beta(t,\tau) Z_{t-\tau} d \tau
+    \E(Z_t \mid Z_0, \dots, Z_{t-1}) &= \sum_{\tau=1}^t \beta(t,\tau) Z_{t-\tau}
     \label{intro:eq:renewal}
@@
 Solving \cref{intro:eq:disease-model} for $Z$ is an ill-posed problem.
-The forwards error of the solution is large when $\max_{i > 0} f(i) / f(0)$~\autocite[chapter 8.2]{highamAccuracy}.
-A large forwards error means that the computed value is far from the true solution~\autocite[6--9]{highamAccuracy}; this is avoided by either using an approach which gives a large $f(0)$ or smoothing $Z_t$.
+The forwards error of the solution is large when $\max_{i > 0} f(i) / f(0)$~\autocite[chapter 8.2]{highamAccuracy} is large.
+A large forwards error means that the computed value is far from the true solution~\autocite[6--9]{highamAccuracy}; this is avoided by either using an approach which gives a large $f(0)$ or by smoothing $Z_t$.
 
@@
 
-\Textcite{abbottCISincidence} applied the backcalculation to CIS data to estimate the incidence of SARS-CoV-2 in the UK.
+\Textcite{abbottCISincidence} applied backcalculation to CIS data to estimate the incidence of SARS-CoV-2 in the UK.
 They used summary statistics of prevalence estimates from the CIS, taken from the ONS website, as data; specifically, the prevalence's posterior point estimate and bounds of the 95\% credible interval.

### 2024-07-29 transmission.tex
Rob typo comments

@@
 A slow increase in incidence is estimated to occur over the autumn 2020 period, peaking in October in most regions (see \cref{transmission:fig:backcalc-regions}).
-An exception is the North East and East of England regions (see \cref{transmission:fig:backcalc-regions}(C) and (D) respectively); these regions show no consistent increase or decrease in incidence during October.
-All regions then decrease during November; partially coincidng with the national lockdown in force in England between 5 Nov and 2 Dec 2020~\autocite{ifgTimeline}.
+An exception is the North East and East of England regions (see \cref{transmission:fig:backcalc-regions}(B)); these regions show no consistent increase or decrease in incidence during October.
+All regions then decrease during November; partially coinciding with the national lockdown in force in England between 5 Nov and 2 Dec 2020~\autocite{ifgTimeline}.
 
@@
 The higher incidence in this age group prior to November means it could be an immunity effect.
-Alternatively, it could be a consequence of schools remaining open in the November lockdown; 20--24 year olds rarely interact with school-aged children (see \cref{E-SEIR:fig:age-contacts})
+Alternatively, it could be a consequence of schools remaining open in the November lockdown; 20--24 year olds rarely interact with school-aged children (see \cref{E-SEIR:fig:age-contacts}).
 Further work is required to determine causality.
@@
 Google mobility data are based on the location of Android users over time, relative to a pre-pandemic baseline.
-The mobility and school data are averaged over a week down or up-weight the number of contacts in each setting, before combining contacts across settings to form $\kappa_{aa'}(t)$.
+The mobility and school data are averaged over a week, used to down or up-weight the number of contacts in each setting, and then combined across settings to form $\kappa_{aa'}(t)$.
 $\kappa_{aa'}(t)$ is assumed to be piecewise constant, changing only at the start of each week.

### 2024-07-29 cis-perfect-testing.tex
Chrstil comment on vague v weakly informative prior
Christl comment on Beta(0.5, 0.5) being uninformative

@@
 An uninformative prior would be of the form $\lambda_t \dist \BetaDist(\alpha, \alpha)$.
-Commonly, $\alpha$ is chosen to be $0.5$ (Jeffreys' Prior) or $1$ (a uniform prior).
+Commonly, $\alpha$ is chosen to be $0.5$ or $1$ (a uniform prior).
+The former, $\alpha = 0.5$, is Jefferys’ prior~\autocite{jeffreysInvariant}, commonly considered an uninformative prior because it is invariant to parameterization~\autocite[53]{gelmanBDA}.
 Even though these priors are uninformative on the hazard, they become highly informative on the survival. 
@@
 
-Instead, I propose a weakly informative (or vague) prior Beta(0.1, 1.9), which has mean 0.05 and minimal information.
+Instead, I propose the vague prior Beta(0.1, 1.9), which has mean 0.05 and minimal information.
 The amount of information in a Beta distribution is related to the sum of its parameters.

### 2024-07-29 biology-data.tex
Rob p46 comment
Christl comment on semicolons

@@
 \Textcite{hakkiOnset} analysed the 57 contacts who tested positive for SARS-CoV-2 and had the growth phase in their viral load measured.
-They excluded the remaining 115 contacts with at least one positive RT-PCR test, ATACCC did not capture the phase of the infection with increasing viral load.
+They excluded the remaining 115 contacts with at least one positive RT-PCR test where ATACCC did not capture the phase of the infection with increasing viral load.
 This exclusion was to ensure viral load trajectories are identifiable.
@@
 $\rho=0$ means the beta-binomial coincides with the binomial distribution.
-As $\rho \to \infty$, the distribution tends towards the uniform distribution on $[0, 1]$~\autocite{hughesUsing}.
+As $\rho \to \infty$, the beta component of the beta-binomial distribution tends towards the uniform distribution on $[0, 1]$~\autocite{hughesUsing}.
 

### 2024-07-29 introduction.tex
Christl comment on semicolons

@@
 All pandemics start as an \emph{outbreak}, a sudden and unexpected rise in the number of people with a disease in a local area; if this spreads over a larger geographic area (\eg a country) it becomes an \emph{epidemic}; if it spreads globally then it is a \emph{pandemic}~\autocite{grennanPandemic}.
-The precise meaning of these terms varies, they are defined implicitly by their usage rather than formal definitions~\autocite{morensPandemic,doshiElusive}.
+The precise meaning of these terms varies; they are defined implicitly by their usage rather than formal definitions~\autocite{morensPandemic,doshiElusive}.
 I will refer to the SARS-CoV-2 epidemic within England (or the UK), and the global impact as the pandemic.

### 2024-07-29 biology-data.tex
"data" is plural

@@
 ATACCC tested individual frequently but had only 57 infections followed for a maximum of 20 days.
-The frequent testing makes estimation for these first 20 days reliable, but longer periods are rare and the data does not exist; therefore, the tails of the distribution cannot be estimated.
+The frequent testing makes estimation for these first 20 days reliable, but longer periods are rare and the data do not exist; therefore, the tails of the distribution cannot be estimated.
 The CIS had more infrequent testing but a large sample size (thousands of infections) and up to years of follow-up.

### 2024-07-29 cis-perfect-testing.tex
Christl's first comment on simulations
"data" is plural

@@
 
-In \cref{perf-test:sec:problem}, I describe the challenges analysing the data presents.
+In \cref{perf-test:sec:problem}, I describe the challenges presented by analysing the data.
 This motivates using a survival analysis framework which I introduce in \cref{perf-test:sec:surv-analysis}, before deriving the statistical model for the CIS setting in \cref{perf-test:sec:model}.
@@
 These curves and the combined curve are compared in \cref{perf-test:fig:duration-dist}.
+Compared to the ATACCC estimate alone, incorporating the CIS portion shifts some probability mass into the tail of the distribution.
+Therefore, the survival probability is initially higher but later lower than ATACCC alone.
 Individual $i$'s duration of detectability, $D_i$, is then an independent draw from this distribution.

### 2024-07-29 introduction.tex
"data" is plural

@@
 
-An example of incidence-type data are the number of cases identified through community testing, which relies on individuals presenting for a test.
+An example of incidence-type data is the number of cases identified through community testing, which relies on individuals presenting for a test.
 Individuals' ability and willingness to be tested varies over time, which needs to be accounted for in order to perform unbiased inference.
@@
 
-However, severe consequences can be rare in some population subgroups (\eg young, otherwise healthy individuals infected with SARS-CoV-2~\autocite{wardSero,bhopalChildren}), meaning they rarely appear in this data.
-Therefore, the data does not provide much information on incidence in those subgroups.
+However, severe consequences can be rare in some population subgroups (\eg young, otherwise healthy individuals infected with SARS-CoV-2~\autocite{wardSero,bhopalChildren}), meaning they rarely appear in these data.
+Therefore, the data do not provide much information on incidence in those subgroups.
 Furthermore, the delay between infection and the severe event occurring means that recent infections cannot be estimated with much precision~\autocite{swallow2022challenges}.
@@
 Where possible, aggregated, non-disclosive versions of the datasets are available alongside the code.
-The original ATACCC data used is available alongside the code.
+The original ATACCC data used are available alongside the code.
 The original CIS data can only be accessed by accredited researchers within the ONS's SRS (Secure Research Service); see \url{https://www.ons.gov.uk/aboutus/whatwedo/statistics/requestingstatistics/secureresearchservice} for details.

### 2024-07-29 transmission.tex
"data" is plural

@@
 In this set-up, the parameters $i^+$, $q_c$, and $\psi$ are fully identified; $\pi_0$ and $\rho$ are not identified at all; the other parameters are partially identified, their posterior has moved away from the prior but there is insufficient information in the data to fully overcome the prior (see \cref{SEIR:fig:true-vs-posterior}).
-The ability to identify $q_c$ is an advantage of using data which includes measurements on children, unlike data on hospitalisations or deaths.
+The ability to identify $q_c$ is an advantage of using data which include measurements on children, unlike data on hospitalisations or deaths.
 
@@
 Such a model would stratify positive tests in the CIS by pre-Alpha or Alpha variant.
-Then, each variant would be explicitly included in the model to fit to this data.
+Then, each variant would be explicitly included in the model to fit to these data.
 

### 2024-07-29 ATACCC.tex
Rob second p75 comment
Rob typo comments
Christl acronym comment

@@
 
-Several features of the \textcite{hakkiOnset} model can be motivated by the desire to apply Stan, a modern and convenient implementation of the NUTS algorithm (see \cref{E-MCMC}).
+Several features of the \textcite{hakkiOnset} model can be motivated by the desire to apply Stan, a modern and convenient implementation of the NUTS (the No U-Turn Sampler) algorithm (see \cref{E-MCMC}).
 The first of these is using a continuous viral load trajectory rather than the more common piecewise linear model.
@@
 \begin{align}
-v_i(t) = v_{\max,i} \frac{a_i+b_i}{b_ie^{-a_i(t-\tau_{\max,i})} + a_ie^{b_i(t-\tau_{max,i})}}. \label{ATACCC:eq:viral-load}
+v_i(t) = v_{\max,i} \frac{a_i+b_i}{b_ie^{-a_i(t-\tau_{\max,i})} + a_ie^{b_i(t-\tau_{\max,i})}}. \label{ATACCC:eq:viral-load}
 \end{align}
@@
 Applying \cref{ATACCC:alg:MonteCarlo} produces a sample of the posterior distribution of $F_{D,g}(d)$.
-One sample is produced for each iteration of the MCMC output used to estimate $\vec{\mu}_{\phi,g}$ and $\matr{\Sigma}_\phi$.
+One sample is produced for each iteration of the MCMC (Markov Chain Monte Carlo) output used to estimate $\vec{\mu}_{\phi,g}$ and $\matr{\Sigma}_\phi$.
 
@@
 
-The original \textcite{hakkiOnset} model has convergence issues (Rhat > 1.07 and ESS < 100) for the parameters of the test failure distribution, $\sigma_0$ and $\rho$.
+The original \textcite{hakkiOnset} model has convergence issues (Rhat > 1.07 and ESS (Effective Sample Size) < 100) for the parameters of the test failure distribution, $\sigma_0$ and $\rho$.
 The simplified model I proposed in \cref{ATACCC:sec:observation-modification} solved this issue, meaning that this analysis converged.
@@
 As the estimates are similar and the modified model converges better, I will use only the results from the modified model.
+In this study, $\vlod = 0$.
 \begin{figure}

### 2024-07-29 biology-data.tex
Christl acronym comment

@@
 
-SARS-CoV-2 is closely related to previous coronaviruses that have caused outbreaks in humans, including SARS-CoV and MERS-CoV.
+SARS-CoV-2 is closely related to previous coronaviruses that have caused outbreaks in humans, including SARS-CoV (Severe Acute Respiratory Syndrome--related CoronaVirus) and MERS-CoV (Middle East Respiratory Syndrome--related CoronaVirus).
 Coronaviruses are RNA (Ribonucleic acid) viruses, which means they use RNA, rather than DNA (Deoxyribonucleic acid), as their genetic material.
@@
     Dots show the unadjusted positivity rate (days with fewer than 100 tests are excluded for readability).
-    Ribbon shows modelled 95\% CrI for the positivity, corrected for non-response bias and smoothed over time (using the methodology described in \cref{E-backcalc:sec:methods}).
+    Ribbon shows modelled 95\% CrI (Credible Interval) for the positivity, corrected for non-response bias and smoothed over time (using the methodology described in \cref{E-backcalc:sec:methods}).
   }
@@
 
-For privacy reasons, the CIS data are stored in the ONS's SRS (Secure Research Service)~\autocite{onsSRS}.
+For privacy reasons, the CIS data are stored in the ONS's SRS~\autocite{onsSRS}.
 The SRS is a TRE (Trusted Research Environment) that allows researchers to access sensitive data in a secure environment.

### 2024-07-29 incidence-prevalence.tex
Christl acronym comment

@@
 The duration is a positive integer, \ie $D_i \in \{1, 2, \dots, \dmax \}$, where $\dmax$ is the maximum duration.
-The distribution of $D_i$ can be expressed through the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
+The distribution of $D_i$ can be expressed through the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cdf (cumulative density function), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
 Alternatively, the \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(d - 1)$ with $S(1) = 1$.
@@
 
-The interactions between strata are included by replacing the scalar $\beta$ with a WAIFW (\emph{Who Acquires Infection From Whom}) matrix, $\matr{\beta}$~\autocite[58]{keelingModeling}.
+The interactions between strata are included by replacing the scalar $\beta$ with a WAIFW (Who Acquires Infection From Whom) matrix, $\matr{\beta}$~\autocite[58]{keelingModeling}.
 The force of infection is now a vector, defined as $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i}_j(t)$ where $\vec{i}_j$ is a vector such that $i_{j,a}$ is the proportion of stratum $a$ in the $j$th infectious state.
@@
 Reproduction numbers in models with structured populations are more complex than the homogeneous case.
-The WAIFM allows calculation of the number of secondary infections generated by a single infection in the $a$th strata,  this follows directly from its definition as $\frac{1}{\gamma} \sum_{a' \in \set{A}} \beta_{a'a}$.
+The WAIFW matrix allows calculation of the number of secondary infections generated by a single infection in the $a$th strata,  this follows directly from its definition as $\frac{1}{\gamma} \sum_{a' \in \set{A}} \beta_{a'a}$.
 However, how to average these contributions to calculate the basic reproduction number is not obvious.

### 2024-07-29 introduction.tex
Christl acronym comment

@@
 
-Infection with the SARS-CoV-2 virus is estimated to have killed approximately 15 million in the most acute phase (to the end of 2021) and 30 million in total~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
+Infection with the SARS-CoV-2 (Severe Acute Respiratory Syndrome CoronaVirus 2) virus is estimated to have killed approximately 15 million in the most acute phase (to the end of 2021) and 30 million in total~\autocite{whoCOVIDExcess,economistCOVIDExcess}.
 Beyond the lives lost, there were massive health, economic, and societal costs.
@@
 Data on the severe consequences of an infection are typically less affected by biases.
-An extreme example is the development of AIDS following infection by HIV.
+An extreme example is the development of AIDS (Acquired ImmunoDeficiency Syndrome) following infection by HIV (Human Immunodeficiency Virus).
 Here, the disease is so severe that it is extremely unlikely for an individual to develop the disease without seeking healthcare~\autocite{evansCompleteness}.
@@
 where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that result in the consequence of interest; and, conditional on being observed, $f(i)$ describes the (discretised) probability that an infection is included in the count $W_t$ at $i$ time units after the infection.
-$f$ is an example of a \emph{delay function}, and $f$ is commonly the pdf of a distribution, or a transformation of it.
+$f$ is an example of a \emph{delay function}, and $f$ is commonly the pdf (probability density function) of a distribution, or a transformation of it.
 
@@
 These were designed to avoid the biases of community testing and observe the general population.
-These surveys use RT-PCR (reverse transcription polymerase chain reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
+These surveys use RT-PCR (Real-Time Polymerase Chain Reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
 The first survey was the CIS (Coronavirus (COVID-19) Infection Survey), run by the ONS (Office for National Statistics)~\autocite{CIS,cisMethodsONS}.
@@
 Many longitudinal studies were conducted throughout the pandemic.
-They aimed to estimate a wide variety of epidemiological and biological quantities of the SARS-CoV2 virus.
+They aimed to estimate a wide variety of epidemiological and biological quantities of the SARS-CoV-2 virus.
 These include the CIS, previously mentioned.
@@
 Another study, which I use in this thesis, is the ATACCC (Assessment of Transmission and Contagiousness of COVID-19 in Contacts) study~\autocite{singanayagamCommunity,hakkiOnset}.
-ATACCC recruited individuals from contacts of known infections reported to NHS Test and Trace, and collected data on them for 20 days.
+ATACCC recruited individuals from contacts of known infections reported to NHS (the National Health Service) Test and Trace, and collected data on them for 20 days.
 This included daily RT-PCR tests, which allowed quantification of their viral load (the amount of virus in an individual), a biomarker.

### 2024-07-29 introduction.tex
Christl typo

@@
 The main data sources informing the response to the SARS-Cov-2 epidemic in England were based on: community testing, severe events, seroprevalence surveys, and prevalence surveys.
-Each of these data sources have strengths and weaknesses, see \textcite{royalSocietyRnumber} and references therein for further details on each of these data sources.
+Each of these data sources has strengths and weaknesses, see \textcite{royalSocietyRnumber} and references therein for further details on each of these data sources.
 

### 2024-07-31 biology-data.tex
Christl p44 minor comment

@@
 The age groups used have varied slightly for different CIS analyses~\autocite[e.g.][]{pouwelsMRPvaccination,pouwelsCommunity,cisMethodsONS,houseInferring,walkerTracking}.
-I will use the age groups: 0--11 (pre-school and primary school), 12--16 (compulsory secondary school), 17--25 (adolescents and young adults who may remain in formal education), 26--50 (younger working age adults), 51--70 (older working age adults), and 71+ (elderly).
+I will use the age groups: 0--11 years old (pre-school and primary school), 12--16 years old (compulsory secondary school), 17--25 years old (adolescents and young adults who may remain in formal education), 26--50 years old (younger working age adults), 51--70 years old (older working age adults), and over 71 years old (elderly).
 

### 2024-07-31 cis-perfect-testing.tex
Rob second p87 comment

@@
 
+For our purposes, an infection episode $j$ consists of the three random variables: $i(j)$, the individual it occurs in; $B_j$, the day it begins (the first day $i(j)$ is detectable due to this infection episode); $E_j$, the day it ends (the last day $i(j)$ is detectable due to this infection episode); the distribution of the episode's duration is a deterministic transformation of these variables, $D_j = E_j - B_j + 1$.
+I assume that $B_j$, $D_j$, and $i(j)$ are independent.
+$D_j$'s distribution is determined by $\vec{\theta}$; both $B_j$ and $i(j)$ have uniform priors over their respective state spaces.
+$O(j)$ is a deterministic function of $B_j$, $E_j$, and $\sched_{i(j)}$.
+
+To make this problem tractable, I model infection episodes as independent conditional on $\ntot$, the total number of them that occur, and $\vec{\theta}$; \ie $(B_j, E_j, i(j)) \ind (B_{j'}, E_{j'}, i(j')) \mid \ntot, \vec{\theta}$ for $j \neq j'$.
+This is true both for infection episodes within the same individual and across individuals; this assumption is discussed in \cref{perf-test:sec:discussion}.
+
 Let $n_k$ denote the number of observations of $\vec{\nu}_k$; $\nnodet$ denote the latent number of undetected episodes; and $\vec{n} = [n_1, \dots, n_{N_E}, \nnodet]^T$.
@@
 Together, $\vec{p} = [p_1, \dots, p_{N_E}, p_u]^T$.
-Assume, for tractability, that the events $O(j) = \vec{\nu}_k$ and $O(j') = \vec{\nu}_{k'}$ are independent for $j \neq j'$; this assumption is discussed in \cref{perf-test:sec:discussion}.
 
-Then:
+As we assume $O(j) = \vec{\nu}_k$ and $O(j') = \vec{\nu}_{k'}$ are independent for $j \neq j'$:
 \begin{align}
@@
 % This is valid as $\prob(O(j) = \nu_k \mid i(j) \neq i_k) = 0$ due to the condition here being equivalent to equating the two vectors' final elements.
-Assume that each infection episode occurs independently and with equal probability in any individual, \ie $\prob(i(j) = i_k) = 1/\Ncis$ for all $j$ and $k$.
+As I assume that each infection episode occurs independently and with equal probability in any individual, $\prob(i(j) = i_k) = 1/\Ncis$ for all $j$ and $k$.
 
@@
 \end{align}
-under the assumption of uniform probability of infection time.
+by the assumption of uniform probability of infection time.
 This is the standard form of the likelihood for doubly interval censored data without truncation~\autocite[e.g.][]{sunEmpirical}.
@@
 
-The derivation in \cref{perf-test:sec:model} relied on an assumption that $O(j)$ and $O(j')$ occur independently for $j \neq j'$.
+The derivation in \cref{perf-test:sec:model} relied on an assumption that infection episodes occur independently.
 This is not the case, as immunity means that an individual previously infected is less likely to be infected in the future.

### 2024-07-31 cis-perfect-testing.tex
add a dgp section to ch 5

@@
 In \cref{perf-test:sec:data}, I introduce the data, defining which quantities are observed and which are latent.
+In \cref{perf-test:sec:dgp}, I introduce the data generating process, defining the model for how infection episodes arise and how these are observed.
 In the remainder of the section, I derive the posterior density, adapting the general approach of \textcite{heiseyModelling}.
@@
 These potential classification can then be compared to the actual observations.
+
+The target of inference is $\vec{\theta}$, the parameters of the survival function $S_{\vec{\theta}}(t)$.
+The contents of $\vec{\theta}$ are discussed in \cref{perf-test:sec:parameters-priors}.
 
@@
 
-I analyse the $\ndet = \numprint{4800}$ episodes with their first positive test between 10 Oct 2020 and 6 Dec 2020 inclusive, with negatives bounding the start and end time of the episode; I refer to these episodes as \emph{detected}; in \cref{perf-test:sec:prob-undetected} I formalise the conditions for an episode to be detected.
+I analyse the $\ndet = \numprint{4800}$ episodes with their first positive test between 10 Oct 2020 and 6 Dec 2020 inclusive, with negatives bounding the start and end time of the episode; I refer to these episodes as \emph{detected}.
 Each day in this period is denoted as $t = 1, \dots, T$; \ie day $1$ is the 10 Oct 2020 and day $T$ is the 6 Dec 2020.
@@
 
-For any episode $j$, define $O(j)$ as the observations of that episode; $j = 1, \dots, \ntot$ where $\ntot$ is the total number of infection episodes in the cohort.
-Test results are deterministic based on the time of the test and the individual's infection status.
-Therefore, $O(j)$ is fully determined by the triplet $(b_j, e_j, i(j))$ consisting of $j$'s beginning time, end time, and the individual in which the episode occurs respectively.
-The episode's duration is $d_j = e_j - b_j + 1$ (see \cref{E-inc-prev:sec:definitions}).
+\subsection{Data generating process} \label{perf-test:sec:dgp}
 
-If $j$ is detected, then $O(j) = [l_j^{(b)}, r_j^{(b)}, l_j^{(e)}, r_j^{(e)}, i(j)]^T$, where $i(j)$ is the individual in which the episode occurs; $l_j^{(b)}$ and $r_j^{(b)}$ are the earliest and latest time the episode could have begun respectively; and $l_j^{(e)}$ and $r_j^{(e)}$ are the earliest and latest time the episode could have ended respectively.
+For our purposes, the $j$th infection episode ($j = 1, \dots, \ntot$) is $W_j = [B_j, E_j, i(j)]^T$ where $\ntot$ is the total number of infection episodes in the cohort; $i(j)$ is the individual the infection episode occurs in; $B_j$ is the day the episode begins (the first day $i(j)$ is detectable due to this infection episode); and $E_j$ is the day the episode ends (the last day $i(j)$ is detectable due to this infection episode).
+The episode's duration is a deterministic transformation of these variables, $D_j = E_j - B_j + 1$; and I assume that $B_j$, $D_j$, and $i(j)$ are independent.
+$D_j$'s distribution is determined by $\vec{\theta}$; both $B_j$ and $i(j)$ have uniform distributions over their respective state spaces.
 
-If $j$ is undetected, there are no observations; in which case, define $O(j) = \emptyset$.
-Let $\nnodet$ denote the (latent) number of undetected episodes experienced across all $\Ncis$ individuals.
-Therefore, the total number of infection episodes in the cohort is $\ntot = \ndet + \nnodet$.
+To make this problem tractable, I model infection episodes as independent conditional on $\ntot$, the total number of them that occur, and $\vec{\theta}$; \ie $W_j \ind W_{j'} \mid \ntot, \vec{\theta}$ for $j \neq j'$.
+This is true both for infection episodes within the same individual and across individuals; this assumption is discussed in \cref{perf-test:sec:discussion}.
 
-% Any infection episode, whether detected or not, is fully characterised by the triplet $(b_j, e_j, i(j))$ where $b_j$ and $e_j$ are the times the episode begin and end respectively; and $i(j)$ is the individual the episode occurred in.
-% $i(j)$ identifies the test schedule.
-% For all $j$, the exact values $b_j$ and $e_j$ are latent.
-% If $j$ is undetected then $i(j)$ is also latent and little is known about $b_j$ or $e_j$.
-% If $j$ is detected, then $i(j)$ is known; $b_j \in [l_j^{(b)}, r_j^{(b)}]$; and $e_j \in [l_j^{(e)}, r_j^{(e)}]$.
-% These latter two properties are what make the data doubly interval censored.
+The test results for any individual is fully determined by any infection episodes that occur in that individual and their test schedule.
+The set of times $i(j)$ tests positive at due to $W_j$ is $\posResults(W_j) = \{ t \in \sched_{i(j)} \ssep B_j \leq t \leq E_j \}$.
+The negative test immediately before the start of the episode (if there is one) is $\negResults^{(l)}(W_j) = \max \{ t \in \sched_{i(j)} \ssep t < B_j \}$.
+The negative test immediately after the end of the episode (if there is one) is $\negResults^{(r)}(W_j) = \min \{ t \in \sched_{i(j)} \ssep t > E_j \}$.
+
+Next, I define $O(j)$ to contain the test results due to $W_j$ that give information on $W_j$'s length, and hence are relevant to inference about $\vec{\theta}$.
+If $\posResults(W_j) = \emptyset$, $\min \posResults(W_j) \notin [1, T]$, or either $\negResults^{(l)}$ or $\negResults^{(r)}$ is undefined then the episode is undetected, in which case $O(j) = \emptyset$.
+Otherwise, $O(j) = [l_j^{(b)}, r_j^{(b)}, l_j^{(e)}, r_j^{(e)}, i(j)]^T$, where $i(j)$ is the individual in which the episode occurs; $l_j^{(b)} = \negResults^{(l)}(W_j) + 1$ and $r_j^{(b)} = \min \posResults(W_j)$ are the earliest and latest time the episode could have begun and produced the observed series of test results respectively; and $l_j^{(e)} = \max \posResults(W_j)$ and $r_j^{(e)} = \negResults^{(r)}(W_j) - 1$ are the earliest and latest time the episode could have ended respectively.
+These definitions coincide with the way the dataset is constructed (see \cref{E-biology-data:sec:cis-episodes}).
 
@@
 
-Define an integer $N_E$ and $\set{E} = \{ \vec{\nu}_1, \dots, \vec{\nu}_{N_E} \}$ as the set of all possible observations of detected episodes; that is, $O(j) \in \set{E}$ if and only if $j$ is a detected infection.
-Let $\vec{\nu}_k = [l^{(b)}_k, r^{(b)}_k, l^{(e)}_k, r^{(e)}_k, i_k]^T$ be an arbitrary vector of possible observations, visualised in \cref{perf-test:fig:partitionSpace}; the conditions for $\vec{\nu}_k \in \set{E}$ are in \cref{perf-test:sec:conditions-nu-E}.
-
-The target of inference is $\vec{\theta}$, the parameters of the survival function $S_{\vec{\theta}}(t)$.
-The contents of $\vec{\theta}$ are discussed in \cref{perf-test:sec:parameters-priors}.
-
-For our purposes, an infection episode $j$ consists of the three random variables: $i(j)$, the individual it occurs in; $B_j$, the day it begins (the first day $i(j)$ is detectable due to this infection episode); $E_j$, the day it ends (the last day $i(j)$ is detectable due to this infection episode); the distribution of the episode's duration is a deterministic transformation of these variables, $D_j = E_j - B_j + 1$.
-I assume that $B_j$, $D_j$, and $i(j)$ are independent.
-$D_j$'s distribution is determined by $\vec{\theta}$; both $B_j$ and $i(j)$ have uniform priors over their respective state spaces.
-$O(j)$ is a deterministic function of $B_j$, $E_j$, and $\sched_{i(j)}$.
-
-To make this problem tractable, I model infection episodes as independent conditional on $\ntot$, the total number of them that occur, and $\vec{\theta}$; \ie $(B_j, E_j, i(j)) \ind (B_{j'}, E_{j'}, i(j')) \mid \ntot, \vec{\theta}$ for $j \neq j'$.
-This is true both for infection episodes within the same individual and across individuals; this assumption is discussed in \cref{perf-test:sec:discussion}.
-
-Let $n_k$ denote the number of observations of $\vec{\nu}_k$; $\nnodet$ denote the latent number of undetected episodes; and $\vec{n} = [n_1, \dots, n_{N_E}, \nnodet]^T$.
-Hence, $\ntot = \sum_{i=1}^{N_E} n_i + \nnodet$.
+Define an integer $N_E$ and $\set{E} = \{ \vec{\nu}_1, \dots, \vec{\nu}_{N_E} \}$ as the set of all possible values $O(j)$ can take, excluding $\emptyset$, conditional on the test schedules but no other data; that is, $O(j) \in \set{E}$ if and only if $j$ is a detected infection; a full list of the formal conditions defining the elements of $\set{E}$ are in \cref{perf-test:sec:conditions-nu-E}.
+Let $\vec{\nu}_k = [l^{(b)}_k, r^{(b)}_k, l^{(e)}_k, r^{(e)}_k, i_k]^T$ be an arbitrary member of $\set{E}$, visualised in \cref{perf-test:fig:partitionSpace}.
+Let $n_k$ denote the number of times that $\vec{\nu}_k$ appears in the episodes dataset (\ie the observed data); $\nnodet$ denote the latent number of undetected episodes; and $\vec{n} = [n_1, \dots, n_{N_E}, \nnodet]^T$.
+Hence, $\ntot = \nnodet + \ndet = \sum_{i=1}^{N_E} n_i + \nnodet$.
 
@@
 
-Recall the definitions $p_k = \prob(O(j) = \vec{\nu}_k \mid \vec{\theta})$ where $O(j)$ are the observations of episode $j$ and $\vec{\nu}_k = [l^{(b)}_k, r^{(b)}_k, l^{(e)}_k, r^{(e)}_k, i_k]^T$.
+Recall the definitions $p_k = \prob(O(j) = \vec{\nu}_k \mid \vec{\theta})$ where $O(j)$ is the random variable containing all relevant information for episode $j$ and $\vec{\nu}_k = [l^{(b)}_k, r^{(b)}_k, l^{(e)}_k, r^{(e)}_k, i_k]^T$ is an arbitrary element from $O(j)$'s state space, excluding undetected infections.
 

### 2024-07-31 cis-imperfect-testing.tex
"data" is plural
Rob first p109 comment (on assumptions)

@@
 
-The CIS data provides evidence for the test sensitivity declining over the course of an episode.
+The CIS data provide evidence for the test sensitivity declining over the course of an episode.
 \Cref{imperf-test:fig:bounding-cis-sensitivity} demonstrates this by using the following classification for test results.
@@
 
-In this section, I introduce a simple model of false negatives (\ie allowing $\psens < 1$) into the posterior density derived in \cref{E-perf-test:sec:model}.
+In this section, I introduce a simple model of false negatives (\ie allowing $\psens < 1$) into the posterior density derived in \cref{E-perf-test:sec:model}; I continue to assume the probability of a false positive is negligible.
 Using a simple model, in particular a constant test sensitivity, means calculating the likelihood remains tractable.
+Additionally, I limit the number of permutations of false negative and true positive tests that I consider have non-negligible probability.
+Specifically, I assume that, if $i(j)$ is tested during infection episode $j$, either: the individual has a single false negative test and the episode ends before their next test, or the first test is a false positive test.
+This assumption is reasonable because false negatives normally occur when the viral load is low, which is normally late in the infection episode.
+For the same reason, I assume there is at most one false negative test following the final true positive test in the episode.
 
@@
 
-For tractability, assume that the negative test bounding the start of the episode, on day $l_k^{(b)}-1$, is a true negative.
-This assumption is reasonable because, since a positive test follows at $r_k^{(b)}$, the negative at $l_k^{(b)}-1$ is likely early in the infection and the test sensitivity is high early in an infection.
+By assumption, the negative test bounding the start of the episode, on day $l_k^{(b)}-1$, is a true negative.
 True negatives occur with probability 1, and hence this test does not contribute to the likelihood.
@@
 For there to be no subsequent positive tests, all tests up until day $e_j$ are false negatives.
-I assume there is a negligible probability of two false negatives.
+By assumption, there is a negligible probability of two false negatives.
 Therefore, this can only occur if the episode ends before another test.

### 2024-07-31 cis-imperfect-testing.tex
Rob second p109 comment (on definitions)

@@
 
-Introducing false negatives means that $O(j)$, the observation vector of episode $j$, is now random, even if $B_j$ and $E_j$ are known.
+Introducing false negatives means that $O(j)$, containing the relevant information for episode $j$, is now random, even if $B_j$ and $E_j$ are known.
 Therefore, the results between the tests that bound episode $j$'s beginning and end times should now enter into the likelihood.
@@
 
-To formalise this, I define analogous variables to those in \cref{E-perf-test:sec:posterior}.
-To start with, define the relevant testing times for any episode $j$ with $O(j) = \vec{\nu}_k$ as $\sched'_k = \{ t \in \sched_{i_k} \ssep r_k^{(b)} \leq t \leq r_k^{(e)} + 1 \}$ and let $m_k$ denote the size of this set.
-Denote the elements of $\sched'_k$ by $t_{k,1} < \dots < t_{k,m_k}$.
-Let $\set{E}' = \{ \vec{\nu}_1', \dots, \vec{\nu}_{N_E'}' \}$ be a new set of possible observations of detected episodes, including the pattern of positive and negative tests.
-Represent the observations such that for any $\vec{\nu}_k' \in \set{E}'$, $\vec{\nu}_k' = [\vec{\nu}_{k}, \vec{y}_k]^T$, where $\vec{y}_k \in \{ 0, 1 \}^{m_k}$ with $y_{k,l}$ being the test result at time $t_{k,l}$.
-
-The elements of $\set{E}'$ are all $\vec{\nu}_k'$ such that:
+To do so, define a random vector $O'(j)$ with state space $\{\emptyset\} \cup \set{E}'$ which will replace $O(j)$ from \cref{E-perf-test}.
+As before, $O'(j) = \emptyset$ if episode $j$ is undetected.
+Otherwise, $O'(j) \in \set{E}'$ where $\set{E}' = \{ \vec{\nu}_1', \dots, \vec{\nu}_{N_E'}' \}$ augments the space $\set{E}$ with test results during the episode.
+The elements of $\set{E}'$ are all $\vec{\nu}_k' = [\vec{\nu}_k, \vec{y}_k]^T$ where $\vec{\nu}_k \in \set{E}$ and $\vec{y}_k$ is a vector of test results for the relevant testing times, $\sched'_k = \{ t \in \sched_{i_k} \ssep r_k^{(b)} \leq t \leq r_k^{(e)} + 1 \}$.
+To formalise the definition of $\vec{y}_k$, let $m_k$ denote the size of $\sched'_k$ and denote the elements of $\sched'_k$ by $t_{k,1} < \dots < t_{k,m_k}$.
+Then, $\vec{y}_k \in \{0, 1\}^{m_k}$ and satisfies the following two conditions.
 \begin{enumerate}
-  \item $\vec{\nu}_k \in \set{E}$.
   \item The elements of $\vec{y}_k$ corresponding to the tests at times $r_k^{(b)}$ and $l_k^{(e)}$ are positive, \ie $y_{k,1} = y_{k,m_k-1} = 1$.
@@
 \end{enumerate}
-The final two conditions are due to the construction of the intervals as positive and negative tests bounding the beginning and end times of the episode.
+These conditions are due to the construction of the intervals as positive and negative tests bounding the beginning and end times of the episode.
 
-Let $O'(j)$ be the observations generated by episode $j$ including the new vector of results.
-As before, $O'(j) \in \set{E}'$ indicates that episode $j$ was detected; otherwise, $O'(j) = \emptyset$.
-
-Similarly, let $p_k'$, $p_{ik}'$, $p_u'$, and $p_{iu}'$ have the same definition as $p_k$, $p_{ik}$, $p_u$, and $p_{iu}$ respectively except replacing $O$ with $O'$ and $\vec{\nu}_k$ with $\vec{\nu}_k'$, and conditioning on $\psens$.
+Let $p_k'$, $p_{ik}'$, $p_u'$, and $p_{iu}'$ have the same definition as $p_k$, $p_{ik}$, $p_u$, and $p_{iu}$ respectively except replacing $O$ with $O'$ and $\vec{\nu}_k$ with $\vec{\nu}_k'$, and conditioning on $\psens$.
 As before, $p'_k = 1/\Ncis p'_{ik}$ and $p'_u = 1/\Ncis \sum_{i=1}^{\Ncis} p'_{iu}$.

### 2024-07-31 cis-perfect-testing.tex
improve figure appearance

@@
 \begin{figure}
-  \centering \includegraphics{cis-perfect-testing/flat-prior}
+  \makebox[\textwidth][c]{\includegraphics{cis-perfect-testing/flat-prior}}
   \caption[Uninformative priors for the hazard]{
@@
 \begin{figure}
-  \centering \includegraphics{cis-perfect-testing/vague-prior}
+  \makebox[\textwidth][c]{\includegraphics{cis-perfect-testing/vague-prior}}
   \caption[Vague prior for the hazard]{
@@
 \begin{figure}
-  \centering \includegraphics{cis-perfect-testing/rw2-prior}
+  \makebox[\textwidth][c]{\includegraphics{cis-perfect-testing/rw2-prior}}
   \caption[RW2 prior for the hazard]{The prior predictive survival time (mean and 95\% central interval) for the RW2 prior specified in the main text.
@@
   % \vspace{-1cm}
-  \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{cis-perfect-testing/ataccc-approximation-hazard}}
+  \makebox[\textwidth][c]{\includegraphics{cis-perfect-testing/ataccc-approximation-hazard}}
   \caption[Approximating the ATACCC posterior hazard as a logit-normal]{Comparison of the posterior estimate of the hazard from \cref{E-ATACCC} and its approximation with a logit-normal distribution (kernel density estimate smoothed) for hazards between day 4 and 15.
@@
 \begin{figure}
-  \centering \includegraphics{cis-perfect-testing/ataccc-approximation-survival}
+  \makebox[\textwidth][c]{\includegraphics{cis-perfect-testing/ataccc-approximation-survival}}
   \caption[Approximating the ATACCC posterior survival]{Comparison of the posterior estimate of the survival from \cref{E-ATACCC} and its approximation with a logit-normal distribution (median and central 95\% interval).
@@
 \begin{figure}
-  \centering \includegraphics{cis-perfect-testing/kt-prior}
+  \makebox[\textwidth][c]{\includegraphics{cis-perfect-testing/kt-prior}}
   \caption{Function used for $k_t$. \label{perf-test:fig:kt}}
@@
 \begin{figure}
-  \centering \includegraphics{cis-perfect-testing/input-duration-dists}
+  \makebox[\textwidth][c]{\includegraphics{cis-perfect-testing/input-duration-dists}}
   \caption[Comparison of duration distributions]{Comparison of the duration distribution used for the simulation.
@@
 \begin{figure}
-  \centering \includegraphics{cis-perfect-testing/survival-results}
+  \makebox[\textwidth][c]{\includegraphics{cis-perfect-testing/survival-results}}
   \caption[Comparison of survival function estimates under different priors]{Comparison of the posterior estimate of the survival distribution (pointwise median and 95\% CrI) when using different priors on the hazard.
@@
 \begin{figure}
-  \centering \includegraphics{cis-perfect-testing/hazard-results}
+  \makebox[\textwidth][c]{\includegraphics{cis-perfect-testing/hazard-results}}
   \caption[Comparison of hazard estimates under different priors]{Comparison of the posterior estimate of the hazard (pointwise median and 95\% CrI) when using different priors on the hazard.

### 2024-07-31 ATACCC/duration.R

@@ -35,13 +35,10 @@ p_durations = bind_rows(
     ) +
     theme_survival_time_series () +
     coord_cartesian(xlim = c(0, 40))
-ggsave(
+save_plot(
     filename = here::here("ATACCC/duration.pdf"),
     plot = p_durations,
-    width = 15,
-    height = 9,
-    units = "cm",
-    dpi = 300
+    height = 9
 )
 
 bind_rows(

### 2024-07-31 ATACCC/goodness_fit.R

@@ -72,13 +72,10 @@ p_55 = posterior_predictive |>
   ) +
   theme_trajectory()
   
-ggsave(
+save_plot(
   filename = "ATACCC/fit_individual_55.pdf",
   plot = p_55,
-  width = 15,
-  height = 9,
-  units = "cm",
-  dpi = 300
+  height = 9
 )
 
 produce_gof_plot = function(individuals) {
@@ -106,20 +103,14 @@ all_individuals = unique(posterior_summary$i)
 selected_individuals = all_individuals[all_individuals > 8 & all_individuals %% 2 == 1]
 appendix_individuals = all_individuals[!all_individuals %in% selected_individuals]
 
-ggsave(
+save_plot(
     filename = here::here("ATACCC/fits.pdf"),
     plot = produce_gof_plot(selected_individuals),
-    height = 15,
-    width = 15,
-    units = "cm",
-    dpi = 300
+    caption_lines = 6
 )
 
-ggsave(
+save_plot(
     filename = here::here("ATACCC/appendix_fits.pdf"),
     plot = produce_gof_plot(appendix_individuals),
-    height = 15,
-    width = 15,
-    units = "cm",
-    dpi = 300
+    caption_lines = 6
 )

### 2024-07-31 ATACCC/posteriors.R

@@ -92,14 +92,12 @@ p_mean = bind_rows(
 p_combined = p_mean / p_cov + plot_layout(
     guides = "collect",
     heights = c(1, 2)
-)
-ggsave(
+) &
+theme(legend.position = "bottom")
+save_plot(
     filename = here::here("ATACCC/compare_hakki_modified.pdf"),
     plot = p_combined,
-    width = 15,
-    height = 15,
-    units = "cm",
-    dpi = 300
+    height = 17
 )
 
 p_mean_trajectories = modified |>
@@ -125,14 +123,12 @@ p_mean_trajectories = modified |>
         x = "Time from peak",
         y = "Log viral load",
     ) +
-    coord_cartesian(ylim = c(0, NA))
-ggsave(
+    coord_cartesian(ylim = c(0, NA)) +
+    theme(legend.position = "bottom")
+save_plot(
     filename = here::here("ATACCC/mean_trajectories.pdf"),
     plot = p_mean_trajectories,
-    width = 15,
-    height = 10,
-    units = "cm",
-    dpi = 300
+    height = 12
 )
 # tibble_diagnostics = function(sims) {
 #   tibble(

### 2024-07-31 ATACCC/trajectories.R

@@ -37,11 +37,8 @@ p_typical_trajectory = ggplot(trajectories, aes(x, y, colour = label)) +
     colour = "Parameter values"
   ) +
   standard_plot_theming()
-ggsave(
+save_plot(
   filename = "ATACCC/typical_trajectory.pdf",
   plot = p_typical_trajectory,
-  width = 15,
-  height = 9,
-  units = "cm",
-  dpi = 300
+  height = 9
 )

### 2024-07-31 SEIR/CIS/attack_rates.R

@@ -64,10 +64,8 @@ p_attack_rate = ggplot(attack_rate, aes(x = age, y = occupancy, colour = age)) +
     labs(x = "", y = "Attack rate", colour = "Age") +
     scale_x_discrete(breaks = c()) +
     coord_cartesian(y = c(0, NA))
-ggsave(
+save_plot(
     filename = here::here("SEIR/CIS/attack_rates.pdf"),
     plot = p_attack_rate,
-    width = 5,
-    height = 5,
-    device = cairo_pdf
+    height = 15
 )

### 2024-07-31 SEIR/CIS/posterior_predictive.R

@@ -93,7 +93,7 @@ create_prev_plot = function(regions, label) {
         facet_grid(age ~ region) +
         standard_plot_theming() +
         labs(
-            x = "Date",
+            x = "Date (2020-1)",
             y = "Prevalence"
         ) +
         theme(
@@ -102,16 +102,12 @@ create_prev_plot = function(regions, label) {
             axis.text.x = element_text(angle = 45, hjust = 1)
         )
 
-    ggsave(
-        device = cairo_pdf,
+    save_plot(
+        plot = plot,
         filename = here::here(
             glue::glue("SEIR/CIS/prev_{label}.pdf")
         ),
-        plot = plot,
-        width = 15,
-        height = 16,
-        units = "cm",
-        dpi = 300
+        caption_lines = 6
     )
 }
 
@@ -141,21 +137,15 @@ p_incidence = prediction_intervals |>
     ) +
     facet_wrap(~region, ncol = 2) +
     labs(
-        x = "Date",
+        x = "Date (2020-1)",
         y = "Incidence"
     ) +
     standard_plot_theming() +
     theme(legend.position = "bottom")
-ggsave(
-    device = cairo_pdf,
-    filename = here::here(
-        glue::glue("SEIR/CIS/incidence.pdf")
-    ),
+save_plot(
+    filename = here::here("SEIR/CIS/incidence.pdf"),
     plot = p_incidence,
-    width = 15,
-    height = 20,
-    units = "cm",
-    dpi = 300
+    caption_lines = 1
 )
 
 # Calculate when incidence peaks for each posterior sample (by age and region)
@@ -176,18 +166,14 @@ p_peak = peak_days |>
     standard_plot_theming() +
     theme(legend.position = "bottom") +
     labs(
-        x = "Day",
+        x = "Date (2020-1)",
         y = "Posterior probability",
         colour = "Age group"
     )
-ggsave(
-    device = cairo_pdf,
+save_plot(
     filename = here::here("SEIR/CIS/p_peak.pdf"),
     plot = p_peak,
-    width = 15,
-    height = 10,
-    units = "cm",
-    dpi = 300
+    height = 13
 )
     
 # Output peak incidence dates as a CSV for humans

### 2024-07-31 SEIR/CIS/random_walk.R

@@ -92,11 +92,8 @@ p_walk = walk |>
         x = "Date",
         y = expression(beta[t])
     )
-ggsave(
+save_plot(
     plot = p_walk,
     filename = here::here("SEIR/CIS/beta_walk.pdf"),
-    width = 15,
-    height = 20,
-    units = "cm",
-    dpi = 300
+    caption_lines = 3
 )

### 2024-07-31 SEIR/sim/data.R

@@ -16,8 +16,7 @@ do_plot = function(base_gg, y_axis_label) {
         geom_line(aes(date, group = sim), alpha = 0.2) +
         standard_plot_theming() +
         facet_wrap(~age) +
-        labs(x = "Date", y = y_axis_label) +
-        theme(text = element_text(size = 20))
+        labs(x = "Date (2020-1)", y = y_axis_label)
     )
 }
 
@@ -31,11 +30,8 @@ p_prevalence = inputs |>
     do_plot("Prevalence") +
     ggtitle("(B) Prevalence")
 
-ggsave(
+save_plot(
     filename = here::here("SEIR/sim/data.pdf"),
     plot = p_incidence / p_prevalence,
-    width = 20,
-    height = 25,
-    dpi = 300,
-    device = cairo_pdf
+    caption_lines = 3
 )

### 2024-07-31 SEIR/sim/predictive_check.R

@@ -19,7 +19,7 @@ make_plot = function(x) {
         standard_plot_theming() +
         facet_wrap(~age) +
         labs(
-            x = "Day",
+            x = "Day (2020-1)",
             y = "Coverage",
             fill = "Interval width",
             colour = "Interval width"
@@ -72,10 +72,8 @@ p_combined = p_incidence / p_prevalence +
     theme(
         legend.position = "bottom",
     )
-ggsave(
+save_plot(
     filename = "SEIR/sim/predictive_coverage.pdf",
     plot = p_combined,
-    width = 15,
-    height = 15,
-    device = cairo_pdf
+    caption_lines = 3
 )

### 2024-07-31 biology-data/CIS-recruitment.R

@@ -42,8 +42,8 @@ p_total = swabs |>
     ) +
     ggtitle("(B) Total swabs")
 
-ggsave(
+save_plot(
     filename = here::here("biology-data", "CIS-recruitment.pdf"),
     plot = p_recruit / p_total,
-    width = 6, height = 6
+    height = 16
 )

### 2024-07-31 cis-imperfect-testing/CIS_survival.R

@@ -69,13 +69,10 @@ p_perfect = posterior_draws |>
         colour = ""
     ) +
     coord_cartesian(xlim = c(0, 45))
-ggsave(
+save_plot(
     filename = here::here("cis-imperfect-testing/CIS_perfect.pdf"),
     plot = p_perfect,
-    width = 15,
-    height = 9,
-    units = "cm",
-    dpi = 300
+    height = 9
 )
 
 #############################################################################################
@@ -100,13 +97,10 @@ p_final = posterior_draws |>
         fill = "",
         colour = ""
     )
-ggsave(
+save_plot(
     filename = here::here("cis-imperfect-testing/CIS_final.pdf"),
     plot = p_final,
-    width = 15,
     height = 9,
-    units = "cm",
-    dpi = 300
 )
 
 #############################################################################################
@@ -174,7 +168,7 @@ sens_curves = sens_draws |>
     theme_survival_time_series() +
     labs(colour = expression(p[sens]), fill = expression(p[sens]))
 
-ggsave(
+save_plot(
     filename = here::here("cis-imperfect-testing/CIS_vary.pdf"),
     plot = (r_medians + r_day_50 + r_curves) / (sens_medians + sens_day_50 + sens_curves) +
         plot_annotation(tag_levels = "A") &
@@ -182,10 +176,7 @@ ggsave(
             text = element_text(size = 8),
             legend.key.size = unit(0.3, "cm")
         ),
-    width = 18,
-    height = 20,
-    units = "cm",
-    dpi = 300
+    caption_lines = 5
 )
 
 posterior_draws |>

### 2024-07-31 cis-imperfect-testing/test_sens_bound.R

@@ -37,13 +37,11 @@ p_bound = bound_data |>
         x = "Days since first positive test",
         y = "Test sensitivity",
         colour = NULL
-    )
+    ) +
+    theme(legend.position = "bottom")
 
-ggsave(
+save_plot(
     filename = here::here("cis-imperfect-testing/test-sens-bound.pdf"),
     plot = p_bound,
-    width = 15,
-    height = 9,
-    units = "cm",
-    dpi = 300
+    height = 10
 )

### 2024-07-31 cis-perfect-testing/input-duration-dists.R

@@ -24,11 +24,8 @@ plot = ataccc_posterior |>
     colour = ""
   )
 
-ggsave(
+save_plot(
   filename = "cis-perfect-testing/input-duration-dists.pdf",
   plot = plot,
-  width = 15,
-  height = 9,
-  units = "cm",
-  dpi = 300
+  height = 9
 )

### 2024-07-31 cis-perfect-testing/regions_diag.R

@@ -1,5 +1,6 @@
 library(dplyr)
 library(ggplot2)
+source(here::here("utils.R"))
 
 neg_test_dates = c(0, 7, 14, 56)
 pos_test_dates = c(28, 21)
@@ -34,6 +35,7 @@ tbl_labels = inner_join(
 
 shade_alpha = 0.3
 plot = ggplot() +
+    standard_plot_theming() +
     test_lines(pos_test_dates, TRUE) +
     test_lines(neg_test_dates, FALSE) +
     geom_hline(aes(yintercept = all_test_dates - 0.5), linetype = "dashed") +
@@ -44,7 +46,6 @@ plot = ggplot() +
         aes(x, y, label = label),
         size = 10
     ) +
-    theme_minimal() +
     scale_x_continuous(breaks = all_test_dates, minor_breaks = -100:100, limits = c(-10, 60), expand = c(0, 0)) +
     scale_y_continuous(breaks = all_test_dates, minor_breaks = -100:100, limits = c(0, 59.5), expand = c(0, 0)) +
     geom_point(
@@ -147,4 +148,8 @@ plot = ggplot() +
         legend.position = "bottom",
         legend.box = "vertical"
     )
-ggsave("cis-perfect-testing/regions_diag.pdf", width = 19, height = 20, device = cairo_pdf, units = "cm")
+save_plot(
+    filename = "cis-perfect-testing/regions_diag.pdf",
+    plot = plot,
+    caption_lines = 8
+)

### 2024-07-31 transmission/backcalc-regions.R

@@ -27,7 +27,7 @@ incidence_plot = function(df) {
         stat_lineribbon(alpha = 0.3, .width = 0.95, linewidth = 0.5) +
         scale_y_continuous(labels = scales::label_percent()) +
         labs(
-            x = "Date",
+            x = "Date (2020-1)",
             y = "Incidence proportion",
         ) +
         standard_plot_theming() +
@@ -51,11 +51,10 @@ p_region = (
 ) +
     plot_layout(heights = c(1, 1.6))
 
-ggsave(
+save_plot(
     filename = here::here("transmission", "backcalc-regions.pdf"),
     plot = p_region,
-    width = 6.5,
-    height = 7
+    caption_lines = 1
 )
 
 tbl_min = tbl_regions |>
@@ -94,14 +93,13 @@ p_alpha = tbl_min |>
     standard_plot_theming() +
     geom_smooth(method = "lm", formula = y ~ x) +
     labs(
-        x = "Date Alpha rise",
-        y = "Date of minimum incidence",
+        x = "Date (2020) of Alpha rise",
+        y = "Date (2020) of minimum incidence",
         colour = "Region"
     ) +
     theme(legend.position = "bottom")
-ggsave(
+save_plot(
     filename = here::here("transmission", "backcalc-alpha.pdf"),
     plot = p_alpha,
-    width = 6,
-    height = 5.5
+    height = 10
 )

### 2024-07-31 transmission/backcalc-start-effect.R

@@ -12,14 +12,14 @@ p_start_effect = readRDS(file.path(output_dir, "region.rds")) |>
     scale_y_continuous(labels = scales::label_percent(), limits = c(0, NA)) +
     scale_fill_brewer() +
     labs(
-        x = "Date",
+        x = "Date (2020)",
         y = "Incidence proportion",
     ) +
     standard_plot_theming() +
     theme(legend.position = "none")
-ggsave(
+save_plot(
     filename = here::here("transmission", "backcalc-start-effect.pdf"),
     plot = p_start_effect,
-    width = 4,
-    height = 3.8
+    width = 10,
+    height = 10
 )

### 2024-07-31 transmission/compare_NE.R

@@ -42,13 +42,11 @@ p_compare = bind_rows(
     theme(legend.position = "bottom") +
     scale_y_continuous(labels = scales::label_percent()) +
     labs(
-        x = "Date",
+        x = "Date (2020-1)",
         y = "Incidence proportion"
     )
-ggsave(
+save_plot(
     filename = here::here("transmission", "compare-NE.pdf"),
-    device = cairo_pdf,
     plot = p_compare,
-    width = 6,
-    height = 4
+    height = 12
 )

### 2024-07-31 transmission/compare_regions.R

@@ -43,12 +43,11 @@ p_compare = bind_rows(
     theme(legend.position = "bottom") +
     scale_y_continuous(labels = scales::label_percent()) +
     labs(
-        x = "Date",
+        x = "Date (2020-1)",
         y = "Incidence proportion"
     )
-ggsave(
+save_plot(
     filename = here::here("transmission", "compare-regions.pdf"),
     plot = p_compare,
-    width = 6,
-    height = 5.5
+    height = 18
 )

### 2024-07-31 utils.R

@@ -1,6 +1,7 @@
 standard_plot_theming = function() {
     rlang::list2(
         theme_minimal(),
+        ggplot2::theme(text = ggplot2::element_text(size = 12)),
     )
 }
 
@@ -21,3 +22,21 @@ theme_survival_time_series = function() {
 
 logit = function(x) log(x) - log(1 - x)
 expit = function(x) 1 / (1 + exp(-x))
+
+# save a plot, defaults to full page
+save_plot = function(
+    filename, plot, caption_lines = 0,
+    width = 18, height = 25.7 - 0.6*caption_lines, units = "cm", dpi = 500,
+    device = cairo_pdf, ...
+) {
+    ggplot2::ggsave(
+        filename = filename,
+        plot = plot,
+        width = width,
+        height = height,
+        units = units,
+        dpi = dpi,
+        device = device,
+        ...
+    )
+}

### 2024-08-01 introduction.tex
give an overview of the English pandemic

@@
 \section{SARS-CoV-2 in England}
+
+On 31 Dec 2019, China informed the WHO (World Health Organization) of cases of pneumonia of unknown cause in Wuhan City, Hubei province~\autocite{whoNovelPneumonia2020}.
+By 7 Jan 2020, the cause of the pneumonia had been identified as a novel coronavirus which would go on to be named SARS-CoV-2~\autocite{whoNovelCov2020}.
+SARS-CoV-2 spread rapidly around the world: on 31 Jan 2020 the first cases were confirmed in England in two Chinese nationals~\autocite{bbcFirstUKCOVID}.
+However, introductions from Spain and France in Mar 2020 were the most important sources of the epidemic in England~\autocite{duplessisEstablishment}.
+
+The evolution of the epidemic in England was characterised by a succession of waves, with the context affected by three major types of changes: the imposition and relaxation of non-pharmaceutical interventions (NPIs), the emergence of new variants of the virus, and the roll-out of vaccination.
+There were many NPIs, but the most dramatic were the three national lockdowns imposed on 23 Mar 2020, 5 Nov 2020, and 4 Jan 2021 which legally required everyone to stay at home except for limited exceptions~\autocite{ifgTimeline}.
+Variants are genetic changes in the virus, the most consequential of which were the three variants of concern, named Alpha, Delta, and Omicron, that spread widely in England due to their growth advantage; prior to the emergence of Alpha, there were many variants in circulation, collectively known as ``pre-Alpha variants''~\autocite{lythgoeLineage}.
+Alpha became the dominant variant in Dec 2020, Delta in May 2021, and Omicron in Dec 2021~\autocite{lythgoeLineage}.
+The first vaccination in England was on 8 Dec 2020, with 68\% of adults having received two doses by Jul 2020 and 85\% by Oct 2021; initially, two doses was a full course, but later additional doses were offered, especially following the rise of the Omicron variant~\autocite{naoCovidVaccination}.
 

### 2024-08-01 cis-perfect-testing.tex
Christl comment on lack of independence within households

@@
 
+A particular violation of independence is the household design of the CIS.
+Individuals in the same household both have correlated infection times and correlated testing times, as they will typically have the same test schedule.
+If the test schedules are truly uninformative, as assumed here, this is unlikely to introduce bias; however, further work is required to confirm this hypothesis.
+
 I have developed a flexible simulation and inference framework for this type of analysis.

### 2024-08-01 conclusion.tex
Christl comment on study design

@@
 Furthermore, it reduces the impact of false negatives because they are likely to be followed by a true positive.
+
+Individuals were not tested exactly on the schedule specified in the CIS protocol.
+This was useful for the estimation of the distribution of the duration of detectability, because it meant that information was available at times other than exact multiples of seven.
+Building in some variations in testing schedules into the protocol (\eg randomising gaps between tests) may also be a promising approach to improve the estimation of the distribution of the duration of detectability.
 I have developed software that could be used to perform simulation studies of alternative designs.

### 2024-08-01 biology-data.tex
improve figure appearance
Christl comment on T+T data
Christl comment on representativeness

@@
 The CIS~\autocite{CIS} was set up in April 2020 to provide gold-standard measurements of the prevalence of SARS-CoV-2 in the community.
-It was a longitudinal study with a representative sample of private households, \ie excluding settings such as care homes and student halls.
+It was a longitudinal study aiming for a representative sample of private households, \ie excluding settings such as care homes and student halls.
 Initially, it was limited to England, but expanded to cover the whole of the UK in September 2020.
 Enrolment was continuous until 31 Jan 2022, with data collected until 13 Mar 2023~\autocite{weiRisk}. 
-The dataset is globally unique in providing a representative, longitudinal, and large-scale study across the pandemic.
+The dataset is globally unique in providing a near-representative, longitudinal, and large-scale study across the pandemic.
 
@@
 In reality, visits were often not on this precise schedule, and occasionally missed altogether.
+Additionally, participants' test results from routine, community-based testing, known as NHS Test and Trace, were linked; I do not make use of this information within this thesis because I am evaluating the potential of the designed portion of CIS alone.
 A full description of the study can be found in the study protocol~\autocite{cisProtocol}.
@@
 \begin{figure}
-  \centering \includegraphics{biology-data/CIS-recruitment}
+  \makebox[\textwidth][c]{\includegraphics{biology-data/CIS-recruitment}}
   \caption[CIS swab numbers]{%
@@
 % \begin{figure}
-%   \centering \includegraphics[width=\textwidth]{biology-data/CIS-num-tests}
+%   \makebox[\textwidth][c]{\includegraphics[width=\textwidth]{biology-data/CIS-num-tests}}
 %   \caption[Number of CIS tests]{%
@@
 \begin{figure}
-  \centering \includegraphics[width=\textwidth]{biology-data/CIS-positivity}
+  \makebox[\textwidth][c]{\includegraphics[width=\textwidth]{biology-data/CIS-positivity}}
   \caption[CIS positivity]{%

### 2024-08-01 introduction.tex
Christl comment on representativeness

@@
 
-The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys in representative samples of the general population.
-These were designed to avoid the biases of community testing and observe the general population.
+The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys which aimed to recruit representative samples of the general population; such surveys are considered the gold-standard.
+They invited individuals selected at random from the population; however, not all individuals selected responded to the invite introducing some non-representativeness (see \cref{E-biology-data:sec:MRP}).
+These designs aim to avoid the biases of community testing and observe the general population.
 These surveys use RT-PCR (Real-Time Polymerase Chain Reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.

### 2024-08-01 transmission.tex
improve figure appearance
Rob/Christl p123 comment
Christl comment on representativeness

@@
 In this chapter, I use the estimate of the distribution of the duration of RT-PCR positivity based on combining information from the ATACCC and CIS studies (see \cref{E-imperf-test:sec:conclusion}) to investigate the feasibility of estimating incidence and transmission of SARS-CoV-2 from only the CIS data (described in \cref{E-intro:sec:cis}).
-I will show that the CIS data are sufficient, demonstrating the value of surveys of this type for pandemic surveillance.
+I will show that this is feasible, and the estimates have narrow enough credible intervals, to be useful for informing the response to a pandemic.
 
@@
 \begin{figure}
-    \includegraphics{transmission/backcalc-regions}
+    \thisfloatpagestyle{empty}
+    \makebox[\textwidth][c]{\includegraphics{transmission/backcalc-regions}}
     \caption[Incidence estimated using the phenomenological approach by region]{%
@@
 \begin{figure}
-    \includegraphics{transmission/backcalc-alpha}
+    \makebox[\textwidth][c]{\includegraphics{transmission/backcalc-alpha}}
     \caption[Correlation of the emergence of Alpha and minimum incidence.]{%
@@
 \begin{figure}
-    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{transmission/backcalc-ages}}
+    \makebox[\textwidth][c]{\includegraphics{transmission/backcalc-ages}}
     \caption[Incidence estimated using the phenomenological approach by age group]{%
@@
 \begin{figure}
-    \centering \includegraphics{transmission/backcalc-start-effect}
+    \makebox[\textwidth][c]{\includegraphics{transmission/backcalc-start-effect}}
     \caption[Edge effects in the phenomenological approach]{%
@@
 The largest modification from the generic framework presented in \cref{E-SEIR} is the parameterization of the WAIFW matrix (\cref{SEIR:sec:WAIFW-application}).
-A novel approach (\cref{SEIR:sec:observation}) links this transmission model to the CIS prevalence data; the approach is generalisable to other studies collecting representative prevalence data, \eg REACT.
+A novel approach (\cref{SEIR:sec:observation}) links this transmission model to the CIS prevalence data; the approach is generalisable to other gold-standard prevalence studies, \eg REACT.
 In the mechanistic modelling literature, this part of the model is referred to as the observation model, however, in the terminology of \cref{E-inc-prev}, it includes both the prevalence process and the observation process.
@@
     \thisfloatpagestyle{empty}
-    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/sim/data}}
+    \makebox[\textwidth][c]{\includegraphics{SEIR/sim/data}}
     \caption[Simulated data]{%
@@
 \begin{figure}
-    \includegraphics{SEIR/sim/coverage}
+    \makebox[\textwidth][c]{\includegraphics{SEIR/sim/coverage}}
     \caption[Coverage of simulation study]{%
@@
 \begin{figure}
-    \makebox[\textwidth][c]{\includegraphics[width=0.9\paperwidth]{SEIR/sim/predictive_coverage}}
+    \thisfloatpagestyle{empty}
+    \makebox[\textwidth][c]{\includegraphics{SEIR/sim/predictive_coverage}}
     \caption[Coverage of simulation study (derived quantities)]{%
@@
 \begin{figure}
+    \thisfloatpagestyle{empty}
     \makebox[\textwidth][c]{\includegraphics{SEIR/sim/true_vs_posterior}}
@@
 \begin{figure}
-    \includegraphics{SEIR/CIS/incidence}
+    \thisfloatpagestyle{empty}
+    \makebox[\textwidth][c]{\includegraphics{SEIR/CIS/incidence}}
     \caption[Posterior estimates of incidence]{%
@@
 \begin{figure}
-    \includegraphics{SEIR/CIS/p_peak}
+    \makebox[\textwidth][c]{\includegraphics{SEIR/CIS/p_peak}}
     \caption[Posterior estimates of peak incidence timing]{%
@@
 \begin{figure}
-    \includegraphics{SEIR/CIS/beta_walk}
+    \thisfloatpagestyle{empty}
+    \makebox[\textwidth][c]{\includegraphics{SEIR/CIS/beta_walk}}
     \caption[Posterior estimates of $\beta_w$]{%
@@
 \begin{figure}
-    % \thisfloatpagestyle{empty}
+    \thisfloatpagestyle{empty}
     % \vspace{-3cm}
-    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_main}}
+    \makebox[\textwidth][c]{\includegraphics{SEIR/CIS/prev_main}}
     \captionsetup{width=0.8\paperwidth}
@@
 \begin{figure}
-    \centering \includegraphics{transmission/compare-regions}
+    \makebox[\textwidth][c]{\includegraphics{transmission/compare-regions}}
     \caption[Comparing each models estimate of regional incidence.]{%
@@
 \begin{figure}
-    \centering \includegraphics{transmission/compare-NE}
+    \makebox[\textwidth][c]{\includegraphics{transmission/compare-NE}}
     \caption[Comparing each models estimate of North East incidence proportion by age.]{%

### 2024-08-01 ATACCC.tex
Christl p73 minor comment
improve figure appearance
Christl p75 comment

@@
 \begin{figure}[tbhp]
-  \centering \includegraphics{ATACCC/typical_trajectory}
+  \makebox[\textwidth][c]{\includegraphics{ATACCC/typical_trajectory}}
   \caption[Example viral load model trajectory.]{%
@@
 $\N_+$ indicates a normal distribution truncated to the non-negative reals.
-Details of the distributions, including their parameterizations, are in \cref{E-distributions}.
+Details of all distributions used, including their parameterizations, are in \cref{E-distributions}.
 The elements of $\vec{\mu}_{\phi,g}$ and $\vec{\sigma}_\phi$ have independent priors placed on them, as specified in the first lines of the table.
@@
 
-The original \textcite{hakkiOnset} model has convergence issues (Rhat > 1.07 and ESS (Effective Sample Size) < 100) for the parameters of the test failure distribution, $\sigma_0$ and $\rho$.
+The original \textcite{hakkiOnset} model has convergence issues (Rhat > 1.07 and ESS (Effective Sample Size) < 100) for the parameters of the test failure distribution, $\sigma_0$ and $\rho$; typically Rhat < 1.05 and ESS of at least a few hundred are recommended to ensure that the inference procedure has converged to the true posterior distribution with adequate precision (see \cref{E-MCMC}).
 The simplified model I proposed in \cref{ATACCC:sec:observation-modification} solved this issue, meaning that this analysis converged.
@@
 \begin{figure}
-  \centering \includegraphics{ATACCC/compare_hakki_modified}
+  \makebox[\textwidth][c]{\includegraphics{ATACCC/compare_hakki_modified}}
   \caption[Comparison of population-level parameters between models.]{Kernel-density smoothed posterior densities for the population mean (top), $\vec{\mu}_{\phi,1}$, and covariance matrix (bottom), $\matr{\Sigma}_\phi$, for both the \textcite{hakkiOnset} model (described in \cref{ATACCC:sec:hakki}) and the model with a modified observation model to improve convergence (described in \cref{ATACCC:sec:observation-modification}).
@@
 \begin{figure}
-  \centering \includegraphics{ATACCC/mean_trajectories}
+  \makebox[\textwidth][c]{\includegraphics{ATACCC/mean_trajectories}}
   \caption[Median viral load trajectories]{Trajectory (posterior median and 95\% CrI) using the population median parameter values, $\exp \vec{\mu}_{\phi,g}$, excluding the within-population variation.
@@
 \begin{figure}
-  \centering \includegraphics{ATACCC/duration}
+  \makebox[\textwidth][c]{\includegraphics{ATACCC/duration}}
   \caption[Duration of RT-PCR positivity.]{Estimated distribution (survival function) of duration of RT-PCR positivity for the two populations.
@@
 \begin{figure}
-  \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{ATACCC/fits}}
+  \makebox[\textwidth][c]{\includegraphics{ATACCC/fits}}
   \caption[Posterior predictive viral load fits]{Posterior predictive observed viral loads (median and 95\% CrI, lines and ribbon) for selected individuals, overlaid with their observed test results (dots).
@@
 \begin{figure}
-  \centering \includegraphics{ATACCC/fit_individual_55}
+  \makebox[\textwidth][c]{\includegraphics{ATACCC/fit_individual_55}}
   \caption[Goodness-of-fit for individual 55]{Fit to individual 55's data coloured by MCMC chain.

### 2024-08-01 cis-perfect-testing.tex
Christl p.86 comment

@@
 
-Bayesian methods have been used for nest survival before~\autocite{heBayesiana,heBayesian,caoModeling} by augmenting the data with the unobserved times of infection.
+A related problem is estimating how long insect and bird nests survive for, as detected events are doubly interval censored and some are undetected.
+In the context of nest survival, Bayesian methods have been by augmenting the data with the unobserved times of infection~\autocite{heBayesiana,heBayesian,caoModeling}.
 The data augmentation approach creates a parameter for each detected infection.

### 2024-08-01 cis-perfect-testing.tex
Christl p.91 comment

@@
 &\propto (b+1-\pnodet)^{-(a+\ndet)} \\
-&= (r/\mu + 1 - \pnodet)^{-(r+\ndet)} &\text{sub in $\mu$ and $r$}\\
+&= (r/\mu + 1 - \pnodet)^{-(r+\ndet)} &\text{substitute in $\mu$ and $r$}\\
 &\propto(r + \mu (1- \pnodet))^{-(r+\ndet)}.
@@
 =& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \prob \left( l_k^{(e)} \leq E_{j} \leq r_k^{(e)} \mid B_{j} = b \right) \prob \left(B_{j} = b \right) \\
-=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \prob \left( l_k^{(e)} - b + 1 \leq D_{j} \leq r_k^{(e)} - b + 1 \right) \prob \left(B_{j} = b \right) &\text{by def of $D_{j}$} \\
-=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right) \prob \left(B_{j} = b \right) &\text{by def of $S_{\vec{\theta}}$} \\
+=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \prob \left( l_k^{(e)} - b + 1 \leq D_{j} \leq r_k^{(e)} - b + 1 \right) \prob \left(B_{j} = b \right) &\text{by definition of $D_{j}$} \\
+=& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right) \prob \left(B_{j} = b \right) &\text{by definition of $S_{\vec{\theta}}$} \\
 \propto& \sum_{b = l_k^{(b)}}^{r_k^{(b)}} \left( S_{\vec{\theta}}(l_k^{(e)} - b + 1) - S_{\vec{\theta}}(r_k^{(e)} - b + 2) \right)

### 2024-08-01 incidence-prevalence.tex
Christl p.91 comment

@@
   &\ll \frac{n_t n_{t'} \lvert \cov(P_t, P_{t'}) \rvert}{\Npop^2 \sqrt{\var(P_t) \var(P_{t'})}} &\text{as $\V(Y_t) \gg \V(P_t)$} \\
-  &= \frac{n_t n_{t'}}{\Npop^2} \lvert \cor(P_t, P_{t'}) \rvert &\text{by def of $\cor$}\\
+  &= \frac{n_t n_{t'}}{\Npop^2} \lvert \cor(P_t, P_{t'}) \rvert &\text{by definition of $\cor$}\\
   &\leq \frac{n_t n_{t'}}{\Npop^2} &\text{as $\lvert \cor(\cdot) \rvert \leq 1$}.

### 2024-08-01 cis-imperfect-testing.tex
improve figure appearance
Christl p.107 comment

@@
 \begin{figure}
-  \centering \includegraphics{cis-imperfect-testing/CIS_perfect}
+  \makebox[\textwidth][c]{\includegraphics{cis-imperfect-testing/CIS_perfect}}
   \caption[Estimating survival using CIS data assuming perfect testing]{Estimates of the survival function using CIS data alongside the ATACCC-based estimates of \cref{E-perf-test}.
@@
 \begin{figure}
-  \centering \includegraphics{cis-imperfect-testing/sim-single-positive-episodes}
+  \makebox[\textwidth][c]{\includegraphics{cis-imperfect-testing/sim-single-positive-episodes}}
   \caption[Single positive episodes in CIS simulation]{%
@@
     y-axis is the number of simulations, out of \numprint{1000}, with the indicated proportion.
-    Test sensitivity is $\psens$.
+    Test sensitivity is $\psens$, with the varied test sensitivity referring to the model proposed in \cref{imperf-test:eq:variable-test-sensitivity}.
     The CIS data are shown as a vertical line.
@@
 \begin{figure}
-  \centering \includegraphics{cis-imperfect-testing/CIS_final}
+  \makebox[\textwidth][c]{\includegraphics{cis-imperfect-testing/CIS_final}}
   \caption[Comparing estimates of the duration distribution.]{
@@
 \begin{figure}
-  \centering \includegraphics{cis-imperfect-testing/CIS_ntot}
+  \makebox[\textwidth][c]{\includegraphics{cis-imperfect-testing/CIS_ntot}}
   \caption[Sensitivity of $\ntot$'s posterior to its prior.]{How the posterior estimate of $\ntot$ changes with the value of $r$ in the prior on $\ntot$.}

### 2024-08-02 cis-imperfect-testing.tex
edits to corrections

@@
 Using a simple model, in particular a constant test sensitivity, means calculating the likelihood remains tractable.
-Additionally, I limit the number of permutations of false negative and true positive tests that I consider have non-negligible probability.
-Specifically, I assume that, if $i(j)$ is tested during infection episode $j$, either: the individual has a single false negative test and the episode ends before their next test, or the first test is a false positive test.
+Additionally, I limit the number of permutations of false negative and true positive tests that I consider to have non-negligible probability.
+Specifically, I assume that, if $i(j)$ is tested during infection episode $j$, either: the individual has a single false negative test and the episode ends before their next test, or the first test is a true positive test.
 This assumption is reasonable because false negatives normally occur when the viral load is low, which is normally late in the infection episode.
@@
 
-Introducing false negatives means that $O(j)$, containing the relevant information for episode $j$, is now random, even if $B_j$ and $E_j$ are known.
+Introducing false negatives means that $O(j)$, containing the test results which would bound the length of episode $j$ in the absence of false negatives, is a random variable even after conditioning on $B_j$ and $E_j$.
 Therefore, the results between the tests that bound episode $j$'s beginning and end times should now enter into the likelihood.

### 2024-08-02 cis-perfect-testing.tex
Christl p.97 comment
edits to corrections

@@
 
-A related problem is estimating how long insect and bird nests survive for, as detected events are doubly interval censored and some are undetected.
-In the context of nest survival, Bayesian methods have been by augmenting the data with the unobserved times of infection~\autocite{heBayesiana,heBayesian,caoModeling}.
+A related problem is estimating how long insect and bird nests survive for
+In this context detected events are doubly interval censored and some events are undetected; Bayesian methods have been developed which augment the data with the unobserved times of infection~\autocite{heBayesiana,heBayesian,caoModeling}.
 The data augmentation approach creates a parameter for each detected infection.
@@
 The set of times $i(j)$ tests positive at due to $W_j$ is $\posResults(W_j) = \{ t \in \sched_{i(j)} \ssep B_j \leq t \leq E_j \}$.
-The negative test immediately before the start of the episode (if there is one) is $\negResults^{(l)}(W_j) = \max \{ t \in \sched_{i(j)} \ssep t < B_j \}$.
-The negative test immediately after the end of the episode (if there is one) is $\negResults^{(r)}(W_j) = \min \{ t \in \sched_{i(j)} \ssep t > E_j \}$.
+The negative test immediately before the start of the episode (if there is one) is $\negResults^{(b)}(W_j) = \max \{ t \in \sched_{i(j)} \ssep t < B_j \}$.
+The negative test immediately after the end of the episode (if there is one) is $\negResults^{(e)}(W_j) = \min \{ t \in \sched_{i(j)} \ssep t > E_j \}$.
 
 Next, I define $O(j)$ to contain the test results due to $W_j$ that give information on $W_j$'s length, and hence are relevant to inference about $\vec{\theta}$.
-If $\posResults(W_j) = \emptyset$, $\min \posResults(W_j) \notin [1, T]$, or either $\negResults^{(l)}$ or $\negResults^{(r)}$ is undefined then the episode is undetected, in which case $O(j) = \emptyset$.
-Otherwise, $O(j) = [l_j^{(b)}, r_j^{(b)}, l_j^{(e)}, r_j^{(e)}, i(j)]^T$, where $i(j)$ is the individual in which the episode occurs; $l_j^{(b)} = \negResults^{(l)}(W_j) + 1$ and $r_j^{(b)} = \min \posResults(W_j)$ are the earliest and latest time the episode could have begun and produced the observed series of test results respectively; and $l_j^{(e)} = \max \posResults(W_j)$ and $r_j^{(e)} = \negResults^{(r)}(W_j) - 1$ are the earliest and latest time the episode could have ended respectively.
+If $\posResults(W_j) = \emptyset$, $\min \posResults(W_j) \notin [1, T]$, or either $\negResults^{(b)}$ or $\negResults^{(e)}$ is undefined then the episode is undetected, in which case $O(j) = \emptyset$.
+Otherwise, $O(j) = [l_j^{(b)}, r_j^{(b)}, l_j^{(e)}, r_j^{(e)}, i(j)]^T$, where $i(j)$ is the individual in which the episode occurs; $l_j^{(b)} = \negResults^{(b)}(W_j) + 1$ and $r_j^{(b)} = \min \posResults(W_j)$ are the earliest and latest time the episode could have begun and produced the observed series of test results respectively; and $l_j^{(e)} = \max \posResults(W_j)$ and $r_j^{(e)} = \negResults^{(e)}(W_j) - 1$ are the earliest and latest time the episode could have ended respectively.
 These definitions coincide with the way the dataset is constructed (see \cref{E-biology-data:sec:cis-episodes}).
@@
 
-As we assume $O(j) = \vec{\nu}_k$ and $O(j') = \vec{\nu}_{k'}$ are independent for $j \neq j'$:
+By the independence assumption:
 \begin{align}
@@
 
-Recall the definitions $p_k = \prob(O(j) = \vec{\nu}_k \mid \vec{\theta})$ where $O(j)$ is the random variable containing all relevant information for episode $j$ and $\vec{\nu}_k = [l^{(b)}_k, r^{(b)}_k, l^{(e)}_k, r^{(e)}_k, i_k]^T$ is an arbitrary element from $O(j)$'s state space, excluding undetected infections.
+Recall the definitions $p_k = \prob(O(j) = \vec{\nu}_k \mid \vec{\theta})$ where $O(j)$ contains the relevant test results for episode $j$ and $\vec{\nu}_k = [l^{(b)}_k, r^{(b)}_k, l^{(e)}_k, r^{(e)}_k, i_k]^T$ is an arbitrary element from $O(j)$'s state space, excluding undetected infections.
 
@@
 Commonly, $\alpha$ is chosen to be $0.5$ or $1$ (a uniform prior).
-The former, $\alpha = 0.5$, is Jefferys’ prior~\autocite{jeffreysInvariant}, commonly considered an uninformative prior because it is invariant to parameterization~\autocite[53]{gelmanBDA}.
+The former, $\alpha = 0.5$, is Jefferys' prior~\autocite{jeffreysInvariant}, commonly considered an uninformative prior because it is invariant to the parameterization used~\autocite[53]{gelmanBDA}.
 Even though these priors are uninformative on the hazard, they become highly informative on the survival. 
@@
 Therefore, $\vec{h}$ represents a sample from the approximation of \cref{E-ATACCC}'s posterior estimate of the hazard before increasing the uncertainty to account for the strong model assumptions.
-Using a multivariate normal, as opposed to multiple univariate distribution for each $h_t$, preserves the correlation between the hazards.
+Using a multivariate normal, as opposed to an independent univariate distribution for each $h_t$, preserves the correlation between the hazards.
 % The logit transformation maps from the $[0, 1]$ interval to the full real line.

### 2024-08-02 incidence-prevalence.tex
edits to corrections

@@
 The duration distribution I will be concerned with is the distribution of the duration of detectability.
-Consider an individual $i$ infected on day $B_i$ that is prevalent during the inclusive interval $[B_i, E_i]$ (as in \cref{E-biology-data:sec:cis-episodes}); if $B_i = E_i$ then $i$ is prevalent for exactly one day during which they are detectable.
+Consider an individual $i$ infected on day $B_i$ and is prevalent during the inclusive interval $[B_i, E_i]$ (as in \cref{E-biology-data:sec:cis-episodes}); if $B_i = E_i$ then $i$ is prevalent for exactly one day during which they are detectable.
 I define this period as an \emph{infection episode}, beginning on day $B_i$ and ending on day $E_i$; therefore, with a duration $D_i = E_i - B_i + 1$.

### 2024-08-02 introduction.tex
edits to corrections

--- 2024-08-01 introduction.tex
@@
 
-On 31 Dec 2019, China informed the WHO (World Health Organization) of cases of pneumonia of unknown cause in Wuhan City, Hubei province~\autocite{whoNovelPneumonia2020}.
+On 31 Dec 2019, China notified the WHO (World Health Organization) that they had detected several cases of pneumonia of unknown cause in Wuhan City, Hubei province~\autocite{whoNovelPneumonia2020}.
 By 7 Jan 2020, the cause of the pneumonia had been identified as a novel coronavirus which would go on to be named SARS-CoV-2~\autocite{whoNovelCov2020}.
@@
 
-The evolution of the epidemic in England was characterised by a succession of waves, with the context affected by three major types of changes: the imposition and relaxation of non-pharmaceutical interventions (NPIs), the emergence of new variants of the virus, and the roll-out of vaccination.
+The evolution of the epidemic in England was characterised by a succession of waves, with the context affected by three major types of changes: the imposition and relaxation of NPIs (non-pharmaceutical interventions), the emergence of new variants of the virus, and the roll-out of vaccination.
+
 There were many NPIs, but the most dramatic were the three national lockdowns imposed on 23 Mar 2020, 5 Nov 2020, and 4 Jan 2021 which legally required everyone to stay at home except for limited exceptions~\autocite{ifgTimeline}.
-Variants are genetic changes in the virus, the most consequential of which were the three variants of concern, named Alpha, Delta, and Omicron, that spread widely in England due to their growth advantage; prior to the emergence of Alpha, there were many variants in circulation, collectively known as ``pre-Alpha variants''~\autocite{lythgoeLineage}.
+
+Variants are genetic changes in the virus, with those causing significant challenges known as \emph{variants of concern}~\autocite{whoVOC} and labelled using a Greek letter.
+Three variants of concern, Alpha, Delta, and Omicron, successively dominated all other variants in England; prior to the emergence of Alpha, there were many variants in circulation, collectively known as ``pre-Alpha variants''~\autocite{lythgoeLineage}.
 Alpha became the dominant variant in Dec 2020, Delta in May 2021, and Omicron in Dec 2021~\autocite{lythgoeLineage}.
+
 The first vaccination in England was on 8 Dec 2020, with 68\% of adults having received two doses by Jul 2020 and 85\% by Oct 2021; initially, two doses was a full course, but later additional doses were offered, especially following the rise of the Omicron variant~\autocite{naoCovidVaccination}.
@@
 
-The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys in representative samples of the general population.
-These were designed to avoid the biases of community testing and observe the general population.
+The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys which aimed to recruit representative samples of the general population; such surveys are considered the gold-standard for conducting population surveillance~\autocite{mccabeCISincidence}.
+To achieve this aim, they invited individuals selected at random from the population; however, not all individuals selected responded to the invite introducing some non-representativeness (see \cref{E-biology-data:sec:MRP}).
+These designs aim to avoid the biases of community testing and observe the general population.
 These surveys use RT-PCR (Real-Time Polymerase Chain Reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.

### 2024-08-02 ATACCC.tex
Christl p78 comment
correct notation

@@
 From \cref{ATACCC:eq:di}, we can see that $d_i$ depends only on $i$'s random effects vector $\vec{\phi}_i$.
-Therefore, the distribution of $d_i$ across the population of group $g$ is determined by $\vec{\phi}_i$'s mean, $\vec{\mu}_{\phi,g}$, and covariance matrix, $\matr{\Sigma}_\phi$.
+Therefore, the distribution of $d_i$ across the population of group $g(i)$ is determined by $\vec{\phi}_i$'s mean, $\vec{\mu}_{\phi,g(i)}$, and covariance matrix, $\matr{\Sigma}_\phi$.
 % Hence, the uncertainty in the posterior distribution of the discretised duration cumulative distribution function for group $g$, $F_{D,g}(d) = \prob(D_i \leq d + 0.5)$ originate from the uncertainty in $\vec{\mu}_\phi$ and $\matr{\Sigma}_\phi$.
@@
 
-A strength of the ATACCC data are its general population sample.
+A strength of the ATACCC data is its general population sample.
 This is unlike most estimates (see \cref{E-intro:sec:previous-duration-estimates}) which are from hospitalised patients or healthcare workers.

### 2024-08-02 cis-perfect-testing.tex
clarify nest stuff

@@
 
-A related problem is estimating how long insect and bird nests survive for
-In this context detected events are doubly interval censored and some events are undetected; Bayesian methods have been developed which augment the data with the unobserved times of infection~\autocite{heBayesiana,heBayesian,caoModeling}.
+Bayesian methods have been used for nest survival before~\autocite{heBayesiana,heBayesian,caoModeling} by augmenting the data with the unobserved times.
 The data augmentation approach creates a parameter for each detected infection.

### 2024-08-02 incidence-prevalence.tex
Rob p.65 comment

@@
     (\matr{S})_{i,j} &= \begin{cases}
-        S(i - j + 1) & i - \dmax - 1 \leq j \leq i \\
+        S(i - j + 1) & \text{if } i - \dmax + 1 \leq j \leq i \\
         0 & \text{otherwise.}

### 2024-08-02 cis-imperfect-testing.tex
Updates from Overleaf

--- 2024-07-31 cis-imperfect-testing.tex
@@
 In this chapter, I identify this issue to be the result of false negatives, a form of misclassification bias, within the CIS (\cref{imperf-test:sec:problem}).
-I then propose several models of false negatives (\cref{imperf-test:sec:simulate}) and extend the model of \cref{E-perf-test} to incorporate one of them (\cref{imperf-test:sec:modelling}).
+I then propose several models of false negatives (\cref{imperf-test:sec:simulate}) and extend the model of \cref{E-perf-test} to incorporate one of them (\cref{imperf-test:sec:modelling})
 After showing through simulation that the new model recovers the true duration distribution (\cref{imperf-test:sec:sim-study-results}), I apply it to the CIS data (\cref{imperf-test:sec:application}).
@@
 
-The CIS data provide evidence for the test sensitivity declining over the course of an episode.
+The CIS data provides evidence for the test sensitivity declining over the course of an episode.
 \Cref{imperf-test:fig:bounding-cis-sensitivity} demonstrates this by using the following classification for test results.
@@
 
-In this section, I introduce a simple model of false negatives (\ie allowing $\psens < 1$) into the posterior density derived in \cref{E-perf-test:sec:model}; I continue to assume the probability of a false positive is negligible.
+In this section, I introduce a simple model of false negatives (\ie allowing $\psens < 1$) into the posterior density derived in \cref{E-perf-test:sec:model}.
 Using a simple model, in particular a constant test sensitivity, means calculating the likelihood remains tractable.
-Additionally, I limit the number of permutations of false negative and true positive tests that I consider have non-negligible probability.
-Specifically, I assume that, if $i(j)$ is tested during infection episode $j$, either: the individual has a single false negative test and the episode ends before their next test, or the first test is a false positive test.
-This assumption is reasonable because false negatives normally occur when the viral load is low, which is normally late in the infection episode.
-For the same reason, I assume there is at most one false negative test following the final true positive test in the episode.
 
@@
 
-Introducing false negatives means that $O(j)$, containing the relevant information for episode $j$, is now random, even if $B_j$ and $E_j$ are known.
+Introducing false negatives means that $O(j)$, the observation vector of episode $j$, is now random, even if $B_j$ and $E_j$ are known.
 Therefore, the results between the tests that bound episode $j$'s beginning and end times should now enter into the likelihood.
@@
 
-To do so, define a random vector $O'(j)$ with state space $\{\emptyset\} \cup \set{E}'$ which will replace $O(j)$ from \cref{E-perf-test}.
-As before, $O'(j) = \emptyset$ if episode $j$ is undetected.
-Otherwise, $O'(j) \in \set{E}'$ where $\set{E}' = \{ \vec{\nu}_1', \dots, \vec{\nu}_{N_E'}' \}$ augments the space $\set{E}$ with test results during the episode.
-The elements of $\set{E}'$ are all $\vec{\nu}_k' = [\vec{\nu}_k, \vec{y}_k]^T$ where $\vec{\nu}_k \in \set{E}$ and $\vec{y}_k$ is a vector of test results for the relevant testing times, $\sched'_k = \{ t \in \sched_{i_k} \ssep r_k^{(b)} \leq t \leq r_k^{(e)} + 1 \}$.
-To formalise the definition of $\vec{y}_k$, let $m_k$ denote the size of $\sched'_k$ and denote the elements of $\sched'_k$ by $t_{k,1} < \dots < t_{k,m_k}$.
-Then, $\vec{y}_k \in \{0, 1\}^{m_k}$ and satisfies the following two conditions.
+To formalise this, I define analogous variables to those in \cref{E-perf-test:sec:posterior}.
+To start with, define the relevant testing times for any episode $j$ with $O(j) = \vec{\nu}_k$ as $\sched'_k = \{ t \in \sched_{i_k} \ssep r_k^{(b)} \leq t \leq r_k^{(e)} + 1 \}$ and let $m_k$ denote the size of this set.
+Denote the elements of $\sched'_k$ by $t_{k,1} < \dots < t_{k,m_k}$.
+Let $\set{E}' = \{ \vec{\nu}_1', \dots, \vec{\nu}_{N_E'}' \}$ be a new set of possible observations of detected episodes, including the pattern of positive and negative tests.
+Represent the observations such that for any $\vec{\nu}_k' \in \set{E}'$, $\vec{\nu}_k' = [\vec{\nu}_{k}, \vec{y}_k]^T$, where $\vec{y}_k \in \{ 0, 1 \}^{m_k}$ with $y_{k,l}$ being the test result at time $t_{k,l}$.
+
+$\vec{\nu}'_k \in \set{E}'$ occurs if and only if all the following conditions are met.
 \begin{enumerate}
+  \item $\vec{\nu}_k \in \set{E}$.
   \item The elements of $\vec{y}_k$ corresponding to the tests at times $r_k^{(b)}$ and $l_k^{(e)}$ are positive, \ie $y_{k,1} = y_{k,m_k-1} = 1$.
@@
 \end{enumerate}
-These conditions are due to the construction of the intervals as positive and negative tests bounding the beginning and end times of the episode.
+The final two conditions are due to the construction of the intervals as positive and negative tests bounding the beginning and end times of the episode.
 
-Let $p_k'$, $p_{ik}'$, $p_u'$, and $p_{iu}'$ have the same definition as $p_k$, $p_{ik}$, $p_u$, and $p_{iu}$ respectively except replacing $O$ with $O'$ and $\vec{\nu}_k$ with $\vec{\nu}_k'$, and conditioning on $\psens$.
+Let $O'(j)$ be the observations generated by episode $j$ including the new vector of results.
+As before, $O'(j) \in \set{E}'$ indicates that episode $j$ was detected; otherwise, $O'(j) = \emptyset$.
+
+Similarly, let $p_k'$, $p_{ik}'$, $p_u'$, and $p_{iu}'$ have the same definition as p_k'$, $p_{ik}'$, $p_u'$, and $p_{iu}'$ respectively except replacing $O$ with $O'$ and $\vec{\nu}_k$ with $\vec{\nu}_k'$, and conditioning on $\psens$.
 As before, $p'_k = 1/\Ncis p'_{ik}$ and $p'_u = 1/\Ncis \sum_{i=1}^{\Ncis} p'_{iu}$.
@@
 
-By assumption, the negative test bounding the start of the episode, on day $l_k^{(b)}-1$, is a true negative.
+For tractability, assume that the negative test bounding the start of the episode, on day $l_k^{(b)}-1$, is a true negative.
+This assumption is reasonable because, since a positive test follows at $r_k^{(b)}$, the negative at $l_k^{(b)}-1$ is likely early in the infection and the test sensitivity is high early in an infection.
 True negatives occur with probability 1, and hence this test does not contribute to the likelihood.
@@
 For there to be no subsequent positive tests, all tests up until day $e_j$ are false negatives.
-By assumption, there is a negligible probability of two false negatives.
+I assume there is a negligible probability of two false negatives.
 Therefore, this can only occur if the episode ends before another test.

### 2024-08-02 biology-data.tex
corrections I spotted on paper

@@
 Other data such as demography and symptoms were also collected.
-For further details see \textcite{singanayagamDuration,hakkiOnset}.
+For further details see \textcite{singanayagamCommunity,hakkiOnset}.
 
@@
 Define a (sub) population of interest as a set of strata $\set{S}$.
-Then the poststratified prevalence in the (sub)population on day $t$ is
+Then the $j$th posterior sample of poststratified prevalence in the (sub)population on day $t$ is
 \begin{align}

### 2024-08-02 incidence-prevalence.tex
corrections I spotted on paper

@@
 The SIR model assumes that individuals are immediately infectious upon infection, neglecting the latent period (see \cref{E-biology-data:sec:natural-history}).
-Ignoring the latent period will again overestimate the reproduction number~\autocite{wearingAppropriate}.
+Ignoring the latent period will again underestimate the reproduction number~\autocite{wearingAppropriate}.
 The SIR model can be extended by adding a latent compartment~\autocite[41]{keelingModeling}, also known as an \emph{exposed} compartment.
@@
 Now, denote by $P_{t_z,t_p}$ the number of individuals infected at $t_z$ and prevalent at $t_p$.
-Formally, $P_{t_z,t_p}$ is the infection episodes with $B_i = t_z$ and $E_i \geq t_p$.
+Formally, $P_{t_z,t_p}$ is the number of infection episodes with $B_i = t_z$ and $E_i \geq t_p$.
 $P_{t_z,t_p}$ is $Z_{t_z}$ minus those that recover before $t_p$ (\ie at or up to $t_p - 1$).

### 2024-08-02 introduction.tex
corrections I spotted on paper

@@
 \end{align}
-where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that result in the consequence of interest; and, conditional on being observed, $f(i)$ describes the (discretised) probability that an infection is included in the count $W_t$ at $i$ time units after the infection.
+where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that result in the consequence of interest; and, conditional on suffering the consequence of interest, $f(i)$ describes the (discretised) probability that an infection is included in the count $W_t$ at $i$ time units after the infection.
 $f$ is an example of a \emph{delay function}, and $f$ is commonly the pdf (probability density function) of a distribution, or a transformation of it.
@@
 These designs aim to avoid the biases of community testing and observe the general population.
-These surveys use RT-PCR (Real-Time Polymerase Chain Reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
+These surveys use RT-PCR (Reverse Transcription Polymerase Chain Reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
 The first survey was the CIS (Coronavirus (COVID-19) Infection Survey), run by the ONS (Office for National Statistics)~\autocite{CIS,cisMethodsONS}.

### 2024-08-02 ATACCC/duration.R

@@ -34,7 +34,11 @@ p_durations = bind_rows(
         .width = 0.95
     ) +
     theme_survival_time_series () +
-    coord_cartesian(xlim = c(0, 40))
+    coord_cartesian(xlim = c(0, 40)) +
+    labs(
+        colour = "Group",
+        fill = "Group"
+    )
 save_plot(
     filename = here::here("ATACCC/duration.pdf"),
     plot = p_durations,

### 2024-08-17 cis-imperfect-testing.tex
proof read corrections

--- 2024-08-02 cis-imperfect-testing.tex
@@
 
-Introducing false negatives means that $O(j)$, containing the test results which would bound the length of episode $j$ in the absence of false negatives, is a random variable even after conditioning on $B_j$ and $E_j$.
-Therefore, the results between the tests that bound episode $j$'s beginning and end times should now enter into the likelihood.
+Recall that, for an undetected infection episode, $O(j) = \emptyset$ and, for a detected episode, $O(j) = [l_j^{(b)}, r_j^{(b)}, l_j^{(e)}, r_j^{(e)}, i(j)]^T$ where $r_j^{(b)}$ is the day of the first positive due to infection episode $j$; $l_j^{(b)}$ is the day after the negative preceding this positive; $l_j^{(e)}$ is the day of the final positive due to infection episode $j$; $r_j^{(e)}$ is the day before the negative test following this positive; and $i(j)$ is the individual episode $j$ occurred in.
+As $O(j)$ contains test results now subject to false negatives, it is now a random variable even after conditioning on $B_j$ and $E_j$ (previously it was deterministic).
+Additionally, the results between the tests that bound episode $j$'s beginning and end times are also random and hence should enter into the likelihood.
 For tractability, I consider only tests between the negative tests providing an upper bound on the length of episode $j$.

### 2024-08-17 cis-perfect-testing.tex
proof read corrections

@@
 The episode's duration is a deterministic transformation of these variables, $D_j = E_j - B_j + 1$; and I assume that $B_j$, $D_j$, and $i(j)$ are independent.
-$D_j$'s distribution is determined by $\vec{\theta}$; both $B_j$ and $i(j)$ have uniform distributions over their respective state spaces.
+$D_j$'s distribution is determined by $\vec{\theta}$; both $B_j$ and $i(j)$ are modelled as draws from uniform distributions over their respective discrete state spaces.
 
@@
 A particular violation of independence is the household design of the CIS.
-Individuals in the same household both have correlated infection times and correlated testing times, as they will typically have the same test schedule.
+Individuals in the same household have both correlated infection times and correlated testing times, as they will typically have the same test schedule.
 If the test schedules are truly uninformative, as assumed here, this is unlikely to introduce bias; however, further work is required to confirm this hypothesis.

### 2024-08-17 conclusion.tex
Christl comment on representativeness
proof read corrections

@@
 
-Overall, I have found that representative prevalence surveys are a promising approach for pandemic surveillance.
-A well-designed survey is sufficient to estimate incidence and transmission, although not currently the duration of detectability.
+Overall, I have found that gold-standard prevalence surveys are a promising approach for pandemic surveillance.
+A well-designed survey alone can be used to estimate incidence and transmission, although not currently the duration of detectability.
 

### 2024-08-17 introduction.tex
proof read corrections

@@
 By 7 Jan 2020, the cause of the pneumonia had been identified as a novel coronavirus which would go on to be named SARS-CoV-2~\autocite{whoNovelCov2020}.
-SARS-CoV-2 spread rapidly around the world: on 31 Jan 2020 the first cases were confirmed in England in two Chinese nationals~\autocite{bbcFirstUKCOVID}.
+SARS-CoV-2 spread rapidly around the world: on 31 Jan 2020 the UK government announced that the first two cases had been confirmed in England~\autocite{bbcFirstUKCOVID}.
 However, introductions from Spain and France in Mar 2020 were the most important sources of the epidemic in England~\autocite{duplessisEstablishment}.
@@
 
-The response to the SARS-CoV-2 pandemic in the UK has been unprecedented, both in terms of the scientific input to policy and the surveillance data made available.
+The response to the SARS-CoV-2 pandemic in the UK was unprecedented, both in terms of the scientific input to policy and the surveillance data made available.
 The most prominent surveillance outputs were produced by members of SPI-M-O (the Scientific Pandemic Influenza Group on Modelling Operational sub-group), a sub-committee of SAGE (the Scientific Advisory Group for Emergencies)~\autocite{medleySPIM,govSPIMO}.
@@
 The extraordinary impact of SARS-CoV-2 led the UK Government to fund two large prevalence surveys which aimed to recruit representative samples of the general population; such surveys are considered the gold-standard for conducting population surveillance~\autocite{mccabeCISincidence}.
-To achieve this aim, they invited individuals selected at random from the population; however, not all individuals selected responded to the invite introducing some non-representativeness (see \cref{E-biology-data:sec:MRP}).
-These designs aim to avoid the biases of community testing and observe the general population.
+To achieve this aim, they invited participants who were randomly selected from the population; however, not all individuals selected responded to the invite and consented to participate; this introduced some non-representativeness (see \cref{E-biology-data:sec:MRP}).
+These designs aim to avoid the biases of community testing.
 These surveys use RT-PCR (Reverse Transcription Polymerase Chain Reaction, described in detail in \cref{E-biology-data:sec:PCR}) testing to detect the presence of the virus in a swab from an individual.
